### 2022-03-19 latex/HP1-TranslComm.tex
initial release

\input{Preamble-Print}
\makepagestyle{HPed}
\makeoddhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
\makeevenhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}

\newcommand{\myfn}[1]{\footnote{\texteng{#1}}}
\setlength{\footmarkwidth}{1.3em}
\setlength{\footmarksep}{0em}
\setlength{\footparindent}{1.3em}
\footmarkstyle{\textsuperscript{#1} }
\usepackage{fnpos}
\makeFNbottom

% additions/changes 2022-03-12 mm:
\SetTEIxmlExport{autopar=false}
\NewDocumentEnvironment{translation}{O{}}{}{}
\NewDocumentEnvironment{philcomm}{O{}}{}{}

\NewDocumentEnvironment{tlg}{O{}}{\begin{ekdverse}\indentpattern{0000}}{\end{ekdverse}}
\RenewDocumentEnvironment{verse}{}{\begin{ekdverse}\indentpattern{0000}\begin{patverse*}}{\end{patverse*}\end{ekdverse}}
\NewDocumentCommand{\tl}{m}{#1}

\begin{document}
\pagestyle{HPed} %
\begin{ekdosis}
  \SetLineation{lineation = none,}

  \chapter{Translation and Philological Commentary}

  \subsection*{1.1}
  \begin{translation}[hp01_001]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_001]
    \begin{description}
      \item[Sources]{\emph{Tantrālokaviveka} 5.126, \emph{Śivasūtravimarśinī} 1.6, \emph{Spandanirṇaya} 1.11, \emph{Pratyabhijñāhṛdaya}}
      \item[Testimonia]{\emph{Amanaska} 2.10, \emph{Candrāvalokana} 1}
    \end{description}

    \noindent
    The verse is variously transmitted, with differing case endings and pronouns, but remains
    consistent substance. It occurs in a group of texts from Kashmirian non-dual Śaivism, the earliest
    being the quotations by Kṣemarāja, who quotes this verse frequently. A name of the sources text in
    never given, but the impression is that it is from an esoteric Tantric source, since Kṣemarāja
    quotes it in his \emph{Spandanirṇaya} on 1.11 as from a \emph{mahārahasyaśāstra}, that is, from a
    ``secret'', meaning heterodox Tantric scriptural text.

    The original Tantric context of the verse rules out Brāhmananda's reading \emph{vedaśāstreṣu}
    instead of \emph{sarvatantreṣu} in Pāda d. Since Brahmānanda's version is hardly ever represented
    in the manuscripts—only C$_{7}$ reads \emph{sarvaśāstreṣu}—, one wonders whether this reading was
    in fact one of his own innovations\footnote{There are other cases, where Brahmānanda frames
      doctrines in a Vedāntic context.}  in order to bring the text more in line with—from his
    perspective—Vedāntic main stream religion.

    It is also worth noting that the claim that Yogic technique is contained in the Veda is also a
    position held by our commentator, who says in his commentary: \emph{vedaśāstreṣu gopitā vedeṣu
      ṛgādiṣu śāstreṣu sāṅkhyapātañjalādiṣu gopitā rakṣitā}.

    The verse has been elaborately rewritten in an interesting small work that goes under the name
    \emph{Anubhavanivedana} and is attributed by the editor K.C. Pandey to
    Abhinavagupta,\footnote{\emph{Abhinavagupta. An Historical and Philosophical Study" K.C. Pandey
        Chaukhamba Amarabharati Prakashan}. Varanasi 1935, p.\,953.} but the mainly Yogic content does
    not suggest this authorship. In verse 2 the supreme reality is called \emph{aśeṣabāhyarahitaṃ paraṃ
      pumāṃsam}, which collides with non-dual Pratyabhijñā Śaivism, and 4c \emph{yogaḥ sa prathate
      yataḥ pravahaṇaṃ prāṇasya saṃkṣīyate} suggests more likely a Haṭhayogic context. The attribution
    of this text is probably based on a short note, as we find them in Kashmirian codices that transmit
    smaller texts as writers' notes in the leaves added between the main texts.\footnote{See Jürgen
      Hanneder: \emph{Sāhib Kaul's Tree of Languages}. Marburg 2021, p.??, for an example for this.}
    In Vaktraśambhu's \emph{Yogasaṃgraha} the first verse is attributed to an otherwise unknown
    \emph{Yogaśāṃkara}.

    % collate with Wellcome Ms.
    \begin{ekdverse}
      \emph{antarlakṣyavilīnacittapavano yogī yadā vartate}\\
      \emph{dṛṣṭyā niścalatārayā bahir asau paśyann apaśyann api}\\
      \emph{mudreyaṃ khalu śāmbhavī bhavati sā yuṣmatprasādād guro}\\
      \emph{śūnyāśūnyavivarjitaṃ bhavati yat tattvaṃ padaṃ śāmbhavam} (1)\\
    \end{ekdverse}

    \begin{ekdverse}
      \emph{ardhodghāṭitalocanaḥ sthiramanā nāsāgradattekṣaṇaś}\\
      \emph{candrārkāv api līnatām upagatau trispandabhāvāntare}\\
      \emph{jyotīrūpam aśeṣabāhyarahitaṃ caikaṃ pumāṃsaṃ param}\\
      \emph{tattvaṃ tatpadam eti vastu paramaṃ vācyaṃ kim atrādhikam} (2)\\
    \end{ekdverse}

    \begin{ekdverse}
      \emph{śabdaḥ kaś cana yo mukhād udayato mantraḥ sa lokottaraḥ}\\
      \emph{saṃsthānaṃ sukhaduḥkhajanmavapuṣo yat kāpi mudraiva sā}\\
      \emph{prāṇasya svarasena yat pravahaṇaṃ yogaḥ sa evādbhutaḥ}\\
      \emph{śāktam dhāma paraṃ mamānubhavataḥ kin nāma na bhrājate} (3)\\
    \end{ekdverse}

    \begin{ekdverse}
      \emph{mantraḥ sa pratibhāti varṇaracanā yasmin na saṃlakṣayate}\\
      \emph{mudrā sā samudeti yatra galitā kṛtsnā kriyā kāyikī}\\
      \emph{yogaḥ sa prathate yataḥ pravahaṇaṃ prāṇasya saṃkṣīyate}\\
      \emph{tvaddhāmādhigamotsaveṣu sudhiyāṃ kiṃ kiṃ na nāmādbhutam} (4)\\
    \end{ekdverse}

    \noindent
    The transformation of this Tantric quote does not end here: It is read as Śāṇḍilya Upaniṣat ??, but
    with 1c as referring to the \emph{khecarī} mudrā rather than the \emph{śāmbhavī}: \emph{mudreyaṃ
      khalu khecarī bhavati sā lakṣyaikatānā śivā}, and both the Śāṇḍilya and the Yogakundali Upaniṣat
    furnish it with a Vaiṣṇava context: \emph{śūnyāśūnyavivarjitaṃ sphurati sā tattvaṃ padaṃ vaiṣṇavī}.

    \begin{itemize}
      \item{Vaktraśambhu: \emph{Yogasārasaṃgraha}. Available as a transcript in Pondichery: IFP T0859 copied from D 4373 belonging to the GOML, Madras. The verse is introduced there as \emph{yogaśāṃkare}.}
      \item{śāṇḍilya upaniṣat śrī upaniṣad brahmayogin Adyar library and research centre. Madras 1920}
      \item{yogakuṇḍalyupaniṣat}
      \item{Narahari: Bodhasara}
    \end{itemize}
  \end{philcomm}

  \subsection*{1.2}
  \begin{translation}[hp01_002]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_002]
    dummy commentary.
  \end{philcomm}

  \subsection*{1.3}
  \begin{translation}[hp01_003]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_003]          
    dummy commentary.
  \end{philcomm}

  \subsection*{1.4}
  \begin{translation}[hp01_004]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_004]          
    dummy commentary.
  \end{philcomm}

  \subsection*{1.5}
  \begin{translation}[hp01_005]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_005]          
    dummy commentary.
  \end{philcomm}

  \subsection*{1.6}
  \begin{translation}[hp01_006]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_006]          
    dummy commentary.
  \end{philcomm}

  \subsection*{1.7}
  \begin{translation}[hp01_007]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_007]          
    dummy commentary.
  \end{philcomm}

  \subsection*{1.8}
  \begin{translation}[hp01_008]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_008]          
    dummy commentary.
  \end{philcomm}

  \subsection*{1.9}
  \begin{translation}[hp01_009]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_009]          
    dummy commentary.
  \end{philcomm}

  \subsection*{1.10}
  \begin{translation}[hp01_010]
    dummy translation.
  \end{translation}

  \begin{philcomm}[hp01_010]          
    dummy commentary.
  \end{philcomm}

\end{ekdosis}
\end{document}

### 2022-03-19 latex/Jyotsna.tex
initial release

% additions/changes 2022-03-12 mm:
\input{Preamble-Print}
\makepagestyle{HPed}
\makeoddhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
\makeevenhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}

\setlength{\footmarkwidth}{1.3em}
\setlength{\footmarksep}{0em}
\setlength{\footparindent}{1.3em}
\footmarkstyle{\textsuperscript{#1} }
\usepackage{fnpos}
\makeFNbottom

\SetTEIxmlExport{autopar=false}
\NewDocumentEnvironment{jyotsna}{O{}}{}{}

\begin{document}
\pagestyle{HPed} %
\begin{ekdosis}
  \SetLineation{lineation = none,}

  \begin{jyotsna}[hp01_001]
    śrīādināthāya namo ’stu tasmai \\
    yenopadiṣṭā haṭhayogavidyā  / \\
    vibhrājate pronnatarājayogam \\
    āroḍhum icchor adhirohiṇīva (1.1)

    guruṃ natvā śivaṃ sākṣādbrahmānandena tanyate /
    haṭhapradīpikājyotsnā yogamārgaprakāśikā //

    idānīṃtanānāṃ subodhārtham asyāḥ     (variant: idānīṃ janānāṃ Lonavla Ed) 	
    suvijñāya gorakṣasiddhāntahārdam /
    mayā meruśāstripramukhyābhiyogāt
    sphuṭaṃ kathyate 'tyantagūḍho 'pi bhāvaḥ //

    mumukṣujanahitārthaṃ rājayogadvārā kaivalyaphalāṃ haṭhapradīpikāṃ vidhitsuḥ paramakāruṇikaḥ
    svātmārāmayogīndras tatpratyūhanivṛttaye haṭhayogapravartakaśrīmadādināthanamaskāralakṣaṇaṃ
    maṅgalaṃ tāvad ācarati—śrīādināthāyetyādinā /

    tasmai śrīādināthāya namo 'stv ity anvayaḥ / ādiś cāsau nāthaś ca ādināthaḥ sarveśvaraḥ śiva ity
    arthaḥ / śrīmān ādināthaḥ śrīādināthaḥ tasmai śrīādināthāya / śrīśabda ādir yasya saḥ śrīādiḥ,
    śrīādiś cāsau nāthaś ca śrīādināthaḥ, tasmai śrīādināthāya / śrīnāthāya viṣṇava iti vārthaḥ / śrī
    ādināthāyety atra yaṇabhāvas tu api māṣaṃ maṣaṃ kuryāc chandobhaṅgaṃ tyajed girām iti chandovidāṃ
    sampradāyād uccāraṇasauṣṭhavāc ceti bodhyam / namaḥ prahvībhāvo 'stu / prārthanāyāṃ loṭ / tasmai /
    kasmai ity apekṣāyāṃ āha yeneti / yena ādināthena upadiṣṭā girijāyai haṭhayogavidyā / haś ca ṭhaś
    ca haṭhau sūryacandrau tayor yogo haṭhayogaḥ / etena haṭhaśabdavācyayoḥ sūryacandrākhyayoḥ
    prāṇāpānayor aikyalakṣaṇaḥ prāṇāyāmo haṭhayoga iti haṭhayogasya lakṣaṇaṃ siddham / yathā coktaṃ
    gorakṣanāthena siddhasiddhāntapaddhatau—

    hakāraḥ kīrtitaḥ sūryaṣ ṭhakāraś candra ucyate /
    sūryacandramasor yogād dhaṭhayogo nigadyate // [si.si.pa. 1.69] iti /

    tatpradīpikā vidyā haṭhayogavidyā haṭhayogaśāstram iti yāvat / girijāyai ādināthakṛto
    haṭhavidyopadeśo mahākālayogaśāstrādau prasiddhaḥ /

    prakarṣeṇa unnataḥ pronnataḥ / mantrayogahaṭhayogādīnām adharabhūmīnām uttarabhūmitvād rājayogasya
    pronnatatvam / rājayogaś ca sarvavṛttinirodhalakṣaṇo 'samprajñātayogaḥ / tam icchor mumukṣor
    adhirohiṇīva adhiruhyate 'nayety adhirohiṇī niśreṇīva vibhrājate viśeṣeṇa bhrājate śobhate / yathā
    pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudhaprāpikā bhavati evaṃ haṭhadīpikāpi
    pronnatarājayogam āroḍhum icchor anāyāsena rājayogaprāpikā bhavatīti / upamālaṅkāraḥ /
    indravajrākhyaṃ vṛttam //1// 
  \end{jyotsna}

  \begin{jyotsna}[hp01_002]
    praṇamya śrīguruṃ nāthaṃ svātmārāmeṇa yoginā /
    kevalaṃ rājayogāya haṭhavidyopadiśyate //2//

    evaṃ paramagurunamaskāralakṣaṇaṃ maṅgalaṃ kṛtvā vighnabāhulye maṅgalabāhulyasyāpy apekṣitatvāt
    svagurunamaskārātmakaṃ maṅgalam ācarann asya granthasya viṣayaprayojanādīn
    pradarśayati—praṇamyeti / śrīmantaṃ guruṃ śrīguruṃ nāthaṃ śrīgurunāthaṃ, svagurum iti yāvat /
    praṇamya prakarṣeṇa bhaktipūrvakaṃ natvā svātmārāmeṇa yoginā yogo'syāstīti yena / kevalaṃ
    rājayogāya kevalaṃ rājayogārthaṃ haṭhavidyopadiśyata ity anvayaḥ / haṭhavidyāyā rājayoga eva
    mukhyaṃ phalaṃ na siddhaya iti kevalapadasyābhiprāyaḥ / siddhayas tv ānuṣaṅgikyaḥ / etena
    rājayogaphalasahito haṭhayogo'sya granthasya viṣayaḥ / rājayogadvārā kaivalyaṃ cāsya phalam /
    tatkāmaś cādhikārī / granthaviṣayayoḥ pratipādyapratipādakabhāvaḥ sambandhaḥ / granthasya
    kaivalyasya ca prayojyaprayojakabhāvaḥ sambandhaḥ / granthābhidheyasya saphalayogasya
    kaivalyasya ca sādhyasādhanabhāvaḥ sambandha ity uktam //2//
  \end{jyotsna}

\end{ekdosis}
\end{document}

### 2022-03-19 compile-tidyup.sh

@@ -1,0 +1,53 @@
+#!/bin/sh
+
+# optional argument 1 is filename, otherwise all relevant *.tex files
+
+xslcmd="java -jar /usr/share/java/saxon/saxon9he.jar -s:xml/HP1_OldMss-tei.xml -xsl:xslt/hp-online.xsl"
+
+xslcmdproc(){
+    ${xslcmd} | \
+	sed "/^\s*$/d" \
+	    > html/hp-online.html 
+}
+
+compile(){
+    pdf="${1%.*}.pdf"
+    tei="${1%.*}-tei.xml"
+    
+    pushd  "latex" || exit
+
+    if [ "${1}" -nt "${pdf}" ]
+    then
+	latexmk -lualatex "${1}"
+	
+	sed -i \
+	    -e "s_</\?lg[^>]*>_\n&\n_g" \
+	    -e "s_</l>\s*<l>_</l>\n<l>_g" \
+	    -e "s_</note>_&\n_g" \
+    	    -e "s_</?item>_&\n_g" \
+	    -e "s_</?list>_&\n_g" \
+	    -e "s_\\\\[+!]\?__g" \
+	    "${tei}"
+
+	cp  -v "${tei}" ../xml/
+    else
+	echo "No changes to ${1}, skipping compilation."
+    fi
+
+    popd || exit
+}
+
+
+if [ -f "${1}" ]
+then
+    compile "${1#*/}"
+elif [ -z "${1:-}" ]
+then
+    for f in "HP1_OldMss.tex" "HP1-TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex"
+    do
+	compile "${f}"
+    done
+fi
+xslcmdproc
+
+exit

### 2022-03-24 compile-tidyup.sh
main text in devanagari, extra container for sources

@@ -27,6 +27,7 @@ compile(){
     	    -e "s_</?item>_&\n_g" \
 	    -e "s_</?list>_&\n_g" \
 	    -e "s_\\\\[+!]\?__g" \
+	    -e "s_'_’_g" \
 	    "${tei}"
 
 	cp  -v "${tei}" ../xml/

### 2022-06-01 compile-tidyup.sh
skm/skp intergrated and better sigla display

@@ -2,7 +2,7 @@
 
 # optional argument 1 is filename, otherwise all relevant *.tex files
 
-xslcmd="java -jar /usr/share/java/saxon/saxon9he.jar -s:xml/HP1_OldMss-tei.xml -xsl:xslt/hp-online.xsl"
+xslcmd="java -jar /usr/share/java/saxon/saxon9he.jar -s:xml/HP_skm_test-tei.xml -xsl:xslt/hp-online.xsl"
 
 xslcmdproc(){
     ${xslcmd} | \
@@ -44,7 +44,7 @@ then
     compile "${1#*/}"
 elif [ -z "${1:-}" ]
 then
-    for f in "HP1_OldMss.tex" "HP1-TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex"
+    for f in "HP_skm_test.tex" "HP1_OldMss.tex" "HP1-TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex"
     do
 	compile "${f}"
     done

### 2022-06-07 latex/HP1_TranslComm.tex
Readings vor avataranikas, new apparatus-commands

\input{Preamble-Print}
\makepagestyle{HPed}
\makeoddhead{HPed}{\small\texteng{HP Transl. \& Comm.}}{}{\small\texteng{\today}}
\makeevenhead{HPed}{\small\texteng{HP Transl. \& Comm.}}{}{\small\texteng{\today}}
\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}

% additions/changes 2022-03-12 mm; 2022-6-1 MD:
\SetTEIxmlExport{autopar=false}
\NewDocumentEnvironment{translation}{O{}}{\textcolor{blue}{\textbf{Transl.:}}}{}
\NewDocumentEnvironment{philcomm}{O{}}{
	\textcolor{blue}{\textbf{Comm.:}}}{}
\NewDocumentEnvironment{sources}{O{}}{
	\textcolor{blue}{\textbf{Testimonia:}}}{}

\setlength{\parskip}{0.3em}
\setlength\parindent{0pt}
\def\startverse{\begin{ekdverse}\begin{itshape}}
\def\endverse{\end{itshape}\end{ekdverse}}
\setvnum{}

\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

\chapter*{Translation \& philological commentary}
\subsection*{1.1}
\begin{translation}[hp01_001]
Homage to the glorious Ādinātha by whom the knowledge of Haṭhayoga was taught. It shines forth like a ladder for one desirous of climbing to the lofty terrace of the royal palace.
\end{translation}

\begin{sources}[hp01_001]
Yogasārasaṃgraha (line 3090), Gheraṇḍasaṃhitā
\end{sources}

%\begin{philcomm}[hp01_001]
%\end{philcomm}

\subsection*{1.2}
\begin{translation}[hp01_002]
Having bowed to the glorious guru, the Lord, the yogi Svātmārāma has taught the doctrine of Haṭhayoga solely for [attaining] Rājayoga.
\end{translation}

%\begin{philcomm}[hp01_002]
%\end{philcomm}

\subsection*{1.3}
\begin{translation}[hp01_003]
For those who are ignorant of Rājayoga because of confusion in the darkness of many opinions the compassionate Svātmārāma composes the Lamp on Haṭha.
\end{translation}

%\begin{philcomm}[hp01_003]          
%\end{philcomm}

\subsection*{1.4}
\begin{translation}[hp01_004]
In fact, Matsyendra, Gorakṣa and other [siddhas] knew the doctrine of Haṭha, and the yogi Svātmārāma knows it owing to their favour.
\end{translation}

%\begin{philcomm}[hp01_004]          
%\end{philcomm}

\subsection*{1.5}
\begin{translation}[hp01_005]
The glorious Ādinātha, Matysendra, Śābara, Ānandabhairava, Cauraṅgī, Mīna, Gorakṣa, Virūpākṣa, Bileśaya,
\end{translation}

%\begin{philcomm}[hp01_005]          
%\end{philcomm}

\subsection*{1.6}
\begin{translation}[hp01_006]
Manthānabhairava, Siddhabuddha, and Kanthaḍi, Pauraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
\end{translation}

%\begin{philcomm}[hp01_006]          
%\end{philcomm}

\subsection*{1.7}
\begin{translation}[hp01_007]
Kānerī, Pūjyapāda, Nityanātha, Nirañjana, Kapālī, Bindunātha, and the one named Kākacaṇḍīśvara.
\end{translation}

%\begin{philcomm}[hp01_007]          
%\end{philcomm}

\subsection*{1.8}
\begin{translation}[hp01_008]
Allamaprabhudeva, Ghoḍācolī, Ṭiṇṭiṇī, Bhālukī and Nāgabodha and Khaṇḍakāpālika.
\end{translation}

%\begin{philcomm}[hp01_008]          
%\end{philcomm}

\subsection*{1.9}
\begin{translation}[hp01_009]
These great siddhas and others have destroyed the rod of death through the power of Haṭhayoga and wander in the universe.
\end{translation}

%\begin{philcomm}[hp01_009]          
%\end{philcomm}

\subsection*{1.10}
\begin{translation}[hp01_010]
Haṭha is considered a refuge for those who are burnt by the pains of transmigration. Haṭha is the foundational tortoise for the worlds of all yogas.
\end{translation}

%\begin{philcomm}[hp01_010]          
%\end{philcomm}

\subsection*{1.11}
\begin{translation}[hp01_011]
The doctrine of Haṭha should be kept very secret by those yogis who are desiring success. When it is secret it becomes potent. However, when it has been revealed, it becomes impotent.
\end{translation}

\begin{sources}[hp01_011]
ŚS 5.254; BKhP 10v4, YCM
\end{sources}

%\begin{philcomm}[hp01_011]          
%\end{philcomm}

\subsection*{1.12}
\begin{translation}[hp01_012]
In well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live remotely in a small hut.
\end{translation}

\begin{sources}[hp01_012]
GŚ 32cd; BKhP 107v1, YCM
\end{sources}

%\begin{philcomm}[hp01_012]          
%\end{philcomm}

\subsection*{1.13}
\begin{translation}[hp01_013]
It has a small door, without cracks and holes, its length is not too high or low, thickly smeared with cow dung in the proper way, clean, free from everything that annoys, adorned with a pavilion, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
\end{translation}

\begin{sources}[hp01_013]
BKhP 107v3, YCM
 \end{sources}

\begin{philcomm}[hp01_013]          
\emph{piṭharaṃ} good
\end{philcomm}

\subsection*{1.14}
\begin{translation}[hp01_014]
Locating oneself in a hut of such a kind, free from all worry, one should constantly practise yoga in the way taught by one's guru.
\end{translation}

%\begin{philcomm}[hp01_014]          
%\end{philcomm}

\subsection*{1.15}
\begin{translation}[hp01_015]
Overeating, exertion, chatter (gossiping/bickering?), sticking to rules, associating with people, inconstancy: through [these] six, yoga will be abandoned.
\end{translation}

\begin{philcomm}[hp01_015]          
Impossible to decide on meaning of \emph{niyamāgraha}, avagraha invisible, Jyotsnā takes it as over-insistence (as if \emph{āgraha} was implied) as he relates it to extreme ascetic practice; will be abandoned if \emph{prahāsyate} is adopted.
\end{philcomm}

\subsection*{1.16}
\begin{translation}[hp01_016]
From zeal, conviction, resolve, understanding of the truth [of yoga], discernment, abandonment of associating with people: by [these] six, on the other hand, yoga is successful.
\end{translation}

\begin{philcomm}[hp01_016]          
\emph{tattva} can sometimes refer to the practices of yoga: e.g., \emph{tritattva} in AS 13.12, 14.2--3.
\end{philcomm}

\subsection*{1.17}
\begin{translation}[hp01_017]
Because it is the first auxiliary of haṭha, āsana is taught first. This (\emph{tad}) āsana brings about steadiness, good health and dexterity.
\end{translation}

%\begin{philcomm}[hp01_017]          
%\end{philcomm}

\subsection*{1.18}
\begin{translation}[hp01_018]
I shall now teach some of the postures which have been accepted by sages (\emph{munis}) such as Vasiṣṭha and yogis such as Matsyendra.
\end{translation}

%\begin{philcomm}[hp01_018]          
%\end{philcomm}

\subsection*{1.19}
\begin{translation}[hp01_019]
Correctly placing the soles of both feet between the knees and thighs [and] sitting up with the body straight: they call that the auspicious [pose].
\end{translation}

\begin{sources}[hp01_019]
VS 1.68 etc
\end{sources}

%\begin{philcomm}[hp01_019]          
%\end{philcomm}

\subsection*{1.20}
\begin{translation}[hp01_020]
One should place one's right heel on the left, at the side of the [lower] back, and on the right the left in the same way. This is the Cow's Mouth [posture]; it is like a cow's mouth.
\end{translation}

\begin{sources}[hp01_020]
VS 1.70, YY? 3.5; BKhP 93v7
\end{sources}

%\begin{philcomm}[hp01_020]          
%\end{philcomm}

\subsection*{1.21}
\begin{translation}[hp01_021]
Having placed one foot on one thigh and the seated thigh on the other foot. Vīrāsana is taught [thus]. 
\end{translation}

\subsection*{1.22}
\begin{translation}[hp01_022]
Having blocked the anus with the crossed ankles, focused [the yogi] remains [thus]. Adepts of yoga know that this is kūrmāsana.

Or: While focused, [the yogi] should bind the anus with crossed ankles. This is kūrmāsana. Experts in yoga know this.
\end{translation}

\subsection*{1.23}
\begin{translation}[hp01_023]
Having assumed padmāsana, inserted the hands between the knees and thighs, and placed [the hands] on the ground, [the yogi] remains in the air. This is kurkuṭāsana

Or: [The yogi] should assume padmāsana, insert his hands between the knees and thighs onto the ground and place himself in the air. This is kurkuṭāsana.
\end{translation}

\begin{philcomm}[hp01_023]
Kurkuṭa/kurkkuṭa variant in \getsiglum{V1,J10ac,V3} is attested in Pañcatantra (M-W).
\end{philcomm}

\subsection*{1.24}
\begin{translation}[hp01_024]
Bound in kurkuṭāsana, [the yogi] binds the neck with the hands and lies supine like a tortoise. This is uttānakūrmāsana. 
\end{translation}

\begin{philcomm}[hp01_024]
\emph{°bandhasthaḥ} or \emph{°vat kṛtvā}? Only V1 has latter, which is simpler. Are the others trying to improve it? Stemmatically ambiguous as \emph{°bandha°} is on one branch (V3/J8, V19) and \emph{°madhya°} the other (J10, J17, N17). V1 looks like an outlier.

Adopt \emph{bandhastho} with note. \emph{°bandhastha} not found in any other texts. \emph{°vat kṛtvā} is not good.
\end{philcomm}

\subsection*{1.25}
\begin{translation}[hp01_025]
Having clasped the big toes with hands, [the yogi] should perform a bow drawing action as far as the ear. This is called dhanurāsana.
\end{translation}

\begin{philcomm}[hp01_025]
At a stretch (!), this could mean drawing a crossbow, with both hands taken back to the ears, but probably just means like a normal bow.

Gheraṇḍasaṃhitā:
\startverse
 prasārya pādau bhuvi daṇḍarūpau |
 karau ca pṛṣṭhaṃ dhṛtapādayugmam |\\
 kṛtvā dhanustulyavivartitāṅgaṃ |
 nigadyate vai dhanurāsanaṃ tat || 2.18 ||
\endverse
\end{philcomm}

\subsection*{1.26}
\begin{translation}[hp01_026]
Having grasped the right foot, which is placed at the base of the left thigh, and the left foot, which is wrapped around the outside of the knee, [the yogi] remains with his body twisted. This āsana was taught by Matsyendranātha.
\end{translation}

\subsection*{1.27}
\begin{translation}[hp01_027]
Through practice, Matsyendra's seat gives people a torch for the abdomen, a destructive weapon for a range of terrible diseases, the awakening of K and steadiness of the spine.
\end{translation}

\begin{philcomm}[hp01_027]
Difficult to construe first half of verse: \emph{jaṭhara} compound needs to be in apposition to \emph{matsyendrapīṭhaṃ}. KDham ed has variant \emph{jaṭharapradīpaṃ} but \emph{pradīpa} is masculine. Perhaps we have to accept \emph{jaṭharapradīptiṃ}, but that crosses stemmatic lines, so \emph{jvalanapradīptiṃ}; or \emph{jaṭharapravṛddhiṃ}. \emph{jvalana} is found at 3.66 to refer to the fire in the body but \emph{jaṭhara}, stomach, is not found without \emph{agni} elsewhere.

YCM has \emph{jaṭharapravṛddhiḥ}.

Adopt \emph{jaṭharapradīptiṃ} (\emph{°dīptiṃ} is an emendation), understood as object of \emph{dadāti}.

Pāda d, J10ac and J17 have \emph{candra} for \emph{daṇḍa}, also in YCM and 6-chapter HP.
\end{philcomm}

\subsection*{1.28}
\begin{translation}[hp01_028]
[The yogi] should stretch out both feet on the ground like staffs, hold the ends of both feet with the hands, place the forehead upon the knees and remain thus. They call this the back-stretch (paścimatānam).
\end{translation}

\begin{philcomm}[hp01_028]
Only V1 has \emph{dorbhyāṃ padāgra}, others have variations on the much inferior \emph{dvābhyāṃ karābhyāṃ}. Vulgate has same as V1, which is a surprise.

V1 reading \emph{dorbhyāṃ pādāgradvitayaṃ} is best, adopt.

V1’s \emph{tāṇabandhaḥ} doesn’t work with \emph{idaṃ}.
\end{philcomm}

\subsection*{1.29}
\begin{translation}[hp01_029]
This back-stretch is the foremost among āsanas. It makes the breath flow in the back [i.e. Central channel], increases the digestive fire, makes the belly thin and prevents diseases in men.
\end{translation}

\begin{philcomm}[hp01_029]
Adopt \emph{arogitāṃ}.
\end{philcomm}

\subsection*{1.30}
\begin{translation}[hp01_030]
Supporting oneself on the ground with both hands, their elbows placed on either side of the navel, with a raised position (? uccāsanaḥ) one is placed up into the air [as straight] as a stick. They call this posture the peacock.
\end{translation}

\begin{philcomm}[hp01_030]
1.30 b tat refers to \emph{karadvaya} (cf. VS).
\emph{uccāsano} to be adopted; \emph{samunnataśiraḥpādo} is reading of VS. which is derived from Vimānārcanākalpa 96: 
\emph{karatale bhūmau samsthāpya kūrparau nābhipārśvayor nyasya na(unnata)
̇ taśira ̄hpādau dandavad vyomni samsthito mayūrāsanam iti}.
\end{philcomm}

\subsection*{1.31}
\begin{translation}[hp01_031]
The glorious mayūra posture gets rid of all diseases of the abdomen such as bloating and overcomes [imbalances of] the doṣas. It completely incinerates food which is bad or has been eaten to excess, it generates digestive fire and it digests strong poison.
\end{translation}

\begin{philcomm}[hp01_031]
\emph{aśeṣaṃ} better
\end{philcomm}

\subsection*{1.32}
\begin{translation}[hp01_032]
Lying on one’s back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] all [other] āsanas and relaxes the mind.
\end{translation}

\subsection*{1.33}
\begin{translation}[hp01_033]
However, Śiva taught eight-four āsanas. I shall take the four essential ones from them and teach them. 
\end{translation}

\subsection*{1.34}
\begin{translation}[hp01_034]
The adept, lotus, lion and auspicious pose are the best tetrad and, among those, one should always sit in a comfortable adept’s pose. 
\end{translation}

\subsection*{1.35}
\begin{translation}[hp01_035]
Having joined the place of the perineum with the heel of the foot, the yogi should firmly fix the [other] foot on the penis. Having held the face and chest together and the body erect, [the yogi] remains still, his senses restrained, gazing between the brows with his eyes unmoving. This breaks open the door to liberation and is called the adept’s pose.
\end{translation}

\begin{philcomm}[hp01_035]
KDh ed has \emph{āsya}, variant in mss kh and gha, both BORI.
Jürgen suggests we might take \emph{āsyahṛdaye} and \emph{vigraham} as the objects of \emph{dhṛtvā} and samam as an adverb (i.e., holding straight the face, chest and body). However, in a different doctrine.
\end{philcomm}

\subsection*{1.36}
\begin{translation}[hp01_036]
Having fixed the left heel on the penis, and put the other heel on that, this is siddhāsana. 
\end{translation}

\subsection*{1.37}
\begin{translation}[hp01_037]
Some proclaim this is siddhāsana, others know it as vajrāsana, a few say it is muktāsana and some guptāsana.
\end{translation}

\subsection*{1.38}
\begin{translation}[hp01_038]
The Siddhas know Siddhasana as the single most important amongst
all postures, in the same way as measured diet amongst rules and non-violence amongst observances. 
\end{translation}

\begin{philcomm}[hp01_038]
\emph{iva} or \emph{eva}? \emph{iva} does work — like \emph{siddhāsana}, \emph{mitāhāra} and \emph{ahiṃsā} are the best, but for it to work properly \emph{mitāhāra} and \emph{ahiṃsā} should be accusative. V19 has acc + \emph{iva}, which seems best, especially with \emph{siddhāḥ viduḥ}, but this might be a correction as V19 often corrects. However, one old KDham BORI (?) ms (pha, 1695 CE) has it, as does Jyotsnā, so adopt.

In pāda d V19 has \emph{siddhāsanam idaṃ viduḥ}, but the reading of all other mss is preferable.

Clearly based on DYŚ 33, which includes \emph{ekaṃ} and \emph{mukhya}.
\end{philcomm}

\subsection*{1.39}
\begin{translation}[hp01_039]
Among the eighty-four postures, one should regularly practise just Siddha; in the same way one should practise Suṣumnā among the 72,000 channels.
\end{translation}

\begin{philcomm}[hp01_039]
Odd to have \emph{suṣumnām} as object of \emph{abhyaset}. YCM has this reading though. Perhaps cd were added somewhat indiscriminately by Svātmārāma (with nominative \emph{suṣumnā}) and then others tried to make sense of it.

Some witnesses, including Jyotsnā, have \emph{nāḍīnāṃ malaśodhanam/e} for pāda d, which is probably a patch (no other texts say siddhāsana clears the channels), but cf. Amṛtasiddhi in which the practices are said to bring about cālana of the nāḍīs (e.g. 11.6).

Good example of early contamination.

[\emph{nāḍiṣu} is better supported \getsiglum{J10ac,V19,J17}.]
\end{philcomm}

\subsection*{1.40}
\begin{translation}[hp01_040]
By meditating upon the self, restricting the diet and regularly practising Siddhāsana for twelve years, the yogi attains the Niṣpatti stage.
What’s the point of lots of exhausting postures when there is Siddhāsana?
\end{translation}

\begin{philcomm}[hp01_040]
Only possible variant is \emph{mitāhāro} in V19.

J8 might be correction of J10’s unmetrical reading.

[\emph{sadāsiddhāsanābhyāsād}? Or maybe read \emph{sadā} with \emph{avāpnuyāt}]

V19 is found in YCM: \emph{śramadair bahubhịh pīṭhaiḥ kiṃ syāt siddhāsane sati}; JM this seems best to me.

Is this notion of āsanas causing \emph{śrama} already current in HY texts?
[It is mentioned in the verse on śavāsana]
\end{philcomm}

\subsection*{1.41}
\begin{translation}[hp01_041]
When the Prāṇa breath has been carefully stopped in kevala kumbhaka, the Unmanī [state] arises automatically, without effort, in the same way that
\end{translation}

\begin{philcomm}[hp01_041]
V1 has 
\startverse
śramādau bahubhiḥ pīṭhais sadā siddhāsane sati  |\\
prāṇānile sāvadhānaṃ baddhe kevalakumbhake  || 1.41 ||
\endverse
Mixing up both versions of the verse — contamination already?
\end{philcomm}

\subsection*{1.42}
\begin{translation}[hp01_042]
when just Siddhāsana alone is always firmly bound, the three bandhas arise automatically, without effort.
\end{translation}

\begin{philcomm}[hp01_042]
\emph{dṛḍhe} goes across stemma, although \emph{ṃ} vs \emph{e} is moot.
\end{philcomm}

\subsection*{1.43}
\begin{translation}[hp01_043]
There is no posture like siddhāsana, no breath like the restrained breath, no mudrā like khecarī, no dissolution like nāda.
\end{translation}

\begin{philcomm}[hp01_043]
\emph{na kumbhasadṛśo nilaḥ} is difficilior lectio and attested by all early witnesses except V19: (almost?) all testimonia have \emph{kumbhaḥ kevalopamaḥ}; Śivasaṃhitā has \emph{kumbhasadṛśaṃ balam}.
\end{philcomm}

\subsection*{1.44}
\begin{translation}[hp01_044]
Now lotus pose.
One should place the right foot on the left thigh, and the left on the right though, hold firmly the big toes with the hands behind the back, place the chin on the chest and gaze at the tip of the nose. This is called lotus pose, which destroys diseases for those who undertake the yamas. 
\end{translation}

\subsection*{1.45–47}
\begin{translation}[hp01_045]
However, in another view,
Having put the upturned feet carefully on the thighs and the upturned hands in the middle of the thighs, one should fix the eyes on the tip of the nose. Having raised the root of the uvula with the tongue, one should place the chin on the chest and gradually [draw in\footnote{The verb \emph{ākṛśya} follows in the next verse in the \emph{Dattātreyayogaśāstra}.}] the breath [...]. This is taught as lotus pose, which cures all diseases. 
\end{translation}

\begin{philcomm}[hp01_045]
\emph{uttabhya} vs \emph{uttambhya}.
The witnesses split predictably along the two main branches of the stemma. The evidence of the DYŚ is important here.

\emph{vakṣasyāsthāpya} is a marmasthāna

%\emph{pavanaṃ śanaiḥ} is left hanging. The readings with \emph{vakṣa sthāpayet} do not offer a solution nor indicate how Svātmārāma may have redacted this to make the syntax complete. Instead, it seems that he quoted these two verses (1.45–46) from the Dattātreyayogaśāstra and simply omitted the next verse that made sense of \emph{pavanaṃ śanaiḥ} because it was not relevant to the posture itself.

\emph{pavanaṃ śanaiḥ} is left hanging, perhaps, because of an eyeskip that happened early in the transmission. The following verse in the DYŚ ends with \emph{pavanaṃ śanaiḥ}.

\startverse
nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\
uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\
yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\
yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\
idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |
\endverse

% Jü:
[JH] The background of the passage \emph{rājadantamūlaṃ ca jihvayā uttambhya} in 1.46 is more complex
than it may appear. Here it is a literal quotation from the DYŚ, but many other Haṭhayogic texts
teach a particular position of the tongue, in which it is in one or the other way turned back in
the direction of the uvula, as we read explicitly in the \emph{Vivekamārtāṇḍa} (REF):
\emph{kapālakuhare jihvā praviṣṭā viparītagā}. Brief references to this practice can become
ambiguous for various reasons, and this has possibly confused Brahmānanda.

One reason is that there is a, probably older, rule for meditation postures according to which the
tongue rests near the teeth. One instance would be \emph{Svacchandatantra}, which teaches a
meditation pose called \emph{divyaṃ karaṇam} (4.365f.), in which the tongue is to rest at the tip
of the teeth (\emph{dantāgre jihvām ādāya}). Other Tantric texts have this or similar rules, in
which the tongue is supposed to rest either on the teeth or the palate,\footnote{This rule is
  found in \emph{Iśānaśivapaddhati}: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān
    asaṃspṛśan ṛjukāyaḥ}. REF} the earliest instance being perhaps \emph{Mrgendrāgama} ?.18.
Placing the tongue where it does not disturb the meditation seems quite appropriate for a
``normal'' meditative practice.

[Jason: There’s also a clear reference to the tip of the tongue being placed in the middle of the palate in 2.27 of the Yogapāda of the Mataṅgapārameśvaratantra (tālumadhyagatenaiva jihvāgreṇa mahāmune). In fact, in works that predate Haṭhayoga (i.e., 11th c.), the most common instruction is to put the tip of the tongue on the palate (tālu).]

We might try to interpret the passage in this manner, however, once Haṭhayogic physiology is at the
background, we must assume that the aim is to reach back to the uvula, to the source of the
"nectar".  For the background and for the crucial references see Mallinson's note on
\emph{Khecarīvidyā} 1.65ab.\footnote{p. 209.}  Confusingly Yogic terminology has used and possibly
invented new names for uvula, and among these especially the term \emph{rājadanta} may give rise to
confusion, since, as we have seen, the tongue might also in some Yoga systems be placed at the
(front) teeth.

Furthermore the details in these descriptions of the \emph{khecarīmudrā} are manifold. The 10th
century \emph{Mokṣopāya} says that the tongue rests at the ``source of the
palate''\footnote{\emph{tālumūlatalālagnajihvā} MU V.55.14c.} and the commentary
\emph{Saṃsārataraṇi} on the parallel passage in \emph{Laghuyogavāsiṣṭha} V.6.155, which reads
\emph{tālumūlāntarālagnajihva-}, explains that this means that the tongue is to be placed in the
middle of the two regions of the palate, and that this is the \emph{nabhomudrā}, alias
``\emph{khecarī}''.\footnote{\emph{tālumūlāntarālagnajihvamūlah. tālumūlayoh. kākudamūladeśayoh. āntare
    lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā / yā hi khecarı̄ty ucyate.}}

A little later in the \emph{Mokṣopāya} it is made clear that one should reach the uvula, ``placed a
the root of the palate''.\footnote{\emph{tālumūlagatāṃ yatnāj jihvayākramya ghaṇṭikām}
  (V.78.24ab)}

In view of this background we must conclude that the author of the \emph{Jyotsnā} was probably not
aware of the Yogic meaning of \emph{rājadanta} and has tried his best to make sense of the passage,
echoing the idea of the two roots of the palate (although his text is not talking about the
palate), but then wisely refers to the instruction of the teacher for practical details, probably
noticing that his literal interpretation is somewhat opaque (I omit the synonyms for clarity):

\begin{quote}
Pressing both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.

\emph{rājadantānāṃ daṃṣṭrāṇāṃ savyadakṣiṇabhāge sthitānāṃ mūle ubhe mūlasthāne jihvayā uttambhya ūrdhvaṃ stambhayitvā | guru-mukhād avagantavyo 'yaṃ jihvā-bandhaḥ |}
\end{quote}

[Jason: also, Brahmānanda’s comments on 3.22 indicate that he thought rājadanta refers to the front teeth. When commenting on rājadantasthajihvāyāṃ, he says kutaḥ ? yato dantānāṃ rājāno rājadantā rājadanteṣu tiṣṭhatīti rājadantasthāḥ, rājadantasthā cāsau jihvā ca tasyāṃ rājadantasthajihvāyāṃ bandhaḥ, taduparibhāgasya sambandhaḥ śastaḥ].
\end{philcomm}

\subsection*{1.47}
\begin{translation}[hp01_047]
It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
\end{translation}

\begin{philcomm}[hp01_047]
Durlabham is ambiguous as to whether the posture is hard to perform or hard to acquire. The latter is the usual reading.
\end{philcomm}

\subsection*{1.48}
\begin{translation}[hp01_048]
Next is taught the doctrine of Matsya.

[The yogi] should put the hands together in a bowl shape, very firmly assume padmāsana, firmly place the chin on the chest and meditation in the mind, [and] raising the Apāna breath over and over again, releasing the held Prāṇa, a man attains unequalled knowledge through the power of the goddess [Kuṇḍalinī].
\end{translation}

\begin{philcomm}[hp01_048]
It is unusual to find Matsyendra referred to as simply Matsya as in the subheading here. This verse is Vivekamārtaṇḍa 36. The teachings of the VM are attributed to Gorakṣa, not Matsyendra, although in the extended recension of it known as Gorakṣaśataka etc. the second verse includes homage to Mīna, i.e. Matsyendra.

The end of pāda b is uncertain. Later witnesses including Brahmānanda have \emph{dhyāyaṃś ca} but none of the early ones does. So we have to take \emph{dhyānaṃ} with \emph{sannidhāya}. That makes tat problematic though: what does it refer to?

The two participles \emph{proccālayan} and \emph{muñcan} imply that the two things are happening at the same time, which is surprising but perhaps possible. Or perhaps one can understand them as sequential.

In the VM this verse comes in the passage on raising Kuṇḍalinī, so tat could refer to visualising K opening the mokṣadvāra.
\end{philcomm}

\subsection*{1.49}
\begin{translation}[hp01_049]
While in padmāsana the yogi should fill himself up [with Prāṇa] through [all] the channels.  He who takes in breath [in this way] is sure to be liberated.
\end{translation}

\subsection*{1.50--52}
\begin{translation}[hp01_050]
Now the lion’s pose
[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. Left ankle is on right, and the right ankle on the left.

Having placed both hands on the knees and spread his fingers, his mouth open, he gazes with his eyes fixed on the tip of the nose. This is the lion’s pose, which is always worshipped by yogis. It unites the three locks and is the best of all postures.
\end{translation}

\subsection*{1.53--54}
\begin{translation}[hp01_053]
Now, the friendly pose.
[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.  Having firmly and motionlessly held the sides of the feet with the hands, this is bhadrāsana, which cures all diseases and poisons. Yogis of the siddha tradition call it Gorakṣāsana. 
\end{translation}

\subsection*{1.55}
\begin{translation}[hp01_055]
When the great yogi does not become tired [while engaged] in these āsanas and bandhas, he should now practise the breath techniques of mudrā and so forth in order to purify the channels.
\end{translation}

\begin{philcomm}[hp01_055]
55d is ra-vipulā with \emph{nāḍi}

\emph{nāḍiśuddhiṃ} is difficult to construe. Jim thinks it is a second accusative of the \emph{abhyaset} and reads it as \emph{nāḍiśuddhyartha}. I wonder if it is not better understood as a bahuvrīhī qualifying \emph{°pavanakriyām} (i.e., the yogi should practise a breath technique that is channel-purifying. In other words, he does not practise just any breath technique). As a bahuvṛīhi, it may also be understood as a \emph{hetuviśeṣaṇa} (i.e., why should the yogi practice a breath-technique such as mudrā? Because it is channel-purifying).
\end{philcomm}

\subsection*{1.55*1}
\begin{translation}[hp01_055_1]
Success arises for one engaged in practice. How can it arise for one who has no practice? Success in yoga does not arise by merely reading scriptures.

Wearing a robe does not bring about success, nor does talking [about yoga]. Practice alone is the cause of success. This is true, there is no doubt. In this tradition, it should not be given to one who wears robes and is devoted to his genitals and stomach.
\end{translation}

\begin{philcomm}[hp01_055_1]
1.55*1–2 are omitted from V1 and V19, so may not have been included in the earliest version of the text. In fact, it is possible that both were added (perhaps initially as marginal notes) to elaborate on kriyā in 1.55d. Both verses are similar to verses from the Dattātreyayogaśāstra, and these verses (except 1.55.2ef) appear in the vulgate, but towards the end of chapter 1 (1.65–66).

And then just half of DYŚ 47 is given, resulting in a near-nonsensical hemistich.

The syntax of 1.55.2ef is corrupt. One has to emend to \emph{deyā} to make sense of it. 
\end{philcomm}

\subsection*{1.55*3}
\begin{translation}[hp01_055_3]
In me, the pure ocean of awakening, is this empty bubble we call the universe dissolved or does it arise? Where does this veil of doubt about this come from?
\end{translation}

\begin{philcomm}[hp01_055_3]
It is very difficult to find a reason why this verse should be inserted here. It is apparently a
\emph{muktaka} that would befit an accomplished spiritual poem more than our \emph{Haṭhapradīpikā},
even here, in what appears as a sort of miscellaneous section at the end of a chapter.  In this
verse, the lyrical subject wonders about why the mind is still able to doubt, despite its insight
into the nature of reality and the reader wonders, how the illusionist verse could be understood to
fit our Yoga text, perhaps the scribe of the archetype of V19 and N17, was fond of it.

The source is, as far as we can say, the \emph{Śāntiśataka} of the Kashmirian poet Sillana or
Silhaṇa,\footnote{The mss.\ of his \emph{Svātmopalabdhiśataka} give the name as Sillana, the mostly
Bengali mss. of the Śāntiśataka read Śilhaṇa, as does Aufrecht in his CC. See Hanneder:
forthcoming.} who cannot be dated with any certainty, but predates the HP by a few centuries. The
edition of this text – where hundred original verses had to be identified – places it into an
appendix of doubtful stanzas,\footnote{Karl Schönfeld: \emph{Das Śāntiśataka}. Leipzig:
Harrassowitz 1910, p.\,90 (A9).} but the editor saw no compelling reason not to regard it as
original except through the fact that it is not transmitted in all manuscripts.  What prevents
further investigation of the matter is the lack of Kashmirian mss.\ for the \emph{Śāntiśataka} and
its compilatory character: one quarter of the material is identical with Bhartṛhari's
\emph{Vairāgyaśataka}. A still superficial glance at Sillana's \emph{Svātmopalabdhiśataka} gives
the impression that our verse would fit there, but not so much in the \emph{Śāntiśataka}.  While
these are only preliminary observations the verse is probably not original in any stage of
development of the HYP, but limited to V19 (and N17).
\end{philcomm}

\subsection*{1.55*4}
\begin{translation}[hp01_055_4]
Realisation from scripture, one's own guru or oneself and the cessation of mind; all these are methods that have been systematised and taught by the wise in this tradition.
\end{translation}

\begin{philcomm}[hp01_055_4]
last greyscale verse (\emph{śrutipratītiḥ…}) is only in V19 of old mss (but some other later ones) and is quoted in Yogacintāmaṇi.

The reading in the YCM (\emph{manaso nirodhaḥ}) is better in a yogic context.
\end{philcomm}

\subsection*{1.56}
\begin{translation}[hp01_056]
The various āsana, breath retention, bodily technique (karaṇa) called seals (mudrā), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.
\end{translation}

\begin{philcomm}[hp01_056]
Verse is omitted from V1, probably deliberately as it doesn’t have chapter 4, which teaches nādānusandhāna. The numbering in V1 suggests that its exemplar had this verse.

Kumbhaka is almost always masculine.

Marmasthāna: not clear whether to adopt \emph{citro} or \emph{citraṃ}, or V19’s \emph{mudrākhyaṃ karaṇaṃ tathā} or the others’ \emph{mudrādikaraṇāni ca}.

See also 1.65, which has \emph{mudrādikaraṇāni ca}, so perhaps it was through confusion with this that the same reading is found in some witnesses of 1.56.

Yes, V19 reading probably best.
\startverse
āsanaṃ kumbhakaś citraṃ mudrākhyaṃ karaṇaṃ tathā |\\
atha nādānusandhānam abhyāsānukramo haṭhe ||
\endverse

It seems that the four aṅgas of Haṭhayoga are being referred to in the singular (hence \emph{āsanaṃ}), whereas in 1.65 the plural is used (i.e., \emph{pīṭhāni}). Therefore, we should adopt \emph{citraṃ [...] karaṇaṃ tathā}.
\end{philcomm}

\subsection*{1.57}
\begin{translation}[hp01_057]
Celibate, restricted in diet and devoted to yoga, a yogi succeeds in upwards of a year. No doubt about this should be entertained.
\end{translation}

%\begin{philcomm}[hp01_057]
%\end{philcomm}

\subsection*{1.58}
\begin{translation}[hp01_058]
When very unctuous and sweet food that is without a quarter portion [of the stomach] is eaten for love of śiva, it is called a restricted diet (\emph{mitāhāra}).
\end{translation}

\begin{philcomm}[hp01_058]
This verse probably derives from the ‘original’ GŚ (12c–13b). It is also found, but reworked to be about the \emph{mitāhārī}, in Nowotny’s GŚ (extended recension of VM) at 55.
\end{philcomm}

\subsection*{1.59}
\begin{translation}[hp01_059]
[Adepts] say the [following] is unwholesome: pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink. Flesh of goats and sheep, curds, diluted buttermilk, poor man's pulse, Jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like. 
\end{translation}

\begin{philcomm}[hp01_059]
59a \emph{kaṭvamla°} is better than \emph{kaṭvāmla°} and well attested elsewhere in lists of tastes and types of foods.
On the meaning of \emph{uṣṇa} (in relation to food) see Meulenbeld 1974: 254 fn. 13: ‘Cakra mentions as a variant: \emph{katvamlalavaṇakṣāra} (pungent. acid, saline and caustic). Cakra remarks that the term `hot' (\emph{uṣṇa}) denotes hot on touch when it occurs the first time, and hot with regard to potency when it occurs for the second time.’

\emph{°hari}(\emph{ī} here for metre?)\emph{taśāka°} in some nighaṇṭus is horseradish, which makes better sense here.
\startverse
śigrur haritaśākaś ca śākapattraḥ supattrakaḥ | \textup{Rājanighaṇṭu 7.26}\\
śigruko haritaśākaś ca mato vai mūlapatrakaḥ | \textup{Sauśrutanighaṇṭu 75ab}\\
\endverse
Brahmānanda’s understanding of \emph{harītaśāka} as \emph{patraśāka} is probably wrong if \emph{patraśāka} was intended as ‘leafy vegetables’. But perhaps \emph{patraśāka} can also mean \emph{śigru} (note \emph{śākapattraḥ} above).

Anusvāra at end of \emph{śāka}?

\emph{°sauvīra°} probably means sour gruel.
Brahmānanda: \emph{sauvīra} = \emph{kāñjika} (fermented rice water).
Meulenbeld, madhavanidāna pp. 516–517
\emph{sauvīra} is sour gruel made from barley and wheat. On the process see, Suśruta 1.44.35--40ab.
(PV Sharma’s translation of this passage:)

\begin{quote}
‘Roots of trivṛt etc., the first group (\emph{vidārigandhādi}), \emph{mahat pañcamūla}, \emph{mūrvā} and \emph{śārṅgaṣṭā}, and also of \emph{snuhī}, \emph{haimavatī}, \emph{triphalā}, \emph{ativiṣā} and \emph{vacā} -- these are taken and divided into two parts out of which one is decocted and the other is powdered; now, crushed barley grains are impregnated with the above decoction several times, dried and then slightly fried. Taking three parts of this and one part of the above powder are put in a jar and mixed with their (of \emph{trivṛt}, etc.) cold decoction and fermented properly. This is known as \emph{sauvīraka}.’
\end{quote}

But according to some nighaṇṭus, \emph{sauvīra} can also mean stibnite (an ingredient in some añjana’s and medicines):
\startverse
añjanaṃ yāmunaṃ kṛṣṇaṃ nādeyaṃ mecakaṃ tathā \\
srotojaṃ dṛkpradaṃ nīlaṃ sauvīraṃ ca suvīrajam //  \textup{Rājanighaṇṭu 13.86}
\endverse
Note also that the Yogaprakāśikā takes \emph{sauvīra} with \emph{taila}, perhaps to solve the problem of \emph{taila} on its own (see below for more on this): \emph{sauvīrataila} -- oil produced in the place Suvīra (\emph{suvīradeśodbhavatailam}).

Suvīra , a country the people of which, also called Suvira (V.79), Sauvira (XVI.21) and Sauvīraka (IV.23) ... S.M. Ali identifies it with the Rohri - Khairpur region of Sind (Geography of the Purānas, Delhi, 1966, p. 144).

\emph{taila} could refer to \emph{tilataila}: 
Śārṅgadharasaṃhitā: 
\emph{anuktāvasthāyāṃ paribhāṣāvidhiḥ [...] taile ’nukte tilodbhavam} 48

Dominik Wujastyk supplied this reference and may be able to comment more on taila in this list.

On the translation of \emph{madya}, see Mchugh (An Unholy Brew) 2021: 8.

\emph{ajāvimāṃsa}: note the variant \emph{ajādimāṃsa}, which makes better sense to me as \emph{aja} might simply indicate that the flesh of animals is meant (as opposed to humans). 
\emph{ājāvi°} is needed for the metre.

\emph{kulattha} is a kind of pulse, translated by Dominik Wujastyk as `poor man's pulse' (see Roots of Ayurveda).

\emph{kola}: Zizyphus Jujuba (Nadkarni 1926: pp. 919-920). Also known as \emph{badara}. This is how Brahmānanda takes it (\emph{kolaṃ kolyāḥ phalaṃ badaram}). According to Nadkarni, the fruit of the wild variety is very acid and astringent. It is believed to purify the blood and assist digestion. The bark is astringent and a simple remedy for diarrhoea. Root is useful as a decoction in fever and delirium.

There are references to \emph{kola} being pungent, though this does not seem to indicate sufficiently why \emph{kola} is mentioned separately as an \emph{apathya} food.

\emph{piṇyāka}: Sharma (Ḍalhaṇa and his commentary on drugs: 1982: 69) says, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’

\emph{hiṅgu}: Asafoetida (Nadkarni 1926: pp 360–361): `If long continued even in moderate doses, it gives rise to alliaceous eructations, acrid irritation in the throat, flatulence, diarrhoea and burning in the urine.'

\emph{laśuna} = garlic (Nadkarni 1926: 45).
\end{philcomm}

\subsection*{1.60}
\begin{translation}[hp01_060]
One should know as unwholesome food that has been reheated and is dry. [Food that is] too salty, the leftover paste of crushed sesame seeds (tilapiṇḍa), spoiled food, [prohibited] vegetables (?), and pepper (?) are to be omitted. 
\end{translation}

\begin{philcomm}[hp01_060]
We have not found any conclusive evidence for the meaning of \emph{tilapiṇḍa}. Brahmānanda glosses it as \emph{piṇyāka} (on which see the notes for the previous verse).

Not sure how to take \emph{kadaśanaśākotkaṭaṃ}. Brahmānanda understands it as a dvandva (i.e., \emph{kadaśana}, \emph{śāka}, \emph{utkaṭa}), where \emph{śāka} is prohibited vegetables and \emph{utkaṭa} is pepper.

The meaning of \emph{utkaṭa} is not clear. The word \emph{utkaṭā} can mean pepper according to some nighaṇṭus (e.g., Rājanighaṇṭu 5.16 \emph{pārvatī śailajā tāmrā lambabījā tathotkaṭā}). But \emph{utkaṭa} can refer to Saccharum Sara and \emph{utkaṭā} also to Laurus Cassia (cinnamon).

Also, \emph{utkaṭa} can be an adjective that means ‘abounding in’ at the end of a compound. So could \emph{kadaśanaśākotkaṭaṃ} have been intended as an adjectival tatpuruṣa along the lines of ‘[food] full of spoiled vegetables’?
\end{philcomm}

\subsection*{1.61}
\begin{translation}[hp01_061]
Similarly a saying by Goraksa:
One should avoid a liking for bad people, frequenting fire, women and roads, early morning bathing, fasting and the like, as well as such things as harming the body.
\end{translation}

\begin{philcomm}[hp01_061]
The vulgate has a parallel from the Amaraugha added. Also, the vulgate reads \emph{kāyakleśavidhiṃ tathā}.
\end{philcomm}

\subsection*{1.62}
\begin{translation}[hp01_062]
Pure food with wheat, rice, śāli rice, barley, sixty-day śāli rice, milk, ghee, candied sugar, unclarified fermented butter, ground sugar and honey. Dried ginger, fruit of the snake gourd and five vegetables, mung beans and so on, and rain water are wholesome for the best of sages.
\end{translation}

\begin{philcomm}[hp01_062]
\emph{khaṇḍa} -- candied sugar (Meul 507, different types of sugar).

\emph{navanīta} (MW fresh butter), Mchugh (2021) unclarified fermented butter.

\emph{sitā} -- ground sugar (Meul 507, different types of sugar) ``sitā is very white and looks like gravel"

\emph{madhu} -- honey.

\emph{paṭola} can refer to at least two different gourds. See Meul. p. 569 for a long list of possibilities, including TRICHOSANTHES DIOICA ROXB. (`pointed gourd'), T. CUCUMERINA LINN (snake gourd).

Nadkarni has two entries on \emph{paṭola}:
\begin{enumerate}
\item snake gourd (Nadkarni p. 863) is common in Bengal and cultivated in Northern India and Punjab. The unripe fruit of this climbing plant is generally used as a culinary vegetable and is very wholesome, specially suited for the convalescent.

\item smooth luffa (Nadkarni p. 518) is a hairy climbing herb extensively cultivated in several parts of India. The fruit is edible. Medicinally it is described as `cool, costive, demulcent, producive of loss of appetite and excitive of wind, bile and phlegm")
\end{enumerate}

Wikipedia : smooth luffa = Luffa aegyptiaca (sponge gourd)\\
Sharma (Syn. Kulaka. Well known (Trichosanthas dioica Roxb.)

Brahmānanda glosses it as \emph{kośātakī} (Meul p. 586 LUFFA ACUTANGULA ROXB), which suggests he thought it was some sort of luffa.

Brahmānanda also mentions the vernacular term \emph{paravara} for \emph{paṭola}, which the Lonavla ed. states is a kind of cucumber. However, Paras remarked that \emph{paravara} is more like a gourd (hard shell, etc.).

On \emph{pañcaśāka}, see GhS  
\startverse
bālaśākaṃ kālaśākaṃ tathā paṭolapatrakam |\\
pañcaśākaṃ praśaṃsīyād vāstūkaṃ hilamocikāṃ || 5.20 ||
\endverse

HTK 4.26
\startverse
pañcaśākastu –\\
kṣīraparṇī ca jīvantī matsyākṣī ca punarnavā \\
meghanādaś ceti budhaiḥ pañcaśākaḥ prakīrtitaḥ || iti || 26
\endverse

Jyotsnā
\startverse
sarvaśākam acākṣuṣyaṃ cākṣuṣyaṃ śākapañcakam |\\
jīvantī-vāstu-matsyākṣī-meghanāda- punarnavāḥ || iti ||
\endverse

It is not entirely clear how we should understand \emph{divya}. Brahmānanda glosses it a \emph{nirdoṣa} and takes it with \emph{udaka}. But could it refer more specifically to \emph{gaṅgāmbu} (as suggested by Paras) or rain water? MW has \emph{divyodaka} n. `divine water' i.e. rainwater L.

The term \emph{divyodaka} appears in Āyurvedic works (but we’re yet to find a gloss in a commentary). E.g.,
Aṣṭāṅgahṛdaya 8.42–43
\startverse
śīlayec chāligodhūmayavaṣaṣṭikajāṅgalam |\\
suniṣaṇṇakajīvantībālamūlakavāstukam || 42 ||\\
pathyāmalakamṛdvīkāpaṭolīmudgaśarkarāḥ |\\
ghṛtadivyodakakṣīrakṣaudradāḍimasaindhavam || 43 ||
\endverse

SriKanta Murty translates \emph{divyodaka} as ‘divyodaka (rain water or pure water)'.

The Rājanighaṇṭu says rainwater:
\startverse
divyodakaṃ kharāri syād ākāśasalilaṃ tathā /\\
vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam // \textup{Rajni 14.4}
\endverse

Kharāri? Maybe \emph{khavāri} was intended.

Vācaspatyam:
\textbf{divyodaka} na° karma°.
1 antarīkṣabhave jale divyaśabde bhāva° pra° vākye tadbhedādi dṛśyam.
ambuśabde vivṛtiḥ .
\end{philcomm}

\subsection*{1.63}
\begin{translation}[hp01_063]
The yogi should eat food that is pure (\emph{mṛṣṭa}, or delicious \emph{miṣṭa}), unctuous, and containing diary. It should nourish the bodily constituents, satiate the mind and be fit [for yoga].
\end{translation}

\begin{philcomm}[hp01_063]
The variants of 1.63a all seem possible: \emph{mṛṣṭaṃ}, \emph{miṣṭaṃ} and \emph{iṣṭaṃ}. Maybe the last is made redundant by \emph{mano 'bhilaṣitaṃ}.
\end{philcomm}

\subsection*{1.64}
\begin{translation}[hp01_064]
Whether young, old, very old, sick or even weak, the diligent yogi succeeds in all yogas through practice. 
\end{translation}

\begin{philcomm}[hp01_064]
Note the different reading in 164d for V1: \emph{sarvaṃ yogī yatendriyaḥ}. Sarvaṃ is not easy to construe, and the testimony of the DYŚ suggests that \emph{yoge sarvo ’py atandritaḥ} was original.
\end{philcomm}

\subsection*{1.65}
\begin{translation}[hp01_065]
The postures, various breath retentions and techniques, beginning with seals, are all to be done in the practice of Haṭha until the goal of Rājayoga [is attained].
\end{translation}

\begin{philcomm}[hp01_065]
\emph{sarvaṇy api} is better than \emph{sarvo pi} ca because it refers to all the practices mentioned in the first hemistich.

Reading of Brahmānanda is different for the third pāda: \emph{divyāni karaṇāni}.
\end{philcomm}

\end{ekdosis}
\end{document}

### 2022-06-07 compile-tidyup.sh

@@ -2,7 +2,7 @@
 
 # optional argument 1 is filename, otherwise all relevant *.tex files
 
-xslcmd="java -jar /usr/share/java/saxon/saxon9he.jar -s:xml/HP_skm_test-tei.xml -xsl:xslt/hp-online.xsl"
+xslcmd="java -jar /usr/share/java/saxon/saxon9he.jar -s:xml/HP1_OldMss-tei.xml -xsl:xslt/hp-online.xsl"
 
 xslcmdproc(){
     ${xslcmd} | \
@@ -39,12 +39,13 @@ compile(){
 }
 
 
-if [ -f "${1}" ]
+if [ -f "latex/${1}" ]
 then
+    rm "latex/${1%.*}.pdf"
     compile "${1#*/}"
 elif [ -z "${1:-}" ]
 then
-    for f in "HP_skm_test.tex" "HP1_OldMss.tex" "HP1-TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex"
+    for f in "HP1_OldMss.tex" "HP1_TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex"
     do
 	compile "${f}"
     done

### 2022-08-23 html/switch.js
script toggle integrated

@@ -1,0 +1,69 @@
+var dev = true;
+var ltn = false;
+
+document.addEventListener("DOMContentLoaded", () => {
+
+    var devbutton = document.getElementById("switchdev");
+    var ltnbutton = document.getElementById("switchltn");
+    var alternchar = document.getElementById("altern");
+
+    setbuttons();
+    
+    devbutton.addEventListener('click', () => {
+	if ( dev && !ltn ) {
+	    
+	} else if ( dev ) {
+	    hide("versdev");
+	    dev=false;
+	    setbuttons();
+	} else {
+	    show("versdev");
+	    dev=true;
+	    setbuttons();
+	};
+    });
+
+    ltnbutton.addEventListener('click', () => {
+	if ( ltn && !dev ) {
+	    
+	} else if ( ltn ) {
+	    hide("versltn");
+	    ltn=false;
+	    setbuttons();
+	} else {
+	    show("versltn");
+	    ltn=true;
+	    setbuttons();
+	};
+    });
+
+    function setbuttons() {
+	if ( dev && ltn ) { alternchar.textContent="&";}
+	else { alternchar.textContent="/";};
+	
+	if ( dev ) {devbutton.className="toggleon"; }
+	else { devbutton.className="toggleoff"; };
+	
+	if ( ltn ) { ltnbutton.className="toggleon"; }
+	else { ltnbutton.className="toggleoff"; }
+    }
+
+    function hide(scriptclass) {
+	let class_a = document.getElementsByClassName(scriptclass);
+	
+	for (var i = class_a.length -1 ; i >= 0; i--) {
+	    let div=class_a[i]
+	    div.style.display="none";
+	} 
+    }
+
+    function show(scriptclass) {
+	let class_a = document.getElementsByClassName(scriptclass);
+
+	for (var i = class_a.length -1 ; i >= 0; i--) {
+	    let div=class_a[i]
+	    div.style.display="block";
+	    
+	}
+    }
+});

### 2022-09-18 marma/Marmasthāna.tex
added routine for checking uncollated mss for each marmasthAna

\input{Preamble-Print}
\makepagestyle{HPed}
\makeoddhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
\makeevenhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}

\newcommand{\myfn}[1]{\footnote{\texteng{#1}}}
\setlength{\footmarkwidth}{1.3em}
\setlength{\footmarksep}{0em}
\setlength{\footparindent}{1.3em}
\footmarkstyle{\textsuperscript{#1} }
\usepackage{fnpos}
\makeFNbottom

% additions/changes 2022-03-12 mm:
\SetTEIxmlExport{autopar=false}
\NewDocumentEnvironment{marma}{O{}}{}{}
\def\sthana{}

\begin{document}
\pagestyle{HPed} %
\begin{ekdosis}
 \SetLineation{lineation = none,}

 \chapter{Variants and Versions}

\sthana{24a}

 \begin{marma}[hp01_024]
        \begin{description}
%        \item[(ku(r)k(k)uṭāsana-)]{}
        \item[-vat kṛtvā]{V1, N22}
        \item[-vat kratvā]{P6}
        \item[-bandhastho]{A1, J8, L1, M1, V6, B1, B2 (in marg.), B3, Bo2, C1, C2, C3, C4
           (\emph{madhyastho} is also given above), C6, C8, J1, J3, J6, J13, J14, K1, Ko, LD1, LD2, N1, V2, N2,
           V3 (about \emph{stho}), N4a, V8, N5, N8, N9, N11, V21, N13, N16, N18, N19, N20, N21, N24,
           N25, N26, P1, P2, P3, P4 (\emph{... sanaṃ}), P7, Tue,V17, V14, R3, R5, V13, V10}
        \item[-bandhasthe]{J7, V19}
        \item[-badhascho]{Ba, Bo3, V28, V22}
        \item[-baṃdhascho]{J5, J11, J15, V15}
        \item[-baddhastho]{N12}
        \item[-bandhastu]{Bo1}
        \item[-madhyastho]{C9, J10, J12, J16 (\emph{baṃ} is given in margin), J17, N17, V4, N6, N7,
           N10, V20, V18, P5, P8, V12, LD1}
        \item[-madhyasthā]{P9}
        \item[-madhyasthau]{V16}
        \item[-baṃdhasthau]{N23}
        \item[-badhestu]{J2}
        \item[(unavailable/illegible)]{Ba5, C7, J4 (\emph{gudaṃ nibadhya gulphābhyāṃ}), C5, R2}
        \end{description}
\end{marma}

\begin{marma}[hp01_027]

 \sthana{27d}

 \begin{description}
\item[jaṭharapradīptaṃ]        V6 (jaraṭha), B1, B2, B3, Bo1, C1, J13("pradīpta"), N1, V22, N13, N21 (No anusvara over "ṭha" of "pīṭha"), P2, R3(...pīṭha)
\item[jaṭharapradīpti]        Ba5, V10
\item[jaṭharapradiptaṃ]        P4(...piṭhaṃ)
\item[jaṭharaprakṛṣtaṃ]        V14
\item[jaṭharapradīpaṃ]        P6("piṭhaṃ" changed to "pīṭhaṃ", "dipaṃ" changed to "dīpaṃ")
\item[jaṭharapradīptiṃ]        Bo3, Ko, N4a, Tue, R5
\item[jaṭharapracaṇḍaṃ]        K1, V1, N22, V15
\item[jaṭharaṃ pracaṇḍaṃ]        J11("jvalanaṃ pradīptaṃ" is also given above),
\item[jaṭharapracaṃḍa]        J5
\item[matsyendra-vīra-jaṭhara-pracaṇḍaṃ]              J4, N19 (V1 (above) also has "jaṭhara-pracaṇḍaṃ") [vīra-ja is unmetrical?]
\item[budhaṃ pracaṇḍa]        V8
\item[jaṭharapravṛddha]        J10, J17, V2, N8, N11
\item[jaṭharapravṛddhau]        J15
\item[jaṭharapravṛddhaṃ]        N24
\item[jaṭharapravṛddhiṃ]        C1
\item[jaṭharaprabuddhaṃ]        C8, J14, N2, N12, V21, N23, V12
\item[jaṭharaprabuddha]        Ba,
\item[jaṭharaprabuddhau]        C4 ("jvalanapradīptam" is also given above), L1, V19, V28, N5, N25, P3
\item[jaṭharaprabodhaṃ]        A1, C6, P7,
\item[...bodhaṃ]           R2(only "bodhaṃ" available)
\item[jaṭharāprabodhaṃ]        V20
\item[jaṭharaṃ prabadhaṃ]        J2
\item[jvalanapradīptaṃ]        LD1, Bo2, C3, C9, J1, J3, J6, J8, N17, V3, V4, N6 ("na" is written as a later addition), N9, N18 (No anusvāra visible on "ṭha"), V18, N26, P1, P5,V17, P8('pradī' is correction at bottom of folio), V13, V16, P9
\item[jvalanaṃ pradīptaṃ]        J7, J12, LD2, N10, N16, N20(It is "di", short "i". Also, "ṭhaṃ" is not there.)
\item[t(?)valanaṃ pradiptaṃ]        J16,
\item[jvalatpradīptaṃ]        N7
\item[(unavailable/illegible)]        C5, C7, M1
        \end{description}
\end{marma}

\begin{marma}[hp01_027]

  \sthana{27d}

        \begin{description}
        \item[candrasthiratvaṃ ca dadāti] Ba5, J7, J10ac, J17, LD1, LD2, V6, B2, Bo1, Bo3, N4a, V4, V22, N10,
         N13, P1, Tue, P8, R5, V13, V10
         \item[candraṃ sthiratvaṃ ca dadāti] P9("ca" in margin)
        \item[candre sthiratvaṃ ca dadāti] C9,
        \item[candrasthiratvaṃ hi dadāti] B1, C2, N1, R3
        \item[cādrasthiratvaṃ hi dadāti] B3
        \item[candrasthiratvaṃ pradadāti] C1, N24
        \item[caṇḍasthiratvaṃ ca(?) dadāti] N7
        \item[daṇḍasthiratvaṃ ca dadāti] A1, Ba, Bo2, C3, C6, C8, J1, J2, J8, J10pc, J11, J12, J14,
         J15, J16("ḍaṃ" is given in margin), K1, Ko, M1, N17, V1, V2, V3, N6, N9, N12, V20, N16, N18,
         N19, N20, N22, N26, P6, V17, V12, R2
        \item[daṇḍāsthiratvaṃ ca dadāti] P5
        \item[daṇḍāsthitatvaṃ ca dadāmi] V14
        \item[daṇḍaṃ sthiratvaṃ ca dadāti] P7
        \item[daṇḍasthiratvaṃ hi dadāti] J13, P2 (anusvāra over "da" is missing)
        \item[daṇḍasthiratvaṃ hi dadātti] P4
        \item[daṇḍasthiratāṃ ca dadāti] V21
        \item[daṃḍaschiratvaṃ dadāti] J5,
        \item[daṇḍasthiratāṃ sa dadāti] J3, J6,
        \item[daṇḍasthiratvaṃ vidadhāti] C4, L1, V28, V18, N25, V16
        \item[daṇḍasthiratvāṃ vidadhāti] P3
        \item[daṇḍasthiratvaṃ pradadāti] N8, N11, V19, N23
        \item[daṃḍe sthiratvaṃ pradadāti] N5
        \item[daṇḍasthiratvaṃ jvalanaṃ] V8
        \item[(daṇḍa?)sthirasvaṃ ce dadāti] N2
        \item[dehasthiratvaṃ ca dadāti] N21
        \item[kāyasthiratvaṃ ca dadāti] V15
        \item[kaḍasthiratvaṃ ca dadāti(?)] J4
        \item[(unavailable/illegible)] C5, C7
        \end{description}
\end{marma}

\begin{marma}[hp01_028]
\sthana{1.28b}
  \begin{description}
  \item[dorbhyāṃ padāgradvitayaṃ] A1, Bo1, C2, C6, C8, J4, V1, M1, B2, B3, N1, N4a, V21, V20, V22,
        V28, N16, N19, N21, N22, N23, N24, P2, P6, R5, V13, R2
  \item[dorbhyāṃ padāgredvitayaṃ] V14
  \item[dorbhyāṃ padāgradvītayaṃ] P7
  \item[dorbhyāṃ padāgnadvittayaṃ] P4 (could be "padāgra")
  \item[dorbhyāṃ padāgraṃ dvitīyaṃ] Ko, V15
  \item[dorbhyāṃ padāgradvitīyaṃ] R3
\item[dorbhyāṃ padagradvitayaṃ] Bo3, Ba5, K1, N13, Tue ("pa" uncertain), V10
\item[dorbhyāṃ padāgryau dvitayaṃ] J5
\item[dorbhyāṃ pādāgrādvitayaṃ] B1
\item[dorbhyāṃ ca pādadvitayaṃ] C4, L1, V19, N5, N7, N8, N11, N12, P3
\item[dorbhyāṃ ca pādaṃ dvitayaṃ] C1
\item[dorbhyāṃ ca pādau dvitayaṃ] N25
\item[dorbhyāṃ ca pādāgradvitayaṃ] J13
\item[dorbhyāṃ sa pādāgradvayaṃ] J2
\item[dvābhyāṃ karābhyāṃ dvitayaṃ] Bo2, C3, C9, J1, J3, J6, J7, J8, J10, J15, J16, J17, LD1, LD2, N17, V6, V4, V3, N6, N9, N10, N18, N20, V18, N26, P1, P5, V17, V16, LD1, P9("karābhyāṃ" in margin)
\item[dvābhyāṃ karābhyāṃ hitayaṃ] J12, V12
\item[dvābhyāṃ padāgraṃ dvitayaṃ] J11
\item[dvābhyāṃ pādāgrādvitīyaṃ] P8
\item[(unavailable/illegible)] Ba, C7
\item[dorbhyāṃ padagraṃthidvayaṃ] J14, N2
\item[.. rbhyāṃ pādaṃbhyāṃ dvitayaṃ] V8
\item[(unavailable/illegible)] C5,
        \end{description}
\end{marma}

\begin{marma}[hp01_035]
\sthana{35b}
  \begin{description}
\item[asyahṛdaye dhṛtvā]        J7("eva niyataṃ" in below margin), J12, A1, V4, V18, V16, P8
\item[asya hṛdayaṃ dhṛtvā]   P9
\item[āsyahṛdaye dhṛtvā]        N10, V1, P1, P6
\item[āsyahridayaṃ dhṛtvā]        J16
\item[eva hṛdaye dhṛtvā] C4 ('vidadhe' is also given above), C7, C8, J5, J10, J11, J15, J17, Ko,
 V19, N2, N5, V21, N25, P3, V15, V14, R8,V12
\item[eva hṛdaye kṛtvā]        Bo3, K1, N4a, V22, N13, N19, N24 (seems later addition), Tue, V10
\item[eva niyataṃ dhṛtvā]        C1, C3, J14, V8, V28
\item[eva niyataṃ kṛtvā] B1, B3, Bo2, C2, C6, J1, J3, J6("sudṛdha" is also given above), J8, J13,
 L1, N17, LD1, N1, V3, N12, V20, N16 (It seems "meḍhra" here), N18, N20, N21, N26, P2, P4, P7, R3,
 R5('hṛdaye' canceled and replaced by 'niyataṃ')
\item[ekahṛdaye dhṛtvā]        J4, N3, N23
\item[ekahṛdayo dhṛtvā]        J2, M1
\item[eva sudṛḍhaṃ kṛtvā]        N8, N11
\item[vāmaniyataṃ dhṛtvā]        V6
\item[samyak(ū?) hṛdaye dhṛtvā]        N9    
\item[āsyahṛdaye kṛtvā]        N22, V17
\item[asya hṛdayeḥ dhṛtvā]        C9 ("āsyaṃ hṛdaye yasya saḥ" is given above)
\item[meḍhre pādam atho hṛdi svacibukaṃ]        N7
\item[meṃḍhraṃ pādam athaikam]                Bo1, LD1, LD2
\item[meḍhre pādam athaikm eva hṛdaye]        Ba5, N7
\item[meṃḍhreṃ pādam athaikam]                LD2
\item[meḍhraṃ pādam athaika samyak meva hṛdaye dhṛtvā]               P5
\item[(unavailable/illegible)]        Ba, C5, V13, R2
        \end{description}
\end{marma}

\begin{marma}[hp01_038]
\sthana{38a}
  \begin{description}
  \item[yameṣv eva mitāhāra(ḥ)] C2, J10, J17, M1, V1, V6, P8
  \item[yameṣv eva sitāhāra] V16
  \item[yameṣv eva mitāhāro] N7
  \item[yameṣv eva mitāhāram] C4, B2, J14, N2, N5
  \item[yameṣv eva ātmacetana] V28
  \item[yameṣv iva mitāhāra(ḥ)] B1, B3, Bo1, Bo2, C3, C9, J3, J6, J8, J13, J15, K1, Ko, LD1, N17, N1,
        V3, V4, N9, N10, N14, N16, N18, N21, N26, P1 (seems to be "mitāhārāḥ"), P2, P4, R3(after
        correction), V12, LD1, P9
  \item[jameṣṭiva mitāhāram] J5
  \item[yameṣv iva mitāhāro] V18
  \item[yameṣv iva mitāhāre] P5
  \item[yameṣv iva mītāhāra] J16,
  \item[yameṣv iva mitāhāram] A1, Bo3, Ba5, C1, C7, C8, J4, J11, L1, LD2, V19, N4a, V22, V21, N13, N23,
        N24('va' of 'iva' not clear), N25, P3, P7("mitah.." earlier), Tue, V15, R5, V13, V10
  \item[yamiṣviva mitāhāram] V14
  \item[yameṣv iva mitāhā(ḥ)] V8
  \item[niyameṣu mitāhāro] N8, N11, N12
  \item[niyameṣu mitāhārā] J12,
  \item[(unavailable/illegible)] Ba, V20, R2
  \item[yameśiva mitāhāram] N19
  \item[yameṣv eva mitāhāra] N20, V17
  \item[yameṣu ca nirāhārī] N22, P6
  \item[yameṣu ca mitāhārm] C6,
  \item[yamaṣviva mitāharm] J1
  \item[niyameṣu mitāhari] J2
  \item[yameṣv api mitahāraḥ] J7("yamesṣv iva" is deleted)
   \item[(unavailable/illegible)] C5,
        \end{description}
\end{marma}

\sthana{1.38b}
  \begin{description}
\item[ahiṃ niyameṣv iva]        B1
\item[ahiṃsā niyameṣu ca]        J4, M1, V1, N22, P6
\item[ahiṃsā niyameṣv api]        B2, J14, N2, N24, V14
\item[ahiṃsā niyameṣv iva] A1, B3, Ba5, Bo1, Bo2, Bo3, C2, C3, C4, C7, C8, C9, J1, J3, J5, J6, J7, J8,
 J10, J11, J13, J15, J17, Ko, LD1, LD2, N17, V3, V4, N5 (there is an extra mātrā after "me" and makes it
 look like "mo"), N9, V21, N13, N14, N16, N19, N20, N21, V18, N25, N26, P1 (was "ivā", the long
 mātrā seems canceled), P2, P4, P7, V16, P8, R3, R5, V13, V12, LD1, P9
\item[ahiṃsā niyāmaṣv iva]        P5
\item[ahiṃsā niyameṣu eva]        V17
\item[āhiṃsā niyameṣv iva]        Ko,
\item[ahiṃsā nīyame sthitā]        K1,
\item[ahisā nīyameṣv iva]        J16,
\item[aṃhiṃsā niyamaṣv iva]        J12,
\item[svāhiṃsā niyameṣv iva]        V6
\item[ahiṃsāṃ niyameṣv iva] L1, V19, N4a, V22, P3, Tue, V15, V10
\item[ahiṃsāṃ niyameṣu ca]        C6,
\item[ahiṃsā niyaṃmai(?)ṣv iva]         N1
\item[(a)hiṃsā ca niyameṣv iva]        N7("a" could be supplied due to its potential lopa due to sandhi)
\item[yathāhiṃsā yameṣv iva]        N8, N11
\item[ahiṃsā niyameṣv iha]        N10
\item[asiṃsā niyame sthata(ḥ)]               V8
\item[yameṣv evātma-cintanaṃ]        N12
\item[na hiṃsāṃ niyameṣv iva]        N23
\item[kṣamasye ātmacintane]        J2
\item[(unavailable/illegible)]        Ba, C5, V28, V20, N18 (only "ahisā" is visible), R2
        \end{description}
\end{marma}

\begin{marma}[hp01_039]
\sthana{39d}           
          \begin{description}
(ṇ/n and ś/s irrelevant)
\item[suṣumṇā iva nāḍikā]        V1, N22 (anusvāra over "nā"), P6
\item[suṣumṇā iva nāḍiṣu]         B2, J4, J14, K1, Ko, N2, V8, N19 (seems "suṣumṇa", short "a"), V12
\item[suṣumṇā ca nāḍiṣu]        J5,
\item[suṣumṇā eva nāḍiṣu ]        V20
\item[suṣumṇā iva nāḍiṣuḥ]        N8
\item[suṣumiva nāḍiṣu]         V28
\item[suṣumṇām eva nāḍiṣu]        J10ac, J17, N24, V14
\item[suṣumṇām iva nāḍiṣu ]        A1, C1, C4 ("nāḍīnāṃ malaśodhane" is given above), C6, C7, C8, C9 (in marg.), J1, J2, J3, J6, J11, J15 (suṣūmṇām), L1, M1, V19, N5, N7, N12, N16, N23 (sukhumṇām), N25, V15, V13
\item[suṣuṇmām iva nāḍiṣu]        P7
\item[suṣumṇām iva ḍinoṣu]        P3
\item[nāḍīnāṃ malaśodhanam]        B1, B3, Ba5, Bo1 (it seems 'मळ'), Bo3, C2, C3, C9, J8, J12, J13, LD2, N1, N4a, N10, N13, N18 (It is "ḍi" here. Short "i"), N20, N21 ("nāṃ" seems written differently), N26, V3, V22, P1 (was "śodhane", mātrā of "e" was canceled and anusvāra was added), P2, P4, Tue("naḍī.." earlier. Mark over "na" probably intends to indicate it as "nā"), V17, R3, R5, V10
\item[nāḍināṃ malasodhanaṃ]        J16,
\item[nāḍīnāṃ malaśodhanā]   P9
\item[nāḍīnāṃ malaśodhane]        J7, J10pc, LD1, N9, N17, V6, V21, N14(seems "dī"), V18, P5, V16, P8, V12, LD1
\item[suṣumṇādiṣu śodhanam ]         V4
\item(unavailable/illegible)]        Ba, Bo2, C5, R2
        \end{description}
\end{marma}

\begin{marma}[hp01_040]
\sthana{1.40e}
  \begin{description}
\item[śramādau bahubhiḥ pīṭhaiḥ] V1
\item[śramadair bahubhiḥ pīṭhaiḥ] B1, B3, Bo1, Bo2, C1, C2, C6, C8, J3, J6, J13, J14, K1, M1,
 N1(not sure if it is 'śra'), N12, N16, N21("daiḥ" instead of "dair"), N22, N23, N24, P2, P4,
 P7(bbahu), V14, R3, V13
\item[śramadair bahubhiḥ pīṭhai]        P6("piṭhai" changed to "pīṭhai")
\item[śramadai bahubhiḥ pīṭhaiḥ]        Ba, C9mg, J5,
\item[śramadair bahubhi pīṭhai]        N8
\item[śramādaiḥ bahubhi pīṭhaiḥ]        V21
\item[śramadairghyādibhiḥ pīṭhaiḥ]        C4("kim ādyair bahubhiḥ pīṭhai" is given above), C7, L1, V19, N5, N25, P3
\item[śramadairdhyādibhiḥ pīṭhaiḥ]        V28
\item[śamardda bahubhiḥ pīṭhaiḥ]        J1
\item[sanastai bahubhyaḥ pīṭai]        V8
\item[(samastair?) bahubhiḥ pīṭhaiḥ]        N2
\item[samastair bahubhiḥ pīṭhaiḥ]        J2, J4, J11, Ko,
\item[samasta bahubhiḥ pīṭhai]        V15
\item[kim ādyair bahubhiḥ pīṭhaiḥ] A1, C9("śramādai bahubhiḥ pīthaiḥ" is also given above), J7, J8,
 J10, J17, LD2, N17, V6, V3, V4, N7, N14(visarga at the end not visible due to bad picture), V20,
 N18(No visarga at the end), N20, V18, N26, P5, V17, P8, V12, LD1, R2, P9("rba" in margin)
\item[kim ādyair bahubhiḥ pīṭheḥ]        N9, V16
\item[kim adyair bahubhiḥ pīṭhaiḥ]        N10(N10 has this pāda as the 1st pāda, not the 5th)
\item[kim ādyair bahubhiḥ piṭhaiḥ]        J15, LD1, P1
\item[kim anyair bahubhiḥ pīṭhaiḥ]        Ba5, C3, N4a, N13, Tue, R5, V10
\item[kim anyair bahuniḥ pīṭhaḥ]        Bo3, V22
\item[kim arthai bahubhiḥ pīṭhai]        J12, J16("adyair" is also given),
\item[kiṃ śramair bahubhiḥ pīṭhaiḥ]        B2
\item[pramadair bahubhiḥ pīṭheḥ]        N19
\item[(unavailable/illegible)] C5,
        \end{description}

\sthana{1.40f}
  \begin{description}
\item[sadā siddhāsanasthiteḥ]        A1, V20, V12, R2
\item[sadā siddhāsane sati] B1, B3, Bo1, C2, C9("kiṃ sadā sevanena tu" is also given above), J7,
 J8, J10, J11("siddhaḥ" is given above), J12, J15, J17, LD1, LD2, N17, V1, V6, N1('kiṃ' and 'śa' seem faded
 and thus, I guess, deleted.), V4, N7, N9, N10(N10 has this pāda as the 2nd pāda, not the 6th),
 N20("sa" of "sati" is not clearly visible), N22, V18, N26, P1, P2, P4, P5, P6, V17, V16, P8, LD1, P9
\item[sidhe siddhāsane sati]        V10
\item[sadā siddhāsane satī]        J16,
\item[sadā siddhāsane satya]                  V3
\item[sadā siddhāsanaṃ(?)]        V8
\item[sadā siddhāsanā nyāsāt]        V22
\item[sadā siddhāsanābhyām sādyogī]        V14
\item[yadā siddhāsane sati]        J2, J4, N2
\item[yadā siṃhāsane sati]        Ko,
\item[siddhe siddhāsane sati]        Ba5, Bo3, C3, N4a, N13, Tue
\item[siddhe siddhāsane miti]        R5('mi' seems tampered with)
\item[kiṃ sadā siddhasādhanaiḥ]        M1
\item[kiṃ syāt siddhāsane sati] Bo2, C1, C4("sadā siddhāsane sati" is also given above), C7, J1,
 J3, J5, J6, J13, L1, V19, B2, V28, N5, N16, N21("sīddh..." long"i"), N25, P3, V15, R3
\item[kiṃ syāt siddhāsane mati(ṃ?)]        N8
\item[kiṃ syāt siddhāsane sthite]        C6, C8, V21
\item[kiṃ syā siddhāsane sthite ]        P7
\item[kiṃ syāt siddhāsana sthite ]        N23
\item[kiṃ syāt siddhāsane schite]        Ba
\item[kiṃ vā siddhāsane sati]        J14, N12, N24
\item[kiṃ syāt siddhāsanai(?)ti]        N18
\item[kiṃ sadā sevanena tu]         K1, N19(1.40h is "baddhe siddhāsane sadā")
\item[(unavailable/illegible/unable to read)]        C5, N14, V13
        \end{description}
\end{marma}

\begin{marma}[hp01_046]
\sthana{1.46cd}
\item[(utta(m)bhya cibukaṃ ~)]
\item[vakṣa sthāpayet ]        J8, J10, J17, LD2, N18, V6, N7, P8, N18, V12, V3, P9("thā" unclear)
\item[vakṣaḥ sthāpayet]         Bo1, B2, C3, J1, J12, J15, J16(sthāpaye),  K1, LD1, V18, P1, P5, V16, N10, N22(vakṣyaḥ), LD1
\item[vandā sthāpayet]        V17
\item[vakṣa sthāpayat]         N17
\item[vakṣe saṃsthāpayat(?)]        Ko,
\item[vakṣe uthāpayet]        V4
\item[vakṣa sthāpya?]        J7(A word is deleted there), C9, N5(sthāpya sevanaṃ)
\item[vatsa sthāpayet ]        N20, N26
\item[vakṣaṃ sthāpayet]         V1
\item[vakṣasy āsthāpya]         C4, L1, M1, V19, P3
\item[vakṣasy utthāya]        Bo2, C6, N16, N8, N11, N16
\item[vakṣasy utthāpya] B5, Bo3, C1, J13, J14, N13, N21, N23, N24(unclear if it is "syu"), Tue,
 N1(unclear), N4a, R5
\item[vakṣasy utṭhāpyo]        J4, J11,
\item[vakṣasy utṭhāgre]        J6
\item[vakṣasy oschāpyo]        J5
\item[vakṣyasy utthāpyot patane...]        N19
\item[vakṣyasy asthāpya]        N25
\item[vakṣyapyutchāpya]        Ba
\item[vakṣe sthāpayet]        N14 (seems "vacche"), P2, P4, N9
\item[vakṣe syu sthāpya pavanaṃ]         V10
\item[vakṣye sthāpayet]        C2
\item[vakṣye pyutthāya]        C8
\item[vakṣyesu uttānya]        V14
\item[vikṣe sthāpayet]        R3
\item[cakṣasya utthāpya]        V15
\item[svakṣasy utthāpya]        P7
\item[urghātopaticotanau karau kṛtvā tato dṛśaṃ?]        J2
\item[dakṣasy utthāgret(?)]        J3
\item[utkuṃcya cibukaṃ vakṣa sthāpayet ]           P6
\item[sannāya cibukaṃ vakṣyasy utthāpya]          N2
\item[yuktaṃ vakṣastha-cibukaṃ mākramet]        N12
\item[(unavailable/illegible/unable to read)]        C5, V13, R2
        \end{description}
\end{marma}

\begin{marma}[hp01_048]
\sthana{1.48b}
  \begin{description}
\item[(gāḍhaṃ vakṣasi sannidhāya cibukaṃ ~)]
\item[dhyānaṃ tataś cetasi]        J2, V1
\item[dhyānaṃ tataś cepsitaṃ]        J15, C9
\item[dhyānaṃ tataś cośitaṃ]        V6
\item[dhyānaṃ tataś conmitaṃ]        P5
\item[dhyānaṃ tataḥ conmī]        N1(some syllables seem deleted, 'nmī' is added later)
\item[dhyānaṃ ca tac cetasi] B2, B3, Ba, Bo1, Bo2, C1, C2, C4, C6, C7, C8, J4, J5(caitasi), J8,
 J11, J12, J13, K1, Ko, L1, V19, N25, N26, P2, P3 ("cubukaṃ"), P4, P6, P7(cetasī), V4(ṃ omitted),
 V17, V16, N3, N5, N8, N11, V15, V14, P8, R3, V13, N14, N16, N17,N19, V12, V3, N21, N23, N24(saṃniyamya), R2, P9
\item[dhyānaṃ ca tecesi]        V8
\item[dhyānaṃ ca tac cepsitaṃ] A1, J1, J3, J6, J10, J14, J16, J17, LD1, P1, N7(later attempt to form 'dhyā'), N9, N10(saṃvidhāya), N12, N18(dhānaṃ), LD1, N20, N22
\item[dhyānaṃ ca yac cepsitaṃ]        C3,
\item[dhyānaṃ ca tac cepsite]        J7,
\item[dhyānaṃ ca tatrepsitaṃ]        LD2, N2
\item[dhyāyaṃś ca tac cetasi]        Ba5, Bo3, M1, Tue, N4a, N13, V10, V22
\item[dhyāyaṃś ca taṃ cepsitaṃ]        R5(something is canceled and 'psitaṃ' is added)
\item[(unavailable/illegible)]   C5,
        \end{description}

\sthana{1.48c}
 \begin{description}
\item[(vāraṃ vāram apānam ūrdhvam anilaṃ ~)]
\item[protsālayan pūrayan]        V1
\item[protsārayan pūrayan]        J7, V6, N7, N12
\item[protsārayaḥ pūrayem]        V8
\item[protsārayan pūritaṃ]        A1, Ba5, Bo3, C1, C6, C8, J10, Tue, N4a, V14, R5, V13, N13, V10, V22, R2("n pū" not legible due to damage)
\item[protsārayat pūritaṃ]        P7
\item[protsyā?rayat pūritaṃ]        Ba
\item[protsārayan pūrayet]        C4("proccālayan pūritaṃ" is also given above),L1, N25, P3
\item[protsārayet pūrayan]        V19
\item[protsārayet pūrayet]        C7
\item[protsāraṇaṃ pūrayet]        N5
\item[protthāya yat pūrayan]        B3
\item[protthāpanṃ pūrayet]        Bo1,
\item[protthāpayan pūrayan]        C2, P2, P4, N1, R3, P9(not sure of "tthā")
\item[procārayet pūritaṃ]        J8
\item[prottāsayan pūritaṃ]        V12
\item[prollāsayan pūritaṃ]        J10, J17, N17, P1, P5
\item[prollāśayet pūritaṃ]        J15, N9
\item[prolāsayet pūrayan]        J12,
\item[prollāsayet pūrayan]        N10
\item[prollosayana pūrayan]   LD2
\item[prollāsayet pūrayet]        J16,
\item[prollāsayan pūrayan]        C9, V16, P8, V4
\item[pro(llāpa?)yan pūrayan] LD1
\item[prollāsat pūritaṃ]        C3,
\item[prollāghayan pūrayan]        LD1
\item[procāraranu pūritaṃ]        J5
\item[proccārayan pūritaṃ]        M1, V17, N3(vālaṃ vāram)
\item[proccārayan pūrayan]        J4,
\item[proccārayat puraya yataṃ(?)]        J1,
\item[proccārayat pūritaṃ]        N2
\item[proccārayet pūritaṃ]        V15, V3, N23
\item[proccārayet pūrayet]        N19
\item[proccālayan]        K1
\item[proccālayan pūrayen]        B2
\item[proccālayan pūrayet]        J13,
\item[proccālayaṃn pūrayan]        P6, N22(vvāraṃ cāram apānamūlapranilaṃ), \item[proccālayan pūrayan]        N18, N24(ayāmūrdh...)
\item[proccālayan pūritaṃ]        J2, J14, Ko, N26, N20(proccālayaṃn)
\item[proccālayet pūritaṃ]        Bo2, J3, J6("proccāṭayan pūrayet" is also given below), N16
\item[proccālitaṃ pūritaṃ]        N14(vālaṃ vālam)
\item[proccānalaṃ pūrayen]        N21
\item[prodbhā(ṭa?)yan pūrayet]        N8
\item[prod(ghā?)ṭayan pūrayet]        N11

\sthana{1.48d}
 \begin{description}
\item[(muñcan prāṇam upaiti bodham atulaṃ ~)]
\item[śaktiprabhāvān naraḥ] B3, Ba5, Bo2, Bo3, C1, C2, C6, C8, J1, J3, J6, J8, J11, J13, J14, K1, Ko,
 M1, V1, P2, P4, Tue, N2, N3, V15, V14, R3, R5, V13, N13, N16('vā' at bottom of page), N19, V10, V22, N23
\item[śaktiprabhāvān nara]        V3
\item[śaktiprabhāvā naraḥ]        V8, N22(mucan...)
\item[śaktiḥ prabhāvān naraḥ]        Ba,
\item[śaktiṃ prabhāvān naraḥ]        Bo1,
\item[śaktiṃ svabhāvān naraḥ]   N21
\item[saktiprabhāvān]        J5
\item[śaktiprabhāvon naraḥ]        N1, N7
\item[śaktiprabhāvīn naraḥ]        P6
\item[śaktiprabhāvaṃ naraḥ]        J12, N4a
\item[śaktiprabhān naraḥ]        B2
\item[śaktiprabhāvāṃtaraḥ]        N18
\item[śaktiprabhāvodayāt] C4, C7, L1, V19, N25 (1st part of pāda: prāṇaṃ muñcyati yāti bo...), V16,
 N5, N14(sakti)('da' seems later correction)
\item[śaktiḥ prasādān naraḥ]        J2,
\item[śaktimabhāvodayāt]        P3 ("ma" could well be "pra)
\item[śaktiḥ prabhāvodayāt]        A1, R2("vo" illegible due to damage)
\item[śaktiprabodhodayāt]        N20
\item[śaktiprabhāvodayaḥ]        J7, V6
\item[śaktiprabhāvād ataḥ]        C3, C9, J10, J16, P5, P7, N9, N10, N11, N12, P8, V12, LD1, N24, P9
\item[śaktīprabhāvad ataḥ]        LD2
\item[śaktiprabhāvad ata]        J15, N8, V4
\item[śaktiprabhāvataḥ]        J17
\item[śaktiprabhāvadhataḥ]        V17
\item[śaktiprabodhodayāt]        N17, N26
\item[śaktiprabhāvodhayaḥ]        P1
\item[bhakti prabhāvodayaḥ]        J4,
\item[(unavailable/illegible)]   C5,

        \end{description}
\end{marma}

\begin{marma}[hp01_049]
\sthana{1.49c}
 \begin{description}
\item[mārutaṃ nayate yas tu]        C6, C8,  
\item[mārutaṃ niyato yas tu]        Ba,
\item[mārutaṃ niyatam yaskta]        V14
\item[mārutaṃ niyatam yas tu]        V8, N23(mār(ū?)taṃ)
\item[mārutaṃ nīyate yas tu]        C1,
\item[mārutaṃ pīyate yas tu]        V1
\item[mārutaṃ pīvyate yas tu]        J8ac, V3
\item[mārutaṃ yaḥ pīvet nitya]        V13
\item[mārutaṃ pibati yas tu]        J8pc
\item[mārutaṃ pibate yas tu]        J11,
\item[mārutaṃ yas tu pibati] A1, C9, J7, J10, J12, J15, J17, LD2, N17, V6, N26, P1, P5, P6, P7, V17, V16,
 N7, N9, N10, P8, N14, V4, LD1, N20(piba?ti), R2(earlier reading cancelled and overwritten), P9
\item[māruto yas tu pibati]        N22
\item[mārutaṃ yaska pibati]        V12
\item[māruta yas tu pibati]         R1
\item[mārutaṃ yas tu pibaṃti]        C3,
\item[mārutaṃ yas tu pībaṃti]        J16,
\item[mārutaṃ dhārayet yas tu] Ba5, Bo2, Bo3, C2, C7, J1, J3, J6, J13, L1, M1, V19, N25 ("dhārayed"), P2
 ("dhārayed"), P3("), P4("), Tue("), N4a("), N11("), N12("), R5("), N13("), N16("), V10, N24("dhārayed")
\item[māruto dhārayed yas tu]        N1, R3, V22
\item[māruta dhāraye?as tu]        N21
\item[mārutaṃ dhāraye yas tu]        N8, N18
\item[marutaṃ dhārayed yas tu]        C4, N5, V15
\item[mārutaṃ māryatd yas tu]        J5
\item[mārutaṃ mriyate yas tu]        N3
\item[mārutaṃ mriyate yasmāt]        Ko,
\item[māruto mriyate yas tu]        J4
\item[māruto bhidyate yasya]        B2, Bo1, J14, K1, N2, N19
\item[mahato dhārayed yastu]        B3
\item[pavanṃ prāsayeta saṃstu]        J2
\item[(unavailable/illegible)]    C5, 
        \end{description}
\end{marma}

\begin{marma}[hp01_051]
\sthana{1.51d}
 \begin{description}
\item[nāsāgre nyastalocanaḥ]        B3, C2, C9, J7, J13, J16, N17, V1, V6, P2, P4, N1, N7, R3, N22
\item[nāsāgre nyastalocanā]        P6
\item[nāsāgra-nyastalocanaḥ]        A1, Bo1, C3, J10, J17, N26, P1, P5, V16, N10, V12, V4, LD1, LD2, N20, R2
\item[nāsāgraṃ nyastalocanaḥ]  P9
\item[nāsāgranyastalocanaṃ]        J8
\item[nāsāgraṃ nyastalocanaṃ]        P7, V17, V3
\item[nāsāgra susamāhitaḥ]        C8,
\item[nāsāgraṃ susamāhitaḥ] Bo2, Bo3, C1, C4("nāsāgra nyastalocanaḥ" is also given), J1, J2, J3, J6,
 J11, J14, J15, K1, Ko, L1, M1, V19, N25, P3, Tue, N2, N4a, N5, N12, V15, V14, V13, N13, N16, V10,
 V8
\item[nāsāgrastu samāhitaḥ]   N23
\item[nāśāgre susamāhitaḥ]        N19,
\item[nāsyagraṃ tu samāhitaḥ]        J5,
\item[nāsāgraṃ susamāhitaṃ]        Ba5, C6, V22
\item[nāsāgryaṃ susamāhitaṃ]        N18
\item[nāsāgraṃ susamāhita]        N8
\item[nāsāṃgraṃ susamāhitaḥ]        N11
\item[nāśāgraṃ tu samāhitaḥ]        N3,
\item[nāsāgraṃ tu samāhitaḥ]   N24
\item[nāsāgre susamāhitaḥ]        B2, Ba, K1, N21
\item[nāsāyāṃ susamāhitaḥ]        J4,
\item[nāsāṃ tu samāhitaḥ]        R5
\item[nāsāgre neṣu nocana]        N14
\item[(unavailable)]        C5, C7,
        \end{description}
\end{marma}

\begin{marma}[hp01_052]
\sthana{1.52d}
 \begin{description}
 \item[kurute cāsanottamam] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J8,
        J10pc, J13, J15, L1, N17, V6, N26, P2, P3, P4, P5, P7, Tue, N1, V17, N4a, N5, N8, N11(anusvāra
        over 'cā' seems canceled), N12, V15, R3(visarga is also there), R5, V13, N13, N14, N18, N19,
        V8, V3, N20, N21, N23, R2
\item[kuruta cāsanottamam]        V22
\item[kurute cāsanomatam]        V10
\item[kurute cāsanottame]        N16
\item[kurute cāsanottamaḥ]        J2,
\item[kurute cāsanotraye]        V14
\item[kurute cāsanottama]        A1,
\item[kurute cāsanontayam]   N24
\item[kurute vāsanottamam]        V1, V19, N3
\item[kurute sādhanottamam]        C3, J12, J10ac, J16, J17, P1, V16, N9, N10, V12, V4, P9
\item[kurute sādhanottaramaṃ]        P8(perhaps 'ra' is attempted to be canceled)
\item[kurute sādhakottamam]        C9, N7,
\item[kurute sādhakottamaḥ]        LD1, LD2
\item[kurute cāsanam uttamaṃ]         N25
\item[sevate cāsanottamaiḥ]        J14,
\item[sevate cāsanottamaṃ]        J11, Ko(casano°), N2
\item[sevyate cāsanottamaṃ]        J4,
\item[(unavailable)]        B2, C5, C7, K1, M1, P6, N22
        \end{description}
\end{marma}

\begin{marma}[hp01_054]
\sthana{1.54b}
 \begin{description}
 \item[sarvavyādhi-viṣāpaham] C3, C9, J5, J7, J8, J10, J12, J15, J16, J17, LD1, V1pc, V6ac, P1,
        P5(sarvva), P6, P7, V17, N3, V16, N9, N10, P8, R1, V12, V3, V4, N22, P9
 \item[sarvavyādhī-viṣāpaham] LD2
 \item[sarvavyādhi-vināśanam] A1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4("viṣāpahaṃ" is also given above), C6, C7, C8, J1, J2, J3, J4, J6, J11, J13, K1, Ko, L1, N17, V1ac, V6pc, V19, N25, N26(Seems ...sanam), P2, P3, P4, Tue, N1, N4a, N5, N7, N11, N12, V15, V14, R3(seems an extra mātrā after 'vyā'), R5, N13, N14(sarvva...sanaṃ), N16, N18(sanaṃ), N19(anusvāra seems erased due to damage), V10, V8, V22, N23, N24(sanaṃ), R2
\item[sarvavyāvināśanam]   N21
\item[sarvadhyādhi-vighātakaṃ]   N20
\item[sarvavyādhi-praṇāśanaṃ]        J14, N2
\item[kaphavātāmayāpahaṃ]        M1
\item[(unclear/unable to read)]        C5, N8(last two syllables seem erased), V13
        \end{description}
\end{marma}

\begin{marma}[hp01_055]
\sthana{1.55 additional verses}

kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet |
na śāstrapāṭhamātreṇa yogasiddhiḥ prajāyate ||
na veṣadhāraṇaṃ siddheḥ kāraṇam na ca tatkathā |
kriyaiva kāraṇaṃ siddheḥ satyam etan na saṃśayaḥ |
śiśnodararatāyeha na deyo veṣadhāriṇaḥ ||

\begin{description}
\item[Yes] B1, B2, B3, Ba, Bo1, Bo2, C1, C3, C4, C6, C8, C9, J1, J3, J6, J7(58 num.), J8, J10, J11,
 J12, J13, J14, J15, J16, J17, K1, L1, LD1, LD2, N5, N17, N25(the first 3 lines only?), V3, V17, V25, P8,
 R1('śiśno....' not there), R3, N14, N16, N18, V12, LD1, N7, N9, N10, N13(appears way later, after the verse "yuvā vṛddho...yogeṣvatandritaḥ", and last line not there), N20, N21, N26, P1, P3, P4, P5, P7, R2, P9
\item[Yes], but the 4th line appears as the 2nd line: C2, J13, N1, P2
\item[No] A1, Ba5, Bo3, C1, C5, C7, J2, J4, J5, Ko, M1, V1, V6, V8, V19, R5, N19(mudrādipavanakriyām also
 not there), V10, V22, V4 (unclear/unable to read) V13, N3, N4a, N8, N11, N12, N22, N23, N24, Tue, P6
        \end{description}

\sthana{1.55 third additional verse}

mayi bodhāmbudhau svacche tuccho 'yaṃ viśvabudbudaḥ |
pralīna udito veti vikalpapaṭalaḥ kutaḥ ||

\begin{description}
\item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, V6, V19,
  R3, N14(bodhibudhau, malīna), N16, N18, N1 (word "tathā" before "pralīna", also some letters before "svacche", perhaps "ttu", seem faded and not taken into consideration), N8, N11, P2, P3, P4, N13(at the end of 1st chapter)
\item[No] A1, B2, Ba5, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, LD1, LD2, M1,
  V1, V3, V8, V17, V25, P8, R1, R5, N19, V12, V10, V22, V4, LD1 (unclear/unable to read) V13, N2, N3, N4a, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, Tue, P1, P5, P6, P7, R2, P9
        \end{description}

\sthana{1.55 fourth additional verse}
% P: What is the difference between this and the previous extra verse?

mayi bodhāmbudhau svacche tuccho 'yaṃ viśvabudbudaḥ |
pralīna udito veti vikalpapaṭalaḥ kutaḥ ||

\begin{description}
\item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, V6, V19, R3,
 N14(bodhibudhau, malīna), N16, N18, N8, N11, P2, P3, P4
\item[No] A1, B2, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, M1,
 V1, V3, V8, V17, V25, P8, R1, R5, N13, N19, V12, V10, V22, V4, LD1 (unclear/unable to read) V13, N2, N3, N4a, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, Tue, P1, P5, P6, P7, R2, P9
        \end{description}

\begin{marma}[hp01_056]
\sthana{1.56a}
 \begin{description}
 \item[āsanaṃ kumbhakaś citro] J8pc, N24
 \item[āsanaṃ kumbhakaś citraṃ] B2, C3, J7, J8ac, J10, J12, J15, N17, P8, N14(ā(?)naṃ), V12, LD1, LD2, N4a, N10, N12, N20, P1, P5
 \item[āsanaṃ kumbhavaś citra] J2,
\item[āsanaṃ kumbhavaś ci(nt?)aṃ] P9
 \item[āsana kumbhakaś citraṃ] V3
\item[āsanaṃ kumbhakāś citrāḥ]  N8, N11
\item[āsanaṃ kumbhakāś citrā]   P7
\item[āsanaṃ kumbhakācitraṃ]   Tue
 \item[āsanaṃ kumbhakaṃś citraṃ] C9, J17
 \item[āsanaṃ kumbhakaṃ citraṃ] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C7, C8, J1, J3, J4, J5, J6, J11, J13, J14(55), K1, Ko, L1, M1, V19, R3, R5, N13, N16, N19, V10, V22, V4, N1, N5, N23, N25, N26, P2, P3, P4, P6, R2
\item[āsanaṃ kumbhakaṃ śvitraṃ]   N21   
 \item[āsanaṃ kumbhaka citrāṃ] J16, N7
 \item[āsanaṃ kumbhaka citaṃ] N18
\item[āsanaṃ kubhakaṃ citraṃ] N2 (appears before kriyāyuktasya....)
\item[asanaṃ kumbhakaṃ citraṃ] N3, N9
 \item[(unclear/unable to read)] V13
\item[(unavailable)]        C5, N22, R1
        \end{description}

\sthana{1.56b}
 \begin{description}
 \item[mudrādikaraṇāni ca] A1, B2, C3, C9, J7, J8pc, J10, J12, J15, J16, J17, LD1, N17, P8, N14, V12, N7, N9, N10, P1("kā" changed to "ka" later), P5, P7, R2, P9
 \item[mudrānikaraṇāni ca] LD2
\item[mudrādikāraṇāni ca]        V4
\item[vicitrīkaraṇāni ca]        J2,
\item[mudrādi pavanakriyā]        B1, B3, Bo1, C2, J5, J8ac, J13, R3, V3, N1, N20, P2, P4
\item[mudrādi pavanakriyāḥ]   N26
\item[mudrākhyaṃ karaṇaṃ tathā] Ba, Ba5, Bo2, Bo3, C1, C3 ("mudrādikāraṇāni ca" is also given), C6, C7,
 C8, J1, J3, J11, L1, M1, V19, R5, N13, N16, N18, N19, V10, V22, N5, N11, N12, N21, N23, N25, Tue, P3, P6
\item[mudrākhaṃ karaṇaṃ tathā]   N24
\item[mudrākhyaṃ karaṇaṃ bhavet]   N3
\item[mudrākhyakaraṇaṃ tathā]        Ko, N4a, N8
\item[mudrādi karaṇaṃ tathā]        K1,
\item[mudrākṣaṃ karaṇaṃ tathā]        J4, J6,
\item[mudrakhecaraṇīṃ tathā]        J14(given at 55)
\item[mudrakhecariṇīṃ tathā]        N2 (appears before kriyāyuktasya....)
\item[(unclear/unable to read)]        V13
\item[(unavailable)]         C5, N22, R1
        \end{description}
\end{marma}

\begin{marma}[hp01_060]
\sthana{1.60cd}
 \begin{description}
\item[(atilavaṇaṃ tilapinḍaṃ kadaśanaśākotkaṭaṃ varjyam |)]
\item[atilavaṇaṃ tilapinḍaṃ]        J17, LD2, N10, R1
\item[atilavaṇaṃ tilapinḍa]        C3, J8pc, J10, J15, J16, LD1, N17, V4, N9, N20, P1, P5, P8, P9
\item[atilavaṇaṃ tilapiṇḍā?]   N14
\item[atilavaṇaṃ tilaṃ pinḍa]        J8ac
\item[atilavaṇaṃ tilaṃ pinḍaṃ]         V3
\item[atilavaṇaṃ tilapiṇyākadarśanam]        N26
\item[atilavaṇam amlayuktaṃ]        A1, B3, Ba, Ba5, Bo3, V22, V10, N1, N4a, N11 (original letters "...sacapalaṃ" striked off and "...amlayuktaṃ"), N13 mentioned in bottom margin), Tue, P2("amla" in margin), R2, R5(some doubt in "amla")
\item[atilavaṇama malayuktaṃ]        C2,
\item[atilavaṇamalayuktaṃ]   P4, R3
\item[atilavaṇāmlayuktaṃ]   P7
\item[atilavaṇa amlayuktaṃ]        C6, C8, V13
\item[atilavaṇa amlayutaṃ]        J11,
\item[atilavatilapi]        V6
\item[atilavaṇatiktapiṇḍa]        B2
\item[atilavaṇatilapiṇḍakaṃ]        C9
\item[atilavaṇaṃ dyuṣṇaṃ]        V1
\item[atilavaṇam apramuṣṇaṃ]        M1
\item[atilavaṇam ca apramitaṃ]        V8
\item[atilavaṇam cāpramitaṃ]   P6
\item[atilavaṇaṃ ca tilapiḍaṃ]   N7
\item[atilavaṇam pramitaṃ]        J12, K1,
\item[atilavaṇādi prasaktaṃ]        J14,
\item[atilavaṇādi proktaṃ]   N22
\item[atilavaṇacāmlapiktaṃ]        N19
\item[atilavaṇakavalṃ cāpamitaṃ]        J1,
\item[atilavaṇasavapalala]        V19
\item[atilavaṇasavapalalaṃ]   N25
\item[atilavaṇaṃ sala(pa?)laṃ]   N8
\item[atilavaṇaṃ palilaṃ]         R4
\item[atilavaṇāsavapalalaṃ]        C1, C7, L1, N5
\item[atilavaṇātiproktaṃ]        J2,
\item[atilavaṇādiyutaṃ]        J4,
\item[atilavaṇādyaptataptaṃ]        J5,
\item[atilavaṇādyapran(t?)aṃ]  N3
\item[atilavaṇādyapratapta]   N12
\item[atilavanā?āna(va?)lala]   N24
\item[atilavaṇamatprayuktaṃ]   N21
\item[atilavaṇāmavapalalaṃ]        C4 ("atilavaṇaṃ tilapinḍaṃ" is also given),
\item[atilavaṇāya priyatavṃ]        Ko
\item[atilavaṇaṃ sapalaṃ ca]        Bo2, J3, J6,
\item[atilavaṇasalilaṃ kadaṣana]        B1, Bo1, J13,
\item[atilavaṇetilapiṃḍakvadaśanaṃ]        J7
\item[atilavaṇaprasūtaṃ]         N2
\item[atilavaṇaṃ kavalaṃ cāpramitaṃ piṇḍa]   N18
\item[atīlavaṇāmlapṛktaṃ]   N23
\item[(illegible/unavailable)]  C5, N16, P3

* * *

\item[śākotkaṭaṃ] A1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C3, C4, C6, C8, C9, J2, J3, J4, J6, J10, J11, J12, J13, J14, J15, J17, K1, Ko, L1, LD1, LD2, V6, V19, V22, V4, V10, V13, N2, N3, N4a, N7, N11, N13, N18, N19, N21, N24, N25, Tue, P2, P5, P6, P7, P8, R1, R2, R3, R4, R5, P9
\item[śākotkaṭhaṃ]        V8
\item[śākotkavaṃ]        J5,
\item[śākotkaṭa]        Bo2, J8pc, J16, N1, N9, N26(??kotkaṭa)
\item[śākotkarṭeṃ]  P4
\item[śākātkaṭa]        V3
\item[śākotchadaṃ]        Ba,
\item[śākolkaṭaṃ]   P1
\item[śākokṣadaṃ]   N23
\item[śāṃkotkaṭaṃ]   N10
\item[śāketkaṭaṃ]        C7
\item[śokātkaṭa]        J8ac, V1
\item[śokotkaṭaṃ]        B1, J7,
\item[śāśotka]        M1
\item[śātkṭaṃ]        J1,
\item[śa?tāsanaḥ]   N14
\item[sākotkaṭa]   N20
\item[sākotkaṭaṃ]  N22
\item[syacotkaṭaṃ] N12
\item[(illegible/unavailable)]        C5, N17, N8, N16, P3
\item[kotkaṭaṃ]  N5

* * *
\item[varjyaṃ] A1, B1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C4, C7, C9, J3, J6, J13, J15, K1, L1, LD2, V3, V6, V19,
 V22, N1, N4a, N5, N11, N12(va?jyaṃ), N13, N22, N25, Tue, P2, P4, P6, P7, R3, R4, R5
\item[varjya] V8, V10
\item[varjyāṃ] J16
\item[varj(j)aṃ] J7, J10, J12, J17, M1, N9, R1
\item[varjjyaṃ]  P5
\item[varj(j)itaṃ] J1, Bo2, J8, LD1, V3, N18, N20, N26
\item[varyyaṃ]   N7, R2, P9
\item[varjjāṃ]   N19
\item[varjet]  N10
\item[varjayet] C3, P1("varjaṃ" earlier)
\item[vayīṃ]   P8
\item[duṣṭaṃ] C6, C8, J4, J5, J11, J14, Ko, V13, N2, N21, N23, N24
\item[du(ṣl?)aṃ]  N3
\item[dṛṣṭaḥ]
\item[durjvaṃ(?)] Ba,
\item[(illegible/unavailable)] C5, N17, V1, V4, N8, N14, N16, P3
        \end{description}
\end{marma}

\begin{marma}[hp01_062]
\sthana{1.62d}
 \begin{description}
 \item[mudgādi divyam udakaṃ ca] A1, B2(dīvyam), Ba, Ba5, Bo3, C6("dgā" is missing), C8, J5, J7, J10, J11, J12, J16, J17, K1, L1pc, LD1, LD2, N17, V1, V6, V8, V22, V4, V13, N1 ("mudgā" correction in folio margin), N10, N12, N13, N14(not sure for "dgā" and "ṃca"), N20, N22, N23(something after "mudgādi" but perhaps it is cancelled), N24(is it "mudrā"?), N26, Tue, P1, P6, P8, R1, R2, P9
\item[mudgādi divyam udakaṃ]   N18
\item[mudgādi divyam udakaṃ ka]   N4a
\item[mudgādi divyam udake ca]   R5
\item[mudgādi divyam udakaṃ yamīndra]        V3, N2
\item[mudgādi divyam udakaṃ hi]        C3, J14, J15, N9, P5
\item[mudgādi pivyam udakaṃ ca]        J2,
\item[mudgājya divyam udakaṃ ca]        C9
\item[mudgādivyam udakaṃ ca]        C4, J8, M1, V10, N3
\item[mudgādi divyagaditaṃ ca]   N7
\item[mudgādi divyamuktaṃ]        Ko,
\item[mudgādi cālpam udakaṃ ca]        B1, B3, Bo1, Bo2, C1, C2, C7, V19, N5, N11 ("ca" in margin), N21, P2(not sure of "cā"), P4, R3, R4
\item[mudgādi cālyam udakaṃ ca]        L1ac, N8(no "ca")
\item[mudgādi cājyam udakaṃ ca]        N16
\item[mudgādi cālpa sudakaṃ ca]         J1
\item[mudgādi dālpam udakaṃ ca]        J3, J6("cālpa" is also given)] J13,
\item[mudgādi dālpamudgaṃ ca]        N25
\item[mudrādi divyam udakaṃ ca]  P7
\item[mudrādi divyam udakaṃ ca mudakaṃ]        N19
\item[(illegible/unavailable)]       C5, J4, P3
        \end{description}

\sthana{1.62d}
 \begin{description}
 \item[munīndrapathyaṃ (or: munindra-)] B1, B2, B3, Bo1, C1, C2, C4, C7, J3, J5, J6, J13, J16, J17, K1, L1, LD2, V1, V19, N1, N5, N11, N19, N21, N22, P2, P4, R4
\item[munīdrapathyaṃ]   N8, R3
\item[munīndrapathaṃ]   N25
\item[mahendrapathyaṃ]  N9, P5
 \item[yamīndrapathyaṃ (or: yamindra-)] A1, Ba, Ba5, Bo3, C6, C8, C9, J7, J8, J10, J11, J14, LD1, M1, N17, V6, V22, V3, V4 (pathyaṃ missing), V10, V13, N2, N3, N4a, N12, N13, N14, N20, N26, Tue, P1, P6, P7, P8, R2, R5, P9
\item[yamīṃdrapraparyya]   N24
\item[yatīṃdrapathyaṃ]        J2, N10, N23
\item[yamendrapathyaṃ]        C3,
\item[yogīdrapathya]        J1,         
\item[mūnidra pathayam]        V8
\item[munidra pathayam]        J12,
\item[munīdramukhyaiḥ]        N7
\item[muṃhanīdrapathya]        J15
\item[suyogipathyaṃ]         N16
\item[varayogipathyaṃ]        N18
\item[jamodūpathyaṃ]        R1
\item[paramaṃ hitaṃ hi]        Ko,
\item[(illegible/unavailable)]   C5, J4, P3
        \end{description}
\end{marma}

\begin{marma}[hp01_063]
\sthana{1.63a}
 \begin{description}
\item[iṣṭaṃ sumadhuraṃ]        A1, J3, J6, J10, N18(appears after one verse), P7, R2, P9
\item[iṣṭaṃ samadhuraṃ]        C9, J7, J17, LD2, N17, V19, V4, LD1, P1, P8, R1
\item[iṣṭaṃ saṃmadhuraṃ]   N7
\item[daṣṭaṃ samudhūraṃ]        J16,
\item[dṛṣṭaṃ samūdhuraṃ]        K1
\item[dṛṣṭaṃ sumūdhuraṃ]        N10
\item[piṣṭaṃ sumadhuraṃ]        C8, N8, N11, N14("śu")
\item[pathyaṃ sumadhuraṃ]        J12,
\item[puṣṭaṃ sumadhuraṃ]        Ba5, Bo3, N4a, N13, Tue
\item[puṣṭaṃ samadhuraṃ]   R5
\item[yeṣṭaṃ sumadhura]        V13
\item[yuṣṭaṃ sumadhuraṃ]        V10(mu°), V22
\item[mṛṣṭaṃ sumadhuraṃ]        M1, V1
\item[mṛṣṭaṃ sumadhurāṃ]        Ko
\item[mṛṣṭaṃ sumadhusnigdhaṃ]   N3
\item[mṛṣṭaṃ samadhuraṃ]        C3,
\item[miṣṭaṃ samadhuraṃ]        J8, Ba, C4, N20(appears after one verse), N26(appears after one verse), P5
\item[miṣṭaṃ sumadhuraṃ]        Bo2, C6, J1, J11, J14, N2, N16(appears after one verse), N19(miṣtaṃ??), N21(appears after one verse), N22, N23(appears after one verse), R4(appears after one verse)
\item[miṣṭaṃ samudhuraṃ]        J5, J15,[%P: How is this reading different from the reading above the reading above this?]
\item[miṣṭaṃ sūmadhuraṃ]  P6(appears after one verse)
\item[miṣṭaṃ sumadhura]   N25
\item[miṣṭaṃ madhuraṃ]        V3
\item[miṣṭaṃ ca madhuraṃ]        B2
\item[maṃ(ṣṭ?)aṃ samudhuraṃ]  N9
\item[sumiṣṭaṃ madhuraṃ]        N12(appears after one verse)
\item[uṣṇaṃ sumadhuraṃ]        B1, Bo1, C2, C7, L1, J13, N1 (seems "uṣṇaṃ"), P2
\item[uṣṇaṃ samadhuraṃ]        C1,
\item[uṣma(?) sumadhuraṃ]        B3
\item[uṣmaṃ sumadhuraṃ]        P4, R3(appears after one verse)
\item[uśmā?mumadhuraṃ]  N5
\item[suvṛkṣaṃ madhuraṃ]        V6
\item[ghṛtam ca madhuraṃ]        V8
\item[vṛṣyaṃ sumadhuraṃ]        J2,
\item[tiṣṭaṃ tu madhuraṃ]  N24
\item[(illegible/unavailable)]   C5, J4, P3
        \end{description}
\end{marma}

\begin{marma}[hp02_003]
\sthana{2.3b}
\item[jīvanam] Ba, Ba5, Bo2, Bo3, C1, C6, J1, J2, J3, J6, J12, J13, J14, J16, N17, V1, N4a, N7, N8, N9, N10, N11, N12, N13, N16, V16, V17, V10, N18, N21, V21, Tue, N23, V6, V7, V20, V28, V18
\item[jīvaṃnam] N14
\item[jīvitam] A1, B1, B2, B3, Bo1, C2, C3, C4, C8, J4, J5, J7, J8, J10, J11, LD1, LD2, N1, N2, N5, V8, N20, N22, V14, V3, N24, N25, V15, V5, V22, V2, V12, V13, V9
\item[jivitam] J15,
\item[jīvita] N26
\item[jīvitum] J17, N19
\item[jītavim] N3
\item[jīvo na] Ko
\item[(unavailable/illegible)] C5, C9,

 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_003]
\sthana{2.3d}
\item[nibandhayet] Ba, Bo2, C6, V1, J2, J4, J6, J8, J11, J14, N12, N16, N19, N22, V14, V3, V13
\item[nirundhayet] A1, B1, B2, B3, C1, C3, C9, J5, J10, J13, J15, J17, LD1, LD2, N17, V19, W4, N2, N7, N9, N10, N14("ye" unclear),V16, V17, N20, V21, N23, N26, V20, V5, V2, V18, V12, V9
\item[nirodhayet] Ba5, Bo1, Bo3, C2, C4, C8, J1, J3, J6pc, J7, V19, N1, N4a, N5, N8(seems unclear), V8, N11, N13, V10, N18(“ye” is doubtful), N21, Tue, N24, N25, V6, V7, V15, V28, V22
\item[nisaṃdhayet] J16,
\item[nirodhanaṃ] J12,
\item[nibandhanāt] V19, N3
\item[nibadhnayāt] Ko,
\item[(unavailable/illegible)] C5,
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_004]
\sthana{2.4d}
\item[kāyasiddhiḥ] A1, B1, B2, C2, C6, J8, J14, V1, N1, N8, N11, V17, N24, V20, V15, V9
\item[kāyāsiddhiḥ] J2, N3,
\item[kāyāsuddhi] V6
\item[kayasiddhiḥ] N20
\item[kāyasiddhiḥ] J5,
\item[kāyaśuddhiḥ] N23
\item[kāryasiddhiḥ] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C8, C9, J3, J4, J6, J10, J11, J13, J17, Ko, LD1, N17, V1, W4, N2, N4a, N5, N9, N12, N13, V16, V10, N19, N21, Tue, N25, N26, V70, V5, V28, V22, V2, V18, V12, V13
\item[kāryasiddhiṃ] J15, 
\item[kāryāsiddhiḥ] J1
\item[kāryasiddhi] J12, LD2, N7, V8, N10, N14("rya" unclear), N16, N18, N22, V21, V3 
\item[kāryasiddhī] J16,
\item[kāryasuddhiḥ] V14
\item[kāyāśuddhiḥ] J7,
\item[(unavailable/illegible)] C5,

 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_006]
\sthana{2.6c}
\item[suṣumnāsu snighdhā] J4, V1, V15
\item[suṣumnānāḍisthā] Ba5, N13, Tue, 
\item[suṣumṇānāḍīsthā] N4a, N21(“ṇā” doubtful), 
\item[suṣumṇānāmadhysthā] V2
\item[suṣumṇānāḍīstho] N24
\item[suṣumnānāḍīstho] J14,
\item[suṣumṇānāḍīschā] Bo3, V28, V22
\item[suṣumṇāṃtasthhā] J15,
\item[suṣumnāṃtaraschā] J11
\item[sukhumnāṃtarasthā] V19
\item[sukhumnāṃtarālasthā] B1,
\item[suṣumnāṃtarālasthā] B3, Bo1, Bo2, C1, C2, J3, J6, J13, N1, N8, N11
\item[suṣumnāṃtarālasyā] J1
\item[suṣumnāṃtarājasthā] N16
\item[suṣumṇāṃtarasthā]  N5, C4
\item[suṣumṇāṃtarāla] V5
\item[suṣumnā pāśvaṃsthā] A1, V20
\item[yathā suṣumṇā pārśvaḥsthā] Ba
\item[yathā suṣumṇā pārśvasthā] C6, C8, J7, N23
\item[suṣumṇāpañcasthā] N19
\item[yathā suṣumnā mukhasthā] V17
\item[yathā suṣumnā nāḍisthamala] V10
\item[yathā suṣumnņayā śvaḥsthā] V8, V21
\item[suṣumņā sustabdhāḥ] N12
\item[yathā suṣumnā susvasthā][a] J5, N22, V14, V3, N25, N3
\item[suṣumņā suṣasthā] N20
\item[suṣumnņayā ścasyā][b] V6, V13
\item[suṣumnāvasthāyaṃ] B1, B2
\item[suṣumnānāḍibhyāṃ] V7
\item[suṣumnāsvasthāyan] N2
\item[suṣama susvasthā] J8
\item[suṣumnāsui svasthā] J8ac %H: I didn’t find this reading in ms. I don’t know who added this.
\item[suṣumnāsaṃsvasthā] J2,
\item[sukham avasthāya] C3, C9, J10, J12, J17, LD2, N17, W4, N7, N9, V16, V18, V12, V9
\item[sukham avasthāyaḥ] J16,
\item[sukham avasthāyā] N14
\item[sukham avasthāpyā] LD1
\item[sukham avaschāṃya] N10
\item[suṣam avasthāya] J8pc, N18,
\item[suṣumṇām avasthāpya] J7pc
\item[suṣumnāṃtvarayāpalaḥ] N25
\item[(unavailable/illegible)] C5, Ko,

 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_009]
\sthana{2.09 additional verses}
dvādaśamātraḥ kanīyānmadhyamo mātracaturvviśaḥ |
uttamaḥ ṣaṭtriṃśanmātro mātrābhedāḥ smṛtā tajñaiḥ |
vaśiṣṭasaṃhitādau tu pūrakaṃ ṣoḍaśamātrabhiḥ |
kuṃbhakaṃ ca catuṣṣaṣṭimātrābhiḥ recakaṃ tu |
dvātriṃśanmātrābhiḥ vidhivatkuṃbhakaṃ kṛtvā punaścaṃdreṇa recayet ||
\begin{description}
\item[Yes] J17
\item[No] A1, B1, B2, B3, Ba, Ba5, Bo1 Bo2, C1, C2, C3, C4, C5, C6, C8, C9, J1, J2, J3, J5, J10, J11, J12, J14, J15, J16, LD2, N17, V1, J4, J6, K7. J8, J13, Ko, LD1, W4, V19, N1, N2, N3, N4a, N5, N7, N8, N9, V28, V8, N10, N11, N12, N13, N14, N16, V16, V17, V10, N18, N19, N20, N21, N22, V14, V21, V3, Tue, N23, N24, N25, N26, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9
        \end{description}

\begin{marma}[hp02_011]
\sthana{2.11 additional verses}
iḍayā piva ṣoḍaśabhiḥ pavanaṃ |
kuru ṣaṣṭicatuṣṭayamaṃtagataṃ |
tyaja piṃgalayā sanakaiḥ sanakaiḥ |
daśabhirdaśabhirdaśabhirdvyadhikaiḥ ||
\begin{description}
\item[Yes] B1, B3, C9 (2.13), V19, N1 (2.13), N5, N10 (some change), N12, N21(some changes), N25, V28
\item[No] A1, B2, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C4, C5, C6, C8, J1, J2, J3, J10, J13, J17, J4, J5, J6, J7, J11, J12, J14, J15, J16, Ko, LD1, LD2, N17, V1, J8, W4, N2, N3, N4a, N7, N8, N9, V8, N11, N13, N14, N16, V16, V17, V10, N18, N19, N20, N22, V14, V21, V3, Tue, N23, N24, N26, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9
        \end{description}

\begin{marma}[hp02_012]
\sthana{2.12cd}
\item[uttānaṃ cottame prāṇarodhe padmāsane muhuḥ] Ba, C1, C6, V14
\item[uttānaṃ cottame prāṇe stebdhe? padmāsane muhuḥ] C8,
\item[uttame sthānam āpnoti tato vāyuṃ nibaṃdhayet] N13, V10, Tue, V7
\item[utiṣṭhe tūtame prāṇā yāvat padmāsane sūmaṃ] J2,
\item[uttiṣṭhaty uttame prāṇaṃ rodheḥ padmāsanaṃ .uhuḥ] V1
\item[uttiṣṭhāṃty uttame prāṃṇā baddhe padmāsane uttama] J16
\item[uttiṣṭhaty uttame prāṇē baddhe padmāsanae dṛḍhe] C2
\item[uttiṣṭhaty uttame prāṇaḥ baddhe padmāsanae dṛḍhe] J13,
\item[uttiṣṭhaty uttame prāṇaṃ baddhā padmāsane dṛḍhe] B1
\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane dṛḍhe] B2, B3,
\item[uttiṣṭhaṃty uttame prāṇā baddhe padmāsane muhuḥ] LD2, J10
\item[uttiṣṭhaṃty ūttame prāṇā baddhe padmāsane muhuḥ] J15,
\item[uttiṣṭhaṃty uttame prāṇā baddha padmāsane muhuḥ] C3, N10
\item[uttisthaṃ cottame prāṇā baddha padmāsane muhu] N14
\item[uttiṣṭhaty uttame prāṇa baddhe padmāsanaṃ muhuḥ] J8ac, J14, V3
\item[uttiṣṭhaty uttame prāṇa baddho padmāsane muhuḥ] V8
\item[uttiṣṭhatm uttame prāṇa baddhe padmāsane muhuḥ] V16, V17, V21, V20, V18, V13
\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane muhuḥ] C9, 
\item[uttiṣṭhaty uttame prāṇo baddha padmāsane muhuḥ] A1
\item[uttiṣṭhaty uttame prāṇo badhye padmāsane muhuḥ] N9
\item[uttiṣṭhaty uttame prāṇo baddhapadmāsane muhuḥ] N7
\item[uttiṣṭhaty uttame prāṇē baddhe padmāsanaṃ muhuḥ] J8pc
\item[uttiṣṭhaty uttame prāṇē baddha padmāsane haye] Bo1
\item[uttiṣṭhaṃty uttare prāṇā baddhe padmāsane muhuḥ] J17
\item[uttiṣṭhaṃty uttame prāṇā baddhe padmāsane muhuḥ] W4
\item[uttiṣṭhaty uttame prāṇā baddhe padmāsane muhuḥ] N20
\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane mukuḥ] N26
\item[uttiṣṭhaty uttame prāṇa rodhe padmāsana sthitaḥ] Bo2, J1, J3, N8, N11, N16, V2
\item[uttiṣṭhaty uttame prāṇa rodhe padmāsana sthite] C4, N25
\item[uttiṣṭhaty uttame prāṇāṃ rodhe padmāsane sthite] V19
\item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthite] V5, V28
\item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthitaḥ] J6, N2
\item[uttiṣṭaty uttame prāṇa rodhe padmāsanasthite]  N5
\item[uttiṣṭhaty uttame prāṇa rodhe padmāsanaṃ muhuḥ] N3, V15
\item[uttiṣṭhaty uttāme prāṇarodhe padmāsanaṃ muhuḥ] J5, J11,
\item[uttiṣṭhaty uttāme prāṇarodhe padmāsano muhuḥ] J7
\item[tattiṣṭhaty uttāmaprāṇarodhya padmāsanaṃ muhuḥ] J4
\item[uttiṣ(?)yaty uttame prāṇa rodho padmāsanaṃ muhuḥ] N24
\item[uttiṣṭhaty uttame kārye rodhe padmāsanae muhu] Ko,
\item[uttiṣṭhaṃty uttame prāṇa rodho padmāsane sthitaḥ] N18
\item[uṭṭiṣṭhaṃ cottare prāṇā baddhe padmāsane muhuḥ] N17
\item[uttiṣṭaty attame prāṇe baddhe padmāsane dṛḍhe] N1
\item[uttiṣṭhaty uttame prāṇe yāvat padmāsane samaḥ] N12
\item[uttiṣṭhaty uttare prāṇā baddhe padmāsane dṛḍhe] J12,
\item[uttiṣṭhaty unname prāṇe baddhe padmāsane dṛḍhe] N19
\item[uttiṣṭhaty uttame prāṇa baddha padmāsanaṃ dṛḍhaṃ] N21
\item[uttiṣṭety uname prāṇaḥ rodhe pavāsane sthite] N22
\item[uttiṣṭhaty uttame prāṇa rāvai padmāsano muhuḥ] N23
\item[uttiṣṭhaty uttame prāṇa rodhe padmāsano muhuḥ] V6
\item[uṭṭiṣṭhnty uttame prāṇā baddhe padmāsane muhuḥ] LD1,
\item[(illegible/unavailable)]   Ba5, Bo3, C5, N4a, V22, V12, V9
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_019]
\sthana{2.19c}
\item[kṛśatā] A1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C6, C8, J1, J4, J5, J6, J7, J11, J13, J14, V19, N1, N2, N4a, N5, N8, N11, N12, N13, N16, N18, N19, N22, C4, V17, V10, V21, Tue, N23, N26, V7, V20, V5, V28, V22, V2, V18
\item[kṛsatā] Ko, J3,  
\item[kṛśate] J10pc
\item[kriyate] C3, J10, J12, J16, J17, LD1, N7, N9, N10, V12, C9, J2, N17, LD2, V19
\item[kṛyate] J15, N14
\item[kṛśatāṃ] J8, V8, N21, V3, N25, V6
\item[kṛṣitāṃ] N24
\item[kṛtā] N3, N20, V14
\item[krūratā] J2
\item[(illegible/unavailable)]   B1, B2, C5, V16, V15, V13, V9
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_021]
\sthana{2.22a}
\item[medaḥśoṣādikaṃ pūrvaṃ] V1
\item[medaḥśleṣmādikaṃ surve] V6
\item[medaśokādikaṃ pūrvaṃ] N22
\item[medaḥśleṣmādi nāśārthe] A1, C3,
\item[medaḥśleṣmādi nāśārtha] LD2,
\item[medaḥśleṣmādi ṣatkarmāṅi] V28
\item[medaḥśleṣmādi nāśārthaṃ] B1, B3, Bo1, C2, J10, J12, J13, J15, J16, J17, LD1, N7, V21, V20, V12
\item[medaḥśleṣmādināśīrṣaṃ] N9
\item[medaḥ tuṣmād adhikei purva śaṭ karmaṃ] V8
\item[medaśleṣmādi nāśārthaṃ] C9, N1, N10, N14, N21, V18
\item[medaśleṣādi nāśārthe] V17
\item[medaśleṣmādi nāsārthe] N20, N26
\item[medaśleṣma nivṛtyarthaṃ] Bo2, C1, J1, J3, J6, N2, N8, N11, N16, N18, V15, V5, V2
\item[medaḥśleṣma nivṛtyarthaṃ] N5, C4
\item[??śleṣma nivṛtyarthaṃ] N25
\item[medaḥśleṣmādhikaḥ pūrvaṃ] Ba, C8, J5, V10, V14, V13
\item[medaḥśleṣmādhikaḥ pūrva] J7,
\item[medaḥśleṣmādhikaḥ arthaṃ] V16
\item[medai śleṣmāṃdhika purṇa] J2
\item[medaśleṣmāṃdhikaḥ pūrvaṃ] Ba5, Bo3, V3
\item[medaḥ ślṣāṃdhikaḥ pūrvaṃ] C6, J14, 
\item[medaśleṣmādhikaḥ pūrvaṃ]  J11, N3, N4a, N12, N13, Tue, V22
\item[medāśleṣmāṣikaḥ pūrvva] N23, V7
\item[medaḥśleṣmaṇisapurṇa] N24 
\item[mitaḥ sthūlādikaiḥ kāryyaṃ] B2
\item[medodhikastu havabhiḥ] Ko,
\item[(unavailable/illegible)] C5, J4, N19, V9
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_021d]
\sthana{2.22d}
\item[apy abhāvataḥ] V1
\item[api bhāvataḥ] C3
\item[samabhāvataḥ] A1, B1, Ba5, Bo1, Bo3, C9, J1, J3, J6, J10, J11, J12, J13, J14, J15, J17, LD1, LD2, N3, N4a, N7, N8, N10, N11, N12, N13, V10, N20, N21, V14,V3, Tue, N26(perhaps left margin says something), V7, V20, V15, V28, V22, V18, V12, V13
\item[samabhāvat] J16,
\item[samupāgataḥ] J7,
\item[samabhyāvataḥ] N24
\item[api samācaret] V16, V17, V21
\item[samabhāvata]  N9
\item[samabhāvanā] N22
\item[samabhāvanāt] N22
\item[smabhāvata]  N14
\item[sabhāgataḥ] Ba
\item[samabhāgataḥ] B2, C6, C8, V8, N18, N23, 
\item[samabhāgatā] Ko,
\item[samabhāgata] V6
\item[samabhāgikaḥ] B3, C2, N1
\item[samabhārat] J2
\item[samatā yataḥ]  Bo2, C1, C4, N2, N16, N25, V2
\item[samatā yathā]   N5,
\item[samatāpataḥ] V5
\item[(unavailable/illegible)] C5, J4, N19
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_022a-c]
\sthana{}
 \begin{description}
        \end{description}
\end{marma}

\begin{marma}[hp02_023c]
\sthana{2.22c}
\item[vicitraguṇasandhāyī(i)] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C3, C8, C9, J1, J2, J3, J4, J5, J6, J7, J11, J13, J14, J15, LD1, LD2, V1, W4, J10, N17, J17, J8ac, Ko, N1, N3, N4a, N7, N9, V8, N10, N16, V16, V17, V10, N18(...saṃdhāī, no “y”), N19(...saṃdhāi, no “y”), N21, N22(...saṃdhāī, no “y”), V14, V3, Tue, N23, V7, V15, V5, V28, V22, V2, V18, V12, V13
\item[vicītraguṇasandhāyi] J16,
\item[vicitraguṇasandhāya]  C6, N12
\item[vicitraguṇasandhāyā] N20, V20
\item[vicitragunasandhāyai]  N14, V21, V6
\item[vicitraguṇasandhāryā]  N13
\item[vicitraguṇasandhāryaṃ]  J12,
\item[vicitraguṇasandhānaṃ] N26
\item[vicitraguṇasadyāpi] N2
\item[vicitraguptasaṃ?jñopi] N5
\item[vicitraguṇasaṃjñopi] C4, N25
\item[vicitraguṇasaṃdhāyoga]  N8("vivi?"), N11 ("yogasiddhidāyi" in brackets, then "yi" of "sandhāyi" )
\item[trividhaṃ guṇasandhāyi] B2
\item[(unavailable/illegible)] C5, N24
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_24c]
\sthana{2.24c}
\item[tataḥ pratyaharec caitad] J12, J16, J17, N23
\item[punaḥ pratyahare caitad] J13, J14, 
\item[punaḥ pratyaharec caitaḥ] J15,
\item[(unavailable/illegible)]
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_24d]
\sthana{2.24d}
\item[prakṣālyaṃ dhautikarma tat] Ba, C6, V21
\item[prakṣālanaṃ dhautikarma tat] V13
\item[prakṣālye? dhautikarma tat] C8
\item[ādāyaṃ dhautikarma tat] V14
\item[udgāraṃ dhautikarma tat] V1,
\item[sodgāraṃ dhautikarmavit] N12
\item[uditaṃ dhautikarma tat] A1, Ba5, Bo1, Bo3, C9, J10pc, J12, J15, LD1, LD2, N4a, N7, N9, N10, N13, N14("ti" added later, anusvara on "ka" seems cancelled), V16, V17, V10, N20, N21(“u” of uditaṃ doubtful), Tue, N26, V7, V20, V15, V18, V12
\item[aditaṃ dhautikarma tat] J17,
\item[uditaṃ dhautikarma kṛt] J10, V22
\item[up(?)itaṃ dhautikarma tat] C3
\item[abhyāsād dhautikarmavit] B1, B2, C4, J1, J3, J6, J13, V5, V28, V2
\item[abhyāsād dhautikarmasu] N25
\item[abhyāsā dhautikarmavit] N18
\item[abhyāso dhautikarmavit] N19
\item[abhyāse dhautakarmavit] J14,
\item[abhyāsād dhotikarmavit] C1
\item[abhyāsād dhautikarmavat] N5
\item[abhyāsaud dhautikarmavit] Bo2, C2,
\item[abhyāsāsau dhautikarmavit] V6
\item[abhyāsau dhautikarmatat] V8
\item[abhyāsād dhautikarmavita] N2
\item[abhyāsād dhautikarma tat] N1 (after correction),
\item[dukhalātā dhautikarmataḥ] V3
\item[duḥkhalṃ dhautikarma tat] Ko,
\item[duṣālaṃ dhautikarma tat] J4, J11, 
\item[uṣālaṃ dhautikarma tat] J5,
\item[abhyāsa dhautakarka tat] N22
\item[abhyāsād dhauti bhavati karma tat] N8("dhīta" instead of "dhauti"), N11 ("bhavati" in brackets)
\item[uṃtthānaṃ dhautikarma tat] N23
\item[dākṣālaṃ dhautikarma tat] N24
\item[utthānaṃ dhautikarma tat] J7,
\item[dugdhāraṃ dhautikarma tat] J2
\item[syā sā dhautikarmakṛt?] B3
\item[?s?vāla dhautikarmakṛt] N3
\item[(unavailable/illegible)] C5, N16
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_25d]
\sthana{2.25d}
\item[chadhyate ca na] J16,
\item[dhāvaty eva] V1, V8, V3, V2
\item[dhāvate ca na] B2
\item[dhāvaṃta ca na] N19
\item[dhāvaṃte ca na] J5,
\item[dhavaṃte ca na] N22
\item[dhruvntye neva] Ko,
\item[gachaty eva na] N21
\item[gachaṃty eva na] Bo1
\item[prayāntyeva na] Ba, N4a (2.26a), Bo2, Bo3, C1, C4, C6, C8, J1, J3, J6, J7, N5, N8, N11, N13, N16, V10, Tue, N25, V28, V22
\item[prayānteva na] N23, V7
\item[śudhyatty eva] A1,
\item[śudhyaty eva] LD2,
\item[śudhyate nātra] J12,
\item[śudhyate ca na] N10
\item[śudhyanty eva] B1, C3, C9, J10, J13, J17, N1 (not sure for "eva")
\item[śudhyant eva] B3,
\item[śudhyaṃteva na] J15, N9, N14, V16, V17
\item[śudhyaṃte ca na] LD1, 
\item[suddhyaṃtevaṃ na] N18
\item[śudhyaṃte ca na] N7
\item[śudhyaṃ sevana] C2,
\item[kṣīyaṃte saklā malā] J14,
\item[?īyatsyeva na] N2
\item[bhavaṃty eva na] N3, V14
\item[naśyaṃteva] N12
\item[naśyaṃte] J11
\item[nāśaṃ yāṃti] J4, 
\item[naśyaṃty eva na] J2,
\item[jayaty eva na] N24
\item[śuṣyaṃte sakalāmalāḥ] N20, N26
\item[(illegible/unavailable)] Ba5,C5, V21, V6, V20, V15, V5, V18, V12, V13
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_26a]
\sthana{2.26a}
\item[udaragatapadārtham udvahatī] J12,
\item[udaragatapadārtham udvahantī] Ba, C3, J1, J7, Ko, V1, V16, N18, N22, N23, V6, V28, V2
\item[udaragatapadārthamūrdhāhatī] J16,
\item[uragatapadārtham udvahantī] B2,
\item[udaragatapadārthān udvahaṃti] C6
\item[udaragatapadārtham uşmahantī] V17
\item[udaragatapadārtham udvamaṃti(ī)] B1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J11, J14, LD1, LD2, N20, N21, V3, N26, V20, V22, V18, V12, V13
\item[udaragatapadārtham udvavaṃtī] J17,
\item[udaragatapadārtham udvavaṃtī] J10,
\item[udaragatapadārtham udvamaṃta] J4,
\item[udaragatapadārtham udvamaṃte] V15
\item[udaragatapadārthām udyamaṃte] N19
\item[udaragatapadārthaṃ śuddhapetaṃ] N25
\item[udaragatapadyarthān udvamaṃtī] A1
\item[udaragatapadārtham uḍhasaṃtī] N24
\item[udaragatapadārthaśodhanaṃ syāt] Bo1
\item[udaragatapadārtha śuddhayeta] C4
\item[udaragatapadārtha śuddhaye vai] J15, 
\item[udaragatapadārtha suddhayeta] J13,
\item[udagatapadārtha uddhanaṃtī] J5,
\item[(illegible/unavailable)] Bo3, C1, J2, V19, V10, V14, Tue, C5, V21, V7, V5, V22
 \begin{description}
        \end{description}
\end{marma}

\begin{marma}[hp02_26b]
\sthana{2.26b}
\item[pavanam apānam udīrya kaṇṭhanāle] V1, A1, B2, Ba, Ba5, C6, C8, C9, J3, J4, J6, J7, J10, J11, J14, J15, J17, Ko, LD1, LD2, N17, W4, J8, V16, N18, N20, N22, V3, N23, V20, V15, V28, V18, V12, V13 
\item[pavanam apānam udārya kaṇṭhanālo] J16,
\item[pavanām apānam udīrya kaṇṭhanālṃ] J1
\item[pavana apānam udīrya kaṇṭhanāle] V6
\item[pavanam upānam udīrya kaṇṭhanāle] C3
\item[pavanam upānam urīrya kaṇṭhanāle] C4
\item[pavanam apānam udīrya kaṇṭhanālo] N19
\item[pavanam apānam udīrya kaṇṭhanālāt] B3, C2, J13, N26, V2
\item[pavanam apām udirya kaṃnāle] N25
\item[pavanam apāsenam udāryaṃ kaṇṭhanāle] J5,
\item[pavanam mudārya kaṇṭhanālo] J12
\item[pavanam udirya kaṇṭhanāle] N21
\item[mapānam udīrya kaṇṭhanāle] B1,
\item[pavanamapānam udīrya kaṇṭhanāle] Bo1, Bo2,
\item[pavanam athānam utiya kaṇṭhanāle] N24
\item[pavanam apānam udīrya karāḍanāle] V17
\item[(illegible/unavailable)] Bo3, C5, J2, V19, V10, V14, V21, Tue, V7, V5, V22        
\end{description}
\end{marma}

\begin{marma}[hp02_26c]
\sthana{2.26c}
\item[karibhir iva jalasya vāyuvegāt] A1, B1, B3, C9, J1, J3, J6, J8, J10, J17, N17, W4, V16, V17, N18, N20, N26, V2, V18, V12 
\item[karibhir iva gajasya vāyuvegāt] V20
\item[karibhir iva jalasya vāyuvegā] B2, C3, J12, J16, 
\item[karibhir iva jalaṃ ca vāyuvegā] Ba
\item[karibhiriva jalaṃsya vāyuvegāt] Bo2, 
\item[karibhiriva jalasya vāyuvegāt] C2, LD1, 
\item[karibhiriva jalasya vāyuvego] J13,
\item[karibhiriva jalasya vāyuyogāt] V28
\item[kramamarivayavasya vāyur yo] V1
\item[kramaparicayavaśya nāḍicakrā] Ba5
\item[kramaparicayavaśyā nāḍīmārgā] C6, J7,
\item[kramaparicayavaśya vāyumārga] C8
\item[kramaparicayavasya vāyumārgre] J14,
\item[kramaparicayavaśya vāyumārgāt] Bo1
\item[kramaparicayavaśya vāyumārgaṃ] N21
\item[kramac pacya paricaya vaśyā vāyumārgaṃ] J5
\item[kramaparicayavaśyaṃ vāthagargāṃ] N24
\item[kramaparicayavaśya mārgte(??)] N23
\item[kramaparicayavasya vāyumārge] N22
\item[kramapārīcita vāyumārge] J15,
\item[kramaparicitavasya vāyuyogo] Ko,
\item[kramapārīcitavaśya vāyumārg(oṃ)?] N19
\item[kramapārīcitavaśya vāyumārgo] J11
\item[kramapārīcitavaśya vāyumārge] J4, 
\item[kramapārecitavaktu? vāyumārge] C4
\item[karamaparicayavaśya nāḍīmārgā] LD2
\item[kramaparicittakakrū vāyumārge] N25
\item[(illegible/unavailable)] Bo3, C1, J2, V19, V10, V14, Tue, C5, V21, V6, V7, V15, V5, V22, V13
 \begin{description}
        \end{description}
\end{marma}

\begin{marma}[hp02_26d]
\sthana{2.26d}
\item[gajakaraṇānti nigadyate haṭhajṇau] Ko,
\item[gajakaraṇīti nigadyate haṭhajñaiḥ] B2, Ba, Ba5, Bo1, C3, C4, C6, J4, J6, J7, J11, J15, V1, N19, N23, N25(“dya” unclear), V20, V15, V13
\item[gajakaraṇīti nigadate haṭhajñaiḥ] J14,
\item[gajakaraṇīti nigadyāta haṭhaj~naiḥ] J5,
\item[gajakaraṇīti nigadyaṃte haṭhajñaiḥ] LD2,
\item[gajakaraṇīti nigadyate] N24, V2
\item[gajakarṇīti nigadyate haṭhaj~naiḥ] A1, C8
\item[gajakaraṇīva nigadyate haṭhaj~naiḥ] B1
\item[gajakariṇīva nigadyate haṭhajñaiḥ] C2
\item[gajakaraṇīti nigachate haṭhaj~naiḥ] B3
\item[gajakāraṇī nigadyate haṭhaj~naiḥ] Bo2,V3, V6
\item[gajakaraṇīti nigadyate haṭhakṣaiḥ] N22
\item[gajakaraṇī ca nigadyate dṛḍhajñaiḥ] J13,
\item[jalagajakaraṇī nigadyate haṭhaj~naiḥ] J1, J3, 
\item[jalakaraṇīti nigadyate haṭhajñaiḥ] C9, J12, J16, LD1, N17, W4, N20, N26, 
\item[jalakaraṇīti nigadyate mahāhaṭhaj~naiḥ] V21
\item[jalakariṇīti nigadyate haṭhajñaiḥ] J10, J17, V16, V28, V18, V12
\item[jalakaraṇānī nigadyate haṭhaj~naiḥ] V17
\item[jalagajakaraṇīti nigadyate haṭhaj~naiḥ] J8, N18
\item[gajakaraṇīti nirādyate haṭhaj~naiḥ] N21(unusual way of writing “dya”)
\item[(illegible/unavailable)] Bo3, C1, J2, C5, V19, V10, V14, Tue, V7, V5, V22
 \begin{description}
        \end{description}
\end{marma}

\begin{marma}[hp02_29c]
\sthana{2.29c}
\item[aseṣadoṣān api yan nihanyād] Ko, 
\item[aśeṣadoṣopacayaṃ nihanyad] J8, N17, V1, W4, V16, V17, V10, V6, V7, V15, 
\item[aśeṣadoṣoyathā nihanyānād] V14
\item[aśeṣadoṣopacayaṃ nihanyād] B1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C6, C8, C9, J2, J3, J5, J6, J7, J10, J11, J13, J14, J15, J17, LD1, LD2, N18(“aseṣa”), N19, N20, Tue, N25, N26, V20, V5, V28, V22, V2. V18, V12
\item[aśeṣadoṣopacaya no hanyād] J16,
\item[aśeṣadoṣopacayā nihanyād] V13
\item[aśeṣadoṣopacayaṃ nihanyā] N23(space for “da” left blank with an overhead line, verse comes before the earlier verse in this list)
\item[aśeṣadoṣopacaya nihanyād] N24(verse comes before the earlier verse in this list)
\item[aśeṣa?oṣovacayaṃ nihanyād] N21(verse comes before the earlier verse in this list)
\item[aśeṣadoṣepacayaṃti hanyād] J12,
\item[aśeṣadoṣepacayaṃ nihanyad] C2,
\item[aśeṣadoṣāpacayaṃ nihanyad] V21, V3
\item[aśeṣadoṣopacitiṃ hathajñād] J4,
\item[aśeṣadoṣopacayaṃ ca hanyad] A1
\item[aśeṣadoṣaprabhavaṃ nihanyad] V19
\item[aśeṣadoṣasya cayana hatyā] N22
\item[(illegible/unavailable)] C5,
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_29d]
\sthana{2.29d}
\item[abhyasyamānaṃ jalavastikarma] Ba5, Bo1, Bo2, Bo3, C1, C3, C6, C8, J1, J2, J4, J5, J6, J7, J8, J11, J12, J13, J14, J15, J16, LD1, N17, W4, V16, V17, V10, N18, N19, N21, V21, V3, Tue, N26, V6, V7, V20, V15, V5, V28, V22, V2, V18, V13
\item[anyasyamānaṃ jalavastikarma] J3,
\item[abhyasyamānaṃ nijavastikarma] V12
\item[abhyasyamānaṃ jalavastikarmaḥ] Ko,
\item[abhyasyamānaṃ khalu vastikarma] C9,
\item[abhyasyamāna ja?vastikarma] N20(something seems added above)
\item[abhyasyamānaṃ jalavastitakramāt] N24
\item[praamāńam jalavastikarma] V14
\item[abhyasyamāno jalavastikarma] B3, C2, 
\item[abhyasyamā jalavastikarma] B1,
\item[abhyāsamānaṃ jalavastikarma] C4, V1, N25
\item[abhyāsyamānaṃ jalavastikarma] A1, Ba
\item[abhyāsamānaṃ jalavasti] V19
\item[abhyasyamānaṃ javastikarma] J10, J17,
\item[abhyasyamānaṃ jvalabastikarmaḥ] LD2,
\item[bhyasyabhānaṃ jalavestikarma] N23(space for “da” left blank with an overhead line)
\item[abhyasya śūnyaṃ jalavastikarma] B2,
\item[dattasya mānaṃ yanabasyakarmaḥ] N22
\item[(illegible/unavailable)] C5,
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_30]
\sthana{2.30}
\item[atha natīḥ] J1
\item[atha netī(ḥ)] Ba, Ba5, Bo1, Bo2, C1, C2, C4, J3, J6, J8, V1, V10, N18, V21, V3, N23, N25, N26, V7, V5, V2
\item[atha neti] J2, Ko, N20, N24, V6, V28
\item[atha netiḥ] J7, Tue, V22, V12
\item[atha netikarma] C3, C9, J10, J17, LD1, W4, V16, V17, N21, V14, V15, V18, V13
\item[atha netikarmaḥ] C8, J12, J15, N22
\item[atha netīkarma] A1, B1, B2, C6, J13, J14, J16, N17, V20
\item[atha netīkarmaḥ] LD2
\item[atha nītī] J5,
\item[atha nītikarma] J4,
\item[atha nītīkarma] N19
\item[prathama netīkarma] B3
\item[(illegible/unavailable)] Bo3, C5, J11, 
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_30c]
\sthana{2.30c}
\item[mukhanirgamanād eva] A1, B2, C3, J10, J17, LD1, N17, V1, W4, V16, N22, V20, V18, V12, V13
\item[mukhanirgamanād evaḥ] J15,
\item[mukhanirgamanād deva] C9
\item[mukhanirgatanād evaṃ] J12,
\item[mukhanirgatamanād eva] V14
\item[mukhanirgamayedeti] J2
\item[mukhan nirgamanād eva] B3,
\item[mukhān nirgamanād eva] C2,
\item[mukha nīrgamanod eva] J16, 
\item[mukhān nirgamayec caiṣā] Ba, Ba5, Bo1, C8, J8, J11, LD2, V17, N19, N20, V21, Tue, N26, 
\item[mukhān nirgamyate caiṣā] J4,
\item[mukhān virgamayed eṣā] N21
\item[mukhā nirgamayec caiṣā] J7, V10, V3, V6, V7, V15, V22
\item[mukhe nirgamayec caiṣā] C6
\item[mu nirgamayai caiṣā] J5,
\item[mukhān nirgamaye doṣā] N24
\item[mukhān nirgamayet sāpi] V19
\item[mukhān nirgāmayet sā hi] B1, Bo2, J13, J14, N18, N25  
\item[mukhān nirgamayet sā hi] J6,
\item[mukhā nirgāmaye sā hi] V2
\item[mukhā nirgāmayet so hi] Ko,
\item[mukhā nirgamanā sā hi] V5
\item[mukhān nigarmayet sā hi] C4, J3,
\item[mukhā nigarmayet sā hi] J1, V28  
\item[mukhān nirgatya nāsāyam] C1,
\item[mukhaśca nirgamec caiṣā] N23
\item[(illegible/unavailable)] Bo3, C5
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_33a]
\sthana{2.33a}
\item[ghaṭanaṃ] J2, 
\item[muñcakaṃ] Ko,
\item[mucyate] J4,
\item[mocanaṃ] B2, Ba5, C8, J7, J12, N18, N21, N22, Tue, N23, V7
\item[mocakaṃ] Bo2, C1, J1, J3, J5, J6, J11, J14, N19, V28
\item[moṭakaṃ] V1, Bo1, N25 
\item[toṭakaṃ] V19, V16
\item[tāṭakaṃ] V17
\item[trāṭakaṃ] A1, C4, C6, V10, V14, V21, V3, V6, V20, V15, V5, V28, V22, V2, V18, V12, V13
\item[troṭakaṃ] N24
\item[rocanaṃ?] Ba,
\item[sphoṭanaṃ] B1, B3, C2, C3, C9, J8, J10, J13, J15, J16, J17, LD2, N17, W4, N20, N26,  
\item[sphoṭakaṃ] LD1,
\item[(illegible/unavailable)] Bo3, C5
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_33b]
\sthana{2.33b}
\item[jadrānāṃ ca kapāṭakaṃ] J16,
\item[tandrādīnāṃ gapāṭa[ka]m] V1,
\item[tandrādīnāṃca pāṭanaṃ] V19,
\item[tandrādīnāṃ ca kapāṭakaṃ] B1,
\item[tandrādīnāṃ kapāṭakam] B2, B3, Ba, Ba5, Bo1, C1, C2, C4, C6, C8, J3, J5, J6, J7, J8, J11, J12, Ko, LD1, LD2, W4, V17, V10, N21, N22, V3, Tue, V7, V5, V28, V22, V2, V13
\item[tandrādīṇāṃ kapāṭakam] J13, 
\item[tandrāṇāṃ ca kapāṭakam] J15, 
\item[tandrādīnāṃ kapālakam] J1
item[tandrāegaṃ kathapāṭakam] V21
\item[tandrādīnāṃ kapāṭakai] Bo2
\item[tandrādiṇāṃ kapāṭakam] J8, N18, N23,
\item[tandrādiṇāṃ ca pāṭakam] J14, N19
\item[tandrādīṇāṃ ca pāṭakam] C9
\item[tandrādīṇāṃ ca pāṭhakam] V15
\item[tandrādīṇāṃ ca pāṭavam] J4,
\item[tandrāṇāṃ ca kapāṭakam] C3, J10, J17, V16, N20, V14, N26, V20, V18
\item[tadvāṇāṃ ca kapāṭakam] A1,
\item[tandrāṇāṃ tu kapāṭakam] N17
\item[sandrādīnāṃ kapātakāṃ] N24
\item[nidrārttīnāṃ kapāṭakam] N25
\item[hetādīnāṃ? kapāṭhakam] J2,
\item[(illegible/unavailable)] Bo3, C5, V6, V12
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_34b]
\sthana{2.34b}
\item[gudaṃ] V1
\item[gopyaṃ] V16, V17, V10, N18, V14, V21, V3,V6, V7, V20, V15, V28, V22, V2, V18, V12, V13
\item[tundaṃ] A1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J7, J8, J13, J14, J16, LD1, LD2, N19, N20, Tue, N26, 
\item[tunduṃ] N23
\item[tunda] N17
\item[tunde] C1, C4,  
\item[tuṃḍe(?)] N25, V5
\item[tundat] B2, J10, J17, 
\item[tundet] J15, 
\item[truṃdaṃ] N21
\item[tuda] W4
\item[tudaṃ] B1, C3, C6, J1, J5, J12,  
\item[tundan] J4,
\item[tudan] Ba, Bo1, J11, 
\item[tulyaṃ] Ko,
\item[tujyaṃ] V19
\item[t(?)uddhaṃ] N22
\item[puṃda] N24
\item[saṃdada] J2
\item[(illegible/unavailable)] Bo3, C5,
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_35a]
\sthana{2.35a}
\item[pāvanādi] V1
\item[pāvanāni] N18,
\item[pāvanāgniḥ] N25
\item[pāvanādi] Ko,
\item[pāvakādi] C3, J10, J16,
\item[pāvakādī] J12,
\item[pāvakāgni] N26
\item[pācakādi] C2, C9, J17, LD1, LD2, N17, W4, V16, V17, N20, V14, V21, V18, V13
\item[pācatādi] Bo2, 
\item[pācanādi] A1, B1, B2, B3, Ba, Ba5, Bo1, C6, C8, J4, J5, J7, J8, J11, J14, V19, V10, N19, N21, V3, Tue, N23, V6, V7, V20, V15, V28, V22
\item[pācanāgni] J3, J6, J13,  
\item[pācakāgniḥ] J15
\item[pācakāgni] V2
\item[pācakāri?] J2,
\item[pācanāṃnī] N24
\item[kārkādeśaṃ] N22
\item[(illegible/unavailable)] Bo3, C5, V5, V12
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_35c]
\sthana{2.35c}
\item[apa śoṣaṇī ca] Ba5,
\item[mala śoṣiṇī va] V1,
\item[mala śoṣaṇī ca] B2, C3, J8, LD1, N17, W4, V16, V14, V3, V7, V18, V12
\item[maya śoṣaṇī ca] V21, V15, V28, V13
\item[maya śodhaṇī yā] N23
\item[maya śodhaṇī cā] V6, 
\item[mala śoṣaṇī va] J10, J17,
\item[mala śoṣaṇī vaḥ] J16,
\item[doṣāmaya śoṣaṇī ca] C1, C2, C4, C6, C8, C9, J3, J4, J11, J13, J14,  Ko, N18, N21, Tue, N25, V22, V2
\item[doṣāmaya śauṣaṇī caḥ] J15, 
\item[doṣān mala śoṣaṇī ca] J12,
\item[doṣāmaya śoṣaṇīya] J1
\item[doṣāmaya śoṣaṇīyaṃ] J2
\item[doṣāmaya śoṣiṇī ca] A1, B3, Ba, Bo1, J6, V19, N19
\item[doṣāmala śoṣaṇī ca] N20
\item[doṣāmaya śoṣiṇāṃ ca] Bo2
\item[doṣāmaya śodhinī ca] J7, 
\item[doṣābhayahā soṣaṇī ca] J5,
\item[dauṣāmaya śoṣaṇī ca] LD2,
\item[doṣāmaya goṣaṇī ca] N24
\item[mala śoṣiṇī ca] V17, N26,
\item[doṣāṇapi śoṣaṇā ca] B1,
\item[doṣāmaya śoṣaṇī ca] V10
\item[doṣapra śoṣiṇī] V5
\item[doṣānala śoṣaṇī bahava] N22
\item[(illegible/unavailable)] Bo3, C5, V20, 
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_35d]
\sthana{2.35d}
\item[mūlam iyaṃ hi] J14, V1
\item[mūlam idaṃ hi] Ko, J4, J11, N19
\item[mūlam ayaṃ ca mauliḥ] N22
\item[maulir iyaṃ hi] A1, C1, C4, J3, J5, V19, V21, N25+, V15, V2
\item[maulir iyaṃ] J1
\item[maulir iyaṃ ca] Ba, Ba5, Bo1,Bo2, C6, C8, J6, J7, N21, Tue, N23(“yaṃ” looks like “paṃ”), V6, V7, V20, V28, V22
\item[saulir iya ca] N24
\item[maulir ayaṃ hi] J2
\item[sau jayatīha] J10, J17, N17, W4, N18, N20, N26
\item[jayatīha] B1, B3, C2, C3, C9, J8, J12, J13, J15, J16, LD1, LD2, V3
\item[sauryakarīha?] B2
\item[(illegible/unavailable)] Bo3, C5, V16, V17, V10, V14, V5, V18, V12, V13
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_38a]
\sthana{2.38a}
\item[prāṇāyāmena vai] Ko, J4, J11, J14, V1, N19, V15
\item[prāṇāyāmair eva] B1, B2, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J12, J13, J15, LD2, N1 (not sure of "eva"), N3, N4a, N5, N8, N11, N12, N13, N16, N18, N22, V3, Tue, N23, N24, N25, V6, V7, V5, V28, V22, V2, V12, V13
\item[prāṇāyāmaiḥ sukhāt] A1, C3, C9, J10, J17, LD1, N2 ("śu"), N9, N10,
\item[prāṇāyāme sukhāt] J16,
\item[prāṇāyāmaiḥ sukhāḥ] N26
\item[prāṇāyāmair mu?khāṃ] N20
\item[prāṇāyāmena sarvepi] N21
\item[praṇāyāme kṛte samyak] J2
\item[prāṇāyāme śukhā] N14, V18
\item[prāṇāyāme tataḥ] V16, V17, V10, V14, V21, V20
\item[praṇāyāmāt sukhāt] N7
\item[(illegible/unavailable)] Bo3, C5,
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_38b]
\sthana{2.38b}
\item[malā api] J4, J11, V19, N21, V15, V28
\item[malād api] Ko,
\item[malā iti] Ba, Ba5, Bo1, Bo2, C1, C6, C8, J1, J2, J3, J5, J6, J7, LD2, N3, N4a, N5, N8, N11, N13, N16, V10, N18, N19, Tue, N23, N24, N25, V6, V7, V22
\item[malā itiḥ] J15,
\item[malā ime] J14, 
\item[malāśayaḥ] J10,
\item[malāśayāḥ] A1, B1, B3, C2, C9, J13, LD1, N1, N2, N10, V17, V14, V20, V18, V13
\item[malākulaṃ] B2, N22
\item[malākulāḥ] N20, N26
\item[malāśayā(ḥ)] C3, C4, J16, J17, N9, V16
\item[marāśayā] N14
\item[malāsayā] J12, V21, V3, V2
\item[malāśrayāḥ] N7
\item[malāyinaḥ] N12
\item[(illegible/unavailable)] Bo3, C5, V5, V12
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_39]
\sthana{2.39}
\item[]
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_40b]
\sthana{2.40b}
\item[cirāśrayam] LD2,
\item[nirāśrayam] A1, Ba, C1, C3, C6, C8, C9, J2, J4, J5, J7, J8, J10, J11, J15, J17, LD1, V1, W4, N3, N9, N10, V17, N20, N22, V21, V3, N23, N26, V6, V15, V28, V12, V13
\item[nirāśrayaḥ] J12,
\item[nirodhā vādham] V8
\item[nirāśayam] Bo1,
\item[nirāsnayām] J16,
\item[nirāmayam] B2, Bo2, C4, J1, J3, J6, J14, Ko, N17, V19, N2, N5, N7 ("ma" seems tampered with later), N8, N11, N12, N14, N16, V16, N18, N19, N21, N24, N25, V20, V5, V2, V18
\item[nirākulam] B1, B3, Ba5, C2, J13, N1 (2.45), N4a, N13, V10, Tue, V7, V22
\item[(illegible/unavailable)] Bo3, C5, V14
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_40c]
\sthana{2.40c}
\item[akṣir?] C8,
\item[īkṣed] Ba, J5, 
\item[īkṣet] C6, V13
\item[īkṣe] C1, J7, 
\item[paśyed] A1, Bo1, C9, J4, J8, J10, J11, J14, J15, J16, J17, Ko, LD1, N17, V1, W4, N9, V8, N10, V16, V17, N20, N22(pasyed), V21, V3, N26, V15, V12
\item[paśyad] J12,
\item[paśyet] V20
\item[paśyeta] LD2,
\item[paśye] N19, N24
\item[paśe] C3
\item[paśyan] N7
\item[pasya?] N14  
\item[dṛṣṭir] B3, Ba5, C2, C4, J6, V19, N1, N2, N4a, N11, N16, V10, N21, N25, V7, V5
\item[dṛṣṭi] B1, B2, Bo2, J1, J2, J3, J13, N5, N8, N12, N13, N18, Tue, V28, V22, V2 
\item[vīkṣed] N3
\item[iched] N23
\item[(illegible/unavailable)] Bo3, C5, V14, V6, V18
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_43a]
\sthana{2.43a}
\item[vidhānajñā] C9, J5, J10, J11, Ko, V1, N2 ("vi" seems faded), N4a, N7, N10, N11, N12, N13, N14, N16, V16, V17, V10, N19, N20, N21, V14, V3, Tue, N23, N26, V6, V7, V15, V22, V18, V13
\item[vidhānaḥjñā] J2
\item[vidhānajñe] V2
\item[vidhanajñā] N18
\item[vidhaustajñā] V20
\item[vidhānujñaś] J1
\item[vidhānajñāś] Bo2, C1, C3, C8, J4, J6, J7, J12, 
\item[vidhānajñaś] B1, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C8, J3, J16, N3, N24,
\item[vidhānajña] N22, V12
\item[vidhānejñaś] N8
\item[vidhijñās tu] C6,
\item[vidhānajño] C4, J15, J17, LD1, LD2, N5, N25, V28
\item[vidhānajñoś] J13, 
\item[vividhānajño] B2, 
\item[vidhāna?ī] N9, V8
\item[vidhānajñaiś] N1 (unsure about "ai" matra)
\item[vidhiṃstajñaś] A1
\item[(illegible/unavailable)] Bo3, C5, J14, V21, V5
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_44cd]
\sthana{2.44cd}
\item[bhastrakā bhrāmarī vedhā kevalaṃ cāṣṭakumbhakāḥ] J2
\item[bhastrikā bhrāmarī mūrcchā kevalī cāṣṭakumbhakāḥ] V1, V16, V17, 
\item[bhastrikā bhrāmarī mūrcchā kevala cāṣṭakumbhakāḥ] V20, V18, V12, V13
\item[bhastrikā bhrāmarī mūrcchā vinītyaṣṭakumbhakāḥ] V10
\item[bhastrikā bhrāmarī mūrcchā vinītyaṣṭakumbhakān] V28
\item[bhastrikā bhrāmarī mūr??ā kevalaś cāṣṭakumbhakā] N1,
\item[bhastrīkā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] C9, J13, J15, LD1, LD2, N21
\item[bhaṃstrikā bhramarī mūrchā kāṃbalā atha? kumbhakāḥ] J5
\item[bhaṃstrīkā bhramarī mūrchā kevala cāṣṭakumbhakāḥ] N22, V21, V3
\item[bhastṛ(?)kā bhrāmarī mūrcchā kevalaś cāṣṭakumbhakāḥ] N26
\item[bhastrikā bhrāmarī mūrchā kaivalaś cā sa kumbhakaḥ] B3,
\item[bhastrikā bhramarī mūrcchā cāṣṭakumbhakāḥ sahita?] N2 (after exchange of words' positions)
\item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakāḥ] A1, B1, C3, J7pc, J10, J12, J16, J17, N3, N7, N10("ri" instead of "rī"),
\item[bhastrikā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] N20
\item[bhastrikā marī mūrchā kevalaṃ cāṣṭakumbhakāḥ] B2,
\item[bhastrikā bhrāmalī mūrchā kevalaś cāṣṭakumbhakā] N14
\item[gachati tiṣṭto kāryā mujjayākhyāṃ cākumbhakāḥ] V8
\item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakā] N9
\item[bhastrikā bhrāmarī mūrchā plāvanī saṣṭhakumbhakāḥ] N4a
\item[bhastrikā bhramarī mūrcchā plāvanī tvaṣṭakumbhakāḥ] N12, V6, V7
\item[bhastrikā bhramarī mūrcchā pāvanīty eṣakumbhakāḥ] Ko,
\item[bhastrikā bhramarī mūrcchā pāvanīty aṣṭakumbhakān] J4,
\item[bhastrikā bhrāmarī mūrchā plāvinītyaṣṭakumbhakāḥ] Ba5, Bo1, C1, N13, V14, Tue(no visarga at the end), V15, V22
\item[bhastrikā bhrāmarī mūrchā prāvatītyaṣṭakumbhakaḥ] N23 
\item[bhastrikā bhramarī mūrchā plāvanītyaṣṭakumbhakāḥ] Ba, C8, J7, 
\item[bhastrikā bhrāmarī mūrcchā pālāvanīty aṣṭakumbhakāḥ] J11,
\item[bhastrikā bhramarī mūrchā pratvanītyaṣṭakumbhakāḥ] C6
\item[bhastrikā bhramarī mūrchā sahitāś cāṣṭakumbhakāḥ] C4, N5
\item[bhastikā bhrāmarī mūrchā sahitāś cāṣṭakumbhakam] N25, V5
\item[bhastrikabhrama ma mūrchā saṃhataṃ cāṣṭakumbhakā] N8
\item[bhastrikā bhramarī mūrchā saṃhataṃ cāṣṭakumbhakāḥ] N11(visarga added later), N16 ("saṃhitaṃ"),
\item[bhastrikā bhramarī mūrchā saṃhitaṃ cāṣṭakumbhakāḥ] Bo2, J3, J6, V2
\item[bhastrikā bhramarī mūrchā saṃprataṃ cāṣṭakumbhakān] N18
\item[bhadrikā? bhramarī mūrchā saṃprataṃ rāṣṭakumbhakāṃ] J1
\item[bhastrikā śītalī bhramarī mūrchā plāvanītyaṣtakumbhakāḥ] N19 
\item[sastrikā bhramarī mūrchā kavalaś cāṣṭakumbhakā] N24
\item[(illegible/unavailable)] Bo3, C5, J14, 
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_46]
\sthana{2.46}
\item[om.] V19,
\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkaḥ ||] Ba5
\item[adhas tāt kuñcane vāyuḥ kaṃṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāmīgaḥ ||] Ko,
\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkagḥ ||] V22
\item[adhas tāt kuñcanenāsuṃ kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkāṃ ||] C6
\item[adhas tā kuñcakīnālau kaṇṭhasaṃkocane kṣate | madhyaṃ paścimatāṇena syāt prāṇo brahmanāḍīkaḥ ||] J2
\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] A1, Bo1, C8, J10, J11, V1, N3, N13, Tue, V7, V20, V15, V28, V12, 
\item[adhastvākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C9
\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṣate | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C1,
\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] C3, LD1, N12, N26
\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] B2, J15, V10, V21
\item[adhasthāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] J7,
\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J6, J17, 
\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] Bo2, N10, V16, V17, V14, V3, V6
\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J3,
\item[adhastā kuñcanenāśu kaṇṭhaśaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍiga ||] J12,
\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāna syāt prāṇo brahmāṇḍanāḍigaḥ ||]  Ba, N4a
\item[adhastākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syā? prāṇo brahmanāḍikaḥ ||]  N5
\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍikaḥ ||] C4
\item[adhastākuñcaṇeṇāśu kaṇṭhaśaṇkocane kṛ(?)te | madhye paścamatānena syāt prāṇo madhye nāḍikā ||] N14
\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] N1
\item[adhastāt kuñcanenaiva kaṇṭhaṃ saṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] B3
\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo madhyanāgiḍaḥ ||] J4,
\item[adhas tāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāgiḍaḥ ||] J4,
\item[adhamān kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ||] J5,
\item[adhastāt kuñcanenaiva kaṇṭhaṃ saṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C2
\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] B1
\item[adhasthāt kuñcanenaiva kaṇṭhasokocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] N21
\item[adhastāt kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanādigaḥ ||] J16, 
\item[adhastā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo madhyanāḍigaḥ ||] LD2,
\item[adhastvā kuñcanenāsu kaṇṭhasaṃkocane kṛtaṃ | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] N7, V13
\item[adhastvā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇe brahmanāḍigaḥ ||] N25
\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] N16
\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ ||] J1, N18, N19
\item[adhastāt kuñcanenāsu kekatvaṃ saṃkocane krate madhye paścimatānena syāt prāṇo vasta(?)nāḍ?gaḥ ||] N9
\item[adhastāt kuñcatenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍitāḥ ||] N20
\item[adhas tā kucanenāmaḥ kaṃcasaṃkocane kṛte | madhye paścimatānena prāṇasyā brahmanāḍīkaḥ ||] N22
\item[adhas ta kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanābhirāt ||] N23
\item[adhas tāt kuñcanaunāśu kaṇṭhakocane kṛte | madhye paścimatāne tasyāt pāno madhyanāḍigaḥ ||] N24
\item[adhas tāt kuñcanaunāśu kaṇṭhakocane kṛte | madhye paścimatāne tasyāt prāno madhyanāḍigaḥ ||] V18
\item[(illegible/unavailable)] Bo3, C5, J14, N2, N8, N11, V5, V2
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_48b]
\sthana{2.48b}
\item[āsthāpya] V1, V19
\item[sansthāpya] V8
\item[utthāpya] A1, B1, B2, B3, Ba, Ba5, Bo1, Bo2, C2, C3, C4, C6, C8, C9, J1, J3, J6, J7, J8, J10, J11, J13, J17, LD1, N17, W4, N2, N3, N4a, N5, N8 (not sure of "pya"), N9, N10(seems "chā" instead of "thā"), N11, N12, N13, N16, V16, V17, V10, N18, N21, V14, V21, V3, Tue, N23, N24, N26, V6, V7, V15, V28, V22, V18, V12, V13,
\item[ūtthāpya] J16,
\item[uthāpya] Ko,
\item[usthāpya] J2, J12, 
\item[udhyāpya] N19
\item[ur(?)thāpya] N14
\item[utthāya] C1, N7, N20,
\item[utprāpya] N22
\item[uschāpya] N1 (not sure)
\item[usthāpya] J4,
\item[uchāpya] J5, J15, 
\item[utchāpya?] LD2,
\item[(illegible/unavailable)] Bo3, C5, J14, N25, V20, V5, V2

 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_49b]
\sthana{2.49b}
\item[virōdhāvadhi kumbhayet]
\item[nīrodhāvadhi kumbhayet] J15, J16, J17, V21
\item[nīrodhānadhi kumbhayet] V17
\item[nirodhaṃ kumbhayet tataḥ] J13, 
\item[(illegible/unavailable)] J14,

 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_54a]
\sthana{2.54a}
\item[śītkaraḥ] B1, Ba, C8, 
\item[śītkaram] C6,
\item[śīkāra] J7,
\item[śītkāraḥ] C1, B3, J13, J15, V17, V21 
\item[śītakārāī] LD1, 
\item[sītkara(ḥ)] B1, Ba, Bo2, C2, C3, C4, J3, J4, 
\item[sītkāraḥ] J6
\item[śītkārī] J16, 
\item[sītkāṃ] N18, N22, Tue, N24, N25, V15, V28, V22, V10
\item[sītkād] V14
\item[sītkā] Ko,
\item[śītkāṃ] A1
\item[sītkarā] J17,
\item[sītkārī] Ba5, Bo1, J5, J12, 
\item[sītakārī] C9
\item[sītakarākarmma] J10, 
\item[śītikārī] LD2,
\item[sītalī] J2
\item[sikta?] J1
\item[sītkīṃ] V19,
\item[sitkāṃ] J11, N19
\item[sinktā?] B2
\item[sātkāṃ] N21
\item[dhikt(?)āṃ] N23
\item[kumbhaṃ] V3,J8, N20,V14, V21, N26, V16  
\item[kumbhakaṃ] V7, V20, V18, V12, V13, 
\item[kumbhakaḥ] V3
\item[]
\item[(illegible/unavailable)] Bo3, C5, J14, V1, V6, V5, V2, 
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_55a]
\sthana{2.55a}
\item[cakram āsādya(ḥ)] A1, B3, Ba, Bo1, C2, V1, J8pc, J10, J17, P28, V12, N21(unusual way of writing “dya”), V14, V20, V18, V16
\item[cakrasāmānya(ḥ)] B1, J8ac, J11, J13, N19, N24, V3, V22
\item[cakrasammānyaḥ] Ba5, Tue, V10
\item[cakrastamānyaṃ] J2
\item[cakrasāmānyaṃ] LD2
\item[cakramādyāya] J12,
\item[cakrasāmānya] J4, J5,
\item[cakraṃ samāsādya] N17, W4
\item[cakrasamāsādya] LD1,
\item[cakranāśāya] N20, N26, V17
\item[cakrabhogyaś ca] N22
\item[cakrasevyas tu] Bo2, C1, C4, C6, C8, J1, J3, J6, J7, J15, V19, N18(“ca” doubtful), V21, N25, V15, V28, V2, V13, V21
\item[cakrasevyas ta] N23
\item[cakrabhogyas tu] N22
\item[cakrābhimānyaḥ] Ko,
\item[vaktram āsādya] C3, C9
\item[cakramadyāya] J16,
\item[(illegible/unavailable)] Bo3, C5, J14, V6, V5
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_56*1]
\sthana{2.55*1[c][d]}
\item[satyāmūlena randhreṇa] C4, J6, J8pc, J10, J17, V21, N25
\item[satyāmūlena randhrena] J15,
\item[sadāmūlena randhreṇa] J3, J6pc
\item[satyāmūlena randhrenayaḥ] V18, V16, V21
\item[spṛśyamūlena randhreṇa] B1, B3, C2, J13,  
\item[rasanātālumūlena] N17, 
\item[rasanāmūla randhreṇa] C9, 
\item[rasavatyāmūla randhre yaḥ] N26
\item[]
\item[]
\item[(illegible/unavailable)] A1, B2, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C5, C6, C8, J1, J2, J4, J5, J7, J11, J12, J14, J16, Ko, LD1, LD2, P28, W4, V3, V6, V20, V15, V5, V28, V22, V2, V12, V13, V10, V14, V17
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_57a]
\sthana{2.57a}
\item[rasanātālumūlena] A1, B1, B2, B3, Bo1, C2, C4, C9, J4, J10, J12, J13, J15, J16, J17, Ko, LD1, N21, N22(tālū), N25, N26, V1, V20, V15, V28
\item[rasanātālumūle yaḥ] Ba,
\item[rasanātālumūlenaya] V14,
\item[rasanātālumūlenayaḥ] V2, V18, V12, V3, V16, V17
\item[rasanāttālumūlena] N20
\item[rasanātālumūlaṃ ca] J5, 
\item[rasanātāluyogena] Bo2, C6, C8, J3, J6, J7, N18, V21, N23, V13
\item[rasanātālayogena] J1,
\item[rasanāṃ tālamūlena] J11
\item[rasanānāṃ tālamūlena] N19
\item[jihvayā vāyum ākṛṣya] Ba5,
\item[satyāṃ mūlena randhreṇa] C3,
\item[]
\item[(illegible/unavailable)] Bo3, C5, J2, J14, Tue, LD2, N24, V6, V5, V22, V10
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_58b]
\sthana{2.58b}
\item[pūrvavat kumbhakād anu] C4, C6, C8, J1, J4, J6pc, J7, J11, Ko, N21, V14, V21,  N24, N25, V6, V19, V15, V28, V2
\item[pūrvavat kūmbhakād anu] Bo2,
\item[pūrvavat kumbhakād anmuṃ] Bo1,
\item[pūrvavat kumbhayed anu] A1, V20
\item[pūrvavat kumbhakād anū] N19
\item[pūrve ce kumbhanād agu] J5,
\item[pūrvavat kumbhakādane] N23, V7
\item[pūrvavat kumbhakeṣv anu] B2
\item[pūrvavat kumbhasādhanaṃ] B1, B3, Ba, Ba5, Bo3, C1, C2, C3, C9, J2, J3, J6, J10, J12, J13, J15, J16, J17, LD1, LD2, N18(“sā” unclear), N22, V3, Tue, V22, V18, V12, V13, V3, V10, V16
\item[kākacaṃcusamannayāt] V17
\item[kākacaṃcusamunnayāt] N26
\item[kākacaṃcusamunnarayāt] N20(mark over “kā”, perhaps “ra” is cancelled), 
\item[(illegible/unavailable)] C5, J14, V5
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_62b]
\sthana{2.62b}
\item[kapāle ca samantataḥ] V1 
\item[kapāle niḥsvanaṃ tathā] Ba, 
\item[kapole niḥsvanaṃ tadā] C1, 
\item[kapole nisvanaṃ tadā] J2, 
\item[kapole svāsanaṃ tataḥ] N26
\item[kapole svasanaṃ tataḥ] V18
\item[kapāla śvasamaṃ tataḥ] J4, 
\item[kapāla śasanaṃ tataḥ] J15, 
\item[kapāle śvasanaṃ tataḥ] J11, V17 
\item[kapāle svaśanaṃ tataḥ] LD1, 
\item[kapāla śparśanaṃ tataḥ] J16, 
\item[kapole svaśanaṃ tataḥ] LD2, V16
\item[kapole śvavaṇe tathā] B2,
\item[kapole śvavaṇe tataḥ] V13, V3
\item[kapola saśvanaṃ tataḥ] J5,
\item[kapole sca sanantataḥ] V3
\item[kayolesvaśanaṃtataḥ] N24
\item[kapāla sasvanaṃ tataḥ] N21, V12
\item[kapālasya samaṃ tataḥ] C3
\item[kapālasyarśanaṃ tataḥ] J10, 
\item[kapāle sasvanaṃ tataḥ] Bo1, V14
\item[kapāle mukhano marut] Ko,
\item[kapālādhi sasvanaṃ tataḥ] V6
\item[kapālasparśanaṃ tataḥ] C9, J17,
\item[kapālasya śanaṃ tataḥ] J12,
\item[kapālādhi niḥsvanaṃ] C6,
\item[kapālāvadhi sasvanaṃ] A1, Ba5, C8, J7, Tue, V7, V20, V28, V22, V10
\item[kapālāvadhi sasanaṃ] V21
\item[kapālāvadhi pūrayet] B1, B3, Bo2, C2, C4, J1, J3, J6, J13, N25, V15, V2
\item[kepālavadhi sasvanaṃ] N23,

\item[(illegible/unavailable)] Bo3, C5, J14, V5
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_65cd]
\sthana{2.65cd}
\item[dhārayen nāsikāmadhye aṅgulībhyāṃ vinā dṛḍham] V1, V21
\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] LD1,
\item[dhārayen nāsikāmadhyaṃ aṅgulībhyāṃ tathā dṛḍham] A1, C9, 
\item[dhāraye nāsikāmadhyaṃ maṅgulānāṃ tathā dṛḍham] J12,
\item[dhārayen nāsikāmadhaṃ aṅgulībhyāṃ tathā dṛḍham] V20
\item[dhārayen nāsikāmadhya aṅgulībhyāṃ tathā dṛḍham] J10,
\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] B2, C3, J17, V18, V12, V16, V17
\item[dhārayen nāsikāmadhyai aṅgulībhyāṃ tathā dṛḍham] J15,
\item[dhārayen nāsīkāmadhye maṃgulībhā tathā dṛḍham] J16,
\item[dhārayen nāsīkāmadhye aṅgulībhyāṃ dṛḍham tathā] LD2,
\item[dhārayen nāsikāmadhye aṅguṣṭhābhyāṃ tathā dṛḍham] N26,
\item[dhārayen nāsikāmadhyṃ tarjanībhyāṃ vinā dṛḍham] Ko,
\item[dhārayen nāsikāmadhyaṃ tarjanībhyāṃ vinā dṛḍham] J11,
\item[dhārayen nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] C6, J5, V3, N24, V6, V7, V28, V22, V2, V3, V14
\item[dhāraye nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] Ba,
\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] Ba5, Bo2, J3,
\item[dhāraye nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] Ba5, V15
\item[dhārayen nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] J7,
\item[dhāraye nāsikāṃ madhye tarjjanībhyāṃ vinā dṛḍham] Bo1,
\item[dhārayen nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] B1, B3, J13, V13
\item[dhārayan nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] C2, 
\item[dhārayen nāsikāmadhya tarjjanībhyāṃ vinā dṛḍhām] J1
\item[dhārayen nāsikā madhye tarjanībhyāṃ vinā madam] J4,
\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ yathā dṛḍham] C8
\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] J6pc,
\item[kārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] J6,
\item[dhārayen nāsikāṃ madhyāt tarjanībhyāṃ vinā dṛḍham] Tue
\item[dhārayen nāsikāmadhyāt tarjanībhyāṃ vinā dṛḍham] C4, N25, V10
\item[dhārayaṃ nāśikāmadhye tarjanabhyāṃ vinā dṛḍham] N23
\item[dhāraae vāsikāmadhye tarjanībhyāṃ vinā dṛḍham] J2,
\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_67ab]
\sthana{2.67ab}
\item[kaṇdalībodhakaṃ viprabhavaghnaṃ sukhadaṃ śubhaṃ] V1
\item[kuṇdalībodhanaṃ sarva doṣaghnaṃ sukhadaṃ śivaṃ] Bo1
\item[kuṇdalībodhanaṃ sarva doṣaghnaṃ sukhadaṃ śuvaṃ] V15
\item[kaṇdalībodhakaṃ ksipraṃpavanaṃ sukhadaṃ hitaṃ] V7
\item[kuṇḍalo bodhaka kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] J2
\item[kuṇḍalībodhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] Ba5, V22
\item[kuṇḍalabodhakaṃ pāpāghnaṃ suṣadaṃ śubhaṃ] Ko,
\item[kaṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] V6, V2
\item[kaṇḍalībodhanaṃ cakaṃ pāpaghnaṃ sukhada śubhaṃ] V14
\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] B2, Bo2, C4, J3, J6, J7, J11,  
\item[kaṇḍalībodhanaṃ kuryāt pāpaghnaṃ subhaṃ tad śukhaṃ] V28
\item[kuṇdalībādhakaṃ ksipraṃ pāpghnaṃ sukhadaṃ hitaṃ] Ba, 
\item[kuṇḍalībodhakaḥ karthūḥ roghnaṃḥ sukhadaḥ śubhaḥ] A1, V20
\item[kuṇḍalībodhanaṃ karttur bhavaghnam sukhadaṃ śubhaṃ] C6,
\item[kuṇḍalībodhanaṃ karttur bhavaghnam sukhaṃ śubhaṃ] V21
\item[kuṇḍalībodhakaṃ miśrabhavaghnam sukhadaṃ śubhaṃ] J4,
\item[kuṇḍalībodhanaṃ kartuṃ bhavaghnam sukhadaṃ śubhadaṃ] C8,
\item[kuṇḍalībodhakaḥ kumbho roghnaḥ sukhadaḥ śubhaḥ] C3, J10, J17, 
\item[kuṇḍalībodhakaṃ kumbhoḥ roghnaḥ sukhadaḥ śubhaḥ] J15, J16,
\item[kuṇḍalībodhakaḥ kumbho roghnaḥ suṣadaḥ śubhaḥ] C9, 
\item[kuṇḍalībodhakaḥ kumbho rogaghnaḥ sukhadaḥ śubhaḥ] J12, LD1, LD2, V16 
\item[kuṇḍalībodhakaḥ kumbho rogaghnaḥ sukhadaḥ śubhaḥ] B1, B3, C2, J13, V18, V12, V13, V3, V17 
\item[kuṇḍalīno bodhacakraṃ? bhāvadaṃ supada śubhāṃ] J5
\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, V10
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_68a]
\sthana{2.68a}
\item[guṇatrayasamudbhūtaṃ] Bo1,
\item[samagātrasamudbhūtaṃ] J4,
\item[samyagātrasamudbhūtaṃ] J7,
\item[samyagātrasamudbhutaṃ] J15,
\item[samyaggātrasamudbhūtaṃ] B2, Bo2, C4, C8, J2, J11, V1, V7, V2, V3, V17
\item[samya gātre samudbhūtaṃ] J5,
\item[samyaggātre samudbhūtaṃ] J6, V28
\item[samyaggātre samudbhūta] J3,
\item[samyag gātraṃ samudbhūtaṃ] Ko,
\item[samyaggātasamudbhūtam] V20, V15
\item[samyagataḥ samudbhūtam] V14
\item[samyak gātrasamudbhūta] Ba, V22
\item[samyag antaḥ samudbhū] A1
\item[samyag antaḥ samudbhūta] C6
\item[samyak bhastrā samudbhūtā] B1, B3, C2, J12, J13, J16,  
\item[samyak bhastrā samudbhūtaḥ] V21  
\item[samyak bhastrā samabhyāsaḥ] V13
\item[samyag bhastrā samudbhūtāṃ] LD1, LD2, 
\item[samyak bhastrā samudbhūto] C3, J10, J17, 
\item[samyag bhastri samudbhūtaṃ] C9
\item[sasyagātrasamudbhūtaṃ] J1
\item[samagātrasamudbhūtaṃ] V6, V16
\item[samagātrasamudbhūtaghnaṃ] V18
\item[(illegible/unavailable)] Ba5, Bo3, C1, C5, J14, V5, V12, V10
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_69a]
\sthana{2.69a}
\item[vegair ghoṣaṃ] V1
\item[vegākṛṣṭaṃ] Ba, J2, V15
\item[vego tad ghoṣa] J1
\item[vego ghoṣam] C3, C9, J16,  
\item[vego ghoṣaḥ] J15,
\item[vegāt ghoṣaṃ] A1, B2, B3, Ba5, C9, LD1, V6, V7, V22, V18, V12, V13, V14, V16, V17
\item[vegā ghoṣaṃ] V3
\item[vegot ghoṣaṃ] J12, V2
\item[vegot ghoṣe] V2,
\item[vegodveṣaḥ] J4,
\item[vaiṣodvaiṣaṃ] J5,
\item[veget vege] V28
\item[vegād ghoṣaṃ] J10, J17, V20
\item[vegod ghoṣaṃ] Bo2, C6, J10pc, J11,
\item[vedod ghoṣaṃ] V21
\item[vegodyoṣaṃ] J7,
\item[vemod ghoṣaṃ] J3, J6, 
\item[vegodvegaṃ] C4, J13, 
\item[vegād aghoṣaṃ] B1,
\item[vegād doṣaṃ] Bo1,
\item[veṇu doṣaṃ] C2,
\item[(illegible/unavailable)] Bo3, C1, C5, J14, Ko, V5, V10
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_73a]
\sthana{2.73a}
\item[ārecyāpūrya yat kuryāt] 
\item[ārecyāpūrya vat kuryāt] J13,
\item[recya vā pūrya sanakaiḥ] J11,
\item[recake pūrakaṃ kuryāt] J12, J16, 
\item[recakaḥ pūrakaṣ kāryyaḥ] J17,
\item[recakaḥ pūrakaḥ kāryaḥ] J15, 
\item[recayitvā pūrakaḥ kāryaḥ] J14, 
\item[recake pūrakaṃ uktwā] V14
\item[recaka pūraka kumbhai] V3, V16
\item[recaka pūraka kumbhakaiḥ] V17
\item[recaka pūraka varjite] V10
\item[(illegible/unavailable)]
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_73b]
\sthana{2.73b}
\item[sa vai sahitakumbhakaḥ] J12, J13, J15, 
\item[sarve sahitakumbhakaḥ] J17,
\item[sa vai sahitakumbhak] J16, 
\item[sa taiḥ sahitakumbhakaḥ] J14,
\item[sa vai kevalakumbhakaḥ] V10
\item[kāryaḥ sahitakumbhakaḥ] J11,
\item[kāryaḥ śanai sahitakumbhakā] V3, V14
\item[kāryaḥ śarveiḥ sahitakumbhakeiḥ] V16
\item[kāryaḥ śarveiḥ sahitakumbhaka] V17
\item[(illegible/unavailable)]
 \begin{description}

        \end{description}
\end{marma}

\begin{marma}[hp02_78a]
\sthana{kumbhakaḥ prāṇarecānte} C3, J2, J3, J13, V1, V6, V28, V18, V16
\sthana{kumbhakāt prāṇarecānte} V17
\item[kumbhaka prāṇarecānte] A1, B1, B2, Bo2, C1, J1, J5, LD1, V2, V12
\item[kumbhake siddha reca puraka] V21
\item[kumbhaka prāṇa cātte?] J10, J17, 
\item[kumbhaka prāṇa recāttaḥ] J16,
\item[kumbhakaṃ prāṇarecānte] B3, C4, C9, J6, 
\item[kumbhakaṃ prāṇarecātteḥ] J15,
\item[kumbhakaṃ prāṇareṃcānte] J12,
\item[kumbhataḥ prāṇarecānte] J7,
\item[kumbhata prāṇarecānte] V3
\item[kumbhakaṃ ghrāṇarecānte] LD2, 
\item[kumbhitaḥ prāṇarecānte] Bo1, V15
\item[kumbhītaḥ prāṇarecānte] Ko, J11, J14, 
\item[kumbhīte prāṇarecānte] V14
\item[kumbhītaprāṇareṣānte] J4,
\item[tad etat prāṇarrecānte] C6, C8, 
\item[kumbhakaḥ prāṇarodhānte] Ba, Ba5, V7, V20, V22, V10
\item[(illegible/unavailable)] Bo3, C1, C5, V5, V13
 \begin{description}

        \end{description}
\end{marma}

\end{ekdosis}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

[a]Perhaps "yathā" can be removed from these readings
[b]Perhaps "yathā" can be removed from these readings
[c]I don't know what part of a verse this number means
[d]I think it is 56*1 @nilsliersch@web.de isn't it? 2nd thing..do we need full verse or just a line or a word here?

### 2022-09-18 marma/print-uncollated-sigla.sh

@@ -1,0 +1,40 @@
+#!/bin/sh
+marma_tex=Marmasthāna.tex
+sig_txt=Sigla-Liste.txt
+
+tsvmarma() { 
+    sed -e 's_\\sthana_\\end{sthana}\n\\begin{sthana}_' ${marma_tex} |
+	sed -n '/\\begin{sthana}{[a-z0-9]/,/\\end/ {p}' |
+	sed -e 's_\s\+_ _g' \
+	    -e 's_\\begin{sthana}{\([^}]*\)}_#\1\&_' |
+	tr '\n' ' ' |
+	sed -e 's_#_\n#_g'  \
+	    -e ':a; s_([^()]*)_ _g; t a' \
+	    -e 's_\\begin{[^}]*}\(\[[^]]*\]\)\?__g' \
+	    -e 's_\\item\[[^]]*\]__g' \
+	    -e 's_\\end{[^}]*}__g' \
+	    -e 's_[%{},;] *_ _g' \
+	    -e 's_ \+_ _g' |
+	sed -e '/^\s*$/d' > marma.tsv 
+}
+
+mapfile -t siga < $sig_txt
+tsvmarma
+mapfile -t marmaa < marma.tsv
+
+printf "Uncollated MSS per \\sthana-command:\n\n"
+
+for ((i=0; i<${#marmaa[@]}; i+=1)); do
+    printf "${marmaa[$i]}\n" | cut -d"&" -f1
+
+    for ((j=0; j<${#siga[@]}; j+=1)); do
+	sig=" ${siga[j]} "
+	if (! $(echo "${marmaa[$i]}\n" | cut -d"&" -f2 | grep -sq "$sig"))
+	then
+	    printf ' -%s\n' "$sig"
+	fi
+    done
+
+    printf "\n"
+    
+done

### 2022-12-16 marma/Marmasthāna.tex
new uncollated-sigla.txt

@@
+Notes
+    • P16 has chapter 1 till 1.45
+    • P22 has only 3 folios of chapter 1
 \input{Preamble-Print}
@@
     \item[-vat kratvā]{P6}
-        \item[-bandhastho]{A1, J8, L1, M1, V6, B1, B2 (in marg.), B3, Bo2, C1, C2, C3, C4
-           (\emph{madhyastho} is also given above), C6, C8, J1, J3, J6, J13, J14, K1, Ko, LD1, LD2, N1, V2, N2,
-           V3 (about \emph{stho}), N4a, V8, N5, N8, N9, N11, V21, N13, N16, N18, N19, N20, N21, N24,
-           N25, N26, P1, P2, P3, P4 (\emph{... sanaṃ}), P7, Tue,V17, V14, R3, R5, V13, V10}
+    \item[-bandhastho]{A1, J8, L1, M1, V6, B1, B2, B3, Bo2, C1, C2, C3, C4, N2, C6, C8, J1, J3, J6, J13, J14, K1, Ko, LD1, LD2, N1, V2, V3 (about \emph{stho}), N4a, V8, N5, N8, N9, N11, V21, N13, N16, N18, N19, N20, N21, N24, N25, N26, P1, P2, P3, P4, P7, P10, P11, P14, P15, P17, P20, P21, P22, P23, R4, Tue, V17, V14, R3, R5, V13, V10, G1
     \item[-bandhasthe]{J7, V19}
     \item[-badhascho]{Ba, Bo3, V28, V22}
-        \item[-baṃdhascho]{J5, J11, J15, V15}
-        \item[-baddhastho]{N12}
+    \item[-baṃdhascho] J5, J11, J15, P12, P13, P18, P27, V15
+    \item[-baddhastho] N12, P25, 
     \item[-bandhastu]{Bo1}
-        \item[-madhyastho]{C9, J10, J12, J16 (\emph{baṃ} is given in margin), J17, N17, V4, N6, N7,
-           N10, V20, V18, P5, P8, V12, LD1}
+    \item[-madhyastho]{C9, J10, J12, J16 (\emph{baṃ} is given in margin), J17, N17, V4, N6, N7, N10, V20, V18, P5, P8, P24, P26, V12, LD1}
      \item[-madhyasthā]{P9}
@@
     \item[-badhestu]{J2}
-        \item[(unavailable/illegible)]{Ba5, C7, J4 (\emph{gudaṃ nibadhya gulphābhyāṃ}), C5, R2}
+    \item[-vadhnastho] P16, 
+    \item[(unavailable/illegible)]{Ba5, C7, J4, N14, C5, G11, R2, T, Ba3
     \end{description}
@@
   \begin{description}
-\item[jaṭharapradīptaṃ]        V6 (jaraṭha), B1, B2, B3, Bo1, C1, J13("pradīpta"), N1, V22, N13, N21 (No anusvara over "ṭha" of "pīṭha"), P2, R3(...pīṭha)
+\item[jaṭharapradīptaṃ]    V6 (jaraṭha), B1, B2, B3, Bo1, C1, J13, N1, V22, N13, N21, P2, P12, P13, P14, P17, R3, R4, 
 \item[jaṭharapradīpti]    Ba5, V10
-\item[jaṭharapradiptaṃ]        P4(...piṭhaṃ)
+\item[jaṭharapradiptaṃ]    P4,
 \item[jaṭharaprakṛṣtaṃ]    V14
-\item[jaṭharapradīpaṃ]        P6("piṭhaṃ" changed to "pīṭhaṃ", "dipaṃ" changed to "dīpaṃ")
-\item[jaṭharapradīptiṃ]        Bo3, Ko, N4a, Tue, R5
-\item[jaṭharapracaṇḍaṃ]        K1, V1, N22, V15
-\item[jaṭharaṃ pracaṇḍaṃ]        J11("jvalanaṃ pradīptaṃ" is also given above),
-\item[jaṭharapracaṃḍa]        J5
-\item[matsyendra-vīra-jaṭhara-pracaṇḍaṃ]              J4, N19 (V1 (above) also has "jaṭhara-pracaṇḍaṃ") [vīra-ja is unmetrical?]
+\item[jaṭharapradīpaṃ]    P6,
+\item[jaṭharapradīptiṃ]    Bo3, Ko, N4a, P18, P22, Tue, R5
+\item[jaṭharāgnidīptiṃ]    P22a,
+\item[jaṭharapracaṇḍaṃ]    K1, V1, N22, V15, G1
+\item[jaṭharaṃ pracaṇḍaṃ]    J11, P16, 
+\item[jaṭharapracaṃḍa]    J5, P11, P27, 
+\item[matsyendra-vīra-jaṭhara-pracaṇḍaṃ]          J4, N19, V1 [vīra-ja is unmetrical?]
 \item[budhaṃ pracaṇḍa]    V8
+\item[jaṭharapravṛdhaḥ]    P21, P23, 
 \item[jaṭharapravṛddha]    J10, J17, V2, N8, N11
-\item[jaṭharapravṛddhau]        J15
+\item[jaṭharapravṛddhau]    J15, P20, 
 \item[jaṭharapravṛddhaṃ]    N24
 \item[jaṭharapravṛddhiṃ]    C1
-\item[jaṭharaprabuddhaṃ]        C8, J14, N2, N12, V21, N23, V12
-\item[jaṭharaprabuddha]        Ba,
+\item[jaṭharaprabuddhaṃ]    C8, J14, N2, N12, V21, N23, V12,
+\item[jaṭharaprabuddha]    Ba, P25, 
 \item[jaṭharaprabuddhau]    C4 ("jvalanapradīptam" is also given above), L1, V19, V28, N5, N25, P3
-\item[jaṭharaprabodhaṃ]        A1, C6, P7,
-\item[...bodhaṃ]           R2(only "bodhaṃ" available)
+\item[jaṭharaprabodhaṃ]    A1, C6, P7, P26,
+\item[...bodhaṃ]       R2,
 \item[jaṭharāprabodhaṃ]    V20
 \item[jaṭharaṃ prabadhaṃ]    J2
-\item[jvalanapradīptaṃ]        LD1, Bo2, C3, C9, J1, J3, J6, J8, N17, V3, V4, N6 ("na" is written as a later addition), N9, N18 (No anusvāra visible on "ṭha"), V18, N26, P1, P5,V17, P8('pradī' is correction at bottom of folio), V13, V16, P9
-\item[jvalanaṃ pradīptaṃ]        J7, J12, LD2, N10, N16, N20(It is "di", short "i". Also, "ṭhaṃ" is not there.)
+\item[jaṭhara pravaddhaṃ] N2, 
+\item[jvalanapradīptaṃ]    LD1, Bo2, C3, C9, J1, J3, J6, J8, N17, V3, V4, N6,, N9, N18, V18, N26, P1, P5, V17, P8, P24, V13, V16, P9
+\item[jvalanaṃ pradīptaṃ]    J7, J12, LD2, N10, N16, N20, P10, 
 \item[t(?)valanaṃ pradiptaṃ]    J16,
-\item[jvalatpradīptaṃ]        N7
-\item[(unavailable/illegible)]        C5, C7, M1
+\item[jvalatpradīptaṃ]    N7,
+\item[vīrapracaṃḍaṃ] P15,
+\item[(unavailable/illegible)]    C5, C7, M1, N14, G11, T, Ba3
     \end{description}
@@
     \begin{description}
-        \item[candrasthiratvaṃ ca dadāti] Ba5, J7, J10ac, J17, LD1, LD2, V6, B2, Bo1, Bo3, N4a, V4, V22, N10,
-         N13, P1, Tue, P8, R5, V13, V10
+    \item[candrasthiratvaṃ ca dadāti] Ba5, J7, J10ac, J17, LD1, LD2, V6, B2, Bo1, Bo3, N4a, V4, V22, N10, N13, P1, Tue, P8, P22, R5, V13, V10
       \item[candraṃ sthiratvaṃ ca dadāti] P9("ca" in margin)
@@
     \item[candrasthiratvaṃ pradadāti] C1, N24
-        \item[caṇḍasthiratvaṃ ca(?) dadāti] N7
-        \item[daṇḍasthiratvaṃ ca dadāti] A1, Ba, Bo2, C3, C6, C8, J1, J2, J8, J10pc, J11, J12, J14,
-         J15, J16("ḍaṃ" is given in margin), K1, Ko, M1, N17, V1, V2, V3, N6, N9, N12, V20, N16, N18,
-         N19, N20, N22, N26, P6, V17, V12, R2
+    \item[caṇḍasthiratvaṃ ca(?) dadāti] N7,
+    \item[daṇḍasthiratvaṃ ca dadāti] A1, Ba, Bo2, C3, C6, C8, J1, J2, J8, J10pc, J11, J12, J14, J15, J16, K1, Ko, N2, M1, N17, V1, V2, V3, N6, N9, N12, P10, V20, N16, N18, N19, N20, N22, N26, P6, P18, P25, V17, V12, R2
     \item[daṇḍāsthiratvaṃ ca dadāti] P5
+    \item[daṇḍasthira.ntya karoti] G1
     \item[daṇḍāsthitatvaṃ ca dadāmi] V14
-        \item[daṇḍaṃ sthiratvaṃ ca dadāti] P7
-        \item[daṇḍasthiratvaṃ hi dadāti] J13, P2 (anusvāra over "da" is missing)
+    \item[daṇḍaṃ sthiratvaṃ ca dadāti] P7, P11, 
+    \item[daṇḍasthiratvaṃ hi dadāti] J13, P2
     \item[daṇḍasthiratvaṃ hi dadātti] P4
+    \item[daṃḍaschiratvaṃ hi dadāti] P12, 
     \item[daṇḍasthiratāṃ ca dadāti] V21
@@
     \item[daṇḍasthiratvaṃ pradadāti] N8, N11, V19, N23
+    \item[daṇḍaṃ sthiratvaṃ pradadāti] P23, 
     \item[daṃḍe sthiratvaṃ pradadāti] N5
@@
     \item[kaḍasthiratvaṃ ca dadāti(?)] J4
-        \item[(unavailable/illegible)] C5, C7
+    \item[vīryasthiratvaṃ ca dadāti] P22a,
+    \item[pracaṃḍahrunmaṃḍalakhaḍanāstraṃ] P21, 
+    \item[pracaṃḍarugmaṃḍalakhaṃḍanāstraṃ] P13, P14, P17, P20, P24, P26, R4, 
+    \item[rugmaṃḍalākhaṃḍanakhaṃḍanāstraṃ] P27,
+    \item[picaṃḍarukmāṃḍanakhaṃḍanāstraṃ] P15,
+    \item[vicaṃḍayugmaṃḍalakhaṃḍanāstraṃ] P16,
+    \item[(unavailable/illegible)] C5, C7, N14, T, Ba3
     \end{description}
@@
    \begin{description}
-  \item[dorbhyāṃ padāgradvitayaṃ] A1, Bo1, C2, C6, C8, J4, V1, M1, B2, B3, N1, N4a, V21, V20, V22,
-        V28, N16, N19, N21, N22, N23, N24, P2, P6, R5, V13, R2
+   \item[dorbhyāṃ padāgradvitayaṃ] A1, Bo1, C2, C6, C8, J4, V1, M1, B2, B3, N1, N4a, P11, P14, P18, P20, V21, V20, V22, G1, V28, N16, N19, N21, N22, N23, N24, P2, P6, P12, P13, P17, P26, R4, R5, V13, R2
    \item[dorbhyāṃ padāgredvitayaṃ] V14
@@
    \item[dorbhyāṃ padāgraṃ dvitīyaṃ] Ko, V15
-  \item[dorbhyāṃ padāgradvitīyaṃ] R3
-\item[dorbhyāṃ padagradvitayaṃ] Bo3, Ba5, K1, N13, Tue ("pa" uncertain), V10
+   \item[dorbhyāṃ padāgradvitīyaṃ] R3, P25, P27, 
+   \item[dorbhyāṃ padāgraṃthidvayaṃ] N2,
+\item[dorbhyāṃ padagradvitayaṃ] Bo3, Ba5, K1, N13, P14, P22, Tue, V10
 \item[dorbhyāṃ padāgryau dvitayaṃ] J5
 \item[dorbhyāṃ pādāgrādvitayaṃ] B1
-\item[dorbhyāṃ ca pādadvitayaṃ] C4, L1, V19, N5, N7, N8, N11, N12, P3
+\item[dorbhyāṃ ca pādadvitayaṃ] C4, L1, V19, N5, N7, N8, N11, N12, P3, P16, P23, 
 \item[dorbhyāṃ ca pādaṃ dvitayaṃ] C1
@@
 \item[dorbhyāṃ sa pādāgradvayaṃ] J2
-\item[dvābhyāṃ karābhyāṃ dvitayaṃ] Bo2, C3, C9, J1, J3, J6, J7, J8, J10, J15, J16, J17, LD1, LD2, N17, V6, V4, V3, N6, N9, N10, N18, N20, V18, N26, P1, P5, V17, V16, LD1, P9("karābhyāṃ" in margin)
+\item[dvābhyāṃ karābhyāṃ dvitayaṃ] Bo2, C3, C9, J1, J3, J6, J7, J8, J10, J15, J16, J17, LD1, LD2, N17, V6, V4, V3, N9, N10, N18, N20, V18, N26, P1, P5, P10, P24, V17, V16, LD1, P9("karābhyāṃ" in margin)
+\item[dorbhyāṃ ca pādadvayakaṃ] P21, 
+\item[dvābhyāṃ karābhyāṃ dvitīyaṃ] N6, 
 \item[dvābhyāṃ karābhyāṃ hitayaṃ] J12, V12
@@
 \item[dvābhyāṃ pādāgrādvitīyaṃ] P8
-\item[(unavailable/illegible)] Ba, C7
+\item[(unavailable/illegible)] Ba, C7, G11
 \item[dorbhyāṃ padagraṃthidvayaṃ] J14, N2
 \item[.. rbhyāṃ pādaṃbhyāṃ dvitayaṃ] V8
-\item[(unavailable/illegible)] C5,
+\item[(unavailable/illegible)] C5, N14, T, Ba3, 
     \end{description}
@@
    \begin{description}
-\item[asyahṛdaye dhṛtvā]        J7("eva niyataṃ" in below margin), J12, A1, V4, V18, V16, P8
+\item[asyahṛdaye dhṛtvā]    J7, J12, A1, V4, V18, V16, P8
 \item[asya hṛdayaṃ dhṛtvā]   P9
@@
 \item[āsyahridayaṃ dhṛtvā]    J16
-\item[eva hṛdaye dhṛtvā] C4 ('vidadhe' is also given above), C7, C8, J5, J10, J11, J15, J17, Ko,
- V19, N2, N5, V21, N25, P3, V15, V14, R8,V12
-\item[eva hṛdaye kṛtvā]        Bo3, K1, N4a, V22, N13, N19, N24 (seems later addition), Tue, V10
-\item[eva niyataṃ dhṛtvā]        C1, C3, J14, V8, V28
-\item[eva niyataṃ kṛtvā] B1, B3, Bo2, C2, C6, J1, J3, J6("sudṛdha" is also given above), J8, J13,
- L1, N17, LD1, N1, V3, N12, V20, N16 (It seems "meḍhra" here), N18, N20, N21, N26, P2, P4, P7, R3,
+\item[hṛdayaṃ kṛtvā] G1
+\item[eva hṛdaye dhṛtvā] C4, C7, C8, J5, J10, J11, J15, J17, Ko, V19, N2, N5, V21, N25, P3, P20, P23, P24, P25, V15, V14, R8, V12
+\item[ava hṛdaye dhṛtvā] P11, 
+\item[eva hṛdaye dhṛtvā]    P21,
+\item[eva hṛdaye kṛtvā]    Bo3, K1, N4a, V22, N13, N19, N24, P14, P27, Tue, V10
+\item[eva niyataṃ dhṛtvā]    C1, C3, J14, R4, V8, V28
+\item[eva niyataṃ kṛtvā] B1, B3, Bo2, C2, C6, J1, J3, J6("sudṛdha" is also given above), J8, J13, L1, N17, LD1, N1, V3, N12, V20, N16, N18, N20, N21, N26, P2, P4, P7, P12, P13, P17, P18, P26, R3,
   R5('hṛdaye' canceled and replaced by 'niyataṃ')
-\item[ekahṛdaye dhṛtvā]        J4, N3, N23
+\item[ekahṛdaye dhṛtvā]    J4, N3, N23, P15, G11
 \item[ekahṛdayo dhṛtvā]    J2, M1
+\item[eksak ahṛdaye dhṛtvā]    P10, 
 \item[eva sudṛḍhaṃ kṛtvā]    N8, N11
+\item[eva dṛḍhayan dhṛtvā]    P16,
 \item[vāmaniyataṃ dhṛtvā]    V6
@@
 \item[meṃḍhraṃ pādam athaikam]            Bo1, LD1, LD2
-\item[meḍhre pādam athaikm eva hṛdaye]        Ba5, N7
+\item[meḍhre pādam athaikm eva hṛdaye]    B2, Ba5,
 \item[meṃḍhreṃ pādam athaikam]            LD2
 \item[meḍhraṃ pādam athaika samyak meva hṛdaye dhṛtvā]           P5
-\item[(unavailable/illegible)]        Ba, C5, V13, R2
+\item[(unavailable/illegible)]    Ba, C5, V13, N6, N14, P22, R2, T, Ba3
     \end{description}
@@
    \item[yameṣv eva sitāhāra] V16
-  \item[yameṣv eva mitāhāro] N7
+   \item[yameṣv eva mitāhāro] N7,
    \item[yameṣv eva mitāhāram] C4, B2, J14, N2, N5
    \item[yameṣv eva ātmacetana] V28
-  \item[yameṣv iva mitāhāra(ḥ)] B1, B3, Bo1, Bo2, C3, C9, J3, J6, J8, J13, J15, K1, Ko, LD1, N17, N1,
-        V3, V4, N9, N10, N14, N16, N18, N21, N26, P1 (seems to be "mitāhārāḥ"), P2, P4, R3(after
-        correction), V12, LD1, P9
+   \item[yameṣv iva mitāhāra(ḥ)] B1, B3, Bo1, Bo2, C3, C9, J3, J6, J8, J13, J15, K1, Ko, LD1, N17, N1, V3, V4, N9, N10, N14, N16, N18, N21, N26, P1, P2, P4, P12, P17, R3, V12, LD1, P9
    \item[jameṣṭiva mitāhāram] J5
-  \item[yameṣv iva mitāhāro] V18
-  \item[yameṣv iva mitāhāre] P5
+   \item[yameṣv iva mitāhāro] P13, R4, V18
+   \item[yameṣv iva mitāhāre] P5, P24, 
    \item[yameṣv iva mītāhāra] J16,
-  \item[yameṣv iva mitāhāram] A1, Bo3, Ba5, C1, C7, C8, J4, J11, L1, LD2, V19, N4a, V22, V21, N13, N23,
-        N24('va' of 'iva' not clear), N25, P3, P7("mitah.." earlier), Tue, V15, R5, V13, V10
+   \item[yameṣv iva mitāhāram] A1, Bo3, Ba5, C1, C7, C8, J4, J11, L1, LD2, V19, N4a, V22, V21, N13, N23, G11, N24, N25, P3, P7, P11, P14, P15, P16, P18, P20, P21, P23, P25, P26, P27, Tue, V15, R5, V13, V10
    \item[yamiṣviva mitāhāram] V14
    \item[yameṣv iva mitāhā(ḥ)] V8
+   \item[yameṣv iva mityāhāraḥ] P10, 
    \item[niyameṣu mitāhāro] N8, N11, N12
@@
    \item[yameṣv api mitahāraḥ] J7("yamesṣv iva" is deleted)
-   \item[(unavailable/illegible)] C5,
+   \item[(unavailable/illegible)] C5, N6, T, Ba3, G1, P22,
     \end{description}
@@
 \item[ahiṃsā niyameṣv api]    B2, J14, N2, N24, V14
-\item[ahiṃsā niyameṣv iva] A1, B3, Ba5, Bo1, Bo2, Bo3, C2, C3, C4, C7, C8, C9, J1, J3, J5, J6, J7, J8,
- J10, J11, J13, J15, J17, Ko, LD1, LD2, N17, V3, V4, N5 (there is an extra mātrā after "me" and makes it
- look like "mo"), N9, V21, N13, N14, N16, N19, N20, N21, V18, N25, N26, P1 (was "ivā", the long
- mātrā seems canceled), P2, P4, P7, V16, P8, R3, R5, V13, V12, LD1, P9
+\item[ahiṃsā niyameṣv iva] A1, B3, Ba5, Bo1, Bo2, Bo3, C2, C3, C4, C7, C8, C9, J1, J3, J5, J6, J7, J8, G11, J10, J11, J13, J15, J17, Ko, LD1, LD2, N17, V3, V4, N5, N9, V21, N13, N14, N16, N19, N20, N21, V18, N25, N26, P1, P2, P4, P7, P10, P11, P12, P13, P15, P16, P17, P18, P23, P24, P25, P26, V16, P8, R3, R5, V13, V12, LD1, P9, P14,
+\item[ahiṃsā niyameṣv ivaṃ] P21,
 \item[ahiṃsā niyāmaṣv iva]    P5
@@
 \item[ahiṃsā nīyame sthitā]    K1,
+\item[ahiṃsāṃ nīyame sthitā]    P27, 
 \item[ahisā nīyameṣv iva]    J16,
@@
 \item[svāhiṃsā niyameṣv iva]    V6
-\item[ahiṃsāṃ niyameṣv iva] L1, V19, N4a, V22, P3, Tue, V15, V10
+\item[ahiṃsāṃ niyameṣv iva] L1, V19, N4a, V22, P3, P20, Tue, V15, V10
 \item[ahiṃsāṃ niyameṣu ca]    C6,
 \item[ahiṃsā niyaṃmai(?)ṣv iva]     N1
-\item[(a)hiṃsā ca niyameṣv iva]        N7("a" could be supplied due to its potential lopa due to sandhi)
+\item[(a)hiṃsā ca niyameṣv iva]    N7,
 \item[yathāhiṃsā yameṣv iva]    N8, N11
@@
 \item[yameṣv evātma-cintanaṃ]    N12
+\item[hiṃsā ca niyameṣv iva] R4, 
 \item[na hiṃsāṃ niyameṣv iva]    N23
 \item[kṣamasye ātmacintane]    J2
-\item[(unavailable/illegible)]        Ba, C5, V28, V20, N18 (only "ahisā" is visible), R2
+\item[(unavailable/illegible)]    Ba, C5, N6, V28, V20, N18, R2, T, G1, Ba3, P22, P25, 
     \end{description}
@@
  (ṇ/n and ś/s irrelevant)
-\item[suṣumṇā iva nāḍikā]        V1, N22 (anusvāra over "nā"), P6
-\item[suṣumṇā iva nāḍiṣu]         B2, J4, J14, K1, Ko, N2, V8, N19 (seems "suṣumṇa", short "a"), V12
+\item[suṣumṇā iva nāḍikā]    V1, N22, P6
+\item[suṣumṇā iva nāḍiṣu]     B2, J4, J14, K1, Ko, N2, V8, N19, V12, G11
+\item[suṣumṇā caiva nāḍiṣu]     P26, 
 \item[suṣumṇā ca nāḍiṣu]    J5,
 \item[suṣumṇā eva nāḍiṣu ]    V20
-\item[suṣumṇā iva nāḍiṣuḥ]        N8
+\item[suṣumṇā iva nāḍiṣuḥ] N8, P15, 
 \item[suṣumiva nāḍiṣu]     V28
 \item[suṣumṇām eva nāḍiṣu]    J10ac, J17, N24, V14
-\item[suṣumṇām iva nāḍiṣu ]        A1, C1, C4 ("nāḍīnāṃ malaśodhane" is given above), C6, C7, C8, C9 (in marg.), J1, J2, J3, J6, J11, J15 (suṣūmṇām), L1, M1, V19, N5, N7, N12, N16, N23 (sukhumṇām), N25, V15, V13
+\item[suṣumṇām iva nāḍiṣu ]    A1, C1, C4, C6, C7, C8, C9, J1, J2, J3, J6, J11, J15, L1, M1, V19, N5, N7, N12, N16, N23, N25, P11, P16, P20, P21, P23, P25, P27, V15, V13
 \item[suṣuṇmām iva nāḍiṣu]    P7
+\item[suṣumṇāśva nāḍiṣu]    N11,
 \item[suṣumṇām iva ḍinoṣu]    P3
-\item[nāḍīnāṃ malaśodhanam]        B1, B3, Ba5, Bo1 (it seems 'मळ'), Bo3, C2, C3, C9, J8, J12, J13, LD2, N1, N4a, N10, N13, N18 (It is "ḍi" here.
-Short "i"), N20, N21 ("nāṃ" seems written differently), N26, V3, V22, P1 (was "śodhane", mātrā of "e" was canceled and anusvāra was added), P2, P4, Tue("naḍī.." earlier.
-Mark over "na" probably intends to indicate it as "nā"), V17, R3, R5, V10
+\item[nāḍīnāṃ malaśodhanam]    B1, B3, Ba5, Bo1, Bo3, C2, C3, C9, J8, J12, J13, LD2, N1, N4a, N10, N13, N18, N20, N21, N26, V3, V22, P1, P2, P4, P12, P13, P14, P17, P18, Tue, V17, R3, R5, V10
+\item[nāḍiṣu malaśodhanam]  R4,
 \item[nāḍināṃ malasodhanaṃ]    J16,
 \item[nāḍīnāṃ malaśodhanā]   P9
-\item[nāḍīnāṃ malaśodhane]        J7, J10pc, LD1, N9, N17, V6, V21, N14(seems "dī"), V18, P5, V16, P8, V12, LD1
+\item[nāḍīnāṃ malaśodhane]    J7, J10pc, LD1, N9, N17, V6, V21, N14(seems "dī"), V18, P5, P10, V16, P8, P24, V12, LD1
 \item[suṣumṇādiṣu śodhanam ]     V4
-\item(unavailable/illegible)]        Ba, Bo2, C5, R2
+\item(unavailable/illegible)]    Ba, Bo2, C5, N6, R2, T, G1, Ba3, P22, 
     \end{description}
@@
 \item[śramādau bahubhiḥ pīṭhaiḥ] V1
-\item[śramadair bahubhiḥ pīṭhaiḥ] B1, B3, Bo1, Bo2, C1, C2, C6, C8, J3, J6, J13, J14, K1, M1,
- N1(not sure if it is 'śra'), N12, N16, N21("daiḥ" instead of "dair"), N22, N23, N24, P2, P4,
- P7(bbahu), V14, R3, V13
-\item[śramadair bahubhiḥ pīṭhai]        P6("piṭhai" changed to "pīṭhai")
+\item[śramadair bahubhiḥ pīṭhaiḥ] B1, B3, Bo1, Bo2, C1, C2, C6, C8, J3, J6, J13, J14, K1, M1, G11, N1, N11, N12, N16, N21, N22, N23, N24, P2, P4, P7, P11, P12, P13, P16, P17, P21, P23, P25, P27, V14, R3, V13
+\item[śramadair bahubhiḥ pīṭhai]    P6,
 \item[śramadai bahubhiḥ pīṭhaiḥ]    Ba, C9mg, J5,
@@
 \item[śramādaiḥ bahubhi pīṭhaiḥ]    V21
-\item[śramadairghyādibhiḥ pīṭhaiḥ]        C4("kim ādyair bahubhiḥ pīṭhai" is given above), C7, L1, V19, N5, N25, P3
+\item[śramadairghyādibhiḥ pīṭhaiḥ]    C4, C7, L1, V19, N5, N25, P3, P20, 
 \item[śramadairdhyādibhiḥ pīṭhaiḥ]    V28
@@
 \item[(samastair?) bahubhiḥ pīṭhaiḥ]    N2
-\item[samastair bahubhiḥ pīṭhaiḥ]        J2, J4, J11, Ko,
+\item[samastair bahubhiḥ pīṭhaiḥ]    J2, J4, J11, Ko, N2, P15, 
 \item[samasta bahubhiḥ pīṭhai]    V15
-\item[kim ādyair bahubhiḥ pīṭhaiḥ] A1, C9("śramādai bahubhiḥ pīthaiḥ" is also given above), J7, J8,
- J10, J17, LD2, N17, V6, V3, V4, N7, N14(visarga at the end not visible due to bad picture), V20,
- N18(No visarga at the end), N20, V18, N26, P5, V17, P8, V12, LD1, R2, P9("rba" in margin)
+\item[kim ādyair bahubhiḥ pīṭhaiḥ] A1, C9, J7, J8, J10, J17, LD2, N17, P26, V6, V3, V4, N7, N14, V20,
+  N18(No visarga at the end), N20, V18, N26, P5, P10, V17, P8, V12, LD1, R2, P9
 \item[kim ādyair bahubhiḥ pīṭheḥ]    N9, V16
-\item[kim adyair bahubhiḥ pīṭhaiḥ]        N10(N10 has this pāda as the 1st pāda, not the 5th)
-\item[kim ādyair bahubhiḥ piṭhaiḥ]        J15, LD1, P1
-\item[kim anyair bahubhiḥ pīṭhaiḥ]        Ba5, C3, N4a, N13, Tue, R5, V10
+\item[kim adyair bahubhiḥ pīṭhaiḥ]    N10,
+\item[kim ādyair bahubhiḥ piṭhaiḥ]    J15, LD1, P1, P24, 
+\item[kim anyair bahubhiḥ pīṭhaiḥ]    Ba5, C3, N4a, N13, P14, P18, Tue, R5, V10
 \item[kim anyair bahuniḥ pīṭhaḥ]    Bo3, V22
@@
 \item[pramadair bahubhiḥ pīṭheḥ]    N19
-\item[(unavailable/illegible)] C5,
+\item[tathaikasmin eva dṛḍham]    R4, 
+\item[(unavailable/illegible)] C5, N6, T, G1, Ba3, P22,
     \end{description}
@@
 \item[sadā siddhāsanasthiteḥ]    A1, V20, V12, R2
-\item[sadā siddhāsane sati] B1, B3, Bo1, C2, C9("kiṃ sadā sevanena tu" is also given above), J7,
- J8, J10, J11("siddhaḥ" is given above), J12, J15, J17, LD1, LD2, N17, V1, V6, N1('kiṃ' and 'śa' seem faded
- and thus, I guess, deleted.), V4, N7, N9, N10(N10 has this pāda as the 2nd pāda, not the 6th),
- N20("sa" of "sati" is not clearly visible), N22, V18, N26, P1, P2, P4, P5, P6, V17, V16, P8, LD1, P9
+\item[sadā siddhāsanasthiteḥ]    P26, 
+\item[sadā siddhāsane sati] B1, B3, Bo1, C2, C9, J7, J8, J10, J11, J12, J15, J17, LD1, LD2, N17, V1, V6, N1, V4, N7, N9, N10, G11, N20, N22, V18, N26, P1, P2, P4, P5, P6, P8, P9, P10, P12, P24, V17, V16, LD1,
 \item[sidhe siddhāsane sati]    V10
+\item[siddhe siddhāsane sati] P18,
 \item[sadā siddhāsane satī]    J16,
@@
 \item[sadā siddhāsanābhyām sādyogī]    V14
-\item[yadā siddhāsane sati]        J2, J4, N2
+\item[yadā siddhāsane sati]    J2, J4, N2, P15,  
 \item[yadā siṃhāsane sati]    Ko,
-\item[siddhe siddhāsane sati]        Ba5, Bo3, C3, N4a, N13, Tue
-\item[siddhe siddhāsane miti]        R5('mi' seems tampered with)
-\item[kiṃ sadā siddhasādhanaiḥ]        M1
-\item[kiṃ syāt siddhāsane sati] Bo2, C1, C4("sadā siddhāsane sati" is also given above), C7, J1,
- J3, J5, J6, J13, L1, V19, B2, V28, N5, N16, N21("sīddh..." long"i"), N25, P3, V15, R3
+\item[siddhe siddhāsane sati]    Ba5, Bo3, C3, N4a, N13, P14, Tue
+\item[siddhe siddhāsane miti]    R5,
+\item[kiṃ sadā siddhasādhanaiḥ]    M1, P11, 
+\item[kiṃ syāt siddhāsane sati] Bo2, C1, C4, C7, J1, J3, J5, J6, J13, L1, P13, P16, P20, V19, B2, V28, N5, N16, N21, N25, P3, P17, P21, P23, V15, R3
 \item[kiṃ syāt siddhāsane mati(ṃ?)]    N8
-\item[kiṃ syāt siddhāsane sthite]        C6, C8, V21
+\item[kiṃ syāt siddhāsane sthite]    C6, C8, P25, V21
 \item[kiṃ syā siddhāsane sthite ]    P7
 \item[kiṃ syāt siddhāsana sthite ]    N23
-\item[kiṃ syāt siddhāsane schite]        Ba
+\item[kiṃ syāt siddhāsane schite]    Ba,
 \item[kiṃ vā siddhāsane sati]    J14, N12, N24
@@
 \item[kiṃ sadā sevanena tu]     K1, N19(1.40h is "baddhe siddhāsane sadā")
-\item[(unavailable/illegible/unable to read)]        C5, N14, V13
+\item[baṃdhe siddhāsane sadā]  P27,
+\item[baddhe siddhāsane sadā]  R4,
+\item[va??t siddhāsane sadā]  N11,
+\item[(unavailable/illegible/unable to read)]    C5, N6, N14, V13, T, G1, Ba3, P22,
     \end{description}
@@
 \item[(utta(m)bhya cibukaṃ ~)]
-\item[vakṣa sthāpayet ]        J8, J10, J17, LD2, N18, V6, N7, P8, N18, V12, V3, P9("thā" unclear)
-\item[vakṣaḥ sthāpayet]         Bo1, B2, C3, J1, J12, J15, J16(sthāpaye),  K1, LD1, V18, P1, P5, V16, N10, N22(vakṣyaḥ), LD1
+\item[vakṣa sthāpayet ]    J8, J10, J17, LD2, V6, N7, P8, N18, V12, V3, P9("thā" unclear)
+\item[vakṣaḥ sthāpayet]     Bo1, B1, B2, C3, J1, J12, J15, J16(sthāpaye),  K1, LD1, V18, P1, P5, P10, P24, R4, V16, N10, N22, LD1
+\item[vakṣaḥ sthāpaye]     P26, 
 \item[vandā sthāpayet]    V17
@@
 \item[vakṣasy āsthāpya]     C4, L1, M1, V19, P3
-\item[vakṣasy utthāya]        Bo2, C6, N16, N8, N11, N16
-\item[vakṣasy utthāpya] B5, Bo3, C1, J13, J14, N13, N21, N23, N24(unclear if it is "syu"), Tue,
- N1(unclear), N4a, R5
+\item[vakṣasy utthāya]    Bo2, C6, N16, N8, N11,
+item[vakṣasy stthāpya] G11
+\item[vakṣasy utthāpya] Ba5, Bo3, C1, J13, J14, N2, N13, N21, N23, N24, P13, Tue, N1, N4a, P14, P18, P23,  R5
+\item[vakṣasy usthāpya] P21,
+\item[vakṣasy utthāyo] P15,
 \item[vakṣasy utṭhāpyo]    J4, J11,
@@
 \item[vakṣasy oschāpyo]    J5
+\item[vakṣasy osthāpyot]    P11, 
 \item[vakṣyasy utthāpyot patane...]    N19
 \item[vakṣyasy asthāpya]    N25
-\item[vakṣyapyutchāpya]        Ba
-\item[vakṣe sthāpayet]        N14 (seems "vacche"), P2, P4, N9
+\item[vakṣyasy āsthāpya]    P20,
+\item[vakṣyapyutchāpya]    Ba, 
+\item[vakṣyapyutthāpya]    P25, 
+\item[vakṣe sthāpayet]    N14, P2, P4, N9
 \item[vakṣe syu sthāpya pavanaṃ]     V10
-\item[vakṣye sthāpayet]        C2
+\item[vakṣye sthāpayet]    B3, C2, P17, 
+\item[vakṣye schāpayet]    P12, 
 \item[vakṣye pyutthāya]    C8
@@
 \item[yuktaṃ vakṣastha-cibukaṃ mākramet]    N12
-\item[(unavailable/illegible/unable to read)]        C5, V13, R2
+\item[(unavailable/illegible/unable to read)]    A1, C5, N6, P16, V13, R2, T, G1, Ba3, P22, P27, 
     \end{description}
@@
 \item[dhyānaṃ tataś cośitaṃ]    V6
-\item[dhyānaṃ tataś conmitaṃ]        P5
+\item[dhyānaṃ tataś conmitaṃ]    P5, P24, 
 \item[dhyānaṃ tataḥ conmī]    N1(some syllables seem deleted, 'nmī' is added later)
-\item[dhyānaṃ ca tac cetasi] B2, B3, Ba, Bo1, Bo2, C1, C2, C4, C6, C7, C8, J4, J5(caitasi), J8,
- J11, J12, J13, K1, Ko, L1, V19, N25, N26, P2, P3 ("cubukaṃ"), P4, P6, P7(cetasī), V4(ṃ omitted),
- V17, V16, N3, N5, N8, N11, V15, V14, P8, R3, V13, N14, N16, N17,N19, V12, V3, N21, N23, N24(saṃniyamya), R2, P9
+\item[dhyānaṃ ca tac cetasi] B1, B2, B3, Ba, Bo1, Bo2, C1, C2, C4, C6, C7, C8, J4, J5, J8, G11, J11, J12, J13, K1, Ko, L1, V19, N25, N26, P2, P3, P4, P6, P7, P12, P13, P15, P17, P26, V4(ṃ omitted), V17, V16, N3, N5, N8, N11, V15, V14, P8, P20, P25, R3, V13, N14, N16, N17,N19, V12, V3, N21, N23,  N24(saṃniyamya), R2, P9
+\item[dhyāyaṃś ca tat cetasi] P14,
 \item[dhyānaṃ ca tecesi]    V8
-\item[dhyānaṃ ca tac cepsitaṃ] A1, J1, J3, J6, J10, J14, J16, J17, LD1, P1, N7(later attempt to form 'dhyā'), N9, N10(saṃvidhāya), N12, N18(dhānaṃ), LD1, N20, N22
+\item[dhyānena tac cetasi] P21, P23, 
+\item[dhyānaṃ ca tac cepsitaṃ] A1, J1, J3, J6, J10, J14, J16, J17, LD1, P1, P10, N7, N9, N10, N12, N18, LD1, N20, N22, P27, 
 \item[dhyānaṃ ca yac cepsitaṃ]    C3,
 \item[dhyānaṃ ca tac cepsite]    J7,
-\item[dhyānaṃ ca tatrepsitaṃ]        LD2, N2
-\item[dhyāyaṃś ca tac cetasi]        Ba5, Bo3, M1, Tue, N4a, N13, V10, V22
-\item[dhyāyaṃś ca taṃ cepsitaṃ]        R5(something is canceled and 'psitaṃ' is added)
-\item[(unavailable/illegible)]   C5,
+\item[dhyānaṃ ca tatrepsitaṃ]    LD2, N2,
+\item[dhyāyaṃś ca tac cetasi]    Ba5, Bo3, M1, Tue, N4a, N13, P18, V10, V22
+\item[dhyāyaṃś ca taṃ cepsitaṃ]    R5
+\item[(unavailable/illegible)]   C5, N6, T, G1, Ba3, P11, P16, P22, R4, 
     \end{description}
@@
   \begin{description}
-\item[(vāraṃ vāram apānam ūrdhvam anilaṃ ~)]
+\item[(vāraṃ vāram apānam ūrdhvam anilaṃ ~)] G11
 \item[protsālayan pūrayan]    V1
@@
 \item[protsārayaḥ pūrayem]    V8
-\item[protsārayan pūritaṃ]        A1, Ba5, Bo3, C1, C6, C8, J10, Tue, N4a, V14, R5, V13, N13, V10, V22, R2("n pū" not legible due to damage)
-\item[protsārayat pūritaṃ]        P7
-\item[protsyā?rayat pūritaṃ]        Ba
+\item[protsārayan pūritaṃ]    A1, Ba5, Bo3, C1, C6, C8, J10, P14, P15, P26, Tue, N4a, V14, R5, V13, N13, P18, V10, V22, R2, G11
+\item[protsārayat pūritaṃ]    P7, P25, 
+\item[protsyā?rayat pūritaṃ]    Ba,
 \item[protsārayan pūrayet]    C4("proccālayan pūritaṃ" is also given above),L1, N25, P3
-\item[protsārayet pūrayan]        V19
+\item[protsārayet pūrayan]    P20, V19
 \item[protsārayet pūrayet]    C7
-\item[protsāraṇaṃ pūrayet]        N5
+\item[protsāraṇaṃ pūrayet]    N5, P21, P23, 
 \item[protthāya yat pūrayan]    B3
 \item[protthāpanṃ pūrayet]    Bo1,
-\item[protthāpayan pūrayan]        C2, P2, P4, N1, R3, P9(not sure of "tthā")
+\item[prodhāyana pūrayet]    B1,
+\item[protthāpayan pūrayan]    C2, P2, P4, N1, R3, P9, P12, P17, 
 \item[procārayet pūritaṃ]    J8
 \item[prottāsayan pūritaṃ]    V12
-\item[prollāsayan pūritaṃ]        J10, J17, N17, P1, P5
+\item[prollāsayan pūritaṃ]    J10, J17, N17, P1, P5, P24, 
 \item[prollāśayet pūritaṃ]    J15, N9
@@
 \item[proccārayat puraya yataṃ(?)]    J1,
-\item[proccārayat pūritaṃ]        N2
+\item[proccārayat pūritaṃ]    N2, P27, 
 \item[proccārayet pūritaṃ]    V15, V3, N23
@@
 \item[proccālitaṃ pūritaṃ]    N14(vālaṃ vālam)
-\item[proccānalaṃ pūrayen]        N21
+\item[….... śayat pūritaṃ]    P10, 
+\item[proccānalaṃ pūrayen]   P13, N21
 \item[prodbhā(ṭa?)yan pūrayet]    N8
-\item[prod(ghā?)ṭayan pūrayet]        N11
+\item[prod(ghā?)ṭayan pūrayet]    N6, N11
+\item[(unavailable/illegible)] T, G1, Ba3, P11, P16, P22, R4,
 
@@
 \item[(muñcan prāṇam upaiti bodham atulaṃ ~)]
-\item[śaktiprabhāvān naraḥ] B3, Ba5, Bo2, Bo3, C1, C2, C6, C8, J1, J3, J6, J8, J11, J13, J14, K1, Ko,
- M1, V1, P2, P4, Tue, N2, N3, V15, V14, R3, R5, V13, N13, N16('vā' at bottom of page), N19, V10, V22, N23
+\item[śaktiprabhāvān naraḥ] B1, B3, Ba5, Bo2, Bo3, C1, C2, C6, C8, J1, J3, J6, J8, J11, J13, J14, K1, Ko, M1, V1, P2, P4, Tue, N2, N3, V15, V14, R3, R5, V13, N13, N16, N19, P12, P13, P14, P15, P17, P18, P25, V10, V22, N23, P27, G11
 \item[śaktiprabhāvān nara]    V3
@@
 \item[śaktiprabhāvāṃtaraḥ]    N18
-\item[śaktiprabhāvodayāt] C4, C7, L1, V19, N25 (1st part of pāda: prāṇaṃ muñcyati yāti bo...), V16,
- N5, N14(sakti)('da' seems later correction)
+\item[śaktiprabhāvodayāt] C4, C7, L1, V19, N25, P20, V16, N5, N14, P21, P23, P26, 
 \item[śaktiḥ prasādān naraḥ]    J2,
@@
 \item[śaktiprabodhodayāt]    N20
-\item[śaktiprabhāvodayaḥ]        J7, V6
+\item[śaktiprabhāvodayaḥ]    J7, P10, V6
 \item[śaktiprabhāvād ataḥ]    C3, C9, J10, J16, P5, P7, N9, N10, N11, N12, P8, V12, LD1, N24, P9
-\item[śaktīprabhāvad ataḥ]        LD2
+\item[śaktīprabhāvad ataḥ]    LD2, P24, 
 \item[śaktiprabhāvad ata]    J15, N8, V4
@@
 \item[bhakti prabhāvodayaḥ]    J4,
-\item[(unavailable/illegible)]   C5,
+\item[(unavailable/illegible)]   C5, N6, T, G1, Ba3, P11, P16, P22, R4,
 
@@
   \begin{description}
-\item[mārutaṃ nayate yas tu]        C6, C8,  
-\item[mārutaṃ niyato yas tu]        Ba,
+\item[mārutaṃ nayate yas tu]    C6, C8, P11, 
+\item[mārutaṃ niyato yas tu]    Ba, P25, 
 \item[mārutaṃ niyatam yaskta]    V14
 \item[mārutaṃ niyatam yas tu]    V8, N23(mār(ū?)taṃ)
+\item[mārutaṃ mriyate yas tu] G11
 \item[mārutaṃ nīyate yas tu]    C1,
@@
 \item[mārutaṃ pibate yas tu]    J11,
-\item[mārutaṃ yas tu pibati] A1, C9, J7, J10, J12, J15, J17, LD2, N17, V6, N26, P1, P5, P6, P7, V17, V16,
- N7, N9, N10, P8, N14, V4, LD1, N20(piba?ti), R2(earlier reading cancelled and overwritten), P9
+\item[mārutaṃ yas tu pibati] A1, C9, J7, J10, J12, J15, J17, LD2, N17, V6, N26, P1, P5, P6, P7, V17, V16, N7, N9, N10, P8, N14, V4, LD1, N20, P24, R2, P9
+\item[mārutaṃ yas tu pibatī] P21, 
+\item[māruto bhidyate yasya]    N2, 
 \item[māruto yas tu pibati]    N22
@@
 \item[mārutaṃ yas tu pībaṃti]    J16,
-\item[mārutaṃ dhārayet yas tu] Ba5, Bo2, Bo3, C2, C7, J1, J3, J6, J13, L1, M1, V19, N25 ("dhārayed"), P2
- ("dhārayed"), P3("), P4("), Tue("), N4a("), N11("), N12("), R5("), N13("), N16("), V10, N24("dhārayed")
-\item[māruto dhārayed yas tu]        N1, R3, V22
+\item[mārutaṃ yas tu vidhivat]    P26,
+\item[mārutaṃ dhārayet yas tu] Ba5, Bo2, Bo3, C2, C7, J1, J3, J6, J13, L1, M1, V19, N25, P2, P3, P4, P14, P18, P20, P23, Tue, N4a, N11, N12, P12, P13, R5("), N13, N16, V10, N24,
+\item[māruto dhārayed yas tu]    N1, P17, R3, V22
 \item[māruta dhāraye?as tu]    N21
@@
 \item[mārutaṃ mriyate yasmāt]    Ko,
-\item[māruto mriyate yas tu]        J4
-\item[māruto bhidyate yasya]        B2, Bo1, J14, K1, N2, N19
+\item[māruto mriyate yas tu]    J4, P15, 
+\item[māruto kṣiyate yasya] P27,
+\item[māruto bhidyate yasya]    B1, B2, Bo1, J14, K1, N2, N19
 \item[mahato dhārayed yastu]    B3
 \item[pavanṃ prāsayeta saṃstu]    J2
-\item[(unavailable/illegible)]    C5, 
+\item[(unavailable/illegible)]    C5, N6, T, G1, Ba3, P10, P16, P22, R4, 
     \end{description}
@@
   \begin{description}
-\item[nāsāgre nyastalocanaḥ]        B3, C2, C9, J7, J13, J16, N17, V1, V6, P2, P4, N1, N7, R3, N22
-\item[nāsāgre nyastalocanā]        P6
-\item[nāsāgra-nyastalocanaḥ]        A1, Bo1, C3, J10, J17, N26, P1, P5, V16, N10, V12, V4, LD1, LD2, N20, R2
+\item[nāsāgre nyastalocanaḥ]    B1, B3, C2, C9, J7, J13, J16, N17, V1, V6, P2, P4, P17, N1, N7, P12, R3, N22
+\item[nāsāgre nyastalocanā]    P6, R4, 
+\item[nāsāgra-nyastalocanaḥ]    A1, Bo1, C3, J10, J17, N26, P1, P5, V16, N10, V12, V4, LD1, LD2, N20, P24, R2
 \item[nāsāgraṃ nyastalocanaḥ]  P9
 \item[nāsāgranyastalocanaṃ]    J8
+\item[nāsāgra vyasalocanaḥ]    P26, 
+\item[nāsāgranyastaloca....]    P10,
 \item[nāsāgraṃ nyastalocanaṃ]    P7, V17, V3
 \item[nāsāgra susamāhitaḥ]    C8,
-\item[nāsāgraṃ susamāhitaḥ] Bo2, Bo3, C1, C4("nāsāgra nyastalocanaḥ" is also given), J1, J2, J3, J6,
- J11, J14, J15, K1, Ko, L1, M1, V19, N25, P3, Tue, N2, N4a, N5, N12, V15, V14, V13, N13, N16, V10,
- V8
+\item[nāsāgraṃ susamāhitaḥ] Bo2, Bo3, C1, C4("nāsāgra nyastalocanaḥ" is also given), J1, J2, J3, J6, J11, J14, J15, K1, Ko, L1, M1, V19, N25, P3, Tue, N2, N4a, N5, N12, V15, V14, V13, N13, N16, P13, P14, P15, P18, P23, V10, V8, G11
 \item[nāsāgrastu samāhitaḥ]   N23
 \item[nāśāgre susamāhitaḥ]    N19,
+\item[nāsāgre susamāhitaḥ]    P27,
 \item[nāsyagraṃ tu samāhitaḥ]    J5,
+\item[nāsāgre susamāhitaṃ]    P21, 
 \item[nāsāgraṃ susamāhitaṃ]    Ba5, C6, V22
@@
 \item[nāśāgraṃ tu samāhitaḥ]    N3,
-\item[nāsāgraṃ tu samāhitaḥ]   N24
-\item[nāsāgre susamāhitaḥ]        B2, Ba, K1, N21
+\item[nāsāgraṃ tu samāhitaḥ]   N24, P20, 
+\item[nāsāgre susamāhitaḥ]    B2, Ba, K1, N21, P25, 
+\item[nāsāyaṃ susamāhitaḥ]    J4, P11,
 \item[nāsāyāṃ susamāhitaḥ]    J4,
@@
 \item[nāsāgre neṣu nocana]    N14
-\item[(unavailable)]        C5, C7,
+\item[(unavailable/illegible)]    C5, C7, N6, T, G1, Ba3, P16, P22,
     \end{description}
@@
   \begin{description}
- \item[kurute cāsanottamam] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J8,
-        J10pc, J13, J15, L1, N17, V6, N26, P2, P3, P4, P5, P7, Tue, N1, V17, N4a, N5, N8, N11(anusvāra
-        over 'cā' seems canceled), N12, V15, R3(visarga is also there), R5, V13, N13, N14, N18, N19,
-        V8, V3, N20, N21, N23, R2
+  \item[kurute cāsanottamam] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J8, J10pc, J13, J15, L1, N17, V6, N26, P2, P3, P4, P5, P7, Tue, N1, V17, N4a, N5, N8, N11, N12, P12,  P13, P14, P17, P18, P20, P21, P23, P24, V15, R3, R5, V13, N13, N14, N18, N19, V8, V3, N20, N21, N23, P25, R2, R4, G11
 \item[kuruta cāsanottamam]    V22
 \item[kurute cāsanomatam]    V10
-\item[kurute cāsanottame]        N16
+\item[kurute cāsanottame]    B1, N16
 \item[kurute cāsanottamaḥ]    J2,
 \item[kurute cāsanotraye]    V14
-\item[kurute cāsanottama]        A1,
+\item[kurute cāsanottama]    A1, P26, 
 \item[kurute cāsanontayam]   N24
-\item[kurute vāsanottamam]        V1, V19, N3
-\item[kurute sādhanottamam]        C3, J12, J10ac, J16, J17, P1, V16, N9, N10, V12, V4, P9
+\item[kurute vāsanottamam]    V1, V19, N3, P11, 
+\item[kurute sādhanottamam]    C3, J12, J10ac, J16, J17, P1, V16, N9, N10, P9, P10, V12, V4,
 \item[kurute sādhanottaramaṃ]    P8(perhaps 'ra' is attempted to be canceled)
@@
 \item[sevate cāsanottamaiḥ]    J14,
-\item[sevate cāsanottamaṃ]        J11, Ko(casano°), N2
+\item[sevate cāsanottamaṃ]    J11, Ko, N2, P15,  
 \item[sevyate cāsanottamaṃ]    J4,
-\item[(unavailable)]        B2, C5, C7, K1, M1, P6, N22
+\item[(unavailable/illegible)]    B2, C5, C7, K1, M1, N6, N22, P6, T, G1, Ba3, P16, P22, P27, 
     \end{description}
@@
   \begin{description}
- \item[sarvavyādhi-viṣāpaham] C3, C9, J5, J7, J8, J10, J12, J15, J16, J17, LD1, V1pc, V6ac, P1,
-        P5(sarvva), P6, P7, V17, N3, V16, N9, N10, P8, R1, V12, V3, V4, N22, P9
+  \item[sarvavyādhi-viṣāpaham] C3, C9, J5, J7, J8, J10, J12, J15, J16, J17, LD1, V1pc, V6ac, P1, P5, P6, P7, P24, V17, N3, V16, N9, N10, P8, R1, V12, V3, V4, N22, P9, P11, G11
   \item[sarvavyādhī-viṣāpaham] LD2
- \item[sarvavyādhi-vināśanam] A1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4("viṣāpahaṃ" is also given above), C6, C7, C8, J1, J2, J3, J4, J6, J11, J13, K1, Ko, L1, N17, V1ac, V6pc, V19, N25, N26(Seems ...sanam), P2, P3, P4, Tue, N1, N4a, N5, N7, N11, N12, V15, V14, R3(seems an extra mātrā after 'vyā'), R5, N13, N14(sarvva...sanaṃ), N16, N18(sanaṃ), N19(anusvāra seems erased due to damage), V10, V8, V22, N23, N24(sanaṃ), R2
+  \item[sarvavyādhi-vināśanam] A1, B1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C7, C8, J1, J2, J3, J4, J6, J11, J13, K1, Ko, L1, N17, V1ac, V6pc, V19, N25, N26, P2, P3, P4, P17, P25, Tue, N1, N4a, N5, N7, N11, N12, P12, P13, P14, P15, P18, P20, V15, V14, R3, R5, N13, N14, N16, N18, N19, V10, V8, V22, N23, N24, P21, P23, P26, P27, R2, R4,
 \item[sarvavyāvināśanam]   N21
 \item[sarvadhyādhi-vighātakaṃ]   N20
-\item[sarvavyādhi-praṇāśanaṃ]        J14, N2
+\item[sarvavyādhi-praṇāśanaṃ]    J14, N2, 
+  \item[sarvavyādhivikṣāpaham] P10, 
 \item[kaphavātāmayāpahaṃ]    M1
-\item[(unclear/unable to read)]        C5, N8(last two syllables seem erased), V13
+\item[(unavailable/illegible)]    C5, N6, N8, V13, T, G1, Ba3, P16, P22,
     \end{description}
@@
 \begin{description}
-\item[Yes] B1, B2, B3, Ba, Bo1, Bo2, C1, C3, C4, C6, C8, C9, J1, J3, J6, J7(58 num.), J8, J10, J11,
- J12, J13, J14, J15, J16, J17, K1, L1, LD1, LD2, N5, N17, N25(the first 3 lines only?), V3, V17, V25, P8,
- R1('śiśno....' not there), R3, N14, N16, N18, V12, LD1, N7, N9, N10, N13(appears way later, after the verse "yuvā vṛddho...yogeṣvatandritaḥ", and last line not there), N20, N21, N26, P1, P3, P4, P5, P7, R2, P9
+\item[Yes] B1, B2, B3, Ba, Bo1, Bo2, C1, C3, C4, C6, C8, C9, J1, J3, J6, J7, J8, J10, J11, J12, J13, J14, J15, J16, J17, K1, L1, LD1, LD2, N5, N17, N25, P10, V3, V17, V25, P8, R1, R3, N14, N16, N18, V12, LD1, N7, N9, N10, N13, N20, N21, N26, P1, P3, P4, P5, P7, P14, P17, P21, R2, P9, P12, P24, P25, P26, P27,
 \item[Yes], but the 4th line appears as the 2nd line: C2, J13, N1, P2
-\item[No] A1, Ba5, Bo3, C1, C5, C7, J2, J4, J5, Ko, M1, V1, V6, V8, V19, R5, N19(mudrādipavanakriyām also
- not there), V10, V22, V4 (unclear/unable to read) V13, N3, N4a, N8, N11, N12, N22, N23, N24, Tue, P6
+\item[No] A1, Ba5, Bo3, C1, C5, C7, J2, J4, J5, Ko, M1, V1, V6, V8, V19, R5, N19, P18, P20, V10, V22, V4 (unclear/unable to read) V13, N2, N3, N4a, N6, N8, N11, N12, N22, N23, N24, Tue, P6, P11, P13, P15, P16, P22, P23, R4, G11, T, G1, Ba3
     \end{description}
@@
  \begin{description}
-\item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, V6, V19,
-  R3, N14(bodhibudhau, malīna), N16, N18, N1 (word "tathā" before "pralīna", also some letters before "svacche", perhaps "ttu", seem faded and not taken into consideration), N8, N11, P2, P3, P4, N13(at the end of 1st chapter)
-\item[No] A1, B2, Ba5, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, LD1, LD2, M1,
-  V1, V3, V8, V17, V25, P8, R1, R5, N19, V12, V10, V22, V4, LD1 (unclear/unable to read) V13, N2, N3, N4a, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, Tue, P1, P5, P6, P7, R2, P9
+ \item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, P12, P17, P20, P21, P23, P26, V6, V19, R3, N14, N16, N18, N1,
+
+ \item[No] A1, B2, Ba5, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, LD1, LD2, M1, V1, V3, V8, V17, V25, P8, R1, R5, N19, V12, V10, V22, V4, LD1, V13, N2, N3, N4a, N6, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, P11, P18, Tue, P1, P5, P6, P7, P9, P10, P13, P14, P15, P16, P24, P27, R2, R4, G11, T, G1, Ba3, P22, P25,
     \end{description}
@@
 \begin{description}
-\item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, V6, V19, R3,
- N14(bodhibudhau, malīna), N16, N18, N8, N11, P2, P3, P4
-\item[No] A1, B2, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, M1,
- V1, V3, V8, V17, V25, P8, R1, R5, N13, N19, V12, V10, V22, V4, LD1 (unclear/unable to read) V13, N2, N3, N4a, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, Tue, P1, P5, P6, P7, R2, P9
+\item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, V6, V19, R3, N14, N16, N18, N8, N11, P2, P3, P4, P12, P21, P23, 
+\item[No] A1, B2, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, M1, V1, V3, V8, V17, V25, P8, R1, R5, N13, N19, V12, V10, V22, V4, LD1, V13, N2, N3, N4a, N6, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, Tue, P1, P5, P6, P7, R2, P9, P10, P11, P13, P14, P15, P16, P17, P18, P20, G11, T, G1, Ba3
     \end{description}
@@
   \item[āsanaṃ kumbhakaś citro] J8pc, N24
- \item[āsanaṃ kumbhakaś citraṃ] B2, C3, J7, J8ac, J10, J12, J15, N17, P8, N14(ā(?)naṃ), V12, LD1, LD2, N4a, N10, N12, N20, P1, P5
+\item[pīṭhāni kuṃbhakāś citro] G11
+\item[pīṭhādi kuṃbhakāś citram] R4, 
+  \item[āsanaṃ kumbhakaś citraṃ] B2, C3, J7, J8ac, J10, J12, J15, N17, P8, N14(ā(?)naṃ), V12, LD1, LD2, N4a, N10, N12, N20, P1, P5, P24, P27, 
   \item[āsanaṃ kumbhavaś citra] J2,
@@
 \item[āsanaṃ kumbhakāś citrāḥ]  N8, N11
+  \item[āsanaṃ kumbhavaś cittaṃ] P10, 
 \item[āsanaṃ kumbhakāś citrā]   P7
@@
   \item[āsanaṃ kumbhakaṃś citraṃ] C9, J17
- \item[āsanaṃ kumbhakaṃ citraṃ] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C7, C8, J1, J3, J4, J5, J6, J11, J13, J14(55), K1, Ko, L1, M1, V19, R3, R5, N13, N16, N19, V10, V22, V4, N1, N5, N23, N25, N26, P2, P3, P4, P6, R2
+  \item[āsanaṃ kumbhakaṃ citraṃ] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C7, C8, J1, J3, J4, J5, J6, J11, J13, J14, K1, Ko, L1, M1, P11, P12, P13, P18, V19, R3, R5, N13, N16, N19, P14, P17, P20, P21, P23, P25, V10, V22, V4, N1, N5, N23, N25, N26, P2, P3, P4, P6, P26, R2
+  \item[āsanaṃ kumbhakaṃ citra] P15,
 \item[āsanaṃ kumbhakaṃ śvitraṃ]   N21   
- \item[āsanaṃ kumbhaka citrāṃ] J16, N7
+  \item[āsanaṃ kumbhaka citrāṃ] J16, N7,
   \item[āsanaṃ kumbhaka citaṃ] N18
-\item[āsanaṃ kubhakaṃ citraṃ] N2 (appears before kriyāyuktasya....)
+\item[āsanaṃ kubhakaṃ citraṃ] N2, 
 \item[asanaṃ kumbhakaṃ citraṃ] N3, N9
   \item[(unclear/unable to read)] V13
-\item[(unavailable)]        C5, N22, R1
+\item[(unavailable/illegible)]    C5, N6, N22, R1, T, G1, Ba3, P16, P22,
     \end{description}
@@
   \begin{description}
- \item[mudrādikaraṇāni ca] A1, B2, C3, C9, J7, J8pc, J10, J12, J15, J16, J17, LD1, N17, P8, N14, V12, N7, N9, N10, P1("kā" changed to "ka" later), P5, P7, R2, P9
+  \item[mudrādikaraṇāni ca] A1, B2, C3, C9, J7, J8pc, J10, J12, J15, J16, J17, LD1, N17, P8, N14, V12, N7, N9, N10, P1, P5, P7, P9, P10, P24, R2, R4, 
   \item[mudrānikaraṇāni ca] LD2
@@
 \item[vicitrīkaraṇāni ca]    J2,
-\item[mudrādi pavanakriyā]        B1, B3, Bo1, C2, J5, J8ac, J13, R3, V3, N1, N20, P2, P4
+  \item[mudrādikaraṇakriyā] P27,
+\item[mudrādi pavanakriyā]    B1, B3, Bo1, C2, J5, J8ac, J13, R3, V3, N1, N20, P2, P4, P12, P17, 
 \item[mudrādi pavanakriyāḥ]   N26
-\item[mudrākhyaṃ karaṇaṃ tathā] Ba, Ba5, Bo2, Bo3, C1, C3 ("mudrādikāraṇāni ca" is also given), C6, C7,
- C8, J1, J3, J11, L1, M1, V19, R5, N13, N16, N18, N19, V10, V22, N5, N11, N12, N21, N23, N25, Tue, P3, P6
+\item[mudrākhyaṃ karaṇaṃ tathā] Ba, Ba5, Bo2, Bo3, C1, C3, C6, C7,  C8, J1, J3, J11, L1, M1, P11, P14, P15, P18, P20, P23, P25, P26, V19, R5, N13, N16, N18, N19, V10, V22, N5, N11, N12, N21, N23, N25, Tue, P3, P6, P21,
 \item[mudrākhaṃ karaṇaṃ tathā]   N24
 \item[mudrākhyaṃ karaṇaṃ bhavet]   N3
-\item[mudrākhyakaraṇaṃ tathā]        Ko, N4a, N8
+\item[mudrākhyakaraṇaṃ tathā]    Ko, N4a, N8, P13, 
 \item[mudrādi karaṇaṃ tathā]    K1,
@@
 \item[mudrakhecaraṇīṃ tathā]    J14(given at 55)
-\item[mudrakhecariṇīṃ tathā]        N2 (appears before kriyāyuktasya....)
+\item[mudrakhecariṇīṃ tathā]    N2, 
 \item[(unclear/unable to read)]    V13
-\item[(unavailable)]         C5, N22, R1
+\item[(unavailable)]     C5, N6, N22, R1, G11, T, G1, Ba3, P16, P22,
     \end{description}
@@
 \item[atilavaṇaṃ tilapiṇḍā?]   N14
+\item[atilavaṇatilapiṇḍakadaśanaṃ]  P10, P24, 
 \item[atilavaṇaṃ tilaṃ pinḍa]    J8ac
@@
 \item[atilavaṇaṃ tilapiṇyākadarśanam]    N26
-\item[atilavaṇam amlayuktaṃ]        A1, B3, Ba, Ba5, Bo3, V22, V10, N1, N4a, N11 (original letters "...sacapalaṃ" striked off and "...amlayuktaṃ"), N13 mentioned in bottom margin), Tue, P2("amla" in margin), R2, R5(some doubt in "amla")
+\item[atilavaṇam amlayuktaṃ]    A1, B3, Ba, Ba5, Bo3, V22, V10, N1, N4a, N11, N13, Tue, P2, P12, P14, P17, P18, P25, P26, R2, R5
 \item[atilavaṇama malayuktaṃ]    C2,
+\item[atilavaṇāmalpayuktaṃ]    P13, 
 \item[atilavaṇamalayuktaṃ]   P4, R3
@@
 \item[atilavaṇam ca apramitaṃ]    V8
-\item[atilavaṇam cāpramitaṃ]   P6
-\item[atilavaṇaṃ ca tilapiḍaṃ]   N7
+\item[atilavaṇam cāpramitaṃ]   P6, P27, 
+\item[atilavaṇaṃ ca tilapiḍaṃ]   N7,
 \item[atilavaṇam pramitaṃ]    J12, K1,
@@
 \item[atilavaṇasavapalalaṃ]   N25
+\item[atilavaṇāsavasalilaṃ]   P21, 
+\item[atilavaṇaṃ sapalalaṃ]   P20,
 \item[atilavaṇaṃ sala(pa?)laṃ]   N8
@@
 \item[atilavaṇāsavapalalaṃ]  C1, C7, L1, N5
+\item[atilavaṇāsavapalaṃla]  P23, 
 \item[atilavaṇātiproktaṃ]    J2,
 \item[atilavaṇādiyutaṃ]    J4,
+\item[atilavaṇādi prayuktaṃ] P15,
+\item[atilavaṇādi puttaṃkdaśanaṃ]    P11,
 \item[atilavaṇādyaptataptaṃ]    J5,
@@
 \item[atilavaṇetilapiṃḍakvadaśanaṃ]    J7
-\item[atilavaṇaprasūtaṃ]         N2
+\item[atilavaṇaprasūtaṃ]     N2, 
 \item[atilavaṇaṃ kavalaṃ cāpramitaṃ piṇḍa]   N18
 \item[atīlavaṇāmlapṛktaṃ]   N23
-\item[(illegible/unavailable)]  C5, N16, P3
+\item[(illegible/unavailable)]  C5, N6, N16, P3, G11, T, G1, Ba3, P16, P22,
 
@@
 
-\item[śākotkaṭaṃ] A1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C3, C4, C6, C8, C9, J2, J3, J4, J6, J10, J11, J12, J13, J14, J15, J17, K1, Ko, L1, LD1, LD2, V6, V19, V22, V4, V10, V13, N2, N3, N4a, N7, N11, N13, N18, N19, N21, N24, N25, Tue, P2, P5, P6, P7, P8, R1, R2, R3, R4, R5, P9
+\item[śākotkaṭaṃ] A1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C3, C4, C6, C8, C9, J2, J3, J4, J6, J10, J11, J12, J13, J14, J15, J17, K1, Ko, L1, LD1, LD2, V6, V19, V22, V4, V10, V13, N2, N3, N4a, N7, N11, N13, N18, N19, N21, N24, N25, Tue, P2, P5, P6, P7, P8, P9, P10, P11, P12, P13, P14, P15, P17, P18, P20, P21, P23, P24, P26, R1, R2, R3, R4, R5,
 \item[śākotkaṭhaṃ]    V8
 \item[śākotkavaṃ]    J5,
-\item[śākotkaṭa]        Bo2, J8pc, J16, N1, N9, N26(??kotkaṭa)
+\item[śākotkaṭa]    Bo2, J8pc, J16, N1, N9, N26, P27,
+\item[śākoktaṃ]    P25, 
 \item[śākotkarṭeṃ]  P4
@@
 \item[syacotkaṭaṃ] N12
-\item[(illegible/unavailable)]        C5, N17, N8, N16, P3
+\item[(illegible/unavailable)]    C5, N6, N17, N8, N16, P3, G11, T, G1, Ba3, P16, P22,
 \item[kotkaṭaṃ]  N5
@@
 * * *
-\item[varjyaṃ] A1, B1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C4, C7, C9, J3, J6, J13, J15, K1, L1, LD2, V3, V6, V19,
- V22, N1, N4a, N5, N11, N12(va?jyaṃ), N13, N22, N25, Tue, P2, P4, P6, P7, R3, R4, R5
+\item[varjyaṃ] A1, B1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C4, C7, C9, J3, J6, J13, J15, K1, L1, LD2, V3, V6, V19, V22, N1, N4a, N5, N11, N12(va?jyaṃ), N13, N22, N25, Tue, P2, P4, P6, P7, P12, P14, P17, P18, P20, P21, P23, P27, R3, R4, R5
 \item[varjya] V8, V10
 \item[varjyāṃ] J16
-\item[varj(j)aṃ] J7, J10, J12, J17, M1, N9, R1
-\item[varjjyaṃ]  P5
+\item[varj(j)aṃ] J7, J10, J12, J17, M1, N9, P10, P26, R1
+\item[varjjyaṃ]  P5, P24, 
 \item[varj(j)itaṃ] J1, Bo2, J8, LD1, V3, N18, N20, N26
@@
 \item[vayīṃ]   P8
-\item[duṣṭaṃ] C6, C8, J4, J5, J11, J14, Ko, V13, N2, N21, N23, N24
+\item[duṣṭaṃ] C6, C8, J4, J5, J11, J14, Ko, V13, N2, N21, N23, N24, P13, P15, P25, 
 \item[du(ṣl?)aṃ]  N3
@@
 \item[durjvaṃ(?)] Ba,
-\item[(illegible/unavailable)] C5, N17, V1, V4, N8, N14, N16, P3
+\item[(illegible/unavailable)] C5, N6, N17, V1, V4, N8, N14, N16, P3, G11, T, G1, Ba3, P11, P16, P22,
     \end{description}
@@
   \begin{description}
- \item[mudgādi divyam udakaṃ ca] A1, B2(dīvyam), Ba, Ba5, Bo3, C6("dgā" is missing), C8, J5, J7, J10, J11, J12, J16, J17, K1, L1pc, LD1, LD2, N17, V1, V6, V8, V22, V4, V13, N1 ("mudgā" correction in folio margin), N10, N12, N13, N14(not sure for "dgā" and "ṃca"), N20, N22, N23(something after "mudgādi" but perhaps it is cancelled), N24(is it "mudrā"?), N26, Tue, P1, P6, P8, R1, R2, P9
-\item[mudgādi divyam udakaṃ]   N18
+  \item[mudgādi divyam udakaṃ ca] A1, B2(dīvyam), Ba, Ba5, Bo3, C6("dgā" is missing), C8, J5, J7, J10, J11, J12, J16, J17, K1, L1pc, LD1, LD2, N17, V1, V6, V8, V22, V4, V13, N1, N10, N12, N13, N14, N20, N22, N23, N24, N26, Tue, P1, P6, P8, P9, P14, P17, P18, P23, P25, P26, P27, R1, R2,
+\item[mudgādi divyam udakaṃ]   N18, P11,
 \item[mudgādi divyam udakaṃ ka]   N4a
@@
 \item[mudgādi divyam udakaṃ yamīndra]    V3, N2
-\item[mudgādi divyam udakaṃ hi]        C3, J14, J15, N9, P5
+\item[mudgādi divyam udakaṃ hi]    C3, J14, J15, N9, P5, P10, P15, P24, 
 \item[mudgādi pivyam udakaṃ ca]    J2,
@@
 \item[mudgādivyam udakaṃ ca]    C4, J8, M1, V10, N3
-\item[mudgādi divyagaditaṃ ca]   N7
+\item[mudgādi divyagaditaṃ ca]   N7,
 \item[mudgādi divyamuktaṃ]    Ko,
-\item[mudgādi cālpam udakaṃ ca]        B1, B3, Bo1, Bo2, C1, C2, C7, V19, N5, N11 ("ca" in margin), N21, P2(not sure of "cā"), P4, R3, R4
+\item[mudgādi cālpam udakaṃ ca]    B1, B3, Bo1, Bo2, C1, C2, C7, V19, N5, N11, N21, P2, P4, P12, P13, P20, P21, R3, R4
 \item[mudgādi cālyam udakaṃ ca]    L1ac, N8(no "ca")
@@
 \item[mudrādi divyam udakaṃ ca mudakaṃ]    N19
-\item[(illegible/unavailable)]       C5, J4, P3
+\item[(illegible/unavailable)]       C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22,
     \end{description}
@@
   \begin{description}
- \item[munīndrapathyaṃ (or: munindra-)] B1, B2, B3, Bo1, C1, C2, C4, C7, J3, J5, J6, J13, J16, J17, K1, L1, LD2, V1, V19, N1, N5, N11, N19, N21, N22, P2, P4, R4
+  \item[munīndrapathyaṃ (or: munindra-)] B1, B2, B3, Bo1, C1, C2, C4, C7, J3, J5, J6, J13, J16, J17, K1, L1, LD2, V1, V19, N1, N5, N11, N19, N21, N22, P2, P4, P12, P13, P17, P20, P21, P23, P26, R4
 \item[munīdrapathyaṃ]   N8, R3
 \item[munīndrapathaṃ]   N25
-\item[mahendrapathyaṃ]  N9, P5
- \item[yamīndrapathyaṃ (or: yamindra-)] A1, Ba, Ba5, Bo3, C6, C8, C9, J7, J8, J10, J11, J14, LD1, M1, N17, V6, V22, V3, V4 (pathyaṃ missing), V10, V13, N2, N3, N4a, N12, N13, N14, N20, N26, Tue, P1, P6, P7, P8, R2, R5, P9
+\item[mahendrapathyaṃ]  N9, P5, P10, P24, 
+  \item[yamīndrapathyaṃ (or: yamindra-)] A1, Ba, Ba5, Bo3, C6, C8, C9, J7, J8, J10, J11, J14, LD1, M1, N17, V6, V22, V3, V4 (pathyaṃ missing), V10, V13, N2, N3, N4a, N12, N13, N14, N20, N26, Tue, P1, P6, P7, P8, P14, P18, P25, P27, R2, R5, P9
 \item[yamīṃdrapraparyya]   N24
+\item[yamīḍyapathyaṃ] P15,
 \item[yatīṃdrapathyaṃ]    J2, N10, N23
+\item[yavatīṃdrapathyaṃ]    P11,
 \item[yamendrapathyaṃ]    C3,
@@
 \item[munidra pathayam]    J12,
-\item[munīdramukhyaiḥ]        N7
+\item[munīdramukhyaiḥ]    N7,
 \item[muṃhanīdrapathya]    J15
@@
 \item[paramaṃ hitaṃ hi]    Ko,
-\item[(illegible/unavailable)]   C5, J4, P3
+\item[(illegible/unavailable)]   C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22,
     \end{description}
@@
 \item[iṣṭaṃ sumadhuraṃ]    A1, J3, J6, J10, N18(appears after one verse), P7, R2, P9
+\item[iṣṭaṃ sumadhura]    P11,
 \item[iṣṭaṃ samadhuraṃ]    C9, J7, J17, LD2, N17, V19, V4, LD1, P1, P8, R1
-\item[iṣṭaṃ saṃmadhuraṃ]   N7
+\item[iṣṭaṃ saṃmadhuraṃ]   N7,
+\item[iṣṭaṃ samakaraṃ]   P10, 
 \item[daṣṭaṃ samudhūraṃ]    J16,
@@
 \item[pathyaṃ sumadhuraṃ]    J12,
-\item[puṣṭaṃ sumadhuraṃ]        Ba5, Bo3, N4a, N13, Tue
+\item[puṣṭaṃ sumadhuraṃ]    Ba5, Bo3, N4a, N13, P14, P18, Tue
 \item[puṣṭaṃ samadhuraṃ]   R5
@@
 \item[mṛṣṭaṃ samadhuraṃ]    C3,
-\item[miṣṭaṃ samadhuraṃ]        J8, Ba, C4, N20(appears after one verse), N26(appears after one verse), P5
-\item[miṣṭaṃ sumadhuraṃ]        Bo2, C6, J1, J11, J14, N2, N16(appears after one verse), N19(miṣtaṃ??), N21(appears after one verse), N22, N23(appears after one verse), R4(appears after one verse)
-\item[miṣṭaṃ samudhuraṃ]        J5, J15,[%P: How is this reading different from the reading above the reading above this?]
-\item[miṣṭaṃ sūmadhuraṃ]  P6(appears after one verse)
+\item[miṣṭaṃ samadhuraṃ]    J8, Ba, C4, N20, N26, P5, P24, 
+\item[miṣṭaṃ sumadhuraṃ]    Bo2, C6, J1, J11, J14, N2, N16, N19, N21, N22, N23, P13, P15, P21, P25, P27, R4
+\item[miṣṭaṃ samudhuraṃ]    J5, J15,
+\item[miṣṭaṃ sūmadhuraṃ]  P6
 \item[miṣṭaṃ sumadhura]   N25
@@
 \item[maṃ(ṣṭ?)aṃ samudhuraṃ]  N9
-\item[sumiṣṭaṃ madhuraṃ]        N12(appears after one verse)
-\item[uṣṇaṃ sumadhuraṃ]        B1, Bo1, C2, C7, L1, J13, N1 (seems "uṣṇaṃ"), P2
+\item[sumiṣṭaṃ madhuraṃ]    N12
+\item[uṣṇaṃ sumadhuraṃ]    B1, Bo1, C2, C7, L1, J13, N1, P2, P17, P23, 
+\item[uṣṇaṃ sumudhuraṃ]    P20,
+\item[uṣṇaṃ sumadhura]    P12, 
 \item[uṣṇaṃ samadhuraṃ]    C1,
@@
 \item[tiṣṭaṃ tu madhuraṃ]  N24
-\item[(illegible/unavailable)]   C5, J4, P3
-        \end{description}
-\end{marma}
-
+\item[(illegible/unavailable)]   C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22, P26,
+    \end{description}
+ \end{marma}
+
+Notes:
+    • P3, R1, has not 2nd chapter
+    • Folios of P4 are scattered 
+    • P16 has only 3 folios of chapter 2 at the end
+    • P22 has only 3 folios of chapter 1
+    • R3 has only 1 folio of chapter 2
+    • 1 & 2 Chapters are mixed in R4
+    • R5 has only one folio of chapter 2
 \begin{marma}[hp02_003]
 \sthana{2.3b}
-\item[jīvanam] Ba, Ba5, Bo2, Bo3, C1, C6, J1, J2, J3, J6, J12, J13, J14, J16, N17, V1, N4a, N7, N8, N9, N10, N11, N12, N13, N16, V16, V17, V10, N18, N21, V21, Tue, N23, V6, V7, V20, V28, V18
+\item[jīvanam] Ba, Ba5, Bo2, Bo3, C1, C6, J1, J2, J3, J6, J12, J13, J14, J16, N17, P1, V1, N4a, N7, N8, N9, N10, N11, N12, N13, N16, P10, P13, P14, V16, V17, V10, N18, N21, P18, P20, P25, P26, P27, R5, V21, Tue, N23, V6, V7, V20, V28, V18
 \item[jīvaṃnam] N14
-\item[jīvitam] A1, B1, B2, B3, Bo1, C2, C3, C4, C8, J4, J5, J7, J8, J10, J11, LD1, LD2, N1, N2, N5, V8, N20, N22, V14, V3, N24, N25, V15, V5, V22, V2, V12, V13, V9
+\item[jīvitam] A1, B1, B2, B3, Bo1, C2, C3, C4, C8, J4, J5, J7, J8, J10, J11, LD1, LD2, N1, N2, N5, V8, N20, N22, P2, P6, P9, P17, P21, P23, V14, V3, N24, N25, P12, P15, R1, R3, V15, V5, V22, V2, V12, V13, V9, G11, Ba3, G1,
+\item[jīvatam] P5, P24,
 \item[jivitam] J15,
 \item[jīvita] N26
+\item[jīviti] P11, 
 \item[jīvitum] J17, N19
 \item[jītavim] N3
-\item[jīvo na] Ko
-\item[(unavailable/illegible)] C5, C9,
+\item[jīvo na] Ko, P7, P8, 
+\item[(unavailable/illegible)] C5, C9, N6, T, P16, P22,
 
@@
 \sthana{2.3d}
-\item[nibandhayet] Ba, Bo2, C6, V1, J2, J4, J6, J8, J11, J14, N12, N16, N19, N22, V14, V3, V13
-\item[nirundhayet] A1, B1, B2, B3, C1, C3, C9, J5, J10, J13, J15, J17, LD1, LD2, N17, V19, W4, N2, N7, N9, N10, N14("ye" unclear),V16, V17, N20, V21, N23, N26, V20, V5, V2, V18, V12, V9
-\item[nirodhayet] Ba5, Bo1, Bo3, C2, C4, C8, J1, J3, J6pc, J7, V19, N1, N4a, N5, N8(seems unclear), V8, N11, N13, V10, N18(“ye” is doubtful), N21, Tue, N24, N25, V6, V7, V15, V28, V22
+\item[nibandhayet] Ba, Bo2, C6, V1, J2, J4, J6, J8, J11, J14, N12, N16, N19, N22, P11, P15, P25, V14, V3, V13
+\item[nibaddhayet] G11, P6, P27,
+\item[nirundhayet] A1, B1, B2, B3, C1, C3, C9, J5, J10, J13, J15, J17, LD1, LD2, N17, P1, V19, W4, N2, N7, N9, N10, N14, P5, P8, P9, P10, P23, V16, V17, N20, V21, N23, N26, P2, P7, P26, V20, V5, V2, V18, V12, V9, Ba3
+\item[nirodhayet] Ba5, Bo1, Bo3, C2, C4, C8, J1, J3, J6pc, J7, V19, N1, N4a, N5, N8, R3, V8, N11, N13, V10, N18, N21, P13, P14, P17, P18, P20, Tue, N24, N25, P12, P21, R5, V6, V7, V15, V28, V22
+\item[nirudhayet] P24,
+\item[niruddhayet] R1, 
 \item[nisaṃdhayet] J16,
 \item[nirodhanaṃ] J12,
-\item[nibandhanāt] V19, N3
+\item[nibandhanāt] V19, N3, G1
 \item[nibadhnayāt] Ko,
-\item[(unavailable/illegible)] C5,
+\item[(unavailable/illegible)] C5, N6, T, P16, P22,
   \begin{description}
@@
 \sthana{2.4d}
-\item[kāyasiddhiḥ] A1, B1, B2, C2, C6, J8, J14, V1, N1, N8, N11, V17, N24, V20, V15, V9
+\item[kāyasiddhiḥ] A1, B1, B2, C2, C6, J5, J8, J14, V1, N1, N8, N11, P2, P12, P17, P26, R3, V17, N24, V20, V15, V9, G11, G1
 \item[kāyāsiddhiḥ] J2, N3,
 \item[kāyāsuddhi] V6
+\item[kāyāsiddhi] P6, P27, 
 \item[kayasiddhiḥ] N20
-\item[kāyasiddhiḥ] J5,
 \item[kāyaśuddhiḥ] N23
-\item[kāryasiddhiḥ] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C8, C9, J3, J4, J6, J10, J11, J13, J17, Ko, LD1, N17, V1, W4, N2, N4a, N5, N9, N12, N13, V16, V10, N19, N21, Tue, N25, N26, V70, V5, V28, V22, V2, V18, V12, V13
+\item[kāryasiddhiḥ] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C8, C9, J3, J4, J6, J10, J11, J13, J17, Ko, LD1, N17, P10, V1, W4, N2, N4a, N5, N9, N12, N13, V16, V10, N19, N21, P1, P5, P6, P7, P8, P9, P11, P13, P14, P15, P18, P20, P21, P23, P24, P25, R1, Tue, N25, N26, V70, V5, V28, V22, V2, V18, V12, V13, Ba3
+\item[kāryasiddhaḥ] R5,
 \item[kāryasiddhiṃ] J15, 
@@
 \item[kāyāśuddhiḥ] J7,
-\item[(unavailable/illegible)] C5,
+\item[(unavailable/illegible)] C5, N6, T, P16, P22,
 
@@
 \item[suṣumnāsu snighdhā] J4, V1, V15
-\item[suṣumnānāḍisthā] Ba5, N13, Tue, 
-\item[suṣumṇānāḍīsthā] N4a, N21(“ṇā” doubtful), 
+\item[suṣumnāḥ snighdhā] P15, 
+\item[suṣumnānāḍisthā] Ba5, N13, P13, P14, P20, Tue, 
+\item[suṣumṇāṃ nāḍisthāḥ] P21, 
+\item[suṣumṇānāḍīsthā] N4a, N21, P18,  
 \item[suṣumṇānāmadhysthā] V2
+\item[suṣumnā nāḍīsthaṃ] G1
 \item[suṣumṇānāḍīstho] N24
@@
 \item[sukhumnāṃtarasthā] V19
-\item[sukhumnāṃtarālasthā] B1,
-\item[suṣumnāṃtarālasthā] B3, Bo1, Bo2, C1, C2, J3, J6, J13, N1, N8, N11
+\item[suṣumnāṃtarālasthā] B3, Bo1, Bo2, C1, C2, J3, J6, J13, N1, N8, N11, P2, P17, P23, 
+\item[suṣumnāṃtarālaschā] P12, 
 \item[suṣumnāṃtarālasyā] J1
@@
 \item[suṣumnā pāśvaṃsthā] A1, V20
-\item[yathā suṣumṇā pārśvaḥsthā] Ba
-\item[yathā suṣumṇā pārśvasthā] C6, C8, J7, N23
+\item[yathā suṣumṇā pārśvaḥsthā] Ba,
+\item[yathā suṣumṇā pārśvasthā] C6, C8, J7, N23, P7, P25, P26, 
 \item[suṣumṇāpañcasthā] N19
 \item[yathā suṣumnā mukhasthā] V17
+\item[yathā suṣumnā sumalāḥ] P27,
 \item[yathā suṣumnā nāḍisthamala] V10
 \item[yathā suṣumnņayā śvaḥsthā] V8, V21
+\item[yathā suṣumņā sustakcya?] P11, 
 \item[suṣumņā sustabdhāḥ] N12
-\item[yathā suṣumnā susvasthā][a] J5, N22, V14, V3, N25, N3
+\item[yathā suṣumnā susvasthā] J5, N22, N26, V14, V3, N3
 \item[suṣumņā suṣasthā] N20
-\item[suṣumnņayā ścasyā][b] V6, V13
+\item[suṣumnņayā ścasyā] V6, V13
 \item[suṣumnāvasthāyaṃ] B1, B2
 \item[suṣumnānāḍibhyāṃ] V7
+\item[suśumnāsusvasthā] G11
 \item[suṣumnāsvasthāyan] N2
 \item[suṣama susvasthā] J8
-\item[suṣumnāsui svasthā] J8ac %H: I didn’t find this reading in ms. I don’t know who added this.
+\item[suṣumnāsui svasthā] J8ac,
 \item[suṣumnāsaṃsvasthā] J2,
-\item[sukham avasthāya] C3, C9, J10, J12, J17, LD2, N17, W4, N7, N9, V16, V18, V12, V9
+\item[suṣumṇā sūmalā] P6,
+\item[sukham avasthāya] C3, C9, J10, J12, J17, LD2, N17, P1, P5, P24, W4, N7, N9, V16, V18, V12, V9, Ba3
+\item[sukhamalāsyāya] P9, 
 \item[sukham avasthāyaḥ] J16,
-\item[sukham avasthāyā] N14
+\item[sukham avasthāyā] N14, P8, 
 \item[sukham avasthāpyā] LD1
+\item[sukham avaschāya] R1,
 \item[sukham avaschāṃya] N10
+\item[sukha........] P10,
 \item[suṣam avasthāya] J8pc, N18,
@@
 \item[suṣumnāṃtvarayāpalaḥ] N25
-\item[(unavailable/illegible)] C5, Ko,
+\item[(unavailable/illegible)] C5, Ko, N6, T, P16, P22, R5,
 
@@
 \begin{description}
-\item[Yes] J17
+\item[Yes] J17, R1, 
 \item[No] A1, B1, B2, B3, Ba, Ba5, Bo1 Bo2, C1, C2, C3, C4, C5, C6, C8, C9, J1, J2, J3, J5, J10, J11, J12, J14, J15, J16, LD2, N17, V1, J4, J6, K7.
-J8, J13, Ko, LD1, W4, V19, N1, N2, N3, N4a, N5, N7, N8, N9, V28, V8, N10, N11, N12, N13, N14, N16, V16, V17, V10, N18, N19, N20, N21, N22, V14, V21, V3, Tue, N23, N24, N25, N26, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9
+J8, J13, Ko, LD1, W4, V19, N1, N2, N3, N4a, N5, N6, N7, N8, N9, V28, V8, N10, N11, N12, N13, N14, N16, P5, P10, V16, V17, V10, N18, N19, N20, N21, N22, P1, P6, P7, P8, P9, P11, P12, P13, P14, P15, P16, P17, P18, P20, P21, P22, P23, P24, P25, P26, P27, V14, V21, V3, Tue, N23, N24, N25, N26, P2, R5, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9, G11, T, Ba3, G1
     \end{description}
@@
 \begin{description}
-\item[Yes] B1, B3, C9 (2.13), V19, N1 (2.13), N5, N10 (some change), N12, N21(some changes), N25, V28
-\item[No] A1, B2, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C4, C5, C6, C8, J1, J2, J3, J10, J13, J17, J4, J5, J6, J7, J11, J12, J14, J15, J16, Ko, LD1, LD2, N17, V1, J8, W4, N2, N3, N4a, N7, N8, N9, V8, N11, N13, N14, N16, V16, V17, V10, N18, N19, N20, N22, V14, V21, V3, Tue, N23, N24, N26, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9
+\item[Yes] B1, B3, C, V19, N1, N5, N10 , N12, N21, N25, P12, P13, P26, P17, P20, P21, V28
+\item[No] A1, B2, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C4, C5, C6, C8, J1, J2, J3, J10, J13, J17, J4, J5, J6, J7, J11, J12, J14, J15, J16, Ko, LD1, LD2, N17, P1, V1, J8, W4, N2, N3, N4a, N6, N7, N8, N9, V8, N11, N13, N14, N16, P5, P6, P8, P15, P18, V16, V17, V10, N18, N19, N20, N22, V14, V21, V3, Tue, N23, N24, N26, P2, P7, P9, P10, P11, P14, P16, P22, P23, P24, P25, P27, R1, R5, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9, G11, T, Ba3, G1
     \end{description}
@@
 \sthana{2.12cd}
-\item[uttānaṃ cottame prāṇarodhe padmāsane muhuḥ] Ba, C1, C6, V14
+\item[uttānaṃ cottame prāṇarodhe padmāsane muhuḥ] Ba, C1, C6, P25, V14
+\item[uttānaṃ cottame prāṇe rodhe padmāsane muhuḥ] P7, 
 \item[uttānaṃ cottame prāṇe stebdhe? padmāsane muhuḥ] C8,
@@
 \item[uttiṣṭhāṃty uttame prāṃṇā baddhe padmāsane uttama] J16
-\item[uttiṣṭhaty uttame prāṇē baddhe padmāsanae dṛḍhe] C2
+\item[uttiṣṭhaty uttame prāṇē baddhe padmāsanae dṛḍhe] C2, P17, 
 \item[uttiṣṭhaty uttame prāṇaḥ baddhe padmāsanae dṛḍhe] J13,
 \item[uttiṣṭhaty uttame prāṇaṃ baddhā padmāsane dṛḍhe] B1
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane dṛḍhe] B2, B3,
-\item[uttiṣṭhaṃty uttame prāṇā baddhe padmāsane muhuḥ] LD2, J10
-\item[uttiṣṭhaṃty ūttame prāṇā baddhe padmāsane muhuḥ] J15,
+\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane dṛḍhe] B2, B3, P2,
+\item[uttiṣṭhaty uttame prāṇa baddhe padmāsanaṃ dṛḍhaṃ] Ba3
+\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane dṛḍhe] P13, 
+\item[uttiṣṭhaty uttame prāṇaḥ badhe padmāsane dṛḍhe] P27,
+\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane sadṛḍhe] P12,
+\item[uttiṣṭhaṃty uttame prāṇā baddhe padmāsane muhuḥ] LD2, J10, P1, P8, P9, P10, P24, 
+\item[uttiṣṭhaṃty ūttame prāṇā baddhe padmāsane muhuḥ] J15, P5, R1, 
 \item[uttiṣṭhaṃty uttame prāṇā baddha padmāsane muhuḥ] C3, N10
@@
 \item[uttiṣṭhaty uttame prāṇē baddhe padmāsanaṃ muhuḥ] J8pc
+\item[uttiṣṭhaty uttame prāṇē baddhapadmāsane muhuḥ] P26, 
 \item[uttiṣṭhaty uttame prāṇē baddha padmāsane haye] Bo1
@@
 \item[uttiṣṭhaty uttame prāṇa rodhe padmāsana sthite] C4, N25
+\item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthite] P21, P23, 
 \item[uttiṣṭhaty uttame prāṇāṃ rodhe padmāsane sthite] V19
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthite] V5, V28
+\item[uttiṣṭhaty uttame prāṇo rodhe padmāsane sthite] P20, V5, V28
 \item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthitaḥ] J6, N2
-\item[uttiṣṭaty uttame prāṇa rodhe padmāsanasthite]  N5
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsanaṃ muhuḥ] N3, V15
-\item[uttiṣṭhaty uttāme prāṇarodhe padmāsanaṃ muhuḥ] J5, J11,
+\item[uttiṣṭaty uttame prāṇa rodhe padmāsanasthite]  N5, P6, 
+\item[uttiṣṭhaty uttame prāṇa rodhe padmāsanaṃ muhuḥ] N3, , P15, V15
+\item[uttiṣṭhaty uttāme prāṇarodhe padmāsanaṃ muhuḥ] J5, J11, G1
 \item[uttiṣṭhaty uttāme prāṇarodhe padmāsano muhuḥ] J7
@@
 \item[uttiṣ(?)yaty uttame prāṇa rodho padmāsanaṃ muhuḥ] N24
+\item[uttiṣṭhaty uttame prāṇaṃ rodhe padmāsanaṃ muhuḥ] P11, 
 \item[uttiṣṭhaty uttame kārye rodhe padmāsanae muhu] Ko,
@@
 \item[uṭṭiṣṭhaṃ cottare prāṇā baddhe padmāsane muhuḥ] N17
+\item[uttiṣṭhatyuttame prāṇarodhe padmāsane muhuḥ] G11
 \item[uttiṣṭaty attame prāṇe baddhe padmāsane dṛḍhe] N1
@@
 \item[uṭṭiṣṭhnty uttame prāṇā baddhe padmāsane muhuḥ] LD1,
-\item[(illegible/unavailable)]   Ba5, Bo3, C5, N4a, V22, V12, V9
+\item[uttame sthānaṃ āpnoti tato vāyuṃ nibandhayet] P14, P18, 
+\item[(illegible/unavailable)]   Ba5, Bo3, C5, N4a, N6, P22, V22, V12, V9, T, P16, R5, 
   \begin{description}
@@
 \sthana{2.19c}
-\item[kṛśatā] A1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C6, C8, J1, J4, J5, J6, J7, J11, J13, J14, V19, N1, N2, N4a, N5, N8, N11, N12, N13, N16, N18, N19, N22, C4, V17, V10, V21, Tue, N23, N26, V7, V20, V5, V28, V22, V2, V18
+\item[kṛśatā] A1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C6, C8, J1, J4, J5, J6, J7, J11, J13, J14, V19, N1, N2, N4a, N5, N8, N11, N12, N13, N16, N18, N19, N22, P2, P5, P6, P7, P11, P12, P13, P14, P15, P17, P18, P20, P23, P25, P26, P27, C4, V17, V10, V21, Tue, N23, N26, V7, V20, V5, V28, V22, V2, V18, G11, G1
 \item[kṛsatā] Ko, J3,  
 \item[kṛśate] J10pc
-\item[kriyate] C3, J10, J12, J16, J17, LD1, N7, N9, N10, V12, C9, J2, N17, LD2, V19
+\item[kriyate] C3, J10, J12, J16, J17, LD1, N7, N9, N10, P1, P8, P9, P10, R1, V12, C9, J2, N17, LD2, V19, Ba3
 \item[kṛyate] J15, N14
-\item[kṛśatāṃ] J8, V8, N21, V3, N25, V6
+\item[kṛśatāṃ] J8, V8, N21, V3, N25, P21, V6
 \item[kṛṣitāṃ] N24
@@
 \item[krūratā] J2
-\item[(illegible/unavailable)]   B1, B2, C5, V16, V15, V13, V9
+\item[(illegible/unavailable)]   B1, B2, C5, N6, V16, V15, V13, V9, T, P16, P22, P24, R5,
   \begin{description}
@@
 \sthana{2.22a}
-\item[medaḥśoṣādikaṃ pūrvaṃ] V1
+\item[medaḥśoṣādikaṃ pūrvaṃ] P6, V1
 \item[medaḥśleṣmādikaṃ surve] V6
 \item[medaśokādikaṃ pūrvaṃ] N22
-\item[medaḥśleṣmādi nāśārthe] A1, C3,
+\item[medaḥśleṣmādi nāśārthe] A1, C3, P26, 
 \item[medaḥśleṣmādi nāśārtha] LD2,
 \item[medaḥśleṣmādi ṣatkarmāṅi] V28
-\item[medaḥśleṣmādi nāśārthaṃ] B1, B3, Bo1, C2, J10, J12, J13, J15, J16, J17, LD1, N7, V21, V20, V12
+\item[medaḥśleṣmādi nāśārthaṃ] B1, B3, Bo1, C2, J10, J12, J13, J15, J16, J17, LD1, N7, N17, P1, P2, P5, P8, P9, P12, P13, P17, R1, V21, V20, V12
+\item[medaḥśleṣmādhi nāśārthaṃ] P27, 
+\item[madaṃ śleṣmādi nāśārthe] P10, 
+\item[nādaśleṣmādinā śāstraṃ] Ba3
 \item[medaḥśleṣmādināśīrṣaṃ] N9
@@
 \item[medaśleṣmādi nāsārthe] N20, N26
-\item[medaśleṣma nivṛtyarthaṃ] Bo2, C1, J1, J3, J6, N2, N8, N11, N16, N18, V15, V5, V2
+\item[medaśleṣma nivṛtyarthaṃ] Bo2, C1, J1, J3, J6, N2, N8, N11, N16, N18, P20, P21, P23, V15, V5, V2
 \item[medaḥśleṣma nivṛtyarthaṃ] N5, C4
 \item[??śleṣma nivṛtyarthaṃ] N25
-\item[medaḥśleṣmādhikaḥ pūrvaṃ] Ba, C8, J5, V10, V14, V13
+\item[medaḥśleṣmādhikaḥ pūrvaṃ] Ba, C8, J5, P7, P11, P14, P18, V10, V14, V13, G11
 \item[medaḥśleṣmādhikaḥ pūrva] J7,
@@
 \item[medaḥ ślṣāṃdhikaḥ pūrvaṃ] C6, J14, 
-\item[medaśleṣmādhikaḥ pūrvaṃ]  J11, N3, N4a, N12, N13, Tue, V22
+\item[medaśleṣmādhikaḥ pūrvaṃ]  J11, N3, N4a, N12, N13, P25, Tue, V22
 \item[medāśleṣmāṣikaḥ pūrvva] N23, V7
@@
 \item[medodhikastu havabhiḥ] Ko,
-\item[(unavailable/illegible)] C5, J4, N19, V9
+\item[(unavailable/illegible)] C5, J4, N6, N19, P15, V9, T, G1, P16, P22, P24, R5,
   \begin{description}
@@
 \item[api bhāvataḥ] C3
-\item[samabhāvataḥ] A1, B1, Ba5, Bo1, Bo3, C9, J1, J3, J6, J10, J11, J12, J13, J14, J15, J17, LD1, LD2, N3, N4a, N7, N8, N10, N11, N12, N13, V10, N20, N21, V14,V3, Tue, N26(perhaps left margin says something), V7, V20, V15, V28, V22, V18, V12, V13
-\item[samabhāvat] J16,
+\item[samabhā] R1,
+\item[samabhāvataḥ] A1, B1, Ba5, Bo1, Bo3, C9, J1, J3, J6, J10, J11, J12, J13, J14, J15, J17, LD1, LD2, N3, N4a, N7, N8, N10, N11, N12, N13, P1, P5, P10, V10, N20, N21, P6, P8, P9, P11, P13, P14, P18, V14, V3, Tue, Ba3, N26, V7, V20, V15, V28, V22, V18, V12, V13, G11
+\item[samabhāvat] J16, N17, 
 \item[samupāgataḥ] J7,
@@
 \item[samabhāvanā] N22
-\item[samabhāvanāt] N22
 \item[smabhāvata]  N14
-\item[sabhāgataḥ] Ba
-\item[samabhāgataḥ] B2, C6, C8, V8, N18, N23, 
+\item[sabhāgataḥ] Ba,
+\item[samabhāgataḥ] B2, C6, C8, P7, V8, N18, N23, P25, P27, 
 \item[samabhāgatā] Ko,
 \item[samabhāgata] V6
-\item[samabhāgikaḥ] B3, C2, N1
+\item[samabhāgikaḥ] B3, C2, N1, P2, P12, P17, 
 \item[samabhārat] J2
-\item[samatā yataḥ]  Bo2, C1, C4, N2, N16, N25, V2
+\item[samahāvata]  P26, 
+\item[samatā yataḥ]  Bo2, C1, C4, N2, N16, N25, P20, P23, V2
+\item[samatā yatāṃ]  P21, 
 \item[samatā yathā]   N5,
 \item[samatāpataḥ] V5
-\item[(unavailable/illegible)] C5, J4, N19
+\item[(unavailable/illegible)] C5, J4, N6, N19, P15, T, G1, P16, P22, P24, R5,
   \begin{description}
@@
 \sthana{2.22c}
-\item[vicitraguṇasandhāyī(i)] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C3, C8, C9, J1, J2, J3, J4, J5, J6, J7, J11, J13, J14, J15, LD1, LD2, V1, W4, J10, N17, J17, J8ac, Ko, N1, N3, N4a, N7, N9, V8, N10, N16, V16, V17, V10, N18(...saṃdhāī, no “y”), N19(...saṃdhāi, no “y”), N21, N22(...saṃdhāī, no “y”), V14, V3, Tue, N23, V7, V15, V5, V28, V22, V2, V18, V12, V13
-\item[vicītraguṇasandhāyi] J16,
+\item[vicitraguṇasandhāyī(i)] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C3, C8, C9, J1, J2, J3, J4, J5, J6, J7, J11, J13, J14, J15, J16, LD1, LD2, P7, V1, W4, J10, N17, J17, J8ac, Ko, N1, N3, N4a, N7, N9, N10, N16, P1, P2, P6, P9, P10, P11, P12, P13, P15, P17, P18, P21, P24, P25, P27, R1, V8, V16, V17, V10, N18, N19, N21, N22, P14, V14, V3, Tue, N23, V7, V15, V5, V28, V22, V2, V18, V12, V13, G11, Ba3
+\item[vicitraguṇasadhāyī] P26,
+\item[vicitraguṇaśandhāyī]  P8, 
+\item[vicītraguṇaṃsandhāyi] P5, 
+\item[vicitraguṇasaṃyāyi] P23,
 \item[vicitraguṇasandhāya]  C6, N12
@@
 \item[vicitraguptasaṃ?jñopi] N5
-\item[vicitraguṇasaṃjñopi] C4, N25
-\item[vicitraguṇasaṃdhāyoga]  N8("vivi?"), N11 ("yogasiddhidāyi" in brackets, then "yi" of "sandhāyi" )
+\item[vicitraguṇasaṃjñopi] C4, P20, N25
+\item[vicitraguṇasaṃdhāyoga]  N8, N11,
 \item[trividhaṃ guṇasandhāyi] B2
-\item[(unavailable/illegible)] C5, N24
+\item[(unavailable/illegible)] C5, N6, N24, T, G1, P16, P22, R5,
   \begin{description}
@@
 \sthana{2.24c}
-\item[tataḥ pratyaharec caitad] J12, J16, J17, N23
+\item[graset pratyāharec caitad] P25, 
+\item[tataḥ pratyāharec caitad] A1, Ba, Ba3, B3, G11, J12, J16, J17, N9, N12, N14, N17, N20, N21, N23, P1, P5, P8, P9, P10, P11, P12, P13, P15, P24, P26, P27, R1,
+\item[tataḥ pratyaharec cetad] N10
+\item[tataḥ pratyaharec caitad] N13, P7, 
+\item[tata pratyaharec caiva] N24,
+\item[punaḥ pratyaharec caitad] B1, B2, P2, P14, P17, P18, Tue,
 \item[punaḥ pratyahare caitad] J13, J14, 
+\item[punaḥ pratyāhare caitad] N19, 
+\item[punaḥ pratyāhare caitā] N22,
+\item[pūnaḥ pratyāharec caityāt] P6,
+\item[punaḥ pratyahared etad] N8, N11, N16, N18, N25, N26, P20, P21, P23, 
 \item[punaḥ pratyaharec caitaḥ] J15,
-\item[(unavailable/illegible)]
+\item[(unavailable/illegible)] N6, N7, T, G1, P16, P22, R5,
   \begin{description}
@@
 \sthana{2.24d}
-\item[prakṣālyaṃ dhautikarma tat] Ba, C6, V21
+\item[prakṣālyaṃ dhautikarma tat] Ba, C6, P7, P25, V21
 \item[prakṣālanaṃ dhautikarma tat] V13
@@
 \item[sodgāraṃ dhautikarmavit] N12
-\item[uditaṃ dhautikarma tat] A1, Ba5, Bo1, Bo3, C9, J10pc, J12, J15, LD1, LD2, N4a, N7, N9, N10, N13, N14("ti" added later, anusvara on "ka" seems cancelled), V16, V17, V10, N20, N21(“u” of uditaṃ doubtful), Tue, N26, V7, V20, V15, V18, V12
+\item[uditaṃ dhautikarma tat] A1, Ba5, Bo1, Bo3, C9, J10pc, J12, J15, LD1, LD2, N4a, N7, N9, N10, N13, N14, N17, P5, P8, P9, P10, P14, P18, P24, V16, V17, V10, N20, N21, Tue, N26, V7, V20, V15, V18, V12
+\item[uditaṃ dhātikarma tat] P13, 
+\item[udita dhaitikarma tat] G11
 \item[aditaṃ dhautikarma tat] J17,
-\item[uditaṃ dhautikarma kṛt] J10, V22
+\item[uditaṃ dhautikarma kṛt] J10, P1, R1, V22
 \item[up(?)itaṃ dhautikarma tat] C3
-\item[abhyāsād dhautikarmavit] B1, B2, C4, J1, J3, J6, J13, V5, V28, V2
+\item[abhyāsād dhautikarmavit] B1, B2, C4, J1, J3, J6, J13, P2, P12, P17, P20, P26, V5, V28, V2
 \item[abhyāsād dhautikarmasu] N25
 \item[abhyāsā dhautikarmavit] N18
+\item[abhyāsā dhautikarma tat] P6,
+\item[abhyāsā dhautikarmavit] P21,
 \item[abhyāso dhautikarmavit] N19
+\item[abhyase dhautikarmavit] P27,
 \item[abhyāse dhautakarmavit] J14,
@@
 \item[duṣālaṃ dhautikarma tat] J4, J11, 
-\item[uṣālaṃ dhautikarma tat] J5,
+\item[duḥkhahaṃ dhautikarma tat] P11, 
+\item[uṣālaṃ dhautikarma tat] J5, P15, 
 \item[abhyāsa dhautakarka tat] N22
-\item[abhyāsād dhauti bhavati karma tat] N8("dhīta" instead of "dhauti"), N11 ("bhavati" in brackets)
+\item[abhyāsād dhauti bhavati karma tat] N8, N11 ("bhavati" in brackets)
 \item[uṃtthānaṃ dhautikarma tat] N23
@@
 \item[?s?vāla dhautikarmakṛt] N3
-\item[(unavailable/illegible)] C5, N16
+\item[(unavailable/illegible)] C5, N6, N16, T, G1, P16, P22, P23, R5,
   \begin{description}
@@
 \sthana{2.25d}
+\item[bādhate yaṃ na] P27,
 \item[chadhyate ca na] J16,
 \item[dhāvaty eva] V1, V8, V3, V2
+\item[dhāvaṃtyeva na] P6, P15, 
 \item[dhāvate ca na] B2
@@
 \item[dhruvntye neva] Ko,
-\item[gachaty eva na] N21
+\item[gachaty eva na] N21, P13, 
 \item[gachaṃty eva na] Bo1
-\item[prayāntyeva na] Ba, N4a (2.26a), Bo2, Bo3, C1, C4, C6, C8, J1, J3, J6, J7, N5, N8, N11, N13, N16, V10, Tue, N25, V28, V22
+\item[prayāntyeva na] Ba, N4a, Bo2, Bo3, C1, C4, C6, C8, J1, J3, J6, J7, N5, N8, N11, N13, N16, P7, P11, P14, P18, P20, P21, P25, V10, Tue, N25, V28, V22
+\item[dhāvatyeva na] G11
 \item[prayānteva na] N23, V7
 \item[śudhyatty eva] A1,
-\item[śudhyaty eva] LD2,
+\item[śudhyaty eva] LD2, P2, P9, 
 \item[śudhyate nātra] J12,
-\item[śudhyate ca na] N10
-\item[śudhyanty eva] B1, C3, C9, J10, J13, J17, N1 (not sure for "eva")
-\item[śudhyant eva] B3,
-\item[śudhyaṃteva na] J15, N9, N14, V16, V17
-\item[śudhyaṃte ca na] LD1, 
+\item[śudhyate ca na] N10, P26, 
+\item[śudhyanty eva] B1, C3, C9, J10, J13, J17, N1, N17, P5, P8, P12, P17, Ba3
+\item[śudhyanteva] B3, P10, 
+\item[śudhyaṃteva na] J15, N9, N14, P24, V16, V17
+\item[śvadhvatyeva na] R1,
+\item[śudhyaṃte ca na] LD1, N7, P1, 
 \item[suddhyaṃtevaṃ na] N18
-\item[śudhyaṃte ca na] N7
 \item[śudhyaṃ sevana] C2,
@@
 \item[śuṣyaṃte sakalāmalāḥ] N20, N26
-\item[(illegible/unavailable)] Ba5,C5, V21, V6, V20, V15, V5, V18, V12, V13
+\item[(illegible/unavailable)] Ba5,C5, N6, V21, V6, V20, V15, V5, V18, V12, V13, T, G1, P16, P22, P23, R5,
   \begin{description}
@@
 \item[udaragatapadārtham udvahatī] J12,
-\item[udaragatapadārtham udvahantī] Ba, C3, J1, J7, Ko, V1, V16, N18, N22, N23, V6, V28, V2
+\item[udaragatapadārthaṃ  utbhavanti] G1
+\item[udaragatapadārtham udvahati] R1,
+\item[udaragatapadārtham udvahantī] Ba, C3, J1, J7, Ko, N10, P6, V1, V16, N12, N18, N22, N23, P20, P27, V6, V28, V2
 \item[udaragatapadārthamūrdhāhatī] J16,
 \item[uragatapadārtham udvahantī] B2,
-\item[udaragatapadārthān udvahaṃti] C6
+\item[udaragatapadārthān udvahaṃti] C6, P7, 
 \item[udaragatapadārtham uşmahantī] V17
-\item[udaragatapadārtham udvamaṃti(ī)] B1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J11, J14, LD1, LD2, N20, N21, V3, N26, V20, V22, V18, V12, V13
-\item[udaragatapadārtham udvavaṃtī] J17,
-\item[udaragatapadārtham udvavaṃtī] J10,
-\item[udaragatapadārtham udvamaṃta] J4,
-\item[udaragatapadārtham udvamaṃte] V15
+\item[udaragatapadārtham udvamaṃti(ī)] B1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J11, J14, LD1, LD2, N7, N13, N16, N16, N20, N21, N26, P2, P8, P9, P11, P12, P13, P14, P17, P18, P21, P25, P26, V3, V20, V22, V18, V12, V13, Ba3
+\item[udaragatapadārtham udvayant] P5,
+\item[udaragatapadārtham uddhayaṃt] P24, 
+\item[udaragatpadārtham udvamaṃtī] N9,  
+\item[udaragatapadārtham udvavaṃtī] J10, J17, P1,
+\item[udaragatapadārtham udvamaṃta] J4, P10, 
+\item[udaragatapadārtham udvamaṃte] P15, V15
 \item[udaragatapadārthām udyamaṃte] N19
@@
 \item[udaragatapadārtham uḍhasaṃtī] N24
+\item[udaragatapadārtham udvapanti] N14, 
 \item[udaragatapadārthaśodhanaṃ syāt] Bo1
@@
 \item[udagatapadārtha uddhanaṃtī] J5,
-\item[(illegible/unavailable)] Bo3, C1, J2, V19, V10, V14, Tue, C5, V21, V7, V5, V22
+\item[(illegible/unavailable)] Bo3, C1, J2, N6, N8, N11, R5, V19, V10, V14, Tue, C5, V21, V7, V5, V22, G11, T, P16, P22, P23, 
   \begin{description}
@@
 \sthana{2.26b}
-\item[pavanam apānam udīrya kaṇṭhanāle] V1, A1, B2, Ba, Ba5, C6, C8, C9, J3, J4, J6, J7, J10, J11, J14, J15, J17, Ko, LD1, LD2, N17, W4, J8, V16, N18, N20, N22, V3, N23, V20, V15, V28, V18, V12, V13 
+\item[pavanam apānam udīrya kaṇṭhanāle] V1, A1, B2, Ba, Ba5, C6, C8, C9, J3, J4, J6, J7, J10, J11, J14, J15, J17, Ko, LD1, LD2, N7, N9, N10, N13, N16, N17, P6, P7, P8, P9, P10, P13, P18, W4, J8, V16, N18, N20, N22, P14, P15, P20, P25, P27, V3, N23, V20, V15, V28, V18, V12, V13, G1
+\item[pevanam apānam udīrya kaṇṭhanāle] P5, P24, 
+\item[pavanam apānam udīrya kaṇṭhānālī] Ba3
+\item[pavanam apānam udīrya kaṇṭhanālaiḥ] N12, 
 \item[pavanam apānam udārya kaṇṭhanālo] J16,
@@
 \item[pavanam upānam urīrya kaṇṭhanāle] C4
-\item[pavanam apānam udīrya kaṇṭhanālo] N19
-\item[pavanam apānam udīrya kaṇṭhanālāt] B3, C2, J13, N26, V2
+\item[pavanam apānam udīrya kaṇṭhanālo] N19, R1,
+\item[pavanam apānam udīrya kaṇṭhanālāt] B3, C2, J13, N26, P1, P2, P12, P17, P21, P26, V2
 \item[pavanam apām udirya kaṃnāle] N25
 \item[pavanam apāsenam udāryaṃ kaṇṭhanāle] J5,
+\item[pavanam apānam udīryakaṃ jāle] P11, 
 \item[pavanam mudārya kaṇṭhanālo] J12
@@
 \item[pavanam apānam udīrya karāḍanāle] V17
-\item[(illegible/unavailable)] Bo3, C5, J2, V19, V10, V14, V21, Tue, V7, V5, V22        
+\item[pavanam apānam udīrya kaṇṭhaḍālai] N14, 
+\item[(illegible/unavailable)] Bo3, C5, J2, N6, N8, N11, R5, V19, V10, V14, V21, Tue, V7, V5, V22, G11, T, P16, P22, P23, 
 \end{description}
@@
 \sthana{2.26c}
-\item[karibhir iva jalasya vāyuvegāt] A1, B1, B3, C9, J1, J3, J6, J8, J10, J17, N17, W4, V16, V17, N18, N20, N26, V2, V18, V12 
+\item[karibhir iva jalasya vāyuvegāt] A1, B1, B3, C2, C9, J1, J3, J6, J8, J10, J17, LD1, N7, N10, N16,  N17, P1, P2, P9, P12, P17, W4, V16, V17, N18, N20, N26, V2, V18, V12, Ba3
+\item[karibhir iva jalasya vāyuvegaj] R1,
 \item[karibhir iva gajasya vāyuvegāt] V20
-\item[karibhir iva jalasya vāyuvegā] B2, C3, J12, J16, 
-\item[karibhir iva jalaṃ ca vāyuvegā] Ba
+\item[karibhir iva jalasya vāyuvegā] B2, C3, J12, J16, N14, P8, 
+\item[karibhir iva jalasya vāyubagā] P26,
+\item[karibhir iva jalasya vāyuvege] N9,
+\item[karibhir iva jalaṃ ca vāyuvegā] Ba,
+\item[karibhir iva jalaṃ ca vāyuvegāt] P25, 
 \item[karibhiriva jalaṃsya vāyuvegāt] Bo2, 
-\item[karibhiriva jalasya vāyuvegāt] C2, LD1, 
 \item[karibhiriva jalasya vāyuvego] J13,
 \item[karibhiriva jalasya vāyuyogāt] V28
+\item[karibhirivadāyikā jalabhya vāyuvegāt] P21, 
 \item[kramamarivayavasya vāyur yo] V1
-\item[kramaparicayavaśya nāḍicakrā] Ba5
-\item[kramaparicayavaśyā nāḍīmārgā] C6, J7,
+\item[kramaparicayavaśya nāḍicakrā] Ba5, N13, P14, P18,  
+\item[kramaparicayavaśyā nāḍīmārgā] C6, J7, LD2, 
+\item[kramaparicayavaśyā nāḍīmārgāt] N12,
 \item[kramaparicayavaśya vāyumārga] C8
+\item[kramaparicayavaśya vāyumārgā] P7, P20, 
+\item[kramaparicayavasya vāyuvegā] P27,
+\item[kramaparicayavaśa nāḍimārgā] P11,  
+\item[kramaḥ paricayavasya vāyuḥ mārgre] P6,
 \item[kramaparicayavasya vāyumārgre] J14,
 \item[kramaparicayavaśya vāyumārgāt] Bo1
-\item[kramaparicayavaśya vāyumārgaṃ] N21
+\item[kramaparicayavaśya vāyumārgaṃ] N21, P13, 
 \item[kramac pacya paricaya vaśyā vāyumārgaṃ] J5
@@
 \item[kramapārīcita vāyumārge] J15,
+\item[kramaparījita vāyuḥ] P15, 
+\item[kramapārīcitavacana vāyumārgā] P5, P24, 
 \item[kramaparicitavasya vāyuyogo] Ko,
@@
 \item[kramapārecitavaktu? vāyumārge] C4
-\item[karamaparicayavaśya nāḍīmārgā] LD2
 \item[kramaparicittakakrū vāyumārge] N25
-\item[(illegible/unavailable)] Bo3, C1, J2, V19, V10, V14, Tue, C5, V21, V6, V7, V15, V5, V22, V13
+\item[kirimir iva jalasya vāyuvegāt] P10, 
+\item[(illegible/unavailable)] Bo3, C1, J2, N6, N8, N11, R5, V19, V10, V14, Tue, C5, V21, V6, V7, V15, V5, V22, V13, G11, T, G1, P16, P22, P23, 
   \begin{description}
@@
 \item[gajakaraṇānti nigadyate haṭhajṇau] Ko,
-\item[gajakaraṇīti nigadyate haṭhajñaiḥ] B2, Ba, Ba5, Bo1, C3, C4, C6, J4, J6, J7, J11, J15, V1, N19, N23, N25(“dya” unclear), V20, V15, V13
+\item[gajakarīṇīti kathyate haṭhavidyā] G1
+\item[gajakaraṇīti nigadyate haṭhajñaiḥ] B2, Ba, Ba5, Bo1, C3, C4, C6, J4, J6, J7, J11, J15, N12, N13, P5, P6, P7, P9, P10, P14, P18, P24, P26, P27, V1, N19, N23, N25, P13, V20, V15, V13
+\item[gajjakaraṇīti nigadyate haṭhajñaiḥ] P8, 
+\item[gajakaraṇī nigadyate haṭhajñaiḥ] N16, 
 \item[gajakaraṇīti nigadate haṭhajñaiḥ] J14,
@@
 \item[gajakaraṇī ca nigadyate dṛḍhajñaiḥ] J13,
+\item[gajakaranīti nigadye haṭhajña] N14, 
+\item[gogajakaṇīti nigadyate haṭhajñaiḥ] P15, 
 \item[jalagajakaraṇī nigadyate haṭhaj~naiḥ] J1, J3, 
-\item[jalakaraṇīti nigadyate haṭhajñaiḥ] C9, J12, J16, LD1, N17, W4, N20, N26, 
+\item[jalakaraṇīti nigadyate haṭhajñaiḥ] C9, J12, J16, LD1, N7, N17, P1, W4, N20, N26, P21, P25, 
+\item[jalakaraṇīti nigadyato haṭhajñaiḥ] N10,
 \item[jalakaraṇīti nigadyate mahāhaṭhaj~naiḥ] V21
-\item[jalakariṇīti nigadyate haṭhajñaiḥ] J10, J17, V16, V28, V18, V12
+\item[jalakariṇīti nigadyate haṭhajñaiḥ] J10, J17, N9, P20, R1, V16, V28, V18, V12, Ba3
 \item[jalakaraṇānī nigadyate haṭhaj~naiḥ] V17
-\item[jalagajakaraṇīti nigadyate haṭhaj~naiḥ] J8, N18
-\item[gajakaraṇīti nirādyate haṭhaj~naiḥ] N21(unusual way of writing “dya”)
-\item[(illegible/unavailable)] Bo3, C1, J2, C5, V19, V10, V14, Tue, V7, V5, V22
+\item[jalagajakaraṇīti nigadyate haṭhaj~naiḥ] J8, N18, P2,
+\item[jalakaraṇīva nigadyate haṭhajñaiḥ] P12,
+\item[jalagajakaraṇīva nigadyate haṭhajñaiḥ] P17, 
+\item[gajakaraṇīti nirādyate haṭhaj~naiḥ] N21,
+\item[najakariṇīti nigadyate haṭhajñaiḥ] P11, 
+\item[(illegible/unavailable)] Bo3, C1, J2, C5, N6, N8, N11, V19, V10, V14, Tue, V7, V5, V22, G11, T, P16, P22, P23, R5,
   \begin{description}
@@
 \item[aśeṣadoṣoyathā nihanyānād] V14
-\item[aśeṣadoṣopacayaṃ nihanyād] B1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C6, C8, C9, J2, J3, J5, J6, J7, J10, J11, J13, J14, J15, J17, LD1, LD2, N18(“aseṣa”), N19, N20, Tue, N25, N26, V20, V5, V28, V22, V2.
+\item[aśeṣadoṣāmaya śoṣaṇāya] G1
+\item[aśeṣadoṣopacayaṃ nihanyād] B1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C6, C8, C9, J2, J3, J5, J6, J7, J10, J11, J13, J14, J15, J17, LD1, LD2, N7, N8, N11, N13, N16, N18, N19, N20, P1, P2, P5, P8, P9, P12, P13, P14, P15, P17, P18, P20, P21, P24, P25, P27, Tue, R1, N25, N26, V20, V5, V28, V22, V2.
 V18, V12
+\item[aśeṣadopacayaṃ nihanyād] P23,
+\item[aśeṣa doṣā pacayan niśamyaha nyāda] G11
+\item[aśeṣadoṣo ca nayaṃti hanyād] N10,
 \item[aśeṣadoṣopacaya no hanyād] J16,
 \item[aśeṣadoṣopacayā nihanyād] V13
-\item[aśeṣadoṣopacayaṃ nihanyā] N23(space for “da” left blank with an overhead line, verse comes before the earlier verse in this list)
-\item[aśeṣadoṣopacaya nihanyād] N24(verse comes before the earlier verse in this list)
-\item[aśeṣa?oṣovacayaṃ nihanyād] N21(verse comes before the earlier verse in this list)
-\item[aśeṣadoṣepacayaṃti hanyād] J12,
+\item[aśeṣadoṣopacayān nihanyād] N12,
+\item[aśeṣadoṣapacayaṃ nihanyād] N9,
+\item[aśeṣadoṣopacayaṃ nihanti] P7, 
+\item[aśeṣadoṣopacayaṃ nihanyā] N23,
+\item[aśeṣadoṣopacaya nihanyād] N24, P10, P26, 
+\item[aśeṣa?oṣovacayaṃ nihanyād] N21,
+\item[aśeṣadoṣasya cayaṃti hanyād] P6,
+\item[aśeṣadoṣepacayaṃti hanyād] J12, N14, 
+\item[aśeṣa doṣo vacayan nihanyād] Ba3
 \item[aśeṣadoṣepacayaṃ nihanyad] C2,
 \item[aśeṣadoṣāpacayaṃ nihanyad] V21, V3
+\item[aśeṣadoṣopacayaṃ nihatyad] P11, 
 \item[aśeṣadoṣopacitiṃ hathajñād] J4,
@@
 \item[aśeṣadoṣasya cayana hatyā] N22
-\item[(illegible/unavailable)] C5,
+\item[(illegible/unavailable)] C5, N6, T, P16, P22, R5,
   \begin{description}
@@
 \sthana{2.29d}
-\item[abhyasyamānaṃ jalavastikarma] Ba5, Bo1, Bo2, Bo3, C1, C3, C6, C8, J1, J2, J4, J5, J6, J7, J8, J11, J12, J13, J14, J15, J16, LD1, N17, W4, V16, V17, V10, N18, N19, N21, V21, V3, Tue, N26, V6, V7, V20, V15, V5, V28, V22, V2, V18, V13
+\item[abhyasyamānaṃ jalavastikarma] Ba5, Bo1, Bo2, Bo3, C1, C3, C6, C8, J1, J2, J4, J5, J6, J7, J8, J11, J12, J13, J14, J15, J16, LD1, N8, N9, N10, N11, N12, N13, N14, N16, N17, P1, P2, P8, P9, P11,  P12, P14, P15, P18, P20, P23, W4, V16, V17, V10, N18, N19, N21, P13, P25, P27, V21, V3, Tue, N26, V6, V7, V20, V15, V5, V28, V22, V2, V18, V13, G11, G1
+\item[abhyasyamānaṃ janabastikarma] R1,
+\item[abhyasyamānaṃ jalavāstikarma] P5, P24, 
 \item[anyasyamānaṃ jalavastikarma] J3,
 \item[abhyasyamānaṃ nijavastikarma] V12
-\item[abhyasyamānaṃ jalavastikarmaḥ] Ko,
+\item[abhyasyamānaṃ jalavastikarmaḥ] Ko, P6, 
+\item[abhyamānaṃ jalabastikarma] P21, 
 \item[abhyasyamānaṃ khalu vastikarma] C9,
@@
 \item[praamāńam jalavastikarma] V14
-\item[abhyasyamāno jalavastikarma] B3, C2, 
+\item[abhyasyamāno jalavastikarma] B3, C2, P17, 
 \item[abhyasyamā jalavastikarma] B1,
-\item[abhyāsamānaṃ jalavastikarma] C4, V1, N25
-\item[abhyāsyamānaṃ jalavastikarma] A1, Ba
+\item[abhyāsamānaṃ jalavastikarma] C4, V1, N7, N25, P10, P26, 
+\item[abhyāsyamānaṃ jalavastikarma] A1, Ba,
 \item[abhyāsamānaṃ jalavasti] V19
+\item[abhyasya mānam jalapastikarma] Ba3
 \item[abhyasyamānaṃ javastikarma] J10, J17,
@@
 \item[abhyasya śūnyaṃ jalavastikarma] B2,
+\item[cābhyasyamānaṃ jalavastikarma] P7, 
 \item[dattasya mānaṃ yanabasyakarmaḥ] N22
-\item[(illegible/unavailable)] C5,
+\item[(illegible/unavailable)] C5, N6, T, P16, P22, R5,
   \begin{description}
@@
 \item[atha natīḥ] J1
-\item[atha netī(ḥ)] Ba, Ba5, Bo1, Bo2, C1, C2, C4, J3, J6, J8, V1, V10, N18, V21, V3, N23, N25, N26, V7, V5, V2
-\item[atha neti] J2, Ko, N20, N24, V6, V28
-\item[atha netiḥ] J7, Tue, V22, V12
-\item[atha netikarma] C3, C9, J10, J17, LD1, W4, V16, V17, N21, V14, V15, V18, V13
-\item[atha netikarmaḥ] C8, J12, J15, N22
-\item[atha netīkarma] A1, B1, B2, C6, J13, J14, J16, N17, V20
+\item[atha ti] P26, 
+\item[atha netī(ḥ)] Ba, Ba5, Bo1, Bo2, C1, C2, C4, J3, J6, J8, N8, N11, N16, V1, V10, N18, P20, P21, P23, V21, V3, N23, N25, N26, V7, V5, V2
+\item[atha neti] J2, Ko, N7, N9, N10, N13, N14, N20, N24, P7, P9, V6, V28
+\item[atha netiḥ] J7, N12, P14, P18, P25, Tue, V22, V12, Ba3
+\item[atha netikarma] C3, C9, J10, J17, LD1, P1, P2, P5, P8, P10, P24, W4, V16, V17, N21, P13, V14, V15, V18, V13
+\item[atha netikarmaḥ] C8, J12, J15, N22, P6, 
+\item[atha netīkarma] A1, B1, B2, C6, J13, J14, J16, N17, P12, P17, P27, R1, V20
 \item[atha netīkarmaḥ] LD2
 \item[atha nītī] J5,
+\item[atha nīti] P11, 
 \item[atha nītikarma] J4,
@@
 \item[prathama netīkarma] B3
-\item[(illegible/unavailable)] Bo3, C5, J11, 
+\item[(illegible/unavailable)] Bo3, C5, N6, J11, G11, P15, T, G1, P16, P22, R5,
   \begin{description}
@@
 \sthana{2.30c}
-\item[mukhanirgamanād eva] A1, B2, C3, J10, J17, LD1, N17, V1, W4, V16, N22, V20, V18, V12, V13
+\item[mukhanirgamanād eva] A1, B2, C3, J10, J17, LD1, N7, N9, N17, P1, P5, P6, P8, P9, P10, P24, P26, V1, W4, V16, N22, V20, V18, V12, V13, Ba3
 \item[mukhanirgamanād evaḥ] J15,
@@
 \item[mukhan nirgamanād eva] B3,
-\item[mukhān nirgamanād eva] C2,
+\item[mukhān nirgamanād eva] C2, P12, P17, 
+\item[muvi nirgamanād eva] R1,
+\item[mukhā nirgamanād eva] P2, 
+\item[mukhṃ nirgamanād eva] N14, 
 \item[mukha nīrgamanod eva] J16, 
-\item[mukhān nirgamayec caiṣā] Ba, Ba5, Bo1, C8, J8, J11, LD2, V17, N19, N20, V21, Tue, N26, 
+\item[mukhān nirgamayec caiṣā] Ba, Ba5, Bo1, C8, J8, J11, LD2, N12, N13, V17, N19, N20, P14, P18, P25, V21, Tue, N26, 
 \item[mukhān nirgamyate caiṣā] J4,
+\item[mukhān nirgamayed eṣā] P13, 
+\item[sukhān nirgamayeccaiṣā] G11
 \item[mukhān virgamayed eṣā] N21
-\item[mukhā nirgamayec caiṣā] J7, V10, V3, V6, V7, V15, V22
-\item[mukhe nirgamayec caiṣā] C6
+\item[mukhā nirgamayec caiṣā] J7, P15, V10, V3, V6, V7, V15, V22
+\item[mukhe nirgamayec caiṣā] C6, P7, 
+\item[mukhe nirgamaye caiṣā] P11, 
 \item[mu nirgamayai caiṣā] J5,
@@
 \item[mukhān nirgamayet sāpi] V19
-\item[mukhān nirgāmayet sā hi] B1, Bo2, J13, J14, N18, N25  
-\item[mukhān nirgamayet sā hi] J6,
+\item[mukhān nirgāmayet sā hi] B1, Bo2, J13, J14, N11, N18, N25  
+\item[mukhān nirgamanāt sā hi] P21, P23, 
+\item[mukhān nirgamayet sā hi] J6, N8, N20, P20, P27
 \item[mukhā nirgāmaye sā hi] V2
@@
 \item[mukhā nirgamanā sā hi] V5
-\item[mukhān nigarmayet sā hi] C4, J3,
+\item[mukhān nigarmayet sā hi] C4, J3, N16, 
 \item[mukhā nigarmayet sā hi] J1, V28  
@@
 \item[mukhaśca nirgamec caiṣā] N23
-\item[(illegible/unavailable)] Bo3, C5
+\item[sukhanirgamanād eva] N10,
+\item[mukhaṃ nirgamayedeṣu nīti] G1
+\item[(illegible/unavailable)] Bo3, C5, N6, T, P16, P22, R5,
+  \begin{description}
+
+    \end{description}
+ \end{marma}
+
+ \begin{marma}[hp02_31cd]
+\sthana{2.30cd}
+\item[] 
+\item[(illegible/unavailable)] 
   \begin{description}
@@
 \item[mucyate] J4,
-\item[mocanaṃ] B2, Ba5, C8, J7, J12, N18, N21, N22, Tue, N23, V7
-\item[mocakaṃ] Bo2, C1, J1, J3, J5, J6, J11, J14, N19, V28
+\item[mocanaṃ] B2, Ba5, C8, J7, J12, N12, N13, N18, N21, N22, P13, P14, P16, P18, P26, Tue, N23, V7
+\item[mocakaṃ] Bo2, C1, J1, J3, J5, J6, J11, J14, N16, N19, P15, P20, P21, P23, V28
 \item[moṭakaṃ] V1, Bo1, N25 
+\item[meṭakaṃ] P11, 
+\item[mokṣakaṃ] N8, N11, 
 \item[toṭakaṃ] V19, V16
 \item[tāṭakaṃ] V17
-\item[trāṭakaṃ] A1, C4, C6, V10, V14, V21, V3, V6, V20, V15, V5, V28, V22, V2, V18, V12, V13
+\item[trāṭakaṃ] A1, C4, C6, P6, P7, P27, V10, V14, V21, V3, V6, V20, V15, V5, V28, V22, V2, V18, V12, V13, G11, Ba3, G1
 \item[troṭakaṃ] N24
-\item[rocanaṃ?] Ba,
-\item[sphoṭanaṃ] B1, B3, C2, C3, C9, J8, J10, J13, J15, J16, J17, LD2, N17, W4, N20, N26,  
+\item[rocanaṃ] Ba, P25, 
+\item[sphoṭanaṃ] B1, B3, C2, C3, C9, J8, J10, J13, J15, J16, J17, LD2, N7, N14, N17, P1, P8, W4, N20, N26, P2, P5, P9, P10, P12, P17, P24, Ba3  
+\item[sphoṭana] R1,
+\item[spheṭanaṃ] N10,
 \item[sphoṭakaṃ] LD1,
-\item[(illegible/unavailable)] Bo3, C5
+\item[sphoṭa] N9, 
+\item[(illegible/unavailable)] Bo3, C5, N6, T, P22, R5,
   \begin{description}
@@
 \item[tandrādīnāṃ gapāṭa[ka]m] V1,
-\item[tandrādīnāṃca pāṭanaṃ] V19,
+\item[tandrādīnāṃca pāṭanaṃ] P15, V19,
+\item[ta.ntryādīnāṃ kavāṭakam] G1
 \item[tandrādīnāṃ ca kapāṭakaṃ] B1,
-\item[tandrādīnāṃ kapāṭakam] B2, B3, Ba, Ba5, Bo1, C1, C2, C4, C6, C8, J3, J5, J6, J7, J8, J11, J12, Ko, LD1, LD2, W4, V17, V10, N21, N22, V3, Tue, V7, V5, V28, V22, V2, V13
+\item[tandrādīnāṃ kapāṭakam] B2, B3, Ba, Ba5, Bo1, C1, C2, C4, C6, C8, J3, J5, J6, J7, J8, J11, J12, Ko, LD1, LD2, N7, N8, N11, N13, N16, P2, P6, P7, P12, P13, P14, P16, P17, P18, P23, P25, P26, P27, W4, V17, V10, N21, N22, V3, Tue, V7, V5, V28, V22, V2, V13, Ba3
+\item[tadrādīnāṃ kapāṭakam] P21,
+\item[tandrādīnāṃ kavāṭakam] G11
+\item[tandrānāṃ tu kapāṭakam] N14,
+\item[tandrādīnāṃ kṛpāṭakam] N12, 
 \item[tandrādīṇāṃ kapāṭakam] J13, 
-\item[tandrāṇāṃ ca kapāṭakam] J15, 
+\item[tandrāṇāṃ ca kapāṭakam] J15, N9, N10, P1, P8, P10, R1, 
+\item[tadbhānaṃ ca kapāṭakam] P5, P24, 
 \item[tandrādīnāṃ kapālakam] J1
@@
 \item[tandrādiṇāṃ kapāṭakam] J8, N18, N23,
+\item[tandrādīnāṃ kaṃpāṭa] P9, 
+\item[tandrādīn kapāṭakam] P11,
 \item[tandrādiṇāṃ ca pāṭakam] J14, N19
@@
 \item[sandrādīnāṃ kapātakāṃ] N24
+\item[nidrādīnāṃ kapāṭakam] P20,
 \item[nidrārttīnāṃ kapāṭakam] N25
 \item[hetādīnāṃ? kapāṭhakam] J2,
-\item[(illegible/unavailable)] Bo3, C5, V6, V12
+\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, V6, V12, T
   \begin{description}
@@
 \item[gudaṃ] V1
-\item[gopyaṃ] V16, V17, V10, N18, V14, V21, V3,V6, V7, V20, V15, V28, V22, V2, V18, V12, V13
-\item[tundaṃ] A1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J7, J8, J13, J14, J16, LD1, LD2, N19, N20, Tue, N26, 
+\item[gopyaṃ] P1, V16, V17, V10, N14, N18, V14, V21, V3,V6, V7, V20, V15, V28, V22, V2, V18, V12, V13, G11, Ba3, G1
+\item[tuda] N10,,
+\item[tundaṃ] A1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J7, J8, J13, J14, J16, LD1, LD2, N8, N9, N11, N13, N16, N19, N20, P2, P17, P18, P20, P26, Tue, N26, G11, P11, 
 \item[tunduṃ] N23
-\item[tunda] N17
-\item[tunde] C1, C4,  
+\item[tunda] N12, N17 P16, 
+\item[tunde] C1, C4, P21, P23,  
 \item[tuṃḍe(?)] N25, V5
-\item[tundat] B2, J10, J17, 
+\item[tundat] B2, J10, J17, P8, P10, R1, 
 \item[tundet] J15, 
-\item[truṃdaṃ] N21
+\item[truṃdaṃ] N21, P15, 
 \item[tuda] W4
-\item[tudaṃ] B1, C3, C6, J1, J5, J12,  
+\item[tudaṃ] B1, C3, C6, J1, J5, J12, P6, P7, P12, P13, P14, P27, 	
 \item[tundan] J4,
-\item[tudan] Ba, Bo1, J11, 
+\item[tudan] Ba, Bo1, J11, P9, P25,  
+\item[tudat] N7, P5, P24, 
 \item[tulyaṃ] Ko,
@@
 \item[saṃdada] J2
-\item[(illegible/unavailable)] Bo3, C5,
+\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, T
   \begin{description}
@@
 \item[pāvanādi] Ko,
-\item[pāvakādi] C3, J10, J16,
+\item[pāvakādi] C3, J10, J16, N10, Ba3, R1, 
 \item[pāvakādī] J12,
 \item[pāvakāgni] N26
-\item[pācakādi] C2, C9, J17, LD1, LD2, N17, W4, V16, V17, N20, V14, V21, V18, V13
+\item[pācakādi] C2, C9, J17, LD1, LD2, N7, N9, N14, N17, P1, P5, P8, P9, P10, P24, W4, V16, V17, N20, V14, V21, V18, V13
 \item[pācatādi] Bo2, 
-\item[pācanādi] A1, B1, B2, B3, Ba, Ba5, Bo1, C6, C8, J4, J5, J7, J8, J11, J14, V19, V10, N19, N21, V3, Tue, N23, V6, V7, V20, V15, V28, V22
-\item[pācanāgni] J3, J6, J13,  
+\item[pācanādi] A1, B1, B2, B3, Ba, Ba5, Bo1, C6, C8, J4, J5, J7, J8, J11, J14, N12, N13, P6, P7, P14, P17, P25, P27, V19, V10, N19, N21, P2, P12, P13, P18, P20, V3, Tue, N23, V6, V7, V20, V15, V28, V22, G11, G1
+\item[pācanādīn P26, 
+\item[pācanāgni] J3, J6, J13, N8, N11, N16, P16, P23, 
+\item[pācanāgnī] P21,
 \item[pācakāgniḥ] J15
@@
 \item[pācakāri?] J2,
+\item[pācanānī] P11, P15, 
 \item[pācanāṃnī] N24
 \item[kārkādeśaṃ] N22
-\item[(illegible/unavailable)] Bo3, C5, V5, V12
+\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, V5, V12, T
   \begin{description}
@@
 \item[mala śoṣiṇī va] V1,
-\item[mala śoṣaṇī ca] B2, C3, J8, LD1, N17, W4, V16, V14, V3, V7, V18, V12
+\item[mala śoṣaṇī ca] B2, C3, J8, LD1, N9, N17, N20, P1, P5, P10, P16, P24, W4, V16, V14, V3, V7, V18, V12
 \item[maya śoṣaṇī ca] V21, V15, V28, V13
@@
 \item[maya śodhaṇī cā] V6, 
-\item[mala śoṣaṇī va] J10, J17,
+\item[mala śoṣaṇīva] J10, J17, P9,
 \item[mala śoṣaṇī vaḥ] J16,
-\item[doṣāmaya śoṣaṇī ca] C1, C2, C4, C6, C8, C9, J3, J4, J11, J13, J14,  Ko, N18, N21, Tue, N25, V22, V2
+\item[doṣāmaya śoṣaṇī ca] C1, C2, C4, C6, C8, C9, J3, J4, J11, J13, J14, Ko, N12, N13, N18, N21, P2, P6, P7, P11, P12, P13, P14, P15, P17, P18, P20, P21, P23, P27, Tue, N25, V22, V2, G11, Ba3
 \item[doṣāmaya śauṣaṇī caḥ] J15, 
-\item[doṣān mala śoṣaṇī ca] J12,
+\item[doṣamala śoṣaṇī ca] P8,  
+\item[doṣāmala soṣaṇi ca] N14, R1,  
+\item[doṣān mala śoṣaṇī ca] J12, N10, 
 \item[doṣāmaya śoṣaṇīya] J1
 \item[doṣāmaya śoṣaṇīyaṃ] J2
-\item[doṣāmaya śoṣiṇī ca] A1, B3, Ba, Bo1, J6, V19, N19
-\item[doṣāmala śoṣaṇī ca] N20
+\item[doṣāmaya śoṣiṇī ca] A1, B3, Ba, Bo1, J6, N8, N11, N16, V19, N19, P25, 
 \item[doṣāmaya śoṣiṇāṃ ca] Bo2
 \item[doṣāmaya śodhinī ca] J7, 
+\item[doṣāmaya śodhanī ca] P26,
 \item[doṣābhayahā soṣaṇī ca] J5,
@@
 \item[doṣāmaya goṣaṇī ca] N24
-\item[mala śoṣiṇī ca] V17, N26,
+\item[mala śoṣiṇī ca] V17, N7, N26,
 \item[doṣāṇapi śoṣaṇā ca] B1,
@@
 \item[doṣānala śoṣaṇī bahava] N22
-\item[(illegible/unavailable)] Bo3, C5, V20, 
+\item[(illegible/unavailable)] Bo3, C5, N6, P22, V20, R5, T, G1
   \begin{description}
@@
 \sthana{2.35d}
-\item[mūlam iyaṃ hi] J14, V1
+\item[mūlam iyaṃ hi] J14, P15, V1
 \item[mūlam idaṃ hi] Ko, J4, J11, N19
+\item[mau.riyaṃ hi] G1
 \item[mūlam ayaṃ ca mauliḥ] N22
-\item[maulir iyaṃ hi] A1, C1, C4, J3, J5, V19, V21, N25+, V15, V2
+\item[maulir iyaṃ hi] A1, C1, C4, J3, J5, N8, N11, N16, P6, P20, P23, P27, V19, V21, N25, V15, V2
 \item[maulir iyaṃ] J1
-\item[maulir iyaṃ ca] Ba, Ba5, Bo1,Bo2, C6, C8, J6, J7, N21, Tue, N23(“yaṃ” looks like “paṃ”), V6, V7, V20, V28, V22
+\item[maulir iyaṃ ca] Ba, Ba5, Bo1,Bo2, C6, C8, J6, J7, N12, N13, N21, Tue, N23, P7, P11, P13, P14, P16, P18, P21, P25, V6, V7, V20, V28, V22, G11
 \item[saulir iya ca] N24
@@
 \item[sau jayatīha] J10, J17, N17, W4, N18, N20, N26
-\item[jayatīha] B1, B3, C2, C3, C9, J8, J12, J13, J15, J16, LD1, LD2, V3
+\item[jayatīha] B1, B3, C2, C3, C9, J8, J12, J13, J15, J16, LD1, LD2, N7, N9, N10, N14, P1, P2, P5, P8, P9, P10, P12, P17, P24, P26, R1, V3
 \item[sauryakarīha?] B2
-\item[(illegible/unavailable)] Bo3, C5, V16, V17, V10, V14, V5, V18, V12, V13
+\item[(illegible/unavailable)] Bo3, C5, N6, P22, V16, V17, V10, V14, V5, V18, V12, V13, T, Ba3, R5,
   \begin{description}
@@
 \sthana{2.38a}
-\item[prāṇāyāmena vai] Ko, J4, J11, J14, V1, N19, V15
-\item[prāṇāyāmair eva] B1, B2, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J12, J13, J15, LD2, N1 (not sure of "eva"), N3, N4a, N5, N8, N11, N12, N13, N16, N18, N22, V3, Tue, N23, N24, N25, V6, V7, V5, V28, V22, V2, V12, V13
-\item[prāṇāyāmaiḥ sukhāt] A1, C3, C9, J10, J17, LD1, N2 ("śu"), N9, N10,
+\item[prāṇāyāmena vai] Ko, J4, J11, J14, V1, N19, P15, V15
+\item[prāṇāyāmair eva] B1, B2, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J12, J13, J15, LD2, N1, N3, N4a, N5, N8, N11, N12, N13, N16, N18, N22, P2, P7, P11, P12, P14, P17, P18, P20, P21, P23, P25, V3, Tue, N23, N24, N25, V6, V7, V5, V28, V22, V2, V12, V13, G11, G1
+\item[prāṇāyāmaiḥ sukhāt] A1, C3, C9, J10, J17, LD1, N2, N9, N10, N17, P1, P5, P8, P9, P10, P24, P26, R1, Ba3
 \item[prāṇāyāme sukhāt] J16,
+\item[prāṇāyāmais tu sā sarvaiḥ] P16,
 \item[prāṇāyāmaiḥ sukhāḥ] N26
 \item[prāṇāyāmair mu?khāṃ] N20
-\item[prāṇāyāmena sarvepi] N21
+\item[prāṇāyāmena sarvepi] N21, P6, P13, P27, 
 \item[praṇāyāme kṛte samyak] J2
@@
 \item[praṇāyāmāt sukhāt] N7
-\item[(illegible/unavailable)] Bo3, C5,
+\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, T
   \begin{description}
@@
 \sthana{2.38b}
-\item[malā api] J4, J11, V19, N21, V15, V28
+\item[malā api] J4, J11, V19, N21, P6, P13, P15, P27, V15, V28
 \item[malād api] Ko,
-\item[malā iti] Ba, Ba5, Bo1, Bo2, C1, C6, C8, J1, J2, J3, J5, J6, J7, LD2, N3, N4a, N5, N8, N11, N13, N16, V10, N18, N19, Tue, N23, N24, N25, V6, V7, V22
+\item[malā iti] Ba, Ba5, Bo1, Bo2, C1, C6, C8, J1, J2, J3, J5, J6, J7, LD2, N3, N4a, N5, N8, N11, N13, N16, P7, P11, P14, P18, P21, P23, P25, V10, N18, N19, P20, Tue, N23, N24, N25, V6, V7, V22, G11, G1
 \item[malā itiḥ] J15,
@@
 \item[malāśayaḥ] J10,
-\item[malāśayāḥ] A1, B1, B3, C2, C9, J13, LD1, N1, N2, N10, V17, V14, V20, V18, V13
+\item[malāśayāḥ] A1, B1, B3, C2, C9, J13, LD1, N1, N2, N10, N17, P5, P10, P12, P17, P24, P26, R1, V17, V14, V20, V18, V13
 \item[malākulaṃ] B2, N22
 \item[malākulāḥ] N20, N26
-\item[malāśayā(ḥ)] C3, C4, J16, J17, N9, V16
+\item[malāśayā] C3, C4, J16, J17, N9, P1, P2, P8, P9, P16, V16
 \item[marāśayā] N14
@@
 \item[malāyinaḥ] N12
-\item[(illegible/unavailable)] Bo3, C5, V5, V12
- \begin{description}
-
-        \end{description}
-\end{marma}
-
-\begin{marma}[hp02_39]
-\sthana{2.39}
-\item[]
+\item[(illegible/unavailable)] Bo3, C5, N6, P22, P24, V5, V12, T, Ba3, R5,
+  \begin{description}
+
+    \end{description}
+ \end{marma}
+
+ \begin{marma}[hp02_39ab]
+\sthana{2.39ab}
+\item[brahmādayo 'pi tridaśāḥ pavanābhyāsatatparāḥ] A1, B2, B3, Ba, G1, P11, P12, P13, P14, P15, P16, P18, P20, P27, P21, P23, P25, P26,
+\item[brahmādayo 'pi tṛdaśāḥ pavanābhyāsatatparāḥ] P17,
+\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparāḥ] B1
+\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparaḥ] R1,
+\item[(illegible/unavailable)] P22, P24, R5, T
+  \begin{description}
+
+    \end{description}
+ \end{marma}
+
+\begin{marma}[hp02_39cd]
+\sthana{2.39cd}
+\item[abhūvannantakabhayāt tasmat pavanam abhyaset] G1,
+\item[abhūvannantakabhayāt tasmāt pavanam abhyaset] A1, Ba, P11, P13, P14, P15, P18, P20a, P27,
+\item[abhūvannātankabhayāt tasmāt pavanam abhyaset] B1,
+\item[abhūvannatakabhayāṃ tasmat pavanam abhyaset] P26,
+\item[abhavannantakabhayāt tasmat pavanam abhyaset] P25,
+\item[abhūvan nanekabhayāt tasmat pavanam abhyaset] P16,
+\item[te ca siddhiṃ gatāḥ sarve tasmat pavanam abhyaset] P12,
+\item[tena siddhi gatās te ca tasmat pavanam abhyaset] P21, P23, 
+\item[tena siddhiṃ gatās te ca tasmat pavanam abhyaset] B2, 
+\item[tena siddhi gatāḥ satyaḥ tasmat pavanam abhyaset] P20,
+\item[naiva siddhiṃ gatāḥ sarve tasmat pavanam abhyaset] B3, P17,
+\item[sumanskāṃtako bhavyaḥ tasmat pavanam abhyaset] R1,
+\item[(illegible/unavailable)] P22, P24, R5, T
   \begin{description}
@@
 \item[cirāśrayam] LD2,
-\item[nirāśrayam] A1, Ba, C1, C3, C6, C8, C9, J2, J4, J5, J7, J8, J10, J11, J15, J17, LD1, V1, W4, N3, N9, N10, V17, N20, N22, V21, V3, N23, N26, V6, V15, V28, V12, V13
+\item[nirāśrayam] A1, Ba, C1, C3, C6, C8, C9, J2, J4, J5, J7, J8, J10, J11, J15, J17, LD1, V1, W4, N3, N9, N10, V17, N20, N22, P1, P5, P6, P8, P9, P10, P11, P15, P20a, P24, P25, P27, V21, V3, N23, N26, R1, V6, V15, V28, V12, V13, G1
 \item[nirāśrayaḥ] J12,
+\item[niśaṃśrayaṃ] G11
 \item[nirodhā vādham] V8
@@
 \item[nirāsnayām] J16,
-\item[nirāmayam] B2, Bo2, C4, J1, J3, J6, J14, Ko, N17, V19, N2, N5, N7 ("ma" seems tampered with later), N8, N11, N12, N14, N16, V16, N18, N19, N21, N24, N25, V20, V5, V2, V18
-\item[nirākulam] B1, B3, Ba5, C2, J13, N1 (2.45), N4a, N13, V10, Tue, V7, V22
-\item[(illegible/unavailable)] Bo3, C5, V14
+\item[nirāmayam] B2, Bo2, C4, J1, J3, J6, J14, Ko, N17, V19, N2, N5, N7, N8, N11, N12, N14, N16, V16, N18, N19, N21, N24, N25, P7, P13, P16, P20, P21, P23, P26, V20, V5, V2, V18, Ba3,
+\item[nirākulam] B1, B3, Ba5, C2, J13, N1, N4a, N13, P2, P12, P14, P17, P18, V10, Tue, V7, V22
+\item[(illegible/unavailable)] Bo3, C5, N6, P22, P24, R5, V14, T
   \begin{description}
@@
 \item[akṣir?] C8,
-\item[īkṣed] Ba, J5, 
-\item[īkṣet] C6, V13
-\item[īkṣe] C1, J7, 
-\item[paśyed] A1, Bo1, C9, J4, J8, J10, J11, J14, J15, J16, J17, Ko, LD1, N17, V1, W4, N9, V8, N10, V16, V17, N20, N22(pasyed), V21, V3, N26, V15, V12
+\item[īkṣed] Ba, J5, P25,
+\item[īkṣet] C6, V13, G1
+\item[īkṣe] C1, J7, P7, 
+\item[īkṣye] P11, 
+\item[paśyed] A1, Bo1, C9, J4, J8, J10, J11, J14, J15, J16, J17, Ko, LD1, N17, P15, V1, W4, N9, V8, N10, P5, P8, P10, P24, P26, R1, V16, V17, N20, N22, V21, V3, N26, V15, V12, Ba3
+\item[paśyer] P1,
 \item[paśyad] J12,
@@
 \item[paśye] N19, N24
-\item[paśe] C3
+\item[paśe] C3,
+\item[paśed] P9, 
 \item[paśyan] N7
+\item[paśca] P6, 
 \item[pasya?] N14  
-\item[dṛṣṭir] B3, Ba5, C2, C4, J6, V19, N1, N2, N4a, N11, N16, V10, N21, N25, V7, V5
-\item[dṛṣṭi] B1, B2, Bo2, J1, J2, J3, J13, N5, N8, N12, N13, N18, Tue, V28, V22, V2 
-\item[vīkṣed] N3
+\item[dṛṣṭir] B3, Ba5, C2, C4, J6, V19, N1, N2, N4a, N11, N16, P2, P12, P14, P17, P18, P20, P23, V10, N21, N25, V7, V5
+\item[dṛṣṭi] B1, B2, Bo2, J1, J2, J3, J13, N5, N8, N12, N13, N18, P13, P21, Tue, V28, V22, V2 
+\item[vīkṣed] N3, G11
 \item[iched] N23
-\item[(illegible/unavailable)] Bo3, C5, V14, V6, V18
+\item[(illegible/unavailable)] Bo3, C5, N6, V14, V6, V18, T, P16, P22, P24, P27, R5,
   \begin{description}
@@
 \sthana{2.43a}
-\item[vidhānajñā] C9, J5, J10, J11, Ko, V1, N2 ("vi" seems faded), N4a, N7, N10, N11, N12, N13, N14, N16, V16, V17, V10, N19, N20, N21, V14, V3, Tue, N23, N26, V6, V7, V15, V22, V18, V13
+\item[vidhānajñā] C9, J5, J10, J11, Ko, V1, N2 ("vi" seems faded), N4a, N7, N10, N11, N12, N13, N14, N16, P6, P8, P9, V16, V17, V10, N19, N20, N21, V14, V3, Tue, N23, N26, V6, V7, V15, V22, V18, V13, G11, G1
+\item[vidhānajñāḥ] Ba3
 \item[vidhānaḥjñā] J2
@@
 \item[vidhānujñaś] J1
-\item[vidhānajñāś] Bo2, C1, C3, C8, J4, J6, J7, J12, 
-\item[vidhānajñaś] B1, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C8, J3, J16, N3, N24,
+\item[vidhānataś] P23, 
+\item[vidhānajñāś] Bo2, C1, C3, C8, J4, J6, J7, J12, P5, P13, P14, P15, P24, P25, P27, 
+\item[vidhānajñaś] B1, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C8, J3, J16, N3, N24, P2, P11, P12, P17, P18, P21, 
 \item[vidhānajña] N22, V12
@@
 \item[vidhijñās tu] C6,
-\item[vidhānajño] C4, J15, J17, LD1, LD2, N5, N25, V28
+\item[vidhānajño] C4, J15, J17, LD1, LD2, N5, N17, N25, P1, P10, P20, V28
 \item[vidhānajñoś] J13, 
+\item[viddhānajño?] R1,
 \item[vividhānajño] B2, 
@@
 \item[vidhānajñaiś] N1 (unsure about "ai" matra)
+\item[vidhistajñāś] P26,
 \item[vidhiṃstajñaś] A1
-\item[(illegible/unavailable)] Bo3, C5, J14, V21, V5
+\item[vidhiṃstajñaiś] P7,
+\item[(illegible/unavailable)] Bo3, C5, N6, J14, V21, V5, T, P16, P22, P24, R5,
   \begin{description}
@@
 \item[bhastrakā bhrāmarī vedhā kevalaṃ cāṣṭakumbhakāḥ] J2
+\item[bhastrikā bhramarikā mūrcchā kevaleṣṭā ca kumbhakā] G1
 \item[bhastrikā bhrāmarī mūrcchā kevalī cāṣṭakumbhakāḥ] V1, V16, V17, 
-\item[bhastrikā bhrāmarī mūrcchā kevala cāṣṭakumbhakāḥ] V20, V18, V12, V13
+\item[bhastrikā bhrāmarī mūrcchā kevala cāṣṭakumbhakāḥ] V20, V18, V12, V13, G11, Ba3
 \item[bhastrikā bhrāmarī mūrcchā vinītyaṣṭakumbhakāḥ] V10
@@
 \item[bhastrīkā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] C9, J13, J15, LD1, LD2, N21
+\item[bhastrīkā bhrāmarī mūrchā kevalaś cāṣṭakumbhakaḥ] P12, 
+\item[bhastikā bhrāmarī mūrchā kevalaś cāṣṭakumbhakaḥ] P26,
 \item[bhaṃstrikā bhramarī mūrchā kāṃbalā atha? kumbhakāḥ] J5
@@
 \item[bhastrikā bhramarī mūrcchā cāṣṭakumbhakāḥ sahita?] N2 (after exchange of words' positions)
-\item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakāḥ] A1, B1, C3, J7pc, J10, J12, J16, J17, N3, N7, N10("ri" instead of "rī"),
-\item[bhastrikā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] N20
+\item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakāḥ] A1, B1, C3, J7pc, J10, J12, J16, J17, N3, N7, N9, N10, N17, P1, P2, P6, P9, P10, P17, P20, P27, R1, 
+\item[bhastrikā bhramarī mūrchā kevalaś caṣṭakumbhakāḥ] P5, P24, 
+\item[bhastrikā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] N20, P8, 
 \item[bhastrikā marī mūrchā kevalaṃ cāṣṭakumbhakāḥ] B2,
@@
 \item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakā] N9
+\item[bhastrakā bhramarī mūrchā kevalāś cāṣṭakumbhakā] P13,
 \item[bhastrikā bhrāmarī mūrchā plāvanī saṣṭhakumbhakāḥ] N4a
 \item[bhastrikā bhramarī mūrcchā plāvanī tvaṣṭakumbhakāḥ] N12, V6, V7
+\item[bhastrikā bhramarī mūrcchā plāvanī tyaṣṭakumbhikā] P11, 
 \item[bhastrikā bhramarī mūrcchā pāvanīty eṣakumbhakāḥ] Ko,
 \item[bhastrikā bhramarī mūrcchā pāvanīty aṣṭakumbhakān] J4,
-\item[bhastrikā bhrāmarī mūrchā plāvinītyaṣṭakumbhakāḥ] Ba5, Bo1, C1, N13, V14, Tue(no visarga at the end), V15, V22
+\item[bhastrikā bhrāmarī mūrchā plāvinītyaṣṭakumbhakāḥ] Ba5, Bo1, C1, N13, P14, P18, V14, Tue, V15, V22
+\item[bhastrīkā bhrāmarī mūrchā plāvinītyaṣṭakumbhakāḥ] P25,
 \item[bhastrikā bhrāmarī mūrchā prāvatītyaṣṭakumbhakaḥ] N23
-\item[bhastrikā bhramarī mūrchā plāvanītyaṣṭakumbhakāḥ] Ba, C8, J7, 
+\item[bhastrikā bhrāmarī mūrchā pratyahītyaṣṭakumbhakāḥ] P7,  
+\item[bhastrikā bhramarī mūrchā plāvanītyaṣṭakumbhakāḥ] Ba, C8, J7, P15,  
 \item[bhastrikā bhrāmarī mūrcchā pālāvanīty aṣṭakumbhakāḥ] J11,
 \item[bhastrikā bhramarī mūrchā pratvanītyaṣṭakumbhakāḥ] C6
+\item[bhastrikā bhrāmarī mūrchā sahita cāṣṭakumbhakā] P23,
 \item[bhastrikā bhramarī mūrchā sahitāś cāṣṭakumbhakāḥ] C4, N5
 \item[bhastikā bhrāmarī mūrchā sahitāś cāṣṭakumbhakam] N25, V5
+\item[bhastikā bhrāmarī mūrchā sahitāś cāṣṭakumbhakān] P21, 
 \item[bhastrikabhrama ma mūrchā saṃhataṃ cāṣṭakumbhakā] N8
@@
 \item[sastrikā bhramarī mūrchā kavalaś cāṣṭakumbhakā] N24
-\item[(illegible/unavailable)] Bo3, C5, J14, 
+\item[(illegible/unavailable)] Bo3, C5, N6, J14, T, P16, P22, R5,
   \begin{description}
@@
 \item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkaḥ ||] Ba5
+\item[adhastāt kuñcanādāśu kaṇṭhasaṅkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ] G1, P11, P18, 
+\item[adhastāt kuñcanādāśu kaṇṭhasaṃcane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ] P25,
+\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkā ||] P7, 
+\item[athastātkuñcanenāśukarṇasaṅkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ] Ba3
 \item[adhas tāt kuñcane vāyuḥ kaṃṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāmīgaḥ ||] Ko,
@@
 \item[adhas tā kuñcakīnālau kaṇṭhasaṃkocane kṣate | madhyaṃ paścimatāṇena syāt prāṇo brahmanāḍīkaḥ ||] J2
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] A1, Bo1, C8, J10, J11, V1, N3, N13, Tue, V7, V20, V15, V28, V12, 
+\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] A1, Bo1, C8, J10, J11, V1, N3, N13, P14, P20, Tue, V7, V20, V15, V28, V12, G11
+\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścamatānena syāt prāṇo brahmanāḍigaḥ ||] P21, 
 \item[adhastvākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C9
@@
 \item[adhasthāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] J7,
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J6, J17, 
+\item[adhasthāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānenya syāt prāṇo brahmanāḍigaḥ ||] P10,
+\item[adhas thāt kuñjakenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇāsthā syāt prāṇo brahmanāḍīgataḥ ||] P15, 
+\item[adhastvākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhyanāḍigaḥ ||] P1, P9, 
+\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J6, J17, P2,
 \item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] Bo2, N10, V16, V17, V14, V3, V6
+\item[addhastāt kuñcanenāsu kaṃṭhasaṃkocane krate | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] R1, 
 \item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J3,
+\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍikaḥ ||] P17, 
+\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] N17,
 \item[adhastā kuñcanenāśu kaṇṭhaśaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍiga ||] J12,
@@
 \item[adhastākuñcaṇeṇāśu kaṇṭhaśaṇkocane kṛ(?)te | madhye paścamatānena syāt prāṇo madhye nāḍikā ||] N14
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] N1
+\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] N1,
+\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] P13, P27, 
+\item[adhastāt kuñcane caiva kaṇṭhasaṃkocane krate | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] P6,
 \item[adhastāt kuñcanenaiva kaṇṭhaṃ saṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] B3
@@
 \item[adhas tāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāgiḍaḥ ||] J4,
+\item[adhas tāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhyanāgiḍaḥ ||] P12,
 \item[adhamān kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ||] J5,
@@
 \item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] B1
+\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] N1
 \item[adhasthāt kuñcanenaiva kaṇṭhasokocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] N21
@@
 \item[adhastā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo madhyanāḍigaḥ ||] LD2,
+\item[adhastāt kuñcānenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ ||] P26,
+\item[adhaścā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhye paścimatānyena syāt prāṇo madhyanāḍigaḥ ||] P5, P24, 
+\item[aṃtastvā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] P8, 
 \item[adhastvā kuñcanenāsu kaṇṭhasaṃkocane kṛtaṃ | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] N7, V13
@@
 \item[adhas tāt kuñcanaunāśu kaṇṭhakocane kṛte | madhye paścimatāne tasyāt prāno madhyanāḍigaḥ ||] V18
-\item[(illegible/unavailable)] Bo3, C5, J14, N2, N8, N11, V5, V2
+\item[(illegible/unavailable)] Bo3, C5, J14, N6, N2, N8, N11, V5, V2, T, P16, P22, P23, R5,
   \begin{description}
@@
 \item[āsthāpya] V1, V19
-\item[sansthāpya] V8
-\item[utthāpya] A1, B1, B2, B3, Ba, Ba5, Bo1, Bo2, C2, C3, C4, C6, C8, C9, J1, J3, J6, J7, J8, J10, J11, J13, J17, LD1, N17, W4, N2, N3, N4a, N5, N8 (not sure of "pya"), N9, N10(seems "chā" instead of "thā"), N11, N12, N13, N16, V16, V17, V10, N18, N21, V14, V21, V3, Tue, N23, N24, N26, V6, V7, V15, V28, V22, V18, V12, V13,
+\item[sansthāpya] P26, P27, V8
+\item[utthāpya] A1, B1, B2, B3, Ba, Ba5, Bo1, Bo2, C2, C3, C4, C6, C8, C9, J1, J3, J6, J7, J8, J10, J11, J13, J17, LD1, N17, W4, N2, N3, N4a, N5, N8, N9, N11, N12, N13, N16, P1, P2, P5, P6, P7, P8, P11, P12, P13, P14, P15, P17, P18, P20, P24, P25, V16, V17, V10, N18, N21, V14, V21, V3, Tue, N23, N24, N26, V6, V7, V15, V28, V22, V18, V12, V13, G11
+\item[utthāya] P23,
+\item[uddhāpya] Ba3
 \item[ūtthāpya] J16,
-\item[uthāpya] Ko,
+\item[uthāpya] Ko, R1, 
 \item[usthāpya] J2, J12, 
@@
 \item[ur(?)thāpya] N14
-\item[utthāya] C1, N7, N20,
+\item[utthāya] C1, N7, N20, P21, 
 \item[utprāpya] N22
-\item[uschāpya] N1 (not sure)
+\item[uschāpya] N1, 
 \item[usthāpya] J4,
-\item[uchāpya] J5, J15, 
-\item[utchāpya?] LD2,
-\item[(illegible/unavailable)] Bo3, C5, J14, N25, V20, V5, V2
+\item[uchāpya] J5, J15, N10, P10, 
+\item[utchāpya?] LD2, P9, 
+\item[(illegible/unavailable)] Bo3, C5, J14, N6, N25, V20, V5, V2, T, G1, P16, P22, R5,
 
@@
 \item[virōdhāvadhi kumbhayet]
-\item[nīrodhāvadhi kumbhayet] J15, J16, J17, V21
-\item[nīrodhānadhi kumbhayet] V17
+\item[nīrādhāvadhi kumbhayet] N23,
+\item[nīrodhāvadhi kambhayet] Ba,
+\item[nīrodhāvadhi kumbhayet] A1, B1, B3, J15, J16, J17, N4a, N10, N13, P1, P2, P5, P8, P9, P10, P11,  P12, P14, P17, P18, P24, P25, P27, Tue, V21, G11, Ba3
+\item[nīrodhāvadhi kumbhata] R1,
+\item[nīrodhavadhi kumbhayet] P13, 
+\item[nīrovadhi kumbhayet] P15,
+\item[nīrodhān adhikumbhayet] N20, 
+\item[nirodho parikumbhake] G1
+\item[nīrodhāvadhi kumbhayat] N3, N7, N12, P20a,   
+\item[nīrodhāvadhi kumbhavat] N21, 
+\item[nīrodhāvadhi kumbhakaṃ] N2, N11,  
+\item[nīrodhāvadhi kumbhake] N18, 
+\item[nīrodhānadhi kumbhayet] N1, N26, V17
+\item[nīrodhāvahi kuṃbhayet] N9, N24,
+\item[nīrodhīvadhi kumbhayet] N19, 
 \item[nirodhaṃ kumbhayet tataḥ] J13, 
-\item[(illegible/unavailable)] J14,
+\item[nirodhāścādhi kumbhayet] P7, 
+\item[nīrodhāvadhikam bhavet] P26,
+\item[śirodhāvadhi kumbhakaṃ] N5, N8, N16, P21, P23,  
+\item[śirodhāvati kumbhakān] P6,
+\item[śirodhyavadhi kumbhakaṃ] N25, P20, 
+\item[śiro ‘pya?vadhi kumbhayet] N17,
+\item[śiroyāvadti kumbhakāt] B2,
+\item[sirodhāvadhi kumbhakāt] N22,
+\item[(illegible/unavailable)] J14, N6, N14, T, P16, P22, R5,
+
+  \begin{description}
+
+    \end{description}
+ \end{marma}
+
+\begin{marma}[hp02_53a]
+\sthana{2.53a}
+\item[jālodare]
+\item[jālodarerā] P24, 
+\item[jalodar] A1,
+\item[jalodarā] P13, P14, P15, P18, P20, P21, P23, P27, 
+\item[jalodaro] B2,
+\item[jalodāraṃ] P17,
+\item[jalodaraṃ] B1, B3, Ba, P25, P26, 
+\item[jālodared] R1,
+\item[(illegible/unavailable)] P16, P22, R5,
+
+  \begin{description}
+
+    \end{description}
+ \end{marma}
+
+\begin{marma}[hp02_55ab]
+\sthana{2.55ab}
+\item[śītkāṃ dadyāt sadā cakre ghraṇe caiva vijṛmbhikām]
+\item[śītkāṃ dadyāt sadā vaktre ghraṇe naiva vijṛmbhikām] P14, P18, 
+\item[sītkāṃ dadyāt sadā vaktre ghraṇe caiva vijṛmbhikām] P27,
+\item[sītkāṃ dadyāt sadā vaktre ghraṇe naiva visarjjayet] B2,
+\item[sātkāṃ dadyāt sadā vaktre prāṇe caiva vijṛmbhite] P13,
+\item[satkāṃ dadyāt sadā vaktre prāṇe cātha vijambhakāṃ] P15,
+\item[śītkāṃ kuryāt sadā vaktrāt ghraṇo caiva vijṛ-kām] A1,
+\item[śītkāṃ kuryāt sadā vaktrāt ghraṇe naiva vijṛmbhakān] P26,
+\item[sītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikām] P25,  
+\item[śītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikām] Ba,  
+\item[śītkāṃ kuryāt yathā vaktre prāṇaṃ caiva visarjjayet] P20, 
+\item[śitkāṃ kuryāt sadā vaktre prāṇenaiva visarjayet] P21,
+\item[śitkāṃ kuryāt tathā vaktre ghrāṇenaiva visarjayet] P23, 
+\item[kuṃbhaṃ kuryāt sadā vaktre grāṇenaiva visarjayet] B1, B3, 
+\item[kuṃbhaṃ kuryāt sadā vaktrat ghrāṇenaiva vijaṃbhitaṃ] R1, 
+\item[(illegible/unavailable)] P16, P22, P24, R5,
+
+\item[]
+\item[(illegible/unavailable)]
 
@@
 \sthana{2.54a}
-\item[śītkaraḥ] B1, Ba, C8, 
+\item[śītkaraḥ] B1, C8, P17, 
 \item[śītkaram] C6,
 \item[śīkāra] J7,
-\item[śītkāraḥ] C1, B3, J13, J15, V17, V21 
+\item[śītkāraḥ] C1, B3, J13, J15, P7, P20, P25, V17, V21 
+\item[śītakāraḥ] N9, 
+\item[śītakāra] P1, R1, 
+\item[śitakāra] P6, 
 \item[śītakārāī] LD1, 
-\item[sītkara(ḥ)] B1, Ba, Bo2, C2, C3, C4, J3, J4, 
-\item[sītkāraḥ] J6
+\item[śīghrā] P11,
+\item[śītalī] P15,  
+\item[sītkara(ḥ)] Ba, Bo2, C2, C3, C4, J3, J4, N6, 
+\item[sītkāraḥ] J6, N8, N11, N16, P2, P12, P13, 
+\item[sītkāra] N17, P5, P24, 
 \item[śītkārī] J16, 
-\item[sītkāṃ] N18, N22, Tue, N24, N25, V15, V28, V22, V10
+\item[śītakārī] P8, P26, 
+\item[śītakarī] P9,
+\item[sītkāṃ] N5, N18, N22, Tue, N24, N25, V15, V28, V22, V10, G11
 \item[sītkād] V14
-\item[sītkā] Ko,
+\item[sitkā] N3, N14, 
+\item[sītkā] Ko, N12, P27, 
 \item[śītkāṃ] A1
+\item[sītātkārī] N13,
+\item[sītkaraḥ] N1,
+\item[sītkāra] P23,
+\item[sītkārā] P21,
 \item[sītkarā] J17, 
-\item[sītkārī] Ba5, Bo1, J5, J12, 
+\item[sītkārī] Ba5, Bo1, J5, J12, N2, N4a, N7, N10, P14, P18,    
 \item[sītakārī] C9
@@
 \item[dhikt(?)āṃ] N23
-\item[kumbhaṃ] V3,J8, N20,V14, V21, N26, V16  
+\item[kumbhaṃ] V3,J8, N20,V14, V21, N26, V16, Ba3 
 \item[kumbhakaṃ] V7, V20, V18, V12, V13, 
 \item[kumbhakaḥ] V3
-\item[]
-\item[(illegible/unavailable)] Bo3, C5, J14, V1, V6, V5, V2, 
+\item[(illegible/unavailable)] Bo3, C5, J14, P10, V1, V6, V5, V2, T, G1, P16, P22, R5,
   \begin{description}
@@
 \sthana{2.55a}
-\item[cakram āsādya(ḥ)] A1, B3, Ba, Bo1, C2, V1, J8pc, J10, J17, P28, V12, N21(unusual way of writing “dya”), V14, V20, V18, V16
-\item[cakrasāmānya(ḥ)] B1, J8ac, J11, J13, N19, N24, V3, V22
-\item[cakrasammānyaḥ] Ba5, Tue, V10
+\item[cakram āsādya(ḥ)] A1, B3, Ba, Bo1, C2, V1, J8pc, J10, J17, N1, N7, N10, N28, V12,
+ N21, P1, P2, P5, P8, P9, P12, P13, P17, P24, P25, P26, P27, V14, V20, V18, V16, Ba3
+\item[cakramāsādyamān] 
+\item[cakrasāmānya(ḥ)] B1, J8ac, J11, J13, N19, N24, P11, P15, P18, V3, V22
+\item[cakrasammānyaḥ] Ba5, N4a, N13, P14, Tue, V10, G1
 \item[cakrastamānyaṃ] J2
+\item[cakramāsādyamān] G11
+\item[cakram āsāya] N9,
 \item[cakrasāmānyaṃ] LD2
@@
 \item[cakrasāmānya] J4, J5,
-\item[cakraṃ samāsādya] N17, W4
+\item[cakrasāmānyaḥ] N3, 
+\item[cakraṃ samāsādya] N6, N14, N17, W4
 \item[cakrasamāsādya] LD1,
-\item[cakranāśāya] N20, N26, V17
-\item[cakrabhogyaś ca] N22
-\item[cakrasevyas tu] Bo2, C1, C4, C6, C8, J1, J3, J6, J7, J15, V19, N18(“ca” doubtful), V21, N25, V15, V28, V2, V13, V21
+\item[cakranāśāya] N20, N26, P10, V17
+\item[cakranāsānyaḥ] N12, 
+\item[cakrabhogyaś ca] B2,
+\item[cakrasevyas tu] Bo2, C1, C4, C6, C8, J1, J3, J6, J7, J15, N2, N5, N8, N11, N16, P7, V19, N18, V P20, P21, P23, 21, N25, V15, V28, V2, V13, V21
 \item[cakrasevyas ta] N23
-\item[cakrabhogyas tu] N22
+\item[cakrabhogyas tu] N22, P6, 
 \item[cakrābhimānyaḥ] Ko,
@@
 \item[cakramadyāya] J16,
-\item[(illegible/unavailable)] Bo3, C5, J14, V6, V5
+\item[(illegible/unavailable)] Bo3, C5, J14, V6, V5, T, P16, P22, R1, R5,
   \begin{description}
@@
  \begin{marma}[hp02_56*1]
-\sthana{2.55*1[c][d]}
-\item[satyāmūlena randhreṇa] C4, J6, J8pc, J10, J17, V21, N25
+\sthana{2.55*1}
+\item[satyāmūlena randhreṇa] C4, J6, J8pc, J10, J17, N5, N9, P1, P10, V21, N25
 \item[satyāmūlena randhrena] J15,
+\item[satyaṃ mūlena randhreṇa] P5, P24, 
 \item[sadāmūlena randhreṇa] J3, J6pc
 \item[satyāmūlena randhrenayaḥ] V18, V16, V21
-\item[spṛśyamūlena randhreṇa] B1, B3, C2, J13,  
-\item[rasanātālumūlena] N17, 
+\item[spṛśyamūlena randhreṇa] B1, B3, C2, J13, P2, P12, P17, 
+\item[rasanātālumūlena] N6, N14, N17, 
 \item[rasanāmūla randhreṇa] C9, 
-\item[rasavatyāmūla randhre yaḥ] N26
+\item[rasavatyāmūla randhre yaḥ] N20, N26
 \item[]
 \item[]
-\item[(illegible/unavailable)] A1, B2, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C5, C6, C8, J1, J2, J4, J5, J7, J11, J12, J14, J16, Ko, LD1, LD2, P28, W4, V3, V6, V20, V15, V5, V28, V22, V2, V12, V13, V10, V14, V17
+\item[(illegible/unavailable)] A1, B2, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C5, C6, C8, J1, J2, J4, J5, J7, J11, J12, J14, J16, Ko, LD1, LD2, N1, N2, N3, N4a, N7, N8, N10, N11, N12, N13, N16, N18, N19, N21, N22, N23, N24, P6, P7, P8, P9, P11, P13, P14, P15, P18, P20, P21, P22, P23, P25, P26, P27, P28, R1, R5, W4, V3, V6, V20, V15, V5, V28, V22, V2, V12, V13, V10, V14, V17, G11, Tue, T, Ba3, G1, P16,
   \begin{description}
@@
 \sthana{2.57a}
-\item[rasanātālumūlena] A1, B1, B2, B3, Bo1, C2, C4, C9, J4, J10, J12, J13, J15, J16, J17, Ko, LD1, N21, N22(tālū), N25, N26, V1, V20, V15, V28
+\item[rasanātālumūlena] A1, B1, B2, B3, Bo1, C2, C4, C9, J4, J10, J12, J13, J15, J16, J17, Ko, LD1, N1, N2, N3, N6, N7, N10, N11, N12, N14, N17, N21, N22, N25, N26, P1, P2, P5, P6, P8, P9, P10, P11, P12, P13, P15, P20, P23, P24, P26, P27, R1, V1, V20, V15, V28, Ba3
 \item[rasanātālumūle yaḥ] Ba,
 \item[rasanātālumūlenaya] V14,
-\item[rasanātālumūlenayaḥ] V2, V18, V12, V3, V16, V17
+\item[rasanātālumūlena yaḥ] V2, V18, V12, V3, V16, V17, G11
+\item[rasanātāgru mūlena yaḥ] G1
 \item[rasanāttālumūlena] N20
 \item[rasanātālumūlaṃ ca] J5, 
-\item[rasanātāluyogena] Bo2, C6, C8, J3, J6, J7, N18, V21, N23, V13
+\item[rasanātāluyogena] Bo2, C6, C8, J3, J6, J7, N8, N16, N18, P7, V21, N23, V13
 \item[rasanātālayogena] J1,
+\item[rasanā tālamūlena] N5, N9, 
 \item[rasanāṃ tālamūlena] J11
+\item[rasanāṃ tālumūlena] P25,
+\item[rasanāṃ tāluyogena] P21, 
 \item[rasanānāṃ tālamūlena] N19
@@
 \item[]
-\item[(illegible/unavailable)] Bo3, C5, J2, J14, Tue, LD2, N24, V6, V5, V22, V10
+\item[(illegible/unavailable)] Bo3, C5, J2, J14, N4a, Tue, LD2, N13, N24, P14, P17, P18, P22, V6, V5, V22, V10, T, P16, R5,
   \begin{description}
@@
 \sthana{2.58b}
-\item[pūrvavat kumbhakād anu] C4, C6, C8, J1, J4, J6pc, J7, J11, Ko, N21, V14, V21,  N24, N25, V6, V19, V15, V28, V2
+\item[pūrvavat kumbhakād anu] C4, C6, C8, J1, J4, J6pc, J7, J11, Ko, N2, N5, N8, N11, N21, P13, P20, P23, V14, V21, N3, N24, N25, P27, V6, V19, V15, V28, V2, Ba3, G1
+\item[pūrvavat kuṃbhanādan] G11
 \item[pūrvavat kūmbhakād anu] Bo2,
+\item[pūrvavat kumbhayed anu] P26,
+\item[pūrvavat kumbhanāda tu] P11,
 \item[pūrvavat kumbhakād anmuṃ] Bo1,
@@
 \item[pūrvavat kumbhakeṣv anu] B2
-\item[pūrvavat kumbhasādhanaṃ] B1, B3, Ba, Ba5, Bo3, C1, C2, C3, C9, J2, J3, J6, J10, J12, J13, J15, J16, J17, LD1, LD2, N18(“sā” unclear), N22, V3, Tue, V22, V18, V12, V13, V3, V10, V16
+\item[pūrvavat kumbhasādhanaṃ] B1, B3, Ba, Ba5, Bo3, C1, C2, C3, C9, J2, J3, J6, J10, J12, J13, J15, J16, J17, N4a, LD1, LD2, N6, N9, N10, N12, N13, N14, N17, N18, N22, P2, P5, P6, P10, P12, P24, P25, R1, V3, Tue, V22, V18, V12, V13, V3, V10, V16
+\item[pūrvaṃ ca  kumbhasādhanaṃ] P1, P8, P9, 
+\item[pūrvava kumbhasādhanaṃ] N7, 
 \item[kākacaṃcusamannayāt] V17
-\item[kākacaṃcusamunnayāt] N26
-\item[kākacaṃcusamunnarayāt] N20(mark over “kā”, perhaps “ra” is cancelled), 
-\item[(illegible/unavailable)] C5, J14, V5
+\item[pūrvavat kumbhasādhayet] N1, P17, 
+\item[pūrvakumbhakavad bhavet] P7,
+\item[kākacaṃcusamunnayāt] N16, N26
+\item[kākacaṃcusamunnarayāt] N20, 
+\item[kākacaṃcusamanvayāt] P21, 
+\item[(illegible/unavailable)] C5, J14, P14, P15, V5, T, P16, P18, P22, R5,
   \begin{description}
@@
 \item[kapāle ca samantataḥ] V1 
+\item[kapāla pavanaṃ tataḥ] N6, 
 \item[kapāle niḥsvanaṃ tathā] Ba,
+\item[kapole niḥsvanaṃ tathā] P25, 
 \item[kapole niḥsvanaṃ tadā] C1, 
@@
 \item[kapole svasanaṃ tataḥ] V18
-\item[kapāla śvasamaṃ tataḥ] J4, 
+\item[kapāla śvasamaṃ tataḥ] J4, P10, 
 \item[kapāla śasanaṃ tataḥ] J15, 
-\item[kapāle śvasanaṃ tataḥ] J11, V17 
+\item[kapāle śasanaṃ tataḥ] Ba3
+\item[kapāle śvasanaṃ tataḥ] J11, N3, P20a, V17 
 \item[kapāle svaśanaṃ tataḥ] LD1, 
-\item[kapāla śparśanaṃ tataḥ] J16, 
+\item[kapāla śparśanaṃ tataḥ] J16, N9, N10, P5, P24, 
 \item[kapole svaśanaṃ tataḥ] LD2, V16
+\item[kapole śvāsanaṃ tataḥ] N20, 
+\item[kapole svasanaṃ tataḥ] N7, P8, 
 \item[kapole śvavaṇe tathā] B2,
 \item[kapole śvavaṇe tataḥ] V13, V3
+\item[kapāle sasvanaṃ tataḥ] P13, P15, P27, 
+\item[kapole sasvanaṃ tataḥ] P9,
+\item[kapolaṃ sasvaramarut] G11
+\item[kapolaṃ sasvanaṃ tataḥ] P11, 
 \item[kapola saśvanaṃ tataḥ] J5,
@@
 \item[kayolesvaśanaṃtataḥ] N24
+\item[kapāla pavanaṃ tata] N14, 
 \item[kapāla sasvanaṃ tataḥ] N21, V12
+\item[kapāla svasanaṃ tataḥ] N19, 
 \item[kapālasya samaṃ tataḥ] C3
@@
 \item[kapālādhi sasvanaṃ tataḥ] V6
-\item[kapālasparśanaṃ tataḥ] C9, J17,
+\item[kapāla pavanaṃ tataḥ] N17,
+\item[kapāle susvanam] G1
+\item[kapālasparśanaṃ tataḥ] C9, J17, P1,
 \item[kapālasya śanaṃ tataḥ] J12,
 \item[kapālādhi niḥsvanaṃ] C6,
-\item[kapālāvadhi sasvanaṃ] A1, Ba5, C8, J7, Tue, V7, V20, V28, V22, V10
+\item[kapālavidhi sasvanaṃ] P26,
+\item[kapālāvadhi sasvanaṃ] A1, Ba5, C8, J7, N4a, N12, N13, P7, P14, P18, Tue, V7, V20, V28, V22, V10
 \item[kapālāvadhi sasanaṃ] V21
-\item[kapālāvadhi pūrayet] B1, B3, Bo2, C2, C4, J1, J3, J6, J13, N25, V15, V2
+\item[kapālāvadhi pūrayet] B1, B3, Bo2, C2, C4, J1, J3, J6, J13, N1, N2, N5, N8, N11, N16, N18, N25, P2, P12, P17, P20, P21, P23, V15, V2
 \item[kepālavadhi sasvanaṃ] N23, 
-
-\item[(illegible/unavailable)] Bo3, C5, J14, V5
+\item[kepālena samaṃ tataḥ] P6, 
+\item[kepālenastha sarīrasthaṃ] N22,
+ \item[(illegible/unavailable)] Bo3, C5, J14, V5, T, P16, P22, R1, R5,
   \begin{description}
@@
 \sthana{2.65cd}
+\item[dhārayan nāsikāmadhyṃ aṃgulībhyāṃ tathā dṛḍham] N10,
+\item[dhārayennasikā madhye aṅgulīnāṃ tathā dṛḍham] Ba3
 \item[dhārayen nāsikāmadhye aṅgulībhyāṃ vinā dṛḍham] V1, V21
 \item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] LD1,
-\item[dhārayen nāsikāmadhyaṃ aṅgulībhyāṃ tathā dṛḍham] A1, C9, 
+\item[dhārayen nāsikāmadhyaṃ aṅgulībhyāṃ tathā dṛḍham] A1, C9, P26, 
 \item[dhāraye nāsikāmadhyaṃ maṅgulānāṃ tathā dṛḍham] J12,
@@
 \item[dhārayen nāsikāmadhya aṅgulībhyāṃ tathā dṛḍham] J10,
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] B2, C3, J17, V18, V12, V16, V17
+\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] B2, C3, J17, N6, N14, N17, P5, P8, P10, P24, V18, V12, V16, V17, G1
+\item[dhāraye nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] N7, 
 \item[dhārayen nāsikāmadhyai aṅgulībhyāṃ tathā dṛḍham] J15,
+\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tadā dṛḍham] N9, P1, 
 \item[dhārayen nāsīkāmadhye maṃgulībhā tathā dṛḍham] J16,
 \item[dhārayen nāsīkāmadhye aṅgulībhyāṃ dṛḍham tathā] LD2,
-\item[dhārayen nāsikāmadhye aṅguṣṭhābhyāṃ tathā dṛḍham] N26,
+\item[dhārayen nāsikāmadhye aṅguṣṭhābhyāṃ tathā dṛḍham] N20, N26,
+\item[dhārayen nāsikāṃ madhye aṅgulībhyāṃ tathā dṛḍham] P9, 
+\item[dhārayen nāsikābhyāṃatarjabhyāṃ tathā dṛḍham] R1,
 \item[dhārayen nāsikāmadhyṃ tarjanībhyāṃ vinā dṛḍham] Ko,
-\item[dhārayen nāsikāmadhyaṃ tarjanībhyāṃ vinā dṛḍham] J11,
-\item[dhārayen nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] C6, J5, V3, N24, V6, V7, V28, V22, V2, V3, V14
+\item[dhārayen nāsikāmadhyaṃ tarjanībhyāṃ vinā dṛḍham] J11, N3, 
+\item[dhārayen nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] C6, J5, N12, N18, N19, P6, V3, N21, N24, P13, P17, P25, P27, V6, V7, V28, V22, V2, V3, V14, G11
+\item[dhārayen nāsikāmadhyaṃ tarjanībhyāṃ vinā dṛḍham] P7,
 \item[dhāraye nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] Ba,
-\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] Ba5, Bo2, J3,
-\item[dhāraye nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] Ba5, V15
-\item[dhārayen nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] J7,
+\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] Ba5, Bo2, J3, N11, N16, P11, P14, P20, 
+\item[dhārayen nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] P23,
+\item[dhāraye nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] P18,
+\item[dhāraye nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] Ba5, N4a, N13, V15
+\item[dhāraye nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] N8, 
+\item[dhārayen nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] J7, P21, 
+\item[dhāvayen nāsikāmadhyāṃ tarjanībhyāṃ vinā dṛḍham] P15,
 \item[dhāraye nāsikāṃ madhye tarjjanībhyāṃ vinā dṛḍham] Bo1,
-\item[dhārayen nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] B1, B3, J13, V13
+\item[dhārayen nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] B1, B3, J13, N1, N2, P2, P12, V13
 \item[dhārayan nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] C2, 
@@
 \item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ yathā dṛḍham] C8
+\item[dhārayen nāśikāmadhye sarjanībhyāṃ yathā dṛḍham] N22,
 \item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] J6pc,
@@
 \item[dhārayen nāsikāṃ madhyāt tarjanībhyāṃ vinā dṛḍham] Tue
-\item[dhārayen nāsikāmadhyāt tarjanībhyāṃ vinā dṛḍham] C4, N25, V10
+\item[dhārayen nāsikāmadhyāt tarjanībhyāṃ vinā dṛḍham] C4, N5, N25, V10
 \item[dhārayaṃ nāśikāmadhye tarjanabhyāṃ vinā dṛḍham] N23
 \item[dhāraae vāsikāmadhye tarjanībhyāṃ vinā dṛḍham] J2,
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5
+\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, T, P16, P22, R5,
   \begin{description}
@@
 \item[kaṇdalībodhakaṃ viprabhavaghnaṃ sukhadaṃ śubhaṃ] V1
+\item[kuṇḍalī bodhakaḥ kubho rogaghnaḥ sukhada śubhaḥ] Ba3
 \item[kuṇdalībodhanaṃ sarva doṣaghnaṃ sukhadaṃ śivaṃ] Bo1
 \item[kuṇdalībodhanaṃ sarva doṣaghnaṃ sukhadaṃ śuvaṃ] V15
-\item[kaṇdalībodhakaṃ ksipraṃpavanaṃ sukhadaṃ hitaṃ] V7
+\item[kuṇdalībodhanaṃ sarva doṣa?? sūkhadaṃ sūkhaṃ] N24,
+\item[kaṇdalībodhakaṃ ksipraṃ pavanaṃ sukhadaṃ hitaṃ] Tue, V7
 \item[kuṇḍalo bodhaka kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] J2
-\item[kuṇḍalībodhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] Ba5, V22
+\item[kuṇḍalībodhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] Ba5, N13, P14, P18, V22
 \item[kuṇḍalabodhakaṃ pāpāghnaṃ suṣadaṃ śubhaṃ] Ko,
@@
 \item[kaṇḍalībodhanaṃ cakaṃ pāpaghnaṃ sukhada śubhaṃ] V14
-\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] B2, Bo2, C4, J3, J6, J7, J11,  
+\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] B2, Bo2, C4, J3, J6, J7, J11, N16, N19, N23, N25, P6, P21, P23, P27, 
+\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ śubhadaṃ sukhaṃ] P20,
+\item[kuḍalībodhanaṃ kuryāt pāpaghna suṣadaṃ subhaṃ] N18, 
+\item[kuṇḍaṃlībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] N8, N11, 
+\item[kuṇḍalabodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] N5, 
 \item[kaṇḍalībodhanaṃ kuryāt pāpaghnaṃ subhaṃ tad śukhaṃ] V28
 \item[kuṇdalībādhakaṃ ksipraṃ pāpghnaṃ sukhadaṃ hitaṃ] Ba,
+\item[kuṇdalībodhakaṃ ksipraṃ pāpaghnaṃ sukhadaṃ hitaṃ] P25, 
+\item[kumḍalīśhodhanaṃ kuyāt pāpaghnaṃ ca sukhaṃ śhubhaṃ] N22, 
+\item[kuṇdalībādhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] N4a,  
+\item[kaṇḍalībodhanaṃ kuryāt tāpaghnaṃ sukhadaṃ śubhaṃ] N2, 
+\item[kuṇḍalībodhakaḥ kuṃbhaḥ rogaghnaḥ sukhadaḥ śubhaḥ] N6,
+\item[kuḍalībodhaka kuṃbhao rogaghnaḥ sukhadaḥ śubhaḥ] P5, P8, P9, P24, 
+\item[kuḍalībodhanaḥ kuṃbhao rogaghnaḥ sukhadaḥ śubhaḥ] P7,
+\item[kuṇḍalībodhakaḥ kuṃbho rogaghnaḥ sukhadaṃ śubhaḥ] N9, N10, N17, N20,  N26, P1, P10, 
+\item[kuḍalībodhakaḥ kuṃbhao rogaghnaṃ sukhadaṃ śubhaṃ] P17,
 \item[kuṇḍalībodhakaḥ karthūḥ roghnaṃḥ sukhadaḥ śubhaḥ] A1, V20
+\item[kuṇḍalībodhaka kumbhao rogaghnaṃ sukhadaḥ śubham] N14, P2,
+\item[kuṇḍalībodhakaṃ cakraṃ bhāvaghnam sukhadaṃ śubhaṃ] N3,
+\item[kuṇḍalībodhanaṃ cakra bhāvaghnam sukhadaṃ śubhaṃ] P11,  
+\item[kuṇḍalībodhakaṃ kṣipram bhavaghnam sukhadaṃ śubhaṃ] N12,
+\item[kuṇḍalībodhakaṃ vipram bhavaghnam sukhadaṃ śubhaṃ] P15,
 \item[kuṇḍalībodhanaṃ karttur bhavaghnam sukhadaṃ śubhaṃ] C6,
@@
 \item[kuṇḍalībodhanaṃ kartuṃ bhavaghnam sukhadaṃ śubhadaṃ] C8,
+\item[kuṇḍalībodhakaḥ kartuḥ rogaghnasukhadaḥ śubhaṃ] P26, 
+\item[kuṇḍalībodhakaḥ kumbhoḥ ragaghnaḥ sukhadaḥ śubhaḥ] N7, 
 \item[kuṇḍalībodhakaḥ kumbho roghnaḥ sukhadaḥ śubhaḥ] C3, J10, J17, 
@@
 \item[kuṇḍalībodhakaḥ kumbho rogaghnaḥ sukhadaḥ śubhaḥ] B1, B3, C2, J13, V18, V12, V13, V3, V17 
+\item[kuṇḍalībodhakaḥ kumbho rogaghnaṃ sukhadaṃ śubhaṃ] P12,
+\item[kuṇḍali bodhakaṃ vakrabhāvaghnaṃ sukhadaṃ śubhaṃ] G11
+\item[kuṇḍalībodhakaṃ sarva doṣaghnaṃ sukhadaṃ hitaṃ] N21, P13, 
+\item[kuṇḍalībodhakaḥ kumbho rogaghnaṃ sukhadaṃ śubhaṃ] N1, 
 \item[kuṇḍalīno bodhacakraṃ? bhāvadaṃ supada śubhāṃ] J5
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, V10
+\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, V10, T, G1, P16, P22, R1, R5,
   \begin{description}
@@
 \item[guṇatrayasamudbhūtaṃ] Bo1,
-\item[samagātrasamudbhūtaṃ] J4,
+\item[guṇatrayasamudbhūta] N21,
+\item[samagātrasamudbhūtaṃ] J4, N7, 
+\item[samagātrasamā....] P10, 
 \item[samyagātrasamudbhūtaṃ] J7,
 \item[samyagātrasamudbhutaṃ] J15,
-\item[samyaggātrasamudbhūtaṃ] B2, Bo2, C4, C8, J2, J11, V1, V7, V2, V3, V17
+\item[samyagātrasamudbhuta] P14,
+\item[samyaggātrasamudbhūtaṃ] B2, Bo2, C4, C8, J2, J11, N3, N9, N16, N19, P13, V1, V7, V2, V3, V17
+\item[samyagatrasamudbhutaṃ] N12,
+\item[samyaggātrasamudbhūta] N2, N4a, N11, N13, N20, N23, N26, P15, P18, P25, P27, Tue, 
+\item[samyagātrā sumuvbhutaḥ?] N24,
 \item[samya gātre samudbhūtaṃ] J5,
-\item[samyaggātre samudbhūtaṃ] J6, V28
+\item[samyaggātre samudbhūtaṃ] J6, N5, N8, N25, P20, P23, V28
 \item[samyaggātre samudbhūta] J3,
+\item[samyaggātre samudbhūtat] P6, 
 \item[samyag gātraṃ samudbhūtaṃ] Ko,
@@
 \item[samyagataḥ samudbhūtam] V14
-\item[samyak gātrasamudbhūta] Ba, V22
+\item[samyak gātrasamudbhūta] Ba, V22, G11, P11, 
 \item[samyag antaḥ samudbhū] A1
 \item[samyag antaḥ samudbhūta] C6
-\item[samyak bhastrā samudbhūtā] B1, B3, C2, J12, J13, J16,  
-\item[samyak bhastrā samudbhūtaḥ] V21  
+\item[samyak bhastrā samudbhūtā] B1, B3, C2, J12, J13, J16, N1, N10, P1, P2, P8, P9, P12, P17, Ba3
+\item[samyak bhastrā samudbhūtaḥ] P26, V21  
 \item[samyak bhastrā samabhyāsaḥ] V13
-\item[samyag bhastrā samudbhūtāṃ] LD1, LD2, 
-\item[samyak bhastrā samudbhūto] C3, J10, J17, 
+\item[samyag bhastrā samudbhūtāṃ] LD1, LD2, N18, P7, P21, 
+\item[samyak bhastrā samudbhūto] C3, J10, J17, N14, N17, P5, P24, 
 \item[samyag bhastri samudbhūtaṃ] C9
+\item[samyag bhastrī samudbhūto] N6, 
 \item[sasyagātrasamudbhūtaṃ] J1
 \item[samagātrasamudbhūtaṃ] V6, V16
+\item[samagātre samudbhūta] N22,
 \item[samagātrasamudbhūtaghnaṃ] V18
-\item[(illegible/unavailable)] Ba5, Bo3, C1, C5, J14, V5, V12, V10
+\item[(illegible/unavailable)] Ba5, Bo3, C1, C5, J14, V5, V12, V10, T, G1, P16, P22, R1, R5,
   \begin{description}
@@
 \item[vegair ghoṣaṃ] V1
-\item[vegākṛṣṭaṃ] Ba, J2, V15
+\item[vegākṛṣṭaṃ] Ba, J2, N21, P25, V15
+\item[vegākṛṣṇaṃ] P13,
+\item[vegā gho] N23,
 \item[vego tad ghoṣa] J1
-\item[vego ghoṣam] C3, C9, J16,  
+\item[vego ghoṣam] C3, C9, J16, N3, N9, N16, N20, P1, P5, P10, P24, P27, 
+\item[vego ghoṣa] N24,
 \item[vego ghoṣaḥ] J15,
-\item[vegāt ghoṣaṃ] A1, B2, B3, Ba5, C9, LD1, V6, V7, V22, V18, V12, V13, V14, V16, V17
-\item[vegā ghoṣaṃ] V3
-\item[vegot ghoṣaṃ] J12, V2
+\item[vegāt ghoṣaṃ] A1, B2, B3, Ba5, C9, N4a, LD1, N13, N26, P14, P17, P18, P26, R1, Tue, V6, V7, V22, V18, V12, V13, V14, V16, V17, Ba3
+\item[vegad ghoṣa] N10,
+\item[vegad ghoṣaṃ] N7, 
+\item[vegā ghoṣaṃ] P8, V3
+\item[vegot ghoṣaṃ] J12, N14, N17, P11, V2, G11
 \item[vegot ghoṣe] V2,
 \item[vegodveṣaḥ] J4,
+\item[vegodvegaṃ] P20a,
+\item[vegodveṣaṃ] N5, P15, P20, 
 \item[vaiṣodvaiṣaṃ] J5,
 \item[veget vege] V28
-\item[vegād ghoṣaṃ] J10, J17, V20
-\item[vegod ghoṣaṃ] Bo2, C6, J10pc, J11,
+\item[vegād ghoṣaṃ] J10, J17, N1, V20
+\item[vegod ghoṣaṃ] Bo2, C6, J10pc, J11, N18, P6, P9, P21, 
+\item[vegod ghoraṃ] P23, 
 \item[vedod ghoṣaṃ] V21
+\item[vedyo ghoṣaṃ] N22,
 \item[vegodyoṣaṃ] J7,
+\item[vegopyoyaṃ] N19, 
 \item[vemod ghoṣaṃ] J3, J6, 
-\item[vegodvegaṃ] C4, J13, 
+\item[vegodvegaṃ] C4, J13, N25, 
+\item[vegod deyaṃ] P7, 
 \item[vegād aghoṣaṃ] B1,
-\item[vegād doṣaṃ] Bo1,
+\item[vegād doṣaṃ] Bo1, N2, 
+\item[vegod doṣaṃ] N6, 
 \item[veṇu doṣaṃ] C2,
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, Ko, V5, V10
+\item[veṇu ghoṣaṃ] P2,
+\item[veṇud ghoṣaṃ] P12, 
+\item[(illegible/unavailable)] Bo3, C1, C5, J14, Ko, N8, N11, N12, V5, V10, T, G1, P16, P22, R5,
   \begin{description}
@@
 \sthana{2.73a}
-\item[ārecyāpūrya yat kuryāt] 
+\item[ārecyāpūrya yat kuryāt] N2, N5, 
+\item[ārecya āpūrya yat kuryāt] P23,
+\item[ārecyāpūrya ?????] N11, 
 \item[ārecyāpūrya vat kuryāt] J13,
+\item[ārecyāpūrvavat kuryāt] N25,
+\item[recakaṃ pūrakaṃ kuryāt] N1, N10, P2, 
 \item[recya vā pūrya sanakaiḥ] J11,
 \item[recake pūrakaṃ kuryāt] J12, J16, 
+\item[recayet pūrakaṃ kuryāt] N16, N19, N23, P6, 
+\item[recakaṃ pūrakaṃ kuryāt] P12, P17, 
+\item[ārecyāpūrya kuṃ?yāṃt] N8,
 \item[recakaḥ pūrakaṣ kāryyaḥ] J17,
-\item[recakaḥ pūrakaḥ kāryaḥ] J15, 
+\item[recakaḥ pūrakaḥ kāryaḥ] A1, N9, J15, N17, N26, P1, P5, P9, P24, R1, 
+\item[recakaḥ pūrakaḥ kuryāt] P21, 
+\item[recanaṃ pūraṇam kuryāt] P13,
+\item[recakaḥ pūrakaḥ kā....] P10,
+\item[recakaḥ pūrakaḥ kāryāḥ] N7, 
+\item[recakaḥ pūrakaḥ kāryaṃ] N20
+\item[recakaḥ pūrakaḥ kāryo] P8, 
+\item[recakaḥ pūrvavat kāryaḥ] N3, 
+\item[recakaṃ pūrakaṃ kuryāt] B1, B3, 
+\item[ārecaṃ pūrvavat kuryāt] P20,
 \item[recayitvā pūrakaḥ kāryaḥ] J14, 
-\item[recake pūrakaṃ uktwā] V14
+\item[reco vā pūrakaḥ kāryaḥ] P11, 
+\item[rocyatvā pūrakaḥ kuṃbhakaḥ] N24,
+\item[recakaḥ pūrakaṃ tyatvā] N18, 
+\item[recakaṃ pūrakaṃ muktā] N22,
+\item[recake pūrakaṃ uktwā] V14, G11, Ba3
+\item[recake pūrṇaṃ kāryaṃ] Ba,
+\item[recake pūraṇaṃ kāryaṃ] P25,
+\item[recanaṃ pūranaṃ kuryāt] N21,
 \item[recaka pūraka kumbhai] V3, V16
 \item[recaka pūraka kumbhakaiḥ] V17
+\item[recakaṃ pūrakaṃ muktyā] G1
 \item[recaka pūraka varjite] V10
-\item[(illegible/unavailable)]
+\item[virecya pūrakaḥ kāryaḥ] P7,
+\item[prāṇāyāmoyam ity uktaḥ] N4a, N12, P14, P18, Tue,
+\item[(illegible/unavailable)] B2, N6, N13, N14, P15, T, P16, P22, P26, P27, R5,
   \begin{description}
@@
 \sthana{2.73b}
-\item[sa vai sahitakumbhakaḥ] J12, J13, J15, 
-\item[sarve sahitakumbhakaḥ] J17,
+\item[śanaiḥ saṃhitakumbhakaḥ] N16, 
+\item[sarve kevalakumbhakam] G1
+\item[śanaiḥ sahitakumbhakaḥ] N23, P7, 
+\item[śanaiḥ sahitakumbhakaiḥ] P11,
+\item[śanaiḥ sahitakumbhakaṃ] P6, 
+\item[sa vai kevalakumbhakaḥ] N4a, N18, N22, P18, P26, Tue, 
+\item[sa vai sahitakumbhakaḥ] A1, B1, B2, B3, J12, J13, J15, N1, N3, N5, N10, N17, N21, P2, P8, P12, P13, P17, P20, P21, P23, P27, 
+\item[sarve sahitakumbhakaḥ] J17, N20, 
+\item[savyai sahitakumbhakaḥ] N19, 
+\item[sarvai sahitakumbhakaḥ] N9, N12, N26, P1, P9, P10, R1, 
+\item[sarvaiḥ sahitakumbhakam] Ba, P25,
+\item[sarvai sā?ti kumbhakaṃ] N25,
+\item[sarvai kevalakumbhaka] N14,
+\item[sarvair ahitakumbhakaḥ] N7, 
 \item[sa vai sahitakumbhak] J16, 
-\item[sa taiḥ sahitakumbhakaḥ] J14,
-\item[sa vai kevalakumbhakaḥ] V10
+\item[sa taiḥ sahitakumbhakaḥ] J14, N24, 
+\item[sa vai kevalakumbhakaḥ] Ba3, G11, P14, V10
+\item[sarvaiḥ sahitapūrakaḥ] P5, P24, 
 \item[kāryaḥ sahitakumbhakaḥ] J11,
+\item[kuryat sahitakumbhakaḥ] N2, N8, 
+\item[kuryat saṃhitakumbhakaḥ] N11, 
 \item[kāryaḥ śanai sahitakumbhakā] V3, V14
@@
 \item[kāryaḥ śarveiḥ sahitakumbhaka] V17
-\item[(illegible/unavailable)]
+\item[(illegible/unavailable)] N6, N13, P15, T, P16, P22, R5,
   \begin{description}
@@
 \begin{marma}[hp02_78a]
-\sthana{kumbhakaḥ prāṇarecānte} C3, J2, J3, J13, V1, V6, V28, V18, V16
-\sthana{kumbhakāt prāṇarecānte} V17
-\item[kumbhaka prāṇarecānte] A1, B1, B2, Bo2, C1, J1, J5, LD1, V2, V12
+\sthana{kumbhakaḥ prāṇarecānte} C3, J2, J3, J13, N8, N11, N16, N21, P5, P24,V1, V6, V28, V18, V16, G11,
+\sthana{kumbhakāt prāṇarecānte} N20, N26, V17
+\item[kumbhaka prāṇarecānte] A1, B1, B2, Bo2, C1, J1, J5, LD1, N2, N7, N9, N10, N12, N14, N17, N22, N25, P1, P2, P9, P10, P12, P13, P17, P20, P21, P23, V2, V12, Ba3
+\item[kumbhaka prāṇarecyaṃte] N1,
 \item[kumbhake siddha reca puraka] V21
@@
 \item[kumbhaka prāṇa recāttaḥ] J16,
-\item[kumbhakaṃ prāṇarecānte] B3, C4, C9, J6, 
+\item[kumbhaka prāṇarecātte] P8, P26,
+\item[kumbhakaṃ prāṇarecānte] B3, C4, C9, J6, N5, N19, P6, P27, G1, R1, 
 \item[kumbhakaṃ prāṇarecātteḥ] J15,
@@
 \item[kumbhakaṃ ghrāṇarecānte] LD2, 
-\item[kumbhitaḥ prāṇarecānte] Bo1, V15
-\item[kumbhītaḥ prāṇarecānte] Ko, J11, J14, 
+\item[kumbhitaḥ prāṇarecānte] Bo1, N23, N24, P11, P15, V15
+\item[kumbhītaḥ prāṇarecānte] Ko, J11, J14, N3,  
 \item[kumbhīte prāṇarecānte] V14
 \item[kumbhītaprāṇareṣānte] J4,
-\item[tad etat prāṇarrecānte] C6, C8, 
-\item[kumbhakaḥ prāṇarodhānte] Ba, Ba5, V7, V20, V22, V10
-\item[(illegible/unavailable)] Bo3, C1, C5, V5, V13
+\item[tad etat prāṇarrecānte] C6, C8, P7, 
+\item[kumbhakaḥ prāṇarodhānte] Ba, Ba5, P14, P18, P25, Tue, V7, V20, V22, V10
+\item[kumbhaka prāṇarodhānte] N4a, N13,
+\item[(illegible/unavailable)] Bo3, C1, C5, N6, N18, V5, V13, T, P16, P22, R5,
   \begin{description}
@@
 %%% End:
-
-[a]Perhaps "yathā" can be removed from these readings
-[b]Perhaps "yathā" can be removed from these readings
-[c]I don't know what part of a verse this number means
-[d]I think it is 56*1 @nilsliersch@web.de isn't it? 2nd thing..do we need full verse or just a line or a word here?

### 2023-04-11 latex/CommStyle.tex
new base text

--- 2022-06-07 latex/HP1_TranslComm.tex
@@
-\input{Preamble-Print}
 \makepagestyle{HPed}
@@
 
-% additions/changes 2022-03-12 mm; 2022-6-1 MD:
 \SetTEIxmlExport{autopar=false}
@@
 \NewDocumentEnvironment{sources}{O{}}{
-	\textcolor{blue}{\textbf{Testimonia:}}}{}
+	\textcolor{blue}{\textbf{Sources:}}\linebreak}{}
+\NewDocumentEnvironment{testimonia}{O{}}{
+	\textcolor{blue}{\textbf{Testimonia:}}\linebreak}{}
+%\newcommand{\var}[1]{\footnotesize\textup{#1}}
+\def\vl{\textit{v.l.}}
+\def\var#1{\\ {\small #1}}
+\def\sl#1{\emph{#1}}
 
+\linespread{1}
 \setlength{\parskip}{0.3em}
 \setlength\parindent{0pt}
+\setlength{\vindent}{0pt}
 \def\startverse{\begin{ekdverse}\begin{itshape}}
-\def\endverse{\end{itshape}\end{ekdverse}}
+\def\endverse{\normalsize\end{itshape}\end{ekdverse}} % \textup
 \setvnum{}
-
-\begin{document}
-\pagestyle{HPed}
-\begin{ekdosis}
-\SetLineation{lineation = none,}
-
-\chapter*{Translation \& philological commentary}
-\subsection*{1.1}
-\begin{translation}[hp01_001]
-Homage to the glorious Ādinātha by whom the knowledge of Haṭhayoga was taught. It shines forth like a ladder for one desirous of climbing to the lofty terrace of the royal palace.
-\end{translation}
-
-\begin{sources}[hp01_001]
-Yogasārasaṃgraha (line 3090), Gheraṇḍasaṃhitā
-\end{sources}
-
-%\begin{philcomm}[hp01_001]
-%\end{philcomm}
-
-\subsection*{1.2}
-\begin{translation}[hp01_002]
-Having bowed to the glorious guru, the Lord, the yogi Svātmārāma has taught the doctrine of Haṭhayoga solely for [attaining] Rājayoga.
-\end{translation}
-
-%\begin{philcomm}[hp01_002]
-%\end{philcomm}
-
-\subsection*{1.3}
-\begin{translation}[hp01_003]
-For those who are ignorant of Rājayoga because of confusion in the darkness of many opinions the compassionate Svātmārāma composes the Lamp on Haṭha.
-\end{translation}
-
-%\begin{philcomm}[hp01_003]          
-%\end{philcomm}
-
-\subsection*{1.4}
-\begin{translation}[hp01_004]
-In fact, Matsyendra, Gorakṣa and other [siddhas] knew the doctrine of Haṭha, and the yogi Svātmārāma knows it owing to their favour.
-\end{translation}
-
-%\begin{philcomm}[hp01_004]          
-%\end{philcomm}
-
-\subsection*{1.5}
-\begin{translation}[hp01_005]
-The glorious Ādinātha, Matysendra, Śābara, Ānandabhairava, Cauraṅgī, Mīna, Gorakṣa, Virūpākṣa, Bileśaya,
-\end{translation}
-
-%\begin{philcomm}[hp01_005]          
-%\end{philcomm}
-
-\subsection*{1.6}
-\begin{translation}[hp01_006]
-Manthānabhairava, Siddhabuddha, and Kanthaḍi, Pauraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
-\end{translation}
-
-%\begin{philcomm}[hp01_006]          
-%\end{philcomm}
-
-\subsection*{1.7}
-\begin{translation}[hp01_007]
-Kānerī, Pūjyapāda, Nityanātha, Nirañjana, Kapālī, Bindunātha, and the one named Kākacaṇḍīśvara.
-\end{translation}
-
-%\begin{philcomm}[hp01_007]          
-%\end{philcomm}
-
-\subsection*{1.8}
-\begin{translation}[hp01_008]
-Allamaprabhudeva, Ghoḍācolī, Ṭiṇṭiṇī, Bhālukī and Nāgabodha and Khaṇḍakāpālika.
-\end{translation}
-
-%\begin{philcomm}[hp01_008]          
-%\end{philcomm}
-
-\subsection*{1.9}
-\begin{translation}[hp01_009]
-These great siddhas and others have destroyed the rod of death through the power of Haṭhayoga and wander in the universe.
-\end{translation}
-
-%\begin{philcomm}[hp01_009]          
-%\end{philcomm}
-
-\subsection*{1.10}
-\begin{translation}[hp01_010]
-Haṭha is considered a refuge for those who are burnt by the pains of transmigration. Haṭha is the foundational tortoise for the worlds of all yogas.
-\end{translation}
-
-%\begin{philcomm}[hp01_010]          
-%\end{philcomm}
-
-\subsection*{1.11}
-\begin{translation}[hp01_011]
-The doctrine of Haṭha should be kept very secret by those yogis who are desiring success. When it is secret it becomes potent. However, when it has been revealed, it becomes impotent.
-\end{translation}
-
-\begin{sources}[hp01_011]
-ŚS 5.254; BKhP 10v4, YCM
-\end{sources}
-
-%\begin{philcomm}[hp01_011]          
-%\end{philcomm}
-
-\subsection*{1.12}
-\begin{translation}[hp01_012]
-In well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live remotely in a small hut.
-\end{translation}
-
-\begin{sources}[hp01_012]
-GŚ 32cd; BKhP 107v1, YCM
-\end{sources}
-
-%\begin{philcomm}[hp01_012]          
-%\end{philcomm}
-
-\subsection*{1.13}
-\begin{translation}[hp01_013]
-It has a small door, without cracks and holes, its length is not too high or low, thickly smeared with cow dung in the proper way, clean, free from everything that annoys, adorned with a pavilion, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
-\end{translation}
-
-\begin{sources}[hp01_013]
-BKhP 107v3, YCM
- \end{sources}
-
-\begin{philcomm}[hp01_013]          
-\emph{piṭharaṃ} good
-\end{philcomm}
-
-\subsection*{1.14}
-\begin{translation}[hp01_014]
-Locating oneself in a hut of such a kind, free from all worry, one should constantly practise yoga in the way taught by one's guru.
-\end{translation}
-
-%\begin{philcomm}[hp01_014]          
-%\end{philcomm}
-
-\subsection*{1.15}
-\begin{translation}[hp01_015]
-Overeating, exertion, chatter (gossiping/bickering?), sticking to rules, associating with people, inconstancy: through [these] six, yoga will be abandoned.
-\end{translation}
-
-\begin{philcomm}[hp01_015]          
-Impossible to decide on meaning of \emph{niyamāgraha}, avagraha invisible, Jyotsnā takes it as over-insistence (as if \emph{āgraha} was implied) as he relates it to extreme ascetic practice; will be abandoned if \emph{prahāsyate} is adopted.
-\end{philcomm}
-
-\subsection*{1.16}
-\begin{translation}[hp01_016]
-From zeal, conviction, resolve, understanding of the truth [of yoga], discernment, abandonment of associating with people: by [these] six, on the other hand, yoga is successful.
-\end{translation}
-
-\begin{philcomm}[hp01_016]          
-\emph{tattva} can sometimes refer to the practices of yoga: e.g., \emph{tritattva} in AS 13.12, 14.2--3.
-\end{philcomm}
-
-\subsection*{1.17}
-\begin{translation}[hp01_017]
-Because it is the first auxiliary of haṭha, āsana is taught first. This (\emph{tad}) āsana brings about steadiness, good health and dexterity.
-\end{translation}
-
-%\begin{philcomm}[hp01_017]          
-%\end{philcomm}
-
-\subsection*{1.18}
-\begin{translation}[hp01_018]
-I shall now teach some of the postures which have been accepted by sages (\emph{munis}) such as Vasiṣṭha and yogis such as Matsyendra.
-\end{translation}
-
-%\begin{philcomm}[hp01_018]          
-%\end{philcomm}
-
-\subsection*{1.19}
-\begin{translation}[hp01_019]
-Correctly placing the soles of both feet between the knees and thighs [and] sitting up with the body straight: they call that the auspicious [pose].
-\end{translation}
-
-\begin{sources}[hp01_019]
-VS 1.68 etc
-\end{sources}
-
-%\begin{philcomm}[hp01_019]          
-%\end{philcomm}
-
-\subsection*{1.20}
-\begin{translation}[hp01_020]
-One should place one's right heel on the left, at the side of the [lower] back, and on the right the left in the same way. This is the Cow's Mouth [posture]; it is like a cow's mouth.
-\end{translation}
-
-\begin{sources}[hp01_020]
-VS 1.70, YY? 3.5; BKhP 93v7
-\end{sources}
-
-%\begin{philcomm}[hp01_020]          
-%\end{philcomm}
-
-\subsection*{1.21}
-\begin{translation}[hp01_021]
-Having placed one foot on one thigh and the seated thigh on the other foot. Vīrāsana is taught [thus]. 
-\end{translation}
-
-\subsection*{1.22}
-\begin{translation}[hp01_022]
-Having blocked the anus with the crossed ankles, focused [the yogi] remains [thus]. Adepts of yoga know that this is kūrmāsana.
-
-Or: While focused, [the yogi] should bind the anus with crossed ankles. This is kūrmāsana. Experts in yoga know this.
-\end{translation}
-
-\subsection*{1.23}
-\begin{translation}[hp01_023]
-Having assumed padmāsana, inserted the hands between the knees and thighs, and placed [the hands] on the ground, [the yogi] remains in the air. This is kurkuṭāsana
-
-Or: [The yogi] should assume padmāsana, insert his hands between the knees and thighs onto the ground and place himself in the air. This is kurkuṭāsana.
-\end{translation}
-
-\begin{philcomm}[hp01_023]
-Kurkuṭa/kurkkuṭa variant in \getsiglum{V1,J10ac,V3} is attested in Pañcatantra (M-W).
-\end{philcomm}
-
-\subsection*{1.24}
-\begin{translation}[hp01_024]
-Bound in kurkuṭāsana, [the yogi] binds the neck with the hands and lies supine like a tortoise. This is uttānakūrmāsana. 
-\end{translation}
-
-\begin{philcomm}[hp01_024]
-\emph{°bandhasthaḥ} or \emph{°vat kṛtvā}?
-Only V1 has latter, which is simpler.
-Are the others trying to improve it?
-Stemmatically ambiguous as \emph{°bandha°} is on one branch (V3/J8, V19) and \emph{°madhya°} the other (J10, J17, N17).
-V1 looks like an outlier.
-
-Adopt \emph{bandhastho} with note. \emph{°bandhastha} not found in any other texts. \emph{°vat kṛtvā} is not good.
-\end{philcomm}
-
-\subsection*{1.25}
-\begin{translation}[hp01_025]
-Having clasped the big toes with hands, [the yogi] should perform a bow drawing action as far as the ear. This is called dhanurāsana.
-\end{translation}
-
-\begin{philcomm}[hp01_025]
-At a stretch (!), this could mean drawing a crossbow, with both hands taken back to the ears, but probably just means like a normal bow.
-
-Gheraṇḍasaṃhitā:
-\startverse
- prasārya pādau bhuvi daṇḍarūpau |
- karau ca pṛṣṭhaṃ dhṛtapādayugmam |\\
- kṛtvā dhanustulyavivartitāṅgaṃ |
- nigadyate vai dhanurāsanaṃ tat || 2.18 ||
-\endverse
-\end{philcomm}
-
-\subsection*{1.26}
-\begin{translation}[hp01_026]
-Having grasped the right foot, which is placed at the base of the left thigh, and the left foot, which is wrapped around the outside of the knee, [the yogi] remains with his body twisted.
-This āsana was taught by Matsyendranātha.
-\end{translation}
-
-\subsection*{1.27}
-\begin{translation}[hp01_027]
-Through practice, Matsyendra's seat gives people a torch for the abdomen, a destructive weapon for a range of terrible diseases, the awakening of K and steadiness of the spine.
-\end{translation}
-
-\begin{philcomm}[hp01_027]
-Difficult to construe first half of verse: \emph{jaṭhara} compound needs to be in apposition to \emph{matsyendrapīṭhaṃ}.
-KDham ed has variant \emph{jaṭharapradīpaṃ} but \emph{pradīpa} is masculine.
-Perhaps we have to accept \emph{jaṭharapradīptiṃ}, but that crosses stemmatic lines, so \emph{jvalanapradīptiṃ}; or \emph{jaṭharapravṛddhiṃ}.
-\emph{jvalana} is found at 3.66 to refer to the fire in the body but \emph{jaṭhara}, stomach, is not found without \emph{agni} elsewhere.
-
-YCM has \emph{jaṭharapravṛddhiḥ}.
-
-Adopt \emph{jaṭharapradīptiṃ} (\emph{°dīptiṃ} is an emendation), understood as object of \emph{dadāti}.
-
-Pāda d, J10ac and J17 have \emph{candra} for \emph{daṇḍa}, also in YCM and 6-chapter HP.
-\end{philcomm}
-
-\subsection*{1.28}
-\begin{translation}[hp01_028]
-[The yogi] should stretch out both feet on the ground like staffs, hold the ends of both feet with the hands, place the forehead upon the knees and remain thus.
-They call this the back-stretch (paścimatānam).
-\end{translation}
-
-\begin{philcomm}[hp01_028]
-Only V1 has \emph{dorbhyāṃ padāgra}, others have variations on the much inferior \emph{dvābhyāṃ karābhyāṃ}. Vulgate has same as V1, which is a surprise.
-
-V1 reading \emph{dorbhyāṃ pādāgradvitayaṃ} is best, adopt.
-
-V1’s \emph{tāṇabandhaḥ} doesn’t work with \emph{idaṃ}.
-\end{philcomm}
-
-\subsection*{1.29}
-\begin{translation}[hp01_029]
-This back-stretch is the foremost among āsanas. It makes the breath flow in the back [i.e. Central channel], increases the digestive fire, makes the belly thin and prevents diseases in men.
-\end{translation}
-
-\begin{philcomm}[hp01_029]
-Adopt \emph{arogitāṃ}.
-\end{philcomm}
-
-\subsection*{1.30}
-\begin{translation}[hp01_030]
-Supporting oneself on the ground with both hands, their elbows placed on either side of the navel, with a raised position (?
-uccāsanaḥ) one is placed up into the air [as straight] as a stick.
-They call this posture the peacock.
-\end{translation}
-
-\begin{philcomm}[hp01_030]
-1.30 b tat refers to \emph{karadvaya} (cf. VS).
-\emph{uccāsano} to be adopted; \emph{samunnataśiraḥpādo} is reading of VS. which is derived from Vimānārcanākalpa 96: 
-\emph{karatale bhūmau samsthāpya kūrparau nābhipārśvayor nyasya na(unnata)
-̇ taśira ̄hpādau dandavad vyomni samsthito mayūrāsanam iti}.
-\end{philcomm}
-
-\subsection*{1.31}
-\begin{translation}[hp01_031]
-The glorious mayūra posture gets rid of all diseases of the abdomen such as bloating and overcomes [imbalances of] the doṣas.
-It completely incinerates food which is bad or has been eaten to excess, it generates digestive fire and it digests strong poison.
-\end{translation}
-
-\begin{philcomm}[hp01_031]
-\emph{aśeṣaṃ} better
-\end{philcomm}
-
-\subsection*{1.32}
-\begin{translation}[hp01_032]
-Lying on one’s back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] all [other] āsanas and relaxes the mind.
-\end{translation}
-
-\subsection*{1.33}
-\begin{translation}[hp01_033]
-However, Śiva taught eight-four āsanas. I shall take the four essential ones from them and teach them. 
-\end{translation}
-
-\subsection*{1.34}
-\begin{translation}[hp01_034]
-The adept, lotus, lion and auspicious pose are the best tetrad and, among those, one should always sit in a comfortable adept’s pose. 
-\end{translation}
-
-\subsection*{1.35}
-\begin{translation}[hp01_035]
-Having joined the place of the perineum with the heel of the foot, the yogi should firmly fix the [other] foot on the penis.
-Having held the face and chest together and the body erect, [the yogi] remains still, his senses restrained, gazing between the brows with his eyes unmoving.
-This breaks open the door to liberation and is called the adept’s pose.
-\end{translation}
-
-\begin{philcomm}[hp01_035]
-KDh ed has \emph{āsya}, variant in mss kh and gha, both BORI.
-Jürgen suggests we might take \emph{āsyahṛdaye} and \emph{vigraham} as the objects of \emph{dhṛtvā} and samam as an adverb (i.e., holding straight the face, chest and body).
-However, in a different doctrine.
-\end{philcomm}
-
-\subsection*{1.36}
-\begin{translation}[hp01_036]
-Having fixed the left heel on the penis, and put the other heel on that, this is siddhāsana. 
-\end{translation}
-
-\subsection*{1.37}
-\begin{translation}[hp01_037]
-Some proclaim this is siddhāsana, others know it as vajrāsana, a few say it is muktāsana and some guptāsana.
-\end{translation}
-
-\subsection*{1.38}
-\begin{translation}[hp01_038]
-The Siddhas know Siddhasana as the single most important amongst
-all postures, in the same way as measured diet amongst rules and non-violence amongst observances. 
-\end{translation}
-
-\begin{philcomm}[hp01_038]
-\emph{iva} or \emph{eva}?
-\emph{iva} does work — like \emph{siddhāsana}, \emph{mitāhāra} and \emph{ahiṃsā} are the best, but for it to work properly \emph{mitāhāra} and \emph{ahiṃsā} should be accusative.
-V19 has acc + \emph{iva}, which seems best, especially with \emph{siddhāḥ viduḥ}, but this might be a correction as V19 often corrects.
-However, one old KDham BORI (?) ms (pha, 1695 CE) has it, as does Jyotsnā, so adopt.
-
-In pāda d V19 has \emph{siddhāsanam idaṃ viduḥ}, but the reading of all other mss is preferable.
-
-Clearly based on DYŚ 33, which includes \emph{ekaṃ} and \emph{mukhya}.
-\end{philcomm}
-
-\subsection*{1.39}
-\begin{translation}[hp01_039]
-Among the eighty-four postures, one should regularly practise just Siddha; in the same way one should practise Suṣumnā among the 72,000 channels.
-\end{translation}
-
-\begin{philcomm}[hp01_039]
-Odd to have \emph{suṣumnām} as object of \emph{abhyaset}.
-YCM has this reading though.
-Perhaps cd were added somewhat indiscriminately by Svātmārāma (with nominative \emph{suṣumnā}) and then others tried to make sense of it.
-
-Some witnesses, including Jyotsnā, have \emph{nāḍīnāṃ malaśodhanam/e} for pāda d, which is probably a patch (no other texts say siddhāsana clears the channels), but cf.
-Amṛtasiddhi in which the practices are said to bring about cālana of the nāḍīs (e.g.
-11.6).
-
-Good example of early contamination.
-
-[\emph{nāḍiṣu} is better supported \getsiglum{J10ac,V19,J17}.]
-\end{philcomm}
-
-\subsection*{1.40}
-\begin{translation}[hp01_040]
-By meditating upon the self, restricting the diet and regularly practising Siddhāsana for twelve years, the yogi attains the Niṣpatti stage.
-What’s the point of lots of exhausting postures when there is Siddhāsana?
-\end{translation}
-
-\begin{philcomm}[hp01_040]
-Only possible variant is \emph{mitāhāro} in V19.
-
-J8 might be correction of J10’s unmetrical reading.
-
-[\emph{sadāsiddhāsanābhyāsād}? Or maybe read \emph{sadā} with \emph{avāpnuyāt}]
-
-V19 is found in YCM: \emph{śramadair bahubhịh pīṭhaiḥ kiṃ syāt siddhāsane sati}; JM this seems best to me.
-
-Is this notion of āsanas causing \emph{śrama} already current in HY texts?
-[It is mentioned in the verse on śavāsana]
-\end{philcomm}
-
-\subsection*{1.41}
-\begin{translation}[hp01_041]
-When the Prāṇa breath has been carefully stopped in kevala kumbhaka, the Unmanī [state] arises automatically, without effort, in the same way that
-\end{translation}
-
-\begin{philcomm}[hp01_041]
-V1 has 
-\startverse
-śramādau bahubhiḥ pīṭhais sadā siddhāsane sati  |\\
-prāṇānile sāvadhānaṃ baddhe kevalakumbhake  || 1.41 ||
-\endverse
-Mixing up both versions of the verse — contamination already?
-\end{philcomm}
-
-\subsection*{1.42}
-\begin{translation}[hp01_042]
-when just Siddhāsana alone is always firmly bound, the three bandhas arise automatically, without effort.
-\end{translation}
-
-\begin{philcomm}[hp01_042]
-\emph{dṛḍhe} goes across stemma, although \emph{ṃ} vs \emph{e} is moot.
-\end{philcomm}
-
-\subsection*{1.43}
-\begin{translation}[hp01_043]
-There is no posture like siddhāsana, no breath like the restrained breath, no mudrā like khecarī, no dissolution like nāda.
-\end{translation}
-
-\begin{philcomm}[hp01_043]
-\emph{na kumbhasadṛśo nilaḥ} is difficilior lectio and attested by all early witnesses except V19: (almost?) all testimonia have \emph{kumbhaḥ kevalopamaḥ}; Śivasaṃhitā has \emph{kumbhasadṛśaṃ balam}.
-\end{philcomm}
-
-\subsection*{1.44}
-\begin{translation}[hp01_044]
-Now lotus pose.
-One should place the right foot on the left thigh, and the left on the right though, hold firmly the big toes with the hands behind the back, place the chin on the chest and gaze at the tip of the nose.
-This is called lotus pose, which destroys diseases for those who undertake the yamas.
-\end{translation}
-
-\subsection*{1.45–47}
-\begin{translation}[hp01_045]
-However, in another view,
-Having put the upturned feet carefully on the thighs and the upturned hands in the middle of the thighs, one should fix the eyes on the tip of the nose.
-Having raised the root of the uvula with the tongue, one should place the chin on the chest and gradually [draw in\footnote{The verb \emph{ākṛśya} follows in the next verse in the \emph{Dattātreyayogaśāstra}.}] the breath [...].
-This is taught as lotus pose, which cures all diseases.
-\end{translation}
-
-\begin{philcomm}[hp01_045]
-\emph{uttabhya} vs \emph{uttambhya}.
-The witnesses split predictably along the two main branches of the stemma. The evidence of the DYŚ is important here.
-
-\emph{vakṣasyāsthāpya} is a marmasthāna
-
-%\emph{pavanaṃ śanaiḥ} is left hanging.
-The readings with \emph{vakṣa sthāpayet} do not offer a solution nor indicate how Svātmārāma may have redacted this to make the syntax complete.
-Instead, it seems that he quoted these two verses (1.45–46) from the Dattātreyayogaśāstra and simply omitted the next verse that made sense of \emph{pavanaṃ śanaiḥ} because it was not relevant to the posture itself.
-
-\emph{pavanaṃ śanaiḥ} is left hanging, perhaps, because of an eyeskip that happened early in the transmission. The following verse in the DYŚ ends with \emph{pavanaṃ śanaiḥ}.
-
-\startverse
-nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\
-uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\
-yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\
-yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\
-idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |
-\endverse
-
-% Jü:
-[JH] The background of the passage \emph{rājadantamūlaṃ ca jihvayā uttambhya} in 1.46 is more complex
-than it may appear. Here it is a literal quotation from the DYŚ, but many other Haṭhayogic texts
-teach a particular position of the tongue, in which it is in one or the other way turned back in
-the direction of the uvula, as we read explicitly in the \emph{Vivekamārtāṇḍa} (REF):
-\emph{kapālakuhare jihvā praviṣṭā viparītagā}. Brief references to this practice can become
-ambiguous for various reasons, and this has possibly confused Brahmānanda.
-
-One reason is that there is a, probably older, rule for meditation postures according to which the
-tongue rests near the teeth. One instance would be \emph{Svacchandatantra}, which teaches a
-meditation pose called \emph{divyaṃ karaṇam} (4.365f.), in which the tongue is to rest at the tip
-of the teeth (\emph{dantāgre jihvām ādāya}). Other Tantric texts have this or similar rules, in
-which the tongue is supposed to rest either on the teeth or the palate,\footnote{This rule is
-  found in \emph{Iśānaśivapaddhati}: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān
-    asaṃspṛśan ṛjukāyaḥ}. REF} the earliest instance being perhaps \emph{Mrgendrāgama} ?.18.
-Placing the tongue where it does not disturb the meditation seems quite appropriate for a
-``normal'' meditative practice.
-
-[Jason: There’s also a clear reference to the tip of the tongue being placed in the middle of the palate in 2.27 of the Yogapāda of the Mataṅgapārameśvaratantra (tālumadhyagatenaiva jihvāgreṇa mahāmune).
-In fact, in works that predate Haṭhayoga (i.e., 11th c.), the most common instruction is to put the tip of the tongue on the palate (tālu).]
-
-We might try to interpret the passage in this manner, however, once Haṭhayogic physiology is at the
-background, we must assume that the aim is to reach back to the uvula, to the source of the
-"nectar".  For the background and for the crucial references see Mallinson's note on
-\emph{Khecarīvidyā} 1.65ab.\footnote{p. 209.}  Confusingly Yogic terminology has used and possibly
-invented new names for uvula, and among these especially the term \emph{rājadanta} may give rise to
-confusion, since, as we have seen, the tongue might also in some Yoga systems be placed at the
-(front) teeth.
-
-Furthermore the details in these descriptions of the \emph{khecarīmudrā} are manifold. The 10th
-century \emph{Mokṣopāya} says that the tongue rests at the ``source of the
-palate''\footnote{\emph{tālumūlatalālagnajihvā} MU V.55.14c.} and the commentary
-\emph{Saṃsārataraṇi} on the parallel passage in \emph{Laghuyogavāsiṣṭha} V.6.155, which reads
-\emph{tālumūlāntarālagnajihva-}, explains that this means that the tongue is to be placed in the
-middle of the two regions of the palate, and that this is the \emph{nabhomudrā}, alias
-``\emph{khecarī}''.\footnote{\emph{tālumūlāntarālagnajihvamūlah. tālumūlayoh. kākudamūladeśayoh. āntare
-    lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā / yā hi khecarı̄ty ucyate.}}
-
-A little later in the \emph{Mokṣopāya} it is made clear that one should reach the uvula, ``placed a
-the root of the palate''.\footnote{\emph{tālumūlagatāṃ yatnāj jihvayākramya ghaṇṭikām}
-  (V.78.24ab)}
-
-In view of this background we must conclude that the author of the \emph{Jyotsnā} was probably not
-aware of the Yogic meaning of \emph{rājadanta} and has tried his best to make sense of the passage,
-echoing the idea of the two roots of the palate (although his text is not talking about the
-palate), but then wisely refers to the instruction of the teacher for practical details, probably
-noticing that his literal interpretation is somewhat opaque (I omit the synonyms for clarity):
-
-\begin{quote}
-Pressing both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.
-
-\emph{rājadantānāṃ daṃṣṭrāṇāṃ savyadakṣiṇabhāge sthitānāṃ mūle ubhe mūlasthāne jihvayā uttambhya ūrdhvaṃ stambhayitvā | guru-mukhād avagantavyo 'yaṃ jihvā-bandhaḥ |}
-\end{quote}
-
-[Jason: also, Brahmānanda’s comments on 3.22 indicate that he thought rājadanta refers to the front teeth.
-When commenting on rājadantasthajihvāyāṃ, he says kutaḥ ?
-yato dantānāṃ rājāno rājadantā rājadanteṣu tiṣṭhatīti rājadantasthāḥ, rājadantasthā cāsau jihvā ca tasyāṃ rājadantasthajihvāyāṃ bandhaḥ, taduparibhāgasya sambandhaḥ śastaḥ].
-\end{philcomm}
-
-\subsection*{1.47}
-\begin{translation}[hp01_047]
-It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
-\end{translation}
-
-\begin{philcomm}[hp01_047]
-Durlabham is ambiguous as to whether the posture is hard to perform or hard to acquire. The latter is the usual reading.
-\end{philcomm}
-
-\subsection*{1.48}
-\begin{translation}[hp01_048]
-Next is taught the doctrine of Matsya.
-
-[The yogi] should put the hands together in a bowl shape, very firmly assume padmāsana, firmly place the chin on the chest and meditation in the mind, [and] raising the Apāna breath over and over again, releasing the held Prāṇa, a man attains unequalled knowledge through the power of the goddess [Kuṇḍalinī].
-\end{translation}
-
-\begin{philcomm}[hp01_048]
-It is unusual to find Matsyendra referred to as simply Matsya as in the subheading here.
-This verse is Vivekamārtaṇḍa 36.
-The teachings of the VM are attributed to Gorakṣa, not Matsyendra, although in the extended recension of it known as Gorakṣaśataka etc.
-the second verse includes homage to Mīna, i.e.
-Matsyendra.
-
-The end of pāda b is uncertain.
-Later witnesses including Brahmānanda have \emph{dhyāyaṃś ca} but none of the early ones does.
-So we have to take \emph{dhyānaṃ} with \emph{sannidhāya}.
-That makes tat problematic though: what does it refer to?
-
-The two participles \emph{proccālayan} and \emph{muñcan} imply that the two things are happening at the same time, which is surprising but perhaps possible.
-Or perhaps one can understand them as sequential.
-
-In the VM this verse comes in the passage on raising Kuṇḍalinī, so tat could refer to visualising K opening the mokṣadvāra.
-\end{philcomm}
-
-\subsection*{1.49}
-\begin{translation}[hp01_049]
-While in padmāsana the yogi should fill himself up [with Prāṇa] through [all] the channels.  He who takes in breath [in this way] is sure to be liberated.
-\end{translation}
-
-\subsection*{1.50--52}
-\begin{translation}[hp01_050]
-Now the lion’s pose
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. Left ankle is on right, and the right ankle on the left.
-
-Having placed both hands on the knees and spread his fingers, his mouth open, he gazes with his eyes fixed on the tip of the nose.
-This is the lion’s pose, which is always worshipped by yogis.
-It unites the three locks and is the best of all postures.
-\end{translation}
-
-\subsection*{1.53--54}
-\begin{translation}[hp01_053]
-Now, the friendly pose.
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
-Having firmly and motionlessly held the sides of the feet with the hands, this is bhadrāsana, which cures all diseases and poisons.
-Yogis of the siddha tradition call it Gorakṣāsana.
-\end{translation}
-
-\subsection*{1.55}
-\begin{translation}[hp01_055]
-When the great yogi does not become tired [while engaged] in these āsanas and bandhas, he should now practise the breath techniques of mudrā and so forth in order to purify the channels.
-\end{translation}
-
-\begin{philcomm}[hp01_055]
-55d is ra-vipulā with \emph{nāḍi}
-
-\emph{nāḍiśuddhiṃ} is difficult to construe.
-Jim thinks it is a second accusative of the \emph{abhyaset} and reads it as \emph{nāḍiśuddhyartha}.
-I wonder if it is not better understood as a bahuvrīhī qualifying \emph{°pavanakriyām} (i.e., the yogi should practise a breath technique that is channel-purifying.
-In other words, he does not practise just any breath technique).
-As a bahuvṛīhi, it may also be understood as a \emph{hetuviśeṣaṇa} (i.e., why should the yogi practice a breath-technique such as mudrā?
-Because it is channel-purifying).
-\end{philcomm}
-
-\subsection*{1.55*1}
-\begin{translation}[hp01_055_1]
-Success arises for one engaged in practice. How can it arise for one who has no practice? Success in yoga does not arise by merely reading scriptures.
-
-Wearing a robe does not bring about success, nor does talking [about yoga].
-Practice alone is the cause of success.
-This is true, there is no doubt.
-In this tradition, it should not be given to one who wears robes and is devoted to his genitals and stomach.
-\end{translation}
-
-\begin{philcomm}[hp01_055_1]
-1.55*1–2 are omitted from V1 and V19, so may not have been included in the earliest version of the text.
-In fact, it is possible that both were added (perhaps initially as marginal notes) to elaborate on kriyā in 1.55d.
-Both verses are similar to verses from the Dattātreyayogaśāstra, and these verses (except 1.55.2ef) appear in the vulgate, but towards the end of chapter 1 (1.65–66).
-
-And then just half of DYŚ 47 is given, resulting in a near-nonsensical hemistich.
-
-The syntax of 1.55.2ef is corrupt. One has to emend to \emph{deyā} to make sense of it. 
-\end{philcomm}
-
-\subsection*{1.55*3}
-\begin{translation}[hp01_055_3]
-In me, the pure ocean of awakening, is this empty bubble we call the universe dissolved or does it arise? Where does this veil of doubt about this come from?
-\end{translation}
-
-\begin{philcomm}[hp01_055_3]
-It is very difficult to find a reason why this verse should be inserted here. It is apparently a
-\emph{muktaka} that would befit an accomplished spiritual poem more than our \emph{Haṭhapradīpikā},
-even here, in what appears as a sort of miscellaneous section at the end of a chapter.  In this
-verse, the lyrical subject wonders about why the mind is still able to doubt, despite its insight
-into the nature of reality and the reader wonders, how the illusionist verse could be understood to
-fit our Yoga text, perhaps the scribe of the archetype of V19 and N17, was fond of it.
-
-The source is, as far as we can say, the \emph{Śāntiśataka} of the Kashmirian poet Sillana or
-Silhaṇa,\footnote{The mss.\ of his \emph{Svātmopalabdhiśataka} give the name as Sillana, the mostly
-Bengali mss. of the Śāntiśataka read Śilhaṇa, as does Aufrecht in his CC. See Hanneder:
-forthcoming.} who cannot be dated with any certainty, but predates the HP by a few centuries. The
-edition of this text – where hundred original verses had to be identified – places it into an
-appendix of doubtful stanzas,\footnote{Karl Schönfeld: \emph{Das Śāntiśataka}. Leipzig:
-Harrassowitz 1910, p.\,90 (A9).} but the editor saw no compelling reason not to regard it as
-original except through the fact that it is not transmitted in all manuscripts.  What prevents
-further investigation of the matter is the lack of Kashmirian mss.\ for the \emph{Śāntiśataka} and
-its compilatory character: one quarter of the material is identical with Bhartṛhari's
-\emph{Vairāgyaśataka}. A still superficial glance at Sillana's \emph{Svātmopalabdhiśataka} gives
-the impression that our verse would fit there, but not so much in the \emph{Śāntiśataka}.  While
-these are only preliminary observations the verse is probably not original in any stage of
-development of the HYP, but limited to V19 (and N17).
-\end{philcomm}
-
-\subsection*{1.55*4}
-\begin{translation}[hp01_055_4]
-Realisation from scripture, one's own guru or oneself and the cessation of mind; all these are methods that have been systematised and taught by the wise in this tradition.
-\end{translation}
-
-\begin{philcomm}[hp01_055_4]
-last greyscale verse (\emph{śrutipratītiḥ…}) is only in V19 of old mss (but some other later ones) and is quoted in Yogacintāmaṇi.
-
-The reading in the YCM (\emph{manaso nirodhaḥ}) is better in a yogic context.
-\end{philcomm}
-
-\subsection*{1.56}
-\begin{translation}[hp01_056]
-The various āsana, breath retention, bodily technique (karaṇa) called seals (mudrā), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.
-\end{translation}
-
-\begin{philcomm}[hp01_056]
-Verse is omitted from V1, probably deliberately as it doesn’t have chapter 4, which teaches nādānusandhāna. The numbering in V1 suggests that its exemplar had this verse.
-
-Kumbhaka is almost always masculine.
-
-Marmasthāna: not clear whether to adopt \emph{citro} or \emph{citraṃ}, or V19’s \emph{mudrākhyaṃ karaṇaṃ tathā} or the others’ \emph{mudrādikaraṇāni ca}.
-
-See also 1.65, which has \emph{mudrādikaraṇāni ca}, so perhaps it was through confusion with this that the same reading is found in some witnesses of 1.56.
-
-Yes, V19 reading probably best.
-\startverse
-āsanaṃ kumbhakaś citraṃ mudrākhyaṃ karaṇaṃ tathā |\\
-atha nādānusandhānam abhyāsānukramo haṭhe ||
-\endverse
-
-It seems that the four aṅgas of Haṭhayoga are being referred to in the singular (hence \emph{āsanaṃ}), whereas in 1.65 the plural is used (i.e., \emph{pīṭhāni}).
-Therefore, we should adopt \emph{citraṃ [...] karaṇaṃ tathā}.
-\end{philcomm}
-
-\subsection*{1.57}
-\begin{translation}[hp01_057]
-Celibate, restricted in diet and devoted to yoga, a yogi succeeds in upwards of a year. No doubt about this should be entertained.
-\end{translation}
-
-%\begin{philcomm}[hp01_057]
-%\end{philcomm}
-
-\subsection*{1.58}
-\begin{translation}[hp01_058]
-When very unctuous and sweet food that is without a quarter portion [of the stomach] is eaten for love of śiva, it is called a restricted diet (\emph{mitāhāra}).
-\end{translation}
-
-\begin{philcomm}[hp01_058]
-This verse probably derives from the ‘original’ GŚ (12c–13b). It is also found, but reworked to be about the \emph{mitāhārī}, in Nowotny’s GŚ (extended recension of VM) at 55.
-\end{philcomm}
-
-\subsection*{1.59}
-\begin{translation}[hp01_059]
-[Adepts] say the [following] is unwholesome: pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink.
-Flesh of goats and sheep, curds, diluted buttermilk, poor man's pulse, Jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like.
-\end{translation}
-
-\begin{philcomm}[hp01_059]
-59a \emph{kaṭvamla°} is better than \emph{kaṭvāmla°} and well attested elsewhere in lists of tastes and types of foods.
-On the meaning of \emph{uṣṇa} (in relation to food) see Meulenbeld 1974: 254 fn.
-13: ‘Cakra mentions as a variant: \emph{katvamlalavaṇakṣāra} (pungent.
-acid, saline and caustic).
-Cakra remarks that the term `hot' (\emph{uṣṇa}) denotes hot on touch when it occurs the first time, and hot with regard to potency when it occurs for the second time.’
-
-\emph{°hari}(\emph{ī} here for metre?)\emph{taśāka°} in some nighaṇṭus is horseradish, which makes better sense here.
-\startverse
-śigrur haritaśākaś ca śākapattraḥ supattrakaḥ | \textup{Rājanighaṇṭu 7.26}\\
-śigruko haritaśākaś ca mato vai mūlapatrakaḥ | \textup{Sauśrutanighaṇṭu 75ab}\\
-\endverse
-Brahmānanda’s understanding of \emph{harītaśāka} as \emph{patraśāka} is probably wrong if \emph{patraśāka} was intended as ‘leafy vegetables’.
-But perhaps \emph{patraśāka} can also mean \emph{śigru} (note \emph{śākapattraḥ} above).
-
-Anusvāra at end of \emph{śāka}?
-
-\emph{°sauvīra°} probably means sour gruel.
-Brahmānanda: \emph{sauvīra} = \emph{kāñjika} (fermented rice water).
-Meulenbeld, madhavanidāna pp. 516–517
-\emph{sauvīra} is sour gruel made from barley and wheat. On the process see, Suśruta 1.44.35--40ab.
-(PV Sharma’s translation of this passage:)
-
-\begin{quote}
-‘Roots of trivṛt etc., the first group (\emph{vidārigandhādi}), \emph{mahat pañcamūla}, \emph{mūrvā} and \emph{śārṅgaṣṭā}, and also of \emph{snuhī}, \emph{haimavatī}, \emph{triphalā}, \emph{ativiṣā} and \emph{vacā} -- these are taken and divided into two parts out of which one is decocted and the other is powdered; now, crushed barley grains are impregnated with the above decoction several times, dried and then slightly fried.
-Taking three parts of this and one part of the above powder are put in a jar and mixed with their (of \emph{trivṛt}, etc.) cold decoction and fermented properly.
-This is known as \emph{sauvīraka}.’
-\end{quote}
-
-But according to some nighaṇṭus, \emph{sauvīra} can also mean stibnite (an ingredient in some añjana’s and medicines):
-\startverse
-añjanaṃ yāmunaṃ kṛṣṇaṃ nādeyaṃ mecakaṃ tathā \\
-srotojaṃ dṛkpradaṃ nīlaṃ sauvīraṃ ca suvīrajam //  \textup{Rājanighaṇṭu 13.86}
-\endverse
-Note also that the Yogaprakāśikā takes \emph{sauvīra} with \emph{taila}, perhaps to solve the problem of \emph{taila} on its own (see below for more on this): \emph{sauvīrataila} -- oil produced in the place Suvīra (\emph{suvīradeśodbhavatailam}).
-
-Suvīra , a country the people of which, also called Suvira (V.79), Sauvira (XVI.21) and Sauvīraka (IV.23) ...
-S.M.
-Ali identifies it with the Rohri - Khairpur region of Sind (Geography of the Purānas, Delhi, 1966, p.
-144).
-
-\emph{taila} could refer to \emph{tilataila}: 
-Śārṅgadharasaṃhitā: 
-\emph{anuktāvasthāyāṃ paribhāṣāvidhiḥ [...] taile ’nukte tilodbhavam} 48
-
-Dominik Wujastyk supplied this reference and may be able to comment more on taila in this list.
-
-On the translation of \emph{madya}, see Mchugh (An Unholy Brew) 2021: 8.
-
-\emph{ajāvimāṃsa}: note the variant \emph{ajādimāṃsa}, which makes better sense to me as \emph{aja} might simply indicate that the flesh of animals is meant (as opposed to humans). 
-\emph{ājāvi°} is needed for the metre.
-
-\emph{kulattha} is a kind of pulse, translated by Dominik Wujastyk as `poor man's pulse' (see Roots of Ayurveda).
-
-\emph{kola}: Zizyphus Jujuba (Nadkarni 1926: pp.
-919-920).
-Also known as \emph{badara}.
-This is how Brahmānanda takes it (\emph{kolaṃ kolyāḥ phalaṃ badaram}).
-According to Nadkarni, the fruit of the wild variety is very acid and astringent.
-It is believed to purify the blood and assist digestion.
-The bark is astringent and a simple remedy for diarrhoea.
-Root is useful as a decoction in fever and delirium.
-
-There are references to \emph{kola} being pungent, though this does not seem to indicate sufficiently why \emph{kola} is mentioned separately as an \emph{apathya} food.
-
-\emph{piṇyāka}: Sharma (Ḍalhaṇa and his commentary on drugs: 1982: 69) says, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’
-
-\emph{hiṅgu}: Asafoetida (Nadkarni 1926: pp 360–361): `If long continued even in moderate doses, it gives rise to alliaceous eructations, acrid irritation in the throat, flatulence, diarrhoea and burning in the urine.'
-
-\emph{laśuna} = garlic (Nadkarni 1926: 45).
-\end{philcomm}
-
-\subsection*{1.60}
-\begin{translation}[hp01_060]
-One should know as unwholesome food that has been reheated and is dry.
-[Food that is] too salty, the leftover paste of crushed sesame seeds (tilapiṇḍa), spoiled food, [prohibited] vegetables (?), and pepper (?) are to be omitted.
-\end{translation}
-
-\begin{philcomm}[hp01_060]
-We have not found any conclusive evidence for the meaning of \emph{tilapiṇḍa}. Brahmānanda glosses it as \emph{piṇyāka} (on which see the notes for the previous verse).
-
-Not sure how to take \emph{kadaśanaśākotkaṭaṃ}.
-Brahmānanda understands it as a dvandva (i.e., \emph{kadaśana}, \emph{śāka}, \emph{utkaṭa}), where \emph{śāka} is prohibited vegetables and \emph{utkaṭa} is pepper.
-
-The meaning of \emph{utkaṭa} is not clear.
-The word \emph{utkaṭā} can mean pepper according to some nighaṇṭus (e.g., Rājanighaṇṭu 5.16 \emph{pārvatī śailajā tāmrā lambabījā tathotkaṭā}).
-But \emph{utkaṭa} can refer to Saccharum Sara and \emph{utkaṭā} also to Laurus Cassia (cinnamon).
-
-Also, \emph{utkaṭa} can be an adjective that means ‘abounding in’ at the end of a compound.
-So could \emph{kadaśanaśākotkaṭaṃ} have been intended as an adjectival tatpuruṣa along the lines of ‘[food] full of spoiled vegetables’?
-\end{philcomm}
-
-\subsection*{1.61}
-\begin{translation}[hp01_061]
-Similarly a saying by Goraksa:
-One should avoid a liking for bad people, frequenting fire, women and roads, early morning bathing, fasting and the like, as well as such things as harming the body.
-\end{translation}
-
-\begin{philcomm}[hp01_061]
-The vulgate has a parallel from the Amaraugha added. Also, the vulgate reads \emph{kāyakleśavidhiṃ tathā}.
-\end{philcomm}
-
-\subsection*{1.62}
-\begin{translation}[hp01_062]
-Pure food with wheat, rice, śāli rice, barley, sixty-day śāli rice, milk, ghee, candied sugar, unclarified fermented butter, ground sugar and honey.
-Dried ginger, fruit of the snake gourd and five vegetables, mung beans and so on, and rain water are wholesome for the best of sages.
-\end{translation}
-
-\begin{philcomm}[hp01_062]
-\emph{khaṇḍa} -- candied sugar (Meul 507, different types of sugar).
-
-\emph{navanīta} (MW fresh butter), Mchugh (2021) unclarified fermented butter.
-
-\emph{sitā} -- ground sugar (Meul 507, different types of sugar) ``sitā is very white and looks like gravel"
-
-\emph{madhu} -- honey.
-
-\emph{paṭola} can refer to at least two different gourds. See Meul. p. 569 for a long list of possibilities, including TRICHOSANTHES DIOICA ROXB. (`pointed gourd'), T. CUCUMERINA LINN (snake gourd).
-
-Nadkarni has two entries on \emph{paṭola}:
-\begin{enumerate}
-\item snake gourd (Nadkarni p.
-863) is common in Bengal and cultivated in Northern India and Punjab.
-The unripe fruit of this climbing plant is generally used as a culinary vegetable and is very wholesome, specially suited for the convalescent.
-
-\item smooth luffa (Nadkarni p.
-518) is a hairy climbing herb extensively cultivated in several parts of India.
-The fruit is edible.
-Medicinally it is described as `cool, costive, demulcent, producive of loss of appetite and excitive of wind, bile and phlegm")
-\end{enumerate}
-
-Wikipedia : smooth luffa = Luffa aegyptiaca (sponge gourd)\\
-Sharma (Syn. Kulaka. Well known (Trichosanthas dioica Roxb.)
-
-Brahmānanda glosses it as \emph{kośātakī} (Meul p. 586 LUFFA ACUTANGULA ROXB), which suggests he thought it was some sort of luffa.
-
-Brahmānanda also mentions the vernacular term \emph{paravara} for \emph{paṭola}, which the Lonavla ed.
-states is a kind of cucumber.
-However, Paras remarked that \emph{paravara} is more like a gourd (hard shell, etc.).
-
-On \emph{pañcaśāka}, see GhS  
-\startverse
-bālaśākaṃ kālaśākaṃ tathā paṭolapatrakam |\\
-pañcaśākaṃ praśaṃsīyād vāstūkaṃ hilamocikāṃ || 5.20 ||
-\endverse
-
-HTK 4.26
-\startverse
-pañcaśākastu –\\
-kṣīraparṇī ca jīvantī matsyākṣī ca punarnavā \\
-meghanādaś ceti budhaiḥ pañcaśākaḥ prakīrtitaḥ || iti || 26
-\endverse
-
-Jyotsnā
-\startverse
-sarvaśākam acākṣuṣyaṃ cākṣuṣyaṃ śākapañcakam |\\
-jīvantī-vāstu-matsyākṣī-meghanāda- punarnavāḥ || iti ||
-\endverse
-
-It is not entirely clear how we should understand \emph{divya}.
-Brahmānanda glosses it a \emph{nirdoṣa} and takes it with \emph{udaka}.
-But could it refer more specifically to \emph{gaṅgāmbu} (as suggested by Paras) or rain water?
-MW has \emph{divyodaka} n.
-`divine water' i.e.
-rainwater L.
-
-The term \emph{divyodaka} appears in Āyurvedic works (but we’re yet to find a gloss in a commentary). E.g.,
-Aṣṭāṅgahṛdaya 8.42–43
-\startverse
-śīlayec chāligodhūmayavaṣaṣṭikajāṅgalam |\\
-suniṣaṇṇakajīvantībālamūlakavāstukam || 42 ||\\
-pathyāmalakamṛdvīkāpaṭolīmudgaśarkarāḥ |\\
-ghṛtadivyodakakṣīrakṣaudradāḍimasaindhavam || 43 ||
-\endverse
-
-SriKanta Murty translates \emph{divyodaka} as ‘divyodaka (rain water or pure water)'.
-
-The Rājanighaṇṭu says rainwater:
-\startverse
-divyodakaṃ kharāri syād ākāśasalilaṃ tathā /\\
-vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam // \textup{Rajni 14.4}
-\endverse
-
-Kharāri? Maybe \emph{khavāri} was intended.
-
-Vācaspatyam:
-\textbf{divyodaka} na° karma°.
-1 antarīkṣabhave jale divyaśabde bhāva° pra° vākye tadbhedādi dṛśyam.
-ambuśabde vivṛtiḥ .
-\end{philcomm}
-
-\subsection*{1.63}
-\begin{translation}[hp01_063]
-The yogi should eat food that is pure (\emph{mṛṣṭa}, or delicious \emph{miṣṭa}), unctuous, and containing diary. It should nourish the bodily constituents, satiate the mind and be fit [for yoga].
-\end{translation}
-
-\begin{philcomm}[hp01_063]
-The variants of 1.63a all seem possible: \emph{mṛṣṭaṃ}, \emph{miṣṭaṃ} and \emph{iṣṭaṃ}. Maybe the last is made redundant by \emph{mano 'bhilaṣitaṃ}.
-\end{philcomm}
-
-\subsection*{1.64}
-\begin{translation}[hp01_064]
-Whether young, old, very old, sick or even weak, the diligent yogi succeeds in all yogas through practice. 
-\end{translation}
-
-\begin{philcomm}[hp01_064]
-Note the different reading in 164d for V1: \emph{sarvaṃ yogī yatendriyaḥ}. Sarvaṃ is not easy to construe, and the testimony of the DYŚ suggests that \emph{yoge sarvo ’py atandritaḥ} was original.
-\end{philcomm}
-
-\subsection*{1.65}
-\begin{translation}[hp01_065]
-The postures, various breath retentions and techniques, beginning with seals, are all to be done in the practice of Haṭha until the goal of Rājayoga [is attained].
-\end{translation}
-
-\begin{philcomm}[hp01_065]
-\emph{sarvaṇy api} is better than \emph{sarvo pi} ca because it refers to all the practices mentioned in the first hemistich.
-
-Reading of Brahmānanda is different for the third pāda: \emph{divyāni karaṇāni}.
-\end{philcomm}
-
-\end{ekdosis}
-\end{document}

### 2023-04-11 latex/HP1_TranslComm.tex
new base text

@@
 \input{Preamble-Print}
-\makepagestyle{HPed}
-\makeoddhead{HPed}{\small\texteng{HP Transl. \& Comm.}}{}{\small\texteng{\today}}
-\makeevenhead{HPed}{\small\texteng{HP Transl. \& Comm.}}{}{\small\texteng{\today}}
-\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
-\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}
-
-% additions/changes 2022-03-12 mm; 2022-6-1 MD:
-\SetTEIxmlExport{autopar=false}
-\NewDocumentEnvironment{translation}{O{}}{\textcolor{blue}{\textbf{Transl.:}}}{}
-\NewDocumentEnvironment{philcomm}{O{}}{
-	\textcolor{blue}{\textbf{Comm.:}}}{}
-\NewDocumentEnvironment{sources}{O{}}{
-	\textcolor{blue}{\textbf{Testimonia:}}}{}
-
-\setlength{\parskip}{0.3em}
-\setlength\parindent{0pt}
-\def\startverse{\begin{ekdverse}\begin{itshape}}
-\def\endverse{\end{itshape}\end{ekdverse}}
-\setvnum{}
-
+\input{CommStyle}
 \begin{document}
@@
 \begin{translation}[hp01_001]
-Homage to the glorious Ādinātha by whom the knowledge of Haṭhayoga was taught. It shines forth like a ladder for one desirous of climbing to the lofty terrace of the royal palace.
-\end{translation}
-
-\begin{sources}[hp01_001]
-Yogasārasaṃgraha (line 3090), Gheraṇḍasaṃhitā
-\end{sources}
-
-%\begin{philcomm}[hp01_001]
-%\end{philcomm}
+Homage to the glorious Ādinātha by whom the system of Haṭhayoga was taught. It shines forth like a ladder for one desirous of climbing to the lofty terrace of the royal palace.
+\end{translation}
+
+\begin{testimonia}[hp01_001]
+Cf. Yogasārasaṃgraha, p. 54.
+\startverse
+sadādināthāya namo'stu tubhyaṃ yenopadiṣṭā haṭhayogavidyā | \\
+virājate pronnatarājayogam āroḍhum icchoradhirohiṇīva ||
+\endverse
+
+Gheraṇḍasaṃhitā 1.1
+\startverse
+ādīśvarāya praṇamāmi tasmai yenopadiṣṭā haṭhayogavidyā | \\
+virājate pronnatarājayogam āroḍhum icchor adhirohiṇīva ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_001]
+The reading \emph{rājasaudha} is preferable for poetical reasons, as explained in Hanneder 2020, p.
+128–130.
+Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison
+(\emph{iva}), (2) the object compared (\emph{upameya}), i.e., the \emph{haṭhayogavidyā} leading to Rājayoga (\emph{rājayogaprāpikā}), (3) the property compared, that is, effortlessness (\emph{anāyāsena}) and, most importantly, (4) an image to compare with, which in this case is \emph{rājasaudha}.
+This idea is confirmed by Brahmānanda, when he sums up and completes the poetical details as: \emph{yathā pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudhaprāpikā bhavati evaṃ haṭhadīpikāpi pronnatarājayogam āroḍhum icchor anāyāsena rājayogaprāpikā bhavatīti upamālaṅkāraḥ}.
+
+The Sanskrit poeticians explain that when some elements of a comparison remains unexpressed we get an incomplete (\emph{lupta}) comparison.
+Often words like \emph{iva} or \emph{yathā} are missing or the common property, but if we read \emph{rājayoga} we lose the \emph{upamāna} (i.e., \emph{rājasaudha}).
+One part of the comparison should not be missing.
+But from the perspective of \emph{alaṃkāraśāstra} the verse has a problem that has surely caused the dilemma: it either leaves the \emph{upameya} or the \emph{upamāna} incomplete.
+Brahmānanda has mentioned both \emph{yathā adhirohiṇī saudhaprāpikā bhavati evaṃ haṭhadīpikā rājayogaprāpikā bhavati}, but had to supply \emph{rājasaudha}, while his text reads \emph{rājayoga}.
+But to have the \emph{upameya} in the text is odd.
+Even Brahmānanda could only know of the image (\emph{rājasaudha}) from the alternative reading he did not accept.
+However, it seems likely that the author would have included the \emph{upamāna} in the text, as in our critical text, so that the reader would know that the upper terrace of the palace is an image for \emph{rājayoga}.
+It is further likely that the substitution of \emph{rājayoga} for \emph{rājasaudha} was the result of the tendency to insert the word \emph{yoga} in the opening verses of the text wherever possible, even where it does not fit, as can be seen in 1.2d (\emph{haṭhayogopadiśyate}) and 1.3b (\emph{rājayogam ajānatām}).
+In the latter case the poetical image has also been lost.
+
+The metre of 1.1 is Indravajrā
+\end{philcomm}
 
@@
 \begin{translation}[hp01_002]
-Having bowed to the glorious guru, the Lord, the yogi Svātmārāma has taught the doctrine of Haṭhayoga solely for [attaining] Rājayoga.
-\end{translation}
-
-%\begin{philcomm}[hp01_002]
-%\end{philcomm}
+Having bowed to the glorious guru, the Lord, the yogi Svātmārāma has taught the system of Haṭhayoga solely for [attaining] Rājayoga.
+\end{translation}
+
+\begin{philcomm}[hp01_002]
+The reading \emph{yogopadiśyate} (J7,J10,N17,W4, etc.) is only possible if one accepts that double \emph{sandhi} is a feature of the style of composition, which it is not.
+Moreover, this reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_003]
-For those who are ignorant of Rājayoga because of confusion in the darkness of many opinions the compassionate Svātmārāma composes the Lamp on Haṭha.
-\end{translation}
-
-%\begin{philcomm}[hp01_003]          
-%\end{philcomm}
+The compassionate Svātmārāma holds the Lamp on Haṭha for those who are ignorant of the royal path because of wandering in the darkness of many opinions.
+\end{translation}
+
+\begin{testimonia}[hp01_003]
+Haṭharatnāvalī 1.4
+\startverse
+bhrāntyā bahumatadhvānte rājayogam ajānatām |\\
+kevalaṃ rājayogāya haṭhavidyopadiśyate || 
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_003] 
+Most witnesses have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b.
+While this reading is well attested by the manuscripts transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note that P,T,t1 of the critical edition [Gharote 2009: 5 n.
+2] have \emph{rājamārgam ajānataḥ}), the reading of a royal path (\emph{rājamārgam}) is consistent with the metaphor of people wandering from a path in the darkness and, therefore, most probably authorial.
+
+In 1.1d both \emph{kṛpākaraḥ} and \emph{kṣamākaraḥ} are attested.
+The former is preserved by the V1 and group 2 whereas the latter by some manuscripts of the delta group.
+Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākara} makes better sense.
+As Brahmānanda notes, this compound can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}).
+In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} may have arisen as a mistake for \emph{kṛ}.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_004]
-In fact, Matsyendra, Gorakṣa and other [siddhas] knew the doctrine of Haṭha, and the yogi Svātmārāma knows it owing to their favour.
-\end{translation}
-
-%\begin{philcomm}[hp01_004]          
-%\end{philcomm}
+In fact, Matsyendra, Gorakṣa and other [perfected yogis] knew the system of Haṭha, and the yogi Svātmārāma knows it owing to their favour.
+\end{translation}
+
+\begin{testimonia}[hp01_004]
+Post-15th Century\\
+Haṭharatnāvali 1.3
+\startverse
+haṭhavidyāṃ hi gorakṣamatsyendrādyā vijānate |\\
+ātmārāmo 'pi jānīte śrīnivāsas tathā svayam || 
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_004]   
+The word \emph{athavā} (‘or’) is difficult to construe here.
+Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it.
+The variant \emph{mahāyogī} in group 4c (G5, J4, J11, Ko, P15) avoids this difficulty, but is probably an attempt to remove the difficulty of understanding \emph{athavā}.
+One could emend to \emph{tathā} but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.
+\end{philcomm}
 
@@
 
-%\begin{philcomm}[hp01_005]          
-%\end{philcomm}
+\begin{testimonia}[hp01_005]
+Haṭharatnāvalī 1.80
+\startverse
+śrīādināthamatsyendraśābarānandabhairavāḥ |\\
+śāraṅgīmīnagorakṣavirūpākṣabileśayāḥ || 
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_005]     
+In Śaiva texts which predate the haṭha corpus, Mīnanātha and Matsyendra are one and the same, but they are differentiated in later Tibetan and Indian lists of siddhas (Mallinson 2019:273 n.35).   
+\end{philcomm}
 
@@
 \begin{translation}[hp01_006]
-Manthānabhairava, Siddhabuddha, and Kanthaḍi, Pauraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
-\end{translation}
-
-%\begin{philcomm}[hp01_006]          
-%\end{philcomm}
+Manthānabhairava, Siddhabuddha, and Kanthaḍi, Koraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
+\end{translation}
+
+\begin{testimonia}[hp01_006]
+Haṭharatnāvalī 1.81
+\startverse
+manthānabhairavo yogī siddhabuddhaś ca kandalī |\\
+korandakaḥ surānandaḥ siddhipādaś ca carpaṭī ||
+\endverse
+
+Caturbhuja Misra's Mugdhāvabodhinī (1.7.8) on the Rasahṛdayatantra
+\startverse
+manthānabhairavo yogī siddhabuddhaś ca kanthaḍī |\\
+koraṇṭakaḥ surānandaḥ siddhapādaś ca carpaṭī ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_006]        
+Many manuscripts preserve the reading \emph{pauraṇṭaka}.
+We have accepted \emph{koraṇṭaka} because this name is associated with the \emph{Haṭhābhyāsapaddhati}.
+This reading is well attested by various manuscripts of the \emph{Haṭhapradīpikā}, including those of group 2, as well as those of the \emph{Haṭharatnāvalī} (Gharote 2009: 35 n.
+3), which also attest \emph{gonandaka}.
+Also, \emph{Goraṇṭakuḍu} is the name of a disciple of Gorakṣanātha in the \emph{Navanāthacaritramu} (Jones 2017:194 n.3).
+The compound \emph{siddhapāda} could be a respectful affix.
+However, it seems unlikely here because it would cross the \emph{pāda} break.
+\end{philcomm}
 
@@
 
-%\begin{philcomm}[hp01_007]          
-%\end{philcomm}
+\begin{testimonia}[hp01_007]
+Haṭharatnāvalī 1.82
+\startverse
+karoṭiḥ pūjyapādaś ca nityanātho nirañjanaḥ |\\
+kapālī bindunāthaś ca kākacaṇḍīśvarāhvayaḥ || 
+\endverse
+
+Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
+​​\startverse
+kaṇerī pūjyapādaśca nityanātho nirañjanaḥ |\\
+kapālī bindunāthaśca kākacaṇḍīśvaro gajaḥ |
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_007]   
+It is possible that \emph{pūjyapāda} could be a respectful affix to the name Kāṇerī.
+The variant \emph{dhvaninātha} may have resulted from a transposition of the first two syllables of \emph{nityanātha}.
+\end{philcomm}
 
@@
 
-%\begin{philcomm}[hp01_008]          
-%\end{philcomm}
+\begin{testimonia}[hp01_008]
+Haṭharatnāvalī 1.83
+\startverse
+allamaḥ prabhudevaś ca naiṭacūṭiś ca ṭiṇṭiṇiḥ |\\
+bhālukir nāgabodhaś ca khaṇḍakāpālikas tathā || 
+\endverse
+
+Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
+\startverse
+āllamaḥ prabhudevaś ca ghoḍācolī ca ṭhiṇṭhinī |\\
+bhālukir nāgadevaś ca khaṇḍī kāpālikas tathā ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_008]  
+The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literaure) is largely transmitted as allamaḥ prabhudevaḥ, as though it were two names, although some manuscripts of the \emph{Haṭhapradīpikā} have \emph{allamaprabhudevaś ca} (i.e., Tue,V3,V8,V13,V16,V22,Vu,N24,N26) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n.
+8).
+
+The names Nāgabodha, Nāgabodhi, Naradeva, Nāgadeva all seem possible in 1.8c.
+
+Many witnesses have khaṇḍa and kāpālika as separate names.
+However, khaṇḍakāpālika is well attested.
+Examples include \emph{Kathāsaritsāgara} 121.5 ff.
+(check), \emph{Bṛhatkathāmañjarī} 10.45 (check) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}).
+It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a `part-time' Kāpālika.
+Alternatively, it could simply refer to one who used a broken skull as a bowl.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_009]
-These great siddhas and others have destroyed the rod of death through the power of Haṭhayoga and wander in the universe.
-\end{translation}
-
-%\begin{philcomm}[hp01_009]          
-%\end{philcomm}
+Having destroyed the rod of death through the power of Haṭhayoga, these great perfected yogis and others wander in the world.
+\end{translation}
+
+\begin{testimonia}[hp01_009]
+Haṭharatnāvalī 1.84
+\startverse
+ityādayo mahāsiddhāḥ haṭhayogaprasādataḥ |\\
+khaṇḍayitvā kāladaṇḍaṃ brahmāṇḍe vicaranti te ||
+\endverse
+
+Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
+\startverse
+ityādayo mahāsiddhā rasabhogaprasādataḥ |\\
+khaṇḍayitvā kāladaṇḍaṃ trilokyāṃ vicaranti te |
+\endverse
+
+Haṭhatattvakaumudī 17.24
+\startverse
+ūrdhvaṃretaḥprabhāvena sanakādyā maharṣayaḥ |\\
+khaṇḍayitvā kāladaṇḍaṃ yathecchaṃ viharanti te || 24 ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_009]        
+The reference to \emph{brahmāṇḍa} (‘the world’) implies liberation-in-life (\emph{jīvanmukti}) and physical immortality.  
+\end{philcomm}
 
@@
 \begin{translation}[hp01_010]
-Haṭha is considered a refuge for those who are burnt by the pains of transmigration. Haṭha is the foundational tortoise for the worlds of all yogas.
-\end{translation}
-
-%\begin{philcomm}[hp01_010]          
-%\end{philcomm}
+Haṭha is considered a refuge for those who are burnt by the scorching torment of transmigration. Haṭha is the tortoise that supports the worlds of all yogas.
+\end{translation}
+
+\begin{testimonia}[hp01_010]
+Yogasārasaṃgraha (YSS), p.53.
+\startverse
+saṃsāratāpataptānāṃ samāśrayahaṭho haṭhaḥ |\\
+aśeṣayogajagatām ādhārakamaṭho haṭhaḥ ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_010] 
+The compound \emph{saṃsāratāpa°} is well attested and found elsewhere (e.g., \emph{Viṣṇupurāṇa} 6.7.62, \emph{Agnipurāṇa} 371.1, \emph{Haṭhatattvakaumudī} 38.92, emph{Haṭhābhyāsapaddhati} ms.
+46/440, f.
+1v).
+The reading of \emph{samāśrayo} in V1 is metrically faulty.
+The accepted reading \emph{āśrayo [']yam} is better metically and well attested.
+N23 preserves the reading \emph{samāśrayamaṭho haṭhaḥ} (similar to the \emph{Yogasārasaṃgraha}), which is similar to the fourth \emph{pāda}.
+Though this reading is possible, it seems more likely that it came about through dittography.
+The word \emph{°jagatāṃ} is probably authorial because it makes good sense with  \emph{ādhārakamaṭha} in light of the cosmological notion that the tortoise supports all the worlds.
+However, this reading may not have been understood by some and was changed in the vulgate and other witnesses to \emph{°yuktānāṃ} instead.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_011]
-The doctrine of Haṭha should be kept very secret by those yogis who are desiring success. When it is secret it becomes potent. However, when it has been revealed, it becomes impotent.
+The doctrine of Haṭha should be kept completely secret by those yogis who are desiring success. When it is secret it becomes potent. However, when it has been revealed, it becomes impotent.
 \end{translation}
@@
 \begin{sources}[hp01_011]
-ŚS 5.254; BKhP 10v4, YCM
-\end{sources}
-
-%\begin{philcomm}[hp01_011]          
-%\end{philcomm}
+Śivasaṃhitā
+\startverse
+haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\% (ab missing in some ŚS witnesses)
+bhaved vīryavatī guptā nirvīryā ca prakāśitā || 5.254
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_011]
+Yogacintāmaṇi
+\startverse
+tathā haṭhapradīpikāyām—\\
+haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\
+bhaved vīryavatī guptā nirvīryā tu prakāśiteti ||
+\endverse
+
+BKhP 10v4
+\end{testimonia}
+
+\begin{philcomm}[hp01_011]        
+Either the singular or plural of yogin could be read here. The singular is well attested among the testimonia. However, the weight of the manuscript evidence favours the plural.  
+\end{philcomm}
 
@@
 \begin{translation}[hp01_012]
-In well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live remotely in a small hut.
+In a well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live remotely in a small hut.
 \end{translation}
@@
 \begin{sources}[hp01_012]
-GŚ 32cd; BKhP 107v1, YCM
-\end{sources}
-
-%\begin{philcomm}[hp01_012]          
-%\end{philcomm}
+References to maṭhikā here.
+\end{sources}
+
+\begin{testimonia}[hp01_012]
+Haṭharatnāvalī 1.66
+\startverse
+surāṣṭre dhārmike deśe subhikṣe nirupadrave |\\
+ekāntamaṭhikāmadhye sthātavyaṃ haṭhayoginā ||
+\endverse
+
+Yogacintāmaṇi
+\startverse
+haṭhapradīpikāyām—\\
+surājye dhārmike deśe subhikṣe nirupadrave |\\
+ekānte maṭhikāmadhye sthātavyaṃ haṭhayoginā ||
+\endverse
+
+BKhP 107v1
+\end{testimonia}
+
+\begin{philcomm}[hp01_012]        
+In the \emph{Jyotsnā} and printed editions of the \emph{Haṭhapradīpikā}, including one by Digambara and Kokaje (1970: 6), this verse has the additional hemistich, \emph{dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite}.
+This hemistich derives from the \emph{Gorakṣaśataka} (32cd), which has \emph{°paryante} instead of \emph{°paryantaṃ}.
+It stipulates that the hut should be built in a place measuring up to a bow length and free from rocks, fire and water.
+None of the early manuscripts have this hemistich which suggests that it was added at a later time.
+Nonetheless, it appears in over a dozen manuscripts that were consulted for this edition.
+These manuscripts are not close to an early hyparchetype of the text.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_013]
-It has a small door, without cracks and holes, its length is not too high or low, thickly smeared with cow dung in the proper way, clean, free from everything that annoys, adorned with a pavilion, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
+It has a small door and is without cracks, holes and potsherds.
+It extends not too high or low, and is thickly smeared with cow dung in the proper way.
+It is clean, free from everything that annoys, furnished on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
 \end{translation}
@@
 \begin{sources}[hp01_013]
-BKhP 107v3, YCM
- \end{sources}
+Cf. Dattātreyayogaśāstra
+\startverse
+suśobhanaṃ maṭhaṃ kuryāt sūkṣmadvāraṃ tu nirvraṇaṃ || 54 ||\\
+suṣṭhu liptaṃ gomayena sudhayā vā prayatnataḥ |\\
+matkuṇair maśakair bhūtair varjitaṃ ca prayatnataḥ || 55 ||\\
+dine dine susammṛṣṭaṃ sammārjanyā hy atandritaḥ |\\
+vāsitaṃ ca sugandhena dhūpitaṃ guggulādibhiḥ || 56 ||\\
+malamūtrādibhir vargair aṣṭādaśabhir eva ca |\\
+varjitaṃ dvārasampannaṃ vastrāvaraṇam eva vā || 57 ||
+\endverse
+
+\end{sources}
+
+\begin{testimonia}[hp01_013]
+Suśruta 6.17.67:
+\emph{gṛhe nirābādhe}
+
+Yogacintāmaṇi
+\startverse
+alpadvāram arandhragartaghaṭitaṃ nāpy uccanīcāyitam |\\
+samyaggomayasāndraliptavimalaṃ niḥśeṣajantūjjhitam |\\
+bāhye maṇḍapakūpavediracitaṃ prākārasaṃveṣṭitam |\\
+proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||
+\endverse
+
+Haṭharatnāvalī 1.67
+\startverse
+alpadvāram arandhragartapiṭharaṃ nātyuccanīcāyataṃ\\
+samyaggomayasāndraliptavimalaṃ niḥśeṣabādhojjhitaṃ |\\
+bāhye maṇḍapavedikūparuciraṃ prākārasaṃveṣṭitam\\
+proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||
+\endverse
+
+BKhP 107v3
+\end{testimonia}
 
 \begin{philcomm}[hp01_013]  
-\emph{piṭharaṃ} good
+The syntax of this verse is rather problematic.
+One would expect the features of the hut, which are listed in the first three pādas of the verse, to be in the nominative case, with the words \emph{idaṃ lakṣaṇaṃ} in the fourth pāda referring back to them.
+However, the compounds in the first three pādas appear to qualify \emph{lakṣaṇa} as though they were adjectives, and this appears to have been the way the verse was composed.
+
+The manuscripts preserve many different readings at the end of the compound beginning with \emph{arandhragarta°}.
+The reading \emph{°piṭharaṃ} has been suggested by Dominic Goodall, who understands it as potsherds (ask ref from Dominic?).
+The sense is that the hut should be free of rubbish, such as potsherds.
+One would expect a word for a defect in a hut that is similar to, but not the same as, cracks (\emph{randhra}) and holes (\emph{garta}).
+For this reason, the reading \emph{°vivaraṃ} looks like a patch, as its meaning does not add anything to \emph{°randhragarta°}.
+The reading \emph{viṭapaṃ} is well attested among the witnesses of the \emph{Haṭhapradīpikā} and \emph{Haṭharatnāvalī}.
+However, it is difficult to construe in this context unless it was intended to refer to creepers or branches that might invade or encroach upon the hut.
+
+Nearly all the early mss.
+have \emph{°bādhojjitaṃ}, whereas the Yogacintāmaṇi and V19 have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’).
+However, \emph{°bādhojjitaṃ} may have been original because it is so well attested and a hut is described similarly in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
 \end{philcomm}
@@
 \begin{translation}[hp01_014]
-Locating oneself in a hut of such a kind, free from all worry, one should constantly practise yoga in the way taught by one's guru.
-\end{translation}
+Locating oneself in a hut of such a kind, free from all worry, [the yogi] should practise only yoga, in the way taught by his guru.
+\end{translation}
+
+\begin{sources}[hp01_014]
+Cf. Amanaska 2.15
+\startverse
+evaṃvidhaṃ guruṃ labdhvā sarvacintāvivarjitaḥ\\
+sthitvā manohare deśe yogam eva samabhyaset
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_014]
+Yogacintāmaṇi
+\startverse
+evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\
+gurūpadiṣṭamārgeṇa yogam eva sadābhyased iti ||
+\endverse
+
+Haṭharatnāvalī 1.68
+\startverse
+evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\
+gurūpadiṣṭamārgeṇa yogam eva sadābhyaset ||
+\endverse
+\end{testimonia}
 
@@
 \begin{translation}[hp01_015]
-Overeating, exertion, chatter (gossiping/bickering?), sticking to rules, associating with people, inconstancy: through [these] six, yoga will be abandoned.
-\end{translation}
+Overeating, exertion, idle chatter, not sticking to rules, socialising, sensuality: through [these] six, yoga is lost.
+\end{translation}
+
+\begin{testimonia}[hp01_015]
+Yogacintāmaṇi
+\startverse
+atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\
+janasaṅgaś ca laulyaṃ ca ṣaḍbhir yogaḥ praṇaśyati ||
+\endverse
+
+Haṭharatnāvalī 1.77
+\startverse
+atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\
+janasaṅgaṃ ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||
+\endverse
+
+Yuktabhavadeva 4.25 (attributed to the śivayoga)
+\startverse
+atyāhāraḥ prayāsaśca prajalpo niyamāgrahaḥ |\\
+janasaṃgaś ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||
+\endverse
+
+Jyotsnā
+\startverse
+śītodakena prātaḥsnānanaktabhojanaphalāhārādirūpaniyamasya grahaṇaṃ niyamagrahaḥ |
+\endverse
+
+Yogaprakāśikā
+\startverse
+niyamāgrahaḥ vakṣyamāṇaniyamāparipālanaṃ
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_015]        
-Impossible to decide on meaning of \emph{niyamāgraha}, avagraha invisible, Jyotsnā takes it as over-insistence (as if \emph{āgraha} was implied) as he relates it to extreme ascetic practice; will be abandoned if \emph{prahāsyate} is adopted.
+It is impossible to be certain about the meaning of \emph{niyamagraha}, as the manuscripts do not indicate whether an \emph{avagraha} (i.e., \emph{prajalpo ’niyamagraha}) was intended.
+Although \emph{yama} and \emph{niyama} are not included in the \emph{Haṭhapradīpikā} as auxiliaries of Haṭhayoga, verse 2.14 implies that \emph{niyama} is necessary at least in the early stages of establishing a practice.
+Furthermore, verse 3.82 suggests that a yogi who does not practice \emph{niyama} might obtain success in yoga through the practice of \textit{vajroli}.
+Ambiguity over the role of \emph{yama} and \emph{niyama} in Haṭhayoga appears to have prompted some to insert verses on ten \emph{yama}s and ten \emph{niyama}s after the next verse (1.16).
+The additional verses may derive from the \emph{Śāradātilakatantra} (25.7–8) or the  \emph{Vasiṣṭhasaṃhitā} (1.38, 1.53).
+In the \emph{Jyotsnā}, Brahmānanda reads \emph{niyamāgraha} and takes it as though \emph{āgraha} was implied, which yields the meaning of ‘over-insistence on rules’, and he relates it to extreme ascetic practice.
 \end{philcomm}
@@
 \begin{translation}[hp01_016]
-From zeal, conviction, resolve, understanding of the truth [of yoga], discernment, abandonment of associating with people: by [these] six, on the other hand, yoga is successful.
-\end{translation}
+However, from zeal, conviction, resolve, contentment, understanding of the truth, and avoiding contact with people, yoga is successful because of [these] six.
+\end{translation}
+
+\begin{sources}[hp01_016]
+Dharmaputrikā 137cd–38ab%Narharināth ed.?
+\startverse
+utsāho niścayo dhairyaṃ santoṣas tattvadarśanam | \\
+kratūnāṃ copasaṃhāraḥ ṣaṭsādhanam iti smṛtam |
+\endverse
+
+Śivadharmottara 10 (W 122r):%JM: what ed or ms is this?
+\startverse
+utsāhān niścayād dhairyāt santoṣāt tattvadarśanāt |\\
+muner janapadatyāgād ṣaḍbhir yogaḥ prasiddhyati |
+\endverse
+
+Jñānārṇava 20.1
+\startverse
+utsāhān niścayād dhairyāt saṃtoṣāt tattvaniścayāt |\\
+muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 1 ||
+\endverse
+
+Yogabindu by Haribhadra
+\startverse
+utsāhān niścayād dhairyāt saṃtopāt tattvadarśanāt |\\
+muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 411 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_016]
+Yogacintāmaṇi
+\startverse
+utsāhāt sāhasād dhairyāt tatvajñānād viniścayāt |\\
+janasaṅgaparityāgāt ṣaḍbhir yogaḥ prasidhyati ||
+\endverse
+
+Haṭharatnāvalī 1.78:
+\startverse
+utsāhān niścayād dhairyāt tattvajñānārthadarśanāt
+[utsāhān niścalād- P,T]\\
+bindusthairyān mitāhārāj janasaṅgavivarjanāt |\\
+nidrātyāgāj jitaśvāsāt pīṭhasthairyād anālasāt\\
+gurvācāryaprasādāc ca ebhir yogas tu sidhyati || 
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_016]    
-\emph{tattva} can sometimes refer to the practices of yoga: e.g., \emph{tritattva} in AS 13.12, 14.2--3.
+For the second \emph{pāda}, the north-Indian manuscripts have something along the lines of \emph{tattvajñānāc ca darśanāt}.
+The problem with this reading is the meaning of \emph{darśana} by itself (i.e., discernment of what?).
+The early sources of this verse, in particular the \emph{Śivadharmottara}, indicate that the second pāda read as \emph{santoṣāt tattvadarśanāt}, which makes much better sense of the word \emph{darśana} (i.e., ‘seeing the truth’).
+It seems that this reading may have been adopted by Svātmārāma because the word \emph{santoṣāt} is present in the south-Indian manuscripts.
+At some point in the transmission, \emph{tattvadarśanāt} became \emph{tattvajñānāt} and \emph{santoṣāt} was dropped.
+The word \emph{darśanāt} was changed to \emph{viniścayāt} in V19 and the \emph{Yogacintāmaṇi}, which indicates that we are not alone in questioning its meaning.
+The problem with \emph{viniścayāt} in \emph{pāda} two is that \emph{niścayāt} is widely attested in \emph{pāda} one.
+It should also be noted that the word \emph{tattva} could have a more specific meaning in the \emph{Haṭhapradīpikā} (4.3) as Svātmārāma states that it is a synonym of \emph{samādhi}.
+In other yoga texts, it can sometimes refer to the practices of yoga (e.g., \emph{tritattva} in \emph{Amṛtasiddhi} 13.12, 14.2--3).
 \end{philcomm}
@@
 \begin{translation}[hp01_017]
-Because it is the first auxiliary of haṭha, āsana is taught first. This (\emph{tad}) āsana brings about steadiness, good health and dexterity.
-\end{translation}
-
-%\begin{philcomm}[hp01_017]          
-%\end{philcomm}
+Because it is the first auxiliary of haṭha, \emph{āsana} is taught first. This type of (\emph{tad}) \emph{āsana} brings about steadiness, good health and physical fitness.
+\end{translation}
+
+\begin{testimonia}[hp01_017]
+Yogacintāmaṇi
+\startverse
+haṭhasya prathamāṅgatvād āsanaṃ pūrvam ucyate |\\
+tat kuryād āsanasthairyam ārogyaṃ cāṅgalāghavaṃ ||
+\endverse
+
+Haṭharatnāvalī 3.5
+\startverse
+haṭhasya prathamāṅgatvād āsanaṃ dārśyate mayā |\\
+tat kuryād āsanaṃ sthairyam ārogyaṃ cāṅgapāṭavam || 
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_017]        
+The reading of \emph{aṅgapāṭavam} is attested among most of the early manuscripts.
+Although this compound rarely appears in other yoga texts, a similar term \emph{śarīrapāṭava} occurs in the \emph{Śivasaṃhitā} (2.35) as one of the benefits bestowed by digestive fire (\emph{vaiśvānarāgni}), which indicates that the word \emph{pāṭava} was used in relation to the body and the benefits of yoga.
+The compound \emph{aṅgapāṭava} seems to imply the optimal functioning of the body.
+However, the alternative reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts, even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}).
+Therefore, it is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by group two (i.e., N23, V19, etc.)
+\end{philcomm}
 
@@
 \begin{translation}[hp01_018]
-I shall now teach some of the postures which have been accepted by sages (\emph{munis}) such as Vasiṣṭha and yogis such as Matsyendra.
-\end{translation}
-
-%\begin{philcomm}[hp01_018]          
-%\end{philcomm}
+I shall now teach some of the postures which have been accepted by both sages (\emph{muni}), such as Vasiṣṭha, and yogis, such as Matsyendra.
+\end{translation}
+
+\begin{testimonia}[hp01_018]
+Yogacintāmaṇi
+\startverse
+haṭhapradīpikāyām—\\
+vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ |\\
+aṅgīkṛtāny āsanāni vakṣyante kānicin mayā ||
+\endverse
+
+Haṭharatnāvalī 3.6
+\startverse
+vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ ||\\
+aṃgīkṛtāny āsanāni lakṣyante kāni cin mayā ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_018]        
+On the historical implications of these two traditions of postural practice in early Haṭhayoga, see Mallinson 2016 (119–122) and Birch 2018 (45–46).
+\end{philcomm}
 
@@
 \begin{translation}[hp01_019]
-Correctly placing the soles of both feet between the knees and thighs [and] sitting up with the body straight: they call that the auspicious [pose].
+Correctly placing the soles of both feet between the knees and thighs [and] sitting up with the body straight: they call that \emph{svastikāsana}.
 \end{translation}
@@
 \begin{sources}[hp01_019]
-VS 1.68 etc
-\end{sources}
-
-%\begin{philcomm}[hp01_019]          
-%\end{philcomm}
+Śāradātilaka 25.12
+\startverse
+jānūrvor antare  samyak  kṛtvā  pādatale ubhe |\\
+ṛjukāyo viśed yogī svastikaṃ tat pracakṣate ||
+\endverse
+
+Vasiṣṭhasaṃhitā 1.68
+\startverse
+jānūrvor antaraṃ samyak kṛtvā pādatale ubhe |\\
+ṛjukāyas tathāsīnaḥ svastikaṃ tat pracakṣate ||
+\endverse
+
+Yogayājñavalkya 3.3
+\startverse
+jānūrvor antare samyak kṛtvā pādatale ubhe\\
+ṛjukāyaḥ sukhāsīnaḥ svastikaṃ tat pracakṣate
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_019]
+Yogacintāmaṇi
+\startverse
+Yājñavalkyaḥ—\\
+jānūrvor antare samyak kṛtvā pādatale ubhe |\\
+ṛjukāyaḥ samāsīnaḥ svastikaṃ tat pracakṣate ||
+\endverse
+
+Haṭharatnāvalī 3.52
+\startverse
+atha svastikāsanam---\\
+jānūrvor antaraṃ samyak kṛtvā padatale ubhe ||\\
+ṛjukāyasamāsīnaḥ svastikaṃ tat pracakṣate || 
+\endverse
+
+\end{testimonia}
+
+\begin{philcomm}[hp01_019]        
+One might wonder how the soles of the feet could be placed between the knees and thighs.
+Brahmānanda explains that the region of the shank near the knee should be understood by the word ‘knee’ in this verse (\emph{atra jānuśabdena jānusaṃnihito jaṅghāpradeśo grāhyaḥ jānusaṃnihito jaṅghāpradeśaḥ}).
+This is consistent with the earliest known description of \emph{svastikāsana} in the \emph{Pātañjalayogaśāstravivaraṇa} (2.46), which states that the big toes of one foot are tucked in between the shank and thigh of the other so it is not seen (\emph{dakṣiṇaṃ pādāṅguṣṭhaṃ savyenorujaṅghena parigṛhyādṛśyaṃ kṛtvā tathā savyaṃ pādāṅguṣṭhuṃ dakṣiṇenorujaṅkgenādṛśyaṃ parihṛhya yathā ca pārṣṇibhyāṃ vṛṣaṇayor apīḍanaṃ tathā yenāste tat svastikam āsanam}).
+For a discussion of \emph{svastikāsana} in the Pātañjalayoga tradition, see Maas 2018: 68–69.
+The descriptions of \emph{svastikāsana} in early Śaiva Tantras do not mention the inserting of the toes between the knees and thighs (see Goodall 2004: 348–350, fn.
+371).
+\end{philcomm}
 
@@
 \begin{translation}[hp01_020]
-One should place one's right heel on the left, at the side of the [lower] back, and on the right the left in the same way. This is the Cow's Mouth [posture]; it is like a cow's mouth.
+[The yogi] should place his right heel on the left side of the [lower] back, and the left [heel] on the right [side], in the same way. This is \emph{gomukhāsana}, which [looks] like a cow's face.
 \end{translation}
@@
 \begin{sources}[hp01_020]
-VS 1.70, YY? 3.5; BKhP 93v7
-\end{sources}
-
-%\begin{philcomm}[hp01_020]          
-%\end{philcomm}
+Cf. Ahirbudhnyasaṃhitā 31.45cd–46
+\startverse
+ubhayor gulphayoḥ kṛtvā pṛṣṭhapārśvāv ubhāv api ||\\
+vyutkrameṇātha pāṇibhyāṃ vinyastābhyāṃ vigṛhya ca |\\
+prṣṭhagābhyāṃ padāṅguṣṭhāv etad gomukham ucyate || 
+\endverse
+
+Vasiṣṭhasaṃhitā 1.70
+\startverse
+savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet |\\
+dakṣiṇe 'pi tathā savyaṃ gomukhaṃ tat pracakṣate ||
+\endverse
+
+Yogayājñavalkya 3.5cd–3.6ab
+\startverse
+savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet\\
+dakṣiṇe ‘pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_020]
+Yogacintāmaṇi
+\startverse
+savye dakṣiṇagulphaṃ tu pṛṣthapārśve niveśayet |\\
+dakṣiṇe'pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā ||
+\endverse
+
+Haṭharatnāvalī 3.53
+\startverse
+atha gomukhāsanam---\\
+savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niyojayet ||\\
+dakṣiṇe 'pi tathā savyaṃ gomukhaṃ gomukhāsanam ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_020]        
+This posture first appears in some Vaiṣṇava \emph{Saṃhitā}s that predate the \emph{Haṭhapradīpikā}, namely, the \emph{Ahirbudhnyasaṃhitā} and the \emph{Vasiṣṭhasaṃhitā}.
+The position of the ankles is the same in all the source texts.
+However, an interesting feature of the description in the \emph{Ahirbudhnyasaṃhitā} is the position of the hands, which are crossed behind the back and hold the big toes.
+The \emph{Haṭhapradīpikā}’s description, which derives from the \emph{Vasiṣṭhasaṃhitā}, does not mention the position of the hands.
+For illustrations of six possible positions of the arms and hands, see Gharote, Jha, Devnath, Sakhalkar 2006: 111–113.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_021]
-Having placed one foot on one thigh and the seated thigh on the other foot. Vīrāsana is taught [thus]. 
-\end{translation}
+By fixing one foot on one thigh and placing the [other] thigh on the other foot, vīrāsana is taught [to arise].
+\end{translation}
+
+\begin{sources}[hp01_021]
+Vasiṣṭhasaṃhitā 1.72
+\startverse
+ekaṃ pādam athaikasmin vinyasyorau ca saṃsthitam |\\
+itarasmiṃs tathaivoruṃ vīrāsanam itīritam ||
+\endverse
+
+Cf. Śāradātilakatantra 25.15cd–16ab
+\startverse
+ekaṃ pādam adhaḥ kṛtvā vinyasyorau tathetaram ||\\
+ṛjukāyo viśed yogī vīrāsanam itīritam |
+\endverse
+
+Yogayājñavalkya 3.8
+\startverse
+ekaṃ pādam athaikasmin vinyasyoruṇi saṃsthitam |\\
+itarasmiṃs tathā coruṃ vīrāsanam udāhṛtam ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_021]
+Yogacintāmaṇi
+\startverse
+ekaṃ pādam athaikasmin vinyasyoruṇi saṃsthitaḥ |\\
+itarasmiṃs tathā coruṃ vīrāsanam udāhṛtam ||
+\endverse
+
+Haṭharatnāvalī 3.54
+\startverse
+atha vīrāsanam---\\
+ekaṃ pādam athaikasmin vinyased ūruṇi sthiram ||\\
+itarasmiṃs tathā coruṃ vīrāsanam īritam || 
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_021]   
+Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by V1, the sources and the testimonia.
+It appears to be verse filler here rather than indicating a temporal sequence of actions.
+Svātmārāma borrowed the verse on \emph{vīrāsana} from the \emph{Vasiṣṭhasaṃhitā}, the redactor of which appears to have adapted its first hemistich from a description of this posture in the \emph{Śāradātilakatantra}.
+This would explain the rather strange syntax of the \emph{Vasiṣṭhasaṃhitā}’s version, in which \emph{adhaḥ kṛtvā} was changed to \emph{athaikasmin}, and \emph{tathetaram} became \emph{ca saṃsthitam}.
+It seems that \emph{saṃsthitaṃ} must be understood with \emph{ūruṃ} in the third \emph{pāda} in the sense of \emph{saṃsthāpya} (i.e., ‘having placed’).
+
+Different versions of \emph{vīrāsana} are found in earlier Tantras, such as the Kiraṇatantra (58.9), Hemacandra’s \emph{Yogaśāstra} and commentaries on the \emph{Pātañjalayogaśāstra}.
+For a discussion of some of these sources, see Maas 2018: 66–68.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_022]
-Having blocked the anus with the crossed ankles, focused [the yogi] remains [thus]. Adepts of yoga know that this is kūrmāsana.
-
-Or: While focused, [the yogi] should bind the anus with crossed ankles. This is kūrmāsana. Experts in yoga know this.
-\end{translation}
+Knowers of yoga know that \emph{kūrmāsana} arises by attentively blocking the anus with turned-out ankles.
+\end{translation}
+
+\begin{sources}[hp01_022]
+Vasiṣṭhasaṃhitā 1.80
+\startverse
+gudaṃ nirudhya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\
+kūrmāsanaṃ bhaved etad iti yogavido viduḥ ||
+\endverse
+
+Cf. Ahirbudhnyasaṃhitā 31.35
+\startverse
+gudaṃ nipīḍya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\
+etat kūrmāsanaṃ proktaṃ yogasiddhikaraṃ param || 
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_022]
+Yogacintāmaṇi
+\startverse
+gudaṃ niyamya gulphābhyaṃ vyutkrameṇa samāhitaḥ |\\
+kūrmāsanaṃ bhaved etad iti yogavido viduḥ ||
+\endverse
+
+Yuktabhavadeva 6.15
+\startverse
+haṭhapradīpikāyām\\
+gudaṃ niyamya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\
+kūrmāsanaṃ bhavedetaditi yogavido viduḥ ||
+\endverse
+
+\end{testimonia}
+
+\begin{philcomm}[hp01_022]   
+In the first \emph{pāda}, the witnesses are split between \emph{nirudhya} (‘having blocked’), \emph{nibadhya} (‘having bound’), \emph{niyamya} (‘having restrained’) and \emph{niṣpīḍya} (‘having pressed’).
+The source, the \emph{Vasiṣṭhasaṃhitā}, supports \emph{nirudhya} whereas the testimonia supports \emph{niyamya}.
+In terms of closing the anus by sitting on the ankles, \emph{nirudhya} makes better sense and has been adopted because it is supported by N23 (an important witness of group 2) and the manuscripts reported in the critical edition of the \emph{Vasiṣṭhasaṃhitā}.
+The word \emph{vyutkrameṇa} appears to describe the position of the ankles.
+Its basic meaning is ‘against the normal direction’, which would suggest that the ankles are turned out or crossed rather than placed together naturally.
+If the yogi is in a kneeling-type position, turning the feet out would bring the ankles together, blocking the perineal area.
+See \emph{Yoga Mīmāṃsā}, vol 8, no.
+2, pp.
+29–30 for a discussion of \emph{vyutkramena} and the position of the ankles in \emph{kūrmāsana}, and vol 8, no.2, Figures 3–6 for photographs of a practitioner performing this \emph{āsana}.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_023]
-Having assumed padmāsana, inserted the hands between the knees and thighs, and placed [the hands] on the ground, [the yogi] remains in the air. This is kurkuṭāsana
-
-Or: [The yogi] should assume padmāsana, insert his hands between the knees and thighs onto the ground and place himself in the air. This is kurkuṭāsana.
-\end{translation}
+[The yogi] adeptly assumes \emph{padmāsana}, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is \emph{kurkuṭāsana}.
+\end{translation}
+
+\begin{sources}[hp01_023]
+Vasiṣṭhasaṃhitā 1.78
+\startverse
+padmāsanaṃ samāsthāya jānūrvor antare karau |\\
+bhūmau niveśya saṃsthāpya vyomasthaṃ kukkuṭāsanam ||\\
+{[}niveśya bhūmau – mss. la, va, śa]
+\endverse
+
+Cf. Ahirbudhnyasaṃhitā 31.38
+\startverse
+kukkuṭāsanam\\
+padmāsanam adhiṣṭhāya jānvantaraviniḥsṛtau |\\
+karau bhūmau niveśyaitad vyomasthaṃ kukkuṭāsanam || 
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_023]
+Yogacintāmaṇi
+\startverse
+padmāsanaṃ tu saṃyojya jānūrvor antare karau | \\
+niveśya bhūmau saṃsthāpya vyomasthaṃ kukkuṭāsanam ||
+\endverse
+
+Haṭharatnāvalī 3.73
+\startverse
+atha kukkuṭāsanam---\\
+padmāsanaṃ susaṃsthāpya jānūrvor antare karau ||\\
+niveśya bhūmau saṃsthāpya vyomasthaḥ kukkuṭāsanam ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_023]
-Kurkuṭa/kurkkuṭa variant in \getsiglum{V1,J10ac,V3} is attested in Pañcatantra (M-W).
+\emph{Kurkuṭa} and \emph{kurkkuṭa} in V1, J10ac, V3 are variant spellings attested in the \emph{Pañcatantra} (M-W).
 \end{philcomm}
@@
 \begin{translation}[hp01_024]
-Bound in kurkuṭāsana, [the yogi] binds the neck with the hands and lies supine like a tortoise. This is uttānakūrmāsana. 
-\end{translation}
+While maintaining kurkuṭāsana, [the yogi] binds the neck with the hands and lies like a tortoise on his back. This is uttānakūrmāsana.
+\end{translation}
+
+\begin{testimonia}[hp01_024]
+Yogacintāmaṇi
+\startverse
+kukkuṭāsanabandhastho dorbhyāṃ saṃbadhya kandharam |\\
+bhavet kūrmavad uttānam etad uttānakūrmakam ||
+\endverse
+
+Haṭharatnāvalī 3.74
+\startverse
+kukkuṭāsanabandhastho dorbhyāṃ sambadhya kandharām ||\\
+śete kūrmavad uttānam etad uttānakūrmakam || 74 ||
+\endverse
+\end{testimonia}
 
@@
 \emph{°bandhasthaḥ} or \emph{°vat kṛtvā}?
-Only V1 has latter, which is simpler.
+Only V1 has the latter, which is simpler.
 Are the others trying to improve it?
@@
 
-Adopt \emph{bandhastho} with note. \emph{°bandhastha} not found in any other texts. \emph{°vat kṛtvā} is not good.
+Adopt \emph{bandhastho} with note. \emph{°bandhastha} not found in any other texts. While \emph{°vat kṛtvā} is possible, it is a singular reading that appears to be unrelated to the other variants.
 \end{philcomm}
@@
 \begin{translation}[hp01_025]
-Having clasped the big toes with hands, [the yogi] should perform a bow drawing action as far as the ear. This is called dhanurāsana.
-\end{translation}
+Clasping the big toes with hands and performing the action of drawing a bow as far as the ear is called dhanurāsana.
+\end{translation}
+
+\begin{testimonia}[hp01_025]
+Yogacintāmaṇi
+\startverse
+pādāṅguṣṭhau ca pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\
+dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam |
+\endverse
+
+Haṭharatnāvalī 3.51
+\startverse
+atha dhanurāsanam---\\
+pādāṅguṣṭhau tu pāṇibhyāṃ gṛhītvā śravaṇāvadhi ||\\
+dhanurākarṣaṇaṃ kṛtvā dhanurāsanam ucyate ||
+\endverse
+
+Cf. Haṭhayogasaṃhitā
+\startverse
+dhanurāsanam |\\
+prasārya pādau bhuvi daṇḍarūpau \\
+karau ca pṛṣṭhe dhṛtapādayugmau |\\
+kṛtvā dhanustulyavivarttitāṅgaṃ \\
+nigadyate vai dhanurāsanaṃ tat || 25 ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_025]
-At a stretch (!), this could mean drawing a crossbow, with both hands taken back to the ears, but probably just means like a normal bow.
-
-Gheraṇḍasaṃhitā:
-\startverse
- prasārya pādau bhuvi daṇḍarūpau |
- karau ca pṛṣṭhaṃ dhṛtapādayugmam |\\
- kṛtvā dhanustulyavivartitāṅgaṃ |
- nigadyate vai dhanurāsanaṃ tat || 2.18 ||
-\endverse
+Comment on kṛtvā, which is the reading of V19, Yogacintāmaṇi and Haṭharatnāvalī.
 \end{philcomm}
@@
 \begin{translation}[hp01_026]
-Having grasped the right foot, which is placed at the base of the left thigh, and the left foot, which is wrapped around the outside of the knee, [the yogi] remains with his body twisted.
+Having grasped the right foot, which is placed at the base of the left thigh, [the yogi’s] left foot is wrapped around the outside of the knee and he remains with his body twisted.
 This āsana was taught by Matsyendranātha.
 \end{translation}
+
+\begin{testimonia}[hp01_026]
+Yogacintāmaṇi
+\startverse
+vāmorumūlārpitadakṣapādaṃ
+jānvor bahir veṣṭitadakṣadoṣṇā |\\
+pragṛhya tiṣṭhet parivartitāṅgaḥ
+śrīmatsyanāthoditam āsanaṃ syāt ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+atha matsyendrāsanam---\\
+vāmorumūlārpitā dakṣapādo jānvor bahir veṣṭitadakṣadoṣṇā ||\\
+pragṛhya tiṣṭhet partivartitāṅgaḥ śrīmatsyanāthoditam āsanaṃ syāt || 3.57 ||
+\endverse
+\end{testimonia}
 
@@
 \begin{translation}[hp01_027]
-Through practice, Matsyendra's seat gives people a torch for the abdomen, a destructive weapon for a range of terrible diseases, the awakening of K and steadiness of the spine.
-\end{translation}
+Matsyendra's seat, by which the digestive fire is kindled, is a destructive weapon for a range of terrible diseases; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
+\end{translation}
+
+\begin{testimonia}[hp01_027]
+Yogacintāmaṇi
+\startverse
+matsyendrapīṭhaṃ jaṭharapravṛddhiṃ\\
+pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\
+abhyāsataḥ kuṇḍalinīprabodhaṃ\\
+daṇḍe sthiratvaṃ pradadāti puṃsām ||
+\endverse
+
+Haṭharatnāvalī 3.58
+\startverse
+matsyendrapīṭhaṃ jaṭharapradīptaṃ \\
+pracaṇḍarugmaṇḍalakhaṇḍanāstram ||\\
+abhyāsataḥ kuṇḍalinīprabodhaṃ \\
+daṇḍasthiratvaṃ ca dadāti puṃsām ||
+\endverse
+
+HTK
+\startverse
+matsyendrapīṭhaṃ jaṭharapracaṇḍa-\\
+ruṅmaṇḍalakhaṇḍanakhaṇḍanāstram |\\
+abhyāsataḥ kuṇḍalinīprabodhaṃ \\
+daṇḍasthiratvaṃ ca dadāti puṃsām || 8 ||
+\endverse
+
+YBD
+\startverse
+matsyendrapīṭhaṃ jaṭharaprabuddhaṃ\\
+pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\
+abhyasataṃ kuṇḍalinīprabodhaṃ\\
+daṇḍasthiratvaṃ ca dadāti puṃsām ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_027]
-Difficult to construe first half of verse: \emph{jaṭhara} compound needs to be in apposition to \emph{matsyendrapīṭhaṃ}.
-KDham ed has variant \emph{jaṭharapradīpaṃ} but \emph{pradīpa} is masculine.
-Perhaps we have to accept \emph{jaṭharapradīptiṃ}, but that crosses stemmatic lines, so \emph{jvalanapradīptiṃ}; or \emph{jaṭharapravṛddhiṃ}.
-\emph{jvalana} is found at 3.66 to refer to the fire in the body but \emph{jaṭhara}, stomach, is not found without \emph{agni} elsewhere.
-
-YCM has \emph{jaṭharapravṛddhiḥ}.
-
-Adopt \emph{jaṭharapradīptiṃ} (\emph{°dīptiṃ} is an emendation), understood as object of \emph{dadāti}.
-
-Pāda d, J10ac and J17 have \emph{candra} for \emph{daṇḍa}, also in YCM and 6-chapter HP.
+Difficult to construe first half of verse:
+We have not found an instance in another verse where \emph{jaṭhara} means \emph{jaṭharāgni}.
+So, the reading \emph{jvalana}° is better and most of the witnesses that have it attest \emph{jvalanapradīptam}.
+This compound has been accepted and understood as a reverse \emph{bahuvrīhi} (i.e., ‘the āsana by which the digestive fire is kindled).
+J10 and J17 have \emph{jaṭharapravṛddha}, which would qualify \emph{pracaṇḍaruṅmaṇḍala}°, but it seems rather to strange to say that a range of terrible diseases have increased in the stomach, rather than arisen in the stomach.
+KDham ed has variant \emph{jaṭharapradīpaṃ} but this is not attested among the oldest dated manuscripts.
+
+Pāda d, J10ac and J17 have \emph{candra} for \emph{daṇḍa}; However,  also in Yogacintāmaṇi and 6-chapter HP.
 \end{philcomm}
@@
 [The yogi] should stretch out both feet on the ground like staffs, hold the ends of both feet with the hands, place the forehead upon the knees and remain thus.
-They call this the back-stretch (paścimatānam).
-\end{translation}
+They call this the back-stretch (\emph{paścimatānam}).
+\end{translation}
+
+\begin{sources}[hp01_028]
+Cf. Śivasaṃhitā
+\startverse
+prasārya caraṇadvandvaṃ parasparasusaṃyutam |\\
+svapāṇibhyāṃ dṛḍhaṃ dhṛtvā jānūpari śiro nyaset || 3.108 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_028]
+Yogacintāmaṇi
+\startverse
+prasārya pādau bhuvi daṇḍarūpau \\
+dvābhyāṃ ca pādadvitayaṃ gṛhītvā |\\
+jānūpari nyastalalāṭadeśo \\
+'bhyased idaṃ paścimatānam āhuḥ ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+atha paścimatānāsanam---\\
+prasārya pādau bhuvi daṇḍarūpau \\
+dorbhyāṃ padāgradvitayaṃ gṛhītvā ||\\
+jānūpari nyastalalāṭadeśo \\
+vased idaṃ paścimatānam āhuḥ || 3.66 ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_028]
-Only V1 has \emph{dorbhyāṃ padāgra}, others have variations on the much inferior \emph{dvābhyāṃ karābhyāṃ}. Vulgate has same as V1, which is a surprise.
-
-V1 reading \emph{dorbhyāṃ pādāgradvitayaṃ} is best, adopt.
-
-V1’s \emph{tāṇabandhaḥ} doesn’t work with \emph{idaṃ}.
+Only V1 has \emph{dorbhyāṃ padāgra}, others have variations on the much inferior \emph{dvābhyāṃ karābhyāṃ}.
+The karābhyaṃ reading was probably introduced to remove dorbhyāṃ, which usually means arms.The reading \emph{tāṇabandhaḥ} in V1 doesn’t work with \emph{idaṃ}.
 \end{philcomm}
@@
 
+\begin{sources}[hp01_029]
+Cf. Śivasaṃhitā
+\startverse
+Āsanāgryam idaṃ proktaṃ jaṭharānaladīpanam |\\
+dehāvasādaharaṇaṃ paścimottānasaṃjñakam || 3.109 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_029]
+Yogacintāmaṇi
+\startverse
+iti paścimatānam āsanāgryaṃ
+pavanaṃ paścimavāhinaṃ karoti |\\
+udayaṃ jaṭharānalasya kuryād
+udare kārśyam arogitāṃ ca puṃsām ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+iti paścimatānam āsanāgryaṃ pavanaṃ paścimavāhinaṃ karoti ||\\
+udayaṃ jaṭharānalasya kuryād udare kārśyam arogatāṃ ca puṃsām || 3.67 ||
+\endverse
+\end{testimonia}
+
 \begin{philcomm}[hp01_029]
-Adopt \emph{arogitāṃ}.
+Adopt \emph{arogitāṃ} (which is well attested), despite the fact that \emph{arogatāṃ} is much more common.
+The use of the word \emph{paścima} to mean the central channel is found in \emph{Yogabīja} 121.
+The  \emph{Yogabīja} (95) also refers to the path of the central channel (\emph{paścimamārga}), and this understanding of \emph{paścima} is found in the \emph{Jyotsnā 1.29}: \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumnāmārgeṇa vahatīti paścimavāhī}.
 \end{philcomm}
@@
 
+\begin{sources}[hp01_030]
+Vimānārcanākalpa 96
+\startverse
+karatale bhūmau saṃsthāpya kūrparau nābhipārśvayor nyasya nataśirāḥ (unnataśirāḥ) pādau ḍaṇḍavad vyomni saṃsthito mayūrāsanam iti |
+\endverse
+
+Pādmasaṃhitā yogapāda 1.21c–22d:
+\startverse
+avaṣṭabhya dharāṃ samyak talābhyāṃ hastayor dvayoḥ ||\\
+kūrparau nābhipārśve ca sthāpayitvā mayūravat |\\
+samunnamya śiraḥpādau mayūrāsanam iṣyate ||
+\endverse
+
+Ahirbudhnyasaṃhitā 31.36–37
+\startverse
+mayūrāsanam\\
+niveśya kūrparau samyaṅ nābhimaṇḍalapārśvayoḥ |\\
+avaṣṭabhya bhuvaṃ pāṇitalābhyāṃ vyomni daṇḍavat || 31--36 ||
+\endverse
+
+Vasiṣṭhasaṃhitā 1.76–77
+\startverse
+avaṣṭabhya dharāṃ samyak talābhyāṃ ca karadvayam |\\
+hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ ||\\
+samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\
+mayūrāsanam etad dhi sarvapāpavināśanam ||
+\endverse
+
+Yogayājñavalkya 3.15–16
+\startverse
+avaṣṭabhya dharāṃ samyak talābhyāṃ tu karadvayoḥ\\
+hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ\\
+samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\
+mayūrāsanam etat tu sarvapāpapraṇāśanam ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_030]
+Yogacintāmaṇi
+\startverse
+dharām avaṣṭabhya punaḥ karābhyāṃ
+tat kūrpare sthāpitanābhipārśvaḥ | \\
+tadāsane daṇḍavad utthitaḥ khe
+mayūram etat pravadanti santaḥ ||
+\endverse
+
+Haṭharatnāvalī 3.42
+\startverse
+atha mayūram\\
+dharām avaṣṭabhya karadvayena tatkūrpare sthāpitanābhipārśvaḥ ||\\
+uccāsano daṇḍavad utthitaḥ khe mayūram etat pravadanti pīṭham ||
+\endverse
+\end{testimonia}
+
 \begin{philcomm}[hp01_030]
-1.30 b tat refers to \emph{karadvaya} (cf. VS).
-\emph{uccāsano} to be adopted; \emph{samunnataśiraḥpādo} is reading of VS. which is derived from Vimānārcanākalpa 96: 
-\emph{karatale bhūmau samsthāpya kūrparau nābhipārśvayor nyasya na(unnata)
-̇ taśira ̄hpādau dandavad vyomni samsthito mayūrāsanam iti}.
+The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77).
+The compound \emph{uccāsano} seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}, which may be derived from earlier Vaiṣṇava sources, such as the \textit{Pādmasaṃhitā}.
+
+1.30b tat refers to \emph{karadvaya} (cf. Vasiṣṭhasaṃhitā).
 \end{philcomm}
@@
 \begin{translation}[hp01_031]
-The glorious mayūra posture gets rid of all diseases of the abdomen such as bloating and overcomes [imbalances of] the doṣas.
+The glorious mayūra posture gets rid of all diseases of the abdomen such as bloating and overcomes humoral imbalances.
 It completely incinerates food which is bad or has been eaten to excess, it generates digestive fire and it digests strong poison.
 \end{translation}
+
+\begin{testimonia}[hp01_031]
+Yogacintāmaṇi
+\startverse
+harati sakalarogān āśu gulmodarādīn\\
+abhibhavati ca doṣān āsanaṃ śrīmayūram |\\
+bahukadaśanabhuktaṃ bhasma kuryād aśeṣam\\
+janayati jaṭharāgniṃ jārayet kālakūṭam ||
+\endverse
+
+Haṭharatnāvalī 3.43
+\startverse
+harati sakalarogān āśu gulmodarādīn\\
+abhibhavati ca doṣān āsanaṃ śrīmayūram ||\\
+bahukadaśanabhuktaṃ bhasma kuryād vicitram\\
+janayati jaṭharāgniṃ jīryate kālakūṭam ||
+\endverse
+\end{testimonia}
 
@@
 \begin{translation}[hp01_032]
-Lying on one’s back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] all [other] āsanas and relaxes the mind.
-\end{translation}
+Lying on one’s back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] any āsana and relaxes the mind.
+\end{translation}
+
+\begin{sources}[hp01_032]
+Cf. Dattātreyayogaśāstra 24cd
+\startverse
+uttānaśavavad bhūmau śayanaṃ coktam uttamam ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_032]
+Yogacintāmaṇi
+\startverse
+uttānaṃ śavavad bhūmau śavāsanam idaṃ smṛtam |\\
+śavāsanaṃ śrāntiharaṃ cittaviśrāntisādhanam ||
+\endverse
+
+Haṭharatnāvalī 3.76
+\startverse
+athāntimaṃ śavāsanam\\
+prasārya hastapādau ca viśrāntyā śayanaṃ tathā ||\\
+sarvāsanaśramaharaṃ śayitaṃ tu śavāsanam || 3.76 ||
+\endverse
+
+Cf. HTK
+\startverse
+śavāsanaṃ hṛtkupitavātagranthivibhedakam |\\
+sarvāsanaśrāntijit hṛtśramaghnaṃ yogi saukhyadam || 12 ||
+\endverse
+
+YBD
+\startverse
+uttānaṃ śavavad bhūmau śayanaṃ tu śavāsanam ||\\
+śavāsanaṃ śrāntiharaṃ cittaviśrāntikārakam ||\\
+iti śavāsanam || 21 ||
+\endverse
+\end{testimonia}
 
@@
 \begin{translation}[hp01_033]
-However, Śiva taught eight-four āsanas. I shall take the four essential ones from them and teach them. 
-\end{translation}
+Śiva taught eighty-four āsanas. I shall take the four best and teach them.
+\end{translation}
+
+\begin{sources}[hp01_033]
+Śivasaṃhitā 3.96
+\startverse
+caturaśīty āsanāni santi nānāvidhāni ca |\\
+tebhyaś catuṣkam ādāya mayoktāni bravīmy aham ||
+\endverse
+
+Cf. Dattātreyayogaśāstra 5
+\startverse
+caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\
+ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||
+\endverse
+
+Cf. Vivekamārtaṇḍa 5
+\startverse
+caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\
+ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_033]
+Yogacintāmaṇi
+\startverse
+haṭhapradīpikāyām—\\
+caturaśīty āsanāni śivena kathitāni vai |\\
+tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+caturaśīty āsanāni śivena kathitāni tu |\\
+tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham || 3.23 ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_033]
+In Pādas ac Svātmārāma may have rewritten Śivasaṃhitā 3.96 to include the information that it was Śiva (\emph{śivena}) who taught the 84 āsanas, whereas in the source Śiva is himself speaking.
+With this
+Svātmārāma confuses the verse, since one would have to understand Pādas cd to be direct speech by Śiva in order to understand \emph{bravīmi} correctly (but cf.
+4.2 which is a verse from the Gorakṣaśataka which includes pravakṣyāmi).
+
+The Śivasaṃhitā follows 3.96 with teachings on siddha, padma, paścimottāna and svastika āsanas, but the HP teaches siddha, padma, siṃha and bhadra.
+
+The word \emph{tu} is often used to introduce a new posture, but in this verse seems to be a verse filler.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_034]
-The adept, lotus, lion and auspicious pose are the best tetrad and, among those, one should always sit in a comfortable adept’s pose. 
-\end{translation}
+The adept, lotus, lion and auspicious pose are the best tetrad and, among those, always sit in the adept’s pose, my dear.
+\end{translation}
+
+\begin{sources}[hp01_034]
+--
+\end{sources}
+
+\begin{testimonia}[hp01_034]
+Yogacintāmaṇi
+\startverse
+siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\
+śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||
+\endverse
+
+Haṭharatnāvalī 3.24
+\startverse
+siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\
+śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_034]
+It is likely that the original version of this verse contained the vocative with the imperative form of the verb.
+There is another instance where Svātmārāma included a verse with the vocative (4.58?) or the subject in the first person (4.2), as though the text were a dialogue.
+In this case, it seems that efforts have been made to write out the vocative and imperative verb.
+However, the \emph{sukhe} and \emph{sukham} is difficult to construe in this context, because the context suggests that the intended meaning was that one should always sit in Siddhāsana (as opposed to the other three), rather than the prescription to always sit in a comfortable Siddhāsana.
+\end{philcomm}
 
@@
 
+\begin{sources}[hp01_035]
+Vivekamārtaṇḍa 7
+\startverse
+yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyase[n] \\
+meḍhre pādam athaikam āsyahṛdaye dhṛtvā samaṃ vigraham |\\
+sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\
+etan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_035]
+Yogacintāmaṇi
+\startverse
+pavanayogasaṃgrahe—\\
+yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\
+meḍhre pādam athaikam ekahṛdaye kṛtvā samaṃ vigraham |\\
+sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyed bhruvor antaraṃ tv\\
+etan mokṣakapāṭabhedanakaraṃ siddhāsanaṃ procyate ||
+\endverse
+
+Haṭharatnāvalī 3.25
+\startverse
+tatra siddhāsanam\\
+yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\
+meḍhre pādam athaikam eva niyataṃ kṛtvā samaṃ vigraham |\\
+sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\
+caitan mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||
+\endverse
+\end{testimonia}
+
 \begin{philcomm}[hp01_035]
@@
 
+\begin{sources}[hp01_036]
+Vasiṣṭhasaṃhitā 1.81
+\startverse
+meḍhrād upari nikṣipya gulphaṃ tathopari |\\
+gulphāntaraṃ vinikṣipya muktāsanam idaṃ smṛtam ||
+\endverse
+
+Yogayājñavalkya 3.15
+\startverse
+meḍhrād upari nikṣipya savyaṃ gulphaṃ tathopari |\\
+gulphāntaraṃ ca nikṣipya muktāsanam idaṃ tu vā ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_036]
+Yogacintāmaṇi
+\startverse
+tathā |\\
+meḍhrād upari vinyasya savyaṃ gulphaṃ tathopari |\\
+gulphāntaraṃ tu vinyasya siddhāsanam idaṃ bhavet ||
+\endverse
+
+Haṭharatnāvalī 3.26
+\startverse
+matāntare tu\\
+meḍhrād upari niḥkṣipya savyaṃ gulphaṃ tathopari |\\
+gulphāntaraṃ ca niḥkṣipya siddhāḥ siddhāsanaṃ viduḥ ||
+\endverse
+\end{testimonia}
+
 \subsection*{1.37}
@@
 \end{translation}
+
+\begin{sources}[hp01_037]
+--
+\end{sources}
+
+\begin{testimonia}[hp01_037]
+Yogacintāmaṇi
+\startverse
+siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\
+śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\
+śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā || 3.24 ||
+\endverse
+
+Cf. Śivayogasāram by Kolani Ganapatideva (date 14th c.)
+\startverse
+siddāsanambunu, gondaru vajrāsanambaniyu | \\
+gondaru muktāsanambaniyu, gondadu  gulbāsanam ||
+\endverse
+
+Cf. A verse by the poet Pingali Surana (active 16th c.)
+\startverse
+kondaru siddāsanamani \\
+kondaru vajrāsanamani koniyādudurī \\
+pondaga dīnini mariyoka \\
+kondaru guptāsamanu kondru mahātmā
+\endverse
+\end{testimonia}
 
@@
 \end{translation}
+
+\begin{sources}[hp01_038]
+Cf. Dattātreyayogaśāstra 33
+\startverse
+laghvāhāras tu teṣv eko mukhyo bhavati nāpare |\\
+ahiṃsā niyameṣv eko mukhyo bhavati nāpare || 33 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_038]
+Yogacintāmaṇi
+\startverse
+niyameṣu mitāharo yathāhịmsā yameṣv iva |\\
+mukhyaṃ sarvāsaneṣv evaṃ siddhāsanaṃ idaṃ viduḥ |
+\endverse
+\end{testimonia}
 
@@
 
+\begin{sources}[hp01_039]
+\end{sources}
+
+\begin{testimonia}[hp01_039]
+Yogacintāmaṇi
+\startverse
+caturaśītipīṭheṣu siddhāsanaṃ samabhyaset | \\
+dvāsaptatisahasreṣu suṣumṇām iva nāḍiṣu ||
+\endverse
+
+YSS
+\startverse
+caturāśītapīṭheṣu siddham eva samabhyaset |\\
+dvisaptatisahasreṣu suṣumnām iva nāḍiṣu ||
+\endverse
+
+Yogacintāmaṇi (in a different passage)
+​​\startverse
+maṇḍalā dṛśyate siddhiḥ kuṇḍalyabhyāsayoginaḥ |\\
+dvisaptatisahasrāṇāṃ nāḍīnāṃ malaśodhanam ||
+\endverse
+
+Cf. Kumbhakapaddhati (effects of practising kumbhaka)
+\startverse
+dvāsaptati sahāsrāṇāṃ nāḍīnāṃ malaśodhanam |\\
+yatheṣṭaṃ dhāraṇaṃ vāyor vikārābhāva eva ca || 120 ||
+\endverse
+\end{testimonia}
+
 \begin{philcomm}[hp01_039]
 Odd to have \emph{suṣumnām} as object of \emph{abhyaset}.
-YCM has this reading though.
+Yogacintāmaṇi has this reading though.
 Perhaps cd were added somewhat indiscriminately by Svātmārāma (with nominative \emph{suṣumnā}) and then others tried to make sense of it.
@@
 
-[\emph{nāḍiṣu} is better supported \getsiglum{J10ac,V19,J17}.]
+[\emph{nāḍiṣu} is better supported (J10ac,V19,J17).]
 \end{philcomm}
@@
 
+\begin{sources}[hp01_040]
+--
+\end{sources}
+
+\begin{testimonia}[hp01_040]
+Yogacintāmaṇi
+\startverse
+ātmadhyāyī mitāhārī yāvad dvādaśavatsaram |\\
+sadā siddhāsanābhyāsād yogī niṣpattim āpnuyāt |\\
+śramadair bahubhiḥ pīṭhaiḥ kiṃ syāt siddhāsane sati ||
+\endverse
+
+YSS
+\startverse
+ātmadhyāyo mitāhārī yāvad dvādaśavatsaram |\\
+sadā siddhāsanābhyāsād yoganiṣpattim āpnuyāt ||\\
+śramadair bahubhiḥ pīṭhair alaṃ siddhāsane sati |
+\endverse
+\end{testimonia}
+
 \begin{philcomm}[hp01_040]
 Only possible variant is \emph{mitāhāro} in V19.
-
 J8 might be correction of J10’s unmetrical reading.
@@
 
-V19 is found in YCM: \emph{śramadair bahubhịh pīṭhaiḥ kiṃ syāt siddhāsane sati}; JM this seems best to me.
+V1 has
+\startverse
+śramādau bahubhiḥ pīṭhais sadā siddhāsane sati |\\
+prāṇānile sāvadhānaṃ baddhe kevalakumbhake || 1.41 ||
+\endverse
+
+Mixing up both versions of the verse — contamination already?
+
+V19 is found in Yogacintāmaṇi: \emph{śramadair bahubhịh pīṭhaiḥ kiṃ syāt siddhāsane sati}; JM this seems best to me.
 
 Is this notion of āsanas causing \emph{śrama} already current in HY texts?
-[It is mentioned in the verse on śavāsana]
+[Yes, it is mentioned above in the verse on śavāsana]
 \end{philcomm}
@@
 \begin{translation}[hp01_041]
-When the Prāṇa breath has been carefully stopped in kevala kumbhaka, the Unmanī [state] arises automatically, without effort, in the same way that
-\end{translation}
+Just as the Unmanī [state] arises automatically, without effort, when the prāṇa breath has been carefully stopped in kevala kumbhaka, [...]
+\end{translation}
+
+\begin{sources}[hp01_041]
+\end{sources}
+
+\begin{testimonia}[hp01_041]
+Yogacintāmaṇi
+\startverse
+prāṇānile sāvadhāne baddhe kevalakumbhake |\\
+utpatsyate nirāyāsāt svayam evonmanī yathā ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_041]
-V1 has 
-\startverse
-śramādau bahubhiḥ pīṭhais sadā siddhāsane sati  |\\
-prāṇānile sāvadhānaṃ baddhe kevalakumbhake  || 1.41 ||
-\endverse
-Mixing up both versions of the verse — contamination already?
+--
 \end{philcomm}
@@
 \begin{translation}[hp01_042]
-when just Siddhāsana alone is always firmly bound, the three bandhas arise automatically, without effort.
-\end{translation}
+So too the three bandhas arise automatically, without effort, when Siddhāsana alone is always firmly adopted,
+\end{translation}
+
+\begin{sources}[hp01_042]
+--
+\end{sources}
+
+\begin{testimonia}[hp01_042]
+Yogacintāmaṇi
+\startverse
+athaikasminn eva dṛḍhaṃ baddhe siddhāsane sadā |\\
+bandhatrayam anāyāsāt svayam evopajāyate |
+\endverse
+
+YSS
+\startverse
+tathaikasminn eva baddhe siddhāsane sadā |\\
+granthitrayam anāyāsāt svayamevopabhidyate |
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_042]
-\emph{dṛḍhe} goes across stemma, although \emph{ṃ} vs \emph{e} is moot.
+\emph{dṛḍhe} goes across stemma.
+It is likely that the adverb (\emph{dṛdhaṃ}) was intended because \emph{dṛdhataraṃ}, which is not ambiguous, is used in 1.48a to qualify how \emph{padmāsana} should be adopted.
 \end{philcomm}
@@
 \end{translation}
+
+\begin{sources}[hp01_043]
+Śivasaṃhitā 5.47
+\startverse
+nāsanaṃ siddhasadṛśaṃ na kumbhasadṛśaṃ balam |\\
+na khecarīsamā mudrā na nādasadṛśo layaḥ || 5.47 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_043]
+Yogacintāmaṇi
+\startverse
+nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ |\\
+na khecarīsamā mudrā na nādasadṛśo layaḥ ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ ||\\
+na khecarīsamā mudrā na nādasadṛśo layaḥ || 3.29 ||
+\endverse
+\end{testimonia}
 
@@
 
-\subsection*{1.45–47}
+\begin{sources}[hp01_044]
+Vivekamārtaṇḍa 8
+\startverse
+vāmorūpari dakṣiṇañ ca caraṇaṃ saṃsthāpya vāmaṃ tathā\\
+yāmyorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\
+aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\
+etad vyādhivikārahāri yamināṃ padmāsanaṃ procyate || 8 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_044]
+Yogacintāmaṇi
+\startverse
+haṭhayoge—\\
+vāmorūpari dakṣiṇaṃ hi caraṇaṃ saṃsthāpya vāmaṃ tathā\\
+dakṣorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham | \\
+aṅguṣṭhau hṛdaye nidhāya civukaṃ nāsāgram ālokayet\\
+ekad vyādhivikāranāśanakaraṃ padmāsanaṃ procyate ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+vāmorūpari dakṣiṇaṃ ca caraṇaṃ saṃsthāpya vāmaṃ tathā\\
+yāmyorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\
+aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\
+etad vyādhivināśakāri yamināṃ padmāsanaṃ procyate || 3.34 ||
+\endverse
+\end{testimonia}
+
+\subsection*{1.45--47}
 \begin{translation}[hp01_045]
@@
 Having raised the root of the uvula with the tongue, one should place the chin on the chest and gradually [draw in\footnote{The verb \emph{ākṛśya} follows in the next verse in the \emph{Dattātreyayogaśāstra}.}] the breath [...].
-This is taught as lotus pose, which cures all diseases.
-\end{translation}
+\end{translation}
+
+\begin{sources}[hp01_045]
+Dattātreyayogaśāstra 35–37
+\startverse
+uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\
+ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 35 ||\\
+nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\
+uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\
+yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\
+yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||
+\endverse
+
+Śivasaṃhitā 3.102–104
+\startverse
+uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ\\
+ūrumadhye tathottānau pāṇī kṛtvā tu tādṛśau 3.102\\
+nāsāgre vinyased dṛṣṭiṃ rājadantaṃ ca jihvayā\\
+uttambhya cibukaṃ vakṣe saṃsthāpya pavanaṃ śanaiḥ 3.103\\
+yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ\\
+yathāśaktyaiva paścāt tu recayed anirodhataḥ 3.104
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_045]
+Yogacintāmaṇi
+\startverse
+dattātreyaḥ—\\
+uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\
+ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || = 1.45\\
+nāsāgre vinyased rājadantamūlaṃ tu jihvayā |\\
+uttabhya civukaṃ vakṣasy utthāpya pavanaṃ śanaiḥ || = 1.46\\
+yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\
+yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+dattātreyo 'pi\\
+uttānau caraṇau kṛtvā ūrvoḥ saṃsthāpya yatnataḥ |\\
+ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 3.36 ||\\
+nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\
+uttabhya cibukaṃ vakṣaḥ saṃsthāpya pavanaṃ śanaiḥ || 3.37 ||
+\endverse
+\end{testimonia}
 
@@
 The witnesses split predictably along the two main branches of the stemma. The evidence of the DYŚ is important here.
-
-\emph{vakṣasyāsthāpya} is a marmasthāna
-
-%\emph{pavanaṃ śanaiḥ} is left hanging.
-The readings with \emph{vakṣa sthāpayet} do not offer a solution nor indicate how Svātmārāma may have redacted this to make the syntax complete.
-Instead, it seems that he quoted these two verses (1.45–46) from the Dattātreyayogaśāstra and simply omitted the next verse that made sense of \emph{pavanaṃ śanaiḥ} because it was not relevant to the posture itself.
-
-\emph{pavanaṃ śanaiḥ} is left hanging, perhaps, because of an eyeskip that happened early in the transmission. The following verse in the DYŚ ends with \emph{pavanaṃ śanaiḥ}.
+%\emph{vakṣasyāsthāpya} is a marmasthāna
+
+The statement ending with \emph{pavanaṃ śanaiḥ} is left hanging, perhaps, because of an eyeskip that happened early in the transmission.
+The subsequent verse in the DYŚ ends with \emph{pavanaṃ śanaiḥ}.
 
@@
 \endverse
+
+The manuscript readings with \emph{vakṣa sthāpayet} (J7, V3, J8, J10, J17, N17) or something similar (V1, W4) do not offer a solution nor indicate how Svātmārāma may have redacted this to make the syntax complete.
+Instead, it seems that he quoted these two verses (1.45–46) from the Dattātreyayogaśāstra and simply omitted the next verse that made sense of \emph{pavanaṃ śanaiḥ} because it was not relevant to the posture itself.
 
@@
 ``\emph{khecarī}''.\footnote{\emph{tālumūlāntarālagnajihvamūlah. tālumūlayoh. kākudamūladeśayoh. āntare
-    lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā / yā hi khecarı̄ty ucyate.}}
+  lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā | yā hi khecarı̄ty ucyate.}}
 
@@
 \begin{translation}[hp01_047]
-It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
-\end{translation}
+This is taught as lotus pose, which cures all diseases. It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
+\end{translation}
+
+\begin{sources}[hp01_047]
+DYŚ
+\startverse
+idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\
+durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 38 ||
+\endverse
+
+Śivasaṃhitā
+\startverse
+idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam\\
+durlabhaṃ yena kenāpi dhīmatā labhyate param 3.105
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_047]
+Yogacintāmaṇi
+\startverse
+idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\
+durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\
+durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 3.38 ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_047]
-Durlabham is ambiguous as to whether the posture is hard to perform or hard to acquire. The latter is the usual reading.
+In this context, the word \emph{durlabham} is ambiguous as to whether the posture is hard to perform or hard to acquire. The latter is the usual reading.
+% Maybe add some remarks by commentators on how they understand it.
 \end{philcomm}
@@
 
+\begin{sources}[hp01_048]
+VM 36
+\startverse
+kṛtvā saṃpuṭitau karau dṛḍhataraṃ baddhvātha padmāsanaṃ\\
+gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyānaṃś ca tac cetasi |\\
+vāraṃ vāram apānam ūrdhvam anilaṃ proccālayan pūritaṃ\\
+muṇcan prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_048]
+Yogacintāmaṇi
+\startverse
+tathāca granthāntare—\\
+kṛtvā saṃpuṭitau karau dṛḍhataraṃ badhvā ca padmāsanam\\
+gāḍhaṃ vakṣasi saṃnidhāya civukaṃ dhyānaṃ ca tac cetasi |\\
+vāraṃ vāram apānam ūrdhvam anilaṃ protsārayet pūrayet\\
+prāṇaṃ muñcati bodham eti niyataṃ śaktiprabodhodayāt ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+kṛtvā saṃpuṭitau karau dṛḍhataraṃ baddhvā tu padmāsanam\\
+gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyānaṃ ca tac cetasi |\\
+vāraṃ vāram apānam ūrdhvam anilaṃ proccārayet pūritam\\
+muñcat prāṇam upaiti bodham atulaṃ śakteḥ prabhāvān naraḥ || 3.39 ||
+\endverse
+\end{testimonia}
+
 \begin{philcomm}[hp01_048]
-It is unusual to find Matsyendra referred to as simply Matsya as in the subheading here.
-This verse is Vivekamārtaṇḍa 36.
-The teachings of the VM are attributed to Gorakṣa, not Matsyendra, although in the extended recension of it known as Gorakṣaśataka etc.
+The reading of \emph{matsyamataṃ} in the subheading is unusual because elsewhere Matsyendra is not referred to as simply Matsya.
+Furthermore, this verse is from the \emph{Vivekamārtaṇḍa} (36), and the teachings of the \emph{Vivekamārtaṇḍa} are attributed to Gorakṣa, not Matsyendra.
+However, in the extended recension of it known as the  \emph{Gorakṣaśataka},  \emph{Gorakṣasaṃhitā}, etc.
 the second verse includes homage to Mīna, i.e.
@@
 The end of pāda b is uncertain.
-Later witnesses including Brahmānanda have \emph{dhyāyaṃś ca} but none of the early ones does.
-So we have to take \emph{dhyānaṃ} with \emph{sannidhāya}.
-That makes tat problematic though: what does it refer to?
+Later witnesses including Brahmānanda have \emph{dhyāyaṃś ca} but none of the early ones has this reading.
+So, we have to take \emph{dhyānaṃ} with \emph{sannidhāya}.
+However, \emph{tat} is problematic because here it has no clear referent.
+In the \emph{Vivekamārtaṇḍa tat} appears to refer to the \emph{mokṣadvāra} broken by Kuṇḍalinī, which is mentioned in the previous verse.
 
@@
 \begin{translation}[hp01_049]
-While in padmāsana the yogi should fill himself up [with Prāṇa] through [all] the channels.  He who takes in breath [in this way] is sure to be liberated.
-\end{translation}
+While in padmāsana the yogi should inhale through the opening to the central channel. He who holds the breath is sure to be liberated.
+\end{translation}
+
+\begin{sources}[hp01_049]
+--
+\end{sources}
+
+\begin{testimonia}[hp01_049]
+Yogacintāmaṇi
+\startverse
+padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\
+mārutaṃ dhārayed yas tu sa mukto nātra saṃśayaḥ ||
+\endverse
+
+Haṭharatnāvalī 3.40
+\startverse
+padmāsane sthito yogī nāḍīdvāreṣu pūrayet |\\
+pūritaṃ dhrīyate yas tu sa mukto nātra saṃśayaḥ || 3.40 ||
+\endverse
+
+Dhyānabindu Up
+\startverse
+padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\
+mārutaṃ kumbhayan yas tu sa mukto nātra saṃśayaḥ || 70 ||
+\endverse
+\end{testimonia}
+
+%\begin{philcomm}[hp01_049]
+% comment on \emph{dvāreṣu} and \emph{dvāreṇa} and explain how the reading with \emph{pibati} in the third pāda might have been understood (i.e., by supplying evam).
+%\end{philcomm}
 
@@
 Now the lion’s pose
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. Left ankle is on right, and the right ankle on the left.
-
-Having placed both hands on the knees and spread his fingers, his mouth open, he gazes with his eyes fixed on the tip of the nose.
+[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
+Having placed the left ankle on right, the right ankle on the left and both hands on the knees and spread his fingers, with his mouth open, he stares at the tip of the nose.
 This is the lion’s pose, which is always worshipped by yogis.
-It unites the three locks and is the best of all postures.
-\end{translation}
+It causes the three locks to arise together and is the best of [all] postures.
+\end{translation}
+
+\begin{sources}[hp01_050]
+Vasiṣṭhasaṃhitā 1.73
+\startverse
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
+dakṣiṇaṃ savyagulphena dakṣiṇenetaretaram ||\\
+hastau jānau ca saṃsthāpya svāṅgulīś ca prasārya ca |\\
+vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ ||\\
+siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā ||
+\endverse
+
+Yogayājñavalkya 3.9
+\startverse
+gulpau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\
+dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram ||\\
+hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca\\
+vyāttavaktro nirīkṣet nāsagraṃ susamāhitaḥ ||\\
+siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā
+\endverse
+
+Sūtasaṃhitā 15.7
+\startverse
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
+dakṣiṇaṃ savyagulphena vāmaṃ dakṣiṇagulphataḥ || 7 ||\\
+hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca |\\
+nāsāgraṃ ca nirīkṣeta bhavet siṃhāsanaṃ hi tat || 8 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_050]
+Yogacintāmaṇi (citing Yogayājñavalkya)
+\startverse
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
+dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram || \\
+hastau jānūpari sthāpya svāṅgulīḥ saṃprasārya ca |\\
+vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ | \\
+siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā |
+\endverse
+
+[Note: The Yogacintāmaṇi attributes its citation of verses on siṃhāsana to the Yogayājñavalkya and does not include HP 1.52cd, which affirms that 1.52cd is not from the Yogayājñavalkya (or Vasiṣṭhasaṃhitā)].
+
+Haṭharatnāvalī
+\startverse
+atha siṃhāsanam\\
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
+dakṣiṇe savyagulphaṃ ca dakṣiṇe tu tathetaram || 3.31 ||\\
+hastau tu jānvoḥ saṃsthāpya svāṅgulīḥ samprasārya ca ||\\
+vyāttavaktro nirīkṣeta nāsāgraṃ tu samāhitaḥ || 3.32 ||\\
+siṃhāsanaṃ bhaved etat sevitaṃ yogibhiḥ sadā ||\\
+bandhatritayasaṃsthānaṃ kurute cāsanottamam || 3.33 ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_050--052]
+% comment on nyastalocanaḥ and susamāhitaḥ.
+The latter is well attested by the sources and testimonia, and is in the V19 and N23 branch.
+However, most of the old mss.
+have nyastalocanaḥ.
+Maybe comment on the meaning of sandhāna, which seems to mean simply that the three locks arise together (cf.
+a similar comment on siddhāsana in 1.42).
+Spreading the fingers and keeping the mouth open seem to be imitating the lion, and this is depicted in some iconography of yoganarasiṃha (for example, Yoga Narasimha, Vishnu's Man-Lion Incarnation, Samuel Eilenberg Collection, Bequest of Samuel Eilenberg, 1998, Accession Number: 2000.284.4.
+https://www.metmuseum.org/art/collection/search/39251).
+\end{philcomm}
 
@@
 
+\begin{philcomm}[hp01_054]
+We have understood \emph{pārśvapāda} (i.e., side of the foot) like the compound \emph{agrapāda} (i.e., front of the foot), which can be classified as a \emph{ekadeśitatpuruṣa}.
+See Sanskrit Vademecum, p.84 (ed.\ Maximilian Mehner and Jürgen Hanneder, Marburg 2019).
+\end{philcomm}
+
+\begin{sources}[hp01_053]
+Vasiṣṭhasaṃhitā 1.79
+\startverse
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipan | \\
+pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam |\\
+bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham ||
+\endverse
+
+Yogayājñavalkya 3.11cd--3.12ab
+\startverse
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\
+pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam\\
+bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_053]
+Yogacintāmaṇi (citing Yājñavalkya)
+\startverse
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
+pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ badhvā suniścalaḥ |\\
+bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham |
+\endverse
+
+Haṭharatnāvalī
+\startverse
+atha bhadrāsanam\\
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet ||\\
+pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam ||\\
+bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham || 3.30 ||
+\endverse
+
+YSS
+\startverse
+gulphau ca vṛṣaṇasyāyasvīvinyāḥ pārśvayoḥ kṣipet |\\
+pārśvapāde ca pāṇibhyāṃ dṛḍhaṃ badhvāsu niścalaḥ ||\\
+bhadrāsanaṃ bhaved etat sarvavyādhivināśanam |\\
+gorakṣāsanam ity āhur idaṃ vai siddhayoginaḥ ||
+\endverse
+\end{testimonia}
+
 \subsection*{1.55}
 \begin{translation}[hp01_055]
-When the great yogi does not become tired [while engaged] in these āsanas and bandhas, he should now practise the breath techniques of mudrā and so forth in order to purify the channels.
-\end{translation}
+When the great yogi does not become tired from adopting the āsanas in this way, he should now practise the breath techniques with mudrās and so forth from which purification of the channels arise.
+\end{translation}
+
+\begin{sources}[hp01_055]
+\end{sources}
+
+\begin{testimonia}[hp01_055]
+Yogacintāmaṇi
+\startverse
+evam āsanabandheṣu yogīndro vijitaśramaḥ |\\
+abhyasen nāḍiśuddhiṃ ca mudrayā pavanakriyām || iti ||
+\endverse
+
+HSC
+\startverse
+evam āsanabandhastho yogīndro vigataśramaḥ |\\
+athābhyasen nāḍiśuddhiṃ mudrādipavanakriyām || 73 ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_055]
-55d is ra-vipulā with \emph{nāḍi}
-
-\emph{nāḍiśuddhiṃ} is difficult to construe.
-Jim thinks it is a second accusative of the \emph{abhyaset} and reads it as \emph{nāḍiśuddhyartha}.
-I wonder if it is not better understood as a bahuvrīhī qualifying \emph{°pavanakriyām} (i.e., the yogi should practise a breath technique that is channel-purifying.
-In other words, he does not practise just any breath technique).
-As a bahuvṛīhi, it may also be understood as a \emph{hetuviśeṣaṇa} (i.e., why should the yogi practice a breath-technique such as mudrā?
-Because it is channel-purifying).
-\end{philcomm}
-
-\subsection*{1.55*1}
+55c is ra-vipulā with \emph{nāḍi}
+The second hemistich can be interpreted in different ways.
+One possibility is to understand \emph{nāḍiśuddhiṃ} as a bahuvrīhi qualifying \emph{mudrādipavanakriyām} in the sense that one should practise a breathing technique with mudrās from which purification of the channels arise.
+Alternatively, one could separate \emph{mudrādi} from \emph{pavanakriyām} and understand three different techniques here, namely, the practice of purifying the channels (perhaps by the alternative nostril method mentioned at the beginning of the second chapter), the \emph{mudrā}s and the breathing techniques of prāṇāyāma.
+\end{philcomm}
+
+\subsection*{1.55*1--2}
 \begin{translation}[hp01_055_1]
@@
 
+\begin{sources}[hp01_055_1]
+Dattātreyayogaśāstra 42cd--47
+\startverse
+kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet |\\
+na śāstrapāṭhamātreṇa kā cit siddhiḥ prajāyate |\\
+na veṣadhāraṇaṃ siddheḥ kāraṇaṃ na ca tatkathā |\\
+kriyaivakāraṇaṃ siddheḥ satyam eva tu sāṃkṛte || 46 ||\\
+śiśnodarārthaṃ yogasya kathayā veṣadhāriṇaḥ |\\
+anuṣṭhānavihīnās tu vañcayanti janān kila || 47 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_055_1]
+\end{testimonia}
+
 \begin{philcomm}[hp01_055_1]
@@
 \end{translation}
+
+\begin{testimonia}[hp01_055_3]
+Vārāhītantra
+\startverse
+mayi bodhībudho svasthe tucho yaṃ viśvabudbudaḥ |\\
+malīna udito vetti vikalpāvasaraḥ kutaḥ |
+\endverse
+
+Siddhāntamuktāvalī
+\startverse
+śiśnodararatāya hi na deyaṃ veṣadhāriṇe ||\\
+mayi bodhyaṃ buddhau svacche tad dheyaṃ viśvabudbudam || 7 ||
+\endverse
+
+Yogaprakāśikā
+\startverse
+“śiśnodararatāyaitan na deya" etat yogajñānam etena śiśnodararatas tyājyo nanv etanmate tyājyapadārtho 'prasiddha iti śaṃkāṃ nirasyati mayi iti svacche bodhasvarūpasamudre budbudatulyasya viśvasya heyatvād iti bhāvaḥ
+\endverse
+\end{testimonia}
 
@@
 \emph{Vairāgyaśataka}. A still superficial glance at Sillana's \emph{Svātmopalabdhiśataka} gives
-the impression that our verse would fit there, but not so much in the \emph{Śāntiśataka}.  While
+the impression that our verse would fit there, but not so much in the \emph{Śāntiśataka}. Perhaps its first citation is in Advayavajra’s \emph{Tattva\-ratnāvalī} (ref??),  While
 these are only preliminary observations the verse is probably not original in any stage of
@@
 \begin{translation}[hp01_055_4]
-Realisation from scripture, one's own guru or oneself and the cessation of mind; all these are methods that have been systematised and taught by the wise in this tradition.
-\end{translation}
+Realisation from scripture, one's own guru or oneself and the cessation of mind; all these methods have been combined and taught by the wise in this tradition.
+\end{translation}
+
+\begin{sources}[hp01_055_4]
+\end{sources}
+
+\begin{testimonia}[hp01_055_4]
+Yogacintāmaṇi
+\startverse
+haṭhapradīpikāyām--\\
+śrutapratītiḥ svagurupratītiḥ, svātmapratītir manaso nirodhaḥ |\\
+etāni sarvāṇi samuccitāni, matāni dhīrair iha sādhanāni ||
+\endverse
+\end{testimonia}
 
@@
 
-The reading in the YCM (\emph{manaso nirodhaḥ}) is better in a yogic context.
+The reading in the Yogacintāmaṇi (\emph{manaso nirodhaḥ}) is better in a yogic context.
 \end{philcomm}
@@
 
+\begin{sources}[hp01_056]
+\end{sources}
+
+\begin{testimonia}[hp01_056]
+Yogacintāmaṇi
+\startverse
+haṭhapradīpikāyām--\\
+āsanaṃ kumbhakaṃ citraṃ mudrākhyaṃ karaṇaṃ tathā |\\
+atha nādānusandhānam abhyāsānukrameṇa ca ||
+\endverse
+\end{testimonia}
+
 \begin{philcomm}[hp01_056]
@@
 \begin{translation}[hp01_057]
-Celibate, restricted in diet and devoted to yoga, a yogi succeeds in upwards of a year. No doubt about this should be entertained.
-\end{translation}
-
-%\begin{philcomm}[hp01_057]
-%\end{philcomm}
+Celibate, restricted in diet and devoted to yoga, the renunciant succeeds in upwards of a year. No doubt about this should be entertained.
+\end{translation}
+
+\begin{sources}[hp01_057]
+Vivekamārtaṇḍa 37
+\startverse
+brahmacārī mitāhārī yogī yogaparāyaṇaḥ |\\
+abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_057]
+Yogacintāmaṇi
+\startverse
+brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ |\\
+abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ ||\\
+abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā || 3.28 ||
+\endverse
+\end{testimonia}
+
+\begin{philcomm}[hp01_057]
+Comment on tyāgī and yogī, and the reading of this verse in the Vivekamārtaṇḍa.
+It seems likely that Svātmārāma was using a version of the VM with tyāgī (as it’s unlikely that yogī would be emended to tyāgī), but whether this was the original reading in the VM is more complicated because there tyāgī may be a dittographical-type mistake.
+\end{philcomm}
 
@@
 \begin{translation}[hp01_058]
-When very unctuous and sweet food that is without a quarter portion [of the stomach] is eaten for love of śiva, it is called a restricted diet (\emph{mitāhāra}).
-\end{translation}
+When very unctuous and sweet food that is without a quarter portion is eaten for love of śiva, it is called a restricted diet. (\emph{mitāhāra}).
+\end{translation}
+
+\begin{sources}[hp01_058]
+Gorakṣaśataka (original)
+\startverse
+susnigdhamadhurāhāraś caturthāṃśavivarjitaḥ |\\
+bhujyate śivasaṃprītyai mitāhāraḥ sa ucyate ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_058]
+Yuktabhavadeva
+\startverse
+tad uktaṃ haṭhapradīpikāyām-\\
+susnigdhamadhurāhārāś caturthāṃśavivarjitaḥ ||\\
+bhujyate śivasamprītyai mitāhāraḥ sa ucyate || 16 ||
+\endverse
+
+Yogacūḍāmaṇyupaniṣat
+\startverse
+susnigdhamadhurāhāraścaturthāṃśavivarjitaḥ |\\
+bhuñjate śivasaṃprītyā mitāhārī sa ucyate || 43 ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_058]
-This verse probably derives from the ‘original’ GŚ (12c–13b). It is also found, but reworked to be about the \emph{mitāhārī}, in Nowotny’s GŚ (extended recension of VM) at 55.
+This verse probably derives from the ‘original’ GŚ (12c–13b).
+It is also found, but reworked to be about the \emph{mitāhārī}, in Nowotny’s GŚ (extended recension of VM) at 55.
+The idea of not eating the fourth portion of one’s food (\emph{caturthāṃśavivarjitaḥ}) can be found in older sources, such as Dharmaputrikā 1.51-52 (ṣaḍrasopetasuṣnigdhasvādusāndrasugandhinā |
+udarasyārdhabhāgan tu bhojanena prapūrayet || pānīyena caturbhāgaṃ taccheṣaṃ śūnyam iṣyate | vāyos sañcāraṇānārtham āhāraniyamaḥ smṛtaḥ) and Aṣṭāṅgahṛdayasaṃhitā Sūtrasthāna 8.46c-47b ( annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet ||āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet).
+As noted in the Jy
+
 \end{philcomm}
@@
 [Adepts] say the [following] is unwholesome: pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink.
-Flesh of goats and sheep, curds, diluted buttermilk, poor man's pulse, Jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like.
-\end{translation}
+Flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, Jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like.
+\end{translation}
+
+\begin{sources}[hp01_059]
+\end{sources}
+
+\begin{testimonia}[hp01_059]
+Yogacintāmaṇi
+\startverse
+haṭhapradīpikāyām--\\
+kaṭvamlatīkṣṇalavaṇoṣṇaharītaśāka-\\
+sauvīratailatilasarṣapamatsyamadyam |\\
+ajādimāṃsadadhitakrakulatthakola-\\
+piṇyākahiṅgulaśunādyam apathyam āhuḥ ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+kaṭvamlatīkṣṇalavaṇoṣṇaharītaśākaṃ \\
+sauvīratailatilasarṣapamatsyamadyam |\\
+ajādimāṃsadadhitakrakulatthakodra-\\
+piṇyākahiṅgulaśunādyam apathyam āhuḥ || 1.72 ||
+\endverse
+
+HTK
+\startverse
+atha varjyāni –\\
+kaṭvamlatīkṣṇalavaṇoṣṇa haritaśāka-\\
+sauvīratailatilasarṣapamatsyamadyam ||\\
+ajāvimāṃsadadhitakrakulatthakola-\\
+piṇyākahiṃgulaśunādyam apathyam āhuḥ || 28 ||
+\endverse
+\end{testimonia}
 
@@
 
-\emph{°hari}(\emph{ī} here for metre?)\emph{taśāka°} in some nighaṇṭus is horseradish, which makes better sense here.
+\emph{°hari}(\emph{ī} here for metre?)\emph{taśāka°} in some nighaṇṭus is horseradish, which makes better sense here than Brahmānanda’s gloss.
 \startverse
@@
 \endverse
+
 Brahmānanda’s understanding of \emph{harītaśāka} as \emph{patraśāka} is probably wrong if \emph{patraśāka} was intended as ‘leafy vegetables’.
@@
 añjanaṃ yāmunaṃ kṛṣṇaṃ nādeyaṃ mecakaṃ tathā \\
-srotojaṃ dṛkpradaṃ nīlaṃ sauvīraṃ ca suvīrajam //  \textup{Rājanighaṇṭu 13.86}
-\endverse
+srotojaṃ dṛkpradaṃ nīlaṃ sauvīraṃ ca suvīrajam ||  \textup{Rājanighaṇṭu 13.86}
+\endverse
+
 Note also that the Yogaprakāśikā takes \emph{sauvīra} with \emph{taila}, perhaps to solve the problem of \emph{taila} on its own (see below for more on this): \emph{sauvīrataila} -- oil produced in the place Suvīra (\emph{suvīradeśodbhavatailam}).
@@
 
-\emph{ajāvimāṃsa}: note the variant \emph{ajādimāṃsa}, which makes better sense to me as \emph{aja} might simply indicate that the flesh of animals is meant (as opposed to humans). 
-\emph{ājāvi°} is needed for the metre.
+\emph{ājādimāṃsa}: āja° is required for the metre, so only ājādi makes sense, not ājāvi because there is no adjective āvi.
+
+Diwakar Acharya: prohibition of goat flesh and fish is aimed at north/east India.
 
@@
 \emph{kola}: Zizyphus Jujuba (Nadkarni 1926: pp.
-919-920).
+919--920).
 Also known as \emph{badara}.
@@
 Root is useful as a decoction in fever and delirium.
+kola can mean banana in some parts of India (Diwakar)
 
@@
 
-\emph{piṇyāka}: Sharma (Ḍalhaṇa and his commentary on drugs: 1982: 69) says, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’
+\emph{piṇyāka}: Sharma (Ḍalhaṇa and his commentary on drugs: 1982: 69) says, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’ Diwakar says it is an oil cake that has a strong flavour.
 
@@
 \begin{translation}[hp01_060]
-One should know as unwholesome food that has been reheated and is dry.
-[Food that is] too salty, the leftover paste of crushed sesame seeds (tilapiṇḍa), spoiled food, [prohibited] vegetables (?), and pepper (?) are to be omitted.
-\end{translation}
+One should know food that has been reheated, is dry, is too salty, the leftover paste of crushed sesame seeds (tilapiṇḍa), spoiled [or] contains an excess of leafy vegetables to be unwholesome.
+It is to be avoided.
+\end{translation}
+
+\begin{sources}[hp01_060]
+\end{sources}
+
+\begin{testimonia}[hp01_060]
+Yogacintāmaṇi
+\startverse
+bhojanam ahitaṃ vidyāt punar <apy> uṣṇīkṛtaṃ tathā rukṣam |\\
+atilavaṇaṃ sapalaṃ vā prasitaṃ śākotkaṭaṃ varjyam ||
+\endverse
+
+HSC
+\startverse
+bhojanam ahitaṃ vidyāt punar uṣṇīkṛtaṃ rūkṣaṃ | \\
+atilavaṇādikayuktaṃ kadaśanaśākotkaṭaṃ duṣṭaṃ ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_060]
+We have understood the compound \emph{kadaśanaśākotkaṭa} as a dvandva referring to spoiled food (\emph{kadaśana}) and an excess of leafy vegetables (\emph{śākotkaṭa}).
 We have not found any conclusive evidence for the meaning of \emph{tilapiṇḍa}. Brahmānanda glosses it as \emph{piṇyāka} (on which see the notes for the previous verse).
@@
 Similarly a saying by Goraksa:
-One should avoid a liking for bad people, frequenting fire, women and roads, early morning bathing, fasting and the like, as well as such things as harming the body.
-\end{translation}
+One should avoid a liking for bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.
+\end{translation}
+
+\begin{sources}[hp01_061]
+Amṛtasiddhi 19.7
+\startverse
+agnisevābalāsevā pathasevā ca sarvadā |\\
+prathamābhyāsakāle tu saṃtyājyā yoginā sadā || 19.7 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_061]
+Yogacintāmaṇi
+\startverse
+haṭhadīpikāyām—\\
+varjayed durjanaprītiṃ vahnistrīpathasevanam |\\
+prātaḥsnānopavāsādi kāyakleśādikaṃ tathā ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+tathā ca gorakṣavacanam---\\
+varjayed durjanaprītivahnistrīpathasevanam |\\
+prātaḥsnānopavāsādi kāyakleśādikaṃ tathā || 1.73 ||
+\endverse
+\end{testimonia}
 
 \begin{philcomm}[hp01_061]
-The vulgate has a parallel from the Amaraugha added. Also, the vulgate reads \emph{kāyakleśavidhiṃ tathā}.
+The vulgate has a parallel from the Amaraugha added. 
 \end{philcomm}
@@
 Pure food with wheat, rice, śāli rice, barley, sixty-day śāli rice, milk, ghee, candied sugar, unclarified fermented butter, ground sugar and honey.
-Dried ginger, fruit of the snake gourd and five vegetables, mung beans and so on, and rain water are wholesome for the best of sages.
-\end{translation}
+Dried ginger, fruit of the snake gourd and so forth, the five vegetables, mung beans and so on, and rain water are wholesome for the best of sages.
+\end{translation}
+
+\begin{sources}[hp01_062]
+\end{sources}
+
+\begin{testimonia}[hp01_062]
+Yogacintāmaṇi
+\startverse
+godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ\\
+kṣīrājyamaṇḍanavanītasitāmadhūni |\\
+śuṇṭhīpaṭolakaphalādikapañcaśākaṃ\\
+mudgādi cālpam udakaṃ ca munīndrapathyam ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+godhūmaśāliyavaṣaṣṭikaśobhanānnaṃ \\
+kṣīrājyamaṇḍanavanītasitāmadhūni |\\
+śuṇṭhīpaṭolaphalapatrajapañcaśākaṃ \\
+mudgādidivyam udakaṃ ca yamīndrapathyam || 1.71 ||
+\endverse
+
+YBD
+\startverse
+tathā ca śivayoge-\\
+godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ \\
+kṣīrājyakhaṇḍanavanītasitāmadhūni ||\\
+śuṇṭhīpaṭolakaphalādi ca pañcaśāka-\\
+mudgādidivyam udakaṃ ca munīndrapathyam || 21 ||
+\endverse
+\end{testimonia}
 
@@
 
-Jyotsnā
+Jyotsnā and Yuktabhavadeva 4.22
 \startverse
@@
 It is not entirely clear how we should understand \emph{divya}.
-Brahmānanda glosses it a \emph{nirdoṣa} and takes it with \emph{udaka}.
+Brahmānanda glosses it with \emph{nirdoṣa} and takes it with \emph{udaka}.
 But could it refer more specifically to \emph{gaṅgāmbu} (as suggested by Paras) or rain water?
@@
 \startverse
-divyodakaṃ kharāri syād ākāśasalilaṃ tathā /\\
-vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam // \textup{Rajni 14.4}
+divyodakaṃ kharāri syād ākāśasalilaṃ tathā |\\
+vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam || Rajni 14.4
 \endverse
@@
 ambuśabde vivṛtiḥ .
+
+62*1 is quoted in the Jyotsna as from a medical work (“vaidyake”).
 \end{philcomm}
@@
 \begin{translation}[hp01_063]
-The yogi should eat food that is pure (\emph{mṛṣṭa}, or delicious \emph{miṣṭa}), unctuous, and containing diary. It should nourish the bodily constituents, satiate the mind and be fit [for yoga].
-\end{translation}
+The yogi should eat food that is sweet, delicious, unctuous, contains cow products, nourishes the bodily constituents, is desired by the mind and is appropriate.
+\end{translation}
+
+\begin{sources}[hp01_063]
+\end{sources}
+
+\begin{testimonia}[hp01_063]
+Yogacintāmaṇi
+\startverse
+piṣṭaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\
+manobhilaṣitaṃ yogyaṃ yogī bhojanam ācared iti ||
+\endverse
+
+YBhD
+\startverse
+śreṣṭhaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam ||\\
+mano'bhilaṣitaṃ yogyaṃ yogī bhojanamācaret || 23 ||
+\endverse
+\end{testimonia}
 
@@
 
+\begin{sources}[hp01_064]
+DYS 40
+\startverse
+yuvāvastho 'pi vṛddho vā vyādhito vā śanaiḥ śanaiḥ |\\
+abhyāsāt siddhim āpnoti yoge sarvo 'py atandritaḥ || 40 ||
+\endverse
+\end{sources}
+
+\begin{testimonia}[hp01_064]
+Yogacintāmaṇi
+\startverse
+haṭhapradīpikāyām—\\
+yuvā bālo 'tivṛddho vā vyādhito durbalo 'pi vā |\\
+abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ ||
+\endverse
+
+Haṭharatnāvalī
+\startverse
+yuvā bhavati vṛddho 'pi vyādhito durbalo 'pi vā |\\
+abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ || 1.23 ||
+\endverse
+\end{testimonia}
+
 \begin{philcomm}[hp01_064]
-Note the different reading in 164d for V1: \emph{sarvaṃ yogī yatendriyaḥ}. Sarvaṃ is not easy to construe, and the testimony of the DYŚ suggests that \emph{yoge sarvo ’py atandritaḥ} was original.
+Note the different reading in 164d for V1: \emph{sarvaṃ yogī yatendriyaḥ}.
+\emph{sarvaṃ} is not easy to construe, and the testimony of the DYŚ suggests that \emph{yoge sarvo ’py atandritaḥ} was original.
 \end{philcomm}
@@
 \begin{translation}[hp01_065]
-The postures, various breath retentions and techniques, beginning with seals, are all to be done in the practice of Haṭha until the goal of Rājayoga [is attained].
-\end{translation}
+The postures, various breath retentions and techniques, beginning with seals, are all [to be done] in the practice of Haṭha until the reward that is Rājayoga [is attained].
+\end{translation}
+
+\begin{sources}[hp01_065]
+\end{sources}
+
+\begin{testimonia}[hp01_065]
+Haṭharatnāvalī
+\startverse
+pīthāni kumbhakāś citrā divyāni karaṇāni ca |\\
+sāṅgo 'pi ca haṭhābhyāso rājayogaphalārthadaḥ || 1.17 ||
+\endverse
+\end{testimonia}
 

### 2023-04-11 compile-tidyup.sh

@@ -2,7 +2,7 @@
 
 # optional argument 1 is filename, otherwise all relevant *.tex files
 
-xslcmd="java -jar /usr/share/java/saxon/saxon9he.jar -s:xml/HP1_OldMss-tei.xml -xsl:xslt/hp-online.xsl"
+xslcmd="java -jar ${HOME}/.nix-profile/saxon9he.jar -s:xml/HP1_edition-tei.xml -xsl:xslt/hp-online.xsl"
 
 xslcmdproc(){
     ${xslcmd} | \
@@ -45,7 +45,7 @@ then
     compile "${1#*/}"
 elif [ -z "${1:-}" ]
 then
-    for f in "HP1_OldMss.tex" "HP1_TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex"
+    for f in "HP1_edition.tex" "HP1_TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex"
     do
 	compile "${f}"
     done

### 2023-04-11 html/switch.js

@@ -52,7 +52,7 @@ document.addEventListener("DOMContentLoaded", () => {
 	let class_a = document.getElementsByClassName(scriptclass);
 	
 	for (var i = class_a.length -1 ; i >= 0; i--) {
-	    let div=class_a[i]
+	    let div=class_a[i];
 	    div.style.display="none";
 	} 
     }
@@ -61,7 +61,7 @@ document.addEventListener("DOMContentLoaded", () => {
 	let class_a = document.getElementsByClassName(scriptclass);
 
 	for (var i = class_a.length -1 ; i >= 0; i--) {
-	    let div=class_a[i]
+	    let div=class_a[i];
 	    div.style.display="block";
 	    
 	}

### 2023-04-13 latex/CommStyle.tex
included testimonia

@@
 \setlength{\vindent}{0pt}
-\def\startverse{\begin{ekdverse}\begin{itshape}}
-\def\endverse{\normalsize\end{itshape}\end{ekdverse}} % \textup
+\def\startverse{\begin{ekdverse}\begin{itshape}} %%% marked for change 2023-04-13 mm
+\def\endverse{\normalsize\end{itshape}\end{ekdverse}} % \textup %%% marked for change 2023-04-13 mm
+\NewDocumentEnvironment{versinnote}{O{}}{\begin{ekdverse}\begin{itshape}}{\end{itshape}\end{ekdverse}} %%% change 2023-04-13 mm
 \setvnum{}

### 2023-04-15 latex/HP1_TranslComm.tex
included testimonia
split sourcas and testimonia

@@
 Cf. Yogasārasaṃgraha, p. 54.
-\startverse
-sadādināthāya namo'stu tubhyaṃ yenopadiṣṭā haṭhayogavidyā | \\
-virājate pronnatarājayogam āroḍhum icchoradhirohiṇīva ||
-\endverse
+
+\begin{versinnote}
+\tl{sadādināthāya namo'stu tubhyaṃ yenopadiṣṭā haṭhayogavidyā |\\+}
+\tl{virājate pronnatarājayogam āroḍhum icchoradhirohiṇīva ||\\!}
+\end{versinnote}
 
 Gheraṇḍasaṃhitā 1.1
-\startverse
-ādīśvarāya praṇamāmi tasmai yenopadiṣṭā haṭhayogavidyā | \\
-virājate pronnatarājayogam āroḍhum icchor adhirohiṇīva ||
-\endverse
+
+\begin{versinnote}
+\tl{ādīśvarāya praṇamāmi tasmai yenopadiṣṭā haṭhayogavidyā |\\+}
+\tl{virājate pronnatarājayogam āroḍhum icchor adhirohiṇīva ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Haṭharatnāvalī 1.4
-\startverse
-bhrāntyā bahumatadhvānte rājayogam ajānatām |\\
-kevalaṃ rājayogāya haṭhavidyopadiśyate || 
-\endverse
+
+\begin{versinnote}
+\tl{bhrāntyā bahumatadhvānte rājayogam ajānatām |\\+}
+\tl{kevalaṃ rājayogāya haṭhavidyopadiśyate || \\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 \begin{testimonia}[hp01_004]
-Post-15th Century\\
+Post-15th Century:
 Haṭharatnāvali 1.3
-\startverse
-haṭhavidyāṃ hi gorakṣamatsyendrādyā vijānate |\\
-ātmārāmo 'pi jānīte śrīnivāsas tathā svayam || 
-\endverse
+
+\begin{versinnote}
+\tl{haṭhavidyāṃ hi gorakṣamatsyendrādyā vijānate |\\+}
+\tl{ātmārāmo 'pi jānīte śrīnivāsas tathā svayam || \\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Haṭharatnāvalī 1.80
-\startverse
-śrīādināthamatsyendraśābarānandabhairavāḥ |\\
-śāraṅgīmīnagorakṣavirūpākṣabileśayāḥ || 
-\endverse
+
+\begin{versinnote}
+\tl{śrīādināthamatsyendraśābarānandabhairavāḥ |\\+}
+\tl{śāraṅgīmīnagorakṣavirūpākṣabileśayāḥ || \\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Haṭharatnāvalī 1.81
-\startverse
-manthānabhairavo yogī siddhabuddhaś ca kandalī |\\
-korandakaḥ surānandaḥ siddhipādaś ca carpaṭī ||
-\endverse
+
+\begin{versinnote}
+\tl{manthānabhairavo yogī siddhabuddhaś ca kandalī |\\+}
+\tl{korandakaḥ surānandaḥ siddhipādaś ca carpaṭī ||\\!}
+\end{versinnote}
 
 Caturbhuja Misra's Mugdhāvabodhinī (1.7.8) on the Rasahṛdayatantra
-\startverse
-manthānabhairavo yogī siddhabuddhaś ca kanthaḍī |\\
-koraṇṭakaḥ surānandaḥ siddhapādaś ca carpaṭī ||
-\endverse
+
+\begin{versinnote}
+\tl{manthānabhairavo yogī siddhabuddhaś ca kanthaḍī |\\+}
+\tl{koraṇṭakaḥ surānandaḥ siddhapādaś ca carpaṭī ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Haṭharatnāvalī 1.82
-\startverse
-karoṭiḥ pūjyapādaś ca nityanātho nirañjanaḥ |\\
-kapālī bindunāthaś ca kākacaṇḍīśvarāhvayaḥ || 
-\endverse
+
+\begin{versinnote}
+\tl{karoṭiḥ pūjyapādaś ca nityanātho nirañjanaḥ |\\+}
+\tl{kapālī bindunāthaś ca kākacaṇḍīśvarāhvayaḥ || \\!}
+\end{versinnote}
 
 Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
-​​\startverse
-kaṇerī pūjyapādaśca nityanātho nirañjanaḥ |\\
-kapālī bindunāthaśca kākacaṇḍīśvaro gajaḥ |
-\endverse
+​​
+\begin{versinnote}
+\tl{kaṇerī pūjyapādaśca nityanātho nirañjanaḥ |\\+}
+\tl{kapālī bindunāthaśca kākacaṇḍīśvaro gajaḥ |\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Haṭharatnāvalī 1.83
-\startverse
-allamaḥ prabhudevaś ca naiṭacūṭiś ca ṭiṇṭiṇiḥ |\\
-bhālukir nāgabodhaś ca khaṇḍakāpālikas tathā || 
-\endverse
+
+\begin{versinnote}
+\tl{allamaḥ prabhudevaś ca naiṭacūṭiś ca ṭiṇṭiṇiḥ |\\+}
+\tl{bhālukir nāgabodhaś ca khaṇḍakāpālikas tathā || \\!}
+\end{versinnote}
 
 Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
-\startverse
-āllamaḥ prabhudevaś ca ghoḍācolī ca ṭhiṇṭhinī |\\
-bhālukir nāgadevaś ca khaṇḍī kāpālikas tathā ||
-\endverse
+
+\begin{versinnote}
+\tl{āllamaḥ prabhudevaś ca ghoḍācolī ca ṭhiṇṭhinī |\\+}
+\tl{bhālukir nāgadevaś ca khaṇḍī kāpālikas tathā ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Haṭharatnāvalī 1.84
-\startverse
-ityādayo mahāsiddhāḥ haṭhayogaprasādataḥ |\\
-khaṇḍayitvā kāladaṇḍaṃ brahmāṇḍe vicaranti te ||
-\endverse
+
+\begin{versinnote}
+\tl{ityādayo mahāsiddhāḥ haṭhayogaprasādataḥ |\\+}
+\tl{khaṇḍayitvā kāladaṇḍaṃ brahmāṇḍe vicaranti te ||\\!}
+\end{versinnote}
 
 Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
-\startverse
-ityādayo mahāsiddhā rasabhogaprasādataḥ |\\
-khaṇḍayitvā kāladaṇḍaṃ trilokyāṃ vicaranti te |
-\endverse
+
+\begin{versinnote}
+\tl{ityādayo mahāsiddhā rasabhogaprasādataḥ |\\+}
+\tl{khaṇḍayitvā kāladaṇḍaṃ trilokyāṃ vicaranti te |\\!}
+\end{versinnote}
 
 Haṭhatattvakaumudī 17.24
-\startverse
-ūrdhvaṃretaḥprabhāvena sanakādyā maharṣayaḥ |\\
-khaṇḍayitvā kāladaṇḍaṃ yathecchaṃ viharanti te || 24 ||
-\endverse
+
+\begin{versinnote}
+\tl{ūrdhvaṃretaḥprabhāvena sanakādyā maharṣayaḥ |\\+}
+\tl{khaṇḍayitvā kāladaṇḍaṃ yathecchaṃ viharanti te || 24 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogasārasaṃgraha (YSS), p.53.
-\startverse
-saṃsāratāpataptānāṃ samāśrayahaṭho haṭhaḥ |\\
-aśeṣayogajagatām ādhārakamaṭho haṭhaḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{saṃsāratāpataptānāṃ samāśrayahaṭho haṭhaḥ |\\+}
+\tl{aśeṣayogajagatām ādhārakamaṭho haṭhaḥ ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Śivasaṃhitā
-\startverse
-haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\% (ab missing in some ŚS witnesses)
-bhaved vīryavatī guptā nirvīryā ca prakāśitā || 5.254
-\endverse
+
+\begin{versinnote}
+\tl{haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\% (ab missing in some ŚS witnesses)
+bhaved vīryavatī guptā nirvīryā ca prakāśitā || 5.254\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-tathā haṭhapradīpikāyām—\\
-haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\
-bhaved vīryavatī guptā nirvīryā tu prakāśiteti ||
-\endverse
+
+\begin{versinnote}
+\tl{tathā haṭhapradīpikāyām—\\+}
+\tl{haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\+}
+\tl{bhaved vīryavatī guptā nirvīryā tu prakāśiteti ||\\!}
+\end{versinnote}
 
@@
 Haṭharatnāvalī 1.66
-\startverse
-surāṣṭre dhārmike deśe subhikṣe nirupadrave |\\
-ekāntamaṭhikāmadhye sthātavyaṃ haṭhayoginā ||
-\endverse
-
-Yogacintāmaṇi
-\startverse
-haṭhapradīpikāyām—\\
-surājye dhārmike deśe subhikṣe nirupadrave |\\
-ekānte maṭhikāmadhye sthātavyaṃ haṭhayoginā ||
-\endverse
+
+\begin{versinnote}
+\tl{surāṣṭre dhārmike deśe subhikṣe nirupadrave |\\+}
+\tl{ekāntamaṭhikāmadhye sthātavyaṃ haṭhayoginā ||\\!}
+\end{versinnote}
+
+Yogacintāmaṇi
+
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{surājye dhārmike deśe subhikṣe nirupadrave |\\+}
+\tl{ekānte maṭhikāmadhye sthātavyaṃ haṭhayoginā ||\\!}
+\end{versinnote}
 
@@
 Cf. Dattātreyayogaśāstra
-\startverse
-suśobhanaṃ maṭhaṃ kuryāt sūkṣmadvāraṃ tu nirvraṇaṃ || 54 ||\\
-suṣṭhu liptaṃ gomayena sudhayā vā prayatnataḥ |\\
-matkuṇair maśakair bhūtair varjitaṃ ca prayatnataḥ || 55 ||\\
-dine dine susammṛṣṭaṃ sammārjanyā hy atandritaḥ |\\
-vāsitaṃ ca sugandhena dhūpitaṃ guggulādibhiḥ || 56 ||\\
-malamūtrādibhir vargair aṣṭādaśabhir eva ca |\\
-varjitaṃ dvārasampannaṃ vastrāvaraṇam eva vā || 57 ||
-\endverse
+
+\begin{versinnote}
+\tl{suśobhanaṃ maṭhaṃ kuryāt sūkṣmadvāraṃ tu nirvraṇaṃ || 54 ||\\+}
+\tl{suṣṭhu liptaṃ gomayena sudhayā vā prayatnataḥ |\\+}
+\tl{matkuṇair maśakair bhūtair varjitaṃ ca prayatnataḥ || 55 ||\\+}
+\tl{dine dine susammṛṣṭaṃ sammārjanyā hy atandritaḥ |\\+}
+\tl{vāsitaṃ ca sugandhena dhūpitaṃ guggulādibhiḥ || 56 ||\\+}
+\tl{malamūtrādibhir vargair aṣṭādaśabhir eva ca |\\+}
+\tl{varjitaṃ dvārasampannaṃ vastrāvaraṇam eva vā || 57 ||\\!}
+\end{versinnote}
 
@@
 Yogacintāmaṇi
-\startverse
-alpadvāram arandhragartaghaṭitaṃ nāpy uccanīcāyitam |\\
-samyaggomayasāndraliptavimalaṃ niḥśeṣajantūjjhitam |\\
-bāhye maṇḍapakūpavediracitaṃ prākārasaṃveṣṭitam |\\
-proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{alpadvāram arandhragartaghaṭitaṃ nāpy uccanīcāyitam |\\+}
+\tl{samyaggomayasāndraliptavimalaṃ niḥśeṣajantūjjhitam |\\+}
+\tl{bāhye maṇḍapakūpavediracitaṃ prākārasaṃveṣṭitam |\\+}
+\tl{proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 1.67
-\startverse
-alpadvāram arandhragartapiṭharaṃ nātyuccanīcāyataṃ\\
-samyaggomayasāndraliptavimalaṃ niḥśeṣabādhojjhitaṃ |\\
-bāhye maṇḍapavedikūparuciraṃ prākārasaṃveṣṭitam\\
-proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{alpadvāram arandhragartapiṭharaṃ nātyuccanīcāyataṃ\\+}
+\tl{samyaggomayasāndraliptavimalaṃ niḥśeṣabādhojjhitaṃ |\\+}
+\tl{bāhye maṇḍapavedikūparuciraṃ prākārasaṃveṣṭitam\\+}
+\tl{proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||\\!}
+\end{versinnote}
 
@@
 Cf. Amanaska 2.15
-\startverse
-evaṃvidhaṃ guruṃ labdhvā sarvacintāvivarjitaḥ\\
-sthitvā manohare deśe yogam eva samabhyaset
-\endverse
+
+\begin{versinnote}
+\tl{evaṃvidhaṃ guruṃ labdhvā sarvacintāvivarjitaḥ\\+}
+\tl{sthitvā manohare deśe yogam eva samabhyaset\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\
-gurūpadiṣṭamārgeṇa yogam eva sadābhyased iti ||
-\endverse
+
+\begin{versinnote}
+\tl{evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\+}
+\tl{gurūpadiṣṭamārgeṇa yogam eva sadābhyased iti ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 1.68
-\startverse
-evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\
-gurūpadiṣṭamārgeṇa yogam eva sadābhyaset ||
-\endverse
+
+\begin{versinnote}
+\tl{evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\+}
+\tl{gurūpadiṣṭamārgeṇa yogam eva sadābhyaset ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\
-janasaṅgaś ca laulyaṃ ca ṣaḍbhir yogaḥ praṇaśyati ||
-\endverse
+
+\begin{versinnote}
+\tl{atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\+}
+\tl{janasaṅgaś ca laulyaṃ ca ṣaḍbhir yogaḥ praṇaśyati ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 1.77
-\startverse
-atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\
-janasaṅgaṃ ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||
-\endverse
+
+\begin{versinnote}
+\tl{atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\+}
+\tl{janasaṅgaṃ ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||\\!}
+\end{versinnote}
 
 Yuktabhavadeva 4.25 (attributed to the śivayoga)
-\startverse
-atyāhāraḥ prayāsaśca prajalpo niyamāgrahaḥ |\\
-janasaṃgaś ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||
-\endverse
+
+\begin{versinnote}
+\tl{atyāhāraḥ prayāsaśca prajalpo niyamāgrahaḥ |\\+}
+\tl{janasaṃgaś ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||\\!}
+\end{versinnote}
 
 Jyotsnā
-\startverse
-śītodakena prātaḥsnānanaktabhojanaphalāhārādirūpaniyamasya grahaṇaṃ niyamagrahaḥ |
-\endverse
+
+\begin{versinnote}
+\tl{śītodakena prātaḥsnānanaktabhojanaphalāhārādirūpaniyamasya grahaṇaṃ niyamagrahaḥ |\\!}
+\end{versinnote}
 
 Yogaprakāśikā
-\startverse
-niyamāgrahaḥ vakṣyamāṇaniyamāparipālanaṃ
-\endverse
+
+\begin{versinnote}
+\tl{niyamāgrahaḥ vakṣyamāṇaniyamāparipālanaṃ\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Dharmaputrikā 137cd–38ab%Narharināth ed.?
-\startverse
-utsāho niścayo dhairyaṃ santoṣas tattvadarśanam | \\
-kratūnāṃ copasaṃhāraḥ ṣaṭsādhanam iti smṛtam |
-\endverse
+
+\begin{versinnote}
+\tl{utsāho niścayo dhairyaṃ santoṣas tattvadarśanam |\\+}
+\tl{kratūnāṃ copasaṃhāraḥ ṣaṭsādhanam iti smṛtam |\\!}
+\end{versinnote}
 
 Śivadharmottara 10 (W 122r):%JM: what ed or ms is this?
-\startverse
-utsāhān niścayād dhairyāt santoṣāt tattvadarśanāt |\\
-muner janapadatyāgād ṣaḍbhir yogaḥ prasiddhyati |
-\endverse
+
+\begin{versinnote}
+\tl{utsāhān niścayād dhairyāt santoṣāt tattvadarśanāt |\\+}
+\tl{muner janapadatyāgād ṣaḍbhir yogaḥ prasiddhyati |\\!}
+\end{versinnote}
 
 Jñānārṇava 20.1
-\startverse
-utsāhān niścayād dhairyāt saṃtoṣāt tattvaniścayāt |\\
-muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 1 ||
-\endverse
+
+\begin{versinnote}
+\tl{utsāhān niścayād dhairyāt saṃtoṣāt tattvaniścayāt |\\+}
+\tl{muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 1 ||\\!}
+\end{versinnote}
 
 Yogabindu by Haribhadra
-\startverse
-utsāhān niścayād dhairyāt saṃtopāt tattvadarśanāt |\\
-muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 411 ||
-\endverse
+
+\begin{versinnote}
+\tl{utsāhān niścayād dhairyāt saṃtopāt tattvadarśanāt |\\+}
+\tl{muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 411 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-utsāhāt sāhasād dhairyāt tatvajñānād viniścayāt |\\
-janasaṅgaparityāgāt ṣaḍbhir yogaḥ prasidhyati ||
-\endverse
+
+\begin{versinnote}
+\tl{utsāhāt sāhasād dhairyāt tatvajñānād viniścayāt |\\+}
+\tl{janasaṅgaparityāgāt ṣaḍbhir yogaḥ prasidhyati ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 1.78:
-\startverse
-utsāhān niścayād dhairyāt tattvajñānārthadarśanāt
-[utsāhān niścalād- P,T]\\
-bindusthairyān mitāhārāj janasaṅgavivarjanāt |\\
-nidrātyāgāj jitaśvāsāt pīṭhasthairyād anālasāt\\
-gurvācāryaprasādāc ca ebhir yogas tu sidhyati || 
-\endverse
+
+\begin{versinnote}
+\tl{utsāhān niścayād dhairyāt tattvajñānārthadarśanāt
+[utsāhān niścalād- P,T]\\+}
+\tl{bindusthairyān mitāhārāj janasaṅgavivarjanāt |\\+}
+\tl{nidrātyāgāj jitaśvāsāt pīṭhasthairyād anālasāt\\+}
+\tl{gurvācāryaprasādāc ca ebhir yogas tu sidhyati || \\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-haṭhasya prathamāṅgatvād āsanaṃ pūrvam ucyate |\\
-tat kuryād āsanasthairyam ārogyaṃ cāṅgalāghavaṃ ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhasya prathamāṅgatvād āsanaṃ pūrvam ucyate |\\+}
+\tl{tat kuryād āsanasthairyam ārogyaṃ cāṅgalāghavaṃ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.5
-\startverse
-haṭhasya prathamāṅgatvād āsanaṃ dārśyate mayā |\\
-tat kuryād āsanaṃ sthairyam ārogyaṃ cāṅgapāṭavam || 
-\endverse
+
+\begin{versinnote}
+\tl{haṭhasya prathamāṅgatvād āsanaṃ dārśyate mayā |\\+}
+\tl{tat kuryād āsanaṃ sthairyam ārogyaṃ cāṅgapāṭavam || \\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-haṭhapradīpikāyām—\\
-vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ |\\
-aṅgīkṛtāny āsanāni vakṣyante kānicin mayā ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ |\\+}
+\tl{aṅgīkṛtāny āsanāni vakṣyante kānicin mayā ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.6
-\startverse
-vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ ||\\
-aṃgīkṛtāny āsanāni lakṣyante kāni cin mayā ||
-\endverse
+
+\begin{versinnote}
+\tl{vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ ||\\+}
+\tl{aṃgīkṛtāny āsanāni lakṣyante kāni cin mayā ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Śāradātilaka 25.12
-\startverse
-jānūrvor antare  samyak  kṛtvā  pādatale ubhe |\\
-ṛjukāyo viśed yogī svastikaṃ tat pracakṣate ||
-\endverse
+
+\begin{versinnote}
+\tl{jānūrvor antare  samyak  kṛtvā  pādatale ubhe |\\+}
+\tl{ṛjukāyo viśed yogī svastikaṃ tat pracakṣate ||\\!}
+\end{versinnote}
 
 Vasiṣṭhasaṃhitā 1.68
-\startverse
-jānūrvor antaraṃ samyak kṛtvā pādatale ubhe |\\
-ṛjukāyas tathāsīnaḥ svastikaṃ tat pracakṣate ||
-\endverse
+
+\begin{versinnote}
+\tl{jānūrvor antaraṃ samyak kṛtvā pādatale ubhe |\\+}
+\tl{ṛjukāyas tathāsīnaḥ svastikaṃ tat pracakṣate ||\\!}
+\end{versinnote}
 
 Yogayājñavalkya 3.3
-\startverse
-jānūrvor antare samyak kṛtvā pādatale ubhe\\
-ṛjukāyaḥ sukhāsīnaḥ svastikaṃ tat pracakṣate
-\endverse
+
+\begin{versinnote}
+\tl{jānūrvor antare samyak kṛtvā pādatale ubhe\\+}
+\tl{ṛjukāyaḥ sukhāsīnaḥ svastikaṃ tat pracakṣate\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-Yājñavalkyaḥ—\\
-jānūrvor antare samyak kṛtvā pādatale ubhe |\\
-ṛjukāyaḥ samāsīnaḥ svastikaṃ tat pracakṣate ||
-\endverse
+
+\begin{versinnote}
+\tl{Yājñavalkyaḥ—\\+}
+\tl{jānūrvor antare samyak kṛtvā pādatale ubhe |\\+}
+\tl{ṛjukāyaḥ samāsīnaḥ svastikaṃ tat pracakṣate ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.52
-\startverse
-atha svastikāsanam---\\
-jānūrvor antaraṃ samyak kṛtvā padatale ubhe ||\\
-ṛjukāyasamāsīnaḥ svastikaṃ tat pracakṣate || 
-\endverse
+
+\begin{versinnote}
+\tl{atha svastikāsanam---\\+}
+\tl{jānūrvor antaraṃ samyak kṛtvā padatale ubhe ||\\+}
+\tl{ṛjukāyasamāsīnaḥ svastikaṃ tat pracakṣate || \\!}
+\end{versinnote}
 
@@
 Cf. Ahirbudhnyasaṃhitā 31.45cd–46
-\startverse
-ubhayor gulphayoḥ kṛtvā pṛṣṭhapārśvāv ubhāv api ||\\
-vyutkrameṇātha pāṇibhyāṃ vinyastābhyāṃ vigṛhya ca |\\
-prṣṭhagābhyāṃ padāṅguṣṭhāv etad gomukham ucyate || 
-\endverse
+
+\begin{versinnote}
+\tl{ubhayor gulphayoḥ kṛtvā pṛṣṭhapārśvāv ubhāv api ||\\+}
+\tl{vyutkrameṇātha pāṇibhyāṃ vinyastābhyāṃ vigṛhya ca |\\+}
+\tl{prṣṭhagābhyāṃ padāṅguṣṭhāv etad gomukham ucyate || \\!}
+\end{versinnote}
 
 Vasiṣṭhasaṃhitā 1.70
-\startverse
-savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet |\\
-dakṣiṇe 'pi tathā savyaṃ gomukhaṃ tat pracakṣate ||
-\endverse
+
+\begin{versinnote}
+\tl{savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet |\\+}
+\tl{dakṣiṇe 'pi tathā savyaṃ gomukhaṃ tat pracakṣate ||\\!}
+\end{versinnote}
 
 Yogayājñavalkya 3.5cd–3.6ab
-\startverse
-savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet\\
-dakṣiṇe ‘pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā
-\endverse
+
+\begin{versinnote}
+\tl{savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet\\+}
+\tl{dakṣiṇe ‘pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-savye dakṣiṇagulphaṃ tu pṛṣthapārśve niveśayet |\\
-dakṣiṇe'pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā ||
-\endverse
+
+\begin{versinnote}
+\tl{savye dakṣiṇagulphaṃ tu pṛṣthapārśve niveśayet |\\+}
+\tl{dakṣiṇe'pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.53
-\startverse
-atha gomukhāsanam---\\
-savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niyojayet ||\\
-dakṣiṇe 'pi tathā savyaṃ gomukhaṃ gomukhāsanam ||
-\endverse
+
+\begin{versinnote}
+\tl{atha gomukhāsanam---\\+}
+\tl{savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niyojayet ||\\+}
+\tl{dakṣiṇe 'pi tathā savyaṃ gomukhaṃ gomukhāsanam ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Vasiṣṭhasaṃhitā 1.72
-\startverse
-ekaṃ pādam athaikasmin vinyasyorau ca saṃsthitam |\\
-itarasmiṃs tathaivoruṃ vīrāsanam itīritam ||
-\endverse
+
+\begin{versinnote}
+\tl{ekaṃ pādam athaikasmin vinyasyorau ca saṃsthitam |\\+}
+\tl{itarasmiṃs tathaivoruṃ vīrāsanam itīritam ||\\!}
+\end{versinnote}
 
 Cf. Śāradātilakatantra 25.15cd–16ab
-\startverse
-ekaṃ pādam adhaḥ kṛtvā vinyasyorau tathetaram ||\\
-ṛjukāyo viśed yogī vīrāsanam itīritam |
-\endverse
+
+\begin{versinnote}
+\tl{ekaṃ pādam adhaḥ kṛtvā vinyasyorau tathetaram ||\\+}
+\tl{ṛjukāyo viśed yogī vīrāsanam itīritam |\\!}
+\end{versinnote}
 
 Yogayājñavalkya 3.8
-\startverse
-ekaṃ pādam athaikasmin vinyasyoruṇi saṃsthitam |\\
-itarasmiṃs tathā coruṃ vīrāsanam udāhṛtam ||
-\endverse
+
+\begin{versinnote}
+\tl{ekaṃ pādam athaikasmin vinyasyoruṇi saṃsthitam |\\+}
+\tl{itarasmiṃs tathā coruṃ vīrāsanam udāhṛtam ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-ekaṃ pādam athaikasmin vinyasyoruṇi saṃsthitaḥ |\\
-itarasmiṃs tathā coruṃ vīrāsanam udāhṛtam ||
-\endverse
+
+\begin{versinnote}
+\tl{ekaṃ pādam athaikasmin vinyasyoruṇi saṃsthitaḥ |\\+}
+\tl{itarasmiṃs tathā coruṃ vīrāsanam udāhṛtam ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.54
-\startverse
-atha vīrāsanam---\\
-ekaṃ pādam athaikasmin vinyased ūruṇi sthiram ||\\
-itarasmiṃs tathā coruṃ vīrāsanam īritam || 
-\endverse
+
+\begin{versinnote}
+\tl{atha vīrāsanam---\\+}
+\tl{ekaṃ pādam athaikasmin vinyased ūruṇi sthiram ||\\+}
+\tl{itarasmiṃs tathā coruṃ vīrāsanam īritam || \\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Vasiṣṭhasaṃhitā 1.80
-\startverse
-gudaṃ nirudhya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\
-kūrmāsanaṃ bhaved etad iti yogavido viduḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{gudaṃ nirudhya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\+}
+\tl{kūrmāsanaṃ bhaved etad iti yogavido viduḥ ||\\!}
+\end{versinnote}
 
 Cf. Ahirbudhnyasaṃhitā 31.35
-\startverse
-gudaṃ nipīḍya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\
-etat kūrmāsanaṃ proktaṃ yogasiddhikaraṃ param || 
-\endverse
+
+\begin{versinnote}
+\tl{gudaṃ nipīḍya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\+}
+\tl{etat kūrmāsanaṃ proktaṃ yogasiddhikaraṃ param || \\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-gudaṃ niyamya gulphābhyaṃ vyutkrameṇa samāhitaḥ |\\
-kūrmāsanaṃ bhaved etad iti yogavido viduḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{gudaṃ niyamya gulphābhyaṃ vyutkrameṇa samāhitaḥ |\\+}
+\tl{kūrmāsanaṃ bhaved etad iti yogavido viduḥ ||\\!}
+\end{versinnote}
 
 Yuktabhavadeva 6.15
-\startverse
-haṭhapradīpikāyām\\
-gudaṃ niyamya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\
-kūrmāsanaṃ bhavedetaditi yogavido viduḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhapradīpikāyām\\+}
+\tl{gudaṃ niyamya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\+}
+\tl{kūrmāsanaṃ bhavedetaditi yogavido viduḥ ||\\!}
+\end{versinnote}
 
@@
 Vasiṣṭhasaṃhitā 1.78
-\startverse
-padmāsanaṃ samāsthāya jānūrvor antare karau |\\
-bhūmau niveśya saṃsthāpya vyomasthaṃ kukkuṭāsanam ||\\
-{[}niveśya bhūmau – mss. la, va, śa]
-\endverse
+
+\begin{versinnote}
+\tl{padmāsanaṃ samāsthāya jānūrvor antare karau |\\+}
+\tl{bhūmau niveśya saṃsthāpya vyomasthaṃ kukkuṭāsanam ||\\+}
+\tl{{[}niveśya bhūmau – mss. la, va, śa]\\!}
+\end{versinnote}
 
 Cf. Ahirbudhnyasaṃhitā 31.38
-\startverse
-kukkuṭāsanam\\
-padmāsanam adhiṣṭhāya jānvantaraviniḥsṛtau |\\
-karau bhūmau niveśyaitad vyomasthaṃ kukkuṭāsanam || 
-\endverse
+
+\begin{versinnote}
+\tl{kukkuṭāsanam\\+}
+\tl{padmāsanam adhiṣṭhāya jānvantaraviniḥsṛtau |\\+}
+\tl{karau bhūmau niveśyaitad vyomasthaṃ kukkuṭāsanam || \\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-padmāsanaṃ tu saṃyojya jānūrvor antare karau | \\
-niveśya bhūmau saṃsthāpya vyomasthaṃ kukkuṭāsanam ||
-\endverse
+
+\begin{versinnote}
+\tl{padmāsanaṃ tu saṃyojya jānūrvor antare karau |\\+}
+\tl{niveśya bhūmau saṃsthāpya vyomasthaṃ kukkuṭāsanam ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.73
-\startverse
-atha kukkuṭāsanam---\\
-padmāsanaṃ susaṃsthāpya jānūrvor antare karau ||\\
-niveśya bhūmau saṃsthāpya vyomasthaḥ kukkuṭāsanam ||
-\endverse
+
+\begin{versinnote}
+\tl{atha kukkuṭāsanam---\\+}
+\tl{padmāsanaṃ susaṃsthāpya jānūrvor antare karau ||\\+}
+\tl{niveśya bhūmau saṃsthāpya vyomasthaḥ kukkuṭāsanam ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-kukkuṭāsanabandhastho dorbhyāṃ saṃbadhya kandharam |\\
-bhavet kūrmavad uttānam etad uttānakūrmakam ||
-\endverse
+
+\begin{versinnote}
+\tl{kukkuṭāsanabandhastho dorbhyāṃ saṃbadhya kandharam |\\+}
+\tl{bhavet kūrmavad uttānam etad uttānakūrmakam ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.74
-\startverse
-kukkuṭāsanabandhastho dorbhyāṃ sambadhya kandharām ||\\
-śete kūrmavad uttānam etad uttānakūrmakam || 74 ||
-\endverse
+
+\begin{versinnote}
+\tl{kukkuṭāsanabandhastho dorbhyāṃ sambadhya kandharām ||\\+}
+\tl{śete kūrmavad uttānam etad uttānakūrmakam || 74 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-pādāṅguṣṭhau ca pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\
-dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam |
-\endverse
+
+\begin{versinnote}
+\tl{pādāṅguṣṭhau ca pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\+}
+\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam |\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.51
-\startverse
-atha dhanurāsanam---\\
-pādāṅguṣṭhau tu pāṇibhyāṃ gṛhītvā śravaṇāvadhi ||\\
-dhanurākarṣaṇaṃ kṛtvā dhanurāsanam ucyate ||
-\endverse
+
+\begin{versinnote}
+\tl{atha dhanurāsanam---\\+}
+\tl{pādāṅguṣṭhau tu pāṇibhyāṃ gṛhītvā śravaṇāvadhi ||\\+}
+\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam ucyate ||\\!}
+\end{versinnote}
 
 Cf. Haṭhayogasaṃhitā
-\startverse
-dhanurāsanam |\\
-prasārya pādau bhuvi daṇḍarūpau \\
-karau ca pṛṣṭhe dhṛtapādayugmau |\\
-kṛtvā dhanustulyavivarttitāṅgaṃ \\
-nigadyate vai dhanurāsanaṃ tat || 25 ||
-\endverse
+
+\begin{versinnote}
+\tl{dhanurāsanam |\\+}
+\tl{prasārya pādau bhuvi daṇḍarūpau\\+}
+\tl{karau ca pṛṣṭhe dhṛtapādayugmau |\\+}
+\tl{kṛtvā dhanustulyavivarttitāṅgaṃ\\+}
+\tl{nigadyate vai dhanurāsanaṃ tat || 25 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-vāmorumūlārpitadakṣapādaṃ
-jānvor bahir veṣṭitadakṣadoṣṇā |\\
-pragṛhya tiṣṭhet parivartitāṅgaḥ
-śrīmatsyanāthoditam āsanaṃ syāt ||
-\endverse
+
+\begin{versinnote}
+\tl{vāmorumūlārpitadakṣapādaṃ
+jānvor bahir veṣṭitadakṣadoṣṇā |\\+}
+\tl{pragṛhya tiṣṭhet parivartitāṅgaḥ
+śrīmatsyanāthoditam āsanaṃ syāt ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-atha matsyendrāsanam---\\
-vāmorumūlārpitā dakṣapādo jānvor bahir veṣṭitadakṣadoṣṇā ||\\
-pragṛhya tiṣṭhet partivartitāṅgaḥ śrīmatsyanāthoditam āsanaṃ syāt || 3.57 ||
-\endverse
+
+\begin{versinnote}
+\tl{atha matsyendrāsanam---\\+}
+\tl{vāmorumūlārpitā dakṣapādo jānvor bahir veṣṭitadakṣadoṣṇā ||\\+}
+\tl{pragṛhya tiṣṭhet partivartitāṅgaḥ śrīmatsyanāthoditam āsanaṃ syāt || 3.57 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-matsyendrapīṭhaṃ jaṭharapravṛddhiṃ\\
-pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\
-abhyāsataḥ kuṇḍalinīprabodhaṃ\\
-daṇḍe sthiratvaṃ pradadāti puṃsām ||
-\endverse
+
+\begin{versinnote}
+\tl{matsyendrapīṭhaṃ jaṭharapravṛddhiṃ\\+}
+\tl{pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\+}
+\tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
+\tl{daṇḍe sthiratvaṃ pradadāti puṃsām ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.58
-\startverse
-matsyendrapīṭhaṃ jaṭharapradīptaṃ \\
-pracaṇḍarugmaṇḍalakhaṇḍanāstram ||\\
-abhyāsataḥ kuṇḍalinīprabodhaṃ \\
-daṇḍasthiratvaṃ ca dadāti puṃsām ||
-\endverse
+
+\begin{versinnote}
+\tl{matsyendrapīṭhaṃ jaṭharapradīptaṃ\\+}
+\tl{pracaṇḍarugmaṇḍalakhaṇḍanāstram ||\\+}
+\tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
+\tl{daṇḍasthiratvaṃ ca dadāti puṃsām ||\\!}
+\end{versinnote}
 
 HTK
-\startverse
-matsyendrapīṭhaṃ jaṭharapracaṇḍa-\\
-ruṅmaṇḍalakhaṇḍanakhaṇḍanāstram |\\
-abhyāsataḥ kuṇḍalinīprabodhaṃ \\
-daṇḍasthiratvaṃ ca dadāti puṃsām || 8 ||
-\endverse
+
+\begin{versinnote}
+\tl{matsyendrapīṭhaṃ jaṭharapracaṇḍa-\\+}
+\tl{ruṅmaṇḍalakhaṇḍanakhaṇḍanāstram |\\+}
+\tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
+\tl{daṇḍasthiratvaṃ ca dadāti puṃsām || 8 ||\\!}
+\end{versinnote}
 
 YBD
-\startverse
-matsyendrapīṭhaṃ jaṭharaprabuddhaṃ\\
-pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\
-abhyasataṃ kuṇḍalinīprabodhaṃ\\
-daṇḍasthiratvaṃ ca dadāti puṃsām ||
-\endverse
+
+\begin{versinnote}
+\tl{matsyendrapīṭhaṃ jaṭharaprabuddhaṃ\\+}
+\tl{pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\+}
+\tl{abhyasataṃ kuṇḍalinīprabodhaṃ\\+}
+\tl{daṇḍasthiratvaṃ ca dadāti puṃsām ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Cf. Śivasaṃhitā
-\startverse
-prasārya caraṇadvandvaṃ parasparasusaṃyutam |\\
-svapāṇibhyāṃ dṛḍhaṃ dhṛtvā jānūpari śiro nyaset || 3.108 ||
-\endverse
+
+\begin{versinnote}
+\tl{prasārya caraṇadvandvaṃ parasparasusaṃyutam |\\+}
+\tl{svapāṇibhyāṃ dṛḍhaṃ dhṛtvā jānūpari śiro nyaset || 3.108 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-prasārya pādau bhuvi daṇḍarūpau \\
-dvābhyāṃ ca pādadvitayaṃ gṛhītvā |\\
-jānūpari nyastalalāṭadeśo \\
-'bhyased idaṃ paścimatānam āhuḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{prasārya pādau bhuvi daṇḍarūpau\\+}
+\tl{dvābhyāṃ ca pādadvitayaṃ gṛhītvā |\\+}
+\tl{jānūpari nyastalalāṭadeśo\\+}
+\tl{'bhyased idaṃ paścimatānam āhuḥ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-atha paścimatānāsanam---\\
-prasārya pādau bhuvi daṇḍarūpau \\
-dorbhyāṃ padāgradvitayaṃ gṛhītvā ||\\
-jānūpari nyastalalāṭadeśo \\
-vased idaṃ paścimatānam āhuḥ || 3.66 ||
-\endverse
+
+\begin{versinnote}
+\tl{atha paścimatānāsanam---\\+}
+\tl{prasārya pādau bhuvi daṇḍarūpau\\+}
+\tl{dorbhyāṃ padāgradvitayaṃ gṛhītvā ||\\+}
+\tl{jānūpari nyastalalāṭadeśo\\+}
+\tl{vased idaṃ paścimatānam āhuḥ || 3.66 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Cf. Śivasaṃhitā
-\startverse
-Āsanāgryam idaṃ proktaṃ jaṭharānaladīpanam |\\
-dehāvasādaharaṇaṃ paścimottānasaṃjñakam || 3.109 ||
-\endverse
+
+\begin{versinnote}
+\tl{Āsanāgryam idaṃ proktaṃ jaṭharānaladīpanam |\\+}
+\tl{dehāvasādaharaṇaṃ paścimottānasaṃjñakam || 3.109 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-iti paścimatānam āsanāgryaṃ
-pavanaṃ paścimavāhinaṃ karoti |\\
-udayaṃ jaṭharānalasya kuryād
-udare kārśyam arogitāṃ ca puṃsām ||
-\endverse
+
+\begin{versinnote}
+\tl{iti paścimatānam āsanāgryaṃ
+pavanaṃ paścimavāhinaṃ karoti |\\+}
+\tl{udayaṃ jaṭharānalasya kuryād
+udare kārśyam arogitāṃ ca puṃsām ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-iti paścimatānam āsanāgryaṃ pavanaṃ paścimavāhinaṃ karoti ||\\
-udayaṃ jaṭharānalasya kuryād udare kārśyam arogatāṃ ca puṃsām || 3.67 ||
-\endverse
+
+\begin{versinnote}
+\tl{iti paścimatānam āsanāgryaṃ pavanaṃ paścimavāhinaṃ karoti ||\\+}
+\tl{udayaṃ jaṭharānalasya kuryād udare kārśyam arogatāṃ ca puṃsām || 3.67 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Vimānārcanākalpa 96
-\startverse
-karatale bhūmau saṃsthāpya kūrparau nābhipārśvayor nyasya nataśirāḥ (unnataśirāḥ) pādau ḍaṇḍavad vyomni saṃsthito mayūrāsanam iti |
-\endverse
+
+\begin{versinnote}
+\tl{karatale bhūmau saṃsthāpya kūrparau nābhipārśvayor nyasya nataśirāḥ (unnataśirāḥ) pādau ḍaṇḍavad vyomni saṃsthito mayūrāsanam iti |\\!}
+\end{versinnote}
 
 Pādmasaṃhitā yogapāda 1.21c–22d:
-\startverse
-avaṣṭabhya dharāṃ samyak talābhyāṃ hastayor dvayoḥ ||\\
-kūrparau nābhipārśve ca sthāpayitvā mayūravat |\\
-samunnamya śiraḥpādau mayūrāsanam iṣyate ||
-\endverse
+
+\begin{versinnote}
+\tl{avaṣṭabhya dharāṃ samyak talābhyāṃ hastayor dvayoḥ ||\\+}
+\tl{kūrparau nābhipārśve ca sthāpayitvā mayūravat |\\+}
+\tl{samunnamya śiraḥpādau mayūrāsanam iṣyate ||\\!}
+\end{versinnote}
 
 Ahirbudhnyasaṃhitā 31.36–37
-\startverse
-mayūrāsanam\\
-niveśya kūrparau samyaṅ nābhimaṇḍalapārśvayoḥ |\\
-avaṣṭabhya bhuvaṃ pāṇitalābhyāṃ vyomni daṇḍavat || 31--36 ||
-\endverse
+
+\begin{versinnote}
+\tl{mayūrāsanam\\+}
+\tl{niveśya kūrparau samyaṅ nābhimaṇḍalapārśvayoḥ |\\+}
+\tl{avaṣṭabhya bhuvaṃ pāṇitalābhyāṃ vyomni daṇḍavat || 31--36 ||\\!}
+\end{versinnote}
 
 Vasiṣṭhasaṃhitā 1.76–77
-\startverse
-avaṣṭabhya dharāṃ samyak talābhyāṃ ca karadvayam |\\
-hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ ||\\
-samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\
-mayūrāsanam etad dhi sarvapāpavināśanam ||
-\endverse
+
+\begin{versinnote}
+\tl{avaṣṭabhya dharāṃ samyak talābhyāṃ ca karadvayam |\\+}
+\tl{hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ ||\\+}
+\tl{samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\+}
+\tl{mayūrāsanam etad dhi sarvapāpavināśanam ||\\!}
+\end{versinnote}
 
 Yogayājñavalkya 3.15–16
-\startverse
-avaṣṭabhya dharāṃ samyak talābhyāṃ tu karadvayoḥ\\
-hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ\\
-samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\
-mayūrāsanam etat tu sarvapāpapraṇāśanam ||
-\endverse
+
+\begin{versinnote}
+\tl{avaṣṭabhya dharāṃ samyak talābhyāṃ tu karadvayoḥ\\+}
+\tl{hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ\\+}
+\tl{samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\+}
+\tl{mayūrāsanam etat tu sarvapāpapraṇāśanam ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-dharām avaṣṭabhya punaḥ karābhyāṃ
-tat kūrpare sthāpitanābhipārśvaḥ | \\
-tadāsane daṇḍavad utthitaḥ khe
-mayūram etat pravadanti santaḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{dharām avaṣṭabhya punaḥ karābhyāṃ
+tat kūrpare sthāpitanābhipārśvaḥ |\\+}
+\tl{tadāsane daṇḍavad utthitaḥ khe
+mayūram etat pravadanti santaḥ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.42
-\startverse
-atha mayūram\\
-dharām avaṣṭabhya karadvayena tatkūrpare sthāpitanābhipārśvaḥ ||\\
-uccāsano daṇḍavad utthitaḥ khe mayūram etat pravadanti pīṭham ||
-\endverse
+
+\begin{versinnote}
+\tl{atha mayūram\\+}
+\tl{dharām avaṣṭabhya karadvayena tatkūrpare sthāpitanābhipārśvaḥ ||\\+}
+\tl{uccāsano daṇḍavad utthitaḥ khe mayūram etat pravadanti pīṭham ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-harati sakalarogān āśu gulmodarādīn\\
-abhibhavati ca doṣān āsanaṃ śrīmayūram |\\
-bahukadaśanabhuktaṃ bhasma kuryād aśeṣam\\
-janayati jaṭharāgniṃ jārayet kālakūṭam ||
-\endverse
+
+\begin{versinnote}
+\tl{harati sakalarogān āśu gulmodarādīn\\+}
+\tl{abhibhavati ca doṣān āsanaṃ śrīmayūram |\\+}
+\tl{bahukadaśanabhuktaṃ bhasma kuryād aśeṣam\\+}
+\tl{janayati jaṭharāgniṃ jārayet kālakūṭam ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.43
-\startverse
-harati sakalarogān āśu gulmodarādīn\\
-abhibhavati ca doṣān āsanaṃ śrīmayūram ||\\
-bahukadaśanabhuktaṃ bhasma kuryād vicitram\\
-janayati jaṭharāgniṃ jīryate kālakūṭam ||
-\endverse
+
+\begin{versinnote}
+\tl{harati sakalarogān āśu gulmodarādīn\\+}
+\tl{abhibhavati ca doṣān āsanaṃ śrīmayūram ||\\+}
+\tl{bahukadaśanabhuktaṃ bhasma kuryād vicitram\\+}
+\tl{janayati jaṭharāgniṃ jīryate kālakūṭam ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Cf. Dattātreyayogaśāstra 24cd
-\startverse
-uttānaśavavad bhūmau śayanaṃ coktam uttamam ||
-\endverse
+
+\begin{versinnote}
+\tl{uttānaśavavad bhūmau śayanaṃ coktam uttamam ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-uttānaṃ śavavad bhūmau śavāsanam idaṃ smṛtam |\\
-śavāsanaṃ śrāntiharaṃ cittaviśrāntisādhanam ||
-\endverse
+
+\begin{versinnote}
+\tl{uttānaṃ śavavad bhūmau śavāsanam idaṃ smṛtam |\\+}
+\tl{śavāsanaṃ śrāntiharaṃ cittaviśrāntisādhanam ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.76
-\startverse
-athāntimaṃ śavāsanam\\
-prasārya hastapādau ca viśrāntyā śayanaṃ tathā ||\\
-sarvāsanaśramaharaṃ śayitaṃ tu śavāsanam || 3.76 ||
-\endverse
+
+\begin{versinnote}
+\tl{athāntimaṃ śavāsanam\\+}
+\tl{prasārya hastapādau ca viśrāntyā śayanaṃ tathā ||\\+}
+\tl{sarvāsanaśramaharaṃ śayitaṃ tu śavāsanam || 3.76 ||\\!}
+\end{versinnote}
 
 Cf. HTK
-\startverse
-śavāsanaṃ hṛtkupitavātagranthivibhedakam |\\
-sarvāsanaśrāntijit hṛtśramaghnaṃ yogi saukhyadam || 12 ||
-\endverse
+
+\begin{versinnote}
+\tl{śavāsanaṃ hṛtkupitavātagranthivibhedakam |\\+}
+\tl{sarvāsanaśrāntijit hṛtśramaghnaṃ yogi saukhyadam || 12 ||\\!}
+\end{versinnote}
 
 YBD
-\startverse
-uttānaṃ śavavad bhūmau śayanaṃ tu śavāsanam ||\\
-śavāsanaṃ śrāntiharaṃ cittaviśrāntikārakam ||\\
-iti śavāsanam || 21 ||
-\endverse
+
+\begin{versinnote}
+\tl{uttānaṃ śavavad bhūmau śayanaṃ tu śavāsanam ||\\+}
+\tl{śavāsanaṃ śrāntiharaṃ cittaviśrāntikārakam ||\\+}
+\tl{iti śavāsanam || 21 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Śivasaṃhitā 3.96
-\startverse
-caturaśīty āsanāni santi nānāvidhāni ca |\\
-tebhyaś catuṣkam ādāya mayoktāni bravīmy aham ||
-\endverse
+
+\begin{versinnote}
+\tl{caturaśīty āsanāni santi nānāvidhāni ca |\\+}
+\tl{tebhyaś catuṣkam ādāya mayoktāni bravīmy aham ||\\!}
+\end{versinnote}
 
 Cf. Dattātreyayogaśāstra 5
-\startverse
-caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\
-ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||
-\endverse
+
+\begin{versinnote}
+\tl{caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\+}
+\tl{ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||\\!}
+\end{versinnote}
 
 Cf. Vivekamārtaṇḍa 5
-\startverse
-caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\
-ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||
-\endverse
+
+\begin{versinnote}
+\tl{caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\+}
+\tl{ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-haṭhapradīpikāyām—\\
-caturaśīty āsanāni śivena kathitāni vai |\\
-tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{caturaśīty āsanāni śivena kathitāni vai |\\+}
+\tl{tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-caturaśīty āsanāni śivena kathitāni tu |\\
-tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham || 3.23 ||
-\endverse
+
+\begin{versinnote}
+\tl{caturaśīty āsanāni śivena kathitāni tu |\\+}
+\tl{tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham || 3.23 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\
-śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||
-\endverse
+
+\begin{versinnote}
+\tl{siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\+}
+\tl{śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.24
-\startverse
-siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\
-śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā ||
-\endverse
+
+\begin{versinnote}
+\tl{siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\+}
+\tl{śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Vivekamārtaṇḍa 7
-\startverse
-yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyase[n] \\
-meḍhre pādam athaikam āsyahṛdaye dhṛtvā samaṃ vigraham |\\
-sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\
-etan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||
-\endverse
+
+\begin{versinnote}
+\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyase[n]\\+}
+\tl{meḍhre pādam athaikam āsyahṛdaye dhṛtvā samaṃ vigraham |\\+}
+\tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
+\tl{etan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-pavanayogasaṃgrahe—\\
-yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\
-meḍhre pādam athaikam ekahṛdaye kṛtvā samaṃ vigraham |\\
-sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyed bhruvor antaraṃ tv\\
-etan mokṣakapāṭabhedanakaraṃ siddhāsanaṃ procyate ||
-\endverse
+
+\begin{versinnote}
+\tl{pavanayogasaṃgrahe—\\+}
+\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
+\tl{meḍhre pādam athaikam ekahṛdaye kṛtvā samaṃ vigraham |\\+}
+\tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyed bhruvor antaraṃ tv\\+}
+\tl{etan mokṣakapāṭabhedanakaraṃ siddhāsanaṃ procyate ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.25
-\startverse
-tatra siddhāsanam\\
-yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\
-meḍhre pādam athaikam eva niyataṃ kṛtvā samaṃ vigraham |\\
-sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\
-caitan mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||
-\endverse
+
+\begin{versinnote}
+\tl{tatra siddhāsanam\\+}
+\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
+\tl{meḍhre pādam athaikam eva niyataṃ kṛtvā samaṃ vigraham |\\+}
+\tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
+\tl{caitan mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Vasiṣṭhasaṃhitā 1.81
-\startverse
-meḍhrād upari nikṣipya gulphaṃ tathopari |\\
-gulphāntaraṃ vinikṣipya muktāsanam idaṃ smṛtam ||
-\endverse
+
+\begin{versinnote}
+\tl{meḍhrād upari nikṣipya gulphaṃ tathopari |\\+}
+\tl{gulphāntaraṃ vinikṣipya muktāsanam idaṃ smṛtam ||\\!}
+\end{versinnote}
 
 Yogayājñavalkya 3.15
-\startverse
-meḍhrād upari nikṣipya savyaṃ gulphaṃ tathopari |\\
-gulphāntaraṃ ca nikṣipya muktāsanam idaṃ tu vā ||
-\endverse
+
+\begin{versinnote}
+\tl{meḍhrād upari nikṣipya savyaṃ gulphaṃ tathopari |\\+}
+\tl{gulphāntaraṃ ca nikṣipya muktāsanam idaṃ tu vā ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-tathā |\\
-meḍhrād upari vinyasya savyaṃ gulphaṃ tathopari |\\
-gulphāntaraṃ tu vinyasya siddhāsanam idaṃ bhavet ||
-\endverse
+
+\begin{versinnote}
+\tl{tathā |\\+}
+\tl{meḍhrād upari vinyasya savyaṃ gulphaṃ tathopari |\\+}
+\tl{gulphāntaraṃ tu vinyasya siddhāsanam idaṃ bhavet ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.26
-\startverse
-matāntare tu\\
-meḍhrād upari niḥkṣipya savyaṃ gulphaṃ tathopari |\\
-gulphāntaraṃ ca niḥkṣipya siddhāḥ siddhāsanaṃ viduḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{matāntare tu\\+}
+\tl{meḍhrād upari niḥkṣipya savyaṃ gulphaṃ tathopari |\\+}
+\tl{gulphāntaraṃ ca niḥkṣipya siddhāḥ siddhāsanaṃ viduḥ ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\
-śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||
-\endverse
+
+\begin{versinnote}
+\tl{siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\+}
+\tl{śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\
-śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā || 3.24 ||
-\endverse
+
+\begin{versinnote}
+\tl{siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\+}
+\tl{śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā || 3.24 ||\\!}
+\end{versinnote}
 
 Cf. Śivayogasāram by Kolani Ganapatideva (date 14th c.)
-\startverse
-siddāsanambunu, gondaru vajrāsanambaniyu | \\
-gondaru muktāsanambaniyu, gondadu  gulbāsanam ||
-\endverse
+
+\begin{versinnote}
+\tl{siddāsanambunu, gondaru vajrāsanambaniyu |\\+}
+\tl{gondaru muktāsanambaniyu, gondadu  gulbāsanam ||\\!}
+\end{versinnote}
 
 Cf. A verse by the poet Pingali Surana (active 16th c.)
-\startverse
-kondaru siddāsanamani \\
-kondaru vajrāsanamani koniyādudurī \\
-pondaga dīnini mariyoka \\
-kondaru guptāsamanu kondru mahātmā
-\endverse
+
+\begin{versinnote}
+\tl{kondaru siddāsanamani\\+}
+\tl{kondaru vajrāsanamani koniyādudurī\\+}
+\tl{pondaga dīnini mariyoka\\+}
+\tl{kondaru guptāsamanu kondru mahātmā\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Cf. Dattātreyayogaśāstra 33
-\startverse
-laghvāhāras tu teṣv eko mukhyo bhavati nāpare |\\
-ahiṃsā niyameṣv eko mukhyo bhavati nāpare || 33 ||
-\endverse
+
+\begin{versinnote}
+\tl{laghvāhāras tu teṣv eko mukhyo bhavati nāpare |\\+}
+\tl{ahiṃsā niyameṣv eko mukhyo bhavati nāpare || 33 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-niyameṣu mitāharo yathāhịmsā yameṣv iva |\\
-mukhyaṃ sarvāsaneṣv evaṃ siddhāsanaṃ idaṃ viduḥ |
-\endverse
+
+\begin{versinnote}
+\tl{niyameṣu mitāharo yathāhịmsā yameṣv iva |\\+}
+\tl{mukhyaṃ sarvāsaneṣv evaṃ siddhāsanaṃ idaṃ viduḥ |\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-caturaśītipīṭheṣu siddhāsanaṃ samabhyaset | \\
-dvāsaptatisahasreṣu suṣumṇām iva nāḍiṣu ||
-\endverse
+
+\begin{versinnote}
+\tl{caturaśītipīṭheṣu siddhāsanaṃ samabhyaset |\\+}
+\tl{dvāsaptatisahasreṣu suṣumṇām iva nāḍiṣu ||\\!}
+\end{versinnote}
 
 YSS
-\startverse
-caturāśītapīṭheṣu siddham eva samabhyaset |\\
-dvisaptatisahasreṣu suṣumnām iva nāḍiṣu ||
-\endverse
+
+\begin{versinnote}
+\tl{caturāśītapīṭheṣu siddham eva samabhyaset |\\+}
+\tl{dvisaptatisahasreṣu suṣumnām iva nāḍiṣu ||\\!}
+\end{versinnote}
 
 Yogacintāmaṇi (in a different passage)
-​​\startverse
-maṇḍalā dṛśyate siddhiḥ kuṇḍalyabhyāsayoginaḥ |\\
-dvisaptatisahasrāṇāṃ nāḍīnāṃ malaśodhanam ||
-\endverse
+​​
+\begin{versinnote}
+\tl{maṇḍalā dṛśyate siddhiḥ kuṇḍalyabhyāsayoginaḥ |\\+}
+\tl{dvisaptatisahasrāṇāṃ nāḍīnāṃ malaśodhanam ||\\!}
+\end{versinnote}
 
 Cf. Kumbhakapaddhati (effects of practising kumbhaka)
-\startverse
-dvāsaptati sahāsrāṇāṃ nāḍīnāṃ malaśodhanam |\\
-yatheṣṭaṃ dhāraṇaṃ vāyor vikārābhāva eva ca || 120 ||
-\endverse
+
+\begin{versinnote}
+\tl{dvāsaptati sahāsrāṇāṃ nāḍīnāṃ malaśodhanam |\\+}
+\tl{yatheṣṭaṃ dhāraṇaṃ vāyor vikārābhāva eva ca || 120 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-ātmadhyāyī mitāhārī yāvad dvādaśavatsaram |\\
-sadā siddhāsanābhyāsād yogī niṣpattim āpnuyāt |\\
-śramadair bahubhiḥ pīṭhaiḥ kiṃ syāt siddhāsane sati ||
-\endverse
+
+\begin{versinnote}
+\tl{ātmadhyāyī mitāhārī yāvad dvādaśavatsaram |\\+}
+\tl{sadā siddhāsanābhyāsād yogī niṣpattim āpnuyāt |\\+}
+\tl{śramadair bahubhiḥ pīṭhaiḥ kiṃ syāt siddhāsane sati ||\\!}
+\end{versinnote}
 
 YSS
-\startverse
-ātmadhyāyo mitāhārī yāvad dvādaśavatsaram |\\
-sadā siddhāsanābhyāsād yoganiṣpattim āpnuyāt ||\\
-śramadair bahubhiḥ pīṭhair alaṃ siddhāsane sati |
-\endverse
+
+\begin{versinnote}
+\tl{ātmadhyāyo mitāhārī yāvad dvādaśavatsaram |\\+}
+\tl{sadā siddhāsanābhyāsād yoganiṣpattim āpnuyāt ||\\+}
+\tl{śramadair bahubhiḥ pīṭhair alaṃ siddhāsane sati |\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 V1 has
-\startverse
-śramādau bahubhiḥ pīṭhais sadā siddhāsane sati |\\
-prāṇānile sāvadhānaṃ baddhe kevalakumbhake || 1.41 ||
-\endverse
+
+\begin{versinnote}
+\tl{śramādau bahubhiḥ pīṭhais sadā siddhāsane sati |\\+}
+\tl{prāṇānile sāvadhānaṃ baddhe kevalakumbhake || 1.41 ||\\!}
+\end{versinnote}
 
@@
 Yogacintāmaṇi
-\startverse
-prāṇānile sāvadhāne baddhe kevalakumbhake |\\
-utpatsyate nirāyāsāt svayam evonmanī yathā ||
-\endverse
+
+\begin{versinnote}
+\tl{prāṇānile sāvadhāne baddhe kevalakumbhake |\\+}
+\tl{utpatsyate nirāyāsāt svayam evonmanī yathā ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-athaikasminn eva dṛḍhaṃ baddhe siddhāsane sadā |\\
-bandhatrayam anāyāsāt svayam evopajāyate |
-\endverse
+
+\begin{versinnote}
+\tl{athaikasminn eva dṛḍhaṃ baddhe siddhāsane sadā |\\+}
+\tl{bandhatrayam anāyāsāt svayam evopajāyate |\\!}
+\end{versinnote}
 
 YSS
-\startverse
-tathaikasminn eva baddhe siddhāsane sadā |\\
-granthitrayam anāyāsāt svayamevopabhidyate |
-\endverse
+
+\begin{versinnote}
+\tl{tathaikasminn eva baddhe siddhāsane sadā |\\+}
+\tl{granthitrayam anāyāsāt svayamevopabhidyate |\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Śivasaṃhitā 5.47
-\startverse
-nāsanaṃ siddhasadṛśaṃ na kumbhasadṛśaṃ balam |\\
-na khecarīsamā mudrā na nādasadṛśo layaḥ || 5.47 ||
-\endverse
+
+\begin{versinnote}
+\tl{nāsanaṃ siddhasadṛśaṃ na kumbhasadṛśaṃ balam |\\+}
+\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ || 5.47 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ |\\
-na khecarīsamā mudrā na nādasadṛśo layaḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ |\\+}
+\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ ||\\
-na khecarīsamā mudrā na nādasadṛśo layaḥ || 3.29 ||
-\endverse
+
+\begin{versinnote}
+\tl{nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ ||\\+}
+\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ || 3.29 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Vivekamārtaṇḍa 8
-\startverse
-vāmorūpari dakṣiṇañ ca caraṇaṃ saṃsthāpya vāmaṃ tathā\\
-yāmyorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\
-aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\
-etad vyādhivikārahāri yamināṃ padmāsanaṃ procyate || 8 ||
-\endverse
+
+\begin{versinnote}
+\tl{vāmorūpari dakṣiṇañ ca caraṇaṃ saṃsthāpya vāmaṃ tathā\\+}
+\tl{yāmyorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\+}
+\tl{aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\+}
+\tl{etad vyādhivikārahāri yamināṃ padmāsanaṃ procyate || 8 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-haṭhayoge—\\
-vāmorūpari dakṣiṇaṃ hi caraṇaṃ saṃsthāpya vāmaṃ tathā\\
-dakṣorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham | \\
-aṅguṣṭhau hṛdaye nidhāya civukaṃ nāsāgram ālokayet\\
-ekad vyādhivikāranāśanakaraṃ padmāsanaṃ procyate ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhayoge—\\+}
+\tl{vāmorūpari dakṣiṇaṃ hi caraṇaṃ saṃsthāpya vāmaṃ tathā\\+}
+\tl{dakṣorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\+}
+\tl{aṅguṣṭhau hṛdaye nidhāya civukaṃ nāsāgram ālokayet\\+}
+\tl{ekad vyādhivikāranāśanakaraṃ padmāsanaṃ procyate ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-vāmorūpari dakṣiṇaṃ ca caraṇaṃ saṃsthāpya vāmaṃ tathā\\
-yāmyorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\
-aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\
-etad vyādhivināśakāri yamināṃ padmāsanaṃ procyate || 3.34 ||
-\endverse
-\end{testimonia}
-
-\subsection*{1.45--47}
+
+\begin{versinnote}
+\tl{vāmorūpari dakṣiṇaṃ ca caraṇaṃ saṃsthāpya vāmaṃ tathā\\+}
+\tl{yāmyorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\+}
+\tl{aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\+}
+\tl{etad vyādhivināśakāri yamināṃ padmāsanaṃ procyate || 3.34 ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\subsection*{1.45--46}
 \begin{translation}[hp01_045]
 However, in another view,
+
 Having put the upturned feet carefully on the thighs and the upturned hands in the middle of the thighs, one should fix the eyes on the tip of the nose.
@@
 Dattātreyayogaśāstra 35–37
-\startverse
-uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\
-ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 35 ||\\
-nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\
-uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\
-yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\
-yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||
-\endverse
+
+\begin{versinnote}
+\tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\+}
+\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 35 ||\\+}
+\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
+\tl{uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\+}
+\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
+\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\!}
+\end{versinnote}
 
 Śivasaṃhitā 3.102–104
-\startverse
-uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ\\
-ūrumadhye tathottānau pāṇī kṛtvā tu tādṛśau 3.102\\
-nāsāgre vinyased dṛṣṭiṃ rājadantaṃ ca jihvayā\\
-uttambhya cibukaṃ vakṣe saṃsthāpya pavanaṃ śanaiḥ 3.103\\
-yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ\\
-yathāśaktyaiva paścāt tu recayed anirodhataḥ 3.104
-\endverse
+
+\begin{versinnote}
+\tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ\\+}
+\tl{ūrumadhye tathottānau pāṇī kṛtvā tu tādṛśau 3.102\\+}
+\tl{nāsāgre vinyased dṛṣṭiṃ rājadantaṃ ca jihvayā\\+}
+\tl{uttambhya cibukaṃ vakṣe saṃsthāpya pavanaṃ śanaiḥ 3.103\\+}
+\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ\\+}
+\tl{yathāśaktyaiva paścāt tu recayed anirodhataḥ 3.104\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-dattātreyaḥ—\\
-uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\
-ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || = 1.45\\
-nāsāgre vinyased rājadantamūlaṃ tu jihvayā |\\
-uttabhya civukaṃ vakṣasy utthāpya pavanaṃ śanaiḥ || = 1.46\\
-yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\
-yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{dattātreyaḥ—\\+}
+\tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\+}
+\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || = 1.45\\+}
+\tl{nāsāgre vinyased rājadantamūlaṃ tu jihvayā |\\+}
+\tl{uttabhya civukaṃ vakṣasy utthāpya pavanaṃ śanaiḥ || = 1.46\\+}
+\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
+\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-dattātreyo 'pi\\
-uttānau caraṇau kṛtvā ūrvoḥ saṃsthāpya yatnataḥ |\\
-ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 3.36 ||\\
-nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\
-uttabhya cibukaṃ vakṣaḥ saṃsthāpya pavanaṃ śanaiḥ || 3.37 ||
-\endverse
+
+\begin{versinnote}
+\tl{dattātreyo 'pi\\+}
+\tl{uttānau caraṇau kṛtvā ūrvoḥ saṃsthāpya yatnataḥ |\\+}
+\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 3.36 ||\\+}
+\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
+\tl{uttabhya cibukaṃ vakṣaḥ saṃsthāpya pavanaṃ śanaiḥ || 3.37 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 
-\startverse
-nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\
-uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\
-yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\
-yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\
-idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |
-\endverse
+\begin{versinnote}
+\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
+\tl{uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\+}
+\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
+\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\+}
+\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\!}
+\end{versinnote}
 
@@
 
+\subsection*{1.46}
+\begin{translation}[hp01_046]
+See 1.45.
+\end{translation}
+
+\begin{sources}[hp01_046]
+See 1.45.
+\end{sources}
+
+\begin{testimonia}[hp01_046]
+See 1.45.
+\end{testimonia}
+
+\begin{philcomm}[hp01_046]
+See 1.45.
+\end{philcomm}
+
 \subsection*{1.47}
@@
 DYŚ
-\startverse
-idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\
-durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 38 ||
-\endverse
+
+\begin{versinnote}
+\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
+\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 38 ||\\!}
+\end{versinnote}
 
 Śivasaṃhitā
-\startverse
-idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam\\
-durlabhaṃ yena kenāpi dhīmatā labhyate param 3.105
-\endverse
+
+\begin{versinnote}
+\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam\\+}
+\tl{durlabhaṃ yena kenāpi dhīmatā labhyate param 3.105\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\
-durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi ||
-\endverse
+
+\begin{versinnote}
+\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
+\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\
-durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 3.38 ||
-\endverse
+
+\begin{versinnote}
+\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
+\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 3.38 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 VM 36
-\startverse
-kṛtvā saṃpuṭitau karau dṛḍhataraṃ baddhvātha padmāsanaṃ\\
-gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyānaṃś ca tac cetasi |\\
-vāraṃ vāram apānam ūrdhvam anilaṃ proccālayan pūritaṃ\\
-muṇcan prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{kṛtvā saṃpuṭitau karau dṛḍhataraṃ baddhvātha padmāsanaṃ\\+}
+\tl{gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyānaṃś ca tac cetasi |\\+}
+\tl{vāraṃ vāram apānam ūrdhvam anilaṃ proccālayan pūritaṃ\\+}
+\tl{muṇcan prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-tathāca granthāntare—\\
-kṛtvā saṃpuṭitau karau dṛḍhataraṃ badhvā ca padmāsanam\\
-gāḍhaṃ vakṣasi saṃnidhāya civukaṃ dhyānaṃ ca tac cetasi |\\
-vāraṃ vāram apānam ūrdhvam anilaṃ protsārayet pūrayet\\
-prāṇaṃ muñcati bodham eti niyataṃ śaktiprabodhodayāt ||
-\endverse
+
+\begin{versinnote}
+\tl{tathāca granthāntare—\\+}
+\tl{kṛtvā saṃpuṭitau karau dṛḍhataraṃ badhvā ca padmāsanam\\+}
+\tl{gāḍhaṃ vakṣasi saṃnidhāya civukaṃ dhyānaṃ ca tac cetasi |\\+}
+\tl{vāraṃ vāram apānam ūrdhvam anilaṃ protsārayet pūrayet\\+}
+\tl{prāṇaṃ muñcati bodham eti niyataṃ śaktiprabodhodayāt ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-kṛtvā saṃpuṭitau karau dṛḍhataraṃ baddhvā tu padmāsanam\\
-gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyānaṃ ca tac cetasi |\\
-vāraṃ vāram apānam ūrdhvam anilaṃ proccārayet pūritam\\
-muñcat prāṇam upaiti bodham atulaṃ śakteḥ prabhāvān naraḥ || 3.39 ||
-\endverse
+
+\begin{versinnote}
+\tl{kṛtvā saṃpuṭitau karau dṛḍhataraṃ baddhvā tu padmāsanam\\+}
+\tl{gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyānaṃ ca tac cetasi |\\+}
+\tl{vāraṃ vāram apānam ūrdhvam anilaṃ proccārayet pūritam\\+}
+\tl{muñcat prāṇam upaiti bodham atulaṃ śakteḥ prabhāvān naraḥ || 3.39 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\
-mārutaṃ dhārayed yas tu sa mukto nātra saṃśayaḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\+}
+\tl{mārutaṃ dhārayed yas tu sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī 3.40
-\startverse
-padmāsane sthito yogī nāḍīdvāreṣu pūrayet |\\
-pūritaṃ dhrīyate yas tu sa mukto nātra saṃśayaḥ || 3.40 ||
-\endverse
+
+\begin{versinnote}
+\tl{padmāsane sthito yogī nāḍīdvāreṣu pūrayet |\\+}
+\tl{pūritaṃ dhrīyate yas tu sa mukto nātra saṃśayaḥ || 3.40 ||\\!}
+\end{versinnote}
 
 Dhyānabindu Up
-\startverse
-padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\
-mārutaṃ kumbhayan yas tu sa mukto nātra saṃśayaḥ || 70 ||
-\endverse
+
+\begin{versinnote}
+\tl{padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\+}
+\tl{mārutaṃ kumbhayan yas tu sa mukto nātra saṃśayaḥ || 70 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 \begin{translation}[hp01_050]
-Now the lion’s pose
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
-Having placed the left ankle on right, the right ankle on the left and both hands on the knees and spread his fingers, with his mouth open, he stares at the tip of the nose.
-This is the lion’s pose, which is always worshipped by yogis.
-It causes the three locks to arise together and is the best of [all] postures.
+Now the lion’s pose.
+
+[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. Having placed the left ankle on right, the right ankle on the left ...
+\end{translation}
+
+\begin{translation}[hp01_051]
+... and both hands on the knees and spread his fingers, with his mouth open, he stares at the tip of the nose.
+\end{translation}
+
+\begin{translation}[hp01_052]
+This is the lion’s pose, which is always worshipped by yogis. It causes the three locks to arise together and is the best of [all] postures.
 \end{translation}
@@
 Vasiṣṭhasaṃhitā 1.73
-\startverse
-gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
-dakṣiṇaṃ savyagulphena dakṣiṇenetaretaram ||\\
-hastau jānau ca saṃsthāpya svāṅgulīś ca prasārya ca |\\
-vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ ||\\
-siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā ||
-\endverse
+
+\begin{versinnote}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
+\tl{dakṣiṇaṃ savyagulphena dakṣiṇenetaretaram ||\\+}
+\tl{hastau jānau ca saṃsthāpya svāṅgulīś ca prasārya ca |\\+}
+\tl{vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ ||\\+}
+\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā ||\\!}
+\end{versinnote}
 
 Yogayājñavalkya 3.9
-\startverse
-gulpau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\
-dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram ||\\
-hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca\\
-vyāttavaktro nirīkṣet nāsagraṃ susamāhitaḥ ||\\
-siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā
-\endverse
+
+\begin{versinnote}
+\tl{gulpau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\+}
+\tl{dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram ||\\+}
+\tl{hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca\\+}
+\tl{vyāttavaktro nirīkṣet nāsagraṃ susamāhitaḥ ||\\+}
+\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā\\!}
+\end{versinnote}
 
 Sūtasaṃhitā 15.7
-\startverse
-gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
-dakṣiṇaṃ savyagulphena vāmaṃ dakṣiṇagulphataḥ || 7 ||\\
-hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca |\\
-nāsāgraṃ ca nirīkṣeta bhavet siṃhāsanaṃ hi tat || 8 ||
-\endverse
+
+\begin{versinnote}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
+\tl{dakṣiṇaṃ savyagulphena vāmaṃ dakṣiṇagulphataḥ || 7 ||\\+}
+\tl{hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca |\\+}
+\tl{nāsāgraṃ ca nirīkṣeta bhavet siṃhāsanaṃ hi tat || 8 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi (citing Yogayājñavalkya)
-\startverse
-gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
-dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram || \\
-hastau jānūpari sthāpya svāṅgulīḥ saṃprasārya ca |\\
-vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ | \\
-siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā |
-\endverse
+
+\begin{versinnote}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
+\tl{dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram ||\\+}
+\tl{hastau jānūpari sthāpya svāṅgulīḥ saṃprasārya ca |\\+}
+\tl{vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ |\\+}
+\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā |\\!}
+\end{versinnote}
 
@@
 Haṭharatnāvalī
-\startverse
-atha siṃhāsanam\\
-gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
-dakṣiṇe savyagulphaṃ ca dakṣiṇe tu tathetaram || 3.31 ||\\
-hastau tu jānvoḥ saṃsthāpya svāṅgulīḥ samprasārya ca ||\\
-vyāttavaktro nirīkṣeta nāsāgraṃ tu samāhitaḥ || 3.32 ||\\
-siṃhāsanaṃ bhaved etat sevitaṃ yogibhiḥ sadā ||\\
-bandhatritayasaṃsthānaṃ kurute cāsanottamam || 3.33 ||
-\endverse
-\end{testimonia}
-
-\begin{philcomm}[hp01_050--052]
+
+\begin{versinnote}
+\tl{atha siṃhāsanam\\+}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
+\tl{dakṣiṇe savyagulphaṃ ca dakṣiṇe tu tathetaram || 3.31 ||\\+}
+\tl{hastau tu jānvoḥ saṃsthāpya svāṅgulīḥ samprasārya ca ||\\+}
+\tl{vyāttavaktro nirīkṣeta nāsāgraṃ tu samāhitaḥ || 3.32 ||\\+}
+\tl{siṃhāsanaṃ bhaved etat sevitaṃ yogibhiḥ sadā ||\\+}
+\tl{bandhatritayasaṃsthānaṃ kurute cāsanottamam || 3.33 ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp01_051]
 % comment on nyastalocanaḥ and susamāhitaḥ.
@@
 a similar comment on siddhāsana in 1.42).
+
 Spreading the fingers and keeping the mouth open seem to be imitating the lion, and this is depicted in some iconography of yoganarasiṃha (for example, Yoga Narasimha, Vishnu's Man-Lion Incarnation, Samuel Eilenberg Collection, Bequest of Samuel Eilenberg, 1998, Accession Number: 2000.284.4.
@@
 Now, the friendly pose.
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
-Having firmly and motionlessly held the sides of the feet with the hands, this is bhadrāsana, which cures all diseases and poisons.
-Yogis of the siddha tradition call it Gorakṣāsana.
-\end{translation}
-
-\begin{philcomm}[hp01_054]
+
+[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.  Having firmly and motionlessly held the sides of the feet with the hands,...
+\end{translation}
+
+\begin{translation}[hp01_054]
+... this is bhadrāsana, which cures all diseases and poisons. Yogis of the siddha tradition call it Gorakṣāsana.
+\end{translation}
+
+\begin{philcomm}[hp01_053]
 We have understood \emph{pārśvapāda} (i.e., side of the foot) like the compound \emph{agrapāda} (i.e., front of the foot), which can be classified as a \emph{ekadeśitatpuruṣa}.
@@
 Vasiṣṭhasaṃhitā 1.79
-\startverse
-gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipan | \\
-pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam |\\
-bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham ||
-\endverse
+
+\begin{versinnote}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipan |\\+}
+\tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam |\\+}
+\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham ||\\!}
+\end{versinnote}
 
 Yogayājñavalkya 3.11cd--3.12ab
-\startverse
-gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\
-pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam\\
-bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham
-\endverse
+
+\begin{versinnote}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\+}
+\tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam\\+}
+\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi (citing Yājñavalkya)
-\startverse
-gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\
-pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ badhvā suniścalaḥ |\\
-bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham |
-\endverse
+
+\begin{versinnote}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
+\tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ badhvā suniścalaḥ |\\+}
+\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham |\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-atha bhadrāsanam\\
-gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet ||\\
-pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam ||\\
-bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham || 3.30 ||
-\endverse
+
+\begin{versinnote}
+\tl{atha bhadrāsanam\\+}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet ||\\+}
+\tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam ||\\+}
+\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham || 3.30 ||\\!}
+\end{versinnote}
 
 YSS
-\startverse
-gulphau ca vṛṣaṇasyāyasvīvinyāḥ pārśvayoḥ kṣipet |\\
-pārśvapāde ca pāṇibhyāṃ dṛḍhaṃ badhvāsu niścalaḥ ||\\
-bhadrāsanaṃ bhaved etat sarvavyādhivināśanam |\\
-gorakṣāsanam ity āhur idaṃ vai siddhayoginaḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{gulphau ca vṛṣaṇasyāyasvīvinyāḥ pārśvayoḥ kṣipet |\\+}
+\tl{pārśvapāde ca pāṇibhyāṃ dṛḍhaṃ badhvāsu niścalaḥ ||\\+}
+\tl{bhadrāsanaṃ bhaved etat sarvavyādhivināśanam |\\+}
+\tl{gorakṣāsanam ity āhur idaṃ vai siddhayoginaḥ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\subsection*{1.54}
+\begin{sources}[hp01_054]
+See 1.53.
+\end{sources}
+\begin{testimonia}[hp01_054]
+See 1.53.
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-evam āsanabandheṣu yogīndro vijitaśramaḥ |\\
-abhyasen nāḍiśuddhiṃ ca mudrayā pavanakriyām || iti ||
-\endverse
+
+\begin{versinnote}
+\tl{evam āsanabandheṣu yogīndro vijitaśramaḥ |\\+}
+\tl{abhyasen nāḍiśuddhiṃ ca mudrayā pavanakriyām || iti ||\\!}
+\end{versinnote}
 
 HSC
-\startverse
-evam āsanabandhastho yogīndro vigataśramaḥ |\\
-athābhyasen nāḍiśuddhiṃ mudrādipavanakriyām || 73 ||
-\endverse
+
+\begin{versinnote}
+\tl{evam āsanabandhastho yogīndro vigataśramaḥ |\\+}
+\tl{athābhyasen nāḍiśuddhiṃ mudrādipavanakriyām || 73 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Dattātreyayogaśāstra 42cd--47
-\startverse
-kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet |\\
-na śāstrapāṭhamātreṇa kā cit siddhiḥ prajāyate |\\
-na veṣadhāraṇaṃ siddheḥ kāraṇaṃ na ca tatkathā |\\
-kriyaivakāraṇaṃ siddheḥ satyam eva tu sāṃkṛte || 46 ||\\
-śiśnodarārthaṃ yogasya kathayā veṣadhāriṇaḥ |\\
-anuṣṭhānavihīnās tu vañcayanti janān kila || 47 ||
-\endverse
+
+\begin{versinnote}
+\tl{kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet |\\+}
+\tl{na śāstrapāṭhamātreṇa kā cit siddhiḥ prajāyate |\\+}
+\tl{na veṣadhāraṇaṃ siddheḥ kāraṇaṃ na ca tatkathā |\\+}
+\tl{kriyaivakāraṇaṃ siddheḥ satyam eva tu sāṃkṛte || 46 ||\\+}
+\tl{śiśnodarārthaṃ yogasya kathayā veṣadhāriṇaḥ |\\+}
+\tl{anuṣṭhānavihīnās tu vañcayanti janān kila || 47 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Vārāhītantra
-\startverse
-mayi bodhībudho svasthe tucho yaṃ viśvabudbudaḥ |\\
-malīna udito vetti vikalpāvasaraḥ kutaḥ |
-\endverse
+
+\begin{versinnote}
+\tl{mayi bodhībudho svasthe tucho yaṃ viśvabudbudaḥ |\\+}
+\tl{malīna udito vetti vikalpāvasaraḥ kutaḥ |\\!}
+\end{versinnote}
 
 Siddhāntamuktāvalī
-\startverse
-śiśnodararatāya hi na deyaṃ veṣadhāriṇe ||\\
-mayi bodhyaṃ buddhau svacche tad dheyaṃ viśvabudbudam || 7 ||
-\endverse
+
+\begin{versinnote}
+\tl{śiśnodararatāya hi na deyaṃ veṣadhāriṇe ||\\+}
+\tl{mayi bodhyaṃ buddhau svacche tad dheyaṃ viśvabudbudam || 7 ||\\!}
+\end{versinnote}
 
 Yogaprakāśikā
-\startverse
-“śiśnodararatāyaitan na deya" etat yogajñānam etena śiśnodararatas tyājyo nanv etanmate tyājyapadārtho 'prasiddha iti śaṃkāṃ nirasyati mayi iti svacche bodhasvarūpasamudre budbudatulyasya viśvasya heyatvād iti bhāvaḥ
-\endverse
+
+\begin{versinnote}
+\tl{“śiśnodararatāyaitan na deya" etat yogajñānam etena śiśnodararatas tyājyo nanv etanmate tyājyapadārtho 'prasiddha iti śaṃkāṃ nirasyati mayi iti svacche bodhasvarūpasamudre budbudatulyasya viśvasya heyatvād iti bhāvaḥ\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-haṭhapradīpikāyām--\\
-śrutapratītiḥ svagurupratītiḥ, svātmapratītir manaso nirodhaḥ |\\
-etāni sarvāṇi samuccitāni, matāni dhīrair iha sādhanāni ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhapradīpikāyām--\\+}
+\tl{śrutapratītiḥ svagurupratītiḥ, svātmapratītir manaso nirodhaḥ |\\+}
+\tl{etāni sarvāṇi samuccitāni, matāni dhīrair iha sādhanāni ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-haṭhapradīpikāyām--\\
-āsanaṃ kumbhakaṃ citraṃ mudrākhyaṃ karaṇaṃ tathā |\\
-atha nādānusandhānam abhyāsānukrameṇa ca ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhapradīpikāyām--\\+}
+\tl{āsanaṃ kumbhakaṃ citraṃ mudrākhyaṃ karaṇaṃ tathā |\\+}
+\tl{atha nādānusandhānam abhyāsānukrameṇa ca ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yes, V19 reading probably best.
-\startverse
-āsanaṃ kumbhakaś citraṃ mudrākhyaṃ karaṇaṃ tathā |\\
-atha nādānusandhānam abhyāsānukramo haṭhe ||
-\endverse
+
+\begin{versinnote}
+\tl{āsanaṃ kumbhakaś citraṃ mudrākhyaṃ karaṇaṃ tathā |\\+}
+\tl{atha nādānusandhānam abhyāsānukramo haṭhe ||\\!}
+\end{versinnote}
 
@@
 Vivekamārtaṇḍa 37
-\startverse
-brahmacārī mitāhārī yogī yogaparāyaṇaḥ |\\
-abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||
-\endverse
+
+\begin{versinnote}
+\tl{brahmacārī mitāhārī yogī yogaparāyaṇaḥ |\\+}
+\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ |\\
-abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||
-\endverse
+
+\begin{versinnote}
+\tl{brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ |\\+}
+\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ ||\\
-abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā || 3.28 ||
-\endverse
+
+\begin{versinnote}
+\tl{brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ ||\\+}
+\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā || 3.28 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Gorakṣaśataka (original)
-\startverse
-susnigdhamadhurāhāraś caturthāṃśavivarjitaḥ |\\
-bhujyate śivasaṃprītyai mitāhāraḥ sa ucyate ||
-\endverse
+
+\begin{versinnote}
+\tl{susnigdhamadhurāhāraś caturthāṃśavivarjitaḥ |\\+}
+\tl{bhujyate śivasaṃprītyai mitāhāraḥ sa ucyate ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yuktabhavadeva
-\startverse
-tad uktaṃ haṭhapradīpikāyām-\\
-susnigdhamadhurāhārāś caturthāṃśavivarjitaḥ ||\\
-bhujyate śivasamprītyai mitāhāraḥ sa ucyate || 16 ||
-\endverse
+
+\begin{versinnote}
+\tl{tad uktaṃ haṭhapradīpikāyām-\\+}
+\tl{susnigdhamadhurāhārāś caturthāṃśavivarjitaḥ ||\\+}
+\tl{bhujyate śivasamprītyai mitāhāraḥ sa ucyate || 16 ||\\!}
+\end{versinnote}
 
 Yogacūḍāmaṇyupaniṣat
-\startverse
-susnigdhamadhurāhāraścaturthāṃśavivarjitaḥ |\\
-bhuñjate śivasaṃprītyā mitāhārī sa ucyate || 43 ||
-\endverse
+
+\begin{versinnote}
+\tl{susnigdhamadhurāhāraścaturthāṃśavivarjitaḥ |\\+}
+\tl{bhuñjate śivasaṃprītyā mitāhārī sa ucyate || 43 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-haṭhapradīpikāyām--\\
-kaṭvamlatīkṣṇalavaṇoṣṇaharītaśāka-\\
-sauvīratailatilasarṣapamatsyamadyam |\\
-ajādimāṃsadadhitakrakulatthakola-\\
-piṇyākahiṅgulaśunādyam apathyam āhuḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhapradīpikāyām--\\+}
+\tl{kaṭvamlatīkṣṇalavaṇoṣṇaharītaśāka-\\+}
+\tl{sauvīratailatilasarṣapamatsyamadyam |\\+}
+\tl{ajādimāṃsadadhitakrakulatthakola-\\+}
+\tl{piṇyākahiṅgulaśunādyam apathyam āhuḥ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-kaṭvamlatīkṣṇalavaṇoṣṇaharītaśākaṃ \\
-sauvīratailatilasarṣapamatsyamadyam |\\
-ajādimāṃsadadhitakrakulatthakodra-\\
-piṇyākahiṅgulaśunādyam apathyam āhuḥ || 1.72 ||
-\endverse
+
+\begin{versinnote}
+\tl{kaṭvamlatīkṣṇalavaṇoṣṇaharītaśākaṃ\\+}
+\tl{sauvīratailatilasarṣapamatsyamadyam |\\+}
+\tl{ajādimāṃsadadhitakrakulatthakodra-\\+}
+\tl{piṇyākahiṅgulaśunādyam apathyam āhuḥ || 1.72 ||\\!}
+\end{versinnote}
 
 HTK
-\startverse
-atha varjyāni –\\
-kaṭvamlatīkṣṇalavaṇoṣṇa haritaśāka-\\
-sauvīratailatilasarṣapamatsyamadyam ||\\
-ajāvimāṃsadadhitakrakulatthakola-\\
-piṇyākahiṃgulaśunādyam apathyam āhuḥ || 28 ||
-\endverse
+
+\begin{versinnote}
+\tl{atha varjyāni –\\+}
+\tl{kaṭvamlatīkṣṇalavaṇoṣṇa haritaśāka-\\+}
+\tl{sauvīratailatilasarṣapamatsyamadyam ||\\+}
+\tl{ajāvimāṃsadadhitakrakulatthakola-\\+}
+\tl{piṇyākahiṃgulaśunādyam apathyam āhuḥ || 28 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 \emph{°hari}(\emph{ī} here for metre?)\emph{taśāka°} in some nighaṇṭus is horseradish, which makes better sense here than Brahmānanda’s gloss.
-\startverse
-śigrur haritaśākaś ca śākapattraḥ supattrakaḥ | \textup{Rājanighaṇṭu 7.26}\\
-śigruko haritaśākaś ca mato vai mūlapatrakaḥ | \textup{Sauśrutanighaṇṭu 75ab}\\
-\endverse
+
+Rājanighaṇṭu 7.26
+
+\begin{versinnote}
+\tl{śigrur haritaśākaś ca śākapattraḥ supattrakaḥ |\\!}
+\end{versinnote}
+
+Sauśrutanighaṇṭu 75ab
+\begin{versinnote}
+\tl{śigruko haritaśākaś ca mato vai mūlapatrakaḥ |\\!}
+\end{versinnote}
 
@@
 But according to some nighaṇṭus, \emph{sauvīra} can also mean stibnite (an ingredient in some añjana’s and medicines):
-\startverse
-añjanaṃ yāmunaṃ kṛṣṇaṃ nādeyaṃ mecakaṃ tathā \\
-srotojaṃ dṛkpradaṃ nīlaṃ sauvīraṃ ca suvīrajam ||  \textup{Rājanighaṇṭu 13.86}
-\endverse
+
+Rājanighaṇṭu 13.86
+
+\begin{versinnote}
+\tl{añjanaṃ yāmunaṃ kṛṣṇaṃ nādeyaṃ mecakaṃ tathā\\+}
+\tl{srotojaṃ dṛkpradaṃ nīlaṃ sauvīraṃ ca suvīrajam ||\\!}
+\end{versinnote}
 
@@
 Yogacintāmaṇi
-\startverse
-bhojanam ahitaṃ vidyāt punar <apy> uṣṇīkṛtaṃ tathā rukṣam |\\
-atilavaṇaṃ sapalaṃ vā prasitaṃ śākotkaṭaṃ varjyam ||
-\endverse
+
+\begin{versinnote}
+\tl{bhojanam ahitaṃ vidyāt punar <apy> uṣṇīkṛtaṃ tathā rukṣam |\\+}
+\tl{atilavaṇaṃ sapalaṃ vā prasitaṃ śākotkaṭaṃ varjyam ||\\!}
+\end{versinnote}
 
 HSC
-\startverse
-bhojanam ahitaṃ vidyāt punar uṣṇīkṛtaṃ rūkṣaṃ | \\
-atilavaṇādikayuktaṃ kadaśanaśākotkaṭaṃ duṣṭaṃ ||
-\endverse
+
+\begin{versinnote}
+\tl{bhojanam ahitaṃ vidyāt punar uṣṇīkṛtaṃ rūkṣaṃ |\\+}
+\tl{atilavaṇādikayuktaṃ kadaśanaśākotkaṭaṃ duṣṭaṃ ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Amṛtasiddhi 19.7
-\startverse
-agnisevābalāsevā pathasevā ca sarvadā |\\
-prathamābhyāsakāle tu saṃtyājyā yoginā sadā || 19.7 ||
-\endverse
+
+\begin{versinnote}
+\tl{agnisevābalāsevā pathasevā ca sarvadā |\\+}
+\tl{prathamābhyāsakāle tu saṃtyājyā yoginā sadā || 19.7 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-haṭhadīpikāyām—\\
-varjayed durjanaprītiṃ vahnistrīpathasevanam |\\
-prātaḥsnānopavāsādi kāyakleśādikaṃ tathā ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhadīpikāyām—\\+}
+\tl{varjayed durjanaprītiṃ vahnistrīpathasevanam |\\+}
+\tl{prātaḥsnānopavāsādi kāyakleśādikaṃ tathā ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-tathā ca gorakṣavacanam---\\
-varjayed durjanaprītivahnistrīpathasevanam |\\
-prātaḥsnānopavāsādi kāyakleśādikaṃ tathā || 1.73 ||
-\endverse
+
+\begin{versinnote}
+\tl{tathā ca gorakṣavacanam---\\+}
+\tl{varjayed durjanaprītivahnistrīpathasevanam |\\+}
+\tl{prātaḥsnānopavāsādi kāyakleśādikaṃ tathā || 1.73 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Yogacintāmaṇi
-\startverse
-godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ\\
-kṣīrājyamaṇḍanavanītasitāmadhūni |\\
-śuṇṭhīpaṭolakaphalādikapañcaśākaṃ\\
-mudgādi cālpam udakaṃ ca munīndrapathyam ||
-\endverse
+
+\begin{versinnote}
+\tl{godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ\\+}
+\tl{kṣīrājyamaṇḍanavanītasitāmadhūni |\\+}
+\tl{śuṇṭhīpaṭolakaphalādikapañcaśākaṃ\\+}
+\tl{mudgādi cālpam udakaṃ ca munīndrapathyam ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-godhūmaśāliyavaṣaṣṭikaśobhanānnaṃ \\
-kṣīrājyamaṇḍanavanītasitāmadhūni |\\
-śuṇṭhīpaṭolaphalapatrajapañcaśākaṃ \\
-mudgādidivyam udakaṃ ca yamīndrapathyam || 1.71 ||
-\endverse
+
+\begin{versinnote}
+\tl{godhūmaśāliyavaṣaṣṭikaśobhanānnaṃ\\+}
+\tl{kṣīrājyamaṇḍanavanītasitāmadhūni |\\+}
+\tl{śuṇṭhīpaṭolaphalapatrajapañcaśākaṃ\\+}
+\tl{mudgādidivyam udakaṃ ca yamīndrapathyam || 1.71 ||\\!}
+\end{versinnote}
 
 YBD
-\startverse
-tathā ca śivayoge-\\
-godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ \\
-kṣīrājyakhaṇḍanavanītasitāmadhūni ||\\
-śuṇṭhīpaṭolakaphalādi ca pañcaśāka-\\
-mudgādidivyam udakaṃ ca munīndrapathyam || 21 ||
-\endverse
+
+\begin{versinnote}
+\tl{tathā ca śivayoge-\\+}
+\tl{godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ\\+}
+\tl{kṣīrājyakhaṇḍanavanītasitāmadhūni ||\\+}
+\tl{śuṇṭhīpaṭolakaphalādi ca pañcaśāka-\\+}
+\tl{mudgādidivyam udakaṃ ca munīndrapathyam || 21 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 
-Wikipedia : smooth luffa = Luffa aegyptiaca (sponge gourd)\\
+Wikipedia : smooth luffa = Luffa aegyptiaca (sponge gourd)
 Sharma (Syn. Kulaka. Well known (Trichosanthas dioica Roxb.)
@@
 On \emph{pañcaśāka}, see GhS
-\startverse
-bālaśākaṃ kālaśākaṃ tathā paṭolapatrakam |\\
-pañcaśākaṃ praśaṃsīyād vāstūkaṃ hilamocikāṃ || 5.20 ||
-\endverse
+
+\begin{versinnote}
+\tl{bālaśākaṃ kālaśākaṃ tathā paṭolapatrakam |\\+}
+\tl{pañcaśākaṃ praśaṃsīyād vāstūkaṃ hilamocikāṃ || 5.20 ||\\!}
+\end{versinnote}
 
 HTK 4.26
-\startverse
-pañcaśākastu –\\
-kṣīraparṇī ca jīvantī matsyākṣī ca punarnavā \\
-meghanādaś ceti budhaiḥ pañcaśākaḥ prakīrtitaḥ || iti || 26
-\endverse
+
+\begin{versinnote}
+\tl{pañcaśākastu –\\+}
+\tl{kṣīraparṇī ca jīvantī matsyākṣī ca punarnavā\\+}
+\tl{meghanādaś ceti budhaiḥ pañcaśākaḥ prakīrtitaḥ || iti || 26\\!}
+\end{versinnote}
 
 Jyotsnā and Yuktabhavadeva 4.22
-\startverse
-sarvaśākam acākṣuṣyaṃ cākṣuṣyaṃ śākapañcakam |\\
-jīvantī-vāstu-matsyākṣī-meghanāda- punarnavāḥ || iti ||
-\endverse
+
+\begin{versinnote}
+\tl{sarvaśākam acākṣuṣyaṃ cākṣuṣyaṃ śākapañcakam |\\+}
+\tl{jīvantī-vāstu-matsyākṣī-meghanāda- punarnavāḥ || iti ||\\!}
+\end{versinnote}
 
@@
 Aṣṭāṅgahṛdaya 8.42–43
-\startverse
-śīlayec chāligodhūmayavaṣaṣṭikajāṅgalam |\\
-suniṣaṇṇakajīvantībālamūlakavāstukam || 42 ||\\
-pathyāmalakamṛdvīkāpaṭolīmudgaśarkarāḥ |\\
-ghṛtadivyodakakṣīrakṣaudradāḍimasaindhavam || 43 ||
-\endverse
+
+\begin{versinnote}
+\tl{śīlayec chāligodhūmayavaṣaṣṭikajāṅgalam |\\+}
+\tl{suniṣaṇṇakajīvantībālamūlakavāstukam || 42 ||\\+}
+\tl{pathyāmalakamṛdvīkāpaṭolīmudgaśarkarāḥ |\\+}
+\tl{ghṛtadivyodakakṣīrakṣaudradāḍimasaindhavam || 43 ||\\!}
+\end{versinnote}
 
@@
 The Rājanighaṇṭu says rainwater:
-\startverse
-divyodakaṃ kharāri syād ākāśasalilaṃ tathā |\\
-vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam || Rajni 14.4
-\endverse
+
+\begin{versinnote}
+\tl{divyodakaṃ kharāri syād ākāśasalilaṃ tathā |\\+}
+\tl{vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam || Rajni 14.4\\!}
+\end{versinnote}
 
@@
 Yogacintāmaṇi
-\startverse
-piṣṭaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\
-manobhilaṣitaṃ yogyaṃ yogī bhojanam ācared iti ||
-\endverse
+
+\begin{versinnote}
+\tl{piṣṭaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\+}
+\tl{manobhilaṣitaṃ yogyaṃ yogī bhojanam ācared iti ||\\!}
+\end{versinnote}
 
 YBhD
-\startverse
-śreṣṭhaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam ||\\
-mano'bhilaṣitaṃ yogyaṃ yogī bhojanamācaret || 23 ||
-\endverse
+
+\begin{versinnote}
+\tl{śreṣṭhaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam ||\\+}
+\tl{mano'bhilaṣitaṃ yogyaṃ yogī bhojanamācaret || 23 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 DYS 40
-\startverse
-yuvāvastho 'pi vṛddho vā vyādhito vā śanaiḥ śanaiḥ |\\
-abhyāsāt siddhim āpnoti yoge sarvo 'py atandritaḥ || 40 ||
-\endverse
+
+\begin{versinnote}
+\tl{yuvāvastho 'pi vṛddho vā vyādhito vā śanaiḥ śanaiḥ |\\+}
+\tl{abhyāsāt siddhim āpnoti yoge sarvo 'py atandritaḥ || 40 ||\\!}
+\end{versinnote}
+
 \end{sources}
@@
 Yogacintāmaṇi
-\startverse
-haṭhapradīpikāyām—\\
-yuvā bālo 'tivṛddho vā vyādhito durbalo 'pi vā |\\
-abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ ||
-\endverse
+
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{yuvā bālo 'tivṛddho vā vyādhito durbalo 'pi vā |\\+}
+\tl{abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ ||\\!}
+\end{versinnote}
 
 Haṭharatnāvalī
-\startverse
-yuvā bhavati vṛddho 'pi vyādhito durbalo 'pi vā |\\
-abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ || 1.23 ||
-\endverse
+
+\begin{versinnote}
+\tl{yuvā bhavati vṛddho 'pi vyādhito durbalo 'pi vā |\\+}
+\tl{abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ || 1.23 ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 Haṭharatnāvalī
-\startverse
-pīthāni kumbhakāś citrā divyāni karaṇāni ca |\\
-sāṅgo 'pi ca haṭhābhyāso rājayogaphalārthadaḥ || 1.17 ||
-\endverse
+
+\begin{versinnote}
+\tl{pīthāni kumbhakāś citrā divyāni karaṇāni ca |\\+}
+\tl{sāṅgo 'pi ca haṭhābhyāso rājayogaphalārthadaḥ || 1.17 ||\\!}
+\end{versinnote}
+
 \end{testimonia}

### 2023-06-07 latex/HP1_TranslComm.tex
added 2nd chapter, changed omission display

@@
 \begin{testimonia}[hp01_004]
-Post-15th Century:
-Haṭharatnāvali 1.3
+Haṭharatnāvali 1.3 (Post-15th Century)
 
@@
 (check), \emph{Bṛhatkathāmañjarī} 10.45 (check) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}).
-It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a `part-time' Kāpālika.
+It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika.
 Alternatively, it could simply refer to one who used a broken skull as a bowl.
@@
 \begin{versinnote}
-\tl{haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\% (ab missing in some ŚS witnesses)
-bhaved vīryavatī guptā nirvīryā ca prakāśitā || 5.254\\!}
+\tl{haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\+}
+\tl{bhaved vīryavatī guptā nirvīryā ca prakāśitā || 5.254\\+}
+\tl{(ab missing in some ŚS witnesses)\\!}
 \end{versinnote}
@@
 Suśruta 6.17.67:
-\emph{gṛhe nirābādhe}
+
+\begin{versinnote}
+gṛhe nirābādhe
+\end{versinnote}
 
@@
 \tl{bhūmau niveśya saṃsthāpya vyomasthaṃ kukkuṭāsanam ||\\+}
-\tl{{[}niveśya bhūmau – mss. la, va, śa]\\!}
+\tl{[niveśya bhūmau – mss. la, va, śa]\\!}
 \end{versinnote}
@@
 \tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\+}
-\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || = 1.45\\+}
+\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau ||\\+}
 \tl{nāsāgre vinyased rājadantamūlaṃ tu jihvayā |\\+}
-\tl{uttabhya civukaṃ vakṣasy utthāpya pavanaṃ śanaiḥ || = 1.46\\+}
+\tl{uttabhya civukaṃ vakṣasy utthāpya pavanaṃ śanaiḥ ||\\+}
 \tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}

### 2023-06-07 latex/HP2_TranslComm.tex
added 2nd chapter, changed omission display

\input{Preamble-Print}
\input{CommStyle}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

\chapter*{Translation \& philological commentary}
\subsection*{2.1}
\begin{translation}[hp02_001]
Now, when posture is firm, the disciplined yogi whose diet is wholesome and measured, should practise breath control according to the path taught by the teacher.
\end{translation}

\begin{sources}[hp02_001]
\end{sources}

\begin{testimonia}[hp02_001]
Haṭharatnāvalī 3.78

\begin{versinnote}
\tl{atha prāṇāyāmaḥ--\\+}
\tl{athāsane dṛḍhe yogī vaśī hitamitāśanaḥ ||\\+}
\tl{gurūpadiṣṭamārgeṇa prānāyāmān samabhyaset || 3.78 ||\\!}
\end{versinnote}

Cf. Yogamārgaprakāśikā

\begin{versinnote}
\tl{athāsane dṛḍhībhūte suśobhanamaṭhe yadā |\\+}
\tl{guruṃ natvā śivaṃ caiva prāṇāyāmaṃ tato 'bhyaset || 47 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_001]
\emph{dṛḍho yogī} seems unlikely as \emph{dṛḍha} usually qualifies a technique, the body, etc.
It is interesting that J10, J17 and N17 have \emph{yogamārgaṃ}. The vulgate has \emph{prāṇāyāmān}.
\end{philcomm}

\subsection*{2.2}
\begin{translation}[hp02_002]
When the wind moves, everything moves and, when it is still, everything is still. Therefore, the yogī obtains the motionless state because of restraining the breath.
\end{translation}

\begin{sources}[hp02_002]
VM 71

\begin{versinnote}
\tl{cale vāte calaṃ sarvaṃ niścale niścal*aṃ* tathā |\\+}
\tl{yogī sthāṇutvam āpnoti tato vāyuni*baṃ*dhanāt || 71 ||\\!}
\end{versinnote}
Cf. Amanaska 92

\begin{versinnote}
\tl{citte calati saṃsāro 'cale mokṣaḥ prajāyate |\\+}
\tl{tasmāc cittaṃ sthirīkuryād audāsīnyaparāyaṇaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_002]
Yogacintāmaṇi

\begin{versinnote}
\tl{tathā ca skandapurāṇe--\\+}
\tl{cale vāte calaṃ cittaṃ niścalaṃ niścale tathā |\\+}
\tl{yogī sthāṇutvam āpnoti tato vāyuṃ nirodhayet ||\\!}
\end{versinnote}
Haṭharatnāvalī

\begin{versinnote}
\tl{cale vāte calaṃ cittaṃ niścale niścalaṃ tathā ||\\+}
\tl{yogī sthāṇutvam āpnoti tato vāyuṃ nirundhayet || 3.79 ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_002]
Comment on \emph{cittaṃ} instead of \emph{sarvaṃ}. The latter is supported by the source and makes better sense as a general statement (i.e., when the wind moves, everything moves, etc.). The reading \emph{citta} only seems to make sense if one understands \emph{sthāṇutvam} as referring to \emph{samādhi} (i.e., motionlessness of the mind). As far as we know, such a meaning of \emph{sthāṇutvam} is not attested elsewhere. Nonetheless, Brahmānanda glosses \emph{sthāṇutvam} as \emph{kāṣṭhavat}, implying that it refers to \emph{samādhi}.
\emph{vāyunibandanāt} makes better sense, but most witnesses have an optative verb. There was probably confusion over the ablative and optative in 2d and 3d. The optative verb works well in 3d.
\end{philcomm}

\subsection*{2.3}
\begin{translation}[hp02_003]
As long as the breath is situated in the body, there is said to be life. The emission of this is death. Therefore, one should stop the breath.
\end{translation}

\begin{sources}[hp02_003]
VM 72

\begin{versinnote}
\tl{yāvad vāyuḥ sthito dehe tāvaj jīvitam ucyate\\+}
\tl{maraṇaṃ tasya niḥkrāntau tato vāyuṃ nirodhayet || 72 ||\\!}
\end{versinnote}
Cf. Mṛgendratantra 1.11.20cd--22ab

\begin{versinnote}
\tl{vyāpārād yasya ceṣṭante  śārīrāḥ pañca vāyavaḥ  || 1,11.20 ||\\+}
\tl{prāṇāpānād ayas te tu bhinnā vṛtter na vastutaḥ  |\\+}
\tl{vṛttiṃ leśān nigadato bharadvāja nibodha me  || 1,11.21 ||\\+}
\tl{vṛttiḥ praṇayanaṃ nāma yat taj jīvanam ucyate  |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_003]
\end{testimonia}

\begin{philcomm}[hp02_003]
We should adopt \emph{jīvitaṃ} because the old witnesses of the VM have it and it is well attested among the old HP mss. However, both \emph{jīvitaṃ} and \emph{jīvanaṃ} are possible. Also, \emph{nirundhayet} may have been the original reading that was  changed to the more correct form of \emph{nirodhayet} and later confused with the final pāda of the previous verse and changed to \emph{nibandhayet}.
\end{philcomm}

\subsection*{2.4}
\begin{translation}[hp02_004]
When the channels are full of impurities, the breath does not go into the central channel. How would the state beyond the mind arise? How would perfection of the body be achieved?
\end{translation}

%\begin{sources}[hp02_004]
%\end{sources}

\begin{testimonia}[hp02_004]
Haṭharatnāvalī 3.81

\begin{versinnote}
\tl{malākulāsu nāḍīṣu māruto naiva madhyagaḥ ||\\+}
\tl{kathaṃ syād unmanībhāvaḥ kāyasiddhiḥ kathaṃ bhavet || 3.81 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_004]
The manuscripts are split between \emph{kāyasiddhi} and \emph{kāryasiddhi}. In the context of physical yoga, \emph{kāyasiddhi} makes better sense, as \emph{kāryasiddhi} can refer more generally to accomplishing anything. 
\end{philcomm}

\subsection*{2.5}
\begin{translation}[hp02_005]
When entire network of channels, which is full of impurities, becomes pure, then the yogi becomes capable of holding the breath.
\end{translation}

\begin{sources}[hp02_005]
VM 95

\begin{versinnote}
\tl{śuddhim eti yadā sarvaṃ nāḍīcakraṃ malākulam |\\+}
\tl{tadaiva jāyate yogī prāṇasaṃgrahaṇakṣamaḥ || 76 ||\\!}
\end{versinnote}
On nāḍīcakra: Sārdhatriśatikālottara 10.1

\begin{versinnote}
\tl{nāḍīcakraṃ paraṃ sūkṣmaṃ pravakṣyāmy anupūrvaśaḥ |\\+}
\tl{nābher adhastād yat kandam aṅkurās tatra nirgatāḥ || 1 ||\\+}
\tl{dvāsaptatisahasrāṇi nābhimadhye vyavasthitāḥ |\\+}
\tl{tiryag ūrdhvam adhaś caiva vyāptaṃ nābheḥ samantataḥ |\\+}
\tl{cakravatsaṃsthitā nāḍyaḥ pradhānā daśa tāsu yāḥ |\\+}
\tl{iḍā ca piṅgalā caiva suṣumnā ca tṛtīyakā\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_005]
Yogacintāmaṇi

\begin{versinnote}
\tl{skandapurāṇe--\\+}
\tl{śuddhim eti yadā sarvaṃ nāḍīcakraṃ malākulam |\\+}
\tl{tadaiva jāyate yogī kṣamaḥ prāṇanibandhane ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_005]

\end{philcomm}

\subsection*{2.6}
\begin{translation}[hp02_006]
Therefore, one should always practice breath [retention] with a pure mind, so that the impurities situated at the sides of Suṣumṇā wither away.
\end{translation}

\begin{sources}[hp02_006]
Gorakṣaśataka 73cd–74ab

\begin{versinnote}
\tl{prāṇābhyāsas tataḥ kāryo nityaṃ sattvāsthayā dhiyā |\\+}
\tl{suṣumnāṃ layate cittaṃ na ca vāyuḥ pradhāvati ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_006]
Jyotsnā

\begin{versinnote}
\tl{prāṇāyāmaṃ tataḥ kuryān nityaṃ sāttvikayā dhiyā |\\+}
\tl{yathā suṣumṇānāḍīsthā malāḥ śuddhiṃ prayānti ca || 6 ||\\!}
\end{versinnote}

Yogakarṇikā 137

\begin{versinnote}
\tl{prāṇāyāmaṃ tataḥ kuryānnityaṃ sāttvikayā dhiyā || 137 ||\\+}
\tl{tathā suṣumnāpārśvasthā malāḥ śodhaṃ prayānti hi |\\!}
\end{versinnote}

Prāṇatoṣaṇī p.788

\begin{versinnote}
\tl{prāṇāyāmaṃ tataḥ kuryān nityaṃ  sāttvikayā dhiyā | \\+}
\tl{tathā suṣumnāpārśvasthā malāḥ śoṣaṃ prayānti hi |\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_006]
There’s a problem in the third pāda. Among the divergent readings, V19 conveys the idea that the impurities are in the central channel  (\emph{suṣumṇāntarasthā}), which is a ra-vipulā without the caesura after the 4th syllable). This is similar in sense to V2, which has the reading \emph{suṣumṇāmadhyasthā}. This meaning was accepted by Brahmānanda. However, we are not aware of a reference in another text to impurities (\emph{mala}) existing in the central channel. Therefore, we have adopted the reading of N23,J7 and A1, which states that the impurites are situated at the sides of Suṣumṇā, which would allude to the secondary channels of \emph{iḍā} and \emph{piṅgalā}.

The \emph{ca} at the end of the 4th pāda suggests that two statements are being made in the second hemistich. The reading of V1 makes sense of the \emph{ca}. However, the meaning of \emph{suṣumṇā susnigdhā} (`\emph{suṣumṇā} becomes well-lubricated') seems implausible. It seems likely that \emph{ca} was understood simply as a verse-filler.
\end{philcomm}

\subsection*{2.7}
\begin{translation}[hp02_007]
Seated in lotus pose, the yogi should breathe in the air through the moon channel, hold according to his capacity and then breathe out through the sun channel.
\end{translation}

\begin{sources}[hp02_007]
VM 77

\begin{versinnote}
\tl{baddhapadmāsano yogī prāṇaṃ candreṇa pūrayet |\\+}
\tl{dhārayitvā yathāśaktyā bhūyaḥ sūryeṇa recayet || 77 ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_007]
Yuktabhavadeva

\begin{versinnote}
\tl{baddhapadmāsano yogī prāṇaṃ candreṇa pūrayet |\\+}
\tl{dhārayitvā yathāśakti bhūyaḥ sūryeṇa recayet || 12 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_007]

\end{philcomm}

\subsection*{2.8}
\begin{translation}[hp02_008]
And having drawn the breath through the sun channel, he should gradually fill the abdomen. Having done the retention as prescribed, he should then exhale through the moon channel.
\end{translation}

\begin{sources}[hp02_008]
VM 79

\begin{versinnote}
\tl{prāṇaṃ sūryeṇa cākṛṣya pūrayed udaraṃ śanaiḥ\\+}
\tl{vidhivat kuṃbhakaṃ kṛtvā punaś candreṇa recayet ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_008]
Yukabhavadeva

\begin{versinnote}
\tl{prāṇaṃ sūryeṇa cākṛṣya pūrayedudaraṃ śanaiḥ\\+}
\tl{kumbhayitvā vidhānena bhūyaś candreṇa recayet ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_008]
\emph{udare} is well-attested (V1 and J10) but the verb needs an object so \emph{udaraṃ} has been adopted.
\end{philcomm}

\subsection*{2.9}
\begin{translation}[hp02_009]
Having inhaled through the [channel] by which one exhales, one should hold it without discomfort. And then he exhales through the other [channel] gently, not forcefully.
\end{translation}

\begin{sources}[hp02_009]
Cf. DYŚ 60--61

\begin{versinnote}
\tl{dhārayitvā yathāśakti recayed iḍayā śanaiḥ || 62 ||\\+}
\tl{yathā tyajet tathāpūrya dhārayed avirodhataḥ |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_009]
Haṭharatnāvalī 3.85cd%–86ab

\begin{versinnote}
\tl{yena tyajet tenāpūrya dhārayed avirodhataḥ ||\\!}
%prāṇaṃ ced iḍayā piben niyamitaṃ bhūyo 'nyayā recayet |
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_009]
The meaning of \emph{avirodhataḥ} (‘without harm’) makes better sense in this context than \emph{anirodhataḥ} (‘without cessation’). One might try to construe \emph{anirodhataḥ} as ‘without stopping the breath’, but verse 2.7 clearly states that the breath should be held as long as possible (\emph{yathāśakti}). According to the apparatus of the critical edition of the \emph{Haṭharatnāvalī}, \emph{avirodhataḥ} is well-attested in the transmission of this work.
\end{philcomm}

\subsection*{2.10}
\begin{translation}[hp02_010]
If [the yogi] takes in the breath by the moon channel, he should then exhale the restrained [breath] through the other. Having taken in the breath through the sun channel and held it, he should then exhale through the left. By this method of sun and moon, the multitudes of channels of yamis who are meditating on both orbs become pure after three months.
\end{translation}

\begin{sources}[hp02_010]
VM 81

\begin{versinnote}
\tl{prāṇaṃ ced iḍayā pibet niyamitaṃ bhūyo 'nyayā recayet |\\+}
\tl{pītvā piṅgalayā samīraṇam alaṃ baddhvā tyajed vāmayā \\+}
\tl{sūryācandramasor anena vidhinā bimbadvayaṃ dhyāyatāṃ \\+}
\tl{śuddhā nāḍigaṇā bhavanti yaminā māsatrayād ūrdhvataḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_010]
Yogacintāmaṇi

\begin{versinnote}
\tl{haṭhayoge 'pi— \\+}
\tl{prāṇaṃ ced iḍayā piben niyamito bhūyo 'nyayā recayet\\+}
\tl{pītvā piṅgalayā samīraṇam atho baddhvā tyajed vāmayā |\\+}
\tl{sūryācandramasor anena vidhinā bimbadvayaṃ dhyāyatām\\+}
\tl{śuddhā nāḍigaṇā bhavanti yamināṃ māsatrayād ūrdhvataḥ ||\\+}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{prāṇaṃ ced iḍayā piben niyamitaṃ bhūyo 'nyayā recayet\\+}
\tl{pītvā piṅgalayā samīraṇaṃ atho baddhvā tyajed vāmayā ||\\+}
\tl{sūryācandramasor anena vidhinā bimbadvayaṃ dhyāyatāṃ\\+}
\tl{śuddhā nāḍigaṇā bhavanti yamināṃ māsatrayād ūrdhvataḥ || 3.86 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_010]
The reading \emph{vidhinābhyāsaṃ} in the third pāda of various witnesses is unmetrical because it lacks the caesura. 
\end{philcomm}

\subsection*{2.11}
\begin{translation}[hp02_011]
One should slowly practise retentions four times, at sunrise, midday, sunset and midnight, up to [a total of] eighty.
\end{translation}

\begin{sources}[hp02_011]
\end{sources}

\begin{testimonia}[hp02_011]
Haṭharatnāvalī 3.87

\begin{versinnote}
\tl{prātar madhyadine sāyam ardharātre ca kumbhakān ||\\+}
\tl{śanair aśītiparyantaṃ caturvāraṃ samabhyaset || 3.87 ||\\!}
\end{versinnote}

Cf. DYŚ 63cd--65ab

\begin{versinnote}
\tl{evaṃ prātaḥ samāsīnaḥ kuryād viṃśati kumbhakān || 63 ||\\+}
\tl{evaṃ madhyāhnasamaye kuryād viṃśati kumbhakān |\\+}
\tl{evaṃ sāyaṃ prakurvīta punar viṃśati kumbhakān || 64 ||\\+}
\tl{evam evārdharātre 'pi kuryād viṃśati kumbhakān |\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_011]
The \emph{caturvāraṃ} is ambiguous, but the parallel verses in the DYŚ make it clear that twenty kumbhakas are to be practised four times a day. In his Jyotsnā, Brahmānanda understands it as eighty kumbhakas four times a day.

Note Brahmānanda’s comment on \emph{śanaiḥ} (i.e., gradually building up to 80 retentions).

\end{philcomm}

\subsection*{2.12}
\begin{translation}[hp02_012]
In the lowest cessation of the breath, sweating arises, in the average, shaking and in the highest [the yogi] rises up again again in lotus pose.
\end{translation}

\begin{sources}[hp02_012]
Cf. VM 87

\begin{versinnote}
\tl{adhame ca ghano gharma kampo bhavati madhyame |\\+}
\tl{uttiṣṭhaty uttamo deho baddhapadmāsano muhuḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_012]
Haṭharatnāvalī 3.88

\begin{versinnote}
\tl{kanīyasi bhavet svedaḥ kampo bhavati madhyame |\\+}
\tl{uttiṣṭhaty uttame prāṇarodhe padmāsane muhuḥ || 3.88 ||\\!}
\end{versinnote}

Yogacintāmaṇi

\begin{versinnote}
\tl{kanīyasi bhavet svedaḥ kampo bhavati madhyame |\\+}
\tl{uttiṣṭhaty uttame prāṇarodhe padmāsanasthitaḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_012]
The manuscripts diverge greatly in the second hemistrich. All have \emph{prāṇa} (instead of \emph{deha} in the VM). J10 (and others) appear to say that the breaths rise up again and again when one is seated in the lotus pose (\emph{uttiṣṭhanty uttame prāṇā baddhe padmāsane muhuḥ}). However, this is a statement about the external signs that arise in \emph{prāṇāyāma}. V1 seems to be stating that \emph{padmāsana} rises up again and again in the highest stage of holding the breath (\emph{uttiṣṭhaty uttame prāṇarodhe padmāsanaṃ muhuḥ}).

Confusion has arisen here from the verse being taken from the VM without its context, which is a classification of different levels of prāṇāyāma. Thus Svātmārāma needed to include prāṇarodhe meaning prāṇāyāma in order for the different adjectives to have something to agree with.
\end{philcomm}

\subsection*{2.13}
\begin{translation}[hp02_013]
The yogī should practise rubbing the limbs with the sweat produced by his exertion. Because of it, firmness and dexterity of the body arise.
\end{translation}

\begin{sources}[hp02_013]
Cf. DYŚ 75

\begin{versinnote}
\tl{prasvedo jāyate pūrvaṃ mardanaṃ tena kārayet |\\+}
\tl{tato ’tidhāraṇād vāyoḥ krameṇaiva śanaiḥ śanaiḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_013]
Haṭharatnāvalī 3.89

\begin{versinnote}
\tl{jalena śramajātena aṅgamardanam ācaret |\\+}
\tl{dṛḍhatā laghutā cāpi tathā gātrasya jāyate || 3.89 ||\\!}
\end{versinnote}

Cf. Śivasaṃhitā 3.46

\begin{versinnote}
\tl{svedaḥ saṃjāyate dehe yoginaḥ prathamodyame |\\+}
\tl{yadā saṃjāyate svedo mardanaṃ kārayet sudhīḥ |\\+}
\tl{anyathā vigrahe dhāturnaṣṭo bhavati yoginaḥ ||\\!}
\end{versinnote}

Yogacintāmaṇi

\begin{versinnote}
\tl{jalena śramajātena gātramardanam ācaret |\\+}
\tl{dṛḍhatā laghutā cāpi tena gātrasya jāyate ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_013]
See Śivasaṃhitā 3.46 on \emph{mardana}. If it is not done, dhātus are lost from the body.
\end{philcomm}

\subsection*{2.14}
\begin{translation}[hp02_014]
At the beginning of the practice, food with milk and ghee is recommended. Then, when the practice has become well established, there is no need to adopt such regulations.
\end{translation}

\begin{sources}[hp02_014]
Śivasaṃhitā 3.43

\begin{versinnote}
\tl{abhyāsakāle prathamaṃ kuryāt kṣīrājyabhojanam\\+}
\tl{tato'bhyāse sthirībhūte na tādṛṅ niyamagrahaḥ 3.43\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_014]
Haṭharatnāvalī 1.24

\begin{versinnote}
\tl{abhyāsakāle prathame śastaṃ kṣīrādibhojanam |\\+}
\tl{tato 'bhyāse dṛḍhībhūte na tāvan niyamagrahaḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_014]
\end{philcomm}

\subsection*{2.15}
\begin{translation}[hp02_015]
Just as a lion, elephant and tiger should be tamed gradually, so the breath should be trained. Otherwise, it kills the practitioner.
\end{translation}

\begin{sources}[hp02_015]
VM 101

\begin{versinnote}
\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śāneḥ |\\+}
\tl{anyathā hanti yāntāraṃ tathā vāyur asevitaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_015]
Haṭharatnāvalī 3.90

\begin{versinnote}
\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śanaiḥ ||\\+}
\tl{tathaiva sevito vāyur bhaved vaśyaḥ śanaiḥ śanaiḥ || 3.90 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_015]
The second hemistich has been rewritten to express the same idea (and simile) in the VM, but the VM expresses it better.
\end{philcomm}

\subsection*{2.16}
\begin{translation}[hp02_016]
The end of all diseases is caused by prāṇāyāma as prescribed. As a result of inappropriate practice any disease can arise.
\end{translation}

\begin{sources}[hp02_016]
VM 99

\begin{versinnote}
\tl{prāṇāyāmena yuktena sarvarogakṣayo bhavet |\\+}
\tl{ayuktābhyāsayogena sarvarogasamudbhavaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_016]
\end{testimonia}

\begin{philcomm}[hp02_016]
\end{philcomm}

\subsection*{2.17}
\begin{translation}[hp02_017]
Hiccups, wheezing, cough, pains in the head, ears and eyes and various [other] diseases arise as a result of the breath going awry.
\end{translation}

\begin{sources}[hp02_017]
VM 100

\begin{versinnote}
\tl{hikkā śvāsaś ca kāsaś ca śiraḥkarṇākṣivedanā |\\+}
\tl{bhavanti vividhā doṣāḥ pavanasya vyatikramāt ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_017]
Cf.  Śivadharmottara 10.124

\begin{versinnote}
\tl{hikkāśvāsapratiśyāyaḥ karṇadantākṣivedanāḥ || 10.124 ||\\+}
\tl{mūkatā jaḍatā kāsaḥ śirorogaḥ śramakṣaraḥ |\\+}
\tl{ityevamādayo doṣā jāyante vyutkrameṇa tu || 10.125 ||\\!}
\end{versinnote}

Cf. Dharmaputrikā 10.265

\begin{versinnote}
\tl{kaphakoṣṭhe yadā vāyur granthir bhūtvāvatiṣṭhate |\\+}
\tl{hallāsahikkikāśvāsaśiraḥ śūlādayo rujāḥ || 265 || [hṛllāsa-]\\+}
\tl{jāyante dhātuvaiṣamyāt tadā kuryāt pratikriyāṃ |\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_017]
\end{philcomm}

\subsection*{2.18}
\begin{translation}[hp02_018]
One should exhale correctly, inhale correctly and hold the breath correctly. Thus, one obtains success.
\end{translation}

\begin{sources}[hp02_018]
VM 102

\begin{versinnote}
\tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ ca pūrayet\\%122ab ; HP2.18ab
yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt || 1\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_018]
Haṭharatnāvalī

\begin{versinnote}
\tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ prapūrayet ||\\+}
\tl{yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt || 3.93 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_018]
V1 has \emph{śuddhiṃ} (instead of \emph{siddhim}) but this may be an error by someone who was anticipating the subject of the next verse.

%AGS: the repetition of yuktaṃ suggests a southern origin for the verse. [search for Proof]
\end{philcomm}

\subsection*{2.19}
\begin{translation}[hp02_019]
When the channels are pure, then external signs occur. Leanness and lustre of the body are certain to arise.
\end{translation}

\begin{sources}[hp02_019]
Cf.  DYŚ 67--68

\begin{versinnote}
\tl{yadā tu nāḍiśuddhiḥ syāt tadā cihnāni bāhyataḥ || 67 ||\\+}
\tl{jāyante yogino dehe tāni vakṣyāmy aśeṣataḥ |\\+}
\tl{śarīralaghutā dīptir jaṭharāgnivivardhanam || 68 ||\\+}
\tl{kṛśatvaṃ ca śarīrasya tadā jāyeta niścitam |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_019]
Haṭharatnāvalī

\begin{versinnote}
\tl{yadā tu nāḍīśuddhiḥ syāt tadā cihnāni bāhyataḥ ||\\+}
\tl{kāyasya kṛśatā kāntir jāyate tasya niścitam || 3.94 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_019]
The idea that \emph{prāṇāyāma} is done to purify the channels (\emph{nāḍī}) can be found in discussions of \emph{prāṇāyāma} in early Śaiva tantras. For example, the \emph{Nayasūtra} of the \emph{Niśvāsatattvasaṃhitā} (4.110) and the \emph{Svacchandatantra} (7.294cd–7.295ab) specifically refer to purifying the channels by inhaling through the left nostril and exhaling through the right (\emph{apasavyena pūryeta savyenaiva virecayet} | \emph{nāḍīsaṃśodhanaṃ caitan mokṣamārgapathasya ca}).
\end{philcomm}

\subsection*{2.20}
\begin{translation}[hp02_020]
Holding of the breath as long as one desires, stimulating the fire, causing the internal resonance to arise and freedom from disease occur when the channels are pure.
\end{translation}

\begin{sources}[hp02_020]
VM 101

\begin{versinnote}
\tl{yatheṣṭaṃ dhāraṇā[ṃ] vāyor analasya pradīpanam |\\+}
\tl{nādābhivyaktir ārogyaṃ jāyate nāḍīśodhanāt ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_020]
Cf. Vasiṣṭhasaṃhitā 2.68--69

\begin{versinnote}
\tl{nāḍīśuddhim avāpnoti pṛthak cihnopalakṣitām | 2.68ab\\+}
\tl{śarīralaghutā dīptir jaṭharāgnivivardhanam || 2.68cd\\+}
\tl{nādābhivyaktir ity etac cihnaṃ tacchuddhisūcakam | 2.69ab\\+}
\tl{yāvad etāni saṃpaśyet tāvad evaṃ samācaret || 2.69cd\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_020]

\end{philcomm}

\subsection*{2.21}
\begin{translation}[hp02_021]
One who has an excess of fat and phlegm should initially practise the six therapeutic interventions. However, anyone else should not practise them when their humours are in balance.
\end{translation}

\begin{sources}[hp02_021]
\end{sources}

\begin{testimonia}[hp02_021]

\end{testimonia}

\begin{philcomm}[hp02_021]
The J10 group also have a valid reading of \emph{medaḥśleṣmādināśārthaṃ} and \emph{anyathā}. However, the \emph{pūrvam} in V1, J8 and V3 seems to the fit the context the ṣaṭkarma as a preliminary practice for prāṇāyāma.
\end{philcomm}

\subsection*{2.22}
\begin{translation}[hp02_022]
Dhauti, basti, neti, trāṭaka, naulika and kapālabhāti. [Experts] says these are the six therapeutic interventions.
\end{translation}

\begin{sources}[hp02_022]
\end{sources}

\begin{testimonia}[hp02_022]
\end{testimonia}

\begin{philcomm}[hp02_022]
The issue with this verse is the spellings of the names of the techniques. Looking at the occurrences of each name in subsequent verses helps to establish this.
\end{philcomm}

\subsection*{2.23}
\begin{translation}[hp02_023]
These six interventions should be kept secret. Purifiers of the body, which bestow various good qualities, they are worshipped by the best yogis.
\end{translation}

\begin{sources}[hp02_023]
\end{sources}

\begin{testimonia}[hp02_023]
\end{testimonia}

\begin{philcomm}[hp02_023]
\end{philcomm}

\subsection*{2.24}
\begin{translation}[hp02_024]
One should swallow a cloth that is four finger-breadths wide and has been soaked. Then one should draw it out. This has been taught as the action of dhauti.
\end{translation}

\begin{sources}[hp02_024]
\end{sources}

\begin{testimonia}[hp02_024]
Satkarmasaṅgraha 56–58

\begin{versinnote}
\tl{atha dhautī
mṛdulaṃ dhavalaṃ śuddhaṃ caturaṅgulavistṛtam |\\+}
\tl{tithihastamitāyāmaṃ dhautīvastrasya lakṣaṇam || 56 || [tithi =15]\\+}
\tl{toyasiktaṃ grased vastraṃ ghrāṇābhyāṃ vāyum utsṛjan |\\+}
\tl{śanaiḥ sanais tu sakalaṃ punaḥ pratyāharec chanaiḥ ||\\+}
\tl{dhautīkarmedam ākhyātaṃ yatra gaṅgādhidaivatam || 57 ||\\+}
\tl{kāsasvāsaphīlakuṣṭhādināśam vahner māndyaṃ viṃśatiḥ śleṣarogān |\\+}
\tl{dūrīkuryāt karṇabādhir tam uccair dhautīkarma praditaṃ śaṅkareṇa || 58 ||\\!}
\end{versinnote}

Yogacintāmaṇi

\begin{versinnote}
\tl{caturaṅgulavistāraṃ hastapañcadaśena tu |\\+}
\tl{svagurūktaprakāreṇa siktaṃ vastraṃ śanair graset |\\+}
\tl{punaḥ pratyāhared etad abhyāsād dhautikarmavit ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{atha dhautiḥ--\\+}
\tl{viṃśaddhastapramāṇena dhautavastraṃ sudīrghitam |\\+}
\tl{caturaṅgulavistāraṃ siktaṃ caiva śanaiḥ graset || 1.37 ||\\+}
\tl{tataḥ pratyāharec caitad abhyāsād dhautir ucyate |\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_024]
At this point, it seems that most mss. have added \emph{pāda}s on the length of the cloth and doing the practice according to the guru’s teachings. These \emph{pāda}s are absent in many of the old manuscripts, including both V1, V19 and group 2, and the compound \emph{hastapañcadaśena} does not seem to fit the syntax of the sentence. The other inserted \emph{pāda} (i.e., \emph{gurūpadiṣṭa°}) appears to be a verse filler.
\end{philcomm}

\subsection*{2.25}
\begin{translation}[hp02_025]
Coughing, wheezing, splenitis and skin diseases, as well as the twenty phlegmatic diseases, undoubtedly disappear through the power of the dhauti technique.
\end{translation}

\begin{sources}[hp02_025]
\end{sources}

\begin{testimonia}[hp02_025]
Yogacintāmaṇi

\begin{versinnote}
\tl{kāsaśvāsaplīhakuṣṭhaṃ kapharogāś ca vidradhiḥ |\\+}
\tl{dhautīkarmaprabhāvena prayānty eva na saṃśayaḥ ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{kāsaśvāsaplīhakuṣṭhaṃ kapharogāś ca viṃśatiḥ |\\+}
\tl{dhautikarmaprabhāvena dhāvanty eva na saṃśayaḥ || 1.39 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_025]
\emph{dhāvanti} is a play on words, using a different root \emph{dhāv}, “to run”, from that of \emph {dhauti}, which is derived from \emph{dhāv}, “to purify”.
Look at lists of the 20 kapha roga (in āyurvedic texts) to see if they are different to those mentioned in 2.25a
\end{philcomm}

\subsection*{2.26}
\begin{translation}[hp02_026]
In water up to the navel, one adopts a squatting pose (\emph{utkaṭa}) with a reed fixed in the anus. One should clench the base. This cleansing is bastikarma.
\end{translation}

\begin{sources}[hp02_026]
\end{sources}

\begin{testimonia}[hp02_026]
Yogacintāmaṇi

\begin{versinnote}
\tl{atha vastī |\\+}
\tl{nābhidaghne jale pāyunyastanālotkaṭāsanaḥ |\\+}
\tl{ādhārākuñcanaṃ kuryād abhyāsād vastikarmavit ||\\!}
\end{versinnote}

Haṭharatnāvalī 1.45

\begin{versinnote}
\tl{nābhidaghne jale sthitvā pāyunāle sthitāṅguliḥ |\\+}
\tl{cakrimārgeṇa jaṭharaṃ pāyunālena pūrayet || 1.45 ||\\+}
\tl{vicitrakaraṇīm kṛtvā nirbhītaḥ recayej jalam |\\+}
\tl{yāvad balaṃ prapūryaiva kṣaṇaṃ sthitvā virecayet || 1.46 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_026]

\end{philcomm}

\subsection*{2.27}
\begin{translation}[hp02_027]
By the power of the basti technique, swelling, splenitis, stomach disorders and all diseases arising from wind, bile and phlegm are removed.
\end{translation}

\begin{sources}[hp02_027]
\end{sources}

\begin{testimonia}[hp02_027]
Haṭharatnāvalī

\begin{versinnote}
\tl{gulmaplīhodaraṃ vāpi vātapittakaphādikam |\\+}
\tl{bastikarmaprabhāvena dhāvanty eva saṃśayaḥ || 1.48 ||\\!}
\end{versinnote}

Yogacintāmaṇi

\begin{versinnote}
\tl{gulmodaraṃ cāpi vātaplīhapittakaphodbhavāḥ |\\+}
\tl{vastikarmaprabhāvena bādhyante sakalāmayāḥ ||\\!}
\end{versinnote}

Cf. Satkarmasaṅgraha 135

\begin{versinnote}
\tl{yāvan malā vinaśyanti vātapittakaphodbhāvāḥ |\\+}
\tl{trivāraṃ vā caturvāraṃ kṛtvā bastim virecayet || 135 ||\\+}
\tl{mahojasvī mahajjyotir jaṭharāgnipradīpanam |\\+}
\tl{gulmaplīhodarādīnāṃ nāśanaṃ sukhavardhanam || 140 ||\\+}
\tl{vātapittakaphottānāṃ doṣāṇāṃ nāśanaṃ paraṃ |\\+}
\tl{kuṣṭhānāṃ nāśanaṃ cāpi bastisiddhe prajāyate || 141 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_027]
The J10 group has \emph{°bhavaṃ}, which would qualify \emph{gulmaplīhodaram}.

Diwakar thinks that \emph{gulmodara} and \emph{plīhodara} should be read. Search āyurvedic commentaries on this. (homework)
\end{philcomm}

\subsection*{2.28}
\begin{translation}[hp02_028]
When practised repeatedly the jalabasti technique bestows purity of the bodily constituents, senses and mind, radiance, [and] stimulation of the digestive fire, and removes [excessive] accumulation of any humour.
\end{translation}

\begin{sources}[hp02_028]
\end{sources}

\begin{testimonia}[hp02_028]
Yogacintāmaṇi

\begin{versinnote}
\tl{dhātvindriyāntaḥkaraṇaprasādaṃ\\+}
\tl{dadhyāc ca kāntiṃ dahanapradīptim |\\+}
\tl{aśeṣadoṣopacayaṃ nihanyād\\+}
\tl{abhyasyamānaṃ jalavastikarma ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{dhātvindriyāntaḥkaraṇaprasādaṃ \\+}
\tl{dadyāc ca kāntiṃ dahanapradīptim |\\+}
\tl{aśeṣadoṣopacayaṃ nihanyād \\+}
\tl{abhyasyamānaṃ jalabastikarma || 1.49 ||\\!}
\end{versinnote}

Cf. SKS

\begin{versinnote}
\tl{tiṣṭhed vaśī mitāhāraḥ sarvāṅgaṃ tena śudhyati |\\+}
\tl{dhātvindriyāntaḥkaraṇaprasādo dehalāghavam || 139 ||\\+}
\tl{mahojasvī mahajjyotir jaṭharāgnipradīpanam |\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_028]
\end{philcomm}

\subsection*{2.29}
\begin{translation}[hp02_029]
[The technique] which, after the apāna wind has been raised into the tube of the throat (i.e., oesophagus), ejects the contents of the stomach from the windpipe (which has been brought under control by cumulative practice) is called the elephant’s action by experts of Haṭha.
\end{translation}

\begin{sources}[hp02_029]
\end{sources}

\begin{testimonia}[hp02_029]
Haṭharatnāvalī 1.51

\begin{versinnote}
\tl{udaragatapadārtham udvamantī \\+}
\tl{pavanam apānam udīrya kaṇṭhanāle |\\+}
\tl{kramaparicayatas tu vāyumārge \\+}
\tl{gajakaraṇīti nigadyate haṭhajñaiḥ || 1.51 ||\\!}
\end{versinnote}

HTK 8.8

\begin{versinnote}
\tl{udaragatapadārthān udvamed eva nityaṃ\\+}
\tl{pavanagamanamārgāt kaṇṭhanālapraveśāt ||\\+}
\tl{kramaparicayavaśyaṃ syāc ca gargādayo hi\\+}
\tl{gajakaraṇam itīha prahur āryā munīndrāḥ || 8 ||\\!}
\end{versinnote}

Satkarmasaṅgraha 108–109

\begin{versinnote}
\tl{atha gajakaraṇī\\+}
\tl{śuddhaṃ toyaṃ nārikelodbhavaṃ vā \\+}
\tl{pītvākaṇṭhaṃ dugdhamiśraṃ jalaṃ vā |\\+}
\tl{vāraṃ vāraṃ māṇibandhaṃ tu kurvan \\+}
\tl{nodgāreṇa prakṣiped bhūmibhāge  || 108 ||\\+}
\tl{eṣā proktā kaphapittāmayeṣu \\+}
\tl{medoghnīva kariṇī hastipūrvā || 109 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_029]
The reference to the elephants in in the J10 group (as well as J8 and V19) appears to be explaining the name of this practice (i.e., ‘because the speed of the breath is like that of water [propelled] by elephants’). However, the syntax is not so easy to construe. The first half of the adopted reading (i.e., \emph{kramaparicayavaśya} is supported by V1 and group 2, and these witnesses preserve \emph{mārga} at the end.

The vulgate and printed editions have a finite verb in the first hemistich and assume a plural subject (i.e., \emph{yoginaḥ}) instead of the feminine singular present participle (agreeing with \emph{gajakaraṇī}).
\end{philcomm}

\subsection*{2.30}
\begin{translation}[hp02_030]
[The yogi] should insert a thread which is handspan [long] and very smooth into the nasal passage. He takes it out through the mouth. This is called neti by the Siddhas.
\end{translation}

\begin{sources}[hp02_030]
\end{sources}

\begin{testimonia}[hp02_030]
Yogacintāmaṇi

\begin{versinnote}
\tl{atha netī |\\+}
\tl{sūtraṃ vitastisusnigdhaṃ nāsānāle praveśayet |\\+}
\tl{mukhān nirgamayet sā hi netī siddhair nigadyate ||\\!}
\end{versinnote}

Cf. Haṭharatnāvalī 1.40–41

\begin{versinnote}
\tl{atha netikarma--\\+}
\tl{ākhupucchākāranibhaṃ sūtraṃ susnigdhanirmitam |\\+}
\tl{ṣaḍvitastimitaṃ sūtraṃ netisūtrasya lakṣanam || 1.40 ||\\+}
\tl{nāsānāle praviśyainaṃ mukhān nirgamayet kramāt |\\+}
\tl{sūtrasyāntaṃ prabaddhvā tu bhrāmayen nāsanālayoḥ |\\!}
\end{versinnote}

Cf. SKS

\begin{versinnote}
\tl{atha netī\\+}
\tl{mṛdu ślakṣṇaṃ sitaṃ sūtraṃ nāsānāle praveśayet |\\+}
\tl{mukhān nirgamayed dasrau cintayen netikā smṛtā || 67 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_030]
The reading \emph{mukhanirgamanād} is difficult to construe. It appears to a vague \emph{nirvacana}. According to Turner’s Comparative and Etymological Dictionary (1966: 427, entry 7588), the word \emph{netī} in Hindi refers to the cord of a churning stick and is cognate with the Sanskrit \emph{netra}. The action of pulling the cord of a churning stick is similar to the way the thread can be pulled back and forth, from side to side, through the nostril and mouth. The reading \emph{mukhān nirgamayec caiṣā} makes better sense in terms of describing the final part of the practise, but it may be a patch because the first hemistich has a finite verb and the \emph{eṣā} must be construed with \emph{netiḥ} in the fourth pāda. The same problem is in the reading V19 (\emph{sāpi}).
\end{philcomm}

\subsection*{2.31}
\begin{translation}[hp02_031]
And neti purifies the skull, bestows devine sight and quickly cures a multitude of diseases that arise above the clavicles.
\end{translation}

\begin{sources}[hp02_031]
\end{sources}

\begin{testimonia}[hp02_031]
Yogacintāmaṇi

\begin{versinnote}
\tl{kapālaśodhanī caiva divyadṛṣṭipradīpinī |\\+}
\tl{jatrūrdhvajātarogaughān jarayaty āśu netivit ||\\!} % cf. J6
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{kapālaśodhinī kāryā divyadṛṣṭipradāyinī | [caiva -P]\\+}
\tl{jatrūrdhvajātarogaghnī jāyate netir uttamā || 1.42 ||\\!} % cf. M1
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_031]
V1 has \emph{kāryā} in the first \emph{pāda}, which is unnecessary because of the main verb in the final \emph{pāda}. Therefore, \emph{caiva} is easier to construe in the first \emph{pāda}.

Different readings exist for the last pāda. Most witnesses have \emph{netir āśu nihaṃti ca}. V1's \emph{jayati sā tu sūtrikā} is possible. The variants with \emph{netivit} seem implausible because of the epithets in the first hemistich, which require \sl{neti} to be understood as the subject of the sentence.
\end{philcomm}

\subsection*{2.32}
\begin{translation}[hp02_032]
The focused [yogi] should look at a small focal point with unmoving gaze until tears fall. This is taught as trāṭaka by teachers.
\end{translation}

\begin{sources}[hp02_032]
\end{sources}

\begin{testimonia}[hp02_032]
Haṭharatnāvalī

\begin{versinnote}
\tl{nirīkṣya niścaladṛśā sūkṣmalakṣyaṃ samāhitaḥ |\\+}
\tl{aśrusampātaparyantam ācāryais trāṭakaṃ smṛtam || 1.54 ||\\!}
\end{versinnote}

Yogacintāmaṇi

\begin{versinnote}
\tl{atha trāṭakam |\\+}
\tl{nirīkṣen niścaladṛśā sūkṣmalakṣyaṃ samāhitaḥ |\\+}
\tl{aśruprapātaparyantam āryais tat trāṭakaṃ matam ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_032]

\end{philcomm}

\subsection*{2.33}
\begin{translation}[hp02_033]
It is the destroyer of eye diseases and the door [blocking] sloth and so forth. Trāṭaka should be carefully concealed just as chest of gold.
\end{translation}

\begin{sources}[hp02_033]
\end{sources}

\begin{testimonia}[hp02_033]
Yogacintāmaṇi

\begin{versinnote}
\tl{moṭanaṃ netrarogānāṃ tandrādīnāṃ kapāṭakam |\\+}
\tl{etac ca trāṭakaṃ gopyaṃ yathā hāṭakapeṭakam ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{sphoṭanaṃ netrarogāṇāṃ tandrādīnāṃ kapāṭakam |\\+}
\tl{prayatnāt trāṭakaṃ gopyaṃ yathā ratnasupeṭakam || 1.55 ||\\!}
\end{versinnote}

Yuktabhavadeva 6.159

\begin{versinnote}
\tl{moṭakaṃ sarvarogāṇāṃ tandrādīnāṃ kapāṭanam ||\\+}
\tl{yatnatas trāṭakaṃ gopyaṃ yathā hāṭakapeṭakam || 159 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_033]
The reading \emph{moṭakaṃ} is found in V1 and the \emph{Yuktabhavadeva}. We understand it to mean “destroyer”. Bohtlingk and Roth (s.v.) and Monier-Williams (s.v.) give medicinal pill as a possible meaning of \emph{moṭaka} (cf. \emph{modaka}) but it appears that this is mainly an inference drawn only from this verse, where the pill is merely a comparison. Several witnesses have \emph{sphoṭanaṃ} (‘destroying’), which is also possible. The reading \emph{kapāṭakam} in pāda b is found in most of the witnesses and testimonia and we have adopted it accordingly. The context indicates that it must mean “blocker” (perhaps, in the sense of a door that blocks something), but we have not found any parallel usages of it in this sense.
\end{philcomm}

\subsection*{2.34}
\begin{translation}[hp02_034]
With the shoulders lowered, one should rotate the stomach to the left and right With speed of a rapid whirlpool. This is called Nauli by people from Gauḍa.
\end{translation}

\begin{sources}[hp02_034]
\end{sources}

\begin{testimonia}[hp02_034]
Yogacintāmaṇi

\begin{versinnote}
\tl{amandāvartavegena tundaṃ savyāpasavyayoḥ |\\+}
\tl{natāṃso bhrāmayed eṣā naulī yoge pracakṣate ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{atha nauliḥ--\\+}
\tl{amandāvartavegena tundaṃ savyāpasavyataḥ |\\+}
\tl{natāṃso bhrāmayed eṣā nauliḥ gauḍaiḥ praśasyate || 1.34 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_034]
\emph{āmandācakravegena} could mean the speed of a slightly slow wheel (Jürgen’s suggestion). But this is strange so we have adopted the usual metaphor of a fast moving whirlpool. The reading \emph{gauḍaiḥ} is found in only two witnesses but none of the others makes sense apart from the \emph{Jyotsnā}’s \emph{siddhaiḥ} which is likely to be a correction.

\end{philcomm}

\subsection*{2.35}
\begin{translation}[hp02_035]
Naulī brings about the stimulation of sluggish bodily fire, digestion and the like, always brings bliss, and makes all humoural disorders and diseases wither away. Naulī is the best of all Haṭha practices.
\end{translation}

\begin{sources}[hp02_035]
\end{sources}

\begin{testimonia}[hp02_035]
HTK

\begin{versinnote}
\tl{mandāgnisandīpanapācanādisandhāyikānandakarī sadaiva |\\+}
\tl{aśeṣadoṣopacayaśoṣaṇīva haṭhakriyā 'sau jayatīha nauliḥ || 12 ||\\!}
\end{versinnote}

Yogacintāmaṇi

\begin{versinnote}
\tl{mandāgnisaṃdīpanapācanāgnisaṃdhāyikānandakarī tathaiva |\\+}
\tl{aśeṣadoṣāmayaśoṣinī ca haṭhakriyāmaulir iyaṃ hi naulī ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{tundāgnisandīpanapācanādisandīpikānandakarī sadaiva |\\+}
\tl{aśeṣadoṣāmayaśoṣaṇī ca haṭhakriyāmaulir iyaṃ ca nauliḥ || 1.35 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_035]
We should adopt \emph{pācana} in the sense of digestion (cooking the food in the intestines). \emph{āmaya} makes better sense in the third pāda, and \emph{mala} might have crept in because of the association of this word with \emph{śoṣiṇī}.

Most witnesses and the testimonia have \emph{maulir iyaṃ} in the fourth \emph{pāda}, which expresses the idea that \emph{nauli} was thought to be the best of the \emph{ṣaṭkarma}, and the assonance of \emph{nauli} and \emph{mauli} may have been intended. The reading of \emph{mūlam iyaṃ} in V1 would suggest that \emph{nauli} is necessary for the other practices, which does not seem to be the case.
\end{philcomm}

\subsection*{2.36}
\begin{translation}[hp02_036]
Like the bellows of a blacksmith, the inhalation and exhalation are fast. It is called kapālabhāti. It dries up all phlegmatic disorders.
\end{translation}

\begin{sources}[hp02_036]
\end{sources}

\begin{testimonia}[hp02_036]
Yogacintāmaṇi

\begin{versinnote}
\tl{atha kapālabhātī |\\+}
\tl{bhastreva lohakārāṇāṃ recapūrau sasaṃbhramau |\\+}
\tl{kapālabhātī vikhyātā kaphadoṣaviśoṣiṇī ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{atha kapālabhastrikā--\\+}
\tl{bhastrival lohakārāṇāṃ recapūrasusambhramau |\\+}
\tl{kapālabhastrī vikhyātā sarvarogaviśoṣaṇī || 1.56 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_036]
The term \emph{bhātī} is derived from \emph{bhastrī} (Turner 1966: 537, entry 9424). The ‘skull bellows’ is the implication of the name. The J10 group has \emph{savyāpasavyataḥ} (‘left and right’) but this doesn’t make sense as this version of the verse does not stipulate what is moving to the left and right.
\end{philcomm}

\subsection*{2.37}
\begin{translation}[hp02_037]
One whose excess weight, phlegm, fat, impurity and the like have been removed by the ṣaṭkarma, should then do prāṇāyāma. It is accomplished without exertion.
\end{translation}

\begin{sources}[hp02_037]
\end{sources}

\begin{testimonia}[hp02_037]
Yogacintāmaṇi

\begin{versinnote}
\tl{tathā cātmārāmaḥ—\\+}
\tl{ṣaṭkarmanirgatasthaulyakaphamedogadādikaḥ |\\+}
\tl{prāṇāyāmaṃ tataḥ kuryād anāyāsena sidhyati ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_037]
Most witnesses support \emph{ṣaṭkarmabhir gata°}, which is somewhat unconventional syntax. It seems more likely that it was corrected to \emph{ṣaṭkarmanirgata°} than the latter being original. The variant reading \emph{°ādhikaḥ} in the second \emph{pāda} is inferior because one would assume that the \emph{ṣaṭkarma} should remove all excess weight (\emph{sthaulya}) and impurities (\emph{mala}). 
\end{philcomm}

\subsection*{2.38}
\begin{translation}[hp02_038]
Some teachers say that all the impurities are dried up by breath-controls alone and do not recommend any other practice.
\end{translation}

\begin{sources}[hp02_038]
\end{sources}

\begin{testimonia}[hp02_038]
Yogacintāmaṇi

\begin{versinnote}
\tl{prāṇāyāmair eva \textup{<2 syllables missing>} praśuṣyanti malā iti |\\+}
\tl{keṣāṃ cid ācāryāṇām anya karma na saṃmatam iti ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_038]
The plural of \emph{prāṇāyāma}, which we have translated here as ‘breath-controls’, probably refers to practising multiple repetitions of breath retentions with alternate nostril breathing. The verb \emph{praśuṣyanti} is well attested and makes good sense in the context of \emph{mala}s. Many witnesses lower on the stemma have \emph{malāśaya} (instead of \emph{malā api} or \emph{malā iti}) but this usually has the more specific meaning of bowels or bladder and so seems inappropriate in a general statement. The witnesses that have \emph{malāśaya} also have the verb \emph{pra+śudh}, which connotes that the place where the impurities accumulate is cleaned (rather than the impurities themselves). 
\end{philcomm}

\subsection*{2.39}
\begin{translation}[hp02_039]
\end{translation}

\begin{sources}[hp02_039]
\end{sources}

\begin{testimonia}[hp02_039]
Haṭharatnāvalī

\begin{versinnote}
\tl{brahmādayo 'pi tridaśāḥ pavanābhyāsatatparaḥ ||\\+}
\tl{abhūvan mṛtyurahitā tasmāt pavanam abhyaset || 3.82 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_039]
\emph{ṣaṭkarmayoga} is somewhat strange and this hemistich is omitted in V1. Perhaps, yoga was intended as ‘method’. It seems likely that the second and fourth hemistiches (as found in V1) were original. \emph{samanaska} is strange and difficult to construe (Jim thinks it might refer to death with the mind active as opposed to the preferable situation of dying while in samādhi). However, the verbs \emph{abhuvann} and \emph{amucann} make better sense. The third hemistich may be an attempt to rewrite the pāda to fix the problem of \emph{samanaska}.
\end{philcomm}

\subsection*{2.40}
\begin{translation}[hp02_040]
So long as the breath is bound in the body, the mind without support, and one looks at the middle of the brow, where is the fear of death?
\end{translation}

\begin{sources}[hp02_040]
VM

\begin{versinnote}
\tl{yāvad baddho marud dehe tāvac cittaṃ nirāśrayam |\\+}
\tl{yāvad dṛṣṭir bhruvor madhye tāvat kālabhayaṃ kutaḥ || 72 ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_040]
Yogacintāmaṇi

\begin{versinnote}
\tl{yāvad baddho marud dehe yāvad vṛttau nirāśrayam |\\+}
\tl{yāvad dṛṣṭir bhruvor madhye tāvat kālabhayaṃ kutaḥ ||\\!}
\end{versinnote}
Yuktabhavadeva 7.8

\begin{versinnote}
\tl{yāvad baddho maruddehe yāvaccittaṃ nirāmayam |\\+}
\tl{yāvaddṛṣṭarbhruvormadhye tāvat kālabhayaṃ kutaḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_040]
The verb \emph{paśyet} in the third \emph{pāda} is well attested among the manuscripts (V1, J10, etc.). However, it often occurs with \emph{bhruvor madhye}, which is rather odd and suggests that the alternative reading \emph{dṛṣṭir} was original. In fact, \emph{dṛṣṭir} is supported by the manuscripts of the source text, the \emph{Vivekamārtaṇḍa}, and the testimonia. 
\end{philcomm}

\subsection*{2.41}
\begin{translation}[hp02_041]
When the network of channels has been purified by breath-controls as prescribed, the breath pierces and easily enters the mouth of suṣumṇā.
\end{translation}

\begin{sources}[hp02_041]
\end{sources}

\begin{testimonia}[hp02_041]
Yogacintāmaṇi

\begin{versinnote}
\tl{haṭhapradīpikāyām—\\+}
\tl{vividhaiḥ prāṇasaṃyāmaiḥ nāḍīcakre viśodhite |\\+}
\tl{suṣumnāvadanaṃ bhitvā sukhād viśati mārutaḥ ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{vidhivat prāṇasaṃyāmaiḥ nāḍicakre viśodhite |\\+}
\tl{suṣumnāvadanaṃ bhitvā sukhād viśati mārutaḥ || 2.2 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_041]
The compound \emph{prāṇasaṃyāmaiḥ} (‘breath-controls’) should be understood here as a synonym for \emph{prāṇāyāmaiḥ}  in 2.38 (see the note on this verse).
\end{philcomm}

\subsection*{2.42}
\begin{translation}[hp02_042]
When the breath moves in the central channel, stillness of the mind arises. The mind’s becoming very still is the ‘mind beyond mind’ (\emph{manonmanī}) state.
\end{translation}

\begin{sources}[hp02_042]
\end{sources}

\begin{testimonia}[hp02_042]
Yogacintāmaṇi

\begin{versinnote}
\tl{mārute madhyasaṃcāre manaḥsthairyaṃ prajāyate |\\+}
\tl{yo manaḥsusthirībhāvaḥ saivāvasthā manonmanīti ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{mārute madhyame jāte manaḥsthairyaṃ prajāyate |\\+}
\tl{manasaḥ susthirībhāvaḥ saivāvasthā manonmanī || 2.3 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_042]
\end{philcomm}

\subsection*{2.43}
\begin{translation}[hp02_043]
In order to achieve that, those who know [their] methods perform various retentions. As a result of the practice of the various different retentions, [the yogi] obtains various different results.
\end{translation}

\begin{sources}[hp02_043]
\end{sources}

\begin{testimonia}[hp02_043]
YBD

\begin{versinnote}
\tl{tatsiddhaye vidhānajñāś citrān kurvanti kumbhakān ||\\+}
\tl{vicitrakumbhakābhyāsād vicitrāṃ siddhim āpnuyāt || 92 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_043]

\end{philcomm}

\subsection*{2.44}
\begin{translation}[hp02_044]
Sūryabhedana, ujjāyī, sītkāra, [etc.] and kevala are the eight kumbhakas.
\end{translation}

\begin{sources}[hp02_044]
\end{sources}

\begin{testimonia}[hp02_044]
Yogacintāmaṇi

\begin{versinnote}
\tl{haṭhayoge—\\+}
\tl{sūryabhedanam ujjāyī tathā sītkāraśītalī |\\+}
\tl{bhastrikā bhramarī mūrcchā sahitaṃ cāṣṭa kumbhakāḥ ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{sūryabhedanam ujjayī tathā sītkāraśītalī |\\+}
\tl{bhastrikā bhrāmarī mūrcchā kevalaś cāṣṭakumbhakāḥ || 2.6 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_044]
In the fourth \emph{pāda}, the manuscripts of group 2 have \emph{plāvanīty aṣṭa}, which makes good sense in so far as most manuscripts contain a verse on \emph{plāvinīkumbhaka} as one of the kumbhakas. Furthermore, the alternative reading \emph{kevalī/kevalaś}, which is supported by V1, J10, etc., is not consistent with the idea of \emph{kevalakumbhaka} standing outside the category of \emph{sahitakumbhaka}, as stated in 2.72–75. This idea of two types of kumbhaka is found in the source text from which Svātmārāma borrowed four of the kumbhakas, namely, the \emph{Gorakṣaśataka}. Therefore, it seems likely the word \emph{plāvinī} dropped out of some manuscripts. In fact, it may have been removed by some scribes because its heading is omitted in many witnesses. The reading in V19 and the \emph{Yogacintāmaṇi} (\emph{sahitāś cāṣṭa}) was probably an attempt to remove \emph{kevalī/kevalaś} from the list.
\end{philcomm}

\subsection*{2.45}
\begin{translation}[hp02_045]
At the end of the inhalation, the lock called Jālandhara should be done. However, at the end of the retention and beginning of the exhalation, uḍḍiyāna should be done.
\end{translation}

\begin{sources}[hp02_045]
Gorakṣaśataka 62ab

\begin{versinnote}
\tl{pūrakānte tu kartavyo bandho jālandharābhidhaḥ |\\!}
\end{versinnote}
Gorakṣaśataka 58ab

\begin{versinnote}
\tl{kuṃbhakānte recakādau kartavyoḍḍiyaṇābhidhaḥ |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_045]
Yogacintāmaṇī

\begin{versinnote}
\tl{pūrakānte tu kartavyo bandho jālandharābhidhaḥ |\\+}
\tl{kumbhakānte recakādau kartavyas tūḍḍiyānakaḥ ||\\!}
\end{versinnote}

Haṭharatnāvalī 2.7

\begin{versinnote}
\tl{pūrakānte tu kartavyo bandho jālandharābhidhaḥ |\\+}
\tl{kumbhakānte recakādau kartavyas tūḍḍiyānakaḥ ||\\!}
\end{versinnote}

Yuktabhavadeva 7.94

\begin{versinnote}
\tl{pūrakānte ca karttavyo bandho jālandharābhidhaḥ |\\+}
\tl{kumbhakānte recakādau karttavyas tūḍḍiyānakaḥ || \\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_045]
\end{philcomm}

\subsection*{2.46}
\begin{translation}[hp02_046]
When the contraction of the throat has been applied, the breath goes into the central channel by quickly contracting below [the abdomen] and stretching back the middle [of the body].
\end{translation}

\begin{sources}[hp02_046]
\end{sources}

\begin{testimonia}[hp02_046]
Yogacintāmaṇī

\begin{versinnote}
\tl{adhas tv ākuñcanenāśu kaṇṭhasaṅkocanena ca |\\+}
\tl{madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_046]
\emph{madhyapaścimatānena} is possible and well attested.
\end{philcomm}

\subsection*{2.47}
\begin{translation}[hp02_047]
Having raised up \emph{apānavāyu}, one should move \emph{prāṇa} down from the throat. The yogi is freed from old age and becomes sixteen years old.
\end{translation}

\begin{sources}[hp02_047]
\end{sources}

\begin{testimonia}[hp02_047]
Yogacintāmaṇī

\begin{versinnote}
\tl{apānam ūrdhvam utthāpya prāṇaṃ kaṇṭhād adho nayet |\\+}
\tl{yogī jarāvimuktaḥ san vayasā ṣoḍaśo bhavet ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_047]
\end{philcomm}

\subsection*{2.48}
\begin{translation}[hp02_048]
Now, piercing the sun---
Having adopted vajrāsana on a comfortable mat, the yogi should gradually draw in the external air through the right nostril,
\end{translation}

\begin{sources}[hp02_048]
\end{sources}

\begin{testimonia}[hp02_048]
Yogacintāmaṇi

\begin{versinnote}
\tl{āsane sukhade yogī baddhavajrāsanas tataḥ |\\+}
\tl{dakṣanāḍyā samākṛṣya bahisthaṃ pavanaṃ śanaiḥ ||\\!}
\end{versinnote} 
\end{testimonia}

\begin{philcomm}[hp02_048]
\emph{āsane sukhade} is supported by the [different but parallel] reading in the Gorakṣa\-śataka.
\end{philcomm}

\subsection*{2.49}
\begin{translation}[hp02_049]
[The yogi] does \emph{kumbhaka} until the cessation [of the breath] as far as the tips of the hair and nails. The wise yogi should then exhale the breath slowly through the left nostril.
\end{translation}

\begin{sources}[hp02_049]
\end{sources}

\begin{testimonia}[hp02_049]
Yogacintāmaṇi 
\begin{versinnote} 
\tl{ākeśāgraṃ nakhāgraṃ ca śirodhāvadhi kumbhakam |\\+}
\tl{tataḥ śanaiḥ savyanāḍyā recayet pavanaṃ sudhīḥ ||\\!}
\end{versinnote}

Yuktabhavadeva 7.99

\begin{versinnote}
\tl{ā keśād ā nakhāgrāc ca nirodhāvadhi kumbhayet |\\+}
\tl{tataḥ śanaiḥ savyanāḍyā recayet pavanaṃ sudhīḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_049]
Refer to Brahmānanda’s commentary for an explanation of \emph{ākeśāgram nakhāgraṃ ca}. It seems that one should hold the breath within the body. If it escapes through the hair follicles or nails then it destroys the body. The ablative with \emph{ā} (\emph{a keśād nakhāgrāc ca}) is also well-attested (group 2 manuscripts), and it would mean the same.

Cf. BDhS 4.1.23
\begin{versinnote}
\tl{āvartayet sadā yuktaḥ prāṇāyāmān punaḥpunaḥ |\\+}
\tl{ā keśāntān nakhāgrāc ca tapas tapyata uttamam ||\\!}
\end{versinnote}

\emph{nirodhāvadhi} is not entirely clear, but all sources and the \emph{Jyotsnā} agree on this reading. The problem is that to practise \emph{kumbhaka} ``up to cessation \emph{nirodha}'' seems to suggest that cessation is not of the physical breath (which by definition ceases in \emph{kumbhaka}), but, if this explanation is not too far-reaching, of the \emph{prāṇa} within the body.

Perhaps, \emph{nirodhāvadhi} is referring to stopping the flow of \emph{prāṇa} at the extremities of the body so that it does not damage the body by exiting through the hair follicles. This idea is alluded to in the following verse quoted in the \emph{Jyotsnā}:

\begin{versinnote}
\tl{haṭhān niruddhaḥ prāṇo'yaṃ romakūpeṣu niḥsaret |\\+}
\tl{dehaṃ vidārayaty eṣa kuṣṭhādi janayaty api ||\\!}
\end{versinnote}
\end{philcomm}

\subsection*{2.50}
\begin{translation}[hp02_050]
This purifies the skull, cures [imbalances] of vātadoṣa [and] gets rid of diseases caused by worms [so] should be done repeatedly. It is called the piercing of the sun.
\end{translation}

\begin{sources}[hp02_050]
Gorakṣaśataka 35a-36b

\begin{versinnote}
\tl{kapālaśodhane vāpi recayet pavanaṃ sudhīḥ |\\+}
\tl{tundasya vātadoṣaghnaḥ kṛmidoṣaṃ nihanti ca ||\\+}
\tl{punaḥ punar idaṃ kāryaṃ sūryābhedam udāhṛtam |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_050]
Yogacintāmaṇi 
\begin{versinnote} 
\tl{kapālaśodhanaṃ vātadoṣaghnaṃ kṛmidoṣahṛt |\\+}
\tl{punaḥ punar idaṃ kuryāt sūryabhedanam uttamam ||\\!}
\end{versinnote}

Haṭharatnāvalī 2.11cd–12

\begin{versinnote}
\tl{kapālaṃ śodhanaṃ cāpi recayet pavanaṃ śanaiḥ ||\\+}
\tl{[kapālaśodhanaṃ … kṛmināśanaṃ – N,n1,n4]\\+}
\tl{ālasyaṃ vātadoṣaghnaṃ kṛmikīṭaṃ nihanti ca |\\+}
\tl{punaḥ punar idaṃ kāryaṃ sūryabhedākhyakumbhakam ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_050]
The \emph{Gorakṣaśataka} and \emph{Yogacintāmaṇi} support \emph{kṛmidoṣa}. The compound \emph{sūryabheda} is metri causa. The J10 group and V19 have attempted to reinstate the name \emph{sūryabhedana}.
%Add to introduction note on action nouns in °nam etc. being used as adjectives.
\end{philcomm}

\subsection*{2.51}
\begin{translation}[hp02_051]
Having closed the mouth, one should gradually draw in the breath through the nostrils so that it comes into contact with the throat as far as the chest and resonates.\end{translation}

\begin{sources}[hp02_051]
Gorakṣaśataka

\begin{versinnote}
\tl{mukhaṃ saṃyamya nāḍībhyāṃ ākṛṣya pavanaṃ śanaiḥ |\\+}
\tl{yathā lagati kaṇṭhāt tu hṛdayāvadhi sasvanam || [T kaṇṭhaṃ tu]\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_051]
Yogacintāmaṇi

\begin{versinnote}
\tl{mukhaṃ saṃyamya nāḍībhyām ākṛṣya pavanaṃ punaḥ |\\+}
\tl{yathā lagati hṛtkaṇṭhād dhṛdayāvadhi sasvanaḥ ||\\!}
\end{versinnote}
Haṭharatnāvalī

\begin{versinnote}
\tl{mukhaṃ saṃyamya nāḍībhyām ākṛṣya pavanaṃ śanaiḥ |
[nāsābhyām J,n1,n4]\\+}
\tl{yathā lagati hṛtkaṇṭhaṃ hṛdayāvadhi svasvanaḥ || 2.13 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_051]
\end{philcomm}

\subsection*{2.52}
\begin{translation}[hp02_052]
As before, he should hold the breath and then exhale through the \emph{iḍā} channel. It cures diseases caused by phlegm in the throat and increases the body’s fire.
\end{translation}

\begin{sources}[hp02_052]
GŚ

\begin{versinnote}
\tl{pūrvavat kumbhayet prāṇaṃ recayed iḍayā tataḥ |\\+}
\tl{śīrṣotthitānalaharaṃ galaśleṣmaharaṃ paraṃ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_052]
Yogacintāmaṇi

\begin{versinnote}
\tl{pūrvavat kumbhayet prāṇaṃ recayed iḍayā tataḥ |\\+}
\tl{śleṣmadoṣaharaṃ kaṇṭhe dehānalavivardhanam ||\\!}
\end{versinnote}
Haṭharatnāvalī

\begin{versinnote}
\tl{pūrvavat kumbhayet prāṇaṃ recayed iḍayā tataḥ |\\+}
\tl{gale śleṣmaharaṃ proktaṃ dehānalavivardhanam || 2.14 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_052]
In the second hemistich of this verse, many of the readings in the oldest manuscripts, such as \emph{dehād analadīptivardhanam} (V1), \emph{dehānaladīptivivardhanam} (J10) and \emph{dehe [’]naladīptivi vardhanam} (P28), are unlikely to be original because both the source and testimonia indicate that \emph{Ujjāyī} is supposed to remove phlegm from the throat. These versions may have arisen from attempts to remove \emph{kaṇṭhe} in the third \emph{pāda}, which was thought to be hanging. Therefore, it seems that \emph{kaṇṭhe dehānalavardhanam} is the better reading and it is well attested among the manuscripts (including group 2).
\end{philcomm}

\subsection*{2.53}
\begin{translation}[hp02_053]
It cures diseases in the bodily constituents inside the network of nāḍīs,
The retention called ujjāyī should be done by one moving or stationary.
\end{translation}

\begin{sources}[hp02_053]
Gorakṣaśataka 38

\begin{versinnote}
\tl{nāḍījalodarādhātugatadoṣavināśanam |\\+}
\tl{gacchatas tiṣṭhataḥ kāryam ujjāyyākhyaṃ ca kumbhakam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_053]
Yogacintāmaṇi

\begin{versinnote}
\tl{nāḍījalodaradhātugatadoṣavināśanam |\\%check collation of YCM to see if mss have the unmetrical °odara°
gacchatas tiṣṭhataḥ kāryam ujjāyyākhyaṃ ca kumbhakam ||\\!}
\end{versinnote}
Haṭharatnāvalī 2.15

\begin{versinnote}
\tl{nāḍījālodarādhātugatadoṣavināśanam |\\+}
\tl{\textup{[}nāḍījalodaradhātu- \textup{J,P;} nāḍījalodaraṃ dhātu- \textup{N, n1, n4]}\\+}
\tl{gacchatā tiṣṭhatā kāryam ujjāyyākhyaṃ hi kumbhakam || \\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_053]
Nearly all the manuscripts have \emph{nāḍījalodara}, which does not make sense in this context because it connotes that a disease or humoral imbalance (\emph{doṣa}) exists in the disease ascites (\emph{jalodara}). This problem is also present in the transmission of the source text for the verse, the \emph{Gorakṣaśataka}. A solution can be found in some of the manuscripts of the \emph{Haṭharatnāvalī}, which read \emph{nāḍījālodarā°} (‘in the network of channels and stomach’). The other problem is \emph{°darādhātu°}. In spite of Brahmānanda’s efforts to explain it as \emph{°dara}, \emph{ā}, \emph{dhātu°}, the \emph{ā} before \emph{dhātu°} appears to have been inserted for metri causa. We have adopted \emph{nāḍījālodare} (‘inside the network of channels’), which we understand to have been changed in the transmission to \emph{°jalodara}, through confusion with the disease of a similar name.

For the compound \emph{dhātudoṣa} see \emph{Tantrāloka} 28.283cd, where worldly concepts are said to arise from it (\emph{dhātudoṣāc ca saṃsārasaṃskārās te ...}), but it is also the source of physical disorders (\emph{dhātudoṣakṛtaṃ mūrcchā} ĪPV on 2.15).

Diwakar Acharya suggests that the name \emph{ujjāyī} may be a Prakrit form from \emph{uddhmāyī} from the verb \emph{ud-dhmā}, “to blow out”.

\end{philcomm}

\subsection*{2.54}
\begin{translation}[hp02_054]
Now śītkāra---
He should constantly produce a \emph{śīt} sound in the mouth and a flaring of the nostrils. By practising thus, one becomes a second god of love.
\end{translation}

\begin{sources}[hp02_054]
Cf. Kaulajñānanirṇaya ??

\begin{versinnote}
\tl{cittan dadyāt tu vaktreṇa nāse dadyād vijṛmbhikā[m] ||\\+}
\tl{vācāsiddhir bhavaty evaṅ kāmadevo ’paraḥ priyaḥ |\\!}
\end{versinnote}

Cf. BHU Ms. of the Jñānasāra

\begin{versinnote}
\tl{hikkā dadyāt sadā vaktre prāyaś caiva vijṛmbhikām ||\\+}
\tl{evamabhyasyamānas tu kāmadevo dvitīyakaḥ |\\!}
\end{versinnote}

Prāṇatoṣiṇī (citing the Jñānasāra)

\begin{versinnote}
\tl{hikkāṃ dadyāt sadā vaktre ghrāṇañ caiva vijṛmbhate |\\+}
\tl{evam abhyāsayogena kāmadevo dvitīyakaḥ |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_054]
Yogacintāmaṇi

\begin{versinnote}
\tl{sītkāṃ kuryāt tathā vaktre ghrāṇenaiva visarjayet |\\+}
\tl{evam abhyāsayogena kāmadevo dvitīyakaḥ ||\\!}
\end{versinnote}

Haṭharatnāvalī 2.16

\begin{versinnote}
\tl{sītkāṃ kuryāt tathā vaktre ghrāṇenaiva visarjayet |\\+}
\tl{evam abhyāsayogena kāmadevo dvitīyakaḥ || 2.16 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_054]
The source texts and V1 have \emph{dadyāt} in the first \emph{pāda}, which seems to be the lectio difficilior. Many of the old manuscripts have \emph{kumbhaṃ} instead of \emph{sītkāṃ}. The latter reading is supported by some manuscripts of the \emph{Haṭhapradīpikā}, the testimonia and the name of the \emph{kumbhaka}, which is stated in 2.44 and the heading of this verse. Various readings for first \emph{pāda} are also seen in the source texts, namely the \emph{Kaulajñānanirṇaya} and \emph{Jñānasāra}.There is also a discrepancy over \emph{śīt} and \emph{sīt}. The result of becoming a second god of love may be connected with the sound \emph{sītkāra}, which is one of the sounds made during sex in \emph{Kāmaśāstra} (ref?).

\end{philcomm}

\subsection*{2.55}
\begin{translation}[hp02_055]
Having become one among the circle of yoginis, he can bring about creation and destruction. Neither hunger nor thirst [nor] sleep nor indolence arise.
\end{translation}

\begin{sources}[hp02_055]
Cf. Kaulajñānanirṇaya 7.18ab

\begin{versinnote}
\tl{yoginīgaṇasāmānyā sṛṣṭisaṃhārakārakaḥ |\\!}
\end{versinnote}

Jñānasāra 13cd–14ab

\begin{versinnote}
\tl{yoginīguṇasāmānyaḥ sṛṣṭisaṃhārakārakaḥ || 13 ||\\+}
\tl{na kṣudhā na ca tṛṇnidrā naiva murchā prajāyate |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_055]
Yogacintāmaṇi

\begin{versinnote}
\tl{yoginīcakrasaṃsevyaḥ sṛṣṭisaṃhārakārakaḥ |\\+}
\tl{na kṣudhā na tṛṣṇā nidrā tandrālasyaṃ na jāyate ||\\!}
\end{versinnote}

Haṭharatnāvalī 2.17

\begin{versinnote}
\tl{yoginīcakrasaṃsevyaḥ sṛṣṭisaṃhārakārakaḥ |\\+}
\tl{na kṣudhā na tṛṣā nidrā naivālasyaṃ prajāyate || 2.17 ||\\!}
\end{versinnote}

Yuktabhavadeva

\begin{versinnote}
\tl{yoginīcakrasāmānyaḥ sṛṣṭisthityantakārakaḥ ||\\+}
\tl{na kṣudhā na tṛṣā nidrā nālasya ca prajāyate || 105 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_055]
There are various possible readings for the first \emph{pāda}, namely, \emph{yoginīcakrasāmānyaḥ} (‘one who is one among the circle of yoginīs'), \emph{yoginīcakram āsādya} (‘having reached the circle of yoginis’) and \emph{yoginīcakrasaṃsevyaḥ} (‘one worshipped by the circle of yoginis’). The first of these is closer to the source texts and is a cliche in Kaula literature.
\end{philcomm}

\subsection*{2.56}
\begin{translation}[hp02_056]
His body is as he wishes, and he is free from all afflictions. By this method, he truly becomes the lord of yogis in the region of the earth.
\end{translation}

\begin{sources}[hp02_056]
Jñānasāra

\begin{versinnote}
\tl{bhavet svacchandadehas tu sarvopadravavarjitaḥ\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_056]
Yogacintāmaṇi

\begin{versinnote}
\tl{bhavet svachandadehas tu sarvopadravavarjitaḥ |\\+}
\tl{anena vidhinā yas tu yogīndro bhūmimaṇḍale ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{bhavet svacchandadehas tu sarvopadravavarjitaḥ |\\+}
\tl{anena vidhinā satyaṃ yogīndro bhāti bhūtale || 2.18 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_056]
The aiśa compound \emph{bhuvimaṇḍale}, which is attested at \emph{Mañjuśrīmūlakalpa} 45.221, is likely the original reading here. The word  \emph{bhuvi} as the first member of a compound is attested elsewhere. The alternative \emph{bhumi} is well-attested and so the change may have happened early in the transmission. 
\end{philcomm}

\subsection*{2.56*1}
\begin{translation}[hp02_056_1]
One who always takes in the breath through the aperture at the roots of the tongue, undoubtedly becomes a receptacle of all \emph{siddhi}s.
\end{translation}

\begin{philcomm}[hp02_056_1]
This verse appears to be a derivative of 2.57, and was not original to the text. Cf. \emph{Siddhāntamuktāvalī}
\begin{versinnote}
\tl{jihvāmūlena randhreṇa yaḥ prāṇaṃ satataṃ pibet |\\+}
\tl{sa bhavet sarvasiddhānāṃ bhājanaṃ nātra saṃśayaḥ || 46 ||\\!}
\end{versinnote}
\end{philcomm}

\subsection*{2.57}
\begin{translation}[hp02_057]
It has also been taught:
He who regularly takes in the breath through the root of the tongue and palate, has all his diseases cured in half a year.
\end{translation}

\begin{sources}[hp02_057]
Vivekamārtaṇḍa 120

\begin{versinnote}
\tl{rasanātālumūlena yaḥ prāṇam anilaṃ pibet |\\+}
\tl{abdārdhena bhavet tasya sarvarogaparikṣayaḥ ||\\!}
\end{versinnote}

Śivasaṃhitā 3.80

\begin{versinnote}
\tl{rasanāṃ tālumūle yaḥ sthāpayitvā vipaścitaḥ | \\+}
\tl{pibet prāṇānilaṃ tasya rogāṇāṃ saṃkṣayo bhavet ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_057]
Yogacintāmaṇi

\begin{versinnote}
\tl{rasanātāluyogena yaḥ prāṇaṃ satataṃ pibet |\\+}
\tl{abdārdhena bhavet tasya sarvarogaparikṣayaḥ ||\\!}
\end{versinnote}

Ānandakanda 1.20.137

\begin{versinnote}
\tl{jihvayā tālumūlena prāṇaṃ yaḥ pibati priye |\\+}
\tl{tasya ṣaṇmāsataḥ sarve rogā naśyanti yoginaḥ || \\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_057]
Verse 2.57 seems to be describing an alternative method of \emph{śītkārakumbhaka}. It could have been included by Svātmārāma. The introductory phrase \emph{uktaṃ ca} suggests that the teaching in this verse is consistent with what preceded it, but one would expect that it is another view (\emph{matāntare}).

The compound \emph{rasanātālumūlena} is difficult to understand. In his \emph{Haṭhasaṅketacandrikā} (ref??), Sundaradeva seems to think that the external air strikes (which might thus make the sound \emph{śīt}) the root of the tongue and palate and the upper part of the uvula (\emph{atra mūhū (muhu?) rasanātālumūlāhataṃ ghaṃṭikordhvabhāgāhataṃ bahiḥsthavāyuṃ vidhāya pibed ity arthaḥ}). More helpful are the remarks of the commentator of the \emph{Yogataraṅgiṇī}. He seems to think that a hole or cavity (\emph{vivara}) is made by the root of the palate with the help of the tongue, and the yogi breathes through it (\emph{evaṃ rasanātālumūlena rasanā jihvā tatsahāyabhūtatālumūlena kṛtaṃ yad vivaraṃ, tena kṛtvā yaḥ yogī prāṇam anilaṃ prāṇavāyuṃ pibet pūrayet, tasya yogino’bdārdhena ṣaṇmāsena sarvarogāṇāṃ nāśaḥ kṣayo bhavet} || 39 |).

This idea might have been intended with the parallel reading of the \emph{Yogacintāmaṇi}: \emph{rasanātāluyogena} (‘by connecting the tongue and palate'). So, we have translated \emph{rasanātālumūlena} as ‘through the tongue and root of the palate’. It’s a vague way of saying that the tongue is turned back to touch the root of the palate to make a hole that one breathes through (when the breath is taken in through the mouth). The \emph{Kumbhakapaddhati} (137) states this more clearly:

\emph{rasanām unmukhīkṛtya sītkāraṃ kurvatā marut} |\\
\emph{pīyante kumbhake yasmin nāsikābhyāṃ virecanam} || [pīyate?]

And this idea is also expressed clearly in \emph{Śivasaṃhitā} 3.80:

\emph{rasanāṃ tālumūle yaḥ sthāpayitvā vipaścitaḥ} |\\
\emph{pibet prāṇānilaṃ tasya rogāṇāṃ saṃkṣayo bhavet} ||

The verse may derive from \emph{Kauljñānanirṇaya} 6.19, which has \emph{rasanā[ṃ] tālumūle tu kṛtvā vāyuṃ pibet…}.
\end{philcomm}

\subsection*{2.58}
\begin{translation}[hp02_058]
Having drawn in the air through the tongue, the method of retention [is done] as before. The wise yogi gradually exhales through the nostrils.
\end{translation}

\begin{sources}[hp02_058]
Gorakṣaśataka

\begin{versinnote}
\tl{jihvayā vāyum ākṛṣya pūrvavat kuṃbhakād anu |\\+}
\tl{śanais tu ghrāṇarandhrābhyāṃ recayed anilaṃ sudhīḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_058]
Cf. Ānandakanda

\begin{versinnote}
\tl{kākacañcuvad āsyaṃ ca kṛtvā vāyuṃ sasūtkṛtam |\\+}
\tl{ādāya nāsārandhreṇa punastaṃ śvasanaṃ tyajet ||\\+}
\tl{śītalīkaraṇākhyo 'yaṃ yogas tu jvarapittahṛt |\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_058]
\end{philcomm}

\subsection*{2.59}
\begin{translation}[hp02_059]
Diseases such as swelling, enlargement of the spleen and the like, and fever, [excess] bile, hunger and thirst; this retention called śītalī cures them.
\end{translation}

\begin{sources}[hp02_059]
GŚ

\begin{versinnote}
\tl{gulmaplīhādikā doṣāḥ kṣayaṃ yānti pittaṃ jvaraṃ |\\+}
\tl{viṣāṇi śītalī nāma kuṃbhako 'yaṃ nihanti ca\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_059]
Yogacintāmaṇi

\begin{versinnote}
\tl{gulmaplīhodaraṃ cāpi vātapittaṃ kṣudhāṃ tṛṣām |\\+}
\tl{viṣāṇi śītalī nāma kumbhako vinihanti ca ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{gulmaplīhodaraṃ doṣaṃ jvarapittakṣudhātṛṣāḥ |\\+}
\tl{viṣāṇi śītalī nāma kumbhako 'yaṃ nihanti ca || 2.20 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_059]
\end{philcomm}

\subsection*{2.60}
\begin{translation}[hp02_060]
Now, bhastrikā---
If [the yogi] places the soles of both feet on the thighs, the lotus pose, which destroys all bad deeds, duly arises.
\end{translation}

\begin{sources}[hp02_060]
Gorakṣaśataka

\begin{versinnote}
\tl{ūrvor upari ced dhatte ubhe pādatale tathā |\\+}
\tl{padmāsanaṃ bhavet samyak sarvapāpapraṇāśanam || 14 ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_060]
Yogacintāmaṇi

\begin{versinnote}
\tl{bhastrikā—\\+}
\tl{ūrvor upari saṃsthāpya ubhe pādatale tathā ||\\+}
\tl{padmāsanaṃ bhavet samyak sarvapāpapraṇāśanam ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{atha bhastrikā---\\+}
\tl{recakaḥ pūrakaś caiva kumbhakaḥ praṇavātmakaḥ |\\+}
\tl{recako 'jasraniḥśvāsaḥ pūrakas tannirodhakaḥ |\\+}
\tl{samānasaṃsthito yo 'sau kumbhakaḥ parikīrtitaḥ || 2.21 ||\\!}
\end{versinnote}

Yuktabhavadeva

\begin{versinnote}
\tl{atha bhastrikā---\\+}
\tl{ūrvor upari cādhatte ubhe pādatale tathā ||\\+}
\tl{padmāsanaṃ bhavet samyak sarvapāpapraṇaśanam || 110 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_060]
The source text, the \emph{Gorakṣaśataka}, has \emph{ced} in the first pāda, and this has been dropped in nearly all of the available manuscripts of the \emph{Haṭhapradīpikā}, as well as the testimonia. It seems likely that \emph{ced dhatte} was the reading reading adopted by Svatmārāma because the \emph{cet} makes sense of the two finite verbs in the description. The first finite verb \emph{dhatte} is also supported by some of the old manuscripts, such as V1 and J10. At some stage, the verse was changed to read \emph{saṃsthāpya} to remove the awkward syntax posed by \emph{saṃdhatte} and \emph{vai dhatte}.
\end{philcomm}

\subsection*{2.61}
\begin{translation}[hp02_061]
Having adopted lotus pose correctly, the wise yogi whose neck and abdomen are straight should close the mouth and exhale the breath through the nose effortfully.
\end{translation}

\begin{sources}[hp02_061]
Gorakṣaśataka 41

\begin{versinnote}
\tl{tataḥ padmāsanaṃ baddhvā samagrīvodaraḥ sudhīḥ |\\+}
\tl{mukhaṃ saṃyamya yatnena prāṇaṃ ghrāṇena recayet |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_061]
Yogacintāmaṇi

\begin{versinnote}
\tl{samyak padmāsanaṃ badhvā samagrīvodaraḥ sudhīḥ |\\+}
\tl{mukhaṃ saṃyamya yatnena prāṇaṃ ghrāṇena recayet ||\\!}
\end{versinnote}

Yuktabhavadeva

\begin{versinnote}
\tl{samyak padmāsanaṃ baddhvā samagrīvodaraḥ śanaiḥ ||\\+}
\tl{mukhaṃ saṃyamya yatnena prāṇaṃ ghrāṇena recayet || 111 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_061]
\end{philcomm}

\subsection*{2.62}
\begin{translation}[hp02_062]
In such a way that [the breath] comes into contact with the chest and throat, and there is then a sound in the skull, he should quickly inhale a small amount of the breath as far as the heart lotus.
\end{translation}

\begin{sources}[hp02_062]
Gorakṣaśataka 43

\begin{versinnote}
\tl{yathā lagati kaṇṭhāt tu kapāle sasvanaṃ tataḥ\\+}
\tl{vegena pūrayet kiṃ cit hṛtpadmāvadhi mārutam\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_062]
Yogacintāmaṇi

\begin{versinnote}
\tl{yathā lagati hṛtkaṇṭhe kapālāvadhi pūrayet |\\+}
\tl{vegena pūrayet samyag hṛtpadmāvadhi mārutam ||\\!}
\end{versinnote}

Yuktabhavadeva

\begin{versinnote}
\tl{yathā lagati hṛtkaṇṭhakapāleṣu ca sasvanam ||\\+}
\tl{vegena pūrayet kiñcit hṛtpadmāvadhi mārutam || 112 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_062]
First hemistich is tricky. None of the old mss preserve \emph{sasvanaṃ}, which is in the GŚ and makes good sense. All witnesses have \emph{hṛtkaṇṭhe}, which we have understand as a dual accusative. One would expect \emph{kapāla} to also be in the accusative, but we have understood it as a locative sg. with \emph{sasvana}.
\end{philcomm}

\subsection*{2.63}
\begin{translation}[hp02_063]
Then, the yogi should exhale and inhale again and again. In the very same way as blacksmiths’ bellows are operated forcefully, [...]
\end{translation}

\begin{sources}[hp02_063]
Gorakṣaśataka 44

\begin{versinnote}
\tl{punar virecayet tadvat pūrayec ca punaḥ punaḥ |\\+}
\tl{yathaiva lohakārāṇāṃ bhastrā vegena cālyate ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_063]
Yogacintāmaṇi

\begin{versinnote}
\tl{punar virecayet tadvat pūrayitvā punaḥ punaḥ |\\+}
\tl{yathaiva lohakārāṇāṃ bhastrā vegena cālyate |\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{yathaiva lohakārāṇāṃ bhastrī vegena cālyate |\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{punar virecayet tadvat pūrayec ca punaḥ punaḥ ||\\+}
\tl{yathaiva lohakārāṇāṃ bhastrā vegena cālyate || 113 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_063]
V1 reading of \emph{lohakāreṇa} fits well with the passive verb, but it is the only witness to have this, and appears to be an attempt to improve what was probably the original reading \emph{lohakārāṇāṃ} (as attested by the group 2 manuscripts, the source and testimonia).
\end{philcomm}

\subsection*{2.64}
\begin{translation}[hp02_064]
[...] the wise [yogi] should move the breath in the body. When fatigue arises in the body, he should inhale by way of the sun
\end{translation}

\begin{sources}[hp02_064]
Gorakṣaśataka 45

\begin{versinnote}
\tl{tathaiva svaśarīrasthaṃ cālayet pavanaṃ sudhīḥ |\\+}
\tl{yadā śramo bhaved dehe tadā sūryeṇa pūrayet |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_064]
Yogacintāmaṇi

\begin{versinnote}
\tl{tathaiva svaśarīrasthaś cālyate pavano dhiyā |\\+}
\tl{yathā śramo bhaved dehe tathā vegena pūrayet |\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{tathaiva svaśarīrasthaṃ cālayet pavanaṃ sudhīḥ || 2.22 ||\\+}
\tl{yathā śramo bhaved dehe tathā sūryeṇa pūrayet |\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{tathaiva svaśarīrasthaṃ cālayet pavanaṃ dhiyā ||\\+}
\tl{yadā śramo bhaved dehe tadā sūryeṇa recayet || 114 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_064]
Most of the manuscripts support \emph{dhiyā} but the manuscripts of the source text and the testimonia support \emph{sudhīḥ}. Since the subject of the simile is \emph{bhastrā}, one would expect the subject of \emph{cālayet}, which must be different, to be stated (as is the case with \emph{sudhīḥ}). Also, one would expect the instrumental of \emph{dhī}to be qualified by some adjective, such as in the case of \emph{sattvāsthayā dhiyā} (Gorakṣaśataka 74b) and \emph{sāttvikayā dhiyā} (Haṭhapradīpikā 2.6b).
\end{philcomm}

\subsection*{2.65}
\begin{translation}[hp02_065]
in such a way that the abdomen is filled quickly by the breath and  hold the nose firmly without using the middle and index fingers.
\end{translation}

\begin{sources}[hp02_065]
Gorakṣaśataka 45cd–46ab

\begin{versinnote}
\tl{yathodaraṃ bhavet pūrṇaṃ pavanena tathā laghu |\\+}
\tl{dhārayan nāsikā madhyaṃ tarjanībhyāṃ vinā dṛḍhaṃ |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_065]
Yogacintāmaṇi

\begin{versinnote}
\tl{yathodaraṃ bhavet pūrṇaṃ pavanena tathā laghu |\\+}
\tl{dhārayen nāsikāṃ madhyatarjanībhyāṃ vinā dṛḍham ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{yathodaraṃ bhavet pūrṇaṃ pavanena tathā laghu || 2.23 ||\\+}
\tl{dhārayen nāsikāṃ madhyātarjanībhyāṃ vinā dṛḍham |
\var{23c madhyā ] madhye vl}\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_065]
Only two witnesses have \emph{madhyātarjanībhyāṃ} and it is not well attested by the manuscripts of the source text and testimonia. To hold the nose without the middle and index fingers is consistent with the way alternate nostril breathing is done in modern yoga (ref??). However, the reading of many manuscripts suggests that the nose was held with only the index fingers of both hands (\emph{nāsikāmadhye … tathā}) or that the nose was held with all the fingers of both hands, except the index fingers (\emph{nāsikāmadhye … vinā}), which seems highly impracticable. It is likely that scribes changed \emph{madhyātarjanībhyāṃ} to \emph{madhye tarjanībhyāṃ} or \emph{madhyaṃ tarjanībhyāṃ} because of the \emph{pāda} break. 
\end{philcomm}

\subsection*{2.66}
\begin{translation}[hp02_066]
Having done the retention as before, the yogi should exhale through the left channel. [Because] it removes [imbalances] in wind, bile and phlegm, increases the body’s fire,

\end{translation}

\begin{sources}[hp02_066]
Gorakṣaśataka

\begin{versinnote}
\tl{kumbhakaṃ pūrvavat kṛtvā recayed iḍayānilam\\+}
\tl{kaṇṭhotthitānalaharaṃ śarīrāgnivivardhanam\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_066]
Yogacintāmaṇi

\begin{versinnote}
\tl{kumbhakaṃ pūrvavat kṛtvā recayed iḍayā tataḥ |\\+}
\tl{vātapittaśleṣmaharaṃ śarīrāgnivivardhanam ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{kumbhakaṃ pūrvavat kṛtvā recayed iḍayānilam || 2.24 |\\+}
\tl{vātapittaśleṣmaharaṃ śarīrāgnivivardhanam |\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_066]
\end{philcomm}

\subsection*{2.67}
\begin{translation}[hp02_067]
is an auspicious thunderbolt that awakens kuṇḍalinī, destroys bad deeds, bestows happiness, and destroys the blockage of phlegm, etc., situated at the mouth of the central channel,
\end{translation}

\begin{sources}[hp02_067]
Gorakṣaśataka

\begin{versinnote}
\tl{kuṇḍalībodhakaṃ vajraṃ pāpaghnaṃ śubhadaṃ sukham |\\+}
\tl{brahmanāḍīmukhāntaḥsthakaphādyargalanāśanam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_067]
Yogacintāmaṇi

\begin{versinnote}
\tl{kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubham |\\+}
\tl{brahmanāḍīmukhe saṃsthaṃ kapāṭārgalanāśanam ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{brahmanāḍīmukhe saṃsthakaphādyargalanāśanam |\\!}
\end{versinnote}

Yuktabhavadeva

\begin{versinnote}
\tl{kuṇḍalībodhanaṃ sarvadoṣaghnaṃ sukhadaṃ śubham ||\\+}
\tl{brahmanāḍīmukhāntasthakaphādyargalanāśanam || 117 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_067]
The word \emph{vipra} in V1 seems to be a mistake for \emph{vajraṃ} in the \emph{Gorakṣaśataka}.

Maybe comment on \emph{pāpaghnaṃ} and \emph{bhavaghnaṃ}

The J10 group has rewritten this verse in the masc sg. with \emph{kumbhaḥ}. And this is possible without the \emph{viśeṣeṇaiva [...] kumbhakaṃ tv idaṃ} line. However, it seems that the \emph{viśeṣeṇaiva} line has dropped out because it is in the Gorakṣaśataka. Therefore, the neuter was probably original.
\end{philcomm}

\subsection*{2.68}
\begin{translation}[hp02_068]
[and] completely pierces the three knots that have arisen from the three \emph{guṇa}s, this retention called ‘the bellows’ in particular is to be done.
\end{translation}

\begin{sources}[hp02_068]
Gorakṣaśataka

\begin{versinnote}
\tl{guṇatrayasamudbhūtagranthitrayavibhedakam || 48 ||\\+}
\tl{viśeṣeṇaiva kartavyaṃ bhastrākhyaṃ kuṃbhakaṃ tv idaṃ\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_068]
Yogacintāmaṇi

\begin{versinnote}
\tl{samyaggātrasamudbhūtagranthitrayavibhedanam |\\+}
\tl{viśeṣeṇaiva kartavyaṃ bhastrākhyaṃ kumbhakaṃ tv idam ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{viśeṣenaiva kartavyaṃ bhastrākhyaṃ kumbhakaṃ tv idam || 2.25 ||\\!}
\end{versinnote}

YBhD

\begin{versinnote}
\tl{samyaggātrasamudbhūtagranthitrayavibhedanam ||\\+}
\tl{viśeṣeṇaiva karttavyaṃ bhastrākhyaṃ kumbhakaṃ tv idam || 118 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_068]
\end{philcomm}

\subsection*{2.69}
\begin{translation}[hp02_069]
Now Bhrāmarī---
The inhalation has a forceful noise and the sound of a male bee, and the exhalation has the sound of the female bee and is very slow. For the best yogis, as a result of practising thus, there arises in the mind an extraordinary blissful playfulness.
\end{translation}

\begin{testimonia}[hp02_069]
Yogacintāmaṇi

\begin{versinnote}
\tl{bhramarī—\\+}
\tl{vegodghoṣaṃ pūrakaṃ bhṛṅganādaṃ \\+}
\tl{bhṛṅgīnādaṃ recakaṃ mandamandam |\\+}
\tl{yogīndrāṇāṃ nityam abhyāsayogāc \\+}
\tl{citte jātā kācid ānandalīlā ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{atha bhrāmarī---\\+}
\tl{vegodghoṣaṃ pūrakaṃ bhṛṅganādaṃ \\+}
\tl{bhṛṅgīnādaṃ recakaṃ mandamandam |\\+}
\tl{yogīndrāṇāṃ nityam abhyāsayogāc \\+}
\tl{citte jātā kā cid ānandalīlā || 2.26 ||\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{atha bhrāmarī---\\+}
\tl{vegodghoṣaṃ pūrakaṃ bhṛṃganādaṃ \\+}
\tl{recakaṃ mandamandam ||\\+}
\tl{yogīndrāṇāmevamabhyāsayogāc\\+}
\tl{citte jātā kācidānandalīlā || 119\\!}
\end{versinnote}

YPr.

\begin{versinnote}
bhrāmarīkumbhakaṃ lakṣayatyatheti ||
vegena sañjāta udghoṣo yasmin pūrake taṃ bhṛṃganādatulyaṃ
\end{versinnote}

HSC

\begin{versinnote}
\tl{vegākṛṣṭiṃ pūrakaṃ bhṛṃganādaṃ \\+}
\tl{bhaṃgānādaṃ recakaṃ maṃdaṃ maṃdaṃ ||\\+}
\tl{yogīdrāṇām evam abhyāsayogac \\+}
\tl{cite jātā kā cid ānaṃdamūrchā ||\\+}
\tl{vegodghoṣam iti vā pāṭhaḥ ||\\!}
\end{versinnote}

Kumbhakapaddhati

\begin{versinnote}
\tl{aliśabdayutaṃ vegāt pūrayet kumbhayet tataḥ |\\+}
\tl{sāliśabdāc chanai rekāt bhrāmarīkumbhako muneḥ ||\\+}
\tl{ānandalīlāṃ kurute bhrāmarīkumbhako muneḥ || 169 ||\\!}
\end{versinnote}

Gheraṇḍasaṃhitā 7.10--11

\begin{versinnote}
\tl{anilaṃ mandavegena bhrāmarīkumbhakaṃ caret |\\+}
\tl{mandaṃ mandaṃ recayed vāyuṃ bhṛṅganādaṃ tato bhavet || 7.10 ||\\+}
\tl{antaḥsthaṃ bhramarīnādaṃ śrutvā tatra mano nayet |\\+}
\tl{samādhir jāyate tatra ānandaḥ so 'ham ity ataḥ || 7.11 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_069]
The first hemistich needs a verb because \emph{recaka} and \emph{pūraka} are in the accusative and it is difficult to construe without one.
\end{philcomm}

\subsection*{2.70}
\begin{translation}[hp02_070]
Now mūrcchā---
Having deeply applied Jālandhara at the end of the inhalation, [the yogi] should exhale slowly. This [kumbhaka] called mūrcchā gives the bliss of the fainting mind.
\end{translation}

\begin{sources}[hp02_070]
Yogacintāmaṇi
​​
\begin{versinnote}
\tl{mūrchā—\\+}
\tl{pūrakānte gāḍhataraṃ bandho jālandharaḥ śanaiḥ |\\+}
\tl{recayen mūrchanākhyo 'yaṃ manomūrchā sukhapradā ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{atha mūrcchā---\\+}
\tl{pūrakānte gāḍhataraṃ baddhva jālandharaṃ śanaiḥ |\\+}
\tl{recayen mūrcchanākhyo 'yaṃ manomūrcchā sukhapradā || 2.27 ||\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{atha mūrcchā---\\+}
\tl{pūrakānte gāḍhataraṃ baddhvā jālandharaṃ śanaiḥ ||\\+}
\tl{recayen mūrcchanākhyeyaṃ manomūrcchā sukhapradā || 120 ||\\!}
\end{versinnote}

Kumbhakapaddhati

\begin{versinnote}
\tl{āpūrya kumbhitaṃ prāṇaṃ badhvā jālandharaṃ śanaiḥ |\\+}
\tl{recayen mūrcchanākumbho manomūrcchā sukhapradā || 170 ||\\!}
\end{versinnote}
\end{sources}

\begin{philcomm}[hp02_070]
The kumbhaka (retention) is not apparent in the description.
\end{philcomm}

\subsection*{2.71}
\begin{translation}[hp02_071]
Now Plāvinī---
[The yogi] whose abdomen is completely filled by the breath of eructation, which has been internally inverted, floats easily like a lotus leaf even on deep water.
\end{translation}

\begin{testimonia}[hp02_071]
{[Not in Yogacintāmaṇi, Haṭharatnāvalī]}

Cf. Kumbhakapaddhati 171

\begin{versinnote}
\tl{yatheṣṭaṃ pūrayed vāyuṃ baddhe jālandhare dṛḍhe |\\+}
\tl{hṛdi dhṛtvā jale suptvā plāvinīkumbhako bhavet || 171 ||\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{antaḥpravarttitādhāramārutāpūritodaraḥ ||\\+}
\tl{payasy agādhe 'pi sukhāt plavate padmapatravat ||\\+}
\tl{ayameva plāvinī kumbhako'pi || 121 ||\\!}
\end{versinnote}

Yogaprakāśikā

\begin{versinnote}
\tl{plāvanīkumbhakaṃ lakṣayati antariti ||\\+}
\tl{antaḥsañcāritenāpānavāyunā pūritamudaraṃ yasyeti vigrahaḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_071]
We have understood \emph{udgāramāruta} to refer to the breath of eructation, i.e. the nāga breath as described in e.g. \emph{Vivekamārtaṇḍa} 36.
We have adopted \emph{payasy agādhe pi sukham} in pāda c because \emph{sukham} gives a better meaning and it is a bha-vipulā
 V19 and P28, Yogacintāmaṇi and the Haṭharatnāvalī omit this verse and accordingly do not mention plāvinī in verse 44, substituting it with kevala.
\end{philcomm}

\subsection*{2.72}
\begin{translation}[hp02_072]
Now kevalakumbhaka---
Prāṇāyāma is said to be threefold with exhalation, inhalation and retention. Retention is considered twofold: sahita and kevala.

\end{translation}

\begin{sources}[hp02_072]
GŚ

\begin{versinnote}
\tl{prāṇaś ca dehajo vāyur āyāmaḥ kumbhakaḥ smṛtaḥ |\\+}
\tl{sa eva dvividhaḥ proktaḥ sahitaḥ kevalas tathā ||\\!}
\end{versinnote}

Vasiṣṭhasaṃhitā 3.2cd

\begin{versinnote}
\tl{prāṇāyāmas tribhiḥ prokto recapūrakakumbhakaiḥ || 2 ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_072]
YBD

\begin{versinnote}
\tl{atha kevalaḥ---\\+}
\tl{prāṇāyāmas tridhā prokto recapūrakakumbhakaiḥ ||\\+}
\tl{sahitaḥ kevalaś ceti kumbhako dvividho mataḥ || 122 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_072]
\end{philcomm}

\subsection*{2.73}
\begin{translation}[hp02_073]
The [kumbhaka] that one performs with exhalation and inhalation is sahita. One should practice sahita until kevala is perfected.
\end{translation}

\begin{sources}[hp02_073]
Cf. DYŚ

\begin{versinnote}
\tl{sahito recapūrābhyāṃ tasmāt sahitakumbhakaḥ |\\!}
\end{versinnote}

GŚ

\begin{versinnote}
\tl{yāvat kevalasiddhiḥ syāt tāvat sahitam abhyaset |\\!}
\end{versinnote}

Vasiṣṭhasaṃhitā 3.28

\begin{versinnote}
\tl{virecyāpūrya yaṃ kuryāt sa vai sahitakumbhakaḥ \\+}
\tl{sahitaṃ kevalaṃ cātha kumbhakaṃ nityam abhyaset ||\\+}
\tl{yāvat kevalasiddhiḥ syāt tāvat sahitam abhyaset |\\!}
\end{versinnote}

Yogayājñavalkya 6.31cd–32

\begin{versinnote}
\tl{recya cāpūrya yaḥ kuryāt sa vai sahitakumbhakaḥ ||\\+}
\tl{sahitaṃ kevalaṃ cātha kumbhakaṃ nityam abhyaset |\\+}
\tl{yāvat kevalasiddhiḥ syāt tāvat sahitam abhyaset ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_073]
Yogacintāmaṇi

\begin{versinnote}
\tl{ārecyāpūrya yat kuryāt sa vai sahitakumbhakaḥ |\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{recya vā pūrakaḥ kāryaḥ śanaiḥ sahitakumbhakaḥ ||\\+}
\tl{yāvat kevalasiddhiḥ syāt sahitaṃ tāvad abhyaset || 123 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_073]
The \emph{Vasiṣṭhasaṃhitā}’s reading, which is not found in any of the HP mss, is the only one that makes sense of 2.73ab so has been adopted.
\end{philcomm}

\subsection*{2.74}
\begin{translation}[hp02_074]
Holding the breath comfortably without exhalation and inhalation is kevalakumbhaka. This is said to be [the true] prāṇāyāma. 
\end{translation}

\begin{sources}[hp02_074]
Vasiṣṭhasaṃhitā 3.27

\begin{versinnote}
\tl{recanaṃ pūraṇaṃ muktvā sukhaṃ yad vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}

Yogayājñavalkya 6.30cd–6.31ab

\begin{versinnote}
\tl{recakaṃ pūrakaṃ muktvā sukhaṃ yad vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_074]
Yogacintāmaṇi

\begin{versinnote}
\tl{recakaṃ pūrakaṃ muktvā yat sukhaṃ vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{atha kevalaḥ---\\+}
\tl{recakaṃ pūrakaṃ muktvā sukhaṃ yad vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{recakaṃ pūrakaṃ muktvā yad vāyudhāraṇam ||\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}

Yogadīpikā 77cd--78ab

\begin{versinnote}
\tl{recakaṃ pūrakaṃ muktvā susukhaṃ vāyudhāraṇaṃ\\+}
\tl{prāṇāyāmoyam ityuktaḥ sa vai kevalakuṃbhakaḥ\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_074]
The relative pronoun is omitted in V1 and the J10 group, which have \emph{vāyunirodhanam}. But the relative is needed to connect the description of kevalakumbhaka in the first hemistich to the last pāda (\emph{sa vai kevalakumbhakaḥ}).
On this verse see Sellmer ?date of article.
\end{philcomm}

\subsection*{2.75}
\begin{translation}[hp02_075]
When kevalakumbhaka without exhalation and inhalation is accomplished, there is nothing in the three worlds that is impossible for the yogi to achieve.
\end{translation}

\begin{sources}[hp02_075]
Vasiṣṭhasaṃhitā 3.30

\begin{versinnote}
\tl{kevale kumbhake siddhe recapūraṇavarjite |\\+}
\tl{na tasya durlabhaṃ kiṃ cit triṣu lokeṣu vidyate |\\!}
\end{versinnote}

DYŚ

\begin{versinnote}
\tl{kevale kumbhake siddhe recapūrakavarjite |\\+}
\tl{na tasya durlabhaṃ kiṃ cit triṣu lokeṣu vidyate || 74 ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_075]
Haṭharatnāvalī

\begin{versinnote}
\tl{kevale kumbhake siddhe recapūrakavarjite |\\+}
\tl{na tasya durlabhaṃ kiñ cit triṣu lokeṣu vidyate ||\\!}
\end{versinnote}
\end{testimonia}

\subsection*{2.76}
\begin{translation}[hp02_076]
He who is empowered by kevalakumbhaka undoubtedly attains [the ability to] hold the breath as long as he wants and the state of Rājayoga.
\end{translation}

\begin{testimonia}[hp02_076]
Haṭharatnāvalī

\begin{versinnote}
\tl{śaktaḥ kevalakumbhena yatheṣṭaṃ vāyudhāraṇam |\\+}
\tl{etādṛśo rājayogo kathito nātra saṃśayaḥ || 2.30 ||\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{śaktaḥ kevalakumbhena yatheṣṭaṃ vāyudhāraṇam ||\\+}
\tl{rājayogapadaṃ samyak labhate nātra saṃśayaḥ || 126 ||\\!}
\end{versinnote}

HTK

\begin{versinnote}
\tl{haṭhapradīpikāyām–\\+}
\tl{śaktaḥ kevalakumbhena yatheṣṭaṃ vāyudhāraṇe |\\+}
\tl{rājayogapadaṃ caiva labhate nātra saṃśayaḥ || 59 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_076]
\end{philcomm}

\subsection*{2.77}
\begin{translation}[hp02_077]
Rājayoga does not succeed without Haṭha nor Haṭha without Rājayoga. So, one should practise both until the niṣpatti [stage].
\end{translation}

\begin{sources}[hp02_077]
Śivasaṃhitā 5.222

\begin{versinnote}
\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
\tl{na sidhyati tato yugmam āniṣpatteḥ samabhyaset\\+}
\tl{\textup{[middle hemistich not in mss. I, III, IV, VII, IX, X, XII, XIV--XVI]}\\+}
\tl{tasmāt pravartate yogī haṭhe sadgurumārgataḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_077]
Yogacintāmaṇi

\begin{versinnote}
\tl{haṭhapradīpikāyām
haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
\tl{na sidhyati tato yugmaṃ manīṣy etau samabhyaset |\\+}
\tl{haṭhaṃ vinā rājayogaṃ rājayogaṃ vinā haṭham |\\+}
\tl{ye vai caranti tān manye prayāsaphalavarjitān iti ||\\!}
\end{versinnote}

Haṭharatnāvalī

\begin{versinnote}
\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
\tl{vyāptiḥ syād avinābhūtā śrīrājahaṭhayogayoḥ || 1.19 ||\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ ||\\+}
\tl{na sidhyati tato yugmam āniṣpatteḥ samācaret || 127 ||\\!}
\end{versinnote}

Śivayogadarpana

\begin{versinnote}
\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
\tl{na sidhyati tato yugmam manīṣī tat samabhyaset || 6 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_077]
\end{philcomm}

\subsection*{2.78}
\begin{translation}[hp02_078]
At the end of exhaling the retained breath, one should make the mind supportless. One reaches the state of Rājayoga by practising thus.
\end{translation}

\begin{testimonia}[hp02_078]
YBD

\begin{versinnote}
\tl{kumbhitaḥ prāṇarecānte kuryyāc cittaṃ nirāmayam ||\\+}
\tl{evamabhyāsayogena rājayogapadaṃ vrajet || 128 ||\\!}
\end{versinnote}

HTK 44.60

\begin{versinnote}
\tl{kumbhitaḥ prāṇarecānte kuryyāc cittaṃ nirāśrayam |\\+}
\tl{evamabhyāsayogena rājayogaṃ labhet punaḥ || 60 ||\\+}
\tl{nirāśrayaṃ saṃkalparahitam ||\\!}
\end{versinnote}

Yogaprakāśikā

\begin{versinnote}
\tl{tad eva visadayati kumbhakam iti || kevalakumbhakābhyāsena cittaṃ dagdhaparṇavat nirvāsanaṃ bhavatītyarthaḥ || anyad vyākhyātam || 67 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_078]
\end{philcomm}

\subsection*{2.79}
\begin{translation}[hp02_079]
Because of a kumbhaka, kuṇḍalinī awakens; from the awakening of kuṇḍalinī, suṣumṇā becomes free of blockages and success in Haṭha arises.
\end{translation}

\begin{testimonia}[hp02_079]
Yogacintāmaṇi

\begin{versinnote}
\tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet |\\+}
\tl{anargalaḥ suṣumṇānto haṭhasiddhiś ca jāyate ||\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet ||\\+}
\tl{anargalā suṣumnā ca haṭhasiddhiśca jāyate || 128 ||\\!}
\end{versinnote}

HTK

\begin{versinnote}
\tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet |\\+}
\tl{anargalā suṣumṇā ca haṭhasiddhiḥ prajāyate || iti || 61 ||\\+}
\tl{kumbhakaprāṇāyāmāt bodho jāgaraṇaṃ | suṣumṇā anargalā bādhakarahitā bhavati | tato yogasiddhir bhavati iti ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_079]
\end{philcomm}

\subsection*{2.80}
\begin{translation}[hp02_080]
Thinness of the body, healthy complexion, clarity of the internal resonance, very bright eyes, freedom from disease, mastery of semen, stimulation of the [body’s] fire [and] purification of the channels are the signs of success in Haṭha.
\end{translation}

\begin{testimonia}[hp02_080]
Yogacintāmaṇi

\begin{versinnote}
\tl{vapuḥkṛśatvaṃ vadane prasannatā \\+}
\tl{nādasphuṭatvaṃ nayane sunirmale |\\+}
\tl{arogitā bindujayo 'gnidīpanaṃ \\+}
\tl{nāḍīviśuddhir haṭhasiddhilakṣaṇam ||\\!}
\end{versinnote}

YBD

\begin{versinnote}
\tl{vapuḥ kṛśatvaṃ vadane prasannatā \\+}
\tl{nādasphuṭatvaṃ nayane ca nirmale ||\\+}
\tl{arogatā bindujayo'gnidīpanaṃ \\+}
\tl{nāḍīviśuddhir haṭhasiddhilakṣaṇam || 130 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_080]
\end{philcomm}

\end{ekdosis}
\end{document}

### 2023-06-07 compile-tidyup.sh

@@ -2,12 +2,31 @@
 
 # optional argument 1 is filename, otherwise all relevant *.tex files
 
-xslcmd="java -jar ${HOME}/.nix-profile/saxon9he.jar -s:xml/HP1_edition-tei.xml -xsl:xslt/hp-online.xsl"
-
 xslcmdproc(){
-    ${xslcmd} | \
-	sed "/^\s*$/d" \
-	    > html/hp-online.html 
+    xslcmd="java -jar ${HOME}/.nix-profile/saxon9he.jar"
+
+    echo "processing xml …"
+    $xslcmd -s:xml/HP1_edition-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp1.html \
+	    chapid="hp1" \
+	    transl="../xml/HP1_TranslComm-tei.xml" \
+	    marma="../xml/Marmasthanas-tei.xml" \
+	    jyotsna="../xml/Jyotsna-tei.xml"
+    
+    $xslcmd -s:xml/HP2_edition-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp2.html \
+	    chapid="hp2" \
+	    transl="../xml/HP2_TranslComm-tei.xml" \
+	    marma="../xml/Marmasthanas-tei.xml" \
+	    jyotsna="../xml/Jyotsna-tei.xml"
+
+    echo "concatenating html …"
+    sed '/<!--content-->/q' html/meta.html > html/hp-online.html 
+    
+    for f in "html/hp1.html" "html/hp2.html"
+    do
+	sed -e '/^\s*$/d' $f >> html/hp-online.html
+    done
+    
+    sed -n '/<!--content-->/,$ {p}' html/meta.html >> html/hp-online.html 
 }
 
 compile(){
@@ -45,7 +64,7 @@ then
     compile "${1#*/}"
 elif [ -z "${1:-}" ]
 then
-    for f in "HP1_edition.tex" "HP1_TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex"
+    for f in "HP1_edition.tex" "HP1_TranslComm.tex" "Jyotsna.tex" "Marmasthanas.tex" "HP2_edition.tex" "HP2_TranslComm.tex"
     do
 	compile "${f}"
     done

### 2023-06-16 compile-tidyup.sh
minor fixes

@@ -54,6 +54,7 @@ compile(){
 	echo "No changes to ${1}, skipping compilation."
     fi
 
+    latexmk -c
     popd || exit
 }
 

### 2023-08-17 latex/HP1_TranslComm.tex
adapted to new files

@@
 \chapter*{Translation \& philological commentary}
+
+%%%%%%%%%%
 \subsection*{1.1}
@@
 
+%%%%%%%%%%
 \subsection*{1.2}
@@
 
+%%%%%%%%%%
 \subsection*{1.3}
@@
 
+%%%%%%%%%%
 \subsection*{1.4}
@@
 
+%%%%%%%%%%
 \subsection*{1.5}
@@
 
+%%%%%%%%%%
 \subsection*{1.6}
@@
 
+%%%%%%%%%%
 \subsection*{1.7}
@@
 
+%%%%%%%%%%
 \subsection*{1.8}
@@
 
+%%%%%%%%%%
 \subsection*{1.9}
@@
 
+%%%%%%%%%%
 \subsection*{1.10}
 \begin{translation}[hp01_010]
-Haṭha is considered a refuge for those who are burnt by the scorching torment of transmigration. Haṭha is the tortoise that supports the worlds of all yogas.
+Haṭha is a hut of refuge for those who are burnt by the scorching torment of transmigration. Haṭha is the tortoise that supports the worlds of all yogas.
 \end{translation}
@@
 The reading of \emph{samāśrayo} in V1 is metrically faulty.
-The accepted reading \emph{āśrayo [']yam} is better metically and well attested.
-N23 preserves the reading \emph{samāśrayamaṭho haṭhaḥ} (similar to the \emph{Yogasārasaṃgraha}), which is similar to the fourth \emph{pāda}.
-Though this reading is possible, it seems more likely that it came about through dittography.
 The word \emph{°jagatāṃ} is probably authorial because it makes good sense with  \emph{ādhārakamaṭha} in light of the cosmological notion that the tortoise supports all the worlds.
@@
 
+%%%%%%%%%%
 \subsection*{1.11}
@@
 
+%%%%%%%%%%
 \subsection*{1.12}
 \begin{translation}[hp01_012]
-In a well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live remotely in a small hut.
+In a well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live in an isolated hut.
 \end{translation}
@@
 
+%%%%%%%%%%
 \subsection*{1.13}
@@
 It extends not too high or low, and is thickly smeared with cow dung in the proper way.
-It is clean, free from everything that annoys, furnished on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
+It is clean, free from everything that annoys, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
 \end{translation}
@@
 
+%%%%%%%%%%
 \subsection*{1.14}
@@
 
+%%%%%%%%%%
 \subsection*{1.15}
@@
 
+%%%%%%%%%%
 \subsection*{1.16}
 \begin{translation}[hp01_016]
-However, from zeal, conviction, resolve, contentment, understanding of the truth, and avoiding contact with people, yoga is successful because of [these] six.
+Yoga is successful as a result of six [qualities]: zeal, daring, resolve, gnosis of the truth, conviction, and avoiding contact with people.
 \end{translation}
@@
 
+%%%%%%%%%%
 \subsection*{1.17}
@@
 
+%%%%%%%%%%
 \subsection*{1.18}
@@
 
+%%%%%%%%%%
 \subsection*{1.19}
@@
 
+%%%%%%%%%%
 \subsection*{1.20}
@@
 \tl{savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet\\+}
-\tl{dakṣiṇe ‘pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā\\!}
+\tl{dakṣiṇe 'pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā\\!}
 \end{versinnote}
@@
 
+%%%%%%%%%%
 \subsection*{1.21}
@@
 
+%%%%%%%%%%
 \subsection*{1.22}
@@
 
+%%%%%%%%%%
 \subsection*{1.23}
@@
 
+%%%%%%%%%%
 \subsection*{1.24}
@@
 
+%%%%%%%%%%
 \subsection*{1.25}
@@
 Comment on kṛtvā, which is the reading of V19, Yogacintāmaṇi and Haṭharatnāvalī.
-\end{philcomm}
-
+Godāvaramiśra has dhanurākarṣavat kṛṣṭaṃ dhanurāsanam ucyate, which is likely to be the original reading.
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.26}
@@
 
+\begin{philcomm}[hp01_026]
+Comment on vāmapadam vs dakṣadoṣṇā
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.27}
 \begin{translation}[hp01_027]
-Matsyendra's seat, by which the digestive fire is kindled, is a destructive weapon for a range of terrible diseases; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
+Matsyendra's seat, is a destructive weapon for the many terrible diseases that develop in the stomach; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
 \end{translation}
@@
 \begin{versinnote}
-\tl{matsyendrapīṭhaṃ jaṭharapravṛddhiṃ\\+}
+\tl{matsyendrapīṭhaṃ jaṭharapravṛddha-\\+}
 \tl{pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\+}
 \tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
-\tl{daṇḍe sthiratvaṃ pradadāti puṃsām ||\\!}
+\tl{daṇḍe sthiratvaṃ pradadāti puṃsām ||
+\var{°pravṛddha ] N : °pravṛddhiṃ L}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_027]
-Difficult to construe first half of verse:
-We have not found an instance in another verse where \emph{jaṭhara} means \emph{jaṭharāgni}.
-So, the reading \emph{jvalana}° is better and most of the witnesses that have it attest \emph{jvalanapradīptam}.
-This compound has been accepted and understood as a reverse \emph{bahuvrīhi} (i.e., ‘the āsana by which the digestive fire is kindled).
-J10 and J17 have \emph{jaṭharapravṛddha}, which would qualify \emph{pracaṇḍaruṅmaṇḍala}°, but it seems rather to strange to say that a range of terrible diseases have increased in the stomach, rather than arisen in the stomach.
-KDham ed has variant \emph{jaṭharapradīpaṃ} but this is not attested among the oldest dated manuscripts.
+Change note: we are adopting jaṭharapravṛddha (translation has been changed).  Cf. Yogacintāmaṇi ms. N. and HP mss.  J10, J17, V2, N8, N11.
 
@@
 
+%%%%%%%%%%
 \subsection*{1.28}
@@
 
+%%%%%%%%%%
 \subsection*{1.29}
@@
 
+%%%%%%%%%%
 \subsection*{1.30}
@@
 
+%%%%%%%%%%
 \subsection*{1.31}
@@
 
+%%%%%%%%%%
 \subsection*{1.32}
@@
 
+%%%%%%%%%%
 \subsection*{1.33}
@@
 
+%%%%%%%%%%
 \subsection*{1.34}
@@
 
+%%%%%%%%%%
 \subsection*{1.35}
@@
 \end{translation}
+% [MD] add translation of "pūrvoktam eva matsaṃmataṃ"
 
@@
 
+%%%%%%%%%%
 \subsection*{1.36}
@@
 
+%%%%%%%%%%
 \subsection*{1.37}
@@
 
+%%%%%%%%%%
 \subsection*{1.38}
@@
 
+%%%%%%%%%%
 \subsection*{1.39}
@@
 
+%%%%%%%%%%
 \subsection*{1.40}
@@
 
+%%%%%%%%%%
 \subsection*{1.41}
@@
 
+%%%%%%%%%%
 \subsection*{1.42}
@@
 
+%%%%%%%%%%
 \subsection*{1.43}
@@
 
+%%%%%%%%%%
 \subsection*{1.44}
@@
 
+%%%%%%%%%%
 \subsection*{1.45--46}
@@
 
+%%%%%%%%%%
 \subsection*{1.46}
@@
 
+%%%%%%%%%%
 \subsection*{1.47}
@@
 \begin{versinnote}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam\\+}
-\tl{durlabhaṃ yena kenāpi dhīmatā labhyate param 3.105\\!}
+\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
+\tl{durlabhaṃ yena kenāpi dhīmatā labhyate param || 3.105 ||\\!}
 \end{versinnote}
@@
 
+%%%%%%%%%%
 \subsection*{1.48}
 \begin{translation}[hp01_048]
-Next is taught the doctrine of Matsya.
+%Next is taught the doctrine of Matsya.
+% [MD] add translation of "paścāduktam eva matsaṃmatam" at the end of the previous section.
 
@@
 \begin{philcomm}[hp01_048]
-The reading of \emph{matsyamataṃ} in the subheading is unusual because elsewhere Matsyendra is not referred to as simply Matsya.
+%The reading of \emph{matsyamataṃ} in the subheading is unusual because elsewhere Matsyendra is not referred to as simply Matsya.
 Furthermore, this verse is from the \emph{Vivekamārtaṇḍa} (36), and the teachings of the \emph{Vivekamārtaṇḍa} are attributed to Gorakṣa, not Matsyendra.
@@
 
+%%%%%%%%%%
 \subsection*{1.49}
 \begin{translation}[hp01_049]
-While in padmāsana the yogi should inhale through the opening to the central channel. He who holds the breath is sure to be liberated.
+The yogi in padmāsana who fills [himself] up through the openings of the channels and holds the breath is sure to be liberated.
 \end{translation}
@@
 
-%\begin{philcomm}[hp01_049]
-% comment on \emph{dvāreṣu} and \emph{dvāreṇa} and explain how the reading with \emph{pibati} in the third pāda might have been understood (i.e., by supplying evam).
-%\end{philcomm}
-
+\begin{philcomm}[hp01_049]
+pūrayan can take the locative (e.g., Amaraugha 21d pūrayen mukhe).
+The reading of N23 niyatam would explain the passive verbs in other witnesses.
+The passive verbs do not make sense with yas tu.
+This yields the idea of filling up the central channel (as opposed to other channels) seen in the Yogabīja.
+Note that a breath retention is not mentioned in the previous verse describing padmāsana (only inhalation and exhalation).
+Also,  māruto mriyate yasya is possible (if one  accepts pūrayet in the second pāda).
+J5 also has a plausible reading (mārutaṃ mārayet yas tu) but J5 is contaminated (so the Gr3 reading dhārayet is more reliable).
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.50--52}
 \begin{translation}[hp01_050]
-Now the lion’s pose.
-
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. Having placed the left ankle on right, the right ankle on the left ...
+Now the lion’s pose. ---
+
+[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. He should place the left ankle on right, the right ankle on the left [...]
 \end{translation}
@@
 \begin{translation}[hp01_051]
-... and both hands on the knees and spread his fingers, with his mouth open, he stares at the tip of the nose.
+[...] and both hands on the knees, spread his fingers, open his mouth and gaze in deep concentration at the tip of the nose.
 \end{translation}
@@
 
+%%%%%%%%%%
+\subsection*{1.51}
+\begin{sources}[hp01_051]
+See 1.50.
+\end{sources}
+
+\begin{testimonia}[hp01_051]
+See 1.50.
+\end{testimonia}
+
+\begin{philcomm}[hp01_051]
+See 1.50.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{1.52}
+\begin{sources}[hp01_052]
+See 1.50.
+\end{sources}
+
+\begin{testimonia}[hp01_052]
+See 1.50.
+\end{testimonia}
+
+\begin{philcomm}[hp01_052]
+See 1.50.
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.53--54}
@@
 
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.  Having firmly and motionlessly held the sides of the feet with the hands,...
+[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.  Having firmly and motionlessly held the sides of the feet with the hands,[...]
 \end{translation}
@@
 \begin{translation}[hp01_054]
-... this is bhadrāsana, which cures all diseases and poisons. Yogis of the siddha tradition call it Gorakṣāsana.
-\end{translation}
-
-\begin{philcomm}[hp01_053]
-We have understood \emph{pārśvapāda} (i.e., side of the foot) like the compound \emph{agrapāda} (i.e., front of the foot), which can be classified as a \emph{ekadeśitatpuruṣa}.
-See Sanskrit Vademecum, p.84 (ed.\ Maximilian Mehner and Jürgen Hanneder, Marburg 2019).
-\end{philcomm}
+[...] this is bhadrāsana, which cures all diseases and poisons. Yogis of the siddha tradition call it Gorakṣāsana.
+\end{translation}
 
@@
 
+\begin{philcomm}[hp01_053]
+We have understood \emph{pārśvapāda} (i.e., side of the foot) like the compound \emph{agrapāda} (i.e., front of the foot), which can be classified as a \emph{ekadeśitatpuruṣa}.
+See Sanskrit Vademecum, p.84 (ed.\ Maximilian Mehner and Jürgen Hanneder, Marburg 2019).
+
+% [MD] comment on the additional hemistich "savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe" (N23,Jyo)?
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.54}
@@
 \end{sources}
+
 \begin{testimonia}[hp01_054]
@@
 
+\begin{philcomm}[hp01_054]
+See 1.53.
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.55}
@@
 
+%%%%%%%%%%
 \subsection*{1.55*1--2}
@@
 Success arises for one engaged in practice. How can it arise for one who has no practice? Success in yoga does not arise by merely reading scriptures.
-
+\end{translation}
+
+\begin{translation}[hp01_055_2]
 Wearing a robe does not bring about success, nor does talking [about yoga].
@@
 
+%%%%%%%%%%
+\subsection*{1.55*2}
+\begin{sources}[hp01_055_2]
+See 1.55*1.
+\end{sources}
+
+\begin{testimonia}[hp01_055_2]
+See 1.55*1.
+\end{testimonia}
+
+\begin{philcomm}[hp01_055_2]
+See 1.55*1.
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.55*3}
@@
 
+%%%%%%%%%%
 \subsection*{1.55*4}
@@
 
+%%%%%%%%%%
 \subsection*{1.56}
@@
 
+%%%%%%%%%%
 \subsection*{1.57}
@@
 
+%%%%%%%%%%
 \subsection*{1.58}
@@
 
+%%%%%%%%%%
 \subsection*{1.59}
@@
 
+%%%%%%%%%%
 \subsection*{1.60}
 \begin{translation}[hp01_060]
-One should know food that has been reheated, is dry, is too salty, the leftover paste of crushed sesame seeds (tilapiṇḍa), spoiled [or] contains an excess of leafy vegetables to be unwholesome.
-It is to be avoided.
+One should know food that has been reheated, is dry, too salty or sour, contains an excess of leafy vegetables that are hard to chew, [or] is spoiled to be unwholesome.
 \end{translation}
@@
 
+%%%%%%%%%%
 \subsection*{1.61}
@@
 Similarly a saying by Goraksa:
-One should avoid a liking for bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.
+One should avoid places near bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.
 \end{translation}
@@
 \begin{philcomm}[hp01_061]
-The vulgate has a parallel from the Amaraugha added. 
-\end{philcomm}
-
+The vulgate has a parallel from the Amaraugha added. Stemma requires prāṃtam, which is difficilior lectio in relationship to prītiṃ.
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.62}
 \begin{translation}[hp01_062]
-Pure food with wheat, rice, śāli rice, barley, sixty-day śāli rice, milk, ghee, candied sugar, unclarified fermented butter, ground sugar and honey.
+Pure food with wheat, rice, śāli rice, barley, sixty-day śāli rice, milk, ghee, cream, unclarified fermented butter, ground sugar and honey.
 Dried ginger, fruit of the snake gourd and so forth, the five vegetables, mung beans and so on, and rain water are wholesome for the best of sages.
@@
 
+%%%%%%%%%%
 \subsection*{1.63}
@@
 
+%%%%%%%%%%
 \subsection*{1.64}
@@
 
+%%%%%%%%%%
 \subsection*{1.65}
 \begin{translation}[hp01_065]
-The postures, various breath retentions and techniques, beginning with seals, are all [to be done] in the practice of Haṭha until the reward that is Rājayoga [is attained].
+The postures, various breath retentions, and heavenly techniques: the whole practice of Haṭha [is to be done] until Rājayoga results. 
 \end{translation}
@@
 \begin{philcomm}[hp01_065]
-\emph{sarvaṇy api} is better than \emph{sarvo pi} ca because it refers to all the practices mentioned in the first hemistich.
-
-Reading of Brahmānanda is different for the third pāda: \emph{divyāni karaṇāni}.
 \end{philcomm}

### 2023-08-17 latex/HP2_TranslComm.tex
adapted to new files

@@
 \chapter*{Translation \& philological commentary}
+%%%%%%%%%%
 \subsection*{2.1}
@@
 \emph{dṛḍho yogī} seems unlikely as \emph{dṛḍha} usually qualifies a technique, the body, etc.
-It is interesting that J10, J17 and N17 have \emph{yogamārgaṃ}. The vulgate has \emph{prāṇāyāmān}.
-\end{philcomm}
-
+The vulgate has \emph{prāṇāyāmān}.
+Other witnesses have the plural \emph{prāṇāyāmān} and this variation between singular and plural recurs through this chapter without a clear split among different branches of the stemma, so we have preferred what seem to us to be the better readings on semantic grounds.
+It is interesting that J10, J17 and N17 have \emph{yogamārgaṃ} but it is unlikely that this was the original reading.
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.2}
@@
 
+%%%%%%%%%%
 \subsection*{2.3}
@@
 
+%%%%%%%%%%
 \subsection*{2.4}
@@
 
+%%%%%%%%%%
 \subsection*{2.5}
@@
 \begin{sources}[hp02_005]
-VM 95
+VM 76
 
@@
 
+%%%%%%%%%%
 \subsection*{2.6}
@@
 
+%%%%%%%%%%
 \subsection*{2.7}
@@
 
+%%%%%%%%%%
 \subsection*{2.8}
@@
 
+%%%%%%%%%%
 \subsection*{2.9}
@@
 
+%%%%%%%%%%
 \subsection*{2.10}
@@
 
+%%%%%%%%%%
 \subsection*{2.11}
@@
 
+%%%%%%%%%%
 \subsection*{2.12}
@@
 \begin{philcomm}[hp02_012]
-The manuscripts diverge greatly in the second hemistrich.
+The manuscripts diverge greatly in the second hemistich.
 All have \emph{prāṇa} (instead of \emph{deha} in the VM).
@@
 
+%%%%%%%%%%
 \subsection*{2.13}
@@
 
+%%%%%%%%%%
 \subsection*{2.14}
@@
 
+%%%%%%%%%%
 \subsection*{2.15}
@@
 
+%%%%%%%%%%
 \subsection*{2.16}
@@
 
+%%%%%%%%%%
 \subsection*{2.17}
@@
 
+%%%%%%%%%%
 \subsection*{2.18}
@@
 \tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ ca pūrayet\\%122ab ; HP2.18ab
-yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt || 1\\!}
+yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt ||\\!}
 \end{versinnote}
@@
 
+%%%%%%%%%%
 \subsection*{2.19}
@@
 
+%%%%%%%%%%
 \subsection*{2.20}
@@
 
+%%%%%%%%%%
 \subsection*{2.21}
@@
 
+%%%%%%%%%%
 \subsection*{2.22}
@@
 
+%%%%%%%%%%
 \subsection*{2.23}
@@
 
+%%%%%%%%%%
 \subsection*{2.24}
@@
 
+%%%%%%%%%%
 \subsection*{2.25}
@@
 
+%%%%%%%%%%
 \subsection*{2.26}
@@
 
+%%%%%%%%%%
 \subsection*{2.27}
@@
 
+%%%%%%%%%%
 \subsection*{2.28}
@@
 
+%%%%%%%%%%
 \subsection*{2.29}
@@
 
+%%%%%%%%%%
 \subsection*{2.30}
@@
 
+%%%%%%%%%%
 \subsection*{2.31}
 \begin{translation}[hp02_031]
-And neti purifies the skull, bestows devine sight and quickly cures a multitude of diseases that arise above the clavicles.
+And neti purifies the skull, bestows divine sight and quickly cures a multitude of diseases that arise above the clavicles.
 \end{translation}
@@
 
+%%%%%%%%%%
 \subsection*{2.32}
@@
 
+%%%%%%%%%%
 \subsection*{2.33}
@@
 
+%%%%%%%%%%
 \subsection*{2.34}
@@
 
+%%%%%%%%%%
 \subsection*{2.35}
@@
 
+%%%%%%%%%%
 \subsection*{2.36}
@@
 
+%%%%%%%%%%
 \subsection*{2.37}
@@
 
+%%%%%%%%%%
 \subsection*{2.38}
@@
 \begin{versinnote}
-\tl{prāṇāyāmair eva \textup{<2 syllables missing>} praśuṣyanti malā iti |\\+}
+\tl{prāṇāyāmair eva {\footnotesize <2 syllables missing>} praśuṣyanti malā iti |\\+}
 \tl{keṣāṃ cid ācāryāṇām anya karma na saṃmatam iti ||\\!}
@@
 
+%%%%%%%%%%
 \subsection*{2.39}
@@
 
+%%%%%%%%%%
 \subsection*{2.40}
@@
 \begin{sources}[hp02_040]
-VM
+VM 73
 
@@
 
+%%%%%%%%%%
 \subsection*{2.41}
@@
 
+%%%%%%%%%%
 \subsection*{2.42}
@@
 
+%%%%%%%%%%
 \subsection*{2.43}
@@
 
+%%%%%%%%%%
 \subsection*{2.44}
@@
 
+%%%%%%%%%%
 \subsection*{2.45}
@@
 
+%%%%%%%%%%
 \subsection*{2.46}
@@
 \begin{sources}[hp02_046]
+Yogabīja ?
 \end{sources}
@@
 
+%%%%%%%%%%
 \subsection*{2.47}
@@
 
+%%%%%%%%%%
 \subsection*{2.48}
@@
 
+%%%%%%%%%%
 \subsection*{2.49}
@@
 
+%%%%%%%%%%
 \subsection*{2.50}
@@
 \begin{sources}[hp02_050]
-Gorakṣaśataka 35a-36b
+Gorakṣaśataka 35ab, 36ab
 
@@
 
+%%%%%%%%%%
 \subsection*{2.51}
@@
 \begin{sources}[hp02_051]
-Gorakṣaśataka
+Gorakṣaśataka 36c–37b
 
@@
 
+%%%%%%%%%%
 \subsection*{2.52}
@@
 \begin{sources}[hp02_052]
-GŚ
+GŚ 37c–38b
 
@@
 
+%%%%%%%%%%
 \subsection*{2.53}
@@
 
+%%%%%%%%%%
 \subsection*{2.54}
@@
 
+%%%%%%%%%%
 \subsection*{2.55}
@@
 
+%%%%%%%%%%
 \subsection*{2.56}
@@
 
+%%%%%%%%%%
 \subsection*{2.56*1}
@@
 
+%%%%%%%%%%
 \subsection*{2.57}
@@
 
-\emph{rasanām unmukhīkṛtya sītkāraṃ kurvatā marut} |\\
-\emph{pīyante kumbhake yasmin nāsikābhyāṃ virecanam} || [pīyate?]
+\begin{versinnote}
+\tl{rasanām unmukhīkṛtya sītkāraṃ kurvatā marut |\\+}
+\tl{pīyante kumbhake yasmin nāsikābhyāṃ virecanam || [pīyate?]\\!}
+\end{versinnote}
 
@@
 
-\emph{rasanāṃ tālumūle yaḥ sthāpayitvā vipaścitaḥ} |\\
-\emph{pibet prāṇānilaṃ tasya rogāṇāṃ saṃkṣayo bhavet} ||
+\begin{versinnote}
+\tl{rasanāṃ tālumūle yaḥ sthāpayitvā vipaścitaḥ |\\+}
+\tl{pibet prāṇānilaṃ tasya rogāṇāṃ saṃkṣayo bhavet ||\\!}
+\end{versinnote}
 
@@
 
+%%%%%%%%%%
 \subsection*{2.58}
@@
 
+%%%%%%%%%%
 \subsection*{2.59}
@@
 
+%%%%%%%%%%
 \subsection*{2.60}
@@
 
+%%%%%%%%%%
 \subsection*{2.61}
@@
 
+%%%%%%%%%%
 \subsection*{2.62}
@@
 
+%%%%%%%%%%
 \subsection*{2.63}
@@
 
+%%%%%%%%%%
 \subsection*{2.64}
@@
 Since the subject of the simile is \emph{bhastrā}, one would expect the subject of \emph{cālayet}, which must be different, to be stated (as is the case with \emph{sudhīḥ}).
-Also, one would expect the instrumental of \emph{dhī}to be qualified by some adjective, such as in the case of \emph{sattvāsthayā dhiyā} (Gorakṣaśataka 74b) and \emph{sāttvikayā dhiyā} (Haṭhapradīpikā 2.6b).
-\end{philcomm}
-
+Also, one would expect the instrumental of \emph{dhī} to be qualified by some adjective, such as in the case of \emph{sattvāsthayā dhiyā} (Gorakṣaśataka 74b) and \emph{sāttvikayā dhiyā} (Haṭhapradīpikā 2.6b).
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.65}
@@
 
+%%%%%%%%%%
 \subsection*{2.66}
@@
 
+%%%%%%%%%%
 \subsection*{2.67}
@@
 
+%%%%%%%%%%
 \subsection*{2.68}
@@
 
+%%%%%%%%%%
 \subsection*{2.69}
@@
 
+%%%%%%%%%%
 \subsection*{2.70}
@@
 
+%%%%%%%%%%
 \subsection*{2.71}
@@
 
+%%%%%%%%%%
 \subsection*{2.72}
@@
 Now kevalakumbhaka---
-Prāṇāyāma is said to be threefold with exhalation, inhalation and retention. Retention is considered twofold: sahita and kevala.
+Prāṇāyāma is said to be threefold, with exhalation, inhalation, and retention. Retention is considered twofold: sahita and kevala.
 
@@
 
+%%%%%%%%%%
 \subsection*{2.73}
@@
 
+%%%%%%%%%%
 \subsection*{2.74}
@@
 
+%%%%%%%%%%
 \subsection*{2.75}
@@
 
+%%%%%%%%%%
 \subsection*{2.76}
@@
 
+%%%%%%%%%%
 \subsection*{2.77}
@@
 
+%%%%%%%%%%
 \subsection*{2.78}
@@
 
+%%%%%%%%%%
 \subsection*{2.79}
@@
 
+%%%%%%%%%%
 \subsection*{2.80}

### 2023-09-02 latex/Jyotsna.tex
new routine for jyotsna

@@
-% additions/changes 2022-03-12 mm:
-\input{Preamble-Print}
-\makepagestyle{HPed}
-\makeoddhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
-\makeevenhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
-\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
-\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}
+\enableregime[utf]
+\definepapersize[IeT][width=155mm, height=220mm]
+\setuppapersize[IeT][A4]       
+\setuplayout[width=fit,height=fit,header=14pt,grid=no,marking=off,backspace=13mm,topspace=15mm,location=middle] 
+\setupheader[leftwidth=.8\hsize, rightwidth=.8\hsize,style=italic]
+\setupheadertexts[][][][]
+\setuppagenumbering[alternative=doublesided,location={footer,center}]
+\definefontfeature[default][default][fakecombining=yes,compose=yes]
 
-\setlength{\footmarkwidth}{1.3em}
-\setlength{\footmarksep}{0em}
-\setlength{\footparindent}{1.3em}
-\footmarkstyle{\textsuperscript{#1} }
-\usepackage{fnpos}
-\makeFNbottom
+\starttypescript [serif] [adobetextpro]
+  \definefontsynonym[AdobeTextPro-Regular]     [file:AdobeTextPro-Regular] [features=default]
+  \definefontsynonym[AdobeTextPro-Italic]          [file:AdobeTextPro-It]  [features=default]
+  \definefontsynonym[AdobeTextPro-Bold]           [file:AdobeTextPro-Bold]  [features=default]
+\stoptypescript
 
-\SetTEIxmlExport{autopar=false}
-\NewDocumentEnvironment{jyotsna}{O{}}{}{}
+\starttypescript [serif] [adobetextpro] [name]
+  \definefontsynonym [Serif]           [AdobeTextPro-Regular]  
+  \definefontsynonym [SerifItalic]     [AdobeTextPro-Italic] 
+  \definefontsynonym [SerifBold]       [AdobeTextPro-Bold] 
+\stoptypescript
 
-\begin{document}
-\pagestyle{HPed} %
-\begin{ekdosis}
-  \SetLineation{lineation = none,}
+\starttypescript [adobetextpro]
+  \definetypeface [adobetextpro] [rm] [serif] [adobetextpro] [default] [features=default]
+\stoptypescript
 
-  \begin{jyotsna}[hp01_001]
-    śrīādināthāya namo ’stu tasmai \\
-    yenopadiṣṭā haṭhayogavidyā  / \\
-    vibhrājate pronnatarājayogam \\
-    āroḍhum icchor adhirohiṇīva (1.1)
+\setupbodyfont[adobetextpro] 
+\mainlanguage[sa]  
+% Critical Edition Setup: Prose with two layers of apparatus
+\setupnotations[alternative=serried]
+\setuplinenumbering[step=5,method=page]
+\definelinenote[var][paragraph=yes]
+\definelinenote[comment][paragraph=yes]   %oder: n=2]
+\definelines[sloka][][indenting={no, small},
+                   before=\blank\startnarrower,after=\stopnarrower\blank]
+\setuptolerance[vertical,verytolerant]
+\setuptolerance[horizontal,verytolerant,stretch]
+\clubpenalty=10000 % keine Schusterjungen / no clubs
+\widowpenalty=10000 % keine Hurenkinder / no widows
+\setupwhitespace[3pt]    %  3pt zwischen Paragraphen
+\setupblank[big]    %	  \blank[2*line],   \blank[medium]
+\def\lem{\kern0em{}]\enskip}
 
-    guruṃ natvā śivaṃ sākṣādbrahmānandena tanyate /
-    haṭhapradīpikājyotsnā yogamārgaprakāśikā //
+\def\Wthree{W\ }  \def\Wthreeac{W\high{ac}\ }  \def\Wthreepc{W\high{pc}\ } 
+\def\Tue{T\ }    \def\Tueac{T\high{ac}\ }  \def\Tuepc{T\high{pc}\ }
+\def\Msix{M}\def\EdLo{Ed\low{Lon}\ }   \def\EdAd{Ed\low{Adyar}\ }
+\def\EdMu{Ed\low{Princ}\ } \def\emend{em.\ }
+\def\om{om.}
 
-    idānīṃtanānāṃ subodhārtham asyāḥ     (variant: idānīṃ janānāṃ Lonavla Ed) 	
-    suvijñāya gorakṣasiddhāntahārdam /
+\starttext
+
+{\tfa\bf  Haṭhapradīpikājyotsnā}
+
+\blank[7mm]
+
+\startlinenumbering
+%\begin{vsid}{hp01_001}
+\startsloka
+{\bf śrīādināthāya namo 'stu tasmai 
+\ \ yenopadiṣṭā haṭhayogavidyā
+vibhrājate pronnatarājayogam 
+\ \ āroḍhum icchor adhirohiṇīva (1.1)}
+\stopsloka
+
+\startsloka
+guruṃ natvā śivaṃ sākṣād brahmānandena tanyate
+\ \ haṭhapradīpikājyotsnā yogamārgaprakāśikā
+
+idānīṃtanānāṃ\var{idānīṃtanānāṃ \Msix] idānīṃ janānāṃ \EdLo} subodhārtham asyāḥ 
+\ \ suvijñāya gorakṣasiddhāntahārdam 
 mayā meruśāstripramukhyābhiyogāt
-    sphuṭaṃ kathyate 'tyantagūḍho 'pi bhāvaḥ //
+\ \ sphuṭaṃ kathyate 'tyantagūḍho 'pi bhāvaḥ
+\stopsloka
 
-    mumukṣujanahitārthaṃ rājayogadvārā kaivalyaphalāṃ haṭhapradīpikāṃ vidhitsuḥ paramakāruṇikaḥ
+\blank
+
+mumukṣujanahitārthaṃ rājayogadvārā kaivalyaphalāṃ haṭhapradīpikāṃ vi\-dhitsuḥ paramakāruṇikaḥ
 svātmārāmayogīndras tatpratyūhanivṛttaye haṭhayogapravartakaśrīmadādināthanamaskāralakṣaṇaṃ
-    maṅgalaṃ tāvad ācarati—śrīādināthāyetyādinā /
+maṅgalaṃ tāvad ācarati śrī\-ādināthetyādinā. tasmai śrīādināthāya namaḥ astu ity anvayaḥ. ādiś
+cāsau nāthaś ca ādināthaḥ sarveśvaraḥ śiva ity arthaḥ. śrīmān ādināthaḥ śrīādināthaḥ tasmai
+śrīādi\-nāthāya. śrīśabda ādir yasya saḥ śrīādiḥ śrīādiś cāsau nāthaś ca śrīādināthaḥ tasmai
+śrīādināthāya śrīnāthāya viṣṇave iti vārthaḥ.  śrīādināthāyety atra yaṇabhāvas tu api māṣaṃ
+maṣaṃ\var{māṣaṃ maṣaṃ \lem māsaṃ masaṃ \Msix (not reported in \EdLo)} kuryāc chandobhaṅgaṃ tyajed girām
+iti chandovidāṃ sampradāyād uccāraṇasauṣṭhavāc ceti bodhyam. namaḥ prahvī\-bhāvo 'stu. prārthanāyāṃ
+loṭ. tasmai. kasmai ity apekṣāyāṃ āha yeneti yena ādināthena upadiṣṭā girijāyai
+haṭhayogavidyā. haś ca ṭhaś ca haṭhau sūryacandrau tayor yogo haṭhayogaḥ. etena haṭhaśabdavācyayoḥ
+sūryacandrākhyayoḥ prāṇāpānayor aikyalakṣaṇaḥ prāṇāyāmo haṭhayoga iti haṭha\-yogasya lakṣaṇaṃ
+siddham. tathā coktaṃ gorakṣanāthena siddha\-siddhāntapaddhatau
 
-    tasmai śrīādināthāya namo 'stv ity anvayaḥ / ādiś cāsau nāthaś ca ādināthaḥ sarveśvaraḥ śiva ity
-    arthaḥ / śrīmān ādināthaḥ śrīādināthaḥ tasmai śrīādināthāya / śrīśabda ādir yasya saḥ śrīādiḥ,
-    śrīādiś cāsau nāthaś ca śrīādināthaḥ, tasmai śrīādināthāya / śrīnāthāya viṣṇava iti vārthaḥ / śrī
-    ādināthāyety atra yaṇabhāvas tu api māṣaṃ maṣaṃ kuryāc chandobhaṅgaṃ tyajed girām iti chandovidāṃ
-    sampradāyād uccāraṇasauṣṭhavāc ceti bodhyam / namaḥ prahvībhāvo 'stu / prārthanāyāṃ loṭ / tasmai /
-    kasmai ity apekṣāyāṃ āha yeneti / yena ādināthena upadiṣṭā girijāyai haṭhayogavidyā / haś ca ṭhaś
-    ca haṭhau sūryacandrau tayor yogo haṭhayogaḥ / etena haṭhaśabdavācyayoḥ sūryacandrākhyayoḥ
-    prāṇāpānayor aikyalakṣaṇaḥ prāṇāyāmo haṭhayoga iti haṭhayogasya lakṣaṇaṃ siddham / yathā coktaṃ
-    gorakṣanāthena siddhasiddhāntapaddhatau—
+\startsloka
+hakāraḥ kīrtitaḥ sūryaṣ ṭhakāraś candra ucyate
+sūryacandramasor yogād dhaṭhayogo nigadyate 
+\stopsloka
 
-    hakāraḥ kīrtitaḥ sūryaṣ ṭhakāraś candra ucyate /
-    sūryacandramasor yogād dhaṭhayogo nigadyate // [si.si.pa. 1.69] iti /
+iti tatpratipādikā vidyā haṭhayogavidyā haṭhayogaśāstram iti yāvat. girijāyai\var{girijāyai \lem
+  girijāyāḥ \Msix} ādināthakṛto haṭhavidyopadeśo mahākālayogaśāstrādau prasiddhaḥ.
 
-    tatpradīpikā vidyā haṭhayogavidyā haṭhayogaśāstram iti yāvat / girijāyai ādināthakṛto
-    haṭhavidyopadeśo mahākālayogaśāstrādau prasiddhaḥ /
+prakarṣeṇa unnataḥ pronnataḥ. mantrayogahaṭhayogādīnām adharabhūmī\-nām uttarabhūmitvād rājayogasya
+pronnatatvam. rājayogaś ca sarvavṛttini\-rodhalakṣaṇo 'samprajñātayogaḥ. tam icchor mumukṣor
+adhirohiṇīva adhi\-ruhyate 'nayety adhirohiṇī niḥśreṇīva vibhrājate viśeṣeṇa bhrājate śobhate. yathā
+pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudha\-prāpi\-kā bhavati evaṃ haṭhapradīpikāpi
+pronnatarājayogam āroḍhum icchor anā\-yāsena rājayogaprāpikā bhavatīti upamālaṅkāraḥ.
+indravajrākhyaṃ vṛttam. 
+%\end{vsid}
 
-    prakarṣeṇa unnataḥ pronnataḥ / mantrayogahaṭhayogādīnām adharabhūmīnām uttarabhūmitvād rājayogasya
-    pronnatatvam / rājayogaś ca sarvavṛttinirodhalakṣaṇo 'samprajñātayogaḥ / tam icchor mumukṣor
-    adhirohiṇīva adhiruhyate 'nayety adhirohiṇī niśreṇīva vibhrājate viśeṣeṇa bhrājate śobhate / yathā
-    pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudhaprāpikā bhavati evaṃ haṭhadīpikāpi
-    pronnatarājayogam āroḍhum icchor anāyāsena rājayogaprāpikā bhavatīti / upamālaṅkāraḥ /
-    indravajrākhyaṃ vṛttam //1// 
-  \end{jyotsna}
+%\begin{vsid}{hp01_002}
+\startsloka
+{\bf praṇamya śrīguruṃ nāthaṃ svātmārāmeṇa yoginā
+kevalaṃ rājayogāya haṭhavidyopadiśyate (1.2)}
+\stopsloka
 
-  \begin{jyotsna}[hp01_002]
-    praṇamya śrīguruṃ nāthaṃ svātmārāmeṇa yoginā /
-    kevalaṃ rājayogāya haṭhavidyopadiśyate //2//
+evaṃ paramagurunamaskāralakṣaṇaṃ maṅgalaṃ kṛtvā vighnabāhulye maṅga\-labāhulyasyāpy\var{apy \lem om
+  \Msix} apekṣitatvāt svagurunamaskārātmakaṃ maṅgalam ācarann asya granthasya viṣayaprayojanādīn
+pradarśayati praṇamyeti.\var{praṇamyeti \lem om \Msix} śrīmantaṃ guruṃ śrīguruṃ nāthaṃ śrīgurunāthaṃ
+svagurum iti yāvat.  praṇamya pra\-karṣeṇa bhaktipūrvakaṃ natvā svātmārāmeṇa\var{svātmārāmeṇa \lem
+  svātmārāmeṇa svātmārāmākhyeṇa \Msix} yoginā yogo 'syāstīti yena\var{yena \lem tena \Msix}. kevalaṃ
+rājayogāya kevalaṃ rājayogārthaṃ haṭhavidyopadiśyata ity anvayaḥ. haṭhavidyāyā rājayoga eva mukhyaṃ
+phalaṃ na siddhaya iti kevalapadasyā\-bhi\-prāyaḥ. siddhayas tv ānuṣaṅgikyaḥ\var{ānuṣaṅgikyaḥ \lem
+  ānuṣaṅgikāḥ \Msix}. etena rājayogaphalasahito haṭhayogo 'sya granthasya viṣayaḥ. rājayogadvārā
+ükaivalyaṃ cāsya phalam.  tatkāmaś cādhikārī. granthaviṣayayoḥ pratipādyapratipādakabhāvaḥ
+sambandhaḥ. granthasya kaivalyasya ca prayojyaprayojakabhāvaḥ sambandhaḥ. granthābhidheyasya
+saphalayogasya kaivalyasya ca sādhyasādhanabhāvaḥ sambandha ity uktam.
+%\end{vsid}
 
-    evaṃ paramagurunamaskāralakṣaṇaṃ maṅgalaṃ kṛtvā vighnabāhulye maṅgalabāhulyasyāpy apekṣitatvāt
-    svagurunamaskārātmakaṃ maṅgalam ācarann asya granthasya viṣayaprayojanādīn
-    pradarśayati—praṇamyeti / śrīmantaṃ guruṃ śrīguruṃ nāthaṃ śrīgurunāthaṃ, svagurum iti yāvat /
-    praṇamya prakarṣeṇa bhaktipūrvakaṃ natvā svātmārāmeṇa yoginā yogo'syāstīti yena / kevalaṃ
-    rājayogāya kevalaṃ rājayogārthaṃ haṭhavidyopadiśyata ity anvayaḥ / haṭhavidyāyā rājayoga eva
-    mukhyaṃ phalaṃ na siddhaya iti kevalapadasyābhiprāyaḥ / siddhayas tv ānuṣaṅgikyaḥ / etena
-    rājayogaphalasahito haṭhayogo'sya granthasya viṣayaḥ / rājayogadvārā kaivalyaṃ cāsya phalam /
-    tatkāmaś cādhikārī / granthaviṣayayoḥ pratipādyapratipādakabhāvaḥ sambandhaḥ / granthasya
-    kaivalyasya ca prayojyaprayojakabhāvaḥ sambandhaḥ / granthābhidheyasya saphalayogasya
-    kaivalyasya ca sādhyasādhanabhāvaḥ sambandha ity uktam //2//
-  \end{jyotsna}
+%\begin{vsid}{hp01_003}
+\startsloka
+{\bf bhrāntyā bahumatadhvānte rājayogam ajānatām
+haṭhapradīpikāṃ dhatte svātmārāmaḥ kṛpākaraḥ (1.3)}
+\stopsloka
 
-\end{ekdosis}
-\end{document}
+nanu mantrayogasaguṇadhyānanirguṇadhyānamudrādibhir eva rājayogasiddhau kiṃ haṭhavidyopadeśenety
+āśaṅkya vyutthitacittānāṃ mantrayogādibhī rājayogāsiddher haṭhayogād eva rājayogasiddhiṃ vadan
+granthaṃ pratijānīte bhrāntyeti. mantrayogādibahumatarūpe dhvānte gāḍhāndhakāre yā bhrāntir bhramas
+tayā\var{tayā \lem om \Msix}. tais tair upāyai rājayogārthaṃ pravṛttasya tatra tatra
+tadalābhāt. vakṣyati ca haṭhaṃ\var{haṭhaṃ \Msix \lem haṭhayogaṃ Ś} vinā rājayogaḥ
+ityādinā. tathā\var{tathā \lem tayā \Msix} rājayogam ajānatāṃ na jānantīty ajānantaḥ teṣām ajānatāṃ
+puṃsāṃ rājayogajñānāyeti śeṣaḥ. karotīti karaḥ kṛpāyāḥ karaḥ kṛpākaraḥ kṛpāyā ākara iti vā
+tādṛśaḥ. anena haṭhapradīpikākaraṇe ajñānukampaiva\var{ajñānukampaiva \lem bhūtānukampaiva \Msix}
+hetur ity uktam. svātmany āramate iti svātmārāmaḥ. haṭhasya haṭhayogasya pradīpikeva prakāśakatvāt
+haṭhapradīpikā tām. athavā haṭha eva pradīpikā haṭhapradīpika rājayogaprakāśakatvāt tāṃ dhatte
+vidhatte karotīti yāvat.
+
+svātmārāma ity anena jñānasya saptamabhūmikāṃ\var{saptama \Wthree \Tuepc \Msix ] sapta \Tueac ,
+saptaṃ \EdLo} prāpto brahmavidvariṣṭha ity uktam. tathā ca śrutiḥ ātmakrīḍa ātmaratiḥ kriyāvān eṣa
+  brahmavidāṃ variṣṭhaḥ\comment{{\em Muṇḍakopaniṣat} 3.1.4} iti.
+
+saptabhūmayaś coktā yogavāsiṣṭhe
+
+\startsloka
+jñānabhūmiḥ śubhecchākhyā prathamā samudāhṛtā
+vicāraṇā dvitīyā syāt tṛtīyā tanumānasā
+
+sattvāpattiś caturthī syāt tato 'saṃsaktināmikā
+parārthābhāvinī ṣaṣṭhī saptamī turyagā smṛtā iti\comment{{\em Laghuyogavāsiṣṭha} 9.114–115}
+\stopsloka
+
+asyārthaḥ śubhecchā ity ākhyā yasyāḥ sā śubhecchākhyā vivekavairāgyayutā\var{yutā \lem om \Msix}
+śamādipūrvikā tīvramumukṣā prathamā jñānasya bhūmiḥ bhūmikā udāhṛtā kathitā yogibhir iti śeṣaḥ.
+
+vicāraṇā śravaṇamananātmikā dvitīyā jñānabhūmiḥ syāt.
+
+anekārthagrāhakaṃ mano yadānekārthān parityajya sadekārthavṛttipra\-vāha\-vad bhavati tadā
+tanumānasaṃ yasyāṃ sā tanumānasā nididhyāsanarūpā tṛtīyā jñānabhūmiḥ syād iti śeṣaḥ.
+
+imās tisraḥ sādhanabhūmikāḥ. āsu bhūmiṣu sādhaka ity ucyate. tisṛbhir bhūmikābhiḥ śuddhasattve
+'ntaḥkaraṇe 'haṃ brahmāsmīty ākārikā aparokṣa\-vṛttirūpā sattvāpattināmikā caturthī jñānabhūmiḥ
+syāt. caturthīyaṃ phalabhūmiḥ. asyāṃ yogī brahmavid ity ucyate. iyaṃ samprajñātayogabhūmikā.
+
+vakṣyamāṇās tisro 'samprajñātayogabhūmayaḥ. sattvāpatter\var{sattvāpatter \lem tataḥ sattvāpatter
+  \Msix} anantarā sattvā\-pattisaṃjñikāyāṃ bhūmāv upasthitāsu siddhiṣu asaṃsaktasyāsaṃsaktināmikā
+pañcamī jñānabhūmiḥ syāt. asyāṃ yogī svayam eva vyuttiṣṭhate. etāṃ\var{etāṃ \lem enāṃ \Msix} bhūmiṃ
+prāpto brahmavidvara ity ucyate.
+
+parabrahmātiriktam arthaṃ na bhāvayati yasyāṃ sā parārthābhāvinī ṣaṣṭhī jñānabhūmiḥ syāt. asyāṃ
+yogī paraprabodhita eva vyutthito bhavati. etāṃ\var{etāṃ \lem enāṃ \Msix} prāpto brahmavidvarīyān ity
+ucyate.
+
+turyagā nāma saptamī bhūmiḥ\var{bhūmiḥ \lem jñānabhūmiḥ \Msix} smṛtā. asyāṃ yogī svataḥ parato vā na
+vyutthānaṃ prāpnoti. etāṃ\var{etāṃ \lem enāṃ \Msix} prāpto brahmavidvariṣṭha ity ucyate.
+
+tatra pramāṇabhūtā śrutir atraivoktā pūrvam. ayam eva jīvanmukta ity ucyate sa evātra
+svātmārāmapadenokta ity alaṃ bahūktena.
+%\end{vsid}
+
+%\begin{vsid}{hp01_004}
+\startsloka
+{\bf haṭhavidyāṃ hi matsyendragorakṣādyā vijānate
+svātmārāmo 'thavā yogī jānīte tatprasādataḥ} (1.4)
+\stopsloka
+
+mahatsevitatvād dhaṭhavidyāṃ praśaṃsan svasyāpi mahatsakāśād dhaṭha\-vidyālābhād gauravaṃ
+dyotayati haṭhavidyāṃ hīti. hīti prasiddham. matsyendraś ca gorakṣaś ca tau ādyau yeṣāṃ te
+matsyendragorakṣādyāḥ ādyaśabdena jālandharanāthabhartṛharigopīcandaprabhrtayo grāhyāḥ. te
+haṭhavidyāṃ haṭhayogavidyāṃ vijānate viśeṣeṇa sādhanalakṣaṇabhedaphalair jānantīty arthaḥ.
+
+svātmārāmaḥ svātmārāmanāmā. athavāśabdaḥ samuccaye. yogī yogavān tatprasādataḥ gorakṣaprasādāj
+jānīta ity anvayaḥ. paramamahatā brahmaṇāpīyaṃ vidyā sevitety atra yogiyājñavalkyasmṛtiḥ\comment{?}
+
+\startsloka
+hiraṇyagarbho yogasya vaktā nānyaḥ purātanaḥ
+\stopsloka
+
+vaktṛtvaṃ ca mānasavyāpārapūrvakaṃ bhavatīti mānaso vyāpāro 'rthād āgamaḥ\var{āgamaḥ \lem āgataḥ
+  \Msix}. tathā ca śrutiḥ yan manasā dhyāyati tad vācā vadati\comment{nṛ.pū.tā.u. 1.1 ?}
+iti. bhagavateyaṃ\var{bhagavateyaṃ \lem bhagavatāpīyaṃ \Msix} vidyā bhāgavatān\var{bhāgavatān \lem
+  bhāgavatādau \Msix} uddhavādīn praty uktā. śivas tu yogī prasiddha eva. evaṃ ca sarvottamair
+brahmaviṣṇuśivaiḥ seviteyaṃ vidyā.  na ca brahmasūtrakṛtā vyāsena yogo nirākṛta iti
+śaṅkanīyam. prakṛtisvātantryacidbhedāṃśamātrasya nirākaraṇāt na tu
+bhāvanāviśeṣarūpayogasya. bhāva\-nāyāś ca sarvasaṃmatatvāt tāṃ vinā sukhasyāpy
+asambhavāt. tathoktaṃ bhagavadgītāsu\comment{Bhagavadgītā 2.66} iti\var{iti \lem om \Msix}
+
+\startsloka
+nāsti buddhir ayuktasya na cāyuktasya bhāvanā
+na cābhāvayataḥ śāntir aśāntasya kutaḥ sukham
+\stopsloka
+
+nārāyaṇatīrthair apy uktam\comment{Yogasiddhāntacandrikā, introductory verses, p.1.}
+
+\startsloka
+svātantryasatyatvamukhaṃ\var{mukhaṃ ] sukhaṃ Yogasiddhāntacandrikā, ed. ??} pradhāne 
+  satyaṃ ca cidbhedam abhedavākyaiḥ\comment{cidbhedam abhedavākyaiḥ \Wthree \Msix ] cidbhedagataṃ ca vākyaiḥ in editio princeps
+  and most manuscripts}
+vyāso nirācaṣṭa na bhāvanākhyaṃ 
+yogaṃ svayaṃ nirmitabrahmasūtraiḥ\comment{The ligature {\em bra}, for which I could not find a variant,
+  breaks the metre here.}
+
+api cātmapradaṃ yogaṃ vyākaron matimān svayam
+bhāṣyādiṣu tatas tatrāpy ācāryapramukhair mataḥ
+mato yogo bhagavatā gītāyām adhiko 'nyataḥ
+kṛtaḥ śukādibhis tasmād atra santo 'tisādarāḥ
+
+iti
+
+vedeṣu yajñeṣu tapaḥsu caiva
+dāneṣu yat puṇyaphalaṃ\var{puṇyaphalaṃ \lem kurmaphalaṃ \Msix} pradiṣṭam 
+atyeti tat sarvam idaṃ viditvā
+yogī paraṃ sthānam upaiti cādyam\comment{BhG 8.28}
+\stopsloka
+
+iti bhagavadukteḥ. kiṃ bahunā jijñāsur api yogasya śabdabrahmātivartate\comment{BhG 6.44} iti
+vadatā bhagavatā yogajijñāsor apy autkṛṣṭyaṃ varṇitaṃ kim uta yoginaḥ. nāradādibhaktaśreṣṭhair
+yājñavalkyādijñānimukhyaiś cāsyāḥ sevanād bhaktajñāninām apy aviruddhety uparamyate.
+%\end{vsid}
+
+%\begin{vsid}{hp01_005}
+\startsloka
+{\bf śrīādināthamatsyendraśābarānandabhairavāḥ
+cauraṅgīmīnagorakṣavirūpākṣabileśayāḥ (1.5)}
+\stopsloka
+
+haṭhayoge pravṛttiṃ janayituṃ haṭhavidyayā prāptaiśvaryān siddhān āha śrīādināthetyādinā.
+ādināthaḥ śivaḥ sarveṣāṃ nāthānāṃ prathamo nāthaḥ. tato nāthasampradāyaḥ pravṛtta iti
+nāthasampradāyino vadanti.
+
+matsyendrākhyaś cādināthaśiṣyaḥ. atraivaṃ kiṃvadantī. kadācid ādi\-nāthaḥ kasmiṃścid dvīpe sthitaḥ
+tatra vijanam iti matvā girijāyai yogam upadiṣṭavān. tīrasamīpanīrasthaḥ kaścana\var{kaścana ]
+kaścan \EdLo} matsyas taṃ yogopadeśaṃ śrutvā ekāgracitto niścalakāyo 'vatasthe. taṃ tādṛśaṃ
+dṛṣṭvānena yogaḥ śrutaḥ iti taṃ matvā kṛpālur ādinātho jalena prokṣitavān. sa ca prokṣaṇamātrād
+divyakāyo matsyendraḥ siddho 'bhūt.  tam eva matsyendranātha iti vadanti.
+
+śābaranāmā kaścit siddhaḥ. ānandabhairavanāmānyaḥ. eteṣām itaretaradvandvaḥ. chinnahastapādaṃ
+puruṣaṃ hindusthānabhāṣāyāṃ cauraṅgīti vadanti. kadācid ādināthāl labdhayogasya bhuvaṃ paryaṭato
+matsyendranāthasya kṛpāvalokanamātrāt kutracid araṇye sthitaś cauraṅgy aṅkuritahastapādo babhūva.
+sa ca tat\var{ca tat \lem caitat \Msix}kṛpayā sañjātahastapādo 'ham iti matvā tatpādayoḥ praṇipatya
+mamānugrahaṃ kurv iti prārthitavān. matsyendro 'pi tam anugṛhītavān. tasyānugrahāc cauraṅgīti
+prasiddhaḥ siddhaḥ so 'bhūt.
+
+mīno mīnanāthaḥ gorakṣo gorakṣanāthaḥ virūpākṣanāmā bileśayanāmā ca. cauraṅgīprabhṛtīnāṃ
+dvandvasamāsaḥ. 
+%\end{vsid}
+
+%\begin{vsid}{hp01_006}
+\startsloka
+{\bf manthāno bhairavo yogī śuddhabuddhaś ca kanthaḍiḥ
+koraṇṭakaḥ surānandaḥ siddhapādaś ca carpaṭiḥ (1.6)}
+\stopsloka
+
+manthāna iti. manthānaḥ bhairavaḥ. yogīti manthānaprabhṛtīnāṃ sarveṣāṃ viśeṣaṇam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_007}
+\startsloka
+{\bf kānerī pūjyapādaś ca nityanātho nirañjanaḥ
+kapālī bindunāthaś ca kākacaṇḍīśvarāhvayaḥ (1.7)}
+\stopsloka
+
+kānerīti.\var{kānerīti \om \Msix} kākacaṇḍīśvara ity āhvayo nāma yasya sa tathā. anye spaṣṭāḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_008}
+\startsloka
+  {\bf allamaḥ\comment{Here and in the commentary mss.\ read short or long ā in various patterns.
+      The text is here standardized following the critical edition of the mūla in the HP project.}
+                                                   prabhudevaś ca ghoḍācolī ca ṭiṇṭiṇiḥ.
+bhālukī nāgadevaś ca khaṇḍakāpālikas tathā (1.8)}
+\stopsloka
+
+allama iti.\var{allama iti \lem om \Msix} tathāśabdaḥ samuccaye.
+%\end{vsid}
+
+%\begin{vsid}{hp01_009}
+\startsloka
+{\bf ityādayo mahāsiddhā haṭhayogaprabhāvataḥ
+khaṇḍayitvā kāladaṇḍaṃ brahmāṇḍe vicaranti te (1.9)}
+\stopsloka
+
+iti pūrvoktā ādayo yeṣāṃ te tathā. ādiśabdena tārānāthādayo grāhyāḥ. mahāntaś ca te siddhāś ca
+apratihataiśvaryā ity arthaḥ. haṭhayogasya prabhāvāt sāmarthyād iti haṭhayogaprabhāvataḥ.
+pañcamyās tasil. kālo mṛtyus tasya daṇḍanaṃ daṇḍaḥ dehaprāṇaviyogānukūlo vyāpāraḥ taṃ
+khaṇḍayitvā chittvā mṛtyuṃ jitvety arthaḥ. brahmāṇḍe brahmāṇḍamadhye vicaranti
+viśeṣeṇāvyāhatagatyā carantīty arthaḥ. tad uktaṃ bhāgavate
+
+\startsloka  % check sources
+yogeśvarāṇāṃ gatim āhur antar
+bahistrilokyāḥ pavanāntarātmanām \comment{Bhāgavatapurāṇa 2.2.23ab} iti.
+%na karmabhis tāṃ gatim āpnuvanti
+%vidyātapoyogasamādhibhājām   the rest of the verse was added in the e-text of unknown provenance. 
+\stopsloka
+%\end{vsid}
+
+%\begin{vsid}{hp01_010}
+\startsloka
+{\bf aśeṣatāpataptānāṃ samāśrayamaṭho haṭhaḥ
+aśeṣayogayuktānām ādhārakamaṭho haṭhaḥ (1.10)}
+\stopsloka
+
+haṭhasyāśeṣatāpanāśakatvam aśeṣayogasādhakatvaṃ ca maṭhakamaṭha\-rūpakeṇāha aśeṣeti. aśeṣāḥ
+ādhyātmikādhibhautikādhidaivikabhedena trividhāḥ. tatrādhyātmikaṃ dvividham śārīraṃ mānasaṃ ca.
+tatra śārīraṃ duḥkhaṃ vyādhijaṃ mānasaṃ duḥkhaṃ kāmādijam. ādhibhautikaṃ
+vyāghrasarpādijanitam. ādhidaivikaṃ grahādijanitam. te ca te tāpāś ca tais taptānāṃ santaptānāṃ
+puṃsāṃ haṭho haṭhayogaḥ. samyag āśrīyata iti samāśrayaḥ āśrayabhūto maṭhaḥ maṭha eva. tathā haṭhaḥ
+aśeṣayogayuktānāṃ aśeṣayogayuktāḥ\var{aśeṣayogayuktāḥ ] aśeṣā ye yogayuktāḥ \Msix \Wthree (variant
+not reported by \EdLo)} mantrayogakarmayogādiyuktās teṣām ādhārabhūtaḥ kamaṭhaḥ.\comment{\EdLo\
+    reports an inserted line here in ms. I. The line occurs also in \Tue, but is there deleted. In
+    VI: kamaṭha eva yathā taraṇikiraṇatāpataptānāṃ puṃsāṃ āśrayo maṭhaḥ} evaṃ trividhatāpataptānāṃ
+  puṃsām\var{puṃsām \lem om \Msix} āśrayo haṭhaḥ. yathā ca viśvādhāraḥ kamaṭha evaṃ
+  nikhilayoginām\var{yoginām \lem yogānām \Msix} ādhāro haṭha ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_011}
+\startsloka
+{\bf haṭhavidyā paraṃ gopyā yoginā siddhim icchatā
+bhaved vīryavatī guptā nirvīryā tu prakāśitā (1.11)}
+\stopsloka
+
+athākhilavidyāpekṣayā haṭhavidyāyā atigopyatvam āha haṭhavidyeti. siddhim aṇimādyaiśvaryam icchatā.
+yad vā siddhiṃ kaivalyasiddhim icchatā vāñchatā yoginā haṭhavidyā haṭhayogavidyā param atyantaṃ
+gopyā gopanīyā\var{gopanīyā \lem om \Msix} gopanārhāstīti. tatra hetum āha yato guptā haṭhavidyā
+vīryavatī bhavet apratihataiśvaryajananasamarthā syāt. kaivalyajananasamarthā
+kaivalyasiddhijananasamarthā vā syāt\var{syāt \lem bhavet \Msix}.  prakāśitā prasiddhiṃ gamitā tu
+nirvīryā. dīrghakālasevitāpi apratihataiśvaryajananāsamarthā kaivalyasiddhijananāsamarthā vā
+syāt. atha yogādhikārī\comment{Source unknown. ??}
+
+\startsloka
+  jitākṣāya śāntāya saktāya muktau
+  vihīnāya doṣair asaktāya bhuktau
+  ahīnāya doṣetarair uktakartre
+  pradeyo na deyo haṭhaś cetarasmai
+\stopsloka
+
+yājñavalkyaḥ
+
+\startsloka
+vidhyuktakarmasaṃyuktaḥ kāmasaṅkalpavarjitaḥ 
+yamaiś ca niyamair yuktaḥ sarvasaṅgavivarjitaḥ
+kṛtavidyo jitakrodhaḥ satyadharmaparāyaṇaḥ
+guruśuśrūṣaṇarataḥ pitṛmātṛparāyaṇaḥ
+svāśramasthaḥ sadācāro vidvadbhiś ca suśikṣitaḥ\comment{Yogayājñavalkya 5.3–5ab} 
+\stopsloka
+
+iti. śiśnodararatāyaiva na deyaṃ veṣadhāriṇe iti kutracit. atra yogacintāmaṇi\-kārāḥ yady api
+
+\startsloka
+brāhmaṇakṣatriyaviśāṃ strīśūdrāṇāṃ ca pāvanam 
+śāntaye karmaṇām anyad yogān nāsti vimuktaye\comment{Yogacintāmaṇi, p. 57.}
+\stopsloka
+
+ityādipurāṇavākyeṣu prāṇimātrasya yoge 'dhikāra upalabhyate tathāpi mokṣarūpaphalavati yoge
+viraktasyaivādhikāra ucitaḥ. tathā ca vāyusaṃhitāyāṃ\comment{Śivapurāṇa 37.12}
+
+\startsloka
+dṛṣṭe tathānuśravike viraktaṃ viṣaye manaḥ
+yasya tasyādhikāro 'smin yoge nānyasya kasyacit
+\stopsloka
+
+sureśvarācāryāḥ\comment{?}
+
+\startsloka
+ihāmutraviraktasya saṃsāraṃ prajihāsataḥ
+jijñāsor eva kasyāpi yoge 'sminn adhikāritā
+\stopsloka
+
+ity āhuḥ. vṛddhair apy uktam—
+
+\startsloka
+naitad deyaṃ durvinītāya jātu 
+jñānaṃ guptaṃ tad dhi samyak phalāya
+asthāne hi sthāpyamānaiva vācāṃ
+devī kopān nirdahen no 'cirāya\var{no 'cirāya \lem taṃ cirāya \Msix} 
+\stopsloka
+%\end{vsid}
+
+iti.
+
+%\begin{vsid}{hp01_012}
+\startsloka
+{\bf surājye dhārmike deśe subhikṣe nirupadrave
+dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite
+ekānte maṭhikāmadhye sthātavyaṃ haṭhayoginā (1.12)}
+\stopsloka
+
+atha haṭhābhyāsayogyaṃ deśam āha sārdhena surājya iti. rājñaḥ karma bhāvo vā rājyaṃ tac chobhanaṃ
+yasmin sa surājyas tasmin surājye. yathā rājā tathā prajāḥ iti mahadukteḥ. rājñaḥ śobhanatvāt
+prajānām api śobhanatvaṃ sūcitam. dhārmike dharmavati\var{dharmavati \Wthree \Tue \Msix \lem
+  dharmayati \EdLo} anena haṭhābhyāsino 'nukūlāhārādilābhaḥ sūcitaḥ. subhikṣa ity anenānāyāsena
+tallābhaḥ sūcitaḥ. nirupadrave cauravyāghrādyupadravarahite. etena deśasya dīrghakālavāsayogyatā
+sūcitā. dhanuṣaḥ pramāṇaṃ dhanuḥpramāṇaṃ caturhastamātraṃ tatparyantam. śilāgnijalavarjite śilā
+prastaraḥ agnir vahniḥ jalaṃ toyaṃ tair varjite rahite.  yatrāsanaṃ tataś caturhastamātre
+śilāgnijalāni na syur ity arthaḥ. tena śītoṣṇādi\var{ śītoṣṇādi \lem śītoṣṇa \Msix}vikārābhāvaḥ
+sūcitaḥ. ekānte vijane. anena\var{anena \lem ity anena \Msix} janasamāgamābhāvāt kalahavārtādyabhāvaḥ
+sūcitaḥ. janasaṃmarde tu kalahādikaṃ syād eva. tad uktaṃ bhāgavate
+
+\startsloka
+  vāse bahūnāṃ kalaho bhaved vārtā dvayor api\comment{Bhāgavatapurāṇa 11.9.10ab} 
+\stopsloka
+
+iti. tādṛśe maṭhikāmadhye\var{tādṛśe maṭhikāmadhye \Wthree \EdMu \lem tādṛśe deśe maṭhikāmadhye
+  \Tue Bo3 \Msix, om \EdLo}.  alpo maṭho maṭhikā alpīyasi kan. tasyāḥ madhye haṭhayoginā haṭhābhyāsī
+yogī haṭhayogī tena.  śākapārthivādivat samāsaḥ. sthātavyaṃ sthātuṃ yogyam. maṭhikāmadhye ity anena
+śītātapādijanitakleśābhāvaḥ sūcitaḥ. atra yuktāhāravihārasya\var{vihārasya \lem vihāreṇa \Msix}
+haṭhayogasya siddhaye ity ardhaṃ kenacit kṣiptatvān na vyākhyātam. mūlaślokānām evātra
+vyākhyānam. evam agre 'pi ye mayā na vyākhyātā ślokā haṭhapradīpikāyām upalabhyeran te sarve kṣiptā
+iti boddhavyam\var{boddhavyam \lem bodhyam \Msix}.
+%\end{vsid}
+
+%\begin{vsid}{hp01_013}
+\startsloka
+{\bf alpadvāram arandhragartavivaraṃ nātyuccanīcāyataṃ
+samyaggomayasāndraliptam amalaṃ niḥśesajantūjjhitam.
+bāhye maṇḍapavedikūparuciraṃ prākārasaṃveṣṭitaṃ
+proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair
+\hfill haṭhābhyāsibhiḥ (1.13)}
+\stopsloka
+
+atha maṭhalakṣaṇam āha alpadvāram iti\var{alpadvāram iti \lem alpadvāreti \Msix}. alpaṃ dvāram iti
+yasmin tat tādṛśam. randhro gavākṣādiḥ garto nimnapradeśaḥ vivaro mūṣakāhibilaṃ\var{mūṣakāhi \lem
+  mūṣakāhi \Msix} te na santi yasmin tat tādṛśam. atyuccaṃ ca tan nīcaṃ cātyuccanīcaṃ tac ca tad
+āyataṃ cātyuccanīcāyataṃ viśeṣaṇaṃ viśeṣyeṇa bahulam\comment{Aṣṭādhyāyī 2.1.57} ity atra
+bahulagrahaṇād viśeṣaṇānāṃ karmadhārayaḥ.
+
+nanūccanīcāyataśabdānāṃ bhinnārthakānāṃ kathaṃ karmadhārayaḥ. tatpuruṣaḥ samānādhikaraṇaḥ
+karmadhārayaḥ\comment{Aṣṭādhyāyī 1.2.42} iti tallakṣaṇād iti cen na. maṭhe teṣāṃ
+sāmānādhikaraṇyāsambhavāt.\var{sāmānādhikaraṇyāsambhavāt II V XII \lem samānādhikāraṇyasambhavāt IV
+  samānādhi\-karaṇy\-asambhavāt VIII X} na cātyucca\var{cātyucca \lem atyucca \Msix}nīcāyataṃ
+nātyuccanīcāyataṃ. naśabdena samāsān nalopābhāvaḥ. neti pṛthakpadaṃ vā. na\var{na \lem om \Msix}
+atyucce ārohaṇe śramaḥ syād atinīce 'varohaṇe śramo bhavet. atyāyate dūraṃ dṛṣṭir gacchet
+tannirākaraṇārtham uktaṃ nātyuccanīcāyatanam iti. samyak samīcīnatayā gomayena
+gopurīṣeṇa\var{gopurīṣeṇa \lem goḥ purīṣeṇa \Msix} sāndraṃ yathā bhavati tathā liptam. amalaṃ
+nirmalam. niḥśesā nikhilā ye jantavo maśakamatkuṇādyās tair ujjhitam tyaktaṃ rahitam\var{tyaktaṃ
+  rahitam \lem tyakte rahite \Msix}.
+
+bāhye maṭhād bahiḥpradeśe. maṇḍapaḥ śālāviśeṣaḥ vediḥ pariṣkṛtā bhūmiḥ kūpo jalāśayaviśeṣaḥ.  tai
+ruciraṃ ramaṇīyam. prākāreṇa vara\-ṇena samyag veṣṭitaṃ parito\var{veṣṭitaṃ parito \lem parito
+  veṣṭitaṃ parito \Msix} bhittiyuktam ity arthaḥ.  haṭhābhyāsibhiḥ haṭha\-yogābhyasanaśīlaiḥ
+siddhaiḥ. idaṃ pūrvoktam alpadvārādikaṃ yogamaṭhasya lakṣaṇaṃ svarūpaṃ proktaṃ kathitam.
+
+nandikeśvarapurāṇe tv evaṃ maṭhalakṣaṇam uktam\comment{Quoted via the Yogacintāmaṇi (p.90)}
+
+\startsloka
+mandiraṃ ramyavinyāsaṃ manojñaṃ gandhavāsitam
+dhūpāmodādisurabhi kusumotkaramaṇḍitam
+munitīrthanadīvṛkṣapadminīśailaśobhitam
+citrakarmanibaddhaṃ ca citrabhedavicitritam
+kuryād yogagṛhaṃ dhīmān suramyaṃ śubhavartmanā
+dṛṣṭvā citragatāṃś chāntān munīn yāti manaḥ śamam
+siddhān dṛṣṭvā citragatān matir abhyudyame bhavet
+madhye yogagṛhasyātha likhet saṃsāramaṇḍalam
+śmaśānaṃ ca mahāghoraṃ narakāṃś ca likhet kvacit
+tān dṛṣṭvā bhīṣaṇākārān saṃsāre sāravarjite\var{saṃsāre sāravarjite \lem saṃsāraṃ sāravarjitaṃ \Msix} 
+anavasādo bhavati yogī siddhyabhilāṣukaḥ
+paśyaṃś ca vyādhitān jantūn natān mattāṃś caladvraṇāt
+\stopsloka
+%\end{vsid}
+
+%\begin{vsid}{hp01_014}
+\startsloka
+{\bf evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ
+gurūpadiṣṭamārgeṇa yogam eva sadābhyaset (1.14)}
+\stopsloka
+
+maṭhalakṣaṇam uktvā maṭhe yat kartavyaṃ tad āha evaṃvidha iti. evaṃ pūrvoktā vidhā prakāro yasya
+sa tathā pūrvoktalakṣaṇa ity arthaḥ. tasmin sthitvā sthitiṃ kṛtvā sarvā yāś cintās tābhir viśeṣeṇa
+varjito rahito 'śeṣacintārahitaḥ. guruṇopadiṣṭo yo mārgaḥ haṭhābhyāsaprakārarūpas tena sadā nityaṃ
+yogam evābhyaset. evaśabdenābhyāsāntarasya yoge vighna\-karatvaṃ sūcitam. gurūpadiṣṭamārgeṇety anena
+gurūpadeśaṃ vinā yogo na sidhyatīti sūcitam. tad uktaṃ yogabīje
+
+\startsloka
+marujjayo yasya siddhas taṃ seveta guruṃ sadā
+guruvaktraprasādena kuryāt prāṇajayaṃ budhaḥ\comment{Yogabīja 91}
+\stopsloka
+
+rājayoge\comment{Amanaskayoga 1.40 = Yogabīja 66}
+
+\startsloka
+vedāntatarkoktibhir āgamaiś ca
+nānāvidhaiḥ śāstrakadambakaiś ca
+dhyānādibhiḥ satkaraṇair na gamyaś 
+cintāmaṇir\var{cintāmaṇir \lem cintāmaṇiṃ \Msix} hy ekaguruṃ vihāya
+\stopsloka
+
+skandapurāṇe
+
+\startsloka
+ācāryād yogasarvasvam avāpya sthiradhīḥ svayam
+yathoktaṃ labhate tena prāpnoty api ca nirvṛtim\comment{nirvṛtim \Wthree \Tue \EdMu \lem nivṛtim \EdLo}
+\stopsloka
+
+sureśvarācāryaḥ
+
+\startsloka
+guruprasādāl labhate yogam aṣṭāṅgasaṃyutam
+śivaprasādāl labhate yogasiddhiṃ ca śāśvatīm
+\stopsloka
+
+śrutiś ca
+
+\startsloka
+yasya deve parā bhaktir yathā deve tathā gurau
+tasyaite kathitā hy arthāḥ prakāśante mahātmanaḥ\comment{Śvetāśvataropaniṣat 6.23} 
+\stopsloka
+
+iti. ācāryavān puruṣo veda\comment{Chāndogyopaniṣat 6.14.2} iti ca.
+%\end{vsid}
+
+%\begin{vsid}{hp01_015}
+\startsloka
+{\bf atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ
+janasaṅgaś ca laulyaṃ ca ṣaḍbhir yogo vinaśyati (1.15)}
+\stopsloka
+
+atha yogābhyāsapratibandhakān āha atyāhāra iti. atiśayita āhāro 'tyāhāraḥ
+kṣudhāpekṣayādhikabhojanam. prayāsaḥ śramajananānukūlo vyāpāraḥ. prakṛṣṭo jalpaḥ prajalpo
+bahubhāṣaṇam. śītodakena prātaḥsnānanakta\var{nakta ] om \EdLo}bhojanaphalāhārādirūpaniyamasya
+grahaṇaṃ niyamagrahaḥ.  janānāṃ saṅgo janasaṅgaḥ kāmādijanakatvāt. lolasya bhāvaḥ laulyaṃ
+cāñcalyam. ṣaḍbhir atyāhārādibhiḥ abhyāsapratibandhād yogo vinaśyati viśeṣeṇa naśyati.
+%\end{vsid}
+
+%\begin{vsid}{hp01_016}
+\startsloka
+{\bf utsāhāt sāhasād dhairyāt tattvajñānāc ca niścayāt
+janasaṅgaparityāgāt ṣaḍbhir yogaḥ prasiddhyati (1.16)}
+\stopsloka
+
+atha yogasiddhikarān āha utsāhād iti. viṣayapravaṇam cittaṃ nirotsyāmy\comment{nirotsyāmy \Msix \lem
+  nirotsāmy \EdLo} evety udyamam utsāhaḥ.  sādhyatvāsādhyatve aparibhāvya sahasā pravṛttiḥ
+sāhasam. yāvajjīvanaṃ setsyaty evety akhedo dhairyam. viṣayā mṛgatṛṣṇājalavad asantaḥ brahmaiva
+satyam\var{satyam \Tue \lem sad \Wthree \Msix} iti vāstavikaṃ jñānaṃ tattvajñānaṃ yogāṅgānāṃ
+vāstavikaṃ jñānaṃ vā. śāstraguruvākyeṣu viśvāso niścayaḥ śraddheti yāvat. janānāṃ
+yogābhyāsa\var{yogābhyāsa \lem yogābhyāsaḥ \EdLo}pratikūlānāṃ yaḥ saṅgas tasya parityāgāt. ṣaḍbhir
+ebhir yogaḥ prakarṣeṇāvilambena siddhyatīty arthaḥ.
+
+% atha yamaniyamāḥ—
+
+% [ahiṃsā satyam asteyaṃ brahmacaryaṃ kṣamā dhṛtiḥ.
+% dayārjavaṃ mitāhāraḥ śaucaṃ caiva yamā daśa.|*||
+% tapaḥ santoṣa āstikyaṃ dānam īśvarapūjanam.
+% siddhāntavākyaśravaṇaṃ hrīmatī ca tapo hutam.
+% niyamā daśa samproktā yogaśāstraviśāradaiḥ.|*||\lem
+%\end{vsid}
+
+%\begin{vsid}{hp01_017}
+\startsloka
+{\bf haṭhasya prathamāṅgatvād āsanaṃ pūrvam ucyate
+kuryāt tad āsanaṃ sthairyam ārogyaṃ cāṅgalāghavam (1.17)}
+\stopsloka
+
+ādāv āsanakathane saṅgatiṃ sāmānyatas tatphalaṃ cāha haṭhasyeti. haṭhasya
+
+\startsloka
+āsanaṃ kumbhakaṃ citraṃ\var{citraṃ \lem om \EdLo} mudrākhyaṃ karaṇaṃ tathā
+atha nādānusandhānam\comment{HP 1.56}
+\stopsloka
+
+iti vakṣyamāṇāni catvāry aṅgāni. pratyāhārādisamādhyantānāṃ nādānusandhāne 'ntarbhāvaḥ. tanmadhye
+āsanasya prathamāṅgatvāt pūrvam āsanam ucyate iti sambandhaḥ.
+
+tad āsanaṃ sthairyaṃ dehasya\var{dehasya \lem om \EdLo} manasaś ca\var{ca \Wthree \lem om \EdLo}
+cāñcalyarūparajodharmanāśakatvena sthiratāṃ kuryāt. āsanena rajo hanti iti vākyāt. ārogyaṃ
+cittavikṣepakarogābhāvaḥ\var{rogābhāvaḥ \lem rogābhāvaṃ \Msix}. rogasya cittavikṣepakatvam uktaṃ
+pātañjalasūtre vyādhistyānasaṃśayapramādālasyāviratibhrāntidarśanālabdhabhūmikatvānavasthitatvāni
+cittavikṣepās te 'ntarāyāḥ\comment{YS 1.30} iti. aṅgānāṃ lāghavaṃ laghutvam.
+gauravarūpatamodharmanāśakatvam apy etenoktam. cakārāt kṣudvṛddhyādikam api bodhyam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_018}
+\startsloka
+{\bf vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ\var{yogibhiḥ \lem munibhiḥ \Msix}
+aṅgīkṛtāny āsanāni kathyante kānicin mayā (1.18)}
+\stopsloka
+
+vasiṣṭhādisaṃmatāsanamadhye śreṣṭhāni mayocyanta ity āha vasiṣṭhādyair iti. vasiṣṭha ādyo yeṣāṃ
+yājñavalkyādīnāṃ tair munibhir mananaśīlaiḥ. cakārān mantrādiparaiḥ. matsyendra ādyo yeṣāṃ
+jālandharanāthādīnāṃ taiḥ. yogibhiḥ\var{yogibhiḥ \lem munibhiḥ \Msix} haṭhābhyāsibhiḥ. cakārān
+mudrādiparaiḥ. aṅgīkṛtāni caturaśīty āsanāni tanmadhye kānicit śreṣṭhāni mayā kathyante. yady apy
+ubhayor api mananahaṭhābhyāsau stas tathāpi vasiṣṭhādīnāṃ mananaṃ mukhyaṃ matsyendrādīnāṃ
+haṭhābhyāso mukhya iti pṛthag grahaṇam.
+%\end{vsid}
+
+%  * 50728.jpg
+
+%\begin{vsid}{hp01_019}
+\startsloka
+{\bf jānūrvor antare samyak kṛtvā pādatale ubhe
+ṛjukāyaḥ samāsīnaḥ svastikaṃ tat pracakṣate (1.19)}
+\stopsloka
+
+tatra sukaratvāt prathamaṃ svastikāsanam āha jānūrvor iti. jānu ca ūruś ca. atra jānuśabdena
+jānusaṃnihito jaṅghāpradeśo grāhyaḥ. jaṅghorvor iti pāṭhas tu sādhīyān. tayor antare madhye ubhe
+pādayos tale talapradeśau kṛtvā ṛjukāyaḥ samakāyaḥ yatra samāsīno bhavet tad āsanaṃ svastikaṃ
+svastikākhyaṃ pracakṣate vadanti. yogina iti śeṣaḥ. śrīdhareṇoktam\var{śrīdharenoktaṃ up to budhāḥ
+  iti \lem om \Msix}
+
+\startsloka
+ūrujaṅghāntar ādhāya prapade jānumadhyage
+yogino yad avasthānaṃ svastikaṃ tad vidur budhāḥ iti
+\stopsloka
+%\end{vsid}
+
+%\begin{vsid}{hp01_020}
+\startsloka
+{\bf savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niyojayet
+dakṣiṇe 'pi tathā savyaṃ gomukhaṃ gomukhākṛti (1.20)}
+\stopsloka
+
+gomukhāsanam āha savya iti. savye vāme pṛṣṭhasya pārśve sampradāyāt kaṭer adhobhāge dakṣiṇaṃ
+gulphaṃ nitarāṃ yojayet. dakṣiṇe 'pi pṛṣṭhapārśve tathā savyavat savyaṃ gulphaṃ niyojayet.
+gomukhasyākṛtir iva ākṛtir yasya tat tādṛśaṃ gomukhaṃ gomukhasaṃjñakam āsanaṃ bhavet.
+%\end{vsid}
+
+%\begin{vsid}{hp01_021}
+\startsloka
+{\bf ekaṃ pādaṃ tathaikasmin vinyased ūruṇi sthiram\var{sthiram \lem sthitam \Msix}
+itarasmiṃs tathā coruṃ vīrāsanam itīritam (1.21)}
+\stopsloka
+
+vīrāsanam āha ekam iti. ekaṃ dakṣiṇaṃ pādam. tathā pādapūraṇe. ekasmin vāmoruṇi\var{vāmoruṇi \lem
+  vāme ūruṇi \Msix} sthiraṃ\var{sthiram \lem sthitam \Msix} vinyaset.  itarasmin vāme pāde ūruṃ
+dakṣiṇaṃ vinyaset. tad vīrāsanam itīritaṃ kathitam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_022}
+\startsloka
+{\bf gudaṃ nirudhya gulphābhyāṃ vyutkrameṇa samāhitaḥ
+kūrmāsanaṃ bhaved etad iti yogavido viduḥ (1.22)}
+\stopsloka
+
+kūrmāsanam āha gudam iti. gulphābhyāṃ gudaṃ nirudhya niyamya\var{niyamya \lem om \EdLo} vyutkrameṇa
+yatra samyag āhitaḥ sthito bhavet etat kūrmāsanaṃ bhavet iti yogavido vidur ity anvayaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_023}
+\startsloka
+{\bf padmāsanaṃ tu saṃsthāpya jānūrvor antare karau
+  niveśya bhūmau saṃsthāpya vyomasthaṃ
+  \hfill kukkuṭāsanam (1.23)}
+\stopsloka
+
+kukkuṭāsanam āha padmāsanaṃ tv iti. padmāsanaṃ tu ūrvor upari uttānacaraṇasthāpanarūpaṃ samyak
+sthāpayitvā. jānupadena jānusaṃnihito jaṅghāpradeśaḥ. tac ca ūruś ca jānūr tayor antare madhye
+karau niveśya bhūmau saṃsthāpya. karāv ity atrāpi sambadhyate. vyomasthaṃ khasthaṃ padmāsanasadṛśaṃ
+yat tat kukkuṭāsanam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_024}
+\startsloka
+{\bf kukkuṭāsanabandhastho dorbhyāṃ sambadhya kandharām
+bhaved kūrmavad uttāna etad uttānakūrmakam (1.24)}
+\stopsloka
+
+uttānakūrmāsanam āha kukkuṭāsaneti. kukkuṭāsanasya yo bandhaḥ pūrvaślokoktas tasmin sthitaḥ
+dorbhyāṃ bāhubhyāṃ kandharāṃ grīvāṃ sambadhya kūrmavad uttāno yasmin bhaved etad āsanam
+uttānakūrmakaṃm nāma.
+%\end{vsid}
+
+%\begin{vsid}{hp01_025}
+\startsloka
+{\bf pādāṅguṣṭhau tu pāṇibhyāṃ gṛhītvā śravaṇāvadhi
+dhanurākarṣaṇaṃ kuryād dhanurāsanam ucyate (1.25)}
+\stopsloka
+
+dhanurāsanam āha pādāṅguṣṭhau tv iti. pāṇibhyāṃ pādayor aṅguṣṭhau gṛhītvā śravaṇāvadhi
+karṇaparyantaṃ dhanuṣa ākarṣaṇaṃ yathā bhavati tathā kuryāt. gṛhītāṅguṣṭham ekaṃ pāṇiṃ prasāritaṃ
+kṛtvā gṛhītāṅguṣṭham itaraṃ pāṇiṃ\var{pāṇiṃ \lem paṇiṃ \Msix} karṇaparyantam ākuñcitaṃ kuryād ity
+arthaḥ. etad dhanurāsanam ucyate.
+%\end{vsid}
+
+%\begin{vsid}{hp01_026}
+\startsloka
+{\bf vāmorumūlārpitadakṣapādaṃ
+jānor bahir veṣṭitavāmapādam
+pragṛhya tiṣṭhet parivartitāṅgaḥ 
+śrīmatysanāthoditam āsanaṃ syāt (1.26)}
+\stopsloka
+
+matyendrāsanam āha vāmorumūleti. vāmorumūle 'rpitaḥ sthāpito yo dakṣapādaḥ taṃ sampradāyāt
+pṛṣṭhatogatavāmapāṇinā gulphasyoparibhāge pragṛhya. jānoḥ dakṣiṇapādajānoḥ bahiḥ\var{bahiḥ \lem
+  bahiḥ bahiḥ \Msix}pradeśe veṣṭito yo vāmapādaḥ taṃ vāmapādajānoḥ bahir veṣṭitadakṣiṇapāṇinā
+aṅguṣṭhe pragṛhya. parivartitāṅgaḥ vāmabhāgena pṛṣṭhatomukhaṃ yathā syād evaṃ parivartitaṃ
+parāvartitam \var{parāvartitam \Wthree \Tue \EdMu \Msix \lem parivartitam \EdLo} aṅgaṃ yena sa tathā
+tādṛśo yatra tiṣṭhet sthitiṃ kuryāt tad āsanaṃ matysendranāthenoditam kathitaṃ syāt. taduditatvāt
+tannāmakam eva vadanti. evaṃ dakṣorumūlārpitavāmapādaṃ pṛṣṭhatogatadakṣiṇapāṇinā pragṛhya vāmajānor
+bahirveṣṭitadakṣapādaṃ dakṣiṇapādajānoḥ bahirveṣṭitavāmapāṇinā pragṛhya dakṣabhāgena pṛṣṭhatomukhaṃ
+yathā syād evaṃ parivartitāṅgaś cābhyaset.
+%\end{vsid}
+
+%\begin{vsid}{hp01_027}
+\startsloka
+{\bf matsyendrapīṭhaṃ jaṭharapradīptiṃ
+pracaṇḍarugmaṇḍalakhaṇḍanāstram
+abhyāsataḥ kuṇḍalinīprabodhaṃ
+candrasthiratvaṃ ca dadāti puṃsām (1.27)}
+\stopsloka
+
+matsyendrāsanasya phalam āha matsyendreti. pracaṇḍaṃ duḥsahaṃ yat rujāṃ rogāṇāṃ maṇḍalaṃ samūhaḥ
+tasya khaṇḍane chedane 'stram astram iva tādṛśaṃ matsyendrapīṭhaṃ matsyendrāsanam. abhyāsataḥ
+pratyahamāvartanarūpād abhyāsāt puṃsāṃ jaṭharasya jaṭharāgneḥ\var{jaṭharāgneḥ \lem jaṭhārāgneḥ
+  \EdLo} prakṛṣṭāṃ dīptiṃ vṛddhim dadāti.  tathā kuṇḍalinyā ādhāraśakteḥ prabodhaṃ nidrābhāvaṃ tathā
+candrasya tālunaḥ uparibhāge sthitasya nityaṃ kṣarataḥ sthiratvaṃ kṣaraṇābhāvaṃ ca dadātīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_028}
+\startsloka
+{\bf prasārya pādau bhuvi daṇḍarūpau
+dorbhyāṃ padagradvitayaṃ gṛhītvā
+jānūparinyastalalāṭadeśo
+vased idaṃ paścimatānam āhuḥ (1.28)}
+\stopsloka
+
+paścimatānāsanam āha prasāryeti. bhuvi bhūmau daṇḍasya rūpam iva rūpaṃ yayos tau daṇḍākārau
+śliṣṭagulphau prasārya prasāritaṃ\var{prasāritaṃ \lem prasāritau \Msix} kṛtvā. dorbhyām
+ākuñcitatarjanībhyāṃ bhujābhyāṃ padoḥ pādayoś cāgre\var{cāgre \lem agre \Msix (better??)} agrabhāgau
+tayor dvitayaṃ dvayam aṅguṣṭhapradeśayugmaṃ balād ākarṣaṇapūrvakaṃ yathā jānvadhobhāgasya bhūmer
+utthānaṃ na syāt tathā gṛhītvā. jānor\var{jānor \lem jānunor \Msix} uparinyasto lalāṭadeśo yena
+tādṛśo yatra vaset. idaṃ paścimatānanāmakam āsanam āhuḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_029}
+\startsloka
+{\bf iti paścimatānam āsanāgryaṃ 
+pavanaṃ paścimavāhinaṃ karoti
+udayaṃ jaṭharānalasya kuryād
+udare kārśyam arogatāṃ ca puṃsām (1.29)}
+\stopsloka
+
+atha tatphalam itīti. iti pūrvoktam āsaneṣv 'py agryaṃ mukhyaṃ paścimatānam pavanaṃ prāṇaṃ
+paścimavāhinaṃ paścimena paścimamārgeṇa suṣumnāmārgeṇa vahatīti paścimavāhī taṃ tādṛśaṃ karoti.
+jaṭharānalasya jaṭhare yo 'nalo 'gnis tasyodayaṃ vṛddhiṃ kuryāt. udare madhyapradeśe kārśyaṃ
+kṛśatvaṃ kuryāt. arogatām ārogyaṃ cakārān nāḍīvalanādisāmyaṃ kuryāt.
+%\end{vsid}
+
+%\begin{vsid}{hp01_030}
+\startsloka
+{\bf dharām avaṣṭabhya karadvayena
+tatkūrparasthāpitanābhipārśvaḥ
+uccāsano daṇḍavad utthitaḥ khe\var{khe \lem syāt \Msix}
+māyūram etat pravadanti pīṭham (1.30)}
+\stopsloka
+
+atha māyūrāsanam āha dharām iti. karadvayena karayor dvayaṃ yugmaṃ tena dharām bhūmim avaṣṭabhya
+avalambya prasāritāṅgulī bhūmisaṃlagnatalau saṃnnihitau karau kṛtvety arthaḥ. tasya karadvayasya
+kūrparayoḥ bhujamadhyasandhibhāgayoḥ sthāpite dhṛte nābheḥ pārśve pārśvabhāgau yena sa\var{sa \lem
+  om \Msix (better!)} uccāsana uccam unnatam āsanaṃ yasyaitādṛśaḥ. khe śūnye daṇḍavad daṇḍena tulyam
+utthita ūrdhvaṃ sthito yatra bhavati tan māyūram mayūrasyedaṃ tatsambandhitvāt tannāmakaṃ
+pravadanti. yogina iti śeṣaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_031}
+\startsloka
+{\bf harati sakalarogān āśu gulmodarādīn
+abhibhavati ca doṣān āsanaṃ śrīmayūram
+bahu kadaśanabhuktaṃ bhasmakuryād aśeṣaṃ
+janayati ca jaṭharāgniṃ jārayet kālakūṭam (1.31)}
+\stopsloka
+
+mayūrāsanaguṇān āha haratīti. gulmo rogaviśeṣaḥ udaraṃ jalodaraṃ te ādinī yeṣāṃ plīhādīnāṃ te tathā
+tān sakalarogān sakalā ye rogās tān āśu jhaṭiti harati nāśayati. śrīmayūram āsanam iti sarvatra
+sambadhyate. doṣān vātapittakaphān ālasyādīṃś cābhibhavati tiraskaroti\var{tiraskaroti \lem
+  tiraskāroti \Msix}. bahv atiśayitaṃ kadaśanaṃ kadannaṃ yad bhuktaṃ tad aśeṣaṃ\var{tad aśeṣaṃ \var
+  om \EdLo} samastaṃ bhasmakuryāt pācayed ity arthaḥ. jaṭharāgniṃ jaṭharānalaṃ janayati
+prādurbhāvayati. kālakūṭam viṣaṃ kālakūṭaśabdo 'tra kālakūṭavad apakārakānnaparaḥ taṃ jārayet
+jīrṇaṃ kuryāt pācayed ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_032}
+\startsloka
+{\bf uttānaṃ śavavad bhūmau śayanaṃ tac chavāsanam
+śavāsanaṃ śrāntiharaṃ cittaviśrāntikārakam (1.32)}
+\stopsloka
+
+śavāsanam āhārdhena uttānam iti. śavena mṛtaśarīreṇa tulyaṃ śavavad uttānaṃ
+bhūmisaṃlagnaṃ\var{saṃlagnaṃ \lem saṃlagna \Msix} pṛṣṭhaṃ yathā syāt tathā śayanaṃ nidrāyām iva
+saṃniveśo yat tac chavāsanaṃ. śavākhyam āsanam.  śavāsanaprayojanam āha uttarārdhena. śavāsanaṃ
+śrāntiharaṃ śrāntiṃ haṭhābhyāsaśramaṃ haratīti śrāntiharaṃ. cittasya viśrāntir viśrāmas tasyāḥ
+kārakam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_033}
+\startsloka
+{\bf caturaśīty āsanāni śivena kathitāni ca
+tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham (1.33)}
+\stopsloka
+
+vakṣyamāṇāsanacatuṣṭayasya śreṣṭhatvaṃ vadann āha caturaśītīti. śiveneśvareṇa
+caturadhikāśītisaṅkhyākāny āsanāni kathitāni. cakārāc caturaśītilakṣāṇi ca. tad uktaṃ
+gorakṣanāthena
+
+\startsloka
+āsanāni ca tāvanti yāvantyo\var{yāvantyo \lem yāvatyo \Msix} jīvajātayaḥ
+eteṣām akhilān bhedān\var{bhedān \lem bhogān \Msix} vijānāti maheśvaraḥ
+caturāśītilakṣāṇi ekaikaṃ samudāhṛtam
+tataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtam\comment{1.89} iti.
+\stopsloka
+
+tebhyaḥ śivoktacaturaśītilakṣāsanānāṃ madhye praśastāni yāni caturaśīty āsanāni tebhya ādāya
+gṛhītvā sārabhūtaṃ śreṣṭhabhūtaṃ catuṣkam ahaṃ bravīmīty anvayaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_035}
+\startsloka
+{\bf siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam
+śreṣṭhaṃ tatrāpi ca sukhe tiṣṭhet siddhāsane sadā (1.34)}
+\stopsloka
+
+tad eva catuṣkaṃ nāmnā nirdiśati siddham iti. siddhaṃ siddhāsanaṃ padmaṃ padmāsanaṃ siṃhaṃ
+siṃhāsanaṃ bhadraṃ bhadrāsanaṃ ca iti catuṣṭayam śreṣṭham atiśayena praśasyam. tatrāpi catuṣṭaye
+sukhe sukhakare siddhāsane sadā tiṣṭhet\comment{{\em siddhāsane sadā tiṣṭhet} echoes HP
+  1.39b} etena siddhāsanaṃ catuṣṭaye 'py utkṛṣṭam iti sūcitam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_035}
+\startsloka
+{\bf yonisthānakam\var{yoni \lem tatra siddhāsaṃ yoni \Msix} aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyaset
+meḍhre pādam athaikam eva hṛdaye kṛtvā hanuṃ susthiram
+sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyed bhruvor antaraṃ
+hy etan mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate
+\hfill (1.35)}
+\stopsloka
+
+āsanacatuṣṭaye 'py utkṛṣṭatvāt prathamaṃ siddhāsanam āha\var{āha \lem āha tatreti \Msix}
+yonisthānakam iti. yonisthānam eva yonisthānakam svārthe kapratyayaḥ. gudopasthayor
+madhyapradeśo\var{madhyapradeśo \EdLo \Msix \lem madhyamapradeśe padaṃ \EdMu} yonisthānaṃ
+tat. aṅghrir vāmaś caraṇas tasya mūlena pārṣṇibhāgena ghaṭitaṃ saṃlagnaṃ kṛtvā. athānantaram ekam
+pādaṃ dakṣiṇaṃ pādaṃ meḍhre indriyasyoparibhāge\var{meḍhre indriyasyoparibhāge \Wthree \Tue \Msix
+  \lem meḍhrendriyasyoparibhāge Ed Mu \EdLo (no variants reported)} dṛḍhaṃ yathā syāt tathā
+vinyaset. hṛdaye hṛdayasamīpe hanuṃ cibukaṃ susthiram samyaksthiraṃ kṛtvā hanuhṛdayayoś
+caturaṅgulam antaraṃ yathā bhavati tathā kṛtveti rahasyam. saṃyamitāni viṣayebhyaḥ parāvṛttāni
+indriyāṇi yena sa tathā. acalā yā dṛk dṛṣṭis tayā bhruvor antaraṃ madhyaṃ paśyet. hi prasiddhaṃ
+mokṣasya yat kapāṭaṃ pratibandhakaṃ tasya bhedaṃ nāśaṃ janayatīti tādṛśam. siddhānāṃ yoginām. āste
+'tra āsyate 'neneti vā āsanaṃ siddhāsananāmakam idaṃ bhaved ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_036}
+\startsloka
+{\bf matāntare tu
+
+meḍhrād upari vinyasya savyaṃ gulphaṃ tathopari
+gulphāntaraṃ ca nikṣipya siddhāsanam idaṃ bhavet (1.36)}
+\stopsloka
+
+matsyendrasammataṃ siddhāsanam uktvānyasaṃmataṃ vaktum āha matāntare tv iti. tad eva darśayati
+meḍhrād iti. meḍhrād upasthād upary ūrdhvabhāge savyaṃ\var{savyaṃ \EdLo \Msix \lem savye \Tue}
+vāmagulphaṃ\var{vāmagulphaṃ \Wthree \Tue \EdMu \lem vāmaṃ gulphaṃ \EdLo \Msix} vinyasya tathā savyavad upari
+savyapādasyopari na tu savyagulphasya. gulphāntaraṃ dakṣiṇagulphaṃ ca nikṣipya vased iti
+śeṣaḥ. idaṃ siddhāsanam matāntarābhimataṃ bhaved ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_037}
+\startsloka
+{\bf pūrvoktam eva matsyendramatam
+
+etat siddhāsanaṃ prāhur anye vajrāsanaṃ viduḥ
+muktāsanaṃ vadanty eke prāhur guptāsanaṃ pare (1.37)}
+\stopsloka
+
+tatra prathamaṃ mahāsiddhasaṃmatam ity āha pūrvoktam iti. asyaiva matabhedān nāmabhedān āha etad
+iti. etat pūrvoktaṃ siddhāsanaṃ siddhāsananāmakaṃ prāhuḥ. kecid ity adhyāhāraḥ. anye vajrāsanaṃ
+vajrāsanasaṃjñakaṃ viduḥ jānanti. eke muktāsanaṃ muktāsanābhidhaṃ vadanti. pare guptāsanaṃ
+guptāsanākhyaṃ prāhuḥ. atrāsanābhijñāḥ. yatra vāmapādapārṣṇiṃ yonisthāne\var{yonisthāne \lem
+  mūlasthāne \Msix} niyojya dakṣiṇapādapārṣṇir meṇḍhrād upari sthāpyate tat siddhāsanam. yatra
+dakṣiṇapādapārṣṇiṃ\var{dakṣiṇapādapārṣṇiṃ \lem vāmapādapārṣṇiṃ \Msix} yonisthāne niyojya
+vāmapādapārṣṇir\var{vāmapādapārṣṇir \lem dakṣiṇapādapārṣṇir \Msix} meḍhrād upari sthāpyate tad
+vajrāsanam\var{vajrāsanam \Wthree \EdMu \Msix \lem siddhāsanam \Tue \EdLo}.  yatra tu
+dakṣiṇasavyapārṣṇidvayam uparyadhobhāgena saṃyojya yonisthānena\var{yonisthānena \lem yonisthāne
+  \Msix} saṃyojyate tan muktāsanam. yatra ca pūrvavat saṃyuktaṃ pārṣṇidvayaṃ meḍhrād upari nidhīyate
+tad guptāsanam iti.
+%\end{vsid}
+
+%\begin{vsid}{hp01_038}
+\startsloka
+{\bf yameṣv iva mitāhāram ahiṃsāṃ niyameṣv iva
+mukhyaṃ sarvāsaneṣv ekaṃ siddhāḥ siddhāsanaṃ viduḥ (1.38)}
+\stopsloka
+
+atha saptabhiḥ ślokaiḥ siddhāsanaṃ praśaṃsati yameṣvityādibhiḥ. yameṣu mitāhāram iva. mitāhāro
+vakṣyamāṇaḥ susnigdhamadhurāhāra\comment{HP 1.58} ityādinā. niyameṣu ahiṃsām iva. sarvāṇi yāny
+āsanāni teṣu siddhāḥ ekaṃ siddhāsanaṃ mukhyaṃ vidur iti sambandhaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_039}
+\startsloka
+{\bf caturaśītipīṭheṣu siddham eva sadābhyaset
+dvāsaptatisahasrāṇāṃ nāḍīnāṃ malaśodhanam (1.39)}
+\stopsloka
+
+caturaśītīti. caturadhikāśītisaṅkhyākāni yāni pīṭhāni teṣu siddham eva siddhāsanam eva sadā
+sarvadābhyaset. siddhāsanasya sadābhyāse hetugarbhaviśeṣaṇaṃ. dvāsaptatisahasrāṇāṃ nāḍīnāṃ
+malaśodhanam malasya śodhanaṃ\var{malasya śodhanaṃ \lem om \Msix} śodhakaṃ yataḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_040}
+\startsloka
+{\bf ātmadhyāyī mitāhārī yāvaddvādaśavatsaram
+sadā siddhāsanābhyāsād yogī niṣpattim āpnuyāt (1.40)}
+\stopsloka
+
+ātmadhyāyīti. ātmānaṃ dhyāyatīty ātmadhyāyī. mita āhāro\var{mita āhāro \lem mitāhāro \Wthree \Msix}
+'syāstīti mitāhārī. yāvanto dvādaśa vatsarāḥ\var{yāvanto dvādaśa vatsarāḥ \lem om \EdLo}
+yāvaddvādaśavatsaram. yāvad avadhāraṇa\comment{Aṣṭādhyāyī 2.1.8.} ity avyayībhāvaḥ samāsaḥ.
+dvādaśavatsaraparyantam ity arthaḥ. sadā sarvadā siddhāsanasyābhyāsād yogī yogābhyāsī niṣpattim
+yogasiddhim āpnuyāt prāpnuyāt. yogāṅgāntarābhyāsam antareṇa siddhāsanābhyāsamātreṇa siddhiṃ
+prāpnuyād ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_041}
+\startsloka
+{\bf kim anyair bahubhiḥ pīṭhaiḥ siddhe siddhāsane sati
+prāṇānile sāvadhāne baddhe kevalakumbhake
+utpadyate nirāyāsāt svayam evonmanī kalā (1.41)}
+\stopsloka
+
+kim anyair iti. siddhāsane siddhe saty anyair bahubhiḥ pīṭhair āsanaiḥ kim. na kim apīty arthaḥ.
+sāvadhāne prāṇānile prāṇavāyau kevalakumbhake baddhe sati unmanī unmanyavasthā. sā
+kalevāhlādakatvāc candralekheva nirāyāsād anāyāsāt svayam evotpadyata udeti.
+%\end{vsid}
+
+%\begin{vsid}{hp01_042}
+\startsloka
+{\bf tathaikāsminn eva dṛḍhe baddhe siddhāsane sati
+bandhatrayam anāyāsāt svayam evopajāyate (1.42)}
+\stopsloka
+
+tatheti. tathoktaprakāreṇaikasminn eva siddhāsane dṛḍhe baddhe sati bandhatrayam
+mūlabandhoḍḍīyānabandhajālandharabandharūpam anāyāsāt
+
+\startsloka
+pārṣṇibhāgena sampīḍya yonim ākuñcayed gudam\comment{HP 3.6}
+\stopsloka
+
+ityādivakṣyamāṇaprakāreṇa\var{prakāreṇa \Wthree \Msix \EdLo \lem om \Tue \EdMu} mūlabandhādiṣu ya
+āyāsas taṃ vinaiva svayam evopajāyate svata evotpadyata ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_043}
+\startsloka
+{\bf nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ
+na khecarīsamā mudrā na nādasadṛśo layaḥ (1.43)}
+\stopsloka
+
+nāsanam iti. siddhena siddhāsanena sadṛśam āsanaṃ na astīti śeṣaḥ. kevalena kevalakumbhakena
+upamīyata iti kevalopamaḥ kumbhaḥ kumbhako nāsti. khecarīmudrāsamā mudrā nāsti. nādasadṛśo\var{nāda
+\lem oṃnāda \Wthree} layo layahetur nāsti.
+%\end{vsid}
+
+%\begin{vsid}{hp01_044}
+\startsloka
+{\bf atha padmāsanam
+
+vāmorūpari dakṣiṇaṃ ca caraṇaṃ saṃsthāpya vāmaṃ tathā
+dakṣorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham
+aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed
+etad vyādhivināśakāri yamināṃ padmāsanaṃ procyate (1.44)}
+\stopsloka
+
+padmāsanaṃ vaktum upakramate atheti. padmāsanam āha vāmorūparīti. vāmo ya ūrus tasyopari dakṣiṇaṃm
+caraṇaṃ. cakāraḥ pādapūraṇe. saṃsthāpya samyag uttānaṃ sthāpayitvā vāmaṃ savyaṃ caraṇaṃ tathā
+dakṣiṇacaraṇavad dakṣo dakṣiṇo ya ūrus tasyopari saṃsthāpya paścimena bhāgena\var{paścimena bhāgena
+  \lem paścimena paścimabhāgena \Msix} pṛṣṭhabhāgeneti.  vidhir vidhānaṃ karayor ity arthāt. tena
+karābhyāṃ hastābhyāṃ dṛḍham yathā syāt tathā aṅguṣṭhau pādāṅguṣṭhau dhṛtvā gṛhītvā. dakṣiṇaṃ karaṃ
+pṛṣṭhataḥ kṛtvā vāmorusthitadakṣiṇacaraṇāṅguṣṭhaṃ gṛhītvā vāmakaraṃ pṛṣṭhataḥ kṛtvā
+dakṣiṇorusthitavāmacaraṇāṅguṣṭhaṃ gṛhītvety arthaḥ. hṛdaye hṛdayasamīpe. sāmīpikādhāre
+saptamī. cibukaṃ hanuṃ nidhāya urasaś caturāṅgulāntare cibukaṃ nidhāyeti rahasyam. nāsāgram
+nāsikāgram ālokayet paśyet. yatraitad yamināṃ yogināṃ\var{yogināṃ \lem om \Msix} vyādher vināśaṃ
+karotīti vyādhivināśakāri padmāsanam etannāmakaṃ procyate siddhair iti śeṣaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_45,46,46}        % JH: Max, geht das ?
+\startsloka
+{\bf  matāntare
+uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ
+ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau (1.45)
+nāsāgre vinyased rājadantamūle tu jihvayā
+uttambhya cibukaṃ vakṣasy utthāpya pavanaṃ śanaiḥ (1.46)
+idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam
+durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi (1.47)}
+\stopsloka
+
+matsyendranāthābhimataṃ padmāsanam āha uttānāv iti. uttānau caraṇau ūrusaṃlagnapṛṣṭhabhāgau caraṇau
+pādau prayatnataḥ prakṛṣṭād yatnād ūrusaṃsthāv ūrvoḥ samyak tiṣṭhata\var{tiṣṭhata \lem tiṣṭheta
+  \Msix} ity ūrusaṃsthau tādṛśau kṛtvā. ūrvor madhye ūrumadhye. tathā cārthe. pāṇī karāv uttānau
+kṛtvā.  ūrusaṃsthottānapādobhayapārṣṇisaṃlagnapṛṣṭhaṃ savyaṃ pāṇim uttānaṃ kṛtvā tadupari dakṣiṇaṃ
+pāṇiṃ cottānaṃ kṛtvety arthaḥ. tatas tadanantaraṃ. dṛśau dṛṣṭī. nāsāgra iti nāsāgre nāsikāgre
+vinyased viśeṣeṇa niścalatayā nyased ity arthaḥ. rājadantānāṃ daṃṣṭrāṇāṃ savyadakṣiṇabhāge
+sthitānāṃ mūle ubhe mūlasthāne jihvayā uttambhya ūrdhvaṃ stambhayitvā. gurumukhād avagantavyo 'yaṃ
+jihvābandhaḥ. cibukaṃ vakṣasi nidhāyeti śeṣaḥ. śanair mandaṃ mandaṃ pavanaṃ vāyum utthāpya. anena
+mūlabandhaḥ proktaḥ.  mūlabandho 'pi gurumukhād evāvagantavyaḥ. vastutas tu jihvābandhenaivāyaṃ
+caritārtha iti haṭharahasyavidaḥ. idam iti\var{idam iti \lem om \Msix}. evaṃ yatrāsyate tad idaṃ
+padmāsanaṃ padmāsanābhidhānaṃ proktam. āsanajñair iti śeṣaḥ. kīdṛśam sarveṣāṃ vyādhīnāṃ viśeṣeṇa
+nāśanaṃ yena kenāpi bhāgyahīnena durlabham. dhīmatā bhuvi bhūmau labhyate prāpyate.
+%\end{vsid}
+
+%\begin{vsid}{hp01_048}
+\startsloka
+paścād uktam eva matsyendramatam
+{\bf kṛtvā sampuṭitau karau dṛḍhataraṃ baddhvā tu padmāsanaṃ
+gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyāyaṃś ca tac cetasi
+vāraṃ vāram apānam ūrdhvam anilaṃ protsārayan pūritaṃ
+nyañcan\var{nyañcan \lem muñcan \Msix} prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ (1.48)}
+\stopsloka
+
+etac ca mahāyogisaṃmatam ity āha paścād iti. anyad api padmāsane kṛtyaviśeṣam āha kṛtveti.
+saṃpuṭitau saṃpuṭīkṛtau karau utsaṅgasthāv iti śeṣaḥ. dṛḍhataram atiśayena dṛḍhaṃ susthiraṃ
+padmāsanaṃ baddhvā kṛtvety arthaḥ. cibukaṃ hanuṃ gāḍhaṃ dṛḍhaṃ yathā syāt tathā vakṣasi
+vakṣaḥsamīpe saṃnnidhāya saṃnnihitaṃ kṛtvā caturaṅgulāntareṇeti yogisampradāyāj jñeyam.
+jālandharabandhaṃ kṛtvety arthaḥ. tat svasyeṣṭa\var{svasyeṣṭa \lem svasva \Msix (better)}devatārūpaṃ
+brahma vā. oṃ tat sad iti nirdeśo brahmaṇas trividhaḥ smṛtaḥ\comment{Bhagavadgītā 17.23.} iti
+bhagavadukteḥ. cetasi citte dhyāyan cintayan. apānam anilam apānavāyum ūrdhvaṃ protsārayan
+mūlabandhaṃ kṛtvā suṣumnāmārgeṇa prāṇam\var{prāṇam \lem om \Msix} ūrdhvaṃ nayan pūritaṃ pūrakeṇa
+antar dhāritaṃ prāṇaṃ nyañcan nīcair adhaḥ añcan gamayan.  antarbhāvitaṇyartho
+'ñcatiḥ. prāṇāpānayor aikyaṃ kṛtvety arthaḥ. naraḥ pumān atulaṃ bodhaṃ nirupamajñānaṃ
+śaktiprabhāvāc chaktir ādhāraśaktiḥ kuṇḍalinī tasyāḥ prabhāvāt sāmarthyād upaiti
+prāpnoti. prāṇāpānayor aikye kuṇḍalinībodho bhavati. kuṇḍalinībodhe suṣumnāmārgeṇa prāṇo
+brahmarandhraṃ gacchati. tatra gate cittasthairyaṃ bhavati. cittasthairye 'saṃśayād\var{'saṃśayād
+  \lem saṃyamād \Msix} ātmasākṣātkāro bhavatīty arthaḥ. % (yo.ta. 1.52, yo.cū.u. 40)
+%\end{vsid}
+
+%\begin{vsid}{hp01_049}
+\startsloka
+{\bf padmāsane sthito yogī nāḍīdvāreṇa pūritam
+mārutaṃ dhārayed yas tu sa mukto nātra saṃśayaḥ (1.49)}
+\stopsloka
+
+padmāsana iti. padmāsane sthito yo yogī yogābhyāsī pūritaṃ pūrakeṇāntarnītaṃ mārutaṃ vāyuṃ
+suṣumnāmārgeṇa mūrdhānaṃ nītveti śeṣaḥ. yaḥ dhārayet sthirīkuryāt sa muktaḥ. atra saṃśayo
+nāstīty anvayaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_050}
+\startsloka
+{\bf atha siṃhāsanam
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet
+dakṣiṇe savyagulphaṃ tu dakṣagulphaṃ tu savyake (1.50)}
+\stopsloka
+
+siṃhāsanam āha gulphau ceti. vṛṣaṇasyādhaḥ adhobhāge sīvanyāḥ pārśvayoḥ sīvanyāḥ ubhayabhāgayoḥ
+kṣipet prerayet sthāpayed iti yāvat. gulphasthāpanaprakāram evāha dakṣiṇa iti. sīvanyāḥ dakṣiṇe
+bhāge savyagulphaṃ sthāpayet. savyake sīvanyāḥ savyabhāge dakṣiṇagulphaṃ sthāpayet.
+%\end{vsid}
+
+%\begin{vsid}{hp01_051}
+\startsloka
+{\bf hastau tu jānvoḥ saṃsthāpya svāṅgulīḥ samprasārya ca
+vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ (1.51)}   %  read nāsāgre !!! ?????
+\stopsloka
+
+hastāv iti. jānvoḥ upari hastau tu saṃsthāpya samyag jānusaṃlagnatalau yathā syātāṃ tathā
+sthāpayitvā. svāṅgulīḥ hastāṅgulīḥ samprasārya samyak prasārayitvā\var{prasārayitvā \Msix \Wthree
+  \EdMu \lem prasārya \EdLo} vyāttavaktraḥ samprasāritalalajjihvamukhaḥ susamāhitaḥ ekāgracittaḥ
+nāsāgraṃ nāsikāgraṃ yasmin nirīkṣeta.   % yasmin means locative, i.e. nāsāgre??? 
+%\end{vsid}
+
+%\begin{vsid}{hp01_052}
+\startsloka
+{\bf siṃhāsanaṃ bhaved etat pūjitaṃ yogipuṅgavaiḥ
+bandhatritayasandhānaṃ kurute cāsanottamam (1.52)}
+\stopsloka
+
+etat siṃhāsanaṃ bhavet. kīdṛśam yogipuṅgavaiḥ yogiśreṣṭhaiḥ pūjitaṃ prastutam āsaneṣūttamaṃ
+siṃhāsanaṃ bandhānāṃ mūlabandhādīnāṃ tritayaṃ trayaṃ tasya sandhānaṃ tasya sannidhānaṃ kurute.
+%\end{vsid}
+
+%\begin{vsid}{hp01_053}
+\startsloka
+{\bf atha bhadrāsanam
+
+gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet
+savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe (1.53)}
+\stopsloka
+
+bhadrāsanam āha gulphāv iti. vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ sīvanyā ubhayataḥ gulphau
+pādagranthī kṣipet. kṣepaṇaprakāram evāha savyagulpham iti. savye sīvanyāḥ pārśve savyagulphaṃ
+kṣipet. tathā pādapūraṇe. dakṣagulphaṃ tu dakṣiṇe sīvanyāḥ pārśve kṣipet (53)
+%\end{vsid}
+
+%\begin{vsid}{hp01_054}
+\startsloka
+{\bf pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam
+bhadrāsanaṃ bhaved etat sarvavyādhivināśanam
+gorakṣāsanam ity āhur idaṃ vai siddhayoginaḥ (1.54)}
+\stopsloka
+
+pārśvapādau ca pārśvasamīpagatau pādau pāṇibhyāṃ bhujābhyāṃ dṛḍhaṃ baddhvā
+parasparasaṃlagnāṅgulibhyām udarasaṃlagnatalābhyāṃ pāṇibhyāṃ baddhvety arthaḥ. etad bhadrāsanaṃ
+bhavet. kīdṛśam sarveṣāṃ vyādhīnāṃ viśeṣeṇa nāśanam.\var{siddhāś \Msix \lem siddhāś gorakṣeti
+  \EdMu} siddhāś ca te yoginaś ca siddhayoginaḥ. idaṃ bhadrāsanaṃ gorakṣāsanam ity āhuḥ.
+gorakṣeṇa prāyaśo 'bhyastatvād gorakṣāsanam iti vadanti.
+%\end{vsid}
+
+%\begin{vsid}{hp01_055}
+\startsloka
+{\bf 
+evam āsanabandheṣu yogīndro vigataśramaḥ
+abhyasen nāḍikāśuddhiṃ mudrādipavanakriyām (1.55)}
+\stopsloka
+
+āsanāny uktvā\var{uktvā \Msix \Wthree \lem uktāni \EdMu \EdLo}. teṣu yat kartavyaṃ tad āha evam
+iti. evam ukteṣv āsanānāṃ bandheṣu bandhanaprakāreṣu vigataḥ śramo yasya sa vigataśramaḥ āsanānāṃ
+bandheṣu śramarahitaḥ. yoginām indro yogīndraḥ. nāḍikānāṃ nāḍīnāṃ śuddhim. prāṇaṃ ced iḍayā piben
+niyamitam\comment{HP 2.10} iti vakṣyamāṇarūpā mudrā ādir yasyāḥ sūryabhedādes tādṛśīm pavanasya
+prāṇavāyoḥ kriyāṃ prāṇāyāmarūpāṃ cābhyaset.
+%\end{vsid}
+
+%\begin{vsid}{hp01_056}
+\startsloka
+{\bf 
+āsanaṃ kumbhakaṃ citraṃ mudrākhyaṃ karaṇaṃ tathā
+atha nādānusandhānam abhyāsānukramo haṭhe (1.56)}
+\stopsloka
+
+atha haṭhābhyasanakramam āha āsanam iti. āsanam uktalakṣaṇaṃ citraṃ nānāvidhaṃ kumbhakaṃ
+sūryabhedanam ujjāyī\comment{HP 2.44} ityādivakṣyamāṇam. mudrā ity akhyā yasya tat mudrākhyam
+mahāmudrādirūpaṃ\var{rūpaṃ \Msix \lem rūpa \EdLo} karaṇaṃ haṭhasiddhau prakṛṣṭopakārakam. tathā
+cārthe. athaitattrayānuṣṭhānānantaraṃ nādasya anāhatadhvaner anusandhānam anucintanaṃ haṭhe
+haṭhayoge abhyāso abhyasanaṃ tasya anukramaḥ paurvāparyakramo 'yam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_057}
+\startsloka
+{\bf 
+brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ
+abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā (1.57)}
+\stopsloka
+
+haṭhasiddher avadhim āha brahmacāryīti. brahmacārī brahmacaryavān mitāhāro vakṣyamāṇaḥ so 'syāstīti
+mitāhārī tyāgī dānaśīlo viṣayaparityāgī vā yogaparāyaṇaḥ yogāṅgābhyasanaparaḥ. abdād varṣād ūrdhvaṃ
+siddhaḥ siddhahaṭho bhavet. atra ukte 'rthe\var{atra ukte 'rthe \Wthree \lem atra uktarthe \Msix}
+vicāraṇā syān na veti saṃśayaprayuktā na kāryā. etan niścitam evety arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_058}
+\startsloka
+{\bf 
+susnigdhamadhurāhāraś caturthāṃśavivarjitaḥ
+bhujyate śivasamprītyai mitāhāraḥ sa ucyate (1.58)}
+\stopsloka
+
+pūrvaśloke mitāhārīty uktam tatra yogināṃ kīdṛśo mitāhāra ity apekṣāyām āha susnigdheti.  susnigdho
+'tisnigdhaḥ sa cāsau madhuraś ca tādṛśa āhāraḥ caturthāṃśavivarjitaḥ caturthabhāgarahitaḥ.  tad
+uktam abhiyuktaiḥ
+
+\startsloka
+dvau bhāgau pūrayed annais toyenaikaṃ prapūrayet
+vāyoḥ sañcaraṇārthāya caturtham avaśeṣayet iti
+\stopsloka
+
+śivo jīva īśvaro vā. bhoktā devo maheśvaraḥ iti vacanāt. tasya samprītyai samyakprītyarthaṃ yo
+bhujyate sa mitāhāra ity ucyate.
+%\end{vsid}
+
+%\begin{vsid}{hp01_059}
+\startsloka
+{\bf 
+kaṭvāmlatīkṣṇalavaṇoṣṇaharītaśāka-
+sauvīratailatilasarṣapamadyamatsyān
+ājādimāṃsadadhitakrakulatthakola-
+piṇyākahiṅgulaśunādyam apathyam āhuḥ (1.59)}
+\stopsloka
+
+atha yoginām apathyam āha dvābhyāṃ kaṭv iti. kaṭu kāravellādi. amlaṃ ciñcāphalādi. tīkṣṇaṃ
+marīcādi. lavaṇaṃ prasiddham. uṣṇaṃ guḍādi. harītaśākaṃ patraśākaṃ. sauvīraṃ kāñjikaṃ. tailaṃ
+tilasarṣapādisnehaḥ. tilāḥ prasiddhāḥ. sarṣapāḥ siddhārthāḥ. madyaṃ surā. matsyo jhaṣaḥ. eṣām
+itaretaradvandvaḥ. etān apathyān āhuḥ. ajasyedam ājaṃ ājaṃ ādir yasya saukarādeḥ tad ājādi tac ca
+tanmāṃsaṃ cājādimāṃsam. dadhi dugdhapariṇāmaviśeṣaḥ. takraṃ gṛhītasāraṃ
+dadhi. kulatthāḥ\var{kulatthāḥ ? \lem kulitthāḥ \Msix} dvidalaviśeṣāḥ. kolaṃ kolyāḥ phalaṃ
+badaram. karkandhūr badarī koliḥ ity\var{koliḥ ity ? \lem kolīty \Msix}
+amaraḥ. piṇyākaṃ\var{piṇyākaṃ ? \lem pinyākaṃ \Msix} tilapiṇḍam. hiṅgu rāmaṭhaṃ laśunam. eṣām
+itaretaradvandvaḥ. etāny ādyāni yasya tat tathā. ādyaśabdena palāṇḍugṛñjanamādakadravyamāṣānnādikaṃ
+grāhyam. apathyam ahitam. yoginām\var{yoginām \lem yiginām \Msix} iti śeṣaḥ. āhur yogina ity
+adhyāhāraḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_060}
+\startsloka
+{\bf 
+bhojanam ahitaṃ vidyāt punar apy uṣṇīkṛtaṃ rūkṣam
+atilavaṇam amlayuktaṃ kadaśanaśākotkaṭaṃ varjyam (1.60)}
+\stopsloka
+
+bhojanam iti. pūrvaṃ pācitaṃ paścād agnisaṃyogenoṣṇīkṛtaṃ yad bhojanaṃ sūpaudanaroṭikādi rūkṣaṃ
+ghṛtādihīnam. atiśayitaṃ lavaṇaṃ yasmin tad atilavaṇam. yad vā lavaṇam atikrāntam atilavaṇaṃ
+cākūvā iti loke prasiddhaṃ śākaṃ yavakṣārādikaṃ ca. lavaṇasya sarvathā varjanīyatvād uttarapakṣaḥ
+sādhuḥ. tathā ca dattātreyaḥ
+
+\startsloka
+atha varjyāni vakṣyāmi yogavighnakarāṇi ca
+lavaṇaṃ sarṣapaṃ cāmlam uṣṇaṃ tīkṣṇaṃ ca rūkṣakam
+atīva bhojanaṃ tyājyam atinidrātibhāṣaṇam  iti.
+\stopsloka
+
+skandapurāṇe 'pi
+
+\startsloka
+tyajet kaṭvamlalavaṇaṃ kṣīrabhojī sadā bhavet iti
+\stopsloka
+
+amlayuktam amladravyeṇa yuktam. amladravyeṇa yuktam api tyājyaṃ kim uta sākṣād
+amlam.% (āmla? All sources checked have amla)
+atra tṛtīyapādaṃ palalaṃ vā tilapiṇḍam iti kecit paṭhanti. tasyāyam arthaḥ palalaṃ māṃsaṃ
+tilapiṇḍaṃ piṇyākaṃ\var{piṇyākaṃ \lem pinyākaṃ \Msix} kadaśanaṃ kadannaṃ yāvanālakodravādi śākaṃ
+vihitetaraśākamātram. utkaṭaṃ vidāhi miracī iti loke prasiddham miracā iti
+hindusthānabhāṣāyām. kadaśanādīnāṃ samāhāradvandvaḥ.  atilavaṇādikaṃ varjyaṃ varjanārham. duṣṭam
+iti pāṭhe duṣṭaṃ pūtiparyuṣitādi. ahitam iti yojanīyam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_061}
+\startsloka
+{\bf 
+vahnistrīpathisevānām ādau varjanam ācaret 
+tathā hi gorakṣavacanam
+varjayed durjanaprāntaṃ vahnistrīpathisevanam
+prātaḥsnānopavāsādi kāyakleśavidhiṃ tathā (1.61)}
+\stopsloka
+
+evaṃ yogināṃ sadā varjyāny uktvābhyāsakāle varjyāny āha ardhena vahnīti. vahniś ca strī ca panthāś
+ca teṣāṃ sevā vahnisevana\var{vahnisevana \EdLo \EdMu vahniseka \Msix \Wthree
+  Tue}strīsaṅgatīrthayātrādau\var{yātrādau \Msix Wai \lem yātrādi \Tue} pathigamanādirūpās tāsāṃ
+varjanam ādāv abhyāsakāle ācaret. siddhe 'bhyāse tu kadācit śīte vahnisevanaṃ gṛhasthasya ṛtau
+svabhāryāgamanaṃ tīrthayātrādau mārgagamanaṃ ca na niṣiddham ity ādipadena sūcyate.
+
+tatra pramāṇaṃ gorakṣavacanam avatārayati tathā hīti. tat paṭhati varjayed iti. durjanaprāntaṃ
+durjanasamīpavāsam. durjanaprītim iti kvacit pāṭhaḥ. vahnistrīpathāṃ\var{pathāṃ \Msix \Wthree \Tueac
+  \lem pathi \Tuepc}sevanaṃ vyākhyātam.  prātaḥsnānam\lem{prātaḥsnānam \Tue \lem prātaḥsnānaṃ ca
+  \Wthree \Msix} upavāsaś cādir yasya phalāhārādeḥ tac ca tayoḥ samāhāradvandvaḥ. prathamābhyāsinaḥ
+prātaḥsnāne śītavikārotpatteḥ\var{vikārotpatteḥ \Tue \Wthree \lem kārotpatteḥ \Msix}. upavāsādinā
+pittādyutpatteḥ. kāyakleśavidhiṃ kāyakleśakaraṃ vidhiṃ kriyāṃ bahusūryanamaskārādirūpāṃ
+bahubhārodvahanādirūpāṃ ca. tathā samuccaye. atra pratipadaṃ varjayed iti kriyāsambandhaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_062}
+\startsloka
+{\bf godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ\var{ṣāṣṭika \Wthree \Tue \Msix \EdMu \lem ṣaṣṭika \EdLo}
+kṣīrājyakhaṇḍanavanītasitāmadhūni
+śuṇṭhīpaṭolakaphalādikapañcaśākaṃ
+mudgādi divyam udakaṃ ca yamīndrapathyam (1.62)}
+\stopsloka
+
+atha yogipathyam āha godhūmetyādinā. godhūmāś ca śālayaś ca yavāś ca ṣāṣṭikāḥ ṣaṣṭhyā dinair ye
+pacyante taṇḍulaviśeṣās \var{taṇḍulaviśeṣās \lem taṇḍulaviśeṣās te \EdLo (against all mss.)}
+śobhanam annaṃ\var{śobhanam annaṃ \lem śobhanānnaṃ \Msix} pavitrānnaṃ śyāmākanīvārādi tac caiteṣāṃ
+samāhāradvandvaḥ. kṣīraṃ dugdham. ājyaṃ ghṛtaṃ. khaṇḍaḥ śarkarā. navanītaṃ mathitadadhisāraḥ. sitā
+tīvrapallī\var{tīvrapallī \Msix \EdLo \lem tīvrapadī \EdMu tīvraparṇī \Wthree}
+khaḍīsākara\var{khaḍīsākara \Msix \EdLo \lem khaṇḍaśarkareti \EdMu khaḍīśākara \Wthree khaṇḍaśākara
+  \Tuepc khaṇḍīśākara \Tueac} iti loke prasiddhā misarī iti
+hindusthānabhāṣāyām\var{hindusthānabhāṣāyām \lem hindusthānabhāṣāyām prasiddham \Wthree}. madhu
+kṣaudram eṣām itaretaradvandvaḥ. śuṇṭhī prasiddhā paṭolakaphalaṃ paravara iti bhāṣāyāṃ prasiddhaṃ
+śākaṃ.  tadādir yasya kośātakyādes\var{kośātakyādes \lem kauśāta\-kyādes \EdLo}
+tatpaṭolakaphalādikaṃ śeṣād vibhāṣā\comment{Aṣṭhādhyāyī 5.4.154} iti kapratyayaḥ. pañcānāṃ śākānāṃ
+samāhāraḥ pañcaśākam. tad uktaṃ vaidyake
+
+\startsloka
+sarvaśākam acākṣuṣyaṃ cākṣuṣyaṃ śākapañcakam
+jīvantīvāstumatsyākṣīmeghanādapunarnavāḥ iti.
+\stopsloka
+
+mudgā\var{mudgā \lem om \EdLo} dvidalaviśeṣā ādir yasya tan mudgādi. ādipadena āḍhakī grāhyā. divyaṃ
+nirdoṣam udakaṃ jalaṃ ca. yama eṣām astīti yaminaḥ teṣv indro iva śreṣṭho\var{iva śreṣṭho \Wthree \Tue
+  \lem devaśreṣṭho \EdMu \EdLo} yo yogīndras tasya pathyaṃ hitam.
+%\end{vsid}
+
+%\begin{vsid}{hp01_063}
+\startsloka
+{\bf puṣṭaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam
+mano'bhilaṣitaṃ yogyaṃ yogī bhojanam ācaret (1.63)}
+\stopsloka
+
+atha yogino bhojananiyamam āha puṣṭam iti. puṣṭaṃ dehapuṣṭikaram odanādi sumadhuraṃ
+śarkarādisahitaṃ snigdhaṃ saghṛtaṃ gavyaṃ godugdhaghṛtādiyuktaṃ gavyālābhe māhiṣaṃ dugdhādi
+grāhyam.  dhātuprapoṣaṇaṃ laḍḍukāpūpādi. mano'bhilaṣitaṃ puṣṭādiṣu yan manorucikaraṃ tad eva yoginā
+bhoktavyam. mano'bhilāṣitam api kim avihitaṃ bhoktavyam nety āha yogyam iti. vihitam evety
+arthaḥ. yogī bhojanaṃ pūrvoktaviśeṣaṇaviśiṣṭam ācaret kuryād ity arthaḥ. na tu saktubharjitānnādinā
+nirvāhaṃ kuryād iti bhāvaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_064}
+\startsloka
+{\bf yuvo vṛddho 'tivṛddho vā vyādhito durbalo 'pi vā
+abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ (1.64)}
+\stopsloka
+
+yogābhyāsino vayoviśeṣārogyādyapekṣā nāstīty āha yuveti. yuvā taruṇaḥ vṛddho vṛddhāvasthāṃ prāptaḥ
+ativṛddho 'tivārddhakaṃ gato vā. abhyāsād āsanakumbhakādīnām abhyasanāt siddhiṃ
+samādhitatphalarūpām āpnoti. abhyāsaprakāram eva vadan viśinaṣṭi sarvayogeṣv iti. sarveṣu yogeṣu
+yogāṅgeṣv atandrito 'nalasaḥ. yogāṅgābhyāsāt siddhim āpnotīty arthaḥ. jīvanasādhane kṛṣivāṇijyādau
+jīvanaśabdaprayogavat sākṣāt paramparayā vā yogasādhaneṣu yogāṅgeṣu\var{yogāṅgeṣu \lem \om \Msix}
+yogaśabdaprayogaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_055_1}
+\startsloka
+{\bf kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet
+na śāstrapāṭhamātreṇa yogasiddhiḥ prajāyate (1.65)}
+\stopsloka
+
+abhyāsād eva siddhir bhavatīti draḍhayann āha dvābhyāṃ kriyāyuktasyeti. kriyā yogāṅgānuṣṭhānarūpā
+tayā yuktasya siddhir yogasiddhiḥ syāt. akriyasya yogāṅgānuṣṭhānarahitasya kathaṃ bhavet. na
+katham apīty arthaḥ. nanu yogaśāstrādhyayanena yogasiddhiḥ syāt nety āha neti. śāstrasya
+yogaśāstrasya pāṭhamātreṇa kevalena pāṭhena yogasya siddhir na prajāyate naiva jāyata ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_055_2}
+\startsloka
+{\bf na veṣadhāraṇaṃ siddheḥ kāraṇaṃ na ca tatkathā
+kriyaiva kāraṇaṃ siddheḥ satyam etan na saṃśayaḥ (1.66)}
+\stopsloka
+
+neti. veṣasya kāṣāyavastrādeḥ dhāraṇaṃ siddher yogasiddheḥ kāraṇaṃ na. tasya yogasya kathā vārttā
+ca na siddheḥ kāraṇaṃ. kiṃ tarhi siddheḥ kāraṇam ity ata āha kriyaiveti. yogāṅgānuṣṭhānarūpā
+kriyaiva siddheḥ kāraṇam etat satyaṃ saṃśayo na atrety adhyāhāraḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp01_0??} %  not in critical edition 
+\startsloka
+{\bf pīṭhāni kumbhakāś citrā divyāni karaṇāni ca
+sarvāṇy api haṭhābhyāse rājayogaphalāvadhi (1.65)}
+\stopsloka
+
+atha\var{atha \Msix \Wthree \ \om \EdMu \EdLo} yogāṅgānuṣṭhānasyāvadhim āha pīṭhānīti. pīṭhāny āsanāni
+citrā aneka\-vidhāḥ kumbhakāḥ sūryabhedādayaḥ divyāny utkṛṣṭāni karaṇāni mahā\-mudrādīni
+haṭhasiddhau\var{haṭhasiddhau \EdMu \Wthree \Tue \lem om \EdLo} haṭhābhyāse haṭhasyābhyāse sarvāṇi
+pīṭhakumbhakakaraṇāni rājayogaphalāvadhi rājayoga eva phalaṃ tadavadhi tatparyantaṃ kartavyānīti
+śeṣaḥ.
+%\end{vsid}
+
+%%%%%%%%%%%%%      (2)
+
+%\begin{vsid}{hp02_001}
+\startsloka
+{\bf athāsane dṛḍhe yogī vaśī hitamitāśanaḥ
+gurūpadiṣṭamārgeṇa prāṇāyāmān samabhyaset (2.1)}
+\stopsloka
+
+athāsanopadeśānantaraṃ prāṇāyāmān vaktum upakramate atheti. atheti maṅgalārthaḥe. āsane dṛḍhe sati
+vaśī jitākṣaḥ hitaṃ pathyaṃ ca tan mitaṃ ca pūrvopadeśoktalakṣaṇaṃ tat tādṛśam aśanaṃ yasya sa
+hitamitāśanaḥ guruṇopadiṣṭo yo mārgaḥ prāṇāyāmābhyāsaprakāras tena prāṇāyāmān vakṣyamāṇān samyag
+utsāhasāhasa\var{sāhasa \Tuepc \EdMu\lem sāha \Msix sa \Tueac \EdLo}dhairyādibhir abhyaset. dṛḍhe
+sthire kukkuṭādivivarjite siddhāsanādāv iti vā yojanā.
+%\end{vsid}
+
+%\begin{vsid}{hp02_002}
+\startsloka
+{\bf cale vāte calaṃ cittaṃ niścale niścalaṃ bhavet
+yogī sthāṇutvam āpnoti tato vāyuṃ nirodhayet (2.2)}
+\stopsloka
+
+prayojanam anuddiśya na mando 'pi pravartate iti mahadukteḥ\var{REF?} prayojanābhāve pravṛttyabhāvāt
+prāṇāyāmaprayojanam āha cale vāte iti. vāte cale sati cittaṃ calaṃ bhavet. niścale vāte niścalaṃ
+bhavet cittam ity atrāpi sambadhyate. vāte citte ca niścale yogī sthāṇutvaṃ sthiratvaṃ
+dīrghajīvitvam iti yāvat. īśatvaṃ vāpnoti. tatas tasmād vāyuṃ prāṇaṃ nirodhayet kumbhayet.
+%\end{vsid}
+
+%\begin{vsid}{hp02_003}
+\startsloka
+{\bf yāvad vāyuḥ sthito dehe tāvaj jīvanam ucyate
+maraṇaṃ tasya niṣkrāntis tato vāyuṃ nirodhayet (2.3)}
+\stopsloka
+
+yāvad iti. dehe śarīre yāvatkālaṃ vāyuḥ prāṇaḥ sthitaḥ tāvatkālaparyantaṃ jīvanam ucyate lokaiḥ.
+dehaprāṇasaṃyogasyaiva jīvanapadārthatvāt. tasya prāṇasya niṣkrāntir dehād viyogaḥ maraṇam ucyate.
+tatas tasmād vāyuṃ nirodhayet.
+%\end{vsid}
+
+%\begin{vsid}{hp02_004}
+\startsloka
+{\bf malākulāsu nāḍīṣu māruto naiva madhyagaḥ
+kathaṃ syād unmanībhāvaḥ kāryasiddhiḥ kathaṃ bhavet (2.4)}
+\stopsloka
+
+malaśuddher haṭhasiddhijanakatvaṃ vyatirekeṇāha malākulāsv iti. nāḍīṣu malair ākulāsu vyāptāsu
+satīṣu\var{satīṣu \Tuepc \EdMu \lem satsu Ba2 \Msix \Wthree \Tueac \EdLo} mārutaḥ prāṇo madhyagaḥ
+suṣumnāmārgavāhī naiva syāt. api tu śuddhamalāsv eva madhyago bhavatīty arthaḥ. unmanībhāva unmanyā
+bhāvo bhavanaṃ kathaṃ syāt na katham apīty arthaḥ.  kāryasya kaivalyarūpasya siddhir niṣpattiḥ
+kathaṃ bhaven na kathañcid apīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_005}
+\startsloka
+{\bf śuddhim eti yadā sarvaṃ nāḍīcakraṃ malākulam
+tadaiva jāyate yogī prāṇasaṃgrahaṇe kṣamaḥ (2.5)}
+\stopsloka
+
+anvayenāpi malaśuddher haṭhasiddhihetutvam āha śuddhim etīti. yadā yasmin kāle malair ākulam
+vyāptaṃ sarvaṃ samastaṃ nāḍīnāṃ cakraṃ samūhaḥ śuddhim malarāhityam eti prāpnoti tadaiva tasminn
+eva kāle yogī yogābhyāsī prāṇasya samyag grahaṇe\var{samyag grahaṇe \Wthree \Msix \lem saṃgrahaṇe
+  \EdLo grahaṇe \EdMu Ba2} kṣamaḥ samartho jāyate.
+%\end{vsid}
+
+%\begin{vsid}{hp02_006}
+\startsloka
+{\bf prāṇāyāmaṃ tataḥ kuryān nityaṃ sāttvikayā dhiyā
+yathā suṣumṇānāḍīsthā malāḥ śuddhiṃ prayānti ca (2.6)}    %  mala in der suṣumṇā ??
+\stopsloka
+
+malaśuddhiḥ kathaṃ bhavatīty ākāṅkṣāyāṃ tacchodhakaṃ prāṇāyāmam āha prāṇāyāmam iti. yato
+malaśuddhiṃ vinā prāṇasaṅgrahaṇe kṣamo na bhavati tatas tasmād
+īśvarapraṇidhānotsāhasāhasādiprayatnābhibhūtavikṣepālasyādirājasatāmasadharmayā sāttvikayā
+prakāśa\-pra\-sādaśīlayā dhiyā buddhyā nityaṃ prāṇāyāmaṃ kuryāt. yathā yena prakāreṇa suṣumnṇānāḍyāṃ
+sthitā malāḥ śuddhim apagamaṃ prayānti naśyantīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_007}
+\startsloka
+{\bf baddhapadmāsano yogī prāṇaṃ candreṇa pūrayet
+dhārayitvā yathāśakti bhūyaḥ sūryeṇa recayet (2.7)}
+\stopsloka
+
+malaśodhakaprāṇāyāmaprakāram āha dvābhyāṃ baddhapadmāsana iti. baddhaṃ padmāsanaṃ yena tādṛśo yogī
+prāṇaṃ vāyuṃ candreṇa candranāḍyā iḍayā pūrayet. śaktim anatikramya yathāśakti dhārayitvā
+kumbhayitvā bhūyaḥ punaḥ sūryeṇa sūryanāḍyā piṅgalayā recayet. bāhyavāyoḥ prayatnaviśeṣād upādānaṃ
+pūrakaḥ. jālandharādibandhapūrvakaṃ prāṇanirodhaḥ kumbhakaḥ. kumbhitasya vāyoḥ prayatnaviśeṣād
+vamanaṃ recakaḥ. prāṇāyāmāṅgarecakapūrakayor\var{recakapūrakayor \Wthree \lem pūrakarecakayor \Msix}
+eveme lakṣaṇe iti. bhastrāval lohakārasya recapūrau sasambhramau\comment{HP 2.35} iti
+gauṇarecakapūrakayor nātivyāptiḥ tayor lakṣyatvābhāvāt.
+%\end{vsid}
+
+%\begin{vsid}{hp02_008}
+\startsloka
+{\bf prāṇaṃ sūryeṇa cākṛṣya pūrayed udaraṃ śanaiḥ
+vidhivat kumbhakaṃ kṛtvā punaś candreṇa recayet (2.8)}  % dhṛtvā \Msix
+\stopsloka
+
+prāṇam iti. sūryeṇa sūryanāḍyā piṅgalayā sūryeṇa prāṇam ākṛṣya gṛhītvā śanair mandaṃ mandam udaraṃ
+jaṭharaṃ pūrayet. vidhivatd bandhapūrvakaṃ kumbhakaṃ kṛtvā punar bhūyaś candreṇeḍayā recayet.
+%\end{vsid}
+
+%\begin{vsid}{hp02_009}
+\startsloka
+{\bf yena tyajet tena pītvā dhārayed atirodhataḥ
+recayec ca tato 'nyena śanair eva na vegataḥ (2.9)}
+\stopsloka
+
+ukte prāṇāyāme viśeṣam āha yeneti. yena candreṇa sūryeṇa vā tyajed recayet tena pītvā tenaiva
+pūrayitvā. atirodhato 'tiśayitena rodhena svedakampādijananaparyantena. sārvavibhaktikas tasil.
+yena pūrakas tato 'nyena śanair eva recayet na tu vegataḥ. vegād recane balahāniḥ syāt. yena
+pūrakaḥ kṛtas tena recako na kartavyaḥ. yena recakaḥ kṛtas tenaiva pūrakaḥ kartavya iti bhāvaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_010}
+\startsloka
+{\bf prāṇaṃ ced iḍayā piben niyamitaṃ bhūyo 'nyayā recayet 
+pītvā piṅgalayā samīraṇam atho baddhvā tyajed vāmayā
+sūryācandramasor anena vidhinābhyāsaṃ sadā tanvatāṃ     % wrong caesura !?
+śuddhā nāḍigaṇā bhavanti yamināṃ māsatrayād ūrdhvataḥ (2.10)}
+\stopsloka
+
+baddhapadmāsana ityādyuktam arthaṃ piṇḍīkṛtyānuvadan prāṇāyāma\-syāvāntara\-phalam āha prāṇam iti ced
+yadi iḍayā vāmanāḍyā prāṇaṃ pibet pūrayet tarhi niyamitaṃ kumbhitaṃ prāṇaṃ bhūyaḥ punar anyayā
+piṅgalayā recayet. piṅgalayā dakṣanāḍyā samīraṇaṃm vāyuṃ pītvā pūrayitvātho pūraṇānantaraṃ baddhvā
+kumbhayitvā vāmayā iḍayā tyajed recayet. sūryaś ca candramāś ca sūryācandramasau tayoḥ
+devatādvandve ca\comment{Aṣṭādhyāyī 7.3.21} ity ānaṅ. anenoktena vidhinā prakāreṇa sadā nityam
+abhyāsaṃ candreṇāpūrya kumbhayitvā sūryeṇa recayet sūryeṇāpūrya kumbhayitvā candreṇa recayed ity
+ākārakaṃ tanvatāṃ vistārayatāṃ yamināṃ yamavatāṃ nāḍīgaṇāḥ nāḍīsamūhāḥ māsatrayād ūrdhvato māsānāṃ
+trayaṃ māsatrayaṃ tasmād upari śuddhā malarahitā bhavanti.
+%\end{vsid}
+
+%\begin{vsid}{hp02_011}
+\startsloka
+{\bf prātar madhyandine sāyam ardharātre ca kumbhakān
+śanair aśītiparyantaṃ caturvāraṃ samabhyaset (2.11)}
+\stopsloka
+
+atha prāṇāyāmābhyāsakālaṃ tadavadhiṃ cāha prātar iti. prātar aruṇodayam ārabhya sūryodayād
+ghaṭikātrayaparyante prātaḥkāle madhyandine madhyāhne pañcadhā vibhaktasya dinasya madhyabhāge
+sāyaṃsandhyā trināḍīpramitārkam astād\var{pramitārkam astād \EdLo \Msix \Tue \lem pramitārkāstād
+  \EdMu} adhastād ūrdhvaṃ cety uktalakṣaṇe sandhyākāle rātrer ardhaṃ ardharātraṃ tasminn ardharātre
+rātrer madhye muhūrtadvaye ca śanair aśītiparyantam aśītisaṅkhyāvadhi caturvāraṃ
+vāracatuṣṭayaṃ. kālādhvanor atyantasaṃyoge\comment{Aṣṭādhyāyī 2.3.5.} iti dvitīyā.  caturṣu kāleṣv
+ekaikasmin kāle aśītiprāṇāyāmāḥ kāryāḥ. ardharātre kartum aśaktaś cet trisandhyaṃ kartavyā iti
+sampradāyaḥ. caturvāraṃ kṛtāś ced dine dine viṃśatyadhikaśatatrayaparimitāḥ prāṇāyāmā
+bhavanti. vāratrayaṃ kṛtāś cet catvāriṃśadadhikaśatadvayaparimitā bhavanti.
+%\end{vsid}
+
+%\begin{vsid}{hp02_012}
+\startsloka
+{\bf kanīyasi bhavet svedaḥ kampo bhavati madhyame
+uttame sthānam āpnoti tato vāyuṃ nibandhayet (2.12)}
+\stopsloka
+
+kaniṣṭhamadhyamottamānāṃ prāṇāyāmānāṃ krameṇa jñāpakaviśeṣān āha kanīyasīti. kanīyasi kaniṣṭhe
+prāṇāyāme svedaḥ prasvedo bhaved bhavati. svedānumeyaḥ kaniṣṭhaḥ. madhyame prāṇāyāme kampo
+'ṅgamejayo bhavati. kampānumeyo madhyamaḥ. uttame prāṇāyāme sthānaṃ brahmarandhram āpnoti.
+sthanaprāptyanumeya uttamaḥ. tatas tasmād vāyuṃ prāṇaṃ nibandhayen nitarāṃ bandhayet.
+kaniṣṭhādīnāṃ lakṣaṇam uktaṃ liṅgapurāṇe
+
+\startsloka
+prāṇāyāmasya mānaṃ tu mātrādvādaśakaṃ smṛtam
+nīco dvādaśamātras tu sakṛd udgdhāta īritaḥ
+madhyamas tu dvirudghātaś caturviṃśatimātrakaḥ
+mukhyas tu yas trirudghātaḥ ṣaṭtriṃśanmātra ucyate
+prasvedakampanotthānajanakaś ca yathākramam
+ānando jāyate cātra nidrā ghūrṇis\var{ghūrṇis \EdAd \lem dhūmas \EdMu \Tueac (?) ghūrmas \EdLo \Wthree \Msix \Tuepc } tathaiva ca
+romāñco dhvanisaṃvittir aṅgamoṭanakampanam
+bhramaṇasvedajalpādyaṃ saṃvinmūrcchā jayed yadā   % bhramaṇa \lem śravaṇa \Tue  ? ādyaṃ \Tue \lem ādya \Msix \Wthree
+tadottama iti proktaḥ prāṇāyāmaḥ suśobhana iti\comment{Liṅgapurāṇa 8.46–50}
+\stopsloka
+
+ghūrṇiś cittāndolanam. gorakṣo 'pi
+
+\startsloka
+adhame dvādaśa proktā madhyame dviguṇāḥ smṛtāḥ
+uttame triguṇā mātrāḥ prāṇāyāme dvijottamaiḥ\comment{2.5?}??
+\stopsloka
+
+udgdhātalakṣaṇaṃ tu
+
+\startsloka
+prāṇenotsāryamāṇena\var{sāryamāṇena \EdLo \lem sārpamāṇena \Msix} apānaḥ pīḍyate yadā
+gatvā cordhvaṃ nivarteta etad uddghātalakṣaṇam
+\stopsloka
+
+mātrām āha yājñavalkyaḥ
+
+\startsloka
+aṅguṣṭhāṅgulimokṣaṃ tris trir jānuparimārjanam
+tālatrayam api prājñā mātrāsaṃjñāṃ pracakṣate\comment{?}
+\stopsloka
+
+skandapurāṇe
+
+\startsloka
+ekaśvāsamayī mātrā prāṇāyāme\var{prāṇāyāme \Msix \Wthree \Tue \lem prāṇāyāmo \EdMu \EdLo} nigadyate 
+\stopsloka
+
+etad vyākhyātaṃ yogacintāmaṇau\comment{}
+
+nidrāvaśaṃ gatasya puṃso yāvatā kālenaikaḥ śvāso gacchaty āgacchati ca tāvān kālaḥ prāṇāyāmasya
+mātrety ucyata iti. ardhaśvāsādhikadvādaśaśvā\-sā\-vacchinnaḥ kālaḥ prāṇāyāmakālaḥ. ṣaḍbhiḥ śvāsair
+ekaṃ palaṃ bhavati. evaṃ ca sārdhaśvāsapaladvayātmakaḥ kālaḥ prāṇāyāmakālaḥ
+siddhaḥ. sārdha\-dvādaśamātrāmitaḥ prāṇāyāmo yaḥ sa evottamaḥ prāṇāyāma ity ucyate.  na ca
+pūrvodāhṛtaliṅgapurāṇagorakṣavākyavirodhas tatra dvādaśa\-mātrakasya prāṇāyāmasyādhamatvokter iti
+śaṅkanīyam.
+
+\startsloka
+jānuṃ pradakṣiṇīkuryān na drutaṃ na vilambitam
+pradadyāc choṭikāṃ yāvat tāvan mātreti gīyate
+\stopsloka
+
+iti skandapurāṇāt.
+
+\startsloka
+aṅguṣṭhāṅgulimokṣaṃ ca jānoś ca parimārjanam
+pradadyāc choṭikām ekāṃ mātrā saṅkhyāyate tadā
+\stopsloka
+
+iti dattātreyavacanāc ca.
+
+liṅgapurāṇagorakṣādivākyeṣv ekacchoṭikāvacchinnasya kālasya mātrātvena vivakṣitatvāt.
+yājñavalkyādivākyeṣu ekachoṭikāvacchinnasya kālasya mātrātvena vivakṣitatvāt
+triguṇasyādhamasyottamatvaṃ tatrāpy uktam ity avirodhaḥ. sarveṣu yogasādhaneṣu prāṇāyāmo mukhyaḥ
+tatsiddhau pratyāhārādīnāṃ siddheḥ. tadasiddhau pratyāhārādyasiddheś ca. vastutas tu prāṇāyāma eva
+pratyāhārādiśabdair nigadyate. tathā coktaṃ yogacintāmaṇau\comment{?}
+
+\startsloka
+prāṇāyāma evābhyāsakrameṇa vardhamānaḥ pratyāhāradhāraṇādhyāna\-samādhiśabdair ucyata iti.
+\stopsloka
+
+tad uktaṃ skandapurāṇe
+
+\startsloka
+prāṇāyāmadviṣaṭkena pratyāhāra udāhṛtaḥ
+pratyāhāradviṣaṭkena dhāraṇā parikīrtitā
+bhaved īśvarasaṅgatyai dhyānaṃ dvādaśadhāraṇam
+dhyānadvādaśakenaiva samādhir abhidhīyate
+yat samādhau paraṃ jyotir anantaṃ svaprakāśakam
+tasmin dṛṣṭe kriyākāṇḍaṃ yātāyātaṃ nivartate iti
+\stopsloka
+
+tathā
+
+\startsloka
+dhāraṇā pañcanāḍībhir dhyānaṃ syāt ṣaṣṭhināḍikam
+dinadvādaśakena syāt samādhiḥ prāṇasaṃyamād  iti ca
+\stopsloka
+
+gorakṣādibhir apy evam evoktam. atraivaṃ vyavasthā kiñcid ūnadvicatvāriṃśadvipalātmakaḥ
+kaniṣṭhaprāṇāyāmakālaḥ. ayam evaikacchoṭikāvacchinnasya kālasya mātrātvavivakṣayā dvādaśamātrakaḥ
+kālaḥ. kiñcid ūnacaturaśītivipalātmako madhyamaprāṇāyāmakālaḥ. ayam ekacchoṭikāvacchinnasya kālasya
+mātrātvavivakṣayā caturviṃśatimātrakaḥ kālaḥ. pañcaviṃśatyuttaraśatavipalātmaka uttamaḥ
+prāṇāyāmakālaḥ. ayam ekacchoṭikāvacchinnasya kālasya mātrātvavivakṣayā ṣaṭtriṃśanmātrakaḥ kālaḥ.
+choṭikātrayāvacchinnasya kālasya mātrātvavivakṣayā tu dvādaśamātraka eva. bandhapūrvakaṃ
+pañcaviṃśatyuttaraśatavipalaparyantaṃ yadā prāṇāyāmasthairyaṃ bhavati tadā prāṇo brahmarandhraṃ
+gacchati. brahmarandhraṃ gataḥ prāṇo yadā pañcaviṃśatipalaparyantaṃ tiṣṭhati tadā pratyāhāraḥ.
+yadā pañcaghaṭikāparyantaṃ tiṣṭhati tadā dhāraṇā. yadā ṣaṣṭhighaṭikāparyantaṃ tiṣṭhati tadā
+dhyānam. yadā dvādaśadinaparyantaṃ tiṣṭhati tadā samādhir bhavatīti sarvaṃ ramaṇīyam.
+%\end{vsid}
+
+%\begin{vsid}{hp02_013}
+\startsloka
+{\bf jalena śramajātena gātramardanam ācaret
+dṛḍhatā laghutā caiva tena gātrasya jāyate (2.13)}
+\stopsloka
+
+prāṇāyāmān abhyasataḥ svede jāte viśeṣam āha jaleneti. śramāt prāṇāyāmābhyāsaśramāj jātaṃ
+śramajātaṃ tena jalena prasvedena gātrasya śarīrasya mardanaṃ tailābhyaṅgavad ācaret kuryāt. tena
+mardanena gātrasya dṛḍhatā dārḍhyaṃ laghutā jāḍyābhāvo jāyate prādurbhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp02_014}
+\startsloka
+{\bf abhyāsakāle prathame śastaṃ kṣīrājyabhojanam
+tato 'bhyāse dṛḍhībhūte na tādṛṅniyamagrahaḥ (2.14)}
+\stopsloka
+
+atha prathamottarābhyāsayoḥ kṣīrādiniyamāniyamau\var{kṣīrādiniyamāniyamau \Msix \Wthree kṣīrādiniyamān
+  \EdLo \EdMu} āha abhyāsakāla iti. kṣīraṃ dugdham ājyaṃ ghṛtaṃ tadyuktaṃ bhojanaṃ
+kṣīrājyabhojanam. śākapārthivādivat samāsaḥ. kevale kumbhake siddhe 'bhyāso dṛḍho bhavati. spaṣṭam
+anyat.
+%\end{vsid}
+
+%\begin{vsid}{hp02_015}
+\startsloka
+{\bf yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śanaiḥ
+tathaiva sevito vāyur anyathā hanti sādhakam (2.15)}
+\stopsloka
+
+siṃhādivac chanair eva prāṇaṃ vaśayen na sahasety āha yatheti. yathā yena prakāreṇa siṃho mṛgendro
+gajo hastī\var{hastī \Msix \Wthree \lem vanahastī \EdLo} vyāghraḥ śārdūlaḥ śanaiḥ śanair eva vaśyaḥ
+svādhīno bhaven na sahasā tathaiva tenaiva prakāreṇa sevito 'bhyasto vāyuḥ prāṇo vaśyo
+bhavet. anyathā sahasā gṛhyamāṇaḥ sādhakam abhyāsinaṃ hanti siṃhādivat.
+%\end{vsid}
+
+%\begin{vsid}{hp02_016}
+\startsloka
+{\bf prāṇāyāmena yuktena sarvarogakṣayo bhavet
+ayuktābhyāsayogena sarvarogasamudbhavaḥ (2.16)}
+\stopsloka
+
+yuktāyuktayoḥ prāṇāyāmayoḥ phalam āha prāṇāyāmeneti. āhārādiyuktipūrvako
+jālandharādibandhayuktiviśiṣṭaḥ prāṇāyāmo yukta ity ucyate. tena sarvarogakṣayaḥ sarveṣāṃ rogāṇāṃ
+kṣayo nāśo bhavet. ayukta uktayuktirahito yo 'bhyāsas tadyuktena prāṇāyāmena sarvarogasamudbhavaḥ
+sarveṣāṃ rogāṇāṃ samyag udbhava utpattir bhavet.
+%\end{vsid}
+
+%\begin{vsid}{hp02_017}
+\startsloka
+{\bf hikkā śvāsaś ca kāsaś ca śiraḥkarṇākṣivedanāḥ
+bhavanti vividhā rogāḥ pavanasya prakopataḥ (2.17)}
+\stopsloka
+
+ayuktena prāṇāyāmena ke rogā bhavantīty apekṣāyām āha hikketi. hikkāśvāsakāsāḥ rogaviśeṣāḥ. śiraś ca
+karṇau cākṣiṇī ca śiraḥkarṇākṣi śiraḥkarṇākṣiṇi vedanāḥ śiraḥkarṇākṣivedanāḥ. vividhā nānāvidhā
+rogāḥ jvarādayaḥ. pavanasya vāyoḥ prakopato bhavanti.
+%\end{vsid}
+
+%\begin{vsid}{hp02_018}
+\startsloka
+{\bf yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ ca pūrayet
+yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt (2.18)}
+\stopsloka
+
+yataḥ pavanasya prakopato vividhā rogā bhavanty ataḥ \var{ataḥ \Wthree \Msix \Tue \EdAd \lem ataḥ kiṃ
+  kartavyam ata āha \EdMu \EdLo} yuktaṃ yuktam iti.  vāyuṃ prāṇaṃ yuktaṃ yuktaṃ tyajet. recakakāle
+śanaiḥ śanair eva recayen na vegataḥ ity arthaḥ.  yuktaṃ yuktaṃ ca na cālpaṃ\var{na cālpaṃ \EdLo
+  \lem nālpaṃ \Msix \Wthree} nādhikaṃ ca pūrayet. yuktaṃ yuktaṃ ca jālandharādibandhayuktaṃ badhnīyāt
+kumbhayet. evam abhyasan siddhiṃ haṭhasiddhim avāpnuyāt prāpnuyāt\var{āpnuyāt prāpnuyāt \Msix \Wthree
+  \Tue \lem avāpnuyāt \EdMu \EdLo}.
+%\end{vsid}
+
+% wwwww 2125
+
+%\begin{vsid}{hp02_019}
+\startsloka
+{\bf yadā tu nāḍīśuddhiḥ syāt tathā cihnāni bāhyataḥ
+kāyasya kṛśatā kāntis tadā jāyeta niścitam (2.19)}  % tathā  tadā   check Lesarten.
+\stopsloka
+
+yuktaṃ prāṇāyāmam abhyasato jāyamānāyā nāḍīśuddher lakṣaṇam āha dvābhyāṃ yadā tv iti. yadā tu
+yasmin kāle tu nāḍīnāṃ śuddhir malarāhitayaṃ syāt tadā bāhyato bāhyāni sārvavibhaktikas tasiḥ
+cihnāni lakṣaṇāni tathāśabdenāntarāṇy api cihnāni bhavantīty arthaḥ. tāny evāha kāyasyeti.
+kāyasya dehasya kṛśatā kārśyaṃ kāntiḥ surucir niścitam jāyeta.
+%\end{vsid}
+
+%\begin{vsid}{hp02_020}
+\startsloka
+{\bf yatheṣṭaṃ dhāraṇaṃ vāyor analasya pradīpanam
+nādābhivyaktir ārogyaṃ jāyate nāḍiśodhanāt (2.20)}
+\stopsloka
+
+yatheṣṭam iti. vāyoḥ prāṇasya yatheṣṭaṃ bahuvāraṃ dhāraṇaṃ kumbhakeṣu. analasya jaṭharāgneḥ
+pradīpanaṃ prakṛṣṭā dīptir nādasya dhvaner abhivyaktiḥ prākaṭyam ārogyam arogatā nāḍīśodhanāt
+nāḍīnāṃ śodhanāt malarāhityāj jāyate.
+%\end{vsid}
+
+%\begin{vsid}{hp02_021}
+\startsloka
+{\bf medaśleṣmādhikaḥ pūrvaṃ ṣaṭkarmāṇi samācaret
+anyas tu nācaret tāni doṣāṇāṃ samabhāvataḥ (2.21)}
+\stopsloka
+
+medādyādhikye upāyāntaram āha medaśleṣmādhika iti. medaś ca śleṣmā ca medaśleṣmāṇau tāv adhikau
+yasya sa tādṛśaḥ puruṣaḥ. pūrvaṃ prāṇāyāmābhyāsāt prāk na tu prāṇāyāmābhyāsakāle. ṣaṭ karmāṇi
+vakṣyamāṇāni samācaret samyag ācaret. anyas tu medaśleṣmādhikyarahitas tu tāni ṣaṭ karmāṇi nācaret.
+tatra hetum āha doṣāṇāṃ vātapittakaphānāṃ samasya bhāvaḥ samabhāvaḥ samatvaṃ tasmād doṣāṇāṃ
+samatvād ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_022}
+\startsloka
+{\bf dhautir bastis tathā netis trāṭakaṃ naulikaṃ tathā
+kapālabhātiś caitāni ṣaṭkarmāṇi pracakṣate (2.22)}
+\stopsloka
+
+ṣaṭkarmāṇy uddiśati dhautir iti. spaṣṭam.
+%\end{vsid}
+
+%\begin{vsid}{hp02_023}
+\startsloka
+{\bf karmaṣaṭkam idaṃ gopyaṃ ghaṭaśodhanakārakam
+vicitraguṇasandhāyi pūjyate yogipuṅgavaiḥ (2.23)}
+\stopsloka
+
+idaṃ rahasyam ity āha karmaṣaṭkam iti. ghaṭasya śarīrasya śodhanaṃ malāpanayanaṃ karotīti
+ghaṭaśodhanakārakam idam uddiṣṭaṃ karmaṇāṃ ṣaṭkam dhautyādikaṃ gopyaṃ gopanīyam. yataḥ
+vicitraguṇasandhāyi\var{sandhāyi \emend \lem sandhāyī \Wthree \Msix \Tue sandhāyīti \EdLo \EdMu}
+vicitraṃ vilakṣaṇaṃ guṇaṃ ṣaṭkarmarūpaṃ sandhātuṃ kartuṃ śīlam asyeti vicitraguṇasandhāyi
+yogipuṅgavair yogiśreṣṭhaiḥ pūjyate satkriyate. gopanābhāve tu karmaṣaṭkam anyair api viditaṃ syād
+iti yoginaḥ pūjyatvābhāvaḥ prasajyeteti bhāvaḥ. etenedam eva karmaṣaṭkasya mukhyaṃ phalam iti
+sūcitam. medaśleṣmādināśasya prāṇāyāmair api sambhavāt. tad uktaṃ ṣaṭkarmayogam āpnoti
+pavanābhyāsatatpara iti pūrvottaragranthasyāpy evam eva svārasyāc ca.
+%\end{vsid}
+
+%\begin{vsid}{hp02_024}
+\startsloka
+{\bf tatra dhautiḥ
+caturaṅgulavistāraṃ hastapañcadaśāyatam
+gurūpadiṣṭamārgeṇa siktaṃ vastraṃ śanair graset
+punaḥ pratyāharec caitad uditaṃ dhautikarma tat (2.24)}
+\stopsloka
+
+dhautikarmāha caturaṅgulam iti. caturṇām aṅgulānāṃ samāhāraś caturaṅgulaṃ caturaṅgulaṃ
+vistāro\var{caturaṅgulaṃ vistāro \Tue \EdMu \Msix \lem caturaṅgulavistāro \Wthree vistāro \EdLo} yasya
+tādṛśaṃ hastānāṃ pañcadaśair āyataṃ dīrghaṃ siktaṃ jalārdraṃ kiñcid uṣṇaṃ vastraṃ paṭaṃ tac ca
+sūkṣmaṃ nūtanoṣṇīṣādeḥ khaṇḍaṃ grāhyam. guruṇā upadiṣṭo yo mārgaḥ vastragrasanaprakāras tena śanair
+mandaṃ mandaṃ kiñcit kiñcid graset.  prathamadine hastamātraṃ graset.\var{prathamadine hastamātraṃ
+  graset \Wthree \Msix \ \om \Tue \EdLo (reports something misssing, but the missing part is not
+  mentioned) \EdMu \EdAd} dvitīye dine hastadvayaṃ tṛtīyadine hastatrayam. evaṃ dinavṛddhyā
+hastapañcadaśamātraṃ graset. punar iti. tasya prāntaṃ rājadantamadhye dṛḍhaṃ saṃlagnaṃ kṛtvā
+naulikarmaṇodarasthaṃ vastraṃ samyak cālayitvā punaḥ śanaiḥ pratyāharec ca etad vastram udgiren
+niṣkāsayec ca. tad dhautikarma uditaṃ kathitaṃ siddhaiḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_025}
+\startsloka
+{\bf kāsaśvāsaplīhakuṣṭhaṃ kapharogāś ca viṃśatiḥ
+dhautikarmaprabhāveṇa prayānty eva na saṃśayaḥ (2.25)}
+\stopsloka
+
+dhautikarmaṇaḥ phalam āha kāsaśvāseti. kāsaś ca śvāsaś ca plīhaś ca kuṣṭhaṃ ca. samāhāradvandvaḥ
+kāsādayo rogaviśeṣāḥ viṃśatiḥ viṃśatisaṅkhyākāḥ kapharogāś ca dhautikarmaṇaḥ prabhāveṇa gacchanty
+eva na saṃśayaḥ niścitam etad ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_026}
+\startsloka
+{\bf atha bastiḥ
+nābhidaghnajale pāyau nyastanālotkaṭāsanaḥ
+ādhārākuñcanaṃ kuryāt kṣālanaṃ bastikarma tat (2.26)}
+\stopsloka
+
+atha bastikarmāha nābhidaghneti. nābhipramāṇaṃ nābhidaghnam. pramāṇe
+daghnacpratyayaḥ.\comment{Compare Aṣṭādhyāyī 5.3.37. A vārttika mentions parimāṇa as a possible
+  sense on the same Sūtra, but not for daghna. Readings of many manuscripts have parimāṇa for
+  pramāṇa.} tasminn nābhidaghne nābhipramāṇe jale nadyāditoye pāyur gudaṃ tasmin nyasto nālo
+vaṃśanālo yena kaniṣṭhikāpraveśayogyarandhrayuktaṃ ṣaḍaṅguladīrghaṃ vaṃśanālaṃ gṛhītvā caturaṅgulaṃ
+pāyau praveśayet. aṅgulidvayamitaṃ bahiḥ sthāpayet. utkaṭam āsanaṃ yasya sa utkaṭāsanaḥ.
+pārṣṇidvaye sphicau vinyasya pādāṅgulibhiḥ sthitir utkaṭāsanam. ādhārasya mūlādhārasya ākuñcanaṃ
+yathā jalam antaḥ praviśet tathā saṅkocanaṃ kuryāt. antaḥpraviṣṭaṃ jalaṃ naulikarmaṇā cālayitvā
+tyajet. tat kṣālanaṃ bastikarmocyate. dhautibastikarmadvayaṃ bhojanāt prāg eva kartavyam.
+tadanantaraṃ bhojane vilambo 'pi na kāryaḥ. kecit tu pūrvaṃ mūlādhāreṇa vāyor ākarṣaṇam abhyasya
+jale sthitvā pāyau nālapraveśanam antareṇaiva bastikarmābhyasyanti. tathākaraṇe sarvaṃ jalaṃ bahir
+nāyāti. ato nānārogadhātukṣayādisambhavāc ca tathā bastikarma naiva vidheyam. kim anyathā
+svātmārāmaḥ pāyau nyastanāla iti brūyāt.
+%\end{vsid}
+
+%\begin{vsid}{hp02_027}
+\startsloka
+{\bf gulmaplīhodaraṃ cāpi vātapittakaphodbhavāḥ
+bastikarmaprabhāveṇa kṣīyante sakalāmayāḥ (2.27)}
+\stopsloka
+
+bastikarmaguṇān āha dvābhyāṃ gulmaplīhodaram iti. gulmaś ca plīhaś ca rogaviśeṣāv udaraṃ jalodaraṃ
+ca teṣāṃ samāhāradvandvaḥ. vātaś ca pittaṃ ca kaphaś ca tebhyaḥ udbhavāḥ ekaikasmād dvābhyāṃ
+sarvebhyo vā jātāḥ sakalāḥ sarve āmayā rogāḥ bastikarmaṇaḥ prabhāvaḥ sāmarthyaṃ śaktiḥ tena
+kṣīyante naśyanti.
+%\end{vsid}
+
+%\begin{vsid}{hp02_028}
+\startsloka
+{\bf dhātvindriyāntaḥkaraṇaprasādaṃ 
+dadyāc ca kāntiṃ dahanapradīptim
+aśeṣadoṣopacayaṃ nihanyād 
+abhyasyamānaṃ jalabastikarma (2.28)}
+\stopsloka
+
+dhātv iti. abhyasyamānam anuṣṭhīyamānaṃ jale bastikarma jalabastikarma kartṛ dadyād anuṣṭhātur
+iti śeṣaḥ. dhātavo
+
+\startsloka
+rasāsṛṅmāṃsamedo'sthimajjāśukrāṇi dhātavaḥ\comment{vāgbhaṭaḥ 1.13?}
+\stopsloka
+
+ity uktāḥ.  indriyāṇi vākpāṇipādapāyūpasthāni pañca karmendriyāṇi śrotratvakcakṣurjihvāghrāṇāni
+pañca jñānendriyāṇi ca antaḥkaraṇāni manobuddhicittāhaṅkārarūpāṇi teṣāṃ
+paritāpavikṣepaśokādi\var{śokādi \Msix \Wthree \Tue \lem śoka
+  \EdLo}mohagauravāvaraṇadainyādirājasatāmasadharmavinivartanena
+sukhaprakāśalāgha\-vā\-disāttvikadharmāvirbhāvaḥ prasādas taṃ kāntiṃ dyutiṃ dahanasya jaṭharāgneḥ
+pradīptiṃ prakṛṣṭāṃ dīptiṃ ca. tathā aśeṣāḥ samastā ye doṣāḥ vātapittakaphās teṣām upacayo
+'pacayasyāpy\var{upacayo 'pacayasyāpy \Wthree \Tue \Msix \lem \EdLo inconclusive}
+upalakṣaṇam. upacayāpacayau nihanyād nitarāṃ hanyāt.  doṣasāmyarūpam ārogyaṃ kuryād ity arthaḥ.
+%\end{vsid}
+
+% Commentary on 2.28:
+% Brahmānanda clarifies the construction as: {\em jalabastikarma kartṛ dadyād anuṣṭhātur
+% iti śeṣaḥ} ``...''.
+
+%   bis hier mit \Msix kollationiert.
+
+%\begin{vsid}{hp02_030}
+\startsloka
+{\bf
+atha netiḥ
+sūtraṃ vitasti susnigdhaṃ nāsānāle praveśayet
+mukhān nirgamayec caiṣā netiḥ siddhair nigadyate (2.29)}
+\stopsloka
+
+atha netikarmāha sūtram iti. vitasti vitastimitaṃ vitastir ity upalakṣaṇam adhikasyāpi. yāvatā
+sūtreṇa samyag netikarma bhavet tāvad grāhyam. susnigdhaṃ suṣṭhu snigdhaṃ granthyādirahitaṃ
+sūtram. tac ca navadhā dvādaśadhā pañcadaśadhā vā guṇitaṃ sudṛḍhaṃ grāhyam. nāsā nāsikā saiva nālaḥ
+sacchidratvāt tasmin praveśayet. mukhān nirgamayen niṣkāsayet.
+
+tatprakāras tv evaṃ. sūtraprāntaṃ nāsānāle praveśyetaranāsāpuṭam aṅgulyā nirudhya pūrakaṃ kuryāt.
+punaś ca mukhena recayet. punaḥ punar evaṃ kurvato mukhe sūtraprāntam āyāti. tatsūtraprāntaṃ
+nāsābahiḥsthasūtraprāntaṃ ca gṛhītvā śanaiś cālayed iti. cakārād ekasmin nāsānāle praveśyetarasmin
+nirgamayed ity uktam. tatprakāras tv ekasmin nāsānāle sūtraprāntaṃ praveśyetaranāsāpuṭam aṅgulyā
+nirudhya pūrakaṃ kuryāt paścād itaranāsānālena recayet. punaḥ punar evaṃ kurvata itaranāsānāle
+sūtraprāntam āyāti tasya pūrvavac cālanaṃ kuryād iti. ayaṃ prakāras tu bahuvāraṃ kurvataḥ kadācid
+bhavati. eṣoktā siddhair aṇimādiguṇasampannaiḥ.
+
+tad uktam
+
+\startsloka
+avāptāṣṭaguṇaiśvaryāḥ siddhāḥ sadbhir nirūpitāḥ
+\stopsloka
+
+iti. netir nigadyate netir iti kathyate.
+%\end{vsid}
+
+%\begin{vsid}{hp02_031}
+\startsloka
+{\bf kapālaśodhanī caiva divyadṛṣṭipradāyinī
+jatrūrdhvajātarogaughaṃ netir āśu nihanti ca (2.30)}
+\stopsloka
+
+netiguṇān āha kapālaśodhanīti. kapālaṃ śodhayati śuddhaṃ malarahitaṃ karotīti kapālaśodhanī.
+cakārān nāsānālādīnām api. evaśabdo 'vadhāraṇe. divyāṃ sūkṣmapadārthagrāhiṇīṃ dṛṣṭiṃ prakarṣeṇa
+dadātīti divyadṛṣṭipradāyinī. netir netikriyā. jatruṇoḥ skandhasandhyayor ūrdhvam uparibhāge jāto
+jatrūrdhvajātaḥ sa cāsau rogāṇām aughaś ca tam āśu jhaṭiti nihanti. cakāraḥ pādapūraṇe.
+
+\startsloka
+  skandhe bhujaśiro 'ṃso 'strī sanṃdhī tasyaiva jatruṇī
+\stopsloka
+
+ity amaraḥ\comment{Amarakośa 2.6.78}.
+%\end{vsid}
+
+%\begin{vsid}{hp02_032}
+\startsloka
+{\bf atha trāṭakam
+nirīkṣen niścaladṛśā sūkṣmalakṣyaṃ samāhitaḥ
+aśrusampātaparyantam ācāryais trāṭakaṃ smṛtam (2.31)}
+\stopsloka
+
+trāṭakam āha nirīkṣed iti. samāhitaḥ ekāgracitto 'bhyāsī niścalā acalā cāsau\var{niścalā acalā
+  cāsau \Wthree \lem niścalā cāsau \EdMu (and six mss), niścalācalatvācāsau \EdLo} dṛk ca dṛṣṭis tayā
+sūkṣmaṃ ca tal lakṣyaṃ ca sūkṣmalakṣyam aśrūṇāṃ samyak pātaḥ patanaṃ tatparyantam.  anena
+nirīkṣaṇasyāvadhir uktaḥ. nirīkṣet paśyet. ācāryair matsyendrādibhir idaṃ trāṭakaṃ trāṭakakarma
+smṛtaṃ kathitam.
+%\end{vsid}
+
+%\begin{vsid}{hp02_033}
+\startsloka
+{\bf mocanaṃ netrarogāṇāṃ tandrādīṇāṃ kapāṭakam
+yatnatas trāṭakaṃ gopyaṃ yathā hāṭakapeṭakam (2.32)}
+\stopsloka
+
+trāṭakaguṇān āha mocanam iti. netrasya rogā netrarogās teṣāṃ mocanaṃ nāśakaṃ tandrā ādir yeṣām
+ālasyādīnāṃ teṣāṃ kapāṭakam kapāṭavad antardhāyakam abhibhāvakam ity arthaḥ. tandrā tāmasaś
+cittavṛttiviśeṣaḥ. trāṭakaṃm trāṭakākhyaṃm karma yatnataḥ prayatnād gopyaṃ gopanīyam.  gopane
+dṛṣṭāntam āha yatheti. hāṭakasya suvarṇasya peṭakaṃ peṭī iti loke prasiddhaṃ. yathā yena prakāreṇa
+gopyate tadvat.
+%\end{vsid}
+
+%\begin{vsid}{hp02_034}
+\startsloka
+{\bf atha nauliḥ
+amandāvartavegena tundaṃ savyāpasavyataḥ
+natāṃso bhrāmayed eṣā nauliḥ siddhaiḥ pracakṣyate (2.33)}
+\stopsloka
+
+naulikarmāha amandeti. natau namrībhūtāv aṃsau skandhau yasya sa natāṃsaḥ pumān amando 'tiśayito ya
+āvartas tasyeva\var{āvartas tasyeva \EdMu I II XII XIII \lem āvartasyeva \EdLo \Wthree} jalabhramasyeva
+vego javas tena tundam udaram
+
+\startsloka
+  picaṇḍakukṣī jaṭharodaraṃ tundaṃ stanau kucau
+\stopsloka
+
+ity amaraḥ\comment{Amarakoṣa 2.6.77}. savyaṃ cāpasavyaṃ ca savyāpasavye dakṣiṇavāmabhāgau tayoḥ
+savyāpasavyataḥ. saptamyarthe tasiḥ. bhrāmayed bhramantaṃ prerayet. siddhair eṣā nauliḥ
+pracakṣyate kathyate.
+%\end{vsid}
+
+%\begin{vsid}{hp02_035}
+\startsloka
+{\bf mandāgnisandīpanapācanādi-
+sandhāyikānandakarī sadaiva
+aśeṣadoṣāmayaśoṣaṇī ca 
+haṭhakriyāmaulir iyaṃ ca nauliḥ (2.34)}
+\stopsloka
+
+nauliguṇān āha mandāgnīti. mandaś cāsāv agniś ca jaṭharāgnis tasya sandīpanaṃ samyag dīpanaṃ ca
+pācanaṃ ca bhuktānnaparipākaś ca mandāgnisandīpanapācane te ādinī yasya tan
+mandāgnisandīpanapācanādi tasya sandhāyikā vidhātrī. ādiśabdena malaśuddhyādi. sadaiva
+sarvadaivānandakarī sukhakarī. aśeṣāḥ samastāś ca te doṣāś ca vātādaya āmayāś ca rogās teṣāṃ śoṣaṇī
+śoṣaṇakartrī. haṭhasya kriyāṇāṃ dhautyādīnāṃ maulir maulir ivottamā dhautibastyor
+naulisāpekṣatvāt. iyam uktā nauliḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_036}
+\startsloka
+{\bf atha kapālabhātiḥ
+bhastrāval lohakārasya recapūrau sasambhramau
+kapālabhātir vikhyātā kaphadoṣaviśoṣaṇī (2.35)}
+\stopsloka
+
+atha kapālabhātiṃ tadguṇaṃ ca āha bhastrāvad iti. lohakārasya bhastrā agner dhamanasādhanībhūtaṃ
+carma tadvat sambhrameṇa saha vartamānau sasambhramāv amandau yau recapūrau recakapūrakau
+kapālabhātir iti vikhyātā. kīdṛśī kaphadoṣaviśoṣaṇī. kaphasya doṣā viṃśatibhedabhinnāḥ. tad uktaṃ
+nidāne kapharogāś ca viṃśatir iti. teṣāṃ viśoṣaṇī vināśinī.
+%\end{vsid}
+
+%\begin{vsid}{hp02_037}
+\startsloka
+{\bf ṣaṭkarmanirgatasthaulyakaphadoṣamalādikaḥ
+prāṇāyāmaṃ tataḥ kuryād anāyāsena siddhyati (2.36)}
+\stopsloka
+
+ṣaṭkarmaṇāṃ prāṇāyāmopakārakatvam āha ṣaṭkarmeti. ṣaṭkarmabhir dhautiprabhṛtibhiḥr nirgatāḥ.
+sthaulyaṃ sthūlasya bhāvaḥ sthūlatvam. kaphadoṣā viṃśatisaṅkhyākāḥ malādayaś ca yasya sa tathā.
+śeṣād vibhāṣā\comment{Aṣṭādhyāyī 5.4.154.} iti kapratyayaḥ. ādiśabdena pittādayaḥ. prāṇāyāmaṃm
+kuryāt. tatas tasmāt ṣaṭkarmapūrvakāt prāṇāyāmād anāyāsenāśrameṇa siddhyati yoga iti
+śeṣaḥ. ṣaṭkarmākaraṇe tu prāṇāyāme śramādhikyaṃ syād iti bhāvaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_038}
+\startsloka
+{\bf prāṇāyāmair eva sarve praśuṣyanti malā iti
+ācāryāṇāṃ tu keṣāñcid anyat karma na samṃmatam (2.37)}
+\stopsloka
+
+matabhedena ṣaṭkarmaṇām anupayogam āha prāṇāyāmair iti. prāṇāyāmair eva. evaśabdaḥ
+ṣaṭkarmavyavacchedārthaḥ. sarve malāḥ praśuṣyanti. malā ity upalakṣaṇaṃ sthaulyakaphapittādīnām.
+iti hetoḥ keṣāñcid ācāryāṇāṃ yājñavalkyādīnām anyat karma ṣaṭkarma na samṃmatam nābhimatam.
+ācāryalakṣaṇam uktaṃ vāyupurāṇe
+
+\startsloka
+ācinoti ca śāstrārtham ācāre sthāpayed api
+svayam ācarate yasmād ācāryas tena cocyate iti
+\stopsloka
+%\end{vsid}
+
+%\begin{vsid}{hp02_29}
+\startsloka
+{\bf atha gajakaraṇī
+udaragatapadārtham udvamanti 
+pavanam apānam udīrya kaṇṭhanāle
+kramaparicayavaśyanāḍicakrā
+gajakaraṇīti nigadyate haṭhajñaiḥ (2.38)}
+\stopsloka
+
+gajakaraṇīm āha udaragateti. apānaṃ pavanam apānavāyuṃ kaṇṭhanāle kaṇṭho nāla iva kaṇṭhanālas
+tasminn udīryotkṣipyodare gataḥ prāptaḥ sa cāsau padārthaś ca bhuktapītānnajalādis taṃ yayodvamanti
+udgiranti yayā. yogina ity adhyāhāraḥ. krameṇa yaḥ paricayo 'bhyāsas tena vaśyaṃ svādhīnaṃ nāḍīnāṃ
+cakraṃ yasyāṃ sā tathā. sā kriyā haṭhajñair haṭhayogābhijñair gajakaraṇīti nigadyate
+kathyate. kramaparicayavaśyanāḍimārgā iti kvacit pāṭhaḥ. tasyāyam arthaḥ kramaparicayena vaśyo
+nāḍyāḥ śaṅkhinyāḥ mārgaḥ kaṇṭhaparyanto yasyāṃ sā tathā.
+%\end{vsid}
+
+%\begin{vsid}{hp02_039}
+\startsloka
+{\bf brahmādayo 'pi tridaśāḥ pavanābhyāsatatparāḥ
+abhūvann antakabhayāt tasmāt pavanam abhyaset (2.39)}
+\stopsloka
+
+prāṇāyāmo 'vaśyam abhyasanīyaḥ sarvottamair abhyastatvān mahāphalatvāc ceti sūcayann āha caturbhiḥ
+brahmādaya iti. brahmā ādir yeṣāṃ te brahmādayas te 'pi kim utānye ity arthaḥ. tridaśāḥ devāḥ.
+antayatīty antakaḥ kālas tasmād bhayam antakabhayaṃ tasmāt pavanasya prāṇavāyoḥ yo 'abhyāsaḥ
+recakapūrakakumbhakabhedabhinnaprāṇāyāmānuṣṭhānarūpas tasmin tatparāḥ avahitā abhūvann āsan.
+tasmāt pavanam abhyaset prāṇam abhyaset.
+%\end{vsid}
+
+%\begin{vsid}{hp02_040}
+\startsloka
+{\bf yāvad baddho marud dehe yāvac cittaṃ nirākulam
+yāvad dṛṣṭir bhruvor madhye tāvat kālabhayaṃ kutaḥ (2.40)}
+\stopsloka
+
+yāvad iti. yāvad yāvatkālaparyantaṃ marut prāṇānilo dehe śarīre baddhaḥ śvāsocchvāsakriyāśūnyaḥ.
+yāvac cittam antaḥkaraṇaṃ nirākulam avikṣiptaṃ samāhitam. yāvad bhruvor madhye dṛṣṭir
+antaḥkaraṇavṛttiḥ dṛśir atra jñānasāmānyārthaḥ. tāvat tāvatkālaparyantaṃ kalayatīti kālo 'ntakas
+tasmād bhayaṃ kutaḥ. na kuto 'pīty arthaḥ. tathā ca vakṣyati
+
+\startsloka
+khādyate na ca kālena bādhyate na ca karmaṇā
+sādhyate na sa kenāpi yogī yuktaḥ samādhinā\comment{HP 4.108.} 
+\stopsloka
+
+iti svādhīnamṛtyur bhavatīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_041}
+\startsloka
+{\bf vidhivat prāṇasaṃyāmair nāḍīcakre viśodhite
+suṣumnṇāvadanaṃ bhittvā sukhād viśati mārutaḥ (2.41)}
+\stopsloka
+
+vidhivad iti. vidhivat prāṇasaṃyāmair āsanajālandharabandhādividhiyuktaprāṇāyāmair nāḍīcakre
+nāḍīnāṃ cakraṃ samūhas tasmin viśodhite viśeṣeṇa śodhite nirmale sati māruto vāyuḥ suṣumṇā
+iḍāpiṅgalayor madhyasthā nāḍī tasyā vadanaṃ mukhaṃ bhittvā sukhād anāyāsād viśati suṣumṇnāntar iti
+śeṣaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_042}
+\startsloka
+{\bf mārute madhyasañcāre manaḥsthairyaṃ prajāyate
+yo manaḥsusthirībhāvaḥ saivāvasthā manonmanī (2.42)}
+\stopsloka
+
+māruta iti. mārute prāṇavāyau madhye suṣumṇnāmadhye sañcāraḥ samyak caraṇaṃ gamanaṃ mūrdhaparyantaṃ
+yasya sa madhyasañcāras tasmin sati manasaḥ sthairyaṃ dhyeyākāravṛttipravāho jāyate
+prādurbhavati. yo manasaḥ susthirībhāvaḥ suṣṭhu sthirībhavanaṃ saiva manonmanyavasthā.
+manonmanīśabda unmanīparyāyaḥ. tathāgre vakṣyati rājayogaḥ samādhiś ca\comment{HP 4.3} ity ādinā.
+%\end{vsid}
+
+%\begin{vsid}{hp02_043}
+\startsloka
+{\bf tatsiddhaye vidhānajñāś citrān kurvanti kumbhakān
+vicitrakumbhakābhyāsād vicitrāṃ siddhim āpnuyāt (2.43)}
+\stopsloka
+
+vicitreṣu kumbhakeṣu pravṛttiṃ janayituṃ teṣāṃ mukhyaphalam avāntaraphalaṃ cāha tatsiddhaya iti.
+vidhānaṃ kumbhakānuṣṭhānaprakāras taj jānantīti vidhānajñās tatsiddhaye unmanyavasthāsiddhaye
+citrān sūryabhedanādibhedena nānāvidhān kumbhakān kurvanti. vicitrāś ca te kumbhakāś ca
+vicitrakumbhakās teṣām abhyāsād anuṣṭhānād vicitrām aṇimādibhedena nānāvidhāṃ vilakṣaṇāṃ vā
+janmauṣadhimantratapojātābhyāṃ siddhim. tad uktaṃ bhāgavate
+
+\startsloka
+janmauṣadhitapomantrair yāvatya iha siddhayaḥ
+yogenāpnoti tāḥ sarvā nānyair yogagatiṃ vrajet\comment{Bhāgavatapurāṇa 11.15.34.} 
+\stopsloka
+
+iti. āpnuyāt pratyāhārādiparamparayeti bhāvaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_044}
+\startsloka
+{\bf atha kumbhakabhedāḥ
+sūryabhedanam ujjāyī sītkārī śītalī tathā
+bhastrikā bhrāmarī mūrcchā plāvinīty aṣṭakumbhakāḥ (2.44)}
+\stopsloka
+
+athāṣṭakumbhakān nāmabhir nirdiśati sūryabhedanam iti spaṣṭam.
+%\end{vsid}
+
+%\begin{vsid}{hp02_045}
+\startsloka
+{\bf pūrakānte tu kartavyo bandho jālandharābhidhaḥ
+kumbhakānte recakādau kartavyas tūḍḍiyānakaḥ (2.45)}
+\stopsloka
+
+atha haṭhasiddhāv ananyathāsiddhāṃ pāramarahasyāṃ sarvakumbhakasādhāraṇayuktim āha tribhiḥ
+pūrakānta iti. jālandhara ity abhidhā nāma yasya sa jālandharābhidho bandho badhnāti prāṇavāyum iti
+bandhaḥ. kaṇṭhākuñcanapūrvakaṃ cibukasya hṛdi sthāpanaṃ jālandharabandhaḥ pūrakānte pūrakasyānte
+pūrakānantaraṃ jhaṭiti kartavyaḥ. tuśabdāt kumbhakādau. uḍḍiyānakas tu kumbhakānte
+kumbhakasyānte kiñcit kumbhakaśeṣe recakasyādau recakādau recakāt pūrvaṃ kartavyaḥ.
+prayatnaviśeṣeṇa nābhipradeśasya pṛṣṭhata ākarṣaṇam uḍḍiyānabandhaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_046}
+\startsloka
+{\bf adhastāt kuñcanenāśu kaṇṭhasaṅkocane kṛte
+madhye paścimatānena syāt prāṇo brahmanāḍigaḥ (2.46)}
+\stopsloka
+
+adhastād iti. kaṇṭhasya saṅkocanaṃ kaṇṭhasaṅkocanaṃ tasmin kṛte sati jālandharabandhe kṛte satīty
+arthaḥ. āśv avyavahitottaram evādhastād adhaḥpradeśād ākuñcanena ādhārākuñcanena mūlabandhenety
+arthaḥ.  madhye nābhipradeśe paścimataḥ pṛṣṭhatas tānaṃ tānanam ākarṣaṇaṃ tenoḍḍiyānabandhenety
+arthaḥ. uktarītyā kṛtena bandhatrayeṇa prāṇo vāyur brahmanāḍiṃ suṣumnṇāṃ gacchatīti brahmanāḍigaḥ
+suṣumṇnānāḍigāmī syād ity arthaḥ.
+
+atredaṃ rahasyaṃ yadi śrīgurumukhāj jihvābandhaḥ samyak parijñātas tarhi jihvābandhapūrvakena
+jālandharabandhenaiva prāṇāyāmaḥ siddhyati. vāyuprakopo naiva bhavati.  vapuḥkṛśatvaṃ vadane
+prasannatā\comment{HP 2.78} ityādīni sarvāṇi lakṣaṇāni jāyanta iti mūlabandhoḍḍiyānabandhau
+nopayuktau. tayor jihvābandhapūrvakeṇa jālandharabandhenānyathāsiddhatvāt. jihvābandho na viditaś
+ced adhastāt kuñcanena iti ślokoktarītyā prāṇāyāmāḥ kartavyāḥ. trayo 'pi bandhā gurumukhāj
+jñātavyāḥ. mūlabandhas tvasamyagjñāto nānārogotpādakaḥ. tathā hi yadi mūlabandhe kṛte dhātukṣayo
+viṣṭambho 'gnimāndyaṃ nādamandyaṃ guṭikāsamūhākāram ajasyeva purīṣaṃ syāt tadā mūlabandhaḥ samyag
+na jñāta iti bodhyam. yadi tu dhātupuṣṭiḥ samyag malaśuddhir agnidīptiḥ samyag nādābhivyaktiś ca
+syāt tadā jñeyaṃ tadā mūlabandhaḥ samyag jñāta iti.
+%\end{vsid}
+
+%\begin{vsid}{hp02_047}
+\startsloka
+{\bf āpānam ūrdhvam utthāpya prāṇaṃ kaṇṭhād adho nayet
+yogī jarāvimuktaḥ san ṣoḍaśābdavayā bhavet (2.47)}
+\stopsloka
+
+āpānam iti. āpānam āpānavāyum ūrdhvam utthāpya ādhārākuñcanena prāṇaṃ prāṇavāyuṃ kaṇṭhād adhaḥ
+adhobhāge nayet yogī prāpayed. yaḥ sa yogī yogo 'syāsti abhyasyatveneti yogī yogābhyāsī jarayā
+vārdhakyena vimukto viśeṣeṇa muktaḥ san ṣoḍaśānām abdānāṃ samāhāraḥ ṣoḍaśābdam ṣoḍaśābdaṃ vayo
+yasya sa tādṛśo bhavet. yady api pūrakānte tu kartavyaḥ\comment{HP 2.45.} ity ādinā trayāṇāṃ
+ślokānām eka evārthaḥ paryavasyati tathāpi pūrakānte tu kartavyaḥ ity anena bandhānāṃ kāla
+uktaḥ. adhastāt kuñcanenety anena bandhānāṃ svarūpam uktam. apānam ūrdhvam utthāpyety anena
+bandhānāṃ phalam uktam iti viśeṣaḥ. jālandharabandhe mūlabandhe ca kṛte uḍḍiyānabandho bhavaty
+evety asmin śloke-noktaḥ. tathā coktaṃ jñāneśvareṇa gītāṣaṣṭhādhyāyavyākhyāyāṃ mūlabandhe
+jālandharabandhe ca kṛte nābher adhobhāga ākarṣaṇākhyo bandhaḥ svayam eva
+bhavati\comment{Jñāneśvarī 6.200?} iti.
+%\end{vsid}
+
+%\begin{vsid}{hp02_048}
+\startsloka
+{\bf atha sūryabhedanam
+āsane sukhade yogī baddhvā caivāsanaṃ tataḥ
+dakṣanāḍyā samākṛṣya bahiḥsthaṃ pavanaṃ śanaiḥ (2.48)}
+\stopsloka
+
+\startsloka
+yogābhyāsakramaṃ vakṣye yogināṃ yogasiddhaye
+uṣaḥkāle samutthāya prātaḥkāle 'thavā budhaḥ
+guruṃ saṃsmṛtya śirasi hṛdaye sveṣṭadevatām
+śaucaṃ kṛtvā dantaśuddhiṃ vidadhyād bhasmadhāraṇam
+śucau deśe maṭhe ramye pratiṣṭhāpyāsanaṃ mṛdu 
+tatropaviśya saṃsmṛtya manasā gurum īśvaram
+deśakālau ca saṅkīrtya saṅkalpavidhipūrvakam
+\stopsloka
+
+adyetyādi śrīparameśvaraprasādapūrvakam samādhitatphalasiddhyartham āsanapūrvakān prāṇāyāmādīn kariṣye
+
+\startsloka
+  anantaṃ praṇamed devaṃ nāgeśaṃ pīṭhasiddhaye
+\stopsloka
+
+maṇibhrājatphaṇāsahasravidhṛtaviśvambharāmaṇḍalāyānantāya nāgarājāya namaḥ
+
+\startsloka
+tato 'bhyased āsanāni śrame jāte śavāsanam
+ante samabhyaset tat tu śramābhāve tu nābhyaset
+karaṇīṃ viparītākhyāṃ kumbhakān pūrvam abhyaset
+jālandharasya dārḍhyārthaṃ kumbhakeṣūpayogataḥ
+vidhāyācamanaṃ kṛtvā karmāṅgaṃ prāṇasaṃyamam 
+yogīndrādīn namaskṛtya kaurmāc ca śivavākyataḥ
+\stopsloka
+
+kūrmapurāṇe śivavākyam
+
+\startsloka
+namaskṛtyātha yogīndrān saśiṣyāṃś ca vināyakam
+guruṃ caivātha māṃ yogī yuñjīta susamāhitaḥ
+baddhvābhyāse siddhapīṭhaṃ kumbhakān bandhapūrvakam
+prathame daśa kartavyāḥ pañcavṛddhyā dine dine
+kāryā aśītiparyantaṃ kumbhakāḥ susamāhitaiḥ
+yogīndraiḥ prathamaṃ kuryād abhyāsaṃ candrasūryayoḥ
+anulomavilomākhyam etaṃ prāhur manīṣiṇaḥ
+sūryabhedanam abhyasya bandhapūrvakam ekadhīḥ
+ujjāyinaṃ tataḥ kuryāt sītkārīṃ śītalīṃ tataḥ 
+bhastrikāṃ ca samabhyasya kuryād anyān na vā parān
+mudrāḥ samabhyased buddhā guruvaktrād yathākramam
+tataḥ padmāsanaṃ baddhvā kuryān nādānucintanam
+abhyāsaṃ sakalaṃ kuryād īśvarārpaṇam ādṛtaḥ
+abhyāsād utthitaḥ snānaṃ kuryād uṣṇena vāriṇā
+snātvā samāpayen nityaṃ karma saṅkṣepataḥ sudhīḥ
+madhyāhne 'pi tathābhyasya kiñcid viśramya bhojanam
+kurvīta yogināṃ pathyam apathyaṃ na kadācana
+elāṃ vāpi lavaṅgaṃ vā bhojanānte ca bhakṣayet
+kecit karpūram icchanti tāmbūlaṃ śobhanaṃ tathā
+cūrṇena rahitaṃ śastaṃ pavanābhyāsayoginām
+iti cintāmaṇer vākyaṃ svārasyaṃ bhajate na hi
+kecit padena yasmāt tu tayoḥ śītauṣṇyahetutā
+bhojanānantaraṃ kuryān mokṣaśāstrāvalokanam
+purāṇaśravaṇaṃ vāpi nāmasaṃkīrtanaṃ vibhoḥ
+sāyaṃ sandhyāvidhiṃ kṛtvā yogaṃ pūrvavad abhyaset
+yadā trighaṭikāśeṣo divaso 'bhyāsam ācaret
+abhyāsānantaraṃ kāryā sāyaṃsandhyā tadā budhaiḥ
+ardharātre haṭhābhyāsaṃ vidadhyāt pūrvavad yamī
+viparītāṃ tu karaṇīṃ sāyaṃkālārdharātrayoḥ
+nābhyased bhojanād ūrdhvaṃ yataḥ sā na praśasyate
+\stopsloka
+
+athoddeśānukrameṇa kumbhakān vivakṣus tatra prathamoditaṃ sūryabhedanaṃ tadguṇāṃś cāha tribhiḥ
+āsana iti. sukhaṃ dadātīti sukhadaṃ tasmin sukhade.
+
+\startsloka
+śucau deśe pratiṣṭhāpya sthiram āsanam ātmanaḥ
+nātyucchritaṃ nātinīcaṃ cailājinakuśottaram\comment{Bhagavadgītā 6.11}
+\stopsloka
+
+ity uktalakṣaṇe viviktadeśe ca sukhāsanasthaḥ śuciḥ samagrīvaśiraḥśarīraḥ\comment{kaivalyopaniṣat
+  5.} iti śruteś ca. cailājinakuśottara āsane. āste 'sminn ity āsanam āsyate 'neneti vā tasmin yogī
+yogābhyāsī.  āsanaṃ svastikavīrasiddhapadmādyanyatamaṃ mukhyatvāt siddhāsanam eva vā baddhvaiva
+bandhanena sampādyaiva kṛtvaivety arthaḥ. tata āsanabandhānantaraṃ dakṣā dakṣiṇabhāgasthā yā nāḍī
+piṅgalā tayā bahiḥsthaṃ dehād bahir vartamānaṃ pavanaṃ vāyuṃ śanair mandamandam ākṛṣya piṅgalayā
+mandamandaṃ pūrakaṃ kṛtvety arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_049}
+\startsloka
+{\bf ākeśād ānakhāgrāc ca nirodhāvadhi kumbhayet
+tataḥ śanaiḥ savyanāḍyā recayet pavanaṃ śanaiḥ (2.49)}
+\stopsloka
+
+ākeśād iti. keśān maryādīkṛtyety ā keśaṃ tasmāt. nakhāgrān maryādī\-kṛtyety ānakhāgraṃ tasmāc
+ca. nirodhasya vāyor avarodhasya avadhir maryādā yasmin karmaṇi tat tathā kumbhayet.  keśaparyantaṃ
+nakhāgraparyantaṃ ca vāyor nirodho yathā bhavet tathātiprayatnena kumbhakaṃ kuryād ity arthaḥ.
+
+nanu
+
+\startsloka
+haṭhān niruddhaḥ prāṇo 'yaṃ romakūpeṣu niḥsaret
+dehaṃ vidārayaty eṣa kuṣṭhādi janayaty api
+tataḥ pratyāyitavyo 'sau krameṇāraṇyahastivat
+vanyo gajo gajārir vā krameṇa mṛdutām iyāt
+karoti śāstṛnirdeśān na ca taṃ parilaṅghayet
+tathā prāṇo hṛdistho'yaṃ yogināṃ kramayogataḥ
+gṛhītaḥ sevyamānas tu viśrambham upagacchati\comment{Quoted in YCM.}    
+\stopsloka
+
+iti vākyaviruddham atiprayatnena kumbhakaṃ kuryād iti katham uktam iti cen na. haṭhān niruddhaḥ
+prāṇo 'yaṃ itivākyasya balād acireṇa prāṇajayaṃ kariṣyāmīti buddhyārambhaḥ eva. evaṃ ca
+bahvabhyāsāsaktaparatvāt krameṇāraṇyahastivad iti dṛṣṭāntasvārasyāc ca. ata eva sūryācandramasor
+abhyāse dhārayitvā yathāśakti\comment{HP 3.21.} nidhārayed atirodhata\comment{HP 2.9.} iti coktaṃ
+saṅgacchate.  tasmāt kumbhakas tv atiprayatnapūrvakaṃ kartavyaḥ. yathā yathātiyatnena kumbhakaḥ
+kriyate tathā tathā tasmin guṇādhikyaṃ bhavet. yathā yathā ca śithilaḥ kumbhakaḥ syāt tathā tathā
+guṇālpatvaṃ syāt. atra yoginām anubhavo 'pi mānam. pūrakas tu śanaiḥ śanaiḥ kāryaḥ vegād vā
+kartavyaḥ.  vegād api kṛte pūrake doṣābhāvāt. recakas tu śanaiḥ śanair eva kartavyaḥ. vegāt kṛte
+recake balahāniprasaṅgāt. tataḥ śanaiḥ śanair eva recayen na tu vegata ityādy anekadhā
+granthakārokteś ca. tato nirodhāvadhi kumbhakānantaraṃ śanaiḥ śanair mandaṃ mandaṃ savye vāmabhāge
+sthitā nāḍī savyanāḍī tayā savyanāḍyā iḍayā pavanaṃ vāyuṃ recayed bahir niḥsārayet. punaḥ śanair
+ity uktas tu śanair eva recayed ity avadhāraṇārthā. tad uktaṃ
+
+\startsloka
+{\bf vismaye ca viṣāde ca dainye caivāvadhāraṇe
+tathā prasādane harṣe vākyam ekaṃ dvir ucyate iti}
+\stopsloka
+%\end{vsid}
+
+%\begin{vsid}{hp02_050}
+\startsloka
+{\bf kapālaśodhanaṃ vātadoṣaghnaṃ kṛmidoṣahṛt
+punaḥ punar idaṃ kāryaṃ sūryabhedanam uttamam (2.50)}
+\stopsloka
+
+kapālaśodhanam iti. kapālasya mastakasya śodhanaṃ śuddhikaraṃ vātajā doṣā vātadoṣā aśītiprakārās
+tān hantīti vātadoṣaghnam. kṛmīṇām udare jātānāṃ doṣo vikāras taṃ haratīti kṛmidoṣahṛt. punaḥ punar
+bhūyo bhūyaḥ kāryam. sūryeṇāpūrya kumbhayitvā candreṇa recanam iti rītyedam uttamam utkṛṣṭaṃ
+sūryabhedanaṃ sūryabhedanākhyam uktaṃ yogibhir iti śeṣaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_051}
+\startsloka
+{\bf atha ujjāyī
+mukhaṃ saṃyamya nāḍībhyām ākṛṣya pavanaṃ śanaiḥ
+yathā lagati kaṇṭhāt tu hṛdayāvadhi sasvanam (2.51)}
+\stopsloka
+
+ujjāyinam āha sārdhena mukham iti. mukham āsyaṃ saṃyamya saṃyataṃ kṛtvā mudrayitvety
+arthaḥ. kaṇṭhāt tu kaṇṭhād ārabhya hṛdayāvadhi hṛdayam avadhir yasmin karmaṇi tat tathā svanena
+sahitaṃ yathā syāt tathā. ubhe kriyāviśeṣaṇe. lagati śliṣyati pavana ity arthāt. tathā tena
+prakāreṇa nāḍībhyām iḍāpiṅgalābhyāṃ pavanaṃ vāyuṃ śanair mandam ākṛṣya ākṛṣṭaṃ kṛtvā pūrayitvety
+arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_052}
+\startsloka
+{\bf pūrvavat kumbhayet prāṇaṃ recayed iḍayā tataḥ
+śleṣmadoṣaharaṃ kaṇṭhe dehānalavivardhanam (2.52)}
+\stopsloka
+
+pūrvavad iti. prāṇaṃ pūrvavat pūrveṇa sūryabhedanena tulyaṃ pūrvavat. ākeśād ānakhāgrāc ca
+nirodhāvadhi kumbhayed ityuktarītyā kumbhayed rodhayet. tataḥ kumbhakānantaram iḍayā vāmanāḍyā
+recayet tyajet.  prāṇaṃ pūrvavat pūrveṇa sūryabhedanena tulyaṃ pūrvavat. ākeśād ānakhāgrāc ca
+nirodhāvadhi kumbhayed ityuktarītyā kumbhayed rodhayet. tataḥ kumbhakānantaram iḍayā vāmanāḍyā
+recayet tyajet.
+
+ujjāyiguṇān āha sārdhaślokena śleṣmadoṣaharam iti. kaṇṭhe kaṇṭhapradeśe śleṣmaṇo doṣāḥ śleṣmadoṣāḥ
+kāsādayas tān haratīti śleṣmadoṣaharas taṃ dehānalasya dehamadhyagatānalasya jāṭharasya vivardhanaṃ
+viśeṣeṇa vardhanaṃ dīpanam ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_053}
+\startsloka
+{\bf nāḍījalodarādhātugatadoṣavināśanam
+gacchatā tiṣṭhatā kāryam ujjāyyākhyaṃ tu kumbhakam (2.53)}
+\stopsloka
+
+nāḍīti. nāḍī śirā jalaṃ pītam udakam udaraṃ tundam āsamantād dehe vartamānā dhātava ādhātavaḥ.
+eṣām itaretaradvandvaḥ. teṣu gataḥ prāpto yo doṣo vikāras taṃ viśeṣeṇa nāśayatīti
+nāḍījalodarādhātugatadoṣavināśanam. gacchatā gamanaṃ kurvatā tiṣṭhatā sthitena vāpi puṃsā
+ujjāyyākhyam ujjāyīty ākhyā yasya tat. tu ity anena asya vaiśiṣṭyaṃ dyotayati. kāryaṃ kartavyam.
+udyāyīti kvacit pāṭhaḥ. gacchatā tiṣṭhatā tu bandharahitaḥ kartavyaḥ. kumbhakaśabdas
+triliṅgaḥ. puṃlliṅgapāṭhe tu viśeṣaṇeṣv api puṃlliṅgaḥ pāṭhaḥ kāryaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_054}
+\startsloka
+{\bf atha sītkārī
+sītkāṃ kuryāt tathā vaktre ghrāṇenaiva vijṛmbhikām
+evam abhyāsayogena kāmadevo dvitīyakaḥ (2.54)}
+\stopsloka
+
+sītkārīkumbhakam āha sītkām iti. vaktre mukhe sītkāṃ sīd eva sītkā sīd iti śabdaḥ sītkāras tāṃ
+kuryāt. oṣṭhayor antare saṃlagnayā jihvayā sītkārapūrvakaṃ mukhena pūrakaṃ kuryād ity arthaḥ.
+ghrāṇenaiva nāsikayaivety anenobhābhyāṃ nāsāpuṭābhyāṃ recakaḥ kāryam ity uktam. evaśabdena
+vaktrasya vyavacchedaḥ. vaktreṇa vāyor niḥsāraṇaṃ tv abhyāsānantaram api na kāryam.
+balahānikaratvāt. vijṛmbhikāṃ recakaṃ kuryād ity atrāpi sambadhyate. kumbhakas tv anukto 'pi
+sītkāryāḥ kumbhakatvād evāvagantavyaḥ. atha sītkāryāḥ praśaṃsā. evam uktaprakāreṇābhyāsaḥ
+paunaḥpunyenānuṣṭhānaṃ sa eva yogaḥ yogasādhanatvāt tena dvitīya eva dvitīyakaḥ kāmadevaḥ
+kandarpaḥ. rūpalāvaṇyātiśayena kāmadevasādṛśyāt.
+%\end{vsid}
+
+%\begin{vsid}{hp02_055}
+\startsloka
+{\bf yoginīcakrasaṃmānyaḥ sṛṣṭisaṃhārakārakaḥ
+na kṣudhā na tṛṣā nidrā naivālasyaṃ prajāyate (2.55)}
+\stopsloka
+
+yoginīnāṃ cakraṃ yoginīcakraṃ yoginīsamūhaḥ. tasya samṃmānyaḥ saṃsevyaḥ. sṛṣṭiḥ prapañcotpattiḥ
+saṃhāras tallayaḥ tayoḥ kārakaḥ kartā. kṣudhā bhoktum icchā na. tṛṣā jalapānecchā na. nidrā
+suṣuptir na. ālasyaṃ kāyacittagauravāt pravṛttyabhāvaḥ. kāyagauravaṃ kaphādinā cittagauravaṃ
+tamoguṇena. naiva prajāyate naiva prādurbhavati. evam abhyāsayogeneti prajāyata iti ca prativākyaṃ
+sambadhyate.
+%\end{vsid}
+
+%\begin{vsid}{hp02_056}
+\startsloka
+{\bf bhavet sattvaṃ ca dehasya sarvopadravavarjitaḥ
+anena vidhinā satyaṃ yogīndro bhūmimaṇḍale (2.56)}
+\stopsloka
+
+bhaved iti. dehasya śarīrasya sattvaṃ balaṃ ca bhavet. anenoktena vidhinābhyāsavidhinā yogīndro
+yoginām indra iva yogīndro bhūmimaṇḍale sarvair upadravair varjitaḥ sarvopadravavarjito bhavet
+satyam. sarvaṃ vākyaṃ sāvadhāraṇam iti nyāyāt. yad uktaṃ phalaṃ tat satyam evety arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_058}
+\startsloka
+{\bf atha śītalī
+jihvayā vāyum ākṛṣya pūrvavat kumbhasādhanam
+śanakair ghrāṇarandhrābhyāṃ recayet pavanaṃ sudhīḥ (2.57)}
+\stopsloka
+
+śītalīkumbhakam āha jihvayeti. jihvayā oṣṭhayor bahir nirgatayā vihaṅgamādharacañcusadṛśayā vāyum
+ākṛṣya śanaiḥ pūrakaṃ kṛtvety arthaḥ. pūrvavat sūryabhedanavat kumbhasya kumbhakasya sādhanaṃm
+vidhānaṃ kṛtvety adhyāhāraḥ. sudhīḥ śobhanā dhīr yasya saḥ. ghrāṇasya randhre ghrāṇarandhre tābhyāṃ
+nāsāpuṭavivarābhyāṃ śanakaiḥ śanair eva. avyayasarvanāmnām\comment{Aṣṭādhyāyī 5.3.71.} ity
+akac. pavanaṃ vāyum recayet.
+%\end{vsid}
+
+%\begin{vsid}{hp02_059}
+\startsloka
+{\bf gulmaplīhādikān rogān jvaraṃ pittaṃ kṣudhāṃ tṛṣām
+viṣāṇi śītalī nāma kumbhikeyaṃ nihanti hi (2.58)}
+\stopsloka
+
+śītalīguṇān āha gulmeti. gulmaś ca plīhaś ca gulmaplīhau rogaviśeṣāv ādī yeṣāsāṃ te gulmaplīhādikās
+tān rogān āmayān jvaraṃ jvarākhyaṃ rogam pittaṃ pittavikāram kṣudhāṃ bhoktum icchām tṛṣām
+jalapānecchām viṣāṇi sarpādiviṣajanitavikārān śītalī nāmeti prasiddhārthakam avyayam iyam uktā
+kumbhikā nihanti nitarāṃ hanti. kumbhaśabdaḥ strīliṅgo 'pi. tathā ca śrīharṣaḥ
+
+\startsloka
+udasya kumbhīr athaha śātakumbhajāḥ\comment{Naiṣadhīyacarita 5.19.}
+\stopsloka
+
+iti svārthe kapratyaye kumbhī eva kumbhikā iti.
+%\end{vsid}
+
+%\begin{vsid}{hp02_060}
+\startsloka
+{\bf atha bhastrikā
+ūrvor upari saṃsthāpya śubhe pādatale ubhe
+padmāsanaṃ bhaved etat sarvapāpapraṇāśanam (2.59)}
+\stopsloka
+
+bhastrākumbhakasya padmāsanapūrvakam evānuṣṭhānāt tadādau padmāsanam āha ūrvor iti. upary uttāne
+śubhe śuddhe ubhe dve pādayos tale 'dhaḥpradeśe ūrvoḥ saṃsthāpya samyak sthāpayitvā vaset. etat
+padmāsanaṃ bhavet. kīdṛśam sarveṣāṃ pāpānāṃ prakarṣeṇa nāśanam. atroparīty avyayam
+uttānavācakam. tathā ca kārakeṣu manoramāyāṃ upary upari buddhīnām\comment{Prauḍhamanoramā (KSS),
+  p. 506.} ity atroparibuddhīnām ity asyottānabuddhīnām iti vyākhyānaṃ kṛtam.
+%\end{vsid}
+
+%\begin{vsid}{hp02_061}
+\startsloka
+{\bf samyak padmāsanaṃ baddhvā samagrīvodaraḥ sudhīḥ
+mukhaṃ saṃyamya yatnena prāṇaṃ ghrāṇena recayet (2.60)}
+\stopsloka
+
+bhastrikākumbhakam āha samyag iti. grīvā ca udaraṃ ca grīvodaram. prāṇyaṅgatvād ekavadbhāvaḥ.
+samaṃ grīvodaraṃ yasya sa samagrīvodaraḥ. susthitā dhīr yasya sa sudhīḥ. padmāsanaṃ samyak sthiraṃ
+baddhvā mukhaṃ saṃyamya saṃyataṃ kṛtvā yatnena prayatnena ghrāṇena ghrāṇasyaikatareṇa randhreṇa
+prāṇaṃ śarīrāntaḥsthitaṃ vāyuṃ recayet.  % !! Kumbhakapaddhati Nils
+%\end{vsid}
+
+%\begin{vsid}{hp02_062}
+\startsloka
+{\bf yathā lagati hṛtkaṇṭhe kapālāvadhi sasvanam
+vegena pūrayec cāpi hṛtpadmāvadhi mārutam (2.61)}
+\stopsloka
+
+recakaprakāram āha yatheti. hṛc ca kaṇṭhaś ca hṛtkaṇṭhaṃ tasmin hṛtkaṇṭhe. samāhāradvandvaḥ.
+kapālāvadhi kapālaparyantaṃ svanena sahitaṃ sasvanaṃm yathā syāt tathā yena prakāreṇa lagati. prāṇa
+iti śeṣaḥ. tathā recayet. hṛtpadmam avadhir yasmin karmaṇi tat hṛtpadmāvadhi vegena tarasā
+mārutaṃ vāyuṃ pūrayet. cāpīti pādapūraṇārtham.
+%\end{vsid}
+
+%\begin{vsid}{hp02_063}
+\startsloka
+{\bf punar virecayet tadvat pūrayec ca punaḥ punaḥ
+yathaiva lohakāreṇa bhastrā vegena cālyate (2.62)}
+\stopsloka
+
+punar iti. tadvat pūrvavat punar virecayet punaḥ punaḥ pūrayec cety anvayaḥ. ukte 'rthe dṛṣṭāntam
+āha yathaiveti. lohakāreṇa lohavikārāṇāṃ kartrā bhastrā agner dhamanasādhanībhūtaṃ carma yathaiva
+yenaiva prakāreṇa vegena cālyate.
+%\end{vsid}
+
+%\begin{vsid}{hp02_064}
+\startsloka
+{\bf tathaiva svaśarīrasthaṃ cālayet pavanaṃ dhiyā
+yadā śramo bhaved dehe tadā sūryeṇa pūrayet (2.63)}
+\stopsloka
+
+tathaiveti. tathaiva tenaiva prakāreṇa svaśarīrasthaṃ svaśarīre sthitaṃ pavanaṃ prāṇaṃ dhiyā
+buddhyā cālayet. recakapūrakayor nirantarāvartanena cālanasyāvadhim āha yadā śrama iti. yadā yasmin
+kāle dehe śarīre śramo recakapūrakayor nirantarāvartanenāyāso bhavet tadā tasmin kāle sūryeṇa
+sūryanāḍyā pūrayet.
+%\end{vsid}
+
+%\begin{vsid}{hp02_065}
+\startsloka
+{\bf yathodaraṃ bhavet pūrṇam anilena tathā laghu
+dhārayen nāsikāṃ madhyātarjanībhyāṃ vinā dṛḍham (2.64)}
+\stopsloka
+
+yatheti. yathā yena prakāreṇa pavanena vāyunā laghu kṣipram evodaraṃ pūrṇaṃ bhavet tathā tena
+prakāreṇa sūryanāḍyā pūrayet. laghu kṣipram araṃ drutam ity amaraḥ\comment{Amarakośa 1.1.66.}.
+pūrakānantaraṃ yat kartavyaṃ tad āha dhārayed iti. madhyātarjanībhyāṃ madhyamātarjanībhyāṃ vinā
+aṅguṣṭhānāmikākaniṣṭhikābhir nāsikāṃ dṛḍhaṃ dhārayet. aṅguṣṭhena dakṣiṇanāsāpuṭaṃ
+nirudhyānāmikākaniṣṭhikābhyāṃ vāmanāsāpuṭaṃ nirudhya nāsikāṃ dṛḍhaṃ gṛhṇīyād ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_066,02.067}
+\startsloka
+{\bf vidhivat kumbhakaṃ kṛtvā recayed iḍayānilam
+vātapittaśleṣmaharaṃ śarīrāgnivivardhanam (2.65)
+kuṇḍalībodhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitam
+brahmanāḍīmukhe saṃsthakaphādyargalanāśanam (2.66)}
+\stopsloka
+
+vidhivad iti. vidhivat bandhapūrvakaṃ kumbhakaṃ kṛtvā iḍayā candranāḍyānilaṃ vāyuṃ recayet.
+bhastrākumbhakasyeyaṃ paripāṭī. vāmanāsikāpuṭaṃ dakṣiṇabhujānāmikākaniṣṭhikābhyāṃ nirudhya
+dakṣiṇānāsikāpuṭena bhastrāvad vegena recakapūrakāḥ kāryāḥ. śrame jāte tenaiva nāsāpuṭena pūrakaṃ
+kṛtvāṅguṣṭhena dakṣiṇaṃ nāsikāpuṭaṃ nirudhya yathāśakti kumbhakaṃ dhārayet. paścād iḍayā
+recayet. punar dakṣiṇanāsāpuṭam aṅguṣṭhena nirudhya vāmanāsikāpuṭena bhastrāvaj jhaṭiti
+recakapūrakāḥ kartavyāḥ. śrame jāte tenaiva nāsikāpuṭena pūrakaṃ kṛtvānāmikākaniṣṭhikābhyāṃ
+vāmanāsikāpuṭaṃ nirudhya yathāśakti kumbhakaṃ kṛtvā piṅgalayā recayed ity ekā
+rītiḥ.
+
+vāmanāsikāpuṭam anāmikākaniṣṭhikābhyāṃ nirudhya dakṣiṇanāsikāpuṭena pūrakaṃ kṛtvā jhaṭity
+aṅguṣṭhena dakṣiṇanāsikāpuṭaṃ nirudhya vāmanāsāpuṭena recayed. evaṃ śatadhā kṛtvā śrame jāte
+tenaiva pūrayet. bandhapūrvakaṃ kṛtveḍayā recayet. punar dakṣiṇanāsāpuṭam aṅguṣṭhena nirudhya
+vāmanāsāpuṭena pūrakaṃ kṛtvā jhaṭiti vāmanāsikāpuṭam anāmikākaniṣṭhikābhyāṃ nirudhya piṅgalayā
+recayed bhastrāvat. punaḥ punar evaṃ kṛtvā recakapūrakāvṛttiśrame jāte vāmanāsāpuṭenaiva pūrakaṃ
+kṛtvānāmikākaniṣṭhikābhyāṃ dhṛtvā kumbhakaṃ kṛtvā piṅgalayā recayed iti dvitīyā rītiḥ.
+
+bhastrikāguṇān āha vātapitteti. vātaś ca pittaṃ ca śleṣmā ca vātapittaśleṣmāṇas tān haratīti tat
+tādṛśaṃ śarīre dehe yo 'gnir jaṭharānalas tasya viśeṣeṇa vardhanaṃ dīpanam. kṣipraṃ śīghraṃ
+kuṇḍalyāḥ suptāyā bodhakaṃ bodhakartṛ punātīti pavanaṃ pavitrakārakaṃ sukhaṃ dadātīti sukhadaṃ
+hitaṃ tridoṣaharatvāt sarveṣāṃ hitaṃ sarvadā ca hitam. sarveṣāṃ kumbhakānāṃ sarvadā hitatve 'pi
+sūryabhedanojjāyināv uṣṇau prāyeṇa śīte hitau. sītkārīśītalyau śītale prāyeṇoṣṇe
+hite. bhastrākumbhakaḥ samaśītoṣṇaḥ sarvadā hitaḥ. sarveṣāṃ kumbhakānāṃ sarvarogaharatve 'pi
+sūryabhedanaṃ prāyeṇa vātaharam. ujjāyī prāyeṇa śleṣmaharaḥ. sītkārīśītalyau prāyeṇa
+pittahare. bhastrākhyaḥ kumbhakaḥ tridoṣaharaḥ iti bodhyam. brahmanāḍī suṣumnṇā
+brahmaprāpakatvāt. tathā ca śrutiḥ
+
+\startsloka
+śataṃ caikā ca hṛdayasya nāḍyas 
+tāsāṃ mūrdhānam abhiniḥsṛtaikā
+tayordhvam āyann amṛtatvam eti 
+viṣvaṅṅ anyā utkramaṇe bhavanti\comment{Chāndogyopaniṣat 8.6.6.}
+\stopsloka
+
+tasyā mukhe 'grabhāge saṃsthaḥ samyak sthito yaḥ kaphādirūpo 'rgalaḥ
+prāṇagatipratibandhakas tasya nāśanaṃ nāśakartṛ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_068}
+\startsloka
+{\bf samyaggātrasamudbhūtagranthitrayavibhedakam
+viśeṣeṇaiva kartavyaṃ bhastrākhyaṃ kumbhakaṃ tv idam (2.67)}
+\stopsloka
+
+samyag iti. samyag dṛḍhībhūtaṃ gātre gātramadhye suṣumṇnāyām eva samyag udbhūtaṃ samudbhūtaṃ jātaṃ
+yad granthīnāṃ trayaṃ granthitrayaṃm brahmagranthiviṣṇugranthirudragranthirūpaṃ tasya viśeṣeṇa
+bhedajanakam. ata eva idam bhastrā ity ākhyāyata iti bhastrākhyaṃ kumbhakaṃ tu viśeṣeṇaiva
+kartavyam avaśyakartavyam ity arthaḥ. sūryabhedanādayas tu yathāsambhavaṃ kartavyāḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_069}
+\startsloka
+{\bf atha bhrāmarī
+vegād ghoṣaṃ pūrakaṃ bhṛṅganādaṃ 
+bhṛṅgīnādaṃ recakaṃ mandamandam
+yogīndrāṇām evam abhyāsayogāc
+citte jātā kācid ānandalīlā (2.68)}
+\stopsloka
+
+bhrāmarīkumbhakam āha vegād iti. vegāt tarasā ghoṣaṃ saśabdaṃ yathā syāt tathā bhṛṅgasya
+bhramarasya nāda iva nādo yasmin karmaṇi tat tathā pūrakaṃ kṛtvā bhṛṅgyā bhramaryā nāda
+
+iva nādo yasmin tat tathā mandaṃ mandaṃ recakaṃ kuryāt. pūrakānantaraṃ kumbhakas tu bhrāmaryāḥ
+kumbhakatvād eva siddho 'viśeṣāc ca noktaḥ. pūrakarecakayos tu viśeṣo 'stīti tāv evoktau. evam
+uktarītyābhyasanam abhyāsas tasya yogo yuktis tasmād yogīndrāṇāṃ citte kācid anirvācyā ānande līlā
+krīḍā ānandalīlā jātotpannā bhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp02_070}
+\startsloka
+{\bf atha mūrcchā
+pūrakānte gāḍhataraṃ baddhvā jālandharaṃ śanaiḥ
+recayen mūrcchanākhyeyaṃ manomūrcchā sukhapradā (2.69)}
+\stopsloka
+
+mūrcchākumbhakam āha pūrakānta iti. pūrakasyānte 'vasāne 'tiśayena gāḍhaṃ gāḍhataraṃ dṛḍhataraṃ
+jālandharākhyaṃ bandhaṃ baddhvā śanair mandaṃ mandaṃ recayet. iyaṃ kumbhikā mūrcchanākhyā mano
+mūrcchayatīti mūrcchanā ity ityākhyāyata iti mūrcchanākhyā. kīdṛśī mano mūrcchayati iti
+manomūrcchā. etena mūrcchanāyā vigrahadarśanapūrvakaṃ phalam uktam. punaḥ kīdṛśī sukhapradā
+sukhaṃ pradadātīti sukhapradā.
+%\end{vsid}
+
+%\begin{vsid}{hp02_071}
+\startsloka
+{\bf atha plāvinī
+antaḥ pravartitodāramārutāpūritodaraḥ
+payasy agādhe 'pi sukhāt plavate padmapatravat (2.70)}
+\stopsloka
+
+plāvinīkumbhakam āha antar iti. antaḥ śarīrāntaḥ pravartitaḥ pūrita udāro 'tiśayito yo mārutaḥ
+samīras tena ā samantāt pūritam udaraṃ yena sa pumān agādhe 'py atalasparśe 'pi payasi jale
+padmapatravat padmapatreṇa tulyaṃ sukhād anāyāsāt plavate tarati jalopari gacchati.
+%\end{vsid}
+
+%\begin{vsid}{hp02_072}
+\startsloka
+{\bf prāṇāyāmas tridhā prokto recapūrakakumbhakaiḥ
+sahitaḥ kevalaś ceti kumbhako dvividho mataḥ (2.71)}
+\stopsloka
+
+atha prāṇāyāmabhedān āha prāṇāyāma iti. prāṇasya śarīrāntaḥsañcārivāyor āyamanaṃ nirodhanam āyāmaḥ
+prāṇāyāmaḥ. prāṇāyāmalakṣaṇam uktaṃ gorakṣanāthena
+
+\startsloka
+prāṇaḥ svadehajo vāyur āyāmas tannirodhanam\comment{Gorakṣaśataka 42.}
+\stopsloka
+
+iti. recakaś ca pūrakaś ca kumbhakaś ca tair bhedais tridhā triprakārakaḥ recakaprāṇāyāmaḥ
+pūrakaprāṇāyāmaḥ kumbhakaprāṇāyāmaś ceti. recakalakṣaṇam āha yājñavalkyaḥ
+
+\startsloka
+  bahir yad recanaṃ vāyor udarād recakaḥ smṛtaḥ\comment{Yogayājñavalkya 6.25ab} 
+\stopsloka
+
+iti. recakaprāṇāyāmalakṣaṇam
+
+\startsloka
+niṣkramya nāsāvivarād aśeṣaṃ 
+prāṇaṃ bahiḥ śūnyam ivānilena
+nirucchvasaṃs tiṣṭhati ruddhavāyuḥ 
+sa recako nāma mahānirodhaḥ\comment{Source unknown.}
+\stopsloka
+
+pūrakalakṣaṇam
+
+\startsloka
+bāhyād āpūraṇaṃ vāyor udare pūrako hi saḥ\comment{Yogayājñavalkya 6.24ab}
+\stopsloka
+
+iti. pūrakaprāṇāyāmalakṣaṇam
+
+\startsloka
+bāhye sthitaṃ prāṇapuṭena vāyum 
+ākṛṣya tenaiva śanaiḥ samantāt
+nāḍīś ca sarvāḥ paripūrayed yaḥ 
+sa pūrako nāma mahānirodhaḥ\comment{Source unknown.}
+\stopsloka
+
+kumbhakalakṣaṇaṃ
+
+\startsloka
+  sampūrya kumbhavad vāyor dhāraṇaṃ kumbhako bhavet\comment{Yogayājñavalkya 6.24cd}
+\stopsloka
+
+ayaṃ kumbhakas tu pūrakaprāṇāyāmād abhinnaḥ. bhinnas tu
+
+\startsloka
+  na recako naiva ca pūrako 'tra
+  nāsāpuṭāntasthitam eva vāyum
+  suniścalaṃ dhārayate krameṇa
+  kumbhākhyam etat pravadanti tajjñāḥ
+\stopsloka
+
+atha prakārāntareṇa prāṇāyāmaṃ vibhajate sahita iti. kumbhako dvividhaḥ sahitaḥ kevalaś ceti.  mato
+'bhimato yoginām iti śeṣaḥ. tatra sahito dvividhaḥ recakapūrvakaḥ pūrakapūrvakaś ca. tad
+uktaṃ
+
+%\begin{vsid}{hp02_073}
+\startsloka     
+  ārecyāpūrya vā kuryāt sa vai sahitakumbhakaḥ\comment{This line was originally part of the
+  HP and was complemented by 2.72ab (yāvat kevalasiddhiḥ ...)} 
+\stopsloka
+%\end{vsid}
+
+tatra recakapūrvako recakaprāṇāyāmād abhinnaḥ. pūrakapūrvakaḥ kumbhakaḥ pūrakaprāṇāyāmād
+abhinnaḥ. kevalakumbhakaḥ kumbhaka\-prāṇā\-yāmād abhinnaḥ. prāg uktāḥ sūryabhedanādayaḥ
+pūraka\-pūrvakasya kumbhakasya bhedā jñātavyāḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp02_073-074}
+\startsloka
+{\bf yāvat kevalasiddhiḥ syāt sahitaṃ tāvad abhyaset
+recakaṃ pūrakaṃ muktvā sukhaṃ yad vāyudhāraṇam (2.72)}
+\stopsloka
+
+sahitakumbhakābhyāsasyāvadhim āha yāvad iti. kevalasya kevalakumbha\-kasya siddhiḥ kevalasiddhir
+yāvat yāvatparyantaṃ syāt tāvat tāvatparyantaṃ sahitaṃ sahitakumbhakaṃ sūryabhedādikam abhyased
+anutiṣṭhet. suṣumṇnābhedānantaraṃ yadā suṣumṇnāntaḥ ghaṭaśabdā\var{ghaṭa \Tue \EdLo etc. \lem ṣaṭ \Wthree
+  etc. (neither reading seems interpretable)} bhavanti tadā kevalakumbhakaḥ siddhyati. tadanantaraṃ
+sahitakumbhakā daśa viṃśatir vā kāryāḥ. aśītisaṅkhyāpūrtiḥ kevalakumbhakair eva kartavyā. sati
+sāmarthye kevalakumbhakā aśīter adhikāḥ kāryāḥ.  kevalakumbhakasya lakṣaṇam āha recakam
+iti. recakaṃ pūrakaṃ muktvā tyaktvā sukham anāyāsaṃ yathā syāt tathā vāyor dhāraṇaṃm vāyudhāraṇaṃm
+yat.
+%\end{vsid}
+
+%\begin{vsid}{hp02_074-075}
+\startsloka
+{\bf prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ
+kumbhake kevale siddhe recapūrakavarjite (2.73)}
+\stopsloka
+
+prāṇāyāma iti. sa vai iti niścitaṃ. kevalakumbhakaḥ prāṇāyāmaḥ ity ayam uktaḥ. kevalaṃ
+praśaṃsanti kevala iti. reco recakaḥ. recaś ca pūrakaś ceti recapūrakau tābhyāṃ varjite rahite
+kevale kumbhake siddhe sati.
+%\end{vsid}
+
+%\begin{vsid}{hp02_075-076}
+\startsloka
+{\bf na tasya durlabhaṃ kiñcit triṣu lokeṣu vidyate
+śaktaḥ kevalakumbhena yatheṣṭaṃ vāyudhāraṇāt (2.74)}
+\stopsloka
+
+neti. tasya yoginas triṣu lokeṣu durlabhaṃ duṣprāpaṃ kim api kiñcit kim api na vidyate. tasya
+sarvaṃ sulabham ity arthaḥ. śakta iti kevalakumbhakena kevalakumbhakābhyāsena śaktaḥ samartho
+yatheṣṭaṃ yathecchaṃ vāyor dhāraṇaṃ vāyudhāraṇaṃ tasmād vāyudhāraṇāt.
+%\end{vsid}
+
+%\begin{vsid}{hp02_076,02.079}
+\startsloka
+{\bf rājayogapadaṃ cāpi labhate nātra saṃśayaḥ
+kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet
+anargalā suṣumṇā ca haṭhasiddhiś ca jāyate (2.75)}
+\stopsloka
+
+rājeti. rājayogapadaṃ rājayogātmakaṃ padaṃ labhate. atra saṃśayo na. niścitam etad ity arthaḥ.
+kumbhakābhyāsasya paramparayā kaivalyahetumatvam āha kumbhakād iti. kumbhakāt kumbhakābhyāsāt
+kuṇḍalī ādhāraśaktis tasyā bodho nidrābhaṅgo bhavet. kuṇḍalyā bodhaḥ kuṇḍalībodhas tasmāt
+kuṇḍalībodhataḥ anargaleti suṣumṇnā madhyanāḍī anargalā kaphādyargalarahitā bhavet. haṭhasya
+haṭhābhyāsasya siddhiḥ pratyāhārādiparamparayā kaivalyarūpā siddhir jāyate.
+%\end{vsid}
+
+%\begin{vsid}{hp02_077}
+\startsloka
+{\bf haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ
+na sidhyati tato yugmam āniṣpatteḥ samabhyaset (2.76)}
+\stopsloka
+
+haṭhayogarājayogasādhanayoḥ parasparopakāryopakārakatvam āha haṭhaṃ vineti. haṭhaṃ haṭhayogaṃ vinā
+rājayogo na siddhyati rājayogaṃ vinā haṭho na siddhyati. tataḥ yato 'nyatarasya siddhir nāsti
+tasmān niṣpattiṃ rājayogasiddhim ā maryādīkṛtya yā niṣpattis tasyā rājayogasiddhiparyantaṃ yugmaṃ
+haṭhayogarājayogadvayam abhyased anutiṣṭhet. haṭhātirikte sākṣāt paramparayā vā rājayogasādhane
+'tra rājayogaśabdaḥ jīvanasādhane lāṅgale jīvanaśabdaprayogavat. rājayogasādhanaṃ ca caturthopadeśe
+vakṣyamāṇam unmanīśāmbhavīmudrādirūpam aparokṣānubhūtāv\comment{REF?} uktaṃ pañcadaśāṅgarūpaṃ
+ca. vākyasudhāyām\comment{REF?} uktaṃ dṛśyānuviddhādirūpaṃ ca.
+%\end{vsid}
+
+%\begin{vsid}{hp02_078}
+\startsloka
+{\bf kumbhakaprāṇarodhānte kuryāc cittaṃ nirāśrayam
+evam abhyāsayogena rājayogapadaṃ vrajet (2.77)}
+\stopsloka
+
+haṭhābhyāsād rājayogaprāptiprakāram āha kumbhaketi. kumbhakena prāṇasya yo rodhas tasyānte madhye
+cittam antaḥkaraṇaṃ nirāśrayaṃ kuryāt. samprajñātasamādhau jātāyāṃ brahmākāravṛtteḥ paravairāgyeṇa
+vilayaṃ kuryād ity arthaḥ. evam uktarītyābhyāsasya yogo yuktis tena. yogaḥ
+saṃnahanopāyadhyānasaṅgatiyuktiṣu iti kośaḥ\comment{Amarakośa 3.3.22?}. rājayogapadaṃ
+rājayogātmakaṃ padaṃ vrajet prāpnuyāt.
+%\end{vsid}
+
+%\begin{vsid}{hp02_080}
+\startsloka
+{\bf vapuḥkṛśatvaṃ vadane prasannatā
+nādasphuṭatvaṃ nayane sunirmale
+arogatā bindujayo 'gnidīpanaṃ
+nāḍīviśuddhir haṭhasiddhilakṣaṇam (2.78)}
+\stopsloka
+
+haṭhasiddhijñāpakam āha vapuḥkṛśatvam iti. vapuṣo dehasya kṛśatvaṃ kārśyaṃ. vadane mukhe
+prasannatā prasādaḥ. nādasya dhvaneḥ sphuṭatvaṃ prākaṭyaṃm. nayane netre suṣṭhu nirmale.
+arogasya bhāvo 'rogatā ārogyam. bindor dhātor jayaḥ kṣayābhāvarūpaḥ. agner udaryasya dīpanaṃ
+dīptiḥ. nāḍīnāṃ viśeṣeṇa śuddhir malāpagamaḥ. etad dhaṭhasya haṭhābhyāsasiddher bhāvinyā
+lakṣyate 'neneti lakṣaṇaṃ jñāpakam.
+%\end{vsid}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+
+%\begin{vsid}{hp03_001}
+\startsloka
+{\bf saśailavanadhātrīṇāṃ yathādhāro 'hināyakaḥ
+sarveṣāṃ yogatantrāṇāṃ tathādhāro hi kuṇḍalī (3.1)}
+\stopsloka
+
+atha kuṇḍalyāḥ sarvayogāśrayatvam āha saśaileti. śailāś ca vanāni ca śailavanāni taiḥ saha
+vartamānāḥ saśailavanāḥ tāś ca tā dhātryaś ca bhūmayas tāsām. dhātryā ekatve 'pi deśabhedād bhedam
+ādāya bahuvacanam. ahīnāṃ sarpāṇāṃ nāyako netāhināyakaḥ śeṣo yathā yadvad ādhāra āśrayas tathā
+tadvat. sarveṣāṃ yogasya tantrāṇi yogatantrāṇi yogopāyās teṣām. kuṇḍālyādhāraśaktir
+āśrayaḥ. kuṇḍalībodhaṃ vinā sarvayogopāyānāṃ vaiyarthyād iti bhāvaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_002}
+\startsloka
+{\bf suptā guruprasādena yadā jāgarti kuṇḍalī
+tadā sarvāṇi padmāni bhidyante granthayo 'pi ca (3.2)}
+\stopsloka
+
+kuṇḍalībodhasya phalam āha dvābhyāṃ supteti. suptā kuṇḍalī guroḥ prasādena yadā jāgarti budhyate
+tadā sarvāṇi padmāni ṣaṭcakrāṇi bhidyante bhinnāni bhavanti. granthayo 'pi ca
+brahmagranthiviṣṇugranthirudragranthayo bhidyante bhedaṃ prāpnuvantīty anvayaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_003}
+\startsloka
+{\bf prāṇasya śūnyapadavī tadā rājapathāyate
+tadā cittaṃ nirālambaṃ tadā kālasya vañcanam (3.3)}
+\stopsloka
+
+prāṇasyeti. tadā śūnyapadavī suṣumnā prāṇasya vāyo rājñāṃ panthā rājapathaḥ rājapatha ivācarati
+rājapathāyate rājamārgāyate. sukhena gamanasambhavāt. tadā cittam ālambanam āśrayas tasmān nirgataṃ
+nirālambaṃ nirviṣayaṃ bhavati. tadā kālasya mṛtyor vañcanaṃ pratāraṇaṃ bhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_004}
+\startsloka
+{\bf suṣumṇā śūnyapadavī brahmarandhraṃ mahāpathaḥ
+śmaśānaṃ śāmbhavī madhyamārgaś cety ekavācakāḥ (3.4)}
+\stopsloka
+
+suṣumnāparyāyān āha suṣumneti. ity uktāḥ śabdā ekasya ekārthasya vācakāḥ ekavācakāḥ. paryāyā ity
+arthaḥ. spaṣṭaḥ ślokārthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_005}
+\startsloka
+{\bf tasmāt sarvaprayatnena prabodhayitum īśvarīm
+brahmadvāramukhe suptāṃ mudrābhyāsaṃ samācaret (3.5)}
+\stopsloka
+
+tasmād iti. yasmāt kuṇḍalībodhanenaiva ṣaṭcakrabhedādikaṃ bhavati tasmāt sarvaprayatnena sarveṇa
+prayatnena brahma saccidānandalakṣaṇaṃ tasya dvāraṃ prāptyupāyaḥ suṣumnā tasya mukhe 'grabhāge
+mukhena suṣumnādvāraṃ pidhāya suptām īśvarīṃ kuṇḍalīṃ prabodhayituṃ prakarṣeṇa bodhayituṃ mudrāṇāṃ
+mahāmudrādīnām abhyāsam āvṛttiṃ samācaret samyag ācaret.
+%\end{vsid}
+
+%\begin{vsid}{hp03_006,03.007}
+\startsloka
+{\bf mahāmudrā mahābandho mahāvedhaś ca khecarī
+uḍḍyānaṃ mūlabandhaś ca bandho jālandharābhidhaḥ (3.6)
+karaṇī viparītākhyā vajrolī śakticālanam
+idaṃ hi mudrādaśakaṃ jarāmaraṇanāśanam (3.7)}
+\stopsloka
+
+mudrā uddiśati mahāmudretyādinā sārdhena. sārdhārthaḥ spaṣṭaḥ. mudrāphalam āha sārdhadvābhyām idam
+iti. idam uktaṃ mudrāṇāṃ daśakaṃ jarā ca maraṇaṃ ca jarāmaraṇe tayor nāśanam nivārakam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_08}
+\startsloka
+{\bf ādināthoditaṃ divyam aṣṭaiśvaryapradāyakam
+vallabhaṃ sarvasiddhānāṃ durlabhaṃ marutām api (3.8)}
+\stopsloka
+
+ādināthena śambhunoditaṃ kathitam. divi bhavaṃ divyam uttamam. aṣṭau ca tāny aiśvaryāṇi
+cāṣṭaiśvaryāṇi aṇimamahimagarimalaghimaprāptiprākāmyeśatāvaśitākhyāni. tatrāṇimā saṅkalpamātreṇa
+prakṛtyapagame paramāṇuvad dehasya sūkṣmatā. mahimā prakṛtyāpūreṇā\-kāśādivan mahadbhāvaḥ. garimā
+laghutarasyāpi tūlādeḥ parvatādivad gurubhāvaḥ. laghimā gurutarasyāpi parvatādes tūlādival
+laghubhāvaḥ. prāptiḥ sarvabhāvasānnidhyam. yathā bhūmistha evāṅgulyagreṇa spṛśati
+candramasam. prākāmyam icchānabhighātaḥ. yathā udaka iva bhūmau nimajjaty unmajjati ca. īśatā
+bhūtabhautikānāṃ prabhavāpyayasaṃsthānaviśeṣasāmarthyam. vaśitvaṃ bhūtabhautikānāṃ
+svādhī\-nī\-karaṇam. teṣāṃ pradāyakaṃ prakarṣeṇa dadātīti tathā taṃ sarve ca te siddhāś ca kapilādayas
+teṣāṃ vallabhaṃ priyaṃ marutāṃ devānām api durlabhaṃ duṣprāpaṃ kim utānyeṣām ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_09}
+\startsloka
+{\bf gopanīyaṃ prayatnena yathā ratnakaraṇḍakam
+kasyacin naiva vaktavyaṃ kulastrīsurataṃ yathā (3.9)}
+\stopsloka
+
+gopanīyam iti. prayatnena prakṛṣṭena yatnena gopanīyaṃ gopanārham. gopane dṛṣṭāntam
+āha yatheti. ratnānāṃ hīrakādīnāṃ karaṇḍakaṃ ratnakaraṇḍakaṃ yathā yena prakāreṇa gopyate
+tadvat. kasyāpi janamātrasya yad vā kasyāpi brahmaṇo 'pi naiva vaktavyaṃ naiva vācyaṃ kim
+utānyasya. tatra dṛṣṭāntaḥ kulastriyaḥ surataṃ kulastrīsurataṃ saṅgamanaṃ yathā yadvat.
+%\end{vsid}
+
+%\begin{vsid}{hp03_10}
+\startsloka
+{\bf atha mahāmudrā
+pādamūlena vāmena yoniṃ sampīḍya dakṣiṇam
+prasāritaṃ padaṃ kṛtvā karābhyāṃ dhārayed dṛḍham (3.10)}
+\stopsloka
+
+mudrāsu\var{mudrāsu \Wthree \EdAd \lem mudrādiṣu \EdLo daśavidhamudrādiṣu \EdMu} prathamoddiṣṭatvena
+mahāmudrāṃ tāvad āha pādamūleneti. vāmena savyena pādasya mūlaṃ pādamūlaṃ pārṣṇis tena pādamūlena
+vāmapādapārṣṇinety arthaḥ. yoniṃ yonisthānaṃ gudameḍhrayor madhyabhāgaṃ
+sampīḍyākuñcitavāmapādapārṣṇinā yonisthānaṃ dṛḍhaṃ saṃyojyety arthaḥ. dakṣiṇaṃ savyetaraṃ padaṃ
+caraṇaṃ prasāritaṃ bhūmisaṃlagnapārṣṇikam ūrdhvāṅgulikaṃ daṇḍavat kṛtvā karābhyāṃ sampradāyād
+ākuñcitakaratarjanībhyāṃ dṛḍhaṃ gāḍhaṃ dhārayed aṅguṣṭhapradeśe gṛhṇīyāt.
+%\end{vsid}
+
+%\begin{vsid}{hp03_11}
+\startsloka
+{\bf kaṇṭhe bandhaṃ samāropya dhārayed vāyum ūrdhvataḥ
+yathā daṇḍahataḥ sarpo daṇḍākāraḥ prajāyate (3.11)}
+\stopsloka
+
+kaṇṭha iti. kaṇṭhe kaṇṭhadeśe bandhaṃ bandhanaṃ samyag āropya kṛtvā. jālandharabandhaṃ kṛtvety
+arthaḥ. vāyuṃ pavanam ūrdhvata upari suṣumnāyāṃ dhārayet. anena mūlabandhaḥ sūcitaḥ. sa tu
+yonisampīḍanena jihvābandhena ca caritārtha iti sāmpradāyikāḥ. yathā daṇḍena hatas tāḍito
+daṇḍahataḥ sarpaḥ kuṇḍalī daṇḍākāraḥ daṇḍasyākāra ivākāro yasya sa tādṛśaḥ. kuṇḍalākāraṃ tyaktvā
+sarala ity arthaḥ. prakarṣeṇa jāyate bhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_12}
+\startsloka
+{\bf ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahasā bhavet
+tadā sā maraṇāvasthā jāyate dvipuṭāśrayā (3.12)}
+\stopsloka
+
+ṛjvīti. tathā kuṇḍaly ādhāraśaktiḥ sahasā śīghram eva ṛjvī sampadyate tathābhūtā ṛjvībhūtā saralā
+bhavet. tadā seti. dve puṭe iḍāpiṅgale āśrayo yasyāḥ sā maraṇāvasthā jāyate. kuṇḍalībodhe sati
+suṣumnāyāṃ praviṣṭe prāṇe dvayoḥ puṭayoḥ prāṇaviyogāt.
+%\end{vsid}
+
+%\begin{vsid}{hp03_13}
+\startsloka
+{\bf tataḥ śanaiḥ śanair eva recayen na tu vegataḥ
+iyaṃ khalu mahāmudrā mahāsiddhaiḥ pradarśitā (3.13)}
+\stopsloka
+
+tatas tadanantaraṃ śanaiḥ śanair eva recayet. vāyum iti sambadhyate. vegatas tu vegān na
+recayet. vegato recane balahāniprasaṅgāt. khalv iti vākyālaṅkāre. iyaṃ mahāmudrā mahāsiddhair
+ādināthādibhiḥ pradarśitā prakarṣeṇa darśitā.
+%\end{vsid}
+
+%\begin{vsid}{hp03_14}
+\startsloka
+{\bf mahākleśādayo doṣāḥ kṣīyante maraṇādayaḥ
+mahāmudrāṃ ca tenaiva vadanti vibudhottamāḥ (3.14)}
+\stopsloka
+
+mahāmudrāyā anvarthatām āha mahākleśeti. mahāntaś ca te kleśāś ca mahākleśā
+avidyāsmitārāgadveṣābhiniveśāḥ pañca. ta ādayo yeṣāṃ tatkāryāṇāṃ śokamohādīnāṃ te doṣāḥ
+kṣīyante. maraṇam ādir yeṣāṃ jarādīnāṃ te 'pi ca kṣīyante naśyanti. yatas tenaiva hetunā viśiṣṭā
+budhā vibudhās teṣūttamā vibudhottamā mahāmudrāṃ vadanti. mahākleśān maraṇādīṃś ca doṣāṇ mudrayati
+śamayatīti mahāmudreti vyutpatter ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_15}
+\startsloka
+{\bf candrāṅge tu samabhyasya sūryāṅge punar abhyaset
+yāvat tulyā bhavet saṅkhyā tato mudrāṃ visarjayet (3.15)}
+\stopsloka
+
+mahāmudrābhyāsakramam āha candrāṅga iti. candreṇa candranāḍyopalakṣitam aṅgaṃ candrāṅgaṃ tasmin
+candrāṅge vāmāṅge. tuśabdaḥ pādapūraṇe. samyag abhyasya sūryeṇa piṅgalayopalakṣitam aṅgaṃ sūryāṅgaṃ
+tasmin sūryāṅge dakṣāṅge punar vāmāṅgābhyāsānantaraṃ yāvad yāvatkālaparyantaṃ tulyā vāmāṅge
+kumbhakakābhyāsa\-saṅkhyā\-samā saṅkhyā bhavet tāvad abhyaset. tataḥ saṅkhyāsāmyānantaraṃ mudrāṃ
+mahāmudrāṃ visarjayet. atrāyaṃ kramaḥ. ākuñcitavāma\-pāda\-pārṣṇiṃ yonisthāne saṃyojya
+prasāritadakṣiṇapādāṅguṣṭham ākuñicitatarjanībhyāṃ gṛhītvābhyāso vāmāṅge 'bhyāsaḥ. asminn abhyāse
+pūrito vāyur vāmāṅge tiṣṭhati. ākuñcitadakṣapādapārṣṇiṃ yonisthāne saṃyojya
+prasāritavāmapādāṅguṣṭham ākuñcitatarjanībhyāṃ gṛhītvābhyāso dakṣāṅge 'bhyāsaḥ. asminn abhyāse
+pūrito vāyur dakṣāṅge tiṣṭhati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf na hi pathyam apathyaṃ vā rasāḥ sarve 'pi nīrasāḥ
+api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam api jīryati (3.16)}
+\stopsloka
+
+mahāmudrāguṇān āha na hīti. hi yasmād mahāmudrābhyāsina ity adhyāhāraḥ. pathyam apathyaṃ vā
+na. pathyāpathyavicāro nāstīty arthaḥ. tasmāt sarve 'pi bhuktā rasāḥ kaṭvamlādayo jīryanta ti
+vibhaktivipariṇāmenānvayah.\var{vibhakti cett. \lem vacana \EdAd} nīrasā nirgato raso yebhyas te
+yātayāmāḥ padārthā jīryanti. ghoram iti durjaraṃ bhuktam annaṃ viṣaṃ kṣveḍam api pīyūṣam ivāmṛtam
+iva jīryati jīrṇaṃ bhavati. kim utānyad iti bhāvaḥ.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf kṣayakuṣṭhagudāvartagulmājīrṇapurogamāḥ
+tasya doṣāḥ kṣayaṃ yānti mahāmudrāṃ tu yo 'bhyaset (3.17)}
+\stopsloka
+
+yaḥ pumān mahāmudrām abhyaset tasya kṣayo rājarogaḥ kuṣṭhagudāvartagulmā rogaviśeṣāḥ. ajīrṇaṃ
+bhuktānnāparipākas tāni purogamāny agresarāṇi yeṣāṃ mahodarajvarādīnāṃ te tathā tādṛśā doṣā
+doṣajanitā rogāḥ kṣayaṃ nāśaṃ yānti prāpnuvanti.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf kathiteyaṃ mahāmudrā mahāsiddhikarī nṝṇām
+gopanīyā prayatnena na deyā yasya kasyacit (3.18)}
+\stopsloka
+
+mahāmudrām upasaṃharan tasyā gopyatvam āha kathiteti. iyam eṣā mahāmudrā kathitoktā. mayā iti
+śeṣaḥ. kīdṛśī nṝṇām abhyasyatāṃ narāṇāṃ mahatyaś ca tāḥ siddhayaś cāṇimādyās tāsāṃ karī
+kartrīyam. prakṛṣṭo yatnaḥ prayatnas tena prayatnena gopanīyā gopanārhā yasya kasyacid yasya
+kasyāpy anadhikāriṇaḥ sambandhasāmānye ṣaṣṭhī. na deyā dātuṃ yogyā na bhavatīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf atha mahābandhaḥ
+pārṣṇiṃ vāmasya pādasya yonisthāne niyojayet 
+vāmorūpari saṃsthāpya dakṣiṇaṃ caraṇaṃ tathā (3.19)}
+\stopsloka
+
+mahābandham āha pārṣṇim iti. vāmasya savyasya pādasya caraṇasya pārṣṇiṃ gulphayor adhobhāgam.
+
+\startsloka
+  tad granthī ghuṭike gulphau pumān pārṣṇis tayor adhaḥ\comment{Amarakośa 2.6.72}
+\stopsloka
+
+ity amaraḥ. yonisthāne gudameṇḍhrayor antarāle niyojayen nitarāṃ yojayet. vāmaḥ savyo ya ūrus
+tasyopari dakṣiṇaṃ caraṇaṃ pādaṃ saṃsthāpya samyak sthāpayitvā. tathāśabdaḥ pādapūraṇe.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf pūrayitvā tato vāyuṃ hṛdaye cibukaṃ dṛḍham 
+niṣpīḍya yonim ākuñcya mano madhye niyojayet (3.20)}
+\stopsloka
+
+pūrayitveti. tatas tadanantaraṃ vāyuṃ pūrayitvā hṛdaye cibukaṃ dṛḍham niṣpīḍya gāḍhaṃ
+saṃsthāpya. etena jālandharabandhaḥ proktaḥ. yoniṃ gudameṇḍhrayor antarālam ākuñcya. anena
+mūlabandhaḥ sūcitaḥ. sa tu jihvābandhena gatārthatvān na kartavyaḥ. manaḥ svāntaṃ madhye
+madhyanāḍyāṃ niyojayet pravartayet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf dhārayitvā yathāśakti recayed anilaṃ śanaiḥ
+savyāṅge tu samabhyasya dakṣāṅge punar abhyaset (3.21)}
+\stopsloka
+
+dhārayitveti. śaktim anatikramya yathāśakti. dhārayitvā kumbhayitvā. śanair mandaṃ mandam anilaṃ
+vāyuṃ recayet. savyāṅge vāmāṅge samabhyasya samyag āvartya dakṣāṅge dakṣiṇāṅge punar yāvat tulyā
+bhavet saṅkhyā tāvad abhyaset.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf matam atra tu keṣāñcit kaṇṭhabandhaṃ vivarjayet
+rājadantasthajihvāyāṃ bandhaḥ śasto bhaved iti (3.22)} 
+\stopsloka
+
+atha jālandharabandhe kaṇṭhasaṅkocanasyānupayogam āha matam iti. keṣāñcit tv ācāryāṇām idaṃ
+matam. kiṃ tat ity āha atra jālandharabandhe kaṇṭhabandhaṃ\var{kaṇṭhabandhaṃ \Wthree}{kaṇṭhabandhe \EdLo}
+kaṇṭhasya bandhanaṃ bandhaḥ saṅkocas taṃ vivarjayet viśeṣeṇa varjayet. kutaḥ yato dantānāṃ rājāno
+rājadantā rājadanteṣu tiṣṭhatīti rājadantasthāḥ rājadantasthā cāsau jihvā ca tasyāṃ
+rājadantasthajihvāyāṃ bandhaḥ taduparibhāgasya sambandhaḥ śastaḥ. kaṇṭhākuñcanāpekṣayā praśasto
+bhaved iti hetoḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf ayaṃ tu sarvanāḍīnām ūrdhvaṃgatinirodhakaḥ\var{ūrdhvaṃ \Tue \EdMu \lem ūrdhva \Wthree \EdLo (unmetrical)}
+ayaṃ khalu mahābandho mahāsiddhipradāyakaḥ (3.23)}
+\stopsloka
+
+ayaṃ tv iti. ayaṃ tu rājadantasthajihvāyāṃ bandhas tu sarvāś ca tā nāḍyaś ca sarvanāḍyo
+dvāsaptatisahasrasaṅkhyākās tāsāṃ suṣumnātiriktānām ūrdhvam upari vāyor gatir ūrdhvagatiḥ tasyā
+nirodhakaḥ pratibandhakaḥ. etena
+
+\startsloka
+badhnāti hi sirājālam adhogāmi nabhojalam\comment{HP 3.71}
+\stopsloka
+
+iti jālandharoktaṃ phalam anenaiva siddham iti sūcitam. mahābandhasya phalam āha ayaṃ khalv
+iti. ayam uktaḥ khalu prasiddhaḥ mahābandho mahāsiddhīḥ prakarṣeṇa dadātīti tathā.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf kālapāśamahābandhavimocanavicakṣaṇaḥ
+triveṇīsaṅgamaṃ dhatte kedāraṃ prāpayen manaḥ (3.24)}
+\stopsloka
+
+kālasya mṛtyoḥ pāśo vāgurā tena yo mahābandho bandhanaṃ tasya viśeṣeṇa mocane mokṣaṇe vicakṣaṇaḥ
+pravīṇaḥ. tisṛṇāṃ nadīnāṃ veṇī samudāyaḥ sa eva saṅgamaḥ prayāgas taṃ dhatte vidhatte. kedāraṃ
+bhruvor madhye śivasthānaṃ kedāraśabdavācyaṃ taṃ manaḥ svāntaṃ
+prāpayet. gatibuddhītyādinā\comment{Aṣṭādhyāyī 1.4.52.} aṇau kartur manaso ṇau karmatvam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf rūpalāvaṇyasampannā yathā strī puruṣaṃ vinā 
+mahāmudrāmahābandhau niṣphalau vedhavarjitau (3.25)}
+\stopsloka
+
+mahāvedham vaktukāmas tasyotkarṣaṃ tāvad āha rūpeti. rūpaṃ saundaryaṃ cakṣuḥpriyo guṇaḥ. lāvaṇyaṃ
+kāntiviśeṣaḥ. tad uktaṃ
+
+\startsloka
+  muktāphaleṣu chāyāyās taralatvam ivāntarā
+  pratibhāti yad aṅgeṣu lāvaṇyaṃ tad ihocyate (3.26)\comment{This definition of
+    {\em lāvaṇya} is regularly quoted, as for instance,
+    in {\em Rasaratnasamuccayabodhinī} (4.15)} 
+\stopsloka
+
+tābhyāṃ sampannā viśiṣṭā strī yuvatī puruṣaṃ bhartāraṃ vinā yathā yādṛśī niṣphalā tathā mahāmudrā
+ca mahābandhaś ca tau vedhena mahāvedhena vināpi. pratyayapūrvottarapadayor lopo vaktavyaḥ iti
+bhāṣyakārokter\comment{Quotation not traced. \EdMu reads {\em vināpi} as part of the quotation.}
+mahacchabdasya lopaḥ. varjitau rahitau niṣphalau vyarthāv ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+  {\bf atha mahāvedhaḥ
+    mahābandhasthito yogī kṛtvā pūrakam ekadhīḥ
+vāyūnāṃ gatim āvṛtya nibhṛtaṃ kaṇṭhamudrayā (3.26)}
+\stopsloka
+
+mahāvedham āha mahābandheti. mahābandhe mahābandhamudrāyāṃ sthito mahābandhasthitaḥ. ekā ekāgrā
+dhīr yasya sa ekadhīḥ. yogī yogābhyāsī pūrakaṃ nāsāpuṭābhyāṃ vāyor grahaṇaṃ kṛtvā kaṇṭhe mudrā
+kaṇṭhamudrā tayā jālandharamudrayā vāyūnāṃ prāṇādīnāṃ gatim ūrdhvādhogamanādirūpāṃ nibhṛtaṃ
+niścalaṃ yathā bhavati tathāvṛtya nirudhya kumbhakaṃ kṛtvety arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf samahastayugo bhūmau sphicau santāḍayec chanaiḥ 
+puṭadvayam atikramya vāyuḥ sphurati madhyagaḥ (3.27)}
+\stopsloka
+
+samahasteti. bhūmau bhuvi hastayor yugaṃ hastayugaṃ samaṃ hastayugaṃ yasya sa samahastayugaḥ
+bhūmisaṃlagnatalau saralau hastau yasya tādṛśaḥ sann ity arthaḥ. sphicau kaṭiprothau. striyām
+sphicau kaṭiprothau ity amaraḥ.\comment{Amarakośa ?} bhūmisaṃlagnatalayor hastayor avalambanena
+yonisthānasaṃlagnapārṣṇinā vāmapādena saha bhūmeḥ kiñcid utthāpitau śanair mandaṃ santāḍayet samyak
+tāḍayet bhūmāv eva. puṭayor dvayam iḍāpiṅgalayor yugmam atikramyollaṅghya madhye suṣumnāmadhye
+gacchatīti madhyago vāyuḥ sphurati.  %  Misprints \EdLo avalambena and vāyu 
+%\end{vsid}
+
+%\begin{vsid}{hp03_028}
+\startsloka
+{\bf somasūryāgnisambandho jāyate cāmṛtāya vai
+mṛtāvasthā samutpannā tato vāyuṃ virecayet (3.28)}
+\stopsloka
+
+someti. somaś ca sūryaś cāgniś ca somasūryāgnayaḥ somasūryāgniśabdais tadadhiṣṭhitā nāḍya
+iḍāpiṅgalāsuṣumnā grāhyās teṣāṃ sambandhaḥ. tadvāyusambandhāt teṣāṃ sambandhaḥ. amṛtāya mokṣāya
+jāyate. vai iti niścaye 'vyayam. mṛtasya prāṇaviyuktasyāvasthā mṛtāvasthā samutpannā bhavati
+iḍāpiṅgalayoḥ prāṇasañcārābhāvāt. tatas tadanantaraṃ vāyuṃ virecayen nāsikāpuṭābhyāṃ śanais tyajet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_029}
+\startsloka
+{\bf mahāvedho 'yam abhyāsān mahāsiddhipradāyakaḥ
+valīpalitavepaghnaḥ sevyate sādhakottamaiḥ (3.29)}
+\stopsloka
+
+mahāvedha iti. ayaṃ mahāvedho 'bhyāsāt punaḥ punar āvartanāt mahāsiddhayo 'ṇimādyās tāsāṃ
+pradāyakaḥ prakarṣeṇa samarpakaḥ. valī jarayā dehe carmasaṅkocaḥ palitaṃ jarasā keśeṣu śauklyaṃ
+vepaḥ kampaḥ tān hantīti valīpalitavepaghnaḥ. ata eva sādhakeṣv abhyāsiṣūttamāḥ sādhakottamās taiḥ
+sevyate 'bhyasyata ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_030}
+\startsloka
+{\bf etat trayaṃ mahāguhyaṃ jarāmṛtyuvināśanam
+vahnivṛddhikaraṃ caiva hy aṇimādiguṇapradam (3.30)}
+\stopsloka
+
+mahāmudrādīnāṃ tisṛṇām atigopyatvam āha etad iti. etat trayaṃ\var{trayaṃ \lem jayaṃ \EdLo}
+mahāmudrāditrayaṃ mahāguhyam atirahasyam. atra hetugarbhāṇi viśeṣaṇāni. hi yasmāt. jarā vārdhakaṃ
+mṛtyuś caramaḥ prāṇadehaviyogaḥ tayor viśeṣeṇa nāśanam. vahner jāṭharasya vṛddhir dīptis tasyāḥ
+karaṃ kartṛ aṇimā ādir yeṣāṃ te 'ṇimādayas te ca te guṇāś ca tān prakarṣeṇa dadātīty
+aṇimādiguṇapradam. cakāra ārogyabindujayādisamuccayārthaḥ. evaśabdo 'vadhāraṇārthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_031}
+\startsloka
+{\bf aṣṭadhā kriyate caiva yāme yāme dine dine
+puṇyasaṃbhārasandhāyi pāpaughabhiduraṃ sadā
+samyakśikṣāvatām evaṃ svalpaṃ prathamasādhanam (3.31)}
+\stopsloka
+
+athaitat trayasya pṛthak sādhanaviśeṣam āha aṣṭadheti. dine dine pratidinam. yāme yāme prahare
+prahare. paunaḥpunye dvirvacanam. aṣṭabhiḥ prakārair aṣṭadhā kriyate. caśabdo viśeṣe evaśabdo
+'vadhāraṇe. etat trayam ity atrāpi sambadhyate. kīdṛśam puṇyasya\var{puṇyasya \Wthree P14 \lem puṇyāṇāṃ
+  \EdLo \EdAd, puṇya \Tue \EdMu} sambhāraḥ samūhas tasya sandhāyi vidhāyi. punaḥ kīdṛśam. pāpānām
+oghaḥ pūraḥ samūha iti yāvat. tasya bhiduraṃ kuliśam iva nāśanaṃ sadā sarvadā yadābhyastaṃ tadaiva
+pāpanāśanam. samyak sāmpradāyikī śikṣā gurūpadeśo vidyate yeṣāṃ te tathā teṣām. evaṃ dine dine yāme
+yāme 'ṣṭadhety uktarītyā pūrvaṃ prathamaṃ sādhanaṃ svalpasvalpam eva kāryam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_034}
+\startsloka
+{\bf atha khecarī
+kapālakuhare jihvā praviṣṭā viparītagā
+bhruvor antargatā dṛṣṭir mudrā bhavati khecarī (3.32)}
+\stopsloka
+
+khecarīṃ vivakṣur ādau tatsvarūpam āha kapāleti. kapāle mūrdhni kuharaṃ suṣiraṃ tasmin kapālakuhare
+viparītaṃ pratīpaṃ gacchatīti viparītagā parāṅmukhībhūtā jihvā rasanā syāt. bhruvor antargatā
+bhruvor madhye praviṣṭā dṛṣṭir darśanaṃ syāt. sā khecarīmudrā bhavati. kapālakuhare
+jihvāpraveśapūrvakaṃ bhruvor antardarśanaṃ khecarīti lakṣaṇaṃ siddham.
+%\end{vsid}
+
+%\begin{vsid}{hp03_033}
+\startsloka
+{\bf chedanacālanadohaiḥ kalāṃ krameṇātha vardhayet tāvat
+sā yāvad bhrūmadhyaṃ spṛśati tadā khecarīsiddhiḥ  (3.33)}
+\stopsloka
+
+khecarīsiddher lakṣaṇam āha chedaneti. chedanam anupadam eva vakṣyamāṇam. cālanaṃ hastayor
+aṅguṣṭhatarjanībhyāṃ rasanāṃ gṛhītvā savyāpasavyataḥ parivartanam. dohaḥ karayor
+aṅguṣṭhatarjanībhyāṃ godohanavat dohanam. taiḥ kalāṃ jihvāṃ tāvat vardhayet dīrghāṃ kuryāt. tāvat
+kiyat yāvat sā kalā bhrūmadhyaṃ bahir bhruvor madhyaṃ spṛśati yadā tadā khecaryāḥ siddhiḥ
+khecarīsiddhir bhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf snuhīpatranibhaṃ śastraṃ sutīkṣṇaṃ snigdhanirmalam
+samādāya tatas tena romamātraṃ samucchinet (3.34)}
+\stopsloka
+
+tatsādhanam āha snuhīti. snuhī guḍā tasyāḥ patraṃ dalaṃ snuhīpatram. snuhīpatreṇa sadṛśaṃ
+snuhīpatranibhaṃ sutīkṣṇam atitīkṣṇaṃ snigdhaṃ ca tan nirmalam ca snigdhanirmalam śastraṃ
+chedanasādhanaṃ samādāya samyag ādāya gṛhītvā tataḥ śastragrahaṇānantaraṃ tena śastreṇa
+romapramāṇaṃ romamātraṃ samucchinet samyag ucchinec chindyāt. rasanāmūlaśirām iti
+karmādhyāhāraḥ.
+
+\startsloka
+miśre yo 'py atha sīhuṇḍo vajradruḥ snuk snuhī guḍā
+\stopsloka
+
+ity amaraḥ\comment{?? CHECK}.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tataḥ saindhavapathyābhyāṃ cūrṇitābhyāṃ pragharṣayet
+punaḥ saptadine prāpte romamātraṃ samucchinet (3.35)}
+\stopsloka
+
+tataś chedanānantaraṃ cūrṇitābhyāṃ cūrṇīkṛtābhyāṃ saindhavaṃ sindhudeśodbhavaṃ lavaṇaṃ pathyā
+harītakī tābhyāṃ pragharṣayet prakarṣeṇa gharṣayet chinnaṃ śirāpradeśam. saptadinaparyantaṃ
+chedanaṃ saindhavapathyābhyāṃ gharṣaṇaṃ ca sāyaṃ prātar vidheyam. yogābhyāsino lavaṇaniṣedhāt
+khadirapathyācūrṇaṃ gṛhṇanti. mūle saindhavoktis tu haṭhābhyāsāt pūrvaṃ
+khecarīsādhanābhiprāyeṇa. saptānāṃ dinānāṃ samāhāraḥ saptadinaṃ tasmin prāpte gate sati aṣṭame dina
+ity arthāt. ye prāptyarthās te gatyarthāḥ. punaḥ pūrvachedanāpekṣayādhikaṃ romamātraṃ samucchinet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf evaṃ krameṇa ṣaṇmāsaṃ nityaṃ yuktaḥ samācaret
+ṣaṇmāsād rasanāmūlaśirābandhaḥ praṇaśyati (3.36)}
+\stopsloka
+
+evam iti. evaṃ krameṇa pūrvaṃ romamātracchedanaṃ saptadinaparyantaṃ tāvad eva sāyaṃprātaś chedenaṃ
+gharṣaṇaṃ ca. aṣṭame dine 'dhikaṃ chedanam ity uktakrameṇa ṣaṇmāsaṃ ṣaṇmāsaparyantaṃ nityaṃ yuktaḥ
+san samācaret samyag ācaret. chedanagharṣaṇe iti karmādhyāhāraḥ. ṣaṇmāsād anantaraṃ rasanā jihvā
+tasyā mūlam adhobhāgo rasanāmūlaṃ tatra yā śirā kapālakuhararasanāsaṃyogapratibandhakībhūtā nāḍī
+tayā bandho bandhanaṃ praṇaśyati prakarṣeṇa naśyati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_035}
+\startsloka
+{\bf kalāṃ parāṅmukhīṃ kṛtvā tripathe pariyojayet
+sā bhavet khecarīmudrā vyomacakraṃ tad ucyate (3.37)}
+\stopsloka
+
+chedanādinā jihvāvṛddhau yat kartavyaṃ tad āha kalām iti. kalāṃ jihvāṃ parāṅmukham agraṃ yasyāḥ sā
+tathā tāṃ parāṅmukhīṃ pratyaṅmukhīṃ kṛtvā tisṛṇāṃ nāḍīnāṃ panthāḥ tripathas tasmiṃstripathe
+kapālakuhare pariyojayet saṃyojayet. sā tripathe rasanāpariyojanarūpā khecarīmudrā tad vyomacakram
+ity ucyate vyomacakraśabdenocyate.
+%\end{vsid}
+
+%\begin{vsid}{hp03_035}
+\startsloka
+{\bf rasanām ūrdhvagāṃ kṛtvā kṣaṇārdham api tiṣṭhati
+viṣair vimucyate yogī vyādhimṛtyujarādibhiḥ (3.38)}
+\stopsloka
+
+atha khecarīguṇāḥ rasanām iti. ūrdhvaṃ tālūpari vivaraṃ gacchatīti ūrdhvagā tāṃ tādṛśīṃ rasanām
+jihvāṃ kṛtvā kṣaṇārdham kṣaṇasya muhūrtasya ardhaṃ kṣaṇārdham ghaṭikāmātram api khecarīmudrā
+tiṣṭhati cet tarhi yogī viṣaiḥ sarpavṛścikādiviṣair vimucyate viśeṣeṇa mucyate. vyādhir
+dhātuvaiṣamyaṃ mṛtyuś caramaḥ prāṇadehaviyogaḥ jarā vṛddhāvasthā tā ādayo yeṣāṃ valyādīnāṃ taiś ca
+vimucyate.
+
+\startsloka
+utsave ca prakoṣṭhe ca muhūrte niyame tathā
+kṣaṇaśabdo vyavasthāyāṃ samaye 'pi nigadyate 
+\stopsloka
+
+iti nānārthaḥ\comment{REF?}
+%\end{vsid}
+
+%\begin{vsid}{hp03_036}
+\startsloka
+{\bf na rogo maraṇaṃ tandrā na nidrā na kṣudhā tṛṣā
+na ca mūrcchā bhavet tasya yo mudrāṃ vetti khecarīm  (3.39)}
+\stopsloka
+
+na roga iti. yaḥ khecarīṃ mudrāṃ vetti tasya rogo na maraṇaṃ na tandrā
+tāmasāntaḥkaraṇavṛttiviśeṣo na nidrā na kṣudhā tṛṣā pipāsā na mūrcchā cittasya
+tamasābhibhūtāvasthāviśeṣaś ca na bhavet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_40}
+\startsloka
+{\bf pīḍyate na sa rogeṇa lipyate na ca karmaṇā
+bādhyate na sa kālena yo mudrāṃ vetti khecarīm (3.40)}
+\stopsloka
+
+pīḍyata iti. yaḥ khecarīṃ mudrāṃ vetti sa rogeṇa jvarādinā na pīḍyate pramādāj jātenāśubhena
+karmaṇā lokasaṅgrahārthakṛtaśubhena ca karmaṇā na lipyate. kālena mṛtyunā sa na bādhyate na
+hanyate.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf cittaṃ carati khe yasmāj jihvā carati khe gatā
+tenaiṣā khecarī nāma mudrā siddhair nirūpitā (3.41)}
+\stopsloka
+
+cittam iti. yasmād dhetoś cittam antaḥkaraṇaṃ khe bhruvor antaravakāśe carati jihvā khe tatraiva
+gatā satī carati. tena hetunā eṣā kathitā mudrā khecarī nāma khecarīti prasiddhā. nāmeti prasiddhāv
+avyayam. siddhaiḥ kapilādibhir nirūpitā. khe bhruvor antar vyomni carati gacchati cittaṃ jihvā ca
+yasyāṃ sā khecarīty avayavaśaktyā vyutpāditā. ukteṣu triṣu ślokeṣu vyādhyādīnāṃ punaruktis tu teṣāṃ
+ślokānāṃ saṃgṛhītatvān na doṣāya.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf khecaryā mudritaṃ yena vivaraṃ lambikordhvataḥ
+na tasya kṣarate binduḥ kāminyāśleṣitasya ca (3.42)}   
+\stopsloka
+
+khecaryeti. yena yoginā khecaryā mudrayā lambikāyā ūrdhvam iti lambikordhvataḥ. sārvibhaktikas
+tasiḥ. lambikā tālu tasyā ūrdhvata uparibhāge sthitaṃ vivaraṃ chidraṃ mudritaṃ pihitam. kāminyā
+yuvatyāśleṣitasyāliṅgitasyāpi. caśabdo 'pyarthe. tasya binduḥ vīryaṃ na kṣarate na skhalati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf calito 'pi yadā binduḥ samprāpto yonimaṇḍalam
+vrajaty ūrdhvaṃ hṛtaḥ śaktyā nibaddho yonimudrayā (3.43)}
+\stopsloka
+
+calita iti. calito 'pi skhalito 'pi bindur yadā yasmin kāle yonimaṇḍalam yonisthānaṃ samprāptaḥ
+saṃgatas tadaiva yonimudrayā meḍhrākuñcanarūpayā. etena vajrolīmudrā sūcitā. nibaddho nitarāṃ
+baddhaḥ śaktyākarṣaṇaśaktyā hṛtaḥ ākṛṣṭa ūrdhvaṃ vrajati. suṣumnāmārgeṇa bindusthānaṃ gacchati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf ūrdhvajihvaḥ sthiro bhūtvā somapānaṃ karoti yaḥ
+māsārdhena na sandeho mṛtyuṃ jayati yogavit (3.44)}
+\stopsloka
+
+ūrdhvajihva iti. ūrdhvā lambikordhvavivaronmukhā jihvā yasya sa ūrdhvajihvaḥ sthiro niścalo bhūtvā
+somasya lambikordhvavivaragalitacandrāmṛtasya pānaṃ somapānaṃ yaḥ pumān karoti yogaṃ vettīti
+yogavit sa māsasyārdhaṃ māsārdhaṃ tena māsārdhena pakṣeṇa mṛtyuṃ maraṇaṃ jayati abhibhavati. na
+sandehaḥ niścitam etad ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf nityaṃ somakalāpūrṇaṃ śarīraṃ yasya yoginaḥ
+takṣakeṇāpi daṣṭasya viṣaṃ tasya na sarpati (3.45)}
+\stopsloka
+
+nityam iti. yasya yoginaḥ śarīraṃ nityaṃ pratidinaṃ somakalāpūrṇaṃ candrakalāmṛtapūrṇaṃ tasya
+takṣakeṇa sarpaviśeṣeṇāpi daṣṭasya daṃśitasya yoginaḥ śarīre viṣaṃ garalaṃ tajjanyaṃ duḥkhaṃ na
+sarpati na prasarati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf indhanāni yathā vahnis tailavartiṃ ca dīpakaḥ
+tathā somakalāpūrṇaṃ dehī dehaṃ na muñcati (3.46)}
+\stopsloka
+
+yathā vahnir agniḥ indhanāni kāṣṭhādīni na muñcati dīpako dīpaḥ tailavartiṃ ca tailayuktāṃ vartiṃ na
+muñcati tathā somakalāpūrṇaṃ candrakalāmṛtapūrṇaṃ dehaṃ śarīraṃ dehī jīvo na muñcati na tyajati.
+%\end{vsid}
+
+% anti-Tantric, for B. it is by Svātmārāma.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf gomāṃsaṃ bhakṣayen nityaṃ pibed amaravāruṇīm
+kulīnaṃ tam ahaṃ manye itare kulaghātakāḥ (3.47)}
+\stopsloka
+
+gomāṃsam iti. gomāṃsaṃ pāribhāṣikaṃ vakṣyamāṇaṃ yo bhakṣayen nityaṃ pratidinam amaravāruṇīm api
+vakṣyamāṇāṃ pibet taṃ yoginam. aham iti granthakāroktiḥ. kule jātaḥ kulīnaḥ tam satkulotpannaṃ
+manye. tad uktaṃ brahmavaivarte\comment{???}
+
+\startsloka
+kṛtārthau pitarau tena dhanyo deśaḥ kulaṃ ca tat 
+jāyate yogavān yatra dattam akṣayyatāṃ vrajet
+dṛṣṭaḥ sambhāṣitaḥ spṛṣṭaḥ puṃprakṛtyor vivekavān
+bhavakoṭiśatopāttaṃ punāti vṛjinaṃ nṛṇām iti 
+\stopsloka
+
+brahmāṇḍapurāṇe\comment{???}
+
+\startsloka
+gṛhasthānāṃ sahasreṇa vānaprasthaśatena ca
+brahmacārisahasreṇa yogābhyāsī viśiṣyate
+\stopsloka
+
+rājayoge vāmadevaṃ prati śivavākyaṃ\comment{???}
+
+\startsloka
+rājayogasya māhātmyaṃ ko vijānāti tattvataḥ
+tajjñānī vasate yatra sa deśaḥ puṇyabhājanam
+darśanād arcanād asya triḥsaptakulasaṃyutāḥ 
+ajñā muktipadaṃ yānti kiṃ punas tatparāyaṇāḥ
+antaryogaṃ bahiryogaṃ yo jānāti viśeṣataḥ
+tvayā mayāpy asau vandyaḥ śeṣair vandyās tu kiṃ punaḥ 
+\stopsloka
+
+iti.
+
+kūrmapurāṇe\comment{???}
+
+\startsloka
+ekakālaṃ dvikālaṃ vā trikālaṃ nityam eva vā
+ye yuñjate mahāyogaṃ vijñeyās te maheśvarāḥ
+\stopsloka
+
+iti. itare vakṣyamāṇagomāṃsabhakṣaṇāmaravāruṇīpānarahitā ayoginas te kulaghātakāḥ
+kulanāśakāḥ satkule jātasya janmano vaiyarthyāt.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf gośabdenoditā jihvā tat praveśo hi tāluni
+gomāṃsabhakṣaṇaṃ tat tu mahāpātakanāśanam  (3.48)}
+\stopsloka
+
+gomāmṣaśabdārtham āha gośabdeneti. gośabdena go ityākārakeṇa śabdena gopadenety arthaḥ. jihvā
+rasanoditā kathitā. tālunīti sāmīpikādhāre saptamī. tālusamīpordhvavivare tasyā jihvāyāḥ praveśo
+gomāṃsabhakṣaṇaṃ gomāṃsabhakṣaṇaśabdavācyam. tat tu tādṛśaṃ gomāṃsabhakṣaṇaṃ tu mahāpātakānāṃ
+svarṇasteyādīnāṃ nāśanam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf jihvāpraveśasambhūtavahninotpāditaḥ khalu
+candrāt sravati yaḥ sāraḥ sā syād amaravāruṇī  (3.49)}
+\stopsloka
+
+amaravāruṇīśabdārtham āha jihveti. jihvāyāḥ praveśo lambikordhvavivare praveśanaṃ tasmāt sambhūto
+jāto yo vahnir ūṣmā tenotpādito niṣpāditaḥ. atra vahniśabdenauṣṇyam upalakṣyate. yaḥ sāraḥ candrād
+bhruvor antarvāmabhāgasthāt somāt sravati galati sā amaravāruṇī syād amaravāruṇīpadavācyā bhavet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf cumbantī yadi lambikāgram aniśaṃ jihvā rasasyandinī 
+sakṣārā kaṭukāmladugdhasadṛśī madhvājyatulyā tathā
+vyādhīnāṃ haraṇaṃ jarāntakaraṇaṃ śastrāgamodīraṇaṃ
+tasya syād amaratvam aṣṭaguṇitaṃ siddhāṅganākarṣaṇam  (3.50)}
+\stopsloka
+
+cumbantīti. yadi cel lambikāgram lambikordhvavivaraṃ cumbantī spṛśantī. aniśaṃ nirantaram. ata eva
+rasasya somakalāmṛtasya syandaḥ syandanaṃ prasravaṇam asyām astīti rasasyandinī yasya
+jihvā. kṣāreṇa lavaṇarasena sahitā sakṣārā. kaṭukaṃ marīcādi. āmlaṃ ciñcāphalādi. dugdhaṃ payas.
+taiḥ sadṛśī samānā. madhu kṣaudram. ājyaṃ ghṛtaṃ. tābhyāṃ tulyā samā. tathāśabdaḥ samuccaye. etair
+viśeṣaṇai rasasyānekarasavattvād madhuratvāt snigdhatvāc ca jihvāyā api rasasyandane tathātvam
+uktam. tarhi tasya vyādhīnāṃ rogāṇāṃ haraṇam apagamo jarāyā vṛddhāvasthāyā antaḥkaraṇaṃ nāśanaṃ
+śastrāṇām āyudhānām āgamaḥ svābhimukhāgamanaṃ tasyodīraṇaṃ nivāraṇam. aṣṭau guṇā aṇimādayas te asya
+sañjātā ity aṣṭaguṇitam amaratvam amarabhāvaḥ siddhānām aṅganāḥ siddhāṅganāḥ siddhāś ca tā
+aṅganāś ceti vā tāsām ākarṣaṇam ākarṣaṇaśaktiḥ syāt.
+%\end{vsid}
+
+% sequence of tastes acc. to Jim. not presence of all tastes. Khecari, Intro, p. 22.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf mūrdhnaḥ ṣoḍaśapatrapadmagalitaṃ prāṇād avāptaṃ haṭhād
+ūrdhvāsyo rasanāṃ niyamya vivare śaktiṃ parāṃ cintayan 
+utkallolakalājalaṃ ca vimalaṃ dhārāmayaṃ yaḥ piben 
+nirvyādhiḥ sa mṛṇālakomalavapur yogī ciraṃ jīvati (3.51)}
+\stopsloka
+
+mūrdhna iti. rasanāṃ jihvāṃ vivare kapālakuhare niyamya saṃyojya. ūrdhvam uttānam āsyaṃ yasya saḥ
+ūrdhvāsya ity anena viparītakaraṇī sūcitā. parāṃ śaktiṃ kuṇḍalinīṃ cintayan dhyāyan san prāṇāt
+sādhanabhūtāt. ṣoḍaśa patrāṇi dalāni yasya tat ṣoḍaśapatraṃ tac ca tatpadmaṃ ca kaṇṭhasthāne
+vartamānaṃ tasmin galitaṃ haṭhād haṭhayogād avāptaṃ prāptaṃ vimalaṃ nirmalam dhārāmayaṃ dhārārūpam
+utkallolam uttaraṅgaṃ ca tat kalājalaṃ somakalārasaṃ yaḥ pumān pibet dhayet sa yogī nirgatā vyādhayo
+jvarādayo yasmāt sa nirvyādhiḥ san yad vā nirgatā vividhā trividhā ādhayo mānasī vyathā yasmāt sa
+tādṛśaḥ san mṛṇālaṃ bisam iva komalaṃ mṛdulaṃ vapuḥ śarīraṃ yasya sa mṛṇālakomalavapuś ca san ciraṃ
+dīrghakālaṃ jīvati prāṇān dhārayati. haṭhād dhaṭhayogāt. prāṇāt sādhanabhūtād avāptam iti vā
+yojanā. prāṇair iti kvacit pāṭhaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf yat prāleyaṃ prahitasuṣiraṃ merumūrdhāntarasthaṃ
+tasmiṃs tattvaṃ pravadati sudhīs tanmukhaṃ nimnagānām
+candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ
+tad badhnīyāt sukaraṇam adho nānyathā kāyasiddhiḥ  (3.52)}
+\stopsloka
+
+yat prāleyam iti. meruvat sarvonnatā suṣumnā merus tasya mūrdhā uparibhāgaḥ tasyāntare madhye
+tiṣṭhatīti merumūrdhāntarasthaṃ yat prāleyaṃ somakalājalaṃ prahitaṃ nihitaṃ yasmin tat tathā tac
+ca tat suṣiraṃ vivaraṃ tasmin vivare sudhīḥ śobhanā rajastamobhyām anabhibhūtasattvā dhīr yasya
+saḥ tattvam ātmatattvaṃ pravadati prakarṣeṇa vadati.
+
+\startsloka
+tasyāḥ śikhāyā madhye paramātmā vyavasthitaḥ\comment{ma.nā.u. 11.13\lem ??}
+\stopsloka
+
+iti śruteḥ.
+
+ātmano vibhutve khecarīmudrāyāṃ tatrābhivyaktis tasmiṃs tattvam ity uktam. nimnagānām
+gaṅgāyamunāsarasvatīnarmadādiśabdavācyānām iḍāpiṅgalāsuṣumnāgāndhārīprabhṛtīnāṃ tat tasmin vivare
+tatsamīpe mukham agram asti candrāt somād vapuṣaḥ śarīrasya sāraḥ sārabhūto rasaḥ sravati kṣarati
+tena candrasārakṣaraṇena narāṇāṃ manuṣyāṇāṃ mṛtyur maraṇaṃ bhavati. ato hetos tatpūrvoditaṃ
+sukaraṇaṃ śobhanaṃ karaṇaṃ khecarīmudrākhyaṃ badhnīyāt. sukaraṇe baddhe candrasārasravaṇābhāvān
+mṛtyur na syād iti bhāvaḥ. anyathā sukaraṇabandhanābhāve kāyasya dehasya siddhī
+rūpalāvaṇyabalavajrasaṃhananarūpā na syāt.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf suṣiraṃ jñānajanakaṃ pañcasrotaḥsamanvitam
+tiṣṭhate khecarīmudrā tasmin śūnye nirañjane  (3.53)}
+\stopsloka
+
+suṣiram iti. pañca yāni srotāṃsi iḍādīnāṃ pravāhāḥ taiḥ samanvitam samyag anugatam
+saptasrotaḥsamanvitam iti kvacit pāṭhaḥ. jñānajanakam alaukikābādhitātmasākṣātkārajanakaṃ yat
+suṣiraṃ vivaraṃ tasmin suṣire 'ñjanam avidyā tatkāryaṃ śokamohādi ca nirgataṃ yasmāt tan nirañjanaṃ
+tasmin nirañjane śūnye suṣirāvakāśe khecarīmudā tiṣṭhate sthirībhavati. prakāśanastheyākhyayoś ca
+\comment{Aṣṭādhyāyī 1.3.23.} ity ātmanepadam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī
+eko devo nirālamba ekāvasthā manonmanī  (3.54)}
+\stopsloka
+
+ekam iti. sṛṣṭimayaṃ sṛṣṭirūpaṃ praṇavākhyaṃ bījam ekaṃ mukhyam. tad uktaṃ māṇḍūkyopaniṣadi
+
+om ity etad akṣaram idaṃ sarvam\comment{Māṇḍūkyopaniṣad 1.}
+
+iti.
+
+khecarīmudrā ekā mukhyā. nirālamba ālambanaśūnya eko mukhyo devaḥ. ālambanaparityāgenātmanaḥ
+svarūpāvasthānāt. unmanyavasthaikā mukhyā. eke mukhyānyakevalāḥ\comment{Amarakośa ??} ity
+amaraḥ. bījādiṣu praṇavādivan mudrāsu khecarī mukhyety arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf atha uḍḍīyānabandhaḥ
+baddho yena suṣumṇāyāṃ prāṇas tūḍḍīyate yataḥ
+tasmād uḍḍīyanākhyo 'yaṃ yogibhiḥ samudāhṛtaḥ (3.55)}
+\stopsloka
+
+uḍḍīyānabandhaṃ vivakṣus tāvad uḍḍīyānaśabdārtham āha baddha iti. yato yasmād dhetor yena bandhena
+baddho niruddhaḥ prāṇaḥ suṣumnāyāṃ madhyanāḍyām uḍḍīyate suṣumnāṃ vihāyasā gacchati tasmāt kāraṇād
+ayaṃ bandho yogibhir matsyendrādibhir ? check uḍḍīya(ā)?  uḍḍīyānam ākhyābhidhā yasya sa
+uḍḍīyānākhyaḥ samudāhṛtaḥ samyag vyutpattyodāhṛtaḥ kathitaḥ. suṣumnāyām uḍḍīyate 'nena baddhaḥ
+prāṇaḥ ity uḍḍīyānam. utpūrvāḍ ḍīṅ vihāyasā gatāv ity asmāt karaṇe lyuṭ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf uḍḍīnaṃ kurute yasmād aviśrāntaṃ mahākhagaḥ
+uḍḍīyānaṃ tad eva syāt tatra bandho 'bhidhīyate (3.56)}
+\stopsloka
+
+uḍḍīṇam iti. mahāṃś cāsau khagaś ca mahākhagaḥ prāṇaḥ. sarvadā dehāvakāśe gatimattvāt. yasmād
+bandhād aviśrāntaṃ yathā syāt tathoḍḍīnaṃ vihaṅgamagatiṃ kurute. suṣumnāyām ity adhyāhāryam. tad
+eva bandhaviśeṣa eva uḍḍīyānam uḍḍīyānanāmakaṃ syāt. tatra tasmin viṣaye bandho 'bhidhīyate
+bandhasvarūpaṃ kathyate mayeti śeṣaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf udare paścimaṃ tānaṃ nābher ūrdhvaṃ ca kārayet
+uḍḍīyāno hy asau bandho mṛtyumātaṅgakesarī (3.57)}
+\stopsloka
+
+uḍḍīyānabandham āha udara iti. udare tunde nābher ūrdhvaṃ cakārād adha uparibhāge 'dhobhāge ca
+paścimaṃ tānaṃ paścimam ākarṣaṇaṃ nābher ūrdhvādhobhāgau yathā pṛṣṭhasaṃlagnau syātāṃ tathā tānaṃ
+tānanaṃ nāmākarṣaṇaṃ kārayet kuryāt. ṇijartho 'vivakṣitaḥ. asau nābher ūrdhvādhobhāgayos tānanarūpa
+uḍḍīyāna uḍḍīyānākhyo bandhaḥ. kīdṛśaḥ mṛtyur eva mātaṅgo gajas tasya kesarī siṃhaḥ siṃha iva
+nivartakaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf uḍḍīyānaṃ tu sahajaṃ guruṇā kathitaṃ sadā
+abhyaset satataṃ yas tu vṛddho 'pi taruṇāyate (3.58)}
+\stopsloka
+
+uḍḍīyānaṃ tv iti. gurur hitopadeṣṭā tena guruṇā uḍḍīyānaṃ tu sadā sarvadā sahajaṃ svābhāvikaṃ
+kathitam. prāṇasya bahirgamanaṃ sarvadā sarvasyaiva jāyamānatvāt. yas tu yaḥ puruṣas tu satataṃ
+nirantaram abhyaset. uḍḍīyānam ity atrāpi sambadhyate. sa tu vṛddho 'pi sthaviro 'pi taruṇāyate
+taruṇa ivācarati taruṇāyate.
+%\end{vsid}
+
+% Rewritten by Haṭharatnāvalī  guruṇā sahajaṃ proktam vṛddho 'pi .. (check)
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf nābher ūrdhvam adhaś cāpi tānaṃ kuryāt prayatnataḥ 
+ṣaṇmāsam abhyasen mṛtyuṃ jayaty eva na saṃśayaḥ (3.59)}
+\stopsloka
+
+nābher iti. nābher ūrdhvam uparibhāge 'dhaś cāpy adhobhāge 'pi prayatnataḥ prakṛṣṭo yatnaḥ
+prayatnas tasmāt prayatnataḥ. yatnaviśeṣāt tānaṃ paścimatānaṃ kuryāt. pūrvārdhena uḍḍīyānasvarūpam
+uktam. atha tatpraśaṃsā yaḥ ṣaṇmāsam ṣaṇmāsaparyantam uḍḍīyānam ity adhyāhāraḥ abhyaset punaḥ
+punar anutiṣṭhet. sa mṛtyuṃ jayaty eva saṃśayo na atra sandeho nāstīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf sarveṣām eva bandhānām uttamo hy uḍḍīyānakaḥ
+uḍḍiyāne dṛḍhe bandhe muktiḥ svābhāvikī bhavet (3.60)}
+\stopsloka
+
+sarveṣām iti. sarveṣāṃ bandhānām ṣoḍaśādhārabandhānāṃ madhye uḍḍīyānaka eva uḍḍīyānabandha eva
+svārthe kapratyayaḥ uttama utkṛṣṭaḥ hi yasmād uḍḍīyāne bandhe dṛḍhe sati svābhāvikī
+svabhāvasiddhaiva muktir bhavet. uḍḍīyānabandhe kṛte vihaṅgamagatyā suṣumnāyāṃ prāṇasya mūrdhni
+gamanāt samādhau mokṣam āpnoti iti vākyāt\comment{Gorakṣaśataka 113c.} sahajaiva muktiḥ syād iti
+bhāvaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf atha mūlabandhaḥ
+pārṣṇibhāgena sampīḍya yonim ākuñcayed gudam
+apānam ūrdhvam ākṛṣya mūlabandho 'bhidhīyate (3.61)}
+\stopsloka
+
+mūlabandham āha pārṣṇibhāgeneti. pārṣṇibhāgaḥ pārṣṇer bhāgo gulphayor adhaḥpradeśas tena yoniṃ
+yonisthānam gudameḍhrayor madhyabhāgaṃ sampīḍya samyak pīḍayitvā gudam pāyum ākuñcayed
+saṅkocayet. apānam adhogatiṃ vāyum ūrdhvam upary ākṛṣyākṛṣṭaṃ kṛtvā mūlabandho 'bhidhīyate
+kathyate. pārṣṇibhāgena yonisthānasampīḍanapūrvakaṃ gudasyākuñcanaṃ mūlabandha ity ucyate ity
+arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf adhogatim apānaṃ vai ūrdhvagaṃ kurute balāt
+ākuñcanena taṃ prāhur mūlabandhaṃ hi yoginaḥ (3.62)}
+\stopsloka
+
+adhogatim iti. yaḥ adhogatim adho 'rvāggatir yasya sa tathā tam apānam apānavāyum ākuñcanena
+mūlādhārasya saṅkocanena balād haṭhād ūrdhvaṃ gacchatīty ūrdhvagaḥ tam ūrdhvagaṃ suṣumnāyām
+ūrdhvagamanaśīlaṃ kurute. vai iti niścaye 'vyayam. yogino yogābhyāsinas taṃ mūlabandhaṃ mūlasya
+mūlasthānasya bandhanaṃ mūlabandhas taṃ mūlabandham ity anvarthaṃ prāhuḥ. anena mūlabandhaśabdārtha
+uktaḥ. pūrvaślokena tu tasya bandhanaprakāra ukta ity apaunaruktyam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf gudaṃ pārṣṇyā tu sampīḍya vāyum ākuñcayed balāt
+vāraṃ vāraṃ yathā cordhvaṃ samāyāti samīraṇaḥ  (3.63)}
+\stopsloka
+
+% B. knows the Southern recension of the Yogabīja (16th) according to Birch.
+
+atha yogabījoktarītyā\comment{Yogabīja 103 (Southern recension).} mūlabandham āha gudam
+iti. pārṣṇyā gulphayor adhobhāgena gudaṃ pāyuṃ sampīḍya samyak pīḍayitvā saṃyojyety
+arthaḥ. tuśabdaḥ pūrvasmād asya viśeṣatvadyotakaḥ. yathā yena prakāreṇa samīraṇo vāyur ūrdhvaṃ
+suṣumnāyā uparibhāge yāti gacchati tathā tena prakāreṇa balād haṭhād vāraṃ vāraṃ punaḥ punar vāyum
+apānam ākuñcayed gudasyākuñcanenākarṣayet. ayaṃ mūlabandha iti vākyādhyāhāraḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf prāṇāpānau nādabindū mūlabandhena caikatām
+gatvā yogasya saṃsiddhiṃ yacchato nātra saṃśayaḥ (3.64)}
+\stopsloka
+
+atha mūlabandhaguṇān āha prāṇāpānau iti. prāṇaś cāpānaś ca prāṇāpānau ūrdhvādhogatī vāyū. nādo
+'nāhatadhvaniḥ bindur anusvāras tau mūlabandhenaikatām gatvaikībhūya yogasya saṃsiddhiṃ samyak
+siddhis tāṃ yogasiddhiṃ yacchato dattaḥ. abhyāsina iti śeṣaḥ. atrāsminn arthe saṃśayo na sandeho
+nāstīty arthaḥ. ayaṃ bhāvaḥ mūlabandhe kṛte 'pānaḥ prāṇena sahaikībhūya suṣumnāyāṃ praviśati. tato
+nādābhivyaktir bhavati. tato nādena saha prāṇāpānau hṛdayopari gatvā nādasya bindunā sahaikyam
+vidhāya mūrdhni gacchataḥ. tato yogasiddhiḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf apānaprāṇayor aikyaṃ kṣayo mūtrapurīṣayoḥ
+yuvā bhavati vṛddho 'pi satataṃ mūlabandhanāt (3.65)}
+\stopsloka
+
+apānaprāṇayor iti. satataṃ santataṃ mūlabandhanān mūlabandhamudrākaraṇād apānaprāṇayor aikyaṃ
+bhavati. mūtrapurīṣayoḥ sañcitayoḥ kṣayaḥ patanaṃ bhavati. vṛddho 'pi sthaviro 'pi yuvā taruṇo
+bhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf apāna ūrdhvage jāte prayāte vahnimaṇḍalam
+tadānalaśikhā dīrghā jāyate vāyunāhatā (3.66)}
+\stopsloka
+
+apāna iti. mūlabandhanād apāne adhogamanaśīle vāyau ūrdhvage ūrdhvaṃ gacchatīty ūrdhvagaḥ tasmin
+tādṛśe sati vahnimaṇḍalam vahner maṇḍalam trikoṇaṃ nābher adhobhāge 'sti. tad uktaṃ
+yājñavalkyena\comment{Yogayājñyavalkya  4.11cd–131b}
+%\end{vsid}
+
+\startsloka
+dehamadhye śikhisthānaṃ taptajāmbūnadaprabham
+trikoṇaṃ tu manuṣyānāṃ  caturasraṃ catuṣpadām
+maṇḍalaṃ tu pataṅgānāṃ  satyam etad bravīmi te
+tanmadhye tu śikhā tanvī sadā tiṣṭhati pāvake
+\stopsloka
+
+iti.  tadā tasmin kāle vāyunā apānenāhatā saṅgatā saty analaśikhā jaṭharāgniśikhā dīrghā āyatā
+jāyate. vardhata iti kvacit pāṭhaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tato yāto vahnyapānau prāṇam uṣṇasvarūpakam
+tenātyantapradīptas tu jvalano dehajas tathā  (3.67)}
+\stopsloka
+
+tata iti. tatas tadanantaraṃ vahniś cāpānaś ca vahnyapānau. uṣṇaṃ svarūpaṃ yasya sa tathā tam
+analaṃ śikhādairghyād uṣṇasvarūpaṃ prāṇam ūrdhvagatim anilaṃ yāto gacchataḥ. tato
+'nalaśikhādairghyād uṣṇasvarūpakam iti vā yojanā. tena prāṇasaṅgamanena dehe jāto dehajo jvalano
+'gnir atyantam adhikaṃ dīpto bhavati. tatheti pādapūraṇe. apānasyordhvagamanena dīpta eva jvalanaḥ
+prāṇasaṅgatyātyantaṃ pradīpto bhavatīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tena kuṇḍalinī suptā santaptā samprabudhyate
+  daṇḍāhatā bhujaṅgīva niśvasya\comment{\EdLo reads {\em niḥśvasya} against
+    all sources checked.} ṛjutāṃ vrajet (3.68)}
+\stopsloka
+
+teneti. tena jvalanasyātyantapradīpanena santaptā samyak taptā satī suptā nidritā kuṇḍalinī śaktiḥ
+samprabudhyate samyak prabuddhā bhavati. daṇḍenāhatā daṇḍāhatā. sā cāsau bhujaṅgīva sarpiṇīva
+niśvasya niśvāsaṃ kṛtvā ṛjutāṃ saralatāṃ vrajed gacchet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf bilaṃ praviṣṭeva tato brahmanāḍyantaraṃ vrajet
+tasmān nityaṃ mūlabandhaḥ kartavyo yogibhiḥ sadā (3.69)}
+\stopsloka
+
+bilaṃ praviṣṭeti. tataḥ ṛjutāprāptyanantaraṃ bilaṃ vivaraṃ praviṣṭā bhujaṅgīva brahmanāḍī suṣumnā
+tasyā antaraṃ madhyaṃ vrajet gacchet. tasmād dhetor yogibhir yogābhyāsibhir mūlabandho nityaṃ
+pratidinaṃ sadā sarvasmin kāle kartavyaḥ kartuṃ yogyaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf atha jālandharabandhaḥ
+kaṇṭham ākuñcya hṛdaye sthāpayec cibukaṃ dṛḍham
+bandho jālandharākhyo 'yaṃ jarāmṛtyuvināśakaḥ (3.70)}
+\stopsloka
+
+jālandharabandham āha kaṇṭham iti. kaṇṭham galabilam ākuñcya saṅkocya hṛdaye vakṣaḥsamīpe
+caturaṅgulāntaritapradeśe cibukaṃ hanuṃ dṛḍham sthiraṃ sthāpayet sthitaṃ kuryāt. ayaṃ
+kaṇṭhākuñcanapūrvakaṃ caturaṅgulāntaritahṛdayasamīpe 'dhonamanayatnapūrvakaṃ cibukasthāpanarūpo
+jālandhara ity ākhyayata iti jālandharākhyo jālandharanāmā bandhaḥ. kīdṛśaḥ jarā vṛddhāvasthā
+mṛtyur maraṇaṃ tayor vināśako viśeṣeṇa nāśayatīti vināśako vināśakartā.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf badhnāti hi śirājālam adhogāmi nabhojalam
+tato jālandharo bandhaḥ kaṇṭhaduḥkhaughanāśanaḥ (3.71)}
+\stopsloka
+
+jālandharapadasyārtham āha badhnātīti. hi yasmāt. śirāṇāṃ nāḍīnāṃ jālam samudāyaṃ badhnāti. adho
+gantuṃ śīlam asyety adhogāmi nabhasaḥ kapālakuharasya jalam amṛtaṃ ca badhnāti pratibadhnāti. tatas
+tasmāj jālandharo jālandharanāmako 'nvartho bandhaḥ. jālaṃ sirājālaṃ (not śirā??) jalānāṃ samūho
+jālaṃ ca dharatīti jālandharaḥ. kīdṛśaḥ kaṇṭhe galapradeśe yo duḥkhaugho vikārajāto duḥkhasamūhas
+tasya nāśano nāśakartā.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf jālandhare kṛte bandhe kaṇṭhasaṃkocalakṣaṇe
+na pīyūṣaṃ pataty agnau na ca vāyuḥ prakupyati (3.72)}
+\stopsloka
+
+jālandharaguṇān āha jālandhara iti. kaṇṭhasya galabilasya saṃkocanaṃ saṅkoca ākuñcanaṃ tad eva
+lakṣaṇaṃ svarūpaṃ yasya saḥ kaṇṭhasaṃkocalakṣaṇaḥ tasmin tādṛśe jālandhare jālandharasaṃjñake
+bandhe kṛte sati pīyūṣam amṛtam agnau jāṭhare 'nale na patati na sarati. vāyuś ca prāṇaś ca na
+prakupyati nāḍyantare vāyor gamanaṃ prakopas taṃ na karotīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf kaṇṭhasaṃkocanenaiva dve nāḍyau stambhayed dṛḍham
+madhyacakram idaṃ jñeyaṃ ṣoḍaśādhārabandhanam (3.73)}
+\stopsloka
+
+kaṇṭhasaṃkocaneneti. dṛḍham gāḍhaṃ kaṇṭhasaṃkocanenaiva kaṇṭha\-saṃkoca\-namātreṇa dve nāḍyau
+iḍāpiṅgale stambhayed bandhayet. ayaṃ jālandhara iti kartṛpadādhyāhāraḥ. idaṃ kaṇṭhasthāne sthitaṃ
+viśuddhākhyaṃ cakram madhyacakram madhyamaṃ cakram jñeyam. kīdṛśam ṣoḍaśādhārabandhanam
+ṣoḍaśasaṅkhyākā ye ādhārā aṅguṣṭhā\-dhārādi\-brahma\-randhrāntās teṣāṃ bandhanaṃ bandhanakārakam.
+
+\startsloka
+aṅguṣṭhagulphajānūrusīvanīliṅganābhayaḥ
+hṛd grīvā kaṇṭhadeśaś ca lambikā\var{lambikā \Wthree \Tue \EdMu \lem lambhikā \EdLo \EdAd} nāsikā tathā
+bhrūmadhyaṃ ca lalāṭaṃ ca mūrdhā ca brahmarandhrakam
+ete hi ṣoḍaśādhārāḥ kathitā yogipuṅgavaiḥ 
+\stopsloka
+
+teṣv ādhareṣu dhāraṇāyāḥ phalaviśeṣas tu gorakṣasiddhāntād\comment{siddhasiddhāntapaddhatiḥ 2.1025
+  ??} avagantavyaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf mūlasthānaṃ samākuñcya uḍḍiyānaṃ tu kārayet
+iḍāṃ ca piṅgalāṃ baddhvā vāhayet paścime pathi (3.74)}
+\stopsloka
+
+uktasya bandhatrayasyopayogam āha mūlasthānam iti. mūlasthānam ādhārabhūtam ādhārasthānaṃ
+samākuñcya samyag ākuñcya uḍḍiyānaṃ nābheḥ paścimatānarūpaṃ bandhaṃ kārayet kuryāt. ṇijartho
+'vivakṣitaḥ. iḍāṃ gaṅgāṃ piṅgalāṃ yamunāṃ ca baddhvā jālandharabandhenety
+arthaḥ. kaṇṭhasaṅkocanenaiva dve nāḍyau stambhayed dṛḍham ity ukteḥ. paścime pathi suṣumnāmārge
+vāhayet gamayet prāṇam iti śeṣaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf anenaiva vidhānena prayāti pavano layam
+tato na jāyate mṛtyur jarārogādikaṃ tathā (3.75)}
+\stopsloka
+
+aneneti. anenaivoktenaiva vidhānena vidhinā pavanaḥ prāṇo layam sthairyaṃ
+prayāti. gatyabhāvapūrvakaṃ brahmarandhre sthitiḥ prāṇasya layaḥ. tataḥ prāṇasya layān mṛtyur
+jarārogādikaṃ. tathā cārthe. na jāyate nodbhavati. ādipadena valīpalitatandrālasyādikaṃ grāhyam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf bandhatrayam idaṃ śreṣṭhaṃ mahāsiddhaiś ca sevitam
+sarveṣāṃ haṭhatantrāṇāṃ sādhanaṃ yogino viduḥ (3.76)}
+\stopsloka
+
+bandhatrayam iti. idaṃ pūrvoktaṃ bandhatrayam śreṣṭhaṃ ṣoḍaśādhārabandheṣv atipraśastaṃ
+mahāsiddhair matsyendrādibhiḥ cakārād vasiṣṭhādimunibhiḥ sevitam. sarveṣāṃ haṭhatantrāṇāṃ
+haṭhopāyānāṃ sādhanaṃ siddhijanakaṃ yogino gorakṣādyā vidur jānanti.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf yat kiñcit sravate candrād amṛtaṃ divyarūpiṇaḥ
+tat sarvaṃ grasate sūryas tena piṇḍo jarāyutaḥ (3.77)}
+\stopsloka
+
+viparītakaraṇīṃ vivakṣus tadupodghātatvena piṇḍasya jarākaraṇaṃ tāvad āha yat kiñcid iti. divyam
+utkṛṣṭaṃ sudhāmayaṃ rūpaṃ yasya sa tathā tasmād divyarūpiṇaś candrāt somāt tālumūlasthād yat kiñcit
+yat kim apy amṛtaṃ pīyūṣaṃ sravate patati tat sarvaṃ tat pīyūṣaṃ sūryo nābhistho 'nalātmakaḥ
+grasate grāsīkaroti. tad uktaṃ gorakṣanāthena
+
+\startsloka
+nābhideśe sthito nityaṃ  bhāskaro dahanātmakaḥ
+amṛtātmā sthito nityaṃ tālumūle ca candramāḥ
+varṣaty adhomukhaś candro grasaty ūrdhvamukho raviḥ
+karaṇaṃ tac ca kartavyaṃ  yena pīyūṣam āpyate\comment{go.śa. 5758, go.pa. 1.3233 ??} 
+\stopsloka
+
+iti. tena sūryakartṛkāmṛtagrasanena piṇḍo deho jarāyutaḥ jarasā yukto bhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tatrāsti karaṇaṃ divyaṃ sūryasya mukhavañcanam
+gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ (3.78)}
+\stopsloka
+
+tatreti. tatra tadviṣaye sūryasya nābhisthānalasya mukhaṃ vañcyate 'neneti tādṛśaṃ divyam uttamaṃ
+karaṇaṃ vakṣyamāṇamudrākhyam asti. tad gurūpadeśato gurūpadeśāj jñeyaṃ jñātuṃ śakyam. śāstrārthānāṃ
+koṭibhiḥ na tu naiva jñātuṃ śakyam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf ūrdhvanābher adhastālor ūrdhvaṃ bhānur adhaḥ śaśī
+karaṇī viparītākhyā guruvākyena labhyate  (3.79)}
+% \Wthree add: karaṇaṃ viparītākhyaṃ sarvaṃ vyādhivināśanam
+% commentary ends abruptly on 208, folio lost?
+\stopsloka
+
+viparītakaraṇīm āha ūrdhvanābher iti. ūrdhvam uparibhāge nābhir yasya sa ūrdhvanābhiḥ
+tasyordhvanābher adhaḥ adhobhāge tālu tālusthānaṃ yasya so 'dhastālus tasyādhastālor yogina ūrdhvam
+uparibhāge bhānur dahanātmakaḥ sūryo bhavati. adhaḥ adhobhāge śaśy amṛtātmā candro
+bhavati.
+
+prathamāntapāṭhe tu yadā ūrdhvanābhir adhastālur yogī bhavati tadordhvaṃ bhānur adhaḥ śaśī bhavati.
+yadā tadā padayor adhyāhāreṇānvayaḥ. iyaṃ viparītākhyā viparītanāmikā karaṇī. ūrdhvādhaḥsthitayoś
+candrasūryayor adhaūrdhvakaraṇenānvarthā guruvākyena guror vākyenaiva labhyate prāpyate nānyathā.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf nityam abhyāsayuktasya jaṭharāgnivivardhinī
+āhāro bahulas tasya sampādyaḥ sādhakasya ca (3.80)}
+\stopsloka
+
+nityam iti. nityam pratidinaṃ abhyāso 'bhyasanaṃ tasmin yuktasyāvahitasya jaṭharāgnir udarāgnis
+tasya vivardhinī viśeṣeṇa vardhinīti viparītakaraṇīviśeṣaṇam. tasya sādhakasya
+viparītakaraṇyabhyāsina āhāro bhojanaṃ bahulo yathecchaḥ sampādyaḥ sampādanīyaḥ. caḥ pādapūraṇe
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf alpāhāro yadi bhaved agnir dahati tatkṣaṇāt
+  adhaḥśirāś cordhvapādaḥ kṣaṇaṃ syāt prathame dine
+  kṣaṇāc ca kiñcid adhikam abhyasec ca dine dine   (3.81)}\var{\Wthree, \Tue and \EdAd
+  read the ṣaṭpadaśloka here.}
+\stopsloka
+
+alpāhāra iti. yady alpāhāraḥ alpo bhoktum iṣṭād ūna āhāro bhojanaṃ yasya tādṛśo bhavet syāt
+tadāgnir jaṭharānalo dehaṃ tatkṣaṇāt kṣaṇamātrād dahet. śīghraṃ dahed ity arthaḥ.
+
+ūrdhvādhaḥsthitayoś candrasūryayor adhaūrdhvakaraṇakriyām āha adhaḥśirā iti. adhaḥ adhobhāge bhūmau
+śiro yasya so 'dhaḥśirāḥ. karābhyāṃ kaṭipradeśam avalambya bāhumūlād ārabhya kūrparaparyantābhyāṃ
+bāhubhyāṃ skandhābhyāṃ galapṛṣṭhabhāgaśiraḥpṛṣṭha\-bhāgābhyāṃ ca bhūmim avaṣṭabhyādhaḥśirā
+bhavet. ūrdhvam upary antarikṣe pādau yasya sa ūrdhvapādaḥ. prathame dine ārambhadine kṣaṇaṃ
+kṣaṇamātraṃ syāt. dine dine pratidinaṃ kṣaṇāt kiñcid adhikaṃ dvikṣaṇaṃ trikṣaṇam evaṃ
+dinakramavṛddhyābhyased abhyāsaṃ kuryāt.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf  valitaṃ palitaṃ caiva ṣaṇmāsordhvaṃ na dṛśyate      % Pāda d Jyo selects reading of testimonia
+yāmamātraṃ tu yo nityam abhyaset sa tu kālajit (3.82)}
+\stopsloka
+
+viparītakaraṇīguṇān āha valitam iti. valitaṃ carmasaṅkocaḥ palitaṃ keśeṣu śāuklyaṃ ca. ṣaṇṇāṃ
+māsānāṃ samāhāraḥ ṣaṇmāsaṃ. tasmād ūrdhvam upari naiva dṛśyate naivāvalokyate. sādhakasya deha iti
+vākyādhyāhāraḥ. yas tu sādhako yāmamātraṃ praharamātraṃ nityam abhyaset sa tu kālajit kālaṃ mṛtyuṃ
+jayatīti kālajin mṛtyujetā bhavet. etena yogasya prārabdha\-karmapratibandhakatvam api sūcitam. tad
+uktaṃ viṣṇudharme\comment{Viṣṇudharma 100.141}
+
+\startsloka
+svadehārambhakasyāpi karmaṇaḥ saṃkṣayāvahaḥ
+yo yogaḥ pṛthivīpāla śṛṇu tasyāpi lakṣaṇam iti 
+\stopsloka
+
+vidyāraṇyair api jīvanmuktāv uktam yathā prārabdhakarma tattvajñānāt prabalaṃ tathā tasmād api
+karmaṇo yogābhyāsaḥ prabalaḥ. ata eva yoginām uddālakavītahavyādīnāṃ svecchayā dehatyāga upapadyate
+iti. bhāgavate 'py uktaṃ daivaṃ\comment{\EdLo reads dehaṃ, as does \EdMu, but according to the
+  apparatus in \EdLo most mss. read daivaṃ.} jahyāt samādhinā\comment{Bhāgavatapurāṇa 7.15.24b.}
+iti.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf atha vajrolī
+svecchayā vartamāno 'pi yogoktair niyamair vinā
+vajrolīṃ yo vijānāti sa yogī siddhibhājanam (3.83)}
+\stopsloka
+
+vajrolyāṃ pravṛttiṃ janayitum ādau tatphalam āha svecchayeti. yo 'bhyāsī vajrolīṃ vajrolīmudrāṃ
+vijānāti viśeṣeṇa svānubhavena jānāti sa yogī yoge yogaśāstre uktā yogoktās tair yogoktair niyamair
+brahmacaryādibhir vinā ṛte svecchayā nijecchayā vartamāno 'pi vyavaharann api siddhibhājanam
+siddhīnām aṇimādīnāṃ bhājanaṃ pātraṃ bhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tatra vastudvayaṃ vakṣye durlabhaṃ yasya kasyacit
+kṣīraṃ caikaṃ dvitīyaṃ tu nārī ca vaśavartinī (3.84)}
+\stopsloka
+
+tatsādhanopayogi vastudvayam āha tatreti. tatra vajrolyabhyāse vastunor dvayaṃ vastudvayaṃ
+padārthayugmaṃ vakṣye kathayiṣye. kīdṛśaṃ vastudvayam yasya kasyacit yasya kasyāpi dhanahīnasya
+durlabhaṃ duḥkhena labdhum śakyaṃ duḥkhenāpi labdhum aśakyam durlabham iti vā. duḥ syāt
+kaṣṭaniṣedhayoḥ iti kośāt\comment{Not indentified.}. kiṃ tad vastudvayam ity apekṣāyām āha kṣīram iti. ekaṃ vastu
+kṣīraṃ dugdhaṃ pānārthaṃ mehanānantaram indriyanairbalyāt tadbalārthaṃ kṣīrapānaṃ yuktam. kecit tu
+abhyāsakāle ākarṣaṇārtham ity āhuḥ.
+%\comment{Compare Haṭhasaṅketacandrikā f. 39r liṃgachidre ['tha  vivṛte kṣīrākṛṣṭiṃ tato bhajed ???}
+tad ayuktam. tasyāntargatasya ghanībhāve nirgamanāsambhavāt. dvitīyaṃ tu vastu vaśavartinī
+svādhīnā nārī vanitā.
+%\end{vsid}
+
+%\begin{vsid}{hp03_85}
+\startsloka
+{\bf mehanena śanaiḥ samyag ūrdhvākuñcanam abhyaset
+puruṣo 'py athavā nārī vajrolīsiddhim āpnuyāt (3.85)}
+\stopsloka
+
+vajrolīmudrāprakāram āha mehaneneti. mehanena strīsaṅgānantaraṃ bindoḥ \var{kṣaraṇena \Tue \lem
+  rakṣaṇena \Wthree} sādhanabhūtena puruṣaḥ pumān athavā nāry api yoṣid api śanair mandaṃ samyak
+yatnapūrvakam ūrdhvākuñcanam ūrdhvam upary ākuñcanaṃ meṇḍhrākuñcanena bindor upary ākarṣaṇam
+abhyased vajrolīmudrāsiddhim āpnuyāt siddhiṃ gacchet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf yatnataḥ śastanālena phūtkāraṃ vajrakandare
+śanaiḥ śanaiḥ prakurvīta vāyusañcārakāraṇāt (3.86)}
+\stopsloka
+
+atha vajrolyāḥ pūrvāṅgaprakriyām āha yatnata iti. śastaḥ praśasto yo nālas tena śastanālena
+sīsakādinirmitena nālena śanaiḥ śanaiḥ mandaṃ mandaṃ yathāgner dhamanārthaṃ phūtkāraḥ kriyate
+tādṛśaṃ phūtkāraṃ vajrakandare meṇḍhravivare vāyoḥ sañcāraḥ samyag vajrakandare caraṇaṃ gamanaṃ
+tatkāraṇāt taddhetoḥ prakurvīta prakarṣeṇa punaḥ punaḥ kurvīta.
+
+atha vajrolīsādhanaprakriyā. sīsakanirmitāṃ snigdhāṃ meṇḍhrapraveśa\-yogyāṃ caturdaśāṅgulamātrāṃ
+śalākāṃ kārayitvā tasyā meṇḍhre praveśanam abhyaset. prathamadine ekāṅgulamātrāṃ praveśayet
+dvitīyadine dvyaṅgulamātrāṃ tṛtīyadine tryaṅgulamātrām. evaṃ krameṇa vṛddhyā\var{vṛddhyā \lem
+  vṛddhau \EdLo (unclear variants)} dvādaśāṅgulamātrāpraveśe meṇḍhramārgaḥ śuddho bhavati. punas
+tādṛśīṃ caturdaśāṅgulamātrāṃ dvyaṅgulamātrā\var{mātrā \lem mātre mātra}vakrām ūrdhvamukhāṃ
+kārayitvā tāṃ dvādaśāṅgulamātrāṃ praveśayet. vakram ūrdhvamukhaṃ dvyaṅgulamātraṃ bahiḥ
+sthāpayet. tataḥ suvarṇakārasya agnidhamanasādhanībhūtanālasadṛśaṃ nālaṃ gṛhītvā tadagraṃ
+meṇḍhrapraveśitadvādaśāṅgulasya nālasya vakrordhvamukhadvyaṅgulamadhye praveśya phūtkāraṃ
+kuryāt. tena samyak mārgaśuddhir bhavati. tataḥ koṣṇasya jalasya meṇḍhreṇākarṣaṇam
+abhyasyet. jalākarṣaṇe siddhe pūrvoktaślokarītyā bindor ūrdhvākarṣaṇam abhyasyet. bindvākarṣaṇe
+siddhe vajrolīmudrāsiddhiḥ. iyaṃ jitaprāṇasyaiva sidhyati nānyasya.
+khecarīmudrāprāṇajayobhayasiddhau tu samyag bhavati.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf nārībhage patadbindum abhyāsenordhvam āharet
+calitaṃ ca nijaṃ bindum ūrdhvam ākṛṣya rakṣayet (3.87)}
+\stopsloka
+
+evaṃ vajrolyabhyāse siddhe taduttaraṃ sādhanam āha nārībhage iti. nārībhage strīyonau patatīti
+patan pataṃś cāsau binduś patadbindus taṃ patadbindum ratikāle patantaṃ bindum abhyāsena
+vajrolīmudrābhyāsenordhvam upary āhared ākarṣayet patanāt pūrvam eva. yadi patanāt pūrvaṃ bindor
+ākarṣaṇaṃ na syāt tarhi patitam ākarṣayed ity āha calitam iti. calitaṃ nārībhage patitaṃ nijaṃ
+svakīyaṃ bindum cakārāt tadrajaḥ ūrdhvam upary ākṛṣyāhṛtya rakṣayet sthāpayet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf evaṃ saṃrakṣayed binduṃ mṛtyuṃ jayati yogavit
+maraṇaṃ bindupātena jīvanaṃ bindudhāraṇāt (3.88)}
+\stopsloka
+
+vajrolīguṇān āha evam iti. evam uktarītyā binduṃ yaḥ saṃrakṣayet samyak rakṣayet sa yogavit
+yogābhijño mṛtyuṃ jayaty abhibhavati. yato bindoḥ śukrasya pātena patanena maraṇaṃ bhavati. bindor
+dhāraṇaṃ bindudhāraṇaṃ tasmād bindudhāraṇāj jīvanaṃ bhavati. tasmād binduṃ saṃrakṣayed ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf sugandho yogino dehe jāyate bindudhāraṇāt
+yāvad binduḥ sthiro dehe tāvat kālabhayaṃ kutaḥ (3.89)}
+\stopsloka
+
+sugandha iti. yogino vajrolyabhyāsino dehe bindoḥ śukrasya dhāraṇaṃ bindudhāraṇaṃ tasmāt sugandhaḥ
+śobhano gandho jāyate prādurbhavati. dehe yāvad binduḥ sthiras tāvat kālabhayaṃ mṛtyubhayaṃ kutaḥ
+na kuto 'pīty arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf cittāyattaṃ nṝṇāṃ śukraṃ śukrāyattaṃ ca jīvitam
+tasmāc chukraṃ manaś caiva rakṣaṇīyaṃ prayatnataḥ (3.90)}
+\stopsloka
+
+cittāyattam iti. hi yasmāt. nṝṇāṃ śukraṃ vīryaṃ cittāyattam. citte cale calatvāt citte sthire
+sthiratvāc cittādhīnam. jīvitaṃ jīvanaṃ śukrāyattaṃ śukre sthire jīvanāc chukre naṣṭe maraṇāt
+śukrādhīnam. tasmāc chukraṃ binduṃ manaś ca mānasaṃ ca prakṛṣṭād yatnād iti prayatnataḥ rakṣaṇīyam
+eva. avaśyaṃ rakṣaṇīyam ity arthaḥ. evaśabdo bhinnakramaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf ṛtumatyā rajo 'py evaṃ nijaṃ binduṃ ca rakṣayet
+meḍhreṇākarṣayed ūrdhvaṃ samyag abhyāsayogavit (3.91)}
+\stopsloka
+
+ṛtumatyā iti. evaṃ pūrvoktenābhyāsena ṛtur vidyate yasyāḥ sā ṛtumatī tasyā ṛtumatyā ṛtusnātāyāḥ
+striyāḥ\var{striyāḥ \EdAd \lem striyoḥ \Tue \EdLo \EdMu} rajaḥ nijaṃ svakīyaṃ binduṃ ca
+rakṣayet. pūrvoktābhyāsaṃ darśayati meṇḍhreṇeti. abhyāso vajrolyabhyāse sa eva yogo yogasādhanatvāt
+taṃ vettīty abhyāsayogavit meḍhreṇa guhyendriyeṇa samyag yatnapūrvakam ūrdhvam upary
+ākarṣayet. rajo binduṃ ceti karmādhyāhāraḥ. ayaṃ ślokaḥ prakṣiptaḥ. % warum?  \Tue liest kṣisaḥ, not reported.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf atha sahajoliḥ
+sahajoliś cāmarolir vajrolyā bheda ekataḥ
+jale subhasma nikṣipya dagdhagomayasambhavam (3.92)}
+\stopsloka
+
+sahajolyamarolyau vivakṣus tayor vajrolīviśeṣatvam āha sahajoliś ceti. vajrolyā bhedo viśeṣaḥ
+sahajolir amaroliś ca. tatra hetuḥ ekata ekatvād ekaphalatvād ity arthaḥ. ekaśabdād bhāvapradhānāt
+pañcamyās tasil. sahajolim āha jala iti. goḥ purīṣāni gomayāni dagdhāni ca tāni gomayāni
+dagdhagomayāni teṣu sambhava utpattir yasya tad dagdhagomayasambhavam śobhanaṃ bhasma subhasma
+vibhūtiḥ tat jale toye nikṣipya toyamiśraṃ kṛtvety uttaraślokenānveti.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf vajrolīmaithunād ūrdhvaṃ strīpuṃsoḥ svāṅgalepanam
+āsīnayoḥ sukhenaiva muktavyāpārayoḥ kṣaṇāt (3.93)}
+\stopsloka
+
+vajrolīti. vajrolīmudrārthaṃ maithunaṃ tasmād ūrdhvam anantaraṃ sukhenaivānandenaivāsīnayor
+upaviṣṭayoḥ kṣaṇād ratyutsavān muktas tyakto vyāpāro ratikriyā yābhyāṃ tau muktavyāpārau tayor
+muktavyāpārayoḥ strī ca pumāṃś ca strīpuṃsau tayoḥ strīpuṃsayoḥ svāṅgalepanam śobhanāny aṅgāni
+mūrdhalalāṭanetrahṛdayaskandhabhujādīni teṣu lepanam. kuryād iti śeṣaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka 
+{\bf sahajolir iyaṃ proktā śraddheyā yogibhiḥ sadā
+ayaṃ śubhakaro yogo bhogayukto 'pi muktidaḥ (3.94)}
+\stopsloka
+
+sahajolir iti. iyam uktā kriyā sahajolir iti proktā kathitā yogibhir matsyendrādibhiḥ. kīdṛśī sadā
+śraddheyā sarvadā śraddhātuṃ yogyā. ayaṃ sahajolyākhyo yoga upāyaḥ śubhakaraḥ śubhaṃ śreyaḥ
+karotīti śubhakaraḥ. yogaḥ saṃnahanopāyadhyānasaṃgatiyuktiṣu ity abhidhānāt.\comment{Amarakośa
+  3.3.22.} kīdṛśo yogaḥ bhogena strīsaṅgena yukto 'pi muktido mokṣadaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf ayaṃ yogaḥ puṇyavatāṃ dhīrāṇāṃ tattvadarśinām
+nirmatsarāṇāṃ vai sidhyen na tu matsaraśālinām (3.95)}
+\stopsloka
+
+ayaṃ yoga iti. ayam ukto yogaḥ puṇyaṃ vidyate yeṣāṃ te puṇyavantaḥ sukṛtinas teṣāṃ puṇyavatāṃ
+dhīrāṇāṃ dhairyavatāṃ tattvaṃ vāstavikaṃ draṣṭuṃ śīlaṃ yeṣāṃ te\var{draṣṭuṃ śīlaṃ yeṣāṃ te \EdAd
+  \lem paśyantīti \EdLo \Wthree} tattvadarśinas teṣāṃ tattvadarśinām. matsarān niṣkrāntā nirmatsarās
+teṣāṃ nirmatsarāṇām anyaguṇadveṣarahitānām. matsaro 'nyaguṇadveṣa ity amaraḥ\comment{Amarakośa
+  3.3.172}. tādṛśānāṃ puṃsāṃ sidhyet siddhiṃ gacchet. matsaraśālinām matsaravatāṃ tu na sidhyet.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf atha amarolī
+pittolbaṇatvāt prathamāmbudhārāṃ 
+vihāya niḥsāratayāntyadhārām
+niṣevyate śītalamadhyadhārā
+kāpālike khaṇḍamate 'marolī  (3.96)}
+\stopsloka
+
+amarolīm āha pittolbaṇatvād iti. pittenolbaṇotkaṭā pittolbaṇā tasyā bhāvaḥ pittolbaṇatvaṃ
+tasmāt. yathā prathamā pūrvā yāmbunaḥ śivāmbuno dhārā tāṃ vihāya śivāmbunirgamanasamaye kiñcit
+pūrvāṃ dhārāṃ tyaktvā. nirgataḥ sāro yasyāḥ sā niḥsārā tasyā bhāvo niḥsāratā tayā niḥsāratayā
+niḥsāratvenāntyadhārā antyā caramā yā dhārā tāṃ vihāya kiñcid antyāṃ dhārāṃ tyaktvā. śītalā
+pittādidoṣaniḥsāratvarahitā yā madhyadhārā madhyamā dhārā sā niṣevyate nitarāṃ sevyate. khaṇḍo
+yogaviśeṣo mato 'bhimato yasya sa khaṇḍamatas tasmin khaṇḍamate kāpālikasyāyaṃ kāpālike
+khaṇḍakāpālikasampradāya ity arthaḥ. amarolī prasiddheti śeṣaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf amarīṃ yaḥ piben nityaṃ nasyaṃ kurvan dine dine
+vajrolīm abhyaset samyak sāmarolīti kathyate  (3.97)}
+\stopsloka
+
+amarīm iti. amarīṃ śivāmbu yaḥ pumāṇ nityaṃ pibet nasyaṃ kurvan śvāsenāmaryā ghrāṇāntar grahaṇaṃ
+kurvan san dine dine pratidinaṃ mehanena śanaiḥ\comment{HP 3.85.} iti ślokenoktāṃ samyag abhyaset
+sāmarolīti kathyate. kāpālikair iti śeṣaḥ. amarīpānāmarīnasyapūrvikā vajroly amarolīśabdenocyate
+ity arthaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf abhyāsān niḥsṛtāṃ cāndrīṃ vibhūtyā saha miśrayet
+dhārayed uttamāṅgeṣu divyadṛṣṭiḥ prajāyate (3.98)}
+\stopsloka
+
+abhyāsād iti. abhyāsād amarolyabhyāsān niḥsṛtāṃ nirgatāṃ cāndrīṃ candrasyeyaṃ cāndrī tāṃ cāndrīṃ
+sudhāṃ vibhūtyā bhasmanā saha sākaṃ miśrayet saṃyojayet uttamāṅgeṣu
+śiraḥkapālanetraskandhakaṇṭhahṛdayabhujādiṣu dhārayet. bhasmamiśritāṃ cāndrīm iti śeṣaḥ. divyā
+atītānāgatavartamānavyavahitaviprakṛṣṭapadārthadarśanayogyā dṛṣṭir yasya sa divyadṛṣṭir divyadṛk
+prajāyate prakarṣeṇa jāyate. amarīsevanaprakāraviśeṣāḥ śivāmbukalpād\comment{???} avagantavyāḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf puṃso binduṃ samākuñcya samyagabhyāsapāṭavāt
+yadi nārī rajo rakṣed vajrolyā sāpi yoginī (3.99)}
+\stopsloka
+
+puṃso vajrolīsādhanam uktvā nāryās tad āha puṃso bindum iti. samyagabhyāsasya samyagabhyasanasya
+pāṭavaṃ paṭutvaṃ tasmāt puṃsaḥ puruṣasya binduṃ vīryaṃ samākuñcya samyag ākṛṣya nārī strī yadi rajo
+vajrolyā vajrolīmudrayā rakṣet sāpi nārī yoginī praśastayogavatī jñeyā. puṃso bindusamāyuktam iti
+pāṭhe tu etad rajaso viśeṣaṇam.
+%\end{vsid}
+
+%yoginī in the sense of Yoga practioner !
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tasyāḥ kiñcid rajo nāśaṃ na gacchati na saṃśayaḥ
+tasyāḥ śarīre nādaś ca bindutām eva gacchati  (3.100)}
+\stopsloka
+
+nārīkṛtāyā vajrolyāḥ phalam āha tasyā iti. tasyā vajrolyabhyasanaśīlāyā nāryā rajaḥ kiñcit kim api
+svalpam api nāśaṃ na gacchati naṣṭaṃ na bhavati patanaṃ na prāpnotīty arthaḥ. atra saṃśayo na
+sandeho na. tasyā nāryāḥ śarīre nādaś ca bindutām eva gacchati mūlādhārād utthito nādo hṛdayopari
+bindubhāvaṃ gacchati. bindunā sahaikībhavatīty arthaḥ. amṛtasiddhau\comment{Amṛtasiddhi 7.8cd–7.9,
+  7.12 (only a–c), 7.5.ab, 7.2ab}
+
+% JH  nāda becomes bindu ?  usage unclear.
+
+\startsloka
+bījaṃ ca pauruṣaṃ proktaṃ rajaś ca strīsamudbhavam 
+anayor bāhyayogena sṛṣṭiḥ sañjāyate nṛṇām
+yadābhyantaryogaḥ syāt tadā yogīti gīyate
+binduś candramasaḥ prokto rajaḥ sūryamayaṃ tathā
+anayoḥ saṅgamād eva jāyate paramaṃ padam
+svargado mokṣado bindur dharmado 'dharmadas tathā
+tanmadhye devatāḥ sarvās tiṣṭhante sūkṣmarūpataḥ
+\stopsloka
+
+iti.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf sa bindus tad rajaś caiva ekībhūya svadehagau
+vajrolyabhyāsayogena sarvasiddhiṃ prayacchataḥ (3.101)}
+\stopsloka
+
+sa bindur iti. sa puṃso bindus tad rajo nāryā rajaś caiva vajrolīmudrāyā abhyāso vajrolyabhyāsaḥ sa
+eva yogas tenaikībhūya militvā svadehagau svadehe gatau sarvasiddhiṃ prayacchataḥ dattaḥ.
+%\end{vsid}
+
+% JH again singular readings of Jyotsnā prayacchataḥ
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf rakṣed ākuñcanād ūrdhvaṃ yā rajaḥ sā hi yoginī
+atītānāgataṃ vetti khecarī ca bhaved dhruvam (3.102)}
+\stopsloka
+
+rakṣed iti. yā nāry ākuñcanād yonisaṅkocanād ūrdhvam uparisthāne nītvā rajo rakṣet hīti prasiddhaṃ
+yogaśāstre sā yoginī atītānāgataṃ bhūtaṃ bhaviṣyaṃ ca vastu vetti jānāti. dhruvam iti
+niścitam. khe 'ntarikṣe caratīti khecary antarikṣacarī ca bhavet.
+%  reading of the HP mss. is meḍhraṃ, which does not make sense. 
+%   women emphasized by Jyotsnā?  not in early versions in this constellation    ????
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf dehasiddhiṃ ca labhate vajrolyabhyāsayogataḥ
+ayaṃ puṇyakaro yogo bhoge bhukte 'pi muktidaḥ (3.103)}
+\stopsloka
+
+dehasiddhim iti. vajrolyā abhyāsasya yogo yuktis tasmād dehasya siddhiṃ
+rūpalāvaṇyabalavajrasaṃhananatvarūpāṃ\var{tvarūpam \EdLo \Tue \lem svarūpāṃ ??} labhate. ayaṃ yogo
+vajrolyabhyāsayogaḥ puṇyakaro 'dṛṣṭaviśeṣajanakaḥ. kīdṛśo yogaḥ  bhujyata iti bhogo viṣayas tasmin
+bhukte 'pi muktido mokṣadaḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf atha śakticālanam
+kuṭilāṅgī kuṇḍalinī bhujaṅgī śaktir īśvarī
+kuṇḍaly arundhatī caite śabdāḥ paryāyavācakāḥ (3.104)}
+\stopsloka
+
+śakticālanaṃ vivakṣus tadupodghātatayā kuṇḍalīparyāyān tayā mokṣadvāravibhedanādikaṃ cāha
+saptabhiḥ kuṭilāṅgīti. kuṭilāṅgī kuṇḍalinī bhujaṅgī śaktiḥ īśvarī kuṇḍalī arundhatī caite
+sapta śabdāḥ paryāyavācakā ekārthavācakāḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf udghāṭayet kapāṭaṃ tu yathā kuñcikayā haṭhāt
+kuṇḍalinyā tathā yogī mokṣadvāraṃ vibhedayet (3.105)}
+\stopsloka
+
+udghāṭayed iti. yathā yena prakāreṇa pumān kuñcikayā kapāṭārgalotsāraṇasādhanībhūtayā haṭhād balāt
+kapāṭam araram udghāṭayed utsārayet. haṭhād iti dehalīdīpanyāyenobhayatra sambadhyate. tathā tena
+prakāreṇa yogī haṭhād haṭhābhyāsāt kuṇḍalinyā śaktyā mokṣadvāraṃ mokṣasya dvāraṃ prāpakaṃ
+suṣumnāmārgaṃ vibhedayed viśeṣeṇa bhedayet. tayordhvam āyann amṛtatvam eti\comment{Chāndogyopaniṣat
+  8.6.6?} iti śruteḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf yena mārgeṇa gantavyaṃ brahmasthānaṃ nirāmayam
+mukhenācchādya tad dvāraṃ prasuptā parameśvarī (3.106)}
+\stopsloka
+
+yeneti. āmayo rogajanyaṃ duḥkhaṃ duḥkhamātropalakṣaṇaṃ tasmān nirgataṃ nirāmayaṃ duḥkhamātrarahitaṃ
+brahmasthānaṃ brahmāvirbhāvajanakaṃ sthānaṃ brahmasthānaṃ brahmarandhram. tasyāḥ śikhāyā madhye
+paramātmā vyavasthitaḥ\comment{Mahānārāyaṇopaniṣat 11.13.} iti śruteḥ. yena mārgeṇa suṣumnāmārgeṇa
+gantavyaṃ gamanārham asti tad dvāraṃ tasya mārgasya dvāraṃ praveśamārgaṃ mukhenāsyenācchādya
+ruddhvā parameśvarī kuṇḍalinī prasuptā nidritāsti.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf kandordhve kuṇḍalī śaktiḥ suptā mokṣāya yoginām
+bandhanāya ca mūḍhānāṃ yas tāṃ vetti sa yogavit (3.107)}
+\stopsloka
+
+kandordhva iti. kuṇḍalī śaktiḥ kandordhve kandasyoparibhāge yogināṃ mokṣāya suptā mūḍhānāṃ
+bandhanāya suptā. yoginas tāṃ cālayitvā muktā bhavanti. mūḍhās tadajñānād baddhās tiṣṭhantīti
+bhāvaḥ. tāṃ kuṇḍalinīṃ yo vetti sa yogavit. sarveṣāṃ yogatantrāṇāṃ kuṇḍalyāśrayatvād ity arthaḥ.
+%\end{vsid}
+
+% Brahmānanda might have elided the verse 3.106 of the critical edition (ambodhiśaila ...)
+% because it duplicates 3.1
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf kuṇḍalī kuṭilākārā sarpavat parikīrtitā
+sā śaktiś cālitā yena sa mukto nātra saṃśayaḥ (3.108)}
+\stopsloka
+
+kuṇḍalīti. kuṇḍalī śaktiḥ sarpavat bhujaṅgavat kuṭila ākāraḥ svarūpaṃ yasyāḥ sā kuṭilākārā
+parikīrtitā kathitā yogibhiḥ. sā kuṇḍalī śaktir yena puṃsā cālitā mūlādhārād ūrdhvaṃ nītā sa mukto
+'jñānabandhān nivṛttaḥ. atrāsminn arthe saṃśayo na sandeho nāstīty arthaḥ. tayordhvam āyann
+amṛtatvam eti\comment{Chāndogyopaniṣat 8.6.6.} iti śruteḥ.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf gaṅgāyamunayor madhye bālaraṇḍāṃ tapasvinīm
+balātkāreṇa gṛhṇīyāt tad viṣṇoḥ paramaṃ padam (3.109)}
+\stopsloka
+
+gaṅgāyamunayor iti. gaṅgāyamunayor ādhārādheyabhāvena tayor bhāvanād gaṅgāyamunayor abhedena
+bhāvanād vā gaṅgāyamune iḍāpiṅgale tayor madhye suṣumnāmārge tapasvinīṃ niraśanasthiteḥ bālaraṇḍāṃ
+bālaraṇḍāśabdavācyāṃ kuṇḍalīṃ balātkāreṇa haṭhena gṛhṇīyāt. tat tasyā gaṅgāyamunayor madhye
+grahaṇaṃ viṣṇor harer vyāpakasyātmano vā paramaṃ padam paramapadaprāpakam.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf iḍā bhagavatī gaṅgā piṅgalā yamunā nadī
+iḍāpiṅgalayor madhye bālaraṇḍā ca kuṇḍalī (3.110)}
+\stopsloka
+
+gaṅgāyamunādipadārtham āha iḍeti. iḍā vāmaniḥśvāsā nāḍī bhagavaty aiśvaryādisampannā gaṅgā
+gaṅgāpadavācyā piṅgalā dakṣiṇaniḥśvāsā yamunā yamunāśabdavācyā nadī. iḍāpiṅgalayor madhye
+madhyagatā yā kuṇḍalī sā bālaraṇḍā bālaraṇḍāśabdavācyā.
+%\end{vsid}
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf pucche pragṛhya bhujagīṃ suptām udbodhayec ca tām
+nidrāṃ vihāya sā śaktir  ūrdhvam uttiṣṭhate haṭhāt (3.111)}
+\stopsloka
+
+śakticālanam āha puccha iti. suptām nidritāṃ tāṃ bhujagīṃ tāṃ kuṇḍalinīṃ pucche pragṛhya
+pragṛhītvodbodhayet prabodhayet yā śaktiḥ kuṇḍalī nidrāṃ vihāya haṭhād ūrdhvam uttiṣṭhata ity
+anvayaḥ. etad rahasyaṃ tu gurumukhād avagantavyam.
+%\end{vsid}
+
+% Only the Jyotsnā reads śakti, even YCM has ṛjvī as most manuscripts.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf avasthitā caiva phaṇāvatī sā 
+prātaś ca sāyaṃ praharārdhamātram
+prapūrya sūryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā (3.112)}
+\stopsloka
+
+avasthiteti. avasthitārvāk sthitā mūlādhārasthitā phaṇāvatī bhujaṅgī sā kuṇḍalī sūryād āpūrya
+sūryāt pūraṇaṃ kṛtvā paridhānasya yuktiḥ paridhānayuktiḥ tayā paridhānayuktyā pragṛhya
+gṛhītvā. sāyaṃ sūryāstasamaye prātaḥ sūryodayavelāyāṃ nityam aharahaḥ praharasya yāmasyārdhaṃ
+praharārdhaṃ praharārdham eva praharārdhamātram muhūrtadvayamātraṃ paricālanīyā paritaś cālayituṃ
+yogyā. paridhānayuktir deśikād bodhyā.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf ūrdhvaṃ vitastimātraṃ tu vistāraṃ caturaṅgulam
+mṛdulaṃ dhavalaṃ proktaṃ veṣṭitāmbaralakṣaṇam (3.113)}
+\stopsloka
+
+% Strange interpretation by B. because of a misplaced next verse (different in ....)
+
+kandasampīḍanena śakticālanaṃ vivakṣur ādau kandasya sthānaṃ svarūpaṃ cāha ūrdhvam iti. mūlasthānād
+vitastimātraṃ vitastipramāṇam ūrdhvam upari nābhimeṇḍhrayor madhye. etena kandasya sthānam
+uktam. tathā coktaṃ gorakṣaśatake
+
+\startsloka
+  ūrdhvaṃ meḍhrād adho nābheḥ kandayoniḥ khagāṇḍavat
+  tatra nāḍyaḥ samutpannāḥ sahasrāṇāṃ dvisaptatiḥ\comment{Vivekamārtaṇḍa 16} %  yo.ta. 1.25?} 
+\stopsloka
+
+iti. yājñavalkyaḥ\comment{Yogayājñavalkya 4.14, 16–17.}
+
+\startsloka
+gudāt tu dvyaṅgulād ūrdhvam adho meḍhrāc ca dvyaṅgulāt
+dehamadhyaṃ tayor madhyaṃ manuṣyāṇām itīritam
+kandasthānaṃ manuṣyāṇāṃ dehamadhyān navāṅgulam 
+caturaṅgulam utsedam āyāmaṃ ca tathāvidham  % āyāmaḥ ? so etext
+aṇḍākṛtivad ākāraṃ bhūṣitaṃ tat tvagādibhiḥ
+catuṣpadāṃ tiraścāṃ ca dvijānāṃ tundamadhyame
+\stopsloka
+
+iti.
+
+gudād dvyaṅgulopary ekāṅgulaṃ madhyaṃ tasmān navāṅgulaṃ kandasthānaṃ militvā dvādaśāṅgulapramāṇaṃ
+vitastimātraṃ jātam. caturṇām aṅgulānāṃ samāhāraś caturaṅgulaṃ caturaṅgulapramāṇaṃ
+vistāram. vistāro dairghyasyāpy upalakṣaṇam. caturaṅgulaṃ dīrghaṃ ca mṛdulaṃ komalaṃ dhavalaṃ
+śubhraṃ veṣṭitaṃ veṣṭanena khagāṇḍākārīkṛtaṃ yad ambaraṃ vastraṃ tasya lakṣaṇaṃ svarūpam iva
+lakṣaṇaṃ svarūpaṃ yasya tādṛśaṃ proktaṃ kathitam. kandasvarūpaṃ yogibhir iti śeṣaḥ.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf sati vajrāsane pādau karābhyāṃ dhārayed dṛḍham
+gulphadeśasamīpe ca kandaṃ tatra prapīḍayet (3.114)}
+\stopsloka
+
+satīti. vajrāsane\var{vajrāsane \Tue \lem vajrāsane siddhāsane \Wthree} kṛte sati karābhyāṃ hastābhyāṃ
+gulphau pādagranthī tayor deśau pradeśau tayoḥ samīpe gulphābhyāṃ kiñcid upari. tadgranthī ghuṭike
+gulphau ity amaraḥ.\comment{Amarakośa ??} pādau caraṇau dṛḍham gāḍham dhārayed gṛhṇīyāt. cakārād
+dhṛtābhyāṃ pādābhyāṃ tatra kandasthāne kandaṃ prapīḍayet prakarṣeṇa pīḍayet. gulphād ūrdhvaṃ
+karābhyāṃ pādau gṛhītvā nābher adhobhāge kandaṃ pīḍayed ity arthaḥ.
+
+% Brahmānanda has redacted the text by lifting out 3.114 from its position in the description of
+% Uḍḍiyānabandha (where it occurs in all manuscripts available to us) and inserted it in
+% Śakticālana, where it fits perfectly.  Furthermore the Jyotsnā completely omits the verse
+% following in all manuscripts (3.65 of our critical edition).
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf vajrāsane sthito yogī cālayitvā ca kuṇḍalīm
+kuryād anantaraṃ bhastrāṃ kuṇḍalīm āśu bodhayet (3.115)}
+\stopsloka
+
+vajrāsana iti. vajrāsane sthito yogī kuṇḍalīm cālayitvā śakticālanamudrāṃ kṛtvety arthaḥ.
+anantaraṃ śakticālanānantaraṃ bhastrāṃ bhastrākhyaṃ kumbhakaṃ kuryāt. evaṃrītyā kuṇḍalīm śaktim
+āśu śīghraṃ bodhayet prabuddhāṃ kuryāt. vajrāsane śakticālanasya pūrvaṃ vidhāne 'pi punar
+vajrāsanopādānaṃ\var{vajrāsanopādānaṃ \Tue \lem vajrāsanopapādānaṃ \EdLo} śakticālanānantaraṃ
+bhastrāyāṃ vajrāsanam eva kartavyam iti niyamārtham.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf bhānor ākuñcanaṃ kuryāt kuṇḍalīṃ cālayet tataḥ
+mṛtyuvaktragatasyāpi tasya mṛtyubhayaṃ kutaḥ (3.116)}
+\stopsloka
+
+bhānor iti. bhānor nābhideśasthasya sūryasyākuñcanaṃ kuryāt. nābher ākuñcanenaiva
+tasyākuñcanaṃ bhavati. tato bhānor ākuñcanāt kuṇḍalīṃ śaktiṃ cālayet. mṛtyor vaktraṃ mukhaṃ
+gatasyāpi prāptasyāpi tasya puṃso mṛtyubhayaṃ kālabhayaṃ kutaḥ.  na kuto 'pīty arthaḥ.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf muhūrtadvayaparyantaṃ nirbhayaṃ cālanād asau
+ūrdhvam ākṛṣyate kiñcit suṣumṇāyāṃ samudgatā (3.117)}
+\stopsloka
+
+muhūrtadvayeti\var{muhūrtadvayeti \Wthreepc \EdAd \lem muhūrtadvayam iti \Wthreeac \Tue \EdLo \EdMu}. muhūrtayor
+dvayaṃ yugmaṃ ghaṭikācatuṣṭayātmakaṃ tatparyantaṃ tadavadhi nirbhayaṃ niḥśaṅkaṃ cālanād asau śaktiḥ
+suṣumnāyāṃ samudgatā satī kiñcid ūrdhvam ākṛṣyate ākṛṣṭā bhavati.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tena kuṇḍalinī tasyāḥ suṣumṇāyā mukhaṃ dhruvam
+jahāti tasmāt prāṇo 'yaṃ suṣumṇāṃ vrajati svataḥ (3.118)}
+\stopsloka
+
+teneti. tenordhvākarṣaṇena kuṇḍalinī tasyāḥ prasiddhāyāḥ suṣumnāyā mukhaṃ praveśamārgaṃ dhruvaṃ
+niścitaṃ jahāti tyajati. tasmān mārgatyāgād ayaṃ prāṇo vāyuḥ svataḥ svayam eva suṣumnāṃ vrajati
+gacchati. suṣumnāmukhāt prāg eva kuṇḍalinyā nirgatatvād iti bhāvaḥ.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tasmāt sañcālayen nityaṃ sukhasuptām arundhatīm
+tasyāḥ sañcālanenaiva yogī rogaiḥ pramucyate (3.119)}
+\stopsloka
+
+tasmād iti. yasmāc chakticālanena prāṇaṃ suṣumnāṃ vrajati tasmāt sukhena suptā sukhasuptā taṃ
+sukhasuptām arundhatīm śaktiṃ nityaṃ pratidinaṃ sañcālayet samyak cālayet. tasyāḥ śakteḥ
+sañcālanenaiva sañcālanamātreṇa yogī rogaiḥ kāsaśvāsajvarā\var{jvarā \Tue \EdAd \lem jarā EdMU
+  \EdLo}dhibhiḥ pramucyate prakarṣeṇa mukto bhavati.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf yena sañcālitā śaktiḥ sa yogī siddhibhājanam 
+kim atra bahunoktena kālaṃ jayati līlayā (3.120)}
+\stopsloka
+
+yeneti. yena yoginā śaktiḥ kuṇḍalī sañcālitā sa yogī siddhīnām aṇimādīnāṃ bhājanam pātraṃ
+bhavati. atrāsminn arthe bahunoktena bahupraśaṃsanena kim  na kim apīty arthaḥ. kālaṃ mṛtyuṃ
+līlayā krīḍayānāyāsena jayaty abhibhavatīty arthaḥ.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf brahmacaryaratasyaiva nityaṃ hitamitāśinaḥ
+maṇḍalād dṛśyate siddhiḥ kuṇḍalyabhyāsayoginaḥ (3.121)}
+\stopsloka
+
+brahmacaryeti. brahmacaryaṃ śrotrādibhiḥ sahopasthasaṃyamas tasmin ratasya tatparasya nityaṃ
+sarvadā hitaṃ pathyaṃ mitaṃ caturthāṃśavarjitam aśnātīti tasya kuṇḍalyabhyāsaḥ śakticālanābhyāsaḥ
+sa eva yogaḥ so 'syāstīti sa tathā tasya maṇḍalāc catvāriṃśaddinātmakād anantaraṃ siddhiḥ
+prāṇāyāmasiddhir dṛśyate.
+
+\startsloka
+{\bf nāsādakṣiṇamārgavāhipavanāt prāṇo 'tidīrghīkṛtaś
+candrābhaḥ paripūritāmṛtatanuḥ prāg ghaṇṭikāyās tataḥ 
+chittvā kālaviśālavahnivaśagaṃ bhrūrandhranāḍīgataṃ
+tat kāyaṃ kurute punar navataraṃ chinnadrumaṃ skandhavat}\comment{drumaṃ neutre?
+         Amaraughaśāsana (part of the text in all manuscripts 4.114) check marmasthānas?? check with V10} 
+\stopsloka
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf kuṇḍalīṃ cālayitvā tu bhastrāṃ kuryād viśeṣataḥ
+evam abhyasato nityaṃ yamino yamabhīḥ kutaḥ (3.122)}
+\stopsloka
+
+% abhyasya vs abhyasa
+
+kuṇḍalīm iti. kuṇḍalīṃ cālayitvā śakticālanaṃ kṛtvā. athānantaram eva bhastrāṃ bhastrākhyaṃ
+kumbhakaṃ kuryāt. nityaṃ pratidinam. evam uktaprakāreṇābhyasato yamino yogino yamabhīr yamād bhayaṃ
+kutaḥ na kuto 'pīty arthaḥ. yogino dehatyāgasya svādhīnatvād iti tātparyam.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf dvāsaptatisahasrāṇāṃ nāḍīnāṃ malaśodhane
+kutaḥ prakṣālanopāyaḥ kuṇḍalyabhyasanād ṛte (3.123)}
+\stopsloka
+
+dvāsaptatīti. dvābhyām adhikā saptatiḥ dvāsaptatisaṅkhyākāni sahasrāṇi dvāsaptatisahasrāṇi
+teṣāṃ tatsaṅkhyākānāṃ nāḍīnāṃ malaśodhane kartavye sati kuṇḍalyabhyasanāc chakticālanābhyāsād
+ṛte vinā kutaḥ prakṣālanopāyaḥ na kuto 'pi. śakticālanābhyāsenaiva sarvāsāṃ nāḍīnāṃ
+malaśodhanaṃ bhavatīty abhiprāyaḥ.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf iyaṃ tu madhyamā nāḍī dṛḍhābhyāsena yoginām
+āsanaprāṇasaṃyāmamudrābhiḥ saralā bhavet (3.124)}
+\stopsloka
+
+iyaṃ tv iti. iyaṃ madhyamā nāḍī suṣumnā yoginām dṛḍhābhyāsenāsanaṃ svastikādi prāṇasaṃyāmaḥ
+prāṇāyāmo mudrā mahāmudrādikās taiḥ saralā ṛvī bhavet.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf abhyāse tu vinidrāṇāṃ mano dhṛtvā samādhinā
+rudrāṇī vāparā mudrā bhadrāṃ siddhiṃ prayacchati (3.125)}
+\stopsloka
+
+abhyāsa iti. samādhinetaravṛttinirodharūpeṇaikāgryeṇa mano dhṛtvāntaḥkaraṇaṃ dhāraṇāniṣṭhaṃ
+kṛtvābhyāse manaḥsthitau yatne vigatā nidrā yeṣāṃ te tathā teṣām. nidrāpadam ālasyopalakṣaṇam
+analasānām ity arthaḥ. rudrāṇī śāmbhavī mudrā vā athavāparānyā unmanyādikā bhadrāṃ śubhāṃ siddhiṃ
+yogasiddhiṃ prayacchati dadāti. etena haṭhayogopakārako rājayogaḥ proktaḥ.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf rājayogaṃ vinā pṛthvī rājayogaṃ vinā niśā
+rājayogaṃ vinā mudrā vicitrāpi na rājate (3.126)}
+\stopsloka
+
+rājayogaṃ vinā āsanādīnāṃ vaiyarthyam aupacārikaślokenāha rājayogam.
+iti. vṛttyantaranirodhapūrvakātmagocaradhārāvāhikanirvikalpavṛttī rājayogaḥ. haṭhaṃ vinā
+rājayogaḥ\comment{HP 2.76} ity atra sūcitas tatsādhanābhyāso vā taṃ vinā tam ṛte. pṛthvīśabdena
+sthairyaguṇayogād āsanaṃ lakṣyate. tan na rājate na śobhate. rājayogaṃ vinā
+paramapuruṣārthaphalāsiddher iti hetur agre 'pi yojanīyaḥ.  rājayogaṃ vinā niśeva niśā kumbhako na
+rājate niśāyāṃ prāyeṇa janasañcārābhāvāt. niśāśabdena prāṇasañcārābhāvalakṣaṇaḥ kumbhako
+lakṣyate. rājayogaṃ vinā mudrā mahāmudrādirūpā vicitrāpi vividhāpi vilakṣaṇāpi vā na rājate na
+śobhate.
+
+pakṣāntare rājño nṛpasya yogo rājayogo rājasambandhas taṃ vinā pṛthvī bhūmir na rājate. śāstāraṃ 
+vinā bhūmau nānopadravasambhavāt. rājā candraḥ. somo 'smākaṃ brāhmaṇānāṃ
+rājā\comment{Taittirīyasaṃhitā 1.8.10.2.} iti śruteḥ. tasya yogaṃ sambandhaṃ vinā niśā rātrir na
+rājate. rājayogaṃ vinā nṛpasambandhaṃ vinā mudrā rājabhiḥ patreṣu kriyamāṇaś
+cihnaviśeṣaḥ. vicitrāpi. pṛthvīpakṣe ratnādijanakatvena vilakṣaṇāpi niśāpakṣe grahanakṣatrādibhir
+vicitrāpi mudrāpakṣe rekhābhir vicitrāpi na rājate.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf mārutasya vidhiṃ sarvaṃ manoyuktaṃ samabhyaset
+itaratra na kartavyā manovṛttir manīṣiṇā (3.127)}
+\stopsloka
+
+mārutasyeti. mārutasya vāyoḥ sarvaṃ vidhiṃ kumbhakamudrāvi\-dhānaṃ manoyuktaṃ manasā yuktaṃ
+samabhyaset samyag abhyaset manīṣiṇā buddhimatā puṃsā. itaratra mārutasya vidher anyasmin viṣaye
+manovṛttir manaso vṛttir na kartavyā na kāryā.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf iti mudrā daśa proktā ādināthena śambhunā
+ekaikā tāsu yamināṃ mahāsiddhipradāyinī (3.128)}
+\stopsloka
+
+mudrā upasaṃharati itīti. ādināthena sarveśvareṇa śambhunā śaṃ sukhaṃ bhavaty asmād iti śambhuḥ
+tena. ity uktarītyā daśa daśasaṅkhyākā mudrāḥ proktāḥ kathitāḥ. tāsu mudrāsu madhye ekaikāpi
+pratyekāpi yā kācana mudrā yamināṃ yamavatāṃ yogināṃ mahāsiddhipradāyinī aṇimādipradātrī
+kaivalyapradātrī vā.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf upadeśaṃ hi mudrāṇāṃ yo datte sāmpradāyikam
+sa eva śrīguruḥ svāmī sākṣād īśvara eva saḥ (3.129)}
+\stopsloka
+
+mudropadeṣṭāraṃ guruṃ praśaṃsati upadeśam iti. yaḥ pumān mudrāṇāṃ mahāmudrādīnāṃ sampradāyāt
+yogināṃ guruparamparārūpād āgataṃ sāmpradāyikam upadeśaṃ datte dadāti sa eva sa pumān eva śrīguruḥ
+śrīmān guruḥ sarvagurubhyaḥ śreṣṭha ity arthaḥ. svāmī prabhuḥ sa eva sākṣāt pratyakṣa īśvara eva
+saḥ īśvarābhinna eva sa ity arthaḥ.
+
+%\begin{vsid}{hp03_00}
+\startsloka
+{\bf tasya vākyaparo bhūtvā mudrābhyāse samāhitaḥ
+aṇimādiguṇaiḥ sārdhaṃ labhate kālavañcanam (3.130)}
+\stopsloka
+
+tasya mudrāṇām upadeṣṭur guror vākyaparo vākyam āsanakumbhakādyanuṣṭhānaviṣayakaṃ
+ca tasmin paras tatparaḥ tatparaś cādaravān. ādaraś cāvahitatayā karaṇam. bhūtvā sambhūya
+mudrāṇāṃ mahāmudrādīnāṃ abhyāsaḥ paunaḥpunyenāvartanaṃ tasmin mudrābhyāse samāhitaḥ sāvadhānaḥ
+puruṣo 'ṇimādiguṇair aṇimādisiddhibhiḥ sārdhaṃ sākaṃ kālasya mṛtyor vañcanaṃ pratāraṇaṃ
+labhate prāpnoti.
+
+% iti śrīsvātmārāmayogīndraviracitāyāṃ haṭhayogapradīpikāyāṃ mudrāvidhānaṃ nāma tṛtīyopadeśaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf namaḥ śivāya gurave nādabindukalātmane
+nirañjanapadaṃ yāti nityaṃ tatra parāyaṇaḥ (4.1)}
+\stopsloka
+
+prathamadvitīyatṛtīyopadeśoktānām āsanakumbhakamudrāṇāṃ phala\-bhūtaṃ rājayogaṃ vivakṣuḥ svātmārāmaḥ
+śreyāṃsi bahuvighnānīti tatra vighnabāhulyasya sambhavāt tannivṛttaye śivābhinnagurunamas\-kārātmakaṃ
+maṅgalam ācarati nama iti.
+
+śivāya sukharūpāyeśvarābhinnāya vā. tad uktaṃ
+
+\startsloka
+  namas te nātha bhagavan śivāya gururūpiṇe\comment{The verse occurs in many texts, e.g.\ {\em
+      Śrīvidyārṇavatantra} 1.383}
+\stopsloka
+
+iti. gurave deśikāya yad vā gurave sarvāntaryāmitayā nikhilopadeṣṭre śivāyeśvarāya. tathā ca
+pātañjalasūtram
+
+\startsloka
+  pūrveṣām api guruḥ kālenānavacchedāt\comment{Yogasūtra 1.26.} iti.
+\stopsloka
+
+namaḥ prahvībhāvo 'stu. kīdṛśāya śivāya gurave nādabindukalātmane. kāṃsyaghaṇṭānirhrādavad
+anuraṇanaṃ nādaḥ bindur anusvārottarabhāvī dhvaniḥ kalā nādaikadeśas tā ātmā svarūpaṃ yasya sa
+tathā tasmai. nādabindukalātmanā vartamānāyety arthaḥ. tatra nādabindukalātmani śive gurau nityaṃ
+pratidinaṃ parāyaṇo 'vahitaḥ pumān. etena nādānusandhānaparāyaṇa ity uktam. pūrvapādena guruśivayor
+abhedaḥ sūcitaḥ. añjanaṃ māyopādhis tadrahitaṃ nirañjanaṃ śuddham padyate gamyate yogibhir iti
+padaṃ brahma yāti prāpnoti. tathā ca vakṣyati nādānusandhānasamādhibhājām\comment{HP 4.81.}
+ityādinā.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf athedānīṃ pravakṣyāmi samādhikramam uttamam
+mṛtyughnaṃ ca sukhopāyaṃ brahmānandakaraṃ param (4.2)}
+\stopsloka
+
+samādhikramaṃ pratijānīte atheti. athāsanakumbhakamudrākathanānantar\-am idānīm asminn avasare
+samādhikramam pratyāhārādirūpaṃ pravakṣyāmi prakarṣeṇa vivicya vakṣyāmīty anvayaḥ. kīdṛśaṃ
+samādhikramam uttamam śrīādināthoktasapādakoṭisamādhiprakāreṣūtkṛṣṭam.
+
+punaḥ kīdṛśam mṛtyuṃ kālaṃ hanti nivārayatīti mṛtyughnaṃ svecchayā dehatyāgajanakaṃ
+tattvajñānodayamanonāśavāsanākṣayaiḥ sukhasya jīvanmuktisukhasyopāyaṃ prāptisādhanam.
+
+punaḥ kīdṛśam paraṃ brahmānandakaraṃ prārabdhakarmakṣaye sati jīvabrahmaṇor
+abhedenātyantikabrahmānandaprāptirūpavidehamuktikaram. tatra nirodhasamādhinā cittasya
+sasaṃskārāśeṣavṛttinirodhe śāntaghoramūḍhāvasthānivṛttau
+
+\startsloka
+jīvann eva hi vidvān harṣaśokābhyāṃ vimucyate\comment{quotation untraced}
+\stopsloka
+
+ity ādiśrutyuktanirvikārasvarūpāvasthitirūpā jīvanmuktir bhavati. parama\-muktis tu
+prārabdha\var{prārabdha \lem prānta \Tue \Wthree prāpta \EdMu \EdLo}bhogānte 'ntaḥkaraṇaguṇānāṃ
+pratiprasavenaupādhikarūpātyantikanivṛttāv ātyantikasvarūpāvasthānaṃ
+pratiprasava\-siddham. vyutthānanirodhasamādhisaṃskārā manasi līyante. mano 'smitāyām asmitā mahati
+mahān pradhāna iti cittaguṇānāṃ pratiprasavaḥ pratisargaḥ svakāraṇe layaḥ.
+
+nanu jīvanmuktasya vyutthāne brāhmaṇo 'haṃ manuṣyo 'ham ity ādi vyavahāradarśanāc cittādibhir
+aupādhikabhāvajananād amlena dugdhasyeva svarūpacyutiḥ syād iti cen na samprajñātasamādhau
+anubhūtātmasaṃskārasya nirodhasaṃskārasya ca tadānīṃ sattvāt. tābhyāṃ ca vyutthānasaṃskārasya
+dagdhabījakalpatvād vyutthānavyavahārasyā\-tāttvi\-ka\-tvaniścayāt. atāttvikānyathābhāvasya
+vikāritvāprayojakatvāt. amlena dugdhasya dadhibhāvas tu tāttvika iti dṛṣṭāntavaiṣamyāc
+ca. puruṣasya tv antaḥkaraṇopādhiko 'haṃ brāhmaṇa ityādivyavahāraḥ sphaṭikasya
+japākusumasaṃnidhānopādhir aruṇimeva na tāttvikaḥ. japākusumāpagame sphaṭikasya svasvarūpasthitivad
+antaḥkaraṇasya sakalavṛttinirodhe svasvarūpāvasthitir acyutaiva puruṣasya.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf rājayogaḥ samādhiś ca unmanī ca manonmanī
+amaratvaṃ layas tattvaṃ śūnyāśūnyaṃ paraṃ padam (4.3)
+amanaskaṃ tathādvaitaṃ nirālambaṃ nirañjanam
+jīvanmuktiś ca sahajā turyā cety ekavācakāḥ (4.4)}
+\stopsloka
+
+samādhiparyāyān viśeṣeṇāha rājayoga ityādinā ślokadvayena. spaṣṭam.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf salile saindhavaṃ yadvat sāmyaṃ bhajati yogataḥ
+tathātmamanasor aikyaṃ samādhir abhidhīyate (4.5)}
+\stopsloka
+
+tribhiḥ samādhim āha salila iti. yadvat yathā saindhavaṃ sindhudeśodbhavaṃ lavaṇaṃ salile jale
+yogataḥ saṃyogāt sāmyaṃ salilasāmyaṃ salilaikyaṃ bhajati prāpnoti tathā tadvad ātmā ca manaś
+cātmamanasī tayor ātmamanasor aikyam ekākāratā. ātmani dhāritaṃ mana ātmākāraṃ sadātmasāmyaṃ
+bhajati tādṛśam ātmamanasor aikyaṃ samādhir abhidhīyate samādhiśabdenocyate ity arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf yadā saṃkṣīyate prāṇo mānasaṃ ca pralīyate
+tadā samarasatvaṃ ca samādhir abhidhīyate (4.6)}
+\stopsloka
+
+yadā saṃkṣīyate iti. yadā yasmin kāle jitaprāṇasya yoginaḥ kumbhakakāle prāṇaḥ śarīrāntarvartī
+vāyuḥ kumbhakena niruddhavṛttikaḥ prāṇaḥ kṣīṇa ity ucyate saṃkṣīyate samyak kṣīṇo bhavati.
+
+mānasaṃ ca manaś ca pralīyate prakarṣeṇa layaṃ prāpnoti ātmani sthitasya manasaḥ ātmākāratā layaḥ
+tadā tasmin kāle samarasatvam ekākāratvaṃ manasaś ca ātmani sthitasya
+ātmākārapariṇāmenātmākāratvam. japākusumasthasya sphaṭikamaṇer japākusumākāratvavat. tathā ca
+pātañjalaṃ sūtraṃ
+
+% Strange interpretation: see also sāmarasyamekībhāvaḥ (Nityaṣoḍaśik), vedyavedakasāmarasyaṃ
+% (śitikaṇṭha),evākhaṇḍacidābhāsarūpo'nākhyakramamayo bhedavigalanena sāmarasyamayo, ekībhūtaṃ
+% sāmarasyaṃ prāptaṃ (Sambapancasika), teṣāṃ mātṛmānameyānāṃ śaṅghaṭṭaḥ" sāmarasyaṃ tataḥ TĀV 6.103
+
+\startsloka
+  kṣīṇavṛtter abhijātasyeva maṇer grahītṛgrahaṇagrāhyeṣu
+  tatsthatadañjanatā samāpattiḥ\comment{Yogasūtra 1.41}
+\stopsloka
+
+cakāreṇa samarasatvasya sthairyaṃ samuccīyate. ata eva pātañjalabhāṣye
+
+\startsloka
+  vikṣipte cetasi vikṣepopasarjanībhūtaḥ samādhir na
+  yogapakṣe vartate\comment{Yogasūtrabhāṣya 1.1}
+\stopsloka
+
+ity uktam. itthaṃ cātmākāramanovṛtteḥ sthirībhāvaḥ samādhir abhidhīyate ity arthaḥ. uktābhyāṃ
+dvābhyāṃ ślokābhyāṃ samprajñātaḥ samādhir uktaḥ. samyak saṃśayaviparyayarahitatvena prajñāyate
+prakarṣeṇāparokṣatayānubhūyate bhāvyasya svarūpam aneneti samprajñātaḥ samādhiḥ
+bhāvanāviśeṣaḥ. bhāvanā ca bhāvyasya viṣayāntaraparihāreṇa cetasi punaḥ punar
+niveśanam samprajñātaḥ sālambanaḥ sabījo layaḥ. ete samprajñātasya prasiddhāḥ paryāyāḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf tatsamaṃ ca dvayor aikyaṃ jīvātmaparamātmanoḥ
+pranaṣṭasarvasaṅkalpaḥ samādhiḥ so 'bhidhīyate (4.7)}   % pranaṣṭa conj Jyo  all VM read samastanaṣṭa-
+\stopsloka
+
+tatsamam iti. tatsamaṃ pūrvoktasamaṃ dvayor ubhayor jīvātmā antaḥkaraṇopahitaṃ caitanyam paramātmā
+saccidānandalakṣaṇaṃ deśakālavastuparicchedaśūnyaṃ caitanyaṃ tayor aikyam abhedaḥ. aikyaṃ
+viśinaṣṭi pranaṣṭasarvasaṅkalpa iti. prakarṣeṇa naṣṭā niruddhāḥ sarve dhyātṛdhyānadhyeyādirūpāḥ
+saṅkalpāḥ samyak kalpanāḥ yasmin tat tādṛśam aikyam. puṃliṅgapāṭhe samādher viśeṣaṇam. sa
+yogaśāstraprasiddhaḥ samādhiḥ nirbījaḥ samādhiḥ abhidhīyate kathyate. tathā ca pātañjalaṃ
+sūtram
+
+\startsloka
+tasyāpi nirodhe sarvanirodhān nirbījaḥ samādhiḥ\comment{Yogasūtra 1.51}
+\stopsloka
+
+iti anenāsamprajñātaḥ samādhir uktaḥ. anvarthaś cāyaṃ na kim api samyag dhyātṛdhyānadhyeyatvādinā
+prajñāyate 'sminn ity asamprajñātaḥ. asamprajñāto nirālambo nirbījo rājayogo nirodhaś caite
+asamprajñātasya prasiddhāḥ paryāyāḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf rājayogasya māhātmyaṃ ko vā jānāti tattvataḥ
+jñānaṃ muktiḥ sthitiḥ siddhir guruvākyena labhyate (4.8)}
+\stopsloka
+
+atha rājayogapraśaṃsā rājayogasyeti. rājayogasyānantaram evoktasya māhātmyaṃ prabhāvaṃ tattvato
+vastutaḥ ko vā jānāti. na ko 'pi jānātīty arthaḥ. tattvato vaktum aśakyatve 'py ekadeśena
+rājayogaprabhāvam āha jñānam iti. jñānam svasvarūpāparokṣānubhavo muktir videhamuktiḥ sthitir
+nirvikārasvarūpāvasthitirūpā jīvanmuktiḥ siddhir aṇimādir guruvākyena guruvacasā
+labhyate. rājayogād iti śeṣaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf durlabho viṣayatyāgo durlabhaṃ tattvadarśanam
+durlabhā sahajāvasthā sadguroḥ karuṇāṃ vinā (4.9)}
+\stopsloka
+
+durlabha iti. viśeṣeṇa sinvanty avabadhnanti pramātāraṃ svasaṅgeneti\var{svasaṅgeneti \Wthree \Tue
+  \EdAd \EdMu \lem svasaṃvegena \EdLo (no variants given)} viṣayā aihikā dārādaya āmuṣmikāḥ
+sudhādayas teṣāṃ tyāgo bhogecchābhāvo durlabhaḥ. tattvadarśanam ātmāparokṣānubhavaḥ
+durlabham. sahajāvasthā turyāvasthā. sadguroḥ
+
+\startsloka
+dṛṣṭiḥ sthirā yasya vinaiva dṛśyam\comment{Amanaska 44}
+\stopsloka
+
+iti vakṣyamāṇalakṣaṇasya karuṇāṃ dayāṃ vineti sarvatra sambadhyate. durlabhā labdhum aśakyā duḥ
+syāt kaṣṭaniṣedhayoḥ iti koṣaḥ. gurukṛpayā tu sarvaṃ sulabham iti bhāvaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf vividhair āsanaiḥ kumbhair vicitraiḥ karaṇair api
+prabuddhāyāṃ mahāśaktau prāṇaḥ śūnye pralīyate (4.10)}
+\stopsloka
+
+vividhair iti. vividhair anekavidhair āsanaiḥ matsyendrādipīṭhaiḥ vicitrair nānāvidhaiḥ kumbhaiḥ
+kumbhakaiḥ. vicitrair iti kākākṣigolakanyāyenobhayatra sambadhyate. vicitrair anekaprakārakaiḥ
+karaṇair haṭhasiddhau prakṛṣṭopakārakair mahāmudrādibhir mahāśaktau kuṇḍalinyāṃ prabuddhāyāṃ
+gatanidrāyāṃ satyāṃ prāṇo vāyuḥ śūnye brahmarandhre pralīyate pralayaṃ prāpnoti. vyāpārābhāvaḥ
+prāṇasya pralayaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf utpannaśaktibodhasya tyaktaniḥśeṣakarmaṇaḥ
+yoginaḥ sahajāvasthā svayam eva prajāyate (4.11)}
+\stopsloka
+
+utpanneti. utpanno jātaḥ śaktibodhaḥ kuṇḍalībodho yasya tasya tyaktāni parihṛtāni niḥśeṣāṇi
+samagrāṇi karmāṇi yena tasya yoginaḥ āsanena kāyikavyāpāre tyakte prāṇendriyeṣu vyāpāras
+tiṣṭhati. kumbhakena prāṇanirodhe prāṇendriyavyāpāre tyakte manasi vyāpāras
+tiṣṭhati. pratyāhāradhāraṇādhyānasamprajñātasamādhibhir mānasikavyāpāre tyakte buddhau vyāpāras
+tiṣṭhati. asaṅgo hy ayaṃ puruṣaḥ\comment{Bṛhad\-āraṇya\-kopani\-ṣat 4.3.15} iti śruter apariṇāmī śuddhaḥ
+puruṣaḥ sattvaguṇātmikā pariṇāminī buddhir iti. paravairāgyeṇa dīrghakālasamprajñātābhyāsenaiva vā
+buddhivyāpāre parityakte nirvikārasvarūpāvasthitir bhavati. saiva sahajāvasthā turyāvasthā
+jīvanmuktiḥ svayam eva prayatnāntaraṃ vinaiva prajāyate prādurbhavati. yena tyajasi tat tyaja iti
+niḥsaṅgaḥ prajñayā bhavet\comment{Māṇḍūkyakārikā 3.45??} iti ca śruteḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf suṣumṇāvāhini prāṇe śūnye viśati mānase
+tadā sarvāṇi karmāṇi nirmūlayati yogavit (4.12)}
+\stopsloka
+
+suṣumṇeti. prāṇe vāyau suṣumṇāvāhini madhyanāḍīpravāhiṇi sati mānase antaḥkaraṇe śūnye
+deśakālavastuparicchedahīne brahmaṇi viśati sati tadā tasmin kāle yogavit cittavṛttinirodhajñaḥ
+sarvāṇi karmāṇi saprārabdhāni nirmūlāni karoti nirmūlayati. nirmūlaśabdāt tat
+karoti\comment{gaṇasūtra?}  iti ṇic.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf amarāya namas tubhyaṃ so 'pi kālas tvayā jitaḥ
+patitaṃ vadane yasya jagad etac carācaram (4.13)}
+\stopsloka
+
+samādhyabhyāsena prārabdhakarmaṇo py abhibhavāj jitakālaṃ yoginaṃ namaskaroti amarāyeti. na mriyate
+ity amaras tasmai amarāya cirajīvine tubhyaṃ yogine namaḥ. so 'pi durvāro 'pi kālo mṛtyus tvayā
+yoginā jito 'bhibhūtaḥ. idaṃ vākyaṃ namaskaraṇe hetuḥ. sa kaḥ yasya kālasya vadane mukhe etad
+dṛśyamānaṃ carācaraṃ sthāvarajaṅgamaṃ jagat saṃsāraḥ patitaḥ so 'pi jagadbhakṣako 'pīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf citte samatvam āpanne vāyau vrajati madhyame
+tadāmarolī vajrolī sahajolī prajāyate (4.14)}
+\stopsloka
+
+pūrvoktam amarolyādikaṃ samādhisiddhāv eva siddhyatīti samādhinirūpa\-ṇānantaraṃ samādhisiddhau
+tatsiddhir ity āha citta iti. citte 'ntaḥkaraṇe samatvam dhyeyākāravṛttipravāhavattvam āpanne
+prāpte sati vāyau prāṇe madhyame suṣumnāyāṃ vrajati satīti cittasamatve hetuḥ. tadā tasmin kāle
+amarolī vajrolī sahajolī ca pūrvoktāḥ prajāyante. nājitaprāṇasya na cājitacittasya siddhyantīti
+bhāvaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf jñānaṃ kuto manasi sambhavatīha tāvat
+prāṇo 'pi jīvati mano mriyate na yāvat
+prāṇo mano dvayam idaṃ vilayaṃ nayed yo
+mokṣaṃ sa gacchati naro na kathañcid anyaḥ (4.15)}
+\stopsloka
+
+haṭhābhyāsaṃ vinā jñānaṃ mokṣaś ca na siddhyatīty āha jñānam iti. yāvat prāṇo jīvati apiśabdād
+indriyāṇi jīvanti na tu mriyante yāvan mano na mriyate kintu jīvaty eva. iḍāpiṅgalābhyāṃ vahanaṃ
+prāṇasya jīvanam svasvaviṣayagrahaṇam indriyāṇāṃ jīvanam nānāviṣayākāravṛttyutpādanaṃ manaso
+jīvanam tattadabhāvaḥ tattanmaraṇam atra vivakṣitam na tu svarūpatas teṣāṃ nāśas. tāvan manasy
+antaḥkaraṇe jñānam ātmāparokṣānubhavaḥ kutaḥ sambhavati na kuto'pi. prāṇendriyamanovṛttīnāṃ
+jñānapratibandhakatvād iti bhāvaḥ. prāṇo manaḥ idaṃ dvayaṃ yo yogī vilayaṃ nāśaṃ nayet sa mokṣam
+ātyantikasvarūpāvasthānalakṣaṇaṃ gacchati prāpnoti. brahmarandhre nirvyāpārasthitiḥ prāṇasya
+layaḥ. dhyeyākārāveśāt viṣayāntareṇāpariṇamanaṃ manaso layaḥ. anyaḥ alīnaprāṇo 'līnamanāś ca
+kathañcid upāyaśatenāpi na mokṣaṃ prāpnotīty arthaḥ. tad uktaṃ yogabīje
+
+\startsloka
+nānāvidhair vicārais tu na sādhyaṃ jāyate manaḥ
+tasmāt tasya jayopāyaḥ prāṇasya jaya eva hi \  iti\comment{Yogabīja 80?}
+nānāmārgaiḥ suduṣprāpaṃ kaivalyam paramaṃ padam
+siddhamārgeṇa labhyeta nānyathā śivabhāṣitam \ iti ca \comment{Yogabīja 8?}
+\stopsloka
+
+siddhamārgo yogamārgaḥ. etena yogaṃ vinā jñānaṃ mokṣaś ca na siddhyatīti
+siddham. śrutismṛtītihāsapurāṇādiṣu cedaṃ prasiddham. tathā hi atha taddarśanābhyupāyo yogaḥ
+iti\comment{atha taddarśanābhyopāyo yogaḥ (quotation not traced)} taddarśanam ātmadarśanam.
+
+\startsloka
+adhyātmayogādhigamena devaṃ
+matvā dhīro harṣaśokau jahāti \comment{Kaṭhopaniṣat 1.2.12 ?} iti.
+
+śraddhābhaktidhyānayogād avehi \comment{Kaivalyopaniṣat 2 ?} iti.
+
+yadā pañcāvatiṣṭhante jñānāni manasā saha
+buddhiś ca na viceṣṭati tām āhuḥ paramāṃ gatim
+tāṃ yogam iti manyante sthirām indriyadhāraṇām
+apramattas tadā bhavati iti \comment{Kaṭhopaniṣat 2.3–10 ?}
+
+yad ātmatattvena tu brahmatattvaṃ
+dīpopameneha yuktaḥ prapaśyet
+ajaṃ dhruvaṃ sarvatattvair viśuddhaṃ
+jñātvā devaṃ mucyate sarvapāśaiḥ \comment{Śvetaśvataropaniṣat 2.15 ?}
+
+brahmaṇe tvā mahasa om ity ātmānaṃ yuñjīta \comment{Mahā\-nārāyaṇo\-pani\-ṣat 24.2 ?}
+
+trir unnataṃ sthāpya samaṃ śarīraṃ
+hṛdīndriyāṇi manasā sannirudhya
+brahmoḍupena pratareta vidvān
+srotāṃsi sarvāṇi bhayāvahāni \comment{Śvetaśvataropaniṣat 2.8 ?}
+
+om ity evaṃ dhyāyatha ātmānaṃ \comment{Muṇḍakopaniṣat 2.2.6 ?} ityādyāḥ śrutayaḥ.
+\stopsloka
+
+yatidharmaprakaraṇe\comment{??} manuḥ
+
+\startsloka
+bhūtabhāvyān avekṣeta yogena paramātmanaḥ
+dehadvayaṃ vihāyāśu mukto bhavati bandhanāt
+\stopsloka
+
+yājñavalkyasmṛtau\comment{Yājñavalkyasmṛti 1.1.8}
+
+\startsloka
+ijyācāradamāhiṃsādānasvādhyāyakarmaṇām
+ayaṃ tu paramo dharmo yad yogenātmadarśanam 
+\stopsloka
+
+maharṣimātaṅgaḥ\comment{??}
+
+\startsloka
+agniṣṭomādikān sarvān vihāya dvijasattamaḥ
+yogābhyāsarataḥ śāntaḥ paraṃ brahmādhigacchati
+brāhmaṇakṣatriyaviśāṃ strīśūdrāṇāṃ ca pāvanam
+śāntaye karmaṇām anyad yogān nāsti vimuktaye
+\stopsloka
+
+dakṣasmṛtau vyatirekamukhenoktam\comment{7.25 / 8.25}
+
+\startsloka
+svasaṃvedyaṃ hi tad brahma kumārī strīsukhaṃ yathā
+ayogī naiva jānāti jātyandho hi yathā ghaṭam ityādyāḥ smṛtayaḥ.
+\stopsloka
+
+mahābhārate yogamārgaprasaṅge vyāsaḥ
+
+\startsloka
+api varṇāvakṛṣṭas tu nārī vā dharmakāṅkṣiṇī  % ?
+tāv apy etena mārgeṇa gacchetāṃ paramāṃ gatim\comment{Mahābhārata 12.232.32 (246.34)?}
+yadi vā sarvadharmajño yadi vāpy akṛtī pumān
+yadi vā dhārmikaḥ śreṣṭho yadi vā pāpakṛttamaḥ 
+yadi vā puruṣavyāghro yadi vā klaibyadhārakaḥ
+taraty eva mahāduḥkhaṃ jarāmaraṇasāgaram 
+api jijñāsamāno 'pi śabdabrahmātivartate\comment{Mahābhārata 12.228.68} iti.
+\stopsloka
+
+bhagavadgītāyām
+
+\startsloka
+yuñjann evaṃ sadātmānaṃ yogī niyatamānasaḥ
+śāntiṃ nirvāṇaparamāṃ matsaṃsthām adhigacchati \comment{gītā 6.15}
+
+yat sāṅkhyaiḥ prāpyate sthānam \comment{gītā 5.5} ity ādi ca. 
+\stopsloka
+
+ādityapurāṇe
+
+\startsloka
+yogāt sañjāyate jñānaṃ yogo mayy ekacittatā 
+\stopsloka
+
+skandapurāṇe \comment{?}
+\startsloka
+ātmajñānena muktiḥ syāt tac ca yogād ṛte nahi
+sa ca yogaś ciraṃ kālam abhyāsād eva siddhyati
+\stopsloka
+
+kūrmapurāṇe śivavākyam \comment{?}
+
+\startsloka
+ataḥ paraṃ pravakṣyāmi yogaṃ paramadurlabham
+yenātmānaṃ prapaśyanti bhānum antam iveśvaram
+yogāgnir dahati kṣipram aśeṣaṃ pāpapañjaram
+prasannaṃ jāyate jñānaṃ jñānān nirvāṇam ṛcchati
+\stopsloka
+
+garuḍapurāṇe \comment{?}
+
+\startsloka
+tathā yateta matimān yathā syān nirvṛtitḥ parā
+yogena labhyate sā tu na cānyena tu kenacit
+bhavatāpena taptānāṃ yogo hi paramauṣadham
+parāvaraprasaktā dhīr yasya nirvedasambhavā
+sa ca yogāgninā dagdhasamastakleśasañcayaḥ
+nirvāṇaṃ paramaṃ nityaṃ prāpnoty eva na saṃśayaḥ
+samprāptayogasiddhis tu pūrṇo yas tv ātmadarśanāt
+na kiñcid dṛśyate kāryaṃ tenaiva sakalaṃ kṛtam
+ātmārāmaḥ sadā pūrṇaḥ sukham ātyantikaṃ gataḥ
+atas tasyāpi nirvedaḥ parānandamayasya ca
+tapasā bhāvitātmāno yoginaḥ saṃyatendriyāḥ
+prataranti mahātmāno yogenaiva mahārṇavam 
+\stopsloka
+
+viṣṇudharmeṣu \comment{?}
+
+\startsloka
+yac chreyaḥ sarvabhūtānāṃ strīṇām atyupakārakam
+api kīṭapataṅgānāṃ tan naḥ śreyaḥ paraṃ vada
+ity uktaḥ kapilaḥ pūrvaṃ devair devarṣibhis tathā
+yoga eva paraṃ śreyas teṣām ity uktavān purā
+\stopsloka
+
+vāsiṣṭhe
+
+\startsloka
+duḥsahā rāma saṃsāraviṣavegaviṣūcikā
+yogagāruḍamantreṇa pāvanenopaśāmyati iti\comment{Laghuyogavāsiṣṭha 2.51}
+\stopsloka
+
+nanu tattvamasyādivākyair apy aparokṣapramā sambhavatīti kimartham atiśramasādhye yoge prayāsaḥ
+kāryaḥ. na ca vākya\-janya\-jñānasyāpa\-rokṣatve pramāṇāsambhava iti
+vācyam. tattvama\-syādi\-vākya\-janyaṃ jñānam aparokṣaṃ aparokṣaviṣayakatvāt
+cākṣuṣaghaṭādi\-pratyakṣa\-vad ity anumānasya pramāṇatvāt. na ca viṣayagatāparokṣatvasya
+durnirūpatvād dhetvasiddhir iti vācyam.
+ajñānāviṣaya\-cittva\-tattādātmyā\-panna\-tvānya\-tara\-rūpasya tasya sunirūpatvāt. yathā hi
+ghaṭādau cakṣuḥsaṃnikarṣeṇāntaḥkaraṇavṛttidaśāyāṃ tadadhiṣṭhānacaitanyājñānanivṛttau tac
+caitanyasyājñānāviṣayatvaṃ tadghaṭasyājñānāviṣayacaitanyatādātmyāpannatvaṃ cāparokṣatvam. tathā
+tattvamasyādivākyena śuddhacaitanyākārāntaḥkaraṇavṛttyutthāpane sati tadajñānasya nivṛttatvenaiva
+caitanyasyājñānāviṣayatvāc caitanyasyāparokṣatvam iti na hetvasiddhiḥ. na
+cāprayojakatvam. jñānagatāparokṣatvaṃ praty aparokṣaviṣayakatvena prayojakatvāt.
+
+nanv indriyajanyatvaṃ manasa indriyatvābhāvena sukhādipratyakṣe
+vyabhicārāt. athavābhivyaktacaitanyābhinnatayā bhāsamānatvaṃ viṣayasyāparokṣatvam. abhivyaktatvaṃ
+ca nivṛttāvaraṇakatvaṃ parokṣavṛttisthale cāvaraṇanivṛttyabhāvān
+nātivyāptiḥ. sarpādibhramajanakadoṣavatas tu nāyaṃ sarpaḥ kintu rajjur iti vākyena jāyamānā vṛttis
+tu nāvaraṇam nivartayatīti tatra parokṣa eva viṣayaḥ. vedāntavākyajanyaṃ ca jñānam
+āvaraṇanivartakatvād aparokṣam eva. mananādeḥ pūrvam utpannaṃ jñānam ajñānānivartakam.
+pramāṇāsambhāvanādidoṣasāmānyābhāvaviśiṣṭasyaiva tasyājñānanivartakatvāt.
+
+kiṃ ca taṃ tv aupaniṣadaṃ puruṣaṃ pṛcchāmi \comment{Bṛhad\-āraṇya\-kopani\-ṣat 3.9.26} iti
+śrutipratipannam upaniṣanmātragamyatvaṃ yogagamyatve nopapannaṃ syāt. tasmāt tattvamasyādivākyād
+evāparokṣam iti cen na. anumānasyāprayojakatvāt. na ca pratyakṣaṃ prati niruktāparokṣaviṣayakatvena
+prayojakatvāt tanmūlakatarkeṇāprayojakaśaṅkānivṛttir iti vācyam. lāghavāj janyapratyakṣasāmānyaṃ
+prati indriyatvena kāraṇatayā tajjanyatvasyaiva prayojakatvāt. nityānityasādhāraṇapratyakṣatve tu
+na kiñcit prayojakam iti. tanmate\comment{read tvanmate \Wthree \Tue etc} tu pratyakṣaviśeṣe indriyaṃ
+kāraṇaṃ tadviśeṣe ca śabdaviśeṣa iti kāryakāraṇabhāvadvayaṃ syāt. na ca manaso 'nindriyatvaṃ manasa
+indriyatve bādhakābhāvāt indriyāṇāṃ mano nāthaḥ \comment{HP 4.29} iti tu manuṣyam evoddiśya
+manuṣyāṇām ayaṃ rājetyādivad indriyeṣv eva kiñcid utkarṣaṃ bravīti. na tu tasyānindriyatvaṃ tattvaṃ
+ca ṣaṭsv akhaṇḍopādhiviśeṣa eva. ata eva karmendriyaṃ tu pāyvādi manonetrādi dhīndriyam
+\comment{Amarakośa 1.5.8} iti pratyakṣaṃ syād aindriyakam apratyakṣam atīndriyam \comment{Amarakośa
+  3.1.79} iti ca śaktipramāṇabhūtakośo 'pīndriyāpramāṇakajñānasyāpratyakṣatvaṃ vadan manasa
+indriyatvajñāpakaḥ saṅgacchate. indriyāṇi daśaikaṃ ca \comment{Bhagavadgītā 13.5} iti gītāvacanam
+api manasa indriyatve pramāṇam.
+
+kiṃ ca tattvamasyādivākyajanyaṃ jñānaṃ śābdam. śabdajanyatvād yajetetyādivākyajanyajñānavad ity
+anenāparokṣatvavirodhiśābdatvasādhakena satpratipakṣaḥ. na cedam aprayojakam. śābdaṃ praty eva
+śabdasya janakatvena lāghavamūlakānukūlatarkāt. tvanmate tu śabdād api pratyakṣasvīkāreṇa
+kāryakāraṇabhāvadvayakalpane gauravam. api ca manananididhyāsanābhyāṃ pūrvam apy utpannaṃ jñānaṃ
+tava mate 'parokṣam api nājñānanivartakam ity ajñānanivṛttiṃ praty apratibaddhajñānatvenaiva
+hetutvam iti gauravam. mama tu        % \W 261
+samādhyabhyāsaparipākenāsambhāvanādisakalamalarahitenāntaḥkaraṇenātmani dṛṣṭe sati darśanamātrād
+evājñānanivṛtter na kaścid gauravāvakāśaḥ.
+
+\startsloka
+eṣa sarveṣu bhūteṣu gūḍho 'tmā na prakāśate
+dṛśyate tv agryayā (\Wthree agrayā) buddhyā sūkṣmayā sūkṣmadarśibhiḥ
+yacched vāṅmanasī prājñaḥ \comment{Kaṭhopaniṣat 1.3.12–13 ?} 
+\stopsloka
+
+ity ārabhyājñānanivṛttyarthakena mṛtyumukhāt pramucyate \comment{Kaṭhopaniṣat 1.3.16} ityantena
+kaṭhavallīsthamṛtyūpadeśena saṃmato 'yam artha iti na kaścid atra vivāda iti. yadi tu mananādeḥ
+pūrvam utpannaṃ jñānaṃ parokṣam eveti nāpratibaddhatvakṛtagauravam iti matam ādriyate. tadāpi
+śravaṇādibhir manaḥsaṃskāre siddhe 'vyavahitottaram ātmadarśanasambhavāt taduttaraṃ
+vākyasmaraṇādikalpanaṃ mahadgauravāpādakam eva.
+
+nanu na vayaṃ kevalena tarkeṇa śabdajanyajñānasyāparokṣatvaṃ vadāmaḥ\var{vadāma \lem vadāma \Wthree}
+kintu śrutyāpi. tathā hi taṃ tv aupaniṣadaṃ puruṣaṃ pṛcchāmi \comment{Bṛhad\-āraṇya\-kopani\-ṣat 3.9.26}
+iti śrutyā\var{śrutyā ca\lem śrutau \Wthree} caupaniṣadatvaṃ puruṣasya
+nopaniṣajjanyabuddhiviṣayatvamātraṃ pratyakṣādigamye 'py aupaniṣadatvavyavahārāpatteḥ. yathā hi
+dvādaśakapāleṣv aṣṭānāṃ kapālānāṃ sattvād\var{sattvād \lem sattve 'pi \Wthree} dvādaśakapāla\var{kapāla
+  \lem om \Wthree}saṃskṛtenāṣṭā\-kapālādivyavahāraḥ yathā dviputrādāv ekaputrādivyavahāraḥ tathātrāpi
+nānyatra tathā vyavahāra iti upaniṣanmātragamyatvam eva pratyayārthaḥ.
+
+tac ca manogamyatve' nupapannam iti cen na. na hi pratyayenopaniṣadbhinnaṃ sarvaṃ
+kāraṇatvena\var{kāraṇatvena \EdLo \Wthree \lem kāraṇatvād \EdAd} vyāvartyate. śābdāparokṣavādinā tvayāpy
+ātmāparokṣe manaādīnāṃ karaṇatvasyāṅgīkārāt. kiṃ tu purāṇādiśabdāntaram eva śrotavyaḥ
+śrutivākyebhyaḥ\comment{???} iti smaraṇāt. sa cārtho mamāpi saṃmata iti na kiñcid
+etat. pramāṇāntaravyāvṛttau tātparyakalpanaṃ cātmāparokṣe śabdasya pramāṇatve siddha eva vaktum
+ucitam. śabdāntaravyāvṛttitātparyaṃ tu śrutyādisaṃmatatvāt kalpayitum ucitam eva. evaṃ sthite
+manasaivānudraṣṭavyam \comment{Bṛhad\-āraṇya\-kopani\-ṣat 4.4.19} manasaivedam āptavyam
+\comment{Kaṭhopaniṣat 2.1.11} ityādiśrutayo 'py āñjasyena pratipāditā bhaveyuḥ.
+
+% wwwwwwwwwww
+
+yat tu kaiścid uktaṃ darśanavṛttiṃ prati manomātrasyopādānatvaparāyattāḥ\var{parāyattāḥ \lem parā
+  etā \Wthree} śrutayo na virudhyanta iti tad atīva vicārāsaham. yataḥ pramāṇākāṅkṣāyāṃ pravṛttās tāḥ
+katham upādānaparā bhaveyuḥ. kāmaḥ saṅkalpo vicikitsā \comment{Bṛhad\-āraṇya\-kopani\-ṣat 1.5.3}
+ityādiśrutyā sāvadhāraṇayā sarvāsāṃ vṛttīnāṃ manomātropādānakatve bodhite
+ākāṅkṣābhāvenopādānatātparyakatvena varṇayituṃ kathaṃ śakyeran. pūrvaṃ dvitīyavallyāṃ praṇavasya
+brahmabodhakatvenoktes tasyāpy aparokṣahetutvam iti śaṅkāṃ nivārayituṃ manasaivānudraṣṭavyam
+ityādisāvadhāraṇavākyānīty evaṃ varṇayituṃ śakyāni syur ity alam ativāgjālena.
+
+vastutas tu yogināṃ samādhau dūraviprakṛṣṭapadārthajñānaṃ sarvaśāstraprasiddhaṃ na
+parokṣam. tadānīṃ parokṣasāmagryabhāvāt. nāpi smaraṇaṃ teṣāṃ pūrvaṃ viśiṣyānanubhavāt. nāpi
+sukhādijñānavat sākṣirūpam apasiddhāntāt. nāpy apramāṇakaṃ pramāsāmānye karaṇaniyamāt. nāpi
+cakṣurādijanyaṃ teṣām asaṃnikarṣāt. tasmān mānasikī pramaiva sā vācyeti manasa indriyatvaṃ
+pramāṇatvaṃ ca durapahnavam eveti. ye 'pi yogaśrutyoḥ samuccayaṃ kalpayanti teṣām api
+pūrvoktadūṣaṇagaṇas tadavastha eva. tasmād yogajanyasaṃskārasacivamanomātragamya ātmeti siddham.
+
+na ca kāminīṃ bhāvayato vyavahitakāminīsākṣātkārasyeva
+bhāvanājanyatvenātmasākṣātkārasyāpramātvaprasaṅgaḥ. abādhitaviṣayatvād doṣajanyatvābhāvāc
+ca. kāminīsākṣātkārasya tu bādhitaviṣayatvād doṣajanyatvāc cāprāmāṇyaṃ na bhāvanājanyatvāt. na ca
+bhāvanāsamādher jñānajanakatve pramāṇāntarāpātaḥ\var{āntarāpātaḥ \Wthree \lem āntartvāpātaḥ ?}. tasyā
+manaḥsahakāritvāt pramāṇanirūpaṇanpuṇair naiyāyikādibhir api yogaja\var{yogaja \Wthree \lem yoga
+  ?}pratyakṣasyālaukikapratyakṣe 'ntarbhāvaḥ kṛtaḥ. yogajālaukikasaṃni\-karṣeṇa yogino
+vyavahitaviprakṛṣṭasūkṣmārtham ātmānam api yathārtham paśyanti.
+
+tathā ca pātañjale sūtre ṛtambharā tatra prajñā śrutānumānaprajñābhyām anyaviṣayā viśeṣārthatvāt.
+\comment{Yogasūtra 1.48–49} tatra samādhau yā prajñā sā ṛtaṃ satyam eva puruṣayāthārthyaṃ
+bibhartīti ṛtambhareti prathamasūtrārthaḥ. sā samādhiprajñā śrutānumānaprajñābhyāṃ śrutaṃ śravaṇaṃ
+śābdabodhaḥ anumananam anumānaṃ yauktika\var{yauktika \lem yauktikaṃ \Wthree}jñānaṃ tadrūpaprajñābhyām
+anyaviṣayā atiriktaviṣayā. kutaḥ viśeṣārthatvāt. viśeṣo nirvikalpo 'rtho viṣayo yasyāḥ sā tathā
+tasyā bhāvas tathātvaṃ tasmāc chabdasya padārthatāvacchedakapuraskāreṇaivānumānasya
+vyāpakatāvacchedakapuraskāreṇaiva dhījanakatvaniyamena tadgrahaṇāyogyaviśeṣyamātraparatvād ity
+arthaḥ.
+
+atra bādarāyaṇakṛtaṃ bhāṣyam śrutam āgamavijñānaṃ tatsāmānyaviṣayam. na hy āgamena śakyo viśeṣo
+'bhidhātum. kasmāt na hi viśeṣeṇa kṛtasaṅketaḥ śabda ity ārabhya
+samādhiprajñānirgrāhya\var{nirgrāhya \EdLo \lem nirgrāhya \Wthree} eva sa viśeṣo bhavati\var{viśeṣo
+  bhavati ?  \lem viśeṣo \EdLo}. bhūtasūkṣmagato vā puruṣagato vā \comment{Vyāsabhāṣya 1.49} iti.
+
+yogabīje
+
+\startsloka
+jñānaniṣṭho virakto 'pi dharmajño 'pi jitendriyaḥ
+vinā yogena devo 'pi na mokṣaṃ labhate priye \comment{Yogabīja ?}
+\stopsloka
+
+kiṃ ca tad eva saktaḥ saha karmaṇa iti liṅgaṃ mano yatra niṣaktam asya
+\comment{Bṛhad\-āraṇya\-kopani\-ṣat 4.4.6} iti śruteḥ.  kāraṇaṃ guṇasaṅgo 'sya sadasadyonijanmasu
+\comment{Bhagavadgītā 13.21} iti smṛteś ca dehāvasānasamaye yatra rāgādyudbuddho bhavati tām eva
+yoniṃ jīvaḥ prāpnotīti yogahīnasya janmāntaraṃ syād eva. maraṇasamaye samudbhūtavaiklavyasyāyoginā
+vārayitum aśakyatvāt. tad uktaṃ yogabīje
+
+\startsloka
+dehāvasānasamaye citte yad yad vibhāvayet
+tat tad eva bhavej jīva ity evaṃ janmakāraṇam
+dehānte kiṃ bhavej janma tan na jānanti mānavāḥ
+tasmāj jñānaṃ ca vairāgyaṃ japaś ca kevalaṃ śramaḥ
+pipīlikā yadā lagnā dehe jñānād vimucyate
+asau kiṃ vṛścikair daṣṭo dehānte vā kathaṃ sukhī
+\stopsloka
+
+iti. yogināṃ tu yogabalenāntakāle 'py ātmabhāvanayā mokṣa eveti na syāj janmāntaram. tad uktaṃ
+bhagavatā
+
+\startsloka
+prayāṇakāle manasācalena
+bhaktyā yukto yogabalena caiva\comment{Bhagavadgītā 8.10}
+\stopsloka
+
+ityādinā.
+
+\startsloka
+  śataṃ caikā ca hṛdayasya nāḍyaḥ \comment{Chāndogyopaniṣat 8.6.6}
+\stopsloka
+
+ityādiśruteś ca. na ca tattvamasyādivākyasyāparokṣajñānājanakatve\var{jñānājanakatve \EdLo \EdAd
+  \lem jñānajanakatve \Wthree} tadvicārasya vaiyarthyam eveti śaṅkyam. vākyavicārajanyajñānasya
+yogadvārāparokṣajñānasādhanatvāt.
+
+atra ca yogabīje gaurīśvarasaṃvādo mahān asti tataḥ kiñcil likhyate
+
+\startsloka
+devy uvāca 
+jñāninas tu mṛtā ye vai teṣāṃ bhavati kīdṛśī
+gatiḥ kathaya deveśa kāruṇyāmṛtavāridhe \comment{Yogabīja ???}
+\stopsloka
+
+\startsloka
+īśvara uvāca
+dehānte jñāninā puṇyāt pāpāt tatphalam āpyate
+yādṛśaṃ tu bhavet tat tat bhuktvā jñānī punar bhavet
+puṇyāt puṇyena labhate siddhena saha saṅgatim
+tataḥ siddhasya kṛpayā yogī bhavati nānyathā
+tato naśyati saṃsāro nānyathā śivabhāṣitam\comment{Yogabīja ???}
+\stopsloka
+
+\startsloka
+devy uvāca
+jñānād eva hi mokṣaṃ ca vadanti jñāninaḥ sadā
+na kathaṃ siddhayogena yogaḥ kiṃ mokṣado bhavet\comment{Yogabīja ???}
+\stopsloka
+
+\startsloka
+īśvara uvāca
+jñānenaiva hi mokṣo hi teṣāṃ vākyaṃ tu nānyathā
+sarve vadanti khaḍgena jayo bhavati tarhi kim
+vinā yuddhena vīryeṇa kathaṃ jayam avāpnuyāt
+tathā yogena rahitaṃ jñānaṃ mokṣāya no bhavet
+\stopsloka
+
+nanu janakādīnāṃ yogam antareṇāpy apratibaddhajñānamokṣayoḥ śravaṇāt kathaṃ yogād
+evāpratibaddhajñānaṃ mokṣaś ceti cet ucyate teṣāṃ pūrvajanmānuṣṭhitayogasaṃskārāj jñānaprāptir iti
+purāṇādau śrūyate. tathā hi
+
+\startsloka
+jaigīṣavyo yathā vipro yathā caivāsitādayaḥ
+kṣatriyā janakādyās tu tulādhārādayo viśaḥ
+samprāptāḥ paramāṃ siddhiṃ pūrvābhyastasvayogataḥ
+dharmavyādhādayaḥ sapta śūdrāḥ pailavakādayaḥ
+maitreyī sulabhā śārṅgī śāṇḍilī ca tapasvinī
+ete cānye ca bahavo nīcayonigatā api
+jñānaniṣṭhāṃ parāṃ prāptāḥ pūrvābhyastasvayogataḥ
+\stopsloka
+
+iti. kiṃ ca pūrvajanmānuṣṭhitayogābhyāsapuṇyatāratamyena kecid brahmatvaṃ kecid brahmaputratvaṃ
+kecid devarṣitvaṃ kecid brahmarṣitvaṃ kecin munitvaṃ kecid bhaktatvaṃ ca prāptāḥ
+santi. tatropadeśam antareṇaivātmasākṣātkāravanto babhūvuḥ. tathā hi
+hiraṇyagarbhavasiṣṭhanāradasanatkumāravāmadevaśukādayo janmasiddhā ity eva purāṇādiṣu śrūyante. yat
+tu brāhmaṇa eva mokṣādhikārīti śrūyate purāṇādau tadayogiparam. tad uktaṃ garuḍapurāṇe
+
+\startsloka
+yogābhyāso nṛṇāṃ yeṣāṃ nāsti janmāntarād ṛtaḥ
+yogasya prāptaye teṣāṃ śūdravaiśyādikakramaḥ
+strītvāc chūdratvam abhyeti tato vaiśyatvam āpnuyāt
+tataś ca kṣatriyo vipraḥ kriyāhīnas tato bhavet
+anūcānaḥ smṛto yajvā karmanyāsī tataḥ param
+tato jñānitvam abhyeti yogī muktiṃ kramāl labhet
+\stopsloka
+
+iti. śūdravaiśyādikramād yogī bhūtvā muktiṃ labhetety arthaḥ. itthaṃ ca yoge sarvādhikāraśravaṇād
+yogotpannatattvajñānena sarva eva mucyanta iti siddham. yoginas tu bhraṣṭasyāpi na śūdrādikramaḥ
+
+\startsloka
+ śucīnāṃ śrīmatāṃ gehe yogabhraṣṭo 'bhijāyate
+\stopsloka
+
+atha vā yoginām eva ityādibhagavadvacanād\comment{Bhagavadgītā 6.41--42} ity alam.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf jñātvā suṣumṇāsadbhedaṃ kṛtvā vāyuṃ ca madhyagam
+sthitvā sadaiva susthāne brahmarandhre nirodhayet (4.16)}
+\stopsloka
+
+prāṇamanasor layaṃ vinā mokṣo na sidhyatītyuktam. tatra prāṇalayena manaso 'pi layaḥ sidhyatīti
+tallayarītim āha jñātveti. sadaiva sarvadaiva susthāne śobhane sthāne surājye dhārmike deśe
+\comment{HP 1.12} ityādyuktalakṣaṇe sthitvā sthitiṃ kṛtvā vasatiṃ kṛtvety arthaḥ. suṣumṇā
+madhyanāḍī tasyāḥ sadbhedaṃ śobhanaṃ bhedanaprakāraṃ jñātvā gurumukhād viditvā vāyuṃ prāṇaṃ
+madhyagam madhyanāḍīsañcāriṇaṃ kṛtvā brahmarandhre mūrdhāvakāśe nirodhayet nitarāṃ ruddhaṃ
+kuryāt. prāṇasya brahmarandhre nirodho layaḥ. prāṇalaye jāte mano 'pi līyate. tad uktaṃ vāsiṣṭhe
+
+\startsloka
+abhyāsena parispande prāṇānāṃ kṣayam āgate  \comment{Laghuyogavāsiṣṭha 9.88}
+manaḥ praśamam āyāti nirvāṇam avaśiṣyate 
+\stopsloka
+
+iti. prāṇamanasor laye sati bhāvanāviśeṣarūpasamādhisahakṛtenāntaḥ\-kara\-ṇena abādhitātmasākṣātkāro bhavati
+tadā jīvann eva muktaḥ puruṣo bhavati.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf sūryācandramasau dhattaḥ kālaṃ rātrindivātmakam
+  bhoktrī suṣumnā kālasya guhyam etad udāhṛtam (4.17)}
+\stopsloka
+
+prāṇalaye kālajayo bhavatīty āha sūryacandramasau iti. sūryaś ca candramāś ca
+sūryācandramasau. devatādvandve ca \comment{Aṣṭādhyāyī 6.3.26} ity ānaṅ. rātriś ca divā ca
+rātrindivam. acatura \comment{Aṣṭādhyāyī 5.4.77} ityādinā nipātitaḥ. rātrindivam ātmā svarūpaṃ
+yasya sa rātrindivātmakas taṃ rātrindivātmakaṃ kālaṃ samayaṃ dhattaḥ vidhattaḥ kurutaḥ. suṣumnā
+sarasvatī. kālasya sūryācandramobhyām kṛtasya rātrindivātmakasya samayasya bhoktrī bhakṣikā
+vināśikā. etad guhyam rahasyam udāhṛtam kathitam.
+
+ayaṃ bhāvaḥ sārdhaghaṭikādvayaṃ sūryo vahati sārdhaghaṭikādvayaṃ candro vahati. yadā sūryo vahati
+tadā dinam ucyate. yadā candro vahati tadā rātrir ucyate. pañcaghaṭikāmadhye rātrindivātmakaḥ kālo
+bhavati. laukikāhorātramadhye yogināṃ dvādaśāhorātrātmakaḥ kālavyavahāro bhavati. tādṛśakālamānena
+jīvānām āyurmānam asti. yadā suṣumnāmārgeṇa vāyur brahmarandhre līno bhavati tadā
+rātrindivātmakasya kālasyābhāvād uktaṃ bhoktrī suṣumnā kālasyeti. yāvad brahmarandhre vāyur līyate
+tāvad yogina āyur vardhate. dīrghakālābhyastasamādhir yogī pūrvam eva maraṇakālaṃ jñātvā
+brahmarandhre vāyuṃ nītvā kālaṃ nivārayati svecchayā dehatyāgaṃ ca karotīti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf dvāsaptatisahasrāṇi nāḍīdvārāṇi pañjare
+suṣumṇā śāmbhavī śaktiḥ śeṣās tv eva nirarthakāḥ (4.18)}
+\stopsloka
+
+dvāsaptatīti. pañjare pañjaravat śirāsthibhir baddhe śarīre dvābhyām adhikā saptatiḥ dvāsaptatiḥ
+dvāsaptatisaṅkhyākāni sahasrāṇi dvāsaptatisahasrāṇi nāḍīnāṃ śirāṇāṃ dvārāṇi vāyupraveśamārgāḥ
+santi. suṣumṇā madhyanāḍī śāmbhavī śaktir asti. śaṃ sukhaṃ bhavaty asmād bhaktānām iti śambhur
+īśvaras tasyeyaṃ śāmbhavī dhyānena śambhuprāpakatvāt. śambhor āvirbhāvajanakatvād vā śāmbhavī. yad
+vā śaṃ sukharūpo bhavati tiṣṭhatīti śambhur ātmā tasyeyaṃ śāmbhavī. cidabhivyaktisthānatvād
+dhyānenātmasākṣātkārahetutvāc ca. śeṣā iḍāpiṅgalādayas tu nirarthakā eva nirgato 'rthaḥ prayojanaṃ
+yāsāṃ tā nirarthakāḥ pūrvoktaprayojanābhāvāt.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf vāyuḥ paricito yasmād agninā saha kuṇḍalīm
+bodhayitvā suṣumṇāyāṃ praviśed anirodhataḥ (4.19)}
+\stopsloka
+
+vāyur iti. yasmāt paricito 'bhyasto vāyus tasmād agninā jaṭharāgninā saha kuṇḍalīm śaktiṃ
+bodhayitvā anirodhato 'pratibandhāt suṣumṇāyāṃ sarasvatyāṃ praviśet vāyoḥ suṣumnāpraveśārtham
+abhyāsaḥ kartavya ity arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+  {\bf suṣumṇāvāhini prāṇe siddhyaty eva manonmanī
+    anyathā tv itarābhyāsāḥ prayāsāyaiva yoginām (4.20)}
+\stopsloka
+
+suṣumneti. prāṇe suṣumṇāvāhini sati manonmanī unmanyavasthā siddhyaty eva. anyathā prāṇe
+suṣumnāvāhiny asati tu itarābhyāsāḥ suṣumnetaranāḍyabhyāsāḥ yoginām yogābhyāsināṃ prayāsāyaiva
+śramāyaiva bhavantīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf pavano badhyate yena manas tenaiva badhyate
+manaś ca badhyate yena pavanas tena badhyate (4.21)}
+\stopsloka
+
+pavana iti. yena yoginā pavanaḥ prāṇavāyur badhyate baddhaḥ kriyate tenaiva yoginā mano
+badhyate. yena mano badhyate tena pavano badhyate. manaḥpavanayor ekatare baddhe ubhayaṃ baddhaṃ
+bhavatīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf hetudvayaṃ tu cittasya vāsanā ca samīraṇaḥ
+tayor vinaṣṭa ekasmin tau dvāv api vinaśyataḥ (4.22)}
+\stopsloka
+
+hetudvayaṃ tu cittasyeti. cittasya pravṛttau hetudvayaṃ kāraṇadvayam asti. kiṃ tad ity āha vāsanā
+bhāvanākhyaḥ saṃskāraḥ samīraṇaḥ prāṇavāyuś ca. tayor vāsanāsamīraṇayor ekasmin vinaṣṭe sati kṣīṇe
+sati tau dvāv api vinaśyataḥ. ayam āśayaḥ vāsanākṣaye samīraṇacitte kṣīṇe bhavataḥ. samīraṇe kṣīṇe
+cittavāsane kṣīṇe bhavataḥ. citte kṣīṇe samīraṇavāsane kṣīṇe bhavataḥ. tad uktaṃ vāsiṣṭhe
+
+\startsloka
+dve bīje rāma cittasya prāṇaspandanavāsane  \comment{Laghuyogavāsiṣṭha 10.64, 110–113, 15}
+ekasmiṃś ca tayor naṣṭe kṣipraṃ dve api naśyataḥ (10.64)
+\stopsloka
+
+tatraiva vyatirekeṇoktam
+
+\startsloka
+yāvad vilīnaṃ na mano na tāvad vāsanākṣayaḥ 
+na kṣīṇā vāsanā yāvac cittaṃ tāvan na śāmyati 
+na yāvat tattvavijñānaṃ na tāvac cittasaṃkṣayaḥ
+yāvan na cittopaśamo na tāvat tattvavedanam
+yāvan na vāsanānāśas tāvat tattvāgamaḥ kutaḥ
+yāvan na tattvasamprāptir na tāvad vāsanākṣayaḥ
+tattvajñānaṃ manonāśo vāsanākṣaya eva ca
+mithaḥ kāraṇatāṃ gatvā duḥsādhyāni sthitāny ataḥ (10.110–113)
+
+traya ete samaṃ yāvan na svabhyastā muhur muhuḥ
+tāvan na tattvasamprāptir bhavaty api samaiḥ śataiḥ (10.15)
+\stopsloka
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf mano yatra vilīyeta pavanas tatra līyate 
+pavano līyate yatra manas tatra vilīyate (4.23)}
+\stopsloka
+
+mana iti. yatra yasminn ādhāre mano vilīyeta pavanas tatra tasminn ādhāre līyate. yatra yasminn
+ādhāre pavano līyate tatra tasminn ādhāre mano vilīyata ity anvayaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf dugdhāmbuvat saṃmilitāv ubhau tau
+tulyakriyau mānasamārutau hi
+yato marut tatra manaḥpravṛttir
+yato manas tatra marutpravṛttiḥ (4.24)}
+\stopsloka
+
+dugdhāmbuvad iti. dugdhāmbuvat kṣīranīravat saṃmilitau samyak militau tāv ubhau dvāv api
+mānasamārutau mānasaṃ ca mārutaś ca mānasamārutau cittaprāṇau. tulyakriyau tulyā samā kriyā
+pravṛttir yayos tādṛśau bhavataḥ. tulyakriyatvam evāha yata iti. yataḥ yatra sārvavibhaktikas
+tasiḥ. yasmin cakre marud vāyuḥ pravartate tatra tasmin cakre manaḥpravṛttiḥ manasaḥ pravṛttir
+bhavati. yato yasmin cakre manaḥ pravartate tatra tasmin cakre marutpravṛttiḥ vāyoḥ pravṛttir
+bhavatīty arthaḥ. tad uktaṃ vāsiṣṭhe
+
+\startsloka
+avinābhāvinī nityaṃ jantūnāṃ prāṇacetasī
+kusumāmodavan miśre tilataile iva sthite (7.21)
+kurutaś ca vināśena kāryaṃ mokṣākhyam uttamam (7.22cd)
+\stopsloka
+
+iti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf tatraikanāśād aparasya nāśa
+ekapravṛtter aparapravṛttiḥ
+adhvastayoś cendriyavargavṛttiḥ
+pradhvastayor mokṣapadasya siddhiḥ (4.25)}
+\stopsloka
+
+tatreti. tatra tayor mānasamārutayor madhye ekasya mānasasya mārutasya vā nāśāl layād
+aparasyānyasya mārutasya mānasasya vā nāśo layo bhavati. ekapravṛtter ekasya mānasasya mārutasya vā
+pravṛtter vyāpārād aparapravṛttiḥ aparasya mārutasya mānasasya pravṛttir vyāpāro
+bhavati. adhvastayor alīnayor mānasamārutayoḥ sator indriyavargavṛttir indriyasamudāyasya
+svasvaviṣaye pravṛttir bhavati. pradhvastayoḥ pralīnayos tayoḥ sator mokṣapadasya mokṣākhyapadasya
+siddhir niṣpattir bhavati. tayor laye puruṣasya svarūpe 'vasthānād ity arthaḥ.
+
+\startsloka
+tatrāpi sādhyaḥ pavanasya nāśaḥ
+ṣaḍaṅgayogādiniṣevaṇena
+manovināśas tu guroḥ prasādān
+nimeṣamātreṇa susādhya eva
+\stopsloka
+
+yogabīje mūlaślokasyāyam uttaraḥ ślokaḥ.\comment{Yogabīja ??}
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf rasasya manasaś caiva cañcalatvaṃ svabhāvataḥ
+raso baddho mano baddhaṃ kiṃ na siddhyati bhūtale (4.26)}
+\stopsloka
+
+rasasyeti. rasasya pāradasya manaso mānasasya svabhāvataḥ svabhāvāc cañcalatvaṃ cāñcalyam
+asti. rasaḥ pārado baddhaś cen manaś cittaṃ baddhaṃ bhavati. tato bhūtale pṛthivītale kiṃ na
+siddhyati. sarvaṃ siddhyatīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf mūrcchito harate vyādhīn mṛto jīvayati svayam
+baddhaḥ khecaratāṃ dhatte raso vāyuś ca pārvati (4.27)}
+\stopsloka
+
+tad evāha mūrcchita iti. oṣadhiviśeṣayogena gatacāpalo raso mūrcchitaḥ. kumbhakānte recakanivṛttau
+vāyur mūrcchita ity ucyate. he pārvatīti sambodhyeśvaravākyam. mūrcchito rasaḥ pāradaḥ vāyuḥ prāṇaś
+ca vyādhīn rogān harate nāśayati. bhasmībhūto rasaḥ brahmarandhre līno vāyuś ca mṛtaḥ svayam
+ātmanā svasāmarthyenety arthaḥ jīvayati dīrghakālaṃ jīvanaṃ karoti. kriyāviśeṣeṇa guṭikākārakṛto
+raso baddhaḥ bhrūmadhyādau dhāraṇāviśeṣeṇa dhṛto vāyuś ca baddhaḥ khecaratām ākāśagatiṃ dhatte
+vidhatte karotīty arthaḥ. tad uktaṃ gorakṣaśatake
+
+\startsloka
+yad bhinnāñjanapuñjasannibham idaṃ vṛttaṃ bhruvor antare
+tattvaṃ vāyumayaṃ yakārasahitaṃ tatreśvaro devatā
+prāṇaṃ tatra vilāpya\var{vilāpya \lem vilīya (preferred by \EdLo)} pañcaghaṭikaṃ cittānvitaṃ dhārayed
+eṣā khe gamanaṃ karoti yamināṃ syād vāyavī dhāraṇā \comment{go.śa. 72, yo.ta. 2.57} iti.
+\stopsloka
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf manaḥsthairye sthiro vāyus tato binduḥ sthiro bhavet
+bindusthairyāt sadā sattvaṃ piṇḍasthairyaṃ prajāyate (4.28)}
+\stopsloka
+
+manaḥsthairya iti. manasaḥ sthairye sati vāyuḥ prāṇaḥ sthiro bhavet. tato vāyusthairyād bindur
+vīryaṃ sthiro bhavet. bindoḥ sthairyāt sadā sarvadā sattvaṃ balaṃ piṇḍasthairyaṃ dehasthairyaṃ
+prajāyate.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf indriyāṇāṃ mano nātho manonāthas tu mārutaḥ
+mārutasya layo nāthaḥ sa layo nādam āśritaḥ (4.29)}
+\stopsloka
+
+indriyāṇām iti. indriyāṇāṃ śrotrādīnāṃ mano 'ntaḥkaraṇaṃ nāthaḥ pravartakaḥ. manonātho manaso nātho
+mārutaḥ prāṇaḥ. mārutasya prāṇasya layo manovilayo nāthaḥ. sa layo manolayaḥ nādam āśrito nāde mano
+līyata ity arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf so 'yam evāstu mokṣākhyo māstu vāpi matāntare
+manaḥprāṇalaye kaścid ānandaḥ sampravartate (4.30)}
+\stopsloka
+
+so 'yam iti. so 'yam eva cittavilaya eva mokṣākhyo mokṣapadavācyaḥ. matāntare 'nyamate māstu
+vā. cittalayasya suṣuptāv api sattvān manaḥprāṇayor laye sati kaścid anirvācya ānandaḥ
+sampravartate samyak pravṛtto bhavati. anirvācyānandāvirbhāve jīvanmuktisukhaṃ bhavaty eveti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf pranaṣṭaśvāsaniśvāsaḥ pradhvastaviṣayagrahaḥ
+niśceṣṭo nirvikāraś ca layo jayati yoginām (4.31)}
+\stopsloka
+
+praṇaṣṭeti. śvāsaś ca niśvāsaś ca śvāsaniśvāsau pranaṣṭau pralīnau śvāsaniśvāsau yasmin sa tathā
+bāhyavāyor antaḥpraveśanaṃ śvāsaḥ antaḥsthitasya vāyor bahirniḥsaraṇaṃ niśvāsaḥ. pradhvastaḥ
+prakarṣeṇa dhvasto naṣṭo viṣayāṇāṃ śabdādīnāṃ graho grahaṇaṃ yasmin nirgatā ceṣṭā kāyakriyā
+yasmin nirgato vikāro 'ntaḥkaraṇakriyā yasmin etādṛśo yoginām layo 'ntaḥkaraṇavṛtter
+dhyeyākārāpattir jayati sarvotkarṣeṇa vartate.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf ucchinnasarvasaṅkalpo niḥśeṣāśeṣaceṣṭitaḥ
+svāvagamyo layaḥ ko'pi jāyate vāgagocaraḥ (4.32)}
+\stopsloka
+
+ucchinneti. ucchinnā naṣṭāḥ sarve saṅkalpā manaḥpariṇāmā yasmin sa tathā nirgataḥ śeṣo yebhyas tāni
+niḥśeṣāṇy aśeṣāṇi ceṣṭitāni yasmin sa tathā svenaivāvagantuṃ boddhuṃ śakyaḥ svāvagamyaḥ vācām
+agocaro 'viṣayaḥ ko 'pi vilakṣaṇo layaḥ jāyate yogināṃ prādurbhavati.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf yatra dṛṣṭir layas tatra bhūtendriyasanātanī
+sā śaktir jīvabhūtānāṃ dve alakṣye layaṃ gate (4.33)}
+\stopsloka
+
+yatra dṛṣṭir iti. yatra yasmin viṣaye brahmaṇi dṛṣṭir antaḥkaraṇavṛttis tatraiva layo
+bhavati. bhūtāni pṛthivyādīni indriyāṇi śrotrādīni sanātanāni śāśvatāni yasyāṃ sā
+satkāryavāde 'vidyāyāṃ kāryajātasya sattvāt. jīvabhūtānāṃ prāṇināṃ śaktir avidyā ime dve alakṣye
+brahmaṇi layaṃ gate yoginām iti śeṣaḥ.
+
+\startsloka
+{\bf layo laya iti prāhuḥ kīdṛśaṃ layalakṣaṇam
+apunarvāsanotthānāl layo viṣayavismṛtiḥ (4.34)}
+\stopsloka
+
+laya iti. layo laya iti prāhur vadanti bahavaḥ. layasya lakṣaṇaṃ layasvarūpaṃ kīdṛśam iti
+praśnapūrvakaṃ layasvarūpam āha apunar iti. apunarvāsanotthānāt punarvāsanotthānābhāvāt
+viṣayavismṛtiḥ viṣayāṇāṃ śabdādīnāṃ dhyeyākārasya viṣayasya vā vismṛtir layo layaśabdārtha ity
+arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf vedaśāstrapurāṇāni sāmānyagaṇikā iva
+ekaiva śāmbhavī mudrā guptā kulavadhūr iva (4.35)}
+\stopsloka
+
+vedeti. vedāś catvāraḥ śāstrāṇi ṣaṭ purāṇāny aṣṭādaśa sāmānyagaṇikā iva veśyā
+iva bahupuruṣagamyatvāt. ekā śāmbhavī mudraiva kulavadhūr iva kulastrīva guptā
+puruṣaviśeṣagamyatvāt.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf atha śāmbhavī
+antarlakṣyaṃ bahir dṛṣṭir nimeṣonmeṣavarjitā
+eṣā sā śāmbhavī mudrā vedaśāstreṣu gopitā (4.36)}
+\stopsloka
+
+cittalayāya prāṇalayasādhanībhūtāṃ mudrāṃ vivakṣus tatra śāmbhavīm mudrām āha antarlakṣyam
+iti. antar ādhārādibrahmarandhrānteṣu cakreṣu madhye svābhimate cakre lakṣyam
+antaḥkaraṇavṛttiḥ. bahir dehād bahiḥ pradeśe dṛṣṭiḥ cakṣuḥsambandhaḥ. kīdṛśī dṛṣṭiḥ
+nimeṣonmeṣavarjitā nimeṣaḥ pakṣasaṃyogaḥ unmeṣaḥ pakṣmasaṃyogaviśleṣaḥ tābhyāṃ varjitā rahitā
+cittasya dhyeyākārāveśe nimeṣonmeṣavarjitā dṛṣṭir bhavati. soktaiṣā mudrā śāmbhavī śabhor iyaṃ
+śāmbhavī śivapriyā śivāvirbhāvajanikā vā bhavati. kīdṛśī vedaśāstreṣu gopitā vedeṣu ṛgādiṣu
+śāstreṣu sāṅkhyapātañjalādiṣu gopitā rakṣitā.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf antar lakṣyavilīnacittapavano yogī yadā vartate
+dṛṣṭyā niścalatārayā bahir adhaḥ paśyann apaśyann api
+mudreyaṃ khalu śāmbhavī bhavati sā labdhā prasādād guroḥ
+śūnyāśūnyavilakṣaṇaṃ sphurati tat tattvaṃ padaṃ śāmbhavam (4.37)}
+\stopsloka
+
+śāmbhavīṃ mudrām abhinīya darśayati antarlakṣyeti. yadā yasyām avasthāyām antaḥ anāhatapadmādau yal
+lakṣyaṃ saguṇeśvaramūrtyādikaṃ tattvamasyādivākyalakṣyaṃ jīveśvarābhinnam ahaṃ brahmāsmīti
+vākyārthabhūtaṃ brahma vā tasmin vilīnau viśeṣeṇa līnau cittapavanau manomārutau yasya sa tathā
+yogī vartate. niścalatārayā niścalā sthirā tārā kanīnikā yasyāṃ tādṛśyā dṛṣṭyā bahir dehād
+bahiḥpradeśe paśyann api cakṣuḥsambandhaṃ kurvann api apaśyan bāhyaviṣayagrahaṇam akurvan vartate
+āste. khalv iti vākyālaṅkāre. iyam uktā śāmbhavī mudrā śāmbhavīnāmikā mudrayati kleśān iti mudrā
+[bhavati sā śāmbhavī mudrā??? \lem \comment{addition in brackets occurs in some manuscripts as in \Wthree}
+guror deśikasya prasādāt prītipūrvakād anugrahāl labdhā prāptā tad idam iti vaktum aśakyaṃ
+śāmbhavam śāmbhavyā idaṃ śāmbhavam śāmbhavīmudrāyāṃ bhāsamānaṃ padaṃ padyate gamyate yogibhir iti
+padam ātmasvarūpaṃ śūnyāśūnyavilakṣaṇaṃ dhyeyākāravṛtteḥ sadbhāvāc chūnyavilakṣaṇaṃ tasyā api
+bhānābhāvād aśūnyavilakṣaṇaṃ tattvaṃ vāstavikaṃ vastu sphurati pratīyate. tathā coktam\comment{source?}
+
+\startsloka
+antarlakṣyam ananyadhīr avirataṃ paśyan mudā saṃyamī
+dṛṣṭyunmeṣanimeṣavarjitam iyaṃ mudrā bhavec chāmbhavī
+gupteyaṃ giriśena tantraviduṣā tantreṣu tattvārthinām
+eṣā syād yamināṃ manolayakarī muktipradā durlabhā
+
+ūrdhvadṛṣṭir adhodṛṣṭir ūrdhvavedho hy adhaḥśirā
+rādhāyantravidhānena jīvanmukto bhavet kṣitau
+\stopsloka
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf śrīśāmbhavyāś ca khecaryā avasthādhāmabhedataḥ
+bhavec cittalayānandaḥ śūnye citsukharūpiṇi (4.38)}
+\stopsloka
+
+śrīśāmbhavyā iti. śrīśāmbhavyāḥ śrīmatyāḥ śāmbhavīmudrāyāḥ khecarīmudrāyāś cāvasthādhāmabhedataḥ
+avasthā avasthitiḥ dhāma sthānaṃ tayor bhedāc chāmbhavyāṃ bahirdṛṣṭyā avasthitiḥ khecaryāṃ
+bhrūmadhyadṛṣṭyāvasthitiḥ. śāmbhavyāṃ hṛdayaṃ bhāvanādeśaḥ khecaryāṃ bhrūmadhya eva deśaḥ tayor
+bhedābhyāṃ śūnye deśakālavastuparicchedaśūnye sajātīyavijātīyasvagatabhedaśūnye vā citsukharūpiṇi
+cidānandasvarūpiṇy ātmani cittalayānando bhavet syāt. śrīśāmbhavīkhecaryor
+avasthādhāmarūpasādhanāṃśe bhedaḥ na tu cittalayānandarūpaphalāṃśa iti bhāvaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf tāre jyotiṣi saṃyojya kiñcid unnamayed bhruvau
+pūrvayogaṃ mano yuñjann unmanīkārakaḥ kṣaṇāt (4.39)}
+\stopsloka
+
+unmanīmudrām āha tāra iti. tāre netrayoḥ kanīnike jyotiṣi tārayor nāsāgre yojanāt prakāśamāne
+tejasi saṃyojya saṃyukte kṛtvā bhruvau kiñcit svalpam unnamayed ūrdhvaṃ namayet. pūrvaḥ pūrvoktaḥ
+antarlakṣyaṃ bahirdṛṣṭiḥ\comment{HP 4.36} ity ākārako yogo yuktir yasmin tat tādṛśaṃ mano
+'ntaḥkaraṇaṃ yuñjan yuktaṃ kurvan yogī kṣaṇān muhūrtād unmanīkāraka unmanyavasthākārako bhavati.
+
+%  read pūrvayogamano acc. to comm.?  -kārakaḥ does not work well.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf kecid āgamajālena kecin nigamasaṅkulaiḥ
+kecit tarkeṇa muhyanti naiva jānanti tārakam (4.40)}
+\stopsloka
+
+unmanīm antarā anyastareṇopāyo nāstīty āha kecid iti. kecic chāstratantrādividaḥ āgacchanti buddhim
+ārohanti arthā ebhya ity āgamāḥ śāstratantrādayaḥ teṣāṃ jālair jālavad bandhanasādhanais taduktaiḥ
+phalair muhyanti mohaṃ prāpnuvanti. tatrāsaktā badhyanta iti bhāvaḥ. kecid vaidikā nigamasaṅkulair
+nigamānāṃ nigamoktānāṃ saṅkulaiḥ phalabāhulyair muhyanti. kecid vaiśeṣikādayas\var{vaināśikādayas
+  \Wthree \Tue \EdLo \lem vaiśeṣikādayas \EdAd} tarkeṇa svakalpitayuktiviśeṣeṇa muhyanti. tārayatīti
+tārakas taṃ tārakaṃ taraṇopāyaṃ naiva jānanti. uktonmany eva taraṇopāyas taṃ na jānantīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf ardhonmīlitalocanaḥ sthiramanā nāsāgradattekṣaṇaś
+candrārkāv api līnatām upanayan nispandabhāvena yaḥ
+jyotīrūpam aśeṣabījam akhilaṃ dedīpyamānaṃ paraṃ
+tattvaṃ tat padam eti vastu paramaṃ vācyaṃ kim atrādhikam (4.41)}
+\stopsloka
+
+ardhonmīliteti. ardham unmīlite ardhonmīlite ardhonmīlite locane yena saḥ
+ardhonmīlitalocanaḥ. ardhodghāṭitalocana ity arthaḥ. sthiraṃ niścalaṃ mano yasya sa sthiramanāḥ
+nāsāyā nāsikāyā agre 'grabhāge nāsikāyāṃ dvādaśāṅgulaparyante vā datte prahite īkṣaṇe netre yena sa
+nāsāgradattekṣaṇaḥ. tad āha vasiṣṭhaḥ\comment{Laghuyogavāsiṣṭha 9.81}
+
+%  Jyotsnā gives meaning with the majority reading:  ardhodghāṭitalocana
+
+\startsloka
+dvādaśāṅgulaparyante nāsāgre vimale 'mbare
+saṃviddṛśoḥ praśāmyantyoḥ prāṇaspando nirudhyate  
+\stopsloka
+
+iti. nispandasya niścalasya bhāvo nispandabhāvaḥ kāyendriyamanasāṃ niścalatvaṃ tena candrārkau
+candrasūryāv api līnatāṃ līnasya bhāvo līnatā layas tām upanayan prāpayan kāyendriyamanasāṃ
+niścalatvena prāṇasañcāram api stambhayann ity arthaḥ. tad uktaṃ prāk mano yatra vilīyate
+\comment{HP 4.23} ityādipūrvoktaviśeṣaṇasampanno yogī jyotīrūpaṃ jyotir ivākhilaprakāśakaṃ rūpaṃ
+yasya sa tathā tam aśeṣabījam ākāśādyutpattidvārā sarvakāraṇam akhilaṃ pūrṇaṃ dedīpyamānam
+atiśayena dīpyata iti dedīpyamānaḥ taṃ tathā svaprakāśaṃ paraṃ kāyendriyamanasāṃ sākṣiṇaṃ tattvam
+anāropitaṃ vāstavikam ity arthaḥ. tad idam iti vaktum aśakyaṃ padyate gamyate yogibhir iti padaṃ
+paramaṃ sarvotkṛṣṭaṃ vastu ātmasvarūpam eti prāpnoti. unmanyavasthāyāṃ svasvarūpāvasthito yogī
+bhavatīty arthaḥ. atrādhikaṃ kiṃ vācyam. aparaṃ vastu na prāpnotīty atra kiṃ vaktavyam ity arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf divā na pūjayel liṅgaṃ rātrau caiva na pūjayet
+sarvadā pūjayel liṅgaṃ divārātrinirodhataḥ (4.42)}
+\stopsloka
+
+unmanībhāvanāyāḥ kālaniyamābhāvam āha divā neti. divā sūryasañcāre liṅgaṃ sarvakāraṇam ātmānam
+etasmād ātmana ākāśaḥ sambhūtaḥ \comment{Taittirīyopaniṣat 2.1} ityādiśruteḥ. na pūjayet na
+bhāvayet. dhyānam evātmapūjanam. tad uktaṃ vāsiṣṭhe\comment{Laghuyogavāsiṣṭha 2.90}
+
+\startsloka
+dhyānopahāra evātmā dhyānam asya mahārcanam
+vinā tenetareṇāyam ātmā labhyate eva no iti
+\stopsloka
+
+rātrau candrasañcāre ca naiva pūjayen naiva bhāvayet. candrasūryasañcāre cittasthairyābhāvāt. cale
+vāte calaṃ cittam \comment{HP 2.2} ity uktatvāt. divārātrinirodhataḥ sūryacandrau
+nirudhya. lyablope pañcamī. tasyās tasil. sarvadā sarvasmin kāle liṅgam ātmānaṃ pūjayed
+bhāvayet. sūryacandrayor nirodhe kṛte suṣumnāntargate prāṇe manaḥsthairyāt. tad uktaṃ
+
+\startsloka
+suṣumnāntargate vāyau manaḥsthairyaṃ prajāyate \comment{quotation untraced.}
+\stopsloka
+
+iti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf atha khecarī
+savyadakṣiṇanāḍistho madhye carati mārutaḥ
+tiṣṭhate khecarīmudrā tasmin sthāne na saṃśayaḥ (4.43)}
+\stopsloka
+
+khecarīm āha savyeti. savyadakṣiṇanāḍistho vāmataditaranāḍistho māruto vāyur yatra madhye carati
+yasmin madhyapradeśe gacchati tasmin sthāne tasmin pradeśe khecarī mudrā tiṣṭhate sthirā
+bhavati. prakāśanastheyākhyayoś ca \comment{Aṣṭādhyāyī 1.3.32} ity ātmanepadam. na saṃśayaḥ ukte
+'rthe sandeho nāstīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf iḍāpiṅgalayor madhye śūnyaṃ caivānilaṃ graset
+tiṣṭhate khecarīmudrā tatra satyaṃ punaḥ punaḥ (4.44)}
+\stopsloka
+
+iḍāpiṅgalayor iti. iḍāpiṅgalayoḥ savyadakṣiṇanāḍyor madhye yac chūnyaṃ khaṃ kartṛ anilaṃ prāṇavāyuṃ
+yatra graset. śūnye prāṇasya sthirībhāva eva grāsaḥ. tatra tasmin śūnye khecarī mudrā
+tiṣṭhate. punaḥ punaḥ satyam iti yojanā.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf sūryācandramasor madhye nirālambāntare punaḥ
+saṃsthitā vyomacakre yā sā mudrā nāma khecarī (4.45)}
+\stopsloka
+
+sūryācandramasor iti. sūryācandramasor iḍāpiṅgalayor madhye nirālambaṃ yadantaram avakāśas
+tatra. punah pādapūraṇe. vyomnāṃ khānāṃ cakre samudāye. bhrūmadhye sarvakhānāṃ samanvayāt. tad
+uktaṃ pañcasrotaḥsamanvitam \comment{3.53} iti. yā saṃsthitā sā mudrā khecarī nāma.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf somād yatroditā dhārā sākṣāt sā śivavallabhā
+pūrayed atulāṃ divyāṃ suṣumṇāṃ paścime mukhe (4.46)}
+\stopsloka
+
+somād iti. somāc candrād yatra yasyāṃ khecaryāṃ dhārā amṛtadhārā uditodgatā sā khecarī sākṣāc
+chivavallabhā śivasya priyeti pūrvenānvayaḥ. atulāṃ nirupamāṃ divyāṃ sarvanāḍyuttamāṃ suṣumṇāṃ
+paścime mukhe pūrayet jihvayā iti śeṣaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf purastāc caiva pūryeta niścitā khecarī bhavet
+abhyastā khecarī mudrāpy unmanī samprajāyate (4.47)}
+\stopsloka
+
+purastāc caiveti. purastāc caiva pūrvata eva pūryeta suṣumnā prāṇeneti śeṣaḥ. yadi tarhi
+niścitāsandigdhā khecarī khecaryākhyā mudrā bhaved iti. yadi tu purastāt prāṇena na pūryeta
+jihvāmātreṇa paścimataḥ pūryeta tarhi mūḍhāvasthājanikā na niścitā khecarī syād iti bhāvaḥ. khecarī
+mudrāpy abhyastā satī unmanī samprajāyate cittasya dhyeyākārāveśāt turyāvasthā bhavatīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf bhruvor madhye śivasthānaṃ manas tatra vilīyate
+jñātavyaṃ tatpadaṃ turyaṃ tatra kālo na vidyate (4.48)}
+\stopsloka
+
+bhruvor iti. bhruvor madhye bhruvor antarāle śivasthānaṃ śivasya īśvarasya sthānaṃ śivasya
+sukharūpasya ātmano vā sthānam astīti śeṣaḥ. tatra tasmin śive mano līyate. śivākāravṛttipravāhavad
+bhavati. tac cittalayarūpaṃ turyaṃ padaṃ jāgratsvapnasuṣuptibhyaś caturthāvasthākhyaṃ
+jñātavyam. tatra tasmin pade kālo mṛtyur na vidyate. yad vā sūryacandrayor nirodhād
+āyuḥkṣayakārakaḥ kālaḥ samayo na vidyata ity arthaḥ. tad uktaṃ bhoktrī suṣumnā kālasya
+\comment{4.17} iti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf abhyaset khecarīṃ tāvad yāvat syād yoganidritaḥ
+samprāptayoganidrasya kālo nāsti kadācana (4.49)}
+\stopsloka
+
+abhyased iti. tāvat khecarīṃ mudrām abhyaset yāvad yoganidritaḥ. yogaḥ sarvavṛttinirodhaḥ saiva
+nidrā yoganidrā yoganidrā asya\var{asya \lem ālasya \Wthree} sañjātā iti yoganidritaḥ tādṛśaḥ
+syāt. samprāptā yoganidrā yena sa samprāptayoganidraḥ tasya kadācana kasmiṃścid api samaye kālo
+mṛtyur nāsti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf nirālambaṃ manaḥ kṛtvā na kiñcid api cintayet
+sa bāhyābhyantaraṃ vyomni ghaṭavat tiṣṭhati dhruvam (4.50)}
+\stopsloka
+
+nirālambam iti. yo nirālambam ālambanaśūnyaṃ manaḥ kṛtvā kiñcid api na cintayet khecarīmudrāyāṃ
+jāyamānāyāṃ brahmākārām api vṛttiṃ paravairāgyeṇa parityajed ity arthaḥ. sa yogī bāhyābhyantare
+bāhye bahirbhave ābhyantare 'bhyantarbhave ca vyomny ākāśe ghaṭavat ghaṭa iva tiṣṭhati. dhruvam
+niścitam ity arthaḥ. yathākāśe ghaṭo bahirantaś cākāśapūrṇo bhavati tathā khecaryām
+ālambanaparityāgena yogī brahmaṇā pūrṇas tiṣthatīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf bāhyavāyur yathā līnas tathā madhyo na saṃśayaḥ
+svasthāne sthiratām eti pavano manasā saha (4.51)}
+\stopsloka
+
+bāhyeti. bāhyo dehād bahirbhavo vāyur yathā līno bhavati khecaryām tasyāntaḥpravṛttyabhāvāt. tathā
+madhyo dehamadhyavartī vāyur līno bhavati. tasya bahiḥpravṛttyabhāvāt. na saṃśayaḥ asminn arthe
+sandeho nāstīty arthaḥ. sthīyate sthirībhūyate 'sminn iti sthānaṃ svasya prāṇasya sthānaṃ
+sthairyādhiṣṭhānaṃ brahmarandhraṃ tatra manasā cittena saha pavanaḥ prāṇaḥ sthiratām niścalatām eti
+prāpnoti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf evam abhyasyamānasya vāyumārge divāniśam
+abhyāsāj jīryate vāyur manas tatraiva līyate (4.52)}
+\stopsloka
+
+evam uktaprakāreṇa vāyumārge prāṇamārge suṣumnāyām ity arthaḥ. divāniśam rātriṃdivam abhyasamanasya
+abhyāsaṃ kurvato yogino 'bhyāsād yatra yasminn ādhāre vāyuḥ prāṇo jīryate kṣīyate līyate ity arthaḥ.
+tatraiva vāyor layādhiṣṭhāne manaś cittaṃ līyate jīryata ity arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf amṛtaiḥ plāvayed deham āpādatalamastakam
+siddhyaty eva mahākāyo mahābalaparākramaḥ (4.53)}
+\stopsloka
+
+amṛtair iti. amṛtaiḥ suṣiranirgataiḥ pādatalaṃ ca mastakaṃ ca pādatalamastakam. dvandvaś ca
+prāṇitūryasenāṅgānāṃ \comment{Aṣṭādhyāyī 2.4.2} ity ekavadbhāvaḥ. pādatalamastakam abhivyāpyety
+āpādatalamastakam deham āplāvayed āplāvitam kuryāt. mahān utkṛṣṭaḥ kāyo yasya sa mahākāyaḥ mahāntau
+balaparākramau yasyety etādṛśo yogī siddhyaty eva amṛtāplāvanena siddho bhavaty eva.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf śaktimadhye manaḥ kṛtvā śaktiṃ mānasamadhyagām
+manasā mana ālokya dhārayet paramaṃ padam (4.54)}
+\stopsloka
+
+śaktir iti. śaktiḥ kuṇḍalinī tasyā madhye śaktimadhye manaḥ kṛtvā tasyāṃ mano dhṛtvā tadākāraṃ
+manaḥ kṛtvety arthaḥ. śaktiṃ mānasamadhyagām kṛtvā. śaktidhyānāveśāc chaktimanasor atyantaikyaṃ
+kṛtvā tena kuṇḍalīṃ bodhayitveti yāvat.
+
+\startsloka
+  prabuddhā vahniyogena manasā marutā saha \comment{Gorakṣaśataka 31, yo.ta. 1.49}
+\stopsloka
+
+iti gorakṣokteḥ. manasāntaḥkaraṇena mana ālokya buddhiṃ manaso'valokanena sthirīkṛtyety
+arthaḥ. paramaṃ padam sarvotkṛṣṭaṃ svarūpaṃ dhārayed dhāraṇāviṣayaṃ kuryād ity arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf khamadhye kuru cātmānam ātmamadhye ca khaṃ kuru
+sarvaṃ ca khamayaṃ kṛtvā na kiñcid api cintayet (4.55)}
+\stopsloka
+
+khamadhya iti. kham iva pūrṇaṃ brahma khaṃ tanmadhye ātmānam svasvarūpaṃ kuru. brahmāham iti
+bhāvayety arthaḥ. ātmamadhye svasvarūpe ca khaṃ pūrṇaṃ brahma kuru. ahaṃ brahmeti ca bhāvayety
+arthaḥ. sarvaṃ ca khamayaṃ kṛtvā brahmamayaṃ vibhāvya kim api na cintayet. ahaṃ brahmeti dhyānam
+api parityajety arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf antaḥ śūnyo bahiḥ śūnyaḥ śūnyaḥ kumbha ivāmbare
+antaḥ pūrṇo bahiḥ pūrṇaḥ pūrṇaḥ kumbha ivārṇave (4.56)}
+\stopsloka
+
+evaṃ samāhitasya svarūpe sthitim āha antaḥśūnya iti. antaḥ antaḥkaraṇe śūnyaḥ. brahmātiriktavṛtter
+abhāvād dvitīyaśūnyaḥ. bahir antaḥkaraṇād bahir api śūnyaḥ. dvitīyādarśanāt. ambare ākāśe kumbho
+ghaṭo yathāntarbahiḥśūnyas tadvad antaḥ antaḥkaraṇe hṛdākāśe vā pūrṇo vyāptatvād brahmākāravṛtteḥ
+sadbhāvāt brahmasattvād vā. bahiḥ pūrṇaḥ antaḥkaraṇād bahir hṛdākāśād bahir vā pūrṇaḥ. sattayā
+brahmātiriktavṛtter abhāvād brahmapūrṇatvād vā. arṇave samudre kumbho ghaṭo yathā sarvato jalapūrṇo
+bhavaty evaṃ samādhiniṣṭho yogī brahmapūrṇo bhavatīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf bāhyacintā na kartavyā tathaivāntaracintanam
+sarvacintāṃ parityajya na kiñcid api cintayet (4.57)}
+\stopsloka
+
+bāhyacinteti. samāhitena yoginety adhyāhāraḥ. bāhyacintā bāhyaviṣayacintā na kartavyā. tathaiva
+bāhyacintākaraṇavad āntaracintanam āntarāṇāṃ manasā parikalpitānām āśāmodakasaudhavāṭikādīnāṃ
+cintanaṃ na kartavyam iti liṅgavipariṇāmenānvayaḥ. sarvacintāṃ bāhyābhyantaracintanaṃ parityajya
+kiñcid api na cintayet paravairāgyeṇātmākāravṛttim api parityajet. tattyāge svarūpāvasthitirūpā
+jīvanmuktir bhavatīti bhāvaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf saṅkalpamātrakalanaiva jagat samagraṃ
+saṅkalpamātrakalanaiva manovilāsaḥ
+saṅkalpamātramatim utsṛja nirvikalpam
+āśritya niścayam avāpnuhi rāma śāntim (4.58)}    % MU/LYV
+\stopsloka
+
+bāhyābhyantaracintāparityāge śāntiś ca bhavatīty atra vasiṣṭhavākyaṃ\comment{The verse is an
+  adaptation of Laghuyogavāsiṣṭha 7.27} pramāṇayati saṅkalpeti. saṅkalpo mānasiko vyāpāraḥ sa eva
+saṅkalpamātraṃ tasya kalanaiva racanaivedaṃ dṛśyamānaṃ samagraṃ jagat
+bāhyaprapañco\var{bāhyaprapañco \Wthree \EdAd \lem bāhyaḥ samagraḥ prapañco Ed Lo \Tue} manomātrakalpita
+ity arthaḥ. manaso mānasasya vilāso nānāviṣayākārakalpanā āśāmodakasaudhavāṭikādikalpanārūpo
+vilāsaḥ saṅkalpamātrakalanaiva. mānasaḥ prapañco 'pi saṅkalpamātraracanaivety arthaḥ. saṅkalpamātre
+bāhyābhyantaraprapañce yā matiḥ satyatvabuddhis tām utsṛja tyaja. tarhi kiṃ kartavyam ity ata āha
+nirvikalpam iti. viśiṣṭā kalpanā\var{viśiṣṭā kalpanā \Tue \Wthree \lem viśiṣṭakalpanā \EdMu \EdLo viśiṣṭaḥ
+  kalpo \EdAd} vikalpaḥ. ātmani
+kartṛtvabhoktṛtvasukhitvaduḥkhitvasajātīyavijātīyasvagatabhedadeśakālavastuparicchedakalpanārūpaḥ
+tasmān niṣkrānto nirvikalpaḥ ātmā tam āśritya dhāraṇādiviṣayaṃ kṛtvā he rāma niścayam asandigdhaṃ
+śāntim paramoparatim avāpnuhi. tataḥ sukham api prāpsyasīti bhāvaḥ. tad uktaṃ bhagavatā
+vyatirekeṇa
+
+\startsloka
+  na cābhāvayataḥ śāntir aśāntasya kutaḥ sukham \comment{Bhagavadgītā 2.66}
+\stopsloka
+
+iti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf karpūram anale yadvat saindhavaṃ salile yathā
+tathā sandhīyamānaṃ ca manas tattve vilīyate (4.59)}
+\stopsloka
+
+karpūram iti. yad yad yathā anale 'gnau sandhīyamānaṃ saṃyojyamānaṃ karpūram vilīyate viśeṣeṇa
+līyate līnaṃ bhavati agnyākāraṃ bhavati. yathā salile jale sandhīyamānaṃ saindhavaṃ lavaṇaṃ
+vilīyate lavaṇākāraṃ parityajya jalākāraṃ bhavati tathā tadvat tattve ātmani sandhīyamānaṃ
+dhāryamāṇaṃ mano vilīyate ātmākāraṃ bhavati.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf jñeyaṃ sarvaṃ pratītaṃ ca jñānaṃ ca mana ucyate
+jñānaṃ jñeyaṃ samaṃ naṣṭaṃ nānyaḥ panthā dvitīyakaḥ (4.60)}
+\stopsloka
+
+manaso vilaye jāte dvaitam api līyata ity āha tribhiḥ jñeyam iti. sarvaṃ sakalaṃ jñeyaṃ jñānārhaṃ
+pratītaṃ jñātaṃ ca jñānaṃ ca idaṃ sarvaṃ mana ucyate. sarvasya manaḥkalpanāmātratvān
+manaḥśabdenocyate. jñānaṃ jñeyaṃ ca samaṃ manaso vilaye manasā sārdhaṃ naṣṭaṃ yadi tarhi dvitīyakaḥ
+dvitīya eva dvitīyakaḥ panthā manoviṣayo nāsti. dvaitaṃ nāstīti phalitārthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf manodṛśyam idaṃ sarvaṃ yat kiñcit sacarācaram
+manaso hy unmanībhāvād dvaitaṃ naivolabhyate (4.61)}
+\stopsloka
+
+manodṛśyam iti. idam upalabhyamānaṃ yat kiñcit yat kim api caraṃ jaṅgamam acaraṃ sthāvaraṃ caraṃ
+cācaraṃ ca carācare tābhyāṃ saha vartata iti sacarācaram yaj jagat tat sarvaṃ manodṛśyaṃ manasā
+dṛśyam manaḥsaṅkalpamātram ity arthaḥ. manaḥkalpanāsattve pratītes tadabhāve cāpratīter bhrama eva
+sarvaṃ jagat bhramasya prātītikaśarīratvāt. na ca bauddhamataprasaṅgaḥ. bhramādhiṣṭhānasya
+brahmaṇaḥ satyatvābhyupagamāt. manasa unmanībhāvād vilayād dvaitaṃ bhedaḥ naivolabhyate naiva
+pratīyate dvaitabhramahetor manaḥsaṅkalpasyābhāvāt. hīti hetāv avyayam.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf jñeyavastuparityāgād vilayaṃ yāti mānasam
+manaso vilaye jāte kaivalyam avaśiṣyate (4.62)}
+\stopsloka
+
+jñeyeti. jñeyaṃ jñānaviṣayaṃ yad vastu sarvaṃ carācaraṃ yad dṛśyaṃ\var{yad dṛśyaṃ \Wthree \Tue \EdMu \lem
+  om \EdLo (no variants given)} tasya parityāgān nāmarūpātmakasya tasya parivarjanād mānasaṃ vilayaṃ
+yāti\var{mānasaṃ vilayaṃ yāti \EdAd \lem vilayaṃ \Tue \Wthree \EdLo \EdMu}
+saccidānandarūpātmākāraṃ\var{saccidānandātmakaṃ \Wthree \lem saccidānandarūpātmākāraṃ \Tue \EdLo \EdAd
+  \EdMu} bhavati. manaso vilaye jāte sati kaivalyam kevalasyātmano bhāvaḥ kaivalyam
+avaśiṣyate. advitīyātmasvarūpam avaśiṣṭaṃ bhavatīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf evaṃ nānāvidhopāyāḥ samyak svānubhavānvitāḥ
+samādhimārgāḥ kathitāḥ pūrvācāryair mahātmabhiḥ (4.63)}
+\stopsloka
+
+evam iti. evaṃ antarlakṣyaṃ bahirdṛṣṭiḥ\comment{HP 4.36} ityādyuktaprakāreṇa mahān
+samādhipariśīlanaśuddha ātmāntaḥkaraṇaṃ yeṣāṃ te mahātmānas tair mahātmabhiḥ pūrve ca te ācāryāś ca
+pūrvācāryā matsyendrādayas taiḥ samādheś cittavṛttinirodhasya mārgāḥ prāptyupāyāḥ kathitāḥ. kīdṛśāḥ
+samādhimārgāḥ nānāvidhopāyāḥ nānāvidhā upāyāḥ sādhanāni yeṣāṃ te tathā samyak samīcīnatayā
+saṃśayaviparyayarāhityena yaḥ svānubhava ātmānubhavas tenānvitā yuktāḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf suṣumṇāyai kuṇḍalinyai sudhāyai candrajanmane
+manonmanyai namas tubhyaṃ mahāśaktyai cidātmane (4.64)}
+\stopsloka
+
+suṣumṇādibhyaḥ kṛtakṛtyas tāḥ praṇamati suṣumṇāyai iti. suṣumṇā madhyanāḍī tasyai kuṇḍalinyai
+ādhāraśaktyai candrād bhrūmadhyasthāj janma yasyās tasyai sudhāyai pīyūṣāya manonmanyai
+turyāvasthāyai cic caitanyam ātmā svarūpaṃ yasyāḥ sā tathā tasyai mahatī jaḍānāṃ kāyendriyamanasāṃ
+caitanyasampādakatvāt sarvottamā yā śaktiś citiśaktiḥ puruṣarūpā tasyai. tubhyam iti pratyekaṃ
+sambadhyate. namaḥ prahvībhāvo 'stu.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf aśakyatattvabodhānāṃ mūḍhānām api saṃmatam
+proktaṃ gorakṣanāthena nādopāsanam ucyate (4.65)}
+\stopsloka
+
+nānāvidhān samādhyupāyān uktvā nādānusandhānarūpaṃ mukhyopāyaṃ pratijānīte aśakyeti. avyutpannatvād
+aśakyas tattvabodhas tattvajñānaṃ yeṣāṃ te tathā teṣāṃ mūḍhānām anadhītānāṃ saṃmatam. apiśabdāt
+kimutādhītānām iti gamyate. gorakṣanāthena proktam ity anena mahaduktatvād upādeyatvaṃ
+gamyate. nādasyānāhatadhvaner upāsanam anusandhānarūpaṃ sevanam ucyate kathyate.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf śrīādināthena sapādakoṭi-
+layaprakārāḥ kathitā jayanti
+nādānusandhānakam ekam eva
+manyāmahe mukhyatamaṃ layānām (4.66)}
+\stopsloka
+
+śrīādinātheneti. śrīādināthena śivena kathitāḥ proktāḥ pādena caturthāṃśena saha vartamānāḥ
+koṭisaṅkhyākā layaprakārāś\var{layaprakārāś \EdMu \EdAd \lem ye layaprakārāś \Tue \Wthree \EdLo}
+cittalayasādhanabhedā jayanti utkarṣeṇa vartante. vayaṃ tu nādānusandhānakam nādānucintanam eva
+ekam kevalaṃ layānām layasādhanānāṃ madhye mukhyatamam atiśayena mukhyaṃ manyāmahe
+jānīmahe. utkṛṣṭānāṃ layasādhanānāṃ madhye utkṛṣṭatamatvād gorakṣābhimatatvād asmadabhimatatvāc ca
+nādānusandhānam eva avaśyaṃ vidheyam iti bhāvaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf muktāsane sthito yogī mudrāṃ sandhāya śāmbhavīm
+śṛṇuyād dakṣiṇe karṇe nādam antastham ekadhīḥ (4.67)}
+\stopsloka
+
+śāmbhavīmudrayā nādānusandhānam āha muktāsana iti. muktāsane siddhāsane sthito yogī śāmbhavīm
+mudrāṃ antarlakṣyaṃ bahirdṛṣṭiḥ \comment{HP 4.36} ityādinoktāṃ sandhāya kṛtvā ekadhīr ekāgracittaḥ
+san dakṣiṇe karṇe 'ntasthaṃ suṣumnānāḍyāṃ santam eva nādaṃ śṛṇuyāt. tad uktaṃ tripurāsārasamuccaye
+
+ādau mattālimālājanitaravasamastārajhaṅkārakārī
+nādo 'sau vāṃśikasyānilabharitasadvaṃśaniḥsvānatulyaḥ
+ghaṇṭānādānukārī tad anu ca jaladhidhvānadhīro gabhīro
+garjatparjanyaghoṣaḥ para iha kuhare vartate brahmanāḍyāḥ \comment{Tripurāsārasamuccaye 3.50} iti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf śravaṇapuṭanayanayugalaghrāṇamukhānāṃ nirodhanaṃ kāryam
+śuddhasuṣumṇāsaraṇau sphuṭam amalaḥ śrūyate nādaḥ (4.68)}  % metre? 
+\stopsloka
+
+ṣaṇmukhīmudrayā nādānusandhānam āha śravaṇeti. śravaṇapuṭe śrotrapuṭe nayanayor netrayor yugalaṃ
+yugmaṃ ghrāṇaśabdena ghrāṇapuṭe mukham āsyam eṣāṃ dvandve prāṇyaṅgatvād ekavadbhāve prāpte 'pi
+sarvasyāpi dvandvaikavadbhāvasya vaikalpikatvān na bhavati. teṣāṃ nirodhanaṃ karāṅgulibhiḥ
+kāryam. nirodhanaṃ cettham
+
+\startsloka
+aṅguṣṭhābhyām ubhau karṇau tarjanībhyāṃ ca cakṣuṣī
+nāsāpuṭau tathānyābhyāṃ pracchādya karaṇāni ca \comment{quotation untraced}
+\stopsloka
+
+iti cakārāt tadanyābhyāṃ mukhaṃ pracchādyeti samuccīyate. śuddhā prāṇāyāmair malarahitā yā
+suṣumnāsaraṇiḥ suṣumṇāpaddhatis tasyām amalo nādaḥ sphuṭam vyaktaṃ śrūyate.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf ārambhaś ca ghaṭaś caiva tathā paricayo 'pi ca
+niṣpattiḥ sarvayogeṣu syād avasthācatuṣṭayam (4.69)}
+\stopsloka
+
+atha nādasya catasro 'vasthāḥ prāha ārambhaś ceti. ārambhāvasthā ghaṭāvasthā paricayāvasthā
+niṣpattyavasthā iti. sarvayogeṣu sarveṣu cittavṛttinirodhopāyeṣu śāmbhavyādiṣu avasthācatuṣṭayam
+syāt. cacaivatathāpicāḥ pādapūraṇārthāḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf atha ārambhāvasthā
+brahmagranther bhaved bhedo hy ānandaḥ śūnyasambhavaḥ
+vicitraḥ kvaṇako dehe 'nāhataḥ śrūyate dhvaniḥ (4.70)}
+\stopsloka
+
+tatrārambhāvasthām āha brahmagranther iti. brahmagranther anāhatacakre vartamānāyā bhedaḥ
+prāṇāyāmābhyāsena bhedanaṃ yadā bhavet tadeti yattador adhyāhāraḥ. ānandayatīty ānandaḥ
+ānandajanakaḥ śūnye hṛdākāśe sambhavatīti śūnyasambhavo hṛdākāśotpanno vicitro nānāvidhaḥ kvaṇo
+bhūṣaṇaninadaḥ sa eva kvaṇakaḥ. bhūṣaṇaninadasadṛśa ity arthaḥ.
+
+\startsloka
+  \ \ \ \ \ \ \ \ \ \ \ \ \ bhūṣaṇānāṃ tu śiñjitam
+  nikvāṇo nikvaṇaḥ kvāṇaḥ kvaṇaḥ kvaṇanam ity api
+\stopsloka
+
+ity amaraḥ.\comment{Amarakośa 1.7.3} anāhataḥ dhvanir anāhato nirhrādo dehe dehamadhye śrūyate
+śravaṇaviṣayo bhavatīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf divyadehaś ca tejasvī divyagandhas tv arogavān
+sampūrṇahṛdayaḥ śūnya ārambhe yogavān bhavet (4.71)}
+\stopsloka
+
+divyadeha iti. śūnye hṛdākāśe ya ārambho nādārambhas tasmin sati
+hṛdākāśaviśuddhākāśa\comment{viśuddhākāśa \Wthree \Tue \EdLo \EdMu \lem viśuddhyākāśa \EdAd}bhrūmadhyākāśāḥ
+śūnyātiśūnyamahāśūnyaśabdair vyavahriyante yogibhiḥ. sampūrṇahṛdayaḥ prāṇavāyunā samyak pūrṇaṃ
+hṛdayaṃ yasya sa tathā ānandena vā pūrṇahṛdayaḥ yogavān yogī divyo rūpalāvaṇyabalasampanno deho
+yasya sa divyadehaḥ tejasvī pratāpavān divyagandho divya uttamo gandho yasya sa tathā arogavān
+rogarahito bhaved iti sambandhaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf atha ghaṭāvasthā
+dvitīyāyāṃ ghaṭīkṛtya vāyur bhavati madhyagaḥ
+dṛḍhāsano bhaved yogī jñānī devasamas tadā (4.72)}
+\stopsloka
+
+ghaṭāvasthām āha dvitīyāyām iti. dvitīyāyāṃ ghaṭāvasthāyāṃ vāyuḥ prāṇaḥ ghaṭīkṛtya ātmanā sahāpānaṃ
+nādabindū caikīkṛtya madhyago madhyacakragataḥ kaṇṭhasthāne madhyacakram. tad uktam atraiva
+jālandharabandhe
+
+\startsloka
+madhyacakram idaṃ jñeyaṃ ṣoḍaśādhārabandhanam \comment{HP 3.73}
+\stopsloka
+
+iti. yadā bhaved ity adhyāhāraḥ. tadā tasyām avasthāyāṃ yogī yogābhyāsī dṛḍham āsanaṃ yasya sa
+dṛḍhāsanaḥ sthirāsano jñānī pūrvāpekṣayā kuśalabuddhir devasamo rūpalāvaṇyādhikyād devatulyo
+bhavet. tad uktam īśvarokte rājayoge\comment{Yogabīja ????}
+
+\startsloka
+prāṇāpānau nādabindū jīvātmaparamātmanoḥ
+militvā ghaṭate yasmāt tasmāt sa ghaṭa ucyate
+\stopsloka
+
+iti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf viṣṇugranthes tato bhedāt paramānandasūcakaḥ
+atiśūnye vimardaś ca bherīśabdas tadā bhavet}
+\stopsloka
+
+viṣṇugranther iti. tato brahmagranthibhedanānantaraṃ viṣṇugrantheḥ kaṇṭhe vartamānāyā bhedāt
+kumbhakair bhedanāt paramānandasya bhāvino brahmānandasya sūcako jñāpakaḥ. atiśūnye kaṇṭhāvakāśe
+vimardo 'nekanādasaṃmardo bheryāḥ śabda iva śabdo bherīśabdo bherīnādaś ca tadā tasmin kāle bhavet.
+
+%  saṃmarda, the explanation in the Jyotsnā is again, the reading of another ms.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+atha paricayāvasthā  
+{\bf tṛtīyāyāṃ tu vijñeyo vihāyomardaladhvaniḥ
+mahāśūnyaṃ tadā yāti sarvasiddhisamāśrayam (4.74)}
+\stopsloka
+
+paricayāvasthām āha sārdhadvābhyāṃ tṛtīyāyām iti. tṛtīyāyāṃ paricayāvasthāyāṃ vihāyomardaladhvanir
+vihāyasi bhrūmadhyākāśe mardalasya vādyaviśeṣasya dhvanir iva dhvanir vijñeyo viśeṣeṇa jñānārho
+bhavati. tadā tasyām avasthāyāṃ sarvasiddhisamāśrayam sarvāsāṃ siddhīnām aṇimādīnāṃ samāśrayam
+sthānam tatra saṃyamād aṇimādiprāpteḥ. mahāśūnyaṃ bhrūmadhyākāśaṃ yāti gacchati prāṇa iti śeṣaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf cittānandaṃ tadā jitvā sahajānandasambhavaḥ
+doṣaduḥkhajarāvyādhikṣudhānidrāvivarjitaḥ (4.75)}
+\stopsloka
+
+cittānandam iti. cittānandaṃ nādaviṣayāntaḥkaraṇavṛttijanyaṃ sukhaṃ jitvābhibhūya
+sahajānandasambhavaḥ sahajānandaḥ svābhāvikam ātmasukhaṃ tasya sambhava āvirbhāvaḥ saḥ. doṣā
+vātapittakaphā duḥkhaṃ tajjanyā vedanā ādhyātmikādi ca jarā vṛddhāvasthā vyādhir jvarādiḥ kṣudhā
+bubhukṣā nidrā svāpa etābhir vivarjito rahitas tadā yogī bhavatīti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf rudragranthiṃ yadā bhittvā śarvapīṭhagato 'nilaḥ
+niṣpattau vaiṇavaḥ śabdaḥ kvaṇadvīṇākvaṇo bhavet (4.76)}
+\stopsloka
+
+%  Interesting singular reading of the Jyotsna: śarva for sarva
+
+tadā. kadā ity apekṣāyām āha rudreti. yadā rudragranthiṃ bhittvā ājñācakre rudragranthiḥ
+śarvasyeśvarasya pīṭhaṃ sthānaṃ bhrūmadhyaṃ tatra gataḥ prāpto 'nilaḥ prāṇo bhavati tadā.
+
+niṣpattyavasthām āha niṣpattāv iti. niṣpattau niṣpattyavasthāyām brahmarandhre gate prāṇe
+niṣpattyavasthā bhavati. vaiṇavaḥ veṇor ayaṃ vaiṇavo vaṃśasambandhī śabdo ninādaḥ kvaṇantī
+śabdāyamānā yā vīṇā tasyāḥ kvaṇaḥ śabdo bhavet.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf ekībhūtaṃ tadā cittaṃ rājayogābhidhānakam
+sṛṣṭisaṃhārakartāsau yogīśvarasamo bhavet (4.77)}
+\stopsloka
+
+tadā tasyām avasthāyāṃ cittam antaḥkaraṇam ekībhūtam ekaviṣayībhūtam viṣayaviṣayiṇor
+abhedopacārāt. tad rājayogābhidhānakam rājayoga ity abhidhānaṃ yasya tad rājayogābhidhānakam
+cittasyaikāgrataiva rājayoga ity arthaḥ. sṛṣṭisaṃhāreti asau nādānusandhānaparo yogī
+sṛṣṭisaṃhārakartā sṛṣṭiṃ saṃhāraṃ ca karotīti tādṛśaḥ. ata eveśvarasama īśvaratulyo bhavet.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf astu vā māstu vā muktir atraivākhaṇḍitaṃ sukham
+layodbhavam idaṃ saukhyaṃ rājayogād avāpyate (4.78)
+rājayogam ajānantaḥ kevalaṃ haṭhakarmiṇaḥ
+etān abhyāsino manye prayāsaphalavarjitān (4.79)}
+\stopsloka
+
+astu veti. rajayogam iti. ubhau prāg vyākhyātau.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf unmanyavāptaye śīghraṃ bhrūdhyānaṃ mama saṃmatam
+rājayogapadaṃ prāptuṃ sukhopāyo 'lpacetasām
+sadyaḥpratyayasandhāyī jāyate nādajo layaḥ (4.80)}
+\stopsloka
+
+% constellation of lines inconclusive in the collation of the original
+
+unmanyavāptaya iti. śīghraṃ tvaritam unmanyā unmanyavasthāyā avāptaye prāptyarthaṃ bhrūdhyānaṃ
+bhruvor dhyānaṃ bhrūmadhye dhyānaṃ mama svātmārāmasya saṃmatam. rājayogo yogānāṃ rājā tad eva padaṃ
+rājayogapadaṃ turyāvasthākhyaṃ prāptuṃ labdhuṃ pūrvoktabhrūdhyānarūpaḥ sukhopāyaḥ sukhasādhya
+upāyaḥ sukhopāyaḥ alpacetasām alpabuddhīnām api. kim utānyeṣām ity abhiprāyaḥ. nādajaḥ nādāj jāto
+layaś cittavilayaḥ sadyaḥ śīghraṃ pratyayaṃ pratītiṃ sandadhātīti pratyayasandhāyī pratītikaro
+jāyate prādurbhavati.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf nādānusandhānasamādhibhājāṃ
+yogīśvarāṇāṃ hṛdi vardhamānam
+ānandam ekaṃ vacasām agamyaṃ
+jānāti taṃ śrīgurunātha ekaḥ (4.81)}
+\stopsloka
+
+nādānusandhāneti. nādasya anāhatadhvaner anusandhānam anucintanaṃ tena samādhiś cittaigāgryaṃ taṃ
+bhajantīti nādānusandhānasamādhibhājaḥ teṣām yogiṣu yogayukteṣu īśvarāḥ samarthāḥ teṣām hṛdi
+hṛdaye vardhata iti vardhamānas taṃ vardhamānam vacasāṃ vācām agamyam idam iti vaktum aśakyaṃ taṃ
+yogaśāstraprasiddham ekaṃ mukhyam ānandam āhlādam eko 'nanyaḥ śrīgurunāthaḥ śrīmān gurur eva nātho
+jānāti vetti. etena nādānusandhānānando gurugamya eveti sūcitam.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf karṇau pidhāya hastābhyāṃ yaṃ śṛṇoti dhvaniṃ muniḥ
+tatra cittaṃ sthirīkuryād yāvat sthirapadaṃ vrajet (4.82)}
+\stopsloka
+
+nādānusandhānāt pratyāhārādikrameṇa samādhim āha karṇāv ity ādibhiḥ munir mananaśīlo yogī
+hastābhyām ity anena hastāṅguṣṭhau lakṣyete tābhyāṃ karṇau śrotre pidhāya hastāṅguṣṭhau
+śrotravivarayoḥ kṛtvety arthaḥ yaṃ dhvanim anāhataniḥsvanaṃ śṛṇoty ākarṇayati tatra tasmin
+dhvanau cittaṃ sthirīkuryād asthiraṃ sthiraṃ sampadyamānaṃ kuryāt. yāvat sthiraṃ padaṃ sthirapadaṃ
+turyākhyaṃ gacchet. tad uktaṃ turyāvasthā cidabhivyañjakanādasya vedanaṃ proktam iti
+nādānusandhānena vāyusthairyam aṇimādayo 'pi bhavanti. ity uktaṃ ca tripurāsārasamuccaye
+
+\startsloka
+vijito bhavatīha tena vāyuḥ sahajo yasya samutthitaḥ praṇādaḥ
+aṇimādiguṇā bhavanti tasyāmitapuṇyasya mahāguṇodayasya
+surarājatanūjavairirandhre vinirudhya svakarāṅgulidvayena
+jaladher iva dhīranādam antaḥ prasarantaṃ sahasā śṛṇoti martyaḥ \comment{Tripurāsārasamuccaya 3.5253} 
+\stopsloka
+
+iti. surarāja indras tasya tanūjo 'rjunas tasya vairī karṇas tadrandhre. spaṣṭam anyat.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf abhyasyamāno nādo 'yaṃ bāhyam āvṛṇute dhvanim
+pakṣād vikṣepam akhilaṃ jitvā yogī sukhī bhavet (4.83)}
+\stopsloka
+
+abhyasyamāna iti. abhyasyamāno 'nusandhīyamāno 'yaṃ nādo 'nāhatākhyo bāhyam dhvanim bahirbhavaṃ
+śabdam āvṛṇute śrutyor aviṣayaṃ karoti. yogī nādābhyāsī pakṣān māsārdhād akhilaṃ sarvaṃ vikṣepam
+cittacāñcalyaṃ jitvābhibhūya sukhī svānandī bhavet.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf śrūyate prathamābhyāse nādo nānāvidho mahān
+tato 'bhyāse vardhamāne śrūyate sūkṣmasūkṣmakaḥ (4.84)}
+\stopsloka
+
+śrūyata iti. prathamābhyāse pūrvābhyāse nānāvidho 'nekavidho mahān jaladhijīmūtabheryādisadṛśo
+nādo 'nāhatasvanaḥ śrūyate ākarṇyate. tato 'nantaram abhyāse nādānusandhānābhyāse vardhamāne sati
+sūkṣmasūkṣmakaḥ sūkṣmaḥ sūkṣma eva śrūyate śravaṇaviṣayo bhavati.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf ādau jaladhijīmūtabherījharjharasambhavāḥ
+madhye mardalaśaṅkhotthā ghaṇṭākāhalajās tathā (4.85)}
+\stopsloka
+
+nānāvidhaṃ nādam āha dvābhyāṃ ādāv iti. ādau vāyor brahmarandhragamanasamaye jaladhiḥ samudro
+jīmūto megho bherī vādyaviśeṣaḥ bherī strī dundubhiḥ pumān ity amaraḥ\comment{Amarakośa
+  ??}. jharjharo vādyaviśeṣaḥ
+
+\startsloka
+vādyaprabhedā ḍamarumaḍḍuḍiṇḍimajharjharāḥ
+mardalaḥ paṇavo 'nye ca\comment{Amarakośa
+  ??} 
+\stopsloka
+
+ity amaraḥ. jaladhipramukhebhyaḥ sambhava iva sambhavo yeṣāṃ te tathā madhye brahmarandhre
+vāyoḥ sthairyānantaraṃ mardalo vādyaviśeṣaḥ śaṅkho jalajas tābhyām utthā iva
+mardalaśaṅkhotthāḥ. ghaṇṭākāhalau vādyaviśeṣau tābhyāṃ jātā iva ghaṇṭākāhalajāḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf ante tu kiṅkiṇīvaṃśavīṇābhramaraniḥsvanāḥ
+iti nānāvidhā nādāḥ śrūyante dehamadhyagāḥ (4.86)}
+\stopsloka
+
+ante tv iti. ante tu prāṇasya brahmarandhre bahusthairyānantaraṃ tu kiṅkiṇī kṣudraghaṇṭikā
+vaṃśo veṇuḥ vīṇā tantrī bhramaro madhupaḥ teṣāṃ niḥsvanā iva niḥsvanā iti pūrvoktā nānāvidhā
+anekaprakārakā dehasya madhye gatāḥ prāptāḥ śrūyante.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf mahati śrūyamāṇe 'pi meghabheryādike dhvanau
+tatra sūkṣmāt sūkṣmataraṃ nādam eva parāmṛśet (4.87)}
+\stopsloka
+
+mahatīti. meghaś ca bherī ca te ādī yasya sa meghabheryādikaḥ tasmin. meghabherīśabdau
+tajjanyanirghoṣaparau. mahati bṛhati dhvanau nināde śrūyamāṇe ākarṇyamāne saty api tatra teṣu
+nādeṣu sūkṣmāt sūkṣmataram atisūkṣmaṃ nādam eva parāmṛśet cintayet. sūkṣmasya nādasya
+cirasthāyitvāt tatrāsaktaṃ cittaṃ ciraṃ sthirībhaved iti bhāvaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf ghanam utsṛjya vā sūkṣme sūkṣmam utsṛjya vā ghane
+ramamāṇam api kṣiptaṃ mano nānyatra cālayet (4.88)}      % = Nādabindūpaniṣat 37
+\stopsloka
+
+ghanam iti. ghanam mahāntaṃ nādaṃ meghabheryādikam utsṛjya tyaktvā sūkṣme kiṅkiṇībhramarādisvane vā
+sūkṣmam utsṛjya ghane vā nāde ramamāṇam ghanasūkṣmānyataranādagrahaṇaparityāgābhyāṃ krīḍantam api
+kṣiptaṃ rajasātyantacañcalaṃ mano 'nyatra viṣayāntare na cālayet na prerayet. kṣiptaṃ mano
+viṣayāntarāsaktaṃ na samādhīyate nādeṣv eva ramamāṇaṃ tu samādhīyate iti bhāvaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf yatra kutrāpi vā nāde lagati prathamaṃ manaḥ
+tatraiva susthirībhūya tena sārdhaṃ vilīyate (4.89)}
+\stopsloka
+
+yatreti. vā athavā yatra kutrāpi nāde yasmin kasmiṃścid ghane sūkṣme vā nāde prathamaṃ pūrvaṃ mano
+lagati lagnaṃ bhavati tatraiva tasminn eva nāde susthirībhūya samyak sthiraṃ bhūtvā tena nādena
+sārdhaṃ sākaṃ vilīyate līnaṃ bhavatīty arthaḥ. atra pūrvavākyena pratyāhāro dvitīyena dhāraṇā
+tṛtīyena dhyānadvārā samādhir uktaḥ.
+
+%  Here Patañjali's aṅgas are used for commentation instead of the four {\em avasthā}s. 
+% Could be Svātmārāma!
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf makarandaṃ piban bhṛṅgo gandhaṃ nāpekṣate yathā
+nādāsaktaṃ tathā cittaṃ viṣayān na hi kāṅkṣate (4.90)}
+\stopsloka
+
+makarandam iti. makarandaṃ puṣparasaṃ piban dhayan bhṛṅgo bhramaro gandhaṃ yathā nāpekṣate necchati
+tathā nādāsaktaṃ nāde āsaktaṃ cittam antaḥkaraṇaṃ viṣayān visinvanty avabadhnanti pramātāraṃ
+svasaṅgeti viṣayāḥ srakcandanavanitādayas tān na kāṅkṣate necchati. hīti niścaye.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf manomattagajendrasya viṣayodyānacāriṇaḥ
+niyamane samartho 'yaṃ ninādaniśitāṅkuśaḥ (4.91)}
+\stopsloka
+
+mana iti. viṣayaḥ śabdādir evodyānaṃ vanaṃ tatra caratīti viṣayodyānacārī tasya mana eva
+mattagajendro durnivāratvāt tasya nināda evānāhatadhvanir eva niśitāṅkuśaḥ tīkṣṇāṅkuśaḥ niyamane
+parāvartane samarthaḥ śaktaḥ. etaiḥ ślokaiḥ indriyāṇāṃ viṣayebhyaḥ pratyāharaṇaṃ pratyāhāraḥ
+
+\startsloka
+caratāṃ cakṣurādīnāṃ viṣayeṣu yathākramam
+yat pratyāharaṇaṃ teṣāṃ pratyāhāraḥ prakīrtitaḥ \comment{source ?}
+\stopsloka
+
+ityuktalakṣaṇaḥ pratyāhāraḥ proktaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf baddhaṃ tu nādabandhena manaḥ santyaktacāpalam
+prayāti sutarāṃ sthairyaṃ chinnapakṣaḥ khago yathā (4.92)}
+\stopsloka
+
+baddhaṃ tv iti. nāda eva bandhaḥ badhyate 'neneti bandhaḥ bandhanasādhanaṃ tena svaśaktyā
+svādhīnakaraṇena baddhaṃ bandhanam iva prāptam. nādadhāraṇādāv āsaktam ity arthaḥ. ata eva samyak
+tyaktaṃ santyaktaṃ cāpalam kṣaṇe kṣaṇe viṣayagrahaṇaparityāgarūpaṃ yena tat tathā manaḥ sutarāṃ
+sthairyaṃ prayāti nitarāṃ dhāraṇām eti. tatra dṛṣṭāntam āha chinnau pakṣau yasya tādṛśaḥ khe
+gacchatīti khagaḥ pakṣī yathā. etena
+
+\startsloka
+prāṇāyāmena pavanaṃ pratyāhāreṇa cendriyam
+vaśīkṛtya tataḥ kuryāc cittasthānaṃ śubhāśraye \comment{source ??}
+\stopsloka
+
+śubhāśraye cittasthāpanaṃ dhāraṇety uktalakṣaṇā dhāraṇā proktā.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf sarvacintāṃ parityajya sāvadhānena cetasā
+nāda evānusandheyo yogasāmrājyam icchatā (4.93)}
+\stopsloka
+
+sarvacintām iti. sarveṣāṃ bāhyabhyantaraviṣayāṇāṃ yā cintā cintanaṃ tāṃ parityajya tyaktvā
+sāvadhānena ekāgreṇa cetasā yogānāṃ sāmrājyam samrājo bhāvaḥ. yogaśabdo 'rśa
+ādyajantaḥ.\comment{explains sequence??} rājayogitvam iti yāvat. icchatā vāñchatā puṃsā nāda
+evānāhatadhvanir evānusandheyo 'nucintanīyaḥ. nādākāravṛttipravāhaḥ kartavya ity arthaḥ. etena
+
+\startsloka
+tadrūpapratyayaikāgryasantatiś cānyanispṛhā
+tad dhyānaṃ prathamair aṅgaiḥ ṣaḍbhir niṣpādyate nṛpa\comment{Viṣṇupurāṇam 6.7.89 ?}
+\stopsloka
+
+tatra pratyayaikatānatā dhyānam \comment{Yogasūtra 3.2} ityuktalakṣaṇaṃ dhyānam uktam.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf nādo 'ntaraṅgasāraṅgabandhane vāgurāyate
+antaraṅgakuraṅgasya vadhe vyādhāyate 'pi ca (4.94)}
+\stopsloka
+
+nādo 'ntaraṅgeti. nādaḥ antaraṅgaṃ mana eva sāraṅgo mṛgas tasya bandhane cāñcalyaharaṇe vāgurāyate
+vāgurevācarati. vāgurā jālam. yathā vāgurā bandhanena sāraṅgasya cāñcalyaṃ harati tathā nādo
+'ntaraṅgasya svaśaktyā cāñcalyaṃ haratīty arthaḥ. antaraṅgaṃ mana eva kuraṅgo hariṇas tasya vadhe
+nānāvṛttyutpādanāpanayanam eva manaso vadhas tasmin vyādhāyate vyādha ivācarati. yathā vyādho
+vāgurābaddhaṃ mṛgaṃ hanti evaṃ nādo 'pi svāsaktaṃ mano hantīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf antaraṅgasya yamino vājinaḥ parighāyate
+nādopāstirato nityam avadhāryā hi yoginā (4.95)}
+\stopsloka
+
+antaraṅgasyeti. yamino yogino 'ntaraṅgaṃ manas tasya capalatvād\var{capalatvād \Tue \EdAd\lem ca
+  palāyatvād \Wthree cāpalatvād \EdLo} vājino 'śvasya parighāyate vājiśālādvāraparigha ivācarati
+nāda iti śeṣaḥ. yathā vājiśālāparigho vājino 'nyatra gatiṃ ruṇaddhi tathā nādo 'ntaraṅgasyety
+arthaḥ. ataḥ kāraṇāt\var{ataḥ kāraṇāt \Wthree \Tue \EdMu \lem antaḥkāraṇād \EdLo (variants
+  unclear), antaḥkaraṇāt} \EdAd yoginā nādasyopāstir upāsanā nityaṃ pratyaham
+avadhāryāvadhāraṇīyā. hīti niścaye 'vyayam.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf baddhaṃ vimuktacāñcalyaṃ nādagandhakajāraṇāt
+manaḥpāradam āpnoti nirālambākhyakhe 'ṭanam (4.96)}
+\stopsloka
+
+baddham iti. nāda eva gandhaka upadhātuviśeṣas tena jāraṇaṃ jīrṇīkaraṇaṃ nādagandhakasambandhena
+cāñcalyaharaṇaṃ tasmād baddhaṃ nādaikāsaktaṃ pakṣe guṭikākṛtiṃ prāptaṃ ata eva vimuktaṃ tyaktaṃ
+cāñcalyam anekaviṣayākārapariṇāmarūpaṃ yena. pakṣe vimuktalaulyaṃ manaḥ pāradaṃ mana eva pāradaṃ
+cañcalaṃ nirālambaṃ brahma tad evākhyā yasya tan nirālambākhyaṃ tad eva kham aparicchinnatvāt
+tasminn aṭanam gamanaṃ tadākāravṛttipravāham. pakṣe ākāśagamanaṃ prāpnoti. yathā baddhaṃ pāradam
+ākāśagamanaṃ karoti evaṃ baddhaṃ mano brahmākāravṛttipravāham avicchinnaṃ karotīty arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf nādaśravaṇataḥ kṣipram antaraṅgabhujaṅgamam
+vismṛtya sarvam ekāgraḥ kutracin na hi dhāvati (4.97)}
+\stopsloka
+
+nādeti. nādasyānāhatasvanasya śravaṇataḥ śravaṇāt kṣipram drutam antaraṅgaṃ mana eva bhujaṅgamaḥ
+sarpaś capalatvān nādapriyatvāc ca bhujaṅgamena rūpakaṃ manasaḥ. sarvaṃ viśvam vismṛtya
+vismṛtiviṣayaṃ kṛtva ekāgro nādākāravṛttipravāhavān san kutrāpi viṣayāntare nahi dhāvati naiva
+dhāvanaṃ karoti. dhyānottaraiḥ ślokaiḥ
+
+\startsloka
+tasyaiva kalpanāhīnaṃ svarūpagrahaṇaṃ hi yat
+manasā dhyānaniṣpādyaṃ samādhiḥ so 'bhidhīyate \comment{Viṣṇupurāṇa 6.7.92}
+\stopsloka
+
+iti viṣṇupurāṇoktalakṣaṇaḥ tad evārthamātranirbhāsaṃ svarūpaśūnyam iva samādhiḥ\comment{Yogasūtra
+  3.3} iti pātañjalasūtroktalakṣaṇaś ca samprajñātalakṣaṇaḥ samādhir uktaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf kāṣṭhe pravartito vahniḥ kāṣṭhena saha śāmyati
+nāde pravartitaṃ cittaṃ nādena saha līyate (4.98)}
+\stopsloka
+
+kāṣṭha iti. kāṣṭhe dāruṇi pravartitaḥ prajvālito vahniḥ kāṣṭhena saha śāmyati jvālārūpaṃ parityajya
+tejomātra\var{tejomātra \Wthree \Tue \lem tanmātra \EdAd}rūpeṇāvatiṣṭhate yathā tathā nāde pravartitaṃ
+cittaṃ nādena saha līyate. rājasatāmasavṛttināśāt sattvamātrāvaśeṣaṃ saṃskāraśeṣaṃ ca
+bhavati. tatra ca maitrāyaṇīmantraḥ
+
+\startsloka
+yathā nirindhano vahniḥ svayonāv upaśāmyati
+tathā vṛttikṣayāc cittaṃ svayonāv upaśāmyati \comment{Maitrāyanīyopaniṣat 6.1} iti.
+\stopsloka
+
+% 
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf ghaṇṭādinādasaktastabdhāntaḥkaraṇahariṇasya
+  praharaṇam api sukaraṃ syāc charasandhānapravīṇaś
+  \hfill cet (4.99)} \comment{{\em syāt} is missing in all
+sources except \EdAd.}
+\stopsloka
+
+ghaṇṭādīti. ghaṇṭā ādir yeṣāṃ śaṅkhamardalajharjharadundubhi\-jīmūtā\-dīnāṃ te ghaṇṭādayaḥ teṣāṃ nādāḥ
+teṣu saktaḥ. ata eva stabdho niścalo yo 'ntaḥkaraṇam eva hariṇo mṛgaḥ tasya praharaṇam
+nānāvṛttyutpattipratibandhanam\var{utpatti \Wthree \Tueac \EdLo \ om \Tuepc \EdAd} antaḥkaraṇapakṣe.
+
+hariṇapakṣe tu praharaṇaṃ hananam api śaravad drutagāmino vāyoḥ sandhānaṃ
+suṣumnāmārgeṇa brahmarandhre nirodhanam\var{nirodhanam \Tue \Wthree \lem nirodhana (probably misprint)
+  \EdLo}. pakṣe śarasya bāṇasya sandhānaṃ dhanuṣi yojanaṃ tasmin pravīṇaḥ kuśalaś cet sukaraṃ
+sukhena kartuṃ śakyaṃ syāt.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf anāhatasya śabdasya dhvanir ya upalabhyate
+dhvaner antargataṃ jñeyaṃ jñeyasyāntargataṃ manaḥ
+manas tatra layaṃ yāti tad viṣṇoḥ paramaṃ padam (4.100)}
+\stopsloka
+
+anāhatasyeti. anāhatasya śabdasyānāhatasvanasya yo dhvanir nirhrāda upalabhyate śrūyate tasya
+dhvaner antargataṃ jñeyaṃ jyotiḥ svaprakāśacaitanyam. jñeyasyāntargataṃ jñeyākāratām āpannaṃ
+mano 'ntaḥkaraṇaṃ tatra jñeye mano vilayaṃ yāti paravairāgyeṇa sakalavṛttiśūnyaṃ saṃskāraśeṣaṃ
+bhavati. tad viṣṇoḥ vibhor ātmanaḥ paramam antaḥkaraṇavṛttyupādhirāhityān nirupādhikaṃ padyate
+gamyate yogibhir iti padam svarūpam.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf tāvad ākāśasaṅkalpo yāvac chabdaḥ pravartate
+niḥśabdaṃ tat paraṃ brahma paramātmeti gīyate (4.101)}
+\stopsloka
+
+tāvad iti. yāvac chabdo 'nāhatadhvaniḥ pravartate śrūyate tāvad ākāśasya samyak kalpanaṃ
+bhavati. śabdasyākāśaguṇatvād guṇaguṇinor abhedād vā manasā saha śabdasya vilayān niḥśabdaṃ
+śabdarahitaṃ yat paraṃ brahma parabrahmaśabdavācyaṃ paramātmeti gīyate paramātmaśabdena sa
+ucyate. sarvavṛttivilaye yaḥ svarūpeṇāvasthitaḥ sa eva parabrahmaparamātmaśabdābhyām ucyate iti
+bhāvaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf yat kiñcin nādarūpeṇa śrūyate śaktir eva sā
+yas tattvānto nirākāraḥ sa eva parameśvaraḥ (4.102)
+ iti nādānusandhānam \var{iti nādānusandhānam \EdAd \Wthree \lem om \Tue \EdLo}}
+\stopsloka
+
+yat kiñcid iti. nādarūpeṇa anāhatadhvanirūpeṇa yat kiñcic chrūyate ākarṇyate sā śaktir eva. yas
+tattvāntas tattvānām anto layo yasmin saḥ. tathā nirākāra ākārarahitaḥ sa eva parameśvaraḥ
+sarvavṛttikṣaye svarūpāvasthito yaḥ sa ātmety arthaḥ kāṣṭhe pravartito vahniḥ \comment{4.98} ity
+ādibhiḥ ślokaiḥ rājayogāparaparyāyo 'samprajñātaḥ samādhir uktaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf sarve haṭhalayopāyā rājayogasya siddhaye
+rājayogasamārūḍhaḥ puruṣaḥ kālavañcakaḥ (4.103)}
+\stopsloka
+
+sarva iti. haṭhaś ca layaś ca haṭhalayau tayor upāyā haṭhalayopāyā haṭhopāyā āsanakumbhakamudrārūpā
+layopāyā nādānusandhānaśāmbhavīmudrādayaḥ. rājayogasya manasaḥ sarvavṛttinirodhalakṣaṇasya
+siddhaye niṣpattaye proktā iti śeṣaḥ. rājayogaṃ samārūḍhaḥ samyag ārūḍhaḥ prāptavān ya puruṣaḥ sa
+kālavañcakaḥ kālaṃ mṛtyuṃ vañcayati jayatīti tādṛśaḥ syād iti śeṣaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf tattvaṃ bījaṃ haṭhaḥ kṣetram audāsīnyaṃ jalaṃ tribhiḥ
+unmanīkalpalatikā sadya eva pravartate (4.104)}
+\stopsloka
+
+tattvam iti. tattvaṃ cittaṃ bījaṃ bījavad unmanyavasthāṅkurākāreṇa pariṇamamānatvāt. haṭhaḥ
+prāṇāpānayor aikyalakṣaṇaḥ prāṇāyāmaḥ kṣetram\var{kṣetraṃ \EdAd \lem om \Wthree \EdLo} kṣetra iva
+prāṇāyāme unmanīkalpalatikotpatteḥ audāsīnyaṃ paraṃ vairāgyaṃ jalaṃ tasyā
+utpattikāraṇatvāt. paravairāgyahetukaḥ saṃskāraviśeṣaś cittasyāsamprajñātaḥ iti tallakṣaṇāt. etais
+tribhir unmanyasamprajñātāvasthā saiva kalpalatikā sakaleṣṭasādhanatvāt sadya eva śīghram eva
+pravartate pravṛttā bhavati utpannā bhavati.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf sadā nādānusandhānāt kṣīyante pāpasañcayāḥ
+nirañjane vilīyete niścitaṃ cittamārutau (4.105)}
+\stopsloka
+
+sadeti. sadā sarvadā nādānusandhānāt nādānucintanāt pāpasañcayāḥ pāpasamūhāḥ kṣīyante
+naśyanti nirañjane nirguṇe caitanye niścitaṃ dhruvaṃ cittamārutau manaḥprāṇau vilīyete vilīnau
+bhavataḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf śaṅkhadundhubhinādaṃ ca na śṛṇoti kadācana
+kāṣṭhavaj jāyate deha unmanyāvasthayā dhruvam (4.106)}
+\stopsloka
+
+unmanyavasthāṃ prāptasya yoginaḥ sthitim āhāṣṭabhiḥ śaṅkhadundhubhīti. śaṅkho jalajo dundhubhir
+vādyaviśeṣaḥ tayor nādaṃ ghoṣaṃ kadācana kasmiṃścid api samaye na śṛṇoti. śaṅkhadundhubhīty
+upalakṣaṇaṃ nādamātrasya. unmanyāvasthayā deho dhruvam kāṣṭhavaj jāyate niśceṣṭatvād ity arthaḥ.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf sarvāvasthāvinirmuktaḥ sarvacintāvivarjitaḥ
+mṛtavat tiṣṭhate yogī sa mukto nātra saṃśayaḥ (4.107)}
+\stopsloka
+
+jāgratsvapnasuṣuptimūrcchāmaraṇalakṣaṇāḥ pañca vyutthānāvasthās tābhir viśeṣeṇa mukto rahitaḥ sarvā
+yāś cintāḥ smṛtayaḥ tābhir vivarjito virahito yaḥ yogaḥ sakalavṛttinirodho 'syāstīti yogī
+turyāvasthāvān sa mukto jīvann eva muktaḥ. sakalavṛttinirodhe ātmanaḥ svarūpāvasthānāt. tad uktaṃ
+pātañjale sūtre tadā draṣṭuḥ svarūpe'vasthānam \comment{Yogasūtra 1.3} iti. spaṣṭam anyat.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf khādyate na ca kālena bādhyate na ca karmaṇā
+sādhyate na sa kenāpi yogī yuktaḥ samādhinā (4.108)}
+\stopsloka
+
+khādyata iti. samādhinā yukto yogī kālena mṛtyunā na khādyate na bhakṣyate na hanyata ity
+arthaḥ. karmaṇā kṛtena śubhenāśubhena vā na bādhyate janmamaraṇādijananena na kleśyate. tathā ca
+samādhiprakaraṇe pātañjalasūtraṃ tataḥ kleśakarmanivṛttiḥ \comment{Yogasūtra 4.30} iti. kenāpi
+puruṣāntareṇa yantramantrādinā vā na sādhyate sādhayituṃ na śakyate.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf na gandhaṃ na rasaṃ rūpaṃ na ca sparśaṃ na niḥsvanam
+nātmānaṃ na paraṃ vetti yogī yuktaḥ samādhinā (4.109)}
+\stopsloka
+
+na gandham iti. samādhinā yukto yogī na gandhaṃ surabhim asurabhiṃ vā na rasaṃ
+madhurāmlalavaṇakaṭukaṣāyatiktabhedāt ṣaḍvidhaṃ na rūpaṃ śuklanīlapītaraktaharitakapiśacitrabhedāt
+saptavidhaṃ na sparśaṃ śītam uṣṇam anuṣṇaśītaṃ vā na niḥsvanam
+śaṅkhadundubhijaladhijīmūtādininādaṃ bāhyam ābhyantaraṃ vā nātmānaṃ dehaṃ na paraṃ puruṣāntaraṃ
+vettīti sarvatrānveti. ātmā dehe ghṛtau jīve svabhāve paramātmani iti vaijayantī.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf cittaṃ na suptaṃ no jāgratsmṛtivismṛtivarjitam
+na cāstam eti nodeti yasyāsau mukta eva saḥ (4.110)}
+\stopsloka
+
+cittam iti. yasya yoginaś cittam antaḥkaraṇaṃ na suptaṃ āvarakasya tamaso 'bhāvāt triguṇe
+'ntaḥkaraṇe yadā sattvarajasī abhibhūya samastakaraṇāvarakaṃ tama āvirbhavati tadāntaḥkaraṇasya
+viṣayākārapariṇāmābhāvāt tat suptam ity ucyate. no jāgrad indriyair arthagrahaṇābhāvāt. smṛtiś ca
+vismṛtiś ca smṛtivismṛtī tābhyāṃ varjitam. vṛttisāmānyābhāvād udbodhakābhāvāc ca
+smṛtivarjitam. smṛtyanukūlasaṃskārābhāvād\var{bhāvād \Wthree \EdLo \lem bhāvābhāvād \Tue}
+vismṛtivarjitam. na cāstam nāśam eti prāpnoti saṃskāraśeṣasya cittasya sattvāt. nodety
+udbhavati vṛttyanutpādanāt. so 'sau mukta eva jīvanmukta eva.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf na vijānāti śītoṣṇaṃ na duḥkhaṃ na sukhaṃ tathā
+na mānaṃ nāpamānaṃ ca yogī yuktaḥ samādhinā (4.111)}
+\stopsloka
+
+na vijānātīti. samādhinā yukto yogī śītaṃ ca uṣṇaṃ ca śītoṣṇaṃ samāhāradvandvaḥ śītam uṣṇaṃ vā
+padārthaṃ na duḥkhaṃ duḥkhajanakam parakṛtaṃ tāḍanādikaṃ na sukhaṃ sukhasādhanaṃ
+surabhicandanādyanulepanādikaṃ na tathā cārthe mānaṃ parakṛtaṃ satkāraṃ na apamānam anādaraṃ ca
+na vijānātīti kriyāpadaṃ prativākyam anveti.
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf svastho jāgradavasthāyāṃ suptavad yo 'vatiṣṭhate
+niḥśvāsocchvāsahīnaś ca niścitaṃ mukta eva saḥ (4.112)}
+\stopsloka
+
+svastha iti. svasthaḥ prasannendriyāntaḥkaraṇaḥ etena tandrāmūrcchādi\-vyāvṛttiḥ jāgradavasthāyām ity
+anena svapnasuṣuptyor nivṛttiḥ suptavat suptena tulyaṃ kāyendriyavyāpāraśūnyo yo yogī avatiṣṭhate
+sthito bhavati samavapravibhya sthaḥ \comment{Aṣṭādhyāyī 1.3.22} ity
+ātmanepadam. niśvāsocchvāsahīnaḥ bāhyavāyoḥ koṣṭhe grahaṇaṃ niḥśvāsaḥ koṣṭhasthitasya vāyor
+bahirniḥsaraṇam ucchvāsas tābhyāṃ hīnaś cāvatiṣṭhata ity atrāpi sambadhyate. sa niścitaṃ
+niḥsandigdhaṃ mukta eva jīvanmukta eva. tat uktaṃ dattatreyeṇa
+
+\startsloka
+nirguṇadhyānasampannaḥ samādhiṃ ca tato 'bhyaset
+dinadvādaśakenaiva samādhiṃ samavāpnuyāt
+vāyuṃ nirudhya medhāvī jīvanmukto bhaved dhruvam \comment{??} iti
+\stopsloka
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf avadhyaḥ sarvaśastrāṇām aśakyaḥ sarvadehinām
+agrāhyo mantrayantrāṇāṃ yogī yuktaḥ samādhinā (4.113)}
+\stopsloka
+
+avadhya iti. samādhinā yukto yogī. sarvaśastrāṇām iti sambandhasāmānye ṣaṣṭhī. sarvaśāstrair ity
+arthaḥ. avadhyo hantum aśakya ity arthaḥ. sarvadehinām ity atrāpi sambandhamātravivakṣāyāṃ
+ṣaṣṭhī. aśakyaḥ sarvadehibhiḥ balena śakyo na bhavatīty arthaḥ. mantrayantrāṇāṃ
+vaśīkaraṇamāraṇoccāṭanādiphalair mantrayantrair agrāhyaḥ vaśīkartum\var{vaśīkartum \EdAd \lem
+  vaśyādikaḥ kartum Edlo \Tue} aśakyaḥ. evaṃ prāptayogasya yogino vighnā bahavaḥ
+samāyānti. tannivāraṇārthaṃ tajjñānasyāpekṣitatvāt te 'pi pradarśyante. dattātreyaḥ
+
+\startsloka
+ālasyaṃ prathamo vighno dvitīyas tu prakathyate
+pūrvoktadhūrtagoṣṭhī ca tṛtīyo mantrasādhanam
+caturtho dhātuvādaḥ syād iti yogavido viduḥ
+\stopsloka
+
+mārkaṇḍeyapurāṇe\var{mārkaṇḍeyapurāṇe \Wthree \lem iti mārkaṇḍeyapurāṇe \Tue}
+
+\startsloka
+upasargāḥ pravartante dṛṣṭā hy ātmani yoginaḥ
+ye tāṃs te sampravakṣyāmi samāsena nibodha me
+kāmyāḥ kriyās tathā kāmān manuṣyo yo 'bhivāñchati\var{manuṣyo yo 'bhivāñchati \EdLo \lem manuṣyān abhivāñchati \Wthree}
+striyo dānaphalaṃ vidyāṃ māyāṃ kupyaṃ dhanaṃ vasu
+devatvam amareśatvaṃ rasāyanavayaḥ\var{vayaḥ \Wthree \lem cayaḥ \EdLo} kriyāḥ\var{kriyāḥ \EdAd \lem kriyām \EdLo \Wthree}
+meruprapatanaṃ yajñaṃ jalāgnyāveśanaṃ tathā
+śrāddhānāṃ sarva\var{sarva \EdAd\lem śakti \EdLo}dānānāṃ phalāni niyamāṃs tathā
+tathopavāsāt pūrtāc ca devapitrarcanād\var{devapitrarcanād \EdLo \lem devatābhyarcanād \EdAd} api
+tebhyas tebhyaś\var{tebhyas tebhyaś \EdAd \lem atitibhyaś \EdLo \Wthree} ca karmabhya upasṛṣṭo 'bhivāñchati 
+cittam\var{cittam \EdAd \lem vighnam \EdLo} itthaṃ vartamānaṃ\var{vartamānaṃ \EdAd \lrm pravarte taṃ } yatnād yogī nivartayet
+brahmasaṅgi manaḥ kurvann upasargaiḥ\var{upasargaiḥ \EdLo \lem upasargāt \EdAd} pramucyate %
+\comment{Mārkaṇḍeyapurāṇa 40.16} 
+\stopsloka
+
+iti.
+
+skandapurāṇe
+
+\startsloka
+yadaibhir antarāyair na kṣipyate 'sya hi mānasam
+tadāgre tam avāpnoti paraṃ brahmātidurlabham 
+\stopsloka
+
+yogabhāskare
+
+\startsloka
+sāttvikīṃ dhṛtim ālambya yogī sattvena susthiraḥ
+nirguṇaṃ manasā dhyāyann upasargaiḥ pramucyate 
+evaṃ yogam upāsīnaḥ śakrādipadanispṛhaḥ
+siddhyādivāsanātyāgī jīvanmukto bhaven muniḥ
+vistarasya bhiyā notkaḥ santi vighnā hy anekaśaḥ
+dhyānena viṣṇuharayor vāraṇīyā hi yoginā iti
+\stopsloka
+
+%\begin{vsid}{hp04_00}
+\startsloka
+{\bf
+yāvan naiva praviśati caran māruto madhyamārge
+yāvad bindur na bhavati dṛḍhaḥ prāṇavātaprabandhāt
+yāvad dhyāne sahajasadṛśaṃ jāyate naiva tattvaṃ
+tāvaj jñānaṃ vadati tad idaṃ dambhamithyāpralāpaḥ (4.114)}
+\stopsloka
+
+iti śrīsahajānandasantānacintāmaṇisvātmārāmayogīndraviracitāyāṃ haṭhapradīpikāyāṃ
+samādhilakṣaṇaṃ nāma caturthopadeśaḥ.
+
+ayogināṃ jñānaṃ nirākurvan yoginām eva jñānaṃ bhavatīty āha yāvad iti. madhyamārge suṣumnāyāṃ caran
+gacchan mārutaḥ prāṇavāyuḥ yāvat yāvatkālaparyantaṃ na praviśati prakarṣeṇa brahmarandhraparyantaṃ
+na viśati. brahmarandhraṃ gatasya sthairyād brahmarandhraṃ gatvā na sthirībhavatīty
+arthaḥ. suṣumnāyām asañcaran vāyur asiddha ity ucyate. tad uktam amṛtasiddhau
+
+\startsloka
+yāvad vimārgago vāyur niścalo naiva madhyagaḥ
+asiddhaṃ taṃ vijānīyād vāyuṃ karmavaśānugam 
+\stopsloka
+
+iti. prāṇayati jīvayati iti prāṇaḥ. sa cāsau vātaś ca prāṇavātaḥ tasya prabandhāt kumbhakena
+sthirīkaraṇād bindur vīryaṃ dṛḍhaḥ sthiro na bhavati. prāṇavātasthairye bindusthairyam uktam
+atraiva prāk.
+
+\startsloka
+manaḥsthairye sthiro vāyus tato binduḥ sthiro bhavet \comment{4.28}
+\stopsloka
+
+iti. tadabhāve tv asiddhatvaṃ yoginaḥ. uktam amṛtasiddhau
+
+\startsloka
+tāvad baddho 'py asiddho 'sau naraḥ sāṃsāriko mataḥ
+yāvad bhavati dehastho rasendro brahmarūpakaḥ
+asiddhaṃ taṃ vijānīyān naram abrahmacāriṇam
+jarāmaraṇasaṅkīrṇaṃ sarvakleśasamāśrayam 
+\stopsloka
+
+iti. yāvat tattvaṃ cittaṃ dhyāne dhyeyacintane sahajasadṛśaṃ svābhāvikadhyeyākāravṛttipravāhavat
+naiva jāyate naiva bhavati prāṇavātaprabandhād iti dehalīdīpanyāyenātrāpi
+sambadhyate. vāyusthairye cittasthairyam uktam amṛtasiddhau
+
+\startsloka
+yadāsau mriyate\var{mriyate \lem śriyate check} vāyur madhyamāṃ madhyayogataḥ
+tadā binduś ca cittaṃ ca mriyate vāyunā saha
+\stopsloka
+
+tadabhāve 'hy asiddhatvam uktam amṛtasiddhau
+
+\startsloka
+yāvat prasyandate cittaṃ bāhyābhyantaravastuṣu
+asiddhaṃ tad vijānīyāc cittaṃ karmaguṇānvitam 
+\stopsloka
+
+iti. tāvad yaj jñānaṃ śābdaṃ vadati kaścit tad idaṃ jñānakathanaṃ dambhamithyāpralāpaḥ dambhena
+jñānakathanenāhaṃ loke pūjyo bhaviṣyāmīti dhiyā mithyāpralāpo mithyābhāṣaṇaṃ dambhapūrvakaṃ
+mithyābhāṣaṇam ity arthaḥ. prāṇabinducittānāṃ jayābhāve jñānasyābhāvāt saṃsṛtir durvārā. tad uktam
+amṛtasiddhau
+
+\startsloka
+calaty eṣa yadā vāyus tadā binduś calaḥ smṛtaḥ
+binduś calati yasyāṅge cittaṃ tasyaiva cañcalam
+cale bindau cale citte cale vāyau ca sarvadā
+jāyate mriyate lokaḥ satyaṃ satyam idaṃ vacaḥ 
+\stopsloka
+
+iti. yogabīje 'py uktam
+
+\startsloka
+cittaṃ pranaṣṭaṃ yadi bhāsate vai
+tatra pratīto maruto 'pi nāśaḥ
+na vā yadi syān na tu tasya śāstraṃ
+nātmapratītir na gurur na mokṣaḥ \comment{Yogabīja 138}  iti.
+\stopsloka
+
+etena prāṇabindumanasāṃ jaye tu jñānadvārā yogino muktiḥ syād eveti sūcitam. tad uktam
+amṛtasiddhau\comment{Amṛtasiddhi 7.6cd, 7.16ab, 7.7ab, 7.7cd, 7.16cd}
+
+\startsloka
+yām avasthāṃ vrajed vāyur bindus tām adhigacchati
+yathā hi sādhyate vāyus tathā binduprasādhanam
+mūrcchito harati vyādhiṃ vṛddhaḥ khecaratāṃ nayet
+sarvasiddhikaro līno niścalo muktidāyakaḥ
+yathāvasthā bhaved bindoś cittāvasthā tathā tathā 
+\stopsloka
+
+iti. nanu
+
+\startsloka
+yogās trayo mayā proktā nṝṇāṃ śreyovidhitsayā
+jñānaṃ karma ca bhaktiś ca nopāyo 'nyo 'sti kutracit \comment{Bhāgavatapurāṇa 11.20.6}
+\stopsloka
+
+iti bhagavaduktās trayo mokṣopāyās teṣu satsu kathaṃ yoga eva mokṣopāyatvenokta iti cen na teṣāṃ
+yogāṅgeṣv antarbhāvāt. tathā hi
+
+\startsloka
+ātmā vā are draṣṭayaḥ śrotavyo mantavyo nididhyāsitavyaḥ \comment{Bṛhad\-āraṇya\-kopani\-ṣat 2.4.5}
+\stopsloka
+
+iti śrutyā paramapuruṣārthasādhanātmasākṣātkārahetutayā śravaṇamanananididhyāsanāny uktāni. tatra
+śravaṇamanane niyamāntargate svādhyāye 'ntarbhavataḥ. svādhyāyaś ca mokṣaśāstrāṇām adhyayanam. sa
+ca tātparyārthanirṇayaparyavasāyī grāhyaḥ. tātparyārthanirṇayaś ca śravaṇamananābhyāṃ bhavatīti
+śravaṇamananayoḥ svādhyāye 'ntarbhāvaḥ. niyamavivaraṇe yājñavalkyena
+
+\startsloka
+  siddhāntaśravaṇaṃ proktaṃ vedāntaśravaṇaṃ budhaiḥ \comment{Yajñavalkya 2.8}
+\stopsloka
+
+iti spaṣṭam eva śravaṇasya niyamāntargatir uktā.
+
+\startsloka
+adhītya vedaṃ sūtraṃ vā purāṇaṃ setihāsakam
+padeṣv adhayayanaṃ yac ca sadābhyāso japaḥ smṛtaḥ \comment{Yajñavalkya 2.13} 
+\stopsloka
+
+iti yuktibhir anavaratam anucintanalakṣaṇasya sadābhyāsarūpasya mananasyāpi niyamāntargatir
+uktā. vijātīyapratyayanirodhapūrvakasajātīyapra\-tyaya\-pravāharūpasya nididhyāsanasya uktalakṣaṇe
+dhyāne 'ntarbhāvaḥ. tasyāpi tatparipākarūpasamādhinātmasākṣātkāradvārā mokṣahetutvam
+īśvarārpaṇabuddhyā niṣkāmakarmānuṣṭhānalakṣaṇasya karmayogasya tapaḥsvādhyāyeśvarapraṇidhānāni
+kriyāyogaḥ\comment{Yogasūtra 2.1} iti patañjaliprokte niyamāntargate kriyāyoge 'ntarbhāvaḥ. tatra
+tapa uktam īśvaragītāyāṃ
+
+\startsloka
+upavāsaparākādi kṛcchracāndrāyaṇādibhiḥ
+śarīraśoṣaṇaṃ prāhus tāpasās tapa uttamam \comment{Īśvaragītā 11.21} 
+\stopsloka
+
+iti. svādhyāyo 'pi tatroktaḥ
+
+\startsloka
+vedāntaśatarudrīyapraṇavādijapaṃ budhāḥ
+sattvaśuddhikaraṃ puṃsāṃ svādhyāyaṃ paricakṣate \comment{Īśvaragītā 11.22} iti.
+\stopsloka
+
+īśvarapraṇidhānaṃ ca tatroktam
+
+\startsloka
+stutismaraṇapūjābhir vāṅmanaḥkāyakarmabhiḥ
+suniścalā bhaved bhaktir etad īśvarapūjanam\comment{Īśvaragītā 11.29} 
+\stopsloka
+
+iti. kriyāyogaś ca paramparayā samādhinātmasākṣātkāradvāraiva mokṣahetur iti
+
+\startsloka
+  samādhibhāvanārthaḥ kleśatanūkaraṇārthaś ca \comment{Yogasūtra 2.2}
+\stopsloka
+
+ity uttarasūtreṇa spaṣṭīkṛtaṃ patañjalinā. bhajyate sevyate bhagavadākāram antaḥkaraṇaṃ
+kriyate 'nayeti bhaktir iti karaṇavyutpattyā
+
+\startsloka
+śravaṇaṃ kīrtanaṃ viṣṇoḥ smaraṇaṃ pādasevanam
+arcanaṃ vandanaṃ dāsyaṃ sakhyam ātmanivedanam \comment{Bhāgavatapurāṇa 7.5.23}
+\stopsloka
+
+iti navadhoktā sādhanabhaktir abhidhīyate. tasyā īśvarapraṇidhānarūpe niyame 'ntarbhāvaḥ. tasyāś ca
+samādhihetutvaṃ coktam patañjalinā īśvarapraṇidhānād vā \comment{Yogasūtra 1.23} iti
+īśvaraviṣayakāt praṇidhānād bhaktiviśeṣāt samādhilābhaḥ samādhiphalaṃ ca bhavatīti 
+sūtrārthaḥ. bhajanam antaḥkaraṇasya bhagavadākāratārūpaṃ bhaktir iti bhāvavyutpattyā phalabhūtā
+bhaktir ity abhidhīyate. saiva premabhaktir ity ucyate. tallakṣaṇam uktaṃ
+nārāyaṇatīrthaiḥ
+
+\startsloka
+premabhaktiyogas tu īśvaracaraṇāravindaviṣayakaikāntikātyantikapremapravāho 'vicchinnaḥ
+\stopsloka
+
+iti. madhusūdanasarasvatībhis tu
+
+\startsloka
+dravībhāvapūrvikā manaso bhagavadākāratārūpā savikalpavṛttir bhaktiḥ \comment{bhaktirasāyaṇa 1.1 (vṛtti)}
+\stopsloka
+
+iti. tasyās tu śraddhābhaktidhyānayogād avehi \comment{Kaivalyopaniṣat 2} iti śruteḥ. bhaktyā mām
+abhijānāti \comment{Bhagavadgītā 18.55} iti smṛteś ca ātmasākṣātkāradvārā mokṣahetutvam. bhaktās
+tu sukhasyaiva puruṣārthatvād duḥkhāsambhinnaniratiśayasukhadhārārūpā premabhaktir eva puruṣārtha
+ity āhuḥ. tasyās tu samprajñātasamādhāv antarbhāvaḥ. evaṃ ca aṣṭāṅgayogātiriktaṃ kim api
+paramapuruṣārthasādhanaṃ nāstīti siddham.
+
+\startsloka
+  grāhyam eva viduṣāṃ hitaṃ yato
+  bhāṣaṇaṃ mama yad apy asaṃskṛtam
+  rakṣa gacchati payo 'nalāhitaṃ
+  hy amba ity abhihitaṃ śiśor yathā
+
+  sadarthadyotanakarī tamaḥstomavināśinī
+  brahmānandena jyotsneyaṃ śivāṅghriyugale 'rpitā
+\stopsloka
+
+%iti śrīhaṭhayogapradīpikāvyākhyāyāṃ brahmānandakṛtāyāṃ jyotsnābhidhāyāṃ samādhinirūpaṇaṃ nāma
+%caturthopadeśaḥ.
+
+\stoplinenumbering\stoptext
+
+%%% Local Variables:
+%%% mode: context
+%%% TeX-master: t
+%%% End:

### 2023-09-02 marma/marma.tex
new routine for jyotsna

--- 2022-12-16 marma/Marmasthāna.tex
@@
     \item[-vadhnastho] P16, 
-    \item[(unavailable/illegible)]{Ba5, C7, J4, N14, C5, G11, R2, T, Ba3
+    \item[(unavailable/illegible)] Ba5, C7, J4, N14, C5, G11, R2, T, Ba3
+    \end{description}
+ \end{marma}
+
+\begin{marma}[hp01_025]
+
+  \sthana{25c}
+
+  \begin{description}
+\item[dhanurākarṣaṇaṃ kṛtvā]
+\item[dhanusaṃkarṣaṇaṃ kuryād] Bo2,
+\item[dhanuḥsaṃkarṣaṇaṃ kuryād] N9,
+\item[dhanurākarkhaṇa kṛtvā] V19,
+\item[dhanurākarṣakākuṣṭaṃ] M3,
+\item[dhanurākarṣaṇā[kṛ]ṣṇaṃ] P7,
+\item[dhanurākarṣaṇaḥ kṛṣṭaṃ] V1,
+\item[dhanurākarṣaṇākarṣaṃ] Ko,
+\item[dhanurākarṣaṇākṛṣṭaṃ] Ba, C6, C8, G5, G11, J5, J11, M1, N12, N20, P15, V15, V2, V25, 
+\item[dhanurākarṣeṇākṛṣṭaṃ] J2,
+\item[dhanurākarṣaṇaṃ kṛtvā] B2, C1, C4, J1, J3, J6, J10, J14, J15, K1, L1, MYO20, N5, N11, N16, N18, N19, N22, N24, N25, P3, P6, P13, P23, Pen,
+\item[dhanurākarṣana kṛtvā] N21,
+\item[dhanurākarṣaṇaṃ kṛṣṭa] J4,
+\item[dhanurākarṣaṇaṃ kṛṣṭaṃ] C6,
+\item[dhanurākarṣaṇaṃ kṛuṣṭaṃ] G4,
+\item[dhanurākarṣaṇaṃ kuryād] A1, B1, B3, Ba5, Bo1, Bo3, C2, C3, C9, J7, J10, J12, J13, J16, J17, Jyo, LD1, LD2, N13, N17, P1, P2, P4, P5, P7, P9, P12, P17, V16, V18, 
+\item[dhanurākarṣaṇāt kaṣṭaṃ] J8, P11, V3, 
+\item[dhanurākarṣaṇāt kṛṣṭaṃ] L2, N2, N23, P11, V14, V17, 
+\item[jānūrākarṣaṇa kṛuṣiḥ] Ba4, 
+\item[(unavailable/illegible)] A1, C5, C7, K2, K3, N3, N14, V9, P10, 
 \end{description}
@@
   \begin{description}
-\item[jaṭharapradīptaṃ]    V6 (jaraṭha), B1, B2, B3, Bo1, C1, J13, N1, V22, N13, N21, P2, P12, P13, P14, P17, R3, R4, 
+\item[jaṭharapradīptaṃ]    V6 (jaraṭha), B1, B2, B3, Bo1, J13, N1, V22, N13, N21, P2, P12, P13, P14, P17, R3, R4, 
 \item[jaṭharapradīpti]    Ba5, V10
@@
 \item[āsyahṛdaye kṛtvā]    N22, V17
-\item[asya hṛdayeḥ dhṛtvā]    C9 ("āsyaṃ hṛdaye yasya saḥ" is given above)
+\item[asya hṛdayeḥ dhṛtvā]    C9,
 \item[meḍhre pādam atho hṛdi svacibukaṃ]    N7
@@
 \item[(unavailable/illegible)]    Ba, C5, V13, N6, N14, P22, R2, T, Ba3
+    \end{description}
+ \end{marma}
+
+\begin{marma}[hp01_037]
+ \sthana{Before 1.37}
+   \begin{description}
+\item[pūrvoktam eva matsyāsanaṃ] N12,
+\item[pūrvoktam eva saṃmataṃ] P11,
+\item[pūrvoktam eva matsaṃmataṃ] J14, Ko,
+\item[pūrvoktam eva matsāsanaṃ] N24,
+\item[pūrvoktam evāsanaṃ mataṃ] J2,
+\item[pūrvoktam eva matsyendramataṃ] B2, Ba5, C6, J5, J11, N23,
+\item[dvāvektimeva matseṃdramataṃ] C8, 
+\item[(unavailable/illegible)]  A1, B1, B3, Ba, Bo1, Bo2, Bo2, C2, C3, C4, C5, C9, J1, J3, J4, J6, J7, J8, J12, J13, J15, J16, J17, K1, K2, LD1, LD2, N13, N14, N16, N17, N18, N20, N21, N22, N25, P1, P2, P3, P4, P5, P6, P8, P9, P10, P11, P12, P13,
 \end{description}
@@
    \item[yameṣu ca mitāhārm] C6,
-   \item[yamaṣviva mitāharm] J1
+   \item[yamaṣviva mitāharm] J1,
    \item[niyameṣu mitāhari] J2
@@
 \item[suṣumṇām eva nāḍiṣu]    J10ac, J17, N24, V14
-\item[suṣumṇām iva nāḍiṣu ]    A1, C1, C4, C6, C7, C8, C9, J1, J2, J3, J6, J11, J15, L1, M1, V19, N5, N7, N12, N16, N23, N25, P11, P16, P20, P21, P23, P25, P27, V15, V13
+\item[suṣumṇām iva nāḍiṣu ]    A1, C1, C4, C6, C7, C8, J1, J2, J3, J6, J11, J15, L1, M1, V19, N5, N7, N12, N16, N23, N25, P11, P16, P20, P21, P23, P25, P27, V15, V13
 \item[suṣuṇmām iva nāḍiṣu]    P7
@@
 \item[śramadairdhyādibhiḥ pīṭhaiḥ]    V28
-\item[śamardda bahubhiḥ pīṭhaiḥ]    J1
+\item[śamardda bahubhiḥ pīṭhaiḥ]    J1,
 \item[sanastai bahubhyaḥ pīṭai]    V8
@@
 \item[marutaṃ dhārayed yas tu]    C4, N5, V15
-\item[mārutaṃ māryatd yas tu]    J5
+\item[mārutaṃ mārayed yas tu]    J5
 \item[mārutaṃ mriyate yas tu]    N3
@@
  \begin{description}
- \item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, P12, P17, P20, P21, P23, P26, V6, V19, R3, N14, N16, N18, N1,
+ \item[Yes] B1, B3, Ba, Bo1, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, P12, P17, P20, P21, P23, P26, V6, V19, R3, N14, N16, N18, N1,
 
@@
 \item[mudgādi cājyam udakaṃ ca]    N16
-\item[mudgādi cālpa sudakaṃ ca]     J1
+\item[mudgādi cālpa sudakaṃ ca]     J1,
 \item[mudgādi dālpam udakaṃ ca]    J3, J6("cālpa" is also given)] J13,
@@
     • R5 has only one folio of chapter 2
+    • C5 has not 1st and 2nd chapters
+    • U1 has only chapter 2
 \begin{marma}[hp02_003]
@@
 \item[jīvaṃnam] N14
-\item[jīvitam] A1, B1, B2, B3, Bo1, C2, C3, C4, C8, J4, J5, J7, J8, J10, J11, LD1, LD2, N1, N2, N5, V8, N20, N22, P2, P6, P9, P17, P21, P23, V14, V3, N24, N25, P12, P15, R1, R3, V15, V5, V22, V2, V12, V13, V9, G11, Ba3, G1,
+\item[jīvitam] A1, B1, B2, B3, Bo1, C2, C3, C4, C8, J4, J5, J7, J8, J10, J11, LD1, LD2, N1, N2, N5, V8, N20, N22, P2, P6, P9, P17, P21, P23, V14, V3, N24, N25, P12, P15, R1, R3, V15, V5, V22, V2, V12, V13, V9, G11, Ba3, G1, U1,
 \item[jīvatam] P5, P24,
@@
 \item[nirundhayet] A1, B1, B2, B3, C1, C3, C9, J5, J10, J13, J15, J17, LD1, LD2, N17, P1, V19, W4, N2, N7, N9, N10, N14, P5, P8, P9, P10, P23, V16, V17, N20, V21, N23, N26, P2, P7, P26, V20, V5, V2, V18, V12, V9, Ba3
-\item[nirodhayet] Ba5, Bo1, Bo3, C2, C4, C8, J1, J3, J6pc, J7, V19, N1, N4a, N5, N8, R3, V8, N11, N13, V10, N18, N21, P13, P14, P17, P18, P20, Tue, N24, N25, P12, P21, R5, V6, V7, V15, V28, V22
+\item[nirodhayet] Ba5, Bo1, Bo3, C2, C4, C8, J1, J3, J6pc, J7, V19, N1, N4a, N5, N8, R3, V8, N11, N13, V10, N18, N21, P13, P14, P17, P18, P20, Tue, N24, N25, P12, P21, R5, V6, V7, V15, V28, V22, U2,
 \item[nirudhayet] P24,
@@
 \item[kāryasiddhiṃ] J15, 
-\item[kāryāsiddhiḥ] J1
+\item[kāryāsiddhiḥ] J1,
 \item[kāryasiddhi] J12, LD2, N7, V8, N10, N14("rya" unclear), N16, N18, N22, V21, V3 
@@
 \item[suṣumnāṃtarālaschā] P12, 
-\item[suṣumnāṃtarālasyā] J1
+\item[suṣumnāṃtarālasyā] J1,
 \item[suṣumnāṃtarājasthā] N16
@@
 \item[Yes] B1, B3, C, V19, N1, N5, N10 , N12, N21, N25, P12, P13, P26, P17, P20, P21, V28
-\item[No] A1, B2, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C4, C5, C6, C8, J1, J2, J3, J10, J13, J17, J4, J5, J6, J7, J11, J12, J14, J15, J16, Ko, LD1, LD2, N17, P1, V1, J8, W4, N2, N3, N4a, N6, N7, N8, N9, V8, N11, N13, N14, N16, P5, P6, P8, P15, P18, V16, V17, V10, N18, N19, N20, N22, V14, V21, V3, Tue, N23, N24, N26, P2, P7, P9, P10, P11, P14, P16, P22, P23, P24, P25, P27, R1, R5, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9, G11, T, Ba3, G1
+\item[No] A1, B2, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C4, C5, C6, C8, C9, J1, J2, J3, J10, J13, J17, J4, J5, J6, J7, J11, J12, J14, J15, J16, Ko, LD1, LD2, N17, P1, V1, J8, W4, N2, N3, N4a, N6, N7, N8, N9, V8, N11, N13, N14, N16, P5, P6, P8, P15, P18, V16, V17, V10, N18, N19, N20, N22, V14, V21, V3, Tue, N23, N24, N26, P2, P7, P9, P10, P11, P14, P16, P22, P23, P24, P25, P27, R1, R5, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9, G11, T, Ba3, G1
     \end{description}
@@
 \item[(illegible/unavailable)]   B1, B2, C5, N6, V16, V15, V13, V9, T, P16, P22, P24, R5,
+
 \begin{description}
@@
 \item[graset pratyāharec caitad] P25, 
-\item[tataḥ pratyāharec caitad] A1, Ba, Ba3, B3, G11, J12, J16, J17, N9, N12, N14, N17, N20, N21, N23, P1, P5, P8, P9, P10, P11, P12, P13, P15, P24, P26, P27, R1,
+\item[tataḥ pratyāharec caitad] A1, Ba, Ba3, Bo1, B3, C8, G11, J12, J16, J17, N9, N12, N14, N17, N20, N21, N23, P1, P5, P8, P9, P10, P11, P12, P13, P15, P24, P26, P27, R1,
+\item[tatao pratyāharec caitad] C9,
+\item[tataḥ pratyāhare caitad] C6,
 \item[tataḥ pratyaharec cetad] N10
@@
 \item[pūnaḥ pratyāharec caityāt] P6,
-\item[punaḥ pratyahared etad] N8, N11, N16, N18, N25, N26, P20, P21, P23, 
+\item[punaḥ pratyahared etad] C1, C2, C3, C4, J1, N8, N11, N16, N18, N25, N26, P20, P21, P23, 
 \item[punaḥ pratyaharec caitaḥ] J15,
-\item[(unavailable/illegible)] N6, N7, T, G1, P16, P22, R5,
+\item[(unavailable/illegible)] C5, J1, N6, N7, T, G1, P16, P22, R5,
   \begin{description}
@@
 \item[up(?)itaṃ dhautikarma tat] C3
-\item[abhyāsād dhautikarmavit] B1, B2, C4, J1, J3, J6, J13, P2, P12, P17, P20, P26, V5, V28, V2
+\item[abhyāsād dhautikarmavit] B1, B2, C2, C4, J1, J3, J6, J13, P2, P12, P17, P20, P26, V5, V28, V2
 \item[abhyāsād dhautikarmasu] N25
@@
 \item[abhyāsād dhautikarmavat] N5
-\item[abhyāsaud dhautikarmavit] Bo2, C2,
+\item[abhyāsaud dhautikarmavit] Bo2,
 \item[abhyāsāsau dhautikarmavit] V6
@@
 \item[pavanam apānam udārya kaṇṭhanālo] J16,
-\item[pavanām apānam udīrya kaṇṭhanālṃ] J1
+\item[pavanām apānam udīrya kaṇṭhanālṃ] J1,
 \item[pavana apānam udīrya kaṇṭhanāle] V6
@@
 \item[pavanam apānam udīrya kaṇṭhaḍālai] N14, 
-\item[(illegible/unavailable)] Bo3, C5, J2, N6, N8, N11, R5, V19, V10, V14, V21, Tue, V7, V5, V22, G11, T, P16, P22, P23, 
+\item[(illegible/unavailable)] Bo3, C1, C5, J2, N6, N8, N11, R5, V19, V10, V14, V21, Tue, V7, V5, V22, G11, T, P16, P22, P23, 
 \end{description}
@@
 V18, V12
+\item[aśeṣadoṣopacayaṃti hanyād] J1,
 \item[aśeṣadopacayaṃ nihanyād] P23,
@@
 \sthana{2.30}
-\item[atha natīḥ] J1
+\item[atha natīḥ] J1,
 \item[atha ti] P26, 
@@
 \item[mukhanirgatamanād eva] V14
-\item[mukhanirgamayedeti] J2
+\item[mukhanirgamayedeti] J2,
 \item[mukhan nirgamanād eva] B3,
@@
 \item[tadbhānaṃ ca kapāṭakam] P5, P24, 
-\item[tandrādīnāṃ kapālakam] J1
+\item[tandrādīnāṃ kapālakam] J1,
 item[tandrāegaṃ kathapāṭakam] V21
@@
 \item[pācanādīn P26, 
-\item[pācanāgni] J3, J6, J13, N8, N11, N16, P16, P23, 
+\item[pācanāgni] C1, C4, J1, J3, J6, J13, N8, N11, N16, P16, P23, 
 \item[pācanāgnī] P21,
@@
 \item[doṣān mala śoṣaṇī ca] J12, N10, 
-\item[doṣāmaya śoṣaṇīya] J1
+\item[doṣāmaya śoṣaṇīya] J1,
 \item[doṣāmaya śoṣaṇīyaṃ] J2
@@
 \item[maulir iyaṃ hi] A1, C1, C4, J3, J5, N8, N11, N16, P6, P20, P23, P27, V19, V21, N25, V15, V2
-\item[maulir iyaṃ] J1
+\item[maulir iyaṃ] J1,
 \item[maulir iyaṃ ca] Ba, Ba5, Bo1,Bo2, C6, C8, J6, J7, N12, N13, N21, Tue, N23, P7, P11, P13, P14, P16, P18, P21, P25, V6, V7, V20, V28, V22, G11
@@
 \sthana{2.39ab}
-\item[brahmādayo 'pi tridaśāḥ pavanābhyāsatatparāḥ] A1, B2, B3, Ba, G1, P11, P12, P13, P14, P15, P16, P18, P20, P27, P21, P23, P25, P26,
+\item[brahmādayo 'pi tridaśāḥ pavanābhyāsatatparāḥ] A1, B2, B3, Ba, Bo1, C1, C4, C6, C8, C9, G1, J1, P11, P12, P13, P14, P15, P16, P18, P20, P27, P21, P23, P25, P26,
 \item[brahmādayo 'pi tṛdaśāḥ pavanābhyāsatatparāḥ] P17,
-\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparāḥ] B1
-\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparaḥ] R1,
-\item[(illegible/unavailable)] P22, P24, R5, T
+\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparāḥ] B1, C2, 
+\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparaḥ] C3, R1,
+\item[(illegible/unavailable)] C5, P22, P24, R5, T
   \begin{description}
@@
 \item[abhūvannantakabhayāt tasmat pavanam abhyaset] G1,
-\item[abhūvannantakabhayāt tasmāt pavanam abhyaset] A1, Ba, P11, P13, P14, P15, P18, P20a, P27,
+\item[abhūvannantakabhayāt tasmāt pavanam abhyaset] A1, Ba, Bo1,C1, C6, C8, P11, P13, P14, P15, P18, P20a, P27,
 \item[abhūvannātankabhayāt tasmāt pavanam abhyaset] B1,
@@
 \item[tena siddhi gatās te ca tasmat pavanam abhyaset] P21, P23, 
-\item[tena siddhiṃ gatās te ca tasmat pavanam abhyaset] B2, 
+\item[tena siddhiṃ gatā yasmāt tasmāt pavanam abhyaset] C4,
+\item[tena siddhiṃ gatās te ca tasmat pavanam abhyaset] B2, C9,  J1, 
 \item[tena siddhi gatāḥ satyaḥ tasmat pavanam abhyaset] P20,
@@
 \item[sumanskāṃtako bhavyaḥ tasmat pavanam abhyaset] R1,
-\item[(illegible/unavailable)] P22, P24, R5, T
+\item[samanskāṃtako bhavyaḥ tasmāt pavanam abhyaset] C2,
+\item[samanskoraho bhavyaḥ tasmāt pavanam abhyaset] C3,
+\item[(illegible/unavailable)] C5, P22, P24, R5, T
   \begin{description}
@@
 \item[vidhaustajñā] V20
-\item[vidhānujñaś] J1
+\item[vidhānujñaś] J1,
 \item[vidhānataś] P23, 
-\item[vidhānajñāś] Bo2, C1, C3, C8, J4, J6, J7, J12, P5, P13, P14, P15, P24, P25, P27, 
-\item[vidhānajñaś] B1, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C8, J3, J16, N3, N24, P2, P11, P12, P17, P18, P21, 
+\item[vidhānajñāś] Bo2, C3, C8, J4, J6, J7, J12, P5, P13, P14, P15, P24, P25, P27, 
+\item[vidhānajñaś] B1, B3, Ba, Ba5, Bo1, Bo2, C1, C2, J3, J16, N3, N24, P2, P11, P12, P17, P18, P21, 
 \item[vidhānajña] N22, V12
@@
 \item[bhastrikā bhramarī mūrcchā cāṣṭakumbhakāḥ sahita?] N2 (after exchange of words' positions)
+\item[bhastrikā bhramarī mūrdhā kevalaś cāṣṭakumbhakāḥ] C2,
 \item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakāḥ] A1, B1, C3, J7pc, J10, J12, J16, J17, N3, N7, N9, N10, N17, P1, P2, P6, P9, P10, P17, P20, P27, R1, 
@@
 \item[bhastrikā bhramarī mūrchā saṃprataṃ cāṣṭakumbhakān] N18
-\item[bhadrikā? bhramarī mūrchā saṃprataṃ rāṣṭakumbhakāṃ] J1
+\item[bhadrikā? bhramarī mūrchā saṃprataṃ rāṣṭakumbhakāṃ] J1,
 \item[bhastrikā śītalī bhramarī mūrchā plāvanītyaṣtakumbhakāḥ] N19 
@@
 \item[nīrodhāvadhi kambhayet] Ba,
-\item[nīrodhāvadhi kumbhayet] A1, B1, B3, J15, J16, J17, N4a, N10, N13, P1, P2, P5, P8, P9, P10, P11,  P12, P14, P17, P18, P24, P25, P27, Tue, V21, G11, Ba3
+\item[nirodhābadhi kumbhayet] C2,
+\item[nirodhāvadhi kumbhayet] A1, B1, B3, C3, C8, C9, J15, J16, J17, N4a, N10, N13, P1, P2, P5, P8, P9, P10, P11,  P12, P14, P17, P18, P24, P25, P27, Tue, V21, G11, Ba3
+\item[nirodhāccādhi kumbhayet] C6,
+\item[nirodhāvadhi kumbhakaḥ] Bo1, 
 \item[nīrodhāvadhi kumbhata] R1,
@@
 \item[nīrodhāvadhi kumbhavat] N21, 
-\item[nīrodhāvadhi kumbhakaṃ] N2, N11,  
+\item[nīrodhāvadhi kumbhakaṃ] J1, N2, N11,  
 \item[nīrodhāvadhi kumbhake] N18, 
@@
 \item[nīrodhāvadhikam bhavet] P26,
-\item[śirodhāvadhi kumbhakaṃ] N5, N8, N16, P21, P23,  
+\item[śirodhāvadhi kumbhakaṃ] C1, C4, N5, N8, N16, P21, P23,  
 \item[śirodhāvati kumbhakān] P6,
@@
 \item[sirodhāvadhi kumbhakāt] N22,
-\item[(illegible/unavailable)] J14, N6, N14, T, P16, P22, R5,
+\item[(illegible/unavailable)] C5, J14, N6, N14, T, P16, P22, R5,
 
@@
 \sthana{2.53a}
-\item[jālodare]
+\item[jālodare] C3, C9,  
 \item[jālodarerā] P24, 
 \item[jalodar] A1,
-\item[jalodarā] P13, P14, P15, P18, P20, P21, P23, P27, 
+\item[jalodara] C6,
+\item[jalodarā] C1, C8, P13, P14, P15, P18, P20, P21, P23, P27, 
 \item[jalodaro] B2,
 \item[jalodāraṃ] P17,
-\item[jalodaraṃ] B1, B3, Ba, P25, P26, 
+\item[jalodaraṃ] B1, B3, Ba, Bo1, C2, C4, J1, P25, P26, 
 \item[jālodared] R1,
-\item[(illegible/unavailable)] P16, P22, R5,
+\item[(illegible/unavailable)] C5, P16, P22, R5,
 
@@
 \item[śītkāṃ dadyāt sadā vaktre ghraṇe naiva vijṛmbhikām] P14, P18, 
+\item[śītkāṃ dadyāt sadā vaktre ghraṇe caiva vijṛmbhikām] C6, 
+\item[chikkāṃ dadyāt sadā valke ghraṇe naiva vijṛmbhikām] Bo1, 
 \item[sītkāṃ dadyāt sadā vaktre ghraṇe caiva vijṛmbhikām] P27,
@@
 \item[śītkāṃ kuryāt sadā vaktrāt ghraṇe naiva vijṛmbhakān] P26,
+\item[sītkāṃ kuryāt tathā vaktre ghraṇe naiva vijṛmbhikām] C1,
+\item[sītkāṃ kuryā tathā vaktre ghraṇe naiva vivarjayet] C4,
 \item[sītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikām] P25,  
 \item[śītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikām] Ba,  
+\item[śītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikā] C8,  
 \item[śītkāṃ kuryāt yathā vaktre prāṇaṃ caiva visarjjayet] P20, 
@@
 \item[śitkāṃ kuryāt tathā vaktre ghrāṇenaiva visarjayet] P23, 
-\item[kuṃbhaṃ kuryāt sadā vaktre grāṇenaiva visarjayet] B1, B3, 
-\item[kuṃbhaṃ kuryāt sadā vaktrat ghrāṇenaiva vijaṃbhitaṃ] R1, 
-\item[(illegible/unavailable)] P16, P22, P24, R5,
+\item[sitkāṃ kuryāt tayā vaktre ….nenaiva visarjayet] J1,
+\item[kuṃbhaṃ kuryāt sadā vaktre grāṇenaiva visarjayet] B1, B3, C2,  
+\item[kuṃbhaṃ kuryāt sadā vaktrāt grāṇe caiva vijṛmbhikām] C3,
+\item[kuṃbhaṃ kuryāt sadā vaktrat ghrāṇenaiva vijaṃbhitaṃ] R1, C9,  
+\item[(illegible/unavailable)] C5, P16, P22, P24, R5,
 
@@
 \item[sītalī] J2
-\item[sikta?] J1
+\item[sikta?] J1,
 \item[sītkīṃ] V19,
@@
 \item[]
-\item[(illegible/unavailable)] A1, B2, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C5, C6, C8, J1, J2, J4, J5, J7, J11, J12, J14, J16, Ko, LD1, LD2, N1, N2, N3, N4a, N7, N8, N10, N11, N12, N13, N16, N18, N19, N21, N22, N23, N24, P6, P7, P8, P9, P11, P13, P14, P15, P18, P20, P21, P22, P23, P25, P26, P27, P28, R1, R5, W4, V3, V6, V20, V15, V5, V28, V22, V2, V12, V13, V10, V14, V17, G11, Tue, T, Ba3, G1, P16,
+\item[(illegible/unavailable)] A1, B2, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C5, C6, C8, J1, J2, J4, J5, J7, J11, J12, J14, J16, Ko, LD1, LD2, N1, N2, N3, N4a, N7, N8, N10, N11, N12, N13, N16, N18, N19, N21, N22, N23, N24, P6, P7, P8, P9, P11, P13, P14, P15, P18, P20, P21, P22, P23, P25, P26, P27, P28, R1, R5, W4, V3, V6, V20, V15, V5, V28, V22, V2, V12, V13, V10, V14, V17, G11, Tue, T, Ba3, G1, P16,
   \begin{description}
@@
 \item[]
-\item[(illegible/unavailable)] Bo3, C5, J2, J14, N4a, Tue, LD2, N13, N24, P14, P17, P18, P22, V6, V5, V22, V10, T, P16, R5,
+\item[(illegible/unavailable)] Bo3, C1, C5, J2, J14, N4a, Tue, LD2, N13, N24, P14, P17, P18, P22, V6, V5, V22, V10, T, P16, R5,
   \begin{description}
@@
 \item[dhārayan nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] C2, 
-\item[dhārayen nāsikāmadhya tarjjanībhyāṃ vinā dṛḍhām] J1
+\item[dhārayen nāsikāmadhya tarjjanībhyāṃ vinā dṛḍhām] J1,
 \item[dhārayen nāsikā madhye tarjanībhyāṃ vinā madam] J4,
@@
 \sthana{2.67ab}
-\item[kaṇdalībodhakaṃ viprabhavaghnaṃ sukhadaṃ śubhaṃ] V1
+\item[kaṇdalībodhakaṃ viprabhavaghnaṃ sukhadaṃ śubhaṃ] V1,
 \item[kuṇḍalī bodhakaḥ kubho rogaghnaḥ sukhada śubhaḥ] Ba3
@@
 \item[kaṇdalībodhakaṃ ksipraṃ pavanaṃ sukhadaṃ hitaṃ] Tue, V7
-\item[kuṇḍalo bodhaka kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] J2
+\item[kuṇḍalo bodhaka kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] J2,
 \item[kuṇḍalībodhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] Ba5, N13, P14, P18, V22
@@
 \item[kaṇḍalībodhanaṃ cakaṃ pāpaghnaṃ sukhada śubhaṃ] V14
-\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] B2, Bo2, C4, J3, J6, J7, J11, N16, N19, N23, N25, P6, P21, P23, P27, 
+\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] B2, Bo2, C4, J1, J3, J6, J7, J11, N16, N19, N23, N25, P6, P21, P23, P27, 
 \item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ śubhadaṃ sukhaṃ] P20,
@@
 \item[kuṇḍalībodhakaḥ kumbho rogaghnaṃ sukhadaṃ śubhaṃ] N1, 
-\item[kuṇḍalīno bodhacakraṃ? bhāvadaṃ supada śubhāṃ] J5
+\item[kuṇḍalīno bodhacakraṃ? bhāvadaṃ supada śubhāṃ] J5,
 \item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, V10, T, G1, P16, P22, R1, R5,
@@
 \item[samyag bhastrī samudbhūto] N6, 
-\item[sasyagātrasamudbhūtaṃ] J1
+\item[sasyagātrasamudbhūtaṃ] J1,
 \item[samagātrasamudbhūtaṃ] V6, V16
@@
 \item[vegā gho] N23,
-\item[vego tad ghoṣa] J1
-\item[vego ghoṣam] C3, C9, J16, N3, N9, N16, N20, P1, P5, P10, P24, P27, 
+\item[vego tad ghoṣa] J1,
+\item[vego ghoṣam] C3, C8, J16, N3, N9, N16, N20, P1, P5, P10, P24, P27, 
 \item[vego ghoṣa] N24,
@@
 \item[ārecyāpūrya yat kuryāt] N2, N5, 
+\item[ārecya pūryaṃ yat kuryāt] C4,
 \item[ārecya āpūrya yat kuryāt] P23,
@@
 \item[ārecyāpūrvavat kuryāt] N25,
+\item[caraikaṃ? pūrakaṃ kuryāt] C2,
 \item[recakaṃ pūrakaṃ kuryāt] N1, N10, P2, 
@@
 \item[recake pūrakaṃ kuryāt] J12, J16, 
-\item[recayet pūrakaṃ kuryāt] N16, N19, N23, P6, 
+\item[recayet pūrakaṃ kuryāt] J1, N16, N19, N23, P6, 
 \item[recakaṃ pūrakaṃ kuryāt] P12, P17, 
@@
 \item[recakaḥ pūrakaṣ kāryyaḥ] J17,
-\item[recakaḥ pūrakaḥ kāryaḥ] A1, N9, J15, N17, N26, P1, P5, P9, P24, R1, 
+\item[recakaḥ pūrakaḥ kāryaḥ] A1, C3, C9, N9, J15, N17, N26, P1, P5, P9, P24, R1, 
 \item[recakaḥ pūrakaḥ kuryāt] P21, 
@@
 \item[recakaḥ pūrakaḥ kā....] P10,
-\item[recakaḥ pūrakaḥ kāryāḥ] N7, 
+\item[recakaḥ pūrakaḥ kāryāḥ] Bo1, N7, 
 \item[recakaḥ pūrakaḥ kāryaṃ] N20
@@
 \item[recaka pūraka varjite] V10
-\item[virecya pūrakaḥ kāryaḥ] P7,
+\item[virecya pūrakaḥ kāryaḥ] C6, C8, P7,
 \item[prāṇāyāmoyam ity uktaḥ] N4a, N12, P14, P18, Tue,
-\item[(illegible/unavailable)] B2, N6, N13, N14, P15, T, P16, P22, P26, P27, R5,
+\item[(illegible/unavailable)] B2, C1, C5, N6, N13, N14, P15, T, P16, P22, P26, P27, R5,
   \begin{description}
@@
 \item[sarve kevalakumbhakam] G1
-\item[śanaiḥ sahitakumbhakaḥ] N23, P7, 
+\item[śanaiḥ sahitakumbhakaḥ] Bo1, C6, C8, J1, N23, P7, 
 \item[śanaiḥ sahitakumbhakaiḥ] P11,
@@
 \item[sa vai kevalakumbhakaḥ] N4a, N18, N22, P18, P26, Tue, 
-\item[sa vai sahitakumbhakaḥ] A1, B1, B2, B3, J12, J13, J15, N1, N3, N5, N10, N17, N21, P2, P8, P12, P13, P17, P20, P21, P23, P27, 
+\item[sa vai sahitakumbhakaḥ] A1, B1, B2, B3, C2, C4, J12, J13, J15, N1, N3, N5, N10, N17, N21, P2, P8, P12, P13, P17, P20, P21, P23, P27, 
 \item[sarve sahitakumbhakaḥ] J17, N20, 
@@
 \item[sarvai sahitakumbhakaḥ] N9, N12, N26, P1, P9, P10, R1, 
+\item[sarvaiḥ sahitakumbhakaḥ] C3, C9, 
 \item[sarvaiḥ sahitakumbhakam] Ba, P25,
@@
 \item[kāryaḥ śarveiḥ sahitakumbhaka] V17
-\item[(illegible/unavailable)] N6, N13, P15, T, P16, P22, R5,
+\item[(illegible/unavailable)] C1, C5, N6, N13, P15, T, P16, P22, R5,
   \begin{description}
@@
 \begin{marma}[hp02_78a]
-\sthana{kumbhakaḥ prāṇarecānte} C3, J2, J3, J13, N8, N11, N16, N21, P5, P24,V1, V6, V28, V18, V16, G11,
-\sthana{kumbhakāt prāṇarecānte} N20, N26, V17
-\item[kumbhaka prāṇarecānte] A1, B1, B2, Bo2, C1, J1, J5, LD1, N2, N7, N9, N10, N12, N14, N17, N22, N25, P1, P2, P9, P10, P12, P13, P17, P20, P21, P23, V2, V12, Ba3
+\sthana{2.78}
+\item[kumhakaḥ prāṇarecānte] C3, J2, J3, J13, N8, N11, N16, N21, P5, P24,V1, V6, V28, V18, V16, G11,
+\item[kumkāt prāṇarecānte] N20, N26, V17
+\item[kumbhaka prāṇarecānte] A1, B1, B2, Bo2, C2, J1, J5, LD1, N2, N7, N9, N10, N12, N14, N17, N22, N25, P1, P2, P9, P10, P12, P13, P17, P20, P21, P23, V2, V12, Ba3
 \item[kumbhaka prāṇarecyaṃte] N1,
@@
 %%% End:
+
+    • Verses are found till 42ab in Bo1 ms
+    • Chapter 3 is not available in Bo2; Bo3; C5; P2;
+    • Many verses of chapter 1 & 4 are given in chapter 3 of J4 i.e. khecarī etc...
+    • K2 has only till sthāna 65d of chapter 3
+    • N4a has only till 76b of chapter 3
+    • N4b starts from the sthana 56ab
+    • N20 has only till sthāna 3.29
+    • O1, O2, O3 have only few folios of chapter 1
+\begin{marma}[hp03_010]
+
+\sthana{3.10cd}
+
+\item[pādaṃ prasāritaṃ dhṛtvā karābhyāṃ pūrayen mukhaṃ] C4, C8, J1, J3, J6, J7, J14, K1, N5, N8, N11, N25, P6, 
+\item[padaṃ prasāritaṃ dhṛtvā karābhyaṃ pūrayen mukhe] N12,
+\item[pādaṃ prasāritaṃ kṛtvā karābhyāṃ pūrayen mukham] C6, C9, J10a, J11,
+\item[pādaṃ prasāritaṃ dhṛtvā karābhyāṃ dhārayan mukhaṃ] N17,
+\item[pādaṃ prasāritaṃ dhṛtvā karābhyāṃ dhārayen mukham] C3, J10, N20, P9, 
+\item[pādaprasaritaṃ dhṛtvā karābhyāṃ dhārayen mukham] P10,
+\item[pādaprasāritaṃ dhṛtvā karābhyāṃ dhārayen mukham] J17, P1, P8, 
+\item[pādaṃ prasāritaṃ dhṛtvāḥ karābhyāṃ dhārayen mukham] J15, N9, 
+\item[pādaṃ prasāritaṃ kṛtvāḥ karābhyāṃ dhārayen mukham] P5,
+\item[pādaṃ prasāritaṃ dhṛtvā karābhyāṃ dhārayed dṛḍhaṃ] A1, C2, LD1, P2, P4, P12, P13, 
+\item[pādaṃ prasāritaṃ kṛtvā hastābhyāṃ dhāraye dṛḍhaṃ] N24,
+\item[pādaṃ prasāritaṃ dhṛtvā karābhyāṃ dhāraye dṛḍhaṃ] J13,
+\item[pādaṃ prasāritaṃ kṛtvā karābhyāṃ dhāraye dṛḍhaṃ] J12, Ko, 
+\item[pādaṃ prasāritaṃ kṛtvā karābhyāṃ dhārayet dṛḍhaṃ] LD2, N4a, N10,
+\item[pādaṃ prasārita kṛtvā kārābhyāṃ dhāraye dṛḍhe] J16,
+\item[pāda prasāritaṃ dhṛtvā karābhyāṃ dhāraye dṛḍhaṃ] B3,
+\item[pādaṃ prasāritaṃ dhṛtvā karābhyāṃ dhārayet sukhaṃ] N6, N7, 
+\item[pādaṃ praśāritaṃ dhṛtvā kalābhyāṃ dhārayet sukhaṃ] N14,
+\item[pādaṃ prasāritaṃ dhṛtvā karābhyāṃ purayed dṛḍhaṃ] B1, 
+\item[pādaḥ prasaritaṃ dhṛtvā karābhyāṃ dhārayen mukhaṃ] N15, N16,
+\item[pādaṃ prasaritaṃ dhṛtvā karābhyāṃ dhārayen mukhe] B2,
+\item[pādaṃ prasāritaṃ kṛtvā karābhyāṃ dhārayed dhiyā] Bo1,  
+\item[padaṃ prasāritaṃ kṛtvā karābhyaṃ pūrayen mukhe] J4, J5, J8, N3, P11, 
+\item[pādaṃ prasāritaṃ kṛtvā karābhyaṃ pūrayen mukhaṃ] N22,
+\item[pādaṃ prasāritaṃ dhṛtvā karābhyaṃ pūrayan mukhaṃ] N18,
+\item[pada prasāritaṃ dhṛtvā karābhyaṃ pūrayen mukhaṃ] N23, P7, 
+\item[padaṃ prasāritaṃ dhṛtvā karābhyaṃ pūrayen sukhaṃ] K2,
+\item[prasāritaṃ padaṃ kṛtvā karābhyaṃ pūrayen mukhe] Ba,
+\item[prasāritaṃ paraṃ kṛtvā karābhyaṃ pūrayen mukhe] J2,
+\item[prasāritaṃ padaṃ kṛtvā karābhyaṃ dhāraye dṛḍhaṃ] N21,
+\item[prasāritaṃ padaṃ kṛtvā karābhyaṃ dhārayed dṛḍhaṃ] Ba5, N13, P3,
+
+\item[(illegible/unavailable)] Bo2, Bo3, C5, N4b, 
+  \begin{description}
+    \end{description}
+
+  \end{marma}
+
+\begin{marma}[hp03_012]
+
+\sthana{3.12cd}
+
+\item[tadāsau maraṇāvasthā jayate dvipuṭāśritā] P11,
+\item[tadāsau maraṇāvasthā jāyate dvipuṭāśritā] A1, J5, N3, 
+\item[tadāsau maraṇāvasthā jāyate dviputāśritā] N24,
+\item[tadāsau maraṇāvasthā jāyate dviputāśritā] N21,
+\item[tadāsau maraṇāvasthā jāyate dviputāśritāḥ] P13,
+\item[tadāsau maraṇāvasthā jāyate dvipuṭāśrayā] C8,
+\item[tadāsau maraṇāvasthā jāyate dvipuṭā hi sā] C6, P7,
+\item[tadāsau maraṇāvaschā jāyate dvipuṭāśritā] Bo1, 
+\item[tadāsau maraṇāvasthā jāyate tripurāśritā] J4, 
+\item[tadāsau maraṇāvasthā harate dvipuṭamāśritā] B1, 
+\item[tadāsau maraṇāvasthā harate dvipūtāśrayī] N23,
+\item[tadāsau maraṇāvasthā harate dvipuṭaśritā] C2,
+\item[tadāsau maraṇāvasthā harate dvipuṭāśritā] B3, J15, 
+\item[tadāsau maraṇāvasthāṃ harate dvipuṭāśritā] J13, 
+\item[tadāsau maraṇāvasthā jāyate tripuṭāśritaḥ] Ko,
+\item[tadāsau maraṇāvasthā jāyate tripuṭāśritāḥ] N12,
+\item[tadāsau maraṇāvasthā jāyate trimuṭāśratāḥ] J2,
+\item[tadāsau ramaṇāvasthā jāyate dvipuṭāśritā] J8,
+\item[tadeyaṃ maraṇāvasthā jāyate dvipuṭāśritā] J14,
+\item[yathāsau maraṇāvatthāṃ harate dvipaṭāśrayaṃ] J1,
+\item[yathāso maraṇāvasthā jāyate dvipūṭāśrītāḥ] J16,
+\item[yathāsau maraṇāvasthāṃ harate dvipaṭāśrayaṃ] J6,
+\item[yathāsau maraṇāvasthāṃ harate dvipuṭāśritāṃ] P2, P4, 
+\item[yathāsau maraṇāvasthāṃ harate dvipadāśrayaṃ] N16,
+\item[tathāsau maraṇāvasthāṃ harate dvipadāśrayaṃ] J6a,
+\item[tathāsau maraṇāvasthāṃ harate dvipuṭāśrayā] K1, K2, N11, N22, 
+\item[tathāsau maraṇāvasthāṃ harate vapurāśrayā] C4,
+\item[tathāsau maraṇāvasthāṃ harate vapurāśrayāṃ] J7,
+\item[tathāsau maraṇāvasthāṃ harate dvipuṭāśrayāṃ] J3,
+\item[tathāsau maraṇāvasthāṃ haraṃte dvipadāśrayā] N5,
+\item[tathāsau maraṇāvasthāṃ harate dvipaṭāśrayā] N8,
+\item[tathā sā maraṇāvasthā harate dvipuṭāśrayā] N25,
+\item[tadāsau maraṇāvasthāṃ harate dvipuṭāśritā] P12,
+\item[tadāsau maraṇāvaschā harate vapurāśrayi] Ba, 
+\item[tada sā maraṇāvasthā jāyate dvipaṭaśritā] N17,
+\item[tadā sā maraṇāvasthā jāyate dvipaṭaśritā] N7,
+\item[tathā sā maraṇāvasthā haste dvipuṭamāsrayāt] B2,
+\item[tadā sā maraṇāvasthā jāyate dvipuṭāśayā] Ba5,
+\item[tadā sā maraṇāvasthā jāyate dvipuṭāśrayā] N4a, N13, P1, 
+\item[tadā sā maraṇāvasthā jāyate dvipuṭāśritā] C9, J10, J17, LD1, N6, N15, N20, P5, P8, P9, P10, 
+\item[tathā sa maraṇāvasthā jāyate dviputāśṛtā] N14,
+\item[tadā sā maraṇāvasthā jāyate dvipaṭāśritā] N9,
+\item[tadā sā maraṇāvasthāṃ harate dvipuṭāśrayā] C9a,
+\item[yadā sā maraṇā ’vasthā jāyate dvipuṭāsthitā] C3,
+\item[yadā sā maraṇāvasthā jāyate dvipuṭāśritā] J12, N10,
+\item[tadā sā maraṇāvasthā jāyate tripuṭāsanā] J11,
+\item[tadā sā maraṇāvasthā jāyate tripuṭīśritā] LD2,
+\item[yathāsau maraṇāvasthāṃ harate dvipuṭaśritā] N18,
+\item[yathāsau maraṇāvasthā jāyate dvipūṭāśrayā] P6,
+  \item[(illegible/unavailable)] Bo2, Bo3, C5, N4b, P3,
+
+  \begin{description}
+    \end{description}
+
+\end{marma}
+
+\begin{marma}[hp02_014]
+
+\sthana{2.14a}
+
+\item[mahākleśādayo doṣā] A1, Ba, C2, C6, C8, J2, J8, J11, J15, J17, LD2, N3, N14, N16, N20, N22, P6, P7, P8, P10,
+\item[mahān kleśādayo doṣā] J10, P1, 
+\item[mahākleśā.........doṣā] K2,
+\item[mahāklesādayo doṣā] C9,
+\item[mahākleśayato doṣā] N23,
+\item[mahākleśā mahādoṣā] C4, J13, N5, N25, 
+\item[mahākleśādayo doṣāḥ] B1, B2, B3, Ba5, C3, J1, J3, J6, J14, LD1, N4a, N6, N7, N9, N10, N13, N15, N17, N21, P2, P4, P5, P12, P13, 
+\item[mahākleśādaye doṣāḥ] P9,
+\item[mahākleśāyate doṣāḥ] N24,
+\item[māhākleśādayo doṣāḥ] N18,
+\item[mahākleśāyato doṣa] J7, 
+\item[mahākleśāpanā doṣā] Bo1,
+\item[mahākleśā yathā doṣā] Ko,
+\item[mahākleśā yathā dehe] J4, 
+\item[mahākleśāyata doṣā] P11,
+\item[mahatkāśādayo doṣāḥ] N12,
+\item[mahatakuśodayo] J5,
+\item[mahīkleśādayo doṣā] J12, J16, 
+  \item[(illegible/unavailable)] Bo2, Bo3, C5, K1, N4b, N8, N11, P3,
+
+  \begin{description}
+    \end{description}
+
+  \end{marma}
+
+\begin{marma}[hp02_018]
+
+\sthana{2.18b}
+
+\item[mahāsiddhikarī nṛṇām] B1, B3, Ba, Bo1, Ba5, C2, C3, J1, J2, J4, J6, J8, J10, J11, J12, J13, J14, J17, Ko, LD1, LD2, N4a, N6, N7, N9, N10, N12, N13, N14, N15, N16, N17, N21, N24, P1, P2, P4, P5, P6, P8, P10, P12, P13,
+\item[mahāsiddhiḥkarī nṛṇām] N18,
+\item[mahāsiddhikari nṛṇām] J16,
+\item[mahāsidhikarī nṛṇām] C9,
+\item[mahāsiddhikarī nṛṇām] J15,
+\item[mahāsiddhikaraṃ nṛṇām] N22,
+\item[mahāsiddhipradāyakaḥ] J3,
+\item[mahāmudrākarī nṛṇām] P9,
+\item[sarvasiddhikarī nṛṇām] N20,
+\item[jarāmṛtyuvināśinī] A1, C4, C6, C8, J7, K1, K2, N3, N5, N8, N11, N23, P7, 
+\item[jarāmṛtyuvivarjitaṃ] N25,
+\item[nṛṇāṃ mṛtyuvināśinī] J5,
+\item[na deyā yasya kasya cit] B2,
+  \item[(illegible/unavailable)] Bo2, Bo3, C5, N4b, P3, P11,
+
+  \begin{description}
+    \end{description}
+
+\end{marma}
+
+\begin{marma}[hp02_022]
+
+\sthana{2.22cd}
+\item[rājadantadvayaṃ tatra jihvayottaṃbhaved iti] J7, N24, 
+\item[rājadadatadvayaṃ tatra jihvayottaṃbhaved iti] N23, 
+\item[rājadaṃtadvaya tatra jihvayotta bhaved iti] J5,
+\item[rājadantadvayaṃ tatra jihvayotaṃbhayed iti] C8, J14, 
+\item[rājadantavilaṃ tatra jihvayottaṃbhayed iti] A1, C6, J6a,
+\item[rājadantavilaṃ tatra jihvāsstaṃbhayed iti] P7,
+\item[rājadantavilaṃ tatra jihvayottaṃbhayod iti] N16,
+\item[rājadaṃtavilaṃ yatra jihvayottaṃbhayed iti] N20,
+\item[rājadantavilaṃ jatra jihvayottaṃbhayed iti] J8,
+\item[rājadantavilaṃ tatra jihvayottaṃtrayed iti] J6,
+\item[rājadantavilaṃ (yaṃtra) jihvayottaṃ bhaved iti] J1,
+\item[rājadantavalaṃ tadvaj jihvayā staṃbhayed iti] J11,
+\item[rājadantavalaṃ tadvaj jihvā staṃbhayed iti] N25,
+\item[rājadantavilaṃ haṃti jihvāyām staṃbhayoditā] Ko,
+\item[rājadantasyabilaṃ tadvaj juhvayāstaṃbhayed iti] P5,
+\item[rājadantadvayaṃ tatra jihvayā (ruraṃ)bhayed iti] Bo1, 
+\item[rājadaṃtasya jihvāyaṃ baṃdhaṃ ca staṃbhayed iti] Ba, 
+\item[rājadaṃtasthajihvāyaṃ baṃdhaṃ saṃbhedayed iti] N12,
+\item[dājadaṃtasthi jihvāyaṃ baṃdhaṃ ca sthaṃbhayed iti] J2,
+\item[rājadaṃtasya jihvāyaṃ baṃdha śasto bhave dvijaḥ] C2,
+\item[rājadaṃta sajīhvāyā baṃdhaś ca staṃbhayed iti] P6,
+\item[rājadaṃtasya jihvāyaṃ baṃdhaḥ śasto bhaveddhitaḥ] B1, B3, J10, J13, J15, J17, LD1, LD2, N6, N7, N9, N15, N17, N18, P1, P2, P4, P8, P9, P12, 
+\item[rājadaṃtasya jihvāyaṃ baṃdhaḥ śasto bhaved iha] J12,
+\item[rājadaṃtasya jihvāyaṃ baṃdhaḥ śasto bhaved ehi] N10,
+\item[rājadaṃtasya jihvāyoṃ baṃdhaḥ śasto bhaved ihaḥ] J16,
+\item[rājadaṃtasya jihvāyaṃ baṃdhaḥ sasto bhaveddhitaḥ] C9,
+\item[rājadaṃtasya jihvāyaṃ baṃdhaḥ sasto bhaved iti] C9a, N21, P13, 
+\item[rājadaṃtasya jihvāyaṃ baṃdhaḥ śasto bhaved iti] N4a, N13, 
+\item[rājadaṃtasya jihvāyaṃ baṃdhaḥ śasnā bhaveddhitaḥ] C3,
+\item[rājadaṃtasya jihvāyaṃ baṃdhaḥ śasto bhaved iti] Ba5,
+\item[rājadaṃtasya jihvāyaṃ baṃdhaṃ śasto bhaveddhitaḥ] B2,
+\item[rātadaṃtasajihvāyaṃ baṃdha śasto bhaveddhitaḥ] K2,
+\item[śatadaṃtasya jihvāyaṃ baṃdhaḥ śasto bhaveddhitaḥ] C4,
+\item[śatadaṃtasya jihvāyaṃ baṃdhaḥ śasto bhavedihaṃ] N5,
+\item[(illegible/unavailable)] Bo2, Bo3, C5, J3, J4, K1, Ko, N3, N4b, N8, N11, N14, N22, P3, P10, P11,
+
+  \begin{description}
+    • It is given at image no. 19 in J5
+    \end{description}
+
+  \end{marma}
+
+\begin{marma}[hp02_024]
+
+\sthana{2.24d}
+
+\item[kedāraṃ prāpayen manaḥ] A1, B1, B2, B3, Ba5, Bo1, C2, C3, C4, C6, C9, J1, J6, J7, J8, J10, J12, J13, J15, J16, J17, K1, K2, LD1, LD2, N3, N4a, N5, N6, N7, N8, N9, N12, N13, N15, N16, N17, N20, N21, N22, N23, N24, N25, P1, P2, P5, P6, P8, P9, P10, P12, P13,
+\item[kedāraṃ prāpayan manaḥ] N5, N18, P11, 
+\item[kedāraṃ prāppayen manaḥ] Ba, C8, 
+\item[kedāraṃ prāpayen mama] N10,
+\item[kedāraṃ prāpayen mamaḥ] P4,
+\item[kedāraṃ prāpayen manī] P7,
+\item[kedāraṃ prāpayen naraṃ] J11,
+\item[keharaṃ prāpayen naraṃ] Ko,
+\item[kehāraṃ prāpayen naraṃ] J4, 
+\item[kedāraṃ prāpyate naraḥ] J5,
+\item[kedāraṃ prāpnuyān naraḥ] J14,
+\item[kedāraṃ prāsayen raḥ] J2,
+  \item[(illegible/unavailable)] Bo2, Bo3, C5, J3, N4b, N11, N14, P3,
+
+  \begin{description}
+    \end{description}
+
+  \end{marma}
+
+  \begin{marma}[hp02_026]
+
+\sthana{2.26cd}
+
+\item[vāyunā gatim āvṛtya nibhṛtaṃ kaṇṭhamudrayā] B2, C6, C8, J8, J12, J17, K2, LD1, N3, N4a, N5, N9, N11, N20, N21, P9, P10, P12, P13, 
+\item[vāyurnā gatim āvṛtya nimṛtaṃ kaṇṭhamudrayā] N24,
+\item[vāyunā gatim ākṛkhya nibhṛtakaṃṭhamudrayā] N18,
+\item[vāyunā gātim āvṛtya nibhūtaṃ kaṇthamuṃdrayā] J15,
+\item[vāyunā gātim āvṛtya nibhṛtaṃ kaṇthamudrayaṃ] C3,
+\item[vāyunā gātim āvṛtya nibhṛtaṃ kaṇthamudrāyā] J10,
+\item[vāyunā gātim āvṛtya nibhṛtaṃ kathamudrayā] Ko,
+\item[vāyunāṃ gātim āvṛtya vibhṛtaṃ kaṃthamudrayā] C9, J14, 
+\item[vāyunā gātim āvṛtya niścitaṃ kaṃḍamudrayā] Ba, J2, 
+\item[vāyunā gatim āvṛtya nivṛtaṃ kaṇṭhamudrayā] P8,
+\item[vāyunāṃ gatim āvṛtya nibhṛtaṃ kaṃṭhamudrayā] J3, J6a, J7, K1, N7, N10, N13, P4, P7, P11, 
+\item[vāyunāṃ gātim āvṛtya nibhūtaṃ kaṃṭhamudrayā] J16,
+\item[vāyunāṃ gātim āvṛtti nibhṛtaṃ kaṃṭhamudrayā] J4, J6,
+\item[vāyunā gatim ākṛṣya nibhṛtaṃ kaṇṭhamudrayā] A1, B1, J13, LD2, P1, P2, 
+\item[vāyunā gatim ākṛṣya nibhṛta kaṇṭhamudrayā] J23, N23,
+\item[dhāvāyunā gatim ākṛṣya nibhṛtaṃ kaṇṭhamudrayā] J5,
+\item[vāyunāṃ gatim ākṛṣya nibhṛtaṃ kaṃṭhamudrayā] B3, P6, 
+\item[vāyūnāṃ gatim ākṛṣya nibhṛtaṃ kaṃṭhamudrayā] C2, N16,
+\item[vāyūnāṃ gatim ākṛṣya nibhṛtaṃ haṭhamudrayā] J1,
+\item[vāyūnām atimāvṛttya nibhṛtaṃ kaṃṭhamudrayā] N6,
+\item[vāyūnāṃ gatim āvṛttya nibhṛtaṃ kaṃṭhamudrayā] C4, J11, N12, N25, 
+\item[vāyunā gamitikṛṣya nibhṛtaṃ kaṇṭhamudrayā] N22,
+\item[vāyunāṃ gatim āvṛtya nibhaṃ taṃ kaṃṭhamudrayā] Bo1, 
+\item[vāyunāmatim āvṛtya nibhṛtaṃ kaṇthamudrayā] N17,
+\item[vāyunā gatim āvṛtya nirbhṛtaṃ kaṇṭhamudrayā] P5,
+\item[(illegible/unavailable)] Bo2, Bo3, C5, N4b, N8, N14, N15, P3,
+
+    \begin{description}
+      The meaning of \emph{nibhṛtam} is not clear. 
+    \end{description}
+
+  \end{marma}
+
+\begin{marma}[hp02_028]
+
+\sthana{3.28ab}
+
+\item[somasūryāgnisambandhāj jāyate cāmṛtāyate]
+\item[somasūryāgnisambandhāt jāyate cāmṛtāya vai] B2, J14, K1, Ko, N3, N21, N22, P6, P11, P13, 
+\item[somasūryāgnisambandho jāyate cāmṛtāya vai] Ba, Ba5, N4a, N12, N13, 
+\item[somasūryāgnisambandho jāyate cāmṛtāya va] N24,
+\item[somasūryāgnisambandhāj jāyate ca mṛtāya vai] J8, J11a, N20,
+\item[somasūryāgnisambandho jāyate cāmṛtāya vau] J2,
+\item[somasūryāgnisambandho jāyate cāmṛtāyate] J12,
+\item[somasūryāgnisambandhāj jāyate cāmṛtāya vau] J5,
+\item[somasūryāgnisambandha jāyate cāmṛtāyate] N16,
+\item[somasūryāgnisambandhāj jāyate cāmṛtāyate] A1, C6, J7, N23, P7, 
+\item[somasūryāgnisambandhāj jāyate cāmṛtāya tu] C8,
+\item[somasūryāgnisambandhāj jāyate cāmṛtāya ca] Bo1, N7,
+\item[somasūryāgnisaṃbaṃdhāj jāyate camṛtiṃ jayaḥ] B1, B3, C2, J13, LD2, P2, P4, P12, 
+\item[somasūryāgnisaṃbaṃdhā jāyate mṛtur jjayaḥ] N15,
+\item[somasūryāgnisaṃbaṃdhāj jāyate ca mṛtur jjayaḥ] J17, LD1, P8, 
+\item[somasūryāgnisaṃbaṃdhāj jāyate ca mṛtejaṃ yaḥ] C3,
+\item[somasūryāgnisaṃbaṃdhā jāyate ca mṛtyuṃjayaḥ] C9,
+\item[somasūryāgnisaṃbaṃdhāj jāyate ca mṛturjayaḥ] J10,
+\item[somasūryāgnisaṃbaṃdhāj jāyate ca mṛtyor jjayaḥ] N17,
+\item[somasūryāgnisaṃbaṃdhāj jāte ca mṛtyor jayaḥ] N9,
+\item[somasūryāgnisaṃbaṃdhāj jāyate ca mṛter jayaḥ] J10a, J15, N10, P1, P5,
+\item[somasūryāgnisaṃbaṃdhāj jāyate ca mṛtter jayaḥ] P9,
+\item[somasūryāgnisaṃbaṃdhā jāyate ca mṛter jayā] J16,
+\item[somasūryāgnisaṃbaṃdhāj jāyate ca mṛtyu jayaḥ] P10,
+\item[somasūryāgnisaṃdhānā jāyate cāmṛtāyate] N25,
+\item[somasūryāgnisaṃdhānāj jāyate cāmṛtāyate] C4,
+\item[somasūryāgnisaṃdhānā jāyate vāmṛtāyate] N5,
+\item[somasūryāgnisaṃdhānāj jāyate cāmṛtāya ca] J11,
+\item[somasūryāgnisaṃdhānāṃ  jāyate cāmṛtāya ca] J4, 
+\item[somasūryāgnisaṃdhānaṃ jāyate cāmṛtāyate] J6, N11, N18,
+\item[somasūryāgnisaṃdhānā jāyate bāmṛtāyate] K2,
+\item[somasūryāgnisaṃdhānaṃ jāyate vāmṛtāyate] J3,
+\item[somasūryāgnisaṃdhānaṃj jāyate cāmṛtāyate] J1,
+\item[somasūryāgnisambandhāj jāyate jāyate tripuṭāśrayaḥ] N6,
+\item[somasūryāgnisambaddha mṛtyur jjayati yogavit] N14, 
+  \item[(illegible/unavailable)] Bo2, Bo3, C5, N4b, N8, P3,
+
+  \begin{description}
+  \end{description}
+
+\sthana{3.28cd}
+
+\item[mṛtāvasthā samutpannā tato mṛtyubhayaṃ kutaḥ] A1, B1, B2, C2, C4, C6, C8, J1, J4, J5, J6, J7, J8, J11, J13, J14, K1, K2, N3, N5, N11, N12, N16, N20, N22, P2, P4, P6, P11, P12, 
+\item[mṛtāvasthā samutpannā tato mṛtyubhayaṃ jayet] J12, J16, N10,
+\item[mṛtāvasthā samutpannā tato mṛtyubhayaḥ kutaḥ] B3,
+\item[mṛtāvaschā samutpaṃnnā tato mṛtyubhayaḥ kutaḥ] Ba,
+\item[mṛtāvasthā samotpannā tato mṛtyubhayaṃ kutaḥ] N25,
+\item[mṛtavasthā samutpannā tato mṛtyubhayaṃ kutaḥ] P7,
+\item[mṛtāvasthā samutpannā tato vāyuṃ virecayet] Ba5, C9, J10, J17, LD1, N4a, N6, N7, N9, N13, N14, N17, N18, N21, P1, P5, P9, 
+\item[mṛtāvasthā samutpannā tato vāyu virecayet] P8, P10, 
+\item[mṛtāvasthāṃ samutpannā tato vāyuṃ viruṃcyatet] LD2,
+\item[mṛtāvasthā samutpannā tato vāyu virsarjjayet] Ko,
+\item[mṛtā ‘vasthā samutpannā tato vāyuṃ virecayet] C3, 
+\item[mṛtāvaschā samutpannā tato vāyuṃ viracayet] Bo1,
+\item[mṛtāvastu samutpanna tato vāyuṃ virecayet] J2,
+\item[samutpannā mṛtāvasthā tato vāyuṃ virecayet] N24,
+\item[mṛtāvasthā samutpannā recayec ca tato vāyu] N15,
+\item[mṛtavasthā tato vāyuṃ niruṃdhayet kuṃbhakena] N23,
+\item[(illegible/unavailable)] Bo2, Bo3, C5, N4b, N8, P3, P13,
+
+  \begin{description}
+    \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_029]
+
+\sthana{3.29c}
+
+\item[valīpalitavedhagnaḥ]
+\item[valīpalitavedhaghnaḥ] P13,
+\item[valipalitavedhaghnau] J2,
+\item[valīpalitavandhaḥ] C3, 
+\item[valīpalitavepaghnaḥ] Ba5, J1, J3, J6, J14, N4a, N11, N12, N13, N16, N18, N23, 
+\item[valipalitaveghaghnaḥ] N22,
+\item[valīpalitaveghaghnaḥ] K1, Ko, N20, N24, P11, 
+\item[valīpalitavelaghnaḥ] N21,
+\item[valīpalitavegāghnaḥ] N25,
+\item[valīpalitabaṃdhaghnaḥ] J7, J10, J12, J16, J17, LD1, N6, N9, N17, P1, P5, P8, P9, 
+\item[valipalitabaṃdhaghnaḥ] N7, N10, N15, 
+\item[valīpalītabaṃdhagghnaḥ] LD2,
+\item[valīpalitavegaghnaḥ] C4, K2, N5,  
+\item[valipalitavegaghnaḥ] J15,
+\item[valipalitavedhaghna] P6, P7,
+\item[valītpalitavedhaghna] P10,
+\item[valīpalitavedhaghnaṃ] J8,
+\item[valīpalitaveghaghnaṃ] N3,
+\item[valīpalīvedhaghnaḥ] Ba,
+\item[valiḥ palitavedhaghnaḥ] J5,
+\item[valīpalitavedhaghnaḥ] C6, J4, J11, 
+\item[vallīpalitavedhaghnaḥ] A1, B2, Bo1, 
+\item[valīpalitarogaghnaḥ] C8,
+\item[valīpalitavaṃdhaghnaḥ] C9,
+\item[vallīpalitanirmuktaḥ] B1, B3, C2, J13, P2, P4, P12, 
+  \item[(illegible/unavailable)] Bo2, Bo3, C5, N4b, N8, N14, P3,
+
+    \begin{description}
+ The occurance of \emph{vedha} in third \emph{pāda} is rather strange as it does not seem related to the other two symptoms of old age (i.e., wrinkles and grey hair) that this \emph{mudrā} can cure.
+The reading \emph{vepa} (`trembling') of the \emph{Yogacintāmaṇi} and \emph{Jyotsnā} is much better in this regard.
+Also, \emph{vega} (`agitation' or `shock') renders a somewhat plausible meaning, but is not strongly attested.
+    \end{description}
+
+  \end{marma}
+
+\begin{marma}[hp03_039]
+
+\sthana{2.39c}
+
+\item[vrajaty ūrdhvaṃ haṭhāc] C3, 
+\item[vrajaty ūrdhvaṃ haṭhāchaktyā] J3, J10, 
+\item[vrajaty ūrdhvaṃ haṭhācchaktyā] J11, J14, N12, 
+\item[vrajaty ūrdhvaṃ haṭha śaktyā] N22,
+\item[vrajaty ūrdhvahaṭhāchaktyā] J7, J12, 
+\item[vrajaty ūrdhvaṃ haṭhā chaktā] J4,
+\item[vrajaṃty ūrhaṃ haṭhā chaktā] J16,
+\item[vrajaty ūrdhvaṃ haṭhā ch] A1, C6, C9,
+\item[vrajaty ūrdha haṭhā chāktā] J1,
+\item[vrajaty ūrdhvaṃ haṭhā śaktyā] C8, P6, 
+\item[vrajaty ūrdhvaṃ haṭhāt śaktyā] P7,
+\item[vrajaty ūrdhvaṃ haṭhaḥ śaktyā] P2, P12, 
+\item[vrajaty ūrdhvaṃ haṭhā chaktyā] J3, J8, N7, N9, N10, N16, N18, P1, P5, P8, P10, 
+\item[vrajaty ūrdhvaṃ haṭhāc chaktyā] J6, J15, J17, N6,
+\item[vrajaty ūrdhvaṃ haṭhāt....tkṣā] P9,
+\item[vrajaty ūrdhvaṃ hi tac chaktyā] J6a, N11, N25, 
+\item[vrajaty urdhaṃ hi chaktyā] N5,
+\item[vrajaty ūrdhvaṃ hitachaktyā] C4, N8,
+\item[vrajaty ūrdhvaṃ tataḥ śaktyā] K1,
+\item[vrajaty ūrdhaṃ tataḥ ṣaktya] B2,
+\item[vrajaty ūrdhaṃ hataḥ śaktyā] Ba, C2,
+\item[vrajaty ūrdhaṃ haṭaḥ śaktyā] P4,
+\item[vrajaty ūrdhaṃ hṛtaḥ śaktyā] Ba5, N13, 
+\item[vrajaty ūrddhvaṃ hṛtaḥ śaktyā] N4a,
+\item[vrajety ūrdhvaṃ haṭhāt chaktyā] J5, 
+\item[vrajety ūrdhvaṃ haṭhāt śhaktyā] J13,
+\item[nayed ūrdhvaṃ haṭhātchaktyā] Bo1,
+\item[vrajaty ūrdhvaṃ hi tachaktyā] K2,
+\item[vṛajaty ūrdhva hatāśatsā] J2,
+\item[vṛjaṃty ū......] N3,
+\item[vrajya urddhe haṭhā śaktyā] LD2,
+\item[nayed urddhvahaṭhāchaktyā] N21, P13, 
+\item[vrajaty ūrdhvamukhe chāktyā] N15,
+\item[jajatyūrdhvaṃ hatāchatkā] N23,
+  \item[(illegible/unavailable)] B1, B3, Bo2, Bo3, C5, Ko, LD1, N4b, N14, N17, N20, N24, P3, P11,
+
+    \begin{description}
+    • It is at 45c in A1; C6;
+    • It is at 62c in C2
+    • It is at 63c in J1; J3;
+    • It is at 64c in J13;
+    • It is at 43c in C8
+    • It is at Image no. 38 in J4
+    • It is at 48c in J12
+      Most of the collated manuscripts have the reading \emph{haṭhāt śaktyā}, which does not make good sense.
+The \emph{Vivekamārtaṇḍa}'s reading \emph{haṭaḥ śaktyā} (`struck by kuṇḍalinī') is not found among the \emph{Haṭhapradīpikā}'s manuscripts, but is close to N23.
+The reading of 3a2 (and the YCM), \emph{hi tacchaktyā} appears to be the most plausible reading of the collated manuscripts.
+It could be understood as `by the power of  khecarī', which would mean that \emph{yonimudrā} blocks \emph{bindu}'s downward course and \emph{khecarī} causes it to go upwards.
+    \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_046]
+
+\sthana{3.46f}
+
+\item[vimalaṃ dhārāmṛtaṃ yaḥ pibet] C2, C3, C6, C8, C9, J3, J4, J5, J8, J13, J14, J15, J17, Ko, LD1, N7, N8, N9, N11, N15, N16, N25, P1, P6, P8, P9, P13, 
+\item[vimalaṃ dhālāmṛtaṃ yaḥ pibet] N14,
+\item[sumūla dhārāmṛtaṃ yaḥ pibet] J12,
+\item[vimalaṃ dhārāmṛtaṃ yaḥ pībe] N22,
+\item[vimala dhārāmṛtaṃ yaḥ pivet] J1, N3, 
+\item[vimalaṃ dhārāmṛtyaṃ yaḥ pibet] P10,
+\item[vimalaṃ dhāraṃ mṛtaṃ yaḥ pibet] B3, 
+\item[vimalaṃ dhārāmṛtaṃ yaḥ piven] A1, B2, C4, J6, J7, J10, J11, K2, N5, N13, N17, N18, N21, N23, P2, P4, P5, P7, P11, P12, 
+\item[vimalaṃ dhārāmṛta yaḥ piven] N6,
+\item[vimalaṃ dhārāmṛtaṃ yaḥ piveṃn] LD2,
+\item[vimalaṃ dhārāmṛtaṃ yaḥ pīvet] K1,
+\item[vimalaṃ dhārāmṛtaṃ yaḥ pīven] J16,
+\item[vimalaṃ dhārāmayaṃ yaḥ pibet] Ba5, N4a, N12, 
+\item[vivare dhārāmayaṃ yaḥ pibet] Ba, J2, 
+\item[vimalaṃ dhārāmṛtaṃ yaḥ pibaṃn] N10,
+\item[vimalaṃ dharamṛtaṃ yaḥ piven] N24,
+\item[(illegible/unavailable)] B1, Bo1, Bo2, Bo3, C5, N4b, N20, P3,
+
+  \begin{description}
+    • It is given at 75 in B3
+    • It is given at 72f in C2
+    • It is given at 47f in C8
+    \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_049]
+
+\sthana{3.49}
+
+\item[pātāle yadviśati suṣiraṃ merumūle yadasti | 
+tadvaccaitat pravadati sudhīs tanmukhaṃ nimnagānām | 
+candrātsāraḥ sravati vapuṣastena mṛtyur narāṇāṃ 
+taṃ badhnīyāt sukaraṇam atho nānyathā kāyasiddhiḥ ||] J14,
+
+\item[pātāle yadvaśati sukhiraṃ merumūle yadasti
+tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrātsāraḥ śravati vapuṣos tena mṛtyur narāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kāryasiddhi] P13,
+
+\item[pātāle ye dvitayasūṣire merumūle tadasmins
+tatvaṃ yattat pravadati sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdraṃ sāraḥ śravaṃtyau vapuṣas tanema mṛtyuṃ nnarāṇāṃ
+bandhiyātakaraṇamamṛtaṃ nānyathā kāryasiddhi ||] P6,
+
+\item[yatpātāle viśati suṣiraṃ merumūle yadasti
+tasmins tatvaṃ pravadati sudhīs taṃnmukhaṃ nimnagānāṃ ||
+caṃdrātsāraḥ sravati vapuṣaḥ stena mṛtyur narāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ ||] P2,
+
+\item[yatpātāle viśati sukhiraṃ merumūle yadasti
+tasmiṃs tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānā ||
+caṃdrātsāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ ||] P4,
+
+\item[yatpātāle viśati sukhiraṃ merumūle yadasti
+tasmiṃs tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdrātsāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kāyasiddhi ||] P12,
+
+\item[yahapāledviśati sukhiramahamūle yadāsti
+taviśve tat pravadati sudhā tanmukhaṃ nimnagānāṃ
+caṃdrāsāra sravati vapuṣaṣto na matyu narāṇāṃ
+taṃ vadhnīyāt sukaraṇamatho nānyathā kāryyasiddhi] N24,
+
+\item[yatprāleyaṃ pihītasuṣiraṃ merūmurddhātathāṃtaṃ
+tasmiṃstatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrātsāraḥ śravati vapuṣas tena mṛtyur nnarāṇāṃ 
+taṃ badhnīyāt sukaraṇamatho nānyathā kāryasiddhi] J15,
+
+\item[yatprāleyaṃ cāpihitamukhisorūmūddhnaṃ tathyaṃ
+tasmiṃs tatvaṃ pravadaṃti sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdrāt sāraḥ śravati vapuṣes tena mṛtyuṃ nnarāṇāṃ
+taṃ vadhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ ||] P10,
+
+\item[pātāle yadvitayuśubhiraṃ merumūlai tadasmin
+tatvaṃ caitat pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃ taṃ prasāraṃ grasati vapuṣā doṣamṛtyunarāṇāṃ
+taṃ badhnīyāt sukaraṇamṛdā nānyathā kāryasiddhi] P11,
+
+\item[yatprāleyaṃ pihitasukhiraṃ merumūrdhdhnaṃ tatathyaṃ
+tasmiṃs tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdrātsāraḥ śravati vapuṣas tena mṛtyur narāṇāṃ
+taṃ vadhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ ||] P9,
+
+\item[yatproleyaṃ pihitasukhiraṃ merumūrdhnyātathyaṃ
+tasmistatvaṃ pravadati sudhīstanmukhaṃ nimnagānāṃ
+candrochāraḥ śravati vapuṣestenamṛtyunnarāṇāṃ
+taṃ pātāle yadviśatisuṣiraṃ merumūle yasti
+tadvaccaitat pravadati sudhīs tanmukhaṃ nimnagānāṃ
+vadhnāyāstu karaṇamatho caṃdrātsāraḥ
+sravati vapuṣestena mṛtyurnnarāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ] P8,
+
+\item[pātāle yadvitaya sukhiraserumūles tadasmin
+statvaṃ yatat pravadaṃti sudhās tanmukhaṃ nimnagānāṃ
+cāṃdraṃ sāraṃ śravati vapuṣas tena mṛtyur narāṇāṃ
+badhnayātat kāraṇamamataṃ nānyathā kāryasiddhiḥ] N22,
+
+\item[pātālād yadviśatiśukhiraṃ merumūle tadasti
+tatvaṃ caitat pravadati sudhīḥ tatsukhaṃ nimnagānāṃ ||42||
+caṃdrāt sāraṃ sravati vapuṣā tena mṛtyur narāṇāṃ
+taṃ badhnīyāt sukaramṛdā nānyathā kāryasiddhiḥ ||] P7,
+
+\item[yatprāleyaṃ cāpihitaṃ suṣiraṃ merumūrddhataḥ 
+tasmiṃs tatvaṃ pravadaṃti sudhīs tanmukhaṃ nimnagānām
+candrāt sravatipīyūṣas tena mṛtyur nnarāṇāṃ
+taṃ vadhnīyāt sukaraṇamatho nānyathā kāryyasiddhi] N16,
+
+\item[yatprāleyaṃ pihitasukhiraṃ merumūrddhnitithyaṃ 
+tasmiṃs tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdrātsāraṃ śravati vapuṣas tena mṛtyuṃ narāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kā prasiddhiḥ ||] P5,
+
+\item[yatprāleyaṃ pihitaśiṣire merūmūrddhaṃtathāṃtaṃ
+tasmiṃstatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrātsāraḥ śravati vapus tena mṛtyur nnarāṇāṃ
+taṃ vadhnīyād varūṇamamṛtaṃ nānyathā kāryasiddhiḥ] N25,
+
+\item[yatprāleyaṃ pihitasukhiraṃ me tu samūrddhye tathyaṃ
+tasmistatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdrātsāraḥ śravati vapuṣas tena mṛtyurnnarāṇāṃ 
+taṃ badhnīyāt sukaraṇamatho nānyathā kāryasiddhiḥ ||] P1,
+
+\item[yatprāleyaṃ pihitasukhiraṃ merumuddhnastitathyaṃ
+tasmin tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ 
+caṃdrātsaraṃ sravati vapuṣas tena mṛtyur narāṇāṃ
+taṃ vadhnīyāt sukaraṇamatho nānyathā kāryyasiddhi] N18,
+
+\item[yatprāleyaṃ pihitasukhiraṃ merumūrddhnāṃ tathyaṃ
+tasmiṃs tatvaṃ pravadati sudhī yas tanmukhaṃ nimnagānāṃ ||
+candrātsāraḥ sravati vapuṣas tena mṛtyur nnarāṇāṃ
+taṃ vadhnīyād varuṇamamṛtaṃ nānyathā kāryyasiddhiḥ ||] N17,
+
+\item[yatprāle pihitasukhiraṃ merumūrddhyaṃ || 
+tathyaṃ tasmīs tatvaṃ pravadati sudhiḥ tanmukhaṃ nimnagānāṃ
+cadrātsāraḥ śravati vapuṣas tena mṛtyur narāṇāṃ ||
+ta vadhnīyāt sukaraṇamano nānyathā kāryasiddhi ||] N15,
+
+\item[yatprāleya prahitasukhiraṃ merūmūrddhāṃtarasthaṃs
+tasmiṃs tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrātsāraḥ sravati vapuṣaḥ tena mṛtyur narāṇāṃ
+tadvadhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ] N13,
+
+\item[pātāle yadvitayasuṣiraṃ merumūle tadasmiṃs
+tatvaṃ caitat pravadati sudhī tanmukhaṃ nimnagānāṃ ||
+candrāsāraṃ sravati vapuṣo doṣamṛtyur narāṇaāṃ
+tad badhnīyāt sukaraṇamaho nānyathā kāryasiddhiḥ ||] N10,
+
+\item[pātāle yad vasati sukhiraṃ merumūle yadasti
+tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrātsāraḥ śravati vapuṣas tena mṛtyur narāṇā
+taṃ vaddhniyāt sukaraṇamatho nānyathā kāryasiddhiḥ] N21,
+
+\item[yatprāleyaṃ cāpihittasukhiraṃ meru mūddhā tathyaṃ
+tasmis tatvaṃ pravati sudhīs tanmukhaṃ nimnāgānā
+caṃdrāt sāraḥ śravati vapuṣes tena mṛtyuṃ narāṇāṃ
+na vadhnīyāt sukaraṇāmadho nānyathā kāyasiddhi] N9,
+
+\item[pātāle yadvicayaṃ suṣiraṃ meruūle tadasmin
+tatvaṃ caitat pravadati sudhīs tanmukhaṃ nimnagāyāḥ |
+candrāsāraṃ grasati vapuṣā doṣamṛtyur narāṇāṃ
+taṃ vadhnīyāt sukaraṇamahā nānyathā kāryasiddhiḥ ||] N12,
+
+\item[yatprāleyaṃ pihitasuṣiraṃ merumūrddhnyāṃ tathyaṃ
+tasmis tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ |
+caṃdrātsāraḥ sravati vapuṣas tena mṛtyur nnarāṇāṃ
+taṃ taṃ vadhnīyād varuṇamamṛtaṃ nānyathā kāryyasiddhi ||] N6,
+
+\item[yatprāyaṃ pihitaḥ sukhiraṃ merumūrddāṃtarasthaṃ
+tasmin tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ |
+caṃdrātsāraḥ sravati vapuṣe tena mṛtyur nnarāṇāṃ
+taṃ taṃ vadhnīyād sukhakaramathā nānyathā kāryasiddhi ||] N7,
+
+\item[yatprāleyaṃ pihitaśikharaṃ merumūrdhnisthitaṃ yat
+tasmin tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrātsāraḥ śravati vapuṣas tena mṛtyur narāṇāṃ
+tad vadhnīyāt karaṇamamṛtaṃ nānyathā kāryasiddhiḥ] Ko,
+
+\item[pātāle yadvitayasukhiraṃ merūmūle tadasmiṃs
+taṃtvaṃ cetat pravadatī sudhāṃ tanmūkhaṃ nimnagānāṃ 
+caṃdrāsāraṃ sravati vapuṣo doṣamṛtyu narāṇāṃ
+tadbadhnīyāt sukarora hī nānyathā kāryaṃ siddhī] J16,
+
+\item[pātāle ya dvitiya sukhiraṃ merumūle tadasmin
+statvaṃ caitat pravadati sudhā tanmukhaṃ nimnagānāṃ
+caṃdrāsāraṃ sravati vapuṣo doṣamṛtyur narāṇāṃ
+tad badhnīyāt sukarāgama hi nānyathā kāryasiddhiḥ] J12,
+
+\item[yatprāleya prahitasuṣiraṃ merumūrdhnāstitathyaṃ
+tasmiṃstatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrātsāraṃ sravati vapuṣo doṣamṛtyur narāṇāṃ
+tadvadhnīyāt sakaraṇamayo nānyathā kāryasiddhiḥ] K1,
+
+\item[pātāle ya dvitayasuṣiraṃ merumūle tadasmin
+tadvac caitā pravadati sudhī tanmukhaṃ nimnagānāṃ
+caṃdrāt sāraḥ sravati vapuṣas tenamṛtyur narāṇāṃ
+taṃ vadhnīyāt sukaraṇahā nānyathā kāryasiddhiḥ] J11,
+
+\item[pātāle yad viśati sukhiraṃ merumūle yadasti 
+tadva caitat pravadati sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ
+tad vadhnīyāt sukhakaramatho nānyathā kāyasiddhiḥ ||] J7,
+
+\item[yatprāleyapihitasukhiraṃ merumūrddhni sthitaṃ
+tasmiṃs tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ |
+caṃdrāt sāraḥ śravati vapuṣas tena mṛtyur nnarāṇāṃ
+taṃ vadhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ ||] J10,
+
+\item[yatprāleyaṃ pihita suṣiraṃ merumūdhnāṃ tathāṃntyaṃ
+tasmiṃstatvaṃ prabadati sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdrātsāraḥ sravati bapuṣas tena mṛtyur jarāṇāṃ |
+taṃ badhnīyadguṇamamṛtaṃ nānyathā kāryasiddhiḥ] K2,
+
+\item[yatprāleyaṃ pihitasukhiraṃ merumūrddhni sthitaṃ
+tasmiṃs tatvaṃ pravadaṃti sudhīs tanmukhaṃ nimnagānāṃ
+candāt sāraḥ śravati vapuṣes tena mṛtyun narāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ] J17,
+
+\item[pātāle yadvitayasukhiraṃ merumūle tadasti
+tatvaṃ caitat pravadati sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrāsāraṃ sravati vapuṣo doṣamtyu narāṇāṃ
+na badhnīyāt sukaraṇamaho nānyathā kāryasiddhiḥ] LD1,
+
+\item[pātāle yadviśati sukhira merūmūle pakṣasti
+tadvac caitat pravadati sudhās tanmukhaṃ niṣagmanāṃ ||
+caṃdrātsāraḥ rapativapuṣas tena mṛtyur narāṇāṃ 
+tacha-yāt sukhakaraṇamartho nāmarthā kāyasiddhiḥ ||] N23,
+
+\item[yatpātāle vrajaddhitapasukhiraṃ merumūle ca
+tasmin tattvaṃ pravadati sudhīs tanmukhaṃ korthaḥ nāḍīnā kāryaḥ |
+caṃdrāt sāraḥ śravati vapuṣas tena mṛtyur nnarāṇāṃ
+taṃ vadhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ ||] J10a,
+
+\item[yatprālepaṃ pihitasukhire merumūle yadastī
+tasmi tvaṃ pravadaṃti sudhīs tanmukhaṃ nimnagānāṃ
+caṃdrātsāraṃ śravati vapuṣaṃ stenamṛtyuṃr nnarāṇāṃ |
+taṃ badhnīyāt sukaraṇamṛdā nānyathā kāryasiddhiḥ ||] J5,
+
+\item[yatprāleyaṃ cāpihitaḥ sukhiraṃ merumūrddhātaḥ 79 ||
+tasmin tatvaṃ pravadati sudhīs tanmukhaṃ nimnagānāṃ ||
+caṃdrotsāraḥ śravati vapuṣas tena mṛtyur narāṇāṃ ||
+taṃ badhnīyāt sukaraṇamatho nānyathā kāryasiddhiḥ ||] J13,
+
+\item[yatprāleyaṃ cāpi hitasukhiraṃ meha mūrddhāṃtaḥ
+tasmin tattvaṃ pravadati sudhīs tanmukhaṃ bhinnagānāṃ
+caṃdrātsāraḥ sravati ca pīyūṣas tenamṛtyur narāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ] J6,
+
+\item[yat prāleyaṃ cāpi hitasukhiraṃ merumūrddhvataḥ ||
+tasmiṃs tatvaṃ pravadati sudhyaṃs tanmukhaṃ nimnagānāṃ ||74||
+candrāt sāraḥ śravati vapayuṣes tena mṛtyun narāṇāṃ |
+taṃ vadhnīyātasukaraṇamatho nānyathā kāyasiddhiḥ ||] J8,
+
+\item[yatprāleyaṃ vahati suṣiraṃ meha mūrddhāṃtaḥ
+tasmin tattvaṃ pravadati sudhīs tanmukhaṃ bhinnagānāṃ
+caṃdrātsāraḥ sravati satataṃ tenamṛtyur narāṇāṃ
+taṃ badhnīyāt sukaraṇamatho nānyathā kāyasiddhiḥ] J6a,
+
+\item[pātāle yannitayasuṣiraṃ merumūle tad asti
+tadvac caitat pravadati dhīstanmuratanimnagānāṃ ||
+caṃdrāsāraṃ sravati vapuṣes tenaćtyur narāṇāṃ
+tad vadhnīyātyakaraṇaćdā nānyathā kāryasiddhi ||] J4,
+
+\item[pātālād yad viśati suśiraṃ merumūle tadasti
+tatvaṃ caitat pravadati sudhīs tatsukhaṃ nimnagānām |
+caṃdrātsāraṃ sravati vapuṣā tenamṛtyur narāṇāṃ
+taṃ badhnīyāt sukaraṇamṛdā nānyathā kāyasiddhiḥ ||] C6,
+
+\item[pātālād yadi viśati sukhiraṃ merumūle tadasthi
+tatvaṃ caitat pravadati sudhīs tatsukhaṃ nimnamānām |
+caṃdrātsāraṃ sravati vapuṣā tena mṛtyur narāṇāṃ
+taṃ badhnīyāt sukaraṇamṛdunā nānyathā kāryasiddhiḥ ||] C8,
+
+\item[yat pātāle viśati suṣiraṃ merumūle yadasti | 
+tasmiṃstatvaṃ pravadati sudhīs tan mukhaṃ nimnagānāṃ
+candrātsāraḥ sravati vapuṣastena mṛtyur narāṇāṃ 
+taṃ badhnīyāt sukaraṇam atho nānyathā kāyasiddhiḥ ||] B3,
+
+\item[yatpātālapihītasuśīraṃ merumūlena yadasti
+tasmiṃstatvaṃ pravadati sudhīs ta(nsu)ṣaṃ nimnagānāṃ ||55||
+caṃdrāsāra śravati vapuṣas tena mṛtyuṃ narāṇāṃ
+taṃ vadhniyāt sukaraṇaṃ yatho nānyathā kāryasiddhiḥ ||56||] LD2,
+
+\item[yat pātāle vikhati suṣiraṃ merumūle yadasti | 
+tasmiṃstatvaṃ pravadati sudhīs tan mukhaṃ nimnagānāṃ
+candrātsāraḥ sravati vapuṣastena mṛtyur narāṇāṃ 
+taṃ badhnīyāt sukaraṇam atho nānyathā kāyasiddhiḥ ||] C2,
+
+\item[prāleya prahita suṣiraṃ merumūrdhāṃtarasthaṃ
+tasmiṃs tatvaṃ pravadati sudhīs tan mukhaṃ nimnagānām |
+caṃdrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ
+tad badhnīyāt sukaraṇamato nānyathā kāyasiddhiḥ ||] Ba5,
+
+\item[yatprāleya prahita suṣiraṃ merumūrdhāntarasthaṃ
+tasmin tatvaṃ pravadati sudhīs tan mukhaṃ nimnagānām |
+candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ
+tad vadhnīyāt sukaraṇamato nānyathā kāyasiddhiḥ ||] N4a,
+
+\item[yatprāleyaṃ pihitasukhiraṃ merumūrddhnāsmitathyaṃ 
+tasmiṃs tatvaṃ pravadati sudhis tanmukhaṃ nimnagāṇāṃ 
+caṃdrāt sāra śravati vapuṣi tenaćtyu narāṇaāṃ
+taṃ vadhniyāt sukaraṇapatho nānyathā kāryasiddhiḥ] C9,
+
+\item[yatprāleyaṃ vahati suṣiraṃ cordhvamūrdhvāṃtasaṃsyaṃ ||
+tasmin tatvaṃ pravadati sudhīs tanmukhaṃ bhinnagānāṃ ||76|
+caṃdrātsāraḥ sravati satataṃ tena mṛtyur narāṇāṃ ||
+taṃ badhnīyāt sukhakaramatho nānyathā kāyasiddhiḥ ||] J3,
+
+\item[yatprāleya prahita sukhiraṃ mesamūrdhāṃ ca tathyaṃ
+tasmiṃs tatvaṃ pravadati sudhīs tan mukhaṃ nimnagānaṃ ||
+caṃdrāt sāraḥ śravati vapuṣas tena mṛtyur narāṇāṃ
+tad badhnīyāt sukaraṇamato nānyathā kāyasiddhiḥ ||] C3,
+
+\item[yatprāleyaṃ pihita suṣiraṃ merumūrdhnā tathāṃ taṃ
+tasmiṃs tatvaṃ pravadati sudhīs tan mukhaṃ nimagnānāṃ ||
+caṃdrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ
+taṃ badhnīyād guruṇmamṛtaṃ nānyathā kāryasiddhiḥ ||] C4,
+
+\item[yatprāleyaṃ pihita suṣiraṃ merumūrdhā? tathāṃtaṃ
+tasmiṃs tatvaṃ pravadati sudhīs tan mukhaṃ nimagnānāṃ ||
+caṃdrāt sāraḥ śravati vapuṣas tena mṛtyur narāṇāṃ
+taṃ badhnīyād guruṇamamṛtaṃ nānyathā kāryasiddhiḥ ||] N5,
+
+\item[pātāle yadvitay suṣi-
+stitatvaṃ caitat pravadati sudhis tanmūkhaṃ ni(śū)gānāṃ
+candrātsāraḥ sravati vapuṣo doṣamṛtyur narāṇāṃ 
+taṃ badhnīyāt suka(ṭha)............nānyathā kāyasiddhi ||] Ba,
+
+\item[yatprāreyaṃ pihitaṃ sukhiraḥ merumurddhyastitathyaṃ
+tasmistatvaṃ pravadaṃti sudhīs tanmukhaṃ nimnagānāṃ ||76||
+candrātsāraḥ śravati ca vapīyūṣastenamṛtyurnarāṇāṃ ||
+taṃ badhnīyāt sukaraṇatho nānyathā kāryasiddhiḥ ||] J1,
+
+\item[(illegible/unavailable)] A1, B1, B2, Bo1, Bo2, Bo3, C5, J2, N3, N4b, N8, N11, N14, N20, P3,
+
+  \begin{description}
+    • It is given at 80 in B3
+    • It is given at 75 in C2
+    • It is given at 41 in C8
+    \end{description}
+
+  \end{marma}
+
+\begin{marma}[hp03_056]
+
+\sthana{3.56ab}
+
+\item[tato jātau vahnyapānau prāṇam uṣmasvarūpakau] J3, J7, J11, 
+\item[tato jātau vahnipānau prāṇamūlasvarūpakau] N18,
+\item[tato jātau vahnyapānau prāṇamūktasvarūpakaṃ] K2,
+\item[tato jātau vahnyapānau prāṇamūktāsvarūpakaṃ] J15,
+\item[tato jātau vahnipānau prāṇemuṣmasvarūpakāṃ] N24,
+\item[tato jātau vahijānau prāṇamūlesvarūpakau] J1,
+\item[tato jātau vahnipānau prāṇam uṣmasvarūpakaṃ] C8,
+\item[tato jātau vahnyapānau prāṇamuktasvarūpakaṃ] C4, N5, 
+\item[tato jātau vahnyapānau prāṇamuktasvarūpakau] J6, J14, N16, 
+\item[tato jātau vaha?pānau prāṇamuktasvarūpakau] N11,
+\item[tato jātau vahnyapānau prāṇamūlasvarūpakaṃ] Ko,
+\item[tato jātau vardhapāne prāṇamuṣmasvarūpakaṃ] J8,
+\item[tato jāto vahipānau prāṇamūlasvarūpaka] N7,
+\item[tato vāyu vahniyogau prāṇamūlasvarūmukaḥ] N17,
+\item[tato yoga vahniyonau prāṇamūlasvarūmukaḥ] P8,
+\item[vato yāto vanhyapānau prāṇam uṣmasvarūpakaṃ] P12,
+\item[vato yātau vahnyapānau prāṇam uṣmasvarūpakaṃ] J4, LD2, P7, P13, 
+\item[vato yātau vahnyapānau prāṇamukhasvarūpakaṃ] N21,
+\item[vato vātau vanhipānau prāṇam uṣmasvarūpakaṃ] K1, 
+\item[vato vāyū vanhipānau prāṇarvddhrcarāvamū] N6,
+\item[tato yātau vahnyapānau prāṇam uṣmasvarūpakaṃ] C2,
+\item[tato yātau brahmapānau prāṇam uṣmasvarūpakaṃ] Ba5, 
+\item[tato yātau vrahmapānau prāṇam uktasvarūpakaṃ] N25,
+\item[tato yātau bāhyapānau prāṇam uṣmasvarūpakaṃ] C6,
+\item[tato jātā vaṃdhapānau prāṇamurasvarūpaṃke] N23,
+\item[tato yātau vāhyāpānau praṇam uṣṇasvarūpakaṃ] P11,
+\item[tato yātau vāhyāpānau prāṇam uṣṇasvarūpakaṃ] J13, P2, P4,  
+\item[tato yāto vahir yonau prāṇamūlasvarūpakaḥ] C9,
+\item[tato yāttau vahnipātau prāṇmūlasvarūpakaṃ] J5,
+\item[tato yātau vahnipānau prāṇmūlasvarūpakau] N8,
+\item[tato yāmau vahnipānau prāṇamūlasvarūpakaṃ] N3,
+\item[tato vahnir apānau ca prāṇmūlasvarūpakaṃ] B2,
+\item[tato cāptau brahmapānau prāṇam uspasvarūpakaṃ] N22,
+\item[tato yo vahni yo vahni yonau prāṇaṃ mūlasvarūpakaḥ] N15,
+\item[tato yāte vahnir yonau prāṇe mūlasvarūpakaḥ] C3,
+\item[tato yāte vahniyonau prāṇa mūlasvarūpakaḥ] J17, P1, P5, P10, 
+\item[tato yāte vahnyar yono prāṇe mūlasvarūpake] J16,  
+\item[tato yāte bahir yonau prāṇe mūlasvarūpake] J12,
+\item[tato yāte vanhipānau prāṇe mūlasvarūpake] N10,
+\item[tato yāne vahnipānau prāṇamūlasvarūpakaḥ] N9,
+\item[tato yāto vahni yonau prāṇamūlasvarūpakaḥ] J10,
+\item[tato yāto vanhyapānau prāṇamūṣmasvarūpakaṃ] N4a, N13,
+\item[tato yāto vanhyapānau prāṇamūṣṇasvarūpakaṃ] N4b,
+\item[tato pātau vahnyāpānau prāṇam uṣṇasvarūpakaṃ] A1, B3,
+\item[tato yāto vahnipānau prāṇamuṣṇasvarūpakaṃ] LD1,
+\item[tenātyaṃta pradīpastu jvalano dehajas tathā] Ba, 
+\item[tato vāyuvahnir yātau prāṇmūrañ caturmukha] N14,
+\item[tato yātau vahnyapārṣṇau prāṇamuhmyasvarūpakaṃ] P6,
+\item[tato sthito vanhiyonau prāṇamūlasvarūpakaḥ] P9,
+\item[(illegible/unavailable)] B1, Bo1, Bo2, Bo3, C5, J2, N12, N20, P3,
+
+  \begin{description}
+    • It is given at 74ab at Ba
+    • It is given at 85ab in C2
+    \end{description}
+
+  \end{marma}
+
+  \begin{marma}[hp03_056]
+
+\sthana{3.57d}
+
+\item[niḥśvasya] A1, B2, Ba, Ba5, C6, J14, LD1, N6, N10, N17, P4, P12,
+\item[niḥśvayād] J13,
+\item[niścitam] J1, J3, J7, Ko, N8, N11, N16, N18, N23, V19, N19, 
+\item[niśvasyād] C4, N25, P5, 
+\item[niścayād] N5,
+\item[niśvasya] B3, C2, C8, C9, J2, J4, J12, J15, J16, J17, Jyo, N3, N4a, N4b, N7, N9, N12, N13, N15, N21, N22, N24, P1, P2, P7, P8, P11, P13, 
+\item[nīśvasya] LD2,
+\item[niśrayād] K2,
+\item[niśvāsya] J5,
+\item[nisvan] K1,
+\item[nisvasya] C3, P9, P10, 
+\item[nisvatya] N14,
+\item[viśvasya] V3, J8, J11, V1, J10, P6, 
+\item[(illegible/unavailable)] B1, Bo1, Bo2, Bo3, C5, J6, N20, P3,
+
+  \begin{description}
+    • It is given at 86d in C2
+    • Folio no. 25 is missing in J6
+    \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_059]
+
+\sthana{3.59}
+
+\item[uḍḍiyānākhyo] B2, C9, J4, J7,J10, J15, J16, LD1, N6, P8, P9,
+\item[uḍḍiyānāṣo] N15,
+\item[uḍiyaṇākhyo] J5,  
+\item[uḍiyanākhyo] J17,
+\item[uḍḍiyaṇākhyo] B3,
+\item[uḍḍiyanākhyo] N7,
+\item[uḍḍiyanākhaṃ] P7,
+\item[uḍiyaṇākhyoyaṃ] P11,
+\item[uḍḍiyaṇākhyoyaṃ] C2, J3, J12, J13, 
+\item[uḍḍiyāṇākhyoyaṃ] K1, P12, 
+\item[uḍḍiyāṇākhyoyāṃ] P13,
+\item[uḍḍiyāṇākhyo] N21,
+\item[uḍḍiyāṇakākhyo] Ko,
+\item[uḍḍiyanākheyaṃ] J1, J11, N3, 
+\item[uḍiyānākhyo] P10,
+\item[uḍiyāṇākhyo] P4,
+\item[uḍḍiyānākhyoyaṃ] N13, N16, N25, 
+\item[uḍiyanākheyaṃ] J8,
+\item[uddiyanākhyoyaṃ] N23,
+\item[uḍḍiyanākhyoyaṃ] C4, K2, N18, 
+\item[uḍḍīyanākhyoyaṃ] N4a, N5, N8, N11, N12, 
+\item[uḍiyanākhyoyaṃ] N10,
+\item[uḍiyakākhyoyaṃ] N24,
+\item[uḍīyāṇākhyo] B1, P2, 
+\item[uḍīyānākhyo] C3, N9, P1, P5, 
+\item[uḍḍīyānākhyo] LD2,
+\item[uḍyāṇākhyo] J2,
+\item[uḍiyānākhyaṃ] C6, C8, 
+\item[uḍḍiyānākhyaṃ] A1, 
+\item[uḍḍiyanākhoyaṃ] Ba, 
+\item[uḍḍiyabaṃdhoyaṃ] J14,
+\item[udiyānākhya] N14,
+\item[udiyānākhyo] N17,
+\item[uḍāyamāṇākhyo] N22,
+\item[uddīyeṇākhyoyaṃ] P6,
+  \item[(illegible/unavailable)] Ba5, Bo1, Bo2, Bo3, C5, J6, N4b, N20, P3,
+
+  \begin{description}
+    • It is given at 93cd in B1
+    \end{description}
+
+  \end{marma}
+
+\begin{marma}[hp03_060]
+
+\sthana{3.60d}
+
+\item[tatra] A1, B2, Ba5, C6, C8, C9, J5, J7, J8, J10a, J11, J12, J14, K1, Ko, LD1, LD2, N4a, N7, N10, N11, N13, N22, N23, N24, P7, P10, V3, N19, P6, P11, V1, Jyo
+\item[atra] B1, B3, C2, J4, J13, N21, P2, P4, P12, P13, 
+\item[kṣatra] N15,
+\item[kṣetra] C3, J10, J15, J16, J17, N6, N9, N14, N17, P1, P8, P9, 
+\item[kṣetraṃ] P5,
+\item[mūla] C4, J1, J3, K2, N3, N5, N8, N12, N16, N18, N25, V19  
+  \item[(illegible/unavailable)] Ba, Bo1, Bo2, Bo3, C5, J2, J6, N4b, N20, P3,
+
+  \begin{description}
+    • It is given at 94cd in B1
+    \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_061]
+
+\sthana{3.61ab}
+
+\item[udare paścimaṃ tānaṃ nābher ūrdhvaṃ ca kārayet] B3, Ba5, C2, C3, C4, C6, J3, J4, J5, J7, J15, K2, N4a, N5, N13, N25, P2, P4, P7, P11, P12, 
+\item[udare paścimetānaṃ nābher ūrdhvaṃ tu kārayet] N6, N14, 
+\item[udare paścimaṃ tānaṃ tānaṃ nābher ūrdhvaṃ tu kārayet] N15,
+\item[udare paścime tāle nābher ūrdhvaṃ ca kārayet] N8,
+\item[udare paścimetānaṃ nābhe nābher ūrdhvaṃ tu kārayet] N7,
+\item[udare paścimatāo nābher ūrdhvaṃ ca kārayet] J16,
+\item[udare paścimaṃ tānaṃ nābher ūrdhve ca kārayet] J13, 
+\item[udare paścime tānaṃ nābher ūrdhvaṃ ca kārayet] N16,
+\item[udare paścime tālaṃ nābher ūrdhve ca kārayet] N11,
+\item[udare paścime tānā nābher ūrdhvaṃ ca kārayet] N18,
+\item[udare paścimaṃ tānāṃ nābher ūrdhvaṃ ca kārayet] J12,
+\item[udare paścime tānaṃ nābher ūrdhvaṃ ca kārayet] J11, N12, 
+\item[udarāt paścime bhāge nābher ūrdhvaṃ ca kārayet] J8,
+\item[udarāt paścime bhāge adho nābher nigadyate] LD2,
+\item[udare paścimatānaṃ nābher ūrdhvaṃ ca kārayet] B1, N3, 
+\item[udare paścime yātuṃ nābhiruddhe ca kārayet] J1,
+\item[udare paścimatānena nābher ūrdhvaṃ ca kārayet] C8,
+\item[udare paścime tānaṃ nābher ūrdhvaṃ tu kārayet] N17,
+\item[udare paścimaṃ tānaṃ nābher ūrdhvaṃ tu kārayet] A1, B2, C9, K1, LD1, N9, N21, N22, N24, P5, P6, P8, P9, P13, 
+\item[udare paścimatānaṃ nābher ūrdhve tu kārayet] J17,
+\item[udare paścimatānaṃ nābher ūrdhvaṃ ca kārayet] N10,
+\item[udare paścimottānaṃ nābher ūrddhe tu kārayet] P1,
+\item[udare paścimaṃ tānaṃ nābher ūrddhe tu kārayet] J10,
+\item[udara paścimantānaṃ nābhebher ūrdhaṃ akārayet] N23,
+\item[udara paścimantānaṃ nābhebher ūrdhaṃ tu kārayet] B1,
+\item[udaraṃ paścimaṃ tāṇaṃ nābher ūrdhve tu kārayet] P10,
+\item[udaraṃ paścimaṃ tāṇaṃ nābher ūrdhvaṃ ca kārayet] J14, Ko, 
+\item[paścimaṃ tānam ure (tā)raye dṛḍhatāṃ gate] Ba, 
+  \item[(illegible/unavailable)] Bo1, Bo2, Bo3, C5, J2, J6, N4b, N20, P3,
+
+    \begin{description}
+      Tāna|tāṇa: the latter is because of influence from vernacular pronunciation.
+    • It is given at 95ab in B1
+    • It is given at 95ab in B3
+    \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_062]
+
+  \sthana{3.62b}
+
+\item[guruṇā kathitaṃ yathā] C4, J1, J3, J7, J14, K2, N5, N8, N11, N16, N18, N23, N25, 
+\item[guruṇā kathitaṃ tathā] B2, K1, N22, P6, 
+\item[guruṇā kathitaṃ tadā] Ba5, 
+\item[guruṇā kathitaṃ sadā] A1, B1, B3, C2, C3, C6, C8, C9, J4, J5, J10, J11, J12, J13, J16, J17, Ko, LD1, N3, N4a, N6, N7, N9, N10, N12, N13, N14, N15, N17, N21, P1, P2, P4, P5, P7, P8, P9, P10, P11, P12, P13,
+\item[guruṇā kathita sadā] J15,
+\item[guruṇāṃ kathitaṃ sadā] J8,
+\item[guruṇāṃ kathitaṃ sadāḥ] LD2,
+\item[kathitaṃ guruṇā sadā] N24,
+\item[(illegible/unavailable)] Ba, Bo1, Bo2, Bo3, C5, J2, J6, N4b, N20, P3,
+
+  \begin{description}
+    The word \emph{sadā} (instead of \emph{yathā}) is also well attested.
+However, the first hemistich seems more difficult to understand with  \emph{sadā}.
+Perhaps, the easiest way to make sense of \emph{sadā} is to read it with the verb \emph{abhyaset}; `Even an old man who regularly and tirlessly practises \emph{uḍḍiyāṇa}, which is easy when taught by a guru, becomes young.'
+    • It is given at 95ab in B1
+    • It is given at 96b in B3
+
+  \end{description}
+
+\end{marma}
+
+  \sthana{3.62cd}
+
+\item[abhyaset tad anaṃdas tu vṛddho’pi taruṇāyate ||] N16,
+\item[abhyaset tad atandras tu vṛddho’pi taruṇāyate ||] C4, J1, J3, J7, K2, N5, N11, N25, 
+\item[abhyase tad atandras tu vṛddho’pi taruṇāyate ||] N8,
+\item[abhyāsed astataṃdras tu vṛddho’pi taruṇo bhavet] J5, P6, 
+\item[abhyāsed astatadras tu vṛddho’pi taruṇo bhavet] N3,
+\item[abhyāset satataṃ yastu vṛddho’pi taruṇo bhavet] P1,
+\item[abhyāset dastataṃdras tu vṛddho’pi taruṇo bhavet] N22,
+\item[abhyāsen na taṃdras tu vṛddhāpi taruṇāyate ||] N23,
+\item[abhyāsataḥ svataṃtras tu vṛddho’pi taruṇāyate] A1, C6, C8, P7, 
+\item[abhyased asya taṃtraṃ tu vṛddho’pi taruṇāyate] J11,
+\item[abhyaset satataṃ yas tu vṛddho’pi tarṇāyate] Ba5, N4a, N13, N18,
+\item[abhyased aniśaṃ yogī vṛddho’pi taruṇāyate ||] J14,
+\item[abhyased aniśaṃ yogī vṛddho’pi taruṇo bhavet ||] N21, P13, 
+\item[abhyaset satataṃ yas tu vṛddas tu taruṇo bhavet] LD1,
+\item[abhyaset satataṃ yas tu vṛddho’pi taruṇo bhavet] B1, C2, C3, J10, J13, J17, LD2, N6, N7, N10, N12, N15, N17, P2, P4, P5, P8, P9, P10, 
+\item[abhyaset satataṃ yas tu vṛddho’pi sa taruṇo bhavet] P12,
+\item[abhiset satataṃ yas tu vṛddho’pi taruṇo bhavet] N9,
+\item[abhyased astadataṃdras tu vṛddho’pi taruṇo bhavet] Ko,
+\item[abhyaset satataṃ yas tu vṛddhopi tarūṃṇau bhavet] J15, J16, 
+\item[abhyaset satataṃ yas tu vṛdho’pi taruṇo bhavet] C9,
+\item[abhyasyat satataṃ yas tu vṛddho’pi taruṇaṃ bhavet] B3,
+\item[abhyāse satataṃ yas tu vṛddho’pi taruṇo bhavet] J12, N14,
+\item[abhyāsāt satataṃ yas tu vṛddho’pi taruṇo bhavet] J8, 
+\item[abhyased asvataṃtran tu vṛddho’pi taruṇo bhavet] B2, 
+\item[abhyased asvataṃtras tu vṛddho’pi taruṇo bhavet] J4, K1, P11, 
+\item[abhyased ataṃ saṃyogād vaddhopi taruṇo bhavet] N24, 
+\item[(illegible/unavailable)] Ba, Bo1, Bo2, Bo3, C5, J2, J6, N4b, N20, P3,
+
+  \begin{description}
+The reading \emph{astatandraḥ} is probably the original one in the \emph{Dattātreyayogaśāstra} and is attested in some manuscripts of the \emph{Haṭhapradīpikā}.
+However, it is not strongly supported by the witnesses of the important groups and so may have changed to something simpler, such as \emph{tad atandras tu}.
+    • It is given at 96cd in B1
+    • It is given at 96cd in B3
+
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_064]
+
+  \sthana{3.64ab}
+
+\item[sati vajrāsane pādau karābhyāṃ kāraye dṛḍham] J4,
+\item[sati vajrāsanaṃ pādau karābhyāṃ kāraye dṛḍham] N3,
+\item[sati vajrāsane pādau karābhyāṃ kārayed dṛḍham] N5,
+\item[sati vajrāsane pādau karābhyāṃ dhārayed dṛḍham] A1, Ba5, C2, C4, C6, J3, J14, J17, LD1, N6, N12, N13, N14, N25, P5, P6, 
+\item[sati vajrāsanaṃ pādau karābhyāṃ dhārayed dṛḍham] P11,
+\item[sata bajrāsane pādau karābhyāṃ dhāraye dṛḍham] K2,
+\item[sati vajrāsane pādau karābhyāṃ dhāraye dṛḍham] C3, C8, J1, J7, J11, J13, J15, K1, N7, N9, N15, N17, N18, N21, N22, N24, P1, P2, P4, P8, P9, P10, P12,
+\item[sati vajrāsane pādau karābhyāṃ dhāraye dṛḍhau] N10,
+\item[satī vajrāsane pādau karābhyāṃ dhāraye dṛḍham] LD2,
+\item[sati vajrāsane pādau karābhyāṃ dhāraye dṛḍhai] J12,
+\item[sati vajrāsane pādo karābhyāṃ dhāraye dṛḍham] J8,
+\item[sati vajrāsane pādau karā- sandhāraye dṛḍhe] N23,
+\item[sati vajrāsane pādau karābhyāṃ dhāraye draḍham] C9,
+\item[sati vajrāsane pādau karābhyāṃ kāraṃye draḍham] Ko,
+\item[sati vajrāsane pādau karābhyāṃ dhārayad draḍham] J10,
+\item[śanai vajrāsane pādau karābhyāṃ dhāraye dṛḍham] N8, N11,
+\item[sati vajrāsane pādau karābhyāṃ dhāraye dhruvam] B3, 
+\item[saṃti vajrāsane pādau kadābhyāṃ dhāraye dṛḍhāṃ] P13,
+\item[saṃti vajñāsane pādau kadābhyāṃ dhāraye dṛḍham] J2, P7, 
+\item[saṃti jjāsane pādau karābhyāṃ kāraye dṛḍham] J5,
+\item[saṃsthi vajrāsane pādau karābhyāṃ dhāraye dṛḍham] N16,
+\item[saṃsthita vajrāsane pādau karābhyāṃ dhārayed dṛḍham] B1,
+\item[satīva āsane pad....rābhyaṃ dhāraye dṛḍhaṃ] B2,
+\item[śatri vajrāsane pādau karābhyāṃ dhāraye dṛḍho] J16,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C5, J6, N4a, N4b, N20, P3,
+
+  \begin{description}
+    • It is given at 98ab in B1
+    • It is given at 116ab in Ba5
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_064cd]
+
+  \sthana{3.64cd}
+
+\item[ulphadeśasamīpe ca udaraṃ tat prapīḍayet] J1, J3, N11,
+\item[ulphadeśe samīpe ca udaraṃ tat prapīḍayet] N16,
+\item[gulphadeśasamīpe ca kandaṃ tatra prapīḍayet] B1, J12, K1, Ko, N10, N13, P7, 
+\item[gulphadeśasamīpe ca kaṃda tatra prapīḍayet] P13,
+\item[gulphadesasamīpe ca kandaṃ tatra prapīdayet] N21, 
+\item[gulphadeśasamīpa ca kaṃdaṃ tatra prapīdayat] N23,
+\item[gulphadeśasamīpe ca kaṃpaṃ tatra prapīḍayet] B3, J7, J16, 
+\item[gulphe deśe samīpe ca kaṃdaṃ tatra prapīḍyate] C3,
+\item[gulphe deśe samīpe ca kaṃpaṃ tatra prapīḍayet] C2, P12, 
+\item[gulphadeśe samīpe ca kaṃpaṃ tatra prapīḍayet] J13, P2, P4, 
+\item[gulphadeśe samīpaś ca kandaṃ tatra prapīḍayet] B2,
+\item[gulphadeśe samīpaṃ ca kaṃdaṃ tatra prapīḍayet] J4,
+\item[gulphadeśasamīpaṃ ca kaṃdaṃ tatra prapīḍayet] N5, N25, 
+\item[gulphadeśasamīpe ca gudaṃ tatra prapīḍayet] J14,
+\item[gulphadeśasamīpaṃ ca skaṃdhaṃ tatra prapīḍayet] K2,
+\item[gulphadeśe samīpaṃ ca skaṃdhaṃ tatra prapīḍayet] C4, 
+\item[gulphadeśasamīpe ca kanda tatra prapīḍayet] A1, 
+\item[gulphadeśasamīpe ca kadaṃ tatra prapīḍayet] Ba5, C6, 
+\item[gulphadeśasamīpe ca kaṇṭhaṃ tatra prapīḍayet] N12, P11, 
+\item[gulphadesasamīpe ca kaṃṭhaṃ tacca prapīḍayet] LD2,
+\item[gulmadeśasamīpe ca kaṃṭha tacca prapīḍyate] N7,
+\item[gulphadesasamīpe ca kaṃdaṃ tacca prapīḍyate] C9, N17, P8, P9, 
+\item[gulphadeśasamīpe ca kaṃdaṃ ta prapīḍyate] P10,
+\item[gulphadeśasamīpe ca kaṃdaṃ tacca prapīḍyate] P5,
+\item[gulmadeśasamīpe ca kaṃdaṃ tatra prapiḍayet] N22,
+\item[gulmadeśasamīpe ca kaṃdaṃ tacca prapīḍyate] J17, N6, N9, P1, 
+\item[gulphadeśasamīpe ca kaṃdaṃ tatra prapīḍyate] N15,
+\item[gulphadesasamīpe ca kaṃdaṃ tatra prapīḍyate] J15,
+\item[gulphadeśasamīpe ca kaṃdaṃ tacca prapīḍyate] J8, J10, 
+\item[gulphadeśasamīpe ca kaṃdaṃ tatra prapīḍayet] J11,
+\item[gulphadeśaṃ samīpe ca kaṃdaṃ tava prapīḍyate] N3,
+\item[gulphadeśasamīpe ca kaṃdhan tasya prapīḍyate] N14,
+\item[gulphadeśasamīpe ca kaṃde tacca prapīḍayet] LD1,
+\item[gulphadeśasamīpe ca kaṃ(dha) tatra prapīḍayet] C8, J6, 
+\item[gulphadeśe samāvaṃtu kaṃṭha tatrāpi pīḍayet] J2,
+\item[gulphadeśe samitye vā kaṃṭhaṃ tatra pradāyate] J5,
+\item[gulphadeśasamīpe ca udaraṃ tat prapīḍayet] N8,
+\item[gudadeśasamīpe ca kaṃdaṃ tacca prapīḍayet] N18,
+\item[gulphadeśasamīpe tu kandaṃ tatra prapīḍayet] N24,
+\item[gulpadeśasamipe ca kaṃdaṃ tatra sapīḍayet] P6,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C5, N4a, N4b, N20, P3,
+
+  \begin{description}
+In the \emph{Gorakṣaśataka} this verse comes before any instruction to perform \emph{uḍḍiyāna}, so it appears to refer to pressing the \emph{kanda} with the feet, which is plausible if the \emph{kanda} is located in the gential region.
+However, Sundaradeva and Bhavadevamiśra think the \emph{kanda} is located near the navel.
+In their view, the yogi presses the \emph{kanda} by holding the legs near the ankles with the hands and performing \emph{uḍḍiyānabandha} (at the end of \emph{kumbhaka} and the beginning of exhalation, according to Sundaradeva).
+In this view, it is the backward stretch in the abdomen (\emph{udare paścimatāna}) that presses the \emph{kanda} in the navel.
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_065]
+
+  \sthana{3.65b}
+
+\item[upīpa? hṛdaye gale] N15,
+\item[kāraye hṛdayaṃ gataḥ] J4,
+\item[kārayed dhṛdaye gale] A1,  
+\item[kārayed hṛdaye gale] B2, C6, J5, J7, J11, LD1, LD2, N3, N6, N12, N17, N21, N22, N24, P7, P11, 
+\item[kāraye hṛdaye gale] C8, J8, K1, P13, 
+\item[kārayed hṛdaye gataiḥ] N23,
+\item[kārayedradaye gale] P6,
+\item[cibukaṃ hṛdi] J1,
+\item[kārayec cibukaṃ hṛdi] B1, C2, C4, J3, J13, J14, K2, N5, N8, N11, N16, N18, N25, P2, P12, 
+\item[kāraye cibukaṃ hṛdi] B3, Ko, 
+\item[kārayec cubukaṃ hṛdi] P4,
+\item[kāraye dṛḍhaye nare] N14,
+\item[pīḍayetadaye gate] J12,
+\item[pīḍayeddhadaye gate] J16,
+\item[pīḍayed hṛdaye gate] N10,
+\item[pīḍayed radaye gale] P8,
+\item[pīḍayed hṛdaye gale] C3, J10, J17, N7, N9, P1, P5, P9, P10, 
+\item[pīḍayed hṛdaye galaṃ] J15,
+\item[pīḍaye hṛdaye gale] C9,
+\item[(illegible/unavailable)] Ba5, Bo1, Bo2, Bo3, C5, J2, J6, N4a, N4b, N13, N20, P3,
+
+  \begin{description}
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_065]
+
+  \sthana{3.65d}
+
+\item[ūrddhvaasaṃdhiṃ na gachati] J11,
+\item[tundasaṃdhiṃ na gacchati] J4, P11, 
+\item[tundasaṃsi na gachati] N22,
+\item[tundasaṃdhiṃ na gachati] J7, N3, 
+\item[(khedaṃ) saṃdhiṃ na gacchati] J1,
+\item[daṃtaṃ gaṃdhi nigachati] N15,
+\item[datta siddhir nigacchati] P7,
+\item[daṃtaṃ gaṃdhiṃ nigacchati] P8,
+\item[daṃtasaṃdhiṃ na gacchati] C9,
+\item[daṃtaṃgaṃdhi nigacchati] P9,
+\item[tanosiṃdhiṃ na gachati] Ko,
+\item[tudasadhi na gachati] N11,
+\item[tudan sādhiṃ na gachati] B2,
+\item[taṃda saṃdhiṃ na ga] N23,
+\item[dasaṃdhiṃ nā gachati] J8,
+\item[tuṇḍasaṃdhiṃ na gacchati] N12,
+\item[tundasiddhiṃ nigacchati] A1, C6, 
+\item[tudasadhi na gachati] N8,
+\item[tudasaṃdhi na gachati] P6,
+\item[(ruṃṭa)siddhaṃ ca gacchati] C8,
+\item[tathā saṃdhiṃ na gachati] N24,
+\item[tadā saṃdhiṃ ca gacchati] C4, K2, N5, N25, 
+\item[nāḍīsaṃdhiṃ na gacchati] C3, J11a, N21, P5, 
+\item[nāḍīsaṃdhi na gachati] J13, N9, P10, P13, 
+\item[nāḍīsiddhiṃ ca gachati] N7,
+\item[nāḍīsiddhiṃ nigachati] P1,
+\item[nāḍīsiddhiṃ na gachati] LD1,
+\item[nāḍīsaṃdhi nigacchati] J10, J17, 
+\item[nāḍīsiddhi niyacchati] J12,
+\item[nāḍīsidhiṃ niyachati] J
+\item[nāḍisaṃdhi nīyacchatī] J16,
+\item[nāḍīśuddhiṃ nigachati] LD2,
+\item[rudasaṃdhiṃ na gachati] K1,
+\item[skaṃdasaṃdhi na gachati] J5,
+\item[skandhasaṃdhin na gacchati] N16,
+\item[skandasaṃdhiṃ na gacchati] J3,
+\item[skandasaṃdhiṃ nirgachati] J14,
+\item[skandasaṃghena gacchati] B1,  
+\item[skandasaṃgena gachati] B3, C2, 
+\item[tadā gaṃ?dhiṃ pragacchati] N6,
+\item[tadā gaṃdhiṃ pragachati] N17,
+\item[uttamodyadyānavandhanāt] N14,
+\item[ṣedaṃ saṃdhi na gakṣati] N18,
+\item[kaṃdasagena gachati] P2,
+\item[kaṃdasaṃgena gachati] P4, P12, 
+\item[(illegible/unavailable)] Ba5, Bo1, Bo2, Bo3, C5, J2, J6, J15, N4a, N4b, N13, N20, P3,
+
+  \begin{description}
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_076]
+
+  \sthana{3.76b}
+
+\item[mukhabandhanam] A1, B1, B2, B3, C2, C4, C6, C8, C9, J1, J2, J3, J6, J7, J8, J10, J11, J12, J13, J14, J15, J16, K1, LD1, LD2, N4a, N5, N7, N8, N10, N11, N12, N16, N18, N21, N22, N23, N25, P2, P4, P6, P7, P10, P11,
+\item[mukhavañcanam] Jyo, N4b, N13, 
+\item[mukha(caṃ)canam] Ba5,
+\item[mukhamaṃdhavaṃ] J5, 
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C3, C5, J4, J17, K2, Ko, N3, N6, N9, N14, N15, N17, N20, N24, P1, P3, P5, P8, P9, P12, P13,
+
+  \begin{description}
+    • It is given at 111c in B1 
+    • It is given at 106b in C2
+    • It is striked in N4b
+No source? Only Brahmānanda reads mukhavañcanam, which makes better sense.
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_081]
+
+  \sthana{3.81b}
+
+\item[ṣaṇmāsārdhena naśyati] J1, J3, J4, J6a, J7, K1, N5, N8, N16, N18, N23, N25, V19
+\item[ṣaṇmāsārdhaṃ praṇasyate] N22,
+\item[ṣaṇmāsārdhena paśyati] J6,
+\item[ṣanmāsānna ca  dṛśyate] C9,
+\item[mūrdham ākṛṣya rakṣayet] B2,
+\item[ṣaṇmāsordhvaṃ na dṛśyate] Ba5, N4b,
+\item[ṣaṇmāsārddhe na dṛśyate] J5,
+\item[ṣaṇmāsārddhaṃ na dṛśyate] J8, N11, N24, 
+\item[ṣaṇmāsārdhān na dṛśyate] A1, C6, C8, J11, J14, N3, N9, P7, P11, 
+\item[ṣaṇmāsārdhaṃ na dṛśyati] Ko,
+\item[ṣaṇyāsorddhaṃ na drasyate] J2,
+\item[ṣaṇmāsorddhvaṃ pranasyate] P6,
+\item[ṣaṇmāsa tu na dṛśyate] N14,
+\item[ṣaṇmāsā tu na dṛśyate] C3, J16, 
+\item[ṣaṇmāsāt tu dṛśyate] J15,
+\item[ṣaṇmāsāṃ tu na dṛśyate] N7,
+\item[ṣaṇmāsān naiva dṛśyate] N12,
+\item[ṣaṇmāsāt tu na dṛśyate] B1, B3, C2, J10, J12, J13, J17, LD1, LD2, N6, N10, N15, N17, N21, P1, P2, P4, P5, P9, P10, P12, P13, 
+\item[parāmāsāt tu na dṛśyate] P8,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C4, C5, K2, N4a, N13, N20, P3,
+
+  \begin{description}
+    • It is given at 116b in B1
+    • It is given at 111b in C2
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_082]
+
+  \sthana{3.82d}
+
+\item[siddhabhājanaṃ] LD2,
+\item[siddhibhājanaṃ] B1, B2, B3, Ba5, C2, C6, C8, J1, J3, J4, J6, J7, J8, J15, N4b, N8, N9, N23, V19, C7, K3, Ko,  N19, V15, J11, J13, K1, N3, N11, N12, N16, N18, N21, N22, N24, N25, P2, P4, P6, P10, P11, P13, P23, V1, V3, Jyo
+\item[siddhibhājana] J5,
+\item[siddhibhājanaḥ] J14, P7, 
+\item[vidhibhājanaṃ] J2,
+\item[siddhibhāg bhavet] J12,
+\item[siddhimān bhavet] A1, C3, C9, J10, J17, LD1, N6, N7, N10, N14, N15, N17, P1, P5, P8, P9, 
+\item[sīddhimān bhavet] J16,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C4, C5, K2, N4a, N5, N13, N20, P3, P12,
+
+  \begin{description}
+    • It is noted in C4 that “vajrolī graṃthāṃtare likhitāsīt”
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_090]
+
+  \sthana{3.90b}
+
+\item[bījaṃ] V19, N19, V15, J11, V1, J10, J17, P1, P8, 
+\item[bijaṃ] J16, LD2,
+\item[bīnduṃ] B2, C6, K1, P7,
+\item[vindaṃ] J2,
+\item[bīndu] C8,
+\item[dhanāṃ] A1, 
+\item[jīvaṃ] C7, P23
+\item[nijaṃ] Ba5, N4b, N13, Jyo
+\item[bījabiṃdu] P9,
+\item[bījabinduṃ] C3,
+\item[bījaṃ binduṃ ca] J1, J3, J6, J12, J14, LD1, N5, N8, N6, N9, N11, N14, N16, N17, P5, 
+\item[bījaṃ bindu ca] N10,
+\item[svīyaṃ bindu ca] P11,
+\item[rajaṃ bīnduṃ ca] P10,
+\item[stribaṃdu caiva] N22,
+\item[striṃ biṃduṃ ca] P6,
+\item[striyaṃ biṃduṃ ca] N12,
+\item[striyo biṃduś ca] N7,
+\item[striyā biṃduṃ ca] N23,
+\item[jayaṃ bindu ca] J8,
+\item[bīrjaṃ binduṃ ca] C9,
+\item[jīvaṃ bindu ca] N18,
+\item[vīryaṃ binduṃ ca] J5,
+\item[nijabinduṃ] B1, B3, C2, Jyo, N21, P2, P4, P12, P13, 
+\item[nījaṃ binduṃ ca] J13,
+\item[nijaṃ binduṃ ca] J15,
+\item[nijaṃ bindu] J4,
+\item[jayaṃ] V3
+\item[striyā] J7,N23
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C4, C5, K2, Ko, N3, N4a, N15, N20, N24, N25, P3,
+
+  \begin{description}
+    • It is given at 126b in B1 
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_092]
+
+  \sthana{3.92c}
+
+\item[jaleṣu bhasma] B2, Ba5, C3, C9, J1, J3, J6, J7, J10, J11, J12, J14, J15, J16, J17, K1, Ko, LD1, LD2, N5, N6, N8, N7, N9, N10, N11, N13, N14, N16, N17, N18, P1, P5, P6, P7, P8, P9, P10, 
+\item[jalesu bhasma] N4b,
+\item[jaleṣu bhaśma] J2,
+\item[jale bhasma] B1, B3, C2, J4, J13, N21, P4, P12, 
+\item[jale tu bhasma] A1,  
+\item[taleṣu tasma] N22,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C4, C5, C6, C8, J5, J8, K2, N3, N4a, N12, N15, N20, N23, N24, N25, P2, P3, P11, P13,
+
+  \begin{description}
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_096]
+
+  \sthana{3.96ab}
+
+\item[amarīṃ yaḥ piben nityaṃ nasyaṃ kurvan dine dine] Ba5, C6, C8, J11, N13, P7,
+\item[amarīṃ yaḥ piven nityaṃ nasyaṃ kurvan dine dine] N4b,
+\item[amarīṃ yaḥ piben nityaṃ naśyaṃ kurvan dine dine] J14,
+\item[amarīṃ yo pive nityaṃ nasyaṃ kuryād dine dine] N17,
+\item[amarī yaḥ pibe nityaṃ nasya kuryā dine dine] J5,
+\item[amarāvyapiva nityaṃ nasyaṃ kurvan dine dine] N22,
+\item[amarī ya piben nityaṃ naśyaṃ kurvaṃ dine dine] J1,
+\item[amarī yaḥ piven nityaṃ nasyaṃ kurvana dine dine] J8,
+\item[amarī yaḥ pīven nītyaṃ nasyaṃ kuryā dīne dīne] LD2,
+\item[amarī yaḥ piben nityaṃ naśya kurvvaṃ dine dine] J2,
+\item[amarī yaḥ ṣpiven nityaṃ naśyaṃ kurvva dine dine] N18,
+\item[amarī yo piben nityaṃ nasyaṃ kuryād dine dine] P9,
+\item[amarī yo piven nityaṃ sammyak kuryād dine dine] N14,
+\item[amarī yaḥ piven nityaṃ taśya kurvan dine dine] P11,
+\item[amarī yaṃ piven nityaṃ paśyan kurvan dine dine] N12,
+\item[amarīṃ yaḥ piben nityaṃ tasya kurvan dine dine] A1, J3, J6, 
+\item[amaṃrī yaḥ piven nityaṃ tasya kuryā dine dine] C3, J7,
+\item[amaṃrī yā piben nityaṃ tapaḥ kuryā dine dinai] J15,
+\item[amarīṃ yo piven nityaṃ tasyaṃ kuryā dine dine] P10,
+\item[amarīṃ yo piven nityaṃ tasyaṃ kuryyād dine dine] J10,
+\item[amarīṃ yaṣ piven nityaṃ tasya kuryyād dine dine] P5,
+\item[amarīṃ yo piben nityaṃ tasya kuryyād dine dine] LD1, N9,
+\item[amarīṃ yo piben nityaṃ tasyāṃ kuryyād dine dine] P1,
+\item[amarī yo pive nityaṃ tasya kuryā dine dine] P8,
+\item[amarīṃ yaḥ piben nityaṃ vasyaṃ kuryyād dine dine] B1, 
+\item[yamarī yo piben nityaṃ tasyāṃ kuryyād dine dine] B2,
+\item[aamarīṃ yo piven nityaṃ nasyaṃ kuryād dine dine] N6,
+\item[amarīṃ yo piven nityaṃ tasyaṃ kuryā dine dine] J17,
+\item[ammuṃ yogī pibe nityaṃ naśyaṃ kuryā dine dine] B3, C2, 
+\item[amarīṃ yaḥ pive nityaṃ nasyaṃ kuryā dine dine] C9,
+\item[amarīṃ yaḥ piven nityaṃ naśyaṃ kuryād dine dine] Ko,
+\item[amarī yaḥ piven nityaṃ nasyaṃ kuryād dine dine] N7,
+\item[amarīṃ jaḥ piven nityaṃ nāśāraṃdhrād dine dine] J10a,
+\item[amarīṃ yaḥ piben nityaṃ kriyāṃ kurvan dine dine] J6a,
+\item[amarīṃ yaḥ piben nityaṃ vaśyaṃ kuryā dine dine] J4, J13, N21, 
+\item[amarī yaḥ piben nityaṃ vaśyaṃ kuryā dine dine] P13,
+\item[amarī yeḥ piben nityaṃ sa bhavet makaradhvajaḥ] J12,
+\item[amariṃ yaḥ pibin nityaṃ sa bhavet makaṃradhvaja] J16,
+\item[amariṃ yaḥ pibin nityaṃ saṃ bhavet makaṃradhvajaḥ] N10,
+\item[aṃvarī ca piven nityaṃ naśyaṃ kurvan dine dine] N16,
+\item[amuṃ yogī piben nityaṃ nasyaṃ kuryād dine dine] P12,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C4, C5, K1, K2, N3, N4a, N5, N8, N11, N15,
+ N20, N23, N24, N25, P2, P3, P4, P6,
+
+  \begin{description}
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_099*1]
+
+  \sthana{3.99*1 additional line}
+
+\item[meḍhram ākuṃcanād ūrdhvaṃ rajasāpi hi yoginī] J17, N17,
+\item[meḍhram akuṃcanād ūrdhvaṃ rajasyāpi hi yoginī] N9, P1,
+\item[meḍhrām ākuṃcanād ūrdhvaṃ rajasāpi hi yoginī] J6, N6,
+\item[meḍhrām ākuṃcanād ūrdhvarajasāpi hi yoginī] P10,
+\item[meḍhrād ākuṃcanād ūrdhvaṃ rajasāpi hi yoginī] J3, J6a,
+\item[meḍhram ākuṃcayed ūrdhvaṃ rajasāpi hi yoginī] C3, J12, 
+\item[meḍhram ākuṃcayed ūrdhvaṃ rajaḥsāpi hi yoginī] N10,
+\item[meḍhram ākuṃcanād ūrdhvaṃ rajasāpi hi yogīnī] J15,
+\item[meḍhrām ākuṃcanād ūrdhvaṃ rajasāpi hi yoginī] J10, 
+\item[meḍhrasyākuṃcanād ūrdhvaṃ rajasyāpi hi yoginī] P5,
+\item[meḍyam ākuṃcanād ūrdhvaṃ rajasopi hi yoginī] J11,
+\item[meḍhreṇākuñcanād ūrdhvaṃ rajasāpi hi yoginī] A1,
+\item[meṃḍham ākuṃcayed ūrdhvaṃ rarajasārpi hi yogīnīḥ] J16,
+\item[meḍhrāeṇākarṣayed ūrdhvaṃ samaggamaṅi pā(ṭa)vāt] C8,
+\item[rakṣed ākuṃcanād ūrdhvaṃ yā rajaḥ sa hi yoginī] N13,
+\item[mehanākuṃcatād ūrdhvaṃ rajasāpi ca yoginī] N16,
+\item[mehanākuñcanād ūrdhvaṃ rajasāpi ca yoginī] B1, J4, J13, J14, 
+\item[mehanākuñcanād ūrdhvaṃ rajasāpī ca yoginī] N12,
+\item[mehanākuñcanād ūrdhvā rajasāpī hi yoginī] P8,
+\item[mehanākuñcanād ūrdhvaṃ rajasāpi hi yoginī] K1, N7, N21, P7, P9, P13, 
+\item[mehanākuṃcanād ūrdhvaṃ na yasyāpi hi yoginī] C2,
+\item[mehenākuñcanād ūrddhaṃ rajasāpi hi yoginī] B2,
+\item[mehenākucanād ūrddhva rajasāpi hi yogīnā] N22,
+\item[meḍhrmākuñcanād ūrddhvaḥ rajasyāpi hi yoginī] N14,
+\item[mehenākuñcanād ūrddhv rajasāpi hi yoginī] J7,
+\item[meruākuñcanād ūrdhvaṃ nayaty apī hi yogīnī] B3, 
+\item[mehanākuñcanād ūrdhvaṃ na yasyāpi hi yoginī] P2, P4, P12, 
+\item[rakṣedākuñcanād ūrdhvaṃ yā rajaḥ sā hi yoginī] N4b, P6, 
+\item[(illegible/unavailable)] Ba5, Bo1, Bo2, Bo3, C4, C5, C6, C9, J1, J2, J5, J8, K2, LD1, LD2, N3, N4a, N5, N8, N11, N15, N18, N20, N23, N24, N25, P3, P11, 
+  \begin{description}
+  \item[Yes] A1; B1; B2; J1; J6; J7; J11; J10; Jyo; P2;
+    \item[No] Ba5, C9; J2; J5; K2; Ko; LD1; LD2; N3; N4a; N5; N8; N10; N11; N15; N20; N23; N24; N25; P3;
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_099]
+
+  \sthana{3.99b}
+
+\item[svadehajau] B1, C9, J1, J2, J4, J7, J6, J8, K1, Ko, LD1, V1, V3, N10, V15, J11, J13, J14, N7, N12, N16, N18, N21, P1, P11, 
+\item[svadehaje] P6,
+\item[svadehajo] LD2,
+\item[svadehagau] Jyo, N4b, N13, 
+\item[svadehajaḥ] B3, C2, C8, P2, P4, P12, 
+\item[svadehajaiḥ] A1, C3, C6, J10, J12, J15, J16, J17, N6, N9, N17, P7, P8, P9, P10, 
+\item[svadehijaḥ] P5,
+\item[svadehaya] N14,
+\item[svadehajaṃ] C7, J5, 
+\item[svadehijau] P13,
+\item[svadehi(nau)] Ba5,
+\item[sadehajaṃ] V19
+\item[sadehabhājāṃ] P23
+\item[śca dehajaiḥ] B2,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C4, C5, J3, K2, N3, N4a, N5, N8, N10, N11, N15, N20, N22, N23, N24, N25, P3,
+
+  \begin{description}
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_101]
+
+  \sthana{3.101b}
+
+\item[bhogabhuktivimuktidaḥ] B1, C2, J4, J13, LD1, P4, 
+\item[bhogamuktinabhuktida] N22,
+\item[bhogībhuktivimuktibhiḥ] K1,
+\item[bhoge mukti vimuktiha] J16,
+\item[bhogamuktivimuktidaḥ] N17, P2, P5, P8, P9, 
+\item[bhogamuktivimuktidaṃ] N14,
+\item[bhogamuktavimuktidaḥ] P1,
+\item[bhogabhuktivimuktadaḥ] C9,
+\item[bhogayuktavimuktidaḥ] LD2,
+\item[bhogair yuktetha muktidaḥ] N12,
+\item[bhoge muktepi muktitaḥ] P13,
+\item[bhoge bhukte ca muktidaḥ] J1, N18,
+\item[bhoge bhuktepi muktidaḥ] J7, J8, N4b, N13, N16, N21, P6, P11, 
+\item[bhoge bhuktopi muktida] J15, N9,
+\item[bhojayuktepi muktidaḥ] J11,
+\item[bhogabhuktepi muktidaḥ] P10,
+\item[bhogabhuktepi muktidā] Ko,
+\item[bhogabhukte vimuktidaḥ] J10,
+\item[bhoge mukte vimuktidaḥ] P12,
+\item[bhoge mukte ca muktidaḥ] J3, J6, 
+\item[bhogayogepi muktidaḥ] C6, 
+\item[bhoge yogepi muktidaḥ] C8,
+\item[bhogamuktivimuktidaḥ] B2, C3, B3, J2, J12, J17, N6, N7,
+\item[bhogo muktepi muktidaḥ] Ba5,
+\item[bhogayukte’pi muktidaḥ] A1, 
+\item[bhogayukto’pi muktidaḥ] J14, P7,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C4, C5, J5, K2, N3, N4a, N5, N8, N10, N11, N15, N20, N23, N24, N25, P3,
+
+  \begin{description}
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_110]
+
+  \sthana{3.110}
+
+\item[paristhitā caiva phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prasārya sā cārpy avidhānayuktyā 
+pragṛhya nityaṃ paricālanīyā ||] C6,
+
+\item[paristhitā caiva phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prasārayet tat paridhānayuktyā
+pragṛhyate tat paripālanīyā ||] LD1,
+
+\item[paristhitāṃ caiva phaṇāvatī sā 
+prātaś ca sāyaṃ praharārdhamātraṃ ||
+prātar ya sūryāyavidhānayuktā
+pragṛhya paripālanīyā ||] P7,
+
+\item[avasthitasyāṃ phaṇārpayatī tāṃ
+prātaś ca sāyaṃ prarārddhamātraṃ 
+prayārya yāttat paridhānamuktā
+pragṛhyate tat paricālanīyā] P10,
+
+\item[paristhitā caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ ||
+prapurya suryāt paridhānayuktā
+pragṛhyato yat paricālanīyā ||] P6,
+
+\item[avasthitasthaṃdaphaṇārpayaṃtī
+prātaś ca sāyaṃ praharārdhamātraṃ ||
+prasāryatte tat paridhānamuktā
+pragṛhyate tat paripālanīyā ||] P9,
+
+\item[paristhitā caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ
+prapūrya sūryāparidhānamuktā
+pragṛhya niryāty avicāni sā] N25,
+
+\item[paristhitā caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddha mātraṃ
+prapūrya sūryā paridhārna muktā
+pragṛhyato yatparicālanīyā] N22,
+
+\item[paristhitā caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ 
+prasāryya yat tat paridhāya yuktyā
+pragṛhyate tat paricālanīyā] N18,
+
+\item[paristhitā caiva phaṇāvatī sā 
+prātaś ca sāyaṃ praharārddhamātraṃ
+prapūryya sauryyāt paridhāyamuktā
+pragṛhya niryāti vicālitā sā] N16,
+
+\item[avasthitasyāṃdaphaṇārpayaṃtī tāṃ
+prātaś ca sāyaṃ praharārddhamātraṃ
+prasāhya yaṃ tatparidhānamuktā
+yad gṛhyate tat paripālanīyā] P8,
+
+\item[praviśya tasyaiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ |
+prapūrya sūryātmavidhānayuktā
+pragṛhya tiryak pravicālitā sā ||] N12,
+
+\item[śayyāsanasthāsya phaṇāvatī sā 
+prātaś ca sāyaṃ praharārdhramātraṃ ||
+prapūrya śauryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā ||] P2,
+
+\item[śayyāsanasthāsvaphaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ ||
+prapūrya śauryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā ||] P12,
+
+\item[avasthita syā phaṇārpayaṃtī tāṃ
+prātaś ca sāyaṃ praharārddhamātraṃ
+prasārya yastāt paridhānamuktā
+pragṛhyate tatparicālanīyā ||] N9,
+
+\item[paristhitā caiva phaṇāvati sā 
+prātaś ca sāyaṃ praharārddhamātraṃ
+prasārya yat tat paridhānayuktyā
+pragṛhyate tat paripālanīyā||] LD2,
+
+\item[avasthitā saiva phaṇāvatī sā 
+prātaśca sāyaṃ praharārddhamātraṃ
+prasārya yāvat paridhānamuktā
+pragṛhya niryāti bicālitī sā] J15,
+
+\item[avaschitasyāṃ phaṇārpayaṃtī tāṃ
+prātaś ca sāyaṃ praharārddhamātraṃ ||
+prasārya yāvat paridhānamukta
+pragṛhyate tat paridhānanīyā ||] P1,
+
+\item[avasthitā caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ
+prapūya sūryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā] N13,
+
+\item[avasthitā caiva phaṇāvati tāṃ
+prātaś ca sāyaṃ praharārddhamātraṃ 
+prasāryya sūryāt paridhānamuktā 
+pragṛhyate te tatparicālanīyā] N17,
+
+\item[avasthitā caiva phanāvatin tāṃ
+prātaś ca sāyaṃ praharārddhamātraṃ
+prasāryya sūryyā paridhānamuktā
+pragṛhyate tat paricālanīyā] N14,
+
+\item[avasthitasya phaṇārpyayantīyaṃ ||
+prātaś ca sāyaṃ praharārddhamātraṃ
+prasārya sūryāt paridhānayuktyā |
+pragṛhya tyakta paricālanīyāṃ] J17,
+
+\item[avasthitā caiva phaṇāvatīṃ tāṃ
+prātaś ca sāyaṃ praharārddhamātraṃ |
+prasāryya sūryāt paridhānamuktā
+pragṛhyate tat paricālanīyā ||] N6,
+
+\item[avasthitā caiva phaṇārpyaṃatīṃ 
+prātaś ca sāyaṃ praharārddhamātraṃ |
+āpūrya sūryāt paridhānamuktā
+prasahyate tat paricālanīyā ||] N10,
+
+\item[paristhitā caiva phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prapūrya sūryāt paridhānayuktā
+pragṛhya nityaṃ paricālanīyā ||] J11,
+
+\item[avasthitasyā phaṇārppayaṃtī tāṃ
+prātaś ca sāyaṃ praharārddhamātraṃ ||
+prasāryya yāttatparidhānamuktā
+pragṛhyate tat paricālanīyāḥ ||] P5,
+
+\item[saṃprasthitā caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ ||
+prapūrya sūryāt paridhānayuktyā
+prasāhaṇī yā paricālanīyā ||] N7,
+
+\item[paristhitā caiva phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prapūrya sūryāt paridhānayuktā
+pragṛhyatīr yat paricālanīyā ||] ko,
+
+\item[paristhi caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ
+prapūrya sūryāt paridhānayuktyā
+pragṛhya niryāt paricālanīyā] K1,
+
+\item[paristhitā caiva phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prapūrya sūryāt paridhānamuktā
+pragṛhya niryāti vicālitā sā ||] N11,
+
+\item[avasthitā syāt phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātraṃ
+prasahyate tat paricālanīyā
+pragṛhya niryāti bicālati sā] J16,
+
+\item[avasthitasyaṃda phaṇāvatī tāṃ
+prātaḥ sāyaṃ praharārddhamātraṃ ||
+prasārya yat tat pradhānamuktā
+pragṛhya tatparicālanīyā ||] N15,
+
+\item[paristhitā caiva phaṇāvatī sā 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prapūrya sūryāt paridhānayuktā 
+pragṛhya nipītyavicālinī sā ||] J7,
+
+\item[avasthitasyaṃdaphaṇārpayāṃtī
+prātaś ca sāyaṃ praharārddhamātraṃ ||
+prasāryarpāṣṇāt paridhānayuktyā
+pragṛhyateta parivālanīyā ||] J10,
+
+\item[avasthitasyaṃdaphaṇārpayāṃtī
+prātaś ca sāyaṃ praharārddhamātraṃ ||
+prasāryasūrjāt paridhānayuktyā
+pragṛhyateta parivālanīyā ||] J10a,
+
+\item[pṛṣṭhisthitasyaiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārdharātram |
+prapūryyasūryyāt paridhāyayuktā
+pragṛhya niryyāt paricālanīyā ||] J8,
+
+\item[pravistṛtasyavaphaṇāvatī sā
+prātaś ca sāyaṃ praharārddharātraṃ
+prasāryya sūryātmavidhānayuktyā
+pragṛhya niryāt paricālanīyā] P11,
+
+\item[paristhitā caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārdharātram |
+prapūrya sūryāt paridhīyayuktā
+pragṛhya niryāti vicālitā sā ||] J14,
+
+\item[paristhitā caika phaṇāvatī sā, 
+prātaś ca sāhaṃ praharārdhamātraṃ | 
+prapūrja śauryāt paridhāyamuktā 
+pragṛhya niryāti vicālitā sā ||] J3,
+
+\item[paristhitā caika phaṇāvatī sā, 
+prātaś ca sāhaṃ praharārdhamātraṃ | 
+prasūrya sauryāt paridhāyamuktā 
+pragṛhya niryāti vicālitā sā ||] J6,
+
+\item[paristhitā caika phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prapūrya śauryāt paridhānamuktā 
+pragṛhya niryāti vicālitā sā ||] J6a,
+
+\item[paristhitā caiva phaṇāvatī sā 
+prātaś ca sāyaṃ harārddhamātraṃ ||
+prapuryāt paridhāyayuktāṃ 
+pragṛhya nirghātivicāritāsāṃ ||] J1,
+
+\item[paristhitā caiva phaṇāvatī sā, 
+paristhitāyaṃ praharārdhamātraṃ | 
+prasārya sūryā(rpya) avidhānayuktyā 
+yat gṛhya nityaṃ paricālanīyā ||] C8,
+
+\item[paristhitā caiva phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prapūrya saudā paridhāna muktā, 
+pragṛhya niryāty avicālinī sā ||] C4,
+
+\item[paristhītā caiva phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prasūrya sauryā paridhāna muktā, 
+pragṛhya niryāty avicālinī sā ||] N5,
+
+\item[paristhītā caiva phaṇāvatī sā, 
+prātaś ca sāyaṃ praharārdhamātraṃ | 
+prapūrya suryā paridhānamuktā, 
+pragṛhya niryāti vicālitā sā ||] N8,
+
+\item[praviśyataviśya tasyaiva phaṇāvatī sā
+prātastvasāyaṃ praharārdhamātraṃ ||
+māpūryasūryāt paridhānivuktā
+pragṛhyatīrthāt paripālanāya] J2,
+
+\item[samyāsanasthasya phaṇāvatī sā 
+prātaś ca sāyaṃ prahārārddhamātraṃ ||
+prapūrya sauryāt paridhānayuktyā
+pragṛhya niryātyavicālitā sā ||] J4, J13,
+
+\item[avasthitā styāt phaṇam arpayantī
+prātaś ca sāyaṃ praharārddhamātram 
+prasāryya yā tat paridhānamuktā
+pragṛhyate tat paricālanīyā ] C3,
+
+\item[avasthitāṃ spaṃda phaṇāvatīṃ tāṃ
+prātaś ca sāyaṃ praharārdhamātraṃ 
+prasārya yatnāt paridhāna(yuktyā)
+pragrahvate tatparicālanīyā] C9,
+
+\item[avasthitā syāt phaṇamṛpiyaṃtī 
+prātaś ca sāyaṃ praharārddhamātraṃ
+āpūrya sūryāt paridhānayuktā
+pravakṣyate tatparimelanīyā] J12,
+
+\item[paristhitā caiva phaṇāvatī sā 
+prātaś ca sāyaṃ praharārdhamātram | 
+prasārya sūryāya vidhānayuktā 
+pragṛhya niryaṃ paricālanīyā ||] A1,
+
+\item[paristhitā caiva phaṇāvatī sā 
+prātaś ca sāyaṃ praharārdhamātram | 
+prapūrya sūryāt paridhānayuktyā 
+pragṛhya niryāt paricālanīyā ||] B2,
+
+\item[śaivāsanasthāsya phaṇavatī sā 
+prātaś ca sāyaṃ praharārddhamātraṃ |
+prapūrya śauryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā ||] B3,
+
+\item[śayyāsanasthasya phaṇāvatī sā
+prātaś ca sāyaṃ praharārtdhamātraṃ
+prapūrya sūryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā] P13,
+
+\item[śayyāsanasthāsya phaṇavatī sā 
+prātaś ca sāyaṃ praharārddhamātraṃ |
+prapūrya śauryāt paridhānayuktā
+pragṛhya nityaṃ paricālanīyā ||] C2,
+
+\item[śayyāsanasthāsya phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamatraṃ ||
+prapūrya śauryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā ||] P4,
+
+\item[avasthitā caiva phaṇāvatī sā
+prātaś ca sāyaṃ praharārddhamātram ||
+prapūrya sūryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā ||] Ba5, N4b,
+
+\item[śayyāśanasthasya phaṇāvatī sā
+prātaś ca sāyaṃ prahārddhamātram
+prasūryasūryāt paridhānayuktyā
+pragṛhya nityaṃ paricālanīyā] N21,
+
+\item[(illegible/unavailable)] B1, Bo1, Bo2, Bo3, C5, J5, K2, N3, N4a, N20, N23, N24, P3,
+
+  \begin{description}
+    • It is given at 114 in C2
+    • It is given at 137 in J1
+We need this verse from every mss.
+
+    No known source.
+Sundaradeva makes sense of this verse by equating the tongue with kuṇḍalinī.
+This enables him to understand the reference to the cloth (\emph{paridhāna}) as the technique of wrapping the tongue in a cloth and milking it, which is a practice associated with khecarī mudrā.
+This interpretation also makes sense of the next verse in the \emph{Haṭhapradīpikā}, which describes the cloth.
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_114]
+
+  \sthana{3.114}
+
+\item[nāsādakṣiṇavāhimārgapavano ghrāṇena dīrghī kṛtaḥ
+candrāmbhaḥ paripūritāmṛtatanuḥ prā ghaṇṭikāyās tathā/
+(chaṃdaḥ) kālaviśālavahnipavanān bhrūrandhranāḍīgaṇān
+tatkāyaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat//] C6,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghīkṛtaś 
+caṃdrāṃtaḥ paripūritāmatatanuḥ prākghaṃṭikāyās tathā ||
+bhiṃdatkālaviśālavanhivaśagān bhrūraṃdhranāḍīgaṇān
+tatkāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat ||] P2,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghīkṛtaś
+caṃdrāṃtaḥ paripuritāmṛtatanuḥ prāgghaṃṭikāyās tathā
+bhiṃdatkālaviśālavanhivaśagān bhrūraṃdhranāḍīgaṇān
+tatkāryaṃ kurute punarnavataraṃ jīrṇadrumaskaṃdhavat ||] P4,
+
+\item[nāsādakṣiṇamārgavāhipavanaḥ prāṇotidīrghikṛtaḥ
+ścaṃdrābhaḥ paripūritāmṛtatanuḥ prākghaṃṭikāyāstathā ||
+chiṃdan kālaviśālavahnivaśagāt bhrūraṃdhranāḍīgaṇāt
+taṃ kāyaṃ kurute pūnar navatanuṃ jīrṇakramaḥ skaṃdhavat ||] P6,
+
+\item[nāsādakṣiṇamārgavāhipavanāṃ prāṇebhidīrghaśkṛteś
+caṃdrāṃbhaḥ paripūratorimṛtatanuḥ prāgghaṃṭikāyās tathā ||
+chinnaṃ kālaviśālavanhivaśagaṃ prāgraṃdhanāḍīgaṇā
+tatkāyaṃ kurute punar navataraṃ chinnadrumaskaṃdhavat ||] P9,
+
+\item[nāsādakṣiṇamārgavāhipavanaḥ prāṇotidīrghākṛtaś
+caṃdrābhaḥ paripūritāmṛtanuḥ prāgghaṃṭikāyās tathā ||
+chidat kālaviśālavahnivaśagaṃ bhrūraṃdhranāḍīgaṇās
+tatkayaṃ kurute punarnavataraṃ chinnadramaskaṃdhavat ||] N7,
+
+\item[nāsādakṣiṇavārgāvāhipavanāṃtaprāṇebhidīghīkṛteś
+caṃdrāṃbha paripūritāmṛtatanuḥ prāḍaghaṃṭikāyās tathā
+chinnatkālaviśālavahnivaśaṃgaṃ prāgurdhanāḍīgaṇāt
+tatkāryaṃ kurute punar nnavattaraṃ chinnadrumaskaṃdhavat] P8,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇobhidīrghākṛteś
+caṃdrābhaḥ paripūritāṃ mṛtatanuḥ prāgbaṃdhakāyāyethā ||
+chinnatkālaviśālavahnivaśagaghyāghragranāḍīgaṇāt
+tatkāryaṃ kurute punarnnavataraṃ chinnadrumaskaṃdhavat ||] P1,
+
+\item[nāsādakṣiṇamārggavāhipavanāt prāṇobhidīrghākṛteś
+caṃdrābhaḥ paripūritāṃmṛtatanuḥ prāgghaṃṭikāyas tathā ||
+chinnatkālaviśālavahnivaśagaṃ prāgadhonāḍīgaṇān
+na tatkāryyāt kurute punarnnavataraṃ sthichinnaṅgulamaskaṃdhavat ||] P5,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghīkṛtaḥ
+caṃdrātaḥ paripūritāmṛtatanu prāgghaṃṭikāyās tathā ||
+bhiṃdat kālaviśālavanhivaśagān bhrūraṃdhranāḍīgaṇān
+tatkāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat ||] P12,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇopi dīrghākṛtaś
+candrāntaḥ paripūritāmṛtatanuḥ prāghaṇṭikāyā yathā |
+chindan kālavikālavahniparamān bhrūrandhranāḍīgaṇān
+taṃ kāyaḥ kurute punarnavataraṃ jīrṇādayaḥ skaṃdhavat ||] N12,
+
+\item[nāśādakṣiṇavartavāhipavano prāṇeti dīrghākṛtaṃ
+caṃdrāṃtaḥ paripūrya pūritatanuḥ prāgghaṃṭikāyās tathā
+bhiṃdat kālaviśālavahnivaśagān bhrūraṃdhranāḍigaṇās
+taṃ kāyaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat] N25,
+
+\item[nāśādakṣiṇamārgravāhipavanāt prāṇopi dīrghīkṛtaḥ
+caṃdrāṃśāt paripūritāmṛtatanuḥ prāgghaṃtikāyāḥ pathā
+chiṃdat kālaviśālavadriparamān bhrūraṃdhranāḍīguṇān
+taṃ kāyaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat] P11,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇebhidīrghākṛteś
+caṃdrāṃtaḥ paripūratāmṛtatanuḥ prāgghaṃṭikāyās tathā
+cchinakālaviśālavahnivaśagaṃ prāgaṃdhvanāḍīgaṇāt
+tatkāryaṃ kurute punarnnavattaraṃ cchinnadramuskaṃdhavat] P10,
+
+\item[nāsādakṣiṇamārgavāhipavanoghrāṇenadīrghākṛtaṃ
+caṃdrāṃgaḥ paripūritāmṛtatanuḥ prāgghaṭhikāyās tathā ||
+chaṃdaḥ kālaviśālavanhipavanād bhrūraṃdhranāḍīgaṇā
+tatkāryaṃ kurute punarnavataraṃ jīrṇadrumaskaṃdhavat ||] P7,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇopi dīrghīkṛtaś
+candrāmbhaḥ paripūritāmṛtatanuḥ prākghaṃṭikāyas tathā
+bhindan kālaviśālavanhivaśagān bhūrandhranāḍīgaṇān
+taṃ kāyaṃ kurute punar navataran jīrṇadrumaskandhavat] N16,
+
+\item[nāsādakṣiṇamārggavāhipavanāt prāṇobhidīrghākṛtauś
+candrāṃbhaḥ paripūritāmṛtatanuḥ prāgghaṭikāyās tathā
+chinnatkālaviśālavahnivaśagā bhrūraṃdhranāḍīgaṇāt
+tatkāryyaṃ kurute punarnnavataraṃ chinnadrumaskaṃdhavat] N17,
+
+\item[nāsādakṣiṇamārgāvāhipavanāt prāṇotidīrghīkṛtaḥ
+caṃdrāṃbhaḥ paripūritāmṛtatanuḥ prāgghaṃṭikāyās tathā 
+bhiṃdan kālaviśālavahnivasagān bhūraṃdhranāḍīgaṇān
+tatkāryyaṃ kurute punar nnavataraṃ chinnabhramaskaṃdhavat] N18,
+
+\item[nasādakṣiṇamārgavāhipavanāt prāṇotidīrghokṛtaś
+caṃdrābhaḥ paripūritāmṛtatanuḥ prāgghaṭikāyāḥ sadā
+bhinatkālaviśālavahnivaśagān bhrūraṃdhranāḍīgaṇān
+tatkālaṃ kurute punar navataraṃ chinnadrumaskaṃdhavat] N21,
+
+\item[nāsādakṣīṇamārggavāhipavanāt prāṇobhidīrghākṛtoś
+candrāmbhaḥ paripūritāmatatanu prāghaṇṭikāyās tathā |
+cihnatkālavisālavahnivasagād bhrūrandhranāḍiganāt
+tatkāryyaṃ kulute puṇarṇṇavataraṃ cihnadrumaskandhavat ||] N14,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghākṛtaḥ
+caṃdrāṃbhaḥ paripūritāmṛtanuḥ prāgghaṃṭikāyāḥ sadā
+bhiṃdan kālaviśālavahniviśagān bhrūraṃdhranāḍīgaṇān
+tatkālaṃ kurute punarvanataraṃ chinnadamaskaṃdhavat] P13,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇopidīghīkrataś
+caṃdrābhaḥ pariparitānanatanuḥ prāgghaṃṭikāyās tathā ||
+chinnaṃ kālaviśālavahnivaśagaṃ bhrūraṃdhranāḍīgaṇān
+tatkayaṃ kurute punarnavataraṃ chinnadramaskaṃdhavat ||] N10,
+
+\item[nāsādakṣiṇavāhimārgapavanāt prāṇotidīrghīkṛtaḥ
+caṃdrābhaḥ paripūrṇatāmṛtatanuḥ prāgghaṃṭikāyāḥ sadā
+chiṃdan kālaviśālavahnivaśagān bhrūraṃdhranāḍīgaṇān
+tatkālaṃ kurute punarnavataraṃ chiṃdan drumaskaṃdhavat] J11,
+
+\item[nāsādakṣiṇamārthethavāhipavanaprāṇotidisīkṛta
+caṃdrābhā paripūritāmrāṃmṛtatanuḥ prāke praghaṃṭikāyās tathā
+vidatkālaviśālavahnivaśagād bhūraṃdhranāḍīgaṇā
+taṃ kāyaṃ kurute punar navatanuṃ jīrṇadrumaskaṃdhavat] N22,
+
+\item[nāsādakṣiṇamārgavāhipavanāṃ prāṇotīdīrghīkṛteś
+caṃdrāṃbhaḥ paripūritāmṛtatanūḥ prāgghaṭikās tathā ||
+chinan kālaviśālavahnivaśagaṃ prāgraṃdhanāḍīgaṇā
+tatkāryaṃ kurute purnavataraṃ chidrumaskaṃdhavat ||] N15,
+
+\item[nāsādakṣiṇamārgavāhapavanaghrāṇotidīrghākṛtaś
+caṃdrābhāparipūritāmṛtatanuḥ prāk ghaṃṭikāyās tathā 
+bhiṃdan kālaviśālavahnivaśagān bhrūraṃdhranāḍīgaṇān
+taṃ kāyaṃ kurute punarnavataraṃ jīrṇadrumaskaṃdhavat ||] N11,
+
+\item[nāsaṃdakṣiṇamārgavāhipavanāt prāṇobhidīrghākṛteś
+caṃdrāṃbhaḥ paripuritātanuḥ prāgghaṭikāyās tathā ||
+chinnakālavisālāvahnivaśagaṃ prāgradhanāḍīgaṇā
+tatkāryaṃ kurute punar navattaraṃ chinnadrumaskaṃdhavat ||] N9,
+
+\item[nāsādakṣiṇavāhimārgapavanāt prāṇotidīrghīkṛtaś
+caṃdrābhaḥ paripūrṇatāmṛtatanuḥ prāghaṃṭikāyās tathā
+chiṃdan kālaviśālavahnivaśagān bhrūraṃdhranāḍīgaṇān
+taṃ kāyaṃ kurute punarnavataraṃ jīrṇaṃ drumaskaṃtvat] K1,
+
+\item[nāsādakṣiṇavāhimārgapavanāt prāpṇotidīrghīkṛtiṃ
+caṃdrābhaḥ paripūrṇatāmṛtatanuḥ prodghāṭikāyās sadā
+chīdan kālavisālavahnivasagān bhruraṃdhranāḍīgaṇān
+tat kāyaṃ kurute punarnavataraṃ chīrdaṃ drumaskaṃdavat] Ko,
+
+\item[nāsādakṣiṇavāhimārgayavanāt prāṇotidīrghīkṛtaś
+caṃdrāṃbhaḥ paripūrṇatāmṛtatanuḥ prāgghaṃṭikāyās tathā
+chiṃdan kālaviśālavahnivaśagān bhrūraṃdhranāḍīgaṇān
+tatkālaṃ kurute punarnavataraṃ chiṃdan drumaskaṃdavat] J14,
+
+item[nāsādakṣiṇavāhimārgayavano prāṇotidīrghīkṛtaś
+caṃdrāṃbhaḥ paripūritāmṛtatanuḥ prāgghaṃṭikāyās tathā
+chinnat kālaviśālavahnivaśagaṃ bhrūraṃdhranāḍīgaṇāt
+svaṃ kāyaṃ kurute punarnavataraṃ chinna drumaskaṃdhavat] J15,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇeti dīrghākṛti |
+ścaṃdrāṃbhaḥ paripūritāmṛtatanu prāgghaṭikāyās tathā |
+chinnatkālaviśālavahnivaśagāṃ prāgraṃdhranāḍīgaṇāt |
+tatkāryaṃ kurute punannavataraṃ chinnadrūmaskaṃdhavat ||] J17,
+
+\item[nāsādakṣiṇamārggavāhipavanāt prāṇebhidīrghākṛtauś
+caṃdrāṃbhaḥ paripūritāmṛtatanuḥ rprāgghaṃṭikāyās tathā |
+cchinnat kālaviśālavahnivaśagāṃ bhrūraṃdhranāḍīgaṇāt
+tatkāryyaṃ kurute punarnnavataraṃ cchinnadrumaskaṃdhavat ||] N6,
+
+\item[nāsādakṣiṇamārgavāhipavano prāṇena dīrghīkṛtoś
+caṃdrāṃbhaḥ paripūritāmṛtatanuṃ prāgghaṃṭikāyās tadā
+chinnat kālaviśālavahnivaśagaṃ bhrūraṃdhranāḍīgaṇās
+tatkāryaṃ kurute punarnavataraṃ chinnadrumaskaṃdhavat] LD1,
+
+\item[nāsādakṣaṇamārgavāhīpavanaprāṇo na dīrghā kṛteś
+caṃdrāṃbhaḥ paripūrītāmṛtatanu prāgghaṃṭikāyās tathā
+chinnat kālaviśālavahnīvaśagaṃ bhrūraṃdhranāḍīgaṇās
+ta kāryaṃ kurute punarnavataraṃ chīnnadrumaskaṃdhavat ||] LD2,
+
+Item\[nāsādakṣiṇamārgavāhipavanāt prāṇotīdīrghākṛtaiś
+caṃdrābha paripūritāmṛtatanu prāg ghaṭikāyās tathā
+chinnaṃ kālaviśālavahnivaśagaṃ bhrūraṃdhranāḍīgaṇāt
+svaṃ kāryaṃ kute punarnavataraṃ chinnadrumaskaṃdhavat] J16,
+
+\item[nāsādakṣiṇamargavāhipavanāt prāṇoditi dīrghate
+caṃdrābhaḥ paripūritāmṛtatanuḥ prāg ghaṭikāyā tathā
+chinnaṃ kālaviśālavahnivaśagaṃ bhrūraṃdhranāḍīgaṇāt
+kāryaṃ tat kurute vapur navataraṃ chinnaṃ drumaskaṃdhavat] J12,
+
+\item[nāsādakṣiṇamārgavāhipavanodghrāṇetidīrghī kṛteś
+candrābhaḥ paripūritāmṛtatanuḥ prāgghaṇṭikāyās tadā |
+bhiṃdankālaviśālavahnivaśagān bhrūrandhranāḍīgaṇān
+taṃ kāyaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat//] J7,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇo dīrghīkṛtaḥ |
+candrāṃgāt paripūritāmṛtyutanuḥ prāgghaṃṭikāyā yathā ||140||
+(bhiṃnat)kālaviśālavahnivaśagān mṛṃraṃdhranāḍīgaṇān ||
+taṃ kāyaṃ kurute punar nnavatarajīrṇaṃ drumaskaṃdhavat ||] J8,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghākṛtiś
+caṃdrāṃbhaḥ paripūritāṃmatatanuḥ prāgvvaṃḍhikāyās tathā||13||
+chinnat kālaviśālavahnivaśagāt prāgraṃdhranāḍīgaṇā ||
+ttatkāryaṃ kurute punar nnavataraṃ chinnadrumaskaṃdhavat ||] J10,
+
+\item[nāsādakṣiṇamārgavāhipavanaḥ prāṇotidīrghākṛtiś
+caṃdrāṃbhaḥ paripūritāṃmatatanuḥ prāgvvaṃḍhikājāḥ yathā||13||
+chinnat kālaviśālavahnivaśagāt prāgraṃdhranāḍīgaṇā ||
+ttatkāryaṃ kurute punar nnavataraṃ chinnadrumaskaṃdhavat ||] J10a,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghī kṛtaḥ ||
+caṃdrāṃtaḥ paritāmṛtanuḥ māgdhaṃṭikāyās tathā ||57||
+bhidat kālaviśālavanhivaśagān bhrūṃraṃdhranāḍīgaṇāt ||
+tatkāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat ||] J4,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghī kṛtaś
+caṃdrāṃbaḥ paripūrṇatāmṛtanu prāpnoti kāye sadā
+biṃdūn kālaviśālavahnivaśagān bhrūraṃdhranāḍīgaṇān ||
+tatkāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat ||] J13,
+
+\item[nāsādakṣiṇavāhimārgapavano prāṇatidīrghākṛtā
+caṃdrābhaḥ paripūrṇatāmṛtatanu prāpnoti kāye sadā |
+viṃdūnakālaviśālapāśavtṛ bhūradhanāḍīgaṇān
+tatkālaṃ kurute punarnavatanu chinne dumaskaṃḍavat ||] J2,
+
+\item[nāsāt dakṣiṇamārgaṃ vāhi pavanān prāṇodīrghītikṛta
+caṃdrābhaḥ paripūritāmṛtatanuḥ prākghaṇṭiyās tathā ||
+bhidan kālaviśālavahi vaśan bhūraṃdhranāḍīgaṇān ||
+tat kāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat ||] J1,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghīkṛtś
+caṃdrābhaḥ paripūritāmṛtatanuḥ prāgghaṃṭikāyās tathā
+chiṃdat kālavisālavanhivasagāt bhrūraṃdhranāḍīgaṇāt
+tatkāryaṃ kurute punar navataṭaṃ chinnadrumaskaṃdhavat] C9,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotra dīrghīkṛtḥ
+aṃtas tat paripūritāmṛtatanuḥ prāk ghaṃṭikāyās tathā
+bhiṃdan kālavisālavahnivasagān bhrūraṃdhranāḍīgaṇān
+taṃ kāyaṃ kurute punar navataṭaṃ jīrṇaṃdrumaskaṃdhavat] J3,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotra dīrghīkṛtaḥ
+aṃtastāt paripūritāćtatanuḥ prāk ghaṃṭikāyās tathā 
+bhiṃdan kālaviśālavahnivaśagān bhrūraṃdhranāḍīgaṇān
+taṃ kāyaṃ kurute punarnavataraṃ jīrṇaṃ drumaṃ skaṃdhavat] J6,
+
+\item[nāsādakṣiṇavāhimārgapavano ghrāṇo na dīrghī kṛtaḥ
+ścandrābhaḥ paripūritāṃmṛtatanuḥ prā ghaṭṭikāyāstathā/
+cheda kālaviśālavahnipavanād bhrūrandhranāḍīgaṇāt
+tatkāyaṃ kurute punar navataraṃ jīrṇaḥ drumaḥ skaṃdhavat//] C8,
+
+\item[nāsādakṣiṇavartmavāhipavano ghrāṇe ca dīrghīkṛtaś
+candrāmtaḥparipūryapūritatanuḥ prāg ghaṃṭikāyās tathā/
+bhiṃdan kālaviśālavahnivaśagāṃ bhrūraṃdhranāḍīgaṇāṃs
+taṃ kāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat//] C4,
+
+\item[nāsādakṣiṇavartmavāhipavano ghrāṇoti dīrghīkṛta
+Candrāmtaḥ paripūryapūritatanuḥ prāg ghaṃṭikās tathā/
+bhidat kālaviśālavahnivaśagāṃ bhrūraṃdhranāḍīgaṇāṃs
+taṃ kāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat//] N5, 
+\item[nāsādakṣiṇamārgavāhipavano ghrāṇeti dīrghrākṛtaś
+candrābhā paripūritāmṛtanuḥ prāk ghaṃṭikāyās tathā/
+bhiṃdaṃn kālaviśālavahnivaśagān bhrūraṃdhranāḍīgaṇāt
+taṃ kāyaṃ kurute punar navataraṃ jīrṇadrumaskaṃdhavat//] N8,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇo ’tidīrghīkṛtaś
+candrāmbhaḥparipūritāmṛtatanuḥ prāg ghaṇṭikāyās tataḥ/
+chitvā kālaviśālavahnivaśagaṃ bhrūrandhranāḍīgataṃ
+tat kā(yaṃ) kurute punar navataraṃ chinnaṃ drumaṃ skandhavat//] Ba5, N4b,
+
+\item[nāsādakṣiṇamārgavāhipavano ghrāṇena dīrghaṃkṛtaś
+candrāmbhaḥ paripūritāmṛtatanuḥ prāg ghaṇṭṭikāyās pathā/
+chinnaṃ kālaviśālavahnivaśagān prāgruddhanāḍīgaṇāt
+tat kāyaṃ kurute punar navataraṃ chinnadrumaskandhavat//] A1,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghākṛtaḥ
+caṃdrātaḥ paripūritāmṛtatanuḥ prāg ghaṇṭakāyās tathā/
+bhindet kālaviśālavanhivaśagān bhrūraṃdhranāḍīgaṇān
+tat kāryaṃ kurute punar navataraṃ jīrṇadrumaskandavat//] B1,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghākṛtaś
+caṃdrātaḥ paripūritāmṛtatanū prāg ghaṃṭikāyaṃs tathā/
+bhiṃdaṃt kālaviśālavanhivaśagān bhrūraṃdhranāḍīgaṇān
+tat kāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdavat//] B3,
+
+\item[nāsādakṣiṇamārggavāhipavanaḥ prāṇo(dgidīrvvī) kṛtaś
+candrāmbhaḥparipūritāmṛtatanuḥ prāg ghaṇṭikāyās tathā/
+chindan kālaviśālavahnivasagān bhrūraṃdhranāḍīgatān
+taṃ kāyaṃ kurute punar navataraṃ chinnadrumaskandhavat//] B2,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prāṇotidīrghākṛtaś
+caṃdrāṃtaḥ paripūritāmṛtatanū prāg ghaṃṭikāyās tathā/
+bhiṃdat kālaviśālavanhivaśagān bhrūraṃdhranāḍīgaṇān
+tat kāryaṃ kurute punar navataraṃ jīrṇadrumaskaṃdavat//] C2,
+
+\item[nāsādakṣiṇamārgavāhipavanāt prā?bhidīrghā kṛte
+ścaṃdrāṃbhaḥ paripūritāmṛtatanuḥ prāg ghaṭikāyās tathā/
+chiṃdat kālaviśālavanhivaśagān prāgraṃdhranāḍīglāt
+tat kāyaṃ kurute punar navataraṃ chinnadrumaskaṃdhavat//] C3,
+
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C5, J5, K2, N3, N4a, N13, N20, N23, N24, P3,
+
+  \begin{description}
+ Please make notes on the position of the verse.
+    • It is given at 158 in B1; B3
+    • It is given at 148 in C2
+    • It is given at 114 in C8; J10; J15; Ko; P6; J8; N17;
+    • It is given at 115 in P8; P9;
+    • It is given at 110 in C9
+    • It is given at 141 in J1; N18;
+    • It is given at 142 in J3
+    • It is given at 139 in J6
+    • It is given at 117 in J7; P5; P7;
+    • It is given at image no.54 in J4
+    • It is given at image no.00016 in J11;
+    • It is given at 128 in J12; N10;
+    • It is given at 125 in N4b
+    • It is given at 135 in N9
+    • It is given at 137 in P10
+    • It is given at 157 in J13
+    • It is given at 119 in J14
+    • It is given at PDF pg no. 95 in J16
+    • It is given at 113 in J17; N6; N14;
+    • It is given at 110 in K1
+    • It is given at 120 in LD1
+    • It is given at 121 in N7
+    • It is given at 122 in LD2
+    • It is given at PDF no. 13 in N11
+    • It is given at JPG no. 1341 lower in N12
+    • It is given at PDF no. 21 in N5
+    • It is given at PDF no. 28 in N8
+    • It is given at PDF no. 09 in N15
+    • It is given at 138 in N16
+    • It is given at 148 in N21; P2; P12; P13;
+    • It is given at 147 in P4;
+    • It is given at 134 in P1
+    • It is given at PDF no. 23 in N22
+    • It is given at PDF no. 20 in N25
+
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_119b]
+
+  \sthana{3.119b}
+
+\item[śambhugarbhām arundhatīm] J7, J11, J14, 
+\item[śambhugarbhām arundhatī] J2, N21, 
+\item[śambhugabham irundhatī] P13,
+\item[śabdagarbhām aruṃdhatī] J5,
+\item[śabdagarbhbhām aruṃddhatī] J8,
+\item[śabdagaṃdhām aruṃddhatī] N3,
+\item[śabdagarbhām sarasvatīm] A1, Ko, P11, 
+\item[śaṃbhugarbhām sarasvatīm] N12, P7,
+\item[śabdagabhām sarasvatīm] C6,
+\item[sraṣṭugarbhā sarasvatī] C8,
+\item[sukhasuptām aruṃdhatī] B1, J15, J17, N6, N9, P9, 
+\item[saṣumnāyām aruṃdhatī] C3, J1, J12, K1, 
+\item[saṣumṇāyām aruṃdhatīṃ] N10,
+\item[saṣumṇāyārūṃdhatī] J16,
+\item[sukhasuptām aruṃdhati] N15,
+\item[sukhasuptām aruṃdhatīḥ] B3,
+\item[sukhasuptām aruṃdhatīm] Ba5, C9, J3, J6, LD1, N5, N13, P4, 
+\item[sukhasuptām aruṃdhatīm] P10,
+\item[susuptām aruṃdhatīm] P5,
+\item[suṣasuptām aruṃdhatī] N7,
+\item[sukhaṃ suptāṃ maruṃdhatīm] J4, J13, 
+\item[sukhaṃ suptāṃ maruṃdhatī] LD2,
+\item[susuptā māmaruṃdhati] N14,
+\item[suptā prāṇaruṃdhati] P1,
+\item[suṣasuptām aruṃdhatī] P8,
+\item[suṣasuptām aruṃdhatīṃ] C2, J10, N17,
+\item[susuptāyām aruṃdhatīṃ] P12,
+\item[suṣumnāyām aruṃdhatīṃ] N18,
+\item[sukhasuptām aruṃdhatām] B2,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C4, C5, K2, N4a, N4b, N8, N11, N16, N20, N22, N23, N24, N25, P2, P3, P6,
+
+  \begin{description}
+Could \emph{śaṃbhugarbhām} here mean that kuṇḍalinī is Śiva's offspring? The literal sense of a bahuvrīhi (in whose womb is Śiva) does not seem to make sense in this context.
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_127]
+
+  \sthana{3.127ab}
+
+\item[iyaṃ tu madhyamā nāḍī dṛḍhābhyāsena yoginām] N4b,
+\item[īḍāvimadhyamā nāḍī iḍhābhyāsena yogini] J2,
+\item[kuṭilā madhyamā nāḍī dṛḍhābhyāsena yoginām] J3, J6a, 
+\item[pittāpi madhyamā nāḍī dṛḍhābhyāsena yoginām] J1, J6, 
+\item[khilāpi madhyamā nāḍī dṛḍhābhyāsena yoginaṃ] N3, N22, 
+\item[khilāpi madhyamā nāḍī dṛḍhābhyāsena yoginām] C2, J8, K1, 
+\item[khilāpi madhyamā nāḍī pri?ḍābhyāsena yogināṃ] N10,
+\item[khilāpi madhyamā nāḍī dṛḍhābhyasepi yogminā] J5,
+\item[khilāpi madhyamā nāḍī dṛḍhābhyāsena yoginā] C6, P6, P7,
+\item[khilāpi madhya nāḍī ca dṛḍhābhyāsena yoginā] P11,
+\item[khilāpi madhyamā nāḍī iḍābhyāsena yoginā] C8,
+\item[khilāpi madhyamā nāḍī dṛḍhābhyāsena yoginaḥ] B2, 
+\item[...lāpi madhyamā nāḍī dṛḍhābhyāsena yoginaḥ] N7,
+\item[kṣīṇāpi madhyamā nāḍī dṛḍhābhyāsena yoginām] N12,
+\item[calāpi madhyamā nāḍī dṛḍhābhyāsena yoginām] J11,
+\item[cālitā madhyamā nāḍī dṛḍhābhyāsena yoginā] Ko,
+\item[pittāpi madhyamā nāḍī dṛḍhābhyāsena yoginām] N16,
+\item[vināpi madhyamā nāḍī dṛḍhābhyāsena yoginaḥ] C9, N17, 
+\item[vināpi madhyamā nāḍīm iḍābhyāsena yoginaḥ] J12, J15, P9, 
+\item[vināpi madhyamā nāḍī dṛḍhābhyāsena yoginā] A1, B3, N14,
+\item[iyaṃ tu madhyamā nāḍī dṛḍhābhyāsena yoginā] Ba5,
+\item[iyaṃ tu madhyamā nāḍī dṛḍhābhyāsena yogināṃ] N13, 
+\item[vināpi madhyamā nāḍī dṛḍhābhyāsena yogināṃ] B1, J4, J7, J13, P2, P4, P12, 
+\item[vināpi madhyamā nāḍī dṛḍhābhyāsena yonināṃ] C4, N5, N18,
+\item[vināpī madhyamā nāḍī driḍhābhyāsena yogīnaḥ] LD2, N6, 
+\item[vināpi madhyamāṃ nāḍī driḍhābhyāsena yoginaḥ] J10, P8, 
+\item[vināpi madhyamāṃ nāḍī drāḍhābhyāsena yogina] J17,
+\item[vināpi madhyamāṃ nāḍīm iḍābhyāsena yoginaḥ] C3, LD1, P5, P10,  
+\item[vināpi madhyamā nāḍiṃ iḍābhyāsena ?????] N15,
+\item[vināpi madhyamāṃ nāḍīm iḍābhyāsena yoginaḥ] P1,
+\item[mināpi madhyamāṃ nāḍīm iḍābhyāsena yoginaḥ] N9,
+\item[vināpī madhyamāṃ nāḍīm iḍābhyāsena yoginīḥ] J16,
+\item[calāpi madhyamā nāḍī dṛḍhābhyāsena yonināṃ] J14,
+\item[ṣilāṃpi madhyamā nāḍī dṛḍhābhyāsena yogināṃ] N8,
+\item[ṣilāpi madhyamā nāḍī dṛḍhābhyāsena yogināṃ] N11,
+\item[kṣīṇāpi madhyamā nāḍī dṛḍhābhyāsena yogināṃ] N21, P13, 
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C5, K2, N4a, N20, N23, N24, N25, P3,
+
+  \begin{description}
+    khila seems odd (could it mean defective or unperfected here?), but the alternative readings are not helpful (except iyaṃ tu, which seems like a patch).
+  \end{description}
+
+\end{marma}
+
+\begin{marma}[hp03_128]
+
+  \sthana{3.128ab}
+
+\item[upāsanavinidrāṇāṃ rājayogaḥ samudrakaḥ]
+\item[upāsanaṃ vinidrāṇāṃ rājayogaḥ samudrakaḥ] J7,
+\item[upāsanavinidrāṇāṃ rājayogaḥ sasadgatiḥ] J6, 
+\item[upāsanavinidrāṇāṃ rājayogaḥ sasadgatiḥ] J3,
+\item[upāsanepinidrāṇāṃ rājayogasamudbhavān] J5,
+\item[upāsane viniprā rājayogasamudravat] N3,
+\item[abhyāseṣu vinidrāṇāṃ manodhrata samādhinā] J2,
+\item[upāsano vinidrāṇāṃ rājayogaḥ samaṃ gati] J1,
+\item[upadeśaṃ hi mudrāṇāṃ yo dhatte sāṃpraḍyikaḥ] C9,
+\item[upadeśaṃ hi mudrāṇāṃ yo datte sāṃpraḍāyakaḥ] P12,
+\item[upadeśaṃ hi mudrāṇāṃ yo dhatte sāṃpraḍāyikaḥ] P13,
+\item[upadeśaṃ hi mudrāṇāṃ yo dhatte sāṃpraḍāyikaṃ] N22, P5, 
+\item[upadeśaṃ hi mudrāṇāṃ yo dhatte sāṃpraḍāyikāṃ] P8,
+\item[upadeśo hi mudrāṇā yo datte ṣāṃpraḍayakaḥ] N21,
+\item[upadeśaṃ hi mudrāṇāṃ yo datte śāṃpradīpikaṃ] N18, 
+\item[upāṃsanavidhinṛṇāṃ yo datte saṃprādāyikaṃ] N8, N11,
+\item[upāsanavinidrāṇāṃ rājayogaḥ samudragaḥ] K1,
+\item[upāsanavinidrāṇāṃ rājayogaḥ samāhitaḥ] C4,
+\item[upāsanavinidrāṇāṃ rājayogaḥ samāhakaḥ] N5,
+\item[upāsanavinidrāṇāṃ rājayogaḥ sasadgatiḥ] N16,
+\item[abhyāse tu vinidrāṇāṃ mano dhṛtvā samādhinā] Ba5, N4b,
+\item[abhyāsena hi mudrāṇāṃ tad udeti samādhinā] A1, B2, C3, J10, J12, LD1, N6, N7, N9, N10, N15, N17, P9, P10, 
+\item[abhyāsena hi mudrāṇāṃ na tūdeti samādhinā] P2, P4, 
+\item[abhyāsena hi mudrāṇāṃ tad udeti samādhinaḥ] P1,
+\item[abhyāsena hi mudrāṇāṃ tad udetī samādhīnā] LD2,
+\item[abhyāsena hi mudrānāṃ tad udetu samādhīnāṃ] N14,
+\item[abhyāsena hī muṃdrāṇāṃ tad udeti samādhīnā] J15,
+\item[abhyāseṣu vinidrāṇāṃ anu(dgata) samādhinām] C6,
+\item[abhyāseṣu vinidrāṇāṃ anudruta samādhiṣu] P11,
+\item[abhyāseṣu vinidrāṇāṃ manodhṛta samādhināṃ] J14,
+\item[abhyāse tu vinidrāṇāṃ mano dhṛtvā samādhināṃ] N13,
+\item[abhyāsena vinidrāṇāṃ manodhṛti samādhinā] N12, 
+\item[abhyāseṣu vinidrāṇāṃ anudbhūta samādhinām] J11,
+\item[abhyāseṣu vinidrāṇāṃ rājayogaṃ samudragaḥ] P6,
+\item[abhyāseṣu vinidrāṇāṃ rājayoga samudrakaṃ] J10a,
+\item[abhyāseṣu vinidrāṇāṃ manovṛti samādhināṃ] Ko,
+\item[abhyāseṣu vinidrāṇāṃ manudbhūta samādhiṣu] C8,
+\item[abhyāseṣu vinidrāṇāṃ manudṛta samādhiṣu] J8,
+\item[abhyāsena hi mudrāṇāṃ tad udeti samādhināṃ] J17,
+\item[abhyāsena hī muṃdrāṇāṃ tad udeti samādhinā] J16,
+\item[abhyāsena hi mudrāṇāṃ na tūdeti samādhinā] C2, J4, J13, 
+\item[abhyāsena hi mudrāṇāṃ na tudeti samādhinā] B1, B3, 
+\item[abhyāse vinimudrāṇāṃ manudbhutasamādhiṣu] P7,
+\item[(illegible/unavailable)] Bo1, Bo2, Bo3, C5, K2, N4a, N20, N23, N24, N25, P3,
+
+  \begin{description}
+    No idea. We need more readings. 
+  \end{description}
+
+\end{marma}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2023-09-02 marma/marma1.tex
new routine for jyotsna

--- 2022-12-16 marma/Marmasthāna.tex
@@
  \end{marma}
-
-Notes:
-    • P3, R1, has not 2nd chapter
-    • Folios of P4 are scattered 
-    • P16 has only 3 folios of chapter 2 at the end
-    • P22 has only 3 folios of chapter 1
-    • R3 has only 1 folio of chapter 2
-    • 1 & 2 Chapters are mixed in R4
-    • R5 has only one folio of chapter 2
-\begin{marma}[hp02_003]
-\sthana{2.3b}
-\item[jīvanam] Ba, Ba5, Bo2, Bo3, C1, C6, J1, J2, J3, J6, J12, J13, J14, J16, N17, P1, V1, N4a, N7, N8, N9, N10, N11, N12, N13, N16, P10, P13, P14, V16, V17, V10, N18, N21, P18, P20, P25, P26, P27, R5, V21, Tue, N23, V6, V7, V20, V28, V18
-\item[jīvaṃnam] N14
-\item[jīvitam] A1, B1, B2, B3, Bo1, C2, C3, C4, C8, J4, J5, J7, J8, J10, J11, LD1, LD2, N1, N2, N5, V8, N20, N22, P2, P6, P9, P17, P21, P23, V14, V3, N24, N25, P12, P15, R1, R3, V15, V5, V22, V2, V12, V13, V9, G11, Ba3, G1,
-\item[jīvatam] P5, P24,
-\item[jivitam] J15,
-\item[jīvita] N26
-\item[jīviti] P11, 
-\item[jīvitum] J17, N19
-\item[jītavim] N3
-\item[jīvo na] Ko, P7, P8, 
-\item[(unavailable/illegible)] C5, C9, N6, T, P16, P22,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_003]
-\sthana{2.3d}
-\item[nibandhayet] Ba, Bo2, C6, V1, J2, J4, J6, J8, J11, J14, N12, N16, N19, N22, P11, P15, P25, V14, V3, V13
-\item[nibaddhayet] G11, P6, P27,
-\item[nirundhayet] A1, B1, B2, B3, C1, C3, C9, J5, J10, J13, J15, J17, LD1, LD2, N17, P1, V19, W4, N2, N7, N9, N10, N14, P5, P8, P9, P10, P23, V16, V17, N20, V21, N23, N26, P2, P7, P26, V20, V5, V2, V18, V12, V9, Ba3
-\item[nirodhayet] Ba5, Bo1, Bo3, C2, C4, C8, J1, J3, J6pc, J7, V19, N1, N4a, N5, N8, R3, V8, N11, N13, V10, N18, N21, P13, P14, P17, P18, P20, Tue, N24, N25, P12, P21, R5, V6, V7, V15, V28, V22
-\item[nirudhayet] P24,
-\item[niruddhayet] R1, 
-\item[nisaṃdhayet] J16,
-\item[nirodhanaṃ] J12,
-\item[nibandhanāt] V19, N3, G1
-\item[nibadhnayāt] Ko,
-\item[(unavailable/illegible)] C5, N6, T, P16, P22,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_004]
-\sthana{2.4d}
-\item[kāyasiddhiḥ] A1, B1, B2, C2, C6, J5, J8, J14, V1, N1, N8, N11, P2, P12, P17, P26, R3, V17, N24, V20, V15, V9, G11, G1
-\item[kāyāsiddhiḥ] J2, N3,
-\item[kāyāsuddhi] V6
-\item[kāyāsiddhi] P6, P27, 
-\item[kayasiddhiḥ] N20
-\item[kāyaśuddhiḥ] N23
-\item[kāryasiddhiḥ] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C8, C9, J3, J4, J6, J10, J11, J13, J17, Ko, LD1, N17, P10, V1, W4, N2, N4a, N5, N9, N12, N13, V16, V10, N19, N21, P1, P5, P6, P7, P8, P9, P11, P13, P14, P15, P18, P20, P21, P23, P24, P25, R1, Tue, N25, N26, V70, V5, V28, V22, V2, V18, V12, V13, Ba3
-\item[kāryasiddhaḥ] R5,
-\item[kāryasiddhiṃ] J15, 
-\item[kāryāsiddhiḥ] J1
-\item[kāryasiddhi] J12, LD2, N7, V8, N10, N14("rya" unclear), N16, N18, N22, V21, V3 
-\item[kāryasiddhī] J16,
-\item[kāryasuddhiḥ] V14
-\item[kāyāśuddhiḥ] J7,
-\item[(unavailable/illegible)] C5, N6, T, P16, P22,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_006]
-\sthana{2.6c}
-\item[suṣumnāsu snighdhā] J4, V1, V15
-\item[suṣumnāḥ snighdhā] P15, 
-\item[suṣumnānāḍisthā] Ba5, N13, P13, P14, P20, Tue, 
-\item[suṣumṇāṃ nāḍisthāḥ] P21, 
-\item[suṣumṇānāḍīsthā] N4a, N21, P18,  
-\item[suṣumṇānāmadhysthā] V2
-\item[suṣumnā nāḍīsthaṃ] G1
-\item[suṣumṇānāḍīstho] N24
-\item[suṣumnānāḍīstho] J14,
-\item[suṣumṇānāḍīschā] Bo3, V28, V22
-\item[suṣumṇāṃtasthhā] J15,
-\item[suṣumnāṃtaraschā] J11
-\item[sukhumnāṃtarasthā] V19
-\item[suṣumnāṃtarālasthā] B3, Bo1, Bo2, C1, C2, J3, J6, J13, N1, N8, N11, P2, P17, P23, 
-\item[suṣumnāṃtarālaschā] P12, 
-\item[suṣumnāṃtarālasyā] J1
-\item[suṣumnāṃtarājasthā] N16
-\item[suṣumṇāṃtarasthā]  N5, C4
-\item[suṣumṇāṃtarāla] V5
-\item[suṣumnā pāśvaṃsthā] A1, V20
-\item[yathā suṣumṇā pārśvaḥsthā] Ba,
-\item[yathā suṣumṇā pārśvasthā] C6, C8, J7, N23, P7, P25, P26, 
-\item[suṣumṇāpañcasthā] N19
-\item[yathā suṣumnā mukhasthā] V17
-\item[yathā suṣumnā sumalāḥ] P27,
-\item[yathā suṣumnā nāḍisthamala] V10
-\item[yathā suṣumnņayā śvaḥsthā] V8, V21
-\item[yathā suṣumņā sustakcya?] P11, 
-\item[suṣumņā sustabdhāḥ] N12
-\item[yathā suṣumnā susvasthā] J5, N22, N26, V14, V3, N3
-\item[suṣumņā suṣasthā] N20
-\item[suṣumnņayā ścasyā] V6, V13
-\item[suṣumnāvasthāyaṃ] B1, B2
-\item[suṣumnānāḍibhyāṃ] V7
-\item[suśumnāsusvasthā] G11
-\item[suṣumnāsvasthāyan] N2
-\item[suṣama susvasthā] J8
-\item[suṣumnāsui svasthā] J8ac,
-\item[suṣumnāsaṃsvasthā] J2,
-\item[suṣumṇā sūmalā] P6,
-\item[sukham avasthāya] C3, C9, J10, J12, J17, LD2, N17, P1, P5, P24, W4, N7, N9, V16, V18, V12, V9, Ba3
-\item[sukhamalāsyāya] P9, 
-\item[sukham avasthāyaḥ] J16,
-\item[sukham avasthāyā] N14, P8, 
-\item[sukham avasthāpyā] LD1
-\item[sukham avaschāya] R1,
-\item[sukham avaschāṃya] N10
-\item[sukha........] P10,
-\item[suṣam avasthāya] J8pc, N18,
-\item[suṣumṇām avasthāpya] J7pc
-\item[suṣumnāṃtvarayāpalaḥ] N25
-\item[(unavailable/illegible)] C5, Ko, N6, T, P16, P22, R5,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_009]
-\sthana{2.09 additional verses}
-dvādaśamātraḥ kanīyānmadhyamo mātracaturvviśaḥ |
-uttamaḥ ṣaṭtriṃśanmātro mātrābhedāḥ smṛtā tajñaiḥ |
-vaśiṣṭasaṃhitādau tu pūrakaṃ ṣoḍaśamātrabhiḥ |
-kuṃbhakaṃ ca catuṣṣaṣṭimātrābhiḥ recakaṃ tu |
-dvātriṃśanmātrābhiḥ vidhivatkuṃbhakaṃ kṛtvā punaścaṃdreṇa recayet ||
-\begin{description}
-\item[Yes] J17, R1, 
-\item[No] A1, B1, B2, B3, Ba, Ba5, Bo1 Bo2, C1, C2, C3, C4, C5, C6, C8, C9, J1, J2, J3, J5, J10, J11, J12, J14, J15, J16, LD2, N17, V1, J4, J6, K7.
-J8, J13, Ko, LD1, W4, V19, N1, N2, N3, N4a, N5, N6, N7, N8, N9, V28, V8, N10, N11, N12, N13, N14, N16, P5, P10, V16, V17, V10, N18, N19, N20, N21, N22, P1, P6, P7, P8, P9, P11, P12, P13, P14, P15, P16, P17, P18, P20, P21, P22, P23, P24, P25, P26, P27, V14, V21, V3, Tue, N23, N24, N25, N26, P2, R5, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9, G11, T, Ba3, G1
-    \end{description}
-
-\begin{marma}[hp02_011]
-\sthana{2.11 additional verses}
-iḍayā piva ṣoḍaśabhiḥ pavanaṃ |
-kuru ṣaṣṭicatuṣṭayamaṃtagataṃ |
-tyaja piṃgalayā sanakaiḥ sanakaiḥ |
-daśabhirdaśabhirdaśabhirdvyadhikaiḥ ||
-\begin{description}
-\item[Yes] B1, B3, C, V19, N1, N5, N10 , N12, N21, N25, P12, P13, P26, P17, P20, P21, V28
-\item[No] A1, B2, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C4, C5, C6, C8, J1, J2, J3, J10, J13, J17, J4, J5, J6, J7, J11, J12, J14, J15, J16, Ko, LD1, LD2, N17, P1, V1, J8, W4, N2, N3, N4a, N6, N7, N8, N9, V8, N11, N13, N14, N16, P5, P6, P8, P15, P18, V16, V17, V10, N18, N19, N20, N22, V14, V21, V3, Tue, N23, N24, N26, P2, P7, P9, P10, P11, P14, P16, P22, P23, P24, P25, P27, R1, R5, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9, G11, T, Ba3, G1
-    \end{description}
-
- \begin{marma}[hp02_012]
-\sthana{2.12cd}
-\item[uttānaṃ cottame prāṇarodhe padmāsane muhuḥ] Ba, C1, C6, P25, V14
-\item[uttānaṃ cottame prāṇe rodhe padmāsane muhuḥ] P7, 
-\item[uttānaṃ cottame prāṇe stebdhe? padmāsane muhuḥ] C8,
-\item[uttame sthānam āpnoti tato vāyuṃ nibaṃdhayet] N13, V10, Tue, V7
-\item[utiṣṭhe tūtame prāṇā yāvat padmāsane sūmaṃ] J2,
-\item[uttiṣṭhaty uttame prāṇaṃ rodheḥ padmāsanaṃ .uhuḥ] V1
-\item[uttiṣṭhāṃty uttame prāṃṇā baddhe padmāsane uttama] J16
-\item[uttiṣṭhaty uttame prāṇē baddhe padmāsanae dṛḍhe] C2, P17, 
-\item[uttiṣṭhaty uttame prāṇaḥ baddhe padmāsanae dṛḍhe] J13,
-\item[uttiṣṭhaty uttame prāṇaṃ baddhā padmāsane dṛḍhe] B1
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane dṛḍhe] B2, B3, P2,
-\item[uttiṣṭhaty uttame prāṇa baddhe padmāsanaṃ dṛḍhaṃ] Ba3
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane dṛḍhe] P13, 
-\item[uttiṣṭhaty uttame prāṇaḥ badhe padmāsane dṛḍhe] P27,
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane sadṛḍhe] P12,
-\item[uttiṣṭhaṃty uttame prāṇā baddhe padmāsane muhuḥ] LD2, J10, P1, P8, P9, P10, P24, 
-\item[uttiṣṭhaṃty ūttame prāṇā baddhe padmāsane muhuḥ] J15, P5, R1, 
-\item[uttiṣṭhaṃty uttame prāṇā baddha padmāsane muhuḥ] C3, N10
-\item[uttisthaṃ cottame prāṇā baddha padmāsane muhu] N14
-\item[uttiṣṭhaty uttame prāṇa baddhe padmāsanaṃ muhuḥ] J8ac, J14, V3
-\item[uttiṣṭhaty uttame prāṇa baddho padmāsane muhuḥ] V8
-\item[uttiṣṭhatm uttame prāṇa baddhe padmāsane muhuḥ] V16, V17, V21, V20, V18, V13
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane muhuḥ] C9, 
-\item[uttiṣṭhaty uttame prāṇo baddha padmāsane muhuḥ] A1
-\item[uttiṣṭhaty uttame prāṇo badhye padmāsane muhuḥ] N9
-\item[uttiṣṭhaty uttame prāṇo baddhapadmāsane muhuḥ] N7
-\item[uttiṣṭhaty uttame prāṇē baddhe padmāsanaṃ muhuḥ] J8pc
-\item[uttiṣṭhaty uttame prāṇē baddhapadmāsane muhuḥ] P26, 
-\item[uttiṣṭhaty uttame prāṇē baddha padmāsane haye] Bo1
-\item[uttiṣṭhaṃty uttare prāṇā baddhe padmāsane muhuḥ] J17
-\item[uttiṣṭhaṃty uttame prāṇā baddhe padmāsane muhuḥ] W4
-\item[uttiṣṭhaty uttame prāṇā baddhe padmāsane muhuḥ] N20
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane mukuḥ] N26
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsana sthitaḥ] Bo2, J1, J3, N8, N11, N16, V2
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsana sthite] C4, N25
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthite] P21, P23, 
-\item[uttiṣṭhaty uttame prāṇāṃ rodhe padmāsane sthite] V19
-\item[uttiṣṭhaty uttame prāṇo rodhe padmāsane sthite] P20, V5, V28
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthitaḥ] J6, N2
-\item[uttiṣṭaty uttame prāṇa rodhe padmāsanasthite]  N5, P6, 
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsanaṃ muhuḥ] N3, , P15, V15
-\item[uttiṣṭhaty uttāme prāṇarodhe padmāsanaṃ muhuḥ] J5, J11, G1
-\item[uttiṣṭhaty uttāme prāṇarodhe padmāsano muhuḥ] J7
-\item[tattiṣṭhaty uttāmaprāṇarodhya padmāsanaṃ muhuḥ] J4
-\item[uttiṣ(?)yaty uttame prāṇa rodho padmāsanaṃ muhuḥ] N24
-\item[uttiṣṭhaty uttame prāṇaṃ rodhe padmāsanaṃ muhuḥ] P11, 
-\item[uttiṣṭhaty uttame kārye rodhe padmāsanae muhu] Ko,
-\item[uttiṣṭhaṃty uttame prāṇa rodho padmāsane sthitaḥ] N18
-\item[uṭṭiṣṭhaṃ cottare prāṇā baddhe padmāsane muhuḥ] N17
-\item[uttiṣṭhatyuttame prāṇarodhe padmāsane muhuḥ] G11
-\item[uttiṣṭaty attame prāṇe baddhe padmāsane dṛḍhe] N1
-\item[uttiṣṭhaty uttame prāṇe yāvat padmāsane samaḥ] N12
-\item[uttiṣṭhaty uttare prāṇā baddhe padmāsane dṛḍhe] J12,
-\item[uttiṣṭhaty unname prāṇe baddhe padmāsane dṛḍhe] N19
-\item[uttiṣṭhaty uttame prāṇa baddha padmāsanaṃ dṛḍhaṃ] N21
-\item[uttiṣṭety uname prāṇaḥ rodhe pavāsane sthite] N22
-\item[uttiṣṭhaty uttame prāṇa rāvai padmāsano muhuḥ] N23
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsano muhuḥ] V6
-\item[uṭṭiṣṭhnty uttame prāṇā baddhe padmāsane muhuḥ] LD1,
-\item[uttame sthānaṃ āpnoti tato vāyuṃ nibandhayet] P14, P18, 
-\item[(illegible/unavailable)]   Ba5, Bo3, C5, N4a, N6, P22, V22, V12, V9, T, P16, R5, 
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_019]
-\sthana{2.19c}
-\item[kṛśatā] A1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C6, C8, J1, J4, J5, J6, J7, J11, J13, J14, V19, N1, N2, N4a, N5, N8, N11, N12, N13, N16, N18, N19, N22, P2, P5, P6, P7, P11, P12, P13, P14, P15, P17, P18, P20, P23, P25, P26, P27, C4, V17, V10, V21, Tue, N23, N26, V7, V20, V5, V28, V22, V2, V18, G11, G1
-\item[kṛsatā] Ko, J3,  
-\item[kṛśate] J10pc
-\item[kriyate] C3, J10, J12, J16, J17, LD1, N7, N9, N10, P1, P8, P9, P10, R1, V12, C9, J2, N17, LD2, V19, Ba3
-\item[kṛyate] J15, N14
-\item[kṛśatāṃ] J8, V8, N21, V3, N25, P21, V6
-\item[kṛṣitāṃ] N24
-\item[kṛtā] N3, N20, V14
-\item[krūratā] J2
-\item[(illegible/unavailable)]   B1, B2, C5, N6, V16, V15, V13, V9, T, P16, P22, P24, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_021]
-\sthana{2.22a}
-\item[medaḥśoṣādikaṃ pūrvaṃ] P6, V1
-\item[medaḥśleṣmādikaṃ surve] V6
-\item[medaśokādikaṃ pūrvaṃ] N22
-\item[medaḥśleṣmādi nāśārthe] A1, C3, P26, 
-\item[medaḥśleṣmādi nāśārtha] LD2,
-\item[medaḥśleṣmādi ṣatkarmāṅi] V28
-\item[medaḥśleṣmādi nāśārthaṃ] B1, B3, Bo1, C2, J10, J12, J13, J15, J16, J17, LD1, N7, N17, P1, P2, P5, P8, P9, P12, P13, P17, R1, V21, V20, V12
-\item[medaḥśleṣmādhi nāśārthaṃ] P27, 
-\item[madaṃ śleṣmādi nāśārthe] P10, 
-\item[nādaśleṣmādinā śāstraṃ] Ba3
-\item[medaḥśleṣmādināśīrṣaṃ] N9
-\item[medaḥ tuṣmād adhikei purva śaṭ karmaṃ] V8
-\item[medaśleṣmādi nāśārthaṃ] C9, N1, N10, N14, N21, V18
-\item[medaśleṣādi nāśārthe] V17
-\item[medaśleṣmādi nāsārthe] N20, N26
-\item[medaśleṣma nivṛtyarthaṃ] Bo2, C1, J1, J3, J6, N2, N8, N11, N16, N18, P20, P21, P23, V15, V5, V2
-\item[medaḥśleṣma nivṛtyarthaṃ] N5, C4
-\item[??śleṣma nivṛtyarthaṃ] N25
-\item[medaḥśleṣmādhikaḥ pūrvaṃ] Ba, C8, J5, P7, P11, P14, P18, V10, V14, V13, G11
-\item[medaḥśleṣmādhikaḥ pūrva] J7,
-\item[medaḥśleṣmādhikaḥ arthaṃ] V16
-\item[medai śleṣmāṃdhika purṇa] J2
-\item[medaśleṣmāṃdhikaḥ pūrvaṃ] Ba5, Bo3, V3
-\item[medaḥ ślṣāṃdhikaḥ pūrvaṃ] C6, J14, 
-\item[medaśleṣmādhikaḥ pūrvaṃ]  J11, N3, N4a, N12, N13, P25, Tue, V22
-\item[medāśleṣmāṣikaḥ pūrvva] N23, V7
-\item[medaḥśleṣmaṇisapurṇa] N24 
-\item[mitaḥ sthūlādikaiḥ kāryyaṃ] B2
-\item[medodhikastu havabhiḥ] Ko,
-\item[(unavailable/illegible)] C5, J4, N6, N19, P15, V9, T, G1, P16, P22, P24, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_021d]
-\sthana{2.22d}
-\item[apy abhāvataḥ] V1
-\item[api bhāvataḥ] C3
-\item[samabhā] R1,
-\item[samabhāvataḥ] A1, B1, Ba5, Bo1, Bo3, C9, J1, J3, J6, J10, J11, J12, J13, J14, J15, J17, LD1, LD2, N3, N4a, N7, N8, N10, N11, N12, N13, P1, P5, P10, V10, N20, N21, P6, P8, P9, P11, P13, P14, P18, V14, V3, Tue, Ba3, N26, V7, V20, V15, V28, V22, V18, V12, V13, G11
-\item[samabhāvat] J16, N17, 
-\item[samupāgataḥ] J7,
-\item[samabhyāvataḥ] N24
-\item[api samācaret] V16, V17, V21
-\item[samabhāvata]  N9	
-\item[samabhāvanā] N22
-\item[smabhāvata]  N14
-\item[sabhāgataḥ] Ba,
-\item[samabhāgataḥ] B2, C6, C8, P7, V8, N18, N23, P25, P27, 
-\item[samabhāgatā] Ko,
-\item[samabhāgata] V6
-\item[samabhāgikaḥ] B3, C2, N1, P2, P12, P17, 
-\item[samabhārat] J2
-\item[samahāvata]  P26, 
-\item[samatā yataḥ]  Bo2, C1, C4, N2, N16, N25, P20, P23, V2
-\item[samatā yatāṃ]  P21, 
-\item[samatā yathā]   N5,
-\item[samatāpataḥ] V5
-\item[(unavailable/illegible)] C5, J4, N6, N19, P15, T, G1, P16, P22, P24, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_022a-c]
-\sthana{}
-  \begin{description}
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_023c]
-\sthana{2.22c}
-\item[vicitraguṇasandhāyī(i)] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C3, C8, C9, J1, J2, J3, J4, J5, J6, J7, J11, J13, J14, J15, J16, LD1, LD2, P7, V1, W4, J10, N17, J17, J8ac, Ko, N1, N3, N4a, N7, N9, N10, N16, P1, P2, P6, P9, P10, P11, P12, P13, P15, P17, P18, P21, P24, P25, P27, R1, V8, V16, V17, V10, N18, N19, N21, N22, P14, V14, V3, Tue, N23, V7, V15, V5, V28, V22, V2, V18, V12, V13, G11, Ba3
-\item[vicitraguṇasadhāyī] P26,
-\item[vicitraguṇaśandhāyī]  P8, 
-\item[vicītraguṇaṃsandhāyi] P5, 
-\item[vicitraguṇasaṃyāyi] P23,
-\item[vicitraguṇasandhāya]  C6, N12
-\item[vicitraguṇasandhāyā] N20, V20
-\item[vicitragunasandhāyai]  N14, V21, V6
-\item[vicitraguṇasandhāryā]  N13
-\item[vicitraguṇasandhāryaṃ]  J12,
-\item[vicitraguṇasandhānaṃ] N26
-\item[vicitraguṇasadyāpi] N2
-\item[vicitraguptasaṃ?jñopi] N5
-\item[vicitraguṇasaṃjñopi] C4, P20, N25
-\item[vicitraguṇasaṃdhāyoga]  N8, N11,
-\item[trividhaṃ guṇasandhāyi] B2
-\item[(unavailable/illegible)] C5, N6, N24, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_24c]
-\sthana{2.24c}
-\item[graset pratyāharec caitad] P25, 
-\item[tataḥ pratyāharec caitad] A1, Ba, Ba3, B3, G11, J12, J16, J17, N9, N12, N14, N17, N20, N21, N23, P1, P5, P8, P9, P10, P11, P12, P13, P15, P24, P26, P27, R1,
-\item[tataḥ pratyaharec cetad] N10
-\item[tataḥ pratyaharec caitad] N13, P7, 
-\item[tata pratyaharec caiva] N24,
-\item[punaḥ pratyaharec caitad] B1, B2, P2, P14, P17, P18, Tue,
-\item[punaḥ pratyahare caitad] J13, J14, 
-\item[punaḥ pratyāhare caitad] N19, 
-\item[punaḥ pratyāhare caitā] N22,
-\item[pūnaḥ pratyāharec caityāt] P6,
-\item[punaḥ pratyahared etad] N8, N11, N16, N18, N25, N26, P20, P21, P23, 
-\item[punaḥ pratyaharec caitaḥ] J15,
-\item[(unavailable/illegible)] N6, N7, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_24d]
-\sthana{2.24d}
-\item[prakṣālyaṃ dhautikarma tat] Ba, C6, P7, P25, V21
-\item[prakṣālanaṃ dhautikarma tat] V13
-\item[prakṣālye? dhautikarma tat] C8
-\item[ādāyaṃ dhautikarma tat] V14
-\item[udgāraṃ dhautikarma tat] V1,
-\item[sodgāraṃ dhautikarmavit] N12
-\item[uditaṃ dhautikarma tat] A1, Ba5, Bo1, Bo3, C9, J10pc, J12, J15, LD1, LD2, N4a, N7, N9, N10, N13, N14, N17, P5, P8, P9, P10, P14, P18, P24, V16, V17, V10, N20, N21, Tue, N26, V7, V20, V15, V18, V12
-\item[uditaṃ dhātikarma tat] P13, 
-\item[udita dhaitikarma tat] G11
-\item[aditaṃ dhautikarma tat] J17,
-\item[uditaṃ dhautikarma kṛt] J10, P1, R1, V22
-\item[up(?)itaṃ dhautikarma tat] C3
-\item[abhyāsād dhautikarmavit] B1, B2, C4, J1, J3, J6, J13, P2, P12, P17, P20, P26, V5, V28, V2
-\item[abhyāsād dhautikarmasu] N25
-\item[abhyāsā dhautikarmavit] N18
-\item[abhyāsā dhautikarma tat] P6,
-\item[abhyāsā dhautikarmavit] P21,
-\item[abhyāso dhautikarmavit] N19
-\item[abhyase dhautikarmavit] P27,
-\item[abhyāse dhautakarmavit] J14,
-\item[abhyāsād dhotikarmavit] C1
-\item[abhyāsād dhautikarmavat] N5
-\item[abhyāsaud dhautikarmavit] Bo2, C2,
-\item[abhyāsāsau dhautikarmavit] V6
-\item[abhyāsau dhautikarmatat] V8
-\item[abhyāsād dhautikarmavita] N2
-\item[abhyāsād dhautikarma tat] N1 (after correction),
-\item[dukhalātā dhautikarmataḥ] V3
-\item[duḥkhalṃ dhautikarma tat] Ko,
-\item[duṣālaṃ dhautikarma tat] J4, J11, 
-\item[duḥkhahaṃ dhautikarma tat] P11, 
-\item[uṣālaṃ dhautikarma tat] J5, P15, 
-\item[abhyāsa dhautakarka tat] N22
-\item[abhyāsād dhauti bhavati karma tat] N8, N11 ("bhavati" in brackets)
-\item[uṃtthānaṃ dhautikarma tat] N23
-\item[dākṣālaṃ dhautikarma tat] N24
-\item[utthānaṃ dhautikarma tat] J7,
-\item[dugdhāraṃ dhautikarma tat] J2
-\item[syā sā dhautikarmakṛt?] B3
-\item[?s?vāla dhautikarmakṛt] N3
-\item[(unavailable/illegible)] C5, N6, N16, T, G1, P16, P22, P23, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_25d]
-\sthana{2.25d}
-\item[bādhate yaṃ na] P27,
-\item[chadhyate ca na] J16,
-\item[dhāvaty eva] V1, V8, V3, V2
-\item[dhāvaṃtyeva na] P6, P15, 
-\item[dhāvate ca na] B2
-\item[dhāvaṃta ca na] N19
-\item[dhāvaṃte ca na] J5,
-\item[dhavaṃte ca na] N22
-\item[dhruvntye neva] Ko,
-\item[gachaty eva na] N21, P13, 
-\item[gachaṃty eva na] Bo1
-\item[prayāntyeva na] Ba, N4a, Bo2, Bo3, C1, C4, C6, C8, J1, J3, J6, J7, N5, N8, N11, N13, N16, P7, P11, P14, P18, P20, P21, P25, V10, Tue, N25, V28, V22
-\item[dhāvatyeva na] G11
-\item[prayānteva na] N23, V7
-\item[śudhyatty eva] A1,
-\item[śudhyaty eva] LD2, P2, P9, 
-\item[śudhyate nātra] J12,
-\item[śudhyate ca na] N10, P26, 
-\item[śudhyanty eva] B1, C3, C9, J10, J13, J17, N1, N17, P5, P8, P12, P17, Ba3
-\item[śudhyanteva] B3, P10, 
-\item[śudhyaṃteva na] J15, N9, N14, P24, V16, V17
-\item[śvadhvatyeva na] R1,
-\item[śudhyaṃte ca na] LD1, N7, P1, 
-\item[suddhyaṃtevaṃ na] N18
-\item[śudhyaṃ sevana] C2,
-\item[kṣīyaṃte saklā malā] J14,
-\item[?īyatsyeva na] N2
-\item[bhavaṃty eva na] N3, V14
-\item[naśyaṃteva] N12
-\item[naśyaṃte] J11
-\item[nāśaṃ yāṃti] J4, 
-\item[naśyaṃty eva na] J2,
-\item[jayaty eva na] N24
-\item[śuṣyaṃte sakalāmalāḥ] N20, N26
-\item[(illegible/unavailable)] Ba5,C5, N6, V21, V6, V20, V15, V5, V18, V12, V13, T, G1, P16, P22, P23, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_26a]
-\sthana{2.26a}
-\item[udaragatapadārtham udvahatī] J12,
-\item[udaragatapadārthaṃ  utbhavanti] G1
-\item[udaragatapadārtham udvahati] R1,
-\item[udaragatapadārtham udvahantī] Ba, C3, J1, J7, Ko, N10, P6, V1, V16, N12, N18, N22, N23, P20, P27, V6, V28, V2
-\item[udaragatapadārthamūrdhāhatī] J16,
-\item[uragatapadārtham udvahantī] B2,
-\item[udaragatapadārthān udvahaṃti] C6, P7, 
-\item[udaragatapadārtham uşmahantī] V17
-\item[udaragatapadārtham udvamaṃti(ī)] B1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J11, J14, LD1, LD2, N7, N13, N16, N16, N20, N21, N26, P2, P8, P9, P11, P12, P13, P14, P17, P18, P21, P25, P26, V3, V20, V22, V18, V12, V13, Ba3
-\item[udaragatapadārtham udvayant] P5,
-\item[udaragatapadārtham uddhayaṃt] P24, 
-\item[udaragatpadārtham udvamaṃtī] N9,  
-\item[udaragatapadārtham udvavaṃtī] J10, J17, P1,
-\item[udaragatapadārtham udvamaṃta] J4, P10, 
-\item[udaragatapadārtham udvamaṃte] P15, V15
-\item[udaragatapadārthām udyamaṃte] N19
-\item[udaragatapadārthaṃ śuddhapetaṃ] N25
-\item[udaragatapadyarthān udvamaṃtī] A1
-\item[udaragatapadārtham uḍhasaṃtī] N24
-\item[udaragatapadārtham udvapanti] N14, 
-\item[udaragatapadārthaśodhanaṃ syāt] Bo1
-\item[udaragatapadārtha śuddhayeta] C4
-\item[udaragatapadārtha śuddhaye vai] J15, 
-\item[udaragatapadārtha suddhayeta] J13,
-\item[udagatapadārtha uddhanaṃtī] J5,
-\item[(illegible/unavailable)] Bo3, C1, J2, N6, N8, N11, R5, V19, V10, V14, Tue, C5, V21, V7, V5, V22, G11, T, P16, P22, P23, 
-  \begin{description}
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_26b]
-\sthana{2.26b}
-\item[pavanam apānam udīrya kaṇṭhanāle] V1, A1, B2, Ba, Ba5, C6, C8, C9, J3, J4, J6, J7, J10, J11, J14, J15, J17, Ko, LD1, LD2, N7, N9, N10, N13, N16, N17, P6, P7, P8, P9, P10, P13, P18, W4, J8, V16, N18, N20, N22, P14, P15, P20, P25, P27, V3, N23, V20, V15, V28, V18, V12, V13, G1
-\item[pevanam apānam udīrya kaṇṭhanāle] P5, P24, 
-\item[pavanam apānam udīrya kaṇṭhānālī] Ba3
-\item[pavanam apānam udīrya kaṇṭhanālaiḥ] N12, 
-\item[pavanam apānam udārya kaṇṭhanālo] J16,
-\item[pavanām apānam udīrya kaṇṭhanālṃ] J1
-\item[pavana apānam udīrya kaṇṭhanāle] V6
-\item[pavanam upānam udīrya kaṇṭhanāle] C3
-\item[pavanam upānam urīrya kaṇṭhanāle] C4
-\item[pavanam apānam udīrya kaṇṭhanālo] N19, R1,
-\item[pavanam apānam udīrya kaṇṭhanālāt] B3, C2, J13, N26, P1, P2, P12, P17, P21, P26, V2
-\item[pavanam apām udirya kaṃnāle] N25
-\item[pavanam apāsenam udāryaṃ kaṇṭhanāle] J5,
-\item[pavanam apānam udīryakaṃ jāle] P11, 
-\item[pavanam mudārya kaṇṭhanālo] J12
-\item[pavanam udirya kaṇṭhanāle] N21
-\item[mapānam udīrya kaṇṭhanāle] B1,
-\item[pavanamapānam udīrya kaṇṭhanāle] Bo1, Bo2,
-\item[pavanam athānam utiya kaṇṭhanāle] N24
-\item[pavanam apānam udīrya karāḍanāle] V17
-\item[pavanam apānam udīrya kaṇṭhaḍālai] N14, 
-\item[(illegible/unavailable)] Bo3, C5, J2, N6, N8, N11, R5, V19, V10, V14, V21, Tue, V7, V5, V22, G11, T, P16, P22, P23, 
-\end{description}
- \end{marma}
-
- \begin{marma}[hp02_26c]
-\sthana{2.26c}
-\item[karibhir iva jalasya vāyuvegāt] A1, B1, B3, C2, C9, J1, J3, J6, J8, J10, J17, LD1, N7, N10, N16,  N17, P1, P2, P9, P12, P17, W4, V16, V17, N18, N20, N26, V2, V18, V12, Ba3
-\item[karibhir iva jalasya vāyuvegaj] R1,
-\item[karibhir iva gajasya vāyuvegāt] V20
-\item[karibhir iva jalasya vāyuvegā] B2, C3, J12, J16, N14, P8, 
-\item[karibhir iva jalasya vāyubagā] P26,
-\item[karibhir iva jalasya vāyuvege] N9,
-\item[karibhir iva jalaṃ ca vāyuvegā] Ba,
-\item[karibhir iva jalaṃ ca vāyuvegāt] P25, 
-\item[karibhiriva jalaṃsya vāyuvegāt] Bo2, 
-\item[karibhiriva jalasya vāyuvego] J13,
-\item[karibhiriva jalasya vāyuyogāt] V28
-\item[karibhirivadāyikā jalabhya vāyuvegāt] P21, 
-\item[kramamarivayavasya vāyur yo] V1
-\item[kramaparicayavaśya nāḍicakrā] Ba5, N13, P14, P18,  
-\item[kramaparicayavaśyā nāḍīmārgā] C6, J7, LD2, 
-\item[kramaparicayavaśyā nāḍīmārgāt] N12,
-\item[kramaparicayavaśya vāyumārga] C8
-\item[kramaparicayavaśya vāyumārgā] P7, P20, 
-\item[kramaparicayavasya vāyuvegā] P27,
-\item[kramaparicayavaśa nāḍimārgā] P11,  
-\item[kramaḥ paricayavasya vāyuḥ mārgre] P6,
-\item[kramaparicayavasya vāyumārgre] J14,
-\item[kramaparicayavaśya vāyumārgāt] Bo1
-\item[kramaparicayavaśya vāyumārgaṃ] N21, P13, 
-\item[kramac pacya paricaya vaśyā vāyumārgaṃ] J5
-\item[kramaparicayavaśyaṃ vāthagargāṃ] N24
-\item[kramaparicayavaśya mārgte(??)] N23
-\item[kramaparicayavasya vāyumārge] N22
-\item[kramapārīcita vāyumārge] J15,
-\item[kramaparījita vāyuḥ] P15, 
-\item[kramapārīcitavacana vāyumārgā] P5, P24, 
-\item[kramaparicitavasya vāyuyogo] Ko,
-\item[kramapārīcitavaśya vāyumārg(oṃ)?] N19
-\item[kramapārīcitavaśya vāyumārgo] J11
-\item[kramapārīcitavaśya vāyumārge] J4, 
-\item[kramapārecitavaktu? vāyumārge] C4
-\item[kramaparicittakakrū vāyumārge] N25
-\item[kirimir iva jalasya vāyuvegāt] P10, 
-\item[(illegible/unavailable)] Bo3, C1, J2, N6, N8, N11, R5, V19, V10, V14, Tue, C5, V21, V6, V7, V15, V5, V22, V13, G11, T, G1, P16, P22, P23, 
-  \begin{description}
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_26d]
-\sthana{2.26d}
-\item[gajakaraṇānti nigadyate haṭhajṇau] Ko,
-\item[gajakarīṇīti kathyate haṭhavidyā] G1
-\item[gajakaraṇīti nigadyate haṭhajñaiḥ] B2, Ba, Ba5, Bo1, C3, C4, C6, J4, J6, J7, J11, J15, N12, N13, P5, P6, P7, P9, P10, P14, P18, P24, P26, P27, V1, N19, N23, N25, P13, V20, V15, V13
-\item[gajjakaraṇīti nigadyate haṭhajñaiḥ] P8, 
-\item[gajakaraṇī nigadyate haṭhajñaiḥ] N16, 
-\item[gajakaraṇīti nigadate haṭhajñaiḥ] J14,
-\item[gajakaraṇīti nigadyāta haṭhaj~naiḥ] J5,
-\item[gajakaraṇīti nigadyaṃte haṭhajñaiḥ] LD2,
-\item[gajakaraṇīti nigadyate] N24, V2
-\item[gajakarṇīti nigadyate haṭhaj~naiḥ] A1, C8
-\item[gajakaraṇīva nigadyate haṭhaj~naiḥ] B1
-\item[gajakariṇīva nigadyate haṭhajñaiḥ] C2
-\item[gajakaraṇīti nigachate haṭhaj~naiḥ] B3
-\item[gajakāraṇī nigadyate haṭhaj~naiḥ] Bo2,V3, V6
-\item[gajakaraṇīti nigadyate haṭhakṣaiḥ] N22
-\item[gajakaraṇī ca nigadyate dṛḍhajñaiḥ] J13,
-\item[gajakaranīti nigadye haṭhajña] N14, 
-\item[gogajakaṇīti nigadyate haṭhajñaiḥ] P15, 
-\item[jalagajakaraṇī nigadyate haṭhaj~naiḥ] J1, J3, 
-\item[jalakaraṇīti nigadyate haṭhajñaiḥ] C9, J12, J16, LD1, N7, N17, P1, W4, N20, N26, P21, P25, 
-\item[jalakaraṇīti nigadyato haṭhajñaiḥ] N10,
-\item[jalakaraṇīti nigadyate mahāhaṭhaj~naiḥ] V21
-\item[jalakariṇīti nigadyate haṭhajñaiḥ] J10, J17, N9, P20, R1, V16, V28, V18, V12, Ba3
-\item[jalakaraṇānī nigadyate haṭhaj~naiḥ] V17
-\item[jalagajakaraṇīti nigadyate haṭhaj~naiḥ] J8, N18, P2,
-\item[jalakaraṇīva nigadyate haṭhajñaiḥ] P12,
-\item[jalagajakaraṇīva nigadyate haṭhajñaiḥ] P17, 
-\item[gajakaraṇīti nirādyate haṭhaj~naiḥ] N21,
-\item[najakariṇīti nigadyate haṭhajñaiḥ] P11, 
-\item[(illegible/unavailable)] Bo3, C1, J2, C5, N6, N8, N11, V19, V10, V14, Tue, V7, V5, V22, G11, T, P16, P22, P23, R5,
-  \begin{description}
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_29c]
-\sthana{2.29c}
-\item[aseṣadoṣān api yan nihanyād] Ko, 
-\item[aśeṣadoṣopacayaṃ nihanyad] J8, N17, V1, W4, V16, V17, V10, V6, V7, V15, 
-\item[aśeṣadoṣoyathā nihanyānād] V14
-\item[aśeṣadoṣāmaya śoṣaṇāya] G1
-\item[aśeṣadoṣopacayaṃ nihanyād] B1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C6, C8, C9, J2, J3, J5, J6, J7, J10, J11, J13, J14, J15, J17, LD1, LD2, N7, N8, N11, N13, N16, N18, N19, N20, P1, P2, P5, P8, P9, P12, P13, P14, P15, P17, P18, P20, P21, P24, P25, P27, Tue, R1, N25, N26, V20, V5, V28, V22, V2.
-V18, V12
-\item[aśeṣadopacayaṃ nihanyād] P23,
-\item[aśeṣa doṣā pacayan niśamyaha nyāda] G11
-\item[aśeṣadoṣo ca nayaṃti hanyād] N10,
-\item[aśeṣadoṣopacaya no hanyād] J16,
-\item[aśeṣadoṣopacayā nihanyād] V13
-\item[aśeṣadoṣopacayān nihanyād] N12,
-\item[aśeṣadoṣapacayaṃ nihanyād] N9,
-\item[aśeṣadoṣopacayaṃ nihanti] P7, 
-\item[aśeṣadoṣopacayaṃ nihanyā] N23,
-\item[aśeṣadoṣopacaya nihanyād] N24, P10, P26, 
-\item[aśeṣa?oṣovacayaṃ nihanyād] N21,
-\item[aśeṣadoṣasya cayaṃti hanyād] P6,
-\item[aśeṣadoṣepacayaṃti hanyād] J12, N14, 
-\item[aśeṣa doṣo vacayan nihanyād] Ba3
-\item[aśeṣadoṣepacayaṃ nihanyad] C2,
-\item[aśeṣadoṣāpacayaṃ nihanyad] V21, V3
-\item[aśeṣadoṣopacayaṃ nihatyad] P11, 
-\item[aśeṣadoṣopacitiṃ hathajñād] J4,
-\item[aśeṣadoṣopacayaṃ ca hanyad] A1
-\item[aśeṣadoṣaprabhavaṃ nihanyad] V19
-\item[aśeṣadoṣasya cayana hatyā] N22
-\item[(illegible/unavailable)] C5, N6, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_29d]
-\sthana{2.29d}
-\item[abhyasyamānaṃ jalavastikarma] Ba5, Bo1, Bo2, Bo3, C1, C3, C6, C8, J1, J2, J4, J5, J6, J7, J8, J11, J12, J13, J14, J15, J16, LD1, N8, N9, N10, N11, N12, N13, N14, N16, N17, P1, P2, P8, P9, P11,  P12, P14, P15, P18, P20, P23, W4, V16, V17, V10, N18, N19, N21, P13, P25, P27, V21, V3, Tue, N26, V6, V7, V20, V15, V5, V28, V22, V2, V18, V13, G11, G1
-\item[abhyasyamānaṃ janabastikarma] R1,
-\item[abhyasyamānaṃ jalavāstikarma] P5, P24, 
-\item[anyasyamānaṃ jalavastikarma] J3,
-\item[abhyasyamānaṃ nijavastikarma] V12
-\item[abhyasyamānaṃ jalavastikarmaḥ] Ko, P6, 
-\item[abhyamānaṃ jalabastikarma] P21, 
-\item[abhyasyamānaṃ khalu vastikarma] C9,
-\item[abhyasyamāna ja?vastikarma] N20(something seems added above)
-\item[abhyasyamānaṃ jalavastitakramāt] N24
-\item[praamāńam jalavastikarma] V14
-\item[abhyasyamāno jalavastikarma] B3, C2, P17, 
-\item[abhyasyamā jalavastikarma] B1,
-\item[abhyāsamānaṃ jalavastikarma] C4, V1, N7, N25, P10, P26, 
-\item[abhyāsyamānaṃ jalavastikarma] A1, Ba,
-\item[abhyāsamānaṃ jalavasti] V19
-\item[abhyasya mānam jalapastikarma] Ba3
-\item[abhyasyamānaṃ javastikarma] J10, J17,
-\item[abhyasyamānaṃ jvalabastikarmaḥ] LD2,
-\item[bhyasyabhānaṃ jalavestikarma] N23(space for “da” left blank with an overhead line)
-\item[abhyasya śūnyaṃ jalavastikarma] B2,
-\item[cābhyasyamānaṃ jalavastikarma] P7, 
-\item[dattasya mānaṃ yanabasyakarmaḥ] N22
-\item[(illegible/unavailable)] C5, N6, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_30]
-\sthana{2.30}
-\item[atha natīḥ] J1
-\item[atha ti] P26, 
-\item[atha netī(ḥ)] Ba, Ba5, Bo1, Bo2, C1, C2, C4, J3, J6, J8, N8, N11, N16, V1, V10, N18, P20, P21, P23, V21, V3, N23, N25, N26, V7, V5, V2
-\item[atha neti] J2, Ko, N7, N9, N10, N13, N14, N20, N24, P7, P9, V6, V28
-\item[atha netiḥ] J7, N12, P14, P18, P25, Tue, V22, V12, Ba3
-\item[atha netikarma] C3, C9, J10, J17, LD1, P1, P2, P5, P8, P10, P24, W4, V16, V17, N21, P13, V14, V15, V18, V13
-\item[atha netikarmaḥ] C8, J12, J15, N22, P6, 
-\item[atha netīkarma] A1, B1, B2, C6, J13, J14, J16, N17, P12, P17, P27, R1, V20
-\item[atha netīkarmaḥ] LD2
-\item[atha nītī] J5,
-\item[atha nīti] P11, 
-\item[atha nītikarma] J4,
-\item[atha nītīkarma] N19
-\item[prathama netīkarma] B3
-\item[(illegible/unavailable)] Bo3, C5, N6, J11, G11, P15, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_30c]
-\sthana{2.30c}
-\item[mukhanirgamanād eva] A1, B2, C3, J10, J17, LD1, N7, N9, N17, P1, P5, P6, P8, P9, P10, P24, P26, V1, W4, V16, N22, V20, V18, V12, V13, Ba3
-\item[mukhanirgamanād evaḥ] J15,
-\item[mukhanirgamanād deva] C9
-\item[mukhanirgatanād evaṃ] J12,
-\item[mukhanirgatamanād eva] V14
-\item[mukhanirgamayedeti] J2
-\item[mukhan nirgamanād eva] B3,
-\item[mukhān nirgamanād eva] C2, P12, P17, 
-\item[muvi nirgamanād eva] R1,
-\item[mukhā nirgamanād eva] P2, 
-\item[mukhṃ nirgamanād eva] N14, 
-\item[mukha nīrgamanod eva] J16, 
-\item[mukhān nirgamayec caiṣā] Ba, Ba5, Bo1, C8, J8, J11, LD2, N12, N13, V17, N19, N20, P14, P18, P25, V21, Tue, N26, 
-\item[mukhān nirgamyate caiṣā] J4,
-\item[mukhān nirgamayed eṣā] P13, 
-\item[sukhān nirgamayeccaiṣā] G11
-\item[mukhān virgamayed eṣā] N21
-\item[mukhā nirgamayec caiṣā] J7, P15, V10, V3, V6, V7, V15, V22
-\item[mukhe nirgamayec caiṣā] C6, P7, 
-\item[mukhe nirgamaye caiṣā] P11, 
-\item[mu nirgamayai caiṣā] J5,
-\item[mukhān nirgamaye doṣā] N24
-\item[mukhān nirgamayet sāpi] V19
-\item[mukhān nirgāmayet sā hi] B1, Bo2, J13, J14, N11, N18, N25  
-\item[mukhān nirgamanāt sā hi] P21, P23, 
-\item[mukhān nirgamayet sā hi] J6, N8, N20, P20, P27
-\item[mukhā nirgāmaye sā hi] V2
-\item[mukhā nirgāmayet so hi] Ko,
-\item[mukhā nirgamanā sā hi] V5
-\item[mukhān nigarmayet sā hi] C4, J3, N16, 
-\item[mukhā nigarmayet sā hi] J1, V28  
-\item[mukhān nirgatya nāsāyam] C1,
-\item[mukhaśca nirgamec caiṣā] N23
-\item[sukhanirgamanād eva] N10,
-\item[mukhaṃ nirgamayedeṣu nīti] G1
-\item[(illegible/unavailable)] Bo3, C5, N6, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_31cd]
-\sthana{2.30cd}
-\item[] 
-\item[(illegible/unavailable)] 
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_33a]
-\sthana{2.33a}
-\item[ghaṭanaṃ] J2, 
-\item[muñcakaṃ] Ko,
-\item[mucyate] J4,
-\item[mocanaṃ] B2, Ba5, C8, J7, J12, N12, N13, N18, N21, N22, P13, P14, P16, P18, P26, Tue, N23, V7
-\item[mocakaṃ] Bo2, C1, J1, J3, J5, J6, J11, J14, N16, N19, P15, P20, P21, P23, V28
-\item[moṭakaṃ] V1, Bo1, N25 
-\item[meṭakaṃ] P11, 
-\item[mokṣakaṃ] N8, N11, 
-\item[toṭakaṃ] V19, V16
-\item[tāṭakaṃ] V17
-\item[trāṭakaṃ] A1, C4, C6, P6, P7, P27, V10, V14, V21, V3, V6, V20, V15, V5, V28, V22, V2, V18, V12, V13, G11, Ba3, G1
-\item[troṭakaṃ] N24
-\item[rocanaṃ] Ba, P25, 
-\item[sphoṭanaṃ] B1, B3, C2, C3, C9, J8, J10, J13, J15, J16, J17, LD2, N7, N14, N17, P1, P8, W4, N20, N26, P2, P5, P9, P10, P12, P17, P24, Ba3  
-\item[sphoṭana] R1,
-\item[spheṭanaṃ] N10,
-\item[sphoṭakaṃ] LD1,
-\item[sphoṭa] N9, 
-\item[(illegible/unavailable)] Bo3, C5, N6, T, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_33b]
-\sthana{2.33b}
-\item[jadrānāṃ ca kapāṭakaṃ] J16,
-\item[tandrādīnāṃ gapāṭa[ka]m] V1,
-\item[tandrādīnāṃca pāṭanaṃ] P15, V19,
-\item[ta.ntryādīnāṃ kavāṭakam] G1
-\item[tandrādīnāṃ ca kapāṭakaṃ] B1,
-\item[tandrādīnāṃ kapāṭakam] B2, B3, Ba, Ba5, Bo1, C1, C2, C4, C6, C8, J3, J5, J6, J7, J8, J11, J12, Ko, LD1, LD2, N7, N8, N11, N13, N16, P2, P6, P7, P12, P13, P14, P16, P17, P18, P23, P25, P26, P27, W4, V17, V10, N21, N22, V3, Tue, V7, V5, V28, V22, V2, V13, Ba3
-\item[tadrādīnāṃ kapāṭakam] P21,
-\item[tandrādīnāṃ kavāṭakam] G11
-\item[tandrānāṃ tu kapāṭakam] N14,
-\item[tandrādīnāṃ kṛpāṭakam] N12, 
-\item[tandrādīṇāṃ kapāṭakam] J13, 
-\item[tandrāṇāṃ ca kapāṭakam] J15, N9, N10, P1, P8, P10, R1, 
-\item[tadbhānaṃ ca kapāṭakam] P5, P24, 
-\item[tandrādīnāṃ kapālakam] J1
-item[tandrāegaṃ kathapāṭakam] V21
-\item[tandrādīnāṃ kapāṭakai] Bo2
-\item[tandrādiṇāṃ kapāṭakam] J8, N18, N23,
-\item[tandrādīnāṃ kaṃpāṭa] P9, 
-\item[tandrādīn kapāṭakam] P11,
-\item[tandrādiṇāṃ ca pāṭakam] J14, N19
-\item[tandrādīṇāṃ ca pāṭakam] C9
-\item[tandrādīṇāṃ ca pāṭhakam] V15
-\item[tandrādīṇāṃ ca pāṭavam] J4,
-\item[tandrāṇāṃ ca kapāṭakam] C3, J10, J17, V16, N20, V14, N26, V20, V18
-\item[tadvāṇāṃ ca kapāṭakam] A1,
-\item[tandrāṇāṃ tu kapāṭakam] N17
-\item[sandrādīnāṃ kapātakāṃ] N24
-\item[nidrādīnāṃ kapāṭakam] P20,
-\item[nidrārttīnāṃ kapāṭakam] N25
-\item[hetādīnāṃ? kapāṭhakam] J2,
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, V6, V12, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_34b]
-\sthana{2.34b}
-\item[gudaṃ] V1
-\item[gopyaṃ] P1, V16, V17, V10, N14, N18, V14, V21, V3,V6, V7, V20, V15, V28, V22, V2, V18, V12, V13, G11, Ba3, G1
-\item[tuda] N10,,
-\item[tundaṃ] A1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J7, J8, J13, J14, J16, LD1, LD2, N8, N9, N11, N13, N16, N19, N20, P2, P17, P18, P20, P26, Tue, N26, G11, P11, 
-\item[tunduṃ] N23
-\item[tunda] N12, N17 P16, 
-\item[tunde] C1, C4, P21, P23,  
-\item[tuṃḍe(?)] N25, V5
-\item[tundat] B2, J10, J17, P8, P10, R1, 
-\item[tundet] J15, 
-\item[truṃdaṃ] N21, P15, 
-\item[tuda] W4
-\item[tudaṃ] B1, C3, C6, J1, J5, J12, P6, P7, P12, P13, P14, P27, 	
-\item[tundan] J4,
-\item[tudan] Ba, Bo1, J11, P9, P25,  
-\item[tudat] N7, P5, P24, 
-\item[tulyaṃ] Ko,
-\item[tujyaṃ] V19
-\item[t(?)uddhaṃ] N22
-\item[puṃda] N24
-\item[saṃdada] J2
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_35a]
-\sthana{2.35a}
-\item[pāvanādi] V1
-\item[pāvanāni] N18,
-\item[pāvanāgniḥ] N25
-\item[pāvanādi] Ko,
-\item[pāvakādi] C3, J10, J16, N10, Ba3, R1, 
-\item[pāvakādī] J12,
-\item[pāvakāgni] N26
-\item[pācakādi] C2, C9, J17, LD1, LD2, N7, N9, N14, N17, P1, P5, P8, P9, P10, P24, W4, V16, V17, N20, V14, V21, V18, V13
-\item[pācatādi] Bo2, 
-\item[pācanādi] A1, B1, B2, B3, Ba, Ba5, Bo1, C6, C8, J4, J5, J7, J8, J11, J14, N12, N13, P6, P7, P14, P17, P25, P27, V19, V10, N19, N21, P2, P12, P13, P18, P20, V3, Tue, N23, V6, V7, V20, V15, V28, V22, G11, G1
-\item[pācanādīn P26, 
-\item[pācanāgni] J3, J6, J13, N8, N11, N16, P16, P23, 
-\item[pācanāgnī] P21,
-\item[pācakāgniḥ] J15
-\item[pācakāgni] V2
-\item[pācakāri?] J2,
-\item[pācanānī] P11, P15, 
-\item[pācanāṃnī] N24
-\item[kārkādeśaṃ] N22
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, V5, V12, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_35c]
-\sthana{2.35c}
-\item[apa śoṣaṇī ca] Ba5,
-\item[mala śoṣiṇī va] V1,
-\item[mala śoṣaṇī ca] B2, C3, J8, LD1, N9, N17, N20, P1, P5, P10, P16, P24, W4, V16, V14, V3, V7, V18, V12
-\item[maya śoṣaṇī ca] V21, V15, V28, V13
-\item[maya śodhaṇī yā] N23
-\item[maya śodhaṇī cā] V6, 
-\item[mala śoṣaṇīva] J10, J17, P9,
-\item[mala śoṣaṇī vaḥ] J16,
-\item[doṣāmaya śoṣaṇī ca] C1, C2, C4, C6, C8, C9, J3, J4, J11, J13, J14, Ko, N12, N13, N18, N21, P2, P6, P7, P11, P12, P13, P14, P15, P17, P18, P20, P21, P23, P27, Tue, N25, V22, V2, G11, Ba3
-\item[doṣāmaya śauṣaṇī caḥ] J15, 
-\item[doṣamala śoṣaṇī ca] P8,  
-\item[doṣāmala soṣaṇi ca] N14, R1,  
-\item[doṣān mala śoṣaṇī ca] J12, N10, 
-\item[doṣāmaya śoṣaṇīya] J1
-\item[doṣāmaya śoṣaṇīyaṃ] J2
-\item[doṣāmaya śoṣiṇī ca] A1, B3, Ba, Bo1, J6, N8, N11, N16, V19, N19, P25, 
-\item[doṣāmaya śoṣiṇāṃ ca] Bo2
-\item[doṣāmaya śodhinī ca] J7, 
-\item[doṣāmaya śodhanī ca] P26,
-\item[doṣābhayahā soṣaṇī ca] J5,
-\item[dauṣāmaya śoṣaṇī ca] LD2,
-\item[doṣāmaya goṣaṇī ca] N24
-\item[mala śoṣiṇī ca] V17, N7, N26,
-\item[doṣāṇapi śoṣaṇā ca] B1,
-\item[doṣāmaya śoṣaṇī ca] V10
-\item[doṣapra śoṣiṇī] V5
-\item[doṣānala śoṣaṇī bahava] N22
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, V20, R5, T, G1
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_35d]
-\sthana{2.35d}
-\item[mūlam iyaṃ hi] J14, P15, V1
-\item[mūlam idaṃ hi] Ko, J4, J11, N19
-\item[mau.riyaṃ hi] G1
-\item[mūlam ayaṃ ca mauliḥ] N22
-\item[maulir iyaṃ hi] A1, C1, C4, J3, J5, N8, N11, N16, P6, P20, P23, P27, V19, V21, N25, V15, V2
-\item[maulir iyaṃ] J1
-\item[maulir iyaṃ ca] Ba, Ba5, Bo1,Bo2, C6, C8, J6, J7, N12, N13, N21, Tue, N23, P7, P11, P13, P14, P16, P18, P21, P25, V6, V7, V20, V28, V22, G11
-\item[saulir iya ca] N24
-\item[maulir ayaṃ hi] J2
-\item[sau jayatīha] J10, J17, N17, W4, N18, N20, N26
-\item[jayatīha] B1, B3, C2, C3, C9, J8, J12, J13, J15, J16, LD1, LD2, N7, N9, N10, N14, P1, P2, P5, P8, P9, P10, P12, P17, P24, P26, R1, V3
-\item[sauryakarīha?] B2
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, V16, V17, V10, V14, V5, V18, V12, V13, T, Ba3, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_38a]
-\sthana{2.38a}
-\item[prāṇāyāmena vai] Ko, J4, J11, J14, V1, N19, P15, V15
-\item[prāṇāyāmair eva] B1, B2, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J12, J13, J15, LD2, N1, N3, N4a, N5, N8, N11, N12, N13, N16, N18, N22, P2, P7, P11, P12, P14, P17, P18, P20, P21, P23, P25, V3, Tue, N23, N24, N25, V6, V7, V5, V28, V22, V2, V12, V13, G11, G1
-\item[prāṇāyāmaiḥ sukhāt] A1, C3, C9, J10, J17, LD1, N2, N9, N10, N17, P1, P5, P8, P9, P10, P24, P26, R1, Ba3
-\item[prāṇāyāme sukhāt] J16,
-\item[prāṇāyāmais tu sā sarvaiḥ] P16,
-\item[prāṇāyāmaiḥ sukhāḥ] N26
-\item[prāṇāyāmair mu?khāṃ] N20
-\item[prāṇāyāmena sarvepi] N21, P6, P13, P27, 
-\item[praṇāyāme kṛte samyak] J2
-\item[prāṇāyāme śukhā] N14, V18
-\item[prāṇāyāme tataḥ] V16, V17, V10, V14, V21, V20
-\item[praṇāyāmāt sukhāt] N7
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_38b]
-\sthana{2.38b}
-\item[malā api] J4, J11, V19, N21, P6, P13, P15, P27, V15, V28
-\item[malād api] Ko,
-\item[malā iti] Ba, Ba5, Bo1, Bo2, C1, C6, C8, J1, J2, J3, J5, J6, J7, LD2, N3, N4a, N5, N8, N11, N13, N16, P7, P11, P14, P18, P21, P23, P25, V10, N18, N19, P20, Tue, N23, N24, N25, V6, V7, V22, G11, G1
-\item[malā itiḥ] J15,
-\item[malā ime] J14, 
-\item[malāśayaḥ] J10,
-\item[malāśayāḥ] A1, B1, B3, C2, C9, J13, LD1, N1, N2, N10, N17, P5, P10, P12, P17, P24, P26, R1, V17, V14, V20, V18, V13
-\item[malākulaṃ] B2, N22
-\item[malākulāḥ] N20, N26
-\item[malāśayā] C3, C4, J16, J17, N9, P1, P2, P8, P9, P16, V16
-\item[marāśayā] N14
-\item[malāsayā] J12, V21, V3, V2
-\item[malāśrayāḥ] N7
-\item[malāyinaḥ] N12
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, P24, V5, V12, T, Ba3, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_39ab]
-\sthana{2.39ab}
-\item[brahmādayo 'pi tridaśāḥ pavanābhyāsatatparāḥ] A1, B2, B3, Ba, G1, P11, P12, P13, P14, P15, P16, P18, P20, P27, P21, P23, P25, P26,
-\item[brahmādayo 'pi tṛdaśāḥ pavanābhyāsatatparāḥ] P17,
-\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparāḥ] B1
-\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparaḥ] R1,
-\item[(illegible/unavailable)] P22, P24, R5, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_39cd]
-\sthana{2.39cd}
-\item[abhūvannantakabhayāt tasmat pavanam abhyaset] G1,
-\item[abhūvannantakabhayāt tasmāt pavanam abhyaset] A1, Ba, P11, P13, P14, P15, P18, P20a, P27,
-\item[abhūvannātankabhayāt tasmāt pavanam abhyaset] B1,
-\item[abhūvannatakabhayāṃ tasmat pavanam abhyaset] P26,
-\item[abhavannantakabhayāt tasmat pavanam abhyaset] P25,
-\item[abhūvan nanekabhayāt tasmat pavanam abhyaset] P16,
-\item[te ca siddhiṃ gatāḥ sarve tasmat pavanam abhyaset] P12,
-\item[tena siddhi gatās te ca tasmat pavanam abhyaset] P21, P23, 
-\item[tena siddhiṃ gatās te ca tasmat pavanam abhyaset] B2, 
-\item[tena siddhi gatāḥ satyaḥ tasmat pavanam abhyaset] P20,
-\item[naiva siddhiṃ gatāḥ sarve tasmat pavanam abhyaset] B3, P17,
-\item[sumanskāṃtako bhavyaḥ tasmat pavanam abhyaset] R1,
-\item[(illegible/unavailable)] P22, P24, R5, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_40b]
-\sthana{2.40b}
-\item[cirāśrayam] LD2,
-\item[nirāśrayam] A1, Ba, C1, C3, C6, C8, C9, J2, J4, J5, J7, J8, J10, J11, J15, J17, LD1, V1, W4, N3, N9, N10, V17, N20, N22, P1, P5, P6, P8, P9, P10, P11, P15, P20a, P24, P25, P27, V21, V3, N23, N26, R1, V6, V15, V28, V12, V13, G1
-\item[nirāśrayaḥ] J12,
-\item[niśaṃśrayaṃ] G11
-\item[nirodhā vādham] V8
-\item[nirāśayam] Bo1,
-\item[nirāsnayām] J16,
-\item[nirāmayam] B2, Bo2, C4, J1, J3, J6, J14, Ko, N17, V19, N2, N5, N7, N8, N11, N12, N14, N16, V16, N18, N19, N21, N24, N25, P7, P13, P16, P20, P21, P23, P26, V20, V5, V2, V18, Ba3,
-\item[nirākulam] B1, B3, Ba5, C2, J13, N1, N4a, N13, P2, P12, P14, P17, P18, V10, Tue, V7, V22
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, P24, R5, V14, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_40c]
-\sthana{2.40c}
-\item[akṣir?] C8,
-\item[īkṣed] Ba, J5, P25,
-\item[īkṣet] C6, V13, G1
-\item[īkṣe] C1, J7, P7, 
-\item[īkṣye] P11, 
-\item[paśyed] A1, Bo1, C9, J4, J8, J10, J11, J14, J15, J16, J17, Ko, LD1, N17, P15, V1, W4, N9, V8, N10, P5, P8, P10, P24, P26, R1, V16, V17, N20, N22, V21, V3, N26, V15, V12, Ba3
-\item[paśyer] P1,
-\item[paśyad] J12,
-\item[paśyet] V20
-\item[paśyeta] LD2,
-\item[paśye] N19, N24
-\item[paśe] C3,
-\item[paśed] P9, 
-\item[paśyan] N7
-\item[paśca] P6, 
-\item[pasya?] N14  
-\item[dṛṣṭir] B3, Ba5, C2, C4, J6, V19, N1, N2, N4a, N11, N16, P2, P12, P14, P17, P18, P20, P23, V10, N21, N25, V7, V5
-\item[dṛṣṭi] B1, B2, Bo2, J1, J2, J3, J13, N5, N8, N12, N13, N18, P13, P21, Tue, V28, V22, V2 
-\item[vīkṣed] N3, G11
-\item[iched] N23
-\item[(illegible/unavailable)] Bo3, C5, N6, V14, V6, V18, T, P16, P22, P24, P27, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_43a]
-\sthana{2.43a}
-\item[vidhānajñā] C9, J5, J10, J11, Ko, V1, N2 ("vi" seems faded), N4a, N7, N10, N11, N12, N13, N14, N16, P6, P8, P9, V16, V17, V10, N19, N20, N21, V14, V3, Tue, N23, N26, V6, V7, V15, V22, V18, V13, G11, G1
-\item[vidhānajñāḥ] Ba3
-\item[vidhānaḥjñā] J2
-\item[vidhānajñe] V2
-\item[vidhanajñā] N18
-\item[vidhaustajñā] V20
-\item[vidhānujñaś] J1
-\item[vidhānataś] P23, 
-\item[vidhānajñāś] Bo2, C1, C3, C8, J4, J6, J7, J12, P5, P13, P14, P15, P24, P25, P27, 
-\item[vidhānajñaś] B1, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C8, J3, J16, N3, N24, P2, P11, P12, P17, P18, P21, 
-\item[vidhānajña] N22, V12
-\item[vidhānejñaś] N8
-\item[vidhijñās tu] C6,
-\item[vidhānajño] C4, J15, J17, LD1, LD2, N5, N17, N25, P1, P10, P20, V28
-\item[vidhānajñoś] J13, 
-\item[viddhānajño?] R1,
-\item[vividhānajño] B2, 
-\item[vidhāna?ī] N9, V8
-\item[vidhānajñaiś] N1 (unsure about "ai" matra)
-\item[vidhistajñāś] P26,
-\item[vidhiṃstajñaś] A1
-\item[vidhiṃstajñaiś] P7,
-\item[(illegible/unavailable)] Bo3, C5, N6, J14, V21, V5, T, P16, P22, P24, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_44cd]
-\sthana{2.44cd}
-\item[bhastrakā bhrāmarī vedhā kevalaṃ cāṣṭakumbhakāḥ] J2
-\item[bhastrikā bhramarikā mūrcchā kevaleṣṭā ca kumbhakā] G1
-\item[bhastrikā bhrāmarī mūrcchā kevalī cāṣṭakumbhakāḥ] V1, V16, V17, 
-\item[bhastrikā bhrāmarī mūrcchā kevala cāṣṭakumbhakāḥ] V20, V18, V12, V13, G11, Ba3
-\item[bhastrikā bhrāmarī mūrcchā vinītyaṣṭakumbhakāḥ] V10
-\item[bhastrikā bhrāmarī mūrcchā vinītyaṣṭakumbhakān] V28
-\item[bhastrikā bhrāmarī mūr??ā kevalaś cāṣṭakumbhakā] N1,
-\item[bhastrīkā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] C9, J13, J15, LD1, LD2, N21
-\item[bhastrīkā bhrāmarī mūrchā kevalaś cāṣṭakumbhakaḥ] P12, 
-\item[bhastikā bhrāmarī mūrchā kevalaś cāṣṭakumbhakaḥ] P26,
-\item[bhaṃstrikā bhramarī mūrchā kāṃbalā atha? kumbhakāḥ] J5
-\item[bhaṃstrīkā bhramarī mūrchā kevala cāṣṭakumbhakāḥ] N22, V21, V3
-\item[bhastṛ(?)kā bhrāmarī mūrcchā kevalaś cāṣṭakumbhakāḥ] N26
-\item[bhastrikā bhrāmarī mūrchā kaivalaś cā sa kumbhakaḥ] B3,
-\item[bhastrikā bhramarī mūrcchā cāṣṭakumbhakāḥ sahita?] N2 (after exchange of words' positions)
-\item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakāḥ] A1, B1, C3, J7pc, J10, J12, J16, J17, N3, N7, N9, N10, N17, P1, P2, P6, P9, P10, P17, P20, P27, R1, 
-\item[bhastrikā bhramarī mūrchā kevalaś caṣṭakumbhakāḥ] P5, P24, 
-\item[bhastrikā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] N20, P8, 
-\item[bhastrikā marī mūrchā kevalaṃ cāṣṭakumbhakāḥ] B2,
-\item[bhastrikā bhrāmalī mūrchā kevalaś cāṣṭakumbhakā] N14
-\item[gachati tiṣṭto kāryā mujjayākhyāṃ cākumbhakāḥ] V8
-\item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakā] N9
-\item[bhastrakā bhramarī mūrchā kevalāś cāṣṭakumbhakā] P13,
-\item[bhastrikā bhrāmarī mūrchā plāvanī saṣṭhakumbhakāḥ] N4a
-\item[bhastrikā bhramarī mūrcchā plāvanī tvaṣṭakumbhakāḥ] N12, V6, V7
-\item[bhastrikā bhramarī mūrcchā plāvanī tyaṣṭakumbhikā] P11, 
-\item[bhastrikā bhramarī mūrcchā pāvanīty eṣakumbhakāḥ] Ko,
-\item[bhastrikā bhramarī mūrcchā pāvanīty aṣṭakumbhakān] J4,
-\item[bhastrikā bhrāmarī mūrchā plāvinītyaṣṭakumbhakāḥ] Ba5, Bo1, C1, N13, P14, P18, V14, Tue, V15, V22
-\item[bhastrīkā bhrāmarī mūrchā plāvinītyaṣṭakumbhakāḥ] P25,
-\item[bhastrikā bhrāmarī mūrchā prāvatītyaṣṭakumbhakaḥ] N23
-\item[bhastrikā bhrāmarī mūrchā pratyahītyaṣṭakumbhakāḥ] P7,  
-\item[bhastrikā bhramarī mūrchā plāvanītyaṣṭakumbhakāḥ] Ba, C8, J7, P15,  
-\item[bhastrikā bhrāmarī mūrcchā pālāvanīty aṣṭakumbhakāḥ] J11,
-\item[bhastrikā bhramarī mūrchā pratvanītyaṣṭakumbhakāḥ] C6
-\item[bhastrikā bhrāmarī mūrchā sahita cāṣṭakumbhakā] P23,
-\item[bhastrikā bhramarī mūrchā sahitāś cāṣṭakumbhakāḥ] C4, N5
-\item[bhastikā bhrāmarī mūrchā sahitāś cāṣṭakumbhakam] N25, V5
-\item[bhastikā bhrāmarī mūrchā sahitāś cāṣṭakumbhakān] P21, 
-\item[bhastrikabhrama ma mūrchā saṃhataṃ cāṣṭakumbhakā] N8
-\item[bhastrikā bhramarī mūrchā saṃhataṃ cāṣṭakumbhakāḥ] N11(visarga added later), N16 ("saṃhitaṃ"),
-\item[bhastrikā bhramarī mūrchā saṃhitaṃ cāṣṭakumbhakāḥ] Bo2, J3, J6, V2
-\item[bhastrikā bhramarī mūrchā saṃprataṃ cāṣṭakumbhakān] N18
-\item[bhadrikā? bhramarī mūrchā saṃprataṃ rāṣṭakumbhakāṃ] J1
-\item[bhastrikā śītalī bhramarī mūrchā plāvanītyaṣtakumbhakāḥ] N19 
-\item[sastrikā bhramarī mūrchā kavalaś cāṣṭakumbhakā] N24
-\item[(illegible/unavailable)] Bo3, C5, N6, J14, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_46]
-\sthana{2.46}
-\item[om.] V19,
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkaḥ ||] Ba5
-\item[adhastāt kuñcanādāśu kaṇṭhasaṅkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ] G1, P11, P18, 
-\item[adhastāt kuñcanādāśu kaṇṭhasaṃcane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ] P25,
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkā ||] P7, 
-\item[athastātkuñcanenāśukarṇasaṅkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ] Ba3
-\item[adhas tāt kuñcane vāyuḥ kaṃṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāmīgaḥ ||] Ko,
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkagḥ ||] V22
-\item[adhas tāt kuñcanenāsuṃ kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkāṃ ||] C6
-\item[adhas tā kuñcakīnālau kaṇṭhasaṃkocane kṣate | madhyaṃ paścimatāṇena syāt prāṇo brahmanāḍīkaḥ ||] J2
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] A1, Bo1, C8, J10, J11, V1, N3, N13, P14, P20, Tue, V7, V20, V15, V28, V12, G11
-\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścamatānena syāt prāṇo brahmanāḍigaḥ ||] P21, 
-\item[adhastvākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C9
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṣate | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C1,
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] C3, LD1, N12, N26
-\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] B2, J15, V10, V21
-\item[adhasthāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] J7,
-\item[adhasthāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānenya syāt prāṇo brahmanāḍigaḥ ||] P10,
-\item[adhas thāt kuñjakenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇāsthā syāt prāṇo brahmanāḍīgataḥ ||] P15, 
-\item[adhastvākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhyanāḍigaḥ ||] P1, P9, 
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J6, J17, P2,
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] Bo2, N10, V16, V17, V14, V3, V6
-\item[addhastāt kuñcanenāsu kaṃṭhasaṃkocane krate | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] R1, 
-\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J3,
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍikaḥ ||] P17, 
-\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] N17,
-\item[adhastā kuñcanenāśu kaṇṭhaśaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍiga ||] J12,
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāna syāt prāṇo brahmāṇḍanāḍigaḥ ||]  Ba, N4a
-\item[adhastākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syā? prāṇo brahmanāḍikaḥ ||]  N5
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍikaḥ ||] C4
-\item[adhastākuñcaṇeṇāśu kaṇṭhaśaṇkocane kṛ(?)te | madhye paścamatānena syāt prāṇo madhye nāḍikā ||] N14
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] N1,
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] P13, P27, 
-\item[adhastāt kuñcane caiva kaṇṭhasaṃkocane krate | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] P6,
-\item[adhastāt kuñcanenaiva kaṇṭhaṃ saṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] B3
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo madhyanāgiḍaḥ ||] J4,
-\item[adhas tāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāgiḍaḥ ||] J4,
-\item[adhas tāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhyanāgiḍaḥ ||] P12,
-\item[adhamān kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ||] J5,
-\item[adhastāt kuñcanenaiva kaṇṭhaṃ saṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C2
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] B1
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] N1
-\item[adhasthāt kuñcanenaiva kaṇṭhasokocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] N21
-\item[adhastāt kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanādigaḥ ||] J16, 
-\item[adhastā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo madhyanāḍigaḥ ||] LD2,
-\item[adhastāt kuñcānenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ ||] P26,
-\item[adhaścā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhye paścimatānyena syāt prāṇo madhyanāḍigaḥ ||] P5, P24, 
-\item[aṃtastvā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] P8, 
-\item[adhastvā kuñcanenāsu kaṇṭhasaṃkocane kṛtaṃ | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] N7, V13
-\item[adhastvā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇe brahmanāḍigaḥ ||] N25
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] N16
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ ||] J1, N18, N19
-\item[adhastāt kuñcanenāsu kekatvaṃ saṃkocane krate madhye paścimatānena syāt prāṇo vasta(?)nāḍ?gaḥ ||] N9
-\item[adhastāt kuñcatenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍitāḥ ||] N20
-\item[adhas tā kucanenāmaḥ kaṃcasaṃkocane kṛte | madhye paścimatānena prāṇasyā brahmanāḍīkaḥ ||] N22
-\item[adhas ta kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanābhirāt ||] N23
-\item[adhas tāt kuñcanaunāśu kaṇṭhakocane kṛte | madhye paścimatāne tasyāt pāno madhyanāḍigaḥ ||] N24
-\item[adhas tāt kuñcanaunāśu kaṇṭhakocane kṛte | madhye paścimatāne tasyāt prāno madhyanāḍigaḥ ||] V18
-\item[(illegible/unavailable)] Bo3, C5, J14, N6, N2, N8, N11, V5, V2, T, P16, P22, P23, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_48b]
-\sthana{2.48b}
-\item[āsthāpya] V1, V19
-\item[sansthāpya] P26, P27, V8
-\item[utthāpya] A1, B1, B2, B3, Ba, Ba5, Bo1, Bo2, C2, C3, C4, C6, C8, C9, J1, J3, J6, J7, J8, J10, J11, J13, J17, LD1, N17, W4, N2, N3, N4a, N5, N8, N9, N11, N12, N13, N16, P1, P2, P5, P6, P7, P8, P11, P12, P13, P14, P15, P17, P18, P20, P24, P25, V16, V17, V10, N18, N21, V14, V21, V3, Tue, N23, N24, N26, V6, V7, V15, V28, V22, V18, V12, V13, G11
-\item[utthāya] P23,
-\item[uddhāpya] Ba3
-\item[ūtthāpya] J16,
-\item[uthāpya] Ko, R1, 
-\item[usthāpya] J2, J12, 
-\item[udhyāpya] N19
-\item[ur(?)thāpya] N14
-\item[utthāya] C1, N7, N20, P21, 
-\item[utprāpya] N22
-\item[uschāpya] N1, 
-\item[usthāpya] J4,
-\item[uchāpya] J5, J15, N10, P10, 
-\item[utchāpya?] LD2, P9, 
-\item[(illegible/unavailable)] Bo3, C5, J14, N6, N25, V20, V5, V2, T, G1, P16, P22, R5,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_49b]
-\sthana{2.49b}
-\item[virōdhāvadhi kumbhayet]
-\item[nīrādhāvadhi kumbhayet] N23,
-\item[nīrodhāvadhi kambhayet] Ba,
-\item[nīrodhāvadhi kumbhayet] A1, B1, B3, J15, J16, J17, N4a, N10, N13, P1, P2, P5, P8, P9, P10, P11,  P12, P14, P17, P18, P24, P25, P27, Tue, V21, G11, Ba3
-\item[nīrodhāvadhi kumbhata] R1,
-\item[nīrodhavadhi kumbhayet] P13, 
-\item[nīrovadhi kumbhayet] P15,
-\item[nīrodhān adhikumbhayet] N20, 
-\item[nirodho parikumbhake] G1
-\item[nīrodhāvadhi kumbhayat] N3, N7, N12, P20a,   
-\item[nīrodhāvadhi kumbhavat] N21, 
-\item[nīrodhāvadhi kumbhakaṃ] N2, N11,  
-\item[nīrodhāvadhi kumbhake] N18, 
-\item[nīrodhānadhi kumbhayet] N1, N26, V17
-\item[nīrodhāvahi kuṃbhayet] N9, N24,
-\item[nīrodhīvadhi kumbhayet] N19, 
-\item[nirodhaṃ kumbhayet tataḥ] J13, 
-\item[nirodhāścādhi kumbhayet] P7, 
-\item[nīrodhāvadhikam bhavet] P26,
-\item[śirodhāvadhi kumbhakaṃ] N5, N8, N16, P21, P23,  
-\item[śirodhāvati kumbhakān] P6,
-\item[śirodhyavadhi kumbhakaṃ] N25, P20, 
-\item[śiro ‘pya?vadhi kumbhayet] N17,
-\item[śiroyāvadti kumbhakāt] B2,
-\item[sirodhāvadhi kumbhakāt] N22,
-\item[(illegible/unavailable)] J14, N6, N14, T, P16, P22, R5,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_53a]
-\sthana{2.53a}
-\item[jālodare]
-\item[jālodarerā] P24, 
-\item[jalodar] A1,
-\item[jalodarā] P13, P14, P15, P18, P20, P21, P23, P27, 
-\item[jalodaro] B2,
-\item[jalodāraṃ] P17,
-\item[jalodaraṃ] B1, B3, Ba, P25, P26, 
-\item[jālodared] R1,
-\item[(illegible/unavailable)] P16, P22, R5,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_55ab]
-\sthana{2.55ab}
-\item[śītkāṃ dadyāt sadā cakre ghraṇe caiva vijṛmbhikām]
-\item[śītkāṃ dadyāt sadā vaktre ghraṇe naiva vijṛmbhikām] P14, P18, 
-\item[sītkāṃ dadyāt sadā vaktre ghraṇe caiva vijṛmbhikām] P27,
-\item[sītkāṃ dadyāt sadā vaktre ghraṇe naiva visarjjayet] B2,
-\item[sātkāṃ dadyāt sadā vaktre prāṇe caiva vijṛmbhite] P13,
-\item[satkāṃ dadyāt sadā vaktre prāṇe cātha vijambhakāṃ] P15,
-\item[śītkāṃ kuryāt sadā vaktrāt ghraṇo caiva vijṛ-kām] A1,
-\item[śītkāṃ kuryāt sadā vaktrāt ghraṇe naiva vijṛmbhakān] P26,
-\item[sītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikām] P25,  
-\item[śītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikām] Ba,  
-\item[śītkāṃ kuryāt yathā vaktre prāṇaṃ caiva visarjjayet] P20, 
-\item[śitkāṃ kuryāt sadā vaktre prāṇenaiva visarjayet] P21,
-\item[śitkāṃ kuryāt tathā vaktre ghrāṇenaiva visarjayet] P23, 
-\item[kuṃbhaṃ kuryāt sadā vaktre grāṇenaiva visarjayet] B1, B3, 
-\item[kuṃbhaṃ kuryāt sadā vaktrat ghrāṇenaiva vijaṃbhitaṃ] R1, 
-\item[(illegible/unavailable)] P16, P22, P24, R5,
-
-\item[]
-\item[(illegible/unavailable)]
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_54a]
-\sthana{2.54a}
-\item[śītkaraḥ] B1, C8, P17, 
-\item[śītkaram] C6,
-\item[śīkāra] J7,
-\item[śītkāraḥ] C1, B3, J13, J15, P7, P20, P25, V17, V21 
-\item[śītakāraḥ] N9, 
-\item[śītakāra] P1, R1, 
-\item[śitakāra] P6, 
-\item[śītakārāī] LD1, 
-\item[śīghrā] P11,
-\item[śītalī] P15,  
-\item[sītkara(ḥ)] Ba, Bo2, C2, C3, C4, J3, J4, N6, 
-\item[sītkāraḥ] J6, N8, N11, N16, P2, P12, P13, 
-\item[sītkāra] N17, P5, P24, 
-\item[śītkārī] J16, 
-\item[śītakārī] P8, P26, 
-\item[śītakarī] P9,
-\item[sītkāṃ] N5, N18, N22, Tue, N24, N25, V15, V28, V22, V10, G11
-\item[sītkād] V14
-\item[sitkā] N3, N14, 
-\item[sītkā] Ko, N12, P27, 
-\item[śītkāṃ] A1
-\item[sītātkārī] N13,
-\item[sītkaraḥ] N1,
-\item[sītkāra] P23,
-\item[sītkārā] P21,
-\item[sītkarā] J17, 
-\item[sītkārī] Ba5, Bo1, J5, J12, N2, N4a, N7, N10, P14, P18,    
-\item[sītakārī] C9
-\item[sītakarākarmma] J10, 
-\item[śītikārī] LD2,
-\item[sītalī] J2
-\item[sikta?] J1
-\item[sītkīṃ] V19,
-\item[sitkāṃ] J11, N19
-\item[sinktā?] B2
-\item[sātkāṃ] N21
-\item[dhikt(?)āṃ] N23
-\item[kumbhaṃ] V3,J8, N20,V14, V21, N26, V16, Ba3 
-\item[kumbhakaṃ] V7, V20, V18, V12, V13, 
-\item[kumbhakaḥ] V3
-\item[(illegible/unavailable)] Bo3, C5, J14, P10, V1, V6, V5, V2, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_55a]
-\sthana{2.55a}
-\item[cakram āsādya(ḥ)] A1, B3, Ba, Bo1, C2, V1, J8pc, J10, J17, N1, N7, N10, N28, V12,
- N21, P1, P2, P5, P8, P9, P12, P13, P17, P24, P25, P26, P27, V14, V20, V18, V16, Ba3
-\item[cakramāsādyamān] 
-\item[cakrasāmānya(ḥ)] B1, J8ac, J11, J13, N19, N24, P11, P15, P18, V3, V22
-\item[cakrasammānyaḥ] Ba5, N4a, N13, P14, Tue, V10, G1
-\item[cakrastamānyaṃ] J2
-\item[cakramāsādyamān] G11
-\item[cakram āsāya] N9,
-\item[cakrasāmānyaṃ] LD2
-\item[cakramādyāya] J12,
-\item[cakrasāmānya] J4, J5,
-\item[cakrasāmānyaḥ] N3, 
-\item[cakraṃ samāsādya] N6, N14, N17, W4
-\item[cakrasamāsādya] LD1,
-\item[cakranāśāya] N20, N26, P10, V17
-\item[cakranāsānyaḥ] N12, 
-\item[cakrabhogyaś ca] B2,
-\item[cakrasevyas tu] Bo2, C1, C4, C6, C8, J1, J3, J6, J7, J15, N2, N5, N8, N11, N16, P7, V19, N18, V P20, P21, P23, 21, N25, V15, V28, V2, V13, V21
-\item[cakrasevyas ta] N23
-\item[cakrabhogyas tu] N22, P6, 
-\item[cakrābhimānyaḥ] Ko,
-\item[vaktram āsādya] C3, C9
-\item[cakramadyāya] J16,
-\item[(illegible/unavailable)] Bo3, C5, J14, V6, V5, T, P16, P22, R1, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_56*1]
-\sthana{2.55*1}
-\item[satyāmūlena randhreṇa] C4, J6, J8pc, J10, J17, N5, N9, P1, P10, V21, N25
-\item[satyāmūlena randhrena] J15,
-\item[satyaṃ mūlena randhreṇa] P5, P24, 
-\item[sadāmūlena randhreṇa] J3, J6pc
-\item[satyāmūlena randhrenayaḥ] V18, V16, V21
-\item[spṛśyamūlena randhreṇa] B1, B3, C2, J13, P2, P12, P17, 
-\item[rasanātālumūlena] N6, N14, N17, 
-\item[rasanāmūla randhreṇa] C9, 
-\item[rasavatyāmūla randhre yaḥ] N20, N26
-\item[]
-\item[]
-\item[(illegible/unavailable)] A1, B2, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C5, C6, C8, J1, J2, J4, J5, J7, J11, J12, J14, J16, Ko, LD1, LD2, N1, N2, N3, N4a, N7, N8, N10, N11, N12, N13, N16, N18, N19, N21, N22, N23, N24, P6, P7, P8, P9, P11, P13, P14, P15, P18, P20, P21, P22, P23, P25, P26, P27, P28, R1, R5, W4, V3, V6, V20, V15, V5, V28, V22, V2, V12, V13, V10, V14, V17, G11, Tue, T, Ba3, G1, P16,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_57a]
-\sthana{2.57a}
-\item[rasanātālumūlena] A1, B1, B2, B3, Bo1, C2, C4, C9, J4, J10, J12, J13, J15, J16, J17, Ko, LD1, N1, N2, N3, N6, N7, N10, N11, N12, N14, N17, N21, N22, N25, N26, P1, P2, P5, P6, P8, P9, P10, P11, P12, P13, P15, P20, P23, P24, P26, P27, R1, V1, V20, V15, V28, Ba3
-\item[rasanātālumūle yaḥ] Ba,
-\item[rasanātālumūlenaya] V14,
-\item[rasanātālumūlena yaḥ] V2, V18, V12, V3, V16, V17, G11
-\item[rasanātāgru mūlena yaḥ] G1
-\item[rasanāttālumūlena] N20
-\item[rasanātālumūlaṃ ca] J5, 
-\item[rasanātāluyogena] Bo2, C6, C8, J3, J6, J7, N8, N16, N18, P7, V21, N23, V13
-\item[rasanātālayogena] J1,
-\item[rasanā tālamūlena] N5, N9, 
-\item[rasanāṃ tālamūlena] J11
-\item[rasanāṃ tālumūlena] P25,
-\item[rasanāṃ tāluyogena] P21, 
-\item[rasanānāṃ tālamūlena] N19
-\item[jihvayā vāyum ākṛṣya] Ba5,
-\item[satyāṃ mūlena randhreṇa] C3,
-\item[]
-\item[(illegible/unavailable)] Bo3, C5, J2, J14, N4a, Tue, LD2, N13, N24, P14, P17, P18, P22, V6, V5, V22, V10, T, P16, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_58b]
-\sthana{2.58b}
-\item[pūrvavat kumbhakād anu] C4, C6, C8, J1, J4, J6pc, J7, J11, Ko, N2, N5, N8, N11, N21, P13, P20, P23, V14, V21, N3, N24, N25, P27, V6, V19, V15, V28, V2, Ba3, G1
-\item[pūrvavat kuṃbhanādan] G11
-\item[pūrvavat kūmbhakād anu] Bo2,
-\item[pūrvavat kumbhayed anu] P26,
-\item[pūrvavat kumbhanāda tu] P11,
-\item[pūrvavat kumbhakād anmuṃ] Bo1,
-\item[pūrvavat kumbhayed anu] A1, V20
-\item[pūrvavat kumbhakād anū] N19
-\item[pūrve ce kumbhanād agu] J5,
-\item[pūrvavat kumbhakādane] N23, V7
-\item[pūrvavat kumbhakeṣv anu] B2
-\item[pūrvavat kumbhasādhanaṃ] B1, B3, Ba, Ba5, Bo3, C1, C2, C3, C9, J2, J3, J6, J10, J12, J13, J15, J16, J17, N4a, LD1, LD2, N6, N9, N10, N12, N13, N14, N17, N18, N22, P2, P5, P6, P10, P12, P24, P25, R1, V3, Tue, V22, V18, V12, V13, V3, V10, V16
-\item[pūrvaṃ ca  kumbhasādhanaṃ] P1, P8, P9, 
-\item[pūrvava kumbhasādhanaṃ] N7, 
-\item[kākacaṃcusamannayāt] V17
-\item[pūrvavat kumbhasādhayet] N1, P17, 
-\item[pūrvakumbhakavad bhavet] P7,
-\item[kākacaṃcusamunnayāt] N16, N26
-\item[kākacaṃcusamunnarayāt] N20, 
-\item[kākacaṃcusamanvayāt] P21, 
-\item[(illegible/unavailable)] C5, J14, P14, P15, V5, T, P16, P18, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_62b]
-\sthana{2.62b}
-\item[kapāle ca samantataḥ] V1 
-\item[kapāla pavanaṃ tataḥ] N6, 
-\item[kapāle niḥsvanaṃ tathā] Ba,
-\item[kapole niḥsvanaṃ tathā] P25, 
-\item[kapole niḥsvanaṃ tadā] C1, 
-\item[kapole nisvanaṃ tadā] J2, 
-\item[kapole svāsanaṃ tataḥ] N26
-\item[kapole svasanaṃ tataḥ] V18
-\item[kapāla śvasamaṃ tataḥ] J4, P10, 
-\item[kapāla śasanaṃ tataḥ] J15, 
-\item[kapāle śasanaṃ tataḥ] Ba3
-\item[kapāle śvasanaṃ tataḥ] J11, N3, P20a, V17 
-\item[kapāle svaśanaṃ tataḥ] LD1, 
-\item[kapāla śparśanaṃ tataḥ] J16, N9, N10, P5, P24, 
-\item[kapole svaśanaṃ tataḥ] LD2, V16
-\item[kapole śvāsanaṃ tataḥ] N20, 
-\item[kapole svasanaṃ tataḥ] N7, P8, 
-\item[kapole śvavaṇe tathā] B2,
-\item[kapole śvavaṇe tataḥ] V13, V3
-\item[kapāle sasvanaṃ tataḥ] P13, P15, P27, 
-\item[kapole sasvanaṃ tataḥ] P9,
-\item[kapolaṃ sasvaramarut] G11
-\item[kapolaṃ sasvanaṃ tataḥ] P11, 
-\item[kapola saśvanaṃ tataḥ] J5,
-\item[kapole sca sanantataḥ] V3
-\item[kayolesvaśanaṃtataḥ] N24
-\item[kapāla pavanaṃ tata] N14, 
-\item[kapāla sasvanaṃ tataḥ] N21, V12
-\item[kapāla svasanaṃ tataḥ] N19, 
-\item[kapālasya samaṃ tataḥ] C3
-\item[kapālasyarśanaṃ tataḥ] J10, 
-\item[kapāle sasvanaṃ tataḥ] Bo1, V14
-\item[kapāle mukhano marut] Ko,
-\item[kapālādhi sasvanaṃ tataḥ] V6
-\item[kapāla pavanaṃ tataḥ] N17,
-\item[kapāle susvanam] G1
-\item[kapālasparśanaṃ tataḥ] C9, J17, P1,
-\item[kapālasya śanaṃ tataḥ] J12,
-\item[kapālādhi niḥsvanaṃ] C6,
-\item[kapālavidhi sasvanaṃ] P26,
-\item[kapālāvadhi sasvanaṃ] A1, Ba5, C8, J7, N4a, N12, N13, P7, P14, P18, Tue, V7, V20, V28, V22, V10
-\item[kapālāvadhi sasanaṃ] V21
-\item[kapālāvadhi pūrayet] B1, B3, Bo2, C2, C4, J1, J3, J6, J13, N1, N2, N5, N8, N11, N16, N18, N25, P2, P12, P17, P20, P21, P23, V15, V2
-\item[kepālavadhi sasvanaṃ] N23, 
-\item[kepālena samaṃ tataḥ] P6, 
-\item[kepālenastha sarīrasthaṃ] N22,
- \item[(illegible/unavailable)] Bo3, C5, J14, V5, T, P16, P22, R1, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_65cd]
-\sthana{2.65cd}
-\item[dhārayan nāsikāmadhyṃ aṃgulībhyāṃ tathā dṛḍham] N10,
-\item[dhārayennasikā madhye aṅgulīnāṃ tathā dṛḍham] Ba3
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ vinā dṛḍham] V1, V21
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] LD1,
-\item[dhārayen nāsikāmadhyaṃ aṅgulībhyāṃ tathā dṛḍham] A1, C9, P26, 
-\item[dhāraye nāsikāmadhyaṃ maṅgulānāṃ tathā dṛḍham] J12,
-\item[dhārayen nāsikāmadhaṃ aṅgulībhyāṃ tathā dṛḍham] V20
-\item[dhārayen nāsikāmadhya aṅgulībhyāṃ tathā dṛḍham] J10,
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] B2, C3, J17, N6, N14, N17, P5, P8, P10, P24, V18, V12, V16, V17, G1
-\item[dhāraye nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] N7, 
-\item[dhārayen nāsikāmadhyai aṅgulībhyāṃ tathā dṛḍham] J15,
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tadā dṛḍham] N9, P1, 
-\item[dhārayen nāsīkāmadhye maṃgulībhā tathā dṛḍham] J16,
-\item[dhārayen nāsīkāmadhye aṅgulībhyāṃ dṛḍham tathā] LD2,
-\item[dhārayen nāsikāmadhye aṅguṣṭhābhyāṃ tathā dṛḍham] N20, N26,
-\item[dhārayen nāsikāṃ madhye aṅgulībhyāṃ tathā dṛḍham] P9, 
-\item[dhārayen nāsikābhyāṃatarjabhyāṃ tathā dṛḍham] R1,
-\item[dhārayen nāsikāmadhyṃ tarjanībhyāṃ vinā dṛḍham] Ko,
-\item[dhārayen nāsikāmadhyaṃ tarjanībhyāṃ vinā dṛḍham] J11, N3, 
-\item[dhārayen nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] C6, J5, N12, N18, N19, P6, V3, N21, N24, P13, P17, P25, P27, V6, V7, V28, V22, V2, V3, V14, G11
-\item[dhārayen nāsikāmadhyaṃ tarjanībhyāṃ vinā dṛḍham] P7,
-\item[dhāraye nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] Ba,
-\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] Ba5, Bo2, J3, N11, N16, P11, P14, P20, 
-\item[dhārayen nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] P23,
-\item[dhāraye nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] P18,
-\item[dhāraye nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] Ba5, N4a, N13, V15
-\item[dhāraye nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] N8, 
-\item[dhārayen nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] J7, P21, 
-\item[dhāvayen nāsikāmadhyāṃ tarjanībhyāṃ vinā dṛḍham] P15,
-\item[dhāraye nāsikāṃ madhye tarjjanībhyāṃ vinā dṛḍham] Bo1,
-\item[dhārayen nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] B1, B3, J13, N1, N2, P2, P12, V13
-\item[dhārayan nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] C2, 
-\item[dhārayen nāsikāmadhya tarjjanībhyāṃ vinā dṛḍhām] J1
-\item[dhārayen nāsikā madhye tarjanībhyāṃ vinā madam] J4,
-\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ yathā dṛḍham] C8
-\item[dhārayen nāśikāmadhye sarjanībhyāṃ yathā dṛḍham] N22,
-\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] J6pc,
-\item[kārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] J6,
-\item[dhārayen nāsikāṃ madhyāt tarjanībhyāṃ vinā dṛḍham] Tue
-\item[dhārayen nāsikāmadhyāt tarjanībhyāṃ vinā dṛḍham] C4, N5, N25, V10
-\item[dhārayaṃ nāśikāmadhye tarjanabhyāṃ vinā dṛḍham] N23
-\item[dhāraae vāsikāmadhye tarjanībhyāṃ vinā dṛḍham] J2,
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_67ab]
-\sthana{2.67ab}
-\item[kaṇdalībodhakaṃ viprabhavaghnaṃ sukhadaṃ śubhaṃ] V1
-\item[kuṇḍalī bodhakaḥ kubho rogaghnaḥ sukhada śubhaḥ] Ba3
-\item[kuṇdalībodhanaṃ sarva doṣaghnaṃ sukhadaṃ śivaṃ] Bo1
-\item[kuṇdalībodhanaṃ sarva doṣaghnaṃ sukhadaṃ śuvaṃ] V15
-\item[kuṇdalībodhanaṃ sarva doṣa?? sūkhadaṃ sūkhaṃ] N24,
-\item[kaṇdalībodhakaṃ ksipraṃ pavanaṃ sukhadaṃ hitaṃ] Tue, V7
-\item[kuṇḍalo bodhaka kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] J2
-\item[kuṇḍalībodhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] Ba5, N13, P14, P18, V22
-\item[kuṇḍalabodhakaṃ pāpāghnaṃ suṣadaṃ śubhaṃ] Ko,
-\item[kaṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] V6, V2
-\item[kaṇḍalībodhanaṃ cakaṃ pāpaghnaṃ sukhada śubhaṃ] V14
-\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] B2, Bo2, C4, J3, J6, J7, J11, N16, N19, N23, N25, P6, P21, P23, P27, 
-\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ śubhadaṃ sukhaṃ] P20,
-\item[kuḍalībodhanaṃ kuryāt pāpaghna suṣadaṃ subhaṃ] N18, 
-\item[kuṇḍaṃlībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] N8, N11, 
-\item[kuṇḍalabodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] N5, 
-\item[kaṇḍalībodhanaṃ kuryāt pāpaghnaṃ subhaṃ tad śukhaṃ] V28
-\item[kuṇdalībādhakaṃ ksipraṃ pāpghnaṃ sukhadaṃ hitaṃ] Ba,
-\item[kuṇdalībodhakaṃ ksipraṃ pāpaghnaṃ sukhadaṃ hitaṃ] P25, 
-\item[kumḍalīśhodhanaṃ kuyāt pāpaghnaṃ ca sukhaṃ śhubhaṃ] N22, 
-\item[kuṇdalībādhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] N4a,  
-\item[kaṇḍalībodhanaṃ kuryāt tāpaghnaṃ sukhadaṃ śubhaṃ] N2, 
-\item[kuṇḍalībodhakaḥ kuṃbhaḥ rogaghnaḥ sukhadaḥ śubhaḥ] N6,
-\item[kuḍalībodhaka kuṃbhao rogaghnaḥ sukhadaḥ śubhaḥ] P5, P8, P9, P24, 
-\item[kuḍalībodhanaḥ kuṃbhao rogaghnaḥ sukhadaḥ śubhaḥ] P7,
-\item[kuṇḍalībodhakaḥ kuṃbho rogaghnaḥ sukhadaṃ śubhaḥ] N9, N10, N17, N20,  N26, P1, P10, 
-\item[kuḍalībodhakaḥ kuṃbhao rogaghnaṃ sukhadaṃ śubhaṃ] P17,
-\item[kuṇḍalībodhakaḥ karthūḥ roghnaṃḥ sukhadaḥ śubhaḥ] A1, V20
-\item[kuṇḍalībodhaka kumbhao rogaghnaṃ sukhadaḥ śubham] N14, P2,
-\item[kuṇḍalībodhakaṃ cakraṃ bhāvaghnam sukhadaṃ śubhaṃ] N3,
-\item[kuṇḍalībodhanaṃ cakra bhāvaghnam sukhadaṃ śubhaṃ] P11,  
-\item[kuṇḍalībodhakaṃ kṣipram bhavaghnam sukhadaṃ śubhaṃ] N12,
-\item[kuṇḍalībodhakaṃ vipram bhavaghnam sukhadaṃ śubhaṃ] P15,
-\item[kuṇḍalībodhanaṃ karttur bhavaghnam sukhadaṃ śubhaṃ] C6,
-\item[kuṇḍalībodhanaṃ karttur bhavaghnam sukhaṃ śubhaṃ] V21
-\item[kuṇḍalībodhakaṃ miśrabhavaghnam sukhadaṃ śubhaṃ] J4,
-\item[kuṇḍalībodhanaṃ kartuṃ bhavaghnam sukhadaṃ śubhadaṃ] C8,
-\item[kuṇḍalībodhakaḥ kartuḥ rogaghnasukhadaḥ śubhaṃ] P26, 
-\item[kuṇḍalībodhakaḥ kumbhoḥ ragaghnaḥ sukhadaḥ śubhaḥ] N7, 
-\item[kuṇḍalībodhakaḥ kumbho roghnaḥ sukhadaḥ śubhaḥ] C3, J10, J17, 
-\item[kuṇḍalībodhakaṃ kumbhoḥ roghnaḥ sukhadaḥ śubhaḥ] J15, J16,
-\item[kuṇḍalībodhakaḥ kumbho roghnaḥ suṣadaḥ śubhaḥ] C9, 
-\item[kuṇḍalībodhakaḥ kumbho rogaghnaḥ sukhadaḥ śubhaḥ] J12, LD1, LD2, V16 
-\item[kuṇḍalībodhakaḥ kumbho rogaghnaḥ sukhadaḥ śubhaḥ] B1, B3, C2, J13, V18, V12, V13, V3, V17 
-\item[kuṇḍalībodhakaḥ kumbho rogaghnaṃ sukhadaṃ śubhaṃ] P12,
-\item[kuṇḍali bodhakaṃ vakrabhāvaghnaṃ sukhadaṃ śubhaṃ] G11
-\item[kuṇḍalībodhakaṃ sarva doṣaghnaṃ sukhadaṃ hitaṃ] N21, P13, 
-\item[kuṇḍalībodhakaḥ kumbho rogaghnaṃ sukhadaṃ śubhaṃ] N1, 
-\item[kuṇḍalīno bodhacakraṃ? bhāvadaṃ supada śubhāṃ] J5
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, V10, T, G1, P16, P22, R1, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_68a]
-\sthana{2.68a}
-\item[guṇatrayasamudbhūtaṃ] Bo1,
-\item[guṇatrayasamudbhūta] N21,
-\item[samagātrasamudbhūtaṃ] J4, N7, 
-\item[samagātrasamā....] P10, 
-\item[samyagātrasamudbhūtaṃ] J7,
-\item[samyagātrasamudbhutaṃ] J15,
-\item[samyagātrasamudbhuta] P14,
-\item[samyaggātrasamudbhūtaṃ] B2, Bo2, C4, C8, J2, J11, N3, N9, N16, N19, P13, V1, V7, V2, V3, V17
-\item[samyagatrasamudbhutaṃ] N12,
-\item[samyaggātrasamudbhūta] N2, N4a, N11, N13, N20, N23, N26, P15, P18, P25, P27, Tue, 
-\item[samyagātrā sumuvbhutaḥ?] N24,
-\item[samya gātre samudbhūtaṃ] J5,
-\item[samyaggātre samudbhūtaṃ] J6, N5, N8, N25, P20, P23, V28
-\item[samyaggātre samudbhūta] J3,
-\item[samyaggātre samudbhūtat] P6, 
-\item[samyag gātraṃ samudbhūtaṃ] Ko,
-\item[samyaggātasamudbhūtam] V20, V15
-\item[samyagataḥ samudbhūtam] V14
-\item[samyak gātrasamudbhūta] Ba, V22, G11, P11, 
-\item[samyag antaḥ samudbhū] A1
-\item[samyag antaḥ samudbhūta] C6
-\item[samyak bhastrā samudbhūtā] B1, B3, C2, J12, J13, J16, N1, N10, P1, P2, P8, P9, P12, P17, Ba3
-\item[samyak bhastrā samudbhūtaḥ] P26, V21  
-\item[samyak bhastrā samabhyāsaḥ] V13
-\item[samyag bhastrā samudbhūtāṃ] LD1, LD2, N18, P7, P21, 
-\item[samyak bhastrā samudbhūto] C3, J10, J17, N14, N17, P5, P24, 
-\item[samyag bhastri samudbhūtaṃ] C9
-\item[samyag bhastrī samudbhūto] N6, 
-\item[sasyagātrasamudbhūtaṃ] J1
-\item[samagātrasamudbhūtaṃ] V6, V16
-\item[samagātre samudbhūta] N22,
-\item[samagātrasamudbhūtaghnaṃ] V18
-\item[(illegible/unavailable)] Ba5, Bo3, C1, C5, J14, V5, V12, V10, T, G1, P16, P22, R1, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_69a]
-\sthana{2.69a}
-\item[vegair ghoṣaṃ] V1
-\item[vegākṛṣṭaṃ] Ba, J2, N21, P25, V15
-\item[vegākṛṣṇaṃ] P13,
-\item[vegā gho] N23,
-\item[vego tad ghoṣa] J1
-\item[vego ghoṣam] C3, C9, J16, N3, N9, N16, N20, P1, P5, P10, P24, P27, 
-\item[vego ghoṣa] N24,
-\item[vego ghoṣaḥ] J15,
-\item[vegāt ghoṣaṃ] A1, B2, B3, Ba5, C9, N4a, LD1, N13, N26, P14, P17, P18, P26, R1, Tue, V6, V7, V22, V18, V12, V13, V14, V16, V17, Ba3
-\item[vegad ghoṣa] N10,
-\item[vegad ghoṣaṃ] N7, 
-\item[vegā ghoṣaṃ] P8, V3
-\item[vegot ghoṣaṃ] J12, N14, N17, P11, V2, G11
-\item[vegot ghoṣe] V2,
-\item[vegodveṣaḥ] J4,
-\item[vegodvegaṃ] P20a,
-\item[vegodveṣaṃ] N5, P15, P20, 
-\item[vaiṣodvaiṣaṃ] J5,
-\item[veget vege] V28
-\item[vegād ghoṣaṃ] J10, J17, N1, V20
-\item[vegod ghoṣaṃ] Bo2, C6, J10pc, J11, N18, P6, P9, P21, 
-\item[vegod ghoraṃ] P23, 
-\item[vedod ghoṣaṃ] V21
-\item[vedyo ghoṣaṃ] N22,
-\item[vegodyoṣaṃ] J7,
-\item[vegopyoyaṃ] N19, 
-\item[vemod ghoṣaṃ] J3, J6, 
-\item[vegodvegaṃ] C4, J13, N25, 
-\item[vegod deyaṃ] P7, 
-\item[vegād aghoṣaṃ] B1,
-\item[vegād doṣaṃ] Bo1, N2, 
-\item[vegod doṣaṃ] N6, 
-\item[veṇu doṣaṃ] C2,
-\item[veṇu ghoṣaṃ] P2,
-\item[veṇud ghoṣaṃ] P12, 
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, Ko, N8, N11, N12, V5, V10, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_73a]
-\sthana{2.73a}
-\item[ārecyāpūrya yat kuryāt] N2, N5, 
-\item[ārecya āpūrya yat kuryāt] P23,
-\item[ārecyāpūrya ?????] N11, 
-\item[ārecyāpūrya vat kuryāt] J13,
-\item[ārecyāpūrvavat kuryāt] N25,
-\item[recakaṃ pūrakaṃ kuryāt] N1, N10, P2, 
-\item[recya vā pūrya sanakaiḥ] J11,
-\item[recake pūrakaṃ kuryāt] J12, J16, 
-\item[recayet pūrakaṃ kuryāt] N16, N19, N23, P6, 
-\item[recakaṃ pūrakaṃ kuryāt] P12, P17, 
-\item[ārecyāpūrya kuṃ?yāṃt] N8,
-\item[recakaḥ pūrakaṣ kāryyaḥ] J17,
-\item[recakaḥ pūrakaḥ kāryaḥ] A1, N9, J15, N17, N26, P1, P5, P9, P24, R1, 
-\item[recakaḥ pūrakaḥ kuryāt] P21, 
-\item[recanaṃ pūraṇam kuryāt] P13,
-\item[recakaḥ pūrakaḥ kā....] P10,
-\item[recakaḥ pūrakaḥ kāryāḥ] N7, 
-\item[recakaḥ pūrakaḥ kāryaṃ] N20
-\item[recakaḥ pūrakaḥ kāryo] P8, 
-\item[recakaḥ pūrvavat kāryaḥ] N3, 
-\item[recakaṃ pūrakaṃ kuryāt] B1, B3, 
-\item[ārecaṃ pūrvavat kuryāt] P20,
-\item[recayitvā pūrakaḥ kāryaḥ] J14, 
-\item[reco vā pūrakaḥ kāryaḥ] P11, 
-\item[rocyatvā pūrakaḥ kuṃbhakaḥ] N24,
-\item[recakaḥ pūrakaṃ tyatvā] N18, 
-\item[recakaṃ pūrakaṃ muktā] N22,
-\item[recake pūrakaṃ uktwā] V14, G11, Ba3
-\item[recake pūrṇaṃ kāryaṃ] Ba,
-\item[recake pūraṇaṃ kāryaṃ] P25,
-\item[recanaṃ pūranaṃ kuryāt] N21,
-\item[recaka pūraka kumbhai] V3, V16
-\item[recaka pūraka kumbhakaiḥ] V17
-\item[recakaṃ pūrakaṃ muktyā] G1
-\item[recaka pūraka varjite] V10
-\item[virecya pūrakaḥ kāryaḥ] P7,
-\item[prāṇāyāmoyam ity uktaḥ] N4a, N12, P14, P18, Tue,
-\item[(illegible/unavailable)] B2, N6, N13, N14, P15, T, P16, P22, P26, P27, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_73b]
-\sthana{2.73b}
-\item[śanaiḥ saṃhitakumbhakaḥ] N16, 
-\item[sarve kevalakumbhakam] G1
-\item[śanaiḥ sahitakumbhakaḥ] N23, P7, 
-\item[śanaiḥ sahitakumbhakaiḥ] P11,
-\item[śanaiḥ sahitakumbhakaṃ] P6, 
-\item[sa vai kevalakumbhakaḥ] N4a, N18, N22, P18, P26, Tue, 
-\item[sa vai sahitakumbhakaḥ] A1, B1, B2, B3, J12, J13, J15, N1, N3, N5, N10, N17, N21, P2, P8, P12, P13, P17, P20, P21, P23, P27, 
-\item[sarve sahitakumbhakaḥ] J17, N20, 
-\item[savyai sahitakumbhakaḥ] N19, 
-\item[sarvai sahitakumbhakaḥ] N9, N12, N26, P1, P9, P10, R1, 
-\item[sarvaiḥ sahitakumbhakam] Ba, P25,
-\item[sarvai sā?ti kumbhakaṃ] N25,
-\item[sarvai kevalakumbhaka] N14,
-\item[sarvair ahitakumbhakaḥ] N7, 
-\item[sa vai sahitakumbhak] J16, 
-\item[sa taiḥ sahitakumbhakaḥ] J14, N24, 
-\item[sa vai kevalakumbhakaḥ] Ba3, G11, P14, V10
-\item[sarvaiḥ sahitapūrakaḥ] P5, P24, 
-\item[kāryaḥ sahitakumbhakaḥ] J11,
-\item[kuryat sahitakumbhakaḥ] N2, N8, 
-\item[kuryat saṃhitakumbhakaḥ] N11, 
-\item[kāryaḥ śanai sahitakumbhakā] V3, V14
-\item[kāryaḥ śarveiḥ sahitakumbhakeiḥ] V16
-\item[kāryaḥ śarveiḥ sahitakumbhaka] V17
-\item[(illegible/unavailable)] N6, N13, P15, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_78a]
-\sthana{kumbhakaḥ prāṇarecānte} C3, J2, J3, J13, N8, N11, N16, N21, P5, P24,V1, V6, V28, V18, V16, G11,
-\sthana{kumbhakāt prāṇarecānte} N20, N26, V17
-\item[kumbhaka prāṇarecānte] A1, B1, B2, Bo2, C1, J1, J5, LD1, N2, N7, N9, N10, N12, N14, N17, N22, N25, P1, P2, P9, P10, P12, P13, P17, P20, P21, P23, V2, V12, Ba3
-\item[kumbhaka prāṇarecyaṃte] N1,
-\item[kumbhake siddha reca puraka] V21
-\item[kumbhaka prāṇa cātte?] J10, J17, 
-\item[kumbhaka prāṇa recāttaḥ] J16,
-\item[kumbhaka prāṇarecātte] P8, P26,
-\item[kumbhakaṃ prāṇarecānte] B3, C4, C9, J6, N5, N19, P6, P27, G1, R1, 
-\item[kumbhakaṃ prāṇarecātteḥ] J15,
-\item[kumbhakaṃ prāṇareṃcānte] J12,
-\item[kumbhataḥ prāṇarecānte] J7,
-\item[kumbhata prāṇarecānte] V3
-\item[kumbhakaṃ ghrāṇarecānte] LD2, 
-\item[kumbhitaḥ prāṇarecānte] Bo1, N23, N24, P11, P15, V15
-\item[kumbhītaḥ prāṇarecānte] Ko, J11, J14, N3,  
-\item[kumbhīte prāṇarecānte] V14
-\item[kumbhītaprāṇareṣānte] J4,
-\item[tad etat prāṇarrecānte] C6, C8, P7, 
-\item[kumbhakaḥ prāṇarodhānte] Ba, Ba5, P14, P18, P25, Tue, V7, V20, V22, V10
-\item[kumbhaka prāṇarodhānte] N4a, N13,
-\item[(illegible/unavailable)] Bo3, C1, C5, N6, N18, V5, V13, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\end{ekdosis}
-\end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:

### 2023-09-02 marma/marma2.tex
new routine for jyotsna

--- 2022-12-16 marma/Marmasthāna.tex
@@
-Notes
-    • P16 has chapter 1 till 1.45
-    • P22 has only 3 folios of chapter 1
-\input{Preamble-Print}
-\makepagestyle{HPed}
-\makeoddhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
-\makeevenhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
-\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
-\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}
-
-\newcommand{\myfn}[1]{\footnote{\texteng{#1}}}
-\setlength{\footmarkwidth}{1.3em}
-\setlength{\footmarksep}{0em}
-\setlength{\footparindent}{1.3em}
-\footmarkstyle{\textsuperscript{#1} }
-\usepackage{fnpos}
-\makeFNbottom
-
-% additions/changes 2022-03-12 mm:
-\SetTEIxmlExport{autopar=false}
-\NewDocumentEnvironment{marma}{O{}}{}{}
-\def\sthana{}
-
-\begin{document}
-\pagestyle{HPed} %
-\begin{ekdosis}
-  \SetLineation{lineation = none,}
-
-  \chapter{Variants and Versions}
-
-\sthana{24a}
-
-  \begin{marma}[hp01_024]
-    \begin{description}
-%    \item[(ku(r)k(k)uṭāsana-)]{}
-    \item[-vat kṛtvā]{V1, N22}
-    \item[-vat kratvā]{P6}
-    \item[-bandhastho]{A1, J8, L1, M1, V6, B1, B2, B3, Bo2, C1, C2, C3, C4, N2, C6, C8, J1, J3, J6, J13, J14, K1, Ko, LD1, LD2, N1, V2, V3 (about \emph{stho}), N4a, V8, N5, N8, N9, N11, V21, N13, N16, N18, N19, N20, N21, N24, N25, N26, P1, P2, P3, P4, P7, P10, P11, P14, P15, P17, P20, P21, P22, P23, R4, Tue, V17, V14, R3, R5, V13, V10, G1
-    \item[-bandhasthe]{J7, V19}
-    \item[-badhascho]{Ba, Bo3, V28, V22}
-    \item[-baṃdhascho] J5, J11, J15, P12, P13, P18, P27, V15
-    \item[-baddhastho] N12, P25, 
-    \item[-bandhastu]{Bo1}
-    \item[-madhyastho]{C9, J10, J12, J16 (\emph{baṃ} is given in margin), J17, N17, V4, N6, N7, N10, V20, V18, P5, P8, P24, P26, V12, LD1}
-     \item[-madhyasthā]{P9}
-    \item[-madhyasthau]{V16}
-    \item[-baṃdhasthau]{N23}
-    \item[-badhestu]{J2}
-    \item[-vadhnastho] P16, 
-    \item[(unavailable/illegible)]{Ba5, C7, J4, N14, C5, G11, R2, T, Ba3
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_027]
-
-  \sthana{27d}
-
-  \begin{description}
-\item[jaṭharapradīptaṃ]    V6 (jaraṭha), B1, B2, B3, Bo1, C1, J13, N1, V22, N13, N21, P2, P12, P13, P14, P17, R3, R4, 
-\item[jaṭharapradīpti]    Ba5, V10
-\item[jaṭharapradiptaṃ]    P4,
-\item[jaṭharaprakṛṣtaṃ]    V14
-\item[jaṭharapradīpaṃ]    P6,
-\item[jaṭharapradīptiṃ]    Bo3, Ko, N4a, P18, P22, Tue, R5
-\item[jaṭharāgnidīptiṃ]    P22a,
-\item[jaṭharapracaṇḍaṃ]    K1, V1, N22, V15, G1
-\item[jaṭharaṃ pracaṇḍaṃ]    J11, P16, 
-\item[jaṭharapracaṃḍa]    J5, P11, P27, 
-\item[matsyendra-vīra-jaṭhara-pracaṇḍaṃ]          J4, N19, V1 [vīra-ja is unmetrical?]
-\item[budhaṃ pracaṇḍa]    V8
-\item[jaṭharapravṛdhaḥ]    P21, P23, 
-\item[jaṭharapravṛddha]    J10, J17, V2, N8, N11
-\item[jaṭharapravṛddhau]    J15, P20, 
-\item[jaṭharapravṛddhaṃ]    N24
-\item[jaṭharapravṛddhiṃ]    C1
-\item[jaṭharaprabuddhaṃ]    C8, J14, N2, N12, V21, N23, V12,
-\item[jaṭharaprabuddha]    Ba, P25, 
-\item[jaṭharaprabuddhau]    C4 ("jvalanapradīptam" is also given above), L1, V19, V28, N5, N25, P3
-\item[jaṭharaprabodhaṃ]    A1, C6, P7, P26,
-\item[...bodhaṃ]       R2,
-\item[jaṭharāprabodhaṃ]    V20
-\item[jaṭharaṃ prabadhaṃ]    J2
-\item[jaṭhara pravaddhaṃ] N2, 
-\item[jvalanapradīptaṃ]    LD1, Bo2, C3, C9, J1, J3, J6, J8, N17, V3, V4, N6,, N9, N18, V18, N26, P1, P5, V17, P8, P24, V13, V16, P9
-\item[jvalanaṃ pradīptaṃ]    J7, J12, LD2, N10, N16, N20, P10, 
-\item[t(?)valanaṃ pradiptaṃ]    J16,
-\item[jvalatpradīptaṃ]    N7,
-\item[vīrapracaṃḍaṃ] P15,
-\item[(unavailable/illegible)]    C5, C7, M1, N14, G11, T, Ba3
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_027]
-
-   \sthana{27d}
-
-    \begin{description}
-    \item[candrasthiratvaṃ ca dadāti] Ba5, J7, J10ac, J17, LD1, LD2, V6, B2, Bo1, Bo3, N4a, V4, V22, N10, N13, P1, Tue, P8, P22, R5, V13, V10
-      \item[candraṃ sthiratvaṃ ca dadāti] P9("ca" in margin)
-    \item[candre sthiratvaṃ ca dadāti] C9,
-    \item[candrasthiratvaṃ hi dadāti] B1, C2, N1, R3
-    \item[cādrasthiratvaṃ hi dadāti] B3
-    \item[candrasthiratvaṃ pradadāti] C1, N24
-    \item[caṇḍasthiratvaṃ ca(?) dadāti] N7,
-    \item[daṇḍasthiratvaṃ ca dadāti] A1, Ba, Bo2, C3, C6, C8, J1, J2, J8, J10pc, J11, J12, J14, J15, J16, K1, Ko, N2, M1, N17, V1, V2, V3, N6, N9, N12, P10, V20, N16, N18, N19, N20, N22, N26, P6, P18, P25, V17, V12, R2
-    \item[daṇḍāsthiratvaṃ ca dadāti] P5
-    \item[daṇḍasthira.ntya karoti] G1
-    \item[daṇḍāsthitatvaṃ ca dadāmi] V14
-    \item[daṇḍaṃ sthiratvaṃ ca dadāti] P7, P11, 
-    \item[daṇḍasthiratvaṃ hi dadāti] J13, P2
-    \item[daṇḍasthiratvaṃ hi dadātti] P4
-    \item[daṃḍaschiratvaṃ hi dadāti] P12, 
-    \item[daṇḍasthiratāṃ ca dadāti] V21
-    \item[daṃḍaschiratvaṃ dadāti] J5,
-    \item[daṇḍasthiratāṃ sa dadāti] J3, J6,
-    \item[daṇḍasthiratvaṃ vidadhāti] C4, L1, V28, V18, N25, V16
-    \item[daṇḍasthiratvāṃ vidadhāti] P3
-    \item[daṇḍasthiratvaṃ pradadāti] N8, N11, V19, N23
-    \item[daṇḍaṃ sthiratvaṃ pradadāti] P23, 
-    \item[daṃḍe sthiratvaṃ pradadāti] N5
-    \item[daṇḍasthiratvaṃ jvalanaṃ] V8
-    \item[(daṇḍa?)sthirasvaṃ ce dadāti] N2
-    \item[dehasthiratvaṃ ca dadāti] N21
-    \item[kāyasthiratvaṃ ca dadāti] V15
-    \item[kaḍasthiratvaṃ ca dadāti(?)] J4
-    \item[vīryasthiratvaṃ ca dadāti] P22a,
-    \item[pracaṃḍahrunmaṃḍalakhaḍanāstraṃ] P21, 
-    \item[pracaṃḍarugmaṃḍalakhaṃḍanāstraṃ] P13, P14, P17, P20, P24, P26, R4, 
-    \item[rugmaṃḍalākhaṃḍanakhaṃḍanāstraṃ] P27,
-    \item[picaṃḍarukmāṃḍanakhaṃḍanāstraṃ] P15,
-    \item[vicaṃḍayugmaṃḍalakhaṃḍanāstraṃ] P16,
-    \item[(unavailable/illegible)] C5, C7, N14, T, Ba3
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_028]
-\sthana{1.28b}
-   \begin{description}
-   \item[dorbhyāṃ padāgradvitayaṃ] A1, Bo1, C2, C6, C8, J4, V1, M1, B2, B3, N1, N4a, P11, P14, P18, P20, V21, V20, V22, G1, V28, N16, N19, N21, N22, N23, N24, P2, P6, P12, P13, P17, P26, R4, R5, V13, R2
-   \item[dorbhyāṃ padāgredvitayaṃ] V14
-   \item[dorbhyāṃ padāgradvītayaṃ] P7
-   \item[dorbhyāṃ padāgnadvittayaṃ] P4 (could be "padāgra")
-   \item[dorbhyāṃ padāgraṃ dvitīyaṃ] Ko, V15
-   \item[dorbhyāṃ padāgradvitīyaṃ] R3, P25, P27, 
-   \item[dorbhyāṃ padāgraṃthidvayaṃ] N2,
-\item[dorbhyāṃ padagradvitayaṃ] Bo3, Ba5, K1, N13, P14, P22, Tue, V10
-\item[dorbhyāṃ padāgryau dvitayaṃ] J5
-\item[dorbhyāṃ pādāgrādvitayaṃ] B1
-\item[dorbhyāṃ ca pādadvitayaṃ] C4, L1, V19, N5, N7, N8, N11, N12, P3, P16, P23, 
-\item[dorbhyāṃ ca pādaṃ dvitayaṃ] C1
-\item[dorbhyāṃ ca pādau dvitayaṃ] N25
-\item[dorbhyāṃ ca pādāgradvitayaṃ] J13
-\item[dorbhyāṃ sa pādāgradvayaṃ] J2
-\item[dvābhyāṃ karābhyāṃ dvitayaṃ] Bo2, C3, C9, J1, J3, J6, J7, J8, J10, J15, J16, J17, LD1, LD2, N17, V6, V4, V3, N9, N10, N18, N20, V18, N26, P1, P5, P10, P24, V17, V16, LD1, P9("karābhyāṃ" in margin)
-\item[dorbhyāṃ ca pādadvayakaṃ] P21, 
-\item[dvābhyāṃ karābhyāṃ dvitīyaṃ] N6, 
-\item[dvābhyāṃ karābhyāṃ hitayaṃ] J12, V12
-\item[dvābhyāṃ padāgraṃ dvitayaṃ] J11
-\item[dvābhyāṃ pādāgrādvitīyaṃ] P8
-\item[(unavailable/illegible)] Ba, C7, G11
-\item[dorbhyāṃ padagraṃthidvayaṃ] J14, N2
-\item[.. rbhyāṃ pādaṃbhyāṃ dvitayaṃ] V8
-\item[(unavailable/illegible)] C5, N14, T, Ba3, 
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_035]
- \sthana{35b}
-   \begin{description}
-\item[asyahṛdaye dhṛtvā]    J7, J12, A1, V4, V18, V16, P8
-\item[asya hṛdayaṃ dhṛtvā]   P9
-\item[āsyahṛdaye dhṛtvā]    N10, V1, P1, P6
-\item[āsyahridayaṃ dhṛtvā]    J16
-\item[hṛdayaṃ kṛtvā] G1
-\item[eva hṛdaye dhṛtvā] C4, C7, C8, J5, J10, J11, J15, J17, Ko, V19, N2, N5, V21, N25, P3, P20, P23, P24, P25, V15, V14, R8, V12
-\item[ava hṛdaye dhṛtvā] P11, 
-\item[eva hṛdaye dhṛtvā]    P21,
-\item[eva hṛdaye kṛtvā]    Bo3, K1, N4a, V22, N13, N19, N24, P14, P27, Tue, V10
-\item[eva niyataṃ dhṛtvā]    C1, C3, J14, R4, V8, V28
-\item[eva niyataṃ kṛtvā] B1, B3, Bo2, C2, C6, J1, J3, J6("sudṛdha" is also given above), J8, J13, L1, N17, LD1, N1, V3, N12, V20, N16, N18, N20, N21, N26, P2, P4, P7, P12, P13, P17, P18, P26, R3,
-  R5('hṛdaye' canceled and replaced by 'niyataṃ')
-\item[ekahṛdaye dhṛtvā]    J4, N3, N23, P15, G11
-\item[ekahṛdayo dhṛtvā]    J2, M1
-\item[eksak ahṛdaye dhṛtvā]    P10, 
-\item[eva sudṛḍhaṃ kṛtvā]    N8, N11
-\item[eva dṛḍhayan dhṛtvā]    P16,
-\item[vāmaniyataṃ dhṛtvā]    V6
-\item[samyak(ū?) hṛdaye dhṛtvā]    N9    
-\item[āsyahṛdaye kṛtvā]    N22, V17
-\item[asya hṛdayeḥ dhṛtvā]    C9 ("āsyaṃ hṛdaye yasya saḥ" is given above)
-\item[meḍhre pādam atho hṛdi svacibukaṃ]    N7
-\item[meṃḍhraṃ pādam athaikam]            Bo1, LD1, LD2
-\item[meḍhre pādam athaikm eva hṛdaye]    B2, Ba5,
-\item[meṃḍhreṃ pādam athaikam]            LD2
-\item[meḍhraṃ pādam athaika samyak meva hṛdaye dhṛtvā]           P5
-\item[(unavailable/illegible)]    Ba, C5, V13, N6, N14, P22, R2, T, Ba3
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_038]
- \sthana{38a}
-   \begin{description}
-   \item[yameṣv eva mitāhāra(ḥ)] C2, J10, J17, M1, V1, V6, P8
-   \item[yameṣv eva sitāhāra] V16
-   \item[yameṣv eva mitāhāro] N7,
-   \item[yameṣv eva mitāhāram] C4, B2, J14, N2, N5
-   \item[yameṣv eva ātmacetana] V28
-   \item[yameṣv iva mitāhāra(ḥ)] B1, B3, Bo1, Bo2, C3, C9, J3, J6, J8, J13, J15, K1, Ko, LD1, N17, N1, V3, V4, N9, N10, N14, N16, N18, N21, N26, P1, P2, P4, P12, P17, R3, V12, LD1, P9
-   \item[jameṣṭiva mitāhāram] J5
-   \item[yameṣv iva mitāhāro] P13, R4, V18
-   \item[yameṣv iva mitāhāre] P5, P24, 
-   \item[yameṣv iva mītāhāra] J16,
-   \item[yameṣv iva mitāhāram] A1, Bo3, Ba5, C1, C7, C8, J4, J11, L1, LD2, V19, N4a, V22, V21, N13, N23, G11, N24, N25, P3, P7, P11, P14, P15, P16, P18, P20, P21, P23, P25, P26, P27, Tue, V15, R5, V13, V10
-   \item[yamiṣviva mitāhāram] V14
-   \item[yameṣv iva mitāhā(ḥ)] V8
-   \item[yameṣv iva mityāhāraḥ] P10, 
-   \item[niyameṣu mitāhāro] N8, N11, N12
-   \item[niyameṣu mitāhārā] J12,
-   \item[(unavailable/illegible)] Ba, V20, R2
-   \item[yameśiva mitāhāram] N19
-   \item[yameṣv eva mitāhāra] N20, V17
-   \item[yameṣu ca nirāhārī] N22, P6
-   \item[yameṣu ca mitāhārm] C6,
-   \item[yamaṣviva mitāharm] J1
-   \item[niyameṣu mitāhari] J2
-   \item[yameṣv api mitahāraḥ] J7("yamesṣv iva" is deleted)
-   \item[(unavailable/illegible)] C5, N6, T, Ba3, G1, P22,
-    \end{description}
- \end{marma}
-
-\sthana{1.38b}
-   \begin{description}
-\item[ahiṃ niyameṣv iva]    B1
-\item[ahiṃsā niyameṣu ca]    J4, M1, V1, N22, P6
-\item[ahiṃsā niyameṣv api]    B2, J14, N2, N24, V14
-\item[ahiṃsā niyameṣv iva] A1, B3, Ba5, Bo1, Bo2, Bo3, C2, C3, C4, C7, C8, C9, J1, J3, J5, J6, J7, J8, G11, J10, J11, J13, J15, J17, Ko, LD1, LD2, N17, V3, V4, N5, N9, V21, N13, N14, N16, N19, N20, N21, V18, N25, N26, P1, P2, P4, P7, P10, P11, P12, P13, P15, P16, P17, P18, P23, P24, P25, P26, V16, P8, R3, R5, V13, V12, LD1, P9, P14,
-\item[ahiṃsā niyameṣv ivaṃ] P21,
-\item[ahiṃsā niyāmaṣv iva]    P5
-\item[ahiṃsā niyameṣu eva]    V17
-\item[āhiṃsā niyameṣv iva]    Ko,
-\item[ahiṃsā nīyame sthitā]    K1,
-\item[ahiṃsāṃ nīyame sthitā]    P27, 
-\item[ahisā nīyameṣv iva]    J16,
-\item[aṃhiṃsā niyamaṣv iva]    J12,
-\item[svāhiṃsā niyameṣv iva]    V6
-\item[ahiṃsāṃ niyameṣv iva] L1, V19, N4a, V22, P3, P20, Tue, V15, V10
-\item[ahiṃsāṃ niyameṣu ca]    C6,
-\item[ahiṃsā niyaṃmai(?)ṣv iva]     N1
-\item[(a)hiṃsā ca niyameṣv iva]    N7,
-\item[yathāhiṃsā yameṣv iva]    N8, N11
-\item[ahiṃsā niyameṣv iha]    N10
-\item[asiṃsā niyame sthata(ḥ)]           V8
-\item[yameṣv evātma-cintanaṃ]    N12
-\item[hiṃsā ca niyameṣv iva] R4, 
-\item[na hiṃsāṃ niyameṣv iva]    N23
-\item[kṣamasye ātmacintane]    J2
-\item[(unavailable/illegible)]    Ba, C5, N6, V28, V20, N18, R2, T, G1, Ba3, P22, P25, 
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_039]
-\sthana{39d}       
-       \begin{description}
- (ṇ/n and ś/s irrelevant)
-\item[suṣumṇā iva nāḍikā]    V1, N22, P6
-\item[suṣumṇā iva nāḍiṣu]     B2, J4, J14, K1, Ko, N2, V8, N19, V12, G11
-\item[suṣumṇā caiva nāḍiṣu]     P26, 
-\item[suṣumṇā ca nāḍiṣu]    J5,
-\item[suṣumṇā eva nāḍiṣu ]    V20
-\item[suṣumṇā iva nāḍiṣuḥ] N8, P15, 
-\item[suṣumiva nāḍiṣu]     V28
-\item[suṣumṇām eva nāḍiṣu]    J10ac, J17, N24, V14
-\item[suṣumṇām iva nāḍiṣu ]    A1, C1, C4, C6, C7, C8, C9, J1, J2, J3, J6, J11, J15, L1, M1, V19, N5, N7, N12, N16, N23, N25, P11, P16, P20, P21, P23, P25, P27, V15, V13
-\item[suṣuṇmām iva nāḍiṣu]    P7
-\item[suṣumṇāśva nāḍiṣu]    N11,
-\item[suṣumṇām iva ḍinoṣu]    P3
-\item[nāḍīnāṃ malaśodhanam]    B1, B3, Ba5, Bo1, Bo3, C2, C3, C9, J8, J12, J13, LD2, N1, N4a, N10, N13, N18, N20, N21, N26, V3, V22, P1, P2, P4, P12, P13, P14, P17, P18, Tue, V17, R3, R5, V10
-\item[nāḍiṣu malaśodhanam]  R4,
-\item[nāḍināṃ malasodhanaṃ]    J16,
-\item[nāḍīnāṃ malaśodhanā]   P9
-\item[nāḍīnāṃ malaśodhane]    J7, J10pc, LD1, N9, N17, V6, V21, N14(seems "dī"), V18, P5, P10, V16, P8, P24, V12, LD1
-\item[suṣumṇādiṣu śodhanam ]     V4
-\item(unavailable/illegible)]    Ba, Bo2, C5, N6, R2, T, G1, Ba3, P22, 
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_040]
- \sthana{1.40e}
-   \begin{description}
-\item[śramādau bahubhiḥ pīṭhaiḥ] V1
-\item[śramadair bahubhiḥ pīṭhaiḥ] B1, B3, Bo1, Bo2, C1, C2, C6, C8, J3, J6, J13, J14, K1, M1, G11, N1, N11, N12, N16, N21, N22, N23, N24, P2, P4, P7, P11, P12, P13, P16, P17, P21, P23, P25, P27, V14, R3, V13
-\item[śramadair bahubhiḥ pīṭhai]    P6,
-\item[śramadai bahubhiḥ pīṭhaiḥ]    Ba, C9mg, J5,
-\item[śramadair bahubhi pīṭhai]    N8
-\item[śramādaiḥ bahubhi pīṭhaiḥ]    V21
-\item[śramadairghyādibhiḥ pīṭhaiḥ]    C4, C7, L1, V19, N5, N25, P3, P20, 
-\item[śramadairdhyādibhiḥ pīṭhaiḥ]    V28
-\item[śamardda bahubhiḥ pīṭhaiḥ]    J1
-\item[sanastai bahubhyaḥ pīṭai]    V8
-\item[(samastair?) bahubhiḥ pīṭhaiḥ]    N2
-\item[samastair bahubhiḥ pīṭhaiḥ]    J2, J4, J11, Ko, N2, P15, 
-\item[samasta bahubhiḥ pīṭhai]    V15
-\item[kim ādyair bahubhiḥ pīṭhaiḥ] A1, C9, J7, J8, J10, J17, LD2, N17, P26, V6, V3, V4, N7, N14, V20,
-  N18(No visarga at the end), N20, V18, N26, P5, P10, V17, P8, V12, LD1, R2, P9
-\item[kim ādyair bahubhiḥ pīṭheḥ]    N9, V16
-\item[kim adyair bahubhiḥ pīṭhaiḥ]    N10,
-\item[kim ādyair bahubhiḥ piṭhaiḥ]    J15, LD1, P1, P24, 
-\item[kim anyair bahubhiḥ pīṭhaiḥ]    Ba5, C3, N4a, N13, P14, P18, Tue, R5, V10
-\item[kim anyair bahuniḥ pīṭhaḥ]    Bo3, V22
-\item[kim arthai bahubhiḥ pīṭhai]    J12, J16("adyair" is also given),
-\item[kiṃ śramair bahubhiḥ pīṭhaiḥ]    B2
-\item[pramadair bahubhiḥ pīṭheḥ]    N19
-\item[tathaikasmin eva dṛḍham]    R4, 
-\item[(unavailable/illegible)] C5, N6, T, G1, Ba3, P22,
-    \end{description}
-
- \sthana{1.40f}
-   \begin{description}
-\item[sadā siddhāsanasthiteḥ]    A1, V20, V12, R2
-\item[sadā siddhāsanasthiteḥ]    P26, 
-\item[sadā siddhāsane sati] B1, B3, Bo1, C2, C9, J7, J8, J10, J11, J12, J15, J17, LD1, LD2, N17, V1, V6, N1, V4, N7, N9, N10, G11, N20, N22, V18, N26, P1, P2, P4, P5, P6, P8, P9, P10, P12, P24, V17, V16, LD1,
-\item[sidhe siddhāsane sati]    V10
-\item[siddhe siddhāsane sati] P18,
-\item[sadā siddhāsane satī]    J16,
-\item[sadā siddhāsane satya]              V3
-\item[sadā siddhāsanaṃ(?)]    V8
-\item[sadā siddhāsanā nyāsāt]    V22
-\item[sadā siddhāsanābhyām sādyogī]    V14
-\item[yadā siddhāsane sati]    J2, J4, N2, P15,  
-\item[yadā siṃhāsane sati]    Ko,
-\item[siddhe siddhāsane sati]    Ba5, Bo3, C3, N4a, N13, P14, Tue
-\item[siddhe siddhāsane miti]    R5,
-\item[kiṃ sadā siddhasādhanaiḥ]    M1, P11, 
-\item[kiṃ syāt siddhāsane sati] Bo2, C1, C4, C7, J1, J3, J5, J6, J13, L1, P13, P16, P20, V19, B2, V28, N5, N16, N21, N25, P3, P17, P21, P23, V15, R3
-\item[kiṃ syāt siddhāsane mati(ṃ?)]    N8
-\item[kiṃ syāt siddhāsane sthite]    C6, C8, P25, V21
-\item[kiṃ syā siddhāsane sthite ]    P7
-\item[kiṃ syāt siddhāsana sthite ]    N23
-\item[kiṃ syāt siddhāsane schite]    Ba,
-\item[kiṃ vā siddhāsane sati]    J14, N12, N24
-\item[kiṃ syāt siddhāsanai(?)ti]    N18
-\item[kiṃ sadā sevanena tu]     K1, N19(1.40h is "baddhe siddhāsane sadā")
-\item[baṃdhe siddhāsane sadā]  P27,
-\item[baddhe siddhāsane sadā]  R4,
-\item[va??t siddhāsane sadā]  N11,
-\item[(unavailable/illegible/unable to read)]    C5, N6, N14, V13, T, G1, Ba3, P22,
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_046]
- \sthana{1.46cd}
-\item[(utta(m)bhya cibukaṃ ~)]
-\item[vakṣa sthāpayet ]    J8, J10, J17, LD2, V6, N7, P8, N18, V12, V3, P9("thā" unclear)
-\item[vakṣaḥ sthāpayet]     Bo1, B1, B2, C3, J1, J12, J15, J16(sthāpaye),  K1, LD1, V18, P1, P5, P10, P24, R4, V16, N10, N22, LD1
-\item[vakṣaḥ sthāpaye]     P26, 
-\item[vandā sthāpayet]    V17
-\item[vakṣa sthāpayat]     N17
-\item[vakṣe saṃsthāpayat(?)]    Ko,
-\item[vakṣe uthāpayet]    V4
-\item[vakṣa sthāpya?]    J7(A word is deleted there), C9, N5(sthāpya sevanaṃ)
-\item[vatsa sthāpayet ]    N20, N26
-\item[vakṣaṃ sthāpayet]     V1
-\item[vakṣasy āsthāpya]     C4, L1, M1, V19, P3
-\item[vakṣasy utthāya]    Bo2, C6, N16, N8, N11,
-item[vakṣasy stthāpya] G11
-\item[vakṣasy utthāpya] Ba5, Bo3, C1, J13, J14, N2, N13, N21, N23, N24, P13, Tue, N1, N4a, P14, P18, P23,  R5
-\item[vakṣasy usthāpya] P21,
-\item[vakṣasy utthāyo] P15,
-\item[vakṣasy utṭhāpyo]    J4, J11,
-\item[vakṣasy utṭhāgre]    J6
-\item[vakṣasy oschāpyo]    J5
-\item[vakṣasy osthāpyot]    P11, 
-\item[vakṣyasy utthāpyot patane...]    N19
-\item[vakṣyasy asthāpya]    N25
-\item[vakṣyasy āsthāpya]    P20,
-\item[vakṣyapyutchāpya]    Ba, 
-\item[vakṣyapyutthāpya]    P25, 
-\item[vakṣe sthāpayet]    N14, P2, P4, N9
-\item[vakṣe syu sthāpya pavanaṃ]     V10
-\item[vakṣye sthāpayet]    B3, C2, P17, 
-\item[vakṣye schāpayet]    P12, 
-\item[vakṣye pyutthāya]    C8
-\item[vakṣyesu uttānya]    V14
-\item[vikṣe sthāpayet]    R3
-\item[cakṣasya utthāpya]    V15
-\item[svakṣasy utthāpya]    P7
-\item[urghātopaticotanau karau kṛtvā tato dṛśaṃ?]    J2
-\item[dakṣasy utthāgret(?)]    J3
-\item[utkuṃcya cibukaṃ vakṣa sthāpayet ]       P6
-\item[sannāya cibukaṃ vakṣyasy utthāpya]      N2
-\item[yuktaṃ vakṣastha-cibukaṃ mākramet]    N12
-\item[(unavailable/illegible/unable to read)]    A1, C5, N6, P16, V13, R2, T, G1, Ba3, P22, P27, 
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_048]
- \sthana{1.48b}
-   \begin{description}
-\item[(gāḍhaṃ vakṣasi sannidhāya cibukaṃ ~)]
-\item[dhyānaṃ tataś cetasi]    J2, V1
-\item[dhyānaṃ tataś cepsitaṃ]    J15, C9
-\item[dhyānaṃ tataś cośitaṃ]    V6
-\item[dhyānaṃ tataś conmitaṃ]    P5, P24, 
-\item[dhyānaṃ tataḥ conmī]    N1(some syllables seem deleted, 'nmī' is added later)
-\item[dhyānaṃ ca tac cetasi] B1, B2, B3, Ba, Bo1, Bo2, C1, C2, C4, C6, C7, C8, J4, J5, J8, G11, J11, J12, J13, K1, Ko, L1, V19, N25, N26, P2, P3, P4, P6, P7, P12, P13, P15, P17, P26, V4(ṃ omitted), V17, V16, N3, N5, N8, N11, V15, V14, P8, P20, P25, R3, V13, N14, N16, N17,N19, V12, V3, N21, N23,  N24(saṃniyamya), R2, P9
-\item[dhyāyaṃś ca tat cetasi] P14,
-\item[dhyānaṃ ca tecesi]    V8
-\item[dhyānena tac cetasi] P21, P23, 
-\item[dhyānaṃ ca tac cepsitaṃ] A1, J1, J3, J6, J10, J14, J16, J17, LD1, P1, P10, N7, N9, N10, N12, N18, LD1, N20, N22, P27, 
-\item[dhyānaṃ ca yac cepsitaṃ]    C3,
-\item[dhyānaṃ ca tac cepsite]    J7,
-\item[dhyānaṃ ca tatrepsitaṃ]    LD2, N2,
-\item[dhyāyaṃś ca tac cetasi]    Ba5, Bo3, M1, Tue, N4a, N13, P18, V10, V22
-\item[dhyāyaṃś ca taṃ cepsitaṃ]    R5
-\item[(unavailable/illegible)]   C5, N6, T, G1, Ba3, P11, P16, P22, R4, 
-    \end{description}
-
-\sthana{1.48c}
-  \begin{description}
-\item[(vāraṃ vāram apānam ūrdhvam anilaṃ ~)] G11
-\item[protsālayan pūrayan]    V1
-\item[protsārayan pūrayan]    J7, V6, N7, N12
-\item[protsārayaḥ pūrayem]    V8
-\item[protsārayan pūritaṃ]    A1, Ba5, Bo3, C1, C6, C8, J10, P14, P15, P26, Tue, N4a, V14, R5, V13, N13, P18, V10, V22, R2, G11
-\item[protsārayat pūritaṃ]    P7, P25, 
-\item[protsyā?rayat pūritaṃ]    Ba,
-\item[protsārayan pūrayet]    C4("proccālayan pūritaṃ" is also given above),L1, N25, P3
-\item[protsārayet pūrayan]    P20, V19
-\item[protsārayet pūrayet]    C7
-\item[protsāraṇaṃ pūrayet]    N5, P21, P23, 
-\item[protthāya yat pūrayan]    B3
-\item[protthāpanṃ pūrayet]    Bo1,
-\item[prodhāyana pūrayet]    B1,
-\item[protthāpayan pūrayan]    C2, P2, P4, N1, R3, P9, P12, P17, 
-\item[procārayet pūritaṃ]    J8
-\item[prottāsayan pūritaṃ]    V12
-\item[prollāsayan pūritaṃ]    J10, J17, N17, P1, P5, P24, 
-\item[prollāśayet pūritaṃ]    J15, N9
-\item[prolāsayet pūrayan]    J12,
-\item[prollāsayet pūrayan]    N10
-\item[prollosayana pūrayan]   LD2
-\item[prollāsayet pūrayet]    J16,
-\item[prollāsayan pūrayan]    C9, V16, P8, V4
-\item[pro(llāpa?)yan pūrayan] LD1
-\item[prollāsat pūritaṃ]    C3,
-\item[prollāghayan pūrayan]    LD1
-\item[procāraranu pūritaṃ]    J5
-\item[proccārayan pūritaṃ]    M1, V17, N3(vālaṃ vāram)
-\item[proccārayan pūrayan]    J4,
-\item[proccārayat puraya yataṃ(?)]    J1,
-\item[proccārayat pūritaṃ]    N2, P27, 
-\item[proccārayet pūritaṃ]    V15, V3, N23
-\item[proccārayet pūrayet]    N19
-\item[proccālayan]    K1
-\item[proccālayan pūrayen]    B2
-\item[proccālayan pūrayet]    J13,
-\item[proccālayaṃn pūrayan]    P6, N22(vvāraṃ cāram apānamūlapranilaṃ), \item[proccālayan pūrayan]    N18, N24(ayāmūrdh...)
-\item[proccālayan pūritaṃ]    J2, J14, Ko, N26, N20(proccālayaṃn)
-\item[proccālayet pūritaṃ]    Bo2, J3, J6("proccāṭayan pūrayet" is also given below), N16
-\item[proccālitaṃ pūritaṃ]    N14(vālaṃ vālam)
-\item[….... śayat pūritaṃ]    P10, 
-\item[proccānalaṃ pūrayen]   P13, N21
-\item[prodbhā(ṭa?)yan pūrayet]    N8
-\item[prod(ghā?)ṭayan pūrayet]    N6, N11
-\item[(unavailable/illegible)] T, G1, Ba3, P11, P16, P22, R4,
-
-\sthana{1.48d}
-  \begin{description}
-\item[(muñcan prāṇam upaiti bodham atulaṃ ~)]
-\item[śaktiprabhāvān naraḥ] B1, B3, Ba5, Bo2, Bo3, C1, C2, C6, C8, J1, J3, J6, J8, J11, J13, J14, K1, Ko, M1, V1, P2, P4, Tue, N2, N3, V15, V14, R3, R5, V13, N13, N16, N19, P12, P13, P14, P15, P17, P18, P25, V10, V22, N23, P27, G11
-\item[śaktiprabhāvān nara]    V3
-\item[śaktiprabhāvā naraḥ]    V8, N22(mucan...)
-\item[śaktiḥ prabhāvān naraḥ]    Ba,
-\item[śaktiṃ prabhāvān naraḥ]    Bo1,
-\item[śaktiṃ svabhāvān naraḥ]   N21
-\item[saktiprabhāvān]    J5
-\item[śaktiprabhāvon naraḥ]    N1, N7
-\item[śaktiprabhāvīn naraḥ]    P6
-\item[śaktiprabhāvaṃ naraḥ]    J12, N4a
-\item[śaktiprabhān naraḥ]    B2
-\item[śaktiprabhāvāṃtaraḥ]    N18
-\item[śaktiprabhāvodayāt] C4, C7, L1, V19, N25, P20, V16, N5, N14, P21, P23, P26, 
-\item[śaktiḥ prasādān naraḥ]    J2,
-\item[śaktimabhāvodayāt]    P3 ("ma" could well be "pra)
-\item[śaktiḥ prabhāvodayāt]    A1, R2("vo" illegible due to damage)
-\item[śaktiprabodhodayāt]    N20
-\item[śaktiprabhāvodayaḥ]    J7, P10, V6
-\item[śaktiprabhāvād ataḥ]    C3, C9, J10, J16, P5, P7, N9, N10, N11, N12, P8, V12, LD1, N24, P9
-\item[śaktīprabhāvad ataḥ]    LD2, P24, 
-\item[śaktiprabhāvad ata]    J15, N8, V4
-\item[śaktiprabhāvataḥ]    J17
-\item[śaktiprabhāvadhataḥ]    V17
-\item[śaktiprabodhodayāt]    N17, N26
-\item[śaktiprabhāvodhayaḥ]    P1
-\item[bhakti prabhāvodayaḥ]    J4,
-\item[(unavailable/illegible)]   C5, N6, T, G1, Ba3, P11, P16, P22, R4,
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_049]
-\sthana{1.49c}
-  \begin{description}
-\item[mārutaṃ nayate yas tu]    C6, C8, P11, 
-\item[mārutaṃ niyato yas tu]    Ba, P25, 
-\item[mārutaṃ niyatam yaskta]    V14
-\item[mārutaṃ niyatam yas tu]    V8, N23(mār(ū?)taṃ)
-\item[mārutaṃ mriyate yas tu] G11
-\item[mārutaṃ nīyate yas tu]    C1,
-\item[mārutaṃ pīyate yas tu]    V1
-\item[mārutaṃ pīvyate yas tu]    J8ac, V3
-\item[mārutaṃ yaḥ pīvet nitya]    V13
-\item[mārutaṃ pibati yas tu]    J8pc
-\item[mārutaṃ pibate yas tu]    J11,
-\item[mārutaṃ yas tu pibati] A1, C9, J7, J10, J12, J15, J17, LD2, N17, V6, N26, P1, P5, P6, P7, V17, V16, N7, N9, N10, P8, N14, V4, LD1, N20, P24, R2, P9
-\item[mārutaṃ yas tu pibatī] P21, 
-\item[māruto bhidyate yasya]    N2, 
-\item[māruto yas tu pibati]    N22
-\item[mārutaṃ yaska pibati]    V12
-\item[māruta yas tu pibati]     R1
-\item[mārutaṃ yas tu pibaṃti]    C3,
-\item[mārutaṃ yas tu pībaṃti]    J16,
-\item[mārutaṃ yas tu vidhivat]    P26,
-\item[mārutaṃ dhārayet yas tu] Ba5, Bo2, Bo3, C2, C7, J1, J3, J6, J13, L1, M1, V19, N25, P2, P3, P4, P14, P18, P20, P23, Tue, N4a, N11, N12, P12, P13, R5("), N13, N16, V10, N24,
-\item[māruto dhārayed yas tu]    N1, P17, R3, V22
-\item[māruta dhāraye?as tu]    N21
-\item[mārutaṃ dhāraye yas tu]    N8, N18
-\item[marutaṃ dhārayed yas tu]    C4, N5, V15
-\item[mārutaṃ māryatd yas tu]    J5
-\item[mārutaṃ mriyate yas tu]    N3
-\item[mārutaṃ mriyate yasmāt]    Ko,
-\item[māruto mriyate yas tu]    J4, P15, 
-\item[māruto kṣiyate yasya] P27,
-\item[māruto bhidyate yasya]    B1, B2, Bo1, J14, K1, N2, N19
-\item[mahato dhārayed yastu]    B3
-\item[pavanṃ prāsayeta saṃstu]    J2
-\item[(unavailable/illegible)]    C5, N6, T, G1, Ba3, P10, P16, P22, R4, 
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_051]
-\sthana{1.51d}
-  \begin{description}
-\item[nāsāgre nyastalocanaḥ]    B1, B3, C2, C9, J7, J13, J16, N17, V1, V6, P2, P4, P17, N1, N7, P12, R3, N22
-\item[nāsāgre nyastalocanā]    P6, R4, 
-\item[nāsāgra-nyastalocanaḥ]    A1, Bo1, C3, J10, J17, N26, P1, P5, V16, N10, V12, V4, LD1, LD2, N20, P24, R2
-\item[nāsāgraṃ nyastalocanaḥ]  P9
-\item[nāsāgranyastalocanaṃ]    J8
-\item[nāsāgra vyasalocanaḥ]    P26, 
-\item[nāsāgranyastaloca....]    P10,
-\item[nāsāgraṃ nyastalocanaṃ]    P7, V17, V3
-\item[nāsāgra susamāhitaḥ]    C8,
-\item[nāsāgraṃ susamāhitaḥ] Bo2, Bo3, C1, C4("nāsāgra nyastalocanaḥ" is also given), J1, J2, J3, J6, J11, J14, J15, K1, Ko, L1, M1, V19, N25, P3, Tue, N2, N4a, N5, N12, V15, V14, V13, N13, N16, P13, P14, P15, P18, P23, V10, V8, G11
-\item[nāsāgrastu samāhitaḥ]   N23
-\item[nāśāgre susamāhitaḥ]    N19,
-\item[nāsāgre susamāhitaḥ]    P27,
-\item[nāsyagraṃ tu samāhitaḥ]    J5,
-\item[nāsāgre susamāhitaṃ]    P21, 
-\item[nāsāgraṃ susamāhitaṃ]    Ba5, C6, V22
-\item[nāsāgryaṃ susamāhitaṃ]    N18
-\item[nāsāgraṃ susamāhita]    N8
-\item[nāsāṃgraṃ susamāhitaḥ]    N11
-\item[nāśāgraṃ tu samāhitaḥ]    N3,
-\item[nāsāgraṃ tu samāhitaḥ]   N24, P20, 
-\item[nāsāgre susamāhitaḥ]    B2, Ba, K1, N21, P25, 
-\item[nāsāyaṃ susamāhitaḥ]    J4, P11,
-\item[nāsāyāṃ susamāhitaḥ]    J4,
-\item[nāsāṃ tu samāhitaḥ]    R5
-\item[nāsāgre neṣu nocana]    N14
-\item[(unavailable/illegible)]    C5, C7, N6, T, G1, Ba3, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_052]
-\sthana{1.52d}
-  \begin{description}
-  \item[kurute cāsanottamam] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J8, J10pc, J13, J15, L1, N17, V6, N26, P2, P3, P4, P5, P7, Tue, N1, V17, N4a, N5, N8, N11, N12, P12,  P13, P14, P17, P18, P20, P21, P23, P24, V15, R3, R5, V13, N13, N14, N18, N19, V8, V3, N20, N21, N23, P25, R2, R4, G11
-\item[kuruta cāsanottamam]    V22
-\item[kurute cāsanomatam]    V10
-\item[kurute cāsanottame]    B1, N16
-\item[kurute cāsanottamaḥ]    J2,
-\item[kurute cāsanotraye]    V14
-\item[kurute cāsanottama]    A1, P26, 
-\item[kurute cāsanontayam]   N24
-\item[kurute vāsanottamam]    V1, V19, N3, P11, 
-\item[kurute sādhanottamam]    C3, J12, J10ac, J16, J17, P1, V16, N9, N10, P9, P10, V12, V4,
-\item[kurute sādhanottaramaṃ]    P8(perhaps 'ra' is attempted to be canceled)
-\item[kurute sādhakottamam]    C9, N7,
-\item[kurute sādhakottamaḥ]    LD1, LD2
-\item[kurute cāsanam uttamaṃ]     N25
-\item[sevate cāsanottamaiḥ]    J14,
-\item[sevate cāsanottamaṃ]    J11, Ko, N2, P15,  
-\item[sevyate cāsanottamaṃ]    J4,
-\item[(unavailable/illegible)]    B2, C5, C7, K1, M1, N6, N22, P6, T, G1, Ba3, P16, P22, P27, 
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_054]
-\sthana{1.54b}
-  \begin{description}
-  \item[sarvavyādhi-viṣāpaham] C3, C9, J5, J7, J8, J10, J12, J15, J16, J17, LD1, V1pc, V6ac, P1, P5, P6, P7, P24, V17, N3, V16, N9, N10, P8, R1, V12, V3, V4, N22, P9, P11, G11
-  \item[sarvavyādhī-viṣāpaham] LD2
-  \item[sarvavyādhi-vināśanam] A1, B1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C7, C8, J1, J2, J3, J4, J6, J11, J13, K1, Ko, L1, N17, V1ac, V6pc, V19, N25, N26, P2, P3, P4, P17, P25, Tue, N1, N4a, N5, N7, N11, N12, P12, P13, P14, P15, P18, P20, V15, V14, R3, R5, N13, N14, N16, N18, N19, V10, V8, V22, N23, N24, P21, P23, P26, P27, R2, R4,
-\item[sarvavyāvināśanam]   N21
-\item[sarvadhyādhi-vighātakaṃ]   N20
-\item[sarvavyādhi-praṇāśanaṃ]    J14, N2, 
-  \item[sarvavyādhivikṣāpaham] P10, 
-\item[kaphavātāmayāpahaṃ]    M1
-\item[(unavailable/illegible)]    C5, N6, N8, V13, T, G1, Ba3, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_055]
-\sthana{1.55 additional verses}
-
-kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet |
-na śāstrapāṭhamātreṇa yogasiddhiḥ prajāyate ||
-na veṣadhāraṇaṃ siddheḥ kāraṇam na ca tatkathā |
-kriyaiva kāraṇaṃ siddheḥ satyam etan na saṃśayaḥ |
-śiśnodararatāyeha na deyo veṣadhāriṇaḥ ||
-
-\begin{description}
-\item[Yes] B1, B2, B3, Ba, Bo1, Bo2, C1, C3, C4, C6, C8, C9, J1, J3, J6, J7, J8, J10, J11, J12, J13, J14, J15, J16, J17, K1, L1, LD1, LD2, N5, N17, N25, P10, V3, V17, V25, P8, R1, R3, N14, N16, N18, V12, LD1, N7, N9, N10, N13, N20, N21, N26, P1, P3, P4, P5, P7, P14, P17, P21, R2, P9, P12, P24, P25, P26, P27,
-\item[Yes], but the 4th line appears as the 2nd line: C2, J13, N1, P2
-\item[No] A1, Ba5, Bo3, C1, C5, C7, J2, J4, J5, Ko, M1, V1, V6, V8, V19, R5, N19, P18, P20, V10, V22, V4 (unclear/unable to read) V13, N2, N3, N4a, N6, N8, N11, N12, N22, N23, N24, Tue, P6, P11, P13, P15, P16, P22, P23, R4, G11, T, G1, Ba3
-    \end{description}
-
-\sthana{1.55 third additional verse}
-
-mayi bodhāmbudhau svacche tuccho 'yaṃ viśvabudbudaḥ |
-pralīna udito veti vikalpapaṭalaḥ kutaḥ ||
-
- \begin{description}
- \item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, P12, P17, P20, P21, P23, P26, V6, V19, R3, N14, N16, N18, N1,
-
- \item[No] A1, B2, Ba5, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, LD1, LD2, M1, V1, V3, V8, V17, V25, P8, R1, R5, N19, V12, V10, V22, V4, LD1, V13, N2, N3, N4a, N6, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, P11, P18, Tue, P1, P5, P6, P7, P9, P10, P13, P14, P15, P16, P24, P27, R2, R4, G11, T, G1, Ba3, P22, P25,
-    \end{description}
-
-\sthana{1.55 fourth additional verse}
-% P: What is the difference between this and the previous extra verse?
-
-mayi bodhāmbudhau svacche tuccho 'yaṃ viśvabudbudaḥ |
-pralīna udito veti vikalpapaṭalaḥ kutaḥ ||
-
-\begin{description}
-\item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, V6, V19, R3, N14, N16, N18, N8, N11, P2, P3, P4, P12, P21, P23, 
-\item[No] A1, B2, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, M1, V1, V3, V8, V17, V25, P8, R1, R5, N13, N19, V12, V10, V22, V4, LD1, V13, N2, N3, N4a, N6, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, Tue, P1, P5, P6, P7, R2, P9, P10, P11, P13, P14, P15, P16, P17, P18, P20, G11, T, G1, Ba3
-    \end{description}
-
-\begin{marma}[hp01_056]
-\sthana{1.56a}
-  \begin{description}
-  \item[āsanaṃ kumbhakaś citro] J8pc, N24
-\item[pīṭhāni kuṃbhakāś citro] G11
-\item[pīṭhādi kuṃbhakāś citram] R4, 
-  \item[āsanaṃ kumbhakaś citraṃ] B2, C3, J7, J8ac, J10, J12, J15, N17, P8, N14(ā(?)naṃ), V12, LD1, LD2, N4a, N10, N12, N20, P1, P5, P24, P27, 
-  \item[āsanaṃ kumbhavaś citra] J2,
-\item[āsanaṃ kumbhavaś ci(nt?)aṃ] P9
-  \item[āsana kumbhakaś citraṃ] V3
-\item[āsanaṃ kumbhakāś citrāḥ]  N8, N11
-  \item[āsanaṃ kumbhavaś cittaṃ] P10, 
-\item[āsanaṃ kumbhakāś citrā]   P7
-\item[āsanaṃ kumbhakācitraṃ]   Tue
-  \item[āsanaṃ kumbhakaṃś citraṃ] C9, J17
-  \item[āsanaṃ kumbhakaṃ citraṃ] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C7, C8, J1, J3, J4, J5, J6, J11, J13, J14, K1, Ko, L1, M1, P11, P12, P13, P18, V19, R3, R5, N13, N16, N19, P14, P17, P20, P21, P23, P25, V10, V22, V4, N1, N5, N23, N25, N26, P2, P3, P4, P6, P26, R2
-  \item[āsanaṃ kumbhakaṃ citra] P15,
-\item[āsanaṃ kumbhakaṃ śvitraṃ]   N21   
-  \item[āsanaṃ kumbhaka citrāṃ] J16, N7,
-  \item[āsanaṃ kumbhaka citaṃ] N18
-\item[āsanaṃ kubhakaṃ citraṃ] N2, 
-\item[asanaṃ kumbhakaṃ citraṃ] N3, N9
-  \item[(unclear/unable to read)] V13
-\item[(unavailable/illegible)]    C5, N6, N22, R1, T, G1, Ba3, P16, P22,
-    \end{description}
-
-\sthana{1.56b}
-  \begin{description}
-  \item[mudrādikaraṇāni ca] A1, B2, C3, C9, J7, J8pc, J10, J12, J15, J16, J17, LD1, N17, P8, N14, V12, N7, N9, N10, P1, P5, P7, P9, P10, P24, R2, R4, 
-  \item[mudrānikaraṇāni ca] LD2
-\item[mudrādikāraṇāni ca]    V4
-\item[vicitrīkaraṇāni ca]    J2,
-  \item[mudrādikaraṇakriyā] P27,
-\item[mudrādi pavanakriyā]    B1, B3, Bo1, C2, J5, J8ac, J13, R3, V3, N1, N20, P2, P4, P12, P17, 
-\item[mudrādi pavanakriyāḥ]   N26
-\item[mudrākhyaṃ karaṇaṃ tathā] Ba, Ba5, Bo2, Bo3, C1, C3, C6, C7,  C8, J1, J3, J11, L1, M1, P11, P14, P15, P18, P20, P23, P25, P26, V19, R5, N13, N16, N18, N19, V10, V22, N5, N11, N12, N21, N23, N25, Tue, P3, P6, P21,
-\item[mudrākhaṃ karaṇaṃ tathā]   N24
-\item[mudrākhyaṃ karaṇaṃ bhavet]   N3
-\item[mudrākhyakaraṇaṃ tathā]    Ko, N4a, N8, P13, 
-\item[mudrādi karaṇaṃ tathā]    K1,
-\item[mudrākṣaṃ karaṇaṃ tathā]    J4, J6,
-\item[mudrakhecaraṇīṃ tathā]    J14(given at 55)
-\item[mudrakhecariṇīṃ tathā]    N2, 
-\item[(unclear/unable to read)]    V13
-\item[(unavailable)]     C5, N6, N22, R1, G11, T, G1, Ba3, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_060]
-\sthana{1.60cd}
-  \begin{description}
-\item[(atilavaṇaṃ tilapinḍaṃ kadaśanaśākotkaṭaṃ varjyam |)]
-\item[atilavaṇaṃ tilapinḍaṃ]    J17, LD2, N10, R1
-\item[atilavaṇaṃ tilapinḍa]    C3, J8pc, J10, J15, J16, LD1, N17, V4, N9, N20, P1, P5, P8, P9
-\item[atilavaṇaṃ tilapiṇḍā?]   N14
-\item[atilavaṇatilapiṇḍakadaśanaṃ]  P10, P24, 
-\item[atilavaṇaṃ tilaṃ pinḍa]    J8ac
-\item[atilavaṇaṃ tilaṃ pinḍaṃ]     V3
-\item[atilavaṇaṃ tilapiṇyākadarśanam]    N26
-\item[atilavaṇam amlayuktaṃ]    A1, B3, Ba, Ba5, Bo3, V22, V10, N1, N4a, N11, N13, Tue, P2, P12, P14, P17, P18, P25, P26, R2, R5
-\item[atilavaṇama malayuktaṃ]    C2,
-\item[atilavaṇāmalpayuktaṃ]    P13, 
-\item[atilavaṇamalayuktaṃ]   P4, R3
-\item[atilavaṇāmlayuktaṃ]   P7
-\item[atilavaṇa amlayuktaṃ]    C6, C8, V13
-\item[atilavaṇa amlayutaṃ]    J11,
-\item[atilavatilapi]    V6
-\item[atilavaṇatiktapiṇḍa]    B2
-\item[atilavaṇatilapiṇḍakaṃ]    C9
-\item[atilavaṇaṃ dyuṣṇaṃ]    V1
-\item[atilavaṇam apramuṣṇaṃ]    M1
-\item[atilavaṇam ca apramitaṃ]    V8
-\item[atilavaṇam cāpramitaṃ]   P6, P27, 
-\item[atilavaṇaṃ ca tilapiḍaṃ]   N7,
-\item[atilavaṇam pramitaṃ]    J12, K1,
-\item[atilavaṇādi prasaktaṃ]    J14,
-\item[atilavaṇādi proktaṃ]   N22
-\item[atilavaṇacāmlapiktaṃ]    N19
-\item[atilavaṇakavalṃ cāpamitaṃ]    J1,
-\item[atilavaṇasavapalala]    V19
-\item[atilavaṇasavapalalaṃ]   N25
-\item[atilavaṇāsavasalilaṃ]   P21, 
-\item[atilavaṇaṃ sapalalaṃ]   P20,
-\item[atilavaṇaṃ sala(pa?)laṃ]   N8
-\item[atilavaṇaṃ palilaṃ]     R4
-\item[atilavaṇāsavapalalaṃ]  C1, C7, L1, N5
-\item[atilavaṇāsavapalaṃla]  P23, 
-\item[atilavaṇātiproktaṃ]    J2,
-\item[atilavaṇādiyutaṃ]    J4,
-\item[atilavaṇādi prayuktaṃ] P15,
-\item[atilavaṇādi puttaṃkdaśanaṃ]    P11,
-\item[atilavaṇādyaptataptaṃ]    J5,
-\item[atilavaṇādyapran(t?)aṃ]  N3
-\item[atilavaṇādyapratapta]   N12
-\item[atilavanā?āna(va?)lala]   N24
-\item[atilavaṇamatprayuktaṃ]   N21
-\item[atilavaṇāmavapalalaṃ]    C4 ("atilavaṇaṃ tilapinḍaṃ" is also given),
-\item[atilavaṇāya priyatavṃ]    Ko
-\item[atilavaṇaṃ sapalaṃ ca]    Bo2, J3, J6,
-\item[atilavaṇasalilaṃ kadaṣana]    B1, Bo1, J13,
-\item[atilavaṇetilapiṃḍakvadaśanaṃ]    J7
-\item[atilavaṇaprasūtaṃ]     N2, 
-\item[atilavaṇaṃ kavalaṃ cāpramitaṃ piṇḍa]   N18
-\item[atīlavaṇāmlapṛktaṃ]   N23
-\item[(illegible/unavailable)]  C5, N6, N16, P3, G11, T, G1, Ba3, P16, P22,
-
-* * *
-
-\item[śākotkaṭaṃ] A1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C3, C4, C6, C8, C9, J2, J3, J4, J6, J10, J11, J12, J13, J14, J15, J17, K1, Ko, L1, LD1, LD2, V6, V19, V22, V4, V10, V13, N2, N3, N4a, N7, N11, N13, N18, N19, N21, N24, N25, Tue, P2, P5, P6, P7, P8, P9, P10, P11, P12, P13, P14, P15, P17, P18, P20, P21, P23, P24, P26, R1, R2, R3, R4, R5,
-\item[śākotkaṭhaṃ]    V8
-\item[śākotkavaṃ]    J5,
-\item[śākotkaṭa]    Bo2, J8pc, J16, N1, N9, N26, P27,
-\item[śākoktaṃ]    P25, 
-\item[śākotkarṭeṃ]  P4
-\item[śākātkaṭa]    V3
-\item[śākotchadaṃ]    Ba,
-\item[śākolkaṭaṃ]   P1
-\item[śākokṣadaṃ]   N23
-\item[śāṃkotkaṭaṃ]   N10
-\item[śāketkaṭaṃ]    C7
-\item[śokātkaṭa]    J8ac, V1
-\item[śokotkaṭaṃ]    B1, J7,
-\item[śāśotka]    M1
-\item[śātkṭaṃ]    J1,
-\item[śa?tāsanaḥ]   N14
-\item[sākotkaṭa]   N20
-\item[sākotkaṭaṃ]  N22
-\item[syacotkaṭaṃ] N12
-\item[(illegible/unavailable)]    C5, N6, N17, N8, N16, P3, G11, T, G1, Ba3, P16, P22,
-\item[kotkaṭaṃ]  N5
-
-* * *
-\item[varjyaṃ] A1, B1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C4, C7, C9, J3, J6, J13, J15, K1, L1, LD2, V3, V6, V19, V22, N1, N4a, N5, N11, N12(va?jyaṃ), N13, N22, N25, Tue, P2, P4, P6, P7, P12, P14, P17, P18, P20, P21, P23, P27, R3, R4, R5
-\item[varjya] V8, V10
-\item[varjyāṃ] J16
-\item[varj(j)aṃ] J7, J10, J12, J17, M1, N9, P10, P26, R1
-\item[varjjyaṃ]  P5, P24, 
-\item[varj(j)itaṃ] J1, Bo2, J8, LD1, V3, N18, N20, N26
-\item[varyyaṃ]   N7, R2, P9
-\item[varjjāṃ]   N19
-\item[varjet]  N10
-\item[varjayet] C3, P1("varjaṃ" earlier)
-\item[vayīṃ]   P8
-\item[duṣṭaṃ] C6, C8, J4, J5, J11, J14, Ko, V13, N2, N21, N23, N24, P13, P15, P25, 
-\item[du(ṣl?)aṃ]  N3
-\item[dṛṣṭaḥ]
-\item[durjvaṃ(?)] Ba,
-\item[(illegible/unavailable)] C5, N6, N17, V1, V4, N8, N14, N16, P3, G11, T, G1, Ba3, P11, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_062]
-\sthana{1.62d} 
-  \begin{description}
-  \item[mudgādi divyam udakaṃ ca] A1, B2(dīvyam), Ba, Ba5, Bo3, C6("dgā" is missing), C8, J5, J7, J10, J11, J12, J16, J17, K1, L1pc, LD1, LD2, N17, V1, V6, V8, V22, V4, V13, N1, N10, N12, N13, N14, N20, N22, N23, N24, N26, Tue, P1, P6, P8, P9, P14, P17, P18, P23, P25, P26, P27, R1, R2,
-\item[mudgādi divyam udakaṃ]   N18, P11,
-\item[mudgādi divyam udakaṃ ka]   N4a
-\item[mudgādi divyam udake ca]   R5
-\item[mudgādi divyam udakaṃ yamīndra]    V3, N2
-\item[mudgādi divyam udakaṃ hi]    C3, J14, J15, N9, P5, P10, P15, P24, 
-\item[mudgādi pivyam udakaṃ ca]    J2,
-\item[mudgājya divyam udakaṃ ca]    C9
-\item[mudgādivyam udakaṃ ca]    C4, J8, M1, V10, N3
-\item[mudgādi divyagaditaṃ ca]   N7,
-\item[mudgādi divyamuktaṃ]    Ko,
-\item[mudgādi cālpam udakaṃ ca]    B1, B3, Bo1, Bo2, C1, C2, C7, V19, N5, N11, N21, P2, P4, P12, P13, P20, P21, R3, R4
-\item[mudgādi cālyam udakaṃ ca]    L1ac, N8(no "ca")
-\item[mudgādi cājyam udakaṃ ca]    N16
-\item[mudgādi cālpa sudakaṃ ca]     J1
-\item[mudgādi dālpam udakaṃ ca]    J3, J6("cālpa" is also given)] J13,
-\item[mudgādi dālpamudgaṃ ca]    N25
-\item[mudrādi divyam udakaṃ ca]  P7
-\item[mudrādi divyam udakaṃ ca mudakaṃ]    N19
-\item[(illegible/unavailable)]       C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22,
-    \end{description}
-
-\sthana{1.62d}
-  \begin{description}
-  \item[munīndrapathyaṃ (or: munindra-)] B1, B2, B3, Bo1, C1, C2, C4, C7, J3, J5, J6, J13, J16, J17, K1, L1, LD2, V1, V19, N1, N5, N11, N19, N21, N22, P2, P4, P12, P13, P17, P20, P21, P23, P26, R4
-\item[munīdrapathyaṃ]   N8, R3
-\item[munīndrapathaṃ]   N25
-\item[mahendrapathyaṃ]  N9, P5, P10, P24, 
-  \item[yamīndrapathyaṃ (or: yamindra-)] A1, Ba, Ba5, Bo3, C6, C8, C9, J7, J8, J10, J11, J14, LD1, M1, N17, V6, V22, V3, V4 (pathyaṃ missing), V10, V13, N2, N3, N4a, N12, N13, N14, N20, N26, Tue, P1, P6, P7, P8, P14, P18, P25, P27, R2, R5, P9
-\item[yamīṃdrapraparyya]   N24
-\item[yamīḍyapathyaṃ] P15,
-\item[yatīṃdrapathyaṃ]    J2, N10, N23
-\item[yavatīṃdrapathyaṃ]    P11,
-\item[yamendrapathyaṃ]    C3,
-\item[yogīdrapathya]    J1,     
-\item[mūnidra pathayam]    V8
-\item[munidra pathayam]    J12,
-\item[munīdramukhyaiḥ]    N7,
-\item[muṃhanīdrapathya]    J15
-\item[suyogipathyaṃ]     N16
-\item[varayogipathyaṃ]    N18
-\item[jamodūpathyaṃ]    R1
-\item[paramaṃ hitaṃ hi]    Ko,
-\item[(illegible/unavailable)]   C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_063]
-\sthana{1.63a}
-  \begin{description}
-\item[iṣṭaṃ sumadhuraṃ]    A1, J3, J6, J10, N18(appears after one verse), P7, R2, P9
-\item[iṣṭaṃ sumadhura]    P11,
-\item[iṣṭaṃ samadhuraṃ]    C9, J7, J17, LD2, N17, V19, V4, LD1, P1, P8, R1
-\item[iṣṭaṃ saṃmadhuraṃ]   N7,
-\item[iṣṭaṃ samakaraṃ]   P10, 
-\item[daṣṭaṃ samudhūraṃ]    J16,
-\item[dṛṣṭaṃ samūdhuraṃ]    K1
-\item[dṛṣṭaṃ sumūdhuraṃ]    N10
-\item[piṣṭaṃ sumadhuraṃ]    C8, N8, N11, N14("śu")
-\item[pathyaṃ sumadhuraṃ]    J12,
-\item[puṣṭaṃ sumadhuraṃ]    Ba5, Bo3, N4a, N13, P14, P18, Tue
-\item[puṣṭaṃ samadhuraṃ]   R5
-\item[yeṣṭaṃ sumadhura]    V13
-\item[yuṣṭaṃ sumadhuraṃ]    V10(mu°), V22
-\item[mṛṣṭaṃ sumadhuraṃ]    M1, V1
-\item[mṛṣṭaṃ sumadhurāṃ]    Ko
-\item[mṛṣṭaṃ sumadhusnigdhaṃ]   N3
-\item[mṛṣṭaṃ samadhuraṃ]    C3,
-\item[miṣṭaṃ samadhuraṃ]    J8, Ba, C4, N20, N26, P5, P24, 
-\item[miṣṭaṃ sumadhuraṃ]    Bo2, C6, J1, J11, J14, N2, N16, N19, N21, N22, N23, P13, P15, P21, P25, P27, R4
-\item[miṣṭaṃ samudhuraṃ]    J5, J15,
-\item[miṣṭaṃ sūmadhuraṃ]  P6
-\item[miṣṭaṃ sumadhura]   N25
-\item[miṣṭaṃ madhuraṃ]    V3
-\item[miṣṭaṃ ca madhuraṃ]    B2
-\item[maṃ(ṣṭ?)aṃ samudhuraṃ]  N9
-\item[sumiṣṭaṃ madhuraṃ]    N12
-\item[uṣṇaṃ sumadhuraṃ]    B1, Bo1, C2, C7, L1, J13, N1, P2, P17, P23, 
-\item[uṣṇaṃ sumudhuraṃ]    P20,
-\item[uṣṇaṃ sumadhura]    P12, 
-\item[uṣṇaṃ samadhuraṃ]    C1,
-\item[uṣma(?) sumadhuraṃ]    B3
-\item[uṣmaṃ sumadhuraṃ]    P4, R3(appears after one verse)
-\item[uśmā?mumadhuraṃ]  N5
-\item[suvṛkṣaṃ madhuraṃ]    V6
-\item[ghṛtam ca madhuraṃ]    V8
-\item[vṛṣyaṃ sumadhuraṃ]    J2,
-\item[tiṣṭaṃ tu madhuraṃ]  N24
-\item[(illegible/unavailable)]   C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22, P26,
-    \end{description}
- \end{marma}
-
 Notes:

### 2023-09-02 marma/marma3.tex
new routine for jyotsna

--- 2023-09-02 marma/marma.tex
@@
-Notes
-    • P16 has chapter 1 till 1.45
-    • P22 has only 3 folios of chapter 1
-\input{Preamble-Print}
-\makepagestyle{HPed}
-\makeoddhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
-\makeevenhead{HPed}{\small\texteng{HP1 OldMss}}{}{\small\texteng{\today}}
-\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
-\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}
-
-\newcommand{\myfn}[1]{\footnote{\texteng{#1}}}
-\setlength{\footmarkwidth}{1.3em}
-\setlength{\footmarksep}{0em}
-\setlength{\footparindent}{1.3em}
-\footmarkstyle{\textsuperscript{#1} }
-\usepackage{fnpos}
-\makeFNbottom
-
-% additions/changes 2022-03-12 mm:
-\SetTEIxmlExport{autopar=false}
-\NewDocumentEnvironment{marma}{O{}}{}{}
-\def\sthana{}
-
-\begin{document}
-\pagestyle{HPed} %
-\begin{ekdosis}
-  \SetLineation{lineation = none,}
-
-  \chapter{Variants and Versions}
-
-\sthana{24a}
-
-  \begin{marma}[hp01_024]
-    \begin{description}
-%    \item[(ku(r)k(k)uṭāsana-)]{}
-    \item[-vat kṛtvā]{V1, N22}
-    \item[-vat kratvā]{P6}
-    \item[-bandhastho]{A1, J8, L1, M1, V6, B1, B2, B3, Bo2, C1, C2, C3, C4, N2, C6, C8, J1, J3, J6, J13, J14, K1, Ko, LD1, LD2, N1, V2, V3 (about \emph{stho}), N4a, V8, N5, N8, N9, N11, V21, N13, N16, N18, N19, N20, N21, N24, N25, N26, P1, P2, P3, P4, P7, P10, P11, P14, P15, P17, P20, P21, P22, P23, R4, Tue, V17, V14, R3, R5, V13, V10, G1
-    \item[-bandhasthe]{J7, V19}
-    \item[-badhascho]{Ba, Bo3, V28, V22}
-    \item[-baṃdhascho] J5, J11, J15, P12, P13, P18, P27, V15
-    \item[-baddhastho] N12, P25, 
-    \item[-bandhastu]{Bo1}
-    \item[-madhyastho]{C9, J10, J12, J16 (\emph{baṃ} is given in margin), J17, N17, V4, N6, N7, N10, V20, V18, P5, P8, P24, P26, V12, LD1}
-     \item[-madhyasthā]{P9}
-    \item[-madhyasthau]{V16}
-    \item[-baṃdhasthau]{N23}
-    \item[-badhestu]{J2}
-    \item[-vadhnastho] P16, 
-    \item[(unavailable/illegible)] Ba5, C7, J4, N14, C5, G11, R2, T, Ba3
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_025]
-
-  \sthana{25c}
-
-  \begin{description}
-\item[dhanurākarṣaṇaṃ kṛtvā]
-\item[dhanusaṃkarṣaṇaṃ kuryād] Bo2,
-\item[dhanuḥsaṃkarṣaṇaṃ kuryād] N9,
-\item[dhanurākarkhaṇa kṛtvā] V19,
-\item[dhanurākarṣakākuṣṭaṃ] M3,
-\item[dhanurākarṣaṇā[kṛ]ṣṇaṃ] P7,
-\item[dhanurākarṣaṇaḥ kṛṣṭaṃ] V1,
-\item[dhanurākarṣaṇākarṣaṃ] Ko,
-\item[dhanurākarṣaṇākṛṣṭaṃ] Ba, C6, C8, G5, G11, J5, J11, M1, N12, N20, P15, V15, V2, V25, 
-\item[dhanurākarṣeṇākṛṣṭaṃ] J2,
-\item[dhanurākarṣaṇaṃ kṛtvā] B2, C1, C4, J1, J3, J6, J10, J14, J15, K1, L1, MYO20, N5, N11, N16, N18, N19, N22, N24, N25, P3, P6, P13, P23, Pen,
-\item[dhanurākarṣana kṛtvā] N21,
-\item[dhanurākarṣaṇaṃ kṛṣṭa] J4,
-\item[dhanurākarṣaṇaṃ kṛṣṭaṃ] C6,
-\item[dhanurākarṣaṇaṃ kṛuṣṭaṃ] G4,
-\item[dhanurākarṣaṇaṃ kuryād] A1, B1, B3, Ba5, Bo1, Bo3, C2, C3, C9, J7, J10, J12, J13, J16, J17, Jyo, LD1, LD2, N13, N17, P1, P2, P4, P5, P7, P9, P12, P17, V16, V18, 
-\item[dhanurākarṣaṇāt kaṣṭaṃ] J8, P11, V3, 
-\item[dhanurākarṣaṇāt kṛṣṭaṃ] L2, N2, N23, P11, V14, V17, 
-\item[jānūrākarṣaṇa kṛuṣiḥ] Ba4, 
-\item[(unavailable/illegible)] A1, C5, C7, K2, K3, N3, N14, V9, P10, 
-\end{description}
- \end{marma}
-
-\begin{marma}[hp01_027]
-
-  \sthana{27d}
-
-  \begin{description}
-\item[jaṭharapradīptaṃ]    V6 (jaraṭha), B1, B2, B3, Bo1, J13, N1, V22, N13, N21, P2, P12, P13, P14, P17, R3, R4, 
-\item[jaṭharapradīpti]    Ba5, V10
-\item[jaṭharapradiptaṃ]    P4,
-\item[jaṭharaprakṛṣtaṃ]    V14
-\item[jaṭharapradīpaṃ]    P6,
-\item[jaṭharapradīptiṃ]    Bo3, Ko, N4a, P18, P22, Tue, R5
-\item[jaṭharāgnidīptiṃ]    P22a,
-\item[jaṭharapracaṇḍaṃ]    K1, V1, N22, V15, G1
-\item[jaṭharaṃ pracaṇḍaṃ]    J11, P16, 
-\item[jaṭharapracaṃḍa]    J5, P11, P27, 
-\item[matsyendra-vīra-jaṭhara-pracaṇḍaṃ]          J4, N19, V1 [vīra-ja is unmetrical?]
-\item[budhaṃ pracaṇḍa]    V8
-\item[jaṭharapravṛdhaḥ]    P21, P23, 
-\item[jaṭharapravṛddha]    J10, J17, V2, N8, N11
-\item[jaṭharapravṛddhau]    J15, P20, 
-\item[jaṭharapravṛddhaṃ]    N24
-\item[jaṭharapravṛddhiṃ]    C1
-\item[jaṭharaprabuddhaṃ]    C8, J14, N2, N12, V21, N23, V12,
-\item[jaṭharaprabuddha]    Ba, P25, 
-\item[jaṭharaprabuddhau]    C4 ("jvalanapradīptam" is also given above), L1, V19, V28, N5, N25, P3
-\item[jaṭharaprabodhaṃ]    A1, C6, P7, P26,
-\item[...bodhaṃ]       R2,
-\item[jaṭharāprabodhaṃ]    V20
-\item[jaṭharaṃ prabadhaṃ]    J2
-\item[jaṭhara pravaddhaṃ] N2, 
-\item[jvalanapradīptaṃ]    LD1, Bo2, C3, C9, J1, J3, J6, J8, N17, V3, V4, N6,, N9, N18, V18, N26, P1, P5, V17, P8, P24, V13, V16, P9
-\item[jvalanaṃ pradīptaṃ]    J7, J12, LD2, N10, N16, N20, P10, 
-\item[t(?)valanaṃ pradiptaṃ]    J16,
-\item[jvalatpradīptaṃ]    N7,
-\item[vīrapracaṃḍaṃ] P15,
-\item[(unavailable/illegible)]    C5, C7, M1, N14, G11, T, Ba3
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_027]
-
-   \sthana{27d}
-
-    \begin{description}
-    \item[candrasthiratvaṃ ca dadāti] Ba5, J7, J10ac, J17, LD1, LD2, V6, B2, Bo1, Bo3, N4a, V4, V22, N10, N13, P1, Tue, P8, P22, R5, V13, V10
-      \item[candraṃ sthiratvaṃ ca dadāti] P9("ca" in margin)
-    \item[candre sthiratvaṃ ca dadāti] C9,
-    \item[candrasthiratvaṃ hi dadāti] B1, C2, N1, R3
-    \item[cādrasthiratvaṃ hi dadāti] B3
-    \item[candrasthiratvaṃ pradadāti] C1, N24
-    \item[caṇḍasthiratvaṃ ca(?) dadāti] N7,
-    \item[daṇḍasthiratvaṃ ca dadāti] A1, Ba, Bo2, C3, C6, C8, J1, J2, J8, J10pc, J11, J12, J14, J15, J16, K1, Ko, N2, M1, N17, V1, V2, V3, N6, N9, N12, P10, V20, N16, N18, N19, N20, N22, N26, P6, P18, P25, V17, V12, R2
-    \item[daṇḍāsthiratvaṃ ca dadāti] P5
-    \item[daṇḍasthira.ntya karoti] G1
-    \item[daṇḍāsthitatvaṃ ca dadāmi] V14
-    \item[daṇḍaṃ sthiratvaṃ ca dadāti] P7, P11, 
-    \item[daṇḍasthiratvaṃ hi dadāti] J13, P2
-    \item[daṇḍasthiratvaṃ hi dadātti] P4
-    \item[daṃḍaschiratvaṃ hi dadāti] P12, 
-    \item[daṇḍasthiratāṃ ca dadāti] V21
-    \item[daṃḍaschiratvaṃ dadāti] J5,
-    \item[daṇḍasthiratāṃ sa dadāti] J3, J6,
-    \item[daṇḍasthiratvaṃ vidadhāti] C4, L1, V28, V18, N25, V16
-    \item[daṇḍasthiratvāṃ vidadhāti] P3
-    \item[daṇḍasthiratvaṃ pradadāti] N8, N11, V19, N23
-    \item[daṇḍaṃ sthiratvaṃ pradadāti] P23, 
-    \item[daṃḍe sthiratvaṃ pradadāti] N5
-    \item[daṇḍasthiratvaṃ jvalanaṃ] V8
-    \item[(daṇḍa?)sthirasvaṃ ce dadāti] N2
-    \item[dehasthiratvaṃ ca dadāti] N21
-    \item[kāyasthiratvaṃ ca dadāti] V15
-    \item[kaḍasthiratvaṃ ca dadāti(?)] J4
-    \item[vīryasthiratvaṃ ca dadāti] P22a,
-    \item[pracaṃḍahrunmaṃḍalakhaḍanāstraṃ] P21, 
-    \item[pracaṃḍarugmaṃḍalakhaṃḍanāstraṃ] P13, P14, P17, P20, P24, P26, R4, 
-    \item[rugmaṃḍalākhaṃḍanakhaṃḍanāstraṃ] P27,
-    \item[picaṃḍarukmāṃḍanakhaṃḍanāstraṃ] P15,
-    \item[vicaṃḍayugmaṃḍalakhaṃḍanāstraṃ] P16,
-    \item[(unavailable/illegible)] C5, C7, N14, T, Ba3
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_028]
-\sthana{1.28b}
-   \begin{description}
-   \item[dorbhyāṃ padāgradvitayaṃ] A1, Bo1, C2, C6, C8, J4, V1, M1, B2, B3, N1, N4a, P11, P14, P18, P20, V21, V20, V22, G1, V28, N16, N19, N21, N22, N23, N24, P2, P6, P12, P13, P17, P26, R4, R5, V13, R2
-   \item[dorbhyāṃ padāgredvitayaṃ] V14
-   \item[dorbhyāṃ padāgradvītayaṃ] P7
-   \item[dorbhyāṃ padāgnadvittayaṃ] P4 (could be "padāgra")
-   \item[dorbhyāṃ padāgraṃ dvitīyaṃ] Ko, V15
-   \item[dorbhyāṃ padāgradvitīyaṃ] R3, P25, P27, 
-   \item[dorbhyāṃ padāgraṃthidvayaṃ] N2,
-\item[dorbhyāṃ padagradvitayaṃ] Bo3, Ba5, K1, N13, P14, P22, Tue, V10
-\item[dorbhyāṃ padāgryau dvitayaṃ] J5
-\item[dorbhyāṃ pādāgrādvitayaṃ] B1
-\item[dorbhyāṃ ca pādadvitayaṃ] C4, L1, V19, N5, N7, N8, N11, N12, P3, P16, P23, 
-\item[dorbhyāṃ ca pādaṃ dvitayaṃ] C1
-\item[dorbhyāṃ ca pādau dvitayaṃ] N25
-\item[dorbhyāṃ ca pādāgradvitayaṃ] J13
-\item[dorbhyāṃ sa pādāgradvayaṃ] J2
-\item[dvābhyāṃ karābhyāṃ dvitayaṃ] Bo2, C3, C9, J1, J3, J6, J7, J8, J10, J15, J16, J17, LD1, LD2, N17, V6, V4, V3, N9, N10, N18, N20, V18, N26, P1, P5, P10, P24, V17, V16, LD1, P9("karābhyāṃ" in margin)
-\item[dorbhyāṃ ca pādadvayakaṃ] P21, 
-\item[dvābhyāṃ karābhyāṃ dvitīyaṃ] N6, 
-\item[dvābhyāṃ karābhyāṃ hitayaṃ] J12, V12
-\item[dvābhyāṃ padāgraṃ dvitayaṃ] J11
-\item[dvābhyāṃ pādāgrādvitīyaṃ] P8
-\item[(unavailable/illegible)] Ba, C7, G11
-\item[dorbhyāṃ padagraṃthidvayaṃ] J14, N2
-\item[.. rbhyāṃ pādaṃbhyāṃ dvitayaṃ] V8
-\item[(unavailable/illegible)] C5, N14, T, Ba3, 
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_035]
- \sthana{35b}
-   \begin{description}
-\item[asyahṛdaye dhṛtvā]    J7, J12, A1, V4, V18, V16, P8
-\item[asya hṛdayaṃ dhṛtvā]   P9
-\item[āsyahṛdaye dhṛtvā]    N10, V1, P1, P6
-\item[āsyahridayaṃ dhṛtvā]    J16
-\item[hṛdayaṃ kṛtvā] G1
-\item[eva hṛdaye dhṛtvā] C4, C7, C8, J5, J10, J11, J15, J17, Ko, V19, N2, N5, V21, N25, P3, P20, P23, P24, P25, V15, V14, R8, V12
-\item[ava hṛdaye dhṛtvā] P11, 
-\item[eva hṛdaye dhṛtvā]    P21,
-\item[eva hṛdaye kṛtvā]    Bo3, K1, N4a, V22, N13, N19, N24, P14, P27, Tue, V10
-\item[eva niyataṃ dhṛtvā]    C1, C3, J14, R4, V8, V28
-\item[eva niyataṃ kṛtvā] B1, B3, Bo2, C2, C6, J1, J3, J6("sudṛdha" is also given above), J8, J13, L1, N17, LD1, N1, V3, N12, V20, N16, N18, N20, N21, N26, P2, P4, P7, P12, P13, P17, P18, P26, R3,
-  R5('hṛdaye' canceled and replaced by 'niyataṃ')
-\item[ekahṛdaye dhṛtvā]    J4, N3, N23, P15, G11
-\item[ekahṛdayo dhṛtvā]    J2, M1
-\item[eksak ahṛdaye dhṛtvā]    P10, 
-\item[eva sudṛḍhaṃ kṛtvā]    N8, N11
-\item[eva dṛḍhayan dhṛtvā]    P16,
-\item[vāmaniyataṃ dhṛtvā]    V6
-\item[samyak(ū?) hṛdaye dhṛtvā]    N9    
-\item[āsyahṛdaye kṛtvā]    N22, V17
-\item[asya hṛdayeḥ dhṛtvā]    C9,
-\item[meḍhre pādam atho hṛdi svacibukaṃ]    N7
-\item[meṃḍhraṃ pādam athaikam]            Bo1, LD1, LD2
-\item[meḍhre pādam athaikm eva hṛdaye]    B2, Ba5,
-\item[meṃḍhreṃ pādam athaikam]            LD2
-\item[meḍhraṃ pādam athaika samyak meva hṛdaye dhṛtvā]           P5
-\item[(unavailable/illegible)]    Ba, C5, V13, N6, N14, P22, R2, T, Ba3
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_037]
- \sthana{Before 1.37}
-   \begin{description}
-\item[pūrvoktam eva matsyāsanaṃ] N12,
-\item[pūrvoktam eva saṃmataṃ] P11,
-\item[pūrvoktam eva matsaṃmataṃ] J14, Ko,
-\item[pūrvoktam eva matsāsanaṃ] N24,
-\item[pūrvoktam evāsanaṃ mataṃ] J2,
-\item[pūrvoktam eva matsyendramataṃ] B2, Ba5, C6, J5, J11, N23,
-\item[dvāvektimeva matseṃdramataṃ] C8, 
-\item[(unavailable/illegible)]  A1, B1, B3, Ba, Bo1, Bo2, Bo2, C2, C3, C4, C5, C9, J1, J3, J4, J6, J7, J8, J12, J13, J15, J16, J17, K1, K2, LD1, LD2, N13, N14, N16, N17, N18, N20, N21, N22, N25, P1, P2, P3, P4, P5, P6, P8, P9, P10, P11, P12, P13,
-\end{description}
- \end{marma}
-
- \begin{marma}[hp01_038]
- \sthana{38a}
-   \begin{description}
-   \item[yameṣv eva mitāhāra(ḥ)] C2, J10, J17, M1, V1, V6, P8
-   \item[yameṣv eva sitāhāra] V16
-   \item[yameṣv eva mitāhāro] N7,
-   \item[yameṣv eva mitāhāram] C4, B2, J14, N2, N5
-   \item[yameṣv eva ātmacetana] V28
-   \item[yameṣv iva mitāhāra(ḥ)] B1, B3, Bo1, Bo2, C3, C9, J3, J6, J8, J13, J15, K1, Ko, LD1, N17, N1, V3, V4, N9, N10, N14, N16, N18, N21, N26, P1, P2, P4, P12, P17, R3, V12, LD1, P9
-   \item[jameṣṭiva mitāhāram] J5
-   \item[yameṣv iva mitāhāro] P13, R4, V18
-   \item[yameṣv iva mitāhāre] P5, P24, 
-   \item[yameṣv iva mītāhāra] J16,
-   \item[yameṣv iva mitāhāram] A1, Bo3, Ba5, C1, C7, C8, J4, J11, L1, LD2, V19, N4a, V22, V21, N13, N23, G11, N24, N25, P3, P7, P11, P14, P15, P16, P18, P20, P21, P23, P25, P26, P27, Tue, V15, R5, V13, V10
-   \item[yamiṣviva mitāhāram] V14
-   \item[yameṣv iva mitāhā(ḥ)] V8
-   \item[yameṣv iva mityāhāraḥ] P10, 
-   \item[niyameṣu mitāhāro] N8, N11, N12
-   \item[niyameṣu mitāhārā] J12,
-   \item[(unavailable/illegible)] Ba, V20, R2
-   \item[yameśiva mitāhāram] N19
-   \item[yameṣv eva mitāhāra] N20, V17
-   \item[yameṣu ca nirāhārī] N22, P6
-   \item[yameṣu ca mitāhārm] C6,
-   \item[yamaṣviva mitāharm] J1,
-   \item[niyameṣu mitāhari] J2
-   \item[yameṣv api mitahāraḥ] J7("yamesṣv iva" is deleted)
-   \item[(unavailable/illegible)] C5, N6, T, Ba3, G1, P22,
-    \end{description}
- \end{marma}
-
-\sthana{1.38b}
-   \begin{description}
-\item[ahiṃ niyameṣv iva]    B1
-\item[ahiṃsā niyameṣu ca]    J4, M1, V1, N22, P6
-\item[ahiṃsā niyameṣv api]    B2, J14, N2, N24, V14
-\item[ahiṃsā niyameṣv iva] A1, B3, Ba5, Bo1, Bo2, Bo3, C2, C3, C4, C7, C8, C9, J1, J3, J5, J6, J7, J8, G11, J10, J11, J13, J15, J17, Ko, LD1, LD2, N17, V3, V4, N5, N9, V21, N13, N14, N16, N19, N20, N21, V18, N25, N26, P1, P2, P4, P7, P10, P11, P12, P13, P15, P16, P17, P18, P23, P24, P25, P26, V16, P8, R3, R5, V13, V12, LD1, P9, P14,
-\item[ahiṃsā niyameṣv ivaṃ] P21,
-\item[ahiṃsā niyāmaṣv iva]    P5
-\item[ahiṃsā niyameṣu eva]    V17
-\item[āhiṃsā niyameṣv iva]    Ko,
-\item[ahiṃsā nīyame sthitā]    K1,
-\item[ahiṃsāṃ nīyame sthitā]    P27, 
-\item[ahisā nīyameṣv iva]    J16,
-\item[aṃhiṃsā niyamaṣv iva]    J12,
-\item[svāhiṃsā niyameṣv iva]    V6
-\item[ahiṃsāṃ niyameṣv iva] L1, V19, N4a, V22, P3, P20, Tue, V15, V10
-\item[ahiṃsāṃ niyameṣu ca]    C6,
-\item[ahiṃsā niyaṃmai(?)ṣv iva]     N1
-\item[(a)hiṃsā ca niyameṣv iva]    N7,
-\item[yathāhiṃsā yameṣv iva]    N8, N11
-\item[ahiṃsā niyameṣv iha]    N10
-\item[asiṃsā niyame sthata(ḥ)]           V8
-\item[yameṣv evātma-cintanaṃ]    N12
-\item[hiṃsā ca niyameṣv iva] R4, 
-\item[na hiṃsāṃ niyameṣv iva]    N23
-\item[kṣamasye ātmacintane]    J2
-\item[(unavailable/illegible)]    Ba, C5, N6, V28, V20, N18, R2, T, G1, Ba3, P22, P25, 
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_039]
-\sthana{39d}       
-       \begin{description}
- (ṇ/n and ś/s irrelevant)
-\item[suṣumṇā iva nāḍikā]    V1, N22, P6
-\item[suṣumṇā iva nāḍiṣu]     B2, J4, J14, K1, Ko, N2, V8, N19, V12, G11
-\item[suṣumṇā caiva nāḍiṣu]     P26, 
-\item[suṣumṇā ca nāḍiṣu]    J5,
-\item[suṣumṇā eva nāḍiṣu ]    V20
-\item[suṣumṇā iva nāḍiṣuḥ] N8, P15, 
-\item[suṣumiva nāḍiṣu]     V28
-\item[suṣumṇām eva nāḍiṣu]    J10ac, J17, N24, V14
-\item[suṣumṇām iva nāḍiṣu ]    A1, C1, C4, C6, C7, C8, J1, J2, J3, J6, J11, J15, L1, M1, V19, N5, N7, N12, N16, N23, N25, P11, P16, P20, P21, P23, P25, P27, V15, V13
-\item[suṣuṇmām iva nāḍiṣu]    P7
-\item[suṣumṇāśva nāḍiṣu]    N11,
-\item[suṣumṇām iva ḍinoṣu]    P3
-\item[nāḍīnāṃ malaśodhanam]    B1, B3, Ba5, Bo1, Bo3, C2, C3, C9, J8, J12, J13, LD2, N1, N4a, N10, N13, N18, N20, N21, N26, V3, V22, P1, P2, P4, P12, P13, P14, P17, P18, Tue, V17, R3, R5, V10
-\item[nāḍiṣu malaśodhanam]  R4,
-\item[nāḍināṃ malasodhanaṃ]    J16,
-\item[nāḍīnāṃ malaśodhanā]   P9
-\item[nāḍīnāṃ malaśodhane]    J7, J10pc, LD1, N9, N17, V6, V21, N14(seems "dī"), V18, P5, P10, V16, P8, P24, V12, LD1
-\item[suṣumṇādiṣu śodhanam ]     V4
-\item(unavailable/illegible)]    Ba, Bo2, C5, N6, R2, T, G1, Ba3, P22, 
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_040]
- \sthana{1.40e}
-   \begin{description}
-\item[śramādau bahubhiḥ pīṭhaiḥ] V1
-\item[śramadair bahubhiḥ pīṭhaiḥ] B1, B3, Bo1, Bo2, C1, C2, C6, C8, J3, J6, J13, J14, K1, M1, G11, N1, N11, N12, N16, N21, N22, N23, N24, P2, P4, P7, P11, P12, P13, P16, P17, P21, P23, P25, P27, V14, R3, V13
-\item[śramadair bahubhiḥ pīṭhai]    P6,
-\item[śramadai bahubhiḥ pīṭhaiḥ]    Ba, C9mg, J5,
-\item[śramadair bahubhi pīṭhai]    N8
-\item[śramādaiḥ bahubhi pīṭhaiḥ]    V21
-\item[śramadairghyādibhiḥ pīṭhaiḥ]    C4, C7, L1, V19, N5, N25, P3, P20, 
-\item[śramadairdhyādibhiḥ pīṭhaiḥ]    V28
-\item[śamardda bahubhiḥ pīṭhaiḥ]    J1,
-\item[sanastai bahubhyaḥ pīṭai]    V8
-\item[(samastair?) bahubhiḥ pīṭhaiḥ]    N2
-\item[samastair bahubhiḥ pīṭhaiḥ]    J2, J4, J11, Ko, N2, P15, 
-\item[samasta bahubhiḥ pīṭhai]    V15
-\item[kim ādyair bahubhiḥ pīṭhaiḥ] A1, C9, J7, J8, J10, J17, LD2, N17, P26, V6, V3, V4, N7, N14, V20,
-  N18(No visarga at the end), N20, V18, N26, P5, P10, V17, P8, V12, LD1, R2, P9
-\item[kim ādyair bahubhiḥ pīṭheḥ]    N9, V16
-\item[kim adyair bahubhiḥ pīṭhaiḥ]    N10,
-\item[kim ādyair bahubhiḥ piṭhaiḥ]    J15, LD1, P1, P24, 
-\item[kim anyair bahubhiḥ pīṭhaiḥ]    Ba5, C3, N4a, N13, P14, P18, Tue, R5, V10
-\item[kim anyair bahuniḥ pīṭhaḥ]    Bo3, V22
-\item[kim arthai bahubhiḥ pīṭhai]    J12, J16("adyair" is also given),
-\item[kiṃ śramair bahubhiḥ pīṭhaiḥ]    B2
-\item[pramadair bahubhiḥ pīṭheḥ]    N19
-\item[tathaikasmin eva dṛḍham]    R4, 
-\item[(unavailable/illegible)] C5, N6, T, G1, Ba3, P22,
-    \end{description}
-
- \sthana{1.40f}
-   \begin{description}
-\item[sadā siddhāsanasthiteḥ]    A1, V20, V12, R2
-\item[sadā siddhāsanasthiteḥ]    P26, 
-\item[sadā siddhāsane sati] B1, B3, Bo1, C2, C9, J7, J8, J10, J11, J12, J15, J17, LD1, LD2, N17, V1, V6, N1, V4, N7, N9, N10, G11, N20, N22, V18, N26, P1, P2, P4, P5, P6, P8, P9, P10, P12, P24, V17, V16, LD1,
-\item[sidhe siddhāsane sati]    V10
-\item[siddhe siddhāsane sati] P18,
-\item[sadā siddhāsane satī]    J16,
-\item[sadā siddhāsane satya]              V3
-\item[sadā siddhāsanaṃ(?)]    V8
-\item[sadā siddhāsanā nyāsāt]    V22
-\item[sadā siddhāsanābhyām sādyogī]    V14
-\item[yadā siddhāsane sati]    J2, J4, N2, P15,  
-\item[yadā siṃhāsane sati]    Ko,
-\item[siddhe siddhāsane sati]    Ba5, Bo3, C3, N4a, N13, P14, Tue
-\item[siddhe siddhāsane miti]    R5,
-\item[kiṃ sadā siddhasādhanaiḥ]    M1, P11, 
-\item[kiṃ syāt siddhāsane sati] Bo2, C1, C4, C7, J1, J3, J5, J6, J13, L1, P13, P16, P20, V19, B2, V28, N5, N16, N21, N25, P3, P17, P21, P23, V15, R3
-\item[kiṃ syāt siddhāsane mati(ṃ?)]    N8
-\item[kiṃ syāt siddhāsane sthite]    C6, C8, P25, V21
-\item[kiṃ syā siddhāsane sthite ]    P7
-\item[kiṃ syāt siddhāsana sthite ]    N23
-\item[kiṃ syāt siddhāsane schite]    Ba,
-\item[kiṃ vā siddhāsane sati]    J14, N12, N24
-\item[kiṃ syāt siddhāsanai(?)ti]    N18
-\item[kiṃ sadā sevanena tu]     K1, N19(1.40h is "baddhe siddhāsane sadā")
-\item[baṃdhe siddhāsane sadā]  P27,
-\item[baddhe siddhāsane sadā]  R4,
-\item[va??t siddhāsane sadā]  N11,
-\item[(unavailable/illegible/unable to read)]    C5, N6, N14, V13, T, G1, Ba3, P22,
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_046]
- \sthana{1.46cd}
-\item[(utta(m)bhya cibukaṃ ~)]
-\item[vakṣa sthāpayet ]    J8, J10, J17, LD2, V6, N7, P8, N18, V12, V3, P9("thā" unclear)
-\item[vakṣaḥ sthāpayet]     Bo1, B1, B2, C3, J1, J12, J15, J16(sthāpaye),  K1, LD1, V18, P1, P5, P10, P24, R4, V16, N10, N22, LD1
-\item[vakṣaḥ sthāpaye]     P26, 
-\item[vandā sthāpayet]    V17
-\item[vakṣa sthāpayat]     N17
-\item[vakṣe saṃsthāpayat(?)]    Ko,
-\item[vakṣe uthāpayet]    V4
-\item[vakṣa sthāpya?]    J7(A word is deleted there), C9, N5(sthāpya sevanaṃ)
-\item[vatsa sthāpayet ]    N20, N26
-\item[vakṣaṃ sthāpayet]     V1
-\item[vakṣasy āsthāpya]     C4, L1, M1, V19, P3
-\item[vakṣasy utthāya]    Bo2, C6, N16, N8, N11,
-item[vakṣasy stthāpya] G11
-\item[vakṣasy utthāpya] Ba5, Bo3, C1, J13, J14, N2, N13, N21, N23, N24, P13, Tue, N1, N4a, P14, P18, P23,  R5
-\item[vakṣasy usthāpya] P21,
-\item[vakṣasy utthāyo] P15,
-\item[vakṣasy utṭhāpyo]    J4, J11,
-\item[vakṣasy utṭhāgre]    J6
-\item[vakṣasy oschāpyo]    J5
-\item[vakṣasy osthāpyot]    P11, 
-\item[vakṣyasy utthāpyot patane...]    N19
-\item[vakṣyasy asthāpya]    N25
-\item[vakṣyasy āsthāpya]    P20,
-\item[vakṣyapyutchāpya]    Ba, 
-\item[vakṣyapyutthāpya]    P25, 
-\item[vakṣe sthāpayet]    N14, P2, P4, N9
-\item[vakṣe syu sthāpya pavanaṃ]     V10
-\item[vakṣye sthāpayet]    B3, C2, P17, 
-\item[vakṣye schāpayet]    P12, 
-\item[vakṣye pyutthāya]    C8
-\item[vakṣyesu uttānya]    V14
-\item[vikṣe sthāpayet]    R3
-\item[cakṣasya utthāpya]    V15
-\item[svakṣasy utthāpya]    P7
-\item[urghātopaticotanau karau kṛtvā tato dṛśaṃ?]    J2
-\item[dakṣasy utthāgret(?)]    J3
-\item[utkuṃcya cibukaṃ vakṣa sthāpayet ]       P6
-\item[sannāya cibukaṃ vakṣyasy utthāpya]      N2
-\item[yuktaṃ vakṣastha-cibukaṃ mākramet]    N12
-\item[(unavailable/illegible/unable to read)]    A1, C5, N6, P16, V13, R2, T, G1, Ba3, P22, P27, 
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp01_048]
- \sthana{1.48b}
-   \begin{description}
-\item[(gāḍhaṃ vakṣasi sannidhāya cibukaṃ ~)]
-\item[dhyānaṃ tataś cetasi]    J2, V1
-\item[dhyānaṃ tataś cepsitaṃ]    J15, C9
-\item[dhyānaṃ tataś cośitaṃ]    V6
-\item[dhyānaṃ tataś conmitaṃ]    P5, P24, 
-\item[dhyānaṃ tataḥ conmī]    N1(some syllables seem deleted, 'nmī' is added later)
-\item[dhyānaṃ ca tac cetasi] B1, B2, B3, Ba, Bo1, Bo2, C1, C2, C4, C6, C7, C8, J4, J5, J8, G11, J11, J12, J13, K1, Ko, L1, V19, N25, N26, P2, P3, P4, P6, P7, P12, P13, P15, P17, P26, V4(ṃ omitted), V17, V16, N3, N5, N8, N11, V15, V14, P8, P20, P25, R3, V13, N14, N16, N17,N19, V12, V3, N21, N23,  N24(saṃniyamya), R2, P9
-\item[dhyāyaṃś ca tat cetasi] P14,
-\item[dhyānaṃ ca tecesi]    V8
-\item[dhyānena tac cetasi] P21, P23, 
-\item[dhyānaṃ ca tac cepsitaṃ] A1, J1, J3, J6, J10, J14, J16, J17, LD1, P1, P10, N7, N9, N10, N12, N18, LD1, N20, N22, P27, 
-\item[dhyānaṃ ca yac cepsitaṃ]    C3,
-\item[dhyānaṃ ca tac cepsite]    J7,
-\item[dhyānaṃ ca tatrepsitaṃ]    LD2, N2,
-\item[dhyāyaṃś ca tac cetasi]    Ba5, Bo3, M1, Tue, N4a, N13, P18, V10, V22
-\item[dhyāyaṃś ca taṃ cepsitaṃ]    R5
-\item[(unavailable/illegible)]   C5, N6, T, G1, Ba3, P11, P16, P22, R4, 
-    \end{description}
-
-\sthana{1.48c}
-  \begin{description}
-\item[(vāraṃ vāram apānam ūrdhvam anilaṃ ~)] G11
-\item[protsālayan pūrayan]    V1
-\item[protsārayan pūrayan]    J7, V6, N7, N12
-\item[protsārayaḥ pūrayem]    V8
-\item[protsārayan pūritaṃ]    A1, Ba5, Bo3, C1, C6, C8, J10, P14, P15, P26, Tue, N4a, V14, R5, V13, N13, P18, V10, V22, R2, G11
-\item[protsārayat pūritaṃ]    P7, P25, 
-\item[protsyā?rayat pūritaṃ]    Ba,
-\item[protsārayan pūrayet]    C4("proccālayan pūritaṃ" is also given above),L1, N25, P3
-\item[protsārayet pūrayan]    P20, V19
-\item[protsārayet pūrayet]    C7
-\item[protsāraṇaṃ pūrayet]    N5, P21, P23, 
-\item[protthāya yat pūrayan]    B3
-\item[protthāpanṃ pūrayet]    Bo1,
-\item[prodhāyana pūrayet]    B1,
-\item[protthāpayan pūrayan]    C2, P2, P4, N1, R3, P9, P12, P17, 
-\item[procārayet pūritaṃ]    J8
-\item[prottāsayan pūritaṃ]    V12
-\item[prollāsayan pūritaṃ]    J10, J17, N17, P1, P5, P24, 
-\item[prollāśayet pūritaṃ]    J15, N9
-\item[prolāsayet pūrayan]    J12,
-\item[prollāsayet pūrayan]    N10
-\item[prollosayana pūrayan]   LD2
-\item[prollāsayet pūrayet]    J16,
-\item[prollāsayan pūrayan]    C9, V16, P8, V4
-\item[pro(llāpa?)yan pūrayan] LD1
-\item[prollāsat pūritaṃ]    C3,
-\item[prollāghayan pūrayan]    LD1
-\item[procāraranu pūritaṃ]    J5
-\item[proccārayan pūritaṃ]    M1, V17, N3(vālaṃ vāram)
-\item[proccārayan pūrayan]    J4,
-\item[proccārayat puraya yataṃ(?)]    J1,
-\item[proccārayat pūritaṃ]    N2, P27, 
-\item[proccārayet pūritaṃ]    V15, V3, N23
-\item[proccārayet pūrayet]    N19
-\item[proccālayan]    K1
-\item[proccālayan pūrayen]    B2
-\item[proccālayan pūrayet]    J13,
-\item[proccālayaṃn pūrayan]    P6, N22(vvāraṃ cāram apānamūlapranilaṃ), \item[proccālayan pūrayan]    N18, N24(ayāmūrdh...)
-\item[proccālayan pūritaṃ]    J2, J14, Ko, N26, N20(proccālayaṃn)
-\item[proccālayet pūritaṃ]    Bo2, J3, J6("proccāṭayan pūrayet" is also given below), N16
-\item[proccālitaṃ pūritaṃ]    N14(vālaṃ vālam)
-\item[….... śayat pūritaṃ]    P10, 
-\item[proccānalaṃ pūrayen]   P13, N21
-\item[prodbhā(ṭa?)yan pūrayet]    N8
-\item[prod(ghā?)ṭayan pūrayet]    N6, N11
-\item[(unavailable/illegible)] T, G1, Ba3, P11, P16, P22, R4,
-
-\sthana{1.48d}
-  \begin{description}
-\item[(muñcan prāṇam upaiti bodham atulaṃ ~)]
-\item[śaktiprabhāvān naraḥ] B1, B3, Ba5, Bo2, Bo3, C1, C2, C6, C8, J1, J3, J6, J8, J11, J13, J14, K1, Ko, M1, V1, P2, P4, Tue, N2, N3, V15, V14, R3, R5, V13, N13, N16, N19, P12, P13, P14, P15, P17, P18, P25, V10, V22, N23, P27, G11
-\item[śaktiprabhāvān nara]    V3
-\item[śaktiprabhāvā naraḥ]    V8, N22(mucan...)
-\item[śaktiḥ prabhāvān naraḥ]    Ba,
-\item[śaktiṃ prabhāvān naraḥ]    Bo1,
-\item[śaktiṃ svabhāvān naraḥ]   N21
-\item[saktiprabhāvān]    J5
-\item[śaktiprabhāvon naraḥ]    N1, N7
-\item[śaktiprabhāvīn naraḥ]    P6
-\item[śaktiprabhāvaṃ naraḥ]    J12, N4a
-\item[śaktiprabhān naraḥ]    B2
-\item[śaktiprabhāvāṃtaraḥ]    N18
-\item[śaktiprabhāvodayāt] C4, C7, L1, V19, N25, P20, V16, N5, N14, P21, P23, P26, 
-\item[śaktiḥ prasādān naraḥ]    J2,
-\item[śaktimabhāvodayāt]    P3 ("ma" could well be "pra)
-\item[śaktiḥ prabhāvodayāt]    A1, R2("vo" illegible due to damage)
-\item[śaktiprabodhodayāt]    N20
-\item[śaktiprabhāvodayaḥ]    J7, P10, V6
-\item[śaktiprabhāvād ataḥ]    C3, C9, J10, J16, P5, P7, N9, N10, N11, N12, P8, V12, LD1, N24, P9
-\item[śaktīprabhāvad ataḥ]    LD2, P24, 
-\item[śaktiprabhāvad ata]    J15, N8, V4
-\item[śaktiprabhāvataḥ]    J17
-\item[śaktiprabhāvadhataḥ]    V17
-\item[śaktiprabodhodayāt]    N17, N26
-\item[śaktiprabhāvodhayaḥ]    P1
-\item[bhakti prabhāvodayaḥ]    J4,
-\item[(unavailable/illegible)]   C5, N6, T, G1, Ba3, P11, P16, P22, R4,
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_049]
-\sthana{1.49c}
-  \begin{description}
-\item[mārutaṃ nayate yas tu]    C6, C8, P11, 
-\item[mārutaṃ niyato yas tu]    Ba, P25, 
-\item[mārutaṃ niyatam yaskta]    V14
-\item[mārutaṃ niyatam yas tu]    V8, N23(mār(ū?)taṃ)
-\item[mārutaṃ mriyate yas tu] G11
-\item[mārutaṃ nīyate yas tu]    C1,
-\item[mārutaṃ pīyate yas tu]    V1
-\item[mārutaṃ pīvyate yas tu]    J8ac, V3
-\item[mārutaṃ yaḥ pīvet nitya]    V13
-\item[mārutaṃ pibati yas tu]    J8pc
-\item[mārutaṃ pibate yas tu]    J11,
-\item[mārutaṃ yas tu pibati] A1, C9, J7, J10, J12, J15, J17, LD2, N17, V6, N26, P1, P5, P6, P7, V17, V16, N7, N9, N10, P8, N14, V4, LD1, N20, P24, R2, P9
-\item[mārutaṃ yas tu pibatī] P21, 
-\item[māruto bhidyate yasya]    N2, 
-\item[māruto yas tu pibati]    N22
-\item[mārutaṃ yaska pibati]    V12
-\item[māruta yas tu pibati]     R1
-\item[mārutaṃ yas tu pibaṃti]    C3,
-\item[mārutaṃ yas tu pībaṃti]    J16,
-\item[mārutaṃ yas tu vidhivat]    P26,
-\item[mārutaṃ dhārayet yas tu] Ba5, Bo2, Bo3, C2, C7, J1, J3, J6, J13, L1, M1, V19, N25, P2, P3, P4, P14, P18, P20, P23, Tue, N4a, N11, N12, P12, P13, R5("), N13, N16, V10, N24,
-\item[māruto dhārayed yas tu]    N1, P17, R3, V22
-\item[māruta dhāraye?as tu]    N21
-\item[mārutaṃ dhāraye yas tu]    N8, N18
-\item[marutaṃ dhārayed yas tu]    C4, N5, V15
-\item[mārutaṃ mārayed yas tu]    J5
-\item[mārutaṃ mriyate yas tu]    N3
-\item[mārutaṃ mriyate yasmāt]    Ko,
-\item[māruto mriyate yas tu]    J4, P15, 
-\item[māruto kṣiyate yasya] P27,
-\item[māruto bhidyate yasya]    B1, B2, Bo1, J14, K1, N2, N19
-\item[mahato dhārayed yastu]    B3
-\item[pavanṃ prāsayeta saṃstu]    J2
-\item[(unavailable/illegible)]    C5, N6, T, G1, Ba3, P10, P16, P22, R4, 
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_051]
-\sthana{1.51d}
-  \begin{description}
-\item[nāsāgre nyastalocanaḥ]    B1, B3, C2, C9, J7, J13, J16, N17, V1, V6, P2, P4, P17, N1, N7, P12, R3, N22
-\item[nāsāgre nyastalocanā]    P6, R4, 
-\item[nāsāgra-nyastalocanaḥ]    A1, Bo1, C3, J10, J17, N26, P1, P5, V16, N10, V12, V4, LD1, LD2, N20, P24, R2
-\item[nāsāgraṃ nyastalocanaḥ]  P9
-\item[nāsāgranyastalocanaṃ]    J8
-\item[nāsāgra vyasalocanaḥ]    P26, 
-\item[nāsāgranyastaloca....]    P10,
-\item[nāsāgraṃ nyastalocanaṃ]    P7, V17, V3
-\item[nāsāgra susamāhitaḥ]    C8,
-\item[nāsāgraṃ susamāhitaḥ] Bo2, Bo3, C1, C4("nāsāgra nyastalocanaḥ" is also given), J1, J2, J3, J6, J11, J14, J15, K1, Ko, L1, M1, V19, N25, P3, Tue, N2, N4a, N5, N12, V15, V14, V13, N13, N16, P13, P14, P15, P18, P23, V10, V8, G11
-\item[nāsāgrastu samāhitaḥ]   N23
-\item[nāśāgre susamāhitaḥ]    N19,
-\item[nāsāgre susamāhitaḥ]    P27,
-\item[nāsyagraṃ tu samāhitaḥ]    J5,
-\item[nāsāgre susamāhitaṃ]    P21, 
-\item[nāsāgraṃ susamāhitaṃ]    Ba5, C6, V22
-\item[nāsāgryaṃ susamāhitaṃ]    N18
-\item[nāsāgraṃ susamāhita]    N8
-\item[nāsāṃgraṃ susamāhitaḥ]    N11
-\item[nāśāgraṃ tu samāhitaḥ]    N3,
-\item[nāsāgraṃ tu samāhitaḥ]   N24, P20, 
-\item[nāsāgre susamāhitaḥ]    B2, Ba, K1, N21, P25, 
-\item[nāsāyaṃ susamāhitaḥ]    J4, P11,
-\item[nāsāyāṃ susamāhitaḥ]    J4,
-\item[nāsāṃ tu samāhitaḥ]    R5
-\item[nāsāgre neṣu nocana]    N14
-\item[(unavailable/illegible)]    C5, C7, N6, T, G1, Ba3, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_052]
-\sthana{1.52d}
-  \begin{description}
-  \item[kurute cāsanottamam] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J8, J10pc, J13, J15, L1, N17, V6, N26, P2, P3, P4, P5, P7, Tue, N1, V17, N4a, N5, N8, N11, N12, P12,  P13, P14, P17, P18, P20, P21, P23, P24, V15, R3, R5, V13, N13, N14, N18, N19, V8, V3, N20, N21, N23, P25, R2, R4, G11
-\item[kuruta cāsanottamam]    V22
-\item[kurute cāsanomatam]    V10
-\item[kurute cāsanottame]    B1, N16
-\item[kurute cāsanottamaḥ]    J2,
-\item[kurute cāsanotraye]    V14
-\item[kurute cāsanottama]    A1, P26, 
-\item[kurute cāsanontayam]   N24
-\item[kurute vāsanottamam]    V1, V19, N3, P11, 
-\item[kurute sādhanottamam]    C3, J12, J10ac, J16, J17, P1, V16, N9, N10, P9, P10, V12, V4,
-\item[kurute sādhanottaramaṃ]    P8(perhaps 'ra' is attempted to be canceled)
-\item[kurute sādhakottamam]    C9, N7,
-\item[kurute sādhakottamaḥ]    LD1, LD2
-\item[kurute cāsanam uttamaṃ]     N25
-\item[sevate cāsanottamaiḥ]    J14,
-\item[sevate cāsanottamaṃ]    J11, Ko, N2, P15,  
-\item[sevyate cāsanottamaṃ]    J4,
-\item[(unavailable/illegible)]    B2, C5, C7, K1, M1, N6, N22, P6, T, G1, Ba3, P16, P22, P27, 
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_054]
-\sthana{1.54b}
-  \begin{description}
-  \item[sarvavyādhi-viṣāpaham] C3, C9, J5, J7, J8, J10, J12, J15, J16, J17, LD1, V1pc, V6ac, P1, P5, P6, P7, P24, V17, N3, V16, N9, N10, P8, R1, V12, V3, V4, N22, P9, P11, G11
-  \item[sarvavyādhī-viṣāpaham] LD2
-  \item[sarvavyādhi-vināśanam] A1, B1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C7, C8, J1, J2, J3, J4, J6, J11, J13, K1, Ko, L1, N17, V1ac, V6pc, V19, N25, N26, P2, P3, P4, P17, P25, Tue, N1, N4a, N5, N7, N11, N12, P12, P13, P14, P15, P18, P20, V15, V14, R3, R5, N13, N14, N16, N18, N19, V10, V8, V22, N23, N24, P21, P23, P26, P27, R2, R4,
-\item[sarvavyāvināśanam]   N21
-\item[sarvadhyādhi-vighātakaṃ]   N20
-\item[sarvavyādhi-praṇāśanaṃ]    J14, N2, 
-  \item[sarvavyādhivikṣāpaham] P10, 
-\item[kaphavātāmayāpahaṃ]    M1
-\item[(unavailable/illegible)]    C5, N6, N8, V13, T, G1, Ba3, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_055]
-\sthana{1.55 additional verses}
-
-kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet |
-na śāstrapāṭhamātreṇa yogasiddhiḥ prajāyate ||
-na veṣadhāraṇaṃ siddheḥ kāraṇam na ca tatkathā |
-kriyaiva kāraṇaṃ siddheḥ satyam etan na saṃśayaḥ |
-śiśnodararatāyeha na deyo veṣadhāriṇaḥ ||
-
-\begin{description}
-\item[Yes] B1, B2, B3, Ba, Bo1, Bo2, C1, C3, C4, C6, C8, C9, J1, J3, J6, J7, J8, J10, J11, J12, J13, J14, J15, J16, J17, K1, L1, LD1, LD2, N5, N17, N25, P10, V3, V17, V25, P8, R1, R3, N14, N16, N18, V12, LD1, N7, N9, N10, N13, N20, N21, N26, P1, P3, P4, P5, P7, P14, P17, P21, R2, P9, P12, P24, P25, P26, P27,
-\item[Yes], but the 4th line appears as the 2nd line: C2, J13, N1, P2
-\item[No] A1, Ba5, Bo3, C1, C5, C7, J2, J4, J5, Ko, M1, V1, V6, V8, V19, R5, N19, P18, P20, V10, V22, V4 (unclear/unable to read) V13, N2, N3, N4a, N6, N8, N11, N12, N22, N23, N24, Tue, P6, P11, P13, P15, P16, P22, P23, R4, G11, T, G1, Ba3
-    \end{description}
-
-\sthana{1.55 third additional verse}
-
-mayi bodhāmbudhau svacche tuccho 'yaṃ viśvabudbudaḥ |
-pralīna udito veti vikalpapaṭalaḥ kutaḥ ||
-
- \begin{description}
- \item[Yes] B1, B3, Ba, Bo1, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, P12, P17, P20, P21, P23, P26, V6, V19, R3, N14, N16, N18, N1,
-
- \item[No] A1, B2, Ba5, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, LD1, LD2, M1, V1, V3, V8, V17, V25, P8, R1, R5, N19, V12, V10, V22, V4, LD1, V13, N2, N3, N4a, N6, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, P11, P18, Tue, P1, P5, P6, P7, P9, P10, P13, P14, P15, P16, P24, P27, R2, R4, G11, T, G1, Ba3, P22, P25,
-    \end{description}
-
-\sthana{1.55 fourth additional verse}
-% P: What is the difference between this and the previous extra verse?
-
-mayi bodhāmbudhau svacche tuccho 'yaṃ viśvabudbudaḥ |
-pralīna udito veti vikalpapaṭalaḥ kutaḥ ||
-
-\begin{description}
-\item[Yes] B1, B3, Ba, Bo2, C1, C2, C4, C7, C8, J1, J3, J6, J7, J13, L1, N5, N17, N25, V6, V19, R3, N14, N16, N18, N8, N11, P2, P3, P4, P12, P21, P23, 
-\item[No] A1, B2, Bo3, C3, C5, C6, C9, J2, J4, J8, J5, J10, J11, J12, J14, J15, J16, J17, K1, Ko, M1, V1, V3, V8, V17, V25, P8, R1, R5, N13, N19, V12, V10, V22, V4, LD1, V13, N2, N3, N4a, N6, N7, N9, N10, N12, N20, N21, N22, N23, N24, N26, Tue, P1, P5, P6, P7, R2, P9, P10, P11, P13, P14, P15, P16, P17, P18, P20, G11, T, G1, Ba3
-    \end{description}
-
-\begin{marma}[hp01_056]
-\sthana{1.56a}
-  \begin{description}
-  \item[āsanaṃ kumbhakaś citro] J8pc, N24
-\item[pīṭhāni kuṃbhakāś citro] G11
-\item[pīṭhādi kuṃbhakāś citram] R4, 
-  \item[āsanaṃ kumbhakaś citraṃ] B2, C3, J7, J8ac, J10, J12, J15, N17, P8, N14(ā(?)naṃ), V12, LD1, LD2, N4a, N10, N12, N20, P1, P5, P24, P27, 
-  \item[āsanaṃ kumbhavaś citra] J2,
-\item[āsanaṃ kumbhavaś ci(nt?)aṃ] P9
-  \item[āsana kumbhakaś citraṃ] V3
-\item[āsanaṃ kumbhakāś citrāḥ]  N8, N11
-  \item[āsanaṃ kumbhavaś cittaṃ] P10, 
-\item[āsanaṃ kumbhakāś citrā]   P7
-\item[āsanaṃ kumbhakācitraṃ]   Tue
-  \item[āsanaṃ kumbhakaṃś citraṃ] C9, J17
-  \item[āsanaṃ kumbhakaṃ citraṃ] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C4, C6, C7, C8, J1, J3, J4, J5, J6, J11, J13, J14, K1, Ko, L1, M1, P11, P12, P13, P18, V19, R3, R5, N13, N16, N19, P14, P17, P20, P21, P23, P25, V10, V22, V4, N1, N5, N23, N25, N26, P2, P3, P4, P6, P26, R2
-  \item[āsanaṃ kumbhakaṃ citra] P15,
-\item[āsanaṃ kumbhakaṃ śvitraṃ]   N21   
-  \item[āsanaṃ kumbhaka citrāṃ] J16, N7,
-  \item[āsanaṃ kumbhaka citaṃ] N18
-\item[āsanaṃ kubhakaṃ citraṃ] N2, 
-\item[asanaṃ kumbhakaṃ citraṃ] N3, N9
-  \item[(unclear/unable to read)] V13
-\item[(unavailable/illegible)]    C5, N6, N22, R1, T, G1, Ba3, P16, P22,
-    \end{description}
-
-\sthana{1.56b}
-  \begin{description}
-  \item[mudrādikaraṇāni ca] A1, B2, C3, C9, J7, J8pc, J10, J12, J15, J16, J17, LD1, N17, P8, N14, V12, N7, N9, N10, P1, P5, P7, P9, P10, P24, R2, R4, 
-  \item[mudrānikaraṇāni ca] LD2
-\item[mudrādikāraṇāni ca]    V4
-\item[vicitrīkaraṇāni ca]    J2,
-  \item[mudrādikaraṇakriyā] P27,
-\item[mudrādi pavanakriyā]    B1, B3, Bo1, C2, J5, J8ac, J13, R3, V3, N1, N20, P2, P4, P12, P17, 
-\item[mudrādi pavanakriyāḥ]   N26
-\item[mudrākhyaṃ karaṇaṃ tathā] Ba, Ba5, Bo2, Bo3, C1, C3, C6, C7,  C8, J1, J3, J11, L1, M1, P11, P14, P15, P18, P20, P23, P25, P26, V19, R5, N13, N16, N18, N19, V10, V22, N5, N11, N12, N21, N23, N25, Tue, P3, P6, P21,
-\item[mudrākhaṃ karaṇaṃ tathā]   N24
-\item[mudrākhyaṃ karaṇaṃ bhavet]   N3
-\item[mudrākhyakaraṇaṃ tathā]    Ko, N4a, N8, P13, 
-\item[mudrādi karaṇaṃ tathā]    K1,
-\item[mudrākṣaṃ karaṇaṃ tathā]    J4, J6,
-\item[mudrakhecaraṇīṃ tathā]    J14(given at 55)
-\item[mudrakhecariṇīṃ tathā]    N2, 
-\item[(unclear/unable to read)]    V13
-\item[(unavailable)]     C5, N6, N22, R1, G11, T, G1, Ba3, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_060]
-\sthana{1.60cd}
-  \begin{description}
-\item[(atilavaṇaṃ tilapinḍaṃ kadaśanaśākotkaṭaṃ varjyam |)]
-\item[atilavaṇaṃ tilapinḍaṃ]    J17, LD2, N10, R1
-\item[atilavaṇaṃ tilapinḍa]    C3, J8pc, J10, J15, J16, LD1, N17, V4, N9, N20, P1, P5, P8, P9
-\item[atilavaṇaṃ tilapiṇḍā?]   N14
-\item[atilavaṇatilapiṇḍakadaśanaṃ]  P10, P24, 
-\item[atilavaṇaṃ tilaṃ pinḍa]    J8ac
-\item[atilavaṇaṃ tilaṃ pinḍaṃ]     V3
-\item[atilavaṇaṃ tilapiṇyākadarśanam]    N26
-\item[atilavaṇam amlayuktaṃ]    A1, B3, Ba, Ba5, Bo3, V22, V10, N1, N4a, N11, N13, Tue, P2, P12, P14, P17, P18, P25, P26, R2, R5
-\item[atilavaṇama malayuktaṃ]    C2,
-\item[atilavaṇāmalpayuktaṃ]    P13, 
-\item[atilavaṇamalayuktaṃ]   P4, R3
-\item[atilavaṇāmlayuktaṃ]   P7
-\item[atilavaṇa amlayuktaṃ]    C6, C8, V13
-\item[atilavaṇa amlayutaṃ]    J11,
-\item[atilavatilapi]    V6
-\item[atilavaṇatiktapiṇḍa]    B2
-\item[atilavaṇatilapiṇḍakaṃ]    C9
-\item[atilavaṇaṃ dyuṣṇaṃ]    V1
-\item[atilavaṇam apramuṣṇaṃ]    M1
-\item[atilavaṇam ca apramitaṃ]    V8
-\item[atilavaṇam cāpramitaṃ]   P6, P27, 
-\item[atilavaṇaṃ ca tilapiḍaṃ]   N7,
-\item[atilavaṇam pramitaṃ]    J12, K1,
-\item[atilavaṇādi prasaktaṃ]    J14,
-\item[atilavaṇādi proktaṃ]   N22
-\item[atilavaṇacāmlapiktaṃ]    N19
-\item[atilavaṇakavalṃ cāpamitaṃ]    J1,
-\item[atilavaṇasavapalala]    V19
-\item[atilavaṇasavapalalaṃ]   N25
-\item[atilavaṇāsavasalilaṃ]   P21, 
-\item[atilavaṇaṃ sapalalaṃ]   P20,
-\item[atilavaṇaṃ sala(pa?)laṃ]   N8
-\item[atilavaṇaṃ palilaṃ]     R4
-\item[atilavaṇāsavapalalaṃ]  C1, C7, L1, N5
-\item[atilavaṇāsavapalaṃla]  P23, 
-\item[atilavaṇātiproktaṃ]    J2,
-\item[atilavaṇādiyutaṃ]    J4,
-\item[atilavaṇādi prayuktaṃ] P15,
-\item[atilavaṇādi puttaṃkdaśanaṃ]    P11,
-\item[atilavaṇādyaptataptaṃ]    J5,
-\item[atilavaṇādyapran(t?)aṃ]  N3
-\item[atilavaṇādyapratapta]   N12
-\item[atilavanā?āna(va?)lala]   N24
-\item[atilavaṇamatprayuktaṃ]   N21
-\item[atilavaṇāmavapalalaṃ]    C4 ("atilavaṇaṃ tilapinḍaṃ" is also given),
-\item[atilavaṇāya priyatavṃ]    Ko
-\item[atilavaṇaṃ sapalaṃ ca]    Bo2, J3, J6,
-\item[atilavaṇasalilaṃ kadaṣana]    B1, Bo1, J13,
-\item[atilavaṇetilapiṃḍakvadaśanaṃ]    J7
-\item[atilavaṇaprasūtaṃ]     N2, 
-\item[atilavaṇaṃ kavalaṃ cāpramitaṃ piṇḍa]   N18
-\item[atīlavaṇāmlapṛktaṃ]   N23
-\item[(illegible/unavailable)]  C5, N6, N16, P3, G11, T, G1, Ba3, P16, P22,
-
-* * *
-
-\item[śākotkaṭaṃ] A1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C3, C4, C6, C8, C9, J2, J3, J4, J6, J10, J11, J12, J13, J14, J15, J17, K1, Ko, L1, LD1, LD2, V6, V19, V22, V4, V10, V13, N2, N3, N4a, N7, N11, N13, N18, N19, N21, N24, N25, Tue, P2, P5, P6, P7, P8, P9, P10, P11, P12, P13, P14, P15, P17, P18, P20, P21, P23, P24, P26, R1, R2, R3, R4, R5,
-\item[śākotkaṭhaṃ]    V8
-\item[śākotkavaṃ]    J5,
-\item[śākotkaṭa]    Bo2, J8pc, J16, N1, N9, N26, P27,
-\item[śākoktaṃ]    P25, 
-\item[śākotkarṭeṃ]  P4
-\item[śākātkaṭa]    V3
-\item[śākotchadaṃ]    Ba,
-\item[śākolkaṭaṃ]   P1
-\item[śākokṣadaṃ]   N23
-\item[śāṃkotkaṭaṃ]   N10
-\item[śāketkaṭaṃ]    C7
-\item[śokātkaṭa]    J8ac, V1
-\item[śokotkaṭaṃ]    B1, J7,
-\item[śāśotka]    M1
-\item[śātkṭaṃ]    J1,
-\item[śa?tāsanaḥ]   N14
-\item[sākotkaṭa]   N20
-\item[sākotkaṭaṃ]  N22
-\item[syacotkaṭaṃ] N12
-\item[(illegible/unavailable)]    C5, N6, N17, N8, N16, P3, G11, T, G1, Ba3, P16, P22,
-\item[kotkaṭaṃ]  N5
-
-* * *
-\item[varjyaṃ] A1, B1, B2, B3, Ba5, Bo1, Bo3, C1, C2, C4, C7, C9, J3, J6, J13, J15, K1, L1, LD2, V3, V6, V19, V22, N1, N4a, N5, N11, N12(va?jyaṃ), N13, N22, N25, Tue, P2, P4, P6, P7, P12, P14, P17, P18, P20, P21, P23, P27, R3, R4, R5
-\item[varjya] V8, V10
-\item[varjyāṃ] J16
-\item[varj(j)aṃ] J7, J10, J12, J17, M1, N9, P10, P26, R1
-\item[varjjyaṃ]  P5, P24, 
-\item[varj(j)itaṃ] J1, Bo2, J8, LD1, V3, N18, N20, N26
-\item[varyyaṃ]   N7, R2, P9
-\item[varjjāṃ]   N19
-\item[varjet]  N10
-\item[varjayet] C3, P1("varjaṃ" earlier)
-\item[vayīṃ]   P8
-\item[duṣṭaṃ] C6, C8, J4, J5, J11, J14, Ko, V13, N2, N21, N23, N24, P13, P15, P25, 
-\item[du(ṣl?)aṃ]  N3
-\item[dṛṣṭaḥ]
-\item[durjvaṃ(?)] Ba,
-\item[(illegible/unavailable)] C5, N6, N17, V1, V4, N8, N14, N16, P3, G11, T, G1, Ba3, P11, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_062]
-\sthana{1.62d} 
-  \begin{description}
-  \item[mudgādi divyam udakaṃ ca] A1, B2(dīvyam), Ba, Ba5, Bo3, C6("dgā" is missing), C8, J5, J7, J10, J11, J12, J16, J17, K1, L1pc, LD1, LD2, N17, V1, V6, V8, V22, V4, V13, N1, N10, N12, N13, N14, N20, N22, N23, N24, N26, Tue, P1, P6, P8, P9, P14, P17, P18, P23, P25, P26, P27, R1, R2,
-\item[mudgādi divyam udakaṃ]   N18, P11,
-\item[mudgādi divyam udakaṃ ka]   N4a
-\item[mudgādi divyam udake ca]   R5
-\item[mudgādi divyam udakaṃ yamīndra]    V3, N2
-\item[mudgādi divyam udakaṃ hi]    C3, J14, J15, N9, P5, P10, P15, P24, 
-\item[mudgādi pivyam udakaṃ ca]    J2,
-\item[mudgājya divyam udakaṃ ca]    C9
-\item[mudgādivyam udakaṃ ca]    C4, J8, M1, V10, N3
-\item[mudgādi divyagaditaṃ ca]   N7,
-\item[mudgādi divyamuktaṃ]    Ko,
-\item[mudgādi cālpam udakaṃ ca]    B1, B3, Bo1, Bo2, C1, C2, C7, V19, N5, N11, N21, P2, P4, P12, P13, P20, P21, R3, R4
-\item[mudgādi cālyam udakaṃ ca]    L1ac, N8(no "ca")
-\item[mudgādi cājyam udakaṃ ca]    N16
-\item[mudgādi cālpa sudakaṃ ca]     J1,
-\item[mudgādi dālpam udakaṃ ca]    J3, J6("cālpa" is also given)] J13,
-\item[mudgādi dālpamudgaṃ ca]    N25
-\item[mudrādi divyam udakaṃ ca]  P7
-\item[mudrādi divyam udakaṃ ca mudakaṃ]    N19
-\item[(illegible/unavailable)]       C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22,
-    \end{description}
-
-\sthana{1.62d}
-  \begin{description}
-  \item[munīndrapathyaṃ (or: munindra-)] B1, B2, B3, Bo1, C1, C2, C4, C7, J3, J5, J6, J13, J16, J17, K1, L1, LD2, V1, V19, N1, N5, N11, N19, N21, N22, P2, P4, P12, P13, P17, P20, P21, P23, P26, R4
-\item[munīdrapathyaṃ]   N8, R3
-\item[munīndrapathaṃ]   N25
-\item[mahendrapathyaṃ]  N9, P5, P10, P24, 
-  \item[yamīndrapathyaṃ (or: yamindra-)] A1, Ba, Ba5, Bo3, C6, C8, C9, J7, J8, J10, J11, J14, LD1, M1, N17, V6, V22, V3, V4 (pathyaṃ missing), V10, V13, N2, N3, N4a, N12, N13, N14, N20, N26, Tue, P1, P6, P7, P8, P14, P18, P25, P27, R2, R5, P9
-\item[yamīṃdrapraparyya]   N24
-\item[yamīḍyapathyaṃ] P15,
-\item[yatīṃdrapathyaṃ]    J2, N10, N23
-\item[yavatīṃdrapathyaṃ]    P11,
-\item[yamendrapathyaṃ]    C3,
-\item[yogīdrapathya]    J1,     
-\item[mūnidra pathayam]    V8
-\item[munidra pathayam]    J12,
-\item[munīdramukhyaiḥ]    N7,
-\item[muṃhanīdrapathya]    J15
-\item[suyogipathyaṃ]     N16
-\item[varayogipathyaṃ]    N18
-\item[jamodūpathyaṃ]    R1
-\item[paramaṃ hitaṃ hi]    Ko,
-\item[(illegible/unavailable)]   C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22,
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp01_063]
-\sthana{1.63a}
-  \begin{description}
-\item[iṣṭaṃ sumadhuraṃ]    A1, J3, J6, J10, N18(appears after one verse), P7, R2, P9
-\item[iṣṭaṃ sumadhura]    P11,
-\item[iṣṭaṃ samadhuraṃ]    C9, J7, J17, LD2, N17, V19, V4, LD1, P1, P8, R1
-\item[iṣṭaṃ saṃmadhuraṃ]   N7,
-\item[iṣṭaṃ samakaraṃ]   P10, 
-\item[daṣṭaṃ samudhūraṃ]    J16,
-\item[dṛṣṭaṃ samūdhuraṃ]    K1
-\item[dṛṣṭaṃ sumūdhuraṃ]    N10
-\item[piṣṭaṃ sumadhuraṃ]    C8, N8, N11, N14("śu")
-\item[pathyaṃ sumadhuraṃ]    J12,
-\item[puṣṭaṃ sumadhuraṃ]    Ba5, Bo3, N4a, N13, P14, P18, Tue
-\item[puṣṭaṃ samadhuraṃ]   R5
-\item[yeṣṭaṃ sumadhura]    V13
-\item[yuṣṭaṃ sumadhuraṃ]    V10(mu°), V22
-\item[mṛṣṭaṃ sumadhuraṃ]    M1, V1
-\item[mṛṣṭaṃ sumadhurāṃ]    Ko
-\item[mṛṣṭaṃ sumadhusnigdhaṃ]   N3
-\item[mṛṣṭaṃ samadhuraṃ]    C3,
-\item[miṣṭaṃ samadhuraṃ]    J8, Ba, C4, N20, N26, P5, P24, 
-\item[miṣṭaṃ sumadhuraṃ]    Bo2, C6, J1, J11, J14, N2, N16, N19, N21, N22, N23, P13, P15, P21, P25, P27, R4
-\item[miṣṭaṃ samudhuraṃ]    J5, J15,
-\item[miṣṭaṃ sūmadhuraṃ]  P6
-\item[miṣṭaṃ sumadhura]   N25
-\item[miṣṭaṃ madhuraṃ]    V3
-\item[miṣṭaṃ ca madhuraṃ]    B2
-\item[maṃ(ṣṭ?)aṃ samudhuraṃ]  N9
-\item[sumiṣṭaṃ madhuraṃ]    N12
-\item[uṣṇaṃ sumadhuraṃ]    B1, Bo1, C2, C7, L1, J13, N1, P2, P17, P23, 
-\item[uṣṇaṃ sumudhuraṃ]    P20,
-\item[uṣṇaṃ sumadhura]    P12, 
-\item[uṣṇaṃ samadhuraṃ]    C1,
-\item[uṣma(?) sumadhuraṃ]    B3
-\item[uṣmaṃ sumadhuraṃ]    P4, R3(appears after one verse)
-\item[uśmā?mumadhuraṃ]  N5
-\item[suvṛkṣaṃ madhuraṃ]    V6
-\item[ghṛtam ca madhuraṃ]    V8
-\item[vṛṣyaṃ sumadhuraṃ]    J2,
-\item[tiṣṭaṃ tu madhuraṃ]  N24
-\item[(illegible/unavailable)]   C5, J4, N6, P3, G11, T, G1, Ba3, P16, P22, P26,
-    \end{description}
- \end{marma}
-
-Notes:
-    • P3, R1, has not 2nd chapter
-    • Folios of P4 are scattered 
-    • P16 has only 3 folios of chapter 2 at the end
-    • P22 has only 3 folios of chapter 1
-    • R3 has only 1 folio of chapter 2
-    • 1 & 2 Chapters are mixed in R4
-    • R5 has only one folio of chapter 2
-    • C5 has not 1st and 2nd chapters
-    • U1 has only chapter 2
-\begin{marma}[hp02_003]
-\sthana{2.3b}
-\item[jīvanam] Ba, Ba5, Bo2, Bo3, C1, C6, J1, J2, J3, J6, J12, J13, J14, J16, N17, P1, V1, N4a, N7, N8, N9, N10, N11, N12, N13, N16, P10, P13, P14, V16, V17, V10, N18, N21, P18, P20, P25, P26, P27, R5, V21, Tue, N23, V6, V7, V20, V28, V18
-\item[jīvaṃnam] N14
-\item[jīvitam] A1, B1, B2, B3, Bo1, C2, C3, C4, C8, J4, J5, J7, J8, J10, J11, LD1, LD2, N1, N2, N5, V8, N20, N22, P2, P6, P9, P17, P21, P23, V14, V3, N24, N25, P12, P15, R1, R3, V15, V5, V22, V2, V12, V13, V9, G11, Ba3, G1, U1,
-\item[jīvatam] P5, P24,
-\item[jivitam] J15,
-\item[jīvita] N26
-\item[jīviti] P11, 
-\item[jīvitum] J17, N19
-\item[jītavim] N3
-\item[jīvo na] Ko, P7, P8, 
-\item[(unavailable/illegible)] C5, C9, N6, T, P16, P22,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_003]
-\sthana{2.3d}
-\item[nibandhayet] Ba, Bo2, C6, V1, J2, J4, J6, J8, J11, J14, N12, N16, N19, N22, P11, P15, P25, V14, V3, V13
-\item[nibaddhayet] G11, P6, P27,
-\item[nirundhayet] A1, B1, B2, B3, C1, C3, C9, J5, J10, J13, J15, J17, LD1, LD2, N17, P1, V19, W4, N2, N7, N9, N10, N14, P5, P8, P9, P10, P23, V16, V17, N20, V21, N23, N26, P2, P7, P26, V20, V5, V2, V18, V12, V9, Ba3
-\item[nirodhayet] Ba5, Bo1, Bo3, C2, C4, C8, J1, J3, J6pc, J7, V19, N1, N4a, N5, N8, R3, V8, N11, N13, V10, N18, N21, P13, P14, P17, P18, P20, Tue, N24, N25, P12, P21, R5, V6, V7, V15, V28, V22, U2,
-\item[nirudhayet] P24,
-\item[niruddhayet] R1, 
-\item[nisaṃdhayet] J16,
-\item[nirodhanaṃ] J12,
-\item[nibandhanāt] V19, N3, G1
-\item[nibadhnayāt] Ko,
-\item[(unavailable/illegible)] C5, N6, T, P16, P22,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_004]
-\sthana{2.4d}
-\item[kāyasiddhiḥ] A1, B1, B2, C2, C6, J5, J8, J14, V1, N1, N8, N11, P2, P12, P17, P26, R3, V17, N24, V20, V15, V9, G11, G1
-\item[kāyāsiddhiḥ] J2, N3,
-\item[kāyāsuddhi] V6
-\item[kāyāsiddhi] P6, P27, 
-\item[kayasiddhiḥ] N20
-\item[kāyaśuddhiḥ] N23
-\item[kāryasiddhiḥ] B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C8, C9, J3, J4, J6, J10, J11, J13, J17, Ko, LD1, N17, P10, V1, W4, N2, N4a, N5, N9, N12, N13, V16, V10, N19, N21, P1, P5, P6, P7, P8, P9, P11, P13, P14, P15, P18, P20, P21, P23, P24, P25, R1, Tue, N25, N26, V70, V5, V28, V22, V2, V18, V12, V13, Ba3
-\item[kāryasiddhaḥ] R5,
-\item[kāryasiddhiṃ] J15, 
-\item[kāryāsiddhiḥ] J1,
-\item[kāryasiddhi] J12, LD2, N7, V8, N10, N14("rya" unclear), N16, N18, N22, V21, V3 
-\item[kāryasiddhī] J16,
-\item[kāryasuddhiḥ] V14
-\item[kāyāśuddhiḥ] J7,
-\item[(unavailable/illegible)] C5, N6, T, P16, P22,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_006]
-\sthana{2.6c}
-\item[suṣumnāsu snighdhā] J4, V1, V15
-\item[suṣumnāḥ snighdhā] P15, 
-\item[suṣumnānāḍisthā] Ba5, N13, P13, P14, P20, Tue, 
-\item[suṣumṇāṃ nāḍisthāḥ] P21, 
-\item[suṣumṇānāḍīsthā] N4a, N21, P18,  
-\item[suṣumṇānāmadhysthā] V2
-\item[suṣumnā nāḍīsthaṃ] G1
-\item[suṣumṇānāḍīstho] N24
-\item[suṣumnānāḍīstho] J14,
-\item[suṣumṇānāḍīschā] Bo3, V28, V22
-\item[suṣumṇāṃtasthhā] J15,
-\item[suṣumnāṃtaraschā] J11
-\item[sukhumnāṃtarasthā] V19
-\item[suṣumnāṃtarālasthā] B3, Bo1, Bo2, C1, C2, J3, J6, J13, N1, N8, N11, P2, P17, P23, 
-\item[suṣumnāṃtarālaschā] P12, 
-\item[suṣumnāṃtarālasyā] J1,
-\item[suṣumnāṃtarājasthā] N16
-\item[suṣumṇāṃtarasthā]  N5, C4
-\item[suṣumṇāṃtarāla] V5
-\item[suṣumnā pāśvaṃsthā] A1, V20
-\item[yathā suṣumṇā pārśvaḥsthā] Ba,
-\item[yathā suṣumṇā pārśvasthā] C6, C8, J7, N23, P7, P25, P26, 
-\item[suṣumṇāpañcasthā] N19
-\item[yathā suṣumnā mukhasthā] V17
-\item[yathā suṣumnā sumalāḥ] P27,
-\item[yathā suṣumnā nāḍisthamala] V10
-\item[yathā suṣumnņayā śvaḥsthā] V8, V21
-\item[yathā suṣumņā sustakcya?] P11, 
-\item[suṣumņā sustabdhāḥ] N12
-\item[yathā suṣumnā susvasthā] J5, N22, N26, V14, V3, N3
-\item[suṣumņā suṣasthā] N20
-\item[suṣumnņayā ścasyā] V6, V13
-\item[suṣumnāvasthāyaṃ] B1, B2
-\item[suṣumnānāḍibhyāṃ] V7
-\item[suśumnāsusvasthā] G11
-\item[suṣumnāsvasthāyan] N2
-\item[suṣama susvasthā] J8
-\item[suṣumnāsui svasthā] J8ac,
-\item[suṣumnāsaṃsvasthā] J2,
-\item[suṣumṇā sūmalā] P6,
-\item[sukham avasthāya] C3, C9, J10, J12, J17, LD2, N17, P1, P5, P24, W4, N7, N9, V16, V18, V12, V9, Ba3
-\item[sukhamalāsyāya] P9, 
-\item[sukham avasthāyaḥ] J16,
-\item[sukham avasthāyā] N14, P8, 
-\item[sukham avasthāpyā] LD1
-\item[sukham avaschāya] R1,
-\item[sukham avaschāṃya] N10
-\item[sukha........] P10,
-\item[suṣam avasthāya] J8pc, N18,
-\item[suṣumṇām avasthāpya] J7pc
-\item[suṣumnāṃtvarayāpalaḥ] N25
-\item[(unavailable/illegible)] C5, Ko, N6, T, P16, P22, R5,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_009]
-\sthana{2.09 additional verses}
-dvādaśamātraḥ kanīyānmadhyamo mātracaturvviśaḥ |
-uttamaḥ ṣaṭtriṃśanmātro mātrābhedāḥ smṛtā tajñaiḥ |
-vaśiṣṭasaṃhitādau tu pūrakaṃ ṣoḍaśamātrabhiḥ |
-kuṃbhakaṃ ca catuṣṣaṣṭimātrābhiḥ recakaṃ tu |
-dvātriṃśanmātrābhiḥ vidhivatkuṃbhakaṃ kṛtvā punaścaṃdreṇa recayet ||
-\begin{description}
-\item[Yes] J17, R1, 
-\item[No] A1, B1, B2, B3, Ba, Ba5, Bo1 Bo2, C1, C2, C3, C4, C5, C6, C8, C9, J1, J2, J3, J5, J10, J11, J12, J14, J15, J16, LD2, N17, V1, J4, J6, K7.
-J8, J13, Ko, LD1, W4, V19, N1, N2, N3, N4a, N5, N6, N7, N8, N9, V28, V8, N10, N11, N12, N13, N14, N16, P5, P10, V16, V17, V10, N18, N19, N20, N21, N22, P1, P6, P7, P8, P9, P11, P12, P13, P14, P15, P16, P17, P18, P20, P21, P22, P23, P24, P25, P26, P27, V14, V21, V3, Tue, N23, N24, N25, N26, P2, R5, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9, G11, T, Ba3, G1
-    \end{description}
-
-\begin{marma}[hp02_011]
-\sthana{2.11 additional verses}
-iḍayā piva ṣoḍaśabhiḥ pavanaṃ |
-kuru ṣaṣṭicatuṣṭayamaṃtagataṃ |
-tyaja piṃgalayā sanakaiḥ sanakaiḥ |
-daśabhirdaśabhirdaśabhirdvyadhikaiḥ ||
-\begin{description}
-\item[Yes] B1, B3, C, V19, N1, N5, N10 , N12, N21, N25, P12, P13, P26, P17, P20, P21, V28
-\item[No] A1, B2, Ba, Ba5, Bo1, Bo2, C1, C2, C3, C4, C5, C6, C8, C9, J1, J2, J3, J10, J13, J17, J4, J5, J6, J7, J11, J12, J14, J15, J16, Ko, LD1, LD2, N17, P1, V1, J8, W4, N2, N3, N4a, N6, N7, N8, N9, V8, N11, N13, N14, N16, P5, P6, P8, P15, P18, V16, V17, V10, N18, N19, N20, N22, V14, V21, V3, Tue, N23, N24, N26, P2, P7, P9, P10, P11, P14, P16, P22, P23, P24, P25, P27, R1, R5, V6, V7, V20, V15, V5, V22, V2, V18, V12, V13, V9, G11, T, Ba3, G1
-    \end{description}
-
- \begin{marma}[hp02_012]
-\sthana{2.12cd}
-\item[uttānaṃ cottame prāṇarodhe padmāsane muhuḥ] Ba, C1, C6, P25, V14
-\item[uttānaṃ cottame prāṇe rodhe padmāsane muhuḥ] P7, 
-\item[uttānaṃ cottame prāṇe stebdhe? padmāsane muhuḥ] C8,
-\item[uttame sthānam āpnoti tato vāyuṃ nibaṃdhayet] N13, V10, Tue, V7
-\item[utiṣṭhe tūtame prāṇā yāvat padmāsane sūmaṃ] J2,
-\item[uttiṣṭhaty uttame prāṇaṃ rodheḥ padmāsanaṃ .uhuḥ] V1
-\item[uttiṣṭhāṃty uttame prāṃṇā baddhe padmāsane uttama] J16
-\item[uttiṣṭhaty uttame prāṇē baddhe padmāsanae dṛḍhe] C2, P17, 
-\item[uttiṣṭhaty uttame prāṇaḥ baddhe padmāsanae dṛḍhe] J13,
-\item[uttiṣṭhaty uttame prāṇaṃ baddhā padmāsane dṛḍhe] B1
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane dṛḍhe] B2, B3, P2,
-\item[uttiṣṭhaty uttame prāṇa baddhe padmāsanaṃ dṛḍhaṃ] Ba3
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane dṛḍhe] P13, 
-\item[uttiṣṭhaty uttame prāṇaḥ badhe padmāsane dṛḍhe] P27,
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane sadṛḍhe] P12,
-\item[uttiṣṭhaṃty uttame prāṇā baddhe padmāsane muhuḥ] LD2, J10, P1, P8, P9, P10, P24, 
-\item[uttiṣṭhaṃty ūttame prāṇā baddhe padmāsane muhuḥ] J15, P5, R1, 
-\item[uttiṣṭhaṃty uttame prāṇā baddha padmāsane muhuḥ] C3, N10
-\item[uttisthaṃ cottame prāṇā baddha padmāsane muhu] N14
-\item[uttiṣṭhaty uttame prāṇa baddhe padmāsanaṃ muhuḥ] J8ac, J14, V3
-\item[uttiṣṭhaty uttame prāṇa baddho padmāsane muhuḥ] V8
-\item[uttiṣṭhatm uttame prāṇa baddhe padmāsane muhuḥ] V16, V17, V21, V20, V18, V13
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane muhuḥ] C9, 
-\item[uttiṣṭhaty uttame prāṇo baddha padmāsane muhuḥ] A1
-\item[uttiṣṭhaty uttame prāṇo badhye padmāsane muhuḥ] N9
-\item[uttiṣṭhaty uttame prāṇo baddhapadmāsane muhuḥ] N7
-\item[uttiṣṭhaty uttame prāṇē baddhe padmāsanaṃ muhuḥ] J8pc
-\item[uttiṣṭhaty uttame prāṇē baddhapadmāsane muhuḥ] P26, 
-\item[uttiṣṭhaty uttame prāṇē baddha padmāsane haye] Bo1
-\item[uttiṣṭhaṃty uttare prāṇā baddhe padmāsane muhuḥ] J17
-\item[uttiṣṭhaṃty uttame prāṇā baddhe padmāsane muhuḥ] W4
-\item[uttiṣṭhaty uttame prāṇā baddhe padmāsane muhuḥ] N20
-\item[uttiṣṭhaty uttame prāṇo baddhe padmāsane mukuḥ] N26
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsana sthitaḥ] Bo2, J1, J3, N8, N11, N16, V2
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsana sthite] C4, N25
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthite] P21, P23, 
-\item[uttiṣṭhaty uttame prāṇāṃ rodhe padmāsane sthite] V19
-\item[uttiṣṭhaty uttame prāṇo rodhe padmāsane sthite] P20, V5, V28
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsane sthitaḥ] J6, N2
-\item[uttiṣṭaty uttame prāṇa rodhe padmāsanasthite]  N5, P6, 
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsanaṃ muhuḥ] N3, , P15, V15
-\item[uttiṣṭhaty uttāme prāṇarodhe padmāsanaṃ muhuḥ] J5, J11, G1
-\item[uttiṣṭhaty uttāme prāṇarodhe padmāsano muhuḥ] J7
-\item[tattiṣṭhaty uttāmaprāṇarodhya padmāsanaṃ muhuḥ] J4
-\item[uttiṣ(?)yaty uttame prāṇa rodho padmāsanaṃ muhuḥ] N24
-\item[uttiṣṭhaty uttame prāṇaṃ rodhe padmāsanaṃ muhuḥ] P11, 
-\item[uttiṣṭhaty uttame kārye rodhe padmāsanae muhu] Ko,
-\item[uttiṣṭhaṃty uttame prāṇa rodho padmāsane sthitaḥ] N18
-\item[uṭṭiṣṭhaṃ cottare prāṇā baddhe padmāsane muhuḥ] N17
-\item[uttiṣṭhatyuttame prāṇarodhe padmāsane muhuḥ] G11
-\item[uttiṣṭaty attame prāṇe baddhe padmāsane dṛḍhe] N1
-\item[uttiṣṭhaty uttame prāṇe yāvat padmāsane samaḥ] N12
-\item[uttiṣṭhaty uttare prāṇā baddhe padmāsane dṛḍhe] J12,
-\item[uttiṣṭhaty unname prāṇe baddhe padmāsane dṛḍhe] N19
-\item[uttiṣṭhaty uttame prāṇa baddha padmāsanaṃ dṛḍhaṃ] N21
-\item[uttiṣṭety uname prāṇaḥ rodhe pavāsane sthite] N22
-\item[uttiṣṭhaty uttame prāṇa rāvai padmāsano muhuḥ] N23
-\item[uttiṣṭhaty uttame prāṇa rodhe padmāsano muhuḥ] V6
-\item[uṭṭiṣṭhnty uttame prāṇā baddhe padmāsane muhuḥ] LD1,
-\item[uttame sthānaṃ āpnoti tato vāyuṃ nibandhayet] P14, P18, 
-\item[(illegible/unavailable)]   Ba5, Bo3, C5, N4a, N6, P22, V22, V12, V9, T, P16, R5, 
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_019]
-\sthana{2.19c}
-\item[kṛśatā] A1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C6, C8, J1, J4, J5, J6, J7, J11, J13, J14, V19, N1, N2, N4a, N5, N8, N11, N12, N13, N16, N18, N19, N22, P2, P5, P6, P7, P11, P12, P13, P14, P15, P17, P18, P20, P23, P25, P26, P27, C4, V17, V10, V21, Tue, N23, N26, V7, V20, V5, V28, V22, V2, V18, G11, G1
-\item[kṛsatā] Ko, J3,  
-\item[kṛśate] J10pc
-\item[kriyate] C3, J10, J12, J16, J17, LD1, N7, N9, N10, P1, P8, P9, P10, R1, V12, C9, J2, N17, LD2, V19, Ba3
-\item[kṛyate] J15, N14
-\item[kṛśatāṃ] J8, V8, N21, V3, N25, P21, V6
-\item[kṛṣitāṃ] N24
-\item[kṛtā] N3, N20, V14
-\item[krūratā] J2
-\item[(illegible/unavailable)]   B1, B2, C5, N6, V16, V15, V13, V9, T, P16, P22, P24, R5,
-
-\begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_021]
-\sthana{2.22a}
-\item[medaḥśoṣādikaṃ pūrvaṃ] P6, V1
-\item[medaḥśleṣmādikaṃ surve] V6
-\item[medaśokādikaṃ pūrvaṃ] N22
-\item[medaḥśleṣmādi nāśārthe] A1, C3, P26, 
-\item[medaḥśleṣmādi nāśārtha] LD2,
-\item[medaḥśleṣmādi ṣatkarmāṅi] V28
-\item[medaḥśleṣmādi nāśārthaṃ] B1, B3, Bo1, C2, J10, J12, J13, J15, J16, J17, LD1, N7, N17, P1, P2, P5, P8, P9, P12, P13, P17, R1, V21, V20, V12
-\item[medaḥśleṣmādhi nāśārthaṃ] P27, 
-\item[madaṃ śleṣmādi nāśārthe] P10, 
-\item[nādaśleṣmādinā śāstraṃ] Ba3
-\item[medaḥśleṣmādināśīrṣaṃ] N9
-\item[medaḥ tuṣmād adhikei purva śaṭ karmaṃ] V8
-\item[medaśleṣmādi nāśārthaṃ] C9, N1, N10, N14, N21, V18
-\item[medaśleṣādi nāśārthe] V17
-\item[medaśleṣmādi nāsārthe] N20, N26
-\item[medaśleṣma nivṛtyarthaṃ] Bo2, C1, J1, J3, J6, N2, N8, N11, N16, N18, P20, P21, P23, V15, V5, V2
-\item[medaḥśleṣma nivṛtyarthaṃ] N5, C4
-\item[??śleṣma nivṛtyarthaṃ] N25
-\item[medaḥśleṣmādhikaḥ pūrvaṃ] Ba, C8, J5, P7, P11, P14, P18, V10, V14, V13, G11
-\item[medaḥśleṣmādhikaḥ pūrva] J7,
-\item[medaḥśleṣmādhikaḥ arthaṃ] V16
-\item[medai śleṣmāṃdhika purṇa] J2
-\item[medaśleṣmāṃdhikaḥ pūrvaṃ] Ba5, Bo3, V3
-\item[medaḥ ślṣāṃdhikaḥ pūrvaṃ] C6, J14, 
-\item[medaśleṣmādhikaḥ pūrvaṃ]  J11, N3, N4a, N12, N13, P25, Tue, V22
-\item[medāśleṣmāṣikaḥ pūrvva] N23, V7
-\item[medaḥśleṣmaṇisapurṇa] N24 
-\item[mitaḥ sthūlādikaiḥ kāryyaṃ] B2
-\item[medodhikastu havabhiḥ] Ko,
-\item[(unavailable/illegible)] C5, J4, N6, N19, P15, V9, T, G1, P16, P22, P24, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_021d]
-\sthana{2.22d}
-\item[apy abhāvataḥ] V1
-\item[api bhāvataḥ] C3
-\item[samabhā] R1,
-\item[samabhāvataḥ] A1, B1, Ba5, Bo1, Bo3, C9, J1, J3, J6, J10, J11, J12, J13, J14, J15, J17, LD1, LD2, N3, N4a, N7, N8, N10, N11, N12, N13, P1, P5, P10, V10, N20, N21, P6, P8, P9, P11, P13, P14, P18, V14, V3, Tue, Ba3, N26, V7, V20, V15, V28, V22, V18, V12, V13, G11
-\item[samabhāvat] J16, N17, 
-\item[samupāgataḥ] J7,
-\item[samabhyāvataḥ] N24
-\item[api samācaret] V16, V17, V21
-\item[samabhāvata]  N9	
-\item[samabhāvanā] N22
-\item[smabhāvata]  N14
-\item[sabhāgataḥ] Ba,
-\item[samabhāgataḥ] B2, C6, C8, P7, V8, N18, N23, P25, P27, 
-\item[samabhāgatā] Ko,
-\item[samabhāgata] V6
-\item[samabhāgikaḥ] B3, C2, N1, P2, P12, P17, 
-\item[samabhārat] J2
-\item[samahāvata]  P26, 
-\item[samatā yataḥ]  Bo2, C1, C4, N2, N16, N25, P20, P23, V2
-\item[samatā yatāṃ]  P21, 
-\item[samatā yathā]   N5,
-\item[samatāpataḥ] V5
-\item[(unavailable/illegible)] C5, J4, N6, N19, P15, T, G1, P16, P22, P24, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_022a-c]
-\sthana{}
-  \begin{description}
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_023c]
-\sthana{2.22c}
-\item[vicitraguṇasandhāyī(i)] A1, B1, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C2, C3, C8, C9, J1, J2, J3, J4, J5, J6, J7, J11, J13, J14, J15, J16, LD1, LD2, P7, V1, W4, J10, N17, J17, J8ac, Ko, N1, N3, N4a, N7, N9, N10, N16, P1, P2, P6, P9, P10, P11, P12, P13, P15, P17, P18, P21, P24, P25, P27, R1, V8, V16, V17, V10, N18, N19, N21, N22, P14, V14, V3, Tue, N23, V7, V15, V5, V28, V22, V2, V18, V12, V13, G11, Ba3
-\item[vicitraguṇasadhāyī] P26,
-\item[vicitraguṇaśandhāyī]  P8, 
-\item[vicītraguṇaṃsandhāyi] P5, 
-\item[vicitraguṇasaṃyāyi] P23,
-\item[vicitraguṇasandhāya]  C6, N12
-\item[vicitraguṇasandhāyā] N20, V20
-\item[vicitragunasandhāyai]  N14, V21, V6
-\item[vicitraguṇasandhāryā]  N13
-\item[vicitraguṇasandhāryaṃ]  J12,
-\item[vicitraguṇasandhānaṃ] N26
-\item[vicitraguṇasadyāpi] N2
-\item[vicitraguptasaṃ?jñopi] N5
-\item[vicitraguṇasaṃjñopi] C4, P20, N25
-\item[vicitraguṇasaṃdhāyoga]  N8, N11,
-\item[trividhaṃ guṇasandhāyi] B2
-\item[(unavailable/illegible)] C5, N6, N24, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_24c]
-\sthana{2.24c}
-\item[graset pratyāharec caitad] P25, 
-\item[tataḥ pratyāharec caitad] A1, Ba, Ba3, Bo1, B3, C8, G11, J12, J16, J17, N9, N12, N14, N17, N20, N21, N23, P1, P5, P8, P9, P10, P11, P12, P13, P15, P24, P26, P27, R1,
-\item[tatao pratyāharec caitad] C9,
-\item[tataḥ pratyāhare caitad] C6,
-\item[tataḥ pratyaharec cetad] N10
-\item[tataḥ pratyaharec caitad] N13, P7, 
-\item[tata pratyaharec caiva] N24,
-\item[punaḥ pratyaharec caitad] B1, B2, P2, P14, P17, P18, Tue,
-\item[punaḥ pratyahare caitad] J13, J14, 
-\item[punaḥ pratyāhare caitad] N19, 
-\item[punaḥ pratyāhare caitā] N22,
-\item[pūnaḥ pratyāharec caityāt] P6,
-\item[punaḥ pratyahared etad] C1, C2, C3, C4, J1, N8, N11, N16, N18, N25, N26, P20, P21, P23, 
-\item[punaḥ pratyaharec caitaḥ] J15,
-\item[(unavailable/illegible)] C5, J1, N6, N7, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_24d]
-\sthana{2.24d}
-\item[prakṣālyaṃ dhautikarma tat] Ba, C6, P7, P25, V21
-\item[prakṣālanaṃ dhautikarma tat] V13
-\item[prakṣālye? dhautikarma tat] C8
-\item[ādāyaṃ dhautikarma tat] V14
-\item[udgāraṃ dhautikarma tat] V1,
-\item[sodgāraṃ dhautikarmavit] N12
-\item[uditaṃ dhautikarma tat] A1, Ba5, Bo1, Bo3, C9, J10pc, J12, J15, LD1, LD2, N4a, N7, N9, N10, N13, N14, N17, P5, P8, P9, P10, P14, P18, P24, V16, V17, V10, N20, N21, Tue, N26, V7, V20, V15, V18, V12
-\item[uditaṃ dhātikarma tat] P13, 
-\item[udita dhaitikarma tat] G11
-\item[aditaṃ dhautikarma tat] J17,
-\item[uditaṃ dhautikarma kṛt] J10, P1, R1, V22
-\item[up(?)itaṃ dhautikarma tat] C3
-\item[abhyāsād dhautikarmavit] B1, B2, C2, C4, J1, J3, J6, J13, P2, P12, P17, P20, P26, V5, V28, V2
-\item[abhyāsād dhautikarmasu] N25
-\item[abhyāsā dhautikarmavit] N18
-\item[abhyāsā dhautikarma tat] P6,
-\item[abhyāsā dhautikarmavit] P21,
-\item[abhyāso dhautikarmavit] N19
-\item[abhyase dhautikarmavit] P27,
-\item[abhyāse dhautakarmavit] J14,
-\item[abhyāsād dhotikarmavit] C1
-\item[abhyāsād dhautikarmavat] N5
-\item[abhyāsaud dhautikarmavit] Bo2,
-\item[abhyāsāsau dhautikarmavit] V6
-\item[abhyāsau dhautikarmatat] V8
-\item[abhyāsād dhautikarmavita] N2
-\item[abhyāsād dhautikarma tat] N1 (after correction),
-\item[dukhalātā dhautikarmataḥ] V3
-\item[duḥkhalṃ dhautikarma tat] Ko,
-\item[duṣālaṃ dhautikarma tat] J4, J11, 
-\item[duḥkhahaṃ dhautikarma tat] P11, 
-\item[uṣālaṃ dhautikarma tat] J5, P15, 
-\item[abhyāsa dhautakarka tat] N22
-\item[abhyāsād dhauti bhavati karma tat] N8, N11 ("bhavati" in brackets)
-\item[uṃtthānaṃ dhautikarma tat] N23
-\item[dākṣālaṃ dhautikarma tat] N24
-\item[utthānaṃ dhautikarma tat] J7,
-\item[dugdhāraṃ dhautikarma tat] J2
-\item[syā sā dhautikarmakṛt?] B3
-\item[?s?vāla dhautikarmakṛt] N3
-\item[(unavailable/illegible)] C5, N6, N16, T, G1, P16, P22, P23, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_25d]
-\sthana{2.25d}
-\item[bādhate yaṃ na] P27,
-\item[chadhyate ca na] J16,
-\item[dhāvaty eva] V1, V8, V3, V2
-\item[dhāvaṃtyeva na] P6, P15, 
-\item[dhāvate ca na] B2
-\item[dhāvaṃta ca na] N19
-\item[dhāvaṃte ca na] J5,
-\item[dhavaṃte ca na] N22
-\item[dhruvntye neva] Ko,
-\item[gachaty eva na] N21, P13, 
-\item[gachaṃty eva na] Bo1
-\item[prayāntyeva na] Ba, N4a, Bo2, Bo3, C1, C4, C6, C8, J1, J3, J6, J7, N5, N8, N11, N13, N16, P7, P11, P14, P18, P20, P21, P25, V10, Tue, N25, V28, V22
-\item[dhāvatyeva na] G11
-\item[prayānteva na] N23, V7
-\item[śudhyatty eva] A1,
-\item[śudhyaty eva] LD2, P2, P9, 
-\item[śudhyate nātra] J12,
-\item[śudhyate ca na] N10, P26, 
-\item[śudhyanty eva] B1, C3, C9, J10, J13, J17, N1, N17, P5, P8, P12, P17, Ba3
-\item[śudhyanteva] B3, P10, 
-\item[śudhyaṃteva na] J15, N9, N14, P24, V16, V17
-\item[śvadhvatyeva na] R1,
-\item[śudhyaṃte ca na] LD1, N7, P1, 
-\item[suddhyaṃtevaṃ na] N18
-\item[śudhyaṃ sevana] C2,
-\item[kṣīyaṃte saklā malā] J14,
-\item[?īyatsyeva na] N2
-\item[bhavaṃty eva na] N3, V14
-\item[naśyaṃteva] N12
-\item[naśyaṃte] J11
-\item[nāśaṃ yāṃti] J4, 
-\item[naśyaṃty eva na] J2,
-\item[jayaty eva na] N24
-\item[śuṣyaṃte sakalāmalāḥ] N20, N26
-\item[(illegible/unavailable)] Ba5,C5, N6, V21, V6, V20, V15, V5, V18, V12, V13, T, G1, P16, P22, P23, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_26a]
-\sthana{2.26a}
-\item[udaragatapadārtham udvahatī] J12,
-\item[udaragatapadārthaṃ  utbhavanti] G1
-\item[udaragatapadārtham udvahati] R1,
-\item[udaragatapadārtham udvahantī] Ba, C3, J1, J7, Ko, N10, P6, V1, V16, N12, N18, N22, N23, P20, P27, V6, V28, V2
-\item[udaragatapadārthamūrdhāhatī] J16,
-\item[uragatapadārtham udvahantī] B2,
-\item[udaragatapadārthān udvahaṃti] C6, P7, 
-\item[udaragatapadārtham uşmahantī] V17
-\item[udaragatapadārtham udvamaṃti(ī)] B1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J11, J14, LD1, LD2, N7, N13, N16, N16, N20, N21, N26, P2, P8, P9, P11, P12, P13, P14, P17, P18, P21, P25, P26, V3, V20, V22, V18, V12, V13, Ba3
-\item[udaragatapadārtham udvayant] P5,
-\item[udaragatapadārtham uddhayaṃt] P24, 
-\item[udaragatpadārtham udvamaṃtī] N9,  
-\item[udaragatapadārtham udvavaṃtī] J10, J17, P1,
-\item[udaragatapadārtham udvamaṃta] J4, P10, 
-\item[udaragatapadārtham udvamaṃte] P15, V15
-\item[udaragatapadārthām udyamaṃte] N19
-\item[udaragatapadārthaṃ śuddhapetaṃ] N25
-\item[udaragatapadyarthān udvamaṃtī] A1
-\item[udaragatapadārtham uḍhasaṃtī] N24
-\item[udaragatapadārtham udvapanti] N14, 
-\item[udaragatapadārthaśodhanaṃ syāt] Bo1
-\item[udaragatapadārtha śuddhayeta] C4
-\item[udaragatapadārtha śuddhaye vai] J15, 
-\item[udaragatapadārtha suddhayeta] J13,
-\item[udagatapadārtha uddhanaṃtī] J5,
-\item[(illegible/unavailable)] Bo3, C1, J2, N6, N8, N11, R5, V19, V10, V14, Tue, C5, V21, V7, V5, V22, G11, T, P16, P22, P23, 
-  \begin{description}
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_26b]
-\sthana{2.26b}
-\item[pavanam apānam udīrya kaṇṭhanāle] V1, A1, B2, Ba, Ba5, C6, C8, C9, J3, J4, J6, J7, J10, J11, J14, J15, J17, Ko, LD1, LD2, N7, N9, N10, N13, N16, N17, P6, P7, P8, P9, P10, P13, P18, W4, J8, V16, N18, N20, N22, P14, P15, P20, P25, P27, V3, N23, V20, V15, V28, V18, V12, V13, G1
-\item[pevanam apānam udīrya kaṇṭhanāle] P5, P24, 
-\item[pavanam apānam udīrya kaṇṭhānālī] Ba3
-\item[pavanam apānam udīrya kaṇṭhanālaiḥ] N12, 
-\item[pavanam apānam udārya kaṇṭhanālo] J16,
-\item[pavanām apānam udīrya kaṇṭhanālṃ] J1,
-\item[pavana apānam udīrya kaṇṭhanāle] V6
-\item[pavanam upānam udīrya kaṇṭhanāle] C3
-\item[pavanam upānam urīrya kaṇṭhanāle] C4
-\item[pavanam apānam udīrya kaṇṭhanālo] N19, R1,
-\item[pavanam apānam udīrya kaṇṭhanālāt] B3, C2, J13, N26, P1, P2, P12, P17, P21, P26, V2
-\item[pavanam apām udirya kaṃnāle] N25
-\item[pavanam apāsenam udāryaṃ kaṇṭhanāle] J5,
-\item[pavanam apānam udīryakaṃ jāle] P11, 
-\item[pavanam mudārya kaṇṭhanālo] J12
-\item[pavanam udirya kaṇṭhanāle] N21
-\item[mapānam udīrya kaṇṭhanāle] B1,
-\item[pavanamapānam udīrya kaṇṭhanāle] Bo1, Bo2,
-\item[pavanam athānam utiya kaṇṭhanāle] N24
-\item[pavanam apānam udīrya karāḍanāle] V17
-\item[pavanam apānam udīrya kaṇṭhaḍālai] N14, 
-\item[(illegible/unavailable)] Bo3, C1, C5, J2, N6, N8, N11, R5, V19, V10, V14, V21, Tue, V7, V5, V22, G11, T, P16, P22, P23, 
-\end{description}
- \end{marma}
-
- \begin{marma}[hp02_26c]
-\sthana{2.26c}
-\item[karibhir iva jalasya vāyuvegāt] A1, B1, B3, C2, C9, J1, J3, J6, J8, J10, J17, LD1, N7, N10, N16,  N17, P1, P2, P9, P12, P17, W4, V16, V17, N18, N20, N26, V2, V18, V12, Ba3
-\item[karibhir iva jalasya vāyuvegaj] R1,
-\item[karibhir iva gajasya vāyuvegāt] V20
-\item[karibhir iva jalasya vāyuvegā] B2, C3, J12, J16, N14, P8, 
-\item[karibhir iva jalasya vāyubagā] P26,
-\item[karibhir iva jalasya vāyuvege] N9,
-\item[karibhir iva jalaṃ ca vāyuvegā] Ba,
-\item[karibhir iva jalaṃ ca vāyuvegāt] P25, 
-\item[karibhiriva jalaṃsya vāyuvegāt] Bo2, 
-\item[karibhiriva jalasya vāyuvego] J13,
-\item[karibhiriva jalasya vāyuyogāt] V28
-\item[karibhirivadāyikā jalabhya vāyuvegāt] P21, 
-\item[kramamarivayavasya vāyur yo] V1
-\item[kramaparicayavaśya nāḍicakrā] Ba5, N13, P14, P18,  
-\item[kramaparicayavaśyā nāḍīmārgā] C6, J7, LD2, 
-\item[kramaparicayavaśyā nāḍīmārgāt] N12,
-\item[kramaparicayavaśya vāyumārga] C8
-\item[kramaparicayavaśya vāyumārgā] P7, P20, 
-\item[kramaparicayavasya vāyuvegā] P27,
-\item[kramaparicayavaśa nāḍimārgā] P11,  
-\item[kramaḥ paricayavasya vāyuḥ mārgre] P6,
-\item[kramaparicayavasya vāyumārgre] J14,
-\item[kramaparicayavaśya vāyumārgāt] Bo1
-\item[kramaparicayavaśya vāyumārgaṃ] N21, P13, 
-\item[kramac pacya paricaya vaśyā vāyumārgaṃ] J5
-\item[kramaparicayavaśyaṃ vāthagargāṃ] N24
-\item[kramaparicayavaśya mārgte(??)] N23
-\item[kramaparicayavasya vāyumārge] N22
-\item[kramapārīcita vāyumārge] J15,
-\item[kramaparījita vāyuḥ] P15, 
-\item[kramapārīcitavacana vāyumārgā] P5, P24, 
-\item[kramaparicitavasya vāyuyogo] Ko,
-\item[kramapārīcitavaśya vāyumārg(oṃ)?] N19
-\item[kramapārīcitavaśya vāyumārgo] J11
-\item[kramapārīcitavaśya vāyumārge] J4, 
-\item[kramapārecitavaktu? vāyumārge] C4
-\item[kramaparicittakakrū vāyumārge] N25
-\item[kirimir iva jalasya vāyuvegāt] P10, 
-\item[(illegible/unavailable)] Bo3, C1, J2, N6, N8, N11, R5, V19, V10, V14, Tue, C5, V21, V6, V7, V15, V5, V22, V13, G11, T, G1, P16, P22, P23, 
-  \begin{description}
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_26d]
-\sthana{2.26d}
-\item[gajakaraṇānti nigadyate haṭhajṇau] Ko,
-\item[gajakarīṇīti kathyate haṭhavidyā] G1
-\item[gajakaraṇīti nigadyate haṭhajñaiḥ] B2, Ba, Ba5, Bo1, C3, C4, C6, J4, J6, J7, J11, J15, N12, N13, P5, P6, P7, P9, P10, P14, P18, P24, P26, P27, V1, N19, N23, N25, P13, V20, V15, V13
-\item[gajjakaraṇīti nigadyate haṭhajñaiḥ] P8, 
-\item[gajakaraṇī nigadyate haṭhajñaiḥ] N16, 
-\item[gajakaraṇīti nigadate haṭhajñaiḥ] J14,
-\item[gajakaraṇīti nigadyāta haṭhaj~naiḥ] J5,
-\item[gajakaraṇīti nigadyaṃte haṭhajñaiḥ] LD2,
-\item[gajakaraṇīti nigadyate] N24, V2
-\item[gajakarṇīti nigadyate haṭhaj~naiḥ] A1, C8
-\item[gajakaraṇīva nigadyate haṭhaj~naiḥ] B1
-\item[gajakariṇīva nigadyate haṭhajñaiḥ] C2
-\item[gajakaraṇīti nigachate haṭhaj~naiḥ] B3
-\item[gajakāraṇī nigadyate haṭhaj~naiḥ] Bo2,V3, V6
-\item[gajakaraṇīti nigadyate haṭhakṣaiḥ] N22
-\item[gajakaraṇī ca nigadyate dṛḍhajñaiḥ] J13,
-\item[gajakaranīti nigadye haṭhajña] N14, 
-\item[gogajakaṇīti nigadyate haṭhajñaiḥ] P15, 
-\item[jalagajakaraṇī nigadyate haṭhaj~naiḥ] J1, J3, 
-\item[jalakaraṇīti nigadyate haṭhajñaiḥ] C9, J12, J16, LD1, N7, N17, P1, W4, N20, N26, P21, P25, 
-\item[jalakaraṇīti nigadyato haṭhajñaiḥ] N10,
-\item[jalakaraṇīti nigadyate mahāhaṭhaj~naiḥ] V21
-\item[jalakariṇīti nigadyate haṭhajñaiḥ] J10, J17, N9, P20, R1, V16, V28, V18, V12, Ba3
-\item[jalakaraṇānī nigadyate haṭhaj~naiḥ] V17
-\item[jalagajakaraṇīti nigadyate haṭhaj~naiḥ] J8, N18, P2,
-\item[jalakaraṇīva nigadyate haṭhajñaiḥ] P12,
-\item[jalagajakaraṇīva nigadyate haṭhajñaiḥ] P17, 
-\item[gajakaraṇīti nirādyate haṭhaj~naiḥ] N21,
-\item[najakariṇīti nigadyate haṭhajñaiḥ] P11, 
-\item[(illegible/unavailable)] Bo3, C1, J2, C5, N6, N8, N11, V19, V10, V14, Tue, V7, V5, V22, G11, T, P16, P22, P23, R5,
-  \begin{description}
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_29c]
-\sthana{2.29c}
-\item[aseṣadoṣān api yan nihanyād] Ko, 
-\item[aśeṣadoṣopacayaṃ nihanyad] J8, N17, V1, W4, V16, V17, V10, V6, V7, V15, 
-\item[aśeṣadoṣoyathā nihanyānād] V14
-\item[aśeṣadoṣāmaya śoṣaṇāya] G1
-\item[aśeṣadoṣopacayaṃ nihanyād] B1, B2, B3, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C4, C6, C8, C9, J2, J3, J5, J6, J7, J10, J11, J13, J14, J15, J17, LD1, LD2, N7, N8, N11, N13, N16, N18, N19, N20, P1, P2, P5, P8, P9, P12, P13, P14, P15, P17, P18, P20, P21, P24, P25, P27, Tue, R1, N25, N26, V20, V5, V28, V22, V2.
-V18, V12
-\item[aśeṣadoṣopacayaṃti hanyād] J1,
-\item[aśeṣadopacayaṃ nihanyād] P23,
-\item[aśeṣa doṣā pacayan niśamyaha nyāda] G11
-\item[aśeṣadoṣo ca nayaṃti hanyād] N10,
-\item[aśeṣadoṣopacaya no hanyād] J16,
-\item[aśeṣadoṣopacayā nihanyād] V13
-\item[aśeṣadoṣopacayān nihanyād] N12,
-\item[aśeṣadoṣapacayaṃ nihanyād] N9,
-\item[aśeṣadoṣopacayaṃ nihanti] P7, 
-\item[aśeṣadoṣopacayaṃ nihanyā] N23,
-\item[aśeṣadoṣopacaya nihanyād] N24, P10, P26, 
-\item[aśeṣa?oṣovacayaṃ nihanyād] N21,
-\item[aśeṣadoṣasya cayaṃti hanyād] P6,
-\item[aśeṣadoṣepacayaṃti hanyād] J12, N14, 
-\item[aśeṣa doṣo vacayan nihanyād] Ba3
-\item[aśeṣadoṣepacayaṃ nihanyad] C2,
-\item[aśeṣadoṣāpacayaṃ nihanyad] V21, V3
-\item[aśeṣadoṣopacayaṃ nihatyad] P11, 
-\item[aśeṣadoṣopacitiṃ hathajñād] J4,
-\item[aśeṣadoṣopacayaṃ ca hanyad] A1
-\item[aśeṣadoṣaprabhavaṃ nihanyad] V19
-\item[aśeṣadoṣasya cayana hatyā] N22
-\item[(illegible/unavailable)] C5, N6, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_29d]
-\sthana{2.29d}
-\item[abhyasyamānaṃ jalavastikarma] Ba5, Bo1, Bo2, Bo3, C1, C3, C6, C8, J1, J2, J4, J5, J6, J7, J8, J11, J12, J13, J14, J15, J16, LD1, N8, N9, N10, N11, N12, N13, N14, N16, N17, P1, P2, P8, P9, P11,  P12, P14, P15, P18, P20, P23, W4, V16, V17, V10, N18, N19, N21, P13, P25, P27, V21, V3, Tue, N26, V6, V7, V20, V15, V5, V28, V22, V2, V18, V13, G11, G1
-\item[abhyasyamānaṃ janabastikarma] R1,
-\item[abhyasyamānaṃ jalavāstikarma] P5, P24, 
-\item[anyasyamānaṃ jalavastikarma] J3,
-\item[abhyasyamānaṃ nijavastikarma] V12
-\item[abhyasyamānaṃ jalavastikarmaḥ] Ko, P6, 
-\item[abhyamānaṃ jalabastikarma] P21, 
-\item[abhyasyamānaṃ khalu vastikarma] C9,
-\item[abhyasyamāna ja?vastikarma] N20(something seems added above)
-\item[abhyasyamānaṃ jalavastitakramāt] N24
-\item[praamāńam jalavastikarma] V14
-\item[abhyasyamāno jalavastikarma] B3, C2, P17, 
-\item[abhyasyamā jalavastikarma] B1,
-\item[abhyāsamānaṃ jalavastikarma] C4, V1, N7, N25, P10, P26, 
-\item[abhyāsyamānaṃ jalavastikarma] A1, Ba,
-\item[abhyāsamānaṃ jalavasti] V19
-\item[abhyasya mānam jalapastikarma] Ba3
-\item[abhyasyamānaṃ javastikarma] J10, J17,
-\item[abhyasyamānaṃ jvalabastikarmaḥ] LD2,
-\item[bhyasyabhānaṃ jalavestikarma] N23(space for “da” left blank with an overhead line)
-\item[abhyasya śūnyaṃ jalavastikarma] B2,
-\item[cābhyasyamānaṃ jalavastikarma] P7, 
-\item[dattasya mānaṃ yanabasyakarmaḥ] N22
-\item[(illegible/unavailable)] C5, N6, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_30]
-\sthana{2.30}
-\item[atha natīḥ] J1,
-\item[atha ti] P26, 
-\item[atha netī(ḥ)] Ba, Ba5, Bo1, Bo2, C1, C2, C4, J3, J6, J8, N8, N11, N16, V1, V10, N18, P20, P21, P23, V21, V3, N23, N25, N26, V7, V5, V2
-\item[atha neti] J2, Ko, N7, N9, N10, N13, N14, N20, N24, P7, P9, V6, V28
-\item[atha netiḥ] J7, N12, P14, P18, P25, Tue, V22, V12, Ba3
-\item[atha netikarma] C3, C9, J10, J17, LD1, P1, P2, P5, P8, P10, P24, W4, V16, V17, N21, P13, V14, V15, V18, V13
-\item[atha netikarmaḥ] C8, J12, J15, N22, P6, 
-\item[atha netīkarma] A1, B1, B2, C6, J13, J14, J16, N17, P12, P17, P27, R1, V20
-\item[atha netīkarmaḥ] LD2
-\item[atha nītī] J5,
-\item[atha nīti] P11, 
-\item[atha nītikarma] J4,
-\item[atha nītīkarma] N19
-\item[prathama netīkarma] B3
-\item[(illegible/unavailable)] Bo3, C5, N6, J11, G11, P15, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_30c]
-\sthana{2.30c}
-\item[mukhanirgamanād eva] A1, B2, C3, J10, J17, LD1, N7, N9, N17, P1, P5, P6, P8, P9, P10, P24, P26, V1, W4, V16, N22, V20, V18, V12, V13, Ba3
-\item[mukhanirgamanād evaḥ] J15,
-\item[mukhanirgamanād deva] C9
-\item[mukhanirgatanād evaṃ] J12,
-\item[mukhanirgatamanād eva] V14
-\item[mukhanirgamayedeti] J2,
-\item[mukhan nirgamanād eva] B3,
-\item[mukhān nirgamanād eva] C2, P12, P17, 
-\item[muvi nirgamanād eva] R1,
-\item[mukhā nirgamanād eva] P2, 
-\item[mukhṃ nirgamanād eva] N14, 
-\item[mukha nīrgamanod eva] J16, 
-\item[mukhān nirgamayec caiṣā] Ba, Ba5, Bo1, C8, J8, J11, LD2, N12, N13, V17, N19, N20, P14, P18, P25, V21, Tue, N26, 
-\item[mukhān nirgamyate caiṣā] J4,
-\item[mukhān nirgamayed eṣā] P13, 
-\item[sukhān nirgamayeccaiṣā] G11
-\item[mukhān virgamayed eṣā] N21
-\item[mukhā nirgamayec caiṣā] J7, P15, V10, V3, V6, V7, V15, V22
-\item[mukhe nirgamayec caiṣā] C6, P7, 
-\item[mukhe nirgamaye caiṣā] P11, 
-\item[mu nirgamayai caiṣā] J5,
-\item[mukhān nirgamaye doṣā] N24
-\item[mukhān nirgamayet sāpi] V19
-\item[mukhān nirgāmayet sā hi] B1, Bo2, J13, J14, N11, N18, N25  
-\item[mukhān nirgamanāt sā hi] P21, P23, 
-\item[mukhān nirgamayet sā hi] J6, N8, N20, P20, P27
-\item[mukhā nirgāmaye sā hi] V2
-\item[mukhā nirgāmayet so hi] Ko,
-\item[mukhā nirgamanā sā hi] V5
-\item[mukhān nigarmayet sā hi] C4, J3, N16, 
-\item[mukhā nigarmayet sā hi] J1, V28  
-\item[mukhān nirgatya nāsāyam] C1,
-\item[mukhaśca nirgamec caiṣā] N23
-\item[sukhanirgamanād eva] N10,
-\item[mukhaṃ nirgamayedeṣu nīti] G1
-\item[(illegible/unavailable)] Bo3, C5, N6, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_31cd]
-\sthana{2.30cd}
-\item[] 
-\item[(illegible/unavailable)] 
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_33a]
-\sthana{2.33a}
-\item[ghaṭanaṃ] J2, 
-\item[muñcakaṃ] Ko,
-\item[mucyate] J4,
-\item[mocanaṃ] B2, Ba5, C8, J7, J12, N12, N13, N18, N21, N22, P13, P14, P16, P18, P26, Tue, N23, V7
-\item[mocakaṃ] Bo2, C1, J1, J3, J5, J6, J11, J14, N16, N19, P15, P20, P21, P23, V28
-\item[moṭakaṃ] V1, Bo1, N25 
-\item[meṭakaṃ] P11, 
-\item[mokṣakaṃ] N8, N11, 
-\item[toṭakaṃ] V19, V16
-\item[tāṭakaṃ] V17
-\item[trāṭakaṃ] A1, C4, C6, P6, P7, P27, V10, V14, V21, V3, V6, V20, V15, V5, V28, V22, V2, V18, V12, V13, G11, Ba3, G1
-\item[troṭakaṃ] N24
-\item[rocanaṃ] Ba, P25, 
-\item[sphoṭanaṃ] B1, B3, C2, C3, C9, J8, J10, J13, J15, J16, J17, LD2, N7, N14, N17, P1, P8, W4, N20, N26, P2, P5, P9, P10, P12, P17, P24, Ba3  
-\item[sphoṭana] R1,
-\item[spheṭanaṃ] N10,
-\item[sphoṭakaṃ] LD1,
-\item[sphoṭa] N9, 
-\item[(illegible/unavailable)] Bo3, C5, N6, T, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_33b]
-\sthana{2.33b}
-\item[jadrānāṃ ca kapāṭakaṃ] J16,
-\item[tandrādīnāṃ gapāṭa[ka]m] V1,
-\item[tandrādīnāṃca pāṭanaṃ] P15, V19,
-\item[ta.ntryādīnāṃ kavāṭakam] G1
-\item[tandrādīnāṃ ca kapāṭakaṃ] B1,
-\item[tandrādīnāṃ kapāṭakam] B2, B3, Ba, Ba5, Bo1, C1, C2, C4, C6, C8, J3, J5, J6, J7, J8, J11, J12, Ko, LD1, LD2, N7, N8, N11, N13, N16, P2, P6, P7, P12, P13, P14, P16, P17, P18, P23, P25, P26, P27, W4, V17, V10, N21, N22, V3, Tue, V7, V5, V28, V22, V2, V13, Ba3
-\item[tadrādīnāṃ kapāṭakam] P21,
-\item[tandrādīnāṃ kavāṭakam] G11
-\item[tandrānāṃ tu kapāṭakam] N14,
-\item[tandrādīnāṃ kṛpāṭakam] N12, 
-\item[tandrādīṇāṃ kapāṭakam] J13, 
-\item[tandrāṇāṃ ca kapāṭakam] J15, N9, N10, P1, P8, P10, R1, 
-\item[tadbhānaṃ ca kapāṭakam] P5, P24, 
-\item[tandrādīnāṃ kapālakam] J1,
-item[tandrāegaṃ kathapāṭakam] V21
-\item[tandrādīnāṃ kapāṭakai] Bo2
-\item[tandrādiṇāṃ kapāṭakam] J8, N18, N23,
-\item[tandrādīnāṃ kaṃpāṭa] P9, 
-\item[tandrādīn kapāṭakam] P11,
-\item[tandrādiṇāṃ ca pāṭakam] J14, N19
-\item[tandrādīṇāṃ ca pāṭakam] C9
-\item[tandrādīṇāṃ ca pāṭhakam] V15
-\item[tandrādīṇāṃ ca pāṭavam] J4,
-\item[tandrāṇāṃ ca kapāṭakam] C3, J10, J17, V16, N20, V14, N26, V20, V18
-\item[tadvāṇāṃ ca kapāṭakam] A1,
-\item[tandrāṇāṃ tu kapāṭakam] N17
-\item[sandrādīnāṃ kapātakāṃ] N24
-\item[nidrādīnāṃ kapāṭakam] P20,
-\item[nidrārttīnāṃ kapāṭakam] N25
-\item[hetādīnāṃ? kapāṭhakam] J2,
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, V6, V12, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_34b]
-\sthana{2.34b}
-\item[gudaṃ] V1
-\item[gopyaṃ] P1, V16, V17, V10, N14, N18, V14, V21, V3,V6, V7, V20, V15, V28, V22, V2, V18, V12, V13, G11, Ba3, G1
-\item[tuda] N10,,
-\item[tundaṃ] A1, B3, Ba5, Bo2, C2, C8, C9, J3, J6, J7, J8, J13, J14, J16, LD1, LD2, N8, N9, N11, N13, N16, N19, N20, P2, P17, P18, P20, P26, Tue, N26, G11, P11, 
-\item[tunduṃ] N23
-\item[tunda] N12, N17 P16, 
-\item[tunde] C1, C4, P21, P23,  
-\item[tuṃḍe(?)] N25, V5
-\item[tundat] B2, J10, J17, P8, P10, R1, 
-\item[tundet] J15, 
-\item[truṃdaṃ] N21, P15, 
-\item[tuda] W4
-\item[tudaṃ] B1, C3, C6, J1, J5, J12, P6, P7, P12, P13, P14, P27, 	
-\item[tundan] J4,
-\item[tudan] Ba, Bo1, J11, P9, P25,  
-\item[tudat] N7, P5, P24, 
-\item[tulyaṃ] Ko,
-\item[tujyaṃ] V19
-\item[t(?)uddhaṃ] N22
-\item[puṃda] N24
-\item[saṃdada] J2
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_35a]
-\sthana{2.35a}
-\item[pāvanādi] V1
-\item[pāvanāni] N18,
-\item[pāvanāgniḥ] N25
-\item[pāvanādi] Ko,
-\item[pāvakādi] C3, J10, J16, N10, Ba3, R1, 
-\item[pāvakādī] J12,
-\item[pāvakāgni] N26
-\item[pācakādi] C2, C9, J17, LD1, LD2, N7, N9, N14, N17, P1, P5, P8, P9, P10, P24, W4, V16, V17, N20, V14, V21, V18, V13
-\item[pācatādi] Bo2, 
-\item[pācanādi] A1, B1, B2, B3, Ba, Ba5, Bo1, C6, C8, J4, J5, J7, J8, J11, J14, N12, N13, P6, P7, P14, P17, P25, P27, V19, V10, N19, N21, P2, P12, P13, P18, P20, V3, Tue, N23, V6, V7, V20, V15, V28, V22, G11, G1
-\item[pācanādīn P26, 
-\item[pācanāgni] C1, C4, J1, J3, J6, J13, N8, N11, N16, P16, P23, 
-\item[pācanāgnī] P21,
-\item[pācakāgniḥ] J15
-\item[pācakāgni] V2
-\item[pācakāri?] J2,
-\item[pācanānī] P11, P15, 
-\item[pācanāṃnī] N24
-\item[kārkādeśaṃ] N22
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, V5, V12, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_35c]
-\sthana{2.35c}
-\item[apa śoṣaṇī ca] Ba5,
-\item[mala śoṣiṇī va] V1,
-\item[mala śoṣaṇī ca] B2, C3, J8, LD1, N9, N17, N20, P1, P5, P10, P16, P24, W4, V16, V14, V3, V7, V18, V12
-\item[maya śoṣaṇī ca] V21, V15, V28, V13
-\item[maya śodhaṇī yā] N23
-\item[maya śodhaṇī cā] V6, 
-\item[mala śoṣaṇīva] J10, J17, P9,
-\item[mala śoṣaṇī vaḥ] J16,
-\item[doṣāmaya śoṣaṇī ca] C1, C2, C4, C6, C8, C9, J3, J4, J11, J13, J14, Ko, N12, N13, N18, N21, P2, P6, P7, P11, P12, P13, P14, P15, P17, P18, P20, P21, P23, P27, Tue, N25, V22, V2, G11, Ba3
-\item[doṣāmaya śauṣaṇī caḥ] J15, 
-\item[doṣamala śoṣaṇī ca] P8,  
-\item[doṣāmala soṣaṇi ca] N14, R1,  
-\item[doṣān mala śoṣaṇī ca] J12, N10, 
-\item[doṣāmaya śoṣaṇīya] J1,
-\item[doṣāmaya śoṣaṇīyaṃ] J2
-\item[doṣāmaya śoṣiṇī ca] A1, B3, Ba, Bo1, J6, N8, N11, N16, V19, N19, P25, 
-\item[doṣāmaya śoṣiṇāṃ ca] Bo2
-\item[doṣāmaya śodhinī ca] J7, 
-\item[doṣāmaya śodhanī ca] P26,
-\item[doṣābhayahā soṣaṇī ca] J5,
-\item[dauṣāmaya śoṣaṇī ca] LD2,
-\item[doṣāmaya goṣaṇī ca] N24
-\item[mala śoṣiṇī ca] V17, N7, N26,
-\item[doṣāṇapi śoṣaṇā ca] B1,
-\item[doṣāmaya śoṣaṇī ca] V10
-\item[doṣapra śoṣiṇī] V5
-\item[doṣānala śoṣaṇī bahava] N22
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, V20, R5, T, G1
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_35d]
-\sthana{2.35d}
-\item[mūlam iyaṃ hi] J14, P15, V1
-\item[mūlam idaṃ hi] Ko, J4, J11, N19
-\item[mau.riyaṃ hi] G1
-\item[mūlam ayaṃ ca mauliḥ] N22
-\item[maulir iyaṃ hi] A1, C1, C4, J3, J5, N8, N11, N16, P6, P20, P23, P27, V19, V21, N25, V15, V2
-\item[maulir iyaṃ] J1,
-\item[maulir iyaṃ ca] Ba, Ba5, Bo1,Bo2, C6, C8, J6, J7, N12, N13, N21, Tue, N23, P7, P11, P13, P14, P16, P18, P21, P25, V6, V7, V20, V28, V22, G11
-\item[saulir iya ca] N24
-\item[maulir ayaṃ hi] J2
-\item[sau jayatīha] J10, J17, N17, W4, N18, N20, N26
-\item[jayatīha] B1, B3, C2, C3, C9, J8, J12, J13, J15, J16, LD1, LD2, N7, N9, N10, N14, P1, P2, P5, P8, P9, P10, P12, P17, P24, P26, R1, V3
-\item[sauryakarīha?] B2
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, V16, V17, V10, V14, V5, V18, V12, V13, T, Ba3, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_38a]
-\sthana{2.38a}
-\item[prāṇāyāmena vai] Ko, J4, J11, J14, V1, N19, P15, V15
-\item[prāṇāyāmair eva] B1, B2, B3, Ba, Ba5, Bo1, Bo2, C1, C2, C4, C6, C8, J1, J3, J5, J6, J7, J12, J13, J15, LD2, N1, N3, N4a, N5, N8, N11, N12, N13, N16, N18, N22, P2, P7, P11, P12, P14, P17, P18, P20, P21, P23, P25, V3, Tue, N23, N24, N25, V6, V7, V5, V28, V22, V2, V12, V13, G11, G1
-\item[prāṇāyāmaiḥ sukhāt] A1, C3, C9, J10, J17, LD1, N2, N9, N10, N17, P1, P5, P8, P9, P10, P24, P26, R1, Ba3
-\item[prāṇāyāme sukhāt] J16,
-\item[prāṇāyāmais tu sā sarvaiḥ] P16,
-\item[prāṇāyāmaiḥ sukhāḥ] N26
-\item[prāṇāyāmair mu?khāṃ] N20
-\item[prāṇāyāmena sarvepi] N21, P6, P13, P27, 
-\item[praṇāyāme kṛte samyak] J2
-\item[prāṇāyāme śukhā] N14, V18
-\item[prāṇāyāme tataḥ] V16, V17, V10, V14, V21, V20
-\item[praṇāyāmāt sukhāt] N7
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, R5, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_38b]
-\sthana{2.38b}
-\item[malā api] J4, J11, V19, N21, P6, P13, P15, P27, V15, V28
-\item[malād api] Ko,
-\item[malā iti] Ba, Ba5, Bo1, Bo2, C1, C6, C8, J1, J2, J3, J5, J6, J7, LD2, N3, N4a, N5, N8, N11, N13, N16, P7, P11, P14, P18, P21, P23, P25, V10, N18, N19, P20, Tue, N23, N24, N25, V6, V7, V22, G11, G1
-\item[malā itiḥ] J15,
-\item[malā ime] J14, 
-\item[malāśayaḥ] J10,
-\item[malāśayāḥ] A1, B1, B3, C2, C9, J13, LD1, N1, N2, N10, N17, P5, P10, P12, P17, P24, P26, R1, V17, V14, V20, V18, V13
-\item[malākulaṃ] B2, N22
-\item[malākulāḥ] N20, N26
-\item[malāśayā] C3, C4, J16, J17, N9, P1, P2, P8, P9, P16, V16
-\item[marāśayā] N14
-\item[malāsayā] J12, V21, V3, V2
-\item[malāśrayāḥ] N7
-\item[malāyinaḥ] N12
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, P24, V5, V12, T, Ba3, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_39ab]
-\sthana{2.39ab}
-\item[brahmādayo 'pi tridaśāḥ pavanābhyāsatatparāḥ] A1, B2, B3, Ba, Bo1, C1, C4, C6, C8, C9, G1, J1, P11, P12, P13, P14, P15, P16, P18, P20, P27, P21, P23, P25, P26,
-\item[brahmādayo 'pi tṛdaśāḥ pavanābhyāsatatparāḥ] P17,
-\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparāḥ] B1, C2, 
-\item[ṣaṭkarmayogam āpnoti pavanābhyāsatatparaḥ] C3, R1,
-\item[(illegible/unavailable)] C5, P22, P24, R5, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_39cd]
-\sthana{2.39cd}
-\item[abhūvannantakabhayāt tasmat pavanam abhyaset] G1,
-\item[abhūvannantakabhayāt tasmāt pavanam abhyaset] A1, Ba, Bo1,C1, C6, C8, P11, P13, P14, P15, P18, P20a, P27,
-\item[abhūvannātankabhayāt tasmāt pavanam abhyaset] B1,
-\item[abhūvannatakabhayāṃ tasmat pavanam abhyaset] P26,
-\item[abhavannantakabhayāt tasmat pavanam abhyaset] P25,
-\item[abhūvan nanekabhayāt tasmat pavanam abhyaset] P16,
-\item[te ca siddhiṃ gatāḥ sarve tasmat pavanam abhyaset] P12,
-\item[tena siddhi gatās te ca tasmat pavanam abhyaset] P21, P23, 
-\item[tena siddhiṃ gatā yasmāt tasmāt pavanam abhyaset] C4,
-\item[tena siddhiṃ gatās te ca tasmat pavanam abhyaset] B2, C9,  J1, 
-\item[tena siddhi gatāḥ satyaḥ tasmat pavanam abhyaset] P20,
-\item[naiva siddhiṃ gatāḥ sarve tasmat pavanam abhyaset] B3, P17,
-\item[sumanskāṃtako bhavyaḥ tasmat pavanam abhyaset] R1,
-\item[samanskāṃtako bhavyaḥ tasmāt pavanam abhyaset] C2,
-\item[samanskoraho bhavyaḥ tasmāt pavanam abhyaset] C3,
-\item[(illegible/unavailable)] C5, P22, P24, R5, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_40b]
-\sthana{2.40b}
-\item[cirāśrayam] LD2,
-\item[nirāśrayam] A1, Ba, C1, C3, C6, C8, C9, J2, J4, J5, J7, J8, J10, J11, J15, J17, LD1, V1, W4, N3, N9, N10, V17, N20, N22, P1, P5, P6, P8, P9, P10, P11, P15, P20a, P24, P25, P27, V21, V3, N23, N26, R1, V6, V15, V28, V12, V13, G1
-\item[nirāśrayaḥ] J12,
-\item[niśaṃśrayaṃ] G11
-\item[nirodhā vādham] V8
-\item[nirāśayam] Bo1,
-\item[nirāsnayām] J16,
-\item[nirāmayam] B2, Bo2, C4, J1, J3, J6, J14, Ko, N17, V19, N2, N5, N7, N8, N11, N12, N14, N16, V16, N18, N19, N21, N24, N25, P7, P13, P16, P20, P21, P23, P26, V20, V5, V2, V18, Ba3,
-\item[nirākulam] B1, B3, Ba5, C2, J13, N1, N4a, N13, P2, P12, P14, P17, P18, V10, Tue, V7, V22
-\item[(illegible/unavailable)] Bo3, C5, N6, P22, P24, R5, V14, T
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_40c]
-\sthana{2.40c}
-\item[akṣir?] C8,
-\item[īkṣed] Ba, J5, P25,
-\item[īkṣet] C6, V13, G1
-\item[īkṣe] C1, J7, P7, 
-\item[īkṣye] P11, 
-\item[paśyed] A1, Bo1, C9, J4, J8, J10, J11, J14, J15, J16, J17, Ko, LD1, N17, P15, V1, W4, N9, V8, N10, P5, P8, P10, P24, P26, R1, V16, V17, N20, N22, V21, V3, N26, V15, V12, Ba3
-\item[paśyer] P1,
-\item[paśyad] J12,
-\item[paśyet] V20
-\item[paśyeta] LD2,
-\item[paśye] N19, N24
-\item[paśe] C3,
-\item[paśed] P9, 
-\item[paśyan] N7
-\item[paśca] P6, 
-\item[pasya?] N14  
-\item[dṛṣṭir] B3, Ba5, C2, C4, J6, V19, N1, N2, N4a, N11, N16, P2, P12, P14, P17, P18, P20, P23, V10, N21, N25, V7, V5
-\item[dṛṣṭi] B1, B2, Bo2, J1, J2, J3, J13, N5, N8, N12, N13, N18, P13, P21, Tue, V28, V22, V2 
-\item[vīkṣed] N3, G11
-\item[iched] N23
-\item[(illegible/unavailable)] Bo3, C5, N6, V14, V6, V18, T, P16, P22, P24, P27, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_43a]
-\sthana{2.43a}
-\item[vidhānajñā] C9, J5, J10, J11, Ko, V1, N2 ("vi" seems faded), N4a, N7, N10, N11, N12, N13, N14, N16, P6, P8, P9, V16, V17, V10, N19, N20, N21, V14, V3, Tue, N23, N26, V6, V7, V15, V22, V18, V13, G11, G1
-\item[vidhānajñāḥ] Ba3
-\item[vidhānaḥjñā] J2
-\item[vidhānajñe] V2
-\item[vidhanajñā] N18
-\item[vidhaustajñā] V20
-\item[vidhānujñaś] J1,
-\item[vidhānataś] P23, 
-\item[vidhānajñāś] Bo2, C3, C8, J4, J6, J7, J12, P5, P13, P14, P15, P24, P25, P27, 
-\item[vidhānajñaś] B1, B3, Ba, Ba5, Bo1, Bo2, C1, C2, J3, J16, N3, N24, P2, P11, P12, P17, P18, P21, 
-\item[vidhānajña] N22, V12
-\item[vidhānejñaś] N8
-\item[vidhijñās tu] C6,
-\item[vidhānajño] C4, J15, J17, LD1, LD2, N5, N17, N25, P1, P10, P20, V28
-\item[vidhānajñoś] J13, 
-\item[viddhānajño?] R1,
-\item[vividhānajño] B2, 
-\item[vidhāna?ī] N9, V8
-\item[vidhānajñaiś] N1 (unsure about "ai" matra)
-\item[vidhistajñāś] P26,
-\item[vidhiṃstajñaś] A1
-\item[vidhiṃstajñaiś] P7,
-\item[(illegible/unavailable)] Bo3, C5, N6, J14, V21, V5, T, P16, P22, P24, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_44cd]
-\sthana{2.44cd}
-\item[bhastrakā bhrāmarī vedhā kevalaṃ cāṣṭakumbhakāḥ] J2
-\item[bhastrikā bhramarikā mūrcchā kevaleṣṭā ca kumbhakā] G1
-\item[bhastrikā bhrāmarī mūrcchā kevalī cāṣṭakumbhakāḥ] V1, V16, V17, 
-\item[bhastrikā bhrāmarī mūrcchā kevala cāṣṭakumbhakāḥ] V20, V18, V12, V13, G11, Ba3
-\item[bhastrikā bhrāmarī mūrcchā vinītyaṣṭakumbhakāḥ] V10
-\item[bhastrikā bhrāmarī mūrcchā vinītyaṣṭakumbhakān] V28
-\item[bhastrikā bhrāmarī mūr??ā kevalaś cāṣṭakumbhakā] N1,
-\item[bhastrīkā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] C9, J13, J15, LD1, LD2, N21
-\item[bhastrīkā bhrāmarī mūrchā kevalaś cāṣṭakumbhakaḥ] P12, 
-\item[bhastikā bhrāmarī mūrchā kevalaś cāṣṭakumbhakaḥ] P26,
-\item[bhaṃstrikā bhramarī mūrchā kāṃbalā atha? kumbhakāḥ] J5
-\item[bhaṃstrīkā bhramarī mūrchā kevala cāṣṭakumbhakāḥ] N22, V21, V3
-\item[bhastṛ(?)kā bhrāmarī mūrcchā kevalaś cāṣṭakumbhakāḥ] N26
-\item[bhastrikā bhrāmarī mūrchā kaivalaś cā sa kumbhakaḥ] B3,
-\item[bhastrikā bhramarī mūrcchā cāṣṭakumbhakāḥ sahita?] N2 (after exchange of words' positions)
-\item[bhastrikā bhramarī mūrdhā kevalaś cāṣṭakumbhakāḥ] C2,
-\item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakāḥ] A1, B1, C3, J7pc, J10, J12, J16, J17, N3, N7, N9, N10, N17, P1, P2, P6, P9, P10, P17, P20, P27, R1, 
-\item[bhastrikā bhramarī mūrchā kevalaś caṣṭakumbhakāḥ] P5, P24, 
-\item[bhastrikā bhrāmarī mūrchā kevalaś cāṣṭakumbhakāḥ] N20, P8, 
-\item[bhastrikā marī mūrchā kevalaṃ cāṣṭakumbhakāḥ] B2,
-\item[bhastrikā bhrāmalī mūrchā kevalaś cāṣṭakumbhakā] N14
-\item[gachati tiṣṭto kāryā mujjayākhyāṃ cākumbhakāḥ] V8
-\item[bhastrikā bhramarī mūrchā kevalaś cāṣṭakumbhakā] N9
-\item[bhastrakā bhramarī mūrchā kevalāś cāṣṭakumbhakā] P13,
-\item[bhastrikā bhrāmarī mūrchā plāvanī saṣṭhakumbhakāḥ] N4a
-\item[bhastrikā bhramarī mūrcchā plāvanī tvaṣṭakumbhakāḥ] N12, V6, V7
-\item[bhastrikā bhramarī mūrcchā plāvanī tyaṣṭakumbhikā] P11, 
-\item[bhastrikā bhramarī mūrcchā pāvanīty eṣakumbhakāḥ] Ko,
-\item[bhastrikā bhramarī mūrcchā pāvanīty aṣṭakumbhakān] J4,
-\item[bhastrikā bhrāmarī mūrchā plāvinītyaṣṭakumbhakāḥ] Ba5, Bo1, C1, N13, P14, P18, V14, Tue, V15, V22
-\item[bhastrīkā bhrāmarī mūrchā plāvinītyaṣṭakumbhakāḥ] P25,
-\item[bhastrikā bhrāmarī mūrchā prāvatītyaṣṭakumbhakaḥ] N23
-\item[bhastrikā bhrāmarī mūrchā pratyahītyaṣṭakumbhakāḥ] P7,  
-\item[bhastrikā bhramarī mūrchā plāvanītyaṣṭakumbhakāḥ] Ba, C8, J7, P15,  
-\item[bhastrikā bhrāmarī mūrcchā pālāvanīty aṣṭakumbhakāḥ] J11,
-\item[bhastrikā bhramarī mūrchā pratvanītyaṣṭakumbhakāḥ] C6
-\item[bhastrikā bhrāmarī mūrchā sahita cāṣṭakumbhakā] P23,
-\item[bhastrikā bhramarī mūrchā sahitāś cāṣṭakumbhakāḥ] C4, N5
-\item[bhastikā bhrāmarī mūrchā sahitāś cāṣṭakumbhakam] N25, V5
-\item[bhastikā bhrāmarī mūrchā sahitāś cāṣṭakumbhakān] P21, 
-\item[bhastrikabhrama ma mūrchā saṃhataṃ cāṣṭakumbhakā] N8
-\item[bhastrikā bhramarī mūrchā saṃhataṃ cāṣṭakumbhakāḥ] N11(visarga added later), N16 ("saṃhitaṃ"),
-\item[bhastrikā bhramarī mūrchā saṃhitaṃ cāṣṭakumbhakāḥ] Bo2, J3, J6, V2
-\item[bhastrikā bhramarī mūrchā saṃprataṃ cāṣṭakumbhakān] N18
-\item[bhadrikā? bhramarī mūrchā saṃprataṃ rāṣṭakumbhakāṃ] J1,
-\item[bhastrikā śītalī bhramarī mūrchā plāvanītyaṣtakumbhakāḥ] N19 
-\item[sastrikā bhramarī mūrchā kavalaś cāṣṭakumbhakā] N24
-\item[(illegible/unavailable)] Bo3, C5, N6, J14, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_46]
-\sthana{2.46}
-\item[om.] V19,
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkaḥ ||] Ba5
-\item[adhastāt kuñcanādāśu kaṇṭhasaṅkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ] G1, P11, P18, 
-\item[adhastāt kuñcanādāśu kaṇṭhasaṃcane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ] P25,
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkā ||] P7, 
-\item[athastātkuñcanenāśukarṇasaṅkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ] Ba3
-\item[adhas tāt kuñcane vāyuḥ kaṃṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāmīgaḥ ||] Ko,
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkagḥ ||] V22
-\item[adhas tāt kuñcanenāsuṃ kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍīkāṃ ||] C6
-\item[adhas tā kuñcakīnālau kaṇṭhasaṃkocane kṣate | madhyaṃ paścimatāṇena syāt prāṇo brahmanāḍīkaḥ ||] J2
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] A1, Bo1, C8, J10, J11, V1, N3, N13, P14, P20, Tue, V7, V20, V15, V28, V12, G11
-\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścamatānena syāt prāṇo brahmanāḍigaḥ ||] P21, 
-\item[adhastvākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C9
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṣate | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C1,
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] C3, LD1, N12, N26
-\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] B2, J15, V10, V21
-\item[adhasthāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] J7,
-\item[adhasthāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānenya syāt prāṇo brahmanāḍigaḥ ||] P10,
-\item[adhas thāt kuñjakenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇāsthā syāt prāṇo brahmanāḍīgataḥ ||] P15, 
-\item[adhastvākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhyanāḍigaḥ ||] P1, P9, 
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J6, J17, P2,
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] Bo2, N10, V16, V17, V14, V3, V6
-\item[addhastāt kuñcanenāsu kaṃṭhasaṃkocane krate | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] R1, 
-\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] J3,
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍikaḥ ||] P17, 
-\item[adhastā kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhadhyanāḍigaḥ ||] N17,
-\item[adhastā kuñcanenāśu kaṇṭhaśaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhadhyanāḍiga ||] J12,
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāna syāt prāṇo brahmāṇḍanāḍigaḥ ||]  Ba, N4a
-\item[adhastākuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syā? prāṇo brahmanāḍikaḥ ||]  N5
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍikaḥ ||] C4
-\item[adhastākuñcaṇeṇāśu kaṇṭhaśaṇkocane kṛ(?)te | madhye paścamatānena syāt prāṇo madhye nāḍikā ||] N14
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] N1,
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] P13, P27, 
-\item[adhastāt kuñcane caiva kaṇṭhasaṃkocane krate | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] P6,
-\item[adhastāt kuñcanenaiva kaṇṭhaṃ saṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍikaḥ ||] B3
-\item[adhas tāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo madhyanāgiḍaḥ ||] J4,
-\item[adhas tāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāgiḍaḥ ||] J4,
-\item[adhas tāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhya paścimatānena syāt prāṇo madhyanāgiḍaḥ ||] P12,
-\item[adhamān kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ||] J5,
-\item[adhastāt kuñcanenaiva kaṇṭhaṃ saṃkocane kṛte | madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||] C2
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] B1
-\item[adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] N1
-\item[adhasthāt kuñcanenaiva kaṇṭhasokocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍigaḥ ||] N21
-\item[adhastāt kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanādigaḥ ||] J16, 
-\item[adhastā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhye paścimatāṇena syāt prāṇo madhyanāḍigaḥ ||] LD2,
-\item[adhastāt kuñcānenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ ||] P26,
-\item[adhaścā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhye paścimatānyena syāt prāṇo madhyanāḍigaḥ ||] P5, P24, 
-\item[aṃtastvā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] P8, 
-\item[adhastvā kuñcanenāsu kaṇṭhasaṃkocane kṛtaṃ | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] N7, V13
-\item[adhastvā kuñcanenāsu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇe brahmanāḍigaḥ ||] N25
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo madhyanāḍigaḥ ||] N16
-\item[adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhye paścimatānena syāt prāṇo madhyanāḍigaḥ ||] J1, N18, N19
-\item[adhastāt kuñcanenāsu kekatvaṃ saṃkocane krate madhye paścimatānena syāt prāṇo vasta(?)nāḍ?gaḥ ||] N9
-\item[adhastāt kuñcatenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanāḍitāḥ ||] N20
-\item[adhas tā kucanenāmaḥ kaṃcasaṃkocane kṛte | madhye paścimatānena prāṇasyā brahmanāḍīkaḥ ||] N22
-\item[adhas ta kuñcanenāśu kaṇṭhasaṃkocane kṛte | madhyapaścimatānena syāt prāṇo brahmanābhirāt ||] N23
-\item[adhas tāt kuñcanaunāśu kaṇṭhakocane kṛte | madhye paścimatāne tasyāt pāno madhyanāḍigaḥ ||] N24
-\item[adhas tāt kuñcanaunāśu kaṇṭhakocane kṛte | madhye paścimatāne tasyāt prāno madhyanāḍigaḥ ||] V18
-\item[(illegible/unavailable)] Bo3, C5, J14, N6, N2, N8, N11, V5, V2, T, P16, P22, P23, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_48b]
-\sthana{2.48b}
-\item[āsthāpya] V1, V19
-\item[sansthāpya] P26, P27, V8
-\item[utthāpya] A1, B1, B2, B3, Ba, Ba5, Bo1, Bo2, C2, C3, C4, C6, C8, C9, J1, J3, J6, J7, J8, J10, J11, J13, J17, LD1, N17, W4, N2, N3, N4a, N5, N8, N9, N11, N12, N13, N16, P1, P2, P5, P6, P7, P8, P11, P12, P13, P14, P15, P17, P18, P20, P24, P25, V16, V17, V10, N18, N21, V14, V21, V3, Tue, N23, N24, N26, V6, V7, V15, V28, V22, V18, V12, V13, G11
-\item[utthāya] P23,
-\item[uddhāpya] Ba3
-\item[ūtthāpya] J16,
-\item[uthāpya] Ko, R1, 
-\item[usthāpya] J2, J12, 
-\item[udhyāpya] N19
-\item[ur(?)thāpya] N14
-\item[utthāya] C1, N7, N20, P21, 
-\item[utprāpya] N22
-\item[uschāpya] N1, 
-\item[usthāpya] J4,
-\item[uchāpya] J5, J15, N10, P10, 
-\item[utchāpya?] LD2, P9, 
-\item[(illegible/unavailable)] Bo3, C5, J14, N6, N25, V20, V5, V2, T, G1, P16, P22, R5,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_49b]
-\sthana{2.49b}
-\item[virōdhāvadhi kumbhayet]
-\item[nīrādhāvadhi kumbhayet] N23,
-\item[nīrodhāvadhi kambhayet] Ba,
-\item[nirodhābadhi kumbhayet] C2,
-\item[nirodhāvadhi kumbhayet] A1, B1, B3, C3, C8, C9, J15, J16, J17, N4a, N10, N13, P1, P2, P5, P8, P9, P10, P11,  P12, P14, P17, P18, P24, P25, P27, Tue, V21, G11, Ba3
-\item[nirodhāccādhi kumbhayet] C6,
-\item[nirodhāvadhi kumbhakaḥ] Bo1, 
-\item[nīrodhāvadhi kumbhata] R1,
-\item[nīrodhavadhi kumbhayet] P13, 
-\item[nīrovadhi kumbhayet] P15,
-\item[nīrodhān adhikumbhayet] N20, 
-\item[nirodho parikumbhake] G1
-\item[nīrodhāvadhi kumbhayat] N3, N7, N12, P20a,   
-\item[nīrodhāvadhi kumbhavat] N21, 
-\item[nīrodhāvadhi kumbhakaṃ] J1, N2, N11,  
-\item[nīrodhāvadhi kumbhake] N18, 
-\item[nīrodhānadhi kumbhayet] N1, N26, V17
-\item[nīrodhāvahi kuṃbhayet] N9, N24,
-\item[nīrodhīvadhi kumbhayet] N19, 
-\item[nirodhaṃ kumbhayet tataḥ] J13, 
-\item[nirodhāścādhi kumbhayet] P7, 
-\item[nīrodhāvadhikam bhavet] P26,
-\item[śirodhāvadhi kumbhakaṃ] C1, C4, N5, N8, N16, P21, P23,  
-\item[śirodhāvati kumbhakān] P6,
-\item[śirodhyavadhi kumbhakaṃ] N25, P20, 
-\item[śiro ‘pya?vadhi kumbhayet] N17,
-\item[śiroyāvadti kumbhakāt] B2,
-\item[sirodhāvadhi kumbhakāt] N22,
-\item[(illegible/unavailable)] C5, J14, N6, N14, T, P16, P22, R5,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_53a]
-\sthana{2.53a}
-\item[jālodare] C3, C9,  
-\item[jālodarerā] P24, 
-\item[jalodar] A1,
-\item[jalodara] C6,
-\item[jalodarā] C1, C8, P13, P14, P15, P18, P20, P21, P23, P27, 
-\item[jalodaro] B2,
-\item[jalodāraṃ] P17,
-\item[jalodaraṃ] B1, B3, Ba, Bo1, C2, C4, J1, P25, P26, 
-\item[jālodared] R1,
-\item[(illegible/unavailable)] C5, P16, P22, R5,
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_55ab]
-\sthana{2.55ab}
-\item[śītkāṃ dadyāt sadā cakre ghraṇe caiva vijṛmbhikām]
-\item[śītkāṃ dadyāt sadā vaktre ghraṇe naiva vijṛmbhikām] P14, P18, 
-\item[śītkāṃ dadyāt sadā vaktre ghraṇe caiva vijṛmbhikām] C6, 
-\item[chikkāṃ dadyāt sadā valke ghraṇe naiva vijṛmbhikām] Bo1, 
-\item[sītkāṃ dadyāt sadā vaktre ghraṇe caiva vijṛmbhikām] P27,
-\item[sītkāṃ dadyāt sadā vaktre ghraṇe naiva visarjjayet] B2,
-\item[sātkāṃ dadyāt sadā vaktre prāṇe caiva vijṛmbhite] P13,
-\item[satkāṃ dadyāt sadā vaktre prāṇe cātha vijambhakāṃ] P15,
-\item[śītkāṃ kuryāt sadā vaktrāt ghraṇo caiva vijṛ-kām] A1,
-\item[śītkāṃ kuryāt sadā vaktrāt ghraṇe naiva vijṛmbhakān] P26,
-\item[sītkāṃ kuryāt tathā vaktre ghraṇe naiva vijṛmbhikām] C1,
-\item[sītkāṃ kuryā tathā vaktre ghraṇe naiva vivarjayet] C4,
-\item[sītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikām] P25,  
-\item[śītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikām] Ba,  
-\item[śītkāṃ kuryāt tathā vaktre ghraṇe caiva vijṛmbhikā] C8,  
-\item[śītkāṃ kuryāt yathā vaktre prāṇaṃ caiva visarjjayet] P20, 
-\item[śitkāṃ kuryāt sadā vaktre prāṇenaiva visarjayet] P21,
-\item[śitkāṃ kuryāt tathā vaktre ghrāṇenaiva visarjayet] P23, 
-\item[sitkāṃ kuryāt tayā vaktre ….nenaiva visarjayet] J1,
-\item[kuṃbhaṃ kuryāt sadā vaktre grāṇenaiva visarjayet] B1, B3, C2,  
-\item[kuṃbhaṃ kuryāt sadā vaktrāt grāṇe caiva vijṛmbhikām] C3,
-\item[kuṃbhaṃ kuryāt sadā vaktrat ghrāṇenaiva vijaṃbhitaṃ] R1, C9,  
-\item[(illegible/unavailable)] C5, P16, P22, P24, R5,
-
-\item[]
-\item[(illegible/unavailable)]
-
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_54a]
-\sthana{2.54a}
-\item[śītkaraḥ] B1, C8, P17, 
-\item[śītkaram] C6,
-\item[śīkāra] J7,
-\item[śītkāraḥ] C1, B3, J13, J15, P7, P20, P25, V17, V21 
-\item[śītakāraḥ] N9, 
-\item[śītakāra] P1, R1, 
-\item[śitakāra] P6, 
-\item[śītakārāī] LD1, 
-\item[śīghrā] P11,
-\item[śītalī] P15,  
-\item[sītkara(ḥ)] Ba, Bo2, C2, C3, C4, J3, J4, N6, 
-\item[sītkāraḥ] J6, N8, N11, N16, P2, P12, P13, 
-\item[sītkāra] N17, P5, P24, 
-\item[śītkārī] J16, 
-\item[śītakārī] P8, P26, 
-\item[śītakarī] P9,
-\item[sītkāṃ] N5, N18, N22, Tue, N24, N25, V15, V28, V22, V10, G11
-\item[sītkād] V14
-\item[sitkā] N3, N14, 
-\item[sītkā] Ko, N12, P27, 
-\item[śītkāṃ] A1
-\item[sītātkārī] N13,
-\item[sītkaraḥ] N1,
-\item[sītkāra] P23,
-\item[sītkārā] P21,
-\item[sītkarā] J17, 
-\item[sītkārī] Ba5, Bo1, J5, J12, N2, N4a, N7, N10, P14, P18,    
-\item[sītakārī] C9
-\item[sītakarākarmma] J10, 
-\item[śītikārī] LD2,
-\item[sītalī] J2
-\item[sikta?] J1,
-\item[sītkīṃ] V19,
-\item[sitkāṃ] J11, N19
-\item[sinktā?] B2
-\item[sātkāṃ] N21
-\item[dhikt(?)āṃ] N23
-\item[kumbhaṃ] V3,J8, N20,V14, V21, N26, V16, Ba3 
-\item[kumbhakaṃ] V7, V20, V18, V12, V13, 
-\item[kumbhakaḥ] V3
-\item[(illegible/unavailable)] Bo3, C5, J14, P10, V1, V6, V5, V2, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_55a]
-\sthana{2.55a}
-\item[cakram āsādya(ḥ)] A1, B3, Ba, Bo1, C2, V1, J8pc, J10, J17, N1, N7, N10, N28, V12,
- N21, P1, P2, P5, P8, P9, P12, P13, P17, P24, P25, P26, P27, V14, V20, V18, V16, Ba3
-\item[cakramāsādyamān] 
-\item[cakrasāmānya(ḥ)] B1, J8ac, J11, J13, N19, N24, P11, P15, P18, V3, V22
-\item[cakrasammānyaḥ] Ba5, N4a, N13, P14, Tue, V10, G1
-\item[cakrastamānyaṃ] J2
-\item[cakramāsādyamān] G11
-\item[cakram āsāya] N9,
-\item[cakrasāmānyaṃ] LD2
-\item[cakramādyāya] J12,
-\item[cakrasāmānya] J4, J5,
-\item[cakrasāmānyaḥ] N3, 
-\item[cakraṃ samāsādya] N6, N14, N17, W4
-\item[cakrasamāsādya] LD1,
-\item[cakranāśāya] N20, N26, P10, V17
-\item[cakranāsānyaḥ] N12, 
-\item[cakrabhogyaś ca] B2,
-\item[cakrasevyas tu] Bo2, C1, C4, C6, C8, J1, J3, J6, J7, J15, N2, N5, N8, N11, N16, P7, V19, N18, V P20, P21, P23, 21, N25, V15, V28, V2, V13, V21
-\item[cakrasevyas ta] N23
-\item[cakrabhogyas tu] N22, P6, 
-\item[cakrābhimānyaḥ] Ko,
-\item[vaktram āsādya] C3, C9
-\item[cakramadyāya] J16,
-\item[(illegible/unavailable)] Bo3, C5, J14, V6, V5, T, P16, P22, R1, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_56*1]
-\sthana{2.55*1}
-\item[satyāmūlena randhreṇa] C4, J6, J8pc, J10, J17, N5, N9, P1, P10, V21, N25
-\item[satyāmūlena randhrena] J15,
-\item[satyaṃ mūlena randhreṇa] P5, P24, 
-\item[sadāmūlena randhreṇa] J3, J6pc
-\item[satyāmūlena randhrenayaḥ] V18, V16, V21
-\item[spṛśyamūlena randhreṇa] B1, B3, C2, J13, P2, P12, P17, 
-\item[rasanātālumūlena] N6, N14, N17, 
-\item[rasanāmūla randhreṇa] C9, 
-\item[rasavatyāmūla randhre yaḥ] N20, N26
-\item[]
-\item[]
-\item[(illegible/unavailable)] A1, B2, Ba, Ba5, Bo1, Bo2, Bo3, C1, C3, C5, C6, C8, J1, J2, J4, J5, J7, J11, J12, J14, J16, Ko, LD1, LD2, N1, N2, N3, N4a, N7, N8, N10, N11, N12, N13, N16, N18, N19, N21, N22, N23, N24, P6, P7, P8, P9, P11, P13, P14, P15, P18, P20, P21, P22, P23, P25, P26, P27, P28, R1, R5, W4, V3, V6, V20, V15, V5, V28, V22, V2, V12, V13, V10, V14, V17, G11, Tue, T, Ba3, G1, P16,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_57a]
-\sthana{2.57a}
-\item[rasanātālumūlena] A1, B1, B2, B3, Bo1, C2, C4, C9, J4, J10, J12, J13, J15, J16, J17, Ko, LD1, N1, N2, N3, N6, N7, N10, N11, N12, N14, N17, N21, N22, N25, N26, P1, P2, P5, P6, P8, P9, P10, P11, P12, P13, P15, P20, P23, P24, P26, P27, R1, V1, V20, V15, V28, Ba3
-\item[rasanātālumūle yaḥ] Ba,
-\item[rasanātālumūlenaya] V14,
-\item[rasanātālumūlena yaḥ] V2, V18, V12, V3, V16, V17, G11
-\item[rasanātāgru mūlena yaḥ] G1
-\item[rasanāttālumūlena] N20
-\item[rasanātālumūlaṃ ca] J5, 
-\item[rasanātāluyogena] Bo2, C6, C8, J3, J6, J7, N8, N16, N18, P7, V21, N23, V13
-\item[rasanātālayogena] J1,
-\item[rasanā tālamūlena] N5, N9, 
-\item[rasanāṃ tālamūlena] J11
-\item[rasanāṃ tālumūlena] P25,
-\item[rasanāṃ tāluyogena] P21, 
-\item[rasanānāṃ tālamūlena] N19
-\item[jihvayā vāyum ākṛṣya] Ba5,
-\item[satyāṃ mūlena randhreṇa] C3,
-\item[]
-\item[(illegible/unavailable)] Bo3, C1, C5, J2, J14, N4a, Tue, LD2, N13, N24, P14, P17, P18, P22, V6, V5, V22, V10, T, P16, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
- \begin{marma}[hp02_58b]
-\sthana{2.58b}
-\item[pūrvavat kumbhakād anu] C4, C6, C8, J1, J4, J6pc, J7, J11, Ko, N2, N5, N8, N11, N21, P13, P20, P23, V14, V21, N3, N24, N25, P27, V6, V19, V15, V28, V2, Ba3, G1
-\item[pūrvavat kuṃbhanādan] G11
-\item[pūrvavat kūmbhakād anu] Bo2,
-\item[pūrvavat kumbhayed anu] P26,
-\item[pūrvavat kumbhanāda tu] P11,
-\item[pūrvavat kumbhakād anmuṃ] Bo1,
-\item[pūrvavat kumbhayed anu] A1, V20
-\item[pūrvavat kumbhakād anū] N19
-\item[pūrve ce kumbhanād agu] J5,
-\item[pūrvavat kumbhakādane] N23, V7
-\item[pūrvavat kumbhakeṣv anu] B2
-\item[pūrvavat kumbhasādhanaṃ] B1, B3, Ba, Ba5, Bo3, C1, C2, C3, C9, J2, J3, J6, J10, J12, J13, J15, J16, J17, N4a, LD1, LD2, N6, N9, N10, N12, N13, N14, N17, N18, N22, P2, P5, P6, P10, P12, P24, P25, R1, V3, Tue, V22, V18, V12, V13, V3, V10, V16
-\item[pūrvaṃ ca  kumbhasādhanaṃ] P1, P8, P9, 
-\item[pūrvava kumbhasādhanaṃ] N7, 
-\item[kākacaṃcusamannayāt] V17
-\item[pūrvavat kumbhasādhayet] N1, P17, 
-\item[pūrvakumbhakavad bhavet] P7,
-\item[kākacaṃcusamunnayāt] N16, N26
-\item[kākacaṃcusamunnarayāt] N20, 
-\item[kākacaṃcusamanvayāt] P21, 
-\item[(illegible/unavailable)] C5, J14, P14, P15, V5, T, P16, P18, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_62b]
-\sthana{2.62b}
-\item[kapāle ca samantataḥ] V1 
-\item[kapāla pavanaṃ tataḥ] N6, 
-\item[kapāle niḥsvanaṃ tathā] Ba,
-\item[kapole niḥsvanaṃ tathā] P25, 
-\item[kapole niḥsvanaṃ tadā] C1, 
-\item[kapole nisvanaṃ tadā] J2, 
-\item[kapole svāsanaṃ tataḥ] N26
-\item[kapole svasanaṃ tataḥ] V18
-\item[kapāla śvasamaṃ tataḥ] J4, P10, 
-\item[kapāla śasanaṃ tataḥ] J15, 
-\item[kapāle śasanaṃ tataḥ] Ba3
-\item[kapāle śvasanaṃ tataḥ] J11, N3, P20a, V17 
-\item[kapāle svaśanaṃ tataḥ] LD1, 
-\item[kapāla śparśanaṃ tataḥ] J16, N9, N10, P5, P24, 
-\item[kapole svaśanaṃ tataḥ] LD2, V16
-\item[kapole śvāsanaṃ tataḥ] N20, 
-\item[kapole svasanaṃ tataḥ] N7, P8, 
-\item[kapole śvavaṇe tathā] B2,
-\item[kapole śvavaṇe tataḥ] V13, V3
-\item[kapāle sasvanaṃ tataḥ] P13, P15, P27, 
-\item[kapole sasvanaṃ tataḥ] P9,
-\item[kapolaṃ sasvaramarut] G11
-\item[kapolaṃ sasvanaṃ tataḥ] P11, 
-\item[kapola saśvanaṃ tataḥ] J5,
-\item[kapole sca sanantataḥ] V3
-\item[kayolesvaśanaṃtataḥ] N24
-\item[kapāla pavanaṃ tata] N14, 
-\item[kapāla sasvanaṃ tataḥ] N21, V12
-\item[kapāla svasanaṃ tataḥ] N19, 
-\item[kapālasya samaṃ tataḥ] C3
-\item[kapālasyarśanaṃ tataḥ] J10, 
-\item[kapāle sasvanaṃ tataḥ] Bo1, V14
-\item[kapāle mukhano marut] Ko,
-\item[kapālādhi sasvanaṃ tataḥ] V6
-\item[kapāla pavanaṃ tataḥ] N17,
-\item[kapāle susvanam] G1
-\item[kapālasparśanaṃ tataḥ] C9, J17, P1,
-\item[kapālasya śanaṃ tataḥ] J12,
-\item[kapālādhi niḥsvanaṃ] C6,
-\item[kapālavidhi sasvanaṃ] P26,
-\item[kapālāvadhi sasvanaṃ] A1, Ba5, C8, J7, N4a, N12, N13, P7, P14, P18, Tue, V7, V20, V28, V22, V10
-\item[kapālāvadhi sasanaṃ] V21
-\item[kapālāvadhi pūrayet] B1, B3, Bo2, C2, C4, J1, J3, J6, J13, N1, N2, N5, N8, N11, N16, N18, N25, P2, P12, P17, P20, P21, P23, V15, V2
-\item[kepālavadhi sasvanaṃ] N23, 
-\item[kepālena samaṃ tataḥ] P6, 
-\item[kepālenastha sarīrasthaṃ] N22,
- \item[(illegible/unavailable)] Bo3, C5, J14, V5, T, P16, P22, R1, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_65cd]
-\sthana{2.65cd}
-\item[dhārayan nāsikāmadhyṃ aṃgulībhyāṃ tathā dṛḍham] N10,
-\item[dhārayennasikā madhye aṅgulīnāṃ tathā dṛḍham] Ba3
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ vinā dṛḍham] V1, V21
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] LD1,
-\item[dhārayen nāsikāmadhyaṃ aṅgulībhyāṃ tathā dṛḍham] A1, C9, P26, 
-\item[dhāraye nāsikāmadhyaṃ maṅgulānāṃ tathā dṛḍham] J12,
-\item[dhārayen nāsikāmadhaṃ aṅgulībhyāṃ tathā dṛḍham] V20
-\item[dhārayen nāsikāmadhya aṅgulībhyāṃ tathā dṛḍham] J10,
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] B2, C3, J17, N6, N14, N17, P5, P8, P10, P24, V18, V12, V16, V17, G1
-\item[dhāraye nāsikāmadhye aṅgulībhyāṃ tathā dṛḍham] N7, 
-\item[dhārayen nāsikāmadhyai aṅgulībhyāṃ tathā dṛḍham] J15,
-\item[dhārayen nāsikāmadhye aṅgulībhyāṃ tadā dṛḍham] N9, P1, 
-\item[dhārayen nāsīkāmadhye maṃgulībhā tathā dṛḍham] J16,
-\item[dhārayen nāsīkāmadhye aṅgulībhyāṃ dṛḍham tathā] LD2,
-\item[dhārayen nāsikāmadhye aṅguṣṭhābhyāṃ tathā dṛḍham] N20, N26,
-\item[dhārayen nāsikāṃ madhye aṅgulībhyāṃ tathā dṛḍham] P9, 
-\item[dhārayen nāsikābhyāṃatarjabhyāṃ tathā dṛḍham] R1,
-\item[dhārayen nāsikāmadhyṃ tarjanībhyāṃ vinā dṛḍham] Ko,
-\item[dhārayen nāsikāmadhyaṃ tarjanībhyāṃ vinā dṛḍham] J11, N3, 
-\item[dhārayen nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] C6, J5, N12, N18, N19, P6, V3, N21, N24, P13, P17, P25, P27, V6, V7, V28, V22, V2, V3, V14, G11
-\item[dhārayen nāsikāmadhyaṃ tarjanībhyāṃ vinā dṛḍham] P7,
-\item[dhāraye nāsikāmadhye tarjanībhyāṃ vinā dṛḍham] Ba,
-\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] Ba5, Bo2, J3, N11, N16, P11, P14, P20, 
-\item[dhārayen nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] P23,
-\item[dhāraye nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] P18,
-\item[dhāraye nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] Ba5, N4a, N13, V15
-\item[dhāraye nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] N8, 
-\item[dhārayen nāsikāṃ madhyā tarjjanībhyāṃ vinā dṛḍham] J7, P21, 
-\item[dhāvayen nāsikāmadhyāṃ tarjanībhyāṃ vinā dṛḍham] P15,
-\item[dhāraye nāsikāṃ madhye tarjjanībhyāṃ vinā dṛḍham] Bo1,
-\item[dhārayen nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] B1, B3, J13, N1, N2, P2, P12, V13
-\item[dhārayan nāsikāmadhye tarjanībhyāṃ tathā dṛḍham] C2, 
-\item[dhārayen nāsikāmadhya tarjjanībhyāṃ vinā dṛḍhām] J1,
-\item[dhārayen nāsikā madhye tarjanībhyāṃ vinā madam] J4,
-\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ yathā dṛḍham] C8
-\item[dhārayen nāśikāmadhye sarjanībhyāṃ yathā dṛḍham] N22,
-\item[dhārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] J6pc,
-\item[kārayen nāsikāṃ madhya tarjjanībhyāṃ vinā dṛḍham] J6,
-\item[dhārayen nāsikāṃ madhyāt tarjanībhyāṃ vinā dṛḍham] Tue
-\item[dhārayen nāsikāmadhyāt tarjanībhyāṃ vinā dṛḍham] C4, N5, N25, V10
-\item[dhārayaṃ nāśikāmadhye tarjanabhyāṃ vinā dṛḍham] N23
-\item[dhāraae vāsikāmadhye tarjanībhyāṃ vinā dṛḍham] J2,
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_67ab]
-\sthana{2.67ab}
-\item[kaṇdalībodhakaṃ viprabhavaghnaṃ sukhadaṃ śubhaṃ] V1,
-\item[kuṇḍalī bodhakaḥ kubho rogaghnaḥ sukhada śubhaḥ] Ba3
-\item[kuṇdalībodhanaṃ sarva doṣaghnaṃ sukhadaṃ śivaṃ] Bo1
-\item[kuṇdalībodhanaṃ sarva doṣaghnaṃ sukhadaṃ śuvaṃ] V15
-\item[kuṇdalībodhanaṃ sarva doṣa?? sūkhadaṃ sūkhaṃ] N24,
-\item[kaṇdalībodhakaṃ ksipraṃ pavanaṃ sukhadaṃ hitaṃ] Tue, V7
-\item[kuṇḍalo bodhaka kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] J2,
-\item[kuṇḍalībodhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] Ba5, N13, P14, P18, V22
-\item[kuṇḍalabodhakaṃ pāpāghnaṃ suṣadaṃ śubhaṃ] Ko,
-\item[kaṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] V6, V2
-\item[kaṇḍalībodhanaṃ cakaṃ pāpaghnaṃ sukhada śubhaṃ] V14
-\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] B2, Bo2, C4, J1, J3, J6, J7, J11, N16, N19, N23, N25, P6, P21, P23, P27, 
-\item[kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ śubhadaṃ sukhaṃ] P20,
-\item[kuḍalībodhanaṃ kuryāt pāpaghna suṣadaṃ subhaṃ] N18, 
-\item[kuṇḍaṃlībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] N8, N11, 
-\item[kuṇḍalabodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubhaṃ] N5, 
-\item[kaṇḍalībodhanaṃ kuryāt pāpaghnaṃ subhaṃ tad śukhaṃ] V28
-\item[kuṇdalībādhakaṃ ksipraṃ pāpghnaṃ sukhadaṃ hitaṃ] Ba,
-\item[kuṇdalībodhakaṃ ksipraṃ pāpaghnaṃ sukhadaṃ hitaṃ] P25, 
-\item[kumḍalīśhodhanaṃ kuyāt pāpaghnaṃ ca sukhaṃ śhubhaṃ] N22, 
-\item[kuṇdalībādhakaṃ kṣipraṃ pavanaṃ sukhadaṃ hitaṃ] N4a,  
-\item[kaṇḍalībodhanaṃ kuryāt tāpaghnaṃ sukhadaṃ śubhaṃ] N2, 
-\item[kuṇḍalībodhakaḥ kuṃbhaḥ rogaghnaḥ sukhadaḥ śubhaḥ] N6,
-\item[kuḍalībodhaka kuṃbhao rogaghnaḥ sukhadaḥ śubhaḥ] P5, P8, P9, P24, 
-\item[kuḍalībodhanaḥ kuṃbhao rogaghnaḥ sukhadaḥ śubhaḥ] P7,
-\item[kuṇḍalībodhakaḥ kuṃbho rogaghnaḥ sukhadaṃ śubhaḥ] N9, N10, N17, N20,  N26, P1, P10, 
-\item[kuḍalībodhakaḥ kuṃbhao rogaghnaṃ sukhadaṃ śubhaṃ] P17,
-\item[kuṇḍalībodhakaḥ karthūḥ roghnaṃḥ sukhadaḥ śubhaḥ] A1, V20
-\item[kuṇḍalībodhaka kumbhao rogaghnaṃ sukhadaḥ śubham] N14, P2,
-\item[kuṇḍalībodhakaṃ cakraṃ bhāvaghnam sukhadaṃ śubhaṃ] N3,
-\item[kuṇḍalībodhanaṃ cakra bhāvaghnam sukhadaṃ śubhaṃ] P11,  
-\item[kuṇḍalībodhakaṃ kṣipram bhavaghnam sukhadaṃ śubhaṃ] N12,
-\item[kuṇḍalībodhakaṃ vipram bhavaghnam sukhadaṃ śubhaṃ] P15,
-\item[kuṇḍalībodhanaṃ karttur bhavaghnam sukhadaṃ śubhaṃ] C6,
-\item[kuṇḍalībodhanaṃ karttur bhavaghnam sukhaṃ śubhaṃ] V21
-\item[kuṇḍalībodhakaṃ miśrabhavaghnam sukhadaṃ śubhaṃ] J4,
-\item[kuṇḍalībodhanaṃ kartuṃ bhavaghnam sukhadaṃ śubhadaṃ] C8,
-\item[kuṇḍalībodhakaḥ kartuḥ rogaghnasukhadaḥ śubhaṃ] P26, 
-\item[kuṇḍalībodhakaḥ kumbhoḥ ragaghnaḥ sukhadaḥ śubhaḥ] N7, 
-\item[kuṇḍalībodhakaḥ kumbho roghnaḥ sukhadaḥ śubhaḥ] C3, J10, J17, 
-\item[kuṇḍalībodhakaṃ kumbhoḥ roghnaḥ sukhadaḥ śubhaḥ] J15, J16,
-\item[kuṇḍalībodhakaḥ kumbho roghnaḥ suṣadaḥ śubhaḥ] C9, 
-\item[kuṇḍalībodhakaḥ kumbho rogaghnaḥ sukhadaḥ śubhaḥ] J12, LD1, LD2, V16 
-\item[kuṇḍalībodhakaḥ kumbho rogaghnaḥ sukhadaḥ śubhaḥ] B1, B3, C2, J13, V18, V12, V13, V3, V17 
-\item[kuṇḍalībodhakaḥ kumbho rogaghnaṃ sukhadaṃ śubhaṃ] P12,
-\item[kuṇḍali bodhakaṃ vakrabhāvaghnaṃ sukhadaṃ śubhaṃ] G11
-\item[kuṇḍalībodhakaṃ sarva doṣaghnaṃ sukhadaṃ hitaṃ] N21, P13, 
-\item[kuṇḍalībodhakaḥ kumbho rogaghnaṃ sukhadaṃ śubhaṃ] N1, 
-\item[kuṇḍalīno bodhacakraṃ? bhāvadaṃ supada śubhāṃ] J5,
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, V5, V10, T, G1, P16, P22, R1, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_68a]
-\sthana{2.68a}
-\item[guṇatrayasamudbhūtaṃ] Bo1,
-\item[guṇatrayasamudbhūta] N21,
-\item[samagātrasamudbhūtaṃ] J4, N7, 
-\item[samagātrasamā....] P10, 
-\item[samyagātrasamudbhūtaṃ] J7,
-\item[samyagātrasamudbhutaṃ] J15,
-\item[samyagātrasamudbhuta] P14,
-\item[samyaggātrasamudbhūtaṃ] B2, Bo2, C4, C8, J2, J11, N3, N9, N16, N19, P13, V1, V7, V2, V3, V17
-\item[samyagatrasamudbhutaṃ] N12,
-\item[samyaggātrasamudbhūta] N2, N4a, N11, N13, N20, N23, N26, P15, P18, P25, P27, Tue, 
-\item[samyagātrā sumuvbhutaḥ?] N24,
-\item[samya gātre samudbhūtaṃ] J5,
-\item[samyaggātre samudbhūtaṃ] J6, N5, N8, N25, P20, P23, V28
-\item[samyaggātre samudbhūta] J3,
-\item[samyaggātre samudbhūtat] P6, 
-\item[samyag gātraṃ samudbhūtaṃ] Ko,
-\item[samyaggātasamudbhūtam] V20, V15
-\item[samyagataḥ samudbhūtam] V14
-\item[samyak gātrasamudbhūta] Ba, V22, G11, P11, 
-\item[samyag antaḥ samudbhū] A1
-\item[samyag antaḥ samudbhūta] C6
-\item[samyak bhastrā samudbhūtā] B1, B3, C2, J12, J13, J16, N1, N10, P1, P2, P8, P9, P12, P17, Ba3
-\item[samyak bhastrā samudbhūtaḥ] P26, V21  
-\item[samyak bhastrā samabhyāsaḥ] V13
-\item[samyag bhastrā samudbhūtāṃ] LD1, LD2, N18, P7, P21, 
-\item[samyak bhastrā samudbhūto] C3, J10, J17, N14, N17, P5, P24, 
-\item[samyag bhastri samudbhūtaṃ] C9
-\item[samyag bhastrī samudbhūto] N6, 
-\item[sasyagātrasamudbhūtaṃ] J1,
-\item[samagātrasamudbhūtaṃ] V6, V16
-\item[samagātre samudbhūta] N22,
-\item[samagātrasamudbhūtaghnaṃ] V18
-\item[(illegible/unavailable)] Ba5, Bo3, C1, C5, J14, V5, V12, V10, T, G1, P16, P22, R1, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_69a]
-\sthana{2.69a}
-\item[vegair ghoṣaṃ] V1
-\item[vegākṛṣṭaṃ] Ba, J2, N21, P25, V15
-\item[vegākṛṣṇaṃ] P13,
-\item[vegā gho] N23,
-\item[vego tad ghoṣa] J1,
-\item[vego ghoṣam] C3, C8, J16, N3, N9, N16, N20, P1, P5, P10, P24, P27, 
-\item[vego ghoṣa] N24,
-\item[vego ghoṣaḥ] J15,
-\item[vegāt ghoṣaṃ] A1, B2, B3, Ba5, C9, N4a, LD1, N13, N26, P14, P17, P18, P26, R1, Tue, V6, V7, V22, V18, V12, V13, V14, V16, V17, Ba3
-\item[vegad ghoṣa] N10,
-\item[vegad ghoṣaṃ] N7, 
-\item[vegā ghoṣaṃ] P8, V3
-\item[vegot ghoṣaṃ] J12, N14, N17, P11, V2, G11
-\item[vegot ghoṣe] V2,
-\item[vegodveṣaḥ] J4,
-\item[vegodvegaṃ] P20a,
-\item[vegodveṣaṃ] N5, P15, P20, 
-\item[vaiṣodvaiṣaṃ] J5,
-\item[veget vege] V28
-\item[vegād ghoṣaṃ] J10, J17, N1, V20
-\item[vegod ghoṣaṃ] Bo2, C6, J10pc, J11, N18, P6, P9, P21, 
-\item[vegod ghoraṃ] P23, 
-\item[vedod ghoṣaṃ] V21
-\item[vedyo ghoṣaṃ] N22,
-\item[vegodyoṣaṃ] J7,
-\item[vegopyoyaṃ] N19, 
-\item[vemod ghoṣaṃ] J3, J6, 
-\item[vegodvegaṃ] C4, J13, N25, 
-\item[vegod deyaṃ] P7, 
-\item[vegād aghoṣaṃ] B1,
-\item[vegād doṣaṃ] Bo1, N2, 
-\item[vegod doṣaṃ] N6, 
-\item[veṇu doṣaṃ] C2,
-\item[veṇu ghoṣaṃ] P2,
-\item[veṇud ghoṣaṃ] P12, 
-\item[(illegible/unavailable)] Bo3, C1, C5, J14, Ko, N8, N11, N12, V5, V10, T, G1, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_73a]
-\sthana{2.73a}
-\item[ārecyāpūrya yat kuryāt] N2, N5, 
-\item[ārecya pūryaṃ yat kuryāt] C4,
-\item[ārecya āpūrya yat kuryāt] P23,
-\item[ārecyāpūrya ?????] N11, 
-\item[ārecyāpūrya vat kuryāt] J13,
-\item[ārecyāpūrvavat kuryāt] N25,
-\item[caraikaṃ? pūrakaṃ kuryāt] C2,
-\item[recakaṃ pūrakaṃ kuryāt] N1, N10, P2, 
-\item[recya vā pūrya sanakaiḥ] J11,
-\item[recake pūrakaṃ kuryāt] J12, J16, 
-\item[recayet pūrakaṃ kuryāt] J1, N16, N19, N23, P6, 
-\item[recakaṃ pūrakaṃ kuryāt] P12, P17, 
-\item[ārecyāpūrya kuṃ?yāṃt] N8,
-\item[recakaḥ pūrakaṣ kāryyaḥ] J17,
-\item[recakaḥ pūrakaḥ kāryaḥ] A1, C3, C9, N9, J15, N17, N26, P1, P5, P9, P24, R1, 
-\item[recakaḥ pūrakaḥ kuryāt] P21, 
-\item[recanaṃ pūraṇam kuryāt] P13,
-\item[recakaḥ pūrakaḥ kā....] P10,
-\item[recakaḥ pūrakaḥ kāryāḥ] Bo1, N7, 
-\item[recakaḥ pūrakaḥ kāryaṃ] N20
-\item[recakaḥ pūrakaḥ kāryo] P8, 
-\item[recakaḥ pūrvavat kāryaḥ] N3, 
-\item[recakaṃ pūrakaṃ kuryāt] B1, B3, 
-\item[ārecaṃ pūrvavat kuryāt] P20,
-\item[recayitvā pūrakaḥ kāryaḥ] J14, 
-\item[reco vā pūrakaḥ kāryaḥ] P11, 
-\item[rocyatvā pūrakaḥ kuṃbhakaḥ] N24,
-\item[recakaḥ pūrakaṃ tyatvā] N18, 
-\item[recakaṃ pūrakaṃ muktā] N22,
-\item[recake pūrakaṃ uktwā] V14, G11, Ba3
-\item[recake pūrṇaṃ kāryaṃ] Ba,
-\item[recake pūraṇaṃ kāryaṃ] P25,
-\item[recanaṃ pūranaṃ kuryāt] N21,
-\item[recaka pūraka kumbhai] V3, V16
-\item[recaka pūraka kumbhakaiḥ] V17
-\item[recakaṃ pūrakaṃ muktyā] G1
-\item[recaka pūraka varjite] V10
-\item[virecya pūrakaḥ kāryaḥ] C6, C8, P7,
-\item[prāṇāyāmoyam ity uktaḥ] N4a, N12, P14, P18, Tue,
-\item[(illegible/unavailable)] B2, C1, C5, N6, N13, N14, P15, T, P16, P22, P26, P27, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_73b]
-\sthana{2.73b}
-\item[śanaiḥ saṃhitakumbhakaḥ] N16, 
-\item[sarve kevalakumbhakam] G1
-\item[śanaiḥ sahitakumbhakaḥ] Bo1, C6, C8, J1, N23, P7, 
-\item[śanaiḥ sahitakumbhakaiḥ] P11,
-\item[śanaiḥ sahitakumbhakaṃ] P6, 
-\item[sa vai kevalakumbhakaḥ] N4a, N18, N22, P18, P26, Tue, 
-\item[sa vai sahitakumbhakaḥ] A1, B1, B2, B3, C2, C4, J12, J13, J15, N1, N3, N5, N10, N17, N21, P2, P8, P12, P13, P17, P20, P21, P23, P27, 
-\item[sarve sahitakumbhakaḥ] J17, N20, 
-\item[savyai sahitakumbhakaḥ] N19, 
-\item[sarvai sahitakumbhakaḥ] N9, N12, N26, P1, P9, P10, R1, 
-\item[sarvaiḥ sahitakumbhakaḥ] C3, C9, 
-\item[sarvaiḥ sahitakumbhakam] Ba, P25,
-\item[sarvai sā?ti kumbhakaṃ] N25,
-\item[sarvai kevalakumbhaka] N14,
-\item[sarvair ahitakumbhakaḥ] N7, 
-\item[sa vai sahitakumbhak] J16, 
-\item[sa taiḥ sahitakumbhakaḥ] J14, N24, 
-\item[sa vai kevalakumbhakaḥ] Ba3, G11, P14, V10
-\item[sarvaiḥ sahitapūrakaḥ] P5, P24, 
-\item[kāryaḥ sahitakumbhakaḥ] J11,
-\item[kuryat sahitakumbhakaḥ] N2, N8, 
-\item[kuryat saṃhitakumbhakaḥ] N11, 
-\item[kāryaḥ śanai sahitakumbhakā] V3, V14
-\item[kāryaḥ śarveiḥ sahitakumbhakeiḥ] V16
-\item[kāryaḥ śarveiḥ sahitakumbhaka] V17
-\item[(illegible/unavailable)] C1, C5, N6, N13, P15, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\begin{marma}[hp02_78a]
-\sthana{2.78}
-\item[kumhakaḥ prāṇarecānte] C3, J2, J3, J13, N8, N11, N16, N21, P5, P24,V1, V6, V28, V18, V16, G11,
-\item[kumkāt prāṇarecānte] N20, N26, V17
-\item[kumbhaka prāṇarecānte] A1, B1, B2, Bo2, C2, J1, J5, LD1, N2, N7, N9, N10, N12, N14, N17, N22, N25, P1, P2, P9, P10, P12, P13, P17, P20, P21, P23, V2, V12, Ba3
-\item[kumbhaka prāṇarecyaṃte] N1,
-\item[kumbhake siddha reca puraka] V21
-\item[kumbhaka prāṇa cātte?] J10, J17, 
-\item[kumbhaka prāṇa recāttaḥ] J16,
-\item[kumbhaka prāṇarecātte] P8, P26,
-\item[kumbhakaṃ prāṇarecānte] B3, C4, C9, J6, N5, N19, P6, P27, G1, R1, 
-\item[kumbhakaṃ prāṇarecātteḥ] J15,
-\item[kumbhakaṃ prāṇareṃcānte] J12,
-\item[kumbhataḥ prāṇarecānte] J7,
-\item[kumbhata prāṇarecānte] V3
-\item[kumbhakaṃ ghrāṇarecānte] LD2, 
-\item[kumbhitaḥ prāṇarecānte] Bo1, N23, N24, P11, P15, V15
-\item[kumbhītaḥ prāṇarecānte] Ko, J11, J14, N3,  
-\item[kumbhīte prāṇarecānte] V14
-\item[kumbhītaprāṇareṣānte] J4,
-\item[tad etat prāṇarrecānte] C6, C8, P7, 
-\item[kumbhakaḥ prāṇarodhānte] Ba, Ba5, P14, P18, P25, Tue, V7, V20, V22, V10
-\item[kumbhaka prāṇarodhānte] N4a, N13,
-\item[(illegible/unavailable)] Bo3, C1, C5, N6, N18, V5, V13, T, P16, P22, R5,
-  \begin{description}
-
-    \end{description}
- \end{marma}
-
-\end{ekdosis}
-\end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:
-
     • Verses are found till 42ab in Bo1 ms

### 2023-09-02 compile-tidyup.sh

@@ -10,13 +10,13 @@ xslcmdproc(){
 	    chapid="hp1" \
 	    transl="../xml/HP1_TranslComm-tei.xml" \
 	    marma="../xml/Marmasthanas-tei.xml" \
-	    jyotsna="../xml/Jyotsna-tei.xml"
+	    jyotsna="../xml/jyotsna.xml"
     
     $xslcmd -s:xml/HP2_edition-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp2.html \
 	    chapid="hp2" \
 	    transl="../xml/HP2_TranslComm-tei.xml" \
 	    marma="../xml/Marmasthanas-tei.xml" \
-	    jyotsna="../xml/Jyotsna-tei.xml"
+	    jyotsna="../xml/jyotsna.xml"
 
     echo "concatenating html …"
     sed '/<!--content-->/q' html/meta.html > html/hp-online.html 
@@ -58,6 +58,34 @@ compile(){
     popd || exit
 }
 
+xmlizejyotsna(){
+    echo "<TEI xmlns=\"http://www.tei-c.org/ns/1.0\"><teiHeader/><body><text>" > xml/jyotsna.xml
+
+    sed -e '1,/\\startlinenumbering/ {d}' \
+    -e '/\\begin{vsid}{hp03_001}/,$ {d}' \
+    -e 's_%*\\begin{vsid}{\(.*\)}_<note type="jyotsna" target="#\1">_' \
+    -e 's_%*\\end{vsid}_</note>_' \
+    -e 's_%.*__' \
+    -e 's_\s\+$__g' \
+    latex/Jyotsna.tex |
+    tr '\n' '\0' |
+    sed -e 's_\\ \+__g' \
+	-e 's_\\-__g' \
+	-e 's_\\blank__g' \
+	-e 's_\\[a-z]*{[^{}]*}__g' \
+	-e 's_{[^{}]\+}__g' \
+	-e 's_\\[a-z]*{[^{}]*}__g' |
+    tr '\0' '\n' |
+    sed -e '/{\\bf/,/)}/ {d}' \
+	-e '/\\startsloka/,/\\stopsloka/ {
+	s_\\startsloka_<lg>_
+       	s_\\stopsloka_</lg>_
+       	s_\(^[^<].\+\)_<l>\1</l>_
+	}' >> xml/jyotsna.xml
+
+    echo "</text></body></TEI>" >> xml/jyotsna.xml
+}
+
 
[…]

### 2023-09-02 marma/print-uncollated-sigla.sh

@@ -1,5 +1,5 @@
 #!/bin/sh
-marma_tex=Marmasthāna.tex
+marma_tex=marma.tex
 sig_txt=Sigla-Liste.txt
 
 tsvmarma() { 

### 2023-09-02 compile-tidyup.sh
refined jyotsna routine

@@ -62,8 +62,8 @@ xmlizejyotsna(){
     echo "<TEI xmlns=\"http://www.tei-c.org/ns/1.0\"><teiHeader/><body><text>" > xml/jyotsna.xml
 
     sed -e '1,/\\startlinenumbering/ {d}' \
-    -e '/\\begin{vsid}{hp03_001}/,$ {d}' \
-    -e 's_%*\\begin{vsid}{\(.*\)}_<note type="jyotsna" target="#\1">_' \
+    -e '/\\begin{vsid}{#hp03_001}/,$ {d}' \
+    -e 's_%*\\begin{vsid}{\(.*\)}_<note type="jyotsna" target="\1">_' \
     -e 's_%*\\end{vsid}_</note>_' \
     -e 's_%.*__' \
     -e 's_\s\+$__g' \

### 2023-10-25 latex/HP1_TranslComm.tex
updated

@@
 \begin{translation}[hp01_001]
-Homage to the glorious Ādinātha by whom the system of Haṭhayoga was taught. It shines forth like a ladder for one desirous of climbing to the lofty terrace of the royal palace.
+Homage to the glorious Ādinātha who taught the system of Haṭhayoga, which shines forth like a stairway for one desirous of climbing to the lofty terrace of the royal palace.%??JM: would much prefer "science" to "system" (vidyā doesn't really mean system either).
+I changed "ladder" to "stairway" (a bit more poetic).
+I'd also prefer "which is like a glorious stairway for one..."
 \end{translation}
@@
 \begin{testimonia}[hp01_001]
-Cf. Yogasārasaṃgraha, p. 54.
+Cf. \emph{Yogasārasaṅgraha}, p. 54.
 
@@
 
-Gheraṇḍasaṃhitā 1.1
+\emph{Gheraṇḍasaṃhitā} 1.1
 
@@
 128–130.
-Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison
-(\emph{iva}), (2) the object compared (\emph{upameya}), i.e., the \emph{haṭhayogavidyā} leading to Rājayoga (\emph{rājayogaprāpikā}), (3) the property compared, that is, effortlessness (\emph{anāyāsena}) and, most importantly, (4) an image to compare with, which in this case is \emph{rājasaudha}.
-This idea is confirmed by Brahmānanda, when he sums up and completes the poetical details as: \emph{yathā pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudhaprāpikā bhavati evaṃ haṭhadīpikāpi pronnatarājayogam āroḍhum icchor anāyāsena rājayogaprāpikā bhavatīti upamālaṅkāraḥ}.
-
+Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison ??
+JM: what's happened here?
+ALso, do we need to repeat what Jürgen has already written elsewhere?
 The Sanskrit poeticians explain that when some elements of a comparison remains unexpressed we get an incomplete (\emph{lupta}) comparison.
@@
 But from the perspective of \emph{alaṃkāraśāstra} the verse has a problem that has surely caused the dilemma: it either leaves the \emph{upameya} or the \emph{upamāna} incomplete.
-Brahmānanda has mentioned both \emph{yathā adhirohiṇī saudhaprāpikā bhavati evaṃ haṭhadīpikā rājayogaprāpikā bhavati}, but had to supply \emph{rājasaudha}, while his text reads \emph{rājayoga}.
+Brahmānanda has mentioned both ??
+\emph{yathā adhirohiṇī saudhaprāpikā bhavati evaṃ haṭhadīpikā rājayogaprāpikā bhavati}, but had to supply \emph{rājasaudha} because his text reads \emph{rājayoga}.
 But to have the \emph{upameya} in the text is odd.
@@
 The metre of 1.1 is Indravajrā
+
+% JB Oct 11: It might be helpful to translate Brahmānanda's comment for the non-Sanskrit reader: yathā pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudhaprāpikā bhavati evaṃ haṭhadīpikāpi pronnatarājayogam āroḍhum icchor anāyāsena rājayogaprāpikā bhavatīti upamālaṅkāraḥ
+"Just like one, who wishes to climb a lofty terrace, reaches the terrace without effort by a
+staircase, in the same way, one, who wishes to climb the lofty royal Yoga, reaches the royal Yoga
+by means of the Haṭhapradīpikā. This is the (poetic figure called) \emph{comparison}."
+
+% JB: [...] but had to supply \emph{rājasaudha} because his text reads \emph{rājayoga}.
+% JB: Even Brahmānanda could only know of the image of the upper terrace of the palace (\emph{rājasaudha})
 \end{philcomm}
@@
 \begin{philcomm}[hp01_002]
-The reading \emph{yogopadiśyate} (J7,J10,N17,W4, etc.) is only possible if one accepts that double \emph{sandhi} is a feature of the style of composition, which it is not.
+The reading \emph{yogopadiśyate} (\getsiglum{J7,J10,N17,W4}, etc.) is only possible if one accepts that double \emph{sandhi} is a feature of the style of composition, which it is not.
 Moreover, this reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.
@@
 \begin{testimonia}[hp01_003]
-Haṭharatnāvalī 1.4
+\emph{Haṭharatnāvalī} 1.4
 
@@
 \tl{bhrāntyā bahumatadhvānte rājayogam ajānatām |\\+}
-\tl{kevalaṃ rājayogāya haṭhavidyopadiśyate || \\!}
+\tl{kevalaṃ rājayogāya haṭhavidyopadiśyate ||\\+}
+\tl{\var{rājayogam ] rājamārgam P,T,t1}\\!}
 \end{versinnote}
@@
 Most witnesses have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b.
-While this reading is well attested by the manuscripts transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note that P,T,t1 of the critical edition [Gharote 2009: 5 n.
-2] have \emph{rājamārgam ajānataḥ}), the reading of a royal path (\emph{rājamārgam}) is consistent with the metaphor of people wandering from a path in the darkness and, therefore, most probably authorial.
-
-In 1.1d both \emph{kṛpākaraḥ} and \emph{kṣamākaraḥ} are attested.
-The former is preserved by the V1 and group 2 whereas the latter by some manuscripts of the delta group.
-Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākara} makes better sense.
+While this reading is well attested by the manuscript transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note the readings of P,T,t1 of the critical edition [Gharote 2009: 5 n.
+2] in the testimonia), the reading of a royal path (\emph{rājamārgam}) is consistent with the metaphor of people wandering from a path in the darkness and, therefore, most probably authorial.
+
+In 1.1d \emph{kṛpākaraḥ} is attested by the most important groups, including alpha, of the available manuscripts.
+The readings \emph{prakāśyate} and \emph{kṣamākaraḥ} are attested by some manuscripts in lower branches of the stemma.
+Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākara} makes good sense.
 As Brahmānanda notes, this compound can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}).
-In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} may have arisen as a mistake for \emph{kṛ}.
+In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} probably arose as a mistake for \emph{kṛ}.
 \end{philcomm}
@@
 \begin{testimonia}[hp01_004]
-Haṭharatnāvali 1.3 (Post-15th Century)
+\emph{Haṭharatnāvalī} 1.3
 
@@
 \begin{philcomm}[hp01_004]   
-The word \emph{athavā} (‘or’) is difficult to construe here.
+The word \emph{athavā} (‘or’) is well attested but difficult to construe here.
 Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it.
-The variant \emph{mahāyogī} in group 4c (G5, J4, J11, Ko, P15) avoids this difficulty, but is probably an attempt to remove the difficulty of understanding \emph{athavā}.
-One could emend to \emph{tathā} but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.
+The variant \emph{mahāyogī} in epsilon one and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}.
+One could emend to \emph{tathā} in light of the attested reading \emph{yathā} (\getsiglum{C7}) but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.
 \end{philcomm}
@@
 \begin{testimonia}[hp01_005]
-Haṭharatnāvalī 1.80
+\emph{Haṭharatnāvalī} 1.80
 
@@
 In Śaiva texts which predate the haṭha corpus, Mīnanātha and Matsyendra are one and the same, but they are differentiated in later Tibetan and Indian lists of siddhas (Mallinson 2019:273 n.35).
+
+Two manuscripts of the alpha and delta groups have the variant reading \emph{°virūpākṣaḥ savālikaḥ} (\getsiglum{N3,J5,V19}) for \emph{°virūpākṣabileśayāḥ}.
+In \getsiglum{N3}, \emph{savālikaḥ} was corrected to \emph{savālmikaḥ}, perhaps in an effort to restore a name similar to Vālmīki, the celebrated author of the \textit{Rāmāyaṇa}.
 \end{philcomm}
@@
 \begin{translation}[hp01_006]
-Manthānabhairava, Siddhabuddha, and Kanthaḍi, Koraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
+Manthānabhairava, Siddhabuddha, and Kanthaḍi, Goraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
 \end{translation}
@@
 \begin{testimonia}[hp01_006]
-Haṭharatnāvalī 1.81
+\emph{Haṭharatnāvalī} 1.81
 
@@
 \tl{manthānabhairavo yogī siddhabuddhaś ca kandalī |\\+}
-\tl{korandakaḥ surānandaḥ siddhipādaś ca carpaṭī ||\\!}
-\end{versinnote}
-
-Caturbhuja Misra's Mugdhāvabodhinī (1.7.8) on the Rasahṛdayatantra
+\tl{korandakaḥ surānandaḥ siddhipādaś ca carpaṭī ||\\+}
+\tl{\var{korandakaḥ ] gonandaka P,T,J,n1,n4}\\!}
+\end{versinnote}
+
+Caturbhuja Misra's \emph{Mugdhāvabodhinī} (1.7.8) on the \emph{Rasahṛdayatantra}
 
@@
 \begin{philcomm}[hp01_006]        
-Many manuscripts preserve the reading \emph{pauraṇṭaka}.
-We have accepted \emph{koraṇṭaka} because this name is associated with the \emph{Haṭhābhyāsapaddhati}.
-This reading is well attested by various manuscripts of the \emph{Haṭhapradīpikā}, including those of group 2, as well as those of the \emph{Haṭharatnāvalī} (Gharote 2009: 35 n.
-3), which also attest \emph{gonandaka}.
-Also, \emph{Goraṇṭakuḍu} is the name of a disciple of Gorakṣanātha in the \emph{Navanāthacaritramu} (Jones 2017:194 n.3).
-The compound \emph{siddhapāda} could be a respectful affix.
-However, it seems unlikely here because it would cross the \emph{pāda} break.
+The alpha manuscripts have \emph{goraṇṭaka}, and several other manuscript groups have \emph{pauraṇṭaka}.
+We are yet to find the name \emph{goraṇṭaka} in other Sanskrit texts but it may be a Sanskrit rendering of \emph{Goraṇṭakuḍu}, which is the name of a disciple of Gorakṣanātha in the \emph{Navanāthacaritramu} (Jones 2017: 194 n.3).
+The spelling \emph{koraṇṭaka} is attested in the \emph{Haṭhābhyāsapaddhati}, and it is reasonably well attested by manuscripts of the \emph{Haṭhapradīpikā}, as well as those of the \emph{Haṭharatnāvalī} (which also has \emph{gonandaka}).
+
+The compound \emph{siddhapāda} could be a respectful affix. However, it seems unlikely here because it would cross the \emph{pāda} break.
+
 \end{philcomm}
@@
 \begin{translation}[hp01_007]
-Kānerī, Pūjyapāda, Nityanātha, Nirañjana, Kapālī, Bindunātha, and the one named Kākacaṇḍīśvara.
+Kāṇerī, Pūjyapāda, Nityanātha, Nirañjana, Kapālī, Bindunātha, and the one named Kākacaṇḍīśvara.
 \end{translation}
@@
 \begin{testimonia}[hp01_007]
-Haṭharatnāvalī 1.82
+\emph{Haṭharatnāvalī} 1.82
 
@@
 
-Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
-​​
+Caturbhuja Misra's \emph{Mugdhāvabodhinī} on the \emph{Rasahṛdayatantra}
+
 \begin{versinnote}
@@
 The variant \emph{dhvaninātha} may have resulted from a transposition of the first two syllables of \emph{nityanātha}.
+
+The alpha group supports \emph{kākacaṇḍīśvaro gayaḥ} but we have not been able to find evidence for a Siddha called Gaya.
 \end{philcomm}
@@
 \begin{testimonia}[hp01_008]
-Haṭharatnāvalī 1.83
+\emph{Haṭharatnāvalī} 1.83
 
@@
 \tl{allamaḥ prabhudevaś ca naiṭacūṭiś ca ṭiṇṭiṇiḥ |\\+}
-\tl{bhālukir nāgabodhaś ca khaṇḍakāpālikas tathā || \\!}
-\end{versinnote}
-
-Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
+\tl{bhālukir nāgabodhaś ca khaṇḍakāpālikas tathā ||\\+}
+\tl{\var{allamaḥ prabhudevaś ] allamaprabhudevaś P,T,t1}\\!}
+\end{versinnote}
+
+Caturbhuja Misra's \emph{Mugdhāvabodhinī} on the \emph{Rasahṛdayatantra}
 
@@
 \begin{philcomm}[hp01_008]  
-The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literaure) is largely transmitted as allamaḥ prabhudevaḥ, as though it were two names, although some manuscripts of the \emph{Haṭhapradīpikā} have \emph{allamaprabhudevaś ca} (i.e., Tue,V3,V8,V13,V16,V22,Vu,N24,N26) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n.
+The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literaure) is largely transmitted as \emph{allamaḥ prabhu\-devaḥ}, as though it were two names, although some manuscripts of the \emph{Haṭha\-pradīpikā} have \emph{allamaprabhudevaś ca} (i.e., \getsiglum{V3,V8,V13,V16,V22,N24,N26,Jyo}) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n.
 8).
 
-The names Nāgabodha, Nāgabodhi, Naradeva, Nāgadeva all seem possible in 1.8c.
-
-Many witnesses have khaṇḍa and kāpālika as separate names.
-However, khaṇḍakāpālika is well attested.
+The names Nāgabodha, Nāgabodhi, Naradeva, Nāgadeva all seem possible in 1.8c. The reading \emph{nāgabodhaś ca} is attested across several primary groups of manuscripts.
+
+Many witnesses have \emph{khaṇḍa} and \emph{kāpālika} as separate names.
+However, Khaṇḍa\-kāpālika is well attested.
 Examples include \emph{Kathāsaritsāgara} 121.5 ff.
-(check), \emph{Bṛhatkathāmañjarī} 10.45 (check) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}).
+(check??), \emph{Bṛhatkathāmañjarī} 10.45 (check??) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}).
 It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika.
@@
 \begin{testimonia}[hp01_009]
-Haṭharatnāvalī 1.84
+\emph{Haṭharatnāvalī} 1.84
 
@@
 
-Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
+Caturbhuja Misra's \emph{Mugdhāvabodhinī} on the \emph{Rasahṛdayatantra}
 
@@
 
-Haṭhatattvakaumudī 17.24
+\emph{Haṭhatattvakaumudī} 17.24
 
@@
 \begin{testimonia}[hp01_010]
-Yogasārasaṃgraha (YSS), p.53.
+\emph{Yogasārasaṅgraha}, p.53.
 
@@
 \begin{philcomm}[hp01_010] 
-The compound \emph{saṃsāratāpa°} is well attested and found elsewhere (e.g., \emph{Viṣṇupurāṇa} 6.7.62, \emph{Agnipurāṇa} 371.1, \emph{Haṭhatattvakaumudī} 38.92, emph{Haṭhābhyāsapaddhati} ms.
+The compound \emph{saṃsāratāpa°} is well attested and found elsewhere (e.g., \emph{Viṣṇupurāṇa} 6.7.62, \emph{Agnipurāṇa} 371.1, \emph{Haṭhatattvakaumudī} 38.92, \emph{Haṭhābhyāsapaddhati} ms.
 46/440, f.
 1v).
-The reading of \emph{samāśrayo} in V1 is metrically faulty.
+
+The reading of \emph{samāśrayo} in eta one is metrically faulty.
+
 The word \emph{°jagatāṃ} is probably authorial because it makes good sense with  \emph{ādhārakamaṭha} in light of the cosmological notion that the tortoise supports all the worlds.
@@
 \begin{sources}[hp01_011]
-Śivasaṃhitā
+\emph{Śivasaṃhitā} 5.254
 
@@
 \tl{haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\+}
-\tl{bhaved vīryavatī guptā nirvīryā ca prakāśitā || 5.254\\+}
-\tl{(ab missing in some ŚS witnesses)\\!}
+\tl{bhaved vīryavatī guptā nirvīryā ca prakāśitā ||\\+}
+\tl{\var{haṭhavidyā ... icchatā ] \emph{om.} III–XII, XIV}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_011]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 141r %(NGMCP 1-1337, reel no: B 39/5)
 
@@
 
-BKhP 10v4
+%BKhP 10v4
 \end{testimonia}
@@
 \begin{sources}[hp01_012]
-References to maṭhikā here.
 \end{sources}
@@
 \begin{testimonia}[hp01_012]
-Haṭharatnāvalī 1.66
+\emph{Haṭharatnāvalī} 1.66
 
@@
 
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 54r
 
@@
 
-BKhP 107v1
 \end{testimonia}
@@
 \begin{philcomm}[hp01_012]  
+The term \emph{maṭhikā} occurs in narrative literature and yoga texts in the sense of a small hut.
+For example, in the \textit{Kathāsaritsāgara} (12.9.14, 29–30), \emph{maṭhikā} refers to the small hut built in a cremation ground by a young Brahmin who makes as his bed the ashes of the dead girl he had hoped to marry.
+In several other stories (\textit{Kathāsaritsāgara} 6.6.132, 10.5.89, 12.25.35), \emph{maṭhikā} is the term used for the hut of an ascetic.
+In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four fore-arm lenghts (\emph{hasta}) high and wide, and it can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).
+
 In the \emph{Jyotsnā} and printed editions of the \emph{Haṭhapradīpikā}, including one by Digambara and Kokaje (1970: 6), this verse has the additional hemistich, \emph{dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite}.
@@
 It extends not too high or low, and is thickly smeared with cow dung in the proper way.
-It is clean, free from everything that annoys, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
+It is clean, free from everything that annoys, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
 \end{translation}
@@
 \begin{sources}[hp01_013]
-Cf. Dattātreyayogaśāstra
-
-\begin{versinnote}
-\tl{suśobhanaṃ maṭhaṃ kuryāt sūkṣmadvāraṃ tu nirvraṇaṃ || 54 ||\\+}
+Cf. \emph{Dattātreyayogaśāstra} 54cd–57
+
+\begin{versinnote}
+\tl{suśobhanaṃ maṭhaṃ kuryāt sūkṣmadvāraṃ tu nirvraṇaṃ ||\\+}
 \tl{suṣṭhu liptaṃ gomayena sudhayā vā prayatnataḥ |\\+}
-\tl{matkuṇair maśakair bhūtair varjitaṃ ca prayatnataḥ || 55 ||\\+}
+\tl{matkuṇair maśakair bhūtair varjitaṃ ca prayatnataḥ ||\\+}
 \tl{dine dine susammṛṣṭaṃ sammārjanyā hy atandritaḥ |\\+}
-\tl{vāsitaṃ ca sugandhena dhūpitaṃ guggulādibhiḥ || 56 ||\\+}
+\tl{vāsitaṃ ca sugandhena dhūpitaṃ guggulādibhiḥ ||\\+}
 \tl{malamūtrādibhir vargair aṣṭādaśabhir eva ca |\\+}
-\tl{varjitaṃ dvārasampannaṃ vastrāvaraṇam eva vā || 57 ||\\!}
+\tl{varjitaṃ dvārasampannaṃ vastrāvaraṇam eva vā ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_013]
-Suśruta 6.17.67:
+\emph{Suśrutasaṃhitā} 6.17.67:
 
@@
 
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} 54r (attr. \emph{Haṭhapradīpikā})
 
@@
 \tl{bāhye maṇḍapakūpavediracitaṃ prākārasaṃveṣṭitam |\\+}
-\tl{proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 1.67
+\tl{proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||\\+}
+\tl{\var{°vimalaṃ ] L, mavilaṃ N}\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 1.67
 
@@
 \tl{bāhye maṇḍapavedikūparuciraṃ prākārasaṃveṣṭitam\\+}
-\tl{proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||\\!}
-\end{versinnote}
-
-BKhP 107v3
+\tl{proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ |||\\+}
+\tl{\var{°piṭharaṃ ] piṭakaṃ J,n2, peṭakaṃ N}\\!}
+\end{versinnote}
+
+%BKhP 107v3
 \end{testimonia}
@@
 The syntax of this verse is rather problematic.
-One would expect the features of the hut, which are listed in the first three pādas of the verse, to be in the nominative case, with the words \emph{idaṃ lakṣaṇaṃ} in the fourth pāda referring back to them.
-However, the compounds in the first three pādas appear to qualify \emph{lakṣaṇa} as though they were adjectives, and this appears to have been the way the verse was composed.
+One would expect the features of the hut, which are listed in the first three quarters of the verse, to be in the nominative case.
+Then, the words \emph{idaṃ lakṣaṇaṃ} in the fourth quarter would refer back to them.
+However, the compounds in the first three verse-quarters appear to qualify \emph{lakṣaṇa} as though they were adjectives, and this seems to have been the way the verse was composed.
 
 The manuscripts preserve many different readings at the end of the compound beginning with \emph{arandhragarta°}.
-The reading \emph{°piṭharaṃ} has been suggested by Dominic Goodall, who understands it as potsherds (ask ref from Dominic?).
-The sense is that the hut should be free of rubbish, such as potsherds.
+The reading \emph{°piṭharaṃ} has been suggested by Dominic Goodall in the sense of potsherds.
+The idea behind this reading is that hut should be free of rubbish, such as potsherds.
 One would expect a word for a defect in a hut that is similar to, but not the same as, cracks (\emph{randhra}) and holes (\emph{garta}).
 For this reason, the reading \emph{°vivaraṃ} looks like a patch, as its meaning does not add anything to \emph{°randhragarta°}.
-The reading \emph{viṭapaṃ} is well attested among the witnesses of the \emph{Haṭhapradīpikā} and \emph{Haṭharatnāvalī}.
+The reading \emph{°viṭapaṃ} (`the young branch of a tree or creeper') is attested among the witnesses of the \emph{Haṭhapradīpikā}.
 However, it is difficult to construe in this context unless it was intended to refer to creepers or branches that might invade or encroach upon the hut.
 
-Nearly all the early mss.
-have \emph{°bādhojjitaṃ}, whereas the Yogacintāmaṇi and V19 have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’).
-However, \emph{°bādhojjitaṃ} may have been original because it is so well attested and a hut is described similarly in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
+(MD: Stemmatically, \emph{piṭaka} ``a basket; a boil" is well attested.
+Is it perhaps possible to interpret the word as ``a bump"?
+\emph{a-randhra-garta-piṭaka} would then mean that the floor or walls of the hut are not uneven.)
+
+Manuscripts of several groups, namely beta, epsilon and eta, have \emph{°bādhojjitaṃ}, whereas delta and the \emph{Yogacintāmaṇi} have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’), which is somewhat suggested by the alpha reading (\emph{jyaṃtyūpsitaṃ} \getsiglum{N3}, \emph{jaṃtūṣṇitaṃ} \getsiglum{J5}, but \emph{bodhojhitaṃ} \getsiglum{G4}).
+However, \emph{°bādhojjitaṃ} may have been original because it is a rather unusual descriptive compound for a hut, and yet a hut is described similarly in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
+
 \end{philcomm}
@@
 \begin{sources}[hp01_014]
-Cf. Amanaska 2.15
+Cf. \emph{Amanaska} 2.15
 
@@
 \begin{testimonia}[hp01_014]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f.54r (attr. \emph{Haṭhapradīpikā})
 
@@
 \tl{evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\+}
-\tl{gurūpadiṣṭamārgeṇa yogam eva sadābhyased iti ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 1.68
+\tl{gurūpadiṣṭamārgeṇa yogam eva sadābhyaset ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 1.68
 
@@
 \begin{testimonia}[hp01_015]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 48v (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī 1.77
+\emph{Haṭharatnāvalī} 1.77
 
@@
 \tl{atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\+}
-\tl{janasaṅgaṃ ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||\\!}
-\end{versinnote}
-
-Yuktabhavadeva 4.25 (attributed to the śivayoga)
+\tl{janasaṅgaṃ ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||\\+}
+\tl{\var{niyamagrahaḥ ] niyamāgrahaḥ N,J}\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 4.25 (attr. śivayoga)
 
@@
 
-Jyotsnā
+\emph{Jyotsnā} 1.15
 
@@
 
-Yogaprakāśikā
+\emph{Yogaprakāśikā} 1.48
 
@@
 Furthermore, verse 3.82 suggests that a yogi who does not practice \emph{niyama} might obtain success in yoga through the practice of \textit{vajroli}.
-Ambiguity over the role of \emph{yama} and \emph{niyama} in Haṭhayoga appears to have prompted some to insert verses on ten \emph{yama}s and ten \emph{niyama}s after the next verse (1.16).
-The additional verses may derive from the \emph{Śāradātilakatantra} (25.7–8) or the  \emph{Vasiṣṭhasaṃhitā} (1.38, 1.53).
+Ambiguity over the role of \emph{yama} and \emph{niyama} in Haṭhayoga may explain why two verses on ten \emph{yama}s and ten \emph{niyama}s were inserted in some manuscripts after the next verse (1.16).
+The additional verses derive from either the \emph{Śāradātilakatantra} (25.7–8) or the  \emph{Vasiṣṭhasaṃhitā} (1.38, 1.53).
 In the \emph{Jyotsnā}, Brahmānanda reads \emph{niyamāgraha} and takes it as though \emph{āgraha} was implied, which yields the meaning of ‘over-insistence on rules’, and he relates it to extreme ascetic practice.
@@
 \begin{sources}[hp01_016]
-Dharmaputrikā 137cd–38ab%Narharināth ed.?
+\emph{Dharmaputrikā} 38cd–39ab
+%Narharināth ed.? Yes
+%Yogī Naraharinātha, Śivadharma PaśupatimatamŚivadharmamahāśāstram Paśupatināthadarśanam.Kathmandu 1998, pp. 679 onwards
 
@@
 
-Śivadharmottara 10 (W 122r):%JM: what ed or ms is this?
+\emph{Śivadharmottara} 10 (W 122r):%JM: what ed or ms is this?
 
@@
 
-Jñānārṇava 20.1
+\emph{Jñānārṇava} 20.1
 
@@
 \tl{utsāhān niścayād dhairyāt saṃtoṣāt tattvaniścayāt |\\+}
-\tl{muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 1 ||\\!}
-\end{versinnote}
-
-Yogabindu by Haribhadra
+\tl{muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati ||\\!}
+\end{versinnote}
+
+\emph{Yogabindu} 411 (by Haribhadra)
 
@@
 \tl{utsāhān niścayād dhairyāt saṃtopāt tattvadarśanāt |\\+}
-\tl{muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 411 ||\\!}
+\tl{muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_016]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 49r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī 1.78:
-
-\begin{versinnote}
-\tl{utsāhān niścayād dhairyāt tattvajñānārthadarśanāt
-[utsāhān niścalād- P,T]\\+}
+\emph{Haṭharatnāvalī} 1.78:
+
+\begin{versinnote}
+\tl{utsāhān niścayād dhairyāt tattvajñānārthadarśanāt |\\+}
 \tl{bindusthairyān mitāhārāj janasaṅgavivarjanāt |\\+}
 \tl{nidrātyāgāj jitaśvāsāt pīṭhasthairyād anālasāt\\+}
-\tl{gurvācāryaprasādāc ca ebhir yogas tu sidhyati || \\!}
+\tl{gurvācāryaprasādāc ca ebhir yogas tu sidhyati ||\\+} 
+\tl{\var{niścayād ] niścalād- P,T}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_016]    
-For the second \emph{pāda}, the north-Indian manuscripts have something along the lines of \emph{tattvajñānāc ca darśanāt}.
+Alpha and several other groups of manuscripts have \emph{tattvajñānāc ca darśanāt} or something very similar in the second quarter of the verse.
 The problem with this reading is the meaning of \emph{darśana} by itself (i.e., discernment of what?).
-The early sources of this verse, in particular the \emph{Śivadharmottara}, indicate that the second pāda read as \emph{santoṣāt tattvadarśanāt}, which makes much better sense of the word \emph{darśana} (i.e., ‘seeing the truth’).
-It seems that this reading may have been adopted by Svātmārāma because the word \emph{santoṣāt} is present in the south-Indian manuscripts.
-At some point in the transmission, \emph{tattvadarśanāt} became \emph{tattvajñānāt} and \emph{santoṣāt} was dropped.
-The word \emph{darśanāt} was changed to \emph{viniścayāt} in V19 and the \emph{Yogacintāmaṇi}, which indicates that we are not alone in questioning its meaning.
-The problem with \emph{viniścayāt} in \emph{pāda} two is that \emph{niścayāt} is widely attested in \emph{pāda} one.
-It should also be noted that the word \emph{tattva} could have a more specific meaning in the \emph{Haṭhapradīpikā} (4.3) as Svātmārāma states that it is a synonym of \emph{samādhi}.
-In other yoga texts, it can sometimes refer to the practices of yoga (e.g., \emph{tritattva} in \emph{Amṛtasiddhi} 13.12, 14.2--3).
+The early sources of this verse, in particular the \emph{Śivadharmottara}, indicate that the second verse quarter read as \emph{santoṣāt tattvadarśanāt}, which makes much better sense of the word \emph{darśana} (i.e., ‘seeing the truth’).
+However, It seems likely that before the time of Svātmārāma other versions of this verse were circulating, in which \emph{santoṣāt} had been dropped, \emph{niścayāt} moved from the first to second verse quarter, \emph{tattvadarśanāt} became \emph{tattvajñānāt} and \emph{sāhasāt} was introduced.
+It should also be noted that the word \emph{tattva} could have a more specific meaning in the \emph{Haṭhapradīpikā} (4.45–46) as Svātmārāma states that it is a synonym of \emph{samādhi}.
+In other yoga texts, it can sometimes refer to the practices of yoga (e.g., \emph{tritattva} in \emph{Amṛtasiddhi} 13.12, 14.2--3) or, more generally, to the highest reality or truth (e.g., \emph{Amanaska} 1.2, 1.20–21, 2.17, etc.).
 \end{philcomm}
@@
 \begin{translation}[hp01_017]
-Because it is the first auxiliary of haṭha, \emph{āsana} is taught first. This type of (\emph{tad}) \emph{āsana} brings about steadiness, good health and physical fitness.
+Because it is the first auxiliary of haṭha, \emph{āsana} is taught first. This type (\emph{tad}) of \emph{āsana} brings about steadiness, good health and physical fitness.
 \end{translation}
@@
 \begin{testimonia}[hp01_017]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī 3.5
-
-\begin{versinnote}
-\tl{haṭhasya prathamāṅgatvād āsanaṃ dārśyate mayā |\\+}
+\emph{Haṭharatnāvalī} 3.5
+
+\begin{versinnote}
+\tl{haṭhasya prathamāṅgatvād āsanaṃ darśyate mayā |\\+}
 \tl{tat kuryād āsanaṃ sthairyam ārogyaṃ cāṅgapāṭavam || \\!}
@@
 \begin{philcomm}[hp01_017]        
-The reading of \emph{aṅgapāṭavam} is attested among most of the early manuscripts.
+The reading of \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the main one of the alpha group.
 Although this compound rarely appears in other yoga texts, a similar term \emph{śarīrapāṭava} occurs in the \emph{Śivasaṃhitā} (2.35) as one of the benefits bestowed by digestive fire (\emph{vaiśvānarāgni}), which indicates that the word \emph{pāṭava} was used in relation to the body and the benefits of yoga.
 The compound \emph{aṅgapāṭava} seems to imply the optimal functioning of the body.
-However, the alternative reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts, even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}).
-Therefore, it is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by group two (i.e., N23, V19, etc.)
+However, the alternative reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts and occurs even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}).
+Therefore, it is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by manuscripts in several early groups (i.e., beta, gamma and delta).
 \end{philcomm}
@@
 \begin{testimonia}[hp01_018]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} 84r
 
@@
 
-Haṭharatnāvalī 3.6
+\emph{Haṭharatnāvalī} 3.6
 
@@
 \begin{sources}[hp01_019]
-Śāradātilaka 25.12
+\emph{Śāradātilaka} 25.12
 
@@
 
-Vasiṣṭhasaṃhitā 1.68
+\emph{Vasiṣṭhasaṃhitā} 1.68
 
@@
 
-Yogayājñavalkya 3.3
+\emph{Yogayājñavalkya} 3.3
 
@@
 \begin{testimonia}[hp01_019]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{Yājñavalkyaḥ—\\+}
+\emph{Yogacintāmaṇi} f. 83v
+
+\begin{versinnote}
+\tl{yājñavalkyaḥ—\\+}
 \tl{jānūrvor antare samyak kṛtvā pādatale ubhe |\\+}
@@
 
-Haṭharatnāvalī 3.52
+\emph{Haṭharatnāvalī} 3.52
 
@@
 Brahmānanda explains that the region of the shank near the knee should be understood by the word ‘knee’ in this verse (\emph{atra jānuśabdena jānusaṃnihito jaṅghāpradeśo grāhyaḥ jānusaṃnihito jaṅghāpradeśaḥ}).
-This is consistent with the earliest known description of \emph{svastikāsana} in the \emph{Pātañjalayogaśāstravivaraṇa} (2.46), which states that the big toes of one foot are tucked in between the shank and thigh of the other so it is not seen (\emph{dakṣiṇaṃ pādāṅguṣṭhaṃ savyenorujaṅghena parigṛhyādṛśyaṃ kṛtvā tathā savyaṃ pādāṅguṣṭhuṃ dakṣiṇenorujaṅkgenādṛśyaṃ parihṛhya yathā ca pārṣṇibhyāṃ vṛṣaṇayor apīḍanaṃ tathā yenāste tat svastikam āsanam}).
+This is consistent with the earliest known description of \emph{svastikāsana} in the \emph{Pātañjalayogaśāstravivaraṇa} (2.46), which states that the big toe of one foot is tucked in between the shank and thigh of the other so it is not seen (\emph{dakṣiṇaṃ pādāṅguṣṭhaṃ savyenorujaṅghena parigṛhyādṛśyaṃ kṛtvā tathā savyaṃ pādāṅguṣṭhuṃ dakṣiṇenorujaṅkgenādṛśyaṃ parihṛhya yathā ca pārṣṇibhyāṃ vṛṣaṇayor apīḍanaṃ tathā yenāste tat svastikam āsanam}).
 For a discussion of \emph{svastikāsana} in the Pātañjalayoga tradition, see Maas 2018: 68–69.
@@
 \begin{sources}[hp01_020]
-Cf. Ahirbudhnyasaṃhitā 31.45cd–46
+Cf. \emph{Ahirbudhnyasaṃhitā} 31.45cd–46
 
@@
 
-Vasiṣṭhasaṃhitā 1.70
+\emph{Vasiṣṭhasaṃhitā} 1.70
 
@@
 
-Yogayājñavalkya 3.5cd–3.6ab
+\emph{Yogayājñavalkya} 3.5cd–3.6ab
 
@@
 \begin{testimonia}[hp01_020]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 83v (attr. Yājñavalkya)
 
@@
 
-Haṭharatnāvalī 3.53
+\emph{Haṭharatnāvalī} 3.53
 
@@
 \begin{sources}[hp01_021]
-Vasiṣṭhasaṃhitā 1.72
+\emph{Vasiṣṭhasaṃhitā} 1.72
 
@@
 
-Cf. Śāradātilakatantra 25.15cd–16ab
+Cf. \emph{Śāradātilakatantra} 25.15cd–16ab
 
@@
 
-Yogayājñavalkya 3.8
+\emph{Yogayājñavalkya} 3.8
 
@@
 \begin{testimonia}[hp01_021]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 83v (attr. Yājñavalkya)
 
@@
 
-Haṭharatnāvalī 3.54
+\emph{Haṭharatnāvalī} 3.54
 
@@
 \tl{ekaṃ pādam athaikasmin vinyased ūruṇi sthiram ||\\+}
-\tl{itarasmiṃs tathā coruṃ vīrāsanam īritam || \\!}
+\tl{itarasmiṃs tathā coruṃ vīrāsanam īritam ||\\+} 
+\tl{\var{sthiram ] sthitam T}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_021]   
-Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by V1, the sources and the testimonia.
+Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by group 4c, the sources and the testimonia.
 It appears to be verse filler here rather than indicating a temporal sequence of actions.
@@
 \begin{sources}[hp01_022]
-Vasiṣṭhasaṃhitā 1.80
+\emph{Vasiṣṭhasaṃhitā} 1.80
 
@@
 
-Cf. Ahirbudhnyasaṃhitā 31.35
+Cf. \emph{Ahirbudhnyasaṃhitā} 31.35
 
@@
 \begin{testimonia}[hp01_022]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Yuktabhavadeva 6.15
+\emph{Yuktabhavadeva} 6.15
 
@@
 \tl{gudaṃ niyamya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\+}
-\tl{kūrmāsanaṃ bhavedetaditi yogavido viduḥ ||\\!}
+\tl{kūrmāsanaṃ bhavedetaditi yogavido viduḥ || iti kūrmāsanam ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_022]   
-In the first \emph{pāda}, the witnesses are split between \emph{nirudhya} (‘having blocked’), \emph{nibadhya} (‘having bound’), \emph{niyamya} (‘having restrained’) and \emph{niṣpīḍya} (‘having pressed’).
-The source, the \emph{Vasiṣṭhasaṃhitā}, supports \emph{nirudhya} whereas the testimonia supports \emph{niyamya}.
-In terms of closing the anus by sitting on the ankles, \emph{nirudhya} makes better sense and has been adopted because it is supported by N23 (an important witness of group 2) and the manuscripts reported in the critical edition of the \emph{Vasiṣṭhasaṃhitā}.
+In the first quarter of the verse, the witnesses are split between \emph{nirudhya} (‘having blocked’), \emph{nibadhya} (‘having bound’), \emph{niyamya} (‘having restrained’) and \emph{niṣpīḍya} (‘having pressed’).
+The source, the \emph{Vasiṣṭhasaṃhitā}, and two manuscripts of the beta and gamma groups support \emph{nirudhya} whereas the testimonia supports \emph{niyamya}.
+In terms of closing the anus by sitting on the ankles, \emph{nirudhya} makes better sense.
+
 The word \emph{vyutkrameṇa} appears to describe the position of the ankles.
-Its basic meaning is ‘against the normal direction’, which would suggest that the ankles are turned out or crossed rather than placed together naturally.
+Its basic meaning is ‘against the normal direction,’ which would suggest that the ankles are turned out or crossed rather than placed together naturally.
 If the yogi is in a kneeling-type position, turning the feet out would bring the ankles together, blocking the perineal area.
@@
 \begin{translation}[hp01_023]
-[The yogi] adeptly assumes \emph{padmāsana}, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is \emph{kurkuṭāsana}.
+[The yogi] adeptly assumes \emph{padmāsana}, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is \emph{kukkuṭāsana}.
 \end{translation}
@@
 \begin{sources}[hp01_023]
-Vasiṣṭhasaṃhitā 1.78
+\emph{Vasiṣṭhasaṃhitā} 1.78
 
@@
 
-Cf. Ahirbudhnyasaṃhitā 31.38
+Cf. \emph{Ahirbudhnyasaṃhitā} 31.38
 
@@
 \begin{testimonia}[hp01_023]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī 3.73
+\emph{Haṭharatnāvalī} 3.73
 
@@
 \tl{atha kukkuṭāsanam---\\+}
-\tl{padmāsanaṃ susaṃsthāpya jānūrvor antare karau ||\\+}
+\tl{padmāsanaṃ susaṃsthāpya jānūrvor antare karau |\\+}
 \tl{niveśya bhūmau saṃsthāpya vyomasthaḥ kukkuṭāsanam ||\\!}
@@
 
+\emph{Yuktabhavadeva} 6.16 (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{padmāsanaṃ tu saṃyojya jānūrvor antare karau |\\+}
+\tl{niveśya bhūmau saṃsthāpya vyomasthaṃ kukkuṭāsanam || iti kukkuṭāsanam ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 \begin{philcomm}[hp01_023]
-\emph{Kurkuṭa} and \emph{kurkkuṭa} in V1, J10ac, V3 are variant spellings attested in the \emph{Pañcatantra} (M-W).
+The names \emph{kurkuṭa} and \emph{kurkkuṭa} in some manuscripts are variant spellings of \emph{kukkuṭa} attested in the \emph{Pañcatantra} (M-W).
 \end{philcomm}
@@
 \begin{translation}[hp01_024]
-While maintaining kurkuṭāsana, [the yogi] binds the neck with the hands and lies like a tortoise on his back. This is uttānakūrmāsana.
+While maintaining \emph{kukkuṭāsana}, [the yogi] binds the neck with the hands and lies like a tortoise on his back. This is \emph{uttānakūrmāsana}.
 \end{translation}
@@
 \begin{testimonia}[hp01_024]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{kukkuṭāsanabandhastho dorbhyāṃ saṃbadhya kandharam |\\+}
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{kukkuṭāsanabandhastho dorbhyāṃ saṃbadhya kandharām |\\+}
 \tl{bhavet kūrmavad uttānam etad uttānakūrmakam ||\\!}
@@
 
-Haṭharatnāvalī 3.74
+\emph{Haṭharatnāvalī} 3.74
 
@@
 
+\emph{Yuktabhavadeva} 6.17 (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{kukkuṭāsanabandhastho dorbhyāṃ sambadhya kandharām |\\+}
+\tl{śete kūrmavad uttānam etad uttānakūrmakam || iti uttānakūrmāsanam ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 \begin{philcomm}[hp01_024]
-\emph{°bandhasthaḥ} or \emph{°vat kṛtvā}?
-Only V1 has the latter, which is simpler.
-Are the others trying to improve it?
-Stemmatically ambiguous as \emph{°bandha°} is on one branch (V3/J8, V19) and \emph{°madhya°} the other (J10, J17, N17).
-V1 looks like an outlier.
-
-Adopt \emph{bandhastho} with note. \emph{°bandhastha} not found in any other texts. While \emph{°vat kṛtvā} is possible, it is a singular reading that appears to be unrelated to the other variants.
+The oldest dated manuscript, eta one, has \emph{kukkuṭāsanavat kṛtvā}, which is a simpler alternative to the widely attested reading \emph{kukkuṭāsanabandhasthaḥ}, which we have accepted.
+Since there is no known source for this verse other than the \emph{Haṭhapradīpikā}, we can only assume that the reading of eta one was an isolated attempt to simplify the syntax, for it is a singular reading that appears to be unrelated to the other variants.
 \end{philcomm}
@@
 \begin{translation}[hp01_025]
-Clasping the big toes with hands and performing the action of drawing a bow as far as the ear is called dhanurāsana.
+Clasping the big toes with hands and performing the action of drawing a bow as far as the ear is called \emph{dhanurāsana}.
 \end{translation}
@@
 \begin{testimonia}[hp01_025]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 \tl{pādāṅguṣṭhau ca pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\+}
-\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam |\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.51
+\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 3.51
 
@@
 \tl{pādāṅguṣṭhau tu pāṇibhyāṃ gṛhītvā śravaṇāvadhi ||\\+}
-\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam ucyate ||\\!}
-\end{versinnote}
-
-Cf. Haṭhayogasaṃhitā
+\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam ucyate ||\\+}
+\tl{\var{ākarṣaṇaṃ kṛtvā ] ākarṣaṇākṛṣṭaṃ P,T,t1}\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 6.18 (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{pādāṅguṣṭhau tu pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\+}
+\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam || iti dhanurāsanam ||\\!}
+\end{versinnote}
+
+Cf. \emph{Haṭhayogasaṃhitā} p. 21
 
@@
 \begin{philcomm}[hp01_025]
-Comment on kṛtvā, which is the reading of V19, Yogacintāmaṇi and Haṭharatnāvalī.
-Godāvaramiśra has dhanurākarṣavat kṛṣṭaṃ dhanurāsanam ucyate, which is likely to be the original reading.
+Since the word \emph{ākarṣaṇa} in one form or other is so well attested in the third verse quarter, the reading \emph{dhanurākarṣaṇaṃ kṛtvā}, which is in epsilon two, as well as the principal testimonia (i.e., the \emph{Yogacintāmaṇi} and \emph{Haṭharatnāvalī}), fits the overall syntax of the verse.
+However, it is curious that \emph{kṛṣtaṃ} (for \emph{kṛtvā}) is well attested in some groups of manuscripts because it seems redundant with \emph{ākarṣaṇaṃ}.
+However, the following reading in
+Godāvaramiśra's \emph{Yogacintāmaṇi} (f. 40r) makes sense of \emph{kṛṣṭaṃ} and might indeed be the original version of the verse: \emph{dhanurākarṣavat kṛṣṭaṃ dhanurāsanam ucyate}.
+
+A different version of \emph{dhanurāsana} is described in the \emph{Haṭhayogasaṃhitā}. On the two versions of \emph{dhanurāsana}, see Hargreaves and Birch 2017.
+%https://www.theluminescent.org/2017/11/dhanurasana-two-versions-of-bow-pose.html
 \end{philcomm}
@@
 \begin{translation}[hp01_026]
-Having grasped the right foot, which is placed at the base of the left thigh, [the yogi’s] left foot is wrapped around the outside of the knee and he remains with his body twisted.
-This āsana was taught by Matsyendranātha.
+Having grasped the right foot, which is placed at the base of the left thigh, with the right arm wrapped around the outside of the knee, [the yogi] remains with his body twisted.
+This \emph{āsana} was taught by Matsyendranātha.
 \end{translation}
@@
 \begin{testimonia}[hp01_026]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 \tl{vāmorumūlārpitadakṣapādaṃ
-jānvor bahir veṣṭitadakṣadoṣṇā |\\+}
+jānvor bahirveṣṭitadakṣadoṣṇā |\\+}
 \tl{pragṛhya tiṣṭhet parivartitāṅgaḥ
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.57
 
@@
 \tl{atha matsyendrāsanam---\\+}
-\tl{vāmorumūlārpitā dakṣapādo jānvor bahir veṣṭitadakṣadoṣṇā ||\\+}
-\tl{pragṛhya tiṣṭhet partivartitāṅgaḥ śrīmatsyanāthoditam āsanaṃ syāt || 3.57 ||\\!}
+\tl{vāmorumūlārpitadakṣapādo jānvor bahirveṣṭitadakṣadoṣṇā |\\+}
+\tl{pragṛhya tiṣṭhet partivartitāṅgaḥ śrīmatsyanāthoditam āsanaṃ syāt~||\\+}
+\tl{\var{°dakṣapādo ] °dakṣapādaṃ P, °dakṣapādau t1}\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 6.19 (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{vāmorumūlārpitadakṣapādaṃ
+jānvor bahirveṣṭitadakṣadoṣṇā |\\+}
+\tl{pragṛhya tiṣṭhan parivartitāṅgaḥ
+śrīmatsyanāthoditam āsanaṃ syāt ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_026]
-Comment on vāmapadam vs dakṣadoṣṇā
+In the second verse quarter, most of the manuscript groups have a compound with \emph{°doṣṇā} at the end, as seen also in the \emph{Yogacintāmaṇi}, \emph{Haṭharatnāvalī} and \emph{Yuktabhavadeva}.
+The instrumental ending (`with the hand') works well with the gerund (\emph{pragṛhya}) in the third verse quarter and the object (\emph{°dakṣpādaṃ}) in the first quarter.
+This reading indicates that the right foot is held by the right hand, the arm of which is wrapped around the outside of the left leg (Figure 1).
+
+The version of this verse in \emph{Jyotsnā} (1.26), which is supported by some manuscripts in two important groups, beta and eta, has two objects of the gerund, namely the left and right feet, without an instrumental or conjuctive particle.
+In his commentarial remarks, Brahmānanda proposes that the left foot is grasped by the right hand and the right foot by the left foot, as seen in Figure 2.
 \end{philcomm}
@@
 \begin{translation}[hp01_027]
-Matsyendra's seat, is a destructive weapon for the many terrible diseases that develop in the stomach; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
+Matsyendra's seat is a destructive weapon for the many terrible diseases that develop in the stomach; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
 \end{translation}
@@
 \begin{testimonia}[hp01_027]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 \tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
-\tl{daṇḍe sthiratvaṃ pradadāti puṃsām ||
-\var{°pravṛddha ] N : °pravṛddhiṃ L}\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.58
+\tl{daṇḍe sthiratvaṃ pradadāti puṃsām ||\\+}
+\tl{\var{°pravṛddha ] N : °pravṛddhiṃ L}\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 3.58
 
@@
 \tl{matsyendrapīṭhaṃ jaṭharapradīptaṃ\\+}
-\tl{pracaṇḍarugmaṇḍalakhaṇḍanāstram ||\\+}
+\tl{pracaṇḍarugmaṇḍalakhaṇḍanāstram |\\+}
 \tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
-\tl{daṇḍasthiratvaṃ ca dadāti puṃsām ||\\!}
-\end{versinnote}
-
-HTK
+\tl{daṇḍasthiratvaṃ ca dadāti puṃsām ||\\+}
+\tl{\var{°pradīptaṃ ] pravṛttaṃ T,t1 °pravṛttaḥ N,n1,n3,J}\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 7.8
 
@@
 \tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
-\tl{daṇḍasthiratvaṃ ca dadāti puṃsām || 8 ||\\!}
-\end{versinnote}
-
-YBD
+\tl{daṇḍasthiratvaṃ ca dadāti puṃsām ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 6.20 (attr. \emph{Haṭhapradīpikā})
 
@@
 \begin{philcomm}[hp01_027]
-Change note: we are adopting jaṭharapravṛddha (translation has been changed).  Cf. Yogacintāmaṇi ms. N. and HP mss.  J10, J17, V2, N8, N11.
-
-Pāda d, J10ac and J17 have \emph{candra} for \emph{daṇḍa}; However,  also in Yogacintāmaṇi and 6-chapter HP.
+The manuscript readings for the compound beginning with \emph{jaṭhara} diverge significantly and include \emph{jaṭharapravṛddha}, \emph{jaṭharaprabuddha°}, \emph{jaṭharapradīpta°} and \emph{jaṭharapracaṇḍa°}.
+As descriptive compounds, none of these make good sense in regard to Matsyendra's seat.
+Since the stomach or abdomen (\emph{jaṭhara}) is the first member of this compound, it seems more likely that it qualifies the terrible diseases (\emph{pracaṇḍarug}) that are mentioned in the next verse quarter, as suggested by the reading \emph{jaṭharapravṛddha°}, which is attested by eta two and the \textit{Yogacintāmaṇi} and suggested by gamma one and delta one (\emph{jaṭharapravuddh°}).
+The reading \emph{jaṭharapracaṇḍa°} appears to be a dittographic error, which in some cases has also resulted in \emph{jaṭharapradaṇḍa°}.
 \end{philcomm}
@@
 \begin{sources}[hp01_028]
-Cf. Śivasaṃhitā
+Cf. \emph{Śivasaṃhitā} 3.108
 
@@
 \tl{prasārya caraṇadvandvaṃ parasparasusaṃyutam |\\+}
-\tl{svapāṇibhyāṃ dṛḍhaṃ dhṛtvā jānūpari śiro nyaset || 3.108 ||\\!}
+\tl{svapāṇibhyāṃ dṛḍhaṃ dhṛtvā jānūpari śiro nyaset ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_028]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.66
 
@@
 \tl{prasārya pādau bhuvi daṇḍarūpau\\+}
-\tl{dorbhyāṃ padāgradvitayaṃ gṛhītvā ||\\+}
+\tl{dorbhyāṃ padāgradvitayaṃ gṛhītvā |\\+}
 \tl{jānūpari nyastalalāṭadeśo\\+}
-\tl{vased idaṃ paścimatānam āhuḥ || 3.66 ||\\!}
+\tl{vased idaṃ paścimatānam āhuḥ ||\\+}
+\tl{\var{dorbhyāṃ padāgradvitayaṃ ] dvābhyāṃ karābhyāṃ dvitayaṃ n1,n3}\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 6.22 (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{prasārya pādau bhuvi daṇḍarūpau\\+} 
+\tl{dorbhyāṃ ca pādadvitayaṃ gṛhītvā |\\+}
+\tl{jānūpari nyastalalāṭapaṭṭo\\+} 
+\tl{nyased idaṃ paścimatānam āhuḥ ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_028]
-Only V1 has \emph{dorbhyāṃ padāgra}, others have variations on the much inferior \emph{dvābhyāṃ karābhyāṃ}.
-The karābhyaṃ reading was probably introduced to remove dorbhyāṃ, which usually means arms.The reading \emph{tāṇabandhaḥ} in V1 doesn’t work with \emph{idaṃ}.
+The reading \emph{dorbhyāṃ padāgradvitayaṃ} is well attested but is somewhat strange because \emph{dos} usually means `the arm' rather than the hands.
+Hence, the other variation `with both hands' (\emph{dvābhyāṃ karābhyāṃ dvitayaṃ}), which appears to be an attempt to remove \emph{dorbhyāṃ}, is inferior because the thing the hands are holding, that is the toes (\emph{padāgra}), is not explicitly stated in this version of the verse.
 \end{philcomm}
@@
 \begin{translation}[hp01_029]
-This back-stretch is the foremost among āsanas. It makes the breath flow in the back [i.e. Central channel], increases the digestive fire, makes the belly thin and prevents diseases in men.
+This back-stretch is the foremost among āsanas. It makes the breath flow in the back (i.e., the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
 \end{translation}
@@
 \begin{sources}[hp01_029]
-Cf. Śivasaṃhitā
-
-\begin{versinnote}
-\tl{Āsanāgryam idaṃ proktaṃ jaṭharānaladīpanam |\\+}
-\tl{dehāvasādaharaṇaṃ paścimottānasaṃjñakam || 3.109 ||\\!}
+Cf. \emph{Śivasaṃhitā} 3.109
+
+\begin{versinnote}
+\tl{āsanāgryam idaṃ proktaṃ jaṭharānaladīpanam |\\+}
+\tl{dehāvasādaharaṇaṃ paścimottānasaṃjñakam ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_029]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{iti paścimatānam āsanāgryaṃ
-pavanaṃ paścimavāhinaṃ karoti |\\+}
-\tl{udayaṃ jaṭharānalasya kuryād
-udare kārśyam arogitāṃ ca puṃsām ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{iti paścimatānam āsanāgryaṃ pavanaṃ paścimavāhinaṃ karoti ||\\+}
-\tl{udayaṃ jaṭharānalasya kuryād udare kārśyam arogatāṃ ca puṃsām || 3.67 ||\\!}
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{iti paścimatānam āsanāgryaṃ\\+}
+\tl{pavanaṃ paścimavāhinaṃ karoti |\\+}
+\tl{udayaṃ jaṭharānalasya kuryād\\+}
+\tl{udare kārśyam arogitāṃ ca puṃsām ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 3.67
+
+\begin{versinnote}
+\tl{iti paścimatānam āsanāgryaṃ\\+} 
+\tl{pavanaṃ paścimavāhinaṃ karoti |\\+}
+\tl{udayaṃ jaṭharānalasya kuryād\\+} 
+\tl{udare kārśyam arogatāṃ ca puṃsām ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_029]
-Adopt \emph{arogitāṃ} (which is well attested), despite the fact that \emph{arogatāṃ} is much more common.
  The use of the word \emph{paścima} to mean the central channel is found in \emph{Yogabīja} 121.
-The  \emph{Yogabīja} (95) also refers to the path of the central channel (\emph{paścimamārga}), and this understanding of \emph{paścima} is found in the \emph{Jyotsnā 1.29}: \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumnāmārgeṇa vahatīti paścimavāhī}.
+The  \emph{Yogabīja} (95) also refers to the path of the central channel (\emph{paścima\-mārga}), and this understanding of \emph{paścima} is found in the \emph{Jyotsnā 1.29}: \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumnāmārgeṇa vahatīti paścimavāhī}.
 \end{philcomm}
@@
 \begin{translation}[hp01_030]
-Supporting oneself on the ground with both hands, their elbows placed on either side of the navel, with a raised position (?
-uccāsanaḥ) one is placed up into the air [as straight] as a stick.
+Supporting oneself on the ground with both hands, [one's] elbows are placed on either side of the navel [and], with a raised position (\emph{uccāsanaḥ}), one is lifted up into the air [as straight] as a stick.
 They call this posture the peacock.
@@
 \begin{sources}[hp01_030]
-Vimānārcanākalpa 96
-
-\begin{versinnote}
-\tl{karatale bhūmau saṃsthāpya kūrparau nābhipārśvayor nyasya nataśirāḥ (unnataśirāḥ) pādau ḍaṇḍavad vyomni saṃsthito mayūrāsanam iti |\\!}
-\end{versinnote}
-
-Pādmasaṃhitā yogapāda 1.21c–22d:
+\emph{Vimānārcanākalpa} 96
+
+\begin{versinnote}
+\tl{karatale bhūmau saṃsthāpya kūrparau nābhipārśvayor nyasya nataśirāḥ (unnataśirāḥ) pādau ḍaṇḍavad vyomni saṃsthito mayūrāsanam iti ||\\!}
+\end{versinnote}
+
+\emph{Pādmasaṃhitā} (\emph{yogapāda}) 1.21c–22d:
 
@@
 
-Ahirbudhnyasaṃhitā 31.36–37
+\emph{Ahirbudhnyasaṃhitā} 31.36–37
 
@@
 \tl{niveśya kūrparau samyaṅ nābhimaṇḍalapārśvayoḥ |\\+}
-\tl{avaṣṭabhya bhuvaṃ pāṇitalābhyāṃ vyomni daṇḍavat || 31--36 ||\\!}
-\end{versinnote}
-
-Vasiṣṭhasaṃhitā 1.76–77
+\tl{avaṣṭabhya bhuvaṃ pāṇitalābhyāṃ vyomni daṇḍavat ||\\!}
+\end{versinnote}
+
+\emph{Vasiṣṭhasaṃhitā} 1.76–77
 
@@
 \tl{samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\+}
-\tl{mayūrāsanam etad dhi sarvapāpavināśanam ||\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.15–16
-
-\begin{versinnote}
-\tl{avaṣṭabhya dharāṃ samyak talābhyāṃ tu karadvayoḥ\\+}
-\tl{hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ\\+}
+\tl{mayūrāsanam etad dhi sarvapāpavināśanam ||\\+}
+\tl{\var{ca karadvayam ] karayor dvayoḥ}\\!}
+\end{versinnote}
+
+\emph{Yogayājñavalkya} 3.15–16
+
+\begin{versinnote}
+\tl{avaṣṭabhya dharāṃ samyak talābhyāṃ tu karadvayoḥ |\\+}
+\tl{hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ ||\\+}
 \tl{samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\+}
@@
 \begin{testimonia}[hp01_030]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{dharām avaṣṭabhya punaḥ karābhyāṃ
-tat kūrpare sthāpitanābhipārśvaḥ |\\+}
-\tl{tadāsane daṇḍavad utthitaḥ khe
-mayūram etat pravadanti santaḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.42
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{dharām avaṣṭabhya punaḥ karābhyāṃ\\+}
+\tl{tatkūrpare sthāpitanābhipārśvaḥ |\\+}
+\tl{tadāsane daṇḍavad utthitaḥ khe\\+}
+\tl{mayūram etat pravadanti santaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 3.42
 
@@
 \tl{atha mayūram\\+}
-\tl{dharām avaṣṭabhya karadvayena tatkūrpare sthāpitanābhipārśvaḥ ||\\+}
-\tl{uccāsano daṇḍavad utthitaḥ khe mayūram etat pravadanti pīṭham ||\\!}
+\tl{dharām avaṣṭabhya karadvayena\\+}
+\tl{tatkūrpare sthāpitanābhipārśvaḥ |\\+}
+\tl{uccāsano daṇḍavad utthitaḥ khe\\+}
+\tl{mayūram etat pravadanti pīṭham ||\\!}
 \end{versinnote}
@@
 The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77).
-The compound \emph{uccāsano} seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}, which may be derived from earlier Vaiṣṇava sources, such as the \textit{Pādmasaṃhitā}.
-
-1.30b tat refers to \emph{karadvaya} (cf. Vasiṣṭhasaṃhitā).
+The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}, which may be derived from earlier Vaiṣṇava sources, such as the \textit{Pādmasaṃhitā}.
+
+In the second verse quarter, the pronoun in \emph{tatkūrpare} refers to the two hands (\emph{karadvaya}).
+This is stated more explicitly (i.e., \emph{hastayoḥ kūrparau}) in \emph{Vasiṣṭhasaṃhitā} 1.76c and \emph{Yogayājñavalkya} 3.15c.
 \end{philcomm}
@@
 \begin{translation}[hp01_031]
-The glorious mayūra posture gets rid of all diseases of the abdomen such as bloating and overcomes humoral imbalances.
+The glorious peacock posture gets rid of all diseases of the abdomen, such as bloating, and overcomes humoral imbalances.
 It completely incinerates food which is bad or has been eaten to excess, it generates digestive fire and it digests strong poison.
@@
 \begin{testimonia}[hp01_031]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī 3.43
+\emph{Haṭharatnāvalī} 3.43
 
@@
 
-\begin{philcomm}[hp01_031]
-\emph{aśeṣaṃ} better
-\end{philcomm}
+%\begin{philcomm}[hp01_031]
+%\end{philcomm}
 
@@
 \begin{translation}[hp01_032]
-Lying on one’s back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] any āsana and relaxes the mind.
+Lying on one's back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] any \emph{āsana} and relaxes the mind.
 \end{translation}
@@
 \begin{sources}[hp01_032]
-Cf. Dattātreyayogaśāstra 24cd
+Cf. \emph{Dattātreyayogaśāstra} 24cd
 
@@
 \begin{testimonia}[hp01_032]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī 3.76
+\emph{Haṭharatnāvalī} 3.76
 
@@
 \tl{athāntimaṃ śavāsanam\\+}
-\tl{prasārya hastapādau ca viśrāntyā śayanaṃ tathā ||\\+}
-\tl{sarvāsanaśramaharaṃ śayitaṃ tu śavāsanam || 3.76 ||\\!}
-\end{versinnote}
-
-Cf. HTK
+\tl{prasārya hastapādau ca viśrāntyā śayanaṃ tathā |\\+}
+\tl{sarvāsanaśramaharaṃ śayitaṃ tu śavāsanam ||\\!}
+\end{versinnote}
+
+Cf. \emph{Haṭhatattvakaumudī} 7.12
 
@@
 \tl{śavāsanaṃ hṛtkupitavātagranthivibhedakam |\\+}
-\tl{sarvāsanaśrāntijit hṛtśramaghnaṃ yogi saukhyadam || 12 ||\\!}
-\end{versinnote}
-
-YBD
-
-\begin{versinnote}
-\tl{uttānaṃ śavavad bhūmau śayanaṃ tu śavāsanam ||\\+}
+\tl{sarvāsanaśrāntijit hṛtśramaghnaṃ yogisaukhyadam ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 6.21
+
+\begin{versinnote}
+\tl{uttānaṃ śavavad bhūmau śayanaṃ tu śavāsanam |\\+}
 \tl{śavāsanaṃ śrāntiharaṃ cittaviśrāntikārakam ||\\+}
-\tl{iti śavāsanam || 21 ||\\!}
+\tl{iti śavāsanam ||\\!}
 \end{versinnote}
@@
 \begin{translation}[hp01_033]
-Śiva taught eighty-four āsanas. I shall take the four best and teach them.
+Śiva taught eighty-four \emph{āsana}s. I shall take the four best and teach them.
 \end{translation}
@@
 \begin{sources}[hp01_033]
-Śivasaṃhitā 3.96
+\emph{Śivasaṃhitā} 3.96
 
@@
 
-Cf. Dattātreyayogaśāstra 5
+Cf. \emph{Dattātreyayogaśāstra} 5
 
@@
 \tl{caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\+}
-\tl{ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||\\!}
-\end{versinnote}
-
-Cf. Vivekamārtaṇḍa 5
+\tl{ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtam ||\\!}
+\end{versinnote}
+
+Cf. \emph{Vivekamārtaṇḍa} 5
 
@@
 \tl{caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\+}
-\tl{ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||\\!}
+\tl{ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtam ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_033]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84v
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.23
 
@@
 \tl{caturaśīty āsanāni śivena kathitāni tu |\\+}
-\tl{tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham || 3.23 ||\\!}
+\tl{tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_033]
-In Pādas ac Svātmārāma may have rewritten Śivasaṃhitā 3.96 to include the information that it was Śiva (\emph{śivena}) who taught the 84 āsanas, whereas in the source Śiva is himself speaking.
-With this
-Svātmārāma confuses the verse, since one would have to understand Pādas cd to be direct speech by Śiva in order to understand \emph{bravīmi} correctly (but cf.
-4.2 which is a verse from the Gorakṣaśataka which includes pravakṣyāmi).
-
-The Śivasaṃhitā follows 3.96 with teachings on siddha, padma, paścimottāna and svastika āsanas, but the HP teaches siddha, padma, siṃha and bhadra.
-
-The word \emph{tu} is often used to introduce a new posture, but in this verse seems to be a verse filler.
+The word \emph{tu} is often used to introduce a new posture, but in this case seems to be a verse filler.
+
+In the first and third verse quarters, Svātmārāma may have rewritten \emph{Śivasaṃhitā} 3.96 to include the information that it was Śiva (\emph{śivena}) who taught the eighty-four \emph{āsana}s, whereas in the source Śiva is himself speaking.
+By doing this, Svātmārāma changes the meaning of the second half of the verse, as the \emph{Śivasaṃhitā} states that Śiva picked out the four best postures and taught them, whereas in the \emph{Haṭhapradīpikā} it reads as though Svātmārāma himself is responsible for picking out the four best postures and teaching them.
+There are other instances in the \emph{Haṭhapradīpikā} where Svātmārāma borrows a verse with a first person verb (e.g., 3.43, 4.2).
+However, in this instance, he may have intended to indicate that he chose the four postures coming after this verse (i.e., \emph{siddha}, \emph{padma}, \emph{siṃha} and \emph{bhadra}) because the \emph{Śivasaṃhitā} follows 3.96 with teachings on the postures called \emph{siddha}, \emph{padma}, \emph{paścimottāna} and \emph{svastika}.
+Another possibility is that Svātmārāma borrowed 1.33–1.34 from an unknown source that contained a dialogue that was different to that of \emph{Śivasaṃhitā}, as indicated by \emph{sakhe} in 1.34.
 \end{philcomm}
@@
 \begin{translation}[hp01_034]
-The adept, lotus, lion and auspicious pose are the best tetrad and, among those, always sit in the adept’s pose, my dear.
-\end{translation}
-
-\begin{sources}[hp01_034]
---
-\end{sources}
+The adept, lotus, lion and auspicious pose: these four are the best and, among those, always sit in the adept’s pose, my dear.
+\end{translation}
+
+%\begin{sources}[hp01_034]
+%--
+%\end{sources}
 
 \begin{testimonia}[hp01_034]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84v (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī 3.24
+\emph{Haṭharatnāvalī} 3.24
 
@@
 \tl{siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\+}
-\tl{śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā ||\\!}
+\tl{śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā ||\\+}
+\tl{\var{tathā ] satve P, sakhe T,t1}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_034]
-It is likely that the original version of this verse contained the vocative with the imperative form of the verb.
-There is another instance where Svātmārāma included a verse with the vocative (4.58?) or the subject in the first person (4.2), as though the text were a dialogue.
-In this case, it seems that efforts have been made to write out the vocative and imperative verb.
-However, the \emph{sukhe} and \emph{sukham} is difficult to construe in this context, because the context suggests that the intended meaning was that one should always sit in Siddhāsana (as opposed to the other three), rather than the prescription to always sit in a comfortable Siddhāsana.
+It is likely that the original version of this verse contained the vocative with the imperative form of the verb (\emph{sakhe tiṣṭha}).
+There are other instances where Svātmārāma included a verse with the vocative (e.g., 4.10, 4.12, 4.20, 4.72, 4.86, 4.88) as though the text were a dialogue.
+Other versions of this verse are transmitted by some manuscripts of the \emph{Haṭhapradīpikā}, in which the vocative and imperative verb have been removed.
+In these cases, the \emph{sukhe} and \emph{sukham} is difficult to construe because the context suggests that the intended meaning was that one should always sit in \emph{siddhāsana} (as opposed to the other three \emph{āsana}s), rather than the prescription to always sit in a comfortable \emph{siddhāsana}.
 \end{philcomm}
@@
 \begin{translation}[hp01_035]
+Now, the adept's pose (\emph{siddhāsana}).\\ 
 Having joined the place of the perineum with the heel of the foot, the yogi should firmly fix the [other] foot on the penis.
@@
 \end{translation}
-% [MD] add translation of "pūrvoktam eva matsaṃmataṃ"
 
 \begin{sources}[hp01_035]
-Vivekamārtaṇḍa 7
-
-\begin{versinnote}
-\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyase[n]\\+}
+\emph{Vivekamārtaṇḍa} 7
+
+\begin{versinnote}
+\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
 \tl{meḍhre pādam athaikam āsyahṛdaye dhṛtvā samaṃ vigraham |\\+}
 \tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
-\tl{etan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\!}
+\tl{etan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\+}
+\tl{\var{7a °mūlaghaṭitaṃ ] °mulaghaṭanaṃ A • vinyasen ] GHT; vinyase VA, vinyaset Y}\\+}
+\tl{\var{7b meḍhre ] meṃḍhre A • athaikam eva hṛdaye ] H; athaikadeśahṛdayo V, athaikam eva niyataṃ AGBGPkHvl, athaikam eva niṣataṃ GL, athaikam eva hṛdayaṃ GP, athaikam āsyahṛdaye HSHvl, athaikaṃ ekahṛdayo T, athaikam ekahṛdayaḥ Y • dhṛtvā ] VGBTHS; kṛtvā HGLGPGPkY • hanuṃ susthiram ] H; samaṃ vigraham VAGTHSY}\\+}
+\tl{\var{7c paśyed ] TH; paśyan VAGHSTvl, paśyad Y • antaraṃ ] VAGPkGLT; antare GBGP}\\+}
+\tl{\var{7d hy etan ] H; caitan VAGBGLGPT, etan GPkHS, tv etat Y • °janakaṃ ] °navidhau A, °jananaṃ GB, °nakaraṃ Y • procyate ] idaṃ bhavet Y (unm.)} ?? check\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_035]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 84v–85r (attr. \emph{Pavanayogasaṅgraha})
 
@@
 \tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
-\tl{meḍhre pādam athaikam ekahṛdaye kṛtvā samaṃ vigraham |\\+}
+\tl{meḍhre pādam athaikam ekahṛdayaḥ kṛtvā samaṃ vigraham |\\+}
 \tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyed bhruvor antaraṃ tv\\+}
@@
 
-Haṭharatnāvalī 3.25
+\emph{Haṭharatnāvalī} 3.25
 
@@
 \tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
-\tl{caitan mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\!}
+\tl{caitan mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\+}
+\tl{\var{yonisthānakam ] yonidvārakam P,T. niyataṃ ] hṛdaye T,t1,n2. °kapāṭa° ] °kavāṭa° P,T,t1}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_035]
-KDh ed has \emph{āsya}, variant in mss kh and gha, both BORI.
-Jürgen suggests we might take \emph{āsyahṛdaye} and \emph{vigraham} as the objects of \emph{dhṛtvā} and samam as an adverb (i.e., holding straight the face, chest and body).
-However, in a different doctrine.
+The second verse quarter has many variations in the \emph{Haṭhapradīpikā} manuscripts, as well as the manuscripts of the sources and testimonia.
+The alpha manuscripts have \emph{athaikam ekahṛdaye dhṛtvā} and most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}.
+The repetition of \emph{eka} appears to be a dittographical error and \emph{hṛdaye} does not make sense without an object of \emph{dhṛtvā} (i.e., having held what on the chest?).
+In the adopted reading \emph{athaikam āsyahṛdaye dhṛtvā}, which is attested by eta one, the oldest dated manuscript, the compound \emph{āsyahṛdaye} can be understood with the gerund \emph{dhṛtvā} and \emph{ekam} in the sense of `having held the face and chest together.' This alludes to the practice of Jālandhara's lock, in which the chin is placed on the chest.
+This meaning is more clearly seen in the \emph{Jyotsnā}'s version of the second verse quarter, \emph{hṛdaye kṛtvā hanuṃ susthiram} (`having put the jaw formly on the chest').
+The compound \emph{āsyahṛdaye} is supported by other manuscripts, such as \getsiglum{N10,P1,P6,J16} and the variants \emph{asyahṛdaye} and \emph{asyahṛdayaṃ} also occur (e.g., \getsiglum{J7,J12,A1,V4,V18,V16,P8,P9}).
+Another possible reading is \emph{ekahṛdayo}, which is not supported by manuscripts of the \emph{Haṭhapradīpikā} in the most important stemmatic groups but is in two other manuscripts (\getsiglum{J2,M1}), the six-chapter \emph{Vivekamārtaṇḍa} and \emph{Yogacintāmaṇi}, which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}.
+In this case, \emph{ekahṛdayaḥ} would qualify the yogi as having a mind focused on one thing.
+The other well-attested reading, \emph{athaikaṃ eva niyataṃ}, was an attempt to fix the problem of \emph{hṛdaye} by replacing it with \emph{niyatam}, which must be read with \emph{meḍhre pādam athaikaṃ} (`having fixed one foot on the penis').
+But \emph{niyatam} is redundant here because of \emph{vinyaset} in the first verse quarter.
 \end{philcomm}
@@
 \begin{translation}[hp01_036]
-Having fixed the left heel on the penis, and put the other heel on that, this is siddhāsana.
+However, in another view:\\
+Having fixed the left heel on the penis, and put the other heel on that, this is \emph{siddhāsana}.\\ 
+Only the first teaching [on \emph{siddhāsana}] is accepted by me.
 \end{translation}
@@
 \begin{sources}[hp01_036]
-Vasiṣṭhasaṃhitā 1.81
+\emph{Vasiṣṭhasaṃhitā} 1.81
 
@@
 
-Yogayājñavalkya 3.15
+\emph{Yogayājñavalkya} 3.15
 
@@
 \begin{testimonia}[hp01_036]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85r (attr. \emph{Pavanayogasaṅgraha})
 
@@
 
-Haṭharatnāvalī 3.26
+\emph{Haṭharatnāvalī} 3.26
 
@@
 
+\begin{philcomm}[hp01_036]
+Svātmārāma's introductory and following remarks to verse 1.36 indicate that he preferred the \emph{siddhāsana} of the \emph{Vivekamārtaṇḍa} over the version taught as \emph{muktāsana} in the \emph{Vasiṣṭhasaṃhitā} and \emph{Yogayājñavalkya}.
+\end{philcomm}
 %%%%%%%%%%
@@
 \begin{translation}[hp01_037]
-Some proclaim this is siddhāsana, others know it as vajrāsana, a few say it is muktāsana and some guptāsana.
-\end{translation}
-
-\begin{sources}[hp01_037]
---
-\end{sources}
+Some proclaim this is \emph{siddhāsana}, others know it as \emph{vajrāsana}, a few say it is \emph{muktāsana} and some \emph{guptāsana}.
+\end{translation}
+
+%\begin{sources}[hp01_037]
+%--
+%\end{sources}
 
 \begin{testimonia}[hp01_037]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\+}
-\tl{śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\+}
-\tl{śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā || 3.24 ||\\!}
-\end{versinnote}
-
-Cf. Śivayogasāram by Kolani Ganapatideva (date 14th c.)
+\emph{Yogacintāmaṇi} f. 85r (attr. \emph{Pavanayogasaṅgraha})
+
+\begin{versinnote}
+\tl{etat siddhāsanaṃ prāhuḥ padmāsanam atho viduḥ |\\+}
+\tl{guptāsanaṃ vadantyeke prāhur vajrāsanaṃ pare |\\+}
+\tl{ke cin muktāsanaṃ prāhur idam āsanam uttamam ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 3.27
+
+\begin{versinnote}
+\tl{etat siddhāsanaṃ prāhur anye vajrāsanaṃ viduḥ |\\+}
+\tl{muktāsanaṃ vadanty eke prāhur guptāsanaṃ par ||\\!}
+\end{versinnote}
+
+Cf. \emph{Śivayogasāram} by Kolani Ganapatideva (date 14th c.)
 
@@
 \begin{translation}[hp01_038]
-The Siddhas know Siddhasana as the single most important amongst
-all postures, in the same way as measured diet amongst rules and non-violence amongst observances.
+The Siddhas know \emph{siddhasana} as the single most important [pose] amongst all postures, in the same way as a measured diet amongst rules and non-violence amongst observances.
 \end{translation}
@@
 \begin{sources}[hp01_038]
-Cf. Dattātreyayogaśāstra 33
+Cf. \emph{Dattātreyayogaśāstra} 33
 
@@
 \begin{testimonia}[hp01_038]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85r (attr. \emph{Haṭhapradīpikā})
 
@@
 \begin{philcomm}[hp01_038]
-\emph{iva} or \emph{eva}?
+%\emph{iva} or \emph{eva}?
 \emph{iva} does work — like \emph{siddhāsana}, \emph{mitāhāra} and \emph{ahiṃsā} are the best, but for it to work properly \emph{mitāhāra} and \emph{ahiṃsā} should be accusative.
@@
 However, one old KDham BORI (?) ms (pha, 1695 CE) has it, as does Jyotsnā, so adopt.
-
-In pāda d V19 has \emph{siddhāsanam idaṃ viduḥ}, but the reading of all other mss is preferable.
-
-Clearly based on DYŚ 33, which includes \emph{ekaṃ} and \emph{mukhya}.
+%In pāda d V19 has \emph{siddhāsanam idaṃ viduḥ}, but the reading of all other mss is preferable.
+% JB: The reading we have adopted is well attested by the new collation of manuscripts. Therefore, I've ignored the above comments.
+
+%Clearly based on DYŚ 33, which includes \emph{ekaṃ} and \emph{mukhya}.
 \end{philcomm}
@@
 \begin{translation}[hp01_039]
-Among the eighty-four postures, one should regularly practise just Siddha; in the same way one should practise Suṣumnā among the 72,000 channels.
-\end{translation}
-
-\begin{sources}[hp01_039]
-\end{sources}
+Among the eighty-four postures, one should regularly practise just \emph{siddhāsana}, in the same way one should practise Suṣumnā among the 72,000 channels.
+\end{translation}
+
+%\begin{sources}[hp01_039]
+%\end{sources}
 
 \begin{testimonia}[hp01_039]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-YSS
+\emph{Yogasārasaṅgraha} p.9 (attr. \emph{Yogasāramañjarī})
 
@@
 
-Yogacintāmaṇi (in a different passage)
-​​
+\emph{Yogacintāmaṇi} f. 79r (attr. \emph{Haṭhayoga})
+
 \begin{versinnote}
@@
 
-Cf. Kumbhakapaddhati (effects of practising kumbhaka)
+Cf. \emph{Kumbhakapaddhati} 120 (on the effects of practising \emph{kumbhaka})
 
@@
 \tl{dvāsaptati sahāsrāṇāṃ nāḍīnāṃ malaśodhanam |\\+}
-\tl{yatheṣṭaṃ dhāraṇaṃ vāyor vikārābhāva eva ca || 120 ||\\!}
+\tl{yatheṣṭaṃ dhāraṇaṃ vāyor vikārābhāva eva ca ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_039]
-Odd to have \emph{suṣumnām} as object of \emph{abhyaset}.
-Yogacintāmaṇi has this reading though.
-Perhaps cd were added somewhat indiscriminately by Svātmārāma (with nominative \emph{suṣumnā}) and then others tried to make sense of it.
-
-Some witnesses, including Jyotsnā, have \emph{nāḍīnāṃ malaśodhanam/e} for pāda d, which is probably a patch (no other texts say siddhāsana clears the channels), but cf.
-Amṛtasiddhi in which the practices are said to bring about cālana of the nāḍīs (e.g.
-11.6).
-
-Good example of early contamination.
-
-[\emph{nāḍiṣu} is better supported (J10ac,V19,J17).]
+It is odd to have \emph{suṣumnām} as the object of the verb \emph{abhyaset}.
+This reading is well attested by manuscripts of the \emph{Haṭhapradīpikā}, as well as the verse cited in the \emph{Yogacintāmaṇi}, which attributes it to the \emph{Yogasāramañjarī}.
+Perhaps, the second hemistich was added somewhat haphazardly by Svātmārāma, and then others have tried to make sense of it by changing \emph{suṣumnām iva nāḍiṣu} to \emph{nāḍīnāṃ malaśodhanam}, which occurs in the \emph{Jyotsnā} (1.39).
+The reading \emph{nāḍīnāṃ malaśodhanam/e} is probably a patch as no other texts say that \emph{siddhāsana} clears the channels.
+However, the idea of purifying the channels can be found in other contexts (e.g., \emph{Kumbhakapaddhati} 120) and may hark back to an earlier notion of flushing (\emph{cālana}) the channels (e.g., \emph{Amṛtasiddhi} 11.6).
+
+%Good example of early contamination.
+%[\emph{nāḍiṣu} is better supported (J10ac,V19,J17).]
 \end{philcomm}
@@
 \begin{translation}[hp01_040]
-By meditating upon the self, restricting the diet and regularly practising Siddhāsana for twelve years, the yogi attains the Niṣpatti stage.
-What’s the point of lots of exhausting postures when there is Siddhāsana?
-\end{translation}
-
-\begin{sources}[hp01_040]
---
-\end{sources}
+By meditating upon the self, restricting the diet and regularly practising \emph{siddhāsana} for twelve years, the yogi attains the \emph{niṣpatti} stage.
+What’s the point of lots of exhausting postures when there is \emph{siddhāsana}?
+\end{translation}
+
+%\begin{sources}[hp01_040]
+%--
+%\end{sources}
 
 \begin{testimonia}[hp01_040]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-YSS
+\emph{Yogasārasaṅgraha} p. 9 (attr. \emph{Yogasāramañjarī})
 
@@
 \begin{philcomm}[hp01_040]
-Only possible variant is \emph{mitāhāro} in V19.
-J8 might be correction of J10’s unmetrical reading.
-
-[\emph{sadāsiddhāsanābhyāsād}? Or maybe read \emph{sadā} with \emph{avāpnuyāt}]
-
-V1 has
-
-\begin{versinnote}
-\tl{śramādau bahubhiḥ pīṭhais sadā siddhāsane sati |\\+}
-\tl{prāṇānile sāvadhānaṃ baddhe kevalakumbhake || 1.41 ||\\!}
-\end{versinnote}
-
-Mixing up both versions of the verse — contamination already?
-
-V19 is found in Yogacintāmaṇi: \emph{śramadair bahubhịh pīṭhaiḥ kiṃ syāt siddhāsane sati}; JM this seems best to me.
-
-Is this notion of āsanas causing \emph{śrama} already current in HY texts?
-[Yes, it is mentioned above in the verse on śavāsana]
+% Only possible variant is \emph{mitāhāro} in V19.
+% J8 might be correction of J10’s unmetrical reading.
+
+%[\emph{sadāsiddhāsanābhyāsād}? Or maybe read \emph{sadā} with \emph{avāpnuyāt}]
+
+% V1 has:
+
+%\begin{versinnote}
+%\tl{śramādau bahubhiḥ pīṭhais sadā siddhāsane sati |\\+}
+%\tl{prāṇānile sāvadhānaṃ baddhe kevalakumbhake || 1.41 ||\\!}
+%\end{versinnote}
+
+%Mixing up both versions of the verse — contamination already?
+
+%V19 is found in Yogacintāmaṇi: \emph{śramadair bahubhịh pīṭhaiḥ kiṃ syāt siddhāsane sati}; JM this seems best to me.
+
+The notion of \emph{āsana}s causing fatigue \emph{śrama} was already suggested above in the verse on corpse pose (\emph{śavāsana}).
 \end{philcomm}
@@
 \begin{translation}[hp01_041]
-Just as the Unmanī [state] arises automatically, without effort, when the prāṇa breath has been carefully stopped in kevala kumbhaka, [...]
-\end{translation}
-
-\begin{sources}[hp01_041]
-\end{sources}
+Just as the [state] beyond mind (\emph{unmanī}) arises automatically, without effort, when the \emph{prāṇa} breath has been carefully stopped in \emph{kevalakumbhaka}, [...]
+\end{translation}
+
+%\begin{sources}[hp01_041]
+%\end{sources}
 
 \begin{testimonia}[hp01_041]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-\begin{philcomm}[hp01_041]
---
-\end{philcomm}
+%\begin{philcomm}[hp01_041]
+%--
+%\end{philcomm}
 
@@
 \begin{translation}[hp01_042]
-So too the three bandhas arise automatically, without effort, when Siddhāsana alone is always firmly adopted,
-\end{translation}
-
-\begin{sources}[hp01_042]
---
-\end{sources}
+[...] so too the three locks (\emph{bandha}) arise automatically without effort, every time \emph{siddhāsana} alone is firmly adopted.
+\end{translation}
+% adopt dṛḍhaṃ
+
+%\begin{sources}[hp01_042]
+%--
+%\end{sources}
 
 \begin{testimonia}[hp01_042]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-YSS
-
-\begin{versinnote}
-\tl{tathaikasminn eva baddhe siddhāsane sadā |\\+}
-\tl{granthitrayam anāyāsāt svayamevopabhidyate |\\!}
-\end{versinnote}
+%YSS
+%\begin{versinnote}
+%\tl{tathaikasminn eva baddhe siddhāsane sadā |\\+}
+%\tl{granthitrayam anāyāsāt svayamevopabhidyate |\\!}
+%\end{versinnote}
 
@@
 \begin{philcomm}[hp01_042]
-\emph{dṛḍhe} goes across stemma.
-It is likely that the adverb (\emph{dṛdhaṃ}) was intended because \emph{dṛdhataraṃ}, which is not ambiguous, is used in 1.48a to qualify how \emph{padmāsana} should be adopted.
+It seems likely that \emph{dṛḍhaṃ} (rather than \emph{dṛḍhe}) was originally intended in 1.42a because \emph{dṛḍhataraṃ}, which is not ambiguous, is used in 1.48a to qualify how \emph{padmāsana} should be adopted.
+Also, \emph{dṛḍhaṃ} complements \emph{sāvadhānaṃ} in 1.41a.
+% J5 and G4 have dṛḍhaṃ (alpha 2 and 3), + P11 (alpha and beta)
 \end{philcomm}
@@
 \begin{translation}[hp01_043]
-There is no posture like siddhāsana, no breath like the restrained breath, no mudrā like khecarī, no dissolution like nāda.
+There is no posture like \emph{siddhāsana}, no retention like \emph{kevala}, no seal like \emph{khecarī}, [and] no dissolution [of mind] like the one on the internal sound (\emph{nāda}).
 \end{translation}
@@
 \begin{sources}[hp01_043]
-Śivasaṃhitā 5.47
+\emph{Śivasaṃhitā} 5.47
 
@@
 \tl{nāsanaṃ siddhasadṛśaṃ na kumbhasadṛśaṃ balam |\\+}
-\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ || 5.47 ||\\!}
+\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_043]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 75r (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.29
 
@@
 \tl{nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ ||\\+}
-\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ || 3.29 ||\\!}
+\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ ||\\+}
+\tl{\var{kumbhaḥ kevalopamaḥ ] kumbhasadṛśo 'nilaḥ N,n1,n2,n3,J}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_043]
-\emph{na kumbhasadṛśo nilaḥ} is difficilior lectio and attested by all early witnesses except V19: (almost?) all testimonia have \emph{kumbhaḥ kevalopamaḥ}; Śivasaṃhitā has \emph{kumbhasadṛśaṃ balam}.
+The reading \emph{na kumbhasadṛśo 'nilaḥ} (`no breath like a retention') is the lectio difficilior and attested by two early witnesses (eta one and two).
+It could be original.
+However, the alpha manuscripts and several other important groups of manuscripts have \emph{kumbhaḥ kevalopamaḥ}, as well as the \emph{Yogacintāmaṇi} and some manuscripts of the \emph{Haṭharatnāvalī}, suggesting that this reading, which makes much better sense, was in the transmission at an early stage.
+% Śivasaṃhitā has \emph{kumbhasadṛśaṃ balam}.
+%
 \end{philcomm}
@@
 \begin{translation}[hp01_044]
-Now lotus pose.
+Now, lotus pose (\emph{padmāsana}).\\
 One should place the right foot on the left thigh, and the left on the right though, hold firmly the big toes with the hands behind the back, place the chin on the chest and gaze at the tip of the nose.
-This is called lotus pose, which destroys diseases for those who undertake the yamas.
+This is called lotus pose, which destroys diseases for those who undertake the observances (\emph{yama}).
 \end{translation}
@@
 \begin{sources}[hp01_044]
-Vivekamārtaṇḍa 8
+\emph{Vivekamārtaṇḍa} 8
 
@@
 \begin{testimonia}[hp01_044]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85v (attr. \emph{Haṭhayoga})
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.34
 
@@
 \tl{aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\+}
-\tl{etad vyādhivināśakāri yamināṃ padmāsanaṃ procyate || 3.34 ||\\!}
+\tl{etad vyādhivināśakāri yamināṃ padmāsanaṃ procyate ||\\!}
 \end{versinnote}
@@
 \begin{translation}[hp01_045]
-However, in another view,
-
+However, in another view:\\
 Having put the upturned feet carefully on the thighs and the upturned hands in the middle of the thighs, one should fix the eyes on the tip of the nose.
-Having raised the root of the uvula with the tongue, one should place the chin on the chest and gradually [draw in\footnote{The verb \emph{ākṛśya} follows in the next verse in the \emph{Dattātreyayogaśāstra}.}] the breath [...].
+Having raised the root of the uvula with the tongue, one should place the chin on the chest and gradually [draw in] the breath [...].
 \end{translation}
@@
 \begin{sources}[hp01_045]
-Dattātreyayogaśāstra 35–37
+\emph{Dattātreyayogaśāstra} 35–37
 
@@
 \tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\+}
-\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 35 ||\\+}
+\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau ||\\+}
 \tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
-\tl{uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\+}
+\tl{uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ ||\\+}
 \tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
-\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\!}
-\end{versinnote}
-
-Śivasaṃhitā 3.102–104
-
-\begin{versinnote}
-\tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ\\+}
-\tl{ūrumadhye tathottānau pāṇī kṛtvā tu tādṛśau 3.102\\+}
-\tl{nāsāgre vinyased dṛṣṭiṃ rājadantaṃ ca jihvayā\\+}
-\tl{uttambhya cibukaṃ vakṣe saṃsthāpya pavanaṃ śanaiḥ 3.103\\+}
-\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ\\+}
-\tl{yathāśaktyaiva paścāt tu recayed anirodhataḥ 3.104\\!}
+\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ ||\\!}
+\end{versinnote}
+
+\emph{Śivasaṃhitā} 3.102–104
+
+\begin{versinnote}
+\tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\+}
+\tl{ūrumadhye tathottānau pāṇī kṛtvā tu tādṛśau ||\\+}
+\tl{nāsāgre vinyased dṛṣṭiṃ rājadantaṃ ca jihvayā |\\+}
+\tl{uttambhya cibukaṃ vakṣe saṃsthāpya pavanaṃ śanaiḥ ||\\+}
+\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
+\tl{yathāśaktyaiva paścāt tu recayed anirodhataḥ ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_045]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85v
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.36–3.37
 
@@
 \tl{uttānau caraṇau kṛtvā ūrvoḥ saṃsthāpya yatnataḥ |\\+}
-\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 3.36 ||\\+}
+\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau ||\\+}
 \tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
-\tl{uttabhya cibukaṃ vakṣaḥ saṃsthāpya pavanaṃ śanaiḥ || 3.37 ||\\!}
+\tl{uttabhya cibukaṃ vakṣaḥ saṃsthāpya pavanaṃ śanaiḥ ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_045]
-\emph{uttabhya} vs \emph{uttambhya}.
-The witnesses split predictably along the two main branches of the stemma. The evidence of the DYŚ is important here.
+%Both \emph{uttabhya} and \emph{uttambhya} are well attested
+%The witnesses split predictably along the two main branches of the stemma. 
 %\emph{vakṣasyāsthāpya} is a marmasthāna
 
-The statement ending with \emph{pavanaṃ śanaiḥ} is left hanging, perhaps, because of an eyeskip that happened early in the transmission.
-The subsequent verse in the DYŚ ends with \emph{pavanaṃ śanaiḥ}.
+The main problem with this passage is the incomplete syntax of verse 1.46, which describes the position of the eyes, tongue, chin and chest in lotus pose.
+The words `the breath gradually' (\emph{pavanaṃ śanaiḥ}) are left hanging.
+The evidence of the source text, the \emph{Dattātreyayogaśāstra}, is important here because it includes a verse that makes sense of \emph{pavanaṃ śanaiḥ}.
+The verse that completes the syntax is missing in the \emph{Haṭhapradīpikā}, perhaps because of a scribal error that happened early in the transmission.
+As seen in the \emph{Dattātreyayogaśāstra} verses 36 and 37 both end with \emph{pavanaṃ śanaiḥ}, the repetition of which likely caused an eyeskip.
 
@@
 \tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
-\tl{uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\+}
+\tl{uttabhya cibukaṃ vakṣasy āsthāpya \textbf{pavanaṃ śanaiḥ} || 36 ||\\+}
 \tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
-\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\+}
+\tl{yathāśaktyaiva paścāt tu recayet \textbf{pavanaṃ śanaiḥ} || 37 ||\\+}
 \tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\!}
@@
 
-The manuscript readings with \emph{vakṣa sthāpayet} (J7, V3, J8, J10, J17, N17) or something similar (V1, W4) do not offer a solution nor indicate how Svātmārāma may have redacted this to make the syntax complete.
-Instead, it seems that he quoted these two verses (1.45–46) from the Dattātreyayogaśāstra and simply omitted the next verse that made sense of \emph{pavanaṃ śanaiḥ} because it was not relevant to the posture itself.
-
-% Jü:
-[JH] The background of the passage \emph{rājadantamūlaṃ ca jihvayā uttambhya} in 1.46 is more complex
-than it may appear. Here it is a literal quotation from the DYŚ, but many other Haṭhayogic texts
-teach a particular position of the tongue, in which it is in one or the other way turned back in
-the direction of the uvula, as we read explicitly in the \emph{Vivekamārtāṇḍa} (REF):
-\emph{kapālakuhare jihvā praviṣṭā viparītagā}. Brief references to this practice can become
-ambiguous for various reasons, and this has possibly confused Brahmānanda.
-
-One reason is that there is a, probably older, rule for meditation postures according to which the
-tongue rests near the teeth. One instance would be \emph{Svacchandatantra}, which teaches a
-meditation pose called \emph{divyaṃ karaṇam} (4.365f.), in which the tongue is to rest at the tip
-of the teeth (\emph{dantāgre jihvām ādāya}). Other Tantric texts have this or similar rules, in
-which the tongue is supposed to rest either on the teeth or the palate,\footnote{This rule is
-found in \emph{Iśānaśivapaddhati}: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān
-  asaṃspṛśan ṛjukāyaḥ}. REF} the earliest instance being perhaps \emph{Mrgendrāgama} ?.18.
-Placing the tongue where it does not disturb the meditation seems quite appropriate for a
-``normal'' meditative practice.
-
-[Jason: There’s also a clear reference to the tip of the tongue being placed in the middle of the palate in 2.27 of the Yogapāda of the Mataṅgapārameśvaratantra (tālumadhyagatenaiva jihvāgreṇa mahāmune).
-In fact, in works that predate Haṭhayoga (i.e., 11th c.), the most common instruction is to put the tip of the tongue on the palate (tālu).]
-
-We might try to interpret the passage in this manner, however, once Haṭhayogic physiology is at the
-background, we must assume that the aim is to reach back to the uvula, to the source of the
-"nectar".  For the background and for the crucial references see Mallinson's note on
-\emph{Khecarīvidyā} 1.65ab.\footnote{p. 209.}  Confusingly Yogic terminology has used and possibly
-invented new names for uvula, and among these especially the term \emph{rājadanta} may give rise to
-confusion, since, as we have seen, the tongue might also in some Yoga systems be placed at the
-(front) teeth.
-
-Furthermore the details in these descriptions of the \emph{khecarīmudrā} are manifold. The 10th
-century \emph{Mokṣopāya} says that the tongue rests at the ``source of the
-palate''\footnote{\emph{tālumūlatalālagnajihvā} MU V.55.14c.} and the commentary
-\emph{Saṃsārataraṇi} on the parallel passage in \emph{Laghuyogavāsiṣṭha} V.6.155, which reads
-\emph{tālumūlāntarālagnajihva-}, explains that this means that the tongue is to be placed in the
-middle of the two regions of the palate, and that this is the \emph{nabhomudrā}, alias
-``\emph{khecarī}''.\footnote{\emph{tālumūlāntarālagnajihvamūlah. tālumūlayoh. kākudamūladeśayoh. āntare
-  lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā | yā hi khecarı̄ty ucyate.}}
-
-A little later in the \emph{Mokṣopāya} it is made clear that one should reach the uvula, ``placed a
-the root of the palate''.\footnote{\emph{tālumūlagatāṃ yatnāj jihvayākramya ghaṇṭikām}
-(V.78.24ab)}
-
-In view of this background we must conclude that the author of the \emph{Jyotsnā} was probably not
-aware of the Yogic meaning of \emph{rājadanta} and has tried his best to make sense of the passage,
-echoing the idea of the two roots of the palate (although his text is not talking about the
-palate), but then wisely refers to the instruction of the teacher for practical details, probably
-noticing that his literal interpretation is somewhat opaque (I omit the synonyms for clarity):
-
+The manuscript readings with \emph{vakṣa sthāpayet} (\getsiglum{J7,V3,J8,J10,J17,N17}) or something similar (\getsiglum{V1,W4}) do not offer a solution to the incomplete syntax nor indicate that Svātmārāma rewrote \emph{Dattātreyayogaśāstra} 36 so that he could omit \emph{Dattātreyayogaśāstra} 37.
+For, apart from the possibility of an eyeskip, it is also conceivable that Svātmārāma intended to quote only the two verses of the \emph{Dattātreyayogaśāstra} (35–36) that describe the physical posture of \emph{padmāsana} and omit the next verse (37) that made sense of \emph{pavanaṃ śanaiḥ} because it was not relevant to the posture itself.
+Be this as it may, in the absence of evidence that Svātmārāma included \emph{Dattātreyayogaśāstra} 37 or wrote a coherent version of \emph{Haṭhapradīpikā} 1.46, we have tried to make sense of \emph{pavanaṃ śanaiḥ} by supplying the gerund \emph{samākṛṣya} in our translation.
+
+Brahmānanda's comment on the statement, `having raised the root of the uvula with the tongue' (\emph{rājadantamūlaṃ ca jihvayā uttabhya}) in 1.46 is worth noting.
+In the context of Haṭhayoga, one would assume this statement to be referring to a type of  \emph{khecarīmudrā}, in which the tongue lifts the root of the uvula, here called the `royal tooth' (\emph{rājadanta}, on the meaning of which see Mallinson 2007: 209 n.
+258).
+However, Brahmānanda understands it differently (synonyms omitted for clarity):
 \begin{quote}
-Pressing both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.
-
-\emph{rājadantānāṃ daṃṣṭrāṇāṃ savyadakṣiṇabhāge sthitānāṃ mūle ubhe mūlasthāne jihvayā uttambhya ūrdhvaṃ stambhayitvā | guru-mukhād avagantavyo 'yaṃ jihvā-bandhaḥ |}
+Supporting both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.
+
+\emph{rājadantānāṃ daṃṣṭrāṇāṃ savyadakṣiṇabhāge sthitānāṃ mūle ubhe mūlasthāne jihvayā uttambhya ūrdhvaṃ stambhayitvā} | \emph{gurumukhād avagantavyo 'yaṃ jihvābandhaḥ} |
 \end{quote}
-
-[Jason: also, Brahmānanda’s comments on 3.22 indicate that he thought rājadanta refers to the front teeth.
-When commenting on rājadantasthajihvāyāṃ, he says kutaḥ ?
-yato dantānāṃ rājāno rājadantā rājadanteṣu tiṣṭhatīti rājadantasthāḥ, rājadantasthā cāsau jihvā ca tasyāṃ rājadantasthajihvāyāṃ bandhaḥ, taduparibhāgasya sambandhaḥ śastaḥ].
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.46}
-\begin{translation}[hp01_046]
-See 1.45.
-\end{translation}
-
-\begin{sources}[hp01_046]
-See 1.45.
-\end{sources}
-
-\begin{testimonia}[hp01_046]
-See 1.45.
-\end{testimonia}
-
-\begin{philcomm}[hp01_046]
-See 1.45.
-\end{philcomm}
+Brahmānanda appears to have had in mind a probably older rule for meditation postures, according to which the tongue rests near the front teeth.
+One example of this is in \emph{Svacchandatantra}, which teaches a meditation pose called \emph{divyaṃ karaṇam} (4.365f.), in which the tongue is to rest at the tip
+of the teeth (\emph{dantāgre jihvām ādāya}).
+Other Tantric texts have this or similar rules, in which the tongue is supposed to rest either on the teeth or the palate, early examples being the \emph{Mrgendrāgama}, \textit{yogapāda} 19 (\emph{dantāgre jihvām ādāya}) and \emph{Mataṅgapārameśvaratantra}, \textit{yogapāda} 2.27 (\emph{tālumadhyagatenaiva jihvāgreṇa}).
+Placing the tongue where it does not disturb the meditation seems quite appropriate for a `normal' meditative practice.\footnote{The rule of placing the tongue at the palate is also found in \emph{Iśānaśivapaddhati} 18.120: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān asaṃspṛśan ṛjukāyaḥ}}
+
+However, when the context is haṭhayogic physiology, we must assume that the aim is to reach the tongue back to the uvula, which is the source of `nectar.' Confusingly, yogic terminology includes many names for the uvula, and among these especially the term \emph{rājadanta} may give rise to confusion, since, as we have seen, the tongue might also in some yoga systems be placed at the front teeth.
+
+Furthermore, \emph{khecarīmudrā} has been described in manifold ways.
+For example, the tenth-century \emph{Mokṣopāya} (V.55.14c) says that the tongue rests at the `source of the palate' (\emph{tālumūlatalālagnajihvā}) and the commentary, the
+\emph{Saṃsārataraṇi}, on the parallel passage in \emph{Laghuyogavāsiṣṭha} V.6.155, which reads \emph{tālumūlāntarālagnajihvā°}, explains that this means that the tongue is to be placed in the middle of the two regions of the palate, and that this is the \emph{nabhomudrā}, alias `\emph{khecarī}' (\emph{tālumūlāntarālagnajihvamūlaḥ tālumūlayoḥ kākudamūladeśayoḥ āntare lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā} | \emph{yā hi khecarı̄ty ucyate}).
+
+A little later in the \emph{Mokṣopāya} (V.78.24ab) it is made clear that one should reach the uvula, `placed at the root of the palate' (\emph{tālumūlagatāṃ yatnāj jihvayākramya ghaṇṭikām}).
+In view of this background we must conclude that the author of the \emph{Jyotsnā} was probably not aware of the yogic meaning of \emph{rājadanta} and has tried his best to make sense of the passage,
+echoing the idea of the two roots of the palate (although his text is not talking about the palate), but then wisely refers to the instruction of the teacher for practical details, probably noticing that his literal interpretation is somewhat opaque.
+In addition to his commentary on 1.46 (translated above), Brahmānanda’s comments on 3.22 (\emph{rājadantasthajihvāyām}) indicate that he thought the \emph{rājadanta} refers to the front teeth (\emph{kutaḥ}?
+\emph{yato dantānāṃ rājāno rājadantā rājadanteṣu tiṣṭhatīti rājadantasthāḥ, rājadantasthā cāsau jihvā ca tasyāṃ rājadantasthajihvāyāṃ bandhaḥ, taduparibhāgasya sambandhaḥ śastaḥ}).
+\end{philcomm}
+
+%%%%%%%%%%
+%\subsection*{1.46}
+%\begin{translation}[hp01_046]
+%See 1.45.
+%\end{translation}
+
+%\begin{sources}[hp01_046]
+%See 1.45.
+%\end{sources}
+
+%\begin{testimonia}[hp01_046]
+%See 1.45.
+%\end{testimonia}
+
+%\begin{philcomm}[hp01_046]
+%See 1.45.
+%\end{philcomm}
 
@@
 \begin{sources}[hp01_047]
-DYŚ
-
-\begin{versinnote}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
-\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 38 ||\\!}
-\end{versinnote}
-
-Śivasaṃhitā
-
-\begin{versinnote}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
-\tl{durlabhaṃ yena kenāpi dhīmatā labhyate param || 3.105 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_047]
-Yogacintāmaṇi
+\emph{Dattātreyayogaśāstra} 38
 
@@
 
-Haṭharatnāvalī
+\emph{Śivasaṃhitā} 3.105
 
@@
 \tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
-\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 3.38 ||\\!}
+\tl{durlabhaṃ yena kenāpi dhīmatā labhyate param ||\\!}
+\end{versinnote}
+
+\end{sources}
+
+\begin{testimonia}[hp01_047]
+\emph{Yogacintāmaṇi} f. 85v (attr. dattātreya)
+
+\begin{versinnote}
+\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
+\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 3.38
+
+\begin{versinnote}
+\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
+\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_047]
-In this context, the word \emph{durlabham} is ambiguous as to whether the posture is hard to perform or hard to acquire. The latter is the usual reading.
+In this context, the word \emph{durlabham} is somewhat ambiguous as to whether the posture is hard to perform or hard to acquire (the more usual meaning).
+In commenting on \emph{durlabham} in \emph{Jyotsnā} 2.74, Brahmānanda glosses it as \emph{duṣprāpam}, which means `difficult to attain' and `inaccessible.'
 % Maybe add some remarks by commentators on how they understand it.
@@
 
-[The yogi] should put the hands together in a bowl shape, very firmly assume padmāsana, firmly place the chin on the chest and meditation in the mind, [and] raising the Apāna breath over and over again, releasing the held Prāṇa, a man attains unequalled knowledge through the power of the goddess [Kuṇḍalinī].
+A man should put the hands together in a bowl shape, very firmly assume \emph{padmāsana}, firmly place the chin on the chest and meditation in the mind.
+Raising the \emph{apāna} breath over and over again [and] releasing the inhaled \emph{prāṇa}, he attains unequalled knowledge through the power of the goddess [Kuṇḍalinī].
 \end{translation}
@@
 \begin{sources}[hp01_048]
-VM 36
+\emph{Vivekamārtaṇḍa} 36
 
@@
 \begin{testimonia}[hp01_048]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 79v
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.39
 
@@
 \tl{vāraṃ vāram apānam ūrdhvam anilaṃ proccārayet pūritam\\+}
-\tl{muñcat prāṇam upaiti bodham atulaṃ śakteḥ prabhāvān naraḥ || 3.39 ||\\!}
+\tl{muñcat prāṇam upaiti bodham atulaṃ śakteḥ prabhāvān naraḥ ||\\+}
+\tl{\var{proccārayet ] proccālayat P,T,t1,n2}\\!}
 \end{versinnote}
@@
 
-The end of pāda b is uncertain.
-Later witnesses including Brahmānanda have \emph{dhyāyaṃś ca} but none of the early ones has this reading.
+The text at end of the second verse quarter is uncertain.
+Later witnesses, including Brahmānanda, have \emph{dhyāyaṃś ca} but none of the early ones has this reading.
 So, we have to take \emph{dhyānaṃ} with \emph{sannidhāya}.
 However, \emph{tat} is problematic because here it has no clear referent.
-In the \emph{Vivekamārtaṇḍa tat} appears to refer to the \emph{mokṣadvāra} broken by Kuṇḍalinī, which is mentioned in the previous verse.
+In the \emph{Vivekamārtaṇḍa, tat} appears to refer to the \emph{mokṣadvāra} broken by Kuṇḍalinī, which is mentioned in the previous verse.
 
 The two participles \emph{proccālayan} and \emph{muñcan} imply that the two things are happening at the same time, which is surprising but perhaps possible.
-Or perhaps one can understand them as sequential.
-
-In the VM this verse comes in the passage on raising Kuṇḍalinī, so tat could refer to visualising K opening the mokṣadvāra.
+Or one might infer that `and' (\emph{ca}) was supposed to be supplied and understand both actions as sequential.
+
+%In the VM this verse comes in the passage on raising Kuṇḍalinī, so tat could refer to visualising K opening the mokṣadvāra.
 \end{philcomm}
@@
 \begin{translation}[hp01_049]
-The yogi in padmāsana who fills [himself] up through the openings of the channels and holds the breath is sure to be liberated.
-\end{translation}
-
-\begin{sources}[hp01_049]
---
-\end{sources}
+The yogi in \emph{padmāsana} who fills [himself] up through the openings of the channels and holds the breath is sure to be liberated.
+\end{translation}
+
+%\begin{sources}[hp01_049]
+%--
+%\end{sources}
 
 \begin{testimonia}[hp01_049]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85v (attr. dattātreya)
 
@@
 
-Haṭharatnāvalī 3.40
+\emph{Haṭharatnāvalī} 3.40
 
@@
 \tl{padmāsane sthito yogī nāḍīdvāreṣu pūrayet |\\+}
-\tl{pūritaṃ dhrīyate yas tu sa mukto nātra saṃśayaḥ || 3.40 ||\\!}
-\end{versinnote}
-
-Dhyānabindu Up
+\tl{pūritaṃ dhrīyate yas tu sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
+
+\emph{Dhyānabindūpaniṣat} 70
 
@@
 \tl{padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\+}
-\tl{mārutaṃ kumbhayan yas tu sa mukto nātra saṃśayaḥ || 70 ||\\!}
+\tl{mārutaṃ kumbhayan yas tu sa mukto nātra saṃśayaḥ ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_049]
-pūrayan can take the locative (e.g., Amaraugha 21d pūrayen mukhe).
-The reading of N23 niyatam would explain the passive verbs in other witnesses.
-The passive verbs do not make sense with yas tu.
-This yields the idea of filling up the central channel (as opposed to other channels) seen in the Yogabīja.
-Note that a breath retention is not mentioned in the previous verse describing padmāsana (only inhalation and exhalation).
-Also,  māruto mriyate yasya is possible (if one  accepts pūrayet in the second pāda).
-J5 also has a plausible reading (mārutaṃ mārayet yas tu) but J5 is contaminated (so the Gr3 reading dhārayet is more reliable).
+It is not unusual to read \emph{pūrayan} with the locative as seen in this verse (cf. \emph{pūrayen mukhe} in \emph{Amaraugha} 21d).
+
+Gamma one's reading of \emph{niyatam} (instead of \emph{dhārayed}) in the third verse quarter explains the passive verbs in other witnesses.
+The passive verbs do not make sense with \emph{yas tu}.
+The passive verbs meaning to take in the breath (e.g., \emph{pīyate}) may have been adopted to remove the reference to holding the breath because a breath retention is not mentioned in the previous verse describing \emph{padmāsana} (only inhalation and exhalation).
+
+In the third verse quarter, the best alpha manuscript reads \emph{māruto mriyate yas tu}, which does not make sense.
+However, if one accepts \emph{pūrayet} in the second verse quarter, one could emend alpha's reading to \emph{māruto mriyate yasya}, which makes good sense (i.e., `the yogi whose breath dies is undoubtedly liberated').
+In the same vein, J5 also has the plausible reading \emph{mārutaṃ mārayet yas tu}.
+%but J5 is contaminated (so the Gr3 reading dhārayet is more reliable).
+
+(MD: Or read \emph{pūrayan mārutaṃ dhriyate yas tu} ``he who remains by inhaling..."?
+This \emph{dhriyate} is rather intransitive than passive. 
+Cf. PW dhṛ 22a ``fest sein, sich ruhig verhalten, stillhalten, verbleiben, sich erhalten, bestehen")
+
+The \emph{Jyotsnā} (1.49) has \emph{naḍīdvāreṇa} instead of \emph{nāḍīdvāreṣu}, and Brahmānanda interprets it as the opening of the central channel (\emph{suṣumnāmārgeṇa}).
+This yields the idea of filling up the central channel (as opposed to other channels), which is described in the \emph{Yogabīja} (94–95).
 \end{philcomm}
@@
 \begin{translation}[hp01_050]
-Now the lion’s pose. ---
-
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. He should place the left ankle on right, the right ankle on the left [...]
-\end{translation}
-
-\begin{translation}[hp01_051]
-[...] and both hands on the knees, spread his fingers, open his mouth and gaze in deep concentration at the tip of the nose.
-\end{translation}
-
-\begin{translation}[hp01_052]
-This is the lion’s pose, which is always worshipped by yogis. It causes the three locks to arise together and is the best of [all] postures.
+Now, the lion’s pose (\emph{siṃhāsana}).
+
+[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
+He should place the left ankle on right, the right ankle on the left and both hands on the knees, spread his fingers, open his mouth and gaze in deep concentration at the tip of the nose.
+This is the lion’s pose, which is always worshipped by yogis.
+It causes the three locks to arise together and is the best of [all] postures.
 \end{translation}
@@
 \begin{sources}[hp01_050]
-Vasiṣṭhasaṃhitā 1.73
+\emph{Vasiṣṭhasaṃhitā} 1.73–1.75ab
 
@@
 \tl{vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ ||\\+}
-\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā ||\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.9
-
-\begin{versinnote}
-\tl{gulpau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\+}
+\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā |\\!}
+\end{versinnote}
+
+\emph{Yogayājñavalkya} 3.9–3.11ab
+
+\begin{versinnote}
+\tl{gulpau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
 \tl{dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram ||\\+}
-\tl{hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca\\+}
+\tl{hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca |\\+}
 \tl{vyāttavaktro nirīkṣet nāsagraṃ susamāhitaḥ ||\\+}
-\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā\\!}
-\end{versinnote}
-
-Sūtasaṃhitā 15.7
+\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā |\\!}
+\end{versinnote}
+
+\emph{Sūtasaṃhitā} 15.7–8
 
@@
 \tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{dakṣiṇaṃ savyagulphena vāmaṃ dakṣiṇagulphataḥ || 7 ||\\+}
+\tl{dakṣiṇaṃ savyagulphena vāmaṃ dakṣiṇagulphataḥ ||\\+}
 \tl{hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca |\\+}
-\tl{nāsāgraṃ ca nirīkṣeta bhavet siṃhāsanaṃ hi tat || 8 ||\\!}
+\tl{nāsāgraṃ ca nirīkṣeta bhavet siṃhāsanaṃ hi tat ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_050]
-Yogacintāmaṇi (citing Yogayājñavalkya)
+\emph{Yogacintāmaṇi} f. 83v (attr. yājñavalkya)
 
@@
 
-[Note: The Yogacintāmaṇi attributes its citation of verses on siṃhāsana to the Yogayājñavalkya and does not include HP 1.52cd, which affirms that 1.52cd is not from the Yogayājñavalkya (or Vasiṣṭhasaṃhitā)].
-
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.31–3.33
 
@@
 \tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{dakṣiṇe savyagulphaṃ ca dakṣiṇe tu tathetaram || 3.31 ||\\+}
-\tl{hastau tu jānvoḥ saṃsthāpya svāṅgulīḥ samprasārya ca ||\\+}
-\tl{vyāttavaktro nirīkṣeta nāsāgraṃ tu samāhitaḥ || 3.32 ||\\+}
-\tl{siṃhāsanaṃ bhaved etat sevitaṃ yogibhiḥ sadā ||\\+}
-\tl{bandhatritayasaṃsthānaṃ kurute cāsanottamam || 3.33 ||\\!}
+\tl{dakṣiṇe savyagulphaṃ ca dakṣiṇe tu tathetaram ||\\+}
+\tl{hastau tu jānvoḥ saṃsthāpya svāṅgulīḥ samprasārya ca |\\+}
+\tl{vyāttavaktro nirīkṣeta nāsāgraṃ tu samāhitaḥ ||\\+}
+\tl{siṃhāsanaṃ bhaved etat sevitaṃ yogibhiḥ sadā |\\+}
+\tl{bandhatritayasaṃsthānaṃ kurute cāsanottamam ||\\!}
 \end{versinnote}
@@
 https://www.metmuseum.org/art/collection/search/39251).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.51}
-\begin{sources}[hp01_051]
-See 1.50.
-\end{sources}
-
-\begin{testimonia}[hp01_051]
-See 1.50.
-\end{testimonia}
-
-\begin{philcomm}[hp01_051]
-See 1.50.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.52}
-\begin{sources}[hp01_052]
-See 1.50.
-\end{sources}
-
-\begin{testimonia}[hp01_052]
-See 1.50.
-\end{testimonia}
-
-\begin{philcomm}[hp01_052]
-See 1.50.
-\end{philcomm}
+
+The \emph{Yogacintāmaṇi} attributes its citation of these verses on \emph{siṃhāsana} to the Yājñavalkya.
+Its citation does not include \emph{Haṭhapradīpikā} 1.52cd, which affirms that 1.52cd is not from the \emph{Yogayājñavalkya} (or \emph{Vasiṣṭhasaṃhitā}).
+As far as we are aware, there is no source for the hemistich mentioning the three locks, so it may have been composed by Svātmārāma or borrowed from a lost work.
+\end{philcomm}
+
+%%%%%%%%%%
+%\subsection*{1.51}
+%\begin{sources}[hp01_051]
+%See 1.50.
+%\end{sources}
+
+%\begin{testimonia}[hp01_051]
+%See 1.50.
+%\end{testimonia}
+
+%\begin{philcomm}[hp01_051]
+%See 1.50.
+%\end{philcomm}
+
+%%%%%%%%%%
+%\subsection*{1.52}
+%\begin{sources}[hp01_052]
+%See 1.50.
+%\end{sources}
+
+%\begin{testimonia}[hp01_052]
+%See 1.50.
+%\end{testimonia}
+
+%\begin{philcomm}[hp01_052]
+%See 1.50.
+%\end{philcomm}
 
@@
 \begin{translation}[hp01_053]
-Now, the friendly pose.
-
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.  Having firmly and motionlessly held the sides of the feet with the hands,[...]
-\end{translation}
-
-\begin{translation}[hp01_054]
-[...] this is bhadrāsana, which cures all diseases and poisons. Yogis of the siddha tradition call it Gorakṣāsana.
+Now, the friendly pose (\emph{bhadrāsana}).
+
+[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
+Having firmly and motionlessly held the sides of the feet with the hands, this is the friendly pose, which cures all diseases and poisons.
+Yogis of the Siddha tradition call it Gorakṣa's pose (\emph{gorakṣāsana}).
 \end{translation}
@@
 \begin{sources}[hp01_053]
-Vasiṣṭhasaṃhitā 1.79
+\emph{Vasiṣṭhasaṃhitā} 1.79
 
@@
 
-Yogayājñavalkya 3.11cd--3.12ab
+\emph{Yogayājñavalkya} 3.11cd--3.12ab
 
@@
 \begin{testimonia}[hp01_053]
-Yogacintāmaṇi (citing Yājñavalkya)
+\emph{Yogacintāmaṇi} f. 83v (citing yājñavalkya)
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 3.30
 
@@
 \tl{atha bhadrāsanam\\+}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet ||\\+}
+\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
 \tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam ||\\+}
-\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham || 3.30 ||\\!}
-\end{versinnote}
-
-YSS
-
-\begin{versinnote}
-\tl{gulphau ca vṛṣaṇasyāyasvīvinyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{pārśvapāde ca pāṇibhyāṃ dṛḍhaṃ badhvāsu niścalaḥ ||\\+}
-\tl{bhadrāsanaṃ bhaved etat sarvavyādhivināśanam |\\+}
-\tl{gorakṣāsanam ity āhur idaṃ vai siddhayoginaḥ ||\\!}
+\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham ||\\!}
 \end{versinnote}
@@
 We have understood \emph{pārśvapāda} (i.e., side of the foot) like the compound \emph{agrapāda} (i.e., front of the foot), which can be classified as a \emph{ekadeśitatpuruṣa}.
-See Sanskrit Vademecum, p.84 (ed.\ Maximilian Mehner and Jürgen Hanneder, Marburg 2019).
+See Sanskrit Vademecum, 2019: 84.
+
+Manuscripts of two early groups, beta and gamma, as well as the \emph{Jyotsnā} (1.53), include an additional hemistich specifying that the left ankle is placed on the left side and the right ankle on the right (\emph{savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe}).
+This hemistich appears to have been added to make it clear that the ankles are not crossed in \emph{bhadrāsana}, unlike the previous pose, \emph{siṃhāsana}.
+
+%(ed.\ Maximilian Mehner and Jürgen Hanneder, Marburg 2019).
 
@@
 %%%%%%%%%%
-\subsection*{1.54}
-\begin{sources}[hp01_054]
-See 1.53.
-\end{sources}
-
-\begin{testimonia}[hp01_054]
-See 1.53.
-\end{testimonia}
-
-\begin{philcomm}[hp01_054]
-See 1.53.
-\end{philcomm}
+%\subsection*{1.54}
+%\begin{sources}[hp01_054]
+%See 1.53.
+%\end{sources}
+
+%\begin{testimonia}[hp01_054]
+%See 1.53.
+%\end{testimonia}
+
+%\begin{philcomm}[hp01_054]
+%See 1.53.
+%\end{philcomm}
 
@@
 \begin{translation}[hp01_055]
-When the great yogi does not become tired from adopting the āsanas in this way, he should now practise the breath techniques with mudrās and so forth from which purification of the channels arise.
-\end{translation}
-
-\begin{sources}[hp01_055]
-\end{sources}
+When the great yogi does not become tired from adopting the \emph{āsana}s in this way, he should now practise the breath techniques with seals and so forth, from which purification of the channels arise.
+\end{translation}
+
+%\begin{sources}[hp01_055]
+%\end{sources}
 
 \begin{testimonia}[hp01_055]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 85v (attr. dattātreya)
 
@@
 
-HSC
+\emph{Haṭhasaṅketacandrikā} f. 23r
 
@@
 \tl{evam āsanabandhastho yogīndro vigataśramaḥ |\\+}
-\tl{athābhyasen nāḍiśuddhiṃ mudrādipavanakriyām || 73 ||\\!}
+\tl{athābhyasen nāḍiśuddhiṃ mudrādipavanakriyām ||\\+}
+\tl{\var{nāḍiśuddhiṃ ] \emph{em.}, nāhiśuddhi ms. no. 2244}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_055]
-55c is ra-vipulā with \emph{nāḍi}
 The second hemistich can be interpreted in different ways.
-One possibility is to understand \emph{nāḍiśuddhiṃ} as a bahuvrīhi qualifying \emph{mudrādipavanakriyām} in the sense that one should practise a breathing technique with mudrās from which purification of the channels arise.
-Alternatively, one could separate \emph{mudrādi} from \emph{pavanakriyām} and understand three different techniques here, namely, the practice of purifying the channels (perhaps by the alternative nostril method mentioned at the beginning of the second chapter), the \emph{mudrā}s and the breathing techniques of prāṇāyāma.
+One possibility is to understand \emph{nāḍiśuddhiṃ} as a bahuvrīhi qualifying \emph{mudrādipavanakriyām} in the sense that the yogi should practise a breathing technique, along with with \emph{mudrā}s and other mechanisms, such as locks (\emph{bandha}), from which purification of the channels arise.
+Alternatively, one could separate \emph{mudrādi} from \emph{pavanakriyām} and understand three different techniques here, namely, the practice of purifying the channels (perhaps by the alternative nostril method mentioned at the beginning of the second chapter), the \emph{mudrā}s and the breathing techniques of \emph{prāṇāyāma}.
+The absence of a conjunctive particle, such as \emph{ca}, makes the second interpretation less likely.
+The version of this verse in the \emph{Yogacintāmaṇi} (cited in the testimonia) was changed to make it clear that \emph{nāḍiśuddhi} and \emph{pavanakriyā} with \emph{mudrā}s, are two distinct things.
+
+The metre of 1.55c is a \emph{ra-vipulā} with \emph{nāḍi}.
 \end{philcomm}
@@
 \begin{sources}[hp01_055_1]
-Dattātreyayogaśāstra 42cd--47
-
-\begin{versinnote}
-\tl{kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet |\\+}
+\emph{Dattātreyayogaśāstra} 42cd–43ab, 46–47
+
+\begin{versinnote}
+\tl{kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet ||42 ||\\+}
 \tl{na śāstrapāṭhamātreṇa kā cit siddhiḥ prajāyate |\\+}
@@
 
-\begin{testimonia}[hp01_055_1]
-\end{testimonia}
+%\begin{testimonia}[hp01_055_1]
+%\end{testimonia}
 
 \begin{philcomm}[hp01_055_1]
-1.55*1–2 are omitted from V1 and V19, so may not have been included in the earliest version of the text.
-In fact, it is possible that both were added (perhaps initially as marginal notes) to elaborate on kriyā in 1.55d.
-Both verses are similar to verses from the Dattātreyayogaśāstra, and these verses (except 1.55.2ef) appear in the vulgate, but towards the end of chapter 1 (1.65–66).
-
-And then just half of DYŚ 47 is given, resulting in a near-nonsensical hemistich.
-
-The syntax of 1.55.2ef is corrupt. One has to emend to \emph{deyā} to make sense of it.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.55*2}
-\begin{sources}[hp01_055_2]
-See 1.55*1.
-\end{sources}
-
-\begin{testimonia}[hp01_055_2]
-See 1.55*1.
-\end{testimonia}
-
-\begin{philcomm}[hp01_055_2]
-See 1.55*1.
-\end{philcomm}
+1.55*1–2 are omitted from the alpha, gamma and eta groups, so it is likely these verses were not in the earliest versions of the \emph{Haṭhapradīpikā}.
+In fact, it is likely that both were added (perhaps initially as marginal notes) to elaborate on the word \emph{kriyā} in 1.55d.
+Both verses are similar to verses from the \emph{Dattātreyayogaśāstra} (cited as the source).
+However, only the first half of \emph{Dattātreyayogaśāstra} 47 is given in these later versions of the \emph{Haṭhapradīpikā}, resulting in a near-nonsensical hemistich.
+Also, the syntax of 1.55.2ef is corrupt.
+One has to emend to \emph{deyā} to make sense of it.
+These verses (except 1.55.2ef) appear in the \emph{Jyotsna} (1.65–66), but towards the end of chapter one.
+\end{philcomm}
+
+%%%%%%%%%%
+%\subsection*{1.55*2}
+%\begin{sources}[hp01_055_2]
+%See 1.55*1.
+%\end{sources}
+
+%\begin{testimonia}[hp01_055_2]
+%See 1.55*1.
+%\end{testimonia}
+
+%\begin{philcomm}[hp01_055_2]
+%See 1.55*1.
+%\end{philcomm}
 
@@
 \begin{translation}[hp01_055_3]
-In me, the pure ocean of awakening, is this empty bubble we call the universe dissolved or does it arise? Where does this veil of doubt about this come from?
-\end{translation}
+Is this empty bubble we call the universe dissolved or did it arise in me, the pure ocean of awakening? Where does [this] veil of doubt come from?
+\end{translation}
+
+\begin{sources}[hp01_055_3]
+\emph{Tattvaratnāvalī} 24
+
+\begin{versinnote}
+\tl{bodhāmbhodhau mayi svacchaṃ tac chāyam viśvabuddhayaḥ |\\+}
+\tl{udito vā pralīno vā na vikalpāya kalpate ||\\!}
+\end{versinnote}
+
+\end{sources}
 
 \begin{testimonia}[hp01_055_3]
-Vārāhītantra
+\emph{Vārāhītantra} p. 158
 
@@
 \tl{mayi bodhībudho svasthe tucho yaṃ viśvabudbudaḥ |\\+}
-\tl{malīna udito vetti vikalpāvasaraḥ kutaḥ |\\!}
-\end{versinnote}
-
-Siddhāntamuktāvalī
+\tl{malīna udito vetti vikalpāvasaraḥ kutaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhapradīpikā} (10 chapters) 3.7
 
@@
 \tl{śiśnodararatāya hi na deyaṃ veṣadhāriṇe ||\\+}
-\tl{mayi bodhyaṃ buddhau svacche tad dheyaṃ viśvabudbudam || 7 ||\\!}
-\end{versinnote}
-
-Yogaprakāśikā
+\tl{mayi bodhyaṃ buddhau svacche tad dheyaṃ viśvabudbudam ||\\!}
+\end{versinnote}
+
+\emph{Yogaprakāśikā} 3.7
 
@@
 \begin{philcomm}[hp01_055_3]
-It is very difficult to find a reason why this verse should be inserted here. It is apparently a
-\emph{muktaka} that would befit an accomplished spiritual poem more than our \emph{Haṭhapradīpikā},
-even here, in what appears as a sort of miscellaneous section at the end of a chapter.  In this
-verse, the lyrical subject wonders about why the mind is still able to doubt, despite its insight
-into the nature of reality and the reader wonders, how the illusionist verse could be understood to
-fit our Yoga text, perhaps the scribe of the archetype of V19 and N17, was fond of it.
-
-The source is, as far as we can say, the \emph{Śāntiśataka} of the Kashmirian poet Sillana or
-Silhaṇa,\footnote{The mss.\ of his \emph{Svātmopalabdhiśataka} give the name as Sillana, the mostly
-Bengali mss. of the Śāntiśataka read Śilhaṇa, as does Aufrecht in his CC. See Hanneder:
-forthcoming.} who cannot be dated with any certainty, but predates the HP by a few centuries. The
-edition of this text – where hundred original verses had to be identified – places it into an
-appendix of doubtful stanzas,\footnote{Karl Schönfeld: \emph{Das Śāntiśataka}. Leipzig:
-Harrassowitz 1910, p.\,90 (A9).} but the editor saw no compelling reason not to regard it as
-original except through the fact that it is not transmitted in all manuscripts.  What prevents
-further investigation of the matter is the lack of Kashmirian mss.\ for the \emph{Śāntiśataka} and
-its compilatory character: one quarter of the material is identical with Bhartṛhari's
-\emph{Vairāgyaśataka}. A still superficial glance at Sillana's \emph{Svātmopalabdhiśataka} gives
-the impression that our verse would fit there, but not so much in the \emph{Śāntiśataka}. Perhaps its first citation is in Advayavajra’s \emph{Tattva\-ratnāvalī} (ref??),  While
-these are only preliminary observations the verse is probably not original in any stage of
-development of the HYP, but limited to V19 (and N17).
+Verse 1.55.3 is only found in manuscripts of the delta group. It is very difficult to find a reason why this verse should be inserted here. It is apparently a
+\emph{muktaka} that would befit an accomplished spiritual poem more than an instructional manual, like the \emph{Haṭhapradīpikā}, even here, in what appears as a sort of miscellaneous section at the end of a chapter.
+In this verse, the lyrical subject wonders about why the mind is still able to doubt, despite its insight into the nature of reality.
+The reader might wonder how this illusionist verse could be understood to fit our Yoga text.
+We can only speculate that perhaps the scribe of the hyparchetype of the delta manuscripts was fond of it.
+
+The source is, as far as we can say, the \emph{Śāntiśataka} of the Kashmirian poet Sillana or Silhaṇa, The manuscripts of the \emph{Svātmopalabdhiśataka} give the name as \emph{Sillana}, the mostly Bengali manuscripts of the \emph{Śāntiśataka} read \emph{Śilhaṇa}, as does Aufrecht in his \emph{Catalogus Catalogorum}, 1891 (for further details see Hanneder, forthcoming).
+Sillana cannot be dated with any certainty but predates the \emph{Haṭhapradīpikā} by a few centuries.
+The
+edition of the \emph{Śāntiśataka} – where a hundred original verses had to be identified – places the verse in question into an appendix of doubtful stanzas (see Karl Schönfeld: \emph{Das Śāntiśataka}.
+Leipzig: Harrassowitz 1910, p.\,90 [A9]).
+However, the editor did not provide a compelling reason to regard it as
+unoriginal except only the fact that it is not transmitted in all manuscripts.
+What prevents further investigation of the matter is the lack of Kashmirian manuscripts for the \emph{Śāntiśataka} and its compilatory character: one quarter of the material is identical with Bhartṛhari's \emph{Vairāgyaśataka}.
+A still superficial glance at Sillana's \emph{Svātmopalabdhiśataka} gives the impression that our verse would fit there, but not so much in the \emph{Śāntiśataka}.
+Perhaps its first citation is in Advayavajra’s \emph{Tattva\-ratnāvalī} (24).
+While
+these are only preliminary observations the verse is likely not original to the \emph{Haṭhapradīpikā}.
 \end{philcomm}
@@
 
-\begin{sources}[hp01_055_4]
-\end{sources}
+%\begin{sources}[hp01_055_4]
+%\end{sources}
 
 \begin{testimonia}[hp01_055_4]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 48v
 
@@
 \tl{haṭhapradīpikāyām--\\+}
-\tl{śrutapratītiḥ svagurupratītiḥ, svātmapratītir manaso nirodhaḥ |\\+}
-\tl{etāni sarvāṇi samuccitāni, matāni dhīrair iha sādhanāni ||\\!}
+\tl{śrutipratītiś ca gurupratītiḥ svātmapratītiś ca manonirodhaḥ |\\+}
+\tl{etāni sarvāṇi samuccitāni matāni dhīrair iha sādhanāni ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_055_4]
-last greyscale verse (\emph{śrutipratītiḥ…}) is only in V19 of old mss (but some other later ones) and is quoted in Yogacintāmaṇi.
-
-The reading in the Yogacintāmaṇi (\emph{manaso nirodhaḥ}) is better in a yogic context.
+Verse 1.55.4 is in some of the delta manuscripts and is quoted in \emph{Yogacintāmaṇi}.
+
+The reading in the \emph{Yogacintāmaṇi} `cessation of mind' (\emph{manonirodhaḥ}) is better than \emph{manaso 'pi bodhaḥ} (the delta reading) in a yogic context.
 \end{philcomm}
@@
 \begin{translation}[hp01_056]
-The various āsana, breath retention, bodily technique (karaṇa) called seals (mudrā), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.
+The various \emph{āsana}, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.
 \end{translation}
@@
 \begin{testimonia}[hp01_056]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 111v
 
@@
 \begin{philcomm}[hp01_056]
-Verse is omitted from V1, probably deliberately as it doesn’t have chapter 4, which teaches nādānusandhāna. The numbering in V1 suggests that its exemplar had this verse.
-
-Kumbhaka is almost always masculine.
-
-Marmasthāna: not clear whether to adopt \emph{citro} or \emph{citraṃ}, or V19’s \emph{mudrākhyaṃ karaṇaṃ tathā} or the others’ \emph{mudrādikaraṇāni ca}.
-
-See also 1.65, which has \emph{mudrādikaraṇāni ca}, so perhaps it was through confusion with this that the same reading is found in some witnesses of 1.56.
-
-Yes, V19 reading probably best.
-
-\begin{versinnote}
-\tl{āsanaṃ kumbhakaś citraṃ mudrākhyaṃ karaṇaṃ tathā |\\+}
-\tl{atha nādānusandhānam abhyāsānukramo haṭhe ||\\!}
-\end{versinnote}
-
-It seems that the four aṅgas of Haṭhayoga are being referred to in the singular (hence \emph{āsanaṃ}), whereas in 1.65 the plural is used (i.e., \emph{pīṭhāni}).
-Therefore, we should adopt \emph{citraṃ [...] karaṇaṃ tathā}.
+This verse was omitted from eta one, the oldest dated manuscript.
+The omission is probably deliberate as that manuscript does not have chapter four, which teaches \emph{nādānusandhāna}.
+The numbering in eta one suggests that its exemplar had this verse.
+
+The term \emph{kumbhaka} is almost always masculine but appears in this verse as a neuter in the majority of manuscripts of the important groups.
+
+%Marmasthāna: not clear whether to adopt \emph{citro} or \emph{citraṃ}, or V19’s \emph{mudrākhyaṃ karaṇaṃ tathā} or the others’ \emph{mudrādikaraṇāni ca}.
+
+This verse is similar to 1.65, which has \emph{mudrādikaraṇāni ca}, so perhaps it was through confusion with 1.65 that the same reading is found in some witnesses of 1.56.
+It seems that the four auxiliaries (\emph{aṅga}) of Haṭhayoga are being referred to in the singular (hence \emph{āsanaṃ}), whereas in 1.65 the plural is used (i.e., \emph{pīṭhāni}).
+Therefore, the reading \emph{citraṃ} [...] \emph{karaṇaṃ tathā} is likely original for this verse.
+
+%Yes, V19 reading probably best.
+%
+%\begin{versinnote}
+%\tl{āsanaṃ kumbhakaś citraṃ mudrākhyaṃ karaṇaṃ tathā |\\+}
+%\tl{atha nādānusandhānam abhyāsānukramo haṭhe ||\\!}
+%\end{versinnote}
 \end{philcomm}
@@
 \begin{translation}[hp01_057]
-Celibate, restricted in diet and devoted to yoga, the renunciant succeeds in upwards of a year. No doubt about this should be entertained.
-\end{translation}
+Celibate, restricted in diet and devoted to yoga, the yogi succeeds in upwards of a year. No doubt about this should be entertained.
+\end{translation}
+% adopt yogī?
 
 \begin{sources}[hp01_057]
-Vivekamārtaṇḍa 37
+\emph{Vivekamārtaṇḍa} 37
 
@@
 \begin{testimonia}[hp01_057]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 111v (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī}  3.28
 
@@
 \tl{brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ ||\\+}
-\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā || 3.28 ||\\!}
+\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||\\+}
+\tl{\var{tyāgī ] yogī P,T,t1}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_057]
-Comment on tyāgī and yogī, and the reading of this verse in the Vivekamārtaṇḍa.
-It seems likely that Svātmārāma was using a version of the VM with tyāgī (as it’s unlikely that yogī would be emended to tyāgī), but whether this was the original reading in the VM is more complicated because there tyāgī may be a dittographical-type mistake.
+The readings \emph{tyāgī} and \emph{yogī} are both well attested in \emph{Haṭhapradīpikā} 1.57b.
+The confusion between the two appears to have started early in the transmission of the \emph{Vivekamārtaṇḍa}.
+The occurrence of \emph{tyāgī} in \emph{Vivekamārtaṇḍa} 37 may be a dittographical type of mistake because the word \emph{tyāgī} is in the previous hemistich of that work.
+But it is more difficult to determine whether Svātmārāma used a manuscript of the \emph{Vivekamārtaṇḍa} with \emph{tyāgī} or \emph{yogī} in verse 37.
+Since the best alpha manuscript has \emph{yogī}, as well as eta one and many others, we have tentatively adopted it bearing in mind that it was changed early in the transmission of the \emph{Haṭhapradīpikā}, most likely by a scribe who knew the reading of \emph{tyāgī} in a manuscript of the \emph{Vivekamārtaṇḍa}.
 \end{philcomm}
@@
 \begin{translation}[hp01_058]
-When very unctuous and sweet food that is without a quarter portion is eaten for love of śiva, it is called a restricted diet. (\emph{mitāhāra}).
+When very unctuous and sweet food that is without a quarter portion is eaten for love of Śiva, it is called a restricted diet (\emph{mitāhāra}).
 \end{translation}
@@
 \begin{sources}[hp01_058]
-Gorakṣaśataka (original)
-
-\begin{versinnote}
-\tl{susnigdhamadhurāhāraś caturthāṃśavivarjitaḥ |\\+}
-\tl{bhujyate śivasaṃprītyai mitāhāraḥ sa ucyate ||\\!}
+\emph{Gorakṣaśataka} 12cd–13ab
+
+\begin{versinnote}
+\tl{susnigdhamadhurāhāraś caturthāṃśavivarjitaḥ ||\\+}
+\tl{bhujyate śivasaṃprītyai mitāhāraḥ sa ucyate |\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_058]
-Yuktabhavadeva
+\emph{Yuktabhavadeva} 4.16
 
@@
 \tl{tad uktaṃ haṭhapradīpikāyām-\\+}
-\tl{susnigdhamadhurāhārāś caturthāṃśavivarjitaḥ ||\\+}
-\tl{bhujyate śivasamprītyai mitāhāraḥ sa ucyate || 16 ||\\!}
-\end{versinnote}
-
-Yogacūḍāmaṇyupaniṣat
+\tl{susnigdhamadhurāhārāś caturthāṃśavivarjitaḥ |\\+}
+\tl{bhujyate śivasamprītyai mitāhāraḥ sa ucyate ||\\!}
+\end{versinnote}
+
+\emph{Yogacūḍāmaṇyupaniṣat} 43
 
@@
 \tl{susnigdhamadhurāhāraścaturthāṃśavivarjitaḥ |\\+}
-\tl{bhuñjate śivasaṃprītyā mitāhārī sa ucyate || 43 ||\\!}
+\tl{bhuñjate śivasaṃprītyā mitāhārī sa ucyate ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_058]
-This verse probably derives from the ‘original’ GŚ (12c–13b).
-It is also found, but reworked to be about the \emph{mitāhārī}, in Nowotny’s GŚ (extended recension of VM) at 55.
-The idea of not eating the fourth portion of one’s food (\emph{caturthāṃśavivarjitaḥ}) can be found in older sources, such as Dharmaputrikā 1.51-52 (ṣaḍrasopetasuṣnigdhasvādusāndrasugandhinā |
-udarasyārdhabhāgan tu bhojanena prapūrayet || pānīyena caturbhāgaṃ taccheṣaṃ śūnyam iṣyate | vāyos sañcāraṇānārtham āhāraniyamaḥ smṛtaḥ) and Aṣṭāṅgahṛdayasaṃhitā Sūtrasthāna 8.46c-47b ( annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet ||āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet).
-As noted in the Jy
+This verse probably derives from the ‘original’ \emph{Gorakṣaśataka} (12c–13b).
+It is also found, but reworked to be about the \emph{mitāhārī}, in Nowotny’s \emph{Gorakṣaśataka} (55), which is an extended recension of the \emph{Vivekamārtaṇḍa}.
+
+The idea of not eating the fourth portion of one’s food (\emph{caturthāṃśavivarjitaḥ}) can be found in older yoga texts, such as \emph{Dharmaputrikā} 1.51-52:
+
+\begin{versinnote}
+\tl{ṣaḍrasopetasuṣnigdhasvādusāndrasugandhinā |\\+}
+\tl{udarasyārdhabhāgan tu bhojanena prapūrayet ||\\+} 
+\tl{pānīyena caturbhāgaṃ taccheṣaṃ śūnyam iṣyate |\\+}
+\tl{vāyos sañcāraṇānārtham āhāraniyamaḥ smṛtaḥ ||\\!} 
+\end{versinnote}
+
+And, as noted by Brahmānanda in \emph{Jyotsnā} 1.58, the idea also occurs in an āyurvedic work called the \emph{Aṣṭāṅgahṛdayasaṃhitā, Sūtrasthāna}, 8.46cd–47ab:
+
+\begin{versinnote}
+\tl{annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet |||\\+}
+\tl{āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet ||\\!} 
+\end{versinnote}
 
@@
 [Adepts] say the [following] is unwholesome: pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink.
-Flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, Jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like.
+Flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like.
 \end{translation}
@@
 \begin{testimonia}[hp01_059]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 54v
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 1.72
 
@@
 \tl{ajādimāṃsadadhitakrakulatthakodra-\\+}
-\tl{piṇyākahiṅgulaśunādyam apathyam āhuḥ || 1.72 ||\\!}
-\end{versinnote}
-
-HTK
+\tl{piṇyākahiṅgulaśunādyam apathyam āhuḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī}
 
@@
 \begin{philcomm}[hp01_059]
-59a \emph{kaṭvamla°} is better than \emph{kaṭvāmla°} and well attested elsewhere in lists of tastes and types of foods.
-On the meaning of \emph{uṣṇa} (in relation to food) see Meulenbeld 1974: 254 fn.
-13: ‘Cakra mentions as a variant: \emph{katvamlalavaṇakṣāra} (pungent.
+The compound \emph{kaṭvamla°} (1.59a) is better than \emph{kaṭvāmla°} and it is well attested by manuscripts of the \emph{Haṭhapradīpikā}, as well as in lists of tastes and types of foods in other texts.
+
+On the meaning of \emph{uṣṇa} (1.59a) in relation to food, see Meulenbeld 1974: 254 fn. 13:
+\begin{quote}
+Cakra mentions as a variant: \emph{katvamlalavaṇakṣāra} (pungent.
 acid, saline and caustic).
-Cakra remarks that the term `hot' (\emph{uṣṇa}) denotes hot on touch when it occurs the first time, and hot with regard to potency when it occurs for the second time.’
-
-\emph{°hari}(\emph{ī} here for metre?)\emph{taśāka°} in some nighaṇṭus is horseradish, which makes better sense here than Brahmānanda’s gloss.
-
-Rājanighaṇṭu 7.26
+Cakra remarks that the term `hot' (\emph{uṣṇa}) denotes hot on touch when it occurs the first time, and hot with regard to potency when it occurs for the second time.
+\end{quote}
+
+The compound \emph{°harītaśāka°} in 1.59a is spelt \emph{°haritaśāka°} in other works.
+The spelling \emph{°harīta} was probably adopted for metrical reasons.
+In some Nighaṇṭus (`thesauruses'), \emph{°haritaśāka°} is glossed as horseradish.
+\begin{quote}
+\emph{Rājanighaṇṭu} 7.26
 
@@
 \end{versinnote}
-
-Sauśrutanighaṇṭu 75ab
+\emph{Sauśrutanighaṇṭu} 75ab
 \begin{versinnote}
@@
 \end{versinnote}
-
-Brahmānanda’s understanding of \emph{harītaśāka} as \emph{patraśāka} is probably wrong if \emph{patraśāka} was intended as ‘leafy vegetables’.
-But perhaps \emph{patraśāka} can also mean \emph{śigru} (note \emph{śākapattraḥ} above).
-
-Anusvāra at end of \emph{śāka}?
-
-\emph{°sauvīra°} probably means sour gruel.
-Brahmānanda: \emph{sauvīra} = \emph{kāñjika} (fermented rice water).
-Meulenbeld, madhavanidāna pp. 516–517
-\emph{sauvīra} is sour gruel made from barley and wheat. On the process see, Suśruta 1.44.35--40ab.
-(PV Sharma’s translation of this passage:)
+\end{quote}
+
+Brahmānanda’s understanding of \emph{harītaśāka} as \emph{pattraśāka} is probably wrong if \emph{pattraśāka} was intended as ‘leafy vegetables.’ But he may have used the term \emph{pattraśāka} in the sense of horseradish (\emph{śigru}) as the dictionary notes that \emph{pattraśāka} is probably equivalent to \emph{śākapattra}, which is mentioned in \emph{Rājanighaṇṭu} 7.26 (above).
+
+The term \emph{°sauvīra°} (1.59b) probably means sour gruel.
+Brahmānanda glosses  \emph{sauvīra} as \emph{kāñjika}, which is `fermented rice water.' On \emph{sauvīra}, Meulenbeld (1974: 516–517) says, `emph{sauvīra} is sour gruel made from barley and wheat.' The process of making it is described in the \emph{Suśrutasaṃhitā} (1.44.35--40ab) as follows:
 
 \begin{quote}
-‘Roots of trivṛt etc., the first group (\emph{vidārigandhādi}), \emph{mahat pañcamūla}, \emph{mūrvā} and \emph{śārṅgaṣṭā}, and also of \emph{snuhī}, \emph{haimavatī}, \emph{triphalā}, \emph{ativiṣā} and \emph{vacā} -- these are taken and divided into two parts out of which one is decocted and the other is powdered; now, crushed barley grains are impregnated with the above decoction several times, dried and then slightly fried.
-Taking three parts of this and one part of the above powder are put in a jar and mixed with their (of \emph{trivṛt}, etc.) cold decoction and fermented properly.
-This is known as \emph{sauvīraka}.’
+Roots of trivṛt etc., the first group (vidārigandhādi), mahat pañcamūla, mūrvā and śārṅgaṣṭā, and also of snuhī, haimavatī, triphalā, ativiṣā and vacā -- these are taken and divided into two parts out of which one is decocted and the other is powdered; now, crushed barley grains are impregnated with the above decoction several times, dried and then slightly fried.
+Taking three parts of this and one part of the above powder are put in a jar and mixed with their (of trivṛt, etc.) cold decoction and fermented properly.
+This is known as sauvīraka.
+(trans.
+Sharma 2018 (vol.1): 406)
 \end{quote}
 
-But according to some nighaṇṭus, \emph{sauvīra} can also mean stibnite (an ingredient in some añjana’s and medicines):
-
-Rājanighaṇṭu 13.86
+However, according to some Nighaṇṭus, \emph{sauvīra} can also mean stibnite (an ingredient in some añjana’s and medicines). For example, in the \emph{Rājanighaṇṭu} (13.86):
 
@@
 
-Note also that the Yogaprakāśikā takes \emph{sauvīra} with \emph{taila}, perhaps to solve the problem of \emph{taila} on its own (see below for more on this): \emph{sauvīrataila} -- oil produced in the place Suvīra (\emph{suvīradeśodbhavatailam}).
-
-Suvīra , a country the people of which, also called Suvira (V.79), Sauvira (XVI.21) and Sauvīraka (IV.23) ...
-S.M.
-Ali identifies it with the Rohri - Khairpur region of Sind (Geography of the Purānas, Delhi, 1966, p.
-144).
-
-\emph{taila} could refer to \emph{tilataila}:
-Śārṅgadharasaṃhitā:
-\emph{anuktāvasthāyāṃ paribhāṣāvidhiḥ [...] taile ’nukte tilodbhavam} 48
-
-Dominik Wujastyk supplied this reference and may be able to comment more on taila in this list.
-
-On the translation of \emph{madya}, see Mchugh (An Unholy Brew) 2021: 8.
-
-\emph{ājādimāṃsa}: āja° is required for the metre, so only ājādi makes sense, not ājāvi because there is no adjective āvi.
-
-Diwakar Acharya: prohibition of goat flesh and fish is aimed at north/east India.
-
-\emph{kulattha} is a kind of pulse, translated by Dominik Wujastyk as `poor man's pulse' (see Roots of Ayurveda).
-
-\emph{kola}: Zizyphus Jujuba (Nadkarni 1926: pp.
+Note also that the \emph{Yogaprakāśikā} (1.53) takes \emph{sauvīra} with \emph{taila}, perhaps to solve the problem of \emph{taila} on its own (see below for more on this).
+The compound \emph{sauvīrataila} is explained as `oil produced in the place Suvīra' (\emph{suvīradeśodbhavatailam}).
+According to Ali (1966: 144), Suvīra is known as a country that was also called Suvira (V.79), Sauvira (XVI.21) and Sauvīraka (IV.23).
+He identifies it with the Rohri/Khairpur region of Sind.
+%( S.M. Ali, Geography of the Purānas, Delhi, 1966, p. 144).
+
+The word \emph{taila} could refer to \emph{tilataila}. This is supported by the following rule (\emph{paribhāṣā}) in the \emph{Śārṅgadharasaṃhitā} (48):
+\emph{anuktāvasthāyāṃ paribhāṣāvidhiḥ} [...] \emph{taile ’nukte tilodbhavam}. We thank Dominik Wujastyk for this reference.
+
+Our translation of \emph{madya} takes into account the following remarks of James McHugh (2021: 8):
+% Mchugh, An Unholy Brew)
+\begin{quote}
+    The most general Sanskrit term to denote drinks that create a drunken state is \emph{madya} “intoxicating [drink].” Translating this word is hard.
+“Inebriating drink” is clumsy to my ear.
+“Intoxicating” contains the unfortunate “toxic” element that is not present in the Sanskrit word, though at least in English this is a common word, applicable to various substances and states and lacking any “toxic” associations in everyday usage.
+\end{quote}
+
+In the compound \emph{ājādimāṃsa}° (1.59c), the word \emph{āja}° is required for the metre.
+So, variants beginning with \emph{aja}° can be dismissed.
+Another well-attested reading is \emph{ājāvimāṃsa}°.
+Although this was probably read as `goat and sheep flesh,' \emph{āvi} is not attested as an adjective of sheep, so this reading was probably not original.
+Moreover, only \emph{ājādimāṃsa}° makes good sense.
+Diwakar Acharya believes that the prohibition of goat flesh and fish in this verse suggests it derives from the North East of India.
+
+The term \emph{kulattha} is a kind of pulse, translated by Dominik Wujastyk (1998: 77) as `poor man's pulse.'
+% Dominik Wujastyk, Roots of Ayurveda, Penguin Books, 1998.
+
+The word \emph{kola} is a name for Zizyphus Jujuba (Nadkarni 1926: pp.
 919--920).
-Also known as \emph{badara}.
-This is how Brahmānanda takes it (\emph{kolaṃ kolyāḥ phalaṃ badaram}).
+It is also known as \emph{badara}.
+This is how Brahmānanda understands it in \emph{Jyotsnā} 1.59 (\emph{kolaṃ kolyāḥ phalaṃ badaram}).
 According to Nadkarni, the fruit of the wild variety is very acid and astringent.
@@
 The bark is astringent and a simple remedy for diarrhoea.
-Root is useful as a decoction in fever and delirium.
-kola can mean banana in some parts of India (Diwakar)
-
-There are references to \emph{kola} being pungent, though this does not seem to indicate sufficiently why \emph{kola} is mentioned separately as an \emph{apathya} food.
-
-\emph{piṇyāka}: Sharma (Ḍalhaṇa and his commentary on drugs: 1982: 69) says, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’ Diwakar says it is an oil cake that has a strong flavour.
-
-\emph{hiṅgu}: Asafoetida (Nadkarni 1926: pp 360–361): `If long continued even in moderate doses, it gives rise to alliaceous eructations, acrid irritation in the throat, flatulence, diarrhoea and burning in the urine.'
-
-\emph{laśuna} = garlic (Nadkarni 1926: 45).
+The root is useful as a decoction in fever and delirium.
+There are references to \emph{kola} being pungent, though this does not seem to indicate sufficiently why \emph{kola} is mentioned separately in the \emph{Haṭhapradīpikā} as an unwholesome food.
+Diwakar Acharya has informed us that \emph{kola} can refer to a type of banana in some parts of India.
+
+According to Sharma (1982: 69), \emph{piṇyāka} is, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’ Diwakar says it is an oil cake that has a strong flavour, which may account for its inclusion in this list of unwholesome foods.
+% P.V. Sharma, Ḍalhaṇa and his commentary on drugs, 1982.
+
+The term \emph{hiṅgu} is Asafoetida (Nadkarni 1926: 360–361). As to why it might be considered unwholesome, the following comments by Nadkarni give some indication:
+\begin{quote}
+If long continued even in moderate doses, it gives rise to alliaceous eructations, acrid irritation in the throat, flatulence, diarrhoea and burning in the urine.
+\end{quote}
+
+%\emph{laśuna} = garlic (Nadkarni 1926: 45).
 \end{philcomm}
@@
 
-\begin{sources}[hp01_060]
-\end{sources}
+%\begin{sources}[hp01_060]
+%\end{sources}
 
 \begin{testimonia}[hp01_060]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{bhojanam ahitaṃ vidyāt punar <apy> uṣṇīkṛtaṃ tathā rukṣam |\\+}
+\emph{Yogacintāmaṇi} f. 55v (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{bhojanam ahitaṃ vidyāt punar uṣṇīkṛtaṃ tathā |\\+}
 \tl{atilavaṇaṃ sapalaṃ vā prasitaṃ śākotkaṭaṃ varjyam ||\\!}
@@
 
-HSC
+\emph{Haṭhasaṅketacondrikā}
 
@@
 \begin{philcomm}[hp01_060]
-We have understood the compound \emph{kadaśanaśākotkaṭa} as a dvandva referring to spoiled food (\emph{kadaśana}) and an excess of leafy vegetables (\emph{śākotkaṭa}).
-We have not found any conclusive evidence for the meaning of \emph{tilapiṇḍa}. Brahmānanda glosses it as \emph{piṇyāka} (on which see the notes for the previous verse).
-
-Not sure how to take \emph{kadaśanaśākotkaṭaṃ}.
-Brahmānanda understands it as a dvandva (i.e., \emph{kadaśana}, \emph{śāka}, \emph{utkaṭa}), where \emph{śāka} is prohibited vegetables and \emph{utkaṭa} is pepper.
-
-The meaning of \emph{utkaṭa} is not clear.
-The word \emph{utkaṭā} can mean pepper according to some nighaṇṭus (e.g., Rājanighaṇṭu 5.16 \emph{pārvatī śailajā tāmrā lambabījā tathotkaṭā}).
-But \emph{utkaṭa} can refer to Saccharum Sara and \emph{utkaṭā} also to Laurus Cassia (cinnamon).
-
-Also, \emph{utkaṭa} can be an adjective that means ‘abounding in’ at the end of a compound.
-So could \emph{kadaśanaśākotkaṭaṃ} have been intended as an adjectival tatpuruṣa along the lines of ‘[food] full of spoiled vegetables’?
+We have not found any conclusive evidence for the meaning of \emph{tilapiṇḍa}. Brahmānanda glosses it as \emph{piṇyāka} (on the meaning of which see the notes for the previous verse).
+
+The meaning of the compound \emph{kadaśanaśākotkaṭaṃ} is not clear.
+Brahmānanda understands it as a list (\emph{dvandva}) consisting of \emph{kadaśana}, \emph{śāka} and \emph{utkaṭa}, which he defines as bad food, prohibited vegetables and pepper, respectively.
+
+There are various possible meanings of \emph{utkaṭa}.
+According to some Nighaṇṭus, the word \emph{utkaṭā} can mean pepper  (e.g., \emph{Rājanighaṇṭu} 5.16 \emph{pārvatī śailajā tāmrā lambabījā tathotkaṭā}) and, according to the dictionary (s.v.
+Monier Williams), \emph{utkaṭa} can refer to Saccharum Sara and \emph{utkaṭā} also to Laurus Cassia (cinnamon).
+
+However, \emph{utkaṭa} can be an adjective that means ‘abounding in’ at the end of a compound.
+Since this verse consists of many adjectives describing food that is unwholesome, it is likely that \emph{kadaśanaśākotkaṭaṃ} was intended as an adjectival \emph{tatpuruṣa}, in which case it means ‘[food] full of vegetables' \emph{śākotkaṭa} that are `bad food' or perhaps, `bad eating' \emph{kadaśana} in the sense of hard to chew.
 \end{philcomm}
@@
 
-\begin{sources}[hp01_061]
-Amṛtasiddhi 19.7
-
-\begin{versinnote}
-\tl{agnisevābalāsevā pathasevā ca sarvadā |\\+}
-\tl{prathamābhyāsakāle tu saṃtyājyā yoginā sadā || 19.7 ||\\!}
-\end{versinnote}
-
-\end{sources}
+%\begin{sources}[hp01_061]
+%Cf. Amṛtasiddhi 19.7 % this should be quoted for the verse 3.32
+
+%\begin{versinnote}
+%\tl{agnisevābalāsevā pathasevā ca sarvadā |\\+}
+%\tl{prathamābhyāsakāle tu saṃtyājyā yoginā sadā || 19.7 ||\\!}
+%\end{versinnote}
+
+%\end{sources}
 
 \begin{testimonia}[hp01_061]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 48v
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 1.73
 
@@
 \tl{varjayed durjanaprītivahnistrīpathasevanam |\\+}
-\tl{prātaḥsnānopavāsādi kāyakleśādikaṃ tathā || 1.73 ||\\!}
+\tl{prātaḥsnānopavāsādi kāyakleśādikaṃ tathā ||\\+}
+\tl{\var{°prīti°] °prāntaṃ P, prāptaṃ T,t1. kāyakleśādikaṃ ] kāyakleśavidhiṃ P,T.}\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 4.18 (attr. \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{varjayed durjanaprītiṃ vahnistrīpathasevanam |\\+}
+\tl{prātaḥsnānopavāsādikāyakleśavidhiṃ tyajet ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_061]
-The vulgate has a parallel from the Amaraugha added. Stemma requires prāṃtam, which is difficilior lectio in relationship to prītiṃ.
+Manuscripts from the alpha, beta and epsilom groups have the reading \emph{durjanaprāntaṃ} (1.61a), which is the lectio difficilior in relation to \emph{durjanaprītiṃ} (`the friendship of wicked people').
+We have understood \emph{durjanaprānta} in line with Brahmānanda's gloss in \emph{Jyotsnā} 1.64, `a dwelling near bad people' (\emph{durjanasamīpavāsa}) or, in other words, living near bad people.
 \end{philcomm}
@@
 \begin{translation}[hp01_062]
-Pure food with wheat, rice, śāli rice, barley, sixty-day śāli rice, milk, ghee, cream, unclarified fermented butter, ground sugar and honey.
+Pure food with wheat, rice, śāli rice, barley, sixty-day \emph{śāli} rice, milk, ghee, cream, fresh butter, ground sugar and honey.
 Dried ginger, fruit of the snake gourd and so forth, the five vegetables, mung beans and so on, and rain water are wholesome for the best of sages.
@@
 
-\begin{sources}[hp01_062]
-\end{sources}
+%\begin{sources}[hp01_062]
+%\end{sources}
 
 \begin{testimonia}[hp01_062]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} f. 54v (attr. \emph{Haṭhapradīpikā})
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 1.71
 
@@
 \tl{śuṇṭhīpaṭolaphalapatrajapañcaśākaṃ\\+}
-\tl{mudgādidivyam udakaṃ ca yamīndrapathyam || 1.71 ||\\!}
-\end{versinnote}
-
-YBD
+\tl{mudgādidivyam udakaṃ ca yamīndrapathyam ||\\+}
+\tl{\var{°phalapatraja° ] plalādika N,n1,J. yamīndra° ] yatīndra° N,n1,J}\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 4.21
 
@@
 \tl{śuṇṭhīpaṭolakaphalādi ca pañcaśāka-\\+}
-\tl{mudgādidivyam udakaṃ ca munīndrapathyam || 21 ||\\!}
+\tl{mudgādidivyam udakaṃ ca munīndrapathyam ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_062]
-\emph{khaṇḍa} -- candied sugar (Meul 507, different types of sugar).
-
-\emph{navanīta} (MW fresh butter), Mchugh (2021) unclarified fermented butter.
-
-\emph{sitā} -- ground sugar (Meul 507, different types of sugar) ``sitā is very white and looks like gravel"
-
-\emph{madhu} -- honey.
-
-\emph{paṭola} can refer to at least two different gourds. See Meul. p. 569 for a long list of possibilities, including TRICHOSANTHES DIOICA ROXB. (`pointed gourd'), T. CUCUMERINA LINN (snake gourd).
-
-Nadkarni has two entries on \emph{paṭola}:
+In 1.62b, \emph{maṇḍa}, which is supported by alpha, beta and gamma, is more likely than \emph{khaṇḍa} (`candied sugar') because fits the context of diary products mentioned in this compound (i.e., \emph{kṣīra, ājya} and \emph{navanīta}).
+The term \emph{navanīta} is discussed in \emph{Suśrutasaṃhitā, sūtrasthāna,} 45.92 as follows:
+\begin{quote}
+Fresh butter (\emph{navanīta}) is light soft, sweet, astringent, slightly sour, cold, intellect-promoting, appetiser, cordial, checking, aphrodisiac, non-burning, pacifies pitta and vāta and alleviates wasting, cough, wound, cinsumption, piles and facial paralysis [...] (trans.
+Sharma 2018 vol.
+1: 434)
+\end{quote}
+
+The word \emph{sitā} is one of many words for ground sugar. Meulenbeld (1974: 507) comments that \emph{sitā} is `very white and looks like gravel.'
+
+Thw term \emph{paṭola} can refer to at least two different gourds.
+Meulenbeld (1974: 569) compiled a list of six possibilities, including \textsc{Trichosanthes dioica} \textsc{Roxb}.
+(`pointed gourd'), \textsc{Trichosanthes cucumerina} \textsc{Linn} (`snake gourd').
+
+Nadkarni (1954: 863, 518) has two entries on \emph{paṭola}:
+%Nadkarni, A.
+K.
+(1954).
+Dr.
+K.
+M.
+Nadkarni’s Indian Materia Medica, with Ayurvedic, Unani-tibbi, Siddha, Allopathic, Homeopathic, Naturopathic & Home Remedies, Appendices & Notes.
+2 vols.
+Bombay: Popular Prakashan.
+url: https: //archive.org/details/IndianMateriaMedicaK.M.Nadkarni (on 11 Aug.
+2017).
+URL is 1926 edition.
 \begin{enumerate}
-\item snake gourd (Nadkarni p.
-863) is common in Bengal and cultivated in Northern India and Punjab.
+\item Snake gourd is common in Bengal and cultivated in Northern India and Punjab.
 The unripe fruit of this climbing plant is generally used as a culinary vegetable and is very wholesome, specially suited for the convalescent.
 
-\item smooth luffa (Nadkarni p.
-518) is a hairy climbing herb extensively cultivated in several parts of India.
+\item Smooth luffa is a hairy climbing herb extensively cultivated in several parts of India.
 The fruit is edible.
-Medicinally it is described as `cool, costive, demulcent, producive of loss of appetite and excitive of wind, bile and phlegm")
+Medicinally it is described as `cool, costive, demulcent, producive of loss of appetite and excitive of wind, bile and phlegm.
 \end{enumerate}
-
-Wikipedia : smooth luffa = Luffa aegyptiaca (sponge gourd)
-Sharma (Syn. Kulaka. Well known (Trichosanthas dioica Roxb.)
-
-Brahmānanda glosses it as \emph{kośātakī} (Meul p. 586 LUFFA ACUTANGULA ROXB), which suggests he thought it was some sort of luffa.
-
-Brahmānanda also mentions the vernacular term \emph{paravara} for \emph{paṭola}, which the Lonavla ed.
-states is a kind of cucumber.
-However, Paras remarked that \emph{paravara} is more like a gourd (hard shell, etc.).
-
-On \emph{pañcaśāka}, see GhS
-
+Sharma (1982: 156) adds that \emph{paṭola} is a synonym of \emph{kulaka} and is well known as \textsc{Trichosanthas dioica Roxb.} Brahmānanda glosses \emph{paṭola} as \emph{kośātakī}, which is \textsc{Luffa acutangula Roxb} (Meulenbeld 1974: 586), suggesting that he thought it was some sort of luffa.
+He also mentions the vernacular term \emph{paravara} for \emph{paṭola}.
+
+Groups of five vegetables (\emph{pañcaśāka} or \emph{śākapañcaka}) have been defined in various yoga texts, but such grouping of vegetables does not seem to occur outside of literature on yoga.
+The earilest reference to a group of five vegetables known to us is the sixteenth-century \emph{Yuktabhavadeva} 4.22, which attributes the verse to the \emph{Śivayoga}.
+The same verse is also quoted in \emph{Jyotsnā} 1.65 with attribution to a medical source (\emph{vaidyaka}):
+
+\begin{quote}
+\begin{versinnote}
+\tl{sarvaśākam acākṣuṣyaṃ cākṣuṣyaṃ śākapañcakam |\\+}
+\tl{jīvantī vāstumatsyākṣī meghanādaḥ punarnavāḥ || iti ||\\!}
+\end{versinnote}
+\end{quote}
+Another verse on a similar fivefold group of vegetables is also cited in the \emph{Haṭhatattvakaumudī} (4.26)
+\begin{quote}
+\begin{versinnote}
+\tl{pañcaśākas tu–\\+}
+\tl{kṣīraparṇī ca jīvantī matsyākṣī ca punarnavā\\+}
+\tl{meghanādaś ceti budhaiḥ pañcaśākaḥ prakīrtitaḥ || iti ||\\!}
+\end{versinnote}
+\end{quote}
+And a group with more significant differences is mentioned in the \emph{Gheraṇḍasaṃhitā} (5.20).
+\begin{quote}
 \begin{versinnote}
 \tl{bālaśākaṃ kālaśākaṃ tathā paṭolapatrakam |\\+}
-\tl{pañcaśākaṃ praśaṃsīyād vāstūkaṃ hilamocikāṃ || 5.20 ||\\!}
-\end{versinnote}
-
-HTK 4.26
-
-\begin{versinnote}
-\tl{pañcaśākastu –\\+}
-\tl{kṣīraparṇī ca jīvantī matsyākṣī ca punarnavā\\+}
-\tl{meghanādaś ceti budhaiḥ pañcaśākaḥ prakīrtitaḥ || iti || 26\\!}
-\end{versinnote}
-
-Jyotsnā and Yuktabhavadeva 4.22
-
-\begin{versinnote}
-\tl{sarvaśākam acākṣuṣyaṃ cākṣuṣyaṃ śākapañcakam |\\+}
-\tl{jīvantī-vāstu-matsyākṣī-meghanāda- punarnavāḥ || iti ||\\!}
-\end{versinnote}
-
-It is not entirely clear how we should understand \emph{divya}.
-Brahmānanda glosses it with \emph{nirdoṣa} and takes it with \emph{udaka}.
-But could it refer more specifically to \emph{gaṅgāmbu} (as suggested by Paras) or rain water?
-MW has \emph{divyodaka} n.
-`divine water' i.e.
-rainwater L.
-
-The term \emph{divyodaka} appears in Āyurvedic works (but we’re yet to find a gloss in a commentary). E.g.,
-Aṣṭāṅgahṛdaya 8.42–43
-
-\begin{versinnote}
-\tl{śīlayec chāligodhūmayavaṣaṣṭikajāṅgalam |\\+}
-\tl{suniṣaṇṇakajīvantībālamūlakavāstukam || 42 ||\\+}
-\tl{pathyāmalakamṛdvīkāpaṭolīmudgaśarkarāḥ |\\+}
-\tl{ghṛtadivyodakakṣīrakṣaudradāḍimasaindhavam || 43 ||\\!}
-\end{versinnote}
-
-SriKanta Murty translates \emph{divyodaka} as ‘divyodaka (rain water or pure water)'.
-
-The Rājanighaṇṭu says rainwater:
-
+\tl{pañcaśākaṃ praśaṃsīyād vāstūkaṃ hilamocikāṃ ||\\!}
+\end{versinnote}
+\end{quote}
+
+It is not entirely clear how one should understand \emph{divya} (1.62d).
+Brahmānanda glosses it with \emph{nirdoṣa} (`defectless') and takes it with \emph{udaka}.
+Ayurvedic sources indicate more clearly that \emph{divyodaka} was understood as rainwater.
+In a section on types of water (\emph{jalavarga}) in the \emph{Sūtrasthāna} of the \emph{Carakasaṃhitā} (27.196–224), rainwater is referred to as `\emph{divyaṃ udaka}' (1.27.198) in a discussion of the properties of water that has fallen from the sky.
+The compound \emph{divyodaka} is used in other Āyurvedic works to refer to the use of rainwater in recipes and treatments  (e.g., \emph{Aṣṭāṅgahṛdaya} 8.42–43).
+Also, the \emph{Rājanighaṇṭu} (14.4) glosses \emph{divyodaka} as rainwater:
+\begin{quote}
 \begin{versinnote}
 \tl{divyodakaṃ kharāri syād ākāśasalilaṃ tathā |\\+}
-\tl{vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam || Rajni 14.4\\!}
-\end{versinnote}
-
-Kharāri? Maybe \emph{khavāri} was intended.
-
-Vācaspatyam:
-\textbf{divyodaka} na° karma°.
-1 antarīkṣabhave jale divyaśabde bhāva° pra° vākye tadbhedādi dṛśyam.
-ambuśabde vivṛtiḥ .
-
-62*1 is quoted in the Jyotsna as from a medical work (“vaidyake”).
+\tl{vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam ||\\!}
+\end{versinnote}
+\end{quote}
+%
+%
+%62*1 is quoted in the Jyotsna as from a medical work (“vaidyake”).
 \end{philcomm}
@@
 \begin{translation}[hp01_063]
-The yogi should eat food that is sweet, delicious, unctuous, contains cow products, nourishes the bodily constituents, is desired by the mind and is appropriate.
-\end{translation}
-
-\begin{sources}[hp01_063]
-\end{sources}
+The yogi should eat food that is sweet, delicious, unctuous, contains cow products, nourishes the bodily constituents (\emph{dhātu}), is desired by the mind and is appropriate.
+\end{translation}
+
+%\begin{sources}[hp01_063]
+%\end{sources}
 
 \begin{testimonia}[hp01_063]
-Yogacintāmaṇi
+
+\emph{Yogacintāmaṇi} f. 54v (attr. \emph{Haṭhapradīpikā})
 
@@
 \tl{piṣṭaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\+}
-\tl{manobhilaṣitaṃ yogyaṃ yogī bhojanam ācared iti ||\\!}
-\end{versinnote}
-
-YBhD
-
-\begin{versinnote}
-\tl{śreṣṭhaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam ||\\+}
-\tl{mano'bhilaṣitaṃ yogyaṃ yogī bhojanamācaret || 23 ||\\!}
+\tl{manobhilaṣitaṃ yogyaṃ yogī bhojanam ācaret || iti ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 1.75
+
+\begin{versinnote}
+\tl{śreṣṭhaṃ samadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\+}
+\tl{manobhilaṣitaṃ yogyaṃ caturthāṃśavivarjitam |\\+}
+\tl{śivārpitaṃ ca naivedyaṃ yogī bhojanam ācaret ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 4.23 (attr. \emph{Śivayoga})
+
+\begin{versinnote}
+\tl{śreṣṭhaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\+}
+\tl{mano'bhilaṣitaṃ yogyaṃ yogī bhojanamācaret ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_063]
-The variants of 1.63a all seem possible: \emph{mṛṣṭaṃ}, \emph{miṣṭaṃ} and \emph{iṣṭaṃ}. Maybe the last is made redundant by \emph{mano 'bhilaṣitaṃ}.
+The variants of 1.63a all seem possible: \emph{mṛṣṭaṃ}, \emph{miṣṭaṃ} and \emph{iṣṭaṃ}.
+The last is made somewhat redundant by \emph{mano 'bhilaṣitaṃ} in 163c.
+Both \emph{mṛṣṭaṃ} (alpha and eta) and \emph{miṣṭaṃ} (beta and gamma) are both well attested by manuscripts of important groups and there is hardly any difference in their meaning in this context.
+We have adopted \emph{mṛṣṭaṃ} as it is supported by the alpha group.
 \end{philcomm}
@@
 \begin{sources}[hp01_064]
-DYS 40
+\emph{Dattātreyayogaśāstra} 40
 
@@
 \begin{testimonia}[hp01_064]
-Yogacintāmaṇi
+\emph{Yogacintāmaṇi} 15r
 
@@
 
-Haṭharatnāvalī
+\emph{Haṭharatnāvalī} 1.23
 
@@
 \tl{yuvā bhavati vṛddho 'pi vyādhito durbalo 'pi vā |\\+}
-\tl{abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ || 1.23 ||\\!}
+\tl{abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_064]
-Note the different reading in 164d for V1: \emph{sarvaṃ yogī yatendriyaḥ}.
-\emph{sarvaṃ} is not easy to construe, and the testimony of the DYŚ suggests that \emph{yoge sarvo ’py atandritaḥ} was original.
+Eta one, the oldest dated manuscript, has a different reading for the last verse quarter (164d): \emph{sarvaṃ yogī yatendriyaḥ} (`the yogi whose senses are restrained wholly succeeds [...]')  .
+Here, \emph{sarvaṃ} is not easy to construe, and the testimony of the \emph{Dattātreyayogaśāstra}, the alpha manuscripts and other important groups of \emph{Haṭhapradīpikā} manuscripts indicates that \emph{yoge sarvo ’py atandritaḥ} was original.
 \end{philcomm}
@@
 \begin{testimonia}[hp01_065]
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{pīthāni kumbhakāś citrā divyāni karaṇāni ca |\\+}
-\tl{sāṅgo 'pi ca haṭhābhyāso rājayogaphalārthadaḥ || 1.17 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_065]
-\end{philcomm}
+\emph{Haṭharatnāvalī} 1.17
+
+\begin{versinnote}
+\tl{pīṭhāni kumbhakāś citrā divyāni karaṇāni ca |\\+}
+\tl{sāṅgo 'pi ca haṭhābhyāso rājayogaphalārthadaḥ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp01_065]
+%\end{philcomm}
 

### 2023-10-25 compile-tidyup.sh

@@ -6,7 +6,7 @@ xslcmdproc(){
     xslcmd="java -jar ${HOME}/.nix-profile/saxon9he.jar"
 
     echo "processing xml …"
-    $xslcmd -s:xml/HP1_edition-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp1.html \
+    $xslcmd -s:xml/HP1_dev-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp1.html \
 	    chapid="hp1" \
 	    transl="../xml/HP1_TranslComm-tei.xml" \
 	    marma="../xml/Marmasthanas-tei.xml" \
@@ -89,11 +89,12 @@ xmlizejyotsna(){
 
 if [ -f "latex/${1}" ]
 then
+    echo "found $1"
     rm "latex/${1%.*}.pdf"
     compile "${1#*/}"
 elif [ -z "${1:-}" ]
 then
-    for f in "HP1_edition.tex" "HP1_TranslComm.tex" "Marmasthanas.tex" "HP2_edition.tex" "HP2_TranslComm.tex"
+    for f in "HP1_dev.tex" "HP1_TranslComm.tex" "Marmasthanas.tex" "HP2_edition.tex" "HP2_TranslComm.tex"
     do
 	compile "${f}"
     done

### 2023-12-02 latex/HP1_comm.tex
adapted to new input files and cleanup

--- 2023-10-25 latex/HP1_TranslComm.tex
@@
 \begin{translation}[hp01_001]
-Homage to the glorious Ādinātha who taught the system of Haṭhayoga, which shines forth like a stairway for one desirous of climbing to the lofty terrace of the royal palace.%??JM: would much prefer "science" to "system" (vidyā doesn't really mean system either).
+Homage to the glorious Ādinātha who taught the system of Haṭhayoga, which shines forth like a stairway for one desirous of climbing to the lofty terrace of the royal palace.%??JM: would much prefer "science" to "system" (vidyā doesn't really mean system).
 I changed "ladder" to "stairway" (a bit more poetic).
-I'd also prefer "which is like a glorious stairway for one..."
+I'd also prefer "which is like a splendid stairway for one who wants to climb to the lofty royal terrace."
 \end{translation}
@@
 \begin{versinnote}
-\tl{sadādināthāya namo'stu tubhyaṃ yenopadiṣṭā haṭhayogavidyā |\\+}
-\tl{virājate pronnatarājayogam āroḍhum icchoradhirohiṇīva ||\\!}
+\tl{sadādināthāya namo 'stu tubhyaṃ yenopadiṣṭā haṭhayogavidyā |\\+}
+\tl{virājate pronnatarājayogam āroḍhum icchor adhirohiṇīva ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_001]
-The reading \emph{rājasaudha} is preferable for poetical reasons, as explained in Hanneder 2020, p.
+%?? JM: now saudha is preferred for stemmatic reasons too. I suggest the following shortened form of the note: 
+The alpha witnesses read \emph{virājate pronnatarājayogam} while several others have either \emph{vibhrājate} or \emph{°rājasaudham} or both.
+As explained in Hanneder 2020, \emph{virājate} is stylistically preferable and the comparison between the science of haṭhayoga and a stairway works only with \emph{°rājasaudham}.
+
+%The reading \emph{rājasaudha} is preferable for poetical reasons, as explained in Hanneder 2020, p.
 128–130.
@@
 ALso, do we need to repeat what Jürgen has already written elsewhere?
-The Sanskrit poeticians explain that when some elements of a comparison remains unexpressed we get an incomplete (\emph{lupta}) comparison.
+%The Sanskrit poeticians explain that when some elements of a comparison remains unexpressed we get an incomplete (\emph{lupta}) comparison.
 Often words like \emph{iva} or \emph{yathā} are missing or the common property, but if we read \emph{rājayoga} we lose the \emph{upamāna} (i.e., \emph{rājasaudha}).
@@
 % JB Oct 11: It might be helpful to translate Brahmānanda's comment for the non-Sanskrit reader: yathā pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudhaprāpikā bhavati evaṃ haṭhadīpikāpi pronnatarājayogam āroḍhum icchor anāyāsena rājayogaprāpikā bhavatīti upamālaṅkāraḥ
-"Just like one, who wishes to climb a lofty terrace, reaches the terrace without effort by a
-staircase, in the same way, one, who wishes to climb the lofty royal Yoga, reaches the royal Yoga
-by means of the Haṭhapradīpikā. This is the (poetic figure called) \emph{comparison}."
+% JB: ``Just like one, who wishes to climb a lofty terrace, reaches the terrace without effort by a staircase, in the same way, one, who wishes to climb the lofty royal Yoga, reaches the royal Yoga by means of the Haṭhapradīpikā.
+This is the (poetic figure called) \emph{comparison}."
 
@@
 \begin{translation}[hp01_002]
-Having bowed to the glorious guru, the Lord, the yogi Svātmārāma has taught the system of Haṭhayoga solely for [attaining] Rājayoga.
+After bowing to the glorious guru, the Lord, the yogi Svātmārāma has taught the system of Haṭhayoga solely for [attaining] Rājayoga.
 \end{translation}
@@
 The reading \emph{yogopadiśyate} (\getsiglum{J7,J10,N17,W4}, etc.) is only possible if one accepts that double \emph{sandhi} is a feature of the style of composition, which it is not.
-Moreover, this reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.
+This reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.%??
+JM has taught?
+surely teaches?
+Again I would prefer science.
 \end{philcomm}
@@
 \begin{translation}[hp01_003]
-The compassionate Svātmārāma holds the Lamp on Haṭha for those who are ignorant of the royal path because of wandering in the darkness of many opinions.
+The compassionate Svātmārāma holds the Lamp on Haṭha for those who are ignorant of the royal path because of wandering in the darkness of many opinions.%??JM: "for those who cannot find the royal path because they are lost in the darkness of a multitude of doctrines".
 \end{translation}
@@
 Most witnesses have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b.
-While this reading is well attested by the manuscript transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note the readings of P,T,t1 of the critical edition [Gharote 2009: 5 n.
-2] in the testimonia), the reading of a royal path (\emph{rājamārgam}) is consistent with the metaphor of people wandering from a path in the darkness and, therefore, most probably authorial.
+While this reading is well attested by the manuscript transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note the readings of P,T,T1 of the critical edition [Gharote 2009: 5 n.
+2] in the testimonia), the reading of a royal path (\emph{rājamārgam}) is consistent with the metaphor of people unable to find% a path in the darkness and, therefore, most probably authorial.
 
@@
 \begin{translation}[hp01_004]
-In fact, Matsyendra, Gorakṣa and other [perfected yogis] knew the system of Haṭha, and the yogi Svātmārāma knows it owing to their favour.
+In fact, Matsyendra, Gorakṣa and other [perfected yogis] knew the system of Haṭha, and the yogi Svātmārāma knows it owing to their favour.%JM For Matsyendra etc. discovered
 \end{translation}
@@
 \begin{philcomm}[hp01_005]     
-In Śaiva texts which predate the haṭha corpus, Mīnanātha and Matsyendra are one and the same, but they are differentiated in later Tibetan and Indian lists of siddhas (Mallinson 2019:273 n.35).
-
-Two manuscripts of the alpha and delta groups have the variant reading \emph{°virūpākṣaḥ savālikaḥ} (\getsiglum{N3,J5,V19}) for \emph{°virūpākṣabileśayāḥ}.
-In \getsiglum{N3}, \emph{savālikaḥ} was corrected to \emph{savālmikaḥ}, perhaps in an effort to restore a name similar to Vālmīki, the celebrated author of the \textit{Rāmāyaṇa}.
+In Śaiva texts which predate the Haṭha corpus, Mīnanātha and Matsyendra are one and the same, but they are differentiated in later Tibetan and Indian lists of siddhas (Mallinson 2019: 273 n.35).
+
+Two manuscripts of the alpha and delta groups have the variant reading \emph{°virūpākṣaḥ savālikaḥ} (\getsiglum{J5,V19}) for \emph{°virūpākṣabileśayāḥ}.
+In another alpha manuscript, \getsiglum{N3}, \emph{savālikaḥ} was corrected to \emph{savālmikaḥ}, perhaps in an effort to restore a name similar to Vālmīki, the celebrated author of the \textit{Rāmāyaṇa}.
 \end{philcomm}
@@
 \begin{versinnote}
-\tl{kaṇerī pūjyapādaśca nityanātho nirañjanaḥ |\\+}
-\tl{kapālī bindunāthaśca kākacaṇḍīśvaro gajaḥ |\\!}
+\tl{kaṇerī pūjyapādaś ca nityanātho nirañjanaḥ |\\+}
+\tl{kapālī bindunāthaś ca kākacaṇḍīśvaro gajaḥ |\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp01_010]
-\emph{Yogasārasaṅgraha}, p.53.
+\emph{Yogasārasaṅgraha}, p. 53.
 
@@
 Nonetheless, it appears in over a dozen manuscripts that were consulted for this edition.
-These manuscripts are not close to an early hyparchetype of the text.
+These manuscripts are not close to an early hyp\-archetype of the text.
 \end{philcomm}
@@
 
-\emph{Yuktabhavadeva} 4.25 (attr. śivayoga)
+\emph{Yuktabhavadeva} 4.25 (attr. \emph{Śivayoga})
 
@@
 \begin{translation}[hp01_017]
-Because it is the first auxiliary of haṭha, \emph{āsana} is taught first. This type (\emph{tad}) of \emph{āsana} brings about steadiness, good health and physical fitness.
+Because it is the first auxiliary of Haṭha, \emph{āsana} is taught first. This type (\emph{tad}) of \emph{āsana} brings about steadiness, good health and physical fitness.
 \end{translation}
@@
 \begin{translation}[hp01_018]
-I shall now teach some of the postures which have been accepted by both sages (\emph{muni}), such as Vasiṣṭha, and yogis, such as Matsyendra.
+I shall now teach some of the postures which have been accepted by sages  such as Vasiṣṭha and yogis such as Matsyendra.
 \end{translation}
@@
 Brahmānanda explains that the region of the shank near the knee should be understood by the word ‘knee’ in this verse (\emph{atra jānuśabdena jānusaṃnihito jaṅghāpradeśo grāhyaḥ jānusaṃnihito jaṅghāpradeśaḥ}).
-This is consistent with the earliest known description of \emph{svastikāsana} in the \emph{Pātañjalayogaśāstravivaraṇa} (2.46), which states that the big toe of one foot is tucked in between the shank and thigh of the other so it is not seen (\emph{dakṣiṇaṃ pādāṅguṣṭhaṃ savyenorujaṅghena parigṛhyādṛśyaṃ kṛtvā tathā savyaṃ pādāṅguṣṭhuṃ dakṣiṇenorujaṅkgenādṛśyaṃ parihṛhya yathā ca pārṣṇibhyāṃ vṛṣaṇayor apīḍanaṃ tathā yenāste tat svastikam āsanam}).
+This is consistent with the earliest known description of \emph{svastikāsana} in the \emph{Pātañjalayogaśāstravivaraṇa} (2.46), which states that the big toe of one foot is tucked in between the shank and thigh of the other so it is not seen (\emph{dakṣiṇaṃ pādāṅguṣṭhaṃ savyenorujaṅghena parigṛhyādṛśyaṃ kṛtvā tathā savyaṃ pādāṅguṣṭhaṃ dakṣiṇenorujaṅghenādṛśyaṃ parigṛhya yathā ca pārṣṇibhyāṃ vṛṣaṇayor apīḍaṇaṃ tathā yenāste tat svastikam āsanam}).
 For a discussion of \emph{svastikāsana} in the Pātañjalayoga tradition, see Maas 2018: 68–69.
@@
 \tl{vyutkrameṇātha pāṇibhyāṃ vinyastābhyāṃ vigṛhya ca |\\+}
-\tl{prṣṭhagābhyāṃ padāṅguṣṭhāv etad gomukham ucyate || \\!}
+\tl{pṛṣṭhagābhyāṃ padāṅguṣṭhāv etad gomukham ucyate || \\!}
 \end{versinnote}
@@
 \begin{versinnote}
-\tl{savye dakṣiṇagulphaṃ tu pṛṣthapārśve niveśayet |\\+}
-\tl{dakṣiṇe'pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā ||\\!}
+\tl{savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet |\\+}
+\tl{dakṣiṇe 'pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā ||\\!}
 \end{versinnote}
@@
 \begin{translation}[hp01_021]
-By fixing one foot on one thigh and placing the [other] thigh on the other foot, vīrāsana is taught [to arise].
+By fixing one foot on one thigh and placing the [other] thigh on the other foot, \emph{vīrāsana} is taught [to arise].
 \end{translation}
@@
 \tl{ekaṃ pādam athaikasmin vinyased ūruṇi sthiram ||\\+}
-\tl{itarasmiṃs tathā coruṃ vīrāsanam īritam ||\\+} 
+\tl{itarasmiṃs tathā coruṃ vīrāsanam itīritam ||\\+} 
 \tl{\var{sthiram ] sthitam T}\\!}
@@
 \begin{philcomm}[hp01_021]   
-Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by group 4c, the sources and the testimonia.
+Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by \getsiglum{G4} (alpha group) and \getsiglum{V1} (eta group), the sources and the testimonia.
 It appears to be verse filler here rather than indicating a temporal sequence of actions.
@@
 
-Different versions of \emph{vīrāsana} are found in earlier Tantras, such as the Kiraṇatantra (58.9), Hemacandra’s \emph{Yogaśāstra} and commentaries on the \emph{Pātañjalayogaśāstra}.
+Different versions of \emph{vīrāsana} are found in earlier Tantras, such as the \emph{Kiraṇatantra} (58.9), Hemacandra’s \emph{Yogaśāstra} and commentaries on the \emph{Pātañjalayogaśāstra}.
 For a discussion of some of these sources, see Maas 2018: 66–68.
@@
 \tl{gudaṃ niyamya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\+}
-\tl{kūrmāsanaṃ bhavedetaditi yogavido viduḥ || iti kūrmāsanam ||\\!}
+\tl{kūrmāsanaṃ bhaved etad iti yogavido viduḥ || iti kūrmāsanam ||\\!}
 \end{versinnote}
@@
 In the first quarter of the verse, the witnesses are split between \emph{nirudhya} (‘having blocked’), \emph{nibadhya} (‘having bound’), \emph{niyamya} (‘having restrained’) and \emph{niṣpīḍya} (‘having pressed’).
-The source, the \emph{Vasiṣṭhasaṃhitā}, and two manuscripts of the beta and gamma groups support \emph{nirudhya} whereas the testimonia supports \emph{niyamya}.
-In terms of closing the anus by sitting on the ankles, \emph{nirudhya} makes better sense.
+The source, the \emph{Vasiṣṭhasaṃhitā}, and two manuscripts of the beta and gamma groups support \emph{nirudhya} whereas one alpha manuscript (\getsiglum{J5}) and the testimonia support \emph{niyamya} and another alpha manuscript (\getsiglum{G4}) has \emph{niṣpīḍya}.
+In terms of blocking or closing the anus by sitting on the ankles, \emph{nirudhya} makes good sense, and \emph{niṣpīḍya} (`having pressed the anus with both ankles') is also possible.
 
@@
 2, pp.
-29–30 for a discussion of \emph{vyutkramena} and the position of the ankles in \emph{kūrmāsana}, and vol 8, no.2, Figures 3–6 for photographs of a practitioner performing this \emph{āsana}.
+29–30 for a discussion of \emph{vyutkrameṇa} and the position of the ankles in \emph{kūrmāsana}, and vol 8, no.
+2, Figures 3–6 for photographs of a practitioner performing this \emph{āsana}.
 \end{philcomm}
@@
 \tl{padmāsanaṃ tu saṃyojya jānūrvor antare karau |\\+}
-\tl{niveśya bhūmau saṃsthāpya vyomasthaṃ kukkuṭāsanam || iti kukkuṭāsanam ||\\!}
+\tl{niveśya bhūmau saṃsthāpya vyomasthaṃ kukkuṭāsanam ||\\+}
+\tl{iti kukkuṭāsanam ||\\!}
 \end{versinnote}
@@
 Since the word \emph{ākarṣaṇa} in one form or other is so well attested in the third verse quarter, the reading \emph{dhanurākarṣaṇaṃ kṛtvā}, which is in epsilon two, as well as the principal testimonia (i.e., the \emph{Yogacintāmaṇi} and \emph{Haṭharatnāvalī}), fits the overall syntax of the verse.
-However, it is curious that \emph{kṛṣtaṃ} (for \emph{kṛtvā}) is well attested in some groups of manuscripts because it seems redundant with \emph{ākarṣaṇaṃ}.
+However, it is curious that \emph{kṛṣṭaṃ} (for \emph{kṛtvā}) is well attested in some groups of manuscripts because it seems redundant with \emph{ākarṣaṇaṃ}.
 However, the following reading in
@@
 %https://www.theluminescent.org/2017/11/dhanurasana-two-versions-of-bow-pose.html
+
+One manuscript of the \emph{Haṭhapradīpikā} (ms.
+no.
+30051, f.
+2v), which was consulted but not collated for this edition, has a scribal comment stating that \emph{dhanurāsana} should be done continuously (\emph{anavarata}) on the left and right sides (\emph{tatra ekam dhanurākarṣaṇāsanam āsanaṃ savyāpasavyapādahastābhyām [abhy]ased anavaratam}).
+This would make \emph{dhanurāsana} a moving practice as shown in \medialink{this video}{dhanurasana.mp4}.
 \end{philcomm}
@@
 Having grasped the right foot, which is placed at the base of the left thigh, with the right arm wrapped around the outside of the knee, [the yogi] remains with his body twisted.
-This \emph{āsana} was taught by Matsyendranātha.
+This \emph{āsana} was taught by Matsyendranātha.%??JM: [The yogi] should hold the right foot, which is placed at the base of the left thigh, with the [hand of] the right arm, which is wrapped around the outside of the knee, and remain [like that] with his body twisted.
+This posture was taught by Matsyendranātha.
 \end{translation}
@@
 \tl{vāmorumūlārpitadakṣapādo jānvor bahirveṣṭitadakṣadoṣṇā |\\+}
-\tl{pragṛhya tiṣṭhet partivartitāṅgaḥ śrīmatsyanāthoditam āsanaṃ syāt~||\\+}
+\tl{pragṛhya tiṣṭhet parivartitāṅgaḥ śrīmatsyanāthoditam āsanaṃ syāt~||\\+}
 \tl{\var{°dakṣapādo ] °dakṣapādaṃ P, °dakṣapādau t1}\\!}
@@
 The instrumental ending (`with the hand') works well with the gerund (\emph{pragṛhya}) in the third verse quarter and the object (\emph{°dakṣpādaṃ}) in the first quarter.
-This reading indicates that the right foot is held by the right hand, the arm of which is wrapped around the outside of the left leg (Figure 1).
+This reading indicates that the right foot is held by the hand of the arm that is wrapped around the outside of the left leg, which would be the right hand (\emph{°dakṣadoṣṇā}) rather than the left (\emph{°vāmadoṣṇā}), as shown in \medialink{Figure 1}{matsyendrasana1.jpeg}.
+
+One manuscript of alpha (\getsiglum{G4}) and most manuscripts of the \textit{Haṭharatnāvalī} have \emph{°dakṣapādo} in the first verse quarter.
+This reading yields the same meaning as the adopted one if read with \emph{°vāmapādaṃ} in the second verse quarter.
+However, \getsiglum{G4} and manuscripts of the \textit{Haṭharatnāvalī} read \emph{°vāmadoṣṇā}, which is not good because it leaves the gerund without an object.
 
 The version of this verse in \emph{Jyotsnā} (1.26), which is supported by some manuscripts in two important groups, beta and eta, has two objects of the gerund, namely the left and right feet, without an instrumental or conjuctive particle.
-In his commentarial remarks, Brahmānanda proposes that the left foot is grasped by the right hand and the right foot by the left foot, as seen in Figure 2.
+In his commentarial remarks, Brahmānanda proposes that the left foot is grasped by the right hand and the right foot by the left foot, as seen in \medialink{Figure 2}{matsyendrasana2.jpg}.
 \end{philcomm}
@@
 \begin{philcomm}[hp01_027]
-The manuscript readings for the compound beginning with \emph{jaṭhara} diverge significantly and include \emph{jaṭharapravṛddha}, \emph{jaṭharaprabuddha°}, \emph{jaṭharapradīpta°} and \emph{jaṭharapracaṇḍa°}.
+The manuscript readings for the compound beginning with \emph{jaṭhara} diverge significantly and include \emph{jaṭharapravṛddha°}, \emph{jaṭharaprabuddha°}, \emph{jaṭharapradīpta°} and \emph{jaṭharapracaṇḍa°}.
 As descriptive compounds, none of these make good sense in regard to Matsyendra's seat.
 Since the stomach or abdomen (\emph{jaṭhara}) is the first member of this compound, it seems more likely that it qualifies the terrible diseases (\emph{pracaṇḍarug}) that are mentioned in the next verse quarter, as suggested by the reading \emph{jaṭharapravṛddha°}, which is attested by eta two and the \textit{Yogacintāmaṇi} and suggested by gamma one and delta one (\emph{jaṭharapravuddh°}).
-The reading \emph{jaṭharapracaṇḍa°} appears to be a dittographic error, which in some cases has also resulted in \emph{jaṭharapradaṇḍa°}.
+
+In 1.27d, the compound \emph{daṇḍasthiratvaṃ} (`steadiness of the spine') is attested by all the important manuscript groups and testimonia, so it was likely original.
+However, the \emph{Jyotsnā} (1.27d) has \emph{candrasthiratvaṃ} (`steadiness of the moon'), and this reading is well-attested in many manuscripts that are lower on the stemma.
+Brahmānanda understands steadiness here as `the absence of flow' (\emph{sthiratvaṃ kṣaraṇābhāvaṃ}), a reference no doubt to the moon retaining its nectar.
 \end{philcomm}
@@
 \begin{translation}[hp01_029]
-This back-stretch is the foremost among āsanas. It makes the breath flow in the back (i.e., the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
+This back-stretch is the foremost among \emph{āsana}s.
+It makes the breath flow in the back (i.e., the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
 \end{translation}
@@
  The use of the word \emph{paścima} to mean the central channel is found in \emph{Yogabīja} 121.
-The  \emph{Yogabīja} (95) also refers to the path of the central channel (\emph{paścima\-mārga}), and this understanding of \emph{paścima} is found in the \emph{Jyotsnā 1.29}: \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumnāmārgeṇa vahatīti paścimavāhī}.
+The  \emph{Yogabīja} (95) also refers to the path of the central channel (\emph{paścima\-mārga}), and this understanding of \emph{paścima} is found in the \emph{Jyotsnā} 1.29: \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumnāmārgeṇa vahatīti paścimavāhī}.
 \end{philcomm}
@@
 Supporting oneself on the ground with both hands, [one's] elbows are placed on either side of the navel [and], with a raised position (\emph{uccāsanaḥ}), one is lifted up into the air [as straight] as a stick.
-They call this posture the peacock.
+They call this posture the peacock.%JM: Supporting oneself on the ground with both hands, the elbows placed on either side of the navel, lifted up into the air in a raised posture [as straight] as a stick: they call this posture the peacock.
+
 \end{translation}
@@
 \begin{translation}[hp01_033]
-Śiva taught eighty-four \emph{āsana}s. I shall take the four best and teach them.
+Śiva has taught eighty-four \emph{āsana}s. I shall take the four best and describe them.
 \end{translation}
@@
 The second verse quarter has many variations in the \emph{Haṭhapradīpikā} manuscripts, as well as the manuscripts of the sources and testimonia.
-The alpha manuscripts have \emph{athaikam ekahṛdaye dhṛtvā} and most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}.
+\getsiglum{N3} of the alpha group has \emph{athaikam ekahṛdaye dhṛtvā} and most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}.
 The repetition of \emph{eka} appears to be a dittographical error and \emph{hṛdaye} does not make sense without an object of \emph{dhṛtvā} (i.e., having held what on the chest?).
@@
 \tl{etat siddhāsanaṃ prāhuḥ padmāsanam atho viduḥ |\\+}
-\tl{guptāsanaṃ vadantyeke prāhur vajrāsanaṃ pare |\\+}
+\tl{guptāsanaṃ vadanty eke prāhur vajrāsanaṃ pare |\\+}
 \tl{ke cin muktāsanaṃ prāhur idam āsanam uttamam ||\\!}
@@
 \tl{etat siddhāsanaṃ prāhur anye vajrāsanaṃ viduḥ |\\+}
-\tl{muktāsanaṃ vadanty eke prāhur guptāsanaṃ par ||\\!}
+\tl{muktāsanaṃ vadanty eke prāhur guptāsanaṃ pare ||\\!}
 \end{versinnote}
@@
 \begin{translation}[hp01_038]
-The Siddhas know \emph{siddhasana} as the single most important [pose] amongst all postures, in the same way as a measured diet amongst rules and non-violence amongst observances.
+The Siddhas know \emph{siddhāsana} as the single most important [pose] amongst all postures, in the same way as a measured diet amongst rules and non-violence amongst observances.
 \end{translation}
@@
 \tl{athaikasminn eva dṛḍhaṃ baddhe siddhāsane sadā |\\+}
-\tl{bandhatrayam anāyāsāt svayam evopajāyate |\\!}
+\tl{bandhatrayam anāyāsāt svayam evopajāyate ||\\!}
 \end{versinnote}
@@
 \tl{aṅguṣṭhau hṛdaye nidhāya civukaṃ nāsāgram ālokayet\\+}
-\tl{ekad vyādhivikāranāśanakaraṃ padmāsanaṃ procyate ||\\!}
+\tl{etad vyādhivikāranāśanakaraṃ padmāsanaṃ procyate ||\\!}
 \end{versinnote}
@@
 of the teeth (\emph{dantāgre jihvām ādāya}).
-Other Tantric texts have this or similar rules, in which the tongue is supposed to rest either on the teeth or the palate, early examples being the \emph{Mrgendrāgama}, \textit{yogapāda} 19 (\emph{dantāgre jihvām ādāya}) and \emph{Mataṅgapārameśvaratantra}, \textit{yogapāda} 2.27 (\emph{tālumadhyagatenaiva jihvāgreṇa}).
+Other Tantric texts have this or similar rules, in which the tongue is supposed to rest either on the teeth or the palate, early examples being the \emph{Mṛgendrāgama}, \textit{yogapāda} 19 (\emph{dantāgre jihvām ādāya}) and \emph{Mataṅgapārameśvaratantra}, \textit{yogapāda} 2.27 (\emph{tālumadhyagatenaiva jihvāgreṇa}).
 Placing the tongue where it does not disturb the meditation seems quite appropriate for a `normal' meditative practice.\footnote{The rule of placing the tongue at the palate is also found in \emph{Iśānaśivapaddhati} 18.120: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān asaṃspṛśan ṛjukāyaḥ}}
@@
 \begin{translation}[hp01_047]
-This is taught as lotus pose, which cures all diseases. It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
-\end{translation}
+This is taught as lotus pose, which cures all diseases.
+It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
+[The view of lotus pose] taught next is the one I accept.
+\end{translation}
+% [MD] add translation of "paścāduktam eva matsaṃmatam".
 
@@
 In commenting on \emph{durlabham} in \emph{Jyotsnā} 2.74, Brahmānanda glosses it as \emph{duṣprāpam}, which means `difficult to attain' and `inaccessible.'
+
+The comment added to this verse by Svātmārāma indicates that he prefers the following version of \emph{padmāsana}, which derives from the \emph{Vivekamārtaṇḍa}, rather than the one he has borrowed from the \emph{Dattātreyayogāśāstra}.
 % Maybe add some remarks by commentators on how they understand it.
@@
 %Next is taught the doctrine of Matsya.
-% [MD] add translation of "paścāduktam eva matsaṃmatam" at the end of the previous section.
 
@@
 \tl{vāraṃ vāram apānam ūrdhvam anilaṃ proccālayan pūritaṃ\\+}
-\tl{muṇcan prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ ||\\!}
+\tl{muñcan prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ ||\\!}
 \end{versinnote}
@@
 \begin{versinnote}
-\tl{tathāca granthāntare—\\+}
+\tl{tathā ca granthāntare—\\+}
 \tl{kṛtvā saṃpuṭitau karau dṛḍhataraṃ badhvā ca padmāsanam\\+}
@@
 
-(MD: Or read \emph{pūrayan mārutaṃ dhriyate yas tu} ``he who remains by inhaling..."?
-This \emph{dhriyate} is rather intransitive than passive. 
-Cf. PW dhṛ 22a ``fest sein, sich ruhig verhalten, stillhalten, verbleiben, sich erhalten, bestehen")
+%(MD: Or read \emph{pūrayan mārutaṃ dhriyate yas tu} ``he who remains by inhaling..."?
+%This \emph{dhriyate} is rather intransitive than passive. 
+%Cf. PW \emph{dhṛ} 22a ``fest sein, sich ruhig verhalten, stillhalten, verbleiben, sich erhalten, bestehen")
 
@@
 %(ed.\ Maximilian Mehner and Jürgen Hanneder, Marburg 2019).
-
-% [MD] comment on the additional hemistich "savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe" (N23,Jyo)?
 \end{philcomm}
@@
 \begin{translation}[hp01_055]
-When the great yogi does not become tired from adopting the \emph{āsana}s in this way, he should now practise the breath techniques with seals and so forth, from which purification of the channels arise.
+When the great yogi does not become tired on adopting the \emph{āsana}s in this way, he should  practise the breath techniques of seals and so forth, from which purification of the channels arise.
 \end{translation}
@@
 The second hemistich can be interpreted in different ways.
-One possibility is to understand \emph{nāḍiśuddhiṃ} as a bahuvrīhi qualifying \emph{mudrādipavanakriyām} in the sense that the yogi should practise a breathing technique, along with with \emph{mudrā}s and other mechanisms, such as locks (\emph{bandha}), from which purification of the channels arise.
+One possibility is to understand \emph{nāḍiśuddhiṃ} as a bahuvrīhi qualifying \emph{mudrādipavanakriyām} in the sense that the yogi should practise a breathing technique, along with \emph{mudrā}s and other mechanisms, such as locks (\emph{bandha}), from which purification of the channels arise.
 Alternatively, one could separate \emph{mudrādi} from \emph{pavanakriyām} and understand three different techniques here, namely, the practice of purifying the channels (perhaps by the alternative nostril method mentioned at the beginning of the second chapter), the \emph{mudrā}s and the breathing techniques of \emph{prāṇāyāma}.
@@
 \tl{na veṣadhāraṇaṃ siddheḥ kāraṇaṃ na ca tatkathā |\\+}
-\tl{kriyaivakāraṇaṃ siddheḥ satyam eva tu sāṃkṛte || 46 ||\\+}
+\tl{kriyaiva kāraṇaṃ siddheḥ satyam eva tu sāṃkṛte || 46 ||\\+}
 \tl{śiśnodarārthaṃ yogasya kathayā veṣadhāriṇaḥ |\\+}
@@
 \begin{philcomm}[hp01_055_1]
-1.55*1–2 are omitted from the alpha, gamma and eta groups, so it is likely these verses were not in the earliest versions of the \emph{Haṭhapradīpikā}.
+1.55*1–2 are omitted from the alpha, gamma, delta and epsilon groups, so it is likely these verses were not in the earliest versions of the \emph{Haṭhapradīpikā}.
 In fact, it is likely that both were added (perhaps initially as marginal notes) to elaborate on the word \emph{kriyā} in 1.55d.
@@
 One has to emend to \emph{deyā} to make sense of it.
-These verses (except 1.55.2ef) appear in the \emph{Jyotsna} (1.65–66), but towards the end of chapter one.
+These verses (except 1.55.2ef) appear in the \emph{Jyotsnā} (1.65–66), but towards the end of chapter one.
 \end{philcomm}
@@
 \begin{translation}[hp01_056]
-The various \emph{āsana}, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.
+The various \emph{āsana}s, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.%??JM: Posture, multifarious breath retention, the bodily attitude called seal and then concentration on the internal sound is the sequence of practice in Haṭha.
 \end{translation}
@@
 \begin{versinnote}
-\tl{tad uktaṃ haṭhapradīpikāyām-\\+}
+\tl{tad uktaṃ haṭhapradīpikāyām--\\+}
 \tl{susnigdhamadhurāhārāś caturthāṃśavivarjitaḥ |\\+}
@@
 \begin{versinnote}
-\tl{susnigdhamadhurāhāraścaturthāṃśavivarjitaḥ |\\+}
+\tl{susnigdhamadhurāhāraś caturthāṃśavivarjitaḥ |\\+}
 \tl{bhuñjate śivasaṃprītyā mitāhārī sa ucyate ||\\!}
@@
 \begin{versinnote}
-\tl{annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet |||\\+}
+\tl{annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet |\\+}
 \tl{āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet ||\\!} 
@@
 \begin{sources}[hp01_059]
+cf. DYŚ 70ab
+\tl{lavaṇaṃ sarṣapaś cāmlam uṣṇaṃ rūkṣaṃ ca tīkṣṇakam|\\+}
 \end{sources}
@@
 \begin{quote}
-Cakra mentions as a variant: \emph{katvamlalavaṇakṣāra} (pungent.
-acid, saline and caustic).
+Cakra mentions as a variant: \emph{katvamlalavaṇakṣāra} (pungent, acid, saline and caustic).
 Cakra remarks that the term `hot' (\emph{uṣṇa}) denotes hot on touch when it occurs the first time, and hot with regard to potency when it occurs for the second time.
@@
 \begin{quote}
-If long continued even in moderate doses, it gives rise to alliaceous eructations, acrid irritation in the throat, flatulence, diarrhoea and burning in the urine.
+If long continued, even in moderate doses, it gives rise to alliaceous eructations, acrid irritation in the throat, flatulence, diarrhoea and burning in the urine.
 \end{quote}
@@
 
-\emph{Haṭhasaṅketacondrikā}
+\emph{Haṭhasaṅketacandrikā}
 
@@
 \begin{philcomm}[hp01_061]
-Manuscripts from the alpha, beta and epsilom groups have the reading \emph{durjanaprāntaṃ} (1.61a), which is the lectio difficilior in relation to \emph{durjanaprītiṃ} (`the friendship of wicked people').
+Manuscripts from the alpha, beta and epsilon groups have the reading \emph{durjanaprāntaṃ} (1.61a), which is the lectio difficilior in relation to \emph{durjanaprītiṃ} (`the friendship of wicked people').
 We have understood \emph{durjanaprānta} in line with Brahmānanda's gloss in \emph{Jyotsnā} 1.64, `a dwelling near bad people' (\emph{durjanasamīpavāsa}) or, in other words, living near bad people.
@@
 \tl{mudgādidivyam udakaṃ ca yamīndrapathyam ||\\+}
-\tl{\var{°phalapatraja° ] plalādika N,n1,J. yamīndra° ] yatīndra° N,n1,J}\\!}
+\tl{\var{°phalapatraja° ] phalādika N,n1,J. yamīndra° ] yatīndra° N,n1,J}\\!}
 \end{versinnote}
@@
 \begin{quote}
-Fresh butter (\emph{navanīta}) is light soft, sweet, astringent, slightly sour, cold, intellect-promoting, appetiser, cordial, checking, aphrodisiac, non-burning, pacifies pitta and vāta and alleviates wasting, cough, wound, cinsumption, piles and facial paralysis [...] (trans.
+Fresh butter (\emph{navanīta}) is light soft, sweet, astringent, slightly sour, cold, intellect-promoting, appetiser, cordial, checking, aphrodisiac, non-burning, pacifies pitta and vāta and alleviates wasting, cough, wound, consumption, piles and facial paralysis [...] (trans.
 Sharma 2018 vol.
@@
 \tl{piṣṭaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\+}
-\tl{manobhilaṣitaṃ yogyaṃ yogī bhojanam ācaret || iti ||\\!}
+\tl{mano'bhilaṣitaṃ yogyaṃ yogī bhojanam ācaret || iti ||\\!}
 \end{versinnote}
@@
 \tl{śreṣṭhaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\+}
-\tl{mano'bhilaṣitaṃ yogyaṃ yogī bhojanamācaret ||\\!}
+\tl{mano'bhilaṣitaṃ yogyaṃ yogī bhojanam ācaret ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_064]
-Eta one, the oldest dated manuscript, has a different reading for the last verse quarter (164d): \emph{sarvaṃ yogī yatendriyaḥ} (`the yogi whose senses are restrained wholly succeeds [...]')  .
-Here, \emph{sarvaṃ} is not easy to construe, and the testimony of the \emph{Dattātreyayogaśāstra}, the alpha manuscripts and other important groups of \emph{Haṭhapradīpikā} manuscripts indicates that \emph{yoge sarvo ’py atandritaḥ} was original.
+Eta one, the oldest dated manuscript, has a different reading for the last verse quarter (164d): \emph{sarvaṃ yogī yatendriyaḥ} (`the yogi whose senses are restrained wholly succeeds [...]').
+Here, \emph{sarvaṃ} is not easy to construe, and the alpha manuscripts and other important groups of \emph{Haṭhapradīpikā} manuscripts indicates that \emph{sarvayogeṣv atandritaḥ} was the reading adopted by Svātmārāma, which is more similar to the \emph{Dattātreyayogaśāstra}'s (i.e., \emph{yoge sarvo 'py atandritaḥ}).
 \end{philcomm}

### 2023-12-02 latex/HP3_comm.tex
adapted to new input files and cleanup

--- 2023-08-17 latex/HP1_TranslComm.tex
@@
 \chapter*{Translation \& philological commentary}
-
 %%%%%%%%%%
-\subsection*{1.1}
-\begin{translation}[hp01_001]
-Homage to the glorious Ādinātha by whom the system of Haṭhayoga was taught. It shines forth like a ladder for one desirous of climbing to the lofty terrace of the royal palace.
+\subsection*{3.1}
+\begin{translation}[hp03_001]
+test.
 \end{translation}
-
-\begin{testimonia}[hp01_001]
-Cf. Yogasārasaṃgraha, p. 54.
-
-\begin{versinnote}
-\tl{sadādināthāya namo'stu tubhyaṃ yenopadiṣṭā haṭhayogavidyā |\\+}
-\tl{virājate pronnatarājayogam āroḍhum icchoradhirohiṇīva ||\\!}
-\end{versinnote}
-
-Gheraṇḍasaṃhitā 1.1
-
-\begin{versinnote}
-\tl{ādīśvarāya praṇamāmi tasmai yenopadiṣṭā haṭhayogavidyā |\\+}
-\tl{virājate pronnatarājayogam āroḍhum icchor adhirohiṇīva ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_001]
-The reading \emph{rājasaudha} is preferable for poetical reasons, as explained in Hanneder 2020, p.
-128–130.
-Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison
-(\emph{iva}), (2) the object compared (\emph{upameya}), i.e., the \emph{haṭhayogavidyā} leading to Rājayoga (\emph{rājayogaprāpikā}), (3) the property compared, that is, effortlessness (\emph{anāyāsena}) and, most importantly, (4) an image to compare with, which in this case is \emph{rājasaudha}.
-This idea is confirmed by Brahmānanda, when he sums up and completes the poetical details as: \emph{yathā pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudhaprāpikā bhavati evaṃ haṭhadīpikāpi pronnatarājayogam āroḍhum icchor anāyāsena rājayogaprāpikā bhavatīti upamālaṅkāraḥ}.
-
-The Sanskrit poeticians explain that when some elements of a comparison remains unexpressed we get an incomplete (\emph{lupta}) comparison.
-Often words like \emph{iva} or \emph{yathā} are missing or the common property, but if we read \emph{rājayoga} we lose the \emph{upamāna} (i.e., \emph{rājasaudha}).
-One part of the comparison should not be missing.
-But from the perspective of \emph{alaṃkāraśāstra} the verse has a problem that has surely caused the dilemma: it either leaves the \emph{upameya} or the \emph{upamāna} incomplete.
-Brahmānanda has mentioned both \emph{yathā adhirohiṇī saudhaprāpikā bhavati evaṃ haṭhadīpikā rājayogaprāpikā bhavati}, but had to supply \emph{rājasaudha}, while his text reads \emph{rājayoga}.
-But to have the \emph{upameya} in the text is odd.
-Even Brahmānanda could only know of the image (\emph{rājasaudha}) from the alternative reading he did not accept.
-However, it seems likely that the author would have included the \emph{upamāna} in the text, as in our critical text, so that the reader would know that the upper terrace of the palace is an image for \emph{rājayoga}.
-It is further likely that the substitution of \emph{rājayoga} for \emph{rājasaudha} was the result of the tendency to insert the word \emph{yoga} in the opening verses of the text wherever possible, even where it does not fit, as can be seen in 1.2d (\emph{haṭhayogopadiśyate}) and 1.3b (\emph{rājayogam ajānatām}).
-In the latter case the poetical image has also been lost.
-
-The metre of 1.1 is Indravajrā
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.2}
-\begin{translation}[hp01_002]
-Having bowed to the glorious guru, the Lord, the yogi Svātmārāma has taught the system of Haṭhayoga solely for [attaining] Rājayoga.
-\end{translation}
-
-\begin{philcomm}[hp01_002]
-The reading \emph{yogopadiśyate} (J7,J10,N17,W4, etc.) is only possible if one accepts that double \emph{sandhi} is a feature of the style of composition, which it is not.
-Moreover, this reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.3}
-\begin{translation}[hp01_003]
-The compassionate Svātmārāma holds the Lamp on Haṭha for those who are ignorant of the royal path because of wandering in the darkness of many opinions.
-\end{translation}
-
-\begin{testimonia}[hp01_003]
-Haṭharatnāvalī 1.4
-
-\begin{versinnote}
-\tl{bhrāntyā bahumatadhvānte rājayogam ajānatām |\\+}
-\tl{kevalaṃ rājayogāya haṭhavidyopadiśyate || \\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_003] 
-Most witnesses have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b.
-While this reading is well attested by the manuscripts transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note that P,T,t1 of the critical edition [Gharote 2009: 5 n.
-2] have \emph{rājamārgam ajānataḥ}), the reading of a royal path (\emph{rājamārgam}) is consistent with the metaphor of people wandering from a path in the darkness and, therefore, most probably authorial.
-
-In 1.1d both \emph{kṛpākaraḥ} and \emph{kṣamākaraḥ} are attested.
-The former is preserved by the V1 and group 2 whereas the latter by some manuscripts of the delta group.
-Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākara} makes better sense.
-As Brahmānanda notes, this compound can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}).
-In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} may have arisen as a mistake for \emph{kṛ}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.4}
-\begin{translation}[hp01_004]
-In fact, Matsyendra, Gorakṣa and other [perfected yogis] knew the system of Haṭha, and the yogi Svātmārāma knows it owing to their favour.
-\end{translation}
-
-\begin{testimonia}[hp01_004]
-Haṭharatnāvali 1.3 (Post-15th Century)
-
-\begin{versinnote}
-\tl{haṭhavidyāṃ hi gorakṣamatsyendrādyā vijānate |\\+}
-\tl{ātmārāmo 'pi jānīte śrīnivāsas tathā svayam || \\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_004]   
-The word \emph{athavā} (‘or’) is difficult to construe here.
-Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it.
-The variant \emph{mahāyogī} in group 4c (G5, J4, J11, Ko, P15) avoids this difficulty, but is probably an attempt to remove the difficulty of understanding \emph{athavā}.
-One could emend to \emph{tathā} but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.5}
-\begin{translation}[hp01_005]
-The glorious Ādinātha, Matysendra, Śābara, Ānandabhairava, Cauraṅgī, Mīna, Gorakṣa, Virūpākṣa, Bileśaya,
-\end{translation}
-
-\begin{testimonia}[hp01_005]
-Haṭharatnāvalī 1.80
-
-\begin{versinnote}
-\tl{śrīādināthamatsyendraśābarānandabhairavāḥ |\\+}
-\tl{śāraṅgīmīnagorakṣavirūpākṣabileśayāḥ || \\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_005]     
-In Śaiva texts which predate the haṭha corpus, Mīnanātha and Matsyendra are one and the same, but they are differentiated in later Tibetan and Indian lists of siddhas (Mallinson 2019:273 n.35).   
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.6}
-\begin{translation}[hp01_006]
-Manthānabhairava, Siddhabuddha, and Kanthaḍi, Koraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
-\end{translation}
-
-\begin{testimonia}[hp01_006]
-Haṭharatnāvalī 1.81
-
-\begin{versinnote}
-\tl{manthānabhairavo yogī siddhabuddhaś ca kandalī |\\+}
-\tl{korandakaḥ surānandaḥ siddhipādaś ca carpaṭī ||\\!}
-\end{versinnote}
-
-Caturbhuja Misra's Mugdhāvabodhinī (1.7.8) on the Rasahṛdayatantra
-
-\begin{versinnote}
-\tl{manthānabhairavo yogī siddhabuddhaś ca kanthaḍī |\\+}
-\tl{koraṇṭakaḥ surānandaḥ siddhapādaś ca carpaṭī ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_006]        
-Many manuscripts preserve the reading \emph{pauraṇṭaka}.
-We have accepted \emph{koraṇṭaka} because this name is associated with the \emph{Haṭhābhyāsapaddhati}.
-This reading is well attested by various manuscripts of the \emph{Haṭhapradīpikā}, including those of group 2, as well as those of the \emph{Haṭharatnāvalī} (Gharote 2009: 35 n.
-3), which also attest \emph{gonandaka}.
-Also, \emph{Goraṇṭakuḍu} is the name of a disciple of Gorakṣanātha in the \emph{Navanāthacaritramu} (Jones 2017:194 n.3).
-The compound \emph{siddhapāda} could be a respectful affix.
-However, it seems unlikely here because it would cross the \emph{pāda} break.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.7}
-\begin{translation}[hp01_007]
-Kānerī, Pūjyapāda, Nityanātha, Nirañjana, Kapālī, Bindunātha, and the one named Kākacaṇḍīśvara.
-\end{translation}
-
-\begin{testimonia}[hp01_007]
-Haṭharatnāvalī 1.82
-
-\begin{versinnote}
-\tl{karoṭiḥ pūjyapādaś ca nityanātho nirañjanaḥ |\\+}
-\tl{kapālī bindunāthaś ca kākacaṇḍīśvarāhvayaḥ || \\!}
-\end{versinnote}
-
-Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
-​​
-\begin{versinnote}
-\tl{kaṇerī pūjyapādaśca nityanātho nirañjanaḥ |\\+}
-\tl{kapālī bindunāthaśca kākacaṇḍīśvaro gajaḥ |\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_007]   
-It is possible that \emph{pūjyapāda} could be a respectful affix to the name Kāṇerī.
-The variant \emph{dhvaninātha} may have resulted from a transposition of the first two syllables of \emph{nityanātha}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.8}
-\begin{translation}[hp01_008]
-Allamaprabhudeva, Ghoḍācolī, Ṭiṇṭiṇī, Bhālukī and Nāgabodha and Khaṇḍakāpālika.
-\end{translation}
-
-\begin{testimonia}[hp01_008]
-Haṭharatnāvalī 1.83
-
-\begin{versinnote}
-\tl{allamaḥ prabhudevaś ca naiṭacūṭiś ca ṭiṇṭiṇiḥ |\\+}
-\tl{bhālukir nāgabodhaś ca khaṇḍakāpālikas tathā || \\!}
-\end{versinnote}
-
-Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
-
-\begin{versinnote}
-\tl{āllamaḥ prabhudevaś ca ghoḍācolī ca ṭhiṇṭhinī |\\+}
-\tl{bhālukir nāgadevaś ca khaṇḍī kāpālikas tathā ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_008]  
-The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literaure) is largely transmitted as allamaḥ prabhudevaḥ, as though it were two names, although some manuscripts of the \emph{Haṭhapradīpikā} have \emph{allamaprabhudevaś ca} (i.e., Tue,V3,V8,V13,V16,V22,Vu,N24,N26) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n.
-8).
-
-The names Nāgabodha, Nāgabodhi, Naradeva, Nāgadeva all seem possible in 1.8c.
-
-Many witnesses have khaṇḍa and kāpālika as separate names.
-However, khaṇḍakāpālika is well attested.
-Examples include \emph{Kathāsaritsāgara} 121.5 ff.
-(check), \emph{Bṛhatkathāmañjarī} 10.45 (check) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}).
-It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika.
-Alternatively, it could simply refer to one who used a broken skull as a bowl.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.9}
-\begin{translation}[hp01_009]
-Having destroyed the rod of death through the power of Haṭhayoga, these great perfected yogis and others wander in the world.
-\end{translation}
-
-\begin{testimonia}[hp01_009]
-Haṭharatnāvalī 1.84
-
-\begin{versinnote}
-\tl{ityādayo mahāsiddhāḥ haṭhayogaprasādataḥ |\\+}
-\tl{khaṇḍayitvā kāladaṇḍaṃ brahmāṇḍe vicaranti te ||\\!}
-\end{versinnote}
-
-Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
-
-\begin{versinnote}
-\tl{ityādayo mahāsiddhā rasabhogaprasādataḥ |\\+}
-\tl{khaṇḍayitvā kāladaṇḍaṃ trilokyāṃ vicaranti te |\\!}
-\end{versinnote}
-
-Haṭhatattvakaumudī 17.24
-
-\begin{versinnote}
-\tl{ūrdhvaṃretaḥprabhāvena sanakādyā maharṣayaḥ |\\+}
-\tl{khaṇḍayitvā kāladaṇḍaṃ yathecchaṃ viharanti te || 24 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_009]        
-The reference to \emph{brahmāṇḍa} (‘the world’) implies liberation-in-life (\emph{jīvanmukti}) and physical immortality.  
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.10}
-\begin{translation}[hp01_010]
-Haṭha is a hut of refuge for those who are burnt by the scorching torment of transmigration. Haṭha is the tortoise that supports the worlds of all yogas.
-\end{translation}
-
-\begin{testimonia}[hp01_010]
-Yogasārasaṃgraha (YSS), p.53.
-
-\begin{versinnote}
-\tl{saṃsāratāpataptānāṃ samāśrayahaṭho haṭhaḥ |\\+}
-\tl{aśeṣayogajagatām ādhārakamaṭho haṭhaḥ ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_010] 
-The compound \emph{saṃsāratāpa°} is well attested and found elsewhere (e.g., \emph{Viṣṇupurāṇa} 6.7.62, \emph{Agnipurāṇa} 371.1, \emph{Haṭhatattvakaumudī} 38.92, emph{Haṭhābhyāsapaddhati} ms.
-46/440, f.
-1v).
-The reading of \emph{samāśrayo} in V1 is metrically faulty.
-The word \emph{°jagatāṃ} is probably authorial because it makes good sense with  \emph{ādhārakamaṭha} in light of the cosmological notion that the tortoise supports all the worlds.
-However, this reading may not have been understood by some and was changed in the vulgate and other witnesses to \emph{°yuktānāṃ} instead.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.11}
-\begin{translation}[hp01_011]
-The doctrine of Haṭha should be kept completely secret by those yogis who are desiring success. When it is secret it becomes potent. However, when it has been revealed, it becomes impotent.
-\end{translation}
-
-\begin{sources}[hp01_011]
-Śivasaṃhitā
-
-\begin{versinnote}
-\tl{haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\+}
-\tl{bhaved vīryavatī guptā nirvīryā ca prakāśitā || 5.254\\+}
-\tl{(ab missing in some ŚS witnesses)\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_011]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{tathā haṭhapradīpikāyām—\\+}
-\tl{haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\+}
-\tl{bhaved vīryavatī guptā nirvīryā tu prakāśiteti ||\\!}
-\end{versinnote}
-
-BKhP 10v4
-\end{testimonia}
-
-\begin{philcomm}[hp01_011]        
-Either the singular or plural of yogin could be read here. The singular is well attested among the testimonia. However, the weight of the manuscript evidence favours the plural.  
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.12}
-\begin{translation}[hp01_012]
-In a well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live in an isolated hut.
-\end{translation}
-
-\begin{sources}[hp01_012]
-References to maṭhikā here.
-\end{sources}
-
-\begin{testimonia}[hp01_012]
-Haṭharatnāvalī 1.66
-
-\begin{versinnote}
-\tl{surāṣṭre dhārmike deśe subhikṣe nirupadrave |\\+}
-\tl{ekāntamaṭhikāmadhye sthātavyaṃ haṭhayoginā ||\\!}
-\end{versinnote}
-
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhapradīpikāyām—\\+}
-\tl{surājye dhārmike deśe subhikṣe nirupadrave |\\+}
-\tl{ekānte maṭhikāmadhye sthātavyaṃ haṭhayoginā ||\\!}
-\end{versinnote}
-
-BKhP 107v1
-\end{testimonia}
-
-\begin{philcomm}[hp01_012]        
-In the \emph{Jyotsnā} and printed editions of the \emph{Haṭhapradīpikā}, including one by Digambara and Kokaje (1970: 6), this verse has the additional hemistich, \emph{dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite}.
-This hemistich derives from the \emph{Gorakṣaśataka} (32cd), which has \emph{°paryante} instead of \emph{°paryantaṃ}.
-It stipulates that the hut should be built in a place measuring up to a bow length and free from rocks, fire and water.
-None of the early manuscripts have this hemistich which suggests that it was added at a later time.
-Nonetheless, it appears in over a dozen manuscripts that were consulted for this edition.
-These manuscripts are not close to an early hyparchetype of the text.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.13}
-\begin{translation}[hp01_013]
-It has a small door and is without cracks, holes and potsherds.
-It extends not too high or low, and is thickly smeared with cow dung in the proper way.
-It is clean, free from everything that annoys, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
-\end{translation}
-
-\begin{sources}[hp01_013]
-Cf. Dattātreyayogaśāstra
-
-\begin{versinnote}
-\tl{suśobhanaṃ maṭhaṃ kuryāt sūkṣmadvāraṃ tu nirvraṇaṃ || 54 ||\\+}
-\tl{suṣṭhu liptaṃ gomayena sudhayā vā prayatnataḥ |\\+}
-\tl{matkuṇair maśakair bhūtair varjitaṃ ca prayatnataḥ || 55 ||\\+}
-\tl{dine dine susammṛṣṭaṃ sammārjanyā hy atandritaḥ |\\+}
-\tl{vāsitaṃ ca sugandhena dhūpitaṃ guggulādibhiḥ || 56 ||\\+}
-\tl{malamūtrādibhir vargair aṣṭādaśabhir eva ca |\\+}
-\tl{varjitaṃ dvārasampannaṃ vastrāvaraṇam eva vā || 57 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_013]
-Suśruta 6.17.67:
-
-\begin{versinnote}
-gṛhe nirābādhe
-\end{versinnote}
-
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{alpadvāram arandhragartaghaṭitaṃ nāpy uccanīcāyitam |\\+}
-\tl{samyaggomayasāndraliptavimalaṃ niḥśeṣajantūjjhitam |\\+}
-\tl{bāhye maṇḍapakūpavediracitaṃ prākārasaṃveṣṭitam |\\+}
-\tl{proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 1.67
-
-\begin{versinnote}
-\tl{alpadvāram arandhragartapiṭharaṃ nātyuccanīcāyataṃ\\+}
-\tl{samyaggomayasāndraliptavimalaṃ niḥśeṣabādhojjhitaṃ |\\+}
-\tl{bāhye maṇḍapavedikūparuciraṃ prākārasaṃveṣṭitam\\+}
-\tl{proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||\\!}
-\end{versinnote}
-
-BKhP 107v3
-\end{testimonia}
-
-\begin{philcomm}[hp01_013]  
-The syntax of this verse is rather problematic.
-One would expect the features of the hut, which are listed in the first three pādas of the verse, to be in the nominative case, with the words \emph{idaṃ lakṣaṇaṃ} in the fourth pāda referring back to them.
-However, the compounds in the first three pādas appear to qualify \emph{lakṣaṇa} as though they were adjectives, and this appears to have been the way the verse was composed.
-
-The manuscripts preserve many different readings at the end of the compound beginning with \emph{arandhragarta°}.
-The reading \emph{°piṭharaṃ} has been suggested by Dominic Goodall, who understands it as potsherds (ask ref from Dominic?).
-The sense is that the hut should be free of rubbish, such as potsherds.
-One would expect a word for a defect in a hut that is similar to, but not the same as, cracks (\emph{randhra}) and holes (\emph{garta}).
-For this reason, the reading \emph{°vivaraṃ} looks like a patch, as its meaning does not add anything to \emph{°randhragarta°}.
-The reading \emph{viṭapaṃ} is well attested among the witnesses of the \emph{Haṭhapradīpikā} and \emph{Haṭharatnāvalī}.
-However, it is difficult to construe in this context unless it was intended to refer to creepers or branches that might invade or encroach upon the hut.
-
-Nearly all the early mss.
-have \emph{°bādhojjitaṃ}, whereas the Yogacintāmaṇi and V19 have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’).
-However, \emph{°bādhojjitaṃ} may have been original because it is so well attested and a hut is described similarly in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.14}
-\begin{translation}[hp01_014]
-Locating oneself in a hut of such a kind, free from all worry, [the yogi] should practise only yoga, in the way taught by his guru.
-\end{translation}
-
-\begin{sources}[hp01_014]
-Cf. Amanaska 2.15
-
-\begin{versinnote}
-\tl{evaṃvidhaṃ guruṃ labdhvā sarvacintāvivarjitaḥ\\+}
-\tl{sthitvā manohare deśe yogam eva samabhyaset\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_014]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\+}
-\tl{gurūpadiṣṭamārgeṇa yogam eva sadābhyased iti ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 1.68
-
-\begin{versinnote}
-\tl{evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\+}
-\tl{gurūpadiṣṭamārgeṇa yogam eva sadābhyaset ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-%\begin{philcomm}[hp01_014]        
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.15}
-\begin{translation}[hp01_015]
-Overeating, exertion, idle chatter, not sticking to rules, socialising, sensuality: through [these] six, yoga is lost.
-\end{translation}
-
-\begin{testimonia}[hp01_015]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\+}
-\tl{janasaṅgaś ca laulyaṃ ca ṣaḍbhir yogaḥ praṇaśyati ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 1.77
-
-\begin{versinnote}
-\tl{atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\+}
-\tl{janasaṅgaṃ ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||\\!}
-\end{versinnote}
-
-Yuktabhavadeva 4.25 (attributed to the śivayoga)
-
-\begin{versinnote}
-\tl{atyāhāraḥ prayāsaśca prajalpo niyamāgrahaḥ |\\+}
-\tl{janasaṃgaś ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||\\!}
-\end{versinnote}
-
-Jyotsnā
-
-\begin{versinnote}
-\tl{śītodakena prātaḥsnānanaktabhojanaphalāhārādirūpaniyamasya grahaṇaṃ niyamagrahaḥ |\\!}
-\end{versinnote}
-
-Yogaprakāśikā
-
-\begin{versinnote}
-\tl{niyamāgrahaḥ vakṣyamāṇaniyamāparipālanaṃ\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_015]        
-It is impossible to be certain about the meaning of \emph{niyamagraha}, as the manuscripts do not indicate whether an \emph{avagraha} (i.e., \emph{prajalpo ’niyamagraha}) was intended.
-Although \emph{yama} and \emph{niyama} are not included in the \emph{Haṭhapradīpikā} as auxiliaries of Haṭhayoga, verse 2.14 implies that \emph{niyama} is necessary at least in the early stages of establishing a practice.
-Furthermore, verse 3.82 suggests that a yogi who does not practice \emph{niyama} might obtain success in yoga through the practice of \textit{vajroli}.
-Ambiguity over the role of \emph{yama} and \emph{niyama} in Haṭhayoga appears to have prompted some to insert verses on ten \emph{yama}s and ten \emph{niyama}s after the next verse (1.16).
-The additional verses may derive from the \emph{Śāradātilakatantra} (25.7–8) or the  \emph{Vasiṣṭhasaṃhitā} (1.38, 1.53).
-In the \emph{Jyotsnā}, Brahmānanda reads \emph{niyamāgraha} and takes it as though \emph{āgraha} was implied, which yields the meaning of ‘over-insistence on rules’, and he relates it to extreme ascetic practice.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.16}
-\begin{translation}[hp01_016]
-Yoga is successful as a result of six [qualities]: zeal, daring, resolve, gnosis of the truth, conviction, and avoiding contact with people.
-\end{translation}
-
-\begin{sources}[hp01_016]
-Dharmaputrikā 137cd–38ab%Narharināth ed.?
-
-\begin{versinnote}
-\tl{utsāho niścayo dhairyaṃ santoṣas tattvadarśanam |\\+}
-\tl{kratūnāṃ copasaṃhāraḥ ṣaṭsādhanam iti smṛtam |\\!}
-\end{versinnote}
-
-Śivadharmottara 10 (W 122r):%JM: what ed or ms is this?
-
-\begin{versinnote}
-\tl{utsāhān niścayād dhairyāt santoṣāt tattvadarśanāt |\\+}
-\tl{muner janapadatyāgād ṣaḍbhir yogaḥ prasiddhyati |\\!}
-\end{versinnote}
-
-Jñānārṇava 20.1
-
-\begin{versinnote}
-\tl{utsāhān niścayād dhairyāt saṃtoṣāt tattvaniścayāt |\\+}
-\tl{muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 1 ||\\!}
-\end{versinnote}
-
-Yogabindu by Haribhadra
-
-\begin{versinnote}
-\tl{utsāhān niścayād dhairyāt saṃtopāt tattvadarśanāt |\\+}
-\tl{muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 411 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_016]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{utsāhāt sāhasād dhairyāt tatvajñānād viniścayāt |\\+}
-\tl{janasaṅgaparityāgāt ṣaḍbhir yogaḥ prasidhyati ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 1.78:
-
-\begin{versinnote}
-\tl{utsāhān niścayād dhairyāt tattvajñānārthadarśanāt
-[utsāhān niścalād- P,T]\\+}
-\tl{bindusthairyān mitāhārāj janasaṅgavivarjanāt |\\+}
-\tl{nidrātyāgāj jitaśvāsāt pīṭhasthairyād anālasāt\\+}
-\tl{gurvācāryaprasādāc ca ebhir yogas tu sidhyati || \\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_016]    
-For the second \emph{pāda}, the north-Indian manuscripts have something along the lines of \emph{tattvajñānāc ca darśanāt}.
-The problem with this reading is the meaning of \emph{darśana} by itself (i.e., discernment of what?).
-The early sources of this verse, in particular the \emph{Śivadharmottara}, indicate that the second pāda read as \emph{santoṣāt tattvadarśanāt}, which makes much better sense of the word \emph{darśana} (i.e., ‘seeing the truth’).
-It seems that this reading may have been adopted by Svātmārāma because the word \emph{santoṣāt} is present in the south-Indian manuscripts.
-At some point in the transmission, \emph{tattvadarśanāt} became \emph{tattvajñānāt} and \emph{santoṣāt} was dropped.
-The word \emph{darśanāt} was changed to \emph{viniścayāt} in V19 and the \emph{Yogacintāmaṇi}, which indicates that we are not alone in questioning its meaning.
-The problem with \emph{viniścayāt} in \emph{pāda} two is that \emph{niścayāt} is widely attested in \emph{pāda} one.
-It should also be noted that the word \emph{tattva} could have a more specific meaning in the \emph{Haṭhapradīpikā} (4.3) as Svātmārāma states that it is a synonym of \emph{samādhi}.
-In other yoga texts, it can sometimes refer to the practices of yoga (e.g., \emph{tritattva} in \emph{Amṛtasiddhi} 13.12, 14.2--3).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.17}
-\begin{translation}[hp01_017]
-Because it is the first auxiliary of haṭha, \emph{āsana} is taught first. This type of (\emph{tad}) \emph{āsana} brings about steadiness, good health and physical fitness.
-\end{translation}
-
-\begin{testimonia}[hp01_017]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhasya prathamāṅgatvād āsanaṃ pūrvam ucyate |\\+}
-\tl{tat kuryād āsanasthairyam ārogyaṃ cāṅgalāghavaṃ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.5
-
-\begin{versinnote}
-\tl{haṭhasya prathamāṅgatvād āsanaṃ dārśyate mayā |\\+}
-\tl{tat kuryād āsanaṃ sthairyam ārogyaṃ cāṅgapāṭavam || \\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_017]        
-The reading of \emph{aṅgapāṭavam} is attested among most of the early manuscripts.
-Although this compound rarely appears in other yoga texts, a similar term \emph{śarīrapāṭava} occurs in the \emph{Śivasaṃhitā} (2.35) as one of the benefits bestowed by digestive fire (\emph{vaiśvānarāgni}), which indicates that the word \emph{pāṭava} was used in relation to the body and the benefits of yoga.
-The compound \emph{aṅgapāṭava} seems to imply the optimal functioning of the body.
-However, the alternative reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts, even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}).
-Therefore, it is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by group two (i.e., N23, V19, etc.)
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.18}
-\begin{translation}[hp01_018]
-I shall now teach some of the postures which have been accepted by both sages (\emph{muni}), such as Vasiṣṭha, and yogis, such as Matsyendra.
-\end{translation}
-
-\begin{testimonia}[hp01_018]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhapradīpikāyām—\\+}
-\tl{vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ |\\+}
-\tl{aṅgīkṛtāny āsanāni vakṣyante kānicin mayā ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.6
-
-\begin{versinnote}
-\tl{vasiṣṭhādyaiś ca munibhir matsyendrādyaiś ca yogibhiḥ ||\\+}
-\tl{aṃgīkṛtāny āsanāni lakṣyante kāni cin mayā ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_018]        
-On the historical implications of these two traditions of postural practice in early Haṭhayoga, see Mallinson 2016 (119–122) and Birch 2018 (45–46).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.19}
-\begin{translation}[hp01_019]
-Correctly placing the soles of both feet between the knees and thighs [and] sitting up with the body straight: they call that \emph{svastikāsana}.
-\end{translation}
-
-\begin{sources}[hp01_019]
-Śāradātilaka 25.12
-
-\begin{versinnote}
-\tl{jānūrvor antare  samyak  kṛtvā  pādatale ubhe |\\+}
-\tl{ṛjukāyo viśed yogī svastikaṃ tat pracakṣate ||\\!}
-\end{versinnote}
-
-Vasiṣṭhasaṃhitā 1.68
-
-\begin{versinnote}
-\tl{jānūrvor antaraṃ samyak kṛtvā pādatale ubhe |\\+}
-\tl{ṛjukāyas tathāsīnaḥ svastikaṃ tat pracakṣate ||\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.3
-
-\begin{versinnote}
-\tl{jānūrvor antare samyak kṛtvā pādatale ubhe\\+}
-\tl{ṛjukāyaḥ sukhāsīnaḥ svastikaṃ tat pracakṣate\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_019]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{Yājñavalkyaḥ—\\+}
-\tl{jānūrvor antare samyak kṛtvā pādatale ubhe |\\+}
-\tl{ṛjukāyaḥ samāsīnaḥ svastikaṃ tat pracakṣate ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.52
-
-\begin{versinnote}
-\tl{atha svastikāsanam---\\+}
-\tl{jānūrvor antaraṃ samyak kṛtvā padatale ubhe ||\\+}
-\tl{ṛjukāyasamāsīnaḥ svastikaṃ tat pracakṣate || \\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_019]        
-One might wonder how the soles of the feet could be placed between the knees and thighs.
-Brahmānanda explains that the region of the shank near the knee should be understood by the word ‘knee’ in this verse (\emph{atra jānuśabdena jānusaṃnihito jaṅghāpradeśo grāhyaḥ jānusaṃnihito jaṅghāpradeśaḥ}).
-This is consistent with the earliest known description of \emph{svastikāsana} in the \emph{Pātañjalayogaśāstravivaraṇa} (2.46), which states that the big toes of one foot are tucked in between the shank and thigh of the other so it is not seen (\emph{dakṣiṇaṃ pādāṅguṣṭhaṃ savyenorujaṅghena parigṛhyādṛśyaṃ kṛtvā tathā savyaṃ pādāṅguṣṭhuṃ dakṣiṇenorujaṅkgenādṛśyaṃ parihṛhya yathā ca pārṣṇibhyāṃ vṛṣaṇayor apīḍanaṃ tathā yenāste tat svastikam āsanam}).
-For a discussion of \emph{svastikāsana} in the Pātañjalayoga tradition, see Maas 2018: 68–69.
-The descriptions of \emph{svastikāsana} in early Śaiva Tantras do not mention the inserting of the toes between the knees and thighs (see Goodall 2004: 348–350, fn.
-371).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.20}
-\begin{translation}[hp01_020]
-[The yogi] should place his right heel on the left side of the [lower] back, and the left [heel] on the right [side], in the same way. This is \emph{gomukhāsana}, which [looks] like a cow's face.
-\end{translation}
-
-\begin{sources}[hp01_020]
-Cf. Ahirbudhnyasaṃhitā 31.45cd–46
-
-\begin{versinnote}
-\tl{ubhayor gulphayoḥ kṛtvā pṛṣṭhapārśvāv ubhāv api ||\\+}
-\tl{vyutkrameṇātha pāṇibhyāṃ vinyastābhyāṃ vigṛhya ca |\\+}
-\tl{prṣṭhagābhyāṃ padāṅguṣṭhāv etad gomukham ucyate || \\!}
-\end{versinnote}
-
-Vasiṣṭhasaṃhitā 1.70
-
-\begin{versinnote}
-\tl{savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet |\\+}
-\tl{dakṣiṇe 'pi tathā savyaṃ gomukhaṃ tat pracakṣate ||\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.5cd–3.6ab
-
-\begin{versinnote}
-\tl{savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niveśayet\\+}
-\tl{dakṣiṇe 'pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_020]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{savye dakṣiṇagulphaṃ tu pṛṣthapārśve niveśayet |\\+}
-\tl{dakṣiṇe'pi tathā savyaṃ gomukhaṃ gomukhaṃ yathā ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.53
-
-\begin{versinnote}
-\tl{atha gomukhāsanam---\\+}
-\tl{savye dakṣiṇagulphaṃ tu pṛṣṭhapārśve niyojayet ||\\+}
-\tl{dakṣiṇe 'pi tathā savyaṃ gomukhaṃ gomukhāsanam ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_020]        
-This posture first appears in some Vaiṣṇava \emph{Saṃhitā}s that predate the \emph{Haṭhapradīpikā}, namely, the \emph{Ahirbudhnyasaṃhitā} and the \emph{Vasiṣṭhasaṃhitā}.
-The position of the ankles is the same in all the source texts.
-However, an interesting feature of the description in the \emph{Ahirbudhnyasaṃhitā} is the position of the hands, which are crossed behind the back and hold the big toes.
-The \emph{Haṭhapradīpikā}’s description, which derives from the \emph{Vasiṣṭhasaṃhitā}, does not mention the position of the hands.
-For illustrations of six possible positions of the arms and hands, see Gharote, Jha, Devnath, Sakhalkar 2006: 111–113.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.21}
-\begin{translation}[hp01_021]
-By fixing one foot on one thigh and placing the [other] thigh on the other foot, vīrāsana is taught [to arise].
-\end{translation}
-
-\begin{sources}[hp01_021]
-Vasiṣṭhasaṃhitā 1.72
-
-\begin{versinnote}
-\tl{ekaṃ pādam athaikasmin vinyasyorau ca saṃsthitam |\\+}
-\tl{itarasmiṃs tathaivoruṃ vīrāsanam itīritam ||\\!}
-\end{versinnote}
-
-Cf. Śāradātilakatantra 25.15cd–16ab
-
-\begin{versinnote}
-\tl{ekaṃ pādam adhaḥ kṛtvā vinyasyorau tathetaram ||\\+}
-\tl{ṛjukāyo viśed yogī vīrāsanam itīritam |\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.8
-
-\begin{versinnote}
-\tl{ekaṃ pādam athaikasmin vinyasyoruṇi saṃsthitam |\\+}
-\tl{itarasmiṃs tathā coruṃ vīrāsanam udāhṛtam ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_021]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{ekaṃ pādam athaikasmin vinyasyoruṇi saṃsthitaḥ |\\+}
-\tl{itarasmiṃs tathā coruṃ vīrāsanam udāhṛtam ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.54
-
-\begin{versinnote}
-\tl{atha vīrāsanam---\\+}
-\tl{ekaṃ pādam athaikasmin vinyased ūruṇi sthiram ||\\+}
-\tl{itarasmiṃs tathā coruṃ vīrāsanam īritam || \\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_021]   
-Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by V1, the sources and the testimonia.
-It appears to be verse filler here rather than indicating a temporal sequence of actions.
-Svātmārāma borrowed the verse on \emph{vīrāsana} from the \emph{Vasiṣṭhasaṃhitā}, the redactor of which appears to have adapted its first hemistich from a description of this posture in the \emph{Śāradātilakatantra}.
-This would explain the rather strange syntax of the \emph{Vasiṣṭhasaṃhitā}’s version, in which \emph{adhaḥ kṛtvā} was changed to \emph{athaikasmin}, and \emph{tathetaram} became \emph{ca saṃsthitam}.
-It seems that \emph{saṃsthitaṃ} must be understood with \emph{ūruṃ} in the third \emph{pāda} in the sense of \emph{saṃsthāpya} (i.e., ‘having placed’).
-
-Different versions of \emph{vīrāsana} are found in earlier Tantras, such as the Kiraṇatantra (58.9), Hemacandra’s \emph{Yogaśāstra} and commentaries on the \emph{Pātañjalayogaśāstra}.
-For a discussion of some of these sources, see Maas 2018: 66–68.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.22}
-\begin{translation}[hp01_022]
-Knowers of yoga know that \emph{kūrmāsana} arises by attentively blocking the anus with turned-out ankles.
-\end{translation}
-
-\begin{sources}[hp01_022]
-Vasiṣṭhasaṃhitā 1.80
-
-\begin{versinnote}
-\tl{gudaṃ nirudhya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\+}
-\tl{kūrmāsanaṃ bhaved etad iti yogavido viduḥ ||\\!}
-\end{versinnote}
-
-Cf. Ahirbudhnyasaṃhitā 31.35
-
-\begin{versinnote}
-\tl{gudaṃ nipīḍya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\+}
-\tl{etat kūrmāsanaṃ proktaṃ yogasiddhikaraṃ param || \\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_022]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{gudaṃ niyamya gulphābhyaṃ vyutkrameṇa samāhitaḥ |\\+}
-\tl{kūrmāsanaṃ bhaved etad iti yogavido viduḥ ||\\!}
-\end{versinnote}
-
-Yuktabhavadeva 6.15
-
-\begin{versinnote}
-\tl{haṭhapradīpikāyām\\+}
-\tl{gudaṃ niyamya gulphābhyāṃ vyutkrameṇa samāhitaḥ |\\+}
-\tl{kūrmāsanaṃ bhavedetaditi yogavido viduḥ ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_022]   
-In the first \emph{pāda}, the witnesses are split between \emph{nirudhya} (‘having blocked’), \emph{nibadhya} (‘having bound’), \emph{niyamya} (‘having restrained’) and \emph{niṣpīḍya} (‘having pressed’).
-The source, the \emph{Vasiṣṭhasaṃhitā}, supports \emph{nirudhya} whereas the testimonia supports \emph{niyamya}.
-In terms of closing the anus by sitting on the ankles, \emph{nirudhya} makes better sense and has been adopted because it is supported by N23 (an important witness of group 2) and the manuscripts reported in the critical edition of the \emph{Vasiṣṭhasaṃhitā}.
-The word \emph{vyutkrameṇa} appears to describe the position of the ankles.
-Its basic meaning is ‘against the normal direction’, which would suggest that the ankles are turned out or crossed rather than placed together naturally.
-If the yogi is in a kneeling-type position, turning the feet out would bring the ankles together, blocking the perineal area.
-See \emph{Yoga Mīmāṃsā}, vol 8, no.
-2, pp.
-29–30 for a discussion of \emph{vyutkramena} and the position of the ankles in \emph{kūrmāsana}, and vol 8, no.2, Figures 3–6 for photographs of a practitioner performing this \emph{āsana}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.23}
-\begin{translation}[hp01_023]
-[The yogi] adeptly assumes \emph{padmāsana}, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is \emph{kurkuṭāsana}.
-\end{translation}
-
-\begin{sources}[hp01_023]
-Vasiṣṭhasaṃhitā 1.78
-
-\begin{versinnote}
-\tl{padmāsanaṃ samāsthāya jānūrvor antare karau |\\+}
-\tl{bhūmau niveśya saṃsthāpya vyomasthaṃ kukkuṭāsanam ||\\+}
-\tl{[niveśya bhūmau – mss. la, va, śa]\\!}
-\end{versinnote}
-
-Cf. Ahirbudhnyasaṃhitā 31.38
-
-\begin{versinnote}
-\tl{kukkuṭāsanam\\+}
-\tl{padmāsanam adhiṣṭhāya jānvantaraviniḥsṛtau |\\+}
-\tl{karau bhūmau niveśyaitad vyomasthaṃ kukkuṭāsanam || \\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_023]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{padmāsanaṃ tu saṃyojya jānūrvor antare karau |\\+}
-\tl{niveśya bhūmau saṃsthāpya vyomasthaṃ kukkuṭāsanam ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.73
-
-\begin{versinnote}
-\tl{atha kukkuṭāsanam---\\+}
-\tl{padmāsanaṃ susaṃsthāpya jānūrvor antare karau ||\\+}
-\tl{niveśya bhūmau saṃsthāpya vyomasthaḥ kukkuṭāsanam ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_023]
-\emph{Kurkuṭa} and \emph{kurkkuṭa} in V1, J10ac, V3 are variant spellings attested in the \emph{Pañcatantra} (M-W).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.24}
-\begin{translation}[hp01_024]
-While maintaining kurkuṭāsana, [the yogi] binds the neck with the hands and lies like a tortoise on his back. This is uttānakūrmāsana.
-\end{translation}
-
-\begin{testimonia}[hp01_024]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{kukkuṭāsanabandhastho dorbhyāṃ saṃbadhya kandharam |\\+}
-\tl{bhavet kūrmavad uttānam etad uttānakūrmakam ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.74
-
-\begin{versinnote}
-\tl{kukkuṭāsanabandhastho dorbhyāṃ sambadhya kandharām ||\\+}
-\tl{śete kūrmavad uttānam etad uttānakūrmakam || 74 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_024]
-\emph{°bandhasthaḥ} or \emph{°vat kṛtvā}?
-Only V1 has the latter, which is simpler.
-Are the others trying to improve it?
-Stemmatically ambiguous as \emph{°bandha°} is on one branch (V3/J8, V19) and \emph{°madhya°} the other (J10, J17, N17).
-V1 looks like an outlier.
-
-Adopt \emph{bandhastho} with note. \emph{°bandhastha} not found in any other texts. While \emph{°vat kṛtvā} is possible, it is a singular reading that appears to be unrelated to the other variants.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.25}
-\begin{translation}[hp01_025]
-Clasping the big toes with hands and performing the action of drawing a bow as far as the ear is called dhanurāsana.
-\end{translation}
-
-\begin{testimonia}[hp01_025]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{pādāṅguṣṭhau ca pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\+}
-\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam |\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.51
-
-\begin{versinnote}
-\tl{atha dhanurāsanam---\\+}
-\tl{pādāṅguṣṭhau tu pāṇibhyāṃ gṛhītvā śravaṇāvadhi ||\\+}
-\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam ucyate ||\\!}
-\end{versinnote}
-
-Cf. Haṭhayogasaṃhitā
-
-\begin{versinnote}
-\tl{dhanurāsanam |\\+}
-\tl{prasārya pādau bhuvi daṇḍarūpau\\+}
-\tl{karau ca pṛṣṭhe dhṛtapādayugmau |\\+}
-\tl{kṛtvā dhanustulyavivarttitāṅgaṃ\\+}
-\tl{nigadyate vai dhanurāsanaṃ tat || 25 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_025]
-Comment on kṛtvā, which is the reading of V19, Yogacintāmaṇi and Haṭharatnāvalī.
-Godāvaramiśra has dhanurākarṣavat kṛṣṭaṃ dhanurāsanam ucyate, which is likely to be the original reading.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.26}
-\begin{translation}[hp01_026]
-Having grasped the right foot, which is placed at the base of the left thigh, [the yogi’s] left foot is wrapped around the outside of the knee and he remains with his body twisted.
-This āsana was taught by Matsyendranātha.
-\end{translation}
-
-\begin{testimonia}[hp01_026]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{vāmorumūlārpitadakṣapādaṃ
-jānvor bahir veṣṭitadakṣadoṣṇā |\\+}
-\tl{pragṛhya tiṣṭhet parivartitāṅgaḥ
-śrīmatsyanāthoditam āsanaṃ syāt ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{atha matsyendrāsanam---\\+}
-\tl{vāmorumūlārpitā dakṣapādo jānvor bahir veṣṭitadakṣadoṣṇā ||\\+}
-\tl{pragṛhya tiṣṭhet partivartitāṅgaḥ śrīmatsyanāthoditam āsanaṃ syāt || 3.57 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_026]
-Comment on vāmapadam vs dakṣadoṣṇā
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.27}
-\begin{translation}[hp01_027]
-Matsyendra's seat, is a destructive weapon for the many terrible diseases that develop in the stomach; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
-\end{translation}
-
-\begin{testimonia}[hp01_027]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{matsyendrapīṭhaṃ jaṭharapravṛddha-\\+}
-\tl{pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\+}
-\tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
-\tl{daṇḍe sthiratvaṃ pradadāti puṃsām ||
-\var{°pravṛddha ] N : °pravṛddhiṃ L}\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.58
-
-\begin{versinnote}
-\tl{matsyendrapīṭhaṃ jaṭharapradīptaṃ\\+}
-\tl{pracaṇḍarugmaṇḍalakhaṇḍanāstram ||\\+}
-\tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
-\tl{daṇḍasthiratvaṃ ca dadāti puṃsām ||\\!}
-\end{versinnote}
-
-HTK
-
-\begin{versinnote}
-\tl{matsyendrapīṭhaṃ jaṭharapracaṇḍa-\\+}
-\tl{ruṅmaṇḍalakhaṇḍanakhaṇḍanāstram |\\+}
-\tl{abhyāsataḥ kuṇḍalinīprabodhaṃ\\+}
-\tl{daṇḍasthiratvaṃ ca dadāti puṃsām || 8 ||\\!}
-\end{versinnote}
-
-YBD
-
-\begin{versinnote}
-\tl{matsyendrapīṭhaṃ jaṭharaprabuddhaṃ\\+}
-\tl{pracaṇḍaruṅmaṇḍalakhaṇḍanāstram |\\+}
-\tl{abhyasataṃ kuṇḍalinīprabodhaṃ\\+}
-\tl{daṇḍasthiratvaṃ ca dadāti puṃsām ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_027]
-Change note: we are adopting jaṭharapravṛddha (translation has been changed).  Cf. Yogacintāmaṇi ms. N. and HP mss.  J10, J17, V2, N8, N11.
-
-Pāda d, J10ac and J17 have \emph{candra} for \emph{daṇḍa}; However,  also in Yogacintāmaṇi and 6-chapter HP.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.28}
-\begin{translation}[hp01_028]
-[The yogi] should stretch out both feet on the ground like staffs, hold the ends of both feet with the hands, place the forehead upon the knees and remain thus.
-They call this the back-stretch (\emph{paścimatānam}).
-\end{translation}
-
-\begin{sources}[hp01_028]
-Cf. Śivasaṃhitā
-
-\begin{versinnote}
-\tl{prasārya caraṇadvandvaṃ parasparasusaṃyutam |\\+}
-\tl{svapāṇibhyāṃ dṛḍhaṃ dhṛtvā jānūpari śiro nyaset || 3.108 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_028]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{prasārya pādau bhuvi daṇḍarūpau\\+}
-\tl{dvābhyāṃ ca pādadvitayaṃ gṛhītvā |\\+}
-\tl{jānūpari nyastalalāṭadeśo\\+}
-\tl{'bhyased idaṃ paścimatānam āhuḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{atha paścimatānāsanam---\\+}
-\tl{prasārya pādau bhuvi daṇḍarūpau\\+}
-\tl{dorbhyāṃ padāgradvitayaṃ gṛhītvā ||\\+}
-\tl{jānūpari nyastalalāṭadeśo\\+}
-\tl{vased idaṃ paścimatānam āhuḥ || 3.66 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_028]
-Only V1 has \emph{dorbhyāṃ padāgra}, others have variations on the much inferior \emph{dvābhyāṃ karābhyāṃ}.
-The karābhyaṃ reading was probably introduced to remove dorbhyāṃ, which usually means arms.The reading \emph{tāṇabandhaḥ} in V1 doesn’t work with \emph{idaṃ}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.29}
-\begin{translation}[hp01_029]
-This back-stretch is the foremost among āsanas. It makes the breath flow in the back [i.e. Central channel], increases the digestive fire, makes the belly thin and prevents diseases in men.
-\end{translation}
-
-\begin{sources}[hp01_029]
-Cf. Śivasaṃhitā
-
-\begin{versinnote}
-\tl{Āsanāgryam idaṃ proktaṃ jaṭharānaladīpanam |\\+}
-\tl{dehāvasādaharaṇaṃ paścimottānasaṃjñakam || 3.109 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_029]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{iti paścimatānam āsanāgryaṃ
-pavanaṃ paścimavāhinaṃ karoti |\\+}
-\tl{udayaṃ jaṭharānalasya kuryād
-udare kārśyam arogitāṃ ca puṃsām ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{iti paścimatānam āsanāgryaṃ pavanaṃ paścimavāhinaṃ karoti ||\\+}
-\tl{udayaṃ jaṭharānalasya kuryād udare kārśyam arogatāṃ ca puṃsām || 3.67 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_029]
-Adopt \emph{arogitāṃ} (which is well attested), despite the fact that \emph{arogatāṃ} is much more common.
-The use of the word \emph{paścima} to mean the central channel is found in \emph{Yogabīja} 121.
-The  \emph{Yogabīja} (95) also refers to the path of the central channel (\emph{paścimamārga}), and this understanding of \emph{paścima} is found in the \emph{Jyotsnā 1.29}: \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumnāmārgeṇa vahatīti paścimavāhī}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.30}
-\begin{translation}[hp01_030]
-Supporting oneself on the ground with both hands, their elbows placed on either side of the navel, with a raised position (?
-uccāsanaḥ) one is placed up into the air [as straight] as a stick.
-They call this posture the peacock.
-\end{translation}
-
-\begin{sources}[hp01_030]
-Vimānārcanākalpa 96
-
-\begin{versinnote}
-\tl{karatale bhūmau saṃsthāpya kūrparau nābhipārśvayor nyasya nataśirāḥ (unnataśirāḥ) pādau ḍaṇḍavad vyomni saṃsthito mayūrāsanam iti |\\!}
-\end{versinnote}
-
-Pādmasaṃhitā yogapāda 1.21c–22d:
-
-\begin{versinnote}
-\tl{avaṣṭabhya dharāṃ samyak talābhyāṃ hastayor dvayoḥ ||\\+}
-\tl{kūrparau nābhipārśve ca sthāpayitvā mayūravat |\\+}
-\tl{samunnamya śiraḥpādau mayūrāsanam iṣyate ||\\!}
-\end{versinnote}
-
-Ahirbudhnyasaṃhitā 31.36–37
-
-\begin{versinnote}
-\tl{mayūrāsanam\\+}
-\tl{niveśya kūrparau samyaṅ nābhimaṇḍalapārśvayoḥ |\\+}
-\tl{avaṣṭabhya bhuvaṃ pāṇitalābhyāṃ vyomni daṇḍavat || 31--36 ||\\!}
-\end{versinnote}
-
-Vasiṣṭhasaṃhitā 1.76–77
-
-\begin{versinnote}
-\tl{avaṣṭabhya dharāṃ samyak talābhyāṃ ca karadvayam |\\+}
-\tl{hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ ||\\+}
-\tl{samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\+}
-\tl{mayūrāsanam etad dhi sarvapāpavināśanam ||\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.15–16
-
-\begin{versinnote}
-\tl{avaṣṭabhya dharāṃ samyak talābhyāṃ tu karadvayoḥ\\+}
-\tl{hastayoḥ kūrparau cāpi sthāpayan nābhipārśvayoḥ\\+}
-\tl{samunnataśiraḥpādo daṇḍavad vyomni saṃsthitaḥ |\\+}
-\tl{mayūrāsanam etat tu sarvapāpapraṇāśanam ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_030]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{dharām avaṣṭabhya punaḥ karābhyāṃ
-tat kūrpare sthāpitanābhipārśvaḥ |\\+}
-\tl{tadāsane daṇḍavad utthitaḥ khe
-mayūram etat pravadanti santaḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.42
-
-\begin{versinnote}
-\tl{atha mayūram\\+}
-\tl{dharām avaṣṭabhya karadvayena tatkūrpare sthāpitanābhipārśvaḥ ||\\+}
-\tl{uccāsano daṇḍavad utthitaḥ khe mayūram etat pravadanti pīṭham ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_030]
-The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77).
-The compound \emph{uccāsano} seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}, which may be derived from earlier Vaiṣṇava sources, such as the \textit{Pādmasaṃhitā}.
-
-1.30b tat refers to \emph{karadvaya} (cf. Vasiṣṭhasaṃhitā).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.31}
-\begin{translation}[hp01_031]
-The glorious mayūra posture gets rid of all diseases of the abdomen such as bloating and overcomes humoral imbalances.
-It completely incinerates food which is bad or has been eaten to excess, it generates digestive fire and it digests strong poison.
-\end{translation}
-
-\begin{testimonia}[hp01_031]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{harati sakalarogān āśu gulmodarādīn\\+}
-\tl{abhibhavati ca doṣān āsanaṃ śrīmayūram |\\+}
-\tl{bahukadaśanabhuktaṃ bhasma kuryād aśeṣam\\+}
-\tl{janayati jaṭharāgniṃ jārayet kālakūṭam ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.43
-
-\begin{versinnote}
-\tl{harati sakalarogān āśu gulmodarādīn\\+}
-\tl{abhibhavati ca doṣān āsanaṃ śrīmayūram ||\\+}
-\tl{bahukadaśanabhuktaṃ bhasma kuryād vicitram\\+}
-\tl{janayati jaṭharāgniṃ jīryate kālakūṭam ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_031]
-\emph{aśeṣaṃ} better
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.32}
-\begin{translation}[hp01_032]
-Lying on one’s back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] any āsana and relaxes the mind.
-\end{translation}
-
-\begin{sources}[hp01_032]
-Cf. Dattātreyayogaśāstra 24cd
-
-\begin{versinnote}
-\tl{uttānaśavavad bhūmau śayanaṃ coktam uttamam ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_032]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{uttānaṃ śavavad bhūmau śavāsanam idaṃ smṛtam |\\+}
-\tl{śavāsanaṃ śrāntiharaṃ cittaviśrāntisādhanam ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.76
-
-\begin{versinnote}
-\tl{athāntimaṃ śavāsanam\\+}
-\tl{prasārya hastapādau ca viśrāntyā śayanaṃ tathā ||\\+}
-\tl{sarvāsanaśramaharaṃ śayitaṃ tu śavāsanam || 3.76 ||\\!}
-\end{versinnote}
-
-Cf. HTK
-
-\begin{versinnote}
-\tl{śavāsanaṃ hṛtkupitavātagranthivibhedakam |\\+}
-\tl{sarvāsanaśrāntijit hṛtśramaghnaṃ yogi saukhyadam || 12 ||\\!}
-\end{versinnote}
-
-YBD
-
-\begin{versinnote}
-\tl{uttānaṃ śavavad bhūmau śayanaṃ tu śavāsanam ||\\+}
-\tl{śavāsanaṃ śrāntiharaṃ cittaviśrāntikārakam ||\\+}
-\tl{iti śavāsanam || 21 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-%%%%%%%%%%
-\subsection*{1.33}
-\begin{translation}[hp01_033]
-Śiva taught eighty-four āsanas. I shall take the four best and teach them.
-\end{translation}
-
-\begin{sources}[hp01_033]
-Śivasaṃhitā 3.96
-
-\begin{versinnote}
-\tl{caturaśīty āsanāni santi nānāvidhāni ca |\\+}
-\tl{tebhyaś catuṣkam ādāya mayoktāni bravīmy aham ||\\!}
-\end{versinnote}
-
-Cf. Dattātreyayogaśāstra 5
-
-\begin{versinnote}
-\tl{caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\+}
-\tl{ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||\\!}
-\end{versinnote}
-
-Cf. Vivekamārtaṇḍa 5
-
-\begin{versinnote}
-\tl{caturāśītilakṣānām ekaikaṃ samudāhṛtaṃ |\\+}
-\tl{ataḥ śivena pīṭhānāṃ ṣoḍaśonaṃ śataṃ kṛtaṃ ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_033]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhapradīpikāyām—\\+}
-\tl{caturaśīty āsanāni śivena kathitāni vai |\\+}
-\tl{tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{caturaśīty āsanāni śivena kathitāni tu |\\+}
-\tl{tebhyaś catuṣkam ādāya sārabhūtaṃ bravīmy aham || 3.23 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_033]
-In Pādas ac Svātmārāma may have rewritten Śivasaṃhitā 3.96 to include the information that it was Śiva (\emph{śivena}) who taught the 84 āsanas, whereas in the source Śiva is himself speaking.
-With this
-Svātmārāma confuses the verse, since one would have to understand Pādas cd to be direct speech by Śiva in order to understand \emph{bravīmi} correctly (but cf.
-4.2 which is a verse from the Gorakṣaśataka which includes pravakṣyāmi).
-
-The Śivasaṃhitā follows 3.96 with teachings on siddha, padma, paścimottāna and svastika āsanas, but the HP teaches siddha, padma, siṃha and bhadra.
-
-The word \emph{tu} is often used to introduce a new posture, but in this verse seems to be a verse filler.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.34}
-\begin{translation}[hp01_034]
-The adept, lotus, lion and auspicious pose are the best tetrad and, among those, always sit in the adept’s pose, my dear.
-\end{translation}
-
-\begin{sources}[hp01_034]
---
-\end{sources}
-
-\begin{testimonia}[hp01_034]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\+}
-\tl{śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.24
-
-\begin{versinnote}
-\tl{siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\+}
-\tl{śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_034]
-It is likely that the original version of this verse contained the vocative with the imperative form of the verb.
-There is another instance where Svātmārāma included a verse with the vocative (4.58?) or the subject in the first person (4.2), as though the text were a dialogue.
-In this case, it seems that efforts have been made to write out the vocative and imperative verb.
-However, the \emph{sukhe} and \emph{sukham} is difficult to construe in this context, because the context suggests that the intended meaning was that one should always sit in Siddhāsana (as opposed to the other three), rather than the prescription to always sit in a comfortable Siddhāsana.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.35}
-\begin{translation}[hp01_035]
-Having joined the place of the perineum with the heel of the foot, the yogi should firmly fix the [other] foot on the penis.
-Having held the face and chest together and the body erect, [the yogi] remains still, his senses restrained, gazing between the brows with his eyes unmoving.
-This breaks open the door to liberation and is called the adept’s pose.
-\end{translation}
-% [MD] add translation of "pūrvoktam eva matsaṃmataṃ"
-
-\begin{sources}[hp01_035]
-Vivekamārtaṇḍa 7
-
-\begin{versinnote}
-\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyase[n]\\+}
-\tl{meḍhre pādam athaikam āsyahṛdaye dhṛtvā samaṃ vigraham |\\+}
-\tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
-\tl{etan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_035]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{pavanayogasaṃgrahe—\\+}
-\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
-\tl{meḍhre pādam athaikam ekahṛdaye kṛtvā samaṃ vigraham |\\+}
-\tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyed bhruvor antaraṃ tv\\+}
-\tl{etan mokṣakapāṭabhedanakaraṃ siddhāsanaṃ procyate ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.25
-
-\begin{versinnote}
-\tl{tatra siddhāsanam\\+}
-\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
-\tl{meḍhre pādam athaikam eva niyataṃ kṛtvā samaṃ vigraham |\\+}
-\tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
-\tl{caitan mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_035]
-KDh ed has \emph{āsya}, variant in mss kh and gha, both BORI.
-Jürgen suggests we might take \emph{āsyahṛdaye} and \emph{vigraham} as the objects of \emph{dhṛtvā} and samam as an adverb (i.e., holding straight the face, chest and body).
-However, in a different doctrine.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.36}
-\begin{translation}[hp01_036]
-Having fixed the left heel on the penis, and put the other heel on that, this is siddhāsana.
-\end{translation}
-
-\begin{sources}[hp01_036]
-Vasiṣṭhasaṃhitā 1.81
-
-\begin{versinnote}
-\tl{meḍhrād upari nikṣipya gulphaṃ tathopari |\\+}
-\tl{gulphāntaraṃ vinikṣipya muktāsanam idaṃ smṛtam ||\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.15
-
-\begin{versinnote}
-\tl{meḍhrād upari nikṣipya savyaṃ gulphaṃ tathopari |\\+}
-\tl{gulphāntaraṃ ca nikṣipya muktāsanam idaṃ tu vā ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_036]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{tathā |\\+}
-\tl{meḍhrād upari vinyasya savyaṃ gulphaṃ tathopari |\\+}
-\tl{gulphāntaraṃ tu vinyasya siddhāsanam idaṃ bhavet ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.26
-
-\begin{versinnote}
-\tl{matāntare tu\\+}
-\tl{meḍhrād upari niḥkṣipya savyaṃ gulphaṃ tathopari |\\+}
-\tl{gulphāntaraṃ ca niḥkṣipya siddhāḥ siddhāsanaṃ viduḥ ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-%%%%%%%%%%
-\subsection*{1.37}
-\begin{translation}[hp01_037]
-Some proclaim this is siddhāsana, others know it as vajrāsana, a few say it is muktāsana and some guptāsana.
-\end{translation}
-
-\begin{sources}[hp01_037]
---
-\end{sources}
-
-\begin{testimonia}[hp01_037]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{siddhaṃ padmaṃ tathā bhadraṃ siṃhaṃ ceti catuṣṭayam |\\+}
-\tl{śreṣṭhaṃ tatrāpi vai padmaṃ tiṣṭhet siddhāsane sadā ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{siddhaṃ padmaṃ tathā siṃhaṃ bhadraṃ ceti catuṣṭayam |\\+}
-\tl{śreṣṭhaṃ tatrāpi ca tathā tiṣṭhet siddhāsane sadā || 3.24 ||\\!}
-\end{versinnote}
-
-Cf. Śivayogasāram by Kolani Ganapatideva (date 14th c.)
-
-\begin{versinnote}
-\tl{siddāsanambunu, gondaru vajrāsanambaniyu |\\+}
-\tl{gondaru muktāsanambaniyu, gondadu  gulbāsanam ||\\!}
-\end{versinnote}
-
-Cf. A verse by the poet Pingali Surana (active 16th c.)
-
-\begin{versinnote}
-\tl{kondaru siddāsanamani\\+}
-\tl{kondaru vajrāsanamani koniyādudurī\\+}
-\tl{pondaga dīnini mariyoka\\+}
-\tl{kondaru guptāsamanu kondru mahātmā\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-%%%%%%%%%%
-\subsection*{1.38}
-\begin{translation}[hp01_038]
-The Siddhas know Siddhasana as the single most important amongst
-all postures, in the same way as measured diet amongst rules and non-violence amongst observances.
-\end{translation}
-
-\begin{sources}[hp01_038]
-Cf. Dattātreyayogaśāstra 33
-
-\begin{versinnote}
-\tl{laghvāhāras tu teṣv eko mukhyo bhavati nāpare |\\+}
-\tl{ahiṃsā niyameṣv eko mukhyo bhavati nāpare || 33 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_038]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{niyameṣu mitāharo yathāhịmsā yameṣv iva |\\+}
-\tl{mukhyaṃ sarvāsaneṣv evaṃ siddhāsanaṃ idaṃ viduḥ |\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_038]
-\emph{iva} or \emph{eva}?
-\emph{iva} does work — like \emph{siddhāsana}, \emph{mitāhāra} and \emph{ahiṃsā} are the best, but for it to work properly \emph{mitāhāra} and \emph{ahiṃsā} should be accusative.
-V19 has acc + \emph{iva}, which seems best, especially with \emph{siddhāḥ viduḥ}, but this might be a correction as V19 often corrects.
-However, one old KDham BORI (?) ms (pha, 1695 CE) has it, as does Jyotsnā, so adopt.
-
-In pāda d V19 has \emph{siddhāsanam idaṃ viduḥ}, but the reading of all other mss is preferable.
-
-Clearly based on DYŚ 33, which includes \emph{ekaṃ} and \emph{mukhya}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.39}
-\begin{translation}[hp01_039]
-Among the eighty-four postures, one should regularly practise just Siddha; in the same way one should practise Suṣumnā among the 72,000 channels.
-\end{translation}
-
-\begin{sources}[hp01_039]
-\end{sources}
-
-\begin{testimonia}[hp01_039]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{caturaśītipīṭheṣu siddhāsanaṃ samabhyaset |\\+}
-\tl{dvāsaptatisahasreṣu suṣumṇām iva nāḍiṣu ||\\!}
-\end{versinnote}
-
-YSS
-
-\begin{versinnote}
-\tl{caturāśītapīṭheṣu siddham eva samabhyaset |\\+}
-\tl{dvisaptatisahasreṣu suṣumnām iva nāḍiṣu ||\\!}
-\end{versinnote}
-
-Yogacintāmaṇi (in a different passage)
-​​
-\begin{versinnote}
-\tl{maṇḍalā dṛśyate siddhiḥ kuṇḍalyabhyāsayoginaḥ |\\+}
-\tl{dvisaptatisahasrāṇāṃ nāḍīnāṃ malaśodhanam ||\\!}
-\end{versinnote}
-
-Cf. Kumbhakapaddhati (effects of practising kumbhaka)
-
-\begin{versinnote}
-\tl{dvāsaptati sahāsrāṇāṃ nāḍīnāṃ malaśodhanam |\\+}
-\tl{yatheṣṭaṃ dhāraṇaṃ vāyor vikārābhāva eva ca || 120 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_039]
-Odd to have \emph{suṣumnām} as object of \emph{abhyaset}.
-Yogacintāmaṇi has this reading though.
-Perhaps cd were added somewhat indiscriminately by Svātmārāma (with nominative \emph{suṣumnā}) and then others tried to make sense of it.
-
-Some witnesses, including Jyotsnā, have \emph{nāḍīnāṃ malaśodhanam/e} for pāda d, which is probably a patch (no other texts say siddhāsana clears the channels), but cf.
-Amṛtasiddhi in which the practices are said to bring about cālana of the nāḍīs (e.g.
-11.6).
-
-Good example of early contamination.
-
-[\emph{nāḍiṣu} is better supported (J10ac,V19,J17).]
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.40}
-\begin{translation}[hp01_040]
-By meditating upon the self, restricting the diet and regularly practising Siddhāsana for twelve years, the yogi attains the Niṣpatti stage.
-What’s the point of lots of exhausting postures when there is Siddhāsana?
-\end{translation}
-
-\begin{sources}[hp01_040]
---
-\end{sources}
-
-\begin{testimonia}[hp01_040]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{ātmadhyāyī mitāhārī yāvad dvādaśavatsaram |\\+}
-\tl{sadā siddhāsanābhyāsād yogī niṣpattim āpnuyāt |\\+}
-\tl{śramadair bahubhiḥ pīṭhaiḥ kiṃ syāt siddhāsane sati ||\\!}
-\end{versinnote}
-
-YSS
-
-\begin{versinnote}
-\tl{ātmadhyāyo mitāhārī yāvad dvādaśavatsaram |\\+}
-\tl{sadā siddhāsanābhyāsād yoganiṣpattim āpnuyāt ||\\+}
-\tl{śramadair bahubhiḥ pīṭhair alaṃ siddhāsane sati |\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_040]
-Only possible variant is \emph{mitāhāro} in V19.
-J8 might be correction of J10’s unmetrical reading.
-
-[\emph{sadāsiddhāsanābhyāsād}? Or maybe read \emph{sadā} with \emph{avāpnuyāt}]
-
-V1 has
-
-\begin{versinnote}
-\tl{śramādau bahubhiḥ pīṭhais sadā siddhāsane sati |\\+}
-\tl{prāṇānile sāvadhānaṃ baddhe kevalakumbhake || 1.41 ||\\!}
-\end{versinnote}
-
-Mixing up both versions of the verse — contamination already?
-
-V19 is found in Yogacintāmaṇi: \emph{śramadair bahubhịh pīṭhaiḥ kiṃ syāt siddhāsane sati}; JM this seems best to me.
-
-Is this notion of āsanas causing \emph{śrama} already current in HY texts?
-[Yes, it is mentioned above in the verse on śavāsana]
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.41}
-\begin{translation}[hp01_041]
-Just as the Unmanī [state] arises automatically, without effort, when the prāṇa breath has been carefully stopped in kevala kumbhaka, [...]
-\end{translation}
-
-\begin{sources}[hp01_041]
-\end{sources}
-
-\begin{testimonia}[hp01_041]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{prāṇānile sāvadhāne baddhe kevalakumbhake |\\+}
-\tl{utpatsyate nirāyāsāt svayam evonmanī yathā ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_041]
---
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.42}
-\begin{translation}[hp01_042]
-So too the three bandhas arise automatically, without effort, when Siddhāsana alone is always firmly adopted,
-\end{translation}
-
-\begin{sources}[hp01_042]
---
-\end{sources}
-
-\begin{testimonia}[hp01_042]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{athaikasminn eva dṛḍhaṃ baddhe siddhāsane sadā |\\+}
-\tl{bandhatrayam anāyāsāt svayam evopajāyate |\\!}
-\end{versinnote}
-
-YSS
-
-\begin{versinnote}
-\tl{tathaikasminn eva baddhe siddhāsane sadā |\\+}
-\tl{granthitrayam anāyāsāt svayamevopabhidyate |\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_042]
-\emph{dṛḍhe} goes across stemma.
-It is likely that the adverb (\emph{dṛdhaṃ}) was intended because \emph{dṛdhataraṃ}, which is not ambiguous, is used in 1.48a to qualify how \emph{padmāsana} should be adopted.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.43}
-\begin{translation}[hp01_043]
-There is no posture like siddhāsana, no breath like the restrained breath, no mudrā like khecarī, no dissolution like nāda.
-\end{translation}
-
-\begin{sources}[hp01_043]
-Śivasaṃhitā 5.47
-
-\begin{versinnote}
-\tl{nāsanaṃ siddhasadṛśaṃ na kumbhasadṛśaṃ balam |\\+}
-\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ || 5.47 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_043]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ |\\+}
-\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{nāsanaṃ siddhasadṛśaṃ na kumbhaḥ kevalopamaḥ ||\\+}
-\tl{na khecarīsamā mudrā na nādasadṛśo layaḥ || 3.29 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_043]
-\emph{na kumbhasadṛśo nilaḥ} is difficilior lectio and attested by all early witnesses except V19: (almost?) all testimonia have \emph{kumbhaḥ kevalopamaḥ}; Śivasaṃhitā has \emph{kumbhasadṛśaṃ balam}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.44}
-\begin{translation}[hp01_044]
-Now lotus pose.
-One should place the right foot on the left thigh, and the left on the right though, hold firmly the big toes with the hands behind the back, place the chin on the chest and gaze at the tip of the nose.
-This is called lotus pose, which destroys diseases for those who undertake the yamas.
-\end{translation}
-
-\begin{sources}[hp01_044]
-Vivekamārtaṇḍa 8
-
-\begin{versinnote}
-\tl{vāmorūpari dakṣiṇañ ca caraṇaṃ saṃsthāpya vāmaṃ tathā\\+}
-\tl{yāmyorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\+}
-\tl{aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\+}
-\tl{etad vyādhivikārahāri yamināṃ padmāsanaṃ procyate || 8 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_044]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhayoge—\\+}
-\tl{vāmorūpari dakṣiṇaṃ hi caraṇaṃ saṃsthāpya vāmaṃ tathā\\+}
-\tl{dakṣorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\+}
-\tl{aṅguṣṭhau hṛdaye nidhāya civukaṃ nāsāgram ālokayet\\+}
-\tl{ekad vyādhivikāranāśanakaraṃ padmāsanaṃ procyate ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{vāmorūpari dakṣiṇaṃ ca caraṇaṃ saṃsthāpya vāmaṃ tathā\\+}
-\tl{yāmyorūpari paścimena vidhinā dhṛtvā karābhyāṃ dṛḍham |\\+}
-\tl{aṅguṣṭhau hṛdaye nidhāya cibukaṃ nāsāgram ālokayed\\+}
-\tl{etad vyādhivināśakāri yamināṃ padmāsanaṃ procyate || 3.34 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-%%%%%%%%%%
-\subsection*{1.45--46}
-\begin{translation}[hp01_045]
-However, in another view,
-
-Having put the upturned feet carefully on the thighs and the upturned hands in the middle of the thighs, one should fix the eyes on the tip of the nose.
-Having raised the root of the uvula with the tongue, one should place the chin on the chest and gradually [draw in\footnote{The verb \emph{ākṛśya} follows in the next verse in the \emph{Dattātreyayogaśāstra}.}] the breath [...].
-\end{translation}
-
-\begin{sources}[hp01_045]
-Dattātreyayogaśāstra 35–37
-
-\begin{versinnote}
-\tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\+}
-\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 35 ||\\+}
-\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
-\tl{uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\+}
-\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
-\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\!}
-\end{versinnote}
-
-Śivasaṃhitā 3.102–104
-
-\begin{versinnote}
-\tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ\\+}
-\tl{ūrumadhye tathottānau pāṇī kṛtvā tu tādṛśau 3.102\\+}
-\tl{nāsāgre vinyased dṛṣṭiṃ rājadantaṃ ca jihvayā\\+}
-\tl{uttambhya cibukaṃ vakṣe saṃsthāpya pavanaṃ śanaiḥ 3.103\\+}
-\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ\\+}
-\tl{yathāśaktyaiva paścāt tu recayed anirodhataḥ 3.104\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_045]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{dattātreyaḥ—\\+}
-\tl{uttānau caraṇau kṛtvā ūrusaṃsthau prayatnataḥ |\\+}
-\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau ||\\+}
-\tl{nāsāgre vinyased rājadantamūlaṃ tu jihvayā |\\+}
-\tl{uttabhya civukaṃ vakṣasy utthāpya pavanaṃ śanaiḥ ||\\+}
-\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
-\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{dattātreyo 'pi\\+}
-\tl{uttānau caraṇau kṛtvā ūrvoḥ saṃsthāpya yatnataḥ |\\+}
-\tl{ūrumadhye tathottānau pāṇī kṛtvā tato dṛśau || 3.36 ||\\+}
-\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
-\tl{uttabhya cibukaṃ vakṣaḥ saṃsthāpya pavanaṃ śanaiḥ || 3.37 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_045]
-\emph{uttabhya} vs \emph{uttambhya}.
-The witnesses split predictably along the two main branches of the stemma. The evidence of the DYŚ is important here.
-%\emph{vakṣasyāsthāpya} is a marmasthāna
-
-The statement ending with \emph{pavanaṃ śanaiḥ} is left hanging, perhaps, because of an eyeskip that happened early in the transmission.
-The subsequent verse in the DYŚ ends with \emph{pavanaṃ śanaiḥ}.
-
-\begin{versinnote}
-\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
-\tl{uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ || 36 ||\\+}
-\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
-\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ || 37 ||\\+}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\!}
-\end{versinnote}
-
-The manuscript readings with \emph{vakṣa sthāpayet} (J7, V3, J8, J10, J17, N17) or something similar (V1, W4) do not offer a solution nor indicate how Svātmārāma may have redacted this to make the syntax complete.
-Instead, it seems that he quoted these two verses (1.45–46) from the Dattātreyayogaśāstra and simply omitted the next verse that made sense of \emph{pavanaṃ śanaiḥ} because it was not relevant to the posture itself.
-
-% Jü:
-[JH] The background of the passage \emph{rājadantamūlaṃ ca jihvayā uttambhya} in 1.46 is more complex
-than it may appear. Here it is a literal quotation from the DYŚ, but many other Haṭhayogic texts
-teach a particular position of the tongue, in which it is in one or the other way turned back in
-the direction of the uvula, as we read explicitly in the \emph{Vivekamārtāṇḍa} (REF):
-\emph{kapālakuhare jihvā praviṣṭā viparītagā}. Brief references to this practice can become
-ambiguous for various reasons, and this has possibly confused Brahmānanda.
-
-One reason is that there is a, probably older, rule for meditation postures according to which the
-tongue rests near the teeth. One instance would be \emph{Svacchandatantra}, which teaches a
-meditation pose called \emph{divyaṃ karaṇam} (4.365f.), in which the tongue is to rest at the tip
-of the teeth (\emph{dantāgre jihvām ādāya}). Other Tantric texts have this or similar rules, in
-which the tongue is supposed to rest either on the teeth or the palate,\footnote{This rule is
-found in \emph{Iśānaśivapaddhati}: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān
-  asaṃspṛśan ṛjukāyaḥ}. REF} the earliest instance being perhaps \emph{Mrgendrāgama} ?.18.
-Placing the tongue where it does not disturb the meditation seems quite appropriate for a
-``normal'' meditative practice.
-
-[Jason: There’s also a clear reference to the tip of the tongue being placed in the middle of the palate in 2.27 of the Yogapāda of the Mataṅgapārameśvaratantra (tālumadhyagatenaiva jihvāgreṇa mahāmune).
-In fact, in works that predate Haṭhayoga (i.e., 11th c.), the most common instruction is to put the tip of the tongue on the palate (tālu).]
-
-We might try to interpret the passage in this manner, however, once Haṭhayogic physiology is at the
-background, we must assume that the aim is to reach back to the uvula, to the source of the
-"nectar".  For the background and for the crucial references see Mallinson's note on
-\emph{Khecarīvidyā} 1.65ab.\footnote{p. 209.}  Confusingly Yogic terminology has used and possibly
-invented new names for uvula, and among these especially the term \emph{rājadanta} may give rise to
-confusion, since, as we have seen, the tongue might also in some Yoga systems be placed at the
-(front) teeth.
-
-Furthermore the details in these descriptions of the \emph{khecarīmudrā} are manifold. The 10th
-century \emph{Mokṣopāya} says that the tongue rests at the ``source of the
-palate''\footnote{\emph{tālumūlatalālagnajihvā} MU V.55.14c.} and the commentary
-\emph{Saṃsārataraṇi} on the parallel passage in \emph{Laghuyogavāsiṣṭha} V.6.155, which reads
-\emph{tālumūlāntarālagnajihva-}, explains that this means that the tongue is to be placed in the
-middle of the two regions of the palate, and that this is the \emph{nabhomudrā}, alias
-``\emph{khecarī}''.\footnote{\emph{tālumūlāntarālagnajihvamūlah. tālumūlayoh. kākudamūladeśayoh. āntare
-  lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā | yā hi khecarı̄ty ucyate.}}
-
-A little later in the \emph{Mokṣopāya} it is made clear that one should reach the uvula, ``placed a
-the root of the palate''.\footnote{\emph{tālumūlagatāṃ yatnāj jihvayākramya ghaṇṭikām}
-(V.78.24ab)}
-
-In view of this background we must conclude that the author of the \emph{Jyotsnā} was probably not
-aware of the Yogic meaning of \emph{rājadanta} and has tried his best to make sense of the passage,
-echoing the idea of the two roots of the palate (although his text is not talking about the
-palate), but then wisely refers to the instruction of the teacher for practical details, probably
-noticing that his literal interpretation is somewhat opaque (I omit the synonyms for clarity):
-
-\begin{quote}
-Pressing both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.
-
-\emph{rājadantānāṃ daṃṣṭrāṇāṃ savyadakṣiṇabhāge sthitānāṃ mūle ubhe mūlasthāne jihvayā uttambhya ūrdhvaṃ stambhayitvā | guru-mukhād avagantavyo 'yaṃ jihvā-bandhaḥ |}
-\end{quote}
-
-[Jason: also, Brahmānanda’s comments on 3.22 indicate that he thought rājadanta refers to the front teeth.
-When commenting on rājadantasthajihvāyāṃ, he says kutaḥ ?
-yato dantānāṃ rājāno rājadantā rājadanteṣu tiṣṭhatīti rājadantasthāḥ, rājadantasthā cāsau jihvā ca tasyāṃ rājadantasthajihvāyāṃ bandhaḥ, taduparibhāgasya sambandhaḥ śastaḥ].
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.46}
-\begin{translation}[hp01_046]
-See 1.45.
-\end{translation}
-
-\begin{sources}[hp01_046]
-See 1.45.
-\end{sources}
-
-\begin{testimonia}[hp01_046]
-See 1.45.
-\end{testimonia}
-
-\begin{philcomm}[hp01_046]
-See 1.45.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.47}
-\begin{translation}[hp01_047]
-This is taught as lotus pose, which cures all diseases. It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
-\end{translation}
-
-\begin{sources}[hp01_047]
-DYŚ
-
-\begin{versinnote}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
-\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 38 ||\\!}
-\end{versinnote}
-
-Śivasaṃhitā
-
-\begin{versinnote}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
-\tl{durlabhaṃ yena kenāpi dhīmatā labhyate param || 3.105 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_047]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
-\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
-\tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi || 3.38 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_047]
-In this context, the word \emph{durlabham} is ambiguous as to whether the posture is hard to perform or hard to acquire. The latter is the usual reading.
-% Maybe add some remarks by commentators on how they understand it.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.48}
-\begin{translation}[hp01_048]
-%Next is taught the doctrine of Matsya.
-% [MD] add translation of "paścāduktam eva matsaṃmatam" at the end of the previous section.
-
-[The yogi] should put the hands together in a bowl shape, very firmly assume padmāsana, firmly place the chin on the chest and meditation in the mind, [and] raising the Apāna breath over and over again, releasing the held Prāṇa, a man attains unequalled knowledge through the power of the goddess [Kuṇḍalinī].
-\end{translation}
-
-\begin{sources}[hp01_048]
-VM 36
-
-\begin{versinnote}
-\tl{kṛtvā saṃpuṭitau karau dṛḍhataraṃ baddhvātha padmāsanaṃ\\+}
-\tl{gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyānaṃś ca tac cetasi |\\+}
-\tl{vāraṃ vāram apānam ūrdhvam anilaṃ proccālayan pūritaṃ\\+}
-\tl{muṇcan prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_048]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{tathāca granthāntare—\\+}
-\tl{kṛtvā saṃpuṭitau karau dṛḍhataraṃ badhvā ca padmāsanam\\+}
-\tl{gāḍhaṃ vakṣasi saṃnidhāya civukaṃ dhyānaṃ ca tac cetasi |\\+}
-\tl{vāraṃ vāram apānam ūrdhvam anilaṃ protsārayet pūrayet\\+}
-\tl{prāṇaṃ muñcati bodham eti niyataṃ śaktiprabodhodayāt ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{kṛtvā saṃpuṭitau karau dṛḍhataraṃ baddhvā tu padmāsanam\\+}
-\tl{gāḍhaṃ vakṣasi sannidhāya cibukaṃ dhyānaṃ ca tac cetasi |\\+}
-\tl{vāraṃ vāram apānam ūrdhvam anilaṃ proccārayet pūritam\\+}
-\tl{muñcat prāṇam upaiti bodham atulaṃ śakteḥ prabhāvān naraḥ || 3.39 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_048]
-%The reading of \emph{matsyamataṃ} in the subheading is unusual because elsewhere Matsyendra is not referred to as simply Matsya.
-Furthermore, this verse is from the \emph{Vivekamārtaṇḍa} (36), and the teachings of the \emph{Vivekamārtaṇḍa} are attributed to Gorakṣa, not Matsyendra.
-However, in the extended recension of it known as the  \emph{Gorakṣaśataka},  \emph{Gorakṣasaṃhitā}, etc.
-the second verse includes homage to Mīna, i.e.
-Matsyendra.
-
-The end of pāda b is uncertain.
-Later witnesses including Brahmānanda have \emph{dhyāyaṃś ca} but none of the early ones has this reading.
-So, we have to take \emph{dhyānaṃ} with \emph{sannidhāya}.
-However, \emph{tat} is problematic because here it has no clear referent.
-In the \emph{Vivekamārtaṇḍa tat} appears to refer to the \emph{mokṣadvāra} broken by Kuṇḍalinī, which is mentioned in the previous verse.
-
-The two participles \emph{proccālayan} and \emph{muñcan} imply that the two things are happening at the same time, which is surprising but perhaps possible.
-Or perhaps one can understand them as sequential.
-
-In the VM this verse comes in the passage on raising Kuṇḍalinī, so tat could refer to visualising K opening the mokṣadvāra.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.49}
-\begin{translation}[hp01_049]
-The yogi in padmāsana who fills [himself] up through the openings of the channels and holds the breath is sure to be liberated.
-\end{translation}
-
-\begin{sources}[hp01_049]
---
-\end{sources}
-
-\begin{testimonia}[hp01_049]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\+}
-\tl{mārutaṃ dhārayed yas tu sa mukto nātra saṃśayaḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī 3.40
-
-\begin{versinnote}
-\tl{padmāsane sthito yogī nāḍīdvāreṣu pūrayet |\\+}
-\tl{pūritaṃ dhrīyate yas tu sa mukto nātra saṃśayaḥ || 3.40 ||\\!}
-\end{versinnote}
-
-Dhyānabindu Up
-
-\begin{versinnote}
-\tl{padmāsanasthito yogī nāḍīdvāreṣu pūrayan |\\+}
-\tl{mārutaṃ kumbhayan yas tu sa mukto nātra saṃśayaḥ || 70 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_049]
-pūrayan can take the locative (e.g., Amaraugha 21d pūrayen mukhe).
-The reading of N23 niyatam would explain the passive verbs in other witnesses.
-The passive verbs do not make sense with yas tu.
-This yields the idea of filling up the central channel (as opposed to other channels) seen in the Yogabīja.
-Note that a breath retention is not mentioned in the previous verse describing padmāsana (only inhalation and exhalation).
-Also,  māruto mriyate yasya is possible (if one  accepts pūrayet in the second pāda).
-J5 also has a plausible reading (mārutaṃ mārayet yas tu) but J5 is contaminated (so the Gr3 reading dhārayet is more reliable).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.50--52}
-\begin{translation}[hp01_050]
-Now the lion’s pose. ---
-
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. He should place the left ankle on right, the right ankle on the left [...]
-\end{translation}
-
-\begin{translation}[hp01_051]
-[...] and both hands on the knees, spread his fingers, open his mouth and gaze in deep concentration at the tip of the nose.
-\end{translation}
-
-\begin{translation}[hp01_052]
-This is the lion’s pose, which is always worshipped by yogis. It causes the three locks to arise together and is the best of [all] postures.
-\end{translation}
-
-\begin{sources}[hp01_050]
-Vasiṣṭhasaṃhitā 1.73
-
-\begin{versinnote}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{dakṣiṇaṃ savyagulphena dakṣiṇenetaretaram ||\\+}
-\tl{hastau jānau ca saṃsthāpya svāṅgulīś ca prasārya ca |\\+}
-\tl{vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ ||\\+}
-\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā ||\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.9
-
-\begin{versinnote}
-\tl{gulpau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\+}
-\tl{dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram ||\\+}
-\tl{hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca\\+}
-\tl{vyāttavaktro nirīkṣet nāsagraṃ susamāhitaḥ ||\\+}
-\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā\\!}
-\end{versinnote}
-
-Sūtasaṃhitā 15.7
-
-\begin{versinnote}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{dakṣiṇaṃ savyagulphena vāmaṃ dakṣiṇagulphataḥ || 7 ||\\+}
-\tl{hastau ca jānvoḥ saṃsthāpya svāṅgulīś ca prasārya ca |\\+}
-\tl{nāsāgraṃ ca nirīkṣeta bhavet siṃhāsanaṃ hi tat || 8 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_050]
-Yogacintāmaṇi (citing Yogayājñavalkya)
-
-\begin{versinnote}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{dakṣiṇaṃ savyagulphena dakṣiṇena tathetaram ||\\+}
-\tl{hastau jānūpari sthāpya svāṅgulīḥ saṃprasārya ca |\\+}
-\tl{vyāttavaktro nirīkṣeta nāsāgraṃ susamāhitaḥ |\\+}
-\tl{siṃhāsanaṃ bhaved etat pūjitaṃ yogibhiḥ sadā |\\!}
-\end{versinnote}
-
-[Note: The Yogacintāmaṇi attributes its citation of verses on siṃhāsana to the Yogayājñavalkya and does not include HP 1.52cd, which affirms that 1.52cd is not from the Yogayājñavalkya (or Vasiṣṭhasaṃhitā)].
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{atha siṃhāsanam\\+}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{dakṣiṇe savyagulphaṃ ca dakṣiṇe tu tathetaram || 3.31 ||\\+}
-\tl{hastau tu jānvoḥ saṃsthāpya svāṅgulīḥ samprasārya ca ||\\+}
-\tl{vyāttavaktro nirīkṣeta nāsāgraṃ tu samāhitaḥ || 3.32 ||\\+}
-\tl{siṃhāsanaṃ bhaved etat sevitaṃ yogibhiḥ sadā ||\\+}
-\tl{bandhatritayasaṃsthānaṃ kurute cāsanottamam || 3.33 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_051]
-% comment on nyastalocanaḥ and susamāhitaḥ.
-The latter is well attested by the sources and testimonia, and is in the V19 and N23 branch.
-However, most of the old mss.
-have nyastalocanaḥ.
-Maybe comment on the meaning of sandhāna, which seems to mean simply that the three locks arise together (cf.
-a similar comment on siddhāsana in 1.42).
-
-Spreading the fingers and keeping the mouth open seem to be imitating the lion, and this is depicted in some iconography of yoganarasiṃha (for example, Yoga Narasimha, Vishnu's Man-Lion Incarnation, Samuel Eilenberg Collection, Bequest of Samuel Eilenberg, 1998, Accession Number: 2000.284.4.
-https://www.metmuseum.org/art/collection/search/39251).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.51}
-\begin{sources}[hp01_051]
-See 1.50.
-\end{sources}
-
-\begin{testimonia}[hp01_051]
-See 1.50.
-\end{testimonia}
-
-\begin{philcomm}[hp01_051]
-See 1.50.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.52}
-\begin{sources}[hp01_052]
-See 1.50.
-\end{sources}
-
-\begin{testimonia}[hp01_052]
-See 1.50.
-\end{testimonia}
-
-\begin{philcomm}[hp01_052]
-See 1.50.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.53--54}
-\begin{translation}[hp01_053]
-Now, the friendly pose.
-
-[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.  Having firmly and motionlessly held the sides of the feet with the hands,[...]
-\end{translation}
-
-\begin{translation}[hp01_054]
-[...] this is bhadrāsana, which cures all diseases and poisons. Yogis of the siddha tradition call it Gorakṣāsana.
-\end{translation}
-
-\begin{sources}[hp01_053]
-Vasiṣṭhasaṃhitā 1.79
-
-\begin{versinnote}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipan |\\+}
-\tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam |\\+}
-\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham ||\\!}
-\end{versinnote}
-
-Yogayājñavalkya 3.11cd--3.12ab
-
-\begin{versinnote}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet\\+}
-\tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam\\+}
-\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_053]
-Yogacintāmaṇi (citing Yājñavalkya)
-
-\begin{versinnote}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ badhvā suniścalaḥ |\\+}
-\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham |\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{atha bhadrāsanam\\+}
-\tl{gulphau ca vṛṣaṇasyādhaḥ sīvanyāḥ pārśvayoḥ kṣipet ||\\+}
-\tl{pārśvapādau ca pāṇibhyāṃ dṛḍhaṃ baddhvā suniścalam ||\\+}
-\tl{bhadrāsanaṃ bhaved etat sarvavyādhiviṣāpaham || 3.30 ||\\!}
-\end{versinnote}
-
-YSS
-
-\begin{versinnote}
-\tl{gulphau ca vṛṣaṇasyāyasvīvinyāḥ pārśvayoḥ kṣipet |\\+}
-\tl{pārśvapāde ca pāṇibhyāṃ dṛḍhaṃ badhvāsu niścalaḥ ||\\+}
-\tl{bhadrāsanaṃ bhaved etat sarvavyādhivināśanam |\\+}
-\tl{gorakṣāsanam ity āhur idaṃ vai siddhayoginaḥ ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_053]
-We have understood \emph{pārśvapāda} (i.e., side of the foot) like the compound \emph{agrapāda} (i.e., front of the foot), which can be classified as a \emph{ekadeśitatpuruṣa}.
-See Sanskrit Vademecum, p.84 (ed.\ Maximilian Mehner and Jürgen Hanneder, Marburg 2019).
-
-% [MD] comment on the additional hemistich "savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe" (N23,Jyo)?
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.54}
-\begin{sources}[hp01_054]
-See 1.53.
-\end{sources}
-
-\begin{testimonia}[hp01_054]
-See 1.53.
-\end{testimonia}
-
-\begin{philcomm}[hp01_054]
-See 1.53.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.55}
-\begin{translation}[hp01_055]
-When the great yogi does not become tired from adopting the āsanas in this way, he should now practise the breath techniques with mudrās and so forth from which purification of the channels arise.
-\end{translation}
-
-\begin{sources}[hp01_055]
-\end{sources}
-
-\begin{testimonia}[hp01_055]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{evam āsanabandheṣu yogīndro vijitaśramaḥ |\\+}
-\tl{abhyasen nāḍiśuddhiṃ ca mudrayā pavanakriyām || iti ||\\!}
-\end{versinnote}
-
-HSC
-
-\begin{versinnote}
-\tl{evam āsanabandhastho yogīndro vigataśramaḥ |\\+}
-\tl{athābhyasen nāḍiśuddhiṃ mudrādipavanakriyām || 73 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_055]
-55c is ra-vipulā with \emph{nāḍi}
-The second hemistich can be interpreted in different ways.
-One possibility is to understand \emph{nāḍiśuddhiṃ} as a bahuvrīhi qualifying \emph{mudrādipavanakriyām} in the sense that one should practise a breathing technique with mudrās from which purification of the channels arise.
-Alternatively, one could separate \emph{mudrādi} from \emph{pavanakriyām} and understand three different techniques here, namely, the practice of purifying the channels (perhaps by the alternative nostril method mentioned at the beginning of the second chapter), the \emph{mudrā}s and the breathing techniques of prāṇāyāma.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.55*1--2}
-\begin{translation}[hp01_055_1]
-Success arises for one engaged in practice. How can it arise for one who has no practice? Success in yoga does not arise by merely reading scriptures.
-\end{translation}
-
-\begin{translation}[hp01_055_2]
-Wearing a robe does not bring about success, nor does talking [about yoga].
-Practice alone is the cause of success.
-This is true, there is no doubt.
-In this tradition, it should not be given to one who wears robes and is devoted to his genitals and stomach.
-\end{translation}
-
-\begin{sources}[hp01_055_1]
-Dattātreyayogaśāstra 42cd--47
-
-\begin{versinnote}
-\tl{kriyāyuktasya siddhiḥ syād akriyasya kathaṃ bhavet |\\+}
-\tl{na śāstrapāṭhamātreṇa kā cit siddhiḥ prajāyate |\\+}
-\tl{na veṣadhāraṇaṃ siddheḥ kāraṇaṃ na ca tatkathā |\\+}
-\tl{kriyaivakāraṇaṃ siddheḥ satyam eva tu sāṃkṛte || 46 ||\\+}
-\tl{śiśnodarārthaṃ yogasya kathayā veṣadhāriṇaḥ |\\+}
-\tl{anuṣṭhānavihīnās tu vañcayanti janān kila || 47 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_055_1]
-\end{testimonia}
-
-\begin{philcomm}[hp01_055_1]
-1.55*1–2 are omitted from V1 and V19, so may not have been included in the earliest version of the text.
-In fact, it is possible that both were added (perhaps initially as marginal notes) to elaborate on kriyā in 1.55d.
-Both verses are similar to verses from the Dattātreyayogaśāstra, and these verses (except 1.55.2ef) appear in the vulgate, but towards the end of chapter 1 (1.65–66).
-
-And then just half of DYŚ 47 is given, resulting in a near-nonsensical hemistich.
-
-The syntax of 1.55.2ef is corrupt. One has to emend to \emph{deyā} to make sense of it.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.55*2}
-\begin{sources}[hp01_055_2]
-See 1.55*1.
-\end{sources}
-
-\begin{testimonia}[hp01_055_2]
-See 1.55*1.
-\end{testimonia}
-
-\begin{philcomm}[hp01_055_2]
-See 1.55*1.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.55*3}
-\begin{translation}[hp01_055_3]
-In me, the pure ocean of awakening, is this empty bubble we call the universe dissolved or does it arise? Where does this veil of doubt about this come from?
-\end{translation}
-
-\begin{testimonia}[hp01_055_3]
-Vārāhītantra
-
-\begin{versinnote}
-\tl{mayi bodhībudho svasthe tucho yaṃ viśvabudbudaḥ |\\+}
-\tl{malīna udito vetti vikalpāvasaraḥ kutaḥ |\\!}
-\end{versinnote}
-
-Siddhāntamuktāvalī
-
-\begin{versinnote}
-\tl{śiśnodararatāya hi na deyaṃ veṣadhāriṇe ||\\+}
-\tl{mayi bodhyaṃ buddhau svacche tad dheyaṃ viśvabudbudam || 7 ||\\!}
-\end{versinnote}
-
-Yogaprakāśikā
-
-\begin{versinnote}
-\tl{“śiśnodararatāyaitan na deya" etat yogajñānam etena śiśnodararatas tyājyo nanv etanmate tyājyapadārtho 'prasiddha iti śaṃkāṃ nirasyati mayi iti svacche bodhasvarūpasamudre budbudatulyasya viśvasya heyatvād iti bhāvaḥ\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_055_3]
-It is very difficult to find a reason why this verse should be inserted here. It is apparently a
-\emph{muktaka} that would befit an accomplished spiritual poem more than our \emph{Haṭhapradīpikā},
-even here, in what appears as a sort of miscellaneous section at the end of a chapter.  In this
-verse, the lyrical subject wonders about why the mind is still able to doubt, despite its insight
-into the nature of reality and the reader wonders, how the illusionist verse could be understood to
-fit our Yoga text, perhaps the scribe of the archetype of V19 and N17, was fond of it.
-
-The source is, as far as we can say, the \emph{Śāntiśataka} of the Kashmirian poet Sillana or
-Silhaṇa,\footnote{The mss.\ of his \emph{Svātmopalabdhiśataka} give the name as Sillana, the mostly
-Bengali mss. of the Śāntiśataka read Śilhaṇa, as does Aufrecht in his CC. See Hanneder:
-forthcoming.} who cannot be dated with any certainty, but predates the HP by a few centuries. The
-edition of this text – where hundred original verses had to be identified – places it into an
-appendix of doubtful stanzas,\footnote{Karl Schönfeld: \emph{Das Śāntiśataka}. Leipzig:
-Harrassowitz 1910, p.\,90 (A9).} but the editor saw no compelling reason not to regard it as
-original except through the fact that it is not transmitted in all manuscripts.  What prevents
-further investigation of the matter is the lack of Kashmirian mss.\ for the \emph{Śāntiśataka} and
-its compilatory character: one quarter of the material is identical with Bhartṛhari's
-\emph{Vairāgyaśataka}. A still superficial glance at Sillana's \emph{Svātmopalabdhiśataka} gives
-the impression that our verse would fit there, but not so much in the \emph{Śāntiśataka}. Perhaps its first citation is in Advayavajra’s \emph{Tattva\-ratnāvalī} (ref??),  While
-these are only preliminary observations the verse is probably not original in any stage of
-development of the HYP, but limited to V19 (and N17).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.55*4}
-\begin{translation}[hp01_055_4]
-Realisation from scripture, one's own guru or oneself and the cessation of mind; all these methods have been combined and taught by the wise in this tradition.
-\end{translation}
-
-\begin{sources}[hp01_055_4]
-\end{sources}
-
-\begin{testimonia}[hp01_055_4]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhapradīpikāyām--\\+}
-\tl{śrutapratītiḥ svagurupratītiḥ, svātmapratītir manaso nirodhaḥ |\\+}
-\tl{etāni sarvāṇi samuccitāni, matāni dhīrair iha sādhanāni ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_055_4]
-last greyscale verse (\emph{śrutipratītiḥ…}) is only in V19 of old mss (but some other later ones) and is quoted in Yogacintāmaṇi.
-
-The reading in the Yogacintāmaṇi (\emph{manaso nirodhaḥ}) is better in a yogic context.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.56}
-\begin{translation}[hp01_056]
-The various āsana, breath retention, bodily technique (karaṇa) called seals (mudrā), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.
-\end{translation}
-
-\begin{sources}[hp01_056]
-\end{sources}
-
-\begin{testimonia}[hp01_056]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhapradīpikāyām--\\+}
-\tl{āsanaṃ kumbhakaṃ citraṃ mudrākhyaṃ karaṇaṃ tathā |\\+}
-\tl{atha nādānusandhānam abhyāsānukrameṇa ca ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_056]
-Verse is omitted from V1, probably deliberately as it doesn’t have chapter 4, which teaches nādānusandhāna. The numbering in V1 suggests that its exemplar had this verse.
-
-Kumbhaka is almost always masculine.
-
-Marmasthāna: not clear whether to adopt \emph{citro} or \emph{citraṃ}, or V19’s \emph{mudrākhyaṃ karaṇaṃ tathā} or the others’ \emph{mudrādikaraṇāni ca}.
-
-See also 1.65, which has \emph{mudrādikaraṇāni ca}, so perhaps it was through confusion with this that the same reading is found in some witnesses of 1.56.
-
-Yes, V19 reading probably best.
-
-\begin{versinnote}
-\tl{āsanaṃ kumbhakaś citraṃ mudrākhyaṃ karaṇaṃ tathā |\\+}
-\tl{atha nādānusandhānam abhyāsānukramo haṭhe ||\\!}
-\end{versinnote}
-
-It seems that the four aṅgas of Haṭhayoga are being referred to in the singular (hence \emph{āsanaṃ}), whereas in 1.65 the plural is used (i.e., \emph{pīṭhāni}).
-Therefore, we should adopt \emph{citraṃ [...] karaṇaṃ tathā}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.57}
-\begin{translation}[hp01_057]
-Celibate, restricted in diet and devoted to yoga, the renunciant succeeds in upwards of a year. No doubt about this should be entertained.
-\end{translation}
-
-\begin{sources}[hp01_057]
-Vivekamārtaṇḍa 37
-
-\begin{versinnote}
-\tl{brahmacārī mitāhārī yogī yogaparāyaṇaḥ |\\+}
-\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_057]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ |\\+}
-\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{brahmacārī mitāhārī tyāgī yogaparāyaṇaḥ ||\\+}
-\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā || 3.28 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_057]
-Comment on tyāgī and yogī, and the reading of this verse in the Vivekamārtaṇḍa.
-It seems likely that Svātmārāma was using a version of the VM with tyāgī (as it’s unlikely that yogī would be emended to tyāgī), but whether this was the original reading in the VM is more complicated because there tyāgī may be a dittographical-type mistake.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.58}
-\begin{translation}[hp01_058]
-When very unctuous and sweet food that is without a quarter portion is eaten for love of śiva, it is called a restricted diet. (\emph{mitāhāra}).
-\end{translation}
-
-\begin{sources}[hp01_058]
-Gorakṣaśataka (original)
-
-\begin{versinnote}
-\tl{susnigdhamadhurāhāraś caturthāṃśavivarjitaḥ |\\+}
-\tl{bhujyate śivasaṃprītyai mitāhāraḥ sa ucyate ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_058]
-Yuktabhavadeva
-
-\begin{versinnote}
-\tl{tad uktaṃ haṭhapradīpikāyām-\\+}
-\tl{susnigdhamadhurāhārāś caturthāṃśavivarjitaḥ ||\\+}
-\tl{bhujyate śivasamprītyai mitāhāraḥ sa ucyate || 16 ||\\!}
-\end{versinnote}
-
-Yogacūḍāmaṇyupaniṣat
-
-\begin{versinnote}
-\tl{susnigdhamadhurāhāraścaturthāṃśavivarjitaḥ |\\+}
-\tl{bhuñjate śivasaṃprītyā mitāhārī sa ucyate || 43 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_058]
-This verse probably derives from the ‘original’ GŚ (12c–13b).
-It is also found, but reworked to be about the \emph{mitāhārī}, in Nowotny’s GŚ (extended recension of VM) at 55.
-The idea of not eating the fourth portion of one’s food (\emph{caturthāṃśavivarjitaḥ}) can be found in older sources, such as Dharmaputrikā 1.51-52 (ṣaḍrasopetasuṣnigdhasvādusāndrasugandhinā |
-udarasyārdhabhāgan tu bhojanena prapūrayet || pānīyena caturbhāgaṃ taccheṣaṃ śūnyam iṣyate | vāyos sañcāraṇānārtham āhāraniyamaḥ smṛtaḥ) and Aṣṭāṅgahṛdayasaṃhitā Sūtrasthāna 8.46c-47b ( annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet ||āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet).
-As noted in the Jy
-
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.59}
-\begin{translation}[hp01_059]
-[Adepts] say the [following] is unwholesome: pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink.
-Flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, Jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like.
-\end{translation}
-
-\begin{sources}[hp01_059]
-\end{sources}
-
-\begin{testimonia}[hp01_059]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhapradīpikāyām--\\+}
-\tl{kaṭvamlatīkṣṇalavaṇoṣṇaharītaśāka-\\+}
-\tl{sauvīratailatilasarṣapamatsyamadyam |\\+}
-\tl{ajādimāṃsadadhitakrakulatthakola-\\+}
-\tl{piṇyākahiṅgulaśunādyam apathyam āhuḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{kaṭvamlatīkṣṇalavaṇoṣṇaharītaśākaṃ\\+}
-\tl{sauvīratailatilasarṣapamatsyamadyam |\\+}
-\tl{ajādimāṃsadadhitakrakulatthakodra-\\+}
-\tl{piṇyākahiṅgulaśunādyam apathyam āhuḥ || 1.72 ||\\!}
-\end{versinnote}
-
-HTK
-
-\begin{versinnote}
-\tl{atha varjyāni –\\+}
-\tl{kaṭvamlatīkṣṇalavaṇoṣṇa haritaśāka-\\+}
-\tl{sauvīratailatilasarṣapamatsyamadyam ||\\+}
-\tl{ajāvimāṃsadadhitakrakulatthakola-\\+}
-\tl{piṇyākahiṃgulaśunādyam apathyam āhuḥ || 28 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_059]
-59a \emph{kaṭvamla°} is better than \emph{kaṭvāmla°} and well attested elsewhere in lists of tastes and types of foods.
-On the meaning of \emph{uṣṇa} (in relation to food) see Meulenbeld 1974: 254 fn.
-13: ‘Cakra mentions as a variant: \emph{katvamlalavaṇakṣāra} (pungent.
-acid, saline and caustic).
-Cakra remarks that the term `hot' (\emph{uṣṇa}) denotes hot on touch when it occurs the first time, and hot with regard to potency when it occurs for the second time.’
-
-\emph{°hari}(\emph{ī} here for metre?)\emph{taśāka°} in some nighaṇṭus is horseradish, which makes better sense here than Brahmānanda’s gloss.
-
-Rājanighaṇṭu 7.26
-
-\begin{versinnote}
-\tl{śigrur haritaśākaś ca śākapattraḥ supattrakaḥ |\\!}
-\end{versinnote}
-
-Sauśrutanighaṇṭu 75ab
-\begin{versinnote}
-\tl{śigruko haritaśākaś ca mato vai mūlapatrakaḥ |\\!}
-\end{versinnote}
-
-Brahmānanda’s understanding of \emph{harītaśāka} as \emph{patraśāka} is probably wrong if \emph{patraśāka} was intended as ‘leafy vegetables’.
-But perhaps \emph{patraśāka} can also mean \emph{śigru} (note \emph{śākapattraḥ} above).
-
-Anusvāra at end of \emph{śāka}?
-
-\emph{°sauvīra°} probably means sour gruel.
-Brahmānanda: \emph{sauvīra} = \emph{kāñjika} (fermented rice water).
-Meulenbeld, madhavanidāna pp. 516–517
-\emph{sauvīra} is sour gruel made from barley and wheat. On the process see, Suśruta 1.44.35--40ab.
-(PV Sharma’s translation of this passage:)
-
-\begin{quote}
-‘Roots of trivṛt etc., the first group (\emph{vidārigandhādi}), \emph{mahat pañcamūla}, \emph{mūrvā} and \emph{śārṅgaṣṭā}, and also of \emph{snuhī}, \emph{haimavatī}, \emph{triphalā}, \emph{ativiṣā} and \emph{vacā} -- these are taken and divided into two parts out of which one is decocted and the other is powdered; now, crushed barley grains are impregnated with the above decoction several times, dried and then slightly fried.
-Taking three parts of this and one part of the above powder are put in a jar and mixed with their (of \emph{trivṛt}, etc.) cold decoction and fermented properly.
-This is known as \emph{sauvīraka}.’
-\end{quote}
-
-But according to some nighaṇṭus, \emph{sauvīra} can also mean stibnite (an ingredient in some añjana’s and medicines):
-
-Rājanighaṇṭu 13.86
-
-\begin{versinnote}
-\tl{añjanaṃ yāmunaṃ kṛṣṇaṃ nādeyaṃ mecakaṃ tathā\\+}
-\tl{srotojaṃ dṛkpradaṃ nīlaṃ sauvīraṃ ca suvīrajam ||\\!}
-\end{versinnote}
-
-Note also that the Yogaprakāśikā takes \emph{sauvīra} with \emph{taila}, perhaps to solve the problem of \emph{taila} on its own (see below for more on this): \emph{sauvīrataila} -- oil produced in the place Suvīra (\emph{suvīradeśodbhavatailam}).
-
-Suvīra , a country the people of which, also called Suvira (V.79), Sauvira (XVI.21) and Sauvīraka (IV.23) ...
-S.M.
-Ali identifies it with the Rohri - Khairpur region of Sind (Geography of the Purānas, Delhi, 1966, p.
-144).
-
-\emph{taila} could refer to \emph{tilataila}:
-Śārṅgadharasaṃhitā:
-\emph{anuktāvasthāyāṃ paribhāṣāvidhiḥ [...] taile ’nukte tilodbhavam} 48
-
-Dominik Wujastyk supplied this reference and may be able to comment more on taila in this list.
-
-On the translation of \emph{madya}, see Mchugh (An Unholy Brew) 2021: 8.
-
-\emph{ājādimāṃsa}: āja° is required for the metre, so only ājādi makes sense, not ājāvi because there is no adjective āvi.
-
-Diwakar Acharya: prohibition of goat flesh and fish is aimed at north/east India.
-
-\emph{kulattha} is a kind of pulse, translated by Dominik Wujastyk as `poor man's pulse' (see Roots of Ayurveda).
-
-\emph{kola}: Zizyphus Jujuba (Nadkarni 1926: pp.
-919--920).
-Also known as \emph{badara}.
-This is how Brahmānanda takes it (\emph{kolaṃ kolyāḥ phalaṃ badaram}).
-According to Nadkarni, the fruit of the wild variety is very acid and astringent.
-It is believed to purify the blood and assist digestion.
-The bark is astringent and a simple remedy for diarrhoea.
-Root is useful as a decoction in fever and delirium.
-kola can mean banana in some parts of India (Diwakar)
-
-There are references to \emph{kola} being pungent, though this does not seem to indicate sufficiently why \emph{kola} is mentioned separately as an \emph{apathya} food.
-
-\emph{piṇyāka}: Sharma (Ḍalhaṇa and his commentary on drugs: 1982: 69) says, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’ Diwakar says it is an oil cake that has a strong flavour.
-
-\emph{hiṅgu}: Asafoetida (Nadkarni 1926: pp 360–361): `If long continued even in moderate doses, it gives rise to alliaceous eructations, acrid irritation in the throat, flatulence, diarrhoea and burning in the urine.'
-
-\emph{laśuna} = garlic (Nadkarni 1926: 45).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.60}
-\begin{translation}[hp01_060]
-One should know food that has been reheated, is dry, too salty or sour, contains an excess of leafy vegetables that are hard to chew, [or] is spoiled to be unwholesome.
-\end{translation}
-
-\begin{sources}[hp01_060]
-\end{sources}
-
-\begin{testimonia}[hp01_060]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{bhojanam ahitaṃ vidyāt punar <apy> uṣṇīkṛtaṃ tathā rukṣam |\\+}
-\tl{atilavaṇaṃ sapalaṃ vā prasitaṃ śākotkaṭaṃ varjyam ||\\!}
-\end{versinnote}
-
-HSC
-
-\begin{versinnote}
-\tl{bhojanam ahitaṃ vidyāt punar uṣṇīkṛtaṃ rūkṣaṃ |\\+}
-\tl{atilavaṇādikayuktaṃ kadaśanaśākotkaṭaṃ duṣṭaṃ ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_060]
-We have understood the compound \emph{kadaśanaśākotkaṭa} as a dvandva referring to spoiled food (\emph{kadaśana}) and an excess of leafy vegetables (\emph{śākotkaṭa}).
-We have not found any conclusive evidence for the meaning of \emph{tilapiṇḍa}. Brahmānanda glosses it as \emph{piṇyāka} (on which see the notes for the previous verse).
-
-Not sure how to take \emph{kadaśanaśākotkaṭaṃ}.
-Brahmānanda understands it as a dvandva (i.e., \emph{kadaśana}, \emph{śāka}, \emph{utkaṭa}), where \emph{śāka} is prohibited vegetables and \emph{utkaṭa} is pepper.
-
-The meaning of \emph{utkaṭa} is not clear.
-The word \emph{utkaṭā} can mean pepper according to some nighaṇṭus (e.g., Rājanighaṇṭu 5.16 \emph{pārvatī śailajā tāmrā lambabījā tathotkaṭā}).
-But \emph{utkaṭa} can refer to Saccharum Sara and \emph{utkaṭā} also to Laurus Cassia (cinnamon).
-
-Also, \emph{utkaṭa} can be an adjective that means ‘abounding in’ at the end of a compound.
-So could \emph{kadaśanaśākotkaṭaṃ} have been intended as an adjectival tatpuruṣa along the lines of ‘[food] full of spoiled vegetables’?
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.61}
-\begin{translation}[hp01_061]
-Similarly a saying by Goraksa:
-One should avoid places near bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.
-\end{translation}
-
-\begin{sources}[hp01_061]
-Amṛtasiddhi 19.7
-
-\begin{versinnote}
-\tl{agnisevābalāsevā pathasevā ca sarvadā |\\+}
-\tl{prathamābhyāsakāle tu saṃtyājyā yoginā sadā || 19.7 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_061]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhadīpikāyām—\\+}
-\tl{varjayed durjanaprītiṃ vahnistrīpathasevanam |\\+}
-\tl{prātaḥsnānopavāsādi kāyakleśādikaṃ tathā ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{tathā ca gorakṣavacanam---\\+}
-\tl{varjayed durjanaprītivahnistrīpathasevanam |\\+}
-\tl{prātaḥsnānopavāsādi kāyakleśādikaṃ tathā || 1.73 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_061]
-The vulgate has a parallel from the Amaraugha added. Stemma requires prāṃtam, which is difficilior lectio in relationship to prītiṃ.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.62}
-\begin{translation}[hp01_062]
-Pure food with wheat, rice, śāli rice, barley, sixty-day śāli rice, milk, ghee, cream, unclarified fermented butter, ground sugar and honey.
-Dried ginger, fruit of the snake gourd and so forth, the five vegetables, mung beans and so on, and rain water are wholesome for the best of sages.
-\end{translation}
-
-\begin{sources}[hp01_062]
-\end{sources}
-
-\begin{testimonia}[hp01_062]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ\\+}
-\tl{kṣīrājyamaṇḍanavanītasitāmadhūni |\\+}
-\tl{śuṇṭhīpaṭolakaphalādikapañcaśākaṃ\\+}
-\tl{mudgādi cālpam udakaṃ ca munīndrapathyam ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{godhūmaśāliyavaṣaṣṭikaśobhanānnaṃ\\+}
-\tl{kṣīrājyamaṇḍanavanītasitāmadhūni |\\+}
-\tl{śuṇṭhīpaṭolaphalapatrajapañcaśākaṃ\\+}
-\tl{mudgādidivyam udakaṃ ca yamīndrapathyam || 1.71 ||\\!}
-\end{versinnote}
-
-YBD
-
-\begin{versinnote}
-\tl{tathā ca śivayoge-\\+}
-\tl{godhūmaśāliyavaṣāṣṭikaśobhanānnaṃ\\+}
-\tl{kṣīrājyakhaṇḍanavanītasitāmadhūni ||\\+}
-\tl{śuṇṭhīpaṭolakaphalādi ca pañcaśāka-\\+}
-\tl{mudgādidivyam udakaṃ ca munīndrapathyam || 21 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_062]
-\emph{khaṇḍa} -- candied sugar (Meul 507, different types of sugar).
-
-\emph{navanīta} (MW fresh butter), Mchugh (2021) unclarified fermented butter.
-
-\emph{sitā} -- ground sugar (Meul 507, different types of sugar) ``sitā is very white and looks like gravel"
-
-\emph{madhu} -- honey.
-
-\emph{paṭola} can refer to at least two different gourds. See Meul. p. 569 for a long list of possibilities, including TRICHOSANTHES DIOICA ROXB. (`pointed gourd'), T. CUCUMERINA LINN (snake gourd).
-
-Nadkarni has two entries on \emph{paṭola}:
-\begin{enumerate}
-\item snake gourd (Nadkarni p.
-863) is common in Bengal and cultivated in Northern India and Punjab.
-The unripe fruit of this climbing plant is generally used as a culinary vegetable and is very wholesome, specially suited for the convalescent.
-
-\item smooth luffa (Nadkarni p.
-518) is a hairy climbing herb extensively cultivated in several parts of India.
-The fruit is edible.
-Medicinally it is described as `cool, costive, demulcent, producive of loss of appetite and excitive of wind, bile and phlegm")
-\end{enumerate}
-
-Wikipedia : smooth luffa = Luffa aegyptiaca (sponge gourd)
-Sharma (Syn. Kulaka. Well known (Trichosanthas dioica Roxb.)
-
-Brahmānanda glosses it as \emph{kośātakī} (Meul p. 586 LUFFA ACUTANGULA ROXB), which suggests he thought it was some sort of luffa.
-
-Brahmānanda also mentions the vernacular term \emph{paravara} for \emph{paṭola}, which the Lonavla ed.
-states is a kind of cucumber.
-However, Paras remarked that \emph{paravara} is more like a gourd (hard shell, etc.).
-
-On \emph{pañcaśāka}, see GhS
-
-\begin{versinnote}
-\tl{bālaśākaṃ kālaśākaṃ tathā paṭolapatrakam |\\+}
-\tl{pañcaśākaṃ praśaṃsīyād vāstūkaṃ hilamocikāṃ || 5.20 ||\\!}
-\end{versinnote}
-
-HTK 4.26
-
-\begin{versinnote}
-\tl{pañcaśākastu –\\+}
-\tl{kṣīraparṇī ca jīvantī matsyākṣī ca punarnavā\\+}
-\tl{meghanādaś ceti budhaiḥ pañcaśākaḥ prakīrtitaḥ || iti || 26\\!}
-\end{versinnote}
-
-Jyotsnā and Yuktabhavadeva 4.22
-
-\begin{versinnote}
-\tl{sarvaśākam acākṣuṣyaṃ cākṣuṣyaṃ śākapañcakam |\\+}
-\tl{jīvantī-vāstu-matsyākṣī-meghanāda- punarnavāḥ || iti ||\\!}
-\end{versinnote}
-
-It is not entirely clear how we should understand \emph{divya}.
-Brahmānanda glosses it with \emph{nirdoṣa} and takes it with \emph{udaka}.
-But could it refer more specifically to \emph{gaṅgāmbu} (as suggested by Paras) or rain water?
-MW has \emph{divyodaka} n.
-`divine water' i.e.
-rainwater L.
-
-The term \emph{divyodaka} appears in Āyurvedic works (but we’re yet to find a gloss in a commentary). E.g.,
-Aṣṭāṅgahṛdaya 8.42–43
-
-\begin{versinnote}
-\tl{śīlayec chāligodhūmayavaṣaṣṭikajāṅgalam |\\+}
-\tl{suniṣaṇṇakajīvantībālamūlakavāstukam || 42 ||\\+}
-\tl{pathyāmalakamṛdvīkāpaṭolīmudgaśarkarāḥ |\\+}
-\tl{ghṛtadivyodakakṣīrakṣaudradāḍimasaindhavam || 43 ||\\!}
-\end{versinnote}
-
-SriKanta Murty translates \emph{divyodaka} as ‘divyodaka (rain water or pure water)'.
-
-The Rājanighaṇṭu says rainwater:
-
-\begin{versinnote}
-\tl{divyodakaṃ kharāri syād ākāśasalilaṃ tathā |\\+}
-\tl{vyomodakaṃ cāntarikṣajalaṃ ceṣvabhidhāhvayam || Rajni 14.4\\!}
-\end{versinnote}
-
-Kharāri? Maybe \emph{khavāri} was intended.
-
-Vācaspatyam:
-\textbf{divyodaka} na° karma°.
-1 antarīkṣabhave jale divyaśabde bhāva° pra° vākye tadbhedādi dṛśyam.
-ambuśabde vivṛtiḥ .
-
-62*1 is quoted in the Jyotsna as from a medical work (“vaidyake”).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.63}
-\begin{translation}[hp01_063]
-The yogi should eat food that is sweet, delicious, unctuous, contains cow products, nourishes the bodily constituents, is desired by the mind and is appropriate.
-\end{translation}
-
-\begin{sources}[hp01_063]
-\end{sources}
-
-\begin{testimonia}[hp01_063]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{piṣṭaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam |\\+}
-\tl{manobhilaṣitaṃ yogyaṃ yogī bhojanam ācared iti ||\\!}
-\end{versinnote}
-
-YBhD
-
-\begin{versinnote}
-\tl{śreṣṭhaṃ sumadhuraṃ snigdhaṃ gavyaṃ dhātuprapoṣaṇam ||\\+}
-\tl{mano'bhilaṣitaṃ yogyaṃ yogī bhojanamācaret || 23 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_063]
-The variants of 1.63a all seem possible: \emph{mṛṣṭaṃ}, \emph{miṣṭaṃ} and \emph{iṣṭaṃ}. Maybe the last is made redundant by \emph{mano 'bhilaṣitaṃ}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.64}
-\begin{translation}[hp01_064]
-Whether young, old, very old, sick or even weak, the diligent yogi succeeds in all yogas through practice.
-\end{translation}
-
-\begin{sources}[hp01_064]
-DYS 40
-
-\begin{versinnote}
-\tl{yuvāvastho 'pi vṛddho vā vyādhito vā śanaiḥ śanaiḥ |\\+}
-\tl{abhyāsāt siddhim āpnoti yoge sarvo 'py atandritaḥ || 40 ||\\!}
-\end{versinnote}
-
-\end{sources}
-
-\begin{testimonia}[hp01_064]
-Yogacintāmaṇi
-
-\begin{versinnote}
-\tl{haṭhapradīpikāyām—\\+}
-\tl{yuvā bālo 'tivṛddho vā vyādhito durbalo 'pi vā |\\+}
-\tl{abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ ||\\!}
-\end{versinnote}
-
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{yuvā bhavati vṛddho 'pi vyādhito durbalo 'pi vā |\\+}
-\tl{abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ || 1.23 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_064]
-Note the different reading in 164d for V1: \emph{sarvaṃ yogī yatendriyaḥ}.
-\emph{sarvaṃ} is not easy to construe, and the testimony of the DYŚ suggests that \emph{yoge sarvo ’py atandritaḥ} was original.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{1.65}
-\begin{translation}[hp01_065]
-The postures, various breath retentions, and heavenly techniques: the whole practice of Haṭha [is to be done] until Rājayoga results. 
-\end{translation}
-
-\begin{sources}[hp01_065]
-\end{sources}
-
-\begin{testimonia}[hp01_065]
-Haṭharatnāvalī
-
-\begin{versinnote}
-\tl{pīthāni kumbhakāś citrā divyāni karaṇāni ca |\\+}
-\tl{sāṅgo 'pi ca haṭhābhyāso rājayogaphalārthadaḥ || 1.17 ||\\!}
-\end{versinnote}
-
-\end{testimonia}
-
-\begin{philcomm}[hp01_065]
-\end{philcomm}
 

### 2023-12-02 compile-tidyup.sh

@@ -6,22 +6,38 @@ xslcmdproc(){
     xslcmd="java -jar ${HOME}/.nix-profile/saxon9he.jar"
 
     echo "processing xml …"
-    $xslcmd -s:xml/HP1_dev-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp1.html \
+    echo "… hp1"
+    $xslcmd -s:xml/HP1_text-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp1.html \
 	    chapid="hp1" \
-	    transl="../xml/HP1_TranslComm-tei.xml" \
+	    transl="../xml/HP1_comm-tei.xml" \
 	    marma="../xml/Marmasthanas-tei.xml" \
 	    jyotsna="../xml/jyotsna.xml"
-    
-    $xslcmd -s:xml/HP2_edition-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp2.html \
+
+    echo "… hp2"
+    $xslcmd -s:xml/HP2_text-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp2.html \
 	    chapid="hp2" \
-	    transl="../xml/HP2_TranslComm-tei.xml" \
+	    transl="../xml/HP2_comm-tei.xml" \
+	    marma="../xml/Marmasthanas-tei.xml" \
+	    jyotsna="../xml/jyotsna.xml"
+
+    echo "… hp3"
+    $xslcmd -s:xml/HP3_text-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp3.html \
+	    chapid="hp3" \
+	    transl="../xml/HP3_comm-tei.xml" \
+	    marma="../xml/Marmasthanas-tei.xml" \
+	    jyotsna="../xml/jyotsna.xml"
+
+    echo "… hp4"
+    $xslcmd -s:xml/HP4_text-tei.xml -xsl:xslt/hp-online.xsl -o:html/hp4.html \
+	    chapid="hp4" \
+	    transl="../xml/HP4_comm-tei.xml" \
 	    marma="../xml/Marmasthanas-tei.xml" \
 	    jyotsna="../xml/jyotsna.xml"
 
     echo "concatenating html …"
     sed '/<!--content-->/q' html/meta.html > html/hp-online.html 
     
-    for f in "html/hp1.html" "html/hp2.html"
[…]

### 2023-12-07 html/switch.js
added altrec

@@ -1,5 +1,6 @@
 var dev = true;
 var ltn = false;
+var altrec = false;
 
 document.addEventListener("DOMContentLoaded", () => {
 
@@ -7,9 +8,12 @@ document.addEventListener("DOMContentLoaded", () => {
     var ltnbutton = document.getElementById("switchltn");
     var alternchar = document.getElementById("altern");
 
+    var altrecbutton = document.getElementById("switchaltrec");
+
     setbuttons();
+    hide("altrec"); /*entbehrlich durch css?*/
     
-    devbutton.addEventListener('click', () => {
+    devbutton.addEventListener('click', ()=> {
 	if ( dev && !ltn ) {
 	    
 	} else if ( dev ) {
@@ -37,6 +41,18 @@ document.addEventListener("DOMContentLoaded", () => {
 	};
     });
 
+    altrecbutton.addEventListener('click', () => {
+	if ( altrec ) {
+	    hide("altrec");
+	    altrec=false;
+	    setbuttons();
+	} else {
+	    show("altrec");
+	    altrec=true;
+	    setbuttons();
+	};
+    });
+
     function setbuttons() {
 	if ( dev && ltn ) { alternchar.textContent="&";}
 	else { alternchar.textContent="/";};
@@ -46,6 +62,9 @@ document.addEventListener("DOMContentLoaded", () => {
 	
 	if ( ltn ) { ltnbutton.className="toggleon"; }
 	else { ltnbutton.className="toggleoff"; }
+	
+	if ( altrec ) { altrecbutton.className="toggleon"; }
+	else { altrecbutton.className="toggleoff"; }
     }
 
     function hide(scriptclass) {

### 2024-01-29 latex/HP1_comm.tex
added metre-specification

@@
-\input{Preamble-Print}
-\input{CommStyle}
+\input{PreambleComm}
 \begin{document}
@@
 
-\chapter*{Translation \& philological commentary}
+%\chapter*{Translation \& philological commentary}
 
@@
 \begin{translation}[hp01_001]
-Homage to the glorious Ādinātha who taught the system of Haṭhayoga, which shines forth like a stairway for one desirous of climbing to the lofty terrace of the royal palace.%??JM: would much prefer "science" to "system" (vidyā doesn't really mean system).
+Homage to the glorious Ādinātha who taught the science of Haṭhayoga which is like a splendid stairway for one who wants to climb to the lofty royal terrace.%??Add note re meaning of "science".JM: would much prefer "science" to "system" (vidyā doesn't mean system).
 I changed "ladder" to "stairway" (a bit more poetic).
 I'd also prefer "which is like a splendid stairway for one who wants to climb to the lofty royal terrace."
+%JB: I much prefer Jim's alternative translation (i.e., "which is like...).
+But I think we should avoid translating vidyā as `science' because it implies that there was some antecedent to the scientific method in premodern yoga traditions.
+Also, in light of the current politics around yoga in India, "science of Haṭha" fosters the rather absurd view that ancient Indian knowledge systems anticipated scientific discoveries of the 20th century.
+JM: But "system" is wrong.
+
 \end{translation}
@@
 \begin{versinnote}
-\tl{sadādināthāya namo 'stu tubhyaṃ yenopadiṣṭā haṭhayogavidyā |\\+}
-\tl{virājate pronnatarājayogam āroḍhum icchor adhirohiṇīva ||\\!}
+\tl{sadādināthāya namo 'stu tubhyaṃ\\+}
+\tl{yenopadiṣṭā haṭhayogavidyā |\\+}
+\tl{virājate pronnatarājayogam\\+} 
+\tl{āroḍhum icchor adhirohiṇīva ||\\!}
 \end{versinnote}
@@
 \begin{versinnote}
-\tl{ādīśvarāya praṇamāmi tasmai yenopadiṣṭā haṭhayogavidyā |\\+}
-\tl{virājate pronnatarājayogam āroḍhum icchor adhirohiṇīva ||\\!}
+\tl{ādīśvarāya praṇamāmi tasmai\\+} 
+\tl{yenopadiṣṭā haṭhayogavidyā |\\+}
+\tl{virājate pronnatarājayogam\\+} 
+\tl{āroḍhum icchor adhirohiṇīva ||\\!}
 \end{versinnote}
@@
 %?? JM: now saudha is preferred for stemmatic reasons too. I suggest the following shortened form of the note: 
-The alpha witnesses read \emph{virājate pronnatarājayogam} while several others have either \emph{vibhrājate} or \emph{°rājasaudham} or both.
-As explained in Hanneder 2020, \emph{virājate} is stylistically preferable and the comparison between the science of haṭhayoga and a stairway works only with \emph{°rājasaudham}.
+% JB: I prefer the longer note for the digital edition and the shorter one (proposed by Jim) for the printed edition.
+How many readers will have access to Hanneder 2020 and read the whole article?
+The longer note nicely summarises a longer discussion in the article.
+The \textalpha\ witnesses read \emph{virājate pronnatarājayogam} while several others have either \emph{vibhrājate} or \emph{°rājasaudham} or both.
+As explained in Hanneder 2020, \emph{virājate} is stylistically preferable and the comparison between the science of \emph{haṭhayoga} and a stairway works only with \emph{°rājasaudham}.
+
+[Place holder] This reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.
 
@@
 
-The metre of 1.1 is Indravajrā
+?? add comment about yoga replacing mārga and saudha here and in the next two verses.
+
+%The metre of 1.1 is Indravajrā.
 
@@
 
+\begin{metre}[hp01_001]
+Upajāti
+\end{metre}
 %%%%%%%%%%
@@
 \begin{translation}[hp01_002]
-After bowing to the glorious guru, the Lord, the yogi Svātmārāma has taught the system of Haṭhayoga solely for [attaining] Rājayoga.
+After bowing to the glorious guru, the Lord, the yogi Svātmārāma teaches the system of Haṭhayoga solely for [attaining] Rājayoga.
 \end{translation}
@@
 \begin{philcomm}[hp01_002]
-The reading \emph{yogopadiśyate} (\getsiglum{J7,J10,N17,W4}, etc.) is only possible if one accepts that double \emph{sandhi} is a feature of the style of composition, which it is not.
-This reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.%??
-JM has taught?
-surely teaches?
-Again I would prefer science.
+%The reading \emph{yogopadiśyate} (\getsiglum{J7,J10,N17,W4}, etc.) is only possible if one accepts that double \emph{sandhi} is a feature of the style of composition, which it is not.
+This reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.
 \end{philcomm}
@@
 \begin{translation}[hp01_003]
-The compassionate Svātmārāma holds the Lamp on Haṭha for those who are ignorant of the royal path because of wandering in the darkness of many opinions.%??JM: "for those who cannot find the royal path because they are lost in the darkness of a multitude of doctrines".
+ For those who cannot find the royal highway because they are lost in the darkness of many doctrines, the compassionate Svātmārāma holds the Lamp on Haṭha.
 \end{translation}
@@
 While this reading is well attested by the manuscript transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note the readings of P,T,T1 of the critical edition [Gharote 2009: 5 n.
-2] in the testimonia), the reading of a royal path (\emph{rājamārgam}) is consistent with the metaphor of people unable to find% a path in the darkness and, therefore, most probably authorial.
-
-In 1.1d \emph{kṛpākaraḥ} is attested by the most important groups, including alpha, of the available manuscripts.
+2] in the testimonia), the reading \emph{rājamārgam ajānatām}, “for those unable to find the royal highway”, is more appropriate to the metaphor of being lost in darkness.
+
+In 1.3d \emph{kṛpākaraḥ} is attested by the most important groups, including \textalpha, of the available manuscripts.
 The readings \emph{prakāśyate} and \emph{kṣamākaraḥ} are attested by some manuscripts in lower branches of the stemma.
-Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākara} makes good sense.
+Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākaraḥ} makes good sense.
 As Brahmānanda notes, this compound can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}).
@@
 \begin{translation}[hp01_004]
-In fact, Matsyendra, Gorakṣa and other [perfected yogis] knew the system of Haṭha, and the yogi Svātmārāma knows it owing to their favour.%JM For Matsyendra etc. discovered
+In fact, Matsyendra, Gorakṣa, and other [perfected yogis] discovered the science of Haṭha, and the yogi Svātmārāma knows it through their favour.
 \end{translation}
@@
 Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it.
-The variant \emph{mahāyogī} in epsilon one and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}.
+The variant \emph{mahāyogī} in \epsilonOne and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}.
 One could emend to \emph{tathā} in light of the attested reading \emph{yathā} (\getsiglum{C7}) but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.
@@
 \begin{translation}[hp01_005]
-The glorious Ādinātha, Matysendra, Śābara, Ānandabhairava, Cauraṅgī, Mīna, Gorakṣa, Virūpākṣa, Bileśaya,
+The glorious Ādinātha, Matsyendra, Śābara, Ānandabhairava, Cauraṅgī, Mīna, Gorakṣa, Virūpākṣa, Bileśaya,
 \end{translation}
@@
 
-Two manuscripts of the alpha and delta groups have the variant reading \emph{°virūpākṣaḥ savālikaḥ} (\getsiglum{J5,V19}) for \emph{°virūpākṣabileśayāḥ}.
-In another alpha manuscript, \getsiglum{N3}, \emph{savālikaḥ} was corrected to \emph{savālmikaḥ}, perhaps in an effort to restore a name similar to Vālmīki, the celebrated author of the \textit{Rāmāyaṇa}.
+Two manuscripts of the \textalpha\ and \textdelta\ groups have the variant reading \emph{°virūpākṣaḥ savālikaḥ} (\getsiglum{J5,V19}) for \emph{°virūpākṣabileśayāḥ}.
+In another \textalpha\ manuscript, \getsiglum{N3}, \emph{savālikaḥ} was corrected to \emph{savālmikaḥ}, perhaps in an effort to restore a name similar to Vālmīki, the celebrated author of the \textit{Rāmāyaṇa}.
 \end{philcomm}
@@
 \begin{translation}[hp01_006]
-Manthānabhairava, Siddhabuddha, and Kanthaḍi, Goraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
+Manthānabhairava, Siddhabuddha, and Kanthaḍi, Goraṇṭaka, Surānanda, Siddhapāda and Carpaṭi,
 \end{translation}
@@
 \begin{philcomm}[hp01_006]        
-The alpha manuscripts have \emph{goraṇṭaka}, and several other manuscript groups have \emph{pauraṇṭaka}.
+The \textalpha\ manuscripts have \emph{goraṇṭaka}, and several other manuscript groups have \emph{pauraṇṭaka}.
 We are yet to find the name \emph{goraṇṭaka} in other Sanskrit texts but it may be a Sanskrit rendering of \emph{Goraṇṭakuḍu}, which is the name of a disciple of Gorakṣanātha in the \emph{Navanāthacaritramu} (Jones 2017: 194 n.3).
@@
 
-The compound \emph{siddhapāda} could be a respectful affix. However, it seems unlikely here because it would cross the \emph{pāda} break.
+The compound \emph{siddhapāda} could be a respectful affix. However, it seems unlikely here because the name would cross the \emph{pāda} break.
 
@@
 
-The alpha group supports \emph{kākacaṇḍīśvaro gayaḥ} but we have not been able to find evidence for a Siddha called Gaya.
+The \textalpha\ group supports \emph{kākacaṇḍīśvaro gayaḥ} but we have not been able to find evidence for a Siddha called Gaya.
 \end{philcomm}
@@
 \begin{philcomm}[hp01_008]  
-The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literaure) is largely transmitted as \emph{allamaḥ prabhu\-devaḥ}, as though it were two names, although some manuscripts of the \emph{Haṭha\-pradīpikā} have \emph{allamaprabhudevaś ca} (i.e., \getsiglum{V3,V8,V13,V16,V22,N24,N26,Jyo}) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n.
+The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literature) is largely transmitted as \emph{allamaḥ prabhu\-devaḥ}, as though it were two names, although some manuscripts of the \emph{Haṭha\-pradīpikā} have \emph{allamaprabhudevaś ca} (i.e., \getsiglum{V3,V8,V13,V16,V22,N24,N26,Jyo}) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n.
 8).
@@
 It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika.
-Alternatively, it could simply refer to one who used a broken skull as a bowl.
+Alternatively, it could simply refer to one who uses a broken skull as a bowl.
 \end{philcomm}
@@
 \begin{translation}[hp01_009]
-Having destroyed the rod of death through the power of Haṭhayoga, these great perfected yogis and others wander in the world.
+These and other great adepts used the power of hathayoga to smash the rod of death and [so] are roaming the worlds.
 \end{translation}
@@
 \end{philcomm}
+
+\begin{metre}[hp01_009]
+Anuṣṭubh (c: ra-vipulā)
+\end{metre}
 
@@
 \begin{philcomm}[hp01_010] 
-The compound \emph{saṃsāratāpa°} is well attested and found elsewhere (e.g., \emph{Viṣṇupurāṇa} 6.7.62, \emph{Agnipurāṇa} 371.1, \emph{Haṭhatattvakaumudī} 38.92, \emph{Haṭhābhyāsapaddhati} ms.
+%The compound \emph{saṃsāratāpa°} is well attested and found elsewhere (e.g., \emph{Viṣṇupurāṇa} 6.7.62, \emph{Agnipurāṇa} 371.1, \emph{Haṭhatattvakaumudī} 38.92, \emph{Haṭhābhyāsapaddhati} ms.
 46/440, f.
@@
 
-The reading of \emph{samāśrayo} in eta one is metrically faulty.
-
-The word \emph{°jagatāṃ} is probably authorial because it makes good sense with  \emph{ādhārakamaṭha} in light of the cosmological notion that the tortoise supports all the worlds.
-However, this reading may not have been understood by some and was changed in the vulgate and other witnesses to \emph{°yuktānāṃ} instead.
-\end{philcomm}
+%The reading of \emph{samāśrayo} in \etaOne\ is metrically faulty.
+
+Both \emph{°jagatāṃ} and \emph{°yuktānāṃ} are well attested by the collated manuscripts, and the \textalpha\ group omits the second half of this verse.
+We have adopted \emph{°jagatāṃ}  because it makes good sense with  \emph{ādhārakamaṭhah} in light of the cosmological notion that the tortoise supports all the worlds.
+This reading may not have been understood by some and was changed in the vulgate and other witnesses to \emph{°yuktānāṃ} instead.
+\end{philcomm}
+
+\begin{metre}[hp01_010]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
 
@@
 \begin{translation}[hp01_011]
-The doctrine of Haṭha should be kept completely secret by those yogis who are desiring success. When it is secret it becomes potent. However, when it has been revealed, it becomes impotent.
+The science of Haṭha should be kept completely secret by yogis who want success. It becomes potent when kept secret but impotent when revealed.
 \end{translation}
@@
 \begin{philcomm}[hp01_011]        
-Either the singular or plural of yogin could be read here. The singular is well attested among the testimonia. However, the weight of the manuscript evidence favours the plural.  
+Either the singular or plural of yogin could be read here. The singular is well attested among the testimonia, but the manuscript transmission favours the plural.  
 \end{philcomm}
@@
 \begin{translation}[hp01_012]
-In a well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live in an isolated hut.
+In a well-ruled, righteous region, with plenty of food and free from upheaval, the Haṭhayogi should live in an isolated hut.%JM changed disturbance to upheaval
 \end{translation}
@@
 In several other stories (\textit{Kathāsaritsāgara} 6.6.132, 10.5.89, 12.25.35), \emph{maṭhikā} is the term used for the hut of an ascetic.
-In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four fore-arm lenghts (\emph{hasta}) high and wide, and it can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).
+In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four fore-arm lengths (\emph{hasta}) high and wide, and it can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).
 
@@
 It stipulates that the hut should be built in a place measuring up to a bow length and free from rocks, fire and water.
-None of the early manuscripts have this hemistich which suggests that it was added at a later time.
+None of the early manuscripts has this hemistich, which suggests that it was added at a later time.
 Nonetheless, it appears in over a dozen manuscripts that were consulted for this edition.
@@
 \begin{translation}[hp01_013]
-It has a small door and is without cracks, holes and potsherds.
-It extends not too high or low, and is thickly smeared with cow dung in the proper way.
-It is clean, free from everything that annoys, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
+It has a small door and is without cracks, holes and bumps.
+It is neither too high nor too low in extent and is thickly smeared with cow dung in the proper way.
+It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
 \end{translation}
@@
 The manuscripts preserve many different readings at the end of the compound beginning with \emph{arandhragarta°}.
-The reading \emph{°piṭharaṃ} has been suggested by Dominic Goodall in the sense of potsherds.
-The idea behind this reading is that hut should be free of rubbish, such as potsherds.
+We have adopted \emph{piṭaka}, which usually means ``a basket'' but can also mean ``a boil or blister," because it is well attested and might here refer to bumps on the floors or walls that would make them uneven.
+Another possibility is \emph{°piṭharaṃ}, which can have the sense of potsherds and would here mean that the hut should be free of rubbish on the floor.
 One would expect a word for a defect in a hut that is similar to, but not the same as, cracks (\emph{randhra}) and holes (\emph{garta}).
 For this reason, the reading \emph{°vivaraṃ} looks like a patch, as its meaning does not add anything to \emph{°randhragarta°}.
-The reading \emph{°viṭapaṃ} (`the young branch of a tree or creeper') is attested among the witnesses of the \emph{Haṭhapradīpikā}.
-However, it is difficult to construe in this context unless it was intended to refer to creepers or branches that might invade or encroach upon the hut.
-
-(MD: Stemmatically, \emph{piṭaka} ``a basket; a boil" is well attested.
+The reading \emph{°viṭapaṃ} (`the young branch of a tree or creeper') attested in some manuscripts of the \emph{Haṭhapradīpikā} is difficult to construe in this context unless it was intended to refer to creepers or branches that might invade or encroach upon the hut.
+
+%??(MD: Stemmatically, \emph{piṭaka} ``a basket; a boil" is well attested.
 Is it perhaps possible to interpret the word as ``a bump"?
-\emph{a-randhra-garta-piṭaka} would then mean that the floor or walls of the hut are not uneven.)
-
-Manuscripts of several groups, namely beta, epsilon and eta, have \emph{°bādhojjitaṃ}, whereas delta and the \emph{Yogacintāmaṇi} have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’), which is somewhat suggested by the alpha reading (\emph{jyaṃtyūpsitaṃ} \getsiglum{N3}, \emph{jaṃtūṣṇitaṃ} \getsiglum{J5}, but \emph{bodhojhitaṃ} \getsiglum{G4}).
-However, \emph{°bādhojjitaṃ} may have been original because it is a rather unusual descriptive compound for a hut, and yet a hut is described similarly in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
-
-\end{philcomm}
+\emph{a-randhra-garta-piṭaka} would then mean that the floor or walls of the hut are not uneven.) JM: yes, sounds good to me.
+%JB Okay I've written the note to reflect this.
+
+Manuscripts of several groups, namely \textbeta, \textepsilon\ and \texteta, have \emph{°bādhojjitaṃ}, whereas \textdelta\ and the \emph{Yogacintāmaṇi} have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’).
+The \textalpha\ group is split on this, with \getsiglum{G4} (\emph{bodhojhitaṃ}) closer to \emph{°bādhojjitaṃ} and \getsiglum{N3} (\emph{jyaṃtyūpsitaṃ} and \getsiglum{J5} (\emph{jaṃtūṣṇitaṃ}) closer to \emph{°jantūjjhitaṃ}.
+We have adopted the more unusual reading of \emph{°bādhojjitaṃ}  with the support of a similar description of a hut in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
+
+\end{philcomm}
+
+\begin{metre}[hp01_013]
+Śārdūlavikrīḍita
+\end{metre}
 
@@
 \begin{translation}[hp01_014]
-Locating oneself in a hut of such a kind, free from all worry, [the yogi] should practise only yoga, in the way taught by his guru.
+Staying in such a hut, free from all worry, in the way taught by his guru [the yogi] should practise only yoga.
 \end{translation}
@@
 \begin{translation}[hp01_015]
-Overeating, exertion, idle chatter, not sticking to rules, socialising, sensuality: through [these] six, yoga is lost.
+Overeating, exertion, idle chatter, not sticking to rules, socialising and sensuality: through [these] six, yoga is lost.
 \end{translation}
@@
 \begin{versinnote}
-\tl{atyāhāraḥ prayāsaśca prajalpo niyamāgrahaḥ |\\+}
+\tl{atyāhāraḥ prayāsaś ca prajalpo niyamāgrahaḥ |\\+}
 \tl{janasaṃgaś ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||\\!}
@@
 \begin{philcomm}[hp01_015]        
-It is impossible to be certain about the meaning of \emph{niyamagraha}, as the manuscripts do not indicate whether an \emph{avagraha} (i.e., \emph{prajalpo ’niyamagraha}) was intended.
+Since many scribes do not use an \emph{avagraha} we cannot be sure whether to understand  \emph{niyamagrahaḥ} in \emph{pāda} b as having a negative prefix.
 Although \emph{yama} and \emph{niyama} are not included in the \emph{Haṭhapradīpikā} as auxiliaries of Haṭhayoga, verse 2.14 implies that \emph{niyama} is necessary at least in the early stages of establishing a practice.
-Furthermore, verse 3.82 suggests that a yogi who does not practice \emph{niyama} might obtain success in yoga through the practice of \textit{vajroli}.
+Furthermore, verse 3.82 suggests that a yogi who does not practice \emph{niyama} might obtain success in yoga through the practice of \textit{vajrolī}.
 Ambiguity over the role of \emph{yama} and \emph{niyama} in Haṭhayoga may explain why two verses on ten \emph{yama}s and ten \emph{niyama}s were inserted in some manuscripts after the next verse (1.16).
@@
 \begin{translation}[hp01_016]
-Yoga is successful as a result of six [qualities]: zeal, daring, resolve, gnosis of the truth, conviction, and avoiding contact with people.
+Zeal, daring, resolve, gnosis of the truth, conviction and avoiding contact with people: by means of six [virtues], yoga is successful.
 \end{translation}
@@
 \begin{philcomm}[hp01_016]    
-Alpha and several other groups of manuscripts have \emph{tattvajñānāc ca darśanāt} or something very similar in the second quarter of the verse.
-The problem with this reading is the meaning of \emph{darśana} by itself (i.e., discernment of what?).
+\textalpha\ and several other groups of manuscripts have \emph{tattvajñānāc ca darśanāt} or something very similar in the second \emph{pāda} of the verse, but \emph{darśana} by itself is problematic: a vision of what?
 The early sources of this verse, in particular the \emph{Śivadharmottara}, indicate that the second verse quarter read as \emph{santoṣāt tattvadarśanāt}, which makes much better sense of the word \emph{darśana} (i.e., ‘seeing the truth’).
-However, It seems likely that before the time of Svātmārāma other versions of this verse were circulating, in which \emph{santoṣāt} had been dropped, \emph{niścayāt} moved from the first to second verse quarter, \emph{tattvadarśanāt} became \emph{tattvajñānāt} and \emph{sāhasāt} was introduced.
+However it seems likely that before the time of Svātmārāma other versions of this verse were circulating in which \emph{santoṣāt} was not found, \emph{niścayāt} had moved from the first to second verse quarter, \emph{tattvadarśanāt} had become \emph{tattvajñānāt} and \emph{sāhasāt} was introduced.
 It should also be noted that the word \emph{tattva} could have a more specific meaning in the \emph{Haṭhapradīpikā} (4.45–46) as Svātmārāma states that it is a synonym of \emph{samādhi}.
-In other yoga texts, it can sometimes refer to the practices of yoga (e.g., \emph{tritattva} in \emph{Amṛtasiddhi} 13.12, 14.2--3) or, more generally, to the highest reality or truth (e.g., \emph{Amanaska} 1.2, 1.20–21, 2.17, etc.).
+In other yoga texts, it can sometimes refer to the practices of yoga (e.g.
+\emph{tritattva} in \emph{Amṛtasiddhi} 13.12, 14.2--3) or, more generally, to the highest reality or truth (e.g.
+\emph{Amanaska} 1.2, 1.20–21, 2.17, etc.).
 \end{philcomm}
@@
 \begin{philcomm}[hp01_017]        
-The reading of \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the main one of the alpha group.
+The reading of \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the main one of the \textalpha\ group.
 Although this compound rarely appears in other yoga texts, a similar term \emph{śarīrapāṭava} occurs in the \emph{Śivasaṃhitā} (2.35) as one of the benefits bestowed by digestive fire (\emph{vaiśvānarāgni}), which indicates that the word \emph{pāṭava} was used in relation to the body and the benefits of yoga.
 The compound \emph{aṅgapāṭava} seems to imply the optimal functioning of the body.
-However, the alternative reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts and occurs even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}).
-Therefore, it is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by manuscripts in several early groups (i.e., beta, gamma and delta).
+The variant reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts and similar formulations occur even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}).
+It is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by manuscripts in several early groups (i.e., \textbeta, \textgamma\ and \textdelta).
 \end{philcomm}
@@
 
+\begin{metre}[hp01_018]
+Anuṣṭubh (a: na-vipulā; c: ra-vipulā)
+\end{metre}
+
 %%%%%%%%%%
@@
 \begin{translation}[hp01_019]
-Correctly placing the soles of both feet between the knees and thighs [and] sitting up with the body straight: they call that \emph{svastikāsana}.
+Placing the soles of both feet well between the knees and thighs [and] sitting up with the body straight: they call that \emph{svastikāsana}.%JM: correctly is tricky.
+It could mean thoroughly, completely — I've looked at some other usages of samyak and that understanding seems to work well quite often, as it does here.
+E.g.
+VS 3.37ab hastābhyāṃ bandhayet samyak karṇādikaraṇāni ca, VM 58cd rasānāṃ śoṣanaṃ samyak mahāmudrābhidhīyate.
+Or Brahmānanda ad 1.35 susthiram samyak sthiraṃ
+% JB thoroughly doesn't make sense here. And is it possible to 'completely' place the feet between the knees and thighs?
 \end{translation}
@@
 \begin{philcomm}[hp01_020]        
-This posture first appears in some Vaiṣṇava \emph{Saṃhitā}s that predate the \emph{Haṭhapradīpikā}, namely, the \emph{Ahirbudhnyasaṃhitā} and the \emph{Vasiṣṭhasaṃhitā}.
+This posture first appears in some Vaiṣṇava \emph{Saṃhitā}s that predate the \emph{Haṭhapradīpikā}, including the \emph{Ahirbudhnyasaṃhitā} and the \emph{Vasiṣṭhasaṃhitā}, which is likely to have been the source of this verse.
 The position of the ankles is the same in all the source texts.
-However, an interesting feature of the description in the \emph{Ahirbudhnyasaṃhitā} is the position of the hands, which are crossed behind the back and hold the big toes.
-The \emph{Haṭhapradīpikā}’s description, which derives from the \emph{Vasiṣṭhasaṃhitā}, does not mention the position of the hands.
+The \emph{Ahirbudhnyasaṃhitā} adds that the hands are crossed behind the back and hold the big toes.
 For illustrations of six possible positions of the arms and hands, see Gharote, Jha, Devnath, Sakhalkar 2006: 111–113.
@@
 \begin{translation}[hp01_021]
-By fixing one foot on one thigh and placing the [other] thigh on the other foot, \emph{vīrāsana} is taught [to arise].
+ Fixing one foot on one thigh and placing the [other] thigh on the other foot is called \emph{vīrāsana}.
 \end{translation}
@@
 \begin{philcomm}[hp01_021]   
-Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by \getsiglum{G4} (alpha group) and \getsiglum{V1} (eta group), the sources and the testimonia.
+Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by \getsiglum{G4} (\textalpha\ group) and \getsiglum{V1} (\texteta\ group), the sources and the testimonia.
 It appears to be verse filler here rather than indicating a temporal sequence of actions.
@@
 \begin{translation}[hp01_022]
-Knowers of yoga know that \emph{kūrmāsana} arises by attentively blocking the anus with turned-out ankles.
+Knowers of yoga know that \emph{kūrmāsana} arises by carefully blocking the anus with the ankles crossed.%??Tweak note to reflect change to crossed. 
 \end{translation}
@@
 In the first quarter of the verse, the witnesses are split between \emph{nirudhya} (‘having blocked’), \emph{nibadhya} (‘having bound’), \emph{niyamya} (‘having restrained’) and \emph{niṣpīḍya} (‘having pressed’).
-The source, the \emph{Vasiṣṭhasaṃhitā}, and two manuscripts of the beta and gamma groups support \emph{nirudhya} whereas one alpha manuscript (\getsiglum{J5}) and the testimonia support \emph{niyamya} and another alpha manuscript (\getsiglum{G4}) has \emph{niṣpīḍya}.
+The source, the \emph{Vasiṣṭhasaṃhitā}, and two manuscripts of the \textbeta\ and \textgamma\ groups support \emph{nirudhya} whereas one \textalpha\ manuscript (\getsiglum{J5}) and the testimonia support \emph{niyamya} and another \textalpha\ manuscript (\getsiglum{G4}) has \emph{niṣpīḍya}.
 In terms of blocking or closing the anus by sitting on the ankles, \emph{nirudhya} makes good sense, and \emph{niṣpīḍya} (`having pressed the anus with both ankles') is also possible.
 
-The word \emph{vyutkrameṇa} appears to describe the position of the ankles.
+??The word \emph{vyutkrameṇa} appears to describe the position of the ankles.
 Its basic meaning is ‘against the normal direction,’ which would suggest that the ankles are turned out or crossed rather than placed together naturally.
@@
 \begin{translation}[hp01_023]
-[The yogi] adeptly assumes \emph{padmāsana}, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is \emph{kukkuṭāsana}.
+[The yogi] correctly assumes \emph{padmāsana}, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is \emph{kukkuṭāsana}.
 \end{translation}
@@
 
+\begin{metre}[hp01_023]
+Anuṣṭubh (c: ma-vipulā)
+\end{metre}
+
 %%%%%%%%%%
@@
 \tl{kukkuṭāsanabandhastho dorbhyāṃ sambadhya kandharām |\\+}
-\tl{śete kūrmavad uttānam etad uttānakūrmakam || iti uttānakūrmāsanam ||\\!}
+\tl{śete kūrmavad uttānam etad uttānakūrmakam ||\\+}
+\tl{iti uttānakūrmāsanam ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_024]
-The oldest dated manuscript, eta one, has \emph{kukkuṭāsanavat kṛtvā}, which is a simpler alternative to the widely attested reading \emph{kukkuṭāsanabandhasthaḥ}, which we have accepted.
-Since there is no known source for this verse other than the \emph{Haṭhapradīpikā}, we can only assume that the reading of eta one was an isolated attempt to simplify the syntax, for it is a singular reading that appears to be unrelated to the other variants.
+The oldest dated manuscript, \etaOne, has \emph{kukkuṭāsanavat kṛtvā}, which is a simpler alternative to the widely attested reading \emph{kukkuṭāsanabandhasthaḥ}, which we have accepted.
+Since there is no known source for this verse other than the \emph{Haṭhapradīpikā}, we can only assume that the reading of \etaOne\ was an isolated attempt to simplify the syntax.%??JM: I guess we discussed this before, but are we not going to translate āsana names?
+%JB if we do, we should put the Sanskrit in round brackets as many enthusiasts use the Sanskrit names.
 \end{philcomm}
@@
 \begin{translation}[hp01_025]
-Clasping the big toes with hands and performing the action of drawing a bow as far as the ear is called \emph{dhanurāsana}.
+Clasping the big toes with the hands and performing the action of drawing a bow as far as the ear is called \emph{dhanurāsana}.
 \end{translation}
@@
 \tl{pādāṅguṣṭhau tu pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\+}
-\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam || iti dhanurāsanam ||\\!}
+\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam ||\\+}
+\tl{iti dhanurāsanam ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_025]
-Since the word \emph{ākarṣaṇa} in one form or other is so well attested in the third verse quarter, the reading \emph{dhanurākarṣaṇaṃ kṛtvā}, which is in epsilon two, as well as the principal testimonia (i.e., the \emph{Yogacintāmaṇi} and \emph{Haṭharatnāvalī}), fits the overall syntax of the verse.
+Since the word \emph{ākarṣaṇa} in one form or other is so well attested in the third verse quarter, the reading \emph{dhanurākarṣaṇaṃ kṛtvā}, which is in \epsilonTwo, as well as the principal testimonia (i.e., the \emph{Yogacintāmaṇi} and \emph{Haṭharatnāvalī}), fits the overall syntax of the verse.
 However, it is curious that \emph{kṛṣṭaṃ} (for \emph{kṛtvā}) is well attested in some groups of manuscripts because it seems redundant with \emph{ākarṣaṇaṃ}.
@@
 2v), which was consulted but not collated for this edition, has a scribal comment stating that \emph{dhanurāsana} should be done continuously (\emph{anavarata}) on the left and right sides (\emph{tatra ekam dhanurākarṣaṇāsanam āsanaṃ savyāpasavyapādahastābhyām [abhy]ased anavaratam}).
-This would make \emph{dhanurāsana} a moving practice as shown in \medialink{this video}{dhanurasana.mp4}.
+This would make \emph{dhanurāsana} a dynamic practice as shown in \medialink{this video}{dhanurasana.mp4}.
 \end{philcomm}
@@
 \begin{translation}[hp01_026]
-Having grasped the right foot, which is placed at the base of the left thigh, with the right arm wrapped around the outside of the knee, [the yogi] remains with his body twisted.
-This \emph{āsana} was taught by Matsyendranātha.%??JM: [The yogi] should hold the right foot, which is placed at the base of the left thigh, with the [hand of] the right arm, which is wrapped around the outside of the knee, and remain [like that] with his body twisted.
+[The yogi] should hold the right foot, which is placed at the base of the left thigh, with the [hand of] the right arm, which is wrapped around the outside of the knee, and remain [like that] with his body twisted.
 This posture was taught by Matsyendranātha.
@@
 
-One manuscript of alpha (\getsiglum{G4}) and most manuscripts of the \textit{Haṭharatnāvalī} have \emph{°dakṣapādo} in the first verse quarter.
-This reading yields the same meaning as the adopted one if read with \emph{°vāmapādaṃ} in the second verse quarter.
+One manuscript of \textalpha\ (\getsiglum{G4}) and most manuscripts of the \textit{Haṭharatnāvalī} have \emph{°dakṣapādo} in the first \emph{pāda}.
+This reading yields the same meaning as the adopted one if read with \emph{°vāmapādaṃ} in the second.
 However, \getsiglum{G4} and manuscripts of the \textit{Haṭharatnāvalī} read \emph{°vāmadoṣṇā}, which is not good because it leaves the gerund without an object.
 
-The version of this verse in \emph{Jyotsnā} (1.26), which is supported by some manuscripts in two important groups, beta and eta, has two objects of the gerund, namely the left and right feet, without an instrumental or conjuctive particle.
+The version of this verse in \emph{Jyotsnā} (1.26), which is supported by some manuscripts in two important groups, \textbeta\ and \texteta, has two objects of the gerund, namely the left and right feet, without an instrumental or conjuctive particle.
 In his commentarial remarks, Brahmānanda proposes that the left foot is grasped by the right hand and the right foot by the left foot, as seen in \medialink{Figure 2}{matsyendrasana2.jpg}.
 \end{philcomm}
+
+\begin{metre}[hp01_026]
+Upajāti
+\end{metre}
 
@@
 \begin{translation}[hp01_027]
-Matsyendra's seat is a destructive weapon for the many terrible diseases that develop in the stomach; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
+Matsyendra's seat is a destructive missile for the many terrible diseases that develop in the stomach; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
 \end{translation}
@@
 The manuscript readings for the compound beginning with \emph{jaṭhara} diverge significantly and include \emph{jaṭharapravṛddha°}, \emph{jaṭharaprabuddha°}, \emph{jaṭharapradīpta°} and \emph{jaṭharapracaṇḍa°}.
-As descriptive compounds, none of these make good sense in regard to Matsyendra's seat.
-Since the stomach or abdomen (\emph{jaṭhara}) is the first member of this compound, it seems more likely that it qualifies the terrible diseases (\emph{pracaṇḍarug}) that are mentioned in the next verse quarter, as suggested by the reading \emph{jaṭharapravṛddha°}, which is attested by eta two and the \textit{Yogacintāmaṇi} and suggested by gamma one and delta one (\emph{jaṭharapravuddh°}).
+As descriptive compounds, none of these makes good sense in regard to Matsyendra's seat.
+Since the stomach or abdomen (\emph{jaṭhara}) is the first member of this compound, it seems more likely that it qualifies the terrible diseases (\emph{pracaṇḍarug}) that are mentioned in the next verse quarter, as suggested by the reading \emph{jaṭharapravṛddha°}, which is attested by \etaTwo\ and the \textit{Yogacintāmaṇi} and suggested by \gammaOne and \deltaOne (\emph{jaṭharapravuddh°}).
 
@@
 However, the \emph{Jyotsnā} (1.27d) has \emph{candrasthiratvaṃ} (`steadiness of the moon'), and this reading is well-attested in many manuscripts that are lower on the stemma.
-Brahmānanda understands steadiness here as `the absence of flow' (\emph{sthiratvaṃ kṣaraṇābhāvaṃ}), a reference no doubt to the moon retaining its nectar.
-\end{philcomm}
+Brahmānanda understands steadiness here as `the absence of flow' (\emph{sthiratvaṃ kṣaraṇābhāvaṃ}), a reference to the moon retaining its nectar.
+\end{philcomm}
+
+\begin{metre}[hp01_027]
+Upajāti
+\end{metre}
 
@@
 \begin{translation}[hp01_028]
-[The yogi] should stretch out both feet on the ground like staffs, hold the ends of both feet with the hands, place the forehead upon the knees and remain thus.
-They call this the back-stretch (\emph{paścimatānam}).
+[The yogi] should stretch out the legs on the ground [as straight] as sticks, hold the toes of both feet with the hands, place the forehead upon the knees and remain thus.
+They call this the back-stretch (\emph{paścimatānam}).%??
+āsana name translated here.
 \end{translation}
@@
 The reading \emph{dorbhyāṃ padāgradvitayaṃ} is well attested but is somewhat strange because \emph{dos} usually means `the arm' rather than the hands.
-Hence, the other variation `with both hands' (\emph{dvābhyāṃ karābhyāṃ dvitayaṃ}), which appears to be an attempt to remove \emph{dorbhyāṃ}, is inferior because the thing the hands are holding, that is the toes (\emph{padāgra}), is not explicitly stated in this version of the verse.
-\end{philcomm}
+The variant \emph{dvābhyāṃ karābhyāṃ dvitayaṃ}, “with both hands”,  appears to be an attempt to remove \emph{dorbhyāṃ}, but it introduces the problem of the toes not being mentioned.
+\end{philcomm}
+
+\begin{metre}[hp01_028]
+Upajāti
+\end{metre}
 
@@
 This back-stretch is the foremost among \emph{āsana}s.
-It makes the breath flow in the back (i.e., the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
+It makes the breath flow to the rear (i.e.~in the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
 \end{translation}
@@
 \begin{sources}[hp01_029]
-Cf. \emph{Śivasaṃhitā} 3.109
+Cf. \emph{Śivasaṃhitā} 3.109–110
 
@@
 \tl{āsanāgryam idaṃ proktaṃ jaṭharānaladīpanam |\\+}
-\tl{dehāvasādaharaṇaṃ paścimottānasaṃjñakam ||\\!}
+\tl{dehāvasādaharaṇaṃ paścimottānasaṃjñakam ||\\+}
+\tl{ya etad āsanaṃ śreṣṭhaṃ pratyahaṃ sādhayet sudhīḥ |\\+}
+\tl{vāyuḥ paścimamārgeṇa tasya saṃcarati dhruvam ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_029]
- The use of the word \emph{paścima} to mean the central channel is found in \emph{Yogabīja} 121.
-The  \emph{Yogabīja} (95) also refers to the path of the central channel (\emph{paścima\-mārga}), and this understanding of \emph{paścima} is found in the \emph{Jyotsnā} 1.29: \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumnāmārgeṇa vahatīti paścimavāhī}.
-\end{philcomm}
+ The use of the word \emph{paścima} to mean the central channel is found at \emph{Yogabīja} 95 (\emph{paścimamārgataḥ}), 108 (\emph{paścime pathi}), 117 (\emph{paścimadvāramārgeṇa}) and 121 (\emph{paścimaṃ}).
+Cf.~the usages of \emph{paścima\-mārga} in \sl{Dattātreyayogaśāstra} 140, \emph{Yogabīja} 95 and \emph{Śivasaṃhitā} 3.110 (from which this verse is likely to be derived).
+Brahmānanda understands \emph{paścima} as referring to the Suṣumṇā (\emph{Jyotsnā} 1.29): \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumṇāmārgeṇa vahatīti paścimavāhī}.
+\end{philcomm}
+
+\begin{metre}[hp01_029]
+Śiśulīlā 
+\end{metre}
 
@@
 \begin{translation}[hp01_030]
-Supporting oneself on the ground with both hands, [one's] elbows are placed on either side of the navel [and], with a raised position (\emph{uccāsanaḥ}), one is lifted up into the air [as straight] as a stick.
-They call this posture the peacock.%JM: Supporting oneself on the ground with both hands, the elbows placed on either side of the navel, lifted up into the air in a raised posture [as straight] as a stick: they call this posture the peacock.
+Supporting oneself on the ground with both hands, the elbows placed on either side of the navel, lifted up into the air in a raised posture [as straight] as a stick: they call this posture the peacock.
 
@@
 \begin{philcomm}[hp01_030]
-The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77).
-The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}, which may be derived from earlier Vaiṣṇava sources, such as the \textit{Pādmasaṃhitā}.
+%??JM: authorial reworking? 
+The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77), which are themselves derived from earlier Vaiṣṇava sources.
+The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}.
 
@@
 
+\begin{metre}[hp01_030]
+Upajāti
+\end{metre}
+
 %%%%%%%%%%
@@
 \begin{translation}[hp01_031]
-The glorious peacock posture gets rid of all diseases of the abdomen, such as bloating, and overcomes humoral imbalances.
-It completely incinerates food which is bad or has been eaten to excess, it generates digestive fire and it digests strong poison.
+The glorious peacock [posture] quickly gets rid of bloating and all other diseases of the abdomen, and overcomes humoral imbalances.
+It reduces to ashes food which is bad or has been eaten to excess, kindles the digestive fire and causes strong poison to be digested.
 \end{translation}
@@
 
+\begin{metre}[hp01_031]
+Mālinī 
+\end{metre}
+
 %%%%%%%%%%
@@
 
+\begin{metre}[hp01_032]
+Anuṣṭubh (c: bha-vipulā)
+\end{metre}
+
 %%%%%%%%%%
@@
 \begin{translation}[hp01_033]
-Śiva has taught eighty-four \emph{āsana}s. I shall take the four best and describe them.
+Śiva has taught eighty-four \emph{āsana}s. I shall take the four best from them and describe them.
 \end{translation}
@@
 In the first and third verse quarters, Svātmārāma may have rewritten \emph{Śivasaṃhitā} 3.96 to include the information that it was Śiva (\emph{śivena}) who taught the eighty-four \emph{āsana}s, whereas in the source Śiva is himself speaking.
-By doing this, Svātmārāma changes the meaning of the second half of the verse, as the \emph{Śivasaṃhitā} states that Śiva picked out the four best postures and taught them, whereas in the \emph{Haṭhapradīpikā} it reads as though Svātmārāma himself is responsible for picking out the four best postures and teaching them.
+Svātmārāma also changes the meaning of the second half of the verse, as the \emph{Śivasaṃhitā} states that Śiva picked out the four best postures and taught them, whereas in the \emph{Haṭhapradīpikā} it reads as though Svātmārāma himself is responsible for picking out the four best postures and teaching them.
 There are other instances in the \emph{Haṭhapradīpikā} where Svātmārāma borrows a verse with a first person verb (e.g., 3.43, 4.2).
@@
 \end{philcomm}
+
+\begin{metre}[hp01_033]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
 
@@
 
+\begin{metre}[hp01_034]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
+
 %%%%%%%%%%
@@
 \begin{translation}[hp01_035]
-Now, the adept's pose (\emph{siddhāsana}).\\ 
-Having joined the place of the perineum with the heel of the foot, the yogi should firmly fix the [other] foot on the penis.
-Having held the face and chest together and the body erect, [the yogi] remains still, his senses restrained, gazing between the brows with his eyes unmoving.
-This breaks open the door to liberation and is called the adept’s pose.
+Now, the adept's pose (\emph{siddhāsana}).
+
+[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, hold the face and chest together and the body erect, and [remain] stock-still, his senses restrained, gazing between the brows with his eyes unmoving.
+This, which breaks open the door to liberation, is called the adept’s pose.
 \end{translation}
@@
 \tl{\var{7a °mūlaghaṭitaṃ ] °mulaghaṭanaṃ A • vinyasen ] GHT; vinyase VA, vinyaset Y}\\+}
-\tl{\var{7b meḍhre ] meṃḍhre A • athaikam eva hṛdaye ] H; athaikadeśahṛdayo V, athaikam eva niyataṃ AGBGPkHvl, athaikam eva niṣataṃ GL, athaikam eva hṛdayaṃ GP, athaikam āsyahṛdaye HSHvl, athaikaṃ ekahṛdayo T, athaikam ekahṛdayaḥ Y • dhṛtvā ] VGBTHS; kṛtvā HGLGPGPkY • hanuṃ susthiram ] H; samaṃ vigraham VAGTHSY}\\+}
+\tl{\var{7b meḍhre ] meṃḍhre A • athaikam āsyahṛdaye ] H; athaikadeśahṛdayo V, athaikam eva niyataṃ AGBGPk, athaikam eva niṣataṃ GL, athaikam eva hṛdayaṃ GP, athaikaṃ ekahṛdayo T, athaikam ekahṛdayaḥ Y • dhṛtvā ] VGBT; kṛtvā GLGPGPkY}\\+}
 \tl{\var{7c paśyed ] TH; paśyan VAGHSTvl, paśyad Y • antaraṃ ] VAGPkGLT; antare GBGP}\\+}
@@
 \begin{philcomm}[hp01_035]
-The second verse quarter has many variations in the \emph{Haṭhapradīpikā} manuscripts, as well as the manuscripts of the sources and testimonia.
-\getsiglum{N3} of the alpha group has \emph{athaikam ekahṛdaye dhṛtvā} and most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}.
+The second verse quarter has many variations in the \emph{Haṭhapradīpikā} manuscripts, as well as in the manuscripts of the sources and testimonia.
+\getsiglum{N3} of the \textalpha\ group has \emph{athaikam ekahṛdaye dhṛtvā} and most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}.
 The repetition of \emph{eka} appears to be a dittographical error and \emph{hṛdaye} does not make sense without an object of \emph{dhṛtvā} (i.e., having held what on the chest?).
-In the adopted reading \emph{athaikam āsyahṛdaye dhṛtvā}, which is attested by eta one, the oldest dated manuscript, the compound \emph{āsyahṛdaye} can be understood with the gerund \emph{dhṛtvā} and \emph{ekam} in the sense of `having held the face and chest together.' This alludes to the practice of Jālandhara's lock, in which the chin is placed on the chest.
+In the adopted reading \emph{athaikam āsyahṛdaye dhṛtvā}, which is attested by \etaOne, the oldest dated manuscript, the compound \emph{āsyahṛdaye} can be understood with the gerund \emph{dhṛtvā} and \emph{ekam} in the sense of `having held the face and chest together.' This alludes to the practice of the Jālandhara lock, in which the chin is placed on the chest.
 This meaning is more clearly seen in the \emph{Jyotsnā}'s version of the second verse quarter, \emph{hṛdaye kṛtvā hanuṃ susthiram} (`having put the jaw formly on the chest').
 The compound \emph{āsyahṛdaye} is supported by other manuscripts, such as \getsiglum{N10,P1,P6,J16} and the variants \emph{asyahṛdaye} and \emph{asyahṛdayaṃ} also occur (e.g., \getsiglum{J7,J12,A1,V4,V18,V16,P8,P9}).
-Another possible reading is \emph{ekahṛdayo}, which is not supported by manuscripts of the \emph{Haṭhapradīpikā} in the most important stemmatic groups but is in two other manuscripts (\getsiglum{J2,M1}), the six-chapter \emph{Vivekamārtaṇḍa} and \emph{Yogacintāmaṇi}, which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}.
-In this case, \emph{ekahṛdayaḥ} would qualify the yogi as having a mind focused on one thing.
+Another possible reading is \emph{ekahṛdayo}, which is not supported by manuscripts of the \emph{Haṭhapradīpikā} in the most important stemmatic groups but is in two other manuscripts (\getsiglum{J2,M1}), the six-chapter \emph{Vivekamārtaṇḍa} and the \emph{Yogacintāmaṇi}, which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}.
+In this case, \emph{ekahṛdayaḥ} would perhaps qualify the yogi as having a mind focused on one thing.
 The other well-attested reading, \emph{athaikaṃ eva niyataṃ}, was an attempt to fix the problem of \emph{hṛdaye} by replacing it with \emph{niyatam}, which must be read with \emph{meḍhre pādam athaikaṃ} (`having fixed one foot on the penis').
-But \emph{niyatam} is redundant here because of \emph{vinyaset} in the first verse quarter.
-\end{philcomm}
+But \emph{niyatam} is redundant here because of \emph{vinyaset} in the first \emph{pāda}.
+
+\end{philcomm}
+
+\begin{metre}[hp01_035]
+Śārdūlavikrīḍita 
+\end{metre}
 
@@
 \begin{translation}[hp01_036]
-However, in another view:\\
-Having fixed the left heel on the penis, and put the other heel on that, this is \emph{siddhāsana}.\\ 
-Only the first teaching [on \emph{siddhāsana}] is accepted by me.
+However, in another school [\emph{siddhāsana} is taught as follows]:\\
+
+Place the left heel on the penis and put the other heel on top: this is \emph{siddhāsana}.\\
+
+Only the first teaching [on \emph{siddhāsana}] is accepted by me.\\
+
 \end{translation}
@@
 \begin{translation}[hp01_037]
-Some proclaim this is \emph{siddhāsana}, others know it as \emph{vajrāsana}, a few say it is \emph{muktāsana} and some \emph{guptāsana}.
+Some call this \emph{siddhāsana}, others know it as \emph{vajrāsana}, a few say it is \emph{muktāsana} and some call it \emph{guptāsana}.
 \end{translation}
@@
 
-Cf. \emph{Śivayogasāram} by Kolani Ganapatideva (date 14th c.)
+Cf.~the Telugu \emph{Śivayogasāramu} by Kolani Ganapatideva (date 14th c.)
 
@@
 
-Cf. A verse by the poet Pingali Surana (active 16th c.)
+and a Telugu verse by the poet Pingali Surana (active 16th c.)
 
@@
 
+The last two references are taken from Reddy 1982: 41–42
+
 \end{testimonia}
@@
 \begin{translation}[hp01_038]
-The Siddhas know \emph{siddhāsana} as the single most important [pose] amongst all postures, in the same way as a measured diet amongst rules and non-violence amongst observances.
+Like measured diet amongst rules and non-violence amongst observances, the adepts know \emph{siddhāsana} to be the single most important of all postures.
 \end{translation}
@@
 \begin{translation}[hp01_039]
-Among the eighty-four postures, one should regularly practise just \emph{siddhāsana}, in the same way one should practise Suṣumnā among the 72,000 channels.
+Of the eighty-four postures, one should regularly practise just \emph{siddhāsana}, in the same way that of the 72,000 channels one should practise Suṣumṇā.
 \end{translation}
@@
 \begin{philcomm}[hp01_039]
-It is odd to have \emph{suṣumnām} as the object of the verb \emph{abhyaset}.
-This reading is well attested by manuscripts of the \emph{Haṭhapradīpikā}, as well as the verse cited in the \emph{Yogacintāmaṇi}, which attributes it to the \emph{Yogasāramañjarī}.
+It is odd to have \emph{suṣumṇām} as the object of the verb \emph{abhyaset}.
+This reading is well attested by manuscripts of the \emph{Haṭhapradīpikā} and is also found in the \emph{Yogacintāmaṇi}, which attributes it to the \emph{Yogasāramañjarī}.
 Perhaps, the second hemistich was added somewhat haphazardly by Svātmārāma, and then others have tried to make sense of it by changing \emph{suṣumnām iva nāḍiṣu} to \emph{nāḍīnāṃ malaśodhanam}, which occurs in the \emph{Jyotsnā} (1.39).
 The reading \emph{nāḍīnāṃ malaśodhanam/e} is probably a patch as no other texts say that \emph{siddhāsana} clears the channels.
-However, the idea of purifying the channels can be found in other contexts (e.g., \emph{Kumbhakapaddhati} 120) and may hark back to an earlier notion of flushing (\emph{cālana}) the channels (e.g., \emph{Amṛtasiddhi} 11.6).
+However, the idea of purifying the channels can be found in other contexts (e.g., \emph{Kumbhakapaddhati} 120) and may hark back to an earlier notion of flushing (\emph{cālana}) the channels (e.g., \emph{Amṛtasiddhi} 11.6).%??
+somewhat haphazardly?
+I don't understand the second part of the note: purifying the channels is found all over the place.
+%JB: The point is: no other texts say that \emph{siddhāsana} clears the channels.
 
@@
 By meditating upon the self, restricting the diet and regularly practising \emph{siddhāsana} for twelve years, the yogi attains the \emph{niṣpatti} stage.
-What’s the point of lots of exhausting postures when there is \emph{siddhāsana}?
+What’s the point of the [other] many tiring postures when there is \emph{siddhāsana}?
 \end{translation}
@@
 
-The notion of \emph{āsana}s causing fatigue \emph{śrama} was already suggested above in the verse on corpse pose (\emph{śavāsana}).
+The notion of \emph{āsana}s causing fatigue (\emph{śrama}) was already suggested in the verse on the corpse pose (1.32).
 \end{philcomm}
@@
 %\end{philcomm}
+
+\begin{metre}[hp01_041]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
 
@@
 \begin{philcomm}[hp01_042]
-It seems likely that \emph{dṛḍhaṃ} (rather than \emph{dṛḍhe}) was originally intended in 1.42a because \emph{dṛḍhataraṃ}, which is not ambiguous, is used in 1.48a to qualify how \emph{padmāsana} should be adopted.
-Also, \emph{dṛḍhaṃ} complements \emph{sāvadhānaṃ} in 1.41a.
+It seems likely that \emph{dṛḍhaṃ} (rather than \emph{dṛḍhe}) was originally intended in 1.42a because \emph{dṛḍhataraṃ}, which is not ambiguous, is used in 1.48a to qualify how \emph{padmāsana} should be adopted, and \emph{dṛḍhaṃ} complements \emph{sāvadhānaṃ} in 1.41a.
 % J5 and G4 have dṛḍhaṃ (alpha 2 and 3), + P11 (alpha and beta)
 \end{philcomm}
+
+\begin{metre}[hp01_042]
+Anuṣṭubh (a: bha-vipulā)
+\end{metre}
 
@@
 \begin{translation}[hp01_043]
-There is no posture like \emph{siddhāsana}, no retention like \emph{kevala}, no seal like \emph{khecarī}, [and] no dissolution [of mind] like the one on the internal sound (\emph{nāda}).
+There is no posture like \emph{siddhāsana}, no breath-retention like \emph{kevala}, no seal like \emph{khecarī}, [and] no [means for the] dissolution [of mind] like the internal sound (\emph{nāda}).
 \end{translation}
@@
 \begin{philcomm}[hp01_043]
-The reading \emph{na kumbhasadṛśo 'nilaḥ} (`no breath like a retention') is the lectio difficilior and attested by two early witnesses (eta one and two).
-It could be original.
-However, the alpha manuscripts and several other important groups of manuscripts have \emph{kumbhaḥ kevalopamaḥ}, as well as the \emph{Yogacintāmaṇi} and some manuscripts of the \emph{Haṭharatnāvalī}, suggesting that this reading, which makes much better sense, was in the transmission at an early stage.
+The reading \emph{na kumbhasadṛśo 'nilaḥ} (`no breath like a retention') is the lectio difficilior and attested by two early witnesses (\etaOne\ and \etaTwo) and is possibly original.
+However, the \textalpha\ manuscripts and several other important witness groups have the adopted reading \emph{kumbhaḥ kevalopamaḥ}, as well as the \emph{Yogacintāmaṇi} and some manuscripts of the \emph{Haṭharatnāvalī}, suggesting that this reading, which makes much better sense, was in the transmission at an early stage.
 % Śivasaṃhitā has \emph{kumbhasadṛśaṃ balam}.
@@
 
+\begin{metre}[hp01_043]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
 %%%%%%%%%%
@@
 \begin{translation}[hp01_044]
-Now, lotus pose (\emph{padmāsana}).\\
-One should place the right foot on the left thigh, and the left on the right though, hold firmly the big toes with the hands behind the back, place the chin on the chest and gaze at the tip of the nose.
-This is called lotus pose, which destroys diseases for those who undertake the observances (\emph{yama}).
+Now the lotus pose (\emph{padmāsana}).\\
+
+Place the right foot on the left thigh, and the left on the right thigh, firmly hold the big toes with the hands crossed behind the back, put the chin on the chest and gaze at the tip of the nose.
+This, which destroys diseases for those who undertake the observances, is called the lotus pose.
 \end{translation}
@@
 \tl{etad vyādhivikārahāri yamināṃ padmāsanaṃ procyate || 8 ||\\!}
+\tl{\var{8a saṃsthāpya ] vinyasya T}\\+}
+\tl{\var{8b yāmyorūpari paścimena vidhinā dhṛtvā ] VT; dakṣorūpari tasya bandhanavid- hau dhṛtvā A, dakṣorūpari paścimena vidhinā dhṛtvā GPGPkY, tasyaivoparitaś ca vandhanavid- hiṃ kṛtvā GB , tato dakṣorūpari tasya vaṃdhanavidhau pṛṣṭhe GL (unm.) • dṛḍhaṃ ] dṛḍha∗ ṃ
+∗ V}\\+}
+\tl{\var{8c ālokayed ] ālokayan GB}\\+} 
+\tl{\var{8d etad ] antar° T • °vikārahāri yamināṃ ] VAT; °vikāranāśanakaraṃ GPk Y, °vikārakaṃdadamanaṇ GB , °vināśakāri yamināṃ GL GP, °vighātahāri yamināṃ Tvl • °nāśanakaraṃ G}\\!}
 \end{versinnote}
@@
 
+\begin{metre}[hp01_044]
+Śārdūlavikrīḍita 
+\end{metre}
+
 %%%%%%%%%%
@@
 \begin{translation}[hp01_045]
-However, in another view:\\
-Having put the upturned feet carefully on the thighs and the upturned hands in the middle of the thighs, one should fix the eyes on the tip of the nose.
-Having raised the root of the uvula with the tongue, one should place the chin on the chest and gradually [draw in] the breath [...].
+However, in another school [\emph{padmāsana} is taught as follows]:\\
+
+Carefully put the upturned feet on the thighs and the upturned hands in the middle of the thighs, fix the eyes on the tip of the nose, raise the root of the uvula with the tongue, place the chin on the chest, gently [draw in] the breath [...].
 \end{translation}
@@
 
-The main problem with this passage is the incomplete syntax of verse 1.46, which describes the position of the eyes, tongue, chin and chest in lotus pose.
-The words `the breath gradually' (\emph{pavanaṃ śanaiḥ}) are left hanging.
-The evidence of the source text, the \emph{Dattātreyayogaśāstra}, is important here because it includes a verse that makes sense of \emph{pavanaṃ śanaiḥ}.
-The verse that completes the syntax is missing in the \emph{Haṭhapradīpikā}, perhaps because of a scribal error that happened early in the transmission.
-As seen in the \emph{Dattātreyayogaśāstra} verses 36 and 37 both end with \emph{pavanaṃ śanaiḥ}, the repetition of which likely caused an eyeskip.
-
-\begin{versinnote}
-\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
-\tl{uttabhya cibukaṃ vakṣasy āsthāpya \textbf{pavanaṃ śanaiḥ} || 36 ||\\+}
-\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
-\tl{yathāśaktyaiva paścāt tu recayet \textbf{pavanaṃ śanaiḥ} || 37 ||\\+}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\!}
-\end{versinnote}
-
-The manuscript readings with \emph{vakṣa sthāpayet} (\getsiglum{J7,V3,J8,J10,J17,N17}) or something similar (\getsiglum{V1,W4}) do not offer a solution to the incomplete syntax nor indicate that Svātmārāma rewrote \emph{Dattātreyayogaśāstra} 36 so that he could omit \emph{Dattātreyayogaśāstra} 37.
-For, apart from the possibility of an eyeskip, it is also conceivable that Svātmārāma intended to quote only the two verses of the \emph{Dattātreyayogaśāstra} (35–36) that describe the physical posture of \emph{padmāsana} and omit the next verse (37) that made sense of \emph{pavanaṃ śanaiḥ} because it was not relevant to the posture itself.
-Be this as it may, in the absence of evidence that Svātmārāma included \emph{Dattātreyayogaśāstra} 37 or wrote a coherent version of \emph{Haṭhapradīpikā} 1.46, we have tried to make sense of \emph{pavanaṃ śanaiḥ} by supplying the gerund \emph{samākṛṣya} in our translation.
+The syntax of this verse as we have presented it is incomplete: at its end \emph{pavanaṃ śanaiḥ}, `the breath gradually', is left hanging.
+In the source text, the \emph{Dattātreyayogaśāstra}, the following verse completes the syntax with \emph{pūrayed}, “one should inhale”.
+Either Svātmārāma chose to leave the verse hanging (the following verse in the \emph{Dattātreyayogaśāstra} adds nothing about the form of the posture, which is the topic here) or the verse that completes the syntax fell out, perhaps because of a scribal error that happened early in the transmission.
+In the \emph{Dattātreyayogaśāstra} verses 36 and 37 both end with \emph{pavanaṃ śanaiḥ}, the repetition of which may have caused an eyeskip.
+
+%\begin{versinnote}
+%\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
+%\tl{uttabhya cibukaṃ vakṣasy āsthāpya \textbf{pavanaṃ śanaiḥ} || 36 ||\\+}
+%\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
+%\tl{yathāśaktyaiva paścāt tu recayet \textbf{pavanaṃ śanaiḥ} || 37 ||\\+}
+%\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\!}
+%\end{versinnote} ??JM: I've removed this, no need to repeat?
+
+The manuscript readings with \emph{vakṣa sthāpayet} (\getsiglum{J7,V3,J8,J10,J17,N17}) or something similar (\getsiglum{V1,W4}) do not offer a solution to the incomplete syntax so do not indicate that Svātmārāma rewrote \emph{Dattātreyayogaśāstra} 36 so that he could omit \emph{Dattātreyayogaśāstra} 37.
+In the absence of evidence that Svātmārāma included \emph{Dattātreyayogaśāstra} 37 or wrote a coherent version of \emph{Haṭhapradīpikā} 1.46, we have made sense of \emph{pavanaṃ śanaiḥ} by adding “[draw in]” in our translation.
 
@@
 \begin{quote}
-Supporting both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.
+Supporting both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.%?? JM: supporting?
 
@@
 Brahmānanda appears to have had in mind a probably older rule for meditation postures, according to which the tongue rests near the front teeth.
-One example of this is in \emph{Svacchandatantra}, which teaches a meditation pose called \emph{divyaṃ karaṇam} (4.365f.), in which the tongue is to rest at the tip
+One example of this is in \emph{Svacchandatantra} (4.365f.), which teaches a meditation pose called \emph{divyaṃ karaṇam} , in which the tongue is to rest at the tip
 of the teeth (\emph{dantāgre jihvām ādāya}).
 Other Tantric texts have this or similar rules, in which the tongue is supposed to rest either on the teeth or the palate, early examples being the \emph{Mṛgendrāgama}, \textit{yogapāda} 19 (\emph{dantāgre jihvām ādāya}) and \emph{Mataṅgapārameśvaratantra}, \textit{yogapāda} 2.27 (\emph{tālumadhyagatenaiva jihvāgreṇa}).
-Placing the tongue where it does not disturb the meditation seems quite appropriate for a `normal' meditative practice.\footnote{The rule of placing the tongue at the palate is also found in \emph{Iśānaśivapaddhati} 18.120: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān asaṃspṛśan ṛjukāyaḥ}}
-
-However, when the context is haṭhayogic physiology, we must assume that the aim is to reach the tongue back to the uvula, which is the source of `nectar.' Confusingly, yogic terminology includes many names for the uvula, and among these especially the term \emph{rājadanta} may give rise to confusion, since, as we have seen, the tongue might also in some yoga systems be placed at the front teeth.
-
-Furthermore, \emph{khecarīmudrā} has been described in manifold ways.
-For example, the tenth-century \emph{Mokṣopāya} (V.55.14c) says that the tongue rests at the `source of the palate' (\emph{tālumūlatalālagnajihvā}) and the commentary, the
+Placing the tongue where it does not disturb the meditation seems quite appropriate for a `normal' meditative practice.
+\footnote{The rule of placing the tongue at the palate is also found in \emph{Īśānaśivagurudevapaddhati} 18.120: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān asaṃspṛśan ṛjukāyaḥ}.
+For similar references in tantric and other works see Mallinson 2007:17–24}
+
+When the context is haṭhayogic physiology, placing the tongue at the uvula, which is the source of `nectar', is more appropriate.
+Confusingly, yogic terminology includes many names for the uvula, and among these especially the term \emph{rājadanta} may give rise to confusion, since, as we have seen, the tongue might also in some yoga systems be placed at the front teeth.
+
+Furthermore, the haṭhayogic \emph{khecarīmudrā} has been described in manifold ways.
+Usually the tongue is said to be inserted into the cavity above the palate but in some cases it is placed at the uvula.
+Thus the tenth-century \emph{Mokṣopāya} (V.55.14c) says that the tongue rests at the `source of the palate' (\emph{tālumūlatalālagnajihvā}) and the commentary, the
 \emph{Saṃsārataraṇi}, on the parallel passage in \emph{Laghuyogavāsiṣṭha} V.6.155, which reads \emph{tālumūlāntarālagnajihvā°}, explains that this means that the tongue is to be placed in the middle of the two regions of the palate, and that this is the \emph{nabhomudrā}, alias `\emph{khecarī}' (\emph{tālumūlāntarālagnajihvamūlaḥ tālumūlayoḥ kākudamūladeśayoḥ āntare lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā} | \emph{yā hi khecarı̄ty ucyate}).
 
-A little later in the \emph{Mokṣopāya} (V.78.24ab) it is made clear that one should reach the uvula, `placed at the root of the palate' (\emph{tālumūlagatāṃ yatnāj jihvayākramya ghaṇṭikām}).
+A little later in the \emph{Mokṣopāya} (V.78.24ab) it is made clear that one should reach the uvula, `at the root of the palate' (\emph{tālumūlagatāṃ yatnāj jihvayākramya ghaṇṭikām}).
 In view of this background we must conclude that the author of the \emph{Jyotsnā} was probably not aware of the yogic meaning of \emph{rājadanta} and has tried his best to make sense of the passage,
-echoing the idea of the two roots of the palate (although his text is not talking about the palate), but then wisely refers to the instruction of the teacher for practical details, probably noticing that his literal interpretation is somewhat opaque.
-In addition to his commentary on 1.46 (translated above), Brahmānanda’s comments on 3.22 (\emph{rājadantasthajihvāyām}) indicate that he thought the \emph{rājadanta} refers to the front teeth (\emph{kutaḥ}?
-\emph{yato dantānāṃ rājāno rājadantā rājadanteṣu tiṣṭhatīti rājadantasthāḥ, rājadantasthā cāsau jihvā ca tasyāṃ rājadantasthajihvāyāṃ bandhaḥ, taduparibhāgasya sambandhaḥ śastaḥ}).
+echoing the idea of the two roots of the palate (although his text is not talking about the palate), but then referring to the instruction of the teacher for practical details, probably noticing that his literal interpretation is somewhat opaque.
+In addition to his commentary on 1.46 (translated above), Brahmānanda’s comments on \emph{rājadantasthajihvāyām} at 3.22 indicate that he thought the \emph{rājadanta} refers to the front teeth (\emph{kutaḥ}?\emph{yato dantānāṃ rājāno rājadantā rājadanteṣu tiṣṭhatīti rājadantasthāḥ, rājadantasthā cāsau jihvā ca tasyāṃ rājadantasthajihvāyāṃ bandhaḥ, taduparibhāgasya sambandhaḥ śastaḥ}).
 \end{philcomm}
@@
 \begin{translation}[hp01_047]
-This is taught as lotus pose, which cures all diseases.
-It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
-[The view of lotus pose] taught next is the one I accept.
-\end{translation}
-% [MD] add translation of "paścāduktam eva matsaṃmatam".
+This is called the lotus pose [and] it cures all diseases. It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.\\
+
+Only [the teaching on the lotus pose] taught next is approved by me.  
+\end{translation}
+% ?? JM: surely "taught second"?
+%JB the second description follows this comment. So, 'next' makes good sense.
 
@@
 \begin{versinnote}
-\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\+}
+\tl{idaṃ padmāsanaṃ nāma sarvavyādhivināśanam |\\+}
 \tl{durlabhaṃ yena kenāpi dhīmatā labhyate bhuvi ||\\!}
+\tl{\var{38a nāma ] P; ∗ma T, proktaṃ cett.}\\+}
+\tl{\var{38d dhīmatā ] dhīmatāṃ A • bhuvi ] yadi M1, hi vai A}\\!}
 \end{versinnote}
@@
 
-A man should put the hands together in a bowl shape, very firmly assume \emph{padmāsana}, firmly place the chin on the chest and meditation in the mind.
+A man should put his hands together in a bowl shape, very firmly assume \emph{padmāsana}, place the chin tight on the chest and meditation in the mind.
 Raising the \emph{apāna} breath over and over again [and] releasing the inhaled \emph{prāṇa}, he attains unequalled knowledge through the power of the goddess [Kuṇḍalinī].
@@
 \tl{muñcan prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ ||\\!}
+\tl{\var{36a °puṭitau ] GL GP GPk U; ghaṭitau VAGB T • baddhvātha ] VTvl U;
+baddhvā tu AGBH, dhyānaṃ tu GL, dhyāyes ta° GP, dhyātvā ca GPk, baddhvā ca T • padmāsanam ] VGBTU; tac cepsitaṃ GL, °taś cepsitaṃ GP, tat prekṣitam GPk}\\+}
+\tl{\var{36b dhyāyaṃś ]
+TH; dhyānaṃ VAGU •  °cetasi ] °cetasaṃ A}\\+} 
+\tl{\var{36c proccālayan ] ∗∗
+T; pro cc ālayan V, prodvārayaṃ A, proccālayet GB, proccārayet GLGPk, prollāsayet GP, proccārayan U • pūri- taṃ ] prerayaṃ A}\\+}
+\tl{\var{36d muñcan prāṇamupaiti bodhamatulaṃ śak- tiprabhāvān naraḥ ] U; prāṇaṃ muṃcati bodham eti śanakaiḥ proktaprabhāvād ataḥ V, pāṇaṃ muṃcati bodham eti śanakaiḥ śaktiḥ prabhāvād ataḥ A, prāṇaṃ muṃcati yāti bodham amalaṃ śaktipradhānoditaḥ GB, muṃcan prāṇam upaiti bodham atulaṃ śaktiprabhāvād ataḥ GLGP, muñcan prāṇam upaiti bodham atulaṃ śaktiprabodhān naraḥ GPk, prāṇaṃ muñcati bodhameti śanakai (ścu?śśa)[sic] ktiprabodhān naraḥ T, muñcan prāṇam upaiti bodham akhilāṃ śaktiṃ prabhāvād ataḥ Tvl}\\!}
 \end{versinnote}
@@
 Later witnesses, including Brahmānanda, have \emph{dhyāyaṃś ca} but none of the early ones has this reading.
-So, we have to take \emph{dhyānaṃ} with \emph{sannidhāya}.
-However, \emph{tat} is problematic because here it has no clear referent.
-In the \emph{Vivekamārtaṇḍa, tat} appears to refer to the \emph{mokṣadvāra} broken by Kuṇḍalinī, which is mentioned in the previous verse.
-
-The two participles \emph{proccālayan} and \emph{muñcan} imply that the two things are happening at the same time, which is surprising but perhaps possible.
-Or one might infer that `and' (\emph{ca}) was supposed to be supplied and understand both actions as sequential.
-
-%In the VM this verse comes in the passage on raising Kuṇḍalinī, so tat could refer to visualising K opening the mokṣadvāra.
-\end{philcomm}
+We are taking \emph{dhyānaṃ} with \emph{sannidhāya}, but this renders \emph{tat} problematic because it has no clear referent.
+In the source text, the \emph{Vivekamārtaṇḍa, tat} appears to refer to the \emph{mokṣadvāra} broken by Kuṇḍalinī, which is mentioned in the previous verse.
+
+The two participles \emph{proccālayan} and \emph{muñcan} imply that the two things are happening at the same time, which is surprising but perhaps possible.%Or one might infer that `and' (\emph{ca}) was supposed to be supplied and understand both actions as sequential.
+??
+JM: but can wouldn't necessarily imply that they are sequential.
+%JB ?
+
+\end{philcomm}
+
+\begin{metre}[hp01_048]
+Śārdūlavikrīḍita 
+\end{metre}
 
@@
 
-Gamma one's reading of \emph{niyatam} (instead of \emph{dhārayed}) in the third verse quarter explains the passive verbs in other witnesses.
+\gammaOne's reading of \emph{niyatam} (instead of \emph{dhārayed}) in the third verse quarter explains the passive verbs in other witnesses.
 The passive verbs do not make sense with \emph{yas tu}.
@@
 
-In the third verse quarter, the best alpha manuscript reads \emph{māruto mriyate yas tu}, which does not make sense.
-However, if one accepts \emph{pūrayet} in the second verse quarter, one could emend alpha's reading to \emph{māruto mriyate yasya}, which makes good sense (i.e., `the yogi whose breath dies is undoubtedly liberated').
+In the third verse quarter, the best \textalpha\ manuscript reads \emph{māruto mriyate yas tu}, which does not make sense, but if one accepts \emph{pūrayet} in the second verse quarter, one could emend \textalpha's reading to \emph{māruto mriyate yasya}, which makes good sense (i.e., `the yogi whose breath dies is undoubtedly liberated').
 In the same vein, J5 also has the plausible reading \emph{mārutaṃ mārayet yas tu}.
@@
 
-The \emph{Jyotsnā} (1.49) has \emph{naḍīdvāreṇa} instead of \emph{nāḍīdvāreṣu}, and Brahmānanda interprets it as the opening of the central channel (\emph{suṣumnāmārgeṇa}).
+The \emph{Jyotsnā} (1.49) has \emph{nāḍīdvāreṇa} instead of \emph{nāḍīdvāreṣu}, and Brahmānanda interprets it as the opening of the central channel (\emph{suṣumnāmārgeṇa}).
 This yields the idea of filling up the central channel (as opposed to other channels), which is described in the \emph{Yogabīja} (94–95).
@@
 [The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
-He should place the left ankle on right, the right ankle on the left and both hands on the knees, spread his fingers, open his mouth and gaze in deep concentration at the tip of the nose.
-This is the lion’s pose, which is always worshipped by yogis.
+He should place the left ankle on right, the right ankle on the left and both hands on the knees, spread his fingers, open his mouth and gaze in deep concentration at the tip of his nose.
+This is the lion’s pose, which is always honoured by yogis.
 It causes the three locks to arise together and is the best of [all] postures.
@@
 
-Spreading the fingers and keeping the mouth open seem to be imitating the lion, and this is depicted in some iconography of yoganarasiṃha (for example, Yoga Narasimha, Vishnu's Man-Lion Incarnation, Samuel Eilenberg Collection, Bequest of Samuel Eilenberg, 1998, Accession Number: 2000.284.4.
+Spreading the fingers and keeping the mouth wide open mimic a lion, and this is depicted in some iconography of yoganarasiṃha (for example, Yoga Narasimha, Vishnu's Man-Lion Incarnation, Samuel Eilenberg Collection, Bequest of Samuel Eilenberg, 1998, Accession Number: 2000.284.4.
 https://www.metmuseum.org/art/collection/search/39251).
 
-The \emph{Yogacintāmaṇi} attributes its citation of these verses on \emph{siṃhāsana} to the Yājñavalkya.
+The \emph{Yogacintāmaṇi} attributes its citation of these verses on \emph{siṃhāsana} to Yājñavalkya.
 Its citation does not include \emph{Haṭhapradīpikā} 1.52cd, which affirms that 1.52cd is not from the \emph{Yogayājñavalkya} (or \emph{Vasiṣṭhasaṃhitā}).
 As far as we are aware, there is no source for the hemistich mentioning the three locks, so it may have been composed by Svātmārāma or borrowed from a lost work.
-\end{philcomm}
+%\textalpha \alphathree %??JM: I don't understand this.
+Don't we know already that it's not in the YY or VS?
+%JB: I think the comment is trying to explain that "there is no source for the hemistich mentioning the three locks".%OK, but the first two sentences of the para seem redundant to me (they are implied by the last)
+\end{philcomm}
+
+\begin{metre}[hp01_051]
+Anuṣṭubh (a: ma-vipulā)
+\end{metre}
 
@@
 [The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
-Having firmly and motionlessly held the sides of the feet with the hands, this is the friendly pose, which cures all diseases and poisons.
+By firmly and very steadily holding the sides of the feet with the hands, the friendly pose arises, which cures all diseases and poisons.
 Yogis of the Siddha tradition call it Gorakṣa's pose (\emph{gorakṣāsana}).
@@
 \begin{philcomm}[hp01_053]
-We have understood \emph{pārśvapāda} (i.e., side of the foot) like the compound \emph{agrapāda} (i.e., front of the foot), which can be classified as a \emph{ekadeśitatpuruṣa}.
-See Sanskrit Vademecum, 2019: 84.
-
-Manuscripts of two early groups, beta and gamma, as well as the \emph{Jyotsnā} (1.53), include an additional hemistich specifying that the left ankle is placed on the left side and the right ankle on the right (\emph{savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe}).
+We have understood \emph{pārśvapāda} as a \emph{ekadeśitatpuruṣa} meaning the side of the foot, like  \emph{agrapāda}, the toes.
+
+Manuscripts of two early groups, \textbeta\ and \textgamma, as well as the \emph{Jyotsnā} (1.53), include an additional hemistich specifying that the left ankle is placed on the left side and the right ankle on the right (\emph{savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe}).
 This hemistich appears to have been added to make it clear that the ankles are not crossed in \emph{bhadrāsana}, unlike the previous pose, \emph{siṃhāsana}.
@@
 \begin{translation}[hp01_055]
-When the great yogi does not become tired on adopting the \emph{āsana}s in this way, he should  practise the breath techniques of seals and so forth, from which purification of the channels arise.
+When the great yogi does not become tired on adopting the \emph{āsana}s in this way, he should  practise the breath techniques of seals and so forth, from which purification of the channels arise.%??JM: work on the breath using seals
+%JB: I think our original translation is better.
+%JM: but not all mudrās work on the breath; also the compound cannot mean "a breathing technique along with mudrās etc." as said in the commentary: "along with" isn't there.
 \end{translation}
@@
 
-The metre of 1.55c is a \emph{ra-vipulā} with \emph{nāḍi}.
-\end{philcomm}
+The metre of 1.55c is a \emph{ra-vipulā} with \emph{nāḍi}.%??JM:are we planning to note all vipulās?
+%JB Maybe Sven can use that software to identify them once we've finalised the text. It might be useful for future editorial work, particuarly on related works.
+\end{philcomm}
+
+\begin{metre}[hp01_055]
+Anuṣṭubh (c: ra-vipulā)
+\end{metre}
 
@@
 This is true, there is no doubt.
-In this tradition, it should not be given to one who wears robes and is devoted to his genitals and stomach.
+In this system, [the practice] should not be given to one who wears robes and is devoted to his genitals and stomach.%??
+JM: In the  DYŚ I have lust and gluttony.
+This sounds a bit odd.
 \end{translation}
@@
 \tl{anuṣṭhānavihīnās tu vañcayanti janān kila || 47 ||\\!}
+\tl{\var{42c °yuktasya ] °yuktaḥsya W1, °yuktaḥ sa PT
+• siddhiḥ ] siddhi M1M2, siddhaḥ P}\\+}
+\tl{\var{43a na śāstra° ] śāstrasya PT • °mātreṇa ] °rūpeṇa
+M1A}\\+}
+\tl{\var{46a na ] sa° BB-
+BP • °dhāraṇaṃ ] °dhāriṇāṃ AM2, °dhāriṇaṃ JVYSS • siddheḥ ] sidhyai M1, siddhiḥ
+A, siddhi° β, kiṃ cit Y}\\+}
+\tl{\var{46b om. M AM (eye-skip) • kāraṇaṃ na ] kāraṇa∗nn∗a T SS 12
+• na ] tañ β • kathā ] W1VHPYCMK; katham PTβ, tathā JW2DYŚPT }\\+}
+\tl{\var{46c kriyaiva ] PTβW1VYCMPTKYSS; om. M1AM2, kṛpaiva JDYŚ, kṛyaiva W2 • siddheḥ ] siddhiḥ JYSS}\\+}
+\tl{\var{46d satyam ] satvam YSS • eva ] PTW1W2VDYŚK; etad M1AM2YCM, ekaṃ J, etan HP, etat PTYSS, etac β • tu sāṃkṛte ] dhi sāṃkṛte M1AM2YCM, na saṃśayaḥ HP, ca sāṃkṛte β}\\+}
+\tl{\var{47a śiśno° ] siśno° JW2 47a °ārthaṃ yogasya ] °āś ca yogāś ca P, ⊔⊔ yogasya T, °ārthaṃ yogaś ca AM2, °ārthayogaś ca β}\\+}
+\tl{\var{47b kathayā veṣadhāriṇaḥ ] W1W2DYŚYCMK; katheyaṃ veṣadhāriṇaḥ PTYSS, kasyeyaṃ veṣyadhāriṇaḥ M1, katheyaṃ veṣadhāriṇāṃ AM2, kaṃd- hāyāṃ yogadhāraṇaḥ BBBP, kathayan yogadhāraṇaḥ BA, kathaṃ yā veṣadhāriṇī J1, kathaṃ vā veṣadhāriṇī J2, kathaṃ ye veṣadhāriṇaḥ V, kathaṃ vā veṣadhāriṇaḥ PT}\\+}
+\tl{\var{47c anuṣṭhāna° ] anuṣṭhānā° M1, annapāna° PT • °vihīnās tu ] svayaṃ nityaṃ M1, °vihīnāś ca AM2}\\+}
+\tl{\var{47d janān ] janāḥ M1βV, janāt J, janā W1 • kila ] sadā M1AM2, iha YSS • vañcayanti janān kila ] vañcayaty akhilān janān YCM}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_055_1]
-1.55*1–2 are omitted from the alpha, gamma, delta and epsilon groups, so it is likely these verses were not in the earliest versions of the \emph{Haṭhapradīpikā}.
+1.55*1–2 are omitted from the \textalpha, \textgamma, \textdelta\ and \textepsilon\ groups, so it is likely these verses were not in the earliest versions of the \emph{Haṭhapradīpikā}.
 In fact, it is likely that both were added (perhaps initially as marginal notes) to elaborate on the word \emph{kriyā} in 1.55d.
@@
 \begin{translation}[hp01_055_3]
-Is this empty bubble we call the universe dissolved or did it arise in me, the pure ocean of awakening? Where does [this] veil of doubt come from?
+Did this empty bubble we call the universe dissolve or arise in me, the pure ocean of awakening? Where does [this] veil of doubt come from?
 \end{translation}
@@
 \begin{philcomm}[hp01_055_3]
-Verse 1.55.3 is only found in manuscripts of the delta group. It is very difficult to find a reason why this verse should be inserted here. It is apparently a
+Verse 1.55.3 is only found in manuscripts of the \textdelta\ group. It is very difficult to find a reason why this verse should be inserted here. It is apparently a
 \emph{muktaka} that would befit an accomplished spiritual poem more than an instructional manual, like the \emph{Haṭhapradīpikā}, even here, in what appears as a sort of miscellaneous section at the end of a chapter.
@@
 The reader might wonder how this illusionist verse could be understood to fit our Yoga text.
-We can only speculate that perhaps the scribe of the hyparchetype of the delta manuscripts was fond of it.
+We can only speculate that perhaps the scribe of the hyparchetype of the \textdelta\ manuscripts was fond of it.
 
@@
 \begin{translation}[hp01_055_4]
-Realisation from scripture, one's own guru or oneself and the cessation of mind; all these methods have been combined and taught by the wise in this tradition.
+Realisation from scripture, one's own guru and oneself and the cessation of mind; all these methods have been combined and taught by the wise in this tradition.%?? repeat "realisation" like the text?
 \end{translation}
@@
 \begin{philcomm}[hp01_055_4]
-Verse 1.55.4 is in some of the delta manuscripts and is quoted in \emph{Yogacintāmaṇi}.
-
-The reading in the \emph{Yogacintāmaṇi} `cessation of mind' (\emph{manonirodhaḥ}) is better than \emph{manaso 'pi bodhaḥ} (the delta reading) in a yogic context.
-\end{philcomm}
+Verse 1.55.4 is in some of the \textdelta\ manuscripts and is quoted in \emph{Yogacintāmaṇi}.
+
+The reading in the \emph{Yogacintāmaṇi} `cessation of mind' (\emph{manonirodhaḥ}) is better than \emph{manaso 'pi bodhaḥ} (the \textdelta\ reading) in a yogic context.
+\end{philcomm}
+
+\begin{metre}[hp01_055_4]
+Upajāti
+\end{metre}
 
@@
 \begin{translation}[hp01_056]
-The various \emph{āsana}s, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.%??JM: Posture, multifarious breath retention, the bodily attitude called seal and then concentration on the internal sound is the sequence of practice in Haṭha.
+The various \emph{āsana}s, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.%??JM: Posture, multifarious breath retention, the bodily attitude called seal, then concentration on the internal sound is the sequence of practice in Haṭha.
 \end{translation}
@@
 \begin{philcomm}[hp01_056]
-This verse was omitted from eta one, the oldest dated manuscript.
+This verse was omitted from \etaOne, the oldest dated manuscript.
 The omission is probably deliberate as that manuscript does not have chapter four, which teaches \emph{nādānusandhāna}.
-The numbering in eta one suggests that its exemplar had this verse.
+The numbering in \etaOne\ suggests that its exemplar had this verse.
 
@@
 \begin{translation}[hp01_057]
-Celibate, restricted in diet and devoted to yoga, the yogi succeeds in upwards of a year. No doubt about this should be entertained.
+Celibate, restricted in diet and devoted to yoga, the yogi becomes an adept after a year. No doubt about this should be entertained.
 \end{translation}
@@
 But it is more difficult to determine whether Svātmārāma used a manuscript of the \emph{Vivekamārtaṇḍa} with \emph{tyāgī} or \emph{yogī} in verse 37.
-Since the best alpha manuscript has \emph{yogī}, as well as eta one and many others, we have tentatively adopted it bearing in mind that it was changed early in the transmission of the \emph{Haṭhapradīpikā}, most likely by a scribe who knew the reading of \emph{tyāgī} in a manuscript of the \emph{Vivekamārtaṇḍa}.
+Since the best \textalpha\ manuscript has \emph{yogī}, as well as \etaOne\ and many others, we have tentatively adopted it bearing in mind that it was changed early in the transmission of the \emph{Haṭhapradīpikā}, most likely by a scribe who knew the reading of \emph{tyāgī} in a manuscript of the \emph{Vivekamārtaṇḍa}.
 \end{philcomm}
@@
 \tl{annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet |\\+}
-\tl{āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet ||\\!} 
+\tl{āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet ||\\!}%??JM: but both these examples say to half-fill the stomach with food and fill a quarter with water.
+It's odd how in the HP/GŚ the stomach isn't mentioned.
+It seems to mean eat 3/4 of the amount eaten in a normal diet.
 \end{versinnote}
@@
 \begin{translation}[hp01_059]
-[Adepts] say the [following] is unwholesome: pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink.
-Flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like.
+ Pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink, flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like: they say that this is unwholesome.
 \end{translation}
@@
 The spelling \emph{°harīta} was probably adopted for metrical reasons.
-In some Nighaṇṭus (`thesauruses'), \emph{°haritaśāka°} is glossed as horseradish.
+In some Nighaṇṭus (`thesauruses'), \emph{°haritaśāka°} is glossed as horseradish (\emph{śigru}).
 \begin{quote}
@@
 The term \emph{°sauvīra°} (1.59b) probably means sour gruel.
-Brahmānanda glosses  \emph{sauvīra} as \emph{kāñjika}, which is `fermented rice water.' On \emph{sauvīra}, Meulenbeld (1974: 516–517) says, `emph{sauvīra} is sour gruel made from barley and wheat.' The process of making it is described in the \emph{Suśrutasaṃhitā} (1.44.35--40ab) as follows:
+Brahmānanda glosses  \emph{sauvīra} as \emph{kāñjika}, which is `fermented rice water.' On \emph{sauvīra}, Meulenbeld (1974: 516–517) says, \emph{sauvīra} is sour gruel made from barley and wheat.' The process of making it is described in the \emph{Suśrutasaṃhitā} (1.44.35--40ab) as follows:
 
@@
 
-In the compound \emph{ājādimāṃsa}° (1.59c), the word \emph{āja}° is required for the metre.
-So, variants beginning with \emph{aja}° can be dismissed.
+In the compound \emph{ājādimāṃsa}° (1.59c), the adjective \emph{āja}° is required for the metre,  so variants beginning with \emph{aja}° can be dismissed.
 Another well-attested reading is \emph{ājāvimāṃsa}°.
-Although this was probably read as `goat and sheep flesh,' \emph{āvi} is not attested as an adjective of sheep, so this reading was probably not original.
+Although this was probably read as `goat and sheep flesh,' \emph{āvi} is not attested as an adjective of sheep,
+so this reading was probably not original.
 Moreover, only \emph{ājādimāṃsa}° makes good sense.
@@
 
+\begin{metre}[hp01_059]
+Vasantatilakā 
+\end{metre}
+
 %%%%%%%%%%
@@
 \begin{translation}[hp01_060]
-One should know food that has been reheated, is dry, too salty or sour, contains an excess of leafy vegetables that are hard to chew, [or] is spoiled to be unwholesome.
+One should know unwholesome food to be that has been reheated, is dry, too salty or sour, contains an excess of leafy vegetables that are hard to chew, [or] is spoiled.
 \end{translation}
@@
 
+\begin{metre}[hp01_060]
+Upagīti 
+\end{metre}
+
 %%%%%%%%%%
@@
 Similarly a saying by Goraksa:
-One should avoid places near bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.
+One should avoid places with bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.%??maybe prānta = borderlands?
+In Hindi it means region but maybe not Skt
 \end{translation}
@@
 \begin{philcomm}[hp01_061]
-Manuscripts from the alpha, beta and epsilon groups have the reading \emph{durjanaprāntaṃ} (1.61a), which is the lectio difficilior in relation to \emph{durjanaprītiṃ} (`the friendship of wicked people').
-We have understood \emph{durjanaprānta} in line with Brahmānanda's gloss in \emph{Jyotsnā} 1.64, `a dwelling near bad people' (\emph{durjanasamīpavāsa}) or, in other words, living near bad people.
+Manuscripts from the \textalpha, \textbeta\ and \textepsilon\ groups have the reading \emph{durjanaprāntaṃ} (1.61a), which is the lectio difficilior in relation to \emph{durjanaprītiṃ} (`the friendship of wicked people').
+We have understood \emph{durjanaprānta} in line with Brahmānanda's gloss in \emph{Jyotsnā} 1.64, `dwelling near bad people' (\emph{durjanasamīpavāsa}).
 \end{philcomm}
@@
 \begin{translation}[hp01_062]
-Pure food with wheat, rice, śāli rice, barley, sixty-day \emph{śāli} rice, milk, ghee, cream, fresh butter, ground sugar and honey.
-Dried ginger, fruit of the snake gourd and so forth, the five vegetables, mung beans and so on, and rain water are wholesome for the best of sages.
+Pure food with wheat, rice, śāli rice, barley, sixty-day \emph{śāli} rice, milk, ghee, cream, fresh butter, ground sugar and honey, dried ginger, fruit of the snake gourd and so forth, the five vegetables, mung beans and so on, and rain water are wholesome for the best of ascetics.%??JM:How about "Pure grains such as wheat..."?
+and semi-colons after each group.
 \end{translation}
@@
 \begin{philcomm}[hp01_062]
-In 1.62b, \emph{maṇḍa}, which is supported by alpha, beta and gamma, is more likely than \emph{khaṇḍa} (`candied sugar') because fits the context of diary products mentioned in this compound (i.e., \emph{kṣīra, ājya} and \emph{navanīta}).
+In 1.62b, \emph{maṇḍa}, which is supported by \textalpha, \textbeta\ and \textgamma, is more likely than \emph{khaṇḍa} (`candied sugar') because it fits the context of diary products mentioned in this compound (i.e., \emph{kṣīra, ājya} and \emph{navanīta}).
 The term \emph{navanīta} is discussed in \emph{Suśrutasaṃhitā, sūtrasthāna,} 45.92 as follows:
@@
 
+\begin{metre}[hp01_062]
+Vasantatilakā 
+\end{metre}
+
 %%%%%%%%%%
@@
 The last is made somewhat redundant by \emph{mano 'bhilaṣitaṃ} in 163c.
-Both \emph{mṛṣṭaṃ} (alpha and eta) and \emph{miṣṭaṃ} (beta and gamma) are both well attested by manuscripts of important groups and there is hardly any difference in their meaning in this context.
-We have adopted \emph{mṛṣṭaṃ} as it is supported by the alpha group.
+Both \emph{mṛṣṭaṃ} (\textalpha\ and \texteta) and \emph{miṣṭaṃ} (\textbeta\ and \textgamma) are well attested by manuscripts of important groups and there is hardly any difference in their meaning in this context.
+We have adopted \emph{mṛṣṭaṃ} as it is supported by the \textalpha\ group.
 \end{philcomm}
@@
 \begin{philcomm}[hp01_064]
-Eta one, the oldest dated manuscript, has a different reading for the last verse quarter (164d): \emph{sarvaṃ yogī yatendriyaḥ} (`the yogi whose senses are restrained wholly succeeds [...]').
-Here, \emph{sarvaṃ} is not easy to construe, and the alpha manuscripts and other important groups of \emph{Haṭhapradīpikā} manuscripts indicates that \emph{sarvayogeṣv atandritaḥ} was the reading adopted by Svātmārāma, which is more similar to the \emph{Dattātreyayogaśāstra}'s (i.e., \emph{yoge sarvo 'py atandritaḥ}).
+\etaOne, the oldest dated manuscript, has a different reading for the last verse quarter (164d): \emph{sarvaṃ yogī yatendriyaḥ} (`the yogi whose senses are restrained wholly succeeds [...]').
+Here, \emph{sarvaṃ} is not easy to construe, and the readings of the \textalpha\ manuscripts and other important groups of \emph{Haṭhapradīpikā} manuscripts indicate that \emph{sarvayogeṣv atandritaḥ} was the reading adopted by Svātmārāma, which is more similar to the \emph{Dattātreyayogaśāstra}'s (i.e., \emph{yoge sarvo 'py atandritaḥ}).
 \end{philcomm}

### 2024-02-20 latex/HP1_comm.tex
new files

@@
 \input{PreambleComm}
+%\def\Sone{{\scriptsize{\'S\lower.4ex\hbox{\textrm{\tiny{1}}}}}}
+%\def\Soneac{{\scriptsize{\'S\lower.4ex\hbox{\textrm{\tiny{1}}}\kern-2.2pt\lower-1ex\hbox{\textrm{\tiny{ac}}\kern1pt}}}}
+%\def\Sonepc{{\scriptsize{\'S\lower.4ex\hbox{\textrm{\tiny{1}}}\kern-2.2pt\lower-1ex\hbox{\textrm{\tiny{pc}}\kern.7pt}}}}
 \begin{document}
@@
 \begin{translation}[hp01_001]
-Homage to the glorious Ādinātha who taught the science of Haṭhayoga which is like a splendid stairway for one who wants to climb to the lofty royal terrace.%??Add note re meaning of "science".JM: would much prefer "science" to "system" (vidyā doesn't mean system).
+Homage to the glorious Ādinātha who taught the science of Haṭhayoga which is like a splendid stairway for one who wants to climb to the lofty royal terrace.%Add note re meaning of "science".JM: would much prefer "science" to "system" (vidyā doesn't mean system).
 I changed "ladder" to "stairway" (a bit more poetic).
@@
 JM: But "system" is wrong.
+JB: doctrine is better than science.
 
@@
 \begin{philcomm}[hp01_001]
-%?? JM: now saudha is preferred for stemmatic reasons too. I suggest the following shortened form of the note: 
+
+In his commentary {\emph Jyotsnā} on \emph{Haṭhapradīpikā} 1.1 Brahmānanda first states that the
+author Svātmārāma starts appropriately with a \emph{maṅgala}, a verse of adoration addressing
+Ādinātha, i.e.\ Śiva, but in accordance with his non-sectarian approach leaves room for a second
+interpretation of the word as Viṣṇu. A sentence later this is contradicted by his explanantion that Ādinātha, who first taught Yoga, taught it to Pārvatī, which limits the scope to Śiva.
+
+%add comment about yoga replacing mārga and saudha here and in the next two verses.
+The reading \emph{°rājasaudham} in the third verse quarter is well attested by the manuscripts, including \alphaOne.
+However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below).
+It apears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.
+
+In light of the variants, which have lead to our critical edition, Brahmānanda's choices and
+interpretation of the second half cannot be upheld. This is Brahmānanda's version and the relevant
+portion of his commentary:
+
+\begin{versinnote}
+śrīādināthāya namo 'stu tasmai\\ 
+\ \ yenopadiṣṭā haṭhayogavidyā |\\
+vibhrājate pronnatarājayogam\\ 
+\ \ āroḍhum icchor adhirohiṇīva ||1.1||
+\end{versinnote}
+
+Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}.
+However, the difference in the sense is so minor that it does alter the meaning.
+
+The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
+\emph{-rājamsaudhaṃ}, which he explains as follows:
+
+\begin{quote}
+  \emph{rājayogaś ca sarvavṛttini\-rodhalakṣaṇo 'samprajñātayogaḥ. tam icchor mumukṣor adhirohiṇīva
+  adhi\-ruhyate 'nayety adhirohiṇī niḥśreṇīva vibhrājate viśeṣeṇa bhrājate śobhate. yathā
+  pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudha\-prāpi\-kā bhavati evaṃ
+  haṭhapradīpikāpi pronnatarājayogam āroḍhum icchor anā\-yāsena rājayogaprāpikā bhavatīti
+  upamālaṅkāraḥ.  indravajrākhyaṃ vṛttam.}
+
+Rājayoga is the yoga without cognition,\footnote{This must refer to the \emph{asaṃprajñātasamādhi}
+  of the \emph{Yogasūtra} via the equation \emph{yogaḥ samādhiḥ} in the \emph{Bhāṣya} on \emph{Yogasūtra}  1.1.} defined as the stopping of all movements of the mind.
+To a liberation seeker desiring this, [the science of Haṭhayoga] shines like a ladder.\footnote{The commentary adds etymological explanations: ``\emph{adhirohiṇī} means that by which one ascends, i.e.\ a ladder'', and a synonym for ``shines'', which are not translated here.} Just as a ladder leads someone desiring to ascend to a lofty mansion easily to this mansion, in the same way also the
+\emph{haṭhapradīpikā} leads someone desiring the lofty Rājayoga easily to Rājayoga. [In this verse]
+the trope is a comparison. The metre is Indravajrā.
+\end{quote}
+
+The interpretation makes good sense: Haṭhayoga leads effortlessly to Rājayoga, just as a ladder
+leads one to the flat on the top floor. And the comparison as outlined by Brahmānanda has all the
+elements deemed necessary by Sanskritic poetology:
+
+\begin{enumerate}
+\item Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva} 
+\item Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā
+    adhirohiṇī}) 
+\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The text itself has
+    \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā
+    haṭhapradīpikā}) 
+\item Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})
+\end{enumerate}
+
+According to classical poetology a comparison containing all four elements is termed a ``full
+comparison'' (\emph{pūrṇopamā}), whereas an elision of one or more elements is called a ``deficient
+comparison'' (\emph{luptopamā}). One element that cannot be omitted without losing the comparison
+is the \emph{upamāna}. In the verse as given by Brahmānanda this would be ``the ladder''. The
+implication is spelt out by Brahmānanda in his commmentary: ``Just as a ladder leads someone
+desiring to ascend to a lofty mansion easily leads him to this mansion, in the same way also the
+\emph{haṭhapradīpikā} leads someone desiring the lofty Rājayoga easily to Rājayoga.'' However, his
+text version spells out only the side of the \emph{upameya}, that is, ``Haṭhayoga leads to
+Rājayoga'', but leaves the \emph{upamāna} open to interpretation. His mention of the word
+\emph{saudha} in the commentary suggests that this was one of the other options he found in
+manuscripts—this reading is very well attested—but was rejected by him. In this way Brahmānanda
+makes sure that the text states the obvious, but at the same time it loses part of the comparison,
+and it loses its dynamics, which is brought out in the reading \emph{rājasaudha} preferred with
+good manuscript evidence in our critical edition. We think that Brahmānanda was eager to state at
+the outset the primacy of Rājayoga, and to this end sacrificed the more poetically elegant original
+reading that mentions the royal mansion as a sort of metaphor for the ``royal yoga''.
+
+% JM: now saudha is preferred for stemmatic reasons too. I suggest the following shortened form of the note: 
 % JB: I prefer the longer note for the digital edition and the shorter one (proposed by Jim) for the printed edition.
@@
 The longer note nicely summarises a longer discussion in the article.
-The \textalpha\ witnesses read \emph{virājate pronnatarājayogam} while several others have either \emph{vibhrājate} or \emph{°rājasaudham} or both.
+%The \textalpha\ witnesses read \emph{virājate pronnatarājayogam} while several others have either \emph{vibhrājate} or \emph{°rājasaudham} or both.
 As explained in Hanneder 2020, \emph{virājate} is stylistically preferable and the comparison between the science of \emph{haṭhayoga} and a stairway works only with \emph{°rājasaudham}.
-
-[Place holder] This reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.
 
@@
 128–130.
-Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison ??
-JM: what's happened here?
+Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison JM: what's happened here?
 ALso, do we need to repeat what Jürgen has already written elsewhere?
@@
 But from the perspective of \emph{alaṃkāraśāstra} the verse has a problem that has surely caused the dilemma: it either leaves the \emph{upameya} or the \emph{upamāna} incomplete.
-Brahmānanda has mentioned both ??
+Brahmānanda has mentioned both ?
 \emph{yathā adhirohiṇī saudhaprāpikā bhavati evaṃ haṭhadīpikā rājayogaprāpikā bhavati}, but had to supply \emph{rājasaudha} because his text reads \emph{rājayoga}.
@@
 
-?? add comment about yoga replacing mārga and saudha here and in the next two verses.
-
 %The metre of 1.1 is Indravajrā.
@@
 \begin{philcomm}[hp01_003] 
-Most witnesses have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b.
-While this reading is well attested by the manuscript transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note the readings of P,T,T1 of the critical edition [Gharote 2009: 5 n.
-2] in the testimonia), the reading \emph{rājamārgam ajānatām}, “for those unable to find the royal highway”, is more appropriate to the metaphor of being lost in darkness.
-
-In 1.3d \emph{kṛpākaraḥ} is attested by the most important groups, including \textalpha, of the available manuscripts.
+Most witnesses (including \alphaTwo) have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b.
+The reading \emph{rājamārgam ajānatām} (\alphaOne and \alphaThree), “for those unable to find the royal highway”, is more appropriate to the metaphor of being lost in darkness.
+
+%In 1.3d \emph{kṛpākaraḥ} is attested by the most important groups, including \textalpha, of the available manuscripts.
 The readings \emph{prakāśyate} and \emph{kṣamākaraḥ} are attested by some manuscripts in lower branches of the stemma.
 Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākaraḥ} makes good sense.
-As Brahmānanda notes, this compound can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}).
+
+As Brahmānanda notes, the compound \emph{kṛpākaraḥ} can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}).
 In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} probably arose as a mistake for \emph{kṛ}.
@@
 \begin{translation}[hp01_004]
-In fact, Matsyendra, Gorakṣa, and other [perfected yogis] discovered the science of Haṭha, and the yogi Svātmārāma knows it through their favour.
+For, Matsyendra, Gorakṣa, and other [perfected yogis] discovered the science of Haṭha, and the yogi Svātmārāma knows it through their favour.
 \end{translation}
@@
 \begin{philcomm}[hp01_008]  
-The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literature) is largely transmitted as \emph{allamaḥ prabhu\-devaḥ}, as though it were two names, although some manuscripts of the \emph{Haṭha\-pradīpikā} have \emph{allamaprabhudevaś ca} (i.e., \getsiglum{V3,V8,V13,V16,V22,N24,N26,Jyo}) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n.
+The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literature) is frequently transmitted as \emph{allamaḥ prabhu\-devaḥ} (cf.
+\alphaTwo), as though it were two names.
+However, manuscripts \alphaOne, \alphaThree\ and others (e.g., \getsiglum{V3,V8,V13,V16,V22,N24,N26,Jyo}) do not have the \emph{visarga} and write it as one name (i.e., \emph{allamaprabhudevaś ca}).
+This is also the case in some manuscripts of the \emph{Haṭharatnāvalī} (P,T,t1 in Gharote 2009: 35 n.
 8).
@@
 
-Many witnesses have \emph{khaṇḍa} and \emph{kāpālika} as separate names.
-However, Khaṇḍa\-kāpālika is well attested.
+The \alphaOne\ and \alphaTwo\ reading of \emph{siddhaḥ kāpālikas} is an exception among the manuscripts and seems too vague to be referring to someone within a lineage.
+However, Khaṇḍa\-kāpālika is well attested by the manuscripts (including \alphaThree) and this name appears in other literature.
 Examples include \emph{Kathāsaritsāgara} 121.5 ff.
@@
 %The reading of \emph{samāśrayo} in \etaOne\ is metrically faulty. 
-
-Both \emph{°jagatāṃ} and \emph{°yuktānāṃ} are well attested by the collated manuscripts, and the \textalpha\ group omits the second half of this verse.
+The \textalpha\ group omits the second hemistich of this verse, but this was probably the result of eyeskip (i.e., \emph{°maṭho haṭhaḥ} is repeated).
+Both \emph{°jagatām} and \emph{°yuktānām} are well attested by the collated manuscripts.
 We have adopted \emph{°jagatāṃ} because it makes good sense with  \emph{ādhārakamaṭhah} in light of the cosmological notion that the tortoise supports all the worlds.
-This reading may not have been understood by some and was changed in the vulgate and other witnesses to \emph{°yuktānāṃ} instead.
+This reading may not have been understood by some and was changed in other witnesses to \emph{°yuktānām}, which was adopted by Brahmānanda in \emph{Jyotsnā} 1.10.
 \end{philcomm}
@@
 It is neither too high nor too low in extent and is thickly smeared with cow dung in the proper way.
-It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
+It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.%JM remove "it is" etc.
+which read oddly, thus: With a small door and no cracks, holes and bumps, neither too high nor too low in size, thickly smeared with cow dung in the proper way, clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
 \end{translation}
@@
 \begin{philcomm}[hp01_013]  
-The syntax of this verse is rather problematic.
+The syntax of this verse is problematic.
 One would expect the features of the hut, which are listed in the first three quarters of the verse, to be in the nominative case.
@@
 
-%??(MD: Stemmatically, \emph{piṭaka} ``a basket; a boil" is well attested.
+% (MD: Stemmatically, \emph{piṭaka} ``a basket; a boil" is well attested.
 Is it perhaps possible to interpret the word as ``a bump"?
@@
 \begin{translation}[hp01_014]
-Staying in such a hut, free from all worry, in the way taught by his guru [the yogi] should practise only yoga.
+Staying in such a hut, free from all worry, in the way taught by his guru [the yogi] should practise nothing but yoga.
 \end{translation}
@@
 \begin{translation}[hp01_019]
-Placing the soles of both feet well between the knees and thighs [and] sitting up with the body straight: they call that \emph{svastikāsana}.%JM: correctly is tricky.
+Placing the soles of both feet well between the knees and thighs [and] sitting up with the body straight: they call that the auspicious pose (\emph{svastikāsana}).%JM: correctly is tricky.
 It could mean thoroughly, completely — I've looked at some other usages of samyak and that understanding seems to work well quite often, as it does here.
@@
 \begin{translation}[hp01_020]
-[The yogi] should place his right heel on the left side of the [lower] back, and the left [heel] on the right [side], in the same way. This is \emph{gomukhāsana}, which [looks] like a cow's face.
+[The yogi] should place his right heel on the left side of the [lower] back, and the left [heel] on the right [side], in the same way.
+This is cow-faced pose (\emph{gomukhāsana}), which [looks] like a cow's face.
 \end{translation}
@@
 \begin{translation}[hp01_021]
- Fixing one foot on one thigh and placing the [other] thigh on the other foot is called \emph{vīrāsana}.
+ Fixing one foot on one thigh and placing the [other] thigh on the other foot is called hero pose (\emph{vīrāsana}).
 \end{translation}
@@
 \begin{translation}[hp01_022]
-Knowers of yoga know that \emph{kūrmāsana} arises by carefully blocking the anus with the ankles crossed.%??Tweak note to reflect change to crossed. 
+Knowers of yoga know that the tortoise pose (\emph{kūrmāsana}) arises by carefully blocking the anus with the ankles crossed.%??Tweak note to reflect change to crossed. 
 \end{translation}
@@
 \begin{translation}[hp01_023]
-[The yogi] correctly assumes \emph{padmāsana}, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is \emph{kukkuṭāsana}.
+[The yogi] correctly assumes lotus pose, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is the wild cock pose (\emph{kukkuṭāsana}).
 \end{translation}
@@
 \begin{translation}[hp01_024]
-While maintaining \emph{kukkuṭāsana}, [the yogi] binds the neck with the hands and lies like a tortoise on his back. This is \emph{uttānakūrmāsana}.
+While maintaining the wild cock pose, [the yogi] binds the neck with the hands and lies [on his back] upturned like a tortoise. This is the upturned tortoise (\emph{uttānakūrmaka}).
 \end{translation}
@@
 \begin{philcomm}[hp01_024]
-The oldest dated manuscript, \etaOne, has \emph{kukkuṭāsanavat kṛtvā}, which is a simpler alternative to the widely attested reading \emph{kukkuṭāsanabandhasthaḥ}, which we have accepted.
-Since there is no known source for this verse other than the \emph{Haṭhapradīpikā}, we can only assume that the reading of \etaOne\ was an isolated attempt to simplify the syntax.%??JM: I guess we discussed this before, but are we not going to translate āsana names?
-%JB if we do, we should put the Sanskrit in round brackets as many enthusiasts use the Sanskrit names.
+The oldest dated manuscript, \etaOne, has \emph{kukkuṭāsanavat kṛtvā}, which is a simpler alternative to the widely attested reading \emph{kukkuṭāsanabandhasthaḥ} (including \alphaTwo\ and \alphaThree), which we have accepted.
+Since there is no known source for this verse other than the \emph{Haṭhapradīpikā}, it appears that the reading of \etaOne\ was an isolated attempt to simplify the syntax.
 \end{philcomm}
@@
 \begin{translation}[hp01_025]
-Clasping the big toes with the hands and performing the action of drawing a bow as far as the ear is called \emph{dhanurāsana}.
+Clasping the big toes with the hands and performing the action of drawing a bow as far as the ear is called the bow pose (\emph{dhanurāsana}).
 \end{translation}
@@
 [The yogi] should hold the right foot, which is placed at the base of the left thigh, with the [hand of] the right arm, which is wrapped around the outside of the knee, and remain [like that] with his body twisted.
-This posture was taught by Matsyendranātha.
+This posture was taught by the revered Matsyendranātha.
 \end{translation}
@@
 \begin{translation}[hp01_028]
-[The yogi] should stretch out the legs on the ground [as straight] as sticks, hold the toes of both feet with the hands, place the forehead upon the knees and remain thus.
-They call this the back-stretch (\emph{paścimatānam}).%??
-āsana name translated here.
+[The yogi] should stretch out the legs on the ground [as straight] as a stick, hold the toes of both feet with the hands, and practise with the forehead placed on the knees.
+They call this the back-stretch (\emph{paścimatānam}).
 \end{translation}
@@
 \begin{translation}[hp01_029]
-This back-stretch is the foremost among \emph{āsana}s.
-It makes the breath flow to the rear (i.e.~in the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
+Foremost among \emph{āsana}s, the back-stretch thus makes the breath flow to the rear (i.e.~in the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
 \end{translation}
@@
  The use of the word \emph{paścima} to mean the central channel is found at \emph{Yogabīja} 95 (\emph{paścimamārgataḥ}), 108 (\emph{paścime pathi}), 117 (\emph{paścimadvāramārgeṇa}) and 121 (\emph{paścimaṃ}).
-Cf.~the usages of \emph{paścima\-mārga} in \sl{Dattātreyayogaśāstra} 140, \emph{Yogabīja} 95 and \emph{Śivasaṃhitā} 3.110 (from which this verse is likely to be derived).
+Cf.~the usages of \emph{paścima\-mārga} in \sl{Dattātreyayogaśāstra} 140 and \emph{Śivasaṃhitā} 3.110 (from which this verse is likely to be derived).
 Brahmānanda understands \emph{paścima} as referring to the Suṣumṇā (\emph{Jyotsnā} 1.29): \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumṇāmārgeṇa vahatīti paścimavāhī}.
@@
 \begin{philcomm}[hp01_030]
-%??JM: authorial reworking? 
+%
 The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77), which are themselves derived from earlier Vaiṣṇava sources.
@@
 \begin{translation}[hp01_032]
-Lying on one's back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] any \emph{āsana} and relaxes the mind.
+Lying with one's back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] any \emph{āsana} and calms the mind.
 \end{translation}
@@
 
-In the first and third verse quarters, Svātmārāma may have rewritten \emph{Śivasaṃhitā} 3.96 to include the information that it was Śiva (\emph{śivena}) who taught the eighty-four \emph{āsana}s, whereas in the source Śiva is himself speaking.
+In the first and third verse quarters, Svātmārāma appears to have rewritten \emph{Śivasaṃhitā} 3.96 to include the information that it was Śiva (\emph{śivena}) who taught the eighty-four \emph{āsana}s, whereas in the source Śiva is himself speaking.
 Svātmārāma also changes the meaning of the second half of the verse, as the \emph{Śivasaṃhitā} states that Śiva picked out the four best postures and taught them, whereas in the \emph{Haṭhapradīpikā} it reads as though Svātmārāma himself is responsible for picking out the four best postures and teaching them.
@@
 
-[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, hold the face and chest together and the body erect, and [remain] stock-still, his senses restrained, gazing between the brows with his eyes unmoving.
+[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, focus the mind, hold the body erect and [remain] motionless, his senses restrained, gazing between the brows with his eyes unmoving.
 This, which breaks open the door to liberation, is called the adept’s pose.
@@
 \tl{\var{7d hy etan ] H; caitan VAGBGLGPT, etan GPkHS, tv etat Y • °janakaṃ ] °navidhau A, °jananaṃ GB, °nakaraṃ Y • procyate ] idaṃ bhavet Y (unm.)} ?? check\\!}
+%?? retain variants for athaikam āsyahṛdaye dhṛtvā samaṃ vigraham
 \end{versinnote}
@@
 \begin{testimonia}[hp01_035]
-\emph{Yogacintāmaṇi} f. 84v–85r (attr. \emph{Pavanayogasaṅgraha})
+\emph{Yogacintāmaṇi} f.~84v–85r (attr.~\emph{Pavanayogasaṅgraha})
 
@@
 \begin{philcomm}[hp01_035]
+%?? Rewrite the note to reflect the adopted reading ekahṛdayo
+
+The adopted reading \emph{ekahṛdayo} is  supported by two manuscripts of the \emph{Haṭhapradīpikā} (\getsiglum{J2,M1}) and is close to the \textalpha\ reading \emph{ekahṛdaye}, which is also attested by the six-chapter \emph{Viveka\-mārtaṇḍa} and the \emph{Yoga\-cintāmaṇi}, ..
+
+which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}. In this case, \emph{ekahṛdayaḥ} would perhaps qualify the yogi as having a mind focused on one thing. 
 The second verse quarter has many variations in the \emph{Haṭhapradīpikā} manuscripts, as well as in the manuscripts of the sources and testimonia.
-\getsiglum{N3} of the \textalpha\ group has \emph{athaikam ekahṛdaye dhṛtvā} and most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}.
+\alphaOne\ has \emph{athaikam ekahṛdaye dhṛtvā} and most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}.
 The repetition of \emph{eka} appears to be a dittographical error and \emph{hṛdaye} does not make sense without an object of \emph{dhṛtvā} (i.e., having held what on the chest?).
 In the adopted reading \emph{athaikam āsyahṛdaye dhṛtvā}, which is attested by \etaOne, the oldest dated manuscript, the compound \emph{āsyahṛdaye} can be understood with the gerund \emph{dhṛtvā} and \emph{ekam} in the sense of `having held the face and chest together.' This alludes to the practice of the Jālandhara lock, in which the chin is placed on the chest.
-This meaning is more clearly seen in the \emph{Jyotsnā}'s version of the second verse quarter, \emph{hṛdaye kṛtvā hanuṃ susthiram} (`having put the jaw formly on the chest').
+This meaning is more clearly seen in the \emph{Jyotsnā}'s version of the second verse quarter, \emph{hṛdaye kṛtvā hanuṃ susthiram} (`having put the jaw firmly on the chest').
 The compound \emph{āsyahṛdaye} is supported by other manuscripts, such as \getsiglum{N10,P1,P6,J16} and the variants \emph{asyahṛdaye} and \emph{asyahṛdayaṃ} also occur (e.g., \getsiglum{J7,J12,A1,V4,V18,V16,P8,P9}).
-Another possible reading is \emph{ekahṛdayo}, which is not supported by manuscripts of the \emph{Haṭhapradīpikā} in the most important stemmatic groups but is in two other manuscripts (\getsiglum{J2,M1}), the six-chapter \emph{Vivekamārtaṇḍa} and the \emph{Yogacintāmaṇi}, which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}.
-In this case, \emph{ekahṛdayaḥ} would perhaps qualify the yogi as having a mind focused on one thing.
+
 The other well-attested reading, \emph{athaikaṃ eva niyataṃ}, was an attempt to fix the problem of \emph{hṛdaye} by replacing it with \emph{niyatam}, which must be read with \emph{meḍhre pādam athaikaṃ} (`having fixed one foot on the penis').
@@
 
-Place the left heel on the penis and put the other heel on top: this is \emph{siddhāsana}.\\
+Place the left heel on the penis and put the other heel on top: this is the adept's pose (\emph{siddhāsana}).\\
 
@@
 \begin{translation}[hp01_037]
-Some call this \emph{siddhāsana}, others know it as \emph{vajrāsana}, a few say it is \emph{muktāsana} and some call it \emph{guptāsana}.
+Some call this the adept's pose (\emph{siddhāsana}), others know it as the thunderbolt pose (\emph{vajrāsana}), a few say it is the pose of the liberated (\emph{muktāsana}) and some call it the secret pose (\emph{guptāsana}).
 \end{translation}
@@
 %\emph{iva} or \emph{eva}?
-\emph{iva} does work — like \emph{siddhāsana}, \emph{mitāhāra} and \emph{ahiṃsā} are the best, but for it to work properly \emph{mitāhāra} and \emph{ahiṃsā} should be accusative.
-V19 has acc + \emph{iva}, which seems best, especially with \emph{siddhāḥ viduḥ}, but this might be a correction as V19 often corrects.
-However, one old KDham BORI (?) ms (pha, 1695 CE) has it, as does Jyotsnā, so adopt.
+\emph{iva} does work — like \emph{siddhāsana}, \emph{mitāhāra} and \emph{ahiṃsā} are the best, but %for it to work properly \emph{mitāhāra} and \emph{ahiṃsā} should be accusative.
+V19 has acc + \emph{iva}, which seems best, %especially with \emph{siddhāḥ viduḥ}, but this might be a correction as V19 often corrects.
+However, one old KDham BORI (?) ms %(pha, 1695 CE) has it, as does Jyotsnā, so adopt.
 %In pāda d V19 has \emph{siddhāsanam idaṃ viduḥ}, but the reading of all other mss is preferable.
-% JB: The reading we have adopted is well attested by the new collation of manuscripts. Therefore, I've ignored the above comments.
+% JB: The reading we have adopted is well attested by the new collation of manuscripts. Therefore, I've ignored the above %comments.
 
@@
 \begin{translation}[hp01_039]
-Of the eighty-four postures, one should regularly practise just \emph{siddhāsana}, in the same way that of the 72,000 channels one should practise Suṣumṇā.
+From among the eighty-four postures, one should regularly practise just \emph{siddhāsana}, in the same way from among the 72,000 channels [one should practise, focusing on] \emph{suṣumṇā}.
 \end{translation}
@@
 The reading \emph{nāḍīnāṃ malaśodhanam/e} is probably a patch as no other texts say that \emph{siddhāsana} clears the channels.
-However, the idea of purifying the channels can be found in other contexts (e.g., \emph{Kumbhakapaddhati} 120) and may hark back to an earlier notion of flushing (\emph{cālana}) the channels (e.g., \emph{Amṛtasiddhi} 11.6).%??
-somewhat haphazardly?
-I don't understand the second part of the note: purifying the channels is found all over the place.
+However, the idea of purifying the channels can be found in other contexts (e.g., \emph{Kumbhakapaddhati} 120) and may hark back to an earlier notion of flushing (\emph{cālana}) the channels (e.g., \emph{Amṛtasiddhi} 11.6).% somewhat haphazardly?
+I don't %understand the second part of the note: purifying the channels is found all over the place.
 %JB: The point is: no other texts say that \emph{siddhāsana} clears the channels.
@@
 
-The notion of \emph{āsana}s causing fatigue (\emph{śrama}) was already suggested in the verse on the corpse pose (1.32).
+The notion of \emph{āsana}s causing fatigue (\emph{śrama}) was mentioned earlier in the verse on the corpse pose (1.32).
 \end{philcomm}
@@
 % adopt dṛḍhaṃ 
-
+%% JB : only when the adept's pose is firmly bound.
 %\begin{sources}[hp01_042]
@@
 \tl{\var{8d etad ] antar° T • °vikārahāri yamināṃ ] VAT; °vikāranāśanakaraṃ GPk Y, °vikārakaṃdadamanaṇ GB , °vināśakāri yamināṃ GL GP, °vighātahāri yamināṃ Tvl • °nāśanakaraṃ G}\\!}
+%?? Variants not really needed as HP text is solid.
 \end{versinnote}
@@
 \tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ ||\\!}
+%?? Variants for uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ ?
 \end{versinnote}
@@
 
-%\begin{versinnote}
-%\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
-%\tl{uttabhya cibukaṃ vakṣasy āsthāpya \textbf{pavanaṃ śanaiḥ} || 36 ||\\+}
-%\tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
-%\tl{yathāśaktyaiva paścāt tu recayet \textbf{pavanaṃ śanaiḥ} || 37 ||\\+}
-%\tl{idaṃ padmāsanaṃ proktaṃ sarvavyādhivināśanam |\\!}
-%\end{versinnote} ??JM: I've removed this, no need to repeat?
-
 The manuscript readings with \emph{vakṣa sthāpayet} (\getsiglum{J7,V3,J8,J10,J17,N17}) or something similar (\getsiglum{V1,W4}) do not offer a solution to the incomplete syntax so do not indicate that Svātmārāma rewrote \emph{Dattātreyayogaśāstra} 36 so that he could omit \emph{Dattātreyayogaśāstra} 37.
@@
 \begin{quote}
-Supporting both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.%?? JM: supporting?
+Pushing against both roots of the front teeth on the left and right with the tongue [\ldots] — this fixation of the tongue has to be understood from the mouth of the teacher.
 
@@
 
-Only [the teaching on the lotus pose] taught next is approved by me.  
-\end{translation}
-% ?? JM: surely "taught second"?
-%JB the second description follows this comment. So, 'next' makes good sense.
+Only the second teaching [on the lotus pose] is approved by me.  
+\end{translation}
 
@@
 \tl{\var{36d muñcan prāṇamupaiti bodhamatulaṃ śak- tiprabhāvān naraḥ ] U; prāṇaṃ muṃcati bodham eti śanakaiḥ proktaprabhāvād ataḥ V, pāṇaṃ muṃcati bodham eti śanakaiḥ śaktiḥ prabhāvād ataḥ A, prāṇaṃ muṃcati yāti bodham amalaṃ śaktipradhānoditaḥ GB, muṃcan prāṇam upaiti bodham atulaṃ śaktiprabhāvād ataḥ GLGP, muñcan prāṇam upaiti bodham atulaṃ śaktiprabodhān naraḥ GPk, prāṇaṃ muñcati bodhameti śanakai (ścu?śśa)[sic] ktiprabodhān naraḥ T, muñcan prāṇam upaiti bodham akhilāṃ śaktiṃ prabhāvād ataḥ Tvl}\\!}
+%?? Retain variants for dhyānaṃś ca tac cetasi, muñcan and proccālayan
 \end{versinnote}
@@
 We are taking \emph{dhyānaṃ} with \emph{sannidhāya}, but this renders \emph{tat} problematic because it has no clear referent.
-In the source text, the \emph{Vivekamārtaṇḍa, tat} appears to refer to the \emph{mokṣadvāra} broken by Kuṇḍalinī, which is mentioned in the previous verse.
-
-The two participles \emph{proccālayan} and \emph{muñcan} imply that the two things are happening at the same time, which is surprising but perhaps possible.%Or one might infer that `and' (\emph{ca}) was supposed to be supplied and understand both actions as sequential.
-??
-JM: but can wouldn't necessarily imply that they are sequential.
-%JB ?
+In the source text, the \emph{Vivekamārtaṇḍa, tat} appears to refer to the \emph{mokṣadvāra} broken by \emph{kuṇḍalinī}, which is mentioned in the previous verse.
+
+The two participles \emph{proccālayan} and \emph{muñcan} imply that the two things are happening at the same time, which is surprising but perhaps possible.
 
@@
 
-In the third verse quarter, the best \textalpha\ manuscript reads \emph{māruto mriyate yas tu}, which does not make sense, but if one accepts \emph{pūrayet} in the second verse quarter, one could emend \textalpha's reading to \emph{māruto mriyate yasya}, which makes good sense (i.e., `the yogi whose breath dies is undoubtedly liberated').
-In the same vein, J5 also has the plausible reading \emph{mārutaṃ mārayet yas tu}.
+In the third verse quarter, \alphaOne\ reads \emph{māruto mriyate yas tu}, which does not make sense, but if one accepts \emph{pūrayet} in the second verse quarter, one could emend \textalpha's reading to \emph{māruto mriyate yasya}, which makes good sense (i.e., `the yogi whose breath dies is undoubtedly liberated').
+In the same vein, \alphaTwo\ also has the plausible reading \emph{mārutaṃ mārayet yas tu}.
 %but J5 is contaminated (so the Gr3 reading dhārayet is more reliable).
@@
 
-Spreading the fingers and keeping the mouth wide open mimic a lion, and this is depicted in some iconography of yoganarasiṃha (for example, Yoga Narasimha, Vishnu's Man-Lion Incarnation, Samuel Eilenberg Collection, Bequest of Samuel Eilenberg, 1998, Accession Number: 2000.284.4.
+Spreading the fingers and keeping the mouth wide open mimic a lion, and this is depicted in some iconography of Yoganarasiṃha (for example, Yoga Narasimha, Vishnu's Man-Lion Incarnation, Samuel Eilenberg Collection, Bequest of Samuel Eilenberg, 1998, Accession Number: 2000.284.4.
 https://www.metmuseum.org/art/collection/search/39251).
 
-The \emph{Yogacintāmaṇi} attributes its citation of these verses on \emph{siṃhāsana} to Yājñavalkya.
+%The \emph{Yogacintāmaṇi} attributes its citation of these verses on \emph{siṃhāsana} to Yājñavalkya.
 Its citation does not include \emph{Haṭhapradīpikā} 1.52cd, which affirms that 1.52cd is not from the \emph{Yogayājñavalkya} (or \emph{Vasiṣṭhasaṃhitā}).
-As far as we are aware, there is no source for the hemistich mentioning the three locks, so it may have been composed by Svātmārāma or borrowed from a lost work.
-%\textalpha \alphathree %??JM: I don't understand this.
-Don't we know already that it's not in the YY or VS?
-%JB: I think the comment is trying to explain that "there is no source for the hemistich mentioning the three locks".%OK, but the first two sentences of the para seem redundant to me (they are implied by the last)
+As far as we are aware, there is no source for the hemistich (1.52cd) mentioning the three locks, so it may have been composed by Svātmārāma or borrowed from a lost work. 
 \end{philcomm}
@@
 
-%(ed.\ Maximilian Mehner and Jürgen Hanneder, Marburg 2019).
 \end{philcomm}
@@
 \begin{translation}[hp01_055]
-When the great yogi does not become tired on adopting the \emph{āsana}s in this way, he should  practise the breath techniques of seals and so forth, from which purification of the channels arise.%??JM: work on the breath using seals
-%JB: I think our original translation is better.
-%JM: but not all mudrās work on the breath; also the compound cannot mean "a breathing technique along with mudrās etc." as said in the commentary: "along with" isn't there.
+When the great yogi does not tire from adopting the \emph{āsana}s in this way, he should  practise the breath techniques with seals and so forth, from which purification of the channels arises.%??
+Write to Haru to ask about nāḍiśuddhiṃ as a b/v
 \end{translation}
@@
 The second hemistich can be interpreted in different ways.
-One possibility is to understand \emph{nāḍiśuddhiṃ} as a bahuvrīhi qualifying \emph{mudrādipavanakriyām} in the sense that the yogi should practise a breathing technique, along with \emph{mudrā}s and other mechanisms, such as locks (\emph{bandha}), from which purification of the channels arise.
+One possibility is to understand \emph{nāḍiśuddhiṃ} as a \emph{bahuvrīhi} qualifying \emph{mudrādipavanakriyām} in the sense that the yogi should practise the breathing techniques by way of the relevant \emph{mudrā}s and locks (taught in the third chapter), from which purification of the channels arise.
 Alternatively, one could separate \emph{mudrādi} from \emph{pavanakriyām} and understand three different techniques here, namely, the practice of purifying the channels (perhaps by the alternative nostril method mentioned at the beginning of the second chapter), the \emph{mudrā}s and the breathing techniques of \emph{prāṇāyāma}.
@@
 
-The metre of 1.55c is a \emph{ra-vipulā} with \emph{nāḍi}.%??JM:are we planning to note all vipulās?
-%JB Maybe Sven can use that software to identify them once we've finalised the text. It might be useful for future editorial work, particuarly on related works.
+% ?? Mention that the mudrās practised with the breathing techniques are the bandhas (2.44-45) 
+%The metre of 1.55c is a \emph{ra-vipulā} with \emph{nāḍi}.%
 \end{philcomm}
@@
 This is true, there is no doubt.
-In this system, [the practice] should not be given to one who wears robes and is devoted to his genitals and stomach.%??
-JM: In the  DYŚ I have lust and gluttony.
-This sounds a bit odd.
+In this system, [the practice] should not be given to one who wears robes and is devoted to sex and food.
 \end{translation}
@@
 \tl{\var{47d janān ] janāḥ M1βV, janāt J, janā W1 • kila ] sadā M1AM2, iha YSS • vañcayanti janān kila ] vañcayaty akhilān janān YCM}\\!}
+%?? Simplify?
 \end{versinnote}
@@
 1.55*1–2 are omitted from the \textalpha, \textgamma, \textdelta\ and \textepsilon\ groups, so it is likely these verses were not in the earliest versions of the \emph{Haṭhapradīpikā}.
-In fact, it is likely that both were added (perhaps initially as marginal notes) to elaborate on the word \emph{kriyā} in 1.55d.
+In fact, it appears that both were added (perhaps initially as marginal notes) to elaborate on the word \emph{kriyā} in 1.55d.
 Both verses are similar to verses from the \emph{Dattātreyayogaśāstra} (cited as the source).
@@
 Leipzig: Harrassowitz 1910, p.\,90 [A9]).
-However, the editor did not provide a compelling reason to regard it as
-unoriginal except only the fact that it is not transmitted in all manuscripts.
+However, the editor did not provide a compelling reason to regard it as unoriginal except only the fact that it is not transmitted in all manuscripts.
 What prevents further investigation of the matter is the lack of Kashmirian manuscripts for the \emph{Śāntiśataka} and its compilatory character: one quarter of the material is identical with Bhartṛhari's \emph{Vairāgyaśataka}.
@@
 \begin{translation}[hp01_055_4]
-Realisation from scripture, one's own guru and oneself and the cessation of mind; all these methods have been combined and taught by the wise in this tradition.%?? repeat "realisation" like the text?
+Realisation from scripture, realisation from one's own guru, realisation from oneself and the cessation of mind; all these methods have been combined and taught by the wise in this tradition.%
 \end{translation}
@@
 \begin{philcomm}[hp01_055_4]
-Verse 1.55.4 is in some of the \textdelta\ manuscripts and is quoted in \emph{Yogacintāmaṇi}.
+Verse 1.55.4 is in some of the \textdelta\ manuscripts and is quoted in \emph{Yogacintāmaṇi} with attribution to the \emph{Haṭhapradīpikā}.
 
@@
 \begin{translation}[hp01_056]
-The various \emph{āsana}s, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then the fusion of the mind with the internal resonance are the sequence of practice in Haṭha.%??JM: Posture, multifarious breath retention, the bodily attitude called seal, then concentration on the internal sound is the sequence of practice in Haṭha.
+%The various \emph{āsana}s, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then concentrating on the internal resonance is the sequence of practice in Haṭha.%??JM:
+Posture, manifold breath retention, the bodily technique called seal, then concentration on the internal sound is the sequence of practice in Haṭha.
 \end{translation}
@@
 \end{translation}
-% adopt yogī?
 
@@
 \tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||\\!}
+%?? variants on yogī??
 \end{versinnote}
@@
 \begin{translation}[hp01_058]
-When very unctuous and sweet food that is without a quarter portion is eaten for love of Śiva, it is called a restricted diet (\emph{mitāhāra}).
+When very unctuous and sweet food is eaten for love of Śiva, leaving a quarter [of the stomach] empty (\emph{caturthāṃśavivarjitaḥ}), it is called a restricted diet (\emph{mitāhāra}).
 \end{translation}
@@
 
-The idea of not eating the fourth portion of one’s food (\emph{caturthāṃśavivarjitaḥ}) can be found in older yoga texts, such as \emph{Dharmaputrikā} 1.51-52:
+The expression `lacking a fourth part' \emph{caturthāṃśavivarjitaḥ}) is somewhat vague but probably refers to the idea of leaving a quarter of one's stomach empty, which is stated more clearly in the \emph{Dharmaputrikā} 1.51-52:
 
@@
 
-And, as noted by Brahmānanda in \emph{Jyotsnā} 1.58, the idea also occurs in an āyurvedic work called the \emph{Aṣṭāṅgahṛdayasaṃhitā, Sūtrasthāna}, 8.46cd–47ab:
+And, as noted by Brahmānanda in \emph{Jyotsnā} 1.58, this idea also occurs in an āyurvedic work called the \emph{Aṣṭāṅgahṛdayasaṃhitā, Sūtrasthāna}, 8.46cd–47ab:
 
@@
 \tl{annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet |\\+}
-\tl{āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet ||\\!}%??JM: but both these examples say to half-fill the stomach with food and fill a quarter with water.
-It's odd how in the HP/GŚ the stomach isn't mentioned.
-It seems to mean eat 3/4 of the amount eaten in a normal diet.
+\tl{āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet ||\\!}% ??translate citations
 \end{versinnote}
@@
 \begin{translation}[hp01_059]
- Pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink, flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like: they say that this is unwholesome.
+ Pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink, flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like: they say that such [food] is unwholesome.
 \end{translation}
@@
 \begin{philcomm}[hp01_059]
-The compound \emph{kaṭvamla°} (1.59a) is better than \emph{kaṭvāmla°} and it is well attested by manuscripts of the \emph{Haṭhapradīpikā}, as well as in lists of tastes and types of foods in other texts.
-
-On the meaning of \emph{uṣṇa} (1.59a) in relation to food, see Meulenbeld 1974: 254 fn. 13:
+%The compound \emph{kaṭvamla°} (1.59a) is better than \emph{kaṭvāmla°} and it is well attested by manuscripts of the \emph{Haṭhapradīpikā}, as well as in lists of tastes and types of foods in other texts.%JM: why this comment?
+
+On the meaning of \emph{uṣṇa} (1.59a) in relation to food,  Meulenbeld writes (1974: 254 fn. 13):
 \begin{quote}
@@
 The spelling \emph{°harīta} was probably adopted for metrical reasons.
-In some Nighaṇṭus (`thesauruses'), \emph{°haritaśāka°} is glossed as horseradish (\emph{śigru}).
+In some Nighaṇṭus, \emph{°haritaśāka°} is glossed as horseradish (\emph{śigru}).
 \begin{quote}
@@
 
-The term \emph{kulattha} is a kind of pulse, translated by Dominik Wujastyk (1998: 77) as `poor man's pulse.'
+The term \emph{kulattha} means a kind of pulse, translated by Dominik Wujastyk (1998: 77) as `poor man's pulse.'
 % Dominik Wujastyk, Roots of Ayurveda, Penguin Books, 1998.
@@
 
-The term \emph{hiṅgu} is Asafoetida (Nadkarni 1926: 360–361). As to why it might be considered unwholesome, the following comments by Nadkarni give some indication:
+The term \emph{hiṅgu} means Asafoetida (Nadkarni 1926: 360–361). As to why it might be considered unwholesome, the following comments by Nadkarni give some indication:
 \begin{quote}
@@
 \begin{translation}[hp01_060]
-One should know unwholesome food to be that has been reheated, is dry, too salty or sour, contains an excess of leafy vegetables that are hard to chew, [or] is spoiled.
+One should know food to be unfit if it has been reheated, is dry, too salty or sour, contains an excess of leafy vegetables that are hard to chew, [or] is spoiled.
 \end{translation}
@@
 There are various possible meanings of \emph{utkaṭa}.
-According to some Nighaṇṭus, the word \emph{utkaṭā} can mean pepper  (e.g., \emph{Rājanighaṇṭu} 5.16 \emph{pārvatī śailajā tāmrā lambabījā tathotkaṭā}) and, according to the dictionary (s.v.
-Monier Williams), \emph{utkaṭa} can refer to Saccharum Sara and \emph{utkaṭā} also to Laurus Cassia (cinnamon).
+According to some Nighaṇṭus, the word \emph{utkaṭā} can mean pepper  (e.g., \emph{Rājanighaṇṭu} 5.16 \emph{pārvatī śailajā tāmrā lambabījā tathotkaṭā}) and, according to Monier Williams, \emph{utkaṭa} can refer to Saccharum Sara and \emph{utkaṭā} also to Laurus Cassia (cinnamon).
 
 However, \emph{utkaṭa} can be an adjective that means ‘abounding in’ at the end of a compound.
-Since this verse consists of many adjectives describing food that is unwholesome, it is likely that \emph{kadaśanaśākotkaṭaṃ} was intended as an adjectival \emph{tatpuruṣa}, in which case it means ‘[food] full of vegetables' \emph{śākotkaṭa} that are `bad food' or perhaps, `bad eating' \emph{kadaśana} in the sense of hard to chew.
+Since this verse consists of many adjectives describing food that is unwholesome, it is likely that \emph{kadaśanaśākotkaṭaṃ} was intended as an adjectival \emph{tatpuruṣa}, in which case it means ‘[food] full of vegetables' \emph{śākotkaṭa} that are `bad food' or, perhaps, `bad eating' (\emph{kadaśana}) in the sense of hard to chew.
 \end{philcomm}
@@
 \begin{translation}[hp01_061]
-Similarly a saying by Goraksa:
-One should avoid places with bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.%??maybe prānta = borderlands?
-In Hindi it means region but maybe not Skt
+In the same vein there is a saying by Goraksa:
+One should avoid places near bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.
 \end{translation}
@@
 \begin{translation}[hp01_062]
-Pure food with wheat, rice, śāli rice, barley, sixty-day \emph{śāli} rice, milk, ghee, cream, fresh butter, ground sugar and honey, dried ginger, fruit of the snake gourd and so forth, the five vegetables, mung beans and so on, and rain water are wholesome for the best of ascetics.%??JM:How about "Pure grains such as wheat..."?
-and semi-colons after each group.
+The pure grains that are wheat, rice, śāli rice, barley, sixty-day \emph{śāli} rice; milk, ghee, cream, fresh butter, ground sugar and honey; dried ginger, fruit of the snake gourd and so forth; the five vegetables; mung beans and so on; and rain water.
+[These] are wholesome for the best of ascetics.%
 \end{translation}
@@
 Sharma 2018 vol.
-1: 434)
+1: 434).
 \end{quote}
@@
 Ayurvedic sources indicate more clearly that \emph{divyodaka} was understood as rainwater.
-In a section on types of water (\emph{jalavarga}) in the \emph{Sūtrasthāna} of the \emph{Carakasaṃhitā} (27.196–224), rainwater is referred to as `\emph{divyaṃ udaka}' (1.27.198) in a discussion of the properties of water that has fallen from the sky.
+In a section on types of water (\emph{jalavarga}) in the \emph{Sūtrasthāna} of the \emph{Carakasaṃhitā} (27.196–224), rainwater is referred to as `\emph{divyaṃ udakam}' (1.27.198) in a discussion of the properties of water that has fallen from the sky.
 The compound \emph{divyodaka} is used in other Āyurvedic works to refer to the use of rainwater in recipes and treatments  (e.g., \emph{Aṣṭāṅgahṛdaya} 8.42–43).
@@
 \tl{abhyāsāt siddhim āpnoti yoge sarvo 'py atandritaḥ || 40 ||\\!}
+%?? variants on yoge sarvo 'py atandritaḥ?
 \end{versinnote}

### 2024-02-20 latex/HP2_comm.tex
new files

\input{PreambleComm}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

%\chapter*{Translation \& philological commentary}
%%%%%%%%%%
\subsection*{2.1}
\begin{translation}[hp02_001] 
Now, when [his] posture is steady, the disciplined yogi whose diet is good and measured should practise breath control in the way taught by [his] teacher.
\end{translation}

\begin{sources}[hp02_001]
\end{sources}

\begin{testimonia}[hp02_001]
\emph{Haṭharatnāvalī} 3.78

\begin{versinnote}
\tl{atha prāṇāyāmaḥ--\\+}
\tl{athāsane dṛḍhe yogī vaśī hitamitāśanaḥ |\\+}
\tl{gurūpadiṣṭamārgeṇa prānāyāmān samabhyaset ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 36.1

\begin{versinnote}
\tl{atha nāḍīśuddhiḥ--\\+}
\tl{tathā coktaṃ yogacandrikāyām--\\+}
\tl{athāsane dṛḍhībhūte vaśī hitamitāśanaḥ |\\+}
\tl{gurūpadiṣṭamārgeṇa prānāyāmān samabhyaset ||\\!}
\end{versinnote}

%Cf. Yogamārgaprakāśikā

%\begin{versinnote}
%\tl{athāsane dṛḍhībhūte suśobhanamaṭhe yadā |\\+}
%\tl{guruṃ natvā śivaṃ caiva prāṇāyāmaṃ tato 'bhyaset || 47 ||\\!}
%\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_001]
%\emph{dṛḍho yogī} seems unlikely as \emph{dṛḍha} usually qualifies a technique, the body, etc. (not worth commenting on as alpha has dṛḍhe).
The \emph{Jyotsnā} (2.1) has the plural \emph{prāṇāyāmān}, which is supported by some manuscripts of the \textbeta, \textgamma\ and \textepsilon\ groups. The plural is possible here as it could refer to the different techniques of retention (\emph{kumbhaka}) taught later in the chapter. This is how Brahmānanda understands it (\emph{prāṇāyāmān vakṣyamāṇān}). The variation between singular and plural recurs through this chapter, and we have followed the \textalpha\ readings, which make good sense. In this case, since the verse is introducing the topic of breath control, the more general sense of the singular is appropriate.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.2}
\begin{translation}[hp02_002]
When the wind moves, everything moves and, when it is still, everything is still, so the yogi attains motionlessness through restraining the breath.
\end{translation}

\begin{sources}[hp02_002]
\emph{Vivekamārtaṇḍa} 71

\begin{versinnote}
\tl{cale vāte calaṃ sarvaṃ niścale niścalaṃ tathā |\\+}
\tl{yogī sthāṇutvam āpnoti tato vāyunibandhanāt ||\\+}
\tl{\var{71d vāyunibandhanāt ] VT; vāyuṃ nibaṃdhayet A, vāyuṃ nirundhayet G}\\!}
\end{versinnote}

Cf.~\emph{Amanaska} 2.92

\begin{versinnote}
\tl{citte calati saṃsāro 'cale mokṣaḥ prajāyate |\\+}
\tl{tasmāc cittaṃ sthirīkuryād audāsīnyaparāyaṇaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_002]
\emph{Haṭharatnāvalī} 3.79

\begin{versinnote}
\tl{cale vāte calaṃ cittaṃ niścale niścalaṃ tathā |\\+}
\tl{yogī sthāṇutvam āpnoti tato vāyuṃ nirundhayet ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~17r

\begin{versinnote}
\tl{tathā ca skandapurāṇe--\\+}
\tl{cale vāte calaṃ cittaṃ niścalaṃ niścale tathā |\\+}
\tl{yogī sthāṇutvam āpnoti tato vāyuṃ nirodhayet ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_002]
%The \textalpha\ manuscripts, as well as those of various other important groups, have \emph{sarvaṃ} instead of \emph{cittaṃ} in the first verse quarter. The stemma aside, the former makes better sense as a general statement (i.e., when the wind moves, everything moves, etc.). The reading \emph{citta} only seems to make sense if one understands \emph{sthāṇutvam} as referring to \emph{samādhi} (i.e., `motionlessness [of the mind]'). As far as we know, such a meaning of \emph{sthāṇutvam} is not attested elsewhere. Nonetheless, Brahmānanda glosses \emph{sthāṇutvam} as \emph{kāṣṭhavat} (`like a piece of wood'), implying that it refers to a yogi in \emph{samādhi} as this simile is found elsewhere in contexts of meditative absorption (e.g., \emph{Amanaska} 1.27).
%(JM: but we have sarvam now and cittam is clearly wrong; I think this note is unnecessary — sthāṇu can mean a piece of wood or a post and is a byword for motionlessness (but never means samādhi), like kāṣṭha, and we don't want to put people off reading the notes by including too many unedifying ones.) 
% JB: Since we often read against the stemma, I dont think it's very convincing to simply say we are reading with the stemma in cases where there are two well-attested plausible readings. In this particular case, cittaṃ is a plausible reading that's well attested among the collated mss., so I think we should firstly bring it to the reader's attention and then explain why we think the adopted reading is better on semantic grounds. Perhaps, my attempts to do the latter (by tidying up a previous note) can be improved?
%JM: should we mention that sthāṇu can mean Śiva?
In the second verse quarter, most of the \textalpha, \textbeta\ and \textgamma\ manuscripts have \emph{dṛḍhabandhanaṃ} instead of the reading we have adopted, \emph{niścalaṃ tathā}, which is supported by some \textbeta, \textdelta\ and \texteta\ manuscripts, and also the source text (i.e. the \emph{Vivekamārtaṇḍa}) and testimonia. The reading \emph{dṛḍhabandhanaṃ} appears to be a dittographical error and does not make sense when read with either \emph{sarvaṃ} (\textalpha) or \emph{cittaṃ} (\textgamma).

As noted by Brahmānanda, in addition to its primary meaning of immobility, \emph{sthāṇutvam} can also mean the state of being Śiva, for whom Sthāṇu is another name.

%The \alpha\ group is split between the reading we have adopted \emph{vāyunibandhanāt} (\alphaThree), which is close to \alphaOne\ (\emph{vāyuṃ nibandhanāt}), and the widely attested readings with a finite verb, including \alphaTwo \ (\emph{vāyuṃ nirundhayet}).

In the fourth verse quarter, most witnesses have something similar to either \emph{vāyuṃ nibandhayet} and \emph{vāyuṃ nirodhayet}, which are likely to have arisen through confusion with 2.3d. The reading we have adopted (\emph{vāyu\-nibandhanāt}) makes better sense with the finite verb in the second verse quarter.%JM: again I wonder whether all this is worth putting in the edition, online or printed. It is our musings that got us to the text we have constituted, but is not very edifying for the reader. JM later note: if any are necessary, it's only the last sentence as the rest is info found in the apparatus
%JB: If edifying means instructive, then I rate our musings highly :) In 2.2b, we are reading against the stemmea (\textalpha\ dṛdhabandhanaṃ) and in 2.2d most witnesses have a finite verb, including \alphaTwo. But perhaps the note can be improved (I dont find it particularly easy to write this one).

\end{philcomm}

%%%%%%%%%%
\subsection*{2.3}
\begin{translation}[hp02_003]
As long as breath is found in the body, there is said to be life. Its leaving is death, so one should restrain the breath.
\end{translation}

\begin{sources}[hp02_003]
\emph{Vivekamārtaṇḍa} 72

\begin{versinnote}
\tl{yāvad vāyuḥ sthito dehe tāvaj jīvitam ucyate |\\+}
\tl{maraṇaṃ tasya niḥkrāntau tato vāyuṃ nirodhayet ||\\+}
\tl{\var{72a sthito ] sthiro A 72b jīvitam ] VAGBGPT; jīvanam GLGPk 72c niṣkrāntau ] VTA; niḥkrāṃtaṃ GBGP, niṣkrāntis GLGPk 72d vāyuṃ nirodhayet ] T; vāyunibandhanāt V, vāyuṃ nibandhayet ATvl, vāyuṃ nirundhayet G}\\!}
\end{versinnote}

Cf.~\emph{Mṛgendratantra} 1.11.20cd--22ab

\begin{versinnote}
\tl{vyāpārād yasya ceṣṭante śārīrāḥ pañca vāyavaḥ ||\\+}
\tl{prāṇāpānād ayas te tu bhinnā vṛtter na vastutaḥ |\\+}
\tl{vṛttiṃ leśān nigadato bharadvāja nibodha me ||\\+}
\tl{vṛttiḥ praṇayanaṃ nāma yat taj jīvanam ucyate |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_003]

\emph{Haṭharatnāvalī} 3.80

\begin{versinnote}
\tl{yāvad vāyuḥ sthito dehe tāvaj jīvitam ucyate |\\+}
\tl{maraṇaṃ tasya niṣkrāntis tato vāyuṃ nirodhayet ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 11.150

\begin{versinnote}
\tl{yāvad vāyuḥ sthito dehe tāvad dehaṃ na muñcati |\\+}
\tl{maraṇaṃ tasya niṣkrāntis tato vāyuṃ nirundhayet ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_003]
%Both \emph{jīvitaṃ} and \emph{jīvanaṃ} are possible in the second verse quarter. The former is well supported by the manuscripts of several important groups (\alphaOne's \emph{jītaviṃ} seems to be the result of a transposition of the \emph{vi} and \emph{ta}), the source text (the \emph{Vivekamārtaṇḍa}) and the testimonia.

We believe that \emph{nirundhayet} was probably the original reading and that it was changed to the more correct form of \emph{nirodhayet} and, in some cases, confused with the final verse quarter of the previous verse and changed to \emph{nibandhanāt} (\textalpha) or \emph{nibandhayet} (\textbeta).%JM: necessary note?
%JB nirundhayet is not supported by alpha. 
% JB I've commmented on jīvitaṃ/jīvanam because it looks like its unsupported by alpha (alpha one has jītaviṃ). I now see that J5 has jīvitaṃ, so we could eliminate the note if we add this to the collation. But a note might still be helpful here because G4 is damaged at this place, so jīvitaṃ is not so well attested and jīvanaṃ is possible. 
\end{philcomm}

%%%%%%%%%%
\subsection*{2.4}
\begin{translation}[hp02_004]
When the channels are full of impurities, the breath does not go into the middle. How would the state beyond mind occur? How would perfection of the body arise?
\end{translation}

%\begin{sources}[hp02_004]
%\end{sources}

\begin{testimonia}[hp02_004]
\emph{Haṭharatnāvalī} 3.81

\begin{versinnote}
\tl{malākulāsu nāḍīṣu māruto naiva madhyagaḥ |\\+}
\tl{kathaṃ syād unmanībhāvaḥ kāyasiddhiḥ kathaṃ bhavet ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_004]
%The reading \emph{kāyasiddhi} in the final verse quarter is supported by manuscripts of the \textalpha, \textbeta, \textepsilon\ and \texteta\ groups whereas other manuscripts have \emph{kāryasiddhi}. In the context of physical yoga, \emph{kāyasiddhi} makes better sense, as \emph{kāryasiddhi} can refer more generally to accomplishing anything.
%JM: necessary for the reader? Goes with the stemma, makes better sense, variant not particularly remarkable. But maybe I have become too parsimony-partial.
% JB. In this case I agree!
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.5}
\begin{translation}[hp02_005]
Only when the entire impure network of channels is cleansed is the yogi able to control the breath.
\end{translation}

\begin{sources}[hp02_005]
\emph{Vivekamartāṇḍa} 76

\begin{versinnote}
\tl{śuddhim eti yadā sarvaṃ nāḍīcakraṃ malākulam |\\+}
\tl{tadaiva jāyate yogī prāṇasaṃgrahaṇe kṣamaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_005]
\emph{Yogacintāmaṇi} f.~90r

\begin{versinnote}
\tl{skandapurāṇe--\\+}
\tl{śuddhim eti yadā sarvaṃ nāḍīcakraṃ malākulam |\\+}
\tl{tadaiva jāyate yogī kṣamaḥ prāṇanibandhane ||\\!}
\end{versinnote}

\emph{Yuktabhavedeva} 7.11
\begin{versinnote}
\tl{śuddhim eti yadā sarvaṃ nāḍīcakraṃ malākulam |\\+}
\tl{tadaiva jāyate yogī prāṇasaṃgrahaṇe kṣamaḥ ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_005]
%We have understood the term \emph{nāḍīcakra} as the network of channels that originate from the bulb in the lower abdomen. This interpretation is consistent with \emph{Sārdhatriśatikālottara} 10.1–3:%JM there are many many possible refs for this, do we need one? And some systems have them coming from the navel, and some from the heart (Shaman's new article).

%\begin{versinnote}
%\tl{nāḍīcakraṃ paraṃ sūkṣmaṃ pravakṣyāmy anupūrvaśaḥ |\\+}
%\tl{nābher adhastād yat kandam aṅkurās tatra nirgatāḥ ||\\+}
%\tl{dvāsaptatisahasrāṇi nābhimadhye vyavasthitāḥ |\\+}
%\tl{tiryag ūrdhvam adhaś caiva vyāptaṃ nābheḥ samantataḥ ||\\+}
%\tl{cakravatsaṃsthitā nāḍyaḥ pradhānā daśa tāsu yāḥ |\\+}
%\tl{iḍā ca piṅgalā caiva suṣumnā ca tṛtīyakā ||\\!}
%\end{versinnote}

\end{philcomm}

%%%%%%%%%%
\subsection*{2.6}
\begin{translation}[hp02_006]
Therefore [the yogi] should regularly practise breath [retention] with a resolute mind, so that the Suṣumṇā is in good condition and the impurities dry up.%pure (sāttvika) doesn't work with impurities (malāḥ); resolute? grounded in resolve?
\end{translation}

\begin{sources}[hp02_006]
Cf. \emph{Gorakṣaśataka} 73cd–74ab

\begin{versinnote}
\tl{prāṇābhyāsas tataḥ kāryo nityaṃ sattvāsthayā dhiyā |\\+}
\tl{suṣumnāṃ layate cittaṃ na ca vāyuḥ pradhāvati ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_006]
%\emph{Jyotsnā}
%\begin{versinnote}
%\tl{prāṇāyāmaṃ tataḥ kuryān nityaṃ sāttvikayā dhiyā |\\+}
%\tl{yathā suṣumṇānāḍīsthā malāḥ śuddhiṃ prayānti ca || 6 ||\\!}
%\end{versinnote}
\emph{Yogakarṇikā} 58 (attr.~to the \emph{Haṭhapradīpa})

\begin{versinnote}
\tl{prāṇāyāmaṃ tataḥ kuryān nityaṃ sāttvikayā dhiyā |\\+}
\tl{suṣumnā cāntarālasthā malāḥ śoṣaṃ prayānti ca ||\\!}
\end{versinnote}

%\emph{Prāṇatoṣaṇī} p.788

%\begin{versinnote}
%\tl{prāṇāyāmaṃ tataḥ kuryān nityaṃ  sāttvikayā dhiyā | \\+}
%\tl{tathā suṣumnāpārśvasthā malāḥ śoṣaṃ prayānti hi |\\!}
%\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_006]
%There is a problem in the third verse quarter. Among the divergent readings, the delta manuscripts convey the idea that the impurities are in the central channel  (\emph{suṣumṇāntarasthā}), which makes a defective \emph{ra-vipulā} (there is no caesura after the fourth syllable). This is similar in sense to V2(siglum), which has the reading \emph{suṣumṇā\-madhya\-sthā}. This meaning was accepted by Brahmānanda, who adopted \emph{suṣumṇā\-nāḍī\-sthā}. We are not aware of a reference in another text to impurities in the central channel, however, so we have adopted the reading of the \gamma \ group and \betaTwo, which states that the impurites are situated at the sides of Suṣumṇā, understanding this to refer to the secondary channels of \emph{iḍā} and \emph{piṅgalā}.

%The \emph{ca} at the end of the fourth verse quarter would suggest that two statements are being made in the second half of the verse. The readings of some manuscripts, such as \emph{suṣumṇā susnigdhā} in \epsilonThree\ and \etaOne, appear to be attempts to make sense of the \emph{ca}. However, the meaning of \emph{suṣumṇā susnigdhā} (`\emph{suṣumṇā} becomes well-lubricated') seems implausible. In light of the reading in the \gamma \ group and \betaTwo, it seems likely that \emph{ca} was simply a verse-filler.
%JM: I think we should adopt alpha1 (and others): suṣumṇā susvasthā. It works, and makes sense of ca. But we'd not be reading alpha1's yāsthā, which looks to be a mistake.
\end{philcomm}

\begin{metre}[hp02_006]
Anuṣṭubh (c: ma-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.7}
\begin{translation}[hp02_007]
Seated in the lotus pose, the yogi should fill himself up with air via the moon [channel], hold it for as long as he can, then expel it through the sun [channel].
\end{translation}
% adopt yathāśaktyā and punaḥ = alpha one readings.

\begin{sources}[hp02_007]
\emph{Vivekamārtaṇḍa} 77

\begin{versinnote}
\tl{baddhapadmāsano yogī prāṇaṃ candreṇa pūrayet |\\+}
\tl{dhārayitvā yathāśakti bhūyaḥ sūryeṇa recayet ||\\+}
\tl{\var{77c yathāśakti ] GT; yathāśaktyā VA 77d bhūyaḥ ] VAG; punaḥ T}}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_007]
\emph{Haṭharatnāvalī} 3.84ab

\begin{versinnote}
\tl{baddhapadmāsano yogī prāṇaṃ candreṇa pūrayet |\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.12

\begin{versinnote}
\tl{baddhapadmāsano yogī prāṇaṃ candreṇa pūrayet |\\+}
\tl{dhārayitvā yathāśakti bhūyaḥ sūryeṇa recayet ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_007]
%The hemistich beginning with \emph{dhārayitvā} appears to have dropped out of the transmission of the \emph{Haṭharatnāvalī}, thus rendering this passage unintelligible in that work (see Gharote et al. 2002: 125).%JM: necessary?
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.8}
\begin{translation}[hp02_008]
And, drawing the breath through the sun [channel], he should gradually fill the abdomen. Having performed the retention as prescribed, he should then exhale through the moon [channel].
\end{translation}

\begin{sources}[hp02_008]
\emph{Vivekamārtaṇḍa} 79

\begin{versinnote}
\tl{prāṇaṃ sūryeṇa cākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
\tl{vidhivat kuṃbhakaṃ kṛtvā punaś candreṇa recayet ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_008]
\emph{Haṭharatnāvalī} 3.84cd–85ab
\begin{versinnote}
\tl{prāṇaṃ sūryeṇa cākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
\tl{vidhivat kumbhakaṃ kṛtvā punaś candreṇa recayet||\\!}
\end{versinnote}

\emph{Yukabhavadeva} 7.14

\begin{versinnote}
\tl{prāṇaṃ sūryeṇa cākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
\tl{kumbhayitvā vidhānena bhūyaś candreṇa recayet ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_008]
%\emph{udare} is well-attested (V1 and J10) but the verb needs an object so \emph{udaraṃ} has been adopted. (not necessary as alpha and most collated witnesses have udaraṃ)
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.9}
\begin{translation}[hp02_009]
[The yogi] should inhale through the [channel] by which he has exhaled and hold [the breath] without discomfort. And then he should exhale through the other [channel] slowly, not quickly.
\end{translation}

\begin{sources}[hp02_009]
Cf. \emph{Dattātreyayogaśāstra} 61

\begin{versinnote}
\tl{yathāśaktyāvirodhena tataḥ kuryāc ca kumbham |\\+}
\tl{punas tyajet piṅgalayā śanair eva na vegataḥ || \\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_009]
\emph{Haṭharatnāvalī} 3.85cd%–86ab

\begin{versinnote}
\tl{yena tyajet tenāpūrya dhārayed avirodhataḥ ||\\+}
\tl{\var{85d avirodhataḥ ] anirodhataḥ P}\\!}
%prāṇaṃ ced iḍayā piben niyamitaṃ bhūyo 'nyayā recayet |
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_009]
The meaning of \emph{avirodhataḥ} (‘without harm/discomfort’) makes better sense in this context than \emph{anirodhataḥ} (‘without cessation’). One might try to construe \emph{anirodhataḥ} as ‘without stopping the breath,’ but verse 2.7 clearly states that the breath should be held as long as possible (\emph{yathāśakti}). According to the apparatus of the critical edition of the \emph{Haṭharatnāvalī}, \emph{avirodhataḥ} is well-attested for the parallel hemistich. Furthermore, \emph{avirodhata} is attested, as well as \emph{virodhahīna}, in a passage of the \emph{Haṭhatattvakaumudī} (36.6–9) that appears to have been loosely based on \emph{Haṭhapradīpikā} 2.7–9:

\begin{versinnote}
\tl{prāṇāyāme padmapīṭhe svadakṣāṃ-\\+}
\tl{guṣṭhenādau sanniruddhyendunāḍīm |\\+}
\tl{vāyuṃ nātidrāk śanair nātiyuktyā\\+} 
\tl{vyākṛṣyordhvaṃ pūrayet svodarānte ||\\+}
\tl{yathā svaśaktyā laghu dhārayitvā\\+} 
\tl{nāḍyā tataḥ piṅgalayā virecayet |\\+}
\tl{virodhahīnaṃ viratītamadhyaṃ\\+} 
\tl{hṛdā sthireṇābhyasanaṃ muniś caret ||\\+}
\tl{yena tyajet tena virodhahīnaṃ\\+} 
\tl{dhṛtvā purānyena virecayec chanaiḥ |\\+}
\tl{yānty evam abhyāsaratasya puṃsaḥ\\+} 
\tl{sthitiṃ svalakṣye calacittavṛttayaḥ ||\\+}
\tl{yathāśaktyākṛṣya khagaṃ pūrayed udaraṃ śanaiḥ |\\+}
\tl{yathāśaktyā dhṛtaṃ paścād recayed avirodhataḥ ||\\!}
\end{versinnote}
\end{philcomm}

\begin{metre}[hp02_009]
Anuṣṭubh (a: ra-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.10}
\begin{translation}[hp02_010]
If [the yogi] breathes in through Iḍā, he should then exhale the restrained [breath] through the other [channel, i.e. Piṅgalā]. Next, he should inhale through Piṅgalā, hold the breath and release it through the left [channel]. The channels of ascetics meditating on the two orbs of the sun and moon using this method are purified after three months.
\end{translation}

\begin{sources}[hp02_010]
\emph{Vivekamārtaṇḍa} 81

\begin{versinnote}
\tl{prāṇaṃ ced iḍayā pibet niyamitaṃ bhūyo 'nyayā recayet \\+}
\tl{pītvā piṅgalayā samīraṇam alaṃ baddhvā tyajed vāmayā |\\+}
\tl{sūryācandramasor anena vidhinā bimbadvayaṃ dhyāyatāṃ \\+}
\tl{śuddhā nāḍigaṇā bhavanti yaminā māsatrayād ūrdhvataḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_010]
\emph{Yogacintāmaṇi} f.~90v

\begin{versinnote}
\tl{haṭhayoge 'pi— \\+}
\tl{prāṇaṃ ced iḍayā piben niyamito bhūyo 'nyayā recayet\\+}
\tl{pītvā piṅgalayā samīraṇam atho baddhvā tyajed vāmayā |\\+}
\tl{sūryācandramasor anena vidhinā bimbadvayaṃ dhyāyatām\\+}
\tl{śuddhā nāḍigaṇā bhavanti yamināṃ māsatrayād ūrdhvataḥ ||\\+}
\end{versinnote}

\emph{Haṭharatnāvalī} 3.86

\begin{versinnote}
\tl{prāṇaṃ ced iḍayā piben niyamitaṃ bhūyo 'nyayā recayet\\+}
\tl{pītvā piṅgalayā samīraṇaṃ atho baddhvā tyajed vāmayā |\\+}
\tl{sūryācandramasor anena vidhinā bimbadvayaṃ dhyāyatāṃ\\+}
\tl{śuddhā nāḍigaṇā bhavanti yamināṃ māsatrayād ūrdhvataḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.16 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{prāṇaṃ ced iḍayā pibet parimitaṃ bhūyo 'nyayā recayet\\+}
\tl{pītvā piṅgalayā samīraṇam amalaṃ baddhvā tyajed vāmayā |\\+}
\tl{sūryācandramasor anena vidhinā bimbadvayaṃ dhyāyatām\\+}
\tl{śuddhā nāḍigaṇā bhavanti yamināṃ māsatrayād ūrdhvataḥ ||\\+}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_010]
%We have translated \emph{yamināṃ} as `yogis' because in this context it could refer to those practising the observances (\emph{yama}) of yoga rather than ascetics more generally. In \emph{Jyotsnā} 2.44, Brahmānanda understands it this way (\emph{yamināṃ yogināṃ}).%JM: It says ascetics, so why not translate thus?
%The reading \emph{vidhinābhyāsaṃ} in the third pāda of various witnesses is unmetrical because it lacks the caesura. (not well attested in the new collation). 
\end{philcomm}

\begin{metre}[hp02_010]
Śārdūlavikrīḍita 
\end{metre}

%%%%%%%%%%
\subsection*{2.11}
\begin{translation}[hp02_011]
[The yogi] should gently practise [twenty] retentions  four times [a day], at sunrise, midday, sunset and midnight, making a total of eighty.
\end{translation}

%\begin{sources}[hp02_011]
%\end{sources}

\begin{testimonia}[hp02_011]
\emph{Haṭharatnāvalī} 3.87

\begin{versinnote}
\tl{prātar madhyadine sāyam ardharātre ca kumbhakān ||\\+}
\tl{śanair aśītiparyantaṃ caturvāraṃ samabhyaset ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~90v (attr.~to the \emph{Haṭhayoga})
\begin{versinnote}
\tl{prātar madhyaṃ dine sāyam ardharātre ca kumbhakān ||\\+}
\tl{śanair aśītiparyantaṃ caturvāraṃ samabhyaset ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_011]
This verse is summarizing the following passage in the \emph{Dattātreyayogaśāstra} (63cd--65ab):

\begin{versinnote}
\tl{evaṃ prātaḥ samāsīnaḥ kuryād viṃśati kumbhakān || 63 ||\\+}
\tl{evaṃ madhyāhnasamaye kuryād viṃśati kumbhakān |\\+}
\tl{evaṃ sāyaṃ prakurvīta punar viṃśati kumbhakān || 64 ||\\+}
\tl{evam evārdharātre 'pi kuryād viṃśati kumbhakān |\\!}
\end{versinnote}

Without reference to the \emph{Dattātreyayogaśāstra}, the meaning of the second half of the verse is ambiguous because it could be understood as saying up to eighty retentions four times a day. In his \emph{Jyotsnā} (2.11), Brahmānanda understands it this way, and takes \emph{śanaiḥ} to mean `gradually' building up to the eighty retentions. However the parallel verses in the \emph{Dattātreyayogaśāstra} make it clear that twenty retentions (\emph{kumbhaka}) are to be practised four times a day. 
\end{philcomm}

%%%%%%%%%%
\subsection*{2.12}
\begin{translation}[hp02_012]
In the lesser cessation of the breath sweating arises, in the middle [cessation], shaking, and in the highest [the yogi] repeatedly rises up in the lotus pose.
\end{translation}

\begin{sources}[hp02_012]
Cf. \emph{Vivekamārtaṇḍa} 87

\begin{versinnote}
\tl{adhame ca ghano gharmaḥ kampo bhavati madhyame |\\+}
\tl{uttiṣṭhaty uttame deho baddhapadmāsano muhuḥ ||\\+}
\tl{\var{87c uttiṣṭhaty uttame deho ] T; uttiṣṭhaty uttamo deho V, uttame nu guṇam āpnoti A, uttame sthāṇum āpnoti GB, uttame sthānam āpnoti GL, uttiṣṭhaṃty uttame prāṇā GP 87d baddhapadmāsano muhuḥ ] VT; tato vāyu nibandhayet A, tato vāyuṃ nirundhayet GBGL, vaddhapadmāsane muhūḥ GP}\\!}%How to change sigla of GB, GL, GP so that second letters are subscript? Do it in PreambleComm file
\end{versinnote}
\end{sources}
% 87c uttiṣṭhaty uttame deho ] T; uttiṣṭhaty uttamo deho V, uttame nu guṇam āpnoti A, uttame sthāṇum āpnoti GB, uttame sthānam āpnoti GL, uttiṣṭhaṃty uttame prāṇā GP 87d baddhapadmāsano muhuḥ ] VT; tato vāyu nibandhayet A, tato vāyuṃ nirundhayet GBGL, vaddhapadmāsane muhūḥ GP

\begin{testimonia}[hp02_012]
\emph{Haṭharatnāvalī} 3.88

\begin{versinnote}
\tl{kanīyasi bhavet svedaḥ kampo bhavati madhyame |\\+}
\tl{uttiṣṭhaty uttame prāṇarodhe padmāsane muhuḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 90v (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{kanīyasi bhavet svedaḥ kampo bhavati madhyame |\\+}
\tl{uttiṣṭhaty uttame prāṇarodhe padmāsanasthitaḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_012]
The manuscript readings diverge greatly in the second hemistich. In the third verse quarter, all of the manuscripts have \emph{prāṇa} in some form (instead of \emph{deha} in the \emph{Vivekamārtaṇḍa}). \textalpha\ and some of the \textbeta, \textepsilon\ and \texteta\ manuscripts seem to be stating that it is \emph{padmāsana} that rises up again and again in the highest stage of holding the breath (\emph{uttiṣṭhaty uttame prāṇarodhe padmāsanaṃ muhuḥ}). We have adopted a similar reading but with \emph{padmāsane} (\betaTwo, \deltaOne\ and \deltaThree, and \etaTwo) because it makes better sense that the yogi rises up while seated in lotus pose. Another version is seen in \etaTwo\ (and others), which appears to say that the breaths rise up again and again when one is seated in the lotus pose (\emph{uttiṣṭhanty uttame prāṇā baddhe padmāsane muhuḥ}). However, the verse is about the external signs that might arise in \emph{prāṇāyāma} rather than internal processes. Such confusion has arisen because the verse was taken from the \emph{Vivekamārtaṇḍa} without its context, which is a classification of different levels of \emph{prāṇāyāma}, so Svātmārāma needed to include \emph{prāṇarodhe} meaning \emph{prāṇāyāma} in order for the different adjectives to have something with which to agree, and he did so despite the infelicity of \emph{prāṇarodhe} crossing the \emph{pāda} break.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.13}
\begin{translation}[hp02_013]
[The yogi] should rub the limbs with the sweat produced through exertion. As a result the body becomes firm and lithe.%JM how about "strong and lithe"? "Firm" sounds a bit odd to me.
\end{translation}

\begin{sources}[hp02_013]
Cf. \emph{Dattātreyayogaśāstra} 75

\begin{versinnote}
\tl{prasvedo jāyate pūrvaṃ mardanaṃ tena kārayet |\\+}
\tl{tato ’tidhāraṇād vāyoḥ krameṇaiva śanaiḥ śanaiḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_013]
\emph{Haṭharatnāvalī} 3.89

\begin{versinnote}
\tl{jalena śramajātena aṅgamardanam ācaret |\\+}
\tl{dṛḍhatā laghutā cāpi tathā gātrasya jāyate || 3.89 ||\\!}
\end{versinnote}

Cf. \emph{Śivasaṃhitā} 3.46

\begin{versinnote}
\tl{svedaḥ saṃjāyate dehe yoginaḥ prathamodyame |\\+}
\tl{yadā saṃjāyate svedo mardanaṃ kārayet sudhīḥ |\\+}
\tl{anyathā vigrahe dhātur naṣṭo bhavati yoginaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 90v (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{jalena śramajātena gātramardanam ācaret |\\+}
\tl{dṛḍhatā laghutā cāpi tena gātrasya jāyate ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_013]
\emph{Śivasaṃhitā} 3.46 adds that if this practice is not done, the body's constituents (\emph{dhātu}s) are lost.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.14}
\begin{translation}[hp02_014]
At the beginning of the practice, food with milk and ghee is recommended. After that, when the practice has become well established, there is no need to adopt such regulations.
\end{translation}

\begin{sources}[hp02_014]
\emph{Śivasaṃhitā} 3.43

\begin{versinnote}
\tl{abhyāsakāle prathamaṃ kuryāt kṣīrājyabhojanam\\+}
\tl{tato'bhyāse sthirībhūte na tādṛṅniyamagrahaḥ 3.43\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_014]
\emph{Haṭharatnāvalī} 1.24

\begin{versinnote}
\tl{abhyāsakāle prathame śastaṃ kṣīrādibhojanam |\\+}
\tl{tato 'bhyāse dṛḍhībhūte na tāvan niyamagrahaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 4.27 (attr.~to the \emph{Śivayoga})
\begin{versinnote}
\tl{abhyāsakāle prathame śastaṃ kṣīrādibhojanam |\\+}
\tl{tato 'bhyāse dṛḍhībhūte na tādṛṅniyamāgrahaḥ ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_014]
\end{philcomm}

\begin{metre}[hp02_014]
Anuṣṭubh (a: bha-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.15}
\begin{translation}[hp02_015]
Just as a lion, an elephant [or] a tiger is tamed gradually, so the breath is cultivated [gradually], otherwise it kills the practitioner.
\end{translation}

\begin{sources}[hp02_015]
\emph{Vivekamārtaṇḍa} 101

\begin{versinnote}
\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śānaiḥ |\\+}
\tl{anyathā hanti yantāraṃ tathā vāyur asevitaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_015]
\emph{Haṭharatnāvalī} 3.90

\begin{versinnote}
\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śanaiḥ |\\+}
\tl{tathaiva sevito vāyur bhaved vaśyaḥ śanaiḥ śanaiḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.28 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śanaiḥ |\\+}
\tl{tathaiva sevito vāyur bhaved vaśyaḥ śanaiḥ śanaiḥ ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_015]
The second hemistich of this verse has been rewritten to express the same idea (and simile) as that found in \emph{Viveka\-mārtaṇḍa} 101, but the author of the \emph{Viveka\-mārtaṇḍa} expresses it more clearly.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.16}
\begin{translation}[hp02_016]
All diseases are destroyed by correct \emph{prāṇāyāma}. As a result of incorrect practice any disease may arise.
\end{translation}

\begin{sources}[hp02_016]
\emph{Vivekamārtaṇḍa} 99

\begin{versinnote}
\tl{prāṇāyāmena yuktena sarvarogakṣayo bhavet |\\+}
\tl{ayuktābhyāsayogena sarvarogasamudbhavaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_016]
\emph{Haṭharatnāvalī} 3.90

\begin{versinnote}
\tl{prāṇāyāmena yuktena sarvarogakṣayo bhavet |\\+}
\tl{ayuktābhyāsayogena sarvarogasamudbhavaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 91v–92r (attr.~to the \emph{Skandapurāṇa})

\begin{versinnote}
\tl{prāṇāyāmena yuktena sarvavyādhikṣayo bhavet |\\+}
\tl{ayuktābhyāsayogena sarvavyādhisamudbhavaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.26 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{prāṇāyāmena yuktena sarvarogasya saṃkṣayaḥ |\\+}
\tl{ayuktābhyāsayogena sarvarogasya sambhavaḥ ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_016]
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.17}
\begin{translation}[hp02_017]
Hiccups, wheezing, cough, pains in the head, ears and eyes: various diseases arise as a result of the breath going awry.
\end{translation}

\begin{sources}[hp02_017]
\emph{Vivekamārtaṇḍa} 100

\begin{versinnote}
\tl{hikkā śvāsaś ca kāsaś ca śiraḥkarṇākṣivedanā |\\+}
\tl{bhavanti vividhā doṣāḥ pavanasya vyatikramāt ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_017]
\emph{Haṭharatnāvalī} 3.92

\begin{versinnote}
\tl{hikkā śvāsaś ca kāsaś ca śiraḥkarṇākṣivedanāḥ |\\+}
\tl{bhavanti vividhā rogāḥ pavanasya vyatikramāt ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 92r (attr.~to the \emph{Skandpurāṇa})

\begin{versinnote}
\tl{hikkā śvāsaś ca kāsaś ca śiraḥkarṇākṣivedanāḥ |\\+}
\tl{bhavanti vividhā rogāḥ pavanasya vyatikramāt ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.27 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{hikkā śvāsas tathā kāsaḥ śiraḥkarṇākṣivedanā |\\+}
\tl{bhavanti vividhā rogāḥ pavanasya vyatikramāt ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_017]
This verse has parallels in verses on the illnesses caused by incorrect breathing in earlier Śaiva works, two examples of which are:

\emph{Śivadharmottara} 10.124cd–125
\begin{versinnote}
\tl{hikkāśvāsapratiśyāyaḥ karṇadantākṣivedanāḥ ||\\+}
\tl{mūkatā jaḍatā kāsaḥ śirorogaḥ śramakṣaraḥ |\\+}
\tl{ityevamādayo doṣā jāyante vyutkrameṇa tu ||\\!}
\end{versinnote}

\emph{Dharmaputrikā} 10.265–266ab
\begin{versinnote}
\tl{kaphakoṣṭhe yadā vāyur granthir bhūtvāvatiṣṭhate |\\+}
\tl{hṛllāsahikkikāśvāsaśiraḥśūlādayo rujāḥ ||\\+}
\tl{jāyante dhātuvaiṣamyāt tadā kuryāt pratikriyāṃ |\\!}
\end{versinnote}

\end{philcomm}

%%%%%%%%%%
\subsection*{2.18}
\begin{translation}[hp02_018]
[The yogi] should exhale correctly, inhale correctly and hold the breath correctly. He thus becomes purified.
\end{translation}

\begin{sources}[hp02_018]
\emph{Vivekamārtaṇḍa} 102

\begin{versinnote}
\tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ ca pūrayet |\\+}%122ab ; HP2.18ab
\tl{yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_018]
\emph{Haṭharatnāvalī} 3.93

\begin{versinnote}
\tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ prapūrayet |\\+}
\tl{yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 92v (attr.~to the \emph{Skandpurāṇa})

\begin{versinnote}
\tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ ca pūrayet |\\+}
\tl{yuktaṃ yuktaṃ ca badhnīyād itthaṃ siddhyati yogavit ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.29 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ tu pūrayet |\\+}
\tl{yuktaṃ yuktaṃ tu badhnīyād evaṃ siddhim avāpnuyāt ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_018]
%Some of the alpha, epsilon and eta manuscripts have \emph{śuddhiṃ} instead of \emph{siddhiṃ} in the fourth verse quarter but this is likely a later change, or even an error, by someone who was anticipating the subject of the next verse. The reading \emph{siddhiṃ} is attested across all branches of the stemma, as well as the source text and most of the testimonia.

%AGS: the repetition of yuktaṃ suggests a southern origin for the verse. [search for Proof]
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.19}
\begin{translation}[hp02_019]
When the channels are pure, external signs occur. Leanness and lustre of the body are certain to arise.
\end{translation}

\begin{sources}[hp02_019]
Cf. \emph{Dattātreyayogaśāstra} 67cd--69ab

\begin{versinnote}
\tl{yadā tu nāḍiśuddhiḥ syāt tadā cihnāni bāhyataḥ ||\\+}
\tl{jāyante yogino dehe tāni vakṣyāmy aśeṣataḥ |\\+}
\tl{śarīralaghutā dīptir jaṭharāgnivivardhanam ||\\+}
\tl{kṛśatvaṃ ca śarīrasya tadā jāyeta niścitam |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_019]
\emph{Haṭharatnāvalī} 3.94

\begin{versinnote}
\tl{yadā tu nāḍīśuddhiḥ syāt tadā cihnāni bāhyataḥ |\\+}
\tl{kāyasya kṛśatā kāntir jāyate tasya niścitam ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 90v (attr.~to the \emph{Hathayoga})

\begin{versinnote}
\tl{yadā nāḍīviśuddhiḥ syāt tadā cintānirākṛtā |\\+}
\tl{kāyasya kṛśatā kāntis tadā jāyeta niścitam  ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_019]
The idea that \emph{prāṇāyāma} is done to purify the channels (\emph{nāḍī}) can be found in discussions of \emph{prāṇāyāma} in early Śaiva tantras. For example, the \emph{Nayasūtra} of the \emph{Niśvāsatattvasaṃhitā} (4.110) and the \emph{Svacchandatantra} (7.294cd–7.295ab) specifically refer to purifying the channels by inhaling through the left nostril and exhaling through the right, as stated in the latter:

\begin{versinnote}
\tl{apasavyena pūryeta savyenaiva virecayet |\\+} 
\tl{nāḍīsaṃśodhanaṃ caitan mokṣamārgapathasya ca ||\\!}
\end{versinnote}
\end{philcomm}

\begin{metre}[hp02_019]
Anuṣṭubh (a: ma-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.20}
\begin{translation}[hp02_020]
The ability to hold the breath as long as one desires, stimulation of the [digestive] fire, manifestation of the inner sound [and] freedom from disease occur as a result of purifying the channels.%JM: I've changed "resonance" to "sound" for nāda. "resonance" isn't right as it's a quality of sound, not sound itself; furthermore, the quality is one of deep reverberation, which not all internal sounds have.
\end{translation}

\begin{sources}[hp02_020]
\emph{Vivekamārtaṇḍa} 101

\begin{versinnote}
\tl{yatheṣṭaṃ dhāraṇāṃ vāyor analasya pradīpanam |\\+}
\tl{nādābhivyaktir ārogyaṃ jāyate nāḍīśodhanāt ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_020]
\emph{Haṭharatnāvalī} 3.95

\begin{versinnote}
\tl{yatheṣṭaṃ dhāraṇaṃ vāyor analasya pradīpanam |\\+}
\tl{nādābhivyaktir ārogyaṃ jāyate nāḍiśodhanāt ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 90v (attr.~to the \emph{Skandapurāṇa})

\begin{versinnote}
\tl{yatheṣṭaṃ dhāraṇaṃ vāyor analasya pradīpanam |\\+}
\tl{nādābhivyaktir ārogyaṃ bhaven nāḍīviśodhanāt ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.17 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{yatheṣṭaṃ dhāraṇaṃ vāyor analasya pradīpanam |\\+}
\tl{nādābhivyaktir ārogyaṃ jāyate nāḍīśodhanāt ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_020]
Similar signs (\emph{cihna}) arising from the purification of the channels are mentioned in the \emph{Vasiṣṭhasaṃhitā} (2.68--69) and subsequent works related to it:

\begin{versinnote}
\tl{nāḍīśuddhim avāpnoti pṛthak cihnopalakṣitām |\\+}
\tl{śarīralaghutā dīptir jaṭharāgnivivardhanam ||\\+}
\tl{nādābhivyaktir ity etac cihnaṃ tacchuddhisūcakam |\\+}
\tl{yāvad etāni saṃpaśyet tāvad evaṃ samācaret ||\\!}
\end{versinnote}

\end{philcomm}

%%%%%%%%%%
\subsection*{2.21}
\begin{translation}[hp02_021]
The person who has an excess of fat and phlegm should first practise the six therapeutic interventions, but anyone else, because their humours are in balance, should not practise them.
\end{translation}

\begin{sources}[hp02_021]
\end{sources}

\begin{testimonia}[hp02_021]
\emph{Yogacintāmaṇi} 8v (attr.~to Ātmārāma)

\begin{versinnote}
\tl{medaḥśleṣmanivṛtyarthaṃ ṣaṭkarmāṇi samācaret |\\+}
\tl{anyathā nācaret tāni doṣāṇāṃ samatā yataḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.147 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{medaślemādisampūrṇaḥ ṣaṭkarmāṇi samācaret |\\+}
\tl{anyas tu nācaret tāni doṣāṇāṃ samabhāgikaḥ ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_021]
Manuscripts of the delta group, as well as \etaTwo\ and \epsilonThree, also have the valid readings of \emph{medaḥśleṣmādināśārthaṃ} and \emph{anyathā} in the first and third verse quarters, respectively. This version of the verse states that one should practise the six therapeutic interventions to remove fat, phlegm and the like, otherwise one should not practise them when the humours are in balance. However, the \emph{pūrvam} (`first'), which is attested in two \textalpha\ manuscripts, as well as manuscripts of the \textbeta\ and \textgamma\ groups, fits the context of these interventions being preliminary practices for \emph{prāṇāyāma}.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.22}
\begin{translation}[hp02_022]
\emph{Dhauti, basti, neti, trāṭaka, naulī} and \emph{kapālabhātī}. These are said to be the six [therapeutic] techniques.%JM: "therapeutic interventions" adds a lot to karmāṇi. I usually translate it with "[cleansing] techniques". We have techniques everywhere else.
%JM: note on alphaone's laulikam? And should we have naulī here: that's what it is everywhere else so this is prob for metre.
\end{translation}

\begin{sources}[hp02_022]
\end{sources}

\begin{testimonia}[hp02_022]
\emph{Haṭharatnāvalī} 1.27

\begin{versinnote}
\tl{haṭhapradīpikāyām-\\+}
\tl{dhautir bastis tathā netis trāṭakaṃ naulikaṃ tathā |\\+}
\tl{kapālabhrāntir etāni ṣaṭkarmāṇi pracakṣate  ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71r (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha ṣaṭkarmāṇi | haṭhapradīpikāyām ||\\+}
\tl{dhautī bastī tathā netī trāṭakaṃ naulikaṃ tathā |\\+}
\tl{kapālabhātī caitāni ṣaṭkarmāṇi pracakṣate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.148 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{dhautir bastiś ca netiś ca trāṭakaṃ naulikaṃ tathā |\\+}
\tl{kapālabhāti caitāni ṣaṭkarmāṇi pracakṣate ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_022]
Manuscripts across all the groups contain many different spellings of the names of these techniques. As well as the requirements of the metre, the spellings we have favoured take into account the occurrences of each name in subsequent verses. 
\end{philcomm}

\begin{metre}[hp02_022]
Anuṣṭubh (c: ma-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.23}
\begin{translation}[hp02_023]
This set of six techniques should be kept secret. Bringing about purification of the body [and] bestowing various good qualities, it is worshipped by the best yogis.
\end{translation}

\begin{sources}[hp02_023]
\end{sources}

\begin{testimonia}[hp02_023]
\emph{Haṭharatnāvalī} 1.28

\begin{versinnote}
\tl{karmāṣṭakam idaṃ gopyaṃ ghaṭaśodhanakārakam |\\+}
\tl{kasya cin naiva vaktavyaṃ kulastrīsurataṃ yathā  ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71r (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{karmaṣaṭkam idaṃ gopyaṃ ghaṭaśodhanakārakam |\\+}
\tl{vicitraguṇasaṃdhāyi pūjyate yogipuṃgavaiḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.149 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{karmaṣaṭkam idaṃ gopyaṃ ghaṭaśodhanakāraṇam |\\+}
\tl{vicitraguṇasandhāyī kriyate yogibhiḥ sadā ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_023]
\end{philcomm}

%%%%%%%%%%
\subsection*{2.24}
\begin{translation}[hp02_024]
Among them is dhauti:\\ 
{}[The yogi] should slowly swallow a moistened cloth  four finger-breadths in width and then draw it out. This ejection [of it] from the mouth is the dhauti technique.
\end{translation}

%\begin{sources}[hp02_024]
%\end{sources}

\begin{testimonia}[hp02_024]
\emph{Haṭharatnāvalī} 1.37–38ab

\begin{versinnote}
\tl{atha dhautiḥ--\\+}
\tl{viṃśaddhastapramāṇena dhautavastraṃ sudīrghitam |\\+}
\tl{caturaṅgulavistāraṃ siktaṃ caiva śanaiḥ graset ||\\+}
\tl{tataḥ pratyāharec caitad abhyāsād dhautir ucyate |\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~71r (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha dhautī |\\+}
\tl{caturaṅgulavistāraṃ siktaṃ vastraṃ śanair graset |\\+}
\tl{punaḥ pratyāhared etad abhyāsād dhautikarmavit ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.150 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{caturaṃgulavistāraṃ siktaṃ vastraṃ śanair graset |\\+}
\tl{tataḥ pratyāharec caitad ākṣālaṃ dhautikarma tat ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 56–57

\begin{versinnote}
\tl{atha dhautī |\\+}
\tl{mṛdulaṃ dhavalaṃ śuddhaṃ caturaṅgulavistṛtam |\\+}
\tl{tithihastamitāyāmaṃ dhautīvastrasya lakṣaṇam ||\\+}
\tl{toyasiktaṃ grased vastraṃ ghrāṇābhyāṃ vāyum utsṛjan |\\+}
\tl{śanaiḥ sanais tu sakalaṃ punaḥ pratyāharec chanaiḥ |\\+}
\tl{dhautīkarmedam ākhyātaṃ yatra gaṅgādhidaivatam ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_024]
The manuscripts contain many different readings for the fourth \emph{pāda}. \alphaThree\ and \etaOne\ have the term \emph{udgāraṃ}, which rarely occurs in yoga texts. The basic meaning of \emph{udgāra} is the act of discharging something from the mouth, which fits the context of \emph{dhauti} in so far as the cloth swallowed into the stomach is drawn back out through the mouth. Some of the other readings, such as \emph{uditaṃ}, \emph{utthānaṃ}, etc., appear to be mistakes or patches that arose possibly because \emph{udgāra} is not normally neuter but masculine.

Many manuscripts have added verse quarters on the length of the cloth (\emph{hastapañcadaśena tu}) and doing the practice according to the guru’s teachings (\emph{gurūpadiṣṭa\-mārgeṇa}). These additional comments are absent in the \textalpha, \textgamma\ and \textepsilon\ groups. Furthermore, the compound \emph{hastapañcadaśena} does not seem to fit the syntax of the sentence. The other addition, on the guru's teaching, is a cliché that is probably being used here as a verse filler.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.25}
\begin{translation}[hp02_025]
Coughing, wheezing, splenitis and skin diseases, as well as the twenty phlegmatic diseases, are sure to disappear through the power of the dhauti technique.
\end{translation}

\begin{sources}[hp02_025]
\end{sources}

\begin{testimonia}[hp02_025]
\emph{Haṭharatnāvalī} 1.39

\begin{versinnote}
\tl{kāsaśvāsaplīhakuṣṭhaṃ kapharogāś ca viṃśatiḥ |\\+}
\tl{dhautikarmaprabhāvena dhāvanty eva na saṃśayaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~71r (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{kāsaśvāsaplīhakuṣṭhaṃ kapharogāś ca vidradhiḥ |\\+}
\tl{dhautīkarmaprabhāvena prayānty eva na saṃśayaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.151 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{plīhā śvāsaś ca kuṣṭhaṃ ca kapharogāś ca viṃśatiḥ |\\+}
\tl{dhautikarmaprabhāvena gacchanty eva na saṃśayaḥ ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 58

\begin{versinnote}
\tl{kāsasvāsaplīhakuṣṭhādināśam\\+} 
\tl{vahner māndyaṃ viṃśatiḥ śleṣarogān |\\+}
\tl{dūrīkuryāt karṇabādhir tam uccair\\+} 
\tl{dhautīkarma praditaṃ śaṅkareṇa ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_025]
The verb \emph{dhāvanti} is a play on words, using a different root \emph{dhāv}, “run”, from that of \emph {dhauti}, which is derived from \emph{dhāv}, “purify”.

Twenty phlegmatic diseases are enumerated in the \emph{Carakasaṃhitā}, \emph{sūtrasthāna} 20.17, a chapter on major diseases (\emph{mahāroga}). %The other diseases mentioned in the \emph{Haṭhapradīpikā}'s verse on \emph{dhauti}, namely cough (\emph{kāsa}), wheezing (\emph{śvāsa}), splenitis (\emph{plīha}) and skin diseases (\emph{kuṣṭha}), are not among them. The likely reason for this is that these particular diseases are not solely caused by phlegm disorders. For example, \emph{Carakasaṃhitā}, \emph{sūtrasthāna} 19.4 states that the five types of spleen diseases are caused by wind, bile, phlegm, commingling of the humours (\emph{sannipāta}) and blood, and the five types of cough arise from wind, bile, phelgm, injury, and wasting.%JM: I don't understand the import of the second part of the note — the text doesn't imply that the first ailments listed should be among the twenty kapha diseases.
\end{philcomm}

\begin{metre}[hp02_025]
Anuṣṭubh (a: ra-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.26}
\begin{translation}[hp02_026]
Now the \emph{basti} technique.\\
Squatting in water up to the navel with a reed inserted in the anus, [the yogi] should contract the perineal region (\emph{ādhāra°}). The [resultant] flushing is the basti technique.
\end{translation}
% Adopt pakhālaṃ — not done in latest version

\begin{sources}[hp02_026]
\end{sources}

\begin{testimonia}[hp02_026]
\emph{Haṭharatnāvalī} 1.45–47

\begin{versinnote}
\tl{nābhidaghne jale sthitvā pāyunāle sthitāṅguliḥ |\\+}
\tl{cakrimārgeṇa jaṭharaṃ pāyunālena pūrayet ||\\+}
\tl{vicitrakaraṇīm kṛtvā nirbhītaḥ recayej jalam |\\+}
\tl{yāvad balaṃ prapūryaiva kṣaṇaṃ sthitvā virecayet ||\\+}
\tl{ghaṭītrayaṃ na bhoktavyaṃ bastim abhyasatā dhruvam |\\+}
\tl{nivātabhūmau santiṣṭhed vaśī hitamitāśanaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~71r (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha vastī |\\+}
\tl{nābhidaghne jale pāyunyastanālotkaṭāsanaḥ |\\+}
\tl{ādhārākuñcanaṃ kuryād abhyāsād vastikarmavit ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.152 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{nābhidaghne jale pāyau nyastanālotkaṭāsanaḥ |\\+}
\tl{ādhārā kuñcanaṃ kuryāt kṣālanaṃ bastikarma tat ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 132

\begin{versinnote}
\tl{naulīkriyāsusaṃpannas tyaktamūtramalaḥ sudhīḥ |\\+}
\tl{jānudaghne jale kuryād bastiṃ bastividhānavit ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_026]
% MD: kṣālanaṃ or prakṣālaṃ? Adopt pakhālaṃ [MD: done]
The reading \emph{pakhālaṃ} that we have adopted in the fourth \emph{pāda} is found in \betaOmega \ and \alphaOne \ and reflects vernacular usage as found in the old Hindi \emph{Aṣṭāṅgayoga} of Caraṇadāsa (6.71ab). Some other witnesses have the Sanskritised  form \emph{prakṣālaṃ}, which is very rare in Sanskrit sources, while others have the more common \emph{kṣālanaṃ}.

\end{philcomm}

%%%%%%%%%%
\subsection*{2.27}
\begin{translation}[hp02_027]
By the power of the basti technique, swelling, splenitis, stomach disorders and all diseases arising from wind, bile and phlegm are removed.
\end{translation}

%\begin{sources}[hp02_027]
%\end{sources}

\begin{testimonia}[hp02_027]
\emph{Haṭharatnāvalī} 1.48

\begin{versinnote}
\tl{gulmaplīhodaraṃ vāpi vātapittakaphādikam |\\+}
\tl{bastikarmaprabhāvena dhāvanty eva saṃśayaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71r (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{gulmodaraṃ cāpi vātaplīhapittakaphodbhavāḥ |\\+}
\tl{vastikarmaprabhāvena bādhyante sakalāmayāḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.153 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{gulmaplīhodaraṃ cāpi vātapittakaphodbhavāḥ |\\+}
\tl{bastikarmaprabhāvena naśyanti sakalāmayāḥ ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 135, 140–141

\begin{versinnote}
\tl{yāvan malā vinaśyanti vātapittakaphodbhāvāḥ |\\+}
\tl{trivāraṃ vā caturvāraṃ kṛtvā bastim virecayet ||\\+}
\tl{mahojasvī mahajjyotir jaṭharāgnipradīpanam |\\+}
\tl{gulmaplīhodarādīnāṃ nāśanaṃ sukhavardhanam ||\\+}
\tl{vātapittakaphottānāṃ doṣāṇāṃ nāśanaṃ paraṃ |\\+}
\tl{kuṣṭhānāṃ nāśanaṃ cāpi bastisiddhe prajāyate ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_027]
%The J10 group has \emph{°bhavaṃ}, which would qualify \emph{gulmaplīhodaram}. (seems to be an improbable reading because then sakalāmayāḥ is unqualified, thus rendering the statement that basti cures all diseases.
We have assumed that the compound \emph{gulmaplīhodara} is referring generally to swelling (\emph{gulma}), spleen disorders (\emph{plīhan}) and stomach diseases (\emph{udara}). However, the terms \emph{gulma} and \emph{plīhodara} can be understood as more specific diseases. In \emph{Suśrutasaṃhitā, uttaratantra} 42.4, \emph{gulma} is defined as a movable or immovable round lump (\emph{granthi}) that might arise between the heart and lower abdomen (\emph{basti}) and might grow or shrink:

\begin{versinnote}
\tl{hṛdbastyorantare granthiḥ saṃcārī yadi vā 'calaḥ |\\+}
\tl{cayāpacayavān vṛttaḥ sa gulma iti kīrtitaḥ ||\\!}
\end{versinnote}

The compound \emph{plīhodara} is the name of a specific disease, which is defined in the \emph{Suśrutasaṃhitā} (\emph{nidānasthāna} 7.14–15) as enlargement of the spleen (\emph{plīhābhivṛddhi}) so \emph{gulmaplīhodara} might be referring more specifically to abdominal lumps and splenomegaly.
%Diwakar thinks that \emph{gulmodara} and \emph{plīhodara} should be read. Search āyurvedic commentaries on this. (homework) JB: I can find strong evidence for a disease called plīhodara, but not gulmodara (which usually occurs in longer compounds and is translated by P.V. Sharma as gulma and udararoga: e.g., Suśruta 1.45.221ab, 1.46.196ef, 6.39.217, 6.40.180) 
\end{philcomm}

%%%%%%%%%%
\subsection*{2.28}
\begin{translation}[hp02_028]
When practised repeatedly, the water enema (\emph{jalabasti}) technique bestows clarity of the bodily constituents, senses and mind, radiance, [and] stimulation of the digestive fire, and removes [excessive] accumulation of all humours.
\end{translation}

%\begin{sources}[hp02_028]
%\end{sources}

\begin{testimonia}[hp02_028]
\emph{Haṭharatnāvalī} 1.49

\begin{versinnote}
\tl{dhātvindriyāntaḥkaraṇaprasādaṃ \\+}
\tl{dadyāc ca kāntiṃ dahanapradīptim |\\+}
\tl{aśeṣadoṣopacayaṃ nihanyād \\+}
\tl{abhyasyamānaṃ jalabastikarma ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71r (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{dhātvindriyāntaḥkaraṇaprasādaṃ\\+}
\tl{dadhyāc ca kāntiṃ dahanapradīptim |\\+}
\tl{aśeṣadoṣopacayaṃ nihanyād\\+}
\tl{abhyasyamānaṃ jalavastikarma ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.154 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{dhātvīndrintaḥ karaṇaprabodhaṃ\\+}
\tl{dadāti kāntiṃ dahanapradīptim |\\+}
\tl{aśeṣadoṣopacayaṃ nihanyād\\+}
\tl{abhyasyamānaṃ jalavastikarma ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 139–140ab

\begin{versinnote}
\tl{tiṣṭhed vaśī mitāhāraḥ sarvāṅgaṃ tena śudhyati |\\+}
\tl{dhātvindriyāntaḥkaraṇaprasādo dehalāghavam ||\\+}
\tl{mahojasvī mahajjyotir jaṭharāgnipradīpanam |\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_028]
%\end{philcomm}

\begin{metre}[hp02_028]
Upajāti
\end{metre}

%%%%%%%%%%
\subsection*{2.29}
\begin{translation}[hp02_029]
Raising the \emph{apāna} wind into the oesophagus (\emph{kaṇṭhanāle}) and ejecting the contents of the stomach from the windpipe, which has been brought under control by cumulative practice, is called the elephant technique by experts in Haṭha.
\end{translation}

\begin{sources}[hp02_029]
\end{sources}

\begin{testimonia}[hp02_029]
\emph{Haṭharatnāvalī} 1.51

\begin{versinnote}
\tl{udaragatapadārtham udvamantī \\+}
\tl{pavanam apānam udīrya kaṇṭhanāle |\\+}
\tl{kramaparicayatas tu vāyumārge \\+}
\tl{gajakaraṇīti nigadyate haṭhajñaiḥ || 1.51 ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.154 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{udaragatapadārtham udvamantī\\+}
\tl{pavanamapānam udīrya kaṇṭhanāle |\\+}
\tl{kramaparicayavaśyavāyumārgā\\+}
\tl{gajakaraṇīti nigadyate haṭhajñaiḥ ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 8.8

\begin{versinnote}
\tl{udaragatapadārthān udvamed eva nityaṃ\\+}
\tl{pavanagamanamārgāt kaṇṭhanālapraveśāt ||\\+}
\tl{kramaparicayavaśyaṃ syāc ca gargādayo hi\\+}
\tl{gajakaraṇam itīha prahur āryā munīndrāḥ ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 108–109

\begin{versinnote}
\tl{atha gajakaraṇī\\+}
\tl{śuddhaṃ toyaṃ nārikelodbhavaṃ vā \\+}
\tl{pītvākaṇṭhaṃ dugdhamiśraṃ jalaṃ vā |\\+}
\tl{vāraṃ vāraṃ māṇibandhaṃ tu kurvan \\+}
\tl{nodgāreṇa prakṣiped bhūmibhāge  ||\\+}
\tl{eṣā proktā kaphapittāmayeṣu \\+}
\tl{medoghnīva kariṇī hastipūrvā ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_029]
Some manuscripts, including \etaTwo\ and \betaOmega, have an alternative reading for the third verse quarter that appears to be explaining the name of the practice. In other words, it is called the elephant technique `because the speed of the breath is like that of water [propelled] by elephants’ (\emph{karibhir iva jalasya vāyuvegāt}). The syntax of this reading is not so easy to construe with the rest of the verse, which suggests that it was not original. The reading we have adopted (i.e., \emph{kramaparicayavaśya}...) is attested by manuscripts of the most important groups (\alphaOne\ and \alphaTwo, \betaTwo, the \textgamma\ group, etc.) and the same witnesses preserve \emph{mārga} (rather than \emph{vega}).

%The \emph{Jyotsnā} and printed editions have a finite verb in the first hemistich and assume a plural subject (i.e., \emph{yoginaḥ}) instead of the feminine singular present participle (agreeing with \emph{gajakaraṇī}). (JB not necessary as alpha, etc. support udvamantī)
\end{philcomm}

\begin{metre}[hp02_029]
Puṣpitāgrā
\end{metre}

%%%%%%%%%%
\subsection*{2.30}
\begin{translation}[hp02_030]
Now \emph{neti}.\\
{[}The yogi] should insert a very smooth thread one handspan [in length] into the nasal passage and take it out through the mouth. This is called neti by the Siddhas.
\end{translation}

%\begin{sources}[hp02_030]
%\end{sources}

\begin{testimonia}[hp02_030]
Cf.~\emph{Haṭharatnāvalī} 1.40–41

\begin{versinnote}
\tl{atha netikarma--\\+}
\tl{ākhupucchākāranibhaṃ sūtraṃ susnigdhanirmitam |\\+}
\tl{ṣaḍvitastimitaṃ sūtraṃ netisūtrasya lakṣanam || 1.40 ||\\+}
\tl{nāsānāle praviśyainaṃ mukhān nirgamayet kramāt |\\+}
\tl{sūtrasyāntaṃ prabaddhvā tu bhrāmayen nāsanālayoḥ |\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71r–71v (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha netī |\\+}
\tl{sūtraṃ vitastisusnigdhaṃ nāsānāle praveśayet |\\+}
\tl{mukhān nirgamayet sā hi netī siddhair nigadyate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.156 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{sūtraṃ vitastisusnigdhaṃ nāsānāle praveśayet |\\+}
\tl{mukhān nirgamayed eṣā netiḥ siddhair nigadyate ||\\!}
\end{versinnote}

Cf.~\emph{Satkarmasaṅgraha} 67

\begin{versinnote}
\tl{atha netī\\+}
\tl{mṛdu ślakṣṇaṃ sitaṃ sūtraṃ nāsānāle praveśayet |\\+}
\tl{mukhān nirgamayed dasrau cintayen netikā smṛtā ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_030]
According to Turner’s Comparative and Etymological Dictionary (1966: 427, entry 7588), the word \emph{netī} in Hindi refers to the cord of a churning stick and is cognate with the Sanskrit \emph{netra}. The action of pulling the cord of a churning stick is similar to the way the thread can be pulled back and forth, from side to side, through the nostril and mouth.

The reading we have adopted in the third verse quarter (\emph{mukhān nirgamayec caiṣā}) is well attested but may not be original because the first hemistich has a finite verb and the \emph{eṣā} must be construed with \emph{netiḥ} in the fourth verse quarter. The alternative reading \emph{mukhanirgamanād eva} (\etaOne\ and \etaTwo) may be original but is difficult to construe and \emph{mukhān nirgamayet} makes better sense in terms of describing the final part of the practice. 
\end{philcomm}
%%%%%%%%%%
\subsection*{2.31}
\begin{translation}[hp02_031]
[The technique] with a small thread purifies the skull, bestows divine sight and quickly cures a multitude of diseases that arise above the jaw.
\end{translation}
%?? expand note on possibility of niti/nīti being original (found in alpha, betaone, epsilonone, delta i.e. in good manuscripts in most groups; vernacular?) Only niti works well in final pāda. Brahmānanda rewrites to have netī as subject, which would be best. sūtrikā is old — eta1 = V1. In 22 alphatwo has nitiś caiva...
% adopt hanūrdhva
% adopt jayaty āśu sa-sūtrikā

% MD: more mss for sūtrikā:
% jayaṃ(one syllable om.) sā sasūtrikā (K1), jayaṃtī sārasūtrikā (P27), jayate sā ca sūtrikā (B2) [gamma]; jayaṃti sārasūtrikā (P6) [contaminated epsilon]
% Perhaps better to read jayaty āśu sa-sūtrikā (= neti)?
% Or: jayate (Middle)/jayantī (Pres. Part.) sā sa-sūtrikā?

%\begin{sources}[hp02_031]
%\end{sources}

\begin{testimonia}[hp02_031]
\emph{Haṭharatnāvalī} 1.42

\begin{versinnote}
\tl{kapālaśodhinī kāryā divyadṛṣṭipradāyinī | [caiva -P]\\+}
\tl{jatrūrdhvajātarogaghnī jāyate netir uttamā || 1.42 ||\\!} % cf. M1
\end{versinnote}

\emph{Yogacintāmaṇi} 71v (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{kapālaśodhanī caiva divyadṛṣṭipradīpinī |\\+}
\tl{jatrūrdhvajātarogaughān jarayaty āśu netivit ||\\!} % cf. J6
\end{versinnote}

\emph{Yuktabhavadeva} 7.156 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{kapālaśodhinī caiva divyadṛṣṭipradāyinī |\\+}
\tl{jatrūrdhvajātarogādyaiḥ jayatyeva suniścitam ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_031]
We have adopted the reading \emph{caiva} in the first verse quarter, which is attested by manuscripts of the \textbeta, \textgamma\ and \textdelta\ groups and is easy to construe. The \textalpha\ manuscripts have \emph{kaṇṭhā} and \emph{vaṭyā}, which may derive from \emph{kaṇṭhyā} (\epsilonOne), but none of these readings makes sense here. \etaOne\ and \etaTwo\ have \emph{kāryā}, which is unnecessary because of the main verb in the final verse quarter.
% MD: kaṇṭhyā (N19,G5,G11 - all epsilon)

Most witnesses, including the \emph{Jyotsnā}, read \emph{jatrūrdhva°} in the third verse quarter, which is generally understood as `above the collar bones' or, as Brahmānanda says, the area above the shoulder joints (\emph{jatruṇoḥ skandhasandhyor ūrdhvam uparibhāge}). On problems concerning the interpretation of \emph{jatru}, see Meulenbeld 1974: 465. We have adopted the reading of \textalpha, \emph{hanūrdhva°}, ‘above the jaw’, which makes good sense in the context of \emph{neti} because it is supposed to cleanse the skull (\emph{kapāla}) or, in other words, the head.

Different readings exist for the last verse quarter. \etaOne\ has \emph{jayati sā tu sūtrikā}, which is unmetrical, but the word \emph{sūtrikā} may have been original because it explains the problematic readings that arose in the other witnesses when attempts were made to replace \emph{sūtrikā} with \emph{neti}. The variants with \emph{netivit} seem implausible because of the epithets in the first hemistich, which require a feminine noun to be understood as the subject of the sentence. The reading \emph{netir āśu nihanti ca} looks like a patch that was adopted later in the transmission. The reading we have conjectured \emph{jayaty āśu tu sūtrikā} retains \emph{sūtrikā} and corrects the metrical fault of \etaOne\ by adopting \emph{jataty āśu}, which is well attested across the stemma.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.32}
\begin{translation}[hp02_032]
Now \emph{trāṭaka}:\\
{}[The yogi] should concentrate and look at a small focal point with a fixed gaze until tears fall. The experts consider this to be trāṭaka.
\end{translation}
% adopt matam

%\begin{sources}[hp02_032]
%\end{sources}

\begin{testimonia}[hp02_032]
\emph{Haṭharatnāvalī} 1.54

\begin{versinnote}
\tl{atha trāṭakam-\\+}
\tl{nirīkṣya niścaladṛśā sūkṣmalakṣyaṃ samāhitaḥ |\\+}
\tl{aśrusampātaparyantam ācāryais trāṭakaṃ smṛtam ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71v (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha trāṭakam |\\+}
\tl{nirīkṣen niścaladṛśā sūkṣmalakṣyaṃ samāhitaḥ |\\+}
\tl{aśruprapātaparyantam āryais tat trāṭakaṃ matam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.158 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha trāṭakam |\\+}
\tl{vīkṣeta niścaladṛśā sulakṣyaṃ ca samāhitaḥ |\\+}
\tl{aśrusampātaparyantam ācāryais trāṭakaṃ smṛtam ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 40cd–41ab

\begin{versinnote}
\tl{atha trāṭakam\\+}
\tl{sūkṣmalakṣye dṛśau sthāpya nirnimeṣaś ciraṃ bhavet |\\+}
\tl{aśrusampātaparyantaṃ karma trāṭakam īritam  ||\\!}
\end{versinnote}

\end{testimonia}

%\begin{philcomm}[hp02_032]

%\end{philcomm}

\begin{metre}[hp02_032]
Anuṣṭubh (a: na-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.33}
\begin{translation}[hp02_033]
It is the destroyer of eye diseases and the door [shutting out] sloth and so forth. Trāṭaka should be carefully concealed like a chest of gold.
\end{translation}

%\begin{sources}[hp02_033]
%\end{sources}

\begin{testimonia}[hp02_033]
\emph{Haṭharatnāvalī} 1.55

\begin{versinnote}
\tl{sphoṭanaṃ netrarogāṇāṃ tandrādīnāṃ kapāṭakam |\\+}
\tl{prayatnāt trāṭakaṃ gopyaṃ yathā ratnasupeṭakam ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71v (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{moṭanaṃ netrarogānāṃ tandrādīnāṃ kapāṭakam |\\+}
\tl{etac ca trāṭakaṃ gopyaṃ yathā hāṭakapeṭakam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 6.159 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{moṭakaṃ sarvarogāṇāṃ tandrādīnāṃ kapāṭanam ||\\+}
\tl{yatnatas trāṭakaṃ gopyaṃ yathā hāṭakapeṭakam ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 41cd–42ab

\begin{versinnote}
\tl{atha trāṭakam\\+}
\tl{vaṅglāvikaraṇasthe 'sminn antarjyotiḥ prakāśyate |\\+}
\tl{netrarogās tathā tandrā naśyantīty āha dhūrjatiḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_033]
The witnesses have many different readings for the first word of this verse. The \textalpha\ manuscripts are split between \emph{modaka} (\alphaOne), \emph{mocaka} and \emph{mohana}. Of these, \emph{modaka} is the most likely, if it is understood as a medicinal pill. However this meaning is rare, even in medical literature, as \emph{modaka} is generally used to refer to a small sweet. We have adopted \emph{moṭakaṃ}, which is attested by \etaOne\ and \deltaTwo, as well as the \emph{Yuktabhavadeva}. We understand it to mean “destroyer.” Bohtlingk and Roth (s.v.) and Monier-Williams (s.v.) give medicinal pill as a possible meaning of \emph{moṭaka} (cf. \emph{modaka}) but it appears that this is mainly an inference drawn only from this verse, where the pill is merely a comparison. Several witnesses have \emph{sphoṭanaṃ} (‘destroying’), which is also possible.

The reading \emph{kapāṭakam} in the second verse quarter is found in most of the witnesses and testimonia and we have adopted it accordingly. The context indicates that it means “shutter” (in the sense of shutting out something), but we have not found any parallel usages of it in this sense.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.34}
\begin{translation}[hp02_034]
Now \emph{naulī}:\\
With the shoulders lowered, [the yogi] should rotate the stomach to the left and right with the speed of a rapid whirlpool. This is called naulī by people from Gauḍa.%
\end{translation}

%\begin{sources}[hp02_034]
%\end{sources}

\begin{testimonia}[hp02_034]
\emph{Haṭharatnāvalī} 1.34

\begin{versinnote}
\tl{atha nauliḥ--\\+}
\tl{amandāvartavegena tundaṃ savyāpasavyataḥ |\\+}
\tl{natāṃso bhrāmayed eṣā nauliḥ gauḍaiḥ praśasyate ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71v (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha naulī\\+}
\tl{amandāvartavegena tundaṃ savyāpasavyayoḥ |\\+}
\tl{natāṃso bhrāmayed eṣā naulī yoge pracakṣate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 6.162 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha naulī\\+}
\tl{amandāvartavegena tundaṃ savyāpasavyataḥ |\\+}
\tl{natāṃso bhrāmayed eṣā naulir gauḍaiḥ praśasyate ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 110cd–111

\begin{versinnote}
\tl{atha naulī\\+}
\tl{amandāvartavegena jaṭharaṃ dakṣavāmayoḥ |\\+}
\tl{cālayec chaṃbhunā proktaṃ tatra lakṣmyadhidevatā  |\\+}
\tl{bāhyanaulir iyaṃ proktā jaṭharānaladīpinī ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_034]
%\emph{amandācakravegena} could mean the speed of a slightly slow wheel (Jürgen’s suggestion). But this is strange so we have adopted the usual metaphor of a fast moving whirlpool. (JB: varta is overwhelming supported, including alpha)
It is worth noting that \alphaOne\ has \emph{laulī} (instead of \emph{naulī}) as the name of this practice in verses 2.34–35 and \emph{laulikaṃ} in 2.22. The vast majority of manuscripts, including \alphaTwo, and the most important testimonia support \emph{naulī} or \emph{nauliḥ} but the names \emph{laulika} and \emph{laulikī} do occur in some more recent works, such as the \emph{Haṭhayogasaṃhitā} (p. 4), \emph{Gheraṇḍasaṃhitā} (1.12), \emph{Yogasārasaṅgraha} (pp. 54–55) and \emph{Yogakarṇikā} (p. 56).

%??% JM to supply Babaji video to be uploaded to our host. Send to Mitsuyo.

The reading \emph{gauḍaiḥ} is found in only one collated manuscript (\etaTwo) but it is close to the readings of two \textalpha\ manuscripts, \emph{gaulaiḥ} (\alphaOne) and \emph{golaiḥ} (\alphaThree), and is also attested by the \emph{Haṭharatnāvalī} and \emph{Yuktabhavadeva}. This version of the verse, which is likely original, suggests that it was thought that the name \emph{naulī} came from the region known today as Bengal. The reading \emph{gauḍaiḥ} explains many of the mistakes in other manuscripts, such as \emph{gaulīḥ}, \emph{maulaiḥ}, etc.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.35}
\begin{translation}[hp02_035]
Naulī brings about stimulation of the fire in the stomach, [good] digestion and the like, always brings bliss, and makes all humoural disorders and diseases wither away. This naulī is the best of all Haṭha techniques.
\end{translation}
%adopt tundāgni

\begin{sources}[hp02_035]
\end{sources}

\begin{testimonia}[hp02_035]
\emph{Haṭharatnāvalī} 1.35

\begin{versinnote}
\tl{tundāgnisandīpanapācanādisandīpikānandakarī sadaiva |\\+}
\tl{aśeṣadoṣāmayaśoṣaṇī ca haṭhakriyāmaulir iyaṃ ca nauliḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71v (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{mandāgnisaṃdīpanapācanāgnisaṃdhāyikānandakarī tathaiva |\\+}
\tl{aśeṣadoṣāmayaśoṣinī ca haṭhakriyāmaulir iyaṃ hi naulī ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.163 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{mandāgnisandīpanapācanādisandhāvanānandakarī sadaiva |\\+}
\tl{aśeṣadoṣāmayaśoṣaṇīyaṃ haṭhakriyāmaulir iyaṃ hi nauliḥ ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 8.12

\begin{versinnote}
\tl{mandāgnisandīpanapācanādisandhāyikānandakarī sadaiva |\\+}
\tl{aśeṣadoṣopacayaśoṣaṇīva haṭhakriyā 'sau jayatīha nauliḥ ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_035]
\alphaOne\ and \alphaTwo, \betaTwo, and \epsilonOne\ have \emph{tundāgni°} in the first verse quarter. This reading is also supported by all of the manuscripts collated for the published edition of the \emph{Haṭharatnāvalī}. The compound \emph{tundāgni} is rare in Sanskrit literature but it was likely used here because the term \emph{tunda} appears in the previous verse, which explains how naulī is done. The compound \emph{tundāgni} seems synonymous with \emph{udarāgni}, \emph{jaṭharāgni}, \emph{śarīrāgni}, etc., which are commonly used in yoga texts to refer to the body's digestive fire. The alternative reading \emph{mandāgni°} (`sluggish fire') is reasonably common in contexts of stimulating poor digestion, and may have been introduced early in the transmission to replace the more unusual \emph{tundāgni°}.

%We should adopt \emph{pācana} in the sense of digestion (cooking the food in the intestines). \emph{āmaya} makes better sense in the third pāda, and \emph{mala} might have crept in because of the association of this word with \emph{śoṣiṇī}. (JB: Both pācana and āmaya are well attested, so no need to comment).

Most witnesses and the testimonia have \emph{maulir iyaṃ} in the fourth verse quarter, which expresses the idea that naulī was thought to be the best of the \emph{ṣaṭkarma}, and the assonance of \emph{naulī} and \emph{mauli} may have been intended. The alternative reading of \emph{mūlam iyaṃ} in \etaOne\ and \epsilonOne\ would suggest that \emph{nauli} is necessary for the other practices, which does not seem to be the case because, according to \emph{Haṭhapradīpikā} 2.21, the \emph{ṣaṭkarma} are more like therapeutic interventions.
\end{philcomm}

\begin{metre}[hp02_035]
Upajāti
\end{metre}

%%%%%%%%%%
\subsection*{2.36}
\begin{translation}[hp02_036]
Now kapālabhātī:\\
Rapid inhalation and exhalation like the bellows of a blacksmith is called kapālabhātī, the skull bellows. It dries up imbalances of phlegm.
\end{translation}

%\begin{sources}[hp02_036]
%\end{sources}

\begin{testimonia}[hp02_036]
\emph{Haṭharatnāvalī} 1.56

\begin{versinnote}
\tl{atha kapālabhastrikā--\\+}
\tl{bhastrival lohakārāṇāṃ recapūrasusambhramau |\\+}
\tl{kapālabhastrī vikhyātā sarvarogaviśoṣaṇī ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} 71v (attr.~to the Haṭhapradīpikā)

\begin{versinnote}
\tl{atha kapālabhātī |\\+}
\tl{bhastreva lohakārāṇāṃ recapūrau sasaṃbhramau |\\+}
\tl{kapālabhātī vikhyātā kaphadoṣaviśoṣiṇī ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.163 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{atha kapālabhātiḥ |\\+}
\tl{bhastrāval lohakārāṇāṃ recapūrau sasambhramau |\\+}
\tl{kapālabhātir vikhyātā kaphadoṣaviśoṣiṇī ||\\!}
\end{versinnote}

Cf. \emph{Satkarmasaṅgraha} 50cd–51

\begin{versinnote}
\tl{atha bhastrā\\+}
\tl{lohakārasya bhastreva recapūrau tu vegataḥ ||\\+}
\tl{punaḥ punaḥ prakurvīta sthiramūrdhnā prayatnataḥ |\\+}
\tl{sthirabhastreti ca khyāta yogināṃ siddhidāyakā ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_036]
The word \emph{bhātī} is derived from \emph{bhastrī} (Turner 1966: 537, entry 9424).

\etaTwo\ and other manuscripts have \emph{kuryāt savyāpasavyataḥ} (‘left and right’) instead of \emph{recapūrau sasambhramau}. Although one might infer that \emph{savyāpasavyataḥ} is referring to performing the skull bellows breathing alternately through the left and right nostrils, it appears to be a secondary reading because there is no indication of what is moving to the left and right. Such a method of alternate nostril breathing is explained as a variation of kapālabhāti in the \emph{Haṭhayogasaṃhitā} (p. 14):
\begin{versinnote}
\tl{iḍayā pūrayed vāyuṃ recayet piṅgalākhyayā |\\+}
\tl{piṅgalayā pūrayitvā punaś candreṇa recayet |\\+}
\tl{pūrakaṃ recakaṃ kṛtvā vegena na tu cālayet ||\\!}
\end{versinnote}

In the second verse quarter, the \textalpha\ manuscripts have \emph{susambhramau} ('great speed') instead of the reading we have adopted \emph{sasambhramau} (`fast'). The syntax requires \emph{sasambhramau} because it is an adjectival compound that must agree with \emph{recapūrau} (`the exhalation and inhalation').  
\end{philcomm}

\begin{metre}[hp02_036]
Anuṣṭubh (c: ma-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.37}
\begin{translation}[hp02_037]
The person whose excess weight, phlegm, fat, impurities and the like have been removed by the six techniques should then perform breath-control. It succeeds without effort.
\end{translation}

%\begin{sources}[hp02_037]
%\end{sources}

\begin{testimonia}[hp02_037]
\emph{Haṭharatnāvalī} 1.60

\begin{versinnote}
\tl{karmāṣṭabhir gatasthaulyaṃ kaphamedomalādikam |\\+}
\tl{prāṇāyāmaṃ tataḥ kuryād anāyāsena siddhyati  ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~8v

\begin{versinnote}
\tl{tathā cātmārāmaḥ\\+}
\tl{ṣaṭkarmanirgatasthaulyakaphamedogadādikaḥ |\\+}
\tl{prāṇāyāmaṃ tataḥ kuryād anāyāsena sidhyati ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.165 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{ṣaṭkarmabhir gatasthaulyaṃ kaphamedomalātigaḥ |\\+}
\tl{prāṇāyāmaṃ tataḥ kuryād anāyāsena sidhyati ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_037]
%Most witnesses support \emph{ṣaṭkarmabhir gata°}, which is somewhat unconventional syntax. It seems more likely that it was corrected to \emph{ṣaṭkarmanirgata°} than the latter being original. (JB: \emph{ṣaṭkarmabhir gata°} is overwhelmingly supported. No need to justify it.).
%The variant reading \emph{°ādhikaḥ} in the second \emph{pāda} is inferior because one would assume that the \emph{ṣaṭkarma} should remove all excess weight (\emph{sthaulya}) and impurities (\emph{mala}). 
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.38}
\begin{translation}[hp02_038]
Some teachers say that all impurities are dried up by means of breath-controls alone and do not recommend any other practice.
\end{translation}

%\begin{sources}[hp02_038]
%\end{sources}

\begin{testimonia}[hp02_038]
\emph{Yogacintāmaṇi} ff.~8v–9r (attr.~to Ātmārāma)

\begin{versinnote}
\tl{prāṇāyāmair eva sarvaiḥ praśuṣyanti malā yataḥ iti |\\+}
\tl{ācāryāṇāṃ tu keṣāṃ cid anya krama na saṃmatam iti ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.166 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{prāṇāyāmair eva sarve praśuṣyanti malā iti |\\+}
\tl{ācāryāṇāṃ tu keṣāñ cid anyat karma na sammatam ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_038]
The plural of \emph{prāṇāyāma}, which we have translated here as ‘breath-controls’, probably refers to practising multiple repetitions of breath retentions with alternate nostril breathing.

%The verb \emph{praśuṣyanti} is well attested and makes good sense in the context of impurities (\emph{mala}), which usually supposed to dry up and wither away because of the practice of yoga. (JB: \emph{praśuṣyanti}  is so well attested and straight forward that there's no need to comment on it) 
Many witnesses lower on the stemma have \emph{malāśaya} (instead of \emph{malā api} or \emph{malā iti}) in the second verse quarter but this usually has the more specific meaning of bowels or bladder and so seems inappropriate in a general statement. The witnesses that have \emph{malāśaya} also have the verb \emph{pra+śudh}, which connotes that the place where the impurities accumulate is cleaned (rather than the impurities themselves). 
\end{philcomm}

\begin{metre}[hp02_038]
Anuṣṭubh (a: ra-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.39}
\begin{translation}[hp02_039]
Even Brahmā and the other gods became devoted to breath practice through fear of death, so one should perform breath practice.
\end{translation}

%\begin{sources}[hp02_039]
%\end{sources}

\begin{testimonia}[hp02_039]
\emph{Haṭharatnāvalī} 3.82

\begin{versinnote}
\tl{brahmādayo 'pi tridaśāḥ pavanābhyāsatatparāḥ |\\+}
\tl{abhūvan mṛtyurahitā tasmāt pavanam abhyaset ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 8.19

\begin{versinnote}
\tl{brahmādayo 'pi tridaśāḥ pavanābhyāsatatparāḥ |\\+}
\tl{tena siddhiṃ gatā yoge tasmāt pavanam abhyaset ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_039]
This verse has been rewritten in \etaTwo\ and the delta group of manuscripts. \etaTwo's reading attempts, somewhat unsuccessfully, to connect this verse more directly to the \emph{ṣaṭkarma}:
\begin{versinnote}
\tl{ṣaṭkarmayogam āpnoti pavanābhyāsatatparaḥ |\\+}
\tl{sumanaskāṃtako bhavya[s] tasmāt pavanam abhyaset ||\\!}
\end{versinnote}
Different versions of this verse occur in \etaTwo\ and \epsilonThree. Neither of these appear to be original as the compound \emph{sumanaskāntaka} (`dying with the mind active'?) is rather odd. The \delta \ group of manuscripts has another reading for the third quarter (\emph{tena siddhiṃ gatās te ca}), which puts a more positive spin on the verse in as much as one should practise \emph{prāṇāyāma} because through it the gods attained perfection. 
%\epsilonThree\ has a similar version of the third quarter (\emph{samanaskāntakabhayaṃ}), which might refer to the fear of dying with the mind active as opposed to the preferable situation of dying while in \emph{samādhi}.
\end{philcomm}

\begin{metre}[hp02_039]
Anuṣṭubh (a: bha-vipulā; c: na-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.40}
\begin{translation}[hp02_040]
So long as the breath is bound in the body, so long as the mind is without support, so long as the gaze is on the middle of the brow, where is the fear of death?
\end{translation}
% adopt vīkṣā from Six-chapter VM

\begin{sources}[hp02_040]
\emph{Vivekamārtaṇḍa} 73

\begin{versinnote}
\tl{yāvad baddho marud dehe tāvac cittaṃ nirāśrayam |\\+}
\tl{yāvad vīkṣā bhruvor madhye tāvat kālabhayaṃ kutaḥ ||\\+}
\tl{\var{vīkṣā] T; dṛṣṭir \emph{cett.}}}

\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_040]
\emph{Yogacintāmaṇi} f.~92r (attr.~to the Skandapurāṇa)

\begin{versinnote}
\tl{yāvad baddho marud dehe yāvad vṛttau nirāśrayam |\\+}
\tl{yāvad dṛṣṭir bhruvor madhye tāvat kālabhayaṃ kutaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.8 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{yāvad baddho marud dehe yāvac cittaṃ nirāmayam |\\+}
\tl{yāvad dṛṣṭir bhruvor madhye tāvat kālabhayaṃ kutaḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_040]
We have adopted the reading \emph{vīkṣā} in the third verse quarter, which is an emendation of \alphaOne\ and \alphaTwo's \emph{vīkṣed}. The verb \emph{vīkṣed} is likely an error as the \emph{yāvat} clauses appear to have been written as nominal phrases. The reading \emph{vīkṣā} occurs in the same verse of the six-chapter version of the \emph{Vivekamārtaṇḍa} (siglum T), which sometimes preserves old readings of that text.

%The verb \emph{paśyet} in the third \emph{pāda} is well attested among the manuscripts (V1, J10, etc.). However, it often occurs with \emph{bhruvor madhye}, which is rather odd and suggests that the alternative reading \emph{dṛṣṭir} was original. In fact, \emph{dṛṣṭir} is supported by the manuscripts of the source text, the \emph{Vivekamārtaṇḍa}, and the testimonia. (JB: This is an old note).
\end{philcomm}

%%%%%%%%%%
\subsection*{2.41}
\begin{translation}[hp02_041]
When the network of channels has been purified by breath-controls as prescribed, the breath pierces the mouth of \emph{suṣumṇā} and enters it with ease.
\end{translation}

%\begin{sources}[hp02_041]
%\end{sources}

\begin{testimonia}[hp02_041]
\emph{Haṭharatnāvalī} 2.2

\begin{versinnote}
\tl{vidhivat prāṇasaṃyāmaiḥ nāḍicakre viśodhite |\\+}
\tl{suṣumnāvadanaṃ bhitvā sukhād viśati mārutaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~18r

\begin{versinnote}
\tl{haṭhapradīpikāyām—\\+}
\tl{vividhaiḥ prāṇasaṃyāmaiḥ nāḍīcakre viśodhite |\\+}
\tl{suṣumnāvadanaṃ bhitvā sukhād viśati mārutaḥ ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_041]
The compound \emph{prāṇasaṃyāmaiḥ} (‘breath-controls’) should be understood here as a synonym for \emph{prāṇāyāmaiḥ} as found in 2.38 (on which see the note on this verse).
\end{philcomm}

%%%%%%%%%%
\subsection*{2.42}
\begin{translation}[hp02_042]
When the breath moves in the middle, stillness of the mind arises. The mind’s becoming very still is the ‘mind beyond mind’ (\emph{manonmanī}) state.
\end{translation}

%\begin{sources}[hp02_042]
%\end{sources}

\begin{testimonia}[hp02_042]
\emph{Haṭharatnāvalī} 2.3

\begin{versinnote}
\tl{mārute madhyame jāte manaḥsthairyaṃ prajāyate |\\+}
\tl{manasaḥ susthirībhāvaḥ saivāvasthā manonmanī ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~18a (attr.~to Haṭhapradīpikā)

\begin{versinnote}
\tl{mārute madhyasaṃcāre manaḥsthairyaṃ prajāyate |\\+}
\tl{yo manaḥsusthirībhāvaḥ saivāvasthā manonmanī ||\\!}
\end{versinnote}

\end{testimonia}

%\begin{philcomm}[hp02_042]
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.43}
\begin{translation}[hp02_043]
In order to achieve that, he who knows [their] methods should perform various retentions. As a result of the practice of the various retentions, [the yogi] obtains various results.
\end{translation}
% adopt the singular °jñaś .. kurvīta

%\begin{sources}[hp02_043]
%\end{sources}

\begin{testimonia}[hp02_043]
\emph{Haṭharatnāvalī} 2.4
\begin{versinnote}
\tl{tatsiddhaye vidhānajñaḥ sadā kurvīta kumbhakān |\\+}
\tl{vicitrakumbhakābhyāsād vicitrāṃ siddhim āpnuyāt ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.92 (attr.~to the \emph{Yājñavalkyagītā})

\begin{versinnote}
\tl{tatsiddhaye vidhānajñāś citrān kurvanti kumbhakān |\\+}
\tl{vicitrakumbhakābhyāsād vicitrāṃ siddhim āpnuyāt ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_043]
The majority of witnesses, including \alphaTwo, have a plural subject (\emph{vidhānajñāḥ}) and verb (\emph{kurvanti}) in the first hemistich. We have adopted the singular, which is attested by \alphaOne, \alphaThree\ and manuscripts of the \emph{Haṭharatnāvalī}, because it corresponds with the singular subject of the second hemistich. 
\end{philcomm}

%%%%%%%%%%
\subsection*{2.44}
\begin{translation}[hp02_044]
Sūryabhedana, ujjāyī, śītkā, śītalī, bhastrikā, bhrāmarī, mūrcchā and plāvanī: these are the eight kumbhakas.
\end{translation}
% adopt śītka
% adopt plāvanī

\begin{sources}[hp02_044]
\end{sources}

\begin{testimonia}[hp02_044]

\emph{Haṭharatnāvalī} 2.6
\begin{versinnote}
\tl{sūryabhedanam ujjayī tathā sītkāraśītalī |\\+}
\tl{bhastrikā bhrāmarī mūrcchā kevalaś cāṣṭa kumbhakāḥ ||\\!}
\end{versinnote}

\emph{Yogalakṣaṇāvalī} f.~32r (attrib. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{sūryabhedanam ujjayī tathā sītkā ca sītalī |\\+}
\tl{bhastrikā bhrāmarī mūrcchā kevalāś cāṣṭa kumbhakāḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101r

\begin{versinnote}
\tl{haṭhayoge—\\+}
\tl{sūryabhedanam ujjāyī tathā sītkāraśītalī |\\+}
\tl{bhastrikā bhramarī mūrcchā sahitaṃ cāṣṭa kumbhakāḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.93 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{sūryabhedanamujjāyī sītkārī śītalī tathā |\\+}
\tl{bhastrikā bhrāmarī mūrcchā kevalāś cāṣṭa kumbhakāḥ ||\\!}    
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_044]
In the fourth verse quarter, only some manuscripts of the \textepsilon\ and \textgamma\ groups have the adopted reading \emph{plāvanīty aṣṭa}. The name \emph{plāvanī} is necessary in this list because most manuscript groups, including \textalpha, and the testimonia contain a verse on \emph{plāvinīkumbhaka} as one of the eight \emph{kumbhaka}s. The alternative reading \emph{kevalī/kevalaś}, which is supported by the \textalpha\ group and other manuscripts, appears to be a mistake because it is not consistent with the idea of \emph{kevalakumbhaka} standing outside the category of \emph{sahitakumbhaka}, as stated in 2.72–75. These witnesses are among those which include a verse on \emph{plāvinī} later in the chapter. 
%
%This idea of two types of kumbhaka is found in the source text from which Svātmārāma borrowed four of the kumbhakas, namely, the \emph{Gorakṣaśataka}. 
%Therefore, it seems likely the word \emph{plāvinī} dropped out of some manuscripts and was replaced by \emph{kevala} because \emph{kevala} is discussed later in the chapter. In fact, \emph{plāvinī} may have been removed by some scribes because its heading is omitted in many witnesses. The reading in V19 and the \emph{Yogacintāmaṇi} (\emph{sahitāś cāṣṭa}) was probably an attempt to remove \emph{kevalī/kevalaś} from the list.
% mention Haṭharatnāvalī (absence of plāvinī)
% [JB I think the second paragraph is unnecessary]
\end{philcomm}

%%%%%%%%%%
\subsection*{2.45}
\begin{translation}[hp02_045]
At the end of the inhalation, the lock called \emph{jālandhara} is to be performed, while at the end of the retention and beginning of the exhalation, \emph{uḍḍiyāna} is to be performed.
\end{translation}

\begin{sources}[hp02_045]
\emph{Gorakṣaśataka} 62ab

\begin{versinnote}
\tl{pūrakānte tu kartavyo bandho jālandharābhidhaḥ |\\!}
\end{versinnote}
\emph{Gorakṣaśataka} 58ab

\begin{versinnote}
\tl{kuṃbhakānte recakādau kartavyoḍḍiyaṇābhidhaḥ |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_045]
\emph{Haṭharatnāvalī} 2.7

\begin{versinnote}
\tl{pūrakānte tu kartavyo bandho jālandharābhidhaḥ |\\+}
\tl{kumbhakānte recakādau kartavyas tūḍḍiyānakaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇī} f.~80r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{pūrakānte tu kartavyo bandho jālandharābhidhaḥ |\\+}
\tl{kumbhakānte recakādau kartavyas tūḍḍiyānakaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.94 (attrib. to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{pūrakānte ca karttavyo bandho jālandharābhidhaḥ |\\+}
\tl{kumbhakānte recakādau karttavyas tūḍḍiyānakaḥ || \\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_045]
%\end{philcomm}

\begin{metre}[hp02_045]
Anuṣṭubh (c: ra-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.46}
\begin{translation}[hp02_046]
By quickly contracting the lower region when the throat has been constricted and stretching back the middle [of the body] the breath goes into the channel of Brahman.
\end{translation}

\begin{sources}[hp02_046]
\emph{Gorakṣaśataka} 62cd–63ab
\begin{versinnote}
\tl{adhastāt kuñcanenaiva kaṇṭhasaṃkocane kṛte |\\+}
\tl{madhye paścimatāṇena syāt prāṇo brahmanāḍigaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_046]
\emph{Haṭharatnāvalī} 2.8
\begin{versinnote}
\tl{adhastāt kuñcanenāśu kaṇṭhasaṅkocane kṛte |\\+}
\tl{madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇī} f.80r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{adhas tv ākuñcanenāśu kaṇṭhasaṅkocanena ca |\\+}
\tl{madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||\\!}
\end{versinnote}

\emph{Yogabīja} 110 (southern recension)
\begin{versinnote}
\tl{adhastāt kuñcanenāśu kaṇṭhasaṅkocane kṛte |\\+}
\tl{madhye paścimatānena syāt prāṇo brahmanāḍigaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.95 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{adhastāt kuñcanenāśu kaṇṭhasaṃkocane kṛte |\\+}
\tl{madhye paścimatānena syāt prāṇo madhyanāḍiga ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 15.25–27
\begin{versinnote}
\tl{adhastāt kuñcanenaiva kaṇṭhasaṃkocanena ca |\\+}
\tl{madhye paścimatānena syāt prāṇo brahmarandhragaḥ ||\\+}
\tl{prāṇaḥ prāṇavāyuḥ brahmarandhragaḥ suṣumnāpathacārī syāt | madhyago bhavet | \\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_046]
%\emph{madhyapaścimatānena} is possible and well attested.
As is clear in the source text, the \emph{Gorakṣaśataka}, the three techniques alluded to here are \sl{mūlabandha, jālandharabandha} and \emph{uḍḍiyānabandha} respectively (on which see chapter three).
\end{philcomm}

%%%%%%%%%%
\subsection*{2.47}
\begin{translation}[hp02_047]
The yogi should raise up \emph{apānavāyu} and lead \emph{prāṇa} down from the throat. Freed from ageing, he becomes sixteen years old.
\end{translation}

%\begin{sources}[hp02_047]
%\end{sources}

\begin{testimonia}[hp02_047]
\emph{Haṭharatnāvalī} 2.9
\begin{versinnote}
\tl{apānam ūrdhvam utthāpya prāṇaṃ kaṇṭhād adho nayet |\\+}
\tl{yogī jarāvimuktaḥ syāt ṣoḍaśo vayasā bhavet ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇī} f.~80r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{apānam ūrdhvam utthāpya prāṇaṃ kaṇṭhād adho nayet |\\+}
\tl{yogī jarāvimuktaḥ san vayasā ṣoḍaśo bhavet ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.96 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{apānam ūrdhvam utthāpya prāṇaṃ kaṇṭhād adho nayet |\\+}
\tl{yogī jarāvinirmuktaḥ ṣoḍaśo vayasā bhavet ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_047]
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.48}
\begin{translation}[hp02_048]
Now, piercing the sun---\\
The yogi should sit in \emph{vajrāsana} on a comfortable mat, slowly draw in external air through the right nostril, [...]
\end{translation}

\begin{sources}[hp02_048]
Cf. \emph{Gorakṣaśataka} 33–34ab
\begin{versinnote}
\tl{pavitre nātyuccanīce hy āsane sukhade śubhe |\\+}
\tl{baddhvā vajrāsanaṃ kṛtvā sarasvatyāś ca cālanam ||\\+}
\tl{dakṣanāḍyāṃ samākṛṣya bahiṣṭhaṃ pavanaṃ śanaiḥ |\\!}
\end{versinnote} 
\end{sources}

\begin{testimonia}[hp02_048]
\emph{Yogalakṣaṇāvalī} f.~32r (attrib. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{baddhavajrāsano dakṣanāḍyākṛṣyānilaṃ śanaiḥ |\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{āsane sukhade yogī baddhavajrāsanas tataḥ |\\+}
\tl{dakṣanāḍyā samākṛṣya bahiḥsthaṃ pavanaṃ śanaiḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.98 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{āsane sukhade yogī baddhvā padmāsanaṃ tataḥ |\\+}
\tl{dakṣanāḍyā samākṛṣya bahiḥsthaṃ pavanaṃ śanaiḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_048]
%\emph{āsane sukhade} is supported by the [different but parallel] reading in the Gorakṣa\-śataka. (JB: I've put the GŚ verses in the source section)
%After citing this verse with attribution to the \emph{Haṭhapradīpikā} in his \emph{Haṭhasaṅketacandrikā} (f.~78v), Sundaradeva remarks that \emph{sukhade} is a \emph{hetugarbhaviśeṣaṇam}, which indicates that he understood the first hemistich as saying that the yogi assumes \emph{vajrāsana} on a mat (\emph{āsana}) because it is comfortable.%JM:needs more explanation.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.49}
\begin{translation}[hp02_049]
and hold the breath as far as the tips of the hair and nails until cessation [of the breath]. The wise man should then exhale the breath slowly through the left nostril.
\end{translation}
% adopt ā keśāgrān nakhāgrāc ca

%\begin{sources}[hp02_049]
%\end{sources}

\begin{testimonia}[hp02_049]
\emph{Yogalakṣaṇāvalī} f.~32r (attrib. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ā nakhāgrālakāgrāntaṃ kumbhayitvā yathāsukham |\\+}
\tl{savyanāḍyā tato mandaṃ recayet pavanaṃ sudhīḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Yogabīja})
\begin{versinnote} 
\tl{ā keśāgraṃ nakhāgraṃ ca śirodhāvadhi kumbhakam |\\+}
\tl{tataḥ śanaiḥ savyanāḍyā recayet pavanaṃ sudhīḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.99 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{ā keśād ā nakhāgrāc ca nirodhāvadhi kumbhayet |\\+}
\tl{tataḥ śanaiḥ savyanāḍyā recayet pavanaṃ sudhīḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_049]
%Three versions of  the first pāda are attested by the manuscripts:\\ 
%1. ā keśāgraṃ nakhāgraṃ ca (Gr3,Gr4c,V1).\\
%2. ā keśāgrān nakhāgrāc ca (N3,V3,J10).\\
%3. ā keśād ā nakhāgrāc ca (Gr2,Jyo).\\
%The second reading is supported by \textalpha\ and 
% (We've read with the stemma)

%When explaining \emph{ā keśād ā nakhāgrāc ca} in his \emph{Jyotsnā} (2.49), Brahmānanda cites a verse stating that if the breath is held forcefully, it escapes through the hair follicles or nails and destroys the body. The ablative with \emph{ā} (\emph{a keśād nakhāgrāc ca}) is also well-attested (group 2 manuscripts), and it would mean the same.
% [JB: this ref is in the discussion below]

An antecedent to the idea of \emph{prāṇāyāma} affecting the whole body (i.e., as far as the tips of the hair and nails) occurs in the \emph{Baudhāyanadharmasūtra} (4.1.23):

\begin{versinnote}
{}[The yogi] who is constantly engaged [in practice] should repeat breath retentions again and again. Extreme heat burns as far as the tips of the hair and nails. 
\end{versinnote}

\begin{versinnote}
\tl{āvartayet sadā yuktaḥ prāṇāyāmān punaḥ punaḥ |\\+}
\tl{ā keśāntān nakhāgrāc ca tapas tapyata uttamam ||\\!}
\end{versinnote}

The meaning of \emph{nirodhāvadhi} is not entirely clear, but all sources and the \emph{Jyotsnā} (2.49) agree on this reading. The problem is that to practise \emph{kumbhaka} ``up to cessation \emph{nirodha}'' seems to suggest that cessation is not that of the physical breath, which by definition ceases in \emph{kumbhaka}, but of the vital wind (\emph{prāṇa}) within the body. By citing a verse from an unnamed text, Brahmānanda seems to understand this verse as saying that the breath should very carefully (\emph{atiprayatnena}) be held as far as the extremities of the body so that it does not damage the body by exiting through the hair follicles:

\begin{versinnote}
When the breath has been stopped forcefully, it flows out through the hair follicles. This destroys the body and also causes skin diseases and the like.
\end{versinnote}

\begin{versinnote}
\tl{haṭhān niruddhaḥ prāṇo 'yaṃ romakūpeṣu niḥsaret |\\+}
\tl{dehaṃ vidārayaty eṣa kuṣṭhādi janayaty api ||\\!}%
\end{versinnote}
%??JM: how about emending to virodhāvadhi, “until it is uncomfortable”, which would make much better sense. Yes, to add to note but not emend.
\end{philcomm}

\begin{metre}[hp02_049]
Anuṣṭubh (c: ra-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.50}
\begin{translation}[hp02_050]
This purifies the skull, cures [imbalances] of the wind humour [and] gets rid of diseases caused by worms [so] should be done repeatedly. It is called the piercing of the sun.
\end{translation}

\begin{sources}[hp02_050]
\emph{Gorakṣaśataka} 35–36ab

\begin{versinnote}
\tl{kapālaśodhane vāpi recayet pavanaṃ sudhīḥ |\\+}
\tl{tundasya vātadoṣaghnaḥ kṛmidoṣaṃ nihanti ca ||\\+}
\tl{punaḥ punar idaṃ kāryaṃ sūryābhedam udāhṛtam |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_050]
\emph{Haṭharatnāvalī} 2.11cd–12

\begin{versinnote}
\tl{kapālaṃ śodhanaṃ cāpi recayet pavanaṃ śanaiḥ ||\\+}
\tl{kapālaṃ ... śanaiḥ ] kapālaśodhanaṃ vātadoṣaghnaṃ kṛmināśanaṃ N,n1,n4.\\+}
\tl{ālasyaṃ vātadoṣaghnaṃ kṛmikīṭaṃ nihanti ca |\\+}
\tl{punaḥ punar idaṃ kāryaṃ sūryabhedākhyakumbhakam ||\\!}
\end{versinnote}

\emph{Yogalakṣaṇāvalī} f.~32r (attrib. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{kapālaśodhanaṃ caitad vātaghnaṃ kṛmidoṣanut ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Yogabīja})
\begin{versinnote} 
\tl{kapālaśodhanaṃ vātadoṣaghnaṃ kṛmidoṣahṛt |\\+}
\tl{punaḥ punar idaṃ kuryāt sūryabhedanam uttamam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.100 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{kapālaśodhanaṃ vātadoṣaghanaṃ kṛmidoṣaham |\\+}
\tl{punaḥ punar idaṃ kāryaṃ sūryabhedam udāhṛtam ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_050]
%The \emph{Gorakṣaśataka} and \emph{Yogacintāmaṇi} support \emph{kṛmidoṣa}. [JB: Not an issue in the new collation]
The compound \emph{sūryabheda} is metri causa, as the name given in 2.44 and the heading of 2.48 is \emph{sūryabhedana}. Some witnesses, such as the delta group and \getsiglum{V19}, have attempted to reinstate the name \emph{sūryabhedana}.

Both \emph{°doṣaham} and \emph{°doṣahṛt} are well attested and possible. We have favoured the former because the \textalpha\ reading (\emph{doṣajam}) appears to be a corruption of it.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.51}
\begin{translation}[hp02_051]
Now ujjāyī:

[The yogi] should close the mouth and gradually draw in the breath through the nostrils so that it comes into contact [with the region] from the throat to the chest and makes a sound.\end{translation}

\begin{sources}[hp02_051]
\emph{Gorakṣaśataka} 36c–37b

\begin{versinnote}
\tl{mukhaṃ saṃyamya nāḍībhyāṃ ākṛṣya pavanaṃ śanaiḥ |\\+}
\tl{yathā lagati kaṇṭhāt tu hṛdayāvadhi sasvanam || \\+}
\tl{\var{kaṇṭhāt tu] kaṇṭhaṃ tu T}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_051]
\emph{Haṭharatnāvalī} 2.13
\begin{versinnote}
\tl{mukhaṃ saṃyamya nāḍībhyām ākṛṣya pavanaṃ śanaiḥ |\\+}
\tl{yathā lagati hṛtkaṇṭhaṃ hṛdayāvadhi svasvanaḥ ||\\+}
\tl{\var{hṛtkaṇṭhaṃ ] hṛtkaṇṭhe N, n1, n4, J} \\!}%
\end{versinnote}

\emph{Yogalakṣaṇāvalī} f.~32r (attrib.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{mukhaṃ saṃyamya nāsābhyāṃ ākṛṣya pavanaṃ śanaiḥ |\\+}
\tl{yathā lagati kaṇṭhe suḥ hṛdayāvadhi sasvanam ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{mukhaṃ saṃyamya nāḍībhyām ākṛṣya pavanaṃ punaḥ |\\+}
\tl{yathā lagati hṛtkaṇṭhād dhṛdayāvadhi sasvanaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.101 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote} 
\tl{mukhaṃ niyamya nāḍībhyām ākṛṣya pavanaṃ śanaiḥ |\\+}
\tl{yathā lagati kaṇṭhāt tu hṛdayāvadhi pūraṇam ||\\!}
\end{versinnote}

Cf.~\emph{Haṭhatattvakaumudī} 10.7
\begin{versinnote}
\tl{athojjāyī kumbhakaḥ\\+}
\tl{āsyaṃ saṃyamya nāsāpuṭayugasuṣirābhyāṃ samākṛṣya vāyuṃ\\+}
\tl{mandaṃ mandaṃ yathāsau lagati galataṭād āhṛdantaḥ saśabdaḥ |\\+}
\tl{ruddhvā keśān nakhāgrāvadhi pavanam amuṃ recayed vāmanāḍyā\\+}
\tl{proktojjāyīti kumbhaḥ kaphagadadalano dīpti kṛjjāṭharāgne ||\\+}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_051]
The use of \emph{lagati} without a locative or direct object (as found in the source text, the \emph{Gorakṣaśataka}) is supported by the paraphrase of the verse in the \emph{Haṭhatattvakaumudī} (10.7). Most witnesses (including \textalpha) have \emph{kaṇṭhāt tu hṛdayāvadhi}, which we have understood in the sense of a locative as it specifies the place within the body where the contact occurs. 
\end{philcomm}

%%%%%%%%%%
\subsection*{2.52}
\begin{translation}[hp02_052]
As before, he should hold the breath and then exhale through Iḍā. [Because] it cures disorders caused by phlegm in the throat and increases the body’s fire, [\dots]
\end{translation}

\begin{sources}[hp02_052]
\emph{Gorakṣaśataka} 37c–38b
\begin{versinnote}
\tl{pūrvavat kumbhayet prāṇaṃ recayed iḍayā tataḥ |\\+}
\tl{śīrṣotthitānalaharaṃ galaśleṣmaharaṃ paraṃ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_052]
\emph{Haṭharatnāvalī} 2.14

\begin{versinnote}
\tl{pūrvavat kumbhayet prāṇaṃ recayed iḍayā tataḥ |\\+}
\tl{gale śleṣmaharaṃ proktaṃ dehānalavivardhanam ||\\!}
\end{versinnote}

\emph{Yogalakṣaṇāvalī} f.~32r (attrib. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{pūrvavat kuṃbhayet prāṇān īḍayā recayet tataḥ |\\+}
\tl{śleṣmadoṣaharaṃ caitad dhāturogavināśanam ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{pūrvavat kumbhayet prāṇaṃ recayed iḍayā tataḥ |\\+}
\tl{śleṣmadoṣaharaṃ kaṇṭhe dehānalavivardhanam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.102 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote} 
\tl{pūrvavat kumbhayet prāṇān recayediḍayā tataḥ |\\+}
\tl{śleṣmoṣaharaṃ dehānaladīptipravardhanam ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_052]
In the second hemistich of this verse, many of the readings in the oldest manuscripts, such as \emph{dehād analadīptivardhanam} (\getsiglum{V1}), \emph{dehānaladīptivivardhanam} (\getsiglum{J10}) and \emph{dehe [’]naladīptivivardhanam} (\getsiglum{P28}), are unlikely to be original because both the source and testimonia indicate that \emph{Ujjāyī} is supposed to remove phlegm from the throat. These versions may have arisen from attempts to remove \emph{kaṇṭhe} in the third \emph{pāda}, which was thought to be hanging. It appears that \emph{kaṇṭhe dehānalavardhanam} is the better reading and it is well attested among the manuscripts (including \textalpha).
\end{philcomm}

%%%%%%%%%%
\subsection*{2.53}
\begin{translation}[hp02_053]
[and] cures diseases in the bodily constituents inside the network of channels,
the retention called \emph{ujjāyī} should be done when [the yogi] is moving or remaining still.
\end{translation}

\begin{sources}[hp02_053]
\emph{Gorakṣaśataka} 38

\begin{versinnote}
\tl{nāḍījalodarādhātugatadoṣavināśanam |\\+}
\tl{gacchatas tiṣṭhataḥ kāryam ujjāyyākhyaṃ ca kumbhakam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_053]
\emph{Haṭharatnāvalī} 2.15

\begin{versinnote}
\tl{nāḍījālodarādhātugatadoṣavināśanam |\\+}
\tl{nāḍījālodarādhātu° ] nāḍījalodaradhātu° \textup{J,P;} nāḍījalodaraṃ dhātu° \textup{N, n1, n4}\\+}
\tl{gacchatā tiṣṭhatā kāryam ujjāyyākhyaṃ hi kumbhakam || \\!}
\end{versinnote}

\emph{Yogalakṣaṇāvalī} f.~32r (attrib. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{dehānaloddīptikaraṃ jalodaravighātakṛt |\\+}
\tl{gachatā tiṣṭhatā kāryaś cojjāyākhyas tu kumbhakaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{nāḍījalodaradhātugatadoṣavināśanam |\\
gacchatas tiṣṭhataḥ kāryam ujjāyyākhyaṃ ca kumbhakam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.103 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote} 
\tl{nāḍījalodarādhātugatadoṣanivāraṇam |\\+}
\tl{gacchatā tiṣṭhatā kāryam ujjākhyaṃ kumbhakaṃ tv idam ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_053]
Nearly all the manuscripts have \emph{nāḍījalodara}°, which does not make sense in this context because \emph{jalodara} is the disease ascites and \emph{°gata°} requires a location. This problem is also present in the transmission of the source text for the verse, the \emph{Gorakṣaśataka}. A solution can be found in some of the manuscripts of the \emph{Haṭharatnāvalī}, which read \emph{nāḍījālodarā°} (‘in the network of channels and stomach’). The other problem is \emph{°darādhātu°}. In spite of Brahmānanda’s efforts to explain it as \emph{°dara}, \emph{ā}, and \emph{dhātu°}, the \emph{ā} before \emph{dhātu°} appears to have been inserted metri causa. We have adopted \emph{nāḍījālodare} (‘inside the network of channels’), which is close to \getsiglum{V3} and \getsiglum{J10} (\emph{nāḍījalodare}). It is likely that \emph{°jālodare} was changed to \emph{°jalodara} in the transmission because of confusion with the disease of a similar name.

For the idea of \emph{doṣa}s being in \emph{dhātu}s see \emph{Tantrāloka} 28.283cd, where worldly concepts are said to arise from it (\emph{dhātudoṣāc ca saṃsārasaṃskārās te ...}), but it is also the source of physical disorders (\emph{dhātudoṣakṛtaṃ mūrcchā} ĪPV on 2.15).

The name \emph{ujjāyī} may be a Prakrit form from \emph{uddhmāyī} from the verb \emph{ud-dhmā}, “to blow out”. We thank Diwakar Acharya for this suggestion.

\end{philcomm}

%%%%%%%%%%
\subsection*{2.54}
\begin{translation}[hp02_054]
Now sītkā---

{}[The yogi] should continuously make \emph{sīt} sound in the mouth and flare his nostrils. By practising in this way he becomes a second god of love.
\end{translation}

\begin{sources}[hp02_054]
Cf.~\emph{Kaulajñānanirṇaya} 14.54

\begin{versinnote}
\tl{cittan dadyāt tu vaktreṇa nāse dadyād vijṛmbhikā[m] |\\+}
\tl{vācāsiddhir bhavaty eva kāmadevo ’paraḥ priyaḥ ||\\!}
\end{versinnote}

Cf. \emph{Jñānasāra} 2.13

\begin{versinnote}
\tl{hikkā dadyāt sadā vaktre prāyaś caiva vijṛmbhikām |\\+}
\tl{evam abhyasyamānas tu kāmadevo dvitīyakaḥ ||\\!}
\end{versinnote}

\emph{Prāṇatoṣiṇī} (part 6) p. 851 (citing the \emph{Jñānasāra})

\begin{versinnote}
\tl{hikkāṃ dadyāt sadā vaktre ghrāṇañ caiva vijṛmbhate |\\+}
\tl{evam abhyāsayogena kāmadevo dvitīyakaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_054]
\emph{Haṭharatnāvalī} 2.16

\begin{versinnote}
\tl{sītkāṃ kuryāt tathā vaktre ghrāṇenaiva visarjayet |\\+}
\tl{evam abhyāsayogena kāmadevo dvitīyakaḥ || 2.16 ||\\!}
\end{versinnote}

\emph{Yogalakṣaṇāvalī} f.~32r (attrib. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{sītkāṃ dadyāt sadā vaktre ghrāṇe caiva vijṛṃbhitām |\\+}
\tl{evam abhyasato na kṣuttṛṭ cālasyādi jāyate ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{sītkāṃ kuryāt tathā vaktre ghrāṇenaiva visarjayet |\\+}
\tl{evam abhyāsayogena kāmadevo dvitīyakaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.104 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote} 
\tl{sītkāṃ dadyāt sadā vaktre ghrāṇe caiva vijṛmbhikām |\\+}
\tl{evam abhyāsayogena kāmadevo dvitīyakaḥ ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_054]
%The source texts and V1 have \emph{dadyāt} in the first \emph{pāda}, which seems to be the lectio difficilior. %JB: agrees with the Stemma

%Many of the old manuscripts have \emph{kumbhaṃ} instead of \emph{sītkāṃ}. The latter reading is supported by some manuscripts of the \emph{Haṭhapradīpikā}, the testimonia and the name of the \emph{kumbhaka}, which is stated in 2.44 and the heading of this verse. %JB this is not an issue with the new collation

There is division between \emph{śītkāṃ} and \emph{sītkāṃ} in all the manuscript groups of the \emph{Haṭhapradīpikā} (note that the likely reading in the source texts was \emph{hikkāṃ}). The result of becoming a second god of love may be connected with the sound \emph{sīt}, which is said to be made during sex in the \emph{Kāmasūtra} (2.7.4–19).

This verse’s source texts are from Kaula tantric milieus and this is reflected in the result of becoming one with the circle of yoginīs described in the next verse.

\end{philcomm}

%%%%%%%%%%
\subsection*{2.55}
\begin{translation}[hp02_055]
He joins the circle of yoginis and brings about creation and destruction. Neither hunger nor thirst [nor] sleep nor indolence arise [for him].
\end{translation}

\begin{sources}[hp02_055]
Cf.~\emph{Kaulajñānanirṇaya} 7.18ab

\begin{versinnote}
\tl{yoginīgaṇasāmānyā sṛṣṭisaṃhārakārakaḥ |\\!}
\end{versinnote}

\emph{Jñānasāra} 2.13cd–14ab

\begin{versinnote}
\tl{yoginīguṇasāmānyaḥ sṛṣṭisaṃhārakārakaḥ ||\\+}
\tl{na kṣudhā na ca tṛṇnidrā naiva murchā prajāyate |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_055]
\emph{Haṭharatnāvalī} 2.17

\begin{versinnote}
\tl{yoginīcakrasaṃsevyaḥ sṛṣṭisaṃhārakārakaḥ |\\+}
\tl{na kṣudhā na tṛṣā nidrā naivālasyaṃ prajāyate ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{yoginīcakrasaṃsevyaḥ sṛṣṭisaṃhārakārakaḥ |\\+}
\tl{na kṣudhā na tṛṣṇā nidrā tandrālasyaṃ na jāyate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.105 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{yoginīcakrasāmānyaḥ sṛṣṭisthityantakārakaḥ |\\+}
\tl{na kṣudhā na tṛṣā nidrā nālasya ca prajāyate ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_055]
%There are various possible readings for the first \emph{pāda}, namely, \emph{yoginīcakrasāmānyaḥ} (‘one who is one among the circle of yoginīs'), \emph{yoginīcakram āsādya} (‘having reached the circle of yoginis’) and \emph{yoginīcakrasaṃsevyaḥ} (‘one worshipped by the circle of yoginis’). The first of these is closer to the source texts and is a cliche in Kaula literature.
%JB I think the above note is informative but alpha one and three support the reading we have adopted, so maybe this note is not needed. But Alpha two omits this verse (should we note this omission?) JM: I don't think we need to.
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.56}
\begin{translation}[hp02_056]
His body is as he wishes, and he is free from all misfortune. By means of this technique, he truly becomes a lord of yogis in the world.
\end{translation}
% We should adopt bhuvimaṇḍale (alpha 3, eta 2) no?

\begin{sources}[hp02_056]
\emph{Jñānasāra} 2.14cd–15ab

\begin{versinnote}
\tl{bhavet svacchandadehas tu sarvopadravavarjitaḥ ||\\+}
\tl{anena vidhinā devi yogīndro bhūmimaṇḍale |\\!}
\end{versinnote}

\emph{Śivasaṃhitā} 3.94
\begin{versinnote}
\tl{anenaiva vidhānena yogīndro 'vanimaṇḍale |\\+}
\tl{bhavet svacchandacārī ca sarvāpatparivarjitaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_056]
\emph{Haṭharatnāvalī} 2.18

\begin{versinnote}
\tl{bhavet svacchandadehas tu sarvopadravavarjitaḥ |\\+}
\tl{anena vidhinā satyaṃ yogīndro bhāti bhūtale ||\\+}
\tl{°dehas tu ] °dehaḥ syāt P; °dehasyāt T,t1\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{bhavet svachandadehas tu sarvopadravavarjitaḥ |\\+}
\tl{anena vidhinā yas tu yogīndro bhūmimaṇḍale ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.106 (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{bhavet svacchandadehaś ca sarvopadravavarjitaḥ |\\+}
\tl{anena vidhinā satyaṃ yogīndro bhuvimaṇḍale ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_056]
The aiśa compound \emph{bhuvimaṇḍale}, which is attested at \emph{Mañjuśrīmūlakalpa} 45.221, is likely the original reading here. The word  \emph{bhuvi} as the first member of a compound is attested elsewhere. The alternative \emph{bhūmi}° is well-attested and so the change may have happened early in the transmission. 
\end{philcomm}

%%%%%%%%%%
%\subsection*{2.56*1}
%\begin{translation}[hp02_056_1]
%One who always takes in the breath through the aperture at the roots of the tongue, undoubtedly becomes a receptacle of all \emph{siddhi}s.
%\end{translation}
%\begin{philcomm}[hp02_056_1]
%This verse appears to be a derivative of 2.57, and was not original to the text. Cf. Ten-chapter \emph{Haṭhapradīpikā} 4.46
%\begin{versinnote}
%\tl{jihvāmūlena randhreṇa yaḥ prāṇaṃ satataṃ pibet |\\+}
%\tl{sa bhavet sarvasiddhānāṃ bhājanaṃ nātra saṃśayaḥ ||\\!}
%\end{versinnote}
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.57}
\begin{translation}[hp02_057]
And the very same has been taught [as follows]:

He who continuously takes in the breath through the tongue and the root of the palate has all his diseases cured in half a year.
\end{translation}

\begin{sources}[hp02_057]
\emph{Kauljñānanirṇaya} 6.19
\begin{versinnote}
\tl{rasanātālumūle tu kṛtvā vāyuṃ pibec chanaiḥ |\\+}
\tl{ṣaṇmāsād abhyased devi mahārogaiḥ pramucyate ||\\!}
\end{versinnote}

\emph{Vivekamārtaṇḍa} 120

\begin{versinnote}
\tl{rasanātālumūlena yaḥ prāṇam anilaṃ pibet |\\+}
\tl{abdārdhena bhavet tasya sarvarogaparikṣayaḥ ||\\!}
\end{versinnote}

\emph{Śivasaṃhitā} 3.80

\begin{versinnote}
\tl{rasanāṃ tālumūle yaḥ sthāpayitvā vipaścitaḥ | \\+}
\tl{pibet prāṇānilaṃ tasya rogāṇāṃ saṃkṣayo bhavet ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_057]
\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{rasanātāluyogena yaḥ prāṇaṃ satataṃ pibet |\\+}
\tl{abdārdhena bhavet tasya sarvarogaparikṣayaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.107 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{etad evoktaṃ gorakṣanāthena-\\+}
\tl{rasanātālumūlena yaḥ prāṇaṃ satataṃ pibete |\\+}
\tl{abdārdhena bhavet tasya sarvarogaparikṣayaḥ ||\\!}
\end{versinnote}

Cf. \emph{Ānandakanda} 1.20.137

\begin{versinnote}
\tl{jihvayā tālumūlena prāṇaṃ yaḥ pibati priye |\\+}
\tl{tasya ṣaṇmāsataḥ sarve rogā naśyanti yoginaḥ || \\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_057]
Verse 2.57 seems to be describing an alternative method of \emph{sītkā kumbhaka}. The introductory phrase \emph{evam eva uktaṃ ca} suggests that the teaching in this verse is consistent with what preceded it. However, the first version of \emph{sītkā} appears to be done at the front of the mouth (\emph{vaktra}), whereas the next version is done at the back of the mouth (see below).%JM Not sure what the note is saying. This practice could be consistent with 2.54 although no mention is made of a sound.

The compound \emph{rasanātālumūlena} is difficult to understand. In his \emph{Haṭhasaṅketacandrikā} (f.~79r–79v), Sundaradeva says that the external air strikes the root of the tongue and palate and the upper part of the uvula (\emph{atra muhū (mūhū codex) rasanātālumūlāhataṃ ghaṇṭikordhvabhāgāhataṃ bahiḥsthavāyuṃ vidhāya pibed ity arthaḥ}), which could make the sound \emph{sīt}. More helpful are the remarks of the commentator of the \emph{Yogataraṅgiṇī} (2.39). He says that a hole or cavity (\emph{vivara}) is made by the root of the palate with the help of the tongue. The yogi breathes through it (\emph{evaṃ rasanātālumūlena rasanā jihvā tatsahāyabhūtatālumūlena kṛtaṃ yad vivaraṃ, tena kṛtvā yaḥ yogī prāṇam anilaṃ prāṇavāyuṃ pibet pūrayet, tasya yogino ’bdārdhena ṣaṇmāsena sarvarogāṇāṃ nāśaḥ kṣayo bhavet} |).

The idea of breathing through a hole between the root of the palate and tongue might have been intended by the parallel reading of the \emph{Yogacintāmaṇi}: \emph{rasanā\-tālu\-yogena} (‘by joining the tongue and palate'). We have thus translated \emph{rasanā\-tālu\-mūlena} as ‘through the tongue and root of the palate’. It could also imply that the tongue is turned up and back to touch the root of the palate to make a hole that one breathes through (when the breath is taken in through the mouth). The \emph{Kumbhaka\-paddhati} (137ab) states this more clearly:

\begin{versinnote}
{}[The yogi] turns the tongue upwards and takes in the breath while making a \emph{sīt} sound.\\ 
\tl{rasanām unmukhīkṛtya sītkāraṃ kurvatā marut |\\!}%JM: could kumbhake tasmin (yasmin prob ok) refer to the practice as a whole, not the retention part of it? I.e. "In this kumbhaka" Otherwise it doesn't make sense.
\end{versinnote}

A similar practice is also described in \emph{Śivasaṃhitā} 3.80:

\begin{versinnote}
When the wise [yogi] places the tongue at the root of the palate and takes in the Prāṇa breath, his diseases are cured.
\end{versinnote}
\begin{versinnote}
\tl{rasanāṃ tālumūle yaḥ sthāpayitvā vipaścitaḥ |\\+}
\tl{pibet prāṇānilaṃ tasya rogāṇāṃ saṃkṣayo bhavet ||\\!}
\end{versinnote}

\end{philcomm}

%%%%%%%%%%
\subsection*{2.58}
\begin{translation}[hp02_058]
Now śītalī:\\
The wise man should draw in air through the tongue and after retaining the breath as before gradually exhale through the nostrils .
\end{translation}

\begin{sources}[hp02_058]
\emph{Gorakṣaśataka} 39cd–40ab

\begin{versinnote}
\tl{jihvayā vāyum ākṛṣya pūrvavat kuṃbhakād anu |\\+}
\tl{śanais tu ghrāṇarandhrābhyāṃ recayed anilaṃ sudhīḥ ||\\!}
\end{versinnote}

Cf. \emph{Vivekamārtaṇḍa} 139
\begin{versinnote}
\tl{kākacañcuvad āsyena śītalaṃ salilaṃ pibet |\\+}
\tl{prāṇaṃ prāṇavidhānajño yogī bhavati nirjaraḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_058]
\emph{Haṭharatnāvalī} 2.19
\begin{versinnote}
\tl{jihvayā vāyum ākṛṣya pūrvavat kumbhakād anu |\\+}
\tl{śanair aśītiparyantaṃ recayed anilaṃ sudhīḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102v (attr.~to the \emph{Yogabīja})
\begin{versinnote}
\tl{jihvayā vāyum ākṛṣya pūrvavat kumbhakād anu |\\+}
\tl{śanais tu ghrāṇārandhrābhyāṃ recayed anilaṃ suddhīḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.108 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{jihvayā vāyum ākṛṣya pūrvavat kumbhakād anu |\\+}
\tl{śanais tu ghrāṇarandhrābhyāṃ recayed anilaṃ sudhīḥ ||\\!}
\end{versinnote}

Cf. \emph{Ānandakanda} 1.20.135–136ab
\begin{versinnote}
\tl{kākacañcuvad āsyaṃ ca kṛtvā vāyuṃ sasūtkṛtam |\\+}
\tl{ādāya nāsārandhreṇa punas taṃ śvasanaṃ tyajet ||\\+}
\tl{śītalīkaraṇākhyo 'yaṃ yogas tu jvarapittahṛt |\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_058]
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.59}
\begin{translation}[hp02_059]
This retention called śītalī cures diseases such as swelling and enlargement of the spleen, fever, [excess] bile, hunger and thirst.
\end{translation}

\begin{sources}[hp02_059]
\emph{Gorakṣaśataka} 41

\begin{versinnote}
\tl{gulmaplīhādikā doṣāḥ kṣayaṃ yānti pittaṃ jvaraṃ |\\+}
\tl{viṣāṇi śītalī nāma kuṃbhako 'yaṃ nihanti ca ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_059]
\emph{Haṭharatnāvalī} 2.20

\begin{versinnote}
\tl{gulmaplīhodaraṃ doṣaṃ jvarapittakṣudhātṛṣāḥ |\\+}
\tl{viṣāṇi śītalī nāma kumbhako 'yaṃ nihanti ca ||\\+}
\tl{\var{°tṛṣāḥ ] °tṛṣā J,n1; °tṛṣaḥ T,t1}\\!}%
\end{versinnote}

\emph{Yogacintāmaṇi} ms.~L, f.~70r
% this verse is incomplete in N (gulmaplīhodaraṃ vāpi ||)

\begin{versinnote}
\tl{gulmaplīhodaraṃ cāpi vātapittaṃ kṣudhāṃ tṛṣām |\\+}
\tl{viṣāṇi śītalī nāma kumbhako vinihanti ca ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.109 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{gulmaplīhādikān doṣān jvaraṃ pittaṃ kṣudhāṃ tṛṣām |\\+}
\tl{anyāṃś ca śītalī nāma kumbhako 'yaṃ nihanti hi ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_059]
An antecedent to a cooling practice involving the tongue can be found in the \emph{Kaulajñānanirṇaya} (6.23–24), which mentions a point between the two front teeth that is cool to touch with the tongue:

\begin{versinnote}
There is a point located between the two ‘royal teeth’(\emph{rājadanta}). One should know this to be [the place of] nectar that destroys wrinkles and grey hair. Putting the tongue in the place cool to the touch, the wise man becomes free of wrinkles and grey hair and devoid of all diseases.
\tl{dvaurājadantamadhyasthaṃ bindurūpaṃ vyavasthitam |\\+}
\tl{amṛtaṃ taṃ vijānīyād valīpalitanāśanam ||\\+}
\tl{śītalasparśasaṃsthāne rasanāṃ kṛtvā tu buddhimān |\\+}
\tl{valīpalitanirmuktaḥ sarvavyādhivivarjitaḥ ||\\!}   
\end{versinnote}
We wish to thank Shaman Hatley for the reference and translation.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.60}
\begin{translation}[hp02_060]
Now bhastrikā:

If [the yogi] places the soles of both feet on the thighs, the lotus pose, which destroys all bad deeds, duly arises.
\end{translation}

\begin{sources}[hp02_060]
\emph{Gorakṣaśataka} 14

\begin{versinnote}
\tl{ūrvor upari ced dhatte ubhe pādatale tathā |\\+}
\tl{padmāsanaṃ bhavet samyak sarvapāpapraṇāśanam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_060]
%\emph{Haṭharatnāvalī} 2.21
%\begin{versinnote}
%\tl{atha bhastrikā---\\+}
%\tl{recakaḥ pūrakaś caiva kumbhakaḥ praṇavātmakaḥ |\\+}
%\tl{recako 'jasraniḥśvāsaḥ pūrakas tannirodhakaḥ |\\+}
%\tl{samānasaṃsthito yo 'sau kumbhakaḥ parikīrtitaḥ ||\\!}
%\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{bhastrikā\\+}
\tl{ūrvor upari saṃsthāpya ubhe pādatale tathā |\\+}
\tl{padmāsanaṃ bhavet samyak sarvapāpapraṇāśanam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.110 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{atha bhastrikā\\+}
\tl{ūrvor upari cādhatte ubhe pādatale tathā |\\+}
\tl{padmāsanaṃ bhavet samyak sarvapāpapraṇaśanam ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_060]
%The source text, the \emph{Gorakṣaśataka}, has \emph{ced} in the first pāda, and this has been dropped in nearly all of the available manuscripts of the \emph{Haṭhapradīpikā}, as well as the testimonia. It seems likely that \emph{ced dhatte} was the reading adopted by Svatmārāma because the \emph{cet} makes sense of the two finite verbs in the description. The first finite verb \emph{dhatte} is also supported by some of the old manuscripts, such as V1 and J10. At some stage, the verse was changed to read \emph{saṃsthāpya} to remove the awkward syntax posed by \emph{saṃdhatte} and \emph{vai dhatte}.
%[JB: alpha one and two supports ced, so there's no problem here]
%  JB add alpha two (ced dhatte) to the collation 
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.61}
\begin{translation}[hp02_061]
Having correctly adopted the lotus pose, with his neck and torso straight the wise man should close the mouth and forcefully exhale the breath through the nose
\end{translation}

\begin{sources}[hp02_061]
\emph{Gorakṣaśataka} 41cd–42ab

\begin{versinnote}
\tl{tataḥ padmāsanaṃ baddhvā samagrīvodaraḥ sudhīḥ |\\+}
\tl{mukhaṃ saṃyamya yatnena prāṇaṃ ghrāṇena recayet ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_061]
\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{samyak padmāsanaṃ badhvā samagrīvodaraḥ sudhīḥ |\\+}
\tl{mukhaṃ saṃyamya yatnena prāṇaṃ ghrāṇena recayet ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.111 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{samyak padmāsanaṃ baddhvā samagrīvodaraḥ śanaiḥ |\\+}
\tl{mukhaṃ saṃyamya yatnena prāṇaṃ ghrāṇena recayet ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_061]
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.62}
\begin{translation}[hp02_062]
in such a way that it comes into contact with the chest, throat and skull, making a sound. He should then quickly inhale a small amount of breath as far as the heart lotus.
\end{translation}

\begin{sources}[hp02_062]
\emph{Gorakṣaśataka} 42cd–43ab

\begin{versinnote}
\tl{yathā lagati kaṇṭhāt tu kapāle sasvanaṃ tataḥ |\\+}
\tl{vegena pūrayet kiṃ cit hṛtpadmāvadhi mārutam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_062]
\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{yathā lagati hṛtkaṇṭhe kapālāvadhi pūrayet |\\+}
\tl{vegena pūrayet samyag hṛtpadmāvadhi mārutam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.112 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{yathā lagati hṛtkaṇṭhakapāleṣu ca sasvanam |\\+}
\tl{vegena pūrayet kiñ cit hṛtpadmāvadhi mārutam ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_062]
%First hemistich is tricky. None of the old mss preserve \emph{sasvanaṃ}, which is in the GŚ and makes good sense. 
% MD: Now we have at least two good witnesses P11(β1),P15(ε1) with the reading sasvanaṃ.

Most witnesses have \emph{hṛtkaṇṭhe}, which we have understand as a dual accusative. One would expect \emph{kapāla} also to be in the accusative, but we have understood it as a locative sg. with \emph{sasvanam}.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.63}
\begin{translation}[hp02_063]
He should then exhale and inhale in that way over and over again. In the very same way as blacksmiths’ bellows are operated forcefully, [...]
\end{translation}

\begin{sources}[hp02_063]
\emph{Gorakṣaśataka} 43cd–44ab

\begin{versinnote}
\tl{punar virecayet tadvat pūrayec ca punaḥ punaḥ |\\+}
\tl{yathaiva lohakārāṇāṃ bhastrā vegena cālyate ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_063]
\emph{Haṭharatnāvalī} 2.22ab

\begin{versinnote}
\tl{yathaiva lohakārāṇāṃ bhastrī vegena cālyate |\\+}
\tl{\var{bhastrī ] bhastrā n4}\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{punar virecayet tadvat pūrayitvā punaḥ punaḥ |\\+}
\tl{yathaiva lohakārāṇāṃ bhastrā vegena cālyate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.113 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{punar virecayet tadvat pūrayec ca punaḥ punaḥ |\\+}
\tl{yathaiva lohakārāṇāṃ bhastrā vegena cālyate || 113 ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_063]
%The \getsiglum{V1} reading of \emph{lohakāreṇa} fits well with the passive verb, but it is the only witness to have this, and therefore appears to be an attempt to improve what was probably the original reading \emph{lohakārāṇāṃ} (as attested by the manuscripts, the source and testimonia).
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.64}
\begin{translation}[hp02_064]
[...] the wise man should move the breath in his body. When fatigue arises in the body he should inhale through the sun [channel]
\end{translation}

\begin{sources}[hp02_064]
\emph{Gorakṣaśataka} 44cd–45ab

\begin{versinnote}
\tl{tathaiva svaśarīrasthaṃ cālayet pavanaṃ sudhīḥ |\\+}
\tl{yadā śramo bhaved dehe tadā sūryeṇa pūrayet |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_064]
\emph{Haṭharatnāvalī} 2.22cd–23ab

\begin{versinnote}
\tl{tathaiva svaśarīrasthaṃ cālayet pavanaṃ sudhīḥ ||\\+}
\tl{yathā śramo bhaved dehe tathā sūryeṇa pūrayet |\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{tathaiva svaśarīrasthaś cālyate pavano dhiyā |\\+}
\tl{yathā śramo bhaved dehe tathā vegena pūrayet ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.114 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{tathaiva svaśarīrasthaṃ cālayet pavanaṃ dhiyā |\\+}
\tl{yadā śramo bhaved dehe tadā sūryeṇa recayet ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_064]
Most of the witnesses (including \textalpha) have \emph{dhiyā} at the end of the second verse quarter, but the manuscripts of the source text and the \emph{Haṭharatnāvalī} have \emph{sudhīḥ}. Since the subject of the simile is \emph{bhastrā}, one would expect the subject of \emph{cālayet}, which must be different, to be stated (as is the case with \emph{sudhīḥ}). %Also, one would expect the instrumental of \emph{dhī} to be qualified by some adjective, such as in the case of \emph{sattvāsthayā dhiyā} (\emph{Gorakṣaśataka} 74b) and \emph{sāttvikayā dhiyā} (\emph{Haṭhapradīpikā} 2.6b).
%  JB could the alphaTwo reading dhiyaḥ (as some sort of aiśa masc. sing.) explain dhiyā?
%JM: dhiyā on its own is fine. I find lots of refs. I've removed that part of the note. 
\end{philcomm}

%%%%%%%%%%
\subsection*{2.65}
\begin{translation}[hp02_065]
in such a way that the abdomen is filled by the breath, and hold the nose quickly [and] firmly without using the middle and index fingers.
\end{translation}

\begin{sources}[hp02_065]
\emph{Gorakṣaśataka} 45cd–46ab

\begin{versinnote}
\tl{yathodaraṃ bhavet pūrṇaṃ pavanena tathā laghu |\\+}
\tl{dhārayan nāsikā madhyātarjanībhyāṃ vinā dṛḍhaṃ |\\!}
%\tl{madhyā° ] em.; °madhyaṃ GU, °madhye T\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_065]
\emph{Haṭharatnāvalī} 2.23cd–24ab

\begin{versinnote}
\tl{yathodaraṃ bhavet pūrṇaṃ pavanena tathā laghu ||\\+}
\tl{dhārayen nāsikāṃ madhyātarjanībhyāṃ vinā dṛḍham |\\+}
\tl{\var{23c madhyā ] madhye N,n1,n3,n4,J,T,t1}\\!}%
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{yathodaraṃ bhavet pūrṇaṃ pavanena tathā laghu |\\+}
\tl{dhārayen nāsikāṃ madhyatarjanībhyāṃ vinā dṛḍham ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.115 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{yathodaraṃ bhavet pūrṇaṃ pavanena tathā laghu |\\+}
\tl{dhārayen nāsikāmadhye tarjanībhyāṃ vinā dṛḍham ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_065]
Only two of the collated witnesses (\getsiglum{J7}, \getsiglum{V15}) have \emph{madhyātarjanībhyāṃ} (`with the middle and index fingers') and this reading is not well attested by the manuscripts of the source text and testimonia. To hold the nose without the middle and index fingers is consistent with the way alternate nostril breathing is done in modern yoga (e.g., Iyengar 1991: 443–444). However, the reading of many manuscripts suggests  that the nose was held by all the fingers of both hands, except the index fingers (\emph{nāsikāmadhye tarjanībhyāṃ vinā}), which seems highly impracticable, or that the nose was held by only the index fingers of both hands (\emph{nāsikāmadhye tarjanībhyāṃ tathā}). It is likely that scribes changed \emph{madhyātarjanībhyāṃ} to \emph{madhye tarjanībhyāṃ} or \emph{madhyaṃ tarjanībhyāṃ} because of the \emph{pāda} break. 
\end{philcomm}
%Iyengar BKS. Light on Yoga: Yoga Dipika. Great Britain: Mandala, 1991.

%%%%%%%%%%
\subsection*{2.66}
\begin{translation}[hp02_066]
[The yogi] should hold the breath as before then exhale through Iḍā. [Because] it removes [imbalances] in wind, bile and phlegm, increases the body’s fire, [\dots]

\end{translation}

\begin{sources}[hp02_066]
\emph{Gorakṣaśataka} 46cd–47ab

\begin{versinnote}
\tl{kumbhakaṃ pūrvavat kṛtvā recayed iḍayānilam ||\\+}
\tl{kaṇṭhotthitānalaharaṃ śarīrāgnivivardhanam |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_066]
\emph{Haṭharatnāvalī} 2.24cd–25ab

\begin{versinnote}
\tl{kumbhakaṃ pūrvavat kṛtvā recayed iḍayānilam ||\\+}
\tl{vātapittaśleṣmaharaṃ śarīrāgnivivardhanam |\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{kumbhakaṃ pūrvavat kṛtvā recayed iḍayā tataḥ |\\+}
\tl{vātapittaśleṣmaharaṃ śarīrāgnivivardhanam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.116 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{kumbhakaṃ pūrvavat kṛtvā recayed iḍayā śanaiḥ |\\+}
\tl{vātapittaśleṣmaharaṃ śarīrāgnivivardhanam ||\\!}
\end{versinnote}

\end{testimonia}

%\begin{philcomm}[hp02_066]
%\end{philcomm}

\begin{metre}[hp02_066]
Anuṣṭubh (c: bha-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.67}
\begin{translation}[hp02_067]
it is an auspicious thunderbolt that awakens \emph{kuṇḍalinī}, destroys bad deeds, bestows happiness, and destroys the blockage of phlegm, etc., situated at the mouth of the central channel,[\dots]
\end{translation}

\begin{sources}[hp02_067]
\emph{Gorakṣaśataka} 47cd–48ab

\begin{versinnote}
\tl{kuṇḍalībodhakaṃ vajraṃ pāpaghnaṃ śubhadaṃ sukham |\\+}
\tl{brahmanāḍīmukhāntaḥsthakaphādyargalanāśanam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_067]
\emph{Haṭharatnāvalī} 2.25cd

\begin{versinnote}
\tl{brahmanāḍīmukhe saṃsthakaphādyargalanāśanam |\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{kuṇḍalībodhanaṃ kuryāt pāpaghnaṃ sukhadaṃ śubham |\\+}
\tl{brahmanāḍīmukhe saṃsthaṃ kapāṭārgalanāśanam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.117 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{kuṇḍalībodhanaṃ sarvadoṣaghnaṃ sukhadaṃ śubham |\\+}
\tl{brahmanāḍīmukhāntasthakaphādyargalanāśanam ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_067]
The reading \emph{vajraṃ} in the first verse quarter is an emendation based on the manuscripts of the \emph{Gorakṣaśataka}, the source text for this verse. Some manuscripts of the \emph{Haṭhapradīpikā} have readings close to \emph{vajraṃ}, such as \emph{vipra} (\getsiglum{V1}, \getsiglum{P15}) and \emph{vakra} (G11), which suggest that \emph{vajraṃ} was changed at an early stage of the transmission. The \textalpha\ group have \emph{cakraṃ}, which could be understood as a weapon (i.e., the discus often associated with Viṣṇu), but such a meaning would be rather unusual in yoga texts, where the term \emph{cakra} is used so frequently in contexts of the yogic body.
%The J10 group has rewritten this verse in the masc sg. with \emph{kumbhaḥ}. And this is possible without the \emph{viśeṣeṇaiva [...] kumbhakaṃ tv idaṃ} line. However, it seems that the \emph{viśeṣeṇaiva} line has dropped out because it is in the Gorakṣaśataka. Therefore, the neuter was probably original.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.68}
\begin{translation}[hp02_068]
[and] pierces the three knots that have arisen from the three \emph{guṇa}s, it is particularly important to perform this retention called ‘the bellows’.
\end{translation}
% Adopt \item[guṇatrayasamudbhūta] N21, Bo1. Tr reflects this but ed has samyaggātra

\begin{sources}[hp02_068]
\emph{Gorakṣaśataka} 48cd–49ab

\begin{versinnote}
\tl{guṇatrayasamudbhūtagranthitrayavibhedakam |\\+}
\tl{viśeṣeṇaiva kartavyaṃ bhastrākhyaṃ kuṃbhakaṃ tv idam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_068]
\emph{Haṭharatnāvalī} 2.25cd

\begin{versinnote}
\tl{viśeṣenaiva kartavyaṃ bhastrākhyaṃ kumbhakaṃ tv idam ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~102r–102v (attr.~to the \emph{Yogabīja})

\begin{versinnote}
\tl{samyaggātrasamudbhūtagranthitrayavibhedanam |\\+}
\tl{viśeṣeṇaiva kartavyaṃ bhastrākhyaṃ kumbhakaṃ tv idam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.118 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{samyaggātrasamudbhūtagranthitrayavibhedanam |\\+}
\tl{viśeṣeṇaiva karttavyaṃ bhastrākhyaṃ kumbhakaṃ tv idam ||\\!}
\end{versinnote}

\emph{Haṭhasaṅketacandrikā} f.~80v (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{guṇatrayasamudbhūtagranthitrayavibhedakam |\\+}
\tl{viśeṣeṇaiva karttavyaṃ bhastrākhyaṃ kumbhakaṃ svayam ||\\!}
\end{versinnote}

\end{testimonia}

\begin{philcomm}[hp02_068]
In the first verse quarter, nearly all of the manuscripts, including the \textalpha\ group, have \emph{°gātrasamudbhūta°} (`arisen in the limbs/body'), which is rather meaningless here since the three knots are in the central channel. It is probable that \emph{samyaggātra°} is a corruption of \emph{guṇatraya°}, which is attested by the manuscripts of the source text, the \emph{Gorakṣaśataka}, two of the available manuscripts of the \emph{Haṭhapradīpikā} (\getsiglum{N21}, Bo1) and the \emph{Haṭhasaṅketacandrikā}. In the \emph{Gorakṣaśataka}, it is stated clearly that each of the knots arise from one of the three \emph{guṇa}s: \emph{brahmagranthi} from \emph{rajas} (78cd), \emph{viṣṇugranthi} from \emph{sattva} (79cd) and \emph{rudragranthi} from \emph{tamas} (80cd). When the verses on \emph{bhastrā} were extracted from the \emph{Gorakṣaśataka} without the context of the knots and \emph{guṇa}s, the meaning of \emph{guṇa\-traya\-sam\-udbhūta\-granthi\-traya°} (`the three knots that have arisen from the three \emph{guṇa}s') was lost early in the transmission of the \emph{Haṭhapradīpikā}.  
\end{philcomm}

%%%%%%%%%%
\subsection*{2.69}
\begin{translation}[hp02_069]
Now bhrāmarī:

Forcibly loud inhalation with the sound of a male bee; very slow exhalation with the sound of a female bee: as a result of practising thus, there arises in the minds of the best yogis an extraordinary blissful playfulness.
\end{translation}

\begin{testimonia}[hp02_069]
\emph{Haṭharatnāvalī} 2.26

\begin{versinnote}
\tl{atha bhrāmarī---\\+}
\tl{vegodghoṣaṃ pūrakaṃ bhṛṅganādaṃ \\+}
\tl{bhṛṅgīnādaṃ recakaṃ mandamandam |\\+}
\tl{yogīndrāṇāṃ nityam abhyāsayogāc \\+}
\tl{citte jātā kā cid ānandalīlā || \\+}
\tl{līlā ]°mūrcchā N,n1,n2,n3,n4,J,T,t1\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{bhramarī—\\+}
\tl{vegodghoṣaṃ pūrakaṃ bhṛṅganādaṃ \\+}
\tl{bhṛṅgīnādaṃ recakaṃ mandamandam |\\+}
\tl{yogīndrāṇāṃ nityam abhyāsayogāc \\+}
\tl{citte jātā kācid ānandalīlā ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.119 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{atha bhrāmarī---\\+}
\tl{vegodghoṣaṃ pūrakaṃ bhṛṃganādaṃ \\+}
\tl{recakaṃ mandamandam |\\+}
\tl{yogīndrāṇāmevamabhyāsayogāc\\+}
\tl{citte jātā kācid ānandalīlā || \\!}
\end{versinnote}

\emph{Yogaprakāśikā} 4.59 (ten-chapter \emph{Haṭhapradīpikā})

\begin{versinnote}
bhrāmarīkumbhakaṃ lakṣayaty atheti | vegena sañjāta udghoṣo yasmin pūrake taṃ bhṛṅganādatulyaṃ bhṛṅgīnādatulyaṃ recakaṃ kuryād ānandalīleti |
\end{versinnote}

\emph{Haṭhasaṅketacandrikā} f.~80v

\begin{versinnote}
\tl{vegākṛṣṭiṃ pūrakaṃ bhṛṅganādaṃ \\+}
\tl{bhaṅgīnādaṃ recakaṃ mandaṃ mandaṃ |\\+}
\tl{yogīdrāṇām evam abhyāsayogac \\+}
\tl{cite jātā kācid ānandamūrchā ||\\+}
\tl{vegodghoṣam iti vā pāṭhaḥ |\\!}
\end{versinnote}

Cf. \emph{Kumbhakapaddhati} 169

\begin{versinnote}
\tl{aliśabdayutaṃ vegāt pūrayet kumbhayet tataḥ |\\+}
\tl{sāliśabdāc chanai rekāt bhrāmarīkumbhako muneḥ ||\\+}
\tl{ānandalīlāṃ kurute bhrāmarīkumbhako muneḥ || 169 ||\\!}
\end{versinnote}

Cf. \emph{Gheraṇḍasaṃhitā} 7.10--11

\begin{versinnote}
\tl{anilaṃ mandavegena bhrāmarīkumbhakaṃ caret |\\+}
\tl{mandaṃ mandaṃ recayed vāyuṃ bhṛṅganādaṃ tato bhavet || 7.10 ||\\+}
\tl{antaḥsthaṃ bhramarīnādaṃ śrutvā tatra mano nayet |\\+}
\tl{samādhir jāyate tatra ānandaḥ so 'ham ity ataḥ || 7.11 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_069]
In the first two \emph{pāda}s we are understanding the usually masculine \emph{pūraka} and \emph{recaka} to be neuter nominatives. When explaining this verse, Brahmānanda (\emph{Jyotsnā} 2.68) supplies the gerund \emph{kṛtvā} with \emph{pūrakaṃ} and \emph{kuryāt} with \emph{recakaṃ}, but it is unlikely that the author intended this.

The compound \emph{vegodghoṣaṃ} (close to \emph{vegoghoṣaṃ} in \alphaOne) is rather unusual. We have understood it according to the commentator Bālakṛṣṇa’s gloss: `an inhalation in which sound is produced by force' (\emph{vegena sañjāta udghoṣo yasmin pūrake}).

Many manuscripts of the \emph{Haṭhapradīpikā}, including the \textalpha\ group, have \emph{ānandamūrcchā} at the end of the verse, instead of \emph{ānandalīlā}, which we have adopted and which is supported by \betaTwo\ and the \textgamma\ group, as well as the most important testimonia (i.e.~the \emph{Haṭharatnāvalī}, \emph{Yogacintāmaṇi} and \emph{Yuktabhavadeva}). The reading \emph{ānanda\-mūrcchā} is probably the result of dittography as the heading \emph{atha mūrcchā} follows this verse. It has persisted in the transmission because  \emph{ānandamūrcchā} makes some sense by itself (`swooning through bliss'), although when it is read with the rest of the verse quarter, the meaning is somewhat odd (`swooning through bliss arises in the mind').
\end{philcomm}

\begin{metre}[hp02_069]
Śālinī 
\end{metre}

%%%%%%%%%%
\subsection*{2.70}
\begin{translation}[hp02_070]
Now mūrcchā:

At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This, which is called \emph{mūrcchā}, bestows the bliss of  fainting.
\end{translation}
% adopt feminine in 2nd hemistich

\begin{sources}[hp02_070]
\emph{Haṭharatnāvalī} 2.27

\begin{versinnote}
\tl{atha mūrcchā---\\+}
\tl{pūrakānte gāḍhataraṃ baddhva jālandharaṃ śanaiḥ |\\+}
\tl{recayen mūrcchanākhyo 'yaṃ manomūrcchāsukhapradā ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~101v (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{mūrchā—\\+}
\tl{pūrakānte gāḍhataraṃ bandho jālandharaḥ śanaiḥ |\\+}
\tl{recayen mūrchanākhyo 'yaṃ manomūrchāsukhapradā ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.120 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{atha mūrcchā---\\+}
\tl{pūrakānte gāḍhataraṃ baddhvā jālandharaṃ śanaiḥ ||\\+}
\tl{recayen mūrcchanākhyeyaṃ manomūrcchāsukhapradā || 120 ||\\!}
\end{versinnote}

\emph{Kumbhakapaddhati} 170

\begin{versinnote}
\tl{āpūrya kumbhitaṃ prāṇaṃ badhvā jālandharaṃ śanaiḥ |\\+}
\tl{recayen mūrcchanākumbho manomūrcchāsukhapradā ||\\!}
\end{versinnote}
\end{sources}

\begin{philcomm}[hp02_070]
The Jālandhara lock is explained at 3.67–70. 
\end{philcomm}

\begin{metre}[hp02_070]
Anuṣṭubh (a: bha-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.71}
\begin{translation}[hp02_071]
Now Plāvanī:\\
With his abdomen completely filled with the wind of eructation, which has been turned inwards, [the yogi] floats easily even in deep water, like a lotus leaf.
\end{translation}

\begin{testimonia}[hp02_071]
%{[Not in \emph{Yogacintāmaṇi}, \emph{Haṭharatnāvalī}]}

\emph{Yuktabhavadeva} 7.121 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{antaḥpravarttitādhāramārutāpūritodaraḥ |\\+}
\tl{payasy agādhe 'pi sukhāt plavate padmapatravat ||\\+}
\tl{ayam eva plāvinī kumbhako'pi |\\!}
\end{versinnote}

Cf.~\emph{Kumbhakapaddhati} 171

\begin{versinnote}
\tl{yatheṣṭaṃ pūrayed vāyuṃ baddhe jālandhare dṛḍhe |\\+}
\tl{hṛdi dhṛtvā jale suptvā plāvinīkumbhako bhavet || 171 ||\\!}
\end{versinnote}

\emph{Yogaprakāśikā} 4.61 (Ten-chapter \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{plāvanīkumbhakaṃ lakṣayati antar iti |\\+}
\tl{antaḥsañcāritenāpānavāyunā pūritam udaraṃ yasyeti vigrahaḥ |\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_071]
The heading \emph{atha plāvanī} is not found in \alphaOne \ (which follows the description with \emph{plāvanīkumbhakaṃ}), but we have adopted it to be consistent with the other \emph{kumbhaka} descriptions.

We have understood \emph{udgāramāruta} to refer to the breath of eructation, i.e.~the nāga breath as described in e.g.~\emph{Vivekamārtaṇḍa} 36.
% We have adopted \emph{payasy agādhe pi sukham} in pāda c because \emph{sukham} gives a better meaning and it is a bha-vipulā

\getsiglum{V19}, \getsiglum{P28}, the \emph{Yogacintāmaṇi} and the \emph{Haṭharatnāvalī} omit this verse and accordingly do not mention \emph{plāvinī} in verse 2.44, substituting it with \emph{kevala}.

\end{philcomm}

\begin{metre}[hp02_071]
Anuṣṭubh (c: bha-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.72}
\begin{translation}[hp02_072]
Now \emph{kevalakumbhaka}:

Breath control is said to be threefold, with exhalation, inhalation, and retention. Retention is considered to be twofold: \emph{sahita} and \emph{kevala}.

\end{translation}

\begin{sources}[hp02_072]
\emph{Gorakṣaśataka} 29

\begin{versinnote}
\tl{prāṇaś ca dehajo vāyur āyāmaḥ kumbhakaḥ smṛtaḥ |\\+}
\tl{sa eva dvividhaḥ proktaḥ sahitaḥ kevalas tathā ||\\!}
\end{versinnote}

\emph{Vasiṣṭhasaṃhitā} 3.2cd

\begin{versinnote}
\tl{prāṇāyāmas tribhiḥ prokto recapūrakakumbhakaiḥ || 2 ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_072]
\emph{Yogalakṣaṇāvalī} f.~32r (attrib. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{prāṇāyāmas tridhā prokto recapūrakakumbhakaiḥ |\\+}
\tl{bahir virocanaṃ vāyor udarād recakaḥ smṛtaḥ ||\\+}
\tl{bāhyād āpūraṇaṃ vāyor udare pūrakaḥ smṛtaḥ |\\+}
\tl{saṃpūrṇakuṃbhavad vāyor dhāraṇaṃ kumbhako bhavet |\\+}
\tl{sahitaḥ kevalaś ceti kumbhako dvividho mataḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.122 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{atha kevalaḥ\\+}
\tl{prāṇāyāmas tridhā prokto recapūrakakumbhakaiḥ |\\+}
\tl{sahitaḥ kevalaś ceti kumbhako dvividho mataḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_072]
The import of the name \emph{sahita}, “accompanied”, is that \emph{kumbhaka} is accompanied by inhalation and exhalation, and of \emph{kevala}, “isolated”, that it is not.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.73}
\begin{translation}[hp02_073]
The [breath retention] that [the yogi] performs with exhalation and inhalation is \emph{sahita}. He should practice \emph{sahita} until \emph{kevala} is perfected.
\end{translation}

\begin{sources}[hp02_073]
\emph{Vasiṣṭhasaṃhitā} 3.28

\begin{versinnote}
\tl{virecyāpūrya yaṃ kuryāt sa vai sahitakumbhakaḥ \\+}
\tl{sahitaṃ kevalaṃ cātha kumbhakaṃ nityam abhyaset ||\\+}
\tl{yāvat kevalasiddhiḥ syāt tāvat sahitam abhyaset |\\+}
\tl{\var{28a virecyāpūrya ] recyavāpūrya yat \emph{la}, ārecyāpūrya yaḥ \emph{śa}}\\!}
\end{versinnote}

\emph{Yogayājñavalkya} 6.31cd–32

\begin{versinnote}
\tl{recya cāpūrya yaḥ kuryāt sa vai sahitakumbhakaḥ ||\\+}
\tl{sahitaṃ kevalaṃ cātha kumbhakaṃ nityam abhyaset |\\+}
\tl{yāvat kevalasiddhiḥ syāt tāvat sahitam abhyaset ||\\!}
\end{versinnote}

Cf.~\emph{Dattātreyayogaśāstra} 66ab

\begin{versinnote}
\tl{sahito recapūrābhyāṃ tasmāt sahitakumbhakaḥ |\\!}
\end{versinnote}

Cf. \emph{Gorakṣaśataka} 30ab

\begin{versinnote}
\tl{yāvat kevalasiddhiḥ syāt tāvat sahitam abhyaset |\\!}
\end{versinnote}

\end{sources}

\begin{testimonia}[hp02_073]
\emph{Yogacintāmaṇi} f.~96v (attr.~to Yājnavalkya)
\begin{versinnote}
\tl{sahitaṃ kevalaṃ vātha kuṃbhakaṃ nityam abhyaset |\\+}
\tl{yāvat kevalasiddhis syāt tāvat sahitam abhyaset ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.123 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{recya vā pūrakaḥ kāryaḥ śanaiḥ sahitakumbhakaḥ |\\+}
\tl{yāvat kevalasiddhiḥ syāt sahitaṃ tāvad abhyaset ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_073]
We have adopted the \emph{Vasiṣṭhasaṃhitā}’s reading, which is very close to that of \deltaOne \ (\emph{ārecyāpūrya yat kuryāt}) and \deltaThree \ (\emph{ārecyāpūrya yaḥ kuryāt}) and similar to \alphaTwo\ (\emph{recapūrya y\emph{[}a\emph{]}t kāryaḥ}). It is the only one that makes sense of 2.73ab. It appears that the relative pronoun dropped out of the first verse quarter early in the transmission of the \emph{Haṭhapradīpikā} and scribes have tried in various unsuccessful ways to restore some sense.

Cf.~Marcinowska-Rosól \& \ Sellmer 2021, p.\,102f.

% MD: Perhaps \emph{recakaḥ pūrakaḥ kāryaḥ} which is closer to the readings of N3 and V3?  Cf. 2.74a \emph{recakaṃ pūrakaṃ muktvā}.
% JB: We need the relative pronoun to pick up on the sa in the second pāda
% MD: Not always. Cf. 1.58, 2.24, 2.26.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.74}
\begin{translation}[hp02_074]
Holding the breath comfortably without exhalation and inhalation is \emph{kevalakumbhaka}. This is said to be [the true] breath control. 
\end{translation}

\begin{sources}[hp02_074]
\emph{Vasiṣṭhasaṃhitā} 3.27

\begin{versinnote}
\tl{recanaṃ pūraṇaṃ muktvā sukhaṃ yad vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}

\emph{Yogayājñavalkya} 6.30cd–6.31ab

\begin{versinnote}
\tl{recakaṃ pūrakaṃ muktvā sukhaṃ yad vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_074]
\emph{Haṭharatnāvalī} 2.28

\begin{versinnote}
\tl{atha kevalaḥ---\\+}
\tl{recakaṃ pūrakaṃ muktvā sukhaṃ yad vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~94v (attr.~to `\emph{tajjñāḥ}')

\begin{versinnote}
\tl{recakaṃ pūrakaṃ muktvā yat sukhaṃ vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.124 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{recakaṃ pūrakaṃ muktvā yad vāyudhāraṇam |\\+}
\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakumbhakaḥ ||\\!}
\end{versinnote}

%Yogadīpikā 77cd--78ab

%\begin{versinnote}
%\tl{recakaṃ pūrakaṃ muktvā susukhaṃ vāyudhāraṇaṃ\\+}
%\tl{prāṇāyāmo 'yam ity uktaḥ sa vai kevalakuṃbhakaḥ\\!}
%\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_074]
%The relative pronoun is omitted in V1 and the J10 group, which have \emph{vāyunirodhanam}. But the relative is needed to connect the description of kevalakumbhaka in the first hemistich to the last pāda (\emph{sa vai kevalakumbhakaḥ}).
On this verse, see Marcinkowska-Rosół and Sellmer 2020: 102–105.
\end{philcomm}
%Maria Marcinkowska-Rosół & Sven Sellmer. “Notes on some difficult passages of the Haṭhapradīpikā”. Zeitschrift der Deutschen Morgenländischen Gesellschaft 171 (2021), pp. 101–21.

%%%%%%%%%%
\subsection*{2.75}
\begin{translation}[hp02_075]
When breath retention is mastered on its own, without exhalation and inhalation, nothing in the three worlds  is impossible for [the yogi].
\end{translation}

\begin{sources}[hp02_075]
\emph{Dattātreyayogaśāstra} 74

\begin{versinnote}
\tl{kevale kumbhake siddhe recapūrakavarjite |\\+}
\tl{na tasya durlabhaṃ kiṃ cit triṣu lokeṣu vidyate ||\\!}
\end{versinnote}

\emph{Vasiṣṭhasaṃhitā} 3.30

\begin{versinnote}
\tl{kevale kumbhake siddhe recapūraṇavarjite |\\+}
\tl{na tasya durlabhaṃ kiṃ cit triṣu lokeṣu vidyate ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_075]
\emph{Haṭharatnāvalī} 2.29

\begin{versinnote}
\tl{kevale kumbhake siddhe recapūrakavarjite |\\+}
\tl{na tasya durlabhaṃ kiñ cit triṣu lokeṣu vidyate ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~97r (attr.~to Yājñavalkya)

\begin{versinnote}
\tl{kevale kumbhake siddhe recapūraṇavarjite |\\+}
\tl{na tasya durlabhaṃ kiṃ cit triṣu lokeṣu vidyate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.125 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{kevale kumbhake siddhe recapūrakavarjite |\\+}
\tl{na tasya durlabhaṃ kiñcit triṣu lokeṣuṃ vidyate ||\\!}
\end{versinnote}

\end{testimonia}

%%%%%%%%%%
\subsection*{2.76}
\begin{translation}[hp02_076]
He who is capable of \emph{kevalakumbhaka} undoubtedly attains [the ability to] hold the breath as long as he wants and the state of Rājayoga.
\end{translation}

\begin{testimonia}[hp02_076]
\emph{Haṭharatnāvalī} 2.30

\begin{versinnote}
\tl{śaktaḥ kevalakumbhena yatheṣṭaṃ vāyudhāraṇam |\\+}
\tl{etādṛśo rājayogo kathito nātra saṃśayaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.126 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{śaktaḥ kevalakumbhena yatheṣṭaṃ vāyudhāraṇam |\\+}
\tl{rājayogapadaṃ samyak labhate nātra saṃśayaḥ ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 44.59

\begin{versinnote}
\tl{haṭhapradīpikāyām–\\+}
\tl{śaktaḥ kevalakumbhena yatheṣṭaṃ vāyudhāraṇe |\\+}
\tl{rājayogapadaṃ caiva labhate nātra saṃśayaḥ || 59 ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_076]
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.77}
\begin{translation}[hp02_077]
Rājayoga does not succeed without Haṭha nor Haṭha without Rājayoga so one should practise them both together until [the] niṣpatti [stage].
\end{translation}

\begin{sources}[hp02_077]
\emph{Śivasaṃhitā} 5.222

\begin{versinnote}
\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
\tl{na sidhyati tato yugmam āniṣpatteḥ samabhyaset\\+}
\tl{\textup{[middle hemistich not in mss. I, III, IV, VII, IX, X, XII, XIV--XVI]}\\+}
\tl{tasmāt pravartate yogī haṭhe sadgurumārgataḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp02_077]
\emph{Haṭharatnāvalī} 1.19

\begin{versinnote}
\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
\tl{vyāptiḥ syād avinābhūtā śrīrājahaṭhayogayoḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~21r (attr.~to the \emph{Haṭhapradīpika})

\begin{versinnote}
\tl{haṭhapradīpikāyām\\+}
\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
\tl{na sidhyati tato yugmaṃ manīṣy etau samabhyaset |\\+}
\tl{haṭhaṃ vinā rājayogaṃ rājayogaṃ vinā haṭham |\\+}
\tl{ye vai caranti tān manye prayāsaphalavarjitān iti ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.127 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
\tl{na sidhyati tato yugmam āniṣpatteḥ samācaret ||\\!}
\end{versinnote}

%Śivayogadarpana
%\begin{versinnote}
%\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
%\tl{na sidhyati tato yugmam manīṣī tat samabhyaset || 6 ||\\!}
%\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_077]
On the niṣpatti stage see 4.23–25.
\end{philcomm}

\begin{metre}[hp02_077]
Anuṣṭubh (a: ra-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{2.78}
\begin{translation}[hp02_078]
At the end of exhaling the retained breath, [the yogi] should make the mind supportless. By practising in this way he reaches the state of Rājayoga .
\end{translation}

\begin{testimonia}[hp02_078]
\emph{Yuktabhavadeva} 7.128 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{kumbhitaḥ prāṇarecānte kuryyāc cittaṃ nirāmayam ||\\+}
\tl{evamabhyāsayogena rājayogapadaṃ vrajet || 128 ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 44.60 (attr.~to the \emph{Haṭhapradīpika})

\begin{versinnote}
\tl{kumbhitaḥ prāṇarecānte kuryyāc cittaṃ nirāśrayam |\\+}
\tl{evamabhyāsayogena rājayogaṃ labhet punaḥ || 60 ||\\+}
\tl{nirāśrayaṃ saṃkalparahitam ||\\!}
\end{versinnote}

\emph{Yogaprakāśikā} 4.67 (Ten-chapter \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{tad eva visadayati kumbhakam iti || kevalakumbhakābhyāsena cittaṃ dagdhaparṇavat nirvāsanaṃ bhavatītyarthaḥ || anyad vyākhyātam || 67 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp02_078]
The reading \emph{kumbhitaḥ}, which is attested by \alphaOne\ and the testimonia, does not make sense here because the subject must be the yogi (\emph{prānaḥ} would be unmetrical). In other texts, the word \emph{kumbhita} is used to qualify the breath and means `retained' (e.g., \emph{Yogabīja} 94, \emph{Kumbhakapaddhati} 127, \emph{Yuktabhavadeva} 8.32, etc.).
\end{philcomm}

%%%%%%%%%%
\subsection*{2.79}
\begin{translation}[hp02_079]
As a result of retaining the breath, kuṇḍalinī awakens; as a result of the awakening of kuṇḍalinī, suṣumṇā becomes free of blockages and success in Haṭha arises.
\end{translation}

\begin{testimonia}[hp02_079]
\emph{Yogacintāmaṇi} f.~97a (attr.~to the \emph{Haṭhayoga})

\begin{versinnote}
\tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet |\\+}
\tl{anargalaḥ suṣumṇānto haṭhasiddhiś ca jāyate ||\\+}
\tl{anargalaḥ ] L, antargataḥ N\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.129 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet |\\+}
\tl{anargalā suṣumnā ca haṭhasiddhiśca jāyate ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 44.61

\begin{versinnote}
\tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet |\\+}
\tl{anargalā suṣumṇā ca haṭhasiddhiḥ prajāyate || iti ||\\+}
\tl{kumbhakaprāṇāyāmāt bodho jāgaraṇam | suṣumṇā anargalā bādhakarahitā bhavati | tato yogasiddhir bhavati iti ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_079]
%\end{philcomm}

%%%%%%%%%%
\subsection*{2.80}
\begin{translation}[hp02_080]
Thinness of the body, clear complexion, clarity of the inner sound, very bright eyes, freedom from disease, mastery of semen, stimulation of the [body’s] fire [and] purification of the channels are the signs of success in Haṭha.
\end{translation}

\begin{testimonia}[hp02_080]
\emph{Haṭharatnāvalī} 1.59

\begin{versinnote}
\tl{vapuḥkṛśatvaṃ vadane prasannatā\\+} 
\tl{nādaspuṭatvaṃ nayane ca nirmale | \\+}
\tl{arogatā bindujayo 'gnidīpanaṃ\\+}  
\tl{nāḍīṣu śuddhir haṭhasiddhilakṣaṇam ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~111v (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{vapuḥkṛśatvaṃ vadane prasannatā \\+}
\tl{nādasphuṭatvaṃ nayane sunirmale |\\+}
\tl{arogitā bindujayo 'gnidīpanaṃ \\+}
\tl{nāḍīviśuddhir haṭhasiddhilakṣaṇam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.129 (attr.~to Gorakṣanātha)

\begin{versinnote}
\tl{vapuḥkṛśatvaṃ vadane prasannatā \\+}
\tl{nādasphuṭatvaṃ nayane ca nirmale |\\+}
\tl{arogatā bindujayo'gnidīpanaṃ \\+}
\tl{nāḍīviśuddhir haṭhasiddhilakṣaṇam ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp02_080]
%\end{philcomm}

\begin{metre}[hp02_080]
Vaṃśamālā
\end{metre}

\end{ekdosis}
\end{document}

### 2024-03-19 latex/HP1_comm.tex
new files and Nagari-Support for Jyotsna

@@
 \begin{translation}[hp01_024]
-While maintaining the wild cock pose, [the yogi] binds the neck with the hands and lies [on his back] upturned like a tortoise. This is the upturned tortoise (\emph{uttānakūrmaka}).
+While in the wild cock pose, [the yogi] binds the neck with the hands and lies [on his back] upturned like a tortoise. This is the upturned tortoise (\emph{uttānakūrmaka}).
 \end{translation}
@@
 \begin{translation}[hp01_027]
-Matsyendra's seat is a destructive missile for the many terrible diseases that develop in the stomach; through practice it brings about in people the awakening of Kuṇḍalinī and steadiness of the spine.
+Matsyendra's seat is a destructive missile for the many terrible diseases that develop in the stomach; through practice it brings about in men the awakening of Kuṇḍalinī and steadiness of the spine.
 \end{translation}
@@
 % ?? Mention that the mudrās practised with the breathing techniques are the bandhas (2.44-45) 
-%The metre of 1.55c is a \emph{ra-vipulā} with \emph{nāḍi}.%
 \end{philcomm}

### 2024-03-19 latex/HP2_comm.tex
new files and Nagari-Support for Jyotsna

@@
 
-Cf. \emph{Satkarmasaṅgraha} 135, 140–141
+Cf.~\emph{Satkarmasaṅgraha} 135, 140–141
 
@@
 As is clear in the source text, the \emph{Gorakṣaśataka}, the three techniques alluded to here are \sl{mūlabandha, jālandharabandha} and \emph{uḍḍiyānabandha} respectively (on which see chapter three).
+
+The stem form \emph{brahma} is found in various compounds in the text.
+We understand it to refer to Brahman, the absolute, in all instances other than \emph{brahmagranthi}, where we take it to mean the deity Brahmā.
+See Mallinson 2007:205 n.240.
 \end{philcomm}
@@
 
-Cf. \emph{Kumbhakapaddhati} 169
+Cf.~\emph{Kumbhakapaddhati} 169
 
@@
 
-Cf. \emph{Gheraṇḍasaṃhitā} 7.10--11
+Cf.~\emph{Gheraṇḍasaṃhitā} 7.10--11
 
@@
 \begin{translation}[hp02_071]
-Now Plāvanī:\\
+Now plāvanī:\\
 With his abdomen completely filled with the wind of eructation, which has been turned inwards, [the yogi] floats easily even in deep water, like a lotus leaf.

### 2024-03-19 latex/HP3_comm.tex
new files and Nagari-Support for Jyotsna

@@
 
-\chapter*{Translation \& philological commentary}
+%\chapter*{Translation \& philological commentary}
 %%%%%%%%%%
@@
 \begin{translation}[hp03_001]
-test.
+Just as the lord of snakes is the foundation of the regions of the earth along with their mountains and forests, so \emph{kuṇḍalinī} is the foundation of all systems of yoga.
 \end{translation}
+
+%\begin{sources}[hp03_001]
+%\end{sources}
+
+\begin{testimonia}[hp03_001]
+\emph{Haṭharatnāvalī} 2.124
+\begin{versinnote}
+\tl{saśailavanadhātryās tu yathādhāro 'hināyakaḥ |\\+}
+\tl{aśeṣayogatantrāṇāṃ tathādhāro hi kuṇḍalī ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~71v
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{saśailavanadhātrīṇāṃ yathādhāro 'hināyakaḥ |\\+}
+\tl{sarveṣāṃ yogatantrāṇāṃ tathādhāro hi kuṇḍalī ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.170
+\begin{versinnote}
+\tl{tatra haṭhapradīpikāyām--\\+}
+\tl{saśailavanadhātrīṇāṃ yathādhāro 'hināyakaḥ |\\+}
+\tl{sarveṣāṃ yogatantrāṇāṃ tathādhāro hi kuṇḍalī ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_001]
+The plural \emph{°dhātrīṇām} is hard to construe.
+Only this world has mountains and forests, but we want a plural for the comparison with \emph{°tantrāṇām}.
+Brahmānanda (\emph{Jyotsnā} 3.1) understands \emph{°dhātrīṇām} to refer to the different regions of the earth, even though the world (\emph{dhātrī}) is a single entity (\emph{dhātryā ekatve 'pi deśabhedād bhedam ādāya bahuvacanam}).
+A similar comment occurs in the \emph{Yogaprakāśikā} 5.1 (\emph{yathā samastadvīpādisahitapṛthvī ādhāraḥ phaṇīndras tathā samastayogādhāraḥ kuṇḍalīty āha saśaileti}).
+The author of the \textit{Haṭharatnāvalī} circumvented this issue by adopting the reading °\textit{dhātryāḥ}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.2}
+\begin{translation}[hp03_002]
+When the sleeping \emph{kuṇḍalinī} awakens through the favour of the guru, then all the lotuses are pierced, and the knots too, [\dots] 
+\end{translation}
+
+\begin{sources}[hp03_002]
+\emph{Śivasaṃhitā} 4.21
+\begin{versinnote}
+\tl{suptā guruprasādena yadā jāgarti kuṇḍalī |\\+}
+\tl{tadā sarvāṇi padmāni bhidyante granthayo 'pi ca ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_002]
+\emph{Yogacintāmaṇi} f.~71v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{suptā guruprasādena bodhitā sukhadā bhavet |\\+}
+\tl{tathā sarvāṇi padmāni bhidyante granthayo 'pi ca ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 171 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{suptā guruprasādena yadā jāgarti kuṇḍalī |\\+}
+\tl{tadā sarvāṇi padmāni bhidyante granthayo 'pi ca ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_002]
+The usual meaning of \emph{jāgarti} would be “is wakeful” rather than “awakens”, which explains the variant \emph{bodhitā}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.3}
+\begin{translation}[hp03_003]
+[\dots] the empty pathway becomes the royal highway for \emph{prāṇa}, the mind becomes free of support, and death is cheated.
+\end{translation}
+
+\begin{sources}[hp03_003]
+\end{sources}
+
+\begin{testimonia}[hp03_003]
+\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{prāṇasya śūnyapadavī tathā rājapathāyate |\\+}
+\tl{tathā cittaṃ nirālambaṃ tathā kālasya vañcanam ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_003]
+%JM: take prāṇasya with śūnyapadavī?
+%JB: tadā in 3b indicates that this verse should be read with 3.2 (i.e., [when kuṇḍalinī has awakened...] then, the empty pathway becomes the royal pathway for prāṇa)
+%\end{philcomm}
+
+\begin{metre}[hp03_003]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.4}
+What is ``the empty pathway"?
+
+\begin{translation}[hp03_004]
+\emph{Suṣumṇā}, the empty pathway, the great path to the aperture of Brahmā, the cremation ground, Śāmbhavī, and the middle path are synonyms.
+\end{translation}
+
+\begin{sources}[hp03_004]
+Cf.~\emph{Amṛtasiddhi} 2.6
+\begin{versinnote}
+\tl{avadhūtīpadaṃ ke cic chmaśānaṃ ca mahāpatham |\\+}
+\tl{ke cid vadanti ādhārāṃ suṣumnāṃ ca sarasvatīm ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Dattātreyayogaśāstra} 109c–110b
+\begin{versinnote}
+\tl{mahāpathaṃ śmaśānaṃ ca suṣumnāpy ekam eva hi ||\\+}
+\tl{nāmnāṃ matāntare bhedaḥ phale bhedo na vidyate |\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_004]
+\emph{Yogacintāmaṇi} f.~59r
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{suṣumṇā śūnyapadavī brahmarandhraṃ mahāpatham |\\+}
+\tl{śmaśānī śāṃbhavī madhyamārgaś cety ekavācakā iti ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.172 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{prāṇasya śūnyapadavī mahārandhraṃ mahāpatham |\\+}
+\tl{śmaśānaṃ śāmbhavī madhyamārgaś cety ekavācakam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_004]
+%In the context of \textit{suṣumṇā}, Svātmārāma seems to understand \emph{śāmbhavī} as the power (\emph{śakti}) of Śambhu (Cf.~\emph{Haṭhapradīpikā} 4.78c \emph{suṣumṇā śāmbhavī śaktiḥ}).
+JM: I'm not convinced by śāmbhavī = power of Śambhu; I think it's more feminine.
+%\end{philcomm}
+
+\begin{metre}[hp03_004]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.5}
+\begin{translation}[hp03_005]
+Therefore, in order to do his utmost to awaken the goddess sleeping in front of the doorway of Brahman, [the yogi] should carry out the practice of \emph{mudrā}.
+\end{translation}
+
+\begin{sources}[hp03_005]
+\emph{Śivasaṃhitā} 4.22
+\begin{versinnote}
+\tl{tasmāt sarvaprayatnena prabodhayitum īśvarīm |\\+}
+\tl{brahmadvāramukhe suptāṃ mudrābhyāsaṃ samācaret ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_005]
+\emph{Yogacintāmaṇi} f.~59r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tasmāt sarvaprayatnena prabodhayitum īśvarīm |\\+}
+\tl{brahmadvāramukhe suptāṃ mudrābhyāsaparo bhavet ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.173 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tasmāt sarvaprayatnena prabodhayitum īśvarīm |\\+}
+\tl{brahmadvāramukhe suptāṃ mudrābhyāsaṃ samācaret ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_005]
+%Consider pathe for mukhe [JB: no support for pathe]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.6}
+\begin{translation}[hp03_006]
+The great seal, the great lock, the great piercing, the sky-roving [seal], the \emph{uḍḍiyāna} [lock], the root lock and the lock called \emph{jālandhara}, [\dots] 
+\end{translation}
+
+\begin{sources}[hp03_006]
+%Cf.~Amṛtasiddhi 13.12
+%\begin{versinnote}
+%\tl{mahāmudrā mahābandho mahāvedhas tṛtīyakaḥ |\\!}
+%\end{versinnote}
+
+\emph{Śivasaṃhitā} 4.23
+\begin{versinnote}
+\tl{mahāmudrā mahābandho mahāvedhaś ca khecarī |\\+}
+\tl{jālandharo mūlabandho viparītakṛtis tathā ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_006]
+\emph{Haṭharatnāvalī} 2.32
+
+\begin{versinnote}
+\tl{mahāmudrā mahābandho mahāvedhas tṛtīyakaḥ |\\+}
+\tl{uḍḍiyānaṃ mūlabandho bandho jālandharābhidhaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mahāmudrā mahābandho mahāvedhaś ca khecarī |\\+}
+\tl{uḍḍiyānaṃ mūlabandho bandho jālandharābhidhaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.174 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mahāmudrā mahābandho mahāvedhaś ca khecarī |\\+}
+\tl{uḍyānaṃ mūlabandhaś ca bandho jālandharas tathā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_006]
+%JM: Do we translate mudrā or not?
+%\end{philcomm}
+
+\begin{metre}[hp03_006]
+Anuṣṭubh (c: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.7}
+\begin{translation}[hp03_007]
+[\ldots] the bodily position called inverted, \emph{vajrolī} [and] the stimulation of the goddess: this group of ten \emph{mudrā}s and other [practices] destroys old age and death. 
+\end{translation}
+
+\begin{sources}[hp03_007]
+\emph{Śivasaṃhitā} 4.24
+\begin{versinnote}
+\tl{uḍyānaṃ caiva vajrolī daśamaṃ śakticālanam |\\+}
+\tl{idaṃ hi mudrādaśakaṃ mudrāṇām uttamottamam ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_007]
+\emph{Haṭharatnāvalī} 2.33
+\begin{versinnote}
+\tl{karaṇī viparītākhyā vajrolī śakticālanam |\\+}
+\tl{sampradāyā khecarī sā daśa mudrāḥ prakīrtitāḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{karaṇī viparītākhyā tathā vai śakticālanam |\\+}
+\tl{etad dhi mudrānavakaṃ jarāmaraṇavarjitam ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.175 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{viparītakṛtiś caiva vajrolī śakticālanam |\\+}
+\tl{idaṃ hi mudrādaśakaṃ mudrāṇām uttamomam ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_007]
+%JB: hi (in 3.7c) is supported only by the \textit{Jyotsnā}. Mss. have tu and ca.
+%\end{philcomm}
+
+\begin{metre}[hp03_007]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.8}
+\begin{translation}[hp03_008]
+It has been taught by Śiva, is divine, bestows the eight supramundane powers, is beloved of all the Siddhas, is difficult for even the gods to obtain, [\dots] 
+\end{translation}
+
+\begin{sources}[hp03_008]
+\end{sources}
+
+\begin{testimonia}[hp03_008]
+\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ādināthoditaṃ divyam aṣṭaiśvaryapradāyakam |\\+}
+\tl{vallabhaṃ sarvasiddhānāṃ durlabhaṃ marutām api ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.176 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ādināthoditaṃ samyagaṣṭaiśvaryapradāyakam |\\+}
+\tl{vallabhaṃ sarvasiddhendradurlabhaṃ marutām api ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_008]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.9}
+\begin{translation}[hp03_009]
+[\ldots] should be carefully kept secret like a casket of gems [and] must not be spoken of to anyone, like sex with a respectable woman.
+\end{translation}
+
+\begin{sources}[hp03_009]
+\end{sources}
+
+\begin{testimonia}[hp03_009]
+\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{gopanīyaṃ prayatnena yathā ratnakaraṇḍakam |\\+}
+\tl{kasyacin naiva vaktavyaṃ kulastrīsurataṃ yathā ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.177 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{gopanīyaṃ prayatnena jarāmaraṇanāśanam |\\+}
+\tl{kasyacinnaiva vaktavyaṃ kulastrīsurataṃ yathā || \\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_009]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.9*1}
+\begin{translation}[hp03_009_1]
+[With] \emph{Amarolī} and \emph{Sahajolī}, Vajrolī is considered to be threefold. I shall teach their characteristics and the details of how they should be performed.
+\end{translation}
+
+\begin{sources}[hp03_009_1]
+\emph{Dattātreyayogaśāstra} 31c–32b
+\begin{versinnote}
+\tl{vajrolī cāmarolī ca sahajolī tridhā matā |\\+}
+\tl{eteṣāṃ lakṣaṇaṃ vakṣye kartavyaṃ ca viśeṣataḥ ||\\+}
+\tl{\var{vajrolī ] vajrolir \vl • cāmarolī ] amaroliś \vl • sahajolī ] sahajolis \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{philcomm}[hp03_009_1]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.10}
+\begin{translation}[hp03_010]
+Now the great seal (\emph{mahāmudrā}):\\
+{}[The yogi] should press the perineum with the heel of the left foot, hold [the foot of] the extended leg with the hands and breathe in through the mouth.
+\end{translation}
+
+\begin{sources}[hp03_010]
+\emph{Amaraugha} 19
+\begin{versinnote}
+\tl{pādamūlena vāmena yoniṃ saṃpīḍya dakṣiṇam |\\+}
+\tl{pādaṃ prasāritaṃ dhṛtvā karābhyāṃ pūrayen mukhe ||\\+}
+\tl{\var{dhṛtvā ] kṛtvā \vl}\\!}
+\end{versinnote}
+
+Cf.~\emph{Amṛtasiddhi} 11.3
+\begin{versinnote}
+\tl{yoniṃ saṃpīḍya vāmena pādamūlena yatnataḥ |\\+}
+\tl{savyaṃ prasāritaṃ pādaṃ karābhyāṃ dhārayed dṛḍham ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_010]
+\emph{Haṭharatnāvalī} 2.37
+\begin{versinnote}
+\tl{pādamūlena vāmena yoniṃ sampīḍya dakṣiṇam |\\+}
+\tl{pādaṃ prasāritaṃ kṛtvā karābhyāṃ pūrayen mukham ||\\+}
+\tl{\var{mukham ] mukhe \vl}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} ff. 72v–73r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pādamūlena vāmena yoniṃ saṃpīḍya dakṣiṇam |\\+}
+\tl{pādaṃ prasāritaṃ dhṛtvā karābhyāṃ pūrayen mukham ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.178 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pādamūlena vāmena yoniṃ sampīḍya dakṣiṇam |\\+}
+\tl{pādaṃ prasāritaṃ kṛtvā karābhyāṃ dhārayed dṛḍham ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.11}
+\begin{translation}[hp03_011]
+[The yogi] should apply a lock to the throat and hold the breath in the upper [part of the body]. Just as a snake hit with a staff assumes the form of a staff, [\dots] 
+\end{translation}
+
+\begin{sources}[hp03_011]
+\emph{Amaraugha} 20
+\begin{versinnote}
+\tl{kaṇṭhe bandhaṃ samāropya dhārayed vāyum ūrdhvataḥ |\\+}
+\tl{yathā daṇḍāhataḥ sarpo daṇḍākāraḥ prajāyate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_011]
+\emph{Haṭharatnāvalī} 2.37cd–38ab
+\begin{versinnote}
+\tl{kaṇṭhe bandhaṃ samāropya pūrayed vāyum ūrdhvataḥ ||\\+}
+\tl{yathā daṇḍāhataḥ sarpo daṇḍākāraḥ prajāyate |\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} ff.~73r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{kaṇṭhe bandhaṃ samāropya dhārayed vāyum ūrdhvataḥ |\\+}
+\tl{yathā daṇḍāhataḥ sarpo daṇḍākāraḥ prajāyate ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.179 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{kaṇṭhe bandhaṃ samāropya dhārayed vāyum ūrdhvataḥ |\\+}
+\tl{yathā daṇḍāhataḥ sarpo daṇḍākāraḥ prajāyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_011]
+The instruction to hold the breath upwards (\emph{ūrdhvataḥ}) is somewhat vague.
+In a commentarial passage on this verse in \emph{Yuktabhavadeva} 7.187, Bhavadevamiśra clarifies this by saying, `one should hold it higher than the heart' (\emph{hṛdayād ūrdhvato dhārayet}).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.12}
+\begin{translation}[hp03_012]
+[\dots] so the goddess \emph{kuṇḍalinī} suddenly becomes straight. Then she becomes still in the vessel with two halves.
+\end{translation}
+
+\begin{sources}[hp03_012]
+\emph{Amaraugha} 21
+\begin{versinnote}
+\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahasā bhavet |\\+}
+\tl{tadāsau maraṇāvasthā jāyate dvipuṭāśritā ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_012]
+\emph{Haṭharatnāvalī} 2.38cd–39ab
+\begin{versinnote}
+\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahajā bhavet ||\\+}
+\tl{tathā sā maraṇāvasthā jāyate dvipuṭīsthitā |\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} ff. 72v–73r (attr.~to the \emph{Haṭhapradīpikā}) % order of the hemistichs has been changed. I don't know if or where I should mention this.
+\begin{versinnote}
+\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahajā bhavet |\\+}
+\tl{tathāsau maraṇāvasthāṃ harate dvipuṭāśrayām ||\\+}
+\tl{dvipuṭā ] em., dvipaṭā° L, dvipadā° N. \\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.180 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahasā bhavet |\\+}
+\tl{tadāsau maraṇāvasthā jāyate dvipuṭāśritā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_012]
+This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of \emph{kuṇḍalinī} in the central channel.
+Drawing on Hellwig 2009:238–240, Mallinson and Szanto (2021:21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert.
+In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when \emph{kuṇḍalinī} is said to be in the state of \emph{maraṇa} the meaning is that she is stilled.
+The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009:342, Mallinson and Szanto (2021:22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”.
+In the yoga of the \emph{Amṛtasiddhi}, the bodily \emph{saṃpuṭa} is formed by applying locks at the top and bottom of the central channel, i.e.~constricting the perineal region and the throat.
+In the \emph{Haṭhapradīpikā} it is formed by pressing the perineum with the heel and constricting the throat.
+
+As Birch (2019: 971) notes, it is unlikely that later non-Buddhist authors understood \textit{maraṇāvasthā} and \emph{dvipuṭa} according to the alchemical metaphors of the \textit{Amṛtasiddhi}.
+Later commentators take \emph{dvipuṭa} as the two nostrils (e.g., \textit{Yuktabhavadeva} 7.187, \textit{dvināsāpuṭa}); the \textit{iḍā} and \textit{piṅgalā} channels (e.g., \textit{Jyotsnā} 3.27, \textit{puṭayor dvayam iḍāpiṅgalayor yugmam}); or the in and out flows of the breath (e.g., \textit{Yogaprakāśikā} 5.16–17, \textit{vāyor bahirnirgamanam antaḥpraveśa iti yat puṭadvayaṃ tam}).
+How these commentators understood \textit{asau māraṇāvasthā} is less clear.
+Brahmānanda seems to take it as the death of \textit{prāṇa}, or in other words, the absence of the breath, in the two nostrils (\textit{maraṇāvasthā jāyate kuṇḍalībodhe sati suṣumnāyāṃ praviṣṭe prāṇe dvayoḥ puṭayoḥ prāṇaviyogāt}).
+Bhavadeva thought that \emph{kuṇḍalinī}, along with \textit{prāṇa} and \textit{apāna}, remains in the two nostrils while the breath is being held (\textit{evaṃ vāyudhāraṇāyāṃ kriyamāṇāyāṃ vyākulā bhūtā kuṇḍalinī apānaprāṇābhyāṃ saha nāsāpuṭadvayāśritā bhavati}).
+Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvipuṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}.
+Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened \emph{kuṇḍalinī} destroys death, which is usually dependent on the in and out breaths.
+
+%JB: I think our translation should reflect a likely Śaiva interpretation rather than the alchemical metaphors in the Amṛtasiddhi (e.g., `Then, that state of death arises in the two nostrils').
+JM: this would imply that we think that Svātmārāma decided to use the verse with this understanding, which seems unlikely to me.
+JB.
+Unlikely?
+All the evidence in commentaries on the HP and later compendiums, like the YCM and Yuktabhavadeva, indicates that the alchemical metaphors of the AS were not understood here.
+It's not even clear that the alchemical metaphors work with raising moving or raising kuṇḍalinī.
+In contrast to this, there's not one commentarial explanation or rewrite of the verse that reveals an alchemical meaning.
+We also have to bear in mind that there's no evidence that Svātmārāma and later commentators knew the AS.
+In the one case where there is some evidence (i.e., Śivānanda quotes it in the YCM), it doesn't seem to have helped the author understand the alchemical metaphor in this instance (as Śivānanda rewrote the verse to say that kuṇḍalinī destroys the state of death).
+
+%The meaning of "Śaiva" here and above is unclear to me. Doesn't it work to say  "later authors"? JB I've changed the note above to non-Buddhist.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.13}
+\begin{translation}[hp03_013]
+{}[The yogi] should then exhale very slowly, not quickly. This is the great seal revealed by the great Siddhas.
+\end{translation}
+
+%\begin{sources}[hp03_013]
+%\end{sources}
+
+\begin{testimonia}[hp03_013]
+\emph{Yuktabhavadeva} 181 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tataḥ śanaiḥ śanair eva recayen na tu vegataḥ |\\+}
+\tl{iyaṃ khalu mahāmudrā mahāsiddhaiḥ pradṛśyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_013]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.14}
+\begin{translation}[hp03_014]
+Problems such as the great afflictions [and] death and so forth are removed [by it], and that is why the most wise call it the `great seal.'
+\end{translation}
+
+\begin{sources}[hp03_014]
+\emph{Amaraugha} 22
+\begin{versinnote}
+\tl{mahākleśādayo doṣā bhidyante maraṇādayaḥ |\\+}
+\tl{mahāmudrāṃ tu tenaiva vadanti vibudhottamāḥ ||\\+}
+\tl{\var{mahākleśādayo doṣā ] mahārogā mahākleśā \vl}\\+}
+\tl{\var{bhidyante ] jīryante \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_014]
+\emph{Yogacintāmaṇi} f.~72v (attr.~to the \emph{Skandapurāṇa})
+\begin{versinnote}
+\tl{mahākleśā yato doṣā jīryante maraṇādayaḥ |\\+}
+\tl{mahāmudrāṃ ca tenaiva vadanti vibudhottamā iti ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.182 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mahākleśā yato doṣā jīryante maraṇādayaḥ |\\+}
+\tl{mahāmudrā ca tām eva vadanti vibudhottamāḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_014]
+%Discuss the group 2 reading \emph{mahākleśā yato}.
+Also the V19 reading \emph{mahākleśā mahādoṣā}, which is similar to the \emph{Amaraughaprabodha}'s reading (\textit{mahārogā mahākleśā}).
+[JB: can't see a need for a note as alpha reading is close to the Amaraugha.]
+%JM: note on assonance with the sounds of first line supporting the name?
+This verse might be explaining the name of \emph{mahāmudrā} through assonance with \emph{mahākleśādayo, doṣā} and \emph{maraṇādayaḥ} in the first line.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.15}
+\begin{translation}[hp03_015]
+After practising on the lunar side of the body, the yogi should then practise on the solar side. [The yogi] should finish practising the seal when the count is even.
+\end{translation}
+
+\begin{sources}[hp03_015]
+\emph{Vivekamārtaṇḍa} 60
+\begin{versinnote}
+\tl{candrāṅge tu samabhyasya sūryāṅge punar abhyaset |\\+}
+\tl{yāvat tulyā bhavet saṅkhyā tato mudrāṃ visarjayet ||\\!}
+%\tl{\var{candrāṅge tu ] candrāṅgena \vl • sūryāṅge punar abhyaset ] sūryāṅgenābhyaset punaḥ \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_015]
+\emph{Yuktabhavadeva} 7.183 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{candrāṅgena samabhyasya sūryāṅgenābhyaset tataḥ |\\+}
+\tl{yāvat tulyā bhavet saṃkhyā tato mudrāṃ visarjayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_015]
+The terms \emph{candrāṅga} and \emph{sūryāṅga} are unusual and not used in other yoga texts outside the context of \emph{mahāmudrā}.
+In \emph{Jyotsnā} 3.15, Brahmānanda glosses \emph{candrāṅga} as \emph{vāmāṅga} (`the left side of the body') and \emph{sūryāṅga} as \emph{dakṣāṅga} (`the right side of the body') and goes on to explain the sequence of practice as follows:
+
+\begin{quote}
+This is the sequence in the [practice].
+Joining the heel of the bent left leg with the region of the perineum and holding the big toe of the extended right leg with bent index fingers bent is the practice, that is, the practice on the left side of the body (\emph{vāmāṅga}).
+In this practice, the inhaled breath remains on the left side of the body.
+Joining the heel of the bent right leg with the region of the perineum and holding the big toe of the extended left leg with bent index fingers is the practice, that is, the practice on the right side of the body.
+In this practice, the inhaled breath remains on the right side of the body.
+\emph{atrāyaṃ kramaḥ} | \emph{ākuñcitavāmapādapārṣṇiṃ yonisthāne saṃyojya prasāritadakṣiṇapādāṅguṣṭham ākuñcitatarjanībhyāṃ gṛhītvābhyāso vāmāṅge 'bhyāsaḥ} | \emph{asminn abhyāse pūrito vāyur vāmāṅge tiṣṭhati} | \emph{ākuñcita\-dakṣa\-pāda\-pārṣṇiṃ yonisthāne saṃyojya prasārita\-vāma\-pādā\-ṅguṣṭham ākuñcita\-tarjanībhyāṃ gṛhītvābhyāso dakṣāṅge 'bhyāsaḥ} | \emph{asminn abhyāse pūrito vāyur dakṣāṅge tiṣṭhati} |
+\end{quote}
+
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.16}
+\begin{translation}[hp03_016]
+{}[For the yogi who practises thus] there is no wholesome or unwholesome [food], and all flavours without exception become flavourless. Even terrible poison, when consumed, is digested like nectar.
+\end{translation}
+
+\begin{sources}[hp03_016]
+\emph{Vivekamārtaṇḍa} 61
+\begin{versinnote}
+\tl{na hi pathyam apathyaṃ vā rasāḥ sārve 'pi nīrasāḥ |\\+}
+\tl{api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam iva jīryati ||\\+}
+\tl{\var{jīryati] jīryate \vl, jāyate \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_016]
+\emph{Haṭharatnāvalī} 2.40
+\begin{versinnote}
+\tl{na hi pathyam apathyaṃ vā rasāḥ sarve 'pi nīrasāḥ |\\+}
+\tl{api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam iva jīryate ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{na hi pathyam apathyaṃ vā rasāḥ sarve 'pi nīrasāḥ |\\+}
+\tl{api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam iva jīryati ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.184 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{iha pathyam apathyaṃ vā sarasaṃ nirasaṃ ca vā |\\+}
+\tl{api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam iva jīryati ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_016]
+%In \emph{Jyotsnā} 3.16, Brahmānanda understands the second verse quarter to mean all flavours –pungent (\emph{kaṭu}), acid (\emph{amlā}), etc.– and stale (\emph{nīrasa}) [food] and takes it with the verb \emph{jīryati}.
+%[JB: This note doesnt seem relevant to our translation]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.17}
+\begin{translation}[hp03_017]
+Diseases such as consumption, skin afflictions, constipation, swelling and indigestion disappear for [the yogi] who practises the great seal.
+\end{translation}
+
+\begin{sources}[hp03_017]
+\emph{Vivekamārtaṇḍa} 62
+\begin{versinnote}
+\tl{kṣayakuṣṭagudāvarttagulmājīrṇajvaravyathāḥ |\\+}
+\tl{tasya doṣāḥ kṣayaṃ yānti mahāmudrāṃ tu yo 'bhyaset ||\\+}
+\tl{\var{tasya doṣāḥ ] sarvarogāḥ \vl, rogās tasya \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_017]
+\emph{Haṭharatnāvalī} 2.41
+\begin{versinnote}
+\tl{kṣayakuṣṭhagudāvartagulmājīrṇapurogamāḥ |\\+}
+\tl{doṣāḥ sarve kṣayaṃ yānti mahāmudrāṃ tu yo 'bhyaset ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{kṣayakuṣṭhagudāvartagulmaplīhapurogamāḥ |\\+}
+\tl{tasya doṣāḥ kṣayaṃ yānti mahāmudrāṃ ca yo 'bhyaset || \\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.185 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{kṣayakuṣṭhagudāvarttagulmājīrṇapurogamāḥ |\\+}
+\tl{tasya rogāḥ kṣayaṃ yānti mahāmudrāṃ tu yo'bhyaset ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_017]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.18}
+\begin{translation}[hp03_018]
+This great seal has been taught. It brings about the great \emph{siddhi} for men: it should be carefully kept secret and should not be given to all and sundry.
+\end{translation}
+
+\begin{sources}[hp03_018]
+\emph{Vivekamārtaṇḍa} 63
+\begin{versinnote}
+\tl{kathiteyaṃ mahāmudrā mahāsiddhikarī nṛṇām |\\+}
+\tl{gopanīyā prayatnena na deyā yasya kasya cit ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_018]
+\emph{Haṭharatnāvalī} 2.42
+\begin{versinnote}
+\tl{kathiteyaṃ mahāmudrā jarāmṛtyuvināśinī |\\+}
+\tl{gopanīyā prayatnena na deyā yasya kasya cit ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.186 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{kathiteyaṃ mahāmudrā mahāsiddhikarī nṛṇām |\\+}
+\tl{gopanīyā prayatnena na deyā yasya kasya cit ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_018]
+Manuscripts of three groups (\textalpha, \textbeta\ and \textgamma), which are important stemmatically, have the reading \emph{jarāmṛtyuvināśinī} (and \getsiglum{J5} has \emph{nṛṇāṃ mṛtyuvināśinī}) for the second verse quarter.
+While this reading is possible, the play on \emph{mahāsiddhi} and \emph{mahāmudrā} seems more likely original, as seen in the source text, the \emph{Vivekamārtaṇḍa} (without significant variants), and the \texteta\ and \textepsilon\ groups.
+
+In \emph{Jyotsnā} 3.18, Brahmānanda understands \emph{mahāsiddhi} as referring to `great \emph{siddhi}s,' but in other works it can mean liberation (Mallinson 2012).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.19}
+\begin{translation}[hp03_019]
+Now, the great lock (\emph{mahābandha}):\\
+{}[The yogi] should place the heel of the left foot on the perineal region. And he should put the right foot on the left thigh, [\dots]
+\end{translation}
+
+\begin{sources}[hp03_019]
+19ab = \emph{Dattātreyayogaśāstra} 132cd (in the section on \emph{mahāmudrā})
+\begin{versinnote}
+\tl{pārṣṇiṃ vāmasya pādasya yonisthāne niyojayet ||\\!}
+\end{versinnote}
+
+19cd = Cf.~\emph{Vivekamārtaṇḍa} 8a (not \emph{anuṣṭubh})
+\begin{versinnote}
+\tl{vāmorūpari dakṣiṇañ ca caraṇaṃ saṃsthāpya\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_019]
+\emph{Haṭharatnāvalī} 2.43
+\begin{versinnote}
+\tl{pārṣṇiṃ vāmasya pādasya yonisthāne niyojayet |\\+}
+\tl{vāmorūpari saṃsthāpya dakṣiṇaṃ caraṇaṃ tathā ||\\!}
+\end{versinnote}
+
+\emph{Yogalakṣanāvalī} f.~31v
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{vāmāṅghripārṣṇibhāgena yonisthānaṃ nipīḍayet |\\+}
+\tl{vāmorupari saṃsthāpya dakṣiṇaṃ caraṇaṃ tathā ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73r
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{pārṣṇiṃ vāmasya pādasya yonisthāne niyojayet |\\+}
+\tl{vāmorūpari saṃsthāpya dakṣiṇaṃ caraṇaṃ tathā ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.190 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pārṣṇivāmasya pādasya yonisthāne niyojayet |\\+}
+\tl{vāmorupari saṃsthāpya dakṣiṇaṃ caraṇaṃ tathā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_019]
+The seated position for \emph{mahābandha} described in this verse is not in the \emph{Amṛtasiddhi} (chapter 12) or \emph{Amaraugha} (25cd–27).
+The \emph{Amṛtasiddhi} instructs the same position for \emph{mahāmudrā} and \emph{mahābandha}, and the \emph{Amaraugha} does not comment on the posture of \emph{mahābandha}, implying that its posture is the same as \emph{mahāmudrā}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.20}
+\begin{translation}[hp03_020]
+[\dots] inhale through the mouth, firmly put the chin on the chest, contract the perineum and fix the mind in the centre.
+\end{translation}
+
+\begin{sources}[hp03_020]
+\emph{Amaraugha} 24
+\begin{versinnote}
+\tl{pūrayitvā mukhe vāyuṃ cibukaṃ hṛdaye dṛḍham |\\+}
+\tl{nibhṛtya yonim ākuñcya mano madhye niyojayet ||\\+}
+\tl{\var{nibhṛtya ] nibhṛtaṃ \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_020]
+\emph{Haṭharatnāvalī} 2.44
+\begin{versinnote}
+\tl{pūrayen mukhato vāyuṃ hṛdaye cibukaṃ dṛḍham |\\+}
+\tl{nibhṛtya yonim ākuñcya mano madhye niyojayet ||\\+}
+\tl{\var{nibhṛtya ] niṣpīḍya \vl}\\!}
+\end{versinnote}
+
+\emph{Yogalakṣanāvalī} f.~31v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pūrayitvā tato vāyuṃ hṛdaye cibukaṃ dṛḍham |\\+}
+\tl{niḥpīḍya yonim ākuṃcya mano madhye niyojayet ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pūrayitvā mukhe vāyuṃ hṛdaye cibukaṃ tathā |\\+}
+\tl{niḥpīḍya yonim ākuñcya mano madhye niyojayet || \\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.190 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pūrayitvā tato vāyuṃ cibukaṃ dṛḍham |\\+}
+\tl{niḥkṣipya yonim ākuñcya mano madhye niyojayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_020]
+The referent of \emph{madhye} is uncertain.
+The verse is derived from the \emph{Amaraugha}, and the \emph{Amṛtasiddhi} makes no mention of a place to focus the mind in its treatment of \emph{mahābandha} (it does however instruct the yogi to place the mind at the \emph{catuṣpatha} in its teachings on \emph{mahāmudrā}).
+Bhavadevamiśra (7.196), Brahmānanda (3.20) and Bālakṛṣṇa (5.24) take it to mean the central channel.
+It could also plausibly mean the region between the chest and perineum, or perhaps the place between the eyebrows.
+At 3.24 this practice is said to make the mind reach Kedāra, which is sometimes located between the eyebrows (see Mallinson 2007:214 n.285, Birch 2019: 967 n.57).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.21}
+\begin{translation}[hp03_021]
+[The yogi] should hold the breath as long as possible and exhale slowly. And having practised it on the left side, he should practise it on the right side. 
+\end{translation}
+
+\begin{sources}[hp03_021]
+\emph{Dattātreyayogaśāstra} 62cd (\emph{padmāsana}), 134cd (\emph{mahāmudrā})
+\begin{versinnote}
+\tl{dhārayitvā yathāśakti recayed iḍayā śanaiḥ ||\\+}
+\tl{...\\+}
+\tl{vāmāṅgena samabhyasya dakṣiṇāṅgena cābhyaset ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_021]
+\emph{Haṭharatnāvalī} 2.44cd–2.45
+\begin{versinnote}
+\tl{recayec ca śanair evaṃ mahābandho 'yam ucyate ||\\+}
+\tl{ayaṃ yogo mahābandhas sarvasiddhipradāyakaḥ |\\+}
+\tl{savyāṅge ca samabhyasya dakṣiṇāṅge samabhyaset ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.192ab (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{dhārayitvā yathāśakti recayed anilaṃ sudhīḥ |\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_021]
+This section in the description of \emph{mahābandha} has been subject to various revisions in the manuscript groups.
+The shortest version occurs in \alphaOne\ and \alphaTwo, as well as the \textbeta\ group and the \emph{Haṭharatnāvalī}, which omit verses 3.21 and 3.22.
+This omission results in the somewhat odd juxtaposition of two lines stating that this is \emph{mahābandha}.
+
+Verse 3.21 is in the \textgamma\ group and some manuscripts of the \textbeta\ and \texteta\ groups.
+It is a composite of hemistichs from the \emph{Dattātreyayogaśāstra} (62cd, 134cd).
+The first is taken from a description of \emph{padmāsana}, and may have been included by Svātmārāma in his description of \emph{mahābandha} to make it clear that the breath is held for as long as possible after the inhalation.
+The second hemistich is from a passage on \emph{mahāmudrā}, and it echoes a similar statement on \emph{mahāmudrā} in \emph{Haṭhapradīpikā} 3.15.
+It also occurs in the \textalpha\ group after 3.22.
+\end{philcomm}
+
+%%%%%%%%%%
+%\subsection*{3.21*1}
+%\begin{translation}[hp03_021*1]
+%After holding the breath for as long as possible, [the yogi] should exhale gradually. And after practising on the left side, he should practise on the right.
+%\end{translation}
+%\begin{testimonia}[hp03_021*1]
+%\end{testimonia}
+
+%\begin{philcomm}[hp03_021*1]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.22}
+\begin{translation}[hp03_022]
+With regard to this [practice] some are of the opinion that [the yogi] should leave out the throat lock, saying that he should lift up the hollow at the uvula with the tongue instead.
+\end{translation}
+
+\begin{sources}[hp03_022]
+Cf.~\emph{Vivekamārtaṇḍa} 126ab
+\begin{versinnote}
+\tl{saṃpīḍya rasanāgreṇa rājadantabilaṃ mahat |\\!}
+\end{versinnote}
+Cf.~\emph{Dattātreyayogaśāstra} 36
+\begin{versinnote}
+\tl{nāsāgre vinyased rājadantamūlaṃ ca jihvayā |\\+}
+\tl{uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_022]
+\emph{Yogacintāmaṇi} f.~73r (attr.~to Īśvara)% not in the \emph{Yogacintāmaṇi} spreadsheet, but in Muktabodha etext (in a passage attributed to īśvara)
+\begin{versinnote}
+\tl{matam atra tu keṣāṃ cit kaṇṭhabandhaṃ vivarjayet |\\+}
+\tl{rājadantadvayaṃ tatra jihvayottambhayed iti ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.192cd (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{rājadantadvayaṃ tatra jihvayonnamayed iti ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_022]
+This verse expresses an alternative to the application of the chin lock in \emph{mahā\-bandha} (mentioned in 3.20).
+It is found in all groups but is notably absent in \alphaOne\ and \alphaTwo, and the \textbeta\ group.
+At present we cannot be certain if this verse was written by Svātmārāma because it does not reflect textual teachings on \emph{mahābandha} in so far as no other text known to us advocates the use of the tongue rather than the chin lock in \emph{mahābandha}.
+If the verse was composed by Svātmārāma, it was omitted early in the transmission by someone who did not agree with the alternative teaching.
+
+This verse occurs in the \emph{Yogacintāmaṇi} in the middle of a quotation attributed to Īśvara.
+The other verses of the quoted passage are found in the \emph{Śivasaṃhitā} (4.37–42), but the verse in question is not reported in the critical edition of the \emph{Śivasaṃhitā} (2018).
+The verse is absent in another passage on \emph{mahābandha} that the author of the \emph{Yogacintāmaṇi} cites and attributes to the \emph{Haṭhapradīpikā}.
+
+We do not find the idea of lifting up the \emph{rājadantabila} with the tongue in other works, but \emph{Vivekamārtaṇḍa} 126ab instructs the yogi to press it with the tip of the tongue and \emph{Dattātreyayogaśāstra} 36 (found at \emph{Haṭhapradīpikā} 1.46) instructs the yogi in \emph{padmāsana} to lift up the `root of the uvula' (\emph{rājadantamūla}) with the tongue.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.23}
+\begin{translation}[hp03_023]
+This is truly the great lock: it bestows the great \emph{siddhi} [and] is adept at loosening the great bond (\emph{°mahābandha°}) that is the noose of time.
+\end{translation}
+
+%\begin{sources}[hp03_023]
+%\end{sources}
+
+\begin{testimonia}[hp03_023]
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ayaṃ khalu mahābandho mahāsiddhipradāyakaḥ |\\+}
+\tl{kālapāśamahābandhavimocanavicakṣaṇaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.193 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ayaṃ khalu mahābandho mahāsiddhipradāyakaḥ |\\+}
+\tl{kālapāśamahābandhavimocanavicakṣaṇaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_023]
+%J5 reading for this passage is different from alpha1 so should be reported.
+%MD: Read ayaṃ yogo mahābandho instead of ayaṃ khalu ...? Alpha has yoga (N3, acc.)/yogī (J5)/yogo (G4).
+%JM: Alpha yogamahābandhaṃ works. Adopt alpha one alternative verse?
+%(this was above) Adopt yogamahābandhaṃ in 3.23a (and report alpha readings). JM: khalu seems ok though. elsewhere it introduces a nirukti like this, and yoga is odd. I say go with khalu.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.24}
+\begin{translation}[hp03_024]
+And this [lock] initiates an upward flow in all the channels. It brings about a confluence at the Triveṇī [and] causes the mind to reach Kedāra.
+\end{translation}
+
+\begin{sources}[hp03_024]
+\emph{Amaraugha} 25
+\begin{versinnote}
+\tl{ayaṃ ca sarvanāḍīnām ūrdhvaṃgativibodhakaḥ |\\+}
+\tl{triveṇīsaṅgame dhatte kedāraṃ prāpayen manaḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_024]
+\emph{Haṭharatnāvalī} 2.46
+\begin{versinnote}
+\tl{ayaṃ ca sarvanāḍīnām ūrdhvagativibodhakaḥ |\\+}
+\tl{triveṇīsaṅgamaṃ dhatte kedāraṃ prāpayen manaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ayaṃ tu sarvanāḍīnām ūrdhvaṃgamanarodhakaḥ |\\+}
+\tl{triveṇīsaṃgamaṃ dhatte kedāraṃ prāpayen manaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.94 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ayaṃ hi sarvanāḍīnām ūrdhvaṃ gativibodhakaḥ |\\+}
+\tl{triveṇīsaṃgamaṃ dhatte kedāraṃ prāpayen manaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_024]
+The reading \emph{ūrdhvaṃgatibodhakaḥ} (`initiates an upward flow') is found in the source (\emph{Amaraugha} 25) and all \emph{Haṭhapradīpikā} witnesses except the \emph{Jyotsnā} (where the line is found earlier).
+While the \emph{Amaraughaprabodha} has the reading \emph{ūrdhvaṃgativiśodhanaḥ} (`purification of the upward flow'), the \emph{Amṛtasiddhi} (12.14) states that the chin-lock prevents the upward flow (\emph{ūrdhvaṃgatinirodhakaḥ}) in all the channels.
+The \textit{Amaraugha} is referring to the idea (likely accepted by Svātmārāma) that the root lock creates an upward flow in all the channels that prevents the elements and essences of the body from escaping (cf.~\emph{Amṛtasiddhi} 12.8–10).
+
+Triveṇī and Kedāra are pilgrimage sites, the former at Prayāga where the Gaṅgā and Yamunā meet, the latter in the Himālaya, near the source of the Gaṅgā.
+The bodily \emph{triveṇī} is located in the navel or heart by earlier Śaiva works (Birch 2019: 967).
+Here it may be the same as the \emph{trikūṭa} and located between the eyebrows (Mallinson 2007: 209 n.259).
+Brahmānanda, who does not identify a location for Triveṇī, understands Kedāra to be between the eyebrows.
+In the \emph{Khecarīvidyā} it is located on the back of the head above the nape of the neck (Mallinson 2007:214 n.285).
+For other references on the location of Kedāra, see Birch 2019:967 n.57.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.25}
+\begin{translation}[hp03_025]
+Now the great piercing (\emph{mahāvedha}):\\
+Like a beautiful and charming woman without a man, the great seal and the great lock are barren without the great piercing.\\
+\end{translation}
+
+\begin{sources}[hp03_025]
+\emph{Amaraugha} 26
+\begin{versinnote}
+\tl{rūpalāvaṇyasampannā yathā strī puruṣaṃ vinā |\\+}
+\tl{mahāmudrāmahābandhau niṣphalau vedhavarjitau  ||\\+}
+\tl{\var{°bandhau niṣphalau vedhavarjitau ] °bandho niṣphalo vedhavarjitaḥ \vl}\\!}
+\end{versinnote}
+
+Cf.~\emph{Śivasaṃhitā} 4.47
+\begin{versinnote}
+\tl{mahāmudrāmahābandhau niṣphalau vedhavarjitau |\\+}
+\tl{tasmād yogī prayatnena karoti tritayaṃ kramāt ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Amṛtasiddhi} 13.3
+\begin{versinnote}
+\tl{guṇarūpavatī nārī niṣphalā puruṣaṃ vinā |\\+}
+\tl{mahāmudrāmahābandhau vinā vedhena niṣphalau ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_025]
+\emph{Haṭharatnāvalī} 2.47
+\begin{versinnote}
+\tl{rūpalāvaṇyasampannā yathā strī puruṣaṃ vinā |\\+}
+\tl{mahāmudrāmahābandhau niṣphalau vedhavarjitau ||\\+}
+\tl{\var{°mahābandhau ] mahābandho \vl}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{atha mahāvedhaḥ—\\+}
+\tl{rūpalāvaṇyasaṃpannā yathā strī puruṣaṃ vinā |\\+}
+\tl{mahāmudrāmahābandhau niṣphalau vedhavarjitau ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.197 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{atha mahāvedhaḥ -\\+}
+\tl{mahāmudrāmahābandhau niṣphalau vedavarjitau |\\+}
+\tl{rūpalāvaṇyasampannā yathā strī puruṣaṃ vinā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_025]
+This verse, which is from the \emph{Amaraugha} and similar to verses in the \emph{Amṛtasiddhi} and \emph{Śivasaṃhitā}, is stating that the great seal, lock and piercing should be practised together.
+This can be done as a sequence as shown in this video.
+%??
+insert video here
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.26}
+\begin{translation}[hp03_026]
+While in the great lock, the yogi should inhale, focus his mind and firmly block the flow of the bodily winds by means of the throat seal.
+\end{translation}
+
+\begin{sources}[hp03_026]
+\emph{Amaraugha} 27
+\begin{versinnote}
+\tl{punar āsphālayet kaṭyāṃ susthiraṃ kaṇṭhamudrayā |\\+}
+\tl{vāyūnāṃ gatim ārudhya kṛtvā pūrakakumbhakau ||\\+}
+\tl{\var{ārudhya ] āvṛtya, āśritya \vl}\\!}
+\end{versinnote}
+
+Cf.~\emph{Śivasaṃhitā}  4.43
+\begin{versinnote}
+\tl{mahābandhasthito yogī kukṣim āpurya vāyunā |\\+}
+\tl{sphicau saṃtāpayed dhīmān vedho 'yaṃ kīrtito mayā ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_026]
+\emph{Haṭharatnāvalī} 2.48
+\begin{versinnote}
+\tl{mahābandhasthito yogī kṛtvā pūrakam ekadhīḥ |\\+}
+\tl{vāyūnāṃ gatim ākṛṣya nibhṛtaṃ kaṇṭhamudrayā ||\\+}
+\tl{\var{ekadhīḥ ] ekadhā \vl}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mahābandhe sthito yogī kṛtvā pūrakam ekadhā |\\+}
+\tl{vāyūnāṃ gatim āvṛtya nibhṛtaṃ kaṇṭhamudrayā ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.198 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mahābandhasthito yogī kṛtvā pūrakam ekadhā |\\+}
+\tl{vāyunāṃ gatim āvṛtya nibhṛtaṃ kaṇṭhamudrayā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_026]
+We are not certain of the meaning of \emph{nibhṛtam}, which is found in all of the collated witnesses and important testimonia. We have understood it as an adverb meaning `firmly.'
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.27}
+\begin{translation}[hp03_027]
+With hands even on the ground, he should gently tap the buttocks [on the ground]. The breath enters the vessel of two halves and quickly flashes forth.
+\end{translation}
+
+\begin{sources}[hp03_027]
+\emph{Amaraugha} 28ab, 29cd
+\begin{versinnote}
+\tl{samahastayugo bhūmau samapādayugas tathā |\\+}
+\tl{[vedhayet kramayogena catuṣpīṭhaṃ tu vāyunā ||\\+}
+\tl{āsphālayen mahāmeruṃ vāyuvajrāgnikoṭibhiḥ | ]\\+}
+\tl{puṭadvayaṃ samākṛṣya vāyuḥ sphurati satvaram ||\\+}
+\tl{\var{samākṛṣya ] samākramya \vl }\\!}
+\end{versinnote}
+
+\emph{Dattātreyayogaśāstra} 136ab
+\begin{versinnote}
+\tl{mahābandhasthito bhūmau sphicau santāḍayec chanaiḥ |\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_027]
+\emph{Haṭharatnāvalī} 2.51
+\begin{versinnote}
+\tl{samahastayugo bhūmau sphicau saṃtāḍayec chanaiḥ |\\+}
+\tl{ayam eva mahāvedhaḥ siddhido 'bhyāsato bhavet \\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{samahastayugo bhūmau sphijau saṃtāḍayec chanaiḥ |\\+}
+\tl{puṭadvayaṃ samākramya vāyuḥ sphurati madhyagaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavedeva} 7.199 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{nyastahastayugo bhūmau sphicau santāḍayec chanaiḥ |\\+}
+\tl{jaṃghādvayaṃ samākṛṣya vāyuḥ spharati satvaram ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_027]
+The term \emph{puṭadvaya} is referring back to the \emph{dvipuṭa} mentioned in verse 3.12. On the alchemical meaning and other interpretations by later commentators, see the note to 3.12.
+
+The greyscale hemistich added to this verse in the \textgamma\ group is taken from the \emph{Śivasaṃhitā} (4.42ab) in the section on \emph{mahābandha}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.28}
+\begin{translation}[hp03_028]
+And the union of the moon, sun and fire arises, which leads to immortality.
+The state of a dead person has arisen, so where is the fear of death?
+\end{translation}
+
+\begin{sources}[hp03_028]
+\emph{Amaraugha} 30
+\begin{versinnote}
+\tl{somasūryāgnisaṃbandhaṃ jānīyād amṛtāya vai |\\+}
+\tl{mṛtāvasthā samutpannā tato mṛtyubhayaṃ kutaḥ ||\\+}
+\tl{\var{°saṃbandhaṃ ] °saṃbandhāj, °saṃbandhā \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_028]
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{somasūryāgnisandhānaṃ jāyate cāmṛtāyate |\\+}
+\tl{mṛtāvasthāsamutpannaṃ tato mṛtyubhayaṃ kutaḥ ||\\+}
+\tl{\var{°samutpannaṃ ] °samutpannā \vl}\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.200 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{somasūryāgnisambandho jāyate cāmṛtāya ca |\\+}
+\tl{samutpannā mṛtāvasthā tato vāyuṃ virecayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_028]
+The \textalpha\ and \textgamma\ groups have °\emph{saṃbandhāj}, which is possible but somewhat awkward (i.e., `because of the union of the moon, sun and fire, the state of a dead person, which has arisen, leads to immortality').
+The adopted reading °\emph{saṃbandho}, which is supported by \epsilonTwo, \emph{Yuktabhavadeva} 7.200 and \emph{Jyotsna} 3.28, makes better sense but its meaning is not as clear as the formulation in the source text (i.e., \emph{Amaraugha} 30).
+
+The compound \emph{mṛtāvasthā} (`the state of death') likely refers to a dead person in the sense that the yogi appears as if dead when the moon, sun and fire have united.
+In the version found in \emph{Jyotsnā} 3.28 and \etaOne, \etaTwo\ and \epsilonThree, the final verse quarter has been rewritten to say that the yogi then exhales the breath (\emph{tato vāyuṃ virecayet}).
+This implies that \emph{mṛtāvasthā} is a breath retention (\emph{kumhaka}), which is apparent in Brahmānanda's explanation:
+\begin{quote}
+The state of death that has arisen is the state of one who has died, [that is,] of one who is devoid of the breath because of the absence of movement of \emph{prāṇa} in the \emph{iḍā} and \emph{piṅgalā} channels.
+Then, immediately after that [state], [the yogi] exhales the breath, [that is,] he gradually releases it through the nostrils.\\
+\emph{mṛtasya prāṇaviyuktasyāvasthā mṛtāvasthā samutpannā bhavati, iḍāpiṅgalayoḥ prāṇasañcārābhāvāt} | \emph{tatas tadanantaraṃ vāyuṃ virecayen nāsikāpuṭābhyāṃ śanais tyajet} |
+\end{quote}
+
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.29}
+\begin{translation}[hp03_029]
+Through practice, this great piercing bestows the great \emph{siddhi} [and] cures wrinkles, grey hair and trembling. It is used by the best practitioners.
+\end{translation}
+
+%\begin{sources}[hp03_029]
+%\end{sources}
+
+\begin{testimonia}[hp03_029]
+\emph{Haṭharatnāvalī} 2.51cd
+\begin{versinnote}
+\tl{ayam eva mahāvedhaḥ siddhido 'bhyāsato bhavet ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mahāvedho 'yam abhyāsān mahāsiddhipradāyakaḥ |\\+}
+\tl{valīpalitavepaghnaḥ sevyate sādhakottamaiḥ ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.201 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mahāvedho 'yam abhyasto mahāsiddhipradāyakaḥ |\\+}
+\tl{valīpalitavepaghnaḥ sevyate sādhakottamaiḥ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_029]
+In the third verse quarter, the reading \emph{°vedhaghnaḥ} (`cures wounds' ?) in \textalpha\ and the other groups, the exception being \textgamma\ (\emph{°vegaghnaḥ}), is strange as it does not seem related to the other two symptoms of old age (i.e., wrinkles and grey hair) that this \emph{mudrā} can cure.
+It is perhaps a misreading of \emph{°vepaghnaḥ}, which occurs in some manuscripts in the \textbeta\ (\getsiglum{J1}, \getsiglum{N12}), \textepsilon\ (\getsiglum{J14}) and \textdelta\ (\getsiglum{J3}, \getsiglum{N16}, \getsiglum{N18}) groups, as well as the \emph{Yogacintāmaṇi} and \emph{Jyotsnā}.
+The meaning of \emph{vepa}, `trembling,' is consistent with wrinkles and grey hair.
+
+%The collocation \emph{valīpalitaroga} that we have adopted, found in \getsiglum{C8}, is quite common (e.g. \emph{Śāradātilaka} \emph{6.104, Ānandakanda} 1.7.13, .1.7.31).
+%JB: I think we should adopt vepa on the strength of the marmasthāna entry (\item[valīpalitavepaghnaḥ] Ba5, J1, J3, J6, J14, N4a, N11, N12, N16, N18, ), and the fact that it's closer to vedha (alpha reading) than roga (which looks like an outlier in the marmasthāna data).
+% J1=6b, J3,J6,N16,N18=6a, J14=4c, N11 =Gr2, N12 =4b,
+% MD: These mss are not very strong.
+Is it not possible to interpret °vedhaghnaḥ as "destroys by piercing"?
+This verse only has no source in the middle of the long citation from Amaraugha, is probably authorial.
+Svātmārāma might have wanted to use the word vedha of mahāvedha, similar to dhauti/dhāvanti.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.30}
+\begin{translation}[hp03_030]
+This triad is a great secret that destroys old and death, increases [the body's] fire and bestows the powers beginning with minimisation.
+\end{translation}
+
+\begin{sources}[hp03_030]
+\emph{Amaraugha} 31
+\begin{versinnote}
+\tl{etat trayaṃ mahāguhyaṃ jarāmṛtyuvināśanam |\\+}
+\tl{vahnivṛddhikarañ caiva aṇimādiguṇapradam ||\\+}
+\tl{\var{°guhyaṃ ] °guṇyaṃ \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_030]
+\emph{Haṭharatnāvalī} 2.52
+\begin{versinnote}
+\tl{etat trayaṃ mahāguhyaṃ jarāmṛtyuvināśanam |\\+}
+\tl{vahnivṛddhikaraṃ caiva hy aṇimādiguṇapradam ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{etat trayaṃ mahāguptaṃ jalāmṛtyuvināśanam |\\+}
+\tl{vahnivṛddhikaraṃ caiva aṇimādiguṇapradam ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.204 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{bandhatrayaṃ mahāguhyaṃ jarāmṛtyuvināśanam |\\+}
+\tl{vahnivṛddhikaraṃ caiva aṇimādiguṇapradam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_030]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.31}
+\begin{translation}[hp03_031]
+It is practised eight times a day, every three hours. It always produces a wealth of merit and destroys an ocean of demerit.
+\end{translation}
+
+\begin{sources}[hp03_031]
+\emph{Amaraugha} 32
+\begin{versinnote}
+\tl{aṣṭadhā kriyate caiva yāme yāme dine dine |\\+}
+\tl{puṇyasañcayasambhāvi pāpaughabhiduraṃ sadā ||\\+}
+\tl{\var{caiva ] caitad \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_031]
+\emph{Haṭharatnāvalī} 2.49
+\begin{versinnote}
+\tl{aṣṭadhā kriyate caitat yāme yāme dine dine |\\+}
+\tl{puṇyasaṅghātasandhāyī pāpaughabhiduraḥ sadā ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{aṣṭadhā kriyate caiva yāme yāme dine dine |\\+}
+\tl{puṇyasaṃbhārasandhāyi pāpaughabhiduraṃ sadā ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.205 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{aṣṭadhā kriyate caiva yāme yāme dine dine |\\+}
+\tl{puṇyasambhārasandhāyi pāpaughabhiduraṃ sadā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_031]
+%JB: It seems to make better sense to read this as saying that the yogi practises the three seals `eight times, every three hours, every day.' This would mean the triad is practised sixty-four times in total a day (commesurate with the eighty kumbhakas per day).
+not accepted at the workshop by Haru and Lubomir.
+(In a post reading session discussion, Haru thinks the interpretation of 64 times a day is possible)
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.32}
+\begin{translation}[hp03_032]
+It is only for those who have received proper instruction.
+It is taught that they should [do it] a little in the first stage of the practice [and] that at the beginning [the yogi] should avoid frequenting fire, women and roads.
+\end{translation}
+
+\begin{sources}[hp03_032]
+\emph{Amaraugha} 33
+\begin{versinnote}
+\tl{samyakśikṣāvatām eva svalpaṃ prathamasādhane |\\+}
+\tl{vahnistrīpathasevānām ādau varjanam ācaret ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_032]
+\emph{Yogacintāmaṇi} f.~73v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{samyakśikṣāvatām eva svalpaṃ prathamasādhane ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.205 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{samyak śikṣāvatām evaṃ svalpaṃ prathamasādhanam |\\+}
+\tl{vahnistrīpathisevānām ādau varjanam ācaret ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_032]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.33}
+\begin{translation}[hp03_033]
+Now, the sky-roving [seal] (\emph{khecarī}):\\
+By cutting, moving and milking, the yogi should gradually lengthen the tongue until it touches the middle of the brows. Then there is success in \emph{khecarī}.
+\end{translation}
+
+%\begin{sources}[hp03_033]
+%\end{sources}
+
+\begin{testimonia}[hp03_033]
+\emph{Haṭharatnāvalī} 2.141
+\begin{versinnote}
+\tl{haṭhapradīpikāyām ||\\+}
+\tl{chedanacālanadohaiḥ kalāṃ krameṇa vardhayet tāvat |\\+}
+\tl{yāvad iyaṃ bhrūmadhye spṛśati tadānīṃ khecarīsiddhiḥ ||\\+}
+\tl{\var{yāvad iyaṃ bhrūmadhye spṛśati ] sā yāti yāvad bhrūmadhyaṃ spṛśati hi \vl}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~74r
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{chedanacālanadohair jihvāṃ saṃvardhayet tāvat |\\+}
+\tl{yāvad iyaṃ bhrūmadhyaṃ spṛśati tadā khecarī siddhiḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 14.18
+\begin{versinnote}
+\tl{haṭhapradīpikāmate tu –\\+}
+\tl{chedanacālanadohaiḥ krameṇa jihvāṃ pravardhayet tāvat |\\+}
+\tl{sā yāvad bhrūmadhyaṃ spṛśati tadā khecarīsiddhiḥ || \\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_033]
+%There are no \emph{āryā} or \emph{gīti} metres elsewhere in the text, but there are verses  in \emph{upagīti} metre (e.g.
+1.60).
+We have therefore adopted variants that fit the \emph{upagīti} metre.
+Perhaps this is authorial, as we don't know a source for it, and we don't know a source for 1.60.
+% JB: this is a weak reason given the readings of alpha!
+
+Various versions of this verse have been transmitted in \emph{āryā}, \emph{gīti} and \emph{anuṣṭubh} metres. We have adopted a version close to \alphaTwo, which has a slight metrical fault:
+\begin{quote}
+ chedanacālanadohaiḥ kalāṃ krameṇa pravardhayet tāvat |\\
+ sā yāvad bhrūmadhyaṃ viśati tadānīṃ khecarīsiddhiḥ ||
+\end{quote}
+% alpha2
+% chedanacālanaṃ dohaiḥ kalā kramaṇa pravardhaye tāvat
+% sā yāvad bhrumadhyaṃ visati tadānī khecarīsiddhiḥ
+The emendation of \emph{tadānīṃ} to \emph{tadā} renders the verse an \emph{āryā}.
+The word \emph{krameṇa} is well attested by manuscripts of the \textalpha, \textbeta, \texteta\ and  \textepsilon\ groups, which all have unmetrical or corrupted versions.
+% krameṇa is well attested
+% close to alpha 2
+% damaged in alpha 1: [damaged]ḥ kalāḥ krameṇa varddayet | sā yāva[d] bhṛmadhyaṃ viśa tadānīṃ khecarīsiddhiḥ || but it looks like someone has unsuccessfully tried to make it anuṣṭubh.
+Perhaps, suggests that a longer metre was original.
+
+The meaning of \emph{kalā} as `tongue' is not attested by any Sanskrit dictionary but \emph{kalā} occurs in the sense of the tongue in a subsequent verse of this chapter (Cf.~3.35a) and it is glossed by Brahmānanda with \emph{jihvā} in \emph{Jyotsnā} 3.33 and 3.37.
+%
+\end{philcomm}
+
+\begin{metre}[hp03_033]
+Upagīti 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.34}
+\begin{translation}[hp03_034]
+When the tongue is turned back and inserted into the cavity of the skull and the gaze is between the brows, \emph{khecarīmudrā} arises.
+\end{translation}
+
+\begin{sources}[hp03_034]
+\emph{Vivekamārtaṇḍa} 47
+\begin{versinnote}
+\tl{kapālakuhare jihvā praviṣṭā viparītagā |\\+}
+\tl{bhruvor antargatā dṛṣṭir mudrā bhavati khecarī ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_034]
+\emph{Haṭharatnāvalī} 2.138
+\begin{versinnote}
+\tl{dattātreyas tu ||\\+}
+\tl{kapālakuhare jihvā praviṣṭā viparītagā |\\+}
+\tl{bhruvor antargatā dṛṣṭir mudrā bhavati khecarī ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~75r
+\begin{versinnote}
+\tl{skandapurāṇe—\\+}
+\tl{kapālakuhare jihvā praviṣṭā viparītagā |\\+}
+\tl{bhruvor antargatā dṛṣṭir mudrā bhavati khecarī ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.207 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{atha khecarī -\\+}
+\tl{kapālakuhare jihvā praviṣṭā viparītagā |\\+}
+\tl{bhruvor antargatā dṛṣṭir mudrā bhavati khecarī ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_034]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.35}
+\begin{translation}[hp03_035]
+If the yogi turns back the tongue and remains [like that] for half an instant, he is instantly freed from disease, death, old age and the like.
+\end{translation}
+
+\begin{sources}[hp03_035]
+\emph{Śivasaṃhitā} 3.91
+\begin{versinnote}
+\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
+\tl{kṣaṇena mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_035]
+\emph{Yogacintāmaṇi} f.~74r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{kalāṃ parāṅmukhīṃ kṛtvā tripathe parivartayet |\\+}
+\tl{sā bhavet khecarī mudrā vyomacakraṃ tad ucyate ||\\+}
+\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
+\tl{viṣayair mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.209 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{jihvāṃ parāṅmukhīṃ kṛtvā kṣaṇārddhaṃ yadi tiṣṭhati |\\+}
+\tl{kṣaṇena mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_035]
+We have adopted the version of 2.35 in \alphaTwo\ and \alphaThree\ (\alphaOne\ omits 2.35 and the greyscaled verses).
+This version is also attested by the \textgamma\ and \texteta\ groups.
+The additional hemistich gives \emph{vyomacakra} as an alternative name for \emph{khecarīmudrā}.
+This alternative name does not occur in any of the source texts known to be have been used by Svātmārāma, but it may have been inspired by the name \emph{nabhomudrā} for \emph{khecarīmudrā} in the \emph{Vivekamārtaṇḍa} (40).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.36}
+\begin{translation}[hp03_036]
+For the yogi who knows \emph{khecarīmudrā} there is no disease, death, sleep, hunger, thirst or fainting.
+\end{translation}
+
+\begin{sources}[hp03_036]
+\emph{Vivekamārtaṇḍa} 48
+\begin{versinnote}
+\tl{na rogo maraṇaṃ tandrā na nidrā na kṣudhā tṛṣā |\\+}
+\tl{na ca mūrchā bhavet tasya yo mudrāṃ vetti khecarīm ||\\+}
+\tl{\var{tandrā ] tasya \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_036]
+\emph{Haṭharatnāvalī} 2.139 (attr.~to Dattātreya)
+\begin{versinnote}
+\tl{na rogo maraṇaṃ tasya na nidrā na kṣudhā tṛṣā |\\+}
+\tl{na ca mūrcchā bhavet tasya yo mudrāṃ vetti khecarīm ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~75v (attr.~to Dattātreya)
+\begin{versinnote}
+\tl{na rogo maraṇaṃ tasya na nidrā na kṣudhā tṛṣā |\\+}
+\tl{na ca mūrcchā bhavet tasya yo mudrāṃ vetti khecarīm ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.210 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{na rogo maraṇaṃ tasya na nidrā na tṛṣā kṣudhā |\\+}
+\tl{na ca mūrcchā bhavet tasya yo mudrāṃ vetti khecarīm ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_036]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.37}
+\begin{translation}[hp03_037]
+[The yogi] who knows \emph{khecarīmudrā} is neither afflicted by disease, nor tainted by action, nor tormented by death.
+\end{translation}
+
+\begin{sources}[hp03_037]
+\emph{Vivekamārtaṇḍa} 49
+\begin{versinnote}
+\tl{pīḍyate na sa rogeṇa lipyate na ca karmaṇā |\\+}
+\tl{bādhyate na sa kālena yo mudrāṃ vetti khecarīṃ ||\\!}
+%JB: Retain variants for lipyate na ca as alpha reads na ca lipyata/lipyati
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_037]
+\emph{Haṭharatnāvalī} 2.140 (attr.~to Dattātreya)
+\begin{versinnote}
+\tl{pīḍyate na sa rogeṇa lipyate na ca karmaṇā|\\+}
+\tl{bādhyate na ca kālena yo mudrāṃ vetti khecarīm||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~75v (attr.~to the \emph{Skandapurāṇa})
+\begin{versinnote}
+\tl{piḍyate na sa rogeṇa na ca lipyeta karmaṇā |\\+}
+\tl{bādhyate na sa kālena yo mudrāṃ vetti khecarīm ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.211 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pīḍyate na sa rogeṇa lipyate na sa karmaṇā |\\+}
+\tl{bādhyate na sa kālena yasya mudrāsti khecarī ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_037]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.38}
+\begin{translation}[hp03_038]
+Because the mind moves (\emph{carati}) in the ether (\emph{khe}) and the tongue moves (\emph{carati}) in the cavity (\emph{khe}), this mudrā is called \emph{khecarī} [and] is worshipped by the Siddhas.
+\end{translation}
+
+\begin{sources}[hp03_038]
+\emph{Vivekamārtaṇḍa} 50
+\begin{versinnote}
+\tl{cittaṃ carati khe yasmāj  jihvā carati khe gatā |\\+}
+\tl{tenaiṣā  khecarī nāma mudrā siddhair namaskṛtā ||\\+}
+\tl{\var{37c tenaiṣā ] tenaiva, teneyaṃ \vl 37cd nāma mudrā ] mudrā sarva° \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_038]
+\emph{Yogacintāmaṇi} (attributed to the Skandapurāṇa)
+\begin{versinnote}
+\tl{cittaṃ carati khe yasmāj jihvā carati khe gatā |\\+}
+\tl{tenaiṣā khecarī nāma mudrā siddhair niṣevitā ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.212 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{cittaṃ carati khe yasmāj jihvā carati khe yataḥ |\\+}
+\tl{teneyaṃ khecarī mudrā sarvasiddhair namaskṛtā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_038]
+%Judit: this mudrā is worshipped by the siddhas as the one called khecarī. (the idea being that this verse is explaining the name, khe + carī.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.39}
+\begin{translation}[hp03_039]
+The yogi who has sealed the cavity above the uvula with \emph{khecarī} does not lose his semen [even if] embraced by an amorous woman.
+\end{translation}
+
+\begin{sources}[hp03_039]
+\emph{Vivekamārtaṇḍa} 51
+\begin{versinnote}
+\tl{khecaryā mudritaṃ yena vivaraṃ lambikordhvataḥ |\\+}
+\tl{na tasya kṣarate binduḥ kāminyāliṅgitasya ca ||\\+}
+\tl{\var{na tasya kṣarate binduḥ ] binduḥ kṣarati no tasya, tasya na kṣarate binduḥ \vl}\\+}
+\tl{\var{°āliṃgitasya ] °āśleṣitasya \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_039]
+\emph{Yogacintāmaṇi} f.~74v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{khecaryā mudritaṃ yena vivaraṃ lambikordhvataḥ |\\+}
+\tl{na tasya kṣarate binduḥ kāminyāliṅgitasya ca ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.213 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{khecaryā mudritaṃ yena vivaraṃ lambikordhvataḥ |\\+}
+\tl{na tasya kṣarate binduḥ kāminyāliṃgitasya ca ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_039]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.40}
+\begin{translation}[hp03_040]
+Even when semen has moved [down] and reached the region of the perineum, it moves upwards having been blocked by \emph{yonimudrā} and struck by the goddess [\emph{kuṇḍalinī}].
+\end{translation}
+
+\begin{sources}[hp03_040]
+\emph{Vivekamārtaṇḍa} 53
+\begin{versinnote}
+\tl{calito 'pi yadā binduḥ saṃprāptaś ca hutāśanam |\\+}
+\tl{gacchaty ūrdhvaṃ hataḥ śaktyā nibaddho yonimudrayā ||\\+}
+
+\tl{53c \var{gacchaty ] vrajaty \vl 53d hataḥ ] tanaṃ, hṛtas, kṛte, kṛtaḥ, tadā, gatā \vl • nibaddho ] niruddho \vl}\\!}
+\end{versinnote}
+
+Cf.~\emph{Śivasaṃhitā} 4.82
+\begin{versinnote}
+\tl{svakaṃ binduṃ ca saṃbodhya liṅgacālanam ācaret |\\+}
+\tl{daivāc calati ced ūrdhvaṃ nibaddho yonimudrayā ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_040]
+\emph{Yogacintāmaṇi} f.~74v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{calito 'pi yadā binduḥ saṃprāptaś ca hutāśanam |\\+}
+\tl{vrajaty ūrdhvaṃ hi tacchaktyā niruddho yonimudrayā ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.278 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{calito'pi mahābinduḥ samprāpte'pi hutāśanam |\\+}
+\tl{vrajaty ūrdhvaṃ haṭhaḥ śaktyā nibaddho yonimudrayā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_040]
+% JB Perhaps we should adopt \emph{yonimaṇḍalam} in 3.39b?
+It is attested by the alpha group (N3 and J5) as well as V1 and V15.
+It makes good sense as `region of perineum', as understood by Brahmānanda in \emph{Jyotsnā} 3.43 (\emph{yonimaṇḍalam yonisthānam}).
+
+The third quarter of this verse has been subjected to much rewriting.
+Most of the collated manuscripts, including \alphaTwo\ (\alphaOne\ is illegible here and \alphaThree\ is missing this verse quarter), have the reading \emph{haṭhāt śaktyā}.
+This reading only makes sense if one infers that \emph{śaktyā} is referring to \emph{khecarīmudrā}, so that the second line means `blocked by \emph{yonimudrā}, semen goes up forcefully by the power [of \emph{khecarīmudrā}].' The reading \emph{hi tacchaktyā} of two delta manuscripts (\getsiglum{K3} and \getsiglum{C7}) and the \emph{Yogacintāmaṇi}), appears to be an attempt to render more clearly the meaning `by the power of \emph{khecarī}.' Such an interpretation suggests that \emph{yonimudrā} blocks \emph{bindu}'s downward course and \emph{khecarī} causes it to go upwards forcefully.
+
+A manuscript of an early recension of the \emph{Vivekamārtaṇḍa} (ms.~no.~4110) has \emph{hataḥ śaktyā}, which is attested by three \emph{Haṭhapradīpikā} manuscripts on lower branches of the stemma (i.e.~\getsiglum{Ba} of \textdelta\ group and \getsiglum{C2} and \getsiglum{P4} of group 7a ??).
+The participle \emph{haṭaḥ} makes sense of the instrumental \emph{śaktyā}, rendering the meaning `struck by \emph{kuṇḍalinī}.'
+% MD: These groups don't have a Greek sigla yet.
+
+Alternatively, the word \emph{haṭaḥ} (as well as the other variants \emph{kṛtaḥ, kṛte}, and even \emph{haṭhāt}) may derive from \emph{hṛtaḥ}, which is attested by \emph{Haṭhapradīpikā} manuscripts from group 4a (i.e., \getsiglum{Ba5}, \getsiglum{N4a} and \getsiglum{N13}).
+The reading \emph{hṛtaḥ śaktyā} renders the verse as saying that semen goes up, carried by \emph{kuṇḍalinī}.
+
+%Adopt hataḥ śaktyā, which is supported by C2,  and the good manuscripts of the \emph{Vivekamārtaṇḍa}.
+%[MD: Text changed. C2 reads haṭ(h)aḥ, J5 haṭhāt saktyā, G4 damaged]
+%
+% JB haṭhāt is very well attested (J5, J7, Gr4c, V1, C6, V3,J10).
+Perhaps, we should reconsider this.
+Since the context is khecarīmudrā, it makes sense to infer it with śaktyā and perhaps haṭhāt fits with Svātmārāma's agenda to foreground Haṭhayoga?
+
+%But Judit/Philipp think hṛtaḥ is better and explains all the variations (kṛtaḥ, kṛte, etc. and even haṭha). So, Kuṇḍalinī carries semen upwards.
+
+%Adopt nibaddho (J5) "bound/stopped by yonimudrā". [MD: Text changed.]
+
+% write a note on yonimudrā
+In \emph{Jyotsnā} 3.43, Brahmānanda explains `\emph{yonimudrā}' as `taking the form of contracting the penis' (\emph{yonimudrayā meḍhrākuñcanarūpayā}).
+He may have had in mind the practice of contracting and drawing the urethra upwards, which is described below in the section on \emph{vajrolimudrā} (\emph{Haṭhapradīpikā} 3.82).
+The author of the \emph{Yogaprakāśikā} (5.66) says that \emph{yonimudrā} is well-known in treatises on \emph{mantra} (\emph{yonimudrayeti mantraśāstraprasiddhayety arthaḥ} | \emph{°prasidhyayety} ed.).
+This is consistent with the \emph{Śivasaṃhitā}'s teaching on \emph{yonimudrā}, where it is described as activating the perineum (\emph{yoni}) by contracting it (4.2) to bring about success in mantra repetition (5.12).
+Later compendiums on yoga reiterate the role of \emph{yonimudrā} in mantra practice (e.g., the \emph{Yogacintāmaṇi} f.~65r, citing the \emph{Pārameśvaratantra}, and \emph{Haṭhatattvakaumudī} 33.12).
+The \emph{Haṭhayogasaṃhitā} (43–48) teaches a different version of \emph{yonimudrā} in its reportoire of twenty-five \emph{mudrā}s.
+In this work, \emph{yonimudrā} is supposed to awaken \emph{kuṇḍalinī} and consists of sitting in \emph{siddhasana}, blocking the ears, eyes, nose and mouth with the thumbs, index, middle and ring fingers, uniting \emph{praṇa} and \emph{apāna}, meditating on the six \emph{cakra}s and repeating the mantra \emph{huṃ haṃsa}.
+
+Manuscripts of the main groups, including \textalpha, \textbeta\ and \textgamma, have an additional hemistich after 3.40, which seems to say that the tongue is in the cavity of the skull by means of a \emph{mudrā} for uniting the \emph{kalā}s (\emph{kapālakuhare jihvā kalāsandhānamudrayā}).
+This hemistich likely derives from a marginal note that was explaining \emph{ūrdhvajihvaḥ} in the next verse.
+The compound \emph{kalāsandhānamudrayā} may have been added as some form of dittography or as a gloss on \emph{yonimudrā}, which is not described elsewhere in the text.
+In a slightly modified form, this line appears in a verse in the six-chapter version of the \emph{Haṭhapradīpikā} (f.
+112r–112v):
+
+\begin{versinnote}
+\tl{kapālakuhare jihvā kalāsaṃdhānavarjitā |\\+}
+\tl{brahmarandhragatā nityāṃ tasya siddhi na dūrataḥ ||\\!}
+\end{versinnote}
+
+%RW: C2 includes greyscale 3.39 and adds another pāda; then skips 3.40 to get to 3.41 (MD: old numbering)
+%The additional pāda is: tasmād idaṃ prakuvīrta nitya yuktuḥ samāhitaḥ
+%It is difficult to understand the first hemistich without the second (and the second is not attested in any HP witnesses).
+%Judit: kalāsaṃdhānamudrayā may have been added as some form of dittography
+%JB: maybe this line was a marginal note attempting to explain yonimudrā, which is not described elsewhere in the text (?).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.41}
+\begin{translation}[hp03_041]
+The knower of yoga who remains with the tongue upwards and drinks nectar certainly conquers death in half a month.
+\end{translation}
+
+\begin{sources}[hp03_041]
+\emph{Vivekamārtaṇḍa} 125
+\begin{versinnote}
+\tl{ūrdhvajihvaḥ sthito bhūtvā somapānaṃ karoti yaḥ |\\+}
+\tl{māsārdhena na sandeho mṛtyuṃ jayati yogavit ||\\+}
+\tl{\var{ūrdhvajihvaḥ sthito bhūtvā ] ūrdhvajihvas tato bhūtvā, ūrdhvaṃ jihvāṃ sthiraṃ kṛtvā, ūrdhvāṃ jihvā sthirāṃ kṛtvā \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_041]
+\emph{Yogacintāmaṇi} f.~75v (attr.~to the \emph{Skandapurāṇa})
+\begin{versinnote}
+\tl{ūrdhvajihvaḥ sthiro bhutvā somapānaṃ karoti yaḥ |\\+}
+\tl{māsārdheṇa na saṃdeho mṛtyuṃ jayati yogavid ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.215 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ūrdhvajihvaḥ sa medhāvī somapānaṃ karoti yaḥ |\\+}
+\tl{māsārddhena na sandeho mṛtyuṃ jayati yogavit ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_041]
+%Alpha and \emph{Vivekamārtaṇḍa} read sthito. Adopt
+%MD: Done.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.42}
+\begin{translation}[hp03_042]
+Poison does not enter the yogi whose body is always filled by [nectar from] the digits of the moon, even if he is bitten by Takṣaka.
+\end{translation}
+
+\begin{sources}[hp03_042]
+\emph{Vivekamārtaṇḍa} 130
+\begin{versinnote}
+\tl{nityaṃ somakalāpūrṇaṃ śarīraṃ yasya yoginaḥ |\\+}
+\tl{takṣakenāpi daṣṭasya viṣaṃ tasya na pīḍayet ||\\+}
+\tl{\var{pīḍayet ] pīḍyate, sarpati, bādhyate \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_042]
+\emph{Yogacintāmaṇi} f.~75v (attr.~to the \emph{Skandapurāṇa})
+\begin{versinnote}
+\tl{nityaṃ somakalāpūrṇaṃ śarīraṃ yasya yoginaḥ |\\+}
+\tl{takṣakenāpi daṣṭasya viṣaṃ taṃ na ca sarpati ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.216 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{nityaṃ somakalāpūrṇaṃ śarīraṃ yasya yoginaḥ |\\+}
+\tl{takṣakenāpi daṣṭasya viṣaṃ tasya na sarpati ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_042]
+In the context of poison, \emph{takṣaka} refers to one of the three kings of the snakes (\emph{nāga}), the other two being Śeṣa and Vāsuki (Mani 1975: 782–783).
+% Purāṇic Encyclopaedia: A Comprehensive Dictionary with Special Reference to the Epic and Purāṇic Literature.
+Vettam Mani.
+Delhi: Motilal Banarsidass.
+
+The verb \emph{sarpati} is well attested by the manuscripts of the \emph{Haṭhapradīpikā} and testimonia. It can take an object, which in this case is the yogi's body.
+%At least two alpha mss have this verse. Therefore, keep it.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.43}
+\begin{translation}[hp03_043]
+Just as fire does not leave its fuel nor light a wick in oil,
+so the embodied person does not leave a body filled by the digits of the moon.
+\end{translation}
+
+\begin{sources}[hp03_043]
+\emph{Vivekamārtaṇḍa} 131
+\begin{versinnote}
+\tl{indhanāni yathā vahnis tailavartiṃ ca dīpakaḥ |\\+}
+\tl{tathā somakalāpūrṇaṃ dehī dehaṃ na muñcati ||\\+}
+\tl{\var{°vartiṃ ca ] °vartīva, °vartti ca \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_043]
+\emph{Yogacintāmaṇi} f.~74v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{indhanāni yathā vahnis tailavartī ca dīpakaḥ |\\+}
+\tl{nityaṃ somakalāpūrṇaṃ dehī dehaṃ na muñcati ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_043]
+%3.42.1 is not found elsewhere (?). 3.42.2 is in the 6ch HP, but nowhere else.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.44}
+\begin{translation}[hp03_044]
+I consider he who regularly eats cow flesh and drinks God's liquor to be of good family. Others are destroyers of the family.
+\end{translation}
+
+%\begin{sources}[hp03_044]
+%\end{sources}
+
+\begin{testimonia}[hp03_044]
+\emph{Haṭharatnāvalī} 2.158
+\begin{versinnote}
+\tl{gomāṃsaṃ bhakṣayen nityaṃ pibed amaravāruṇīṃ |\\+}
+\tl{kulīnaṃ tam ahaṃ manye anye tu kulaghātakāḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~74v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{gomāṃsaṃ bhakṣayen nityaṃ pibed amaravāruṇīm |\\+}
+\tl{kulīnaṃ tam ahaṃ manye netarān kulaghātakān ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_044]
+%Brahmānanda identifies this (and next two verses?) as compositions of Svātmārāma, presumably because of \emph{manye}.
+[JB: I don't think this note is worth mentioning as it is not relevant to an editorial decision]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.45}
+\begin{translation}[hp03_045]
+By the word 'cow' is meant the tongue, for its insertion into the palate is the eating of cow's flesh. That [eating] is the destroyer of great sin.
+\end{translation}
+
+%\begin{sources}[hp03_045]
+%\end{sources}
+
+\begin{testimonia}[hp03_045]
+\emph{Haṭharatnāvalī} 2.157
+\begin{versinnote}
+\tl{gośabdenoditā jihvā tatpraveśo hi tāluni |\\+}
+\tl{gomāṃsabhakṣaṇaṃ tat tu mahāpātakanāśanaṃ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~74v–75r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{gośabdenoditā jihvā tatpraveśo hi tāluni |\\+}
+\tl{gomāṃsabhakṣaṇaṃ tat tu mahāpātakanāśanam ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_045]
+%44cd is in N3 and J5, so retain.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.46}
+\begin{translation}[hp03_046]
+The essence produced by the fire caused by the insertion of the tongue which flows from the moon is the liquor of the gods.
+\end{translation}
+
+%\begin{sources}[hp03_046]
+%\end{sources}
+
+\begin{testimonia}[hp03_046]
+\emph{Haṭharatnāvalī} 2.159
+
+\begin{versinnote}
+\tl{jihvāpraveśasaṃbhūtavahninotthāpitā khalu |\\+}
+\tl{candrāt sravati yaḥ sāraḥ sā syād amaravāruṇī ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~75r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{jihvāpraveśasaṃbhūtavahninotpāditaḥ khalu |\\+}
+\tl{candrāt sravati yaḥ sāraḥ sā syād amaravāruṇī ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_046]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.47}
+\begin{translation}[hp03_047]
+With his face upwards and his tongue fixed in the aperture [of the skull], the yogi should visualise as the supreme \emph{śakti} [the nectar] that is forcibly obtained from the breath as it drips from the head into the sixteen petals of the lotus.
+And he who drinks the gushing nectar, the pure fluid [surging] from the [moon's] digits in waves, is free of disease, has a body as soft as lotus fibre, and lives a long time.
+\end{translation}
+
+\begin{sources}[hp03_047]
+\emph{Vivekamārtaṇḍa} 118
+\begin{versinnote}
+\tl{mūrdhnaḥ ṣoḍaśapatrapadmagalitaṃ prāṇād avāptaṃ haṭhād \\+}
+\tl{ūrdhvāsyo rasanāṃ niyamya vivare śaktiṃ parāṃ cintayet |\\+}
+\tl{utkallolakalājalaṃ ca vimalaṃ dhārāmayaṃ yaḥ piben \\+}
+\tl{nirdoṣaḥ sa mṛṇālakomalatanur yogī ciraṃ jīvati ||\\+}
+\tl{\var{118b niyamya ] nidhāya, vidhāya \vl • cintayet ] cālayet \vl 118c utkallolakalājalaṃ ] ca vimalaṃ dhārāmayaṃ ] (from HP); utkallolakalākalaṃ, utkallolajalākulaṃ, utkallolajalāmṛtaṃ, tat kallolakalājalaṃ, tat kallolajalākulaṃ \vl • ca vimalaṃ ] suvimalaṃ \vl • dhārāmayaṃ ] (from HP); jīvākulaṃ, jihvākulaṃ, dhārājalaṃ vl 118d tanur ] vapur \vl}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_047]
+Cf.~\emph{Haṭharatnāvalī} 2.150
+\begin{versinnote}
+\tl{utkallolakalāmṛtaṃ ca vimalaṃ dhārāmṛtaṃ yaḥ pibet |\\+}
+\tl{nirdoṣaḥ sa mṛnālakomalatanur yogī ciraṃ jīvati ||\\+}
+\tl{\var{utkallola° ] tatkallola° \vl, °tanur ] °vapur \vl}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~75r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mūrdhnaḥ ṣoḍaśapadmapatragalitaṃ prāṇād avāptaṃ haṭhād\\+}
+\tl{ūrdhvāsyo rasanāṃ niyamya vivare śaktiṃ parāṃ cintayan |\\+}
+\tl{tatkallolakalājalaṃ ca vimalaṃ jihvākulaṃ yaḥ piben\\+}
+\tl{nirdoṣaḥ sa mṛṇālakomalatanur yogī ciraṃ jīvati ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.217 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{mūrdhnaḥ ṣoḍaśapatrapadmagalitaṃ prāṇād avāptaṃ haṭhāt \\+}
+\tl{ūrdhvāsyo rasanāṃ niyamya vivare śaktiṃ parāṃ cintayet |\\+}
+\tl{tatkallolakalājalaṃ suvimalaṃ dhārāmṛtaṃ yaḥ pibet\\+}
+\tl{nirddoṣaḥ sa mṛṇālakomalatanur yogī paraṃ jīvati ||\\+}
+\tl{\var{cintayet ] cintayan \vl}\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 14.24 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ūrdhvaṃ ṣoḍaśapatrapadmagalitaṃ prāṇād avāptaṃ haṭhād \\+}
+\tl{ūrdhvāsyo rasanāṃ niyamya kuhare śaktiṃ parāṃ cintayan |\\+}
+\tl{utkallolakalājalaṃ suvimalaṃ dhārāmṛtaṃ yaḥ piben \\+}
+\tl{nirddoṣaḥ sa mṛṇālakomalavapur yogī ciraṃ jīvati ||\\+}
+\tl{\var{ūrdhvaṃ ] mūrdhvaṃ \vl}\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_047]
+The meaning of \emph{prāṇāt} (`from the breath') in the first verse quarter is not easy to understand without the context of this verse in the source text, the \emph{Vivekamārtaṇḍa}.
+In the verse preceding this one in the \emph{Vivekamārtaṇḍa} (117), the breath, on reaching the “great lotus”, is said to turn into nectar (\emph{amṛta}).
+In \emph{Jyotsnā} 3.51, Brahmānanda says that there is a variant \emph{prāṇaiḥ} (`by means of the breaths'), which is easier to understand than \emph{prāṇāt}.
+He nonetheless accepts \emph{prāṇat} and understands it as being a means (\emph{prāṇāt sādhanabhūtād avāptam}).
+He also understands the sixteen-petalled lotus to be the lotus in the throat, into which the nectar drips.
+%Adopt patrapadma.
+\end{philcomm}
+
+\begin{metre}[hp03_047]
+Śārdūlavikrīḍita 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.48}
+\begin{translation}[hp03_048]
+If the tongue, while oozing nectar and constantly kissing the tip of the uvula, is salty, pungent, like  milk or the same as honey and ghee, diseases are eliminated for [the yogi], old age is stopped, he can recite treatises and scriptures, attains immortality together with the eightfold powers, and attracts Siddha women.
+\end{translation}
+
+\begin{sources}[hp03_048]
+\emph{Vivekamārtaṇḍa} 128
+\begin{versinnote}
+\tl{cumbantī yadi lambikāgram aniśaṃ  jihvā rasasyandinī\\+}
+\tl{sakṣārā kaṭukātha dugdhasadṛśā madhvājyatulyāthavā  |\\+}
+\tl{vyādhīnāṃ haraṇaṃ  jaropaśamanaṃ śāstrāgamodīraṇaṃ\\+}
+\tl{tasya syād amaratvaṃ aṣṭaguṇitaṃ siddhāṅgānākarṣaṇam ||\\+}
+\tl{\var{128a rasasyandanī ] rasaspandanī VTG, rasāsvādinī A 128c jaropaśamanaṃ ] AGBGL; jarāpaharaṇaṃ V, jarāntakaranaṃ TGP • °odīraṇaṃ ] VA; °odgīraṇaṃ TGBGL, °occāraṇaṃ GP}\\!}
+%
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_048]
+\emph{Yogacintāmaṇi} f.~75r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{cumbantī yadi lambikāgram aniśaṃ jihvā rasasyandinī\\+}
+\tl{sakṣārā kaṭukāmladugdhasadṛśaṃ madhvājyatulyaṃ yadā |\\+}
+\tl{vyādhīnāṃ haraṇaṃ jarāntakaraṇaṃ śāstrāgamoddhāraṇaṃ\\+}
+\tl{tasya syād iha siddhir aṣṭaguṇitā siddhāṅgaṇākarṣaṇam ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.218 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{cumbantī yadi lambikāgram aniśaṃ jihvā rasasyandinī \\+}
+\tl{sakṣārā kaṭukāmladugdhasadṛśī madhvājyatulyāthavā |\\+}
+\tl{vyādhīnāṃ haraṇaṃ jarāmbutaraṇaṃ śāstrāgamodgīraṇaṃ\\+}
+\tl{tasya syād amaratvam aṣṭaguṇitaṃ siddhāṅganākarṣaṇam ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 14.25 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{cumbantī yadi lambikāgram anilaṃ jihvā rasasyandinī\\+}
+\tl{sakṣārā kaṭukāmladugdhasadṛśāṃ madhvājyatulyā tathā |\\+}
+\tl{vyādhīnāṃ haraṇaṃ jarāntakaraṇaṃ śāstrāgamodgīraṇaṃ\\+}
+\tl{tasya syād amaratvam aṣṭaguṇavat siddhāṅganākarṣaṇam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_048]
+%kaṭukāmla° or kaṭukātha. Adopt kaṭukātha (N3,J5, \emph{Vivekamārtaṇḍa})
+%guṇavat (N3, J5 [-va])
+%MD: Done.
+%\end{philcomm}
+
+\begin{metre}[hp03_048]
+Śārdūlavikrīḍita 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.49}
+\begin{translation}[hp03_049]
+There is one seed [syllable], which contains creation, one \emph{mudrā}, \emph{khecarī}, one god, the unsupported, and one state, beyond mind.
+\end{translation}
+
+\begin{sources}[hp03_049]
+\emph{Timirodghāṭana} 5.14c–15b (NGMPP A35/3)
+\begin{versinnote}
+\tl{eka[ṃ] sṛṣṭimayaṃ bījaṃ ek[ā] mudrā tu khecarī |\\+}
+\tl{dvāv etau jñāyate yena so pi śāntapade sthitam ||\\!}
+\end{versinnote}
+
+Quotation by Jayaratha \emph{ad} \emph{Tantrāloka} 32.63, introduced with \emph{yad āgamaḥ}
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī |\\+}
+\tl{dvāv ekaṃ yo vijānāti sa vai pūjyaḥ kulāgame ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_049]
+\emph{Haṭharatnāvalī} 4.28
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī |\\+}
+\tl{eko devo nirālambaḥ ekāvasthā manonmanī  ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~75r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī |\\+}
+\tl{eko devo nirālamba ekāvasthā manonmanī ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.219 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ bījaṃ ekā mudrā ca khecarī |\\+}
+\tl{eko devo nirālamba ekāvasthā manonmanī ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_049]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.49*1}
+
+%%%%%%%%%%
+\subsection*{3.49*2}
+\begin{translation}[hp03_049_2]
+That which enters the aperture into the underworld exists at the base of Meru.
+The wise [yogi] says that is the truth, the source of [all] rivers.
+The essence of the body flows from the moon.
+Because of that people die.
+[The yogi] should block it with the clay of the excellent [\emph{khecarī}] technique.
+Bodily perfection [arises] no other way.
+\end{translation}% greyscale in ed.
+
+%\begin{sources}[hp03_049_2]
+%\end{sources}
+
+\begin{testimonia}[hp03_049_2]
+\emph{Haṭharatnāvalī} 4.30
+\begin{versinnote}
+\tl{pātāle yad viśati suṣiraṃ merūmūlaṃ tad asti\\+}
+\tl{tattvaṃ caitad vadati sudhā tanmukhaṃ nimnagānām |\\+}
+\tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
+\tl{tad badhnīyāt sukharatimṛdur nānyathā kāryasiddhiḥ ||\\!}
+\end{versinnote}
+% after (ekaṃ sṛṣṭimayaṃ + mano yatra vilīyeta) and before (śarīraṃ tāvad eva)
+%the context of this verse in the HR is rājayoga.
+
+\emph{Yogacintāmaṇi} f.~75r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tat pātālād viyati śikhare merumūle tad asti \\+}
+\tl{tattvaṃ caitat pravadati sudhīḥ saṃmukhe nimnagānām |\\+}
+\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
+\tl{tad badhnīyāt svakaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
+\end{versinnote}
+% after the verse (cittāyattaṃ nṛṇāṃ śukraṃ) and before (indhanāni yathā vahnis)
+% context is khecarī (3rd chapter material)
+
+\emph{Yuktabhavadeva} 7.220 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{pātāle yad viśati suṣiraṃ merumūle yad asti \\+}
+\tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
+\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
+\tl{badhnīyāt tat sukaraṇam atho nānyathā kāyasiddhiḥ ||\\!}
+\end{versinnote}
+% context is khecarī, (3rd chapter material)
+% after (ekaṃ sṛṣṭimayaṃ bījaṃ) and before (atha uḍḍiyānabandhaḥ)
+
+\emph{Haṭhatattvakaumudī} 14.26 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pātāle yad vitatasuṣiraṃ merumūle tad asmin \\+}
+\tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
+\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
+\tl{taṃ badhnīyāt sukaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
+\end{versinnote}
+% context is khecarī 3rd chapter material
+% after (cumbantī yadi lambikā) and before (cittaṃ vicarati gagane)
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_049_2]
+%Workshop notes
+% Sven: the first line could refer to Kuṇḍalinī
+% J5 has this verse here.
+% mṛdā = using clay to make a dam?
+% Also see Netratantra 7.32
+% Could this verse be about mūlabandha? (reading pātāle and mūle)
+%J5: yat prāleyam pihitasukhire merumūle yad astī — this seems helpful.
+% PN has a nice-ish reading: yat prāleyaṃ pihita sukhiraṃ merumūrdhnaṃ. mūrdhni etc. is much better.
+% J10 has merumūrdhni sthitaṃ yat, which is good.
+% If going with prāleyam, adopt yat prāleyam pihitasuṣiraṃ merumūrdhny asti tathyaṃ (~K1 Grp2)
+% ... tasmin tattvaṃ pravadati
+%Haru: Perhaps, praleyam was corrupted to pātāle and then the rest of the line changed.
+
+This verse occurs in various places in the different recensions of the text.
+It is here in the \alphaTwo\ and \alphaThree\ manuscripts, in chapter four (4.8*3) in \alphaOne\ and the delta group, after 3.39 in \betaOne\ and \betaTwo, and in both the third and fourth chapters in the \textgamma\ manuscripts.
+It must have entered the transmission of the \emph{Haṭhapradīpikā} at an early stage, and has moved around, perhaps because the name of the technique alluded to is not stated and the meaning of the first half of the verse is somewhat vague without its original context (we are yet to identify a source).
+
+The first verse quarter has many variants among the witnesses. \alphaTwo\ has:
+\begin{versinnote}
+\tl{yat prāleyaṃ pihitasukhire merumūle yad astī \\+}
+\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\!}
+\end{versinnote}
+This version of the verse is close to that of other important manuscripts in the \textbeta, \textgamma\ and \texteta\ groups.
+With the help of these manuscripts (in particular \getsiglum{K1} of the \textgamma\ group), the above can be emended and understood as follows:
+\begin{versinnote}
+\tl{yat prāleyaṃ pihitasuṣiraṃ merumūrdhny asti tathyaṃ \\+}
+\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\+}
+\tl{°suṣiraṃ ] \getsiglum{K1}, °sukhire \alphaTwo. °mūrdhny ] \getsiglum{P8}. °mūle \alphaTwo. asti tathyaṃ ] \getsiglum{K1}, yad astī \alphaTwo.\\!}
+“That cool liquid by which the aperture is filled at the top of Meru and exists as the truth, the wise [yogi] says that is the source of [all] rivers.” 
+\end{versinnote}
+
+%K1 in Mitsuyo's notes
+%yat prāleya prahitasuṣiraṃ merumūrdhnāsti tathyaṃ
+%tasmiṃs tatvaṃ pravadati sudhīs tan mukhaṃ nimnagānāṃ
+
+% Other manuscripts have something more similar to:
+% \begin{versinnote}
+% \tl{pātāle yad viśati suṣiraṃ merumūlaṃ tad asti \\+}
+% \tl{tattvaṃ caitat pravadati sudhīs tan mukhaṃ nimnagānām\\!}
+% \end{versinnote}
+
+\end{philcomm}
+
+\begin{metre}[hp03_049_2]
+Mandākrāntā 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.50}
+\begin{translation}[hp03_050]
+The root lock (\emph{mūlabandha})—\\
+When [the yogi] presses the perineum with the side of the heel, clenches the anus and draws up \emph{apāna}, it is called the root lock.
+\end{translation}
+
+\begin{sources}[hp03_050]
+\emph{Vivekamārtaṇḍa} 42
+\begin{versinnote}
+\tl{pārṣṇibhāgena saṃpīḍya yonim ākuñcayed gudam |\\+}
+\tl{apānam ūrdhvam ākṛṣya mūlabandho 'yam ucyate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_050]
+\emph{Haṭharatnāvalī} 2.58
+\begin{versinnote}
+\tl{pārṣṇibhāgena sampīḍya yonim ākuñcayed gudaṃ |\\+}
+\tl{apānam ūrdhvam ākuñcya mūlabandho 'yam ucyate ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pārṣṇibhāgena saṃpīḍya yonim ākuñcayed gudam |\\+}
+\tl{apānam ūrdhvam ākṛṣya mūlabandho 'yam ucyate ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_050]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.51}
+\begin{translation}[hp03_051]
+[The yogi] forces the downward-moving \emph{apāna} breath to move upwards by contracting [the anus]. Yogis call that \emph{mūlabandha}.
+\end{translation}
+
+\begin{sources}[hp03_051]
+\emph{Gorakṣaśataka} 53
+\begin{versinnote}
+\tl{adhogatim apānaṃ vai ūrdhvagaṃ kurute balāt |\\+}
+\tl{ākuñcanena taṃ prāhur mūlabandhaṃ tu yoginaḥ ||\\+}
+\tl{\var{53a °gatiṃ ] TU; °gataṃ G 53c ākuñcanena taṃ ] GU ākuñcane ca tat T}\\+}
+\tl{\var{53d mūlabandhaṃ tu yoginaḥ ] T; mūlabandho yam ucyate GU}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_051]
+\emph{Haṭharatnāvalī} 2.59
+\begin{versinnote}
+\tl{adhogatim apānaṃ vai ūrdhvagaṃ kurute balāt |\\+}
+\tl{ākuñcanena taṃ prāhur mūlabandhaṃ hi yoginaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{adhogatam apānaṃ ca tad ūrdhvaṃ kurute haṭhāt |\\+}
+\tl{ākuñcanena taṃ prāhur mūlabandhaṃ tu yoginaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_051]
+%Adopt balāt (N3). haṭhāt seems to be an early alt. reading.
+%MD: Text changed.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.52}
+\begin{translation}[hp03_052]
+[The yogi] should press his anus with his heel and forcefully contract the [\emph{apāna}] wind over and over again so that the breath goes upwards.
+\end{translation}
+
+\begin{sources}[hp03_052]
+\emph{Dattātreyayogaśāstra} 144
+\begin{versinnote}
+\tl{gudaṃ pārṣṇyā tu saṃpīḍya yonim ākuñcayed balāt | \\+}
+\tl{vāraṃ vāraṃ yathā cordhvaṃ samāyāti samīraṇaḥ ||\\+}
+\tl{\var{yonim ] from \emph{Jyotsnā}, vāyum \emph{codd.}}\\!}
+\end{versinnote}
+
+Cf.~\emph{Śārṅgadharapaddhati} 4416
+\begin{versinnote}
+\tl{gudaṃ pārṣṇyā tu sampīḍya vāyum ākuñcayed balāt |\\+}
+\tl{bāraṃ bāraṃ yathā cordhvaṃ samāyāti samīraṇaḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_052]
+\emph{Haṭharatnāvalī} 2.60
+\begin{versinnote}
+\tl{gudaṃ pārṣṇyā ca saṃpīḍya vāyum ākuñcayed balāt |\\+}
+\tl{vāraṃ vāraṃ yathā cordhvaṃ samāyāti samīraṇaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.76r (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{gudaṃ pārṣṇyā tu saṃpīḍya vāyum ākuñcayed balāt |\\+}
+\tl{vāraṃ vāraṃ tathā cordhvaṃ samāyāti samīraṇaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogabīja} 103 (south-Indian recension)
+\begin{versinnote}
+\tl{gudaṃ pārṣṇyā tu saṃpīḍya vāyum ākuñcayed balāt |\\+}
+\tl{vāraṃ vāraṃ yathā cordhvaṃ samāyāti samīraṇaḥ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_052]
+The instruction to `contract the wind' (\emph{vāyum ākuñcayet}) is odd (especially with \emph{samīraṇaḥ} in the fourth \emph{pāda}) and not found in other texts.
+Mallin\-son has adopted \emph{yonim} for \emph{vāyum} (cf.~\emph{Haṭhapradīpikā} 3.50d) in his edition of this verse in its source text, the \emph{Dattātreyayogaśāstra}, which is not found in the manuscripts of that text but is in the \emph{Haṭhapradīpikā}’s \delta \ manuscripts.
+
+In the context of the root lock, \emph{vāyum ākuñcayet} can be understood as an instruction to contract \emph{apāna\-vāyu}, which is mentioned in the previous verse (3.51a).
+Instructions to contract \emph{apāna\-vāyu} are found in other yoga texts, such as \emph{Yoga\-tārā\-valī} 7b (\emph{ākuñcanaiḥ śaśvad apānavāyoḥ}), \emph{Śiva\-saṃhitā} 4.84cd (\emph{apāna\-vāyum ākuñcya balād...}), \emph{Śiva\-yoga\-pradīpikā} 2.53ab (\emph{athordhva\-madhya\-sthira\-bandha\-nābhyām ākuñcanād ūrdhvam apānavāyoḥ}) and \emph{Yukta\-bhava\-deva} 7.297 (\emph{ādhāra\-kamale suptāṃ cālayet kuṇḍalīṃ dṛḍhām} | \emph{apāna\-vāyum ākṛṣya balād ākuñcya buddhimān}).
+In \emph{Jyotsnā} 3.63, Brahmānanda understands \emph{vāyu} in this verse as \emph{apāna} when he says that `one should contract the wind, \emph{apāna}' (\emph{vāyum apānam ākuñcayed}), which he explains as, `one should pull it by contractions of the anus' (\emph{gudasyā\-kuñcanenā\-karṣayet}).
+In the same vein, Bhava\-deva\-miśra glosses `pulling \emph{apāna}' (\emph{apānā\-karṣaṇam}) as `contracting the anus' (\emph{gudā\-kuñcanam}), when commenting on `having pulled \emph{apāna\-vāyu} and forcefully contracted it ...' (\emph{apāna\-vāyum ākṛṣya balād ākuñcya...}) in \emph{Yukta\-bhava\-deva} 297 and 301.
+\end{philcomm} %?? JB: I've added references to the commentators Brahmānanda and Bhavadevamiśra to clarify the meaning of vāyum ākuñcayet.
+
+%%%%%%%%%%
+\subsection*{3.53}
+\begin{translation}[hp03_053]
+When \emph{prāṇa} and \emph{apāna} [and] \emph{nāda} and \emph{bindu} become united by means of the root lock they are sure to bestow complete success in yoga.
+\end{translation}
+
+\begin{sources}[hp03_053]
+\emph{Dattātreyayogaśāstra} 145
+\begin{versinnote}
+\tl{prāṇāpānau nādabindū mūlabandhena caikatām |\\+}
+\tl{gatvā yogasya saṃsiddhiṃ yacchato nātra saṃśayaḥ ||\\+}
+\tl{\var{145c gatvā yogasya saṃ° ] gacchato yogasaṃ° M1A}\\+}
+\tl{\var{145d yacchato ] gacchato M1, kurute AM2, gachate \pi}\\!}
+\end{versinnote}
+
+%\emph{Śārṅgadharapaddhati} 4417
+%\begin{versinnote}
+%\tl{prāṇāpānau nādabindū mūlabandhena caikatām |\\+}
+%\tl{gatvā yogasya saṃsiddhiṃ gacchatau nātra saṃśayaḥ ||\\!}
+%\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_053]
+\emph{Haṭharatnāvalī} 2.61
+\begin{versinnote}
+\tl{prāṇāpānau nādabindū mūlabandhena caikatām |\\+}
+\tl{gatau tadā yogasiddhiṃ prāpnoty eva na saṃśayaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.76r (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{prāṇāpānau nādabindū mūlabandhena caikatām |\\+}
+\tl{gatvā yogasya saṃsiddhiṃ gacchato nātra saṃśayaḥ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_053]
+Since the term \emph{nāda} usually means `internal sound' in Haṭha and Rājayoga texts, it is possible that \emph{bindu} here was understood by some to have the tantric connotations of sonic and visual foci (Mallinson 2007:219 n.325) or two levels of sonic emanation in \emph{mantroccāra}, where \emph{nāda} is an unvoiced sound and \emph{bindu} is the slightly coarser sound of inner murmuring (see \emph{Tāntrikābhidhānakośa} vol.
+3, 2013: 278–279).
+However there is a passage in the \emph{Amaraugha} (10–12) where \emph{nāda} and \emph{bindu} are paired and it is clear that \emph{bindu} means generative fluid.
+\end{philcomm}
+
+\begin{metre}[hp03_053]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.54}
+\begin{translation}[hp03_054]
+\textit{Prāṇa} and \textit{apāna} unite, urine and faeces diminish, [and] even an old man becomes young as a result of the continuous application of the root lock.
+\end{translation}
+
+\begin{sources}[hp03_054]
+\emph{Vivekamārtaṇḍa} 41
+\begin{versinnote}
+\tl{apānaprāṇayor aikyaṃ kṣayo mūtrapurīṣayoḥ |\\+}
+\tl{yuvā bhavati vṛddho ’pi satataṃ mūlabandhanāt ||\\!}
+\end{versinnote}
+
+%\emph{Śārṅgadharapaddhati} 4418
+%\begin{versinnote}
+%\tl{apānaprāṇayor aikyaṃ kṣayo mūtrapurīṣayoḥ |\\+}
+%\tl{yuvā bhavati vṛddho 'pi satataṃ mūlabandhanāt ||\\!}
+%\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_054]
+\emph{Haṭharatnāvalī} 2.62
+\begin{versinnote}
+\tl{apānaprāṇayor aikyaṃ kṣayo mūtrapurīṣayoḥ |\\+}
+\tl{yuvā bhavati vṛddho 'pi satataṃ mūlabandhanāt ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_054]
+The diminishing of urine and faeces as a result of success in yoga is mentioned in the \emph{Amanaska} (1.50c) and \emph{Dattātreyayogaśāstra} (80a).
+%The term \emph{kṣaya} can mean either the end or diminishment of something.
+In the context of urine and faeces (\emph{mūtrapurīṣa}), their diminishment is the likely meaning intended as there are references in other texts to the successful practice of a yoga technique resulting in a reduced amount of urine and faeces, such as \emph{Amanaska} 1.50c (\emph{svalpamūtrapurīṣatvaṃ}) and \emph{Dattātreyayogaśāstra} 80a (\emph{alpamūtrapurīṣaḥ}), which were both known to Svātmārāma.
+In \emph{Jyotsnā} 3.65, Brahmānanda understands \emph{kṣaya} in this verse as `decline' (\emph{kṣayaḥ patanam}).%JM: I don't recall discussing this, but I think patanam here means excretion.
+Brahmānanda has  mūtrapurīṣayoḥ saṃcitayoḥ kṣayaḥ patanam.
+He's surely wrong in his understanding of the pāda.
+I have changed the note to just give the parallels.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.55}
+\begin{translation}[hp03_055]
+When \emph{apāna} has turned upwards and reached the orb of fire, then the flame of the fire, fanned by the wind, grows tall.
+\end{translation}
+
+\begin{sources}[hp03_055]
+\emph{Gorakṣaśataka} 54
+\begin{versinnote}
+\tl{apāne cordhvage jāte saṃprāpte vahnimaṇḍalam |\\+}
+\tl{tato 'nalaśikhā dīrghā vardhate vāyunāhatā ||\\+}
+\tl{\var{54a °maṇḍalam ] \emph{from Haṭhapradīpikā witnesses}, maṇḍale \emph{codd.}}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_055]
+\emph{Haṭharatnāvalī} 2.63
+\begin{versinnote}
+\tl{apāne cordhvage jāte prayāte vahnimaṇḍale |\\+}
+\tl{tathānalaśikhādīptir vāyunā preritā yathā ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76r (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{apāne cordhvage jāte saṃprāpte vahnimaṇḍale |\\+}
+\tl{tathānalaśikhā dīrghā vardhate vāyunāhatā ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_055]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.56}
+\begin{translation}[hp03_056]
+As a result, fire and \textit{apāna} reach \textit{prāṇa}, which is hot by nature, and the \emph{prāṇa} makes the fire in the body extremely hot.
+\end{translation}
+
+\begin{sources}[hp03_056]
+\emph{Gorakṣaśataka} 55
+\begin{versinnote}
+\tl{tato yātau vahnyapānau prāṇam uṣṇasvarūpakam |\\+}
+\tl{tenātyantapradīptena jvalano dehajas tathā ||\\+}
+\tl{\var{dehajas ] T; dehagas GU}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_056]
+\emph{Haṭharatnāvalī} 2.64
+\begin{versinnote}
+\tl{yātāyātau vahnyapānau mūlarūpasvarūpakau |\\+}
+\tl{tenābhyantaḥ pradīptas tu jvalano dehajas tathā || 2.64 ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76r–76v (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{tato yātau vahnyapānau prāṇam uktasvarūpakau |\\+}
+\tl{tenātyantapradīptas tu jvalano dehajas tathā ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_056]
+The second verse quarter has been rewritten in the \textalpha\ manuscripts as \emph{prāṇamūlasvarūpakam}, which is similar to \etaOne\ (\emph{prāṇamūlasvarūpakau}) and the \emph{Haṭharatnāvalī} (\emph{mūlarūpasvarūpakau}).
+Manuscripts from the delta group, as well as the \emph{Yogacintāmaṇi}, have \emph{prāṇam uktasvarūpakam} or \emph{prāṇam uktasvarūpakau}.
+It appears that the intention behind these rewrites was to avoid the reading in the source text, `\textit{prāṇa} is hot by nature' (\emph{prāṇam uṣṇasvarūpakam}), which was likely accepted by Svātmārāma because it is in some witnesses of the \textbeta, \textgamma\ and \textepsilon\ groups.
+References to \emph{prāṇa} being hot by nature (and \emph{apāna} being cold) occur in other works, such as \emph{Mokṣopāya} 6.85.111-112 and \emph{Haṭhatattvakaumudī} 4.14, 41.2.
+The commentators, Bālakṛṣṇa (\emph{Yogaprakāśikā} 5.85) and Brahmānanda (\emph{Jyotsnā} 3.67) accept the idea that \emph{prāṇa} is hot by nature.
+%
+
+\end{philcomm}
+
+\begin{metre}[hp03_056]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.57}
+\begin{translation}[hp03_057]
+Heated by that [blaze], the sleeping \emph{kuṇḍalinī} wakes up. Like a snake struck by a stick, she hisses and becomes straight.
+\end{translation}
+
+\begin{sources}[hp03_057]
+\emph{Gorakṣaśataka} 56
+\begin{versinnote}
+\tl{tena kuṇḍalinī suptā saṃtaptā saṃprabudhyate |\\+}
+\tl{daṇḍāhatā bhujaṃgīva niśvasya ṛjutāṃ vrajet ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_057]
+\emph{Haṭharatnāvalī} 2.65ab
+\begin{versinnote}
+\tl{daṇḍāhatā bhujaṅgīva niścitaṃ ṛjutām iyāt |\\+}
+\tl{\var{niścitaṃ ] niśvasya T,P,t1}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{tena kuṇḍalinī suptā satataṃ saṃprabodhyate |\\+}
+\tl{daṇḍāhatabhujaṅgīva niścitam ṛjutāṃ vrajet ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_057]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.58}
+\begin{translation}[hp03_058]
+Then, like [a snake] that has entered a hole, she goes into the channel of Brahman. Therefore, yogis should regularly practise the root lock.
+\end{translation}
+
+\begin{sources}[hp03_058]
+\emph{Gorakṣaśataka} 57
+\begin{versinnote}
+\tl{bile praviṣṭe ca tato brahmanāḍyantaraṃ vrajet |\\+}
+\tl{tasmān nityaṃ mūlabandhaḥ kartavyo yogibhiḥ sadā ||\\+}
+\tl{\var{57a bile ] bil*e*T, bila° G2U, bilaṃ G1 • °praviṣṭe ca tato ] T; °praveśato yatra GU}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_058]
+\emph{Haṭharatnāvalī} 2.65c-f
+\begin{versinnote}
+\tl{bilaṃ praviṣṭeva tato brahmanāḍyantaraṃ vrajet |\\+}
+\tl{tasmān nityaṃ mūlabandhaḥ kartavyo yogibhiḥ sadā ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{bilaṃ praviṣṭeva tathā brahmanāḍyantaraṃ vrajet |\\+}
+\tl{tasmān nityaṃ mūlabandhaḥ kartavyo yogipuṅgavaiḥ ||\\+}
+\tl{\var{praviṣṭeva ] praviṣṭaiva \vl}\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_058]
+%\end{philcomm}
+
+\begin{metre}[hp03_058]
+Anuṣṭubh (a: bha-vipulā; c: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.59}
+\begin{translation}[hp03_059]
+ Now the Uḍḍīyaṇa lock:\\
+Because the breath flies up (\emph{uḍḍīyate}) into the \emph{suṣumṇā} when it is bound by it, yogis say that this [practice] is called Uḍḍīyaṇa.
+\end{translation}
+
+\begin{sources}[hp03_059]
+\emph{Gorakṣaśataka} 58c–59b
+\begin{versinnote}
+\tl{baddho yena suṣumṇāyāṃ prāṇas tūḍḍīyate yataḥ |\\+}
+\tl{tasmād uḍḍīyaṇākhyo 'yaṃ yogibhiḥ samudāhṛtaḥ ||\\+}
+\tl{\var{58c baddho ] \emph{em.~from HP}; vajro G, bandho TU 58d yataḥ ] TU; tataḥ G}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_059]
+\emph{Haṭharatnāvalī} 2.53
+\begin{versinnote}
+\tl{baddho yena suṣumnāyāṃ prāṇas tūḍḍiyate yataḥ |\\+}
+\tl{tasmād uḍḍiyānākhyo 'yaṃ yogibhiḥ samudāhṛtaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{baddho yena suṣumṇāyāṃ prāṇas tūḍḍīyate yataḥ |\\+}
+\tl{tasmād uḍḍiyānākhyo 'yaṃ yogibhiḥ samudāhṛtaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogabīja} 104cd–105ab (South-Indian recension)
+\begin{versinnote}
+\tl{baddho yena suṣumṇāyāṃ prāṇas tūḍḍīyate tataḥ|\\+}
+\tl{tasmād uḍḍīyānākhyo 'yaṃ yogibhiḥ samudāhṛtaḥ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_059]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.60}
+\begin{translation}[hp03_060]
+Because the great bird tirelessly flies up (\emph{uḍḍīnaṃ kurute}), it is [called] 'flying up'. In it, the [root] lock is applied.
+\end{translation}
+
+\begin{sources}[hp03_060]
+\emph{Vivekamārtaṇḍa} 43
+\begin{versinnote}
+\tl{uḍḍīṇaṃ kurute yasmād aviśrāntaṃ mahākhagaḥ |\\+}
+\tl{uḍḍiyānaṃ tad eva syāt tatra bandho vidhīyate ||\\+}
+\tl{\var{43b aviśrāntaṃ ] \emph{em.~from HP}; aviśrānta VTU, aviśrānto G}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_060]
+\emph{Haṭharatnāvalī} 2.54
+\begin{versinnote}
+\tl{uḍḍīnaṃ kurute yasmād aviśrāntaṃ mahākhagaḥ |\\+}
+\tl{uḍḍiyānaṃ tad eva syāt tatra bandho 'bhidhīyate ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{uḍḍīnaṃ kurute yasmād aviśrāntaṃ mahākhagaḥ |\\+}
+\tl{uḍḍiyānaṇ tad eva syān mūlabandho 'bhidhīyate ||\\!}
+\end{versinnote}
+
+%Skandapurāṇa 4.41.47
+%\begin{versinnote}
+%\tl{uḍḍīnaṃ kurute yasmād ahorātraṃ mahākhagaḥ |\\+}
+%\tl{uḍḍīyānaṃ tataḥ proktaṃ tatra bandho vidhīyate ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_060]
+In the \emph{Vivekamārtaṇḍa} this verse is preceded by a passage on \emph{mūlabandha}, so the likely meaning of the fourth verse quarter is that (\emph{mūla})\emph{bandha} is to be performed in this practice.
+The \textbeta\ group and many other manuscripts of the \emph{Haṭhapradīpika} have `the root [lock] is applied' (\emph{mūlaṃ vidhīyate}), which appears to be an attempt to clarify the meaning of the original verse.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.61}
+\begin{translation}[hp03_061]
+ [The yogi] should perform a rearward and upward stretching of the navel into the abdomen. That is the Uḍḍiyāna lock, a lion to the elephant of death.
+\end{translation}
+
+\begin{sources}[hp03_061]
+\emph{Vivekamārtaṇḍa} 44
+\begin{versinnote}
+\tl{udare paścime tāṇaṃ nābher ūrdhvaṃ ca kārayet |\\+}
+\tl{uḍḍīyāno hy asau bandho mṛtyumātaṅgakesarī ||\\!}
+\end{versinnote}
+% tāna is not in M-W so prob comes from Marathi. Molesworth:  ताण tāṇa (p. 374)
+% ताण tāṇa m (तन S) The state of being stretched or strained; stretchedness or strain (as of a rope, cloth &c.) v दे, बस, भर.
+2 fig.
+Intense anger, a rage, a passion.
+v ये.
+Ex.
+वर्माची गोष्ट काढतांच कसा ताण आला.
+3...
+
+\emph{Śivasaṃhitā} 4.73
+\begin{versinnote}
+\tl{udare paṣcimaṃ tānaṃ nābher ūrdhvaṃ tu kārayet |\\+}
+\tl{uḍyānākhyo'tra bandho'yaṃ mṛtyumātaṅgakesarī ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_061]
+\emph{Haṭharatnāvalī} 2.55
+\begin{versinnote}
+\tl{udare paścimaṃ tānaṃ nābher ūrdhvaṃ ca dhārayet | \\+}
+\tl{uḍḍiyāno hy asau bandho mṛtyumātaṅgakesarī || \\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{udare paścimaṃ tānaṃ nābher ūrdhvaṃ samācaret |\\+}
+\tl{uḍḍiyāno hy asau bandho mṛtyumātaṅgakesarī ||\\!}
+\end{versinnote}
+
+%Skandapurāṇa 4.41.48
+%\begin{versinnote}
+%\tl{jaṭhare paścimaṃ tānaṃ nābher ūrdhvaṃ ca dhārayet |\\+}
+%\tl{uḍḍīyāno hy ayaṃ bandho mṛtyor api bhayaṃ tyajet || \\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_061]
+The variant spelling \sl{tāṇa} (for \emph{tāna}) reflects vernacular pronunciation (see e.g.~Molesworth 1857 s.v.~\emph{tāṇa}). 
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.62}
+\begin{translation}[hp03_062]
+\emph{Uḍḍiyāna} is easy, but it is always taught by a guru. If he practises it tirelessly, even an old man becomes young.
+\end{translation}
+
+\begin{sources}[hp03_062]
+\emph{Dattātreyayogaśāstra} 141c–142b
+\begin{versinnote}
+\tl{uḍḍiyāṇaṃ tu sahajaṃ guruṇā kathitaṃ sadā |\\+}
+\tl{abhyased astatandras tu vṛddho ’pi taruṇo bhavet ||\\+}
+\tl{\var{141d guruṇā ] guṇaughāt \pi}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_062]
+\emph{Haṭharatnāvalī} 2.56
+\begin{versinnote}
+\tl{guruṇā sahajaṃ proktaṃ vṛddho 'pi taruṇo bhavet |\\+}
+\tl{bāhyoḍyāṇaṃ ca kurute bāhyālaṅkāravardhanam ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{uḍḍiyānaṃ tu sahajaṃ guruṇā kathitaṃ yathā |\\+}
+\tl{abhyaset tad atandras tu vṛddho 'pi taruṇāyate ||\\!}
+\end{versinnote}
+
+\emph{Yogabīja} 106cd–107ab (South Indian recension)
+\begin{versinnote}
+\tl{uḍḍiyāṇaṃ tu sahajaṃ guruṇā kathitaṃ sadā  |\\+}
+\tl{abhyased asvatantras tu vṛddho 'pi taruṇo bhavet ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_062]
+At the end of the second \emph{pāda} \emph{sadā} is attested by witnesses of \textalpha, \texteta, \textepsilon\ and \delta \ groups, as well as the \emph{Dattātreyayogaśāstra}.
+We have understood the first line to mean that even though the basics of the practice of \emph{uḍḍiyāna} are easy, it still needs to be taught by the guru.
+Some witnesses read \emph{yathā} for \emph{sadā}, perhaps as a deliberate substitution of the more difficult \emph{sadā}, making the verse mean that \emph{uḍḍiyāna} is easy in the way that is taught by the guru.
+%The reading \emph{astatandraḥ} is probably the original one in the \emph{Dattātreyayogaśāstra} and is attested in some manuscripts of the \emph{Haṭhapradīpikā}.
+However, it is not strongly supported by the witnesses of the important groups and so may have changed to something simpler, such as \emph{tad atandras tu}.
+%  JB: collate J5 reading astatandras tu, so that last note is not necessary
+\end{philcomm}
+
+\begin{metre}[hp03_062]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.63}
+\begin{translation}[hp03_063]
+[The yogi] should carefully stretch [the region of the abdomen] above and below the navel. If he practises [like this] for six months, he is sure to conquer death.
+\end{translation}
+
+\begin{sources}[hp03_063]
+\emph{Dattātreyayogaśāstra} 142c–143b
+\begin{versinnote}
+\tl{nābher ūrdhvam ataḥ paścāt tānaṃ kuryāt prayatnataḥ || 142 ||\\+}
+\tl{ṣaṇmāsam abhyasen mṛtyuṃ jayaty eva na saṃśayaḥ |\\+}
+\tl{\var{142c ataḥ paścāt ] PT, adhaś cāpi \emph{cett.}}\\!}
+\end{versinnote}
+
+Cf.~\emph{Śivasaṃhitā} 4.72
+\begin{versinnote}
+\tl{nābher ūrdhvam adhaś cāpi tānaṃ paścimam ācaret |\\+}
+\tl{uḍyānabandha eṣaḥ syāt sarvaduḥkhaughanāśanaḥ  ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_063]
+\emph{Haṭharatnāvalī} 2.57
+\begin{versinnote}
+\tl{nābher ūrdhvam adho vāpi tānaṃ kuryāt prayatnataḥ |\\+}
+\tl{ṣaṇmāsam abhyasen mṛtyuṃ jayaty eva na saṃśayaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{nābher ūrdhvam adho vāpi tānaṃ kuryāt prayatnataḥ |\\+}
+\tl{ṣaṇmāsam abhyasen mṛtyuṃ jayaty eva na saṃśayaḥ ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Yuktabhavadeva} f.~76v (attr.~to the \emph{Śivayoga})
+\begin{versinnote}
+\tl{nābher ūrdhvam adhaś cāpi tānaṃ nirbharam ācaret |\\+}
+\tl{uḍḍiyāno hy ayaṃ bandhaḥ sarvaduḥkhaughanāśanaḥ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_063]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.64}
+\begin{translation}[hp03_064]
+%Sitting in \emph{vajrāsana}, [the yogi] should hold his feet firmly with his hands. And [now] the bulb is in the region of the ankles. He should press on it.
+Sitting in \emph{vajrāsana}, [the yogi] should hold his feet firmly with his hands near the region of the ankles and press the bulb (\emph{kanda}) there.
+\end{translation}
+
+\begin{sources}[hp03_064]
+\emph{Gorakṣaśataka} 59c–60b
+\begin{versinnote}
+\tl{sati vajrāsane jānū karābhyāṃ dhārayed dṛḍham ||\\+}
+\tl{gulphadeśasamīpe ca kandaṃ tatra prapīḍayet |\\+}
+\tl{\var{59c jānū ] em.; pādau GU, jānu T, prādau V}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_064]
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{sati vajrāsane pādau karābhyāṃ dhārayed dṛḍham |\\+}
+\tl{gulphadeśasamīpe ca udaraṃ tat prapīḍayet ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Yuktabhavadeva} 7.224 (commenting on \emph{uḍḍiyānabandha})
+\begin{versinnote}
+\tl{dṛḍham āsanaṃ baddhvā gulphadeśasamīpe karābhyāṃ pādau datvā nābhisamīpasthaṃ kandaṃ pīḍayann udare paścimatāṇaṃ tathā kuryād yathā vāyuḥ kukṣisandhiṃ na gacchaty evam uḍḍiyānabandho jarāmṛtyuvināśanaḥ sampadyate ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (ms.~no.~2244) f.~36r
+\begin{versinnote}
+\tl{tathā coktaṃ haṭhapradīpikāyām |\\+}
+\tl{sati vajrāsane pādau karābhyāṃ dharayed dṛḍhaṃ |\\+}
+\tl{gulphadeśasamīpe ca kandaṃ tatra nipīḍayet ||... \\+}
+\tl{siddhāsane sthitvā hastābhyāṃ pādau gulphapradeśasamīpe dṛḍhaṃ dhṛtvā tunde nālotthāna[ṃ] sādhu vidhāya samāhitamanasā sudṛḍhamūlabandhajālandharabandhavatābhyāsinā sādhakena recakādau kuṃbhakānte udare paścimatāne kriyamāṇe nitarāṃ tadā tatra nābhikandanipīḍane paścimatānena sati nābhikandotthānāḍaya urdhvamukhā vikasitā viralā vimalā asaṃhatā vāyugrahasamarthā bhavanti tadā sakuṃbhitaḥ prāṇavāyuḥ śanaiḥ [||]\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_064]
+In \emph{Haṭhapradīpikā} this verse seems to instruct the yogi to press the bulb (\emph{kanda}) with the feet while holding them with the hands.
+The adopted reading of the source text, the \emph{Gorakṣaśataka}, which is only found in witness T (and there in the singular \emph{jānu}, which has been emended to the dual \emph{jānū} in Mallinson's edition) says that it is the knees that are to be held, which would still allow for the feet to press the bulb.
+In the \emph{Yuktabhavadeva} (7.224), Bhavadevamiśra says that the bulb is near the navel and the legs are held near the ankles, suggesting that the bulb is pressed by using the hands to pull the feet into the lower abdomen.
+Holding the ankles with the hands and pressing the \emph{kanda} with the feet is also the view of Brahmānanda (\emph{Jyotsnā} 3.114), who follows the \emph{Yogayājñavalkya} (4.14, 4.16) in thinking that the place of the \emph{kanda} is nine fingerbreadths above the middle of the body, which is two fingerbreaths above the anus (\emph{Jyotsnā} 3.113).
+In the \emph{Haṭhasaṅketacandrikā} (f.~36r), Sundaradeva explains that the yogi presses the \emph{kanda} in the navel by performing \emph{uḍḍiyānabandha}, along with the root and chin locks, at the end of \emph{kumbhaka} and the beginning of exhalation.
+It is thus the backward stretch in the abdomen (\emph{udare paścimatāna}) that presses the \emph{kanda} in the navel.
+
+%M3 and G5 (gr 4c)
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.65}
+\begin{translation}[hp03_065]
+[The yogi] should very gently stretch back his stomach, chest and †neck† in such a way that the breath does not come into contact with the stomach.
+\end{translation}
+
+\begin{sources}[hp03_065]
+\emph{Gorakṣaśataka} 60c–61b
+\begin{versinnote}
+\tl{paścimaṃ tānam udare dhārayed dhṛdaye gate |\\+}
+\tl{śanaiḥ śanair yathā prāṇas tundasaṃdhiṃ na gacchati ||\\+}
+\tl{\var{60d dhārayedd°] GU; kārye*raṃ*T, kuryāñ ca V dhṛdaye gate] \emph{em.}~Sathyanarayanan; dhṛdaye gale GUT, civukaṃ hṛdi V}\\!}%gatam ?
+conj.
+em.
+hṛdayaṃ gataṃ?
+or hṛdayaṃ gate, with gate agreeing with udare?
+gate is found in several HP mss.
+One instance of hṛdayaṃ gataḥ.
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_065]
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{paścimaṃ tānam udare kārayec cibukaṃ hṛdi |\\+}
+\tl{śanaiḥ śanair yathā prāṇas tundasiddhiṃ na gacchati ||\\!}
+\end{versinnote}
+
+\emph{Haṭhsaṅketacandrikā}, f.~36r–36v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{paścime tānam udare ku[r]yac c[a] cibukaṃ hṛdi |\\+}
+\tl{śanaiḥ śanair yathā prāṇaḥ kandasaṃdhi[ṃ] nigacchati ||\\+}
+\tl{yathābhyāsānurūpaṃ kandasaṃdhiṃ kandagatanāḍisaṃghavimalavivaraprānteṣu nigachati nitarāṃ gacchati tathā tathā kuṃbhakavṛddhiḥ sukharūpodbhavati sādhakasya yenoḍḍīyānena bandhena vāyuḥ proḍḍīyāste brahmanāḍyāṃ yato 'sau uḍḍīyānākhyaḥ smṛto bandha ārthaḥ sevyas tasmād yogibhiḥ siddhasevyaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogaprakāśikā} 5.96
+\begin{versinnote}
+\tl{paścimaṃ tānam udare kārayec cibukaṃ hṛdi |\\+}
+\tl{śanaiḥ śanairyathā prāṇaḥ skandhasaṅge na gacchati ||\\+}
+\tl{uktalakṣaṇe vajrāsane baddhe satī gulphadeśasamīpena meḍhraṃ vā prapīḍayet cibukaṃ hṛdi kṛtvā prāṇasyordhvasañcalanaṃ kārayet tena prāṇ[a]ḥ skandhasandhiṃ gacchatīty arthaḥ || \\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_065]
+In 3.65b, the reading \emph{gale} (`in the neck') is very well attested by manuscripts of the source text, the \emph{Gorakṣaśataka}, and the \emph{Hathapradīpikā} (including all three \textalpha\ witnesses).
+Its meaning is not entirely clear to us as the `backward stretch' (\emph{paścimaṃ tānam}) usually occurs above and below the navel when the \emph{uḍḍiyāna} lock is applied, as stated above in verse 3.61.
+In 3.65, the mention of the stretch in the chest (\emph{hṛdaya}) may also be consistent with 3.61 in so far as `above the navel' might include the lower region of the chest.
+Drawing on x-ray experiments on \emph{uḍḍiyānabandha} conducted at the Kaivalyadhama Yoga Institute and published in the \emph{Yoga Mīmāṃsā} Journal (e.g.~vol.1, issues 1-2), Dr M.\,M.\,Gore (2005:144) mentions a sub-atmospheric (negative) pressure in visceral cavities, such as the oesophagus and stomach, as a physiological effect of applying \emph{uḍḍiyāna}.
+So, it may be possible that a `backward stretch' in the throat was intended in 3.65b.
+However, we have not seen the neck mentioned in this regard in any other premodern work and the absence of \emph{ca} suggests that \emph{gale} may be a corruption.
+The alternative reading \emph{cibukaṃ hṛdi} in manuscripts of the \emph{Hathapradīpikā} on lower branches of the stemma and in the testimonia is a reference to \emph{jālandharabandha} and appears to be a patch.
+In his edition of the \emph{Gorakṣaśataka}, Mallinson has adopted the emendation \emph{gate} suggested by Dr Sathyanaryanan, which he understands to mean that the rearward stretch reaches as far as the heart.
+\end{philcomm}
+% `Anatomy and Physiology of Yogic Practices,' Dr. Makarand Madhukar Gore. New Delhi: Shri Jainendra Press. 2005.
+% Remove crux marks in the edition as gale is not gibberish
+
+\begin{metre}[hp03_065]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.66}
+\begin{translation}[hp03_066]
+\emph{Uḍḍiyāna} is the best of all the locks. When the \emph{uḍḍiyāna} lock is firm, liberation becomes easy.
+\end{translation}
+
+%\begin{sources}[hp03_066]
+%\end{sources}
+
+\begin{testimonia}[hp03_066]
+\emph{Yogacintāmaṇi} f.~76v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{sarveṣām eva bandhānāṃ hy uttamo hy uḍḍiyānakaḥ |\\+}
+\tl{uḍḍiyāne dṛḍhe bandhe mūlaḥ svābhāviko bhavet ||\\!}
+\end{versinnote}
+
+\emph{Yogaprakāśikā} 5.97
+\begin{versinnote}
+\tl{sarveṣām eva bandhānām uttamo hy uḍḍiyāṇakaḥ |\\+}
+\tl{uḍḍiyāṇe dṛḍhe bandhe mūlaṃ svābhāvikaṃ bhavet || \\+}
+\tl{uḍḍīyānabandham upasamharati sarveṣām iti || mūlam iti mūlabandho 'nāyāsena sidhyatīty arthaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_066]
+%The reading \textit{mukti} is very well attested (alpha group, etc.) and often occurs with \textit{svābhāvikī}.
+%The reading \textit{mūla} also makes sense as it elucidates on the relationship between these two locks.
+% uḍḍiyāne kṛte bandhe mūlaṃ svābhāvikaṃ bhavet (Gr2,Gr3)
+% uḍḍiyāṇe dṛḍhe bandhe muktiḥ svābhāvikī bhavet
+%[JB: no issue now as alpha supports the readings we have adopted]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.67}
+\begin{translation}[hp03_067]
+Now the \sl{jālandhara} lock–\\
+{}[The yogi] should contract the throat and firmly place the chin on the chest. This is the lock called \sl{jālandhara}. It prevents loss of the nectar of immortality.
+\end{translation}
+
+\begin{sources}[hp03_067]
+\emph{Dattātreyayogaśāstra} 138
+\begin{versinnote}
+\tl{kaṇṭham ākuñcya hṛdaye sthāpayec cibukaṃ dṛḍham |\\+}
+\tl{jālandharo bandha eṣa amṛtāvyayakārakaḥ ||\\+}
+\tl{\var{138b sthāpayec cibukaṃ dṛḍham] HRPTPT; sthāpayed dṛḍhayā dhiyā YTU, sthāpayed dṛḍham icchayā \emph{cett.}}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_067]
+\emph{Haṭharatnāvalī} 2.66
+\begin{versinnote}
+\tl{kaṇṭham ākuñcya hṛdaye sthāpayec cibukaṃ dṛḍham |\\+}
+\tl{bandho jālandharākhyo 'yaṃ jarāmṛtyuvināśakaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~77r (attr.~to the \emph{Yogabīja})
+\begin{versinnote}
+\tl{kaṇṭham ākuñcya hṛdaye sthāpayed dṛḍham icchayā |\\+}
+\tl{bandho jālandharākhyo 'yam amṛtāvyayakārakaḥ ||\\!}
+\end{versinnote}
+
+Yogabīja 109 (south Indian recension)
+\begin{versinnote}
+\tl{kaṇṭham ākuñcya hṛdaye sthāpayed dṛḍham icchayā |\\+}
+\tl{bandho jālandharākhyo 'yam amṛtāvyayakārakaḥ ||\\+}
+\tl{\var{amṛtāvyaya°] amṛtavyaya° \vl}\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_067]
+Manuscripts of the \textalpha, \textbeta, \texteta\ and delta groups have \emph{sthāpayed dṛḍham icchayā} (`one should place it firmly as desired') in the second verse quarter, which is also well-attested in the transmission of the source text, the \emph{Dattātreyayogaśāstra}.
+This reading seems secondary because, in a subsequent verse (3.69), contracting the throat is the main feature of \emph{jālandharabandha}, so it seems contradictory to say that it may be done `as one likes' in 3.67b.
+The word \emph{icchayā} may have crept in to this verse because someone wanted to make this practice optional in light of 3.22, or it might be a corruption of \emph{hṛdaye sthāpayed dṛḍhaṃ niścayam} which is found in some other manuscripts.
+\end{philcomm}
+
+\begin{metre}[hp03_067]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.68}
+\begin{translation}[hp03_068]
+Because it binds all the channels in which the liquid from the void flows down it is [called] the \sl{jālandhara} lock. It gets rid of all problems in the throat.%
+\end{translation}
+
+\begin{sources}[hp03_068]
+\emph{Vivekamārtaṇḍa} 45
+\begin{versinnote}
+\tl{badhnāti hi śirājālam adhogāminabhojalam |\\+}
+\tl{tato jālandharo bandhaḥ kaṇṭhaduḥkhaughanāśanaḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_068]
+\emph{Haṭharatnāvalī} 2.66ef–2.67ab
+\begin{versinnote}
+\tl{badhnāti hi śirājālaṃ nādho yāti nabhojalam |\\+}
+\tl{tato jālandharo bandhaḥ kaṇṭhasaṅkocane kṛte  ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~77r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{badhnātīha śirājālam adhogāminabhojalam |\\+}
+\tl{tato jālandharaḥ proktaḥ kaṇṭhe duḥkhaughanāśanaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.230 (attr.~to the \emph{Śivayoga})
+\begin{versinnote}
+\tl{badhnāti hi śirājālaṃ nādho yāti nabhojalam |\\+}
+\tl{tato jālandharo bandhaḥ kaṇṭhaduḥkhaughanāśanaḥ ||\\!}
+\end{versinnote}
+
+%Skandapurāṇa 4.41.149
+%\begin{versinnote}
+%\tl{badhnāti hi śirājālam adhogāmi nabhojalam |\\+}
+%\tl{eṣa jālandharo bandhaḥ kaṇṭhe duḥkhaughanāśanaḥ || \\!}
+%\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_068]
+%Judit: Read adhogāminabhojalam in which the fluid of the void flows down.
+% JB: I've tweaked the translation to reflect Judit's suggestion.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.69}
+\begin{translation}[hp03_069]
+When the \sl{jālandhara} lock is performed, its defining feature being the contraction of the throat, nectar does not fall in the fire and the breath does not escape.
+\end{translation}
+
+\begin{sources}[hp03_069]
+\emph{Vivekamārtaṇḍa} 46
+\begin{versinnote}
+\tl{jālandhare kṛte bandhe kaṇṭhasaṃkocalakṣaṇe |\\+}
+\tl{na pīyūṣaṃ pataty agnau na ca vāyuḥ pradhāvati ||\\+}
+\tl{\var{pradhāvati ] ; prakupyati AGT}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_069]
+\emph{Yogacintāmaṇi} f.~77v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{jālandhare kṛte bandhe kaṇṭhasaṃkocalakṣaṇe |\\+}
+\tl{na pīyūṣaṃ pataty agnau na ca vāyuḥ prakupyati ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.231 (attr.~to the \emph{Śivayoga})
+\begin{versinnote}
+\tl{jālandhare kṛte bandhe kaṇṭhasaṃkocalakṣaṇe |\\+}
+\tl{na pīyūṣaṃ pataty agnau na ca vāyuḥ prakupyati ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_069]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.70}
+\begin{translation}[hp03_070]
+By contracting the throat, the yogi firmly blocks the two channels. This should be known as the middle cakra, which binds [the mind to] the sixteen supports [in the body].
+\end{translation}
+
+%\begin{sources}[hp03_070]
+%\end{sources}
+
+\begin{testimonia}[hp03_070]
+\emph{Yogaviṣaya} 19ab % probably not a source text as it seems to be a fragment
+\begin{versinnote}
+\tl{kaṇṭhasaṃkocanaṃ kṛtvā dve nāḍyau stambhayed dṛḍham |\\+}
+\tl{rasanāpīḍyamānās tu ṣoḍaśaś cordhvagāminī ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~77v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{madhyacakram idaṃ jñeyaṃ ṣoḍaśādhārabandhanam |\\!}
+\end{versinnote}
+
+\emph{Yogakarṇikā} 85
+\begin{versinnote}
+\tl{kaṇṭhasaṅkocanenaiva dve nāḍye kumbhayed dṛḍham |\\!}
+\end{versinnote}
+
+\emph{Yogasarasaṅgraha} p.58
+\begin{versinnote}
+\tl{kaṇṭhasaṃkocanekaṇṭhasaṃkocane nanaiva dvināḍyo stambhayed dṛḍham~|\\+}
+\tl{ayaṃ bandho mayā proktaḥ ṣoḍaśādhārabandhanam ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 23
+\begin{versinnote}
+\tl{kaṇṭhasaṅkocanaṃ kṛtvā cibukaṃ hṛdaye nyaset |\\+}
+\tl{jālandhare kṛte bandhe ṣoḍaśādhārabandhanam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_070]
+The import of the second line of this verse is obscure to us.
+In \emph{Jyotsnā} 3.73, Brahmānanda thinks that the middle cakra (\emph{madhyacakra}) is \emph{viśuddha} cakra.
+The main reason for this appears to be that this cakra is located in the throat and the salient feature of the \sl{jālandhara} lock is contracting the throat.
+However, he also seems to connect the \emph{viśuddha} cakra to the sixteen supports (\emph{ṣoḍaśādhāra}) at the end of this verse, perhaps because this cakra has sixteen petals (as mentioned in 3.47).
+On the meaning of \emph{ādhāra} in yogic contexts, see entry no.~3 in the \emph{Tāntrikābhidhānakośa} vol.~1 2000: 191.
+%Jyotsnā: kaṇṭhasaṃkocaneneti.
+dṛḍham gāḍhaṃ kaṇṭhasaṃkocanenaiva kaṇṭhasaṃkocanamātreṇa dve nāḍyau iḍāpiṅgale stambhayed bandhayet.
+ayaṃ jālandhara iti kartṛpadādhyāhāraḥ.
+idaṃ kaṇṭhasthāne sthitaṃ viśuddhākhyaṃ cakram madhyacakram madhyamaṃ cakram jñeyam | kīdṛśam ?
+ṣoḍaśādhārabandhanam ṣoḍaśasaṅkhyākā ye ādhārā aṅguṣṭhādhārādibrahmarandhrāntās teṣāṃ bandhanaṃ bandhanaprakāram |
+%Yogaprakāśikā: uktalakṣaṇaṃ nāḍīstambhanaṃ madhyacakrasaṃjñakaṃ bhavati || siddhasiddhāntapaddhatau pratipāditānāṃ ṣoḍaśādhārāṇāṃ nirodhane sādhanaṃ ca ||
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.71}
+\begin{translation}[hp03_071]
+This triad of locks is the best [and] has been practised by the great Siddhas. Yogis know it to be a method of all systems of Haṭha.
+\end{translation}
+
+%\begin{sources}[hp03_071]
+%\end{sources}
+
+\begin{testimonia}[hp03_071]
+\emph{Haṭharatnāvalī} 2.68
+\begin{versinnote}
+\tl{bandhatrayam idaṃ śreṣṭhaṃ mahāsiddhaiś ca sevitam |\\+}
+\tl{sarveṣāṃ yogatantrāṇāṃ sādhanaṃ yogino viduḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~77v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{bandhatrayam idaṃ śreṣṭhaṃ mahāsiddhaniṣevitam |\\+}
+\tl{sarveṣāṃ haṭhatantrāṇāṃ sādhane yoginām iti ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 15.24
+\begin{versinnote}
+\tl{idaṃ bandhatrayaṃ śreṣṭhaṃ marujjayasusiddhadam |\\+}
+\tl{sarveṣāṃ yogatantrāṇāṃ sādhanaṃ yogino viduḥ ||\\!}
+\end{versinnote}
+
+%Yogakarṇikā 85
+%\begin{versinnote}
+%\tl{bandhatrayam idaṃ guhyaṃ mahāsiddhiniṣevitam || \\!}
+%\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_071]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.71*1}
+\begin{translation}[hp03_071_1]
+By immediately contracting the lower [part of the body] (i.e.~by the root lock) when the neck has been contracted (i.e.~by the \emph{jālandhara} lock) and by stretching the abdomen backwards in the middle [of the body] (i.e.~by the \emph{uḍḍiyāna} lock), the breath enters the channel of Brahman.
+\end{translation}
+% greyscaled as alpha
+
+\begin{sources}[hp03_071_1]
+\emph{Gorakṣaśataka} 63 (see 2.46)
+\end{sources}
+
+\begin{testimonia}[hp03_071_1]
+\emph{Haṭharatnāvalī} 2.8, \emph{Yogacintāmaṇī} f.80r, \emph{Yuktabhavadeva} 7.95 and \emph{Haṭhatattvakaumudī} 15.25–27 (see 2.46).
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_071_1]
+%In the second chapter, Group 2 has the reading \emph{kuñcanenāśu kaṇṭhasaṃkocane kṛte}.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.72}
+\begin{translation}[hp03_072]
+[The yogi] should contract the place of the root  and do the \emph{uḍḍiyāna} [lock]. He should [then] block the \emph{iḍā} and \emph{piṅgalā} [channels] and make [the breath] flow in the rear pathway.
+\end{translation}
+
+%\begin{sources}[hp03_072]
+%\end{sources}
+
+\begin{testimonia}[hp03_072]
+\emph{Haṭharatnāvalī} 2.70
+\begin{versinnote}
+\tl{mūlasthānaṃ samākuñcya uḍḍiyānaṃ tu kārayet |\\+}
+\tl{iḍāṃ ca piṅgalāṃ baddhvā vāhayet paścimaṃ pathaṃ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~79v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mūlasthānaṃ samākṛṣya uḍḍiyānaṃ tu kārayet |\\+}
+\tl{iḍāṃ ca piṅgalāṃ baddhvā vāhayet paścime pathi ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 15.25–27
+\begin{versinnote}
+\tl{mūlasthānaṃ samākuñcya uḍyānaṃ tu kārayet |\\+}
+\tl{iḍāṃ ca piṃgalāṃ baddhvā vāhayet paścimāpathaṃ ||\\+}
+\tl{mūlasthānaṃ mārgasaṃkocanaṃ vidhāya uḍyānam udarasaṃkocanaṃ tataḥ | iḍāṃ piṃgalāṃ baddhvā kaṇṭhasaṃkocanena paścimapathaṃ pṛṣṭhavaṃśamārge pavanaṃ vāhayet kuryāt ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_072]
+%No known source?
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.73}
+\begin{translation}[hp03_073]
+By this method alone, the breath attains dissolution. Then death does not arise nor old age, disease and the like.
+\end{translation}
+
+\begin{sources}[hp03_073]
+\end{sources}
+
+\begin{testimonia}[hp03_073]
+\emph{Haṭharatnāvalī} 2.71
+\begin{versinnote}
+\tl{anenaiva vidhānena prayāti pavano layam |\\+}
+\tl{tato na jāyate mṛtyur jarārogādikaṃ tathā || \\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~79v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{anenaiva vidhānena sevayet pavanālayam |\\+}
+\tl{tato na jāyate mṛtyur jarārogādikaṃ tathā ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_073]
+%The \textalpha, \textbeta\ and delta groups have the reading \emph{sevayet pavanālayam} as the second verse quarter of this verse.
+It renders the meaning, `by this method alone, one should honour the abode of the breath.' As far as we know, the compound \emph{pavanālaya} does not occur in other yoga texts.
+The similar compound \emph{prāṇālaya}) is mentioned in other yoga texts, such as the \emph{Yogayājñavalkya} (4.52–53), but it refers to the locations in the body where \emph{prāṇa} resides, as opposed to the other bodiy winds.
+% JB: reconsider the alpha reading, which is well attested in other groups? Maybe \emph{sevayet pavanālayam} is just some general laudatory statement?
+%Judit: ādikam is okay.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.73*1}
+\begin{translation}[hp03_073_1]
+Now the inverted bodily position–\\
+The sun devours whatever divine nectar flows from the moon. As a result, the body perishes.
+\end{translation}
+%3.75-76 gray-scaled.
+
+%\begin{sources}[hp03_073_1]
+%\end{sources}
+
+\begin{testimonia}[hp03_073_1]
+\emph{Haṭharatnāvalī} 2.72 
+\begin{versinnote}
+\tl{atha viparītakaraṇī–\\+}
+\tl{yat kiñ cit sravate candrād amṛtaṃ divyarūpi ca |\\+}
+\tl{tatsarvaṃ grasate sūryas tena piṇḍaṃ vināśi ca ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~77v
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{yat kiṃ cin sravate candrād amṛtaṃ divyarūpi ca |\\+}
+\tl{tat sarvaṃ grasate sūryas tena piṇḍaṃ vināśi ca ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Haṭhayogasaṃhitā} 38 (p. 26)
+\begin{versinnote}
+\tl{nābhimūle vaset sūryas tālumūle ca candramāḥ |\\+}
+\tl{amṛtaṃ grasate sūryas tato mṛtyuvaśo naraḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_073_1]
+%No source?
+%The term \emph{piṇḍa} is rarely neuter, hence the change to the masculine in the \emph{Jyotsnā}.
+%Both ca-s are infelicitous J5 has jīryate for vināśi ca
+The \textalpha\ group do not have 3.73*1 and 3.73*2 in the third chapter (but rather in the fourth) and other manuscripts omit them as well (notably the \texteta\ group).
+It appears that they have been inserted at the beginning of the section on \emph{viparītakaraṇī} as a kind of preamble, which is unusual as the other techniques in this chapter do not have such introductions.
+
+%are found in alpha in ch.4, but should they be there?
+They are very physical and not suited to ch.4.
+But here they are unusual because they are preamble and the practices already described do not have preamble.
+Perhaps drop them altogether.
+But then there is no explanation of how VK works.
+
+%MD: 3.75-76 gray-scaled. Keep \textit{divyarūpi ca} und \textit{vināśi ca} or change them to \textit{divyarūpiṇaḥ} and \textit{jarāyutam}?
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.73*2}
+\begin{translation}[hp03_073_2]
+There is a divine bodily position for this, which blocks the mouth of the sun. It is to be known from the teaching of a guru and not through the countless interpretations of scriptures.
+\end{translation}
+% greyscaled
+
+\begin{sources}[hp03_073_2]
+\end{sources}
+
+\begin{testimonia}[hp03_073_2]
+Haṭharatnāvali 2.73 (on \emph{viparītakaraṇī})
+\begin{versinnote}
+\tl{tatrāsti divyaṃ karaṇaṃ sūryasya mukhabandhanam |\\+}
+\tl{gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ || \\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~77v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tatrāsti karaṇaṃ divyaṃ sūryasya mukhabandhanam |\\+}
+\tl{gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ ||\\!}
+\end{versinnote}
+
+%Yogakarṇikā 147c–148b
+%\begin{versinnote}
+%\tl{tat sarvaṃ grasate sūryas tena piṇḍo jarāyutaḥ || 147 ||\\+}
+%\tl{tatrāsti karaṇaṃ divyaṃ sūryasya mukhavañcanam |\\!}
+%\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_073_2]
+%No source? Only Brahmānanda reads mukhavañcanam, which makes better sense. Judit; bandhanaṃ makes sense with grasate sūryas
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.74}
+\begin{translation}[hp03_074]
+The navel is up, the palate down; the sun up, the moon down: the bodily position called “inverted” is obtained through the teaching of a guru.
+\end{translation}
+%
+
+\begin{sources}[hp03_074]
+\emph{Vivekamārtaṇḍa} 115
+\begin{versinnote}
+\tl{ūrdhvaṃ nābhir adhas tālur ūrdhvaṃ bhānur adhaḥ śaśī |\\+}
+\tl{karaṇī viparītākhyā guruvākyena labhyate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_074]
+\emph{Haṭharatnāvalī} 2.74
+\begin{versinnote}
+\tl{ūrdhvaṃ nābhir adhas tālur ūrdhvaṃ bhānur adhaḥ śaśī |\\+}
+\tl{karaṇī viparītākhyā guruvākyena labhyate ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~73r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ūrdhvanābhir adhastālur ūrdhvabhānur adhaḥśaśī |\\+}
+\tl{karaṇī viparītākhyā sarvavyādhivināśinī ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.236 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{ūrdhvaṃ nābhir adhas tālur ūrdhvaṃ bhānur adhaḥ śaśī |\\+}
+\tl{karaṇī viparītākhyā guruvaktreṇa gamyate  ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_074]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.75}
+\begin{translation}[hp03_075]
+The bodily position called “inverted" destroys all diseases. For [the yogi] who regularly engages in [its] practice, it increases the digestive fire.
+\end{translation}
+
+\begin{sources}[hp03_075]
+\emph{Dattātreyayogaśāstra} 146
+\begin{versinnote}
+\tl{karaṇaṃ viparītākhyaṃ sarvavyādhivināśanam |\\+}
+\tl{nityam abhyāsayuktasya jaṭharāgnir vivardhate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_075]
+\emph{Haṭharatnāvalī} 2.75
+\begin{versinnote}
+\tl{karaṇī viparītākhyā sarvavyādhivināśinī |\\+}
+\tl{nityam abhyāsayuktasya jaṭharāgnivivardinī  ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~78r (attr.~to Dattātreya)
+\begin{versinnote}
+\tl{nityam abhyāsayuktasya jaṭharāgnivivardhanam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_075]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.76}
+\begin{translation}[hp03_076]
+A lot of food should be provided for the practitioner. If the practitioner eats little, the fire will quickly consume his body.
+\end{translation}
+
+\begin{sources}[hp03_076]
+\emph{Dattātreyayogaśāstra} 147
+\begin{versinnote}
+\tl{āhāro bahulas tasya saṃpādyaḥ sāṃkṛte dhruvam |\\+}
+\tl{alpāhāro yadi bhaved agnir dehaṃ dahet kṣaṇāt ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_076]
+
+\emph{Haṭharatnāvalī} 2.76
+\begin{versinnote}
+\tl{āhāro bahulas tasya sampādyaḥ sādhakena vai |\\+}
+\tl{alpāhāro yadi bhaved deham agnir dahet kramāt ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~78r (attr.~to Dattātreya)
+\begin{versinnote}
+\tl{āhāro bahulas tasya saṃpādyaḥ sāṃkṛte dhruvam |\\+}
+\tl{alpāhāro yadi bhaved agnir dāhaṃ karoti vai ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Yuktabhavadeva} 7.238
+\begin{versinnote}
+\tl{asyāṃ kriyamāṇāyāṃ sādhakasya bhakṣyaṃ bahulaṃ sampādyam anyathā pravṛddho jāṭharānalo dhātuṃ dahatīti ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_076]
+Svātmārāma has removed the vocative from the \emph{Dattātreyayogaśāstra}, changing \emph{sāṃkṛte dhruvam} to \emph{sādhakasya tu}.
+\end{philcomm}
+
+\begin{metre}[hp03_076]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.77}
+\begin{translation}[hp03_077]
+On the first day [the yogi] should keep his head down and his feet up for a short while, and he should [then] practise for a little longer every day.
+\end{translation}
+
+\begin{sources}[hp03_077]
+\emph{Dattātreyayogaśāstra} 148c–149b
+\begin{versinnote}
+\tl{adhaḥśirāś cordhvapādaḥ kṣaṇaṃ syāt prathame dine ||\\+}
+\tl{kṣaṇāc ca kiṃ cid adhikam abhyasec ca dine dine |\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_077]
+\emph{Haṭhratnāvalī} 2.77
+\begin{versinnote}
+\tl{adhaḥ śiraś cordhvapādau kṣaṇaṃ syāt prathame dine |\\+}
+\tl{kṣaṇāc ca kiñ cid adhikam abhyasec ca dine dine ||\\+}
+\tl{\var{cordhvapādau ] cordhvapādaḥ \vl}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~78r (attr.~to Dattātreya)
+\begin{versinnote}
+\tl{adhaḥśirāś cordhvapādaḥ kṣaṇaṃ syāt prathame dine |\\+}
+\tl{kṣaṇāc ca kiñ cid adhikam abhyasec ca dine dine ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Yuktabhavadeva} 7.237
+\begin{versinnote}
+\tl{sa ca prathamadine kṣaṇamātraṃ vidheyā dvitīyadine | kiñcidadhikaṃ kālam evaṃ yāmaparyantaṃ vidheyā |\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_077]
+%\end{philcomm}
+
+\begin{metre}[hp03_077]
+Anuṣṭubh (a: ra-vipulā; c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.78}
+\begin{translation}[hp03_078]
+After six months grey hair and wrinkles disappear. [The yogi] who regularly practises for three hours conquers death.
+\end{translation}
+
+\begin{sources}[hp03_078]
+\emph{Dattātreyayogaśāstra} 149c–150b
+\begin{versinnote}
+\tl{valiś ca palitaṃ caiva ṣaṇmāsordhvaṃ na dṛśyate ||\\+}
+\tl{yāmamātraṃ hi yo nityam abhyaset sa tu kālajit |\\+}
+\tl{\var{°māsordhvaṃ na ] °māsāṃ hi na M1, °māsāc ca na AM2, °māsārdhān na YTU, °māsān na tu HR , °māsārddhena \emph{Yogacintāmaṇi} 150b  kālajit ] yogavit πDYŚPT}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_078]
+\emph{Haṭharatnāvalī} 2.78
+\begin{versinnote}
+\tl{valitaṃ palitaṃ caiva ṣaṇmāsān na tu dṛśyate |\\+}
+\tl{yāmamātraṃ tu yo nityam abhyaset sa tu kālajit || \\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~78r (attr.~to Dattātreya)
+\begin{versinnote}
+\tl{valiś ca palitaṃ caiva ṣaṇmāsārdhe na dṛśyate |\\+}
+\tl{yāmamātraṃ tu yo nityam abhyaset sa tu kālajit ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.238 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{valitaṃ palitaṃ caiva ṣaṇmāsārdhān na dṛśyate |\\+}
+\tl{yāmamātraṃ tu yo nityam abhyaset sa tu kālajit ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Haṭhatattvakaumudī} 14.3
+ \begin{versinnote}
+\tl{ūrdhvapādo hy adhomastakaḥ syāt kṣaṇaṃ\\+}
+\tl{vāsare 'thādime 'bhyāsaṃ vṛddhyā dhayet |\\+}
+\tl{evam abhyāsato yāmamātraṃ sadā \\+}
+\tl{mṛtyujit syāj jarājic ca ṣaṇmāsataḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_078]
+We have adopted the reading \emph{ṣaṇmāsordhvaṃ} in the second verse quarter.
+It is attested by manuscripts of the \emph{Dattātreyayogaśāstra} (the source text) and the \emph{Jyotsnā} (3.82).
+It makes good sense and explains the rather odd readings in \textalpha\ and other manuscripts, \emph{ṣaṇmāsārdhān}, \emph{ṣaṇmāsārdhaṃ} and \emph{ṣaṇmāsārdhe}.
+The \textgamma\ and delta groups have a different verb as well, \emph{ṣaṇmāsārdhena naśyati}.
+The original reading was likely \emph{ṣaṇmāsordhvaṃ na dṛśyate} because the compound \emph{ṣaṇmāsārdha} (`half of six months') is very strange and \emph{dṛśyate} is better attested.
+
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.79}
+\begin{translation}[hp03_079]
+Now \emph{vajrolī}:\\
+Even if he behaves as he wishes without [following] the observances (\emph{niyama}) taught in yoga, the [yogi] who knows \emph{vajrolī} is worthy of success.
+\end{translation}
+
+\begin{sources}[hp03_079]
+\emph{Dattātreyayogaśāstra} 152
+\begin{versinnote}
+\tl{svecchayā varttamāno ’pi yogoktaniyamair vinā |\\+}
+\tl{vajroliṃ yo vijānāti sa yogī siddhibhājanaḥ ||\\+}
+\tl{\var{152d °bhājanaḥ ] °mān bhavet M1AM2, °bhājanam YTU}\\!}
+\end{versinnote}
+
+Cf.~\emph{Śivasaṃhitā} 4.79
+\begin{versinnote}
+\tl{svecchayā vartamāno 'pi yogoktaniyamair vinā |\\+}
+\tl{mukto bhaved gṛhastho'pi vajrolyabhyāsayogataḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_079]
+\emph{Haṭharatnāvalī} 2.79 (on \emph{viparītakaraṇī})
+\begin{versinnote}
+\tl{svasthaṃ yo vartamāno 'pi yogoktair niyamair vinā |\\+}
+\tl{karaṇī viparītākhyā śrīnivāsena lakṣitā ||\\!}
+\end{versinnote}
+
+\emph{Yogalakṣaṇāvalī} f.~31r
+\begin{versinnote}
+\tl{svecchayā vartamāno 'pi yogoktaniyamair vinā |\\+}
+\tl{vajrolyabhyāsayogena yogī siddhim avāpnuyāt ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 38
+\begin{versinnote}
+\tl{svecchayā varttamāno 'pi yogoktair niyamair vinā |\\+}
+\tl{vajrolīṃ yo vijānāti sa yogī siddhibhājanam || \\!}
+\end{versinnote}
+
+Cf.~\emph{Yuktabhavadeva} 7.240 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{vajrolīṃ kathayiṣyāmi gopitāṃ sarvayogibhiḥ|\\+}
+\tl{tyaktayogoktaniyamā yayā sidhyanti yoginaḥ || \\!}
+\end{versinnote}
+
+Cf.~\emph{Haṭhatattvakaumudī} 16.3
+\begin{versinnote}
+\tl{svecchayā varttamāno 'pi yogoditaiḥ\\+}
+\tl{sadvidhānair vinā sādhakaḥ sābalaḥ |\\+}
+\tl{mucyate 'sau suvajrolikābhyāsataḥ\\+}
+\tl{sarvasiddhyāspadaṃ yāti bhūmaṇḍale ||\\!}
+\end{versinnote}
+
+%Vajroliyoga 1
+%\begin{versinnote}
+%\tl{svecchayā vartamāno ’pi yogoktair niyamair vinā |\\+}
+%\tl{vajrolī yo vijānāti sa yogī siddhibhājanaṃ ||\\!}
+%\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_079]
+In manuscripts of the delta group, the \emph{vajrolī} section is placed at the end of the work and the following comment is inserted at this place in the third chapter:
+\begin{versinnote}
+\emph{Vajrolī}, which is [usually] here, has been copied at the end of the text. Even though it is found in the sequence here, it has been left out because it is be practised by special individuals.\\
+\emph{atratyā vajrolī granthānte likhitā} | \emph{kramaprāptāpy atra tyaktā} | \emph{asādhāraṇaprāṇyanuṣṭheyatvāt tasyāḥ} |
+\end{versinnote}
+
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.80}
+\begin{translation}[hp03_080]
+I shall teach you two substances [needed] for it which are hard for just anyone to obtain. One is milk and the second is an obedient woman.
+\end{translation}
+
+\begin{sources}[hp03_080]
+\emph{Dattātreyayogaśāstra} 153ab-154ab
+\begin{versinnote}
+\tl{tatra vastudvayaṃ vakṣye durlabhaṃ yena kena cit |\\+}
+\tl{kṣīraṃ caikaṃ dvitīyaṃ ca nārī ca vaśavartinī |\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_080]
+\emph{Yuktabhavadeva} 7.241 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{atra vastudvayaṃ manye durlabhaṃ yasya kasyacit |\\+}
+\tl{kṣīram ekaṃ dvitīyaṃ tu nārī svavaśavarttinī || \\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 39
+\begin{versinnote}
+\tl{tatra vastudvayaṃ vakṣye durlabhaṃ yasya kasya cit |\\+}
+\tl{kṣīraṃ caikaṃ dvitīyaṃ tu nārī ca vaśavarttinī || \\!}
+\end{versinnote}
+
+%Vajroliyoga 2
+%\begin{versinnote}
+%\tl{tatra vastudvayaṃ vakṣye durlabhaṃ yasya kasya cit |\\+}
+%\tl{kṣīraṃ caikaṃ dvitīyaṃ tu nārī ca vaśavartinī ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_080]
+On the possible referents of \emph{kṣīra}, see Mallinson 2024 on \emph{Dattātreyayogaśāstra} 154.
+According to Brahmānanda (\emph{Jyotsnā} 3.84), the compound \emph{vaśavartinī}, which we have translated as `an obedient woman,' could be a wife (\emph{vaśavartinī svādhīnā nārī vanitā}).
+In 3.83, the reading \emph{bhāryābhage} in \getsiglum{V1}, \getsiglum{V3} and \getsiglum{J10} supports Brahmānanda's view that the woman is the yogi's wife.
+%Is kṣīra semen? Or a sap that is hard to find? Difficult to say since nothing more is said about kṣīra.
+%In DYŚ it's kṣīra and aṅgirasa, so maybe this is addressed to male and female practitioners, who can get one or the other. But what about the celibate yogi?
+%LO: in Bengali texts vastu can mean sexual fluid.
+% JB should we mention that 3.80–83ab are missing from Alpha Three. Also, Alpha Three is missing 3.84ab, 3.86cd
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.81}
+\begin{translation}[hp03_081]
+[The yogi] should gently practise a full upward contraction through the urethra. Either a man or a woman may obtain success in \emph{vajrolī}.
+\end{translation}
+
+%\begin{sources}[hp03_081]
+%\end{sources}
+
+\begin{testimonia}[hp03_081]
+\emph{Haṭhatattvakaumudī} 16.4
+\begin{versinnote}
+\tl{apānamārgataḥ samyag ūrdhvakuñcanam abhyaset |\\+}
+\tl{puruṣo vāpi nārī vā vajrolīsiddhibhājanam ||\\+}
+\tl{apānamārgato gudadeśena ūrdhvam upari kuñcanaṃ saṃkocanam ūrdhvam ākarṣaṇaṃ vā abhyaset || iti ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} 53 (p. 39)
+\begin{versinnote}
+\tl{mehanena śanaiḥ samyag ūrdhvākuñcanam abhyaset |\\+}
+\tl{puruṣo 'py athavā nārī vajrolīsiddhim āpnuyāt ||\\!}
+\end{versinnote}
+
+%Vajroliyoga 3
+%\begin{versinnote}
+%\tl{mehanena śanaiḥ samyag ūrdhvākuñcanam abhyaset |\\+}
+%\tl{puruṣo vāpi nārī vā vajrolīsiddhim āpnuyāt ||\\!}
+%\end{versinnote}
+
+\emph{Yogaprakāśikā} 118ab
+\begin{versinnote}
+\tl{mehanena śanaiḥ samyag ūrdhvaṃ kuñcanam abhyaset |\\+}
+\tl{ūrdhvaṃ yathā syāt tathā bindor ākarṣaṇaṃ meḍhreṇābhyased ity arthaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_081]
+ The \sl{Haṭhatattvakaumudī} says that this upward contraction of the urethra, which is the method by which fluids are drawn up it, is done in the region of \emph{apānavāyu} and the anus (\emph{gudadeśa}).
+Brahmānanda states that this practice is done immediately after sex (\emph{strīsaṅgānantaram}).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.82}
+\begin{translation}[hp03_082]
+Carefully using a hollow stalk of bamboo grass, [the yogi] should very gently blow into the opening of the penis in order to make air move [into the urethra].
+\end{translation}
+
+\begin{sources}[hp03_082]
+\emph{Dattātreyayogaśāstra} 165
+\begin{versinnote}
+\tl{tatas tu śaranālena phūtkāraṃ vajrakandare |\\+}
+\tl{śanaiḥ śanaiḥ prakurvīta vāyusaṃcārakāraṇāt ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_082]
+\emph{Haṭharatnāvalī} 2.86–2.87
+\begin{versinnote}
+\tl{haṭhapradīpikākāras tu\\+}
+\tl{yatnataḥ śaranālena phūtkāraṃ vajrakandare |\\+}
+\tl{śanaiḥ śanaiḥ prakurvīta vāyusaṃcārakāraṇāt ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (f. 39r)
+\begin{versinnote}
+\tl{taduktaṃ haṭhapradīpikāyāṃ \\+}
+\tl{yantritaḥ śaranālena phūtkāraṃ vajrakandare |\\+}
+\tl{śanaiḥ śanaiḥ prakurv[ī]ta vāyusaṃcārakāraṇād iti ||\\+}
+\tl{asyārthaḥ ||\\+}
+\tl{ṣoḍaśāṃgulamānāṃ 16 tu prakuryād vaṃśanālikāṃ sūkṣmā'g[r]amūlāntāṃ li[ṃ]gachi[dra]mukhe datvā svāsye 'nu tanmukhaṃ dhṛtvā phūtkāram ante syāḥ k[u]ryād bāḍhaṃ muhur muhuḥ pratyahaṃ tena vivṛtaṃ liṅgadvāraṃ kramād bhavet [|] \\+}
+\tl{tato †nālyānayāto† yam alpaṃ phūtkārato 'ntare [|]\\+}%nālpānn apāno
+\tl{liṅgara[n]dhreṇa gṛhṇīyāt kramavṛddhyā susādhakaḥ [|]\\+}
+\tl{liṃgachidre 'tha vivṛte kṣīrākṛṣṭiṃ tato bhajed iti [|] \\+}
+\tl{vajrakandare liṅgachidre [||]\\!}
+\end{versinnote}
+
+\emph{Yogaprakāśikā} 118cd–ef
+\begin{versinnote}
+\tl{yatnataḥ śaranālena phūtkāraṃ vajrakandare |\\+}
+\tl{śanaiḥ śanaiḥ prakurvīta vāyusañcārakāraṇāt ||\\+}
+\tl{śareti meḍhranālenety arthaḥ || vāyusañcārakāraṇam iti bindor ākarṣaṇaṃ kāraṇam ity arthaḥ \\!}
+\end{versinnote}
+
+Cf.~\emph{Yuktabhavadeva} 7.248cd--249ab
+\begin{versinnote}
+\tl{rasanālena phūtkāraṃ vāyoḥ sañcārakāraṇāt || \\+}
+\tl{kuryāt śanaiḥ śanair yogī yāvac chaktiḥ prajāyate |\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_082]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.83}
+\begin{translation}[hp03_083]
+With practice, [the yogi] may draw up semen which is falling into a woman’s vagina. And [even] if his own semen has moved [down], he may draw it upwards and retain it.
+\end{translation}
+
+\begin{sources}[hp03_083]
+\emph{Dattātreyayogaśāstra} 166
+\begin{versinnote}
+\tl{tadbhage patitaṃ bindum abhyāsenordhvam āharet | \\+}
+\tl{calitaṃ ca tathā bindum ūrdhvam ākṛṣya rakṣayet ||\\!}
+\end{versinnote}
+% there are differences bw the DYS and HP in the 1st and 3rd pādas, but the HP text is solid
+\end{sources}
+
+\begin{testimonia}[hp03_083]
+\emph{Haṭharatnāvalī} 2.96cd--2.97ab
+\begin{versinnote}
+\tl{nāryā bhagāt patadbindum abhyāsenordhvam āharet ||\\+}
+\tl{calitaṃ ca nijaṃ bindum ūrdhvam ākṛṣya rakṣayet |\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 39
+\begin{versinnote}
+\tl{nārībhage pated bindum abhyāsenordhvam āharet |\\+}
+\tl{calitaṃ ca nijaṃ bindum ūrdhvam ākṛṣya rakṣayet ||\\!}
+\end{versinnote}
+
+\emph{Yogaprakāśikā} 5.120
+\begin{versinnote}
+\tl{nāryā bhage patadbindum abhyāsenordhvam āharet |\\+}
+\tl{calitaṃ ca svayaṃ bindum ūrdhvam ākṛṣya rakṣayet ||\\+}
+\tl{nārīsaṃyoge bindupatanaṃ syād ity āśaṅkya nirasyati nāryā iti || patato bindor ūrdhvam āhared āhīyamāṇaṃ svayaṃ calitaṃ bindum ākṛṣyety anvayaḥ || \\!}
+\end{versinnote}
+
+Cf.~\emph{Haṭhasaṅketacandrikā} f.~39r
+\begin{versinnote}
+\tl{apānam ākuñcya tato 'balenordhvaṃ dugdham ākṛṣṭividhikrameṇa |\\+}
+\tl{samabhyasen niścalam alpam alpaṃ bhage patadbindum athārdhvam āharet ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Yuktabhavadeva} 7.249cd, 259
+\begin{versinnote}
+\tl{tato maithunakāle tu patadbinduṃ samunnayet ||\\+}
+\tl{{[}...] patadbindum apānena huṃ huṃkārasahitena balād ūrdhvam ākṛṣya kiñcit kālaṃ vilambya ramet punaḥ || yadā tu na dhārayituṃ śakyate tadā bahiḥskhalitena bindunā saha prasvedenāṅgaṃ marddayet ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_083]
+%The reading \emph{bhāryābhage} in V1, V3 and J10 affirms Brahmānanda's view that the women is the yogi's wife. [JB: I've moved this comment to our commentary on 3.83.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.84}
+\begin{translation}[hp03_084]
+[If] the knower of yoga preserves his semen thus, he conquers death. Death arises through the loss of semen and life from retaining semen.
+\end{translation}
+
+\begin{sources}[hp03_084]
+\emph{Dattātreyayogaśāstra} 167
+\begin{versinnote}
+\tl{evaṃ ca rakṣito bindur mṛtyuṃ jayati tattvataḥ | \\+}
+\tl{maraṇaṃ bindupātena jīvanaṃ bindudhāraṇāt ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Amṛtasiddhi} 3.87cd
+\begin{versinnote}
+\tl{maraṇaṃ bindupātena jīvanaṃ bindudhāraṇāt ||\\+}
+\end{versinnote}
+
+\end{sources}
+
+\begin{testimonia}[hp03_084]
+\emph{Haṭhratnāvalī} 2.97cd-2.98ab
+\begin{versinnote}
+\tl{evaṃ saṃrakṣayed binduṃ mṛtyuṃ jayati yogavit ||\\+}
+\tl{maraṇaṃ bindupātena jīvitaṃ bindudhāraṇāt |\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 252cd-253ab
+\begin{versinnote}
+\tl{evaṃ bindau sthire jāte mṛtyuṃ jayati sarvathā ||\\+}
+\tl{maraṇaṃ bindupātena jīvanaṃ bindudhāraṇāt |\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 39
+\begin{versinnote}
+\tl{evaṃ saṃrakṣayed binduṃ mṛtyuṃ jayati yogavit |\\+}
+\tl{maraṇaṃ bindupātena jīvanaṃ bindudhāraṇāt || \\!}
+\end{versinnote}
+
+%Vajroliyoga 22
+%\begin{versinnote}
+%\tl{evaṃ bindau sthire yāte mṛtyuṃ jayati sarvathā |\\+}
+%\tl{maraṇaṃ bindupātena jīvanaṃ bindudhāraṇāt ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_084]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.85}
+\begin{translation}[hp03_085]
+As a result of the retention of semen, the yogi's body becomes fragrant. As long as semen is steady in the body then why fear death?%
+\end{translation}
+
+\begin{sources}[hp03_085]
+\emph{Dattātreyayogaśāstra} 86cd:
+\begin{versinnote}
+\tl{yogino ’ṅge sugandhaḥ syāt satataṃ bindudhāraṇāt ||\\!}
+\end{versinnote}
+
+\emph{Vivekamārtaṇḍa} 52ad
+\begin{versinnote}
+\tl{yāvad binduḥ sthito dehe tāvad mṛtyubhayaṃ kutaḥ |\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_085]
+\emph{Haṭharatnāvalī} 2.112ab
+\begin{versinnote}
+\tl{sugandhir yogino dehe jāyate bindudhāraṇāt ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 39
+\begin{versinnote}
+\tl{sugandho yogino dehe jāyate bindudhāraṇāt |\\+}
+\tl{yāvad binduḥ sthiro dehe tāvat kālabhayaṃ kutaḥ || \\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 16.10
+\begin{versinnote}
+\tl{tathā coktaṃ granthāntare –\\+}
+\tl{calitaṃ tu svakaṃ bindum ūrdhvam ākuñcya rakṣayet |\\+}
+\tl{sugandho yogināṃ dehe jāyate bindudhāraṇād || iti ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_085]
+The omission of 3.85ab in the \texteta\ group and \deltaThree\ is likely to be the result of haplography (\emph{bindudhāraṇāt} is repeated).
+% MD: I have deleted "in \alphaThree", because the damaged part has enough space for 3.84cd and 3.85ab.
+
+%sthito makes better sense (adopt?), but J5 has sthiro (missing in N3) nad G4 has kṣīro  (closer to sthiro)
+%MD: Does 4.88 speak in favor of sthiro? Or is this bindu something different?
+% JB: it looks like \alphaTwo and other mss are preserving a version of 3.85 where deha is neuter (I've come across this in the early recension of the Yogabīja.) maybe we should consider it: J5 sugandhi yogino dehaṃ jāyate bindudhāraṇāt.
+The J5 reading means that we dont have to emend to °gandhir
+
+The readings \emph{mṛtyubhayaṃ} (\alphaThree, \textbeta, \texteta) and \emph{kālabhayaṃ} (\alphaTwo, \textgamma) are well attested by the main manuscript groups, but \emph{mṛtyubhayaṃ} is in the important witnesses of the source text, the \emph{Vivekamārtaṇḍa}.
+% MD: Which reading does ms T have?
+%  JB: record that alphaThree has mṛtyubhayaṃ in 85d? alphaTwo has kālabhayaṃ, so the stemma (as a whole) seems split. [MD: done]
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.86}
+\begin{translation}[hp03_086]
+In men semen is dependent on the mind and life is dependent on semen, so semen and the mind should be carefully guarded.
+\end{translation}
+
+%\begin{sources}[hp03_086]
+%\begin{versinnote}
+%\end{versinnote}
+%\end{sources}
+
+\begin{testimonia}[hp03_086]
+\emph{Haṭharatnāvalī} 2.98
+\begin{versinnote}
+\tl{cittāyattaṃ nṛṇāṃ śukraṃ śukrāyattaṃ ca jīvitam |\\+}
+\tl{tasmāc chukraṃ manaś caiva rakṣanīyaṃ prayatnataḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~74v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{cittāyattaṃ nṛṇāṃ śukraṃ śukrāyattaṃ ca jīvitam |\\+}
+\tl{tasmāc cittaṃ ca śukraṃ ca rakṣaṇīyaṃ prayatnataḥ ||\\!}
+\end{versinnote}
+
+%Vajroliyoga 4
+%\begin{versinnote}
+%\tl{cittāyattaṃ nṛṇāṃ śukraṃ śukrāyattaṃ tu jīvitam |\\+}
+%\tl{tasmāc chukraṃ manaś caiva rakṣaṇīyaṃ prayatnataḥ ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_086]
+%No known source.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.87}
+\begin{translation}[hp03_087]
+In this way a [the yogi] may also hold on to [both] the menses of a menstruating woman and his semen. By practising correctly he may draw up [both] through the urethra by the proper practice.
+\end{translation}
+
+%\begin{sources}[hp03_087]
+%\end{sources}
+
+\begin{testimonia}[hp03_087]
+\emph{Haṭhratnāvalī} 2.100cd
+\begin{versinnote}
+\tl{ṛtumatyā rajo 'py evaṃ rajo binduṃ ca rakṣayet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p.39
+\begin{versinnote}
+\tl{ṛtumatyā rajo 'py evaṃ bijaṃ binduṃ ca rakṣayet |\\+}
+\tl{meḍhreṇākarṣayed ūrdhvaṃ samyagabhyāsayogavit || 59 || \\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_087]
+%No known source. 
+We have understood the reading of \emph{ṛtumatyā} as qualifying \emph{striyāḥ} (i.e.,`a menstruating woman'). Alternatively, the term \emph{ṛtumati} could mean a post-pubescent woman.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.87*1}
+\begin{translation}[hp03_087_1]
+This yoga succeeds for those who have merit, are fortunate, abide in truth, and are without jealousy, not for those who are jealous.
+\end{translation}
+%greyscaled
+
+\begin{sources}[hp03_087_1]
+\emph{Dattātreyayogaśāstra} 176
+\begin{versinnote}
+\tl{ayaṃ yogaḥ puṇyavatāṃ dhanyānāṃ tattvaśalinām |\\+}
+\tl{nirmatsarāṇāṃ sidhyeta na tu mātsaryaśālinām ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_087_1]
+\emph{Haṭharatnāvalī} 2.110
+\begin{versinnote}
+\tl{ayaṃ yogaḥ puṇyavatāṃ dhanyānāṃ tattvaśālinām |\\+}
+\tl{nirmatsarāṇāṃ sidhyeta na tu matsaraśālinām ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} pp. 40-41
+\begin{versinnote}
+\tl{ayaṃ yāgaḥ puṇyavatāṃ dhīrāṇāṃ tattvadarśinām | \\+}
+\tl{nirmatsarāṇāṃ sidhyeta na tu mātsaryaśālinām || \\!}
+\end{versinnote}
+
+%Vajroliyoga 27
+%\begin{versinnote}
+%\tl{ayaṃ yogaḥ puṇyavatāṃ siddhe saṃsāriṇāṃ na hi |\\+}
+%\tl{amunāṃ siddhim āpnoti yogād yogaḥ pravartate ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_087_1]
+This verse is omitted in \alphaTwo\ and \alphaThree\ (and the folio on which it would be found is missing in \alphaOne).
+\end{philcomm}
+
+\begin{metre}[hp03_087_1]
+Anuṣṭubh (a: bha-vipulā; c: ma-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.88}
+\begin{translation}[hp03_088]
+\emph{Sahajolī} and \emph{amarolī} are varieties of \emph{vajrolī}. 
+\end{translation}
+
+\begin{sources}[hp03_088]
+Cf. \emph{Dattātreyayogaśāstra} 31cd
+\begin{versinnote}
+\tl{vajrolir amaroliś ca sahajolis tridhā matā |\\!}
+\end{versinnote}
+
+\emph{Śivasaṃhitā} 4.95ab
+\begin{versinnote}
+\tl{sahajolyamarolī ca vajrolyā bhedato bhavet |\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_088]
+\emph{Haṭharatnāvalī} 2.113cd
+\begin{versinnote}
+\tl{atha sahajoliḥ -\\+}
+\tl{sahajolī cāmarolī vajrolyā eva bhedataḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasamhitā} p.40
+\begin{versinnote}
+\tl{sahajoliś cāmarolir vajrolyā bheda eva te |\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_088]
+These two \emph{pāda}s introduce the practices of \emph{sahajolī} and \emph{amarolī}, which are described in the verses that follow it.
+The \textalpha\ and \textbeta\ groups omit the headings for \emph{sahajolī} and \emph{amarolī}.
+Since 3.88 introduces these practices, the headings are probably not original.
+
+\end{philcomm}
+
+\begin{metre}[hp03_088]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.89}
+\begin{translation}[hp03_089]
+After intercourse using \emph{vajrolī}, the woman and man should put ash made from burnt cow dung in water [and] smear their bodies [with it...]
+\end{translation}
+% greyscaled
+
+\begin{sources}[hp03_089]
+\emph{Dattātreyayogaśāstra} 182
+\begin{versinnote}
+\tl{tajjale bhasma saṃkṣipya dagdhagomayasaṃbhavam |\\+}
+\tl{vajrolīmaithunād ūrdhvaṃ strīpuṃsor aṅgalepanam ||\\+}
+\tl{\var{182a tajjale bhasma saṃkṣipya ] M2; tajjale bhasmasāt kṣipya M1, tajjale bhasma saddravyaṃ A}\\!}
+\end{versinnote}
+
+\end{sources}
+
+\begin{testimonia}[hp03_089]
+\emph{Haṭharatnāvalī} 2.114
+\begin{versinnote}
+\tl{jale subhasma nikṣipya dagdhagomayasaṃbhavam |\\+}
+\tl{vajrolīmaithunād ūrdhvaṃ strīpuṃsoś cāṅgalepanam ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasamhitā} p.40
+\begin{versinnote}
+\tl{jale subhasma nikṣipya dagdhagomayasambhavam ||\\+}
+\tl{vajrolī maithunād ūrdhvaṃ strīpuṃsoḥ svāṅgalepanam | \\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_089]
+Some manuscripts, including \alphaTwo\ and \alphaThree\ (missing in \alphaOne), omit 3.89ab.
+We have included it because in the \emph{Dattātreyayogaśāstra}, the source of this verse, 3.89ab specifies the substance mentioned in 3.89cd that the man and woman are supposed to rub into their bodies after sexual intercourse.
+
+In the \emph{Dattātreyayogaśāstra}’s teaching on \emph{sahajolī} (163 and 181–183) a rag is used to wipe up the residue of a mixture of semen and sweat that has been rubbed into the body, and then soaked in a paste of water and ash before being rubbed over the body.
+
+Although the plural in 3.89a is awkward, \emph{jaleṣu} was probably the result of Svātmārāma removing the pronoun from the compound \emph{tajjale} in the \emph{Dattātreyayogaśāstra}'s verse because it has no referent in the \emph{Haṭhapradīpikā}'s compilation.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.90}
+\begin{translation}[hp03_090]
+[...] while sitting at complete ease, having just finished intercourse. This is called \emph{sahajolī}. It is always to be trusted by yogis. 
+\end{translation}
+
+\begin{sources}[hp03_090]
+\emph{Dattātreyayogaśāstra} 183
+\begin{versinnote}
+\tl{āsīnayoḥ sukhenaiva muktavyāparayoḥ kṣaṇam |\\+}
+\tl{sahajolī ca saṃproktā śraddheyā yogibhiḥ sadā ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_090]
+\emph{Haṭharatnāvalī} 2.115
+\begin{versinnote}
+\tl{āsīnayoḥ sukhenaiva muktavyāpārayoḥ kṣaṇam |\\+}
+\tl{sahajolir iyaṃ proktā kartavyā yogibhiḥ sadā ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṃhitā} p. 40
+\begin{versinnote}
+\tl{āsīnayoḥ sukhenaiva muktavyāpārayoḥ kṣaṇāt ||\\+}
+\tl{sahajolir iyaṃ proktā śraddheyā yogibhiḥ sadā \\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_090]
+We have understood the \emph{repha} in \emph{sahajolīr iyam} as a hiatus bridge. Elsewhere the nominative of this name is found only as \emph{sahajolī} or \emph{sahajoliḥ}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.90*1}
+\begin{translation}[hp03_090_1]
+This auspicious yoga bestows liberation even when pleasure has been enjoyed.
+\end{translation}
+%greyscaled
+
+%\begin{sources}[hp03_090_1]
+%\end{sources}
+
+\begin{testimonia}[hp03_090_1]
+\emph{Haṭhayogasaṃhitā} p. 40
+\begin{versinnote}
+\tl{ayaṃ śubhakaro yogī bhogayukto'pi muktidaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_090_1]
+This line is absent in \alphaTwo,\ \alphaThree\ and \gammaOne\ (missing in \alphaOne).
+It may have been adapted from \emph{Dattātreyayogaśāstra} 179cd (\emph{tasmād ayaṃ vakṣyamāṇo bhoge bhukte ’pi muktidaḥ}).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.91}
+\begin{translation}[hp03_091]
+Leaving out the first flow of urine because of its excessive heat and the last flow because it is worthless, the cool middle flow of urine is used by Kāpālikas of the Khaṇḍa school.
+\end{translation}
+
+\begin{sources}[hp03_091]
+\end{sources}
+
+\begin{testimonia}[hp03_091]
+\emph{Haṭharatnāvalī} 2.116
+\begin{versinnote}
+\tl{athāmarolī\\+}
+\tl{vihāya nityāṃ prathamāṃ ca dhārāṃ \\+}
+\tl{vihāya niḥsāratayāntyadhārām |\\+}
+\tl{niṣevyate śītalamadhyadhārāṃ \\+}
+\tl{kāpālikaiḥ khaṇḍamatair anarghyām ||\\+}
+\tl{\var{vihāya nityāṃ ] pittolbaṇatvāt \vl}\\+}
+\tl{\var{anarghyām ] anarghyā}\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 16.17
+\begin{versinnote}
+\tl{athāmarolī –\\+}
+\tl{pittolbaṇatvāt prathamāṃ ca dhārāṃ \\+}
+\tl{vihāya niḥsāratayāntyadhārām |\\+}
+\tl{niṣevyate śītamadhyadhārā \\+}
+\tl{kāpālikaiḥ khaṇḍamate 'marolī ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 41
+\begin{versinnote}
+\tl{pittolvaṇatvāt prathamāmbudhārāṃ \\+}
+\tl{niṣevyate śītalamadhyadhārā |\\+}
+\tl{vihāya niḥsāratayāntyadhārāṃ \\+}
+\tl{kāpālike khaṇḍamate 'marolī ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_091]
+We understand `Kāpālikas of the Khaṇḍa school' (\emph{kāpālikair khaṇḍamataiḥ}) to be referring to followers of the Khaṇḍakāpālika who is mentioned in the list of siddhas given at 1.5–9, \emph{pace} Marcinkowska-Rosół and Sellmer (2021: 105–108) who understand \emph{khaṇḍamataiḥ} to mean `whose doctrine is defective'.
+
+%Maria Marcinkowska-Rosół & Sven Sellmer. “Notes on some difficult passages of the Haṭhapradīpikā”. Zeitschrift der Deutschen Morgenländischen Gesellschaft 171 (2021), pp. 101–21
+\end{philcomm}
+
+\begin{metre}[hp03_091]
+Upajāti
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.92}
+\begin{translation}[hp03_092]
+[The yogi] who regularly imbibes urine, taking it by the nose every day, practises \emph{vajrolī} thus. This is called \emph{amarolī}.
+\end{translation}
+
+\begin{sources}[hp03_092]
+\emph{Dattātreyayogaśāstra} 180c–181b
+\begin{versinnote}
+\tl{amarīṃ yaḥ piben nityaṃ nasyaṃ kurvan dine dine ||\\+}
+\tl{vajrolīm abhyasec ceyam amarolīti kathyate |\\+}
+\tl{\var{181a abhyaset ceyam] \emph{em.}; abhyasec chrayam M1, abhyaset yeyam A, abhyasec caivam M2}}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_092]
+\emph{Haṭharatnāvalī} 2.117
+\begin{versinnote}
+\tl{amarīṃ yaḥ piben nityaṃ nasyaṃ kuryād dine dine  |\\+}
+\tl{vajrolīm abhyasen nityam amarolīti kathyate ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} 65 (p.41)
+\begin{versinnote}
+\tl{amarīṃ yaḥ piben nityaṃ nasyaṃ kurvan dine dine |\\+}
+\tl{vajrolīm abhyaset samyag amarolīti kathyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_092]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.92*1}
+\begin{translation}[hp03_092_1]
+If a woman draws up the semen of a man through skillfulness in the correct practice and retains her menses by means of \emph{vajrolī}, it is she who is a [true] yoginī.
+\end{translation}
+%greyscaled
+
+\begin{sources}[hp03_092_1]
+\emph{Dattātreyayogaśāstra} 169cd
+\begin{versinnote}
+\tl{yadi nārī rajo rakṣed vajrolyā sā hi yoginī ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_092_1]
+\emph{Haṭhayogasaṃhitā} p. 41
+\begin{versinnote}
+\tl{puṃso binduṃ samākuñcya samyagabhyāsapāṭavāt |\\+}
+\tl{yadi nārī rajo rakṣed vajrolyā sā'pi yoginī ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_092_1]
+In the first verse quarter, the gerund \emph{samākṛṣya} (\textgamma) has been adopted, instead of \emph{samākuñcya}, as it yields a better meaning and is used similarly to \emph{ākṛṣya} in 3.83.
+
+Verses 3.92*1–3 have been greyscaled because they are absent in \alphaThree \ (and missing in  \alphaOne), and appear to have been added to the text from the \emph{Dattātreyayogaśāstra}'s section on \emph{vajrolī} to provide further details of how a woman practises \emph{vajrolī} to those found in 3.93–94.
+The verses are present in \alphaTwo\ after verse 3.87ab where the verse quarter \emph{vajrolyā saha yoginī} occurs twice (also at 3.93b), which suggests that the version of \emph{vajrolī} in \alphaTwo\ has been subject to further revision.
+The fact that 3.92*1–3 are in groups \textbeta, \textgamma\ and \texteta\ indicates that they were added early in the transmission of the \emph{Haṭhapradīpikā}.
+% J5
+% ṛtumatyā rajo 'py evaṃ vīryaṃ binduṃ ca rakṣayet |
+% yadi nārī rajo rakṣet vajrolyā saha yoginī || (note the difference with our current 3.90c)
+% tasyāḥ kiñ cid rajo nāśaṃ na gacchati na saṃśayaḥ |
+% tasyāḥ śarīre nādas tu bindu[t]ām eva gacchati |
+% sa bindus tad rajaś caiva ekībhūtaṃ svadehagau | 
+% vajrolyābhyāsayogena sarvasiddhiḥ prakurvati [prajāyate] ||
+% sahajolī cāmarolī [ca?] vajrolyā eva bhedataḥ
+% 3.93
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.92*2}
+\begin{translation}[hp03_092_2]
+Assuredly none of her menses is lost. The \emph{nāda} in her body turns into \emph{bindu}.
+\end{translation}
+
+\begin{sources}[hp03_092_2]
+\emph{Dattātreyayogaśāstra} 174
+\begin{versinnote}
+\tl{tasyās tadā rajo nāśaṃ na gacchati na saṃśayaḥ | \\+}
+\tl{tasyāḥ śarīre nādas tu bindutām eva gacchati || 174 ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_092_2]
+\emph{Haṭharatnāvalī} 2.108ab
+\begin{versinnote}
+\tl{tasyāḥ śarīre nādas tu bindutām eva gacchati |\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} pp. 41--42
+\begin{versinnote}
+\tl{tasyāḥ kiñ cid rajo nāśaṃ na gacchati na saṃśayaḥ |\\+}
+\tl{tasyāḥ śarīre nādaś ca bindutām eva gacchati ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_092_2]
+On why this verse is in greyscale, see the note to 3.92*1.
+
+On \emph{nāda} and \emph{bindu} see the note to 3.53. 
+%Brahmānanda identifies nāda with rajas. LO: bindu is source of everything, nāda is going back to its source.
+%
+\end{philcomm}
+
+\begin{metre}[hp03_092_2]
+Anuṣṭubh (c: ma-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.92*3}
+\begin{translation}[hp03_092_3]
+The \emph{bindu} and \emph{rajas}, which are produced in her own body, become one through \emph{vajrolī} and bring about complete perfection by means of practice.%
+\end{translation}
+
+\begin{sources}[hp03_092_3]
+\emph{Dattātreyayogaśāstra} 175
+\begin{versinnote}
+\tl{sa bindus tad rajaś caiva ekībhūya svadehagau | \\+}
+\tl{vajrolyābhyāsayogena sarvasiddhiḥ prajāyate || 175 ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_092_3]
+\emph{Haṭharatnāvalī} 2.108cd--109ab
+\begin{versinnote}
+\tl{sa bindus tad rajaś caiva ekīkṛtya svadehajau ||\\+}
+\tl{vajrolyabhyāsayogena yogasiddhiḥ kare sthitā |\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 42
+\begin{versinnote}
+\tl{sa bindus tad rajaś caiva ekībhūya svadehagau |\\+}
+\tl{vajrolyabhyāsayogena sarvasiddhiṃ prayacchataḥ || \\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_092_3]
+On why this verse is in greyscale, see the note to 3.92*1.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.93}
+\begin{translation}[hp03_093]
+It is she who preserves her menses by means of the upward contraction who is the [true] yoginī.
+She knows the past and the future, and is sure to become a sky-rover (\emph{khecarī}).
+\end{translation}
+
+\begin{sources}[hp03_093]
+\emph{Dattātreyayogaśāstra} 170ab
+\begin{versinnote}
+\tl{atītānāgataṃ vetti khecarī vā bhaved dhruvam |\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_093]
+\emph{Haṭhayogasaṃhitā} p. 42
+\begin{versinnote}
+\tl{rakṣed ākuñcanād ūrdhvaṃ yā rajaḥ sā hi yoginī | \\+}
+\tl{atītānāgatajñānaṃ khecarī ca bhaved dhruvam ||\\!}
+\end{versinnote}
+
+%Haṭhapradīpikā 5.140--141ab
+%\begin{versinnote}
+%\tl{mehanākuñcanād ūrdhvaṃ rajasāpi ca yoginī |\\+}
+%\tl{atītānāgataṃ vetti khecarī ca bhaved dhruvam ||\\+}
+%\tl{dehasiddhiṃ ca labhate vajrolyabhyāsayogataḥ |\\!}
+%\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_093]
+%\end{philcomm}
+
+\begin{metre}[hp03_093]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.94}
+\begin{translation}[hp03_094]
+And she attains perfection of the body as a result of the practice of \emph{vajrolī}. This auspicious yoga bestows liberation even when pleasure has been enjoyed.
+\end{translation}
+%Therefore this yoga succeeds only for those who have merit.
+
+\begin{sources}[hp03_094]
+\emph{Dattātreyayogaśāstra} 179
+\begin{versinnote}
+\tl{dehasiddhiṃ ca labhate vajrolyabhyāsayogataḥ |\\+}
+\tl{tasmād ayaṃ vakṣyamāṇo bhoge bhukte ’pi muktidaḥ \\+}
+\tl{\var{179d bhoge bhukte ’pi muktidaḥ ] conj.; bhoge bhukte tv abhuktidaḥ M1, bhogo yogaś ca muktidaḥ AM2}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_094]
+\emph{Haṭharatnāvalī} 2.111
+\begin{versinnote}
+\tl{sarveṣām eva yogānām ayaṃ yogaḥ śubhaṅkaraḥ |\\+}
+\tl{tasmād ayaṃ variṣṭho 'sau bhuktimuktiphalapradaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 42
+\begin{versinnote}
+\tl{dehasiddhiṃ ca labhate vajrolyabhyāsayogataḥ |\\+}
+\tl{ayaṃ puṇyakaro yogo bhoge bhukte 'pi muktidaḥ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_094]
+%tasmād ayaṃ sādhakāya hemistich is not in Gr2 and seems like an unnecessary repetition (that may have occurred to change a 6 pāda verse into two verses).
+%?? JB we havent translated 3.101*1. Are we going to keep it in the edition or put it in a footnote in the apparatus of the edition (it seems quite repetitive as 3.101*1b= 3.101d)
+% MD: 3.101*1ab and 3.101cd by the old numbering are now merged as 3.94*cd.
+%\end{philcomm}
+
+\begin{metre}[hp03_094]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.94*1}
+\begin{philcomm}[hp03_094_1]
+This verse is not in \textalpha\ and \textgamma, and seems like an unnecessary repetition of 3.101.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.94*2}
+\begin{translation}[hp03_094_2]
+Now the Stimulation of the Goddess (\emph{śakticālanam}):\\ 
+She whose body is bent (\emph{kuṭilāṅgī}), she who is coiled (\emph{kuṇḍalinī}), the female snake (\emph{bhujaṅgī}), the power (\emph{śakti}), the goddess (\emph{īśvarī}), she who is coiled (\emph{kuṇḍalī}) and Arundhatī: these words are synonyms.
+\end{translation}
+% ?? JB. the heading is regular typeface and 3.094_2 is greyscale. So we may need to separate them in our translation to show that the heading is part of the original text.
+
+%\begin{sources}[hp03_094_2]
+%\end{sources}
+
+\begin{testimonia}[hp03_094_2]
+
+\emph{Haṭharatnāvalī} 2.125–127
+\begin{versinnote}
+\tl{phaṇī kuṇḍalinī nāgī cakrī vakrī sarasvatī |\\+}
+\tl{lalanā rasanā kṣatrī lalāṭī śaktiḥ śaṃkhinī ||\\+}
+\tl{rajvī bhujaṅgī śeṣā ca kuṇḍalī sarpiṇī maṇiḥ |\\+}
+\tl{ādhāraśaktiḥ kuṭilā karālī prāṇavāhinī ||\\+}
+\tl{aṣṭavakrā ṣaḍādhārā vyāpinī kalanādharā ||\\+}
+\tl{kurīty evaṃ ca vikhyātāḥ śabdāḥ paryāyavācakāḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~78v (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{kuṇḍalāṅgī kuṇḍalinī bhujaṅgī śaktir īśvarī |\\+}
+\tl{kuṭilārundhatī devī śabdāḥ paryāyavācakāḥ ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.300 (attr.~to the \emph{Śivayoga})
+\begin{versinnote}
+\tl{kuṭilāṃgī kuṇḍalinī bhujaṅgī śaktir īśvarī |\\+}
+\tl{kuṇḍaly arundhatī devī śabdāḥ paryāyavācakāḥ ||\\!}
+\end{versinnote}
+
+%Haṃsavilāsa (p. 46)
+%\begin{versinnote}
+%\tl{āgame nāmāntarāṇi ca |\\+}
+%\tl{kuṭilāṅgī kuṇḍalinī bhujaṅgī śaktir īśvarī |\\+}
+%\tl{kuṇḍalyā ceti ruddhanti śabdāḥ paryāyavācakāḥ ||\\!}
+%\end{versinnote}
+
+%Prāṇatoṣiṇī Part 6 (possiby attr.~to the Dattātreyasaṃhitā) %line 3723
+%\begin{versinnote}
+%\tl{kuṇḍalāṅgī kuṇḍalinī bhujaṅgī śaktirīśvarī | \\+}
+%\tl{kuṇḍalyarundhatī devī śabdāḥ paryāyavācakāḥ |\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_094_2]
+%Authorial?
+%3.4 and 4.4 have cety ekavācakāḥ, but devī paryāyavācakāḥ is well attested among the HP manuscripts (Groups 2 and 3) and the testimonia.
+The \alphaThree\ manuscript has a significantly shorter and more coherent version of \emph{śakticālana}.
+It omits five introductory verses, of which three are from the \emph{Vivekamārtaṇḍa} or one of its longer recensions and three have no known source, including one that contains a list of synonyms for \emph{kuṇḍalinī}.
+This section is missing in \alphaOne\ (3.83–3.98) and \alphaTwo\ adds these verses (except 3.94*6) after 3.97, which suggests that they have been inserted from elsewhere.
+Generally speaking, it appears that some redactors have taken the section on \emph{śakticālana} in the \emph{Haṭhapradīpikā} as an opportunity to add material on \emph{kuṇḍalinī}, in particular her location, shape, and soteriological importance.
+
+\end{philcomm}
+
+\begin{metre}[hp03_094_2]
+Anuṣṭubh (a: bha-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.94*3}
+\begin{translation}[hp03_094_3]
+Just as one might use a key to force open a double door, so the yogi breaks open the door to liberation with \emph{kuṇḍalinī}.
+\end{translation}
+
+\begin{sources}[hp03_094_3]
+\emph{Vivekamārtaṇḍa} 35
+\begin{versinnote}
+\tl{udghāṭayet kapāṭaṃ tu yathā kuñcikayā haṭhāt |\\+}
+\tl{kuṇḍalinyā tathā yogī mokṣadvāraṃ vibhedayet ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_094_3]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{haṭhayoge\\+}
+\tl{udghāṭayet kapāṭaṃ tu yathā kuñcikayā haṭhāt |\\+}
+\tl{kuṇḍalinyā tathā yogī mokṣadvāraṃ vibhedayet ||\\!}
+\end{versinnote}
+
+Haṭhasaṅketacandrikā f.~110r
+\begin{versinnote}
+\tl{tathā coktaṃ haṭhapradīpikāyāṃ |\\+}
+\tl{udghāṭayet kapāṭaṃ tu yathā kuñcikayā haṭhāt |\\+}
+\tl{kuṇḍalinyā tathā yogī mokṣadvāraṃ vibhedayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_094_3]
+%SS: von Hinuber 1992 booklet has something relevant to this. Can provide.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.94*4}
+\begin{translation}[hp03_094_4]
+The supreme goddess sleeps with her mouth covering the opening of the pathway by which the wholesome place of Brahman is reached.
+\end{translation}
+%
+
+\begin{sources}[hp03_094_4]
+\emph{Vivekamārtaṇḍa} 33
+\begin{versinnote}
+\tl{yena mārgeṇa gantavyaṃ brahmasthānaṃ nirāmayam |\\+}
+\tl{mukhenācchādya taddvāraṃ prasuptā parameśvarī ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_094_4]
+\emph{Yogacintāmaṇi} f.~78v (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{yena dvāreṇa gantavyaṃ brahmasthānaṃ nirāmayam |\\+}
+\tl{mukhenācchādya taddvāraṃ prasuptā parameśvarī ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~110r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{yena mārgeṇa gaṃtavyaṃ brahmasthānaṃ nirāmayaṃ |\\+}
+\tl{mukhenācchādya taddvāraṃ prasuptā parameśvarī ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_094_4]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.94*5}
+\begin{translation}[hp03_094_5]
+The coiled goddess, who sleeps above the bulb [in the abdomen], leads to liberation for yogis and bondage for the deluded. He who knows her knows yoga.
+\end{translation}
+
+\begin{sources}[hp03_094_5]
+\emph{Vivekamārtaṇḍa} 39
+\begin{versinnote}
+\tl{kandordhvaṃ kuṇḍalī śaktir suptā mokṣāya yoginām |\\+}
+\tl{bandhanāya ca mūḍhānāṃ yas tāṃ vetti sa yogavit ||\\+}
+\tl{\var{suptā mokṣāya yoginām ] VTvlH; aṣṭadhā kuṃḍalīkṛtā A, aṣṭadhā kuṇḍalākṛtiḥ GLGPk, śubhamokṣāpradāyinī GB, śubhā mokṣapradāyinī GP, aṣṭadhā kuṭilīkṛtā T}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_094_5]
+\emph{Yogacintāmaṇi} f.~78v (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{kandordhve kuṇḍalī śaktir buddhā mokṣāya yoginām |\\+}
+\tl{bandhanāya ca mūḍhānāṃ yas tām vetti sa yogavit ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~110r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{kandordhvaṃ kuḍalī śakti suptā mokṣāya yogināṃ |\\+}
+\tl{bandhanāya ca mūḍhānāṃ yas taṃ vetti sa yogavit ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_094_5]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.94*6}
+\begin{translation}[hp03_094_6]
+[Just as] the coiled serpent Ananta (\emph{śeṣa\-kuṇḍalī}) is the foundation of the oceans, mountains and islands, so kuṇḍalinī is the foundation of all systems of yoga.
+\end{translation}
+
+%\begin{sources}[hp03_094_6]
+%\end{sources}
+
+\begin{testimonia}[hp03_094_6]
+%\emph{Haṭharatnāvalī} 2.124
+%\begin{versinnote}
+%\tl{saśailavanadhātryās tu yathādhāro 'hināyakaḥ |\\+}
+%\tl{aśeṣayogatantrāṇāṃ tathādhāro hi kuṇḍalī ||\\!}
+%\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~78v (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{ambhodhiśailadvīpānām ādhāraḥ śeṣakuṇḍalī |\\+}
+\tl{aśeṣayogatantrāṇām ādhāraḥ kuṇḍalī tathā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_094_6]
+%No known source. Authorial? 
+This verse is similar to 3.1.
+\end{philcomm}
+
+\begin{metre}[hp03_094_6]
+Anuṣṭubh (a: ma-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.94*7}
+\begin{translation}[hp03_094_7]
+\emph{Kuṇḍalinī} is said to have a curved shape like a snake. The person who makes that goddess move is sure to be liberated.
+\end{translation}
+
+%\begin{sources}[hp03_094_7]
+%\end{sources}
+
+\begin{testimonia}[hp03_094_7]
+\emph{Yogacintāmaṇi} f.~78v–79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{kuṇḍalī kuṭilākārā sarpavat parikīrtitā |\\+}
+\tl{sā śaktiś cālitā yena sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_094_7]
+%No known source.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.95}
+\begin{translation}[hp03_095]
+Between the Gaṅgā and Yamuna is the wretched young widow. [The yogi] should forcefully take [her]. That is the supreme state of Viṣṇu.
+\end{translation}
+
+\begin{sources}[hp03_095]
+Cf.~\emph{Śivasaṃhitā} 5.169
+\begin{versinnote}
+\tl{gaṅgāyamunayor madhye vahaty eṣā sarasvatī |\\+}
+\tl{tāsāṃ tu saṃgame snātvā dhanyo yāti parāṃ gatim\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_095]
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{gaṅgāyamunayor madhye bālaraṇḍā tapasvinī |\\+}
+\tl{balātkāreṇa gṛhṇīyāt tad viṣṇoḥ paramaṃ padam ||\\!}
+\end{versinnote}
+
+%Haṃsavilāsa (p. 46)
+%\begin{versinnote}
+%\tl{gaṅgāyamunayor madhye bālaraṇḍāṃ tapasvinīm |\\+}
+%\tl{balātkāreṇa gṛhṇīyāt tad viṣṇoḥ paramaṃ padam ||\\!}
+%\end{versinnote}
+
+Prāṇatoṣiṇī Part 6 (attr.~to the \emph{Dattātreyasaṃhitā})
+\begin{versinnote}
+\tl{gaṅgāyamunayor madhye bālāraṇḍā tapasvinī | \\+}
+\tl{balād ākṛṣya gṛhnīyāt tad viṣṇoḥ paramaṃ padam ||\\!}
+\end{versinnote}
+
+%Bodhasāra 39.34, 12.3 (check, has commentary too):
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_095]
+%No known source. 
+The referent of \emph{bālaraṇḍā tapasvinī}, here is unclear.
+In some manuscripts this verse is followed by one in which \emph{bālaraṇḍā} is identified as \emph{sarasvatī}, which in the context of \emph{śakticālana} could refer to the tongue.
+She could also be\emph{kuṇḍalinī}, who in 3.94*5 is located at the navel, which is said to be the location of Viṣṇu (e.g.~\emph{Dhyāna\-bindū\-paniṣat} 28-30).
+In his commentary on this verse in the \emph{Bodhasāra} (1906: 137), Divākara says that the seizing of \emph{kuṇḍalinī} itself is the highest state of Viṣṇu (...
+\emph{bālaraṇḍāṃ ...
+gṛhṇīyād vaśīkuryāt tat tasyā vaśīkaraṇam eva viṣṇor vyāpana\-lakṣa\-nasya paramā\-tmanaḥ paramaṃ kevalaṃ ...
+padaṃ svarūpaṃ jñeyam}).
+% Narahari N Divākara Divākara Dayānanda Sarasvatī. Bodhasāraḥ = Bodhsār : A Treatise on Vedanta. Arthadīptyā Sahitaḥ. Benares: Vidya Vilas Pr; 1906.
+
+% comment on 3.95*1, which is similar to 3.95?
+Verse 3.95*1, which has no known source, simply identifies the technical terms in 3.95, namely, \emph{gaṅgā}, \emph{yamunā} and \emph{bālaraṇḍā} as \emph{iḍā, piṅgalā} and \emph{sarasvatī}.
+It is absent in the \textalpha\ manuscripts and probably crept into the text as a marginal note early in the transmission.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.95*1}
+% iḍā bhagavatī gaṅgā
+
+%%%%%%%%%%
+\subsection*{3.96}
+\begin{translation}[hp03_096]
+Seizing her tail, the fearless [yogi] wakes the sleeping serpent.  She shakes off sleep and is forced to stand up straight.
+\end{translation}
+
+%\begin{sources}[hp03_096]
+%\end{sources}
+
+\begin{testimonia}[hp03_096]
+\emph{Haṭharatnāvalī} 2.118
+\begin{versinnote}
+\tl{pucche pragṛhya bhujagīṃ suptām udbodhayed abhīḥ  |\\+}
+\tl{nidrāṃ vihāya sā ṛjvī ūrdhvam uttiṣṭhate haṭhāt ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{pucchaṃ pragṛhya bhujagīṃ suptām udbodhayed abhi |\\+}
+\tl{nidrāṃ vihāya sā ṛjvī ūrdhvam uttiṣṭhate haṭhāt ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp03_096]
+%No known source.
+%\end{philcomm}
+
+\begin{metre}[hp03_096]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.96*1}
+\begin{translation}[hp03_096_1]
+The yogi should regularly move the coiled (\emph{paristhitā}), hooded [serpent] by breathing in through the sun channel and holding her using a cloth for an hour and a half in the morning and evening.
+\end{translation}
+%greyscaled
+
+%\begin{sources}[hp03_096_1]
+%\end{sources}
+
+\begin{testimonia}[hp03_096_1]
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{paristhitā caiva phaṇāvatī sā\\+}
+\tl{prātaś ca sāyaṃ praharārdhamātram |\\+}
+\tl{prapūrya sauryā paridhānamuktā\\+}
+\tl{pragṛhya niryāti vicālitā sā ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~110v-111r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tadvidhim āha |\\+}
+\tl{paristhitā caiva phaṇāvatī sā\\+}
+\tl{prātaś ca sāyaṃ praharārdhamātraṃ |\\+}
+\tl{prapūrya sūryāt paridhānayuktā\\+}
+\tl{pragṛhya tīrthāt paricālanīyā ||\\+}
+\tl{paridhān[a]yukteti dvādaśāṅgulapramitasitasūkṣmacaturaṅgulavisṛtaśuddhavastrakhaṇḍena dṛḍhaṃ veṣṭatā sā prasiddhā [ph]aṇāvatī suṣumṇātmakā arundhatī jihvaiva kuṇḍalinī || uktaṃ ca ||\\+}
+\tl{arundhatī bhavej jihvā dhruvo nāsāgramaṇḍalam iti || \\+}
+\tl{tāṃ jihvāṃ laṃbikāyogenordhvaṃ tālvantarbhrūmadhyadeśe vihitāṃ tatas tīrthād bhrūmadhyāt pragṛhya adhaḥ kṛtvā tasyā gurūpadiṣṭavartmanā cālanaṃ vidheyam iti saṃketaḥ [|] cālanaṃ tu khecarī mudrā sādhanavad vidheyaṃ [|] tīrthaṃ bhrūmadhyaḥ [|] \\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 44.5
+\begin{versinnote}
+\tl{paristhitasyeha phaṇāvatī sā\\+}
+\tl{prātastu sāyaṃ praharārdhamātram |\\+}
+\tl{prapūrya sūryāt paridhānayuktā\\+} 
+\tl{pragṛhya niryāt paricālanīyā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_096_1]
+Verses 3.96*1–2 are absent in \alphaOne\ and \alphaTwo.
+They introduce the idea of awakening \emph{kụṇḍalinī} by moving the tongue with a cloth, which is a practice called \emph{sarasvatīcālana} in the \emph{Gorakṣaśataka} (16–25).
+These verses do not have a known source and are somewhat obscure unless one is one aware of the more coherent explanation of this practice in the \emph{Gorakṣaśataka}.
+In his \emph{Haṭhasaṅketacandrikā} (see testimonia), Sundaradeva makes sense of this verse by equating the tongue with \emph{kuṇḍalinī}.
+This enables him to understand the reference to the cloth (\emph{paridhāna}) as the technique of wrapping the tongue in a cloth and milking it, which is a practice called \emph{sarasvatīcālana} in the \emph{Gorakṣaśataka} (16–25).
+This interpretation also makes sense of the next verse in the \emph{Haṭhapradīpikā} (3.11), which describes the cloth.
+
+%%?? JB: We have included paristhitā in our translation (next to `coiled") and it is an unusual meaning of paristhitā. Did anyone take down a reference for understanding it as `coiled'?
+
+%C2 - nityaṃ
+%adopt yuktyā, nityam paricālanīyā (C6 - 4b group) [MD: done]
+%GŚ 22 says to move the tongue for two muhūrtas
+%adopt?:
+%paristhitā caiva phaṇāvatī sā
+%prātaś ca sāyaṃ praharārdhamātram
+%prapūrya sūryāt paridhānayuktyā
+%pragṛhya nityaṃ paricālanīyā
+\end{philcomm}
+
+\begin{metre}[hp03_096_1]
+Upajāti
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.96*2}
+\begin{translation}[hp03_096_2]
+It is said that the characteristics of the cloth for wrapping around [the tongue] are  that it measures a handspan in length and four fingerbreadths in width, and is soft and white.
+\end{translation}
+%greyscaled
+
+\begin{sources}[hp03_096_2]
+Cf.~\emph{Gorakṣaśataka} 20cd
+\begin{versinnote}
+\tl{dvādaśāṅguladairghyaṃ cāmbaraṃ caturaṅgulam\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_096_2]
+\emph{Yogabīja} 81 (South Indian recension)
+\begin{versinnote}
+\tl{vitastipramitaṃ dairghyaṃ vistāre caturaṅgulam |\\+}
+\tl{mṛdulaṃ dhavalaṃ proktaṃ veṣṭanāmbaralakṣaṇam  ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~74r (attr.~to the \emph{Yogabīja} in the context of \emph{khecarīmudrā})
+\begin{versinnote}
+\tl{yogabīje—\\+}
+\tl{vitastipramitaṃ dīrgha[ṃ] vistāraṃ caturaṅgulam |\\+}
+\tl{mṛdulaṃ dhavalaṃ proktaṃ veṣṭanādhāralakṣaṇam ||\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 44
+\begin{versinnote}
+\tl{vitastipramitaṃ dīrghaṃ vistāre caturaṅgulam |\\+}
+\tl{mṛdulaṃ dhavalaṃ sūkṣmaṃ veṣṭanāmbaralakṣaṇam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_096_2]
+%adopt vistāraṃ ?? This was written here (maybe at the time of the workshop?) JB we dont have vistāraṃ as the lemma in the edition?
+This verse was likely added to explain `by the method of the cloth' (\emph{paridhānayuktyā}) in the previous verse.
+One would expect to read \emph{caturaṅgulavistāram} in the second verse quarter, and the current reading is probably a result of the metre.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.97}
+\begin{translation}[hp03_097]
+Sitting in \emph{vajrāsana}, the yogī should make \emph{kuṇḍalinī} move and immediately afterwards perform \emph{bhastrī}. He quickly awakens \emph{kuṇḍalinī}.
+\end{translation}
+
+%\begin{sources}[hp03_097]
+%\end{sources}
+
+\begin{testimonia}[hp03_097]
+\emph{Yogabīja} 111 (South Indian recension)
+\begin{versinnote}
+\tl{vajrāsanasthito yogī cālayitvā tu kuṇḍalīm |\\+}
+\tl{kuryād anantaraṃ bhastrāṃ kuṇḍalīṃ āśu bodhayet ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{vajrāsanasthito yogī cālayitvā tu kuṇḍalīm |\\+}
+\tl{sūryād anantaraṃ bhastrā kuṇḍalīm āśu bodhayet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~111r
+\begin{versinnote}
+\tl{vajrāsanasthito yogī cālayitvā tu kuṃḍalīṃ |\\+}
+\tl{sūryād anantaraṃ bhastrī kuṃḍalīm āśu bodhayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_097]
+\emph{Bhastrī} or \emph{bhastrikā kumbhaka} is taught at 2.60–68.The reference to \emph{vajrāsana} may be pointing to the practice of \emph{uḍḍiyāna}, which was described earlier in the chapter and is supposed to awaken \emph{kuṇḍalinī}.
+The contraction of the sun mentioned in the next verse supports this.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.98}
+\begin{translation}[hp03_098]
+[The yogi] should contract the sun and then make kuṇḍalinī move. Even if he is in the jaws of death, why would he fear death?%??JM: ...there is no danger of death.
+\end{translation}
+
+%\begin{sources}[hp03_098]
+%\end{sources}
+
+\begin{testimonia}[hp03_098]
+\emph{Yogabīja} 83 (South Indian recension)
+\begin{versinnote}
+\tl{bhānor ākuñcanaṃ kuryāt kuṇḍalīṃ cālayet tataḥ | \\+}
+\tl{mṛtyuvartmagatasyāpi tasya mṛtyubhayaṃ kutaḥ ||\\+}
+\tl{\var{mṛtyuvartma° ] mṛtyuvaktra°,  mṛtyuvajra°, mṛtyupadma° \vl}\\!}
+% mṛtyuvaktra° is supported by the North Indian recension and printed editions
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{bhānor ākuñcanaṃ kuryāt kuṇḍalīṃ cālayet tataḥ |\\+}
+\tl{mṛtyuvaktragatasyāpi tasya mṛtyubhayaṃ kutaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (of Sundaradeva) f.~111r
+\begin{versinnote}
+\tl{bhānor ākuñcanaṃ kuryāt kuṃḍalīṃ cālayet svataḥ |\\+}
+\tl{mṛtyuvakragatasyāpi tasya mṛtyu[bha]yaṃ kutaḥ ||\\+}
+\tl{asyārthaḥ sūryanāḍyākarṣaṇena vahneḥ prācuryaṃ tasmāj jvalanatejasā apānavāyvākarṣaṇena vā kuṇḍalyābodho bhavati [|] tasya śakticālanakṛtābhyāsasya mṛtyubhayaṃ kuta iti [||]\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_098]
+In \emph{Jyotsnā} 3.116, Brahmānanda understands \emph{bhānor ākuñcanaṃ} as a contraction of the navel, whereas Sundaradeva in his \emph{Haṭhasaṃketacandrikā} (see testimonia) interprets it as drawing \emph{prāṇa} through the sun channel (\emph{sūryanāḍyākarṣaṇa}), thereby intensifiying the bodily fire.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.98*1}
+\begin{translation}[hp03_098_1]
+When \emph{prāṇa} has been greatly extended, [the yogi's] breath flows through the path of the right nostril and [his] body is immortal, filled with the nectar of the moon from the uvula for the first time.
+Sprinkling [with nectar] the network of channels at the aperture at [the centre of the] brow that is subjugated by the mighty fire of time, [the yogi] then makes his body completely new again like [the regeneration of] the trunk of an ancient tree.
+\end{translation}
+%greyscaled
+
+\begin{sources}[hp03_098_1]
+\end{sources}%?? Aś verse in sources?
+
+\begin{testimonia}[hp03_098_1]
+\emph{Amaraughaśāsana} 6.1-2 (sic; a single śārdūlavikrīḍita verse is numbered thus)
+\begin{versinnote}
+\tl{nāsāpaścimamārgavāhapavanāt prāṇe 'tidīrghīkṛte\\+}
+\tl{candrāmbupratisāraṇāṃ sukṛtinaḥ prāg ghaṇṭikāyāḥ pathaḥ |\\+}
+\tl{siñcan kālaviśālavahnivaśagaṃ bhūtvā sa nāḍīśataṃ \\+}
+\tl{tat kāryaṃ kurute punar navatanuṃ jīrṇadrumaskandhavat ||\\+}
+\tl{pratisāraṇānantaraṃ śaṅkhasāraṇā kathyate\\+}
+\tl{\var{nāḍīśataṃ ] nāḍīgataṃ, nāḍīgaṇaṃ \vl}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{nāsādakṣiṇamārgavāhipavano ghrāṇe 'tidīrghīkṛtaḥ\\+}
+\tl{candrābhaḥparipūritāmṛtatanuḥ prāg ghaṇṭikāyās tataḥ |\\+}
+\tl{bhindan kālaviśālavahnivaśagān bhrūrandhranāḍīgaṇān \\+}
+\tl{taṃ kāyaṃ kurute punar navataraṃ jīrṇadrumaskandhavat ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (f. 111r–111v)
+\begin{versinnote}
+\tl{nāsādakṣiṇamārgavāhipavano ghrāṇe tidī[r]ghīkṛtaś\\+}
+\tl{caṃdrāṃ'bhaḥparipūritā'mṛtatanuḥ prāgh ghaṃṭikāyās tataḥ [|]\\+}
+\tl{bhindan kālaviśālavahnivaśagān bh[r]ū[ran]dhranāḍīgaṇāṃs \\+}
+\tl{taṃ kāyaṃ kurute punar navataraṃ jīrṇadrumaskandhavat [||]\\+}
+\tl{dakṣiṇe ghrāṇe nāsikāyām atidī[r]ghīkṛtaś ciraṃ kuṃbhakīkṛtaḥ prāk prathamaṃ caṃdrāṃ'bhaḥparipūritā'mṛtatanuḥ sādhakaḥ kartā pūrvam iḍayā dhṛtakuṃbhakenātisukhakarasudhopamena saṃtṛptiṃ samupagataḥ satatas tadanantaraṃ sūryanāḍyā cirāyā kalitaṃ kuṃbhauṣṇyātīkṣṇatarakuṃ tenaivauṣṇasvabhāvād ghaṇṭikāyāṃ jatruṇaḥ pañcamy arthaḥ tatsaṃbadhikān ity arthaḥ [|] evaṃvidhān kālabījān bh[r]ūrandhragataśirāpuñjān bhindan saṃchedayan svaṃ nijaśarīraṃ punar navataraṃ kuruta iti asyārthaḥ {|}\\!}
+\end{versinnote}
+
+%Yogasārasaṃgraha has this too, check
+\end{testimonia}
+
+\begin{philcomm}[hp03_098_1]
+Verse 3.98*1 is absent in the \textalpha\ group.
+It is very close to a verse in the \textit{Amaraughaśāsana}, which is likely to be its source.
+Its import of rejuvenating the body by flooding it with nectar is not directly connected with the verses proceeding it.
+
+%Adopt °pavano prāṇe 'tidīrghīkṛte
+%Adopt siñcan from AŚ, or (JT) understand bhindan to mean cutting off channels that are under the influence of time
+%Adopt tat pāda 4 and understand as tasmāt
+%If yogi is subject taṃ doesn't work, adopt tat
+%Simile of tree coming back to life after being burned or withering, through nectar passing through the channels
+\end{philcomm}
+
+\begin{metre}[hp03_098_1]
+Śārdūlavikrīḍita
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.98*2}
+\begin{translation}[hp03_098_2]
+After making kuṇḍalinị move, the yogi should perform \emph{bhastrī} in particular. The god of death is afraid of the ascetic who regularly practises in this way.
+\end{translation}
+%greyscaled
+
+%\begin{sources}[hp03_098_2]
+%\end{sources}
+
+\begin{testimonia}[hp03_098_2]
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{kuṇḍalīṃ cālayitvā tu kuryād bhastrāṃ viśeṣataḥ |\\+}
+\tl{evamabhyāsato nityaṃ yaminaḥ śaṅkate yamaḥ ||\\!}
+\end{versinnote}
+
+Haṭhasaṅketacandrikā (f. 111v–112r)
+\begin{versinnote}
+\tl{kuṃḍalīṃ cālayitvā tu kuryād bhastrīṃ viśeṣataḥ |\\+}
+\tl{evamabhyāsato nityaṃ yaminaḥ śaṃphate manaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_098_2]
+Verses 3.98*2–3 are absent in the \textalpha\ group and have no known source.
+They elaborate further on the instruction to practise \emph{bhastrī kumbhaka} in 3.97.
+The practice of other \emph{kumbhaka}s for moving \emph{kuṇḍalinī} is mentioned in 3.98*3.
+
+The reading \emph{abhyāsato} in 3.98*2c, which is attested by manuscripts of the \textgamma\ group and the testimonia, is possible but seems to be a later corruption of the participle, \emph{abhyasyato}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.98*3}
+\begin{translation}[hp03_098_3]
+Then [the yogi] should practise \emph{sūryabheda}, \emph{ujjāyī} and also \emph{śītalī}. Where is the god of death for the ascetic engaged in the practice in this way?
+\end{translation}
+%greyscaled
+
+%\begin{sources}[hp03_098_3]
+%\end{sources}
+
+\begin{testimonia}[hp03_098_3]
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{tadābhyaset sūryabhedam ujjāyīṃ cāpi śītalīm |\\+}
+\tl{evam abhyāsayuktasya yamas tu yaminaḥ kutaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~112r
+\begin{versinnote}
+\tl{tadābhyaset sūryabhedam ujjāyīṃ vāpi śītalīm |\\+}
+\tl{evamabhyāsayuktasya yamas tu yaminaḥ kutaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_098_3]
+On why this verse is in greyscale, see the note to 3.98*2.
+
+%No known source. the readings \emph{cāpi} and \emph{vāpi} are both possible, but the former is better attested.
+\end{philcomm}
+
+\begin{metre}[hp03_098_3]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.99}
+\begin{translation}[hp03_099]
+As a result of fearlessly moving [\emph{kuṇḍalinī}] for a total of two \emph{muhūrta}s (one hour 46 minutes), \emph{kuṇḍalinī}, which is in \emph{suṣumṇā}, is drawn up slightly.
+\end{translation}
+% ?? JB, since there's a lot of Sanskrit in this translation and non-specialists are unlikely to know what a muhūrta is, I think we should translate it as 'for up to an hour and a half.'
+
+\begin{sources}[hp03_099]
+\emph{Gorakṣaśataka} 22c–23b
+\begin{versinnote}
+\tl{muhūrtadvayaparyantaṃ nirbhayaś cālayed imām |\\+}
+\tl{ūrdhvam ākṛṣyate kiṃcit suṣumnā kuṇḍalīgatā ||\\+}
+\tl{\var{22c nirbhayaś ] YL ; nirbharaś T, nirbhayāc GU 23a ākṛṣyate YL ; ākarṣayet TGU 23b suṣumnā kuṇḍalīgatā ] TG; suṣumnāṃ kuṇḍalīgatāṃ U,  suṣumnā kuṇḍalīyutā YL}\\!}
+\end{versinnote}
+
+Cf.~\emph{Śivasamhitā} 4.109
+\begin{versinnote}
+\tl{gurūpadeśavidhinā tasya mṛtyubhayaṃ kutaḥ |\\+}
+\tl{muhūrtadvayaparyantaṃ vidhinā śakticālanam ||\\!}
+\end{versinnote}
+
+\end{sources}
+
+\begin{testimonia}[hp03_099]
+\emph{Haṭharatnāvalī} 2.121
+\begin{versinnote}
+\tl{muhūrtadvayaparyantaṃ nirbhītaś cālayed asau |\\+}
+\tl{ūrdhvam ākṛṣyate kiṃ cit suṣumṇāṃ kuṇḍalīgatām |\\+}
+\tl{ṣaṇmāsāc cālanenaiva śaktis tasyordhvagā bhavet ||\\+}
+\tl{\var{kuṇḍalīgatām ] kuṇḍalī gatā \vl}\\!}
+\end{versinnote}
+
+\emph{Yogalakṣaṇāvalī} (f.~31r) (attr.~to the \emph{Gorakṣaśata})
+\begin{versinnote}
+\tl{muhūrtadvayaparyantaṃ nirbhayaś cālayed imām |\\+}
+\tl{ūrdhvam ākṛṣyate kiṃcit suṣumnā kuṇḍalīyutā ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{muhūrtadvayaparyantaṃ nirbharaṃ cālanād dhi vai |\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (f. 112r)
+\begin{versinnote}
+\tl{muhūrtadvayaparyantaṃ nirbharaṃ cālanād asau |\\+}
+\tl{ūrdhvam ākṛṣyate kiṃcit suṣumnākuṇḍalīgatam ||\\+}
+\tl{ku[ṇ]ḍalyās tadānīṃ gatam upari yātaṃ kiṃ cit svalpamātra abhyāsasadṛśam ity arthaḥ [|] akṛṣyate uccaiḥ karoti [|] \\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_099]
+%MD: nirbharaṃ vs nirbhayaś discussed?
+% ?? JB we should adopt nirbhayaś as alphaOne, alphaTwo and the source text have it. Yes. JM: also we should adopt alpha reading, so it's suṣumṇā that is lifted up, as in GŚ.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.100}
+\begin{translation}[hp03_100]
+Extracted from \emph{suṣumnā} by this [practice], \emph{kuṇḍalinī} leaves it. As a result of this, \emph{prāṇa} automatically enters \emph{suṣumṇā}.
+\end{translation}
+
+\begin{sources}[hp03_100]
+\emph{Gorakṣaśataka} 23c-24b
+\begin{versinnote}
+\tl{tena kuṇḍalinī tasyāḥ suṣumnāyā mukhaṃ dhruvam |\\+}
+\tl{jahāti tasmāt prāno’yaṃ suṣumnāṃ vrajati svataḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_100]
+\emph{Haṭhatattvakaumudī} 44.25
+\begin{versinnote}
+\tl{tadā kuṇḍalinī tasyāḥ suṣumṇāyā mukhaṃ dhruvam |\\+}
+\tl{jahāti tasmāt prāṇo 'yaṃ suṣumṇāṃ vrajati svataḥ || \\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~112r
+\begin{versinnote}
+\tl{tena proktaśakticālanena vidhinā kuṇḍalinī tasyāḥ suṣumṇāyāḥ samuddhṛtā jahāti tasmāt prāṇo [']yaṃ suṣumṇāṃ vrajati svataḥ [|] suṣumṇāyā antaḥ kiṃ cit tatka[r]tṛkordhvākarṣaṇena samyag ūrdhvavihitā yadā kuṇḍalī bhūry antaḥ praviṣṭety arthaḥ [|]   tadāyaṃ va[hny]āpānamanobhiḥ sārdhaṃ vijitaḥ kuṃḍalīpadaṃ prāptaḥ prāṇavāyuḥ svataḥ svasmāt pārthivarājasavikāraśoṣam iti śeṣaḥ [|] jahāti kuṃḍalībodhe suṣumṇāṃtaḥ pātaprabhāvād vigatāśeṣabāhyavāhaprasaṃ[ga] iti bhāvaḥ || tasmād dhetoḥ suṣumṇaṃ gacchatīti kevalakuṃbhako bhavatīty arthaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_100]
+We have supplied `mouth' (\emph{mukha}) in our translation on the basis of the reading found in the source text (the \emph{Gorakṣaśataka}).
+\end{philcomm}
+
+\begin{metre}[hp03_100]
+Anuṣṭubh (c: ma-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.101}
+\begin{translation}[hp03_101]
+Therefore [the yogi] should regularly make \emph{arundhatī} move, she who contains speech. By making her move the yogi is freed from diseases.
+\end{translation}
+% ?? JB add testimony of J5 to apparatus: śabdagarbhām arundhatīm
+
+\begin{sources}[hp03_101]
+\emph{Gorakṣaśataka} 26cd–27ab
+\begin{versinnote}
+\tl{tasmāt saṃcālayen nityaṃ śabdagarbhāṃ sarasvatīm |\\+}
+\tl{yasyāḥ saṃcālanenaiva yogī rogaiḥ pramucyate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_101]
+Cf.~\emph{Haṭharatnāvalī} 2.122
+\begin{versinnote}
+\tl{sūryeṇa pūrayed vāyuṃ sarasvatyās tu cālayet |\\+}
+\tl{śabdagarbhācālanena yogī rogaiḥ pramucyate ||\\!}
+\end{versinnote}
+
+\emph{Yogalakṣaṇāvalī} f.~31r  (attr.~to the \emph{Gorakṣaśata})
+\begin{versinnote}
+\tl{tasmāt saṃcālayen nityaṃ śabdagarbhāṃ sarasvatī |\\+}
+\tl{asyāḥ saṃcālanenaivaṃ rogā naśyaṃti niścitaṃ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~112r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tasmāt saṃcālayen nityaṃ śabdagarbhām arundhatīm ||\\+}
+\tl{yasyāḥ saṃcālanenāśu yogī rogaiḥ pramucyate ||\\!}
+\end{versinnote}
+
+%Yogasārasaṅgraha (p.59)
+%\begin{versinnote}
+%\tl{tasmāt saṃcālayen nityaṃ śaṃbhugarbhām aruṃdhatīm |\\+}
+%\tl{yasyāḥ sañcālanenāśu yogī rogaiḥ pramucyate ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_101]
+%comment on śabdagarbhām aruṃdhatīṃ.
+In the \emph{Gorakṣaśataka} (26cd–27ab), the source text, this verse occurs in a passage on \emph{sarasvatīcālana}, which is the practice of moving the tongue (i.e.~\emph{sarasvatī}) by wrapping a cloth  around it and tugging it in order to raise \emph{kuṇḍalinī}.
+In the \emph{Gorakṣaśataka}, \emph{sarasvatī} is also called \emph{arundhatī} and, since the tongue is instrumental for speech and \emph{sarasvatī} is the name of a Goddess identified with speech (\emph{vāc}), the \emph{Gorakṣaśataka}'s reading of \emph{śabdagarbhāṃ sarasvatīm} makes good sense.
+However, it seems that Svātmārāma has changed 3.101b to read \emph{śabdagarbhām arundhatīm} and has understood \emph{arundhatī} as \emph{kuṇḍalinī}.
+This is affirmed in longer versions of the \emph{Haṭhapradīpikā} (3.102??) that contain a verse on synonyms of \emph{kuṇḍalinī}, which includes \emph{arundhatī}.
+We are yet to find \emph{arundhatī} equated with  \emph{kuṇḍalinī} in a text composed before the \emph{Haṭhapradīpikā} but this identification is found in subsequent compendiums and commentaries (e.g.~\emph{Yogacintāmaṇi} f.~78v, \emph{Yuktabhavadeva} 7.300, \emph{Jyotsnā} 104, 119, \emph{Yogaprakāśikā} 5.166).%??JM: I think Svātmārāma prob understood the practice as the GŚ does, but later scribes etc.
+didn't.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.102}
+\begin{translation}[hp03_102]
+The yogi who has made kuṇḍalinī move is worthy of success. There is no point in speaking at length about this. He easily conquers death.
+\end{translation}
+
+%\begin{sources}[hp03_102]
+%\end{sources}
+
+\begin{testimonia}[hp03_102]
+\emph{Haṭharatnāvalī} 2.123
+\begin{versinnote}
+\tl{yena saṃcālitā śaktiḥ sa yogī siddhibhājanaḥ |\\+}
+\tl{kim atra bahunoktena mṛtyuṃ jayati līlayā || 2.123 ||\\!}
+\end{versinnote}
+
+\emph{Yogalakṣaṇāvalī} (f. 31r)  (attrib. Gorakṣaśata)
+\begin{versinnote}
+\tl{yena saṃcālitā śaktiḥ śabdagarbhā tv aruṃdhatī |\\+}
+\tl{kim atra bahunoktena tasya kālabhayaṃ na hi ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~112r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{yena saṃcālitā śaktiḥ sa yogī siddhibhājanaṃ |\\+}
+\tl{kim atra bahunoktena kālaṃ jayati līlayā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_102]
+%siddhibhājana is also possible.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.102*1}
+\begin{translation}[hp03_102_1]
+For [the yogi] devoted to celibacy and always eating a healthy and measured diet, success appears after forty days (\emph{maṇḍalāt}) as a result of engaging in the practice of \emph{kuṇḍalinī}.
+\end{translation}
+%greyscaled
+
+%\begin{sources}[hp03_102_1]
+%\end{sources}
+
+\begin{testimonia}[hp03_102_1]
+\emph{Yogacintāmaṇi} f.~79r (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{brahmacaryaratasyaiva nityaṃ hitamitāśinaḥ |\\+}
+\tl{maṇḍalād dṛśyate siddhiḥ kuṇḍalyabhyāsayoginaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogalakṣaṇāvalī} (f. 31r)  (attrib. to Gorakṣaśataka)
+\begin{versinnote}
+\tl{brahmacaryavratasyaiva kuṃḍalyabhyāsayoginaḥ ||\\+}
+\tl{maṇḍalād dṛśyate siddhir iti yogavido viduḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā}  f.~112r--112v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{brahmacaryajatasyaiva nityaṃ hitamitāṃ śanaiḥ |\\+}
+\tl{maṇḍalād dṛśyate siddhiḥ kuṇḍalyabhyāsayoginaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_102_1]
+Verse 3.102*1 has been omitted by \alphaOne\ and \alphaThree.
+It is in \alphaTwo\ at the end of a block of verses (3.94*2–5, 94*7) that is excluded by \alphaThree.
+This block appears after 3.97 and appears to have been inserted from elsewhere.
+3.102*1 has no known source and appears to have been added as a general laudatory statement on the benefits of practising with \emph{kuṇḍalinī}.
+
+The meaning of \emph{maṇḍalād} in 3.102*1c is not clear. Brahmānanda understands it as a period of time (i.e., forty days) but we are yet to find this attested elsewhere.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.103}
+\begin{translation}[hp03_103]
+The yogi should mix with ash the fluid of the moon emitted as a result of the practice. Wearing that [mixture] on the head bestows divine sight.
+\end{translation}
+
+%\begin{sources}[hp03_103]
+%\end{sources}
+
+\begin{testimonia}[hp03_103]
+\emph{Yogalakṣaṇāvalī} f.~31r  (attr.~to \emph{Gorakṣaśataka})
+\begin{versinnote}
+\tl{abhyāsaniḥsṛtāṃ cāndrīṃ vibhūtyā saha miśrayet |\\+}
+\tl{taddhāraṇaṃ cottamāṃge divyadṛṣṭipradāyakaṃ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~112v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{abhyāsaniḥsṛtāṃ cāndrīṃ vibhūtyā saha miśrayet [|]\\+}
+\tl{taddhāraṇaṃ tattamāṃge divyadṛṣṭipradāyakaṃ [||] 19\\+}
+\tl{cāndrīṃ lalāṭacandrān niḥsṛtāṃ abhyāse śramajātāṃ gharmadhārāṃ tāṃ vibhūtyā vimiśrayet | tām uttamāṃge śirasi dhārayed asau sādhakasya divyadṛṣṭipradā bhravatīty arthaḥ [|]\\!}
+\end{versinnote}
+
+\emph{Haṭhayogasaṃhitā} p. 41 (on \emph{amarolī})
+\begin{versinnote}
+\tl{abhyāsān niḥsṛtāṃ cāndrīṃ vibhūtyā saha miśrayet |\\+}
+\tl{dhārayed uttamāṅgeṣu divyadṛṣṭiḥ prajāyate || \\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_103]
+In the important manuscripts of the \emph{Haṭhapradīpikā}, including those of \textalpha, this verse occurs in the section on \emph{śakticālana}.
+This is also the case in the \emph{Yogalakṣaṇāvalī}, \emph{Haṭhasaṅketacandrikā} and the longer recensions of the \emph{Haṭhapradīpikā} with six and ten chapters.
+However, in the context of \emph{śakticālana}, the referent of \emph{cāndrī} is unclear.
+It appears to be understood as some sort of lunar fluid.
+Sundaradeva (see the \emph{Haṭhasaṅketacandrikā} in the testimonia) defines it as a flow of perspiration (\emph{gharmadhārā}) that arises from exertion in the practice and is emitted from moon in forehead (\emph{lalāṭacandra}).
+In \emph{Yogaprakāśikā}, Bālakṛṣṇa glosses it simply as nectar (\emph{sudhā}) (5.182).
+In the \emph{Haṭhayogasaṃhitā} (p.
+41) and \emph{Jyotsnā} (3.98), this verse is in the section on \emph{amarolī}, which provides a clear referent of \emph{cāndrī} as the cool middle flow of urine (see 3.95).
+
+%Verse 120 is more appropriate for finishing the śakticālana section.
+%Note Brahmānanda's interpretation of uttamāṅgeṣu, but usually uttamāṅga refers to the head.
+\end{philcomm}
+
+\begin{metre}[hp03_103]
+Anuṣṭubh (c: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.103*1}
+\begin{translation}[hp03_103_1]
+For purifiying the seventy-two thousand channels, there is no method of cleansing without the practice of \emph{kuṇḍalinī}.
+\end{translation}
+%greyscaled
+% ??MD: better to adapt abhyasanād ṛte? 
+%\begin{sources}[hp03_103_1]
+%\end{sources}
+
+\begin{testimonia}[hp03_103_1]
+\emph{Yogacintāmaṇi} f.~79v (attr.~to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{dvisaptatisahasrāṇāṃ nāḍīnāṃ malaśodhanam |\\+}
+\tl{kutaḥ prakṣālanopāyaḥ kuṇḍalyabhyāsato vinā ||\\!}
+\end{versinnote}
+
+\emph{Yogalakṣaṇāvalī} f.~31r  (attrib. to Gorakṣaśataka)
+\begin{versinnote}
+\tl{dvisaptatisahasrāṇāṃ nāḍīnām api śodhanaṃ |\\+}
+\tl{asatkalpaṃ smṛtaṃ siddhaiḥ kuṃḍalobhyasānād ṛte ||\\!}
+\end{versinnote}
+
+%\emph{Upāsanāsārasaṅgraha} (p. 36) % IFP Transcript T1095
+%\begin{versinnote}
+%\tl{dvāsaptatisahasrāṇāṃ nāḍīnāṃ malaśodhane | \\+}
+%\tl{kutaḥ prakṣālanopāyaḥ kuṇḍalyabhyasanād ṛte |\\!}
+%\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~112v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{dvisaptatisahastrāṇāṃ nāḍīnāṃ malaśodhanaṃ |\\+}
+\tl{kutaḥ prakṣālanopāyaḥ kuṃḍalyābhyāsanād ṛte ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_103_1]
+This verse is omitted by the \textalpha\ group, and was probably added to the original text as a further laudatory statement on the practice of \emph{kuṇḍalinī}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.104}
+\begin{translation}[hp03_104]
+Thus have the ten \emph{mudrā}s been taught by Śiva Ādinātha. Each one among them can bestow liberation for ascetics.
+\end{translation}
+%?? JB put iti śakticālana below the line in the apparatus (as its not in alpha)?
+
+\begin{sources}[hp03_104]
+\end{sources}
+
+\begin{testimonia}[hp03_104]
+
+\emph{Haṭharatnāvalī} 2.35
+\begin{versinnote}
+\tl{iti mudrā daśa proktā ādināthena śambhunā |\\+}
+\tl{ekaikā tāsu mukhyā syān mahāsiddhipradāyinī ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~79v
+\begin{versinnote}
+\tl{iti mudrā nava proktā ādināthena śambhunā |\\+}
+\tl{ekaikā tāsu yamināṃ mahāsiddhipradāyinī ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp03_104]
+Manuscripts of the \textbeta, \texteta, and \textepsilon\ groups have a different reading for the second line; `each mudrā is capable of being a cause of all powers' (\emph{kāraṇaṃ sarvasiddhīnāṃ ekaikāpi kṣamaiva sā}).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.105}
+\begin{translation}[hp03_105]
+Without a king the earth is not resplendent (\emph{rājate}), without the moon the night does not sparkle (\emph{rājate}), without Rājayoga even the wonderful [practice of] \emph{mudrā} does not shine(\emph{rājate}).
+\end{translation}
+
+%\begin{sources}[hp03_105]
+
+%\end{sources}
+
+\begin{testimonia}[hp03_105]
+\emph{Haṭharatnāvalī} 1.16
+\begin{versinnote}
+\tl{rājayogaṃ vinā pṛthvī rājayogaṃ vinā niśā |\\+}
+\tl{rājayogaṃ vinā mudrā vicitrāpi na rājate ||\\!}
+\end{versinnote}
+
+%Yogasārasaṅgraha (p. 59)
+%\begin{versinnote}
+%\tl{rājayogaṃ vinā pṛthvī rājayogaṃ vinā niśi |\\+}
+%\tl{rājayogaṃ vinā mudrā vicitrāpi na rājate ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_105]
+The \emph{Yogaprakāśikā} (5.186) interprets the similes in this verse as we have translated them, `Just as without a king [and] moon, the earth and night do not shine...' (\emph{yathā mahīpālaṃ candramasaṃ vinā pṛthvīniśe na rājete}...).
+However, in \emph{Jyotsnā} 3.126, Brahmānanda interprets the earth (\emph{pṛthvī}) as \emph{āsana} because both are connected by the quality of steadiness (\emph{sthairyaguṇayogāt}), and the night (\emph{niśā}) as breath retention (\emph{kumbhaka}) because both are characterised by the absence of movement of people and wind (\emph{prāṇasañcārābhāvalakṣaṇaḥ}).
+Brahmānanda's interpretation seems somewhat far-fetched.
+\end{philcomm}
+
+\begin{metre}[hp03_105]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{3.106}
+\begin{translation}[hp03_106]
+[The yogi] should carry out all breath practice with his mind engaged. The wise man must not let his attention wander.
+\end{translation}
+
+%\begin{sources}[hp03_106]
+%\end{sources}
+
+\begin{testimonia}[hp03_106]
+\emph{Haṭhasaṅketacandrikā} f.~92v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{mārutābhyasanaṃ kiṃ cin manoyuktaṃ samācaret |\\+}
+\tl{itaratra na kartavyā manovṛttir manīṣiṇā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp03_106]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.107}
+\begin{translation}[hp03_107]
+By means of postures, breath retentions and \emph{mudrā}s, the central channel, even though untraversed/blocked, becomes straight through yogis' firm practice.
+\end{translation}
+%??
+JB: khila can mean uncultivated (see note below) but I havent found a reference for the meaning of `blocked' or 'untraversed'?
+JM: I guess untraversed was getting at that meaning; it needs to contrast with sarala; untraversed is quite good in fact.
+
+%\begin{sources}[hp03_107]
+%\end{sources}
+
+\begin{testimonia}[hp03_107]
+\emph{Upāsanāsārasaṅgraha} p. 36 % IFP Transcript T1095
+\begin{versinnote}
+\tl{iyaṃ tu madhyamā nāḍī dṛḍhābhyāsena yoginām  |\\+}
+\tl{āsanaprāṇasaṃyāmamudrābhiḥ saralā bhavet  ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_107]
+The reading of \emph{khilāpi} in the first verse quarter is unusual but well attested by the witnesses including the \textalpha\ group of manuscripts.
+In the \emph{Abhidhānacintāmaṇi} (940), \emph{khila} is defined as something uncultivated such as field (\emph{kṣetrādyaprahataṃ khilam}) or, as the \emph{Amarapadavivṛti} (2.1.5) puts it, `not marked by a plough' (\emph{lāṅgalena na likhitam iti khilam}).
+In the context of \emph{Haṭhapradīpikā} 3.107, qualifying \emph{suṣumnā} with \emph{khila} implies that the central channel has yet to be cultivated by a yoga practice.
+
+%khila seems odd (could it mean defective or unperfected here?), but the alternative readings are not helpful (except iyaṃ tu, which seems like a patch).
+%Philipp: Apte: khila can mean blocked or a wedge
+%Haru: khila - untraversed (aprahata), common in kośas.
+%Group 6 has kuṭilā for khilāpi. Unlikely to be original but makes sense.
+
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.108}
+\begin{translation}[hp03_108]
+For those who are tireless in their dedication, Rājayoga has a \emph{mudrā}. That is the supreme \emph{rudrāṇi} \emph{mudrā}, which bestows beneficial success.
+\end{translation}
+
+%\begin{sources}[hp03_108]
+%\end{sources}
+
+%\begin{testimonia}[hp03_108]
+%\end{testimonia}
+
+\begin{philcomm}[hp03_108]
+%Consider J5 emendation rājayogasamudbhavā ('produced from Rājayoga')
+%samudraka : rājayoga has a mudrā, is sealed, has the seal of approval,
+\emph{Rudrāṇī} may be the \emph{mudrā} usually called \emph{śāmbhavī}.
+
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.108*1}
+\begin{translation}[hp03_108_1]
+May [the yogi] who offers the traditional teaching of the \emph{mudrā}s be the guru, the master. He is none but the Lord himself.
+\end{translation}
+%greyscaled
+
+%\begin{sources}[hp03_108_1]
+%\end{sources}
+
+\begin{testimonia}[hp03_108_1]
+\emph{Upāsanāsārasaṅgraha} p. 40 % IFP Transcript T1095
+\begin{versinnote}
+\tl{upadeśaṃ hi mudrāṇāṃ yo datte sāṃpradāyikam |\\+}
+\tl{sa eva śrīguruḥ svāmī sākṣād īśvara eva saḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp03_108_1]
+Verses 3.108*1–2 have no known source and are absent in the \textalpha\ group. It is likely both were added to the original text as further praise of those practising the haṭhayogic \emph{mudrā}s.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.108*2}
+\begin{translation}[hp03_108_2]
+The yogi who has become intent on that [guru's] teaching and practises with a focused mind obtains mastery of the powers beginning with minimisation and the cheating of death.
+\end{translation}
+%greyscaled
+
+%\begin{sources}[hp03_108_2]
+%\end{sources}
+
+%\begin{testimonia}[hp03_108_2]
+%\end{testimonia}
+
+\begin{philcomm}[hp03_108_2]
+No version of this verse is entirely satisfactory. See the note on 3.108*1 for why it is in greyscale.
+\end{philcomm}
 

### 2024-03-19 latex/HP4_comm.tex
adapted to new input files and cleanup
new files and Nagari-Support for Jyotsna

--- 2023-12-02 latex/HP3_comm.tex
@@
 
-\chapter*{Translation \& philological commentary}
+%\chapter*{Translation \& philological commentary}
+
 %%%%%%%%%%
-\subsection*{3.1}
-\begin{translation}[hp03_001]
-test.
+\subsection*{4.0*1}
+\begin{translation}[hp04_000_1] 
+Homage to the guru, Śiva, who consists of \emph{nāda}, \emph{bindu} and \emph{kalā}. [The yogi] who is constantly devoted to him attains the untainted state (\emph{nirañjanapada}).
+\end{translation}
+
+%\begin{sources}[hp04_000_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_000_1]
+\emph{Haṃsavilāsa} 14 (p. 47)
+\begin{versinnote}
+\tl{namaḥ śivāya gurave nādabindukalātmane |\\+}
+\tl{nirañjanapadaṃ yāti yatra yogī parāyaṇaḥ iti dhyānam || 14 ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_000_1]
+%The compound \emph{yatnaparāyaṇa} is rather unusual and generally not attested elsewhere (the only exception known  to us is the \emph{Amarasiddhi}).
+The reading  \emph{yogaparāyaṇa} in some secondary manuscripts makes much better sense (i.e., 'one devoted to yoga') but is likely a substitution for a more difficult reading.
+%adopt yatraparāyaṇaḥ [MD: done], and assume tasmai in 4.1ab
+
+%Look into referent of nirañjana in these texts. Judit: best to take it generally (like nirāmaya, etc.).
+Verses 4.0*1–4.0*16 are omitted by the \textalpha\ group and are likely not to be original.
+The first additional verse resembles a \emph{maṅgala} verse that one might expect to see at the beginning of a text.
+The second is a verse from the \emph{Gorakṣaśataka} that introduces the topic of \emph{samadhi}.
+Some manuscripts of the \textbeta, \textepsilon\ and \texteta\ groups have the two verses on the synonyms of \emph{rājayoga} here (on their position in the \textalpha\ group and other manuscripts, see the note to 4.29).
+The rest of the additional verses (4.0*5–4.0*16) are a motley collection on \emph{samādhi}, \emph{rājayoga}, the importance of the guru, dissolving the breath, \emph{suṣumṇā}, etc.
+In contrast to this, the \textalpha\ group begins with a cohesive discusson on absorption (4.1–3) that transitions to the gaze (4.4) and a brief discussion of \emph{śāmbhavī} and \emph{khecarī mudrā}s (4.5–4.8).
+The main topic of the chapter, which is meditating on the internal sound (\emph{nādānusandhāna}), begins at 4.9 in the \textalpha\ group (whereas in other groups it begins after fifty or so verses).
+The emphasis on \emph{nādānusandhāna} in the fourth chapter of the \textalpha\ group is consistent with the statement in verse 1.56 that \emph{nādānusandhāna} is the fourth component of Haṭhayoga.
+
+Its likely that the term \emph{nirañjanapada} was understood here as \emph{samādhi} because \emph{nirañjana} is included in a list of synonyms of \emph{samādhi} later in this chapter (4.29).
+
+The triad \emph{nāda}, \emph{bindu} and \emph{kalā} occurs in earlier works, in particular Śaiva Tantras, where it appears in contexts of enunciating mantras (\emph{mantroccāra}, e.g., \emph{Kubjikāmatatantra} 7.65, \emph{Jñānārṇavatantra} 2.4, \emph{Īśānaśivagurudevapaddhati} 18-110, etc.) and sometimes qualifies dieties (e.g., \emph{Parākhyatantra} 5.156ab) and gurus (e.g., \emph{Gurugītā} 64).
+The context can change the meaning of these terms, so we have chosen not to translate them.
+For a discussion of their various meanings, see \emph{Tāntrikābhidhānakośa} 2004 vol.
+2: 68–73, 2013 vol.
+3, 277–279.
+
+%nāda-bindu-kalā, referring to inner sound, generative fluid, and digits of the moon (or unmanīkalā) in the context of HY? Judit: kalā is like śakti (initial source)?
+%Refer to TĀK
+
+\end{philcomm}
+
+\begin{metre}[hp04_000_1]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.0*2}
+\begin{translation}[hp04_000_2]
+So now I will teach the best way to \emph{samādhi}.It destroys death, has an easy method and brings about the bliss of Brahman.
+\end{translation}
+
+\begin{sources}[hp04_000_2]
+\emph{Gorakṣaśataka} 64
+\begin{versinnote}
+\tl{athedānīṃ pravakṣyāmi samādhikramam uttamam | \\+}
+\tl{mṛtyughnaṃ tu sukhopāyair brahmānandakaraṃ sadā ||\\+}
+\tl{\var{64c tu sukhopāyair ] T; sukhadopāyaṃ GU}\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_000_2]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_000_2]
+%It appears that the \emph{Gorakṣaśataka}'s verse was changed by whoever added it to reflect the discursive nature of the discussion on \emph{samādhi} that follows.
+By changing \emph{samādhikramam uttamam} to \emph{samādhikramalakṣaṇam} (and perhaps \emph{sadā} to \emph{param}) the verse foreshadows a broader discussion of the topic samādhi.
+But it doesn't really work, so adopt uttamam.
+[JB: We have uttamam now, so this note is not needed]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*5}
+\begin{translation}[hp04_000_5]
+The unity of the self and mind arises in the same way that salt becomes identical with water through contact [with it]. That is called \emph{samādhi}.
+\end{translation}
+
+\begin{sources}[hp04_000_5]
+\emph{Vivekamārtaṇḍa} 161
+\begin{versinnote}
+\tl{ambusaindhavayoḥ sāmyaṃ yathā bhavati yogataḥ |\\+}
+\tl{tathātmamanasor aikyaṃ samādhiḥ so’bhidhīyate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_000_5]
+\emph{Haṭharatnāvalī} 4.1
+\begin{versinnote}
+\tl{salile saindhavaṃ yadvat sāmyaṃ bhavati yogavit |\\+}
+\tl{tathātmamanasor aikyaṃ samādhiḥ so 'bhidhīyate ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 11.29 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{ambusaindhavayor aikyaṃ yathā bhavati yogataḥ |\\+}
+\tl{tathātmamanasor aikyaṃ samādhir abhidhīyate ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 51.72
+\begin{versinnote}
+\tl{tad uktaṃ yogacandrikāyām –\\+}
+\tl{ambusaindhavayor aikyaṃ yathā bhavati yogataḥ |\\+}
+\tl{tathātmanasor aikyaṃ samādhiḥ sa vidhīyate ||\\+}
+\tl{aikyaṃ abhinnatvam | yogataḥ yogābhyāsāt ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_000_5]
+%The change \emph{salile saindhavaṃ} appears to have occured because of the verb \emph{bhajati}, which is different to the \emph{Vivekamartaṇḍa}'s reading of \emph{ambusaindhavayor aikyaṃ [...] bhavati [...]}.
+However, since groups 2 and 3 support \emph{salile}, it is possible that Svātmārāma accepted it.
+%SS sees this verse as a nirukti of samādhi using sāmyam
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*6}
+\begin{translation}[hp04_000_6]
+The identity of the individual and universal selves is called \emph{samādhi}, in which all thoughts disappear.
+\end{translation}
+
+\begin{sources}[hp04_000_6]
+\emph{Vivekamārtaṇḍa} 163
+\begin{versinnote}
+\tl{yat samatvaṃ dvayor atra jīvātmaparamātmanoḥ |\\+}
+\tl{samastanaṣṭasaṃkalpaḥ samādhiḥ so’bhidhīyate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_000_6]
+\emph{Haṭharatnāvalī} 4.2
+\begin{versinnote}
+\tl{tat samatvaṃ bhaved atra jīvātmaparamātmanoḥ |\\+}
+\tl{samastanaṣṭasaṃkalpaḥ samādhiḥ so 'bhidhīyate ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 11.28 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{yat sarvadvandvayor aikyaṃ jīvātmaparamātmanoḥ |\\+}
+\tl{samastanaṣṭasaṃkalpaḥ samādhiḥ so 'bhidhīyate ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp04_000_6]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*7}
+\begin{translation}[hp04_000_7]
+Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation †arises† [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
+\end{translation}
+
+\begin{sources}[hp04_000_7]
+\emph{Amanaska} 2.5
+\begin{versinnote}
+\tl{rājayogasya māhātmyaṃ ko vā jānāti tattvataḥ |\\+}
+\tl{jñānāt siddhir muktir iti guror jñānaṃ ca labhyate ||\\+}
+\tl{\var{5c jñānāt siddhir muktir iti ] Bl Jb K Ma Va:
+jñānāt siddhimuktir iti Pa Pc Tr Ua Ea:
+jñānasiddhir muktir iti Vb Vd:
+jñānān mukteḥ siddhir iti Ja:
+jñānāt sidhyati muktir hi AllN (except Na Nm Ve Ea):
+jñānāt sidhyate muktir hi Ve:
+jñānāt sidhyanti muktiṃ hi Na:
+jñānāt sidhyati muktiṃ hi Nm:
+nānāsiddhir muktir iti Tha:
+jñānasiddhir bhavaty eva Mb:
+jñānasiddhimuktisiddhi Pb:
+jñānāt siddhimuktor iti Je}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_000_7]
+\emph{Yogacintāmaṇi} f.~37v (citing the \emph{Rājayoga} [aka \emph{Amanaska}])
+\begin{versinnote}
+\tl{rājayogasya māhātmyaṃ ko hi jānāti tattvataḥ |\\+}
+\tl{tajjñānī vasate yatra sadeśaḥ puṇyabhājanam ||\\!}
+\end{versinnote}
+
+%\emph{Yogamārgaprakāśikā} 4.11
+%\begin{versinnote}
+%\tl{rājayogasya māhātmyaṃ ko vā śaknoti varṇitum | \\+}
+%\tl{yogasyāsya ca kartāro vijñeyās te maheśvarāḥ ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_000_7]
+The third verse quarter of this verse has been subject to much revision in the \emph{Haṭhapradīpikā} and the source text, the \emph{Amanaska}.
+Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}.
+We are unable to make sense of \emph{sthiti} here, and so have cruxed this verse quarter and emended to \emph{muktiḥ sthitā} (`liberation has happened').
+%??
+JB.
+Rather than emend to muktiḥ sthitā, I think we should consider muktisthitiḥ (etaTwo and deltaOne), which is similar to jīvanmuktasthitiṃ in the Mokṣopāya, LYV, MSS, etc.
+We could translate it as Sternbach below ('state of liberation').
+JM: I also don't think we should emend.
+muktisthitiḥ is possible (but MU parallel not great because of mukti vs mukta).
+We could also consider muktiḥ sthitiḥ “liberation [and] stability”
+% Mokṣopāya
+% jīvanmuktasthitiṃ brahman kṛtvā rāghavam āditaḥ|\\
+% kramāt kathaya me nityaṃ bhaviṣyāmi sukhī yathā||\,1||
+%  "jīvanmuktasthitiṃ" jīvanmuktamaryādāṃ
+% Yogavāsiṣṭha
+% vicārajā nijā jīvanmuktasthitir ihoditā |
+% muninābhyāsahīnasya vyādhasya tvanavasthitiḥ || VII. 
+% MSS_9744-1 
+% kāryākārye kim api satataṃ naiva kartṛtvam asti  
+% jīvanmuktasthitir avagato dagdhavastrāvabhāsaḥ /<BR>
+% There is not always an agent in every action, good or bad ; for, the state of a jivanmukta [one who is released from bondage while living in the world] is considered to be similar to that of the splendour of a burnt cloth [which can neither be clean nor dirty].
+(Sternbach's translation)
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*8}
+\begin{translation}[hp04_000_8]
+Letting go of sense objects, seeing the truth, [and] realising the innate state are difficult without the compassion of a good guru.
+\end{translation}
+
+%\begin{sources}[hp04_000_8]
+%\end{sources}
+
+%\begin{testimonia}[hp04_000_8]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_000_8]
+%No source?
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*9}
+\begin{translation}[hp04_000_9]
+When the primal \emph{śakti} (i.e.~\emph{kuṇḍalinī}) has been woken up by means of the various postures, retentions and wonderful techniques [i.e.~\emph{mudrā}s], the breath dissolves into the void.
+\end{translation}
+
+%\begin{sources}[hp04_000_9]
+%\end{sources}
+
+\begin{testimonia}[hp04_000_9]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{vividhair āsanaiḥ kumbhair vicitrakaraṇair api |\\+}
+\tl{prabuddhāyām ādiśaktau prāṇaḥ śūnye vilīyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_000_9]
+In \emph{Jyotsnā} 4.10, Brahmānanda understands `the void' (\emph{śūnya}) as the central channel. In \emph{Haṭhapradīpikā} 3.4, \emph{śūnyapādavī} is a synonym of \emph{suṣumnā}.
+\end{philcomm}
+%?? MD: vilīyate or pralīyate?
+
+\begin{metre}[hp04_000_9]
+Anuṣṭubh (c: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.0*10}
+\begin{translation}[hp04_000_10]
+For the yogi whose \emph{kuṇḍalinī} has awakened and who has given up all activity, the innate state automatically arises.
+\end{translation}
+
+%\begin{sources}[hp04_000_10]
+%\end{sources}
+
+\begin{testimonia}[hp04_000_10]
+\emph{Yogacintāmaṇi} f.~9r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{utpannaśaktibodhasya tyaktaniḥśeṣakarmaṇaḥ |\\+}
+\tl{yoginaḥ sahajāvasthā svayam eva prajāyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_000_10]
+%\end{philcomm}
+
+%?? MD: prajāyate or prakāśate?
+
+%%%%%%%%%%
+\subsection*{4.0*11}
+\begin{translation}[hp04_000_11]
+When the breath is flowing in the central channel and the mind enters the void, the knower of yoga uproots all [past] actions.
+\end{translation}
+%?? JB: all [fruits of past] actions
+
+\begin{sources}[hp04_000_11]
+\end{sources}
+
+\begin{testimonia}[hp04_000_11]
+\emph{Yogacintāmaṇi} f.~9r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{suṣumṇāvāhini prāṇe śūnye viśati mārute |\\+}
+\tl{tathā samastakarmāṇi nirmūlayati yogavit ||\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~66 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{suṣumṇāvāhini prāṇe śūnye viśati mānase |\\+}
+\tl{tadā samastakarmāṇi nirmūlayati marmavit ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_000_11]
+%\emph{tathā} may connect the two locative absolutes.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*12}
+\begin{translation}[hp04_000_12]
+Homage to you, the immortal one. You have slain even death, into whose mouth this world, with everything that is moving and unmoving, has fallen.
+\end{translation}
+
+\begin{sources}[hp04_000_12]
+\end{sources}
+
+\begin{testimonia}[hp04_000_12]
+\end{testimonia}
+
+%\begin{philcomm}[hp04_000_12]
+%It appears that this verse, which is in manuscripts of the \textbeta\ and \textepsilon\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
+%?? JB should we include this note on gamma's omission? Is it necessary? JM: no, I've percented it
+
+%JT: alt 3 is original, makes best sense, others lost pādas b and c and tried to correct. Adopt. [MD: done]
+%Eye skip might explain how amarāya namas became amano nirmanaḥ śūnyaṃ jagad ... and the tadāmaraugha was dropped because it was difficult to understand or was corrupt.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*13}
+\begin{translation}[hp04_000_13]
+When the mind has obtained equanimity and the breath is moving into the central channel, then the \emph{vajrolī} of the lineage of immortals arises even for the person whose hope is for that.
+\end{translation}
+% ??
+JB: we didnt translate this verse in the workshop.
+I've adapted my translation from a similar verse in the Amaraugha.
+JM: equanimity is tautologous with mind, “When equanimity has been obtained"?
+Also I've changed tr.
+to reflect tadā.
+
+\begin{sources}[hp04_000_13]
+\emph{Amaraugha} 7
+\begin{versinnote}
+\tl{citte tu sattvam āpanne vāyau vrajati madhyame |\\+}
+\tl{tadāmaraughavajrolis tadāśājīvitasya ca ||\\+}
+\tl{\var{citte tu sattvam ] $\beta$ : citte samatvam Ba Ad Ed : cittau samatvam Gb Ta.
+tadāmaraughavajrolis ] \textit{conj.} :
+tadāmaraughavajroli Ga :
+tadāmaraudhavajrolī Ae :
+eṣāmarauli vajroli Gb :
+eṣāmarāli vajroli Ba :
+eṣāmarauli vajrauli Ta :
+yeṣāmarāli vajrauli Ad :
+eṣāmarolī vajrolī Ed.
+tadāśājīvitasya ca ] $\beta$ :
+tadā meni mameti ca Ba :
+tadā mati mateti ca Ed :
+mameti ca mameti ca Ad :
+tadā meti mameti mateti ca Gb (\textit{unmetr.}) :
+mateti mame++ Ta.
+}\\!}
+\end{versinnote}
+
+\end{sources}
+
+%\begin{testimonia}[hp04_000_13]
+%\end{testimonia}
+
+\begin{philcomm}[hp04_000_13]
+The second line of this verse is problematic in both the source text (the \emph{Amaraugha}) and the manuscripts of the \emph{Haṭhapradīpikā}.
+It seems that the \textepsilon\ group has preserved the most plausible version of this line, which is also quite close to the source text.
+The line was rewritten in \etaTwo\ and other manuscripts, as well as the \emph{Jyotsnā} (4.14cd), to include the two variations of \emph{vajrolī}, namely \emph{amarolī} and \emph{sahajolī}.%??JM: comment on pāda d?
+It's odd, worth cruxing even.
+%C2 (Grp 7) tadāmarolī ... sahajolī prajāyate (=Jyotsnā)
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*14}
+\begin{translation}[hp04_000_14]
+How can knowledge exist, when the mind is alive, O goddess?
+So long as the breath lives, the mind does not die.
+Breath and mind: he who leads this pair to dissolution attains liberation; in no way [does any] other man.
+\end{translation}
+
+\begin{sources}[hp04_000_14]
+\emph{Candrāvalokana} 6
+\begin{versinnote}
+\tl{jñānaṃ kuto manasi jīvati devi yāvat \\+}
+\tl{prāṇo pi jīvati mano mriyate na yāvat |\\+}
+\tl{prāṇo mano dvayam idaṃ vilayaṃ prayāti\\+}
+\tl{mokṣaṃ sa gacchati naro na kathaṃcid anyaḥ ||\\+}
+\tl{\var{devi ] 7970 4340 T00788, kūdṛṣṭi D4345 (unmetr.)}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_000_14]
+\emph{Yogacintāmaṇi} f.~17r
+\begin{versinnote}
+\tl{atra śivavākyam |\\+}
+\tl{jñānaṃ kuto manasi jīvati devi tāvat\\+}
+\tl{prāṇo 'pi jīvati mano mriyate na yāvat |\\+}
+\tl{prāṇo mano dvayam idaṃ vilayaṃ nayed yo \\+}
+\tl{mokṣaṃ sa gacchati naro na kathañcid anyaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~2r
+\begin{versinnote}
+\tl{jñānaṃ kuto manasi jīvati durvikalpe \\+}
+\tl{prāṇe 'pi jīvati mano mṛyate na yāvat |\\+}
+\tl{prāṇo mano dvayam idaṃ na vilīyate 'tra \\+}
+\tl{mokṣa na gacchati naro 'tra kathaṃ cid eva ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_000_14]
+%\end{philcomm}
+%?? MD: adopt prayāti (beta) in Pāda C?
+
+\begin{metre}[hp04_000_14]
+Vasantatilakā 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.0*15}
+\begin{translation}[hp04_000_15]
+Both mercury and the mind are unsteady by nature. [When either] mercury is fixed [or] the mind is fixed, nothing in the world is impossible.
+\end{translation}
+
+%\begin{sources}[hp04_000_15]
+%\end{sources}
+
+\begin{testimonia}[hp04_000_15]
+\emph{Yogacintāmaṇi} f.~103v
+\begin{versinnote}
+\tl{śāstrāntare śivavākyam—\\+}
+\tl{rasasya manasaś caiva cañcalatvaṃ samāsataḥ |\\+}
+\tl{raso baddho mano baddhaṃ kiṃ na sidhyati bhūtale ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_000_15]
+%LO: rasa implies bindu (and the next verse is about bindu when in the AS). But the next verse is more likely from the Rasārṇava (look at the context there).
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.0*16}
+\begin{translation}[hp04_000_16]
+Stabilised, mercury and breath cure disease; stilled, they automatically bring back life; and bound they bestow the ability to fly, O Bhairavī.
+\end{translation}
+
+\begin{sources}[hp04_000_16]
+\emph{Rasārṇava} 1.19
+\begin{versinnote}
+\tl{mūrchito harati vyādhiṃ mṛto jīvayati svayam |\\+}
+\tl{baddhaḥ khecaratāṃ kuryāt raso vāyuś ca bhairavi ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Amṛtasiddhi} 7.7
+\begin{versinnote}
+\tl{mūrcchito harate vyādhiṃ baddhaḥ khecaratāṃ nayet |\\+}
+\tl{sarvasiddhikaro līno niścalo muktidāyakaḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_000_16]
+\emph{Yogacintāmaṇi} f.~103v (attr.~to \emph{śāstrāntare śivavākyam})
+\begin{versinnote}
+\tl{mūrchito harate vyādhiṃ mṛto jīvayati svayam |\\+}
+\tl{baddhaḥ khecaratāṃ dhatte mano vāyuś ca bhairavi ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_000_16]
+The \emph{Rasārṇava} is the likely source of this verse given the vocative, \emph{bhairavi}. Similar verses are found in other yoga texts (see Mallinson and Szántó 2021: 57).
+
+Disregarding the technical meaning of the terms, the verse plays with apparent contradictions (\emph{virodhābhāsa}) in saying literally that, when mercury and the breath are unconscious, they cure disease; when dead, they restore life and, when bound, they enable one to fly up.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.1}
+\begin{translation}[hp04_001]
+%Now, \emph{samādhi}–
+Mind is the master of the senses; breath is the master of the mind [and] dissolution [of the mind] is the master of the breath. [The yogi] should take refuge in that master, dissolution [of the mind].
+\end{translation}
+%??
+JB: Both \alphaOne and \alphaTwo have atha samādhiḥ as a heading before this verse.
+Shouldnt we include this in our edition or at least note it in the apparatus?
+We've added atha headings elsewhere from alpha.
+There's also the issue of where to place atha samādhi in relation to the greyscale verses.
+
+%\begin{sources}[hp04_001]
+%\end{sources}
+
+\begin{testimonia}[hp04_001]
+\emph{Haṭharatnāvalī} 4.4
+\begin{versinnote}
+\tl{indriyāṇāṃ mano nātho manonāthas tu mārutaḥ |\\+}
+\tl{mārutasya layo nāthaḥ sa layo nādam āśritaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~23r
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{indriyāṇāṃ mano nātho manonāthaś ca mārutaḥ |\\+}
+\tl{mārutasya layo nāthaḥ sa layo nādam āśritaḥ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp04_001]
+%
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.1*1}
+\begin{translation}[hp04_001_1]
+Whether or not this [dissolution] is called liberation  in another teaching, there is no difference in the bliss arising from the dissolution of mind and breath.
+\end{translation}
+%?? JB: 'in another doctrine' for matāntare?
+%?? JB: extraordinary bliss (kaścid?)
+%\begin{sources}[hp04_001_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_001_1]
+\emph{Yogacintāmaṇi} f.~23r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ayam eva tu mokṣākhyo astu vāpi matāntare |\\+}
+\tl{manaḥprāṇalayo nādo na ca kaś cit vibhidyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_001_1]
+Some of the \textbeta, \textepsilon\ and \texteta\ groups have a different version of the second half of this verse, the most plausibe of which is, 'the extraordinary bliss of dissolution of mind and breath arises in me' (\emph{manaḥprāṇalayānando mayi kaścit pravartate}).
+The first person pronoun suggests that this verse derives from a dialogue of some kind before it was changed to the version in the \textgamma\ and \textdelta\ manuscripts.
+\end{philcomm}
+
+% ?? MD: comment on the variant reading "manaḥprāṇalayānando mayi kaścit pravartate"?
+% JB: maybe this is the lectio difficilor and the first person pronoun was later written out.
+
+%%%%%%%%%%
+\subsection*{4.2}
+\begin{translation}[hp04_002]
+A motionless and unchanging dissolution [of mind], in which inhalation and exhalation have disappeared [and] perception of sense objects has ceased, is supreme for yogis.%??JM: or "The yogis’ motionless and unchanging dissolution...
+is the best."
+\end{translation}
+% on the translation of jayati as an intransitive verb, see Alexis' footnote on jayati in his forthcoming book on the Tantrāloka āhnika 1
+
+\begin{sources}[hp04_002]
+\emph{Amanaska} 2.21
+\begin{versinnote}
+\tl{pranaṣṭocchvāsaniśvāsaḥ pradhvastaviṣayagrahaḥ |\\+}
+\tl{niśceṣṭo nirgatārambho hy ānandaṃ yāti yogavit ||\\+}
+\tl{\var{pranaṣṭocchvāsa°  ] Jb Pa Va Bl Ja Ad AllSI  AllN, 
+pranaṣṭasvāsa°  Mb,
+pranaṣṭāsvāsa° Ua,
+praṇaśocchvāsa° Ma,
+pranaṣṭaḥ svāsa° Je Vb,
+pranaṣṭo śvāsa°ḥ Pc,
+prāṇastho śvāsa° Pb,
+pranaṣṭe svāsa° K}\\!} % retained because we've read against alpha.??jM: simplify? It's only the ucch that's different. JB: done.
+%\tl{\var{hy ānandaṃ yāti yogavit ]  AllNI (except  Ma Pa Pb Pc Tr Va),
+%hy ānandaṃ yāti tattvavit Pa Tr Va Ea,
+%svānandaṃ yāti yogavit Vd,
+%hy ānandayati yoginaḥ Ma Tha,
+%hy ānanday[a]ti yogī kaḥ Pc,
+%hy ānanda iva yoginaḥ AllN (except  Na Nm Ea Ve),
+%hy ānanda iva yogi Na Nm,
+%hy ānandaś ca yoginaḥ Ve,
+%hy ānandasukhayogavit Pb,
+%layo jayati yoginaḥ Ad,
+%layo jayati yoginām Cc}\\!} %retained to show that jayati was in the South Indian transmissin of the Amanaska.??JM: but it's in almost all HP witnesses too so perhaps variants not needed here, just mention it in comm?
+Or simplify?
+JB done
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_002]
+\emph{Yogacintāmaṇi} f.~27v (attr.~to the \textit{Rājayoga})
+\begin{versinnote}
+\tl{praṇaṣṭocchvāsaniḥśvāsa[ḥ] vidhvastaviṣayagrahaḥ |\\+}
+\tl{niśceṣṭo nirgatārambho hy ānandayati yogikaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~117 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pranaṣṭaśvāsaniḥśvāsaḥ pradhvastaviṣayajvaraḥ |\\+}
+\tl{niśceṣṭo nirvikāraś ca layo jayati yoginā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_002]
+%Adopt alpha reading. [MD: Done.]
+The fourth verse quarter of this verse differs from that of the North Indian recension of the \emph{Amanaska} (quoted as the source) but is found in two South Indian manuscripts of the \emph{Amanaska} (Ad and Cc).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.3}
+\begin{translation}[hp04_003]
+May an extraordinary absorption prevail, in which all volition has been cut off and all activity ceased, intelligible [only] by means of itself and ineffable.%??JM: which is intelligible > intelligible, and other tweaks.
+Prevail not great but can't think of anything else.
+%% My translation in the Amanaska (reading jāyate): [This] extraordinary absorption by which all volition has been cut off and in which all movement has ceased, is intelligible [only] by means of itself and is beyond the scope of words.
+% According to the Amanaska's transmission, jayatāṃ (and its similar readings) arose at a later time.  
+\end{translation}
+
+\begin{sources}[hp04_003]
+\emph{Amanaska} 2.22
+\begin{versinnote}
+\tl{ucchinnasarvasaṅkalpo niḥśeṣāśeṣaceṣṭitaḥ |\\+}
+\tl{svāvagamyo layaḥ ko 'pi jāyate vāgagocaraḥ ||\\+}
+\tl{\var{jāyate ] AllNI (except  Ma Pc), AllSI (except Ad Tha) Ea, jayatāṃ Ma Ad Nb Eb, jayatā Pc, jñāyatāṃ Tha, jagatāṃ AllN (except Nb Nu Ea Eb), japatāṃ Ca, layatāṃ Nu}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_003]
+\emph{Yogacintāmaṇi} f.~27v (attr.~to the \textit{Rājayoga})
+\begin{versinnote}
+\tl{ucchinnasarvasaṃkalpo nirgatāśeṣaceṣṭakaḥ |\\+}
+\tl{svāvagamyo layaḥ ko 'pi jayatāṃ vāgagocaraḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_003]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.4}
+\begin{translation}[hp04_004]
+Dissolution, which is eternal in the elements and senses, takes place where the gaze is.
+The gaze that has dissolved into the focus becomes the power of living beings.%%??JM: changed meditative focal point to focus (as tr.
+in v.6)
+%JB 'occurs' for 'takes place'
+\end{translation}
+
+\begin{sources}[hp04_004]
+\emph{Candrāvalokana} 8cd–9ab
+\begin{versinnote}
+\tl{yatra dṛṣṭir manas tatra bhūtendriyasanātanaṃ |\\+}
+\tl{sā śaktis sarvabhūtānāṃ dṛṣṭir lakṣye layaṃ gatā ||\\+}
+\tl{\var{sā śaktis ] 4340 4345 T00788, sa śāntis  7970}\\!}
+\end{versinnote}
+
+\emph{Kaulajñānanirṇaya} 3.2
+\begin{versinnote}
+\tl{yatra dṛṣṭir manas tatra bhūtendriya sapudgalaḥ \\+} 
+\tl{svaśaktijīvabhūtā hi dṛṣṭilakṣair layaṃ gatāḥ \\+}
+\tl{°bhūtā hi ] conj. Hatley, °bhūtāni ABCed\\!} 
+% ?? Consult Shaman's critical edition.
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_004]
+\emph{Yogacintāmaṇi} f.~24r
+\begin{versinnote}
+\tl{rājayoge—\\+}
+\tl{yatra dṛṣṭir layas tatra bhūtendriyasanātanī |\\+}
+\tl{syāc chaktiḥ sarvabhūtānāṃ dṛṣṭir lakṣeṇa saṃgatā ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~128r
+\begin{versinnote}
+\tl{tad uktaṃ haṭhapradīpikāyām– \\+}
+\tl{yatra dṛṣṭir layas tatra bhūtendriyasanātanī |\\+}
+\tl{syāc chaktiḥ sarvabhūtānāṃ dṛṣṭir lakṣeṇa saṃgatā ||\\+}
+\tl{\var{syāc chaktiḥ sarva°] B220, syārā śakti serva° 2244}\\+}
+\tl{\var{lakṣeṇa saṃgatā ] 2244, lakṣyakṣayaṃ gatā B220}\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 49.25
+\begin{versinnote}
+\tl{tathā coktaṃ haṭhapradīpikāyām –\\+}
+\tl{yatra dṛṣṭir layas tatra bhūtendriyasanātanaḥ |\\+}
+\tl{na sā śaktir bhāvabhūtānāṃ dṛṣṭe lakṣe kṣayaṃ gatā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_004]
+%Adopt dṛṣṭir lakṣye layaṃ gatā. [MD: Done.]
+%PYŚ: 2.18 on bhūtendriya and drś
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.5}
+\begin{translation}[hp04_005]
+The Vedas, Shastras and Puranas are like common prostitutes. \emph{Śāmbhavī mudrā} is unique and guarded like a woman of a good family.
+\end{translation}
+
+\begin{sources}[hp04_005]
+\emph{Amanaska} 2.9
+\begin{versinnote}
+\tl{vedaśāstrapurāṇāni sāmānyagaṇikā iva |\\+}
+\tl{ekaiva śāmbhavī mudrā guptā kulavadhūr iva ||\\!}
+%\tl{\var{vedaśāstrapurāṇāni ] AllNI (except Jb Mb) AllSI AllN (except Nb Nu), vedaśāstrapurāṇādyāḥ Mb, vedaśāstrāṇi sarvāṇi Jb: vedaśāstrapurāṇādi Nb Nu}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_005]
+\emph{Haṭharatnāvalī} 4.27
+\begin{versinnote}
+\tl{vedaśāstrapurāṇāni sāmānyagaṇikā iva |\\+}
+\tl{ekaiva śāmbhavī mudrā guptā kulavadhūr iva ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~24v (attr.~to the \textit{Rājayoga})
+\begin{versinnote}
+\tl{vedaśāstrapurāṇaughāḥ sāmānyagaṇikā iva |\\+}
+\tl{ekaiva śāmbhavī mudrā guptā kulavadhūr iva ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 49.26 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{vedaśāstrapurāṇāni sāmānyagaṇikā iva |\\+}
+\tl{ekaiva śāmbhavī mudrā sarvatantreṣu gopitā || 49.26\\!}
+\end{versinnote}
+
+%\emph{Vimalāgama} 73-005
+%\begin{versinnote}
+%\tl{skānde-\\+}
+%\tl{vedaśāstrapurāṇādyāḥ aṣṭā veśy[ā]ṅganā iva |\\+}
+%\tl{yā punaḥ śāṅkarī vidyā guptā kulavadhūr iva ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_005]
+See Birch 2013:286 for more parallel verses.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.6}
+\begin{translation}[hp04_006]
+The focus internal, the gaze external, unblinking: this is the \emph{śāmbhavī mudrā} concealed in all the Tantras.%??JM: tr. tweaked
+% JB my preference would be: The focus is internal [yet] the gaze is external and free from blinking.  
+\end{translation}
+
+\begin{sources}[hp04_006]
+\emph{Amanaska} 2.10
+\begin{versinnote}
+\tl{antar lakṣyaṃ bahir dṛṣṭir nimeṣonmeṣavarjitā |\\+}
+\tl{eṣā hi śāmbhavī mudrā sarvatantreṣu gopitā ||\\+}
+\tl{\var{antar lakṣyaṃ ] AllNI (except Ja K Pb Tr Ua Vb) Tha Vd, antarlakṣya Ad Cc K: antarlakṣaṃ Tr Nw Ve: antarlakṣā Ja Ua Ea Eb, antarlakṣo Na Nq Nu, antarlakṣyo Nb Nm: antaryogaṃ Vb: antarlīno Pb}\\!}%??
+JM: delete variants?
+% JB: the lakṣyaṃ/lakṣa spelling is an issue in both the Amanaska and HP mss (inlcuding alpha).
+\end{versinnote}
+
+\emph{Candrāvalokana} 1
+\begin{versinnote}
+\tl{antar lakṣyaṃ bahir dṛṣṭir nimeṣonmeṣavarjitaḥ |\\+}
+\tl{iyaṃ sā śāṃbhavī mudrā sarvatantreṣu gopitā |\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_006]
+\emph{Yogacintāmaṇi} f.~24v (attr.~to the \textit{Rājayoga})
+\begin{versinnote}
+\tl{antar lakṣyaṃ bahir dṛṣṭir nimeṣonmeṣavarjitā |\\+}
+\tl{eṣā tu śāmbhavī mudrā sarvaśāstreṣu gopitā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_006]
+See Birch 2013: 287 for more parallel verses.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.7}
+\begin{translation}[hp04_007]
+When the yogi's mind and breath have dissolved in the internal focus while he is looking outwards and down with a gaze in which the pupils are unmoving, even though he's not looking [at anything], this indeed is \emph{khecarīmudrā}.
+O guru, it manifests because of your favour and is that reality which is Śambhu's state, free from what is void and not void.
+\end{translation}
+
+\begin{sources}[hp04_007]
+\emph{Candrāvalokana} 2
+\begin{versinnote}
+\tl{antarlakṣyavilīnacittapavano yogī yadā vartate \\+}
+\tl{dṛṣṭyā niścalatārayā bahir adhaḥ paśyan na paśyet sadā |\\+}
+\tl{mudreyaṃ khalu khecarī bhavati sā yuktaprasādāt guroḥ\\+}
+\tl{śūnyāśūnyavivarjitaṃ sphurati yat tattvaṃ padaṃ śāṃbhavaṃ ||\\+}
+\tl{\var{°pavano ] 4340 4345 T00788, °pavane 7970}\\+}
+\tl{\var{sā yukta°] 4340 4345, sā yuktā  T00788, satyuṣṭa° 7970}\\+}
+\tl{\var{prasādād guroḥ ] 4340 4345 T00788, prasādāt guro 7970}\\+}
+\tl{\var{yat tattvaṃ ] 7970 4340 4345, cet tatvaṃ T00788}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_007]
+\emph{Yogacintāmaṇi} f.~24v
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{antarlakṣyavilīnacittapavano yogī yadā vartate\\+}
+\tl{dṛṣṭyā niścalatārayā bahir asau paśyann apaśyann api |\\+}
+\tl{mudreyaṃ khalu śāṃbhavī bhavati sā yuṣmatprasādād guro\\+}
+\tl{śūnyāśūnyavivarjitaṃ sphurati yat tatvaṃ padaṃ śāṃbhavam ||\\!}
+\end{versinnote}
+
+\emph{Anubhavanivedana} 1
+\begin{versinnote}
+\tl{antarlakṣyavilīnacittapavano yogī yadā vartate\\+}
+\tl{dṛṣṭyā niścalatārayā bahir asau paśyann apaśyann api |\\+}
+\tl{mudreyaṃ khalu śāmbhavī bhavati sā yuṣmatprasādād guro\\+}
+\tl{śūnyāśūnyavivarjitaṃ bhavati yat tattvaṃ padaṃ śāmbhavam ||1 ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 49.27 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{antarlakṣavilīnacittapavano yogī yadā varttate\\+}
+\tl{dṛṣṭyā niścalatārayā bahir adhaḥ paśyann apaśyann api |\\+}
+\tl{mudreyaṃ khalu śāmbhavī bhavati sā yuṣmatprasādāt guroḥ\\+}
+\tl{śūnyāśūnyavivarjitaṃ sphurati yat tatvaṃ padaṃ śāmbhavam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_007]
+On the similarity of \emph{śāmbhavī} and \emph{khecarī mudrās} in 4.5–7, see the note on 4.8. 
+\end{philcomm}
+
+\begin{metre}[hp04_007]
+Śārdūlavikrīḍita 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.8}
+\begin{translation}[hp04_008]
+There is no difference between the states of \emph{śāmbhavī} and \emph{khecarī}.
+\end{translation}
+% Adopt avasthāyām abhedatā (N10, N12, T11) (Cf.~avasthāyām abhedataḥ C6)
+%
+% śrīśāmbhavyāś ca khecaryā avasthāyām abhedataḥ |M1,J10
+% bhavec cittalayākārā unmanī citsvarūpiṇī | J10
+% The beyond-mind [state], whose form is the dissolution of mind and own nature is consciousness, is identical in the state of Śāmbhavī and Khecarī.
+
+%\begin{sources}[hp04_008]
+%\end{sources}
+
+\begin{testimonia}[hp04_008]
+\emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{śrīśāṃbhavyā khecaryāś ca avasthāṃ ca labhed ataḥ |\\+}
+\tl{tāre jyotiṣi saṃyojya kiñ cid uccālayed bhruvau ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~128r–128v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tāre jyotiṣi saṃyojya kiṃ cid unnamayed bhuvau [|]\\+}
+\tl{śrīśāṃbhavyāś ca khecaryā avasthā[ṃ] tu labhed ataḥ ||\\+}
+\tl{\var{unnamayed ] 2244, saṃcālayed B220}\\+}
+\tl{\var{unmanīkaraṇaṃ kṣaṇāt ] 2244, unmanīkāralakṣaṇam B220}\\+}
+\tl{\var{tu labhed ataḥ ] 2244, labhate daśa B220}\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp04_008]
+This line may be authorial because it is explaining the fact that 4.6 and 4.7 teach similar techniques called \emph{śāmbhavī mudrā} and \emph{khecarīmudrā} respectively.
+One would expect a verse on \emph{khecarī mudrā} to be explaining the insertion of the tongue in the nasopharyngeal cavity (cf.~3.33–48), but the practice of meditating by fixing the gaze is called \emph{khecarī mudrā} in 4.7 (as attested by \textalpha, \textbeta\ and \texteta\ manuscripts) and manuscripts of the \emph{Candrāvalokana}, the text from which Svātmārāma borrowed this verse.
+It is therefore likely that he added 4.8 to explain that \emph{śāmbhavī} and \emph{khecarī mudrās} are the same with respect to the gaze.
+
+%Judit would greyscale 4.8, thinking that it is a marginal note. And greyscale the pātāle verses.
+%avasthā: Brahmānanda explains this as referring to the place (dhāman) of khecarī (the middle of the brow) and śāmbhavī (the heart).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*1}
+\begin{translation}[hp04_008_1]
+[The yogi] should fix the pupils on a light and slightly raise the eyebrows. This is the way of the preliminary yoga, which immediately brings about the beyond-mind state.
+\end{translation}
+% unmanīkārakaḥ ?
+
+\begin{sources}[hp04_008_1]
+\emph{Amanaska} 1.8 (South Indian Recension)
+\begin{versinnote}
+\tl{netre jyotiṣi saṃyojya kiṃ cid unnamayed bhruvau |\\+}
+\tl{pūrvayogasya mārgo 'yam unmanīkārakaḥ kṣaṇāt ||\\+}
+\tl{\var{unnamayed ] Cb Vd: unnamayan Bb Tha: unmīlaye W: unmilayet Uc: unmanaya Pe: unmīlya Cc}\\!}
+%\tl{\var{unmanīkārakaḥ kṣaṇāt ] Bb Tha W Vd Pe: unmunikārakakṣaṇāt Uc: unmanīkāraṇaṃ kṣaṇāt Cc: unmanākāśakaḥ kṣaṇāt Ad Ca: unmanākāśakakarṣaṇāt Cb}\\!}%??
+variants needed?
+JB I've deleted the variants for the last pāda but we need those on unnamayed (see below)
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_008_1]
+\emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tāre jyotiṣi saṃyojya kiṃ cid uccālayed bhruvau |\\+}
+\tl{pūrvayogasya mārgo 'yam unmanīkaraṇaḥ kṣaṇāt ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} ff. 128r-128v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tāre jyotiṣi saṃyojya kiṃ cid unnamayed bhuvau |\\+}
+\tl{pūrvayogasya mārgo [']yam unmanīkaraṇaṃ kṣaṇāt ||\\!}
+\end{versinnote}
+
+%Rājayogāmṛta 3.7c-d $-$ 3.8a-b
+%\begin{versinnote}
+%\tl{tārāj jyotiṣi saṃyojya kiṃcid unmīlayed bhruvau |\\+}
+%\tl{pūrvayogasya mārgo 'yaṃ unmanitārakaṃ kṣaṇāt ||\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_1]
+%MD: Better to adopt unnamayed and kārakaḥ kṣaṇāt?
+% ??
+JB Since unnamayed in the beta witnesses and source, and makes better sense, I think we should adopt.
+The dictionary states that uc-cal means 'to go or move away from; to free or loosen one's self from', which is not so good here.
+
+This and the next verse (4.8*1–2) are from the South Indian recension of the \emph{Amanaska}, a later rewriting of that work which incorporates these verses in an additional passage on Tārakayoga, added here to elaborate on the meditation of fixing the gaze that is taught in 4.4–7.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*2}
+\begin{translation}[hp04_008_2]
+Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning.
+They do not know what causes one to cross over.%??JM: put Tāraka and explain in note?
+Tārakayoga is mentioned in the previous note.
+% JB in the source 
+\end{translation}
+% Maybe grey scale
+
+\begin{sources}[hp04_008_2]
+\emph{Amanaska} 1.11 (South Indian Recension)
+\begin{versinnote}
+\tl{ke cid āgamajālena ke cin nigamasaṅkulaiḥ |\\+}
+\tl{ke cit tarkeṇa muhyanti naiva jānanti tārakam ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_008_2]
+\emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ke cid āgamajālena ke cin niyamasaṃkulāḥ |\\+}
+\tl{ke cit tarkeṇa muhyanti naiva jānanti tārakam ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 49.29 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ke cid āgamajālena ke cin nigamasaṃkule |\\+}
+\tl{ke cit tarkeṇa muhyanti naiva jānanti tārakam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_2]
+On why this verse is in greyscale, see the note to 4.8*1.
+% ??JB we should adopt nigama in pāda b because it is attested by several manuscript groups and the source.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*3}
+\begin{translation}[hp04_008_3]
+With eyes half open, mind steady, gaze placed at the tip of the nose and inner state motionless, [the yogi] should lead the moon and sun to dissolution.
+He attains the state of the highest reality, the supreme essence, whose form is light and which is devoid of anything external and shining intensely.
+What more is to be said here?
+\end{translation}
+
+\begin{sources}[hp04_008_3]
+\emph{Candrāvalokana} 3
+\begin{versinnote}
+\tl{ardhotghāṭitalocana[ḥ] sthiramanā nāsāgradattekṣaṇaś\\+}
+\tl{candrārkāv api līnatām upagatau niṣpandarūpaṃ vapuḥ |\\+}
+\tl{jyotīrūpam aśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ\\+}
+\tl{tatvaṃ tatpadam eti yat tu paramaṃ vācyaṃ kim atrādhikaṃ ||\\+}
+\tl{\var{sthiramanā ] 4340 4345 T00788, sphuramanā 7970}\\+}
+\tl{\var{nāsāgradattekṣaṇaś ] 7970 4345 T00788, nāsāgradaṭhattakṣaṇaś 4340}\\+}
+\tl{\var{tatvaṃ tatpadam eti yat tu] 4340 4345 T00788, tatvaṃ-n-tatpadam eti yat kṛ 7970}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_008_3]
+\emph{Yogacintāmaṇi} ff. 24v–25r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ardhodghāṭitalocanaḥ sthiramanā nāsāgradattekṣaṇaḥ\\+}
+\tl{candrārkāv api līnatām upanayen niḥspandabhāvottare |\\+}
+\tl{jyotīrūpam aśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ  \\+}
+\tl{tatvaṃ tat padam eti vastu paramaṃ vācyaṃ kim atrādhikam ||\\!}
+\end{versinnote}
+
+\emph{Anubhavanivedana} 2
+\begin{versinnote}
+\tl{ardhodghāṭitalocanaḥ sthiramanā nāsāgradattekṣaṇaś\\+}
+\tl{candrārkāv api līnatām upagatau trispandabhāvāntare |\\+}
+\tl{jyotīrūpam aśeṣabāhyarahitaṃ caikaṃ pumāṃsaṃ param\\+}
+\tl{tattvaṃ tatpadam eti vastu paramaṃ vācyaṃ kim atrādhikam ||2 ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 49.30 (citing the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ardhodghāṭitalocanaḥ sthiramanā nāsāgradattekṣaṇaḥ\\+}
+\tl{candrārkāvapi līnatāmupanayennispandavācyaṃ tataḥ |\\+}
+\tl{jyotīrūpaviśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ\\+}
+\tl{tatvaṃ tatparamasti vastu paramaṃ vācyaṃ kimatrādhikam || iti ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_3]
+This verse is not in the \textalpha\ manuscripts and was probably added because it elaborates on the gaze.
+In the \emph{Candrāvalokana} (i.e., the source text) and the \emph{Anubhavanivedana}, it follows 4.7, which may explain its position in the \textgamma\ and \textdelta\ groups.
+\end{philcomm}
+% ?? MD: I have changed the verse order according to the epsilon group. Please adjust the commentary. JB Done
+
+\begin{metre}[hp04_008_3]
+Śārdūlavikrīḍita 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.8*4}
+%\begin{translation}[hp04_008_4]
+%That which enters the aperture into the underworld, which exists at the base of Meru,  the wise yogin say that is the truth, the source of [all] rivers.
+The essence of the body flows from the moon.
+Because of that, death arises for people.
+[This] good mudrā should block that [essence].
+Otherwise, there is no bodily perfection.
+%\end{translation}
+% See 3.49
+
+%\begin{sources}[hp04_008_4]
+%\end{sources}
+
+% \begin{testimonia}[hp04_008_4]
+% \emph{Haṭharatnāvalī} 4.30
+% \begin{versinnote}
+% \tl{pātāle yad viśati suṣiraṃ merūmūlaṃ tad asti \\+}
+% \tl{tattvaṃ caitad vadati sudhā tanmukhaṃ nimnagānām |\\+}
+% \tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ\\+}
+% \tl{tad badhnīyāt sukharatimṛdur nānyathā kāryasiddhiḥ ||\\!}
+% \end{versinnote}
+
+% \emph{Yogacintāmaṇi} f.~74v (attr.~to the \emph{Haṭhapradīpikā})
+% \begin{versinnote}
+% \tl{pātāle yad viyati śikhare merumūle tad asti\\+}
+% \tl{tattvaṃ caitat pravadati sudhīs sammukhe nimnagānām |\\+}
+% \tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ\\+}
+% \tl{taṃ badhnīyāt svakaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
+% \end{versinnote}
+
+% \emph{Yuktabhavadeva} 7.220 (attr.~to Gorakṣanātha)
+% \begin{versinnote}
+% \tl{pātāle yad viśati suṣiraṃ merumūle yad asti \\+}
+% \tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
+% \tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
+% \tl{badhnīyāttat sukaraṇam atho nānyathā kāyasiddhiḥ ||\\!}
+% \end{versinnote}
+
+% \emph{Haṭhatattvakaumudī} 14.26 (attr.~to the \emph{Haṭhapradīpikā})
+% \begin{versinnote}
+% \tl{pātāle yadvitatasuṣiraṃ merumūle tad asmin\\+}
+% \tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
+% \tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
+% \tl{taṃ badhnīyāt sukaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
+% \end{versinnote}
+% \end{testimonia}
+
+\begin{philcomm}[hp04_008_4]
+For the translation and testimonia, as well as an explanation of the various places and versions of this verse in the text, see 3.49.
+\end{philcomm}
+
+%\begin{metre}[hp04_008_4]
+%Mandākrāntā 
+%\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.8*5}
+\begin{philcomm}[hp04_008_5]
+See 3.75, where this verse is also found.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*6}
+\begin{philcomm}[hp04_008_6]
+See 3.76, where this verse is also found.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*7}
+\begin{translation}[hp04_008_7]
+[The yogi] should not worship the \emph{liṅga} by day, nor should he worship it by night. He should worship the \emph{liṅga} constantly, by suppressing day and night.
+\end{translation}
+
+\begin{sources}[hp04_008_7]
+\emph{Khecarīvidyā} 3.19
+\begin{versinnote}
+\tl{na divā pūjayel liṅgaṃ na rātrau ca maheśvari |\\+}
+\tl{sarvadā pūjayel liṅgaṃ divārātrinirodhataḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_008_7]
+\emph{Haṭhasaṅketacandrikā} f.~128v (attr.~Haṭhapradīpikā)
+\begin{versinnote}
+\tl{divā na pūjayel liṃga[ṃ] rātrau naiva prapūjayet [|]\\+}
+\tl{satataṃ pūjayel liṃga[ṃ] divārātrau ca pūjayet [||]\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_7]
+%?? JM: nirodhataḥ reading much better. Early corruption?
+% JB I think we should note nirodhataḥ as the better reading in the comments but adopt divārātrau (or divārātraṃ) na pūjayet (he should not worship at day and night (i.e., when prāṇa is moving in the sun and mmoon channels)).
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*8}
+\begin{translation}[hp04_008_8]
+Now \emph{khecarī}:\\
+There is a hollow that generates knowledge and has five streams. \emph{Khecarīmudrā} is situated in that pure void.
+\end{translation}
+% ?? compound khecarīmudrā in 4.8*8c and 4.8*9c?
+
+%\begin{sources}[hp04_008_8]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_8]
+\emph{Haṭhasaṅketacandrikā} f.~128v
+\begin{versinnote}
+\tl{suṣiraṃ jñānajanakaṃ paṃcastr[o]taḥsamanvitaṃ |\\+}
+\tl{tiṣṭhate khecarī mudrā tasmin [ś]ūnye niraṃjane  ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_8]
+This verse may not be referring to the cavity in which the tongue is placed but perhaps to a more esoteric sense based on the meaning of \textit{pañcasrotas} as the five streams of tantric Śaiva teachings, which in this case generate knowledge.
+On \textit{pañcasrotas}, see \emph{Tāntrikābhidhānakośa} 2013, vol.
+3: 361.
+\end{philcomm}
+
+\begin{metre}[hp04_008_8]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.8*9}
+\begin{translation}[hp04_008_9]
+The breath in the left and right channels moves into the middle. Without doubt, \emph{khecarīmudrā} abides in that place.
+\end{translation}
+
+%\begin{sources}[hp04_008_9]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_9]
+\emph{Haṭhasaṅketacandrikā} f.~??
+\begin{versinnote}
+\tl{savyadakṣiṇanāḍīstho madhye carati mārutaḥ |\\+}
+\tl{tiṣṭhati khecarī mudrā tasmin sthāne na saṃśayaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_9]
+%The term \emph{madhye} appears to be referring to the central channel here. 
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*10}
+%\begin{translation}[hp04_008_10]
+%The mind moves in the void because the tongue moves in the void, which is why that \emph{mudrā} is worshipped by the adepts with the name “she who moves in the void” (\emph{khecarī}).
+%\end{translation}
+
+% \begin{sources}[hp04_008_10]
+% \emph{Vivekamārtaṇḍa} 50
+% \begin{versinnote}
+% \tl{cittaṃ carati khe yasmāj jihvā carati khe gatā |\\+}
+% \tl{[tenai]ṣā khecarī nāma mudrā siddhair namaskṛtā ||\\!}
+% \end{versinnote}
+
+% %\emph{Skandapurāṇa} 4.41.45%??JM: is it worth putting this in? Verse is prob found all over the place and this isn't the old SP. Just YCM will do.
+% %\begin{versinnote}
+% %\tl{cittaṃ carati khe yasmā jihvā carati khe gatā |\\+}
+% %\tl{tenaiṣā khecarī nāma mudrā siddhair niṣevitā। ||\\!}
+% %\end{versinnote}
+% \end{sources}
+
+% \begin{testimonia}[hp04_008_10]
+% \emph{Yogacintāmaṇi} ?? [attrib. Skandapurāṇa]
+% %\begin{versinnote}
+% %\tl{cittaṃ carati khe yasmāj jihvā carati khe gatā |\\+}
+% %\tl{tenaiṣā khecarī nāma mudrā siddhair niṣevitā ||\\!}
+% %\end{versinnote}
+
+% \emph{Yuktabhavadeva} 212
+
+% \end{testimonia}
+
+\begin{philcomm}[hp04_008_10]
+See 3.37, where this verse is also found.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*11}
+\begin{translation}[hp04_008_11]
+At the juncture of Iḍā and Piṅgalā, the void devours the breath. \emph{Khecarīmudrā} abides there. This is true, time and time again.
+\end{translation}
+
+%\begin{sources}[hp04_008_11]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_11]
+\emph{Upāsanāsārasaṅgraha} (IFP Transcript T1095) p.~42.
+\begin{versinnote}
+\tl{iḍāpiṅgalayor yoge śūnye caivānilaṃ graset |\\+}
+\tl{tiṣṭhate khecarī mudrā tatra satyaṃ punaḥ punaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_11]
+Cf.~Rāghavabhaṭṭa \emph{ad Śāradātilaka} 25.43: \emph{suṣumṇāyām eteṣu parvasu iḍāpiṅgalayor yogo bhavatīti jñeyam}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*12}
+\begin{translation}[hp04_008_12]
+The \emph{mudrā} situated in the cakra of the void (\emph{vyomacakre}) in the middle of the sun and moon on an unsupported surface is [the \emph{mudrā}] called \emph{khecarī}.
+\end{translation}
+
+%\begin{sources}[hp04_008_12]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_12]
+\emph{Upāsanāsārasaṅgraha} (IFP Transcript T1095) p.~41.
+\begin{versinnote}
+\tl{somasūryadvayor madhye nirālambe tale punaḥ |\\+}
+\tl{saṃsthitā vyomacakre sā mudrā nāma ca khecarī ||\\!}
+\end{versinnote}
+
+\emph{Gorakṣasiddhāntasaṅgraha} p.~37
+\begin{versinnote}
+\tl{sūryācandramasormadhye nirālambe'nile punaḥ |\\+}
+\tl{saṃsthitā vyomacakre yā sā mudrā nāma khecarī ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_12]
+The reading \emph{tale} is suspect, as is \emph{nirālambāntare}, the conjecture of Brahmānanda.
+%  bhruvormadhye dvyasraṃ vyomacakraṃ tad dhi cinmayasya jyotiṣa upalabdhisthānam  Commentary on LYV 9.82.
+% Haṭhasaṅketacandrika (4192): śūnyaṃ vyomacakravyavasthitam | iḍāpiṃgalāṃtargataṃ brahmaraṃdhrākhyaṃ vyomacakraṃ [...]. 
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*13}
+\begin{translation}[hp04_008_13]
+Brought forth by me, lovely, and clearly beloved of Śiva, the Suṣumṇā should fill herself with the divine air through her rear mouth.
+\end{translation}
+
+%\begin{sources}[hp04_008_13]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_13]
+\emph{Yogasārasaṅgraha} p. 61 (attrib. Praṇavacintāmaṇi)
+\begin{versinnote}
+\tl{sā mayā viditā yā māyā sākṣācchivavallabhā |\\+}
+\tl{pūrayen mārutaṃ divyaṃ suṣumnā paścime mukhe ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_13]
+The referent of \emph{mayā} is unspecified and no source text has been identified.
+It most probably refers to a goddess.%to understand here because the referent of \emph{mayā} is not known.
+If it refers to a yogi, then it might mean that \emph{suṣumnā} has been pierced.
+Otherwise, if \emph{udbheditā} means `brought forth', one would assume the speaker to be a god or goddess.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*14}
+\begin{translation}[hp04_008_14]
+And if [she] fills herself from the front, \emph{khecarī} definitely arises. [The yogi] should practise \emph{khecarīmudrā}. The state beyond mind arises.
+\end{translation}
+
+%\begin{sources}[hp04_008_14]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_14]
+\emph{Upāsanāsārasaṅgraha} p.~135
+purastāc caiva pūryeta niścitā khecarī bhavet |
+abhyaset khecarīmudrāmunmanī saṃprajāyate ||
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_14]
+We have understood \emph{niścitā} as an adverb. No witnesses have \emph{niścitam}, but we see no other way of taking \emph{niścitā}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*15}
+\begin{translation}[hp04_008_15]
+[The yogi] should practise \emph{khecarīmudrā} until he falls into a yogic sleep. For one who has attained yogic sleep, death never arises.
+\end{translation}
+
+%\begin{sources}[hp04_008_15]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_15]
+\emph{Upāsanāsārasaṅgraha} p.~135\\
+abhyaset khecarīṃ tāvat yāvat syād yoganidrataḥ |
+saṃprāptayoganidrasya kālo nāsti kadācana ||
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_15]
+\end{philcomm}
+
+%%%%%%%%%%
+
+\subsection*{4.8*16}
+\begin{translation}[hp04_008_16]
+Between the eyebrows is the place of Śiva. The mind dissolves there. That level should be known as the fourth state. Death does not exist there.
+\end{translation}
+%?? JB: pada here means place/site (level sounds strange)
+% abode of Śiva ... that place should be known
+
+%\begin{sources}[hp04_008_16]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_16]
+\emph{Upāsanāsārasaṅgraha} p.~135
+bhruvor madhye śivasthānaṃ manas tatra vilīyate |
+jñātavyaṃ tatparaṃ turyaṃ tatra kālo na vidyate ||
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_16]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*17}
+\begin{translation}[hp04_008_17]
+Between the moon and the sun, [the yogi] should apply \emph{khecarīmudrā}, which is situated in the supportless, great void, the \emph{vyomacakra}.
+\end{translation}
+
+\begin{sources}[hp04_008_17]
+\emph{Jñānasāra} 3.3cd
+\begin{versinnote}
+\tl{candrasūryadvayor madhye muḍādadyā? tu khecarīm |\\!}
+\end{versinnote}
+
+\end{sources}
+
+%\begin{testimonia}[hp04_008_17]
+%\end{testimonia}
+
+\begin{philcomm}[hp04_008_17]
+The \emph{vyomacakra} ('the cakra of space') is also mentioned in a half-verse that was added to some later recensions of the \emph{Haṭhapradīpikā} (see 3.35) and states that \emph{vyomacakra} is another name for \emph{khecarīmudrā}.
+In \emph{Jyotsnā} 4.45, Brahmānanda states that the \emph{vyomacakra} is associated with all the voids in the middle of the brow (\emph{bhrūmadhye sarvakhānāṃ samanvayāt}) and, in the \emph{Haṭhasaṅketacandrikā} (f.
+129v), Sundaradeva says that it is called the Brahmarandhra, which is between Iḍā and Piṅgalā (\emph{iḍāpiṅgalāntargataṃ brahmarandhrākhyaṃ
+vyomacakraṃ tat khecarīmudrāṃ śaktiṃ kuryād ...}). 
+%?? Write a note about epsilon 3 vyomacakram vyavasthitaṃ: Cf.~the greyscale verse in 3.35.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*18}
+\begin{translation}[hp04_008_18]
+[The yogi] should make the mind supportless and think of nothing at all. He assuredly remains like a pot in the ether, [empty] inside and outside.
+\end{translation}
+
+\begin{sources}[hp04_008_18]
+\emph{Jñānasāra} 3.4
+\begin{versinnote}
+\tl{nirālambaṃ manaḥ kṛtvā na kiñcid api cintayet || 3 ||\\+}
+\tl{sa bāhyābhyantare yogī ghaṭavat tiṣṭhate priye |\\!}
+\end{versinnote}
+
+% \emph{Śivasaṃhitā} 5.210
+% \begin{versinnote}
+% \tl{nirālambo bhavej jīvo jñātvā vedāntayuktitaḥ |\\+}
+% \tl{nirālambaṃ manaḥ kṛtvā na kiṃ cic cintayet sudhīḥ ||\\!}
+% \end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_008_18]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_008_18]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*19}
+\begin{translation}[hp04_008_19]
+When the external air has dissolved into the void, the breath is sure to go to its place †with the mind on the side of the sun†.
+%When the external air has dissolved into the void, the breath surely goes to its own place, † and the sun together with the mind to the fire.†
+\end{translation}
+
+%\begin{sources}[hp04_008_19]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_19]
+\emph{Haṭhapradīpikā} (with ten-chapters) 7.52 
+\begin{versinnote}
+\tl{bāhyavāyur yathā līnaḥ svasya madhye na saṃśayaḥ |\\+}
+\tl{svasthānaṃ gacchati prāṇaḥ sūryo 'gnau pavane tathā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_19]
+% sūryāgnir manasā tathā M1
+It is hard to make sense of \emph{sūryāṅge} here (`on the side of the sun' ?).
+The terms \emph{sūryāṅge} and \emph{candrāṅge} occur in the third chapter (3.15) in the sense of the right and left sides of the body, respectively.
+However, this meaning does not seem relevant here.
+Since we are not sure of the meaning of \emph{sūryāṅge} and since the variant readings with \emph{pavana} are not clear either, we have cruxed the fourth verse quarter.
+
+%\textit{Maitrī Upaniṣad} 6.1-2 might provide a clue to the meaning of \emph{sūryāṅga}.
+This Upaniṣad explains that the \textit{ātman} manifests itself in two ways: as \textit{prāṇa} internally and as the sun externally.
+\textit{Prāṇa} is identified as the inner aspect of the \textit{ātman}.
+It is associated with breath, etc.
+The sun is depicted as the external manifestation of the \textit{ātman}.
+It symbolizes the cosmic source of energy, etc.
+The text emphasizes that the movements of \textit{prāṇa} and the sun are synchronized and states that the golden \textit{puruṣa}, who dwells in the sun, also resides in the heart as \textit{prāṇa}.
+If one is willing to accept a connection between this verse and the concept mentioned in the \textit{Maitrī Upaniṣad}, then \textit{sūryāṅge} possibly has even two layers of meaning.
+On the microcosmic level, \textit{prāṇa} would return home to the heart.
+Moreover, on the macrocosmic level, the \textit{ātman} would return to its macrocosmic home - the sun.
+% ?? Does sūryāṅge occur in the Maitrī Upaniṣad? If not, the connection with this Upaniṣad seems tenuous. 
+\end{philcomm}
+
+%?? MD: yadā (epsilon) or yathā (beta)?
+% JB separate yadā līnaḥ in the edition (yadā seems easier to understand than yathā)
+
+%%%%%%%%%%
+\subsection*{4.8*20}
+\begin{translation}[hp04_008_20]
+For [the yogi] practising in this way day and night on the path of the breath, as a result of the practice the breath is consumed, [and] the mind dissolves into it.
+\end{translation}
+
+%\begin{sources}[hp04_008_20]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_20]
+\emph{Haṭhasaṅketacandrikā} f. 129v
+\begin{versinnote}
+\tl{abhyāsāl līyate vāyuḥ manas tatra vilīyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+%??JM: I can't find these HSC refs
+
+\begin{philcomm}[hp04_008_20]
+Brahmānanda identifies the path of the breath (\emph{vāyumārga}) with Suṣumṇā.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*21}
+\begin{translation}[hp04_008_21]
+[The yogi] should flood the body with nectar from the soles of the feet to the head. He is sure to become perfected, with a great body and great strength and valour.\\Thus ends \emph{khecarī}.
+\end{translation}
+
+%\begin{sources}[hp04_008_21]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_21]
+\emph{Haṭhasaṅketacandrikā} f. 129v
+\begin{versinnote}
+\tl{siddhaty evaṃ tadā kāyo mahābalaparākramaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+%??JM: can't find. ref?
+
+%\begin{philcomm}[hp04_008_21]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*22}
+\begin{translation}[hp04_008_22]
+Now \emph{śāmbhavī}:\\
+{[}The yogi] should [put] the mind in Śakti and Śakti in the mind, observe the mind with the mind, and meditate on it as the supreme state.
+\end{translation}
+%V3 is closer to the CA reading
+
+\begin{sources}[hp04_008_22]
+\emph{Candrāvalokana} 27
+\begin{versinnote}
+\tl{śaktimadhye manaḥ kṛtvā manaś śaktes tu madhyamam |\\+}
+\tl{manasā mana ālokya taṃ dhyāyet paramaṃ padaṃ ||\\+}
+\tl{\var{madhyamam ] madhyagaṃ 4340}\\+}
+\tl{\var{taṃ dhyāyet ] tad dhyāyet 4340}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_008_22]
+\emph{Haṭhasaṅketacandrikā} f.~129v–130r
+\begin{versinnote}
+\tl{śaktimadhyo manaḥ kṛtvā śaktiṃ ca svāntamadhyagāṃ |\\+}
+\tl{manasā mana ālokya tad dhyāyet paramaṃ padam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_22]
+In the second verse quarter, \emph{mana} for \emph{mano} is for the metre.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*23}
+\begin{translation}[hp04_008_23]
+Put the self in space and put space in the self. [The yogi] should make everything consist of space and think of nothing at all.
+\end{translation}
+
+\begin{sources}[hp04_008_23]
+\emph{Uttaragītā} 1.9%??JM: bibliog for texts cited — where?
+\begin{versinnote}
+\tl{khamadhye kuru cātmānam ātmamadhye ca khaṃ kuru |\\+}
+\tl{ātmānaṃ khamayaṃ kṛtvā na kiṃ cid api cintayet  ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_008_23]
+\emph{Haṭharatnāvalī} 4.45
+\begin{versinnote}
+\tl{khamadhye kuru cātmānam ātmamadhye ca khaṃ kuru |\\+}
+\tl{sarvaṃ ca khamayaṃ kṛtvā na kiṃ cid api cintayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_23]
+%?? Pāda c: sarvaṃ or ātmānaṃ?
+%\end{philcomm}
+% J up to here! 11 March
+
+%%%%%%%%%%
+\subsection*{4.8*24}
+\begin{translation}[hp04_008_24]
+Like an empty pot in air, [the yogī] is empty on the inside and empty on the outside.
+Like a full pot in the ocean, [the yogi] is full on the inside and full on the outside.%??JM: I changed the sky to air.
+Not sure, but I don't think we are meant to think of a pot floating in space.
+\end{translation}
+
+\begin{sources}[hp04_008_24]
+\emph{Laghuyogavāsiṣṭha} 6.15.79 (\emph{Mokṣopāya} 6.155.25)
+\begin{versinnote}
+\tl{antaḥśūnyo bahiḥśūnyaḥ śūnyakumbha ivāmbare |\\+}
+\tl{antaḥpūrṇo bahiḥpūrṇaḥ pūrṇakumbha ivārṇave ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_008_24]
+\emph{Yogacintāmaṇi} (attrib.~Vasiṣṭha, = \emph{Laghuyogavāsiṣṭha} 15.79)
+\begin{versinnote}
+\tl{antaḥśūnyo bahiḥśūnyaḥ śūnyakumbha ivāmbare |\\+}
+\tl{antaḥpūrṇo bahiḥpūrṇaḥ pūrnakumbha ivārṇave ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.46
+\begin{versinnote}
+\tl{antaḥpūrṇo bahiḥpūrṇaḥ pūrṇakumbha ivāmbhasi |\\+}
+\tl{antaḥśūnyaṃ bahiḥśūnyaṃ śūnyakumbha ivāmbare ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_24]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*25}
+\begin{translation}[hp04_008_25]
+Do not think about the external or internal. [The yogi] should abandon all thought and think of nothing at all.
+\end{translation}
+
+%\begin{sources}[hp04_008_25]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_25]
+\emph{Haṃsavilāsa} p.~48
+\begin{versinnote}
+\tl{bāhyacintā na kartavyā tathaivāntaracintanam |\\+}
+\tl{sarvacintāṃ parityajya na kiñcid api cintayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_25]
+%°cintanā is not attested elsewhere.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*26}
+\begin{translation}[hp04_008_26]
+The whole world is but a construct of mere ideation.
+A construct of mere ideation is an affectation of the mind.
+Jettison thought that is only ideation, take refuge in a resolve that is free of ideation, and obtain peace, o Rāma.
+\end{translation}
+
+\begin{sources}[hp04_008_26]
+\emph{Laghuyogavāsiṣṭha} 7.27
+\begin{versinnote}
+\tl{saṃkalpajālakalanaiva jagat samagraṃ\\+}
+\tl{saṃkalpajālakalanāt tu manovilāsaḥ |\\+}
+\tl{saṃkalpamātram alam utsṛja nirvikalpam\\+}
+\tl{āśritya niścayam avāpnuhi rāma śāntim ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_008_26]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_008_26]
+%\end{philcomm}
+
+\begin{metre}[hp04_008_26]
+Vasantatilakā
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.8*27}
+\begin{translation}[hp04_008_27]
+Just as camphor in fire and salt in water, so the mind, when being joined with the highest reality, dissolves.
+\end{translation}
+
+%\begin{sources}[hp04_008_27]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_27]
+\emph{Haṭharatnāvalī} 4.43
+\begin{versinnote}
+\tl{karpūram anale yadvat saindhavaṃ salile yathā |\\+}
+\tl{tathā sandhīyamānaṃ hi manas tatraiva līyate||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_27]
+%?? pāda c: hi better. Not in HP sources only HR. Adopt?
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*28}
+\begin{translation}[hp04_008_28]
+Mind is said to be all that is to be known, [all] that has been perceived, and [all] knowledge of that. Knowledge and what is to be known are destroyed together. There is no other path.
+\end{translation}
+
+%\begin{sources}[hp04_008_28]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_28]
+\emph{Haṭhatattvakaumudī} 51.35
+\begin{versinnote}
+\tl{jñeyaṃ sarvapratītaṃ ca tajjñānaṃ mana ucyate |\\+}
+\tl{jñānaṃ jñeyaṃ samaṃ naṣṭaṃ nānyaḥ panthā dvitīyakaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_28]
+%Function of ca?
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*29}
+\begin{translation}[hp04_008_29]
+All this, everything moving and unmoving, is [just] a vision of the mind. When the mind has become free of the mind, it is called the absence of duality.
+\end{translation}
+
+\begin{sources}[hp04_008_29]
+Gauḍapāda's \emph{Māṇḍūkyopaniṣatkārikā} 3.31
+\begin{versinnote}
+\tl{manodṛśyam idaṃ dvaitaṃ yat kiṃ cit sacarācaram |\\+}
+\tl{manaso hy amanībhāve dvaitaṃ naivopalabhyate ||\\!}
+\end{versinnote}
+
+\emph{Amanaska} 2.79
+\begin{versinnote}
+\tl{manodṛśyam idaṃ sarvaṃ yat kim cit sacarācaram |\\+}
+\tl{manaso hy unmanībhāve 'dvaitabhāvaṃ pracakṣate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_008_29]
+\emph{Yogacintāmaṇi} (attr.~to the \emph{Amanaska})
+\begin{versinnote}
+\tl{manodṛśyam idaṃ sarvaṃ yat kiṃ cit sacarācaram |\\+}
+\tl{manasas tūnmanībhāvo 'dvaitābhāvaṃ pracakṣate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_29]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*30}
+\begin{translation}[hp04_008_30]
+As a result of abandoning the things that are to be known the mind attains dissolution (\emph{layam}). When the mind has attained dissolution, isolation (\textit{kaivalyam}) remains.
+\end{translation}
+
+\begin{sources}[hp04_008_30]
+\end{sources}
+
+\begin{testimonia}[hp04_008_30]
+\emph{Haṭharatnāvalī} 4.44
+\begin{versinnote}
+\tl{jñeyavastuparityāgād vilayaṃ yāti mānasaḥ |\\+}
+\tl{mānase vilayaṃ yāte kaivalyam upajāyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_008_30]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*31}
+\begin{translation}[hp04_008_31]
+“Dissolution, dissolution”, they say. What kind of characteristics does dissolution have? Because subliminal impressions do not arise again, dissolution is the forgetting of the objects of the senses.
+\end{translation}
+
+%\begin{sources}[hp04_008_31]
+%\end{sources}
+
+\begin{testimonia}[hp04_008_31]
+\emph{Haṭharatnāvalī} 1.13
+\begin{versinnote}
+\tl{layo laya iti prāhuḥ kīdṛśaṃ layalakṣaṇam |\\+}
+\tl{apunarbhavasaṃsthānaṃ layo viṣayavismṛtiḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogasārasaṅgraha} ??
+\begin{versinnote}
+\tl{śrīdattaḥ |\\+}
+\tl{layo laya iti prāhur īdṛśaṃ lakṣaṇaṃ sphuṭam |\\+}
+\tl{tatra sarvasamādhāne layo viṣayavismṛtiḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_008_31]
+This verse may have been inspired by the \emph{Mokṣopāya} (e.g.~1.2.2).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*32}
+\begin{translation}[hp04_008_32]
+Various methods like these, which are understood properly by one's own experience, are taught as paths to \textit{samādhi} by the great-souled teachers of former times.
+\end{translation}
+
+%\begin{sources}[hp04_008_32]
+%\end{sources}
+
+%\begin{testimonia}[hp04_008_32]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_008_32]
+%\end{philcomm}
+
+\begin{metre}[hp04_008_32]
+Anuṣṭubh (c: bha-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.8*33}
+\begin{translation}[hp04_008_33]
+Homage to Suṣumṇā, to Kuṇḍalinī, to the nectar arising from the moon, to the mind beyond mind state, to you whose nature is consciousness, the great Śakti.%??
+or whose nature is the consciousness of the great Śakti?
+JM: I have forgotten how we are writing nāḍīs etc.
+I like how we have everything here.
+\end{translation}
+
+%\begin{sources}[hp04_008_33]
+%\end{sources}
+
+%\begin{testimonia}[hp04_008_33]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_008_33]
+%\end{philcomm}
+
+\begin{metre}[hp04_008_33]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.8*34}
+\begin{translation}[hp04_008_34]
+The cultivation of the inner sound taught by Gorakṣa [and] approved even for deluded people for whom realisation of the highest reality is impossible is now taught.
+\end{translation}
+
+%\begin{sources}[hp04_008_34]
+%\end{sources}
+
+%\begin{testimonia}[hp04_008_34]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_008_34]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.9}
+\begin{translation}[hp04_009]
+The twelve and a half million methods of dissolution taught by  glorious Śiva are victorious.
+We think only the singular concentration on the inner sound is the most venerable of [all] the dissolutions.%??JM: How about "Glory to..." instead of "are victorious"?
+And then ""We consider one of the dissolutions in particular to be especially worthy of honour, concentration on the inner sound."
+\end{translation}
+
+\begin{sources}[hp04_009]
+\emph{Yogatārāvalī} 2
+\begin{versinnote}
+\tl{sadā śivoktāni sapādalakṣa-\\+}
+\tl{layāvadhānāni lasantu loke |\\+}
+\tl{nādānusandhānasamādhim ekaṃ\\+}
+\tl{manyāmahe mānyatamaṃ layānām ||\\+}
+\tl{\var{lasantu ] Sa : lasanti Pa : vasanti Ad W Eb Ec Ed : ca santi  Ea : hi sarva Ma}\\+}
+\tl{\var{loke ]  Ma  W Eall : bhūmau  Ad Pa Sa}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_009]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{śrīādināthena sapādakoṭi-\\+}
+\tl{layaprakārāḥ kathitā jayanti |\\+}
+\tl{nādānusandhānakam eva kāryam\\+}
+\tl{manyāmahe nānyatamaṃ layānām ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 1.12
+\begin{versinnote}
+\tl{śrīādināthena sapādakoṭi-\\+}
+\tl{layaprakārāḥ kathitā jayantu  |\\+}
+\tl{nādānusandhānakam eva kāryaṃ\\+}
+\tl{manyāmahe mānyatamaṃ layānām ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_009]
+%\end{philcomm}
+
+\begin{metre}[hp04_009]
+Upajāti
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.10}
+\begin{translation}[hp04_010]
+Seated in the pose of the liberated ones, the yogi should adopt \emph{śāṃbhavī mudrā} and, with his mind one-pointed, listen to the inner sound in his right ear.
+\end{translation}
+
+%\begin{sources}[hp04_010]
+%\end{sources}
+
+\begin{testimonia}[hp04_010]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{muktāsanasthito yogī mudrāṃ sandhāya śāṃbhavīm |\\+}
+\tl{śṛṇuyād dakṣiṇe karṇe nādam antargataṃ sadā ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (ms.~2244 f.~124r)
+\begin{versinnote}
+\tl{muktāsanasthito yogī mudrāṃ saṃdhāya śāṃbhavīṃ [|]\\+}
+\tl{śṛṇuyād dakṣiṇe karṇe nādam ekāṃtike sudhīḥ  [||]\\+}
+\tl{\var{dakṣiṇe karṇe ] B220,  dakṣirṇe 2244}\\+}
+\tl{\var{ekāṃtike ] 2244, ekāṃtate B220}\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_010]
+We read this verse here rather than at 4.50 because it follows on from the Śāmbhavī section and is unnecessary and repetitive at 4.50 because 4.51-52 explain the practice.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11}
+\begin{translation}[hp04_011]
+A fire that has been set on wood disappears together with the wood; the mind set on the [inner] sound dissolves together with the [inner] sound.
+\end{translation}
+
+%\begin{sources}[hp04_011]
+%\end{sources}
+
+\begin{testimonia}[hp04_011]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{kāṣṭhe pravartito vahniḥ kāṣṭhena saha śāmyati |\\+}
+\tl{nāde pravartitaṃ cittaṃ nādena saha līyate ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.15
+\begin{versinnote}
+\tl{kāṣṭhe pravartito vahniḥ kāṣṭhena saha līyate |\\+}
+\tl{nāde pravartitaṃ cittaṃ nādena saha śāmyati ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (ms. 2244 f.~124r)
+\begin{versinnote}
+\tl{kāṣṭhaiḥ pravartito vahniḥ kāṣṭhena saha śāmyati |\\+}
+\tl{nāde pravartitaṃ cittaṃ nādena saha līyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_011]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.12}
+\begin{translation}[hp04_012]
+Having forgotten everything external, the mind becomes one with the inner sound like milk and water, then quickly dissolves into the space of consciousness.
+\end{translation}
+
+%\begin{sources}[hp04_012]
+%\end{sources}
+
+\begin{testimonia}[hp04_012]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{vismṛtya sakalaṃ bāhyaṃ nāde dugdhāmbuvan naraḥ |\\+}
+\tl{ekībhūyātha sahasā cidākāśe vilīyate ||\\!}
+\end{versinnote}
+
+Upāsanāsārasaṅgraha (p. 106)
+\begin{versinnote}
+\tl{vismṛtya sakalaṃ bāhyaṃ nāde dagdhāṃbuvan manaḥ |\\+}
+\tl{ekībhūtaṃ tanyā cittaṃ rājayogābhidānakaṃ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (ms. 2244 f.~124r)
+\begin{versinnote}
+\tl{vismṛtya sakalaṃ bāhyaṃ nāde dugdhāṃbuvan manaḥ |\\+}
+\tl{ekībhūyātha sahasā cidākāśe vilīyate ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 39
+\begin{versinnote}
+\tl{vismṛtya sakalaṃ bāhyaṃ nāde dugdhāmbuvan manaḥ |\\+}
+\tl{ekībhūyātha sahasā cidākāśe vilīyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_012]
+%\end{philcomm}
+
+\begin{metre}[hp04_012]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.13}
+\begin{translation}[hp04_013]
+Having become intent on indifference through regular practice, the ascetic should immediately concentrate on nothing but the inner sound, which brings about the [state] beyond mind.%??JM: sadyas is odd.
+Apparently it has an archaic meaning of "daily".
+Consider?
+Immediately doesn't make sense.
+Or take it with unmanīkaraṇam.
+And sadābhyāsena could go with avadhārayet.
+\end{translation}
+
+%\begin{sources}[hp04_013]
+%\end{sources}
+
+\begin{testimonia}[hp04_013]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{audāsīnyaparo bhūtvā sadābhyāsena saṃyamī |\\+}
+\tl{unmanīkaraṇaṃ sadyo nādam evāvadhārayet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.38
+\begin{versinnote}
+\tl{audāsīnyaparo bhūtvā sadābhyāsena saṃyamī |\\+}
+\tl{unmanīkārakaṃ sadyo nādam eva sadābhyaset ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 40
+\begin{versinnote}
+\tl{udāsīnas tato bhūtvā sadābhyāsena saṃyamī |\\+}
+\tl{unmanīkārakaṃ sadyo nādam evāvadhārayet ||\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~106
+\begin{versinnote}
+\tl{audāsīnyaparo bhūtvā sadābhyāsena samyamī |\\+}
+\tl{unmanīkaraṇam sadyo nādam evānu [...] yet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_013]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.14}
+\begin{translation}[hp04_014]
+What kind of indifference?\\
+In the cold season, [indifference towards] whether [one has] an open tent or a quilt, with regard to good nourishment, whether it is cow's milk or water, with regard to food, whether it is lots of alms or forest roots, and with regard to the vessel for food, whether it is the hand, bucket or beggar's bowl.
+\end{translation}
+
+%\begin{sources}[hp04_014]
+%\end{sources}
+
+\begin{testimonia}[hp04_014]
+\emph{Yogacintāmaṇi} ref??
+\begin{versinnote}
+\tl{kīdṛśaṃ caudāsīnyam—\\+}
+\tl{śīte kāle kāpaṭī vā paṭī vā
+pathyāhāre gopayo vā payo vā |\\+}
+\tl{bhakṣye bhikṣāvṛndam āraṇyakandam
+pāṇau droṇī kā parā bhojapātram ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.7
+\begin{versinnote}
+\tl{śīte kāle dvau paṭī vā paṭī vā pathyāhāre gopayo vā payo vā |\\+}
+\tl{bhakṣye bhojye vṛttim āraṇyakaṃ vā pāṇī droṇī ko 'pi vā bhakṣyapātre ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī}  54.39
+\begin{versinnote}
+\tl{audāsīnyaṃ śītakāle paṭī vā
+pathyāhāro gopayo vā payo vā |\\+}
+\tl{bhojyaṃ bhikṣāvṛndam āraṇyakandaṃ
+pāṇī droṇī kāpi vā bhojyapātram ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp04_014]
+On \emph{caupaṭī} in the first \emph{pāda}, McGregor (1994:s.v.) and Callewaert (2009:s.v.) in their dictionaries of modern and old Hindi both give the meaning "open all around" for \emph{caupaṭ}.
+McGregor derives it from Sanskrit \emph{catuṣpaṭṭa}.
+Molesworth (1857:s.v.) in his Marathi dictionary gives "A quadrangular expanse or space, esp.
+as open and extended: also a broad and level tract" for \emph{caupaṭṭā}
+\end{philcomm}
+
+\begin{metre}[hp04_014]
+Śālinī 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.15}
+\begin{translation}[hp04_015]
+Having forever abandoned all worry and all activity, as a result of focusing on nothing but the inner sound the mind dissolves into the inner sound.
+\end{translation}
+
+%\begin{sources}[hp04_015]
+%\end{sources}
+
+\begin{testimonia}[hp04_015]
+\emph{Haṭharatnāvalī} 4.13
+\begin{versinnote}
+\tl{sarvacintāṃ samutsṛjya sarvaceṣṭāṃ ca sarvadā |\\+}
+\tl{nādam evānusandhānān nāde cittaṃ vilīyate ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (ms. 2244 f.~124r)
+\begin{versinnote}
+\tl{sarvaciṃtā[ṃ] parityajya sarvakāle ca sarvadā |\\+}
+\tl{nādam evānusandhatte nāde cittaṃ vilīyate ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 41
+\begin{versinnote}
+\tl{sarvacintāṃ samutsṛjya sarvaceṣṭāvivarjitaḥ |\\+}
+\tl{nādam evānusaṃdadhyān nāde cittaṃ vilīyate  ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_015]
+The occurrence of two finite verbs in the second hemistich of some witnesses (groups 2 and 3) is infelicitous and perhaps points to an error that arose from the \textit{nādam evānusandhānān} reading in the group one??
+mss.
+It seems the original reading was probably \textit{nāda evānusandhānāt}, and the -m- crept in as a hiatus break.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.16}
+\begin{translation}[hp04_016]
+Beginning (\emph{ārambhaḥ}), activation (\emph{ghaṭaḥ}), accumulation (\emph{paricayaḥ}) and completion (\emph{niṣpattiḥ}): they are the stages of yoga in all yogas.
+\end{translation}
+
+\begin{sources}[hp04_016]
+\emph{Amaraugha} 34
+\begin{versinnote}
+\tl{ārambhaś ca ghaṭaś caiva paricayaḥ prakīrtitaḥ |\\+}
+\tl{niṣpattiḥ sarvayogeṣu cetasas tasya bhūmikāḥ ||\\+}
+\tl{\var{paricaryaḥ prakīrtitaḥ Ga :
+tathā paricaya samṛtaḥ Ae (\textit{unmetr.}) :
+pracayaś ca tritīyakaḥ Ad Gb :
+pracayaś ca tṛtīryakaḥ  Ba :
+pracayaṃ ca tritīyakaṃ  Ta}\\+}
+\tl{\var{cetasas tasya bhūmikāḥ ] $\beta$ : yogāvasthā caturvidhā $\theta$}\\!}%?? variants necessary?
+\end{versinnote}
+
+\emph{Amṛtasiddhi} 19.2
+\begin{versinnote}
+\tl{ārambhaś ca ghaṭaś caiva paricayas tṛtīyakaḥ |\\+}
+\tl{niṣpannaḥ sarvaśeṣeṣu yogāvasthāḥ prakīrtitāḥ ||\\!}
+\end{versinnote}
+
+\emph{Śivasaṃhitā} 3.31
+\begin{versinnote}
+\tl{ārambhaś ca ghaṭaś caiva tathā paricayas tathā |\\+}
+\tl{niṣpattiḥ sarvayogeṣu yogāvasthā bhavanti tāḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_016]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{īśvaraprokte—\\+}
+\tl{ārambhaś ca ghaṭaś caiva tathā paricayo 'pi ca |\\+}
+\tl{niṣpattiḥ sarvayogeṣu yogāvasthā bhavanti tāḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.17
+\begin{versinnote}
+\tl{ārambhaś ca ghaṭaś caiva tathā paricayas tathā |\\+}
+\tl{niṣpattiḥ sarvayogeṣu yogāvasthā bhavanti tāḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_016]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.17}
+\begin{translation}[hp04_017]
+Now the beginning stage:\\
+As a result of the piercing of the knot of Brahmā, bliss arises in the void [and] an unstruck sound which has various tones is heard in the body.
+\end{translation}
+%?? MD: better to adopt "tatra ārambhaḥ"? The first component is often introduced with "tatra" after the enumeration in the important mss.
+
+\begin{sources}[hp04_017]
+\emph{Amaraugha} 35
+\begin{versinnote}
+\tl{brahmagranthes tathā bhedād ānandaḥ śūnyasambhavaḥ |\\+}
+\tl{vicitrakvaṇako dehe 'nāhataḥ śrūyate dhvaniḥ ||\\+}
+\tl{\var{°granthes tathā ] Ga : °granthes tato Ae : °granther bhaved Ba Gb Ta : °granthir bhaved Ad}\\+}
+\tl{\var{bhedād ] Ga : bhedāt Ae : vedhād Ba Gb Ta : vedhā Ad}\\!}%?? variants needed?
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_017]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{brahmagranthir bhaved bhinna ānandaḥ śūnyasaṃbhavaḥ |\\+}
+\tl{vicitrakṣaṇiko deho 'nāhataḥ śrūyate dhvaniḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.18
+\begin{versinnote}
+\tl{brahmarandhre bhaved bhedo yo nādaḥ sūryasaṃbhavaḥ |\\+}
+\tl{vicitrakvaṇado dehe 'nāhataḥ śrūyate dhvaniḥ ||\\+}
+\tl{\var{yo nādaḥ ] ānandaḥ \vl}\\+}
+\tl{\var{vicitrakvaṇado ] vicitrakvaṇako \vl}\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.15
+\begin{versinnote}
+\tl{tatra ārambhaḥ –\\+}
+\tl{brahmagranthir bhaved bhinnād ānandaḥ śūnyasambhavaḥ |\\+}
+\tl{vicitrakvaṇiko dehe 'nāhataḥ śrūyate dhvaniḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_017]
+Brahmānanda understands the voids associated with each stage to be places in the body.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.18}
+\begin{translation}[hp04_018]
+In the beginning [stage], which is [the first phase of] voidness and in which the heart is full [of bliss], one becomes a yogi with a divine body, splendour, a divine smell and freedom from disease.%??
+space after śūnye in edition.
+??
+JM:  With a divine body, radiant, smelling heavenly, free from disease [and] his heart full [of bliss], in the void in the beginning [stage] the practitioner attains yoga.
+\end{translation}
+
+\begin{sources}[hp04_018]
+\emph{Amaraugha} verse 36
+\begin{versinnote}
+\tl{divyadehaś ca tejasvī divyagandho hy arogavān |\\+}
+\tl{saṃpūrṇahṛdaye śūnye tv ārambhe yogavān bhavet ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_018]
+\emph{Haṭharatnāvalī} 4.19
+\begin{versinnote}
+\tl{divyadehaḥ sutejasvī divyagandhas tv arogavān |\\+}
+\tl{saṃpūrṇahṛdaye śūnye tv ārambhe yogavān bhavet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.18
+\begin{versinnote}
+\tl{tejasvī divyagandhaś ca divyadeho 'py arogavān |\\+}
+\tl{sampūrṇahṛdaye śūnye tv ārambhe yogavān bhavet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_018]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.19}
+\begin{translation}[hp04_019]
+Now the activation stage:\\
+In the second stage the breath, after bringing about union, goes into the middle [channel]. Then the yogi has a firm posture [and] he becomes a gnostic and equal to a god.
+\end{translation}
+
+\begin{sources}[hp04_019]
+\emph{Amaraugha} 37
+\begin{versinnote}
+\tl{dvitīye saṅghaṭīkṛtya vāyur bhavati madhyagaḥ |\\+}
+\tl{dṛḍhāsano bhaved yogī jñānī devasamas tadā ||\\+}
+\tl{\var{dvitīye saṅ- ] $\beta$ : dvitīyāyāṃ $\theta$}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_019]
+\emph{Haṭharatnāvalī} 4.20
+\begin{versinnote}
+\tl{dvitīyāyāṃ ghaṭīkṛtya vāyur bhavati madhyagaḥ |\\+}
+\tl{dṛḍhāsano bhaved yogī kāmadevasamas tadā ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.15
+\begin{versinnote}
+\tl{atha ghaṭāvasthā –\\+}
+\tl{dvitīyāyāṃ ghaṭīṃ kṛtvā vāyur bhavati madhyagaḥ |\\+}
+\tl{dṛḍhāsano bhaved yogī jñānī devasamas tathā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_019]
+The name of the second stage, \emph{ghaṭa} (and the related form \emph{ghaṭīkṛtya} found in the first \emph{pāda} of this verse) can be understood in at least three ways, as union, activation or pot (the latter with an alchemical connotation, on which see 3.12 and Mallinson and Szanto 2021:20–23).
+In the \emph{Dattātreyayogaśāstra} (verse 90) its primary meaning is union and the united pairs are \emph{prāṇa} and \emph{apāna}, \emph{jīvātman} and \emph{paramātman}, and \emph{nāda} and \emph{bindu}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.20}
+\begin{translation}[hp04_020]
+Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a \textit{bherī} drum occurs.%??
+in ed join śabdas and tadā
+\end{translation}
+
+\begin{sources}[hp04_020]
+\emph{Amaraugha} 38
+\begin{versinnote}
+\tl{viṣṇugranthes tato bhedāt paramānandasūcakaḥ |\\+}
+\tl{atiśūnye vimardaś ca bherīśabdas tato bhavet ||\\+}
+\tl{\var{atiśūnye ] Ad Gb Ta : atiśūnyo $\beta$ Ba}\\+}
+\tl{\var{tato ] $\beta$ : tathā  Ba : tadā Ad Gb Ta}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_020]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{viṣṇugranthir yadā bhinnaḥ paramānandasūcakaḥ |\\+}
+\tl{atiśūnyavibhedaś ca bherīśabdas tadā bhavet ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.21
+\begin{versinnote}
+\tl{viṣṇugranthes tathā bhedaḥ paramānandasūcakaḥ |\\+}
+\tl{atiśūnye vimardaś ca bherīśabdas tathā bhavet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.21
+\begin{versinnote}
+\tl{viṣṇugranthir yadā bhinnā paramānandasūcikā |\\+}
+\tl{atiśūnyavibhedaś ca bherīśabdas tathā bhavet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_020]
+V3??
+and HR have the reading of the \textit{Amaraugha} (\textit{atiśūnye vimardaś ca}), which was adapted from the Buddhist concept of a series of four void-like meditative states and moments.
+In the \emph{Amaraugha},  two of the moments were left out and two were reinterpreted (as in the case of \emph{vimarda}, which seems to have be retained with the more general sense of a pounding sound).
+%It is possible that Svātmārāma adopted this reading but he may have also changed it to \textit{atiśūnyavibheda} (as attested by groups 2 and 3)  in an attempt to make sense of the pāda (i.e., the variety, atiśūnya, arises).??JM: I percented this as we've gone with alpha1.
+In fact do we need a note here?
+How about a more general note on the stages at the beginning of this section, referencing Amṛtasiddhi, Amaraugha and DYŚ?
+See Birch 2019: 968–969; Mallinson and Szanto 2021:19–20.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.21}
+\begin{translation}[hp04_021]
+Now the accumulation stage:\\
+In the third [stage], having pierced [the knot of Viṣṇu], the sound of a \textit{mardala} drum arises in the sky.
+Then [the yogi] reaches the great void, the abode of all supernatural powers.%??
+vihāyas = sky, not void so I've changed tr.
+\end{translation}
+%
+
+\begin{sources}[hp04_021]
+\emph{Amaraugha} 39
+\begin{versinnote}
+\tl{tṛtīyāyāṃ tato bhittvā ninādo mardaladhvaniḥ |\\+}
+\tl{mahāśūnyaṃ tato jātaṃ sarvasiddhisamāśrayam ||\\+}
+\tl{\var{bhittvā ninādo ] Ga : bhītvādinādau Ae : bhittvā vipāko Ba Ad Gb Ta }\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_021]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{tṛtīyāyāṃ tato bhitvā vimāyo mardaladhvaniḥ |\\+}
+\tl{mahāśūnyaṃ tathā yāti sarvasiddhisamāśrayam ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.22
+\begin{versinnote}
+\tl{tṛtīyāyāṃ tato nityaṃ āviṣkāro marddladhvaniḥ |\\+}
+\tl{mahāśūnyaṃ tato yāti sarvasiddhisamāśrayaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.22
+\begin{versinnote}
+\tl{atha paricayāvasthā ||\\+}
+\tl{tṛtīyāyāṃ tato jitvā sahajānandasambhavaḥ |\\+}
+\tl{doṣaduḥkhajarāmṛtyuḥ kṣudhānidrāvivarjitaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_021]
+Here the object of piercing is unspecified, but it is likely to refer to the knot of Viṣṇu mentioned in the previous verse.
+%?? MD: Pāda c - tathā (Gr2) to be changed to tato (Gr3) or tadā (alpha*,V3,Jyo)? JM: tadā is good and tr. already translates that way
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.22}
+\begin{translation}[hp04_022]
+Now the perfection stage:\\
+Having overcome the [supreme] bliss of the mind, there arises innate bliss. [The yogi becomes] free of disease, suffering, old age, death, hunger and sleep.
+\end{translation}
+
+\begin{sources}[hp04_022]
+\emph{Amaraugha} 40
+\begin{versinnote}
+\tl{paramānandarocitvāt sahajānandasambhavaḥ |\\+}
+\tl{doṣaduḥkhajarāmṛtyukṣudhānidrāvivarjitaḥ ||\\+}
+\tl{\var{40a paramānanda° ] $\beta$; cittānandaṃ $\theta$ • °rocitvāt ] $\beta$; tato jitvā $\theta$}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_022]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{cittānandaṃ tato jitvā sahajānandasaṃbhavaḥ |\\+}
+\tl{doṣaduḥkhakṣudhānidrājarāmṛtyuvivarjitaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.23
+\begin{versinnote}
+\tl{cidānandaṃ tato jitvā paramānandasaṃbhavaḥ |\\+}
+\tl{doṣaduḥkhajarāmṛtyukṣudhānidrāvivarjitaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.22
+\begin{versinnote}
+\tl{tṛtīyāyāṃ tato jitvā sahajānandasambhavaḥ |\\+}
+\tl{doṣaduḥkhajarāmṛtyuḥ kṣudhānidrāvivarjitaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_022]
+22a’s \emph{cittānāndaṃ tato jitvā} is perhaps a patch for the somewhat obscure \emph{paramānandarocitvāt} in the source text.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.23}
+\begin{translation}[hp04_023]
+Now the perfection stage:\\
+Then, having pierced Rudra's knot, the breath goes to all the  [of deities in the body]. In the perfected [stage] the sound of a flute becomes the sound of a resonating lute.
+\end{translation}
+
+\begin{sources}[hp04_023]
+\emph{Amaraugha} 41
+\begin{versinnote}
+\tl{rudragranthiṃ tato bhittvā sarvapīṭhagato 'nilaḥ |\\+}
+\tl{niṣpanno vaiṇavaḥ śabdaḥ kvaṇadvīṇākvaṇo bhavet ||\\+}
+\tl{\var{sarva° ] Ae : sarvaṃ Ga : sattva° $\theta$}\\+}
+\tl{\var{niṣpanno ] Ga : niṣpannā Ae : niṣpattau Ad Gb T : ++ttau Ba }\\+}
+\tl{\var{kvaṇadvīṇākvaṇo ] conj. : kvaṇañ cailakvaṇo S1 : kvaṇañ caiva kvaṇo S2 : kvaṇadvitakvaṇo L3 : kvaṇanvitakvaṇo L2 : kvaṇanvītakva+ L4 : kvaṇatbhakvaṇo L1}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_023]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{rudragranthiṃ tato bhitvā sarvapīṭhagato 'nilaḥ |\\+}
+\tl{niṣṭhāto vaiṇavaḥ śabdaḥ kvaṇadvīṇākvaṇo bhavet ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.24
+\begin{versinnote}
+\tl{rudragranthiṃ tato bhitvā śarvapīṭhagato 'nilaḥ |\\+}
+\tl{niṣpattau vaiṇavaḥ śabdaḥ kvanadvīṇākvaṇo bhavet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.24
+\begin{versinnote}
+\tl{atha niṣpattiḥ –\\+}
+\tl{rudragranthiṃ tato bhitvā śarvapīṭhagato'nalaḥ |\\+}
+\tl{niṣpanno vaiṇavaḥ śabdo kvaṇadvīṇākvaṇo bhavet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_023]
+The reading we have adopted for the fourth \emph{pāda}, \emph{kvaṇadvīṇākvaṇo}, Birch’s conjecture in his edition of the \emph{Amaraugha} is supported by several testimonia and \emph{Amṛtasiddhi} 31.2, where it is said that the sound of a \emph{vīṇā} arises in the fourth stage.
+
+The reading \emph{śarva}, i.e.~Śiva, found in some testimonia and the \emph{Jyotsnā} makes good sense, but in its description of this stage, the \emph{Amṛtasiddhi}, which is the ultimate source of this passage has \emph{sarva} (30.1).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.24}
+\begin{translation}[hp04_024]
+Then the mind having become one [with the sound??] is called Rājayoga. That yogi becomes a creator and destroyer, an equal to god.%??see note
+\end{translation}
+
+\begin{sources}[hp04_024]
+\emph{Amaraugha} 42
+\begin{versinnote}
+\tl{ekībhūtaṃ tadā cittaṃ rājayogābhidhānakam |\\+}
+\tl{sṛṣṭisaṃhārakartāsau yogīśvarasamo bhavet ||+}
+\tl{\var{rājayogābhidhānakam ] Ga Ba Ad Gb Ed ; rājayogo 'bhidhīyate Ae}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_024]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{ekībhūtaṃ tathā cittaṃ rājayogābhidhāyakam |\\+}
+\tl{sṛṣṭisaṃhārakartāsau yogīśvarasamo bhavet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_024]
+The bahuvrīhi rājayogābhidhānakaṃ means that the mind is called rājayoga, which is very unusual. Perhaps °bhidhāyakaṃ is better, meaning "expresses rājayoga", but it is only in J7 and V3.
+
+%??SS: Then the mind becomes one. This is called Rājayoga. JM: looks best to me
+
+%??JM: yogīśvarasamo = equal to a yogi sovereign? Odd to take asau with yogī.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.24*1}
+%\begin{translation}[hp04_024_1]
+%\end{translation}
+
+%\begin{sources}[hp04_024_1]
+%\end{sources}
+
+%\begin{testimonia}[hp04_024_1]
+%\end{testimonia}
+
+\begin{philcomm}[hp04_024_1]
+See 4.32.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.25}
+\begin{translation}[hp04_025]
+Whether or not this is liberation, in this very state a great, unbroken pleasure, which is rich in the nectar of absorption, is attained from Rājayoga.
+\end{translation}
+
+%\begin{sources}[hp04_025]
+%\end{sources}
+
+\begin{testimonia}[hp04_025]
+\emph{Yogacintāmaṇi} ??
+\begin{versinnote}
+\tl{astu vā māstu vā siddhir atraivākhaṇḍitaṃ sukham |\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.16
+\begin{versinnote}
+\tl{astu vā māstu vā muktir atraivākhaṇḍitaṃ mahat |\\+}
+\tl{layāmṛtaṃ laye saukhyaṃ rājayogād avāpyate ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.35
+\begin{versinnote}
+\tl{astu vā māstu vā muktir atraivākhaṇḍitaṃ mahat |\\+}
+\tl{layāmṛtamayaṃ saukhyaṃ rājayogād avāpyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_025]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.26}
+\begin{translation}[hp04_026]
+Rājayoga without Haṭha and Haṭha without Rājayoga do not succeed so [the yogi] should practise both until the perfection stage.
+\end{translation}
+
+\begin{sources}[hp04_026]
+\emph{Śivasaṃhitā} 5.222
+\begin{versinnote}
+\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
+\tl{na sidhyati tato yugmam āniṣpatteḥ samabhyaset |\\+}
+\tl{tasmāt pravartate yogī haṭhe sadgurumārgataḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_026]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{haṭhaṃ vinā rājayogaṃ rājayogaṃ vinā haṭham |\\+}
+\tl{ye vai caranti tān manye prayāsaphalavarjitān iti ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 1.19
+\begin{versinnote}
+\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
+\tl{vyāptiḥ syād avinābhūtā śrīrājahaṭhayogayoḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 55.1
+\begin{versinnote}
+\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ  |\\+}
+\tl{na sidhyati tato yugmam āniṣpatteḥ samabhyaset  ||\\!} % = HP 2.77
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_026]
+%\end{philcomm}
+
+\begin{metre}[hp04_026]
+Anuṣṭubh (a: ra-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.27}
+\begin{translation}[hp04_027]
+I consider those who are ignorant of Rājayoga and engage in only Haṭha to be labourers who get no reward for their efforts.%??JM: farm labourers?
+ploughmen?
+It would be good to get the agricultural meaning in.
+\end{translation}
+
+%\begin{sources}[hp04_027]
+%\end{sources}
+
+%\begin{testimonia}[hp04_027]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_027]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.28}
+\begin{translation}[hp04_028]
+The supreme reality is the seed, Haṭha the ground and indifference water. With these three the creeper of good fortune, the no-mind state, immediately grows.
+\end{translation}
+
+%\begin{sources}[hp04_028]
+%\end{sources}
+
+\begin{testimonia}[hp04_028]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{nādo bījaṃ haṭhaḥ kṣetram audāsīnyaṃ jalaṃ smṛtam |\\+}
+\tl{unmanīkalpalatikā sadya evodbhaviṣyati ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_028]
+The meaning of \emph{tattva} here is unclear.
+It is a synonym for \emph{samādhi} and \emph{unmanī} in the next verses.
+Brahmānanda says that \emph{tattva} means \emph{citta}, which works well: the mind grows to the beyond-mind state.
+
+The reading \emph{smṛtam} is also possible and well-attested (groups 2 and 3).%??JM: not in apparatus.
+
+\end{philcomm}
+
+\begin{metre}[hp04_028]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.29}
+\begin{translation}[hp04_029]
+Rājayoga, samādhi, unmanī, manonmanī, amaraugha, advaita, nirālamba, nirañjana...%?? format. Italic? Capitals?
+\end{translation}
+
+%\begin{sources}[hp04_029]
+%\end{sources}
+
+\begin{testimonia}[hp04_029]
+\emph{Yogacintāmaṇi} ??
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{rājayogaḥ samādhiś ca unmanī ca manonmanī |\\+}
+\tl{amaraughaughacāndrīva nirālambaṃ nirañjanam ||\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~106
+\begin{versinnote}
+\tl{haṭhapradīpikāyāṃ |\\+}
+\tl{rājayogas samādhiś cāpy unmanī ca manonmanī |\\+}
+\tl{amaraugho pi cādvaitaṃ nirālambaṃ niraṃjanaṃ  ||\\!}
+\end{versinnote}
+
+\emph{Yogasārasaṅgraha} p.~60
+\begin{versinnote}
+\tl{rājayogaḥ samādhiś conmanī ca manonmanī |\\+}
+\tl{śivayogo layas tatvaṃ śūnyāśūnyaṃ nirañjanam |\\!}
+\end{versinnote}
+
+\emph{Haṃsavilāsa} (p. 47)
+\begin{versinnote}
+\tl{rājayogaḥ samādhiśca unmanī ca manonmanī |\\+}
+\tl{amaraughālayas tatra śūnyāc chūnya paraṃ padam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_029]
+%\emph{amaraughaughacāndrī} appears to be a mistake perhaps because the word \emph{amaraugha} became \emph{amarolī}, hence prompting the change to \emph{candrī}.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.30}
+\begin{translation}[hp04_030]
+amanaska, laya, tattva, śūnyāśūnya, para pada, jīvanmukti, sahaja and turya are synonyms.
+\end{translation}
+
+\begin{sources}[hp04_030]
+\emph{Yogacintāmaṇi} ??
+\begin{versinnote}
+\tl{amanasko layaś caiva śūnyāśūnyaṃ parāparam |\\+}
+\tl{jīvanmuktiś ca sahajaṃ turyaṃ cety ekavācakam iti ||\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~106
+\begin{versinnote}
+\tl{ama[na]sko layas tatvaṃ śūnyāśūnyaparaṃ  padaṃ |\\+}
+\tl{jīvanmuktiś ca sahajaṃ turyaṃ cety ekavācakaṃ  ||\\!}
+\end{versinnote}
+
+\emph{Yogasārasaṅgraha} p.~60
+\begin{versinnote}
+\tl{amanaskaṃ yathā caitan nirālambaṃ nirañjanam |\\+}
+\tl{jīvanmuktiś ca sahajam ity adir hy ekavācakam ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_030]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_030]
+%tattva may have dropped out of Gr2 and 3 because of the awkwardness of including it after tattvaṃ bījam in the previous verse (if tattva was considered something else).
+%\end{philcomm}
+
+\begin{metre}[hp04_030]
+Anuṣṭubh (c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.31}
+\begin{translation}[hp04_031]
+Two paths for the quick attainment of Unmanī are approved by me: tattva, the supreme bliss, and honouring the inner sound...
+\end{translation}
+
+%\begin{sources}[hp04_031]
+%\end{sources}
+
+\begin{testimonia}[hp04_031]
+\emph{Yogacintāmaṇi} ??
+\begin{versinnote}
+\tl{unmanyavāptaye śīghraṃ dvau mārgau mama saṃmatau |\\+}
+\tl{tatvaṃ paramasaukhyaṃ vā nādopāsanam eva vā ||\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~106
+\begin{versinnote}
+\tl{unmanyavāptaye śīghraṃ mārgau dvau mama sammatau |\\+}
+\tl{tatvaṃ paramasākhyaṃ vā nādopāsanam eva vā  ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_031]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.32}
+\begin{translation}[hp04_032]
+which is approved even for stupid people whose minds are immersed in [worldly] pleasures. Dissolution which instantly bestows bliss arises from the inner sound.
+\end{translation}
+
+%\begin{sources}[hp04_032]
+%\end{sources}
+
+\begin{testimonia}[hp04_032]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{saukhyapraviṣṭacittānāṃ mūḍhānām api saṃmatam |\\+}
+\tl{sadya ānandasandhāyī jāyate nādajo layaḥ ||\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~106
+\begin{versinnote}
+\tl{sāṅkhye praviṣṭacittānāṃ mūḍhānām api saṃmateḥ |\\+}
+\tl{tasya svānaṃdasa [...] ryo jāyate nādajo layaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_032]
+%J5 has saṃmataḥ, agreeing with layaḥ
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.32*1}
+\begin{translation}[hp04_032_1]
+There is one seed [syllable] consisting of creation; one \emph{mudrā}, \emph{khecarī}, one god, the unsupported, [and] one state, mind beyond the mind.
+\end{translation}
+
+\begin{sources}[hp04_032_1]
+\emph{Tantrāloka} 32.64
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ bījaṃ yadvīryaṃ sarvamantragam |\\+}
+\tl{ekā mudrā khecarī ca mudraughaḥ prāṇito yayā ||\\!}
+\end{versinnote}
+
+\emph{Tantrālokaviveka} 32.63
+\begin{versinnote}
+\tl{yad āgamaḥ –\\+}
+\tl{ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī	|\\+}
+\tl{dvāvekaṃ yo vijānāti sa vai pūjyaḥ kulāgame	||\\!}
+\end{versinnote}
+\emph{Śivasūtravimarśinī} 5
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ [sṛṣṭimayaṃ bījam iti mantravīryarūpam aham iti bījam | 
+mudrā parabhairavīyātmā |] bījam ekā mudrā ca khecarī |\\+}
+\tl{dvāv etau yasya jāyete so'tiśāntapade sthitaḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_032_1]
+\emph{Haṭharatnāvalī} 4.28
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī |\\+}
+\tl{eko devo nirālambaḥ ekāvasthā manonmanī||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣad} 52cd-53ab
+\begin{versinnote}
+\tl{śaṅkhadundubhinādaṃ ca na śruṇoti kadācana || 52 ||\\+}
+\tl{kāṣṭhavajjñāyate deha unmanyāvasthayā dhruvam |\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_032_1]
+See 3.48.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.32*2}
+\begin{translation}[hp04_032_2]
+[The yogi] never hears the sounds of [even] conch shells and large drums. As a result of the state of no mind, the body assuredly becomes as [insentient as a piece of] wood.
+\end{translation}
+
+\begin{sources}[hp04_032_2]
+Jñānasāra 3.7
+\begin{versinnote}
+\tl{śaṅkhadundubhinādena na śṛṇoti kadācana |\\+}
+\tl{kāṣṭava[j] jñāyate yogī notpattyā vai prajāyate ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_032_2]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_032_2]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.32*3}
+\begin{translation}[hp04_032_3]
+Free from all states of mind and all thought, the yogi is as if dead. He is liberated. In this there is no doubt.
+\end{translation}
+
+%\begin{sources}[hp04_032_3]
+%\end{sources}
+
+\begin{testimonia}[hp04_032_3]
+\emph{Nādabindūpaniṣad} 51cd-52ab
+\begin{versinnote}
+\tl{sarvāvasthāvinirmuktaḥ sarvacintāvivarjitaḥ || 51 ||\\+}
+\tl{mṛtavat tiṣṭhate yogī sa mukto nātra saṃśayaḥ |\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 51.75
+\begin{versinnote}
+\tl{sarvāvasthāvinirmuktaḥ sarvacintāvivarjitaḥ |\\+}
+\tl{mṛtavattiṣṭhate yogī sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_032_3]
+%Maybe kāṣṭhavat crept in as a repetition of the previous verse.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.32*4}
+\begin{translation}[hp04_032_4]
+The yogi in \textit{samādhi} experiences neither cold nor heat, neither suffering nor pleasure, neither praise nor scorn.
+\end{translation}
+
+\begin{sources}[hp04_032_4]
+\emph{Vivekamārtaṇḍa} 166
+\begin{versinnote}
+\tl{nābhijānāti śītoṣṇaṃ na duḥkhaṃ na sukhaṃ tathā |\\+}
+\tl{na mānaṃ nāpamānaṃ ca yogī yuktaḥ samādhinā ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_032_4]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_032_4]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.32*5}
+\begin{translation}[hp04_032_5]
+The yogi in \textit{samādhi} cannot be wounded by any weapon, killed by any person or overpowered by mantras and magic.
+\end{translation}
+
+\begin{sources}[hp04_032_5]
+\emph{Vivekamārtaṇḍa} 168
+\begin{versinnote}
+\tl{abhedyaḥ sarvaśastrāṇām avadhyaḥ sarvadehinām |\\+}
+\tl{agrāhyo mantratantrāṇāṃ yogī yuktaḥ samādhinā ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_032_5]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_032_5]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.32*6}
+\begin{translation}[hp04_032_6]
+The yogi in \textit{samādhi} does not experience smell, taste, form, touch, sound, himself nor anyone else.
+\end{translation}
+
+\begin{sources}[hp04_032_6]
+\emph{Vivekamārtaṇḍa} 165
+\begin{versinnote}
+\tl{na gandhaṃ na rasaṃ rūpaṃ na ca sparśaṃ na nisvanam |\\+}
+\tl{nātmānaṃ na paraṃ vetti yogī yuktaḥ samādhinā ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_032_6]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_032_6]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.32*7}
+\begin{translation}[hp04_032_7]
+He is indeed truly liberated whose mind is neither asleep nor awake, neither mindful nor otherwise, and neither stops nor starts.
+\end{translation}
+
+\begin{sources}[hp04_032_7]
+\emph{Gorakṣaśataka} 7
+\begin{versinnote}
+\tl{cittaṃ na suptaṃ no jāgrat smṛtiman na ca nānyathā |\\+}
+\tl{nāstam eti na codeti yasyāsau mukta eva sa ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_032_7]
+%\end{testimonia}
+
+\begin{philcomm}[hp04_032_7]
+The first line is corrupt in the \emph{Gorakṣaśataka} witnesses; its text has been constituted from the \emph{Haṭhapradīpikā}.
+\end{philcomm}
+
+\begin{metre}[hp04_032_7]
+Anuṣṭubh (a: ma-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.32*8}
+\begin{translation}[hp04_032_8]
+[The yogi] who remains at ease as though asleep in the waking state, without breathing in and out, is definitely liberated.
+\end{translation}
+
+\begin{sources}[hp04_032_8]
+\emph{Amanaska} 2.59
+\begin{versinnote}
+\tl{sadā jāgradavasthāyāṃ suptavad yo 'vatiṣṭhate |\\+}
+\tl{niśvāsocchvāsahīnas ca niścitaṃ mukta eva saḥ ||\\+}
+\tl{\var{sadā jāgradavasthāyāṃ ] \vl
+sadā jāgṛvadasthāyāṃ,
+sadā jāgrat apasthāyāṃ,
+sadā jāgradavasther ya,
+sa jāgras tadavasthāyāṃ,
+yadā jāgṛdavasthāyāṃ,
+yo jāgrad yad avasthāyāṃ,
+svapnajāgradavasthāyāṃ,
+suptajāgradavasthāyāṃ}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_032_8]
+\emph{Kulārṇavatantra} 9.11
+\begin{versinnote}
+\tl{svapnajāgradavasthāyāṃ suptavat yo 'avatiṣṭhate | \\+}
+\tl{niśvāsocchvāsahīnaśca niścitaṃ mukta eva saḥ || 11 ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 1.64
+\begin{versinnote}
+\tl{svapnajāgradavasthāyāṃ suptavat yo 'vatiṣṭhate |\\+}
+\tl{niśvāsocchvāsahīnaś ca niścitaṃ mukta eva saḥ ||64||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 55.24
+\begin{versinnote}
+\tl{sadā jāgradavasthāyāṃ suptavad yo ‘vatiṣṭhate |\\+}
+\tl{niśvāsocchvāsavihīnaś ca niścitaṃ mukta eva saḥ || 55.24 = AY 2.59\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{sadā jāgradavasthāyāṃ suptavada yo'vatiṣṭhate ||\\+}
+\tl{niḥśvāsocchāsahīnaśca niścitaṃ mukta eva saḥ|\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_032_8]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.33}
+\begin{translation}[hp04_033]
+Only the glorious guru lord knows the unique ineffable bliss that has arisen in the hearts of yogi lords who experience \textit{samādhi} by concentrating on the inner sound.
+\end{translation}
+
+\begin{sources}[hp04_033]
+Yogatārāvalī 3
+\begin{versinnote}
+\tl{nādānusandhānasamādhibhājāṃ yogīśvarāṇāṃ hṛdaye pragūḍham |\\+}
+\tl{ānandamātraṃ vacasām avācyaṃ jānāti taṃ śrīgurunātha ekaḥ ||\\+}
+\tl{\var{ekaḥ ] Pa : eva Ad}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_033]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{nādānusandhānasamādhibhājāṃ
+yogīśvarāṇāṃ hṛdaye prarūḍhaṃ |\\+}
+\tl{ānandam ekaṃ vacasām avācyaṃ
+jānāti tatvaṃ gurunātha eva ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.5
+\begin{versinnote}
+\tl{nādānusandhānasamādhibhājāṃ yogīśvarāṇāṃ hṛdaye prarūḍham |\\+}
+\tl{ānandam ekaṃ vacaso 'py agamyaṃ jānāti taṃ śrīgurunātha eva ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_033]
+%Check consistency on tr. of anusandhāna.
+%\end{philcomm}
+
+\begin{metre}[hp04_033]
+Upajāti
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.33*1}
+\begin{translation}[hp04_033_1]
+Seated in the pose of the liberated ones, the yogi should adopt \emph{śāṃbhavī mudrā} and listen continuously to the inner sound in his right ear.
+\end{translation}
+%
+
+%\begin{sources}[hp04_033_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_033_1]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{muktāsanasthito yogī mudrāṃ sandhāya śāṃbhavīm |\\+}
+\tl{śṛṇuyād dakṣiṇe karṇe nādam antargataṃ sadā ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (ms. 2244 f.~124r)
+\begin{versinnote}
+\tl{muktāsanasthito yogī mudrāṃ saṃdhāya śāṃbhavīṃ [|]\\+}
+\tl{śṛṇuyād dakṣiṇe karṇe nādam ekāṃtike sudhīḥ  [||]\\+}
+\tl{\var{dakṣiṇe karṇe ] B220,  dakṣirṇe 2244}\\+}
+\tl{\var{ekāṃtike ] 2244, ekāṃtate B220}\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 31
+\begin{versinnote}
+\tl{siddhāsane sthito yogī mudrāṃ sandhāya vaiṣṇavīm |\\+}
+\tl{śṛṇuyād dakṣiṇe karṇe nādam antargataṃ sadā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_033_1]
+See 4.10.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.34}
+\begin{translation}[hp04_034]
+[The yogi] who desires yogic sovereignty should abandon all thought and concentrate with an attentive mind on nothing but the internal sound.
+\end{translation}
+
+%\begin{sources}[hp04_034]
+%\end{sources}
+
+\begin{testimonia}[hp04_034]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{sarvacintāṃ parityajya sāvadhānena cetasā |\\+}
+\tl{nādam evānusandhatte yogasāmrājyam icchatā ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.14
+\begin{versinnote}
+\tl{sarvacintāṃ parityajya sāvadhānena cetasā |\\+}
+\tl{nāda evānusandheyaḥ yogasāmrājyasiddhaye ||\\!}
+\end{versinnote}
+
+Haṭhasaṅketacandikā (f. 124r)
+\begin{versinnote}
+\tl{sarvacittaṃ parityajya sāvadhānena cetasā |\\+}
+\tl{nāda evānusaṃdheyo yogasāmrājyam icchatā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_034]
+%N3 icchatā
+%yogasāmrājya = rājayogasiddhi
+%numbering is array in the apparatus
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.35}
+\begin{translation}[hp04_035]
+The sage should block his ears with cotton and fix the mind on the sound which he hears, until he attains a state of stillness.
+\end{translation}
+
+%\begin{sources}[hp04_035]
+%\end{sources}
+
+\begin{testimonia}[hp04_035]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{karṇau pidhāya hastena yaḥ śṛṇoti dhvaniṃ muniḥ |\\+}
+\tl{tāvac cittaṃ sthiraṃ kuryād yāvat sthirapadaṃ vrajet ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.8
+\begin{versinnote}
+\tl{karṇau pidhāya tūlena yaḥ śṛṇoti dhvaniṃ yamī |\\+}
+\tl{tatra cittaṃ sthiraṃ kuryād yāvat sthirapadaṃ vrajet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (f. 124r)
+\begin{versinnote}
+\tl{karṇau pidhāya hastābhyāṃ yaś śṛṇoti dhvaniṃ muniḥ |\\+}
+\tl{tatra cittaṃ sthiraṃ kuryād yāvat sthirapadaṃ vrajet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_035]
+%adopt sthirīkuryād [MD: Done.]%??JM: remove space after sthirī  in ed.
+%mūlena (alpha) probably came from tūlena.
+The hasta variations seem impractible because the yogi would have to hold the arms up, but may have been adopted by yogis who did not have access to cotton (?).??JM: I think these comments are unnecessary (and surely it is possible/practicable to use the fingers?).
+
+Several witnesses have \emph{hastābhyām} or \emph{hastena} here.
+Blocking the ears with the fingers in order to listen to the inner sounds is attested as early as the \textit{Svacchandatantra}, in a practice called \textit{ṣaṇmukhīkaraṇa}, in which the other openings of the head are also blocked with the fingers (Vasudeva 2004: 272 n.66).
+\emph{Śivasaṃhitā} 5.36–46 teaches a similar practice.
+%aṅguṣṭābhyām ubhe śrotre tarjanībhyāṃ vilocane
+%nāsārandhre ca madhyābhyām anāmābhyāṃ mukhaṃ dṛḍham 5.36
+%nirudhya mārutaṃ yogī yadaivaṃ kurute bhṛśam
+%tadā tatkṣaṇamātmānaṃ jyotīrūpaṃ sa paśyati 5.37
+%tattejo dṛśyate yena kṣaṇamātraṃ nirākulam
+%sarvapāpavinirmuktaḥ sa yāti paramāṃ gatim 5.38
+%nirantarakṛtābhyāsād yogī vigatakalmaṣaḥ
+%sarvaṃ dehādi vismṛtya tadbhinnas tu svayaṃ bhavet 5.39
+%yaḥ karoti sadābhyāsaṃ guptācāreṇa mānavaḥ
+%sa vai brahmaṇi līnaḥ syāt pāpakarmarato yadi 5.40
+%gopanīyaḥ prayatnena sadyaḥpratyayakārakaḥ
+%nirvāṇadāyako loke yogo 'yaṃ mama vallabhaḥ 5.41
+%nādaḥ saṃjāyate tasya krameṇābhyāsataś ca vai
+%mattabhṛṅgāvalīvīṇāsadṛśaḥ prathamo dhvaniḥ 5.42
+%evam abhyāsataḥ paścāt saṃsāradhvāntanāśanaḥ
+%ghaṇṭāravasamaḥ paścād dhvanir megharavopamaḥ 5.43
+%dhvanau tasmin mano dattvā yadā tiṣṭhati nirbharam
+%tadā saṃjāyate siddhir layasya mama vallabhā 5.44
+%tatra nāde yadā cittaṃ ramate yogino bhṛśam
+%vismṛtya sakalaṃ bāhyaṃ nādena saha śāmyati 5.45
+%etadabhyāsayogena jitvā samyag guṇān bahūn
+%sarvārambhaparityāgī cidākāśe vilīyate 5.46
+In the \textit{Haṭhayogasaṃhitā} (p. 68), the practice of blocking the ears with the hands is stipulated for \textit{bhrāmarī kumbhaka}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.36}
+\begin{translation}[hp04_036]
+When this inner sound is being cultivated, it drowns out external sound. After a fortnight the yogi overcomes all distraction and becomes happy.
+\end{translation}
+
+%\begin{sources}[hp04_036]
+%\end{sources}
+
+\begin{testimonia}[hp04_036]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{abhyasyamāno nādo 'yaṃ bāhyam āvartayed dhvanim |\\+}
+\tl{paścād vikṣepam akhilaṃ jitvā yogī sukhī bhavet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (f. 124r)
+\begin{versinnote}
+\tl{abhyasyamāno nādo yaṃ bāhyam āvarttayet dhvaniṃ |\\+}
+\tl{pakṣād vikṣepam akhilaṃ jitvā yogī sukhībhavet ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 32
+\begin{versinnote}
+\tl{abhyasyamāno nādo 'yaṃ bāhyam āvṛṇute dhvanim  |\\+}
+\tl{pakṣād vipakṣam akhilaṃ jitvā turyapadaṃ vrajet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_036]
+%\end{philcomm}
+
+\begin{metre}[hp04_036]
+Anuṣṭubh (a: ma-vipulā; c: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.37}
+\begin{translation}[hp04_037]
+In the first stage of practice, a loud sound of various kinds is heard. Then, as the practice progresses, a quieter and quieter sound is heard.
+\end{translation}
+
+%\begin{sources}[hp04_037]
+%\end{sources}
+
+\begin{testimonia}[hp04_037]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{śrūyate prathamābhyāse nādo nānāvidho mahān |\\+}
+\tl{vartamāne tato 'bhyāse śrūyate sūkṣmasūkṣmataḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.9
+\begin{versinnote}
+\tl{śrūyate prathamābhyāse nādo nānāvidho bahuḥ |\\+}
+\tl{vardhamāne tato 'bhyāse śrūyate sūkṣmasūkṣmataḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.31
+\begin{versinnote}
+\tl{śrūyate prathamābhyāse nādo nānāvidho bahuḥ |\\+}
+\tl{vardhamāne tato 'bhyāse śrūyate sūkṣmasūkṣmataḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_037]
+%Adopt sūkṣmasūkṣmataḥ (alpha)[MD: Done.]
+%sūkṣma seems to mean more than just quiet.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.38}
+\begin{translation}[hp04_038]
+In the first stage, [the sounds] are those that are produced by the ocean, a [storm] cloud, a kettle drum and a waterfall.
+In the intermediate stage, they are [the sounds] produced by a bass drum and conch and a bell and trumpet.
+\end{translation}
+
+%\begin{sources}[hp04_038]
+%\end{sources}
+
+\begin{testimonia}[hp04_038]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{ādau jaladhijīmūtabherījharjharasaṃbhavaḥ |\\+}
+\tl{madhye mardalaśaṅkhottha ghaṇṭākāhalakās tathā ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.10
+\begin{versinnote}
+\tl{ādau jaladhijīmūtabherīnirjharasaṃbhavāḥ |\\+}
+\tl{madhye marddalaśaṃkhotthā ghaṇṭākāhalakās tathā ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.32
+\begin{versinnote}
+\tl{ādau jaladhijīmūtabherīnirjharasambhavaḥ |\\+}
+\tl{madhye marddalaśaṃkhotthā ghaṃṭākāhalakās tathā ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_038]
+??
+more refs there are different options for kāhala, such as a trumpet or drum that looks like datura flower.
+See Schmidt, Nachträge: "eine Art Trommel (kāhalāsu dhattūrapuṣpākāramukhabherīṣu), Yaśastilakam (commentary).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.39}
+\begin{translation}[hp04_039]
+In the final stage, there are the sounds of little bells, a bamboo flute, a veena and a bee.
+
+[These] various sound are heard in the body.
+\end{translation}
+
+%\begin{sources}[hp04_039]
+%\end{sources}
+
+\begin{testimonia}[hp04_039]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{anye tu kiṅkiṇīvṛndavīṇābhramaraniḥsvanāḥ |\\+}
+\tl{iti nānāvidho nādaḥ śrūyate dehamadhyagaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.11
+\begin{versinnote}
+\tl{ante tu kiṃkiṇīvṛndavīṇābhramaraniḥsvanāḥ |\\+}
+\tl{iti nānāvidhā nādāḥ śrūyante dehamadhyataḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.33
+\begin{versinnote}
+\tl{ante tu kiṃkiṇī vaṃśanādā bhramaraniḥsvanāḥ |\\+}
+\tl{iti nānāvidhā nādāḥ śrūyante yatra madhyataḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_039]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.40}
+\begin{translation}[hp04_040]
+Even if a loud noise such as that of a [storm] cloud or kettle drum is being heard, the [yogi] should concentrate on only the very quietest sound in it.
+\end{translation}
+
+%\begin{sources}[hp04_040]
+%\end{sources}
+
+\begin{testimonia}[hp04_040]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{mahati śrūyamāṇe 'pi meghabheryādike dhvanau |\\+}
+\tl{tataḥ sūkṣmāt sūkṣmataraṃ nādam eva parāmṛśet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.34
+\begin{versinnote}
+\tl{mahati śrūyamāṇe 'pi meghabheryādike svane |\\+}
+\tl{tatra sūkṣmāt sūkṣmataraṃ nādam eva parāmṛśet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_040]
+%\end{philcomm}
+
+\begin{metre}[hp04_040]
+Anuṣṭubh (c: bha-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.41}
+\begin{translation}[hp04_041]
+Or, the mind should filter out the gross sound for the subtle, or the subtle for the gross, or, abandoning both, be [focused] on [a sound in] the middle, [and the] should not move it elsewhere.%??JM: here gross and subtle do seem to make more sense
+\end{translation}
+
+%\begin{sources}[hp04_041]
+%\end{sources}
+
+\begin{testimonia}[hp04_041]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{ghanam utsṛjya vā sūkṣmaṃ sūkṣmaṃ pramṛjya vā ghanam |\\+}
+\tl{paraṃ tatraiva niḥkṣipya mano nānyatra cālayet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.35
+\begin{versinnote}
+\tl{ghanam utsṛjya vā sūkṣme sūkṣmam utsṛjya vā ghane |\\+}
+\tl{ramamāṇam api kṣipraṃ mano nātra pracālayet ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 37
+\begin{versinnote}
+\tl{ghanam utsṛjya vā sūkṣme sūkṣmam utsṛjya vā ghane |\\+}
+\tl{ramamāṇam api kṣiptaṃ mano nānyatra cālayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_041]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.42}
+\begin{translation}[hp04_042]
+Alternatively, the mind fixes upon whatever sound it first attaches to and dissolves together with it.
+\end{translation}
+
+%\begin{sources}[hp04_042]
+%\end{sources}
+
+\begin{testimonia}[hp04_042]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{yatra kutrāpi vā nāde prathamaṃ viśate manaḥ |\\+}
+\tl{tatraiva susthiraṃ kuryāt tena sārdhaṃ vilīyate ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.36
+\begin{versinnote}
+\tl{yatra kutrāpi vā nāde lagati prathamaṃ manaḥ |\\+}
+\tl{tatraiva susthirībhūtvā tena sārdhaṃ vilīyate ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 37
+\begin{versinnote}
+\tl{yatra kutrāpi vā nāde lagati prathamaṃ manaḥ  |\\+}
+\tl{tatra tatra sthirībhūtvā tena sārdhaṃ vilīyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_042]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.43}
+\begin{translation}[hp04_043]
+Just as a bee drinking nectar has no regard for fragrances, so the mind attached to the inner sound does not desire the objects of the senses.
+\end{translation}
+
+%\begin{sources}[hp04_043]
+%\end{sources}
+
+\begin{testimonia}[hp04_043]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{makarandaṃ piban bhṛṅgo gandhān nāpekṣate yathā |\\+}
+\tl{nādāsaktaṃ tathā cittaṃ viṣayān na hi kāṅkṣati ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.12
+\begin{versinnote}
+\tl{makarandaṃ pibed bhṛṅgo gandho na prekṣyate yathā |\\+}
+\tl{nādāsaktaṃ tathā cittaṃ viṣayān na hi kāṃkṣate ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.41
+\begin{versinnote}
+\tl{makarandaṃ piban bhṛṃgo gandhaṃ nāpekṣate yathā |\\+}
+\tl{nādāsaktaṃ tathā cittaṃ viśayān naiva kāṃkṣati ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 42
+\begin{versinnote}
+\tl{makarandaṃ piban bhṛṅgo gandhān nāpekṣate tathā |\\+}
+\tl{nādāsaktaṃ sadā cittaṃ viṣayaṃ na hi kāṅkṣati ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_043]
+%gandhān is the alpha reading and is possible. [MD: adopted]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.44}
+\begin{translation}[hp04_044]
+When the mercury of the mind is bound and has cast off its fickle nature because it has been assimilated with the sulphur of the internal resonance, it attains the immobility called the unsupported [\textit{samādhi}].
+\end{translation}
+
+%\begin{sources}[hp04_044]
+%\end{sources}
+
+\begin{testimonia}[hp04_044]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{purā matsyendrabodhāya ādināthoditavacaḥ |\\+}
+\tl{manaḥ pākam avāpnoti nirālambākhyaghoṭanam ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.42
+\begin{versinnote}
+\tl{baddhaṃ viyuktaṃ cāpalyaṃ nādagandhakajāraṇāt |\\+}
+\tl{manaḥ pāradam āpnoti nirālambākhyakheṭakam ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_044]
+See Hellwig 2009:204–206 on \emph{khoṭa}, “lame”, which in alchemy is a technical term use to describe mercury that has been processed many times using the \emph{māraṇa} technique and no longer moves.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.44*1}
+\begin{translation}[hp04_044_1]
+Bound by the sulphur of the inner sound, the lord that is the mercury of the mind immediately casts off its fickle nature and attains glory, known as “[the bird] with clipped wings”.
+\end{translation}
+
+%\begin{sources}[hp04_044_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_044_1]
+\end{testimonia}
+
+\begin{philcomm}[hp04_044_1]
+
+Cf.~\emph{Rasendracūḍāmaṇi} 16.52--55
+\begin{versinnote}
+\tl{evaṃ ca pañcamo grāsaḥ pradātavyo'ṣṭamāṃśataḥ |\\+}
+\tl{sa pātrastho 'gnisaṃtapto na gacchati kathañcana ||16.52||\\+}
+\tl{sa \textit{pakṣacchinna} ity uktaḥ sa mukto  'khiladurguṇaiḥ |\\+}
+\tl{so'yaṃ niṣevitaḥ sūtas trimāsaṃ rājikāmitaḥ ||16.53||\\+}
+\tl{viḍaṅgatriphalākṣaudraiḥ khe devaiḥ saha saṅgamam |\\+}
+\tl{ghrāṇamātreṇa \textit{sūtendraḥ} sarvaroganikṛntanaḥ ||16.54||\\+}
+\tl{guṇā ete vinirdiṣṭā rasasya rasavādibhiḥ |\\+}
+\tl{sakalāste guṇāḥ satyā bhairaveṇa prakīrtitāḥ ||16.55||\\!}
+\end{versinnote}
+
+Cf.~also NWS (??ref) s.v.~\textit{pakṣaccheda}.
+
+\end{philcomm}
+
+\begin{metre}[hp04_044_1]
+Anuṣṭubh (c: ma-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.45}
+\begin{translation}[hp04_045]
+As a result of listening to the inner sound, the  snake that is the mind forgets everything and, one-pointed, does not slip away anywhere.
+\end{translation}
+
+%\begin{sources}[hp04_045]
+%\end{sources}
+
+\begin{testimonia}[hp04_045]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{nādaśravaṇataś cittam antaraṅgaturaṅgamaḥ |\\+}
+\tl{viśūnyaṃ sarvam ekāgryaṃ kutra cin na hi dhāvati ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.44
+\begin{versinnote}
+\tl{nādaśravaṇataś cittam antaraṃgakuraṃgakaḥ |\\+}
+\tl{vismṛtya viśvam ekāgraḥ kutra cin na hi dhāvati ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_045]
+The metaphor is that of snake charming: beguiled by the inner sound, the snake that is the mind becomes transfixed.
+Witnesses of the \gamma \ and \delta \ groups have \emph{turaṅgamaḥ} instead of \emph{bhujaṅgamaḥ}, perhaps because forms from \emph{dhāv} are unusual with the latter, but they are found at e.g.~\emph{Padmapurāṇa}, \emph{Garuḍapurāṇa} (refs ??).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.46}
+\begin{translation}[hp04_046]
+This inner sound is a sharpened goad with the power to restrain the bull elephant in must that is the mind as it wanders about in the garden of the sense objects.
+\end{translation}
+
+%\begin{sources}[hp04_046]
+%\end{sources}
+
+\begin{testimonia}[hp04_046]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{manomattagajendrasya viṣayodyānacāriṇaḥ |\\+}
+\tl{niyāmanasamartho 'yaṃ ninādo niśitāṅkuśaḥ ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 44cd–45ab
+\begin{versinnote}
+\tl{manomattagajendrasya viṣayodyānacāriṇaḥ ||\\+}
+\tl{niyāmanasamartho 'yaṃ ninādo niśitāṅkuśaḥ |\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_046]
+The unusual form \emph{niyāmana} is also found in Rasaśāstra works where it occurs in the context of restraining mercury and is a topic of discussion (e.g.~\emph{Rasaprakāśasudhākara} 1.23, \emph{Ānandakanda} 1.4.58–59).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.47}
+\begin{translation}[hp04_047]
+The inner sound is a bolt for the swift horse of the mind so the yogi should regularly concentrate on cultivating it.
+\end{translation}
+% 4
+
+%\begin{sources}[hp04_047]
+%\end{sources}
+
+\begin{testimonia}[hp04_047]
+\emph{Haṭhatattvakaumudī} 54.46
+\begin{versinnote}
+\tl{antaraṃgaturaṃgasya vājinaḥ paridhāvataḥ |\\+}
+\tl{nādopāstikhalīnaṃ hi niyāmanakaraṃ dṛḍham ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_047]
+We understand \textit{ninādaḥ}, which is the subject of the previous line, as the subject in the first line of this verse.
+Several witness have \emph{paridhāyate} instead of \emph{parighāyate}.
+The latter is unattested but could mean "is a halter" based on \emph{abhidhānī}’s meaning of “halter”.
+\end{philcomm}
+
+\begin{metre}[hp04_047]
+Anuṣṭubh (a: na-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.47*1}
+\begin{translation}[hp04_047_1]
+The inner sound is a net for trapping the deer of the mind and a hunter for trapping the antelope of the mind.
+\end{translation}
+% nādo’ntaraṅgasāraṅgabandhane vāgurāyate
+%V15 antaraṅgaturaṅgasya nādo vyādhāyate pi ca
+
+%\begin{sources}[hp04_047_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_047_1]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{nādo 'ntaraṅgasāraṅgabandhane vāgurāyate ||\\+}
+\tl{antaraṅgaturaṅgasya bandhane līyate 'pi ca ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 47
+\begin{versinnote}
+\tl{nādo'ntaraṃgasāraṃgabandhane vāgurāyate |\\+}
+\tl{antaraṃgakuraṃgasya nādo vyādhāyate 'pi ca  ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_047_1]
+Adopt V15 67cd
+
+??
+MD: rodhe may be correct.
+In a drive hunt, the animals are driven in a certain direction (often in a confined space)  by means of loud noises, etc., so that another hunter can easily shoot them.
+JM: yes, adopt, I have altered tr.
+to "trapping"
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.48}
+\begin{translation}[hp04_048]
+Striking the deer of the mind when it is transfixed by its attachment to inner sounds such as that of a bell, is very easy if the archer is skilful.
+\end{translation}
+% ghaṇṭādinādasaktastabdhāntaḥkaraṇahariṇasya |
+% praharaṇam atisukaraṃ syāc charasaṃdhātā pravīṇaś cet ||
+%
+
+%\begin{sources}[hp04_048]
+%\end{sources}
+
+%\begin{testimonia}[hp04_048]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_048]
+%N3 has a problem with the metre.
+%\end{philcomm}
+
+\begin{metre}[hp04_048]
+Upagīti 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.49}
+\begin{translation}[hp04_049]
+When the mind dissolves into that which is the most subtle object of perception in the unstruck sound, that is the supreme state of Viṣṇu.
+
+The tone of that sound is that of the unstruck sound.
+The object of perception is inside the sound [and] the mind is inside the object of perception.
+When the mind dissolves [in it], that is the supreme state of Viṣṇu.
+\end{translation}
+
+% anāhatadhvaner antar jñeyaṃ yat sūkṣmasūkṣmakaṃ/
+% manas tatra layaṃ yāti tad viṣṇoḥ paramaṃ padam//
+
+\begin{sources}[hp04_049]
+Uttaragītā 41cd–42 (source??)
+\begin{versinnote}
+\tl{anāhatasya śabdasya tasya śabdasya yo dhvaniḥ ||\\+}
+\tl{dhvaner antargataṃ jyotir jyotirantargataṃ manaḥ |\\+}
+\tl{tan mano vilayaṃ yāti tad viṣṇoḥ paramaṃ padam  ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_049]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{anāhatasya śabdasya tasya śabdasya yo dhvaniḥ ||\\+}
+\tl{dhvaner antargataṃ jyotir jyotirantargataṃ manaḥ |\\+}
+\tl{yan mano vilayaṃ yāti tad viṣṇoḥ paramaṃ padam ||\\!}
+\end{versinnote}
+
+Haṭhayogasaṃhitā p.~68
+\begin{versinnote}
+\tl{anāhatasya śabdasya tasya śabdasya yo dhvaniḥ |\\+}
+\tl{dhvaner antargataṃ jyotir jyotiṣo 'ntargataṃ manaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_049]
+%??
+to discuss.
+It appears that the following lines were added (in Gr2 and Gr3) as a note or supplement to 4.68.
+In particular, the hemistich anāhatasya śabdasya tasya...
+seems to comment on anāhatadhvaner.
+\begin{versinnote}
+\tl{anāhatasya śabdasya tasya śabdasya yo dhvaniḥ ||\\+}
+\tl{dhvaner antargataṃ jyotir jyotirantargataṃ manaḥ |\\+}
+\tl{tanmano vilayaṃ yāti tad viṣṇoḥ paramaṃ padam  || \\!}
+\end{versinnote}
+The source of them may be the Uttaragītā. However, in NGMPP E 2098-11 Uttaragītā ms these three lines are omitted from Kṛṣṇa's words, which start with oṃkāra.
+%HI: tan could mean tatra.
+%Looks like the last line was rewritten to include tatra.
+
+%?? MD: I adopted the version transmitted in Gr4c and J10 (anāhatadhvaner antar jñeyaṃ yat sūkṣmasūkṣmakaṃ ...).
+%?? Which is the more likely reading of the second line for the alternative version?
+%A - dhvaner antargataṃ jyotir jyotirantargataṃ manaḥ (J5,G4,Gr2,Gr3)
+%B - dhvaner antargataṃ jñeyaṃ jñeyasyāntargataṃ manaḥ (N3?,Jyo) JM: the second I think
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.50}
+\begin{translation}[hp04_050]
+As long as sound exists, there is a concept of space. That which is soundless is the supreme Brahman and is called the supreme self.
+\end{translation}
+% adopt samīryate J7, V3
+
+%\begin{sources}[hp04_050]
+%\end{sources}
+
+\begin{testimonia}[hp04_050]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{tāvad ākāśasaṃkalpo yāvac chabdaḥ pravartate |\\+}
+\tl{niḥśabdaṃ tatparaṃ brahma paramātmā samīryate ||\\!}
+\end{versinnote}
+
+Six-chapter \emph{Vivekamārtaṇḍa} 5.15
+\begin{versinnote}
+\tl{tāvad ākāśasaṅkalpo yāvac chabdaḥ pravartate |\\+}
+\tl{niḥśabdaṃ tat parabrahma paramātmā sa gīyate || 15 ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 47cd–48ab
+\begin{versinnote}
+\tl{tāvadākāśasaṅkalpo yāvacchabdaḥ pravatate |\\+}
+\tl{niḥśabdaṃ tatparaṃ brahma paramātmā samīryate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_050]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.51}
+\begin{translation}[hp04_051]
+Whatever is heard as the inner sound is nothing but Śakti. The formless one which hears it is nothing but the supreme lord.
+\end{translation}
+% adopt V3 reading nādarupeṇa
+
+%\begin{sources}[hp04_051]
+%\end{sources}
+
+\begin{testimonia}[hp04_051]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{yat kiñ cin nāmarūpeṇa śrūyate śaktir eva sā |\\+}
+\tl{yas tacchrotā nirākāraḥ sa eva parameśvaraḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~235r–235v
+\begin{versinnote}
+\tl{yat kiṃ cin nāmarūpeṇa śrūyate śaktir eva sā |\\+}
+\tl{yasya śrottā nirākaraḥ sa eva parameśvaraḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_051]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.52}
+\begin{translation}[hp04_052]
+Blocking of the ears, mouth, eyes, and nose should not be performed [when] a clear inner sound is heard distinctly in the purified Suṣumṇā channel.%?? negative in first line? Adopt naiva?
+\end{translation}
+
+%\begin{sources}[hp04_052]
+%\end{sources}
+
+\begin{testimonia}[hp04_052]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{śravaṇapuṭanayananāsāpuṭarodhanaṃ kāryam |\\+}
+\tl{śrīśuddhasuṣumṇāsaraṇau sphuṭam amalaḥ śrūyate nādaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} 123v–124r
+\begin{versinnote}
+\tl{haṭhapradīpikāyāṃ \\+}
+\tl{śravaṇapuṭanayanayugulanāsāmukharodham eva kartavyaṃ |\\+}
+\tl{śuddhasaṣumṇāśaraṇe sphuṭam amalaḥ śrūyate nādaḥ ||\\!}
+\end{versinnote}
+
+\emph{Saubhāgyalakṣmyupaniṣad} 4
+\begin{versinnote}
+\tl{śravaṇamukhanayananāsānirodhanenaiva kartavyam |\\+}
+\tl{śuddhasuṣumnāsaraṇau sphuṭam amalaṃ śrūyate nādaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_052]
+%??N3 is missing an akṣara (nirodhaṃ aiva)
+%??J5 has nirodhaṃ naiva
+??Adopt alpha reading mukha°, naiva. HI: it is strongly attested (groups 4c, C6, etc) and more likely explains the change to °saṃrodhanaṃ kāryam
+\end{philcomm}
+
+\begin{metre}[hp04_052]
+Upagīti 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.52*1}
+\begin{translation}[hp04_052_1]
+Nāda is called Śakti. And gnosis of nāda is Sadāśiva. When gnosis of nāda has disappeared, only unmanī remains.
+\end{translation}
+
+%\begin{sources}[hp04_052_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_052_1]
+\emph{Haṭhatattvakaumudī} 54.50
+\begin{versinnote}
+\tl{nādaḥ śaktir iti khyāto nādajñānaṃ sadāśivaḥ |\\+}
+\tl{nādajñāne vinaṣṭe ca tad unmany eva śiṣyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_052_1]
+%??V15 reading for the third pāda is better than N19.
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.52*2}
+\begin{translation}[hp04_052_2]
+As long as there is the inner sound there is mind. At the end of the inner sound the mind beyond mind state [arises]. The void is said to be sonorous and Brahma is silent.
+\end{translation}
+
+%\begin{sources}[hp04_052_2]
+%\end{sources}
+
+\begin{testimonia}[hp04_052_2]
+\emph{Haṭhatattvakaumudī} 54.51
+\begin{versinnote}
+\tl{nādo yāvan manas tāvan nādānte ca manonmanī |\\+}
+\tl{saśabdaṃ kathitaṃ vāte niḥśabdaṃ brahma kathyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_052_2]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.52*3}
+\begin{translation}[hp04_052_3]
+When the accumulated subliminal impressions have been destroyed as a result of the continuous fusing of the mind with the internal resonance, the mind and breath are sure to dissolve into the unadorned [god] (\emph{nirañjane}).
+\end{translation}
+
+%\begin{sources}[hp04_052_3]
+%\end{sources}
+
+\begin{testimonia}[hp04_052_3]
+\emph{Upāsanāsārasaṅgraha} ??
+\begin{versinnote}
+\tl{sadā nādānusandhānāt saṃkṣīṇe vāsanā bhavet |\\+}
+\tl{nirañjane vilīneti niścalitaṃ mārutaṃ manaḥ ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 49
+\begin{versinnote}
+\tl{saśabdaś cākṣare kṣīṇe niḥśabdaṃ paramaṃ padam |\\+}
+\tl{sadā nādānusandhānāt saṃkṣīṇā vāsanā bhavet ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.52
+\begin{versinnote}
+\tl{sadā nādānusandhānāt saṃkṣīṇe vāsanākṣaye |\\+}
+\tl{nirañjane ca līyate niścitaṃ cittamārutau ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_052_3]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.52*4}
+\begin{translation}[hp04_052_4]
+Thousands of crores of inner sounds and hundreds of crores of visual focal points all dissolve into the place where the unadorned god (\emph{devo nirañjana}) is.
+
+%There are thousands of Crores of resonances and thousands of crores of sonic matter (bindu). All dissolve into where the spotless god exists.
+%?? tr. the spotless god, Nirañjana. ? and in previous
+\end{translation}
+
+%\begin{sources}[hp04_052_4]
+%\end{sources}
+
+\begin{testimonia}[hp04_052_4]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{nādakoṭisahasrāṇi bindukoṭiśatāni ca |\\+}
+\tl{sarve tatra layaṃ yānti yatra devo nirañjana iti ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.53
+\begin{versinnote}
+\tl{nādakoṭisahasrāṇi bindukoṭiśatāni ca |\\+}
+\tl{sarve tatra layaṃ yānti yatra devo nirañjanaḥ ||\\!}
+\end{versinnote}
+
+\emph{Nādabindūpaniṣat} 50cd--51ab
+\begin{versinnote}
+\tl{nādakoṭisahasrāṇi bindukoṭiśatāni ca ||\\+}
+\tl{sarve tatra layaṃ yānti brahmapraṇavanādake |\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_052_4]
+The expression \textit{devo nirañjanaḥ} (`unadorned god') refers to Śiva.
+
+Compare the quotation in the Śabdakalpadruma, s.v.~\emph{dharmaghaṭa}:
+\begin{quote}
+“\dots ante yāti paraṃ sthānaṃ yatra devo nirañjanaḥ||"
+iti bhaviṣyapurāṇoktā dharmaghaṭavratakathā samāptā ||
+\end{quote}
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.52*5}
+%\begin{translation}[hp04_052_5]
+%\end{translation}
+
+%\begin{sources}[hp04_052_5]
+%\end{sources}
+
+%\begin{testimonia}[hp04_052_5]
+%\end{testimonia}
+
+\begin{philcomm}[hp04_052_5]
+See 4.74.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.53}
+\begin{translation}[hp04_053]
+Enough of the verbosity of a learned gathering! O friend, listen to this instruction formerly taught by Śiva for awakening Matsyendra.
+
+\end{translation}
+
+%\begin{sources}[hp04_053]
+%\end{sources}
+
+\begin{testimonia}[hp04_053]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{kāṣṭhagoṣṭhīprasaṅgena nādam antargataṃ śṛṇu |\\+}
+\tl{purā matsyendrabodhāya ādināthoditaṃ vacaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_053]
+This verse is likely to be authorial, composed to introduce next two verses, which are from the \emph{Candrāvalokana}, a dialogue between Matsyendra and Śiva.
+1.34 is another probable authorial verse and, like this one, it has the vocative \emph{sakhe}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.54}
+\begin{translation}[hp04_054]
+As long as the moving breath does not enter the middle path, as long as bindu is not stable, restrained by the breath, as long as the ultimate truth (\emph{tattvam}) does not become as natural as the sky, then all that one says is deceitful, and false prattle.
+\end{translation}
+
+\begin{sources}[hp04_054]
+\emph{Candrāvalokana} 14
+\begin{versinnote}
+\tl{yāvan naiva praviśati caran māruto madhyamārge\\+}
+\tl{yāvad bindur na bhavati dṛḍhaḥ prāṇavātaprabandhaḥ |\\+}
+\tl{yāvad vyomnā sadṛśa sarasaṃ jāyate nonmanatvaṃ {[}em.~vyomnā sadṛśam arasaṃ?]\\+}
+\tl{tāvat sarvaṃ yadi ca vadate dambhamithyāpralāpaḥ ||\\+}
+\tl{\var{14b prāṇavātaprabandhaḥ ] 4345 : prāṇaghātaprabuddhaḥ 75278 : prāṇavāyuḥ prabuddhaḥ 7970 : prāṇavātaprabuddhaḥ T00788
+14d sarvam] 7970, T00788, 75278 : vighnaṃ 4345}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_054]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{haṭhapradīpikāyāṃ—\\+}
+\tl{yāvan naiva praviśati caran māruto madhyamārge\\+}
+\tl{yāvad bindur na bhavati dṛḍhaḥ prāṇavātaprabaddhaḥ |\\+}
+\tl{yāvad vyomnaḥ sahajasadṛśaṃ jñāyate naiva tatvam \\+}
+\tl{tāvat sarvaṃ vadati yad idaṃ dambhamithyāpralāpa iti\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 2.2
+\begin{versinnote}
+\tl{yāvan naiva praviśati caran māruto madhyamārgaṃ\\+}
+\tl{yāvat sūkṣmo na bhavati dṛḍhaḥ prāṇavātaprabandhaḥ |\\+}
+\tl{yāvad vyomnā sahajasadṛśaṃ jāyate naiva cittaṃ\\+}
+\tl{tāvat sarvaṃ vadati tad idaṃ dambhamithyāpralāpaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_054]
+%J5 has sthiraḥ (instead of dṛḍhaḥ) supporting the visarga (rather than dṛdhaṃ N3)
+%tattva has been understood here (and in the verse with bīja) as the mind.
+%\end{philcomm}
+
+\begin{metre}[hp04_054]
+Mandākrāntā 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.55}
+\begin{translation}[hp04_055]
+Having learnt the correct piercing of Suṣumṇā, [the yogi] should make the breath go into the central channel, take it to the place of the moon and stop [the flow] in the nostrils.
+\end{translation}
+%nītvāsāv aindavasthāne
+
+\begin{sources}[hp04_055]
+\emph{Candrāvalokana} 32
+\begin{versinnote}
+\tl{jñātvā suṣumnāsadbhedaṃ kṛtvā vāyuṃ ca madhyagam |\\+}
+\tl{kṛtvāsāv aindave sthāne ghrāṇarandhre nirodhayet ||\\+}
+\tl{\var{satbhedaṃ ] 4345, 4340, T00788 : tatbhedaṃ  7970.}\\+}
+\tl{\var{kṛtvāsāv aindave ] 7970, 4340, T00788 : kṛtvā*d*baindave 4345}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_055]
+\emph{Upāsanāsārasaṅgraha} p.~31 (attributed to the \emph{Candrāvalokana})
+\begin{versinnote}
+\tl{jñatvā suṣumnāṃ tadbhedaṃ kṛtvā vāyuṃ ca madhyagam |\\+}
+\tl{kṛtvāsau baindavasthāne ghraṇarandhre nirodhayet ||\\!}
+\end{versinnote}
+
+\emph{Yogakuṇḍalinyupaniṣat} 7cd–8ab
+\begin{versinnote}
+\tl{jñātvā suṣumnāṃ tadbhedaṃ kṛtvā vāyuṃ ca madhyagam ||\\+}
+\tl{sthitvāsau baindavasthāne ghrāṇarandhre nirodhayet |\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~107v–108r
+\begin{versinnote}
+\tl{jñātvā suṣumnāsadbhedaṃ kṛtvā vāyuṃ ca madhyagam |\\+}
+\tl{sthitvā sadaiva svasthena prāṇarandhraṃ nirodhayet ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_055]
+%??ghrāṇarandhra is too well supported to ignore, but the meaning is difficult to construe. JM: it's in alpha etc., why not adopt it? vāyum is object in pādas bcd.
+%\end{philcomm}
+
+\begin{metre}[hp04_055]
+Anuṣṭubh (a: ma-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.56}
+\begin{translation}[hp04_056]
+The moon and sun move in iḍā and piṅgalā. The moon is said to be tamasic and the sun rajasic.
+\end{translation}
+
+\begin{sources}[hp04_056]
+\emph{Vasiṣṭhasaṃhitā} 2.28–29ab
+\begin{versinnote}
+\tl{iḍāyāṃ piṅgalāyāṃ ca carataś candrabhāskarau |\\+}
+\tl{iḍāyāṃ candramā jñeyaḥ piṅgalāyāṃ raviḥ smṛtaḥ ||\\+}
+\tl{candras tāmasa ity uktaḥ sūryo rājasa ucyate |\\!}
+\end{versinnote}
+
+\emph{Yogayājñavalkya} 4.32cd–33
+\begin{versinnote}
+\tl{iḍāyāṃ piṅgalāyāṃ ca carataś candrabhāskarau ||\\+}
+\tl{iḍāyāṃ candramā jñeyaḥ piṅglāyāṃ raviḥ smṛtaḥ |\\+}
+\tl{candras tāmasa ity uktaḥ sūryo rājasa ucyate ||\\!}
+\end{versinnote}
+
+\emph{Matsyendrasaṃhitā} 4.41cd
+\begin{versinnote}
+\tl{iḍāyāṃ piṅgalāyāṃ ca parataś candrabhāskarau ||\\!} 
+\end{versinnote}
+
+\end{sources}
+
+\begin{testimonia}[hp04_056]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{iḍāyāṃ piṃgalāyāṃ ca carataś candrabhāskarau |\\+}
+\tl{iḍāyāṃ candramā jñeyaḥ piṃgalāyāṃ raviḥ smṛtaḥ ||\\+}
+\tl{candras tāmasa ity uktas sūryo rājasa ucyate |\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_056]
+%??JM: rajasic and tamasic not in Eng dictionary, "associated with the \emph{rajas guṇa}" ?
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.57}
+\begin{translation}[hp04_057]
+Those two bring about the entirety of time, which consists of night and day. Suṣumṇā consumes time. This which has been taught is secret.
+\end{translation}
+
+\begin{sources}[hp04_057]
+Vasiṣṭhasaṃhitā 2.29cd–30ab
+\begin{versinnote}
+\tl{tāv eva sakalaṃ dhattaḥ kālaṃ rātrindivātmakam |\\+}
+\tl{bhoktrī suṣumṇā kālasya guhyam etad udāhṛtam ||\\!}
+\end{versinnote}
+
+Yogayājñavalkya 4.34cd–35ab
+\begin{versinnote}
+\tl{tāv eva dhattaḥ sakalaṃ kālaṃ rātridivātmakam |\\+}
+\tl{bhoktrī suṣumnā kālasya guhyam etad udāhṛtam ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_057]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{tāv eva dhattas sakalaṃ kālaṃ rātriṃ divātmakam |\\+}
+\tl{bhoktrī suṣumṇā kālasya guhyam etad udāhṛtam ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} (check in ms. no. 2244)
+\begin{versinnote}
+\tl{tathā coktaṃ haṭhapradīkāyāṃ–\\+}
+\tl{sūryācandramasau dhattaḥ lagnaḥ kāla rātridinātmakaṃ ||\\+}
+\tl{bhoktrī suṣumnā kālasya guhyate tad udīritaṃ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_057]
+\end{philcomm}
+
+\begin{metre}[hp04_057]
+Anuṣṭubh (a: bha-vipulā; c: ma-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.58}
+\begin{translation}[hp04_058]
+For as the tetrad of verses called the Saubhadra has it:
+
+There are six cakras, sixteen supports, three focal points and three guṇas. Everything else is [just] the prolixity of texts. Trikūṭa is the supreme place.
+\end{translation}
+
+%\begin{sources}[hp04_058]
+%\end{sources}
+
+\begin{testimonia}[hp04_058]
+Cf.~6-chapter \emph{Vivekamārtaṇḍa} 6.3
+\begin{versinnote}
+\tl{ṣaṭcakraṃ ṣoḍaśādhāraṃ trilakṣaṃ vyomapañcakam |\\+}
+\tl{svadehe ye na jānanti kathaṃ sidhyanti yoginaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_058]
+We do not know why this tetrad of verses is called Saubhadra.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.59}
+\begin{translation}[hp04_059]
+Kuṇḍalinī is taught to have a curved shape like a snake. She is Śakti. [The yogi] who has made her move is undoubtedly liberated.
+\end{translation}
+
+%\begin{sources}[hp04_059]
+%\end{sources}
+
+%\begin{testimonia}[hp04_059]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_059]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.60}
+\begin{translation}[hp04_060]
+When kūṭa is situated at trikūṭa [then] the mind is wonderful and uninterrupted. By means of kuṇḍalinī, [the yogi] is undoubtedly liberated.
+\end{translation}
+
+%\begin{sources}[hp04_060]
+%\end{sources}
+
+\begin{testimonia}[hp04_060]
+?? Upāsanāsārasaṅgraha !!!
+\end{testimonia}
+
+\begin{philcomm}[hp04_060]
+We are unsure of the referent of \emph{kūṭa} here.
+As suggested to us by Sven Sellmer, it may mean the tip of the tongue, which in e.g.~\emph{Khecarīvidyā} 1.65–67 and 3.16–17, is to be placed at \emph{trikūṭa} as part of the practice of \emph{khecarīmudrā}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.61}
+\begin{translation}[hp04_061]
+There are seventy-two thousand openings  of the channels in the cage [that is the body]. Suṣumṇā is the Śāmbhavī Śakti while the other [channels] are pointless.
+\end{translation}
+
+%\begin{sources}[hp04_061]
+%\end{sources}
+
+\begin{testimonia}[hp04_061]
+\emph{Haṭhasaṅketacandrikā} f.~108r
+\begin{versinnote}
+\tl{sūryā[c]andram asau kṛtvā viditvā karapaṃjare [|] \\+}
+\tl{suṣumṇā śāṃbhavī śaktiḥ śeṣās tv eva nirarthakāḥ ||\\!}
+\end{versinnote}
+
+Yogaśikhopaniṣat 6.17cd–18ab
+\begin{versinnote}
+\tl{dvisaptatisahasrāṇi nāḍīdvārāṇi pañjare ||\\+}
+\tl{suṣumnā śāmbhavī śaktiḥ śeṣās tv anye nirarthakāḥ |\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_061]
+The compound \emph{nāḍīdvāra} is not found elsewhere (other than as \emph{nāḍīdvāreṇa}) and its meaning here is unclear.
+Brahmānanda understands \emph{dvārāṇi} to refer to routes by which breath enters the body (\emph{dvārāṇi vāyupraveśamārgāḥ}) and we have translated \emph{nāḍīdvārāṇi} accordingly.
+%Consider °nāḍīnāṃ dehapañjare ??JM: probably a patch?
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.62}
+\begin{translation}[hp04_062]
+The breath, having been carefully accumulated, together with fire awakens kuṇḍalinī and enters Suṣumṇā without obstruction.
+\end{translation}
+
+\begin{sources}[hp04_062]
+\emph{Dattātreyayogaśāstra} 108
+\begin{versinnote}
+\tl{vāyuḥ paricito yatnād agninā saha kuṇḍalīm |\\+}
+\tl{bodhayitvā suṣumnāyāṃ praviśed avirodhataḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_062]
+\emph{Śārṅgadharapaddhati} 4399
+\begin{versinnote}
+\tl{vāyuḥ paricito yasmād agninā saha kuṇḍalīm  |\\+}
+\tl{bodhayitvā suṣumnāyāṃ praviśed anirodhataḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} ff.~197v–180r (attributed to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{vāyuḥ paricito yasmād agninā saha kuṃḍalī |\\+}
+\tl{bodhayitvā suṣumnāyāṃ praviśed anirodhata iti\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_062]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.63}
+\begin{translation}[hp04_063]
+When the breath is flowing in Suṣumṇā, the no-mind state is sure to be attained.
+Otherwise [i.e.~if the breath is not flowing in Suṣumṇā] the various practices [of yoga] lead to nothing but exertion for yogis.
+\end{translation}
+
+%\begin{sources}[hp04_063]
+%\end{sources}
+
+\begin{testimonia}[hp04_063]
+\emph{Upāsanāsārasaṅgraha} p.~108
+\begin{versinnote}
+\tl{haṭhapradīpikāyām–\\+}
+\tl{suṣumnāvāhinī prāṇe siddhaty eva manonmani |\\+}
+\tl{anyathā vividhābhyāso prayāsāyaiva yoginām ||\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~113v
+\begin{versinnote}
+\tl{suṣumṇāvāhini prāṇe sidhyaty eva manonmanī  |\\+}
+\tl{anye ye vividhābhyāsāḥ prayāsā eva yogināṃ  ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_063]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.64}
+\begin{translation}[hp04_064]
+The mind is bound by the very same thing that binds the breath and the breath is bound by that which binds the mind.
+\end{translation}
+
+%\begin{sources}[hp04_064]
+%\end{sources}
+
+\begin{testimonia}[hp04_064]
+
+\emph{Haṭhasaṅketacandrikā} f.~67r
+\begin{versinnote}
+\tl{tathā coktaṃ haṭhapradīpikāyāṃ–\\+}
+\tl{pavano badhyate yena manas tenaiva b[a]dhyate \\+}
+\tl{mānaś ca badhyate yena pavanas tena badhyate [|]\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_064]
+Brahmānanda understands \emph{yena} here to refer to the yogi. We have taken it to refer to a practice.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.65}
+\begin{translation}[hp04_065]
+The mind has two impulses: past impression (\emph{vāsanā}) and the breath. When one of those two disappears, both soon disappear.%??JM: tr. vāsanā as plural? Any better ideas?
+\end{translation}
+
+\begin{sources}[hp04_065]
+\emph{Gorakṣaśataka} 9
+\begin{versinnote}
+\tl{hetudvayañ ca cittasya vāsanā ca samīraṇaḥ |\\+}
+\tl{tayor vinaṣṭa ekasmin drutaṃ dvāv api naśyataḥ |\\+}
+\tl{\var{9d drutaṃ dvāv api ] em.; dhṛtaṃ dvāv api T, tasmai dvāv api G1, nasmai dvāv api G2, tad dvāv api vi° U}\\!}
+\end{versinnote}
+
+\emph{Mokṣopāya} V.92.48
+\begin{versinnote}
+\tl{dve bīje rāma cittasya prāṇaspandanavāsane |\\+}
+\tl{ekasmiṃś ca tayoẖ kṣīṇe kṣipraṃ dve api naśyataḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_065]
+\emph{Yogakuṇḍalinyupaniṣat} 1
+\begin{versinnote}
+\tl{hetudvayaṃ hi cittasya vāsanā ca samīraṇaḥ |\\+}
+\tl{tayor vinaṣṭa ekasmiṃs tad dvāv api vinaśyataḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_065]
+It is difficult to say whether the \emph{Gorakṣaśataka}'s reading \emph{cittasya} was changed to \emph{manaso} by the author when this verse was borrowed in order to make the terminology consistent with the previous verse, or whether this change occurred at a later time.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.66}
+\begin{translation}[hp04_066]
+The breath dissolves where the mind dissolves; the mind dissolves right where the breath dissolves.
+\end{translation}
+
+%\begin{sources}[hp04_066]
+%\end{sources}
+
+\begin{testimonia}[hp04_066]
+\emph{Haṭharatnāvalī} 4.29
+\begin{versinnote}
+\tl{mano yatra vilīyeta pavanas tatra līyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_066]
+%JM?? do we need this note? I think it's about something other than samādhi.
+These verses seem to elevate the possible role of the mind in achieving \emph{samādhi} (as opposed to the standard Haṭha method of stopping the breath to achieve samādhi).
+
+%Brahmānanda understands this verse as referring to specific locations into which the breath and mind dissolve, rather tha"when... then...".
+%HI: tatra = tāvat.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.67}
+\begin{translation}[hp04_067]
+ Like water mixed with milk, mind and breath are always joined, behaving in the same way.  As long as there is mind the breath is active, and as long as there is breath the mind is active.
+\end{translation}
+
+\begin{sources}[hp04_067]
+\emph{Amanaska} 2.27
+\begin{versinnote}
+\tl{dugdhāṃbuvat sammilitau sadaiva\\+}
+\tl{tulyakriyau mānasamārutau ca |\\+}
+\tl{yāvan manas tatra marutpravṛttir \\+}
+\tl{yāvan maruc cāpi manaḥpravṛttiḥ |\\+}
+\tl{\var{sadaiva ] NI, S, N : tathaiva Cc : sad eva Na}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_067]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{rājayoge—\\+}
+\tl{dugdhāmbuvat saṃmilitau sadaiva\\+}
+\tl{tulyakriyau mānasamārutau ca |\\+}
+\tl{yāvan manas tatra marutpravṛttir \\+}
+\tl{yāvan marut tatra manaḥpravṛttiḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 2.5
+\begin{versinnote}
+\tl{dugdhāmbuvat saṃmilitau sadaiva \\+}
+\tl{tulyakriyau mānasamārutau ca | \\+}
+\tl{yāvan manas tatra marutpravṛttis \\+}
+\tl{tatraikanāśād aparasya nāśaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_067]
+Complementing his understanding of the previous verse, Brahmānanda has \emph{yato\dots tatra} in \emph{pāda}s c and d, taking it to mean \emph{yatra\dots tatra} and to be referring to cakras.
+\end{philcomm}
+
+\begin{metre}[hp04_067]
+Upajāti
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.68}
+\begin{translation}[hp04_068]
+As a result of one of those two disappearing the other disappears and as a result of one being active the other is active.
+And when neither has disappeared there is perception through all the sense faculties.
+When both have disappeared the state of liberation is attained.
+\end{translation}
+
+\begin{sources}[hp04_068]
+\emph{Amanaska} 2.28
+\begin{versinnote}
+\tl{tatraikanāśād aparasya nāśa\\+}
+\tl{ekapravṛtter aparapravṛttiḥ |\\+}
+\tl{adhvastayoś cendriyavargabuddhir \\+}
+\tl{vidhvastayor mokṣapadasya siddhiḥ ||\\+}
+\tl{\var{vargabuddhir ] NI and S : vargavṛttir Pa Tr Va Nb Ea Eb : vargavṛddhiḥ VbVd: vargavṛddhir N : vargavidhi Pc : sargabuddhir Cc: sargaviddhir Nu}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_068]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{tatraikanāśād aparasya nāśaḥ\\+}
+\tl{ekapravṛtter aparapravṛttiḥ |\\+}
+\tl{adhvastayoḥ svendriyavargavṛddhir\\+}
+\tl{vidhvastayor mokṣapadasya siddhiḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 2.6
+\begin{versinnote}
+\tl{ekapravṛttāv aparapravṛttir \\+}
+\tl{ekasya nāśād aparasya nāśaḥ |\\+}
+\tl{adhvastayor indriyavargavṛttir \\+}
+\tl{vidhvastayoḥ mokṣapadasya siddhiḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_068]
+%Destroy etc. for disappear?
+%J5 śuddhir (for buddhir)
+%\end{philcomm}
+
+\begin{metre}[hp04_068]
+Upajāti
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.69}
+\begin{translation}[hp04_069]
+[The yogi becomes] a traveller by the way of the wind [and] obtains the whole earth and the eight lordly powers. True, true is this, O beautiful one.
+\end{translation}
+
+\begin{sources}[hp04_069]
+\emph{Jñānasāra} 3.5–6
+\begin{versinnote}
+\tl{vāyuvegena deveśi sakalāṃ bhramate mahīm |\\+}
+\tl{aṣṭadhāguṇam aiśvaryaṃ satyaṃ satyaṃ na cānyathā ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_069]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_069]
+%Unidentified source. Jnānasāra removes vocative.
+%vāyumārga is an expression for flying (Rāmāyaṇa 7.33.3, etc., etc.)
+%labhate needed for aṣṭaguṇaṃ
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.70}
+\begin{translation}[hp04_070]
+Thus Viśvarūpācārya:\\
+When the breath is destroyed and the mind dissolves, all experience is the same. That is called Samādhi.
+\end{translation}
+
+\begin{sources}[hp04_070]
+\emph{Vivekamārtaṇḍa} 163
+\begin{versinnote}
+\tl{yadā saṃkṣīyate prāṇo mānasaṃ ca vilīyate |\\+}
+\tl{tadā samarasatvaṃ ca samādhiḥ so 'bhidhīyate ||\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_070]
+\emph{Yuktabhavadeva} 30
+\begin{versinnote}
+\tl{yadā saṃkṣīyate prāṇo mānasaṃ ca pralīyate |\\+}
+\tl{yadā samarasatvaṃ ca samādhiḥ procyate tadā ||\\+}
+\tl{\var{yadā ] tadā}\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_070]
+The six-chapter \emph{Vivekamārtaṇḍa} is attributed to Viśvarūpa, which perhaps explains the attribution of this verse to him.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.71}
+\begin{translation}[hp04_071]
+When the mind is still the breath is still, from which bindu becomes still.  As a result of bindu becoming still, my son, the body becomes still.
+\end{translation}
+
+%\begin{sources}[hp04_071]
+%\end{sources}
+
+\begin{testimonia}[hp04_071]
+\emph{Yogacintāmaṇi}
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{manaḥsthairye sthiro vāyus tato binduḥ sthiro bhavet |\\+}
+\tl{bindusthairyād athāpannaṃ piṇḍasthairyaṃ prajāyata iti ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 43.19
+\begin{versinnote}
+\tl{cittasthairye mārutasusthiraḥ syāt
+tasmād bindususthiro yogino 'ṃge~|\\+}
+\tl{bindusthairye syād dayā satvam ojaḥ
+piṇḍasthairyaṃ kāyasampad balaṃ ca || 19 ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_071]
+%88c āpanna not well attested and N3 has bindusthairyodayāt putra: perhaps this comes from an unknown source text addressed to a son.
+%?? putra
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.72}
+\begin{translation}[hp04_072]
+Only he whose gaze is steady without a visible object, whose breath is steady without effort [and] whose mind is steady without a support is a yogi, is a guru, is to be served.
+\end{translation}
+
+\begin{sources}[hp04_072]
+\emph{Amanaska} 2.44
+\begin{versinnote}
+\tl{dṛṣṭiḥ sthirā yasya vinaiva dṛśyād\\+}
+\tl{vāyuḥ sthiro yasya vinā prayatnāt |\\+}
+\tl{cittaṃ sthiraṃ yasya vināvalambāt\\+}
+\tl{sa eva yogī sa guruḥ sa sevyaḥ ||\\+}
+\tl{\var{vinaiva dṛśyād  AllNI AllSI Eb: vinā sudṛśyād Ua: vinā nimeṣād Pb AllN : vinaiva lakṣyād Cc: vinaiva dṛṣṭyād Jb: vinakadṛśyad Ma: vinānaṅgame Nb}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_072]
+\emph{Yogacintāmaṇi} (attributed to the \emph{Rājayoga})
+\begin{versinnote}
+\tl{dṛṣṭiḥ sthirā yasya vinaiva dṛśyaṃ\\+}
+\tl{vāyuḥ sthiro yasya vinā prayatnaṃ |\\+}
+\tl{cittaṃ sthiraṃ yasya vināvalaṃbaṃ\\+}
+\tl{sa eva yogī sa guruḥ saṃsevyaḥ ||\\!}
+\end{versinnote}
+
+\emph{Kulārṇavatantra} 13.70
+\begin{versinnote}
+\tl{dṛśyaṃ vinā sthirā dṛṣṭir manaś cālambanaṃ  vinā |\\+}
+\tl{vināyāsaṃ sthiro vāyur yasya syāt sa guruḥ priye ||\\!}
+\end{versinnote}
+
+\emph{Haṭharatnāvalī} 4.25
+\begin{versinnote}
+\tl{dṛṣṭiḥ sthirā yasya vinaiva lakṣyāt\\+}
+\tl{vāyuḥ sthiro yasya vinā prayatnāt |\\+}
+\tl{cittaṃ sthiraṃ yasya vināvalambanāt\\+}
+\tl{sa eva yogī sa guruḥ sa sevyaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 16
+\begin{versinnote}
+\tl{dṛṣṭiḥ sthirā yasya vinaiva dṛśyād\\+}
+\tl{vāyuḥ sthiro yasya vinā prayatnāt |\\+}
+\tl{cittaṃ sthiraṃ yasya vinā'valambāt \\+}
+\tl{sa eva yogī sa guruḥ sa sevyaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_072]
+%\end{philcomm}
+
+\begin{metre}[hp04_072]
+Upajāti
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.73}
+\begin{translation}[hp04_073]
+[The yogi] whose breath does not move in, out, left, right, up or down is liberated. In this there is no doubt.
+\end{translation}
+
+\begin{sources}[hp04_073]
+\emph{Gorakṣaśataka} 8
+\begin{versinnote}
+\tl{praveśe nirgame vāme dakṣiṇe cordhvam apy adhaḥ |\\+}
+\tl{na yasya vāyur vrajati sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
+\end{sources}
+
+%\begin{testimonia}[hp04_073]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_073]
+%\end{philcomm}
+
+\begin{metre}[hp04_073]
+Anuṣṭubh (c: bha-vipulā)
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.74}
+\begin{translation}[hp04_074]
+All the methods of Haṭha and Laya are for mastering Rājayoga. The man who has ascended to Rājayoga cheats death.
+\end{translation}
+
+%\begin{sources}[hp04_074]
+%\end{sources}
+
+\begin{testimonia}[hp04_074]
+\emph{Yogacintāmaṇi} (attrib.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{sarve layahaṭhābhyāsā rājayogasya siddhaye |\\+}
+\tl{rājayogaṃ samārūḍhaḥ puruṣaḥ kālavañcaka iti ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 55.34
+\begin{versinnote}
+\tl{haṭhapradīpikāyām–\\+}
+\tl{sarve haṭhalayopāyā rājayogasya siddhaye |\\+}
+\tl{rājayogasamārūḍhaḥ puruṣaḥ kālavañcakaḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṃsavilāsa} p.~49
+\begin{versinnote}
+\tl{sarve haṭhalayopāyā rājayogāya kevalam |\\+}
+\tl{rājayogaṃ samārūḍhaḥ puruṣaḥ kālavañcakaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_074]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.74*1}
+\begin{translation}[hp04_074_1]
+Iḍā is the divine Gaṅgā, Piṅgalā is the river Yamunā. Between those two is Suṣumṇā, who is to be recognised as Sarasvatī.
+\end{translation}
+
+%\begin{sources}[hp04_074_1]
+%\end{sources}
+
+%\begin{testimonia}[hp04_074_1]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_074_1]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.74*2}
+\begin{translation}[hp04_074_2]
+The place of the Triveṇī confluence is called the king of sacred sites. One should bathe there [and] be freed from all sins.
+\end{translation}
+
+%\begin{sources}[hp04_074_2]
+%\end{sources}
+
+%\begin{testimonia}[hp04_074_2]
+%\end{testimonia}
+
+%\begin{philcomm}[hp04_074_2]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.75}
+\begin{translation}[hp04_075]
+O great yogis, experience the nectar of Haṭha, the essence extracted from the ocean of all yoga scriptures after it has thus been churned, if you wish not to grow old and die.
+\end{translation}
+
+%\begin{sources}[hp04_075]
+%\end{sources}
+
+\begin{testimonia}[hp04_075]
+\emph{Haṭhasaṅketacandrikā} f.~145v
+\begin{versinnote}
+\tl{iti tu sakalayogaśāstrasindhoḥ \\+}
+\tl{parimathitād avakṛṣya sārabhūtaṃ [|]\\+}
+\tl{anubhavata haṭhāmṛtaṃ yamīndrā\\+}
+\tl{yadi bhavatām ajarāmaratvavāṃchā [||]\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_075]
+\end{philcomm}
+
+\begin{metre}[hp04_075]
+Puṣpitāgrā
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.75*1}
+\begin{translation}[hp04_75_1]
+The wise people in the world wash away sin at the sacred site of knowledge (\emph{vidyātīrthe}), the virtuous at the sacred site of truth (\emph{satyatīrthe}), the impure-minded at the sacred site of Gaṅgā (\emph{gaṅgātīrthe}), yogis at the sacred site of knowledge (\emph{jñānatīrthe}), kings at the sacred site of the streams (\emph{dhārātīrthe}), the rich at the sacred site of charity (\emph{dānatīrthe}) [and] women of good family at the sacred site of modesty.
+(\emph{lajjātīrthe})
+\end{translation}
+
+%\begin{sources}[hp04_75_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_75_1]
+\emph{Yogacintāmaṇi} ??
+\end{testimonia}
+
+%\begin{philcomm}[hp04_75_1]
+%\end{philcomm}
+
+\begin{metre}[hp04_075_1]
+Mandākrāntā 
+\end{metre}
+
+\subsection*{Colophon}
+\begin{translation}[hp04_col]
+Thus ends the fourth chapter in the \emph{Haṭhapradīpikā} composed by the glorious lord amongst yogis Svātmārāma.
 \end{translation}

### 2024-03-19 html/switch.js

@@ -18,10 +18,12 @@ document.addEventListener("DOMContentLoaded", () => {
 	    
 	} else if ( dev ) {
 	    hide("versdev");
+	    hide("jyotsnadev");
 	    dev=false;
 	    setbuttons();
 	} else {
 	    show("versdev");
+	    show("jyotsnadev");
 	    dev=true;
 	    setbuttons();
 	};
@@ -32,10 +34,12 @@ document.addEventListener("DOMContentLoaded", () => {
 	    
 	} else if ( ltn ) {
 	    hide("versltn");
+	    hide("jyotsnaltn");
 	    ltn=false;
 	    setbuttons();
 	} else {
 	    show("versltn");
+	    show("jyotsnaltn");
 	    ltn=true;
 	    setbuttons();
 	};

### 2024-03-27 export-for-publish.sh
fixed sub

@@ -1,0 +1,49 @@
+#!/bin/sh
+
+# optional argument 1 is filename, otherwise all relevant *.tex files
+
+xslcmd="java -jar ${HOME}/.nix-profile/saxon-he-11.5.jar"
+
+echo "processing xml …"
+echo "… hp1"
+$xslcmd -s:xml/HP1_text-tei.xml -xsl:xslt/hp-online-pub.xsl -o:html/hp1.html \
+	chapid="hp1" \
+	transl="../xml/HP1_comm-tei.xml" \
+	marma="../xml/Marmasthanas-tei.xml" \
+	jyotsna="../xml/jyotsna.xml"
+
+echo "… hp2"
+$xslcmd -s:xml/HP2_text-tei.xml -xsl:xslt/hp-online-pub.xsl -o:html/hp2.html \
+	chapid="hp2" \
+	transl="../xml/HP2_comm-tei.xml" \
+	marma="../xml/Marmasthanas-tei.xml" \
+	jyotsna="../xml/jyotsna.xml"
+
+echo "… hp3"
+$xslcmd -s:xml/HP3_text-tei.xml -xsl:xslt/hp-online-pub.xsl -o:html/hp3.html \
+	chapid="hp3" \
+	transl="../xml/HP3_comm-tei.xml" \
+	marma="../xml/Marmasthanas-tei.xml" \
+	jyotsna="../xml/jyotsna.xml"
+
+echo "… hp4"
+$xslcmd -s:xml/HP4_text-tei.xml -xsl:xslt/hp-online-pub.xsl -o:html/hp4.html \
+	chapid="hp4" \
+	transl="../xml/HP4_comm-tei.xml" \
+	marma="../xml/Marmasthanas-tei.xml" \
+	jyotsna="../xml/jyotsna.xml"
+
+echo "concatenating html …"
+sed '/<!--content-->/q' html/meta.html > html/hp-online-pub.html 
+
+for f in "html/hp1.html" "html/hp2.html" "html/hp3.html" "html/hp4.html"
+do
+    sed -e '/^\s*$/d' $f >> html/hp-online-pub.html
+done
+
+sed -n '/<!--content-->/,$ {p}' html/meta.html >> html/hp-online-pub.html
+
+#postprocess straddle consonants in 4line meteres
[…]

### 2024-07-02 latex/HP1_comm.tex
new input files

@@
 However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below).
-It apears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.
+It appears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.
 
@@
 Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}.  
-However, the difference in the sense is so minor that it does alter the meaning.
+The difference in meaning is negligable.
 
 The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
-\emph{-rājamsaudhaṃ}, which he explains as follows:
+\emph{-rājasaudhaṃ}, which he explains as follows:
 
@@
     adhirohiṇī}) 
-\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The text itself has
+\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The mūla text has
     \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā
@@
 the outset the primacy of Rājayoga, and to this end sacrificed the more poetically elegant original
-reading that mentions the royal mansion as a sort of metaphor for the ``royal yoga''.
+reading that mentions the royal mansion as a metaphor for the ``royal yoga''.
 
@@
 As Brahmānanda notes, the compound \emph{kṛpākaraḥ} can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}).
-In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} probably arose as a mistake for \emph{kṛ}.
+In the Devanāgarī transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} probably arose as a mistake for \emph{kṛ}.
 \end{philcomm}
@@
 Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it.
-The variant \emph{mahāyogī} in \epsilonOne and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}.
+The variant \emph{mahāyogī} in \epsilonOne\ and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}.
 One could emend to \emph{tathā} in light of the attested reading \emph{yathā} (\getsiglum{C7}) but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.
@@
 
-Caturbhuja Misra's \emph{Mugdhāvabodhinī} (1.7.8) on the \emph{Rasahṛdayatantra}
+Caturbhuja Miśra's \emph{Mugdhāvabodhinī} (1.7.8) on the \emph{Rasahṛdayatantra}
 
@@
 The \alphaOne\ and \alphaTwo\ reading of \emph{siddhaḥ kāpālikas} is an exception among the manuscripts and seems too vague to be referring to someone within a lineage.
-However, Khaṇḍa\-kāpālika is well attested by the manuscripts (including \alphaThree) and this name appears in other literature.
-Examples include \emph{Kathāsaritsāgara} 121.5 ff.
-(check??), \emph{Bṛhatkathāmañjarī} 10.45 (check??) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}).
-It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika.
+Khaṇḍa\-kāpālika is well attested by the remaining manuscripts (including \alphaThree) and this name appears in other texts, e.g.~Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45, where Khaṇḍakāpālika is the first of the 24 Vīras (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}).
+It is likely to refer to an ascetic who carries a broken skull.
+\emph{Matsyendrasaṃhitā} 33.2 mentions a practice for which one needs a \emph{khaṇḍakapāla} and in the \emph{Saṃvaramaṇḍala} of the \emph{Niṣpannayogāvalī}, p.26, Vajravārāhī is \emph{kapālakhaṇḍakṛtakaṭibhūṣaṇā}.
+The compound \emph{khaṇḍakāpālika} is found at \emph{Kathāsaritsāgara} 18.2.6, but there \sl{khaṇḍa} is being used in a derogatory sense (18.2.15 refers to the same character as a \sl{duṣṭakāpālika}).
+%It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika.
 Alternatively, it could simply refer to one who uses a broken skull as a bowl.
+% Reference to the KSS completed and text changed: before: Examples include \emph{Kathāsaritsāgara} 121.5 ff. (check).  Why "part-time".
 \end{philcomm}
@@
 \begin{translation}[hp01_009]
-These and other great adepts used the power of hathayoga to smash the rod of death and [so] are roaming the worlds.
+These and other great adepts used the power of haṭhayoga to smash the rod of death and [so] are roaming the worlds.
 \end{translation}
@@
 %The reading of \emph{samāśrayo} in \etaOne\ is metrically faulty. 
-The \textalpha\ group omits the second hemistich of this verse, but this was probably the result of eyeskip (i.e., \emph{°maṭho haṭhaḥ} is repeated).
+The \textalpha\ group omits the second line of this verse, but this was probably the result of eyeskip (i.e., \emph{°maṭho haṭhaḥ} is repeated).
 Both \emph{°jagatām} and \emph{°yuktānām} are well attested by the collated manuscripts.
@@
 \begin{philcomm}[hp01_011]        
-Either the singular or plural of yogin could be read here. The singular is well attested among the testimonia, but the manuscript transmission favours the plural.  
+Either the singular or plural of \emph{yogin} could be read here. The singular is well attested among the testimonia, but the manuscript transmission favours the plural.  
 \end{philcomm}
@@
 \begin{translation}[hp01_012]
-In a well-ruled, righteous region, with plenty of food and free from upheaval, the Haṭhayogi should live in an isolated hut.%JM changed disturbance to upheaval
-\end{translation}
-
-\begin{sources}[hp01_012]
-\end{sources}
+In a well-ruled, righteous region, with plenty of food and free from upheaval, the Haṭhayogi should live in an isolated hut.
+\end{translation}
+
+%\begin{sources}[hp01_012]
+%\end{sources}
 
@@
 In several other stories (\textit{Kathāsaritsāgara} 6.6.132, 10.5.89, 12.25.35), \emph{maṭhikā} is the term used for the hut of an ascetic.
-In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four fore-arm lengths (\emph{hasta}) high and wide, and it can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).
-
-In the \emph{Jyotsnā} and printed editions of the \emph{Haṭhapradīpikā}, including one by Digambara and Kokaje (1970: 6), this verse has the additional hemistich, \emph{dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite}.
-This hemistich derives from the \emph{Gorakṣaśataka} (32cd), which has \emph{°paryante} instead of \emph{°paryantaṃ}.
+In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four "\emph{hasta}s" high and wide (there are various definitions of the term: 18 inches according to the Larger Petrograd Dictionary and Monier-Williams, 48 inches according to the Smaller Petrograd Dictionary; the NWS lists even more variations, but the stipulation that the hut is the length of a bow found in the \emph{Gorakṣaśataka} (on which see below) suggests that 18 inches was meant).
+The hut can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).
+
+In the \emph{Jyotsnā} and printed editions of the \emph{Haṭhapradīpikā}, including one by Digambara and Kokaje (1970: 6), this verse has the additional line, \emph{dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite}.
+This line derives from the \emph{Gorakṣaśataka} (32cd), which has \emph{°paryante} instead of \emph{°paryantaṃ}.
 It stipulates that the hut should be built in a place measuring up to a bow length and free from rocks, fire and water.
-None of the early manuscripts has this hemistich, which suggests that it was added at a later time.
+None of the early manuscripts has this line, which suggests that it was added at a later time.
 Nonetheless, it appears in over a dozen manuscripts that were consulted for this edition.
@@
 \begin{translation}[hp01_013]
-It has a small door and is without cracks, holes and bumps.
+With a small door and no cracks, holes and bumps, neither too high nor too low in size, thickly smeared with cow dung in the proper way, clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
+%JM changed tr from : It has a small door and is without cracks, holes and bumps.
 It is neither too high nor too low in extent and is thickly smeared with cow dung in the proper way.
-It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.%JM remove "it is" etc.
-which read oddly, thus: With a small door and no cracks, holes and bumps, neither too high nor too low in size, thickly smeared with cow dung in the proper way, clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
+It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
 \end{translation}
@@
 \begin{philcomm}[hp01_017]        
-The reading of \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the main one of the \textalpha\ group.
+The reading \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the main one of the \textalpha\ group.
 Although this compound rarely appears in other yoga texts, a similar term \emph{śarīrapāṭava} occurs in the \emph{Śivasaṃhitā} (2.35) as one of the benefits bestowed by digestive fire (\emph{vaiśvānarāgni}), which indicates that the word \emph{pāṭava} was used in relation to the body and the benefits of yoga.
@@
 It appears to be verse filler here rather than indicating a temporal sequence of actions.
-Svātmārāma borrowed the verse on \emph{vīrāsana} from the \emph{Vasiṣṭhasaṃhitā}, the redactor of which appears to have adapted its first hemistich from a description of this posture in the \emph{Śāradātilakatantra}.
+Svātmārāma borrowed the verse on \emph{vīrāsana} from the \emph{Vasiṣṭhasaṃhitā}, the redactor of which appears to have adapted its first line from a description of this posture in the \emph{Śāradātilakatantra}.
 This would explain the rather strange syntax of the \emph{Vasiṣṭhasaṃhitā}’s version, in which \emph{adhaḥ kṛtvā} was changed to \emph{athaikasmin}, and \emph{tathetaram} became \emph{ca saṃsthitam}.
@@
 \begin{translation}[hp01_022]
-Knowers of yoga know that the tortoise pose (\emph{kūrmāsana}) arises by carefully blocking the anus with the ankles crossed.%??Tweak note to reflect change to crossed. 
+Knowers of yoga know that the tortoise pose (\emph{kūrmāsana}) arises by carefully blocking the anus with the ankles crossed. 
 \end{translation}
@@
 
-??The word \emph{vyutkrameṇa} appears to describe the position of the ankles.
-Its basic meaning is ‘against the normal direction,’ which would suggest that the ankles are turned out or crossed rather than placed together naturally.
-If the yogi is in a kneeling-type position, turning the feet out would bring the ankles together, blocking the perineal area.
+The word \emph{vyutkrameṇa} describes the position of the ankles.
+Its basic meaning is ‘against the normal direction’.
+In \emph{āsana} descriptions it usually means ‘crossed’ (see e.g.~\emph{Vasiṣṭhasaṃhitā} 1.71), which is how we have understood it here.
+It could also mean ‘turned out’: if the yogi is in a kneeling-type position, turning the feet out would bring the ankles together, blocking the perineal area.
 See \emph{Yoga Mīmāṃsā}, vol 8, no.
@@
 29–30 for a discussion of \emph{vyutkrameṇa} and the position of the ankles in \emph{kūrmāsana}, and vol 8, no.
-2, Figures 3–6 for photographs of a practitioner performing this \emph{āsana}.
+2, figures 3–6 for photographs of a practitioner performing this \emph{āsana}.
 \end{philcomm}
@@
 \begin{translation}[hp01_023]
-[The yogi] correctly assumes lotus pose, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is the wild cock pose (\emph{kukkuṭāsana}).
-\end{translation}
+[The yogi] correctly assumes the lotus pose, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is the wild cock pose (\emph{kukkuṭāsana}).
+\end{translation}
+%% JH "the lotus pose" for "lotus pose"
 
@@
 \begin{philcomm}[hp01_023]
-The names \emph{kurkuṭa} and \emph{kurkkuṭa} in some manuscripts are variant spellings of \emph{kukkuṭa} attested in the \emph{Pañcatantra} (M-W).
+The names \emph{kurkuṭa} and \emph{kurkkuṭa} in some manuscripts are variant spellings of \emph{kukkuṭa} attested in the \emph{Pañcatantra} (MW).
 \end{philcomm}
@@
 \begin{translation}[hp01_030]
-Supporting oneself on the ground with both hands, the elbows placed on either side of the navel, lifted up into the air in a raised posture [as straight] as a stick: they call this posture the peacock.
-
+Supporting oneself on the ground with both palms, the elbows placed on either side of the navel, lifted up into the air in a raised posture [as straight] as a stick: they call this posture the peacock.
+%Maybe we should adopt karadvayābhyāṃ in 30a.
+It is supported by alpha and explains many other variants ending in abhyāṃ, and perhaps its strangeness suggests it was original.
+JM: yes, but  karasthalābhyāṃ of alpha3 is even better and reflects the °talābhyāṃ of the parallels
 \end{translation}
@@
 \begin{sources}[hp01_030]
-\emph{Vimānārcanākalpa} 96
+Cf. \emph{Vimānārcanākalpa} 96
 
@@
 
-\emph{Pādmasaṃhitā} (\emph{yogapāda}) 1.21c–22d:
+Cf. \emph{Pādmasaṃhitā} (\emph{yogapāda}) 1.21c–22d:
 
@@
 
-\emph{Ahirbudhnyasaṃhitā} 31.36–37
+Cf. \emph{Ahirbudhnyasaṃhitā} 31.36–37
 
@@
 
-\emph{Vasiṣṭhasaṃhitā} 1.76–77
+Cf. \emph{Vasiṣṭhasaṃhitā} 1.76–77
 
@@
 \begin{testimonia}[hp01_030]
-\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
-
-\begin{versinnote}
-\tl{dharām avaṣṭabhya punaḥ karābhyāṃ\\+}
-\tl{tatkūrpare sthāpitanābhipārśvaḥ |\\+}
-\tl{tadāsane daṇḍavad utthitaḥ khe\\+}
-\tl{mayūram etat pravadanti santaḥ ||\\!}
-\end{versinnote}
-
 \emph{Haṭharatnāvalī} 3.42
@@
 
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{dharām avaṣṭabhya punaḥ karābhyāṃ\\+}
+\tl{tatkūrpare sthāpitanābhipārśvaḥ |\\+}
+\tl{tadāsane daṇḍavad utthitaḥ khe\\+}
+\tl{mayūram etat pravadanti santaḥ ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 \begin{philcomm}[hp01_030]
-%
-The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77), which are themselves derived from earlier Vaiṣṇava sources.
+There is no direct source of this verse, but it has the same elements as two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77), which are themselves derived from earlier Vaiṣṇava sources.
 The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}.
 
-In the second verse quarter, the pronoun in \emph{tatkūrpare} refers to the two hands (\emph{karadvaya}).
+In the second verse quarter, the pronoun in \emph{tatkūrpare} refers to the two hands.
 This is stated more explicitly (i.e., \emph{hastayoḥ kūrparau}) in \emph{Vasiṣṭhasaṃhitā} 1.76c and \emph{Yogayājñavalkya} 3.15c.
@@
 \subsection*{1.35}
+\begin{translation}[hp01_035a]
+Now, the adept's pose (\emph{siddhāsana}).
+\end{translation}
+
 \begin{translation}[hp01_035]
-Now, the adept's pose (\emph{siddhāsana}).
-
 [The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, focus the mind, hold the body erect and [remain] motionless, his senses restrained, gazing between the brows with his eyes unmoving.
@@
 \tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
-\tl{meḍhre pādam athaikam āsyahṛdaye dhṛtvā samaṃ vigraham |\\+}
+\tl{meḍhre pādam athaikam ekahṛdayo dhṛtvā samaṃ vigraham |\\+}
 \tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
-\tl{etan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\+}
-\tl{\var{7a °mūlaghaṭitaṃ ] °mulaghaṭanaṃ A • vinyasen ] GHT; vinyase VA, vinyaset Y}\\+}
-\tl{\var{7b meḍhre ] meṃḍhre A • athaikam āsyahṛdaye ] H; athaikadeśahṛdayo V, athaikam eva niyataṃ AGBGPk, athaikam eva niṣataṃ GL, athaikam eva hṛdayaṃ GP, athaikaṃ ekahṛdayo T, athaikam ekahṛdayaḥ Y • dhṛtvā ] VGBT; kṛtvā GLGPGPkY}\\+}
-\tl{\var{7c paśyed ] TH; paśyan VAGHSTvl, paśyad Y • antaraṃ ] VAGPkGLT; antare GBGP}\\+}
-\tl{\var{7d hy etan ] H; caitan VAGBGLGPT, etan GPkHS, tv etat Y • °janakaṃ ] °navidhau A, °jananaṃ GB, °nakaraṃ Y • procyate ] idaṃ bhavet Y (unm.)} ?? check\\!}
-%?? retain variants for athaikam āsyahṛdaye dhṛtvā samaṃ vigraham
+\tl{caitan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\+}
+%\tl{\var{7a °mūlaghaṭitaṃ ] °mulaghaṭanaṃ A • vinyasen ] GHT; vinyase VA, vinyaset Y}\\+}
+\tl{\var{7b athaikaṃ ekahṛdayo ] T; athaikadeśahṛdayo V, athaikam eva niyataṃ AGBGPk, athaikam eva niṣataṃ GL, athaikam eva hṛdayaṃ GP, athaikam ekahṛdayaḥ Y • dhṛtvā ] VGBT; kṛtvā GLGPGPkY}\\+}
+\tl{\var{7c paśyan ] VAGHSTvl; paśyed T paśyad Y} \\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_035]
-%?? Rewrite the note to reflect the adopted reading ekahṛdayo
-
-The adopted reading \emph{ekahṛdayo} is  supported by two manuscripts of the \emph{Haṭhapradīpikā} (\getsiglum{J2,M1}) and is close to the \textalpha\ reading \emph{ekahṛdaye}, which is also attested by the six-chapter \emph{Viveka\-mārtaṇḍa} and the \emph{Yoga\-cintāmaṇi}, ..
-
-which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}. In this case, \emph{ekahṛdayaḥ} would perhaps qualify the yogi as having a mind focused on one thing. 
-The second verse quarter has many variations in the \emph{Haṭhapradīpikā} manuscripts, as well as in the manuscripts of the sources and testimonia.
-\alphaOne\ has \emph{athaikam ekahṛdaye dhṛtvā} and most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}.
-The repetition of \emph{eka} appears to be a dittographical error and \emph{hṛdaye} does not make sense without an object of \emph{dhṛtvā} (i.e., having held what on the chest?).
-In the adopted reading \emph{athaikam āsyahṛdaye dhṛtvā}, which is attested by \etaOne, the oldest dated manuscript, the compound \emph{āsyahṛdaye} can be understood with the gerund \emph{dhṛtvā} and \emph{ekam} in the sense of `having held the face and chest together.' This alludes to the practice of the Jālandhara lock, in which the chin is placed on the chest.
-This meaning is more clearly seen in the \emph{Jyotsnā}'s version of the second verse quarter, \emph{hṛdaye kṛtvā hanuṃ susthiram} (`having put the jaw firmly on the chest').
-The compound \emph{āsyahṛdaye} is supported by other manuscripts, such as \getsiglum{N10,P1,P6,J16} and the variants \emph{asyahṛdaye} and \emph{asyahṛdayaṃ} also occur (e.g., \getsiglum{J7,J12,A1,V4,V18,V16,P8,P9}).
-
+In the second verse quarter, the adopted reading \emph{ekahṛdayo} is supported by two manuscripts of the \emph{Haṭhapradīpikā} (\getsiglum{J2,M1}) and is close to the \textalpha\ reading \emph{ekahṛdaye}.
+It is also attested by the six-chapter \emph{Viveka\-mārtaṇḍa} and the \emph{Yoga\-cintāmaṇi}, which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}.
+In this case, \emph{ekahṛdayaḥ} appears to describe the yogi as having his mind focused on one thing.
+There are many variations of this verse quarter in the \emph{Haṭhapradīpikā} manuscripts, as well as in the manuscripts of the sources and testimonia.
+Most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}, which is close to the adopted and \textalpha\ reading.
+Other readings allude here to the practice of the Jālandhara lock, in which the chin is placed on the chest.
+This is most clearly seen in the \emph{Jyotsnā}'s version, \emph{hṛdaye kṛtvā hanuṃ susthiram} (`having put the jaw firmly on the chest').
 The other well-attested reading, \emph{athaikaṃ eva niyataṃ}, was an attempt to fix the problem of \emph{hṛdaye} by replacing it with \emph{niyatam}, which must be read with \emph{meḍhre pādam athaikaṃ} (`having fixed one foot on the penis').
-But \emph{niyatam} is redundant here because of \emph{vinyaset} in the first \emph{pāda}.
-
+But \emph{niyatam} is redundant here because of \emph{vinyaset} in the first verse quarter.
+
+%The compound \emph{āsyahṛdaye} is supported by other manuscripts, such as \getsiglum{N10,P1,P6,J16} and the variants \emph{asyahṛdaye} and \emph{asyahṛdayaṃ} also occur (e.g., \getsiglum{J7,J12,A1,V4,V18,V16,P8,P9}).
+
+%MD 2024-5-14: adopt caitan (alpha, beta, and most of the mss of VM) in Pāda d? JM: yes
 \end{philcomm}
@@
 
-The last two references are taken from Reddy 1982: 41–42
+The last two references are taken from Reddy 1982: 41–42.
 
@@
 \begin{translation}[hp01_039]
-From among the eighty-four postures, one should regularly practise just \emph{siddhāsana}, in the same way from among the 72,000 channels [one should practise, focusing on] \emph{suṣumṇā}.
+From among the eighty-four postures, one should regularly practise just \emph{siddhāsana}, in the same way from among the 72,000 channels [one should practise, focusing on] Suṣumṇā.
 \end{translation}
@@
 This reading is well attested by manuscripts of the \emph{Haṭhapradīpikā} and is also found in the \emph{Yogacintāmaṇi}, which attributes it to the \emph{Yogasāramañjarī}.
-Perhaps, the second hemistich was added somewhat haphazardly by Svātmārāma, and then others have tried to make sense of it by changing \emph{suṣumnām iva nāḍiṣu} to \emph{nāḍīnāṃ malaśodhanam}, which occurs in the \emph{Jyotsnā} (1.39).
+Perhaps, the second line was added somewhat haphazardly by Svātmārāma, and then others have tried to make sense of it by changing \emph{suṣumnām iva nāḍiṣu} to \emph{nāḍīnāṃ malaśodhanam}, which occurs in the \emph{Jyotsnā} (1.39).
 The reading \emph{nāḍīnāṃ malaśodhanam/e} is probably a patch as no other texts say that \emph{siddhāsana} clears the channels.
@@
 \tl{etad vyādhivikārahāri yamināṃ padmāsanaṃ procyate || 8 ||\\!}
-\tl{\var{8a saṃsthāpya ] vinyasya T}\\+}
-\tl{\var{8b yāmyorūpari paścimena vidhinā dhṛtvā ] VT; dakṣorūpari tasya bandhanavid- hau dhṛtvā A, dakṣorūpari paścimena vidhinā dhṛtvā GPGPkY, tasyaivoparitaś ca vandhanavid- hiṃ kṛtvā GB , tato dakṣorūpari tasya vaṃdhanavidhau pṛṣṭhe GL (unm.) • dṛḍhaṃ ] dṛḍha∗ ṃ
-∗ V}\\+}
-\tl{\var{8c ālokayed ] ālokayan GB}\\+} 
-\tl{\var{8d etad ] antar° T • °vikārahāri yamināṃ ] VAT; °vikāranāśanakaraṃ GPk Y, °vikārakaṃdadamanaṇ GB , °vināśakāri yamināṃ GL GP, °vighātahāri yamināṃ Tvl • °nāśanakaraṃ G}\\!}
-%?? Variants not really needed as HP text is solid.
-\end{versinnote}
-
+\tl{\var{8d °vikārahāri yamināṃ ] VAT; °vikāranāśanakaraṃ GPk Y, °vikārakaṃdadamanaṇ GB , °vināśakāri yamināṃ GL GP, °vighātahāri yamināṃ Tvl}\\!}
+\end{versinnote}
 \end{sources}
@@
 \tl{yathāśaktyā samākṛṣya pūrayed udaraṃ śanaiḥ |\\+}
-\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ ||\\!}
-%?? Variants for uttabhya cibukaṃ vakṣasy āsthāpya pavanaṃ śanaiḥ ?
+\tl{yathāśaktyaiva paścāt tu recayet pavanaṃ śanaiḥ ||\\+}
+\tl{35cd \var{uttabhya ] PTβW1; uttabya M1, yuttamā° A, uttama J1, uttamā M2J2V, uttama J1 • cibukaṃ ] cubukaṃ TM1β, °bhyāṃ ca cu° AM2 • vakṣasy ] PT; vakṣaḥ M1π, °bukaṃ AM2, vakṣe Ba, vakṣya BbBp • āsthāpya ] P; āsthāya T, sthāpayet M1, vakṣastha° AM2, saṃsthāpya \emph{cett.}}}
 \end{versinnote}
@@
 
+%(MD 2024-5-14): It is remarkable that α2 and β1 have osthāpyot and ε3 otthāpya. Perhaps o = ava as in the Middle Indic? avasthāpya would make the perfect sense.
+
 The syntax of this verse as we have presented it is incomplete: at its end \emph{pavanaṃ śanaiḥ}, `the breath gradually', is left hanging.
@@
 
-The manuscript readings with \emph{vakṣa sthāpayet} (\getsiglum{J7,V3,J8,J10,J17,N17}) or something similar (\getsiglum{V1,W4}) do not offer a solution to the incomplete syntax so do not indicate that Svātmārāma rewrote \emph{Dattātreyayogaśāstra} 36 so that he could omit \emph{Dattātreyayogaśāstra} 37.
-In the absence of evidence that Svātmārāma included \emph{Dattātreyayogaśāstra} 37 or wrote a coherent version of \emph{Haṭhapradīpikā} 1.46, we have made sense of \emph{pavanaṃ śanaiḥ} by adding “[draw in]” in our translation.
+The manuscript readings with \emph{vakṣa sthāpayet} (\getsiglum{J7,V3,J8,J10,J17,N17}) or something similar (\getsiglum{V1,W4}) do not offer a solution to the incomplete syntax and do not indicate that Svātmārāma rewrote \emph{Dattātreyayogaśāstra} 36 so that he could omit \emph{Dattātreyayogaśāstra} 37.
+(The readings \emph{osthāpyot} (α2 and β1) and \emph{otthāpya} (ε3) are surprising and are perhaps Middle Indic forms in which Sanskrit \emph{ava-} becomes \emph{o-}.) In the absence of evidence that Svātmārāma included \emph{Dattātreyayogaśāstra} 37 or wrote a coherent version of \emph{Haṭhapradīpikā} 1.46, we have made sense of \emph{pavanaṃ śanaiḥ} by adding “[draw in]” in our translation.
 
@@
 Brahmānanda appears to have had in mind a probably older rule for meditation postures, according to which the tongue rests near the front teeth.
-One example of this is in \emph{Svacchandatantra} (4.365f.), which teaches a meditation pose called \emph{divyaṃ karaṇam} , in which the tongue is to rest at the tip
+One example of this is in \emph{Svacchandatantra} (4.365f.), which teaches a meditation pose called \emph{divyaṃ karaṇam}, in which the tongue is to rest at the tip
 of the teeth (\emph{dantāgre jihvām ādāya}).
 Other Tantric texts have this or similar rules, in which the tongue is supposed to rest either on the teeth or the palate, early examples being the \emph{Mṛgendrāgama}, \textit{yogapāda} 19 (\emph{dantāgre jihvām ādāya}) and \emph{Mataṅgapārameśvaratantra}, \textit{yogapāda} 2.27 (\emph{tālumadhyagatenaiva jihvāgreṇa}).
-Placing the tongue where it does not disturb the meditation seems quite appropriate for a `normal' meditative practice.
-\footnote{The rule of placing the tongue at the palate is also found in \emph{Īśānaśivagurudevapaddhati} 18.120: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān asaṃspṛśan ṛjukāyaḥ}.
+Placing the tongue where it does not disturb the meditation seems quite appropriate for a `normal' meditative practice.\footnote{The rule of placing the tongue at the palate is also found in \emph{Īśānaśivagurudevapaddhati} 18.120: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān asaṃspṛśan ṛjukāyaḥ}.
 For similar references in tantric and other works see Mallinson 2007:17–24}
@@
 \tl{muñcan prāṇam upaiti bodham atulaṃ śaktiprabhāvān naraḥ ||\\!}
-\tl{\var{36a °puṭitau ] GL GP GPk U; ghaṭitau VAGB T • baddhvātha ] VTvl U;
-baddhvā tu AGBH, dhyānaṃ tu GL, dhyāyes ta° GP, dhyātvā ca GPk, baddhvā ca T • padmāsanam ] VGBTU; tac cepsitaṃ GL, °taś cepsitaṃ GP, tat prekṣitam GPk}\\+}
 \tl{\var{36b dhyāyaṃś ]
@@
 \tl{\var{36c proccālayan ] ∗∗
-T; pro cc ālayan V, prodvārayaṃ A, proccālayet GB, proccārayet GLGPk, prollāsayet GP, proccārayan U • pūri- taṃ ] prerayaṃ A}\\+}
-\tl{\var{36d muñcan prāṇamupaiti bodhamatulaṃ śak- tiprabhāvān naraḥ ] U; prāṇaṃ muṃcati bodham eti śanakaiḥ proktaprabhāvād ataḥ V, pāṇaṃ muṃcati bodham eti śanakaiḥ śaktiḥ prabhāvād ataḥ A, prāṇaṃ muṃcati yāti bodham amalaṃ śaktipradhānoditaḥ GB, muṃcan prāṇam upaiti bodham atulaṃ śaktiprabhāvād ataḥ GLGP, muñcan prāṇam upaiti bodham atulaṃ śaktiprabodhān naraḥ GPk, prāṇaṃ muñcati bodhameti śanakai (ścu?śśa)[sic] ktiprabodhān naraḥ T, muñcan prāṇam upaiti bodham akhilāṃ śaktiṃ prabhāvād ataḥ Tvl}\\!}
-%?? Retain variants for dhyānaṃś ca tac cetasi, muñcan and proccālayan
+T; pro cc ālayan V, prodvārayaṃ A, proccālayet GB, proccārayet GLGPk, prollāsayet GP, proccārayan U}\\+}
+\tl{\var{36d muñcan prāṇamupaiti bodhamatulaṃ śak- tiprabhāvān naraḥ ] U; prāṇaṃ muṃcati bodham eti śanakaiḥ proktaprabhāvād ataḥ V, pāṇaṃ muṃcati bodham eti śanakaiḥ śaktiḥ prabhāvād ataḥ A, prāṇaṃ muṃcati yāti bodham amalaṃ śaktipradhānoditaḥ GB, muṃcan prāṇam upaiti bodham atulaṃ śaktiprabhāvād ataḥ GLGP, muñcan prāṇam upaiti bodham atulaṃ śaktiprabodhān naraḥ GPk, prāṇaṃ muñcati bodhameti śanakaiḥ śaktiprabodhān naraḥ T, muñcan prāṇam upaiti bodham akhilāṃ śaktiṃ prabhāvād ataḥ Tvl}\\!}
 \end{versinnote}
@@
 Its citation does not include \emph{Haṭhapradīpikā} 1.52cd, which affirms that 1.52cd is not from the \emph{Yogayājñavalkya} (or \emph{Vasiṣṭhasaṃhitā}).
-As far as we are aware, there is no source for the hemistich (1.52cd) mentioning the three locks, so it may have been composed by Svātmārāma or borrowed from a lost work. 
+As far as we are aware, there is no source for the line (1.52cd) mentioning the three locks, so it may have been composed by Svātmārāma or borrowed from a lost work. 
 \end{philcomm}
@@
 
-Manuscripts of two early groups, \textbeta\ and \textgamma, as well as the \emph{Jyotsnā} (1.53), include an additional hemistich specifying that the left ankle is placed on the left side and the right ankle on the right (\emph{savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe}).
-This hemistich appears to have been added to make it clear that the ankles are not crossed in \emph{bhadrāsana}, unlike the previous pose, \emph{siṃhāsana}.
+Manuscripts of two early groups, \textbeta\ and \textgamma, as well as the \emph{Jyotsnā} (1.53), include an additional line specifying that the left ankle is placed on the left side and the right ankle on the right (\emph{savyagulphaṃ tathā savye dakṣagulphaṃ tu dakṣiṇe}).
+This line appears to have been added to make it clear that the ankles are not crossed in \emph{bhadrāsana}, unlike the previous pose, \emph{siṃhāsana}.
 
@@
 \begin{translation}[hp01_055]
-When the great yogi does not tire from adopting the \emph{āsana}s in this way, he should  practise the breath techniques with seals and so forth, from which purification of the channels arises.%??
-Write to Haru to ask about nāḍiśuddhiṃ as a b/v
+When the great yogi does not tire from adopting the \emph{āsana}s in this way, he should  practise the breath techniques with seals and so forth, from which purification of the channels arises.% Write to Haru to ask about nāḍiśuddhiṃ as a b/v
 \end{translation}
@@
 \begin{philcomm}[hp01_055]
-The second hemistich can be interpreted in different ways.
+The second line can be interpreted in different ways.
 One possibility is to understand \emph{nāḍiśuddhiṃ} as a \emph{bahuvrīhi} qualifying \emph{mudrādipavanakriyām} in the sense that the yogi should practise the breathing techniques by way of the relevant \emph{mudrā}s and locks (taught in the third chapter), from which purification of the channels arise.
@@
 
-% ?? Mention that the mudrās practised with the breathing techniques are the bandhas (2.44-45) 
+%  
 \end{philcomm}
@@
 \tl{anuṣṭhānavihīnās tu vañcayanti janān kila || 47 ||\\!}
-\tl{\var{42c °yuktasya ] °yuktaḥsya W1, °yuktaḥ sa PT
-• siddhiḥ ] siddhi M1M2, siddhaḥ P}\\+}
-\tl{\var{43a na śāstra° ] śāstrasya PT • °mātreṇa ] °rūpeṇa
-M1A}\\+}
-\tl{\var{46a na ] sa° BB-
-BP • °dhāraṇaṃ ] °dhāriṇāṃ AM2, °dhāriṇaṃ JVYSS • siddheḥ ] sidhyai M1, siddhiḥ
-A, siddhi° β, kiṃ cit Y}\\+}
-\tl{\var{46b om. M AM (eye-skip) • kāraṇaṃ na ] kāraṇa∗nn∗a T SS 12
-• na ] tañ β • kathā ] W1VHPYCMK; katham PTβ, tathā JW2DYŚPT }\\+}
-\tl{\var{46c kriyaiva ] PTβW1VYCMPTKYSS; om. M1AM2, kṛpaiva JDYŚ, kṛyaiva W2 • siddheḥ ] siddhiḥ JYSS}\\+}
-\tl{\var{46d satyam ] satvam YSS • eva ] PTW1W2VDYŚK; etad M1AM2YCM, ekaṃ J, etan HP, etat PTYSS, etac β • tu sāṃkṛte ] dhi sāṃkṛte M1AM2YCM, na saṃśayaḥ HP, ca sāṃkṛte β}\\+}
-\tl{\var{47a śiśno° ] siśno° JW2 47a °ārthaṃ yogasya ] °āś ca yogāś ca P, ⊔⊔ yogasya T, °ārthaṃ yogaś ca AM2, °ārthayogaś ca β}\\+}
-\tl{\var{47b kathayā veṣadhāriṇaḥ ] W1W2DYŚYCMK; katheyaṃ veṣadhāriṇaḥ PTYSS, kasyeyaṃ veṣyadhāriṇaḥ M1, katheyaṃ veṣadhāriṇāṃ AM2, kaṃd- hāyāṃ yogadhāraṇaḥ BBBP, kathayan yogadhāraṇaḥ BA, kathaṃ yā veṣadhāriṇī J1, kathaṃ vā veṣadhāriṇī J2, kathaṃ ye veṣadhāriṇaḥ V, kathaṃ vā veṣadhāriṇaḥ PT}\\+}
-\tl{\var{47c anuṣṭhāna° ] anuṣṭhānā° M1, annapāna° PT • °vihīnās tu ] svayaṃ nityaṃ M1, °vihīnāś ca AM2}\\+}
-\tl{\var{47d janān ] janāḥ M1βV, janāt J, janā W1 • kila ] sadā M1AM2, iha YSS • vañcayanti janān kila ] vañcayaty akhilān janān YCM}\\!}
-%?? Simplify?
 \end{versinnote}
@@
 Both verses are similar to verses from the \emph{Dattātreyayogaśāstra} (cited as the source).
-However, only the first half of \emph{Dattātreyayogaśāstra} 47 is given in these later versions of the \emph{Haṭhapradīpikā}, resulting in a near-nonsensical hemistich.
+However, only the first half of \emph{Dattātreyayogaśāstra} 47 is given in these later versions of the \emph{Haṭhapradīpikā}, resulting in a near-nonsensical line.
 Also, the syntax of 1.55.2ef is corrupt.
@@
 \begin{translation}[hp01_056]
-%The various \emph{āsana}s, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then concentrating on the internal resonance is the sequence of practice in Haṭha.%??JM:
+%The various \emph{āsana}s, breath retention, bodily technique (\emph{karaṇa}) called seals (\emph{mudrā}), and then concentrating on the internal resonance is the sequence of practice in Haṭha.%JM: 
 Posture, manifold breath retention, the bodily technique called seal, then concentration on the internal sound is the sequence of practice in Haṭha.
@@
 \begin{sources}[hp01_057]
-\emph{Vivekamārtaṇḍa} 37
+\emph{Vivekamārtaṇḍa} 38
 
@@
 \tl{brahmacārī mitāhārī yogī yogaparāyaṇaḥ |\\+}
-\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||\\!}
-%?? variants on yogī??
+\tl{abdād ūrdhvaṃ bhavet siddho nātra kāryā vicāraṇā ||\\+}
+\tl{\var{38b yogī ] VT; tyāgī AG}}
 \end{versinnote}
@@
 The confusion between the two appears to have started early in the transmission of the \emph{Vivekamārtaṇḍa}.
-The occurrence of \emph{tyāgī} in \emph{Vivekamārtaṇḍa} 37 may be a dittographical type of mistake because the word \emph{tyāgī} is in the previous hemistich of that work.
+The occurrence of \emph{tyāgī} in \emph{Vivekamārtaṇḍa} 37 may be a dittographical type of mistake because the word \emph{tyāgī} is in the previous line of that work.
 But it is more difficult to determine whether Svātmārāma used a manuscript of the \emph{Vivekamārtaṇḍa} with \emph{tyāgī} or \emph{yogī} in verse 37.
@@
 \tl{annena kukṣer dvāv aṃśau pānenaikaṃ prapūrayet |\\+}
-\tl{āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet ||\\!}% ??translate citations
+\tl{āśrayaṃ pavanādīnāṃ caturtham avaśeṣayet ||\\!}% 
 \end{versinnote}
@@
 \begin{translation}[hp01_064]
-Whether young, old, very old, sick or even weak, the diligent yogi succeeds in all yogas through practice.
+Whether young, old, very old, sick or even weak, the diligent [yogi] succeeds in all yogas through practice.
 \end{translation}
@@
 \tl{yuvāvastho 'pi vṛddho vā vyādhito vā śanaiḥ śanaiḥ |\\+}
-\tl{abhyāsāt siddhim āpnoti yoge sarvo 'py atandritaḥ || 40 ||\\!}
-%?? variants on yoge sarvo 'py atandritaḥ?
+\tl{abhyāsāt siddhim āpnoti sarvayogeṣv atandritaḥ ||\\!}
 \end{versinnote}

### 2024-07-02 latex/HP2_comm.tex
new input files

@@
 Piṅgalā].
-Next, he should inhale through Piṅgalā, hold the breath and release it through the left [channel].
+He should inhale through Piṅgalā, hold the breath sufficiently long and release it through the left [channel].
 The channels of ascetics meditating on the two orbs of the sun and moon using this method are purified after three months.
 \end{translation}
+
+%
 
@@
 %??
-expand note on possibility of niti/nīti being original (found in alpha, betaone, epsilonone, delta i.e.
+Change tr, JM to write note expand note on possibility of niti/nīti being original (found in alpha, betaone, epsilonone, delta i.e.
 in good manuscripts in most groups; vernacular?) Only niti works well in final pāda.
@@
 In 22 alphatwo has nitiś caiva...
+Tricky; it seems that the spelling varies here between nīti and niti (in the header it's unanimously nīti in alpha)
 % adopt hanūrdhva
@@
 
-%??% JM to supply Babaji video to be uploaded to our host. Send to Mitsuyo.
+\medialink{Mahant Bālyogī Rām Bālak Dās performing nauli at Dīrgheśvar Mahādev, U.P., in 2012.}{BabajiNauli.m4v}
 
@@
 \begin{translation}[hp02_041]
-When the network of channels has been purified by breath-controls as prescribed, the breath pierces the mouth of \emph{suṣumṇā} and enters it with ease.
+When the network of channels has been purified by breath-controls as prescribed, the breath pierces the mouth of Suṣumṇā and enters it with ease.
 \end{translation}
@@
 
-%\begin{philcomm}[hp02_045]
-%\end{philcomm}
+\begin{philcomm}[hp02_045]
+The term \emph{uḍḍīyāṇa} is spelt various ways among manuscripts of the \emph{Haṭhapradīpikā} and other texts.
+Generally speaking, the \textalpha\ group have \emph{uḍḍīyāṇa}, \textbeta\ \emph{uḍḍiyāṇa}, \textgamma\ and \textdelta\ \emph{uḍḍiyāna} and the \emph{Jyotsnā} \emph{uḍḍīyāna}.
+Other spellings, such as \emph{uḍyāṇa} and \emph{uḍiyāṇa}, also occur.
+We have adopted the spelling of \textalpha\ except in cases where the metre of a verse requires otherwise, as in 2.45d (\emph{uḍḍiyāṇakaḥ}).
+As noted by Sanderson (2007: 265), diverse spellings of the place by the same name occur, including Uḍḍiyāna, Oḍḍiyāna, U/Oḍiyāna, U/Oḍyāna, or U/Oḍḍayana.
+\end{philcomm}
 
@@
 The meaning of \emph{nirodhāvadhi} is not entirely clear, but all sources and the \emph{Jyotsnā} (2.49) agree on this reading.
-The problem is that to practise \emph{kumbhaka} ``up to cessation \emph{nirodha}'' seems to suggest that cessation is not that of the physical breath, which by definition ceases in \emph{kumbhaka}, but of the vital wind (\emph{prāṇa}) within the body.
+The original reading may well have been \emph{virodhāvadhi}, “until it is uncomfortable”, which makes better sense but we have decide not to emend accordingly.
+The problem with \emph{nirodhāvadhi} is that to practise \emph{kumbhaka} ``up to cessation \emph{nirodha}'' seems to suggest that cessation is not that of the physical breath, which by definition ceases in \emph{kumbhaka}, but of the vital wind (\emph{prāṇa}) within the body.
 By citing a verse from an unnamed text, Brahmānanda seems to understand this verse as saying that the breath should very carefully (\emph{atiprayatnena}) be held as far as the extremities of the body so that it does not damage the body by exiting through the hair follicles:
@@
 \end{versinnote}
-%??JM: how about emending to virodhāvadhi, “until it is uncomfortable”, which would make much better sense. Yes, to add to note but not emend.
 \end{philcomm}
@@
 \tl{anena vidhinā satyaṃ yogīndro bhāti bhūtale ||\\+}
-\tl{°dehas tu ] °dehaḥ syāt P; °dehasyāt T,t1\\!}
+\tl{\var{°dehas tu ] °dehaḥ syāt P; °dehasyāt T,t1}\\!}
 \end{versinnote}
@@
 \begin{translation}[hp02_067]
-it is an auspicious thunderbolt that awakens \emph{kuṇḍalinī}, destroys bad deeds, bestows happiness, and destroys the blockage of phlegm, etc., situated at the mouth of the central channel,[\dots]
+it is an auspicious thunderbolt that awakens Kuṇḍalinī, destroys bad deeds, bestows happiness, and destroys the blockage of phlegm, etc., situated at the mouth of the central channel,[\dots]
 \end{translation}
@@
 
-Forcibly loud inhalation with the sound of a male bee; very slow exhalation with the sound of a female bee: as a result of practising thus, there arises in the minds of the best yogis an extraordinary blissful playfulness.
+Forcibly loud inhalation with the sound of a male bee; very slow exhalation with the sound of a female bee: as a result of practising thus, there arises in the minds of the best yogis a certain blissful stupefaction.
 \end{translation}
@@
 \tl{citte jātā kā cid ānandalīlā || \\+}
-\tl{līlā ]°mūrcchā N,n1,n2,n3,n4,J,T,t1\\!}
+\tl{\var{līlā ]°mūrcchā N,n1,n2,n3,n4,J,T,t1}\\!}
 \end{versinnote}
@@
 
-Many manuscripts of the \emph{Haṭhapradīpikā}, including the \textalpha\ group, have \emph{ānandamūrcchā} at the end of the verse, instead of \emph{ānandalīlā}, which we have adopted and which is supported by \betaTwo\ and the \textgamma\ group, as well as the most important testimonia (i.e.~the \emph{Haṭharatnāvalī}, \emph{Yogacintāmaṇi} and \emph{Yuktabhavadeva}).
-The reading \emph{ānanda\-mūrcchā} is probably the result of dittography as the heading \emph{atha mūrcchā} follows this verse.
-It has persisted in the transmission because  \emph{ānandamūrcchā} makes some sense by itself (`swooning through bliss'), although when it is read with the rest of the verse quarter, the meaning is somewhat odd (`swooning through bliss arises in the mind').
+Witnesses of the \gamma \ and \delta \ branches of the stemma, and Brahmānanda, have \emph{ānandalīlā} instead of \emph{ānandamūrchā}, perhaps in order to avoid the repetition of \emph{mūrchā}, which is also found in the following verse.
+
+%Old note before mūrcchā adopted.
+Many manuscripts of the \emph{Haṭhapradīpikā}, including the \textalpha\ group, have \emph{ānandamūrchā} at the end of the verse, instead of \emph{ānandalīlā}, which we have adopted and which is supported by \betaTwo\ and the \textgamma\ group, as well as the most important testimonia (i.e.~the \emph{Haṭharatnāvalī}, \emph{Yogacintāmaṇi} and \emph{Yuktabhavadeva}).
+The reading \emph{ānanda\-mūrcchā} is probably the result of dittography as the heading \emph{atha mūrchā} follows this verse.
+It has persisted in the transmission because  \emph{ānandamūrchā} makes some sense by itself (`swooning through bliss'), although when it is read with the rest of the verse quarter, the meaning is somewhat odd (`swooning through bliss arises in the mind').
 \end{philcomm}
@@
 \begin{translation}[hp02_070]
-Now mūrcchā:
-
-At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This, which is called \emph{mūrcchā}, bestows the bliss of  fainting.
-\end{translation}
+Now mūrchā:
+
+At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This fainting of the mind, which is called \emph{mūrchanā}, gives pleasure.
+\end{translation}
+%name is given as mūrchanā but headings in alpha etc. are mūrchā. Some, including epsilon1 do have mūrchanā.
 % adopt feminine in 2nd hemistich
+% JB: the meaning of manomūrcchā is not conveyed by the current translation.
+Fainting of the mind is different to normal fainting where a person falls to the ground unconsciousness.
+It's unlikely that the person in mūrcchā kumbhaka was supposed to collapse....
+bestows bliss, when the mind faints.
 
@@
 
-% MD: Perhaps \emph{recakaḥ pūrakaḥ kāryaḥ} which is closer to the readings of N3 and V3?  Cf. 2.74a \emph{recakaṃ pūrakaṃ muktvā}.
+% MD: Perhaps \emph{recakaḥ pūrakaḥ kāryaḥ} which is closer to the readings of N3 and V3?  Cf. also 2.74a \emph{recakaṃ pūrakaṃ muktvā}.
 % JB: We need the relative pronoun to pick up on the sa in the second pāda
-% MD: Not always. Cf. 1.58, 2.24, 2.26.
+% MD: Not necessarily. Cf. 1.58, 2.24, 2.26.
 \end{philcomm}
@@
 \begin{translation}[hp02_078]
-At the end of exhaling the retained breath, [the yogi] should make the mind supportless. By practising in this way he reaches the state of Rājayoga .
+At the end of exhaling the retained breath, [the yogi] should make the mind supportless. By practising in this way he reaches the state of Rājayoga.
 \end{translation}
@@
 \begin{translation}[hp02_079]
-As a result of retaining the breath, kuṇḍalinī awakens; as a result of the awakening of kuṇḍalinī, suṣumṇā becomes free of blockages and success in Haṭha arises.
+As a result of retaining the breath, Kuṇḍalinī awakens; as a result of the awakening of Kuṇḍalinī, Suṣumṇā becomes free of blockages and success in Haṭha arises.
 \end{translation}
@@
 \tl{anargalaḥ suṣumṇānto haṭhasiddhiś ca jāyate ||\\+}
-\tl{anargalaḥ ] L, antargataḥ N\\!}
+\tl{\var{anargalaḥ ] L, antargataḥ N}\\!}
 \end{versinnote}
@@
 \begin{translation}[hp02_080]
-Thinness of the body, clear complexion, clarity of the inner sound, very bright eyes, freedom from disease, mastery of semen, stimulation of the [body’s] fire [and] purification of the channels are the signs of success in Haṭha.
-\end{translation}
+Thinness of the body, clear complexion, clarity of the inner sound,  bright eyes, freedom from disease, mastery of semen, stimulation of the [body’s] fire and purification of the channels are the signs of success in Haṭha.
+\end{translation}
+%MD: Pāda d: from a stemmatic point of view, nāḍīṣu śuddhir is better than nāḍīviśuddhir.JM: adopt
+% and Pāda b: ca nirmale is better than sunirmale. JM: adopt
 

### 2024-07-02 latex/HP3_comm.tex
new input files

@@
 \begin{translation}[hp03_001]
-Just as the lord of snakes is the foundation of the regions of the earth along with their mountains and forests, so \emph{kuṇḍalinī} is the foundation of all systems of yoga.
+Just as the lord of snakes is the foundation of the regions of the earth along with their mountains and forests, so Kuṇḍalinī is the foundation of all systems of yoga.
 \end{translation}
@@
 \begin{translation}[hp03_002]
-When the sleeping \emph{kuṇḍalinī} awakens through the favour of the guru, then all the lotuses are pierced, and the knots too, [\dots] 
+When the sleeping Kuṇḍalinī awakens through the favour of the guru, then all the lotuses are pierced, and the knots too, [\dots] 
 \end{translation}
@@
 \begin{translation}[hp03_004]
-\emph{Suṣumṇā}, the empty pathway, the great path to the aperture of Brahmā, the cremation ground, Śāmbhavī, and the middle path are synonyms.
+Suṣumṇā, the empty pathway, the great path to the aperture of Brahmā, the cremation ground, Śāmbhavī, and the middle path are synonyms.
 \end{translation}
@@
 \begin{translation}[hp03_006]
-The great seal, the great lock, the great piercing, the sky-roving [seal], the \emph{uḍḍiyāna} [lock], the root lock and the lock called \emph{jālandhara}, [\dots] 
+The great seal, the great lock, the great piercing, the sky-roving [seal], the \emph{uḍḍiyāna} [lock], the root lock, then [the lock] called \emph{jālandhara}, [\dots] 
 \end{translation}
@@
 \begin{translation}[hp03_012]
-[\dots] so the goddess \emph{kuṇḍalinī} suddenly becomes straight. Then she becomes still in the vessel with two halves.
+[\dots] so the goddess Kuṇḍalinī suddenly becomes straight. Then she becomes still in the vessel with two halves.
 \end{translation}
@@
 \tl{tathāsau maraṇāvasthāṃ harate dvipuṭāśrayām ||\\+}
-\tl{dvipuṭā ] em., dvipaṭā° L, dvipadā° N. \\!}
+\tl{\var{dvipuṭā ] em., dvipaṭā° L, dvipadā° N.} \\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp03_012]
-This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of \emph{kuṇḍalinī} in the central channel.
+This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of Kuṇḍalinī in the central channel.
 Drawing on Hellwig 2009:238–240, Mallinson and Szanto (2021:21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert.
-In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when \emph{kuṇḍalinī} is said to be in the state of \emph{maraṇa} the meaning is that she is stilled.
+In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when Kuṇḍalinī is said to be in the state of \emph{maraṇa} the meaning is that she is stilled.
 The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009:342, Mallinson and Szanto (2021:22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”.
@@
 Brahmānanda seems to take it as the death of \textit{prāṇa}, or in other words, the absence of the breath, in the two nostrils (\textit{maraṇāvasthā jāyate kuṇḍalībodhe sati suṣumnāyāṃ praviṣṭe prāṇe dvayoḥ puṭayoḥ prāṇaviyogāt}).
-Bhavadeva thought that \emph{kuṇḍalinī}, along with \textit{prāṇa} and \textit{apāna}, remains in the two nostrils while the breath is being held (\textit{evaṃ vāyudhāraṇāyāṃ kriyamāṇāyāṃ vyākulā bhūtā kuṇḍalinī apānaprāṇābhyāṃ saha nāsāpuṭadvayāśritā bhavati}).
-Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvipuṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}.
-Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened \emph{kuṇḍalinī} destroys death, which is usually dependent on the in and out breaths.
+Bhavadeva thought that Kuṇḍalinī, along with \textit{prāṇa} and \textit{apāna}, remains in the two nostrils while the breath is being held (\textit{evaṃ vāyudhāraṇāyāṃ kriyamāṇāyāṃ vyākulā bhūtā Kuṇḍalinī apānaprāṇābhyāṃ saha nāsāpuṭadvayāśritā bhavati}).
+Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvi\-puṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}.
+Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened Kuṇḍalinī destroys death, which is usually dependent on the in and out breaths.
 
@@
 \begin{translation}[hp03_013]
-{}[The yogi] should then exhale very slowly, not quickly. This is the great seal revealed by the great Siddhas.
+{[}The yogi] should then exhale very slowly, not quickly. This is the great seal revealed by the great Siddhas.
 \end{translation}
@@
 This verse, which is from the \emph{Amaraugha} and similar to verses in the \emph{Amṛtasiddhi} and \emph{Śivasaṃhitā}, is stating that the great seal, lock and piercing should be practised together.
-This can be done as a sequence as shown in this video.
-%??
-insert video here
+This can be done as a sequence as shown in \medialink{this video}{the_three_mudrās.mp4} (© Mark Robberds).
 \end{philcomm}
@@
 Now, the sky-roving [seal] (\emph{khecarī}):\\
-By cutting, moving and milking, the yogi should gradually lengthen the tongue until it touches the middle of the brows. Then there is success in \emph{khecarī}.
+By cutting, moving, and milking, the yogi should gradually lengthen the tongue until it touches the middle of the brows. Then there is success in \emph{khecarī}.
 \end{translation}
@@
 \begin{metre}[hp03_033]
-Upagīti 
+Āryā
 \end{metre}
@@
 %%%%%%%%%%
-\subsection*{3.35}
-\begin{translation}[hp03_035]
+\subsection*{3.34*1}
+\begin{translation}[hp03_034_1]
 If the yogi turns back the tongue and remains [like that] for half an instant, he is instantly freed from disease, death, old age and the like.
@@
 
-\begin{sources}[hp03_035]
+\begin{sources}[hp03_034_1]
 \emph{Śivasaṃhitā} 3.91
@@
 
-\begin{testimonia}[hp03_035]
+\begin{testimonia}[hp03_034_1]
 \emph{Yogacintāmaṇi} f.~74r (attr.~to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp03_035]
-We have adopted the version of 2.35 in \alphaTwo\ and \alphaThree\ (\alphaOne\ omits 2.35 and the greyscaled verses).
-This version is also attested by the \textgamma\ and \texteta\ groups.
-The additional hemistich gives \emph{vyomacakra} as an alternative name for \emph{khecarīmudrā}.
-This alternative name does not occur in any of the source texts known to be have been used by Svātmārāma, but it may have been inspired by the name \emph{nabhomudrā} for \emph{khecarīmudrā} in the \emph{Vivekamārtaṇḍa} (40).
-\end{philcomm}
+\begin{philcomm}[hp03_034_1]
+This verse is absent in \alphaOne\ and was likely added to a block of verses that Svātmārāma borrowed from the \textit{Vivekamārtaṇḍa} (47–51).
+3.34*1 is in \alphaTwo\ and \alphaThree\ (\alphaOne\ omits 2.35 and the greyscaled verses) and also the \textgamma\ and \texteta\ groups.
+Other manuscripts have an additional line that gives \emph{vyomacakra} as an alternative name for \emph{khecarīmudrā}.
+This alternative name does not occur in any of the source texts known to have been used by Svātmārāma, but it may have been inspired by the name \emph{nabhomudrā}, which is what the \emph{Vivekamārtaṇḍa} calls \emph{khecarīmudrā} (\emph{Vivekamārtaṇḍa} 40).
+\end{philcomm}
+
+% (MD 2024-5-15): Should we perhaps take the absence of this verse in α1 seriously and regard it as inauthentic?
+HP 3.34-43 are all from VM, but not this 3.35.
+In α2 and α3 it may have been added independently by contamination.
+There are considerable differences in the readings between the two.
+JM: seems good to me.
+% Pāda  α2 - α3
+% e: nītvā - kṛtvā*
+% f: api - yadi*
+% g: viṣaye - kṣaṇena*
+% The asterisk indicates that the reading has been adopted for our edition.
 
@@
 %JB: Retain variants for lipyate na ca as alpha reads na ca lipyata/lipyati
+%MD(2024-5-14): Perhaps we should read na ca lipyati karmaṇā as epic Skt.
 \end{versinnote}
@@
 
-%\begin{philcomm}[hp03_037]
-%\end{philcomm}
+\begin{philcomm}[hp03_037]
+The form of \emph{lipyati}, which is found in the \alpha \ manuscripts and has been adopted, is widely attested in epic Sanskrit.
+\end{philcomm}
 
@@
 \begin{translation}[hp03_040]
-Even when semen has moved [down] and reached the region of the perineum, it moves upwards having been blocked by \emph{yonimudrā} and struck by the goddess [\emph{kuṇḍalinī}].
+Even when semen has moved [down] and reached the region of the perineum, it moves upwards having been blocked by \emph{yonimudrā} and struck by the goddess [Kuṇḍalinī].
 \end{translation}
@@
 
-\tl{53c \var{gacchaty ] vrajaty \vl 53d hataḥ ] tanaṃ, hṛtas, kṛte, kṛtaḥ, tadā, gatā \vl • nibaddho ] niruddho \vl}\\!}
+\tl{\var{53c gacchaty ] vrajaty \vl 53d hataḥ ] tanaṃ, hṛtas, kṛte, kṛtaḥ, tadā, gatā \vl • nibaddho ] niruddho \vl}\\!}
 \end{versinnote}
@@
 
-A manuscript of an early recension of the \emph{Vivekamārtaṇḍa} (ms.~no.~4110) has \emph{hataḥ śaktyā}, which is attested by three \emph{Haṭhapradīpikā} manuscripts on lower branches of the stemma (i.e.~\getsiglum{Ba} of \textdelta\ group and \getsiglum{C2} and \getsiglum{P4} of group 7a ??).
-The participle \emph{haṭaḥ} makes sense of the instrumental \emph{śaktyā}, rendering the meaning `struck by \emph{kuṇḍalinī}.'
+A manuscript of an early recension of the \emph{Vivekamārtaṇḍa} (ms.~no.~4110) has \emph{hataḥ śaktyā}, which is attested by three \emph{Haṭhapradīpikā} manuscripts on lower branches of the stemma (i.e.~\getsiglum{Ba} of \textdelta\ group and \getsiglum{C2} and \getsiglum{P4} of group 7a).
+The participle \emph{haṭaḥ} makes sense of the instrumental \emph{śaktyā}, rendering the meaning `struck by Kuṇḍalinī.'
 % MD: These groups don't have a Greek sigla yet.
@@
 Alternatively, the word \emph{haṭaḥ} (as well as the other variants \emph{kṛtaḥ, kṛte}, and even \emph{haṭhāt}) may derive from \emph{hṛtaḥ}, which is attested by \emph{Haṭhapradīpikā} manuscripts from group 4a (i.e., \getsiglum{Ba5}, \getsiglum{N4a} and \getsiglum{N13}).
-The reading \emph{hṛtaḥ śaktyā} renders the verse as saying that semen goes up, carried by \emph{kuṇḍalinī}.
+The reading \emph{hṛtaḥ śaktyā} renders the verse as saying that semen goes up, carried by Kuṇḍalinī.
 
@@
 The \emph{Haṭhayogasaṃhitā} (43–48) teaches a different version of \emph{yonimudrā} in its reportoire of twenty-five \emph{mudrā}s.
-In this work, \emph{yonimudrā} is supposed to awaken \emph{kuṇḍalinī} and consists of sitting in \emph{siddhasana}, blocking the ears, eyes, nose and mouth with the thumbs, index, middle and ring fingers, uniting \emph{praṇa} and \emph{apāna}, meditating on the six \emph{cakra}s and repeating the mantra \emph{huṃ haṃsa}.
+In this work, \emph{yonimudrā} is supposed to awaken Kuṇḍalinī and consists of sitting in \emph{siddhasana}, blocking the ears, eyes, nose and mouth with the thumbs, index, middle and ring fingers, uniting \emph{praṇa} and \emph{apāna}, meditating on the six \emph{cakra}s and repeating the mantra \emph{huṃ haṃsa}.
 
@@
 He also understands the sixteen-petalled lotus to be the lotus in the throat, into which the nectar drips.
-%Adopt patrapadma.
-\end{philcomm}
-
+%Adopt patrapadma. % MD: VM-T has also padmapattra...
+\end{philcomm}
+
+% MD(2024-5-14): adopt vapur (α1,β1,ε2 etc.) instead of tanur (α2,β2,βω etc.)?
 \begin{metre}[hp03_047]
@@
 %\begin{philcomm}[hp03_048]
-%kaṭukāmla° or kaṭukātha. Adopt kaṭukātha (N3,J5, \emph{Vivekamārtaṇḍa})
-%guṇavat (N3, J5 [-va])
-%MD: Done.
+%MD(2024-5-14): adopt sadṛśā? (alpha and VM)
 %\end{philcomm}
@@
 \tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\+}
-\tl{°suṣiraṃ ] \getsiglum{K1}, °sukhire \alphaTwo. °mūrdhny ] \getsiglum{P8}. °mūle \alphaTwo. asti tathyaṃ ] \getsiglum{K1}, yad astī \alphaTwo.\\!}
+\tl{\var{°suṣiraṃ ] \getsiglum{K1}, °sukhire \alphaTwo. °mūrdhny ] \getsiglum{P8}. °mūle \alphaTwo. asti tathyaṃ ] \getsiglum{K1}, yad astī \alphaTwo.}\\!}
 “That cool liquid by which the aperture is filled at the top of Meru and exists as the truth, the wise [yogi] says that is the source of [all] rivers.” 
@@
 The root lock (\emph{mūlabandha})—\\
-When [the yogi] presses the perineum with the side of the heel, clenches the anus and draws up \emph{apāna}, it is called the root lock.
-\end{translation}
+When [the yogi] presses the perineum with part of the heel, clenches the anus and draws up \emph{apāna}, it is called the root lock.
+\end{translation}
+% JB: the back part of the heel presses the perineum. It would be better to say, 'with part of the heel.'
 
@@
 \begin{translation}[hp03_051]
-[The yogi] forces the downward-moving \emph{apāna} breath to move upwards by contracting [the anus]. Yogis call that \emph{mūlabandha}.
-\end{translation}
+It forces the downward-moving \emph{apāna} breath to move upwards by contraction [of the anus]. Yogis call that \emph{mūlabandha}.
+\end{translation}
+% JB: Perhaps it is better to take MB as the subject of 51ab as MB is the subject of the previous verse?
+i.e., It forces the downward-moving \emph{apāna} breath to move upwards because of the contraction.
 
@@
 In the same vein, Bhava\-deva\-miśra glosses `pulling \emph{apāna}' (\emph{apānā\-karṣaṇam}) as `contracting the anus' (\emph{gudā\-kuñcanam}), when commenting on `having pulled \emph{apāna\-vāyu} and forcefully contracted it ...' (\emph{apāna\-vāyum ākṛṣya balād ākuñcya...}) in \emph{Yukta\-bhava\-deva} 297 and 301.
-\end{philcomm} %?? JB: I've added references to the commentators Brahmānanda and Bhavadevamiśra to clarify the meaning of vāyum ākuñcayet.
+\end{philcomm} %
 
@@
 \begin{translation}[hp03_057]
-Heated by that [blaze], the sleeping \emph{kuṇḍalinī} wakes up. Like a snake struck by a stick, she hisses and becomes straight.
+Heated by that [blaze], the sleeping Kuṇḍalinī wakes up. Like a snake struck by a stick, she hisses and becomes straight.
 \end{translation}
@@
 Now the Uḍḍīyaṇa lock:\\
-Because the breath flies up (\emph{uḍḍīyate}) into the \emph{suṣumṇā} when it is bound by it, yogis say that this [practice] is called Uḍḍīyaṇa.
+Because the breath flies up (\emph{uḍḍīyate}) into the Suṣumṇā when it is bound by it, yogis say that this [practice] is called Uḍḍīyaṇa.
 \end{translation}
@@
 \begin{philcomm}[hp03_061]
-The variant spelling \sl{tāṇa} (for \emph{tāna}) reflects vernacular pronunciation (see e.g.~Molesworth 1857 s.v.~\emph{tāṇa}). 
+The spelling \sl{tāṇa} (where many witnesses have \emph{tāna}) reflects vernacular pronunciation (see e.g.~Molesworth 1857 s.v.~\emph{tāṇa}). 
 \end{philcomm}
@@
 \begin{translation}[hp03_062]
-\emph{Uḍḍiyāna} is easy, but it is always taught by a guru. If he practises it tirelessly, even an old man becomes young.
+\emph{Uḍḍiyāṇa} is easy, but it is always taught by a guru. If he practises it tirelessly, even an old man becomes young.%
 \end{translation}
@@
 
-Yogabīja 109 (south Indian recension)
+\emph{Yogabīja} 109 (south Indian recension)
 \begin{versinnote}
@@
 \begin{translation}[hp03_072]
-[The yogi] should contract the place of the root  and do the \emph{uḍḍiyāna} [lock]. He should [then] block the \emph{iḍā} and \emph{piṅgalā} [channels] and make [the breath] flow in the rear pathway.
+[The yogi] should contract the place of the root  and do the \emph{uḍḍiyāna} [lock]. He should [then] block the Iḍā and Piṅgalā [channels] and make [the breath] flow in the rear pathway.
 \end{translation}
@@
 Now the inverted bodily position–\\
-The sun devours whatever divine nectar flows from the moon. As a result, the body perishes.
+The sun devours whatever nectar flows from the divine moon. As a result, the body is afflicted by old age.
 \end{translation}
@@
 
-%MD: 3.75-76 gray-scaled. Keep \textit{divyarūpi ca} und \textit{vināśi ca} or change them to \textit{divyarūpiṇaḥ} and \textit{jarāyutam}?
+%
 \end{philcomm}
@@
 
-%\begin{philcomm}[hp03_086]
-%No known source.
-%\end{philcomm}
+\begin{philcomm}[hp03_086]
+Both \alphaTwo\ and \alphaThree\ indicate that \textit{manas} instead of \textit{citta} was the reading of the initial compound.
+Therefore, we have conjectured \textit{manāyattaṃ}, assuming double \textit{sandhi} from \emph{manas-āyattam}.
+\end{philcomm}
+%MD (2024-06-14): The original reading may have been mana-āyattaṃ as supported by α3 (anā°).
+manomayaṃ (α2), mano'dhīnaṃ (B) and cittāyattaṃ (the others) seem to have  resulted from an attempt to avoid the middle Indic form mana or the double Saṃdhi.
+And we have manas in Pāda c, not citta!
+Cf.
+also 4.8*22b.
 
@@
 %tasmād ayaṃ sādhakāya hemistich is not in Gr2 and seems like an unnecessary repetition (that may have occurred to change a 6 pāda verse into two verses).
-%?? JB we havent translated 3.101*1. Are we going to keep it in the edition or put it in a footnote in the apparatus of the edition (it seems quite repetitive as 3.101*1b= 3.101d)
+%
 % MD: 3.101*1ab and 3.101cd by the old numbering are now merged as 3.94*cd.
@@
 \end{translation}
-% ?? JB. the heading is regular typeface and 3.094_2 is greyscale. So we may need to separate them in our translation to show that the heading is part of the original text.
-
+% 
 %\begin{sources}[hp03_094_2]
@@
 \begin{translation}[hp03_094_3]
-Just as one might use a key to force open a double door, so the yogi breaks open the door to liberation with \emph{kuṇḍalinī}.
+Just as one might use a key to force open a double door, so the yogi breaks open the door to liberation with Kuṇḍalinī.
 \end{translation}
@@
 \begin{translation}[hp03_094_6]
-[Just as] the coiled serpent Ananta (\emph{śeṣa\-kuṇḍalī}) is the foundation of the oceans, mountains and islands, so kuṇḍalinī is the foundation of all systems of yoga.
+[Just as] the coiled serpent Ananta (\emph{śeṣa\-kuṇḍalī}) is the foundation of the oceans, mountains and islands, so Kuṇḍalinī is the foundation of all systems of yoga.
 \end{translation}
@@
 \begin{translation}[hp03_094_7]
-\emph{Kuṇḍalinī} is said to have a curved shape like a snake. The person who makes that goddess move is sure to be liberated.
+Kuṇḍalinī is said to have a curved shape like a snake. The person who makes that goddess move is sure to be liberated.
 \end{translation}
@@
 In some manuscripts this verse is followed by one in which \emph{bālaraṇḍā} is identified as \emph{sarasvatī}, which in the context of \emph{śakticālana} could refer to the tongue.
-She could also be\emph{kuṇḍalinī}, who in 3.94*5 is located at the navel, which is said to be the location of Viṣṇu (e.g.~\emph{Dhyāna\-bindū\-paniṣat} 28-30).
-In his commentary on this verse in the \emph{Bodhasāra} (1906: 137), Divākara says that the seizing of \emph{kuṇḍalinī} itself is the highest state of Viṣṇu (...
+She could also be Kuṇḍalinī, who in 3.94*5 is located at the navel, which is said to be the location of Viṣṇu (e.g.~\emph{Dhyāna\-bindū\-paniṣat} 28-30).
+In his commentary on this verse in the \emph{Bodhasāra} (1906: 137), Divākara says that the seizing of Kuṇḍalinī itself is the highest state of Viṣṇu (...
 \emph{bālaraṇḍāṃ ...
@@
 
-% comment on 3.95*1, which is similar to 3.95?
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.95*1}
+% \begin{translation}[hp03_095_1]
+% \end{translation}
+
+\begin{philcomm}[hp03_095_1]
 Verse 3.95*1, which has no known source, simply identifies the technical terms in 3.95, namely, \emph{gaṅgā}, \emph{yamunā} and \emph{bālaraṇḍā} as \emph{iḍā, piṅgalā} and \emph{sarasvatī}.
@@
 \end{philcomm}
-
-%%%%%%%%%%
-\subsection*{3.95*1}
-% iḍā bhagavatī gaṅgā
+% \alphaTwo\ has this verse in a correct position. Its second half is different from that of the other recensions:
+% iḍā bhagavatī gaṅgā piṅgalā yamunā nadī |
+% tayor madhye prayāgaṃ tu yas taṃ veda sa vedavit ||
+% Cf. Mahābhārata Bhīṣmaparvan
+%06,040.078d@003A_0041 iḍā bhagavatī gaṅgā piṅgalā yamunā nadī |
+%06,040.078d@003A_0042 tayor madhye tṛtīyā tu tat prayāgam anusmaret ||
 
@@
 These verses do not have a known source and are somewhat obscure unless one is one aware of the more coherent explanation of this practice in the \emph{Gorakṣaśataka}.
-In his \emph{Haṭhasaṅketacandrikā} (see testimonia), Sundaradeva makes sense of this verse by equating the tongue with \emph{kuṇḍalinī}.
+In his \emph{Haṭhasaṅketacandrikā} (see testimonia), Sundaradeva makes sense of this verse by equating the tongue with Kuṇḍalinī.
 This enables him to understand the reference to the cloth (\emph{paridhāna}) as the technique of wrapping the tongue in a cloth and milking it, which is a practice called \emph{sarasvatīcālana} in the \emph{Gorakṣaśataka} (16–25).
@@
 
-%%?? JB: We have included paristhitā in our translation (next to `coiled") and it is an unusual meaning of paristhitā. Did anyone take down a reference for understanding it as `coiled'?
+%
 
@@
 \begin{translation}[hp03_096_2]
-It is said that the characteristics of the cloth for wrapping around [the tongue] are  that it measures a handspan in length and four fingerbreadths in width, and is soft and white.
+It is said that the characteristics of the cloth for wrapping around [the tongue] are a length measuring a handspan and a width of four fingerbreadths, and it is soft and white.
 \end{translation}
@@
 \begin{philcomm}[hp03_096_2]
-%adopt vistāraṃ ?? This was written here (maybe at the time of the workshop?) JB we dont have vistāraṃ as the lemma in the edition?
 This verse was likely added to explain `by the method of the cloth' (\emph{paridhānayuktyā}) in the previous verse.
@@
 \begin{translation}[hp03_097]
-Sitting in \emph{vajrāsana}, the yogī should make \emph{kuṇḍalinī} move and immediately afterwards perform \emph{bhastrī}. He quickly awakens \emph{kuṇḍalinī}.
+Sitting in \emph{vajrāsana}, the yogī should make Kuṇḍalinī move and immediately afterwards perform \emph{bhastrī}. He quickly awakens Kuṇḍalinī.
 \end{translation}
@@
 \begin{philcomm}[hp03_097]
-\emph{Bhastrī} or \emph{bhastrikā kumbhaka} is taught at 2.60–68.The reference to \emph{vajrāsana} may be pointing to the practice of \emph{uḍḍiyāna}, which was described earlier in the chapter and is supposed to awaken \emph{kuṇḍalinī}.
+\emph{Bhastrī} or \emph{bhastrikā kumbhaka} is taught at 2.60–68.The reference to \emph{vajrāsana} may be pointing to the practice of \emph{uḍḍiyāna}, which was described earlier in the chapter and is supposed to awaken Kuṇḍalinī.
 The contraction of the sun mentioned in the next verse supports this.
@@
 \begin{translation}[hp03_098]
-[The yogi] should contract the sun and then make kuṇḍalinī move. Even if he is in the jaws of death, why would he fear death?%??JM: ...there is no danger of death.
+[The yogi] should contract the sun and then make Kuṇḍalinī move. Even if he were in the jaws of death, why would he fear death?%
 \end{translation}
@@
 \begin{sources}[hp03_098_1]
-\end{sources}%?? Aś verse in sources?
-
-\begin{testimonia}[hp03_098_1]
 \emph{Amaraughaśāsana} 6.1-2 (sic; a single śārdūlavikrīḍita verse is numbered thus)
@@
 \end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp03_098_1]
 
@@
 Verse 3.98*1 is absent in the \textalpha\ group.
-It is very close to a verse in the \textit{Amaraughaśāsana}, which is likely to be its source.
-Its import of rejuvenating the body by flooding it with nectar is not directly connected with the verses proceeding it.
+It is very close to a verse in the \textit{Amaraughaśāsana}, which is likely to be its source, although the date of the \textit{Amaraughaśāsana} is yet to be firmly established.
+This verse's import of rejuvenating the body by flooding it with nectar is not directly connected with those proceeding it.
 
@@
 \begin{translation}[hp03_098_2]
-After making kuṇḍalinị move, the yogi should perform \emph{bhastrī} in particular. The god of death is afraid of the ascetic who regularly practises in this way.
+After making Kuṇḍalinị move, the yogi should perform \emph{bhastrī} in particular. The god of death is afraid of the ascetic who regularly practises in this way.
 \end{translation}
@@
 
-Haṭhasaṅketacandrikā (f. 111v–112r)
+\emph{Haṭhasaṅketacandrikā} (f. 111v–112r)
 \begin{versinnote}
@@
 They elaborate further on the instruction to practise \emph{bhastrī kumbhaka} in 3.97.
-The practice of other \emph{kumbhaka}s for moving \emph{kuṇḍalinī} is mentioned in 3.98*3.
+The practice of other \emph{kumbhaka}s for moving Kuṇḍalinī is mentioned in 3.98*3.
 
@@
 %No known source. the readings \emph{cāpi} and \emph{vāpi} are both possible, but the former is better attested.
+
+%MD: adopt śamano in Pāda d? śamana = Yama.
 \end{philcomm}
@@
 \begin{translation}[hp03_099]
-As a result of fearlessly moving [\emph{kuṇḍalinī}] for a total of two \emph{muhūrta}s (one hour 46 minutes), \emph{kuṇḍalinī}, which is in \emph{suṣumṇā}, is drawn up slightly.
-\end{translation}
-% ?? JB, since there's a lot of Sanskrit in this translation and non-specialists are unlikely to know what a muhūrta is, I think we should translate it as 'for up to an hour and a half.'
+That fearless [yogi] should move [Kuṇḍalinī] for up to one hour 36 minutes (two \emph{muhūrta}s). Suṣumṇā at Kuṇḍalinī is drawn up slightly.
+\end{translation}
 
@@
 \tl{ūrdhvam ākṛṣyate kiṃcit suṣumnā kuṇḍalīgatā ||\\+}
-\tl{\var{22c nirbhayaś ] YL ; nirbharaś T, nirbhayāc GU 23a ākṛṣyate YL ; ākarṣayet TGU 23b suṣumnā kuṇḍalīgatā ] TG; suṣumnāṃ kuṇḍalīgatāṃ U,  suṣumnā kuṇḍalīyutā YL}\\!}
+\tl{\var{22c nirbhayaś ] YL ; nirbharaś T, nirbhayāc GU}\\+}
+\tl{\var{23a ākṛṣyate ] YL ; ākarṣayet TGU}\\+}
+\tl{\var{23b suṣumnā kuṇḍalīgatā ] TG; suṣumnāṃ kuṇḍalīgatāṃ U,  suṣumnā kuṇḍalīyutā YL}\\!}
 \end{versinnote}
@@
 \tl{ṣaṇmāsāc cālanenaiva śaktis tasyordhvagā bhavet ||\\+}
-\tl{\var{kuṇḍalīgatām ] kuṇḍalī gatā \vl}\\!}
+\tl{\var{ākṛṣyate] ākṛṣya tau P, T, t1. kuṇḍalīgatām ] kuṇḍalī gatā P,T,t1}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp03_099]
-%MD: nirbharaṃ vs nirbhayaś discussed?
-% ?? JB we should adopt nirbhayaś as alphaOne, alphaTwo and the source text have it. Yes. JM: also we should adopt alpha reading, so it's suṣumṇā that is lifted up, as in GŚ.
+%?? Add reference to introduction
+As noted in our introduction, Svātmārāma appears not to have understood the practice of \emph{śakticālana} in the same way as his primary source text for its description, the \emph{Gorakṣaśataka}, in which a cloth is wrapped around the tongue so that it can be repeatedly pulled, thereby lifting up the base of the central channel.
+He does not include the \emph{Gorakṣaśataka} verses which mention the tongue or the cloth (but some later recensions of the \emph{Haṭhapradīpikā} do introduce them).
+Verses 98 and 99 suggest that he understood the practice to involve repeated contraction of the region of the sun at the lower end of the central channel.
+The result is the same, namely that Kuṇḍalinī is awakened and uncoils herself, thereby allowing Prāṇa to enter the central channel.
 \end{philcomm}
@@
 \begin{translation}[hp03_100]
-Extracted from \emph{suṣumnā} by this [practice], \emph{kuṇḍalinī} leaves it. As a result of this, \emph{prāṇa} automatically enters \emph{suṣumṇā}.
+Extracted from Suṣumṇā by this [practice], Kuṇḍalinī leaves it. As a result of this, \emph{prāṇa} automatically enters Suṣumṇā.
 \end{translation}
@@
 \end{translation}
-% ?? JB add testimony of J5 to apparatus: śabdagarbhām arundhatīm
+%
 
@@
 %comment on śabdagarbhām aruṃdhatīṃ.
-In the \emph{Gorakṣaśataka} (26cd–27ab), the source text, this verse occurs in a passage on \emph{sarasvatīcālana}, which is the practice of moving the tongue (i.e.~\emph{sarasvatī}) by wrapping a cloth  around it and tugging it in order to raise \emph{kuṇḍalinī}.
+In the \emph{Gorakṣaśataka} (26cd–27ab), the source text, this verse occurs in a passage on \emph{sarasvatīcālana}, which is the practice of moving the tongue (i.e.~\emph{sarasvatī}) by wrapping a cloth around it and tugging it in order to raise Kuṇḍalinī.
 In the \emph{Gorakṣaśataka}, \emph{sarasvatī} is also called \emph{arundhatī} and, since the tongue is instrumental for speech and \emph{sarasvatī} is the name of a Goddess identified with speech (\emph{vāc}), the \emph{Gorakṣaśataka}'s reading of \emph{śabdagarbhāṃ sarasvatīm} makes good sense.
-However, it seems that Svātmārāma has changed 3.101b to read \emph{śabdagarbhām arundhatīm} and has understood \emph{arundhatī} as \emph{kuṇḍalinī}.
-This is affirmed in longer versions of the \emph{Haṭhapradīpikā} (3.102??) that contain a verse on synonyms of \emph{kuṇḍalinī}, which includes \emph{arundhatī}.
-We are yet to find \emph{arundhatī} equated with  \emph{kuṇḍalinī} in a text composed before the \emph{Haṭhapradīpikā} but this identification is found in subsequent compendiums and commentaries (e.g.~\emph{Yogacintāmaṇi} f.~78v, \emph{Yuktabhavadeva} 7.300, \emph{Jyotsnā} 104, 119, \emph{Yogaprakāśikā} 5.166).%??JM: I think Svātmārāma prob understood the practice as the GŚ does, but later scribes etc.
-didn't.
+However, it seems that Svātmārāma has changed 3.101b to read \emph{śabdagarbhām arundhatīm} and has understood \emph{arundhatī} as Kuṇḍalinī.
+This is affirmed in longer versions of the \emph{Haṭhapradīpikā} (3.94*2) that contain a verse on synonyms of Kuṇḍalinī, which includes \emph{arundhatī}.
+We are yet to find \emph{arundhatī} equated with Kuṇḍalinī in a text composed before the \emph{Haṭhapradīpikā} but this identification is found in subsequent compendiums and commentaries (e.g.~\emph{Yogacintāmaṇi} f.~78v, \emph{Yuktabhavadeva} 7.300, \emph{Jyotsnā} 104, 119, \emph{Yogaprakāśikā} 5.166).
 \end{philcomm}
@@
 \begin{translation}[hp03_102]
-The yogi who has made kuṇḍalinī move is worthy of success. There is no point in speaking at length about this. He easily conquers death.
+The yogi who has made Kuṇḍalinī move is worthy of success. There is no point in speaking at length about this. He easily conquers death.
 \end{translation}
@@
 \begin{translation}[hp03_102_1]
-For [the yogi] devoted to celibacy and always eating a healthy and measured diet, success appears after forty days (\emph{maṇḍalāt}) as a result of engaging in the practice of \emph{kuṇḍalinī}.
+For [the yogi] devoted to celibacy and always eating a healthy and measured diet, success appears after forty days (\emph{maṇḍalāt}) as a result of engaging in the practice of Kuṇḍalinī.
 \end{translation}
@@
 This block appears after 3.97 and appears to have been inserted from elsewhere.
-3.102*1 has no known source and appears to have been added as a general laudatory statement on the benefits of practising with \emph{kuṇḍalinī}.
+3.102*1 has no known source and appears to have been added as a general laudatory statement on the benefits of practising with Kuṇḍalinī.
 
@@
 \begin{translation}[hp03_103_1]
-For purifiying the seventy-two thousand channels, there is no method of cleansing without the practice of \emph{kuṇḍalinī}.
+For purifiying the seventy-two thousand channels, there is no method of cleansing without the practice of Kuṇḍalinī.
 \end{translation}
 %greyscaled
-% ??MD: better to adapt abhyasanād ṛte? 
 %\begin{sources}[hp03_103_1]
@@
 \begin{philcomm}[hp03_103_1]
-This verse is omitted by the \textalpha\ group, and was probably added to the original text as a further laudatory statement on the practice of \emph{kuṇḍalinī}.
+This verse is omitted by the \textalpha\ group, and was probably added to the original text as a further laudatory statement on the practice of Kuṇḍalinī.
 \end{philcomm}
@@
 \end{translation}
-%?? JB put iti śakticālana below the line in the apparatus (as its not in alpha)?
 
@@
 \begin{translation}[hp03_107]
-By means of postures, breath retentions and \emph{mudrā}s, the central channel, even though untraversed/blocked, becomes straight through yogis' firm practice.
-\end{translation}
-%??
-JB: khila can mean uncultivated (see note below) but I havent found a reference for the meaning of `blocked' or 'untraversed'?
-JM: I guess untraversed was getting at that meaning; it needs to contrast with sarala; untraversed is quite good in fact.
+By means of postures, breath retentions and \emph{mudrā}s, the central channel, even though untraversed, becomes straight through yogis' firm practice.
+\end{translation}
+%
 
@@
 The reading of \emph{khilāpi} in the first verse quarter is unusual but well attested by the witnesses including the \textalpha\ group of manuscripts.
-In the \emph{Abhidhānacintāmaṇi} (940), \emph{khila} is defined as something uncultivated such as field (\emph{kṣetrādyaprahataṃ khilam}) or, as the \emph{Amarapadavivṛti} (2.1.5) puts it, `not marked by a plough' (\emph{lāṅgalena na likhitam iti khilam}).
-In the context of \emph{Haṭhapradīpikā} 3.107, qualifying \emph{suṣumnā} with \emph{khila} implies that the central channel has yet to be cultivated by a yoga practice.
+In the \emph{Abhidhāna\-cintāmaṇi} (940), \emph{khila} is defined as something uncultivated such as field (\emph{kṣetrādyaprahataṃ khilam}) or, as the \emph{Amarapadavivṛti} (2.1.5) puts it, `not marked by a plough' (\emph{lāṅgalena na likhitam iti khilam}).
+In the context of \emph{Haṭhapradīpikā} 3.107, qualifying \emph{suṣumnā} with \emph{khila} implies that the central channel has yet to be cultivated (i.e., traversed).
 

### 2024-07-02 latex/HP4_comm.tex
new input files

@@
 
-%nāda-bindu-kalā, referring to inner sound, generative fluid, and digits of the moon (or unmanīkalā) in the context of HY? Judit: kalā is like śakti (initial source)?
+%nāda-bindu-kalā, referring to inner sound, generative fluid, and digits of the moon (or unmanīkalā) in the context of HY. Judit: kalā is like śakti (initial source)?
 %Refer to TĀK
@@
 \begin{translation}[hp04_000_7]
-Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation †arises† [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
+Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation is established [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
 \end{translation}
@@
 Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}.
-We are unable to make sense of \emph{sthiti} here, and so have cruxed this verse quarter and emended to \emph{muktiḥ sthitā} (`liberation has happened').
-%??
-JB.
-Rather than emend to muktiḥ sthitā, I think we should consider muktisthitiḥ (etaTwo and deltaOne), which is similar to jīvanmuktasthitiṃ in the Mokṣopāya, LYV, MSS, etc.
-We could translate it as Sternbach below ('state of liberation').
-JM: I also don't think we should emend.
-muktisthitiḥ is possible (but MU parallel not great because of mukti vs mukta).
-We could also consider muktiḥ sthitiḥ “liberation [and] stability”
-% Mokṣopāya
-% jīvanmuktasthitiṃ brahman kṛtvā rāghavam āditaḥ|\\
-% kramāt kathaya me nityaṃ bhaviṣyāmi sukhī yathā||\,1||
-%  "jīvanmuktasthitiṃ" jīvanmuktamaryādāṃ
-% Yogavāsiṣṭha
-% vicārajā nijā jīvanmuktasthitir ihoditā |
-% muninābhyāsahīnasya vyādhasya tvanavasthitiḥ || VII. 
-% MSS_9744-1 
-% kāryākārye kim api satataṃ naiva kartṛtvam asti  
-% jīvanmuktasthitir avagato dagdhavastrāvabhāsaḥ /<BR>
-% There is not always an agent in every action, good or bad ; for, the state of a jivanmukta [one who is released from bondage while living in the world] is considered to be similar to that of the splendour of a burnt cloth [which can neither be clean nor dirty].
-(Sternbach's translation)
+The compound \emph{muktisthitiḥ} (literally, `the establishment of liberation') has been adopted.
+%  attested by \etaTwo, 
+% adopt muktisthitiḥ and rewrite note. 
+%
 \end{philcomm}
@@
 \begin{philcomm}[hp04_000_9]
-In \emph{Jyotsnā} 4.10, Brahmānanda understands `the void' (\emph{śūnya}) as the central channel. In \emph{Haṭhapradīpikā} 3.4, \emph{śūnyapādavī} is a synonym of \emph{suṣumnā}.
-\end{philcomm}
-%?? MD: vilīyate or pralīyate?
+In \emph{Jyotsnā} 4.10, Brahmānanda understands `the void' (\emph{śūnya}) as the central channel. In \emph{Haṭhapradīpikā} 3.4, \emph{śūnyapādavī} is a synonym of Suṣumnā.
+\end{philcomm}
 
@@
 \begin{translation}[hp04_000_10]
-For the yogi whose \emph{kuṇḍalinī} has awakened and who has given up all activity, the innate state automatically arises.
+For the yogi whose Kuṇḍalinī has awakened and who has given up all activity, the innate state automatically shines forth.
 \end{translation}
@@
 
-%?? MD: prajāyate or prakāśate?
+%
 
@@
 \begin{translation}[hp04_000_11]
-When the breath is flowing in the central channel and the mind enters the void, the knower of yoga uproots all [past] actions.
-\end{translation}
-%?? JB: all [fruits of past] actions
+When the breath is flowing in the central channel, and the mind enters the void, the knower of yoga destroys all actions.
+\end{translation}
+%?
 
@@
 %It appears that this verse, which is in manuscripts of the \textbeta\ and \textepsilon\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
-%?? JB should we include this note on gamma's omission? Is it necessary? JM: no, I've percented it
+%
 
@@
 \begin{translation}[hp04_000_13]
-When the mind has obtained equanimity and the breath is moving into the central channel, then the \emph{vajrolī} of the lineage of immortals arises even for the person whose hope is for that.
-\end{translation}
-% ??
-JB: we didnt translate this verse in the workshop.
-I've adapted my translation from a similar verse in the Amaraugha.
-JM: equanimity is tautologous with mind, “When equanimity has been obtained"?
-Also I've changed tr.
-to reflect tadā.
+When equanimity has been obtained, and the breath is moving into the central channel, then the \emph{vajrolī} of the lineage of immortals arises †even for the person whose hope is for that†.
+\end{translation}
+%
 
@@
 \tl{tadāmaraughavajrolis tadāśājīvitasya ca ||\\+}
-\tl{\var{citte tu sattvam ] $\beta$ : citte samatvam Ba Ad Ed : cittau samatvam Gb Ta.
+\tl{\var{citte tu sattvam ] Ga, Ae (=\emph{Amaraugha}) : citte samatvam Ba Ad Ed : cittau samatvam Gb Ta.
 tadāmaraughavajrolis ] \textit{conj.} :
@@
 eṣāmarolī vajrolī Ed.
-tadāśājīvitasya ca ] $\beta$ :
+tadāśājīvitasya ca ] Ga, Ae (=\emph{Amaraugha}) :
 tadā meni mameti ca Ba :
@@
 The second line of this verse is problematic in both the source text (the \emph{Amaraugha}) and the manuscripts of the \emph{Haṭhapradīpikā}.
-It seems that the \textepsilon\ group has preserved the most plausible version of this line, which is also quite close to the source text.
-The line was rewritten in \etaTwo\ and other manuscripts, as well as the \emph{Jyotsnā} (4.14cd), to include the two variations of \emph{vajrolī}, namely \emph{amarolī} and \emph{sahajolī}.%??JM: comment on pāda d?
-It's odd, worth cruxing even.
+For the last \emph{pāda}, we have adopted the closest reading to that of the source text (the \emph{Amaraugha}) but cannot make sense of it so have cruxed it.
+The line was rewritten in \etaTwo\ and other manuscripts, as well as the \emph{Jyotsnā} (4.14cd), to include the two variations of \emph{vajrolī}, namely \emph{amarolī} and \emph{sahajolī}.%
 %C2 (Grp 7) tadāmarolī ... sahajolī prajāyate (=Jyotsnā)
@@
 So long as the breath lives, the mind does not die.
-Breath and mind: he who leads this pair to dissolution attains liberation; in no way [does any] other man.
+Breath and mind: [when] this pair goes to dissolution, a man attains liberation; in no way [does any] other [man].
 \end{translation}
@@
 %\end{philcomm}
-%?? MD: adopt prayāti (beta) in Pāda C?
 
@@
 \end{translation}
-%??
-JB: Both \alphaOne and \alphaTwo have atha samādhiḥ as a heading before this verse.
-Shouldnt we include this in our edition or at least note it in the apparatus?
-We've added atha headings elsewhere from alpha.
-There's also the issue of where to place atha samādhi in relation to the greyscale verses.
+%
 
@@
 \begin{translation}[hp04_001_1]
-Whether or not this [dissolution] is called liberation  in another teaching, there is no difference in the bliss arising from the dissolution of mind and breath.
-\end{translation}
-%?? JB: 'in another doctrine' for matāntare?
-%?? JB: extraordinary bliss (kaścid?)
+Whether or not this [dissolution] is called liberation in another school, an extraordinary bliss from the dissolution of mind and breath arises in me.
+\end{translation}
 %\begin{sources}[hp04_001_1]
@@
 \begin{philcomm}[hp04_001_1]
-Some of the \textbeta, \textepsilon\ and \texteta\ groups have a different version of the second half of this verse, the most plausibe of which is, 'the extraordinary bliss of dissolution of mind and breath arises in me' (\emph{manaḥprāṇalayānando mayi kaścit pravartate}).
-The first person pronoun suggests that this verse derives from a dialogue of some kind before it was changed to the version in the \textgamma\ and \textdelta\ manuscripts.
-\end{philcomm}
-
-% ?? MD: comment on the variant reading "manaḥprāṇalayānando mayi kaścit pravartate"?
-% JB: maybe this is the lectio difficilor and the first person pronoun was later written out.
+The manuscripts attest two versions of the second line.
+It appears that the reading of \textbeta, \textepsilon\ and \texteta\ groups, which we have adopted, was changed to remove the first person pronoun.
+The modified reading is prevalent in the \textgamma\ and \textdelta\ manuscripts (\emph{manaḥprāṇalayānando nāpi kaś cit vibhidyate}).
+\end{philcomm}
+%
 
@@
 \begin{translation}[hp04_002]
-A motionless and unchanging dissolution [of mind], in which inhalation and exhalation have disappeared [and] perception of sense objects has ceased, is supreme for yogis.%??JM: or "The yogis’ motionless and unchanging dissolution...
-is the best."
-\end{translation}
-% on the translation of jayati as an intransitive verb, see Alexis' footnote on jayati in his forthcoming book on the Tantrāloka āhnika 1
+The yogis' motionless and unchanging dissolution [of mind], in which inhalation and exhalation have disappeared [and] perception of sense objects has ceased, is supreme.
+\end{translation}
 
@@
 prāṇastho śvāsa° Pb,
-pranaṣṭe svāsa° K}\\!} % retained because we've read against alpha.??jM: simplify? It's only the ucch that's different. JB: done.
+pranaṣṭe svāsa° K}\\!} % retained because we've read against alpha.
 %\tl{\var{hy ānandaṃ yāti yogavit ]  AllNI (except  Ma Pa Pb Pc Tr Va),
@@
 %layo jayati yoginaḥ Ad,
-%layo jayati yoginām Cc}\\!} %retained to show that jayati was in the South Indian transmissin of the Amanaska.??JM: but it's in almost all HP witnesses too so perhaps variants not needed here, just mention it in comm?
-Or simplify?
-JB done
+%layo jayati yoginām Cc}\\!} %retained to show that jayati was in the South Indian transmissin of the Amanaska.
 \end{versinnote}
@@
 \begin{philcomm}[hp04_002]
-%Adopt alpha reading. [MD: Done.]
 The fourth verse quarter of this verse differs from that of the North Indian recension of the \emph{Amanaska} (quoted as the source) but is found in two South Indian manuscripts of the \emph{Amanaska} (Ad and Cc).
@@
 \begin{translation}[hp04_003]
-May an extraordinary absorption prevail, in which all volition has been cut off and all activity ceased, intelligible [only] by means of itself and ineffable.%??JM: which is intelligible > intelligible, and other tweaks.
-Prevail not great but can't think of anything else.
+May an extraordinary absorption reign supreme, in which all volition has been cut off and all activity ceased, intelligible [only] by means of itself and ineffable.%
 %% My translation in the Amanaska (reading jāyate): [This] extraordinary absorption by which all volition has been cut off and in which all movement has ceased, is intelligible [only] by means of itself and is beyond the scope of words.
@@
 \begin{translation}[hp04_004]
-Dissolution, which is eternal in the elements and senses, takes place where the gaze is.
-The gaze that has dissolved into the focus becomes the power of living beings.%%??JM: changed meditative focal point to focus (as tr.
-in v.6)
-%JB 'occurs' for 'takes place'
+Dissolution, which is eternal in the elements and senses, occurs where the gaze is. The gaze that has dissolved into the focus becomes the power of living beings.%%
+%
 \end{translation}
@@
 \tl{svaśaktijīvabhūtā hi dṛṣṭilakṣair layaṃ gatāḥ \\+}
-\tl{°bhūtā hi ] conj. Hatley, °bhūtāni ABCed\\!} 
-% ?? Consult Shaman's critical edition.
+\tl{\var{°bhūtā hi ] conj. Hatley, °bhūtāni ABCed}\\!} 
+% 
 \end{versinnote}
@@
 \begin{translation}[hp04_006]
-The focus internal, the gaze external, unblinking: this is the \emph{śāmbhavī mudrā} concealed in all the Tantras.%??JM: tr. tweaked
-% JB my preference would be: The focus is internal [yet] the gaze is external and free from blinking.  
+The focus is internal, the gaze external, unblinking: this is the \emph{śāmbhavī mudrā} concealed in all the Tantras.%
+
 \end{translation}
@@
 \tl{eṣā hi śāmbhavī mudrā sarvatantreṣu gopitā ||\\+}
-\tl{\var{antar lakṣyaṃ ] AllNI (except Ja K Pb Tr Ua Vb) Tha Vd, antarlakṣya Ad Cc K: antarlakṣaṃ Tr Nw Ve: antarlakṣā Ja Ua Ea Eb, antarlakṣo Na Nq Nu, antarlakṣyo Nb Nm: antaryogaṃ Vb: antarlīno Pb}\\!}%??
-JM: delete variants?
+\tl{\var{antar lakṣyaṃ ] AllNI (except Ja K Pb Tr Ua Vb) Tha Vd, antarlakṣya Ad Cc K: antarlakṣaṃ Tr Nw Ve: antarlakṣā Ja Ua Ea Eb, antarlakṣo Na Nq Nu, antarlakṣyo Nb Nm: antaryogaṃ Vb: antarlīno Pb}\\!}%JM: delete variants?
 % JB: the lakṣyaṃ/lakṣa spelling is an issue in both the Amanaska and HP mss (inlcuding alpha).
@@
 \begin{philcomm}[hp04_006]
-See Birch 2013: 287 for more parallel verses.
+As seen in the witnesses of the \emph{Amanaska}, the source text, the spellings \emph{lakṣya} and \emph{lakṣa} occur randomly in the manuscripts. See Birch 2013: 287 for more parallel verses.
 \end{philcomm}
@@
 One would expect a verse on \emph{khecarī mudrā} to be explaining the insertion of the tongue in the nasopharyngeal cavity (cf.~3.33–48), but the practice of meditating by fixing the gaze is called \emph{khecarī mudrā} in 4.7 (as attested by \textalpha, \textbeta\ and \texteta\ manuscripts) and manuscripts of the \emph{Candrāvalokana}, the text from which Svātmārāma borrowed this verse.
-It is therefore likely that he added 4.8 to explain that \emph{śāmbhavī} and \emph{khecarī mudrās} are the same with respect to the gaze.
-
+It is, therefore, likely that he added 4.8 to explain that \emph{śāmbhavī} and \emph{khecarī mudrās} are the same with respect to the gaze.
+A similar conflation occurs in the \textit{Śivayogapradīpikā} (5.3).
+% NJ: Maybe mention Śivayogapradīpikā 5.3 and 5.4 
+%% 5.3ab lakṣyalīnamanasānilena yo vartate 'calitatārako bhavet |
+%% 5.3cd khecarīyam atha saiva śāṃbhavīmudrayā sakṛtayā jagadguruḥ 
+%% (5.3) For one whose eyes remain unmoving [while open], with the mind and breath dissolved in the [inner] lakṣya, this khecarī indeed is śāmbhavī.
+Through the proper practice of this mudrā, he shall be the guru of the world.
+%% 5.4ab loke 'antaḥkhecarī mudrām ajñātvā bāhyakhecarīm
+%% 5.4cd avalambya tvarāḥ kecij jihvāchedanakarmaṇā
+%% (5.4) Not knowing the inner khecarīmudrā in this world, some have hastily taken to the external khecarī through the practice of cutting the tongue.
+%%%%
 %Judit would greyscale 4.8, thinking that it is a marginal note. And greyscale the pātāle verses.
@@
 \tl{\var{unnamayed ] Cb Vd: unnamayan Bb Tha: unmīlaye W: unmilayet Uc: unmanaya Pe: unmīlya Cc}\\!}
-%\tl{\var{unmanīkārakaḥ kṣaṇāt ] Bb Tha W Vd Pe: unmunikārakakṣaṇāt Uc: unmanīkāraṇaṃ kṣaṇāt Cc: unmanākāśakaḥ kṣaṇāt Ad Ca: unmanākāśakakarṣaṇāt Cb}\\!}%??
-variants needed?
-JB I've deleted the variants for the last pāda but we need those on unnamayed (see below)
+%\tl{\var{unmanīkārakaḥ kṣaṇāt ] Bb Tha W Vd Pe: unmunikārakakṣaṇāt Uc: unmanīkāraṇaṃ kṣaṇāt Cc: unmanākāśakaḥ kṣaṇāt Ad Ca: unmanākāśakakarṣaṇāt Cb}\\!} 
 \end{versinnote}
@@
 \begin{philcomm}[hp04_008_1]
-%MD: Better to adopt unnamayed and kārakaḥ kṣaṇāt?
-% ??
-JB Since unnamayed in the beta witnesses and source, and makes better sense, I think we should adopt.
-The dictionary states that uc-cal means 'to go or move away from; to free or loosen one's self from', which is not so good here.
-
 This and the next verse (4.8*1–2) are from the South Indian recension of the \emph{Amanaska}, a later rewriting of that work which incorporates these verses in an additional passage on Tārakayoga, added here to elaborate on the meditation of fixing the gaze that is taught in 4.4–7.
@@
 \begin{translation}[hp04_008_2]
-Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning.
-They do not know what causes one to cross over.%??JM: put Tāraka and explain in note?
-Tārakayoga is mentioned in the previous note.
-% JB in the source 
-\end{translation}
-% Maybe grey scale
+Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
+\end{translation}
 
@@
 \begin{philcomm}[hp04_008_2]
+In the source text \emph{tārakam} refers to  Tāraka yoga, one of two yogas taught in the South Indian recension of the \emph{Amanaska}.
+
 On why this verse is in greyscale, see the note to 4.8*1.    
-% ??JB we should adopt nigama in pāda b because it is attested by several manuscript groups and the source.
 \end{philcomm}
@@
 \begin{versinnote}
-\tl{ardhotghāṭitalocana[ḥ] sthiramanā nāsāgradattekṣaṇaś\\+}
+\tl{ardhodghāṭitalocana[ḥ] sthiramanā nāsāgradattekṣaṇaś\\+}
 \tl{candrārkāv api līnatām upagatau niṣpandarūpaṃ vapuḥ |\\+}
@@
 \end{philcomm}
-% ?? MD: I have changed the verse order according to the epsilon group. Please adjust the commentary. JB Done
 
@@
 
-%\begin{philcomm}[hp04_008_7]
-%?? JM: nirodhataḥ reading much better. Early corruption?
-% JB I think we should note nirodhataḥ as the better reading in the comments but adopt divārātrau (or divārātraṃ) na pūjayet (he should not worship at day and night (i.e., when prāṇa is moving in the sun and mmoon channels)).
-%\end{philcomm}
+\begin{philcomm}[hp04_008_7]
+The manuscripts transmit readings for the last verse quarter that either contradict or repeat the statements in the first line.
+This problem likely occurred through some kind of dittographical error.
+We have therefore adopted the reading of \emph{Jyotsnā} 4.42d (°nirodhataḥ), which is the same as the source text.
+\end{philcomm}
 
@@
 \end{translation}
-% ?? compound khecarīmudrā in 4.8*8c and 4.8*9c?
-
+%
 %\begin{sources}[hp04_008_8]
@@
 
-\begin{testimonia}[hp04_008_9]
-\emph{Haṭhasaṅketacandrikā} f.~??
-\begin{versinnote}
-\tl{savyadakṣiṇanāḍīstho madhye carati mārutaḥ |\\+}
-\tl{tiṣṭhati khecarī mudrā tasmin sthāne na saṃśayaḥ ||\\!}
-\end{versinnote}
-\end{testimonia}
+%\begin{testimonia}[hp04_008_9]
+%\emph{Haṭhasaṅketacandrikā} f. 246v (Mysore ORI Manuscript B220) 
+%\begin{versinnote}
+%\tl{savyadakṣiṇanāḍīstho madhye carati mārutaḥ |\\+}
+%\tl{tiṣṭhati khecarī mudrā tasmin sthāne na saṃśayaḥ ||\\!}
+%\end{versinnote}
+%\end{testimonia}
 
@@
 
-% %\emph{Skandapurāṇa} 4.41.45%??JM: is it worth putting this in? Verse is prob found all over the place and this isn't the old SP. Just YCM will do.
+% %\emph{Skandapurāṇa} 4.41.45
 % %\begin{versinnote}
@@
 % \begin{testimonia}[hp04_008_10]
-% \emph{Yogacintāmaṇi} ?? [attrib. Skandapurāṇa]
+% \emph{Yogacintāmaṇi} [attrib. Skandapurāṇa]
 % %\begin{versinnote}
@@
 \end{translation}
-%?? JB: pada here means place/site (level sounds strange)
-% abode of Śiva ... that place should be known
+%
 
@@
 bhruvor madhye śivasthānaṃ manas tatra vilīyate |
-jñātavyaṃ tatparaṃ turyaṃ tatra kālo na vidyate ||
+jñātavyaṃ tat paraṃ turyaṃ tatra kālo na vidyate ||
 \end{testimonia}
@@
 \begin{philcomm}[hp04_008_17]
-The \emph{vyomacakra} ('the cakra of space') is also mentioned in a half-verse that was added to some later recensions of the \emph{Haṭhapradīpikā} (see 3.35) and states that \emph{vyomacakra} is another name for \emph{khecarīmudrā}.
+The \emph{vyomacakra} (`the cakra of space') is also mentioned in a half-verse that was added to some later recensions of the \emph{Haṭhapradīpikā} (see 3.35) and states that \emph{vyomacakra} is another name for \emph{khecarīmudrā}.
 In \emph{Jyotsnā} 4.45, Brahmānanda states that the \emph{vyomacakra} is associated with all the voids in the middle of the brow (\emph{bhrūmadhye sarvakhānāṃ samanvayāt}) and, in the \emph{Haṭhasaṅketacandrikā} (f.
-129v), Sundaradeva says that it is called the Brahmarandhra, which is between Iḍā and Piṅgalā (\emph{iḍāpiṅgalāntargataṃ brahmarandhrākhyaṃ
-vyomacakraṃ tat khecarīmudrāṃ śaktiṃ kuryād ...}). 
-%?? Write a note about epsilon 3 vyomacakram vyavasthitaṃ: Cf.~the greyscale verse in 3.35.
+129v), Sundaradeva says that it is called the Brahmarandhra, which is between Iḍā and Piṅgalā (\emph{iḍāpiṅgalāntargataṃ brahmarandhrākhyaṃ vyomacakraṃ tat khecarīmudrāṃ śaktiṃ kuryād ...}).
+%
 \end{philcomm}
@@
 \begin{translation}[hp04_008_19]
-When the external air has dissolved into the void, the breath is sure to go to its place †with the mind on the side of the sun†.
+Just as the external air has dissolved into the void, the breath is sure to go to its place †with the mind on the side of the sun†.
 %When the external air has dissolved into the void, the breath surely goes to its own place, † and the sun together with the mind to the fire.†
@@
 Since we are not sure of the meaning of \emph{sūryāṅge} and since the variant readings with \emph{pavana} are not clear either, we have cruxed the fourth verse quarter.
-
-%\textit{Maitrī Upaniṣad} 6.1-2 might provide a clue to the meaning of \emph{sūryāṅga}.
-This Upaniṣad explains that the \textit{ātman} manifests itself in two ways: as \textit{prāṇa} internally and as the sun externally.
-\textit{Prāṇa} is identified as the inner aspect of the \textit{ātman}.
-It is associated with breath, etc.
-The sun is depicted as the external manifestation of the \textit{ātman}.
-It symbolizes the cosmic source of energy, etc.
-The text emphasizes that the movements of \textit{prāṇa} and the sun are synchronized and states that the golden \textit{puruṣa}, who dwells in the sun, also resides in the heart as \textit{prāṇa}.
-If one is willing to accept a connection between this verse and the concept mentioned in the \textit{Maitrī Upaniṣad}, then \textit{sūryāṅge} possibly has even two layers of meaning.
-On the microcosmic level, \textit{prāṇa} would return home to the heart.
-Moreover, on the macrocosmic level, the \textit{ātman} would return to its macrocosmic home - the sun.
-% ?? Does sūryāṅge occur in the Maitrī Upaniṣad? If not, the connection with this Upaniṣad seems tenuous. 
-\end{philcomm}
-
-%?? MD: yadā (epsilon) or yathā (beta)?
-% JB separate yadā līnaḥ in the edition (yadā seems easier to understand than yathā)
+\end{philcomm}
+
+%
 
@@
 \end{testimonia}
-%??JM: I can't find these HSC refs
 
@@
 \end{testimonia}
-%??JM: can't find. ref?
 
@@
 \begin{translation}[hp04_008_23]
-Put the self in space and put space in the self. [The yogi] should make everything consist of space and think of nothing at all.
+Put the self in space and put space in the self. [The yogi] should make the self consist of space and think of nothing at all.
 \end{translation}
@@
 \begin{sources}[hp04_008_23]
-\emph{Uttaragītā} 1.9%??JM: bibliog for texts cited — where?
+\emph{Uttaragītā} 1.9
 \begin{versinnote}
@@
 %\begin{philcomm}[hp04_008_23]
-%?? Pāda c: sarvaṃ or ātmānaṃ?
-%\end{philcomm}
-% J up to here! 11 March
+% Critical edition of the Mbh: 
+% 06,040.078d@003A_0073 ātmānaṃ khamayaṃ kṛtvā na kiṃ cid api cintayet || (36) Add to testimonia khamadhye kuru cātmānam ātmamadhyaṃ ca khaṃ kuru || (35)
+%06,040.078d@003A_0072 khamadhye ca praveṣṭavyaṃ khaṃ ca brahma sanātanam |
+%06,040.078d@003A_0073 ātmānaṃ khamayaṃ kṛtvā na kiṃ cid api cintayet || (36)
+%\end{philcomm}
 
@@
 \begin{translation}[hp04_008_24]
-Like an empty pot in air, [the yogī] is empty on the inside and empty on the outside.
-Like a full pot in the ocean, [the yogi] is full on the inside and full on the outside.%??JM: I changed the sky to air.
-Not sure, but I don't think we are meant to think of a pot floating in space.
+Like an empty pot in air, [the yogī] is empty on the inside and empty on the outside. Like a full pot in the ocean, [the yogi] is full on the inside and full on the outside.%
 \end{translation}
@@
 \begin{testimonia}[hp04_008_24]
-\emph{Yogacintāmaṇi} (attrib.~Vasiṣṭha, = \emph{Laghuyogavāsiṣṭha} 15.79)
-\begin{versinnote}
-\tl{antaḥśūnyo bahiḥśūnyaḥ śūnyakumbha ivāmbare |\\+}
-\tl{antaḥpūrṇo bahiḥpūrṇaḥ pūrnakumbha ivārṇave ||\\!}
-\end{versinnote}
-
 \emph{Haṭharatnāvalī} 4.46
@@
 \end{versinnote}
+
+%\emph{Yogacintāmaṇi} f. 47v (attrib.~Vasiṣṭha, = \emph{Laghuyogavāsiṣṭha} 15.79)
+%\begin{versinnote}
+%\tl{antaḥśūnyo bahiḥśūnyaḥ śūnyakumbha ivāmbare |\\+}
+%\tl{antaḥpūrṇo bahiḥpūrṇaḥ pūrnakumbha ivārṇave ||\\!}
+%\end{versinnote}
+
 \end{testimonia}
@@
 A construct of mere ideation is an affectation of the mind.
-Jettison thought that is only ideation, take refuge in a resolve that is free of ideation, and obtain peace, o Rāma.
+Jettison thought that is only ideation, take refuge in a resolve that is free of ideation, and obtain peace, O Rāma.
 \end{translation}
@@
 \begin{translation}[hp04_008_27]
-Just as camphor in fire and salt in water, so the mind, when being joined with the highest reality, dissolves.
+Just as camphor in fire and salt in water, so the mind, on being brought into contact with the highest reality, dissolves into it.
 \end{translation}
@@
 \tl{karpūram anale yadvat saindhavaṃ salile yathā |\\+}
-\tl{tathā sandhīyamānaṃ hi manas tatraiva līyate||\\!}
+\tl{tathā sandhīyamānaṃ hi manas tatraiva līyate ||\\!}
 \end{versinnote}
@@
 %\begin{philcomm}[hp04_008_27]
-%?? pāda c: hi better. Not in HP sources only HR. Adopt?
 %\end{philcomm}
@@
 \begin{sources}[hp04_008_29]
-Gauḍapāda's \emph{Māṇḍūkyopaniṣatkārikā} 3.31
+\emph{Amanaska} 2.79
+\begin{versinnote}
+\tl{manodṛśyam idaṃ sarvaṃ yat kim cit sacarācaram |\\+}
+\tl{manaso hy unmanībhāve 'dvaitabhāvaṃ pracakṣate ||\\!}
+\end{versinnote}
+
+Cf. Gauḍapāda's \emph{Māṇḍūkyopaniṣatkārikā} 3.31
 \begin{versinnote}
@@
 
-\emph{Amanaska} 2.79
-\begin{versinnote}
-\tl{manodṛśyam idaṃ sarvaṃ yat kim cit sacarācaram |\\+}
-\tl{manaso hy unmanībhāve 'dvaitabhāvaṃ pracakṣate ||\\!}
-\end{versinnote}
 \end{sources}
@@
 \begin{testimonia}[hp04_008_29]
-\emph{Yogacintāmaṇi} (attr.~to the \emph{Amanaska})
+\emph{Yogacintāmaṇi} f. 27r (attr.~to the \emph{Rājayoga}, aka. \emph{Amanaska})
 \begin{versinnote}
@@
 \begin{translation}[hp04_008_30]
-As a result of abandoning the things that are to be known the mind attains dissolution (\emph{layam}). When the mind has attained dissolution, isolation (\textit{kaivalyam}) remains.
-\end{translation}
-
-\begin{sources}[hp04_008_30]
-\end{sources}
+As a result of abandoning the things that are to be known, the mind attains dissolution. When the mind has attained dissolution, liberation (\textit{kaivalyam}) remains.
+\end{translation} %
+
+%\begin{sources}[hp04_008_30]
+%\end{sources}
 
@@
 
-\emph{Yogasārasaṅgraha} ??
-\begin{versinnote}
-\tl{śrīdattaḥ |\\+}
+\emph{Yogasārasaṅgraha} p.~52 (attr. to Śrīdatta)
+\begin{versinnote}
 \tl{layo laya iti prāhur īdṛśaṃ lakṣaṇaṃ sphuṭam |\\+}
@@
 \begin{translation}[hp04_008_33]
-Homage to Suṣumṇā, to Kuṇḍalinī, to the nectar arising from the moon, to the mind beyond mind state, to you whose nature is consciousness, the great Śakti.%??
-or whose nature is the consciousness of the great Śakti?
-JM: I have forgotten how we are writing nāḍīs etc.
-I like how we have everything here.
+Homage to Suṣumṇā, to Kuṇḍalinī, to the nectar arising from the moon, to the mind beyond mind state, to you whose nature is consciousness, the great Śakti.%
 \end{translation}
@@
 \begin{translation}[hp04_008_34]
-The cultivation of the inner sound taught by Gorakṣa [and] approved even for deluded people for whom realisation of the highest reality is impossible is now taught.
-\end{translation}
+The cultivation of the inner sound taught by Gorakṣanātha [and] approved even for deluded people for whom realisation of the highest reality is impossible is [now] taught.
+\end{translation} %
 
@@
 \begin{translation}[hp04_009]
-The twelve and a half million methods of dissolution taught by  glorious Śiva are victorious.
-We think only the singular concentration on the inner sound is the most venerable of [all] the dissolutions.%??JM: How about "Glory to..." instead of "are victorious"?
-And then ""We consider one of the dissolutions in particular to be especially worthy of honour, concentration on the inner sound."
+The twelve and a half million methods of dissolution taught by glorious Śiva are the best.
+We consider one of the dissolutions in particular to be especially worthy of honour, concentration on the inner sound.
 \end{translation}
@@
 \tl{nādānusandhānasamādhim ekaṃ\\+}
-\tl{manyāmahe mānyatamaṃ layānām ||\\+}
-\tl{\var{lasantu ] Sa : lasanti Pa : vasanti Ad W Eb Ec Ed : ca santi  Ea : hi sarva Ma}\\+}
-\tl{\var{loke ]  Ma  W Eall : bhūmau  Ad Pa Sa}\\!}
+\tl{manyāmahe mānyatamaṃ layānām ||\\!}
+%\tl{\var{lasantu ] Sa : lasanti Pa : vasanti Ad W Eb Ec Ed : ca santi  Ea : hi sarva Ma}\\+}
+%\tl{\var{loke ]  Ma  W Eall : bhūmau  Ad Pa Sa}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_009]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī}~1.12
+\begin{versinnote}
+\tl{śrīādināthena sapādakoṭi-\\+}
+\tl{layaprakārāḥ kathitā jayantu  |\\+}
+\tl{nādānusandhānakam eva kāryaṃ\\+}
+\tl{manyāmahe mānyatamaṃ layānām ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 1.12
-\begin{versinnote}
-\tl{śrīādināthena sapādakoṭi-\\+}
-\tl{layaprakārāḥ kathitā jayantu  |\\+}
-\tl{nādānusandhānakam eva kāryaṃ\\+}
-\tl{manyāmahe mānyatamaṃ layānām ||\\!}
-\end{versinnote}
 \end{testimonia}
@@
 \begin{testimonia}[hp04_010]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭhasaṅketacandrikā} (ms.~2244 f.~124r)
+\emph{Haṭhasaṅketacandrikā} f.~124r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
 \tl{muktāsanasthito yogī mudrāṃ saṃdhāya śāṃbhavīṃ [|]\\+}
-\tl{śṛṇuyād dakṣiṇe karṇe nādam ekāṃtike sudhīḥ  [||]\\+}
+\tl{śṛṇuyād dakṣiṇe karṇe nādam ekāntike sudhīḥ  [||]\\+}
 \tl{\var{dakṣiṇe karṇe ] B220,  dakṣirṇe 2244}\\+}
@@
 \begin{philcomm}[hp04_010]
-We read this verse here rather than at 4.50 because it follows on from the Śāmbhavī section and is unnecessary and repetitive at 4.50 because 4.51-52 explain the practice.
+We read this verse here (as attested by \textalpha, etc.) rather than at 4.50 because it follows on from the Śāmbhavī section and is unnecessary and repetitive at 4.50 because 4.51-52 explain the practice.
 \end{philcomm}
@@
 \begin{translation}[hp04_011]
-A fire that has been set on wood disappears together with the wood; the mind set on the [inner] sound dissolves together with the [inner] sound.
+A fire that has been set on wood disappears together with the wood; the mind set on the inner sound dissolves together with the inner sound.
 \end{translation}
@@
 \begin{testimonia}[hp04_011]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.15
+\begin{versinnote}
+\tl{kāṣṭhe pravartito vahniḥ kāṣṭhena saha līyate |\\+}
+\tl{nāde pravartitaṃ cittaṃ nādena saha śāmyati ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.15
-\begin{versinnote}
-\tl{kāṣṭhe pravartito vahniḥ kāṣṭhena saha līyate |\\+}
-\tl{nāde pravartitaṃ cittaṃ nādena saha śāmyati ||\\!}
-\end{versinnote}
-
-\emph{Haṭhasaṅketacandrikā} (ms. 2244 f.~124r)
+\emph{Haṭhasaṅketacandrikā} f.~124r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{testimonia}[hp04_012]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-Upāsanāsārasaṅgraha (p. 106)
+\emph{Haṭhasaṅketacandrikā} f.~124r (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{vismṛtya sakalaṃ bāhyaṃ nāde dugdhāṃbuvan manaḥ |\\+}
+\tl{ekībhūyātha sahasā cidākāśe vilīyate ||\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~106 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \tl{ekībhūtaṃ tanyā cittaṃ rājayogābhidānakaṃ ||\\!}
-\end{versinnote}
-
-\emph{Haṭhasaṅketacandrikā} (ms. 2244 f.~124r)
-\begin{versinnote}
-\tl{vismṛtya sakalaṃ bāhyaṃ nāde dugdhāṃbuvan manaḥ |\\+}
-\tl{ekībhūyātha sahasā cidākāśe vilīyate ||\\!}
 \end{versinnote}
@@
 \begin{translation}[hp04_013]
-Having become intent on indifference through regular practice, the ascetic should immediately concentrate on nothing but the inner sound, which brings about the [state] beyond mind.%??JM: sadyas is odd.
-Apparently it has an archaic meaning of "daily".
-Consider?
-Immediately doesn't make sense.
-Or take it with unmanīkaraṇam.
-And sadābhyāsena could go with avadhārayet.
+Having become intent on indifference through regular practice, the ascetic should concentrate on nothing but the inner sound, which immediately brings about the [state] beyond mind.%.
 \end{translation}
@@
 \begin{testimonia}[hp04_013]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭhatattvakaumudī} 54.38
+\emph{Haṭhasaṅketacandrikā} f.~124r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
 \tl{audāsīnyaparo bhūtvā sadābhyāsena saṃyamī |\\+}
-\tl{unmanīkārakaṃ sadyo nādam eva sadābhyaset ||\\!}
+\tl{unmanīkārakaṃ sadyo nādam evāvadhārayet ||\\+}
+\tl{\var{evāvadhārayet ] emend., evaṃ vadhārayet ms. 2244}\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~106 (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{audāsīnyaparo bhūtvā sadābhyāsena saṃyamī |\\+}
+\tl{unmanīkaraṇam sadyo nādam evānu[...]yet ||\\!}
 \end{versinnote}
@@
 
-\emph{Upāsanāsārasaṅgraha} p.~106
-\begin{versinnote}
-\tl{audāsīnyaparo bhūtvā sadābhyāsena samyamī |\\+}
-\tl{unmanīkaraṇam sadyo nādam evānu [...] yet ||\\!}
-\end{versinnote}
 \end{testimonia}
@@
 What kind of indifference?\\
-In the cold season, [indifference towards] whether [one has] an open tent or a quilt, with regard to good nourishment, whether it is cow's milk or water, with regard to food, whether it is lots of alms or forest roots, and with regard to the vessel for food, whether it is the hand, bucket or beggar's bowl.
+In the cold season, [indifference towards] whether [one has] an open tent or a quilt, with regard to good nourishment, whether it is cow's milk or water, with regard to food, whether it is lots of alms or forest roots, and with regard to the vessel for food, whether it is the hand or some kind of bucket.
+%
 \end{translation}
@@
 \begin{testimonia}[hp04_014]
-\emph{Yogacintāmaṇi} ref??
+\emph{Haṭharatnāvalī} 4.7
+\begin{versinnote}
+\tl{śīte kāle dvau paṭī vā paṭī vā pathyāhāre gopayo vā payo vā |\\+}
+\tl{bhakṣye bhojye vṛttim āraṇyakaṃ vā pāṇī droṇī ko 'pi vā bhakṣyapātre ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.7
-\begin{versinnote}
-\tl{śīte kāle dvau paṭī vā paṭī vā pathyāhāre gopayo vā payo vā |\\+}
-\tl{bhakṣye bhojye vṛttim āraṇyakaṃ vā pāṇī droṇī ko 'pi vā bhakṣyapātre ||\\!}
-\end{versinnote}
-
 \emph{Haṭhatattvakaumudī}  54.39
@@
 \begin{philcomm}[hp04_014]
-On \emph{caupaṭī} in the first \emph{pāda}, McGregor (1994:s.v.) and Callewaert (2009:s.v.) in their dictionaries of modern and old Hindi both give the meaning "open all around" for \emph{caupaṭ}.
+On \emph{caupaṭī} in the first \emph{pāda}, McGregor (1994: s.v.) and Callewaert (2009: s.v.) in their dictionaries of modern and old Hindi both give the meaning ``open all around" for \emph{caupaṭ}.
 McGregor derives it from Sanskrit \emph{catuṣpaṭṭa}.
-Molesworth (1857:s.v.) in his Marathi dictionary gives "A quadrangular expanse or space, esp.
-as open and extended: also a broad and level tract" for \emph{caupaṭṭā}
+Molesworth (1857: s.v.) in his Marathi dictionary gives ``A quadrangular expanse or space, esp.
+as open and extended: also a broad and level tract" for \emph{caupaṭṭā}.
 \end{philcomm}
@@
 \begin{translation}[hp04_015]
-Having forever abandoned all worry and all activity, as a result of focusing on nothing but the inner sound the mind dissolves into the inner sound.
+Having forever abandoned all worry and all activity, as a result of meditating on nothing but the inner sound the mind dissolves into the inner sound.
 \end{translation}
@@
 
-\emph{Haṭhasaṅketacandrikā} (ms. 2244 f.~124r)
+\emph{Haṭhasaṅketacandrikā} f.~124r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_015]
-The occurrence of two finite verbs in the second hemistich of some witnesses (groups 2 and 3) is infelicitous and perhaps points to an error that arose from the \textit{nādam evānusandhānān} reading in the group one??
-mss.
+%The occurrence of two finite verbs in the second hemistich of some witnesses (groups 2 and 3) is infelicitous and perhaps points to an error that arose from the \textit{nādam evānusandhānān} reading in the group one mss.
 It seems the original reading was probably \textit{nāda evānusandhānāt}, and the -m- crept in as a hiatus break.
+
+The third verse quarter varies considerably among the witnesses and testimonia.
+We have accepted the reading of \alphaOne\ (\emph{nādam evānusandhānān}), which is also attested by the \textbeta\ group and manuscripts of the \emph{Haṭharatnāvalī}, on the assumption that the -\emph{m}- at the end of \emph{nādam} is a hiatus break, the intended reading being \textit{nāda evānusandhānāt}.
+A hiatus break is also at 3.90.
 \end{philcomm}
@@
 \begin{translation}[hp04_016]
-Beginning (\emph{ārambhaḥ}), activation (\emph{ghaṭaḥ}), accumulation (\emph{paricayaḥ}) and completion (\emph{niṣpattiḥ}): they are the stages of yoga in all yogas.
+Beginning (\emph{ārambhaḥ}), union (\emph{ghaṭaḥ}), accumulation (\emph{paricayaḥ}) and completion (\emph{niṣpattiḥ}): those are the stages of yoga in all yogas.
 \end{translation}
@@
 \begin{versinnote}
-\tl{ārambhaś ca ghaṭaś caiva paricayaḥ prakīrtitaḥ |\\+}
-\tl{niṣpattiḥ sarvayogeṣu cetasas tasya bhūmikāḥ ||\\+}
-\tl{\var{paricaryaḥ prakīrtitaḥ Ga :
-tathā paricaya samṛtaḥ Ae (\textit{unmetr.}) :
-pracayaś ca tritīyakaḥ Ad Gb :
-pracayaś ca tṛtīryakaḥ  Ba :
-pracayaṃ ca tritīyakaṃ  Ta}\\+}
-\tl{\var{cetasas tasya bhūmikāḥ ] $\beta$ : yogāvasthā caturvidhā $\theta$}\\!}%?? variants necessary?
+\tl{ārambhaś ca ghaṭaś caiva paricayas tṛtīyakaḥ |\\+}
+\tl{niṣpattiḥ sarvayogeṣu yogāvasthā caturvidhā ||\\!}
 \end{versinnote}
@@
 \tl{ārambhaś ca ghaṭaś caiva tathā paricayas tathā |\\+}
+\tl{niṣpattiḥ sarvayogeṣu yogāvasthā bhavanti tāḥ ||\\+}
+\tl{\var{tathā ] tataḥ, tadā, smṛtāḥ, tv athā}\\!}
+\end{versinnote}
+\end{sources}
+
+\begin{testimonia}[hp04_016]
+\emph{Haṭharatnāvalī} 4.17
+\begin{versinnote}
+\tl{ārambhaś ca ghaṭaś caiva tathā paricayas tathā |\\+}
 \tl{niṣpattiḥ sarvayogeṣu yogāvasthā bhavanti tāḥ ||\\!}
 \end{versinnote}
-\end{sources}
-
-\begin{testimonia}[hp04_016]
-\emph{Yogacintāmaṇi}
+
+\emph{Yogacintāmaṇi} f. 111v
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.17
-\begin{versinnote}
-\tl{ārambhaś ca ghaṭaś caiva tathā paricayas tathā |\\+}
-\tl{niṣpattiḥ sarvayogeṣu yogāvasthā bhavanti tāḥ ||\\!}
-\end{versinnote}
-\end{testimonia}
-
-%\begin{philcomm}[hp04_016]
-%\end{philcomm}
+\emph{Yuktabhavadeva} 7.135
+\begin{versinnote}
+\tl{tad uktaṃ śivayoge—\\+}
+\tl{ārambhaś ca ghaṭaś caiva tathā paricayaḥ punaḥ |\\+}
+\tl{niṣpattiś ceti yogasya syād avasthācatuṣṭayam ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.14 (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ārambhaś ca ghaṭaś caiva tathā paricayas tathā|\\+}
+\tl{niṣpattiḥ sarvayogeṣu syād avasthācatuṣṭayam  ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+\begin{philcomm}[hp04_016]
+On these four stages, see Birch 2019: 968–969; Mallinson and Szanto 2021:19–20.
+\end{philcomm}
 
@@
 \begin{translation}[hp04_017]
-Now the beginning stage:\\
+Among these, the beginning stage is [as follows]:\\
 As a result of the piercing of the knot of Brahmā, bliss arises in the void [and] an unstruck sound which has various tones is heard in the body.
 \end{translation}
-%?? MD: better to adopt "tatra ārambhaḥ"? The first component is often introduced with "tatra" after the enumeration in the important mss.
 
@@
 \tl{brahmagranthes tathā bhedād ānandaḥ śūnyasambhavaḥ |\\+}
-\tl{vicitrakvaṇako dehe 'nāhataḥ śrūyate dhvaniḥ ||\\+}
-\tl{\var{°granthes tathā ] Ga : °granthes tato Ae : °granther bhaved Ba Gb Ta : °granthir bhaved Ad}\\+}
-\tl{\var{bhedād ] Ga : bhedāt Ae : vedhād Ba Gb Ta : vedhā Ad}\\!}%?? variants needed?
+\tl{vicitrakvaṇako dehe 'nāhataḥ śrūyate dhvaniḥ ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_017]
-\emph{Yogacintāmaṇi}
-\begin{versinnote}
-\tl{brahmagranthir bhaved bhinna ānandaḥ śūnyasaṃbhavaḥ |\\+}
-\tl{vicitrakṣaṇiko deho 'nāhataḥ śrūyate dhvaniḥ ||\\!}
-\end{versinnote}
-
 \emph{Haṭharatnāvalī} 4.18
@@
 
-\emph{Haṭhatattvakaumudī} 54.15
+\emph{Yogacintāmaṇi} f. 25r
+\begin{versinnote}
+\tl{brahmagranthir bhaved bhinna ānandaḥ śūnyasaṃbhavaḥ |\\+}
+\tl{vicitrakṣaṇiko deho 'nāhataḥ śrūyate dhvaniḥ ||\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.15 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_017]
-Brahmānanda understands the voids associated with each stage to be places in the body.
+In \emph{Jyotsnā} 4.70, Brahmānanda understands the voids associated with each stage to be places in the body.
+However, in the \emph{Amṛtasiddhi} and \emph{Amaraugha}, the source text of this verse, the series of voids (along with their respective blisses and sounds) derives from a tetrad of meditative voids in Vajrayāna traditions (Mallinson and Szanto 2021: 18, Birch 2019: 968).
 \end{philcomm}
@@
 \begin{translation}[hp04_018]
-In the beginning [stage], which is [the first phase of] voidness and in which the heart is full [of bliss], one becomes a yogi with a divine body, splendour, a divine smell and freedom from disease.%??
-space after śūnye in edition.
-??
-JM:  With a divine body, radiant, smelling heavenly, free from disease [and] his heart full [of bliss], in the void in the beginning [stage] the practitioner attains yoga.
+With a divine body, radiant, smelling heavenly, free from disease [and] his heart full [of bliss], in the void in the beginning [stage] the practitioner becomes a yogi.
+
 \end{translation}
@@
 \tl{divyadehaś ca tejasvī divyagandho hy arogavān |\\+}
-\tl{saṃpūrṇahṛdaye śūnye tv ārambhe yogavān bhavet ||\\!}
+\tl{saṃpūrṇahṛdaye śūnye tv ārambhe yogavān bhavet ||\\+}
+\tl{\var{°hṛdaye ] \emph{Amaraugha}, °hṛdayaḥ \emph{Amaraughaprabodha}}\\!}
 \end{versinnote}
@@
 
-\emph{Haṭhatattvakaumudī} 54.18
+\emph{Haṭhatattvakaumudī} 54.18 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{translation}[hp04_019]
-Now the activation stage:\\
+Now the unified stage:\\
 In the second stage the breath, after bringing about union, goes into the middle [channel]. Then the yogi has a firm posture [and] he becomes a gnostic and equal to a god.
@@
 \tl{dṛḍhāsano bhaved yogī jñānī devasamas tadā ||\\+}
-\tl{\var{dvitīye saṅ- ] $\beta$ : dvitīyāyāṃ $\theta$}\\!}
+\tl{\var{dvitīye saṅ- ] \emph{Amaraugha} : dvitīyāyāṃ \emph{Amaraughaprabodha}}\\!}
 \end{versinnote}
@@
 
-\emph{Haṭhatattvakaumudī} 54.15
+\emph{Haṭhatattvakaumudī} 54.15 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{translation}[hp04_020]
-Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a \textit{bherī} drum occurs.%??
-in ed join śabdas and tadā
+Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a \textit{bherī} drum occurs.%
 \end{translation}
@@
 \tl{atiśūnye vimardaś ca bherīśabdas tato bhavet ||\\+}
-\tl{\var{atiśūnye ] Ad Gb Ta : atiśūnyo $\beta$ Ba}\\+}
-\tl{\var{tato ] $\beta$ : tathā  Ba : tadā Ad Gb Ta}\\!}
+\tl{\var{atiśūnye ] Ad Gb Ta : atiśūnyo \emph{Amaraugha}, Ba}\\+}
+\tl{\var{tato ] \emph{Amaraugha} : tathā  Ba : tadā Ad Gb Ta}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_020]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.21
+\begin{versinnote}
+\tl{viṣṇugranthes tathā bhedaḥ paramānandasūcakaḥ |\\+}
+\tl{atiśūnye vimardaś ca bherīśabdas tathā bhavet ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f. 25r
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.21
-\begin{versinnote}
-\tl{viṣṇugranthes tathā bhedaḥ paramānandasūcakaḥ |\\+}
-\tl{atiśūnye vimardaś ca bherīśabdas tathā bhavet ||\\!}
-\end{versinnote}
-
-\emph{Haṭhatattvakaumudī} 54.21
+\emph{Haṭhatattvakaumudī} 54.21 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_020]
-V3??
-and HR have the reading of the \textit{Amaraugha} (\textit{atiśūnye vimardaś ca}), which was adapted from the Buddhist concept of a series of four void-like meditative states and moments.
+%V3 and HR have the reading of the \textit{Amaraugha} (\textit{atiśūnye vimardaś ca}), which was adapted from the Buddhist concept of a series of four void-like meditative states and moments.
 In the \emph{Amaraugha},  two of the moments were left out and two were reinterpreted (as in the case of \emph{vimarda}, which seems to have be retained with the more general sense of a pounding sound).
-%It is possible that Svātmārāma adopted this reading but he may have also changed it to \textit{atiśūnyavibheda} (as attested by groups 2 and 3)  in an attempt to make sense of the pāda (i.e., the variety, atiśūnya, arises).??JM: I percented this as we've gone with alpha1.
-In fact do we need a note here?
-How about a more general note on the stages at the beginning of this section, referencing Amṛtasiddhi, Amaraugha and DYŚ?
-See Birch 2019: 968–969; Mallinson and Szanto 2021:19–20.
+%It is possible that Svātmārāma adopted this reading but he may have also changed it to \textit{atiśūnyavibheda} (as attested by groups 2 and 3)  in an attempt to make sense of the pāda (i.e., the variety, atiśūnya, arises).
 \end{philcomm}
@@
 Now the accumulation stage:\\
-In the third [stage], having pierced [the knot of Viṣṇu], the sound of a \textit{mardala} drum arises in the sky.
-Then [the yogi] reaches the great void, the abode of all supernatural powers.%??
-vihāyas = sky, not void so I've changed tr.
+In the third [stage], having pierced [the knot of Viṣṇu], the sound of a \textit{mardala} drum arises in space. Then [the yogi] reaches the great void, the abode of all supernatural powers.%
 \end{translation}
@@
 \begin{testimonia}[hp04_021]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.22
+\begin{versinnote}
+\tl{tṛtīyāyāṃ tato nityaṃ āviṣkāro marddladhvaniḥ |\\+}
+\tl{mahāśūnyaṃ tato yāti sarvasiddhisamāśrayaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f. 25r
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.22
-\begin{versinnote}
-\tl{tṛtīyāyāṃ tato nityaṃ āviṣkāro marddladhvaniḥ |\\+}
-\tl{mahāśūnyaṃ tato yāti sarvasiddhisamāśrayaḥ ||\\!}
-\end{versinnote}
-
-\emph{Haṭhatattvakaumudī} 54.22
+\emph{Haṭhatattvakaumudī} 54.22 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 Here the object of piercing is unspecified, but it is likely to refer to the knot of Viṣṇu mentioned in the previous verse.
-%?? MD: Pāda c - tathā (Gr2) to be changed to tato (Gr3) or tadā (alpha*,V3,Jyo)? JM: tadā is good and tr. already translates that way
+
+We have understood \emph{vihāyo}° (`space') as referring to the state of total emptiness (\emph{atiśunya}) that was mentioned in the previous verse.
+
 \end{philcomm}
@@
 \begin{translation}[hp04_022]
-Now the perfection stage:\\
-Having overcome the [supreme] bliss of the mind, there arises innate bliss. [The yogi becomes] free of disease, suffering, old age, death, hunger and sleep.
+Having overcome the [supreme] bliss of the mind, there arises innate bliss. [The yogi] becomes free of disease, suffering, old age, death, hunger and sleep.
 \end{translation}
@@
 \tl{doṣaduḥkhajarāmṛtyukṣudhānidrāvivarjitaḥ ||\\+}
-\tl{\var{40a paramānanda° ] $\beta$; cittānandaṃ $\theta$ • °rocitvāt ] $\beta$; tato jitvā $\theta$}\\!}
+\tl{\var{paramānanda° ] \emph{Amaraugha}; cittānandaṃ \emph{Amaraughaprabodha}.}\\+}
+\tl{\var{°rocitvāt ] \emph{Amaraugha}; tato jitvā \emph{Amaraughaprabodha}}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_022]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.23
+\begin{versinnote}
+\tl{cidānandaṃ tato jitvā paramānandasaṃbhavaḥ |\\+}
+\tl{doṣaduḥkhajarāmṛtyukṣudhānidrāvivarjitaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f. 25r
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.23
-\begin{versinnote}
-\tl{cidānandaṃ tato jitvā paramānandasaṃbhavaḥ |\\+}
-\tl{doṣaduḥkhajarāmṛtyukṣudhānidrāvivarjitaḥ ||\\!}
-\end{versinnote}
-
-\emph{Haṭhatattvakaumudī} 54.22
+\emph{Haṭhatattvakaumudī} 54.22 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_022]
-22a’s \emph{cittānāndaṃ tato jitvā} is perhaps a patch for the somewhat obscure \emph{paramānandarocitvāt} in the source text.
+The reading of the first verse quarter, \emph{cittānāndaṃ tato jitvā}, is likely a patch for the somewhat obscure \emph{paramānandarocitvāt} in the \emph{Amaraugha} (40a), which is the source text.
 \end{philcomm}
@@
 Now the perfection stage:\\
-Then, having pierced Rudra's knot, the breath goes to all the  [of deities in the body]. In the perfected [stage] the sound of a flute becomes the sound of a resonating lute.
+Then, having pierced Rudra's knot, the breath goes to all the seats [of the deities in the body]. In the perfected [stage] the sound of a flute becomes the sound of a resonating lute.
 \end{translation}
@@
 \tl{niṣpanno vaiṇavaḥ śabdaḥ kvaṇadvīṇākvaṇo bhavet ||\\+}
-\tl{\var{sarva° ] Ae : sarvaṃ Ga : sattva° $\theta$}\\+}
+\tl{\var{sarva° ] Ae : sarvaṃ Ga : sattva° \emph{Amaraughaprabodha}}\\+}
 \tl{\var{niṣpanno ] Ga : niṣpannā Ae : niṣpattau Ad Gb T : ++ttau Ba }\\+}
@@
 \begin{testimonia}[hp04_023]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.24
+\begin{versinnote}
+\tl{rudragranthiṃ tato bhitvā śarvapīṭhagato 'nilaḥ |\\+}
+\tl{niṣpattau vaiṇavaḥ śabdaḥ kvanadvīṇākvaṇo bhavet ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f. 25r
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.24
-\begin{versinnote}
-\tl{rudragranthiṃ tato bhitvā śarvapīṭhagato 'nilaḥ |\\+}
-\tl{niṣpattau vaiṇavaḥ śabdaḥ kvanadvīṇākvaṇo bhavet ||\\!}
-\end{versinnote}
-
-\emph{Haṭhatattvakaumudī} 54.24
+\emph{Haṭhatattvakaumudī} 54.24 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_023]
-The reading we have adopted for the fourth \emph{pāda}, \emph{kvaṇadvīṇākvaṇo}, Birch’s conjecture in his edition of the \emph{Amaraugha} is supported by several testimonia and \emph{Amṛtasiddhi} 31.2, where it is said that the sound of a \emph{vīṇā} arises in the fourth stage.
+The reading we have adopted for the fourth verse quarter, \emph{kvaṇadvīṇākvaṇo}, Birch’s conjecture in his edition of the \emph{Amaraugha} is supported by several testimonia and \emph{Amṛtasiddhi} 31.2, where it is said that the sound of a \emph{vīṇā} arises in the fourth stage.
 
@@
 \begin{translation}[hp04_024]
-Then the mind having become one [with the sound??] is called Rājayoga. That yogi becomes a creator and destroyer, an equal to god.%??see note
+Then the mind having become one [with the sound]. This is called Rājayoga. He becomes a creator and destroyer, an equal to a lord among yogis.%
 \end{translation}
@@
 \tl{ekībhūtaṃ tadā cittaṃ rājayogābhidhānakam |\\+}
-\tl{sṛṣṭisaṃhārakartāsau yogīśvarasamo bhavet ||+}
-\tl{\var{rājayogābhidhānakam ] Ga Ba Ad Gb Ed ; rājayogo 'bhidhīyate Ae}\\!}
+\tl{sṛṣṭisaṃhārakartāsau yogīśvarasamo bhavet ||\\+}
+\tl{\var{rājayogābhidhānakam ] rājayogo 'bhidhīyate Ae}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_024]
-\emph{Yogacintāmaṇi}
-\begin{versinnote}
-\tl{ekībhūtaṃ tathā cittaṃ rājayogābhidhāyakam |\\+}
-\tl{sṛṣṭisaṃhārakartāsau yogīśvarasamo bhavet ||\\!}
-\end{versinnote}
+\emph{Yogacintāmaṇi} f. 25
+\begin{versinnote}
+\tl{ekībhūtaṃ tathā cittaṃ rājayogābhidhāyakam |\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} p.~106
+\begin{versinnote}
+\tl{vismṛtya sakalaṃ bāhyaṃ nāde dagdhāmbuvan manaḥ |\\+}
+\tl{ekībhūtaṃ tathā cittaṃ rājayogābhidhānakam \\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 \begin{philcomm}[hp04_024]
-The bahuvrīhi rājayogābhidhānakaṃ means that the mind is called rājayoga, which is very unusual. Perhaps °bhidhāyakaṃ is better, meaning "expresses rājayoga", but it is only in J7 and V3.
-
-%??SS: Then the mind becomes one. This is called Rājayoga. JM: looks best to me
-
-%??JM: yogīśvarasamo = equal to a yogi sovereign? Odd to take asau with yogī.
+%The bahuvrīhi rājayogābhidhānakaṃ means that the mind is called rājayoga, which is very unusual. Perhaps °bhidhāyakaṃ is better, meaning "expresses rājayoga", but it is only in J7 and V3.
+
+According to Brahmānanda (\emph{Jyotsnā} 4.77), the idea that the yogi becomes a creator and destroyer means that he becomes equal to God.
+Consequently he construes the line as \emph{asau yogī īśvarasamo bhavet}.
 \end{philcomm}
@@
 \begin{testimonia}[hp04_025]
-\emph{Yogacintāmaṇi} ??
-\begin{versinnote}
-\tl{astu vā māstu vā siddhir atraivākhaṇḍitaṃ sukham |\\!}
-\end{versinnote}
-
 \emph{Haṭharatnāvalī} 4.16
@@
 
-\emph{Haṭhatattvakaumudī} 54.35
+\emph{Yogacintāmaṇi} f. 113v (attr. to īśvara)
+\begin{versinnote}
+\tl{astu vā māstu vā siddhir atraivākhaṇḍitaṃ sukham |\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.35 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \tl{na sidhyati tato yugmam āniṣpatteḥ samabhyaset |\\+}
-\tl{tasmāt pravartate yogī haṭhe sadgurumārgataḥ ||\\!}
+\tl{tasmāt pravartate yogī haṭhe sadgurumārgataḥ ||\\+}
+\tl{\var{na ... samabhyaset ] \emph{om.} I, III, IV, VII, IX, X, XII, XIV–XVI}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_026]
-\emph{Yogacintāmaṇi}
-\begin{versinnote}
-\tl{haṭhaṃ vinā rājayogaṃ rājayogaṃ vinā haṭham |\\+}
-\tl{ye vai caranti tān manye prayāsaphalavarjitān iti ||\\!}
-\end{versinnote}
-
 \emph{Haṭharatnāvalī} 1.19
@@
 \tl{vyāptiḥ syād avinābhūtā śrīrājahaṭhayogayoḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f. 21r (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{haṭhaṃ vinā rājayogaṃ rājayogaṃ vinā haṭham |\\+}
+\tl{na siddhyati tato yugmaṃ manīṣyetau samabhyaset ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.127 (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{haṭhaṃ vinā rājayogo rājayogaṃ vinā haṭhaḥ |\\+}
+\tl{na sidhyati tato yugmamāniṣpatteḥ samācaret ||\\!}
 \end{versinnote}
@@
 \begin{translation}[hp04_027]
-I consider those who are ignorant of Rājayoga and engage in only Haṭha to be labourers who get no reward for their efforts.%??JM: farm labourers?
-ploughmen?
-It would be good to get the agricultural meaning in.
+I consider those who are ignorant of Rājayoga and only work at Haṭha to be like farmhands who get no reward for their efforts.%
 \end{translation}
@@
 
-%\begin{testimonia}[hp04_027]
-%\end{testimonia}
+\begin{testimonia}[hp04_027]
+\emph{Haṃsavilasa} p.~49
+\begin{versinnote}
+\tl{rājayogam ajānantaḥ kevalaṃ haṭhakarmagāḥ|\\+}
+\tl{ye tān karmavaśān manye prayāsaphalavarjitāḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
 
@@
 \begin{testimonia}[hp04_028]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f. 24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 It is a synonym for \emph{samādhi} and \emph{unmanī} in the next verses.
-Brahmānanda says that \emph{tattva} means \emph{citta}, which works well: the mind grows to the beyond-mind state.
-
-The reading \emph{smṛtam} is also possible and well-attested (groups 2 and 3).%??JM: not in apparatus.
+However, this meaning seems unlikely here as \emph{tattva} is one of three factors that is supposed to lead to \emph{unmanī}.
+In \emph{Jyotsnā} 4.104, Brahmānanda says that \emph{tattva} means \emph{citta}, which makes sense in so far as the mind grows to the beyond-mind state.
+However, this interpretation seems somewhat contrived.
+We have understood \emph{tattva} in the sense of the 'highest reality' (\emph{paramatattva}) on the assumption that, as the seed, it is the latent cause of the no-mind state.
 
@@
 \begin{translation}[hp04_029]
-Rājayoga, samādhi, unmanī, manonmanī, amaraugha, advaita, nirālamba, nirañjana...%?? format. Italic? Capitals?
-\end{translation}
+The sovereign yoga (\emph{rājayoga}), meditative absorption (\emph{samādhi}), beyond mind (\emph{unmanī}), transmental (\textit{manonmanī}), [the sovereign yoga of] the lineage of immortals (\emph{amaraugha}), non-dual (\emph{advaita}), without support (\emph{nirālamba}), pure (\emph{nirañjana}), [\dots]
+\end{translation}
+%
 
@@
 \begin{testimonia}[hp04_029]
-\emph{Yogacintāmaṇi} ??
+\emph{Yogacintāmaṇi} f.~6r
 \begin{versinnote}
@@
 
-\emph{Yogasārasaṅgraha} p.~60
+\emph{Yogasārasaṅgraha} p.~60 (attr. to the \emph{Nandikeśvaratārāvalī})
 \begin{versinnote}
@@
 
-\emph{Haṃsavilāsa} (p. 47)
+\emph{Haṃsavilāsa} p.~47
 \begin{versinnote}
@@
 \begin{translation}[hp04_030]
-amanaska, laya, tattva, śūnyāśūnya, para pada, jīvanmukti, sahaja and turya are synonyms.
-\end{translation}
+[\dots] the no-mind [state] (\emph{amanaska}), dissolution [of mind] (\emph{laya}), the [supreme] reality (\emph{tattva}), void and not void (\textit{śūnyāśūnya}), the highest state (\emph{para pada}), liberation in life (\emph{jīvanmukti}), innate (\emph{sahaja}) and the fourth [state] (\emph{turya}) are synonyms.
+\end{translation}
+%
 
 \begin{sources}[hp04_030]
-\emph{Yogacintāmaṇi} ??
+\emph{Yogacintāmaṇi} f.~6r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Upāsanāsārasaṅgraha} p.~106
+\emph{Upāsanāsārasaṅgraha} p.~106 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{translation}[hp04_031]
-Two paths for the quick attainment of Unmanī are approved by me: tattva, the supreme bliss, and honouring the inner sound...
+Two paths for the quick attainment of \emph{unmanī} are approved by me: \emph{tattva}, the supreme bliss, and cultivating the inner sound...
 \end{translation}
@@
 \begin{testimonia}[hp04_031]
-\emph{Yogacintāmaṇi} ??
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Upāsanāsārasaṅgraha} p.~106
+\emph{Upāsanāsārasaṅgraha} p.~106 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
 \tl{unmanyavāptaye śīghraṃ mārgau dvau mama sammatau |\\+}
-\tl{tatvaṃ paramasākhyaṃ vā nādopāsanam eva vā  ||\\!}
+\tl{tattvaṃ paramasākhyaṃ vā nādopāsanam eva vā  ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_032]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{sources}[hp04_032_1]
-\emph{Tantrāloka} 32.64
+Cf. \emph{Tantrāloka} 32.64
 \begin{versinnote}
@@
 \emph{Tantrālokaviveka} 32.63
-\begin{versinnote}
+Cf \begin{versinnote}
 \tl{yad āgamaḥ –\\+}
@@
 \end{versinnote}
+
 \emph{Śivasūtravimarśinī} 5
@@
 \tl{ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī |\\+}
-\tl{eko devo nirālambaḥ ekāvasthā manonmanī||\\!}
-\end{versinnote}
-
-\emph{Nādabindūpaniṣad} 52cd-53ab
-\begin{versinnote}
-\tl{śaṅkhadundubhinādaṃ ca na śruṇoti kadācana || 52 ||\\+}
-\tl{kāṣṭhavajjñāyate deha unmanyāvasthayā dhruvam |\\!}
-\end{versinnote}
+\tl{eko devo nirālambaḥ ekāvasthā manonmanī ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~75r (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī |\\+}
+\tl{eko deśo nirālamba ekāvasthā manonmanī ||\\!}
+\end{versinnote}
+
+\emph{Yuktabhavadeva} 7.219 (attr. to Gorakṣanātha)
+\begin{versinnote}
+\tl{ekaṃ sṛṣṭimayaṃ bījam ekā mudrā ca khecarī |\\+}
+\tl{eko devo nirālamba ekāvasthā manonmanī ||\\!}
+\end{versinnote}
+
+%\emph{Nādabindūpaniṣad} 52cd-53ab
+%\begin{versinnote}
+%\tl{śaṅkhadundubhinādaṃ ca na śruṇoti kadācana || 52 ||\\+}
+%\tl{kāṣṭhavaj jñāyate deha unmanyāvasthayā dhruvam |\\!}
+%\end{versinnote}
 \end{testimonia}
@@
 \end{translation}
+%
 
 \begin{sources}[hp04_032_2]
-Jñānasāra 3.7
+\emph{Jñānasāra} 3.7
 \begin{versinnote}
@@
 
-%\begin{testimonia}[hp04_032_2]
-%\end{testimonia}
+\begin{testimonia}[hp04_032_2]
+\emph{Haṭhasaṅketacandrika} f.~120v (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{haṭhapradīpikāyāṃ \\+}
+\tl{śaṃkhaduṃdubhinādaṃ ca n[a] śṛṇoti kadācana |\\+}
+\tl{kāṣṭavaj jāyate dehe unmanyā'vasthayāś dhruvaṃ ||\\!}
+\end{versinnote}
+
+\end{testimonia}
 
@@
 \begin{translation}[hp04_032_3]
-Free from all states of mind and all thought, the yogi is as if dead. He is liberated. In this there is no doubt.
+Free from all states [of mind] and all thought, the yogi is as if dead. He is liberated. In this there is no doubt.
 \end{translation}
@@
 
-\emph{Haṭhatattvakaumudī} 51.75
+\emph{Haṭhatattvakaumudī} 51.75 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-%\begin{philcomm}[hp04_032_5]
-%\end{philcomm}
+\begin{philcomm}[hp04_032_5]
+The collated manuscripts have \emph{avadhyaḥ} in the first and second verse quarters.
+Although \emph{avadhya} can be construed in both quarters, the repetition appears to be a dittographical error that changed \emph{avedyaḥ}, which is close to the reading of the source text, into \emph{avadhyaḥ}.
+We have therefore emended accordingly.
+\end{philcomm}
 
@@
 
-%\begin{testimonia}[hp04_032_6]
-%\end{testimonia}
+\begin{testimonia}[hp04_032_6]
+\emph{Yuktabhavadeva} 11.31 (attr. to Gorakṣanātha)
+\begin{versinnote}
+\tl{na gandhaṃ na rasaṃ sparśaṃ na rūpaṃ na ca niḥsvanam |\\+}
+\tl{nātmānaṃ ca paraṃ vetti yogī yuktaḥ samādhinā ||\\!}
+\end{versinnote}
+\end{testimonia}
 
@@
 \begin{translation}[hp04_032_7]
-He is indeed truly liberated whose mind is neither asleep nor awake, neither mindful nor otherwise, and neither stops nor starts.
+He is indeed truly liberated whose mind is neither asleep nor awake, has no memory nor otherwise, and neither stops nor starts.
 \end{translation}
@@
 \begin{versinnote}
-\tl{cittaṃ na suptaṃ no jāgrat smṛtiman na ca nānyathā |\\+}
-\tl{nāstam eti na codeti yasyāsau mukta eva sa ||\\!}
-\end{versinnote}
-\end{sources}
-
-%\begin{testimonia}[hp04_032_7]
-%\end{testimonia}
+\tl{cittaṃ prasuptaṃ yogena jāgratsuptaṃ na cānyathā |\\+}
+\tl{nāstam eti na codeti yasyāsau mukta eva hi ||\\+}
+\tl{\var{7a cittaṃ prasuptaṃ yogena ] T; cittaṃ na suptaṃ no jāgrac G.}\\+}
+\tl{\var{7b jāgratsuptaṃ na cānyathā ] em.; jāgratsūtir na *nyathā T, chrutimadvacanasya ca G.}\\!}
+\end{versinnote}
+%
+\end{sources}
+
+\begin{testimonia}[hp04_032_7]
+\emph{Haṭhasaṅketacandrika} f.~120v (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{cittaṃ na suptaṃ no jāgrat smṛtivarṇaṃ na cānyathā |\\+}
+\tl{nāstam eti na codeti yasyāsau mukta eva saḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
 
 \begin{philcomm}[hp04_032_7]
-The first line is corrupt in the \emph{Gorakṣaśataka} witnesses; its text has been constituted from the \emph{Haṭhapradīpikā}.
+%The first line is corrupt in the \emph{Gorakṣaśataka} witnesses; its text has been constituted from the \emph{Haṭhapradīpikā}. 
+The first line of this verse is different to the version in the source text, the \emph{Gorakṣaśataka}, as the words \emph{jāgrat} and \emph{suptaṃ} are found in the first quarter.
+The second quarter is corrupt in many of the \emph{Haṭhapradīpikā} manuscripts but \emph{smṛti} and \emph{nānyathā} are well attested.
+We have therefore adopted the reading of G11 (group 4c), which in this case is close to the readings of the \textBeta\ group.
+%
 \end{philcomm}
@@
 \tl{svapnajāgradavasthāyāṃ suptavat yo 'avatiṣṭhate | \\+}
-\tl{niśvāsocchvāsahīnaśca niścitaṃ mukta eva saḥ || 11 ||\\!}
-\end{versinnote}
-
-\emph{Yuktabhavadeva} 1.64
-\begin{versinnote}
-\tl{svapnajāgradavasthāyāṃ suptavat yo 'vatiṣṭhate |\\+}
-\tl{niśvāsocchvāsahīnaś ca niścitaṃ mukta eva saḥ ||64||\\!}
-\end{versinnote}
-
-\emph{Haṭhatattvakaumudī} 55.24
+\tl{niśvāsocchvāsahīnaś ca niścitaṃ mukta eva saḥ || 11 ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~27v (attr. to the \emph{Rājayoga} [aka. \emph{Amanaska}])
+\begin{versinnote}
+\tl{sadā jāgradavasthāyāṃ suptavada yo'vatiṣṭhate ||\\+}
+\tl{niḥśvāsocchāsahīnaś ca niścitaṃ mukta eva saḥ|\\!}
+\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 55.24 (attr. to the \emph{Rājayoga} [aka. \emph{Amanaska}])
 \begin{versinnote}
@@
 
-\emph{Yogacintāmaṇi}
-\begin{versinnote}
-\tl{sadā jāgradavasthāyāṃ suptavada yo'vatiṣṭhate ||\\+}
-\tl{niḥśvāsocchāsahīnaśca niścitaṃ mukta eva saḥ|\\!}
-\end{versinnote}
+%\emph{Yuktabhavadeva} 1.64 (attr. to the \emph{Kulārṇava})
+%\begin{versinnote}
+%\tl{svapnajāgradavasthāyāṃ suptavat yo 'vatiṣṭhate |\\+}
+%\tl{niśvāsocchvāsahīnaś ca niścitaṃ mukta eva saḥ ||64||\\!}
+%\end{versinnote}
 \end{testimonia}
@@
 \begin{translation}[hp04_033]
-Only the glorious guru lord knows the unique ineffable bliss that has arisen in the hearts of yogi lords who experience \textit{samādhi} by concentrating on the inner sound.
-\end{translation}
+Only the glorious guru lord knows the unique ineffable bliss that has arisen in the hearts of lords among yogis who experience \textit{samādhi} by concentrating on the inner sound.
+\end{translation}
+% JB: Perhaps a note is needed on yogīśvara (here or 4.24). Brahmānanda takes yogīśvarāḥ as yogiṣu yogayukteṣu īśvarāḥ samarthāḥ. i.e., 'lords among yogis' or perhaps 'adepts among yogis'?
+%% Jayaratha (TĀ 2.43) yogīśvaraḥ = yoginām apīśvaraḥ 'a lord of even yogis'
 
 \begin{sources}[hp04_033]
-Yogatārāvalī 3
+\emph{Yogatārāvalī} 3
 \begin{versinnote}
@@
 \begin{testimonia}[hp04_033]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.5
+\begin{versinnote}
+\tl{nādānusandhānasamādhibhājāṃ yogīśvarāṇāṃ hṛdaye prarūḍham |\\+}
+\tl{ānandam ekaṃ vacaso 'py agamyaṃ jānāti taṃ śrīgurunātha eva ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.5
-\begin{versinnote}
-\tl{nādānusandhānasamādhibhājāṃ yogīśvarāṇāṃ hṛdaye prarūḍham |\\+}
-\tl{ānandam ekaṃ vacaso 'py agamyaṃ jānāti taṃ śrīgurunātha eva ||\\!}
-\end{versinnote}
 \end{testimonia}
@@
 \begin{translation}[hp04_033_1]
-Seated in the pose of the liberated ones, the yogi should adopt \emph{śāṃbhavī mudrā} and listen continuously to the inner sound in his right ear.
-\end{translation}
-%
+Seated in the pose of the liberated, the yogi should adopt \emph{śāṃbhavī mudrā} and listen continuously to the inner sound in his right ear.
+\end{translation}
+%%JB: I've changed our translation of muktāsana to the pose of the liberated (instead of 'liberated ones') in keeping with the way we translated it in ch. 1
 
@@
 \begin{testimonia}[hp04_033_1]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \tl{śṛṇuyād dakṣiṇe karṇe nādam ekāṃtike sudhīḥ  [||]\\+}
-\tl{\var{dakṣiṇe karṇe ] B220,  dakṣirṇe 2244}\\+}
-\tl{\var{ekāṃtike ] 2244, ekāṃtate B220}\\!}
+\tl{\var{dakṣiṇe karṇe ] B220,  dakṣirṇe 2244}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_034]
-\emph{Yogacintāmaṇi}
-\begin{versinnote}
-\tl{sarvacintāṃ parityajya sāvadhānena cetasā |\\+}
-\tl{nādam evānusandhatte yogasāmrājyam icchatā ||\\!}
-\end{versinnote}
-
 \emph{Haṭharatnāvalī} 4.14
@@
 
-Haṭhasaṅketacandikā (f. 124r)
+\emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{sarvacintāṃ parityajya sāvadhānena cetasā |\\+}
+\tl{nādam evānusandhatte yogasāmrājyam icchatā ||\\+}
+\tl{\var{°sāmrājyam icchatā ] U, °sāmrājyadhiṣṭhitaḥ N}\\!}
+\end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f. 124r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{testimonia}[hp04_035]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.8
+\begin{versinnote}
+\tl{karṇau pidhāya tūlena yaḥ śṛṇoti dhvaniṃ yamī |\\+}
+\tl{tatra cittaṃ sthiraṃ kuryād yāvat sthirapadaṃ vrajet ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.8
-\begin{versinnote}
-\tl{karṇau pidhāya tūlena yaḥ śṛṇoti dhvaniṃ yamī |\\+}
-\tl{tatra cittaṃ sthiraṃ kuryād yāvat sthirapadaṃ vrajet ||\\!}
-\end{versinnote}
-
-\emph{Haṭhasaṅketacandrikā} (f. 124r)
+\emph{Haṭhasaṅketacandrikā} f.~124r
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_035]
-%adopt sthirīkuryād [MD: Done.]%??JM: remove space after sthirī  in ed.
+%
 %mūlena (alpha) probably came from tūlena.
-The hasta variations seem impractible because the yogi would have to hold the arms up, but may have been adopted by yogis who did not have access to cotton (?).??JM: I think these comments are unnecessary (and surely it is possible/practicable to use the fingers?).
-
-Several witnesses have \emph{hastābhyām} or \emph{hastena} here.
-Blocking the ears with the fingers in order to listen to the inner sounds is attested as early as the \textit{Svacchandatantra}, in a practice called \textit{ṣaṇmukhīkaraṇa}, in which the other openings of the head are also blocked with the fingers (Vasudeva 2004: 272 n.66).
+The hasta variations seem impractible because the yogi would have to hold the arms up, but may have been adopted by yogis who did not have access to cotton (?).JM: I think these comments are unnecessary (and surely it is possible/practicable to use the fingers?).
+
+%JB: I've tweaked the note below to inform the reader why we've adopted tūlena.
+The reading \emph{tūlena} (`with cotton'), which is attested by \alphaThree, makes good sense and is close to the reading of \emph{mūlena} in \alphaOne\ and \alphaTwo.
+Manuscripts of several other groups, including \textgamma, instead have \emph{hastābhyām} (`with the hands') or \emph{hastena} (`with the hands').
+This reading was inspired by the technique of blocking the ears and other orifices with the fingers in order to listen to the inner sounds.
+This practice is attested as early as the \textit{Svacchandatantra} in which it is called \textit{ṣaṇmukhīkaraṇa} (Vasudeva 2004: 272 n.66).
+In this \emph{karaṇa}, the other openings of the head are also blocked with the fingers.
 \emph{Śivasaṃhitā} 5.36–46 teaches a similar practice.
-%aṅguṣṭābhyām ubhe śrotre tarjanībhyāṃ vilocane
-%nāsārandhre ca madhyābhyām anāmābhyāṃ mukhaṃ dṛḍham 5.36
-%nirudhya mārutaṃ yogī yadaivaṃ kurute bhṛśam
-%tadā tatkṣaṇamātmānaṃ jyotīrūpaṃ sa paśyati 5.37
-%tattejo dṛśyate yena kṣaṇamātraṃ nirākulam
-%sarvapāpavinirmuktaḥ sa yāti paramāṃ gatim 5.38
-%nirantarakṛtābhyāsād yogī vigatakalmaṣaḥ
-%sarvaṃ dehādi vismṛtya tadbhinnas tu svayaṃ bhavet 5.39
-%yaḥ karoti sadābhyāsaṃ guptācāreṇa mānavaḥ
-%sa vai brahmaṇi līnaḥ syāt pāpakarmarato yadi 5.40
-%gopanīyaḥ prayatnena sadyaḥpratyayakārakaḥ
-%nirvāṇadāyako loke yogo 'yaṃ mama vallabhaḥ 5.41
-%nādaḥ saṃjāyate tasya krameṇābhyāsataś ca vai
-%mattabhṛṅgāvalīvīṇāsadṛśaḥ prathamo dhvaniḥ 5.42
-%evam abhyāsataḥ paścāt saṃsāradhvāntanāśanaḥ
-%ghaṇṭāravasamaḥ paścād dhvanir megharavopamaḥ 5.43
-%dhvanau tasmin mano dattvā yadā tiṣṭhati nirbharam
-%tadā saṃjāyate siddhir layasya mama vallabhā 5.44
-%tatra nāde yadā cittaṃ ramate yogino bhṛśam
-%vismṛtya sakalaṃ bāhyaṃ nādena saha śāmyati 5.45
-%etadabhyāsayogena jitvā samyag guṇān bahūn
-%sarvārambhaparityāgī cidākāśe vilīyate 5.46
-In the \textit{Haṭhayogasaṃhitā} (p. 68), the practice of blocking the ears with the hands is stipulated for \textit{bhrāmarī kumbhaka}.
+In the \textit{Haṭhayogasaṃhitā} (p.
+68), the practice of blocking the ears with the hands is stipulated for \textit{bhrāmarī kumbhaka}.
 \end{philcomm}
@@
 \begin{testimonia}[hp04_036]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭhasaṅketacandrikā} (f. 124r)
-\begin{versinnote}
-\tl{abhyasyamāno nādo yaṃ bāhyam āvarttayet dhvaniṃ |\\+}
+\emph{Haṭhasaṅketacandrikā} f.~124r
+\begin{versinnote}
+\tl{abhyasyamāno nādo 'yaṃ bāhyam āvarttayet dhvaniṃ |\\+}
 \tl{pakṣād vikṣepam akhilaṃ jitvā yogī sukhībhavet ||\\!}
@@
 \begin{testimonia}[hp04_037]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.9
+\begin{versinnote}
+\tl{śrūyate prathamābhyāse nādo nānāvidho bahuḥ |\\+}
+\tl{vardhamāne tato 'bhyāse śrūyate sūkṣmasūkṣmataḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \tl{vartamāne tato 'bhyāse śrūyate sūkṣmasūkṣmataḥ ||\\!}
-\end{versinnote}
-
-\emph{Haṭharatnāvalī} 4.9
-\begin{versinnote}
-\tl{śrūyate prathamābhyāse nādo nānāvidho bahuḥ |\\+}
-\tl{vardhamāne tato 'bhyāse śrūyate sūkṣmasūkṣmataḥ ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_038]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.10
+\begin{versinnote}
+\tl{ādau jaladhijīmūtabherīnirjharasaṃbhavāḥ |\\+}
+\tl{madhye marddalaśaṃkhotthā ghaṇṭākāhalakās tathā ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.10
-\begin{versinnote}
-\tl{ādau jaladhijīmūtabherīnirjharasaṃbhavāḥ |\\+}
-\tl{madhye marddalaśaṃkhotthā ghaṇṭākāhalakās tathā ||\\!}
-\end{versinnote}
-
 \emph{Haṭhatattvakaumudī} 54.32
@@
 \begin{philcomm}[hp04_038]
-??
-more refs there are different options for kāhala, such as a trumpet or drum that looks like datura flower.
-See Schmidt, Nachträge: "eine Art Trommel (kāhalāsu dhattūrapuṣpākāramukhabherīṣu), Yaśastilakam (commentary).
+We have translated \emph{kāhala} as `trumpet' on the basis of e.g.
+\emph{Viśvalocanakośa}, \emph{lāntavarga} 161 (\emph{dhvaninālā tu vīṇāyāṃ veṇukāhalayor api}), but it can also mean a type of drum (see e.g.
+\emph{Śabdakalpadruma} s.v.
+\emph{kāhala} where it is said to be a \emph{bṛhaḍḍhakkā}, a big drum).%whose mouth looks like a datura flower.
+NWS: "a musical instrument; a trumpet.
+EI 24 (Index) (Sircar 1966: 138)".
+%Schmidt, Nachträge: "eine Art Trommel (\emph{kāhalāsu dhattūrapuṣpākāramukhabherīṣu}), \emph{Yaśastilakam}  (nach dem Ko.) = asphuṭaśabda" [\ldots]. 
 \end{philcomm}
@@
 \begin{translation}[hp04_039]
-In the final stage, there are the sounds of little bells, a bamboo flute, a veena and a bee.
-
-[These] various sound are heard in the body.
+In the final stage, there are the sounds of little bells, a bamboo flute, a veena and a bee. [These] various sound are heard in the body.
 \end{translation}
@@
 \begin{testimonia}[hp04_039]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.11 
+\begin{versinnote}
+\tl{ante tu kiṃkiṇīvṛndavīṇābhramaraniḥsvanāḥ |\\+}
+\tl{iti nānāvidhā nādāḥ śrūyante dehamadhyataḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \tl{iti nānāvidho nādaḥ śrūyate dehamadhyagaḥ ||\\!}
-\end{versinnote}
-
-\emph{Haṭharatnāvalī} 4.11
-\begin{versinnote}
-\tl{ante tu kiṃkiṇīvṛndavīṇābhramaraniḥsvanāḥ |\\+}
-\tl{iti nānāvidhā nādāḥ śrūyante dehamadhyataḥ ||\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_040]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{translation}[hp04_041]
-Or, the mind should filter out the gross sound for the subtle, or the subtle for the gross, or, abandoning both, be [focused] on [a sound in] the middle, [and the] should not move it elsewhere.%??JM: here gross and subtle do seem to make more sense
-\end{translation}
+Or the [yogi] should filter out the gross sound for the subtle, or the subtle for the gross, or, abandoning both, be [focused] on [a sound in] the middle [and] not move the mind elsewhere.%
+\end{translation} %
 
@@
 \begin{testimonia}[hp04_041]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{testimonia}[hp04_042]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{testimonia}[hp04_043]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.12
+\begin{versinnote}
+\tl{makarandaṃ pibed bhṛṅgo gandho na prekṣyate yathā |\\+}
+\tl{nādāsaktaṃ tathā cittaṃ viṣayān na hi kāṃkṣate ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.12
-\begin{versinnote}
-\tl{makarandaṃ pibed bhṛṅgo gandho na prekṣyate yathā |\\+}
-\tl{nādāsaktaṃ tathā cittaṃ viṣayān na hi kāṃkṣate ||\\!}
-\end{versinnote}
-
 \emph{Haṭhatattvakaumudī} 54.41
@@
 \begin{translation}[hp04_044]
-When the mercury of the mind is bound and has cast off its fickle nature because it has been assimilated with the sulphur of the internal resonance, it attains the immobility called the unsupported [\textit{samādhi}].
+When the mercury of the mind is bound and has cast off its fickle nature because it has been assimilated with the sulphur of the internal resonance, it attains the immobility called the unsupported (i.e., \textit{samādhi}).
 \end{translation}
@@
 \begin{testimonia}[hp04_044]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \tl{baddhaṃ viyuktaṃ cāpalyaṃ nādagandhakajāraṇāt |\\+}
-\tl{manaḥ pāradam āpnoti nirālambākhyakheṭakam ||\\!}
+\tl{manaḥpāradam āpnoti nirālambākhyakheṭakam ||\\!}
 \end{versinnote}
@@
 \begin{translation}[hp04_044_1]
-Bound by the sulphur of the inner sound, the lord that is the mercury of the mind immediately casts off its fickle nature and attains glory, known as “[the bird] with clipped wings”.
-\end{translation}
+Bound by the sulphur of the inner sound, the lord that is the mercury of the mind immediately casts off its fickle nature and attains fame as “[the bird] with clipped wings”.
+\end{translation} %
 
@@
 \begin{testimonia}[hp04_044_1]
+\emph{Haṭhatattvakaumudī} 54.43
+\begin{versinnote}
+\tl{baddhaḥ sugandhanādena sadyaḥ santyaktacāpalaḥ |\\+}
+\tl{prayāti sūtacittendraḥ pakṣacchinna ivāprabhaḥ ||\\!}
+\end{versinnote}
 \end{testimonia}
@@
 \begin{philcomm}[hp04_044_1]
-
-Cf.~\emph{Rasendracūḍāmaṇi} 16.52--55
-\begin{versinnote}
-\tl{evaṃ ca pañcamo grāsaḥ pradātavyo'ṣṭamāṃśataḥ |\\+}
-\tl{sa pātrastho 'gnisaṃtapto na gacchati kathañcana ||16.52||\\+}
-\tl{sa \textit{pakṣacchinna} ity uktaḥ sa mukto  'khiladurguṇaiḥ |\\+}
-\tl{so'yaṃ niṣevitaḥ sūtas trimāsaṃ rājikāmitaḥ ||16.53||\\+}
-\tl{viḍaṅgatriphalākṣaudraiḥ khe devaiḥ saha saṅgamam |\\+}
-\tl{ghrāṇamātreṇa \textit{sūtendraḥ} sarvaroganikṛntanaḥ ||16.54||\\+}
-\tl{guṇā ete vinirdiṣṭā rasasya rasavādibhiḥ |\\+}
-\tl{sakalāste guṇāḥ satyā bhairaveṇa prakīrtitāḥ ||16.55||\\!}
-\end{versinnote}
-
-Cf.~also NWS (??ref) s.v.~\textit{pakṣaccheda}.
+On \emph{pakṣaccheda} in alchemical processes of immobilizing mercury and for references in Rasaśāstra, see Hellwig 2009: 276–278.
+
+% Hellwig, O. (2009). Wörterbuch der mittelalterlichen indischen Alchemie. Barkhuis ; University of Groningen.
+
+%Cf.~\emph{Rasendracūḍāmaṇi} 16.52--55
+%\begin{versinnote}
+%\tl{evaṃ ca pañcamo grāsaḥ pradātavyo 'ṣṭamāṃśataḥ |\\+} 
+%\tl{sa pātrastho 'gnisaṃtapto na gacchati kathañ cana ||16.52||\\+}
+%\tl{sa \textit{pakṣacchinna} ity uktaḥ sa mukto 'khiladurguṇaiḥ |\\+}
+%\tl{so 'yaṃ niṣevitaḥ sūtas trimāsaṃ rājikāmitaḥ ||16.53||\\+}
+%\tl{viḍaṅgatriphalākṣaudraiḥ khe devaiḥ saha saṅgamam |\\+}
+%\tl{ghrāṇamātreṇa \textit{sūtendraḥ} sarvaroganikṛntanaḥ ||16.54||\\+}
+%\tl{guṇā ete vinirdiṣṭā rasasya rasavādibhiḥ |\\+}
+%\tl{sakalāste guṇāḥ satyā bhairaveṇa prakīrtitāḥ ||16.55||\\!}
+%\end{versinnote}
+
+%Cf.~also NWS (ref) s.v.~\textit{pakṣaccheda}.
 
@@
 \begin{translation}[hp04_045]
-As a result of listening to the inner sound, the  snake that is the mind forgets everything and, one-pointed, does not slip away anywhere.
+As a result of listening to the inner sound, the snake that is the mind forgets everything and, one-pointed, does not dart off anywhere.
 \end{translation}
@@
 \begin{testimonia}[hp04_045]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 The metaphor is that of snake charming: beguiled by the inner sound, the snake that is the mind becomes transfixed.
-Witnesses of the \gamma \ and \delta \ groups have \emph{turaṅgamaḥ} instead of \emph{bhujaṅgamaḥ}, perhaps because forms from \emph{dhāv} are unusual with the latter, but they are found at e.g.~\emph{Padmapurāṇa}, \emph{Garuḍapurāṇa} (refs ??).
+Witnesses of the \gamma \ and \delta \ groups have \emph{turaṅgamaḥ} instead of \emph{bhujaṅgamaḥ}, perhaps because forms from \emph{dhāv} are unusual with the latter, but it is found at e.g.~\emph{Garuḍapurāṇa} (1.113.33ab).
+% Padmapurāṇa
 \end{philcomm}
@@
 This inner sound is a sharpened goad with the power to restrain the bull elephant in must that is the mind as it wanders about in the garden of the sense objects.
-\end{translation}
+\end{translation} %
 
@@
 \begin{testimonia}[hp04_046]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~23r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{translation}[hp04_047]
-The inner sound is a bolt for the swift horse of the mind so the yogi should regularly concentrate on cultivating it.
-\end{translation}
-% 4
+The inner sound is a bolt for the swift horse of the mind so the yogi should regularly focus on its cultivation.
+\end{translation}
 
@@
 \begin{translation}[hp04_047_1]
-The inner sound is a net for trapping the deer of the mind and a hunter for trapping the antelope of the mind.
+The inner sound is a net for trapping the deer of the mind and a hunter for corraling the antelope of the mind.
 \end{translation}
@@
 \begin{testimonia}[hp04_047_1]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_047_1]
-Adopt V15 67cd
-
-??
-MD: rodhe may be correct.
-In a drive hunt, the animals are driven in a certain direction (often in a confined space)  by means of loud noises, etc., so that another hunter can easily shoot them.
-JM: yes, adopt, I have altered tr.
-to "trapping"
+%Adopt V15 67cd
+%
 \end{philcomm}
@@
 \begin{translation}[hp04_049]
-When the mind dissolves into that which is the most subtle object of perception in the unstruck sound, that is the supreme state of Viṣṇu.
-
-The tone of that sound is that of the unstruck sound.
-The object of perception is inside the sound [and] the mind is inside the object of perception.
-When the mind dissolves [in it], that is the supreme state of Viṣṇu.
-\end{translation}
-
-% anāhatadhvaner antar jñeyaṃ yat sūkṣmasūkṣmakaṃ/
-% manas tatra layaṃ yāti tad viṣṇoḥ paramaṃ padam//
+The tone of that sound is that of the unstruck sound. A light is inside the tone [and] the mind is inside the light. That mind dissolves. That is the supreme state of Viṣṇu.
+\end{translation}
 
 \begin{sources}[hp04_049]
-Uttaragītā 41cd–42 (source??)
+\emph{Uttaragītā} 41cd–42 %(source) JB: I think we can say that the UG is the source of 4.49*1 (but not 4.49)
 \begin{versinnote}
@@
 \begin{testimonia}[hp04_049]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-Haṭhayogasaṃhitā p.~68
+\emph{Haṭhayogasaṃhitā} p.~68
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_049]
-%??
-to discuss.
+% to discuss.
 It appears that the following lines were added (in Gr2 and Gr3) as a note or supplement to 4.68.
@@
 seems to comment on anāhatadhvaner.
-\begin{versinnote}
-\tl{anāhatasya śabdasya tasya śabdasya yo dhvaniḥ ||\\+}
-\tl{dhvaner antargataṃ jyotir jyotirantargataṃ manaḥ |\\+}
-\tl{tanmano vilayaṃ yāti tad viṣṇoḥ paramaṃ padam  || \\!}
-\end{versinnote}
-The source of them may be the Uttaragītā. However, in NGMPP E 2098-11 Uttaragītā ms these three lines are omitted from Kṛṣṇa's words, which start with oṃkāra.
+%\begin{versinnote}
+%\tl{anāhatasya śabdasya tasya śabdasya yo dhvaniḥ ||\\+}
+%\tl{dhvaner antargataṃ jyotir jyotirantargataṃ manaḥ |\\+}
+%\tl{tanmano vilayaṃ yāti tad viṣṇoḥ paramaṃ padam  || \\!}
+%\end{versinnote}
+The source of these lines may be the \emph{Uttaragītā} as they occur in a published version.
+However, in a manuscript of the \emph{Uttaragītā} (NGMPP E 2098-11) these three lines are omitted from Kṛṣṇa's words, which start with \emph{oṃkāra}.
+The author of the \emph{Upāsanāsārasangraha} (f.
+111) quoted these lines and attributed them to the \emph{Gītāsāra}.
 %HI: tan could mean tatra.
@@
 
-%?? MD: I adopted the version transmitted in Gr4c and J10 (anāhatadhvaner antar jñeyaṃ yat sūkṣmasūkṣmakaṃ ...).
-%?? Which is the more likely reading of the second line for the alternative version?
+% MD: I adopted the version transmitted in Gr4c and J10 (anāhatadhvaner antar jñeyaṃ yat sūkṣmasūkṣmakaṃ ...).
+% Which is the more likely reading of the second line for the alternative version?
 %A - dhvaner antargataṃ jyotir jyotirantargataṃ manaḥ (J5,G4,Gr2,Gr3)
-%B - dhvaner antargataṃ jñeyaṃ jñeyasyāntargataṃ manaḥ (N3?,Jyo) JM: the second I think
-\end{philcomm}
+%B - dhvaner antargataṃ jñeyaṃ jñeyasyāntargataṃ manaḥ (N3?,Jyo) JM: the second I think the second. J
+
+% & Note about the note for JB
+% switch 4.49 with 4.49*1
+% Explain why we have adopted Alt2 (probably borrowed from the Uttaragītā) and the likelihood that Alt 1 was an attempt to replace it with something more coherent.
+% We are not sure what the difference between dhvani and śabda is... and have translated them as ...
+\end{philcomm}
+
+% N3 (α1)
+% anāhatasya śabdasya tasya śabdasya ca dhvaniḥ |
+% dhvaner aṃtargataṃ jñeyaṃ yasyāṃ tv aṃtargataṃ manaḥ |
+% tan mano vilayaṃ yāṃti tadviṣṇo paramaṃ padaṃ |
+
+%%%%%%%%%%
+\subsection*{4.49*1}
+\begin{translation}[hp04_049]
+When the mind dissolves into that which is the most subtle object of perception in the unstruck sound, that is the supreme state of Viṣṇu.
+\end{translation}
+% anāhatadhvaner antar jñeyaṃ yat sūkṣmasūkṣmakaṃ/
+% manas tatra layaṃ yāti tad viṣṇoḥ paramaṃ padam//
+
+%\begin{sources}[hp04_049]
+%\end{sources}
+
+\begin{testimonia}[hp04_049*1]
+\emph{Hathatattvakaumudī} 54.48
+\begin{versinnote}
+\tl{anāhatadhvaner antar jñeyaṃ yat sūkṣmasūkṣmakaṃ ||\\+}
+\tl{manas tatra layaṃ yāti tad viṣṇoḥ paramaṃ padam ||\\!}
+\end{versinnote}
+
+\end{testimonia}
+
+%\begin{philcomm}[hp04_049*1]
+%HI: tan could mean tatra.
+%Looks like the last line was rewritten to include tatra
+% MD: I adopted the version transmitted in Gr4c and J10 (anāhatadhvaner antar jñeyaṃ yat sūkṣmasūkṣmakaṃ ...).
+% JB I have divided 4.49 and 4.49*1, and changed the sources and testimonia accordingly.
+But it seems that alpha has 4.49*1 and not 4.49, so have we made a mistake in greyscaling 4.49*1?
+A note is definately needed, but I'm not sure what our thinking is here
+%\end{philcomm}
 
@@
 
-%\begin{sources}[hp04_050]
-%\end{sources}
+\begin{sources}[hp04_050]
+\emph{Vivekamārtaṇḍa} (six-chapters) 5.15 %JB: This was in testimonia but it must be the source as we now know that Svātmārāma used the 6ch VM.
+\begin{versinnote}
+\tl{tāvad ākāśasaṅkalpo yāvac chabdaḥ pravartate |\\+}
+\tl{niḥśabdaṃ tat parabrahma paramātmā sa gīyate || 15 ||\\!}
+\end{versinnote}
+\end{sources}
 
 \begin{testimonia}[hp04_050]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~27r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-Six-chapter \emph{Vivekamārtaṇḍa} 5.15
-\begin{versinnote}
-\tl{tāvad ākāśasaṅkalpo yāvac chabdaḥ pravartate |\\+}
-\tl{niḥśabdaṃ tat parabrahma paramātmā sa gīyate || 15 ||\\!}
-\end{versinnote}
-
 \emph{Nādabindūpaniṣat} 47cd–48ab
@@
 %\end{philcomm}
-
 %%%%%%%%%%
@@
 \begin{testimonia}[hp04_051]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~27r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
-\emph{Haṭhasaṅketacandrikā} f.~235r–235v
+\emph{Haṭhasaṅketacandrikā} f.~123r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{translation}[hp04_052]
-Blocking of the ears, mouth, eyes, and nose should not be performed [when] a clear inner sound is heard distinctly in the purified Suṣumṇā channel.%?? negative in first line? Adopt naiva?
+Blocking of the ears, mouth, eyes, and nose should not be performed [when] a clear inner sound is heard distinctly in the purified Suṣumṇā channel.
 \end{translation}
@@
 \begin{testimonia}[hp04_052]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~26v
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_052]
-%??N3 is missing an akṣara (nirodhaṃ aiva)
-%??J5 has nirodhaṃ naiva
-??Adopt alpha reading mukha°, naiva. HI: it is strongly attested (groups 4c, C6, etc) and more likely explains the change to °saṃrodhanaṃ kāryam
+%N3 is missing an akṣara (nirodhaṃ aiva)
+%J5 has nirodhaṃ naiva
+% Adopt alpha reading mukha°, naiva. HI: it is strongly attested (groups 4c, C6, etc) and more likely explains the change to °saṃrodhanaṃ kāryam. 
+% & Note: blocking of ears, etc., has not been mentioned until now. 
 \end{philcomm}
@@
 \begin{translation}[hp04_052_1]
-Nāda is called Śakti. And gnosis of nāda is Sadāśiva. When gnosis of nāda has disappeared, only unmanī remains.
-\end{translation}
+The inner sound is called Śakti. And gnosis of the inner sound is Sadāśiva. When gnosis of the inner sound has disappeared, only the beyond-mind [state] remains.
+\end{translation}
+%
 
@@
 \begin{testimonia}[hp04_052_1]
+\emph{Upāsanāsārasaṅgraha} f. 107 (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{nādaga...r iti jñeyaṃ nādo jñānaṃ sadāśivaḥ |\\+}
+\tl{jñeyajñāne vilīne [']ṃtaḥ sonmany evāviśiṣyate ||\\!}
+\end{versinnote}
+
 \emph{Haṭhatattvakaumudī} 54.50
@@
 %\begin{philcomm}[hp04_052_1]
-%??V15 reading for the third pāda is better than N19.
+%V15 reading for the third pāda is better than N19.
+JB: If we adopt the epsilon 3 (V15) for the third pāda, then we have to adopt tad unmany eva śiṣyate in the fourth (otherwise 4th pāda is unmetrical with evāvaśiṣyate).
 %\end{philcomm}
@@
 \begin{translation}[hp04_052_3]
-When the accumulated subliminal impressions have been destroyed as a result of the continuous fusing of the mind with the internal resonance, the mind and breath are sure to dissolve into the unadorned [god] (\emph{nirañjane}).
-\end{translation}
+When the store of subliminal impressions has been destroyed as a result of continuously concentrating on the inner sound, the mind and breath are sure to dissolve into the untainted [god] (\emph{nirañjane}).
+\end{translation}%. 
+%
 
@@
 \begin{testimonia}[hp04_052_3]
-\emph{Upāsanāsārasaṅgraha} ??
-\begin{versinnote}
-\tl{sadā nādānusandhānāt saṃkṣīṇe vāsanā bhavet |\\+}
-\tl{nirañjane vilīneti niścalitaṃ mārutaṃ manaḥ ||\\!}
+%\emph{Upāsanāsārasaṅgraha} f. 107 Jb: problematic as most of the first line is missing in ms. no. 12170 and we havent been using the IFP transcript (which does have this verse)
+%\begin{versinnote}
+%\tl{sadā nādānusandhānāt saṃkṣīṇe vāsanā bhavet |\\+}
+%\tl{nirañjane vilīneti niścalitaṃ mārutaṃ manaḥ ||\\!}
+%\end{versinnote}
+
+\emph{Haṭhatattvakaumudī} 54.52
+\begin{versinnote}
+\tl{sadā nādānusandhānāt saṃkṣīṇe vāsanākṣaye |\\+}
+\tl{nirañjane ca līyate niścitaṃ cittamārutau ||\\!}
 \end{versinnote}
@@
 \end{versinnote}
-
-\emph{Haṭhatattvakaumudī} 54.52
-\begin{versinnote}
-\tl{sadā nādānusandhānāt saṃkṣīṇe vāsanākṣaye |\\+}
-\tl{nirañjane ca līyate niścitaṃ cittamārutau ||\\!}
-\end{versinnote}
-\end{testimonia}
-
-%\begin{philcomm}[hp04_052_3]
-%\end{philcomm}
+\end{testimonia}
+
+\begin{philcomm}[hp04_052_3]
+The term \emph{nirañjanaḥ} likely refers to \emph{devo nirañjanaḥ} (`untainted god') in the next verse (4.52*4), an expression which refers to the supreme deity.
+\end{philcomm}
 
@@
 \begin{translation}[hp04_052_4]
-Thousands of crores of inner sounds and hundreds of crores of visual focal points all dissolve into the place where the unadorned god (\emph{devo nirañjana}) is.
-
-%There are thousands of Crores of resonances and thousands of crores of sonic matter (bindu). All dissolve into where the spotless god exists.
-%?? tr. the spotless god, Nirañjana. ? and in previous
+Thousands of crores of inner sounds and hundreds of crores of visual focal points all dissolve into the place of the untainted god (\emph{devo nirañjanaḥ}).
+%
 \end{translation}
@@
 \begin{testimonia}[hp04_052_4]
-\emph{Yogacintāmaṇi}
-\begin{versinnote}
-\tl{nādakoṭisahasrāṇi bindukoṭiśatāni ca |\\+}
-\tl{sarve tatra layaṃ yānti yatra devo nirañjana iti ||\\!}
-\end{versinnote}
-
-\emph{Haṭhatattvakaumudī} 54.53
+\emph{Yogacintāmaṇi} f.~27r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 
+\emph{Haṭhatattvakaumudī} 54.53
+\begin{versinnote}
+\tl{nādakoṭisahasrāṇi bindukoṭiśatāni ca |\\+}
+\tl{sarve tatra layaṃ yānti yatra devo nirañjanaḥ ||\\!}
+\end{versinnote}
+
 \emph{Nādabindūpaniṣat} 50cd--51ab
@@
 \end{versinnote}
-\end{testimonia}
-
-\begin{philcomm}[hp04_052_4]
-The expression \textit{devo nirañjanaḥ} (`unadorned god') refers to Śiva.
-
-Compare the quotation in the Śabdakalpadruma, s.v.~\emph{dharmaghaṭa}:
+
+Cf. \emph{Śabdakalpadruma} (s.v.~\emph{dharmaghaṭa})
 \begin{quote}
-“\dots ante yāti paraṃ sthānaṃ yatra devo nirañjanaḥ||"
+\dots ante yāti paraṃ sthānaṃ yatra devo nirañjanaḥ |
 iti bhaviṣyapurāṇoktā dharmaghaṭavratakathā samāptā ||
 \end{quote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_052_4]
+On the meaning of \emph{devo nirañjana}, see the note to 4.52.3.
+
+%Compare the quotation in the \emph{Śabdakalpadruma}, s.v.~\emph{dharmaghaṭa}:
+%\begin{quote}
+%“\dots ante yāti paraṃ sthānaṃ yatra devo nirañjanaḥ||"
+%iti bhaviṣyapurāṇoktā dharmaghaṭavratakathā samāptā ||
+%\end{quote}
+%JB. I'm not sure I see the point of this comparison. Delete? The hemistich seems generic and not necessarily related to the dharmaghaṭavrata. 
 \end{philcomm}
@@
 Enough of the verbosity of a learned gathering! O friend, listen to this instruction formerly taught by Śiva for awakening Matsyendra.
-
 \end{translation}
@@
 \begin{testimonia}[hp04_053]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_053]
-This verse is likely to be authorial, composed to introduce next two verses, which are from the \emph{Candrāvalokana}, a dialogue between Matsyendra and Śiva.
-1.34 is another probable authorial verse and, like this one, it has the vocative \emph{sakhe}.
+This verse may have been composed by Svātmārāma to introduce the next two verses, which are from the \emph{Candrāvalokana}, a dialogue between Matsyendra and Śiva.
+Verse 1.34, which may also be authorial like this one, has the vocative \emph{sakhe}.
+% JB: In support of this, we could add that other yoga texts which are not dialogues, like the Yogatārāvalī (26), Śivayogapradīpikā (2.59, 3.41, etc.) and Yogabhāskara (2), appear to use the vocative sakhe to address the reader.
 \end{philcomm}
@@
 \begin{translation}[hp04_054]
-As long as the moving breath does not enter the middle path, as long as bindu is not stable, restrained by the breath, as long as the ultimate truth (\emph{tattvam}) does not become as natural as the sky, then all that one says is deceitful, and false prattle.
-\end{translation}
+As long as the moving breath does not enter the middle path, as long as bindu is not stable, restrained by the breath, as long as [realisation of] the ultimate truth (\emph{tattvam}), which is as natural as the sky, does not arise, then all that one says is deceitful, and false prattle.
+\end{translation}%?? tr of 3rd pāda. JM thinks it's "doesn't become as natural as...". Current formulation doesn't work for me (if it's sahaja, it has arisen already)
 
@@
 \tl{tāvat sarvaṃ yadi ca vadate dambhamithyāpralāpaḥ ||\\+}
-\tl{\var{14b prāṇavātaprabandhaḥ ] 4345 : prāṇaghātaprabuddhaḥ 75278 : prāṇavāyuḥ prabuddhaḥ 7970 : prāṇavātaprabuddhaḥ T00788
-14d sarvam] 7970, T00788, 75278 : vighnaṃ 4345}\\!}
+\tl{\var{14b prāṇavātaprabandhaḥ~] 4345, prāṇaghātaprabuddhaḥ 75278, prāṇavāyuḥ prabuddhaḥ 7970, prāṇavātaprabuddhaḥ T00788}\\!}
+%14d sarvam] 7970, T00788, 75278 : vighnaṃ 4345
 \end{versinnote}
@@
 \begin{testimonia}[hp04_054]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~22a
 \begin{versinnote}
@@
 \tl{yāvad vyomnaḥ sahajasadṛśaṃ jñāyate naiva tatvam \\+}
-\tl{tāvat sarvaṃ vadati yad idaṃ dambhamithyāpralāpa iti\\!}
+\tl{tāvat sarvaṃ vadati yad idaṃ dambhamithyāpralāpaḥ\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} f.~110–111
+\begin{versinnote}
+\tl{haṭhapradīpikāyāṃ—\\+}
+\tl{yāvan naiva praviśati caran māruto madhyamārge\\+}
+\tl{yāvad bindur na bhavati dṛḍhaḥ prāṇavātaprabaddhe |\\+}
+\tl{yāvad vyomnā sahajasadṛśaṃ jāyate nātmatattvaṃ \\+}
+\tl{tāvat sarvaṃ vadati yad idaṃ dambhamithyāpralāpaḥ\\!}
 \end{versinnote}
@@
 \begin{translation}[hp04_055]
-Having learnt the correct piercing of Suṣumṇā, [the yogi] should make the breath go into the central channel, take it to the place of the moon and stop [the flow] in the nostrils.
+Having learnt the correct piercing of Suṣumṇā, [the yogi] should make the breath go into the central channel, take it to the place of the moon and block the nostrils.
 \end{translation}
 %nītvāsāv aindavasthāne
+%
 
@@
 \begin{testimonia}[hp04_055]
-\emph{Upāsanāsārasaṅgraha} p.~31 (attributed to the \emph{Candrāvalokana})
+\emph{Upāsanāsārasaṅgraha} p.~31 (attr. to the \emph{Candrāvalokana})
 \begin{versinnote}
@@
 
+\emph{Haṭhasaṅketacandrikā} f.~107v–108r (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{jñātvā suṣumnāsadbhedaṃ kṛtvā vāyuṃ ca madhyagam |\\+}
+\tl{sthitvā sadaiva svasthena prāṇarandhraṃ nirodhayet ||\\!}
+\end{versinnote}
+
 \emph{Yogakuṇḍalinyupaniṣat} 7cd–8ab
@@
 \end{versinnote}
-
-\emph{Haṭhasaṅketacandrikā} f.~107v–108r
-\begin{versinnote}
-\tl{jñātvā suṣumnāsadbhedaṃ kṛtvā vāyuṃ ca madhyagam |\\+}
-\tl{sthitvā sadaiva svasthena prāṇarandhraṃ nirodhayet ||\\!}
-\end{versinnote}
 \end{testimonia}
@@
 %\begin{philcomm}[hp04_055]
-%??ghrāṇarandhra is too well supported to ignore, but the meaning is difficult to construe. JM: it's in alpha etc., why not adopt it? vāyum is object in pādas bcd.
 %\end{philcomm}
@@
 \end{metre}
-
 %%%%%%%%%%
@@
 \begin{translation}[hp04_056]
-The moon and sun move in iḍā and piṅgalā. The moon is said to be tamasic and the sun rajasic.
+And so, Vasiṣṭha [said]:\\
+The moon and sun move in Iḍā and Piṅgalā. The moon is said to be of the nature of \emph{tamas} and the sun of \emph{rajas}.
 \end{translation}
@@
 
-\emph{Yogayājñavalkya} 4.32cd–33
+Cf.~\emph{Yogayājñavalkya} 4.32cd–33
 \begin{versinnote}
@@
 
-\emph{Matsyendrasaṃhitā} 4.41cd
+Cf.~\emph{Matsyendrasaṃhitā} 4.41cd
 \begin{versinnote}
@@
 \begin{testimonia}[hp04_056]
-\emph{Yogacintāmaṇi}
+\emph{Haṭharatnāvalī} 4.36cd–37ab
+\begin{versinnote}
+\tl{iḍāyāṃ piṃgalāyāṃ ca somasūryau pratiṣṭhitau ||\\+}
+\tl{tāmaso rājasaś caiva savyadakṣinasaṃsthitau |\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~59v (attr. to Yājñavalkya)
 \begin{versinnote}
@@
 %\begin{philcomm}[hp04_056]
-%??JM: rajasic and tamasic not in Eng dictionary, "associated with the \emph{rajas guṇa}" ?
 %\end{philcomm}
@@
 \begin{sources}[hp04_057]
-Vasiṣṭhasaṃhitā 2.29cd–30ab
+\emph{Vasiṣṭhasaṃhitā} 2.29cd–30ab
 \begin{versinnote}
@@
 
-Yogayājñavalkya 4.34cd–35ab
+Cf. \emph{Yogayājñavalkya} 4.34cd–35ab
 \begin{versinnote}
@@
 \begin{testimonia}[hp04_057]
-\emph{Yogacintāmaṇi}
-\begin{versinnote}
-\tl{tāv eva dhattas sakalaṃ kālaṃ rātriṃ divātmakam |\\+}
+\emph{Yogacintāmaṇi} (attr. to Yājñavalkya)
+\begin{versinnote}
+\tl{tāv eva dhattaḥ sakalaṃ kālaṃ rātriṃ divātmakam |\\+}
 \tl{bhoktrī suṣumṇā kālasya guhyam etad udāhṛtam ||\\!}
@@
 
-\emph{Haṭhasaṅketacandrikā} (check in ms. no. 2244)
+\emph{Haṭhasaṅketacandrikā} f. 95v
 \begin{versinnote}
 \tl{tathā coktaṃ haṭhapradīkāyāṃ–\\+}
-\tl{sūryācandramasau dhattaḥ lagnaḥ kāla rātridinātmakaṃ ||\\+}
+\tl{sūryācandramasau dhattaḥ kālaṃ rātridinātmakam ||\\+}
 \tl{bhoktrī suṣumnā kālasya guhyate tad udīritaṃ ||\\!}
@@
 
-\begin{philcomm}[hp04_057]
-\end{philcomm}
+%\begin{philcomm}[hp04_057]
+%\end{philcomm}
 
@@
 \begin{translation}[hp04_058]
-For as the tetrad of verses called the Saubhadra has it:
-
-There are six cakras, sixteen supports, three focal points and three guṇas. Everything else is [just] the prolixity of texts. Trikūṭa is the supreme place.
+For as the tetrad of verses called the Saubhadra has it:\\
+There are six cakras, sixteen supports, three focal points and three \emph{guṇa}s. Everything else is [just] the prolixity of texts. Trikūṭa is the supreme place.
 \end{translation}
@@
 \begin{philcomm}[hp04_058]
-We do not know why this tetrad of verses is called Saubhadra.
+We do not know why this tetrad of verses is called Saubhadra.\\
+
+The three components of the yogic body listed here are found together in other texts, the earliest being \emph{Netratantra} 7.1ab (\emph{ṛtucakraṃ svarādhāraṃ trilakṣyaṃ vyomapañcakam}).
+However, we are yet to find a source for this list that includes the three \emph{guṇa}s.
 \end{philcomm}
@@
 
-%\begin{testimonia}[hp04_059]
-%\end{testimonia}
+\begin{testimonia}[hp04_059]
+\emph{Yogacintāmaṇi} f.~79r (attr. to the \emph{Haṭhayoga})
+\begin{versinnote}
+\tl{kuṇḍalī kuṭilākārā sarpavat parikīrtitā |\\+}
+\tl{sā śaktiścālitā yena sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
+
+\emph{Upāsanāsārasaṅgraha} f. 51 (attr. to a \emph{yogaśāstra})
+\begin{versinnote}
+\tl{kuṇḍalī kuṭilākārā sarpavat parikīrtitā |\\+}
+\tl{sā śaktiścālitā yena sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
 
@@
 \begin{translation}[hp04_060]
-When kūṭa is situated at trikūṭa [then] the mind is wonderful and uninterrupted. By means of kuṇḍalinī, [the yogi] is undoubtedly liberated.
+When the \emph{kūṭa} is situated at Trikūṭa [then] the mind is wonderful and uninterrupted. By means of Kuṇḍalinī, [the yogi] is undoubtedly liberated.
 \end{translation}
@@
 \begin{testimonia}[hp04_060]
-?? Upāsanāsārasaṅgraha !!!
+\emph{Upāsanāsārasaṅgraha} f. 51 (attr. to a \emph{yogaśāstra})
+\begin{versinnote}
+\tl{yadā kūṭaṃ trikūṭasthaṃ cittaṃ cittaṃ niraṃtaram |\\+}
+\tl{kuṇḍalyās tu prayogeṇa sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
 \end{testimonia}
@@
 \begin{philcomm}[hp04_060]
-We are unsure of the referent of \emph{kūṭa} here.
-As suggested to us by Sven Sellmer, it may mean the tip of the tongue, which in e.g.~\emph{Khecarīvidyā} 1.65–67 and 3.16–17, is to be placed at \emph{trikūṭa} as part of the practice of \emph{khecarīmudrā}.
+We are unsure of the meaning of \emph{kūṭa} here.
+As suggested to us by Sven Sellmer, it may mean the tip of the tongue which in, for example \emph{Khecarīvidyā} 1.65–67 and 3.16–17, is to be placed at \emph{trikūṭa} as part of the practice of \emph{khecarīmudrā}.
 \end{philcomm}
@@
 There are seventy-two thousand openings  of the channels in the cage [that is the body]. Suṣumṇā is the Śāmbhavī Śakti while the other [channels] are pointless.
-\end{translation}
+\end{translation}%
 
@@
 \begin{testimonia}[hp04_061]
-\emph{Haṭhasaṅketacandrikā} f.~108r
-\begin{versinnote}
-\tl{sūryā[c]andram asau kṛtvā viditvā karapaṃjare [|] \\+}
+\emph{Haṭhasaṅketacandrikā} f.~108r (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{sūryā[c]andramasau kṛtvā viditvā karapaṃjare | \\+}
 \tl{suṣumṇā śāṃbhavī śaktiḥ śeṣās tv eva nirarthakāḥ ||\\!}
@@
 
-Yogaśikhopaniṣat 6.17cd–18ab
+\emph{Upāsanāsārasaṅgraha} f.~111 (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{dvāsaptatisahasrāṇi nāḍīdvārāṇi paṃjare |\\+}
+\tl{suṣumnā śāṃbhavī śaktiḥ śeṣās tv eva nirarthakāḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogaśikhopaniṣat} 6.17cd–18ab
 \begin{versinnote}
@@
 Brahmānanda understands \emph{dvārāṇi} to refer to routes by which breath enters the body (\emph{dvārāṇi vāyupraveśamārgāḥ}) and we have translated \emph{nāḍīdvārāṇi} accordingly.
-%Consider °nāḍīnāṃ dehapañjare ??JM: probably a patch?
+%Consider °nāḍīnāṃ dehapañjare JM: probably a patch? JB: Agree
 \end{philcomm}
@@
 \begin{translation}[hp04_062]
-The breath, having been carefully accumulated, together with fire awakens kuṇḍalinī and enters Suṣumṇā without obstruction.
+The breath, having been carefully accumulated, together with fire awakens Kuṇḍalinī and enters Suṣumṇā without obstruction.
+
 \end{translation}
@@
 \tl{vāyuḥ paricito yatnād agninā saha kuṇḍalīm |\\+}
-\tl{bodhayitvā suṣumnāyāṃ praviśed avirodhataḥ ||\\!}
+\tl{bodhayitvā suṣumnāyāṃ praviśed anirodhataḥ ||\\!}
 \end{versinnote}
@@
 
-\emph{Haṭhasaṅketacandrikā} ff.~197v–180r (attributed to the \emph{Haṭhapradīpikā})
+\emph{Haṭhasaṅketacandrikā} ff.~197v–180r (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{translation}[hp04_063]
-When the breath is flowing in Suṣumṇā, the no-mind state is sure to be attained.
-Otherwise [i.e.~if the breath is not flowing in Suṣumṇā] the various practices [of yoga] lead to nothing but exertion for yogis.
-\end{translation}
+When the breath is flowing in Suṣumṇā, the no-mind state is attained. Otherwise [i.e.~if the breath is not flowing in Suṣumṇā], the various practices [of yoga] lead to nothing but exertion for yogis.
+\end{translation}%
 
@@
 \begin{testimonia}[hp04_063]
-\emph{Upāsanāsārasaṅgraha} p.~108
+\emph{Upāsanāsārasaṅgraha} p.~108 (attr. to the \emph{Haṭhapradīpikā})
 \begin{versinnote}
@@
 \begin{translation}[hp04_065]
-The mind has two impulses: past impression (\emph{vāsanā}) and the breath. When one of those two disappears, both soon disappear.%??JM: tr. vāsanā as plural? Any better ideas?
+The mind has two impulses: past impressions (\emph{vāsanā}) and the breath. When one of those two disappears, both soon disappear.%
 \end{translation}
@@
 
-\emph{Mokṣopāya} V.92.48
+Cf. \emph{Mokṣopāya} V.92.48
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_065]
+The emendation of \emph{drutaṃ} in the last verse quarter has been made to restore the faulty readings of the \textalpha\ manuscripts (i.e., \emph{dguttaṃ, dhṛtaṃ, dṛtaṃ}), which are similar to an incorrect reading in an important witness of the source text, the \emph{Gorakṣaśataka} (T, \emph{dhṛtaṃ}).
+The emendation to \emph{drutaṃ} is based on the parallel verse in the \emph{Mokṣopāya} and its related recensions, which have \emph{kṣipraṃ} instead.
+
 It is difficult to say whether the \emph{Gorakṣaśataka}'s reading \emph{cittasya} was changed to \emph{manaso} by the author when this verse was borrowed in order to make the terminology consistent with the previous verse, or whether this change occurred at a later time.
@@
 \begin{philcomm}[hp04_066]
-%JM?? do we need this note? I think it's about something other than samādhi.
-These verses seem to elevate the possible role of the mind in achieving \emph{samādhi} (as opposed to the standard Haṭha method of stopping the breath to achieve samādhi).
-
-%Brahmānanda understands this verse as referring to specific locations into which the breath and mind dissolve, rather tha"when... then...".
 %HI: tatra = tāvat.
@@
 \begin{testimonia}[hp04_067]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~19r (attr. to the \emph{Rājayoga} (aka. \emph{Amanaska})
 \begin{versinnote}
@@
 \begin{philcomm}[hp04_067]
-Complementing his understanding of the previous verse, Brahmānanda has \emph{yato\dots tatra} in \emph{pāda}s c and d, taking it to mean \emph{yatra\dots tatra} and to be referring to cakras.
+%These verses (4.66–67) seem to elevate the possible role of the mind in achieving \emph{samādhi} (as opposed to the standard Haṭha method of stopping the breath to achieve samādhi). 
+Complementing his understanding of the previous verse, Brahmānanda has \emph{yato\dots tatra} in 4.67cd, taking it to mean \emph{yatra\dots tatra} and to be referring to cakras.
 \end{philcomm}
@@
 \begin{testimonia}[hp04_068]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi} f.~19r (attr. to the \emph{Rājayoga} (aka. \emph{Amanaska})
 \begin{versinnote}
@@
 
-%\begin{testimonia}[hp04_069]
-%\end{testimonia}
+\begin{testimonia}[hp04_069]
+\emph{Haṭhasaṅketacandrikā} f.~117r
+\begin{versinnote}
+\tl{tathā coktaṃ haṭhapradīpikāyāṃ–\\+}
+\tl{vāyumārge [']py asaṃcāre sakalāṃ bhramate mahīm | \\+}
+\tl{tathā'ṣṭāguṇam aiśvaryam ity āha bhagavān śiva iti || \\!}
+\end{versinnote}
+\end{testimonia}
 
@@
 \begin{translation}[hp04_070]
-Thus Viśvarūpācārya:\\
-When the breath is destroyed and the mind dissolves, all experience is the same. That is called Samādhi.
+Thus, Viśvarūpācārya [said]:\\
+When the breath is destroyed and the mind dissolves, all experience is the same (\emph{samarasatva}). That is called \emph{samādhi}.
 \end{translation}
@@
 \begin{testimonia}[hp04_070]
-\emph{Yuktabhavadeva} 30
+\emph{Yuktabhavadeva} 11.30 (attr. to Gorakṣanātha)
 \begin{versinnote}
@@
 \end{versinnote}
+
+\emph{Haṭhasaṅketacandrikā} f.~117v
+\begin{versinnote}
+\tl{tathā ca viśvarūpāyāryāḥ–\\+}
+\tl{yadā saṃkṣīyate prāṇo mānasaṃ ca pralīyate | \\+}
+\tl{tadā samarasatvaṃ yat samādhiḥ so 'bhidhīyata iti || \\+}
+\tl{\var{mānasaṃ ca ] B220, mānaseva 2244}\\!}
+\end{versinnote}
 \end{testimonia}
@@
 \begin{translation}[hp04_071]
-When the mind is still the breath is still, from which bindu becomes still.  As a result of bindu becoming still, my son, the body becomes still.
+When the mind is still the breath is still, from which semen becomes still.  As a result of semen becoming still, my son, the body becomes still.
 \end{translation}
@@
 \begin{testimonia}[hp04_071]
-\emph{Yogacintāmaṇi}
+\emph{Yogacintāmaṇi}~f.19v
 \begin{versinnote}
@@
 
-\emph{Haṭhatattvakaumudī} 43.19
+Cf. \emph{Haṭhatattvakaumudī} 43.19
 \begin{versinnote}
@@
 %\begin{philcomm}[hp04_071]
+The vocative \emph{putra} in \emph{pāda} c suggests that this verse is from a source text that we are yet to identify.
 %88c āpanna not well attested and N3 has bindusthairyodayāt putra: perhaps this comes from an unknown source text addressed to a son.
-%?? putra
 %\end{philcomm}
@@
 \tl{cittaṃ sthiraṃ yasya vināvalambāt\\+}
-\tl{sa eva yogī sa guruḥ sa sevyaḥ ||\\+}
-\tl{\var{vinaiva dṛśyād  AllNI AllSI Eb: vinā sudṛśyād Ua: vinā nimeṣād Pb AllN : vinaiva lakṣyād Cc: vinaiva dṛṣṭyād Jb: vinakadṛśyad Ma: vinānaṅgame Nb}\\!}
+\tl{sa eva yogī sa guruḥ sa sevyaḥ ||\\!}
+%\tl{\var{vinaiva dṛśyād  AllNI AllSI Eb: vinā sudṛśyād Ua: vinā nimeṣād Pb AllN : vinaiva lakṣyād Cc: vinaiva dṛṣṭyād Jb: vinakadṛśyad Ma: vinānaṅgame Nb}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_072]
-\emph{Yogacintāmaṇi} (attributed to the \emph{Rājayoga})
+\emph{Haṭharatnāvalī} 4.25
+\begin{versinnote}
+\tl{dṛṣṭiḥ sthirā yasya vinaiva lakṣyāt\\+}
+\tl{vāyuḥ sthiro yasya vinā prayatnāt |\\+}
+\tl{cittaṃ sthiraṃ yasya vināvalambanāt\\+}
+\tl{sa eva yogī sa guruḥ sa sevyaḥ ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f. 48r (attr. to the \emph{Rājayoga})
 \begin{versinnote}
@@
 
-\emph{Kulārṇavatantra} 13.70
+Cf. \emph{Kulārṇavatantra} 13.70
 \begin{versinnote}
@@
 
-\emph{Haṭharatnāvalī} 4.25
-\begin{versinnote}
-\tl{dṛṣṭiḥ sthirā yasya vinaiva lakṣyāt\\+}
+\emph{Haṭhasaṅketacandrikā}~f. 3v (attr. to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{atha gurulakṣaṇam |\\+}
+\tl{dṛṣṭi[ḥ] sthirā yasya vinaiva dṛśyād\\+}
 \tl{vāyuḥ sthiro yasya vinā prayatnāt |\\+}
-\tl{cittaṃ sthiraṃ yasya vināvalambanāt\\+}
-\tl{sa eva yogī sa guruḥ sa sevyaḥ ||\\!}
-\end{versinnote}
-
-\emph{Haṭhatattvakaumudī} 16
-\begin{versinnote}
-\tl{dṛṣṭiḥ sthirā yasya vinaiva dṛśyād\\+}
-\tl{vāyuḥ sthiro yasya vinā prayatnāt |\\+}
-\tl{cittaṃ sthiraṃ yasya vinā'valambāt \\+}
-\tl{sa eva yogī sa guruḥ sa sevyaḥ ||\\!}
+\tl{cittaṃ sthiraṃ yasya vināvalambaṃ \\+}
+\tl{sa rājayogī sa guruḥ sa sevyaḥ ||\\+}
+\tl{\var{rāja°] B220, rāva° 2244}\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp04_074]
-\emph{Yogacintāmaṇi} (attrib.~to the \emph{Haṭhapradīpikā})
-\begin{versinnote}
-\tl{sarve layahaṭhābhyāsā rājayogasya siddhaye |\\+}
+\emph{Yogacintāmaṇi} f. 8r (attrib.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{sarve haṭhalayābhyāsād rājayogasya siddhaye |\\+}
 \tl{rājayogaṃ samārūḍhaḥ puruṣaḥ kālavañcaka iti ||\\!}
@@
 O great yogis, experience the nectar of Haṭha, the essence extracted from the ocean of all yoga scriptures after it has thus been churned, if you wish not to grow old and die.
-\end{translation}
+\end{translation} %JB: if you do not wish for old age and death.
 
@@
 \subsection*{4.75*1}
-\begin{translation}[hp04_75_1]
-The wise people in the world wash away sin at the sacred site of knowledge (\emph{vidyātīrthe}), the virtuous at the sacred site of truth (\emph{satyatīrthe}), the impure-minded at the sacred site of Gaṅgā (\emph{gaṅgātīrthe}), yogis at the sacred site of knowledge (\emph{jñānatīrthe}), kings at the sacred site of the streams (\emph{dhārātīrthe}), the rich at the sacred site of charity (\emph{dānatīrthe}) [and] women of good family at the sacred site of modesty.
+\begin{translation}[hp04_75_1] The wise people in the world wash away sin at the sacred site of knowledge (\emph{vidyātīrthe}), the virtuous at the sacred site of truth (\emph{satyatīrthe}), the impure-minded at the sacred site of the Gaṅgā (\emph{gaṅgātīrthe}), yogis at the sacred site of knowledge (\emph{jñānatīrthe}), kings at the sacred site of the streams (\emph{dhārātīrthe}), the rich at the sacred site of charity (\emph{dānatīrthe}) [and] women of good family at the sacred site of modesty.
 (\emph{lajjātīrthe})
 \end{translation}
+%
 
@@
 
-\begin{testimonia}[hp04_75_1]
-\emph{Yogacintāmaṇi} ??
-\end{testimonia}
+%\begin{testimonia}[hp04_75_1]
+%\emph{Yogacintāmaṇi} 
+%\end{testimonia}
 
@@
 
-\subsection*{Colophon}
+%%%%%%%%%%
+\subsection*{colophon}
 \begin{translation}[hp04_col]
-Thus ends the fourth chapter in the \emph{Haṭhapradīpikā} composed by the glorious lord amongst yogis Svātmārāma.
+Thus ends the fourth chapter in the \emph{Haṭhapradīpikā} composed by the glorious lord among yogis Svātmārāma.    
 \end{translation}

### 2024-07-08 latex/HP1_comm.tex
new input files

@@
 
-In his commentary {\emph Jyotsnā} on \emph{Haṭhapradīpikā} 1.1 Brahmānanda first states that the
+In his commentary \emph{Jyotsnā} on \emph{Haṭhapradīpikā} 1.1 Brahmānanda first states that the
 author Svātmārāma starts appropriately with a \emph{maṅgala}, a verse of adoration addressing
 Ādinātha, i.e.\ Śiva, but in accordance with his non-sectarian approach leaves room for a second
-interpretation of the word as Viṣṇu. A sentence later this is contradicted by his explanantion that Ādinātha, who first taught Yoga, taught it to Pārvatī, which limits the scope to Śiva.
+interpretation of the word as Viṣṇu. A sentence later this is contradicted by his explanantion that Ādinātha, who first taught Yoga, taught it to Pārvatī, which limits the scope to Śiva.\lb
 
@@
 However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below).
-It appears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.
+It appears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.\lb
 
@@
 \begin{versinnote}
-śrīādināthāya namo 'stu tasmai\\ 
-\ \ yenopadiṣṭā haṭhayogavidyā |\\
-vibhrājate pronnatarājayogam\\ 
-\ \ āroḍhum icchor adhirohiṇīva ||1.1||
+\tl{śrīādināthāya namo 'stu tasmai\\+} 
+\tl{\ \ yenopadiṣṭā haṭhayogavidyā\\+}
+\tl{vibhrājate pronnatarājayogam\\+}
+\tl{\ \ āroḍhum icchor adhirohiṇīva ||1.1||}
 \end{versinnote}
@@
 Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}.  
-The difference in meaning is negligable.
+The difference in meaning is negligable.\lb
 
 The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
-\emph{-rājasaudhaṃ}, which he explains as follows:
+\emph{-rājasaudhaṃ}, which he explains as follows:\lb
 
@@
   haṭhapradīpikāpi pronnatarājayogam āroḍhum icchor anā\-yāsena rājayogaprāpikā bhavatīti
-  upamālaṅkāraḥ.  indravajrākhyaṃ vṛttam.}
+  upamālaṅkāraḥ.  indravajrākhyaṃ vṛttam.}\lb
 
@@
 leads one to the flat on the top floor. And the comparison as outlined by Brahmānanda has all the
-elements deemed necessary by Sanskritic poetology:
+elements deemed necessary by Sanskritic poetology:\lb
 
 \begin{enumerate}
-\item Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva} 
-\item Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā
-    adhirohiṇī}) 
-\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The mūla text has
-    \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā
-    haṭhapradīpikā}) 
-\item Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})
+\item{Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva}}
+\item{Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā adhirohiṇī})}
+\item{Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The mūla text has \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā haṭhapradīpikā})}
+\item{Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})}
 \end{enumerate}
@@
 %%%%%%%%%%
+% \subsection*{1.17 heading}
+% athāsanāni/
+% \begin{translation}[hp01_017a]
+% \end{translation}
+% \begin{philcomm}[hp01_017a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.17}
@@
 %%%%%%%%%%
-\subsection*{1.35}
+\subsection*{1.35 heading}
 \begin{translation}[hp01_035a]
@@
 
+% \begin{philcomm}[hp01_035a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{1.35}
 \begin{translation}[hp01_035]
@@
 %%%%%%%%%%
+\subsection*{1.36 heading}
+\begin{translation}[hp01_036a]
+However, in another school [\emph{siddhāsana} is taught as follows]:
+\end{translation}
+
+% \begin{philcomm}[hp01_036a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.36}
 \begin{translation}[hp01_036]
-However, in another school [\emph{siddhāsana} is taught as follows]:\\
-
-Place the left heel on the penis and put the other heel on top: this is the adept's pose (\emph{siddhāsana}).\\
-
-Only the first teaching [on \emph{siddhāsana}] is accepted by me.\\
-
+Place the left heel on the penis and put the other heel on top: this is the adept's pose (\emph{siddhāsana}).
 \end{translation}
@@
 \end{philcomm}
+%%%%%%%%%%
+\subsection*{1.36 ending}
+\begin{translation}[hp01_036p]
+Only the first teaching [on \emph{siddhāsana}] is accepted by me.
+\end{translation}
+
+% \begin{philcomm}[hp01_036p]
+% \end{philcomm}
+
 %%%%%%%%%%
@@
 %%%%%%%%%%
+\subsection*{1.44 heading}
+\begin{translation}[hp01_044a]
+Now the lotus pose (\emph{padmāsana}).
+\end{translation}
+
+% \begin{philcomm}[hp01_044a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.44}
 \begin{translation}[hp01_044]
-Now the lotus pose (\emph{padmāsana}).\\
-
 Place the right foot on the left thigh, and the left on the right thigh, firmly hold the big toes with the hands crossed behind the back, put the chin on the chest and gaze at the tip of the nose.
@@
 %%%%%%%%%%
+\subsection*{1.45 heading}
+\begin{translation}[hp01_045a]
+However, in another school [\emph{padmāsana} is taught as follows]:
+\end{translation}
+
+% \begin{philcomm}[hp01_045a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.45--46}
 \begin{translation}[hp01_045]
-However, in another school [\emph{padmāsana} is taught as follows]:\\
-
 Carefully put the upturned feet on the thighs and the upturned hands in the middle of the thighs, fix the eyes on the tip of the nose, raise the root of the uvula with the tongue, place the chin on the chest, gently [draw in] the breath [...].
@@
 \begin{translation}[hp01_047]
-This is called the lotus pose [and] it cures all diseases. It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.\\
-
-Only the second teaching [on the lotus pose] is approved by me.  
+This is called the lotus pose [and] it cures all diseases. It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
 \end{translation}
@@
 In commenting on \emph{durlabham} in \emph{Jyotsnā} 2.74, Brahmānanda glosses it as \emph{duṣprāpam}, which means `difficult to attain' and `inaccessible.'
-
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{1.47 ending}
+\begin{translation}[hp01_047p]
+Only the second teaching [on the lotus pose] is approved by me.  
+\end{translation}
+
+\begin{philcomm}[hp01_047p]
 The comment added to this verse by Svātmārāma indicates that he prefers the following version of \emph{padmāsana}, which derives from the \emph{Vivekamārtaṇḍa}, rather than the one he has borrowed from the \emph{Dattātreyayogāśāstra}.
@@
 %%%%%%%%%%
+\subsection*{1.50 heading}
+\begin{translation}[hp01_050a]
+Now, the lion’s pose (\emph{siṃhāsana}).
+\end{translation}
+
+% \begin{philcomm}[hp01_050a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.50--52}
 \begin{translation}[hp01_050]
-Now, the lion’s pose (\emph{siṃhāsana}).
-
 [The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
@@
 %%%%%%%%%%
+\subsection*{1.53 heading}
+\begin{translation}[hp01_053a]
+Now, the friendly pose (\emph{bhadrāsana}).
+\end{translation}
+
+% \begin{philcomm}[hp01_053a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.53--54}
 \begin{translation}[hp01_053]
-Now, the friendly pose (\emph{bhadrāsana}).
-
 [The yogi] should put both ankles at the sides of the perineal seam below the scrotum.
@@
 %\end{philcomm}
+
+%%%%%%%%%%
+% \subsection*{1.54 endding}
+% \begin{translation}[hp01_054p]
+% % ity āsanāni
+% \end{translation}
+% \begin{philcomm}[hp01_054p]
+% \end{philcomm}
 
@@
 %%%%%%%%%%
+\subsection*{1.61 heading}
+\begin{translation}[hp01_061a]
+In the same vein there is a saying by Goraksa:
+\end{translation}
+
+% \begin{philcomm}[hp01_061a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{1.61}
 \begin{translation}[hp01_061]
-In the same vein there is a saying by Goraksa:
 One should avoid places near bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.

### 2024-07-08 latex/HP2_comm.tex
new input files

@@
 %%%%%%%%%%
+\subsection*{2.24 heading}
+\begin{translation}[hp02_024a]
+Among them is dhauti:
+\end{translation}
+
+% \begin{philcomm}[hp02_024a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.24}
 \begin{translation}[hp02_024]
-Among them is dhauti:\\ 
 {}[The yogi] should slowly swallow a moistened cloth  four finger-breadths in width and then draw it out. This ejection [of it] from the mouth is the dhauti technique.
@@
 %%%%%%%%%%
+\subsection*{2.26 heading}
+\begin{translation}[hp02_026a]
+Now the \emph{basti} technique.
+\end{translation}
+
+% \begin{philcomm}[hp02_026a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.26}
 \begin{translation}[hp02_026]
-Now the \emph{basti} technique.\\
 Squatting in water up to the navel with a reed inserted in the anus, [the yogi] should contract the perineal region (\emph{ādhāra°}). The [resultant] flushing is the basti technique.
@@
 %%%%%%%%%%
+\subsection*{2.30 heading}
+\begin{translation}[hp02_030a]
+Now \emph{neti}.
+\end{translation}
+
+% \begin{philcomm}[hp02_030a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.30}
 \begin{translation}[hp02_030]
-Now \emph{neti}.\\
 {[}The yogi] should insert a very smooth thread one handspan [in length] into the nasal passage and take it out through the mouth. This is called neti by the Siddhas.
@@
 %%%%%%%%%%
+\subsection*{2.32 heading}
+\begin{translation}[hp02_032a]
+Now \emph{trāṭaka}:
+\end{translation}
+
+% \begin{philcomm}[hp02_032a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.32}
 \begin{translation}[hp02_032]
-Now \emph{trāṭaka}:\\
 {}[The yogi] should concentrate and look at a small focal point with a fixed gaze until tears fall. The experts consider this to be trāṭaka.
@@
 %%%%%%%%%%
+\subsection*{2.34 heading}
+\begin{translation}[hp02_034a]
+Now \emph{naulī}:
+\end{translation}
+
+% \begin{philcomm}[hp02_034a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.34}
 \begin{translation}[hp02_034]
-Now \emph{naulī}:\\
 With the shoulders lowered, [the yogi] should rotate the stomach to the left and right with the speed of a rapid whirlpool. This is called naulī by people from Gauḍa.%
@@
 %%%%%%%%%%
+\subsection*{2.36 heading}
+\begin{translation}[hp02_036a]
+Now kapālabhātī:
+\end{translation}
+
+% \begin{philcomm}[hp02_036a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.36}
 \begin{translation}[hp02_036]
-Now kapālabhātī:\\
 Rapid inhalation and exhalation like the bellows of a blacksmith is called kapālabhātī, the skull bellows. It dries up imbalances of phlegm.
@@
 %%%%%%%%%%
+\subsection*{2.48 heading}
+\begin{translation}[hp02_048a]
+Now, piercing the sun:
+\end{translation}
+
+% \begin{philcomm}[hp02_048a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.48}
 \begin{translation}[hp02_048]
-Now, piercing the sun---\\
 The yogi should sit in \emph{vajrāsana} on a comfortable mat, slowly draw in external air through the right nostril, [...]
@@
 %%%%%%%%%%
+\subsection*{2.51 heading}
+\begin{translation}[hp02_051a]
+Now ujjāyī:
+\end{translation}
+
+% \begin{philcomm}[hp02_051a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.51}
 \begin{translation}[hp02_051]
-Now ujjāyī:
-
-[The yogi] should close the mouth and gradually draw in the breath through the nostrils so that it comes into contact [with the region] from the throat to the chest and makes a sound.\end{translation}
+[The yogi] should close the mouth and gradually draw in the breath through the nostrils so that it comes into contact [with the region] from the throat to the chest and makes a sound.
+\end{translation}
 
@@
 %%%%%%%%%%
+\subsection*{2.54 heading}
+\begin{translation}[hp02_054a]
+Now sītkā:
+\end{translation}
+
+% \begin{philcomm}[hp02_054a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.54}
 \begin{translation}[hp02_054]
-Now sītkā---
-
 {}[The yogi] should continuously make \emph{sīt} sound in the mouth and flare his nostrils. By practising in this way he becomes a second god of love.
@@
 %%%%%%%%%%
+\subsection*{2.57 heading}
+\begin{translation}[hp02_057a]
+And the very same has been taught [as follows]:
+\end{translation}
+
+% \begin{philcomm}[hp02_057a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.57}
 \begin{translation}[hp02_057]
-And the very same has been taught [as follows]:
-
 He who continuously takes in the breath through the tongue and the root of the palate has all his diseases cured in half a year.
@@
 \begin{philcomm}[hp02_057]
-Verse 2.57 seems to be describing an alternative method of \emph{sītkā kumbhaka}.
-The introductory phrase \emph{evam eva uktaṃ ca} suggests that the teaching in this verse is consistent with what preceded it.
-However, the first version of \emph{sītkā} appears to be done at the front of the mouth (\emph{vaktra}), whereas the next version is done at the back of the mouth (see below).%JM Not sure what the note is saying.
-This practice could be consistent with 2.54 although no mention is made of a sound.
 
@@
 %%%%%%%%%%
+\subsection*{2.58 heading}
+\begin{translation}[hp02_058a]
+Now śītalī:
+\end{translation}
+
+% \begin{philcomm}[hp02_058a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.58}
 \begin{translation}[hp02_058]
-Now śītalī:\\
 The wise man should draw in air through the tongue and after retaining the breath as before gradually exhale through the nostrils .
@@
 %%%%%%%%%%
+\subsection*{2.60 heading}
+\begin{translation}[hp02_060a]
+Now bhastrikā:
+\end{translation}
+
+% \begin{philcomm}[hp02_060a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.60}
 \begin{translation}[hp02_060]
-Now bhastrikā:
-
 If [the yogi] places the soles of both feet on the thighs, the lotus pose, which destroys all bad deeds, duly arises.
@@
 %%%%%%%%%%
+\subsection*{2.69 heading}
+\begin{translation}[hp02_069a]
+Now bhrāmarī:
+\end{translation}
+
+% \begin{philcomm}[hp02_069a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.69}
 \begin{translation}[hp02_069]
-Now bhrāmarī:
-
 Forcibly loud inhalation with the sound of a male bee; very slow exhalation with the sound of a female bee: as a result of practising thus, there arises in the minds of the best yogis a certain blissful stupefaction.
@@
 %%%%%%%%%%
+\subsection*{2.70 heading}
+\begin{translation}[hp02_070a]
+Now mūrchā:
+\end{translation}
+
+% \begin{philcomm}[hp02_070a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.70}
 \begin{translation}[hp02_070]
-Now mūrchā:
-
 At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This fainting of the mind, which is called \emph{mūrchanā}, gives pleasure.
@@
 %%%%%%%%%%
+\subsection*{2.71 heading}
+\begin{translation}[hp02_071a]
+Now plāvanī:
+\end{translation}
+
+% \begin{philcomm}[hp02_071a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.71}
 \begin{translation}[hp02_071]
-Now plāvanī:\\
 With his abdomen completely filled with the wind of eructation, which has been turned inwards, [the yogi] floats easily even in deep water, like a lotus leaf.
@@
 %%%%%%%%%%
+% \subsection*{2.71 ending}
+% iti plāvanīkumbhakam/
+% \begin{translation}[hp02_071p]
+% \end{translation}
+% \begin{philcomm}[hp02_071p]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{2.72 heading}
+\begin{translation}[hp02_072a]
+Now \emph{kevalakumbhaka}:
+\end{translation}
+
+% \begin{philcomm}[hp02_072a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{2.72}
 \begin{translation}[hp02_072]
-Now \emph{kevalakumbhaka}:
 

### 2024-07-08 latex/HP3_comm.tex
new input files

@@
 %%%%%%%%%%
+\subsection*{3.4 heading}
+\begin{translation}[hp03_004a]
+What is ``the empty pathway"?
+\end{translation}
+
+% \begin{philcomm}[hp03_004a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.4}
-What is ``the empty pathway"?
-
 \begin{translation}[hp03_004]
@@
 %%%%%%%%%%
+\subsection*{3.10 heading}
+\begin{translation}[hp03_010a]
+Now the great seal (\emph{mahāmudrā}):
+\end{translation}
+
+% \begin{philcomm}[hp03_010a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.10}
 \begin{translation}[hp03_010]
-Now the great seal (\emph{mahāmudrā}):\\
 {}[The yogi] should press the perineum with the heel of the left foot, hold [the foot of] the extended leg with the hands and breathe in through the mouth.
@@
 %%%%%%%%%%
+\subsection*{3.19 heading}
+\begin{translation}[hp03_019a]
+Now, the great lock (\emph{mahābandha}):
+\end{translation}
+
+% \begin{philcomm}[hp03_019a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.19}
 \begin{translation}[hp03_019]
-Now, the great lock (\emph{mahābandha}):\\
 {}[The yogi] should place the heel of the left foot on the perineal region. And he should put the right foot on the left thigh, [\dots]
@@
 \begin{translation}[hp03_025]
-Now the great piercing (\emph{mahāvedha}):\\
 Like a beautiful and charming woman without a man, the great seal and the great lock are barren without the great piercing.\\
@@
 \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.26 heading}
+\begin{translation}[hp03_026a]
+Now the great piercing (\emph{mahāvedha}):
+\end{translation}
+
+% \begin{philcomm}[hp03_026a]
+% \end{philcomm}
 
@@
 %%%%%%%%%%
+\subsection*{3.33 heading}
+\begin{translation}[hp03_033a]
+Now, the sky-roving [seal] (\emph{khecarī}):
+\end{translation}
+
+% \begin{philcomm}[hp03_033a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.33}
 \begin{translation}[hp03_033]
-Now, the sky-roving [seal] (\emph{khecarī}):\\
 By cutting, moving, and milking, the yogi should gradually lengthen the tongue until it touches the middle of the brows. Then there is success in \emph{khecarī}.
@@
 %%%%%%%%%%
+\subsection*{3.50 heading}
+\begin{translation}[hp03_050a]
+The root lock (\emph{mūlabandha}):
+\end{translation}
+
+% \begin{philcomm}[hp03_050a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.50}
 \begin{translation}[hp03_050]
-The root lock (\emph{mūlabandha})—\\
 When [the yogi] presses the perineum with part of the heel, clenches the anus and draws up \emph{apāna}, it is called the root lock.
@@
 %%%%%%%%%%
+\subsection*{3.59 heading}
+\begin{translation}[hp03_059a]
+Now the Uḍḍīyaṇa lock:
+\end{translation}
+
+% \begin{philcomm}[hp03_059a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.59}
 \begin{translation}[hp03_059]
-Now the Uḍḍīyaṇa lock:\\
 Because the breath flies up (\emph{uḍḍīyate}) into the Suṣumṇā when it is bound by it, yogis say that this [practice] is called Uḍḍīyaṇa.
@@
 %%%%%%%%%%
+\subsection*{3.67 heading}
+\begin{translation}[hp03_067a]
+Now the \sl{jālandhara} lock:
+\end{translation}
+
+% \begin{philcomm}[hp03_067a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.67}
 \begin{translation}[hp03_067]
-Now the \sl{jālandhara} lock–\\
 {}[The yogi] should contract the throat and firmly place the chin on the chest. This is the lock called \sl{jālandhara}. It prevents loss of the nectar of immortality.
@@
 %%%%%%%%%%
+\subsection*{3.74 heading}
+\begin{translation}[hp03_074a]
+Now the inverted bodily position:
+\end{translation}
+
+% \begin{philcomm}[hp03_074a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.73*1}
 \begin{translation}[hp03_073_1]
-Now the inverted bodily position–\\
 The sun devours whatever nectar flows from the divine moon. As a result, the body is afflicted by old age.
@@
 %%%%%%%%%%
+\subsection*{3.79 heading}
+\begin{translation}[hp03_079a]
+Now \emph{vajrolī}:
+\end{translation}
+
+% \begin{philcomm}[hp03_079a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.79}
 \begin{translation}[hp03_079]
-Now \emph{vajrolī}:\\
 Even if he behaves as he wishes without [following] the observances (\emph{niyama}) taught in yoga, the [yogi] who knows \emph{vajrolī} is worthy of success.
@@
 %%%%%%%%%%
+\subsection*{3.88 heading}
+\begin{translation}[hp03_088a]
+Now \emph{sahajolī}:
+\end{translation}
+% atha sahajolī/
+% \begin{philcomm}[hp03_088a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.88}
@@
 \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{3.91 heading}
+\begin{translation}[hp03_091a]
+Now \emph{amarolī}:
+\end{translation}
+
+% \begin{philcomm}[hp03_091a]
+% \end{philcomm}
 
@@
 %%%%%%%%%%
+\subsection*{3.94*2 heading}
+\begin{translation}[hp03_094_2a]
+Now the Stimulation of the Goddess (\emph{śakticālanam}):
+\end{translation}
+
+% \begin{philcomm}[hp03_094_2a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.94*2}
 \begin{translation}[hp03_094_2]
-Now the Stimulation of the Goddess (\emph{śakticālanam}):\\ 
 She whose body is bent (\emph{kuṭilāṅgī}), she who is coiled (\emph{kuṇḍalinī}), the female snake (\emph{bhujaṅgī}), the power (\emph{śakti}), the goddess (\emph{īśvarī}), she who is coiled (\emph{kuṇḍalī}) and Arundhatī: these words are synonyms.
@@
 %%%%%%%%%%
+\subsection*{3.103*1 ending}
+% iti śakticālanaṃ/
+\begin{translation}[hp03_103_1p]
+\end{translation}
+
+% \begin{philcomm}[hp03_103_1p]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{3.104}

### 2024-07-08 latex/HP4_comm.tex
new input files

@@
 %\chapter*{Translation \& philological commentary}
+
+%%%%%%%%%%
+\subsection*{4.1 heading}
+\begin{translation}[hp04_001a]
+Now \emph{samādhi}:
+\end{translation}
+
+% \begin{philcomm}[hp04_001a]
+% \end{philcomm}
 
@@
 \begin{translation}[hp04_008_2]
-Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
+Some are confused by a multitude of tantric texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
 \end{translation}
@@
 %%%%%%%%%%
+\subsection*{4.8*8 heading}
+\begin{translation}[hp04_008_8a]
+Now \emph{khecarī}:
+\end{translation}
+
+% \begin{philcomm}[hp04_008_8a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.8*8}
 \begin{translation}[hp04_008_8]
-Now \emph{khecarī}:\\
 There is a hollow that generates knowledge and has five streams. \emph{Khecarīmudrā} is situated in that pure void.
@@
 %%%%%%%%%%
+% \subsection*{4.8*21 ending}
+% iti khecarī/
+% \begin{translation}[hp04_008_21p]
+% \end{translation}
+
+% \begin{philcomm}[hp04_008_21p]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.8*22 heading}
+\begin{translation}[hp04_008_22a]
+Now \emph{śāmbhavī}:
+\end{translation}
+
+% \begin{philcomm}[hp04_008_22a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.8*22}
 \begin{translation}[hp04_008_22]
-Now \emph{śāmbhavī}:\\
 {[}The yogi] should [put] the mind in Śakti and Śakti in the mind, observe the mind with the mind, and meditate on it as the supreme state.
@@
 \end{metre}
+
+%%%%%%%%%%
+% \subsection*{4.33_8 heading}
+% atha viśrāntiḥ/
+% \begin{translation}[hp04_033_8a]
+% \end{translation}
+
+% \begin{philcomm}[hp04_033_8a]
+% \end{philcomm}
 
@@
 %%%%%%%%%%
+\subsection*{4.14 heading}
+\begin{translation}[hp04_014a]
+What kind of indifference?
+\end{translation}
+
+% \begin{philcomm}[hp04_014a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.14}
 \begin{translation}[hp04_014]
-What kind of indifference?\\
 In the cold season, [indifference towards] whether [one has] an open tent or a quilt, with regard to good nourishment, whether it is cow's milk or water, with regard to food, whether it is lots of alms or forest roots, and with regard to the vessel for food, whether it is the hand or some kind of bucket.
@@
 %%%%%%%%%%
+\subsection*{4.17 heading}
+\begin{translation}[hp04_017a]
+Among these, the beginning stage is [as follows]:
+\end{translation}
+
+% \begin{philcomm}[hp04_017a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.17}
 \begin{translation}[hp04_017]
-Among these, the beginning stage is [as follows]:\\
 As a result of the piercing of the knot of Brahmā, bliss arises in the void [and] an unstruck sound which has various tones is heard in the body.
@@
 %%%%%%%%%%
+\subsection*{4.19 heading}
+\begin{translation}[hp04_019a]
+Now the unified stage:
+\end{translation}
+
+% \begin{philcomm}[hp04_019a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.19}
 \begin{translation}[hp04_019]
-Now the unified stage:\\
 In the second stage the breath, after bringing about union, goes into the middle [channel]. Then the yogi has a firm posture [and] he becomes a gnostic and equal to a god.
@@
 \begin{translation}[hp04_020]
-Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a \textit{bherī} drum occurs.%
+Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a kettle drum occurs.%
 \end{translation}
@@
 %%%%%%%%%%
+\subsection*{4.21 heading}
+\begin{translation}[hp04_021a]
+Now the accumulation stage:
+\end{translation}
+
+% \begin{philcomm}[hp04_021a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.21}
 \begin{translation}[hp04_021]
-Now the accumulation stage:\\
-In the third [stage], having pierced [the knot of Viṣṇu], the sound of a \textit{mardala} drum arises in space. Then [the yogi] reaches the great void, the abode of all supernatural powers.%
+In the third [stage], having pierced [the knot of Viṣṇu], the sound of a bass drum arises in space. Then [the yogi] reaches the great void, the abode of all supernatural powers.%
 \end{translation}
@@
 %%%%%%%%%%
+\subsection*{4.23 heading}
+\begin{translation}[hp04_023a]
+Now the perfection stage:
+\end{translation}
+
+% \begin{philcomm}[hp04_023a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.23}
 \begin{translation}[hp04_023]
-Now the perfection stage:\\
 Then, having pierced Rudra's knot, the breath goes to all the seats [of the deities in the body]. In the perfected [stage] the sound of a flute becomes the sound of a resonating lute.
@@
 \begin{translation}[hp04_024]
-Then the mind having become one [with the sound]. This is called Rājayoga. He becomes a creator and destroyer, an equal to a lord among yogis.%
+Then the mind becomes one [with the sound]. This is called Rājayoga. He becomes a creator and destroyer, an equal to a lord among yogis.%
 \end{translation}
@@
 \tl{purā matsyendrabodhāya ādināthoditavacaḥ |\\+}
-\tl{manaḥ pākam avāpnoti nirālambākhyaghoṭanam ||\\!}
+\tl{manaḥpākam avāpnoti nirālambākhyaghoṭanam ||\\!}
 \end{versinnote}
@@
 \end{translation}
-% nādo’ntaraṅgasāraṅgabandhane vāgurāyate
-%V15 antaraṅgaturaṅgasya nādo vyādhāyate pi ca
+% ?? JB: we have not translated antaraṅga° in pāda a. Change to 'trapping the inner deer that is the mind'?
 
@@
 %%%%%%%%%%
+\subsection*{4.52*4 ending}
+% iti nādānusaṃdhānam/
+\begin{translation}[hp04_052_4p]
+\end{translation}
+
+% \begin{philcomm}[hp04_052_4p]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.52*5}
@@
 \end{metre}
+
+%%%%%%%%%%
+\subsection*{4.56 heading}
+\begin{translation}[hp04_056a]
+And so, Vasiṣṭha [said]:
+\end{translation}
+
+% \begin{philcomm}[hp04_056a]
+% \end{philcomm}
+
 %%%%%%%%%%
@@
 \begin{translation}[hp04_056]
-And so, Vasiṣṭha [said]:\\
 The moon and sun move in Iḍā and Piṅgalā. The moon is said to be of the nature of \emph{tamas} and the sun of \emph{rajas}.
@@
 %%%%%%%%%%
+\subsection*{4.58 heading}
+\begin{translation}[hp04_058a]
+For as the tetrad of verses called the Saubhadra has it:
+\end{translation}
+
+\begin{philcomm}[hp04_058a]
+We do not know why this tetrad of verses is called Saubhadra.
+\end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.58}
 \begin{translation}[hp04_058]
-For as the tetrad of verses called the Saubhadra has it:\\
 There are six cakras, sixteen supports, three focal points and three \emph{guṇa}s. Everything else is [just] the prolixity of texts. Trikūṭa is the supreme place.
@@
 \begin{philcomm}[hp04_058]
-We do not know why this tetrad of verses is called Saubhadra.\\
-
 The three components of the yogic body listed here are found together in other texts, the earliest being \emph{Netratantra} 7.1ab (\emph{ṛtucakraṃ svarādhāraṃ trilakṣyaṃ vyomapañcakam}).
@@
 %%%%%%%%%%
+\subsection*{4.70 heading}
+\begin{translation}[hp04_070a]
+Thus, Viśvarūpācārya [said]:
+\end{translation}
+
+% \begin{philcomm}[hp04_070a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.70}
 \begin{translation}[hp04_070]
-Thus, Viśvarūpācārya [said]:\\
 When the breath is destroyed and the mind dissolves, all experience is the same (\emph{samarasatva}). That is called \emph{samādhi}.

### 2025-04-04 latex/HP1_comm.tex
new files

@@
 Ādinātha, i.e.\ Śiva, but in accordance with his non-sectarian approach leaves room for a second
-interpretation of the word as Viṣṇu. A sentence later this is contradicted by his explanantion that Ādinātha, who first taught Yoga, taught it to Pārvatī, which limits the scope to Śiva.\lb
+interpretation of the word as Viṣṇu. A sentence later this is contradicted by his explanantion that Ādinātha, who first taught Yoga, taught it to Pārvatī, which limits the scope to Śiva. \lb
 
@@
 However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below).
-It appears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.\lb
+It appears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.2d and 1.3b), even at the expense of the poetic imagery.
+\lb
 
@@
 \begin{versinnote}
-\tl{śrīādināthāya namo 'stu tasmai\\+} 
-\tl{\ \ yenopadiṣṭā haṭhayogavidyā\\+}
-\tl{vibhrājate pronnatarājayogam\\+}
-\tl{\ \ āroḍhum icchor adhirohiṇīva ||1.1||}
+śrīādināthāya namo 'stu tasmai\\ 
+\ \ yenopadiṣṭā haṭhayogavidyā |\\
+vibhrājate pronnatarājayogam\\ 
+\ \ āroḍhum icchor adhirohiṇīva ||1.1||
 \end{versinnote}
@@
 Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}.  
-The difference in meaning is negligable.\lb
+The difference in meaning is negligable.
 
 The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
-\emph{-rājasaudhaṃ}, which he explains as follows:\lb
+\emph{-rājasaudhaṃ}, which he explains as follows:
 
@@
   haṭhapradīpikāpi pronnatarājayogam āroḍhum icchor anā\-yāsena rājayogaprāpikā bhavatīti
-  upamālaṅkāraḥ.  indravajrākhyaṃ vṛttam.}\lb
+  upamālaṅkāraḥ.  indravajrākhyaṃ vṛttam.}
 
@@
 leads one to the flat on the top floor. And the comparison as outlined by Brahmānanda has all the
-elements deemed necessary by Sanskritic poetology:\lb
+elements deemed necessary by Sanskritic poetology:
 
 \begin{enumerate}
-\item{Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva}}
-\item{Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā adhirohiṇī})}
-\item{Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The mūla text has \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā haṭhapradīpikā})}
-\item{Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})}
+\item Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva} 
+\item Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā
+    adhirohiṇī}) 
+\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The mūla text has
+    \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā
+    haṭhapradīpikā}) 
+\item Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})
 \end{enumerate}
@@
 The reading \emph{rājamārgam ajānatām} (\alphaOne and \alphaThree), “for those unable to find the royal highway”, is more appropriate to the metaphor of being lost in darkness.
+\lb
 
@@
 Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it.
-The variant \emph{mahāyogī} in \epsilonOne\ and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}.
+The variant \emph{mahāyogī} in \zetaOne\ and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}.
 One could emend to \emph{tathā} in light of the attested reading \emph{yathā} (\getsiglum{C7}) but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.
@@
 \begin{philcomm}[hp01_005]     
-In Śaiva texts which predate the Haṭha corpus, Mīnanātha and Matsyendra are one and the same, but they are differentiated in later Tibetan and Indian lists of siddhas (Mallinson 2019: 273 n.35).
+In Śaiva texts which predate the Haṭha corpus, Mīnanātha and Matsyendra are one and the same, but they are differentiated in later Tibetan and Indian lists of siddhas (Mallinson 2019: 273 n.35). \lb
 
@@
 We are yet to find the name \emph{goraṇṭaka} in other Sanskrit texts but it may be a Sanskrit rendering of \emph{Goraṇṭakuḍu}, which is the name of a disciple of Gorakṣanātha in the \emph{Navanāthacaritramu} (Jones 2017: 194 n.3).
-The spelling \emph{koraṇṭaka} is attested in the \emph{Haṭhābhyāsapaddhati}, and it is reasonably well attested by manuscripts of the \emph{Haṭhapradīpikā}, as well as those of the \emph{Haṭharatnāvalī} (which also has \emph{gonandaka}).
+The spelling \emph{koraṇṭaka} is attested in the \emph{Haṭhābhyāsapaddhati}, and it is reasonably well attested by manuscripts of the \emph{Haṭhapradīpikā}, as well as those of the \emph{Haṭharatnāvalī} (which also has \emph{gonandaka}).\lb
 
@@
 The variant \emph{dhvaninātha} may have resulted from a transposition of the first two syllables of \emph{nityanātha}.
+\lb
 
@@
 \begin{translation}[hp01_008]
-Allamaprabhudeva, Ghoḍācolī, Ṭiṇṭiṇī, Bhālukī and Nāgabodha and Khaṇḍakāpālika.
+Allamaprabhudeva, Ghoḍācolī, Ṭiṇṭiṇi, Bhāluki and Nāgabodha and Khaṇḍakāpālika.
 \end{translation}
@@
 8).
-
-The names Nāgabodha, Nāgabodhi, Naradeva, Nāgadeva all seem possible in 1.8c. The reading \emph{nāgabodhaś ca} is attested across several primary groups of manuscripts.
+\lb
+
+The names Nāgabodha, Nāgabodhi, Nāradeva, Nāgadeva all seem possible in 1.8c. The reading \emph{nāgabodhaś ca} is attested across several primary groups of manuscripts. \lb
 
@@
 In several other stories (\textit{Kathāsaritsāgara} 6.6.132, 10.5.89, 12.25.35), \emph{maṭhikā} is the term used for the hut of an ascetic.
-In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four "\emph{hasta}s" high and wide (there are various definitions of the term: 18 inches according to the Larger Petrograd Dictionary and Monier-Williams, 48 inches according to the Smaller Petrograd Dictionary; the NWS lists even more variations, but the stipulation that the hut is the length of a bow found in the \emph{Gorakṣaśataka} (on which see below) suggests that 18 inches was meant).
-The hut can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).
+In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four ``\emph{hasta}s" high and wide (there are various definitions of the term: 18 inches according to the Larger Petrograd Dictionary and Monier-Williams, 48 inches according to the Smaller Petrograd Dictionary; the NWS lists even more variations, but the stipulation that the hut is the length of a bow found in the \emph{Gorakṣaśataka} (on which see below) suggests that 18 inches was meant).
+The hut can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).\lb
 
@@
 However, the compounds in the first three verse-quarters appear to qualify \emph{lakṣaṇa} as though they were adjectives, and this seems to have been the way the verse was composed.
+\lb
 
@@
 For this reason, the reading \emph{°vivaraṃ} looks like a patch, as its meaning does not add anything to \emph{°randhragarta°}.
-The reading \emph{°viṭapaṃ} (`the young branch of a tree or creeper') attested in some manuscripts of the \emph{Haṭhapradīpikā} is difficult to construe in this context unless it was intended to refer to creepers or branches that might invade or encroach upon the hut.
+The reading \emph{°viṭapaṃ} (`the young branch of a tree or creeper') attested in some manuscripts of the \emph{Haṭhapradīpikā} is difficult to construe in this context unless it was intended to refer to creepers or branches that might invade or encroach upon the hut.\lb
 
@@
 %JB Okay I've written the note to reflect this. 
-
-Manuscripts of several groups, namely \textbeta, \textepsilon\ and \texteta, have \emph{°bādhojjitaṃ}, whereas \textdelta\ and the \emph{Yogacintāmaṇi} have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’).
-The \textalpha\ group is split on this, with \getsiglum{G4} (\emph{bodhojhitaṃ}) closer to \emph{°bādhojjitaṃ} and \getsiglum{N3} (\emph{jyaṃtyūpsitaṃ} and \getsiglum{J5} (\emph{jaṃtūṣṇitaṃ}) closer to \emph{°jantūjjhitaṃ}.
-We have adopted the more unusual reading of \emph{°bādhojjitaṃ}  with the support of a similar description of a hut in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
+% MD 2024-8: amalaṃ or vimalaṃ? amalaṃ best
+
+Manuscripts of several groups, namely \textbeta, \textepsilon\ and \texteta, have \emph{°bādhojjitaṃ}, whereas \textgamma, \textdelta\ and the \emph{Yogacintāmaṇi} have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’).
+The \textalpha\ group is split on this, with \alphaThree\ (\emph{bodhojhitaṃ}) closer to \emph{°bādhojjhitaṃ} and \alphaOne\ (\emph{jyaṃtyūpsitaṃ} and \alphaTwo\ (\emph{jaṃtūṣṇitaṃ}) closer to \emph{°jantūjjhitaṃ}.
+We have adopted the more unusual reading of \emph{°bādhojjhitaṃ}  with the support of a similar description of a hut in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
 
@@
 \begin{translation}[hp01_014]
-Staying in such a hut, free from all worry, in the way taught by his guru [the yogi] should practise nothing but yoga.
+Staying in such a hut, free from all worry, [the yogi] should regularly practise yoga in exactly the way taught by his guru.
 \end{translation}
@@
 \begin{testimonia}[hp01_014]
-\emph{Yogacintāmaṇi} f.54r (attr. \emph{Haṭhapradīpikā})
+\emph{Haṭharatnāvalī} 1.68
 
@@
 
-\emph{Haṭharatnāvalī} 1.68
+\emph{Yogacintāmaṇi} f.54r (attr. \emph{Haṭhapradīpikā})
 
@@
 %\end{philcomm}
+% MD 2024-8: samabhyaset or sadābhyaset?
 
@@
 Although \emph{yama} and \emph{niyama} are not included in the \emph{Haṭhapradīpikā} as auxiliaries of Haṭhayoga, verse 2.14 implies that \emph{niyama} is necessary at least in the early stages of establishing a practice.
-Furthermore, verse 3.82 suggests that a yogi who does not practice \emph{niyama} might obtain success in yoga through the practice of \textit{vajrolī}.
+Furthermore, verse 3.79 suggests that a yogi who does not practice \emph{niyama} might obtain success in yoga through the practice of \textit{vajrolī}.
 Ambiguity over the role of \emph{yama} and \emph{niyama} in Haṭhayoga may explain why two verses on ten \emph{yama}s and ten \emph{niyama}s were inserted in some manuscripts after the next verse (1.16).
@@
 \begin{translation}[hp01_016]
-Zeal, daring, resolve, gnosis of the truth, conviction and avoiding contact with people: by means of six [virtues], yoga is successful.
-\end{translation}
+Zeal, conviction, resolve, contentment, gnosis of the truth, and avoiding contact with people: through these six [virtues], yoga is successful.
+\end{translation}
+% JB; change to 'through [these] six, ...' (to be consistent with the previous verse?) And should we reflect the different syntax of this verse (i.e., the ablatives)?
 
@@
 \begin{testimonia}[hp01_016]
-\emph{Yogacintāmaṇi} f. 49r (attr. \emph{Haṭhapradīpikā})
-
-\begin{versinnote}
-\tl{utsāhāt sāhasād dhairyāt tatvajñānād viniścayāt |\\+}
-\tl{janasaṅgaparityāgāt ṣaḍbhir yogaḥ prasidhyati ||\\!}
-\end{versinnote}
-
-\emph{Haṭharatnāvalī} 1.78:
+\emph{Haṭharatnāvalī} 1.78
 
@@
 
+\emph{Yogacintāmaṇi} f. 49r (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{utsāhāt sāhasād dhairyāt tatvajñānād viniścayāt |\\+}
+\tl{janasaṅgaparityāgāt ṣaḍbhir yogaḥ prasidhyati ||\\!}
+\end{versinnote}
+
 \end{testimonia}
@@
 \begin{philcomm}[hp01_016]    
-\textalpha\ and several other groups of manuscripts have \emph{tattvajñānāc ca darśanāt} or something very similar in the second \emph{pāda} of the verse, but \emph{darśana} by itself is problematic: a vision of what?
+\textalpha\ and several other groups of manuscripts have \emph{tattvajñānārthadarśanāt} (as found in \emph{Bhagavadgītā} 13.11), \emph{tattvajñānāc ca darśanāt} or something very similar in the second \emph{pāda} of the verse, but \emph{darśana} (\alphaOne) by itself is problematic: a vision of what?
 The early sources of this verse, in particular the \emph{Śivadharmottara}, indicate that the second verse quarter read as \emph{santoṣāt tattvadarśanāt}, which makes much better sense of the word \emph{darśana} (i.e., ‘seeing the truth’).
-However it seems likely that before the time of Svātmārāma other versions of this verse were circulating in which \emph{santoṣāt} was not found, \emph{niścayāt} had moved from the first to second verse quarter, \emph{tattvadarśanāt} had become \emph{tattvajñānāt} and \emph{sāhasāt} was introduced.
-It should also be noted that the word \emph{tattva} could have a more specific meaning in the \emph{Haṭhapradīpikā} (4.45–46) as Svātmārāma states that it is a synonym of \emph{samādhi}.
+\lb
+%However it seems likely that before the time of Svātmārāma other versions of this verse were circulating in which \emph{santoṣāt} was not found, \emph{niścayāt} had moved from the first to second verse quarter, \emph{tattvadarśanāt} had become \emph{tattvajñānāt} and \emph{sāhasāt} was introduced.
+
+It should also be noted that the word \emph{tattva} could have a more specific meaning in the \emph{Haṭhapradīpikā} (4.32–33) as Svātmārāma states that it is a synonym of \emph{samādhi}.
 In other yoga texts, it can sometimes refer to the practices of yoga (e.g.
@@
 \end{philcomm}
+
+% MD 2024-8: Are we really remaining with the reading sāhasād of gamma/delta?
+The compound tattvajñānārthadarśana occurs in Bhagavadgītā 13.11.
+Does it not work either?
+In BhG it is a single thing and we need two.
 
@@
 \begin{philcomm}[hp01_017]        
-The reading \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the main one of the \textalpha\ group.
+The reading \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the \textalpha\ group.
 Although this compound rarely appears in other yoga texts, a similar term \emph{śarīrapāṭava} occurs in the \emph{Śivasaṃhitā} (2.35) as one of the benefits bestowed by digestive fire (\emph{vaiśvānarāgni}), which indicates that the word \emph{pāṭava} was used in relation to the body and the benefits of yoga.
@@
 The variant reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts and similar formulations occur even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}).
-It is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by manuscripts in several early groups (i.e., \textbeta, \textgamma\ and \textdelta).
+It is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early in the transmission, as the latter is attested by manuscripts in several early groups (i.e., \textbeta, \textgamma\ and \textdelta).
 \end{philcomm}
@@
 \begin{philcomm}[hp01_021]   
-Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by \getsiglum{G4} (\textalpha\ group) and \getsiglum{V1} (\texteta\ group), the sources and the testimonia.
+Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by \alphaThree\ and \etaOne, the sources and the testimonia.
 It appears to be verse filler here rather than indicating a temporal sequence of actions.
@@
 This would explain the rather strange syntax of the \emph{Vasiṣṭhasaṃhitā}’s version, in which \emph{adhaḥ kṛtvā} was changed to \emph{athaikasmin}, and \emph{tathetaram} became \emph{ca saṃsthitam}.
-It seems that \emph{saṃsthitaṃ} must be understood with \emph{ūruṃ} in the third \emph{pāda} in the sense of \emph{saṃsthāpya} (i.e., ‘having placed’).
+It seems that \emph{saṃsthitaṃ} must be understood with \emph{ūruṃ} in the third \emph{pāda} in the sense of \emph{saṃsthāpya} (i.e., ‘having placed’).\lb
 
@@
 \begin{translation}[hp01_022]
-Knowers of yoga know that the tortoise pose (\emph{kūrmāsana}) arises by carefully blocking the anus with the ankles crossed. 
+Knowers of yoga know that the tortoise pose (\emph{kūrmāsana}) arises by carefully pressing the anus with the ankles crossed. 
 \end{translation}
@@
 In the first quarter of the verse, the witnesses are split between \emph{nirudhya} (‘having blocked’), \emph{nibadhya} (‘having bound’), \emph{niyamya} (‘having restrained’) and \emph{niṣpīḍya} (‘having pressed’).
-The source, the \emph{Vasiṣṭhasaṃhitā}, and two manuscripts of the \textbeta\ and \textgamma\ groups support \emph{nirudhya} whereas one \textalpha\ manuscript (\getsiglum{J5}) and the testimonia support \emph{niyamya} and another \textalpha\ manuscript (\getsiglum{G4}) has \emph{niṣpīḍya}.
-In terms of blocking or closing the anus by sitting on the ankles, \emph{nirudhya} makes good sense, and \emph{niṣpīḍya} (`having pressed the anus with both ankles') is also possible.
+The source, the \emph{Vasiṣṭhasaṃhitā}, and two manuscripts of the \textbeta\ and \textgamma\ groups support \emph{nirudhya} whereas \alphaTwo\ and the testimonia support \emph{niyamya} and \alphaThree\ has \emph{niṣpīḍya}.
+While \emph{nirudhya} makes sense here (i.e., `having blocked or closed the anus...'), we have adopted \emph{niṣpīḍya} because it is better attested among the \textalpha, \epsilon\ and \eta\ groups.\lb
 
@@
 \begin{testimonia}[hp01_025]
-\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
-
-\begin{versinnote}
-\tl{pādāṅguṣṭhau ca pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\+}
-\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam ||\\!}
-\end{versinnote}
-
 \emph{Haṭharatnāvalī} 3.51
@@
 \tl{\var{ākarṣaṇaṃ kṛtvā ] ākarṣaṇākṛṣṭaṃ P,T,t1}\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})
+
+\begin{versinnote}
+\tl{pādāṅguṣṭhau ca pāṇibhyāṃ gṛhītvā śravaṇāvadhi |\\+}
+\tl{dhanurākarṣaṇaṃ kṛtvā dhanurāsanam īritam ||\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp01_025]
-Since the word \emph{ākarṣaṇa} in one form or other is so well attested in the third verse quarter, the reading \emph{dhanurākarṣaṇaṃ kṛtvā}, which is in \epsilonTwo, as well as the principal testimonia (i.e., the \emph{Yogacintāmaṇi} and \emph{Haṭharatnāvalī}), fits the overall syntax of the verse.
-However, it is curious that \emph{kṛṣṭaṃ} (for \emph{kṛtvā}) is well attested in some groups of manuscripts because it seems redundant with \emph{ākarṣaṇaṃ}.
-However, the following reading in
-Godāvaramiśra's \emph{Yogacintāmaṇi} (f. 40r) makes sense of \emph{kṛṣṭaṃ} and might indeed be the original version of the verse: \emph{dhanurākarṣavat kṛṣṭaṃ dhanurāsanam ucyate}.
-
-A different version of \emph{dhanurāsana} is described in the \emph{Haṭhayogasaṃhitā}. On the two versions of \emph{dhanurāsana}, see Hargreaves and Birch 2017.
+We have adopted the reading \emph{dhanurākarṣaṇaṃ kṛtvā}, which is in the \textdelta\ group, as well as the principal testimonia (i.e., the \emph{Yogacintāmaṇi} and \emph{Haṭharatnāvalī}), because it fits the overall syntax of the verse, unlike the readings with \emph{kṛṣṭaṃ} (for \emph{kṛtvā}) that are found in the early manuscripts.
+It is curious that \emph{kṛṣṭaṃ} is so well attested because it seems redundant with \emph{ākarṣaṇaṃ}.
+The following reading in Godāvaramiśra's \emph{Yogacintāmaṇi} (f.
+40r) appears to be an attempt to make sense of \emph{kṛṣṭaṃ}: \emph{dhanurākarṣavat kṛṣṭaṃ dhanurāsanam ucyate}.\lb
+% MD: The well attested reading is dhanurākarṣaṇākṛṣṭaṃ as one word. The commentary should be slightly modified.
+
+A different version of \emph{dhanurāsana} is described in the \emph{Haṭhayogasaṃhitā}. On the two versions of \emph{dhanurāsana}, see Hargreaves and Birch 2017.\lb
 %https://www.theluminescent.org/2017/11/dhanurasana-two-versions-of-bow-pose.html
@@
 The instrumental ending (`with the hand') works well with the gerund (\emph{pragṛhya}) in the third verse quarter and the object (\emph{°dakṣpādaṃ}) in the first quarter.
-This reading indicates that the right foot is held by the hand of the arm that is wrapped around the outside of the left leg, which would be the right hand (\emph{°dakṣadoṣṇā}) rather than the left (\emph{°vāmadoṣṇā}), as shown in \medialink{Figure 1}{matsyendrasana1.jpeg}.
-
-One manuscript of \textalpha\ (\getsiglum{G4}) and most manuscripts of the \textit{Haṭharatnāvalī} have \emph{°dakṣapādo} in the first \emph{pāda}.
+This reading indicates that the right foot is held by the hand of the arm that is wrapped around the outside of the left leg, which would be the right hand (\emph{°dakṣadoṣṇā}) rather than the left (\emph{°vāmadoṣṇā}), as shown in \medialink{Figure 1}{matsyendrasana1.jpeg}.\lb
+
+\alphaThree\ and most manuscripts of the \textit{Haṭharatnāvalī} have \emph{°dakṣapādo} in the first \emph{pāda}.
 This reading yields the same meaning as the adopted one if read with \emph{°vāmapādaṃ} in the second.
-However, \getsiglum{G4} and manuscripts of the \textit{Haṭharatnāvalī} read \emph{°vāmadoṣṇā}, which is not good because it leaves the gerund without an object.
+However, \alphaThree\ and manuscripts of the \textit{Haṭharatnāvalī} read \emph{°vāmadoṣṇā}, which is not good because it leaves the gerund without an object.\lb
 
@@
 \begin{translation}[hp01_027]
-Matsyendra's seat is a destructive missile for the many terrible diseases that develop in the stomach; through practice it brings about in men the awakening of Kuṇḍalinī and steadiness of the spine.
+Matsyendra's seat is a missile for destroying the many chronic and painful diseases of the stomach; through practice it brings about in men the awakening of Kuṇḍalinī and steadiness of the spine.
 \end{translation}
@@
 \begin{philcomm}[hp01_027]
+??
+Write new note by JM to reflect emendation to jaraṭha.
 The manuscript readings for the compound beginning with \emph{jaṭhara} diverge significantly and include \emph{jaṭharapravṛddha°}, \emph{jaṭharaprabuddha°}, \emph{jaṭharapradīpta°} and \emph{jaṭharapracaṇḍa°}.
 As descriptive compounds, none of these makes good sense in regard to Matsyendra's seat.
-Since the stomach or abdomen (\emph{jaṭhara}) is the first member of this compound, it seems more likely that it qualifies the terrible diseases (\emph{pracaṇḍarug}) that are mentioned in the next verse quarter, as suggested by the reading \emph{jaṭharapravṛddha°}, which is attested by \etaTwo\ and the \textit{Yogacintāmaṇi} and suggested by \gammaOne and \deltaOne (\emph{jaṭharapravuddh°}).
+Since the stomach or abdomen (\emph{jaṭhara}) is the first member of this compound, it seems more likely that it qualifies the terrible diseases (\emph{pracaṇḍarug}) that are mentioned in the next verse quarter, as suggested by the reading \emph{jaṭharapravṛddha°}, which is attested by \etaTwo\ and the \textit{Yogacintāmaṇi} and suggested by \gammaOne and \deltaOne (\emph{jaṭharapravuddha°}).\lb
+% MD 2024-8: The best-attested reading is now jaṭhara-pracaṇḍa-picaṇḍa-ruṅ-maṇḍala°.
+Perhaps, jaṭhara is an adjective here like jaraṭha: "a troop of tough/persistent (jaṭhara/jaraṭha) and terrible abdominal (picaṇḍa) diseases"?
 
@@
 JM: yes, but  karasthalābhyāṃ of alpha3 is even better and reflects the °talābhyāṃ of the parallels
+
+%MD 2024-8: Could we adopt the reading mayūram instead of māyūram im Pāda d? Yes.
+% Add comment on karadvayābhyāṃ and °sthalābhyāṃ?
 \end{translation}
@@
 \tl{uccāsano daṇḍavad utthitaḥ khe\\+}
-\tl{mayūram etat pravadanti pīṭham ||\\!}
+\tl{mayūram etat pravadanti pīṭham ||\\+}
+\tl{\var{mayūram ] cett., māyūram P, T, t1}\\!}
 \end{versinnote}
@@
 The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}.
+\lb
 
@@
 \begin{translation}[hp01_031]
-The glorious peacock [posture] quickly gets rid of bloating and all other diseases of the abdomen, and overcomes humoral imbalances.
+The glorious peacock posture quickly gets rid of bloating and all other diseases of the abdomen, and overcomes humoral imbalances.
 It reduces to ashes food which is bad or has been eaten to excess, kindles the digestive fire and causes strong poison to be digested.
 \end{translation}
+% MD: In 1.60 kadaśanaśākōtkaṭa was translated as  "an excess of leafy vegetables that are hard to chew". Perhaps it is better to interpret kadaśana in a similar sense. Indigestive food?
 
@@
 \begin{philcomm}[hp01_033]
-The word \emph{tu} is often used to introduce a new posture, but in this case seems to be a verse filler.
+The word \emph{tu} is often used to introduce a new posture, but in this case seems to be a verse filler.\lb
 
@@
 Svātmārāma also changes the meaning of the second half of the verse, as the \emph{Śivasaṃhitā} states that Śiva picked out the four best postures and taught them, whereas in the \emph{Haṭhapradīpikā} it reads as though Svātmārāma himself is responsible for picking out the four best postures and teaching them.
-There are other instances in the \emph{Haṭhapradīpikā} where Svātmārāma borrows a verse with a first person verb (e.g., 3.43, 4.2).
+There are other instances in the \emph{Haṭhapradīpikā} where Svātmārāma borrows a verse with a first person verb (e.g., 3.44).
 However, in this instance, he may have intended to indicate that he chose the four postures coming after this verse (i.e., \emph{siddha}, \emph{padma}, \emph{siṃha} and \emph{bhadra}) because the \emph{Śivasaṃhitā} follows 3.96 with teachings on the postures called \emph{siddha}, \emph{padma}, \emph{paścimottāna} and \emph{svastika}.
@@
 It is likely that the original version of this verse contained the vocative with the imperative form of the verb (\emph{sakhe tiṣṭha}).
-There are other instances where Svātmārāma included a verse with the vocative (e.g., 4.10, 4.12, 4.20, 4.72, 4.86, 4.88) as though the text were a dialogue.
+There are other instances where Svātmārāma included a verse with the vocative (e.g., 4.7c, 4.56b, 4.72d, 4.74c) as though the text were a dialogue.
+% MD: I have deleted the instances 4.0*14a and *16d since they are secondary.
 Other versions of this verse are transmitted by some manuscripts of the \emph{Haṭhapradīpikā}, in which the vocative and imperative verb have been removed.
@@
 \begin{philcomm}[hp01_035]
-In the second verse quarter, the adopted reading \emph{ekahṛdayo} is supported by two manuscripts of the \emph{Haṭhapradīpikā} (\getsiglum{J2,M1}) and is close to the \textalpha\ reading \emph{ekahṛdaye}.
+In the second verse quarter, the adopted reading \emph{ekahṛdayo} is supported by two manuscripts of the \emph{Haṭhapradīpikā} (\getsiglum{J2,M1}) and is close to the \alphaOne\ reading \emph{ekahṛdaye}.
 It is also attested by the six-chapter \emph{Viveka\-mārtaṇḍa} and the \emph{Yoga\-cintāmaṇi}, which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}.
@@
 There are many variations of this verse quarter in the \emph{Haṭhapradīpikā} manuscripts, as well as in the manuscripts of the sources and testimonia.
-Most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}, which is close to the adopted and \textalpha\ reading.
+Most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}, which is close to the adopted.
 Other readings allude here to the practice of the Jālandhara lock, in which the chin is placed on the chest.
@@
 \begin{translation}[hp01_039]
-From among the eighty-four postures, one should regularly practise just \emph{siddhāsana}, in the same way from among the 72,000 channels [one should practise, focusing on] Suṣumṇā.
-\end{translation}
+From among the eighty-four postures, it is \emph{siddhāsana} that one should always prefer; in the same way that from among the 72,000 channels [one should prefer] Suṣumṇā.
+\end{translation}
+
+% MD 2024-8: Perhaps "siddham eva sadā bhajet" (ε-mss,IFP) is better than "samabhyaset". α2 sadā bhavet and α3 sadā paṭhet support the former.
 
@@
 \begin{philcomm}[hp01_039]
-It is odd to have \emph{suṣumṇām} as the object of the verb \emph{abhyaset}.
-This reading is well attested by manuscripts of the \emph{Haṭhapradīpikā} and is also found in the \emph{Yogacintāmaṇi}, which attributes it to the \emph{Yogasāramañjarī}.
-Perhaps, the second line was added somewhat haphazardly by Svātmārāma, and then others have tried to make sense of it by changing \emph{suṣumnām iva nāḍiṣu} to \emph{nāḍīnāṃ malaśodhanam}, which occurs in the \emph{Jyotsnā} (1.39).
+%Most witnesses have \emph{suṣumṇām} as the object of the verb \emph{abhyaset}, which is odd. %  
+% MD: we have finally adopated the reading "bhajet".
+%This reading is well attested by manuscripts of the \emph{Haṭhapradīpikā} and is also found in the \emph{Yogacintāmaṇi}, which attributes it to the \emph{Yogasāramañjarī}.
+Perhaps, the second line was added somewhat haphazardly by Svātmārāma, and then
+%
+Others have tried to make sense of this by changing \emph{suṣumnām iva nāḍiṣu} to \emph{nāḍīnāṃ malaśodhanam}, which occurs in the \emph{Jyotsnā} (1.39).
 The reading \emph{nāḍīnāṃ malaśodhanam/e} is probably a patch as no other texts say that \emph{siddhāsana} clears the channels.
-However, the idea of purifying the channels can be found in other contexts (e.g., \emph{Kumbhakapaddhati} 120) and may hark back to an earlier notion of flushing (\emph{cālana}) the channels (e.g., \emph{Amṛtasiddhi} 11.6).% somewhat haphazardly?
+It could have been borrowed from verse 3.103*1ab, which has the same hemistich and was added to the section on \textit{śakticālana}.
+%However, the idea of purifying the channels can be found in other contexts (e.g., \emph{Kumbhakapaddhati} 120) and may hark back to an earlier notion of flushing (\emph{cālana}) the channels (e.g., \emph{Amṛtasiddhi} 11.6).% somewhat haphazardly?
 I don't %understand the second part of the note: purifying the channels is found all over the place.
 %JB: The point is: no other texts say that \emph{siddhāsana} clears the channels. 
+% MD: the same half verse with nāḍīnāṃ malaśodhanam/e appears as 3.103*1ab. Better to mention?
 
@@
 \begin{philcomm}[hp01_043]
-The reading \emph{na kumbhasadṛśo 'nilaḥ} (`no breath like a retention') is the lectio difficilior and attested by two early witnesses (\etaOne\ and \etaTwo) and is possibly original.
+%The reading \emph{na kumbhasadṛśo 'nilaḥ} (`no breath like a retention') is the lectio difficilior and attested by two early witnesses (\etaOne\ and \etaTwo) and is possibly original.
 However, the \textalpha\ manuscripts and several other important witness groups have the adopted reading \emph{kumbhaḥ kevalopamaḥ}, as well as the \emph{Yogacintāmaṇi} and some manuscripts of the \emph{Haṭharatnāvalī}, suggesting that this reading, which makes much better sense, was in the transmission at an early stage.
 % Śivasaṃhitā has \emph{kumbhasadṛśaṃ balam}.
-%
+% MD: I think kevalakumbha is indispensable after the comparison in 1.41-42.
+It seems more likely to me that this is the original text of Svātmārāma and that the eta group was influenced by the source text.
 \end{philcomm}
@@
 
-%(MD 2024-5-14): It is remarkable that α2 and β1 have osthāpyot and ε3 otthāpya. Perhaps o = ava as in the Middle Indic? avasthāpya would make the perfect sense.
+%(MD 2024-5-14): It is remarkable that α2 and β1 have osthāpyot and ζ3 otthāpya. Perhaps o = ava as in the Middle Indic? avasthāpya would make the perfect sense.
 
@@
 Either Svātmārāma chose to leave the verse hanging (the following verse in the \emph{Dattātreyayogaśāstra} adds nothing about the form of the posture, which is the topic here) or the verse that completes the syntax fell out, perhaps because of a scribal error that happened early in the transmission.
-In the \emph{Dattātreyayogaśāstra} verses 36 and 37 both end with \emph{pavanaṃ śanaiḥ}, the repetition of which may have caused an eyeskip.
-
-The manuscript readings with \emph{vakṣa sthāpayet} (\getsiglum{J7,V3,J8,J10,J17,N17}) or something similar (\getsiglum{V1,W4}) do not offer a solution to the incomplete syntax and do not indicate that Svātmārāma rewrote \emph{Dattātreyayogaśāstra} 36 so that he could omit \emph{Dattātreyayogaśāstra} 37.
-(The readings \emph{osthāpyot} (α2 and β1) and \emph{otthāpya} (ε3) are surprising and are perhaps Middle Indic forms in which Sanskrit \emph{ava-} becomes \emph{o-}.) In the absence of evidence that Svātmārāma included \emph{Dattātreyayogaśāstra} 37 or wrote a coherent version of \emph{Haṭhapradīpikā} 1.46, we have made sense of \emph{pavanaṃ śanaiḥ} by adding “[draw in]” in our translation.
+In the \emph{Dattātreyayogaśāstra} verses 36 and 37 both end with \emph{pavanaṃ śanaiḥ}, the repetition of which may have caused an eyeskip.\lb
+
+The manuscript readings with \emph{vakṣa sthāpayet} (\betaOmega\ and \etaTwo) or something similar (\etaOne) do not offer a solution to the incomplete syntax and do not indicate that Svātmārāma rewrote \emph{Dattātreyayogaśāstra} 36 so that he could omit \emph{Dattātreyayogaśāstra} 37.
+(The readings \emph{osthāpyot} (\alphaTwo\ and \betaOne) and \emph{otthāpya} (\zetaThree) are surprising and are perhaps Middle Indic forms in which Sanskrit \emph{ava-} becomes \emph{o-}.) In the absence of evidence that Svātmārāma included \emph{Dattātreyayogaśāstra} 37 or wrote a coherent version of \emph{Haṭhapradīpikā} 1.46, we have made sense of \emph{pavanaṃ śanaiḥ} by adding “[draw in]” in our translation.\lb
 
@@
 \end{quote}
+
 Brahmānanda appears to have had in mind a probably older rule for meditation postures, according to which the tongue rests near the front teeth.
@@
 Placing the tongue where it does not disturb the meditation seems quite appropriate for a `normal' meditative practice.\footnote{The rule of placing the tongue at the palate is also found in \emph{Īśānaśivagurudevapaddhati} 18.120: \emph{tāluke jihvāṃ saṃyojya kiñcidvivṛtavaktro dantair dantān asaṃspṛśan ṛjukāyaḥ}.
-For similar references in tantric and other works see Mallinson 2007:17–24}
+For similar references in tantric and other works see Mallinson 2007:17–24}\lb
 
 When the context is haṭhayogic physiology, placing the tongue at the uvula, which is the source of `nectar', is more appropriate.
-Confusingly, yogic terminology includes many names for the uvula, and among these especially the term \emph{rājadanta} may give rise to confusion, since, as we have seen, the tongue might also in some yoga systems be placed at the front teeth.
+Confusingly, yogic terminology includes many names for the uvula, and among these especially the term \emph{rājadanta} may give rise to confusion, since, as we have seen, the tongue might also in some yoga systems be placed at the front teeth.\lb
 
@@
 Thus the tenth-century \emph{Mokṣopāya} (V.55.14c) says that the tongue rests at the `source of the palate' (\emph{tālumūlatalālagnajihvā}) and the commentary, the
-\emph{Saṃsārataraṇi}, on the parallel passage in \emph{Laghuyogavāsiṣṭha} V.6.155, which reads \emph{tālumūlāntarālagnajihvā°}, explains that this means that the tongue is to be placed in the middle of the two regions of the palate, and that this is the \emph{nabhomudrā}, alias `\emph{khecarī}' (\emph{tālumūlāntarālagnajihvamūlaḥ tālumūlayoḥ kākudamūladeśayoḥ āntare lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā} | \emph{yā hi khecarı̄ty ucyate}).
+\emph{Saṃsārataraṇi}, on the parallel passage in \emph{Laghuyogavāsiṣṭha} V.6.155, which reads \emph{tālumūlāntarālagnajihvā°}, explains that this means that the tongue is to be placed in the middle of the two regions of the palate, and that this is the \emph{nabhomudrā}, alias `\emph{khecarī}' (\emph{tālumūlāntarālagnajihvamūlaḥ tālumūlayoḥ kākudamūladeśayoḥ āntare lagnam ālagnam jihvāmūlam yasyety anena nabhomudrā darśitā} | \emph{yā hi khecarı̄ty ucyate}).\lb
 
@@
 \begin{philcomm}[hp01_047p]
-The comment added to this verse by Svātmārāma indicates that he prefers the following version of \emph{padmāsana}, which derives from the \emph{Vivekamārtaṇḍa}, rather than the one he has borrowed from the \emph{Dattātreyayogāśāstra}.
+The comment added to this verse by Svātmārāma indicates that he prefers the second version of \emph{padmāsana}, which derives from the \emph{Dattātreyayogaśāstra}, rather than the one he has borrowed from the \emph{Vivekamārtaṇḍa}.
 % Maybe add some remarks by commentators on how they understand it.
@@
 In the source text, the \emph{Vivekamārtaṇḍa, tat} appears to refer to the \emph{mokṣadvāra} broken by \emph{kuṇḍalinī}, which is mentioned in the previous verse.
+\lb
 
@@
 \begin{translation}[hp01_049]
-The yogi in \emph{padmāsana} who fills [himself] up through the openings of the channels and holds the breath is sure to be liberated.
+The yogi in \emph{padmāsana} who fills [himself] up through the openings of the channels and stabilises the breath is sure to be liberated.
 \end{translation}
@@
 \begin{philcomm}[hp01_049]
-It is not unusual to read \emph{pūrayan} with the locative as seen in this verse (cf. \emph{pūrayen mukhe} in \emph{Amaraugha} 21d).
+It is not unusual to read \emph{pūrayet} with the locative as seen in this verse (cf. \emph{pūrayen mukhe} in \emph{Amaraugha} 21d).\lb
 
@@
 The passive verbs meaning to take in the breath (e.g., \emph{pīyate}) may have been adopted to remove the reference to holding the breath because a breath retention is not mentioned in the previous verse describing \emph{padmāsana} (only inhalation and exhalation).
-
-In the third verse quarter, \alphaOne\ reads \emph{māruto mriyate yas tu}, which does not make sense, but if one accepts \emph{pūrayet} in the second verse quarter, one could emend \textalpha's reading to \emph{māruto mriyate yasya}, which makes good sense (i.e., `the yogi whose breath dies is undoubtedly liberated').
-In the same vein, \alphaTwo\ also has the plausible reading \emph{mārutaṃ mārayet yas tu}.
+\lb
+
+In the third verse quarter, \alphaOne\ and \alphaThree\ read \emph{mārutaṃ mriyate yas tu}, which does not make sense, but if one accepts \emph{pūrayet} in the second verse quarter, one could emend their reading to \emph{māruto mriyate yasya}, which makes good sense: `the yogi whose breath dies (i.e.
+becomes still) is undoubtedly liberated'.
+In the same vein, \alphaTwo\ also has the plausible reading \emph{mārutaṃ mārayed yas tu}.
+\lb
 %but J5 is contaminated (so the Gr3 reading dhārayet is more reliable).
-
-%(MD: Or read \emph{pūrayan mārutaṃ dhriyate yas tu} ``he who remains by inhaling..."?
-%This \emph{dhriyate} is rather intransitive than passive. 
-%Cf. PW \emph{dhṛ} 22a ``fest sein, sich ruhig verhalten, stillhalten, verbleiben, sich erhalten, bestehen")
+% some  mss from the full collation do have yasya ??: JM rewrite note, we are adopting pūrayet | māruto mriyate yasya... also check tr.
+%MD 2024-8: the reading of α1 is "pūrayet | mārutaṃ (not māruto) mriyate yas tu".
 
@@
 Spreading the fingers and keeping the mouth wide open mimic a lion, and this is depicted in some iconography of Yoganarasiṃha (for example, Yoga Narasimha, Vishnu's Man-Lion Incarnation, Samuel Eilenberg Collection, Bequest of Samuel Eilenberg, 1998, Accession Number: 2000.284.4.
-https://www.metmuseum.org/art/collection/search/39251).
+https://www.metmuseum.org/art/collection/search/39251).\lb
 
@@
 \begin{philcomm}[hp01_053]
-We have understood \emph{pārśvapāda} as a \emph{ekadeśitatpuruṣa} meaning the side of the foot, like  \emph{agrapāda}, the toes.
+We have understood \emph{pārśvapāda} as a \emph{ekadeśitatpuruṣa} meaning the side of the foot, like  \emph{agrapāda}, the toes.\lb
 
@@
 \begin{philcomm}[hp01_055_1]
-1.55*1–2 are omitted from the \textalpha, \textgamma, \textdelta\ and \textepsilon\ groups, so it is likely these verses were not in the earliest versions of the \emph{Haṭhapradīpikā}.
+1.55*1–2 are omitted from the \textalpha, \textgamma, \textdelta\ and \textzeta\ groups, so it is likely these verses were not in the earliest versions of the \emph{Haṭhapradīpikā}.
 In fact, it appears that both were added (perhaps initially as marginal notes) to elaborate on the word \emph{kriyā} in 1.55d.
@@
 The reader might wonder how this illusionist verse could be understood to fit our Yoga text.
-We can only speculate that perhaps the scribe of the hyparchetype of the \textdelta\ manuscripts was fond of it.
+We can only speculate that perhaps the scribe of the hyparchetype of the \textdelta\ manuscripts was fond of it.\lb
 
@@
 \tl{haṭhapradīpikāyām--\\+}
-\tl{śrutipratītiś ca gurupratītiḥ svātmapratītiś ca manonirodhaḥ |\\+}
+\tl{śrutipratītiḥ svagurupratītiḥ svātmapratītir manaso nirodhaḥ |\\+}
 \tl{etāni sarvāṇi samuccitāni matāni dhīrair iha sādhanāni ||\\!}
@@
 \begin{philcomm}[hp01_055_4]
-Verse 1.55.4 is in some of the \textdelta\ manuscripts and is quoted in \emph{Yogacintāmaṇi} with attribution to the \emph{Haṭhapradīpikā}.
-
-The reading in the \emph{Yogacintāmaṇi} `cessation of mind' (\emph{manonirodhaḥ}) is better than \emph{manaso 'pi bodhaḥ} (the \textdelta\ reading) in a yogic context.
+Verse 1.55.4 is in some of the \textdelta\ manuscripts and is quoted in \emph{Yogacintāmaṇi} with attribution to the \emph{Haṭhapradīpikā}.\lb
+
+The reading in the \emph{Yogacintāmaṇi} `cessation of mind' (\emph{manaso nirodhaḥ}) is better than \emph{manaso 'pi bodhaḥ} (the \textdelta\ reading) in a yogic context.
 \end{philcomm}
@@
 The omission is probably deliberate as that manuscript does not have chapter four, which teaches \emph{nādānusandhāna}.
-The numbering in \etaOne\ suggests that its exemplar had this verse.
-
-The term \emph{kumbhaka} is almost always masculine but appears in this verse as a neuter in the majority of manuscripts of the important groups.
+The numbering in \etaOne\ suggests that its exemplar had this verse.\lb
+
+%The term \emph{kumbhaka} is almost always masculine [ neuter also in 2.53, 2.68 and the heading before 2.72 (MD)] but appears in this verse as a neuter in the majority of manuscripts of the important groups.\lb
 
@@
 \begin{translation}[hp01_058]
-When very unctuous and sweet food is eaten for love of Śiva, leaving a quarter [of the stomach] empty (\emph{caturthāṃśavivarjitaḥ}), it is called a restricted diet (\emph{mitāhāra}).
+When very unctuous and sweet food is eaten for love of Śiva, leaving a quarter [of the stomach] empty (\emph{caturthāṃśavivarjitaḥ}), that is called a restricted diet (\emph{mitāhāra}).
 \end{translation}
@@
 It is also found, but reworked to be about the \emph{mitāhārī}, in Nowotny’s \emph{Gorakṣaśataka} (55), which is an extended recension of the \emph{Vivekamārtaṇḍa}.
-
-The expression `lacking a fourth part' \emph{caturthāṃśavivarjitaḥ}) is somewhat vague but probably refers to the idea of leaving a quarter of one's stomach empty, which is stated more clearly in the \emph{Dharmaputrikā} 1.51-52:
+\lb
+
+The expression `lacking a fourth part' \emph{caturthāṃśavivarjitaḥ}) is somewhat vague but probably refers to the idea of leaving a quarter of one's stomach empty, which is stated more clearly in the \emph{Dharmaputrikā} 1.51–52:
 
@@
 \begin{translation}[hp01_059]
- Pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink, flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like: they say that such [food] is unwholesome.
+ Pungent, sour, bitter, salty and hot foods, horseradish, sour gruel, [sesame] oil, sesame and mustard seeds, fish and intoxicating drink, flesh of goats and so forth, curds, diluted buttermilk, poor man's pulse, jujube fruit, the leftover paste of oily seeds, asafoetida, garlic and the like: they say that [such food] is unwholesome.
 \end{translation}
@@
 
-Brahmānanda’s understanding of \emph{harītaśāka} as \emph{pattraśāka} is probably wrong if \emph{pattraśāka} was intended as ‘leafy vegetables.’ But he may have used the term \emph{pattraśāka} in the sense of horseradish (\emph{śigru}) as the dictionary notes that \emph{pattraśāka} is probably equivalent to \emph{śākapattra}, which is mentioned in \emph{Rājanighaṇṭu} 7.26 (above).
+Brahmānanda’s understanding of \emph{harītaśāka} as \emph{pattraśāka} is probably wrong if \emph{pattraśāka} was intended as ‘leafy vegetables.’ But he may have used the term \emph{pattraśāka} in the sense of horseradish (\emph{śigru}) as the dictionary notes that \emph{pattraśāka} is probably equivalent to \emph{śākapattra}, which is mentioned in \emph{Rājanighaṇṭu} 7.26 (above).\lb
 
@@
 He identifies it with the Rohri/Khairpur region of Sind.
+\lb
 %( S.M. Ali, Geography of the Purānas, Delhi, 1966, p. 144).
@@
 The word \emph{taila} could refer to \emph{tilataila}. This is supported by the following rule (\emph{paribhāṣā}) in the \emph{Śārṅgadharasaṃhitā} (48):
-\emph{anuktāvasthāyāṃ paribhāṣāvidhiḥ} [...] \emph{taile ’nukte tilodbhavam}. We thank Dominik Wujastyk for this reference.
+\emph{anuktāvasthāyāṃ paribhāṣāvidhiḥ} [...] \emph{taile ’nukte tilodbhavam}. We thank Dominik Wujastyk for this reference.  \lb
 
@@
 Moreover, only \emph{ājādimāṃsa}° makes good sense.
-Diwakar Acharya believes that the prohibition of goat flesh and fish in this verse suggests it derives from the North East of India.
-
-The term \emph{kulattha} means a kind of pulse, translated by Dominik Wujastyk (1998: 77) as `poor man's pulse.'
+Diwakar Acharya believes that the prohibition of goat flesh and fish in this verse suggests it derives from the North East of India.\lb
+
+The term \emph{kulattha} means a kind of pulse, translated by Dominik Wujastyk (1998: 77) as `poor man's pulse.'\lb
 % Dominik Wujastyk, Roots of Ayurveda, Penguin Books, 1998.
@@
 There are references to \emph{kola} being pungent, though this does not seem to indicate sufficiently why \emph{kola} is mentioned separately in the \emph{Haṭhapradīpikā} as an unwholesome food.
-Diwakar Acharya has informed us that \emph{kola} can refer to a type of banana in some parts of India.
-
-According to Sharma (1982: 69), \emph{piṇyāka} is, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’ Diwakar says it is an oil cake that has a strong flavour, which may account for its inclusion in this list of unwholesome foods.
+Diwakar Acharya has informed us that \emph{kola} can refer to a type of banana in some parts of India.\lb
+
+According to Sharma (1982: 69), \emph{piṇyāka} is, ‘The remnant paste of oily seeds after pressing out the oil content is called \emph{piṇyāka}.’ Diwakar says it is an oil cake that has a strong flavour, which may account for its inclusion in this list of unwholesome foods.\lb
 % P.V. Sharma, Ḍalhaṇa and his commentary on drugs, 1982.
@@
 \begin{philcomm}[hp01_060]
-We have not found any conclusive evidence for the meaning of \emph{tilapiṇḍa}. Brahmānanda glosses it as \emph{piṇyāka} (on the meaning of which see the notes for the previous verse).
+We have not found any conclusive evidence for the meaning of \emph{tilapiṇḍa}. Brahmānanda glosses it as \emph{piṇyāka} (on the meaning of which see the notes for the previous verse).\lb
 
 The meaning of the compound \emph{kadaśanaśākotkaṭaṃ} is not clear.
-Brahmānanda understands it as a list (\emph{dvandva}) consisting of \emph{kadaśana}, \emph{śāka} and \emph{utkaṭa}, which he defines as bad food, prohibited vegetables and pepper, respectively.
+Brahmānanda understands it as a list (\emph{dvandva}) consisting of \emph{kadaśana}, \emph{śāka} and \emph{utkaṭa}, which he defines as bad food, prohibited vegetables and pepper, respectively.\lb
 
@@
 According to some Nighaṇṭus, the word \emph{utkaṭā} can mean pepper  (e.g., \emph{Rājanighaṇṭu} 5.16 \emph{pārvatī śailajā tāmrā lambabījā tathotkaṭā}) and, according to Monier Williams, \emph{utkaṭa} can refer to Saccharum Sara and \emph{utkaṭā} also to Laurus Cassia (cinnamon).
+\lb
 
@@
 \begin{philcomm}[hp01_061]
-Manuscripts from the \textalpha, \textbeta\ and \textepsilon\ groups have the reading \emph{durjanaprāntaṃ} (1.61a), which is the lectio difficilior in relation to \emph{durjanaprītiṃ} (`the friendship of wicked people').
+Manuscripts \alphaOne\alphaThree\betaOne\zetaOne\ have the reading \emph{durjanaprāntaṃ} (1.61a), which is the lectio difficilior in relation to \emph{durjanaprītiṃ} (`the friendship of wicked people').
 We have understood \emph{durjanaprānta} in line with Brahmānanda's gloss in \emph{Jyotsnā} 1.64, `dwelling near bad people' (\emph{durjanasamīpavāsa}).
@@
 
-The word \emph{sitā} is one of many words for ground sugar. Meulenbeld (1974: 507) comments that \emph{sitā} is `very white and looks like gravel.'
+The word \emph{sitā} is one of many words for ground sugar. Meulenbeld (1974: 507) comments that \emph{sitā} is `very white and looks like gravel.'\lb
 
@@
 Meulenbeld (1974: 569) compiled a list of six possibilities, including \textsc{Trichosanthes dioica} \textsc{Roxb}.
-(`pointed gourd'), \textsc{Trichosanthes cucumerina} \textsc{Linn} (`snake gourd').
+(`pointed gourd'), \textsc{Trichosanthes cucumerina} \textsc{Linn} (`snake gourd').\lb
 
@@
 Sharma (1982: 156) adds that \emph{paṭola} is a synonym of \emph{kulaka} and is well known as \textsc{Trichosanthas dioica Roxb.} Brahmānanda glosses \emph{paṭola} as \emph{kośātakī}, which is \textsc{Luffa acutangula Roxb} (Meulenbeld 1974: 586), suggesting that he thought it was some sort of luffa.
-He also mentions the vernacular term \emph{paravara} for \emph{paṭola}.
+He also mentions the vernacular term \emph{paravara} for \emph{paṭola}.\lb
 
@@
 \begin{translation}[hp01_063]
-The yogi should eat food that is sweet, delicious, unctuous, contains cow products, nourishes the bodily constituents (\emph{dhātu}), is desired by the mind and is appropriate.
+The yogi should eat food that is sweet, delicious, unctuous, contains cow products, nourishes the bodily constituents, is desired by the mind and is appropriate.
 \end{translation}
@@
 The variants of 1.63a all seem possible: \emph{mṛṣṭaṃ}, \emph{miṣṭaṃ} and \emph{iṣṭaṃ}.
-The last is made somewhat redundant by \emph{mano 'bhilaṣitaṃ} in 163c.
-Both \emph{mṛṣṭaṃ} (\textalpha\ and \texteta) and \emph{miṣṭaṃ} (\textbeta\ and \textgamma) are well attested by manuscripts of important groups and there is hardly any difference in their meaning in this context.
-We have adopted \emph{mṛṣṭaṃ} as it is supported by the \textalpha\ group.
+The last is made somewhat redundant by \emph{mano'bhilaṣitaṃ} in 16.3c.
+Both \emph{mṛṣṭaṃ} and \emph{miṣṭaṃ} are well attested by manuscripts of important groups and there is hardly any difference in their meaning in this context.
+We have adopted \emph{mṛṣṭaṃ} as it is supported by \alphaOne\ and \alphaThree.
 \end{philcomm}
@@
 
-\begin{philcomm}[hp01_064]
-\etaOne, the oldest dated manuscript, has a different reading for the last verse quarter (164d): \emph{sarvaṃ yogī yatendriyaḥ} (`the yogi whose senses are restrained wholly succeeds [...]').
+%\begin{philcomm}[hp01_064]
+%\etaOne, the oldest dated manuscript, has a different reading for the last verse quarter (164d): \emph{sarvaṃ yogī yatendriyaḥ} (`the yogi whose senses are restrained wholly succeeds [...]').
 Here, \emph{sarvaṃ} is not easy to construe, and the readings of the \textalpha\ manuscripts and other important groups of \emph{Haṭhapradīpikā} manuscripts indicate that \emph{sarvayogeṣv atandritaḥ} was the reading adopted by Svātmārāma, which is more similar to the \emph{Dattātreyayogaśāstra}'s (i.e., \emph{yoge sarvo 'py atandritaḥ}).
-\end{philcomm}
+% MD: This commentary may be not necessary.
+%%\end{philcomm}
 

### 2025-04-04 latex/HP2_comm.tex
new files

@@
 \tl{athāsane dṛḍhe yogī vaśī hitamitāśanaḥ |\\+}
-\tl{gurūpadiṣṭamārgeṇa prānāyāmān samabhyaset ||\\!}
+\tl{gurūpadiṣṭamārgeṇa prāṇāyāmān samabhyaset ||\\!}
 \end{versinnote}
@@
 \tl{athāsane dṛḍhībhūte vaśī hitamitāśanaḥ |\\+}
-\tl{gurūpadiṣṭamārgeṇa prānāyāmān samabhyaset ||\\!}
+\tl{gurūpadiṣṭamārgeṇa prāṇāyāmān samabhyaset ||\\!}
 \end{versinnote}
@@
 %\emph{dṛḍho yogī} seems unlikely as \emph{dṛḍha} usually qualifies a technique, the body, etc. (not worth commenting on as alpha has dṛḍhe).
-The \emph{Jyotsnā} (2.1) has the plural \emph{prāṇāyāmān}, which is supported by some manuscripts of the \textbeta, \textgamma\ and \textepsilon\ groups.
+The \emph{Jyotsnā} (2.1) has the plural \emph{prāṇāyāmān}, which is supported by some manuscripts of the \textbeta, \textgamma\ and \textzeta\ groups.
 The plural is possible here as it could refer to the different techniques of retention (\emph{kumbhaka}) taught later in the chapter.
 This is how Brahmānanda understands it (\emph{prāṇāyāmān vakṣyamāṇān}).
-The variation between singular and plural recurs through this chapter, and we have followed the \textalpha\ readings, which make good sense.
+The variation between singular and plural recurs through this chapter, and we have followed the \alphaOne\ readings, which make good sense.
 In this case, since the verse is introducing the topic of breath control, the more general sense of the singular is appropriate.
@@
 \begin{translation}[hp02_002]
-When the wind moves, everything moves and, when it is still, everything is still, so the yogi attains motionlessness through restraining the breath.
+When the wind is moving, everything moves [and] when it is still, everything is firmly fixed, so the yogi attains motionlessness through restraining the breath.
 \end{translation}
@@
 The reading \emph{dṛḍhabandhanaṃ} appears to be a dittographical error and does not make sense when read with either \emph{sarvaṃ} (\textalpha) or \emph{cittaṃ} (\textgamma).
-
-As noted by Brahmānanda, in addition to its primary meaning of immobility, \emph{sthāṇutvam} can also mean the state of being Śiva, for whom Sthāṇu is another name.
+\lb
+
+%MD 2024-8: The most plausible reconstruction of the text, based on our manuscripts, is "cale vāte calet sarvaṃ niścale dṛḍhabandhanam" (When the wind moves, everything would move.
+When [the wind] is still, [everything] would have firm bond.) Would this work?
+Yes, with calaṃ ??
+JM to rewrite note.
+
+As noted by Brahmānanda, in addition to its primary meaning of immobility, \emph{sthāṇutvam} can also mean the state of being Śiva, for whom Sthāṇu is another name. \lb
 
@@
 \tl{yāvad vāyuḥ sthito dehe tāvaj jīvitam ucyate |\\+}
-\tl{maraṇaṃ tasya niḥkrāntau tato vāyuṃ nirodhayet ||\\+}
+\tl{maraṇaṃ tasya niṣkrāntau tato vāyuṃ nirodhayet ||\\+}
 \tl{\var{72a sthito ] sthiro A 72b jīvitam ] VAGBGPT; jīvanam GLGPk 72c niṣkrāntau ] VTA; niḥkrāṃtaṃ GBGP, niṣkrāntis GLGPk 72d vāyuṃ nirodhayet ] T; vāyunibandhanāt V, vāyuṃ nibandhayet ATvl, vāyuṃ nirundhayet G}\\!}
@@
 \tl{vyāpārād yasya ceṣṭante śārīrāḥ pañca vāyavaḥ ||\\+}
-\tl{prāṇāpānād ayas te tu bhinnā vṛtter na vastutaḥ |\\+}
+\tl{prāṇāpānādayas te tu bhinnā vṛtter na vastutaḥ |\\+}
 \tl{vṛttiṃ leśān nigadato bharadvāja nibodha me ||\\+}
-\tl{vṛttiḥ praṇayanaṃ nāma yat taj jīvanam ucyate |\\!}
+\tl{vṛttiḥ prāṇāyānaṃ nāma yat taj jīvanam ucyate |\\!}
 \end{versinnote}
@@
 \begin{testimonia}[hp02_003]
-
 \emph{Haṭharatnāvalī} 3.80
@@
 
-We believe that \emph{nirundhayet} was probably the original reading and that it was changed to the more correct form of \emph{nirodhayet} and, in some cases, confused with the final verse quarter of the previous verse and changed to \emph{nibandhanāt} (\textalpha) or \emph{nibandhayet} (\textbeta).%JM: necessary note?
+%We believe that \emph{nirundhayet} was probably the original reading and that it was changed to the more correct form of \emph{nirodhayet} and, in some cases, confused with the final verse quarter of the previous verse and changed to \emph{nibandhanāt} (\alphaOne) or \emph{nibandhayet} (\alphaThree\textbeta).%JM: necessary note?
 %JB nirundhayet is not supported by alpha.
+%MD: Why is the reading vāyuṃ nibandhayet not possible, when we have vāyunibandhanāt in 2.2d?
+See also samīraṇam alaṃ baddhvā in 2.10b.
+The development of the variant readings "nibandhayet -> nirundhayet -> nirodhayet" is much more plausible.
 % JB I've commmented on jīvitaṃ/jīvanam because it looks like its unsupported by alpha (alpha one has jītaviṃ).
@@
 %\begin{philcomm}[hp02_004]
-%The reading \emph{kāyasiddhi} in the final verse quarter is supported by manuscripts of the \textalpha, \textbeta, \textepsilon\ and \texteta\ groups whereas other manuscripts have \emph{kāryasiddhi}.
+%The reading \emph{kāyasiddhi} in the final verse quarter is supported by manuscripts of the \textalpha, \textbeta, \textzeta\ and \texteta\ groups whereas other manuscripts have \emph{kāryasiddhi}.
 In the context of physical yoga, \emph{kāyasiddhi} makes better sense, as \emph{kāryasiddhi} can refer more generally to accomplishing anything.
@@
 
-\begin{philcomm}[hp02_005]
+%\begin{philcomm}[hp02_005]
 %We have understood the term \emph{nāḍīcakra} as the network of channels that originate from the bulb in the lower abdomen.
@@
 
-\end{philcomm}
+%\end{philcomm}
 
@@
 
-\begin{philcomm}[hp02_006]
+%\begin{philcomm}[hp02_006]
 %There is a problem in the third verse quarter.
@@
 %The \emph{ca} at the end of the fourth verse quarter would suggest that two statements are being made in the second half of the verse.
-The readings of some manuscripts, such as \emph{suṣumṇā susnigdhā} in \epsilonThree\ and \etaOne, appear to be attempts to make sense of the \emph{ca}.
+The readings of some manuscripts, such as \emph{suṣumṇā susnigdhā} in \zetaThree\ and \etaOne, appear to be attempts to make sense of the \emph{ca}.
 However, the meaning of \emph{suṣumṇā susnigdhā} (`\emph{suṣumṇā} becomes well-lubricated') seems implausible.
@@
 %JM: I think we should adopt alpha1 (and others): suṣumṇā susvasthā. It works, and makes sense of ca. But we'd not be reading alpha1's yāsthā, which looks to be a mistake.
-\end{philcomm}
+%\end{philcomm}
 
@@
 %\begin{philcomm}[hp02_007]
-%The hemistich beginning with \emph{dhārayitvā} appears to have dropped out of the transmission of the \emph{Haṭharatnāvalī}, thus rendering this passage unintelligible in that work (see Gharote et al.
+%The line beginning with \emph{dhārayitvā} appears to have dropped out of the transmission of the \emph{Haṭharatnāvalī}, thus rendering this passage unintelligible in that work (see Gharote et al.
 2002: 125).%JM: necessary?
@@
 
-\emph{Yukabhavadeva} 7.14
+\emph{Yuktabhavadeva} 7.14
 
@@
 \begin{sources}[hp02_009]
-Cf. \emph{Dattātreyayogaśāstra} 61
+Cf.~\emph{Dattātreyayogaśāstra} 61
 
@@
 One might try to construe \emph{anirodhataḥ} as ‘without stopping the breath,’ but verse 2.7 clearly states that the breath should be held as long as possible (\emph{yathāśakti}).
-According to the apparatus of the critical edition of the \emph{Haṭharatnāvalī}, \emph{avirodhataḥ} is well-attested for the parallel hemistich.
+According to the apparatus of the critical edition of the \emph{Haṭharatnāvalī}, \emph{avirodhataḥ} is well-attested for the parallel line.
 Furthermore, \emph{avirodhata} is attested, as well as \emph{virodhahīna}, in a passage of the \emph{Haṭhatattvakaumudī} (36.6–9) that appears to have been loosely based on \emph{Haṭhapradīpikā} 2.7–9:
@@
 
-\begin{philcomm}[hp02_010]
+%\begin{philcomm}[hp02_010]
 %We have translated \emph{yamināṃ} as `yogis' because in this context it could refer to those practising the observances (\emph{yama}) of yoga rather than ascetics more generally.
@@
 %The reading \emph{vidhinābhyāsaṃ} in the third pāda of various witnesses is unmetrical because it lacks the caesura. (not well attested in the new collation). 
-\end{philcomm}
+%\end{philcomm}
 
@@
 \begin{philcomm}[hp02_012]
-The manuscript readings diverge greatly in the second hemistich.
+The manuscript readings diverge greatly in the second line.
 In the third verse quarter, all of the manuscripts have \emph{prāṇa} in some form (instead of \emph{deha} in the \emph{Vivekamārtaṇḍa}).
-\textalpha\ and some of the \textbeta, \textepsilon\ and \texteta\ manuscripts seem to be stating that it is \emph{padmāsana} that rises up again and again in the highest stage of holding the breath (\emph{uttiṣṭhaty uttame prāṇarodhe padmāsanaṃ muhuḥ}).
-We have adopted a similar reading but with \emph{padmāsane} (\betaTwo, \deltaOne\ and \deltaThree, and \etaTwo) because it makes better sense that the yogi rises up while seated in lotus pose.
-Another version is seen in \etaTwo\ (and others), which appears to say that the breaths rise up again and again when one is seated in the lotus pose (\emph{uttiṣṭhanty uttame prāṇā baddhe padmāsane muhuḥ}).
+\textalpha\ and some of the \textbeta, \textzeta\ and \texteta\ manuscripts seem to be stating that it is \emph{padmāsana} that rises up again and again in the highest stage of holding the breath (\emph{uttiṣṭhaty uttame prāṇarodhe padmāsanaṃ muhuḥ}).
+We have adopted a similar reading but with \emph{padmāsane} (\betaTwo, \deltaOne, and \etaTwo) because it makes better sense that the yogi rises up while seated in lotus pose.
+Another version is seen in \etaTwo, which appears to say that the breaths rise up again and again when one is seated in the lotus pose (\emph{uttiṣṭhanty uttame prāṇā baddhe padmāsane muhuḥ}).
 However, the verse is about the external signs that might arise in \emph{prāṇāyāma} rather than internal processes.
@@
 
-\begin{philcomm}[hp02_014]
-\end{philcomm}
+% \begin{philcomm}[hp02_014]
+% \end{philcomm}
 
@@
 \begin{versinnote}
-\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śānaiḥ |\\+}
+\tl{yathā siṃho gajo vyāghro bhaved vaśyaḥ śanaiḥ śanaiḥ |\\+}
 \tl{anyathā hanti yantāraṃ tathā vāyur asevitaḥ ||\\!}
@@
 \begin{philcomm}[hp02_015]
-The second hemistich of this verse has been rewritten to express the same idea (and simile) as that found in \emph{Viveka\-mārtaṇḍa} 101, but the author of the \emph{Viveka\-mārtaṇḍa} expresses it more clearly.
+The second line of this verse has been rewritten to express the same idea (and simile) as that found in \emph{Viveka\-mārtaṇḍa} 101, but the author of the \emph{Viveka\-mārtaṇḍa} expresses it more clearly.
 \end{philcomm}
@@
 \begin{translation}[hp02_018]
-[The yogi] should exhale correctly, inhale correctly and hold the breath correctly. He thus becomes purified.
+[The yogi] should exhale correctly, inhale correctly and hold the breath correctly. [His channels] thus become purified.
 \end{translation}
@@
 
-\emph{Yogacintāmaṇi} 92v (attr.~to the \emph{Skandpurāṇa})
+\emph{Yogacintāmaṇi} 92v (attr.~to the \emph{Skandapurāṇa})
 
@@
 %\begin{philcomm}[hp02_018]
-%Some of the alpha, epsilon and eta manuscripts have \emph{śuddhiṃ} instead of \emph{siddhiṃ} in the fourth verse quarter but this is likely a later change, or even an error, by someone who was anticipating the subject of the next verse.
+%Some of the alpha, zeta and eta manuscripts have \emph{śuddhiṃ} instead of \emph{siddhiṃ} in the fourth verse quarter but this is likely a later change, or even an error, by someone who was anticipating the subject of the next verse.
 The reading \emph{siddhiṃ} is attested across all branches of the stemma, as well as the source text and most of the testimonia.
@@
 \begin{philcomm}[hp02_021]
-Manuscripts of the delta group, as well as \etaTwo\ and \epsilonThree, also have the valid readings of \emph{medaḥśleṣmādināśārthaṃ} and \emph{anyathā} in the first and third verse quarters, respectively.
+Manuscripts of the \textdelta\ group, as well as \etaTwo\ and \zetaThree, also have the valid readings of \emph{medaḥśleṣmanivṛtyarthaṃ} (\deltaOne\deltaTwo\zetaThree)/\emph{medaḥśleṣmādināśārthaṃ} (\etaTwo) and \emph{anyathā} in the first and third verse quarters, respectively.
 This version of the verse states that one should practise the six therapeutic interventions to remove fat, phlegm and the like, otherwise one should not practise them when the humours are in balance.
-However, the \emph{pūrvam} (`first'), which is attested in two \textalpha\ manuscripts, as well as manuscripts of the \textbeta\ and \textgamma\ groups, fits the context of these interventions being preliminary practices for \emph{prāṇāyāma}.
+However, the \emph{pūrvam} (`first'), which is attested in the \textalpha, \textbeta\ and \textgamma\ groups, fits the context of these interventions being preliminary practices for \emph{prāṇāyāma}.
 \end{philcomm}
@@
 \begin{translation}[hp02_022]
-\emph{Dhauti, basti, neti, trāṭaka, naulī} and \emph{kapālabhātī}.
+\emph{Dhauti, basti, nīti, trāṭaka, naulī} and \emph{kapālabhātī}.
 These are said to be the six [therapeutic] techniques.%JM: "therapeutic interventions" adds a lot to karmāṇi.
@@
 \tl{dhautir bastis tathā netis trāṭakaṃ naulikaṃ tathā |\\+}
-\tl{kapālabhrāntir etāni ṣaṭkarmāṇi pracakṣate  ||\\!}
+\tl{kapālabhātir etāni ṣaṭkarmāṇi pracakṣate  ||\\!}     %  kapālabhrāntir   corrected
 \end{versinnote}
@@
 
-\begin{philcomm}[hp02_023]
-\end{philcomm}
+% \begin{philcomm}[hp02_023]
+% \end{philcomm}
 
@@
 \begin{translation}[hp02_024]
-{}[The yogi] should slowly swallow a moistened cloth  four finger-breadths in width and then draw it out. This ejection [of it] from the mouth is the dhauti technique.
+{[}The yogi] should slowly swallow a moistened cloth  four finger-breadths in width and then draw it out. This ejection [of it] from the mouth is the dhauti technique.
 \end{translation}
@@
 Some of the other readings, such as \emph{uditaṃ}, \emph{utthānaṃ}, etc., appear to be mistakes or patches that arose possibly because \emph{udgāra} is not normally neuter but masculine.
+\lb
+% MD: Why don't we adopt the reading ukhāla?
+ukhāla seems to mean "vomiting" in Hindi, Marathi, Nepali etc.
+See also https://www.wisdomlib.org/definition/ukhala-pakhala.
+We have accepted pakhāla in 2.26.
+Apparently they belong together.
+??JM to write note
 
 Many manuscripts have added verse quarters on the length of the cloth (\emph{hastapañcadaśena tu}) and doing the practice according to the guru’s teachings (\emph{gurūpadiṣṭa\-mārgeṇa}).
-These additional comments are absent in the \textalpha, \textgamma\ and \textepsilon\ groups.
+These additional comments are absent in the \textalpha, \textgamma\ and \textzeta\ groups.
 Furthermore, the compound \emph{hastapañcadaśena} does not seem to fit the syntax of the sentence.
@@
 \begin{translation}[hp02_025]
-Coughing, wheezing, splenitis and skin diseases, as well as the twenty phlegmatic diseases, are sure to disappear through the power of the dhauti technique.
+Coughing, wheezing, splenitis and skin diseases, as well as the twenty phlegmatic diseases, are sure to flee (\textit{dhāvanti}) through the power of the dhauti technique.
 \end{translation}
@@
 \begin{philcomm}[hp02_025]
-The verb \emph{dhāvanti} is a play on words, using a different root \emph{dhāv}, “run”, from that of \emph {dhauti}, which is derived from \emph{dhāv}, “purify”.
+The verb \emph{dhāvanti} is a play on words, using a different root \emph{dhāv}, “run”, from that of \emph {dhauti}, which is derived from \emph{dhāv}, “purify”. \lb
 
@@
 \end{translation}
-% Adopt pakhālaṃ — not done in latest version
 
@@
 \begin{philcomm}[hp02_026]
-% MD: kṣālanaṃ or prakṣālaṃ? Adopt pakhālaṃ [MD: done]
-The reading \emph{pakhālaṃ} that we have adopted in the fourth \emph{pāda} is found in \betaOmega \ and \alphaOne \ and reflects vernacular usage as found in the old Hindi \emph{Aṣṭāṅgayoga} of Caraṇadāsa (6.71ab).
+The reading \emph{pakhālaṃ} that we have adopted in the fourth \emph{pāda} is found in \alphaTwo\ and \betaOmega\ and is close to the reading of \alphaOne. %  not exactly: pākhālaṃ α1, paṣālaṃ α2
+It reflects vernacular usage as found in the old Hindi \emph{Aṣṭāṅgayoga} of Caraṇadāsa (6.71ab).
 Some other witnesses have the Sanskritised form \emph{prakṣālaṃ}, which is very rare in Sanskrit sources, while others have the more common \emph{kṣālanaṃ}.
@@
 \tl{gulmaplīhodarādīnāṃ nāśanaṃ sukhavardhanam ||\\+}
-\tl{vātapittakaphottānāṃ doṣāṇāṃ nāśanaṃ paraṃ |\\+}
+\tl{vātapittakaphotthānāṃ doṣāṇāṃ nāśanaṃ paraṃ |\\+}      %  kaphottānāṃ
 \tl{kuṣṭhānāṃ nāśanaṃ cāpi bastisiddhe prajāyate ||\\!}
@@
 \begin{versinnote}
-\tl{dhātvīndrintaḥ karaṇaprabodhaṃ\\+}
+\tl{dhātvīndriyāntaḥkaraṇaprabodhaṃ\\+}       %  dhātvīndrintaḥ corrected JH
 \tl{dadāti kāntiṃ dahanapradīptim |\\+}
@@
 \tl{udaragatapadārtham udvamantī\\+}
-\tl{pavanamapānam udīrya kaṇṭhanāle |\\+}
+\tl{pavanam apānam udīrya kaṇṭhanāle |\\+}
 \tl{kramaparicayavaśyavāyumārgā\\+}
@@
 \begin{philcomm}[hp02_029]
-Some manuscripts, including \etaTwo\ and \betaOmega, have an alternative reading for the third verse quarter that appears to be explaining the name of the practice.
+\etaTwo\ and \betaOmega\ have an alternative reading for the third verse quarter that appears to be explaining the name of the practice.
 In other words, it is called the elephant technique `because the speed of the breath is like that of water [propelled] by elephants’ (\emph{karibhir iva jalasya vāyuvegāt}).
 The syntax of this reading is not so easy to construe with the rest of the verse, which suggests that it was not original.
-The reading we have adopted (i.e., \emph{kramaparicayavaśya}...) is attested by manuscripts of the most important groups (\alphaOne\ and \alphaTwo, \betaTwo, the \textgamma\ group, etc.) and the same witnesses preserve \emph{mārga} (rather than \emph{vega}).
-
-%The \emph{Jyotsnā} and printed editions have a finite verb in the first hemistich and assume a plural subject (i.e., \emph{yoginaḥ}) instead of the feminine singular present participle (agreeing with \emph{gajakaraṇī}).
+The reading we have adopted (i.e., \emph{kramaparicayavaśya}...) is attested by manuscripts of the most important groups and the same witnesses preserve \emph{mārga} (rather than \emph{vega}).
+
+%The \emph{Jyotsnā} and printed editions have a finite verb in the first line and assume a plural subject (i.e., \emph{yoginaḥ}) instead of the feminine singular present participle (agreeing with \emph{gajakaraṇī}).
 (JB not necessary as alpha, etc.
@@
 \begin{translation}[hp02_030a]
-Now \emph{neti}.
+Now \emph{nīti}.
 \end{translation}
@@
 \begin{translation}[hp02_030]
-{[}The yogi] should insert a very smooth thread one handspan [in length] into the nasal passage and take it out through the mouth. This is called neti by the Siddhas.
+{[}The yogi] should insert a thread that is firm for one handspan [in length] into the nasal passage and take it out through the mouth. This is called nīti by the Siddhas.
 \end{translation}
@@
 \tl{mṛdu ślakṣṇaṃ sitaṃ sūtraṃ nāsānāle praveśayet |\\+}
-\tl{mukhān nirgamayed dasrau cintayen netikā smṛtā ||\\!}
+\tl{mukhān nirgamayed dasrau cintayen netikā smṛtā ||\\!}      % dasrau cintayen ? JH
 \end{versinnote}
@@
 \begin{philcomm}[hp02_030]
-According to Turner’s Comparative and Etymological Dictionary (1966: 427, entry 7588), the word \emph{netī} in Hindi refers to the cord of a churning stick and is cognate with the Sanskrit \emph{netra}.
+??
+JM: note on nīti/neti   According to Turner’s Comparative and Etymological Dictionary (1966: 427, entry 7588), the word \emph{netī} in Hindi refers to the cord of a churning stick and is cognate with the Sanskrit \emph{netra}.
 The action of pulling the cord of a churning stick is similar to the way the thread can be pulled back and forth, from side to side, through the nostril and mouth.
 
-The reading we have adopted in the third verse quarter (\emph{mukhān nirgamayec caiṣā}) is well attested but may not be original because the first hemistich has a finite verb and the \emph{eṣā} must be construed with \emph{netiḥ} in the fourth verse quarter.
+%The reading we have adopted in the third verse quarter (\emph{mukhān nirgamayec caiṣā}) is well attested but may not be original because the first line has a finite verb and the \emph{eṣā} must be construed with \emph{netiḥ} in the fourth verse quarter.
 The alternative reading \emph{mukhanirgamanād eva} (\etaOne\ and \etaTwo) may be original but is difficult to construe and \emph{mukhān nirgamayet} makes better sense in terms of describing the final part of the practice.
 \end{philcomm}
+%MD 2024-8: The second part of the commentary may be unnecessary. In 2.34, too, eṣā is construed with naulī in the forth Pāda.
+% Should we take vitastisusnigdhaṃ as a compound or separate it?
 %%%%%%%%%%
@@
 \begin{translation}[hp02_031]
-[The technique] with a small thread purifies the skull, bestows divine sight and quickly cures a multitude of diseases that arise above the jaw.
+The excellent Niti [technique] purifies the skull, bestows divine sight and quickly cures a multitude of diseases that arise above the collarbone.
 \end{translation}
 %??
-Change tr, JM to write note expand note on possibility of niti/nīti being original (found in alpha, betaone, epsilonone, delta i.e.
+JM to write note expand note on possibility of niti/nīti being original (found in alpha, betaone, zetaone, delta i.e.
 in good manuscripts in most groups; vernacular?) Only niti works well in final pāda.
@@
 Tricky; it seems that the spelling varies here between nīti and niti (in the header it's unanimously nīti in alpha)
-% adopt hanūrdhva
-% adopt jayaty āśu sa-sūtrikā
-
-% MD: more mss for sūtrikā:
-% jayaṃ(one syllable om.) sā sasūtrikā (K1), jayaṃtī sārasūtrikā (P27), jayate sā ca sūtrikā (B2) [gamma]; jayaṃti sārasūtrikā (P6) [contaminated epsilon]
-% Perhaps better to read jayaty āśu sa-sūtrikā (= neti)?
-% Or: jayate (Middle)/jayantī (Pres. Part.) sā sa-sūtrikā?
 
@@
 \tl{kapālaśodhinī caiva divyadṛṣṭipradāyinī |\\+}
-\tl{jatrūrdhvajātarogādyaiḥ jayatyeva suniścitam ||\\!}
+\tl{jatrūrdhvajātarogādyaiḥ jayaty eva suniścitam ||\\!}
 \end{versinnote}
@@
 We have adopted the reading \emph{caiva} in the first verse quarter, which is attested by manuscripts of the \textbeta, \textgamma\ and \textdelta\ groups and is easy to construe.
-The \textalpha\ manuscripts have \emph{kaṇṭhā} and \emph{vaṭyā}, which may derive from \emph{kaṇṭhyā} (\epsilonOne), but none of these readings makes sense here.
+\alphaOne\ and \alphaTwo\ have \emph{vaṭyā} and \emph{kaṃṭhā} respectively, which may derive from \emph{kaṇṭhyā} (\alphaThree, \epsilonOne, \zetaOne), but none of these readings makes sense here.
 \etaOne\ and \etaTwo\ have \emph{kāryā}, which is unnecessary because of the main verb in the final verse quarter.
-% MD: kaṇṭhyā (N19,G5,G11 - all epsilon)
-
-Most witnesses, including the \emph{Jyotsnā}, read \emph{jatrūrdhva°} in the third verse quarter, which is generally understood as `above the collar bones' or, as Brahmānanda says, the area above the shoulder joints (\emph{jatruṇoḥ skandhasandhyor ūrdhvam uparibhāge}).
+\lb
+
+Most witnesses, including the \emph{Jyotsnā}, read \emph{jatrūrdhva°} in the third verse quarter, which is generally understood as `above the collar bones' or, as Brahmānanda says, the area above the shoulder joints (\emph{jatruṇoḥ skandhasandhyor ūrdhvam upari\-bhāge}).
 On problems concerning the interpretation of \emph{jatru}, see Meulenbeld 1974: 465.
-We have adopted the reading of \textalpha, \emph{hanūrdhva°}, ‘above the jaw’, which makes good sense in the context of \emph{neti} because it is supposed to cleanse the skull (\emph{kapāla}) or, in other words, the head.
+We have adopted it against the reading of \alphaOne \ and \alphaTwo, \emph{hanūrdhva°}, ‘above the jaw’, and \alphaThree, \emph{kaṇṭhordhva°}, ‘above the throat’.\lb%, which makes good sense in the context of \emph{neti} because it is supposed to cleanse the skull (\emph{kapāla}) or, in other words, the head.
+% MD: Now I'm not quite sure whether hanu or jatru is the original, because jatru is lectio difficilior.
+And ūrdhvajatru seems to be known as the term for the part of the body above the torso, namely the neck and head.
 
@@
 \etaOne\ has \emph{jayati sā tu sūtrikā}, which is unmetrical, but the word \emph{sūtrikā} may have been original because it explains the problematic readings that arose in the other witnesses when attempts were made to replace \emph{sūtrikā} with \emph{neti}.
-The variants with \emph{netivit} seem implausible because of the epithets in the first hemistich, which require a feminine noun to be understood as the subject of the sentence.
+The variants with \emph{netivit} seem implausible because of the epithets in the first line, which require a feminine noun to be understood as the subject of the sentence.
 The reading \emph{netir āśu nihanti ca} looks like a patch that was adopted later in the transmission.
-The reading we have conjectured \emph{jayaty āśu tu sūtrikā} retains \emph{sūtrikā} and corrects the metrical fault of \etaOne\ by adopting \emph{jataty āśu}, which is well attested across the stemma.
+The reading we have conjectured \emph{jayaty āśu tu sūtrikā} retains \emph{sūtrikā} and corrects the metrical fault of \etaOne\ by adopting \emph{jayaty āśu}, which is well attested across the stemma.
+%?? MD: We have adopted the reading jayaty āśu satī nitiḥ. JM to rewrite the note.
 \end{philcomm}
@@
 \begin{translation}[hp02_032]
-{}[The yogi] should concentrate and look at a small focal point with a fixed gaze until tears fall. The experts consider this to be trāṭaka.
+{[}The yogi] should concentrate and look at a small focal point with a fixed gaze until tears fall. The experts consider this to be trāṭaka.
 \end{translation}
@@
 The witnesses have many different readings for the first word of this verse.
-The \textalpha\ manuscripts are split between \emph{modaka} (\alphaOne), \emph{mocaka} and \emph{mohana}.
+The \textalpha\ manuscripts are split between \emph{modaka} (\alphaOne), \emph{mocaka} (\alphaTwo) and \emph{movana} (\alphaThree).
 Of these, \emph{modaka} is the most likely, if it is understood as a medicinal pill.
 However this meaning is rare, even in medical literature, as \emph{modaka} is generally used to refer to a small sweet.
-We have adopted \emph{moṭakaṃ}, which is attested by \etaOne\ and \deltaTwo, as well as the \emph{Yuktabhavadeva}.
-We understand it to mean “destroyer.” Bohtlingk and Roth (s.v.) and Monier-Williams (s.v.) give medicinal pill as a possible meaning of \emph{moṭaka} (cf.
+We have adopted \emph{moṭakaṃ}, which is attested by \deltaTwo, \epsilonOne\ and \etaOne, as well as the \emph{Yuktabhavadeva}, and which we understand to mean “destroyer” (\emph{muṭa pramardane}, \emph{Dhātupātha} 1.346).
+Bohtlingk and Roth (s.v.) and Monier-Williams (s.v.) give medicinal pill as a possible meaning of \emph{moṭaka} (cf.
 \emph{modaka}) but it appears that this is mainly an inference drawn only from this verse, where the pill is merely a comparison.
 Several witnesses have \emph{sphoṭanaṃ} (‘destroying’), which is also possible.
+\lb
 
@@
 56).
-
-\medialink{Mahant Bālyogī Rām Bālak Dās performing nauli at Dīrgheśvar Mahādev, U.P., in 2012.}{BabajiNauli.m4v}
-
-The reading \emph{gauḍaiḥ} is found in only one collated manuscript (\etaTwo) but it is close to the readings of two \textalpha\ manuscripts, \emph{gaulaiḥ} (\alphaOne) and \emph{golaiḥ} (\alphaThree), and is also attested by the \emph{Haṭharatnāvalī} and \emph{Yuktabhavadeva}.
+\lb
+
+\medialink{Mahant Bālyogī Rām Bālak Dās performing nauli at Dīrgheśvar Mahādev, U.P., in 2012.}{BabajiNauli.m4v}\lb
+
+%The reading \emph{gauḍaiḥ} is found in only one collated manuscript (\etaTwo) but it is close to the readings of two \textalpha\ manuscripts, \emph{gaulaiḥ} (\alphaOne) and \emph{golaiḥ} (\alphaThree), and is also attested by the \emph{Haṭharatnāvalī} and \emph{Yuktabhavadeva}.
 This version of the verse, which is likely original, suggests that it was thought that the name \emph{naulī} came from the region known today as Bengal.
 The reading \emph{gauḍaiḥ} explains many of the mistakes in other manuscripts, such as \emph{gaulīḥ}, \emph{maulaiḥ}, etc.
+% MD: Now we have three manuscripts with gauḍaiḥ (β1,γ2,ε1).
+Perhaps not necessary to discuss the variant readings.
 \end{philcomm}
@@
 \begin{philcomm}[hp02_035]
-\alphaOne\ and \alphaTwo, \betaTwo, and \epsilonOne\ have \emph{tundāgni°} in the first verse quarter.
+\alphaOne\ and \alphaTwo, \betaTwo, \epsilonOne\ and \zetaOne\ have \emph{tundāgni°} in the first verse quarter.
 This reading is also supported by all of the manuscripts collated for the published edition of the \emph{Haṭharatnāvalī}.
@@
 The alternative reading \emph{mandāgni°} (`sluggish fire') is reasonably common in contexts of stimulating poor digestion, and may have been introduced early in the transmission to replace the more unusual \emph{tundāgni°}.
+\lb
 
@@
 Most witnesses and the testimonia have \emph{maulir iyaṃ} in the fourth verse quarter, which expresses the idea that naulī was thought to be the best of the \emph{ṣaṭkarma}, and the assonance of \emph{naulī} and \emph{mauli} may have been intended.
-The alternative reading of \emph{mūlam iyaṃ} in \etaOne\ and \epsilonOne\ would suggest that \emph{nauli} is necessary for the other practices, which does not seem to be the case because, according to \emph{Haṭhapradīpikā} 2.21, the \emph{ṣaṭkarma} are more like therapeutic interventions.
+The alternative reading of \emph{mūlam iyaṃ} in \zetaOne\ and \etaOne\ would suggest that \emph{nauli} is necessary for the other practices, which does not seem to be the case because, according to \emph{Haṭhapradīpikā} 2.21, the \emph{ṣaṭkarma} are more like therapeutic interventions.
 \end{philcomm}
@@
 \begin{translation}[hp02_036]
-Rapid inhalation and exhalation like the bellows of a blacksmith is called kapālabhātī, the skull bellows. It dries up imbalances of phlegm.
+Very rapid inhalation and exhalation like the bellows of a blacksmith is called kapāla\-bhātī, the skull bellows. It dries up imbalances of phlegm.
 \end{translation}
@@
 \begin{philcomm}[hp02_036]
-The word \emph{bhātī} is derived from \emph{bhastrī} (Turner 1966: 537, entry 9424).
-
-\etaTwo\ and other manuscripts have \emph{kuryāt savyāpasavyataḥ} (‘left and right’) instead of \emph{recapūrau sasambhramau}.
+The word \emph{bhātī} is derived from \emph{bhastrī} (Turner 1966: 537, entry 9424).\lb
+
+\etaTwo\ and other manuscripts have \emph{kuryāt savyāpasavyataḥ} (`left and right’) instead of \emph{recapūrau sasambhramau}.
 Although one might infer that \emph{savyāpasavyataḥ} is referring to performing the skull bellows breathing alternately through the left and right nostrils, it appears to be a secondary reading because there is no indication of what is moving to the left and right.
-Such a method of alternate nostril breathing is explained as a variation of kapālabhāti in the \emph{Haṭhayogasaṃhitā} (p.
-14):
+Such a method of alternate nostril breathing is explained as a variation of kapālabhātī in the \emph{Haṭhayogasaṃhitā} (p.
+14):   % kapālabhāti changed to long i JH
 \begin{versinnote} 
@@
 
-In the second verse quarter, the \textalpha\ manuscripts have \emph{susambhramau} ('great speed') instead of the reading we have adopted \emph{sasambhramau} (`fast').
+In the second verse quarter, the \textalpha\ manuscripts have \emph{susambhramau} (`great speed') instead of the reading we have adopted \emph{sasambhramau} (`fast').
 The syntax requires \emph{sasambhramau} because it is an adjectival compound that must agree with \emph{recapūrau} (`the exhalation and inhalation').
+%??JM to change note on why we have adopted susambhramau, which is rare (unique?) as an adjectival form whereas sasambhrama is very common.
 \end{philcomm}
@@
 \begin{philcomm}[hp02_038]
-The plural of \emph{prāṇāyāma}, which we have translated here as ‘breath-controls’, probably refers to practising multiple repetitions of breath retentions with alternate nostril breathing.
+The plural of \emph{prāṇāyāma}, which we have translated here as ‘breath-controls’, probably refers to practising multiple repetitions of breath retentions with alternate nostril breathing. \lb
 
@@
 (JB: \emph{praśuṣyanti}  is so well attested and straight forward that there's no need to comment on it)
-Many witnesses lower on the stemma have \emph{malāśaya} (instead of \emph{malā api} or \emph{malā iti}) in the second verse quarter but this usually has the more specific meaning of bowels or bladder and so seems inappropriate in a general statement.
+% MD: In the present set of the witnesses "malā iti" and "praśuṣyanti" are well attested. There may be no need to comment on the inferior reading.
+%Many witnesses lower on the stemma have \emph{malāśaya} (instead of \emph{malā api} or \emph{malā iti}) in the second verse quarter but this usually has the more specific meaning of bowels or bladder and so seems inappropriate in a general statement.
 The witnesses that have \emph{malāśaya} also have the verb \emph{pra+śudh}, which connotes that the place where the impurities accumulate is cleaned (rather than the impurities themselves).
@@
 \end{versinnote}
-Different versions of this verse occur in \etaTwo\ and \epsilonThree.
+Different versions of this verse occur in \etaTwo\ and \zetaThree.
 Neither of these appear to be original as the compound \emph{sumanaskāntaka} (`dying with the mind active'?) is rather odd.
-The \delta \ group of manuscripts has another reading for the third quarter (\emph{tena siddhiṃ gatās te ca}), which puts a more positive spin on the verse in as much as one should practise \emph{prāṇāyāma} because through it the gods attained perfection.
-%\epsilonThree\ has a similar version of the third quarter (\emph{samanaskāntakabhayaṃ}), which might refer to the fear of dying with the mind active as opposed to the preferable situation of dying while in \emph{samādhi}.
+The \textdelta\ group of manuscripts has another reading for the third quarter (\emph{tena siddhiṃ gatās te ca}), which puts a more positive spin on the verse in as much as one should practise \emph{prāṇāyāma} because through it the gods attained perfection.
+%\zetaThree\ has a similar version of the third quarter (\emph{samanaskāntakabhayaṃ}), which might refer to the fear of dying with the mind active as opposed to the preferable situation of dying while in \emph{samādhi}.
 \end{philcomm}
@@
 \tl{vidhivat prāṇasaṃyāmaiḥ nāḍicakre viśodhite |\\+}
-\tl{suṣumnāvadanaṃ bhitvā sukhād viśati mārutaḥ ||\\!}
+\tl{suṣumnāvadanaṃ bhittvā sukhād viśati mārutaḥ ||\\!}    %  bhitvā corrected JH
 \end{versinnote}
@@
 \begin{philcomm}[hp02_043]
-The majority of witnesses, including \alphaTwo, have a plural subject (\emph{vidhānajñāḥ}) and verb (\emph{kurvanti}) in the first hemistich.
-We have adopted the singular, which is attested by \alphaOne, \alphaThree\ and manuscripts of the \emph{Haṭharatnāvalī}, because it corresponds with the singular subject of the second hemistich.
+The majority of witnesses, including \alphaTwo, have a plural subject (\emph{vidhānajñāḥ}) and verb (\emph{kurvanti}) in the first line.
+We have adopted the singular, which is attested by \alphaOne, \alphaThree\ and manuscripts of the \emph{Haṭharatnāvalī}, because it corresponds with the singular subject of the second line.
 \end{philcomm}
@@
 \begin{translation}[hp02_044]
-Sūryabhedana, ujjāyī, śītkā, śītalī, bhastrikā, bhrāmarī, mūrcchā and plāvanī: these are the eight kumbhakas.
+Sūryabhedana, ujjāyī, śītkā, śītalī, bhastrikā, bhramarī, mūrcchā and kevala: these are the eight kumbhakas.
 \end{translation}
@@
 % adopt plāvanī
+%MD 2024-8: alpha and some other good mss have usually bhramarī, not bhrāmarī.
 
@@
 
+\emph{Haṭhayogasaṃhitā} p. 60
+\begin{versinnote}
+\tl{sahitaḥ sūryabhedī ca ujjāyī śītalī tathā |\\+}
+\tl{bhastrikā bhrāmarī mūrcchā kevalī cāṣṭa kumbhakāḥ  ||\\!}
+\end{versinnote}
 \end{testimonia}
@@
 \begin{philcomm}[hp02_044]
-In the fourth verse quarter, only some manuscripts of the \textepsilon\ and \textgamma\ groups have the adopted reading \emph{plāvanīty aṣṭa}.
+On the occurrence of \emph{plāvanī} in this list in many manuscript groups, including the \textit{Jyotsnā}, see the note on verse 2.71.
+%In the fourth verse quarter, only some manuscripts of the \textbeta, \textzeta\ and \textgamma\ groups have the adopted reading \emph{plāvanīty aṣṭa}.
 The name \emph{plāvanī} is necessary in this list because most manuscript groups, including \textalpha, and the testimonia contain a verse on \emph{plāvinīkumbhaka} as one of the eight \emph{kumbhaka}s.
@@
 
+% Cf. Gheraṇḍasaṃhitā 5.47 (testimonia)
+% sahitaḥ sūryabhedaś ca ujjāyī śītalī tathā |
+% bhastrikā bhrāmarī mūrcchā kevalī cāṣṭa kumbhikāḥ ||5.47||
+
+% MD: It seems to me that the plāvanī verse was added at a very early stage of the transmission.
+In the manuscripts of the alpha group the heading of plāvanī is missing.
+Instead, (iti) plāvanīkumbhakam is found after the verse.
+Very unusual.
+Is it not possible that originally kevala was counted as the eighth kumbhaka, as in the Gheraṇḍasaṃhitā?
+
 %%%%%%%%%%
@@
 \begin{translation}[hp02_045]
-At the end of the inhalation, the lock called \emph{jālandhara} is to be performed, while at the end of the retention and beginning of the exhalation, \emph{uḍḍiyāna} is to be performed.
+At the end of the inhalation, the lock called \emph{jālandhara} is to be performed, while at the end of the retention and beginning of the exhalation, \emph{uḍḍiyāṇa} is to be performed.
 \end{translation}
@@
 %\emph{madhyapaścimatānena} is possible and well attested.
-As is clear in the source text, the \emph{Gorakṣaśataka}, the three techniques alluded to here are \sl{mūlabandha, jālandharabandha} and \emph{uḍḍiyānabandha} respectively (on which see chapter three).
+As is clear in the source text, the \emph{Gorakṣaśataka}, the three techniques alluded to here are \sl{mūlabandha, jālandharabandha} and \emph{uḍḍiyānabandha} respectively (on which see chapter three).\lb
 
@@
 We understand it to refer to Brahman, the absolute, in all instances other than \emph{brahmagranthi}, where we take it to mean the deity Brahmā.
-See Mallinson 2007:205 n.240.
+See Mallinson 2007: 205 n.
+240.
 \end{philcomm}
@@
 \begin{translation}[hp02_048a]
-Now, piercing the sun:
+Of these [breath retentions], piercing the sun [is now taught]: 
 \end{translation}
@@
 \begin{sources}[hp02_048]
-Cf. \emph{Gorakṣaśataka} 33–34ab
+Cf.~\emph{Gorakṣaśataka} 33–34ab
 \begin{versinnote}
@@
 
-\begin{philcomm}[hp02_048]
+%\begin{philcomm}[hp02_048]
 %\emph{āsane sukhade} is supported by the [different but parallel] reading in the Gorakṣa\-śataka. (JB: I've put the GŚ verses in the source section)
-%After citing this verse with attribution to the \emph{Haṭhapradīpikā} in his \emph{Haṭhasaṅketacandrikā} (f.~78v), Sundaradeva remarks that \emph{sukhade} is a \emph{hetugarbhaviśeṣaṇam}, which indicates that he understood the first hemistich as saying that the yogi assumes \emph{vajrāsana} on a mat (\emph{āsana}) because it is comfortable.%JM:needs more explanation.
-\end{philcomm}
+%After citing this verse with attribution to the \emph{Haṭhapradīpikā} in his \emph{Haṭhasaṅketacandrikā} (f.~78v), Sundaradeva remarks that \emph{sukhade} is a \emph{hetugarbhaviśeṣaṇam}, which indicates that he understood the first line as saying that the yogi assumes \emph{vajrāsana} on a mat (\emph{āsana}) because it is comfortable.%JM:needs more explanation.
+%\end{philcomm}
 
@@
 \begin{versinnote}
-{}[The yogi] who is constantly engaged [in practice] should repeat breath retentions again and again. Extreme heat burns as far as the tips of the hair and nails. 
+{[}The yogi] who is constantly engaged [in practice] should repeat breath retentions again and again. Extreme heat burns as far as the tips of the hair and nails. 
 \end{versinnote}
@@
 The meaning of \emph{nirodhāvadhi} is not entirely clear, but all sources and the \emph{Jyotsnā} (2.49) agree on this reading.
-The original reading may well have been \emph{virodhāvadhi}, “until it is uncomfortable”, which makes better sense but we have decide not to emend accordingly.
-The problem with \emph{nirodhāvadhi} is that to practise \emph{kumbhaka} ``up to cessation \emph{nirodha}'' seems to suggest that cessation is not that of the physical breath, which by definition ceases in \emph{kumbhaka}, but of the vital wind (\emph{prāṇa}) within the body.
+The original reading may well have been \emph{virodhāvadhi}, “until it is uncomfortable”, which makes better sense but we have decided not to emend accordingly.
+The problem with \emph{nirodhāvadhi} is that to practise \emph{kumbhaka} ``up to cessation (\emph{nirodha})'' seems to suggest that cessation is not that of the physical breath, which by definition ceases in \emph{kumbhaka}, but of the vital wind (\emph{prāṇa}) within the body.
 By citing a verse from an unnamed text, Brahmānanda seems to understand this verse as saying that the breath should very carefully (\emph{atiprayatnena}) be held as far as the extremities of the body so that it does not damage the body by exiting through the hair follicles:
@@
 \tl{tundasya vātadoṣaghnaḥ kṛmidoṣaṃ nihanti ca ||\\+}
-\tl{punaḥ punar idaṃ kāryaṃ sūryābhedam udāhṛtam |\\!}
+\tl{punaḥ punar idaṃ kāryaṃ sūryabhedam udāhṛtam |\\!}
 \end{versinnote}
@@
 \tl{kapālaṃ śodhanaṃ cāpi recayet pavanaṃ śanaiḥ ||\\+}
-\tl{kapālaṃ ... śanaiḥ ] kapālaśodhanaṃ vātadoṣaghnaṃ kṛmināśanaṃ N,n1,n4.\\+}
+\tl{\var{kapālaṃ ... śanaiḥ ] kapālaśodhanaṃ vātadoṣaghnaṃ kṛmināśanaṃ N,n1,n4.}\\+}
 \tl{ālasyaṃ vātadoṣaghnaṃ kṛmikīṭaṃ nihanti ca |\\+}
@@
 %The \emph{Gorakṣaśataka} and \emph{Yogacintāmaṇi} support \emph{kṛmidoṣa}. [JB: Not an issue in the new collation]
-The compound \emph{sūryabheda} is metri causa, as the name given in 2.44 and the heading of 2.48 is \emph{sūryabhedana}.
-Some witnesses, such as the delta group and \getsiglum{V19}, have attempted to reinstate the name \emph{sūryabhedana}.
-
-Both \emph{°doṣaham} and \emph{°doṣahṛt} are well attested and possible. We have favoured the former because the \textalpha\ reading (\emph{doṣajam}) appears to be a corruption of it.
+%The compound \emph{sūryabheda} is metri causa, % MD: or simply followed the source text?
+%as the name given in 2.44 and the heading of 2.48 is \emph{sūryabhedana}. Some witnesses, such as the delta group and \getsiglum{V19}, have attempted to reinstate the name \emph{sūryabhedana}.\lb
+
+Both \emph{°doṣaham} and \emph{°doṣahṛt} are well attested and possible.
+We have favoured the former because the \textalpha\ reading (\emph{doṣajam}) appears to be a corruption of it, and °ha is from the same root as the verb in the source text (i.e., \textit{nihanti}).% MD: and also because the source text has nihanti.
 \end{philcomm}
@@
 \tl{mukhaṃ saṃyamya nāsābhyāṃ ākṛṣya pavanaṃ śanaiḥ |\\+}
-\tl{yathā lagati kaṇṭhe suḥ hṛdayāvadhi sasvanam ||\\!}
+\tl{yathā lagati kaṇṭhe suḥ hṛdayāvadhi sasvanam ||\\!}       % suḥ  ? JH
 \end{versinnote}
@@
 \tl{ruddhvā keśān nakhāgrāvadhi pavanam amuṃ recayed vāmanāḍyā\\+}
-\tl{proktojjāyīti kumbhaḥ kaphagadadalano dīpti kṛjjāṭharāgne ||\\+}
+\tl{proktojjāyīti kumbhaḥ kaphagadadalano dīptikṛjjāṭharāgne ||\\+}  % dīptikṛj, not dīpti kṛj   JH
 \end{versinnote}
@@
 \begin{translation}[hp02_052]
-As before, he should hold the breath and then exhale through Iḍā. [Because] it cures disorders caused by phlegm in the throat and increases the body’s fire, [\dots]
+As before, he should hold the breath and then exhale through Iḍā. [Because] it cures disorders caused by phlegm in the throat and increases the body’s fire, [\ldots]
 \end{translation}
@@
 \begin{versinnote} 
-\tl{pūrvavat kumbhayet prāṇān recayediḍayā tataḥ |\\+}
+\tl{pūrvavat kumbhayet prāṇān recayed iḍayā tataḥ |\\+}
 \tl{śleṣmoṣaharaṃ dehānaladīptipravardhanam ||\\!}
@@
 \begin{philcomm}[hp02_052]
-In the second hemistich of this verse, many of the readings in the oldest manuscripts, such as \emph{dehād analadīptivardhanam} (\getsiglum{V1}), \emph{dehānaladīptivivardhanam} (\getsiglum{J10}) and \emph{dehe [’]naladīptivivardhanam} (\getsiglum{P28}), are unlikely to be original because both the source and testimonia indicate that \emph{Ujjāyī} is supposed to remove phlegm from the throat.
+%In the second line of this verse, many of the readings in the oldest manuscripts, such as \emph{dehād analadīptivardhanam} (\getsiglum{V1}), \emph{dehānaladīptivivardhanam} (\getsiglum{J10}) and \emph{dehe [’]naladīptivivardhanam} (\getsiglum{P28}), are unlikely to be original because both the source and testimonia indicate that \emph{Ujjāyī} is supposed to remove phlegm from the throat.
 These versions may have arisen from attempts to remove \emph{kaṇṭhe} in the third \emph{pāda}, which was thought to be hanging.
@@
 \end{philcomm}
+% MD: Do we need this comment? JM: I don't think so.
 
@@
 \begin{translation}[hp02_053]
-[and] cures diseases in the bodily constituents inside the network of channels,
-the retention called \emph{ujjāyī} should be done when [the yogi] is moving or remaining still.
+[and] cures diseases in the channels, fluids, stomach, and as far as all the bodily constituents, the retention called \emph{ujjāyī} should be done when [the yogi] is moving or remaining still.
 \end{translation}
@@
 \begin{philcomm}[hp02_053]
-Nearly all the manuscripts have \emph{nāḍījalodara}°, which does not make sense in this context because \emph{jalodara} is the disease ascites and \emph{°gata°} requires a location.
-This problem is also present in the transmission of the source text for the verse, the \emph{Gorakṣaśataka}.
-A solution can be found in some of the manuscripts of the \emph{Haṭharatnāvalī}, which read \emph{nāḍījālodarā°} (‘in the network of channels and stomach’).
-The other problem is \emph{°darādhātu°}.
-In spite of Brahmānanda’s efforts to explain it as \emph{°dara}, \emph{ā}, and \emph{dhātu°}, the \emph{ā} before \emph{dhātu°} appears to have been inserted metri causa.
-We have adopted \emph{nāḍījālodare} (‘inside the network of channels’), which is close to \getsiglum{V3} and \getsiglum{J10} (\emph{nāḍījalodare}).
-It is likely that \emph{°jālodare} was changed to \emph{°jalodara} in the transmission because of confusion with the disease of a similar name.
-
-For the idea of \emph{doṣa}s being in \emph{dhātu}s see \emph{Tantrāloka} 28.283cd, where worldly concepts are said to arise from it (\emph{dhātudoṣāc ca saṃsārasaṃskārās te ...}), but it is also the source of physical disorders (\emph{dhātudoṣakṛtaṃ mūrcchā} ĪPV on 2.15).
+Nearly all the manuscripts have \emph{nāḍījalodarādhātu}°, which is difficult to understand because of \emph{jala} (`water') in this context and \emph{ā}+\emph{dhātu} within the compound.
+This reading is also present in the transmission of the source text for the verse, the \emph{Gorakṣaśataka}, and most of the testimonia.
+Given the strength of the evidence, we have adopted the reading and understood it as a list of things in which diseases (\emph{doṣa}) may occur, which is consistent with Brahmānanda's explanation (\emph{Jyotsnā} 2.53).
+He explains \emph{ā}+\emph{dhātu} as \emph{samantāt ...
+dhātavaḥ} (`the bodily constitutents altogether').\lb
+
+There are a couple of alternatives to these problems but they are not well-attested.
+For example, some of the manuscripts of the \emph{Haṭharatnāvalī} read \emph{nāḍījālo°} (‘in the network of channels’), which makes better sense than \emph{nāḍījalo°}, as \emph{jala} usually means water rather than bodily fluid.
+Brahmānanda understands \emph{jala} as water that has been drunk (\emph{pītam udakam}), which seems far-fetched as the site of a disease.
+Also, °\emph{dare dhātu}° is attested in two old manuscripts, which avoids the need to read °\emph{darādhātu}°.
+If one were to conjecture \emph{nāḍījālodare}, the hemistich would mean, `it cures diseases of the bodily constitutents inside the network of channels.'\lb
+
+%  \emph{nāḍījālodare} (‘inside the network of channels’), which is close to \getsiglum{V3} and \getsiglum{J10} (\emph{nāḍījalodare}).
+It is likely that \emph{°jālodare} was changed to \emph{°jalodara} in the transmission because of confusion with the disease of a similar name.\lb%JM: we could keep jalo, "in the fluid in the channels"
+% ?? JB New Note on Brahmānanda's reading and interpretation: done!
+%%
+
+For the idea of \emph{doṣa}s being in \emph{dhātu}s see \emph{Tantrāloka} 28.283cd, where worldly concepts are said to arise from it (\emph{dhātudoṣāc ca saṃsārasaṃskārās te ...}), but it is also the source of physical disorders (\emph{dhātudoṣakṛtaṃ mūrcchā} ĪPV on 2.15).\lb
 
@@
 \begin{translation}[hp02_054]
-{}[The yogi] should continuously make \emph{sīt} sound in the mouth and flare his nostrils. By practising in this way he becomes a second god of love.
+{[}The yogi] should continuously make \emph{sīt} sound in the mouth and flare his nostrils. By practising in this way he becomes a second god of love.
 \end{translation}
@@
 The result of becoming a second god of love may be connected with the sound \emph{sīt}, which is said to be made during sex in the \emph{Kāmasūtra} (2.7.4–19).
+\lb
 
@@
 \tl{yoginīguṇasāmānyaḥ sṛṣṭisaṃhārakārakaḥ ||\\+}
-\tl{na kṣudhā na ca tṛṇnidrā naiva murchā prajāyate |\\!}
+\tl{na kṣudhā na ca tṛṇ nidrā naiva murchā prajāyate |\\!}    % tṛṇ nidrā for tṛṇnidrā JH
 \end{versinnote}
@@
 % We should adopt bhuvimaṇḍale (alpha 3, eta 2) no? 
+%MD 2024-8: I don't know whether we have to accept this rather rare compound form when α1 and α2 have the normal bhūmimaṇḍale.
+Perhaps it is easier to adopt bhūmi° and note that the alternative bhuvi° is also possible.
 
@@
 \begin{sources}[hp02_057]
-\emph{Kauljñānanirṇaya} 6.19
+\emph{Kaulajñānanirṇaya} 6.19
 \begin{versinnote}
@@
 \tl{etad evoktaṃ gorakṣanāthena-\\+}
-\tl{rasanātālumūlena yaḥ prāṇaṃ satataṃ pibete |\\+}
+\tl{rasanātālumūlena yaḥ prāṇaṃ satataṃ pibet |\\+}      % pibete
 \tl{abdārdhena bhavet tasya sarvarogaparikṣayaḥ ||\\!}
@@
 He says that a hole or cavity (\emph{vivara}) is made by the root of the palate with the help of the tongue.
-The yogi breathes through it (\emph{evaṃ rasanātālumūlena rasanā jihvā tatsahāyabhūtatālumūlena kṛtaṃ yad vivaraṃ, tena kṛtvā yaḥ yogī prāṇam anilaṃ prāṇavāyuṃ pibet pūrayet, tasya yogino ’bdārdhena ṣaṇmāsena sarvarogāṇāṃ nāśaḥ kṣayo bhavet} |).
+The yogi breathes through it (\emph{evaṃ rasanātālumūlena rasanā jihvā tatsahāyabhūtatālumūlena kṛtaṃ yad vivaraṃ, tena kṛtvā yaḥ yogī prāṇam anilaṃ prāṇavāyuṃ pibet pūrayet, tasya yogino ’bdārdhena ṣaṇmāsena sarvarogāṇāṃ nāśaḥ kṣayo bhavet} |).\lb
 
@@
 \begin{versinnote}
-{}[The yogi] turns the tongue upwards and takes in the breath while making a \emph{sīt} sound.\\ 
+{[}The yogi] turns the tongue upwards and \\
+takes in the breath while making a \emph{sīt} sound.
+
 \tl{rasanām unmukhīkṛtya sītkāraṃ kurvatā marut |\\!}%JM: could kumbhake tasmin (yasmin prob ok) refer to the practice as a whole, not the retention part of it?
@@
 \begin{translation}[hp02_058]
-The wise man should draw in air through the tongue and after retaining the breath as before gradually exhale through the nostrils .
-\end{translation}
+The wise man should draw in air through the tongue and after retaining the breath as before gradually exhale through the nostrils.
+\end{translation}
+%MD 2024-8: I believe that kumbhanād anu may be a preferable option, not only in view of the stemma, but also from the context. In combination with pūrvavat, the action noun kumbhana may fit better.
 
@@
 
+%add m virāma at end of line in ed
 \emph{Yogacintāmaṇi} ms.~L, f.~70r
@@
 \begin{translation}[hp02_061]
-Having correctly adopted the lotus pose, with his neck and torso straight the wise man should close the mouth and forcefully exhale the breath through the nose
+Having correctly adopted the lotus pose, with his neck and torso straight the wise man should close the mouth and forcefully exhale the breath through the nose [\ldots]
 \end{translation}
@@
 \begin{translation}[hp02_062]
-in such a way that it comes into contact with the chest, throat and skull, making a sound. He should then quickly inhale a small amount of breath as far as the heart lotus.
-\end{translation}
+[\ldots] in such a way that the breath comes into contact with the chest, throat, and skull. He should then quickly inhale a small amount of breath as far as the heart lotus.
+\end{translation}%JM: check  revised GŚ reading and change testimonia, perhaps with variants
 
@@
 \begin{philcomm}[hp02_062]
-%First hemistich is tricky. None of the old mss preserve \emph{sasvanaṃ}, which is in the GŚ and makes good sense. 
-% MD: Now we have at least two good witnesses P11(β1),P15(ε1) with the reading sasvanaṃ.
-
-Most witnesses have \emph{hṛtkaṇṭhe}, which we have understand as a dual accusative.
-One would expect \emph{kapāla} also to be in the accusative, but we have understood it as a locative sg.
+%First line is tricky. None of the old mss preserve \emph{sasvanaṃ}, which is in the GŚ and makes good sense. 
+%MD 2024-8: Is "making a sound" an important component here? α1 has śvasanaṃ "wind, breath" instead of sasvanaṃ.
+
+%Most witnesses have \emph{hṛtkaṇṭhe}, which we have understood as a dual accusative.
+One would expect %\emph{kapāla} also to be in the accusative, but we have understood it as a locative sg.
 with \emph{sasvanam}.
+
+Apart from \alphaOne, the reading \emph{śvasanaṃ} is not well attested.
+We have adopted it because it provides a plausible subject for the verb \emph{lagati} (i.e., `breath comes into contact with').
+Most manuscripts, including the \emph{Jyotsnā}, have \emph{sasvanam} (`with sound'), which requires that one interpret this compound as an adverb and infer a subject that is different to that of the previous sentence.
+Breathing `noisily' is generally understood to characterise the rapid breathing of \emph{bhastrikā}.
+The term \emph{śvasana} is rarely used in yoga texts and may have been used here to connote audible breathing, for in some other contexts it can mean hissing, sighing and clearing the throat (s.v.
+MW: 1899).\lb
+
+%??
+JB write a note on śvasanaṃ (better for the syntax, but the idea of breathing with sound –as described in the GŚ– is lost).
+However, śvasana may (according to the dictionary) denote an audible breath, such as hissing, sighing and clearing the throat.
+done!
 \end{philcomm}
@@
 \begin{translation}[hp02_064]
-[...] the wise man should move the breath in his body. When fatigue arises in the body he should inhale through the sun [channel]
+[... the yogi] should attentively move the breath in his body. When fatigue arises in the body he should inhale through the sun [channel] [\ldots]
 \end{translation}
@@
 \begin{versinnote}
-\tl{tathaiva svaśarīrasthaṃ cālayet pavanaṃ sudhīḥ |\\+}
+\tl{tathaiva svaśarīrasthaṃ cālayet pavanaṃ dhiyā |\\+}
 \tl{yadā śramo bhaved dehe tadā sūryeṇa pūrayet |\\!}
@@
 \end{sources}
-
+% JM: adopt dhiyā in pāda b: it's what I now have in GŚ. Done
 \begin{testimonia}[hp02_064]
@@
 \begin{philcomm}[hp02_064]
-Most of the witnesses (including \textalpha) have \emph{dhiyā} at the end of the second verse quarter, but the manuscripts of the source text and the \emph{Haṭharatnāvalī} have \emph{sudhīḥ}.
+%Most of the witnesses (including \textalpha) have \emph{dhiyā} at the end of the second verse quarter, but the manuscripts of the source text and the \emph{Haṭharatnāvalī} have \emph{sudhīḥ}.
 Since the subject of the simile is \emph{bhastrā}, one would expect the subject of \emph{cālayet}, which must be different, to be stated (as is the case with \emph{sudhīḥ}).
@@
 \begin{translation}[hp02_065]
-in such a way that the abdomen is filled by the breath, and hold the nose quickly [and] firmly without using the middle and index fingers.
+{[}\ldots] in such a way that the abdomen is filled by the breath, and hold the nose quickly [and] firmly without using the middle and index fingers.
 \end{translation}
@@
 \begin{philcomm}[hp02_065]
-Only two of the collated witnesses (\getsiglum{J7}, \getsiglum{V15}) have \emph{madhyātarjanībhyāṃ} (`with the middle and index fingers') and this reading is not well attested by the manuscripts of the source text and testimonia.
-To hold the nose without the middle and index fingers is consistent with the way alternate nostril breathing is done in modern yoga (e.g., Iyengar 1991: 443–444).
+Only two of the collated witnesses (\gammaTwo\ and \zetaThree) have \emph{madhyātarjanībhyāṃ} (`with the middle and index fingers') and this reading is not well attested by the manuscripts of the source text and testimonia.
+To hold the nose without the middle and index fingers is consistent with the way alternate nostril breathing is done in modern yoga (e.g., Iyengar 1991: 443–444) as well as in some ritual manuals, as for instance in
+Sāhib Kaula's \emph{Śyāmāpaddhati}: \emph{dakṣānāmākaniṣṭābhyāṃ vāmaṃ puṭaṃ ca niruddhya} (see Hanneder: forthcoming).   % JH addition
 However, the reading of many manuscripts suggests  that the nose was held by all the fingers of both hands, except the index fingers (\emph{nāsikāmadhye tarjanībhyāṃ vinā}), which seems highly impracticable, or that the nose was held by only the index fingers of both hands (\emph{nāsikāmadhye tarjanībhyāṃ tathā}).
@@
 \begin{translation}[hp02_067]
-it is an auspicious thunderbolt that awakens Kuṇḍalinī, destroys bad deeds, bestows happiness, and destroys the blockage of phlegm, etc., situated at the mouth of the central channel,[\dots]
+[\ldots] it awakens Kuṇḍalinī, it stops her from being coiled, bestows happiness, is auspicious, and destroys the blockage of phlegm, etc., situated at the mouth of the central channel,[\dots]
 \end{translation}
@@
 \begin{philcomm}[hp02_067]
-The reading \emph{vajraṃ} in the first verse quarter is an emendation based on the manuscripts of the \emph{Gorakṣaśataka}, the source text for this verse.
+%The reading \emph{vajraṃ} in the first verse quarter is an emendation based on the manuscripts of the \emph{Gorakṣaśataka}, the source text for this verse.
 Some manuscripts of the \emph{Haṭhapradīpikā} have readings close to \emph{vajraṃ}, such as \emph{vipra} (\getsiglum{V1}, \getsiglum{P15}) and \emph{vakra} (G11), which suggest that \emph{vajraṃ} was changed at an early stage of the transmission.
-The \textalpha\ group have \emph{cakraṃ}, which could be understood as a weapon (i.e., the discus often associated with Viṣṇu), but such a meaning would be rather unusual in yoga texts, where the term \emph{cakra} is used so frequently in contexts of the yogic body.
+
+In the first verse quarter, we have adopted \textit{vakra}° (\epsilonOne), which is close to the reading of the \textalpha\ group (i.e., \emph{cakraṃ}).
+In compound with \textit{°bhāvaghnaṃ} (\alphaOne, \alphaThree, etc.) and following the awakening of Kuṇḍalinī, \textit{vakra}° makes good sense.
+According to the manuscripts available to us, the reading of the \emph{Gorakṣaśataka} (\emph{vajraṃ}) is not in the transmission of the \textit{Haṭhapradīpikā}.
+
+% MD: rewrite the comment. We have adopted the reading vakra-bhāvaghnaṃ.
+
 %The J10 group has rewritten this verse in the masc sg.
@@
 \begin{translation}[hp02_068]
-[and] pierces the three knots that have arisen from the three \emph{guṇa}s, it is particularly important to perform this retention called ‘the bellows’.
+[\ldots and] pierces the three knots that have arisen from the three \emph{guṇa}s, it is particularly important to perform this retention called ‘the bellows’.
 \end{translation}
@@
 In the \emph{Gorakṣaśataka}, it is stated clearly that each of the knots arise from one of the three \emph{guṇa}s: \emph{brahmagranthi} from \emph{rajas} (78cd), \emph{viṣṇugranthi} from \emph{sattva} (79cd) and \emph{rudragranthi} from \emph{tamas} (80cd).
-When the verses on \emph{bhastrā} were extracted from the \emph{Gorakṣaśataka} without the context of the knots and \emph{guṇa}s, the meaning of \emph{guṇa\-traya\-sam\-udbhūta\-granthi\-traya°} (`the three knots that have arisen from the three \emph{guṇa}s') was lost early in the transmission of the \emph{Haṭhapradīpikā}.
-\end{philcomm}
+When the verses on \emph{bhastrā} were extracted from the \emph{Gorakṣaśataka} without the context of the knots and \emph{guṇa}s, the meaning of \emph{guṇa\-traya\-sam\-udbhūta\-granthi\-traya°} (`the three knots that have arisen from the three \emph{guṇa}s') may have been lost early in the transmission of the \emph{Haṭhapradīpikā}.
+Alternatively it could be that Svātmārāma deliberately changed the compound in order to remove the reference to the \emph{granthi}s being produced from the \emph{guṇa}s.
+\end{philcomm}
+% MD: Many mss including alpha-mss read °samudbhūtaṃ. Perhaps pāda a is to be construed with kumbhakam.
 
@@
 \begin{translation}[hp02_069a]
-Now bhrāmarī:
+Now bhramarī:
 \end{translation}
@@
 \tl{recakaṃ mandamandam |\\+}
-\tl{yogīndrāṇāmevamabhyāsayogāc\\+}
+\tl{yogīndrāṇām evam abhyāsayogāc\\+}
 \tl{citte jātā kācid ānandalīlā || \\!}
@@
 \tl{vegākṛṣṭiṃ pūrakaṃ bhṛṅganādaṃ \\+}
-\tl{bhaṅgīnādaṃ recakaṃ mandaṃ mandaṃ |\\+}
-\tl{yogīdrāṇām evam abhyāsayogac \\+}
+\tl{bhaṅgīnādaṃ recakaṃ mandaṃ mandaṃ |\\+}  %  add  (recte bhṛṅgī-) ? unless it is a typo  JH
+\tl{yogīdrāṇām evam abhyāsayogac \\+}         %  add (recte yogīndrāṇām) ? JH
 \tl{cite jātā kācid ānandamūrchā ||\\+}
@@
 When explaining this verse, Brahmānanda (\emph{Jyotsnā} 2.68) supplies the gerund \emph{kṛtvā} with \emph{pūrakaṃ} and \emph{kuryāt} with \emph{recakaṃ}, but it is unlikely that the author intended this.
+\lb
 
 The compound \emph{vegodghoṣaṃ} (close to \emph{vegoghoṣaṃ} in \alphaOne) is rather unusual.
-We have understood it according to the commentator Bālakṛṣṇa’s gloss: `an inhalation in which sound is produced by force' (\emph{vegena sañjāta udghoṣo yasmin pūrake}).
-
-Witnesses of the \gamma \ and \delta \ branches of the stemma, and Brahmānanda, have \emph{ānandalīlā} instead of \emph{ānandamūrchā}, perhaps in order to avoid the repetition of \emph{mūrchā}, which is also found in the following verse.
+We have understood it according to the commentator Bālakṛṣṇa’s gloss: `an inhalation in which sound is produced by force' (\emph{vegena sañjāta udghoṣo yasmin pūrake}).\lb
+
+Witnesses of the \textgamma \ and \textdelta \ branches of the stemma, and Brahmānanda, have \emph{ānandalīlā} instead of \emph{ānandamūrchā}, perhaps in order to avoid the repetition of \emph{mūrchā}, which is also found in the following verse.
 
@@
 \begin{translation}[hp02_070]
-At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This fainting of the mind, which is called \emph{mūrchanā}, gives pleasure.
-\end{translation}
-%name is given as mūrchanā but headings in alpha etc. are mūrchā. Some, including epsilon1 do have mūrchanā.
-% adopt feminine in 2nd hemistich
+At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This loss of consciousness, which is called \emph{mūrchanā}, gives pleasure.
+\end{translation}
+%name is given as mūrchanā but headings in alpha etc. are mūrchā. Some, including zeta1 do have mūrchanā.
+% adopt feminine in 2nd line
 % JB: the meaning of manomūrcchā is not conveyed by the current translation.
@@
 bestows bliss, when the mind faints.
+JM: Meulenbeld 2011:127 (on the meaning of mūrch) "Mūrchita designates in literary Sanskrit, and also many times in medical Sanskrit, the state of having fainted/swooned, lost consciousness for a short time.
+Mūrchā means fainting."
 
@@
 \begin{philcomm}[hp02_071]
-The heading \emph{atha plāvanī} is not found in \alphaOne \ (which follows the description with \emph{plāvanīkumbhakaṃ}), but we have adopted it to be consistent with the other \emph{kumbhaka} descriptions.
-
-We have understood \emph{udgāramāruta} to refer to the breath of eructation, i.e.~the nāga breath as described in e.g.~\emph{Vivekamārtaṇḍa} 36.
+Although the important manuscript groups, including \textalpha, have this verse on \emph{plāvinī}, there is important evidence suggesting that \emph{plāvinī} was incorporated into the \emph{Haṭhapradīpikā} and its group of eight breath retentions (\emph{kumbhaka}) sometime after the text was initially composed.
+Firstly, in most manuscript groups, the name \emph{plāvinī} is absent in the list of \emph{kumbhaka}s in 2.44.
+The better attested reading, supported by \textalpha, is that the eighth \emph{kumbhaka} is \emph{kevala} and not \emph{plāvinī}.
+Most of the important testimonia, namely, the \emph{Haṭharatnāvalī}, \emph{Yogalakṣaṇāvalī} and \emph{Yuktabhavadeva}, also include \emph{kevalakumbhaka} and omit \emph{plāvinī} in the list of eight retentions.
+And secondly, the heading \emph{atha plāvanī} is not found in the \textalpha\ group as well as many other manuscripts.
+Instead, the \textalpha\ manuscripts have \emph{iti plāvinīkumbhakaṃ} after the verse on \emph{plāvinī}, which is not consistent with the other \emph{kumbhaka} descriptions.
+So although \emph{plāvinī} was introduced early in the tranmission of the \emph{Haṭhapradīpikā}, it may not have been in the text composed by Svātmārāma.\lb
+
+We have understood \emph{udgāramāruta} to refer to the breath of eructation, i.e.~the nāga breath as described in e.g.~\emph{Vivekamārtaṇḍa} 36.\lb
 % We have adopted \emph{payasy agādhe pi sukham} in pāda c because \emph{sukham} gives a better meaning and it is a bha-vipulā
 
-\getsiglum{V19}, \getsiglum{P28}, the \emph{Yogacintāmaṇi} and the \emph{Haṭharatnāvalī} omit this verse and accordingly do not mention \emph{plāvinī} in verse 2.44, substituting it with \emph{kevala}.
+%\getsiglum{V19}, \getsiglum{P28}, the \emph{Yogacintāmaṇi} and the \emph{Haṭharatnāvalī} omit this verse and accordingly do not mention \emph{plāvinī} in verse 2.44, substituting it with \emph{kevala}.
 
@@
 \begin{translation}[hp02_073]
-The [breath retention] that [the yogi] performs with exhalation and inhalation is \emph{sahita}. He should practice \emph{sahita} until \emph{kevala} is perfected.
+The [breath retention] which is done with exhalation and inhalation is the \emph{sahita} [breath retention]. [The yogi] should practice \emph{sahita} until \emph{kevala} is perfected.
 \end{translation}
@@
 \tl{yāvat kevalasiddhiḥ syāt tāvat sahitam abhyaset |\\+}
-\tl{\var{28a virecyāpūrya ] recyavāpūrya yat \emph{la}, ārecyāpūrya yaḥ \emph{śa}}\\!}
+\tl{\var{28a virecyāpūrya yaṃ ] recyavāpūrya yat \emph{la}, ārecyāpūrya yaḥ \emph{śa}}\\!}
 \end{versinnote}
@@
 \begin{philcomm}[hp02_073]
-We have adopted the \emph{Vasiṣṭhasaṃhitā}’s reading, which is very close to that of \deltaOne \ (\emph{ārecyāpūrya yat kuryāt}) and \deltaThree \ (\emph{ārecyāpūrya yaḥ kuryāt}) and similar to \alphaTwo\ (\emph{recapūrya y\emph{[}a\emph{]}t kāryaḥ}).
-It is the only one that makes sense of 2.73ab.
-It appears that the relative pronoun dropped out of the first verse quarter early in the transmission of the \emph{Haṭhapradīpikā} and scribes have tried in various unsuccessful ways to restore some sense.
+In the first \textit{pāda}, we have adopted a reading similar to that of the \emph{Vasiṣṭhasaṃhitā}, which is very close to that of \deltaOne \ (\emph{ārecyāpūrya yat kuryāt}) and \deltaThree \ (\emph{ārecyāpūrya yaḥ kuryāt}) and similar to \alphaTwo’s hypometrical reading (\emph{recapūrya y\emph{[}a\emph{]}t kāryaḥ}).
+It appears that the relative pronoun dropped out of the first verse quarter early in the transmission of the \emph{Haṭhapradīpikā} and scribes have tried in various ways to restore some sense, with several emending to \emph{recakaḥ pūrakaḥ kāryaḥ}.
 
@@
 
-% MD: Perhaps \emph{recakaḥ pūrakaḥ kāryaḥ} which is closer to the readings of N3 and V3?  Cf. also 2.74a \emph{recakaṃ pūrakaṃ muktvā}.
 % JB: We need the relative pronoun to pick up on the sa in the second pāda
 % MD: Not necessarily. Cf. 1.58, 2.24, 2.26.
+
+% Perhaps: recya vā pūrya vā kāryaḥ sa vai sahitakumbhakaḥ "The sahitakumbhaka is to be done either after exhalation or after inhalation." (Cf.
+Jyotsnā 2.71: tatra sahito dvividhaḥ/ recakapūrvakaḥ pūrakapūrvakaś ca/).
+That would be closer to the readings of ε1 and α2.
 \end{philcomm}
@@
 %The relative pronoun is omitted in V1 and the J10 group, which have \emph{vāyunirodhanam}.
-But the relative is needed to connect the description of kevalakumbhaka in the first hemistich to the last pāda (\emph{sa vai kevalakumbhakaḥ}).
+But the relative is needed to connect the description of kevalakumbhaka in the first line to the last pāda (\emph{sa vai kevalakumbhakaḥ}).
 On this verse, see Marcinkowska-Rosół and Sellmer 2020: 102–105.
@@
 \tl{kevale kumbhake siddhe recapūrakavarjite |\\+}
-\tl{na tasya durlabhaṃ kiñcit triṣu lokeṣuṃ vidyate ||\\!}
+\tl{na tasya durlabhaṃ kiñcit triṣu lokeṣuṃ vidyate ||\\!}    read  lokeṣu ? JH
 \end{versinnote}
@@
 \tl{na sidhyati tato yugmam āniṣpatteḥ samabhyaset\\+}
-\tl{\textup{[middle hemistich not in mss. I, III, IV, VII, IX, X, XII, XIV--XVI]}\\+}
+\tl{\var{[middle line not in mss. I, III, IV, VII, IX, X, XII, XIV--XVI]}\\+}
 \tl{tasmāt pravartate yogī haṭhe sadgurumārgataḥ ||\\!}
@@
 \begin{philcomm}[hp02_077]
-On the niṣpatti stage see 4.23–25.
+On the niṣpatti stage see 4.26–28.
 \end{philcomm}
@@
 \begin{versinnote}
-\tl{tad eva visadayati kumbhakam iti || kevalakumbhakābhyāsena cittaṃ dagdhaparṇavat nirvāsanaṃ bhavatītyarthaḥ || anyad vyākhyātam || 67 ||\\!}
+\tl{tad eva visadayati kumbhakam iti || kevalakumbhakābhyāsena cittaṃ dagdhaparṇavat nirvāsanaṃ bhavatītyarthaḥ || anyad vyākhyātam || 67 ||\\!}     %  read viśadayati  JH 
 \end{versinnote}
@@
 \begin{philcomm}[hp02_078]
-The reading \emph{kumbhitaḥ}, which is attested by \alphaOne\ and the testimonia, does not make sense here because the subject must be the yogi (\emph{prānaḥ} would be unmetrical).
+The reading \emph{kumbhitaḥ}, which is attested by \alphaOne\ and the testimonia, does not make sense here  % do not ? JH 
+because the subject must be the yogi (\emph{prānaḥ} would be unmetrical).
 In other texts, the word \emph{kumbhita} is used to qualify the breath and means `retained' (e.g., \emph{Yogabīja} 94, \emph{Kumbhakapaddhati} 127, \emph{Yuktabhavadeva} 8.32, etc.).
@@
 \tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet |\\+}
-\tl{anargalā suṣumnā ca haṭhasiddhiśca jāyate ||\\!}
+\tl{anargalā suṣumnā ca haṭhasiddhiś ca jāyate ||\\!}
 \end{versinnote}
@@
 \tl{nādasphuṭatvaṃ nayane ca nirmale |\\+}
-\tl{arogatā bindujayo'gnidīpanaṃ \\+}
+\tl{arogatā bindujayo 'gnidīpanaṃ \\+}
 \tl{nāḍīviśuddhir haṭhasiddhilakṣaṇam ||\\!}

### 2025-04-04 latex/HP3_comm.tex
new files

@@
 \begin{translation}[hp03_003]
-[\dots] the empty pathway becomes the royal highway for \emph{prāṇa}, the mind becomes free of support, and death is cheated.
+{[}\dots] the empty pathway becomes the royal highway for \emph{prāṇa}, the mind becomes free of support, and death is cheated.
 \end{translation}
@@
 %\begin{philcomm}[hp03_007]
+The reading of \textalpha\, \emph{idaṃ mudrādi°}, is not attested by the other manuscript groups but it makes sense in so far as 'locks' (\emph{bandha}) and 'actions' (\emph{karaṇa}) figure among the ten techniques taught in this chapter.
+It is also consistent with the reference to \emph{mudrādi} in 1.55.
+However, most manuscripts have readings, such as \emph{idaṃ hi mudrādaśakam}, that refer to the techniques of this chapter as only \emph{mudrā}s, and this is consistent with 3.104.
+
 %JB: hi (in 3.7c) is supported only by the \textit{Jyotsnā}. Mss. have tu and ca.
@@
 \begin{translation}[hp03_010a]
-Now the great seal (\emph{mahāmudrā}):
+Of these, the great seal (\emph{mahāmudrā}) [is now taught]: 
 \end{translation}
@@
 This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of Kuṇḍalinī in the central channel.
-Drawing on Hellwig 2009:238–240, Mallinson and Szanto (2021:21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert.
+Drawing on Hellwig 2009: 238–240, Mallinson and Szanto (2021: 21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert.
 In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when Kuṇḍalinī is said to be in the state of \emph{maraṇa} the meaning is that she is stilled.
-The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009:342, Mallinson and Szanto (2021:22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”.
+The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009: 342, Mallinson and Szanto (2021: 22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”.
 In the yoga of the \emph{Amṛtasiddhi}, the bodily \emph{saṃpuṭa} is formed by applying locks at the top and bottom of the central channel, i.e.~constricting the perineal region and the throat.
-In the \emph{Haṭhapradīpikā} it is formed by pressing the perineum with the heel and constricting the throat.
+In the \emph{Haṭhapradīpikā} it is formed by pressing the perineum with the heel and constricting the throat.\lb
 
@@
 Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvi\-puṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}.
-Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened Kuṇḍalinī destroys death, which is usually dependent on the in and out breaths.
+Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened Kuṇḍalinī destroys death, which is usually dependent on the in and out breaths.\lb
+
+The form \textit{ṛjvībhūtā} is non-Paninian (it should be \textit{ṛjūbhūtā}) but all witnesses have \textit{ṛjvī} or variants thereof.
+\alphaOne\ has \textit{ṛjvī bhūtvā} which is correct morphologically, but it does not make sense with \textit{bhavet} in the next \textit{pāda} and may be an attempt at correction by the redactor.
 
@@
 We also have to bear in mind that there's no evidence that Svātmārāma and later commentators knew the AS.
-In the one case where there is some evidence (i.e., Śivānanda quotes it in the YCM), it doesn't seem to have helped the author understand the alchemical metaphor in this instance (as Śivānanda rewrote the verse to say that kuṇḍalinī destroys the state of death).
+In the one case where there is some evidence (i.e., Śivānanda quotes it in the YCM), it doesn't seem to have helped the author understand the alchemical metaphor in this instance (as Śivānanda rewrote the verse to say that kuṇḍalinī destroys the state of death).JM:but the commentators etc.
+are all a lot later than the HP.
+I think it's better to give Svātmārāma the benefit of the doubt.
+I don't think he would have used the verse if he didn't understand it correctly — as you (JB) point out in the note the later attempts to understand maraṇa in a non-alchemical way are unconvincing.
 
@@
 \begin{translation}[hp03_014]
-Problems such as the great afflictions [and] death and so forth are removed [by it], and that is why the most wise call it the `great seal.'
-\end{translation}
+Problems such as the great afflictions [and] death and so forth dissolve, and that is why the most wise call it the `great seal.'
+\end{translation}
+% MD: Now I prefer jīryante to hīyante.
+The latter is suppoted only α1 and α2, may be an outlier.
+As an explanation of the name mahāmudrā, √jṝ seems to make more sense.
+The kleśa etc.
+decay after being confined (mudrita) or so.
+done
 
@@
 \begin{philcomm}[hp03_018]
-Manuscripts of three groups (\textalpha, \textbeta\ and \textgamma), which are important stemmatically, have the reading \emph{jarāmṛtyuvināśinī} (and \getsiglum{J5} has \emph{nṛṇāṃ mṛtyuvināśinī}) for the second verse quarter.
-While this reading is possible, the play on \emph{mahāsiddhi} and \emph{mahāmudrā} seems more likely original, as seen in the source text, the \emph{Vivekamārtaṇḍa} (without significant variants), and the \texteta\ and \textepsilon\ groups.
+Two readings of the second quarter are well-attested: \emph{jarāmṛtyuvināśinī} (\alphaOne, etc.) and \emph{mahāsiddhikarī nṛṇām} (\alphaThree, etc.).
+While the first reading is possible, the play on \emph{mahāsiddhi} and \emph{mahāmudrā} seems more likely original, as seen in the source text, the \emph{Vivekamārtaṇḍa} (without significant variants).\lb
+
+%Manuscripts of three groups (\textalpha, \textbeta\ and \textgamma), which are important stemmatically, have the reading \emph{jarāmṛtyuvināśinī} (and \alphaTwo\ has \emph{nṛṇāṃ mṛtyuvināśinī}) for the second verse quarter.
+While this reading is possible, the play on \emph{mahāsiddhi} and \emph{mahāmudrā} seems more likely original, as seen in the source text, the \emph{Vivekamārtaṇḍa} (without significant variants), and the \texteta\ and \textzeta\ groups.
+\lb %  check the stemma tree -- α1+α2 vs α3
 
@@
 It could also plausibly mean the region between the chest and perineum, or perhaps the place between the eyebrows.
-At 3.24 this practice is said to make the mind reach Kedāra, which is sometimes located between the eyebrows (see Mallinson 2007:214 n.285, Birch 2019: 967 n.57).
+At 3.24 this practice is said to make the mind reach Kedāra, which is sometimes located between the eyebrows (see Mallinson 2007: 214 n.285; Birch 2019: 967 n.57).
 \end{philcomm}
@@
 \end{translation}
+% MD: pāda a: adopt yathāśaktyā as in 2.7?
+% pāda c: dakṣāṅge ca samabhyaset or dakṣāṅge punar abhyaset instead of dakṣiṇāṅge samabhyaset?
 
@@
 This omission results in the somewhat odd juxtaposition of two lines stating that this is \emph{mahābandha}.
+\lb
 
@@
 This verse expresses an alternative to the application of the chin lock in \emph{mahā\-bandha} (mentioned in 3.20).
-It is found in all groups but is notably absent in \alphaOne\ and \alphaTwo, and the \textbeta\ group.
+It is found in all groups but is notably absent in \alphaOne\ and \alphaTwo, and the \textdelta\ group.
 At present we cannot be certain if this verse was written by Svātmārāma because it does not reflect textual teachings on \emph{mahābandha} in so far as no other text known to us advocates the use of the tongue rather than the chin lock in \emph{mahābandha}.
-If the verse was composed by Svātmārāma, it was omitted early in the transmission by someone who did not agree with the alternative teaching.
+If the verse was composed by Svātmārāma, it was omitted early in the transmission by someone who did not agree with the alternative teaching.\lb
 
@@
 The other verses of the quoted passage are found in the \emph{Śivasaṃhitā} (4.37–42), but the verse in question is not reported in the critical edition of the \emph{Śivasaṃhitā} (2018).
-The verse is absent in another passage on \emph{mahābandha} that the author of the \emph{Yogacintāmaṇi} cites and attributes to the \emph{Haṭhapradīpikā}.
+The verse is absent in another passage on \emph{mahābandha} that the author of the \emph{Yogacintāmaṇi} cites and attributes to the \emph{Haṭhapradīpikā}.\lb
 
@@
 While the \emph{Amaraughaprabodha} has the reading \emph{ūrdhvaṃgativiśodhanaḥ} (`purification of the upward flow'), the \emph{Amṛtasiddhi} (12.14) states that the chin-lock prevents the upward flow (\emph{ūrdhvaṃgatinirodhakaḥ}) in all the channels.
-The \textit{Amaraugha} is referring to the idea (likely accepted by Svātmārāma) that the root lock creates an upward flow in all the channels that prevents the elements and essences of the body from escaping (cf.~\emph{Amṛtasiddhi} 12.8–10).
+The \textit{Amaraugha} is referring to the idea (likely accepted by Svātmārāma) that the root lock creates an upward flow in all the channels that prevents the elements and essences of the body from escaping (cf.~\emph{Amṛtasiddhi} 12.8–10).\lb
 
@@
 Brahmānanda, who does not identify a location for Triveṇī, understands Kedāra to be between the eyebrows.
-In the \emph{Khecarīvidyā} it is located on the back of the head above the nape of the neck (Mallinson 2007:214 n.285).
-For other references on the location of Kedāra, see Birch 2019:967 n.57.
+In the \emph{Khecarīvidyā} it is located on the back of the head above the nape of the neck (Mallinson 2007: 214 n.285).
+For other references on the location of Kedāra, see Birch 2019: 967 n.57.
 \end{philcomm}
@@
 \begin{philcomm}[hp03_027]
-The term \emph{puṭadvaya} is referring back to the \emph{dvipuṭa} mentioned in verse 3.12. On the alchemical meaning and other interpretations by later commentators, see the note to 3.12.
+The term \emph{puṭadvaya} is referring back to the \emph{dvipuṭa} mentioned in verse 3.12. On the alchemical meaning and other interpretations by later commentators, see the note to 3.12.\lb
 
@@
 The \textalpha\ and \textgamma\ groups have °\emph{saṃbandhāj}, which is possible but somewhat awkward (i.e., `because of the union of the moon, sun and fire, the state of a dead person, which has arisen, leads to immortality').
-The adopted reading °\emph{saṃbandho}, which is supported by \epsilonTwo, \emph{Yuktabhavadeva} 7.200 and \emph{Jyotsna} 3.28, makes better sense but its meaning is not as clear as the formulation in the source text (i.e., \emph{Amaraugha} 30).
+The adopted reading °\emph{saṃbandho}, which is supported by \zetaTwo, \emph{Yuktabhavadeva} 7.200 and \emph{Jyotsnā} 3.28, makes better sense but its meaning is not as clear as the formulation in the source text (i.e., \emph{Amaraugha} 30).\lb
 
 The compound \emph{mṛtāvasthā} (`the state of death') likely refers to a dead person in the sense that the yogi appears as if dead when the moon, sun and fire have united.
-In the version found in \emph{Jyotsnā} 3.28 and \etaOne, \etaTwo\ and \epsilonThree, the final verse quarter has been rewritten to say that the yogi then exhales the breath (\emph{tato vāyuṃ virecayet}).
-This implies that \emph{mṛtāvasthā} is a breath retention (\emph{kumhaka}), which is apparent in Brahmānanda's explanation:
+In the version found in \emph{Jyotsnā} 3.28 and \etaOne, \etaTwo\ and \zetaThree, the final verse quarter has been rewritten to say that the yogi then exhales the breath (\emph{tato vāyuṃ virecayet}).
+This implies that \emph{mṛtāvasthā} is a breath retention (\emph{kumbhaka}), which is apparent in Brahmānanda's explanation:
 \begin{quote}
 The state of death that has arisen is the state of one who has died, [that is,] of one who is devoid of the breath because of the absence of movement of \emph{prāṇa} in the \emph{iḍā} and \emph{piṅgalā} channels.
-Then, immediately after that [state], [the yogi] exhales the breath, [that is,] he gradually releases it through the nostrils.\\
+Then, immediately after that [state], [the yogi] exhales the breath, [that is,] he gradually releases it through the nostrils.
+\end{quote}
+\begin{quote}
 \emph{mṛtasya prāṇaviyuktasyāvasthā mṛtāvasthā samutpannā bhavati, iḍāpiṅgalayoḥ prāṇasañcārābhāvāt} | \emph{tatas tadanantaraṃ vāyuṃ virecayen nāsikāpuṭābhyāṃ śanais tyajet} |
@@
 \begin{translation}[hp03_029]
-Through practice, this great piercing bestows the great \emph{siddhi} [and] cures wrinkles, grey hair and trembling. It is used by the best practitioners.
+Through practice, this great piercing bestows the great \emph{siddhi} [and] cures wrinkles, grey hair, and trembling. It is used by the best practitioners.
 \end{translation}
@@
 \begin{philcomm}[hp03_029]
-In the third verse quarter, the reading \emph{°vedhaghnaḥ} (`cures wounds' ?) in \textalpha\ and the other groups, the exception being \textgamma\ (\emph{°vegaghnaḥ}), is strange as it does not seem related to the other two symptoms of old age (i.e., wrinkles and grey hair) that this \emph{mudrā} can cure.
-It is perhaps a misreading of \emph{°vepaghnaḥ}, which occurs in some manuscripts in the \textbeta\ (\getsiglum{J1}, \getsiglum{N12}), \textepsilon\ (\getsiglum{J14}) and \textdelta\ (\getsiglum{J3}, \getsiglum{N16}, \getsiglum{N18}) groups, as well as the \emph{Yogacintāmaṇi} and \emph{Jyotsnā}.
-The meaning of \emph{vepa}, `trembling,' is consistent with wrinkles and grey hair.
+In the third verse quarter, the reading \emph{°vedhaghnaḥ} (`cures wounds'?) in \textalpha\ and the other groups, the exception being \textgamma\ (\emph{°vegaghnaḥ}), is strange as it does not seem related to the other two symptoms of old age (i.e., wrinkles and grey hair) that this \emph{mudrā} can cure.
+It is perhaps a misreading of \emph{°vepaghnaḥ},
+%which occurs in some manuscripts in the \textbeta\ (\getsiglum{J1}, \getsiglum{N12}), \textzeta\ (\getsiglum{J14}) and \textdelta\ (\getsiglum{J3}, \getsiglum{N16}, \getsiglum{N18}) groups, 
+which occurs in some manuscripts in the \deltaOmega\ (\getsiglum{J1}, \getsiglum{J3}, \getsiglum{N16}, \getsiglum{N18}) and \textbeta\ (\getsiglum{N12} contaminated) and \textzeta\ (\getsiglum{J14} contaminated) groups, % by MD
+as well as the \emph{Yogacintāmaṇi} and \emph{Jyotsnā}. The meaning of \emph{vepa}, `trembling,' is consistent with wrinkles and grey hair.
 
 %The collocation \emph{valīpalitaroga} that we have adopted, found in \getsiglum{C8}, is quite common (e.g. \emph{Śāradātilaka} \emph{6.104, Ānandakanda} 1.7.13, .1.7.31).
-%JB: I think we should adopt vepa on the strength of the marmasthāna entry (\item[valīpalitavepaghnaḥ] Ba5, J1, J3, J6, J14, N4a, N11, N12, N16, N18, ), and the fact that it's closer to vedha (alpha reading) than roga (which looks like an outlier in the marmasthāna data).
-% J1=6b, J3,J6,N16,N18=6a, J14=4c, N11 =Gr2, N12 =4b,
+%JB: I think we should adopt vepa on the strength of the marmasthāna entry (\item[valīpalitavepaghnaḥ] Ba5, J1, J3, J6, J14, N4a, N11, N12, N16, N18), and the fact that it's closer to vedha (alpha reading) than roga (which looks like an outlier in the marmasthāna data).
+% J1,J3,J6,N16,N18=6a, J14=4c, N11 =Gr2, N12 =4b,
 % MD: These mss are not very strong.
 Is it not possible to interpret °vedhaghnaḥ as "destroys by piercing"?
-This verse only has no source in the middle of the long citation from Amaraugha, is probably authorial.
+(like daṇḍaghna "striking with a stick" in the Manusmṛti) This verse only has no source in the middle of the long citation from Amaraugha, is probably authorial.
 Svātmārāma might have wanted to use the word vedha of mahāvedha, similar to dhauti/dhāvanti.
@@
 The emendation of \emph{tadānīṃ} to \emph{tadā} renders the verse an \emph{āryā}.
-The word \emph{krameṇa} is well attested by manuscripts of the \textalpha, \textbeta, \texteta\ and  \textepsilon\ groups, which all have unmetrical or corrupted versions.
+The word \emph{krameṇa} is well attested by manuscripts of the \textalpha, \textbeta, \texteta\ and  \textzeta\ groups, which all have unmetrical or corrupted versions.\lb
 % krameṇa is well attested
@@
 %%%%%%%%%%
-\subsection*{3.34*1}
-\begin{translation}[hp03_034_1]
+\subsection*{3.35}
+\begin{translation}[hp03_035]
 If the yogi turns back the tongue and remains [like that] for half an instant, he is instantly freed from disease, death, old age and the like.
@@
 
-\begin{sources}[hp03_034_1]
+\begin{sources}[hp03_035]
 \emph{Śivasaṃhitā} 3.91
@@
 
-\begin{testimonia}[hp03_034_1]
+\begin{testimonia}[hp03_035]
 \emph{Yogacintāmaṇi} f.~74r (attr.~to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp03_034_1]
-This verse is absent in \alphaOne\ and was likely added to a block of verses that Svātmārāma borrowed from the \textit{Vivekamārtaṇḍa} (47–51).
-3.34*1 is in \alphaTwo\ and \alphaThree\ (\alphaOne\ omits 2.35 and the greyscaled verses) and also the \textgamma\ and \texteta\ groups.
+\begin{philcomm}[hp03_035]
+Although 3.35 is absent in \alphaOne, it is in \alphaTwo\ and \alphaThree, and also the \textgamma\ and \texteta\ groups.
 Other manuscripts have an additional line that gives \emph{vyomacakra} as an alternative name for \emph{khecarīmudrā}.
@@
 JM: seems good to me.
-% Pāda  α2 - α3
-% e: nītvā - kṛtvā*
-% f: api - yadi*
-% g: viṣaye - kṣaṇena*
-% The asterisk indicates that the reading has been adopted for our edition.
-
+
+% MD: Sorry, now I want to reactivate the verse, since α4 has the verse too:
+% kalāṃ tu prāṅmukhī nītvā kṣaṇārddhaṃ <ya>di tiṣṭhati
+% kṣaṇena mucyate vyādhi-jarāmṛtyumayādikaiḥ
+% Fortunately, I have not yet changed the verse numbers.
 %%%%%%%%%%
@@
 \begin{philcomm}[hp03_037]
-The form of \emph{lipyati}, which is found in the \alpha \ manuscripts and has been adopted, is widely attested in epic Sanskrit.
+The form of \emph{lipyati}, which is found in the \textalpha\ manuscripts and has been adopted, is widely attested in epic Sanskrit.
 \end{philcomm}
@@
 Most of the collated manuscripts, including \alphaTwo\ (\alphaOne\ is illegible here and \alphaThree\ is missing this verse quarter), have the reading \emph{haṭhāt śaktyā}.
-This reading only makes sense if one infers that \emph{śaktyā} is referring to \emph{khecarīmudrā}, so that the second line means `blocked by \emph{yonimudrā}, semen goes up forcefully by the power [of \emph{khecarīmudrā}].' The reading \emph{hi tacchaktyā} of two delta manuscripts (\getsiglum{K3} and \getsiglum{C7}) and the \emph{Yogacintāmaṇi}), appears to be an attempt to render more clearly the meaning `by the power of \emph{khecarī}.' Such an interpretation suggests that \emph{yonimudrā} blocks \emph{bindu}'s downward course and \emph{khecarī} causes it to go upwards forcefully.
-
-A manuscript of an early recension of the \emph{Vivekamārtaṇḍa} (ms.~no.~4110) has \emph{hataḥ śaktyā}, which is attested by three \emph{Haṭhapradīpikā} manuscripts on lower branches of the stemma (i.e.~\getsiglum{Ba} of \textdelta\ group and \getsiglum{C2} and \getsiglum{P4} of group 7a).
-The participle \emph{haṭaḥ} makes sense of the instrumental \emph{śaktyā}, rendering the meaning `struck by Kuṇḍalinī.'
-% MD: These groups don't have a Greek sigla yet.
-
-Alternatively, the word \emph{haṭaḥ} (as well as the other variants \emph{kṛtaḥ, kṛte}, and even \emph{haṭhāt}) may derive from \emph{hṛtaḥ}, which is attested by \emph{Haṭhapradīpikā} manuscripts from group 4a (i.e., \getsiglum{Ba5}, \getsiglum{N4a} and \getsiglum{N13}).
+This reading only makes sense if one infers that \emph{śaktyā} is referring to \emph{khecarīmudrā}, so that the second line means `blocked by \emph{yonimudrā}, semen goes up forcefully by the power [of \emph{khecarīmudrā}].' The reading \emph{hi tacchaktyā} of \deltaTwo\ and the \emph{Yogacintāmaṇi}), appears to be an attempt to render more clearly the meaning `by the power of \emph{khecarī}.' Such an interpretation suggests that \emph{yonimudrā} blocks \emph{bindu}'s downward course and \emph{khecarī} causes it to go upwards forcefully.\lb
+
+A manuscript of an early recension of the \emph{Vivekamārtaṇḍa} (ms.~no.~4110) has \emph{hataḥ śaktyā}, which is attested by three \emph{Haṭhapradīpikā} manuscripts on lower branches of the stemma (i.e.~\getsiglum{Ba}, % MD: this ms does not belong to any group
+\getsiglum{C2} and \getsiglum{P4}). The participle \emph{hataḥ} makes sense of the instrumental \emph{śaktyā}, rendering the meaning `struck by Kuṇḍalinī.' \lb
+
+Alternatively, the word \emph{hataḥ} (as well as the other variants \emph{kṛtaḥ, kṛte}, and even \emph{haṭhāt}) may derive from \emph{hṛtaḥ}, which is attested by manuscripts of the \emph{Jyotsnā} (i.e., \getsiglum{Ba5}, \getsiglum{N4a} and \getsiglum{N13}).
+% MD: These are Jyotsnā mss.
 The reading \emph{hṛtaḥ śaktyā} renders the verse as saying that semen goes up, carried by Kuṇḍalinī.
@@
 In this work, \emph{yonimudrā} is supposed to awaken Kuṇḍalinī and consists of sitting in \emph{siddhasana}, blocking the ears, eyes, nose and mouth with the thumbs, index, middle and ring fingers, uniting \emph{praṇa} and \emph{apāna}, meditating on the six \emph{cakra}s and repeating the mantra \emph{huṃ haṃsa}.
+\lb
 
@@
 \begin{philcomm}[hp03_042]
-In the context of poison, \emph{takṣaka} refers to one of the three kings of the snakes (\emph{nāga}), the other two being Śeṣa and Vāsuki (Mani 1975: 782–783).
-% Purāṇic Encyclopaedia: A Comprehensive Dictionary with Special Reference to the Epic and Purāṇic Literature.
+In the context of poison, \emph{takṣaka} refers to one of the three kings of the snakes (\emph{nāga}), the other two being Śeṣa and Vāsuki (Mani 1975: 782–783).\lb % Purāṇic Encyclopaedia: A Comprehensive Dictionary with Special Reference to the Epic and Purāṇic Literature.
 Vettam Mani.
@@
 \begin{translation}[hp03_045]
-By the word 'cow' is meant the tongue, for its insertion into the palate is the eating of cow's flesh. That [eating] is the destroyer of great sin.
+By the word `cow' is meant the tongue, for its insertion into the palate is the eating of cow's flesh. That [eating] is the destroyer of great sin.
 \end{translation}
@@
 In \emph{Jyotsnā} 3.51, Brahmānanda says that there is a variant \emph{prāṇaiḥ} (`by means of the breaths'), which is easier to understand than \emph{prāṇāt}.
-He nonetheless accepts \emph{prāṇat} and understands it as being a means (\emph{prāṇāt sādhanabhūtād avāptam}).
+He nonetheless accepts \emph{prāṇāt} and understands it as being a means (\emph{prāṇāt sādhanabhūtād avāptam}).
 He also understands the sixteen-petalled lotus to be the lotus in the throat, into which the nectar drips.
-%Adopt patrapadma. % MD: VM-T has also padmapattra...
-\end{philcomm}
-
-% MD(2024-5-14): adopt vapur (α1,β1,ε2 etc.) instead of tanur (α2,β2,βω etc.)?
+%Adopt patrapadma.
+% MD: The important mss and also VM-T have padmapattra.
+Perhaps we should adopt this.
+We can dissolve the compound into ṣoḍaśāni padmapattrāṇi, not necessarily ṣoḍaśānāṃ padmānāṃ pattrāṇi.
+\end{philcomm}
+
 \begin{metre}[hp03_047]
@@
 \subsection*{3.49*2}
-\begin{translation}[hp03_049_2]
-That which enters the aperture into the underworld exists at the base of Meru.
+% \begin{translation}[hp03_049_2]
+% That which enters the aperture into the underworld exists at the base of Meru.
 The wise [yogi] says that is the truth, the source of [all] rivers.
@@
 Bodily perfection [arises] no other way.
-\end{translation}% greyscale in ed.
-
-%\begin{sources}[hp03_049_2]
-%\end{sources}
-
-\begin{testimonia}[hp03_049_2]
-\emph{Haṭharatnāvalī} 4.30
-\begin{versinnote}
-\tl{pātāle yad viśati suṣiraṃ merūmūlaṃ tad asti\\+}
-\tl{tattvaṃ caitad vadati sudhā tanmukhaṃ nimnagānām |\\+}
-\tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
-\tl{tad badhnīyāt sukharatimṛdur nānyathā kāryasiddhiḥ ||\\!}
-\end{versinnote}
-% after (ekaṃ sṛṣṭimayaṃ + mano yatra vilīyeta) and before (śarīraṃ tāvad eva)
-%the context of this verse in the HR is rājayoga.
-
-\emph{Yogacintāmaṇi} f.~75r (attr.~to the \emph{Haṭhapradīpikā})
-\begin{versinnote}
-\tl{tat pātālād viyati śikhare merumūle tad asti \\+}
-\tl{tattvaṃ caitat pravadati sudhīḥ saṃmukhe nimnagānām |\\+}
-\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
-\tl{tad badhnīyāt svakaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
-\end{versinnote}
-% after the verse (cittāyattaṃ nṛṇāṃ śukraṃ) and before (indhanāni yathā vahnis)
-% context is khecarī (3rd chapter material)
-
-\emph{Yuktabhavadeva} 7.220 (attr.~to Gorakṣanātha)
-\begin{versinnote}
-\tl{pātāle yad viśati suṣiraṃ merumūle yad asti \\+}
-\tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
-\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
-\tl{badhnīyāt tat sukaraṇam atho nānyathā kāyasiddhiḥ ||\\!}
-\end{versinnote}
-% context is khecarī, (3rd chapter material)
-% after (ekaṃ sṛṣṭimayaṃ bījaṃ) and before (atha uḍḍiyānabandhaḥ)
-
-\emph{Haṭhatattvakaumudī} 14.26 (attr.~to the \emph{Haṭhapradīpikā})
-\begin{versinnote}
-\tl{pātāle yad vitatasuṣiraṃ merumūle tad asmin \\+}
-\tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
-\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
-\tl{taṃ badhnīyāt sukaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
-\end{versinnote}
-% context is khecarī 3rd chapter material
-% after (cumbantī yadi lambikā) and before (cittaṃ vicarati gagane)
-
-\end{testimonia}
+% \end{translation}% greyscale in ed.
 
 \begin{philcomm}[hp03_049_2]
-%Workshop notes
-% Sven: the first line could refer to Kuṇḍalinī
-% J5 has this verse here.
-% mṛdā = using clay to make a dam?
-% Also see Netratantra 7.32
-% Could this verse be about mūlabandha? (reading pātāle and mūle)
-%J5: yat prāleyam pihitasukhire merumūle yad astī — this seems helpful.
-% PN has a nice-ish reading: yat prāleyaṃ pihita sukhiraṃ merumūrdhnaṃ. mūrdhni etc. is much better.
-% J10 has merumūrdhni sthitaṃ yat, which is good.
-% If going with prāleyam, adopt yat prāleyam pihitasuṣiraṃ merumūrdhny asti tathyaṃ (~K1 Grp2)
-% ... tasmin tattvaṃ pravadati
-%Haru: Perhaps, praleyam was corrupted to pātāle and then the rest of the line changed.
-
-This verse occurs in various places in the different recensions of the text.
-It is here in the \alphaTwo\ and \alphaThree\ manuscripts, in chapter four (4.8*3) in \alphaOne\ and the delta group, after 3.39 in \betaOne\ and \betaTwo, and in both the third and fourth chapters in the \textgamma\ manuscripts.
-It must have entered the transmission of the \emph{Haṭhapradīpikā} at an early stage, and has moved around, perhaps because the name of the technique alluded to is not stated and the meaning of the first half of the verse is somewhat vague without its original context (we are yet to identify a source).
-
-The first verse quarter has many variants among the witnesses. \alphaTwo\ has:
-\begin{versinnote}
-\tl{yat prāleyaṃ pihitasukhire merumūle yad astī \\+}
-\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\!}
-\end{versinnote}
-This version of the verse is close to that of other important manuscripts in the \textbeta, \textgamma\ and \texteta\ groups.
-With the help of these manuscripts (in particular \getsiglum{K1} of the \textgamma\ group), the above can be emended and understood as follows:
-\begin{versinnote}
-\tl{yat prāleyaṃ pihitasuṣiraṃ merumūrdhny asti tathyaṃ \\+}
-\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\+}
-\tl{\var{°suṣiraṃ ] \getsiglum{K1}, °sukhire \alphaTwo. °mūrdhny ] \getsiglum{P8}. °mūle \alphaTwo. asti tathyaṃ ] \getsiglum{K1}, yad astī \alphaTwo.}\\!}
-“That cool liquid by which the aperture is filled at the top of Meru and exists as the truth, the wise [yogi] says that is the source of [all] rivers.” 
-\end{versinnote}
-
-%K1 in Mitsuyo's notes
-%yat prāleya prahitasuṣiraṃ merumūrdhnāsti tathyaṃ
-%tasmiṃs tatvaṃ pravadati sudhīs tan mukhaṃ nimnagānāṃ
-
-% Other manuscripts have something more similar to:
-% \begin{versinnote}
-% \tl{pātāle yad viśati suṣiraṃ merumūlaṃ tad asti \\+}
-% \tl{tattvaṃ caitat pravadati sudhīs tan mukhaṃ nimnagānām\\!}
-% \end{versinnote}
-
+For the translation and testimonia, as well as an explanation of the various places and versions of this verse in the text, see \manuref{4.9}. 
 \end{philcomm}
@@
 % JB: the back part of the heel presses the perineum. It would be better to say, 'with part of the heel.'
+% MD: α1 and β-mss have 'yam iṣyate in place of 'yam ucyate.
 
@@
 %\begin{philcomm}[hp03_051]
-%Adopt balāt (N3). haṭhāt seems to be an early alt. reading.
-%MD: Text changed.
+% MD: pāda d: hi is better attested than tu. 
 %\end{philcomm}
@@
 The instruction to `contract the wind' (\emph{vāyum ākuñcayet}) is odd (especially with \emph{samīraṇaḥ} in the fourth \emph{pāda}) and not found in other texts.
-Mallin\-son has adopted \emph{yonim} for \emph{vāyum} (cf.~\emph{Haṭhapradīpikā} 3.50d) in his edition of this verse in its source text, the \emph{Dattātreyayogaśāstra}, which is not found in the manuscripts of that text but is in the \emph{Haṭhapradīpikā}’s \delta \ manuscripts.
+Mallin\-son has adopted \emph{yonim} for \emph{vāyum} (cf.~\emph{Haṭhapradīpikā} 3.50d) in his edition of this verse in its source text, the \emph{Dattātreyayogaśāstra}, which is not found in the manuscripts of that text but is in the \emph{Haṭhapradīpikā}’s \textdelta \ manuscripts.
+\lb
 
@@
 \begin{translation}[hp03_056]
-As a result, fire and \textit{apāna} reach \textit{prāṇa}, which is hot by nature, and the \emph{prāṇa} makes the fire in the body extremely hot.
-\end{translation}
+As a result, fire and \textit{apāna} reach \textit{prāṇa}, which is hot by nature. Then [\emph{prāṇa}] makes the fire in the body extremely hot.
+\end{translation}
+
+% MD: tathā or tadā in pāda d?
 
@@
 \begin{philcomm}[hp03_056]
-The second verse quarter has been rewritten in the \textalpha\ manuscripts as \emph{prāṇamūlasvarūpakam}, which is similar to \etaOne\ (\emph{prāṇamūlasvarūpakau}) and the \emph{Haṭharatnāvalī} (\emph{mūlarūpasvarūpakau}).
+The second verse quarter has been rewritten in \alphaOne\ and \alphaTwo\ as \emph{prāṇamūlasvarūpakam}, 
+% \alphaOne\ and \alphaTwo\ only; \alphaThree\ has uṣṇa.
+which is similar to \etaOne\ (\emph{prāṇamūlasvarūpakau}) and the \emph{Haṭharatnāvalī} (\emph{mūlarūpasvarūpakau}).
 Manuscripts from the delta group, as well as the \emph{Yogacintāmaṇi}, have \emph{prāṇam uktasvarūpakam} or \emph{prāṇam uktasvarūpakau}.
-It appears that the intention behind these rewrites was to avoid the reading in the source text, `\textit{prāṇa} is hot by nature' (\emph{prāṇam uṣṇasvarūpakam}), which was likely accepted by Svātmārāma because it is in some witnesses of the \textbeta, \textgamma\ and \textepsilon\ groups.
+It appears that the intention behind these rewrites was to avoid the reading in the source text, `\textit{prāṇa} is hot by nature' (\emph{prāṇam uṣṇasvarūpakam}), which was likely accepted by Svātmārāma because it is in some witnesses of the \textbeta, \textgamma\ and \textzeta\ groups.
 References to \emph{prāṇa} being hot by nature (and \emph{apāna} being cold) occur in other works, such as \emph{Mokṣopāya} 6.85.111-112 and \emph{Haṭhatattvakaumudī} 4.14, 41.2.
@@
 \begin{translation}[hp03_060]
-Because the great bird tirelessly flies up (\emph{uḍḍīnaṃ kurute}), it is [called] 'flying up'. In it, the [root] lock is applied.
+Because the great bird tirelessly flies up (\emph{uḍyāṇaṃ kurute}), it is [called] `flying up'. In it, the [root] lock is applied.
 \end{translation}
@@
 \begin{versinnote}
-\tl{uḍḍīṇaṃ kurute yasmād aviśrāntaṃ mahākhagaḥ |\\+}
+\tl{uḍyāṇaṃ kurute yasmād aviśrāntaṃ mahākhagaḥ |\\+}
 \tl{uḍḍiyānaṃ tad eva syāt tatra bandho vidhīyate ||\\+}
@@
 In the \emph{Vivekamārtaṇḍa} this verse is preceded by a passage on \emph{mūlabandha}, so the likely meaning of the fourth verse quarter is that (\emph{mūla})\emph{bandha} is to be performed in this practice.
-The \textbeta\ group and many other manuscripts of the \emph{Haṭhapradīpika} have `the root [lock] is applied' (\emph{mūlaṃ vidhīyate}), which appears to be an attempt to clarify the meaning of the original verse.
-\end{philcomm}
-
+The \textdelta\ group of the \emph{Haṭhapradīpikā} witnesses has `the root [lock] is applied' (\emph{mūlaṃ vidhīyate}), which appears to be an attempt to clarify the meaning of the original verse.\lb
+
+The word \emph{uḍyāṇaṃ} in the first \emph{pāda} is attested in manuscripts of the \emph{Vivekamārtaṇḍa}, the source of this verse, and two manuscripts of the \textalpha\ group.
+This spelling is attested in several vernacular works (e.g., \emph{Aṣṭāṅgayoga} of Caraṇadāsa 144, 275, 307–8, \emph{Jogapradīpyaka} 588, 635, 686, etc.), as well as other Sanskrit yoga texts (e.g., \emph{Yogatārāvalī} 6, \emph{Dhyānabindūpaniṣat} 75, etc.).
+Other witnesses, including the \emph{Jyotsnā}, have \emph{uḍḍīnaṃ}, the usual form for the verbal noun from \emph{uḍ-ḍī}.
+
+%There is considerable variation among the witnesses with regard to the spelling of \emph{uḍḍīnaṃ} in \emph{pāda} a.
+We have adopted \emph{uḍḍīnaṃ} because it is the correct form for the verbal noun from \emph{uḍ-ḍī}, as widely attested elsewhere.
+\end{philcomm}
+
+% MD: Or is uḍyāṇaṃ a wrong Sanskritization of the vernacular uḍḍāṇaṃ (= Skt. uḍḍayana)?
+
+% uḍyāna is attested in the Aṣṭāṅgayoga of Caraṇadāsa 144, 275, 307–8
+% Jogapradīpyaka: uḍyāṇa:  588, 635, 686
+% Yogatattvabindu V: uḍyānapīṭha°
 %%%%%%%%%%
@@
 \begin{translation}[hp03_062]   
-\emph{Uḍḍiyāṇa} is easy, but it is always taught by a guru. If he practises it tirelessly, even an old man becomes young.%
-\end{translation}
+\emph{Uḍḍiyāṇa} is easy, but it is always taught by a guru. Even an old person becomes young if they practise it tirelessly.%
+\end{translation}
+% MD: In view of the position of the second tu, wouldn't it be better to separate the sentences after abhyaset?
 
@@
 \begin{philcomm}[hp03_062]
-At the end of the second \emph{pāda} \emph{sadā} is attested by witnesses of \textalpha, \texteta, \textepsilon\ and \delta \ groups, as well as the \emph{Dattātreyayogaśāstra}.
+At the end of the second \emph{pāda} \emph{sadā} is attested by witnesses of \textalpha, \texteta, \textzeta\ and \delta \ groups, %??
+all but the gamma/delta groups; JM: but do we need to mention this?
+as well as the \emph{Dattātreyayogaśāstra}.
 We have understood the first line to mean that even though the basics of the practice of \emph{uḍḍiyāna} are easy, it still needs to be taught by the guru.
-Some witnesses read \emph{yathā} for \emph{sadā}, perhaps as a deliberate substitution of the more difficult \emph{sadā}, making the verse mean that \emph{uḍḍiyāna} is easy in the way that is taught by the guru.
+Some witnesses read \emph{yathā} for \emph{sadā}, % = gamma/delta
+perhaps as a deliberate substitution of the more difficult \emph{sadā}, making the verse mean that \emph{uḍḍiyāna} is easy in the way that is taught by the guru.
 %The reading \emph{astatandraḥ} is probably the original one in the \emph{Dattātreyayogaśāstra} and is attested in some manuscripts of the \emph{Haṭhapradīpikā}.
@@
 \begin{philcomm}[hp03_064]
-In \emph{Haṭhapradīpikā} this verse seems to instruct the yogi to press the bulb (\emph{kanda}) with the feet while holding them with the hands.
+In the \emph{Haṭhapradīpikā} this verse seems to instruct the yogi to press the bulb (\emph{kanda}) with the feet while holding them with the hands.
 The adopted reading of the source text, the \emph{Gorakṣaśataka}, which is only found in witness T (and there in the singular \emph{jānu}, which has been emended to the dual \emph{jānū} in Mallinson's edition) says that it is the knees that are to be held, which would still allow for the feet to press the bulb.
 In the \emph{Yuktabhavadeva} (7.224), Bhavadevamiśra says that the bulb is near the navel and the legs are held near the ankles, suggesting that the bulb is pressed by using the hands to pull the feet into the lower abdomen.
-Holding the ankles with the hands and pressing the \emph{kanda} with the feet is also the view of Brahmānanda (\emph{Jyotsnā} 3.114), who follows the \emph{Yogayājñavalkya} (4.14, 4.16) in thinking that the place of the \emph{kanda} is nine fingerbreadths above the middle of the body, which is two fingerbreaths above the anus (\emph{Jyotsnā} 3.113).
+Holding the ankles with the hands and pressing the \emph{kanda} with the feet is also the view of Brahmānanda (\emph{Jyotsnā} 3.114), who follows the \emph{Yogayājñavalkya} (4.14, 4.16) in thinking that the place of the \emph{kanda} is nine fingerbreadths above the middle of the body, which is two fingerbreadths above the anus (\emph{Jyotsnā} 3.113).
 In the \emph{Haṭhasaṅketacandrikā} (f.~36r), Sundaradeva explains that the yogi presses the \emph{kanda} in the navel by performing \emph{uḍḍiyānabandha}, along with the root and chin locks, at the end of \emph{kumbhaka} and the beginning of exhalation.
@@
 \begin{translation}[hp03_065]
-[The yogi] should very gently stretch back his stomach, chest and †neck† in such a way that the breath does not come into contact with the stomach.
-\end{translation}
+[The yogi] should very gently stretch back his stomach, chest and neck in such a way that the breath does not come into contact with the stomach.
+\end{translation}%??JM relevant to note too, isn't gala more throat than neck?
 
@@
 \begin{versinnote}
-\tl{paścimaṃ tānam udare dhārayed dhṛdaye gate |\\+}
+\tl{paścimaṃ tānam udare kārayed dhṛdaye gale |\\+}
 \tl{śanaiḥ śanair yathā prāṇas tundasaṃdhiṃ na gacchati ||\\+}
@@
 However, we have not seen the neck mentioned in this regard in any other premodern work and the absence of \emph{ca} suggests that \emph{gale} may be a corruption.
+% śanaiḥ śanaiḥ "gradually" for ca?
 The alternative reading \emph{cibukaṃ hṛdi} in manuscripts of the \emph{Hathapradīpikā} on lower branches of the stemma and in the testimonia is a reference to \emph{jālandharabandha} and appears to be a patch.
@@
 % Remove crux marks in the edition as gale is not gibberish
-
+%??
+Is the KDham experiment relevant?
+It reports an effect of uddiyana but the text is saying to perform a stretch of the throat, ie the action not the result.
+JB: the negative pressure in the oesophagus might explain why a 'stretch' is felt in the throat (i.e., the idea of stretching the throat by doing uḍḍiyāṇa is plausible).
 \begin{metre}[hp03_065]
@@
 \begin{philcomm}[hp03_067]
-Manuscripts of the \textalpha, \textbeta, \texteta\ and delta groups have \emph{sthāpayed dṛḍham icchayā} (`one should place it firmly as desired') in the second verse quarter, which is also well-attested in the transmission of the source text, the \emph{Dattātreyayogaśāstra}.
+Manuscripts of the \textalpha, \textbeta, \textdelta\ and \texteta\ groups have \emph{sthāpayed dṛḍham icchayā} (`one should place it firmly as desired') in the second verse quarter, which is also well-attested in the transmission of the source text, the \emph{Dattātreyayogaśāstra}.
 This reading seems secondary because, in a subsequent verse (3.69), contracting the throat is the main feature of \emph{jālandharabandha}, so it seems contradictory to say that it may be done `as one likes' in 3.67b.
@@
 
+% MD 2024-8: I would again suggest that we consider the reading "jarārogādi kā kathā (not to mention age, sickness, etc.)".
+In terms of content, it fits better than "jarārogādikaṃ tathā".
+The only problem is that "kā kathā" is usually constructed with loc., gen.
+or prati with acc.
+Perhaps "jarārogādi" can be treated like "jarārogādi prati"?
+% Cf. Yogabīja 67ab: vyādhayas tasya naśyanti chedaghātādikās tathā |
+% °ghātādikās tathā~] \Sone\Sfour\Sfive\YCM\YSU, 
+% -ghātādi kā kathā \Etwo\Nthree\Nfour\Nn, (vyādhayas tasya naśyanti chedaghātādi kā kathā)
+% -dhyānādi+ kathaṃ \Eone, 
+% -yātādi kā kathā \Ptwo, 
+% -khātādikās tathā \Stwo\Sthree, 
+% -khātādi kā kathā \Sy, 
+% -ghātādikāḥ kathā \None\Ntwo\Nfive, 
+% -ghātādikā vyathāḥ \Pthree,
+
 %%%%%%%%%%
@@
 \begin{translation}[hp03_074a]
-Now the inverted bodily position:
+Now, the inverted bodily position:
 \end{translation}
@@
 \subsection*{3.73*1}
+
 \begin{translation}[hp03_073_1]
@@
 \end{translation}
-%3.75-76 gray-scaled.
-
-%\begin{sources}[hp03_073_1]
-%\end{sources}
-
-\begin{testimonia}[hp03_073_1]
-\emph{Haṭharatnāvalī} 2.72 
-\begin{versinnote}
-\tl{atha viparītakaraṇī–\\+}
-\tl{yat kiñ cit sravate candrād amṛtaṃ divyarūpi ca |\\+}
-\tl{tatsarvaṃ grasate sūryas tena piṇḍaṃ vināśi ca ||\\!}
-\end{versinnote}
-
-\emph{Yogacintāmaṇi} f.~77v
-\begin{versinnote}
-\tl{haṭhapradīpikāyām—\\+}
-\tl{yat kiṃ cin sravate candrād amṛtaṃ divyarūpi ca |\\+}
-\tl{tat sarvaṃ grasate sūryas tena piṇḍaṃ vināśi ca ||\\!}
-\end{versinnote}
-
-Cf.~\emph{Haṭhayogasaṃhitā} 38 (p. 26)
-\begin{versinnote}
-\tl{nābhimūle vaset sūryas tālumūle ca candramāḥ |\\+}
-\tl{amṛtaṃ grasate sūryas tato mṛtyuvaśo naraḥ ||\\!}
-\end{versinnote}
-\end{testimonia}
 
 \begin{philcomm}[hp03_073_1]
-%No source?
-%The term \emph{piṇḍa} is rarely neuter, hence the change to the masculine in the \emph{Jyotsnā}.
-%Both ca-s are infelicitous J5 has jīryate for vināśi ca
+See \manuref{4.10}, where this verse is also found.\lb
+
 The \textalpha\ group do not have 3.73*1 and 3.73*2 in the third chapter (but rather in the fourth) and other manuscripts omit them as well (notably the \texteta\ group).
 It appears that they have been inserted at the beginning of the section on \emph{viparītakaraṇī} as a kind of preamble, which is unusual as the other techniques in this chapter do not have such introductions.
-
-%are found in alpha in ch.4, but should they be there?
-They are very physical and not suited to ch.4.
-But here they are unusual because they are preamble and the practices already described do not have preamble.
-Perhaps drop them altogether.
-But then there is no explanation of how VK works.
-
-%
 \end{philcomm}
@@
 \subsection*{3.73*2}
+
 \begin{translation}[hp03_073_2]
@@
 
-\begin{sources}[hp03_073_2]
-\end{sources}
-
-\begin{testimonia}[hp03_073_2]
-Haṭharatnāvali 2.73 (on \emph{viparītakaraṇī})
-\begin{versinnote}
-\tl{tatrāsti divyaṃ karaṇaṃ sūryasya mukhabandhanam |\\+}
-\tl{gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ || \\!}
-\end{versinnote}
-
-\emph{Yogacintāmaṇi} f.~77v (attr.~to the \emph{Haṭhapradīpikā})
-\begin{versinnote}
-\tl{tatrāsti karaṇaṃ divyaṃ sūryasya mukhabandhanam |\\+}
-\tl{gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ ||\\!}
-\end{versinnote}
-
-%Yogakarṇikā 147c–148b
-%\begin{versinnote}
-%\tl{tat sarvaṃ grasate sūryas tena piṇḍo jarāyutaḥ || 147 ||\\+}
-%\tl{tatrāsti karaṇaṃ divyaṃ sūryasya mukhavañcanam |\\!}
-%\end{versinnote}
-\end{testimonia}
-
-%\begin{philcomm}[hp03_073_2]
-%No source? Only Brahmānanda reads mukhavañcanam, which makes better sense. Judit; bandhanaṃ makes sense with grasate sūryas
-%\end{philcomm}
+\begin{philcomm}[hp03_073_2]
+See \manuref{4.11}, where this verse is also found.
+\end{philcomm}
 
@@
 \end{translation}
+% MD: yogokta- or yogoktair?
 
@@
 According to Brahmānanda (\emph{Jyotsnā} 3.84), the compound \emph{vaśavartinī}, which we have translated as `an obedient woman,' could be a wife (\emph{vaśavartinī svādhīnā nārī vanitā}).
-In 3.83, the reading \emph{bhāryābhage} in \getsiglum{V1}, \getsiglum{V3} and \getsiglum{J10} supports Brahmānanda's view that the woman is the yogi's wife.
+In 3.83, the reading \emph{bhāryābhage} in \etaOne, \etaTwo\ and \betaOmega\ supports Brahmānanda's view that the woman is the yogi's wife.
 %Is kṣīra semen? Or a sap that is hard to find? Difficult to say since nothing more is said about kṣīra.
@@
 \begin{philcomm}[hp03_085]
-The omission of 3.85ab in the \texteta\ group and \deltaThree\ is likely to be the result of haplography (\emph{bindudhāraṇāt} is repeated).
+The omission of 3.85ab in the \texteta\ group and \deltaThree\ is likely to be the result of haplography (\emph{bindudhāraṇāt} is repeated).\lb
 % MD: I have deleted "in \alphaThree", because the damaged part has enough space for 3.84cd and 3.85ab.
@@
 
-The readings \emph{mṛtyubhayaṃ} (\alphaThree, \textbeta, \texteta) and \emph{kālabhayaṃ} (\alphaTwo, \textgamma) are well attested by the main manuscript groups, but \emph{mṛtyubhayaṃ} is in the important witnesses of the source text, the \emph{Vivekamārtaṇḍa}.
-% MD: Which reading does ms T have?
-%  JB: record that alphaThree has mṛtyubhayaṃ in 85d? alphaTwo has kālabhayaṃ, so the stemma (as a whole) seems split. [MD: done]
+The readings \emph{mṛtyubhayaṃ} (\alphaThree, \textbeta, \textepsilon, \texteta) and \emph{kālabhayaṃ} (\alphaTwo, \textgamma) are well attested by the main manuscript groups, but \emph{mṛtyubhayaṃ} is in the important witnesses of the source text, the \emph{Vivekamārtaṇḍa}.
 \end{philcomm}
@@
 \begin{translation}[hp03_087]
-In this way a [the yogi] may also hold on to [both] the menses of a menstruating woman and his semen. By practising correctly he may draw up [both] through the urethra by the proper practice.
-\end{translation}
+In this way a [the yogi] may also hold on to [both] the menses of a menstruating woman and his own semen. He who has mastered yoga through correct practice may draw up [both] through the urethra.
+\end{translation}
+
+% MD: pāda b: striyā must be a wrong emendation of gamma group. If the best attested bījaṃ does not work, shall we read svīyaṃ (β1)? Maybe, eva svīyaṃ > evamvījaṃ > evaṃ bījaṃ?
+% pāda d: abhyāsayogavān is better attested than °yogataḥ.
 
@@
 %No known source. 
-We have understood the reading of \emph{ṛtumatyā} as qualifying \emph{striyāḥ} (i.e.,`a menstruating woman'). Alternatively, the term \emph{ṛtumati} could mean a post-pubescent woman.
+%We have understood the reading of \emph{ṛtumatyā} as qualifying \emph{striyāḥ} (i.e.,`a menstruating woman'). Alternatively, the term \emph{ṛtumati} could mean a post-pubescent woman.
 \end{philcomm}
@@
 Some manuscripts, including \alphaTwo\ and \alphaThree\ (missing in \alphaOne), omit 3.89ab.
-We have included it because in the \emph{Dattātreyayogaśāstra}, the source of this verse, 3.89ab specifies the substance mentioned in 3.89cd that the man and woman are supposed to rub into their bodies after sexual intercourse.
-
-In the \emph{Dattātreyayogaśāstra}’s teaching on \emph{sahajolī} (163 and 181–183) a rag is used to wipe up the residue of a mixture of semen and sweat that has been rubbed into the body, and then soaked in a paste of water and ash before being rubbed over the body.
+We have included it because in the \emph{Dattātreyayogaśāstra}, the source of this verse, 3.89ab specifies the substance mentioned in 3.89cd that the man and woman are supposed to rub into their bodies after sexual intercourse.\lb
+
+In the \emph{Dattātreyayogaśāstra}’s teaching on \emph{sahajolī} (163 and 181–183) a rag is used to wipe up the residue of a mixture of semen and sweat that has been rubbed into the body, and then soaked in a paste of water and ash before being rubbed over the body.\lb
 
@@
 \begin{translation}[hp03_091]
-Leaving out the first flow of urine because of its excessive heat and the last flow because it is worthless, the cool middle flow of urine is used by Kāpālikas of the Khaṇḍa school.
-\end{translation}
+Leaving out the first flow because of its excessive heat and the last flow because it is worthless, the cool middle flow of urine is used by Kāpālikas of the Khaṇḍa school.
+\end{translation}
+% MD: adopt the reading "prathamāṃ ca dhārāṃ"? "ambu" was probably introduced in the mss where the last word "amaryāḥ" had been corrupted.
 
@@
 \begin{philcomm}[hp03_092_1]
-In the first verse quarter, the gerund \emph{samākṛṣya} (\textgamma) has been adopted, instead of \emph{samākuñcya}, as it yields a better meaning and is used similarly to \emph{ākṛṣya} in 3.83.
+In the first verse quarter, the gerund \emph{samākṛṣya} (\textgamma) has been adopted, instead of \emph{samākuñcya}, as it yields a better meaning and is used similarly to \emph{ākṛṣya} in 3.83.\lb
 
@@
 \begin{philcomm}[hp03_094_2]
-%Authorial?
 %3.4 and 4.4 have cety ekavācakāḥ, but devī paryāyavācakāḥ is well attested among the HP manuscripts (Groups 2 and 3) and the testimonia.
 The \alphaThree\ manuscript has a significantly shorter and more coherent version of \emph{śakticālana}.
-It omits five introductory verses, of which three are from the \emph{Vivekamārtaṇḍa} or one of its longer recensions and three have no known source, including one that contains a list of synonyms for \emph{kuṇḍalinī}.
-This section is missing in \alphaOne\ (3.83–3.98) and \alphaTwo\ adds these verses (except 3.94*6) after 3.97, which suggests that they have been inserted from elsewhere.
+It omits six introductory verses, of which three are from the \emph{Vivekamārtaṇḍa} or one of its longer recensions and three have no known source, including one that contains a list of synonyms for \emph{kuṇḍalinī}.
+This section is missing in \alphaOne\ (3.84–3.97) and \alphaTwo\ adds these verses (except 3.94*6) after 3.97, which suggests that they have been inserted from elsewhere.
 Generally speaking, it appears that some redactors have taken the section on \emph{śakticālana} in the \emph{Haṭhapradīpikā} as an opportunity to add material on \emph{kuṇḍalinī}, in particular her location, shape, and soteriological importance.
@@
 %\end{philcomm}
+%??
+MD: It seems that the redactor (= the creator of the extended version of HP) first moved 4.62 here, as it deals with kuṇḍalinī and fits the topic here, together with the three adjacent verses (4.61, 63 and 64), as these four form a set (ślokacatuṣṭaya).
+In the \textepsion\ they are still found in the order 4.62 -> heading of the ślokacatuṣṭaya -> 4.61 -> 4.63 -> 4.64.
+However, except for 4.62, they are omitted in the mss of the lower branches of the stemma (zeta, eta), probably because they were considered out of place.
 
@@
 \begin{translation}[hp03_095]
-Between the Gaṅgā and Yamuna is the wretched young widow. [The yogi] should forcefully take [her]. That is the supreme state of Viṣṇu.
+Between the Gaṅgā and Yamunā is the wretched young widow. [The yogi] should forcefully take [her]. That is the supreme state of Viṣṇu.
 \end{translation}
@@
 %No known source. 
-The referent of \emph{bālaraṇḍā tapasvinī}, here is unclear.
+The referent of \emph{bālaraṇḍā tapasvinī} here is unclear.
 In some manuscripts this verse is followed by one in which \emph{bālaraṇḍā} is identified as \emph{sarasvatī}, which in the context of \emph{śakticālana} could refer to the tongue.
@@
 \end{philcomm}
-% \alphaTwo\ has this verse in a correct position. Its second half is different from that of the other recensions:
+% ?? MD: α1 is unavailable. 
+% α3 is damaged, so it is uncertain, but most probably it didn't have the verse. There is not enough space for it.
+% α2 has this verse in a correct position, but its second half is different from that of the other recensions:
 % iḍā bhagavatī gaṅgā piṅgalā yamunā nadī |
@@
 \begin{translation}[hp03_096_1]
-The yogi should regularly move the coiled (\emph{paristhitā}), hooded [serpent] by breathing in through the sun channel and holding her using a cloth for an hour and a half in the morning and evening.
+The yogi should breathe in through the solar channel, take hold of the open-mouthed hooded [serpent] by wrapping a cloth around [her] and move her sideways for an hour and a half in the morning and evening.
 \end{translation}
@@
 \tl{paridhān[a]yukteti dvādaśāṅgulapramitasitasūkṣmacaturaṅgulavisṛtaśuddhavastrakhaṇḍena dṛḍhaṃ veṣṭatā sā prasiddhā [ph]aṇāvatī suṣumṇātmakā arundhatī jihvaiva kuṇḍalinī || uktaṃ ca ||\\+}
-\tl{arundhatī bhavej jihvā dhruvo nāsāgramaṇḍalam iti || \\+}
+\tl{arundhatī bhavej jihvā dhruvo nāsāgramaṇḍalam iti ||\\+}
 \tl{tāṃ jihvāṃ laṃbikāyogenordhvaṃ tālvantarbhrūmadhyadeśe vihitāṃ tatas tīrthād bhrūmadhyāt pragṛhya adhaḥ kṛtvā tasyā gurūpadiṣṭavartmanā cālanaṃ vidheyam iti saṃketaḥ [|] cālanaṃ tu khecarī mudrā sādhanavad vidheyaṃ [|] tīrthaṃ bhrūmadhyaḥ [|] \\!}
@@
 \begin{philcomm}[hp03_096_1]
-Verses 3.96*1–2 are absent in \alphaOne\ and \alphaTwo.
+Verses 3.96*1–2 are absent in \alphaTwo\ and \alphaThree. 
 They introduce the idea of awakening \emph{kụṇḍalinī} by moving the tongue with a cloth, which is a practice called \emph{sarasvatīcālana} in the \emph{Gorakṣaśataka} (16–25).
-These verses do not have a known source and are somewhat obscure unless one is one aware of the more coherent explanation of this practice in the \emph{Gorakṣaśataka}.
+These verses do not have a known source and are somewhat obscure unless one is aware of the more coherent explanation of this practice in the \emph{Gorakṣaśataka}.
 In his \emph{Haṭhasaṅketacandrikā} (see testimonia), Sundaradeva makes sense of this verse by equating the tongue with Kuṇḍalinī.
-This enables him to understand the reference to the cloth (\emph{paridhāna}) as the technique of wrapping the tongue in a cloth and milking it, which is a practice called \emph{sarasvatīcālana} in the \emph{Gorakṣaśataka} (16–25).
+This enables him to understand the reference to the cloth (\emph{paridhāna}) as the technique of wrapping the tongue in a cloth and milking it (i.e., \emph{sarasvatīcālana} in the \emph{Gorakṣaśataka}).
 This interpretation also makes sense of the next verse in the \emph{Haṭhapradīpikā} (3.11), which describes the cloth.
-
-%
-
-%C2 - nityaṃ
-%adopt yuktyā, nityam paricālanīyā (C6 - 4b group) [MD: done]
+%?? MD: partly repeated. JB: okay, i've changed this.
+
 %GŚ 22 says to move the tongue for two muhūrtas
@@
 \begin{philcomm}[hp03_098]
-In \emph{Jyotsnā} 3.116, Brahmānanda understands \emph{bhānor ākuñcanaṃ} as a contraction of the navel, whereas Sundaradeva in his \emph{Haṭhasaṃketacandrikā} (see testimonia) interprets it as drawing \emph{prāṇa} through the sun channel (\emph{sūryanāḍyākarṣaṇa}), thereby intensifiying the bodily fire.
+In \emph{Jyotsnā} 3.116, Brahmānanda understands \emph{bhānor ākuñcanaṃ} as a contraction of the navel.
+In contrast, Sundaradeva, in his \emph{Haṭhasaṃketacandrikā} (see testimonia), interprets it as drawing \emph{prāṇa} through the sun channel (\emph{sūryanāḍyākarṣaṇa}), thereby intensifying the bodily fire.
 \end{philcomm}
@@
 \begin{translation}[hp03_098_1]
-When \emph{prāṇa} has been greatly extended, [the yogi's] breath flows through the path of the right nostril and [his] body is immortal, filled with the nectar of the moon from the uvula for the first time.
-Sprinkling [with nectar] the network of channels at the aperture at [the centre of the] brow that is subjugated by the mighty fire of time, [the yogi] then makes his body completely new again like [the regeneration of] the trunk of an ancient tree.
+%Through the breath that is flowing through the path of the right nostril the [the yogi's] breath is greatly extended. 
+When \textit{prāṇa} has been greatly extended through the breath that is flowing through the path of the right nostril, with [his] body already immortal, filled with the nectar of the moon through the path from the uvula,
+%[his] body is immortal, filled with the nectar of the moon from the uvula for the first time. 
+[the yogi], sprinkling [with nectar] the network of channels at the aperture at [the centre of the] brow that have been subjugated by the mighty fire of time, he then makes his body completely new again like the trunk of a withered tree [when it is regenerated].
 \end{translation}
 %greyscaled
+% Suggested translation with proposed reading °pavanāt, etc
+% The vital force is greatly extended because of the breath flowing through the path of the right nostril.
+The [yogi's] body is immortal, already filled with the nectar of the moon through the path from the uvula.
 
@@
 This verse's import of rejuvenating the body by flooding it with nectar is not directly connected with those proceeding it.
+Its meaning is not clear in the published edtion of the \textit{Amaraughaśāsana} and we are not confident of the readings adopted in our edition nor the meaning of the verse.
 
@@
 \begin{translation}[hp03_098_2]
-After making Kuṇḍalinị move, the yogi should perform \emph{bhastrī} in particular. The god of death is afraid of the ascetic who regularly practises in this way.
+After making Kuṇḍalinī move, the yogi should perform \emph{bhastrī} in particular. The god of death is afraid of the ascetic who regularly practises in this way.
 \end{translation}
@@
 The practice of other \emph{kumbhaka}s for moving Kuṇḍalinī is mentioned in 3.98*3.
+\lb
 
 The reading \emph{abhyāsato} in 3.98*2c, which is attested by manuscripts of the \textgamma\ group and the testimonia, is possible but seems to be a later corruption of the participle, \emph{abhyasyato}.
+%??
+MD: abhyasyato seems to be overcorrect in this text in which always abhyaset is used instead of abhyasyet.
+The participle abhyasato (cf.
+3.63 abhyasan, though this might be wrong for abhyasen) or the adverb abhyāsato (cf.
+1.27c) may be more prabable.
 \end{philcomm}
@@
 \begin{translation}[hp03_099]
-That fearless [yogi] should move [Kuṇḍalinī] for up to one hour 36 minutes (two \emph{muhūrta}s). Suṣumṇā at Kuṇḍalinī is drawn up slightly.
-\end{translation}
+As a result of fearlessly moving [Kuṇḍalinī] for one hour 36 minutes (two \emph{muhūrta}s), Suṣumṇā at Kuṇḍalinī is drawn up slightly.
+\end{translation}
+% MD: better to read "nirbhayaṃ cālanād asau" instead of "nirbhayaś cālayed asau".
+nirbhayaṃ is an adverb here.
+The variant reading nirbhayaś probably originates from the misreading of °bhayañcā° as °bhayaścā° in a South Indian ms.
 
@@
 \begin{philcomm}[hp03_100]
-We have supplied `mouth' (\emph{mukha}) in our translation on the basis of the reading found in the source text (the \emph{Gorakṣaśataka}).
 \end{philcomm}
@@
 \begin{translation}[hp03_102_1]
-For [the yogi] devoted to celibacy and always eating a healthy and measured diet, success appears after forty days (\emph{maṇḍalāt}) as a result of engaging in the practice of Kuṇḍalinī.
+For the yogi whose practice is with Kuṇḍalinī, who has a vow of celibacy and is always eating a healthy and measured diet, success appears after forty days (\emph{maṇḍalāt}).
 \end{translation}
@@
 This block appears after 3.97 and appears to have been inserted from elsewhere.
-3.102*1 has no known source and appears to have been added as a general laudatory statement on the benefits of practising with Kuṇḍalinī.
+3.102*1 has no known source and appears to have been added as a general laudatory statement on the benefits of practising with Kuṇḍalinī.\lb
 
@@
 In the \emph{Haṭhayogasaṃhitā} (p.
-41) and \emph{Jyotsnā} (3.98), this verse is in the section on \emph{amarolī}, which provides a clear referent of \emph{cāndrī} as the cool middle flow of urine (see 3.95).
+41) and \emph{Jyotsnā} (3.98), this verse is in the section on \emph{amarolī}, which provides a clear referent of \emph{cāndrī} as the cool middle flow of urine (see 3.96 [3.91 in our edition]).
 
@@
 \begin{philcomm}[hp03_104]
-Manuscripts of the \textbeta, \texteta, and \textepsilon\ groups have a different reading for the second line; `each mudrā is capable of being a cause of all powers' (\emph{kāraṇaṃ sarvasiddhīnāṃ ekaikāpi kṣamaiva sā}).
+Manuscripts of the \textbeta, \textepsilon, \texteta, and \textzeta\ groups have a different reading for the second line; `each mudrā is capable of being a cause of all powers' (\emph{kāraṇaṃ sarvasiddhīnāṃ ekaikāpi kṣamaiva sā}).%??
+MD: karaṇe of ε may be the better reading: "capable of bringing about all siddhis"
 \end{philcomm}
@@
 \begin{translation}[hp03_108]
-For those who are tireless in their dedication, Rājayoga has a \emph{mudrā}. That is the supreme \emph{rudrāṇi} \emph{mudrā}, which bestows beneficial success.
-\end{translation}
+For those who are tireless in their dedication, Rājayoga has a \emph{mudrā}, and this other \emph{mudrā}, \emph{Rudrāṇī}, bestows great perfection.
+\end{translation}
+%??
+MD: Wouldn't the alpha reading rājayoga-samudra-vat work?
+"Like the ocean of rājayoga (as the ocean yields jewels?), Rudrāṇī bestows great perfection on those who are tireless..." I am not fully convinced myself, but the present reading is not nice either.
+And I have a strong feeling that the verse consists of only one sentence.
 
@@
 \end{philcomm}
+% α1:   upāsane vinidrāṇāṃ rājayogasamudravat/
+%       rudrāṇī cāparā mudrā bhadrāṃ siddhiṃ prayacchati//
+% IFP:  upāsane vinidrāṇāṃ rājayogasamudragāṃ/
+%       rudrāṇīvāparāṃ mudrāṃ bhadrāṃ siddhiṃ prayacchati//
+% G11:  abhyāseṣu vinidrāṇām anudrutasamādhiṣu/
+%       rudrāṇīva parā mudrā bhadrāṃ siddhiṃ prayacchati//
 
@@
 \begin{translation}[hp03_108_2]
-The yogi who has become intent on that [guru's] teaching and practises with a focused mind obtains mastery of the powers beginning with minimisation and the cheating of death.
+The yogi who has become intent on that [guru's] teaching and practises a \textit{mudrā} with a focused mind cheats death with the powers beginning with minimisation.
 \end{translation}

### 2025-04-04 latex/HP4_comm.tex
new files

--- 2024-07-02 latex/HP4_comm.tex
@@
 %%%%%%%%%%
+\subsection*{4.1 heading}
+\begin{translation}[hp04_001a]
+Now \emph{samādhi}:
+\end{translation}
+
+% \begin{philcomm}[hp04_001a]
+% \end{philcomm}
+
+%%%%%%%%%%
 \subsection*{4.0*1}
@@
 The first additional verse resembles a \emph{maṅgala} verse that one might expect to see at the beginning of a text.
-The second is a verse from the \emph{Gorakṣaśataka} that introduces the topic of \emph{samadhi}.
-Some manuscripts of the \textbeta, \textepsilon\ and \texteta\ groups have the two verses on the synonyms of \emph{rājayoga} here (on their position in the \textalpha\ group and other manuscripts, see the note to 4.29).
+The second is a verse from the \emph{Gorakṣaśataka} that introduces the topic of \emph{samādhi}.
+Some manuscripts of the \textbeta, \textepsilon, \textzeta\ and \texteta\ groups have the two verses on the synonyms of \emph{rājayoga} here (on their position in the \textalpha\ group and other manuscripts, see the note to 4.32).
 The rest of the additional verses (4.0*5–4.0*16) are a motley collection on \emph{samādhi}, \emph{rājayoga}, the importance of the guru, dissolving the breath, \emph{suṣumṇā}, etc.
 In contrast to this, the \textalpha\ group begins with a cohesive discusson on absorption (4.1–3) that transitions to the gaze (4.4) and a brief discussion of \emph{śāmbhavī} and \emph{khecarī mudrā}s (4.5–4.8).
-The main topic of the chapter, which is meditating on the internal sound (\emph{nādānusandhāna}), begins at 4.9 in the \textalpha\ group (whereas in other groups it begins after fifty or so verses).
-The emphasis on \emph{nādānusandhāna} in the fourth chapter of the \textalpha\ group is consistent with the statement in verse 1.56 that \emph{nādānusandhāna} is the fourth component of Haṭhayoga.
-
-Its likely that the term \emph{nirañjanapada} was understood here as \emph{samādhi} because \emph{nirañjana} is included in a list of synonyms of \emph{samādhi} later in this chapter (4.29).
+The main topic of the chapter, which is meditating on the internal sound (\emph{nādānusandhāna}), begins at 4.12 in the \textalpha\ group (whereas in other groups it begins after fifty or so verses).
+The emphasis on \emph{nādānusandhāna} in the fourth chapter of the \textalpha\ group is consistent with the statement in verse 1.56 that \emph{nādānusandhāna} is the fourth component of Haṭhayoga.\lb
+
+Its likely that the term \emph{nirañjanapada} was understood here as \emph{samādhi} because \emph{nirañjana} is included in a list of synonyms of \emph{samādhi} later in this chapter (4.32).\lb
 
@@
 \begin{translation}[hp04_000_2]
-So now I will teach the best way to \emph{samādhi}.It destroys death, has an easy method and brings about the bliss of Brahman.
+So now I will teach the best way to \emph{samādhi}. It destroys death, has an easy method and brings about the bliss of Brahman.
 \end{translation}
@@
 %%%%%%%%%%
+\subsection*{4.0*3}
+\begin{translation}[hp04_000_3]
+The sovereign yoga (\emph{rājayoga}), meditative absorption (\emph{samādhi}), the beyond mind state (\emph{unmanī}), the transmental state (\textit{manonmanī}), [the sovereign yoga of] the lineage of immortals (\emph{amaraugha}), dissolution [of mind] (\emph{laya}), the [ultimate] reality (\emph{tattva}), void and not void (\textit{śūnyāśūnya}), the highest state (\emph{para pada}), [\dots]
+\end{translation}
+
+\subsection*{4.0*4}
+\begin{translation}[hp04_000_4]
+[\dots] no-mind (\emph{amanaska}), non-duality (\emph{advaita}), without support (\emph{nirālamba}), pure (\emph{nirañjana}), liberation in life (\emph{jīvanmukti}), innate (\emph{sahaja}) and the fourth [state] (\emph{turya}) are synonyms.
+\end{translation}
+
+%%%%%%%%%%
 \subsection*{4.0*5}
@@
 \begin{translation}[hp04_000_7]
-Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation is established [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
+Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation becomes steady [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
 \end{translation}
@@
 The third verse quarter of this verse has been subject to much revision in the \emph{Haṭhapradīpikā} and the source text, the \emph{Amanaska}.
-Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}.
-The compound \emph{muktisthitiḥ} (literally, `the establishment of liberation') has been adopted.
+Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthirā}, \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}.
+We have adopted \emph{muktiḥ sthirā}, the reading of \textepsilon, an important group for the grey-scaled verses.
 %  attested by \etaTwo, 
 % adopt muktisthitiḥ and rewrite note. 
-%
+% MD: The actual reading is muktiḥ sthirā.
 \end{philcomm}
@@
 \begin{translation}[hp04_000_11]
-When the breath is flowing in the central channel, and the mind enters the void, the knower of yoga destroys all actions.
+When the breath is flowing in the central channel, and the mind enters the void, the expert destroys all actions.
 \end{translation}
@@
 
-\begin{sources}[hp04_000_11]
-\end{sources}
+% \begin{sources}[hp04_000_11]
+% \end{sources}
 
@@
 \begin{translation}[hp04_000_12]
-Homage to you, the immortal one. You have slain even death, into whose mouth this world, with everything that is moving and unmoving, has fallen.
-\end{translation}
+O Lord of the Immortals, homage to you. You have slain even death, into whose mouth this world, with everything that is moving and unmoving, has fallen.
+\end{translation}
+%??
+MD: Should we perhaps read 'amaraugha' instead of 'amareśa'?
+Considering 4.11*30, it is possible for inanimate things to become objects of worship.
+And it may not be a coincidence that the next verse is a quotation from the Amaraugha.
 
@@
 %\begin{philcomm}[hp04_000_12]
-%It appears that this verse, which is in manuscripts of the \textbeta\ and \textepsilon\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
+%It appears that this verse, which is in manuscripts of the \textbeta\ and \textzeta\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
 %
@@
 \end{translation}
-%
+%?? MD: tadâśā jīvite'pi ca? "then there is also the prospect of staying alive"?
 
@@
 \begin{translation}[hp04_000_15]
-Both mercury and the mind are unsteady by nature. [When either] mercury is fixed [or] the mind is fixed, nothing in the world is impossible.
+Both mercury and the mind are flighty by nature. When mercury is stabilised [or] the mind is stabilised, nothing in the world is impossible.
 \end{translation}
@@
 \begin{philcomm}[hp04_000_16]
-The \emph{Rasārṇava} is the likely source of this verse given the vocative, \emph{bhairavi}. Similar verses are found in other yoga texts (see Mallinson and Szántó 2021: 57).
+The \emph{Rasārṇava} is the likely source of this verse given the vocative, \emph{bhairavi}. Similar verses are found in other yoga texts (see Mallinson and Szántó 2021: 57).\lb
 
@@
 The manuscripts attest two versions of the second line.
-It appears that the reading of \textbeta, \textepsilon\ and \texteta\ groups, which we have adopted, was changed to remove the first person pronoun.
+It appears that the reading of \textbeta, \textepsilon\ and \textzeta\ groups, which we have adopted, was changed to remove the first person pronoun.
 The modified reading is prevalent in the \textgamma\ and \textdelta\ manuscripts (\emph{manaḥprāṇalayānando nāpi kaś cit vibhidyate}).
@@
 \begin{philcomm}[hp04_005]
-See Birch 2013:286 for more parallel verses.
+See Birch 2013: 286 for more parallel verses.
 \end{philcomm}
@@
 %%%%%%%%%%
-\subsection*{4.8*1}
-\begin{translation}[hp04_008_1]
+\subsection*{4.9}
+\begin{translation}[hp04_009]
+That which enters the aperture into the underworld, which exists at the base of Meru, the wise yogin say that is the truth, the source of [all] rivers.
+The essence of the body flows from the moon.
+Because of that, death arises for people.
+One should dam that [essence] with the earth that is the good \textit{mudrā}.
+Otherwise, there is no bodily perfection.
+\end{translation}
+% See 3.49*2
+
+% The commentary has been moved from the third chapter. Should be modified.
+
+\begin{testimonia}[hp04_009]
+\emph{Haṭharatnāvalī} 4.30
+\begin{versinnote}
+\tl{pātāle yad viśati suṣiraṃ merūmūlaṃ tad asti\\+}
+\tl{tattvaṃ caitad vadati sudhā tanmukhaṃ nimnagānām |\\+}
+\tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
+\tl{tad badhnīyāt sukharatimṛdur nānyathā kāryasiddhiḥ ||\\!}
+\end{versinnote}
+% after (ekaṃ sṛṣṭimayaṃ + mano yatra vilīyeta) and before (śarīraṃ tāvad eva)
+%the context of this verse in the HR is rājayoga.
+
+\emph{Yogacintāmaṇi} f.~75r (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tat pātālād viyati śikhare merumūle tad asti \\+}
+\tl{tattvaṃ caitat pravadati sudhīḥ saṃmukhe nimnagānām |\\+}
+\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
+\tl{tad badhnīyāt svakaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
+\end{versinnote}
+% after the verse (cittāyattaṃ nṛṇāṃ śukraṃ) and before (indhanāni yathā vahnis)
+% context is khecarī (3rd chapter material)
+
+\emph{Yuktabhavadeva} 7.220 (attr.~to Gorakṣanātha)
+\begin{versinnote}
+\tl{pātāle yad viśati suṣiraṃ merumūle yad asti \\+}
+\tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
+\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
+\tl{badhnīyāt tat sukaraṇam atho nānyathā kāyasiddhiḥ ||\\!}
+\end{versinnote}
+% context is khecarī, (3rd chapter material)
+% after (ekaṃ sṛṣṭimayaṃ bījaṃ) and before (atha uḍḍiyānabandhaḥ)
+
+\emph{Haṭhatattvakaumudī} 14.26 (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{pātāle yad vitatasuṣiraṃ merumūle tad asmin \\+}
+\tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
+\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
+\tl{taṃ badhnīyāt sukaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
+\end{versinnote}
+% context is khecarī 3rd chapter material
+% after (cumbantī yadi lambikā) and before (cittaṃ vicarati gagane)
+
+\end{testimonia}
+
+\begin{philcomm}[hp04_009]
+%Workshop notes
+% Sven: the first line could refer to Kuṇḍalinī
+% J5 has this verse here.
+% mṛdā = using clay to make a dam?
+% Also see Netratantra 7.32
+% Could this verse be about mūlabandha? (reading pātāle and mūle)
+%J5: yat prāleyam pihitasukhire merumūle yad astī — this seems helpful.
+% PN has a nice-ish reading: yat prāleyaṃ pihita sukhiraṃ merumūrdhnaṃ. mūrdhni etc. is much better.
+% J10 has merumūrdhni sthitaṃ yat, which is good.
+% If going with prāleyam, adopt yat prāleyam pihitasuṣiraṃ merumūrdhny asti tathyaṃ (~K1 Grp2)
+% ... tasmin tattvaṃ pravadati
+%Haru: Perhaps, praleyam was corrupted to pātāle and then the rest of the line changed.
+
+% α1,δ1 - in Ch. 4 only
+% β1,β2 - after 3.39
+% all the other mss - at 3.49*2
+% among them α2,α3,gamma have it in both Ch. 3 and 4
+On the position of these verses in the various recensions of the text, see the introduction (ref??).\lb
+
+%This verse occurs in various places in the different recensions of the text.
+It is here in chapter four in the \textalpha\ and \textgamma\ groups (\alphaTwo\ and \alphaThree\ also have it in the third chapter at 3.49*2).
+It occurs after 3.39 in \betaOne\ and \betaTwo, and in both the third and fourth chapters in the \textgamma\ manuscripts.
+It must have entered the transmission of the \emph{Haṭhapradīpikā} at an early stage, and has moved around, perhaps because the name of the technique alluded to is not stated and the meaning of the first half of the verse is somewhat vague without its original context (we are yet to identify a source).
+
+The first half of the verse has many variants among the witnesses. \alphaTwo\ has:
+\begin{versinnote}
+\tl{yat prāleyaṃ pihitasukhire merumūle yad astī \\+}
+\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\!}
+\end{versinnote}
+This version is close to that of other important manuscripts in the \textbeta, \textgamma\ and \texteta\ groups.
+With the help of these manuscripts (in particular \getsiglum{K1} of the \textgamma\ group), the above can be emended and understood as follows:
+\begin{versinnote}
+\tl{yat prāleyaṃ pihitasuṣiraṃ merumūrdhny asti tathyaṃ \\+}
+\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\+}
+\tl{\var{°suṣiraṃ ] \getsiglum{K1}, °sukhire \alphaTwo. °mūrdhny ] \getsiglum{P8}. °mūle \alphaTwo. asti tathyaṃ ] \getsiglum{K1}, yad astī \alphaTwo.}\\!}
+“That cool liquid by which the aperture is filled at the top of Meru and exists as the truth, the wise [yogi] says that is the source of [all] rivers.” 
+\end{versinnote}
+
+% Other manuscripts have something more similar to:
+% \begin{versinnote}
+% \tl{pātāle yad viśati suṣiraṃ merumūlaṃ tad asti \\+}
+% \tl{tattvaṃ caitat pravadati sudhīs tan mukhaṃ nimnagānām\\!}
+% \end{versinnote}
+
+\end{philcomm}
+
+% MD: It is most likely that these and two other stanzas were here first and were moved to the 3rd chapter by the person who reorganised the HP.
+I would therefore suggest reactivating these stanzas here.
+Agreed.
+
+\begin{metre}[hp04_009]
+Mandākrāntā 
+\end{metre}
+
+%%%%%%%%%%
+\subsection*{4.10}
+\begin{translation}[hp04_010]
+The sun devours whatever nectar flows from the divine moon. As a result, the body is afflicted by old age.
+\end{translation}
+
+%\begin{sources}[hp04_010]
+%\end{sources}
+
+\begin{testimonia}[hp04_010]
+\emph{Haṭharatnāvalī} 2.72 
+\begin{versinnote}
+\tl{atha viparītakaraṇī–\\+}
+\tl{yat kiñ cit sravate candrād amṛtaṃ divyarūpi ca |\\+}
+\tl{tatsarvaṃ grasate sūryas tena piṇḍaṃ vināśi ca ||\\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~77v
+\begin{versinnote}
+\tl{haṭhapradīpikāyām—\\+}
+\tl{yat kiṃ cin sravate candrād amṛtaṃ divyarūpi ca |\\+}
+\tl{tat sarvaṃ grasate sūryas tena piṇḍaṃ vināśi ca ||\\!}
+\end{versinnote}
+
+Cf.~\emph{Haṭhayogasaṃhitā} 38 (p. 26) % Gheraṇḍasaṃhitā 3.33
+\begin{versinnote}
+\tl{nābhimūle vaset sūryas tālumūle ca candramāḥ |\\+}
+\tl{amṛtaṃ grasate sūryas tato mṛtyuvaśo naraḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_010]
+%No source? %?? MD: VM 113-114 (Jim's working edition)
+% nābhideśe vasaty eva bhāskaro dahanātmakaḥ |
+% amṛtātmā sthito nityaṃ tālumadhye ca candramāḥ || 113 ||
+% varṣaty adhomukhaś candro gṛhṇāty ūrdhvamukho raviḥ |
+% jñātavyaṃ kāraṇaṃ tatra yena pīyūṣam āpyate || 114 ||
+%(cited also in Jyotsnā 3.77)
+
+%The term \emph{piṇḍa} is rarely neuter, hence the change to the masculine in the \emph{Jyotsnā}.
+
+%are found in alpha in ch.4, but should they be there?
+They are very physical and not suited to ch.4.
+But here they are unusual because they are preamble and the practices already described do not have preamble.
+Perhaps drop them altogether.
+But then there is no explanation of how VK works.
+
+%Supposedly these verses were originally here as in the alpha mss, first moved to the end of the Viparītakaraṇī section (as in the epsilon), and finally moved to the beginning of the same section (as in many other groups).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11}
+\begin{translation}[hp04_011]
+There is a divine bodily position for this, which blocks the mouth of the sun. It is to be known from the teaching of a guru and not through the countless interpretations of scriptures.
+\end{translation}
+% greyscaled
+
+%\begin{sources}[hp04_011]
+%\end{sources}
+
+\begin{testimonia}[hp04_011]
+Haṭharatnāvali 2.73 (on \emph{viparītakaraṇī})
+\begin{versinnote}
+\tl{tatrāsti divyaṃ karaṇaṃ sūryasya mukhabandhanam |\\+}
+\tl{gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ || \\!}
+\end{versinnote}
+
+\emph{Yogacintāmaṇi} f.~77v (attr.~to the \emph{Haṭhapradīpikā})
+\begin{versinnote}
+\tl{tatrāsti karaṇaṃ divyaṃ sūryasya mukhabandhanam |\\+}
+\tl{gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ ||\\!}
+\end{versinnote}
+
+%Yogakarṇikā 147c–148b
+%\begin{versinnote}
+%\tl{tat sarvaṃ grasate sūryas tena piṇḍo jarāyutaḥ || 147 ||\\+}
+%\tl{tatrāsti karaṇaṃ divyaṃ sūryasya mukhavañcanam |\\!}
+%\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_011]
+%See the note on the previous verse.
+%No source? Only Brahmānanda reads mukhavañcanam, which makes better sense. Judit; bandhanaṃ makes sense with grasate sūryas
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*1}
+\begin{translation}[hp04_011_1]
 [The yogi] should fix the pupils on a light and slightly raise the eyebrows. This is the way of the preliminary yoga, which immediately brings about the beyond-mind state.
@@
 
-\begin{sources}[hp04_008_1]
+\begin{sources}[hp04_011_1]
 \emph{Amanaska} 1.8 (South Indian Recension)
@@
 
-\begin{testimonia}[hp04_008_1]
+\begin{testimonia}[hp04_011_1]
 \emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_008_1]
-This and the next verse (4.8*1–2) are from the South Indian recension of the \emph{Amanaska}, a later rewriting of that work which incorporates these verses in an additional passage on Tārakayoga, added here to elaborate on the meditation of fixing the gaze that is taught in 4.4–7.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*2}
-\begin{translation}[hp04_008_2]
-Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
-\end{translation}
-
-\begin{sources}[hp04_008_2]
+\begin{philcomm}[hp04_011_1]
+This and the next verse (4.11*1–2) are from the South Indian recension of the \emph{Amanaska}, a later rewriting of that work which incorporates these verses in an additional passage on Tārakayoga, added here to elaborate on the meditation of fixing the gaze that is taught in 4.4–7.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*2}
+\begin{translation}[hp04_011_2]
+Some are confused by a multitude of tantric texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
+\end{translation}
+
+\begin{sources}[hp04_011_2]
 \emph{Amanaska} 1.11 (South Indian Recension)
@@
 
-\begin{testimonia}[hp04_008_2]
+\begin{testimonia}[hp04_011_2]
 \emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_008_2]
-In the source text \emph{tārakam} refers to  Tāraka yoga, one of two yogas taught in the South Indian recension of the \emph{Amanaska}.
-
-On why this verse is in greyscale, see the note to 4.8*1.    
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*3}
-\begin{translation}[hp04_008_3]
-With eyes half open, mind steady, gaze placed at the tip of the nose and inner state motionless, [the yogi] should lead the moon and sun to dissolution.
-He attains the state of the highest reality, the supreme essence, whose form is light and which is devoid of anything external and shining intensely.
-What more is to be said here?
-\end{translation}
-
-\begin{sources}[hp04_008_3]
+\begin{philcomm}[hp04_011_2]
+In the source text \emph{tārakam} refers to Tāraka yoga, one of two yogas taught in the South Indian recension of the \emph{Amanaska}.\lb
+
+On why this verse is in greyscale, see the note to 4.11*1.    
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*3}
+\begin{translation}[hp04_011_3]
+By leading the moon and sun to dissolution in a motionless state, the [yogi], his eyes half open, mind steady, and gaze placed at the tip of the nose, attains the supreme reality (\textit{vastu}), the state that is the highest principle (\textit{tattva}),
+% attains the state of the highest reality, the supreme essence (\textit{vastu}), 
+whose form is light and which is devoid of anything external and shining intensely. What more is to be said here?
+\end{translation}
+
+\begin{sources}[hp04_011_3]
 \emph{Candrāvalokana} 3
@@
 
-\begin{testimonia}[hp04_008_3]
+\begin{testimonia}[hp04_011_3]
 \emph{Yogacintāmaṇi} ff. 24v–25r (attr.~to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_008_3]
+\begin{philcomm}[hp04_011_3]
 This verse is not in the \textalpha\ manuscripts and was probably added because it elaborates on the gaze.
 In the \emph{Candrāvalokana} (i.e., the source text) and the \emph{Anubhavanivedana}, it follows 4.7, which may explain its position in the \textgamma\ and \textdelta\ groups.
-\end{philcomm}
-
-\begin{metre}[hp04_008_3]
+The attentive reader will notice that we have decided against manuscripts that agree with the source text.
+The reason is based on our stemmatic considerations.
+Cumulative evidence supports the origin of the additional verses at the beginning of the fourth chapter in the hyparchetype of the \textepsilon\ manuscripts.
+We suspect these verses contaminated the \textbeta\ group's copies very early.
+To complicate matters, the scribe was familiar with the source text and corrected the readings accordingly.
+\end{philcomm}
+
+\begin{metre}[hp04_011_3]
 Śārdūlavikrīḍita 
@@
 %%%%%%%%%%
-\subsection*{4.8*4}
-%\begin{translation}[hp04_008_4]
-%That which enters the aperture into the underworld, which exists at the base of Meru,  the wise yogin say that is the truth, the source of [all] rivers.
-The essence of the body flows from the moon.
-Because of that, death arises for people.
-[This] good mudrā should block that [essence].
-Otherwise, there is no bodily perfection.
-%\end{translation}
-% See 3.49
-
-%\begin{sources}[hp04_008_4]
-%\end{sources}
-
-% \begin{testimonia}[hp04_008_4]
-% \emph{Haṭharatnāvalī} 4.30
-% \begin{versinnote}
-% \tl{pātāle yad viśati suṣiraṃ merūmūlaṃ tad asti \\+}
-% \tl{tattvaṃ caitad vadati sudhā tanmukhaṃ nimnagānām |\\+}
-% \tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ\\+}
-% \tl{tad badhnīyāt sukharatimṛdur nānyathā kāryasiddhiḥ ||\\!}
-% \end{versinnote}
-
-% \emph{Yogacintāmaṇi} f.~74v (attr.~to the \emph{Haṭhapradīpikā})
-% \begin{versinnote}
-% \tl{pātāle yad viyati śikhare merumūle tad asti\\+}
-% \tl{tattvaṃ caitat pravadati sudhīs sammukhe nimnagānām |\\+}
-% \tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ\\+}
-% \tl{taṃ badhnīyāt svakaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
-% \end{versinnote}
-
-% \emph{Yuktabhavadeva} 7.220 (attr.~to Gorakṣanātha)
-% \begin{versinnote}
-% \tl{pātāle yad viśati suṣiraṃ merumūle yad asti \\+}
-% \tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
-% \tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
-% \tl{badhnīyāttat sukaraṇam atho nānyathā kāyasiddhiḥ ||\\!}
-% \end{versinnote}
-
-% \emph{Haṭhatattvakaumudī} 14.26 (attr.~to the \emph{Haṭhapradīpikā})
-% \begin{versinnote}
-% \tl{pātāle yadvitatasuṣiraṃ merumūle tad asmin\\+}
-% \tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
-% \tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
-% \tl{taṃ badhnīyāt sukaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
-% \end{versinnote}
-% \end{testimonia}
-
-\begin{philcomm}[hp04_008_4]
-For the translation and testimonia, as well as an explanation of the various places and versions of this verse in the text, see 3.49.
-\end{philcomm}
-
-%\begin{metre}[hp04_008_4]
-%Mandākrāntā 
-%\end{metre}
-
-%%%%%%%%%%
-\subsection*{4.8*5}
-\begin{philcomm}[hp04_008_5]
-See 3.75, where this verse is also found.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*6}
-\begin{philcomm}[hp04_008_6]
-See 3.76, where this verse is also found.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*7}
-\begin{translation}[hp04_008_7]
+\subsection*{4.11*4}
+\begin{translation}[hp04_011_4]
 [The yogi] should not worship the \emph{liṅga} by day, nor should he worship it by night. He should worship the \emph{liṅga} constantly, by suppressing day and night.
-\end{translation}
-
-\begin{sources}[hp04_008_7]
+% he should worship it day and night. Beta 2
+\end{translation}
+
+\begin{sources}[hp04_011_4]
 \emph{Khecarīvidyā} 3.19
@@
 
-\begin{testimonia}[hp04_008_7]
+\begin{testimonia}[hp04_011_4]
 \emph{Haṭhasaṅketacandrikā} f.~128v (attr.~Haṭhapradīpikā)
@@
 
-\begin{philcomm}[hp04_008_7]
+\begin{philcomm}[hp04_011_4]
 The manuscripts transmit readings for the last verse quarter that either contradict or repeat the statements in the first line.
@@
 %%%%%%%%%%
-\subsection*{4.8*8}
-\begin{translation}[hp04_008_8]
-Now \emph{khecarī}:\\
+\subsection*{4.11*5 heading}
+\begin{translation}[hp04_011_5a]
+Now \emph{khecarī}:
+\end{translation}
+
+% \begin{philcomm}[hp04_011_5a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*5}
+\begin{translation}[hp04_011_5]
 There is a hollow that generates knowledge and has five streams. \emph{Khecarīmudrā} is situated in that pure void.
@@
 %
-%\begin{sources}[hp04_008_8]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_8]
+%\begin{sources}[hp04_011_5]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_5]
 \emph{Haṭhasaṅketacandrikā} f.~128v
@@
 
-\begin{philcomm}[hp04_008_8]
+\begin{philcomm}[hp04_011_5]
 This verse may not be referring to the cavity in which the tongue is placed but perhaps to a more esoteric sense based on the meaning of \textit{pañcasrotas} as the five streams of tantric Śaiva teachings, which in this case generate knowledge.
@@
 
-\begin{metre}[hp04_008_8]
+\begin{metre}[hp04_011_5]
 Anuṣṭubh (a: na-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.8*9}
-\begin{translation}[hp04_008_9]
+\subsection*{4.11*6}
+\begin{translation}[hp04_011_6]
 The breath in the left and right channels moves into the middle. Without doubt, \emph{khecarīmudrā} abides in that place.
@@
 
-%\begin{sources}[hp04_008_9]
-%\end{sources}
-
-%\begin{testimonia}[hp04_008_9]
+%\begin{sources}[hp04_011_6]
+%\end{sources}
+
+%\begin{testimonia}[hp04_011_6]
 %\emph{Haṭhasaṅketacandrikā} f. 246v (Mysore ORI Manuscript B220) 
@@
 
-%\begin{philcomm}[hp04_008_9]
+%\begin{philcomm}[hp04_011_6]
 %The term \emph{madhye} appears to be referring to the central channel here. 
@@
 %%%%%%%%%%
-\subsection*{4.8*10}
-%\begin{translation}[hp04_008_10]
+\subsection*{4.11*7}
+%\begin{translation}[hp04_011_7]
 %The mind moves in the void because the tongue moves in the void, which is why that \emph{mudrā} is worshipped by the adepts with the name “she who moves in the void” (\emph{khecarī}).
@@
 
-% \begin{sources}[hp04_008_10]
+% \begin{sources}[hp04_011_7]
 % \emph{Vivekamārtaṇḍa} 50
@@
 
-% \begin{testimonia}[hp04_008_10]
+% \begin{testimonia}[hp04_011_7]
 % \emph{Yogacintāmaṇi} [attrib. Skandapurāṇa]
@@
 
-\begin{philcomm}[hp04_008_10]
+\begin{philcomm}[hp04_011_7]
 See 3.37, where this verse is also found.
@@
 %%%%%%%%%%
-\subsection*{4.8*11}
-\begin{translation}[hp04_008_11]
-At the juncture of Iḍā and Piṅgalā, the void devours the breath. \emph{Khecarīmudrā} abides there. This is true, time and time again.
-\end{translation}
-
-%\begin{sources}[hp04_008_11]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_11]
+\subsection*{4.11*8}
+\begin{translation}[hp04_011_8]
+At the juncture of Iḍā and Piṅgalā, the void devours the breath. \emph{Khecarīmudrā} abides there. This is undoubtedly true.
+\end{translation}
+
+%\begin{sources}[hp04_011_8]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_8]
 \emph{Upāsanāsārasaṅgraha} (IFP Transcript T1095) p.~42.
@@
 
-\begin{philcomm}[hp04_008_11]
+\begin{philcomm}[hp04_011_8]
 Cf.~Rāghavabhaṭṭa \emph{ad Śāradātilaka} 25.43: \emph{suṣumṇāyām eteṣu parvasu iḍāpiṅgalayor yogo bhavatīti jñeyam}.
@@
 %%%%%%%%%%
-\subsection*{4.8*12}
-\begin{translation}[hp04_008_12]
-The \emph{mudrā} situated in the cakra of the void (\emph{vyomacakre}) in the middle of the sun and moon on an unsupported surface is [the \emph{mudrā}] called \emph{khecarī}.
-\end{translation}
-
-%\begin{sources}[hp04_008_12]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_12]
+\subsection*{4.11*9}
+\begin{translation}[hp04_011_9]
+The \emph{mudrā} situated in the cakra of the void (\emph{vyomacakre}) in the middle of the moon and sun on an unsupported surface is [the \emph{mudrā}] called \emph{khecarī}.
+\end{translation}
+
+%\begin{sources}[hp04_011_9]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_9]
 \emph{Upāsanāsārasaṅgraha} (IFP Transcript T1095) p.~41.
@@
 
-\begin{philcomm}[hp04_008_12]
+\begin{philcomm}[hp04_011_9]
 The reading \emph{tale} is suspect, as is \emph{nirālambāntare}, the conjecture of Brahmānanda.
-%  bhruvormadhye dvyasraṃ vyomacakraṃ tad dhi cinmayasya jyotiṣa upalabdhisthānam  Commentary on LYV 9.82.
-% Haṭhasaṅketacandrika (4192): śūnyaṃ vyomacakravyavasthitam | iḍāpiṃgalāṃtargataṃ brahmaraṃdhrākhyaṃ vyomacakraṃ [...]. 
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*13}
-\begin{translation}[hp04_008_13]
-Brought forth by me, lovely, and clearly beloved of Śiva, the Suṣumṇā should fill herself with the divine air through her rear mouth.
-\end{translation}
-
-%\begin{sources}[hp04_008_13]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_13]
-\emph{Yogasārasaṅgraha} p. 61 (attrib. Praṇavacintāmaṇi)
-\begin{versinnote}
-\tl{sā mayā viditā yā māyā sākṣācchivavallabhā |\\+}
-\tl{pūrayen mārutaṃ divyaṃ suṣumnā paścime mukhe ||\\!}
-\end{versinnote}
-\end{testimonia}
-
-\begin{philcomm}[hp04_008_13]
-The referent of \emph{mayā} is unspecified and no source text has been identified.
-It most probably refers to a goddess.%to understand here because the referent of \emph{mayā} is not known.
-If it refers to a yogi, then it might mean that \emph{suṣumnā} has been pierced.
-Otherwise, if \emph{udbheditā} means `brought forth', one would assume the speaker to be a god or goddess.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*14}
-\begin{translation}[hp04_008_14]
-And if [she] fills herself from the front, \emph{khecarī} definitely arises. [The yogi] should practise \emph{khecarīmudrā}. The state beyond mind arises.
-\end{translation}
-
-%\begin{sources}[hp04_008_14]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_14]
-\emph{Upāsanāsārasaṅgraha} p.~135
-purastāc caiva pūryeta niścitā khecarī bhavet |
-abhyaset khecarīmudrāmunmanī saṃprajāyate ||
-\end{testimonia}
-
-\begin{philcomm}[hp04_008_14]
-We have understood \emph{niścitā} as an adverb. No witnesses have \emph{niścitam}, but we see no other way of taking \emph{niścitā}.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*15}
-\begin{translation}[hp04_008_15]
-[The yogi] should practise \emph{khecarīmudrā} until he falls into a yogic sleep. For one who has attained yogic sleep, death never arises.
-\end{translation}
-
-%\begin{sources}[hp04_008_15]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_15]
-\emph{Upāsanāsārasaṅgraha} p.~135\\
-abhyaset khecarīṃ tāvat yāvat syād yoganidrataḥ |
-saṃprāptayoganidrasya kālo nāsti kadācana ||
-\end{testimonia}
-
-\begin{philcomm}[hp04_008_15]
-\end{philcomm}
-
-%%%%%%%%%%
-
-\subsection*{4.8*16}
-\begin{translation}[hp04_008_16]
-Between the eyebrows is the place of Śiva. The mind dissolves there. That level should be known as the fourth state. Death does not exist there.
-\end{translation}
-%
-
-%\begin{sources}[hp04_008_16]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_16]
-\emph{Upāsanāsārasaṅgraha} p.~135
-bhruvor madhye śivasthānaṃ manas tatra vilīyate |
-jñātavyaṃ tat paraṃ turyaṃ tatra kālo na vidyate ||
-\end{testimonia}
-
-%\begin{philcomm}[hp04_008_16]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*17}
-\begin{translation}[hp04_008_17]
-Between the moon and the sun, [the yogi] should apply \emph{khecarīmudrā}, which is situated in the supportless, great void, the \emph{vyomacakra}.
-\end{translation}
-
-\begin{sources}[hp04_008_17]
-\emph{Jñānasāra} 3.3cd
-\begin{versinnote}
-\tl{candrasūryadvayor madhye muḍādadyā? tu khecarīm |\\!}
-\end{versinnote}
-
-\end{sources}
-
-%\begin{testimonia}[hp04_008_17]
-%\end{testimonia}
-
-\begin{philcomm}[hp04_008_17]
+
 The \emph{vyomacakra} (`the cakra of space') is also mentioned in a half-verse that was added to some later recensions of the \emph{Haṭhapradīpikā} (see 3.35) and states that \emph{vyomacakra} is another name for \emph{khecarīmudrā}.
@@
 129v), Sundaradeva says that it is called the Brahmarandhra, which is between Iḍā and Piṅgalā (\emph{iḍāpiṅgalāntargataṃ brahmarandhrākhyaṃ vyomacakraṃ tat khecarīmudrāṃ śaktiṃ kuryād ...}).
+%  bhruvormadhye dvyasraṃ vyomacakraṃ tad dhi cinmayasya jyotiṣa upalabdhisthānam  Commentary on LYV 9.82.
+% Haṭhasaṅketacandrika (4192): śūnyaṃ vyomacakravyavasthitam | iḍāpiṃgalāṃtargataṃ brahmaraṃdhrākhyaṃ vyomacakraṃ [...]. 
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*10}
+\begin{translation}[hp04_011_10]
+Pierced by me, lovely, and clearly beloved of Śiva, the Suṣumṇā should fill herself with the divine air through her rear mouth.
+\end{translation}
+
+%\begin{sources}[hp04_011_10]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_10]
+\emph{Yogasārasaṅgraha} p. 61 (attrib. Praṇavacintāmaṇi)
+\begin{versinnote}
+\tl{sā mayā viditā yā māyā sākṣācchivavallabhā |\\+}% unmetrical
+\tl{pūrayen mārutaṃ divyaṃ suṣumnā paścime mukhe ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_011_10]
+The referent of \emph{mayā} is unspecified, and no source text has been identified.
+It is most probably the goddess.
+The \textepsilon\ manuscripts have \emph{māyodbheditā} for \emph{mayodbheditā}, but it is hard to make sense of.
+If correct, \emph{udbheditā} would more likely mean “produced” or “made manifest”.
+Alternatively, the first line of this verse could be referring to \emph{khecarīmudrā}, in which case its being produced by \emph{māyā} would make more sense.
+In the second line, Brahmānanda’s reading \emph{suṣumnāṃ} is tempting.
+The line would then mean, “[The yogi] should fill Suṣumnā with divine breath through the rear opening”.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*11}
+\begin{translation}[hp04_011_11]
+And if [she] fills herself from the front, \emph{khecarī} definitely arises. [The yogi] should practise \emph{khecarīmudrā}. The state beyond mind arises.
+\end{translation}
+
+%\begin{sources}[hp04_011_11]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_11]
+\emph{Upāsanāsārasaṅgraha} p.~135
+\begin{versinnote}
+\tl{purastāc caiva pūryeta niścitā khecarī bhavet |\\+}
+\tl{abhyaset khecarīmudrāmunmanī saṃprajāyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_011_11]
+We have understood \emph{niścitā} as an adverb. No witnesses have \emph{niścitam}, but we see no other way of taking \emph{niścitā}.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*12}
+\begin{translation}[hp04_011_12]
+[The yogi] should practise \emph{khecarīmudrā} until he falls into a yogic sleep. For one who has attained yogic sleep, death never arises.
+\end{translation}
+
+%\begin{sources}[hp04_011_12]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_12]
+\emph{Upāsanāsārasaṅgraha} p.~135
+\begin{versinnote}
+\tl{abhyaset khecarīṃ tāvat yāvat syād yoganidrataḥ |\\+}
+\tl{saṃprāptayoganidrasya kālo nāsti kadācana ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+\begin{philcomm}[hp04_011_12]
+\end{philcomm}
+
+%%%%%%%%%%
+
+\subsection*{4.11*13}
+\begin{translation}[hp04_011_13]
+Between the eyebrows is the place of Śiva. The mind dissolves there. That level should be known as the fourth state. Death does not exist there.
+\end{translation}
 %
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*18}
-\begin{translation}[hp04_008_18]
+
+%\begin{sources}[hp04_011_13]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_13]
+\emph{Upāsanāsārasaṅgraha} p.~135
+\begin{versinnote}
+\tl{bhruvor madhye śivasthānaṃ manas tatra vilīyate |\\+}
+\tl{jñātavyaṃ tat paraṃ turyaṃ tatra kālo na vidyate ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_011_13]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*14}
+\begin{translation}[hp04_011_14]
+Between the moon and the sun, [the yogi] should apply \emph{khecarīmudrā}, which is situated in the supportless, great void, the \emph{vyomacakra}.
+\end{translation}
+
+\begin{sources}[hp04_011_14]
+\emph{Jñānasāra} 3.3cd
+\begin{versinnote}
+\tl{candrasūryadvayor madhye muḍādadyā? tu khecarīm |\\!}
+\end{versinnote}
+
+\end{sources}
+
+%\begin{testimonia}[hp04_011_14]
+%\end{testimonia}
+
+% \begin{philcomm}
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*15}
+\begin{translation}[hp04_011_15]
 [The yogi] should make the mind supportless and think of nothing at all. He assuredly remains like a pot in the ether, [empty] inside and outside.
@@
 
-\begin{sources}[hp04_008_18]
+\begin{sources}[hp04_011_15]
 \emph{Jñānasāra} 3.4
@@
 
-%\begin{testimonia}[hp04_008_18]
+%\begin{testimonia}[hp04_011_15]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_008_18]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*19}
-\begin{translation}[hp04_008_19]
+%\begin{philcomm}[hp04_011_15]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*16}
+\begin{translation}[hp04_011_16]
 Just as the external air has dissolved into the void, the breath is sure to go to its place †with the mind on the side of the sun†.
@@
 
-%\begin{sources}[hp04_008_19]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_19]
+%\begin{sources}[hp04_011_16]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_16]
 \emph{Haṭhapradīpikā} (with ten-chapters) 7.52 
@@
 
-\begin{philcomm}[hp04_008_19]
+\begin{philcomm}[hp04_011_16]
 % sūryāgnir manasā tathā M1
@@
 %%%%%%%%%%
-\subsection*{4.8*20}
-\begin{translation}[hp04_008_20]
+\subsection*{4.11*17}
+\begin{translation}[hp04_011_17]
 For [the yogi] practising in this way day and night on the path of the breath, as a result of the practice the breath is consumed, [and] the mind dissolves into it.
@@
 
-%\begin{sources}[hp04_008_20]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_20]
+%\begin{sources}[hp04_011_17]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_17]
 \emph{Haṭhasaṅketacandrikā} f. 129v
@@
 
-\begin{philcomm}[hp04_008_20]
+\begin{philcomm}[hp04_011_17]
 Brahmānanda identifies the path of the breath (\emph{vāyumārga}) with Suṣumṇā.
@@
 %%%%%%%%%%
-\subsection*{4.8*21}
-\begin{translation}[hp04_008_21]
-[The yogi] should flood the body with nectar from the soles of the feet to the head. He is sure to become perfected, with a great body and great strength and valour.\\Thus ends \emph{khecarī}.
-\end{translation}
-
-%\begin{sources}[hp04_008_21]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_21]
+\subsection*{4.11*18}
+\begin{translation}[hp04_011_18]
+[The yogi] should flood the body with nectar from the soles of the feet to the head. [His] body is perfected forever, and he has great strength and valour.\\Thus ends \emph{khecarī}.
+\end{translation}
+
+%\begin{sources}[hp04_011_18]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_18]
 \emph{Haṭhasaṅketacandrikā} f. 129v
@@
 
-%\begin{philcomm}[hp04_008_21]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*22}
-\begin{translation}[hp04_008_22]
-Now \emph{śāmbhavī}:\\
+%\begin{philcomm}[hp04_011_18]
+%\end{philcomm}
+
+%%%%%%%%%%
+% \subsection*{4.11*18 ending}
+% iti khecarī/
+% \begin{translation}[hp04_011_18p]
+% \end{translation}
+
+% \begin{philcomm}[hp04_011_18p]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*19 heading}
+\begin{translation}[hp04_011_19a]
+Now \emph{śāmbhavī}:
+\end{translation}
+
+% \begin{philcomm}[hp04_011_19a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*19}
+\begin{translation}[hp04_011_19]
 {[}The yogi] should [put] the mind in Śakti and Śakti in the mind, observe the mind with the mind, and meditate on it as the supreme state.
@@
 
-\begin{sources}[hp04_008_22]
+\begin{sources}[hp04_011_19]
 \emph{Candrāvalokana} 27
@@
 
-\begin{testimonia}[hp04_008_22]
+\begin{testimonia}[hp04_011_19]
 \emph{Haṭhasaṅketacandrikā} f.~129v–130r
@@
 
-\begin{philcomm}[hp04_008_22]
+\begin{philcomm}[hp04_011_19]
 In the second verse quarter, \emph{mana} for \emph{mano} is for the metre.
@@
 %%%%%%%%%%
-\subsection*{4.8*23}
-\begin{translation}[hp04_008_23]
+\subsection*{4.11*20}
+\begin{translation}[hp04_011_20]
 Put the self in space and put space in the self. [The yogi] should make the self consist of space and think of nothing at all.
@@
 
-\begin{sources}[hp04_008_23]
+\begin{sources}[hp04_011_20]
 \emph{Uttaragītā} 1.9
@@
 
-\begin{testimonia}[hp04_008_23]
+\begin{testimonia}[hp04_011_20]
 \emph{Haṭharatnāvalī} 4.45
@@
 
-%\begin{philcomm}[hp04_008_23]
+%\begin{philcomm}[hp04_011_20]
 % Critical edition of the Mbh: 
@@
 %%%%%%%%%%
-\subsection*{4.8*24}
-\begin{translation}[hp04_008_24]
+\subsection*{4.11*21}
+\begin{translation}[hp04_011_21]
 Like an empty pot in air, [the yogī] is empty on the inside and empty on the outside. Like a full pot in the ocean, [the yogi] is full on the inside and full on the outside.%
@@
 
-\begin{sources}[hp04_008_24]
+\begin{sources}[hp04_011_21]
 \emph{Laghuyogavāsiṣṭha} 6.15.79 (\emph{Mokṣopāya} 6.155.25)
@@
 
-\begin{testimonia}[hp04_008_24]
+\begin{testimonia}[hp04_011_21]
 \emph{Haṭharatnāvalī} 4.46
@@
 
-%\begin{philcomm}[hp04_008_24]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*25}
-\begin{translation}[hp04_008_25]
+%\begin{philcomm}[hp04_011_21]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*22}
+\begin{translation}[hp04_011_22]
 Do not think about the external or internal. [The yogi] should abandon all thought and think of nothing at all.
@@
 
-%\begin{sources}[hp04_008_25]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_25]
+%\begin{sources}[hp04_011_22]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_22]
 \emph{Haṃsavilāsa} p.~48
@@
 
-%\begin{philcomm}[hp04_008_25]
+%\begin{philcomm}[hp04_011_22]
 %°cintanā is not attested elsewhere.
@@
 %%%%%%%%%%
-\subsection*{4.8*26}
-\begin{translation}[hp04_008_26]
+\subsection*{4.11*23}
+\begin{translation}[hp04_011_23]
 The whole world is but a construct of mere ideation.
 A construct of mere ideation is an affectation of the mind.
-Jettison thought that is only ideation, take refuge in a resolve that is free of ideation, and obtain peace, O Rāma.
-\end{translation}
-
-\begin{sources}[hp04_008_26]
+So jettison this ideation.
+Take refuge in a resolve that is free of ideation, and obtain peace, O Rāma.
+\end{translation}
+
+\begin{sources}[hp04_011_23]
 \emph{Laghuyogavāsiṣṭha} 7.27
@@
 
-%\begin{testimonia}[hp04_008_26]
+%\begin{testimonia}[hp04_011_23]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_008_26]
-%\end{philcomm}
-
-\begin{metre}[hp04_008_26]
+\begin{philcomm}[hp04_011_23]
+
+\end{philcomm}
+
+\begin{metre}[hp04_011_23]
 Vasantatilakā
@@
 %%%%%%%%%%
-\subsection*{4.8*27}
-\begin{translation}[hp04_008_27]
+\subsection*{4.11*24}
+\begin{translation}[hp04_011_24]
 Just as camphor in fire and salt in water, so the mind, on being brought into contact with the highest reality, dissolves into it.
@@
 
-%\begin{sources}[hp04_008_27]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_27]
+%\begin{sources}[hp04_011_24]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_24]
 \emph{Haṭharatnāvalī} 4.43
@@
 
-%\begin{philcomm}[hp04_008_27]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*28}
-\begin{translation}[hp04_008_28]
+%\begin{philcomm}[hp04_011_24]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*25}
+\begin{translation}[hp04_011_25]
 Mind is said to be all that is to be known, [all] that has been perceived, and [all] knowledge of that. Knowledge and what is to be known are destroyed together. There is no other path.
@@
 
-%\begin{sources}[hp04_008_28]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_28]
+%\begin{sources}[hp04_011_25]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_25]
 \emph{Haṭhatattvakaumudī} 51.35
@@
 
-%\begin{philcomm}[hp04_008_28]
+%\begin{philcomm}[hp04_011_25]
 %Function of ca?
@@
 %%%%%%%%%%
-\subsection*{4.8*29}
-\begin{translation}[hp04_008_29]
-All this, everything moving and unmoving, is [just] a vision of the mind. When the mind has become free of the mind, it is called the absence of duality.
-\end{translation}
-
-\begin{sources}[hp04_008_29]
+\subsection*{4.11*26}
+\begin{translation}[hp04_011_26]
+All this, everything moving and unmoving, is [just] a vision of the mind. For when the mind has become free of the mind, it is called the absence of duality.
+\end{translation}
+
+\begin{sources}[hp04_011_26]
 \emph{Amanaska} 2.79
@@
 
-\begin{testimonia}[hp04_008_29]
+\begin{testimonia}[hp04_011_26]
 \emph{Yogacintāmaṇi} f. 27r (attr.~to the \emph{Rājayoga}, aka. \emph{Amanaska})
@@
 
-%\begin{philcomm}[hp04_008_29]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*30}
-\begin{translation}[hp04_008_30]
+%\begin{philcomm}[hp04_011_26]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*27}
+\begin{translation}[hp04_011_27]
 As a result of abandoning the things that are to be known, the mind attains dissolution. When the mind has attained dissolution, liberation (\textit{kaivalyam}) remains.
@@
 
-%\begin{sources}[hp04_008_30]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_30]
+%\begin{sources}[hp04_011_27]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_27]
 \emph{Haṭharatnāvalī} 4.44
@@
 
-%\begin{philcomm}[hp04_008_30]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.8*31}
-\begin{translation}[hp04_008_31]
+%\begin{philcomm}[hp04_011_27]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*28}
+\begin{translation}[hp04_011_28]
 “Dissolution, dissolution”, they say. What kind of characteristics does dissolution have? Because subliminal impressions do not arise again, dissolution is the forgetting of the objects of the senses.
@@
 
-%\begin{sources}[hp04_008_31]
-%\end{sources}
-
-\begin{testimonia}[hp04_008_31]
+%\begin{sources}[hp04_011_28]
+%\end{sources}
+
+\begin{testimonia}[hp04_011_28]
 \emph{Haṭharatnāvalī} 1.13
@@
 
-\begin{philcomm}[hp04_008_31]
+\begin{philcomm}[hp04_011_28]
 This verse may have been inspired by the \emph{Mokṣopāya} (e.g.~1.2.2).
@@
 %%%%%%%%%%
-\subsection*{4.8*32}
-\begin{translation}[hp04_008_32]
+\subsection*{4.11*29}
+\begin{translation}[hp04_011_29]
 Various methods like these, which are understood properly by one's own experience, are taught as paths to \textit{samādhi} by the great-souled teachers of former times.
@@
 
-%\begin{sources}[hp04_008_32]
-%\end{sources}
-
-%\begin{testimonia}[hp04_008_32]
+%\begin{sources}[hp04_011_29]
+%\end{sources}
+
+%\begin{testimonia}[hp04_011_29]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_008_32]
-%\end{philcomm}
-
-\begin{metre}[hp04_008_32]
+%\begin{philcomm}[hp04_011_29]
+%\end{philcomm}
+
+\begin{metre}[hp04_011_29]
 Anuṣṭubh (c: bha-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.8*33}
-\begin{translation}[hp04_008_33]
-Homage to Suṣumṇā, to Kuṇḍalinī, to the nectar arising from the moon, to the mind beyond mind state, to you whose nature is consciousness, the great Śakti.%
-\end{translation}
-
-%\begin{sources}[hp04_008_33]
-%\end{sources}
-
-%\begin{testimonia}[hp04_008_33]
+% \subsection*{4.11*30 heading}
+% atha viśrāntiḥ/
+% \begin{translation}[hp04_011_30a]
+% \end{translation}
+
+% \begin{philcomm}[hp04_011_30a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.11*30}
+\begin{translation}[hp04_011_30]
+Homage to Suṣumṇā, to Kuṇḍalinī, to the nectar in the orb of the moon, to the mind beyond mind state, to you whose nature is consciousness, the great Śakti.%
+\end{translation}
+
+%\begin{sources}[hp04_011_30]
+%\end{sources}
+
+%\begin{testimonia}[hp04_011_30]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_008_33]
-%\end{philcomm}
-
-\begin{metre}[hp04_008_33]
+%\begin{philcomm}[hp04_011_30]
+%\end{philcomm}
+
+\begin{metre}[hp04_011_30]
 Anuṣṭubh (a: ra-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.8*34}
-\begin{translation}[hp04_008_34]
+\subsection*{4.11*31}
+\begin{translation}[hp04_011_31]
 The cultivation of the inner sound taught by Gorakṣanātha [and] approved even for deluded people for whom realisation of the highest reality is impossible is [now] taught.
@@
 
-%\begin{sources}[hp04_008_34]
-%\end{sources}
-
-%\begin{testimonia}[hp04_008_34]
+%\begin{sources}[hp04_011_31]
+%\end{sources}
+
+%\begin{testimonia}[hp04_011_31]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_008_34]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.9}
-\begin{translation}[hp04_009]
+%\begin{philcomm}[hp04_011_31]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.12}
+\begin{translation}[hp04_012]
 The twelve and a half million methods of dissolution taught by glorious Śiva are the best.
@@
 \end{translation}
-
-\begin{sources}[hp04_009]
+% MD: From a stemmatic point of view, jayante (not -ti) is slightly preferable.
+
+\begin{sources}[hp04_012]
 \emph{Yogatārāvalī} 2
@@
 
-\begin{testimonia}[hp04_009]
+\begin{testimonia}[hp04_012]
 \emph{Haṭharatnāvalī}~1.12
@@
 
-%\begin{philcomm}[hp04_009]
-%\end{philcomm}
-
-\begin{metre}[hp04_009]
+%\begin{philcomm}[hp04_012]
+%\end{philcomm}
+
+\begin{metre}[hp04_012]
 Upajāti
@@
 %%%%%%%%%%
-\subsection*{4.10}
-\begin{translation}[hp04_010]
-Seated in the pose of the liberated ones, the yogi should adopt \emph{śāṃbhavī mudrā} and, with his mind one-pointed, listen to the inner sound in his right ear.
-\end{translation}
-
-%\begin{sources}[hp04_010]
-%\end{sources}
-
-\begin{testimonia}[hp04_010]
+\subsection*{4.13}
+\begin{translation}[hp04_013]
+Seated in the pose of the liberated ones (\emph{muktāsanasthitaḥ}), the yogi should adopt \emph{śāṃbhavī mudrā} and, with his mind one-pointed, listen to the inner sound in his right ear.
+\end{translation}
+
+%\begin{sources}[hp04_013]
+%\end{sources}
+
+\begin{testimonia}[hp04_013]
 \emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_010]
-We read this verse here (as attested by \textalpha, etc.) rather than at 4.50 because it follows on from the Śāmbhavī section and is unnecessary and repetitive at 4.50 because 4.51-52 explain the practice.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.11}
+\begin{philcomm}[hp04_013]
+In verse 1.37, \emph{muktāsana} is said to be the same as \emph{siddhāsana}.
+We read this verse here (as attested by \textalpha, etc.) rather than at \manuref{4.36*1} because it follows on from the Śāmbhavī section and is unnecessary and repetitive at \manuref{4.36*1} because \manuref{4.37–38} explain the practice.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.14}
 \begin{translation}[hp04_011]
@@
 %%%%%%%%%%
-\subsection*{4.12}
+\subsection*{4.15}
 \begin{translation}[hp04_012]
@@
 %%%%%%%%%%
-\subsection*{4.13}
-\begin{translation}[hp04_013]
+\subsection*{4.16}
+\begin{translation}[hp04_016]
 Having become intent on indifference through regular practice, the ascetic should concentrate on nothing but the inner sound, which immediately brings about the [state] beyond mind.%.
@@
 
-%\begin{sources}[hp04_013]
-%\end{sources}
-
-\begin{testimonia}[hp04_013]
+%\begin{sources}[hp04_016]
+%\end{sources}
+
+\begin{testimonia}[hp04_016]
 \emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_013]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.14}
-\begin{translation}[hp04_014]
-What kind of indifference?\\
-In the cold season, [indifference towards] whether [one has] an open tent or a quilt, with regard to good nourishment, whether it is cow's milk or water, with regard to food, whether it is lots of alms or forest roots, and with regard to the vessel for food, whether it is the hand or some kind of bucket.
-%
-\end{translation}
-
-%\begin{sources}[hp04_014]
-%\end{sources}
-
-\begin{testimonia}[hp04_014]
+%\begin{philcomm}[hp04_016]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.17 heading}
+\begin{translation}[hp04_017a]
+What kind of indifference?
+\end{translation}
+
+% \begin{philcomm}[hp04_017a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.17}
+\begin{translation}[hp04_017]
+In the cold season, [indifference towards] whether [one has] an open tent or a quilt, with regard to good nourishment, whether it is cow's milk or water, with regard to food, whether it is lots of alms [or] forest roots, and with regard to the vessel for food, whether it is the hand or some kind of bowl.
+\end{translation}
+
+% MD: pāda c: better to read "bhojye bhikṣā kandam āraṇyakaṃ vā"?
+% JB: kāpi droṇī: a fine wooden vessel
+
+%\begin{sources}[hp04_017]
+%\end{sources}
+
+\begin{testimonia}[hp04_017]
 \emph{Haṭharatnāvalī} 4.7
@@
 
-\begin{philcomm}[hp04_014]
+\begin{philcomm}[hp04_017]
 On \emph{caupaṭī} in the first \emph{pāda}, McGregor (1994: s.v.) and Callewaert (2009: s.v.) in their dictionaries of modern and old Hindi both give the meaning ``open all around" for \emph{caupaṭ}.
@@
 
-\begin{metre}[hp04_014]
+\begin{metre}[hp04_017]
 Śālinī 
@@
 %%%%%%%%%%
-\subsection*{4.15}
-\begin{translation}[hp04_015]
+\subsection*{4.18}
+\begin{translation}[hp04_018]
 Having forever abandoned all worry and all activity, as a result of meditating on nothing but the inner sound the mind dissolves into the inner sound.
@@
 
-%\begin{sources}[hp04_015]
-%\end{sources}
-
-\begin{testimonia}[hp04_015]
+%\begin{sources}[hp04_018]
+%\end{sources}
+
+\begin{testimonia}[hp04_018]
 \emph{Haṭharatnāvalī} 4.13
@@
 
-\begin{philcomm}[hp04_015]
+\begin{philcomm}[hp04_018]
 %The occurrence of two finite verbs in the second hemistich of some witnesses (groups 2 and 3) is infelicitous and perhaps points to an error that arose from the \textit{nādam evānusandhānān} reading in the group one mss.
@@
 %%%%%%%%%%
-\subsection*{4.16}
-\begin{translation}[hp04_016]
+\subsection*{4.19}
+\begin{translation}[hp04_019]
 Beginning (\emph{ārambhaḥ}), union (\emph{ghaṭaḥ}), accumulation (\emph{paricayaḥ}) and completion (\emph{niṣpattiḥ}): those are the stages of yoga in all yogas.
@@
 
-\begin{sources}[hp04_016]
+\begin{sources}[hp04_019]
 \emph{Amaraugha} 34
@@
 
-\begin{testimonia}[hp04_016]
+\begin{testimonia}[hp04_019]
 \emph{Haṭharatnāvalī} 4.17
@@
 
-\begin{philcomm}[hp04_016]
+\begin{philcomm}[hp04_019]
 On these four stages, see Birch 2019: 968–969; Mallinson and Szanto 2021:19–20.
@@
 %%%%%%%%%%
-\subsection*{4.17}
-\begin{translation}[hp04_017]
-Among these, the beginning stage is [as follows]:\\
+\subsection*{4.20 heading}
+\begin{translation}[hp04_020a]
+Among these, the beginning stage is [as follows]:
+\end{translation}
+
+% \begin{philcomm}[hp04_020a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.20}
+\begin{translation}[hp04_020]
 As a result of the piercing of the knot of Brahmā, bliss arises in the void [and] an unstruck sound which has various tones is heard in the body.
@@
 
-\begin{sources}[hp04_017]
+\begin{sources}[hp04_020]
 \emph{Amaraugha} 35
@@
 
-\begin{testimonia}[hp04_017]
+\begin{testimonia}[hp04_020]
 \emph{Haṭharatnāvalī} 4.18
@@
 
-\begin{philcomm}[hp04_017]
+\begin{philcomm}[hp04_020]
 In \emph{Jyotsnā} 4.70, Brahmānanda understands the voids associated with each stage to be places in the body.
-However, in the \emph{Amṛtasiddhi} and \emph{Amaraugha}, the source text of this verse, the series of voids (along with their respective blisses and sounds) derives from a tetrad of meditative voids in Vajrayāna traditions (Mallinson and Szanto 2021: 18, Birch 2019: 968).
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.18}
-\begin{translation}[hp04_018]
+However, in the \emph{Amṛtasiddhi} and \emph{Amaraugha}, the source text of this verse, the series of voids (along with their respective blisses and sounds) derives from a tetrad of meditative voids in Vajrayāna traditions (Mallinson and Szanto 2021: 18; Birch 2019: 968).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.21}
+\begin{translation}[hp04_021]
 With a divine body, radiant, smelling heavenly, free from disease [and] his heart full [of bliss], in the void in the beginning [stage] the practitioner becomes a yogi.
@@
 
-\begin{sources}[hp04_018]
+\begin{sources}[hp04_021]
 \emph{Amaraugha} verse 36
@@
 
-\begin{testimonia}[hp04_018]
+\begin{testimonia}[hp04_021]
 \emph{Haṭharatnāvalī} 4.19
@@
 
-%\begin{philcomm}[hp04_018]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.19}
-\begin{translation}[hp04_019]
-Now the unified stage:\\
+%\begin{philcomm}[hp04_021]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.22 heading}
+\begin{translation}[hp04_022a]
+Now the unified stage:
+\end{translation}
+
+% \begin{philcomm}[hp04_022a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.23}
+\begin{translation}[hp04_022]
 In the second stage the breath, after bringing about union, goes into the middle [channel]. Then the yogi has a firm posture [and] he becomes a gnostic and equal to a god.
@@
 
-\begin{sources}[hp04_019]
+\begin{sources}[hp04_022]
 \emph{Amaraugha} 37
@@
 
-\begin{testimonia}[hp04_019]
+\begin{testimonia}[hp04_022]
 \emph{Haṭharatnāvalī} 4.20
@@
 
-\begin{philcomm}[hp04_019]
-The name of the second stage, \emph{ghaṭa} (and the related form \emph{ghaṭīkṛtya} found in the first \emph{pāda} of this verse) can be understood in at least three ways, as union, activation or pot (the latter with an alchemical connotation, on which see 3.12 and Mallinson and Szanto 2021:20–23).
+\begin{philcomm}[hp04_022]
+The name of the second stage, \emph{ghaṭa} (and the related form \emph{ghaṭīkṛtya} found in the first \emph{pāda} of this verse) can be understood in at least three ways, as union, activation or pot (the latter with an alchemical connotation, on which see 3.12 and Mallinson and Szanto 2021: 20–23).
 In the \emph{Dattātreyayogaśāstra} (verse 90) its primary meaning is union and the united pairs are \emph{prāṇa} and \emph{apāna}, \emph{jīvātman} and \emph{paramātman}, and \emph{nāda} and \emph{bindu}.
@@
 %%%%%%%%%%
-\subsection*{4.20}
-\begin{translation}[hp04_020]
-Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a \textit{bherī} drum occurs.%
-\end{translation}
-
-\begin{sources}[hp04_020]
+\subsection*{4.23}
+\begin{translation}[hp04_021]
+Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a kettle drum occurs.%
+\end{translation}
+
+\begin{sources}[hp04_021]
 \emph{Amaraugha} 38
@@
 
-\begin{testimonia}[hp04_020]
+\begin{testimonia}[hp04_021]
 \emph{Haṭharatnāvalī} 4.21
@@
 
-\begin{philcomm}[hp04_020]
+\begin{philcomm}[hp04_021]
 %V3 and HR have the reading of the \textit{Amaraugha} (\textit{atiśūnye vimardaś ca}), which was adapted from the Buddhist concept of a series of four void-like meditative states and moments.
@@
 %%%%%%%%%%
-\subsection*{4.21}
-\begin{translation}[hp04_021]
-Now the accumulation stage:\\
-In the third [stage], having pierced [the knot of Viṣṇu], the sound of a \textit{mardala} drum arises in space. Then [the yogi] reaches the great void, the abode of all supernatural powers.%
+\subsection*{4.24 heading}
+\begin{translation}[hp04_024a]
+Now the accumulation stage:
+\end{translation}
+
+% \begin{philcomm}[hp04_024a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.24}
+\begin{translation}[hp04_024]
+In the third [stage], having pierced [the knot of Viṣṇu], the sound of a bass drum arises in space. Then [the yogi] reaches the great void, the abode of all supernatural powers.%
 \end{translation}
@@
 
-\begin{sources}[hp04_021]
+\begin{sources}[hp04_024]
 \emph{Amaraugha} 39
@@
 
-\begin{testimonia}[hp04_021]
+\begin{testimonia}[hp04_024]
 \emph{Haṭharatnāvalī} 4.22
@@
 
-\begin{philcomm}[hp04_021]
-Here the object of piercing is unspecified, but it is likely to refer to the knot of Viṣṇu mentioned in the previous verse.
+\begin{philcomm}[hp04_024]
+Here the object of piercing is unspecified, but it is likely to refer to the knot of Viṣṇu mentioned in the previous verse.\lb
 
@@
 %%%%%%%%%%
-\subsection*{4.22}
-\begin{translation}[hp04_022]
+\subsection*{4.25}
+\begin{translation}[hp04_025]
 Having overcome the [supreme] bliss of the mind, there arises innate bliss. [The yogi] becomes free of disease, suffering, old age, death, hunger and sleep.
 \end{translation}
-
-\begin{sources}[hp04_022]
+% MD: cittānanda or cidānanda?
+
+\begin{sources}[hp04_025]
 \emph{Amaraugha} 40
@@
 
-\begin{testimonia}[hp04_022]
+\begin{testimonia}[hp04_025]
 \emph{Haṭharatnāvalī} 4.23
@@
 
-\begin{philcomm}[hp04_022]
+\begin{philcomm}[hp04_025]
 The reading of the first verse quarter, \emph{cittānāndaṃ tato jitvā}, is likely a patch for the somewhat obscure \emph{paramānandarocitvāt} in the \emph{Amaraugha} (40a), which is the source text.
@@
 %%%%%%%%%%
-\subsection*{4.23}
-\begin{translation}[hp04_023]
-Now the perfection stage:\\
+\subsection*{4.26 heading}
+\begin{translation}[hp04_026a]
+Now the perfection stage:
+\end{translation}
+
+% MD: niṣpattyavasthā -> niṣpattiḥ?
+
+% \begin{philcomm}[hp04_026a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.26}
+\begin{translation}[hp04_026]
 Then, having pierced Rudra's knot, the breath goes to all the seats [of the deities in the body]. In the perfected [stage] the sound of a flute becomes the sound of a resonating lute.
@@
 
-\begin{sources}[hp04_023]
+\begin{sources}[hp04_026]
 \emph{Amaraugha} 41
@@
 
-\begin{testimonia}[hp04_023]
+\begin{testimonia}[hp04_026]
 \emph{Haṭharatnāvalī} 4.24
@@
 
-\begin{philcomm}[hp04_023]
-The reading we have adopted for the fourth verse quarter, \emph{kvaṇadvīṇākvaṇo}, Birch’s conjecture in his edition of the \emph{Amaraugha} is supported by several testimonia and \emph{Amṛtasiddhi} 31.2, where it is said that the sound of a \emph{vīṇā} arises in the fourth stage.
+\begin{philcomm}[hp04_026]
+The reading we have adopted for the fourth verse quarter, \emph{kvaṇadvīṇākvaṇo}, Birch’s conjecture in his edition of the \emph{Amaraugha} is supported by several testimonia and \emph{Amṛtasiddhi} 31.2, where it is said that the sound of a \emph{vīṇā} arises in the fourth stage.\lb
 
@@
 %%%%%%%%%%
-\subsection*{4.24}
-\begin{translation}[hp04_024]
-Then the mind having become one [with the sound]. This is called Rājayoga. He becomes a creator and destroyer, an equal to a lord among yogis.%
-\end{translation}
-
-\begin{sources}[hp04_024]
+\subsection*{4.27}
+\begin{translation}[hp04_027]
+Then the mind becomes one [with the sound]. This is called Rājayoga. He becomes a creator and destroyer, an equal to a lord among yogis.%
+\end{translation}
+
+\begin{sources}[hp04_027]
 \emph{Amaraugha} 42
@@
 
-\begin{testimonia}[hp04_024]
+\begin{testimonia}[hp04_027]
 \emph{Yogacintāmaṇi} f. 25
@@
 
-\begin{philcomm}[hp04_024]
+\begin{philcomm}[hp04_027]
 %The bahuvrīhi rājayogābhidhānakaṃ means that the mind is called rājayoga, which is very unusual. Perhaps °bhidhāyakaṃ is better, meaning "expresses rājayoga", but it is only in J7 and V3.
@@
 %%%%%%%%%%
-\subsection*{4.24*1}
-%\begin{translation}[hp04_024_1]
-%\end{translation}
-
-%\begin{sources}[hp04_024_1]
-%\end{sources}
-
-%\begin{testimonia}[hp04_024_1]
+\subsection*{4.27*1}
+\begin{translation}[hp04_027_1]
+The dissolution which arises from the inner sound  instantly gives proof of [its efficacy] [and] is an easy method for attaining the state of Rājayoga [even] for foolish people.
+\end{translation}
+
+%\begin{sources}[hp04_027_1]
+%\end{sources}
+
+%\begin{testimonia}[hp04_027_1]
 %\end{testimonia}
 
-\begin{philcomm}[hp04_024_1]
-See 4.32.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.25}
-\begin{translation}[hp04_025]
+%\begin{philcomm}[hp04_027_1]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.28}
+\begin{translation}[hp04_028]
 Whether or not this is liberation, in this very state a great, unbroken pleasure, which is rich in the nectar of absorption, is attained from Rājayoga.
@@
 
-%\begin{sources}[hp04_025]
-%\end{sources}
-
-\begin{testimonia}[hp04_025]
+%\begin{sources}[hp04_028]
+%\end{sources}
+
+\begin{testimonia}[hp04_028]
 \emph{Haṭharatnāvalī} 4.16
@@
 
-%\begin{philcomm}[hp04_025]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.26}
-\begin{translation}[hp04_026]
+%\begin{philcomm}[hp04_028]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.29}
+\begin{translation}[hp04_029]
 Rājayoga without Haṭha and Haṭha without Rājayoga do not succeed so [the yogi] should practise both until the perfection stage.
@@
 
-\begin{sources}[hp04_026]
+\begin{sources}[hp04_029]
 \emph{Śivasaṃhitā} 5.222
@@
 
-\begin{testimonia}[hp04_026]
+\begin{testimonia}[hp04_029]
 \emph{Haṭharatnāvalī} 1.19
@@
 
-%\begin{philcomm}[hp04_026]
-%\end{philcomm}
-
-\begin{metre}[hp04_026]
+%\begin{philcomm}[hp04_029]
+%\end{philcomm}
+
+\begin{metre}[hp04_029]
 Anuṣṭubh (a: ra-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.27}
-\begin{translation}[hp04_027]
+\subsection*{4.30}
+\begin{translation}[hp04_030]
 I consider those who are ignorant of Rājayoga and only work at Haṭha to be like farmhands who get no reward for their efforts.%
@@
 
-%\begin{sources}[hp04_027]
-%\end{sources}
-
-\begin{testimonia}[hp04_027]
+%\begin{sources}[hp04_030]
+%\end{sources}
+
+\begin{testimonia}[hp04_030]
 \emph{Haṃsavilasa} p.~49
@@
 
-%\begin{philcomm}[hp04_027]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.28}
-\begin{translation}[hp04_028]
-The supreme reality is the seed, Haṭha the ground and indifference water. With these three the creeper of good fortune, the no-mind state, immediately grows.
-\end{translation}
-
-%\begin{sources}[hp04_028]
-%\end{sources}
-
-\begin{testimonia}[hp04_028]
+%\begin{philcomm}[hp04_030]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.31}
+\begin{translation}[hp04_031]
+The supreme reality is the seed, Haṭha the ground and indifference water. With these three the wish-fulfilling vine that is the beyond-mind state immediately grows.
+\end{translation}
+%  JB: wish-Fulfilling vine for kalpalatikā?
+% PhD thesis (2011).
+Lin, Nancy Grace.
+"Adapting the Buddha's Biographies: A Cultural History of the Wish-Fulfilling Vine in Tibet, Seventeenth to Eighteenth Centuries." A cultural history of the Kṣemendra's Bodhisattvāvadānakalpalatā in Tibet.
+
+%\begin{sources}[hp04_031]
+%\end{sources}
+
+\begin{testimonia}[hp04_031]
 \emph{Yogacintāmaṇi} f. 24r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_028]
+\begin{philcomm}[hp04_031]
 The meaning of \emph{tattva} here is unclear.
@@
 However, this interpretation seems somewhat contrived.
-We have understood \emph{tattva} in the sense of the 'highest reality' (\emph{paramatattva}) on the assumption that, as the seed, it is the latent cause of the no-mind state.
-
-\end{philcomm}
-
-\begin{metre}[hp04_028]
+We have understood \emph{tattva} in the sense of the `highest reality' (\emph{paramatattva}) on the assumption that, as the seed, it is the latent cause of the state beyond mind.
+
+\end{philcomm}
+
+\begin{metre}[hp04_031]
 Anuṣṭubh (c: na-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.29}
-\begin{translation}[hp04_029]
-The sovereign yoga (\emph{rājayoga}), meditative absorption (\emph{samādhi}), beyond mind (\emph{unmanī}), transmental (\textit{manonmanī}), [the sovereign yoga of] the lineage of immortals (\emph{amaraugha}), non-dual (\emph{advaita}), without support (\emph{nirālamba}), pure (\emph{nirañjana}), [\dots]
+\subsection*{4.32}
+\begin{translation}[hp04_032]
+The sovereign yoga (\emph{rājayoga}), meditative absorption (\emph{samādhi}), the beyond mind state (\emph{unmanī}), transmental state (\textit{manonmanī}), [the sovereign yoga of] the lineage of immortals (\emph{amaraugha}), non-duality (\emph{advaita}), without support (\emph{nirālamba}), pure (\emph{nirañjana}), [\dots]
 \end{translation}
@@
 
-%\begin{sources}[hp04_029]
-%\end{sources}
-
-\begin{testimonia}[hp04_029]
+%\begin{sources}[hp04_032]
+%\end{sources}
+
+\begin{testimonia}[hp04_032]
 \emph{Yogacintāmaṇi} f.~6r
@@
 
-%\begin{philcomm}[hp04_029]
+%\begin{philcomm}[hp04_032]
 %\emph{amaraughaughacāndrī} appears to be a mistake perhaps because the word \emph{amaraugha} became \emph{amarolī}, hence prompting the change to \emph{candrī}.
@@
 %%%%%%%%%%
-\subsection*{4.30}
-\begin{translation}[hp04_030]
-[\dots] the no-mind [state] (\emph{amanaska}), dissolution [of mind] (\emph{laya}), the [supreme] reality (\emph{tattva}), void and not void (\textit{śūnyāśūnya}), the highest state (\emph{para pada}), liberation in life (\emph{jīvanmukti}), innate (\emph{sahaja}) and the fourth [state] (\emph{turya}) are synonyms.
+\subsection*{4.33}
+\begin{translation}[hp04_033]
+[\dots]  no-mind (\emph{amanaska}), dissolution [of mind] (\emph{laya}), the [ultimate] reality (\emph{tattva}), void and not void (\textit{śūnyāśūnya}), the highest state (\emph{para pada}), liberation in life (\emph{jīvanmukti}), innate (\emph{sahaja}) and the fourth [state] (\emph{turya}) are synonyms.
 \end{translation}
@@
 
-\begin{sources}[hp04_030]
+\begin{sources}[hp04_033]
 \emph{Yogacintāmaṇi} f.~6r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{testimonia}[hp04_030]
+%\begin{testimonia}[hp04_033]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_030]
+%\begin{philcomm}[hp04_033]
 %tattva may have dropped out of Gr2 and 3 because of the awkwardness of including it after tattvaṃ bījam in the previous verse (if tattva was considered something else).
@@
 
-\begin{metre}[hp04_030]
+\begin{metre}[hp04_033]
 Anuṣṭubh (c: na-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.31}
-\begin{translation}[hp04_031]
-Two paths for the quick attainment of \emph{unmanī} are approved by me: \emph{tattva}, the supreme bliss, and cultivating the inner sound...
-\end{translation}
-
-%\begin{sources}[hp04_031]
-%\end{sources}
-
-\begin{testimonia}[hp04_031]
+\subsection*{4.34}
+\begin{translation}[hp04_034]
+Two paths for the quick attainment of the beyond-mind state are approved by me: [cultivating] the ultimate reality (\emph{tattva}) or supreme pleasure. And focusing on the inner sound...
+\end{translation}
+
+% MD: pāda d: comment on ca vs vā?
+% JB: the ultimate reality [of Rājayoga], which is supreme bliss, or the cultivation of inner resonance, (either .... or... if we read vā in c and d)
+
+%\begin{sources}[hp04_034]
+%\end{sources}
+
+\begin{testimonia}[hp04_034]
 \emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_031]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.32}
-\begin{translation}[hp04_032]
-which is approved even for stupid people whose minds are immersed in [worldly] pleasures. Dissolution which instantly bestows bliss arises from the inner sound.
-\end{translation}
-
-%\begin{sources}[hp04_032]
-%\end{sources}
-
-\begin{testimonia}[hp04_032]
+%\begin{philcomm}[hp04_034]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.35}
+\begin{translation}[hp04_035]
+ is approved even for foolish people whose minds are intent upon pleasure. The dissolution which arises from the inner sound instantly bestows bliss.
+\end{translation}
+% JB [worldly] pleasures -> [worldly] bliss ... translate saukhya the same way as previous verse?
+
+%\begin{sources}[hp04_035]
+%\end{sources}
+
+\begin{testimonia}[hp04_035]
 \emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_032]
+%\begin{philcomm}[hp04_035]
 %J5 has saṃmataḥ, agreeing with layaḥ
@@
 %%%%%%%%%%
-\subsection*{4.32*1}
-\begin{translation}[hp04_032_1]
-There is one seed [syllable] consisting of creation; one \emph{mudrā}, \emph{khecarī}, one god, the unsupported, [and] one state, mind beyond the mind.
-\end{translation}
-
-\begin{sources}[hp04_032_1]
+\subsection*{4.35*1}
+\begin{translation}[hp04_035_1]
+There is one seed [syllable] consisting of creation and one \emph{mudrā}, \emph{khecarī}, one god, the unsupported, [and] one state, mind beyond the mind.
+\end{translation}
+
+\begin{sources}[hp04_035_1]
 Cf. \emph{Tantrāloka} 32.64
@@
 
-\begin{testimonia}[hp04_032_1]
+\begin{testimonia}[hp04_035_1]
 \emph{Haṭharatnāvalī} 4.28
@@
 
-\begin{philcomm}[hp04_032_1]
-See 3.48.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.32*2}
-\begin{translation}[hp04_032_2]
+\begin{philcomm}[hp04_035_1]
+See 3.49.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.35*2}
+\begin{translation}[hp04_035_2]
 [The yogi] never hears the sounds of [even] conch shells and large drums. As a result of the state of no mind, the body assuredly becomes as [insentient as a piece of] wood.
@@
 
-\begin{sources}[hp04_032_2]
+\begin{sources}[hp04_035_2]
 \emph{Jñānasāra} 3.7
@@
 
-\begin{testimonia}[hp04_032_2]
+\begin{testimonia}[hp04_035_2]
 \emph{Haṭhasaṅketacandrika} f.~120v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_032_2]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.32*3}
-\begin{translation}[hp04_032_3]
+%\begin{philcomm}[hp04_035_2]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.35*3}
+\begin{translation}[hp04_035_3]
 Free from all states [of mind] and all thought, the yogi is as if dead. He is liberated. In this there is no doubt.
@@
 
-%\begin{sources}[hp04_032_3]
-%\end{sources}
-
-\begin{testimonia}[hp04_032_3]
+%\begin{sources}[hp04_035_3]
+%\end{sources}
+
+\begin{testimonia}[hp04_035_3]
 \emph{Nādabindūpaniṣad} 51cd-52ab
@@
 \tl{sarvāvasthāvinirmuktaḥ sarvacintāvivarjitaḥ |\\+}
-\tl{mṛtavattiṣṭhate yogī sa mukto nātra saṃśayaḥ ||\\!}
-\end{versinnote}
-\end{testimonia}
-
-%\begin{philcomm}[hp04_032_3]
+\tl{mṛtavat tiṣṭhate yogī sa mukto nātra saṃśayaḥ ||\\!}
+\end{versinnote}
+\end{testimonia}
+
+%\begin{philcomm}[hp04_035_3]
 %Maybe kāṣṭhavat crept in as a repetition of the previous verse.
@@
 %%%%%%%%%%
-\subsection*{4.32*4}
-\begin{translation}[hp04_032_4]
+\subsection*{4.35*4}
+\begin{translation}[hp04_035_4]
 The yogi in \textit{samādhi} experiences neither cold nor heat, neither suffering nor pleasure, neither praise nor scorn.
@@
 
-\begin{sources}[hp04_032_4]
+\begin{sources}[hp04_035_4]
 \emph{Vivekamārtaṇḍa} 166
@@
 
-%\begin{testimonia}[hp04_032_4]
+%\begin{testimonia}[hp04_035_4]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_032_4]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.32*5}
-\begin{translation}[hp04_032_5]
+%\begin{philcomm}[hp04_035_4]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.35*5}
+\begin{translation}[hp04_035_5]
 The yogi in \textit{samādhi} cannot be wounded by any weapon, killed by any person or overpowered by mantras and magic.
@@
 
-\begin{sources}[hp04_032_5]
+\begin{sources}[hp04_035_5]
 \emph{Vivekamārtaṇḍa} 168
@@
 
-%\begin{testimonia}[hp04_032_5]
+%\begin{testimonia}[hp04_035_5]
 %\end{testimonia}
 
-\begin{philcomm}[hp04_032_5]
+\begin{philcomm}[hp04_035_5]
 The collated manuscripts have \emph{avadhyaḥ} in the first and second verse quarters.
@@
 %%%%%%%%%%
-\subsection*{4.32*6}
-\begin{translation}[hp04_032_6]
+\subsection*{4.35*6}
+\begin{translation}[hp04_035_6]
 The yogi in \textit{samādhi} does not experience smell, taste, form, touch, sound, himself nor anyone else.
@@
 
-\begin{sources}[hp04_032_6]
+\begin{sources}[hp04_035_6]
 \emph{Vivekamārtaṇḍa} 165
@@
 
-\begin{testimonia}[hp04_032_6]
+\begin{testimonia}[hp04_035_6]
 \emph{Yuktabhavadeva} 11.31 (attr. to Gorakṣanātha)
@@
 
-%\begin{philcomm}[hp04_032_6]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.32*7}
-\begin{translation}[hp04_032_7]
+%\begin{philcomm}[hp04_035_6]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.35*7}
+\begin{translation}[hp04_035_7]
 He is indeed truly liberated whose mind is neither asleep nor awake, has no memory nor otherwise, and neither stops nor starts.
@@
 
-\begin{sources}[hp04_032_7]
+\begin{sources}[hp04_035_7]
 \emph{Gorakṣaśataka} 7
@@
 
-\begin{testimonia}[hp04_032_7]
+\begin{testimonia}[hp04_035_7]
 \emph{Haṭhasaṅketacandrika} f.~120v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_032_7]
+\begin{philcomm}[hp04_035_7]
 %The first line is corrupt in the \emph{Gorakṣaśataka} witnesses; its text has been constituted from the \emph{Haṭhapradīpikā}. 
@@
 The second quarter is corrupt in many of the \emph{Haṭhapradīpikā} manuscripts but \emph{smṛti} and \emph{nānyathā} are well attested.
-We have therefore adopted the reading of G11 (group 4c), which in this case is close to the readings of the \textBeta\ group.
-%
-\end{philcomm}
-
-\begin{metre}[hp04_032_7]
+% We have therefore adopted the reading of G11 (group 4c), which in this case is close to the readings of the \textBeta\ group.
+% MD: to delete the last sentence? G11 = ε1 is now fully collated.
+\end{philcomm}
+
+\begin{metre}[hp04_035_7]
 Anuṣṭubh (a: ma-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.32*8}
-\begin{translation}[hp04_032_8]
+\subsection*{4.35*8}
+\begin{translation}[hp04_035_8]
 [The yogi] who remains at ease as though asleep in the waking state, without breathing in and out, is definitely liberated.
@@
 
-\begin{sources}[hp04_032_8]
+\begin{sources}[hp04_035_8]
 \emph{Amanaska} 2.59
@@
 
-\begin{testimonia}[hp04_032_8]
+\begin{testimonia}[hp04_035_8]
 \emph{Kulārṇavatantra} 9.11
@@
 
-%\begin{philcomm}[hp04_032_8]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.33}
-\begin{translation}[hp04_033]
+%\begin{philcomm}[hp04_035_8]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.36}
+\begin{translation}[hp04_036]
 Only the glorious guru lord knows the unique ineffable bliss that has arisen in the hearts of lords among yogis who experience \textit{samādhi} by concentrating on the inner sound.
 \end{translation}
-% JB: Perhaps a note is needed on yogīśvara (here or 4.24). Brahmānanda takes yogīśvarāḥ as yogiṣu yogayukteṣu īśvarāḥ samarthāḥ. i.e., 'lords among yogis' or perhaps 'adepts among yogis'?
+% JB: Perhaps a note is needed on yogīśvara (here or 4.27). Brahmānanda takes yogīśvarāḥ as yogiṣu yogayukteṣu īśvarāḥ samarthāḥ. i.e., 'lords among yogis' or perhaps 'adepts among yogis'?
 %% Jayaratha (TĀ 2.43) yogīśvaraḥ = yoginām apīśvaraḥ 'a lord of even yogis'
 
-\begin{sources}[hp04_033]
+\begin{sources}[hp04_036]
 \emph{Yogatārāvalī} 3
@@
 
-\begin{testimonia}[hp04_033]
+\begin{testimonia}[hp04_036]
 \emph{Haṭharatnāvalī} 4.5
@@
 
-%\begin{philcomm}[hp04_033]
+%\begin{philcomm}[hp04_036]
 %Check consistency on tr. of anusandhāna.
@@
 
-\begin{metre}[hp04_033]
+\begin{metre}[hp04_036]
 Upajāti
@@
 %%%%%%%%%%
-\subsection*{4.33*1}
-\begin{translation}[hp04_033_1]
+\subsection*{4.36*1}
+\begin{translation}[hp04_036_1]
 Seated in the pose of the liberated, the yogi should adopt \emph{śāṃbhavī mudrā} and listen continuously to the inner sound in his right ear.
@@
 
-%\begin{sources}[hp04_033_1]
-%\end{sources}
-
-\begin{testimonia}[hp04_033_1]
+%\begin{sources}[hp04_036_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_036_1]
 \emph{Yogacintāmaṇi} f.~23v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_033_1]
-See 4.10.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.34}
-\begin{translation}[hp04_034]
+\begin{philcomm}[hp04_036_1]
+See 4.13.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.37}
+\begin{translation}[hp04_037]
 [The yogi] who desires yogic sovereignty should abandon all thought and concentrate with an attentive mind on nothing but the internal sound.
@@
 
-%\begin{sources}[hp04_034]
-%\end{sources}
-
-\begin{testimonia}[hp04_034]
+%\begin{sources}[hp04_037]
+%\end{sources}
+
+\begin{testimonia}[hp04_037]
 \emph{Haṭharatnāvalī} 4.14
@@
 
-%\begin{philcomm}[hp04_034]
+%\begin{philcomm}[hp04_037]
 %N3 icchatā
@@
 %%%%%%%%%%
-\subsection*{4.35}
-\begin{translation}[hp04_035]
+\subsection*{4.38}
+\begin{translation}[hp04_038]
 The sage should block his ears with cotton and fix the mind on the sound which he hears, until he attains a state of stillness.
@@
 
-%\begin{sources}[hp04_035]
-%\end{sources}
-
-\begin{testimonia}[hp04_035]
+%\begin{sources}[hp04_038]
+%\end{sources}
+
+\begin{testimonia}[hp04_038]
 \emph{Haṭharatnāvalī} 4.8
@@
 
-\begin{philcomm}[hp04_035]
+\begin{philcomm}[hp04_038]
 %
@@
 The reading \emph{tūlena} (`with cotton'), which is attested by \alphaThree, makes good sense and is close to the reading of \emph{mūlena} in \alphaOne\ and \alphaTwo.
-Manuscripts of several other groups, including \textgamma, instead have \emph{hastābhyām} (`with the hands') or \emph{hastena} (`with the hands').
+Manuscripts of several other groups instead have \emph{hastābhyām} (`with the hands') or \emph{hastena} (`with the hands').
 This reading was inspired by the technique of blocking the ears and other orifices with the fingers in order to listen to the inner sounds.
@@
 %%%%%%%%%%
-\subsection*{4.36}
-\begin{translation}[hp04_036]
+\subsection*{4.39}
+\begin{translation}[hp04_039]
 When this inner sound is being cultivated, it drowns out external sound. After a fortnight the yogi overcomes all distraction and becomes happy.
@@
 
-%\begin{sources}[hp04_036]
-%\end{sources}
-
-\begin{testimonia}[hp04_036]
+%\begin{sources}[hp04_039]
+%\end{sources}
+
+\begin{testimonia}[hp04_039]
 \emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_036]
-%\end{philcomm}
-
-\begin{metre}[hp04_036]
+%\begin{philcomm}[hp04_039]
+%\end{philcomm}
+
+\begin{metre}[hp04_039]
 Anuṣṭubh (a: ma-vipulā; c: na-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.37}
-\begin{translation}[hp04_037]
+\subsection*{4.40}
+\begin{translation}[hp04_040]
 In the first stage of practice, a loud sound of various kinds is heard. Then, as the practice progresses, a quieter and quieter sound is heard.
@@
 
-%\begin{sources}[hp04_037]
-%\end{sources}
-
-\begin{testimonia}[hp04_037]
+%\begin{sources}[hp04_040]
+%\end{sources}
+
+\begin{testimonia}[hp04_040]
 \emph{Haṭharatnāvalī} 4.9
@@
 
-%\begin{philcomm}[hp04_037]
+%\begin{philcomm}[hp04_040]
 %Adopt sūkṣmasūkṣmataḥ (alpha)[MD: Done.]
@@
 %%%%%%%%%%
-\subsection*{4.38}
-\begin{translation}[hp04_038]
+\subsection*{4.41}
+\begin{translation}[hp04_041]
 In the first stage, [the sounds] are those that are produced by the ocean, a [storm] cloud, a kettle drum and a waterfall.
@@
 
-%\begin{sources}[hp04_038]
-%\end{sources}
-
-\begin{testimonia}[hp04_038]
+%\begin{sources}[hp04_041]
+%\end{sources}
+
+\begin{testimonia}[hp04_041]
 \emph{Haṭharatnāvalī} 4.10
@@
 
-\begin{philcomm}[hp04_038]
+\begin{philcomm}[hp04_041]
 We have translated \emph{kāhala} as `trumpet' on the basis of e.g.
@@
 %%%%%%%%%%
-\subsection*{4.39}
-\begin{translation}[hp04_039]
-In the final stage, there are the sounds of little bells, a bamboo flute, a veena and a bee. [These] various sound are heard in the body.
-\end{translation}
-
-%\begin{sources}[hp04_039]
-%\end{sources}
-
-\begin{testimonia}[hp04_039]
+\subsection*{4.42}
+\begin{translation}[hp04_042]
+In the final stage, there are the sounds of little bells, a bamboo flute, a veena and a bee. These various sound are heard in the body.
+\end{translation}
+
+%\begin{sources}[hp04_042]
+%\end{sources}
+
+\begin{testimonia}[hp04_042]
 \emph{Haṭharatnāvalī} 4.11 
@@
 
-%\begin{philcomm}[hp04_039]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.40}
-\begin{translation}[hp04_040]
+%\begin{philcomm}[hp04_042]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.43}
+\begin{translation}[hp04_043]
 Even if a loud noise such as that of a [storm] cloud or kettle drum is being heard, the [yogi] should concentrate on only the very quietest sound in it.
@@
 
-%\begin{sources}[hp04_040]
-%\end{sources}
-
-\begin{testimonia}[hp04_040]
+%\begin{sources}[hp04_043]
+%\end{sources}
+
+\begin{testimonia}[hp04_043]
 \emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_040]
-%\end{philcomm}
-
-\begin{metre}[hp04_040]
+%\begin{philcomm}[hp04_043]
+%\end{philcomm}
+
+\begin{metre}[hp04_043]
 Anuṣṭubh (c: bha-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.41}
-\begin{translation}[hp04_041]
-Or the [yogi] should filter out the gross sound for the subtle, or the subtle for the gross, or, abandoning both, be [focused] on [a sound in] the middle [and] not move the mind elsewhere.%
+\subsection*{4.44}
+\begin{translation}[hp04_044]
+Or, the [yogi] should filter out the gross sound for the subtle, or the subtle for the gross, or, abandoning both, be [focused] on [a sound] in the middle [and] not move the mind elsewhere.%
 \end{translation} %
-
-%\begin{sources}[hp04_041]
-%\end{sources}
-
-\begin{testimonia}[hp04_041]
+% MD: how did we understand syādvā?
+%\begin{sources}[hp04_044]
+%\end{sources}
+
+\begin{testimonia}[hp04_044]
 \emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_041]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.42}
-\begin{translation}[hp04_042]
+%\begin{philcomm}[hp04_044]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.45}
+\begin{translation}[hp04_045]
 Alternatively, the mind fixes upon whatever sound it first attaches to and dissolves together with it.
@@
 
-%\begin{sources}[hp04_042]
-%\end{sources}
-
-\begin{testimonia}[hp04_042]
+%\begin{sources}[hp04_045]
+%\end{sources}
+
+\begin{testimonia}[hp04_045]
 \emph{Yogacintāmaṇi} f.~24r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_042]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.43}
-\begin{translation}[hp04_043]
+%\begin{philcomm}[hp04_045]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.46}
+\begin{translation}[hp04_046]
 Just as a bee drinking nectar has no regard for fragrances, so the mind attached to the inner sound does not desire the objects of the senses.
@@
 
-%\begin{sources}[hp04_043]
-%\end{sources}
-
-\begin{testimonia}[hp04_043]
+%\begin{sources}[hp04_046]
+%\end{sources}
+
+\begin{testimonia}[hp04_046]
 \emph{Haṭharatnāvalī} 4.12
@@
 
-%\begin{philcomm}[hp04_043]
+%\begin{philcomm}[hp04_046]
 %gandhān is the alpha reading and is possible. [MD: adopted]
@@
 %%%%%%%%%%
-\subsection*{4.44}
-\begin{translation}[hp04_044]
+\subsection*{4.47}
+\begin{translation}[hp04_047]
 When the mercury of the mind is bound and has cast off its fickle nature because it has been assimilated with the sulphur of the internal resonance, it attains the immobility called the unsupported (i.e., \textit{samādhi}).
@@
 
-%\begin{sources}[hp04_044]
-%\end{sources}
-
-\begin{testimonia}[hp04_044]
+%\begin{sources}[hp04_047]
+%\end{sources}
+
+\begin{testimonia}[hp04_047]
 \emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_044]
-See Hellwig 2009:204–206 on \emph{khoṭa}, “lame”, which in alchemy is a technical term use to describe mercury that has been processed many times using the \emph{māraṇa} technique and no longer moves.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.44*1}
-\begin{translation}[hp04_044_1]
+\begin{philcomm}[hp04_047]
+See Hellwig 2009: 204–206 on \emph{khoṭa}, “lame”, which in alchemy is a technical term use to describe mercury that has been processed many times using the \emph{māraṇa} technique and no longer moves.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.47*1}
+\begin{translation}[hp04_047_1]
 Bound by the sulphur of the inner sound, the lord that is the mercury of the mind immediately casts off its fickle nature and attains fame as “[the bird] with clipped wings”.
@@
 
-%\begin{sources}[hp04_044_1]
-%\end{sources}
-
-\begin{testimonia}[hp04_044_1]
+%\begin{sources}[hp04_047_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_047_1]
 \emph{Haṭhatattvakaumudī} 54.43
@@
 
-\begin{philcomm}[hp04_044_1]
+\begin{philcomm}[hp04_047_1]
 On \emph{pakṣaccheda} in alchemical processes of immobilizing mercury and for references in Rasaśāstra, see Hellwig 2009: 276–278.
@@
 
-\begin{metre}[hp04_044_1]
+\begin{metre}[hp04_047_1]
 Anuṣṭubh (c: ma-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.45}
-\begin{translation}[hp04_045]
+\subsection*{4.48}
+\begin{translation}[hp04_048]
 As a result of listening to the inner sound, the snake that is the mind forgets everything and, one-pointed, does not dart off anywhere.
@@
 
-%\begin{sources}[hp04_045]
-%\end{sources}
-
-\begin{testimonia}[hp04_045]
+% MD: Why do some good mss have saṃsmṛtya for vismṛtya
+
+%\begin{sources}[hp04_048]
+%\end{sources}
+
+\begin{testimonia}[hp04_048]
 \emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_045]
+\begin{philcomm}[hp04_048]
 The metaphor is that of snake charming: beguiled by the inner sound, the snake that is the mind becomes transfixed.
-Witnesses of the \gamma \ and \delta \ groups have \emph{turaṅgamaḥ} instead of \emph{bhujaṅgamaḥ}, perhaps because forms from \emph{dhāv} are unusual with the latter, but it is found at e.g.~\emph{Garuḍapurāṇa} (1.113.33ab).
+Witnesses of the \textgamma \ and \textdelta \ groups have \emph{turaṅgamaḥ} instead of \emph{bhujaṅgamaḥ}, perhaps because forms from \emph{dhāv} are unusual with the latter, but it is found at e.g.~\emph{Garuḍapurāṇa} (1.113.33ab).\lb
+
+In the third \emph{pāda} we have read against \alphaOne\ and \alphaTwo\ (as well as \betaTwo\ and \epsilonOne), which have \emph{saṃsmṛtya sarvam} instead of \emph{vismṛtya sarvam}.
+The former could be understood to mean “with complete concentration” but we have adopted \emph{vismṛtya} on semantic grounds.
+% JB: but we have adopted \emph{vismṛtya} because its meaning here is more consistent with the description of \textit{nādānusandhāna} in 4.15ab. 
 % Padmapurāṇa
@@
 %%%%%%%%%%
-\subsection*{4.46}
-\begin{translation}[hp04_046]
+\subsection*{4.49}
+\begin{translation}[hp04_049]
 This inner sound is a sharpened goad with the power to restrain the bull elephant in must that is the mind as it wanders about in the garden of the sense objects.
@@
 
-%\begin{sources}[hp04_046]
-%\end{sources}
-
-\begin{testimonia}[hp04_046]
+%\begin{sources}[hp04_049]
+%\end{sources}
+
+\begin{testimonia}[hp04_049]
 \emph{Yogacintāmaṇi} f.~23r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_046]
+\begin{philcomm}[hp04_049]
 The unusual form \emph{niyāmana} is also found in Rasaśāstra works where it occurs in the context of restraining mercury and is a topic of discussion (e.g.~\emph{Rasaprakāśasudhākara} 1.23, \emph{Ānandakanda} 1.4.58–59).
@@
 %%%%%%%%%%
-\subsection*{4.47}
-\begin{translation}[hp04_047]
-The inner sound is a bolt for the swift horse of the mind so the yogi should regularly focus on its cultivation.
-\end{translation}
-
-%\begin{sources}[hp04_047]
-%\end{sources}
-
-\begin{testimonia}[hp04_047]
+\subsection*{4.50}
+\begin{translation}[hp04_050]
+Cultivation of the inner sound is a bolt for [the stable door of] the swift horse of the mind so the yogi should regularly focus on it.
+\end{translation}
+
+%\begin{sources}[hp04_050]
+%\end{sources}
+
+\begin{testimonia}[hp04_050]
 \emph{Haṭhatattvakaumudī} 54.46
@@
 
-\begin{philcomm}[hp04_047]
-We understand \textit{ninādaḥ}, which is the subject of the previous line, as the subject in the first line of this verse.
+\begin{philcomm}[hp04_050]
+%We understand \textit{ninādaḥ}, which is the subject of the previous line, as the subject in the first line of this verse.
 Several witness have \emph{paridhāyate} instead of \emph{parighāyate}.
-The latter is unattested but could mean "is a halter" based on \emph{abhidhānī}’s meaning of “halter”.
-\end{philcomm}
-
-\begin{metre}[hp04_047]
+The latter is unattested but could mean “is a halter" based on \emph{abhidhānī}’s meaning of “halter”.
+\end{philcomm}
+
+\begin{metre}[hp04_050]
 Anuṣṭubh (a: na-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.47*1}
-\begin{translation}[hp04_047_1]
+\subsection*{4.50*1}
+\begin{translation}[hp04_050_1]
 The inner sound is a net for trapping the deer of the mind and a hunter for corraling the antelope of the mind.
 \end{translation}
-% nādo’ntaraṅgasāraṅgabandhane vāgurāyate
-%V15 antaraṅgaturaṅgasya nādo vyādhāyate pi ca
-
-%\begin{sources}[hp04_047_1]
-%\end{sources}
-
-\begin{testimonia}[hp04_047_1]
+%  JB: we have not translated antaraṅga° in pāda a. Change to 'trapping the inner deer that is the mind'? JM: antaraṅga = mind
+% JB: acts as a net (vāgurāyate) or becomes a net
+%?? MD: Pāda c turaṅga is better attested than kuraṅga.
+
+%\begin{sources}[hp04_050_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_050_1]
 \emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_047_1]
+\begin{philcomm}[hp04_050_1]
 %Adopt V15 67cd
@@
 %%%%%%%%%%
-\subsection*{4.48}
-\begin{translation}[hp04_048]
-Striking the deer of the mind when it is transfixed by its attachment to inner sounds such as that of a bell, is very easy if the archer is skilful.
+\subsection*{4.51}
+\begin{translation}[hp04_051]
+Striking the deer of the mind when, focused upon inner sounds such as that of a bell, it is transfixed, is very easy if the archer is skilful.
 \end{translation}
@@
 % praharaṇam atisukaraṃ syāc charasaṃdhātā pravīṇaś cet ||
-%
-
-%\begin{sources}[hp04_048]
-%\end{sources}
-
-%\begin{testimonia}[hp04_048]
+%MD: I would prefer nādasakta to °sakti.
+The latter is not well attested and sounds unnatural to me.
+If you take ghaṇṭā°...°sakta and stabdhāntaḥkaraṇa as a dvandva compound, there is no grammatical problem.
+
+%\begin{sources}[hp04_051]
+%\end{sources}
+
+%\begin{testimonia}[hp04_051]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_048]
+%\begin{philcomm}[hp04_051]
 %N3 has a problem with the metre.
@@
 
-\begin{metre}[hp04_048]
+\begin{metre}[hp04_051]
 Upagīti 
@@
 %%%%%%%%%%
-\subsection*{4.49}
-\begin{translation}[hp04_049]
+\subsection*{4.52}
+\begin{translation}[hp04_052]
 The tone of that sound is that of the unstruck sound. A light is inside the tone [and] the mind is inside the light. That mind dissolves. That is the supreme state of Viṣṇu.
@@
 
-\begin{sources}[hp04_049]
-\emph{Uttaragītā} 41cd–42 %(source) JB: I think we can say that the UG is the source of 4.49*1 (but not 4.49)
+\begin{sources}[hp04_052]
+\emph{Uttaragītā} 41cd–42 %(source) JB: I think we can say that the UG is the source of 4.52*1 (but not 4.52)
 \begin{versinnote}
@@
 
-\begin{testimonia}[hp04_049]
+\begin{testimonia}[hp04_052]
 \emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_049]
+\begin{philcomm}[hp04_052]
 % to discuss.
@@
 %A - dhvaner antargataṃ jyotir jyotirantargataṃ manaḥ (J5,G4,Gr2,Gr3)
-%B - dhvaner antargataṃ jñeyaṃ jñeyasyāntargataṃ manaḥ (N3?,Jyo) JM: the second I think the second. J
+%B - dhvaner antargataṃ jñeyaṃ jñeyasyāntargataṃ manaḥ (N3?,F,Jyo) JM: the second I think the second. J
 
@@
 % We are not sure what the difference between dhvani and śabda is... and have translated them as ...
-\end{philcomm}
-
-% N3 (α1)
-% anāhatasya śabdasya tasya śabdasya ca dhvaniḥ |
-% dhvaner aṃtargataṃ jñeyaṃ yasyāṃ tv aṃtargataṃ manaḥ |
-% tan mano vilayaṃ yāṃti tadviṣṇo paramaṃ padaṃ |
-
-%%%%%%%%%%
-\subsection*{4.49*1}
-\begin{translation}[hp04_049]
+% MD 2024-09: G11 has both versions! It seems that the editor of the extended HP created the 4-pāda version, but kept the 6-pāda version.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.52*1}
+\begin{translation}[hp04_052]
 When the mind dissolves into that which is the most subtle object of perception in the unstruck sound, that is the supreme state of Viṣṇu.
@@
 
-%\begin{sources}[hp04_049]
-%\end{sources}
-
-\begin{testimonia}[hp04_049*1]
+%\begin{sources}[hp04_052]
+%\end{sources}
+
+\begin{testimonia}[hp04_052*1]
 \emph{Hathatattvakaumudī} 54.48
@@
 
-%\begin{philcomm}[hp04_049*1]
+%\begin{philcomm}[hp04_052*1]
 %HI: tan could mean tatra.
@@
 %%%%%%%%%%
-\subsection*{4.50}
-\begin{translation}[hp04_050]
+\subsection*{4.53}
+\begin{translation}[hp04_053]
 As long as sound exists, there is a concept of space. That which is soundless is the supreme Brahman and is called the supreme self.
@@
 
-\begin{sources}[hp04_050]
+\begin{sources}[hp04_053]
 \emph{Vivekamārtaṇḍa} (six-chapters) 5.15 %JB: This was in testimonia but it must be the source as we now know that Svātmārāma used the 6ch VM.
@@
 
-\begin{testimonia}[hp04_050]
+\begin{testimonia}[hp04_053]
 \emph{Yogacintāmaṇi} f.~27r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_050]
-%\end{philcomm}
-%%%%%%%%%%
-\subsection*{4.51}
-\begin{translation}[hp04_051]
+%\begin{philcomm}[hp04_053]
+%\end{philcomm}
+%%%%%%%%%%
+\subsection*{4.54}
+\begin{translation}[hp04_054]
 Whatever is heard as the inner sound is nothing but Śakti. The formless one which hears it is nothing but the supreme lord.
@@
 
-%\begin{sources}[hp04_051]
-%\end{sources}
-
-\begin{testimonia}[hp04_051]
+%\begin{sources}[hp04_054]
+%\end{sources}
+
+\begin{testimonia}[hp04_054]
 \emph{Yogacintāmaṇi} f.~27r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_051]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.52}
-\begin{translation}[hp04_052]
+%\begin{philcomm}[hp04_054]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.55}
+\begin{translation}[hp04_055]
 Blocking of the ears, mouth, eyes, and nose should not be performed [when] a clear inner sound is heard distinctly in the purified Suṣumṇā channel.
@@
 
-%\begin{sources}[hp04_052]
-%\end{sources}
-
-\begin{testimonia}[hp04_052]
+%\begin{sources}[hp04_055]
+%\end{sources}
+
+\begin{testimonia}[hp04_055]
 \emph{Yogacintāmaṇi} f.~26v
@@
 
-\begin{philcomm}[hp04_052]
+\begin{philcomm}[hp04_055]
 %N3 is missing an akṣara (nirodhaṃ aiva)
@@
 
-\begin{metre}[hp04_052]
+\begin{metre}[hp04_055]
 Upagīti 
@@
 %%%%%%%%%%
-\subsection*{4.52*1}
-\begin{translation}[hp04_052_1]
-The inner sound is called Śakti. And gnosis of the inner sound is Sadāśiva. When gnosis of the inner sound has disappeared, only the beyond-mind [state] remains.
+\subsection*{4.55*1}
+\begin{translation}[hp04_055_1]
+The inner sound is called Śakti; knowledge of the inner sound is Sadāśiva. But when knowledge and the object of knowledge have disappeared, only the beyond-mind [state] remains.
 \end{translation}
@@
 
-%\begin{sources}[hp04_052_1]
-%\end{sources}
-
-\begin{testimonia}[hp04_052_1]
+%\begin{sources}[hp04_055_1]
+%\end{sources}
+
+\begin{testimonia}[hp04_055_1]
 \emph{Upāsanāsārasaṅgraha} f. 107 (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_052_1]
+%\begin{philcomm}[hp04_055_1]
 %V15 reading for the third pāda is better than N19.
-JB: If we adopt the epsilon 3 (V15) for the third pāda, then we have to adopt tad unmany eva śiṣyate in the fourth (otherwise 4th pāda is unmetrical with evāvaśiṣyate).
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.52*2}
-\begin{translation}[hp04_052_2]
+JB: If we adopt the zeta 3 (V15) for the third pāda, then we have to adopt tad unmany eva śiṣyate in the fourth (otherwise 4th pāda is unmetrical with evāvaśiṣyate).
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.55*2}
+\begin{translation}[hp04_055_2]
 As long as there is the inner sound there is mind. At the end of the inner sound the mind beyond mind state [arises]. The void is said to be sonorous and Brahma is silent.
@@
 
-%\begin{sources}[hp04_052_2]
-%\end{sources}
-
-\begin{testimonia}[hp04_052_2]
+%\begin{sources}[hp04_055_2]
+%\end{sources}
+
+\begin{testimonia}[hp04_055_2]
 \emph{Haṭhatattvakaumudī} 54.51
@@
 
-%\begin{philcomm}[hp04_052_2]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.52*3}
-\begin{translation}[hp04_052_3]
+%\begin{philcomm}[hp04_055_2]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.55*3}
+\begin{translation}[hp04_055_3]
 When the store of subliminal impressions has been destroyed as a result of continuously concentrating on the inner sound, the mind and breath are sure to dissolve into the untainted [god] (\emph{nirañjane}).
@@
 
-%\begin{sources}[hp04_052_3]
-%\end{sources}
-
-\begin{testimonia}[hp04_052_3]
+%\begin{sources}[hp04_055_3]
+%\end{sources}
+
+\begin{testimonia}[hp04_055_3]
 %\emph{Upāsanāsārasaṅgraha} f. 107 Jb: problematic as most of the first line is missing in ms. no. 12170 and we havent been using the IFP transcript (which does have this verse)
@@
 
-\begin{philcomm}[hp04_052_3]
-The term \emph{nirañjanaḥ} likely refers to \emph{devo nirañjanaḥ} (`untainted god') in the next verse (4.52*4), an expression which refers to the supreme deity.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.52*4}
-\begin{translation}[hp04_052_4]
+\begin{philcomm}[hp04_055_3]
+The term \emph{nirañjanaḥ} likely refers to \emph{devo nirañjanaḥ} (`untainted god') in the next verse (4.52*4), an expression which refers to the supreme deity.\\
+The compound \emph{manamārutau} with the \emph{aiśa} form \emph{mana} is attested in both the \texteta\ and \textepsilon\ groups.
+Cf.~4.11*22b \emph{manamadhyagām} (in a verse taken from the \emph{Candrāvalokana}).
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.55*4}
+\begin{translation}[hp04_055_4]
 Thousands of crores of inner sounds and hundreds of crores of visual focal points all dissolve into the place of the untainted god (\emph{devo nirañjanaḥ}).
@@
 \end{translation}
-
-%\begin{sources}[hp04_052_4]
-%\end{sources}
-
-\begin{testimonia}[hp04_052_4]
+%% JB: does bindu 
+%\begin{sources}[hp04_055_4]
+%\end{sources}
+
+\begin{testimonia}[hp04_055_4]
 \emph{Yogacintāmaṇi} f.~27r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_052_4]
-On the meaning of \emph{devo nirañjana}, see the note to 4.52.3.
+\begin{philcomm}[hp04_055_4]
+On the meaning of \emph{devo nirañjana}, see the note to 4.55*3.
 
@@
 %%%%%%%%%%
-\subsection*{4.52*5}
-%\begin{translation}[hp04_052_5]
-%\end{translation}
-
-%\begin{sources}[hp04_052_5]
-%\end{sources}
-
-%\begin{testimonia}[hp04_052_5]
+\subsection*{4.55*4 ending}
+% iti nādānusaṃdhānam/
+\begin{translation}[hp04_055_4p]
+\end{translation}
+
+% \begin{philcomm}[hp04_055_4p]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.55*5}
+\begin{translation}[hp04_055_5]
+All the methods of Haṭha and Laya [should be practised] until the attainment of the state of Rājayoga. Having attained the state of Rājayoga, [the yogi] becomes untainted.
+\end{translation}
+
+%\begin{sources}[hp04_055_5]
+%\end{sources}
+
+%\begin{testimonia}[hp04_055_5]
 %\end{testimonia}
 
-\begin{philcomm}[hp04_052_5]
-See 4.74.
-\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.53}
-\begin{translation}[hp04_053]
+% \begin{philcomm}[hp04_055_5]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.56}
+\begin{translation}[hp04_056]
 Enough of the verbosity of a learned gathering! O friend, listen to this instruction formerly taught by Śiva for awakening Matsyendra.
 \end{translation}
-
-%\begin{sources}[hp04_053]
-%\end{sources}
-
-\begin{testimonia}[hp04_053]
+%JB shouldnt it be kāṣṭhāgoṣṭhī - as kāṣṭhagoṣṭhī means a wooden assembly [or now I'm thinking... wooden cowpen). I think we need a note.
+
+%\begin{sources}[hp04_056]
+%\end{sources}
+
+\begin{testimonia}[hp04_056]
 \emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_053]
+\begin{philcomm}[hp04_056]
 This verse may have been composed by Svātmārāma to introduce the next two verses, which are from the \emph{Candrāvalokana}, a dialogue between Matsyendra and Śiva.
@@
 %%%%%%%%%%
-\subsection*{4.54}
-\begin{translation}[hp04_054]
+\subsection*{4.57}
+\begin{translation}[hp04_057]
 As long as the moving breath does not enter the middle path, as long as bindu is not stable, restrained by the breath, as long as [realisation of] the ultimate truth (\emph{tattvam}), which is as natural as the sky, does not arise, then all that one says is deceitful, and false prattle.
-\end{translation}%?? tr of 3rd pāda. JM thinks it's "doesn't become as natural as...". Current formulation doesn't work for me (if it's sahaja, it has arisen already)
-
-\begin{sources}[hp04_054]
+\end{translation}% tr of 3rd pāda. JM thinks it's "doesn't become as natural as...". Current formulation doesn't work for me (if it's sahaja, it has arisen already)
+%% JB: Maybe we should understand tattva here as samādhi (Brahmānanda takes it as citta: yāvat tattvaṃ cittaṃ dhyāne dhyeyacintane sahajasadṛśaṃ svābhāvikadhyeyākāravṛttipravāhavan naiva jāyate).
+Nonetheless, I think the idea is that all one says is deceitful if the ultimate reality/samādhi, which is naturally similar to the sky, hasn't arisen (in the case of the ultimate reality, hasn't revealed itself to that person).
+%?? MD: Do we keep °prabaddhaḥ in the second line? Or adopt °prabandhaḥ as in the source text?
+
+\begin{sources}[hp04_057]
 \emph{Candrāvalokana} 14
@@
 
-\begin{testimonia}[hp04_054]
+\begin{testimonia}[hp04_057]
 \emph{Yogacintāmaṇi} f.~22a
@@
 
-%\begin{philcomm}[hp04_054]
+%\begin{philcomm}[hp04_057]
 %J5 has sthiraḥ (instead of dṛḍhaḥ) supporting the visarga (rather than dṛdhaṃ N3)
@@
 
-\begin{metre}[hp04_054]
+\begin{metre}[hp04_057]
 Mandākrāntā 
@@
 %%%%%%%%%%
-\subsection*{4.55}
-\begin{translation}[hp04_055]
-Having learnt the correct piercing of Suṣumṇā, [the yogi] should make the breath go into the central channel, take it to the place of the moon and block the nostrils.
+\subsection*{4.58}
+\begin{translation}[hp04_058]
+Having learnt the correct piercing of Suṣumṇā, [the yogi] should make the breath go into the central channel, put it in the place of the moon and block the nostrils.
 \end{translation}
@@
 % 
-
-\begin{sources}[hp04_055]
+% MD 2024-8: The IFP-ms has dhṛtvā taṃ caindavasthāne, dhṛtvā is also supported by α3. "dhṛtvā taṃ caindave sthāne" may be the original reading.
+
+\begin{sources}[hp04_058]
 \emph{Candrāvalokana} 32
@@
 
-\begin{testimonia}[hp04_055]
+\begin{testimonia}[hp04_058]
 \emph{Upāsanāsārasaṅgraha} p.~31 (attr. to the \emph{Candrāvalokana})
@@
 
-%\begin{philcomm}[hp04_055]
-%\end{philcomm}
-
-\begin{metre}[hp04_055]
+%\begin{philcomm}[hp04_058]
+%\end{philcomm}
+
+\begin{metre}[hp04_058]
 Anuṣṭubh (a: ma-vipulā)
 \end{metre}
-%%%%%%%%%%
-\subsection*{4.56}
-\begin{translation}[hp04_056]
-And so, Vasiṣṭha [said]:\\
+
+%%%%%%%%%%
+\subsection*{4.59 heading}
+\begin{translation}[hp04_059a]
+And so, Vasiṣṭha [said]:
+\end{translation}
+
+% \begin{philcomm}[hp04_059a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.59}
+\begin{translation}[hp04_059]
 The moon and sun move in Iḍā and Piṅgalā. The moon is said to be of the nature of \emph{tamas} and the sun of \emph{rajas}.
@@
 
-\begin{sources}[hp04_056]
+\begin{sources}[hp04_059]
 \emph{Vasiṣṭhasaṃhitā} 2.28–29ab
@@
 
-\begin{testimonia}[hp04_056]
+\begin{testimonia}[hp04_059]
 \emph{Haṭharatnāvalī} 4.36cd–37ab
@@
 
-%\begin{philcomm}[hp04_056]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.57}
-\begin{translation}[hp04_057]
+%\begin{philcomm}[hp04_059]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.60}
+\begin{translation}[hp04_060]
 Those two bring about the entirety of time, which consists of night and day. Suṣumṇā consumes time. This which has been taught is secret.
@@
 
-\begin{sources}[hp04_057]
+\begin{sources}[hp04_060]
 \emph{Vasiṣṭhasaṃhitā} 2.29cd–30ab
@@
 
-\begin{testimonia}[hp04_057]
+\begin{testimonia}[hp04_060]
 \emph{Yogacintāmaṇi} (attr. to Yājñavalkya)
@@
 
-%\begin{philcomm}[hp04_057]
-%\end{philcomm}
-
-\begin{metre}[hp04_057]
+\begin{philcomm}[hp04_060]
+The variant readings of \emph{pāda} a which name the sun and moon are likely to have arisen due to the absence of the preceding verse in 
+\textepsilon, \texteta, and \textzeta.
+\end{philcomm}
+
+\begin{metre}[hp04_060]
 Anuṣṭubh (a: bha-vipulā; c: ma-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.58}
-\begin{translation}[hp04_058]
-For as the tetrad of verses called the Saubhadra has it:\\
+\subsection*{4.61 heading}
+\begin{translation}[hp04_061a]
+For as the tetrad of verses called the Saubhadra has it:
+\end{translation}
+
+\begin{philcomm}[hp04_061a]
+We do not know why this tetrad of verses is called Saubhadra.
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.61}
+\begin{translation}[hp04_061]
 There are six cakras, sixteen supports, three focal points and three \emph{guṇa}s. Everything else is [just] the prolixity of texts. Trikūṭa is the supreme place.
@@
 
-%\begin{sources}[hp04_058]
-%\end{sources}
-
-\begin{testimonia}[hp04_058]
+%\begin{sources}[hp04_061]
+%\end{sources}
+
+\begin{testimonia}[hp04_061]
 Cf.~6-chapter \emph{Vivekamārtaṇḍa} 6.3
@@
 
-\begin{philcomm}[hp04_058]
-We do not know why this tetrad of verses is called Saubhadra.\\
-
+\begin{philcomm}[hp04_061]
 The three components of the yogic body listed here are found together in other texts, the earliest being \emph{Netratantra} 7.1ab (\emph{ṛtucakraṃ svarādhāraṃ trilakṣyaṃ vyomapañcakam}).
@@
 %%%%%%%%%%
-\subsection*{4.59}
-\begin{translation}[hp04_059]
+\subsection*{4.62}
+\begin{translation}[hp04_062]
 Kuṇḍalinī is taught to have a curved shape like a snake. She is Śakti. [The yogi] who has made her move is undoubtedly liberated.
@@
 
-%\begin{sources}[hp04_059]
-%\end{sources}
-
-\begin{testimonia}[hp04_059]
+%\begin{sources}[hp04_062]
+%\end{sources}
+
+\begin{testimonia}[hp04_062]
 \emph{Yogacintāmaṇi} f.~79r (attr. to the \emph{Haṭhayoga})
@@
 
-%\begin{philcomm}[hp04_059]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.60}
-\begin{translation}[hp04_060]
+%\begin{philcomm}[hp04_062]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.63}
+\begin{translation}[hp04_061]
 When the \emph{kūṭa} is situated at Trikūṭa [then] the mind is wonderful and uninterrupted. By means of Kuṇḍalinī, [the yogi] is undoubtedly liberated.
@@
 
-%\begin{sources}[hp04_060]
-%\end{sources}
-
-\begin{testimonia}[hp04_060]
+%\begin{sources}[hp04_061]
+%\end{sources}
+
+\begin{testimonia}[hp04_061]
 \emph{Upāsanāsārasaṅgraha} f. 51 (attr. to a \emph{yogaśāstra})
@@
 
-\begin{philcomm}[hp04_060]
+\begin{philcomm}[hp04_061]
 We are unsure of the meaning of \emph{kūṭa} here.
@@
 %%%%%%%%%%
-\subsection*{4.61}
-\begin{translation}[hp04_061]
+\subsection*{4.64}
+\begin{translation}[hp04_064]
 There are seventy-two thousand openings  of the channels in the cage [that is the body]. Suṣumṇā is the Śāmbhavī Śakti while the other [channels] are pointless.
@@
 
-%\begin{sources}[hp04_061]
-%\end{sources}
-
-\begin{testimonia}[hp04_061]
+%\begin{sources}[hp04_064]
+%\end{sources}
+
+\begin{testimonia}[hp04_064]
 \emph{Haṭhasaṅketacandrikā} f.~108r (attr. to the \emph{Haṭhapradīpikā})
@@
 
-\begin{philcomm}[hp04_061]
+\begin{philcomm}[hp04_064]
 The compound \emph{nāḍīdvāra} is not found elsewhere (other than as \emph{nāḍīdvāreṇa}) and its meaning here is unclear.
@@
 %%%%%%%%%%
-\subsection*{4.62}
-\begin{translation}[hp04_062]
+\subsection*{4.65}
+\begin{translation}[hp04_065]
 The breath, having been carefully accumulated, together with fire awakens Kuṇḍalinī and enters Suṣumṇā without obstruction.
@@
 
-\begin{sources}[hp04_062]
+\begin{sources}[hp04_065]
 \emph{Dattātreyayogaśāstra} 108
@@
 
-\begin{testimonia}[hp04_062]
+\begin{testimonia}[hp04_065]
 \emph{Śārṅgadharapaddhati} 4399
@@
 
-%\begin{philcomm}[hp04_062]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.63}
-\begin{translation}[hp04_063]
-When the breath is flowing in Suṣumṇā, the no-mind state is attained. Otherwise [i.e.~if the breath is not flowing in Suṣumṇā], the various practices [of yoga] lead to nothing but exertion for yogis.
+%\begin{philcomm}[hp04_065]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.66}
+\begin{translation}[hp04_066]
+When the breath is flowing in Suṣumṇā, the transmental state is attained.
+Otherwise [i.e.~if the breath is not flowing in Suṣumṇā], the various practices [of yoga] lead to nothing but exertion for yogis.
 \end{translation}%
 
-%\begin{sources}[hp04_063]
-%\end{sources}
-
-\begin{testimonia}[hp04_063]
+%\begin{sources}[hp04_066]
+%\end{sources}
+
+\begin{testimonia}[hp04_066]
 \emph{Upāsanāsārasaṅgraha} p.~108 (attr. to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_063]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.64}
-\begin{translation}[hp04_064]
+%\begin{philcomm}[hp04_066]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.67}
+\begin{translation}[hp04_067]
 The mind is bound by the very same thing that binds the breath and the breath is bound by that which binds the mind.
@@
 
-%\begin{sources}[hp04_064]
-%\end{sources}
-
-\begin{testimonia}[hp04_064]
+%\begin{sources}[hp04_067]
+%\end{sources}
+
+\begin{testimonia}[hp04_067]
 
@@
 
-\begin{philcomm}[hp04_064]
+\begin{philcomm}[hp04_067]
 Brahmānanda understands \emph{yena} here to refer to the yogi. We have taken it to refer to a practice.
@@
 %%%%%%%%%%
-\subsection*{4.65}
-\begin{translation}[hp04_065]
+\subsection*{4.68}
+\begin{translation}[hp04_068]
 The mind has two impulses: past impressions (\emph{vāsanā}) and the breath. When one of those two disappears, both soon disappear.%
@@
 
-\begin{sources}[hp04_065]
+\begin{sources}[hp04_068]
 \emph{Gorakṣaśataka} 9
@@
 
-\begin{testimonia}[hp04_065]
+\begin{testimonia}[hp04_068]
 \emph{Yogakuṇḍalinyupaniṣat} 1
@@
 
-\begin{philcomm}[hp04_065]
+\begin{philcomm}[hp04_068]
 The emendation of \emph{drutaṃ} in the last verse quarter has been made to restore the faulty readings of the \textalpha\ manuscripts (i.e., \emph{dguttaṃ, dhṛtaṃ, dṛtaṃ}), which are similar to an incorrect reading in an important witness of the source text, the \emph{Gorakṣaśataka} (T, \emph{dhṛtaṃ}).
-The emendation to \emph{drutaṃ} is based on the parallel verse in the \emph{Mokṣopāya} and its related recensions, which have \emph{kṣipraṃ} instead.
+The emendation to \emph{drutaṃ} is based on the parallel verse in the \emph{Mokṣopāya} and its related recensions, which have \emph{kṣipraṃ} instead.\lb
 
@@
 %%%%%%%%%%
-\subsection*{4.66}
-\begin{translation}[hp04_066]
+\subsection*{4.69}
+\begin{translation}[hp04_069]
 The breath dissolves where the mind dissolves; the mind dissolves right where the breath dissolves.
@@
 
-%\begin{sources}[hp04_066]
-%\end{sources}
-
-\begin{testimonia}[hp04_066]
+%\begin{sources}[hp04_069]
+%\end{sources}
+
+\begin{testimonia}[hp04_069]
 \emph{Haṭharatnāvalī} 4.29
@@
 
-\begin{philcomm}[hp04_066]
+\begin{philcomm}[hp04_069]
 %HI: tatra = tāvat.
@@
 %%%%%%%%%%
-\subsection*{4.67}
-\begin{translation}[hp04_067]
+\subsection*{4.70}
+\begin{translation}[hp04_070]
  Like water mixed with milk, mind and breath are always joined, behaving in the same way.  As long as there is mind the breath is active, and as long as there is breath the mind is active.
@@
 
-\begin{sources}[hp04_067]
+\begin{sources}[hp04_070]
 \emph{Amanaska} 2.27
@@
 
-\begin{testimonia}[hp04_067]
+\begin{testimonia}[hp04_070]
 \emph{Yogacintāmaṇi} f.~19r (attr. to the \emph{Rājayoga} (aka. \emph{Amanaska})
@@
 
-\begin{philcomm}[hp04_067]
+\begin{philcomm}[hp04_070]
 %These verses (4.66–67) seem to elevate the possible role of the mind in achieving \emph{samādhi} (as opposed to the standard Haṭha method of stopping the breath to achieve samādhi). 
-Complementing his understanding of the previous verse, Brahmānanda has \emph{yato\dots tatra} in 4.67cd, taking it to mean \emph{yatra\dots tatra} and to be referring to cakras.
-\end{philcomm}
-
-\begin{metre}[hp04_067]
+Complementing his understanding of the previous verse, Brahmānanda has \emph{yato\dots tatra} in 4.70cd, taking it to mean \emph{yatra\dots tatra} and to be referring to cakras.
+\end{philcomm}
+
+\begin{metre}[hp04_070]
 Upajāti
@@
 %%%%%%%%%%
-\subsection*{4.68}
-\begin{translation}[hp04_068]
+\subsection*{4.71}
+\begin{translation}[hp04_071]
 As a result of one of those two disappearing the other disappears and as a result of one being active the other is active.
@@
 
-\begin{sources}[hp04_068]
+\begin{sources}[hp04_071]
 \emph{Amanaska} 2.28
@@
 
-\begin{testimonia}[hp04_068]
+\begin{testimonia}[hp04_071]
 \emph{Yogacintāmaṇi} f.~19r (attr. to the \emph{Rājayoga} (aka. \emph{Amanaska})
@@
 
-%\begin{philcomm}[hp04_068]
+%\begin{philcomm}[hp04_071]
 %Destroy etc. for disappear?
@@
 
-\begin{metre}[hp04_068]
+\begin{metre}[hp04_071]
 Upajāti
@@
 %%%%%%%%%%
-\subsection*{4.69}
-\begin{translation}[hp04_069]
-[The yogi becomes] a traveller by the way of the wind [and] obtains the whole earth and the eight lordly powers. True, true is this, O beautiful one.
-\end{translation}
-
-\begin{sources}[hp04_069]
+\subsection*{4.72}
+\begin{translation}[hp04_072]
+When there is no movement in the path of the wind, [the yogi]
+obtains the whole earth and the eight lordly powers. True, true is this, O beautiful one.
+\end{translation}
+
+\begin{sources}[hp04_072]
 \emph{Jñānasāra} 3.5–6
@@
 
-\begin{testimonia}[hp04_069]
+\begin{testimonia}[hp04_072]
 \emph{Haṭhasaṅketacandrikā} f.~117r
@@
 
-%\begin{philcomm}[hp04_069]
+\begin{philcomm}[hp04_072]
+As it is found in its source text, the \emph{Jñānasāra}, this verse says that the yogi flies around the world with the speed of the wind (\emph{vāyuvegena}).
+We have understood Svātmārāma to have edited the verse to reflect the subject of the previous two verses in which the breath is to be stopped.
 %Unidentified source. Jnānasāra removes vocative.
@@
 %labhate needed for aṣṭaguṇaṃ
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.70}
-\begin{translation}[hp04_070]
-Thus, Viśvarūpācārya [said]:\\
+\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.73 heading}
+\begin{translation}[hp04_073a]
+Thus, Viśvarūpācārya [said]:
+\end{translation}
+
+% \begin{philcomm}[hp04_073a]
+% \end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.73}
+\begin{translation}[hp04_073]
 When the breath is destroyed and the mind dissolves, all experience is the same (\emph{samarasatva}). That is called \emph{samādhi}.
@@
 
-\begin{sources}[hp04_070]
+\begin{sources}[hp04_073]
 \emph{Vivekamārtaṇḍa} 163
@@
 
-\begin{testimonia}[hp04_070]
+\begin{testimonia}[hp04_073]
 \emph{Yuktabhavadeva} 11.30 (attr. to Gorakṣanātha)
@@
 
-\begin{philcomm}[hp04_070]
+\begin{philcomm}[hp04_073]
 The six-chapter \emph{Vivekamārtaṇḍa} is attributed to Viśvarūpa, which perhaps explains the attribution of this verse to him.
@@
 %%%%%%%%%%
-\subsection*{4.71}
-\begin{translation}[hp04_071]
+\subsection*{4.74}
+\begin{translation}[hp04_074]
 When the mind is still the breath is still, from which semen becomes still.  As a result of semen becoming still, my son, the body becomes still.
@@
 
-%\begin{sources}[hp04_071]
-%\end{sources}
-
-\begin{testimonia}[hp04_071]
+%\begin{sources}[hp04_074]
+%\end{sources}
+
+\begin{testimonia}[hp04_074]
 \emph{Yogacintāmaṇi}~f.19v
@@
 
-%\begin{philcomm}[hp04_071]
+%\begin{philcomm}[hp04_074]
 The vocative \emph{putra} in \emph{pāda} c suggests that this verse is from a source text that we are yet to identify.
@@
 %%%%%%%%%%
-\subsection*{4.72}
-\begin{translation}[hp04_072]
+\subsection*{4.75}
+\begin{translation}[hp04_075]
 Only he whose gaze is steady without a visible object, whose breath is steady without effort [and] whose mind is steady without a support is a yogi, is a guru, is to be served.
@@
 
-\begin{sources}[hp04_072]
+\begin{sources}[hp04_075]
 \emph{Amanaska} 2.44
@@
 
-\begin{testimonia}[hp04_072]
+\begin{testimonia}[hp04_075]
 \emph{Haṭharatnāvalī} 4.25
@@
 
-%\begin{philcomm}[hp04_072]
-%\end{philcomm}
-
-\begin{metre}[hp04_072]
+%\begin{philcomm}[hp04_075]
+%\end{philcomm}
+
+\begin{metre}[hp04_075]
 Upajāti
@@
 %%%%%%%%%%
-\subsection*{4.73}
-\begin{translation}[hp04_073]
+\subsection*{4.76}
+\begin{translation}[hp04_076]
 [The yogi] whose breath does not move in, out, left, right, up or down is liberated. In this there is no doubt.
@@
 
-\begin{sources}[hp04_073]
+\begin{sources}[hp04_076]
 \emph{Gorakṣaśataka} 8
@@
 
-%\begin{testimonia}[hp04_073]
+%\begin{testimonia}[hp04_076]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_073]
-%\end{philcomm}
-
-\begin{metre}[hp04_073]
+%\begin{philcomm}[hp04_076]
+%\end{philcomm}
+
+\begin{metre}[hp04_076]
 Anuṣṭubh (c: bha-vipulā)
@@
 %%%%%%%%%%
-\subsection*{4.74}
-\begin{translation}[hp04_074]
+\subsection*{4.77}
+\begin{translation}[hp04_077]
 All the methods of Haṭha and Laya are for mastering Rājayoga. The man who has ascended to Rājayoga cheats death.
@@
 
-%\begin{sources}[hp04_074]
-%\end{sources}
-
-\begin{testimonia}[hp04_074]
+%\begin{sources}[hp04_077]
+%\end{sources}
+
+\begin{testimonia}[hp04_077]
 \emph{Yogacintāmaṇi} f. 8r (attrib.~to the \emph{Haṭhapradīpikā})
@@
 
-%\begin{philcomm}[hp04_074]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.74*1}
-\begin{translation}[hp04_074_1]
+%\begin{philcomm}[hp04_077]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.77*1}
+\begin{translation}[hp04_077_1]
 Iḍā is the divine Gaṅgā, Piṅgalā is the river Yamunā. Between those two is Suṣumṇā, who is to be recognised as Sarasvatī.
@@
 
-%\begin{sources}[hp04_074_1]
-%\end{sources}
-
-%\begin{testimonia}[hp04_074_1]
+%\begin{sources}[hp04_077_1]
+%\end{sources}
+
+%\begin{testimonia}[hp04_077_1]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_074_1]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.74*2}
-\begin{translation}[hp04_074_2]
+%\begin{philcomm}[hp04_077_1]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.77*2}
+\begin{translation}[hp04_077_2]
 The place of the Triveṇī confluence is called the king of sacred sites. One should bathe there [and] be freed from all sins.
@@
 
-%\begin{sources}[hp04_074_2]
-%\end{sources}
-
-%\begin{testimonia}[hp04_074_2]
+%\begin{sources}[hp04_077_2]
+%\end{sources}
+
+%\begin{testimonia}[hp04_077_2]
 %\end{testimonia}
 
-%\begin{philcomm}[hp04_074_2]
-%\end{philcomm}
-
-%%%%%%%%%%
-\subsection*{4.75}
-\begin{translation}[hp04_075]
+%\begin{philcomm}[hp04_077_2]
+%\end{philcomm}
+
+%%%%%%%%%%
+\subsection*{4.78}
+\begin{translation}[hp04_078]
 O great yogis, experience the nectar of Haṭha, the essence extracted from the ocean of all yoga scriptures after it has thus been churned, if you wish not to grow old and die.
@@
 
-%\begin{sources}[hp04_075]
-%\end{sources}
-
-\begin{testimonia}[hp04_075]
+%\begin{sources}[hp04_078]
+%\end{sources}
+
+\begin{testimonia}[hp04_078]
 \emph{Haṭhasaṅketacandrikā} f.~145v
@@
 
-\begin{philcomm}[hp04_075]
-\end{philcomm}
-
-\begin{metre}[hp04_075]
+\begin{philcomm}[hp04_078]
+\end{philcomm}
+
+\begin{metre}[hp04_078]
 Puṣpitāgrā
@@
 %%%%%%%%%%
-\subsection*{4.75*1}
+\subsection*{4.78*1}
 \begin{translation}[hp04_75_1] The wise people in the world wash away sin at the sacred site of knowledge (\emph{vidyātīrthe}), the virtuous at the sacred site of truth (\emph{satyatīrthe}), the impure-minded at the sacred site of the Gaṅgā (\emph{gaṅgātīrthe}), yogis at the sacred site of knowledge (\emph{jñānatīrthe}), kings at the sacred site of the streams (\emph{dhārātīrthe}), the rich at the sacred site of charity (\emph{dānatīrthe}) [and] women of good family at the sacred site of modesty.
@@
 
-\begin{metre}[hp04_075_1]
+\begin{metre}[hp04_078_1]
 Mandākrāntā 

### 2025-04-04 export-publish.sh

@@ -44,6 +44,6 @@ done
 sed -n '/<!--content-->/,$ {p}' html/meta.html >> html/hp-online-pub.html
 
 #postprocess straddle consonants in 4line meteres
-sed -i '/<p class="hpvers">.*-\s*<\/p>/{N}; x; s_-\(\s*</p>\s*<p class="hpvers">\s*\(<mark>\)\?\s*\)\([kgṅcjñṭḍṇtdnpbmyrlv]\) _\3\1_g' html/hp-online.html
+#sed -i '/<p class="hpvers">.*-\s*<\/p>/{N}; x; s_-\(\s*</p>\s*<p class="hpvers">\s*\(<mark>\)\?\s*\)\([kgṅcjñṭḍṇtdnpbmyrlv]h\) _\3\1_g' html/hp-online.html
 
 exit

### 2025-04-08 export-publish.sh
minor fixes

@@ -44,6 +44,6 @@ done
 sed -n '/<!--content-->/,$ {p}' html/meta.html >> html/hp-online-pub.html
 
 #postprocess straddle consonants in 4line meteres
-#sed -i '/<p class="hpvers">.*-\s*<\/p>/{N}; x; s_-\(\s*</p>\s*<p class="hpvers">\s*\(<mark>\)\?\s*\)\([kgṅcjñṭḍṇtdnpbmyrlv]h\) _\3\1_g' html/hp-online.html
+sed -i '/<p class="vers-line">.*-\s*<\/p>/{N}; x; s_-\(\s*</p>\s*<p class="vers-line">\s*\(<mark>\)\?\s*\)\([kgṅcjñṭḍṇtdnpbmyrlv]h\) _\3\1_g' html/hp-online-pub.html
 
 exit
